### 2020-05-15 free-theorems-nesteds/issues-with-fold.tex
writeup of fold examples / bush reverse investigation

% For double-blind review submission, w/o CCS and ACM Reference (max
% submission space)
\documentclass[acmsmall,review,anonymous]{acmart}
\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

%% Journal information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{POPL} % CONF = POPL or ICFP or OOPSLA
\acmArticle{1}
\acmYear{2020}
\acmMonth{1}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2018}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations
%\citestyle{acmnumeric}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Note: Authors migrating a paper from PACMPL format to traditional
%% SIGPLAN proceedings format must update the '\documentclass' and
%% topmatter commands above; see 'acmart-sigplanproc-template.tex'.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage[utf8]{inputenc}
\usepackage{ccicons}

\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amscd}
%\usepackage{MnSymbol}
\usepackage{xcolor}

\usepackage{bbold}
\usepackage{url}
\usepackage{upgreek}
%\usepackage{stmaryrd}

\usepackage{lipsum}
\usepackage{tikz-cd}
\usetikzlibrary{cd}
\usetikzlibrary{calc}
\usetikzlibrary{arrows}

\usepackage{bussproofs}
\EnableBpAbbreviations

\input{macros}
\input{mytheorem}

\theoremstyle{definition}
\newtheorem{exmpl}{Example}

\renewcommand{\greyout}[1]{} %{{\color{gray} {#1}}} -- toggle to remove greyed text

\newcommand{\emptyfun}{{[]}}
\newcommand{\cal}{\mathcal}
%\newcommand{\fold}{\mathit{fold}}
\newcommand{\F}{\mathcal{F}}
\renewcommand{\G}{\mathcal{G}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\B}{\mathcal{B}}
\renewcommand{\P}{\mathcal{A}}
\newcommand{\pred}{\mathsf{Fam}}
\newcommand{\env}{\mathsf{Env}}
\newcommand{\set}{\mathsf{Set}}
\renewcommand{\S}{\mathcal S}
\renewcommand{\C}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\A}{\mathcal{A}}
\renewcommand{\id}{\mathit{id}}
\newcommand{\map}{\mathsf{map}}
\newcommand{\pid}{\underline{\mathit{id}}}
\newcommand{\pcirc}{\,\underline{\circ}\,}
\newcommand{\pzero}{\underline{0}}
\newcommand{\pone}{\underline{1}}
\newcommand{\psum}{\,\underline{+}\,}
%\newcommand{\inl}{\mathit{inL}\,}
%\newcommand{\inr}{\mathit{inR}\,}
\newcommand{\pinl}{\underline{\mathit{inL}}\,}
\newcommand{\pinr}{\underline{\mathit{inR}}\,}
\newcommand{\ptimes}{\,\underline{\times}\,}
\newcommand{\ppi}{\underline{\pi_1}}
\newcommand{\pppi}{\underline{\pi_2}}
\newcommand{\pmu}{\underline{\mu}}

\title[Free Theorems for Nested Types]{Free Theorems for 
Nested Types} %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
%\subtitle{Subtitle}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

%% Author information
%% Contents and number of authors suppressed with 'anonymous'.
%% Each author should be introduced by \author, followed by
%% \authornote (optional), \orcid (optional), \affiliation, and
%% \email.
%% An author may have multiple affiliations and/or emails; repeat the
%% appropriate command.
%% Many elements are not rendered, but should be provided for metadata
%% extraction tools.

%% Author with single affiliation.
\author{Patricia Johann and Andrew Polonsky}
%\authornote{with author1 note}          %% \authornote is optional;
%                                        %% can be repeated if necessary
%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
%  \position{Position1}
%  \department{Department1}              %% \department is recommended
  \institution{Appalachian State University}            %% \institution is required
%  \streetaddress{Street1 Address1}
%  \city{City1}
%  \state{State1}
%  \postcode{Post-Code1}
%  \country{Country1}                    %% \country is recommended
}
\email{johannp@appstate.edu, andrew.polonsky@gmail.com}          %% \email is recommended

\begin{document}

\begin{abstract}
\end{abstract}

\maketitle
%% sstart 
\section{Limitations of folds and issues defining \texttt{bushreverse}}

We would like to define a function \texttt{bushreverse} which reverses a bush. 
Bushes are not linear, so it is not immediately clear what reversing a bush should even mean. 
First consider the flattening of bushes as a natural transformation from Bush to List (see bush2list). 
The flattening of a reversed bush should be the reversal (on lists) of the flattening of the original bush, 
i.e., \texttt{forall b. bush2list (bushreverse b) = reverse (bush2list b)} 
or \texttt{bush2list $\circ$ bushreverse = reverse $\circ$ bush2list}.
We would also like \texttt{bushreverse} to be an involution on bushes.

Clearly this function will have a recursive definition, so to define it in our calculus we will use a fold. 
Before defining \texttt{bushreverse} as a fold, let's consider how to define list reversal as a fold. There are several ways to do this. 
One method is to replace nil with nil and replace cons with a function which inserts an element at the back of a list.
This cons replacement function can be defined in several ways, e.g.,

  \texttt{$\lambda \; x \; xs \to$ xs ++ [ x ]}, or

  \texttt{$\lambda \; x \to$ foldr (::) [ x ] }.

The first version makes use of the append function on lists. 
The second version simply replaces the nil constructor with a singleton list. 
The first cannot be defined in our system because we cannot define an append function for bushes (see next section). 
The second cannot be defined because the fold for bushes requires a fully polymorphic algebra, 
so the only possible replacement for bnil is bnil in this context. Writing the insertion function with foldr is
much simpler than the case for bushes because there is only one nil constructor in a list, but a bush
can have many occurences of bnil.
We can also give a recursive definition without using append or foldr:
\begin{align*}
  &\texttt{insert : $\forall$ \{A : Set\} $\to$ A $\to$ List A $\to$ List A} \\
  &\texttt{insert x [] = [ x ] } \\
  &\texttt{insert x (y :: ys) = y :: (insert x xs) } 
\end{align*}

and we can define an analogous function for bushes:
\begin{align*}
  &\texttt{binsert : $\forall$ \{A : Set\} $\to$ A $\to$ Bush (Bush A) $\to$ Bush A} \\
  &\texttt{binsert x bnil = bcons x bnil } \\
  &\texttt{binsert x (bcons bnil bbbx) = bcons x (binsert bnil bbbx)} \\
  &\texttt{binsert x (bcons (bcons y bx) bbbx) = bcons y (binsert (binsert x bx) bbbx)}
\end{align*}
and feed this as an argument to the fold for bushes to get a reversal function: 
\begin{align*}
  &\texttt{bushreverse : $\forall$ \{A : Set\} $\to$ Bush A $\to$ Bush A} \\
  &\texttt{bushreverse \{A\} = bfold \{A\} \{Bush\} bnil binsert}
\end{align*}

Note that the first explicit argument to bfold is completely determined by the types. 
The argument must have type \texttt{($\forall$ \{A : Set\} $\to$ Bush A)} and the only inhabitant of this type is bnil.
Folds for (non first-order) nested types will require arguments of the form \texttt{($\forall$ \{A : Set\} $\to$ F A)} whenever 
the type has constructors with no arguments, and due to the universal quantification, the only inhabitants of this type 
will be these constructors themselves (e.g. pnil, bnil) with some other possible combinations 
if there are constructors of type \texttt{($\forall$ \{A : Set\} $\to$ F A $\to$ F A $\to$ F A)}, for example.

This restriction means that the behavior of our reversal function must be specified entirely by binsert. 
The question is, can we define binsert without arbitrary term level recursion? 
If we try to define it as a fold over the \texttt{Bush (Bush A)} parameter, then we have no 
way to include x in that (fully polymorphic) fold. We could even try to define it as a fold over a 
bush built up from x such as \texttt{(bcons x bnil)}, but the same issue arises.

The fundamental issue here is that the binsert function takes two arguments, 
and there is no way to combine them together with a single fold. We can fold
over each argument separately, but then we still have results to combine.

\subsection{What can we define with folds over nested types?}
These issues are not specific to the Bush datatype; they are inherent limitations of folds
over nested types. In fact we are quite restricted in the types of functions we can define recursively. 
To define a function using a fold, its domain must be a fixed point, therefore
types such as \texttt{($\forall$ \{A : Set\} $\to$ (Bush A $\times$ Bush A) $\to$ Bush A)} cannot be defined as
a fold over the pair. Of course, we can fold over each bush in the pair individually, but we cannot interleave
the computations in any way because the polymorphism required by the folds prohibits this.

Additionally, the
return type of the result of the fold must be a functor. So even if we curry the previous type to get
\texttt{($\forall$ \{A : Set\} $\to$ Bush A $\to$ (Bush A $\to$ Bush A))}, we can not define it as a fold
over the first bush because \texttt{(Bush A $\to$ Bush A)} is not a functor.
These two restrictions indicate that we cannot define many interesting functions which have
multiple nested types as arguments.

The restriction stretches even further though, because sometimes the algebras supplied to a fold
must have a recursive definition involving nested types. This is the case with \texttt{bushreverse}
because even though it has a suitable type, \texttt{($\forall$ \{A : Set\} $\to$ Bush A $\to$ Bush A)},
the binsert function has a type which makes it incompatible with fold.

\subsection{Generalized Folds}
We can make our calculus slightly more expressive using generalized folds \cite{?}.
However, these generalized folds are mostly helpful for defining \textit{reductions}, 
functions which take data of a monomorphic type as input and have a monomorphic return type. 
Specifically, generalized folds relax the constraint on the domain of the function being defined as a fold.
Instead of requiring the domain to be a fixed point, the domain can be a fixed point composed with another functor. 
However, this does not help us in defining functions which have multiple nested types as arguments.

\end{document}

### 2020-06-21 free-theorems-nesteds/paper3.tex
paper3 adds new rule to restrict mu-types so their bodies don't contain free functorial variables

% double-blind review submission, w/o CCS and ACM Reference (max
% submission space)
\documentclass[acmsmall,review,anonymous]{acmart}
\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

%% Journal information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{POPL} % CONF = POPL or ICFP or OOPSLA
\acmArticle{1}
\acmYear{2020}
\acmMonth{1}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2018}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations
%\citestyle{acmnumeric}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Note: Authors migrating a paper from PACMPL format to traditional
%% SIGPLAN proceedings format must update the '\documentclass' and
%% topmatter commands above; see 'acmart-sigplanproc-template.tex'.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage[utf8]{inputenc}
\usepackage{ccicons}
\usepackage{verbatim}

\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amscd}
%\usepackage{MnSymbol}
\usepackage{xcolor}

\usepackage{bbold}
\usepackage{url}
\usepackage{upgreek}
%\usepackage{stmaryrd}

\usepackage{lipsum}
\usepackage{tikz-cd}
\usetikzlibrary{cd}
\usetikzlibrary{calc}
\usetikzlibrary{arrows}

\usepackage{bussproofs}
\EnableBpAbbreviations

\input{macros}
\input{mytheorem}

\theoremstyle{definition}
\newtheorem{exmpl}{Example}

\renewcommand{\greyout}[1]{} %{{\color{gray} {#1}}} -- toggle to remove greyed text

\newcommand{\emptyfun}{{[]}}
\newcommand{\cal}{\mathcal}
%\newcommand{\fold}{\mathit{fold}}
\newcommand{\F}{\mathcal{F}}
\renewcommand{\G}{\mathcal{G}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\B}{\mathcal{B}}
\renewcommand{\P}{\mathcal{A}}
\newcommand{\pred}{\mathsf{Fam}}
\newcommand{\env}{\mathsf{Env}}
\newcommand{\set}{\mathsf{Set}}
\renewcommand{\S}{\mathcal S}
\renewcommand{\C}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\A}{\mathcal{A}}
\renewcommand{\id}{\mathit{id}}
\newcommand{\map}{\mathsf{map}}
\newcommand{\pid}{\underline{\mathit{id}}}
\newcommand{\pcirc}{\,\underline{\circ}\,}
\newcommand{\pzero}{\underline{0}}
\newcommand{\pone}{\underline{1}}
\newcommand{\psum}{\,\underline{+}\,}
\newcommand{\pinl}{\underline{\mathit{inL}}\,}
\newcommand{\pinr}{\underline{\mathit{inR}}\,}
\newcommand{\ptimes}{\,\underline{\times}\,}
\newcommand{\ppi}{\underline{\pi_1}}
\newcommand{\pppi}{\underline{\pi_2}}
\newcommand{\pmu}{\underline{\mu}}
\newcommand{\semmap}{\mathit{map}}
\newcommand{\subst}{\mathit{subst}}

\newcommand{\tb}[1]{~~ \mbox{#1} ~~}
\newcommand{\listt}[1]{(\mu \phi. \lambda \beta . \onet + \beta \times
  \phi \beta) #1} 
\newcommand{\filtype}{\Nat^\emptyset 
 (\Nat^\emptyset \, \alpha \, \mathit{Bool})\, (\Nat^\emptyset 
  (List \, \alpha) \, (List \, \alpha))} 
\newcommand{\filtypeGRose}{\Nat^\emptyset 
 (\Nat^\emptyset \, \alpha \, \mathit{Bool})\, (\Nat^\emptyset 
  (\mathit{GRose}\,\psi \, \alpha) \, (\mathit{GRose}\,\psi \, (\alpha
  + \onet)))} 
\newcommand{\maplist}{\mathit{map}_{\lambda A. \setsem{\emptyset; \alpha
      \vdash \mathit{List} \, \alpha} \rho[\alpha := A]}} 
\newcommand{\PLeaves}{\mathsf{PLeaves}}
\newcommand{\swap}{\mathsf{swap}}
\newcommand{\reverse}{\mathsf{reverse}}
\newcommand{\Bcons}{\mathit{Bcons}}
\newcommand{\Bnil}{\mathit{Bnil}}

\title[Free Theorems for Nested Types]{Free Theorems for
  %Primitive
  Nested Types} %% [Short Title] is optional; when present,
                         %% will be used in header instead of Full
                         %% Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
%\subtitle{Subtitle}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

%% Author information
%% Contents and number of authors suppressed with 'anonymous'.
%% Each author should be introduced by \author, followed by
%% \authornote (optional), \orcid (optional), \affiliation, and
%% \email.
%% An author may have multiple affiliations and/or emails; repeat the
%% appropriate command.
%% Many elements are not rendered, but should be provided for metadata
%% extraction tools.

%% Author with single affiliation.
\author{Patricia Johann, Enrico Ghiorzi, and Daniel Jeffries}
%\authornote{with author1 note}          %% \authornote is optional;
%                                        %% can be repeated if necessary
%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
%  \position{Position1}
%  \department{Department1}              %% \department is recommended
  \institution{Appalachian State University}            %% \institution is required
%  \streetaddress{Street1 Address1}
%  \city{City1}
%  \state{State1}
%  \postcode{Post-Code1}
%  \country{Country1}                    %% \country is recommended
}
\email{johannp@appstate.edu, ghiorzie@appstate.edu, jeffriesd@appstate.edu}          %% \email is recommended

\begin{document}

\begin{abstract}
This paper considers parametricity and its consequent free theorems
for nested data types.  Rather than representing nested types via
their Church encodings into a higher-kinded extension of System F, we
adopt a functional programming perspective and design a polymorphic
calculus that provides primitives for constructing a robust class of
nested types, including ADTs and all (truly) nested types appearing in
the literature, directly as fixpoints \`{a} la Pitts' PolyPCF.  At the
term level, our calculus supports the usual pattern matching, map
functions, and stylized recursion via folds for all such types. Using
a variation on Atkey's 2012 construction
%of a parametric model
for $F_\omega$,
%that explicitly constructs equality operators rather than merely
%postulating them,
we construct a parametric model for our calculus. However, to ensure
the existence of the fixpoints interpreting nested types, cocontinuity
conditions on the functors underlying these fixpoints must be threaded
throughout our model construction. This turns out to be both delicate
and challenging. Overall, this paper offers a blueprint for
deriving principled parametricity results 
%constructing parametric models
for practical programming languages,
such as Haskell and Agda, that directly construct nested types as
fixpoints rather than encoding them. It also shows how parametric
models can be used to derive useful free theorems for programs over
nested types. Like their counterparts for ADTs, the free theorems our
model derives provide, in particular, a principled alternative to
proving properties of programs over nested types using, e.g., the deep
induction rules for nested types recently developed by Johann and
Polonsky. This is illustrated with several examples. 
\end{abstract}

%\begin{CCSXML}
%<ccs2012>
%<concept>
%<concept_id>10011007.10011006.10011008</concept_id>
%<concept_desc>Software and its engineering~General programming languages</concept_desc>
%<concept_significance>500</concept_significance>
%</concept>
%<concept>
%<concept_id>10003456.10003457.10003521.10003525</concept_id>
%<concept_desc>Social and professional topics~History of programming languages</concept_desc>
%<concept_significance>300</concept_significance>
%</concept>
%</ccs2012>
%\end{CCSXML}
%
%\ccsdesc[500]{Software and its engineering~General programming languages}
%\ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code

%% Keywords
%% comma separated list
%\keywords{keyword1, keyword2, keyword3}  %% \keywords is optional

\maketitle

\section{Introduction}\label{sec:intro}

Suppose we want to prove some property of programs over an algebraic
data type (ADT) such as that of lists, coded in Agda
%\footnote{All code appearing in this paper are written in Agda.}
as

\vspace*{-0.15in}

{\small
\[\begin{array}{l}
\mathtt{data\; List \;(A : Set)\;:\;Set\;where}\\
\hspace*{0.4in}\mathtt{nil\;:\; List\;A}\\
\hspace*{0.4in}\mathtt{Cons\;:\;A \rightarrow List\;A \rightarrow List\;A}
\end{array}\]}
A natural approach to the problem uses structural induction on the
input data structure in question. This requires knowing not just the
definition of the ADT of which the input data structure is an
instance, but also the program text for the functions involved in the
properties to be proved. For example, to prove by induction that
mapping a polymorphic function over a list and then reversing the
resulting list is the same as reversing the original list and then
mapping the function over the result, we unwind the (recursive)
definitions of the $\mathtt{reverse}$ and $\mathtt{map}$ functions
over lists to according to the inductive structure of the input
list. Such data-driven induction proofs over ADTs are so routine that
they are often included in, say, undergraduate functional programming
courses.

An alternative technique for proving results like the above
$\mathtt{map}$-$\mathtt{reverse}$ property for lists is to use
parametricity, a formalization of extensional type-uniformity in
polymorphic languages. Parametricity captures the intuition that a
polymorphic program must act uniformly on all of its possible type
instantiations; it is formalized as the requirement that every
polymorphic program preserves all relations between any pair of types
that it is instantiated with.  Parametricity was originally put forth
by Reynolds~\cite{rey83} for System F~\cite{gr89}, the formal calculus
at the core of all polymorphic functional languages. It was later
popularized as Wadler's ``theorems for free''~\cite{wad89} because it
allows the deduction of many properties of programs in such languages
solely from their types, i.e., with no knowledge whatsoever of the
text of the programs involved. To get interesting free theorems,
Wadler's calculus included, implicitly, built-in list types; indeed,
most of the free theorems in~\cite{wad89} are consequences of
naturality for polymorphic list-processing functions. However,
parametricity can also be used to prove naturality properties for
non-list ADTs, as well as results, like correctness of program
optimizations like {\em short cut fusion}~\cite{glp93,joh02,joh03},
that go beyond simple naturality.
%In addition, parametricity in the presence of ADTs
%{\color{blue} check!}  has further been used to prove i) {\em
%  properties of programs}, e.g., that they perform no illegal
%operations, satisfy certain security criteria~\cite{ep03,rp10}, or are
%observationally equivalent to their compiled forms~\cite{bh09,hd11};
%ii) {\em whole-language properties}, e.g., that they support data
%abstraction and modularity via representation independence,
%\cite{bm05,dr04,jac99,kat11,mr92}, enforce information flow
%policies~\cite{ss01,ts04}, or guarantee privacy~\cite{rp10}; and iii)
%{\em properties of implementations}, e.g., compiler
%correctness~\cite{ab08,bh09,hd11,glpj93}. Parametricity is thus an
%important tool for reasoning about programs in polymorphic languages
%that natively support ADTs. {\color{blue} Be clear about what this
%  means. Re-word?}

This paper is about parametricity and free theorems for a polymorphic
calculus with explicit syntax not just for ADTs, but for nested types
as well. An ADT defines a {\em family of inductive data types}, one
for each input type. For example, the $\mathtt{List}$ data type
definition above defines a collection of data types $\mathtt{List\;
  A}$, $\mathtt{List\; B}$, $\mathtt{List\; (A \times B)}$,
$\mathtt{List\; (\List\;A)}$, etc., each independent of all the
others. By contrast, a nested type~\cite{bm98} is an {\em inductive
  family of data types} that is defined over, or is defined mutually
recursively with, (other) such data types. Since the structures of the
data type at one type can depend on those at other types, the entire
family of types must be defined at once. Examples of nested types
include, trivially, ordinary ADTs, such as list and tree types; simple
nested types, such as the data type {\small
\[\begin{array}{l}
\mathtt{data\; PTree\;(A : Set)\;:\;Set\;where}\\
\hspace*{0.4in}\mathtt{pleaf\;:\;A \rightarrow PTree\;A}\\
\hspace*{0.4in}\mathtt{pnode\;:\;PTree\;(A \times A) \rightarrow PTree\;A}
\end{array}\]}
\hspace{-0.04in}of perfect trees, whose recursive occurrences never
appear below other type constructors; ``deep'' nested
types~\cite{jp20}, such as the data type {\small
\[\begin{array}{l}
\mathtt{data\; Forest\;(A : Set)\;:\;Set\;where}\\
\hspace*{0.4in}\mathtt{fempty\;:\;Forest\;A}\\
\hspace*{0.4in}\mathtt{fnode\;:\; A \rightarrow PTree\;(Forest\;A) \to
Forest\;A}
\end{array}\]}
\hspace{-0.04in}of perfect forests, whose recursive occurrences appear
below type constructors for other nested types; and truly nested
types\footnote{Nested types that are defined over themselves are known
  as {\em truly nested types}.}, such as the data type {\small
\[\begin{array}{l}
\mathtt{data\; Bush\;(A : Set)\;:\;Set\;where}\\
\hspace*{0.4in}\mathtt{bnil\;:\; Bush\;A}\\
\hspace*{0.4in}\mathtt{bcons\;:\;A \rightarrow Bush\;(Bush \; A)
  \rightarrow Bush\;A} 
\end{array}\]}
\hspace{-0.04in}of bushes (also called {\em bootstrapped heaps}
in~\cite{oka99}), whose recursive occurrences appear below their own
type constructors.

Suppose we now want to prove properties of functions over nested
types. We might, for example, want to prove a
$\mathtt{map}$-$\mathtt{reverse}$ property for the functions on
perfect trees in Figure~\ref{fig:funs}, or for those on
bushes\footnote{To define the $\mathtt{foldBush}$ and
  $\mathtt{mapBush}$ functions in Figure~\ref{fig:funs2} it is
  necessary to turn off Agda's termination checker.} in
Figure~\ref{fig:funs2}. A few well-chosen examples quickly convince us
that such a property should indeed hold for perfect trees, and,
drawing inspiration from the situation for ADTs, we easily construct a
proof by induction on the input perfect tree. To formally establish
this result, we could even prove it in Coq or Agda: each of these
provers actually generates an induction rule for perfect trees, and
the rule it generats gives the expected result because proving
properties of perfect trees requires only that we induct over the
top-level perfect tree in the recursive position, leaving any data
internal to the input tree untouched.

%want to prove that $\mathtt{reversePTree}$ commutes with
%$\mathtt{mapPTree}$ or that $\mathtt{reverseBush}$ commutes with
%$\mathtt{mapBush}$. It is not hard to convince oneself of the former:
%a few well-chosen examples show clearly why mapping a polymorphic
%%function over a perfect tree and then reversing the data at the leaves
%of the resulting perfect tree is indeed the same result as mapping
%that same function over the result of reversing the data at the leaves
%of the original perfect tree.

Unfortunately, it is nowhere near as clear that analogous intuitive or
formal inductive arguments can be made for the
$\mathtt{map}$-$\mathtt{reverse}$ property for bushes. Indeed, a proof
by induction on the input bush must recursively induct over the bushes
that are internal to the top-level bush in the recursive position.
This is sufficiently delicate that no induction rule for bushes or
other truly nested types was known until very recently, when {\em deep
  induction}~\cite{jp20} was developed as a way to induct over {\em
  all} of the structured data present in an input. Deep induction thus
not only gave the first principled and practically useful structural
induction rules for bushes and other truly nested types, and has also
opened the way for incorporating automatic generation of such rules
for (truly) nested data types --- and, eventually, even GADTs --- into
modern proof assistants.
\begin{figure*}
\hspace*{-0.5in}
\resizebox{0.35\linewidth}{!}{
\begin{minipage}[t]{0.5\textwidth}
\[\begin{array}{l}
\mathtt{reversePTree : \forall \{A : Set\}  \rightarrow PTree\; A
  \rightarrow PTree \;A}\\  
\mathtt{reversePTree \; \{A\} = foldPTree \; \{A\}\; \{PTree\}}\\
\hspace*{1.3in} \mathtt{pleaf}\\
\hspace*{1.3in} \mathtt{( \lambda p \rightarrow pnode\; (mapPTree\;
  swap\; p) )}\\  
\\
\mathtt{foldPTree : \forall \{A : Set\} \rightarrow \{F : Set \rightarrow Set \}
  \rightarrow}\\
\hspace*{0.8in} \mathtt{( \{B : Set\} \rightarrow B \rightarrow F B) \rightarrow}\\
\hspace*{0.8in} \mathtt{(\{B : Set\} \rightarrow F (B \times B)
  \rightarrow F B) \rightarrow}\\
\hspace*{0.8in}\mathtt{PTree \;A \rightarrow F\;A}\\
\mathtt{foldPTree\; n\; c\; (pleaf\; x) = n \;x}\\
\mathtt{foldPTree\; n\; c \;(pnode\; p) = c\; (foldPTree \;n \;c \;p)}\\
\\
\mathtt{mapPTree : \forall \{A\, B : Set\} \rightarrow (A \rightarrow B)
  \rightarrow PLeaves\; A \rightarrow PLeaves\; B}\\ 
\mathtt{mapPTree \;f\; (pleaf \;x) = pleaf\; (f\; x)}\\
\mathtt{mapPTree\; f \;(pnode\; p) = pnode\; (mapPTree \;(\lambda p
  \rightarrow (f (\pi_1 \,p), f (\pi_2\,p)))\; p)}\\ 
\\
\mathtt{swap : \forall \{A : Set\} \rightarrow (A \times A)
  \rightarrow (A \times A)}\\ 
\mathtt{swap \;(x,y) = (y,x)}
\end{array}\]
\caption{$\mathtt{reversePTree}$ and auxiliary functions in Agda}\label{fig:funs} 
\end{minipage}}
\quad\quad\quad\quad\quad\quad\quad\quad
\resizebox{0.35\linewidth}{!}{
\begin{minipage}[t]{0.5\textwidth}
\[\begin{array}{l}
  \mathtt{reverseBush : \forall \{A : Set\} \rightarrow Bush\; A
  \rightarrow Bush\; A}\\ 
\mathtt{reverseBush \;\{A\} = foldBush\; \{A\}\; \{Bush\}\; bnil \;balg}\\
\\
\mathtt{foldBush : \forall \{A : Set\} \rightarrow \{F : Set \rightarrow
  Set\} \rightarrow}\\
\hspace*{0.5in} \mathtt{(\{B : Set\} \rightarrow F B) \rightarrow}\\
\hspace*{0.5in} \mathtt{(\{B : Set\} \rightarrow B \rightarrow F\;
  (F\; B) \rightarrow F\; B) \rightarrow}\\ 
\hspace*{0.5in} \mathtt{Bush\; A \rightarrow F \;A}\\
\mathtt{foldBush\; bn\; bc\; bnil = bn}\\
\mathtt{foldBush \;bn \;bc \;(bcons\; x \;bb) =}\\
\hspace*{0.5in} \mathtt{bc\; x \;(foldBush \;bn\; bc
  \;(mapBush \;(foldBush\; bn\; bc) \;bb))}\\
\\
\mathtt{mapBush : \forall \{A \,B : Set\} \to (A \to B) \to (Bush\, A) \to
  (Bush\, B)}\\ 
\mathtt{mapBush\; \_ \; bnil = bnil}\\
\mathtt{mapBush\; f\; (bcons\; x \;bb) = bcons \;(f\, x) \; (mapBush\;
  (mapBush\, f) \; bb)}\\ 
\\
\mathtt{balg : \forall \{B : Set\} \rightarrow B \rightarrow Bush
  \;(Bush\; B) \rightarrow Bush\; B}\\
\mathtt{balg \;x \;bnil = bcons\; x \;bnil}\\
\mathtt{balg \;x\; (bcons \;bnil \;bbbx) = bcons\; x \;(bcons \;bnil\; bbbx)}\\
\mathtt{balg\; x \;(bcons \;(bcons\; y \;bx)\; bbbx) =}\\
\hspace*{0.5in} \mathtt{bcons\; y\; (bcons\; (bcons \;x \;bx) \;bbbx)}\\
\end{array}\]
\caption{$\mathtt{reverseBush}$ and auxiliary
  functions in Agda}\label{fig:funs2} 
\end{minipage}}
\end{figure*}

\vspace*{-0.2in}

Of course it is great to know that we {\em can}, at last, prove
properties of programs over (truly) nested types by induction.  But
recalling that inductive proofs over ADTs can sometimes be
circumvented in the presence of parametricity, we might naturally ask:
\begin{verse}
{\em Can we derive properties of functions over (truly) nested types
  from parametricity?}
\end{verse}
This paper answers the above question in the affirmative by
constructing a parametric model for a polymorphic calculus providing
primitives for {\em constructing} nested types
%(including ADTs)
directly via
%type-level
recursion --- rather than representing them indirectly by Church
encodings as in most polymorphic calculi.

We introduce our calculus in Section~\ref{sec:calculus}.  At the type
level, it is the level-2-truncation of the higher-kinded calculus
from~\cite{jp19}, augmented with a primitive type of natural
transformations. To construct nested types, it constructs type
expressions not just from standard type variables, but also from type
constructor variables of various arities, and includes an explicit
$\mu$-construct for type-level recursion with respect to these
variables. The class of nested types thus constructed is very robust
and includes all (truly) nested types known from the literature. In
Section~\ref{sec:type-interp} we give set and relational
interpretations for the types From Section~\ref{sec:calculus}. As is
usual when modeling parametricity, types are interpreted as functors
from environments interpreting their type variable contexts to set or
relations, as appropriate. But in order to ensure that these functors
satisfy the cocontinuity properties needed to know that the fixpoints
interpreting $\mu$-types exist, set environments must map each $k$-ary
type constructor variable to an appropriately cocontinuous $k$-ary
functor on sets and relation environments must map each $k$-ary type
constructor variable to an appropriately cocontinuous $k$-ary relation
transformer, and these cocontinuity conditions must be threaded
throughout the type interpretations in such a way that the resulting
model is guaranteed to satisfy an appropriate Identity Extension Lemma
(Theorem~\ref{thm:iel}). Properly progagating the cocontinuity
conditions turns out to be both subtle and challenging, and
Section~\ref{sec:iel}, where it is done, is where the bulk of the work
in constructing our model lies.  At the term level, our calculus
includes primitive constructs for the actions on morphisms of the
functors interpreting types, initial algebras for fixpoints of these
functors, and structured recursion over elements of these initial
algebras (i.e., $\map$, $\tin$, and $\fold$ constructs,
respectively). While our calculus does not support general recursion
at the term level, it is strongly normalizing, so does perhaps edge us
toward the kind of provably total practical programming language
proposed at the end of~\cite{wad89}. In Section~\ref{sec:term-interp},
we give set and relational interpretations for the terms of our
calculus. As usual in parametric models, terms are interpreted as
natural transformations from interpretations of the term contexts in
which they are formed to the interpretations of their types, and these
must cohere in what is essentially a fibrational way. Immediately from
the definitions of our interpretations we prove in
Section~\ref{sec:Nat-type-terms} a scheme deriving free theorems that
are consequences of naturality of functions that are polymorphic over
nested types. This scheme is very general, is parameterized over both
the data type and the polymorphic function at hand, and has each of
the above $\mathtt{map}$-$\mathtt{reverse}$ theorems as instances. The
relationship between naturality and parametricity has long been of
interest, and our inclusion of a primitive type of natural
transformations allows us to clearly delineate those free theorems
that are consequences of naturality, and thus hold even in
non-parametric models, from those that require the full power of
parametricity. In Section~\ref{sec:thms} we prove that our model
satisfies an Abstraction Theorem (Theorem~\ref{thm:abstraction}), and
we derive several of this latter kind of free theorem from it in
Section~\ref{sec:ftnt}. Specifically, we state and prove
(non-)inhabitation results in Sections~\ref{sec:bottom}
and~\ref{sec:identity}, a free theorem for the type of a filter
function on generalized rose trees in Section~\ref{sec:ft-adt}, the
correctness of short cut fusion for lists in
Section~\ref{sec:short-cut}, and its generalization to nested types in
Section~\ref{sec:short-cut-nested}.

There is a long line of work on categorical models of parametricity
for System F; see,
e.g.,~\cite{bfss90,bm05,dr04,gjfor15,has94,jac99,mr92,rr94}.  To our
knowledge, all categorical models that treat (algebraic) data types do
so via their Church encodings, verifying in the
%Much of this work does ultimately extend parametricity to data types
%that are modeled as fixpoints of first-order functors, but to our
%knowledge this has never been done by building the data types directly
%into the calculus whose parametricity is to be modeled
%categorically. Indeed, all extensions we know of use the categorical
%equivalent of Atkey's approach. That is, they verify in the
just-constructed parametric model that
%the Church encoding constructed
%from a type constructor is interpreted as the least fixpoint of the
%functor interpreting that type constructor.
the Church encoding of each such type is interpreted as the least
fixpoint of the (first-order) functor interpreting the type
constructor from which its Church encoding was constructed.
%This is the categorical analogue of the approach in~\cite{atk12}
%described above.
The present paper draws on this rich tradition of categorical models
of parametricity for System F, but treats nested types as well as
ADTs, and is the first to treat these types by direct construction via
primitives rather than by Church encodings. This requires that we
modify the type calculus to ensure that
%System F's full impredicative polymorphism interacts reasonably with
%functoriality,
functoriality is guaranteed syntactically, and that functoriality is
reflected in the standard model construction so that the existence of
the fixpoints by which nested types are to be interpreted is ensured.
%parametricity to primitive nested types {\em as the model is being
%constructed} by taking care to ensure that System F's full
%impredicative polymorphism does not interfere with the functoriality
%needed to support primitive nested types.

Like our calculus, Pitts' PolyPCF~\cite{pit98,pit00}
%extends parametricity from pure System F to System F augmented with
%to a calculus that
includes primitives for constructing data types directly. Only list
types are added in~\cite{pit00}, but other polynomial ADTs are easily
included as in~\cite{pit98}. Part of Pitts' motivation is to show that
ADTs and their Church encodings have the same operational behavior in
his calculus. We cannot even ask this question about our calculus,
which cannot express Church encodings of even simple ADTs such as
list, or pair, or sum types, since these are not functorial;
nevertheless, we do prove the correctness of short cut fusion for
nested types, whose operational analogue is valid in the parametric
model Pitts constructs and is the means by which he proves his
equivalence result. It would be interesting to know what operational
analogues of functoriality and cocontinuity are needed to extend
Pitts' parametricity results from a calculus with primitives for
constructing data types modeled as fixpoints of first-order functors
to one that also provides such primitives for data types modeled as
fixpoints of higher-order functors.

We are not the first to consider parametricity at higher
kinds. Atkey~\cite{atk12} constructs a parametric model for full
System $F_\omega$, but within the impredicative Calculus of Inductive
Constructions (iCIC) rather than in a semantic category. His
construction is in some ways similar to ours, but it represents data
types using Church encodings rather than constructing them via
primitives. Since his model is entirely syntactic, his syntactic
``functors'', whose associated $\mathit{fmap}$ functions representing
their functorial actions must be {\em given} together with their
underlying type constructors, cannot be interpreted as semantic
functors. The associated Church encodings therefore cannot be
interpreted as their fixpoints, so Atkey need not, and does not,
impose cocontinuity conditions on his model to ensure that these
fixpoints exist.  Nevertheless, Atkey does verify the existence of
initial algebras for ``functors'' in iCIC. He does not indicate which
type constructors support the kinds of $\mathit{fmap}$ functions
needed to ensure they are ``functors'', but we suspect spelling this
out explicitly would result in a full higher-kinded extension of the
calculus presented here.

\section{The Calculus}\label{sec:calculus}

\subsection{Types}
For each $k \ge 0$, we assume countable sets $\tvars^k$ of \emph{type
  constructor variables of arity $k$} and $\fvars^k$ of
\emph{functorial variables of arity $k$}, all mutually disjoint.
%disjoint for distinct $k$ and disjoint from each other.
The sets of all type constructor variables and functorial variables
are $\tvars = \bigcup_{k \ge 0} \tvars^k$ and $\fvars = \bigcup_{k \ge
  0} \fvars^k$, respectively, and a \emph{type variable} is any
element of $\tvars \cup \fvars$.  We use lower case Greek letters for
type variables, writing $\phi^k$ to indicate that $\phi \in \tvars^k
\cup \fvars^k$, and omitting the arity indicator $k$ when convenient,
unimportant, or clear from context. We reserve letters from the
beginning of the alphabet to denote type variables of arity $0$, i.e.,
elements of $\tvars^0 \cup \fvars^0$. We write $\overline{\zeta}$ for
either a set $\{\zeta_1,...,\zeta_n\}$ of type constructor variables
or a set of functorial variables when the cardinality $n$ of the set
is unimportant or clear from context. If $\Pos\,$ is a set of type
variables we write $\Pos, \overline{\phi}$ for $\Pos\, \cup
\overline{\phi}$ when $\Pos\, \cap \overline{\phi} = \emptyset$.  We
omit the vector notation for a singleton set, thus writing $\phi$,
instead of $\overline{\phi}$, for $\{\phi\}$.
\begin{dfn}
Let $V$ be a finite subset of\, $\tvars$, $\Pos$ be a finite
subset of\, $\fvars$, $\overline{\alpha}$ be a finite subset of\,
$\fvars^0$ disjoint from $\Pos$, and $\phi^k \in \fvars^k
\setminus \Pos$.  The set $\mcF^\Pos(V)$ of {\em functorial
  expressions} over $\Pos$ and $V$ are given by
\begin{align*}
  \mcF^\Pos(V) \; ::= \; %\T(V)
  &\mid
\zerot \mid \onet 
\mid \Nat^{\ol{\alpha}} \, \mcF^{\overline{\alpha}}(V) \;
\mcF^{P,\ol{\alpha}}(V) %\mid V \ol{\T(V)}
\mid \Pos\; \ol{\mcF^\Pos(V)}  \,
\mid V\, \ol{\mcF^\Pos(V)}  
\mid \mcF^{\Pos}(V) + \mcF^\Pos(V)\\
&\mid \mcF^{\Pos}(V) \times \mcF^\Pos(V) \mid \left(\mu
\phi^{~k}. \lambda \alpha_1...\alpha_k.  
\mcF^{\Pos,\alpha_1,...,\alpha_k,\phi}(V)\right)
  \ol{\mcF^{\Pos}(V)}
\end{align*}
\end{dfn}
\noindent
A \emph{type} over $\Pos$ and $V$ is any element of % $\T(V) \cup
$\F^\Pos(V)$.

The notation for types entails that an application $F F_1...F_k$ is
allowed only when $F$ is a type variable of arity $k$, or $F$ is a
subexpression of the form $\mu \phi^{k}.\lambda
\alpha_1...\alpha_k.F'$. Moreover, if $F$ has arity $k$ then $F$ must
be applied to exactly $k$ arguments.  Accordingly, an overbar
indicates a sequence of subexpressions whose length matches the arity
of the type applied to it.  The fact that types are always in
\emph{$\eta$-long normal form} avoids having to consider
$\beta$-conversion at the level of types. In a subexpression
$\Nat^{\ol{\alpha}}F\,G$, the $\Nat$ operator binds all occurrences of
the variables in $\ol{\alpha}$ in $F$ and $G$. Similarly, in a
subexpression $\mu \phi^k.\lambda \ol{\alpha}.F$, the $\mu$ operator
binds all occurrences of the variable $\phi$, and the $\lambda$
operator binds all occurrences of the variables in $\ol{\alpha}$, in
the body $F$.

A {\em type constructor context} is a finite set $\Gamma$ of type
constructor variables, and a {\em functorial context} is a finite set
$\Phi$ of functorial variables. In Definition~\ref{def:wftypes}, a
judgment of the form $\Gamma;\Phi \vdash F$ indicates that the
type $F$ is intended to be functorial in the variables in $\Phi$
but not necessarily in those in $\Gamma$.
\begin{dfn}\label{def:wftypes}
The formation rules for the set $\F \subseteq \bigcup_{V \subseteq
  \tvars, \Pos\, \subseteq \fvars}\F^\Pos(V)$ of\, {\em well-formed
  type expressions} are

\vspace*{-0.2in}

\[\begin{array}{cc}
\AXC{\phantom{$\Gamma,\Phi$}}
\UIC{$\Gamma;\Phi \vdash \zerot$}
\DisplayProof
&
\AXC{\phantom{$\Gamma,\Phi$}}
\UIC{$\Gamma;\Phi \vdash \onet$}
\DisplayProof
\end{array}\]
\[\begin{array}{c}
\AXC{$\Gamma;\ol{\alpha^0} \vdash F$}
\AXC{$\Gamma;\Phi,\ol{\alpha^0}  \vdash G$}
\BIC{$\Gamma;\Phi \vdash \Nat^{\ol{\alpha^0}}F \,G$}
\DisplayProof
\\[3ex]
\AXC{$\phi^k \in \Gamma \cup \Phi$}
\AXC{$\quad\quad\ol{\Gamma;\Phi \vdash F}$}
\BIC{$\Gamma;\Phi \vdash \phi^k \ol{F}$}
\DisplayProof
\\[2ex]
\AXC{$\Gamma;\ol{\beta^0},\ol{\alpha^0},\phi^k \vdash F$}
\AXC{$\quad\quad\ol{\Gamma;\Phi,\ol{\beta^0} \vdash G}$}
\BIC{$\Gamma;\Phi,\ol{\beta^0} \vdash (\mu \phi^k.\lambda
  \ol{\alpha^0}. \,F)\,\ol{G}$} 
\DisplayProof
\end{array}\]

\vspace*{0.04in}

\[\begin{array}{cc}
\AXC{$\Gamma;\Phi \vdash F$}
\AXC{$\Gamma;\Phi \vdash G$}
\BIC{$\Gamma; \Phi \vdash F + G$}
\DisplayProof
&
\AXC{$\Gamma;\Phi \vdash F$}
\AXC{$\Gamma;\Phi \vdash G$}
\BIC{$\Gamma; \Phi \vdash F \times G$}
\DisplayProof
\end{array}\]
%\noindent
%A type $\tau$ is {\em well-formed} if it is
%either a well-formed type constructor expression or a
%well-formed functorial expression.
\end{dfn}
If $F$ is a closed type we may write $\vdash F$, rather than
$\emptyset;\emptyset \vdash F$, for the judgment that it is
well-formed.  Definition~\ref{def:wftypes} ensures that the expected
weakening rules for well-formed types hold, although weakening does
not change the contexts in which types can be formed. If
$\Gamma;\emptyset \vdash F$ and $\Gamma;\emptyset \vdash G$,
then our rules allow formation of the type $\Gamma;\emptyset \vdash
\Nat^\emptyset F \,G$. Since $\Gamma; \emptyset \vdash
\Nat^{\ol \alpha}F\,G$ represents a natural transformation in
$\ol \alpha$ from $F$ to $G$, the type $\Gamma;\emptyset
\vdash \Nat^\emptyset F \,G$ represents the arrow type
$\Gamma \vdash F \to G$ in our calculus. Our calculus
similarly represents the $\forall$-type $\Gamma; \emptyset
\vdash \forall \ol\alpha . F$ as $\Gamma; \emptyset \vdash
\Nat^{\ol\alpha} \,\onet \,F$.
%However, if $\ol\alpha$ is non-empty then $\tau$ cannot be of the
%form $\Nat^{\ol\beta} H \, K$ since $\Gamma; \ol\alpha \vdash
%\Nat^{\ol\beta} H \, K$ is not a valid type judgment in our calculus
%(except by weakening).

Definition~\ref{def:wftypes} allows the formation of all of the
(closed) nested types from the introduction:
\[\begin{array}{lll}
\mathit{List}\, \alpha & = & \mu \beta. \,\onet + \alpha \times
\beta\; = \; (\mu \phi. \lambda \beta.\,\onet + \beta \times \phi
\beta)\,\alpha\\ 
\mathit{PTree}\,\alpha & = & (\mu \phi. \lambda \beta.\,\beta +
\phi\,(\beta \times \beta))\,\alpha\\
\mathit{Forest}\,\alpha & = & (\mu \phi. \lambda \beta. \,\onet +
\beta \times \mathit{PTree}\,(\phi \beta))\,\alpha\\ 
\mathit{Bush}\,\alpha & = & (\mu \phi.\lambda \beta. \,\onet + \beta
\times \phi\,(\phi\beta))\,\alpha 
\end{array}\]
Each of these types can be considered either functorial in $\alpha$ or
not, according to whether $\alpha$ is in the functorial or
non-functorial context in which it is formed. For example, if
$\emptyset;\alpha \vdash \mathit{List} \,\alpha$, then $\vdash
\Nat^\alpha \onet\, (\mathit{List}\,\alpha)$ is well-formed; if
$\alpha;\emptyset \vdash \List \,\alpha$, then it is not.  If
$\mathit{Tree}\,\alpha\,\gamma = \mu \,\beta.\, \alpha + \beta \times
\gamma \times \beta$, then both $\gamma;\emptyset \vdash \Nat^\alpha
(\mathit{List}\, \alpha) \, (\mathit{Tree} \, \alpha\, \gamma)$ and
$\emptyset;\gamma \vdash \Nat^\alpha (\mathit{List}\, \alpha) \,
(\mathit{Tree} \, \alpha\, \gamma)$ are well-formed.  Although types
can be functorial in variables of arity greater than $0$, the body $F$
of a type $(\mu \phi.\lambda \ol\alpha.F)\,\ol G$ cannot. For example,
the type $\mathit{GRose}\, \phi \, \alpha = \mu\beta . \alpha \times
\phi \beta$ can be functorial or not in $\alpha$, but cannot be
functorial in $\phi$.  As a result, $\mathit{GRose}\,\phi\,\alpha$
cannot be the (co)domain of a $\Nat$ type representing a natural
transformation in $\phi$, and the type $\mu \phi. \lambda
\alpha. \mathit{GRose}\,\phi\alpha$ is not well-formed in our
system. These restrictions on higher-arity functoriality are necessary
for constructing our parametric model.  Fortunately, they are entirely
natural, and {\color{blue} preclude only truly bizarre nested types}.
%do not preclude any nested
%types {\color{blue} occurring in practice}.

%If $\mathit{Tree}\,\alpha\,\gamma = \mu \,\beta.\, \alpha + \beta \times
%\gamma \times \beta$, then Definition~\ref{def:wftypes} also allows
%the derivation of, e.g., the type $\gamma; \emptyset \vdash
%\Nat^\alpha (\mathit{List}\, \alpha) \, (\mathit{Tree} \, \alpha\,
%\gamma)$ representing a natural transformation from lists to trees
%that is natural in $\alpha$ but not necessarily in $\gamma$. We
%emphasize that types can be functorial in variables of arity greater
%than $0$. For example, the type $\mathit{GRose}\, \phi \, \alpha =
%\mu\beta . \alpha \times \phi \beta$ can be functorial in $\phi$ if
%$\phi \in \Phi$. As usual, whether $\phi \in \Gamma$ or $\phi \in
%\Phi$ determines whether types such as $\Nat^\alpha ( \mathit{GRose}
%\, \phi \, \alpha)\, (List\, \alpha)$ are well-formed. But even if
%$\mathit{GRose}$ is functorial in $\phi$, it still cannot be the
%domain of a $\Nat$ type representing a natural transformation in
%$\phi$. This is because our calculus does not allow naturality in
%variables of arity greater than $0$.

Definition~\ref{def:wftypes} allows well-formed types to be 
%%considers types in $\T$ to be types in $\F$ that are
functorial in no variables.
%This allows the formation of types such as
%$\mathit{List}\,(\sigma \to \tau)$ and $\mathit{PTree}\, (\forall
%\alpha.\tau)$.
Functorial variables
%in a well-formed type $\tau$
can also be demoted to non-functorial status. The proof is by
induction on the structure of $F$.

\begin{lemma}
If\, $\Gamma; \Phi, \phi^k \vdash F$, then $\Gamma, \psi^k; \Phi
\vdash F[\phi^k :== \psi^k]$ is also derivable. Here, $F[\phi :==
  \psi]$ is the textual replacement of $\phi$ in $F$, meaning that all
occurences of $\phi\ol\sigma$ in $F$ become $\psi\ol\sigma$.
\end{lemma}

In addition to textual replacement, we also have a proper substitution
operation on types. If $F$ is a type over $P$ and $V$, if $\Pos$
and $V$ contain only type variables of arity $0$, and if $k=0$ for
every occurrence of $\phi^k$ bound by $\mu$ in $F$, then we say
that $F$ is {\em first-order}; otherwise we say that $F$ is {\em
  second-order}.  Substitution for first-order types is the usual
capture-avoiding textual substitution. We write $F[\alpha :=
  \sigma]$ for the result of substituting $\sigma$ for $\alpha$ in
$F$, and $F[\alpha_1 := F_1,...,\alpha_k := F_k]$, or
$F[\ol{\alpha := F}]$ when convenient, for $F[\alpha_1 :=
  F_1][\alpha_2 := F_2,...,\alpha_k := F_k]$. Substitution
for second-order types is defined below, where we adopt a similar
notational convention for vectors of types. We note that it is not
correct to substitute along non-functorial variables.

\begin{dfn}\label{def:second-order-subst}
If \,$\Gamma; \Phi,\phi^k \vdash H$
%with $k \geq 1$,
and if\, $\Gamma;\Phi, \ol{\alpha} \vdash F$ with
$|\ol\alpha| = k$, then $\Gamma;\Phi \vdash H[\phi :=_{\ol{\alpha}}
  F]$.  Similarly, if \,$\Gamma, \phi^k; \Phi \vdash H$,
%with $k \geq 1$,
and if\, $\Gamma; \Psi,\ol{\alpha} \vdash F$ with $|\ol\alpha| = k$
and $\Phi \cap \Psi = \emptyset$, then $\Gamma,\Psi;\Phi \vdash H[\phi
  :=_{\ol{\alpha}} F]$. Here, the operation $(\cdot)[\phi :=_{\ol \alpha}
  F]$ of {\em second-order type substitution (along $\ol\alpha$)} is
defined by: 
\[\begin{array}{lll}
\zerot[\phi :=_{\ol{\alpha}} F] & = & \zerot\\[0.5ex]
\onet[\phi :=_{\ol{\alpha}} F] & = & \onet\\[0.25ex]
(\Nat^{\ol\beta} G \,K)[\phi :=_{\ol{\alpha}} F]
& = & \Nat^{\ol\beta}\, (G[\phi :=_{\ol{\alpha}} F]) \,(K[\phi
  :=_{\ol{\alpha}} F])\\
(\psi\ol{G})[\phi :=_{\ol{\alpha}} F] & = &
\left\{\begin{array}{ll}
\psi \,\ol{G[\phi :=_{\ol{\alpha}} F]} & \mbox{if } \psi \not = \phi\\
  F[\ol{\alpha  := G[\phi :=_{\ol{\alpha}} F]}] 
  & \mbox{if } \psi = \phi
\end{array}\right.\\[2.8ex]
(G + K)[\phi :=_{\ol{\alpha}} F] & = & G[\phi
  :=_{\ol{\alpha}} F] + K[\phi :=_{\ol{\alpha}} F]\\[0.5ex] 
(G \times K)[\phi :=_{\ol{\alpha}} F] & = &
G[\phi :=_{\ol{\alpha}} F] \times K[\phi
  :=_{\ol{\alpha}} F]\\[0.5ex]   
((\mu \psi. \lambda \ol{\beta}.\, G)\ol{K})[\phi :=_{\ol{\alpha}}
  F] & = & (\mu \psi. \lambda \ol{\beta}. \,G[\phi :=_{\ol{\alpha}}
  F])\, \ol{K[\phi :=_{\ol{\alpha}} F]}
\end{array}\]
\end{dfn}
\noindent
We note that $(\cdot)[\phi^0 :=_\emptyset F]$ coincides with
first-order substitution. We also omit $\ol\alpha$ when convenient.

\subsection{Terms}\label{sec:terms}

We now define our term calculus. To do so we assume an infinite set
$\cal V$ of term variables disjoint from $\tvars$ and $\fvars$. If
$\Gamma$ is a type constructor context and $\Phi$ is a functorial
context, then a {\em term context for $\Gamma$ and $\Phi$} is a finite
set of bindings of the form $x : F$, where $x \in {\cal V}$ and
$\Gamma; \Phi \vdash F$. We adopt the same conventions for denoting
disjoint unions and for vectors in term contexts as for type
constructor contexts and functorial contexts.

\begin{dfn}\label{def:well-formed-terms}
Let $\Delta$ be a term context for $\Gamma$ and $\Phi$.  The formation
rules for the set of\, {\em well-formed terms over $\Delta$} are
\[\begin{array}{ccc}
%\AXC{$\Gamma;\emptyset \vdash \tau : \T$}
%\UIC{$\Gamma;\emptyset\,|\, \Delta,x :\tau \vdash x : \tau$}
%\DisplayProof
\AXC{$\Gamma;\Phi \vdash F$}
\UIC{$\Gamma;\Phi \,|\, \Delta,x :F \vdash x : F$}
\DisplayProof
&
\AXC{$\Gamma;\Phi \,|\, \Delta \vdash t : \zerot$}
\AXC{$\Gamma;\Phi \vdash F$}
\BIC{$\Gamma;\Phi \,|\, \Delta \vdash \bot_F t  : F$}
\DisplayProof
&
\AXC{$\phantom{\Gamma;\Phi}$}
\UIC{$\Gamma;\Phi \,|\, \Delta \vdash \top : \onet$}
\DisplayProof\\\\
\end{array}\]
\[\begin{array}{cc}
\AXC{$\Gamma;\Phi \,|\, \Delta \vdash s: F$}
\UIC{$\Gamma;\Phi \,|\, \Delta \vdash \inl \,s: F + G$}
\DisplayProof
&
\AXC{$\Gamma;\Phi \,|\, \Delta \vdash t : G$}
\UIC{$\Gamma;\Phi \,|\, \Delta \vdash \inr \,t: F + G$}
\DisplayProof\\\\
\end{array}\]
\[\begin{array}{c}
\AXC{$\Gamma; \Phi \vdash F,G$}
\AXC{$\Gamma;\Phi \,|\, \Delta \vdash t : F+G$}
\AXC{$\Gamma;\Phi \,|\, \Delta, x : F \vdash l : K \hspace{0.3in} \Gamma;\Phi \,|\, \Delta, y : G \vdash r : K$}
\TIC{$\Gamma;\Phi~|~\Delta \vdash \case{t}{x \mapsto l}{y \mapsto r} : K$}
\DisplayProof
\end{array}\]

\vspace*{0.05in}

\[\begin{array}{lll}
\AXC{$\Gamma;\Phi \,|\, \Delta \vdash s: F$}
\AXC{$\Gamma;\Phi \,|\, \Delta \vdash t : G$}
\BIC{$\Gamma;\Phi \,|\, \Delta \vdash (s,t) : F \times G$}
\DisplayProof
&
\AXC{$\Gamma;\Phi \,|\, \Delta \vdash t : F \times G$}
\UIC{$\Gamma;\Phi \,|\, \Delta \vdash \pi_1 t : F$}
\DisplayProof
&
\AXC{$\Gamma;\Phi \,|\, \Delta \vdash t : F \times G$}
\UIC{$\Gamma;\Phi \,|\, \Delta \vdash \pi_2 t : G$}
\DisplayProof
\end{array}\]

\vspace*{0.05in}

\[\begin{array}{c}
%\AXC{$\Gamma; \emptyset \vdash \Nat^{\ol{\alpha}} \,F\,G : \T$}
\AXC{$\Gamma; \ol{\alpha} \vdash F$}
\AXC{$\Gamma; \Phi, \ol{\alpha} \vdash G$}
\AXC{$\Gamma; \Phi, \ol{\alpha} \,|\, \Delta, x : F \vdash t: G$} 
\TIC{$\Gamma; \Phi
  %\emptyset
  \,|\, \Delta \vdash L_{\ol{\alpha}} x.t : \Nat^{\ol{\alpha}} \,F \,G$}
\DisplayProof
\\\\
\AXC{$\Gamma; \Phi
  %\emptyset
  \,|\, \Delta \vdash t : \Nat^{\ol{\alpha}} \,F \,G$}
\AXC{$\ol{\Gamma;\Phi \vdash K}$}
\AXC{$\Gamma;\Phi \,|\, \Delta \vdash s: F[\overline{\alpha := K}]$}
\TIC{$\Gamma;\Phi \,|\, \Delta \vdash t_{\ol K} s:
  G[\overline{\alpha := K}]$}
\DisplayProof
\\\\
\AXC{$\Gamma; \ol{\phi}, \ol{\gamma} \vdash H$}
\AXC{$\ol{\Gamma; \ol{\beta},\ol{\gamma} \vdash F}$}
\AXC{$\ol{\Gamma; \Phi, \ol{\beta},\ol{\gamma} \vdash
    G}$\hspace*{0.3in}$\ol{\Gamma;\Phi~|~\Delta \vdash t :
  \Nat^{\ol{\beta},\ol{\gamma}}\,F\,G}$}
\TIC{$\Gamma; \Phi
  %\emptyset
  ~|~\Delta
%  \ol{ x : \Nat^{\ol{\beta},\ol{\gamma}}\,F\,G}
  %  \emptyset
  \vdash \map^{\ol{F},\ol{G}}_H \, \ol t :
%  \Nat^\emptyset\;(\ol{\Nat^{\ol{\beta},\ol{\gamma}}\,F\,G})\;
  \Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\;H[\ol{\phi
      :=_{\ol{\beta}} G}]$} 
\DisplayProof
\\\\
%\AXC{$\Gamma;\Phi \,|\, \Delta \vdash t : H[\phi := (\mu \phi.\lambda 
%  \ol{\alpha}.H)\ol{\beta}][\ol{\alpha := \tau}]$}
%\AXC{$\ol{\Gamma;\Phi \vdash \tau : \F}$}
\AXC{$\Gamma; \phi, \ol{\alpha},\ol{\gamma} \vdash H$}
\UIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \tin_H :
  \Nat^{\ol{\beta},\ol{\gamma}} H[\phi :=_{\ol{\beta}} (\mu
    \phi.\lambda \ol{\alpha}.H)\ol{\beta}][\ol{\alpha := \beta}]\,(\mu
  \phi.\lambda \ol{\alpha}.H)\ol{\beta}$}
\DisplayProof
\\\\
\AXC{$\Gamma; \phi,\ol{\alpha}, \ol{\gamma} \vdash H$}
\AXC{$\Gamma; \ol{\beta}, \ol{\gamma} \vdash F$}
%\AXC{$\Gamma;\emptyset \,|\, \Delta \vdash t :
%  \Nat^{\ol{\beta}, \ol{\gamma}}\,H[\phi :=_{\ol{\beta}} F][\ol{\alpha
%      := \beta}]\,F$} 
%\TIC{$\Gamma;\emptyset \,|\, \Delta \vdash
%  \fold_{H,F}\, t : \Nat^{\ol{\beta}, \ol{\gamma}}\,((\mu
%  \phi.\lambda \ol{\alpha}.H)\ol{\beta})\,F$} 
\BIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \fold^F_H :
  \Nat^\emptyset\; (\Nat^{\ol{\beta}, \ol{\gamma}}\,H[\phi
    :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\; (\Nat^{\ol{\beta},
    \ol{\gamma}}\,(\mu \phi.\lambda \ol{\alpha}.H)\ol{\beta})\,F)$}
\DisplayProof
\end{array}\]
\end{dfn}

In the rule for $L_{\ol{\alpha}}x.t$, the $L$ operator binds all
occurrences of the type variables in $\ol{\alpha}$ in the type of the
term variable $x$ and in the body $t$, as well as all occurrences of
$x$ in $t$. In the rule for $t_{\ol K} s$ there is one functorial
expression $K$ for every functorial variable $\alpha$. In the rule
for $\map^{\ol{F},\ol{G}}_H$ there is one functorial expression $F$
and one functorial expression $G$ for each functorial variable in
$\ol\phi$. Moreover, for each $\phi^k \in \ol\phi$ the number of
functorial variables $\beta$ in the judgments for its corresponding
functorial expresssions $F$ and $G$ is $k$. In the rules for $\tin_H$
and $\fold^F_H$, the functorial variables in $\ol{\beta}$ are fresh
with respect to $H$, and there is one $\beta$ for every
$\alpha$. (Recall from above that, in order for the types of $\tin_H$
and $\fold^F_H$ to be well-formed, the length of $\alpha$ must equal
the arity of $\phi$.) Substitution for terms is the obvious extension
of the usual capture-avoiding textual substitution, and
Definition~\ref{def:well-formed-terms} ensures that the expected
weakening rules for well-formed terms hold.

Using Definition~\ref{def:well-formed-terms} we can represent the
$\mathtt{reversePTree}$ function from Figure~\ref{fig:funs} in our
calculus as
\[\vdash \fold_{\beta + \phi(\beta \times \beta)}^{\mathit{PTree}\,
  \alpha} \, (\tin_{\beta + \phi(\beta \times \beta)} \circ s) :
\Nat^{\alpha} (\mathit{PTree}\,\alpha)\,(\mathit{PTree}\,\alpha)\] 
where
\[\begin{array}{lll}
\fold_{\beta + \phi(\beta \times \beta)}^{\mathit{PTree}\,\alpha} & :
& \Nat^{\emptyset} (\Nat^{\alpha} (\alpha + \mathit{PTree}\,(\alpha
\times \alpha)) \; (\mathit{PTree}\,\alpha))\; (\Nat^{\alpha}
(\mathit{PTree}\,\alpha) \; (\mathit{PTree}\,\alpha))\\
\tin_{\beta + \phi(\beta \times \beta)} & : & \Nat^{\alpha} (\alpha +
\mathit{PTree}\,(\alpha \times \alpha)) \; (\mathit{PTree}\, \alpha)\\
\map_{\mathit{PTree}\,\alpha}^{\alpha \times \alpha, \alpha \times
  \alpha} t & : & \Nat^{\alpha} (\mathit{PTree}\,(\alpha \times
\alpha))\, (\mathit{PTree}\,(\alpha \times \alpha))
%\[\begin{array}{lll}
%\vdash \fold_{\beta + \phi(\beta \times \beta)}^{\mathit{PTree}\,\alpha} &
%: & \Nat^{\emptyset} (\Nat^{\alpha} (\alpha + \mathit{PTree}\,(\alpha \times
%\alpha)) \; (\mathit{PTree}\,\alpha))\; (\Nat^{\alpha}
%(\mathit{PTree}\,\alpha) \; (\mathit{PTree}\,\alpha))\\
%\vdash \tin_{\beta + \phi(\beta \times \beta)} & : & \Nat^{\alpha}
%(\alpha + \mathit{PTree}\,(\alpha \times \alpha)) \; (\mathit{PTree}\,
%\alpha)\\ 
%\vdash \map_{\mathit{PTree}\,\alpha}^{\alpha \times \alpha, \alpha \times
%  \alpha} & : & \Nat^{\emptyset} (\Nat^{\alpha} (\alpha \times \alpha)\,
%(\alpha \times \alpha))\; (\Nat^{\alpha} (\mathit{PTree}\,(\alpha \times
%\alpha))\, (\mathit{PTree}\,(\alpha \times \alpha)))
\end{array}\]
for $t : \Nat^{\alpha} (\alpha \times \alpha)\, (\alpha \times
\alpha)$, 
and 
$\mathit{swap}$ and $s$ are the terms
\[ \vdash L_{\alpha} p.\, (\pi_2 p, \pi_1 p) :
\Nat^{\alpha} (\alpha \times \alpha)\, (\alpha \times \alpha)\]
and
\[\begin{array}{l}
\vdash L_{\alpha} t. \,\case{t}{b \mapsto
  b}{t' \mapsto  
    \map_{\mathit{PTree}\,\alpha}^{\alpha \times \alpha, \alpha \times
      \alpha}\, \mathit{swap}\, t'} )
:  \Nat^{\alpha} (\alpha + \mathit{PTree}\,(\alpha \times
\alpha))\; (\alpha + \mathit{PTree}\,(\alpha \times \alpha))
\end{array}\]
respectively. We can similarly represent the $\mathtt{reverseBush}$
function from Figure~\ref{fig:funs2} as
\[\vdash \fold_{\onet + \beta \times
\phi (\phi\beta)}^{\mathit{Bush}\,\alpha}\, ( \tin_{\onet + \beta \times
  \phi (\phi\beta)} \circ (\onet + t \circ i \circ i') ) :
\Nat^{\alpha} (\mathit{Bush}\,\alpha)\,(\mathit{Bush}\,\alpha)\] 
where
\[\begin{array}{lll}
\vdash \fold_{\onet + \beta \times \phi
  (\phi\beta)}^{\mathit{Bush}\,\alpha} : \Nat^{\emptyset}\, (\Nat^{\alpha}\, 
(\onet + \alpha \times \mathit{Bush}\, (\mathit{Bush}\, \alpha))) \;
(\mathit{Bush}\,\alpha))\; (\Nat^{\alpha} \,(\mathit{Bush}\,\alpha) \;
(\mathit{Bush}\,\alpha))\\ 
\vdash \tin_{\onet + \beta \times \phi (\phi\beta)} : \Nat^{\alpha}\,
(\onet + \alpha \times \mathit{Bush}\, (\mathit{Bush} \, \alpha))
\;(\mathit{Bush}\, \alpha)\\
\end{array}\]
and $\mathit{bnil}$, $\mathit{bcons}$, $\tin^{-1}_{\onet + \beta
  \times \phi (\phi\beta)}$, $t$, $i$, and 
$i'$ are the terms
\[\begin{array}{l}
\vdash \tin_{\onet + \beta \times \phi (\phi\beta)} \circ (
L_{\alpha}\, x.\, \inl\, x) \; : \; \Nat^{\alpha}\, \onet\;
(\mathit{Bush}\,\alpha)\\ 
\vdash \tin_{\onet + \beta \times \phi (\phi\beta)} \circ (
L_{\alpha}\, x.\, \inr\, x ) \; : \; \Nat^{\alpha}\, (\alpha \times
\mathit{Bush}\,(\mathit{Bush}\,\alpha))\; (\mathit{Bush}\,\alpha)\\
\vdash \fold_{\onet + \beta \times \phi (\phi\beta)}^{(\onet + \beta
  \times \phi (\phi\beta))[\phi := \mathit{Bush}\,\alpha]}\,
(\map_{\onet + \beta \times \phi (\phi\beta)}^{(\onet + \beta \times
  \phi (\phi\beta))[\phi := \mathit{Bush}\,\alpha][\beta := \alpha],
  \mathit{Bush}\,\alpha}\, \tin_{\onet + \beta \times \phi
  (\phi\beta)})\\
\hspace*{0.2in} : \Nat^{\alpha}\, (\mathit{Bush}\,\alpha)\; (\onet +
\alpha \times \mathit{Bush}\,(\mathit{Bush}\,\alpha))\\ 
\vdash L_{\alpha}\, (b, s). \,\mathsf{case}\;s\,\mathsf{of}\, \{ \hspace*{0.21in}
        \ast \mapsto
\mathit{bcons}_{\alpha}\, b\, (\mathit{bnil}_{\alpha}\, \ast); \\
\hspace*{1.15in}\,(s', u) \mapsto \mathsf{case}\,{s'}\,\mathsf{of}\,
\{\hspace*{0.3in} \ast
\mapsto \mathit{bcons}_{\alpha}\, b \,(\mathit{bcons}_{\mathit{Bush}\,
  \alpha}\, (\mathit{bnil}_{\alpha}\, \ast)\, u); \\
\hspace*{2.3in}(b', u') \mapsto \mathit{bcons}_{\alpha}\, b'
(\mathit{bcons}_{\mathit{Bush}\, \alpha}\, (\mathit{bcons}_{\alpha}\,
b\, u )\, u')\}\}\\
\hspace{0.2in} : \Nat^{\alpha} (\alpha \times (\onet + (\onet + \alpha
\times \mathit{Bush}\, (\mathit{Bush} \, \alpha))) \times
\mathit{Bush}\, (\mathit{Bush} \, (\mathit{Bush}\, \alpha))) \;
(\alpha \times \mathit{Bush}\, (\mathit{Bush} \, \alpha))\\
\vdash \alpha \times (\onet + \tin^{-1}_{\onet + \beta \times \phi
  (\phi\beta)} \times \mathit{Bush}\, (\mathit{Bush} \,
(\mathit{Bush}\, \alpha)))\\
\hspace{0.2in} : \Nat^{\alpha} (\alpha \times (\onet + \mathit{Bush}
\, \alpha \times \mathit{Bush}\,( \mathit{Bush} \, (\mathit{Bush}\,
\alpha))))\\
\hspace*{0.57in}(\alpha \times (\onet + (\onet + \alpha \times
\mathit{Bush}\, (\mathit{Bush} \, \alpha)) \times \mathit{Bush}\,
(\mathit{Bush} \, (\mathit{Bush}\, \alpha))))\\
\vdash \alpha \times (L_{\alpha}\,x.\, (\tin^{-1}_{\onet +
  \beta \times \phi (\phi\beta)})_{\mathit{Bush}\,\alpha}\, x)\\
\hspace{0.2in} : \Nat^{\alpha} (\alpha
\times \mathit{Bush} \, (\mathit{Bush}\, \alpha)) \; (\alpha \times (\onet
+ \mathit{Bush} \, (\alpha \times \mathit{Bush}\, (\mathit{Bush} \,
(\mathit{Bush}\, \alpha)))))
\end{array}\]
respectively. Here, $\Gamma; \emptyset \,|\, \Delta \vdash H+
\eta : \Nat^{\ol{\alpha}} (H+ F)\; (H+ G)$ and $\Gamma;
\emptyset \,|\, \Delta \vdash H\times \eta : \Nat^{\ol{\alpha}}
(H\times F)\; (H\times G)$ for $H+ \eta :=
L_{\ol{\alpha}}\, x.\, \case{x}{s \mapsto s}{t
  \mapsto  
  (\eta_{\ol{\alpha}} t)}$ and $H\times \eta := L_{\ol{\alpha}}\,
x.\, (\pi_1 x, \eta_{\ol{\alpha}} (\pi_2 x))$ for $\Gamma; \emptyset
\,|\, \Delta \vdash \eta : \Nat^{\ol{\alpha}} F\; G$ and $\Gamma;
\ol{\alpha} \vdash H$.

Because our calclus is designed to ensure functoriality, and
functoriality does not interact with polymorphism as well as might be
hoped, our calculus cannot express nontrivial recursive functions, such
as a concatenation function for perfect trees, that take not just one
nested type parameterized over a variable $\alpha$ as input, but other
inputs parameterized over $\alpha$ as well. The fundamental issue is
that recursion is expressible only via folds, and it is well-known
that a fold over a nested type must take a natural transformation as
an argument and return a natural transformation as its result.
%and, unfortunately, the polymorphism natural transformations entail
%leaves no nontrivial way to incorporate data from the second argument
%to a function with a type like that above into a fold over its first
%argument.
Even for ADTs there is a difference between which folds over them we
can write when ADTs are viewed as proper nested types --- i.e., as
fixpoints of types interpreted as first-order functors --- and which
ones we can write when ADTs are viewed as nested types --- i.e., as
fixpoints of types that are interpreted as higher-order functors. This
is because enforcing functoriality requires that the arity of the
recursion variable $\phi$ in the fixpoint in the return type of
$\fold$ must equal the number of variables bound by its $\Nat$
construct. For ADTs this number can be $0$, making it possible to
write as folds functions of types such as $\alpha; \emptyset \vdash
\Nat^\emptyset \ (\List\,\alpha)\,(\Nat^\emptyset \alpha\,(\List
    {\color{blue} \nat}))$, all of whose argument types are
    parameterized over the same type $\alpha$. This is not possible
    for nested types, even when those nested types are semantically
    equivalent to ADTs.
%Currying the type of the function does not help either, since the
%codomain of a fold applied to an algebra must be a functor, and this
%clearly is not the case for a function on nested types that takes
%more than one input involving the same type variable.
We can, however, represent types of functions that take
%    as arguments
%one proper nested type and as many other (possibly proper)
%as many proper nested types as we wish
multiple nested types as arguments, provided the types of those
arguments are parameterized over disjoint sets of type variables. The
return type of the function can be parameterized over any or all of
the type variables occurring in its argument types.  For example, we
can represent the type of a zip-like function for bushes.
%A well-known deeper limitation on folds is that a fold over a nested
%type must take a natural transformation as an argument and return a
%natural transformation; this issue is discussed in more detail in
%Section~\ref{sec:conclusion},
%Unfortunately, the polymorphism natural transformations
%entail leaves no nontrivial way to incorporate data from the second
%argument to a function with a type like that above into a fold over
%its first argument.
Note, however, that even some recursive functions of a single proper
nested type --- e.g., a $\mathtt{reverseBush}$ function that is a true
involution --- cannot be expressed as folds because the algebra
arguments needed to define them are again recursive functions
%  cannot be defined. This happens, for
%  example, when the algebra argument to a fold must be a recursive
%  function
%, and thus must itself be defined as a fold
with type signatures of the same problematic forms.
%More sophisticated combinators for stylized recursion (e.g.,
%generalized folds)
%%  ~\cite{bp99})
%don't mitigate these difficulties, either. Indeed, they appear to be a
%fundamental consequence of the interaction of functoriality and
%polymorphism, not just some peculiarity of our particular calculus.
Adding combinators for more sophisticated stylized recursion may
ultimately enhance the expressivity of our calculus but, as discussed
in Section~\ref{sec:conclusion}, the expressivity issue for folds for
nested types long pre-dates, and is somewhat orthogonal to, the issue
of parametricity for languages that construct nested types as
fixpoints studied here.
%, and is orthogonal to, our focus on
%constructing parametric models for calculi in which nested types are
%constructed as fixpoints.

The presence of the ``extra'' functorial variables in $\ol{\gamma}$ in
the rules for $\map^{\ol{F},\ol{G}}_H$, $\tin_H$, and $\fold^F_H$
merit special mention. They allow us to map or fold polymorphic
functions over nested types. Consider, for example, the polymorphic
function $\mathit{flatten} : \Nat^\beta
(\mathit{PTree}\,\beta)\,(\mathit{List}\,\beta)$ that uniformly maps
perfect trees to lists. Even in the absence of extra variables the
instance of $\map$ required to map each non-functorial monomorphic
instantiation of $\mathit{flatten}$ over a list of perfect trees is
well-typed:
\[\begin{array}{l}
\AXC{$\Gamma;\alpha \vdash \mathit{List} \, \alpha$}
\AXC{$\Gamma;\emptyset \vdash \mathit{PTree}\,F$}
\AXC{$\Gamma;\emptyset \vdash \mathit{List}\,G$ \hspace*{0.3in}
  $\Gamma;\emptyset~|~\Delta \vdash t
  : \Nat^\emptyset\,(\mathit{PTree}\, F)\,(\mathit{List}\,
  G)$}
\TIC{$\Gamma;\emptyset \,|\, \Delta \vdash \map^{\mathit{PTree}\,F,
    \,\mathit{List}\,G}_{\mathit{List}\,\alpha} \,t :
  \Nat^\emptyset\,(\mathit{List}\, (\mathit{PTree}\,
  F))\,(\mathit{List}\, (\mathit{List}\, G))$} \DisplayProof
\end{array}\]
But in the absence of $\ol \gamma$, if $\Gamma;
\emptyset~|~\Delta\vdash t : \Nat^\beta(\mathit{PTree}\,
\beta)\,(\mathit{List}\, \beta)$ then the instance
\[\Gamma;\emptyset~|~\Delta \vdash
\map^{\mathit{PTree}\,\beta,\mathit{List}\,\beta}_{\mathit{List}\,\alpha}\,t
: \Nat^\beta\,(\mathit{List}\, (\mathit{PTree}\,
\beta))\,(\mathit{List}\, (\mathit{List}\, \beta))\]
%\[\Gamma;\emptyset \,|\, x : \Nat^\beta(\mathit{PTree}\,
%\beta)\,(\mathit{List}\, \beta) \vdash
%\map^{\mathit{PTree}\,\beta,\mathit{List}\,\beta}_{\mathit{List}\,\alpha}\,x
%: \Nat^\beta\,(\mathit{List}\,
%(\mathit{PTree}\, \beta))\,(\mathit{List}\, (\mathit{List}\,
%\beta))\] of $\mathtt{map}$
required to map the {\em polymorphic}
$\mathit{flatten}$ function over a list of perfect trees is not: in
that setting the functorial contexts for $F$ and $G$ in the rule for
$\map^{F,G}_H$ would have to be empty, but the fact that the
polymorphic $\mathit{flatten}$ function is functorial in some
variable, say $\delta$, means that it cannot possibly have a type of
the form $\Nat^\emptyset F\, G$ that would be required for it to be
the function input to $\map$. Since untypeability of this instance of
$\map$ is unsatisfactory in a polymorphic calculus, where we naturally
expect to be able to manipulate entire polymorphic functions rather
than just their monomorphic instances, we use the ``extra'' variables
in $\ol \gamma$ to remedy the situation. Specifically, the rules from
Definition~\ref{def:well-formed-terms} ensure that the instance of
$\map$ needed to map the polymorphic $\mathit{flatten}$ function is
typeable as follows:
\[\begin{array}{l}
\AXC{$\Gamma;\alpha,\beta \vdash \mathit{List} \, \alpha$}
\AXC{$\Gamma;\beta \vdash \mathit{PTree}\,\beta \hspace*{0.3in}
  \Gamma;\beta \vdash \mathit{List}\,\beta$\hspace*{0.3in}
$\Gamma;\emptyset~|~\Delta \vdash t : \Nat^\beta(\mathit{PTree}\,
  \beta)\,(\mathit{List}\, \beta)$}
\BIC{$\Gamma;\emptyset~|~\Delta \vdash
  \map^{F,G}_{\mathit{List}\,\alpha} \, t :
 \Nat^\beta\,(\mathit{List}\,
  (\mathit{PTree}\, \beta))\,(\mathit{List}\, (\mathit{List}\,
  \beta))$}
\DisplayProof
\end{array}\]
Similar remarks explain the appearance of $\ol \gamma$ in the typing
rules for $\tin$ and $\fold$.

\vspace*{0.05in}

\section{Interpreting Types}\label{sec:type-interp}

We denote the category of sets and functions by $\set$. The category
$\rel$ has as its objects triples $(A,B,R)$ where $R$ is a relation
between the objects $A$ and $B$ in $\set$, i.e., a subset of $A \times
B$, and has as its morphisms from $(A,B,R)$ to $(A',B',R')$ pairs $(f
: A \to A',g : B \to B')$ of morphisms in $\set$ such that $(f a,g\,b)
\in R'$ whenever $(a,b) \in R$. We write $R : \rel(A,B)$ in place of
$(A,B,R)$ when convenient.  If $R : \rel(A,B)$ we write $\pi_1 R$ and
$\pi_2 R$ for the {\em domain} $A$ of $R$ and the {\em codomain} $B$
of $R$, respectively.  If $A : \set$, then we write $\Eq_A =
(A,A,\{(x,x)~|~ x \in A\})$ for the {\em equality relation} on $A$.

The key idea underlying Reynolds' parametricity is to give each type
$F(\alpha)$ with one free variable $\alpha$ both an {\em object
  interpretation} $F_0$ taking sets to sets and a \emph{relational
  interpretation} $F_1$ taking relations $R : \rel(A,B)$ to relations
$F_1 (R) : \rel(F_0 (A), F_0 (B))$, and to interpret each term
$t(\alpha,x) : F(\alpha)$ with one free term variable $x : G(\alpha)$
as a map $t_0$ associating to each set $A$ a function $t_0(A) : G_0(A)
\to F_0(A)$. These interpretations are to be given inductively on the
structures of $F$ and $t$ in such a way that they imply two
fundamental theorems. The first is an \emph{Identity Extension Lemma},
which states that $F_1(\Eq_A) = \Eq_{F_0(A)}$, and is the essential
property that makes a model relationally parametric rather than just
induced by a logical relation. The second is an \emph{Abstraction
  Theorem}, which states that, for any $R :\rel(A, B)$,
$(t_0(A),t_0(B))$ is a morphism in $\rel$ from
$(G_0(A),G_0(B),G_1(R))$ to $(F_0(A),F_0(B),F_1(R))$. The Identity
Extension Lemma is similar to the Abstraction Theorem except that it
holds for {\em all} elements of a type's interpretation, not just
those that are interpretations of terms.
%i.e., $t_0(A)$ and $t_0(B)$ map related arguments to
%related results.
Similar results are expected to hold for types and terms with any
number of free variables.

The key to proving the Identity Extension Lemma
(Theorem~\ref{thm:iel}) in our setting is a familiar ``cutting down''
of the interpretations of universally quantified types, such as our
$\Nat$-types, to include only the ``parametric'' elements.
%(See, e.g.,~\cite{atk12,bfss90,gjfor15,rey83}).
This requires that set interpretations of types are defined
simultaneously with their relational interpretations. We give set
interpretations for our types in Section~\ref{sec:set-interp} and give
their relational interpretations in Section~\ref{sec:rel-interp}.
While the set interpretations are relatively straightforward, their
relation interpretations are less so, mainly because of the
cocontinuity conditions we must impose to ensure that they are
well-defined. We take some effort to develop conditions in
Section~\ref{sec:rel-interp}, which separates
Definitions~\ref{def:set-sem} and~\ref{def:rel-sem} in space, but
otherwise has no impact on the fact that they are given by mutual
induction.

\subsection{Interpreting Types as Sets}\label{sec:set-interp}

%A poset $\mcD = (D,\leq)$ is \emph{$\omega$-directed} if every finite
%subset of $D$ has an upper bound. When $\mcD$ is considered as a
%category, we write $d \in \mcD$ to indicate that $d$ is an object of
%$\mcD$ (i.e., $d \in D$). An {\em $\omega$-directed colimit} in a
%category $\mcC$ is a colimit of a diagram $F : {\mathcal D} \to \mcC$,
%where $\mathcal D$ is an $\omega$-directed poset. A category $\mcC$ is
%{\em $\omega$-cocomplete} if it has all $\omega$-directed colimits. A
%cocomplete category is one that has all colimits.
%
%If $\mcA$ and $\mcC$ are $\omega$-cocomplete, then the functor $F :
%\mcA \to \mcC$ is {\em $\omega$-cocontinuous} if it preserves
%$\omega$-directed colimits.  If $\mcA$ is locally small, then an
%object $A$ of $\mcA$ is {\em finitely presentable} if the functor
%$\mathsf{Hom}_\mcA(A, -) : \mcA \to \Set$ preserves $\omega$-directed
%colimits, i.e., if for every $\omega$-directed poset $\mathcal D$ and
%every functor $F : {\mathcal D} \to \mcC$, there is a canonical
%isomorphism $\colim{d \in \mcD}{\mathsf{Hom}_\mcA(A,Fd)} \simeq
%\mathsf{Hom}_{\mcA}(A, \colim{d \in \mcD}{Fd})$. A category $\mcA$ is
%       {\em finitely accessible} if it is $\omega$-cocomplete and has
%       a set $\mcA_0$ of finitely presentable objects such that every
%%       object is an $omega$-directed colimit of objects in $\mcA_0$;
%       it is {\em locally finitely presentable} if it is finitely
%       accessible and cocomplete.
We will interpret the types in our calculus as $\omega$-cocontinuous
functors on locally finitely presentable categories~\cite{ar94}. Since
functor categories of locally finitely presentable categories are
again locally finitely presentable, this will ensure, in particular,
that the fixpoints interpreting $\mu$-types in $\set$ and $\rel$
exist, and thus that both the set and relational interpretations of
all of the types in Definition~\ref{def:wftypes} are
well-defined~\cite{jp19}. To bootstrap this process, we interpret type
variables themselves as $\omega$-cocontinuous functors in
Definitions~\ref{def:set-env} and~\ref{def:reln-env}. If $\C$ and $\D$
are locally finitely presentable categories, we write $[\C,\D]$ for
the set of $\omega$-cocontinuous functors from $\C$ to $\D$.
%Note that the categories $\set$ and $\rel$ are both locally finitely
%presentable.

\begin{dfn}\label{def:set-env}
A {\em set environment} maps each type variable in $\tvars^k \cup
\fvars^k$ to an element of $[\set^k,\set]$.  A morphism $f : \rho \to
\rho'$ for set environments $\rho$ and $\rho'$ with $\rho|_\tvars =
\rho'|_\tvars$ maps each type constructor variable $\psi^k \in \tvars$
to the identity natural transformation on $\rho \psi^k = \rho'\psi^k$
and each functorial variable $\phi^k \in \fvars$ to a natural
transformation from the $k$-ary functor $\rho \phi^k$ on $\set$ to the
$k$-ary functor $\rho' \phi^k$ on $\set$.  Composition of morphisms on
set environments is given componentwise, with the identity morphism
mapping each set environment to itself. This gives a category of set
environments and morphisms between them, which we denote $\setenv$.
\end{dfn}
When convenient we identify a functor $F : [\set^0, \set]$ with the
set that is its codomain and consider a set environment to map
a type variable of arity $0$ to
%an $\omega$-cocontinuous functor from $\set^0$ to $\set$, i.e., to
a set.  If $\ol{\alpha} = \{\alpha_1,...,\alpha_k\}$ and $\ol{A} =
\{A_1,...,A_k\}$, then we write $\rho[\ol{\alpha := A}]$ for the set
environment $\rho'$ such that $\rho' \alpha_i = A_i$ for $i = 1,...,k$
and $\rho' \alpha = \rho \alpha$ if $\alpha \not \in
\{\alpha_1,...,\alpha_k\}$.  If $\rho$ is a set environment we write
$\Eq_\rho$ for the relation environment (see
Definition~\ref{def:reln-env}) such that $\Eq_\rho v = \Eq_{\rho v}$
for every type variable $v$.
%; see Definition~\ref{def:reln-env} for the complete definition of a
%relation environment.
The relational interpretations appearing in the second clause of
Definition~\ref{def:set-sem} are given in full in
Definition~\ref{def:rel-sem}.

\begin{dfn}\label{def:set-sem}
%Let $\rho$ be a set environment.
The {\em set interpretation} $\setsem{\cdot} : \F \to [\setenv, \set]$
is defined by
\begin{align*}
  \setsem{\Gamma;\Phi \vdash \zerot}\rho &= 0\\
  \setsem{\Gamma;\Phi \vdash \onet}\rho &= 1\\
%  \setsem{\Gamma;\emptyset \vdash v} \rho &= \rho v \mbox{
%    if } v \in \tvars^0 \\ 
  \setsem{\Gamma; \Phi
    %\emptyset
    \vdash \Nat^{\ol{\alpha}}
    \,F\,G}\rho &= \{\eta : \lambda \ol{A}. \,\setsem{\Gamma;
    \ol{\alpha} \vdash
    F}\rho[\ol{\alpha := A}] 
      \Rightarrow \lambda \ol{A}.\,\setsem{\Gamma; \Phi,
        \ol{\alpha} \vdash G}\rho[\ol{\alpha := A}] \\ 
      &\hspace{0.3in}|~\forall \overline{A}, \overline{B} :
      \set. \forall \overline{R : \rel(A, B)}.\\ 
      &\hspace{0.4in}(\eta_{\overline{A}}, \eta_{\overline{B}})
      : \relsem{\Gamma; \ol{\alpha} \vdash F}\Eq_{\rho}[\ol{\alpha := R}]
      \rightarrow \relsem{\Gamma;\Phi, \ol{\alpha} \vdash
        G}\Eq_{\rho}[\ol{\alpha := R}] \} \\
  \setsem{\Gamma;\Phi \vdash \phi\ol{F}}\rho &=
  (\rho\phi)\,\ol{\setsem{\Gamma;\Phi \vdash
    F}\rho}\\
  \setsem{\Gamma;\Phi \vdash F+G}\rho &=
  \setsem{\Gamma;\Phi \vdash F}\rho +
  \setsem{\Gamma;\Phi \vdash G}\rho\\
  \setsem{\Gamma;\Phi \vdash F\times G}\rho &=
  \setsem{\Gamma;\Phi \vdash F}\rho \times
  \setsem{\Gamma;\Phi \vdash G}\rho\\ 
  \setsem{\Gamma;\Phi \vdash (\mu \phi.\lambda
    \ol{\alpha}. H)\ol{G}}\rho &= (\mu
    T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi \vdash G}\rho}\\
    \text{where } T^\set_{H,\rho}\,F & = \lambda
  \ol{A}. \setsem{\Gamma;\Phi,\phi, \ol{\alpha} \vdash
    H}\rho[\phi :=  F][\ol{\alpha := A}]\\
  \text{and } T^\set_{H,\rho}\,\eta &= \lambda
  \ol{A}. \setsem{\Gamma;\Phi,\phi, \ol{\alpha} \vdash
    H}\id_\rho[\phi := \eta][\ol{\alpha := \id_{A}}]
\end{align*}
\end{dfn}
The interpretations in Definition~\ref{def:set-sem} respect weakening,
i.e., a type and its weakenings have the same set interpretations.
The same holds for the actions of these interpretations on morphisms
in Definition~\ref{def:set-sem-funcs} below.
%= \setsem{\Gamma \vdash \sigma} \rho \to
%\setsem{\Gamma \vdash \tau} \rho$, as expected {\color{blue} by IEL}.
If $\rho$ is a set environment and $\vdash F$ then we may
write $\setsem{\vdash F}$ instead of $\setsem{\vdash F}\rho$
since the environment is immaterial.
We note that the third clause of
Definition~\ref{def:set-sem} does indeed define a set: local finite
presentability of $\set$ and $\omega$-cocontinuity of
$\setsem{\Gamma;\ol{\alpha} \vdash F}\rho$ ensure that $\{\eta :
\setsem{\Gamma;\ol{\alpha} \vdash F}\rho \Rightarrow
\setsem{\Gamma;\Phi,\ol{\alpha} \vdash G}\rho\}$ (which contains
$\setsem{\Gamma;\Phi \vdash \Nat^{\ol{\alpha}}\,F\,G}\rho$) is a
subset of $\big\{({\setsem{\Gamma;\Phi,\ol{\alpha} \vdash G}\rho[\ol{\alpha
      := S}]})^{(\setsem{\Gamma;\ol{\alpha} \vdash F}\rho[\ol{\alpha
      := S}])}~\big|~ \ol{S} = (S_1,...,S_{|\ol{\alpha}|}), \mbox{ and
}$ $S_i \mbox{ is a finite set for } i =
1,...,|\ol{\alpha}|\big\}$. There are countably many choices for
tuples $\ol{S}$, and each of these gives rise to a morphism from
${\setsem{\Gamma;\ol{\alpha} \vdash F}\rho[ \ol{\alpha := S}]}$ to
${\setsem{\Gamma;\Phi,\ol{\alpha} \vdash G}\rho[\ol{\alpha := S}]}$. But
there are only $\set$-many choices of morphisms between any
%these (or any)
two objects since $\set$ is locally small. Finally, interpretations of
$\Nat$ types in Definition~\ref{def:set-sem} ensure $\setsem{\Gamma
  \vdash F \to G}$ and $\setsem{\Gamma \vdash \forall
  \alpha. F}$ are as expected in any parametric model.

In order to make sense of the last clause in
Definition~\ref{def:set-sem}, we need to know that, for each $\rho \in
\setenv$, $T^\set_{H,\rho}$ is an $\omega$-cocontinuous endofunctor on
$[\set^k, \set]$, and thus admits a fixpoint.  Since
$T_{H,\rho}^\set$ is defined in terms of $\setsem{\Gamma;\Phi,\phi,
  \ol{\alpha} \vdash H}$, this means that interpretations of types
must be such functors, which in turn means that the actions of set
interpretations of types on objects and on morphisms in $\setenv$ are
intertwined. Fortunately, we know from~\cite{jp19} that, for every
$\Gamma; \ol{\alpha} \vdash G$, $\setsem{\Gamma; \ol{\alpha} \vdash
  G}$ is actually in $[\set^k,\set]$ where $k = |\ol \alpha|$. This
means that for each $\setsem{\Gamma; \Phi, \phi^k, \ol{\alpha} \vdash
  H}$, the corresponding operator $T^\set_{H}$ can be extended to a
{\em functor} from $\setenv$ to $[[\set^k,\set],[\set^k,\set]]$. The
action of $T^\set_H$ on an object $\rho \in \setenv$ is given by the
higher-order functor $T_{H,\rho}^\set$, whose actions on objects
(functors in $[\set^k, \set]$) and morphisms (natural transformations
between such functors) are given in Definition~\ref{def:set-sem}. Its
action on a morphism $f : \rho \to \rho'$ is the higher-order natural
transformation $T^\set_{H,f} : T^\set_{H,\rho} \to T^\set_{H,\rho'}$
whose action on $F : [\set^k,\set]$ is the natural transformation
$T^\set_{H,f}\, F : T^\set_{H,\rho}\,F \to T^\set_{H,\rho'}\,F$ whose
component at $\ol{A}$ is $(T^\set_{H,f}\, F)_{\ol{A}} =
\setsem{\Gamma; \Phi,\phi,\ol{\alpha} \vdash H}f[\phi :=
  \id_F][\ol{\alpha := \id_A}]$. The next definition uses the functor
$T^\set_H$ to define the actions of functors interpreting types on
morphisms between set environments.

\begin{dfn}\label{def:set-sem-funcs}
Let $f: \rho \to \rho'$ for set environments $\rho$ and $\rho'$ (so
that $\rho|_\tvars = \rho'|_\tvars$). The action $\setsem{\Gamma;\Phi
  \vdash F}f$ of\, $\setsem{\Gamma;\Phi \vdash F}$ on the
morphism $f$ is given as follows:
\begin{itemize}
\item If \,$\Gamma;\Phi \vdash \zerot$ then $\setsem{\Gamma;\Phi \vdash
  \zerot}f = \id_0$
\item If \,$\Gamma;\Phi \vdash \onet$ then $\setsem{\Gamma;\Phi \vdash
  \onet}f = \id_1$
%\item If \,$\Gamma,v;\emptyset \vdash v$ then
%  $\setsem{\Gamma,v;\emptyset \vdash v}f = \id_{\rho v}$
\item If \,$\Gamma; \Phi
  %\emptyset
  \vdash \Nat^{\ol{\alpha}}\,F\,G$ then
  $\setsem{\Gamma; \Phi
    %\emptyset
    \vdash \Nat^{\ol{\alpha}}\,F\,G} f =$\\
%  \id_{\setsem{\Gamma; \Phi
      %\emptyset
%      \vdash \Nat^{\ol{\alpha}}\,F\,G}\rho}$
\hspace*{0.2in} $\lambda t : \setsem{\Gamma;\Phi \vdash
    \Nat^{\ol{\alpha}}\,F\,G}\rho.\, 
  (\lambda \ol{A}. \setsem{\Gamma; \Phi,\ol\alpha \vdash G}f[\ol{\alpha
      := id_A}]) \circ t$  
\item If \,$\Gamma;\Phi \vdash \phi \ol{F}$ then
$\setsem{\Gamma;\Phi \vdash \phi \ol{F}} f : \setsem{\Gamma;\Phi
  \vdash \phi \ol{F}}\rho \to \setsem{\Gamma;\Phi \vdash
  \phi\ol{F}}\rho' = (\rho\phi) \ol{\setsem{\Gamma;\Phi \vdash
    F}\rho} \to (\rho'\phi) \ol{\setsem{\Gamma;\Phi \vdash
    F}\rho'}$ is defined by $\setsem{\Gamma;\Phi \vdash \phi
  \ol{F}} f = (f\phi)_{\ol{\setsem{\Gamma;\Phi \vdash
      F}\rho'}}\, \circ\, (\rho\phi) {\ol{\setsem{\Gamma;\Phi
      \vdash F}f}} = (\rho'\phi) {\ol{\setsem{\Gamma;\Phi \vdash
      F}f}}\, \circ\, (f \phi)_{\ol{\setsem{\Gamma;\Phi \vdash
      \F}\rho}}$.  The latter equality holds because $\rho\phi$ and
  $\rho'\phi$ are functors and $f\phi : \rho\phi \to \rho'\phi$ is a
  natural transformation, so the following naturality square commutes:
{\footnotesize\begin{equation}\label{eq:cd2}
\begin{CD}
  (\rho\phi) \ol{\setsem{\Gamma;\Phi \vdash F}\rho} @> (f\phi)_{
    \ol{\setsem{\Gamma;\Phi \vdash F}\rho}} >> (\rho'\phi)
  \ol{\setsem{\Gamma;\Phi \vdash F}\rho} \\ @V(\rho\phi)
  \ol{\setsem{\Gamma;\Phi \vdash F}f}VV @V (\rho'\phi)
  \ol{\setsem{\Gamma;\Phi \vdash F}f} VV \\ (\rho\phi)
  \ol{\setsem{\Gamma;\Phi \vdash F}\rho'} @>(f\phi)_{
    \ol{\setsem{\Gamma;\Phi \vdash F}\rho'}}>> (\rho'\phi)
  \ol{\setsem{\Gamma;\Phi \vdash F}\rho'}
\end{CD}
\end{equation}}
\item If\, $\Gamma;\Phi \vdash F + G$ then $\setsem{\Gamma;\Phi
  \vdash F + G}f$ is defined by $\setsem{\Gamma;\Phi \vdash
  F + G}f(\inl\,x) = \inl\,(\setsem{\Gamma;\Phi \vdash
  F}f x)$ and $\setsem{\Gamma;\Phi \vdash F +
  G}f(\inr\,y) = \inr\,(\setsem{\Gamma;\Phi \vdash G}f y)$
\item If \,$\Gamma;\Phi\vdash F \times G$ then
  $\setsem{\Gamma;\Phi \vdash F \times G}f = 
  \setsem{\Gamma;\Phi \vdash F}f \times \setsem{\Gamma;\Phi \vdash
    G}f$
\item If \,$\Gamma;\Phi \vdash (\mu \phi.\lambda
  \ol{\alpha}. H)\ol{G}$ then
  \[\begin{array}{lll}
  \setsem{\Gamma;\Phi \vdash (\mu  \phi.\lambda
    \ol{\alpha}. H)\ol{G}} f &: &\setsem{\Gamma;\Phi 
    \vdash (\mu \phi.\lambda \ol{\alpha}. H)\ol{G}} \rho \to
  \setsem{\Gamma;\Phi \vdash (\mu
    \phi.\lambda\ol{\alpha}. H)\ol{G}} \rho'\\
  &= &(\mu
  T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi \vdash G}\rho} \to (\mu
  T^\set_{H,\rho'})\ol{\setsem{\Gamma;\Phi \vdash G}\rho'}
  \end{array}\]
  is
  defined by
  \[\begin{array}{ll} & (\mu T^\set_{H,f})\ol{\setsem{\Gamma;\Phi \vdash
      G}\rho'} \circ (\mu T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi
      \vdash G}f}\\ = & (\mu T^\set_{H,\rho'})\ol{\setsem{\Gamma;\Phi
      \vdash G}f} \circ (\mu T^\set_{H,f})\ol{\setsem{\Gamma;\Phi
      \vdash G}\rho}\end{array}\]  The latter equality holds because $\mu
  T^\set_{H,\rho}$ and $\mu T^\set_{H,\rho'}$ are functors and $\mu
  T_{H,f}^\set : \mu T_{H,\rho}^\set \to \mu T_{H,\rho'}^\set$ is a
  natural transformation, so the following naturality square commutes:
{\footnotesize\begin{equation}\label{eq:cd3}
\begin{CD}
 (\mu T^\set_{H,\rho}) \ol{\setsem{\Gamma;\Phi \vdash G}\rho} @> (\mu
  T^\set_{H,f})_{\ol{\setsem{\Gamma;\Phi \vdash G}\rho}} >> (\mu
  T^\set_{H,\rho'}) \ol{\setsem{\Gamma;\Phi \vdash G}\rho} \\ 
 @V(\mu T^\set_{H,\rho}) \ol{\setsem{\Gamma;\Phi \vdash G}f}VV @V  (\mu
 T^\set_{H,\rho'}) \ol{\setsem{\Gamma;\Phi \vdash G}f} VV \\ 
(\mu T^\set_{H,\rho}) \ol{\setsem{\Gamma;\Phi \vdash G}\rho'} @>(\mu
 T^\set_{H,f})_{\ol{\setsem{\Gamma;\Phi \vdash G}\rho'}}>> (\mu
 T^\set_{H,\rho'}) \ol{\setsem{\Gamma;\Phi \vdash G}\rho'}
\end{CD}
\end{equation}}
\end{itemize}
\end{dfn}

\subsection{Interpreting Types as Relations}\label{sec:rel-interp}

\begin{dfn}\label{def:rel-transf}
A {\em $k$-ary relation transformer} $F$ is a triple $(F^1, F^2,F^*)$,
where $F^1,F^2 : [\set^k,\set]$ are functors, $F^* : [\rel^k, \rel]$
is a functor, if $R_1:\rel(A_1,B_1),...,R_k:\rel(A_k,B_k)$, then $F^*
\ol{R} : \rel(F^1 \ol{A}, F^2 \ol{B})$, and if $(\alpha_1, \beta_1)
\in \Homrel(R_1,S_1),..., (\alpha_k, \beta_k) \in \Homrel(R_k,S_k)$
then $F^* \ol{(\alpha, \beta)} = (F^1 \ol{\alpha}, F^2 \ol{\beta})$.
We define $F\ol{R}$ to be $F^*\overline{R}$ and
$F\overline{(\alpha,\beta)}$ to be $F^*\overline{(\alpha,\beta)}$.
\end{dfn}
The last clause of Definition~\ref{def:rel-transf} expands to: if
$\ol{(a,b) \in R}$ implies $\ol{(\alpha\,a,\beta\,b) \in S}$ then
$(c,d) \in F^*\ol{R}$ implies $(F^1 \ol{\alpha}\,c,F^2 \ol{\beta}\,d)
\in F^*\ol{S}$. When convenient we identify a $0$-ary relation
transformer $(A,B,R)$ with $R : \rel(A,B)$. We may also write $\pi_1
F$ for $F^1$ and $\pi_2 F$ for $F^2$. We extend these conventions to
relation environments, introduced in Definition~\ref{def:reln-env}
below, in the obvious way.

\begin{dfn}
The category $RT_k$ of $k$-ary relation transformers is given by the
following data:
\begin{itemize}
\item An object of $RT_k$ is a relation transformer.
\item A morphism $\delta : (G^1,G^2,G^*) \to (H^1,H^2,H^*)$ in $RT_k$
  is a pair of natural transformations $(\delta^1, \delta^2)$ where
  $\delta^1 : G^1 \to H^1$, $\delta^2 : G^2 \to H^2$ such that, for
  all $\ol{R : \rel(A, B)}$, if $(x, y) \in G^*\ol{R}$ then
  $(\delta^1_{\ol{A}}x, \delta^2_{\ol{B}}y) \in H^*\ol{R}$.
\item Identity morphisms and composition are inherited from the
  category of functors on $\set$.
\end{itemize}
\end{dfn}

\begin{dfn}\label{def:RT-functor}
An endofunctor $H$ on $RT_k$ is a triple $H = (H^1,H^2,H^*)$, where
\begin{itemize}
\item $H^1$ and $H^2$ are functors from $[\set^k,\set]$ to $[\set^k,\set]$
\item $H^*$ is a functor from $RT_k$ to $[\rel^k,\rel]$
\item for all $\overline{R : \rel(A,B)}$,
  $\pi_1((H^*(\delta^1,\delta^2))_{\overline{R}}) = (H^1
  \delta^1)_{\overline{A}}$ and
  $\pi_2((H^*(\delta^1,\delta^2))_{\overline{R}}) = (H^2
  \delta^2)_{\overline{B}}$
\item The action of $H$ on objects is given by $H\,(F^1,F^2,F^*) =
  (H^1F^1,\,H^2F^2,\,H^*(F^1,F^2,F^*))$
\item The action of $H$ on morphisms is given by
  $H\,(\delta^1,\delta^2) = (H^1\delta^1,H^2\delta^2)$ for
  $(\delta^1,\delta^2) : (F^1,F^2,F^*)\to (G^1,G^2,G^*)$
\end{itemize}
\end{dfn}
Since the results of applying an endofunctor $H$ to $k$-ary relation
transformers and morphisms between them must again be $k$-ary relation
transformers and morphisms between them, respectively,
Definition~\ref{def:RT-functor} implicitly requires that the following
three conditions hold:\,{\em i})
if $R_1:\rel(A_1,B_1),...,R_k:\rel(A_k,B_k)$, then
  $H^*(F^1,F^2,F^*) \ol{R} : \rel(H^1F^1 \ol{A}, H^2F^2 
  \ol{B})$;
  %In other words, $\pi_1 (H^*(F^1,F^2,F^*) \ol{R}) = H^1F^1
  %\ol{A}$ and $\pi_2 (H^*(F^1,F^2,F^*) \ol{R}) = H^2F^2 \ol{B}$.
{\em ii}) if $(\alpha_1, \beta_1) \in \Homrel(R_1,S_1),..., (\alpha_k,
  \beta_k) \in \Homrel(R_k,S_k)$, then
  $H^*(F^1,F^2,F^*)\, \ol{(\alpha, \beta)} = (H^1F^1\ol{\alpha}, H^2F^2
  \ol{\beta})$; and
  %In other words, $\pi_1 (H^*(F^1,F^2,F^*) \ol{(\alpha,
  %  \beta)}) = H^1F^1 \ol{\alpha}$ and $\pi_2 (H^*(F^1,F^2,F^*) 
  % \ol{(\alpha, \beta)}) = H^2F^2 \ol{\beta}$.  {\em iii}) if
  $(\delta^1,\delta^2) : (F^1,F^2,F^*)\to (G^1,G^2,G^*)$ and
  $R_1:\rel(A_1,B_1),...,R_k:\rel(A_k,B_k)$, then
  $((H^1\delta^1)_{\ol{A}}x, (H^2\delta^2)_{\ol{B}}y) \in
  H^*(G^1,G^2,G^*)\ol{R}$ whenever $(x, y) \in
  H^*(F^1,F^2,F^*)\ol{R}$. Note, however, that this last condition is
  automatically satisfied because it is implied by the third bullet
  point of Definition~\ref{def:RT-functor}.

\begin{dfn}\label{def:RT-nat-trans}
If $H$ and $K$ are endofunctors on $RT_k$, then a {\em natural
  transformation} $\sigma : H \to K$ is a pair $\sigma = (\sigma^1,
\sigma^2)$, where $\sigma^1 : H^1 \to K^1$ and $\sigma^2 : H^2 \to
K^2$ are natural transformations between endofunctors on
$[\set^k,\set]$ and the component of $\sigma$ at $F
%= (F^1,F^2,F^*)
\in RT_k$ is given by $\sigma_F = (\sigma^1_{F^1}, \sigma^2_{F^2})$.
\end{dfn}
Definition~\ref{def:RT-nat-trans} entails that $\sigma^i_{F^i}$ must
be natural in $F^i : [\set^k,\set]$, and, for every $F$, both
$(\sigma^1_{F^1})_{\overline{A}}$ and
$(\sigma^2_{F^2})_{\overline{A}}$ must be natural in $\overline{A}$.
Moreover, since the results of applying $\sigma$ to $k$-ary relation
transformers must be morphisms of $k$-ary relation transformers,
Definition~\ref{def:RT-nat-trans} implicitly requires that
$(\sigma_F)_{\overline{R}} = ( (\sigma^1_{F^1})_{\overline{A}},
(\sigma^2_{F^2})_{\overline{B}})$ is a morphism in $\rel$ for any
$k$-tuple of relations $\overline{R : \rel(A, B)}$, i.e., that if $(x,
y) \in H^*F\overline{R}$, then $((\sigma^1_{F^1})_{\overline{A}} x,
(\sigma^2_{F^2})_{\overline{B}} y) \in K^*F\overline{R}$.

\vspace*{0.1in}

%{\color{blue} More context? Weave cocontinuity requirement for fixed
%  points through text better.}
Critically, we can compute $\omega$-directed colimits in $RT_k$: it is
not hard to see that if $\cal D$ is an $\omega$-directed set then
$\colim{d \in {\cal D}}{(F^1_d, F^2_d,F^*_d)} = (\colim{d \in {\cal
    D}}{F^1_d}, \colim{d \in {\cal D}}{F^2_d}, \colim{d \in {\cal
    D}}{F^*_d})$.  We define an endofunctor $T = (T^1,T^2,T^*)$ on
$RT_k$ to be {\em $\omega$-cocontinuous} if $T^1$ and $T^2$ are
$\omega$-cocontinuous endofunctors on $[\set^k,\set]$ and $T^*$ is an
$\omega$-cocontinuous functor from $RT_k$ to $[\rel^k,\rel]$, i.e., is
in $[RT_k,[\rel^k,\rel]]$.
%\begin{lemma}\label{lem:colimits}
%$\colim{d \in {\cal D}}{(F^1_d, F^2_d,F^*_d)} = (\colim{d \in {\cal
%      D}}{F^1_d}, \colim{d \in {\cal D}}{F^2_d}, \colim{d \in {\cal
%      D}}{F^*_d})$
%\end{lemma}
%\begin{proof}
%We first observe that $(\colim{d \in {\cal D}}{F^1_d}, \colim{d \in
%  {\cal D}}{F^2_d}, \colim{d \in {\cal D}}{F^*_d})$ is in $RT_k$.  If
%$R_1:\rel(A_1,B_1),...,R_k:\rel(A_k,B_k)$, then $\colim{d \in \cal
%  D}{F^*_d\ol{R}} : \rel(\colim{d \in {\cal D}}{F^1_d\ol{A}},
%\,\colim{d \in {\cal D}}{F^2_d\ol{B}})$ because of how colimits are
%computed in $\rel$. Moreover, if $(\alpha_1, \beta_1) \in
%\Homrel(R_1,S_1),..., (\alpha_k, \beta_k) \in \Homrel(R_k,S_k)$, then
%\[\begin{array}{ll}
%  & (\colim{d \in \cal D}{F_d^*})\ol{(\alpha,\beta)}\\
%= & \colim{d \in \cal D}{F_d^*\ol{(\alpha,\beta)}}\\
%= & \colim{d \in \cal D}{(F_d^1\ol{\alpha},\, F_d^2\ol{\beta})}\\
%= & (\colim{d \in \cal D}{F^1_d \ol{\alpha}}, \,\colim{d \in \cal D}{
%  F^2_d \ol{\beta}})
%\end{array}\]
%so $(\colim{d \in {\cal D}}{F^1_d}, \colim{d \in {\cal D}}{F^2_d},
%\colim{d \in {\cal D}}{F^*_d})$ actually is in $RT_k$.
%
%Now to see that $\colim{d \in {\cal D}}{(F^1_d, F^2_d,F^*_d)} =
%(\colim{d \in {\cal D}}{F^1_d}, \colim{d \in {\cal D}}{F^2_d},
%\colim{d \in {\cal D}}{F^*_d})$, let $\gamma^1_d : F^1_d \to \colim{d
%  \in {\cal D}}{F^1_d}$ and $\gamma^2_d : F^2_d \to \colim{d \in {\cal
%    D}}{F^2_d}$ be the injections for the colimits $\colim{d \in {\cal
%    D}}{F^1_d}$ and $\colim{d \in {\cal D}}{F^2_d}$,
%respectively. Then $(\gamma^1_d, \gamma^2_d) : (F^1_d, F^2_d,F^*_d)
%\to \colim{d \in {\cal D}}{(F^1_d, F^2_d,F^*_d)}$ is a morphism in
%$RT_k$ because, for all $\ol{R : \rel(A, B)}$,
%$((\gamma^1_d)_{\ol{A}}, (\gamma^2_d)_{\ol{B}}) : F^*_d \ol{R} \to
%\colim{d \in {\cal D}}{F^*_d \ol{R}}$ is a morphism in $\rel$. So
%$\{(\gamma^1_d, \gamma^2_d)\}_{d \in {\cal D}}$ are the mediating
%morphisms of a cocone in $RT_k$ with vertex $\colim{d \in {\cal
%    D}}{(F^1_d, F^2_d,F^*_d)}$. To see that this cocone is a
%colimiting cocone, let $C = (C^1,C^2,C^*)$ be the vertex of a cocone
%for $\{(F^1_d,F^2_d,F^*_d)\}_{d \in {\cal D}}$ with injections
%$(\delta^1_d,\delta^2_d) : (F^1_d,F^2_d,F^*_d) \to C$. If $\eta^1 :
%\colim{d \in {\cal D}}{F^1_d} \to C^1$ and $\eta^2 : \colim{d \in
%  {\cal D}}{F^2_d} \to C^2$ are the mediating morphisms in
%$[\set^k,\set]$, then $\eta^1$ and $\eta^2$ are unique such that
%$\delta^1_d = \eta^1 \circ \gamma^1_d$ and $\delta^2_d = \eta^2 \circ
%\gamma^2_d$.  We therefore have that $(\eta^1,\eta^2) : \colim{d \in
%  {\cal D}}{(F^1_d, F^2_d,F^*_d)} \to C$ is the mediating morphism in
%$RT_k$. Indeed, for all $\ol{R : \rel(A,B)}$ and $(x,y) \in \colim{d
%  \in {\cal D}}{F^*_d \ol{R}}$, there exist $d$ and $(x',y') \in
%F^*_d\ol{R}$ such that $(\gamma^1_d)_{\ol{A}}x' = x$ and
%$(\gamma^2_d)_{\ol{B}}y' = y$. But then $(\eta^1_{\ol{A}}x,
%\eta^2_{\ol{B}}y) = (\eta^1_{\ol{A}}((\gamma^1_d)_{\ol{A}}x'),
%\eta^2_{\ol{B}}((\gamma^2_d)_{\ol{B}}y')) = ((\delta^1_d)_{\ol{A}}x',
%(\delta^2_d)_{\ol{B}}y')$, and this pair is in $C^*\ol{R}$ because
%$(\delta^1_d, \delta^2_d)$ is a morphism from $(F^1_d,F^2_d,F^*_d)$ to
%$C$ in $RT_k$.
%\end{proof}
\begin{comment}
\begin{dfn}\label{def:omega-cocont}
An endofunctor $T = (T^1,T^2,T^*)$ on $RT_k$ is {\em
$\omega$-cocontinuous} if $T^1$ and $T^2$ are $\omega$-cocontinuous
endofunctors on $[\set^k,\set]$ and $T^*$ is an $\omega$-cocontinuous
functor from $RT_k$ to $[\rel^k,\rel]$, i.e., is in
$[RT_k,[\rel^k,\rel]]$.
\end{dfn}
\end{comment}
Now, for any $k$, any $A : \set$, and any $R : \rel(A, B)$, let
$K^\set_A$ be the constantly $A$-valued functor from $\set^k$ to
$\set$ and $K^\rel_R$ be the constantly $R$-valued functor from
$\rel^k$ to $\rel$.  Also let $0$ denote either the initial object of
$\set$ or the initial object of $\rel$, as appropriate.  Observing
that, for every $k$, $K^\set_0$ is initial in $[\set^k,\set]$, and
$K^\rel_0$ is initial in $[\rel^k,\rel]$, we have that, for each $k$,
$K_0 = (K^\set_0,K^\set_0,K^\rel_0)$ is initial in $RT_k$. Thus, if $T
= (T^1,T^2,T^*) : RT_k \to RT_k$ is an endofunctor on $RT_k$ then we
can define the relation transformer $\mu T$ to be $\colim{n \in
  \nat}{T^n K_0}$.
%Then Lemma~\ref{lem:colimits} shows $\mu T$ is indeed a relation
%transformer, and that it is given explicitly by
It is not hard to see that $\mu T$ is given explicitly as 
\begin{equation}\label{eq:mu}
  %\colim{n \in \nat}{T^n K_0}
  \mu T = (\mu T^1,\mu T^2,
\colim{n \in \nat}{(T^nK_0)^*})
\end{equation}
and that, as our notation suggests, it really is a fixpoint for $T$ if
$T$ is $\omega$-cocontinuous:
\begin{lemma}\label{lem:fp}
For any $T : [RT_k,RT_k]$, $\mu T \cong T(\mu T)$.
\end{lemma}
\noindent
The isomorphism is given by the morphisms $(\mathit{in}_1,
\mathit{in}_2) : T(\mu T) \to \mu T$ and $(in_1^{-1}, in_2^{-1}) : \mu
T \to T(\mu T)$ in $RT_k$. The latter is always a morphism in $RT_k$,
but the former need not be if $T$ is not $\omega$-cocontinuous.

\begin{comment}
\begin{proof}
We have $T(\mu T) = T(\colim{n \in \nat}{(T^nK_0)}) \cong \colim{n \in
\nat}{T(T^nK_0)} =
\mu T$.
\end{proof}
In fact, the isomorphism in Lemma~\ref{lem:fp} is given by the
morphisms $(in_1, in_2) : T(\mu T) \to \mu T$ and $(in_1^{-1},
in_2^{-1}) : \mu T \to T(\mu T)$ in $RT_k$. It is worth noting that the
latter is always a morphism in $RT_k$, but the former isn't necessarily
a morphism in $RT_k$ unless $T$ is $\omega$-cocontinuous.
\end{comment}

It is worth noting that the third component in Equation~(\ref{eq:mu})
is the colimit in $[\rel^k,\rel]$ of third components of relation
transformers, rather than a fixpoint of an endofunctor on
$[\rel^k,\rel]$. That there is an asymmetry between the first two
components of $\mu T$ and its third reflects the important conceptual
observation that the third component of an endofunctor on $RT_k$ need
not be a functor on all of $[\rel^k,\rel]$. In particular, although we
can define $T_{H,\rho}\, F$ for a relation transformer $F$ in
Definition~\ref{def:rel-sem} below, it is not clear how we could
define it for $F : [\rel^k,\rel]$.

\begin{dfn}\label{def:reln-env}
A {\em relation environment} maps each each type variable in $\tvars^k
\cup \fvars^k$ to a $k$-ary relation transformer.  A morphism $f :
\rho \to \rho'$ for relation environments $\rho$ and $\rho'$ with
$\rho|_\tvars = \rho'|_\tvars$ maps each type constructor variable
$\psi^k \in \tvars$ to the identity morphism on $\rho \psi^k = \rho'
\psi^k$ and each functorial variable $\phi^k \in \fvars$ to a morphism
from the $k$-ary relation transformer $\rho \phi$ to the $k$-ary
relation transformer $\rho' \phi$. Composition of morphisms on
relation environments is given componentwise, with the identity
morphism mapping each relation environment to itself. This gives a
category of relation environments and morphisms between them, which we
denote $\relenv$.
\end{dfn}
When convenient we identify a $0$-ary relation transformer with the
relation (transformer) that is its codomain and consider
% With this convention,
a relation environment to map a type variable of arity $0$
%to a $0$-ary relation transformer, i.e.,
to a relation.  We write $\rho[\ol{\alpha := R}]$ for the relation
environment $\rho'$ such that $\rho' \alpha_i \, = R_i$ for $i =
1,...,k$ and $\rho' \alpha = \rho\alpha$ if $\alpha \not \in
\{\alpha_1,...,\alpha_k\}$.  If $\rho$ is a relation environment, we
write $\pi_1 \rho$ and $\pi_2 \rho$ for the set environments mapping
each type variable $\phi$ to the functors $(\rho\phi)^1$ and
$(\rho\phi)^2$, respectively.

We define, for each $k$, the notion of an $\omega$-cocontinuous
functor from $\relenv$ to $RT_k$:
\begin{dfn}\label{def:relenv-functor}
A functor $H : [\relenv, RT_k]$ is a triple $H = (H^1,H^2,H^*)$,
where
\begin{itemize}
\item $H^1$ and $H^2$ are objects in $[\setenv,[\set^k,\set]]$
\item $H^*$ is a an object in $[\relenv,[\rel^k,\rel]]$
\item for all $\overline{R : \rel(A,B)}$ and morphisms $f$ in
  $\relenv$, $\pi_1(H^*f \,{\overline{R}}) = H^1 (\pi_1
  f)\,{\overline{A}}$ and $\pi_2(H^*f \,{\overline{R}}) = H^2 (\pi_2
  f)\,{\overline{B}}$
\item The action of $H$ on $\rho$ in $\relenv$ is given by $H \rho = (H^1
  (\pi_1 \rho),\,H^2 (\pi_2 \rho),\,H^*\rho)$
\item The action of $H$ on morphisms $f : \rho \to \rho'$ in $\relenv$
  is given by $Hf = (H^1 (\pi_1 f),H^2 (\pi_2 f))$
\end{itemize}
\end{dfn}
\noindent Spelling out the last two bullet points above gives the
following analogues of the three conditions immediately following
Definition~\ref{def:RT-functor}: {\em i}) if $R_1 :
\rel(A_1,B_1),...,R_k : \rel(A_k,B_k)$, then $H^*\rho\, \ol{R} :
\rel(H^1(\pi_1 \rho)\, \ol{A}, H^2(\pi_2 \rho)\, \ol{B})$;
%In other words, $\pi_1 (H^*\rho \ol{R}) = H^1(\pi_1 \rho)
%  \,\ol{A}$ and $\pi_2 (H^*\rho \ol{R}) = H^2 (\pi_2 \rho) \,\ol{B}$;
{\em ii}) if $(\alpha_1, \beta_1) \in \Homrel(R_1,S_1),..., (\alpha_k,
\beta_k) \in \Homrel(R_k,S_k)$, then $H^*\rho\, \ol{(\alpha, \beta)} =
(H^1(\pi_1 \rho)\,\ol{\alpha}, H^2(\pi_2 \rho)\, \ol{\beta})$;
%  In other words, $\pi_1 (H^*\rho\, \ol{(\alpha, \beta)}) = H^1(\pi_1
%  \rho)\,\ol{\alpha}$ and $\pi_2 (H^*\rho\, \ol{(\alpha, \beta)}) =
%  H^2(\pi_2 \rho)\,\ol{\beta}$;
and {\em iii}) if $f : \rho \to \rho'$ and
$R_1:\rel(A_1,B_1),...,R_k:\rel(A_k,B_k)$, then $(H^1(\pi_1
f)\,{\ol{A}}\,x, H^2(\pi_2 f)\,{\ol{B}}\,y) \in H^*\rho'\,\ol{R}$
whenever $(x, y) \in H^*\rho\,\ol{R}$. As before, the last
condition is automatically satisfied because it is implied by the
third bullet point of Definition~\ref{def:relenv-functor}.

Considering $\relenv$ as a product $\Pi_{\phi^k \in \tvars \cup
  \fvars} RT_k$, we extend the computation of $\omega$-directed
colimits in $RT_k$ to compute colimits in $\relenv$ componentwise. We
similarly extend the notion of an $\omega$-cocontinuous endofunctor on
$RT_k$ componentwise to give a notion of $\omega$-cocontinuity for
functors from $\relenv$ to $RT_k$.  Recalling from the start of this
subsection that Definition~\ref{def:rel-sem} is given mutually
inductively with Definition~\ref{def:set-sem} we can, at last, define:

\begin{dfn}\label{def:rel-sem}
The {\em relational interpretation} $\relsem{\cdot} : \F \to [\relenv,
 \rel]$ is defined by
\begin{align*}
  \relsem{\Gamma;\Phi \vdash \zerot}\rho &= 0\\
  \relsem{\Gamma;\Phi \vdash \onet}\rho &= 1\\
  \end{align*}
  \begin{align*}
  \relsem{\Gamma; \Phi \vdash \Nat^{\ol{\alpha}} \,F\,G}\rho &= \{\eta
  : \lambda \ol{R}.\,\relsem{\Gamma; \ol{\alpha} \vdash
    F}\rho[\ol{\alpha := R}] \Rightarrow \lambda \ol{R}. \,\relsem{
    \Gamma; \Phi,\ol{\alpha} \vdash G}\rho[\ol{\alpha := R}]\}\\
  &=
  \{(t,t') \in \setsem{\Gamma; \Phi
    \vdash \Nat^{\ol{\alpha}}
    \,F\,G} (\pi_1 \rho) \times \setsem{ 
    \Gamma;\Phi
    \vdash \Nat^{\ol{\alpha}} \,F\,G} (\pi_2
  \rho)~|~\\ 
  & \hspace{0.3in} \forall {R_1 : \rel(A_1,B_1)}\,...\,{R_k : \rel(A_k,B_k)}.\\
  & \hspace{0.4in} (t_{\ol{A}},t'_{\ol{B}}) \in
  (\relsem{\Gamma; \Phi,\ol{\alpha} \vdash G}\rho[\ol{\alpha :=
      R}])^{\relsem{\Gamma;\ol{\alpha}\vdash F}\rho[\ol{\alpha := R}]} \}\\  
  \relsem{\Gamma;\Phi \vdash \phi \ol{F}}\rho &=
  (\rho\phi)\ol{\relsem{\Gamma;\Phi \vdash 
    F}\rho}\\
  \relsem{\Gamma;\Phi \vdash F+G}\rho &=
  \relsem{\Gamma;\Phi \vdash F}\rho +
  \relsem{\Gamma;\Phi \vdash G}\rho\\
  \relsem{\Gamma;\Phi \vdash F\times G}\rho &=
  \relsem{\Gamma;\Phi \vdash F}\rho \times
  \relsem{\Gamma;\Phi \vdash G}\rho\\  
   \relsem{\Gamma;\Phi \vdash (\mu \phi.\lambda
    \ol{\alpha}. H)\ol{G}}\rho
  &= (\mu T_{H,\rho})\ol{\relsem{\Gamma;\Phi \vdash
     G}\rho}\\
  \text{where }	T_{H,\rho}
    &= (T^\set_{H,\pi_1\rho}, T^\set_{H,\pi_2\rho}, T^\rel_{H,\rho}) \\
  \text{and } T^\rel_{H,\rho}\,F
    &= \lambda \ol{R}. \relsem{
      \Gamma;\Phi,\phi,\ol{\alpha} \vdash H}\rho[\phi :=
    F][\ol{\alpha := R}]\\
  \text{and } T^\rel_{H,\rho}\,\delta
    &= \lambda \ol{R}. \relsem{
      \Gamma;\Phi,\phi,\ol{\alpha} \vdash H}\id_\rho[\phi :=
    \delta][\ol{\alpha := \id_{\ol{R}}}]
\end{align*}
\end{dfn}

The interpretations in Definition~\ref{def:rel-sem}, as well as in
Definition~\ref{def:rel-sem-funcs} below, respect weakening, and also
ensure that $\relsem{\Gamma \vdash F \to G}$ and $\relsem{\Gamma
  \vdash \forall \alpha. F}$ are as expected in any parametric model.
%$\relsem{\Gamma \vdash
%  \sigma \to \tau}\rho = \relsem{\Gamma \vdash \sigma} \rho \to
%\relsem{\Gamma \vdash \tau} \rho$.
If $\rho$ is a relational environment and $\vdash F$, then we write
$\relsem{\vdash F}$ instead of $\relsem{\vdash F}\rho$ as for set
interpretations.  For the last clause in Definition~\ref{def:rel-sem}
to be well-defined we need $T_{H,\rho}$ to be an $\omega$-cocontinuous
endofunctor on $RT$ so that, by Lemma~\ref{lem:fp}, it admits a
fixpoint. Since $T_{H,\rho}$ is defined in terms of
$\relsem{\Gamma;\Phi,\phi^k, \ol{\alpha} \vdash H}$, this means that
relational interpretations of types must be $\omega$-cocontinuous
functors from $\relenv$ to $RT_0$, which in turn entails that the
actions of relational interpretations of types on objects and on
morphisms in $\relenv$ are intertwined. As for set interpretations, we
know from~\cite{jp19} that, for every $\Gamma; \ol{\alpha} \vdash F$,
$\setsem{\Gamma; \ol{\alpha} \vdash F}$ is actually in
$[\rel^k,\rel]$ where $k = |\ol \alpha|$.
%In fact, we already know from~\cite{jp19} that, for every $\Gamma;
%\ol{\alpha} \vdash \tau : \F$, $\relsem{\Gamma; \ol{\alpha} \vdash
%  \tau}$ is actually functorial in $\ol{\alpha}$ and
%$\omega$-cocontinuous.
We first define the actions of each of these functors on morphisms
between environments in Definition~\ref{def:rel-sem-funcs}, and then
argue that the functors given by Definitions~\ref{def:rel-sem}
and~\ref{def:rel-sem-funcs} are well-defined and have the required
properties. To do this, we extend $T_H$ to a {\em functor} from
$\relenv$ to $[[\rel^k,\rel],[\rel^k,\rel]]$. Its action on an object
$\rho \in \relenv$ is given by the higher-order functor $T_{H,\rho}$
whose actions on objects and morphisms are given in
Definition~\ref{def:rel-sem-funcs}. Its action on a morphism $f : \rho
\to \rho'$ is the higher-order natural transformation $T_{H,f} :
T_{H,\rho} \to T_{H,\rho'}$ whose action on any $F : [\rel^k,\rel]$ is
the natural transformation $T_{H,f}\, F : T_{H,\rho}\, F \to
T_{H,\rho'}\, F$ whose component at $\ol{R}$ is $(T_{H,f}\,
F)_{\ol{R}} = \relsem{\Gamma; \Phi,\phi,\ol{\alpha} \vdash H}f[\phi :=
  \id_F][\ol{\alpha := \id_R}]$.  The next definition uses the functor
$T_H$ to define the actions of functors interpreting types on
morphisms between relation environments.

\begin{dfn}\label{def:rel-sem-funcs}
Let $f: \rho \to \rho'$ for relation environments $\rho$ and $\rho'$
(so that $\rho|_\tvars = \rho'|_\tvars$). The action
$\relsem{\Gamma;\Phi \vdash F}f$ of $\relsem{\Gamma;\Phi \vdash F}$ on
the morphism $f$ is given exactly as in
Definition~\ref{def:set-sem-funcs}, except that all interpretations
are relational interpretations and all occurrences of $T^\set_{H,f}$
are replaced by $T_{H,f}$.
\begin{comment}
as follows:
\begin{itemize}
\item If \,$\Gamma;\Phi \vdash \zerot$ then $\relsem{\Gamma;\Phi \vdash
  \zerot}f = \id_0$
\item If\, $\Gamma;\Phi \vdash \onet$ then $\relsem{\Gamma;\Phi \vdash
  \onet}f = \id_1$
\item If \,$\Gamma; \Phi
  \vdash \Nat^{\ol{\alpha}}\,F\,G$ then
  $\relsem{\Gamma; \Phi
    \vdash \Nat^{\ol{\alpha}}\,F\,G} f =$\\
\hspace*{0.2in} $\lambda t : \relsem{\Gamma;\Phi \vdash
    \Nat^{\ol{\alpha}}\,F\,G}\rho.\, 
  (\lambda \ol{A}. \relsem{\Gamma; \Phi,\ol\alpha \vdash G}f[\ol{\alpha
      := id_A}]) \circ t$  
\item If \,$\Gamma;\Phi \vdash \phi \ol{\tau}$, then
  $\relsem{\Gamma;\Phi \vdash \phi \ol{\tau}} f : \relsem{\Gamma;\Phi
  \vdash \phi \ol{\tau}}\rho \to \relsem{\Gamma;\Phi \vdash \phi
  \ol{\tau}}\rho' = (\rho\phi) \ol{\relsem{\Gamma;\Phi \vdash
    \tau}\rho} \to (\rho'\phi) \ol{\relsem{\Gamma;\Phi \vdash
    \tau}\rho'}$ is defined by $\relsem{\Gamma;\Phi \vdash \phi
  \tau{A}} f = (f\phi)_{\ol{\relsem{\Gamma;\Phi \vdash \tau}\rho'}}
  \,\circ\, (\rho\phi) \ol{\relsem{\Gamma;\Phi \vdash \tau}f} =
  (\rho'\phi) \ol{\relsem{\Gamma;\Phi \vdash \tau}f} \,\circ\, (f
  \phi)_{\ol{\relsem{\Gamma;\Phi \vdash \tau}\rho}}$
\item If\, $\Gamma;\Phi\vdash \sigma + \tau$ then $\relsem{\Gamma;\Phi
  \vdash \sigma + \tau}f$ is defined by $\relsem{\Gamma;\Phi \vdash
  \sigma + \tau}f(\inl\,x) = \inl\,(\relsem{\Gamma;\Phi \vdash
  \sigma}f x)$ and $\relsem{\Gamma;\Phi \vdash \sigma +
  \tau}f(\inr\,y) = \inr\,(\relsem{\Gamma;\Phi \vdash \tau}f y)$
\item If\, $\Gamma;\Phi\vdash \sigma \times \tau$ then
  $\relsem{\Gamma;\Phi \vdash \sigma \times \tau}f =
  \relsem{\Gamma;\Phi \vdash \sigma}f \times \relsem{\Gamma;\Phi
    \vdash \tau}f$
\item If\, $\Gamma;\Phi \vdash (\mu \phi^k.\lambda
  \ol{\alpha}. H)\ol{\tau}$ then $\relsem{\Gamma;\Phi \vdash (\mu
    \phi.\lambda \ol{\alpha}. H)\ol{\tau}} f = (\mu
  T_{H,f})\ol{\relsem{\Gamma;\Phi \vdash \tau}\rho'} \circ (\mu
  T_{H,\rho})\ol{\relsem{\Gamma;\Phi \vdash \tau}f} = (\mu
  T_{H,\rho'})\ol{\relsem{\Gamma;\Phi \vdash \tau}f} \circ (\mu
  T_{H,f})\ol{\relsem{\Gamma;\Phi \vdash \tau}\rho}$
\end{itemize}
\end{comment}
\end{dfn}

To see that the functors given by Definitions~\ref{def:rel-sem}
and~\ref{def:rel-sem-funcs} are well-defined we must show that, for
every $H$, $T_{H,\rho}\,F$ is a relation transformer for any relation
transformer $F$, and that $T_{H,f}\, F : T_{H,\rho}\, F \to
T_{H,\rho'}\, F$ is a morphism of relation transformers for every
relation transformer $F$ and every morphism $f : \rho \to \rho'$ in
$\relenv$. This is an immediate consequence of 
\begin{lemma}\label{lem:rel-transf-morph}
%The interpretations in Definitions~\ref{def:rel-sem}
%and~\ref{def:rel-sem-funcs} are well-defined and,
For every $\Gamma;\Phi \vdash F$, $\sem{\Gamma;\Phi \vdash F} =
(\setsem{\Gamma;\Phi \vdash F}, \setsem{\Gamma;\Phi \vdash
  F},\relsem{\Gamma;\Phi \vdash F}) \in
%is an $\omega$-cocontinuous functor from $\relenv$ to $RT_0$, i.e., is an
%element of
[\relenv,RT_0]$.
\end{lemma}
\noindent
The proof is a straightforward induction on the structure of $F$,
%%each case of which first shows that $(\setsem{\Gamma;\Phi \vdash
%  \tau}, \setsem{\Gamma;\Phi \vdash \tau},\relsem{\Gamma;\Phi \vdash
%  \tau})$ is indeed a functor from $\relenv$ to $RT_0$, and then uses
using an appropriate result from~\cite{jp19} to deduce
$\omega$-cocontinuity of $\sem{\Gamma;\Phi \vdash F}$ in each case,
together with Lemma~\ref{lem:fp} and Equation~\ref{eq:mu} in the
$\mu$-case.

\begin{comment}
\begin{proof}
By induction on the structure of $\tau$. The only interesting cases
are when $\tau = \phi\ol{\tau}$ and when $\tau = (\mu \phi^k. \lambda
\overline{\alpha}. H)\overline{\tau}$. We consider each in turn.

\begin{itemize}
\item When $\tau = \Gamma; \Phi \vdash \phi\ol{\tau}$, we have    
\[\begin{array}{ll}
  & \pi_i (\relsem{\Gamma; \Phi \vdash \phi\ol{\tau}}\rho)\\
= & \pi_i ((\rho \phi)\overline{\relsem{\Gamma; \Phi \vdash \tau}\rho})\\
= & (\pi_i (\rho \phi)) (\pi_i(\overline{\relsem{\Gamma; \Phi \vdash \tau}\rho}))\\
= & ((\pi_i \rho) \phi) (\overline{\setsem{\Gamma; \Phi \vdash
    \tau}(\pi_i \rho)})\\
= & \setsem{\Gamma; \Phi \vdash \phi\ol{\tau}}(\pi_i \rho)
\end{array}\]
and, for $f : \rho \to \rho'$ in $\relenv$,
\[\begin{array}{ll}
  & \pi_i (\relsem{\Gamma; \Phi \vdash \phi\ol{\tau}}f)\\
= & \pi_i ((f\phi)_{\overline{\relsem{\Gamma; \Phi \vdash \tau}\rho'}})
\circ \pi_i((\rho\phi)({\overline{\relsem{\Gamma; \Phi \vdash \tau}f}}))\\
= & (\pi_i (f\phi))_{\overline{\pi_i (\relsem{\Gamma; \Phi \vdash \tau}\rho')}}
\circ (\pi_i(\rho\phi))({\overline{\pi_i (\relsem{\Gamma; \Phi \vdash \tau}f}}))\\
= & ((\pi_i f)\phi)_{\overline{\setsem{\Gamma; \Phi \vdash \tau}(\pi_i\rho')}}
\circ ((\pi_i\rho)\phi)({\overline{\setsem{\Gamma; \Phi \vdash \tau}(\pi_i f)}})\\
= & \setsem{\Gamma; \Phi \vdash \phi\ol{\tau}}(\pi_i f)
\end{array}\]
The third equalities of each of the above derivations are by the
induction hypothesis. That $\sem{\Gamma; \Phi \vdash
  \phi\ol{\tau}}$ is $\omega$-cocontinuous is an immediate
consequence of the facts that $\set$ and $\rel$ are locally finitely
presentable, together with Corollary~12 of~\cite{jp19}.
\item When $\tau = (\mu \phi. \lambda
  \overline{\alpha}. H)\overline{\tau}$ we first show that $\sem{ (\mu
    \phi. \lambda \overline{\alpha}. H)\overline{\tau}}$ is
  well-defined.
\begin{itemize}
\item \underline{$T_\rho : [RT_k,RT_k]$}:\/ We must show that, for any
  relation transformer $F = 
  (F^1, F^2, F^*)$, the triple $T_{\rho} F = (T^\set_{\pi_1 \rho}F^1,
  T^\set_{\pi_2 \rho}F^2, T^\rel_{\rho}F)$ is also a relation
  transformer.  Let $\overline{R : \rel(A, B)}$. Then for
  $i = 1, 2$, we have
\[\begin{split}
\pi_i(T^\rel_{\rho}\,F\,\overline{R})
&= \pi_i(\relsem{\Gamma;\Phi,\phi,\overline{\alpha} \vdash H}\rho[\phi := F]\overline{[\alpha := R]}) \\
&= \setsem{\Gamma;\Phi,\phi,\overline{\alpha} \vdash H} (\pi_i (\rho[\phi := F]\overline{[\alpha := R]})) \\
&= \setsem{\Gamma;\Phi,\phi,\overline{\alpha} \vdash H} (\pi_i \rho)[\phi := \pi_i F]\overline{[\alpha := \pi_i R]}) \\
&= T^\set_{\pi_i \rho} (\pi_i F) (\overline{\pi_i R})
\end{split}\]
and 
\[\begin{split}
\pi_i(T^\rel_{\rho}\,F\,\overline{\gamma})
&= \pi_i(\relsem{\Gamma;\Phi,\phi,\overline{\alpha} \vdash H}\id_\rho[\phi
  := \id_F]\overline{[\alpha := \gamma]}) \\
&= \setsem{\Gamma;\Phi,\phi,\overline{\alpha} \vdash H} (\pi_i (\id_\rho[\phi := \id_F]\overline{[\alpha := \gamma]})) \\
&= \setsem{\Gamma;\Phi,\phi,\overline{\alpha} \vdash H} \id_{\pi_i \rho}[\phi := \id_{\pi_i F}]\overline{[\alpha := \pi_i \gamma]} \\
&= T^\set_{\pi_i \rho} (\pi_i F) (\overline{\pi_i \gamma})
\end{split}\]
Here, the second equality in each of the above chains of equalities is
by the induction hypothesis.

We also have that, for every morphism $\delta = (\delta^1, \delta^2) :
F \to G$ in $RT_k$ and all $\overline{R : \rel(A, B)}$,
\[\begin{array}{ll}
  & \pi_i((T^\rel_\rho \delta)_{\overline{R}})\\
= & \pi_i(\relsem{\Gamma;\Phi,\phi,\overline{\alpha} \vdash H}\id_\rho[\phi :=
  \delta]\overline{[\alpha := \id_R]})\\
= & \setsem{\Gamma;\Phi,\phi,\overline{\alpha} \vdash H}\id_{\pi_i\rho}[\phi :=
  \pi_i \delta]\overline{[\alpha := \id_{\pi_i R}]}\\
= & (T^\set_{\pi_i \rho} (\pi_i \delta))_{\overline{\pi_i R}}
\end{array}\]

Here, the second equality is by the induction hypothesis.  That
$T_\rho$ is $\omega$-cocontinuous follows immediately from the
induction hypothesis on $\sem{\Gamma;\Phi,\phi,\ol{\alpha} \vdash H}$
and the fact that colimts are computed componentwise in $RT$.
\item \underline{$\sigma_f = (\sigma^\set_{\pi_1 f},
  \sigma^\set_{\pi_2 f})$ is a natural transformation from $T_{\rho}$
  to $T_{\rho'}$}:\/ We must show that $(\sigma_f)_F =
  ((\sigma^\set_{\pi_1 f})_{F^1}, (\sigma^\set_{\pi_2 f})_{F^2})$ is a
  morphism in $RT_k$ for all relation transformers $F = (F^1, F^2,
  F^*)$, i.e., that $((\sigma_f)_F)_{\overline{R}}
  =(((\sigma^\set_{\pi_1 f})_{F^1})_{\overline{A}},
  ((\sigma^\set_{\pi_2 f})_{F_2})_{\overline{B}})$ is a morphism in
  $\rel$ for all relations $\overline{R : \rel(A, B)}$. Indeed, we
  have that
\[
((\sigma_f)_F)_{\overline{R}} =
\relsem{\Gamma;\Phi,\phi,\overline{\alpha} \vdash H}f[\phi :=
  \id_F]\overline{[\alpha := \id_R]}
\]
is a morphism in $RT_0$ (and thus in $\rel$) by the induction hypothesis.
\end{itemize}

\vspace*{0.05in}

The relation transformer $\mu T_\rho$ is therefore a fixpoint of
$T_\rho$ by Lemma~\ref{lem:fp}, and $\mu \sigma_f$ is a morphism in
$RT_k$ from $\mu T_\rho$ to $\mu T_{\rho'}$. ($\mu$ is shown to be a
functor in~\cite{jp19}.)  So $\relsem{\Gamma; \Phi \vdash (\mu
  \phi. \lambda \overline{\alpha}. H)\overline{\tau}}$, and thus
$\sem{\Gamma; \Phi \vdash (\mu \phi. \lambda
  \overline{\alpha}. H)\overline{\tau}}$, is well-defined.

\vspace*{0.1in}

To see that $\sem{\Gamma; \Phi \vdash (\mu \phi. \lambda
  \overline{\alpha}. H)\overline{\tau}} : [\relenv,RT_0]$, we must
verify three conditions:
\begin{itemize}
\item Condition (1) after Definition~\ref{def:relenv-functor} is
  satisfied since
\[\begin{split}
\pi_i(\relsem{\Gamma;\Phi \vdash (\mu \phi. \lambda
  \overline{\alpha}. H) \overline{\tau}}\rho) 
&= \pi_i( (\mu T_\rho) (\overline{\relsem{\Gamma;\Phi \vdash
    \tau}\rho})) \\ 
&= \pi_i(\mu T_{\rho}) (\overline{\pi_i(\relsem{\Gamma;\Phi \vdash
    \tau}\rho})) \\ 
&= \mu T^\set_{\pi_i\rho} (\overline{\setsem{\Gamma;\Phi \vdash
    \tau}(\pi_i\rho)}) \\ 
&= \setsem{\Gamma;\Phi \vdash (\mu \phi. \lambda \overline\alpha. H)
  \overline{\tau}}(\pi_i\rho) 
\end{split}\]
The third equality is by Equation ~\ref{eq:mu} and the induction
hypothesis.
\item Condition (2) after Definition~\ref{def:relenv-functor} is
  satisfied since it is subsumed by the previous condition because $k
  = 0$.
\item The third bullet point of Definition~\ref{def:relenv-functor} is
  satisfied because
\[\begin{split}
& \pi_i(\relsem{\Gamma;\Phi \vdash (\mu \phi. \lambda
  \overline\alpha. H)\overline{\tau}}f)\\
&= \pi_i((\mu T_{\rho'})(\overline{\relsem{\Gamma;\Phi \vdash
    \tau}f}) \circ (\mu \sigma_f)_{\overline{\relsem{\Gamma;\Phi
      \vdash \tau}\rho}}) \\ 
&= \pi_i((\mu T_{\rho'})(\overline{\relsem{\Gamma;\Phi \vdash
    \tau}f})) \circ \pi_i((\mu
\sigma_f)_{\overline{\relsem{\Gamma;\Phi \vdash \tau}\rho}}) \\  
&= \pi_i(\mu T_{\rho'})(\overline{\pi_i(\relsem{\Gamma;\Phi
    \vdash \tau}f})) \circ \pi_i(\mu
\sigma_f)_{\overline{\pi_i(\relsem{\Gamma;\Phi \vdash 
      \tau}\rho})} \\ 
&= (\mu T^\set_{\pi_i \rho'})(\overline{\setsem{\Gamma;\Phi \vdash
    \tau}(\pi_i f)}) \circ (\mu \sigma^\set_{\pi_i
  f})_{\overline{\setsem{\Gamma;\Phi \vdash \tau}(\pi_i\rho)}} \\ 
&= \setsem{\Gamma;\Phi \vdash (\mu \phi. \lambda
  \overline\alpha. H)\overline{\tau}}(\pi_i f). 
\end{split}\]
The fourth equality is by~\ref{eq:mu} and the induction hypothesis.
\end{itemize}
As before, that $\sem{\Gamma; \Phi \vdash (\mu \phi. \lambda
  \overline{\alpha}. H)\overline{\tau}}$ is $\omega$-concontinuous
follows from the facts that $\set$ and $\rel$ are locally finitely
presentable, and that colimits in $\relenv$ are computed
componentwise, together with Corollary~12 of~\cite{jp19}.
\end{itemize}
\end{proof}
\end{comment}

%We need
%$\relsem{\Gamma; \Phi, \phi \vdash \tau}\rho$ $=
%\relsem{\Gamma, \psi; \Phi \vdash \tau[\phi :==
%    \psi]}\rho[\psi:=\rho\phi]$, and that $\relsem{\Gamma \vdash
%  \sigma \to \tau}$ and
%$\setsem{\Gamma; \Phi, \phi \vdash \tau}\rho$ $=
%\setsem{\Gamma, \psi; \Phi \vdash \tau[\phi :==
%    \psi]}\rho[\psi:=\rho\phi]$. 
We can also prove by simultaneous induction that our interpretations
of types interact well with demotion of functorial variables. Indeed,
we have that, for any $\rho,\rho'$ and $f: \rho \to \rho'$ in
$\setenv$, 
\begin{gather}
\label{thm:demotion-objects}
\setsem{\Gamma; \Phi, \phi \vdash F} \rho = \setsem{\Gamma, \psi; \Phi
  \vdash F[\phi :== \psi] } \rho[\psi := \rho \phi]\\
\label{thm:demotion-morphisms}
\setsem{\Gamma; \Phi, \phi \vdash F} f = \setsem{\Gamma, \psi; \Phi
  \vdash F[\phi :== \psi]} f[\psi \mapsto f\phi]
\end{gather}
Moreover,
%\begin{lemma}\label{lem:demotion-objects}
if $\rho, \rho' : \setenv$ are such that $\rho \phi = \rho \psi =
\rho' \phi = \rho' \psi$, \,$f : \rho \to \rho'$ in $\setenv$ is such
that $f \phi = f \psi = \id_{\rho \phi}$,\, $\Gamma; \Phi, \phi^k
\vdash F$,\, $\Gamma;\Phi,\ol{\alpha} \vdash G$,\,
$\Gamma;\Phi,\alpha_1...\alpha_k \vdash H$, and $\ol{\Gamma;\Phi \vdash
F}$, then
\begin{gather}
%\label{thm:demotion-objects}
%\setsem{\Gamma; \Phi, \phi \vdash F} \rho = \setsem{\Gamma, \psi; \Phi
%  \vdash F[\phi :== \psi] } \rho{\color{blue}[\psi := \rho \phi]}\\
%\label{thm:demotion-morphisms}
%\setsem{\Gamma; \Phi, \phi \vdash F} f = \setsem{\Gamma, \psi; \Phi
%  \vdash F[\phi :== \psi]} f{\color{blue}[\psi \mapsto f\phi]}\\
\label{eq:subs-var}
\setsem{\Gamma;\Phi \vdash G[\ol{\alpha := F}]}\rho =
\setsem{\Gamma;\Phi,\ol{\alpha} \vdash G}\rho[\ol{\alpha := 
\setsem{\Gamma;\Phi \vdash F}\rho}]\\
\label{eq:subs-var-morph}
\setsem{\Gamma;\Phi \vdash G[\ol{\alpha := F}]}f =
\setsem{\Gamma;\Phi,\ol{\alpha} \vdash G}f[\ol{\alpha :=
\setsem{\Gamma;\Phi \vdash F}f}]\\
\label{eq:subs-const}
\setsem{\Gamma; \Phi \vdash F[\phi := H]}\rho
= \setsem{\Gamma; \Phi, \phi \vdash F}\rho
[\phi := \lambda \ol{A}.\, \setsem{\Gamma;\Phi,\overline{\alpha}\vdash
    H}\rho[\overline{\alpha := A}]] \\ 
\label{eq:subs-const-morph}
\setsem{\Gamma; \Phi \vdash F[\phi := H]}f
= \setsem{\Gamma; \Phi, \phi \vdash F}f
[\phi := \lambda \ol{A}.\,\setsem{\Gamma;\Phi,\overline{\alpha}\vdash
    H}f[\overline{\alpha := \id_{\ol{A}}}]] 
\end{gather}
Identities analogous to (\ref{thm:demotion-objects}) through
(\ref{eq:subs-const-morph}) hold for relational interpretations as well.
%\end{lemma}

\begin{comment}
\begin{lemma}\label{lem:substitution}
Let $\rho$ and $\rho'$ be set environments and let $f : \rho \to \rho'$ be a
morphism of set environments. 
\begin{itemize}
\item If $\Gamma;\Phi,\ol{\alpha} \vdash F$ and $\Gamma;\Phi \vdash \tau$,
  then  
\begin{gather}\label{eq:subs-var}
\setsem{\Gamma;\Phi \vdash F[\ol{\alpha := \tau}]}\rho =
\setsem{\Gamma;\Phi,\ol{\alpha} \vdash F}\rho[\ol{\alpha :=
\setsem{\Gamma;\Phi \vdash \tau}\rho}]\\
\intertext{and}\label{eq:subs-var-morph}
\setsem{\Gamma;\Phi \vdash F[\ol{\alpha := \tau}]}f =
\setsem{\Gamma;\Phi,\ol{\alpha} \vdash F}f[\ol{\alpha :=
\setsem{\Gamma;\Phi \vdash \tau}f}]
\end{gather}
\item If $\Gamma;\Phi,\phi^k\vdash F$ and
  $\Gamma;\Phi,\alpha_1...\alpha_k \vdash H$, then 
\begin{gather}\label{eq:subs-const}
\setsem{\Gamma; \Phi \vdash F[\phi := H]}\rho
= \setsem{\Gamma; \Phi, \phi \vdash F}\rho
[\phi := \lambda \ol{A}.\, \setsem{\Gamma;\Phi,\overline{\alpha}\vdash
    H}\rho[\overline{\alpha := A}]] \\ 
\intertext{and}\label{eq:subs-const-morph}
\setsem{\Gamma; \Phi \vdash F[\phi := H]}f
= \setsem{\Gamma; \Phi, \phi \vdash F}f
[\phi := \lambda \ol{A}.\,\setsem{\Gamma;\Phi,\overline{\alpha}\vdash
    H}f[\overline{\alpha := \id_{\ol{A}}}]] 
\end{gather}
\end{itemize}
Analogous identities hold for relation environments and morphisms
between them.
\end{lemma}

\end{comment}

\begin{comment}

\begin{thm}\label{thm:demotion-objects}
Let \, $\Gamma; \Phi, \phi \vdash \tau : \F$. If $\rho, \rho' : \setenv$
are such that $\rho \phi = \rho \psi = \rho' \phi = \rho' \psi$, and
if $f : \rho \to \rho'$ is a morphism of set environments such that $f
\phi = f \psi = \id_{\rho \phi}$, then
\[\setsem{\Gamma; \Phi, \phi \vdash \tau} \rho = \setsem{\Gamma, \psi;
  \Phi \vdash \tau[\phi :== \psi] } \rho\]
and
\[\setsem{\Gamma; \Phi, \phi \vdash \tau} f = \setsem{\Gamma, \psi;
  \Phi \vdash \tau[\phi :== \psi]} f\]

\vspace*{0.1in}

\noindent
Analogously, if $\rho, \rho' : \relenv$ are such that $\rho \phi =
\rho \psi = \rho' \phi = \rho' \psi$, and if $f : \rho \to \rho'$ is a
morphism of relation environments such that $f \phi = f \psi =
\id_{\rho \phi}$, then
\[\relsem{\Gamma; \Phi, \phi \vdash \tau} \rho = \relsem{\Gamma, \psi;
  \Phi \vdash \tau[\phi :== \psi] } \rho\]
and
\[\relsem{\Gamma; \Phi, \phi \vdash \tau} f = \relsem{\Gamma, \psi;
  \Phi \vdash \tau[\phi :== \psi]} f\]
\end{thm}

\end{comment}

\begin{comment}
---
The next two theorems are proven by simultaneous induction. We are
actually only interested in using Theorem~\ref{thm:demotion-objects},
but in order to prove the $\mu$ case for this theorem, we need
Theorem~\ref{thm:demotion-morph} to show that two functors have equal
actions on morphisms.

\begin{thm}\label{thm:demotion-objects}
Let $\Gamma; \Phi, \phi \vdash \tau : \F$. If $\rho, \rho' : \setenv$
are such that $\rho \phi = \rho \psi = \rho' \phi = \rho' \psi$, and
if $f : \rho \to \rho'$ is a morphism of set environments such that $f
\phi = f \psi = \id_{\rho \phi}$, then
\[\setsem{\Gamma; \Phi, \phi \vdash \tau} \rho = \setsem{\Gamma, \psi;
  \Phi \vdash \tau[\phi :== \psi] } \rho\]
and
\[\setsem{\Gamma; \Phi, \phi \vdash \tau} f = \setsem{\Gamma, \psi;
  \Phi \vdash \tau[\phi :== \psi]} f\]

\vspace*{0.1in}

\noindent
Analogously, if $\rho, \rho' : \relenv$ are such that $\rho \phi =
\rho \psi = \rho' \phi = \rho' \psi$, and if $f : \rho \to \rho'$ is a
morphism of relation environments such that $f \phi = f \psi =
\id_{\rho \phi}$, then
\[\relsem{\Gamma; \Phi, \phi \vdash \tau} \rho = \relsem{\Gamma, \psi;
  \Phi \vdash \tau[\phi :== \psi] } \rho\]
and
\[\relsem{\Gamma; \Phi, \phi \vdash \tau} f = \relsem{\Gamma, \psi;
  \Phi \vdash \tau[\phi :== \psi]} f\]
\end{thm}
\begin{proof}
We prove the result for set interpretations by induction on the
structure of $\tau$.  The case for relational interpretations proceeds
analogously. Since the only interesting cases are the application case
and the $\mu$-case, we elide the others.
\begin{itemize}
\item If $\Gamma; \Phi, \phi \vdash \phi \ol\tau : \F$, then the
  induction hypothesis gives that  
\[\setsem{\Gamma; \Phi, \phi \vdash \tau } \rho = \setsem{\Gamma,
  \psi; \Phi \vdash \tau[\phi :== \psi] } \rho\] 
and
\[\setsem{\Gamma; \Phi, \phi \vdash \tau} f = \setsem{\Gamma, \psi;
  \Phi \vdash \tau[\phi :== \psi]} f\]
for each $\tau$. Then
\begin{align*}
 & \setsem{\Gamma; \Phi, \phi \vdash \phi \ol\tau } \rho \\
= \; & (\rho \phi) \ol{\setsem{\Gamma; \Phi, \phi \vdash \tau } \rho} \\
= \; & (\rho \phi) \ol{\setsem{\Gamma, \psi; \Phi \vdash \tau[\phi :==
      \psi] } \rho} \\ 
= \; & (\rho \psi) \ol{\setsem{\Gamma, \psi; \Phi \vdash \tau[\phi :==
      \psi] } \rho} \\ 
= \; & \setsem{\Gamma, \psi; \Phi \vdash \psi \ol{\tau[\phi :== \psi]}
} \rho \\ 
= \; & \setsem{\Gamma, \psi; \Phi \vdash (\phi \ol\tau)[\phi :== \psi]
} \rho 
\end{align*}
Here, the first and fifth equalities are by
Definition~\ref{def:set-sem}, and the fourth equality is by equality
of the functors $\rho \phi$ and $\rho \psi$. We also have that
\begin{align*}
& \setsem{\Gamma; \Phi, \phi \vdash \phi \ol\tau} f \\
= \; & (f \phi)_{\ol{\setsem{\Gamma; \Phi, \phi \vdash \tau} \rho'}} 
    \circ (\rho \phi) \ol{\setsem{\Gamma; \Phi, \phi \vdash \tau} f} \\
= \; & (\id_{\rho \phi})_{\ol{\setsem{\Gamma; \Phi, \phi \vdash \tau} \rho'}} 
    \circ (\rho \phi) \ol{\setsem{\Gamma; \Phi, \phi \vdash \tau} f} \\
= \; & (\rho \phi) \ol{\setsem{\Gamma; \Phi, \phi \vdash \tau} f} \\
= \; & (\rho \psi) \ol{\setsem{\Gamma, \psi; \Phi \vdash \tau[\phi :==
      \psi]} f} \\ 
= \; & (\id_{\rho \psi})_{\ol{\setsem{\Gamma, \psi; \Phi \vdash
      \tau[\phi :== \psi]} \rho'}} \circ (\rho \psi)
\ol{\setsem{\Gamma, \psi; \Phi \vdash \tau[\phi :== \psi]} f} \\ 
= \; & (f \psi)_{\ol{\setsem{\Gamma, \psi; \Phi \vdash \tau[\phi :==
        \psi]} \rho'}} \circ (\rho \psi) \ol{\setsem{\Gamma, \psi;
    \Phi \vdash \tau[\phi :== \psi]} f} \\ 
= \; & \setsem{\Gamma, \psi; \Phi \vdash \psi \ol{\tau[\phi :== \psi]}} f \\
= \; & \setsem{\Gamma, \psi; \Phi \vdash (\phi \ol\tau) [\phi :== \psi]} f
\end{align*}
\item If $\Gamma; \Phi, \phi \vdash (\mu \phi'. \lambda
  \ol\alpha. H)\ol\tau : \F$, then the induction hypothesis gives that
\[\setsem{\Gamma; \Phi, \phi', \ol{\alpha}, \phi \vdash H } \rho =
\setsem{\Gamma, \psi; \Phi, \phi', \ol{\alpha} \vdash H[\phi :== \psi]
} \rho\]
and 
\[\setsem{\Gamma; \Phi, \phi \vdash \tau } \rho = \setsem{\Gamma,
  \psi; \Phi \vdash \tau[\phi :== \psi] } \rho\]
as well as that
\[\setsem{\Gamma; \Phi, \phi', \ol{\alpha}, \phi \vdash H } f =
\setsem{\Gamma, \psi; \Phi, \phi', \ol{\alpha} \vdash H[\phi :== \psi]
} f\]
and 
\[\setsem{\Gamma; \Phi, \phi \vdash \tau } f = \setsem{\Gamma, \psi;
  \Phi \vdash \tau[\phi :== \psi] } f\]
for each $\tau$. Then
\begin{align*}
& \setsem{\Gamma; \Phi, \phi \vdash (\mu \phi'. \lambda
    \ol\alpha. H)\ol\tau } \rho \\ 
= \; & (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma; \Phi, \phi',
  \ol{\alpha}, \phi \vdash H} \rho[\phi' := F][\ol{\alpha := A}]))
\ol{\setsem{ \Gamma; \Phi, \phi \vdash \tau} \rho} \\ 
= \; & (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma, \psi; \Phi,
  \phi', \ol{\alpha} \vdash H[\phi :== \psi]} \rho[\phi' :=
  F][\ol{\alpha := A}])) \ol{\setsem{ \Gamma; \Phi, \phi \vdash \tau}
  \rho} \\ 
= \; &  (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma, \psi; \Phi,
  \phi', \ol{\alpha} \vdash H[\phi :== \psi]} \rho[\phi' :=
  F][\ol{\alpha := A}])\ol{\setsem{ \Gamma, \psi; \Phi \vdash
    \tau[\phi :== \psi]} \rho} \\ 
= \; & \setsem{\Gamma, \psi; \Phi \vdash (\mu \phi'. \lambda
  \ol\alpha. H[\phi :== \psi]) \ol{\tau[\phi :== \psi]} } \rho \\ 
= \; & \setsem{\Gamma, \psi; \Phi \vdash ((\mu \phi'. \lambda
  \ol\alpha. H) \ol\tau)[\phi :== \psi] } \rho   
\end{align*} 
The first and fifth equalities are by Definition~\ref{def:set-sem}.
The second equality follows from the following equality:
\begin{align*}
&\lambda F. \lambda \ol{A}. \setsem{\Gamma; \Phi, \phi', \ol{\alpha},
    \phi \vdash H} \rho[\phi' := F][\ol{\alpha := A}] \\ 
& = \lambda F. \lambda \ol{A}. \setsem{\Gamma, \psi; \Phi, \phi',
    \ol{\alpha} \vdash H[\phi :== \psi]} \rho[\phi' := F][\ol{\alpha
      := A}] 
\end{align*}
These two maps have the same actions on objects and morphisms by the
induction hypothesis on $H$, and the fact that the extended
environment $\rho[\phi' := F][\ol{\alpha := A}]$ satisfies the
required hypothesis. They are thus equal as functors and so have the
same fixpoint. We also have that 
\begin{align*}
& \setsem{\Gamma; \Phi, \phi \vdash (\mu \phi'. \lambda
    \ol\alpha. H)\ol\tau } f \\ 
= \; & (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma; \Phi, \phi',
  \ol{\alpha}, \phi \vdash H} f [\phi' := \id_F][\ol{\alpha :=
    \id_A}]))_{\ol{\setsem{\Gamma; \Phi, \phi \vdash \tau} \rho'}} \\ 
\; & \circ (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma; \Phi,
  \phi', \ol{\alpha}, \phi \vdash H} \rho[\phi' := F][\ol{\alpha :=
    A}])) \ol{\setsem{ \Gamma; \Phi, \phi \vdash \tau} f} \\ 
= \; & (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma; \Phi, \phi',
  \ol{\alpha}, \phi \vdash H} f [\phi' := \id_F][\ol{\alpha :=
    \id_A}]))_{\ol{\setsem{\Gamma, \psi; \Phi \vdash \tau[\phi :==
        \psi]} \rho'}} \\ 
\; & \circ (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma; \Phi,
  \phi', \ol{\alpha}, \phi \vdash H} \rho[\phi' := F][\ol{\alpha :=
    A}])) \ol{\setsem{ \Gamma, \psi; \Phi \vdash \tau[\phi :== \psi]}
  f} \\ 
= \; & (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma, \psi; \Phi,
  \phi', \ol{\alpha} \vdash H[\phi :== \psi]} f [\phi' :=
  \id_F][\ol{\alpha := \id_A}]))_{\ol{\setsem{\Gamma, \psi; \Phi
      \vdash \tau[\phi :== \psi]} \rho'}} \\ 
\; & \circ (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma; \Phi,
  \phi', \ol{\alpha}, \phi \vdash H} \rho[\phi' := F][\ol{\alpha :=
    A}])) \ol{\setsem{ \Gamma, \psi; \Phi \vdash \tau[\phi :== \psi]}
  f} \\ 
= \; & (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma, \psi; \Phi,
  \phi', \ol{\alpha} \vdash H[\phi :== \psi]} f [\phi' :=
  \id_F][\ol{\alpha := \id_A}]))_{\ol{\setsem{\Gamma, \psi; \Phi
      \vdash \tau[\phi :== \psi]} \rho'}} \\ 
\; & \circ (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma, \psi;
  \Phi, \phi', \ol{\alpha} \vdash H[\phi :== \psi]} \rho[\phi' :=
  F][\ol{\alpha := A}])) \ol{\setsem{ \Gamma, \psi; \Phi \vdash
    \tau[\phi :== \psi]} f} \\ 
= \; & \setsem{\Gamma, \psi; \Phi \vdash (\mu \phi'. \lambda
  \ol\alpha. H[\phi :== \psi]) \ol{\tau[\phi :== \psi]} } f \\ 
= \; & \setsem{\Gamma, \psi; \Phi \vdash ((\mu \phi'. \lambda
  \ol\alpha. H) \ol\tau)[\phi :== \psi] } f 
\end{align*} 
The first and fifth equalities are by Definition~\ref{def:set-sem}.
The third equality is by the equality of the arguments to the first
$\mu$ operator:
\begin{align*}
& \lambda F. \lambda \ol{A}. \setsem{\Gamma; \Phi, \phi', \ol{\alpha},
    \phi \vdash H} f [\phi' := \id_F][\ol{\alpha := \id_A}] \\ 
& = \lambda F. \lambda \ol{A}. \setsem{\Gamma, \psi; \Phi, \phi',
    \ol{\alpha} \vdash H[\phi :== \psi]} f [\phi' := \id_F][\ol{\alpha
      := \id_A}] 
\end{align*}
By the induction hypothesis on $H$ and the fact that the morphism
$f[\phi' := \id_F][\ol{\alpha := \id_A}] : \rho[\phi' := F][\ol{\alpha
    := A}] \to \rho'[\phi' := F][\ol{\alpha := A}]$ still satisfies
the required hypotheses.  The fourth equality is by the equality of
the arguments to the second $\mu$ operator:
\begin{align*}
& \lambda F. \lambda \ol{A}. \setsem{\Gamma; \Phi, \phi', \ol{\alpha},
    \phi \vdash H} \rho[\phi' := F][\ol{\alpha := A}] \\ 
& = \lambda F. \lambda \ol{A}. \setsem{\Gamma, \psi; \Phi, \phi',
    \ol{\alpha} \vdash H[\phi :== \psi]} \rho[\phi' := F][\ol{\alpha
      := A}] 
\end{align*}
By the same reasoning as above, these two maps are equal as functors,
and thus have the same fixpoint.
\end{itemize}
\end{proof}

---

The following lemma ensures that substitution interacts well with type
interpretations.  It is a consequence of
Definitions~\ref{def:second-order-subst},~\ref{def:set-interp},
and~\ref{def:rel-interp}.

\begin{lemma}\label{lem:substitution}
Let $\rho$ and $\rho'$ be set environments and let $f : \rho \to \rho'$ be a
morphism of set environments. 
\begin{itemize}
\item If $\Gamma;\Phi,\ol{\alpha} \vdash F$ and $\Gamma;\Phi \vdash \tau$,
  then  
\begin{gather}\label{eq:subs-var}
\setsem{\Gamma;\Phi \vdash F[\ol{\alpha := \tau}]}\rho =
\setsem{\Gamma;\Phi,\ol{\alpha} \vdash F}\rho[\ol{\alpha :=
\setsem{\Gamma;\Phi \vdash \tau}\rho}]\\
\intertext{and}\label{eq:subs-var-morph}
\setsem{\Gamma;\Phi \vdash F[\ol{\alpha := \tau}]}f =
\setsem{\Gamma;\Phi,\ol{\alpha} \vdash F}f[\ol{\alpha :=
\setsem{\Gamma;\Phi \vdash \tau}f}]
\end{gather}
\item If $\Gamma;\Phi,\phi^k\vdash F$ and
  $\Gamma;\Phi,\alpha_1...\alpha_k \vdash H$, then 
\begin{gather}\label{eq:subs-const}
\setsem{\Gamma; \Phi \vdash F[\phi := H]}\rho
= \setsem{\Gamma; \Phi, \phi \vdash F}\rho
[\phi := \lambda \ol{A}.\, \setsem{\Gamma;\Phi,\overline{\alpha}\vdash
    H}\rho[\overline{\alpha := A}]] \\ 
\intertext{and}\label{eq:subs-const-morph}
\setsem{\Gamma; \Phi \vdash F[\phi := H]}f
= \setsem{\Gamma; \Phi, \phi \vdash F}f
[\phi := \lambda \ol{A}.\,\setsem{\Gamma;\Phi,\overline{\alpha}\vdash
    H}f[\overline{\alpha := \id_{\ol{A}}}]] 
\end{gather}
\end{itemize}
Analogous identities hold for relation environments and morphisms
between them.
\end{lemma}
\end{comment}

\begin{comment}
\begin{proof}
The proofs for the set and relational interpretations are completely
analogous, so we just prove the former. Likewise, we only prove
Equations~\ref{eq:subs-var} and~\ref{eq:subs-const}, since the proofs
for Equations~\ref{eq:subs-var-morph} and~\ref{eq:subs-const-morph}
are again analogous. Finally, we prove Equation~\ref{eq:subs-var} for
substitution for just a single type variable since the proof for
multiple simultaneous substitutions proceeds similarly.

Although Equation~\ref{eq:subs-var} is a special case of
Equation~\ref{eq:subs-const}, it is convenient to prove
Equation~\ref{eq:subs-var} first, and then use it to prove
Equation~\ref{eq:subs-const}. We prove Equation~\ref{eq:subs-var} by
induction on the structure of $F$ as follows:
\begin{itemize}
\item If $\Gamma; \emptyset \vdash F : \T$, or if $F$ is $\onet$ or
  $\zerot$, then $F$ does not contain any functorial variables to
  replace, so there is nothing to prove.
\item If $F$ is $F_1 \times F_2$ or $F_1 + F_2$, then the substitution
  distributes over the product or coproduct as appropriate, so the
  result follows immediately from the induction hypothesis.
\item If $F = \beta$ with $\beta \neq \alpha$, then there is nothing to
  prove.
\item If $F = \alpha$, then
\[\setsem{\Gamma;\Phi\vdash\alpha[\alpha := \tau]}\rho \, = \,
\setsem{\Gamma;\Phi\vdash\tau}\rho \, = \,
\setsem{\Gamma;\Phi,\alpha\vdash\alpha}\rho[\alpha :=
  \setsem{\Gamma;\Phi\vdash\tau}\rho]\]
\item If $F = \phi \overline{\sigma}$ with $\phi \neq \alpha$, then
\[\begin{array}{ll}
 &\setsem{\Gamma; \Phi \vdash (\phi \overline{\sigma})[\alpha := \tau]}\rho\\
=&\setsem{\Gamma; \Phi \vdash \phi (\overline{\sigma[\alpha := \tau]})}\rho \\
=&(\rho\phi)\overline{\setsem{\Gamma;\Phi\vdash\sigma[\alpha := \tau]}\rho} \\
=&(\rho\phi)\overline{\setsem{\Gamma;\Phi,\alpha\vdash\sigma}
      \rho [\alpha := \setsem{\Gamma;\Phi\vdash\tau}] } \\
=&\setsem{\Gamma; \Phi, \alpha \vdash \phi \overline{\sigma}}
      \rho [\alpha := \setsem{\Gamma;\Phi\vdash\tau}]
\end{array}\]
Here, the third equality is by the induction hypothesis.
\item If $F = (\mu \phi. \lambda \overline{\beta}. G)
  \overline{\sigma}$, then
\[\begin{array}{ll}
  & \setsem{\Gamma; \Phi \vdash ((\mu \phi. \lambda
  \overline{\beta}. G) \overline{\sigma}) [\alpha := \tau]}\rho \\
=&\setsem{\Gamma; \Phi \vdash (\mu \phi. \lambda
  \overline{\beta}. G[\alpha := \tau]) (\overline{\sigma [\alpha :=
      \tau]})}\rho \\ 
=&\mu(\setsem{\Gamma; \Phi, \phi, \overline{\beta} \vdash G[\alpha := \tau]}
      \rho[\phi := \text{--}][\overline{\beta := \text{--}}])
      (\overline{\setsem{\Gamma;\Phi \vdash \sigma [\alpha := \tau]}\rho}) \\
=&\mu(\setsem{\Gamma; \Phi, \phi, \overline{\beta}, \alpha \vdash G}
\rho [\alpha := \setsem{\Gamma;\Phi \vdash \tau}\rho] [\phi := \text{--}]
[\overline{\beta := \text{--}}]) \\
 &\hspace{0.5in} (\overline{\setsem{\Gamma;\Phi,\alpha \vdash \sigma}
   \rho [\alpha := \setsem{\Gamma;\Phi\vdash \tau}\rho]}) \\
=&\setsem{\Gamma;\Phi,\alpha\vdash(\mu\phi.\lambda\overline{\beta}.G)
  \overline{\sigma}}\rho  [\alpha := \setsem{\Gamma;\Phi \vdash
   \tau}\rho] 
\end{array}  \]
Here, the third equality is by the induction hypothesis and weakening.
\end{itemize}

We now prove Equation~\ref{eq:subs-const}, again by induction on the
structure of $F$.
\begin{itemize}
\item If $\Gamma; \emptyset \vdash F : \T$, or if $F$ is $\onet$ or
  $\zerot$, then $F$ does not contain any functorial variables to
  replace, so there is nothing to prove.
\item If $F$ is $F_1 \times F_2$ or $F_1 + F_2$, then the substitution
  distributes over the product or coproduct as appropriate, so the
  result follows immediately from the induction hypothesis.
\item If $F = \phi \overline{\tau}$, then
\[\begin{array}{ll}
 &\setsem{\Gamma; \Phi \vdash (\phi \overline{\tau})[\phi := H]}\rho \\
=&\setsem{\Gamma; \Phi \vdash H[\overline{\alpha:= \tau[\phi := H]}]}\rho \\
=&\setsem{\Gamma; \Phi \vdash H}\rho [\overline{\alpha:=
    \setsem{\Gamma;\Phi\vdash\tau[\phi := H]}\rho}] \\ 
=&\setsem{\Gamma; \Phi \vdash H}\rho [\overline{\alpha:=
    \setsem{\Gamma;\Phi,\phi\vdash\tau} \rho [\phi :=
      \setsem{\Gamma;\Phi,\overline{\alpha}\vdash
        H}\rho[\overline{\alpha := \text{--}}]]}] \\
=&\setsem{\Gamma; \Phi, \phi \vdash \phi \overline{\tau}}\rho [\phi :=
  \setsem{\Gamma;\Phi,\overline{\alpha}\vdash H}\rho[\overline{\alpha
      := \text{--}}]] 
  \end{array}\]
Here, the first equality is by
Definition~\ref{def:second-order-subst}, the second is by
Equation~\ref{eq:subs-var}, the third is by the induction hypothesis,
and the fourth is by Definition~\ref{def:set-sem}.
\item If $F = \psi \overline{\tau}$ with $\psi \neq \phi$, then the
  proof is similar to that for the previous case, but simpler, because
  $\phi$ only needs to be substituted in the arguments $\ol{\tau}$ of
  $\psi$.
\item If $F = (\mu \psi. \lambda \overline{\beta}. G)
  \overline{\tau}$, then
\[\begin{array}{ll}
 &\setsem{\Gamma; \Phi \vdash ((\mu \psi. \lambda \overline{\beta}. G)
 \overline{\tau})[\phi := H]}\rho \\ 
=&\setsem{\Gamma; \Phi \vdash (\mu \psi. \lambda
  \overline{\beta}. G[\phi := H]) (\overline{\tau[\phi := H]})}\rho \\
=&\mu(\setsem{\Gamma; \Phi, \psi, \overline{\beta} \vdash G[\phi := H]}
    \rho[\psi := \text{--}][\overline{\beta := \text{--}}])
    (\overline{\setsem{\Gamma;\Phi \vdash \tau[\phi := H]}\rho}) \\
=&\mu(\setsem{\Gamma; \Phi, \psi, \overline{\beta}, \phi \vdash G}
\rho [\phi := \setsem{\Gamma;\Phi,\overline{\alpha}\vdash
    H}\rho[\overline{\alpha := \text{--}}]][\psi := \text{--}]
     [\overline{\beta := \text{--}}]) \\
&\hspace{0.5in} (\overline{\setsem{\Gamma;\Phi,\phi \vdash \tau} \rho
       [\phi := \setsem{\Gamma;\Phi,\overline{\alpha}\vdash H}
         \rho[\overline{\alpha := \text{--}}]]}) \\
=&\setsem{\Gamma;\Phi,\phi\vdash(\mu\psi.\lambda\overline{\beta}.G)
  \overline{\tau}}\rho [\phi :=
  \setsem{\Gamma;\Phi,\overline{\alpha}\vdash H}\rho[\overline{\alpha
      := \text{--}}]] 
\end{array}\]
Here, the first equality is by
Definition~\ref{def:second-order-subst}, the second and fourth are by
Definition~\ref{def:set-sem}, and the third is by the induction
hypothesis and weakening.
\end{itemize}
\end{proof}
\end{comment}

\section{The Identity Extension Lemma}\label{sec:iel}

In most treatments of parametricity, equality relations on sets are
taken as {\em given} --- either directly as diagonal relations, or
perhaps via reflexive graphs if kinds are also being tracked --- and
the graph relations used to validate existence of initial algebras are
defined in terms of them. We take a different approach here, giving a
categorical definition of graph relations for morphisms (i.e., natural
transformations) between functors and {\em constructing} equality
relations as particular graph relations. Our definitions specialize to
the usual ones for the graph relation for a morphism between sets and
the equality relation for a set. In light of its novelty, we spell out
our construction in detail.

The standard definition of the graph for a morphism $f : A \to B$ in
$\set$ is the relation $\graph{f} : \rel(A,B)$ defined by $(x,y) \in
\graph{f}$ iff $fx = y$. This definition naturally generalizes to
associate to each natural transformation between $k$-ary functors on
$\set$ a $k$-ary relation transformer as follows:

\begin{dfn}\label{dfn:graph-nat-transf}
If $F, G: \Set^k \to \Set$
%are $k$-ary functors
and $\alpha : F \to G$ is a natural transformation, then the functor
$\graph{\alpha}^*: \rel^k \to \rel$ is defined as follows. Given $R_1
: \rel(A_1, B_1),...,R_k : \rel(A_k,B_k)$, let $\iota_{R_i} : R_i
\hookrightarrow A_i \times B_i$, for $i = 1,...,k$, be the inclusion
of $R_i$ as a subset of $A_i \times B_i$,
%. By the universal property of the product, there exists a
let $h_{\overline{A \times B}}$ be the unique morphism making the diagram
{\footnotesize\[\begin{tikzcd}[row sep = large]
        F\overline{A}
        &F(\overline{A \times B})
        \ar[l, "{F\overline{\pi_1}}"']
        \ar[r, "{F\overline{\pi_2}}"]
        \ar[d, dashed, "{h_{\overline{A \times B}}}"]
        &F\overline{B}
        \ar[r, "{\alpha_{\ol{B}}}"]
        &G\overline{B} \\
        &F\overline{A} \times G\overline{B}
        \ar[ul, "{\pi_1}"] \ar[urr, "{\pi_2}"']
\end{tikzcd}\]}

\noindent
commute, and let $h_{\overline{R}} : F\overline{R} \to F\overline{A}
\times G\overline{B}$ be $h_{\overline{A \times B}} \circ
F\overline{\iota_R}$. Further, let $\alpha^\wedge\overline{R}$ be the
subobject through which $h_{\overline{R}}$ is factorized by the
mono-epi factorization system in $\set$, as shown in the
following diagram:
{\footnotesize\[\begin{tikzcd}
        F\overline{R}
        \ar[rr, "{h_{\overline{R}}}"]
        \ar[dr, twoheadrightarrow, "{q_{\alpha^\wedge\overline{R}}}"']
        &&F\overline{A} \times G\overline{B} \\
        &\alpha^\wedge\overline{R}
        \ar[ur, hookrightarrow, "{\iota_{\alpha^\wedge\overline{R}}}"']
\end{tikzcd}\]}

\noindent
Then $\alpha^\wedge\overline{R} : \rel(F\overline{A}, G\overline{B})$
by construction, so the action of $\langle \alpha \rangle^*$ on
objects can be given by $\langle \alpha \rangle^* \overline{(A,B,R)} =
(F\overline{A}, G\overline{B}, \iota_{\alpha^\wedge
  \overline{R}}\alpha^\wedge\overline{R})$. Its action on morphisms is
given by $\graph{\alpha}^*\overline{(\beta, \beta')} =
(F\overline\beta, G\overline\beta')$.
\end{dfn}

The data in Definition~\ref{dfn:graph-nat-transf} yield the {\em graph relation
transformer for $\alpha$}, denoted $\graph{\alpha} = (F, G,
\graph{\alpha}^*)$.
% yield a relation transformer $\graph{\alpha} = (F, G,
%\graph{\alpha}^*)$, called the {\em graph relation transformer for
%$\alpha$}.

\begin{lemma}\label{lem:graph-reln-functors}
If $F,G : [\set^k,\set]$, and if $\alpha : F \to G$ is a natural
transformation, then $\graph{\alpha}$ is in $RT_k$.
\end{lemma}
\begin{proof}
Clearly, $\graph{\alpha}^*$ is $\omega$-cocontinuous, so
$\graph{\alpha}^* : [\rel^k,\rel]$. Now, suppose $\overline{R :
  \rel(A, B)}$, $\overline{S : \rel(C, D)}$, and $\overline{(\beta,
  \beta') : R \to S}$. We want to show that there exists a morphism
$\epsilon : \alpha^\wedge\overline{R} \to \alpha^\wedge\overline{S}$
such that the diagram on the left below commutes. Since
$\ol{(\beta,\beta') : R \to S}$, there exist $\overline{\gamma : R \to
  S}$ such that each diagram in the middle commutes.
Moreover, since both $h_{\overline{C \times D}} \circ
F(\overline{\beta \times \beta'})$ and $(F\overline{\beta} \times
G\overline{\beta'}) \circ h_{\overline{A \times B}}$ make the diagram
on the right commute, they must be equal.
\begin{figure*}[ht]
\vspace*{-0.1in}
  \hspace*{-0.85in}
  \begin{minipage}[b]{0.25\linewidth}
 {\small    \[
    \begin{tikzcd}
        \alpha^\wedge\overline{R}
        \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{R}}}"]
        \ar[d, "{\epsilon}"']
        & F\overline{A} \times G\overline{B}
        \ar[d, "{F\overline{\beta} \times G\overline{\beta'}}"] \\
        \alpha^\wedge\overline{S}
        \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{S}}}"']
        & F\overline{C} \times G\overline{D}
    \end{tikzcd}
    \]}
\end{minipage}
\begin{minipage}[b]{0.25\linewidth}
{\small    \[
    \begin{tikzcd}
        R_i
        \ar[d, "{\gamma_i}"']
        \ar[r, hookrightarrow, "{\iota_{R_i}}"]
        &A_i \times B_i
        \ar[d, "{\beta_i \times \beta'_i}"] \\
        S_i
        \ar[r, hookrightarrow, "{\iota_{S_i}}"]
        &C_i \times D_i
    \end{tikzcd}
    \]}
\end{minipage}
\begin{minipage}[b]{0.25\linewidth}
{\footnotesize \[
      \begin{tikzcd}[row sep = large]
          F\overline{C}
          &F\overline{C} \times F\overline{D}
          \ar[l, "{\pi_1}"'] \ar[r, "{\pi_2}"]
          &F\overline{D}
          \ar[r, "{\alpha_{\ol{D}}}"]
          &G\overline{D}\\
          &F(\overline{A \times B})
          \ar[u, dashed, "{\exists !}"]
          \ar[ul, "{F\ol{\pi_1} \circ F(\overline{\beta \times \beta'})}"]
          \ar[urr, "{\alpha_{\ol{D}} \circ F\ol{\pi_2} \circ F(\overline{\beta \times \beta'})}"']
      \end{tikzcd}
      \]}
\end{minipage}
\end{figure*}
%{\footnotesize    \[
%    \begin{tikzcd}
%        \alpha^\wedge\overline{R}
%        \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{R}}}"]
%        \ar[d, "{\epsilon}"']
%        & F\overline{A} \times G\overline{B}
%        \ar[d, "{F\overline{\beta} \times G\overline{\beta'}}"] \\
%        \alpha^\wedge\overline{S}
%        \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{S}}}"']
%        & F\overline{C} \times G\overline{D}
%    \end{tikzcd}
%    \]}
%    commutes.
%    Since $\ol{(\beta,\beta') : R \to S}$, there exist $\overline{\gamma : R \to S}$
%    such that each diagram
%{\footnotesize    \[
%    \begin{tikzcd}
%        R_i
%        \ar[d, "{\gamma_i}"']
%        \ar[r, hookrightarrow, "{\iota_{R_i}}"]
%        &A_i \times B_i
%        \ar[d, "{\beta_i \times \beta'_i}"] \\
%        S_i
%        \ar[r, hookrightarrow, "{\iota_{S_i}}"]
%        &C_i \times D_i
%    \end{tikzcd}
%    \]}
%    commutes.
%Moreover, since both $h_{\overline{C \times D}} \circ
%F(\overline{\beta \times \beta'})$ and $(F\overline{\beta} \times
%G\overline{\beta'}) \circ h_{\overline{A \times B}}$ make
%{\footnotesize \[
%      \begin{tikzcd}[row sep = large]
%          F\overline{C}
%          &F\overline{C} \times F\overline{D}
%          \ar[l, "{\pi_1}"'] \ar[r, "{\pi_2}"]
%          &F\overline{D}
%          \ar[r, "{\alpha_{\ol{D}}}"]
%          &G\overline{D}\\
%          &F(\overline{A \times B})
%          \ar[u, dashed, "{\exists !}"]
%          \ar[ul, "{F\ol{\pi_1} \circ F(\overline{\beta \times \beta'})}"]
%          \ar[urr, "{\alpha_{\ol{D}} \circ F\ol{\pi_2} \circ F(\overline{\beta \t%imes \beta'})}"']
%      \end{tikzcd}
%      \]}
%      commute, they must be equal.

\vspace*{-0.1in}

\noindent
We therefore get that the right-hand square in the diagram on the left
below commutes, and thus that the entire diagram does as well.
Finally, by the left-lifting property of $q_{F^\wedge\overline{R}}$
with respect to $\iota_{F^\wedge\overline{S}}$ given by the epi-mono
factorization system, there exists an $\epsilon$ such that the diagram
on the right below commutes.
\begin{figure*}[ht]
  \vspace*{-0.1in}
  \hspace*{-0.5in}
  \begin{minipage}[b]{0.45\linewidth}
{\footnotesize \[
      \begin{tikzcd}
          F\overline{R}
          \ar[d, "{F\overline{\gamma}}"']
          \ar[r, hookrightarrow, "{F\overline{\iota_R}}"]
          \ar[rr, bend left, "{h_{\overline{R}}}"]
          &F(\overline{A \times B})
          \ar[d, "{F(\overline{\beta \times \beta'})}"]
          \ar[r, "{h_{\overline{A \times B}}}"]
          &F\overline{A} \times G\overline{B}
          \ar[d, "{F\overline{\beta} \times F\overline{\beta'}}"] \\
          F\overline{S}
          \ar[r, hookrightarrow, "{F\overline{\iota_S}}"']
          \ar[rr, bend right, "{h_{\overline{S}}}"']
          &F(\overline{C \times D})
          \ar[r, "{h_{\overline{C \times D}}}"']
          &F\overline{C} \times G\overline{D}
      \end{tikzcd}
      \]}
\end{minipage}
  \vspace*{-0.5in}
  \begin{minipage}[b]{0.45\linewidth}
      {\footnotesize
        \[  \vspace*{0.4in}
      \begin{tikzcd}
          F\overline{R}
          \ar[d, "{F\overline{\gamma}}"']
          \ar[r, twoheadrightarrow, "{q_{\alpha^\wedge\overline{R}}}"]
          &\alpha^\wedge\overline{R}
          \ar[d, dashed, "{\epsilon}"]
          \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{R}}}"]
          &F\overline{A} \times G\overline{B}
          \ar[d, "{F\overline{\beta} \times G\overline{\beta'}}"] \\
          F\overline{S}
          \ar[r, twoheadrightarrow, "{q_{\alpha^\wedge\overline{S}}}"']
          &\alpha^\wedge\overline{S}
          \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{S}}}"']
          &F\overline{C} \times G\overline{D}
      \end{tikzcd}
      \]}
\end{minipage}
\end{figure*}

\begin{comment}

Finally, by the left-lifting property of $q_{F^\wedge\overline{R}}$
      with respect to $\iota_{F^\wedge\overline{S}}$ given by the epi-mono
      factorization system, there exists an $\epsilon$ such that the
      following diagram commutes:
      {\footnotesize
        \[
      \begin{tikzcd}
          F\overline{R}
          \ar[d, "{F\overline{\gamma}}"']
          \ar[r, twoheadrightarrow, "{q_{\alpha^\wedge\overline{R}}}"]
          &\alpha^\wedge\overline{R}
          \ar[d, dashed, "{\epsilon}"]
          \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{R}}}"]
          &F\overline{A} \times G\overline{B}
          \ar[d, "{F\overline{\beta} \times G\overline{\beta'}}"] \\
          F\overline{S}
          \ar[r, twoheadrightarrow, "{q_{\alpha^\wedge\overline{S}}}"']
          &\alpha^\wedge\overline{S}
          \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{S}}}"']
          &F\overline{C} \times G\overline{D}
      \end{tikzcd}
      \]}

      \vspace*{-0.1in}
\end{comment}
\end{proof}

\vspace*{0.1in}

If $f : A \to B$ is a morphism in $\set$ then the definition of the
graph relation transformer $\langle f \rangle$ for $f$ as a natural
transformation between $0$-ary functors $A$ and $B$ coincides with its
standard definition.  Graph relation transformers are thus a
reasonable extension of graph relations to functors.
\begin{comment}

\vspace*{0.2in}

If $f : A \to B$ is a function with graph relation $\graph{f} = (A, B,
\graph{f}^*)$, then $\langle \id_{A}, f \rangle : A \to A \times B$
and $\langle \id_{A}, f \rangle\, A = \graph{f}^*$.  Moreover, if
$\iota_{\graph{f}} : \graph{f}^* \hookrightarrow A \times B$ is the
inclusion of $\graph{f}^*$ into $A \times B$ then there is an
isomorphism of subobjects
\[\begin{tikzcd}
A \ar[rr, "{\cong}"] \ar[dr, "{\langle \id_{A}, f \rangle}"']
&&{\graph{f}^*} \ar[dl, "{\iota_{\graph{f}}}"]\\
&A \times B
\end{tikzcd}\]

We also note that if $f : A \to B$ is a function seen as a natural
transformation between 0-ary functors, then $\graph{f}$ is (the 0-ary
relation transformer associated with) the graph relation of $f$.
Indeed, we need to apply Definition~\ref{dfn:graph-nat-transf} with $k
= 0$, i.e., to the degenerate relation $\ast : \rel(\ast, \ast)$.  As
degenerate $0$-ary functors, $A$ and $B$ are constant functors, i.e.,
$A\, \ast = A$ and $B\, \ast = B$.  By the universal property of the
product, there exists a unique $h$ making the diagram
\[ \begin{tikzcd}[row sep = large]
        A
        &A
        \ar[l, equal]
        \ar[r, equal]
        \ar[d, dashed, "{h}"]
        &A
        \ar[r, "{f}"]
        &B \\
        &A \times B
        \ar[ul, "{\pi_1}"] \ar[urr, "{\pi_2}"']
\end{tikzcd}\]
commute. Since $\iota_\ast : \ast \to \ast$ is the identity on $\ast$,
and $A\, \id_{\ast} = \id_{A}$, we have $h_{\ast} = h$.  Moreover,
$h_{\overline{A \times B}} = \langle \id_{A}, f \rangle$ is a
monomorphism in $\set$ because $\id_{A}$ is.  Then,
$\iota_{f^\wedge\ast} = \langle \id_{A}, f \rangle$ and $f^\wedge\ast
= A$, from which we deduce that $\iota_{f^\wedge\ast} f^\wedge\ast =
\langle \id_{A}, f \rangle\, A = \graph{f}^*$. This ensures that the
graph of $f$ as a 0-ary natural transformation coincides with the
graph of $f$ as a morphism in $\set$, and so that
Definition~\ref{dfn:graph-nat-transf} is a reasonable generalization
of Definition~\ref{def:graph}.

Just as the equality relation $\Eq_B$ on a set $B$ coincides with
$\graph{\id_B}$, the graph of the identity on the set, so we can
define the equality relation transformer to be the graph of the
identity natural transformation. This gives

\begin{dfn}
Let $F : [\set^k, \set]$.  The equality relation transformer on $F$ is
defined to be $\Eq_F = \graph{\id_{F}}$. This entails that $Eq_F = (F,
F, \Eq_F^*)$ with $\Eq_F^* = \graph{\id_{F}}^*$.
\end{dfn}
\end{comment}

To prove the IEL, we will need to know that the equality relation
transformer preserves equality relations; see
Equation~\ref{eq:eq-pres-eq} below. This will follow from the next
lemma, which shows how to compute the action of a graph relation
transformer on any graph relation.

\begin{lemma}\label{lem:eq-reln-equalities}
If $\alpha : F \to G$ is a morphism in $[\Set^k, \Set]$
and $f_1: A_1 \to B_1, ..., f_k : A_k \to B_k$,
then $\graph{\alpha}^* \graph{\overline{f}}
= \langle G \ol{f} \circ \alpha_{\ol{A}} \rangle
= \langle \alpha_{\ol{B}} \circ F \ol{f} \rangle$.
\end{lemma}
\begin{proof}
Since $h_{\overline{A \times B}}$ is the unique morphism making the
bottom triangle of the diagram on the left below commute, and since
$h_{\graph{\overline{f}}} = h_{\overline{A \times B}} \circ F
\,\ol{\iota_{\graph{f}}} = h_{\overline{A \times B}} \circ F
\overline{\langle \id_A, f \rangle}$, the universal property of the
product depicted in the diagram on the right gives
$h_{\graph{\overline{f}}} = \langle \id_{F \ol{A}}, \alpha_{\ol{B}}
\circ F\ol{f} \rangle : F \ol{A} \to F \ol{A} \times G \ol{B}$.

\begin{figure*}[ht]
  \vspace*{-0.15in}
%  \hspace*{-0.5in}
  \begin{minipage}[b]{0.45\linewidth}
{\footnotesize
\[\begin{tikzcd}[row sep = large]
        &F\overline{A}
        \ar[d, "{F \overline{\langle \id_A, f \rangle}}" description]
        \ar[dl, equal]
        \ar[dr, "{F\ol{f}}"]\\
        F\overline{A}
        &F(\overline{A \times B})
        \ar[l, "{F\overline{\pi_1}}"']
        \ar[r, "{F\overline{\pi_2}}"]
        \ar[d, "{h_{\overline{A \times B}}}"]
        &F\overline{B}
        \ar[r, "{\alpha_{\ol{B}}}"]
        &G\overline{B}\\
        &F\overline{A} \times G\overline{B}
        \ar[ul, "{\pi_1}"] \ar[urr, "{\pi_2}"']
\end{tikzcd}\]}
\end{minipage}
%  \vspace*{-0.5in}
  \begin{minipage}[b]{0.45\linewidth}
{\footnotesize
\[
      \begin{tikzcd}[row sep = large]
          F\overline{A}
          &F\overline{A} \times G\overline{B}
          \ar[l, "{\pi_1}"'] \ar[r, "{\pi_2}"]
          &G\overline{B}\\
          &F\overline{A}
          \ar[u, dashed, "{\exists !}"]
          \ar[ul, equal]
          \ar[r, "{F\ol{f}}"']
          &F{\ol{B}}
          \ar[u, "{\alpha_{\ol{B}}}"']
      \end{tikzcd}
      \]}
\end{minipage}
\end{figure*}

\vspace*{-0.1in}

\noindent
Moreover, $\langle \id_{F \ol{A}}, \alpha_{\ol{B}} \circ
F\ol{f} \rangle$ is a monomorphism in $\set$ because $\id_{F \ol{A}}$
is, so its epi-mono factorization gives $\iota_{\alpha^\wedge
  \graph{\overline{f}}} = \langle \id_{F \ol{A}}, \alpha_{\ol{B}}
\circ F\ol{f} \rangle$, and thus $\alpha^\wedge \graph{\overline{f}} =
%the domain of $\iota_{\alpha^\wedge \graph{\overline{f}}}$, is equal
%to
F\overline{A}$.  Then $\iota_{\alpha^\wedge
  \graph{\overline{f}}} \alpha^\wedge \graph{\overline{f}} = \langle
\id_{F \ol{A}}, \alpha_{\ol{B}} \circ F\ol{f} \rangle (F \ol{A}) =
\graph{ \alpha_{\ol{B}} \circ F\ol{f} }^*$,
%(where the last equality is by Remark~\ref{rmk:graph-fn}).
%We therefore conclude that
so that $\graph{ \alpha }^* \graph{ \overline{f} } = (F\overline{A},
G\overline{B}, \iota_{\alpha^\wedge \graph{\overline{f}}}\,
\alpha^\wedge \graph{\overline{f}}) = (F\overline{A}, G\overline{B},
\graph{ \alpha_{\ol{B}} \circ F\ol{f} }^*) = \graph{ \alpha_{\ol{B}}
  \circ F\ol{f} }$.  Finally, $\alpha_{\ol{B}} \circ F\ol{f} = G\ol{f}
\circ \alpha_{\ol{A}}$ by naturality of $\alpha$.
\end{proof}

The {\em equality relation transformer} on $F : [\set^k,\set]$ is
defined to be $\Eq_F = \graph{\id_{F}}$. Specifically, $Eq_F = (F, F,
\Eq_F^*)$ with $\Eq_F^* = \graph{\id_{F}}^*$, and
Lemma~\ref{lem:eq-reln-equalities} indeed ensures that
\begin{equation}\label{eq:eq-pres-eq}
\Eq^*_F \ol{\Eq_A}
= \graph{\id_F}^* \graph{\id_{\ol{A}}}
= \graph{F \id_{\ol{A}} \circ (\id_F)_{\ol{A}}}
= \graph{\id_{F\ol{A}} \circ \id_{F\ol{A}}}
= \graph{\id_{F\ol{A}}}
= \Eq_{F\ol{A}}
\end{equation}
%$\Eq^*_F \,\ol{\Eq_A} = \Eq_{F\ol{A}}$
for all $\ol{A : \set}$.
\begin{comment}
  \begin{lemma}
$\Eq^*_F \,\ol{\Eq_A}
= \Eq_{F\ol{A}}$ for all $\ol{A : \set}$.
  \end{lemma}
\begin{proof}
We have that\
\begin{equation}\label{eq:eq-pres-eq}
\Eq^*_F \ol{\Eq_A}
= \graph{\id_F}^* \graph{\id_{\ol{A}}}
= \graph{F \id_{\ol{A}} \circ (\id_F)_{\ol{A}}}  %by your Lemma 5
= \graph{\id_{F\ol{A}} \circ \id_{F\ol{A}}}
%= \graph{\id_F \ol{A} \circ \id_{\ol{A}}}
= \graph{\id_{F\ol{A}}}
= \Eq_{F\ol{A}}
\end{equation}
The second identity here is by Lemma~\ref{lem:eq-reln-equalities}.
\end{proof}
\end{comment}
Graph relation transformers in general, and equality relation
transformers in particular, extend to relation environments in the
obvious ways.
%\begin{definition}
Indeed, if $\rho, \rho' : \setenv$ and $f : \rho \to \rho'$, then the
{\em graph relation environment} $\graph{f}$ is defined pointwise by
$\graph{f} \phi = \graph{f \phi}$ for every $\phi$. This entails that
$\pi_1 \graph{f} = \rho$ and $\pi_2 \graph{f} = \rho'$. In particular,
the {\em equality relation environment} $\Eq_\rho$ is defined to be
$\graph{\id_{\rho}}$. This entails that $\Eq_\rho \phi = \Eq_{\rho
  \phi}$ for every $\phi$.
%\end{definition}
With these definitions in hand, we can state and prove both an
Identity Extension Lemma and a Graph Lemma for our calculus.
\begin{thm}[IEL]\label{thm:iel}
  If $\rho : \setenv$ and $\Gamma; \Phi \vdash F$ then
  $\relsem{\Gamma; \Phi \vdash F} \Eq_\rho = \Eq_{\setsem{\Gamma;
      \Phi \vdash F}\rho}$.
\end{thm}
The proof is by induction on the structure of $F$. Only the
application and fixpoint cases are non-routine. Both use
Lemma~\ref{lem:eq-reln-equalities}. The latter also uses the
observation that, for every $n \in \nat$, the following intermediate
results can be proved by simultaneous induction with
Theorem~\ref{thm:iel}: For any $H$, $\rho$, $A$, and any subformula
$J$ of $H$, both
%\begin{equation}\label{eq:iel-fix-point-intermediate1}
$T^n_{H,\Eq_{\rho}} K_0\, \ol{\Eq_A}
%{\setsem{\Gamma; \Phi \vdash F}\rho}}
= (\Eq_{(T^\set_{H,\rho})^n K_0})^*
\ol{\Eq_A}$\;
  %{\setsem{\Gamma; \Phi \vdash F}\rho}}$\;
%\end{equation}
and\;
%\begin{equation}\label{eq:iel-fix-point-intermediate2}
%\begin{split}
\[\begin{array}{ll}
&  \relsem{\Gamma; \Phi, \phi, \ol{\alpha} \vdash J} \Eq_{\rho} [\phi
  := T^{n}_{H,\Eq_{\rho}} K_0] \overline{[\alpha := \Eq_A]}\\
%{\setsem{\Gamma; \Phi \vdash F}\rho}]}\\
=  &
\relsem{\Gamma; \Phi, \phi,
  \ol{\alpha} \vdash J} \Eq_{\rho} [\phi := \Eq_{(T^\set_{H,\rho})^n K_0}]
\overline{[\alpha := \Eq_A]}
%         {\setsem{\Gamma; \Phi \vdash F}\rho}]}.
\end{array}\]
%\end{split}
%\end{equation}
hold. With these results in hand, the proof is straightforward. A
detailed proof is given in the accompanying anonymous supplementary
material.

\begin{comment}
\begin{proof}
By induction on the structure of $\tau$.
\begin{itemize}
\item $\relsem{\Gamma; \emptyset \vdash v}\Eq_{\rho} = \Eq_{\rho} v =
  \Eq_{\rho v} = \Eq_{\setsem{\Gamma; \emptyset \vdash v}\rho}$ where
  $v \in \Gamma$.
\item By definition, $\relsem{\Gamma; \emptyset \vdash
  \Nat^{\overline\alpha} \,F\,G} \Eq_{\rho}$ is the relation on
  $\setsem{\Gamma; \emptyset \vdash \Nat^{\overline\alpha} \,F\,G}
  \rho$ relating $t$ and $t'$ if, for all ${R_1 :
    \rel(A_1,B_1)},...,{R_k : \rel(A_k,B_k)}$, $(t_{\overline{A}},
  t'_{\overline{B}})$ is a morphism from $\relsem{\Gamma;
    \overline\alpha \vdash F} \Eq_{\rho}\overline{[\alpha := R]}$ to
  $\relsem{\Gamma ; \overline\alpha \vdash G}
  \Eq_{\rho}\overline{[\alpha := R]}$ in $\rel$.  To prove that this
  is equal to $\Eq_{\setsem{\Gamma; \emptyset \vdash
      \Nat^{\overline\alpha} \,F\,G} \rho}$ we need to show that
  $(t_{\overline{A}}, t'_{\overline{B}})$ is a morphism from
  $\relsem{\Gamma; \overline\alpha \vdash F}
  \Eq_{\rho}\overline{[\alpha := R]}$ to $\relsem{\Gamma ;
    \overline\alpha \vdash G} \Eq_{\rho}\overline{[\alpha := R]}$ in
  $\rel$ for all ${R_1 : \rel(A_1,B_1)},...,{R_k : \rel(A_k,B_k)}$ if
  and only if $t = t'$ and $(t_{\overline{A}}, t_{\overline{B}})$ is a
  morphism from $\relsem{\Gamma; \overline\alpha \vdash F}
  \Eq_{\rho}\overline{[\alpha := R]}$ to $\relsem{\Gamma ; \overline
    \alpha \vdash G} \Eq_{\rho}\overline{[\alpha := R]}$ in $\rel$ for
  all ${R_1 : \rel(A_1,B_1)}, ...,$ ${R_k : \rel(A_k,B_k)}$. The only
  interesting part of this equivalence is to show that if
  $(t_{\overline{A}}, t'_{\overline{B}})$ is a morphism from
  $\relsem{\Gamma; \overline\alpha \vdash F}
  \Eq_{\rho}\overline{[\alpha := R]}$ to $\relsem{\Gamma ;
    \overline\alpha \vdash G} \Eq_{\rho}\overline{[\alpha := R]}$ in
  $\rel$ for all ${R_1 : \rel(A_1,B_1),}$ $...,{R_k : \rel(A_k,B_k)}$,
  then $t = t'$.  By hypothesis, $(t_{\overline{A}},
  t'_{\overline{A}})$ is a morphism from $\relsem{\Gamma;
    \overline\alpha \vdash F} \Eq_{\rho}\overline{[\alpha :=
      \Eq_{A}]}$ to $\relsem{\Gamma ; \overline\alpha \vdash G}
  \Eq_{\rho}\overline{[\alpha := \Eq_{A}]}$ in $\rel$ for all
  $A_1\,...\,A_k : \set$. By the induction hypothesis, it is therefore
  a morphism from $\Eq_{\setsem{\Gamma; \overline\alpha \vdash F}
    \rho\overline{[\alpha := A]}}$ to $\Eq_{\setsem{\Gamma ;
      \overline\alpha \vdash G} \rho\overline{[\alpha := A]}}$ in
  $\rel$. This means that, for every $x : \Eq_{\setsem{\Gamma;
      \overline\alpha \vdash F} \rho\overline{[\alpha := A]}}$,
  $t_{\overline{A}}x = t'_{\overline{A}}x$.  Then, by extensionality,
  $t = t'$.
\item $\relsem{\Gamma; \Phi \vdash \zerot} \Eq_{\rho} = 0_\rel =
  \Eq_{0_\set} = \Eq_{\setsem{\Gamma; \Phi \vdash \zerot}\rho}$
\item $\relsem{\Gamma; \Phi \vdash \onet} \Eq_{\rho} = 1_\rel =
  \Eq_{1_\set} = \Eq_{\setsem{\Gamma; \Phi \vdash \onet}\rho}$
\item The application case is proved by the following sequence of
  equalities, where the second equality is by the induction hypothesis
  and the definition of the relation environment $\Eq_\rho$, the third
  is by the definition of application of relation transformers, and
  the fourth is by Lemma~\ref{lem:eq-reln-equalities}:
\[
\begin{split}
\relsem{\Gamma; \Phi \vdash \phi\ol{\tau}}\Eq_{\rho} &=
(\Eq_{\rho}\phi)\ol{\relsem{\Gamma; \Phi \vdash \tau}
\Eq_{\rho}}\\
&= \Eq_{\rho \phi}\, \ol{\Eq_{\setsem{\Gamma; \Phi \vdash \tau}
  \rho}}\\
&= (\Eq_{\rho \phi})^* \,\ol{\Eq_{\setsem{\Gamma; \Phi \vdash \tau}
  \rho}}\\
&= \Eq_{(\rho \phi) \,\ol{\setsem{\Gamma; \Phi \vdash \tau} \rho}}\\
&= \Eq_{\setsem{\Gamma; \Phi \vdash \phi\ol{\tau}}\rho}
\end{split}
\]
\item The fixpoint case is proven by the sequence of equalities
\[
\begin{split}
\relsem{\Gamma; \Phi \vdash (\mu \phi.\lambda
  \ol{\alpha}. H)\ol{\tau}}\Eq_{\rho} 
&=(\mu {T_{\Eq_{\rho}}}) \,\ol{\relsem{\Gamma; \Phi \vdash \tau}\Eq_{\rho}}\\ 
&= \colim{n \in \nat}{T^n_{\Eq_{\rho}} K_0}\, \ol{\relsem{\Gamma; \Phi
  \vdash \tau}\Eq_{\rho}}\\
&= \colim{n \in \nat}{ T^n_{\Eq_{\rho}} K_0 \,\ol{\Eq_{\setsem{\Gamma;
    \Phi \vdash \tau}\rho}}}\\
&= \colim{n \in \nat}{(\Eq_{(T^\set_\rho)^n K_0})^*
  \ol{\Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}}}\\
&= \colim{n \in \nat}{\Eq_{(T^\set_\rho)^n K_0 \,\ol{\setsem{\Gamma;
        \Phi \vdash \tau}\rho}}}\\ 
&= \Eq_{\colim{n \in \nat}{ (T^\set_{\rho})^n K_0\,
    \ol{\setsem{\Gamma; \Phi \vdash \tau}\rho}}}\\
&= \Eq_{\setsem{\Gamma; \Phi \vdash (\mu \phi.\lambda
      \ol{\alpha}. H)\ol{\tau}}\rho}
\end{split}
\]
Here, the third equality is by induction hypothesis, the fifth is by
Lemma~\ref{lem:eq-reln-equalities} and the fourth equality is because,
for every $n \in \nat$, the following two statements can be proved by
simultaneous induction:
\begin{equation}\label{eq:iel-fix-point-intermediate1}
T^n_{\Eq_{\rho}} K_0\, \ol{\Eq_{\setsem{\Gamma; \Phi \vdash
      \tau}\rho}} = (\Eq_{(T^\set_\rho)^n K_0})^*
\ol{\Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}}
\end{equation}
and
\begin{equation}\label{eq:iel-fix-point-intermediate2}
\begin{split}
  \relsem{\Gamma; \Phi, \phi, \ol{\alpha} \vdash H}
\Eq_{\rho} [\phi := 
 & T^{n}_{\Eq_{\rho}} K_0] \overline{[\alpha :=
    \Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}]} \\
=\;\; & \relsem{\Gamma; \Phi, \phi, \ol{\alpha} \vdash H} \Eq_{\rho} [\phi
  := \Eq_{(T^\set_\rho)^n K_0}] \overline{[\alpha :=
    \Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}]}
\end{split}
\end{equation}
We prove~\eqref{eq:iel-fix-point-intermediate1}.  The case $n=0$ is
trivial, because $T^0_{\Eq_{\rho}} K_0 = K_0$ and
$(T^\set_\rho)^0 K_0 = K_0$; the inductive step is
proved by the following sequence of equalities:
\[
\begin{split}
T^{n+1}_{\Eq_{\rho}} K_0\, \overline{\Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}}
&= T^\rel_{\Eq_{\rho}} (T^{n}_{\Eq_{\rho}} K_0)
\overline{\Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}} \\ 
&= \relsem{\Gamma; \Phi, \phi, \ol{\alpha} \vdash H} \Eq_{\rho} [\phi
  := T^{n}_{\Eq_{\rho}} K_0] \overline{[\alpha :=
    \Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}]} \\ 
&= \relsem{\Gamma; \Phi, \phi, \ol{\alpha} \vdash H} \Eq_{\rho} [\phi
  := \Eq_{(T^\set_\rho)^n K_0}] \overline{[\alpha :=
    \Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}]} \\ 
&= \relsem{\Gamma; \Phi, \phi, \ol{\alpha} \vdash H} \Eq_{\rho [\phi
    := (T^\set_\rho)^n K_0] \overline{[\alpha :=
      \setsem{\Gamma; \Phi \vdash \tau}\rho]}} \\ 
&= \Eq_{\setsem{\Gamma; \Phi, \phi, \ol{\alpha} \vdash H} \rho [\phi
    := (T^\set_\rho)^n K_0] \overline{[\alpha :=
      \setsem{\Gamma; \Phi \vdash \tau}\rho]}} \\ 
&= \Eq_{(T^\set_\rho)^{n+1} K_0 \overline{\setsem{\Gamma; \Phi
      \vdash \tau}\rho}} \\ 
&= (\Eq_{(T^\set_\rho)^{n+1} K_0})^*\, \overline{\Eq_{\setsem{\Gamma;
      \Phi \vdash \tau}\rho}} 
\end{split}
\]
Here, the third equality is by~\eqref{eq:iel-fix-point-intermediate2},
the fifth by the induction hypothesis on $H$, and the last by
Lemma~\ref{lem:eq-reln-equalities}.  We prove the induction step
of~\eqref{eq:iel-fix-point-intermediate2} by structural induction on
$H$: the only interesting case, though, is when $\phi$ is applied,
i.e., for $H = \phi \ol{\sigma}$, which is proved by the sequence of
equalities:
\[
\begin{split}
& \relsem{\Gamma; \Phi, \phi, \ol{\alpha} \vdash \phi
    \ol{\sigma}} \Eq_{\rho} [\phi := T^{n}_{\Eq_{\rho}} K_0]
  \overline{[\alpha := \Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}]}
  \\
&= T^{n}_{\Eq_{\rho}} K_0\, \overline{\relsem{\Gamma; \Phi,
      \phi, \ol{\alpha} \vdash \sigma} \Eq_{\rho} [\phi :=
      T^{n}_{\Eq_{\rho}} K_0] \overline{[\alpha :=
        \Eq_{\setsem{\Gamma; \Phi \vdash \tau} \rho}]}} \\ 
&= T^{n}_{\Eq_{\rho}} K_0\, \overline{\relsem{\Gamma; \Phi,
      \phi, \ol{\alpha} \vdash \sigma} \Eq_{\rho} [\phi :=
      \Eq_{(T^\set_\rho)^{n} K_0}] \overline{[\alpha :=
        \Eq_{\setsem{\Gamma; \Phi \vdash \tau} \rho}]}} \\ 
&= T^{n}_{\Eq_{\rho}} K_0\, \overline{\relsem{\Gamma; \Phi,
      \phi, \ol{\alpha} \vdash \sigma} \Eq_{\rho [\phi := (T^\set_\rho)^{n}
        K_0] \overline{[\alpha := \setsem{\Gamma; \Phi \vdash
            \tau} \rho]}}} \\ 
&= T^{n}_{\Eq_{\rho}} K_0\, \overline{\Eq_{\setsem{\Gamma;
        \Phi, \phi, \ol{\alpha} \vdash \sigma} \rho [\phi :=
        (T^\set_\rho)^{n} K_0] \overline{[\alpha :=
          \setsem{\Gamma; \Phi \vdash \tau} \rho]}}} \\ 
&= (\Eq_{(T^\set_\rho)^{n} K_0})^* \,\overline{\Eq_{\setsem{\Gamma;
        \Phi, \phi, \ol{\alpha} \vdash \sigma} \rho [\phi :=
        (T^\set_\rho)^{n} K_0] \overline{[\alpha :=
          \setsem{\Gamma; \Phi \vdash \tau} \rho]}}} \\ 
&= (\Eq_{(T^\set_{\rho})^{n} K_0})^* \overline{\relsem{\Gamma;
      \Phi, \phi, \ol{\alpha} \vdash \sigma} \Eq_{\rho} [\phi :=
      \Eq_{(T^\set_{\rho})^{n} K_0}] \overline{[\alpha :=
        \Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}]}} \\ 
&= \relsem{\Gamma; \Phi, \phi, \ol{\alpha} \vdash \phi \ol{\sigma}}
  \Eq_{\rho} [\phi := \Eq_{(T^\set_{\rho})^{n} K_0}]
  \overline{[\alpha := \Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}]} 
\end{split}
\]
Here, the second equality is by the induction hypothesis
for~\eqref{eq:iel-fix-point-intermediate2} on the $\sigma$s, the
fourth is by the induction hypothesis for Theorem~\ref{thm:iel} on the
$\sigma$s, and the fifth is by the induction hypothesis on $n$
for~\eqref{eq:iel-fix-point-intermediate1}.
\item $\relsem{\Gamma; \Phi \vdash \sigma + \tau} \Eq_{\rho} =
  \relsem{\Gamma; \Phi \vdash \sigma} \Eq_{\rho} + \relsem{\Gamma;
    \Phi \vdash \tau} \Eq_{\rho} = \Eq_{\setsem{\Gamma; \Phi \vdash
      \sigma}\rho} + \Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho} =
  \Eq_{\setsem{\Gamma; \Phi \vdash \sigma}\rho + \setsem{\Gamma; \Phi
      \vdash \tau}\rho} = \Eq_{\setsem{\Gamma; \Phi \vdash \sigma +
      \tau}\rho}$
\item $\relsem{\Gamma; \Phi \vdash \sigma \times \tau} \Eq_{\rho} =
  \relsem{\Gamma; \Phi \vdash \sigma}\Eq_{\rho} \times \relsem{\Gamma;
    \Phi \vdash \tau}\Eq_{\rho} = \Eq_{\setsem{\Gamma; \Phi \vdash
      \sigma}\rho} \times \Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}
  = \Eq_{\setsem{\Gamma; \Phi \vdash \sigma}\rho \times
    \setsem{\Gamma; \Phi \vdash \tau}\rho} = \Eq_{\setsem{\Gamma;
      \Phi \vdash \sigma \times \tau}\rho}$
\end{itemize}
\end{proof}
\end{comment}

%It follows from Theorem~\ref{thm:iel} that $\relsem{\Gamma \vdash
%  \sigma \to \tau} \Eq_{\rho} = \Eq_{\setsem{\Gamma \vdash \sigma \to
%    \tau}\rho}$, as expected.  Moreover,

%\vspace*{0.05in}
%
%The Graph Lemma appropriate to our setting is now easily obtained:
\begin{lemma}[Graph Lemma]\label{lem:graph}
If $\rho, \rho' : \setenv$, $f : \rho \to \rho'$, and $\Gamma; \Phi
\vdash F$, then $\graph{\setsem{\Gamma; \Phi \vdash F} f} =
\relsem{\Gamma; \Phi \vdash F}\graph{f}$.
\end{lemma}
\begin{proof}
Applying Lemma~\ref{lem:rel-transf-morph} to the morphisms $(f,
\id_{\rho'}) : \graph{f} \to \Eq_{\rho'}$ and $(\id_{\rho}, f) :
\Eq_{\rho} \to \graph{f}$ of relation environments gives that
$(\setsem{\Gamma; \Phi \vdash F}f, \setsem{\Gamma; \Phi \vdash
  F}\id_{\rho'}) = \relsem{\Gamma; \Phi \vdash F} (f, \id_{\rho'}) :
\relsem{\Gamma; \Phi \vdash F}\graph{f} \to \relsem{\Gamma; \Phi
  \vdash F}\Eq_{\rho'}$ and $(\setsem{\Gamma; \Phi \vdash
  F}\id_{\rho}, \setsem{\Gamma; \Phi \vdash F}f) = \relsem{\Gamma;
  \Phi \vdash F} (\id_{\rho}, f) : \relsem{\Gamma; \Phi \vdash
  F}\Eq_{\rho} \to \relsem{\Gamma; \Phi \vdash F}\graph{f}$.
Expanding the first equation gives that if $(x,y) \in \relsem{\Gamma;
  \Phi \vdash F}\graph{f}$ then $(\setsem{\Gamma; \Phi \vdash F} f\,
x, \setsem{\Gamma; \Phi \vdash F}\id_{\rho'}\, y) \in \relsem{\Gamma;
  \Phi \vdash F}\Eq_{\rho'}$.  Then $\setsem{\Gamma; \Phi \vdash
  F}\id_{\rho'}\, y = \id_{\setsem{\Gamma; \Phi \vdash F}\rho'}\, y =
y$ and $\relsem{\Gamma; \Phi \vdash F}\Eq_{\rho'} =
\Eq_{\setsem{\Gamma; \Phi \vdash F}\rho'}$, so if $(x,y) \in
\relsem{\Gamma; \Phi \vdash F}\graph{f}$ then $(\setsem{\Gamma; \Phi
  \vdash F} f\, x, y) \in \Eq_{\setsem{\Gamma; \Phi \vdash F}\rho'}$,
i.e., $\setsem{\Gamma; \Phi \vdash F} f\, x = y$, i.e., $(x, y) \in
\graph{\setsem{\Gamma; \Phi \vdash F} f}$.  So, we have that
$\relsem{\Gamma; \Phi \vdash F}\graph{f} \subseteq
\graph{\setsem{\Gamma; \Phi \vdash F}f}$.  Expanding the second
equation gives that if $x \in \setsem{\Gamma; \Phi \vdash F}\rho$ then
$(\setsem{\Gamma; \Phi \vdash F}\id_{\rho}\, x, \setsem{\Gamma; \Phi
  \vdash F} f\, x) \in \relsem{\Gamma; \Phi \vdash F}\graph{f}$.  Then
$\setsem{\Gamma; \Phi \vdash F}\id_{\rho}\, x = \id_{\setsem{\Gamma;
    \Phi \vdash F}\rho} x = x$, so for any $x \in \setsem{\Gamma; \Phi
  \vdash F}\rho$ we have that $(x, \setsem{\Gamma; \Phi \vdash F}f\,
x) \in \relsem{\Gamma; \Phi \vdash F}\graph{f}$.  Moreover, $x \in
\setsem{\Gamma; \Phi \vdash F}\rho$ if and only if $(x,
\setsem{\Gamma; \Phi \vdash F} f\, x) \in \graph{\setsem{\Gamma; \Phi
    \vdash F}f}$ and, if $x \in \setsem{\Gamma; \Phi \vdash F}\rho$
then $(x, \setsem{\Gamma; \Phi \vdash F} f\, x) \in \relsem{\Gamma;
  \Phi \vdash F} \graph{f}$, so if $(x, \setsem{\Gamma; \Phi \vdash F}
f\, x) \in \graph{\setsem{\Gamma; \Phi \vdash F}f}$ then $(x,
\setsem{\Gamma; \Phi \vdash F} f\, x) \in \relsem{\Gamma; \Phi \vdash
  F} \graph{f}$, i.e., $\graph{\setsem{\Gamma; \Phi \vdash F}f}
\subseteq \relsem{\Gamma; \Phi \vdash F} \graph{f}$.

\end{proof}

\begin{comment}

\begin{proof}
First observe that $(f, \id_{\rho'}) : \graph{f} \to \Eq_{\rho'}$ and
$(\id_{\rho}, f) : \Eq_{\rho} \to \graph{f}$ are morphisms of relation
environments.  Applying Lemma~\ref{lem:rel-transf-morph} to each of
these observations gives that
\begin{equation}\label{eq:graph-one}
(\setsem{\Gamma; \Phi \vdash F}f, \setsem{\Gamma; \Phi \vdash
    F}\id_{\rho'}) = \relsem{\Gamma; \Phi \vdash F} (f, \id_{\rho'}) :
  \relsem{\Gamma; \Phi \vdash F}\graph{f} \to \relsem{\Gamma; \Phi
    \vdash F}\Eq_{\rho'}
\end{equation}
and
\begin{equation}\label{eq:graph-two}
(\setsem{\Gamma; \Phi \vdash F}\id_{\rho}, \setsem{\Gamma; \Phi \vdash F}f)
= \relsem{\Gamma; \Phi \vdash F} (\id_{\rho}, f)
: \relsem{\Gamma; \Phi \vdash F}\Eq_{\rho} \to \relsem{\Gamma; \Phi \vdash F}\graph{f}
\end{equation}
Expanding Equation~\ref{eq:graph-one} gives that if
$(x,y) \in \relsem{\Gamma; \Phi \vdash F}\graph{f}$
then
\[(\setsem{\Gamma; \Phi \vdash F} f\, x, \setsem{\Gamma; \Phi \vdash
  F}\id_{\rho'}\, y) \in \relsem{\Gamma; \Phi \vdash F}\Eq_{\rho'}\]
Observe that $\setsem{\Gamma; \Phi \vdash F}\id_{\rho'}\, y =
\id_{\setsem{\Gamma; \Phi \vdash F}\rho'}\, y = y$ and
$\relsem{\Gamma; \Phi \vdash F}\Eq_{\rho'} = \Eq_{\setsem{\Gamma; \Phi
    \vdash F}\rho'}$. So, if $(x,y) \in \relsem{\Gamma; \Phi \vdash
  F}\graph{f}$ then $(\setsem{\Gamma; \Phi \vdash F} f\, x, y) \in
\Eq_{\setsem{\Gamma; \Phi \vdash F}\rho'}$, i.e., $\setsem{\Gamma;
  \Phi \vdash F} f\, x = y$, i.e., $(x, y) \in \graph{\setsem{\Gamma;
    \Phi \vdash F} f}$.  So, we have that $\relsem{\Gamma; \Phi \vdash
  F}\graph{f} \subseteq \graph{\setsem{\Gamma; \Phi \vdash F}f}$

Expanding Equation~\ref{eq:graph-two} gives that, for any
$x \in \setsem{\Gamma; \Phi \vdash F}\rho$,
then
\[
(\setsem{\Gamma; \Phi \vdash F}\id_{\rho}\, x, \setsem{\Gamma; \Phi
  \vdash F} f\, x) \in \relsem{\Gamma; \Phi \vdash F}\graph{f} 
\]
Observe that $\setsem{\Gamma; \Phi \vdash F}\id_{\rho}\, x =
\id_{\setsem{\Gamma; \Phi \vdash F}\rho} x = x$ so, for any $x \in
\setsem{\Gamma; \Phi \vdash F}\rho$, we have that $(x, \setsem{\Gamma;
  \Phi \vdash F}f\, x) \in \relsem{\Gamma; \Phi \vdash F}\graph{f}$.
Moreover, $x \in \setsem{\Gamma; \Phi \vdash F}\rho$ if and only if
$(x, \setsem{\Gamma; \Phi \vdash F} f\, x) \in \graph{\setsem{\Gamma;
    \Phi \vdash F}f}$ and, if $x \in \setsem{\Gamma; \Phi \vdash
  F}\rho$ then $(x, \setsem{\Gamma; \Phi \vdash F} f\, x) \in
\relsem{\Gamma; \Phi \vdash F} \graph{f}$, so if $(x, \setsem{\Gamma;
  \Phi \vdash F} f\, x) \in \graph{\setsem{\Gamma; \Phi \vdash F}f}$
then $(x, \setsem{\Gamma; \Phi \vdash F} f\, x) \in \relsem{\Gamma;
  \Phi \vdash F} \graph{f}$, i.e., $\graph{\setsem{\Gamma; \Phi \vdash
    F}f} \subseteq \relsem{\Gamma; \Phi \vdash F} \graph{f}$. We
conclude that $\relsem{\Gamma; \Phi \vdash F}\graph{f} =
\graph{\setsem{\Gamma; \Phi \vdash F}f}$ as desired.
\end{proof}
\end{comment}

\vspace*{-.2in}

\section{Interpreting Terms}\label{sec:term-interp}

If $\Delta = x_1 : F_1,...,x_n : F_n$ is a term context for $\Gamma$
and $\Phi$, then the interpretations $\setsem{\Gamma;\Phi \vdash \Delta}$ and
$\relsem{\Gamma;\Phi \vdash \Delta}$ are defined by
\[\begin{array}{lll}
\setsem{\Gamma;\Phi \vdash \Delta} & = & \setsem{\Gamma;\Phi \vdash
  F_1} \times ... \times \setsem{\Gamma;\Phi \vdash F_n}\\ 
\relsem{\Gamma;\Phi \vdash \Delta} & = & \relsem{\Gamma;\Phi \vdash
  F_1} \times ... \times \relsem{\Gamma;\Phi \vdash F_n}\\ 
\end{array}\]
Every well-formed term $\Gamma;\Phi~|~\Delta \vdash t : F$ then
has, for every $\rho \in \setenv$, set interpretations
$\setsem{\Gamma;\Phi~|~\Delta \vdash t : F}\rho$ as natural
transformations from $\setsem{\Gamma; \Phi \vdash \Delta}\rho$ to
$\setsem{\Gamma; \Phi \vdash F}\rho$, and, for every $\rho \in
\relenv$, relational interpretations $\relsem{\Gamma;\Phi~|~\Delta
  \vdash t : F}\rho$ as natural transformations from
$\relsem{\Gamma; \Phi \vdash \Delta}\rho$ to $\relsem{\Gamma; \Phi
  \vdash F}\rho$. These are given in the next (two) definitions.

\begin{dfn}\label{def:set-interp}
If $\rho$ is a set (resp., relation) environment and
$\Gamma;\Phi~|~\Delta \vdash t : F$ then
$\setsem{\Gamma;\Phi~|~\Delta \vdash t : F}\rho$ (resp.,
$\relsem{\Gamma;\Phi~|~\Delta \vdash t : F}\rho$) is defined as in
Figure~\ref{fig:term-sem}, where $\mathsf D$ is either $\set$ or
$\rel$ as appropriate.
\end{dfn}

\subsection{Basic Properties of Term Interpretations}

The interpretations in Definition~\ref{def:set-interp} respect
weakening, i.e., a term and its weakenings all have the same set and
relational interpretations. Specifically, for any $\rho \in \setenv$,
$\setsem{\Gamma;\Phi \,|\, \Delta, x : F \vdash t : G}\rho =
(\setsem{\Gamma;\Phi \,|\, \Delta \vdash t : G}\rho) \circ
\pi_{\Delta}$, where $\pi_{\Delta}$ is the projection
$\setsem{\Gamma;\Phi \vdash \Delta, x : F} \to
\setsem{\Gamma;\Phi \vdash \Delta}$. A similar result holds for
relational interpretations.

\begin{figure*}
\hspace*{-3.1in}
\resizebox{0.46\linewidth}{!}{
\begin{minipage}[t]{0.5\textwidth}
\[\begin{array}{lll}
\dsem{\Gamma;\emptyset \,|\, \Delta,x :F \vdash x : F} \rho& = &
\pi_{|\Delta|+1}\\
%\dsem{\Gamma;\emptyset \,|\, \Delta \vdash \lambda x.t : \sigma \to \tau}\rho &
%= & \curry (\dsem{\Gamma;\emptyset \,|\, \Delta, x : \sigma \vdash t :
%  \tau}\rho)\\ 
%\dsem{\Gamma;\emptyset \,|\, \Delta \vdash st: \tau} \rho & = &
%\eval \circ
% \langle \dsem{\Gamma;\emptyset \,|\, \Delta \vdash s: \sigma \to
%  \tau}\rho, \dsem{\Gamma;\emptyset \,|\, \Delta \vdash t: \sigma}\rho
%\rangle\\
\dsem{\Gamma;\Phi \,|\, \Delta \vdash L_{\overline \alpha} x.t : \Nat^{\overline
    \alpha} \,F \,G}\rho & = &  \curry (\dsem{\Gamma;\Phi,\overline \alpha
  \,|\, \Delta, x : F \vdash t: G}\rho[\overline{\alpha := \_}])\\
\dsem{\Gamma;\Phi \,|\, \Delta \vdash t_{\overline K} s:
  G [\overline{\alpha := K}]}\rho & = & \eval \circ \langle
  \lambda d.\,(\dsem{\Gamma;\Phi \,|\, \Delta \vdash t :
  \Nat^{\overline{\alpha}} \,F \,G}\rho\; d)_{\overline{\dsem{\Gamma;\Phi
      \vdash K}\rho}},\\ 
 & & \hspace*{0.5in} \dsem{\Gamma;\Phi \,|\,
    \Delta \vdash s: F [\overline{\alpha := K}]}\rho \rangle\\ 
& & \\
%% \color{red} \mbox{Add rules for } \forall \mbox{ if we include it} & & \\
\dsem{\Gamma;\Phi \,|\, \Delta,x :F \vdash x : F} \rho& = &
\pi_{|\Delta|+1}\\
\dsem{\Gamma;\Phi \,|\, \Delta \vdash \bot_F t : F} \rho& = &
!^0_{\dsem{\Gamma;\Phi \vdash F}\rho} \circ
  \dsem{\Gamma;\Phi~|~\Delta \vdash t : \zerot}\rho, \mbox{ where } \\
 & & \hspace*{0.1in} !^0_{\dsem{\Gamma;\Phi \vdash F}\rho}
\mbox{ is the unique morphism from } 0\\
 & & \hspace*{0.1in} \mbox{ to } \dsem{\Gamma;\Phi \vdash F}\rho\\
\dsem{\Gamma;\Phi \,|\, \Delta \vdash \top : \onet}\rho & = &
!^{\dsem{\Gamma;\Phi\vdash \Delta}\rho}_1, \mbox{ where }
!^{\dsem{\Gamma;\Phi\vdash \Delta}\rho}_1\\ 
& & \hspace*{0.1in} \mbox{ is the unique morphism from }
\dsem{\Gamma;\Phi\vdash \Delta}\rho \mbox{ to } 1\\ 
\dsem{\Gamma;\Phi \,|\, \Delta \vdash (s,t) : F \times G} \rho& = &
\dsem{\Gamma;\Phi \,|\, \Delta \vdash s: F} \rho\times
\dsem{\Gamma;\Phi \,|\, \Delta \vdash t : G} \rho\\
\dsem{\Gamma;\Phi \,|\, \Delta \vdash \pi_1 t : F} \rho& = &
\pi_1 \circ \dsem{\Gamma;\Phi \,|\, \Delta \vdash t : F \times G}\rho\\
\dsem{\Gamma;\Phi \,|\, \Delta \vdash \pi_2 t : G}\rho & = &
\pi_2 \circ \dsem{\Gamma;\Phi \,|\, \Delta \vdash t : F \times
  G} \rho\\
\dsem{\Gamma;\Phi~|~\Delta \vdash \case{t}{x \mapsto l}{y \mapsto r} :
  K}\rho & = & \eval \circ \langle \curry \,[\dsem{\Gamma;\Phi
    \,|\, \Delta, x : F \vdash l : K}\rho,\\
   & & \hspace*{0.79in} \dsem{\Gamma;\Phi \,|\, \Delta, y
    : G \vdash r : K}\rho],\\
   & &  \hspace*{0.5in} \dsem{\Gamma;\Phi \,|\, \Delta \vdash t :
  F + G} \rho\rangle\\   
\dsem{\Gamma;\Phi \,|\, \Delta \vdash \inl \,s: F + G} \rho& = &
\inl \circ \dsem{\Gamma;\Phi \,|\, \Delta \vdash s: F}\rho\\
\dsem{\Gamma;\Phi \,|\, \Delta \vdash \inr \,t: F + G}\rho & = & 
\inr \circ \dsem{\Gamma;\Phi \,|\, \Delta \vdash t : G}\rho\\
\dsem{\Gamma;\Phi \,|\, \Delta \vdash \map^{\ol{F},\ol{G}}_H
\,\ol t: \Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\,H[\ol{\phi
      :=_{\ol{\beta}} G}]} \rho
& = & \lambda d\, \ol{C}.\,
\dsem{\Gamma; \ol{\phi},\ol{\gamma}\vdash H}\\ 
\hspace*{0.79in}  & & \hspace*{0.1in}\id_{\rho[\ol{\gamma :=
      C}]}[\ol{\phi := \lambda
    \ol{B}. (\dsem{\Gamma;\Phi\,|\,\Delta \vdash t : 
    \Nat^{\ol{\beta},\ol{\gamma}}\,F\,G}\rho\,d)_{\ol{B}\,\ol{C}}}]\\
\llbracket \Gamma;\emptyset \,|\, \emptyset \vdash \tin_H :
Nat^{\ol{\beta},\ol{\gamma}} \, H[\phi := (\mu \phi.\lambda {\overline
    \alpha}.H){\overline \beta}][\ol{\alpha := \beta}] & = &
\lambda d\,\ol{B}\,
\ol{C}.\,(\mathit{in}_{{T}^X_{H,\rho[\ol{\gamma :=
        C}]}})_{\ol{B}} \;\;\mbox{ where } X \mbox{ is } \set \mbox{ when } \\ 
\hspace*{0.79in}(\mu \phi.\lambda {\overline \alpha}.H){\overline
  \beta} \rrbracket^{\mathsf{D}} \rho & & \hspace*{0.2in}  
\mathsf{D} = \set \mbox{ and not present when }
\mathsf{D} = \rel\\  
\llbracket \Gamma;\emptyset \,|\, \emptyset \vdash
  \fold^F_H : \Nat^\emptyset\;(\Nat^{\ol{\beta}, \ol{\gamma}}\,H[\phi
    :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F) & = &  
\lambda d\,\eta\,\ol{B}\,\ol{C}.\,
(\mathit{fold}_{T^X_{H,\rho[\ol{\gamma := C}]}} \, (\lambda
\ol{A}.\,\eta_{\ol{A}\,\ol{C}}))_{\ol{B}}\\ 
\hspace*{0.79in}(\Nat^{{\ol{\beta},\ol{\gamma}} }\,(\mu
  \phi.\lambda \overline \alpha.H)\overline \beta\;F)
\rrbracket^{\mathsf{D}} \rho & & \hspace*{0.2in} \mbox{ where } X \mbox{ is as above
}\vspace*{-0.1in} 
\end{array}\]
\caption{Term semantics}\label{fig:term-sem} 
\end{minipage}}\vspace*{-0.05in}
\end{figure*}

\begin{comment}
\begin{dfn}\label{def:rel-interp}
If $\rho$ is a relation environment and $\Gamma;\Phi~|~\Delta \vdash t :
\tau$ then $\relsem{\Gamma;\Phi~|~\Delta \vdash t : \tau}\rho$ is
defined as follows:
\[\begin{array}{lll}
\relsem{\Gamma;\emptyset \,|\, \Delta,x :\tau \vdash x : \tau} \rho& = &
\pi_{|\Delta|+1}\\
%\relsem{\Gamma;\emptyset \,|\, \Delta \vdash \lambda x.t : \sigma \to \tau}\rho &
%= & \curry (\relsem{\Gamma;\emptyset \,|\, \Delta, x : \sigma \vdash t :
%  \tau}\rho)\\ 
%\relsem{\Gamma;\emptyset \,|\, \Delta \vdash st: \tau} \rho & = & \eval
%\circ \langle \relsem{\Gamma;\emptyset \,|\, \Delta \vdash s: \sigma \to
%  \tau}\rho, \relsem{\Gamma;\emptyset \,|\, \Delta \vdash t: \sigma}\rho
%\rangle\\
\relsem{\Gamma;\emptyset \,|\, \Delta \vdash L_{\overline \alpha} x.t : \Nat^{\overline
    \alpha} \,F \,G}\rho & = &  \curry (\relsem{\Gamma;\overline \alpha
  \,|\, \Delta, x : F \vdash t: G}\rho[\overline{\alpha := \_}])\\
\relsem{\Gamma;\Phi \,|\, \Delta \vdash t_{\overline{\tau}} s:
  G [\overline{\alpha := \tau}]}\rho & = & \eval \circ \langle
  \lambda e. \,(\relsem{\Gamma;\emptyset \,|\, \Delta \vdash t :
  \Nat^{\overline{\alpha}} \,F \,G}\rho\; e)_{\overline{\relsem{\Gamma;\Phi
      \vdash \tau}\rho}},\\ 
 & & \hspace*{0.5in} \relsem{\Gamma;\Phi \,|\,
    \Delta \vdash s: F [\overline{\alpha := \tau}]}\rho \rangle\\ 
& & \\
%% \color{red} \mbox{Add rules for } \forall \mbox{ if we include it} & & \\
\relsem{\Gamma;\Phi \,|\, \Delta,x :\tau \vdash x : \tau} \rho& = &
\pi_{|\Delta|+1}\\
\relsem{\Gamma;\Phi \,|\, \Delta \vdash \bot_\tau t : \tau} \rho& = &
!^0_{\relsem{\Gamma;\Phi \vdash \tau}\rho} \circ
  \relsem{\Gamma;\Phi~|~\Delta \vdash t : \zerot}\rho, \mbox{ where } \\
 & & \hspace*{0.1in} !^0_{\relsem{\Gamma;\Phi \vdash \tau}\rho}
\mbox{ is the unique morphism from } 0\\
 & & \hspace*{0.1in} \mbox{ to } \relsem{\Gamma;\Phi \vdash \tau}\rho\\
\relsem{\Gamma;\Phi \,|\, \Delta \vdash \top : \onet}\rho & = &
!^{\relsem{\Gamma;\Phi\vdash \Delta}\rho}_1, \mbox{ where }
!^{\relsem{\Gamma;\Phi\vdash \Delta}\rho}_1\\ 
& & \hspace*{0.1in} \mbox{ is the unique morphism from }
\relsem{\Gamma;\Phi\vdash \Delta}\rho \mbox{ to } 1\\ 
\relsem{\Gamma;\Phi \,|\, \Delta \vdash (s,t) : \sigma \times \tau} \rho& = &
\relsem{\Gamma;\Phi \,|\, \Delta \vdash s: \sigma} \rho\times
\relsem{\Gamma;\Phi \,|\, \Delta \vdash t: \tau} \rho\\
\relsem{\Gamma;\Phi \,|\, \Delta \vdash \pi_1 t : \sigma} \rho& = &
\pi_1 \circ \relsem{\Gamma;\Phi \,|\, \Delta \vdash t : \sigma \times \tau}\rho\\
\relsem{\Gamma;\Phi \,|\, \Delta \vdash \pi_2 t : \sigma}\rho & = &
\pi_2 \circ \relsem{\Gamma;\Phi \,|\, \Delta \vdash t : \sigma \times
  \tau} \rho\\
\relsem{\Gamma;\Phi~|~\Delta \vdash \case{t}{x \mapsto l}{y \mapsto r} :
  \gamma}\rho & = & \eval \circ \langle \curry \,[\relsem{\Gamma;\Phi
    \,|\, \Delta, x : \sigma \vdash l : \gamma}\rho,\\
   & & \hspace*{0.79in} \relsem{\Gamma;\Phi \,|\, \Delta, y
    : \tau \vdash r : \gamma}\rho],\\
   & &  \hspace*{0.5in} \relsem{\Gamma;\Phi \,|\, \Delta \vdash t :
  \sigma + \tau} \rho\rangle \\
\relsem{\Gamma;\Phi \,|\, \Delta \vdash \inl \,s: \sigma + \tau} \rho& = &
\inl \circ \relsem{\Gamma;\Phi \,|\, \Delta \vdash s: \sigma}\rho\\
\relsem{\Gamma;\Phi \,|\, \Delta \vdash \inr \,t: \sigma + \tau}\rho & = & 
\inr \circ \relsem{\Gamma;\Phi \,|\, \Delta \vdash t : \tau}\rho\\
\llbracket \Gamma;\emptyset \,|\, \emptyset \vdash \map^{\ol{F},\ol{G}}_H :
\Nat^\emptyset\;(\ol{\Nat^{\ol{\beta},\ol{\gamma}}\,F\,G})\;&
= & \lambda d\, \ol{\eta}\,\ol{R}.\,
\relsem{\Gamma; \ol{\phi},\ol{\gamma}\vdash H}\id_{\rho[\ol{\gamma :=
      R}]}[\ol{\phi := \lambda \ol{S}.\eta_{\ol{S}\,\ol{R}}}]\\
\hspace*{0.79in} (\Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\,H[\ol{\phi
      :=_{\ol{\beta}} G}]) \rrbracket^\rel \rho & & \\
\llbracket \Gamma;\emptyset \,|\, \emptyset \vdash \tin_H :
Nat^{\ol{\beta},\ol{\gamma}} \, H[\phi := (\mu \phi.\lambda {\overline
    \alpha}.H){\overline \beta}][\ol{\alpha := \beta}] & = &
\lambda d\,\ol{R}\, \ol{S}.\,(\mathit{in}_{T_{\rho[\ol{\gamma := S}]}})_{\ol{R}}\\
\hspace*{0.79in}(\mu \phi.\lambda {\overline \alpha}.H){\overline
  \beta} \rrbracket^\rel \rho & & \\
\llbracket \Gamma;\emptyset \,|\, \emptyset \vdash
  \fold^F_H : \Nat^\emptyset\;(\Nat^{\ol{\beta}, \ol{\gamma}}\,H[\phi
    :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F) & = &  
\lambda d\,\eta\,\ol{R}\,\ol{S}.\, (\mathit{fold}_{T_{\rho[\ol{\gamma
        := S}]}} \, (\lambda
\ol{Z}.\,\eta_{\ol{Z}\,\ol{S}}))_{\ol{R}}\\
\hspace*{0.79in}(\Nat^{{\ol{\beta},\ol{\gamma}} }\,(\mu
  \phi.\lambda \overline \alpha.H)\overline \beta\;F)
\rrbracket^\rel \rho & & 
\end{array}\]
\end{dfn}
{\color{blue} Add return type for fold in last clause? Should be
  $\relsem{\Gamma;\ol\beta,\ol\gamma \vdash F}\rho[\ol{\gamma :=
  C}]$.}

This interpretation gives that $\relsem{\Gamma;\emptyset \,|\, \Delta
  \vdash \lambda x.t : \sigma \to \tau}\rho = \curry
(\relsem{\Gamma;\emptyset \,|\, \Delta, x : \sigma \vdash t :
  \tau}\rho)$ and $\relsem{\Gamma;\emptyset \,|\, \Delta \vdash st:
  \tau} \rho = \eval \circ \langle \relsem{\Gamma;\emptyset \,|\,
  \Delta \vdash s: \sigma \to \tau}\rho, \relsem{\Gamma;\emptyset
  \,|\, \Delta \vdash t: \sigma}\rho \rangle$, as expected.

If $t$ is closed, i.e., if $\emptyset; \emptyset~|~\emptyset \vdash t
: \tau$, then we write $\setsem{\vdash t : \tau}$ instead of
$\setsem{\emptyset; \emptyset~|~\emptyset \vdash t : \tau}$.
\end{comment}

\begin{comment}
  , and for any $\rho \in \relenv$, 
\[ \relsem{\Gamma;\Phi \,|\, \Delta, x : \sigma \vdash t : \tau}\rho
  = (\relsem{\Gamma;\Phi \,|\, \Delta \vdash t : \tau}\rho) \circ
  \pi_{\Delta}\] where $\pi_{\Delta}$ is the projection
  $\relsem{\Gamma;\Phi \vdash \Delta, x : \sigma} \to
  \relsem{\Gamma;\Phi \vdash \Delta}$.
\end{comment}

The return type for the semantic fold in the last clause is
  $\dsem{\Gamma;\ol\beta,\ol\gamma \vdash F}\rho[\ol{\gamma := C}]$.
  This interpretation gives $\dsem{\Gamma;\emptyset \,|\, \Delta
    \vdash \lambda x.t : F \to G}\rho = \curry
  (\dsem{\Gamma;\emptyset \,|\, \Delta, x : F \vdash t :
    G}\rho)$ and $\dsem{\Gamma;\emptyset \,|\, \Delta \vdash st:
    G} \rho = \eval \circ \langle \dsem{\Gamma;\emptyset \,|\,
    \Delta \vdash s: F \to G}\rho, \dsem{\Gamma;\emptyset
    \,|\, \Delta \vdash t: F}\rho \rangle$, so it specializes to
  the standard interpretations for System F terms.  If $t$ is closed,
  i.e., if $\emptyset; \emptyset~|~\emptyset \vdash t : F$, then we
  write $\dsem{\vdash t : F}$ instead of $\dsem{\emptyset;
    \emptyset~|~\emptyset \vdash t : F}$.
%, and similarly for $\relsem{\emptyset; \emptyset~|~\emptyset \vdash
%t : \tau}$.
Moreover, if
  $\Gamma,\alpha;\Phi \,|\, \Delta \vdash t : F$ and
  $\Gamma;\Phi,\alpha \,|\, \Delta \vdash t' : F$ and $\Gamma;\Phi
  \vdash G$ then
\begin{itemize}
\item $\setsem{\Gamma;\Phi \,|\, \Delta[\alpha := G] \vdash
  t[\alpha := G] : \tau[\alpha := G]}\rho =
  \setsem{\Gamma,\alpha;\Phi \,|\, \Delta \vdash t : F }\rho [
    \alpha := \setsem{\Gamma;\Phi\vdash G}\rho ]$
%\item $\relsem{\Gamma;\Phi \,|\, \Delta[\alpha := \sigma] \vdash
%  t[\alpha := \sigma] : \tau[\alpha := \sigma]}\rho =
%  \relsem{\Gamma,\alpha;\Phi \,|\, \Delta \vdash t : \tau }\rho [
%    \alpha := \relsem{\Gamma;\Phi\vdash\sigma}\rho ]$
\item $\setsem{\Gamma;\Phi \,|\, \Delta[\alpha := G] \vdash
  t'[\alpha := G] : \tau[\alpha := G]}\rho =
  \setsem{\Gamma;\Phi,\alpha \,|\, \Delta \vdash t' : F }\rho [
    \alpha := \setsem{\Gamma;\Phi\vdash G}\rho ]$
%\item $\relsem{\Gamma;\Phi \,|\, \Delta[\alpha := \sigma] \vdash
%  t'[\alpha := \sigma] : \tau[\alpha := \sigma]}\rho =
%  \relsem{\Gamma;\Phi,\alpha \,|\, \Delta \vdash t' : \tau }\rho [
%    \alpha := \relsem{\Gamma;\Phi\vdash\sigma}\rho ]$
\end{itemize}
and if $\Gamma;\Phi \,|\, \Delta, x: G \vdash t : F$ and
$\Gamma;\Phi \,|\, \Delta \vdash s : G$ then
\begin{itemize}
\item $\lambda A.\, \setsem{\Gamma;\Phi \,|\, \Delta \vdash t[x := s]
  : F }\rho \, A = \lambda A.\,\setsem{\Gamma;\Phi \,|\, \Delta, x:
  G \vdash t : F}\rho \,(A, \setsem{\Gamma;\Phi \,|\,
  \Delta\vdash s: G}\rho\, A)$
%\item $\lambda R.\,\relsem{\Gamma;\Phi \,|\, \Delta \vdash t[x := s] :
%  \tau }\rho\,R = \lambda R.\,\relsem{\Gamma;\Phi \,|\, \Delta, x:
%  \sigma \vdash t : \tau}\rho \,(R, \relsem{\Gamma;\Phi \,|\,
%  \Delta\vdash s: \sigma}\rho\,R)$
\end{itemize}
Direct calculation reveals that the set interpretations of terms also
satisfy
\begin{itemize}
\item $\setsem{\Gamma; \Phi~|~\Delta \vdash
  (L_{\ol{\alpha}}x.t)_{\ol{K}}s} = \setsem{\Gamma; \Phi~|~\Delta
  \vdash t [\ol{\alpha := K}][x := s]}$
\end{itemize}
Term extensionality with respect to types and terms --- i.e.,
$\setsem{\Gamma;\Phi\vdash (L_\alpha x.t)_\alpha \top : F} =
\setsem{\Gamma;\Phi \vdash t : F}$ and $\setsem{\Gamma;\Phi\vdash
  (L_\alpha x.t)_\alpha x : F} = \setsem{\Gamma;\Phi \vdash t : F}$
%$\relsem{\Gamma;\Phi\vdash (L_\alpha x.t)_\alpha t} =
%\relsem{\Gamma;\Phi \vdash t}$, as well as
%term extensionality
%and $\relsem{\Gamma;\Phi\vdash
%  (L_\alpha x.t)_\alpha \top} = \relsem{\Gamma;\Phi \vdash t}$
--- follow (when both sides of the equations are defined). Similar
properties hold for relational interpretations.

\subsection{Properties of Terms of $\Nat$-Type}\label{sec:Nat-type-terms}

Define, for $\Gamma; \ol{\alpha} \vdash F$, the term {\color{blue}
  $\id_F$} to be $\Gamma;\emptyset~|~\emptyset \vdash
L_{\ol{\alpha}}x.x : \Nat^{\ol{\alpha}} F\,F$ and, for terms $\Gamma;
\emptyset \,|\, \Delta \vdash t: \Nat^{\overline{\alpha}} F\,G$ and
$\Gamma; \emptyset \,|\, \Delta \vdash s: \Nat^{\overline{\alpha}}
G\,H$, the {\em composition} $s \circ t$ of $t$ and $s$ to be $\Gamma;
\emptyset \,|\, \Delta \vdash L_{\overline{\alpha}}
x. s_{\overline{\alpha}}(t_{\overline{\alpha}}x):
\Nat^{\overline{\alpha}} F\,H$. Then $\setsem{\Gamma; \emptyset \,|\,
  \emptyset \vdash \id_{F} : \Nat^{\ol{\alpha}} F\,F} \rho\, \ast =
\id_{\lambda \ol{A}. \setsem{\Gamma; \ol{\alpha} \vdash F} \rho
  [\ol{\alpha := A}]}$ for any set environment $\rho$, and
$\setsem{\Gamma; \emptyset \,|\, \Delta \vdash s \circ t:
  \Nat^{\overline{\alpha}} F\,H} = \setsem{\Gamma; \emptyset \,|\,
  \Delta \vdash s: \Nat^{\overline{\alpha}} G\,H} \circ
\setsem{\Gamma; \emptyset \,|\, \Delta \vdash t:
  \Nat^{\overline{\alpha}} F\,G}$.  Straightforward calculation using
these equalities shows that terms of $\Nat$ type behave as
natural transformations with respect to their source and target
functorial types, i.e., we have
\begin{thm}\label{eq:ft-from-nat}
\[\begin{array}{l}
\hspace*{0.135in}\setsem{\Gamma; \emptyset \,|\, x :
  \Nat^{\overline{\alpha}, 
    \overline{\gamma}} F\,G, \overline{y : \Nat^{\overline{\gamma}}
    K\, H} \vdash (\map_G^{\overline{K},
    \overline{H}} \,\overline{y}) \circ
  (L_{\overline{\gamma}} z. x_{\overline{K}, \overline{\gamma}}
  z) : \Nat^{\overline{\gamma}} F[\overline{\alpha := K}]\,
  G[\overline{\alpha := H}]}\\
= \setsem{ \Gamma; \emptyset \,|\, x
  : \Nat^{\overline{\alpha}, \overline{\gamma}} F\, G, \overline{y :
    \Nat^{\overline{\gamma}} K\, H} \vdash
  (L_{\overline{\gamma}} z. x_{\overline{H}, \overline{\gamma}} z)
  \circ (\map_F^{\overline{K}, \overline{\tau}} \,
  \overline{y}) : \Nat^{\overline{\gamma}} F[\overline{\alpha :=
      K}]\, G[\overline{\alpha := H}]}
\end{array}\]
\end{thm}
When $x = \mathit{in}_H$ and $\xi = \Nat^{\overline{\gamma}} H[\phi :=
  (\mu \phi. \lambda \overline{\alpha}. H)
  \overline{\beta}][\overline{\alpha := F}]\, (\mu \phi. \lambda
\overline{\alpha}. H)\overline{G}$ Theorem~\ref{eq:ft-from-nat}
specializes to
\begin{equation}\label{thm:subst}
\begin{array}{l}
\hspace*{0.135in}\setsem{\Gamma; \emptyset \,|\, \overline{y :
    \Nat^{\overline{\gamma}} 
    F\, G} \vdash (\map_{(\mu \phi. \lambda
    \overline{\alpha}. H) \overline{\beta}}^{\overline{F},
    \overline{G}}\, \overline{y}) \circ
  (L_{\overline{\gamma}} z. (\tin_{H})_{\overline{F},
    \overline{\gamma}} z) : \xi}\\
= \setsem{\Gamma; \emptyset \,|\, \overline{y :
  \Nat^{\overline{\gamma}} F\, G} 
  \vdash (L_{\overline{\gamma}} z. (\tin_H)_{\overline{G}, \overline{\gamma}} z)
  \circ (\map_{H[\phi := (\mu \phi. \lambda \overline{\alpha}. H)
      \overline{\beta}]}^{\overline{F},
    \overline{G}}\, \overline{y}) 
  : \xi}
\end{array}
\end{equation}
\noindent
Theorem~\ref{eq:ft-from-nat} gives a family of free theorems that are
consequences of naturality, and thus do not require the full power of
parametricity. Other specific instances include the three
$\mathtt{map}$/$\mathtt{reverse}$ theorems from
Section~\ref{sec:intro}. The $\mathtt{map}$/$\mathtt{reverse}$ theorem
for bushes, e.g., is
\begin{align*}
  &\setsem{\emptyset; \alpha~|~ y : \Nat^\emptyset F \, G
    \vdash (\map_{\mathit{Bush}\,\alpha}^{{F},
      {G}}\, y) \circ (L_{{\emptyset}}
    z. \mathit{reverseBush}_{F}\, z) : \Nat^{{\emptyset}}
    (\mathit{Bush} \, F) \, (\mathit{Bush} \, G) }\\ =\;
  &\setsem{\emptyset; \alpha~|~ y : \Nat^\emptyset F\, G
    \vdash (L_{\emptyset} z. \mathit{reverseBush}_{G} \,z) \circ
    (\map_{\mathit{Bush} \, \alpha}^{F, G}\,
           {y}) : \Nat^{\emptyset} (\mathit{Bush} \, F ) \,
           (\mathit{Bush} \, G)}
\end{align*}
for any term $\vdash \mathit{reverseBush} : \Nat^\alpha (\mathit{Bush}
\,\alpha) \, (\mathit{Bush} \, \alpha)$. We can also prove a theorem
for the data type $\emptyset;\alpha,\gamma \vdash
\mathit{Tree}\,\alpha\,\gamma$ from Section~\ref{sec:calculus} that is
similar but only maps along some of its functorial variables:
%  \begin{align*}
%  &\setsem{\emptyset; \alpha~|~ y : \Nat^\emptyset \sigma \, \tau
%    \vdash ((\map_{\mathit{PTree}\,\alpha}^{{\sigma},
%      {\tau}})_{\emptyset} {y}) \circ (L_{{\emptyset}}
%    z. \mathit{reverse}_{\sigma} z) : \Nat^{{\emptyset}}
%    (\mathit{PTree} \, \sigma) \, (\mathit{PTree} \, \tau) }\\ =\;
%  &\setsem{\emptyset; \alpha~|~ y : \Nat^\emptyset \sigma \, \tau
%    \vdash (L_{\emptyset} z. \mathit{reverse}_{\tau} z) \circ
%    ((\map_{\mathit{PTree} \, \alpha}^{{\sigma}, {\tau}})_{\emptyset} {y}) :
%    \Nat^{\emptyset} (\mathit{PTree} \, \sigma ) \, (\mathit{PTree} \,
%    \tau)}
%\end{align*}
%for any term $\vdash \mathit{reverse} : \Nat^\alpha (\mathit{PTree}
%\,\alpha) \, (\mathit{PTree} \, \alpha)$,
if $\vdash \mathit{reverseTree} : \Nat^{\alpha, \gamma} (\mathit{Tree}
\,\alpha \, \gamma) \, (\mathit{Tree} \, \alpha \, \gamma)$ then
\begin{align*}
  &\setsem{\emptyset; \alpha,\gamma~|~y : \Nat^\emptyset F \,
    G \vdash (\map_{\mathit{Tree}\,\alpha \, \gamma}^{F,
      G}\, y) \circ (L_{{\gamma}}
    z. \mathit{reverseTree}_{F, \gamma}\, z) : \Nat^{{\gamma}}
    (\mathit{Tree} \, F \, \gamma) \, (\mathit{Tree} \, G \,
    \gamma) }\\ =\; &\setsem{\emptyset; \alpha,\gamma~|~y :
    \Nat^\emptyset F \, G \vdash (L_{\gamma}
    z. \mathit{reverseTree}_{G, \gamma} \,z) \circ
    (\map_{\mathit{Tree} \, \alpha\, \gamma}^{{F},
      {G}}\, y) : \Nat^{\gamma} (\mathit{Tree} \,
    F \, \gamma) \, (\mathit{Tree} \, G\, \gamma)}
\end{align*}
\noindent
%Note this instance uses the following version of map:
%$$\map_{\mathit{Tree} \,\alpha \, \gamma}^{\sigma, \tau} : 
%\Nat^{\emptyset} (\Nat^\gamma \, \sigma \,\tau)\,
%(\Nat^\gamma(\mathit{Tree}\,\sigma\,\gamma)\,(\mathit{Tree}\,\tau\,\gamma))$$
%as opposed to 
%$$\map_{\mathit{Tree} \,\alpha \, \gamma}^{\sigma, \sigma', \tau, \tau'} : 
%\Nat^{\emptyset} ((\Nat^\emptyset \, \sigma \,\tau) \times
%(\Nat^\emptyset\,\sigma'\,\tau')) \,
%(\Nat^\emptyset\,(\mathit{Tree}\,\sigma\,\sigma')\,(\mathit{Tree}\,\tau\,\tau'))$$
Finally, in Section~\ref{sec:calculus} we observed that $\emptyset;
\phi, \alpha \vdash \mathit{GRose}\,\phi\,\alpha$ cannot appear as
the (co)domain of a $\Nat$ type binding both $\phi$ and $\alpha$. As a
result, there are no instances of, say, a
$\mathtt{map}$/$\mathtt{reverse}$ free theorem for this data type that
map replacements for both $\phi$ and $\alpha$. On the other hand,
$\phi; \alpha \vdash \mathit{GRose}\,\phi\,\alpha$ does support a
$\mathtt{map}$/$\mathtt{reverse}$ theorem similar to those from
Section~\ref{sec:intro}.

As usual, analogous results hold for relation environments and
relational interpretations.

\subsection{Properties of Initial Algebraic Constructs}\label{sec:iaps}

Our semantics interprets applications of $\map$ as applications of
semantic $\mathit{map}$ functions.
\begin{comment}
Indeed, if \(\Gamma; \ol{\phi},
\ol{\gamma} \vdash H\), \(\ol{\Gamma; \ol{\beta}, \ol{\gamma} \vdash
  F}\), and \(\ol{\Gamma; \Phi, \ol{\beta}, \ol{\gamma} \vdash G}\),
%and $\Gamma;\Phi\,|\,\Delta \vdash t :
%\Nat^{\ol{\beta},\ol{\gamma}}\,F\,G$
then 
Definition~\ref{def:set-interp} gives
%\begin{multline}\label{eq:map-sem-def}
\begin{align}
& \setsem{
\Gamma;\Phi \,|\, \Delta
\vdash \map^{\ol{F},\ol{G}}_H \,\ol t
: %\Nat^\emptyset\;(\ol{\Nat^{\ol{\beta},\ol{\gamma}}\,F\,G})\; {}
\Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\,H[\ol{\phi :=_{\ol{\beta}} G}]
} \rho \nonumber \\ 
=\;\; & \lambda d\, \ol{\eta}\,\ol{B}.\,
\setsem{\Gamma; \ol{\phi},\ol{\gamma}\vdash H}
\id_{\rho[\ol{\gamma := B}]}[\ol{\phi := \lambda
    \ol{A}.\eta_{\ol{A}\,\ol{B}}}] \label{eq:map-sem-def}
\end{align}
%\end{multline}
\end{comment}
In particular, if $\Gamma; \ol{\alpha} \vdash H$, $\ol{\Gamma;
  \emptyset \vdash F}$, and $\ol{\Gamma; \emptyset \vdash G}$, then
Definition~\ref{def:set-interp} gives, for all $\ol{f :
  \setsem{\Gamma;\emptyset \vdash \Nat^\emptyset F\,G}\rho}$, that
\[\begin{array}{rl}
&\setsem{ \Gamma;\emptyset\,|\,\ol{x : \Nat^\emptyset F \, G}
  \vdash \map^{\ol{F},\ol{G}}_H\,\ol x :
  \Nat^{\emptyset}\,H[\ol{\alpha := F}]\,H[\ol{\alpha := G}]}
\rho\,\ol f\\ =& \setsem{\Gamma; \ol{\alpha}\vdash H}
\id_{\rho}[\ol{\alpha := f}] \\ =&
%\lambda \ol{f : \setsem{\Gamma; \emptyset \vdash \sigma}\rho \to \setsem{\Gamma%; \emptyset \vdash \tau}\rho}.\,
\mathit{map}_{\lambda \ol{A}.\,\setsem{\Gamma; \ol{\alpha}\vdash H} \rho [\ol{\alpha := A}]} \ol{f} \\
%=& \mathit{map}_{\lambda \ol{A}.\,\setsem{\Gamma; \ol{\alpha}\vdash H} \rho [\ol{\alpha := A}]}
\end{array}
\]
\begin{comment}
\[\begin{array}{rl}
&\setsem{
\Gamma;\emptyset \,|\, \emptyset
\vdash \map^{\ol{\sigma},\ol{\tau}}_H
: \Nat^\emptyset\;(\ol{\Nat^{\emptyset}\,\sigma\,\tau})\; {}
(\Nat^{\emptyset}\,H[\ol{\alpha := \sigma}]\,H[\ol{\alpha := \tau}])
} \rho\, \ast \\
=& \lambda \ol{f : \setsem{\Gamma; \emptyset \vdash \sigma}\rho \to \setsem{\Gamma; \emptyset \vdash \tau}\rho}.\,
\setsem{\Gamma; \ol{\alpha}\vdash H}
\id_{\rho}[\ol{\alpha := f}] \\
=& \lambda \ol{f : \setsem{\Gamma; \emptyset \vdash \sigma}\rho \to \setsem{\Gamma; \emptyset \vdash \tau}\rho}.\,
\mathit{map}_{\lambda \ol{A}.\,\setsem{\Gamma; \ol{\alpha}\vdash H} \rho [\ol{\alpha := A}]} \ol{f} \\
=& \mathit{map}_{\lambda \ol{A}.\,\setsem{\Gamma; \ol{\alpha}\vdash H} \rho [\ol{\alpha := A}]}
\end{array}
\]
\end{comment}
Here, we obtain the first equality from the appropriate instance of
Definition~\ref{def:set-interp}, and the second one by noting that
$\lambda \ol{A}.\,\setsem{\Gamma; \ol{\alpha}\vdash H} \rho
[\ol{\alpha := A}]$ is a functor in $\alpha$ and using
$\mathit{map}_G\,\ol f$ to denote the action of the semantic functor
$G$ on morphisms $\ol f$.

\vspace*{0.1in}

We also have the expected relationships between the interpetations of
$\map$, $\tin$, and $\fold$, namely

\vspace*{0.1in}

\noindent
$\bullet$\;  If
$\Gamma; \ol{\psi}, \ol{\gamma} \vdash H$,\;
$\ol{\Gamma; \ol{\alpha}, \ol{\gamma}, \ol{\phi} \vdash K}$,\;
$\ol{\Gamma; \ol{\beta}, \ol{\gamma} \vdash F}$, 
$\ol{\Gamma; \Phi, \ol{\beta}, \ol{\gamma} \vdash G}$, and
$\ol{\Gamma;\Phi\,|\,\Delta \vdash t : \Nat^{\ol{\alpha}, \ol{\beta},
    \ol{\gamma}} F\, G}$
then
\[\setsem{\Gamma; \Phi \,|\, \Delta \vdash
\map_{H[\ol{\psi := K}]}^{\ol{F}, \ol{G}}\,\ol t : \xi} = \setsem{\Gamma;
  \Phi \,|\, \Delta \vdash \map_H^{\ol{K[\ol{\phi := F}]},
    \ol{K[\ol{\phi := G}]}}\, (\ol{\map_K^{\ol{F}, \ol{G}}}\, \ol t) :
  \xi}\]
for $\xi = \Nat^{\ol{\gamma}} H[\ol{\psi :=
    K}][\ol{\phi := F}] H[\ol{\psi := K}][\ol{\phi := G}]$

\vspace*{0.18in}

\noindent
$\bullet$\; If $\Gamma; \overline{\beta}, \overline{\gamma} \vdash
H$,\; $\Gamma; \overline{\beta}, \overline{\gamma} \vdash K$,\;
$\overline{\Gamma; \overline{\alpha}, \overline{\gamma} \vdash F}$,\;
$\overline{\Gamma; \overline{\alpha}, \overline{\gamma} \vdash G}$,\;
$\overline{\Gamma; \phi, \overline{\psi}, \overline{\gamma} \vdash
  M}$,\; $\overline{I}$ is the sequence $\overline{F}, H$,\;
$\overline{J}$ is the sequence $\overline{G}, K$, and $\Gamma;
\Phi\,|\,\Delta \vdash s : \Nat^{\overline{\beta},
  \overline{\gamma}} H\, K$ and $\ol{\Gamma; \Phi\,|\,\Delta \vdash
  t : \Nat^{\overline{\alpha}, \overline{\gamma}} F\, G}$, then
\[\hspace*{0.13in}\setsem{\Gamma; \Phi \,|\, \Delta
    \vdash L_{\overline{\gamma}} z.\,
    s_{\overline{M [\overline{\psi := G}] [\phi := K]},
      \overline{\gamma}} \Big(\big( \map_{H}^{\overline{M
        [\overline{\psi := F}] [\phi := H]}, \overline{M
        [\overline{\psi := G}] [\phi := K]}}
    (\overline{\map_{M}^{\overline{I}, \overline{J}}
      (s, \overline{t})}) \big)_{\overline{\gamma}} z \Big) : \xi}\]
\[= \setsem{\Gamma; \Phi \,|\, \Delta \vdash
    L_{\overline{\gamma}} z.\, \big( \map_{K}^{\overline{M
      [\overline{\psi := F}] [\phi := H]}, \overline{M [\overline{\psi
          := G}] [\phi := K]}} (\overline{\map_{M}^{\overline{I},
    \overline{J}} (s, \overline{t})}) \big)_{\overline{\gamma}} \Big(
s_{\overline{M [\overline{\psi := F}] [\phi := H]}, \overline{\gamma}}
z \Big) : \xi}\] for $\xi = \Nat^{\overline{\gamma}}\,
  H[\overline{\beta:= M}] [\overline{\psi := F}] [\phi := H]\;
  K[\overline{\beta := M}] [\overline{\psi := G}] [\phi := K]$

\vspace*{0.18in}

\noindent
$\bullet$\; If $\xi = \Nat^{\overline{\beta}, \overline{\gamma}}\,
H[\phi := (\mu \phi. \lambda
  \overline{\alpha}. H)\overline{\beta}][\overline{\alpha := \beta}]\;
F$, then
\[\hspace*{-0.72in}\setsem{\Gamma; \emptyset \,|\, x:
  \Nat^{\overline{\beta}, \overline{\gamma}} H[\phi :=
    F][\overline{\alpha := \beta}]\, F \vdash ((\fold_{H,
    F})_{\emptyset} x) \circ \tin_{H} : \xi}\]

\vspace*{-0.3in}

\[= \setsem{ \Gamma; \emptyset \,|\, x: \Nat^{\overline{\beta},
    \overline{\gamma}} H[\phi := F][\overline{\alpha := \beta}]\, F
  \vdash x \circ \big( \map_{H [\ol{\alpha := \beta}]}^{(\mu
    \phi. \lambda \overline{\alpha} H) \overline{\beta},
    F} ((\fold_{H, F})_{\emptyset} x) \big) : \xi}\]

\vspace*{0.18in}

\noindent
$\bullet$\; If $\xi = \Nat^{\overline{\beta}, \overline{\gamma}}\, (\mu
\phi. \lambda \overline{\alpha}. H)\overline{\beta}\; (\mu
\phi. \lambda \overline{\alpha}. H)\overline{\beta}$, then
\[\setsem{\Gamma; \emptyset \,|\, \emptyset
  \vdash \tin_H \circ (\fold_{H, H[\phi := (\mu \phi. \lambda
      \overline{\alpha}. H)\overline{\beta}]})_{\emptyset}
  (\map_{H}^{H[\phi := (\mu \phi. \lambda
      \overline{\alpha}. H)\overline{\beta}][\overline{\alpha :=
        \beta}], (\mu \phi. \lambda
    \overline{\alpha}. H)\overline{\beta}} \tin_H) :
  \xi}\]

\vspace*{-0.2in}

\[\hspace*{-3.6in} = \setsem{\Gamma; \emptyset \,|\, \emptyset \vdash \Id_{(\mu
    \phi. \lambda \overline{\alpha}. H)\overline{\beta}} : \xi}\]

\vspace*{0.18in}

\noindent
$\bullet$\; If $\xi = \Nat^{\overline{\beta}, \overline{\gamma}}
H[\phi := (\mu \phi. \lambda \overline{\alpha}. H)\overline{\beta}]\,
H[\phi := (\mu \phi. \lambda \overline{\alpha}. H)\overline{\beta}]$,
then
\[\setsem{\Gamma; \emptyset \,|\, \emptyset \vdash
(\fold_{H, H[\phi := (\mu \phi. \lambda
      \overline{\alpha}. H)\overline{\beta}]})_{\emptyset}
  (\map_{H}^{H[\phi := (\mu \phi. \lambda
      \overline{\alpha}. H)\overline{\beta}][\overline{\alpha :=
        \beta}], (\mu \phi. \lambda
    \overline{\alpha}. H)\overline{\beta}} \tin_H) \circ
  \tin_H : \xi}\]

\vspace*{-0.2in}

\[\hspace*{-3.32in} =\setsem{ \Gamma; \emptyset \,|\, \emptyset \vdash
  \Id_{H[\phi := (\mu \phi. \lambda
      \overline{\alpha}. H)\overline{\beta}]} : \xi}\]

\vspace*{0.1in}

\noindent
Analogous results hold for relational interpretations of terms and
relational environments.

\begin{comment}
\subsection{Free Theorems Derived from Naturality}\label{sec:ft-nat}

In our setting, free theorems of a certain familiar form are derivable
directly from the semantic properties of $\Nat$-types. Indeed, the
following theorem is immediate from Equation~\ref{thm:subst}, by taking
$\gamma$ to be empty, expanding the interpretations on the left- and
right-hand sides, and using the fact from Section~\ref{sec:iaps} that
$\map$ terms are interpreted as semantic $\mathit{map}$ functions:
\begin{thm}\label{thm:ft-nat}
If\, $\Gamma; \ol\alpha \vdash F : \F$,\, $\Gamma; \ol\alpha \vdash G :
\F$, $\rho : \setenv$, $x \in \setsem{\Gamma; \emptyset \vdash
  \Nat^{\ol\alpha} F\, G } \rho$, and, for vectors of length
$|\ol\alpha|$, \, $\ol{\Gamma; \emptyset \vdash \sigma : \F}$,\,
$\ol{\Gamma; \emptyset \vdash \tau: \F}$, and $\ol{y \in
  \setsem{\Gamma; \emptyset \vdash \Nat^{\emptyset} \sigma\, \tau }
  \rho}$, then
\[\mathit{map}_{\setsem{\Gamma; \ol\alpha \vdash G} \rho[\ol{\alpha := \_}]} \ol{y}
  \circ x_{\ol{{\setsem{\Gamma; \emptyset \vdash \sigma} \rho}}} =
  x_{\ol{\setsem{\Gamma; \emptyset \vdash \tau} \rho} } \circ
  \mathit{map}_{\setsem{\Gamma; \ol\alpha \vdash F} \rho[\ol{\alpha :=
        \_}]} \, \ol{y}\]
\end{thm}
\noindent
%INCORRECT
%In particular, for $\ol{A, B: \set}$, variables $\ol{v}$ and $\ol{w}$
%with $\ol{\rho v = A}$ and $\ol{\rho w = B}$, and functions $\ol{f : A
%  \to B}$, the following holds: $\mathit{map}_{\setsem{\Gamma; \ol\alpha \vdash G}
%  \rho[\ol{\alpha := \_}]} \ol{f} \circ x_{\ol{A}} = x_{\ol{B}} \circ
%\mathit{map}_{\setsem{\Gamma; \ol\alpha \vdash F} \rho[\ol{\alpha :=
%      \_}]} \, \ol{f}$.
A similar result holds for relational interpretations.

Specific instances of Theorem~\ref{thm:ft-nat} include {\color{blue}
  Wadler free theorems, map-reverse theorems from intro, ???}
\end{comment}

\begin{comment}
We already know from Equation~\ref{thm:subst} that 
\begin{multline}
\setsem{\Gamma; \emptyset \,|\, x :
  \Nat^{\overline{\alpha}, 
    \overline{\gamma}} F\,G, \overline{y : \Nat^{\overline{\gamma}}
    \sigma\, \tau} \vdash ((\map_G^{\overline{\sigma},
    \overline{\tau}})_{\emptyset} \overline{y}) \circ
  (L_{\overline{\gamma}} z. x_{\overline{\sigma}, \overline{\gamma}}
  z) : \Nat^{\overline{\gamma}} F[\overline{\alpha := \sigma}]\,
  G[\overline{\alpha := \tau}]} \\
= \setsem{ \Gamma; \emptyset \,|\, x
  : \Nat^{\overline{\alpha}, \overline{\gamma}} F\, G, \overline{y :
    \Nat^{\overline{\gamma}} \sigma\, \tau} \vdash
  (L_{\overline{\gamma}} z. x_{\overline{\tau}, \overline{\gamma}} z)
  \circ ((\map_F^{\overline{\sigma}, \overline{\tau}})_{\emptyset}
  \overline{y}) : \Nat^{\overline{\gamma}} F[\overline{\alpha :=
      \sigma}]\, G[\overline{\alpha := \tau}]}
\end{multline}

In particular, for any set environment $\rho$, natural transformation 
$x \in \setsem{\Gamma; \emptyset \vdash \Nat^{\ol\alpha} F\, G } \rho$, and functions
$\ol{y \in \setsem{\Gamma; \emptyset \vdash \Nat^{\emptyset} \sigma\, \tau } \rho}$
(thus specializing to the case where there are no $\gamma$’s), we have
\begin{multline}
\setsem{\Gamma; \emptyset \,|\, x :
  \Nat^{\overline{\alpha}} F\,G, \overline{y : \Nat^{\emptyset}
    \sigma\, \tau} \vdash ((\map_G^{\overline{\sigma},
    \overline{\tau}})_{\emptyset} \overline{y}) \circ
  (L_{\emptyset} z. x_{\overline{\sigma}}
  z) : \Nat^{\emptyset} F[\overline{\alpha := \sigma}]\,
  G[\overline{\alpha := \tau}]} \rho (x, \ol{y}) \\
= \setsem{ \Gamma; \emptyset \,|\, x
  : \Nat^{\overline{\alpha}} F\, G, \overline{y :
    \Nat^{\emptyset} \sigma\, \tau} \vdash
  (L_{\emptyset} z. x_{\overline{\tau}} z)
  \circ ((\map_F^{\overline{\sigma}, \overline{\tau}})_{\emptyset}
  \overline{y}) : \Nat^{\emptyset} F[\overline{\alpha :=
      \sigma}]\, G[\overline{\alpha := \tau}]} \rho (x, \ol{y})
\end{multline}

Unwinding these interpretations yields
\begin{multline}
\setsem{\Gamma; \emptyset \,|\, x :
  \Nat^{\overline{\alpha}} F\,G, \overline{y : \Nat^{\emptyset}
    \sigma\, \tau} \vdash ((\map_G^{\overline{\sigma},
    \overline{\tau}})_{\emptyset} \overline{y}) \circ
  (L_\emptyset z. x_{\overline{\sigma}}
  z) : \Nat^{\emptyset} F[\overline{\alpha := \sigma}]\,
  G[\overline{\alpha := \tau}]} \rho \, (x , \ol{y}) \\
= 
\setsem{\Gamma; \emptyset \,|\, x : \Nat^{\overline{\alpha}} F\,G, 
  \overline{y : \Nat^{\emptyset}
    \sigma\, \tau} \vdash (\map_G^{\overline{\sigma},
    \overline{\tau}})_{\emptyset} \overline{y} } \rho \, (x, \ol{y})
   \circ
  \setsem{\Gamma; \emptyset \,|\, x : \Nat^{\overline{\alpha}} F\,G 
   , \overline{y : \Nat^{\emptyset}
      \sigma\, \tau} \vdash
   L_{\emptyset} z. x_{\overline{\sigma}}
  z } \rho \, (x, \ol{y}) \\
= 
\setsem{\Gamma; \emptyset \,|\, \overline{y : \Nat^{\emptyset}
    \sigma\, \tau} \vdash (\map_G^{\overline{\sigma},
    \overline{\tau}})_{\emptyset} \overline{y} } \rho \, \ol{y}
   \circ
  \setsem{\Gamma; \emptyset \,|\, x : \Nat^{\overline{\alpha}} F\,G \vdash
    L_{\emptyset} z. x_{\overline{\sigma}}
  z } \rho \, x \\
= \map_{\setsem{\Gamma; \ol\alpha \vdash G} \rho[\ol{\alpha := \_}]} \ol{y}
  \circ 
  x_{\ol{{\setsem{\Gamma; \emptyset \vdash \sigma} \rho}}}
\end{multline}

and 
\begin{multline}
\setsem{ \Gamma; \emptyset \,|\, x
  : \Nat^{\overline{\alpha}} F\, G, \overline{y :
    \Nat^{\emptyset} \sigma\, \tau} \vdash
  (L_{\emptyset} z. x_{\overline{\tau}} z)
  \circ ((\map_F^{\overline{\sigma}, \overline{\tau}})_{\emptyset}
  \overline{y}) : \Nat^{\emptyset} F[\overline{\alpha :=
      \sigma}]\, G[\overline{\alpha := \tau}]} \rho (x, \ol{y}) \\
=
\setsem{ \Gamma; \emptyset \,|\, x
  : \Nat^{\overline{\alpha}} F\, G, \overline{y :
    \Nat^{\emptyset} \sigma\, \tau} \vdash
  L_{\emptyset} z. x_{\overline{\tau}} z)} \rho (x, \ol{y})
  \circ 
  \setsem{\Gamma; \emptyset
  \,|\, x : \Nat^{\overline{\alpha}} F\, G, \overline{y :
    \Nat^{\emptyset} \sigma\, \tau} \vdash 
  (\map_F^{\overline{\sigma}, \overline{\tau}})_{\emptyset}
  \overline{y} } \rho (x, \ol{y}) \\
= \setsem{\Gamma; \emptyset \,|\, x
  : \Nat^{\overline{\alpha}} F\, G \vdash 
  L_{\emptyset} z. x_{\overline{\tau}} z} \rho \, x
  \circ 
  \setsem{\Gamma;\emptyset \,|\, \overline{y :
    \Nat^{\emptyset} \sigma\, \tau} \vdash
    (\map_F^{\overline{\sigma}, \overline{\tau}})_{\emptyset}
  \overline{y} } \rho \, \ol{y}  \\
  = x_{\ol{\setsem{\Gamma; \emptyset \vdash \tau} \rho} }
  \circ \map_{\setsem{\Gamma; \ol\alpha \vdash F} 
    \rho[\ol{\alpha := \_}]} \, \ol{y}
\end{multline}

So, we can conclude that
\begin{equation}
\map_{\setsem{\Gamma; \ol\alpha \vdash G} \rho[\ol{\alpha := \_}]} \ol{y}
  \circ 
  x_{\ol{{\setsem{\Gamma; \emptyset \vdash \sigma} \rho}}}
= x_{\ol{\setsem{\Gamma; \emptyset \vdash \tau} \rho} }
\circ \map_{\setsem{\Gamma; \ol\alpha \vdash F} 
  \rho[\ol{\alpha := \_}]} \, \ol{y}
\end{equation}

Moreover, for any $\ol{A, B: \set}$,
we can choose $\ol{\sigma = v}$ and $\ol{\tau = w}$ to be variables such that
$\ol{\rho v = A}$ and $\ol{\rho w = B}$.
Then for any functions $\ol{f : A \to B}$ we have that
\begin{equation}
\map_{\setsem{\Gamma; \ol\alpha \vdash G} \rho[\ol{\alpha := \_}]} \ol{f}
  \circ 
  x_{\ol{A}}
  = x_{\ol{B}}
\circ \map_{\setsem{\Gamma; \ol\alpha \vdash F} 
  \rho[\ol{\alpha := \_}]} \, \ol{f}
\end{equation}
\end{proof}
\end{comment}

\subsection{The Abstraction Theorem}\label{sec:thms}

To go beyond naturality and get {\em all} consequences of
parametricity, we prove an Abstraction Theorem for our calculus. As
usual for such theorems, we prove a more general result in
Theorem~\ref{thm:at-gen} that handles possibly open terms. We then
recover the Abstraction Theorem (Theorem~\ref{thm:abstraction}) as the
special case Theorem~\ref{thm:at-gen} for closed terms of closed type.

\begin{thm}\label{thm:at-gen}
Every well-formed term $\Gamma;\Phi~|~\Delta \vdash t : F$ induces
a natural transformation from $\sem{\Gamma; \Phi \vdash \Delta}$ to
$\sem{\Gamma; \Phi \vdash F}$, i.e., a triple of natural
transformations 
\[(\setsem{\Gamma;\Phi~|~\Delta \vdash t : F},\,
\setsem{\Gamma;\Phi~|~\Delta \vdash t : F},\,
\relsem{\Gamma;\Phi~|~\Delta \vdash t : F})\]
where
\[\begin{array}{lll}
\setsem{\Gamma;\Phi~|~\Delta \vdash t : F} & : & \setsem{\Gamma;
  \Phi \vdash \Delta} \to \setsem{\Gamma; \Phi \vdash F}
\end{array}\]
has as its component at $\rho : \setenv$ a morphism
\[\begin{array}{lll}
\setsem{\Gamma;\Phi~|~\Delta \vdash t : F}\rho & : & \setsem{\Gamma;
  \Phi \vdash \Delta}\rho \to \setsem{\Gamma; \Phi \vdash F}\rho
\end{array}\]
in $\set$,
\[\begin{array}{lll}
\relsem{\Gamma;\Phi~|~\Delta \vdash t : F} & : & \relsem{\Gamma;
  \Phi \vdash \Delta} \to \relsem{\Gamma; \Phi \vdash F}
\end{array}\]
has as its component at $\rho : \relenv$ a morphism
\[\begin{array}{lll}
\relsem{\Gamma;\Phi~|~\Delta \vdash t : F}\rho & : & \relsem{\Gamma;
  \Phi \vdash \Delta}\rho \to \relsem{\Gamma; \Phi \vdash F}\rho
\end{array}\]
in $\rel$,
and, for all $\rho : \relenv$,
\begin{equation}\label{eq:projs}
\relsem{\Gamma;\Phi~|~\Delta \vdash t : F}\rho =
(\setsem{\Gamma;\Phi~|~\Delta \vdash t : F}(\pi_1 \rho),\,
\setsem{\Gamma;\Phi~|~\Delta \vdash t : F}(\pi_2 \rho))
\end{equation}
\end{thm}
\begin{proof}
By structural induction on the type judgment for $t$. The overall
challenge of our development manifests in this proof. It lies in
showing that the set and relational interpretations of each term
judgment are natural transformations, and that all set interpretations
of terms of $\Nat$-types satisfy the appropriate {\color{blue}
  equality preservation} conditions from Definition~\ref{def:set-sem}.
For the interesting cases of abstraction, application, $\map$, $\tin$,
and $\fold$ terms, propagating the naturality conditions is quite
involved; the latter two cases especially require some rather
involved diagram chasing. That this is possible provides strong
evidence that our definitions and development are sensible,
natural, and at an appropriate level of abstraction. Detailed proofs
for the set interpretations for the five interesting forms of terms
are included in the accompanying anonymous supplementary
material. Even in the supplementary material we omit the other,
straightforward, cases.  The proofs for the relational interpretations
of all term judgments are entirely analogous to the proofs for their
set interpretations. The proof that Equation~\ref{eq:projs} holds in
each case is by direct calculation, using the facts that projections
are surjective and that the set and relational interpretations are
defined ``in parallel'', i.e., are fibred.
\end{proof}

The following theorem is an immediate consequence of
Theorem~\ref{thm:at-gen}:
\begin{thm}\label{thm:at-gen-rel}
If \,$\Gamma; \Phi \,|\, \Delta \vdash t : F$ and \,$\rho \in \relenv$,
  and if \,$(a, b) \in \relsem{\Gamma; \Phi \vdash \Delta} \rho$,
  then \\
  $(\setsem{\Gamma; \Phi \,|\, \Delta \vdash t : F} (\pi_1 \rho) \,a \, ,
      \setsem{\Gamma; \Phi \,|\, \Delta \vdash t : F } (\pi_2 \rho) \,b) \in 
    \relsem{\Gamma; \Phi \vdash F} \rho$
\end{thm}
\noindent
Finally, the Abstraction Theorem is the instantiation of
Theorem~\ref{thm:at-gen-rel} to closed terms of closed type:
\begin{thm}[Abstraction Theorem]\label{thm:abstraction}
If $\vdash t : F$, then $(\setsem{\vdash t : F},\setsem{\vdash t
  : F}) \in \relsem{\vdash F}$.
\end{thm}

\section{Free Theorems for Nested Types}\label{sec:ftnt}

In this section we show how Theorem~\ref{thm:abstraction} can be used
to prove more advanced free theorems than those of
Section~\ref{sec:Nat-type-terms} in the presence of directly
constructed nested types.

\subsection{Free Theorem for Type of Polymorphic
  Bottom}\label{sec:bottom}

Suppose $ \vdash g : \Nat^\alpha \,\onet\,\alpha$, let $G^\set =
\setsem{\vdash g : \Nat^\alpha \,\onet\,\alpha}$, and let $G^\rel =
\relsem{\vdash g : \Nat^\alpha \,\onet\,\alpha}$.  By
Theorem~\ref{thm:abstraction}, $(G^\set(\pi_1\rho),G^\set(\pi_2\rho))
= G^\rel\rho$. Thus, for all $\rho \in \relenv$ and any $(a, b) \in
\relsem{\vdash \emptyset}\rho = 1$, eliding the only possible
instantiations of $a$ and $b$ gives that
%\[\begin{array}{lll}
$(G^\set,G^\set) \;= \; (G^\set(\pi_1 \rho), G^\set (\pi_2 \rho))$
$ \in  \relsem{\vdash \Nat^\alpha \,\onet\,\alpha}\rho$
$ =  \{\eta : K_1 \Rightarrow \id\}$
$ =  \{(\eta_1 : K_1 \Rightarrow \id, \eta_2 : K_1 \Rightarrow
\id)\}$ 
%\end{array}\]
That is, $G^\set$ is a natural transformation from the constantly
$1$-valued functor to the identity functor in $\set$. In particular,
for every $S : \set$, $G^\set_S : 1 \to S$. Note, however, that if $S
= \emptyset$, then there can be no such morphism, so no such natural
transformation can exist,
%in $\set$,
and thus no term $\vdash g : \Nat^\alpha \onet \,\alpha$ can exist.
%in our calculus.
That is, our calculus does not admit any terms with the closed type
$\Nat^\alpha \onet \,\alpha$ of the polymorphic bottom.

\subsection{Free Theorem for Type of Polymorphic
  Identity}\label{sec:identity}

Suppose $ \vdash g : \Nat^\alpha \,\alpha\,\alpha$, let $G^\set =
\setsem{\vdash g : \Nat^\alpha \,\alpha\,\alpha}$, and let $G^\rel =
\relsem{\vdash g : \Nat^\alpha \,\alpha\,\alpha}$.  By
Theorem~\ref{thm:abstraction}, $(G^\set(\pi_1\rho),G^\set(\pi_2\rho))
= G^\rel\rho$. Thus, for all $\rho \in \relenv$ and any $(a, b) \in
\relsem{\vdash \emptyset}\rho = 1$, eliding the only possible
instantiations of $a$ and $b$ gives that
%\[\begin{array}{lll}
$(G^\set, G^\set) \; = \; (G^\set(\pi_1 \rho), G^\set (\pi_2 \rho))$
$ \in \relsem{\vdash \Nat^\alpha \,\alpha\,\alpha}\rho$
$ =  \{\eta : \id \Rightarrow \id\}$
$ =  \{(\eta_1 : \id \Rightarrow \id, \eta_2 : \id \Rightarrow
\id)\}$
%\end{array}\]
That is, $G^\set$ is a natural transformation from the identity
functor on $\set$ to itself. Now let $S$ be any set.  If $S =
\emptyset$, then there is exactly one morphism $\id_S: S \to S$, so
$G^\set_S : S \to S$ must be $\id_S$. If $S \not = \emptyset$, then if
$a$ is any element of $S$ and $K_a :S \to S$ is the constantly
$a$-valued morphism on $S$, then instantiating the naturality square
implied by the above equality gives that $G^\set_S \circ K_a = K_a
\circ G^\set_S$, i.e., $G^\set_S \, a = a$, i.e., $G^\set_S = \id_S$.
Putting these two cases together we have that for every $S : \set$,
$G^\set_S = \id_S$, i.e., $G^\set$ is the identity natural
transformation for the identity functor on $\set$. So every closed
term $g$ of closed type $\Nat^\alpha\alpha\,\alpha$ always denotes the
identity natural transformation for the identity functor on $\set$,
i.e., every closed term $g$ of type $\Nat^\alpha\alpha\,\alpha$
denotes the polymorphic identity function.

\subsection{Standard Free Theorems for ADTs and Their Analogues for
  Nested types}\label{sec:ft-adt}

We can derive in our calculus even those free theorems for polymorphic
functions over ADTs that are not consequences of naturality.  We can,
e.g., prove the free theorem for $\mathit{filter}$'s type as follows:

%\begin{lemma}\label{lem:list-graph}
%If $g : A \to B$, $\rho : \relenv$, and $\rho\alpha = (A, B,
%\graph{g})$, then $\relsem{\alpha; \emptyset \vdash List \, \alpha}
%\rho = \graph{map_{\mathit{List}} \, g}$.
%\end{lemma}
%\begin{proof}
%The result follows by straightforward calculation using
%Definition~\ref{def:rel-sem} and Equation~\ref{eq:mu}, together with
%the facts that $\mathit{List} \,(A, B, \graph{g}) =
%\graph{\mathit{map}_{\mathit{List}} \,g}$ and, if $H = \onet + \beta
%\times\phi \beta$ {\color{blue} Why not $H = \onet + \alpha \times
%  \beta$? Does it matter?}, then $(T_{H,\rho}^n K_0)^* R =
%\Sigma_{i=0}^n R^i$ for all $n \in \nat$ and $R \in \rel$ by induction
%on $n$.
%\end{proof}

\begin{thm} 
If $g : A \to B$, $\rho : \relenv$, $\rho \alpha = (A, B, \graph{g})$,
$(a, b) \in \relsem{\alpha ;\emptyset \vdash \Delta} \rho$, $(s \circ
g, s) \in \relsem{\alpha; \emptyset \vdash \Nat^\emptyset \alpha \,
  \mathit{Bool}} \rho$, and % for $s : A \to \mathit{Bool}$, and
%$\alpha; \emptyset \,|\, \Delta \vdash t :
%\filtype$, if
$\mathit{filter} = \setsem{\alpha; \emptyset \,|\, \Delta \vdash t :
  \filtype}$ for some $t$, then
\[  \mathit{map}_{\mathit{List}} \,g \circ \mathit{filter} \, (\pi_1
\rho) \, a \, (s \circ g) = \mathit{filter}\, (\pi_2\rho) \, b \, s
\circ \mathit{map}_{\mathit{List}} \,g\]
\end{thm}
\begin{proof}
By Theorem~\ref{thm:at-gen-rel}, \[(\mathit{filter}\, (\pi_1 \rho)\, a,
\mathit{filter}\, (\pi_2 \rho)\, b) \in \relsem{\alpha; \emptyset
  \vdash \filtype} \rho\] so if $(s, s') \in \relsem{\alpha; \emptyset
  \vdash \Nat^\emptyset \alpha \, \mathit{Bool}} \rho = \rho\alpha \to
\Eq_{\mathit{Bool}}$ and $(xs, xs') \in \relsem{\alpha; \emptyset
  \vdash List \, \alpha} \rho$ then
% then
%  \begin{align*}
%    (t (\pi_1 \rho) \,a \,s, t (\pi_2 \rho) \,b \,s') &\in \relsem{\alpha; \emptys%et \vdash \Nat^\emptyset (List \, \alpha) \, (List \, \alpha)} \rho \\
%   &= \relsem{\alpha; \emptyset \vdash List \, \alpha} \rho \to \relsem{\alpha; \e%mptyset \vdash List \, \alpha} \rho \\
%  \end{align*}
%  So if $(xs, xs') \in \relsem{\alpha; \emptyset \vdash List \, \alpha} \rho$ then,
\begin{equation}\label{eq:filter-thm-list}
  (\mathit{filter}\, (\pi_1\rho) \,a \,s \,xs, \mathit{filter} \,
  (\pi_2\rho) \,b \,s' \,xs') \in \relsem{\alpha; \emptyset \vdash
    \mathit{List} \, \alpha} \rho
\end{equation}
If $\rho\alpha = (A, B, \graph{g})$, then $\relsem{\alpha; \emptyset
  \vdash \mathit{List} \, \alpha} \rho =
\graph{\mathit{map}_{\mathit{List}} \, g}$ by Lemma~\ref{lem:graph}
and Equation~\ref{thm:demotion-objects}.  Moreover, $xs' =
\mathit{map}_{\mathit{List}} \,g \,xs$ and $(s', s) \in \graph{g} \to
\Eq_{\mathit{Bool}}$, so $s' = s \circ g$. The result
follows from % Instantiating
Equation~\ref{eq:filter-thm-list}.\end{proof}
%gives the result.
% \begin{align*}
%    &(t (\pi_1\rho) \,a \,(s' \circ g) \,xs, t (\pi_2\rho) \,b \,s' \,(\mathit{map%}_{\mathit{List}} \,g \,xs)) \in \graph{\mathit{map}_{\mathit{List}} \,g}, \\ 
%    & i.e., \\ 
%    &\mathit{map}_{\mathit{List}} \,g \, (t (\pi_1\rho) \,a \,(s' \circ g) \,xs) =% t (\pi_2\rho) \,b \,s' \,(\mathit{map}_{\mathit{List}} \,g \,xs), \\ 
%    & i.e., \\
%    &\mathit{map}_{\mathit{List}} \,g \circ t (\pi_1 \rho) \, a \, (s' \circ g) = %t (\pi_2\rho) \, b \, s' \circ \mathit{map}_{\mathit{List}} \,g
%  \end{align*}
%  as desired.
%\end{proof}

\begin{comment}
\begin{thm}\label{thm:at-gen-rel}
If \,$\Gamma; \Phi \,|\, \Delta \vdash t : \tau$ and \,$\rho \in \relenv$,
  and if \,$(a, b) \in \relsem{\Gamma; \Phi \vdash \Delta} \rho$,
  then \\
  $(\setsem{\Gamma; \Phi \,|\, \Delta \vdash t : \tau} (\pi_1 \rho) \,a \, ,
      \setsem{\Gamma; \Phi \,|\, \Delta \vdash t : \tau } (\pi_2 \rho) \,b) \in 
    \relsem{\Gamma; \Phi \vdash \tau} \rho$
\end{thm}
\begin{proof}
 Immediate from Theorem~\ref{thm:at-gen}.
\end{proof}
\end{comment}

\begin{comment}
\begin{thm} 
  If $g : A \to B$, 
  $\rho : \relenv$, 
  $\rho \alpha = (A, B, \graph{g})$,
  $(a, b) \in \relsem{\alpha ;\emptyset \vdash \Delta} \rho$, 
  $(s \circ g, s) \in \relsem{\alpha; \emptyset \vdash \Nat^\emptyset \alpha \, \mathit{Bool}} \rho$,
  and, for some well-formed term $\mathit{filter}$, \\
  $$t = \setsem{\alpha; \emptyset \,|\, \Delta \vdash \mathit{filter} : \filtype} \emph{, then }$$
  \begin{align*}
  \mathit{map}_{\mathit{List}} \,g \circ t (\pi_1 \rho) \, a \, (s \circ g) = t (\pi_2\rho) \, b \, s \circ \mathit{map}_{\mathit{List}} \,g
  \end{align*}
\end{thm}
\begin{proof}
  By Theorem~\ref{thm:at-gen-rel},
  $(t (\pi_1 \rho) a, t (\pi_2 \rho) b) \in \relsem{\alpha; \emptyset \vdash \filtype} \rho$.
  Thus if 
  $(s, s') \in \relsem{\alpha; \emptyset \vdash \Nat^\emptyset \alpha \, \mathit{Bool}} \rho = \rho\alpha \to \Eq_{\mathit{Bool}}$,
  then 
  \begin{align*}
    (t (\pi_1 \rho) \,a \,s, t (\pi_2 \rho) \,b \,s') &\in \relsem{\alpha; \emptyset \vdash \Nat^\emptyset (List \, \alpha) \, (List \, \alpha)} \rho \\
   &= \relsem{\alpha; \emptyset \vdash List \, \alpha} \rho \to \relsem{\alpha; \emptyset \vdash List \, \alpha} \rho \\
  \end{align*}
  So if $(xs, xs') \in \relsem{\alpha; \emptyset \vdash List \, \alpha} \rho$ then,
  \begin{equation}\label{eq:filter-thm}
  (t (\pi_1\rho) \,a \,s \,xs, t (\pi_2\rho) \,b \,s' \,xs') \in \relsem{\alpha; \emptyset \vdash List \, \alpha} \rho
  \end{equation}

  Consider the case in which $\rho\alpha = (A, B, \graph{g})$. Then $\relsem{\alpha; \emptyset \vdash List \, \alpha} \rho
  = \graph{\mathit{map}_{\mathit{List}} \, g}$, by Lemma~\ref{lem:list-graph}, and $(xs, xs') \in \graph{\mathit{map}_{\mathit{List}} \,g}$ 
  implies $xs' = \mathit{map}_{\mathit{List}} \,g \,xs$. We also have that 
  $(s, s') \in \graph{g} \to \Eq_{\mathit{Bool}}$ implies 
  $\forall (x, g x) \in \graph{g}. \,\, s x = s' (g x)$ and thus
  $s = s' \circ g$ due to the definition of morphisms between relations.
  %% TODO show (s, s') \in <g> -> \mathit{Bool} implies s = s' \circ g %% 
  With these instantiations, Equation~\ref{eq:filter-thm} becomes
  \begin{align*}
    &(t (\pi_1\rho) \,a \,(s' \circ g) \,xs, t (\pi_2\rho) \,b \,s' \,(\mathit{map}_{\mathit{List}} \,g \,xs)) \in \graph{\mathit{map}_{\mathit{List}} \,g}, \\ 
    & i.e., \\ 
    &\mathit{map}_{\mathit{List}} \,g \, (t (\pi_1\rho) \,a \,(s' \circ g) \,xs) = t (\pi_2\rho) \,b \,s' \,(\mathit{map}_{\mathit{List}} \,g \,xs), \\ 
    & i.e., \\
    &\mathit{map}_{\mathit{List}} \,g \circ t (\pi_1 \rho) \, a \, (s' \circ g) = t (\pi_2\rho) \, b \, s' \circ \mathit{map}_{\mathit{List}} \,g
  \end{align*}
  as desired.
\end{proof}
\end{comment}

%\subsection{Free Theorem for Type of $\mathit{filter}$ for
%  $\GRose$}\label{sec:filter-grose}

A similar proof establishes the analogous result for, say, generalized
rose trees. 
\begin{thm} 
  If $g : A \to B$,
$F, G : \set \to \set$,
  $\eta : F \to G$ in $\set$, $\rho : \relenv$, $\rho \alpha =
 (A, B, \graph{g})$, $\rho \psi = (F, G, \graph{\eta})$, $(a, b) \in
 \relsem{\alpha, \psi ;\emptyset \vdash \Delta} \rho$, $(s \circ
 g, s) \in \relsem{\alpha; \emptyset \vdash \Nat^\emptyset \alpha \,
   \mathit{Bool}} \rho$, and
 \[ \mathit{filter} = \setsem{\alpha, \psi; \emptyset \,|\, \Delta
      \vdash t : \filtypeGRose}  \]
for some $t$, then
\[ \semmap_{\mathit{GRose}}\, \eta\, (g + 1) \circ \mathit{filter} \,
(\pi_1 \rho) \, a \, (s \circ g) = \mathit{filter} \, (\pi_2\rho) \, b
\, s \circ 
\semmap_{\mathit{GRose}}\, \eta\, g\]
\end{thm}

\noindent
This is not surprising since rose trees are essentially
ADT-like. However, as noted in Section~\ref{sec:terms}, our calculus
cannot express the type of a polymorphic filter function for a proper
nested type.

\subsection{Short Cut Fusion for Lists}\label{sec:short-cut}

We can recover standard short cut fusion for lists~\cite{glp93} in our
calculus: 
\begin{thm}
If $\vdash F$, $\vdash H$, and 
%$\beta; \emptyset \,|\, \emptyset \vdash g : \Nat^{\emptyset}
%(\Nat^{\emptyset} (\onet + \tau \times \beta)\, \beta)\, \beta$. 
%If
$G \, = \, \setsem{\beta; \emptyset \,|\, \emptyset \vdash g :
  \Nat^{\emptyset} (\Nat^{\emptyset} (\onet + F \times \beta)\,
  \beta)\, \beta}$ for some $g$, then
\[\mathit{fold}_{1 + F \times \_}\, n\, c\; (G\; (\mathit{List}\,
\setsem{\vdash F})\,\mathit{nil} \,\mathit{cons}) = G \,\setsem{\vdash
  H}\, n\, c \]
\end{thm}
\begin{proof}
%Let $\vdash \tau : \F$ and $\vdash \tau' : \F$, let
%\[\beta; \emptyset \,|\, \emptyset \vdash g : \Nat^{\emptyset}
%(\Nat^{\emptyset} (\onet + \tau \times \beta)\, \beta)\, \beta\]
%and let
%\[G = \setsem{\beta; \emptyset \,|\, \emptyset
%  \vdash g : \Nat^{\emptyset} (\Nat^{\emptyset} (\onet + \tau \times
  %  \beta)\, \beta)\, \beta}\] Then
Theorem~\ref{thm:abstraction} gives
that, for any $\rho : \relenv$,
%and any $(a, b) \in
%\relsem{\beta; \emptyset \vdash \emptyset}\rho = 1$, then (eliding the
%only possible instantiations of $a$ and $b$) we have
\[(G \,(\pi_1 \rho), G\, (\pi_2 \rho))
 \in  \relsem{\beta; \emptyset \vdash \Nat^{\emptyset} (\Nat^{\emptyset}
  (\onet + F \times \beta)\, \beta)\, \beta} \rho\\
\cong (((\relsem{\vdash F} \rho \times
\rho\beta) \to \rho\beta) \times \rho\beta) \to \rho\beta\] 
%\[\begin{array}{lll}
%(G \,(\pi_1 \rho), G\, (\pi_2 \rho))
% & \in & \relsem{\beta; \emptyset \vdash \Nat^{\emptyset} (\Nat^{\emptyset}
%  (\onet + \tau \times \beta)\, \beta)\, \beta} \rho\\
%& \cong& (((\relsem{\vdash \tau} \rho \times
%\rho\beta) \to \rho\beta) \times \rho\beta) \to \rho\beta 
%\end{array}\]
%
%Since
%\[\begin{array}{rl}
% & \relsem{\beta; \emptyset \vdash \Nat^{\emptyset}
%  (\Nat^{\emptyset} (\onet + \tau \times \beta)\, \beta)\, \beta} \rho\\  
%=& \relsem{\beta; \emptyset \vdash \Nat^{\emptyset} (\onet +
%  \tau \times \beta)\, \beta} \rho \to \rho\beta \\ 
%=& (\relsem{\beta; \emptyset \vdash \onet + \tau \times \beta}
%\rho \to \rho\beta) \to \rho\beta \\ 
%=& ((\onet + \relsem{\vdash \tau} \rho \times
%\rho\beta) \to \rho\beta) \to \rho\beta \\ 
%\cong& (((\relsem{\vdash \tau} \rho \times
%\rho\beta) \to \rho\beta) \times \rho\beta) \to \rho\beta 
%\end{array}\]
%we have that
so if $(c', c) \in \relsem{\vdash F} \rho \times
\rho\beta \to \rho\beta$ and $(n', n) \in \rho\beta$ then
$(G \,(\pi_1 \rho)\, n'\, c', G\, (\pi_2 \rho)\, n\, c)
\in \rho \beta$.
In addition,
\[\setsem{\vdash \fold_{\onet + F
    \times \beta}^{H} : \Nat^{\emptyset} (\Nat^{\emptyset} (\onet
  + F \times H)\, \tau')\, (\Nat^{\emptyset} (\mu \alpha. \onet
  + F \times \alpha)\, H)} =
\mathit{fold}_{1 + F \times \_}\] so that if $c \in
\setsem{\vdash F} \,\times\, \setsem{\vdash H} \to
\setsem{\vdash H}$ and $n \in \setsem{\vdash H}$, then
$(n, c) \in \setsem{\vdash \Nat^{\emptyset} (\onet + F \times H)\,
  H}$. The instantiation
$\pi_1 \rho \beta = \setsem{\vdash \mu \alpha. \onet + F \times
  \alpha}=\mathit{List}\,\setsem{\vdash F}$,\, 
$\pi_2 \rho \beta = \setsem{\vdash H}$,\,
$\rho \beta = \graph{\mathit{fold}_{1 + F \times \_}\, n\, c} :
\rel(\pi_1 \rho \beta, \pi_2 \rho \beta)$,\,
$c' = \mathit{cons}$,\, and 
$n' = \mathit{nil}$
%Consider the instantiation:
%\[\begin{array}{rcl}
%\pi_1 \rho \beta &=& \setsem{\vdash \mu \alpha. \onet + \tau \times
%  \alpha} \;=\;\List\,\setsem{\vdash \tau}\\   
%\pi_2 \rho \beta &=& \setsem{\vdash \tau'}\\
%\rho \beta &=& \graph{\mathit{fold}_{1 + \tau \times \_}\, n\, c} :
%\rel(\pi_1 \rho \beta, \pi_2 \rho \beta) \\ 
%c' &=& \mathit{cons}\\
%n' &=& \mathit{nil}
%\end{array}\]
%Clearly, $(\mathit{nil}, n) \in \rho\beta =
%\graph{\mathit{fold}_{1 + \tau \times \_}\, n\, c}$ because $
%\mathit{fold}_{1 + \tau \times \_}\, n\, c \,\mathit{nil} = n$.  Moreover,
%$(\mathit{cons}, c) \in \relsem{\vdash \tau} \,\times\, \rho\beta \to
%\rho\beta$ since if $(x, x') \in \relsem{\vdash \tau}$, i.e., $x =
%x'$, and if $(y, y') \in \rho\beta =
%\graph{\mathit{fold}_{1 + \tau \times \_} n\, c}$, i.e., $y' =
%\mathit{fold}_{1 + \tau \times \_} n\, c\, y$, then
%\[(\mathit{cons}\, x\, y, c\, x\, (\mathit{fold}_{1 + \tau \times \_}\, n\,
%c\, y)) \in \graph{\mathit{fold}_{1 + \tau \times \_}\, n\, c}\]
%i.e.,
%\[c\, x\, (\mathit{fold}_{1 + \tau \times \_}\, n\, c\, y)
%= \mathit{fold}_{1 + \tau \times \_}\, n\, c\, (\mathit{cons} \,x\, y)\]
%holds by definition of $\mathit{fold}_{1 + \tau \times \_}$.  We
%therefore conclude that
thus gives that $(G \;(\mathit{List}\,\setsem{\vdash
  F})\,\mathit{nil} \,\mathit{cons}, G \,\setsem{\vdash H} \,
n\, c) \in \graph{\mathit{fold}_{1 + F \times \_}\, n\, c}$, i.e.,
that $\mathit{fold}_{1 + F \times \_}\, n\, c\; (G\;
(\mathit{List}\,\setsem{\vdash F})\, \mathit{nil}\, \mathit{cons})
= G \,\setsem{\vdash H}\, n\, c$.
\end{proof}

Using Equation~\ref{thm:demotion-objects} we can extend short
cut fusion results to arbitrary ADTs, as in~\cite{joh02}.

\begin{comment}
\subsection{Short Cut Fusion for Arbitrary ADTs}\label{sec:short-cut-adt}

\begin{thm}
Let $\ol{\vdash \tau}$, let $\vdash \tau'$, let $\ol{\alpha}; \beta
\vdash F$, and let $\beta; \emptyset  \,|\,
\emptyset \vdash g : \Nat^{\emptyset} (\Nat^{\emptyset} F[\ol{\alpha
    := \tau}]\, \beta)\, \beta$.  If we regard
\[\begin{array}{lll}
H & = & \setsem{\emptyset;\beta \vdash F[\ol{\alpha :=
      \tau}]}\\
G & = & \setsem{\beta; \emptyset \,|\, \emptyset \vdash g :
  \Nat^{\emptyset} (\Nat^{\emptyset}\, F[\ol{\alpha := \tau}]\,
  \beta)\, \beta}
\end{array}\]
as functors in $\beta$, then for every $B \in H \setsem{\vdash \tau'}
\rightarrow \setsem{\vdash \tau'}$ we have
\[\mathit{fold}_H\, B \; (G\; \mu H \; in_H) = G \,\setsem{\vdash \tau'}\, B\]
\end{thm}

\vspace*{0.2in}

\begin{proof}
  We first note that the type of $g$ is well-formed, since
  $\emptyset;\beta \vdash F[\ol{\alpha := \tau}]$ so our
  promotion theorem gives that $\beta;\emptyset\vdash F[\ol{\alpha :=
      \tau}]$, and $\emptyset;\beta\vdash\beta$ so that our
  promotion theorem gives $\beta;\emptyset\vdash\beta$. From
  these facts we deduce that $\beta;\emptyset \vdash
  \Nat^\emptyset\,F[\ol{\alpha := \tau}]\,\beta : \T$, and thus that
  $\beta; \emptyset \vdash \Nat^{\emptyset} (\Nat^{\emptyset}
  F[\ol{\alpha := \tau}]\, \beta)\, \beta : \T$.

  Theorem~\ref{thm:abstraction} gives that, for any $\rho : \relenv$
  and any $(a, b) \in \relsem{\beta; \emptyset \vdash \emptyset}\rho =
  1$, eliding the only possible instantiations of $a$ and $b$ gives
  that
\[(G \,(\pi_1 \rho), G\, (\pi_2 \rho))
\in \relsem{\beta; \emptyset \vdash \Nat^{\emptyset} (\Nat^{\emptyset}
  F[\ol{\alpha := \tau}]\, \beta)\, \beta} \rho\]
Since
\[\begin{array}{rl}
 & \relsem{\beta; \emptyset \vdash \Nat^{\emptyset}
  (\Nat^{\emptyset} F[\ol{\alpha := \tau}]\, \beta)\, \beta} \rho\\  
=& \relsem{\beta; \emptyset \vdash \Nat^{\emptyset} F[\ol{\alpha :=
      \tau}] \, \beta} \rho \to \rho\beta \\ 
\end{array}\]
we have that if $(A, B) \in \relsem{\beta; \emptyset \vdash
  \Nat^{\emptyset} F[\ol{\alpha := \tau}] \, \beta} \rho$
then
\[(G \,(\pi_1 \rho)\, A, G\, (\pi_2 \rho)\, B)
\in \rho \beta\]
Now note that
\[\setsem{\vdash \fold_{F[\ol{\alpha := \tau}]}^{\tau'} :
  \Nat^{\emptyset}\, (\Nat^{\emptyset}\, F[\ol{\alpha := \tau}][\beta
    := \tau'] \, \tau')\, (\Nat^{\emptyset} \,(\mu
  \beta. F[\ol{\alpha := \tau}] \, \tau')} =
\mathit{fold}_H\]
and consider the instantiation
\[\begin{array}{lll}
A &=& \mathit{in}_H :  H (\mu H) \to \mu H\\
B & : & H\setsem{\vdash \tau'} \to \setsem{\vdash \tau'}\\
\rho \beta &=& \graph{\mathit{fold}_H\, B}
\end{array}\]
(Note that all the types here are well-formed.)  This gives
\[\begin{array}{lll}
\pi_1 \rho \beta &=& \setsem{\vdash \mu \beta. F[\ol{\alpha := \tau}]}
\; = \; \mu H\\
\pi_2 \rho \beta &=& \setsem{\vdash \tau'}\\
\rho \beta &:& \rel(\pi_1 \rho \beta, \pi_2 \rho \beta)\\
A & : & \setsem{\beta; \emptyset \vdash \Nat^{\emptyset}
  F[\ol{\alpha := \tau}] \, \beta} (\pi_1 \rho)\\
B & : & \setsem{\beta; \emptyset \vdash \Nat^{\emptyset} F[\ol{\alpha
      := \tau}] \, \beta}(\pi_2 \rho)\\
\end{array}\]
since
\[\begin{array}{lll}
A \; = \; \mathit{in}_H & : & H(\mu H) \to \mu H\\
 & = & \setsem{\emptyset;\beta \vdash F[\ol{\alpha := \tau}]}(\mu
\setsem{\emptyset;\beta \vdash F[\ol{\alpha := \tau}]}) \to \mu
\setsem{\emptyset;\beta \vdash F[\ol{\alpha := \tau}]}\\
& = & \setsem{\emptyset;\beta \vdash F[\ol{\alpha :=
      \tau}]}(\pi_1\rho) \to \setsem{\emptyset;\beta \vdash
  \beta}(\pi_1 \rho)\\
& = & \setsem{\beta;\emptyset \vdash F[\ol{\alpha :=
      \tau}]}(\pi_1\rho) \to \setsem{\beta; \emptyset \vdash
  \beta}(\pi_1 \rho) \hspace*{0.25in} {\color{red} \mbox{Daniel's
    trick; now a theorem}} \\ 
& = & \setsem{\beta;\emptyset \vdash \Nat^\emptyset F[\ol{\alpha :=
      \tau}]\,\beta}(\pi_1 \rho) 
\end{array}\]
{\color{red} where ``Daniel's trick'' is the observation that a
  functor can be seen as non-functorial when we only care about its
  action on objects. This is now a theorem.}  We also have
\[\begin{array}{lll}
(A,B) \,=\, (\mathit{in}_H,B) & \in & \relsem{\beta; \emptyset \vdash
  \Nat^{\emptyset} F[\ol{\alpha := \tau}] \, \beta} \rho \\
 & = & \relsem{\beta; \emptyset \vdash F[\ol{\alpha :=
      \tau}]}\rho[\beta := \graph{\mathit{fold}_H\, B}] \to
\graph{\mathit{fold}_H\, B}\\
 & = & \relsem{\beta; \emptyset \vdash F[\ol{\alpha :=
      \tau}]}\graph{\mathit{fold}_H\, B} \to
\graph{\mathit{fold}_H\, B}\\
 & = & \relsem{\emptyset; \beta\vdash F[\ol{\alpha :=
      \tau}]}\graph{\mathit{fold}_H\, B} \to
\graph{\mathit{fold}_H\, B} \hspace*{0.35in} {\color{red}
  \mbox{Daniel's trick; now a theorem}}\\  
& = & \graph{\setsem{\emptyset; \beta \vdash F[\ol{\alpha :=
      \tau}]}\,(\mathit{fold}_H\, B)} \to
\graph{\mathit{fold}_H\, B} \hspace*{0.25in} {\color{red} \mbox{by the
    graph lemma}}\\ 
& = & \graph{\mathit{map}_H \,(\mathit{fold}_H\,B)} \to
\graph{\mathit{fold}_H\, B}\\
\end{array}\]
since if $(x,y) \in \graph{\mathit{map}_H \,(\mathit{fold}_H\,B)}$,
i.e., if $\mathit{map}_H \,(\mathit{fold}_H\,B) \, x = y$, then
$\mathit{fold}_H\, B\, (\mathit{in}_H\,x) = B\,y = B\, (\mathit{map}_H
\,(\mathit{fold}_H\,B) \, x)$ by the definition of $\mathit{fold}_H$
as a (indeed, the unique) morphism from $\mathit{in}_H$ to $B$. Thus,
\[(G \,(\pi_1 \rho)\, A, G\, (\pi_2 \rho)\, B) \in
\graph{\mathit{fold}_H\, B}\]
i.e.,
\[\mathit{fold}_H \, B \, (G\, (\pi_1 \rho) \, \mathit{in}_H) =
G\,(\pi_2 \rho)\,B\]

\vspace*{0.1in}

\noindent
Since $\beta$ is the only free variable in $G$, this simplifies to
\[\mathit{fold}_H\, B \, (G\, \mu H\, \mathit{in}_H) =
G\,\setsem{\vdash \tau'}\,B\]  
\end{proof}
\end{comment}

\subsection{Short Cut Fusion for Arbitrary Nested
  Types}\label{sec:short-cut-nested}

%fold/build rule shows that Church encodings for nested types are iso
%(have the same interp) as the nested types themselves. Just like Ex
%2.8 in parpoe, but categorical rather than operational semantics. But
%built-in data types are more versatile in implementations. They can be
%inducted on, e.g. And they are stored in the heap, not the run-time
%stack, and are therefore more efficient.
%``Most higher order type languages meant for human programmers eschew
%fully impredicative polymorphism'' says Andy on p322.

We can also, finally, formally prove correctness of the categorically
inspired short cut fusion for nested types from~\cite{jg10}. Indeed,
we have:
\begin{thm}\label{thm:short-cut-nested}
If $\emptyset;\phi,\alpha \vdash F$, \,$\emptyset; \alpha
\vdash K$, \,
%let $\phi ;\emptyset\,|\,\emptyset\vdash g :
%\Nat^\emptyset\,(\Nat^\alpha\,F\,(\phi\alpha)) \,
%(\Nat^\alpha\,\onet \, (\phi\alpha))$. If we let
$H : [\set,\set] \to [\set,\set]$ is defined by
\[\begin{array}{lll}
H\,f\,x & = & \setsem{\emptyset; \phi, \alpha \vdash F}[\phi :=
  f][\alpha := x]\\
\end{array}\]
and 
\[G = \setsem{\phi;\emptyset\,|\,\emptyset \vdash g :
\Nat^\emptyset\,(\Nat^\alpha\,F\,(\phi\alpha))\,(\Nat^\alpha\,\onet \,
(\phi\alpha))}\] for some $g$, then, for every $B \in H
\setsem{\emptyset;\alpha \vdash K} \rightarrow \setsem{\emptyset;
  \alpha \vdash K}$,
$\mathit{fold}_{H}\, B \, (G\; \mu H \; \mathit{in}_{H}) = G
\,\setsem{\emptyset;\alpha \vdash K}\, B$.
\end{thm}

\begin{proof}
%We first note that the type of $g$ is well-formed since
%$\emptyset;\phi,\alpha \vdash F : \F$ so our promotion theorem gives
%that $\phi;\alpha \vdash F : \F$, and $\phi;\alpha \vdash \phi\alpha :
%\F$, so that $\phi;\emptyset \vdash \Nat^\alpha F\,(\phi\alpha) : \T$
%and $\phi;\emptyset \vdash \Nat^\alpha \onet \,(\phi\alpha) : \T$.
%Then $\phi;\emptyset \vdash \Nat^\alpha F\,(\phi\alpha) : \F$ and
%$\phi;\emptyset \vdash \Nat^\alpha \onet \,(\phi\alpha) : \F$ also
%%%hold, and, finally, 
%$\phi ;\emptyset\vdash 
%\Nat^\emptyset\,(\Nat^\alpha\,F\,(\phi\alpha)) \,
%(\Nat^\alpha\,\onet \, (\phi\alpha)) : \T$
%  
Theorem~\ref{thm:abstraction} gives that, for any 
$\rho : \relenv$,
%and any $(a, b) \in \relsem{\phi,\alpha;\emptyset \vdash
%  \emptyset}\rho = 1$, eliding the only possible instantiations
%of $a$ and $b$ gives that
\[\begin{array}{lll}
(G \,(\pi_1 \rho), G\, (\pi_2 \rho))
& \in & \relsem{\phi;\emptyset\vdash \Nat^{\emptyset} (\Nat^\alpha
  F\, (\phi\alpha))\, (\Nat^\alpha\,\onet \, (\phi\alpha))}
\rho\\ 
& = & \relsem{\phi;\emptyset\vdash \Nat^\alpha F\,
  (\phi\alpha)}\rho \to \relsem{\phi;\emptyset\vdash
  \Nat^\alpha\,\onet \, (\phi\alpha)}\rho\\ 
%& = & \relsem{\phi;\emptyset\vdash \Nat^\alpha F\,
%  (\phi\alpha)}\rho \to (\,\lambda A. 1 \Rightarrow \lambda A.\,(\rho
%\phi) A\,)\\   
%& = & \relsem{\phi;\emptyset\vdash \Nat^\alpha F\,
%  (\phi\alpha)}\rho \to (1 \Rightarrow \rho \phi)\\ 
& = & \relsem{\phi;\emptyset\vdash \Nat^\alpha F\,
  (\phi\alpha)}\rho \to \rho \phi
\end{array}\]
\noindent
so if $(A, B) \in \relsem{\phi;\emptyset\vdash \Nat^\alpha F\,
  (\phi\alpha)}\rho$ then $(G \,(\pi_1 \rho)\, A, G\, (\pi_2 \rho)\,
B) \in \rho \phi$.
In addition,
%$\mathit{fold}_H = \setsem{\vdash \fold_F^K :
%  \Nat^{\emptyset}\, (\Nat^\alpha F[\phi := K]\,K)\, (\Nat^\alpha
%  ((\mu \phi.\lambda\alpha.F)\alpha)\,K)}$.
\[\setsem{\vdash \fold_F^K :
  \Nat^{\emptyset}\, (\Nat^\alpha F[\phi := K]\,K)\, (\Nat^\alpha
  ((\mu \phi.\lambda\alpha.F)\alpha)\,K)} = \mathit{fold}_H\]
Consider the instantiation $A = \mathit{in}_H : H (\mu H) \Rightarrow
\mu H$,\, $B : H\setsem{\emptyset;\alpha\vdash K} \Rightarrow
\setsem{\emptyset;\alpha \vdash K}$,\, $\rho \phi =
\graph{\mathit{fold}_H\, B}$,\, $\pi_1 \rho \phi = \mu H$,\, $\pi_2
\rho \phi = \setsem{\emptyset;\alpha\vdash K}$,\, $\rho \phi :
\rel(\pi_1 \rho \phi, \pi_2 \rho \phi)$,\, $A : \setsem{\phi;
  \emptyset \vdash \Nat^\alpha F \, (\phi \alpha)} (\pi_1 \rho)$,\,
and $B : \setsem{\phi; \emptyset \vdash \Nat^\alpha F \, (\phi
  \alpha)} (\pi_2 \rho)$.
%\[\begin{array}{lll}
%A &=& \mathit{in}_H :  H (\mu H) \Rightarrow \mu H\\
%B & : & H\setsem{\emptyset;\alpha\vdash K} \Rightarrow
%\setsem{\emptyset;\alpha \vdash K}\\
%\rho \phi &=& \graph{\mathit{fold}_H\, B} \hspace*{0.2in} {
%  \color{red} \mbox{a graph of a natural transformation, defined in
%    Enrico's notes}} 
%\end{array}\]
%(Note that all the types here are well-formed.) This gives
%\[\begin{array}{lll}
%\pi_1 \rho \phi &=& \mu H\\
%\pi_2 \rho \phi &=& \setsem{\emptyset;\alpha\vdash K}\\
%\rho \phi &:& \rel(\pi_1 \rho \phi, \pi_2 \rho \phi)\\
%A & : & \setsem{\phi; \emptyset \vdash \Nat^\alpha
%  F \, (\phi \alpha)} (\pi_1 \rho)\\
%B & : & \setsem{\phi; \emptyset \vdash \Nat^\alpha
%  F \, (\phi \alpha)} (\pi_2 \rho)\\
%\end{array}\]
%since
Equation~\ref{thm:demotion-objects} ensures that $A = \mathit{in}_H :
H(\mu H) \Rightarrow \mu H = \setsem{\phi;\emptyset \vdash \Nat^\alpha
  F \,(\phi\alpha)}(\pi_1\rho)$,
%\[\begin{array}{lll}
%A \; = \; \mathit{in}_H & : & H(\mu H) \Rightarrow \mu H\\
%& = & \setsem{\emptyset; \phi, \alpha \vdash F}[\phi := \mu
%  \setsem{\emptyset; \phi, \alpha \vdash F}] \Rightarrow \mu
%  \setsem{\emptyset; \phi, \alpha \vdash F}\\  
%& = & \setsem{\emptyset; \phi, \alpha \vdash F}(\pi_1\rho)
%  \Rightarrow \setsem{\emptyset; \phi,\alpha \vdash \phi\alpha}(\pi_1 \rho)\\
%& = & \setsem{\phi; \alpha \vdash F}(\pi_1\rho)
%  \Rightarrow \setsem{\phi;\alpha \vdash \phi\alpha}(\pi_1
%  \rho) \hspace*{0.25in} {\color{red} \mbox{Daniel's trick; now a theorem}}\\
%& = & \setsem{\phi;\emptyset \vdash \Nat^\alpha F
%    \,(\phi\alpha)}(\pi_1\rho) 
%\end{array}\]
and Equation~\ref{thm:demotion-objects} and Lemma~\ref{lem:graph}
together give that
\[\begin{array}{lll}
(A,B) \,=\, (\mathit{in}_H,B) & \in & \relsem{\phi;\emptyset\vdash
  \Nat^\alpha F\, (\phi\alpha)}\rho\\
% & = & \lambda A. \relsem{\phi;\alpha\vdash F}\rho[\alpha := A]
%\Rightarrow \lambda A. (\rho\phi) A\\
& = & \lambda A. \relsem{\phi;\alpha\vdash F}[\phi :=
  \graph{\mathit{fold}_H\, B}][\alpha := A] \Rightarrow 
 \graph{\mathit{fold}_H\, B} \\ 
%& = & \lambda A. \relsem{\emptyset;\phi,\alpha\vdash F}[\phi :=
%  \graph{\mathit{fold}_H\, B}][\alpha := A] \Rightarrow 
%\graph{\mathit{fold}_H\, B}  \hspace*{0.25in} {\color{red}
%  \mbox{Daniel's trick; now a theorem}}\\  
& = & \relsem{\emptyset;\phi,\alpha\vdash F}
  \graph{\mathit{fold}_H\, B} \Rightarrow \graph{\mathit{fold}_H\,
    B}\\
  & = & \graph{\setsem{\emptyset;\phi,\alpha\vdash F}
    \,(\mathit{fold}_H\,B)} \Rightarrow \graph{\mathit{fold}_H\, B}\\
%\hspace*{0.25in} {\color{red} \mbox{Graph Lemma}}\\
  & = & \graph{\mathit{map}_H \,(\mathit{fold}_H\,B)} \Rightarrow
\graph{\mathit{fold}_H\, B}\\
\end{array}\]
since if $(x,y) \in \graph{\mathit{map}_H \,(\mathit{fold}_H\,B)}$,
%i.e., if $\mathit{map}_H \,(\mathit{fold}_H\,B) \, x = y$,
then $\mathit{fold}_H\, B\, (\mathit{in}_H\,x) = B\,y = B\,
(\mathit{map}_H \,(\mathit{fold}_H\,B) \, x)$ by the definition of
$\mathit{fold}_H$ as a (indeed, the unique) morphism from
$\mathit{in}_H$ to $B$.  Thus, $(G \,(\pi_1 \rho)\, A, G\, (\pi_2
\rho)\, B) \in \graph{\mathit{fold}_H\, B}$, i.e., $\mathit{fold}_H \,
B \, (G\, (\pi_1 \rho) \, \mathit{in}_H) = G\,(\pi_2 \rho)\,B$.  But
since $\phi$ is the only free variable in $G$, this simplifies to
$\mathit{fold}_H\, B \, (G\, \mu H\, \mathit{in}_H) =
G\,\setsem{\emptyset;\alpha\vdash K}\,B$. 
\end{proof}

\vspace*{-0.07in}

As in~\cite{jg10}, replacing $\onet$ with any type $\emptyset;\alpha
\vdash C$ generalizes
%with $\setsem{\emptyset;\alpha \vdash c}\rho = C$
Theorem~\ref{thm:short-cut-nested} to deliver more general a free
theorem whose conclusion is $\mathit{fold}_{H}\, B \; \circ \; G\; \mu
H \; \mathit{in}_{H} = G \,\setsem{\emptyset;\alpha \vdash K}\, B$.

\vspace*{0.02in}

It is also standard to prove that the parametric model constructed
verifies the existence of initial algebras. This is, however,
unnecessary here since initial algebras are built directly into our
model.

\section{Conclusion and Directions for Future Work}\label{sec:conclusion}

In this paper we have constructed a parametric model for a calculus
providing primitives for constructing nested types directly as
fixpoints rather than representing them via their Chuch encodings. We
have also used the Abstraction Theorem for this model to prove free
theorems in the presence of nested types. This was not possible
before~\cite{jp19} because the nested types definable in our syntax
were not previously known to have well-defined interpretations in
locally finitely presentable categories like $\set$ and $\rel$. No
calculus for terms of these types existed, either. The key to
obtaining our parametric model is the delicate threading of
functoriality and its accompanying naturality conditions throughout
the model construction.

Note that all recursion in our calculus must be expressed using
folds. Expressivity of standard folds for nested types has long been
known to be a vexing issue; indeed, such folds can only express
computations that return natural transformations. Of course our
calculus inherits the constraints standard folds for nested types
impose, and thus is perhaps not as expressive as we would ideally
like. But since this expressivity issue is orthogonal to constructing
parametric models in the presence of nested types built from
primitives, and since it would detreact from the conceptual clarity of
our presentation, we have deliberately chosen not to address it in
this paper. (However, we discuss some ideas for increasing
expressivity in the next paragraph.) Despite the limitations of
standard folds, our results provide a blueprint for deriving
principled parametricity results for practical programming languages,
such as Haskell and Agda, that directly construct nested types as
fixpoints.

Many constructions that enhance standard folds have been
proposed. Perhaps the most significant of these are generalized
folds~\cite{bp99},
%and Mendler iterators~\cite{amu05}.
which are used to express computations whose results are not natural
transformations. Generalized folds are shown in~\cite{jg10} to be
equivalent to standard folds in the presence of {\em right} Kan
extensions.  This suggests extending our term calculus %presented here
with an object-level right Kan extension construct; interpreting such
a construct in $\set$ and $\rel$ should be unproblematic {\color{blue}
  since these categories have all(?) limits}. It was also shown
in~\cite{jg08} that GADTs can be represented using an object-level
{\em left} Kan extension construct, and in~\cite{jp19} that augmenting
the types in our calculus with a carefully designed such construct
preserves the cocontinuity needed for GADTs to have well-defined
interpretations in $\set$ and $\rel$. These observations suggest
augmenting our type formation rules with such a construct.
%natural directions for future work.
Another possibility is to add
term-level fixpoints
%, as in~\cite{pit00},
to our calculus. However,
this would require the categories interpreting types to be not just
locally finitely presentable categories, but to support some kind of
domain structure as well. Unfortunately, $\omega$-CPO, the most
obvious category of domains to replace $\set$, is not locally finitely
presentable.

The constructions of this paper should all be possible in any locally
presentable cartesian closed category (lpccc) $\cal C$ whose category
of (abstract) relations, obtained by pullback as in~\cite{jac99}, is
also a lpccc and is appropriately fibred over $C$.  This would
%extend the results presented here along the lines of~\cite{gjfor15},
%thus giving
result in a framework for constructing parametric models for calculi
supporting nested types built from primitives that is based on
``locally presentable fibrations'', for some appropriate definition
thereof.

\vspace*{0.1in}

\noindent

%Won't print in anonymous mode.
\begin{acks}
We thank Andrew Polonsky for valuable feedback and fruitful
discussions. Supported by NSF awards CCR-1906388 and 1420175.
\end{acks}

\bibliography{references}

\end{document}

fixpoints at term level ala Pitts:

I'm not sure if we actually want to do this or not (because not all
fixpoints will necessarily exist, and this will have consequences, and
because of the restriction to strict functions it will entail and the
fact that I haven't yet thought about how those will thread through
our calculus), but another idea we might consider is adding fixpoints
at the term level. This is done for System F in Wadler's section
7. [It is also done in Pitts' Parametric Polymorphism and Operational
  Equivalence paper, where the term introduction rule is SEE SOURCE

%Gamma \vdash F : \tau \to \tau
%-------------------------------
%Gamma \vdash fix (F) : tau

But of course the model there is entirely operational rather than
categorical.]

On the other hand, check out the last paragraph of Wadler's paper, which
suggests that (at least in 1989) there was some interest in calculi like
ours:

"The desire to derive theorems from types therefore suggests that it would be
valuable to explore programming languages that prohibit recursion or only
allow its restricted use. In theory, this is well understood; we have already
noted that any computable function that is provably total in second-order
Peano arithmetic can be defined in the pure polymorphic lambda calculus,
without using the fixpoint as a primitive. However, practical languages based
on this notion remain terra incognita."

Of course we are not working in System F, but perhaps the same desire applies
here, and we do get the same kinds of theorems as Wadler, many just as a
consequence of our interpretation, without invoking parametricity.

In any case, I guess Wadler could be cited as justifying restricting
our attention to our calculus since it could be seen as a way to have
a practical language that ensures that all programs are
terminating. This has been considered in the past, e.g., in ADL,
Charity, PolyP.

TERMINATING RECURSION
Another possible solution to the expressivity issue is extend folds to
track of deep structure of nested types as in~\cite{jp19}, so that
specialization of fixpoint can be computed as fixpoint of
specialization.

Add more polymorphism (all foralls), even though most free theorems
only use one level (or maybe two, like short cut).

### 2020-06-21 free-theorems-nesteds/paper3.tex
small update to paper3.tex; also merged eqn2.tex with old supp.tex to get new supp.tex for submission

@@
 cannot be the (co)domain of a $\Nat$ type representing a natural
-transformation in $\phi$, and the type $\mu \phi. \lambda
+transformation functorial in $\phi$, and the type $\mu \phi. \lambda
 \alpha. \mathit{GRose}\,\phi\alpha$ is not well-formed in our

### 2020-06-21 free-theorems-nesteds/paper3.tex
modifications with Enrico

@@
 \AXC{$\Gamma;\ol{\alpha^0} \vdash F$}
-\AXC{$\Gamma;\Phi,\ol{\alpha^0}  \vdash G$}
-\BIC{$\Gamma;\Phi \vdash \Nat^{\ol{\alpha^0}}F \,G$}
+\AXC{$\Gamma;\ol{\gamma^0},\ol{\alpha^0}  \vdash G$}
+\BIC{$\Gamma;\ol{\gamma^0} \vdash \Nat^{\ol{\alpha^0}}F \,G$}
 \DisplayProof
@@
 \\[2ex]
-\AXC{$\Gamma;\ol{\beta^0},\ol{\alpha^0},\phi^k \vdash F$}
-\AXC{$\quad\quad\ol{\Gamma;\Phi,\ol{\beta^0} \vdash G}$}
-\BIC{$\Gamma;\Phi,\ol{\beta^0} \vdash (\mu \phi^k.\lambda
+\AXC{$\Gamma;\ol{\gamma^0},\ol{\alpha^0},\phi^k \vdash F$}
+\AXC{$\quad\quad\ol{\Gamma;\Phi,\ol{\gamma^0} \vdash G}$}
+\BIC{$\Gamma;\Phi,\ol{\gamma^0} \vdash (\mu \phi^k.\lambda
   \ol{\alpha^0}. \,F)\,\ol{G}$} 
@@
 transformation functorial in $\phi$, and the type $\mu \phi. \lambda
-\alpha. \mathit{GRose}\,\phi\alpha$ is not well-formed in our
+\alpha. \mathit{GRose}\,\phi\,\alpha$ is not well-formed in our
 system. These restrictions on higher-arity functoriality are necessary
@@
 \AXC{$\Gamma; \ol{\alpha} \vdash F$}
-\AXC{$\Gamma; \Phi, \ol{\alpha} \vdash G$}
-\AXC{$\Gamma; \Phi, \ol{\alpha} \,|\, \Delta, x : F \vdash t: G$} 
-\TIC{$\Gamma; \Phi
+\AXC{$\Gamma; \ol\gamma, \ol{\alpha} \vdash G$}
+\AXC{$\Gamma; \ol\gamma, \ol{\alpha} \,|\, \Delta, x : F \vdash t: G$} 
+\TIC{$\Gamma; \ol\gamma
   %\emptyset
@@
 \\\\
-\AXC{$\Gamma; \Phi
+\AXC{$\Gamma; \ol\gamma
   %\emptyset
   \,|\, \Delta \vdash t : \Nat^{\ol{\alpha}} \,F \,G$}
-\AXC{$\ol{\Gamma;\Phi \vdash K}$}
-\AXC{$\Gamma;\Phi \,|\, \Delta \vdash s: F[\overline{\alpha := K}]$}
-\TIC{$\Gamma;\Phi \,|\, \Delta \vdash t_{\ol K} s:
+\AXC{$\ol{\Gamma;\Phi,\ol\gamma \vdash K}$}
+\AXC{$\Gamma;\Phi,\ol\gamma \,|\, \Delta \vdash s: F[\overline{\alpha := K}]$}
+\TIC{$\Gamma;\Phi,\ol\gamma \,|\, \Delta \vdash t_{\ol K} s:
   G[\overline{\alpha := K}]$}
@@
 \AXC{$\ol{\Gamma; \ol{\beta},\ol{\gamma} \vdash F}$}
-\AXC{$\ol{\Gamma; \Phi, \ol{\beta},\ol{\gamma} \vdash
-    G}$\hspace*{0.3in}$\ol{\Gamma;\Phi~|~\Delta \vdash t :
+\AXC{$\ol{\Gamma; \ol\alpha, \ol{\beta},\ol{\gamma} \vdash
+    G}$\hspace*{0.3in}$\ol{\Gamma;\ol\alpha~|~\Delta \vdash t :
   \Nat^{\ol{\beta},\ol{\gamma}}\,F\,G}$}
-\TIC{$\Gamma; \Phi
+\TIC{$\Gamma; \ol\alpha
   %\emptyset
@@
 %    if } v \in \tvars^0 \\ 
-  \setsem{\Gamma; \Phi
+  \setsem{\Gamma; \ol\gamma
     %\emptyset
@@
     F}\rho[\ol{\alpha := A}] 
-      \Rightarrow \lambda \ol{A}.\,\setsem{\Gamma; \Phi,
+      \Rightarrow \lambda \ol{A}.\,\setsem{\Gamma; \ol\gamma,
         \ol{\alpha} \vdash G}\rho[\ol{\alpha := A}] \\ 
@@
       : \relsem{\Gamma; \ol{\alpha} \vdash F}\Eq_{\rho}[\ol{\alpha := R}]
-      \rightarrow \relsem{\Gamma;\Phi, \ol{\alpha} \vdash
+      \rightarrow \relsem{\Gamma;\ol\gamma, \ol{\alpha} \vdash
         G}\Eq_{\rho}[\ol{\alpha := R}] \} \\
@@
   \setsem{\Gamma;\Phi \vdash G}\rho\\ 
-  \setsem{\Gamma;\Phi \vdash (\mu \phi.\lambda
+  \setsem{\Gamma;\Phi,\ol\gamma \vdash (\mu \phi.\lambda
     \ol{\alpha}. H)\ol{G}}\rho &= (\mu
-    T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi \vdash G}\rho}\\
+    T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho}\\
     \text{where } T^\set_{H,\rho}\,F & = \lambda
-  \ol{A}. \setsem{\Gamma;\Phi,\phi, \ol{\alpha} \vdash
+  \ol{A}. \setsem{\Gamma;\ol\gamma,\phi, \ol{\alpha} \vdash
     H}\rho[\phi :=  F][\ol{\alpha := A}]\\
   \text{and } T^\set_{H,\rho}\,\eta &= \lambda
-  \ol{A}. \setsem{\Gamma;\Phi,\phi, \ol{\alpha} \vdash
+  \ol{A}. \setsem{\Gamma;\ol\gamma,\phi, \ol{\alpha} \vdash
     H}\id_\rho[\phi := \eta][\ol{\alpha := \id_{A}}]
@@
 \setsem{\Gamma;\ol{\alpha} \vdash F}\rho \Rightarrow
-\setsem{\Gamma;\Phi,\ol{\alpha} \vdash G}\rho\}$ (which contains
-$\setsem{\Gamma;\Phi \vdash \Nat^{\ol{\alpha}}\,F\,G}\rho$) is a
-subset of $\big\{({\setsem{\Gamma;\Phi,\ol{\alpha} \vdash G}\rho[\ol{\alpha
+\setsem{\Gamma;\ol\gamma,\ol{\alpha} \vdash G}\rho\}$ (which contains
+$\setsem{\Gamma;\ol\gamma \vdash \Nat^{\ol{\alpha}}\,F\,G}\rho$) is a
+subset of $\big\{({\setsem{\Gamma;\ol\gamma,\ol{\alpha} \vdash G}\rho[\ol{\alpha
       := S}]})^{(\setsem{\Gamma;\ol{\alpha} \vdash F}\rho[\ol{\alpha
@@
 ${\setsem{\Gamma;\ol{\alpha} \vdash F}\rho[ \ol{\alpha := S}]}$ to
-${\setsem{\Gamma;\Phi,\ol{\alpha} \vdash G}\rho[\ol{\alpha := S}]}$. But
+${\setsem{\Gamma;\ol\gamma,\ol{\alpha} \vdash G}\rho[\ol{\alpha := S}]}$. But
 there are only $\set$-many choices of morphisms between any
 %these (or any)
-two objects since $\set$ is locally small. Finally, interpretations of
+two objects since $\set$ is locally small. Interpretations of
 $\Nat$ types in Definition~\ref{def:set-sem} ensure $\setsem{\Gamma
@@
 $[\set^k, \set]$, and thus admits a fixpoint.  Since
-$T_{H,\rho}^\set$ is defined in terms of $\setsem{\Gamma;\Phi,\phi,
+$T_{H,\rho}^\set$ is defined in terms of $\setsem{\Gamma;\ol\gamma,\phi,
   \ol{\alpha} \vdash H}$, this means that interpretations of types
@@
   G}$ is actually in $[\set^k,\set]$ where $k = |\ol \alpha|$. This
-means that for each $\setsem{\Gamma; \Phi, \phi^k, \ol{\alpha} \vdash
+means that for each $\setsem{\Gamma; \ol\gamma, \phi^k, \ol{\alpha} \vdash
   H}$, the corresponding operator $T^\set_{H}$ can be extended to a
@@
 component at $\ol{A}$ is $(T^\set_{H,f}\, F)_{\ol{A}} =
-\setsem{\Gamma; \Phi,\phi,\ol{\alpha} \vdash H}f[\phi :=
+\setsem{\Gamma; \ol\gamma,\phi,\ol{\alpha} \vdash H}f[\phi :=
   \id_F][\ol{\alpha := \id_A}]$. The next definition uses the functor
@@
 %  $\setsem{\Gamma,v;\emptyset \vdash v}f = \id_{\rho v}$
-\item If \,$\Gamma; \Phi
+\item If \,$\Gamma; \ol\gamma
   %\emptyset
   \vdash \Nat^{\ol{\alpha}}\,F\,G$ then
-  $\setsem{\Gamma; \Phi
+  $\setsem{\Gamma; \ol\gamma
     %\emptyset
@@
 %      \vdash \Nat^{\ol{\alpha}}\,F\,G}\rho}$
-\hspace*{0.2in} $\lambda t : \setsem{\Gamma;\Phi \vdash
+\hspace*{0.2in} $\lambda t : \setsem{\Gamma;\ol\gamma \vdash
     \Nat^{\ol{\alpha}}\,F\,G}\rho.\, 
-  (\lambda \ol{A}. \setsem{\Gamma; \Phi,\ol\alpha \vdash G}f[\ol{\alpha
+  (\lambda \ol{A}. \setsem{\Gamma; \ol\gamma,\ol\alpha \vdash G}f[\ol{\alpha
       := id_A}]) \circ t$  
@@
     G}f$
-\item If \,$\Gamma;\Phi \vdash (\mu \phi.\lambda
+\item If \,$\Gamma;\Phi,\ol\gamma \vdash (\mu \phi.\lambda
   \ol{\alpha}. H)\ol{G}$ then
   \[\begin{array}{lll}
-  \setsem{\Gamma;\Phi \vdash (\mu  \phi.\lambda
-    \ol{\alpha}. H)\ol{G}} f &: &\setsem{\Gamma;\Phi 
+  \setsem{\Gamma;\Phi,\ol\gamma \vdash (\mu  \phi.\lambda
+    \ol{\alpha}. H)\ol{G}} f &: &\setsem{\Gamma;\Phi,\ol\gamma 
     \vdash (\mu \phi.\lambda \ol{\alpha}. H)\ol{G}} \rho \to
-  \setsem{\Gamma;\Phi \vdash (\mu
+  \setsem{\Gamma;\Phi,\ol\gamma \vdash (\mu
     \phi.\lambda\ol{\alpha}. H)\ol{G}} \rho'\\
   &= &(\mu
-  T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi \vdash G}\rho} \to (\mu
-  T^\set_{H,\rho'})\ol{\setsem{\Gamma;\Phi \vdash G}\rho'}
+  T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho} \to (\mu
+  T^\set_{H,\rho'})\ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho'}
   \end{array}\]
@@
   defined by
-  \[\begin{array}{ll} & (\mu T^\set_{H,f})\ol{\setsem{\Gamma;\Phi \vdash
-      G}\rho'} \circ (\mu T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi
-      \vdash G}f}\\ = & (\mu T^\set_{H,\rho'})\ol{\setsem{\Gamma;\Phi
-      \vdash G}f} \circ (\mu T^\set_{H,f})\ol{\setsem{\Gamma;\Phi
+  \[\begin{array}{ll} & (\mu T^\set_{H,f})\ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash
+      G}\rho'} \circ (\mu T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi,\ol\gamma
+      \vdash G}f}\\ = & (\mu T^\set_{H,\rho'})\ol{\setsem{\Gamma;\Phi,\ol\gamma
+      \vdash G}f} \circ (\mu T^\set_{H,f})\ol{\setsem{\Gamma;\Phi,\ol\gamma
       \vdash G}\rho}\end{array}\]  The latter equality holds because $\mu
@@
 \begin{CD}
- (\mu T^\set_{H,\rho}) \ol{\setsem{\Gamma;\Phi \vdash G}\rho} @> (\mu
-  T^\set_{H,f})_{\ol{\setsem{\Gamma;\Phi \vdash G}\rho}} >> (\mu
-  T^\set_{H,\rho'}) \ol{\setsem{\Gamma;\Phi \vdash G}\rho} \\ 
- @V(\mu T^\set_{H,\rho}) \ol{\setsem{\Gamma;\Phi \vdash G}f}VV @V  (\mu
- T^\set_{H,\rho'}) \ol{\setsem{\Gamma;\Phi \vdash G}f} VV \\ 
-(\mu T^\set_{H,\rho}) \ol{\setsem{\Gamma;\Phi \vdash G}\rho'} @>(\mu
- T^\set_{H,f})_{\ol{\setsem{\Gamma;\Phi \vdash G}\rho'}}>> (\mu
- T^\set_{H,\rho'}) \ol{\setsem{\Gamma;\Phi \vdash G}\rho'}
+ (\mu T^\set_{H,\rho}) \ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho} @> (\mu
+  T^\set_{H,f})_{\ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho}} >> (\mu
+  T^\set_{H,\rho'}) \ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho} \\ 
+ @V(\mu T^\set_{H,\rho}) \ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}f}VV @V  (\mu
+ T^\set_{H,\rho'}) \ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}f} VV \\ 
+(\mu T^\set_{H,\rho}) \ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho'} @>(\mu
+ T^\set_{H,f})_{\ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho'}}>> (\mu
+ T^\set_{H,\rho'}) \ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho'}
 \end{CD}
@@
   \begin{align*}
-  \relsem{\Gamma; \Phi \vdash \Nat^{\ol{\alpha}} \,F\,G}\rho &= \{\eta
+  \relsem{\Gamma; \ol\gamma \vdash \Nat^{\ol{\alpha}} \,F\,G}\rho &= \{\eta
   : \lambda \ol{R}.\,\relsem{\Gamma; \ol{\alpha} \vdash
     F}\rho[\ol{\alpha := R}] \Rightarrow \lambda \ol{R}. \,\relsem{
-    \Gamma; \Phi,\ol{\alpha} \vdash G}\rho[\ol{\alpha := R}]\}\\
+    \Gamma; \ol\gamma,\ol{\alpha} \vdash G}\rho[\ol{\alpha := R}]\}\\
   &=
-  \{(t,t') \in \setsem{\Gamma; \Phi
+  \{(t,t') \in \setsem{\Gamma; \ol\gamma
     \vdash \Nat^{\ol{\alpha}}
     \,F\,G} (\pi_1 \rho) \times \setsem{ 
-    \Gamma;\Phi
+    \Gamma;\ol\gamma
     \vdash \Nat^{\ol{\alpha}} \,F\,G} (\pi_2
@@
   & \hspace{0.4in} (t_{\ol{A}},t'_{\ol{B}}) \in
-  (\relsem{\Gamma; \Phi,\ol{\alpha} \vdash G}\rho[\ol{\alpha :=
+  (\relsem{\Gamma; \ol\gamma,\ol{\alpha} \vdash G}\rho[\ol{\alpha :=
       R}])^{\relsem{\Gamma;\ol{\alpha}\vdash F}\rho[\ol{\alpha := R}]} \}\\  
@@
   \relsem{\Gamma;\Phi \vdash G}\rho\\  
-   \relsem{\Gamma;\Phi \vdash (\mu \phi.\lambda
+   \relsem{\Gamma;\Phi,\ol\gamma \vdash (\mu \phi.\lambda
     \ol{\alpha}. H)\ol{G}}\rho
-  &= (\mu T_{H,\rho})\ol{\relsem{\Gamma;\Phi \vdash
+  &= (\mu T_{H,\rho})\ol{\relsem{\Gamma;\Phi,\ol\gamma \vdash
      G}\rho}\\
@@
     &= \lambda \ol{R}. \relsem{
-      \Gamma;\Phi,\phi,\ol{\alpha} \vdash H}\rho[\phi :=
+      \Gamma;\ol\gamma,\phi,\ol{\alpha} \vdash H}\rho[\phi :=
     F][\ol{\alpha := R}]\\
@@
     &= \lambda \ol{R}. \relsem{
-      \Gamma;\Phi,\phi,\ol{\alpha} \vdash H}\id_\rho[\phi :=
+      \Gamma;\ol\gamma,\phi,\ol{\alpha} \vdash H}\id_\rho[\phi :=
     \delta][\ol{\alpha := \id_{\ol{R}}}]
@@
 fixpoint. Since $T_{H,\rho}$ is defined in terms of
-$\relsem{\Gamma;\Phi,\phi^k, \ol{\alpha} \vdash H}$, this means that
+$\relsem{\Gamma;\ol\gamma,\phi^k, \ol{\alpha} \vdash H}$, this means that
 relational interpretations of types must be $\omega$-cocontinuous
@@
 T_{H,\rho'}\, F$ whose component at $\ol{R}$ is $(T_{H,f}\,
-F)_{\ol{R}} = \relsem{\Gamma; \Phi,\phi,\ol{\alpha} \vdash H}f[\phi :=
+F)_{\ol{R}} = \relsem{\Gamma; \ol\gamma,\phi,\ol{\alpha} \vdash H}f[\phi :=
   \id_F][\ol{\alpha := \id_R}]$.  The next definition uses the functor
@@
 %\rangle\\
-\dsem{\Gamma;\Phi \,|\, \Delta \vdash L_{\overline \alpha} x.t : \Nat^{\overline
-    \alpha} \,F \,G}\rho & = &  \curry (\dsem{\Gamma;\Phi,\overline \alpha
+\dsem{\Gamma;\ol\gamma \,|\, \Delta \vdash L_{\overline \alpha} x.t : \Nat^{\overline
+    \alpha} \,F \,G}\rho & = &  \curry (\dsem{\Gamma;\ol\gamma,\overline \alpha
   \,|\, \Delta, x : F \vdash t: G}\rho[\overline{\alpha := \_}])\\
-\dsem{\Gamma;\Phi \,|\, \Delta \vdash t_{\overline K} s:
+\dsem{\Gamma;\Phi,\ol\gamma \,|\, \Delta \vdash t_{\overline K} s:
   G [\overline{\alpha := K}]}\rho & = & \eval \circ \langle
-  \lambda d.\,(\dsem{\Gamma;\Phi \,|\, \Delta \vdash t :
-  \Nat^{\overline{\alpha}} \,F \,G}\rho\; d)_{\overline{\dsem{\Gamma;\Phi
+  \lambda d.\,(\dsem{\Gamma;\ol\gamma \,|\, \Delta \vdash t :
+  \Nat^{\overline{\alpha}} \,F \,G}\rho\; d)_{\overline{\dsem{\Gamma;\Phi,\ol\gamma
       \vdash K}\rho}},\\ 
- & & \hspace*{0.5in} \dsem{\Gamma;\Phi \,|\,
+ & & \hspace*{0.5in} \dsem{\Gamma;\Phi,\ol\gamma \,|\,
     \Delta \vdash s: F [\overline{\alpha := K}]}\rho \rangle\\ 
@@
 \inr \circ \dsem{\Gamma;\Phi \,|\, \Delta \vdash t : G}\rho\\
-\dsem{\Gamma;\Phi \,|\, \Delta \vdash \map^{\ol{F},\ol{G}}_H
+\dsem{\Gamma;\ol\alpha \,|\, \Delta \vdash \map^{\ol{F},\ol{G}}_H
 \,\ol t: \Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\,H[\ol{\phi
@@
       C}]}[\ol{\phi := \lambda
-    \ol{B}. (\dsem{\Gamma;\Phi\,|\,\Delta \vdash t : 
+    \ol{B}. (\dsem{\Gamma;\ol\alpha\,|\,\Delta \vdash t : 
     \Nat^{\ol{\beta},\ol{\gamma}}\,F\,G}\rho\,d)_{\ol{B}\,\ol{C}}}]\\
@@
 \[\begin{array}{l}
-\hspace*{0.135in}\setsem{\Gamma; \emptyset \,|\, x :
+\hspace*{0.135in}\setsem{\Gamma; {\color{blue} \ol\alpha} \,|\, x :
   \Nat^{\overline{\alpha}, 
@@
   G[\overline{\alpha := H}]}\\
-= \setsem{ \Gamma; \emptyset \,|\, x
+= \setsem{ \Gamma; {\color{blue} \ol\alpha} \,|\, x
   : \Nat^{\overline{\alpha}, \overline{\gamma}} F\, G, \overline{y :
@@
 \begin{array}{l}
-\hspace*{0.135in}\setsem{\Gamma; \emptyset \,|\, \overline{y :
+\hspace*{0.135in}\setsem{\Gamma; {\color{blue} \ol\alpha} \,|\, \overline{y :
     \Nat^{\overline{\gamma}} 
@@
     \overline{\gamma}} z) : \xi}\\
-= \setsem{\Gamma; \emptyset \,|\, \overline{y :
+= \setsem{\Gamma; {\color{blue} \ol\alpha} \,|\, \overline{y :
   \Nat^{\overline{\gamma}} F\, G} 
@@
 %(\Nat^\emptyset\,(\mathit{Tree}\,\sigma\,\sigma')\,(\mathit{Tree}\,\tau\,\tau'))$$
-Finally, in Section~\ref{sec:calculus} we observed that $\emptyset;
-\phi, \alpha \vdash \mathit{GRose}\,\phi\,\alpha$ cannot appear as
-the (co)domain of a $\Nat$ type binding both $\phi$ and $\alpha$. As a
-result, there are no instances of, say, a
-$\mathtt{map}$/$\mathtt{reverse}$ free theorem for this data type that
-map replacements for both $\phi$ and $\alpha$. On the other hand,
-$\phi; \alpha \vdash \mathit{GRose}\,\phi\,\alpha$ does support a
-$\mathtt{map}$/$\mathtt{reverse}$ theorem similar to those from
-Section~\ref{sec:intro}.
+Finally, in Section~\ref{sec:calculus} we observed that $ \phi; \alpha
+\vdash \mathit{GRose}\,\phi\,\alpha$ cannot appear as the (co)domain
+of a $\Nat$ type binding both $\phi$ and $\alpha$. As a result, there
+are no instances of, say, a $\mathtt{map}$/$\mathtt{reverse}$ free
+theorem for this data type that maps replacements for both $\phi$ and
+$\alpha$. On the other hand, $\phi; \alpha \vdash
+\mathit{GRose}\,\phi\,\alpha$ does support a
+$\mathtt{map}$/$\mathtt{reverse}$ theorem that maps only a replacement
+for $\alpha$.
 
@@
 judgment are natural transformations, and that all set interpretations
-of terms of $\Nat$-types satisfy the appropriate {\color{blue}
-  equality preservation} conditions from Definition~\ref{def:set-sem}.
-For the interesting cases of abstraction, application, $\map$, $\tin$,
-and $\fold$ terms, propagating the naturality conditions is quite
-involved; the latter two cases especially require some rather
-involved diagram chasing. That this is possible provides strong
-evidence that our definitions and development are sensible,
-natural, and at an appropriate level of abstraction. Detailed proofs
-for the set interpretations for the five interesting forms of terms
-are included in the accompanying anonymous supplementary
-material. Even in the supplementary material we omit the other,
-straightforward, cases.  The proofs for the relational interpretations
-of all term judgments are entirely analogous to the proofs for their
-set interpretations. The proof that Equation~\ref{eq:projs} holds in
-each case is by direct calculation, using the facts that projections
-are surjective and that the set and relational interpretations are
-defined ``in parallel'', i.e., are fibred.
+of terms of $\Nat$-types satisfy the appropriate equality preservation
+conditions from Definition~\ref{def:set-sem}.  For the interesting
+cases of abstraction, application, $\map$, $\tin$, and $\fold$ terms,
+propagating the naturality conditions is quite involved; the latter
+two cases especially require some rather involved diagram
+chasing. That this is possible provides strong evidence that our
+definitions and development are sensible, natural, and at an
+appropriate level of abstraction. Detailed proofs for the set
+interpretations for the five interesting forms of terms are included
+in the accompanying anonymous supplementary material. Even in the
+supplementary material we omit the other, straightforward, cases.  The
+proofs for the relational interpretations of all term judgments are
+entirely analogous to the proofs for their set interpretations. The
+proof that Equation~\ref{eq:projs} holds in each case is by direct
+calculation, using the facts that projections are surjective and that
+the set and relational interpretations are defined ``in parallel'',
+i.e., are fibred.
 \end{proof}

### 2020-06-23 free-theorems-nesteds/paper3.tex
changes to paper3

@@
 \end{dfn}
-The interpretations in Definition~\ref{def:set-sem} respect weakening,
-i.e., a type and its weakenings have the same set interpretations.
-The same holds for the actions of these interpretations on morphisms
-in Definition~\ref{def:set-sem-funcs} below.
+%The interpretations in Definition~\ref{def:set-sem} respect weakening,
+%i.e., a type and its weakenings have the same set interpretations.
+%
 %= \setsem{\Gamma \vdash \sigma} \rho \to
 %\setsem{\Gamma \vdash \tau} \rho$, as expected {\color{blue} by IEL}.
-If $\rho$ is a set environment and $\vdash F$ then we may
-write $\setsem{\vdash F}$ instead of $\setsem{\vdash F}\rho$
-since the environment is immaterial.
-We note that the third clause of
-Definition~\ref{def:set-sem} does indeed define a set: local finite
-presentability of $\set$ and $\omega$-cocontinuity of
-$\setsem{\Gamma;\ol{\alpha} \vdash F}\rho$ ensure that $\{\eta :
-\setsem{\Gamma;\ol{\alpha} \vdash F}\rho \Rightarrow
-\setsem{\Gamma;\ol\gamma,\ol{\alpha} \vdash G}\rho\}$ (which contains
-$\setsem{\Gamma;\ol\gamma \vdash \Nat^{\ol{\alpha}}\,F\,G}\rho$) is a
-subset of $\big\{({\setsem{\Gamma;\ol\gamma,\ol{\alpha} \vdash G}\rho[\ol{\alpha
+If $\rho \in \setenv$ and $\vdash F$ then we write $\setsem{\vdash F}$
+instead of $\setsem{\vdash F}\rho$ since the environment is
+immaterial. The third clause of Definition~\ref{def:set-sem} does
+indeed define a set: local finite presentability of $\set$ and
+$\omega$-cocontinuity of $\setsem{\Gamma;\ol{\alpha} \vdash F}\rho$
+ensure that $\{\eta : \setsem{\Gamma;\ol{\alpha} \vdash F}\rho
+\Rightarrow \setsem{\Gamma;\ol\gamma,\ol{\alpha} \vdash G}\rho\}$
+(which contains $\setsem{\Gamma;\ol\gamma \vdash
+  \Nat^{\ol{\alpha}}\,F\,G}\rho$) is a subset of
+$\big\{({\setsem{\Gamma;\ol\gamma,\ol{\alpha} \vdash G}\rho[\ol{\alpha
       := S}]})^{(\setsem{\Gamma;\ol{\alpha} \vdash F}\rho[\ol{\alpha
@@
 }$ $S_i \mbox{ is a finite set for } i =
-1,...,|\ol{\alpha}|\big\}$. There are countably many choices for
-tuples $\ol{S}$, and each of these gives rise to a morphism from
+1,...,|\ol{\alpha}|\big\}$. There are countably many choices of tuples
+$\ol{S}$, each of which gives rise to a morphism from
 ${\setsem{\Gamma;\ol{\alpha} \vdash F}\rho[ \ol{\alpha := S}]}$ to
-${\setsem{\Gamma;\ol\gamma,\ol{\alpha} \vdash G}\rho[\ol{\alpha := S}]}$. But
-there are only $\set$-many choices of morphisms between any
-%these (or any)
-two objects since $\set$ is locally small. Interpretations of
-$\Nat$ types in Definition~\ref{def:set-sem} ensure $\setsem{\Gamma
-  \vdash F \to G}$ and $\setsem{\Gamma \vdash \forall
-  \alpha. F}$ are as expected in any parametric model.
-
-In order to make sense of the last clause in
+${\setsem{\Gamma;\ol\gamma,\ol{\alpha} \vdash G}\rho[\ol{\alpha :=
+      S}]}$. But there are only $\set$-many choices of morphisms
+between any two objects since $\set$ is locally small.  That
+$\setsem{\Gamma; \ol\gamma \vdash \Nat^\alpha F\,G}$ is
+$\omega$-cocontinuous follows from $\omega$-cocontinuity of
+$\setsem{\Gamma;\ol\alpha \vdash F}$ and
+$\setsem{\Gamma;\ol\gamma,\ol\alpha \vdash G}$. Interpretations of
+$\Nat$-types ensure that $\setsem{\Gamma \vdash F \to G}$ and
+$\setsem{\Gamma \vdash \forall \alpha. F}$ are as expected in any
+parametric model.
+
+To make sense of the last clause in
 Definition~\ref{def:set-sem}, we need to know that, for each $\rho \in
@@
 morphisms between set environments.
+
+\begin{comment}
+To make sense of the last clause in
+Definition~\ref{def:set-sem}, we need to know that, for each $\rho \in
+\setenv$, $T^\set_{H,\rho}$ is an $\omega$-cocontinuous endofunctor on
+$[\set^k, \set]$, and thus admits a fixpoint.  Since
+$T_{H,\rho}^\set$ is defined in terms of $\setsem{\Gamma;\ol\gamma,\phi,
+  \ol{\alpha} \vdash H}$, this means that interpretations of types
+must be such functors, which in turn means that the actions of set
+interpretations of types on objects and on morphisms in $\setenv$ are
+intertwined. Fortunately, we know from~\cite{jp19} that, for every
+$\Gamma; \ol{\alpha} \vdash G$, $\setsem{\Gamma; \ol{\alpha} \vdash
+  G}$ is actually in $[\set^k,\set]$ where $k = |\ol \alpha|$. This
+means that for each $\setsem{\Gamma; \ol\gamma, \phi^k, \ol{\alpha} \vdash
+  H}$, the corresponding operator $T^\set_{H}$ can be extended to a
+{\em functor} from $\setenv$ to $[[\set^k,\set],[\set^k,\set]]$. The
+action of $T^\set_H$ on an object $\rho \in \setenv$ is given by the
+higher-order functor $T_{H,\rho}^\set$, whose actions on objects
+(functors in $[\set^k, \set]$) and morphisms (natural transformations
+between such functors) are given in Definition~\ref{def:set-sem}. Its
+action on a morphism $f : \rho \to \rho'$ is the higher-order natural
+transformation $T^\set_{H,f} : T^\set_{H,\rho} \to T^\set_{H,\rho'}$
+whose action on $F : [\set^k,\set]$ is the natural transformation
+$T^\set_{H,f}\, F : T^\set_{H,\rho}\,F \to T^\set_{H,\rho'}\,F$ whose
+component at $\ol{A}$ is $(T^\set_{H,f}\, F)_{\ol{A}} =
+\setsem{\Gamma; \ol\gamma,\phi,\ol{\alpha} \vdash H}f[\phi :=
+  \id_F][\ol{\alpha := \id_A}]$. The next definition uses the functor
+$T^\set_H$ to define the actions of functors interpreting types on
+morphisms between set environments.
+\end{comment}
 
@@
 \end{dfn}
+
+Definitions~\ref{def:set-sem} and~\ref{def:set-sem-funcs} respect
+weakening, i.e., ensure that a type and its weakenings have the same
+set interpretations.
 
@@
 
-The interpretations in Definition~\ref{def:rel-sem}, as well as in
-Definition~\ref{def:rel-sem-funcs} below, respect weakening, and also
-ensure that $\relsem{\Gamma \vdash F \to G}$ and $\relsem{\Gamma
-  \vdash \forall \alpha. F}$ are as expected in any parametric model.
+The interpretations in Definitions~\ref{def:rel-sem}
+and~\ref{def:rel-sem-funcs} below respect weakening, and also ensure
+that $\relsem{\Gamma \vdash F \to G}$ and $\relsem{\Gamma \vdash
+  \forall \alpha. F}$ are as expected in any parametric model.
 %$\relsem{\Gamma \vdash
@@
 %\relsem{\Gamma \vdash \tau} \rho$.
-If $\rho$ is a relational environment and $\vdash F$, then we write
-$\relsem{\vdash F}$ instead of $\relsem{\vdash F}\rho$ as for set
-interpretations.  For the last clause in Definition~\ref{def:rel-sem}
-to be well-defined we need $T_{H,\rho}$ to be an $\omega$-cocontinuous
-endofunctor on $RT$ so that, by Lemma~\ref{lem:fp}, it admits a
-fixpoint. Since $T_{H,\rho}$ is defined in terms of
-$\relsem{\Gamma;\ol\gamma,\phi^k, \ol{\alpha} \vdash H}$, this means that
-relational interpretations of types must be $\omega$-cocontinuous
-functors from $\relenv$ to $RT_0$, which in turn entails that the
-actions of relational interpretations of types on objects and on
-morphisms in $\relenv$ are intertwined. As for set interpretations, we
-know from~\cite{jp19} that, for every $\Gamma; \ol{\alpha} \vdash F$,
-$\setsem{\Gamma; \ol{\alpha} \vdash F}$ is actually in
-$[\rel^k,\rel]$ where $k = |\ol \alpha|$.
+That $\relsem{\Gamma; \ol\gamma \vdash \Nat^\alpha F\,G}$ is
+$\omega$-cocontinuous follows from $\omega$-cocontinuity of
+$\relsem{\Gamma;\ol\alpha \vdash F}$ and
+$\relsem{\Gamma;\ol\gamma,\ol\alpha \vdash G}$.  If $\rho \in \relenv$
+and $\vdash F$, then we write $\relsem{\vdash F}$ instead of
+$\relsem{\vdash F}\rho$.  For the last clause in
+Definition~\ref{def:rel-sem} to be well-defined we need $T_{H,\rho}$
+to be an $\omega$-cocontinuous endofunctor on $RT$ so that, by
+Lemma~\ref{lem:fp}, it admits a fixpoint. Since $T_{H,\rho}$ is
+defined in terms of $\relsem{\Gamma;\ol\gamma,\phi^k, \ol{\alpha}
+  \vdash H}$, this means that relational interpretations of types must
+be $\omega$-cocontinuous functors from $\relenv$ to $RT_0$, which in
+turn entails that the actions of relational interpretations of types
+on objects and on morphisms in $\relenv$ are intertwined. As for set
+interpretations, we know from~\cite{jp19} that, for every $\Gamma;
+\ol{\alpha} \vdash F$, $\relsem{\Gamma; \ol{\alpha} \vdash F}$ is
+actually in $[\rel^k,\rel]$ where $k = |\ol \alpha|$.
 %In fact, we already know from~\cite{jp19} that, for every $\Gamma;
@@
 T_{H,\rho'}\, F$ whose component at $\ol{R}$ is $(T_{H,f}\,
-F)_{\ol{R}} = \relsem{\Gamma; \ol\gamma,\phi,\ol{\alpha} \vdash H}f[\phi :=
-  \id_F][\ol{\alpha := \id_R}]$.  The next definition uses the functor
+F)_{\ol{R}} = \relsem{\Gamma; \ol\gamma,\phi,\ol{\alpha} \vdash
+  H}f[\phi := \id_F][\ol{\alpha := \id_R}]$. The next definition uses
 $T_H$ to define the actions of functors interpreting types on
@@
 
-Define, for $\Gamma; \ol{\alpha} \vdash F$, the term {\color{blue}
-  $\id_F$} to be $\Gamma;\emptyset~|~\emptyset \vdash
-L_{\ol{\alpha}}x.x : \Nat^{\ol{\alpha}} F\,F$ and, for terms $\Gamma;
-\emptyset \,|\, \Delta \vdash t: \Nat^{\overline{\alpha}} F\,G$ and
-$\Gamma; \emptyset \,|\, \Delta \vdash s: \Nat^{\overline{\alpha}}
-G\,H$, the {\em composition} $s \circ t$ of $t$ and $s$ to be $\Gamma;
-\emptyset \,|\, \Delta \vdash L_{\overline{\alpha}}
-x. s_{\overline{\alpha}}(t_{\overline{\alpha}}x):
-\Nat^{\overline{\alpha}} F\,H$. Then $\setsem{\Gamma; \emptyset \,|\,
-  \emptyset \vdash \id_{F} : \Nat^{\ol{\alpha}} F\,F} \rho\, \ast =
-\id_{\lambda \ol{A}. \setsem{\Gamma; \ol{\alpha} \vdash F} \rho
-  [\ol{\alpha := A}]}$ for any set environment $\rho$, and
-$\setsem{\Gamma; \emptyset \,|\, \Delta \vdash s \circ t:
-  \Nat^{\overline{\alpha}} F\,H} = \setsem{\Gamma; \emptyset \,|\,
-  \Delta \vdash s: \Nat^{\overline{\alpha}} G\,H} \circ
-\setsem{\Gamma; \emptyset \,|\, \Delta \vdash t:
-  \Nat^{\overline{\alpha}} F\,G}$.  Straightforward calculation using
-these equalities shows that terms of $\Nat$ type behave as
+Straightforward calculation shows that terms of $\Nat$ type behave as
 natural transformations with respect to their source and target
@@
 \begin{thm}\label{eq:ft-from-nat} 
+If\, $\Gamma; \ol\beta \,|\, \Delta \vdash s : \Nat^{\overline{\alpha},
+  \overline{\gamma}} F\,G$ and $\overline{\Gamma; \ol\beta \,|\,
+  \Delta \vdash t : \Nat^{\overline{\gamma}} K\, H}$, then
   \[\begin{array}{l}
-\hspace*{0.135in}\setsem{\Gamma; {\color{blue} \ol\alpha} \,|\, x :
-  \Nat^{\overline{\alpha}, 
-    \overline{\gamma}} F\,G, \overline{y : \Nat^{\overline{\gamma}}
-    K\, H} \vdash (\map_G^{\overline{K},
-    \overline{H}} \,\overline{y}) \circ
-  (L_{\overline{\gamma}} z. x_{\overline{K}, \overline{\gamma}}
-  z) : \Nat^{\overline{\gamma}} F[\overline{\alpha := K}]\,
+  \hspace*{0.135in}
+\setsem{\Gamma; \ol\beta \,|\, \Delta
+  \vdash L_{\overline{\gamma}} z. 
+  (\map_G^{\overline{K}, \overline{H}} \,\overline{t})_{\ol\gamma}
+  (s_{\overline{K}, \overline{\gamma}} z)
+  : \Nat^{\overline{\gamma}} F[\overline{\alpha := K}]\,
   G[\overline{\alpha := H}]}\\
-= \setsem{ \Gamma; {\color{blue} \ol\alpha} \,|\, x
-  : \Nat^{\overline{\alpha}, \overline{\gamma}} F\, G, \overline{y :
-    \Nat^{\overline{\gamma}} K\, H} \vdash
-  (L_{\overline{\gamma}} z. x_{\overline{H}, \overline{\gamma}} z)
-  \circ (\map_F^{\overline{K}, \overline{\tau}} \,
-  \overline{y}) : \Nat^{\overline{\gamma}} F[\overline{\alpha :=
+= \setsem{ \Gamma; \ol\beta \,|\, \Delta \vdash
+  L_{\overline{\gamma}} z. s_{\overline{H}, \overline{\gamma}}
+  ((\map_F^{\overline{K}, \overline{H}} \,
+  \overline{t})_{\ol\gamma} z) : \Nat^{\overline{\gamma}} F[\overline{\alpha :=
       K}]\, G[\overline{\alpha := H}]}
@@
 \end{thm}
-When $x = \mathit{in}_H$ and $\xi = \Nat^{\overline{\gamma}} H[\phi :=
+\noindent
+When $s = \mathsf{in}_H$, and
+$\overline{\Gamma; \ol\delta \,|\, t :
+ \Nat^{\overline{\gamma}} F\, G}$,
+and $\xi = \Nat^{\overline{\gamma}} H[\phi :=
   (\mu \phi. \lambda \overline{\alpha}. H)
   \overline{\beta}][\overline{\alpha := F}]\, (\mu \phi. \lambda
-\overline{\alpha}. H)\overline{G}$ Theorem~\ref{eq:ft-from-nat}
+\overline{\alpha}. H)\overline{G}$, Theorem~\ref{eq:ft-from-nat}
 specializes to
@@
 \begin{array}{l}
-\hspace*{0.135in}\setsem{\Gamma; {\color{blue} \ol\alpha} \,|\, \overline{y :
-    \Nat^{\overline{\gamma}} 
-    F\, G} \vdash (\map_{(\mu \phi. \lambda
+\hspace*{0.135in}\setsem{\Gamma; \ol\delta \,|\, \Delta \vdash
+L_{\overline{\gamma}} z. (\map_{(\mu \phi. \lambda
     \overline{\alpha}. H) \overline{\beta}}^{\overline{F},
-    \overline{G}}\, \overline{y}) \circ
-  (L_{\overline{\gamma}} z. (\tin_{H})_{\overline{F},
+    \overline{G}}\, \overline{t})_{\ol\gamma}
+  (\tin_{H})_{\overline{F},
     \overline{\gamma}} z) : \xi}\\
-= \setsem{\Gamma; {\color{blue} \ol\alpha} \,|\, \overline{y :
-  \Nat^{\overline{\gamma}} F\, G} 
-  \vdash (L_{\overline{\gamma}} z. (\tin_H)_{\overline{G}, \overline{\gamma}} z)
-  \circ (\map_{H[\phi := (\mu \phi. \lambda \overline{\alpha}. H)
+= \setsem{\Gamma; \ol\delta \,|\, \Delta
+\vdash L_{\overline{\gamma}} z. (\tin_H)_{\overline{G}, \overline{\gamma}}
+  ((\map_{H[\phi := (\mu \phi. \lambda \overline{\alpha}. H)
       \overline{\beta}]}^{\overline{F},
-    \overline{G}}\, \overline{y}) 
+    \overline{G}}\, \overline{t})_{\ol\gamma} z)
   : \xi}
@@
 \begin{align*}
-  &\setsem{\emptyset; \alpha~|~ y : \Nat^\emptyset F \, G
-    \vdash (\map_{\mathit{Bush}\,\alpha}^{{F},
-      {G}}\, y) \circ (L_{{\emptyset}}
-    z. \mathit{reverseBush}_{F}\, z) : \Nat^{{\emptyset}}
+  &\setsem{\emptyset; \emptyset~|~ y : \Nat^\emptyset F \, G
+    \vdash L_{{\emptyset}}
+    z. (\map_{\mathit{Bush}\,\alpha}^{{F},
+      {G}}\, y)_{\emptyset} (\mathit{reverseBush}_{F}\, z) : \Nat^{{\emptyset}}
     (\mathit{Bush} \, F) \, (\mathit{Bush} \, G) }\\ =\;
-  &\setsem{\emptyset; \alpha~|~ y : \Nat^\emptyset F\, G
-    \vdash (L_{\emptyset} z. \mathit{reverseBush}_{G} \,z) \circ
-    (\map_{\mathit{Bush} \, \alpha}^{F, G}\,
-           {y}) : \Nat^{\emptyset} (\mathit{Bush} \, F ) \,
+  &\setsem{\emptyset; \emptyset~|~ y : \Nat^\emptyset F\, G
+    \vdash L_{\emptyset} z. \mathit{reverseBush}_{G}
+    ((\map_{\mathit{Bush} \, \alpha}^{F, G}\,
+           {y})_{\emptyset} z) : \Nat^{\emptyset} (\mathit{Bush} \, F ) \,
            (\mathit{Bush} \, G)}
@@
 \begin{align*}
-  &\setsem{\emptyset; \alpha,\gamma~|~y : \Nat^\emptyset F \,
-    G \vdash (\map_{\mathit{Tree}\,\alpha \, \gamma}^{F,
-      G}\, y) \circ (L_{{\gamma}}
-    z. \mathit{reverseTree}_{F, \gamma}\, z) : \Nat^{{\gamma}}
+  &\setsem{\emptyset; \emptyset~|~y : \Nat^\gamma F \, G 
+  \vdash L_{{\gamma}} z.
+    (\map_{\mathit{Tree}\,\alpha \, \gamma}^{F,
+      G}\, y)_{\gamma} (\mathit{reverseTree}_{F, \gamma}\, z) : \Nat^{{\gamma}}
     (\mathit{Tree} \, F \, \gamma) \, (\mathit{Tree} \, G \,
-    \gamma) }\\ =\; &\setsem{\emptyset; \alpha,\gamma~|~y :
-    \Nat^\emptyset F \, G \vdash (L_{\gamma}
-    z. \mathit{reverseTree}_{G, \gamma} \,z) \circ
-    (\map_{\mathit{Tree} \, \alpha\, \gamma}^{{F},
-      {G}}\, y) : \Nat^{\gamma} (\mathit{Tree} \,
+    \gamma) }\\ =\; &\setsem{\emptyset; \emptyset~|~y :
+    \Nat^\gamma F \, G \vdash L_{\gamma}
+    z. \mathit{reverseTree}_{G, \gamma}\,
+    ((\map_{\mathit{Tree} \, \alpha\, \gamma}^{{F},
+      {G}}\, y)_{\gamma} z) : \Nat^{\gamma} (\mathit{Tree} \,
     F \, \gamma) \, (\mathit{Tree} \, G\, \gamma)}
@@
 In particular, if $\Gamma; \ol{\alpha} \vdash H$, $\ol{\Gamma;
-  \emptyset \vdash F}$, and $\ol{\Gamma; \emptyset \vdash G}$, then
+  \emptyset \vdash F}$, and $\ol{\Gamma; \ol\beta \vdash G}$, then
 Definition~\ref{def:set-interp} gives, for all $\ol{f :
-  \setsem{\Gamma;\emptyset \vdash \Nat^\emptyset F\,G}\rho}$, that
+  \setsem{\Gamma;\ol\beta \vdash \Nat^\emptyset F\,G}\rho}$, that
 \[\begin{array}{rl}
-&\setsem{ \Gamma;\emptyset\,|\,\ol{x : \Nat^\emptyset F \, G}
+&\setsem{ \Gamma;\ol\beta\,|\,\ol{x : \Nat^\emptyset F \, G}
   \vdash \map^{\ol{F},\ol{G}}_H\,\ol x :
@@
 \end{comment}
+
+\vspace*{0.05in}
+
+\noindent
 Here, we obtain the first equality from the appropriate instance of
@@
 
-We also have the expected relationships between the interpetations of
-$\map$, $\tin$, and $\fold$, namely
+Define, for $\Gamma; \ol{\alpha} \vdash F$, the term $\id_F$ to be
+$\Gamma;\emptyset~|~\emptyset \vdash L_{\ol{\alpha}}x.x :
+\Nat^{\ol{\alpha}} F\,F$ and, for terms $\Gamma; \emptyset \,|\,
+\Delta \vdash t: \Nat^{\overline{\alpha}} F\,G$ and $\Gamma; \ol\beta
+\,|\, \Delta \vdash s: \Nat^{\overline{\alpha}} G\,H$, the {\em
+  composition} $s \circ t$ of $t$ and $s$ to be $\Gamma; \ol\beta
+\,|\, \Delta \vdash L_{\overline{\alpha}}
+x. s_{\overline{\alpha}}(t_{\overline{\alpha}}x):
+\Nat^{\overline{\alpha}} F\,H$. Then $\setsem{\Gamma; \emptyset \,|\,
+  \emptyset \vdash \id_{F} : \Nat^{\ol{\alpha}} F\,F} \rho\, \ast =
+\id_{\lambda \ol{A}. \setsem{\Gamma; \ol{\alpha} \vdash F} \rho
+  [\ol{\alpha := A}]}$ for any set environment $\rho$, and
+$\setsem{\Gamma; \ol\beta \,|\, \Delta \vdash s \circ t:
+  \Nat^{\overline{\alpha}} F\,H} = \setsem{\Gamma; \ol\beta \,|\,
+  \Delta \vdash s: \Nat^{\overline{\alpha}} G\,H} \circ
+\setsem{\Gamma; \emptyset \,|\, \Delta \vdash t:
+  \Nat^{\overline{\alpha}} F\,G}$.  We have the expected relationships
+between the interpetations of $\map$, $\tin$, and $\fold$:
 
@@
 $\ol{\Gamma; \ol{\beta}, \ol{\gamma} \vdash F}$, 
-$\ol{\Gamma; \Phi, \ol{\beta}, \ol{\gamma} \vdash G}$, and
-$\ol{\Gamma;\Phi\,|\,\Delta \vdash t : \Nat^{\ol{\alpha}, \ol{\beta},
+$\ol{\Gamma; \ol\delta, \ol{\beta}, \ol{\gamma} \vdash G}$, and
+$\ol{\Gamma;\ol\delta\,|\,\Delta \vdash t : \Nat^{\ol{\alpha}, \ol{\beta},
     \ol{\gamma}} F\, G}$
 then
-\[\setsem{\Gamma; \Phi \,|\, \Delta \vdash
+\[\setsem{\Gamma; \ol\delta \,|\, \Delta \vdash
 \map_{H[\ol{\psi := K}]}^{\ol{F}, \ol{G}}\,\ol t : \xi} = \setsem{\Gamma;
-  \Phi \,|\, \Delta \vdash \map_H^{\ol{K[\ol{\phi := F}]},
+  \ol\delta \,|\, \Delta \vdash \map_H^{\ol{K[\ol{\phi := F}]},
     \ol{K[\ol{\phi := G}]}}\, (\ol{\map_K^{\ol{F}, \ol{G}}}\, \ol t) :
   \xi}\]
-for $\xi = \Nat^{\ol{\gamma}} H[\ol{\psi :=
-    K}][\ol{\phi := F}] H[\ol{\psi := K}][\ol{\phi := G}]$
+for $\xi = \Nat^{\ol{\gamma}} \,H[\ol{\psi :=
+    K}][\ol{\phi := F}]\, H[\ol{\psi := K}][\ol{\phi := G}]$
 
@@
 $\overline{\Gamma; \overline{\alpha}, \overline{\gamma} \vdash F}$,\;
-$\overline{\Gamma; \overline{\alpha}, \overline{\gamma} \vdash G}$,\;
+$\overline{\Gamma; \ol\delta, \overline{\alpha}, \overline{\gamma}
+  \vdash G}$,\; 
 $\overline{\Gamma; \phi, \overline{\psi}, \overline{\gamma} \vdash
@@
 $\overline{J}$ is the sequence $\overline{G}, K$, and $\Gamma;
-\Phi\,|\,\Delta \vdash s : \Nat^{\overline{\beta},
-  \overline{\gamma}} H\, K$ and $\ol{\Gamma; \Phi\,|\,\Delta \vdash
+\emptyset\,|\,\Delta \vdash s : \Nat^{\overline{\beta},
+  \overline{\gamma}} H\, K$ and $\ol{\Gamma; \ol\delta\,|\,\Delta \vdash
   t : \Nat^{\overline{\alpha}, \overline{\gamma}} F\, G}$, then
-\[\hspace*{0.13in}\setsem{\Gamma; \Phi \,|\, \Delta
+\[\hspace*{0.13in}\setsem{\Gamma; \ol\delta \,|\, \Delta
     \vdash L_{\overline{\gamma}} z.\,
@@
       (s, \overline{t})}) \big)_{\overline{\gamma}} z \Big) : \xi}\]
-\[= \setsem{\Gamma; \Phi \,|\, \Delta \vdash
+\[= \setsem{\Gamma; \ol\delta\,|\, \Delta \vdash
     L_{\overline{\gamma}} z.\, \big( \map_{K}^{\overline{M
@@
 
-\vspace*{-0.3in}
+\vspace*{-0.15in}
 
@@
 
-\vspace*{0.18in}
+\vspace*{0.1in}
 
@@
 
-\vspace*{-0.2in}
-
-\[\hspace*{-3.6in} = \setsem{\Gamma; \emptyset \,|\, \emptyset \vdash \Id_{(\mu
+\vspace*{-0.15in}
+
+\[\hspace*{-3.6in} = \setsem{\Gamma; \emptyset \,|\, \emptyset \vdash \mathit{id}_{(\mu
     \phi. \lambda \overline{\alpha}. H)\overline{\beta}} : \xi}\]
 
-\vspace*{0.18in}
+\vspace*{0.15in}
 
@@
 
-\vspace*{-0.2in}
+\vspace*{-0.15in}
 
 \[\hspace*{-3.32in} =\setsem{ \Gamma; \emptyset \,|\, \emptyset \vdash
-  \Id_{H[\phi := (\mu \phi. \lambda
+  \mathit{id}_{H[\phi := (\mu \phi. \lambda
       \overline{\alpha}. H)\overline{\beta}]} : \xi}\]
@@
 only use one level (or maybe two, like short cut).
+
+%test

### 2020-06-24 free-theorems-nesteds/paper3.tex
added explanation for coc of Nat types

@@
 $\omega$-cocontinuous follows from $\omega$-cocontinuity of
-$\setsem{\Gamma;\ol\alpha \vdash F}$ and
-$\setsem{\Gamma;\ol\gamma,\ol\alpha \vdash G}$. Interpretations of
-$\Nat$-types ensure that $\setsem{\Gamma \vdash F \to G}$ and
-$\setsem{\Gamma \vdash \forall \alpha. F}$ are as expected in any
-parametric model.
+$\setsem{\Gamma;\ol\alpha \vdash F}$,
+$\setsem{\Gamma;\ol\gamma,\ol\alpha \vdash G}$, and the equality
+relation environment $\Eq$. Interpretations of $\Nat$-types ensure
+that $\setsem{\Gamma \vdash F \to G}$ and $\setsem{\Gamma \vdash
+  \forall \alpha. F}$ are as expected in any parametric model.
 

### 2020-06-24 free-theorems-nesteds/paper4.tex
added paper4

--- 2020-06-24 free-theorems-nesteds/paper3.tex
@@
 \AXC{$\Gamma;\ol{\alpha^0} \vdash F$}
-\AXC{$\Gamma;\ol{\gamma^0},\ol{\alpha^0}  \vdash G$}
-\BIC{$\Gamma;\ol{\gamma^0} \vdash \Nat^{\ol{\alpha^0}}F \,G$}
+\AXC{$\Gamma;\ol{\alpha^0}  \vdash G$}
+\BIC{$\Gamma;\emptyset \vdash \Nat^{\ol{\alpha^0}}F \,G$}
 \DisplayProof
@@
 $\mathit{Tree}\,\alpha\,\gamma = \mu \,\beta.\, \alpha + \beta \times
-\gamma \times \beta$, then both $\gamma;\emptyset \vdash \Nat^\alpha
-(\mathit{List}\, \alpha) \, (\mathit{Tree} \, \alpha\, \gamma)$ and
-$\emptyset;\gamma \vdash \Nat^\alpha (\mathit{List}\, \alpha) \,
-(\mathit{Tree} \, \alpha\, \gamma)$ are well-formed.  Although types
+\gamma \times \beta$, then $\gamma;\emptyset \vdash \Nat^\alpha
+(\mathit{List}\, \alpha) \, (\mathit{Tree} \, \alpha\, \gamma)$ is
+well-formed, but $\emptyset;\gamma \vdash \Nat^\alpha (\mathit{List}\,
+\alpha) \, (\mathit{Tree} \, \alpha\, \gamma)$ is not.  Although types
 can be functorial in variables of arity greater than $0$, the body $F$
@@
 \phi \beta$ can be functorial or not in $\alpha$, but cannot be
-functorial in $\phi$.  As a result, $\mathit{GRose}\,\phi\,\alpha$
+functorial in $\phi$. In particular, $\mathit{GRose}\,\phi\,\alpha$
 cannot be the (co)domain of a $\Nat$ type representing a natural
@@
 \alpha. \mathit{GRose}\,\phi\,\alpha$ is not well-formed in our
-system. These restrictions on higher-arity functoriality are necessary
-for constructing our parametric model.  Fortunately, they are entirely
-natural, and {\color{blue} preclude only truly bizarre nested types}.
+system. The restrictions imposed by our typing rules will ensure that
+set interpretations of types are functors. In
+Section~\ref{sec:conclusion} we discuss how to lessen the expressivity
+constraints they impose.
+%These restrictions on higher-arity functoriality are necessary
+%for constructing our parametric model. Fortunately, they are entirely
+%natural, and {\color{blue} preclude only truly bizarre nested types}.
 %do not preclude any nested
@@
 %the derivation of, e.g., the type $\gamma; \emptyset \vdash
-%\Nat^\alpha (\mathit{List}\, \alpha) \, (\mathit{Tree} \, \alpha\,
+%\Nat \alpha (\mathit{List}\, \alpha) \, (\mathit{Tree} \, \alpha\,
 %\gamma)$ representing a natural transformation from lists to trees
@@
 \AXC{$\Gamma; \ol{\alpha} \vdash F$}
-\AXC{$\Gamma; \ol\gamma, \ol{\alpha} \vdash G$}
-\AXC{$\Gamma; \ol\gamma, \ol{\alpha} \,|\, \Delta, x : F \vdash t: G$} 
-\TIC{$\Gamma; \ol\gamma
-  %\emptyset
+\AXC{$\Gamma; \ol{\alpha} \vdash G$}
+\AXC{$\Gamma; \ol{\alpha} \,|\, \Delta, x : F \vdash t: G$} 
+\TIC{$\Gamma; \emptyset
   \,|\, \Delta \vdash L_{\ol{\alpha}} x.t : \Nat^{\ol{\alpha}} \,F \,G$}
@@
 \\\\
-\AXC{$\Gamma; \ol\gamma
-  %\emptyset
+\AXC{$\Gamma; \emptyset
   \,|\, \Delta \vdash t : \Nat^{\ol{\alpha}} \,F \,G$}
-\AXC{$\ol{\Gamma;\Phi,\ol\gamma \vdash K}$}
-\AXC{$\Gamma;\Phi,\ol\gamma \,|\, \Delta \vdash s: F[\overline{\alpha := K}]$}
-\TIC{$\Gamma;\Phi,\ol\gamma \,|\, \Delta \vdash t_{\ol K} s:
+\AXC{$\ol{\Gamma;\Phi \vdash K}$}
+\AXC{$\Gamma;\Phi \,|\, \Delta \vdash s: F[\overline{\alpha := K}]$}
+\TIC{$\Gamma;\Phi\,|\, \Delta \vdash t_{\ol K} s:
   G[\overline{\alpha := K}]$}
@@
 \AXC{$\ol{\Gamma; \ol{\beta},\ol{\gamma} \vdash F}$}
-\AXC{$\ol{\Gamma; \ol\alpha, \ol{\beta},\ol{\gamma} \vdash
-    G}$\hspace*{0.3in}$\ol{\Gamma;\ol\alpha~|~\Delta \vdash t :
-  \Nat^{\ol{\beta},\ol{\gamma}}\,F\,G}$}
-\TIC{$\Gamma; \ol\alpha
-  %\emptyset
-  ~|~\Delta
+\AXC{$\ol{\Gamma; \ol{\beta},\ol{\gamma} \vdash
+    G}$}
+\TIC{$\Gamma; \emptyset
+  ~|~\emptyset
 %  \ol{ x : \Nat^{\ol{\beta},\ol{\gamma}}\,F\,G}
   %  \emptyset
-  \vdash \map^{\ol{F},\ol{G}}_H \, \ol t :
-%  \Nat^\emptyset\;(\ol{\Nat^{\ol{\beta},\ol{\gamma}}\,F\,G})\;
-  \Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\;H[\ol{\phi
-      :=_{\ol{\beta}} G}]$} 
+  \vdash \map^{\ol{F},\ol{G}}_H :
+  \Nat^\emptyset\;(\ol{\Nat^{\ol{\beta},\ol{\gamma}}\,F\,G})\;
+  (\Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\;H[\ol{\phi
+      :=_{\ol{\beta}} G}])$} 
 \DisplayProof
@@
 \map_{\mathit{PTree}\,\alpha}^{\alpha \times \alpha, \alpha \times
-  \alpha} t & : & \Nat^{\alpha} (\mathit{PTree}\,(\alpha \times
-\alpha))\, (\mathit{PTree}\,(\alpha \times \alpha))
+  \alpha} & : & \Nat^\emptyset
+(\Nat^{\alpha} (\alpha \times \alpha)\, (\alpha \times
+\alpha))\,
+(\Nat^{\alpha} (\mathit{PTree}\,(\alpha \times
+\alpha))\, (\mathit{PTree}\,(\alpha \times \alpha)))
 %\[\begin{array}{lll}
@@
 \end{array}\]
-for $t : \Nat^{\alpha} (\alpha \times \alpha)\, (\alpha \times
-\alpha)$, 
 and 
@@
 \Nat^\emptyset \ (\List\,\alpha)\,(\Nat^\emptyset \alpha\,(\List
-    {\color{blue} \nat}))$, all of whose argument types are
+    (\mu \beta. 1 + \beta)))$, all of whose argument types are
     parameterized over the same type $\alpha$. This is not possible
-    for nested types, even when those nested types are semantically
+    for nested types, even when they are semantically
     equivalent to ADTs.
@@
 multiple nested types as arguments, provided the types of those
-arguments are parameterized over disjoint sets of type variables. The
-return type of the function can be parameterized over any or all of
-the type variables occurring in its argument types.  For example, we
-can represent the type of a zip-like function for bushes.
+arguments are parameterized over disjoint sets of type variables; the
+return type of such a function can be parameterized over only 
+%any or all of
+the type variables occurring in its final argument's type.
+%
+%For example, we can represent the type of a zip-like function for bushes.
+%
 %A well-known deeper limitation on folds is that a fold over a nested
@@
 %its first argument.
-Note, however, that even some recursive functions of a single proper
+Moreover, even some recursive functions of a single proper
 nested type --- e.g., a $\mathtt{reverseBush}$ function that is a true
@@
 of parametricity for languages that construct nested types as
-fixpoints studied here.
+fixpoints studied here. Enhancing the expressivity of our calculus is
+discussed further in Section~\ref{sec:conclusion}.
 %, and is orthogonal to, our focus on
@@
 \AXC{$\Gamma;\emptyset \vdash \mathit{PTree}\,F$}
-\AXC{$\Gamma;\emptyset \vdash \mathit{List}\,G$ \hspace*{0.3in}
-  $\Gamma;\emptyset~|~\Delta \vdash t
-  : \Nat^\emptyset\,(\mathit{PTree}\, F)\,(\mathit{List}\,
-  G)$}
-\TIC{$\Gamma;\emptyset \,|\, \Delta \vdash \map^{\mathit{PTree}\,F,
-    \,\mathit{List}\,G}_{\mathit{List}\,\alpha} \,t :
-  \Nat^\emptyset\,(\mathit{List}\, (\mathit{PTree}\,
-  F))\,(\mathit{List}\, (\mathit{List}\, G))$} \DisplayProof
+\AXC{$\Gamma;\emptyset \vdash \mathit{List}\,G$}
+\TIC{$\Gamma;\emptyset \,|\, \emptyset \vdash \map^{\mathit{PTree}\,F,
+    \,\mathit{List}\,G}_{\mathit{List}\,\alpha} :
+  \Nat^\emptyset (\Nat^\emptyset\,(\mathit{PTree}\, F)\,(\mathit{List}\, G))\,
+  (\Nat^\emptyset\,(\mathit{List}\, (\mathit{PTree}\,
+  F))\,(\mathit{List}\, (\mathit{List}\, G)))$} \DisplayProof
 \end{array}\]
-But in the absence of $\ol \gamma$, if $\Gamma;
-\emptyset~|~\Delta\vdash t : \Nat^\beta(\mathit{PTree}\,
-\beta)\,(\mathit{List}\, \beta)$ then the instance
-\[\Gamma;\emptyset~|~\Delta \vdash
-\map^{\mathit{PTree}\,\beta,\mathit{List}\,\beta}_{\mathit{List}\,\alpha}\,t
-: \Nat^\beta\,(\mathit{List}\, (\mathit{PTree}\,
-\beta))\,(\mathit{List}\, (\mathit{List}\, \beta))\]
+But in the absence of $\ol \gamma$, the instance
+\[\Gamma;\emptyset~|~\emptyset \vdash
+\map^{\mathit{PTree}\,\beta,\mathit{List}\,\beta}_{\mathit{List}\,\alpha}
+: \Nat^\emptyset ( \Nat^\beta(\mathit{PTree}\,
+\beta)\,(\mathit{List}\, \beta))\, (\Nat^\beta\,(\mathit{List}\,
+(\mathit{PTree}\, 
+\beta))\,(\mathit{List}\, (\mathit{List}\, \beta)))\]
 %\[\Gamma;\emptyset \,|\, x : \Nat^\beta(\mathit{PTree}\,
@@
 \AXC{$\Gamma;\beta \vdash \mathit{PTree}\,\beta \hspace*{0.3in}
-  \Gamma;\beta \vdash \mathit{List}\,\beta$\hspace*{0.3in}
-$\Gamma;\emptyset~|~\Delta \vdash t : \Nat^\beta(\mathit{PTree}\,
-  \beta)\,(\mathit{List}\, \beta)$}
-\BIC{$\Gamma;\emptyset~|~\Delta \vdash
-  \map^{F,G}_{\mathit{List}\,\alpha} \, t :
- \Nat^\beta\,(\mathit{List}\,
+  \Gamma;\beta \vdash \mathit{List}\,\beta$}
+\BIC{$\Gamma;\emptyset~|~\emptyset \vdash
+  \map^{F,G}_{\mathit{List}\,\alpha}  : \Nat^\emptyset
+  (\Nat^\beta(\mathit{PTree}\, 
+  \beta)\,(\mathit{List}\, \beta))\,
+ (\Nat^\beta\,(\mathit{List}\,
   (\mathit{PTree}\, \beta))\,(\mathit{List}\, (\mathit{List}\,
-  \beta))$}
+  \beta)))$}
 \DisplayProof
@@
 %    if } v \in \tvars^0 \\ 
-  \setsem{\Gamma; \ol\gamma
-    %\emptyset
+  \setsem{\Gamma; \emptyset
     \vdash \Nat^{\ol{\alpha}}
@@
     F}\rho[\ol{\alpha := A}] 
-      \Rightarrow \lambda \ol{A}.\,\setsem{\Gamma; \ol\gamma,
+      \Rightarrow \lambda \ol{A}.\,\setsem{\Gamma; 
         \ol{\alpha} \vdash G}\rho[\ol{\alpha := A}] \\ 
@@
       : \relsem{\Gamma; \ol{\alpha} \vdash F}\Eq_{\rho}[\ol{\alpha := R}]
-      \rightarrow \relsem{\Gamma;\ol\gamma, \ol{\alpha} \vdash
+      \rightarrow \relsem{\Gamma; \ol{\alpha} \vdash
         G}\Eq_{\rho}[\ol{\alpha := R}] \} \\
@@
 ensure that $\{\eta : \setsem{\Gamma;\ol{\alpha} \vdash F}\rho
-\Rightarrow \setsem{\Gamma;\ol\gamma,\ol{\alpha} \vdash G}\rho\}$
-(which contains $\setsem{\Gamma;\ol\gamma \vdash
+\Rightarrow \setsem{\Gamma;\ol{\alpha} \vdash G}\rho\}$ (which
+contains $\setsem{\Gamma;\emptyset \vdash
   \Nat^{\ol{\alpha}}\,F\,G}\rho$) is a subset of
-$\big\{({\setsem{\Gamma;\ol\gamma,\ol{\alpha} \vdash G}\rho[\ol{\alpha
-      := S}]})^{(\setsem{\Gamma;\ol{\alpha} \vdash F}\rho[\ol{\alpha
-      := S}])}~\big|~ \ol{S} = (S_1,...,S_{|\ol{\alpha}|}), \mbox{ and
-}$ $S_i \mbox{ is a finite set for } i =
+$\big\{({\setsem{\Gamma;\ol{\alpha} \vdash G}\rho[\ol{\alpha :=
+      S}]})^{(\setsem{\Gamma;\ol{\alpha} \vdash F}\rho[\ol{\alpha :=
+      S}])}~\big|~ \ol{S} = (S_1,...,S_{|\ol{\alpha}|}), \mbox{ and }$
+$S_i \mbox{ is a finite set for } i =
 1,...,|\ol{\alpha}|\big\}$. There are countably many choices of tuples
@@
 ${\setsem{\Gamma;\ol{\alpha} \vdash F}\rho[ \ol{\alpha := S}]}$ to
-${\setsem{\Gamma;\ol\gamma,\ol{\alpha} \vdash G}\rho[\ol{\alpha :=
-      S}]}$. But there are only $\set$-many choices of morphisms
-between any two objects since $\set$ is locally small.  That
-$\setsem{\Gamma; \ol\gamma \vdash \Nat^\alpha F\,G}$ is
-$\omega$-cocontinuous follows from $\omega$-cocontinuity of
-$\setsem{\Gamma;\ol\alpha \vdash F}$,
-$\setsem{\Gamma;\ol\gamma,\ol\alpha \vdash G}$, and the equality
-relation environment $\Eq$. Interpretations of $\Nat$-types ensure
+${\setsem{\Gamma;\ol{\alpha} \vdash G}\rho[\ol{\alpha := S}]}$. But
+there are only $\set$-many choices of morphisms between any two
+objects since $\set$ is locally small.  Note that $\setsem{\Gamma;
+  \emptyset \vdash \Nat^\alpha F\,G}$ is $\omega$-cocontinuous because
+it is a constant functor.
+%follows from $\omega$-cocontinuity of $\setsem{\Gamma;\ol\alpha
+%\vdash F}$, $\setsem{\Gamma;\ol\alpha \vdash G}$, and the equality
+%relation environment $\Eq$.
+Interpretations of $\Nat$-types ensure
 that $\setsem{\Gamma \vdash F \to G}$ and $\setsem{\Gamma \vdash
@@
 %  $\setsem{\Gamma,v;\emptyset \vdash v}f = \id_{\rho v}$
-\item If \,$\Gamma; \ol\gamma
-  %\emptyset
+\item If \,$\Gamma; \emptyset
   \vdash \Nat^{\ol{\alpha}}\,F\,G$ then
-  $\setsem{\Gamma; \ol\gamma
-    %\emptyset
-    \vdash \Nat^{\ol{\alpha}}\,F\,G} f =$\\
-%  \id_{\setsem{\Gamma; \Phi
-      %\emptyset
-%      \vdash \Nat^{\ol{\alpha}}\,F\,G}\rho}$
-\hspace*{0.2in} $\lambda t : \setsem{\Gamma;\ol\gamma \vdash
-    \Nat^{\ol{\alpha}}\,F\,G}\rho.\, 
-  (\lambda \ol{A}. \setsem{\Gamma; \ol\gamma,\ol\alpha \vdash G}f[\ol{\alpha
-      := id_A}]) \circ t$  
+  $\setsem{\Gamma; \emptyset
+    \vdash \Nat^{\ol{\alpha}}\,F\,G} f =
+  \id_{\setsem{\Gamma; \emptyset
+      \vdash \Nat^{\ol{\alpha}}\,F\,G}\rho}$
+%\hspace*{0.2in} $\lambda t : \setsem{\Gamma;\emptyset\vdash
+%    \Nat^{\ol{\alpha}}\,F\,G}\rho.\, 
+%  (\lambda \ol{A}. \setsem{\Gamma; \ol\alpha \vdash G}f[\ol{\alpha
+%      := id_A}]) \circ t$  
 \item If \,$\Gamma;\Phi \vdash \phi \ol{F}$ then
@@
   \begin{align*}
-  \relsem{\Gamma; \ol\gamma \vdash \Nat^{\ol{\alpha}} \,F\,G}\rho &= \{\eta
+  \relsem{\Gamma; \emptyset \vdash \Nat^{\ol{\alpha}} \,F\,G}\rho &= \{\eta
   : \lambda \ol{R}.\,\relsem{\Gamma; \ol{\alpha} \vdash
     F}\rho[\ol{\alpha := R}] \Rightarrow \lambda \ol{R}. \,\relsem{
-    \Gamma; \ol\gamma,\ol{\alpha} \vdash G}\rho[\ol{\alpha := R}]\}\\
+    \Gamma; \ol{\alpha} \vdash G}\rho[\ol{\alpha := R}]\}\\
   &=
-  \{(t,t') \in \setsem{\Gamma; \ol\gamma
+  \{(t,t') \in \setsem{\Gamma; \emptyset
     \vdash \Nat^{\ol{\alpha}}
     \,F\,G} (\pi_1 \rho) \times \setsem{ 
-    \Gamma;\ol\gamma
+    \Gamma;\emptyset
     \vdash \Nat^{\ol{\alpha}} \,F\,G} (\pi_2
@@
   & \hspace{0.4in} (t_{\ol{A}},t'_{\ol{B}}) \in
-  (\relsem{\Gamma; \ol\gamma,\ol{\alpha} \vdash G}\rho[\ol{\alpha :=
+  (\relsem{\Gamma; \ol{\alpha} \vdash G}\rho[\ol{\alpha :=
       R}])^{\relsem{\Gamma;\ol{\alpha}\vdash F}\rho[\ol{\alpha := R}]} \}\\  
@@
 %\relsem{\Gamma \vdash \tau} \rho$.
-That $\relsem{\Gamma; \ol\gamma \vdash \Nat^\alpha F\,G}$ is
-$\omega$-cocontinuous follows from $\omega$-cocontinuity of
-$\relsem{\Gamma;\ol\alpha \vdash F}$ and
-$\relsem{\Gamma;\ol\gamma,\ol\alpha \vdash G}$.  If $\rho \in \relenv$
+As for set interpretations, $\relsem{\Gamma; \emptyset \vdash
+  \Nat^\alpha F\,G}$ is $\omega$-cocontinuous because it is a constant
+functor.
+%follows from $\omega$-cocontinuity of
+%$\relsem{\Gamma;\ol\alpha \vdash F}$ and
+%$\relsem{\Gamma;\ol\alpha \vdash G}$.
+If $\rho \in \relenv$
 and $\vdash F$, then we write $\relsem{\vdash F}$ instead of
@@
 \[\begin{array}{lll}
-\dsem{\Gamma;\emptyset \,|\, \Delta,x :F \vdash x : F} \rho& = &
+\dsem{\Gamma;\Phi \,|\, \Delta,x :F \vdash x : F} \rho& = &
 \pi_{|\Delta|+1}\\
@@
 %\rangle\\
-\dsem{\Gamma;\ol\gamma \,|\, \Delta \vdash L_{\overline \alpha} x.t : \Nat^{\overline
-    \alpha} \,F \,G}\rho & = &  \curry (\dsem{\Gamma;\ol\gamma,\overline \alpha
+\dsem{\Gamma;\emptyset \,|\, \Delta \vdash L_{\overline \alpha} x.t : \Nat^{\overline
+    \alpha} \,F \,G}\rho & = &  \curry (\dsem{\Gamma;\overline \alpha
   \,|\, \Delta, x : F \vdash t: G}\rho[\overline{\alpha := \_}])\\
-\dsem{\Gamma;\Phi,\ol\gamma \,|\, \Delta \vdash t_{\overline K} s:
+\dsem{\Gamma;\Phi \,|\, \Delta \vdash t_{\overline K} s:
   G [\overline{\alpha := K}]}\rho & = & \eval \circ \langle
-  \lambda d.\,(\dsem{\Gamma;\ol\gamma \,|\, \Delta \vdash t :
-  \Nat^{\overline{\alpha}} \,F \,G}\rho\; d)_{\overline{\dsem{\Gamma;\Phi,\ol\gamma
+  \lambda d.\,(\dsem{\Gamma;\emptyset \,|\, \Delta \vdash t :
+  \Nat^{\overline{\alpha}} \,F \,G}\rho\; d)_{\overline{\dsem{\Gamma;\Phi
       \vdash K}\rho}},\\ 
- & & \hspace*{0.5in} \dsem{\Gamma;\Phi,\ol\gamma \,|\,
+ & & \hspace*{0.5in} \dsem{\Gamma;\Phi \,|\,
     \Delta \vdash s: F [\overline{\alpha := K}]}\rho \rangle\\ 
@@
 %% \color{red} \mbox{Add rules for } \forall \mbox{ if we include it} & & \\
-\dsem{\Gamma;\Phi \,|\, \Delta,x :F \vdash x : F} \rho& = &
-\pi_{|\Delta|+1}\\
+%\dsem{\Gamma;\Phi \,|\, \Delta,x :F \vdash x : F} \rho& = &
+%\pi_{|\Delta|+1}\\
 \dsem{\Gamma;\Phi \,|\, \Delta \vdash \bot_F t : F} \rho& = &
@@
 \inr \circ \dsem{\Gamma;\Phi \,|\, \Delta \vdash t : G}\rho\\
-\dsem{\Gamma;\ol\alpha \,|\, \Delta \vdash \map^{\ol{F},\ol{G}}_H
-\,\ol t: \Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\,H[\ol{\phi
-      :=_{\ol{\beta}} G}]} \rho
-& = & \lambda d\, \ol{C}.\,
-\dsem{\Gamma; \ol{\phi},\ol{\gamma}\vdash H}\\ 
-\hspace*{0.79in}  & & \hspace*{0.1in}\id_{\rho[\ol{\gamma :=
-      C}]}[\ol{\phi := \lambda
-    \ol{B}. (\dsem{\Gamma;\ol\alpha\,|\,\Delta \vdash t : 
-    \Nat^{\ol{\beta},\ol{\gamma}}\,F\,G}\rho\,d)_{\ol{B}\,\ol{C}}}]\\
+\llbracket \Gamma;\emptyset \,|\, \emptyset \vdash \map^{\ol{F},\ol{G}}_H
+  : \Nat^\emptyset (\Nat^{\ol\beta,\ol\gamma} F\,G)
+& = & \lambda d\, \ol\eta\,\ol{C}.\,
+\dsem{\Gamma; \ol{\phi},\ol{\gamma}\vdash H}\id_{\rho[\ol{\gamma:=
+      C}]}[\ol{\phi := \lambda \ol B. \eta_{\ol B\, \ol C}}]\\ 
+\hspace*{0.5in}
+  (\Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\,H[\ol{\phi
+      :=_{\ol{\beta}} G}]) \rrbracket^{\mathsf D} \rho & &\\
+%\hspace*{0.79in}  & & \hspace*{0.1in}\id_{\rho[\ol{\gamma :=
+%      C}]}[\ol{\phi := \lambda
+%    \ol{B}. (\dsem{\Gamma;\ol\alpha\,|\,\Delta \vdash t : 
+%    \Nat^{\ol{\beta},\ol{\gamma}}\,F\,G}\rho\,d)_{\ol{B}\,\ol{C}}}]\\
 \llbracket \Gamma;\emptyset \,|\, \emptyset \vdash \tin_H :
@@
 
-Straightforward calculation shows that terms of $\Nat$ type behave as
-natural transformations with respect to their source and target
-functorial types, i.e., we have
+Define, for $\Gamma; \ol{\alpha} \vdash F$, the term $\id_F$ to be
+$\Gamma;\emptyset~|~\emptyset \vdash L_{\ol{\alpha}}x.x :
+\Nat^{\ol{\alpha}} F\,F$ and, for terms $\Gamma; \emptyset \,|\,
+\Delta \vdash t: \Nat^{\overline{\alpha}} F\,G$ and $\Gamma; \emptyset
+\,|\, \Delta \vdash s: \Nat^{\overline{\alpha}} G\,H$, the {\em
+  composition} $s \circ t$ of $t$ and $s$ to be $\Gamma;
+\emptyset\,|\, \Delta \vdash L_{\overline{\alpha}}
+x. s_{\overline{\alpha}}(t_{\overline{\alpha}}x):
+\Nat^{\overline{\alpha}} F\,H$. Then $\setsem{\Gamma; \emptyset \,|\,
+  \emptyset \vdash \id_{F} : \Nat^{\ol{\alpha}} F\,F} \rho\, \ast =
+\id_{\lambda \ol{A}. \setsem{\Gamma; \ol{\alpha} \vdash F} \rho
+  [\ol{\alpha := A}]}$ for any set environment $\rho$, and
+$\setsem{\Gamma; \emptyset \,|\, \Delta \vdash s \circ t:
+  \Nat^{\overline{\alpha}} F\,H} = \setsem{\Gamma; \emptyset \,|\,
+  \Delta \vdash s: \Nat^{\overline{\alpha}} G\,H} \circ
+\setsem{\Gamma; \emptyset \,|\, \Delta \vdash t:
+  \Nat^{\overline{\alpha}} F\,G}$. Straightforward calculation shows
+that terms of $\Nat$ type behave as natural transformations with
+respect to their source and target functorial types, i.e., we have
 \begin{thm}\label{eq:ft-from-nat} 
-If\, $\Gamma; \ol\beta \,|\, \Delta \vdash s : \Nat^{\overline{\alpha},
-  \overline{\gamma}} F\,G$ and $\overline{\Gamma; \ol\beta \,|\,
+  If\, $\Gamma; \emptyset \,|\, \Delta \vdash s : \Nat^{\overline{\alpha},
+  \overline{\gamma}} F\,G$ and $\overline{\Gamma; \emptyset \,|\,
   \Delta \vdash t : \Nat^{\overline{\gamma}} K\, H}$, then
@@
   \hspace*{0.135in}
-\setsem{\Gamma; \ol\beta \,|\, \Delta
-  \vdash L_{\overline{\gamma}} z. 
-  (\map_G^{\overline{K}, \overline{H}} \,\overline{t})_{\ol\gamma}
-  (s_{\overline{K}, \overline{\gamma}} z)
+\setsem{\Gamma; \emptyset\,|\, \Delta
+  \vdash 
+  (\map_G^{\overline{K}, \overline{H}} \,\overline{t})_{\ol\gamma} \circ
+(L_{\overline{\gamma}} z. s_{\overline{K}, \overline{\gamma}}
+  z)
   : \Nat^{\overline{\gamma}} F[\overline{\alpha := K}]\,
   G[\overline{\alpha := H}]}\\
-= \setsem{ \Gamma; \ol\beta \,|\, \Delta \vdash
-  L_{\overline{\gamma}} z. s_{\overline{H}, \overline{\gamma}}
-  ((\map_F^{\overline{K}, \overline{H}} \,
-  \overline{t})_{\ol\gamma} z) : \Nat^{\overline{\gamma}} F[\overline{\alpha :=
+= \setsem{ \Gamma; \emptyset \,|\, \Delta \vdash
+(L_{\overline{\gamma}} z.
+  s_{\overline{H}, \overline{\gamma}} z)\circ
+  (\map_F^{\overline{K}, \overline{H}} \,
+  \overline{t})_{\ol\gamma}  : \Nat^{\overline{\gamma}} F[\overline{\alpha :=
       K}]\, G[\overline{\alpha := H}]}
@@
 When $s = \mathsf{in}_H$, and
-$\overline{\Gamma; \ol\delta \,|\, t :
+$\overline{\Gamma; \emptyset \,|\, t :
  \Nat^{\overline{\gamma}} F\, G}$,
@@
 \begin{array}{l}
-\hspace*{0.135in}\setsem{\Gamma; \ol\delta \,|\, \Delta \vdash
+\hspace*{0.135in}\setsem{\Gamma; \emptyset \,|\, \Delta \vdash
 L_{\overline{\gamma}} z. (\map_{(\mu \phi. \lambda
@@
     \overline{\gamma}} z) : \xi}\\
-= \setsem{\Gamma; \ol\delta \,|\, \Delta
+= \setsem{\Gamma; \emptyset \,|\, \Delta
 \vdash L_{\overline{\gamma}} z. (\tin_H)_{\overline{G}, \overline{\gamma}}
@@
 Indeed, if \(\Gamma; \ol{\phi},
-\ol{\gamma} \vdash H\), \(\ol{\Gamma; \ol{\beta}, \ol{\gamma} \vdash
-  F}\), and \(\ol{\Gamma; \Phi, \ol{\beta}, \ol{\gamma} \vdash G}\),
+\ol{\gamma} \vdash H\), \(\ol{\Gamma; \emptyset, \ol{\gamma} \vdash
+  F}\), and \(\ol{\Gamma; \Phi, \emptyset, \ol{\gamma} \vdash G}\),
 %and $\Gamma;\Phi\,|\,\Delta \vdash t :
@@
 In particular, if $\Gamma; \ol{\alpha} \vdash H$, $\ol{\Gamma;
-  \emptyset \vdash F}$, and $\ol{\Gamma; \ol\beta \vdash G}$, then
+  \emptyset \vdash F}$, and $\ol{\Gamma; \emptyset \vdash G}$, then
 Definition~\ref{def:set-interp} gives, for all $\ol{f :
-  \setsem{\Gamma;\ol\beta \vdash \Nat^\emptyset F\,G}\rho}$, that
+  \setsem{\Gamma;\emptyset \vdash \Nat^\emptyset F\,G}\rho}$, that
 \[\begin{array}{rl}
-&\setsem{ \Gamma;\ol\beta\,|\,\ol{x : \Nat^\emptyset F \, G}
+&\setsem{ \Gamma;\emptyset\,|\,\ol{x : \Nat^\emptyset F \, G}
   \vdash \map^{\ol{F},\ol{G}}_H\,\ol x :
@@
 
-Define, for $\Gamma; \ol{\alpha} \vdash F$, the term $\id_F$ to be
-$\Gamma;\emptyset~|~\emptyset \vdash L_{\ol{\alpha}}x.x :
-\Nat^{\ol{\alpha}} F\,F$ and, for terms $\Gamma; \emptyset \,|\,
-\Delta \vdash t: \Nat^{\overline{\alpha}} F\,G$ and $\Gamma; \ol\beta
-\,|\, \Delta \vdash s: \Nat^{\overline{\alpha}} G\,H$, the {\em
-  composition} $s \circ t$ of $t$ and $s$ to be $\Gamma; \ol\beta
-\,|\, \Delta \vdash L_{\overline{\alpha}}
-x. s_{\overline{\alpha}}(t_{\overline{\alpha}}x):
-\Nat^{\overline{\alpha}} F\,H$. Then $\setsem{\Gamma; \emptyset \,|\,
-  \emptyset \vdash \id_{F} : \Nat^{\ol{\alpha}} F\,F} \rho\, \ast =
-\id_{\lambda \ol{A}. \setsem{\Gamma; \ol{\alpha} \vdash F} \rho
-  [\ol{\alpha := A}]}$ for any set environment $\rho$, and
-$\setsem{\Gamma; \ol\beta \,|\, \Delta \vdash s \circ t:
-  \Nat^{\overline{\alpha}} F\,H} = \setsem{\Gamma; \ol\beta \,|\,
-  \Delta \vdash s: \Nat^{\overline{\alpha}} G\,H} \circ
-\setsem{\Gamma; \emptyset \,|\, \Delta \vdash t:
-  \Nat^{\overline{\alpha}} F\,G}$.  We have the expected relationships
+We have the expected relationships
 between the interpetations of $\map$, $\tin$, and $\fold$:
@@
 $\ol{\Gamma; \ol{\alpha}, \ol{\gamma}, \ol{\phi} \vdash K}$,\;
-$\ol{\Gamma; \ol{\beta}, \ol{\gamma} \vdash F}$, 
-$\ol{\Gamma; \ol\delta, \ol{\beta}, \ol{\gamma} \vdash G}$, and
-$\ol{\Gamma;\ol\delta\,|\,\Delta \vdash t : \Nat^{\ol{\alpha}, \ol{\beta},
-    \ol{\gamma}} F\, G}$
+$\ol{\Gamma; \ol{\beta}, \ol{\gamma} \vdash F}$,\; and
+$\ol{\Gamma; \ol{\beta}, \ol{\gamma} \vdash G}$,
 then
-\[\setsem{\Gamma; \ol\delta \,|\, \Delta \vdash
-\map_{H[\ol{\psi := K}]}^{\ol{F}, \ol{G}}\,\ol t : \xi} = \setsem{\Gamma;
-  \ol\delta \,|\, \Delta \vdash \map_H^{\ol{K[\ol{\phi := F}]},
-    \ol{K[\ol{\phi := G}]}}\, (\ol{\map_K^{\ol{F}, \ol{G}}}\, \ol t) :
+\[\setsem{\Gamma; \emptyset \,|\, \emptyset \vdash
+\map_{H[\ol{\psi := K}]}^{\ol{F}, \ol{G}} : \xi} = \setsem{\Gamma;
+  \emptyset \,|\, \emptyset \vdash \map_H^{\ol{K[\ol{\phi := F}]},
+    \ol{K[\ol{\phi := G}]}} \circ \ol{\map_K^{\ol{F}, \ol{G}}} :
   \xi}\]
-for $\xi = \Nat^{\ol{\gamma}} \,H[\ol{\psi :=
-    K}][\ol{\phi := F}]\, H[\ol{\psi := K}][\ol{\phi := G}]$
+for $\xi = \Nat^{\emptyset} (\ol{\Nat^{\ol{\alpha}, \ol{\beta},
+    \ol{\gamma}} F\, G})\, (\Nat^{\ol{\gamma}} H[\ol{\psi :=
+    K}][\ol{\phi := F}]\, H[\ol{\psi := K}][\ol{\phi := G}])$
 
@@
 $\overline{\Gamma; \overline{\alpha}, \overline{\gamma} \vdash F}$,\;
-$\overline{\Gamma; \ol\delta, \overline{\alpha}, \overline{\gamma}
+$\overline{\Gamma; \emptyset, \overline{\alpha}, \overline{\gamma}
   \vdash G}$,\; 
@@
 \emptyset\,|\,\Delta \vdash s : \Nat^{\overline{\beta},
-  \overline{\gamma}} H\, K$ and $\ol{\Gamma; \ol\delta\,|\,\Delta \vdash
+  \overline{\gamma}} H\, K$ and $\ol{\Gamma; \emptyset \,|\,\Delta \vdash
   t : \Nat^{\overline{\alpha}, \overline{\gamma}} F\, G}$, then
-\[\hspace*{0.13in}\setsem{\Gamma; \ol\delta \,|\, \Delta
+\[\hspace*{0.13in}\setsem{\Gamma; \emptyset \,|\, \Delta
     \vdash L_{\overline{\gamma}} z.\,
@@
       (s, \overline{t})}) \big)_{\overline{\gamma}} z \Big) : \xi}\]
-\[= \setsem{\Gamma; \ol\delta\,|\, \Delta \vdash
+\[= \setsem{\Gamma; \emptyset\,|\, \Delta \vdash
     L_{\overline{\gamma}} z.\, \big( \map_{K}^{\overline{M
@@
 equivalent to standard folds in the presence of {\em right} Kan
-extensions.  This suggests extending our term calculus %presented here
-with an object-level right Kan extension construct; interpreting such
-a construct in $\set$ and $\rel$ should be unproblematic {\color{blue}
-  since these categories have all(?) limits}. It was also shown
-in~\cite{jg08} that GADTs can be represented using an object-level
-{\em left} Kan extension construct, and in~\cite{jp19} that augmenting
-the types in our calculus with a carefully designed such construct
-preserves the cocontinuity needed for GADTs to have well-defined
-interpretations in $\set$ and $\rel$. These observations suggest
-augmenting our type formation rules with such a construct.
-%natural directions for future work.
-Another possibility is to add
-term-level fixpoints
-%, as in~\cite{pit00},
-to our calculus. However,
-this would require the categories interpreting types to be not just
-locally finitely presentable categories, but to support some kind of
-domain structure as well. Unfortunately, $\omega$-CPO, the most
-obvious category of domains to replace $\set$, is not locally finitely
-presentable.
-
-The constructions of this paper should all be possible in any locally
-presentable cartesian closed category (lpccc) $\cal C$ whose category
-of (abstract) relations, obtained by pullback as in~\cite{jac99}, is
-also a lpccc and is appropriately fibred over $C$.  This would
+extensions.  
+%; interpreting such
+%a construct in $\set$ and $\rel$ should be unproblematic {\color{blue}
+%  since these categories have all(?) limits}.
+It was also shown in~\cite{jg08} that GADTs can be represented using
+an object-level {\em left} Kan extension construct. This suggests
+extending our term calculus with such object-level constructs.  It was
+shown in~\cite{jp19} that augmenting the types in our calculus with a
+carefully designed left Kan extension preserves the cocontinuity
+needed for GADTs to have well-defined interpretations --- but not
+necessarily in
+%However, this will not, in general, be
+%the case in
+locally finitely presentable catgories. Indeed, as shown
+in~\cite{jp19}, local $\lambda$-presentability for $\lambda >
+\aleph_1$ is required to interpret even commonly used GADTs. This
+suggests
+%augmenting our type formation rules with such a construct.
+%The constructions of this paper should all be possible
+carrying out our model construction in a locally $\lambda$-presentable
+cartesian closed category (lpccc) $\cal C$ whose category of
+(abstract) relations, obtained by pullback as in~\cite{jac99}, is also
+a lpccc and is appropriately fibred over $\cal C$.  This would
 %extend the results presented here along the lines of~\cite{gjfor15},
@@
 supporting nested types built from primitives that is based on
-``locally presentable fibrations'', for some appropriate definition
-thereof.
+``locally $\lambda$-presentable fibrations'', for some appropriate
+definition thereof.
+
+Another possibility is to add term-level fixpoints
+%, as in~\cite{pit00},
+to the calculus presented here. However, this would require the
+categories interpreting types to be not just locally finitely
+presentable categories, but to support some kind of domain structure
+as well. Unfortunately, $\omega$-CPO, the most obvious category of
+domains to replace $\set$, is not locally finitely presentable.
 

### 2020-06-25 free-theorems-nesteds/paper5.tex
added paper5.tex

--- 2020-06-24 free-theorems-nesteds/paper4.tex
@@
 This paper considers parametricity and its consequent free theorems
-for nested data types.  Rather than representing nested types via
-their Church encodings into a higher-kinded extension of System F, we
-adopt a functional programming perspective and design a polymorphic
-calculus that provides primitives for constructing a robust class of
-nested types, including ADTs and all (truly) nested types appearing in
-the literature, directly as fixpoints \`{a} la Pitts' PolyPCF.  At the
-term level, our calculus supports the usual pattern matching, map
-functions, and stylized recursion via folds for all such types. Using
-a variation on Atkey's 2012 construction
+for nested data types. Rather than representing nested types via their
+Church encodings into a higher-kinded extension of System F, we adopt
+a functional programming perspective and design a polymorphic calculus
+that provides primitives for constructing a robust class of nested
+types directly as fixpoints.
+%\`{a} la Pitts' PolyPCF.
+At the term level, our calculus supports
+%the usual
+pattern matching, map functions, and stylized recursion via folds for
+all such types.
+%Using a variation on Atkey's 2012 construction for $F_\omega$,
+%
 %of a parametric model
-for $F_\omega$,
 %that explicitly constructs equality operators rather than merely
 %postulating them,
-we construct a parametric model for our calculus. However, to ensure
-the existence of the fixpoints interpreting nested types, cocontinuity
-conditions on the functors underlying these fixpoints must be threaded
-throughout our model construction. This turns out to be both delicate
-and challenging. Overall, this paper offers a blueprint for
-deriving principled parametricity results 
-%constructing parametric models
-for practical programming languages,
+Our main technical contribution is the construction of a parametric
+model for this calculus. This is both delicate and challenging.  To
+establish a suitable Identity Extension Lemma for our calculus, mild
+restrictions are imposed on the formation of fixpoints; all ADTs and
+all (truly) nested types appearing in the literature can nevertheless
+be expressed. Moreover, to ensure the existence of the fixpoints
+interpreting nested types, cocontinuity conditions on the functors
+underlying these fixpoints must be threaded throughout our model
+construction.  Overall, this paper offers a blueprint for deriving
+principled parametricity results for practical programming languages,
 such as Haskell and Agda, that directly construct nested types as
 fixpoints rather than encoding them. It also shows how parametric
-models can be used to derive useful free theorems for programs over
-nested types. Like their counterparts for ADTs, the free theorems our
-model derives provide, in particular, a principled alternative to
-proving properties of programs over nested types using, e.g., the deep
-induction rules for nested types recently developed by Johann and
-Polonsky. This is illustrated with several examples. 
+models can derive useful free theorems for programs over nested
+types. Like their counterparts for ADTs, these free theorems
+%our model derives
+provide
+%, in particular,
+a principled alternative to proving
+properties of programs over nested types via induction, 
+e.g., using the deep
+induction rules
+%for nested types
+recently developed by
+Johann and Polonsky.
+%This is illustrated with several examples.
 \end{abstract}
@@
 from~\cite{jp19}, augmented with a primitive type of natural
-transformations. To construct nested types, it constructs type
-expressions not just from standard type variables, but also from type
-constructor variables of various arities, and includes an explicit
-$\mu$-construct for type-level recursion with respect to these
-variables. The class of nested types thus constructed is very robust
-and includes all (truly) nested types known from the literature. In
-Section~\ref{sec:type-interp} we give set and relational
-interpretations for the types From Section~\ref{sec:calculus}. As is
-usual when modeling parametricity, types are interpreted as functors
-from environments interpreting their type variable contexts to set or
+transformations.  To construct nested types, it carefully builds up
+type expressions not just from standard type variables, but also from
+type constructor variables of various arities, and includes an
+explicit $\mu$-construct for type-level recursion with respect to
+these variables. The class of nested types it constructs is very
+robust and includes all (truly) nested types known from the
+literature.  In Section~\ref{sec:type-interp} we give set and
+relational interpretations
+%{\color{blue} locally finitely presentable}
+for the types from Section~\ref{sec:calculus}. As is usual when
+modeling parametricity, types are interpreted as functors from
+environments interpreting their type variable contexts to set or
 relations, as appropriate. But in order to ensure that these functors
@@
 model is guaranteed to satisfy an appropriate Identity Extension Lemma
-(Theorem~\ref{thm:iel}). Properly progagating the cocontinuity
-conditions turns out to be both subtle and challenging, and
-Section~\ref{sec:iel}, where it is done, is where the bulk of the work
-in constructing our model lies.  At the term level, our calculus
-includes primitive constructs for the actions on morphisms of the
-functors interpreting types, initial algebras for fixpoints of these
-functors, and structured recursion over elements of these initial
-algebras (i.e., $\map$, $\tin$, and $\fold$ constructs,
+(Theorem~\ref{thm:iel}).
+%{\color{blue} Syntactic restrictions on
+%  $\mu$-types are essential to proving IEL.}
+Properly progagating the cocontinuity conditions turns out to be both
+subtle and challenging, and Section~\ref{sec:iel}, where it is done,
+is where the bulk of the work in constructing our model lies.  At the
+term level, our calculus includes primitive constructs for the actions
+on morphisms of the functors interpreting types, initial algebras for
+fixpoints of these functors, and structured recursion over elements of
+these initial algebras (i.e., $\map$, $\tin$, and $\fold$ constructs,
 respectively). While our calculus does not support general recursion
@@
 toward the kind of provably total practical programming language
-proposed at the end of~\cite{wad89}. In Section~\ref{sec:term-interp},
-we give set and relational interpretations for the terms of our
-calculus. As usual in parametric models, terms are interpreted as
-natural transformations from interpretations of the term contexts in
-which they are formed to the interpretations of their types, and these
-must cohere in what is essentially a fibrational way. Immediately from
-the definitions of our interpretations we prove in
-Section~\ref{sec:Nat-type-terms} a scheme deriving free theorems that
-are consequences of naturality of functions that are polymorphic over
-nested types. This scheme is very general, is parameterized over both
-the data type and the polymorphic function at hand, and has each of
-the above $\mathtt{map}$-$\mathtt{reverse}$ theorems as instances. The
+proposed at the end of~\cite{wad89}.  In
+Section~\ref{sec:term-interp}, we give set and relational
+interpretations for the terms of our calculus. As usual in parametric
+models, terms are interpreted as natural transformations from
+interpretations of the term contexts in which they are formed to the
+interpretations of their types, and these must cohere in what is
+essentially a fibrational way. Immediately from the definitions of our
+interpretations we prove in Section~\ref{sec:Nat-type-terms} a scheme
+deriving free theorems that are consequences of naturality of
+functions that are polymorphic over nested types. This scheme is very
+general, is parameterized over both the data type and the polymorphic
+function at hand, and has each of the above
+$\mathtt{map}$-$\mathtt{reverse}$ theorems as instances.  The
 relationship between naturality and parametricity has long been of
@@
 
-Because our calclus is designed to ensure functoriality, and
-functoriality does not interact with polymorphism as well as might be
-hoped, our calculus cannot express nontrivial recursive functions, such
-as a concatenation function for perfect trees, that take not just one
-nested type parameterized over a variable $\alpha$ as input, but other
-inputs parameterized over $\alpha$ as well. The fundamental issue is
-that recursion is expressible only via folds, and it is well-known
-that a fold over a nested type must take a natural transformation as
-an argument and return a natural transformation as its result.
+Our rule for forming $\Nat$-types means that 
+%Because our calculus is designed to ensure functoriality, and
+%functoriality does not interact with polymorphism as well as might be
+%hoped,
+our calculus cannot express types of recursive functions, such as a
+concatenation function for perfect trees, that take as inputs both a
+nested type and an argument of another type, both of which are
+parameterized over the same variable $\alpha$.
+%not just one nested type parameterized over a variable $\alpha$ as
+%input, but other inputs parameterized over $\alpha$ as well.
+Similarly, our calculus cannot express the type of a zip function for
+bushes. The fundamental issue is that recursion is expressible only
+via folds, and a fold over a nested type must take a natural
+transformation as its argument and return a natural transformation as
+its result.
 %and, unfortunately, the polymorphism natural transformations entail
@@
 Even for ADTs there is a difference between which folds over them we
-can write when ADTs are viewed as proper nested types --- i.e., as
-fixpoints of types interpreted as first-order functors --- and which
-ones we can write when ADTs are viewed as nested types --- i.e., as
-fixpoints of types that are interpreted as higher-order functors. This
-is because enforcing functoriality requires that the arity of the
-recursion variable $\phi$ in the fixpoint in the return type of
-$\fold$ must equal the number of variables bound by its $\Nat$
+can type when ADTs are viewed as ADTs --- i.e., as fixpoints of types
+interpreted as first-order functors --- and which folds we can type
+when ADTs are viewed as proper nested types --- i.e., as fixpoints of
+types that are interpreted as higher-order functors --- because the
+arity of the recursion variable $\phi$ in the fixpoint in the return
+type of $\fold$ must equal the number of variables bound by its $\Nat$
 construct. For ADTs this number can be $0$, making it possible to
-write as folds functions of types such as $\alpha; \emptyset \vdash
-\Nat^\emptyset \ (\List\,\alpha)\,(\Nat^\emptyset \alpha\,(\List
-    (\mu \beta. 1 + \beta)))$, all of whose argument types are
-    parameterized over the same type $\alpha$. This is not possible
-    for nested types, even when they are semantically
-    equivalent to ADTs.
+write types such as $\alpha; \emptyset \vdash \Nat^\emptyset
+\ (\List\,\alpha)\,(\Nat^\emptyset \alpha\,(\List (\mu \beta. 1 +
+\beta)))$, all of whose argument types are parameterized over the same
+type $\alpha$. This is not possible even for nested types that are
+semantically equivalent to ADTs.
 %Currying the type of the function does not help either, since the
@@
 %more than one input involving the same type variable.
-We can, however, represent types of functions that take
-%    as arguments
-%one proper nested type and as many other (possibly proper)
-%as many proper nested types as we wish
-multiple nested types as arguments, provided the types of those
-arguments are parameterized over disjoint sets of type variables; the
-return type of such a function can be parameterized over only 
-%any or all of
-the type variables occurring in its final argument's type.
 %
@@
 %its first argument.
-Moreover, even some recursive functions of a single proper
+Our calculus can, however, express types of functions
+that take
+%    as arguments
+%one proper nested type and as many other (possibly proper)
+%as many proper nested types as we wish
+multiple nested types as arguments, provided those nested types are
+parameterized over disjoint sets of type variables and the return type of
+the function is parameterized over only
+%any or all of
+the variables occurring in the type of its final argument.
+Interestingly, even some recursive functions of a single proper
 nested type --- e.g., a $\mathtt{reverseBush}$ function that is a true
@@
 %, and thus must itself be defined as a fold
-with type signatures of the same problematic forms.
+with types of the same problematic forms.
 %More sophisticated combinators for stylized recursion (e.g.,
@@
 %polymorphism, not just some peculiarity of our particular calculus.
-Adding combinators for more sophisticated stylized recursion may
-ultimately enhance the expressivity of our calculus but, as discussed
-in Section~\ref{sec:conclusion}, the expressivity issue for folds for
-nested types long pre-dates, and is somewhat orthogonal to, the issue
-of parametricity for languages that construct nested types as
-fixpoints studied here. Enhancing the expressivity of our calculus is
-discussed further in Section~\ref{sec:conclusion}.
-%, and is orthogonal to, our focus on
-%constructing parametric models for calculi in which nested types are
-%constructed as fixpoints.
+Allowing extra functorial variables in the codomains of $\Nat$-types
+would recover typeability of, say, zip for bushes, but such extended
+$\Nat$-types need not have semantics as $\omega$-cocontinuous functors
+in $\set$, as will be required in Section~\ref{sec:type-interp}. For
+example,
+%as shown in Example D, Section VI of~\cite{jp19},
+the type $\emptyset; \gamma \vdash \Nat^\emptyset (\onet + \onet)\,
+\gamma$ does not. We would therefore have to construct our models
+from semantic categories that are locally $\lambda$-presentable for
+$\lambda > \omega$.
+%It is tempting to try to recover typeability of, say, a zip function
+%for bushes by allowing extra functorial variables in the codomain type
+%of $\Nat$-types. But such extended $\Nat$-types need not have
+%semantics as $\omega$-cocontinuous functors, as is required in
+%Section~\ref{sec:type-interp}. For example, as shown in Example D,
+%Section VI of~\cite{jp19}, the type $\emptyset; \gamma \vdash
+%\Nat^\emptyset (\onet + \onet)\, \gamma$ does not.
+%
+Adding more expressive stylized recursion combinators could
+also help,
+%enhance the expressivity of our calculus
+but the expressivity issue for folds for nested types long pre-dates,
+and is somewhat orthogonal to, the issue of parametricity for
+languages that construct nested types as fixpoints studied here.
+Enhancing the expressivity of our calculus is discussed further in
+Section~\ref{sec:conclusion}.
 
@@
 %\beta))\] of $\mathtt{map}$
-required to map the {\em polymorphic}
-$\mathit{flatten}$ function over a list of perfect trees is not: in
-that setting the functorial contexts for $F$ and $G$ in the rule for
-$\map^{F,G}_H$ would have to be empty, but the fact that the
-polymorphic $\mathit{flatten}$ function is functorial in some
-variable, say $\delta$, means that it cannot possibly have a type of
-the form $\Nat^\emptyset F\, G$ that would be required for it to be
-the function input to $\map$. Since untypeability of this instance of
-$\map$ is unsatisfactory in a polymorphic calculus, where we naturally
-expect to be able to manipulate entire polymorphic functions rather
-than just their monomorphic instances, we use the ``extra'' variables
-in $\ol \gamma$ to remedy the situation. Specifically, the rules from
-Definition~\ref{def:well-formed-terms} ensure that the instance of
-$\map$ needed to map the polymorphic $\mathit{flatten}$ function is
-typeable as follows:
+required to map the {\em polymorphic} $\mathit{flatten}$ function over
+a list of perfect trees is not: in that setting the functorial
+contexts for $F$ and $G$ in the rule for $\map^{F,G}_H$ would have to
+be empty, but because the polymorphic $\mathit{flatten}$ function is
+functorial in some variable, say $\delta$, it cannot possibly have a
+type of the form $\Nat^\emptyset F\, G$ that would be required for it
+to be the function input to $\map$. Since untypeability of this
+instance of $\map$ is unsatisfactory in a polymorphic calculus, where
+we naturally expect to be able to manipulate entire polymorphic
+functions rather than just their monomorphic instances, we use the
+``extra'' variables in $\ol \gamma$ to remedy the
+situation. Specifically, Definition~\ref{def:well-formed-terms}
+ensures that the instance of $\map$ needed to map the polymorphic
+$\mathit{flatten}$ function is typeable as follows:
 \[\begin{array}{l}
@@
 cocontinuity conditions we must impose to ensure that they are
-well-defined. We take some effort to develop conditions in
+well-defined. We develop these conditions in
 Section~\ref{sec:rel-interp}, which separates
@@
 %       accessible and cocomplete.
-We will interpret the types in our calculus as $\omega$-cocontinuous
-functors on locally finitely presentable categories~\cite{ar94}. Since
-functor categories of locally finitely presentable categories are
-again locally finitely presentable, this will ensure, in particular,
-that the fixpoints interpreting $\mu$-types in $\set$ and $\rel$
-exist, and thus that both the set and relational interpretations of
-all of the types in Definition~\ref{def:wftypes} are
-well-defined~\cite{jp19}. To bootstrap this process, we interpret type
-variables themselves as $\omega$-cocontinuous functors in
-Definitions~\ref{def:set-env} and~\ref{def:reln-env}. If $\C$ and $\D$
-are locally finitely presentable categories, we write $[\C,\D]$ for
-the set of $\omega$-cocontinuous functors from $\C$ to $\D$.
+We interpret types in our calculus as $\omega$-cocontinuous functors
+on locally finitely presentable categories~\cite{ar94}. Since functor
+categories of locally finitely presentable categories are again
+locally finitely presentable, this ensures that the fixpoints
+interpreting $\mu$-types in $\set$ and $\rel$ exist, and thus that
+both the set and relational interpretations of all of the types in
+Definition~\ref{def:wftypes} are well-defined~\cite{jp19}. To
+bootstrap this process, we interpret type variables as
+$\omega$-cocontinuous functors in Definitions~\ref{def:set-env}
+and~\ref{def:reln-env}. If $\C$ and $\D$ are locally finitely
+presentable categories, we write $[\C,\D]$ for the set of
+$\omega$-cocontinuous functors from $\C$ to $\D$.
 %Note that the categories $\set$ and $\rel$ are both locally finitely
@@
 \end{thm}
-The proof is by induction on the structure of $F$. Only the
-application and fixpoint cases are non-routine. Both use
-Lemma~\ref{lem:eq-reln-equalities}. The latter also uses the
+The proof is by induction on the structure of $F$. Only the $\Nat$,
+application, and fixpoint cases are non-routine. The latter two use
+Lemma~\ref{lem:eq-reln-equalities}. The fixpoint case also uses the
 observation that, for every $n \in \nat$, the following intermediate
 results can be proved by simultaneous induction with
-Theorem~\ref{thm:iel}: For any $H$, $\rho$, $A$, and any subformula
+Theorem~\ref{thm:iel}: for any $H$, $\rho$, $A$, and any subformula
 $J$ of $H$, both
@@
 detailed proof is given in the accompanying anonymous supplementary
-material.
+material. As noted there, if functorial variables of arity greater
+than $0$ were allowed to appear in the bodies of $\mu$-types, the IEL
+would fail.
 
@@
 x, \setsem{\Gamma; \Phi \vdash F}\id_{\rho'}\, y) \in \relsem{\Gamma;
-  \Phi \vdash F}\Eq_{\rho'}$.  Then $\setsem{\Gamma; \Phi \vdash
+  \Phi \vdash F}\Eq_{\rho'}$. So $\setsem{\Gamma; \Phi \vdash
   F}\id_{\rho'}\, y = \id_{\setsem{\Gamma; \Phi \vdash F}\rho'}\, y =
 y$ and $\relsem{\Gamma; \Phi \vdash F}\Eq_{\rho'} =
-\Eq_{\setsem{\Gamma; \Phi \vdash F}\rho'}$, so if $(x,y) \in
+\Eq_{\setsem{\Gamma; \Phi \vdash F}\rho'}$, and if $(x,y) \in
 \relsem{\Gamma; \Phi \vdash F}\graph{f}$ then $(\setsem{\Gamma; \Phi
@@
 equation gives that if $x \in \setsem{\Gamma; \Phi \vdash F}\rho$ then
-$(\setsem{\Gamma; \Phi \vdash F}\id_{\rho}\, x, \setsem{\Gamma; \Phi
+$(\setsem{\Gamma; \Phi \vdash F}\id_{\rho}\, x,$\\
+$\setsem{\Gamma; \Phi
   \vdash F} f\, x) \in \relsem{\Gamma; \Phi \vdash F}\graph{f}$.  Then
@@
 
-\vspace*{-0.07in}
+\vspace*{-0.08in}
 
@@
 H \; \mathit{in}_{H} = G \,\setsem{\emptyset;\alpha \vdash K}\, B$.
-
-\vspace*{0.02in}
-
-It is also standard to prove that the parametric model constructed
-verifies the existence of initial algebras. This is, however,
-unnecessary here since initial algebras are built directly into our
-model.
+Although it is standard to prove that the parametric model constructed
+verifies the existence of initial algebras, this is unnecessary here
+since initial algebras are built directly into our model.
 
@@
 
-In this paper we have constructed a parametric model for a calculus
+We have constructed a parametric model for a calculus
 providing primitives for constructing nested types directly as
-fixpoints rather than representing them via their Chuch encodings. We
-have also used the Abstraction Theorem for this model to prove free
+fixpoints.
+%rather than representing them via their Church encodings. We
+We have also used the Abstraction Theorem for this model to prove free
 theorems in the presence of nested types. This was not possible
@@
 locally finitely presentable categories like $\set$ and $\rel$. No
-calculus for terms of these types existed, either. The key to
-obtaining our parametric model is the delicate threading of
-functoriality and its accompanying naturality conditions throughout
-the model construction.
-
-Note that all recursion in our calculus must be expressed using
-folds. Expressivity of standard folds for nested types has long been
-known to be a vexing issue; indeed, such folds can only express
-computations that return natural transformations. Of course our
-calculus inherits the constraints standard folds for nested types
-impose, and thus is perhaps not as expressive as we would ideally
-like. But since this expressivity issue is orthogonal to constructing
-parametric models in the presence of nested types built from
-primitives, and since it would detreact from the conceptual clarity of
-our presentation, we have deliberately chosen not to address it in
-this paper. (However, we discuss some ideas for increasing
-expressivity in the next paragraph.) Despite the limitations of
-standard folds, our results provide a blueprint for deriving
-principled parametricity results for practical programming languages,
-such as Haskell and Agda, that directly construct nested types as
-fixpoints.
-
-Many constructions that enhance standard folds have been
-proposed. Perhaps the most significant of these are generalized
-folds~\cite{bp99},
+term calculus for terms of these types existed, either. The key to
+obtaining our parametric model is
+%{\color{blue} adjusting our type syntax to respect the accessibility
+%  cardinal for the semantic category and}
+the delicate threading of functoriality and its accompanying
+naturality conditions throughout the model construction
+
+Even when nested types have well-defined interpretations in a semantic
+category, the types of standard functions over them may not be.  All
+recursion in our calculus must be coded using folds.  Since standard
+folds for nested types can only express computations that return
+natural transformations, their expressivity has long been a vexing
+issue, which our calculus naturally inherits.
+%indeed, such folds
+%Our calculus naturally inherits these constraints.
+%Of course our calculus inherits the constraints
+%standard folds for nested types impose, and thus is perhaps not as
+%expressive as we would ideally like.
+In addition, while there are many papers on higher-kinded types, we
+know of none that considers expressiblity issues arising from the
+local $\lambda$-presentability cardinal $\lambda$ of the categories in
+which higher-kinded types are interpreted. Because $\lambda = \omega$
+for $\set$ and $\rel$, the expressivity of our
+%$\set$/$\rel$-based parametric model
+calculus is not easily improved there. Construction of a
+$\set$/$\rel$-based parametric model for our calculus is nevertheless
+worthwhile.  For one thing, the expressivity issue is orthogonal to
+constructing parametric models,
+%in the presence of nested types built from primitives,
+and trying to address both in a single paper would
+detract from the conceptual clarity of our construction.
+%our current focus on deriving principled parametricity
+%results for practical programming languages, such as Haskell and Agda,
+%that directly construct nested types as fixpoints.
+%
+%conceptual clarity of this paper.
+For another, if the construction of this paper
+%we couldn't
+%construct a $\set$/$\rel$-based parametric model for the calculus
+%presented here,
+weren't possible then trying to construct a parametric model from
+robust semantic categories for more data types and/or recursion
+schemes would likely be futile. We naturally expect the construction
+reported here to guide its generalization to the locally
+$\lambda$-presentable categories for $\lambda > \omega$ that will be
+needed to accommodate more general data types and more general
+recursion over them.
+
+% As is, the results reported here provide a
+%blueprint for deriving principled parametricity results for practical
+%programming languages, such as Haskell and Agda, that directly
+%construct nested types as fixpoints.
+
+%{\color{blue} Perhaps surprisingly, the accessibility cardinal
+%$\omega$ for the semantic categories $\set$ and $\rel$ also impacts
+%the types of functions (and thus the functions themselves), even over
+%expressible data types, that can be expressed. Although there are
+%many papers on higher-kinded types, no other work that we know of
+%even considers expressiblity issues resulting from the semantic
+%settings in which they are interpreted.}
+%{\color{blue} In light of the aforementioned
+%  restrictions imposed by the accessibility cardinal of $\set$ and
+%  $\rel$, this paper is perhaps best read as a proof-of-concept for
+%  constructing parametric models in the locally finitely presentable
+%  categories $\set$ and $\rel$ for calculi that construct nested types
+%  and other advanced data types as fixed points (rather than
+%  representing them via their Church encodings) and that support
+%  structured recursion over them. If we couldn't construct such a
+%  model even in this restricted setting, then there'd be no point in
+%  trying to do so for more robust categories, data types, and/or
+%  recursion schemes. Moreover, our constructions in this simplest
+%  possible setting will guide their generalizations to the locally
+%  $\lambda$-presentable categories fopr $\lambda > \omega$ required to
+%  handle more general data types, structured recursive functions over
+%  then, and more general fixpoints.}
+
+%But since this expressivity issue is orthogonal to constructing
+%parametric models in the presence of nested types built from
+%primitives, and since it would detract from the conceptual clarity of
+%our presentation, we have deliberately chosen not to address it in
+%this paper. (However, we discuss some ideas for increasing
+%expressivity in the next paragraph.) Despite the limitations of
+%standard folds, our results provide a blueprint for deriving
+%principled parametricity results for practical programming languages,
+%such as Haskell and Agda, that directly construct nested types as
+%fixpoints.
+
+%All recursion in our calculus must be coded using folds. Expressivity
+%of standard folds for nested types has long been a vexing issue,
+%%indeed, such folds
+%since they can only express computations that return natural
+%transformations; our calculus naturally inherits these constraints.
+%%Of course our calculus inherits the constraints
+%%standard folds for nested types impose, and thus is perhaps not as
+%%expressive as we would ideally like.
+%But since this expressivity issue
+%is orthogonal to constructing parametric models in the presence of
+%nested types built from primitives, and since it would detract from
+%the conceptual clarity of our presentation, we have deliberately
+%chosen not to address it in this paper. (However, we discuss some
+%ideas for increasing expressivity in the next paragraph.) Despite the
+%limitations of standard folds, our results provide a blueprint for
+%deriving principled parametricity results for practical programming
+%languages, such as Haskell and Agda, that directly construct nested
+%types as fixpoints.
+
+%Many extensions of standard folds have been
+%proposed. Perhaps the most significant of these are
+Generalized folds~\cite{bp99} are an extension of standard folds that
 %and Mendler iterators~\cite{amu05}.
-which are used to express computations whose results are not natural
-transformations. Generalized folds are shown in~\cite{jg10} to be
-equivalent to standard folds in the presence of {\em right} Kan
+%which
+can express computations whose results are not natural
+transformations. These were shown in~\cite{jg10} to be equivalent to
+standard folds in the presence of object-level {\em right} Kan
 extensions.
@@
 It was also shown in~\cite{jg08} that GADTs can be represented using
-an object-level {\em left} Kan extension construct. This suggests
-extending our term calculus with such object-level constructs.  It was
-shown in~\cite{jp19} that augmenting the types in our calculus with a
-carefully designed left Kan extension preserves the cocontinuity
-needed for GADTs to have well-defined interpretations --- but not
-necessarily in
+object-level {\em left} Kan extensions.
+%
+%This suggests
+%extending our term calculus with such object-level constructs.
+%
+%{\color{blue} We may need to impose restrictions inspired
+%  by those for $\mu$-types on LKEs.}
+Augmenting the types in our calculus with a carefully designed left
+Kan extension was shown in~\cite{jp19} to preserve the cocontinuity
+needed for GADTs to have well-defined interpretations --- although
+%not necessarily in
 %However, this will not, in general, be
-%the case in
-locally finitely presentable catgories. Indeed, as shown
-in~\cite{jp19}, local $\lambda$-presentability for $\lambda >
-\aleph_1$ is required to interpret even commonly used GADTs. This
-suggests
+%the case in locally finitely presentable catgories. Indeed, as shown
+%in~\cite{jp19},
+local $\lambda$-presentabile categories for $\lambda > \aleph_1$ are
+required to interpret even commonly used GADTs. This suggests
+extending our term calculus with Kan extensions and
 %augmenting our type formation rules with such a construct.
@@
 result in a framework for constructing parametric models for calculi
-supporting nested types built from primitives that is based on
-``locally $\lambda$-presentable fibrations'', for some appropriate
-definition thereof.
-
-Another possibility is to add term-level fixpoints
+with nested types built from primitives that is based on ``locally
+$\lambda$-presentable fibrations'', for some appropriate definition
+thereof.  Another option is to add term-level fixpoints
 %, as in~\cite{pit00},
 to the calculus presented here. However, this would require the
-categories interpreting types to be not just locally finitely
-presentable categories, but to support some kind of domain structure
-as well. Unfortunately, $\omega$-CPO, the most obvious category of
-domains to replace $\set$, is not locally finitely presentable.
-
-\vspace*{0.1in}
+categories interpreting types to be not just locally
+$\lambda$-presentable, but to support some kind of domain structure as
+well.
+%Unfortunately, $\omega$-CPO, the most obvious category
+%of domains to replace $\set$, is not locally finitely presentable.
+
+\pagebreak
 

### 2020-06-26 free-theorems-nesteds/paper5.tex
final tweaks to paper5.tex; ready for final proofread

@@
 model for this calculus. This is both delicate and challenging.  To
-establish a suitable Identity Extension Lemma for our calculus, mild
-restrictions are imposed on the formation of fixpoints; all ADTs and
-all (truly) nested types appearing in the literature can nevertheless
-be expressed. Moreover, to ensure the existence of the fixpoints
-interpreting nested types, cocontinuity conditions on the functors
-underlying these fixpoints must be threaded throughout our model
-construction.  Overall, this paper offers a blueprint for deriving
-principled parametricity results for practical programming languages,
-such as Haskell and Agda, that directly construct nested types as
-fixpoints rather than encoding them. It also shows how parametric
-models can derive useful free theorems for programs over nested
-types. Like their counterparts for ADTs, these free theorems
+establish a suitable Identity Extension Lemma for our calculus, extra
+care is taken in the formation of fixpoints; all ADTs and all (truly)
+nested types appearing in the literature can nevertheless be expressed
+in our calculus. In addition, to ensure the existence of the fixpoints
+interpreting nested types, cocontinuity conditions on their underlying
+functors must be threaded throughout our model construction.  Overall,
+this paper offers a blueprint for deriving principled parametricity
+results for practical programming languages, such as Haskell and Agda,
+that directly construct nested types as fixpoints rather than
+representing them via Church encodings.
+%encoding them.
+It also shows how parametric models can derive useful free
+theorems for programs over nested types. Like their counterparts for
+ADTs, these free theorems
 %our model derives
@@
 %, in particular,
-a principled alternative to proving
-properties of programs over nested types via induction, 
-e.g., using the deep
-induction rules
+a principled alternative to proving program properties via induction,
+e.g., using the deep induction rules for nested types
 %for nested types
@@
 not, according to whether $\alpha$ is in the functorial or
-non-functorial context in which it is formed. For example, if
+non-functorial context in which it is formed. In particular, if
 $\emptyset;\alpha \vdash \mathit{List} \,\alpha$, then $\vdash
 \Nat^\alpha \onet\, (\mathit{List}\,\alpha)$ is well-formed; if
-$\alpha;\emptyset \vdash \List \,\alpha$, then it is not.  If
-$\mathit{Tree}\,\alpha\,\gamma = \mu \,\beta.\, \alpha + \beta \times
-\gamma \times \beta$, then $\gamma;\emptyset \vdash \Nat^\alpha
+$\alpha;\emptyset \vdash \List \,\alpha$, then it is not.  Similarly,
+if $\mathit{Tree}\,\alpha\,\gamma = \mu \,\beta.\, \alpha + \beta
+\times \gamma \times \beta$, then $\gamma;\emptyset \vdash \Nat^\alpha
 (\mathit{List}\, \alpha) \, (\mathit{Tree} \, \alpha\, \gamma)$ is
@@
 \phi \beta$ can be functorial or not in $\alpha$, but cannot be
-functorial in $\phi$. In particular, $\mathit{GRose}\,\phi\,\alpha$
-cannot be the (co)domain of a $\Nat$ type representing a natural
-transformation functorial in $\phi$, and the type $\mu \phi. \lambda
-\alpha. \mathit{GRose}\,\phi\,\alpha$ is not well-formed in our
-system. The restrictions imposed by our typing rules will ensure that
-set interpretations of types are functors. In
-Section~\ref{sec:conclusion} we discuss how to lessen the expressivity
-constraints they impose.
+functorial in $\phi$. Moreover, $\mathit{GRose}\,\phi\,\alpha$ cannot
+be the (co)domain of a $\Nat$ type representing a natural
+transformation functorial in $\phi$. Finally, the type $\mu
+\phi. \lambda \alpha. \mathit{GRose}\,\phi\,\alpha$ is not well-formed
+in our system. These restrictions ensure that the set and relational
+interpretations of types in Section~\ref{sec:type-interp} are
+functors. In Section~\ref{sec:conclusion} we discuss how GADTs and
+other more expressive types might be accommodated.
+%consider ways to lessen the expressivity constraints they impose.
+%
 %These restrictions on higher-arity functoriality are necessary
@@
 $x$ in $t$. In the rule for $t_{\ol K} s$ there is one functorial
-expression $K$ for every functorial variable $\alpha$. In the rule
-for $\map^{\ol{F},\ol{G}}_H$ there is one functorial expression $F$
-and one functorial expression $G$ for each functorial variable in
-$\ol\phi$. Moreover, for each $\phi^k \in \ol\phi$ the number of
-functorial variables $\beta$ in the judgments for its corresponding
-functorial expresssions $F$ and $G$ is $k$. In the rules for $\tin_H$
-and $\fold^F_H$, the functorial variables in $\ol{\beta}$ are fresh
-with respect to $H$, and there is one $\beta$ for every
+expression $K$ for every functorial variable $\alpha$. In the rule for
+$\map^{\ol{F},\ol{G}}_H$ there is one functorial expression $F$ and
+one functorial expression $G$ for each functorial variable in
+$\ol\phi$. Moreover, for each $\phi^k$ in $\ol\phi$ the number of
+functorial variables in $\ol\beta$ in the judgments for its
+corresponding functorial expresssions $F$ and $G$ is $k$. In the rules
+for $\tin_H$ and $\fold^F_H$, the functorial variables in $\ol{\beta}$
+are fresh with respect to $H$, and there is one $\beta$ for every
 $\alpha$. (Recall from above that, in order for the types of $\tin_H$
@@
 our calculus cannot express types of recursive functions, such as a
-concatenation function for perfect trees, that take as inputs both a
-nested type and an argument of another type, both of which are
-parameterized over the same variable $\alpha$.
+concatenation function for perfect trees, that take as inputs a nested
+type and an argument of another type, both of which are parameterized
+over the same variable $\alpha$.
 %not just one nested type parameterized over a variable $\alpha$ as
@@
 %argument.
-Even for ADTs there is a difference between which folds over them we
+In fact, even for ADTs there is a difference between which folds over them we
 can type when ADTs are viewed as ADTs --- i.e., as fixpoints of types
@@
 \beta)))$, all of whose argument types are parameterized over the same
-type $\alpha$. This is not possible even for nested types that are
+type $\alpha$. This is not possible for nested types, even if they are
 semantically equivalent to ADTs.
@@
 the type $\emptyset; \gamma \vdash \Nat^\emptyset (\onet + \onet)\,
-\gamma$ does not. We would therefore have to construct our models
-from semantic categories that are locally $\lambda$-presentable for
-$\lambda > \omega$.
+\gamma$ does not. To remedy this, we would need to construct our
+models from semantic categories that are locally $\lambda$-presentable
+for $\lambda > \omega$.
 %It is tempting to try to recover typeability of, say, a zip function
@@
 %
-Adding more expressive stylized recursion combinators could
-also help,
+Adding more expressive recursion combinators could help, too,
 %enhance the expressivity of our calculus
@@
 languages that construct nested types as fixpoints studied here.
-Enhancing the expressivity of our calculus is discussed further in
+%Nevertheless, improving expressivity is
+These and other extensions are discussed in
 Section~\ref{sec:conclusion}.
@@
 %rather than representing them via their Church encodings. We
-We have also used the Abstraction Theorem for this model to prove free
-theorems in the presence of nested types. This was not possible
-before~\cite{jp19} because the nested types definable in our syntax
-were not previously known to have well-defined interpretations in
-locally finitely presentable categories like $\set$ and $\rel$. No
-term calculus for terms of these types existed, either. The key to
-obtaining our parametric model is
+We have also used the Abstraction Theorem for this model to derive
+free theorems in the presence of nested types. This was not possible
+before~\cite{jp19} because the nested types in our calculus were not
+previously known to have well-defined interpretations in locally
+finitely presentable categories like $\set$ and $\rel$. No calculus
+for terms of these types existed, either. The key to obtaining our
+parametric model is
 %{\color{blue} adjusting our type syntax to respect the accessibility
@@
 the delicate threading of functoriality and its accompanying
-naturality conditions throughout the model construction
+naturality conditions throughout the model construction.
 
 Even when nested types have well-defined interpretations in a semantic
-category, the types of standard functions over them may not be.  All
-recursion in our calculus must be coded using folds.  Since standard
-folds for nested types can only express computations that return
-natural transformations, their expressivity has long been a vexing
-issue, which our calculus naturally inherits.
+category, the types of standard functions over them may not. This is
+because all recursion in our calculus must be coded using standard
+folds.  Since folds for nested types can only express computations
+that return natural transformations, their expressivity has long been
+a vexing issue, and this is naturally inherited by our calculus.
 %indeed, such folds
@@
 %$\set$/$\rel$-based parametric model
-calculus is not easily improved there. Construction of a
-$\set$/$\rel$-based parametric model for our calculus is nevertheless
-worthwhile.  For one thing, the expressivity issue is orthogonal to
-constructing parametric models,
+calculus is, unfortunately, not easily improved there. Construction of
+a $\set$/$\rel$-based parametric model for our calculus is
+nevertheless worthwhile.  For one thing, the expressivity issue is
+orthogonal to constructing parametric models,
 %in the presence of nested types built from primitives,
 and trying to address both in a single paper would
-detract from the conceptual clarity of our construction.
+detract from the conceptual clarity of our model construction.
 %our current focus on deriving principled parametricity
@@
 %conceptual clarity of this paper.
-For another, if the construction of this paper
+For another, if the construction in this paper
 %we couldn't
@@
 %presented here,
-weren't possible then trying to construct a parametric model from
-robust semantic categories for more data types and/or recursion
-schemes would likely be futile. We naturally expect the construction
-reported here to guide its generalization to the locally
-$\lambda$-presentable categories for $\lambda > \omega$ that will be
-needed to accommodate more general data types and more general
-recursion over them.
+weren't possible, then trying to construct a parametric model from
+richer semantic categories for fancier data types and/or recursion
+schemes would be futile. We naturally expect the construction reported
+here to guide its generalization to the locally $\lambda$-presentable
+categories for $\lambda > \omega$ that will be needed to accommodate
+more general data types and recursion schemes for them.
 
@@
 It was also shown in~\cite{jg08} that GADTs can be represented using
-object-level {\em left} Kan extensions.
+object-level {\em left} Kan extensions, and
 %
@@
 %  by those for $\mu$-types on LKEs.}
-Augmenting the types in our calculus with a carefully designed left
-Kan extension was shown in~\cite{jp19} to preserve the cocontinuity
-needed for GADTs to have well-defined interpretations --- although
+adding a carefully designed left Kan extension construct to our
+calculus was shown in~\cite{jp19} to preserve the cocontinuity needed
+for GADTs to have well-defined interpretations. We note, however, that
 %not necessarily in
@@
 %in~\cite{jp19},
-local $\lambda$-presentabile categories for $\lambda > \aleph_1$ are
+locally $\lambda$-presentable categories for $\lambda > \aleph_1$ are
 required to interpret even commonly used GADTs. This suggests
-extending our term calculus with Kan extensions and
+extending our calculus with Kan extension constructs and
 %augmenting our type formation rules with such a construct.
@@
 cartesian closed category (lpccc) $\cal C$ whose category of
-(abstract) relations, obtained by pullback as in~\cite{jac99}, is also
-a lpccc and is appropriately fibred over $\cal C$.  This would
+(abstract) relations, obtained by pullback as in~\cite{jac99},
+is also a lpccc and appropriately fibred over $\cal C$.  This would
 %extend the results presented here along the lines of~\cite{gjfor15},
 %thus giving
-result in a framework for constructing parametric models for calculi
+give a framework for constructing parametric models for calculi
 with nested types built from primitives that is based on ``locally
@@
 %, as in~\cite{pit00},
-to the calculus presented here. However, this would require the
+to the calculus presented here. This would, however, require the
 categories interpreting types to be not just locally

### 2020-06-29 free-theorems-nesteds/paper5.tex
slight revisions on paper5.tex

@@
 
+\newcommand{\ininv}[2]{(\tin^{-1}_{\onet +
+  \beta \times \phi (\phi\beta)})_{#1}\, #2}
 \newcommand{\emptyfun}{{[]}}
@@
 establish a suitable Identity Extension Lemma for our calculus, extra
-care is taken in the formation of fixpoints; all ADTs and all (truly)
-nested types appearing in the literature can nevertheless be expressed
-in our calculus. In addition, to ensure the existence of the fixpoints
-interpreting nested types, cocontinuity conditions on their underlying
-functors must be threaded throughout our model construction.  Overall,
-this paper offers a blueprint for deriving principled parametricity
-results for practical programming languages, such as Haskell and Agda,
-that directly construct nested types as fixpoints rather than
-representing them via Church encodings.
+care must be taken in the formation of fixpoint types; all ADTs and
+all (truly) nested types appearing in the literature can nevertheless
+be expressed in our calculus. In addition, to ensure the existence of
+semantic fixpoints interpreting nested types, cocontinuity conditions
+on the functors interpreting all of the types in our calculus must
+be threaded throughout our model construction.  Overall, this paper
+provides a blueprint for deriving principled parametricity results for
+practical programming languages, such as Haskell and Agda, that
+directly construct nested types as fixpoints rather than representing
+them via Church encodings.
 %encoding them.
@@
 a principled alternative to proving program properties via induction,
-e.g., using the deep induction rules for nested types
-%for nested types
-recently developed by
-Johann and Polonsky.
-%This is illustrated with several examples.
+e.g., using the deep induction rules for nested types recently
+developed by Johann and Polonsky.
+%We derive a number of free theorems for nested types.
+%To illustrate, we derive a number of free theorems.
+Our approach is illustrated with several examples. 
 \end{abstract}
@@
 over lists to according to the inductive structure of the input
-list. Such data-driven induction proofs over ADTs are so routine that
-they are often included in, say, undergraduate functional programming
-courses.
-
-An alternative technique for proving results like the above
+list. Such data-driven induction proofs over ADTs are routine enough
+to be included in, say, an undergraduate functional programming
+course.
+
+An alternative approach to proving results like the above
 $\mathtt{map}$-$\mathtt{reverse}$ property for lists is to use
@@
 text of the programs involved. To get interesting free theorems,
-Wadler's calculus included, implicitly, built-in list types; indeed,
-most of the free theorems in~\cite{wad89} are consequences of
-naturality for polymorphic list-processing functions. However,
-parametricity can also be used to prove naturality properties for
-non-list ADTs, as well as results, like correctness of program
-optimizations like {\em short cut fusion}~\cite{glp93,joh02,joh03},
+Wadler's calculus included built-in ADTs; indeed, most of the free
+theorems in~\cite{wad89} are consequences of naturality for
+polymorphic list-processing functions. However, the techniques
+of~\cite{wad89} show that parametricity can also be used to derive 
+naturality properties for non-list ADTs, as well as results, like
+correctness of program optimizations like {\em short cut
+  fusion}~\cite{glp93},
+%,joh02,joh03},
 that go beyond simple naturality.
@@
 This paper is about parametricity and free theorems for a polymorphic
-calculus with explicit syntax not just for ADTs, but for nested types
-as well. An ADT defines a {\em family of inductive data types}, one
-for each input type. For example, the $\mathtt{List}$ data type
-definition above defines a collection of data types $\mathtt{List\;
-  A}$, $\mathtt{List\; B}$, $\mathtt{List\; (A \times B)}$,
-$\mathtt{List\; (\List\;A)}$, etc., each independent of all the
-others. By contrast, a nested type~\cite{bm98} is an {\em inductive
-  family of data types} that is defined over, or is defined mutually
-recursively with, (other) such data types. Since the structures of the
-data type at one type can depend on those at other types, the entire
-family of types must be defined at once. Examples of nested types
-include, trivially, ordinary ADTs, such as list and tree types; simple
-nested types, such as the data type {\small
+calculus with explicit syntax not just for ADTs, but for the
+generalization of ADTs known as {\em nested types}~\cite{bm98}, as
+well. An ADT defines a {\em family of inductive data types}, one for
+each input type. For example, the $\mathtt{List}$ data type definition
+above defines a collection of data types $\mathtt{List\; A}$,
+$\mathtt{List\; B}$, $\mathtt{List\; (A \times B)}$, $\mathtt{List\;
+  (\List\;A)}$, etc., each independent of all the others. By contrast,
+a nested type is an {\em inductive family of data types} that is
+defined over, or is defined mutually recursively with, (other) such
+data types. Since the structures of the data type at one input type
+can depend on those at other input types, the entire family of types
+must be defined at once. Examples of nested types include, trivially,
+ordinary ADTs, such as list and tree types; simple nested types, such
+as the data type {\small
 \[\begin{array}{l}
@@
 provers actually generates an induction rule for perfect trees, and
-the rule it generats gives the expected result because proving
+the rule it generates gives the expected result because proving
 properties of perfect trees requires only that we induct over the
-top-level perfect tree in the recursive position, leaving any data
-internal to the input tree untouched.
+top-level perfect tree in the recursive position. That is, any data
+internal to the input tree is left untouched.
 
@@
 \mathtt{foldBush\; bn\; bc\; bnil = bn}\\
-\mathtt{foldBush \;bn \;bc \;(bcons\; x \;bb) =}\\
+\mathtt{foldBush \;bn \;bc \;(bcons\; x \;bbx) =}\\
 \hspace*{0.5in} \mathtt{bc\; x \;(foldBush \;bn\; bc
-  \;(mapBush \;(foldBush\; bn\; bc) \;bb))}\\
+  \;(mapBush \;(foldBush\; bn\; bc) \;bbx))}\\
 \\
@@
 \mathtt{mapBush\; \_ \; bnil = bnil}\\
-\mathtt{mapBush\; f\; (bcons\; x \;bb) = bcons \;(f\, x) \; (mapBush\;
-  (mapBush\, f) \; bb)}\\ 
+\mathtt{mapBush\; f\; (bcons\; x \;bbx) = bcons \;(f\, x) \; (mapBush\;
+  (mapBush\, f) \; bbx)}\\ 
 \\
@@
 \mathtt{balg \;x\; (bcons \;bnil \;bbbx) = bcons\; x \;(bcons \;bnil\; bbbx)}\\
-\mathtt{balg\; x \;(bcons \;(bcons\; y \;bx)\; bbbx) =}\\
-\hspace*{0.5in} \mathtt{bcons\; y\; (bcons\; (bcons \;x \;bx) \;bbbx)}\\
+\mathtt{balg\; x \;(bcons \;(bcons\; y \;bbx)\; bbbx) =}\\
+\hspace*{0.5in} \mathtt{bcons\; y\; (bcons\; (bcons \;x \;bbx) \;bbbx)}\\
 \end{array}\]
@@
 properties of programs over (truly) nested types by induction.  But
-recalling that inductive proofs over ADTs can sometimes be
-circumvented in the presence of parametricity, we might naturally ask:
+recalling that parametricity often provides a viable alternative to
+inductive proofs for ADTs,
+%inductive proofs over ADTs can sometimes be circumvented in the
+%presence of parametricity,
+we might naturally ask:
 \begin{verse}
@@
 %(including ADTs)
-directly via
-%type-level
-recursion --- rather than representing them indirectly by Church
-encodings as in most polymorphic calculi.
+directly via type-level recursion --- rather than representing them
+indirectly by Church encodings as in, e.g., System F.
 
@@
 %{\color{blue} locally finitely presentable}
-for the types from Section~\ref{sec:calculus}. As is usual when
-modeling parametricity, types are interpreted as functors from
-environments interpreting their type variable contexts to set or
-relations, as appropriate. But in order to ensure that these functors
-satisfy the cocontinuity properties needed to know that the fixpoints
-interpreting $\mu$-types exist, set environments must map each $k$-ary
-type constructor variable to an appropriately cocontinuous $k$-ary
-functor on sets and relation environments must map each $k$-ary type
-constructor variable to an appropriately cocontinuous $k$-ary relation
-transformer, and these cocontinuity conditions must be threaded
-throughout the type interpretations in such a way that the resulting
-model is guaranteed to satisfy an appropriate Identity Extension Lemma
-(Theorem~\ref{thm:iel}).
+for these types. As is standard when modeling parametricity, types are
+interpreted as functors from environments interpreting their type
+variable contexts to set or relations, as appropriate. But to ensure
+that these functors satisfy the cocontinuity properties needed for the
+semantic fixpoints interpreting $\mu$-types to exist, set environments
+must map each $k$-ary type constructor variable to an appropriately
+cocontinuous $k$-ary functor on sets, relation environments must map
+each $k$-ary type constructor variable to an appropriately
+cocontinuous $k$-ary relation transformer, and these cocontinuity
+conditions must be threaded throughout the type interpretations in
+such a way that the resulting model is guaranteed to satisfy an
+appropriate Identity Extension Lemma (Theorem~\ref{thm:iel}).
 %{\color{blue} Syntactic restrictions on
 %  $\mu$-types are essential to proving IEL.}
-Properly progagating the cocontinuity conditions turns out to be both
+Properly propagating the cocontinuity conditions turns out to be both
 subtle and challenging, and Section~\ref{sec:iel}, where it is done,
@@
 functions that are polymorphic over nested types. This scheme is very
-general, is parameterized over both the data type and the polymorphic
-function at hand, and has each of the above
+general, is parameterized over both the data type and the type of the
+polymorphic function at hand, and has each of the above
 $\mathtt{map}$-$\mathtt{reverse}$ theorems as instances.  The
@@
 transformations allows us to clearly delineate those free theorems
-that are consequences of naturality, and thus hold even in
+that are consequences of naturality, and may thus hold even in
 non-parametric models, from those that require the full power of
@@
 satisfies an Abstraction Theorem (Theorem~\ref{thm:abstraction}), and
-we derive several of this latter kind of free theorem from it in
+we derive several of this latter kind of free theorems from it in
 Section~\ref{sec:ftnt}.  Specifically, we state and prove
@@
 primitives rather than by Church encodings. This requires that we
-modify the type calculus to ensure that
+%modify the type calculus to ensure that
 %System F's full impredicative polymorphism interacts reasonably with
 %functoriality,
-functoriality is guaranteed syntactically, and that functoriality is
-reflected in the standard model construction so that the existence of
-the fixpoints by which nested types are to be interpreted is ensured.
+design our calculus in such a way that its types all have
+interpretations as appropriately cocontinuous functors, so that the
+existence of the fixpoints by which nested types are to be interpreted
+is ensured.
+%
+%functoriality is guaranteed syntactically, and that it is reflected in
+%our model construction so that the existence of the fixpoints by which
+%nested types are to be interpreted is ensured.
+%
 %parametricity to primitive nested types {\em as the model is being
@@
 be the (co)domain of a $\Nat$ type representing a natural
-transformation functorial in $\phi$. Finally, the type $\mu
-\phi. \lambda \alpha. \mathit{GRose}\,\phi\,\alpha$ is not well-formed
-in our system. These restrictions ensure that the set and relational
+transformation functorial in $\phi$, and the type $\mu \phi. \lambda
+\alpha. \mathit{GRose}\,\phi\,\alpha$ is not well-formed in our
+system. These restrictions ensure that the set and relational
 interpretations of types in Section~\ref{sec:type-interp} are
-functors. In Section~\ref{sec:conclusion} we discuss how GADTs and
-other more expressive types might be accommodated.
+functors. In Section~\ref{sec:conclusion} we discuss how
+%GADTs and other
+more expressive types might be accommodated.
 %consider ways to lessen the expressivity constraints they impose.
@@
   \psi]$ is the textual replacement of $\phi$ in $F$, meaning that all
-occurences of $\phi\ol\sigma$ in $F$ become $\psi\ol\sigma$.
+occurrences of $\phi\ol\sigma$ in $F$ become $\psi\ol\sigma$.
 \end{lemma}
@@
   :=_{\ol{\alpha}} F]$. Here, the operation $(\cdot)[\phi :=_{\ol \alpha}
-  F]$ of {\em second-order type substitution (along $\ol\alpha$)} is
+  F]$ of {\em second-order type substitution along $\ol\alpha$} is
 defined by: 
@@
 \\\\
+\AXC{$\ol{\Gamma;\Phi \vdash K}$}
 \AXC{$\Gamma; \emptyset
   \,|\, \Delta \vdash t : \Nat^{\ol{\alpha}} \,F \,G$}
-\AXC{$\ol{\Gamma;\Phi \vdash K}$}
 \AXC{$\Gamma;\Phi \,|\, \Delta \vdash s: F[\overline{\alpha := K}]$}
@@
 $x$ in $t$. In the rule for $t_{\ol K} s$ there is one functorial
-expression $K$ for every functorial variable $\alpha$. In the rule for
-$\map^{\ol{F},\ol{G}}_H$ there is one functorial expression $F$ and
-one functorial expression $G$ for each functorial variable in
-$\ol\phi$. Moreover, for each $\phi^k$ in $\ol\phi$ the number of
-functorial variables in $\ol\beta$ in the judgments for its
+expression in $\ol K$ for every functorial variable in $\ol \alpha$. In
+the rule for $\map^{\ol{F},\ol{G}}_H$ there is one functorial
+expression $F$ and one functorial expression $G$ for each functorial
+variable in $\ol\phi$. Moreover, for each $\phi^k$ in $\ol\phi$ the
+number of functorial variables in $\ol\beta$ in the judgments for its
 corresponding functorial expresssions $F$ and $G$ is $k$. In the rules
@@
 function from Figure~\ref{fig:funs2} as
-\[\vdash \fold_{\onet + \beta \times
-\phi (\phi\beta)}^{\mathit{Bush}\,\alpha}\, ( \tin_{\onet + \beta \times
-  \phi (\phi\beta)} \circ (\onet + t \circ i \circ i') ) :
-\Nat^{\alpha} (\mathit{Bush}\,\alpha)\,(\mathit{Bush}\,\alpha)\] 
+\[\vdash \fold^{\mathit{Bush}\,\alpha}_{\onet + \beta \times \phi (\phi \beta)}\, balg
+: \Nat^\alpha (\mathit{Bush}\,\alpha)\,(\mathit{Bush}\,\alpha)\]
+%\[\vdash \fold_{\onet + \beta \times
+%\phi (\phi\beta)}^{\mathit{Bush}\,\alpha}\, ( \tin_{\onet + \beta \times
+%  \phi (\phi\beta)} \circ (\onet + t \circ i \circ i') ) :
+%\Nat^{\alpha} (\mathit{Bush}\,\alpha)\,(\mathit{Bush}\,\alpha)\] 
 where
@@
 (\mathit{Bush}\,\alpha))\\
-\vdash \tin_{\onet + \beta \times \phi (\phi\beta)} : \Nat^{\alpha}\,
-(\onet + \alpha \times \mathit{Bush}\, (\mathit{Bush} \, \alpha))
-\;(\mathit{Bush}\, \alpha)\\
 \end{array}\]
 and $\mathit{bnil}$, $\mathit{bcons}$, $\tin^{-1}_{\onet + \beta
-  \times \phi (\phi\beta)}$, $t$, $i$, and 
-$i'$ are the terms
+  \times \phi (\phi\beta)}$, $\mathit{balg}$, and $\mathit{consalg}$
+are the terms
 \[\begin{array}{l}
@@
 L_{\alpha}\, x.\, \inr\, x ) \; : \; \Nat^{\alpha}\, (\alpha \times
-\mathit{Bush}\,(\mathit{Bush}\,\alpha))\; (\mathit{Bush}\,\alpha)\\
+\mathit{Bush}\,(\mathit{Bush}\,\alpha))\; (\mathit{Bush}\,\alpha)\\[0.05in]
 \vdash \fold_{\onet + \beta \times \phi (\phi\beta)}^{(\onet + \beta
@@
   \mathit{Bush}\,\alpha}\, \tin_{\onet + \beta \times \phi
-  (\phi\beta)})\\
+  (\phi\beta)})\\[0.05in]
 \hspace*{0.2in} : \Nat^{\alpha}\, (\mathit{Bush}\,\alpha)\; (\onet +
 \alpha \times \mathit{Bush}\,(\mathit{Bush}\,\alpha))\\ 
-\vdash L_{\alpha}\, (b, s). \,\mathsf{case}\;s\,\mathsf{of}\, \{ \hspace*{0.21in}
-        \ast \mapsto
-\mathit{bcons}_{\alpha}\, b\, (\mathit{bnil}_{\alpha}\, \ast); \\
-\hspace*{1.15in}\,(s', u) \mapsto \mathsf{case}\,{s'}\,\mathsf{of}\,
-\{\hspace*{0.3in} \ast
-\mapsto \mathit{bcons}_{\alpha}\, b \,(\mathit{bcons}_{\mathit{Bush}\,
-  \alpha}\, (\mathit{bnil}_{\alpha}\, \ast)\, u); \\
-\hspace*{2.3in}(b', u') \mapsto \mathit{bcons}_{\alpha}\, b'
-(\mathit{bcons}_{\mathit{Bush}\, \alpha}\, (\mathit{bcons}_{\alpha}\,
-b\, u )\, u')\}\}\\
-\hspace{0.2in} : \Nat^{\alpha} (\alpha \times (\onet + (\onet + \alpha
-\times \mathit{Bush}\, (\mathit{Bush} \, \alpha))) \times
-\mathit{Bush}\, (\mathit{Bush} \, (\mathit{Bush}\, \alpha))) \;
-(\alpha \times \mathit{Bush}\, (\mathit{Bush} \, \alpha))\\
-\vdash \alpha \times (\onet + \tin^{-1}_{\onet + \beta \times \phi
-  (\phi\beta)} \times \mathit{Bush}\, (\mathit{Bush} \,
-(\mathit{Bush}\, \alpha)))\\
-\hspace{0.2in} : \Nat^{\alpha} (\alpha \times (\onet + \mathit{Bush}
-\, \alpha \times \mathit{Bush}\,( \mathit{Bush} \, (\mathit{Bush}\,
-\alpha))))\\
-\hspace*{0.57in}(\alpha \times (\onet + (\onet + \alpha \times
-\mathit{Bush}\, (\mathit{Bush} \, \alpha)) \times \mathit{Bush}\,
-(\mathit{Bush} \, (\mathit{Bush}\, \alpha))))\\
-\vdash \alpha \times (L_{\alpha}\,x.\, (\tin^{-1}_{\onet +
-  \beta \times \phi (\phi\beta)})_{\mathit{Bush}\,\alpha}\, x)\\
-\hspace{0.2in} : \Nat^{\alpha} (\alpha
-\times \mathit{Bush} \, (\mathit{Bush}\, \alpha)) \; (\alpha \times (\onet
-+ \mathit{Bush} \, (\alpha \times \mathit{Bush}\, (\mathit{Bush} \,
-(\mathit{Bush}\, \alpha)))))
+%\vdash L_{\alpha}\, (b, s). \,\mathsf{case}\;s\,\mathsf{of}\, \{ \hspace*{0.21in}
+%        \ast \mapsto
+%\mathit{bcons}_{\alpha}\, b\, (\mathit{bnil}_{\alpha}\, \ast); \\
+%\hspace*{1.15in}\,(s', u) \mapsto \mathsf{case}\,{s'}\,\mathsf{of}\,
+%\{\hspace*{0.3in} \ast
+%\mapsto \mathit{bcons}_{\alpha}\, b \,(\mathit{bcons}_{\mathit{Bush}\,
+%  \alpha}\, (\mathit{bnil}_{\alpha}\, \ast)\, u); \\
+%\hspace*{2.3in}(b', u') \mapsto \mathit{bcons}_{\alpha}\, b'
+%(\mathit{bcons}_{\mathit{Bush}\, \alpha}\, (\mathit{bcons}_{\alpha}\,
+%b\, u )\, u')\}\}\\
+%\hspace{0.2in} : \Nat^{\alpha} (\alpha \times (\onet + (\onet + \alpha
+%\times \mathit{Bush}\, (\mathit{Bush} \, \alpha))) \times
+%\mathit{Bush}\, (\mathit{Bush} \, (\mathit{Bush}\, \alpha))) \;
+%(\alpha \times \mathit{Bush}\, (\mathit{Bush} \, \alpha))\\
+%\vdash \alpha \times (\onet + \tin^{-1}_{\onet + \beta \times \phi
+%  (\phi\beta)} \times \mathit{Bush}\, (\mathit{Bush} \,
+%(\mathit{Bush}\, \alpha)))\\
+%\hspace{0.2in} : \Nat^{\alpha} (\alpha \times (\onet + \mathit{Bush}
+%\, \alpha \times \mathit{Bush}\,( \mathit{Bush} \, (\mathit{Bush}\,
+%\alpha))))\\
+%\hspace*{0.57in}(\alpha \times (\onet + (\onet + \alpha \times
+%\mathit{Bush}\, (\mathit{Bush} \, \alpha)) \times \mathit{Bush}\,
+%(\mathit{Bush} \, (\mathit{Bush}\, \alpha))))\\
+%\vdash \alpha \times (L_{\alpha}\,x.\, (\tin^{-1}_{\onet +
+%  \beta \times \phi (\phi\beta)})_{\mathit{Bush}\,\alpha}\, x)\\
+%\hspace{0.2in} : \Nat^{\alpha} (\alpha
+%\times \mathit{Bush} \, (\mathit{Bush}\, \alpha)) \; (\alpha \times (\onet
+%+ \mathit{Bush} \, (\alpha \times \mathit{Bush}\, (\mathit{Bush} \,
+%(\mathit{Bush}\, \alpha)))))
+%\[\begin{array}{l}
+  \vdash L_{\alpha} \, s. 
+  \case{s}{* \mapsto \mathit{bnil}_\alpha *}{(a, bba) \mapsto consalg_\alpha (a, bba)}
+  : \Nat^\alpha \, (\onet + \alpha \times \mathit{Bush} (\mathit{Bush} \,\alpha)) \, (\mathit{Bush}\,\alpha) \\
+  \vdash L_{\alpha} \, (a, bba).  
+  \mathsf{case}\, (\ininv{\mathit{Bush}\,\alpha}{bba})\, \mathsf{of}\,\{
+  \\
+  \hspace{8em}\, * \mapsto \mathit{bcons}_{\alpha} (a,\mathit{bnil}_{\mathit{Bush}\,\alpha} *);  \\
+  \hspace{8em} (ba, bbba) \mapsto \mathsf{case}\, (\ininv{\alpha}{ba})
+  \, \mathsf{of}\,\{  \\ 
+  \hspace{16em}* \mapsto \mathit{bcons}_\alpha (a, \mathit{bcons}_{\mathit{Bush}\,\alpha} 
+        (\mathit{bnil}_\alpha *, bbba)) ;  \\ 
+  \hspace{16em}(a', bba') \mapsto \mathit{bcons}_\alpha (a', \mathit{bcons}_{\mathit{Bush}\,\alpha}
+  (\mathit{bcons}_\alpha (a, bba'), bbba)) \} \! \}\\    
+%  \quad \}  
+\hspace*{0.2in} : \Nat^\alpha \, (\alpha \times \mathit{Bush} (\mathit{Bush} \,\alpha)) \, (\mathit{Bush}\,\alpha) \\
 \end{array}\]
-respectively. Here, $\Gamma; \emptyset \,|\, \Delta \vdash H+
-\eta : \Nat^{\ol{\alpha}} (H+ F)\; (H+ G)$ and $\Gamma;
-\emptyset \,|\, \Delta \vdash H\times \eta : \Nat^{\ol{\alpha}}
-(H\times F)\; (H\times G)$ for $H+ \eta :=
-L_{\ol{\alpha}}\, x.\, \case{x}{s \mapsto s}{t
-  \mapsto  
-  (\eta_{\ol{\alpha}} t)}$ and $H\times \eta := L_{\ol{\alpha}}\,
-x.\, (\pi_1 x, \eta_{\ol{\alpha}} (\pi_2 x))$ for $\Gamma; \emptyset
-\,|\, \Delta \vdash \eta : \Nat^{\ol{\alpha}} F\; G$ and $\Gamma;
-\ol{\alpha} \vdash H$.
+respectively.
+%Here, $\Gamma; \emptyset \,|\, \Delta \vdash H+
+%\eta : \Nat^{\ol{\alpha}} (H+ F)\; (H+ G)$ and $\Gamma;
+%\emptyset \,|\, \Delta \vdash H\times \eta : \Nat^{\ol{\alpha}}
+%(H\times F)\; (H\times G)$ for $H+ \eta :=
+%L_{\ol{\alpha}}\, x.\, \case{x}{s \mapsto s}{t
+%  \mapsto  
+%  (\eta_{\ol{\alpha}} t)}$ and $H\times \eta := L_{\ol{\alpha}}\,
+%x.\, (\pi_1 x, \eta_{\ol{\alpha}} (\pi_2 x))$ for $\Gamma; \emptyset
+%\,|\, \Delta \vdash \eta : \Nat^{\ol{\alpha}} F\; G$ and $\Gamma;
+%\ol{\alpha} \vdash H$.
 
@@
 %input, but other inputs parameterized over $\alpha$ as well.
-Similarly, our calculus cannot express the type of a zip function for
-bushes. The fundamental issue is that recursion is expressible only
-via folds, and a fold over a nested type must take a natural
-transformation as its argument and return a natural transformation as
-its result.
+For the same reason, our calculus cannot express, e.g., the type of a
+zip function for bushes. The fundamental
+issue is that recursion is expressible only via folds, and a fold over
+a nested type must take a natural transformation as its argument and
+return a natural transformation as its result.
 %and, unfortunately, the polymorphism natural transformations entail
@@
 %argument.
-In fact, even for ADTs there is a difference between which folds over them we
-can type when ADTs are viewed as ADTs --- i.e., as fixpoints of types
-interpreted as first-order functors --- and which folds we can type
-when ADTs are viewed as proper nested types --- i.e., as fixpoints of
-types that are interpreted as higher-order functors --- because the
-arity of the recursion variable $\phi$ in the fixpoint in the return
-type of $\fold$ must equal the number of variables bound by its $\Nat$
-construct. For ADTs this number can be $0$, making it possible to
-write types such as $\alpha; \emptyset \vdash \Nat^\emptyset
-\ (\List\,\alpha)\,(\Nat^\emptyset \alpha\,(\List (\mu \beta. 1 +
-\beta)))$, all of whose argument types are parameterized over the same
-type $\alpha$. This is not possible for nested types, even if they are
-semantically equivalent to ADTs.
+In fact, even for ADTs there is a difference between which folds over
+them we can type when ADTs are viewed as ADTs --- i.e., as fixpoints
+of types interpreted as first-order functors --- and which folds we
+can type when ADTs are viewed as proper nested types --- i.e., as
+fixpoints of types that are interpreted as higher-order functors ---
+because the arity of the recursion variable $\phi$ in the fixpoint in
+the return type of $\fold$ must equal the number of variables bound by
+its $\Nat$ construct. For ADTs this number can be $0$, making it
+possible to write types such as $\alpha; \emptyset \vdash
+\Nat^\emptyset \ (\List\,\alpha)\,(\Nat^\emptyset \alpha\,(\List (\mu
+\beta. 1 + \beta)))$, all of whose argument types are parameterized
+over the same type $\alpha$, which are not possible for nested types
+even when they are semantically equivalent to ADTs.
 %Currying the type of the function does not help either, since the
@@
 %its first argument.
-Our calculus can, however, express types of functions
+Our calculus can, however, express types of recursive functions
 that take
@@
 the type $\emptyset; \gamma \vdash \Nat^\emptyset (\onet + \onet)\,
-\gamma$ does not. To remedy this, we would need to construct our
-models from semantic categories that are locally $\lambda$-presentable
-for $\lambda > \omega$.
+\gamma$ has no such semantics. To remedy this, we would need to
+construct our models from semantic categories that are locally
+$\lambda$-presentable for $\lambda > \omega$.
 %It is tempting to try to recover typeability of, say, a zip function
@@
 %
-Adding more expressive recursion combinators could help, too,
+Adding more expressive recursion combinators could help, too.
 %enhance the expressivity of our calculus
-but the expressivity issue for folds for nested types long pre-dates,
-and is somewhat orthogonal to, the issue of parametricity for
-languages that construct nested types as fixpoints studied here.
-%Nevertheless, improving expressivity is
-These and other extensions are discussed in
-Section~\ref{sec:conclusion}.
+But since the expressivity issue for folds for nested types long
+pre-dates, and is somewhat orthogonal to, the issue of parametricity
+for languages that construct nested types as fixpoints, we do not
+consider it further in this paper other than to identify, in
+Section~\ref{sec:conclusion}, some ideas for addressing it in future
+work.
+%These and other extensions are discussed
 
@@
 %related results.
-Similar results are expected to hold for types and terms with any
+Similar theorems are expected to hold for types and terms with any
 number of free variables.
@@
 (Theorem~\ref{thm:iel}) in our setting is a familiar ``cutting down''
-of the interpretations of universally quantified types, such as our
-$\Nat$-types, to include only the ``parametric'' elements.
+of the interpretations of universally quantified types to include only
+the ``parametric'' elements; the relevant types in our calculus are
+the $\Nat$-types.
 %(See, e.g.,~\cite{atk12,bfss90,gjfor15,rey83}).
-This requires that set interpretations of types are defined
-simultaneously with their relational interpretations. We give set
-interpretations for our types in Section~\ref{sec:set-interp} and give
-their relational interpretations in Section~\ref{sec:rel-interp}.
-While the set interpretations are relatively straightforward, their
-relation interpretations are less so, mainly because of the
-cocontinuity conditions we must impose to ensure that they are
-well-defined. We develop these conditions in
+This cutting down requires, as usual, that the set interpretations of
+our types are defined simultaneously with their relational
+interpretations. We give the set interpretations for our types in
+Section~\ref{sec:set-interp} and give their relational interpretations
+in Section~\ref{sec:rel-interp}.  While the set interpretations are
+relatively straightforward, their relation interpretations are less
+so, mainly because of the cocontinuity conditions required to ensure
+that they are well-defined. We develop these conditions in
 Section~\ref{sec:rel-interp}, which separates
@@
 there are only $\set$-many choices of morphisms between any two
-objects since $\set$ is locally small.  Note that $\setsem{\Gamma;
+objects since $\set$ is locally small.  We note that $\setsem{\Gamma;
   \emptyset \vdash \Nat^\alpha F\,G}$ is $\omega$-cocontinuous because
-it is a constant functor.
+it is a constant functor (and that allowing functorial variables not
+in $\ol \alpha$ in $G$ fails here, as shown in
+Section~\ref{sec:terms}).
 %follows from $\omega$-cocontinuity of $\setsem{\Gamma;\ol\alpha
@@
 $\omega$-cocontinuity of $\sem{\Gamma;\Phi \vdash F}$ in each case,
-together with Lemma~\ref{lem:fp} and Equation~\ref{eq:mu} in the
-$\mu$-case.
+together with Lemma~\ref{lem:fp} and Equation~\ref{eq:mu} for
+$\mu$-types.
 
@@
 As before, that $\sem{\Gamma; \Phi \vdash (\mu \phi. \lambda
-  \overline{\alpha}. H)\overline{\tau}}$ is $\omega$-concontinuous
+  \overline{\alpha}. H)\overline{\tau}}$ is $\omega$-cocontinuous
 follows from the facts that $\set$ and $\rel$ are locally finitely
@@
 {\em graph relation environment} $\graph{f}$ is defined pointwise by
-$\graph{f} \phi = \graph{f \phi}$ for every $\phi$. This entails that
+$\graph{f} \phi = \graph{f \phi}$ for every $\phi$, which entails that
 $\pi_1 \graph{f} = \rho$ and $\pi_2 \graph{f} = \rho'$. In particular,
 the {\em equality relation environment} $\Eq_\rho$ is defined to be
-$\graph{\id_{\rho}}$. This entails that $\Eq_\rho \phi = \Eq_{\rho
+$\graph{\id_{\rho}}$, which entails that $\Eq_\rho \phi = \Eq_{\rho
   \phi}$ for every $\phi$.
@@
 %\end{equation}
-hold. With these results in hand, the proof is straightforward. A
-detailed proof is given in the accompanying anonymous supplementary
+hold. With these results in hand, the proof follows easily. It is
+given in detail in the accompanying anonymous supplementary
 material. As noted there, if functorial variables of arity greater
-than $0$ were allowed to appear in the bodies of $\mu$-types, the IEL
-would fail.
+than $0$ were allowed to appear in the bodies of $\mu$-types, then the
+IEL would fail.
 
@@
 $\relsem{\Gamma; \Phi \vdash \Delta}\rho$ to $\relsem{\Gamma; \Phi
-  \vdash F}\rho$. These are given in the next (two) definitions.
+  \vdash F}\rho$. These are given in Definition~\ref{def:set-interp}.
 
@@
 %(\Nat^\emptyset\,(\mathit{Tree}\,\sigma\,\sigma')\,(\mathit{Tree}\,\tau\,\tau'))$$
-Finally, in Section~\ref{sec:calculus} we observed that $ \phi; \alpha
-\vdash \mathit{GRose}\,\phi\,\alpha$ cannot appear as the (co)domain
-of a $\Nat$ type binding both $\phi$ and $\alpha$. As a result, there
-are no instances of, say, a $\mathtt{map}$/$\mathtt{reverse}$ free
-theorem for this data type that maps replacements for both $\phi$ and
+We observed in Section~\ref{sec:calculus} that $ \phi; \alpha \vdash
+\mathit{GRose}\,\phi\,\alpha$ cannot appear as the (co)domain of a
+$\Nat$-type binding both $\phi$ and $\alpha$. As a result, there are
+no instances of, say, a $\mathtt{map}$/$\mathtt{reverse}$ free theorem
+for this data type that maps replacements for both $\phi$ and
 $\alpha$. On the other hand, $\phi; \alpha \vdash
@@
 
-We have the expected relationships
-between the interpetations of $\map$, $\tin$, and $\fold$:
+We have the expected properties of the interpretations of $\map$,
+$\tin$, and $\fold$, namely:
 
@@
 propagating the naturality conditions is quite involved; the latter
-two cases especially require some rather involved diagram
+two cases especially require some rather delicate diagram
 chasing. That this is possible provides strong evidence that our
@@
 = \emptyset$, then there can be no such morphism, so no such natural
-transformation can exist,
+transformation,
 %in $\set$,
-and thus no term $\vdash g : \Nat^\alpha \onet \,\alpha$ can exist.
+and thus no term $\vdash g : \Nat^\alpha \onet \,\alpha$, can exist.
 %in our calculus.
-That is, our calculus does not admit any terms with the closed type
-$\Nat^\alpha \onet \,\alpha$ of the polymorphic bottom.
+That is, our calculus admits no non-terminating terms, i.e., terms
+with the closed type $\Nat^\alpha \onet \,\alpha$ of the polymorphic
+bottom.
 
@@
 In addition, while there are many papers on higher-kinded types, we
-know of none that considers expressiblity issues arising from the
+know of none that considers expressibility issues arising from the
 local $\lambda$-presentability cardinal $\lambda$ of the categories in
@@
 %many papers on higher-kinded types, no other work that we know of
-%even considers expressiblity issues resulting from the semantic
+%even considers expressibility issues resulting from the semantic
 %settings in which they are interpreted.}
@@
 calculus was shown in~\cite{jp19} to preserve the cocontinuity needed
-for GADTs to have well-defined interpretations. We note, however, that
+for GADTs to have well-defined interpretations. (Note, however, that
 %not necessarily in
 %However, this will not, in general, be
-%the case in locally finitely presentable catgories. Indeed, as shown
+%the case in locally finitely presentable categories. Indeed, as shown
 %in~\cite{jp19},
 locally $\lambda$-presentable categories for $\lambda > \aleph_1$ are
-required to interpret even commonly used GADTs. This suggests
+required to interpret even commonly used GADTs.) This suggests
 extending our calculus with Kan extension constructs and
@@
 give a framework for constructing parametric models for calculi
-with nested types built from primitives that is based on ``locally
-$\lambda$-presentable fibrations'', for some appropriate definition
+with nested types built from primitives that is based on locally
+$\lambda$-presentable fibrations, for some appropriate definition
 thereof.  Another option is to add term-level fixpoints
 %, as in~\cite{pit00},
-to the calculus presented here. This would, however, require the
+to the calculus presented here. Of course, this would require the
 categories interpreting types to be not just locally

### 2020-06-30 free-theorems-nesteds/paper5.tex
small text changes to supp.tex
today's changes to paper5

@@
 \hspace*{0.4in}\mathtt{nil\;:\; List\;A}\\
-\hspace*{0.4in}\mathtt{Cons\;:\;A \rightarrow List\;A \rightarrow List\;A}
+\hspace*{0.4in}\mathtt{cons\;:\;A \rightarrow List\;A \rightarrow List\;A}
 \end{array}\]}
@@
 \begin{figure*}
-\hspace*{-0.5in}
+\hspace*{-0.7in}
 \resizebox{0.35\linewidth}{!}{
@@
 \mathtt{mapPTree : \forall \{A\, B : Set\} \rightarrow (A \rightarrow B)
-  \rightarrow PLeaves\; A \rightarrow PLeaves\; B}\\ 
+  \rightarrow PTree\; A \rightarrow PTree\; B}\\ 
 \mathtt{mapPTree \;f\; (pleaf \;x) = pleaf\; (f\; x)}\\
-\mathtt{mapPTree\; f \;(pnode\; p) = pnode\; (mapPTree \;(\lambda p
+\mathtt{mapPTree\; f \;(pnode\; p) =}\\
+\hspace*{0.7in}  \mathtt{pnode\; (mapPTree \;(\lambda p
   \rightarrow (f (\pi_1 \,p), f (\pi_2\,p)))\; p)}\\ 
@@
 \end{minipage}}
-\quad\quad\quad\quad\quad\quad\quad\quad
+\quad\quad\quad\quad\quad%\quad\quad\quad
 \resizebox{0.35\linewidth}{!}{
@@
 \\
-\mathtt{mapBush : \forall \{A \,B : Set\} \to (A \to B) \to (Bush\, A) \to
-  (Bush\, B)}\\ 
+\mathtt{mapBush : \forall \{A \,B : Set\} \to (A \to B) \to Bush\, A \to
+  Bush\, B}\\ 
 \mathtt{mapBush\; \_ \; bnil = bnil}\\
@@
   \mcF^\Pos(V) \; ::= \; %\T(V)
-  &\mid
+  & \hspace*{0.15in}
 \zerot \mid \onet 
 \mid \Nat^{\ol{\alpha}} \, \mcF^{\overline{\alpha}}(V) \;
-\mcF^{P,\ol{\alpha}}(V) %\mid V \ol{\T(V)}
+\mcF^{\ol{\alpha}}(V) %\mid V \ol{\T(V)}
 \mid \Pos\; \ol{\mcF^\Pos(V)}  \,
@@
   \tvars, \Pos\, \subseteq \fvars}\F^\Pos(V)$ of\, {\em well-formed
-  type expressions} are
+  types} are
 
@@
 not change the contexts in which types can be formed. If
-$\Gamma;\emptyset \vdash F$ and $\Gamma;\emptyset \vdash G$,
-then our rules allow formation of the type $\Gamma;\emptyset \vdash
-\Nat^\emptyset F \,G$. Since $\Gamma; \emptyset \vdash
-\Nat^{\ol \alpha}F\,G$ represents a natural transformation in
-$\ol \alpha$ from $F$ to $G$, the type $\Gamma;\emptyset
-\vdash \Nat^\emptyset F \,G$ represents the arrow type
-$\Gamma \vdash F \to G$ in our calculus. Our calculus
-similarly represents the $\forall$-type $\Gamma; \emptyset
-\vdash \forall \ol\alpha . F$ as $\Gamma; \emptyset \vdash
-\Nat^{\ol\alpha} \,\onet \,F$.
+$\Gamma;\emptyset \vdash F$ and $\Gamma;\emptyset \vdash G$, then our
+rules allow formation of the type $\Gamma;\emptyset \vdash
+\Nat^\emptyset F \,G$. Since $\Gamma; \emptyset \vdash \Nat^{\ol
+  \alpha}F\,G$ represents a natural transformation in $\ol \alpha$
+from $F$ to $G$, the type $\Gamma;\emptyset \vdash \Nat^\emptyset F
+\,G$ represents the arrow type $\Gamma \vdash F \to G$ in our
+calculus. The type $\Gamma; \emptyset \vdash \Nat^{\ol\alpha} \,\onet
+\,F$ similarly represents the $\forall$-type $\Gamma; \emptyset \vdash
+\forall \ol\alpha . F$.  Note, however, that
+%not even all
+%Hindley-Milner types of System F are representable in our calculus.
+not all System F types are 
+%only those System F types that can be interpreted as functors are
+representable in our calculus.
+%
 %However, if $\ol\alpha$ is non-empty then $\tau$ cannot be of the
@@
 for second-order types is defined below, where we adopt a similar
-notational convention for vectors of types. We note that it is not
+notational convention for vectors of types. Note that it is not
 correct to substitute along non-functorial variables.
@@
 %with $k \geq 1$,
-and if\, $\Gamma; \Psi,\ol{\alpha} \vdash F$ with $|\ol\alpha| = k$
-and $\Phi \cap \Psi = \emptyset$, then $\Gamma,\Psi;\Phi \vdash H[\phi
-  :=_{\ol{\alpha}} F]$. Here, the operation $(\cdot)[\phi :=_{\ol \alpha}
-  F]$ of {\em second-order type substitution along $\ol\alpha$} is
-defined by: 
+and if\, $\Gamma; \ol\psi,\ol{\alpha} \vdash F$ with $|\ol\alpha| = k$
+and $\Phi \cap \ol\psi = \emptyset$, then $\Gamma,\ol\psi';\Phi \vdash
+H[\phi :=_{\ol{\alpha}} F[\ol{\psi :== \psi'}]]$. Here, the operation
+  $(\cdot)[\phi :=_{\ol \alpha} F]$ of {\em second-order type
+    substitution along $\ol\alpha$} is defined by:
 \[\begin{array}{lll}
@@
     :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\; (\Nat^{\ol{\beta},
-    \ol{\gamma}}\,(\mu \phi.\lambda \ol{\alpha}.H)\ol{\beta})\,F)$}
+    \ol{\gamma}}\,(\mu \phi.\lambda \ol{\alpha}.H)\ol{\beta}\,F)$}
 \DisplayProof
@@
 calculus as
-\[\vdash \fold_{\beta + \phi(\beta \times \beta)}^{\mathit{PTree}\,
-  \alpha} \, (\tin_{\beta + \phi(\beta \times \beta)} \circ s) :
+\[\vdash (\fold_{\beta + \phi(\beta \times \beta)}^{\mathit{PTree}\,
+  \alpha})_\emptyset \,
+%(\tin_{\beta + \phi(\beta \times \beta)} \circ
+s :
 \Nat^{\alpha} (\mathit{PTree}\,\alpha)\,(\mathit{PTree}\,\alpha)\] 
@@
 and 
-$\mathit{swap}$ and $s$ are the terms
-\[ \vdash L_{\alpha} p.\, (\pi_2 p, \pi_1 p) :
-\Nat^{\alpha} (\alpha \times \alpha)\, (\alpha \times \alpha)\]
-and
+$\mathit{pleaf}$,$\mathit{pnode}$, $\mathit{swap}$, and $s$ are the terms
 \[\begin{array}{l}
+\vdash \tin_{\beta + \phi (\beta \times \beta)} \circ (
+L_{\alpha}\, x.\, \inl\, x) \; : \; \Nat^{\alpha}\, \alpha\,
+(\mathit{PTree}\,\alpha)\\ 
+\vdash \tin_{\beta + \phi (\beta \times \beta)} \circ (
+L_{\alpha}\, x.\, \inr\, x) \; : \; \Nat^{\alpha}\, \mathit{PTree}
+(\alpha \times \alpha)\,(\mathit{PTree}\,\alpha)\\ 
+\vdash L_{\alpha} p.\, (\pi_2 p, \pi_1 p) :
+\Nat^{\alpha} (\alpha \times \alpha)\, (\alpha \times \alpha)\\
 \vdash L_{\alpha} t. \,\case{t}{b \mapsto
-  b}{t' \mapsto  
-    \map_{\mathit{PTree}\,\alpha}^{\alpha \times \alpha, \alpha \times
-      \alpha}\, \mathit{swap}\, t'} )
+ \mathit{pleaf}\, b}{t' \mapsto \mathit{pnode}\, 
+    (((\map_{\mathit{PTree}\,\alpha}^{\alpha \times \alpha, \alpha \times
+      \alpha})_\emptyset\, \mathit{swap})_\alpha\, t')}
 :  \Nat^{\alpha} (\alpha + \mathit{PTree}\,(\alpha \times
-\alpha))\; (\alpha + \mathit{PTree}\,(\alpha \times \alpha))
+\alpha))\; \mathit{PTree}\,\alpha
 \end{array}\]
@@
 function from Figure~\ref{fig:funs2} as
-\[\vdash \fold^{\mathit{Bush}\,\alpha}_{\onet + \beta \times \phi (\phi \beta)}\, balg
+\[\vdash (\fold^{\mathit{Bush}\,\alpha}_{\onet + \beta \times \phi (\phi \beta)})_\emptyset\, balg
 : \Nat^\alpha (\mathit{Bush}\,\alpha)\,(\mathit{Bush}\,\alpha)\]
@@
 \mathit{Bush}\,(\mathit{Bush}\,\alpha))\; (\mathit{Bush}\,\alpha)\\[0.05in]
-\vdash \fold_{\onet + \beta \times \phi (\phi\beta)}^{(\onet + \beta
-  \times \phi (\phi\beta))[\phi := \mathit{Bush}\,\alpha]}\,
-(\map_{\onet + \beta \times \phi (\phi\beta)}^{(\onet + \beta \times
+\vdash (\fold_{\onet + \beta \times \phi (\phi\beta)}^{(\onet + \beta
+  \times \phi (\phi\beta))[\phi := \mathit{Bush}\,\alpha]})_\emptyset\,
+((\map_{\onet + \beta \times \phi (\phi\beta)}^{(\onet + \beta \times
   \phi (\phi\beta))[\phi := \mathit{Bush}\,\alpha][\beta := \alpha],
-  \mathit{Bush}\,\alpha}\, \tin_{\onet + \beta \times \phi
+  \mathit{Bush}\,\alpha})_\emptyset\, \tin_{\onet + \beta \times \phi
   (\phi\beta)})\\[0.05in]
@@
 %as shown in Example D, Section VI of~\cite{jp19},
-the type $\emptyset; \gamma \vdash \Nat^\emptyset (\onet + \onet)\,
+the type $\emptyset; \gamma \vdash \Nat^\emptyset\, (\mu \beta. \onet + \beta)\,
 \gamma$ has no such semantics. To remedy this, we would need to
@@
 be empty, but because the polymorphic $\mathit{flatten}$ function is
-functorial in some variable, say $\delta$, it cannot possibly have a
+natural in some variable, say $\delta$, it cannot possibly have a
 type of the form $\Nat^\emptyset F\, G$ that would be required for it
@@
 and~\ref{def:reln-env}. If $\C$ and $\D$ are locally finitely
-presentable categories, we write $[\C,\D]$ for the set of
+presentable categories, we write $[\C,\D]$ for the category of
 $\omega$-cocontinuous functors from $\C$ to $\D$.
@@
 \end{dfn}
-When convenient we identify a functor $F : [\set^0, \set]$ with the
-set that is its codomain and consider a set environment to map
-a type variable of arity $0$ to
+When convenient we identify a functor in $[\set^0, \set]$ with its
+value on $\ast$ and consider a set environment to map a type
+variable of arity $0$ to
 %an $\omega$-cocontinuous functor from $\set^0$ to $\set$, i.e., to
@@
 objects since $\set$ is locally small.  We note that $\setsem{\Gamma;
-  \emptyset \vdash \Nat^\alpha F\,G}$ is $\omega$-cocontinuous because
+  \emptyset \vdash \Nat^{\ol\alpha} F\,G}$ is $\omega$-cocontinuous because
 it is a constant functor (and that allowing functorial variables not
@@
 that $\setsem{\Gamma \vdash F \to G}$ and $\setsem{\Gamma \vdash
-  \forall \alpha. F}$ are as expected in any parametric model.
+  \forall \ol\alpha. F}$ are as expected in any parametric model.
 
@@
 \begin{dfn}\label{def:set-sem-funcs}
-Let $f: \rho \to \rho'$ for set environments $\rho$ and $\rho'$ (so
-that $\rho|_\tvars = \rho'|_\tvars$). The action $\setsem{\Gamma;\Phi
-  \vdash F}f$ of\, $\setsem{\Gamma;\Phi \vdash F}$ on the
-morphism $f$ is given as follows:
+Let $f: \rho \to \rho'$ be a morphism between set environments $\rho$
+and $\rho'$ (so that $\rho|_\tvars = \rho'|_\tvars$). The action
+$\setsem{\Gamma;\Phi \vdash F}f$ of\, $\setsem{\Gamma;\Phi \vdash F}$
+on the morphism $f$ is given as follows:
 \begin{itemize}
@@
       F}f}}\, \circ\, (f \phi)_{\ol{\setsem{\Gamma;\Phi \vdash
-      \F}\rho}}$.  The latter equality holds because $\rho\phi$ and
+      F}\rho}}$.  The latter equality holds because $\rho\phi$ and
   $\rho'\phi$ are functors and $f\phi : \rho\phi \to \rho'\phi$ is a
@@
 \begin{dfn}\label{def:reln-env}
-A {\em relation environment} maps each each type variable in $\tvars^k
-\cup \fvars^k$ to a $k$-ary relation transformer.  A morphism $f :
-\rho \to \rho'$ for relation environments $\rho$ and $\rho'$ with
+A {\em relation environment} maps each type variable in $\tvars^k \cup
+\fvars^k$ to a $k$-ary relation transformer.  A morphism $f : \rho \to
+\rho'$ between relation environments $\rho$ and $\rho'$ with
 $\rho|_\tvars = \rho'|_\tvars$ maps each type constructor variable
@@
 that $\relsem{\Gamma \vdash F \to G}$ and $\relsem{\Gamma \vdash
-  \forall \alpha. F}$ are as expected in any parametric model.
+  \forall \ol \alpha. F}$ are as expected in any parametric model.
 %$\relsem{\Gamma \vdash
@@
 As for set interpretations, $\relsem{\Gamma; \emptyset \vdash
-  \Nat^\alpha F\,G}$ is $\omega$-cocontinuous because it is a constant
+  \Nat^{\ol\alpha} F\,G}$ is $\omega$-cocontinuous because it is a constant
 functor.
@@
 of types interact well with demotion of functorial variables. Indeed,
-we have that, for any $\rho,\rho'$ and $f: \rho \to \rho'$ in
-$\setenv$, 
+we have that, if $\rho, \rho' : \setenv$, \,$f : \rho \to \rho'$,
+\,$\rho \phi = \rho \psi = \rho' \phi = \rho' \psi$, \, $f \phi = f
+\psi = \id_{\rho \phi}$,\, $\Gamma; \Phi, \phi^k \vdash F$,\,
+$\Gamma;\Phi,\ol{\alpha} \vdash G$,\, $\Gamma;\Phi,\alpha_1...\alpha_k
+\vdash H$, and $\ol{\Gamma;\Phi \vdash K}$, then
+%for any $\rho,\rho'$ and $f: \rho \to \rho'$ in
+%$\setenv$, 
 \begin{gather}
@@
 \setsem{\Gamma; \Phi, \phi \vdash F} \rho = \setsem{\Gamma, \psi; \Phi
-  \vdash F[\phi :== \psi] } \rho[\psi := \rho \phi]\\
+  \vdash F[\phi :== \psi] } \rho\\
 \label{thm:demotion-morphisms}
 \setsem{\Gamma; \Phi, \phi \vdash F} f = \setsem{\Gamma, \psi; \Phi
-  \vdash F[\phi :== \psi]} f[\psi \mapsto f\phi]
-\end{gather}
-Moreover,
+  \vdash F[\phi :== \psi]} f\\
+%\end{gather}
+%Moreover,
 %\begin{lemma}\label{lem:demotion-objects}
-if $\rho, \rho' : \setenv$ are such that $\rho \phi = \rho \psi =
-\rho' \phi = \rho' \psi$, \,$f : \rho \to \rho'$ in $\setenv$ is such
-that $f \phi = f \psi = \id_{\rho \phi}$,\, $\Gamma; \Phi, \phi^k
-\vdash F$,\, $\Gamma;\Phi,\ol{\alpha} \vdash G$,\,
-$\Gamma;\Phi,\alpha_1...\alpha_k \vdash H$, and $\ol{\Gamma;\Phi \vdash
-F}$, then
-\begin{gather}
+%\begin{gather}
 %\label{thm:demotion-objects}
@@
 \label{eq:subs-var}
-\setsem{\Gamma;\Phi \vdash G[\ol{\alpha := F}]}\rho =
+\setsem{\Gamma;\Phi \vdash G[\ol{\alpha := K}]}\rho =
 \setsem{\Gamma;\Phi,\ol{\alpha} \vdash G}\rho[\ol{\alpha := 
-\setsem{\Gamma;\Phi \vdash F}\rho}]\\
+\setsem{\Gamma;\Phi \vdash K}\rho}]\\
 \label{eq:subs-var-morph}
-\setsem{\Gamma;\Phi \vdash G[\ol{\alpha := F}]}f =
+\setsem{\Gamma;\Phi \vdash G[\ol{\alpha := K}]}f =
 \setsem{\Gamma;\Phi,\ol{\alpha} \vdash G}f[\ol{\alpha :=
-\setsem{\Gamma;\Phi \vdash F}f}]\\
+\setsem{\Gamma;\Phi \vdash K}f}]\\
 \label{eq:subs-const}
@@
 The {\em equality relation transformer} on $F : [\set^k,\set]$ is
-defined to be $\Eq_F = \graph{\id_{F}}$. Specifically, $Eq_F = (F, F,
+defined to be $\Eq_F = \graph{\id_{F}}$. Specifically, $\Eq_F = (F, F,
 \Eq_F^*)$ with $\Eq_F^* = \graph{\id_{F}}^*$, and
@@
 
-The return type for the semantic fold in the last clause is
-  $\dsem{\Gamma;\ol\beta,\ol\gamma \vdash F}\rho[\ol{\gamma := C}]$.
+The return type for the semantic fold is
+  $\dsem{\Gamma;\ol\beta,\ol\gamma \vdash F}\rho[\ol{\gamma :=
+    C}][\ol{\beta := B}]$.
   This interpretation gives $\dsem{\Gamma;\emptyset \,|\, \Delta
@@
 \item $\setsem{\Gamma;\Phi \,|\, \Delta[\alpha := G] \vdash
-  t[\alpha := G] : \tau[\alpha := G]}\rho =
+  t[\alpha := G] : F[\alpha := G]}\rho =
   \setsem{\Gamma,\alpha;\Phi \,|\, \Delta \vdash t : F }\rho [
@@
 \item $\setsem{\Gamma;\Phi \,|\, \Delta[\alpha := G] \vdash
-  t'[\alpha := G] : \tau[\alpha := G]}\rho =
+  t'[\alpha := G] : F[\alpha := G]}\rho =
   \setsem{\Gamma;\Phi,\alpha \,|\, \Delta \vdash t' : F }\rho [
@@
   \vdash 
-  (\map_G^{\overline{K}, \overline{H}} \,\overline{t})_{\ol\gamma} \circ
+  ((\map_G^{\overline{K}, \overline{H}})_\emptyset \,\overline{t}) \circ
 (L_{\overline{\gamma}} z. s_{\overline{K}, \overline{\gamma}}
@@
   s_{\overline{H}, \overline{\gamma}} z)\circ
-  (\map_F^{\overline{K}, \overline{H}} \,
-  \overline{t})_{\ol\gamma}  : \Nat^{\overline{\gamma}} F[\overline{\alpha :=
+  ((\map_F^{\overline{K}, \overline{H}})_\emptyset \,
+  \overline{t})  : \Nat^{\overline{\gamma}} F[\overline{\alpha :=
       K}]\, G[\overline{\alpha := H}]}
@@
 \hspace*{0.135in}\setsem{\Gamma; \emptyset \,|\, \Delta \vdash
-L_{\overline{\gamma}} z. (\map_{(\mu \phi. \lambda
+((\map_{(\mu \phi. \lambda
     \overline{\alpha}. H) \overline{\beta}}^{\overline{F},
-    \overline{G}}\, \overline{t})_{\ol\gamma}
+    \overline{G}})_\emptyset\, \overline{t}) \circ (L_{\overline{\gamma}} z.
   (\tin_{H})_{\overline{F},
@@
 = \setsem{\Gamma; \emptyset \,|\, \Delta
-\vdash L_{\overline{\gamma}} z. (\tin_H)_{\overline{G}, \overline{\gamma}}
+\vdash (L_{\overline{\gamma}} z. (\tin_H)_{\overline{G},
+  \overline{\gamma}} z) \circ
   ((\map_{H[\phi := (\mu \phi. \lambda \overline{\alpha}. H)
       \overline{\beta}]}^{\overline{F},
-    \overline{G}}\, \overline{t})_{\ol\gamma} z)
+    \overline{G}})_\emptyset\, \overline{t})
   : \xi}
@@
   &\setsem{\emptyset; \emptyset~|~ y : \Nat^\emptyset F \, G
-    \vdash L_{{\emptyset}}
-    z. (\map_{\mathit{Bush}\,\alpha}^{{F},
-      {G}}\, y)_{\emptyset} (\mathit{reverseBush}_{F}\, z) : \Nat^{{\emptyset}}
+    \vdash ((\map_{\mathit{Bush}\,\alpha}^{{F},
+      {G}}P)_\emptyset\, y) \circ (L_{{\emptyset}}
+    z. \mathit{reverseBush}_{F}\, z) : \Nat^{{\emptyset}}
     (\mathit{Bush} \, F) \, (\mathit{Bush} \, G) }\\ =\;
   &\setsem{\emptyset; \emptyset~|~ y : \Nat^\emptyset F\, G
-    \vdash L_{\emptyset} z. \mathit{reverseBush}_{G}
-    ((\map_{\mathit{Bush} \, \alpha}^{F, G}\,
-           {y})_{\emptyset} z) : \Nat^{\emptyset} (\mathit{Bush} \, F ) \,
+    \vdash (L_{\emptyset} z. \mathit{reverseBush}_{G}\,z) \circ
+    ((\map_{\mathit{Bush} \, \alpha}^{F, G})_\emptyset\,
+           {y}) : \Nat^{\emptyset} (\mathit{Bush} \, F ) \,
            (\mathit{Bush} \, G)}
@@
   &\setsem{\emptyset; \emptyset~|~y : \Nat^\gamma F \, G 
-  \vdash L_{{\gamma}} z.
-    (\map_{\mathit{Tree}\,\alpha \, \gamma}^{F,
-      G}\, y)_{\gamma} (\mathit{reverseTree}_{F, \gamma}\, z) : \Nat^{{\gamma}}
+  \vdash  ((\map_{\mathit{Tree}\,\alpha \, \gamma}^{F,
+      G})_\emptyset\, y) \circ (L_{{\gamma}} z. \mathit{reverseTree}_{F,
+      \gamma}\, z) : \Nat^{{\gamma}} 
     (\mathit{Tree} \, F \, \gamma) \, (\mathit{Tree} \, G \,
     \gamma) }\\ =\; &\setsem{\emptyset; \emptyset~|~y :
-    \Nat^\gamma F \, G \vdash L_{\gamma}
-    z. \mathit{reverseTree}_{G, \gamma}\,
+    \Nat^\gamma F \, G \vdash (L_{\gamma}
+    z. \mathit{reverseTree}_{G, \gamma}\,z) \circ
     ((\map_{\mathit{Tree} \, \alpha\, \gamma}^{{F},
-      {G}}\, y)_{\gamma} z) : \Nat^{\gamma} (\mathit{Tree} \,
+      {G}})_\emptyset\, y) : \Nat^{\gamma} (\mathit{Tree} \,
     F \, \gamma) \, (\mathit{Tree} \, G\, \gamma)}
@@
 &\setsem{ \Gamma;\emptyset\,|\,\ol{x : \Nat^\emptyset F \, G}
-  \vdash \map^{\ol{F},\ol{G}}_H\,\ol x :
+  \vdash (\map^{\ol{F},\ol{G}}_H)_\emptyset\,\ol x :
   \Nat^{\emptyset}\,H[\ol{\alpha := F}]\,H[\ol{\alpha := G}]}
@@
 \map_{H[\ol{\psi := K}]}^{\ol{F}, \ol{G}} : \xi} = \setsem{\Gamma;
-  \emptyset \,|\, \emptyset \vdash \map_H^{\ol{K[\ol{\phi := F}]},
-    \ol{K[\ol{\phi := G}]}} \circ \ol{\map_K^{\ol{F}, \ol{G}}} :
+  \emptyset \,|\, \emptyset \vdash L_\emptyset \ol
+  z. (\map_H^{\ol{K[\ol{\phi := F}]}, 
+    \ol{K[\ol{\phi := G}]}})_\emptyset \,(\ol{(\map_K^{\ol{F}, \ol{G}})_\emptyset \, \ol z}) :
   \xi}\]
@@
     s_{\overline{M [\overline{\psi := G}] [\phi := K]},
-      \overline{\gamma}} \Big(\big( \map_{H}^{\overline{M
+      \overline{\gamma}} \Big(\big( (\map_{H}^{\overline{M
         [\overline{\psi := F}] [\phi := H]}, \overline{M
-        [\overline{\psi := G}] [\phi := K]}}
-    (\overline{\map_{M}^{\overline{I}, \overline{J}}
+        [\overline{\psi := G}] [\phi := K]}})_\emptyset
+    (\overline{(\map_{M}^{\overline{I}, \overline{J}})_\emptyset
       (s, \overline{t})}) \big)_{\overline{\gamma}} z \Big) : \xi}\]
 \[= \setsem{\Gamma; \emptyset\,|\, \Delta \vdash
-    L_{\overline{\gamma}} z.\, \big( \map_{K}^{\overline{M
+    L_{\overline{\gamma}} z.\, \big( (\map_{K}^{\overline{M
       [\overline{\psi := F}] [\phi := H]}, \overline{M [\overline{\psi
-          := G}] [\phi := K]}} (\overline{\map_{M}^{\overline{I},
-    \overline{J}} (s, \overline{t})}) \big)_{\overline{\gamma}} \Big(
+          := G}] [\phi := K]}})_\emptyset (\overline{(\map_{M}^{\overline{I},
+    \overline{J}})_\emptyset (s, \overline{t})}) \big)_{\overline{\gamma}} \Big(
 s_{\overline{M [\overline{\psi := F}] [\phi := H]}, \overline{\gamma}}
@@
     \overline{\gamma}} H[\phi := F][\overline{\alpha := \beta}]\, F
-  \vdash x \circ \big( \map_{H [\ol{\alpha := \beta}]}^{(\mu
+  \vdash x \circ \big( (\map_{H [\ol{\alpha := \beta}]}^{(\mu
     \phi. \lambda \overline{\alpha} H) \overline{\beta},
-    F} ((\fold_{H, F})_{\emptyset} x) \big) : \xi}\]
+    F})_\emptyset ((\fold_{H, F})_{\emptyset} x) \big) : \xi}\]
 
@@
       \overline{\alpha}. H)\overline{\beta}]})_{\emptyset}
-  (\map_{H}^{H[\phi := (\mu \phi. \lambda
+  ((\map_{H}^{H[\phi := (\mu \phi. \lambda
       \overline{\alpha}. H)\overline{\beta}][\overline{\alpha :=
         \beta}], (\mu \phi. \lambda
-    \overline{\alpha}. H)\overline{\beta}} \tin_H) :
+    \overline{\alpha}. H)\overline{\beta}})_\emptyset \tin_H) :
   \xi}\]
@@
 
-\[\hspace*{-3.6in} = \setsem{\Gamma; \emptyset \,|\, \emptyset \vdash \mathit{id}_{(\mu
+\[\hspace*{-3.46in} = \setsem{\Gamma; \emptyset \,|\, \emptyset \vdash \mathit{id}_{(\mu
     \phi. \lambda \overline{\alpha}. H)\overline{\beta}} : \xi}\]
@@
       \overline{\alpha}. H)\overline{\beta}]})_{\emptyset}
-  (\map_{H}^{H[\phi := (\mu \phi. \lambda
+  ((\map_{H}^{H[\phi := (\mu \phi. \lambda
       \overline{\alpha}. H)\overline{\beta}][\overline{\alpha :=
         \beta}], (\mu \phi. \lambda
-    \overline{\alpha}. H)\overline{\beta}} \tin_H) \circ
+    \overline{\alpha}. H)\overline{\beta}})_\emptyset \tin_H) \circ
   \tin_H : \xi}\]
@@
 
-\[\hspace*{-3.32in} =\setsem{ \Gamma; \emptyset \,|\, \emptyset \vdash
+\[\hspace*{-3.15in} =\setsem{ \Gamma; \emptyset \,|\, \emptyset \vdash
   \mathit{id}_{H[\phi := (\mu \phi. \lambda
@@
 \relsem{\vdash g : \Nat^\alpha \,\onet\,\alpha}$.  By
-Theorem~\ref{thm:abstraction}, $(G^\set(\pi_1\rho),G^\set(\pi_2\rho))
+Theorem~\ref{thm:at-gen}, $(G^\set(\pi_1\rho),G^\set(\pi_2\rho))
 = G^\rel\rho$. Thus, for all $\rho \in \relenv$ and any $(a, b) \in
@@
 \relsem{\vdash g : \Nat^\alpha \,\alpha\,\alpha}$.  By
-Theorem~\ref{thm:abstraction}, $(G^\set(\pi_1\rho),G^\set(\pi_2\rho))
+Theorem~\ref{thm:at-gen}, $(G^\set(\pi_1\rho),G^\set(\pi_2\rho))
 = G^\rel\rho$. Thus, for all $\rho \in \relenv$ and any $(a, b) \in
@@
 \emptyset$, then there is exactly one morphism $\id_S: S \to S$, so
-$G^\set_S : S \to S$ must be $\id_S$. If $S \not = \emptyset$, then if
-$a$ is any element of $S$ and $K_a :S \to S$ is the constantly
+$G^\set_S : S \to S$ must be $\id_S$. If $S \not = \emptyset$, if
+$a$ is any element of $S$, and if $K_a :S \to S$ is the constantly
 $a$-valued morphism on $S$, then instantiating the naturality square
@@
 \mathit{filter}\, (\pi_2 \rho)\, b) \in \relsem{\alpha; \emptyset
-  \vdash \filtype} \rho\] so if $(s, s') \in \relsem{\alpha; \emptyset
+  \vdash \filtype} \rho\] so if $(s', s) \in \relsem{\alpha; \emptyset
   \vdash \Nat^\emptyset \alpha \, \mathit{Bool}} \rho = \rho\alpha \to
-\Eq_{\mathit{Bool}}$ and $(xs, xs') \in \relsem{\alpha; \emptyset
+\Eq_{\mathit{Bool}}$ and $(xs', xs) \in \relsem{\alpha; \emptyset
   \vdash List \, \alpha} \rho$ then
@@
 \begin{equation}\label{eq:filter-thm-list}
-  (\mathit{filter}\, (\pi_1\rho) \,a \,s \,xs, \mathit{filter} \,
-  (\pi_2\rho) \,b \,s' \,xs') \in \relsem{\alpha; \emptyset \vdash
+  (\mathit{filter}\, (\pi_1\rho) \,a \,s' \,xs', \mathit{filter} \,
+  (\pi_2\rho) \,b \,s \,xs) \in \relsem{\alpha; \emptyset \vdash
     \mathit{List} \, \alpha} \rho
@@
 \graph{\mathit{map}_{\mathit{List}} \, g}$ by Lemma~\ref{lem:graph}
-and Equation~\ref{thm:demotion-objects}.  Moreover, $xs' =
-\mathit{map}_{\mathit{List}} \,g \,xs$ and $(s', s) \in \graph{g} \to
+and Equation~\ref{thm:demotion-objects}.  Moreover, $xs =
+\mathit{map}_{\mathit{List}} \,g \,xs'$ and $(s', s) \in \graph{g} \to
 \Eq_{\mathit{Bool}}$, so $s' = s \circ g$. The result
@@
 \begin{thm}
-If $\vdash F$, $\vdash H$, and 
+If \,$\vdash F$, $\vdash H$, and 
 %$\beta; \emptyset \,|\, \emptyset \vdash g : \Nat^{\emptyset}
@@
   \beta)\, \beta}$ for some $g$, then
-\[\mathit{fold}_{1 + F \times \_}\, n\, c\; (G\; (\mathit{List}\,
+\[\mathit{fold}_{1 + \setsem{\vdash F} \times \_}\, n\, c\; (G\; (\mathit{List}\,
 \setsem{\vdash F})\,\mathit{nil} \,\mathit{cons}) = G \,\setsem{\vdash
@@
     \times \beta}^{H} : \Nat^{\emptyset} (\Nat^{\emptyset} (\onet
-  + F \times H)\, \tau')\, (\Nat^{\emptyset} (\mu \alpha. \onet
+  + F \times H)\, H)\, (\Nat^{\emptyset} (\mu \alpha. \onet
   + F \times \alpha)\, H)} =
-\mathit{fold}_{1 + F \times \_}\] so that if $c \in
+\mathit{fold}_{1 + \setsem{\vdash F} \times \_}\] so that if $c \in
 \setsem{\vdash F} \,\times\, \setsem{\vdash H} \to
@@
 $\pi_2 \rho \beta = \setsem{\vdash H}$,\,
-$\rho \beta = \graph{\mathit{fold}_{1 + F \times \_}\, n\, c} :
+$\rho \beta = \graph{\mathit{fold}_{1 + \setsem{\vdash F} \times \_}\, n\, c} :
 \rel(\pi_1 \rho \beta, \pi_2 \rho \beta)$,\,
@@
   F})\,\mathit{nil} \,\mathit{cons}, G \,\setsem{\vdash H} \,
-n\, c) \in \graph{\mathit{fold}_{1 + F \times \_}\, n\, c}$, i.e.,
-that $\mathit{fold}_{1 + F \times \_}\, n\, c\; (G\;
+n\, c) \in \graph{\mathit{fold}_{1 + \setsem{\vdash F} \times \_}\, n\, c}$, i.e.,
+that $\mathit{fold}_{1 + \setsem{\vdash F} \times \_}\, n\, c\; (G\;
 (\mathit{List}\,\setsem{\vdash F})\, \mathit{nil}\, \mathit{cons})
@@
 
-Using Equation~\ref{thm:demotion-objects} we can extend short
-cut fusion results to arbitrary ADTs, as in~\cite{joh02}.
+We can extend short cut fusion results to arbitrary ADTs, as
+in~\cite{joh02,pit98}.
 

### 2020-07-01 free-theorems-nesteds/paper5.tex
small changes/typos to intro/abstract

@@
 for nested data types. Rather than representing nested types via their
-Church encodings into a higher-kinded extension of System F, we adopt
+Church encodings in a higher-kinded extension of System F, we adopt
 a functional programming perspective and design a polymorphic calculus
@@
 definitions of the $\mathtt{reverse}$ and $\mathtt{map}$ functions
-over lists to according to the inductive structure of the input
+over lists according to the inductive structure of the input
 list. Such data-driven induction proofs over ADTs are routine enough
@@
 not only gave the first principled and practically useful structural
-induction rules for bushes and other truly nested types, and has also
+induction rules for bushes and other truly nested types, but has also
 opened the way for incorporating automatic generation of such rules
@@
   \rightarrow}\\
-\hspace*{0.8in} \mathtt{( \{B : Set\} \rightarrow B \rightarrow F B) \rightarrow}\\
-\hspace*{0.8in} \mathtt{(\{B : Set\} \rightarrow F (B \times B)
-  \rightarrow F B) \rightarrow}\\
+\hspace*{0.8in} \mathtt{( \{B : Set\} \rightarrow B \rightarrow F\;B) \rightarrow}\\
+\hspace*{0.8in} \mathtt{(\{B : Set\} \rightarrow F\;(B \times B)
+  \rightarrow F \;B) \rightarrow}\\
 \hspace*{0.8in}\mathtt{PTree \;A \rightarrow F\;A}\\
@@
   Set\} \rightarrow}\\
-\hspace*{0.5in} \mathtt{(\{B : Set\} \rightarrow F B) \rightarrow}\\
+\hspace*{0.5in} \mathtt{(\{B : Set\} \rightarrow F \;B) \rightarrow}\\
 \hspace*{0.5in} \mathtt{(\{B : Set\} \rightarrow B \rightarrow F\;
@@
 interpreted as functors from environments interpreting their type
-variable contexts to set or relations, as appropriate. But to ensure
+variable contexts to sets or relations, as appropriate. But to ensure
 that these functors satisfy the cocontinuity properties needed for the
@@
 satisfies an Abstraction Theorem (Theorem~\ref{thm:abstraction}), and
-we derive several of this latter kind of free theorems from it in
+we derive several of this latter kind of free theorem from it in
 Section~\ref{sec:ftnt}.  Specifically, we state and prove
@@
 types using Church encodings rather than constructing them via
-primitives. Since his model is entirely syntactic, his syntactic
-``functors'', whose associated $\mathit{fmap}$ functions representing
-their functorial actions must be {\em given} together with their
-underlying type constructors, cannot be interpreted as semantic
-functors. The associated Church encodings therefore cannot be
-interpreted as their fixpoints, so Atkey need not, and does not,
-impose cocontinuity conditions on his model to ensure that these
-fixpoints exist.  Nevertheless, Atkey does verify the existence of
-initial algebras for ``functors'' in iCIC. He does not indicate which
-type constructors support the kinds of $\mathit{fmap}$ functions
-needed to ensure they are ``functors'', but we suspect spelling this
+primitives. However, the $\mathit{fmap}$ functions associated to 
+Atkey's functors must be {\em given} together with their
+underlying type constructors, and he need not 
+impose cocontinuity conditions on his model to ensure that 
+fixpoints of his functors exist. Unfortunately, Atkey does not 
+indicate which
+type constructors support the $\mathit{fmap}$ functions
+needed for them to be functors, but we suspect spelling this
 out explicitly would result in a full higher-kinded extension of the
 calculus presented here.
-
 \section{The Calculus}\label{sec:calculus}

### 2020-07-01 submitted-POPL-2021/paper-final.tex
submitted version

--- 2020-07-01 free-theorems-nesteds/paper5.tex
@@
 
+It is well-known that there are no set-based parametric models of
+System F~\cite{rey84}. So it is worth noting that our calculus does
+not contain all System F types, and in fact is not impredicative, so a
+set-based parametric model is not {\em a priori} precluded. In fact,
+the local finite presentability of $\set$ makes it an obvious choice
+for interpreting nested types. We return to this point in
+Section~\ref{sec:conclusion}.
+
 There is a long line of work on categorical models of parametricity
@@
 %functor interpreting that type constructor.
-the Church encoding of each such type is interpreted as the least
+each Church encoding is interpreted as the least
 fixpoint of the (first-order) functor interpreting the type
-constructor from which its Church encoding was constructed.
+constructor from which it was constructed.
 %This is the categorical analogue of the approach in~\cite{atk12}
@@
 Clearly, $\graph{\alpha}^*$ is $\omega$-cocontinuous, so
-$\graph{\alpha}^* : [\rel^k,\rel]$. Now, suppose $\overline{R :
+$\graph{\alpha}^* : [\rel^k,\rel]$. Now, let $\overline{R :
   \rel(A, B)}$, $\overline{S : \rel(C, D)}$, and $\overline{(\beta,
@@
 \subseteq \relsem{\Gamma; \Phi \vdash F} \graph{f}$.
-
 \end{proof}
@@
 \end{comment}
-
-\vspace*{-.2in}
 

### 2020-07-02 submitted-POPL-2021/paper-final.tex
updated submission

@@
 System F~\cite{rey84}. So it is worth noting that our calculus does
-not contain all System F types, and in fact is not impredicative, so a
+not contain all System F types, and is not impredicative, so a
 set-based parametric model is not {\em a priori} precluded. In fact,
-the local finite presentability of $\set$ makes it an obvious choice
-for interpreting nested types. We return to this point in
-Section~\ref{sec:conclusion}.
+local finite presentability of $\set$ makes it an obvious choice for
+interpreting nested types. We return to this point at several places
+below.
 
@@
 System $F_\omega$, but within the impredicative Calculus of Inductive
-Constructions (iCIC) rather than in a semantic category. His
+Constructions rather than in a semantic category. His
 construction is in some ways similar to ours, but it represents data

### 2020-07-06 20-gadts/inf-rules.tex
added 20-gadts

--- 2020-07-01 free-theorems-nesteds/paper5.tex
@@
-% double-blind review submission, w/o CCS and ACM Reference (max
+% For double-blind review submission, w/o CCS and ACM Reference (max
 % submission space)
@@
 \renewcommand{\greyout}[1]{} %{{\color{gray} {#1}}} -- toggle to remove greyed text
-
-\newcommand{\ininv}[2]{(\tin^{-1}_{\onet +
-  \beta \times \phi (\phi\beta)})_{#1}\, #2}
+\newcommand{\new}[1]{\color{blue}#1}
+
 \newcommand{\emptyfun}{{[]}}
@@
 
-\title[Free Theorems for Nested Types]{Free Theorems for
+\title[Practical Parametricity for GADTs]{Practical Parametricity for GADTs}
   %Primitive
-  Nested Types} %% [Short Title] is optional; when present,
+  %% [Short Title] is optional; when present,
                          %% will be used in header instead of Full
@@
 \begin{abstract}
-This paper considers parametricity and its consequent free theorems
-for nested data types. Rather than representing nested types via their
-Church encodings in a higher-kinded extension of System F, we adopt
-a functional programming perspective and design a polymorphic calculus
-that provides primitives for constructing a robust class of nested
-types directly as fixpoints.
-%\`{a} la Pitts' PolyPCF.
-At the term level, our calculus supports
-%the usual
-pattern matching, map functions, and stylized recursion via folds for
-all such types.
-%Using a variation on Atkey's 2012 construction for $F_\omega$,
-%
-%of a parametric model
-%that explicitly constructs equality operators rather than merely
-%postulating them,
-Our main technical contribution is the construction of a parametric
-model for this calculus. This is both delicate and challenging.  To
-establish a suitable Identity Extension Lemma for our calculus, extra
-care must be taken in the formation of fixpoint types; all ADTs and
-all (truly) nested types appearing in the literature can nevertheless
-be expressed in our calculus. In addition, to ensure the existence of
-semantic fixpoints interpreting nested types, cocontinuity conditions
-on the functors interpreting all of the types in our calculus must
-be threaded throughout our model construction.  Overall, this paper
-provides a blueprint for deriving principled parametricity results for
-practical programming languages, such as Haskell and Agda, that
-directly construct nested types as fixpoints rather than representing
-them via Church encodings.
-%encoding them.
-It also shows how parametric models can derive useful free
-theorems for programs over nested types. Like their counterparts for
-ADTs, these free theorems
-%our model derives
-provide
-%, in particular,
-a principled alternative to proving program properties via induction,
-e.g., using the deep induction rules for nested types recently
-developed by Johann and Polonsky.
-%We derive a number of free theorems for nested types.
-%To illustrate, we derive a number of free theorems.
-Our approach is illustrated with several examples. 
+Abstract goes here
 \end{abstract}
@@
 
-\section{Introduction}\label{sec:intro}
-
-Suppose we want to prove some property of programs over an algebraic
-data type (ADT) such as that of lists, coded in Agda
-%\footnote{All code appearing in this paper are written in Agda.}
-as
-
-\vspace*{-0.15in}
-
-{\small
-\[\begin{array}{l}
-\mathtt{data\; List \;(A : Set)\;:\;Set\;where}\\
-\hspace*{0.4in}\mathtt{nil\;:\; List\;A}\\
-\hspace*{0.4in}\mathtt{cons\;:\;A \rightarrow List\;A \rightarrow List\;A}
-\end{array}\]}
-A natural approach to the problem uses structural induction on the
-input data structure in question. This requires knowing not just the
-definition of the ADT of which the input data structure is an
-instance, but also the program text for the functions involved in the
-properties to be proved. For example, to prove by induction that
-mapping a polymorphic function over a list and then reversing the
-resulting list is the same as reversing the original list and then
-mapping the function over the result, we unwind the (recursive)
-definitions of the $\mathtt{reverse}$ and $\mathtt{map}$ functions
-over lists according to the inductive structure of the input
-list. Such data-driven induction proofs over ADTs are routine enough
-to be included in, say, an undergraduate functional programming
-course.
-
-An alternative approach to proving results like the above
-$\mathtt{map}$-$\mathtt{reverse}$ property for lists is to use
-parametricity, a formalization of extensional type-uniformity in
-polymorphic languages. Parametricity captures the intuition that a
-polymorphic program must act uniformly on all of its possible type
-instantiations; it is formalized as the requirement that every
-polymorphic program preserves all relations between any pair of types
-that it is instantiated with.  Parametricity was originally put forth
-by Reynolds~\cite{rey83} for System F~\cite{gr89}, the formal calculus
-at the core of all polymorphic functional languages. It was later
-popularized as Wadler's ``theorems for free''~\cite{wad89} because it
-allows the deduction of many properties of programs in such languages
-solely from their types, i.e., with no knowledge whatsoever of the
-text of the programs involved. To get interesting free theorems,
-Wadler's calculus included built-in ADTs; indeed, most of the free
-theorems in~\cite{wad89} are consequences of naturality for
-polymorphic list-processing functions. However, the techniques
-of~\cite{wad89} show that parametricity can also be used to derive 
-naturality properties for non-list ADTs, as well as results, like
-correctness of program optimizations like {\em short cut
-  fusion}~\cite{glp93},
-%,joh02,joh03},
-that go beyond simple naturality.
-%In addition, parametricity in the presence of ADTs
-%{\color{blue} check!}  has further been used to prove i) {\em
-%  properties of programs}, e.g., that they perform no illegal
-%operations, satisfy certain security criteria~\cite{ep03,rp10}, or are
-%observationally equivalent to their compiled forms~\cite{bh09,hd11};
-%ii) {\em whole-language properties}, e.g., that they support data
-%abstraction and modularity via representation independence,
-%\cite{bm05,dr04,jac99,kat11,mr92}, enforce information flow
-%policies~\cite{ss01,ts04}, or guarantee privacy~\cite{rp10}; and iii)
-%{\em properties of implementations}, e.g., compiler
-%correctness~\cite{ab08,bh09,hd11,glpj93}. Parametricity is thus an
-%important tool for reasoning about programs in polymorphic languages
-%that natively support ADTs. {\color{blue} Be clear about what this
-%  means. Re-word?}
-
-This paper is about parametricity and free theorems for a polymorphic
-calculus with explicit syntax not just for ADTs, but for the
-generalization of ADTs known as {\em nested types}~\cite{bm98}, as
-well. An ADT defines a {\em family of inductive data types}, one for
-each input type. For example, the $\mathtt{List}$ data type definition
-above defines a collection of data types $\mathtt{List\; A}$,
-$\mathtt{List\; B}$, $\mathtt{List\; (A \times B)}$, $\mathtt{List\;
-  (\List\;A)}$, etc., each independent of all the others. By contrast,
-a nested type is an {\em inductive family of data types} that is
-defined over, or is defined mutually recursively with, (other) such
-data types. Since the structures of the data type at one input type
-can depend on those at other input types, the entire family of types
-must be defined at once. Examples of nested types include, trivially,
-ordinary ADTs, such as list and tree types; simple nested types, such
-as the data type {\small
-\[\begin{array}{l}
-\mathtt{data\; PTree\;(A : Set)\;:\;Set\;where}\\
-\hspace*{0.4in}\mathtt{pleaf\;:\;A \rightarrow PTree\;A}\\
-\hspace*{0.4in}\mathtt{pnode\;:\;PTree\;(A \times A) \rightarrow PTree\;A}
-\end{array}\]}
-\hspace{-0.04in}of perfect trees, whose recursive occurrences never
-appear below other type constructors; ``deep'' nested
-types~\cite{jp20}, such as the data type {\small
-\[\begin{array}{l}
-\mathtt{data\; Forest\;(A : Set)\;:\;Set\;where}\\
-\hspace*{0.4in}\mathtt{fempty\;:\;Forest\;A}\\
-\hspace*{0.4in}\mathtt{fnode\;:\; A \rightarrow PTree\;(Forest\;A) \to
-Forest\;A}
-\end{array}\]}
-\hspace{-0.04in}of perfect forests, whose recursive occurrences appear
-below type constructors for other nested types; and truly nested
-types\footnote{Nested types that are defined over themselves are known
-  as {\em truly nested types}.}, such as the data type {\small
-\[\begin{array}{l}
-\mathtt{data\; Bush\;(A : Set)\;:\;Set\;where}\\
-\hspace*{0.4in}\mathtt{bnil\;:\; Bush\;A}\\
-\hspace*{0.4in}\mathtt{bcons\;:\;A \rightarrow Bush\;(Bush \; A)
-  \rightarrow Bush\;A} 
-\end{array}\]}
-\hspace{-0.04in}of bushes (also called {\em bootstrapped heaps}
-in~\cite{oka99}), whose recursive occurrences appear below their own
-type constructors.
-
-Suppose we now want to prove properties of functions over nested
-types. We might, for example, want to prove a
-$\mathtt{map}$-$\mathtt{reverse}$ property for the functions on
-perfect trees in Figure~\ref{fig:funs}, or for those on
-bushes\footnote{To define the $\mathtt{foldBush}$ and
-  $\mathtt{mapBush}$ functions in Figure~\ref{fig:funs2} it is
-  necessary to turn off Agda's termination checker.} in
-Figure~\ref{fig:funs2}. A few well-chosen examples quickly convince us
-that such a property should indeed hold for perfect trees, and,
-drawing inspiration from the situation for ADTs, we easily construct a
-proof by induction on the input perfect tree. To formally establish
-this result, we could even prove it in Coq or Agda: each of these
-provers actually generates an induction rule for perfect trees, and
-the rule it generates gives the expected result because proving
-properties of perfect trees requires only that we induct over the
-top-level perfect tree in the recursive position. That is, any data
-internal to the input tree is left untouched.
-
-%want to prove that $\mathtt{reversePTree}$ commutes with
-%$\mathtt{mapPTree}$ or that $\mathtt{reverseBush}$ commutes with
-%$\mathtt{mapBush}$. It is not hard to convince oneself of the former:
-%a few well-chosen examples show clearly why mapping a polymorphic
-%%function over a perfect tree and then reversing the data at the leaves
-%of the resulting perfect tree is indeed the same result as mapping
-%that same function over the result of reversing the data at the leaves
-%of the original perfect tree.
-
-Unfortunately, it is nowhere near as clear that analogous intuitive or
-formal inductive arguments can be made for the
-$\mathtt{map}$-$\mathtt{reverse}$ property for bushes. Indeed, a proof
-by induction on the input bush must recursively induct over the bushes
-that are internal to the top-level bush in the recursive position.
-This is sufficiently delicate that no induction rule for bushes or
-other truly nested types was known until very recently, when {\em deep
-  induction}~\cite{jp20} was developed as a way to induct over {\em
-  all} of the structured data present in an input. Deep induction thus
-not only gave the first principled and practically useful structural
-induction rules for bushes and other truly nested types, but has also
-opened the way for incorporating automatic generation of such rules
-for (truly) nested data types --- and, eventually, even GADTs --- into
-modern proof assistants.
-\begin{figure*}
-\hspace*{-0.7in}
-\resizebox{0.35\linewidth}{!}{
-\begin{minipage}[t]{0.5\textwidth}
-\[\begin{array}{l}
-\mathtt{reversePTree : \forall \{A : Set\}  \rightarrow PTree\; A
-  \rightarrow PTree \;A}\\  
-\mathtt{reversePTree \; \{A\} = foldPTree \; \{A\}\; \{PTree\}}\\
-\hspace*{1.3in} \mathtt{pleaf}\\
-\hspace*{1.3in} \mathtt{( \lambda p \rightarrow pnode\; (mapPTree\;
-  swap\; p) )}\\  
-\\
-\mathtt{foldPTree : \forall \{A : Set\} \rightarrow \{F : Set \rightarrow Set \}
-  \rightarrow}\\
-\hspace*{0.8in} \mathtt{( \{B : Set\} \rightarrow B \rightarrow F\;B) \rightarrow}\\
-\hspace*{0.8in} \mathtt{(\{B : Set\} \rightarrow F\;(B \times B)
-  \rightarrow F \;B) \rightarrow}\\
-\hspace*{0.8in}\mathtt{PTree \;A \rightarrow F\;A}\\
-\mathtt{foldPTree\; n\; c\; (pleaf\; x) = n \;x}\\
-\mathtt{foldPTree\; n\; c \;(pnode\; p) = c\; (foldPTree \;n \;c \;p)}\\
-\\
-\mathtt{mapPTree : \forall \{A\, B : Set\} \rightarrow (A \rightarrow B)
-  \rightarrow PTree\; A \rightarrow PTree\; B}\\ 
-\mathtt{mapPTree \;f\; (pleaf \;x) = pleaf\; (f\; x)}\\
-\mathtt{mapPTree\; f \;(pnode\; p) =}\\
-\hspace*{0.7in}  \mathtt{pnode\; (mapPTree \;(\lambda p
-  \rightarrow (f (\pi_1 \,p), f (\pi_2\,p)))\; p)}\\ 
-\\
-\mathtt{swap : \forall \{A : Set\} \rightarrow (A \times A)
-  \rightarrow (A \times A)}\\ 
-\mathtt{swap \;(x,y) = (y,x)}
-\end{array}\]
-\caption{$\mathtt{reversePTree}$ and auxiliary functions in Agda}\label{fig:funs} 
-\end{minipage}}
-\quad\quad\quad\quad\quad%\quad\quad\quad
-\resizebox{0.35\linewidth}{!}{
-\begin{minipage}[t]{0.5\textwidth}
-\[\begin{array}{l}
-  \mathtt{reverseBush : \forall \{A : Set\} \rightarrow Bush\; A
-  \rightarrow Bush\; A}\\ 
-\mathtt{reverseBush \;\{A\} = foldBush\; \{A\}\; \{Bush\}\; bnil \;balg}\\
-\\
-\mathtt{foldBush : \forall \{A : Set\} \rightarrow \{F : Set \rightarrow
-  Set\} \rightarrow}\\
-\hspace*{0.5in} \mathtt{(\{B : Set\} \rightarrow F \;B) \rightarrow}\\
-\hspace*{0.5in} \mathtt{(\{B : Set\} \rightarrow B \rightarrow F\;
-  (F\; B) \rightarrow F\; B) \rightarrow}\\ 
-\hspace*{0.5in} \mathtt{Bush\; A \rightarrow F \;A}\\
-\mathtt{foldBush\; bn\; bc\; bnil = bn}\\
-\mathtt{foldBush \;bn \;bc \;(bcons\; x \;bbx) =}\\
-\hspace*{0.5in} \mathtt{bc\; x \;(foldBush \;bn\; bc
-  \;(mapBush \;(foldBush\; bn\; bc) \;bbx))}\\
-\\
-\mathtt{mapBush : \forall \{A \,B : Set\} \to (A \to B) \to Bush\, A \to
-  Bush\, B}\\ 
-\mathtt{mapBush\; \_ \; bnil = bnil}\\
-\mathtt{mapBush\; f\; (bcons\; x \;bbx) = bcons \;(f\, x) \; (mapBush\;
-  (mapBush\, f) \; bbx)}\\ 
-\\
-\mathtt{balg : \forall \{B : Set\} \rightarrow B \rightarrow Bush
-  \;(Bush\; B) \rightarrow Bush\; B}\\
-\mathtt{balg \;x \;bnil = bcons\; x \;bnil}\\
-\mathtt{balg \;x\; (bcons \;bnil \;bbbx) = bcons\; x \;(bcons \;bnil\; bbbx)}\\
-\mathtt{balg\; x \;(bcons \;(bcons\; y \;bbx)\; bbbx) =}\\
-\hspace*{0.5in} \mathtt{bcons\; y\; (bcons\; (bcons \;x \;bbx) \;bbbx)}\\
-\end{array}\]
-\caption{$\mathtt{reverseBush}$ and auxiliary
-  functions in Agda}\label{fig:funs2} 
-\end{minipage}}
-\end{figure*}
-
-\vspace*{-0.2in}
-
-Of course it is great to know that we {\em can}, at last, prove
-properties of programs over (truly) nested types by induction.  But
-recalling that parametricity often provides a viable alternative to
-inductive proofs for ADTs,
-%inductive proofs over ADTs can sometimes be circumvented in the
-%presence of parametricity,
-we might naturally ask:
-\begin{verse}
-{\em Can we derive properties of functions over (truly) nested types
-  from parametricity?}
-\end{verse}
-This paper answers the above question in the affirmative by
-constructing a parametric model for a polymorphic calculus providing
-primitives for {\em constructing} nested types
-%(including ADTs)
-directly via type-level recursion --- rather than representing them
-indirectly by Church encodings as in, e.g., System F.
-
-We introduce our calculus in Section~\ref{sec:calculus}.  At the type
-level, it is the level-2-truncation of the higher-kinded calculus
-from~\cite{jp19}, augmented with a primitive type of natural
-transformations.  To construct nested types, it carefully builds up
-type expressions not just from standard type variables, but also from
-type constructor variables of various arities, and includes an
-explicit $\mu$-construct for type-level recursion with respect to
-these variables. The class of nested types it constructs is very
-robust and includes all (truly) nested types known from the
-literature.  In Section~\ref{sec:type-interp} we give set and
-relational interpretations
-%{\color{blue} locally finitely presentable}
-for these types. As is standard when modeling parametricity, types are
-interpreted as functors from environments interpreting their type
-variable contexts to sets or relations, as appropriate. But to ensure
-that these functors satisfy the cocontinuity properties needed for the
-semantic fixpoints interpreting $\mu$-types to exist, set environments
-must map each $k$-ary type constructor variable to an appropriately
-cocontinuous $k$-ary functor on sets, relation environments must map
-each $k$-ary type constructor variable to an appropriately
-cocontinuous $k$-ary relation transformer, and these cocontinuity
-conditions must be threaded throughout the type interpretations in
-such a way that the resulting model is guaranteed to satisfy an
-appropriate Identity Extension Lemma (Theorem~\ref{thm:iel}).
-%{\color{blue} Syntactic restrictions on
-%  $\mu$-types are essential to proving IEL.}
-Properly propagating the cocontinuity conditions turns out to be both
-subtle and challenging, and Section~\ref{sec:iel}, where it is done,
-is where the bulk of the work in constructing our model lies.  At the
-term level, our calculus includes primitive constructs for the actions
-on morphisms of the functors interpreting types, initial algebras for
-fixpoints of these functors, and structured recursion over elements of
-these initial algebras (i.e., $\map$, $\tin$, and $\fold$ constructs,
-respectively). While our calculus does not support general recursion
-at the term level, it is strongly normalizing, so does perhaps edge us
-toward the kind of provably total practical programming language
-proposed at the end of~\cite{wad89}.  In
-Section~\ref{sec:term-interp}, we give set and relational
-interpretations for the terms of our calculus. As usual in parametric
-models, terms are interpreted as natural transformations from
-interpretations of the term contexts in which they are formed to the
-interpretations of their types, and these must cohere in what is
-essentially a fibrational way. Immediately from the definitions of our
-interpretations we prove in Section~\ref{sec:Nat-type-terms} a scheme
-deriving free theorems that are consequences of naturality of
-functions that are polymorphic over nested types. This scheme is very
-general, is parameterized over both the data type and the type of the
-polymorphic function at hand, and has each of the above
-$\mathtt{map}$-$\mathtt{reverse}$ theorems as instances.  The
-relationship between naturality and parametricity has long been of
-interest, and our inclusion of a primitive type of natural
-transformations allows us to clearly delineate those free theorems
-that are consequences of naturality, and may thus hold even in
-non-parametric models, from those that require the full power of
-parametricity.  In Section~\ref{sec:thms} we prove that our model
-satisfies an Abstraction Theorem (Theorem~\ref{thm:abstraction}), and
-we derive several of this latter kind of free theorem from it in
-Section~\ref{sec:ftnt}.  Specifically, we state and prove
-(non-)inhabitation results in Sections~\ref{sec:bottom}
-and~\ref{sec:identity}, a free theorem for the type of a filter
-function on generalized rose trees in Section~\ref{sec:ft-adt}, the
-correctness of short cut fusion for lists in
-Section~\ref{sec:short-cut}, and its generalization to nested types in
-Section~\ref{sec:short-cut-nested}.
-
-There is a long line of work on categorical models of parametricity
-for System F; see,
-e.g.,~\cite{bfss90,bm05,dr04,gjfor15,has94,jac99,mr92,rr94}.  To our
-knowledge, all categorical models that treat (algebraic) data types do
-so via their Church encodings, verifying in the
-%Much of this work does ultimately extend parametricity to data types
-%that are modeled as fixpoints of first-order functors, but to our
-%knowledge this has never been done by building the data types directly
-%into the calculus whose parametricity is to be modeled
-%categorically. Indeed, all extensions we know of use the categorical
-%equivalent of Atkey's approach. That is, they verify in the
-just-constructed parametric model that
-%the Church encoding constructed
-%from a type constructor is interpreted as the least fixpoint of the
-%functor interpreting that type constructor.
-the Church encoding of each such type is interpreted as the least
-fixpoint of the (first-order) functor interpreting the type
-constructor from which its Church encoding was constructed.
-%This is the categorical analogue of the approach in~\cite{atk12}
-%described above.
-The present paper draws on this rich tradition of categorical models
-of parametricity for System F, but treats nested types as well as
-ADTs, and is the first to treat these types by direct construction via
-primitives rather than by Church encodings. This requires that we
-%modify the type calculus to ensure that
-%System F's full impredicative polymorphism interacts reasonably with
-%functoriality,
-design our calculus in such a way that its types all have
-interpretations as appropriately cocontinuous functors, so that the
-existence of the fixpoints by which nested types are to be interpreted
-is ensured.
-%
-%functoriality is guaranteed syntactically, and that it is reflected in
-%our model construction so that the existence of the fixpoints by which
-%nested types are to be interpreted is ensured.
-%
-%parametricity to primitive nested types {\em as the model is being
-%constructed} by taking care to ensure that System F's full
-%impredicative polymorphism does not interfere with the functoriality
-%needed to support primitive nested types.
-
-Like our calculus, Pitts' PolyPCF~\cite{pit98,pit00}
-%extends parametricity from pure System F to System F augmented with
-%to a calculus that
-includes primitives for constructing data types directly. Only list
-types are added in~\cite{pit00}, but other polynomial ADTs are easily
-included as in~\cite{pit98}. Part of Pitts' motivation is to show that
-ADTs and their Church encodings have the same operational behavior in
-his calculus. We cannot even ask this question about our calculus,
-which cannot express Church encodings of even simple ADTs such as
-list, or pair, or sum types, since these are not functorial;
-nevertheless, we do prove the correctness of short cut fusion for
-nested types, whose operational analogue is valid in the parametric
-model Pitts constructs and is the means by which he proves his
-equivalence result. It would be interesting to know what operational
-analogues of functoriality and cocontinuity are needed to extend
-Pitts' parametricity results from a calculus with primitives for
-constructing data types modeled as fixpoints of first-order functors
-to one that also provides such primitives for data types modeled as
-fixpoints of higher-order functors.
-
-We are not the first to consider parametricity at higher
-kinds. Atkey~\cite{atk12} constructs a parametric model for full
-System $F_\omega$, but within the impredicative Calculus of Inductive
-Constructions (iCIC) rather than in a semantic category. His
-construction is in some ways similar to ours, but it represents data
-types using Church encodings rather than constructing them via
-primitives. However, the $\mathit{fmap}$ functions associated to 
-Atkey's functors must be {\em given} together with their
-underlying type constructors, and he need not 
-impose cocontinuity conditions on his model to ensure that 
-fixpoints of his functors exist. Unfortunately, Atkey does not 
-indicate which
-type constructors support the $\mathit{fmap}$ functions
-needed for them to be functors, but we suspect spelling this
-out explicitly would result in a full higher-kinded extension of the
-calculus presented here.
-\section{The Calculus}\label{sec:calculus}
-
-\subsection{Types}
-For each $k \ge 0$, we assume countable sets $\tvars^k$ of \emph{type
-  constructor variables of arity $k$} and $\fvars^k$ of
-\emph{functorial variables of arity $k$}, all mutually disjoint.
-%disjoint for distinct $k$ and disjoint from each other.
-The sets of all type constructor variables and functorial variables
-are $\tvars = \bigcup_{k \ge 0} \tvars^k$ and $\fvars = \bigcup_{k \ge
-  0} \fvars^k$, respectively, and a \emph{type variable} is any
-element of $\tvars \cup \fvars$.  We use lower case Greek letters for
-type variables, writing $\phi^k$ to indicate that $\phi \in \tvars^k
-\cup \fvars^k$, and omitting the arity indicator $k$ when convenient,
-unimportant, or clear from context. We reserve letters from the
-beginning of the alphabet to denote type variables of arity $0$, i.e.,
-elements of $\tvars^0 \cup \fvars^0$. We write $\overline{\zeta}$ for
-either a set $\{\zeta_1,...,\zeta_n\}$ of type constructor variables
-or a set of functorial variables when the cardinality $n$ of the set
-is unimportant or clear from context. If $\Pos\,$ is a set of type
-variables we write $\Pos, \overline{\phi}$ for $\Pos\, \cup
-\overline{\phi}$ when $\Pos\, \cap \overline{\phi} = \emptyset$.  We
-omit the vector notation for a singleton set, thus writing $\phi$,
-instead of $\overline{\phi}$, for $\{\phi\}$.
+\section{Typing Rules}\label{sec:types}
+
 \begin{dfn}
-Let $V$ be a finite subset of\, $\tvars$, $\Pos$ be a finite
-subset of\, $\fvars$, $\overline{\alpha}$ be a finite subset of\,
-$\fvars^0$ disjoint from $\Pos$, and $\phi^k \in \fvars^k
-\setminus \Pos$.  The set $\mcF^\Pos(V)$ of {\em functorial
-  expressions} over $\Pos$ and $V$ are given by
-\begin{align*}
-  \mcF^\Pos(V) \; ::= \; %\T(V)
-  & \hspace*{0.15in}
-\zerot \mid \onet 
-\mid \Nat^{\ol{\alpha}} \, \mcF^{\overline{\alpha}}(V) \;
-\mcF^{\ol{\alpha}}(V) %\mid V \ol{\T(V)}
-\mid \Pos\; \ol{\mcF^\Pos(V)}  \,
-\mid V\, \ol{\mcF^\Pos(V)}  
-\mid \mcF^{\Pos}(V) + \mcF^\Pos(V)\\
-&\mid \mcF^{\Pos}(V) \times \mcF^\Pos(V) \mid \left(\mu
-\phi^{~k}. \lambda \alpha_1...\alpha_k.  
-\mcF^{\Pos,\alpha_1,...,\alpha_k,\phi}(V)\right)
-  \ol{\mcF^{\Pos}(V)}
-\end{align*}
-\end{dfn}
-\noindent
-A \emph{type} over $\Pos$ and $V$ is any element of % $\T(V) \cup
-$\F^\Pos(V)$.
-
-The notation for types entails that an application $F F_1...F_k$ is
-allowed only when $F$ is a type variable of arity $k$, or $F$ is a
-subexpression of the form $\mu \phi^{k}.\lambda
-\alpha_1...\alpha_k.F'$. Moreover, if $F$ has arity $k$ then $F$ must
-be applied to exactly $k$ arguments.  Accordingly, an overbar
-indicates a sequence of subexpressions whose length matches the arity
-of the type applied to it.  The fact that types are always in
-\emph{$\eta$-long normal form} avoids having to consider
-$\beta$-conversion at the level of types. In a subexpression
-$\Nat^{\ol{\alpha}}F\,G$, the $\Nat$ operator binds all occurrences of
-the variables in $\ol{\alpha}$ in $F$ and $G$. Similarly, in a
-subexpression $\mu \phi^k.\lambda \ol{\alpha}.F$, the $\mu$ operator
-binds all occurrences of the variable $\phi$, and the $\lambda$
-operator binds all occurrences of the variables in $\ol{\alpha}$, in
-the body $F$.
-
-A {\em type constructor context} is a finite set $\Gamma$ of type
-constructor variables, and a {\em functorial context} is a finite set
-$\Phi$ of functorial variables. In Definition~\ref{def:wftypes}, a
-judgment of the form $\Gamma;\Phi \vdash F$ indicates that the
-type $F$ is intended to be functorial in the variables in $\Phi$
-but not necessarily in those in $\Gamma$.
-\begin{dfn}\label{def:wftypes}
 The formation rules for the set $\F \subseteq \bigcup_{V \subseteq
@@
 
-\vspace*{-0.2in}
-
 \[\begin{array}{cc}
@@
 \end{array}\]
+
+\vspace*{0.1in}
+
+\[\begin{array}{cc}
+\AXC{$\Gamma;\Phi \vdash F$}
+\AXC{$\Gamma;\Phi \vdash G$}
+\BIC{$\Gamma; \Phi \vdash F + G$}
+\DisplayProof
+&
+\AXC{$\Gamma;\Phi \vdash F$}
+\AXC{$\Gamma;\Phi \vdash G$}
+\BIC{$\Gamma; \Phi \vdash F \times G$}
+\DisplayProof
+\end{array}\]
+
+\vspace*{0.05in}
+
 \[\begin{array}{c}
-\AXC{$\Gamma;\ol{\alpha^0} \vdash F$}
-\AXC{$\Gamma;\ol{\alpha^0}  \vdash G$}
-\BIC{$\Gamma;\emptyset \vdash \Nat^{\ol{\alpha^0}}F \,G$}
-\DisplayProof
-\\[3ex]
+\new{\AXC{$\Gamma;\Phi \vdash F$}
+\AXC{$\Gamma;\Phi  \vdash G$}
+\BIC{$\Gamma;\emptyset \vdash \Nat^\Phi F \,G$}
+\DisplayProof}
+\\[4ex]
 \AXC{$\phi^k \in \Gamma \cup \Phi$}
@@
 \DisplayProof
-\\[2ex]
+\\[3ex]
 \AXC{$\Gamma;\ol{\gamma^0},\ol{\alpha^0},\phi^k \vdash F$}
@@
 \DisplayProof
-\end{array}\]
-
-\vspace*{0.04in}
-
-\[\begin{array}{cc}
-\AXC{$\Gamma;\Phi \vdash F$}
-\AXC{$\Gamma;\Phi \vdash G$}
-\BIC{$\Gamma; \Phi \vdash F + G$}
-\DisplayProof
-&
-\AXC{$\Gamma;\Phi \vdash F$}
-\AXC{$\Gamma;\Phi \vdash G$}
-\BIC{$\Gamma; \Phi \vdash F \times G$}
-\DisplayProof
-\end{array}\]
-%\noindent
-%A type $\tau$ is {\em well-formed} if it is
-%either a well-formed type constructor expression or a
-%well-formed functorial expression.
+\\[5ex]
+\new{
+\AXC{$\Gamma; \Phi,\ol\alpha^0 \vdash F$}
+\AXC{$\ol{\Gamma;\ol\alpha^0 \vdash K}$}
+\AXC{$\ol{\Gamma;\Phi \vdash A}$}
+\TIC{$\Gamma;\Phi \vdash (\Lan^{\ol{\alpha^0}}_{\ol K} F)\, \ol A$}
+\DisplayProof}
+\end{array}\]
 \end{dfn}
-If $F$ is a closed type we may write $\vdash F$, rather than
-$\emptyset;\emptyset \vdash F$, for the judgment that it is
-well-formed.  Definition~\ref{def:wftypes} ensures that the expected
-weakening rules for well-formed types hold, although weakening does
-not change the contexts in which types can be formed. If
-$\Gamma;\emptyset \vdash F$ and $\Gamma;\emptyset \vdash G$, then our
-rules allow formation of the type $\Gamma;\emptyset \vdash
-\Nat^\emptyset F \,G$. Since $\Gamma; \emptyset \vdash \Nat^{\ol
-  \alpha}F\,G$ represents a natural transformation in $\ol \alpha$
-from $F$ to $G$, the type $\Gamma;\emptyset \vdash \Nat^\emptyset F
-\,G$ represents the arrow type $\Gamma \vdash F \to G$ in our
-calculus. The type $\Gamma; \emptyset \vdash \Nat^{\ol\alpha} \,\onet
-\,F$ similarly represents the $\forall$-type $\Gamma; \emptyset \vdash
-\forall \ol\alpha . F$.  Note, however, that
-%not even all
-%Hindley-Milner types of System F are representable in our calculus.
-not all System F types are 
-%only those System F types that can be interpreted as functors are
-representable in our calculus.
-%
-%However, if $\ol\alpha$ is non-empty then $\tau$ cannot be of the
-%form $\Nat^{\ol\beta} H \, K$ since $\Gamma; \ol\alpha \vdash
-%\Nat^{\ol\beta} H \, K$ is not a valid type judgment in our calculus
-%(except by weakening).
-
-Definition~\ref{def:wftypes} allows the formation of all of the
-(closed) nested types from the introduction:
-\[\begin{array}{lll}
-\mathit{List}\, \alpha & = & \mu \beta. \,\onet + \alpha \times
-\beta\; = \; (\mu \phi. \lambda \beta.\,\onet + \beta \times \phi
-\beta)\,\alpha\\ 
-\mathit{PTree}\,\alpha & = & (\mu \phi. \lambda \beta.\,\beta +
-\phi\,(\beta \times \beta))\,\alpha\\
-\mathit{Forest}\,\alpha & = & (\mu \phi. \lambda \beta. \,\onet +
-\beta \times \mathit{PTree}\,(\phi \beta))\,\alpha\\ 
-\mathit{Bush}\,\alpha & = & (\mu \phi.\lambda \beta. \,\onet + \beta
-\times \phi\,(\phi\beta))\,\alpha 
-\end{array}\]
-Each of these types can be considered either functorial in $\alpha$ or
-not, according to whether $\alpha$ is in the functorial or
-non-functorial context in which it is formed. In particular, if
-$\emptyset;\alpha \vdash \mathit{List} \,\alpha$, then $\vdash
-\Nat^\alpha \onet\, (\mathit{List}\,\alpha)$ is well-formed; if
-$\alpha;\emptyset \vdash \List \,\alpha$, then it is not.  Similarly,
-if $\mathit{Tree}\,\alpha\,\gamma = \mu \,\beta.\, \alpha + \beta
-\times \gamma \times \beta$, then $\gamma;\emptyset \vdash \Nat^\alpha
-(\mathit{List}\, \alpha) \, (\mathit{Tree} \, \alpha\, \gamma)$ is
-well-formed, but $\emptyset;\gamma \vdash \Nat^\alpha (\mathit{List}\,
-\alpha) \, (\mathit{Tree} \, \alpha\, \gamma)$ is not. Although types
-can be functorial in variables of arity greater than $0$, the body $F$
-of a type $(\mu \phi.\lambda \ol\alpha.F)\,\ol G$ cannot. For example,
-the type $\mathit{GRose}\, \phi \, \alpha = \mu\beta . \alpha \times
-\phi \beta$ can be functorial or not in $\alpha$, but cannot be
-functorial in $\phi$. Moreover, $\mathit{GRose}\,\phi\,\alpha$ cannot
-be the (co)domain of a $\Nat$ type representing a natural
-transformation functorial in $\phi$, and the type $\mu \phi. \lambda
-\alpha. \mathit{GRose}\,\phi\,\alpha$ is not well-formed in our
-system. These restrictions ensure that the set and relational
-interpretations of types in Section~\ref{sec:type-interp} are
-functors. In Section~\ref{sec:conclusion} we discuss how
-%GADTs and other
-more expressive types might be accommodated.
-%consider ways to lessen the expressivity constraints they impose.
-%
-%These restrictions on higher-arity functoriality are necessary
-%for constructing our parametric model. Fortunately, they are entirely
-%natural, and {\color{blue} preclude only truly bizarre nested types}.
-%do not preclude any nested
-%types {\color{blue} occurring in practice}.
-
-%If $\mathit{Tree}\,\alpha\,\gamma = \mu \,\beta.\, \alpha + \beta \times
-%\gamma \times \beta$, then Definition~\ref{def:wftypes} also allows
-%the derivation of, e.g., the type $\gamma; \emptyset \vdash
-%\Nat \alpha (\mathit{List}\, \alpha) \, (\mathit{Tree} \, \alpha\,
-%\gamma)$ representing a natural transformation from lists to trees
-%that is natural in $\alpha$ but not necessarily in $\gamma$. We
-%emphasize that types can be functorial in variables of arity greater
-%than $0$. For example, the type $\mathit{GRose}\, \phi \, \alpha =
-%\mu\beta . \alpha \times \phi \beta$ can be functorial in $\phi$ if
-%$\phi \in \Phi$. As usual, whether $\phi \in \Gamma$ or $\phi \in
-%\Phi$ determines whether types such as $\Nat^\alpha ( \mathit{GRose}
-%\, \phi \, \alpha)\, (List\, \alpha)$ are well-formed. But even if
-%$\mathit{GRose}$ is functorial in $\phi$, it still cannot be the
-%domain of a $\Nat$ type representing a natural transformation in
-%$\phi$. This is because our calculus does not allow naturality in
-%variables of arity greater than $0$.
-
-Definition~\ref{def:wftypes} allows well-formed types to be 
-%%considers types in $\T$ to be types in $\F$ that are
-functorial in no variables.
-%This allows the formation of types such as
-%$\mathit{List}\,(\sigma \to \tau)$ and $\mathit{PTree}\, (\forall
-%\alpha.\tau)$.
-Functorial variables
-%in a well-formed type $\tau$
-can also be demoted to non-functorial status. The proof is by
-induction on the structure of $F$.
-
-\begin{lemma}
-If\, $\Gamma; \Phi, \phi^k \vdash F$, then $\Gamma, \psi^k; \Phi
-\vdash F[\phi^k :== \psi^k]$ is also derivable. Here, $F[\phi :==
-  \psi]$ is the textual replacement of $\phi$ in $F$, meaning that all
-occurrences of $\phi\ol\sigma$ in $F$ become $\psi\ol\sigma$.
-\end{lemma}
-
-In addition to textual replacement, we also have a proper substitution
-operation on types. If $F$ is a type over $P$ and $V$, if $\Pos$
-and $V$ contain only type variables of arity $0$, and if $k=0$ for
-every occurrence of $\phi^k$ bound by $\mu$ in $F$, then we say
-that $F$ is {\em first-order}; otherwise we say that $F$ is {\em
-  second-order}.  Substitution for first-order types is the usual
-capture-avoiding textual substitution. We write $F[\alpha :=
-  \sigma]$ for the result of substituting $\sigma$ for $\alpha$ in
-$F$, and $F[\alpha_1 := F_1,...,\alpha_k := F_k]$, or
-$F[\ol{\alpha := F}]$ when convenient, for $F[\alpha_1 :=
-  F_1][\alpha_2 := F_2,...,\alpha_k := F_k]$. Substitution
-for second-order types is defined below, where we adopt a similar
-notational convention for vectors of types. Note that it is not
-correct to substitute along non-functorial variables.
-
-\begin{dfn}\label{def:second-order-subst}
-If \,$\Gamma; \Phi,\phi^k \vdash H$
-%with $k \geq 1$,
-and if\, $\Gamma;\Phi, \ol{\alpha} \vdash F$ with
-$|\ol\alpha| = k$, then $\Gamma;\Phi \vdash H[\phi :=_{\ol{\alpha}}
-  F]$.  Similarly, if \,$\Gamma, \phi^k; \Phi \vdash H$,
-%with $k \geq 1$,
-and if\, $\Gamma; \ol\psi,\ol{\alpha} \vdash F$ with $|\ol\alpha| = k$
-and $\Phi \cap \ol\psi = \emptyset$, then $\Gamma,\ol\psi';\Phi \vdash
-H[\phi :=_{\ol{\alpha}} F[\ol{\psi :== \psi'}]]$. Here, the operation
-  $(\cdot)[\phi :=_{\ol \alpha} F]$ of {\em second-order type
-    substitution along $\ol\alpha$} is defined by:
-\[\begin{array}{lll}
-\zerot[\phi :=_{\ol{\alpha}} F] & = & \zerot\\[0.5ex]
-\onet[\phi :=_{\ol{\alpha}} F] & = & \onet\\[0.25ex]
-(\Nat^{\ol\beta} G \,K)[\phi :=_{\ol{\alpha}} F]
-& = & \Nat^{\ol\beta}\, (G[\phi :=_{\ol{\alpha}} F]) \,(K[\phi
-  :=_{\ol{\alpha}} F])\\
-(\psi\ol{G})[\phi :=_{\ol{\alpha}} F] & = &
-\left\{\begin{array}{ll}
-\psi \,\ol{G[\phi :=_{\ol{\alpha}} F]} & \mbox{if } \psi \not = \phi\\
-  F[\ol{\alpha  := G[\phi :=_{\ol{\alpha}} F]}] 
-  & \mbox{if } \psi = \phi
-\end{array}\right.\\[2.8ex]
-(G + K)[\phi :=_{\ol{\alpha}} F] & = & G[\phi
-  :=_{\ol{\alpha}} F] + K[\phi :=_{\ol{\alpha}} F]\\[0.5ex] 
-(G \times K)[\phi :=_{\ol{\alpha}} F] & = &
-G[\phi :=_{\ol{\alpha}} F] \times K[\phi
-  :=_{\ol{\alpha}} F]\\[0.5ex]   
-((\mu \psi. \lambda \ol{\beta}.\, G)\ol{K})[\phi :=_{\ol{\alpha}}
-  F] & = & (\mu \psi. \lambda \ol{\beta}. \,G[\phi :=_{\ol{\alpha}}
-  F])\, \ol{K[\phi :=_{\ol{\alpha}} F]}
-\end{array}\]
-\end{dfn}
-\noindent
-We note that $(\cdot)[\phi^0 :=_\emptyset F]$ coincides with
-first-order substitution. We also omit $\ol\alpha$ when convenient.
-
-\subsection{Terms}\label{sec:terms}
-
-We now define our term calculus. To do so we assume an infinite set
-$\cal V$ of term variables disjoint from $\tvars$ and $\fvars$. If
-$\Gamma$ is a type constructor context and $\Phi$ is a functorial
-context, then a {\em term context for $\Gamma$ and $\Phi$} is a finite
-set of bindings of the form $x : F$, where $x \in {\cal V}$ and
-$\Gamma; \Phi \vdash F$. We adopt the same conventions for denoting
-disjoint unions and for vectors in term contexts as for type
-constructor contexts and functorial contexts.
 
@@
 \[\begin{array}{ccc}
-%\AXC{$\Gamma;\emptyset \vdash \tau : \T$}
-%\UIC{$\Gamma;\emptyset\,|\, \Delta,x :\tau \vdash x : \tau$}
-%\DisplayProof
 \AXC{$\Gamma;\Phi \vdash F$}
@@
 \[\begin{array}{c}
-%\AXC{$\Gamma; \emptyset \vdash \Nat^{\ol{\alpha}} \,F\,G : \T$}
-\AXC{$\Gamma; \ol{\alpha} \vdash F$}
-\AXC{$\Gamma; \ol{\alpha} \vdash G$}
-\AXC{$\Gamma; \ol{\alpha} \,|\, \Delta, x : F \vdash t: G$} 
+\new{
+\AXC{$\Gamma; \Phi \vdash F$}
+\AXC{$\Gamma; \Phi \vdash G$}
+\AXC{$\Gamma; \Phi \,|\, \Delta, x : F \vdash t: G$} 
 \TIC{$\Gamma; \emptyset
-  \,|\, \Delta \vdash L_{\ol{\alpha}} x.t : \Nat^{\ol{\alpha}} \,F \,G$}
-\DisplayProof
-\\\\
-\AXC{$\ol{\Gamma;\Phi \vdash K}$}
+  \,|\, \Delta \vdash L_\Phi x.t : \Nat^\Phi \,F \,G$}
+\DisplayProof}
+\\\\
+\new{\AXC{$\ol{\Gamma;\Phi,\ol\beta \vdash K}$}
 \AXC{$\Gamma; \emptyset
-  \,|\, \Delta \vdash t : \Nat^{\ol{\alpha}} \,F \,G$}
-\AXC{$\Gamma;\Phi \,|\, \Delta \vdash s: F[\overline{\alpha := K}]$}
+  \,|\, \Delta \vdash t : \Nat^{\ol\psi} \,F \,G$}
+\AXC{$\Gamma;\Phi \,|\, \Delta \vdash s: F[\overline{\psi :=_{\ol\beta} K}]$}
 \TIC{$\Gamma;\Phi\,|\, \Delta \vdash t_{\ol K} s:
-  G[\overline{\alpha := K}]$}
-\DisplayProof
-\\\\
-\AXC{$\Gamma; \ol{\phi}, \ol{\gamma} \vdash H$}
-\AXC{$\ol{\Gamma; \ol{\beta},\ol{\gamma} \vdash F}$}
-\AXC{$\ol{\Gamma; \ol{\beta},\ol{\gamma} \vdash
+  G[\overline{\psi :=_{\ol\beta} K}]$}
+\DisplayProof}
+\\\\
+\new{
+\AXC{$\Gamma; \Phi, \ol{\phi} \vdash H$}
+\AXC{$\ol{\Gamma; \Phi,\ol{\beta} \vdash F}$}
+\AXC{$\ol{\Gamma; \Phi,\ol{\beta} \vdash
     G}$}
@@
   ~|~\emptyset
-%  \ol{ x : \Nat^{\ol{\beta},\ol{\gamma}}\,F\,G}
-  %  \emptyset
   \vdash \map^{\ol{F},\ol{G}}_H :
-  \Nat^\emptyset\;(\ol{\Nat^{\ol{\beta},\ol{\gamma}}\,F\,G})\;
-  (\Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\;H[\ol{\phi
+  \Nat^\emptyset\;(\ol{\Nat^{\Phi,\ol{\beta}}\,F\,G})\;
+  (\Nat^\Phi\,H[\ol{\phi :=_{\ol{\beta}} F}]\;H[\ol{\phi
       :=_{\ol{\beta}} G}])$} 
-\DisplayProof
-\\\\
-%\AXC{$\Gamma;\Phi \,|\, \Delta \vdash t : H[\phi := (\mu \phi.\lambda 
-%  \ol{\alpha}.H)\ol{\beta}][\ol{\alpha := \tau}]$}
-%\AXC{$\ol{\Gamma;\Phi \vdash \tau : \F}$}
-\AXC{$\Gamma; \phi, \ol{\alpha},\ol{\gamma} \vdash H$}
+\DisplayProof}
+\\\\
+\new{\AXC{$\Gamma; \Phi, \phi, \ol{\alpha} \vdash H$}
 \UIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \tin_H :
-  \Nat^{\ol{\beta},\ol{\gamma}} H[\phi :=_{\ol{\beta}} (\mu
+  \Nat^{\Phi,\ol{\beta}} H[\phi :=_{\ol{\beta}} (\mu
     \phi.\lambda \ol{\alpha}.H)\ol{\beta}][\ol{\alpha := \beta}]\,(\mu
   \phi.\lambda \ol{\alpha}.H)\ol{\beta}$}
-\DisplayProof
-\\\\
-\AXC{$\Gamma; \phi,\ol{\alpha}, \ol{\gamma} \vdash H$}
-\AXC{$\Gamma; \ol{\beta}, \ol{\gamma} \vdash F$}
-%\AXC{$\Gamma;\emptyset \,|\, \Delta \vdash t :
-%  \Nat^{\ol{\beta}, \ol{\gamma}}\,H[\phi :=_{\ol{\beta}} F][\ol{\alpha
-%      := \beta}]\,F$} 
-%\TIC{$\Gamma;\emptyset \,|\, \Delta \vdash
-%  \fold_{H,F}\, t : \Nat^{\ol{\beta}, \ol{\gamma}}\,((\mu
-%  \phi.\lambda \ol{\alpha}.H)\ol{\beta})\,F$} 
+\DisplayProof}
+\\\\
+\new{\AXC{$\Gamma; \phi,\Phi,\ol{\alpha} \vdash H$}
+\AXC{$\Gamma; \Phi, \ol{\beta} \vdash F$}
 \BIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \fold^F_H :
-  \Nat^\emptyset\; (\Nat^{\ol{\beta}, \ol{\gamma}}\,H[\phi
-    :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\; (\Nat^{\ol{\beta},
-    \ol{\gamma}}\,(\mu \phi.\lambda \ol{\alpha}.H)\ol{\beta}\,F)$}
-\DisplayProof
+  \Nat^\emptyset\; (\Nat^{\Phi,\ol{\beta}}\,H[\phi
+    :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\; (\Nat^{\Phi,\ol{\beta}}
+    \,(\mu \phi.\lambda \ol{\alpha}.H)\ol{\beta}\,F)$}
+\DisplayProof}
+\\\\
+\new{
+\AXC{$\Gamma; \Phi,\ol{\alpha}\vdash F$}
+\AXC{$\ol{\Gamma; \ol{\alpha} \vdash K}$}
+\AXC{$\ol{\Gamma;\Phi \vdash A} \hspace*{0.3in} \Gamma;\Phi~|~\Delta \vdash t : F[\ol{\alpha :=A}]$}
+\TIC{$\Gamma;\Phi~|~\Delta \vdash \int_{K,F} t :
+  (\Lan^{\ol\alpha}_{\ol K}\,  F)\,\ol{K[\ol{\alpha := A}]}$}  
+\DisplayProof}
+\\\\
+\new{
+\AXC{$\Gamma;\emptyset~|~\Delta \vdash \eta : \Nat^{\Phi,\ol\alpha}\,
+  F\;G[\ol{\beta := K}]$}
+\AXC{$\ol{\Gamma;\Phi \vdash A}$}
+\AXC{$\Gamma;\Phi~|~\Delta \vdash t : (\Lan^{\ol\alpha}_{\ol K} F)\,\ol A $}
+\TIC{$\Gamma;\Phi~|~\Delta \vdash \partial^{G, \ol K}_F \eta\,t :
+  G[\ol{\beta := A}]$}
+\DisplayProof}
 \end{array}\]
@@
 
-In the rule for $L_{\ol{\alpha}}x.t$, the $L$ operator binds all
-occurrences of the type variables in $\ol{\alpha}$ in the type of the
-term variable $x$ and in the body $t$, as well as all occurrences of
-$x$ in $t$. In the rule for $t_{\ol K} s$ there is one functorial
-expression in $\ol K$ for every functorial variable in $\ol \alpha$. In
-the rule for $\map^{\ol{F},\ol{G}}_H$ there is one functorial
-expression $F$ and one functorial expression $G$ for each functorial
-variable in $\ol\phi$. Moreover, for each $\phi^k$ in $\ol\phi$ the
-number of functorial variables in $\ol\beta$ in the judgments for its
-corresponding functorial expresssions $F$ and $G$ is $k$. In the rules
-for $\tin_H$ and $\fold^F_H$, the functorial variables in $\ol{\beta}$
-are fresh with respect to $H$, and there is one $\beta$ for every
-$\alpha$. (Recall from above that, in order for the types of $\tin_H$
-and $\fold^F_H$ to be well-formed, the length of $\alpha$ must equal
-the arity of $\phi$.) Substitution for terms is the obvious extension
-of the usual capture-avoiding textual substitution, and
-Definition~\ref{def:well-formed-terms} ensures that the expected
-weakening rules for well-formed terms hold.
-
-Using Definition~\ref{def:well-formed-terms} we can represent the
-$\mathtt{reversePTree}$ function from Figure~\ref{fig:funs} in our
-calculus as
-\[\vdash (\fold_{\beta + \phi(\beta \times \beta)}^{\mathit{PTree}\,
-  \alpha})_\emptyset \,
-%(\tin_{\beta + \phi(\beta \times \beta)} \circ
-s :
-\Nat^{\alpha} (\mathit{PTree}\,\alpha)\,(\mathit{PTree}\,\alpha)\] 
-where
-\[\begin{array}{lll}
-\fold_{\beta + \phi(\beta \times \beta)}^{\mathit{PTree}\,\alpha} & :
-& \Nat^{\emptyset} (\Nat^{\alpha} (\alpha + \mathit{PTree}\,(\alpha
-\times \alpha)) \; (\mathit{PTree}\,\alpha))\; (\Nat^{\alpha}
-(\mathit{PTree}\,\alpha) \; (\mathit{PTree}\,\alpha))\\
-\tin_{\beta + \phi(\beta \times \beta)} & : & \Nat^{\alpha} (\alpha +
-\mathit{PTree}\,(\alpha \times \alpha)) \; (\mathit{PTree}\, \alpha)\\
-\map_{\mathit{PTree}\,\alpha}^{\alpha \times \alpha, \alpha \times
-  \alpha} & : & \Nat^\emptyset
-(\Nat^{\alpha} (\alpha \times \alpha)\, (\alpha \times
-\alpha))\,
-(\Nat^{\alpha} (\mathit{PTree}\,(\alpha \times
-\alpha))\, (\mathit{PTree}\,(\alpha \times \alpha)))
-%\[\begin{array}{lll}
-%\vdash \fold_{\beta + \phi(\beta \times \beta)}^{\mathit{PTree}\,\alpha} &
-%: & \Nat^{\emptyset} (\Nat^{\alpha} (\alpha + \mathit{PTree}\,(\alpha \times
-%\alpha)) \; (\mathit{PTree}\,\alpha))\; (\Nat^{\alpha}
-%(\mathit{PTree}\,\alpha) \; (\mathit{PTree}\,\alpha))\\
-%\vdash \tin_{\beta + \phi(\beta \times \beta)} & : & \Nat^{\alpha}
-%(\alpha + \mathit{PTree}\,(\alpha \times \alpha)) \; (\mathit{PTree}\,
-%\alpha)\\ 
-%\vdash \map_{\mathit{PTree}\,\alpha}^{\alpha \times \alpha, \alpha \times
-%  \alpha} & : & \Nat^{\emptyset} (\Nat^{\alpha} (\alpha \times \alpha)\,
-%(\alpha \times \alpha))\; (\Nat^{\alpha} (\mathit{PTree}\,(\alpha \times
-%\alpha))\, (\mathit{PTree}\,(\alpha \times \alpha)))
-\end{array}\]
-and 
-$\mathit{pleaf}$,$\mathit{pnode}$, $\mathit{swap}$, and $s$ are the terms
-\[\begin{array}{l}
-\vdash \tin_{\beta + \phi (\beta \times \beta)} \circ (
-L_{\alpha}\, x.\, \inl\, x) \; : \; \Nat^{\alpha}\, \alpha\,
-(\mathit{PTree}\,\alpha)\\ 
-\vdash \tin_{\beta + \phi (\beta \times \beta)} \circ (
-L_{\alpha}\, x.\, \inr\, x) \; : \; \Nat^{\alpha}\, \mathit{PTree}
-(\alpha \times \alpha)\,(\mathit{PTree}\,\alpha)\\ 
-\vdash L_{\alpha} p.\, (\pi_2 p, \pi_1 p) :
-\Nat^{\alpha} (\alpha \times \alpha)\, (\alpha \times \alpha)\\
-\vdash L_{\alpha} t. \,\case{t}{b \mapsto
- \mathit{pleaf}\, b}{t' \mapsto \mathit{pnode}\, 
-    (((\map_{\mathit{PTree}\,\alpha}^{\alpha \times \alpha, \alpha \times
-      \alpha})_\emptyset\, \mathit{swap})_\alpha\, t')}
-:  \Nat^{\alpha} (\alpha + \mathit{PTree}\,(\alpha \times
-\alpha))\; \mathit{PTree}\,\alpha
-\end{array}\]
-respectively. We can similarly represent the $\mathtt{reverseBush}$
-function from Figure~\ref{fig:funs2} as
-\[\vdash (\fold^{\mathit{Bush}\,\alpha}_{\onet + \beta \times \phi (\phi \beta)})_\emptyset\, balg
-: \Nat^\alpha (\mathit{Bush}\,\alpha)\,(\mathit{Bush}\,\alpha)\]
-%\[\vdash \fold_{\onet + \beta \times
-%\phi (\phi\beta)}^{\mathit{Bush}\,\alpha}\, ( \tin_{\onet + \beta \times
-%  \phi (\phi\beta)} \circ (\onet + t \circ i \circ i') ) :
-%\Nat^{\alpha} (\mathit{Bush}\,\alpha)\,(\mathit{Bush}\,\alpha)\] 
-where
-\[\begin{array}{lll}
-\vdash \fold_{\onet + \beta \times \phi
-  (\phi\beta)}^{\mathit{Bush}\,\alpha} : \Nat^{\emptyset}\, (\Nat^{\alpha}\, 
-(\onet + \alpha \times \mathit{Bush}\, (\mathit{Bush}\, \alpha))) \;
-(\mathit{Bush}\,\alpha))\; (\Nat^{\alpha} \,(\mathit{Bush}\,\alpha) \;
-(\mathit{Bush}\,\alpha))\\
-\end{array}\]
-and $\mathit{bnil}$, $\mathit{bcons}$, $\tin^{-1}_{\onet + \beta
-  \times \phi (\phi\beta)}$, $\mathit{balg}$, and $\mathit{consalg}$
-are the terms
-\[\begin{array}{l}
-\vdash \tin_{\onet + \beta \times \phi (\phi\beta)} \circ (
-L_{\alpha}\, x.\, \inl\, x) \; : \; \Nat^{\alpha}\, \onet\;
-(\mathit{Bush}\,\alpha)\\ 
-\vdash \tin_{\onet + \beta \times \phi (\phi\beta)} \circ (
-L_{\alpha}\, x.\, \inr\, x ) \; : \; \Nat^{\alpha}\, (\alpha \times
-\mathit{Bush}\,(\mathit{Bush}\,\alpha))\; (\mathit{Bush}\,\alpha)\\[0.05in]
-\vdash (\fold_{\onet + \beta \times \phi (\phi\beta)}^{(\onet + \beta
-  \times \phi (\phi\beta))[\phi := \mathit{Bush}\,\alpha]})_\emptyset\,
-((\map_{\onet + \beta \times \phi (\phi\beta)}^{(\onet + \beta \times
-  \phi (\phi\beta))[\phi := \mathit{Bush}\,\alpha][\beta := \alpha],
-  \mathit{Bush}\,\alpha})_\emptyset\, \tin_{\onet + \beta \times \phi
-  (\phi\beta)})\\[0.05in]
-\hspace*{0.2in} : \Nat^{\alpha}\, (\mathit{Bush}\,\alpha)\; (\onet +
-\alpha \times \mathit{Bush}\,(\mathit{Bush}\,\alpha))\\ 
-%\vdash L_{\alpha}\, (b, s). \,\mathsf{case}\;s\,\mathsf{of}\, \{ \hspace*{0.21in}
-%        \ast \mapsto
-%\mathit{bcons}_{\alpha}\, b\, (\mathit{bnil}_{\alpha}\, \ast); \\
-%\hspace*{1.15in}\,(s', u) \mapsto \mathsf{case}\,{s'}\,\mathsf{of}\,
-%\{\hspace*{0.3in} \ast
-%\mapsto \mathit{bcons}_{\alpha}\, b \,(\mathit{bcons}_{\mathit{Bush}\,
-%  \alpha}\, (\mathit{bnil}_{\alpha}\, \ast)\, u); \\
-%\hspace*{2.3in}(b', u') \mapsto \mathit{bcons}_{\alpha}\, b'
-%(\mathit{bcons}_{\mathit{Bush}\, \alpha}\, (\mathit{bcons}_{\alpha}\,
-%b\, u )\, u')\}\}\\
-%\hspace{0.2in} : \Nat^{\alpha} (\alpha \times (\onet + (\onet + \alpha
-%\times \mathit{Bush}\, (\mathit{Bush} \, \alpha))) \times
-%\mathit{Bush}\, (\mathit{Bush} \, (\mathit{Bush}\, \alpha))) \;
-%(\alpha \times \mathit{Bush}\, (\mathit{Bush} \, \alpha))\\
-%\vdash \alpha \times (\onet + \tin^{-1}_{\onet + \beta \times \phi
-%  (\phi\beta)} \times \mathit{Bush}\, (\mathit{Bush} \,
-%(\mathit{Bush}\, \alpha)))\\
-%\hspace{0.2in} : \Nat^{\alpha} (\alpha \times (\onet + \mathit{Bush}
-%\, \alpha \times \mathit{Bush}\,( \mathit{Bush} \, (\mathit{Bush}\,
-%\alpha))))\\
-%\hspace*{0.57in}(\alpha \times (\onet + (\onet + \alpha \times
-%\mathit{Bush}\, (\mathit{Bush} \, \alpha)) \times \mathit{Bush}\,
-%(\mathit{Bush} \, (\mathit{Bush}\, \alpha))))\\
-%\vdash \alpha \times (L_{\alpha}\,x.\, (\tin^{-1}_{\onet +
-%  \beta \times \phi (\phi\beta)})_{\mathit{Bush}\,\alpha}\, x)\\
-%\hspace{0.2in} : \Nat^{\alpha} (\alpha
-%\times \mathit{Bush} \, (\mathit{Bush}\, \alpha)) \; (\alpha \times (\onet
-%+ \mathit{Bush} \, (\alpha \times \mathit{Bush}\, (\mathit{Bush} \,
-%(\mathit{Bush}\, \alpha)))))
-%\[\begin{array}{l}
-  \vdash L_{\alpha} \, s. 
-  \case{s}{* \mapsto \mathit{bnil}_\alpha *}{(a, bba) \mapsto consalg_\alpha (a, bba)}
-  : \Nat^\alpha \, (\onet + \alpha \times \mathit{Bush} (\mathit{Bush} \,\alpha)) \, (\mathit{Bush}\,\alpha) \\
-  \vdash L_{\alpha} \, (a, bba).  
-  \mathsf{case}\, (\ininv{\mathit{Bush}\,\alpha}{bba})\, \mathsf{of}\,\{
-  \\
-  \hspace{8em}\, * \mapsto \mathit{bcons}_{\alpha} (a,\mathit{bnil}_{\mathit{Bush}\,\alpha} *);  \\
-  \hspace{8em} (ba, bbba) \mapsto \mathsf{case}\, (\ininv{\alpha}{ba})
-  \, \mathsf{of}\,\{  \\ 
-  \hspace{16em}* \mapsto \mathit{bcons}_\alpha (a, \mathit{bcons}_{\mathit{Bush}\,\alpha} 
-        (\mathit{bnil}_\alpha *, bbba)) ;  \\ 
-  \hspace{16em}(a', bba') \mapsto \mathit{bcons}_\alpha (a', \mathit{bcons}_{\mathit{Bush}\,\alpha}
-  (\mathit{bcons}_\alpha (a, bba'), bbba)) \} \! \}\\    
-%  \quad \}  
-\hspace*{0.2in} : \Nat^\alpha \, (\alpha \times \mathit{Bush} (\mathit{Bush} \,\alpha)) \, (\mathit{Bush}\,\alpha) \\
-\end{array}\]
-respectively.
-%Here, $\Gamma; \emptyset \,|\, \Delta \vdash H+
-%\eta : \Nat^{\ol{\alpha}} (H+ F)\; (H+ G)$ and $\Gamma;
-%\emptyset \,|\, \Delta \vdash H\times \eta : \Nat^{\ol{\alpha}}
-%(H\times F)\; (H\times G)$ for $H+ \eta :=
-%L_{\ol{\alpha}}\, x.\, \case{x}{s \mapsto s}{t
-%  \mapsto  
-%  (\eta_{\ol{\alpha}} t)}$ and $H\times \eta := L_{\ol{\alpha}}\,
-%x.\, (\pi_1 x, \eta_{\ol{\alpha}} (\pi_2 x))$ for $\Gamma; \emptyset
-%\,|\, \Delta \vdash \eta : \Nat^{\ol{\alpha}} F\; G$ and $\Gamma;
-%\ol{\alpha} \vdash H$.
-
-Our rule for forming $\Nat$-types means that 
-%Because our calculus is designed to ensure functoriality, and
-%functoriality does not interact with polymorphism as well as might be
-%hoped,
-our calculus cannot express types of recursive functions, such as a
-concatenation function for perfect trees, that take as inputs a nested
-type and an argument of another type, both of which are parameterized
-over the same variable $\alpha$.
-%not just one nested type parameterized over a variable $\alpha$ as
-%input, but other inputs parameterized over $\alpha$ as well.
-For the same reason, our calculus cannot express, e.g., the type of a
-zip function for bushes. The fundamental
-issue is that recursion is expressible only via folds, and a fold over
-a nested type must take a natural transformation as its argument and
-return a natural transformation as its result.
-%and, unfortunately, the polymorphism natural transformations entail
-%leaves no nontrivial way to incorporate data from the second argument
-%to a function with a type like that above into a fold over its first
-%argument.
-In fact, even for ADTs there is a difference between which folds over
-them we can type when ADTs are viewed as ADTs --- i.e., as fixpoints
-of types interpreted as first-order functors --- and which folds we
-can type when ADTs are viewed as proper nested types --- i.e., as
-fixpoints of types that are interpreted as higher-order functors ---
-because the arity of the recursion variable $\phi$ in the fixpoint in
-the return type of $\fold$ must equal the number of variables bound by
-its $\Nat$ construct. For ADTs this number can be $0$, making it
-possible to write types such as $\alpha; \emptyset \vdash
-\Nat^\emptyset \ (\List\,\alpha)\,(\Nat^\emptyset \alpha\,(\List (\mu
-\beta. 1 + \beta)))$, all of whose argument types are parameterized
-over the same type $\alpha$, which are not possible for nested types
-even when they are semantically equivalent to ADTs.
-%Currying the type of the function does not help either, since the
-%codomain of a fold applied to an algebra must be a functor, and this
-%clearly is not the case for a function on nested types that takes
-%more than one input involving the same type variable.
-%
-%For example, we can represent the type of a zip-like function for bushes.
-%
-%A well-known deeper limitation on folds is that a fold over a nested
-%type must take a natural transformation as an argument and return a
-%natural transformation; this issue is discussed in more detail in
-%Section~\ref{sec:conclusion},
-%Unfortunately, the polymorphism natural transformations
-%entail leaves no nontrivial way to incorporate data from the second
-%argument to a function with a type like that above into a fold over
-%its first argument.
-Our calculus can, however, express types of recursive functions
-that take
-%    as arguments
-%one proper nested type and as many other (possibly proper)
-%as many proper nested types as we wish
-multiple nested types as arguments, provided those nested types are
-parameterized over disjoint sets of type variables and the return type of
-the function is parameterized over only
-%any or all of
-the variables occurring in the type of its final argument.
-Interestingly, even some recursive functions of a single proper
-nested type --- e.g., a $\mathtt{reverseBush}$ function that is a true
-involution --- cannot be expressed as folds because the algebra
-arguments needed to define them are again recursive functions
-%  cannot be defined. This happens, for
-%  example, when the algebra argument to a fold must be a recursive
-%  function
-%, and thus must itself be defined as a fold
-with types of the same problematic forms.
-%More sophisticated combinators for stylized recursion (e.g.,
-%generalized folds)
-%%  ~\cite{bp99})
-%don't mitigate these difficulties, either. Indeed, they appear to be a
-%fundamental consequence of the interaction of functoriality and
-%polymorphism, not just some peculiarity of our particular calculus.
-Allowing extra functorial variables in the codomains of $\Nat$-types
-would recover typeability of, say, zip for bushes, but such extended
-$\Nat$-types need not have semantics as $\omega$-cocontinuous functors
-in $\set$, as will be required in Section~\ref{sec:type-interp}. For
-example,
-%as shown in Example D, Section VI of~\cite{jp19},
-the type $\emptyset; \gamma \vdash \Nat^\emptyset\, (\mu \beta. \onet + \beta)\,
-\gamma$ has no such semantics. To remedy this, we would need to
-construct our models from semantic categories that are locally
-$\lambda$-presentable for $\lambda > \omega$.
-%It is tempting to try to recover typeability of, say, a zip function
-%for bushes by allowing extra functorial variables in the codomain type
-%of $\Nat$-types. But such extended $\Nat$-types need not have
-%semantics as $\omega$-cocontinuous functors, as is required in
-%Section~\ref{sec:type-interp}. For example, as shown in Example D,
-%Section VI of~\cite{jp19}, the type $\emptyset; \gamma \vdash
-%\Nat^\emptyset (\onet + \onet)\, \gamma$ does not.
-%
-Adding more expressive recursion combinators could help, too.
-%enhance the expressivity of our calculus
-But since the expressivity issue for folds for nested types long
-pre-dates, and is somewhat orthogonal to, the issue of parametricity
-for languages that construct nested types as fixpoints, we do not
-consider it further in this paper other than to identify, in
-Section~\ref{sec:conclusion}, some ideas for addressing it in future
-work.
-%These and other extensions are discussed
-
-The presence of the ``extra'' functorial variables in $\ol{\gamma}$ in
-the rules for $\map^{\ol{F},\ol{G}}_H$, $\tin_H$, and $\fold^F_H$
-merit special mention. They allow us to map or fold polymorphic
-functions over nested types. Consider, for example, the polymorphic
-function $\mathit{flatten} : \Nat^\beta
-(\mathit{PTree}\,\beta)\,(\mathit{List}\,\beta)$ that uniformly maps
-perfect trees to lists. Even in the absence of extra variables the
-instance of $\map$ required to map each non-functorial monomorphic
-instantiation of $\mathit{flatten}$ over a list of perfect trees is
-well-typed:
-\[\begin{array}{l}
-\AXC{$\Gamma;\alpha \vdash \mathit{List} \, \alpha$}
-\AXC{$\Gamma;\emptyset \vdash \mathit{PTree}\,F$}
-\AXC{$\Gamma;\emptyset \vdash \mathit{List}\,G$}
-\TIC{$\Gamma;\emptyset \,|\, \emptyset \vdash \map^{\mathit{PTree}\,F,
-    \,\mathit{List}\,G}_{\mathit{List}\,\alpha} :
-  \Nat^\emptyset (\Nat^\emptyset\,(\mathit{PTree}\, F)\,(\mathit{List}\, G))\,
-  (\Nat^\emptyset\,(\mathit{List}\, (\mathit{PTree}\,
-  F))\,(\mathit{List}\, (\mathit{List}\, G)))$} \DisplayProof
-\end{array}\]
-But in the absence of $\ol \gamma$, the instance
-\[\Gamma;\emptyset~|~\emptyset \vdash
-\map^{\mathit{PTree}\,\beta,\mathit{List}\,\beta}_{\mathit{List}\,\alpha}
-: \Nat^\emptyset ( \Nat^\beta(\mathit{PTree}\,
-\beta)\,(\mathit{List}\, \beta))\, (\Nat^\beta\,(\mathit{List}\,
-(\mathit{PTree}\, 
-\beta))\,(\mathit{List}\, (\mathit{List}\, \beta)))\]
-%\[\Gamma;\emptyset \,|\, x : \Nat^\beta(\mathit{PTree}\,
-%\beta)\,(\mathit{List}\, \beta) \vdash
-%\map^{\mathit{PTree}\,\beta,\mathit{List}\,\beta}_{\mathit{List}\,\alpha}\,x
-%: \Nat^\beta\,(\mathit{List}\,
-%(\mathit{PTree}\, \beta))\,(\mathit{List}\, (\mathit{List}\,
-%\beta))\] of $\mathtt{map}$
-required to map the {\em polymorphic} $\mathit{flatten}$ function over
-a list of perfect trees is not: in that setting the functorial
-contexts for $F$ and $G$ in the rule for $\map^{F,G}_H$ would have to
-be empty, but because the polymorphic $\mathit{flatten}$ function is
-natural in some variable, say $\delta$, it cannot possibly have a
-type of the form $\Nat^\emptyset F\, G$ that would be required for it
-to be the function input to $\map$. Since untypeability of this
-instance of $\map$ is unsatisfactory in a polymorphic calculus, where
-we naturally expect to be able to manipulate entire polymorphic
-functions rather than just their monomorphic instances, we use the
-``extra'' variables in $\ol \gamma$ to remedy the
-situation. Specifically, Definition~\ref{def:well-formed-terms}
-ensures that the instance of $\map$ needed to map the polymorphic
-$\mathit{flatten}$ function is typeable as follows:
-\[\begin{array}{l}
-\AXC{$\Gamma;\alpha,\beta \vdash \mathit{List} \, \alpha$}
-\AXC{$\Gamma;\beta \vdash \mathit{PTree}\,\beta \hspace*{0.3in}
-  \Gamma;\beta \vdash \mathit{List}\,\beta$}
-\BIC{$\Gamma;\emptyset~|~\emptyset \vdash
-  \map^{F,G}_{\mathit{List}\,\alpha}  : \Nat^\emptyset
-  (\Nat^\beta(\mathit{PTree}\, 
-  \beta)\,(\mathit{List}\, \beta))\,
- (\Nat^\beta\,(\mathit{List}\,
-  (\mathit{PTree}\, \beta))\,(\mathit{List}\, (\mathit{List}\,
-  \beta)))$}
-\DisplayProof
-\end{array}\]
-Similar remarks explain the appearance of $\ol \gamma$ in the typing
-rules for $\tin$ and $\fold$.
-
-\vspace*{0.05in}
-
-\section{Interpreting Types}\label{sec:type-interp}
-
-We denote the category of sets and functions by $\set$. The category
-$\rel$ has as its objects triples $(A,B,R)$ where $R$ is a relation
-between the objects $A$ and $B$ in $\set$, i.e., a subset of $A \times
-B$, and has as its morphisms from $(A,B,R)$ to $(A',B',R')$ pairs $(f
-: A \to A',g : B \to B')$ of morphisms in $\set$ such that $(f a,g\,b)
-\in R'$ whenever $(a,b) \in R$. We write $R : \rel(A,B)$ in place of
-$(A,B,R)$ when convenient.  If $R : \rel(A,B)$ we write $\pi_1 R$ and
-$\pi_2 R$ for the {\em domain} $A$ of $R$ and the {\em codomain} $B$
-of $R$, respectively.  If $A : \set$, then we write $\Eq_A =
-(A,A,\{(x,x)~|~ x \in A\})$ for the {\em equality relation} on $A$.
-
-The key idea underlying Reynolds' parametricity is to give each type
-$F(\alpha)$ with one free variable $\alpha$ both an {\em object
-  interpretation} $F_0$ taking sets to sets and a \emph{relational
-  interpretation} $F_1$ taking relations $R : \rel(A,B)$ to relations
-$F_1 (R) : \rel(F_0 (A), F_0 (B))$, and to interpret each term
-$t(\alpha,x) : F(\alpha)$ with one free term variable $x : G(\alpha)$
-as a map $t_0$ associating to each set $A$ a function $t_0(A) : G_0(A)
-\to F_0(A)$. These interpretations are to be given inductively on the
-structures of $F$ and $t$ in such a way that they imply two
-fundamental theorems. The first is an \emph{Identity Extension Lemma},
-which states that $F_1(\Eq_A) = \Eq_{F_0(A)}$, and is the essential
-property that makes a model relationally parametric rather than just
-induced by a logical relation. The second is an \emph{Abstraction
-  Theorem}, which states that, for any $R :\rel(A, B)$,
-$(t_0(A),t_0(B))$ is a morphism in $\rel$ from
-$(G_0(A),G_0(B),G_1(R))$ to $(F_0(A),F_0(B),F_1(R))$. The Identity
-Extension Lemma is similar to the Abstraction Theorem except that it
-holds for {\em all} elements of a type's interpretation, not just
-those that are interpretations of terms.
-%i.e., $t_0(A)$ and $t_0(B)$ map related arguments to
-%related results.
-Similar theorems are expected to hold for types and terms with any
-number of free variables.
-
-The key to proving the Identity Extension Lemma
-(Theorem~\ref{thm:iel}) in our setting is a familiar ``cutting down''
-of the interpretations of universally quantified types to include only
-the ``parametric'' elements; the relevant types in our calculus are
-the $\Nat$-types.
-%(See, e.g.,~\cite{atk12,bfss90,gjfor15,rey83}).
-This cutting down requires, as usual, that the set interpretations of
-our types are defined simultaneously with their relational
-interpretations. We give the set interpretations for our types in
-Section~\ref{sec:set-interp} and give their relational interpretations
-in Section~\ref{sec:rel-interp}.  While the set interpretations are
-relatively straightforward, their relation interpretations are less
-so, mainly because of the cocontinuity conditions required to ensure
-that they are well-defined. We develop these conditions in
-Section~\ref{sec:rel-interp}, which separates
-Definitions~\ref{def:set-sem} and~\ref{def:rel-sem} in space, but
-otherwise has no impact on the fact that they are given by mutual
-induction.
-
-\subsection{Interpreting Types as Sets}\label{sec:set-interp}
-
-%A poset $\mcD = (D,\leq)$ is \emph{$\omega$-directed} if every finite
-%subset of $D$ has an upper bound. When $\mcD$ is considered as a
-%category, we write $d \in \mcD$ to indicate that $d$ is an object of
-%$\mcD$ (i.e., $d \in D$). An {\em $\omega$-directed colimit} in a
-%category $\mcC$ is a colimit of a diagram $F : {\mathcal D} \to \mcC$,
-%where $\mathcal D$ is an $\omega$-directed poset. A category $\mcC$ is
-%{\em $\omega$-cocomplete} if it has all $\omega$-directed colimits. A
-%cocomplete category is one that has all colimits.
-%
-%If $\mcA$ and $\mcC$ are $\omega$-cocomplete, then the functor $F :
-%\mcA \to \mcC$ is {\em $\omega$-cocontinuous} if it preserves
-%$\omega$-directed colimits.  If $\mcA$ is locally small, then an
-%object $A$ of $\mcA$ is {\em finitely presentable} if the functor
-%$\mathsf{Hom}_\mcA(A, -) : \mcA \to \Set$ preserves $\omega$-directed
-%colimits, i.e., if for every $\omega$-directed poset $\mathcal D$ and
-%every functor $F : {\mathcal D} \to \mcC$, there is a canonical
-%isomorphism $\colim{d \in \mcD}{\mathsf{Hom}_\mcA(A,Fd)} \simeq
-%\mathsf{Hom}_{\mcA}(A, \colim{d \in \mcD}{Fd})$. A category $\mcA$ is
-%       {\em finitely accessible} if it is $\omega$-cocomplete and has
-%       a set $\mcA_0$ of finitely presentable objects such that every
-%%       object is an $omega$-directed colimit of objects in $\mcA_0$;
-%       it is {\em locally finitely presentable} if it is finitely
-%       accessible and cocomplete.
-We interpret types in our calculus as $\omega$-cocontinuous functors
-on locally finitely presentable categories~\cite{ar94}. Since functor
-categories of locally finitely presentable categories are again
-locally finitely presentable, this ensures that the fixpoints
-interpreting $\mu$-types in $\set$ and $\rel$ exist, and thus that
-both the set and relational interpretations of all of the types in
-Definition~\ref{def:wftypes} are well-defined~\cite{jp19}. To
-bootstrap this process, we interpret type variables as
-$\omega$-cocontinuous functors in Definitions~\ref{def:set-env}
-and~\ref{def:reln-env}. If $\C$ and $\D$ are locally finitely
-presentable categories, we write $[\C,\D]$ for the category of
-$\omega$-cocontinuous functors from $\C$ to $\D$.
-%Note that the categories $\set$ and $\rel$ are both locally finitely
-%presentable.
-
-\begin{dfn}\label{def:set-env}
-A {\em set environment} maps each type variable in $\tvars^k \cup
-\fvars^k$ to an element of $[\set^k,\set]$.  A morphism $f : \rho \to
-\rho'$ for set environments $\rho$ and $\rho'$ with $\rho|_\tvars =
-\rho'|_\tvars$ maps each type constructor variable $\psi^k \in \tvars$
-to the identity natural transformation on $\rho \psi^k = \rho'\psi^k$
-and each functorial variable $\phi^k \in \fvars$ to a natural
-transformation from the $k$-ary functor $\rho \phi^k$ on $\set$ to the
-$k$-ary functor $\rho' \phi^k$ on $\set$.  Composition of morphisms on
-set environments is given componentwise, with the identity morphism
-mapping each set environment to itself. This gives a category of set
-environments and morphisms between them, which we denote $\setenv$.
-\end{dfn}
-When convenient we identify a functor in $[\set^0, \set]$ with its
-value on $\ast$ and consider a set environment to map a type
-variable of arity $0$ to
-%an $\omega$-cocontinuous functor from $\set^0$ to $\set$, i.e., to
-a set.  If $\ol{\alpha} = \{\alpha_1,...,\alpha_k\}$ and $\ol{A} =
-\{A_1,...,A_k\}$, then we write $\rho[\ol{\alpha := A}]$ for the set
-environment $\rho'$ such that $\rho' \alpha_i = A_i$ for $i = 1,...,k$
-and $\rho' \alpha = \rho \alpha$ if $\alpha \not \in
-\{\alpha_1,...,\alpha_k\}$.  If $\rho$ is a set environment we write
-$\Eq_\rho$ for the relation environment (see
-Definition~\ref{def:reln-env}) such that $\Eq_\rho v = \Eq_{\rho v}$
-for every type variable $v$.
-%; see Definition~\ref{def:reln-env} for the complete definition of a
-%relation environment.
-The relational interpretations appearing in the second clause of
-Definition~\ref{def:set-sem} are given in full in
-Definition~\ref{def:rel-sem}.
-
-\begin{dfn}\label{def:set-sem}
-%Let $\rho$ be a set environment.
-The {\em set interpretation} $\setsem{\cdot} : \F \to [\setenv, \set]$
-is defined by
-\begin{align*}
-  \setsem{\Gamma;\Phi \vdash \zerot}\rho &= 0\\
-  \setsem{\Gamma;\Phi \vdash \onet}\rho &= 1\\
-%  \setsem{\Gamma;\emptyset \vdash v} \rho &= \rho v \mbox{
-%    if } v \in \tvars^0 \\ 
-  \setsem{\Gamma; \emptyset
-    \vdash \Nat^{\ol{\alpha}}
-    \,F\,G}\rho &= \{\eta : \lambda \ol{A}. \,\setsem{\Gamma;
-    \ol{\alpha} \vdash
-    F}\rho[\ol{\alpha := A}] 
-      \Rightarrow \lambda \ol{A}.\,\setsem{\Gamma; 
-        \ol{\alpha} \vdash G}\rho[\ol{\alpha := A}] \\ 
-      &\hspace{0.3in}|~\forall \overline{A}, \overline{B} :
-      \set. \forall \overline{R : \rel(A, B)}.\\ 
-      &\hspace{0.4in}(\eta_{\overline{A}}, \eta_{\overline{B}})
-      : \relsem{\Gamma; \ol{\alpha} \vdash F}\Eq_{\rho}[\ol{\alpha := R}]
-      \rightarrow \relsem{\Gamma; \ol{\alpha} \vdash
-        G}\Eq_{\rho}[\ol{\alpha := R}] \} \\
-  \setsem{\Gamma;\Phi \vdash \phi\ol{F}}\rho &=
-  (\rho\phi)\,\ol{\setsem{\Gamma;\Phi \vdash
-    F}\rho}\\
-  \setsem{\Gamma;\Phi \vdash F+G}\rho &=
-  \setsem{\Gamma;\Phi \vdash F}\rho +
-  \setsem{\Gamma;\Phi \vdash G}\rho\\
-  \setsem{\Gamma;\Phi \vdash F\times G}\rho &=
-  \setsem{\Gamma;\Phi \vdash F}\rho \times
-  \setsem{\Gamma;\Phi \vdash G}\rho\\ 
-  \setsem{\Gamma;\Phi,\ol\gamma \vdash (\mu \phi.\lambda
-    \ol{\alpha}. H)\ol{G}}\rho &= (\mu
-    T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho}\\
-    \text{where } T^\set_{H,\rho}\,F & = \lambda
-  \ol{A}. \setsem{\Gamma;\ol\gamma,\phi, \ol{\alpha} \vdash
-    H}\rho[\phi :=  F][\ol{\alpha := A}]\\
-  \text{and } T^\set_{H,\rho}\,\eta &= \lambda
-  \ol{A}. \setsem{\Gamma;\ol\gamma,\phi, \ol{\alpha} \vdash
-    H}\id_\rho[\phi := \eta][\ol{\alpha := \id_{A}}]
-\end{align*}
-\end{dfn}
-%The interpretations in Definition~\ref{def:set-sem} respect weakening,
-%i.e., a type and its weakenings have the same set interpretations.
-%
-%= \setsem{\Gamma \vdash \sigma} \rho \to
-%\setsem{\Gamma \vdash \tau} \rho$, as expected {\color{blue} by IEL}.
-If $\rho \in \setenv$ and $\vdash F$ then we write $\setsem{\vdash F}$
-instead of $\setsem{\vdash F}\rho$ since the environment is
-immaterial. The third clause of Definition~\ref{def:set-sem} does
-indeed define a set: local finite presentability of $\set$ and
-$\omega$-cocontinuity of $\setsem{\Gamma;\ol{\alpha} \vdash F}\rho$
-ensure that $\{\eta : \setsem{\Gamma;\ol{\alpha} \vdash F}\rho
-\Rightarrow \setsem{\Gamma;\ol{\alpha} \vdash G}\rho\}$ (which
-contains $\setsem{\Gamma;\emptyset \vdash
-  \Nat^{\ol{\alpha}}\,F\,G}\rho$) is a subset of
-$\big\{({\setsem{\Gamma;\ol{\alpha} \vdash G}\rho[\ol{\alpha :=
-      S}]})^{(\setsem{\Gamma;\ol{\alpha} \vdash F}\rho[\ol{\alpha :=
-      S}])}~\big|~ \ol{S} = (S_1,...,S_{|\ol{\alpha}|}), \mbox{ and }$
-$S_i \mbox{ is a finite set for } i =
-1,...,|\ol{\alpha}|\big\}$. There are countably many choices of tuples
-$\ol{S}$, each of which gives rise to a morphism from
-${\setsem{\Gamma;\ol{\alpha} \vdash F}\rho[ \ol{\alpha := S}]}$ to
-${\setsem{\Gamma;\ol{\alpha} \vdash G}\rho[\ol{\alpha := S}]}$. But
-there are only $\set$-many choices of morphisms between any two
-objects since $\set$ is locally small.  We note that $\setsem{\Gamma;
-  \emptyset \vdash \Nat^{\ol\alpha} F\,G}$ is $\omega$-cocontinuous because
-it is a constant functor (and that allowing functorial variables not
-in $\ol \alpha$ in $G$ fails here, as shown in
-Section~\ref{sec:terms}).
-%follows from $\omega$-cocontinuity of $\setsem{\Gamma;\ol\alpha
-%\vdash F}$, $\setsem{\Gamma;\ol\alpha \vdash G}$, and the equality
-%relation environment $\Eq$.
-Interpretations of $\Nat$-types ensure
-that $\setsem{\Gamma \vdash F \to G}$ and $\setsem{\Gamma \vdash
-  \forall \ol\alpha. F}$ are as expected in any parametric model.
-
-To make sense of the last clause in
-Definition~\ref{def:set-sem}, we need to know that, for each $\rho \in
-\setenv$, $T^\set_{H,\rho}$ is an $\omega$-cocontinuous endofunctor on
-$[\set^k, \set]$, and thus admits a fixpoint.  Since
-$T_{H,\rho}^\set$ is defined in terms of $\setsem{\Gamma;\ol\gamma,\phi,
-  \ol{\alpha} \vdash H}$, this means that interpretations of types
-must be such functors, which in turn means that the actions of set
-interpretations of types on objects and on morphisms in $\setenv$ are
-intertwined. Fortunately, we know from~\cite{jp19} that, for every
-$\Gamma; \ol{\alpha} \vdash G$, $\setsem{\Gamma; \ol{\alpha} \vdash
-  G}$ is actually in $[\set^k,\set]$ where $k = |\ol \alpha|$. This
-means that for each $\setsem{\Gamma; \ol\gamma, \phi^k, \ol{\alpha} \vdash
-  H}$, the corresponding operator $T^\set_{H}$ can be extended to a
-{\em functor} from $\setenv$ to $[[\set^k,\set],[\set^k,\set]]$. The
-action of $T^\set_H$ on an object $\rho \in \setenv$ is given by the
-higher-order functor $T_{H,\rho}^\set$, whose actions on objects
-(functors in $[\set^k, \set]$) and morphisms (natural transformations
-between such functors) are given in Definition~\ref{def:set-sem}. Its
-action on a morphism $f : \rho \to \rho'$ is the higher-order natural
-transformation $T^\set_{H,f} : T^\set_{H,\rho} \to T^\set_{H,\rho'}$
-whose action on $F : [\set^k,\set]$ is the natural transformation
-$T^\set_{H,f}\, F : T^\set_{H,\rho}\,F \to T^\set_{H,\rho'}\,F$ whose
-component at $\ol{A}$ is $(T^\set_{H,f}\, F)_{\ol{A}} =
-\setsem{\Gamma; \ol\gamma,\phi,\ol{\alpha} \vdash H}f[\phi :=
-  \id_F][\ol{\alpha := \id_A}]$. The next definition uses the functor
-$T^\set_H$ to define the actions of functors interpreting types on
-morphisms between set environments.
-
-\begin{comment}
-To make sense of the last clause in
-Definition~\ref{def:set-sem}, we need to know that, for each $\rho \in
-\setenv$, $T^\set_{H,\rho}$ is an $\omega$-cocontinuous endofunctor on
-$[\set^k, \set]$, and thus admits a fixpoint.  Since
-$T_{H,\rho}^\set$ is defined in terms of $\setsem{\Gamma;\ol\gamma,\phi,
-  \ol{\alpha} \vdash H}$, this means that interpretations of types
-must be such functors, which in turn means that the actions of set
-interpretations of types on objects and on morphisms in $\setenv$ are
-intertwined. Fortunately, we know from~\cite{jp19} that, for every
-$\Gamma; \ol{\alpha} \vdash G$, $\setsem{\Gamma; \ol{\alpha} \vdash
-  G}$ is actually in $[\set^k,\set]$ where $k = |\ol \alpha|$. This
-means that for each $\setsem{\Gamma; \ol\gamma, \phi^k, \ol{\alpha} \vdash
-  H}$, the corresponding operator $T^\set_{H}$ can be extended to a
-{\em functor} from $\setenv$ to $[[\set^k,\set],[\set^k,\set]]$. The
-action of $T^\set_H$ on an object $\rho \in \setenv$ is given by the
-higher-order functor $T_{H,\rho}^\set$, whose actions on objects
-(functors in $[\set^k, \set]$) and morphisms (natural transformations
-between such functors) are given in Definition~\ref{def:set-sem}. Its
-action on a morphism $f : \rho \to \rho'$ is the higher-order natural
-transformation $T^\set_{H,f} : T^\set_{H,\rho} \to T^\set_{H,\rho'}$
-whose action on $F : [\set^k,\set]$ is the natural transformation
-$T^\set_{H,f}\, F : T^\set_{H,\rho}\,F \to T^\set_{H,\rho'}\,F$ whose
-component at $\ol{A}$ is $(T^\set_{H,f}\, F)_{\ol{A}} =
-\setsem{\Gamma; \ol\gamma,\phi,\ol{\alpha} \vdash H}f[\phi :=
-  \id_F][\ol{\alpha := \id_A}]$. The next definition uses the functor
-$T^\set_H$ to define the actions of functors interpreting types on
-morphisms between set environments.
-\end{comment}
-
-\begin{dfn}\label{def:set-sem-funcs}
-Let $f: \rho \to \rho'$ be a morphism between set environments $\rho$
-and $\rho'$ (so that $\rho|_\tvars = \rho'|_\tvars$). The action
-$\setsem{\Gamma;\Phi \vdash F}f$ of\, $\setsem{\Gamma;\Phi \vdash F}$
-on the morphism $f$ is given as follows:
+\vspace*{0.2in}
+
+Things to check:
 \begin{itemize}
-\item If \,$\Gamma;\Phi \vdash \zerot$ then $\setsem{\Gamma;\Phi \vdash
-  \zerot}f = \id_0$
-\item If \,$\Gamma;\Phi \vdash \onet$ then $\setsem{\Gamma;\Phi \vdash
-  \onet}f = \id_1$
-%\item If \,$\Gamma,v;\emptyset \vdash v$ then
-%  $\setsem{\Gamma,v;\emptyset \vdash v}f = \id_{\rho v}$
-\item If \,$\Gamma; \emptyset
-  \vdash \Nat^{\ol{\alpha}}\,F\,G$ then
-  $\setsem{\Gamma; \emptyset
-    \vdash \Nat^{\ol{\alpha}}\,F\,G} f =
-  \id_{\setsem{\Gamma; \emptyset
-      \vdash \Nat^{\ol{\alpha}}\,F\,G}\rho}$
-%\hspace*{0.2in} $\lambda t : \setsem{\Gamma;\emptyset\vdash
-%    \Nat^{\ol{\alpha}}\,F\,G}\rho.\, 
-%  (\lambda \ol{A}. \setsem{\Gamma; \ol\alpha \vdash G}f[\ol{\alpha
-%      := id_A}]) \circ t$  
-\item If \,$\Gamma;\Phi \vdash \phi \ol{F}$ then
-$\setsem{\Gamma;\Phi \vdash \phi \ol{F}} f : \setsem{\Gamma;\Phi
-  \vdash \phi \ol{F}}\rho \to \setsem{\Gamma;\Phi \vdash
-  \phi\ol{F}}\rho' = (\rho\phi) \ol{\setsem{\Gamma;\Phi \vdash
-    F}\rho} \to (\rho'\phi) \ol{\setsem{\Gamma;\Phi \vdash
-    F}\rho'}$ is defined by $\setsem{\Gamma;\Phi \vdash \phi
-  \ol{F}} f = (f\phi)_{\ol{\setsem{\Gamma;\Phi \vdash
-      F}\rho'}}\, \circ\, (\rho\phi) {\ol{\setsem{\Gamma;\Phi
-      \vdash F}f}} = (\rho'\phi) {\ol{\setsem{\Gamma;\Phi \vdash
-      F}f}}\, \circ\, (f \phi)_{\ol{\setsem{\Gamma;\Phi \vdash
-      F}\rho}}$.  The latter equality holds because $\rho\phi$ and
-  $\rho'\phi$ are functors and $f\phi : \rho\phi \to \rho'\phi$ is a
-  natural transformation, so the following naturality square commutes:
-{\footnotesize\begin{equation}\label{eq:cd2}
-\begin{CD}
-  (\rho\phi) \ol{\setsem{\Gamma;\Phi \vdash F}\rho} @> (f\phi)_{
-    \ol{\setsem{\Gamma;\Phi \vdash F}\rho}} >> (\rho'\phi)
-  \ol{\setsem{\Gamma;\Phi \vdash F}\rho} \\ @V(\rho\phi)
-  \ol{\setsem{\Gamma;\Phi \vdash F}f}VV @V (\rho'\phi)
-  \ol{\setsem{\Gamma;\Phi \vdash F}f} VV \\ (\rho\phi)
-  \ol{\setsem{\Gamma;\Phi \vdash F}\rho'} @>(f\phi)_{
-    \ol{\setsem{\Gamma;\Phi \vdash F}\rho'}}>> (\rho'\phi)
-  \ol{\setsem{\Gamma;\Phi \vdash F}\rho'}
-\end{CD}
-\end{equation}}
-\item If\, $\Gamma;\Phi \vdash F + G$ then $\setsem{\Gamma;\Phi
-  \vdash F + G}f$ is defined by $\setsem{\Gamma;\Phi \vdash
-  F + G}f(\inl\,x) = \inl\,(\setsem{\Gamma;\Phi \vdash
-  F}f x)$ and $\setsem{\Gamma;\Phi \vdash F +
-  G}f(\inr\,y) = \inr\,(\setsem{\Gamma;\Phi \vdash G}f y)$
-\item If \,$\Gamma;\Phi\vdash F \times G$ then
-  $\setsem{\Gamma;\Phi \vdash F \times G}f = 
-  \setsem{\Gamma;\Phi \vdash F}f \times \setsem{\Gamma;\Phi \vdash
-    G}f$
-\item If \,$\Gamma;\Phi,\ol\gamma \vdash (\mu \phi.\lambda
-  \ol{\alpha}. H)\ol{G}$ then
-  \[\begin{array}{lll}
-  \setsem{\Gamma;\Phi,\ol\gamma \vdash (\mu  \phi.\lambda
-    \ol{\alpha}. H)\ol{G}} f &: &\setsem{\Gamma;\Phi,\ol\gamma 
-    \vdash (\mu \phi.\lambda \ol{\alpha}. H)\ol{G}} \rho \to
-  \setsem{\Gamma;\Phi,\ol\gamma \vdash (\mu
-    \phi.\lambda\ol{\alpha}. H)\ol{G}} \rho'\\
-  &= &(\mu
-  T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho} \to (\mu
-  T^\set_{H,\rho'})\ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho'}
-  \end{array}\]
-  is
-  defined by
-  \[\begin{array}{ll} & (\mu T^\set_{H,f})\ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash
-      G}\rho'} \circ (\mu T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi,\ol\gamma
-      \vdash G}f}\\ = & (\mu T^\set_{H,\rho'})\ol{\setsem{\Gamma;\Phi,\ol\gamma
-      \vdash G}f} \circ (\mu T^\set_{H,f})\ol{\setsem{\Gamma;\Phi,\ol\gamma
-      \vdash G}\rho}\end{array}\]  The latter equality holds because $\mu
-  T^\set_{H,\rho}$ and $\mu T^\set_{H,\rho'}$ are functors and $\mu
-  T_{H,f}^\set : \mu T_{H,\rho}^\set \to \mu T_{H,\rho'}^\set$ is a
-  natural transformation, so the following naturality square commutes:
-{\footnotesize\begin{equation}\label{eq:cd3}
-\begin{CD}
- (\mu T^\set_{H,\rho}) \ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho} @> (\mu
-  T^\set_{H,f})_{\ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho}} >> (\mu
-  T^\set_{H,\rho'}) \ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho} \\ 
- @V(\mu T^\set_{H,\rho}) \ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}f}VV @V  (\mu
- T^\set_{H,\rho'}) \ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}f} VV \\ 
-(\mu T^\set_{H,\rho}) \ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho'} @>(\mu
- T^\set_{H,f})_{\ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho'}}>> (\mu
- T^\set_{H,\rho'}) \ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho'}
-\end{CD}
-\end{equation}}
+\item \new{Are these $\Nat$-types $\omega$CPOs? An argument similar to the
+  one in the POPL submission should hold.}
+\item \new{Does IEL hold if we allow functoriality in variables of arity $> 0$?}
+\item \new{Does AT hold if we allow functoriality in variables of arity $> 0$?}
 \end{itemize}
-\end{dfn}
-
-Definitions~\ref{def:set-sem} and~\ref{def:set-sem-funcs} respect
-weakening, i.e., ensure that a type and its weakenings have the same
-set interpretations.
-
-\subsection{Interpreting Types as Relations}\label{sec:rel-interp}
-
-\begin{dfn}\label{def:rel-transf}
-A {\em $k$-ary relation transformer} $F$ is a triple $(F^1, F^2,F^*)$,
-where $F^1,F^2 : [\set^k,\set]$ are functors, $F^* : [\rel^k, \rel]$
-is a functor, if $R_1:\rel(A_1,B_1),...,R_k:\rel(A_k,B_k)$, then $F^*
-\ol{R} : \rel(F^1 \ol{A}, F^2 \ol{B})$, and if $(\alpha_1, \beta_1)
-\in \Homrel(R_1,S_1),..., (\alpha_k, \beta_k) \in \Homrel(R_k,S_k)$
-then $F^* \ol{(\alpha, \beta)} = (F^1 \ol{\alpha}, F^2 \ol{\beta})$.
-We define $F\ol{R}$ to be $F^*\overline{R}$ and
-$F\overline{(\alpha,\beta)}$ to be $F^*\overline{(\alpha,\beta)}$.
-\end{dfn}
-The last clause of Definition~\ref{def:rel-transf} expands to: if
-$\ol{(a,b) \in R}$ implies $\ol{(\alpha\,a,\beta\,b) \in S}$ then
-$(c,d) \in F^*\ol{R}$ implies $(F^1 \ol{\alpha}\,c,F^2 \ol{\beta}\,d)
-\in F^*\ol{S}$. When convenient we identify a $0$-ary relation
-transformer $(A,B,R)$ with $R : \rel(A,B)$. We may also write $\pi_1
-F$ for $F^1$ and $\pi_2 F$ for $F^2$. We extend these conventions to
-relation environments, introduced in Definition~\ref{def:reln-env}
-below, in the obvious way.
-
-\begin{dfn}
-The category $RT_k$ of $k$-ary relation transformers is given by the
-following data:
+
+Sum and product intro and elim rules should be annotated with
+  constituent types for consistency.
+
+\vspace*{0.2in}
+
+We should have a computation rule along the lines of: If $\eta :
+\Nat^{\ol\alpha} F \,G[\ol{\beta := K}]$ then
+\[\begin{array}{ll}
+& (\partial_F^{G,\ol K} \eta)_{\ol{K[\ol{\alpha := A}]}} \circ (\int_{K,F})_{\ol
+  A}\; \rightarrow \; \eta_{\ol A}\\
+: & F[\ol{\alpha := A}] \to G[\ol{\beta := K[\ol{\alpha :=A}]}]\\
+= & F[\ol{\alpha := A}] \to G[\ol{\beta := K}][\ol{\alpha := A}]
+\end{array}\]
+
+Some questions/issues:
 \begin{itemize}
-\item An object of $RT_k$ is a relation transformer.
-\item A morphism $\delta : (G^1,G^2,G^*) \to (H^1,H^2,H^*)$ in $RT_k$
-  is a pair of natural transformations $(\delta^1, \delta^2)$ where
-  $\delta^1 : G^1 \to H^1$, $\delta^2 : G^2 \to H^2$ such that, for
-  all $\ol{R : \rel(A, B)}$, if $(x, y) \in G^*\ol{R}$ then
-  $(\delta^1_{\ol{A}}x, \delta^2_{\ol{B}}y) \in H^*\ol{R}$.
-\item Identity morphisms and composition are inherited from the
-  category of functors on $\set$.
+\item Can we write zipBush and appendBush with $\partial$ and $\int$?
+  We could already represent the uncurried type of appendBush
+  (although not its curried type), but couldn't recurse over both
+  input bushes because folds take natural transformations as inputs.
+\item More generally, how do we compute with $\partial$ and $\int?$
+  Can we use the colimit formulation of Lans (see Lemma 6.3.7
+  of~\cite{rie16}) to get a handle on this?
+\item What is the connection between exponentials and natural
+  transformations? (Should we assume only small objects are
+  exponentiable?) Do we want the former or the latter for
+  computational purposes? (I suspect the latter.)
+
+  [From nlab:
+  In a functor category $D^C$ , a natural transformation $\alpha : F \to
+  G$ is exponentiable if (though probably not ``only if'') it is
+  cartesian and each component $\alpha_c : F c \to G c$ is
+  exponentiable in $D$.  Given $H \to F$ we define $(\Pi_\alpha H) c =
+  \Pi_{\alpha_c}(Hc)$; then for $u : c \to c'$ to obtain a map
+  $\Pi_{\alpha_c} (Hc) \to \Pi_{\alpha_{c'}} (Hc')$ we need a map
+  $\alpha^*_{c'}(\Pi_{\alpha_c}(Hc)) \to Hc'$. But since $\alpha$ is
+  cartesian, $\alpha^*_{c'}(\Pi_{\alpha_c}(Hc)) \cong
+  \alpha^*_{c}(\Pi_{\alpha_c}(Hc))$, so we have the counit
+  $\alpha^*_{c}(\Pi_{\alpha_c}(Hc)) \to Hc$ that we can compose with
+  $Hu$.]
+\item After we understand what we can do with Lans and folds on GADTs
+  we might want to try to extend calculus with term-level
+  fixpoints. This would give a categorical analogue for GADTs
+  of~\cite{pit98,pit00} for ADTs. Would it also more accurately
+  reflect how GADTs are used in practice, or are functions over GADTs
+  usually folds? Investigate applications in the literature and/or in
+  implementations.
+\item What kind of category can handle $\Lan$ types? Can $\omega$CPO?
+  This caregory is locally $\omega_1$-presentable (but not locally
+  finitely presentable); see 1.18(2) of~\cite{ar94}). The
+  $\omega_1$-presentable objects are the $\omega$CPOs cardinality
+  less than $\omega_1$, i.e., the countable $\omega$CPOs.
+\item $\omega$CPO is a natural choice for modeling general
+  recursion. We know $(Lan^\gamma_C \onet)D$ is $C \to D$ for any
+  closed type $C$. (Also for select classes of open types?) So can
+  model $\Nat \to \gamma$. But the functor $NX = \Nat \to X$ isn't
+  $\omega$-cocontinuous. It also doesn't preserve
+  $\omega_1$-presentable objects, i.e., countable $\omega$CPOs since
+  $\Nat \to \Nat$ is not countable. So we cannot have a functor like
+  $N$ as the subscript to $\Lan$ and expect the resulting $\Lan$ to be
+  $\omega_1$-cocontinuous.
+\item What functors can be subscripts to $\Lan$ and produce
+  $\omega_1$-cocontinuous functors? We can use functors that preserve
+  presentable objects by theorem in~\cite{jp19}, and possibly others
+  as well. These include polynomial functors, ADTs and nested types
+  seen as functors, certain (which?) GADTs seen as functors? How big
+  can GADTs get? 
 \end{itemize}
-\end{dfn}
-
-\begin{dfn}\label{def:RT-functor}
-An endofunctor $H$ on $RT_k$ is a triple $H = (H^1,H^2,H^*)$, where
-\begin{itemize}
-\item $H^1$ and $H^2$ are functors from $[\set^k,\set]$ to $[\set^k,\set]$
-\item $H^*$ is a functor from $RT_k$ to $[\rel^k,\rel]$
-\item for all $\overline{R : \rel(A,B)}$,
-  $\pi_1((H^*(\delta^1,\delta^2))_{\overline{R}}) = (H^1
-  \delta^1)_{\overline{A}}$ and
-  $\pi_2((H^*(\delta^1,\delta^2))_{\overline{R}}) = (H^2
-  \delta^2)_{\overline{B}}$
-\item The action of $H$ on objects is given by $H\,(F^1,F^2,F^*) =
-  (H^1F^1,\,H^2F^2,\,H^*(F^1,F^2,F^*))$
-\item The action of $H$ on morphisms is given by
-  $H\,(\delta^1,\delta^2) = (H^1\delta^1,H^2\delta^2)$ for
-  $(\delta^1,\delta^2) : (F^1,F^2,F^*)\to (G^1,G^2,G^*)$
-\end{itemize}
-\end{dfn}
-Since the results of applying an endofunctor $H$ to $k$-ary relation
-transformers and morphisms between them must again be $k$-ary relation
-transformers and morphisms between them, respectively,
-Definition~\ref{def:RT-functor} implicitly requires that the following
-three conditions hold:\,{\em i})
-if $R_1:\rel(A_1,B_1),...,R_k:\rel(A_k,B_k)$, then
-  $H^*(F^1,F^2,F^*) \ol{R} : \rel(H^1F^1 \ol{A}, H^2F^2 
-  \ol{B})$;
-  %In other words, $\pi_1 (H^*(F^1,F^2,F^*) \ol{R}) = H^1F^1
-  %\ol{A}$ and $\pi_2 (H^*(F^1,F^2,F^*) \ol{R}) = H^2F^2 \ol{B}$.
-{\em ii}) if $(\alpha_1, \beta_1) \in \Homrel(R_1,S_1),..., (\alpha_k,
-  \beta_k) \in \Homrel(R_k,S_k)$, then
-  $H^*(F^1,F^2,F^*)\, \ol{(\alpha, \beta)} = (H^1F^1\ol{\alpha}, H^2F^2
-  \ol{\beta})$; and
-  %In other words, $\pi_1 (H^*(F^1,F^2,F^*) \ol{(\alpha,
-  %  \beta)}) = H^1F^1 \ol{\alpha}$ and $\pi_2 (H^*(F^1,F^2,F^*) 
-  % \ol{(\alpha, \beta)}) = H^2F^2 \ol{\beta}$.  {\em iii}) if
-  $(\delta^1,\delta^2) : (F^1,F^2,F^*)\to (G^1,G^2,G^*)$ and
-  $R_1:\rel(A_1,B_1),...,R_k:\rel(A_k,B_k)$, then
-  $((H^1\delta^1)_{\ol{A}}x, (H^2\delta^2)_{\ol{B}}y) \in
-  H^*(G^1,G^2,G^*)\ol{R}$ whenever $(x, y) \in
-  H^*(F^1,F^2,F^*)\ol{R}$. Note, however, that this last condition is
-  automatically satisfied because it is implied by the third bullet
-  point of Definition~\ref{def:RT-functor}.
-
-\begin{dfn}\label{def:RT-nat-trans}
-If $H$ and $K$ are endofunctors on $RT_k$, then a {\em natural
-  transformation} $\sigma : H \to K$ is a pair $\sigma = (\sigma^1,
-\sigma^2)$, where $\sigma^1 : H^1 \to K^1$ and $\sigma^2 : H^2 \to
-K^2$ are natural transformations between endofunctors on
-$[\set^k,\set]$ and the component of $\sigma$ at $F
-%= (F^1,F^2,F^*)
-\in RT_k$ is given by $\sigma_F = (\sigma^1_{F^1}, \sigma^2_{F^2})$.
-\end{dfn}
-Definition~\ref{def:RT-nat-trans} entails that $\sigma^i_{F^i}$ must
-be natural in $F^i : [\set^k,\set]$, and, for every $F$, both
-$(\sigma^1_{F^1})_{\overline{A}}$ and
-$(\sigma^2_{F^2})_{\overline{A}}$ must be natural in $\overline{A}$.
-Moreover, since the results of applying $\sigma$ to $k$-ary relation
-transformers must be morphisms of $k$-ary relation transformers,
-Definition~\ref{def:RT-nat-trans} implicitly requires that
-$(\sigma_F)_{\overline{R}} = ( (\sigma^1_{F^1})_{\overline{A}},
-(\sigma^2_{F^2})_{\overline{B}})$ is a morphism in $\rel$ for any
-$k$-tuple of relations $\overline{R : \rel(A, B)}$, i.e., that if $(x,
-y) \in H^*F\overline{R}$, then $((\sigma^1_{F^1})_{\overline{A}} x,
-(\sigma^2_{F^2})_{\overline{B}} y) \in K^*F\overline{R}$.
-
-\vspace*{0.1in}
-
-%{\color{blue} More context? Weave cocontinuity requirement for fixed
-%  points through text better.}
-Critically, we can compute $\omega$-directed colimits in $RT_k$: it is
-not hard to see that if $\cal D$ is an $\omega$-directed set then
-$\colim{d \in {\cal D}}{(F^1_d, F^2_d,F^*_d)} = (\colim{d \in {\cal
-    D}}{F^1_d}, \colim{d \in {\cal D}}{F^2_d}, \colim{d \in {\cal
-    D}}{F^*_d})$.  We define an endofunctor $T = (T^1,T^2,T^*)$ on
-$RT_k$ to be {\em $\omega$-cocontinuous} if $T^1$ and $T^2$ are
-$\omega$-cocontinuous endofunctors on $[\set^k,\set]$ and $T^*$ is an
-$\omega$-cocontinuous functor from $RT_k$ to $[\rel^k,\rel]$, i.e., is
-in $[RT_k,[\rel^k,\rel]]$.
-%\begin{lemma}\label{lem:colimits}
-%$\colim{d \in {\cal D}}{(F^1_d, F^2_d,F^*_d)} = (\colim{d \in {\cal
-%      D}}{F^1_d}, \colim{d \in {\cal D}}{F^2_d}, \colim{d \in {\cal
-%      D}}{F^*_d})$
-%\end{lemma}
-%\begin{proof}
-%We first observe that $(\colim{d \in {\cal D}}{F^1_d}, \colim{d \in
-%  {\cal D}}{F^2_d}, \colim{d \in {\cal D}}{F^*_d})$ is in $RT_k$.  If
-%$R_1:\rel(A_1,B_1),...,R_k:\rel(A_k,B_k)$, then $\colim{d \in \cal
-%  D}{F^*_d\ol{R}} : \rel(\colim{d \in {\cal D}}{F^1_d\ol{A}},
-%\,\colim{d \in {\cal D}}{F^2_d\ol{B}})$ because of how colimits are
-%computed in $\rel$. Moreover, if $(\alpha_1, \beta_1) \in
-%\Homrel(R_1,S_1),..., (\alpha_k, \beta_k) \in \Homrel(R_k,S_k)$, then
-%\[\begin{array}{ll}
-%  & (\colim{d \in \cal D}{F_d^*})\ol{(\alpha,\beta)}\\
-%= & \colim{d \in \cal D}{F_d^*\ol{(\alpha,\beta)}}\\
-%= & \colim{d \in \cal D}{(F_d^1\ol{\alpha},\, F_d^2\ol{\beta})}\\
-%= & (\colim{d \in \cal D}{F^1_d \ol{\alpha}}, \,\colim{d \in \cal D}{
-%  F^2_d \ol{\beta}})
-%\end{array}\]
-%so $(\colim{d \in {\cal D}}{F^1_d}, \colim{d \in {\cal D}}{F^2_d},
-%\colim{d \in {\cal D}}{F^*_d})$ actually is in $RT_k$.
-%
-%Now to see that $\colim{d \in {\cal D}}{(F^1_d, F^2_d,F^*_d)} =
-%(\colim{d \in {\cal D}}{F^1_d}, \colim{d \in {\cal D}}{F^2_d},
-%\colim{d \in {\cal D}}{F^*_d})$, let $\gamma^1_d : F^1_d \to \colim{d
-%  \in {\cal D}}{F^1_d}$ and $\gamma^2_d : F^2_d \to \colim{d \in {\cal
-%    D}}{F^2_d}$ be the injections for the colimits $\colim{d \in {\cal
-%    D}}{F^1_d}$ and $\colim{d \in {\cal D}}{F^2_d}$,
-%respectively. Then $(\gamma^1_d, \gamma^2_d) : (F^1_d, F^2_d,F^*_d)
-%\to \colim{d \in {\cal D}}{(F^1_d, F^2_d,F^*_d)}$ is a morphism in
-%$RT_k$ because, for all $\ol{R : \rel(A, B)}$,
-%$((\gamma^1_d)_{\ol{A}}, (\gamma^2_d)_{\ol{B}}) : F^*_d \ol{R} \to
-%\colim{d \in {\cal D}}{F^*_d \ol{R}}$ is a morphism in $\rel$. So
-%$\{(\gamma^1_d, \gamma^2_d)\}_{d \in {\cal D}}$ are the mediating
-%morphisms of a cocone in $RT_k$ with vertex $\colim{d \in {\cal
-%    D}}{(F^1_d, F^2_d,F^*_d)}$. To see that this cocone is a
-%colimiting cocone, let $C = (C^1,C^2,C^*)$ be the vertex of a cocone
-%for $\{(F^1_d,F^2_d,F^*_d)\}_{d \in {\cal D}}$ with injections
-%$(\delta^1_d,\delta^2_d) : (F^1_d,F^2_d,F^*_d) \to C$. If $\eta^1 :
-%\colim{d \in {\cal D}}{F^1_d} \to C^1$ and $\eta^2 : \colim{d \in
-%  {\cal D}}{F^2_d} \to C^2$ are the mediating morphisms in
-%$[\set^k,\set]$, then $\eta^1$ and $\eta^2$ are unique such that
-%$\delta^1_d = \eta^1 \circ \gamma^1_d$ and $\delta^2_d = \eta^2 \circ
-%\gamma^2_d$.  We therefore have that $(\eta^1,\eta^2) : \colim{d \in
-%  {\cal D}}{(F^1_d, F^2_d,F^*_d)} \to C$ is the mediating morphism in
-%$RT_k$. Indeed, for all $\ol{R : \rel(A,B)}$ and $(x,y) \in \colim{d
-%  \in {\cal D}}{F^*_d \ol{R}}$, there exist $d$ and $(x',y') \in
-%F^*_d\ol{R}$ such that $(\gamma^1_d)_{\ol{A}}x' = x$ and
-%$(\gamma^2_d)_{\ol{B}}y' = y$. But then $(\eta^1_{\ol{A}}x,
-%\eta^2_{\ol{B}}y) = (\eta^1_{\ol{A}}((\gamma^1_d)_{\ol{A}}x'),
-%\eta^2_{\ol{B}}((\gamma^2_d)_{\ol{B}}y')) = ((\delta^1_d)_{\ol{A}}x',
-%(\delta^2_d)_{\ol{B}}y')$, and this pair is in $C^*\ol{R}$ because
-%$(\delta^1_d, \delta^2_d)$ is a morphism from $(F^1_d,F^2_d,F^*_d)$ to
-%$C$ in $RT_k$.
-%\end{proof}
-\begin{comment}
-\begin{dfn}\label{def:omega-cocont}
-An endofunctor $T = (T^1,T^2,T^*)$ on $RT_k$ is {\em
-$\omega$-cocontinuous} if $T^1$ and $T^2$ are $\omega$-cocontinuous
-endofunctors on $[\set^k,\set]$ and $T^*$ is an $\omega$-cocontinuous
-functor from $RT_k$ to $[\rel^k,\rel]$, i.e., is in
-$[RT_k,[\rel^k,\rel]]$.
-\end{dfn}
-\end{comment}
-Now, for any $k$, any $A : \set$, and any $R : \rel(A, B)$, let
-$K^\set_A$ be the constantly $A$-valued functor from $\set^k$ to
-$\set$ and $K^\rel_R$ be the constantly $R$-valued functor from
-$\rel^k$ to $\rel$.  Also let $0$ denote either the initial object of
-$\set$ or the initial object of $\rel$, as appropriate.  Observing
-that, for every $k$, $K^\set_0$ is initial in $[\set^k,\set]$, and
-$K^\rel_0$ is initial in $[\rel^k,\rel]$, we have that, for each $k$,
-$K_0 = (K^\set_0,K^\set_0,K^\rel_0)$ is initial in $RT_k$. Thus, if $T
-= (T^1,T^2,T^*) : RT_k \to RT_k$ is an endofunctor on $RT_k$ then we
-can define the relation transformer $\mu T$ to be $\colim{n \in
-  \nat}{T^n K_0}$.
-%Then Lemma~\ref{lem:colimits} shows $\mu T$ is indeed a relation
-%transformer, and that it is given explicitly by
-It is not hard to see that $\mu T$ is given explicitly as 
-\begin{equation}\label{eq:mu}
-  %\colim{n \in \nat}{T^n K_0}
-  \mu T = (\mu T^1,\mu T^2,
-\colim{n \in \nat}{(T^nK_0)^*})
-\end{equation}
-and that, as our notation suggests, it really is a fixpoint for $T$ if
-$T$ is $\omega$-cocontinuous:
-\begin{lemma}\label{lem:fp}
-For any $T : [RT_k,RT_k]$, $\mu T \cong T(\mu T)$.
-\end{lemma}
-\noindent
-The isomorphism is given by the morphisms $(\mathit{in}_1,
-\mathit{in}_2) : T(\mu T) \to \mu T$ and $(in_1^{-1}, in_2^{-1}) : \mu
-T \to T(\mu T)$ in $RT_k$. The latter is always a morphism in $RT_k$,
-but the former need not be if $T$ is not $\omega$-cocontinuous.
-
-\begin{comment}
-\begin{proof}
-We have $T(\mu T) = T(\colim{n \in \nat}{(T^nK_0)}) \cong \colim{n \in
-\nat}{T(T^nK_0)} =
-\mu T$.
-\end{proof}
-In fact, the isomorphism in Lemma~\ref{lem:fp} is given by the
-morphisms $(in_1, in_2) : T(\mu T) \to \mu T$ and $(in_1^{-1},
-in_2^{-1}) : \mu T \to T(\mu T)$ in $RT_k$. It is worth noting that the
-latter is always a morphism in $RT_k$, but the former isn't necessarily
-a morphism in $RT_k$ unless $T$ is $\omega$-cocontinuous.
-\end{comment}
-
-It is worth noting that the third component in Equation~(\ref{eq:mu})
-is the colimit in $[\rel^k,\rel]$ of third components of relation
-transformers, rather than a fixpoint of an endofunctor on
-$[\rel^k,\rel]$. That there is an asymmetry between the first two
-components of $\mu T$ and its third reflects the important conceptual
-observation that the third component of an endofunctor on $RT_k$ need
-not be a functor on all of $[\rel^k,\rel]$. In particular, although we
-can define $T_{H,\rho}\, F$ for a relation transformer $F$ in
-Definition~\ref{def:rel-sem} below, it is not clear how we could
-define it for $F : [\rel^k,\rel]$.
-
-\begin{dfn}\label{def:reln-env}
-A {\em relation environment} maps each type variable in $\tvars^k \cup
-\fvars^k$ to a $k$-ary relation transformer.  A morphism $f : \rho \to
-\rho'$ between relation environments $\rho$ and $\rho'$ with
-$\rho|_\tvars = \rho'|_\tvars$ maps each type constructor variable
-$\psi^k \in \tvars$ to the identity morphism on $\rho \psi^k = \rho'
-\psi^k$ and each functorial variable $\phi^k \in \fvars$ to a morphism
-from the $k$-ary relation transformer $\rho \phi$ to the $k$-ary
-relation transformer $\rho' \phi$. Composition of morphisms on
-relation environments is given componentwise, with the identity
-morphism mapping each relation environment to itself. This gives a
-category of relation environments and morphisms between them, which we
-denote $\relenv$.
-\end{dfn}
-When convenient we identify a $0$-ary relation transformer with the
-relation (transformer) that is its codomain and consider
-% With this convention,
-a relation environment to map a type variable of arity $0$
-%to a $0$-ary relation transformer, i.e.,
-to a relation.  We write $\rho[\ol{\alpha := R}]$ for the relation
-environment $\rho'$ such that $\rho' \alpha_i \, = R_i$ for $i =
-1,...,k$ and $\rho' \alpha = \rho\alpha$ if $\alpha \not \in
-\{\alpha_1,...,\alpha_k\}$.  If $\rho$ is a relation environment, we
-write $\pi_1 \rho$ and $\pi_2 \rho$ for the set environments mapping
-each type variable $\phi$ to the functors $(\rho\phi)^1$ and
-$(\rho\phi)^2$, respectively.
-
-We define, for each $k$, the notion of an $\omega$-cocontinuous
-functor from $\relenv$ to $RT_k$:
-\begin{dfn}\label{def:relenv-functor}
-A functor $H : [\relenv, RT_k]$ is a triple $H = (H^1,H^2,H^*)$,
-where
-\begin{itemize}
-\item $H^1$ and $H^2$ are objects in $[\setenv,[\set^k,\set]]$
-\item $H^*$ is a an object in $[\relenv,[\rel^k,\rel]]$
-\item for all $\overline{R : \rel(A,B)}$ and morphisms $f$ in
-  $\relenv$, $\pi_1(H^*f \,{\overline{R}}) = H^1 (\pi_1
-  f)\,{\overline{A}}$ and $\pi_2(H^*f \,{\overline{R}}) = H^2 (\pi_2
-  f)\,{\overline{B}}$
-\item The action of $H$ on $\rho$ in $\relenv$ is given by $H \rho = (H^1
-  (\pi_1 \rho),\,H^2 (\pi_2 \rho),\,H^*\rho)$
-\item The action of $H$ on morphisms $f : \rho \to \rho'$ in $\relenv$
-  is given by $Hf = (H^1 (\pi_1 f),H^2 (\pi_2 f))$
-\end{itemize}
-\end{dfn}
-\noindent Spelling out the last two bullet points above gives the
-following analogues of the three conditions immediately following
-Definition~\ref{def:RT-functor}: {\em i}) if $R_1 :
-\rel(A_1,B_1),...,R_k : \rel(A_k,B_k)$, then $H^*\rho\, \ol{R} :
-\rel(H^1(\pi_1 \rho)\, \ol{A}, H^2(\pi_2 \rho)\, \ol{B})$;
-%In other words, $\pi_1 (H^*\rho \ol{R}) = H^1(\pi_1 \rho)
-%  \,\ol{A}$ and $\pi_2 (H^*\rho \ol{R}) = H^2 (\pi_2 \rho) \,\ol{B}$;
-{\em ii}) if $(\alpha_1, \beta_1) \in \Homrel(R_1,S_1),..., (\alpha_k,
-\beta_k) \in \Homrel(R_k,S_k)$, then $H^*\rho\, \ol{(\alpha, \beta)} =
-(H^1(\pi_1 \rho)\,\ol{\alpha}, H^2(\pi_2 \rho)\, \ol{\beta})$;
-%  In other words, $\pi_1 (H^*\rho\, \ol{(\alpha, \beta)}) = H^1(\pi_1
-%  \rho)\,\ol{\alpha}$ and $\pi_2 (H^*\rho\, \ol{(\alpha, \beta)}) =
-%  H^2(\pi_2 \rho)\,\ol{\beta}$;
-and {\em iii}) if $f : \rho \to \rho'$ and
-$R_1:\rel(A_1,B_1),...,R_k:\rel(A_k,B_k)$, then $(H^1(\pi_1
-f)\,{\ol{A}}\,x, H^2(\pi_2 f)\,{\ol{B}}\,y) \in H^*\rho'\,\ol{R}$
-whenever $(x, y) \in H^*\rho\,\ol{R}$. As before, the last
-condition is automatically satisfied because it is implied by the
-third bullet point of Definition~\ref{def:relenv-functor}.
-
-Considering $\relenv$ as a product $\Pi_{\phi^k \in \tvars \cup
-  \fvars} RT_k$, we extend the computation of $\omega$-directed
-colimits in $RT_k$ to compute colimits in $\relenv$ componentwise. We
-similarly extend the notion of an $\omega$-cocontinuous endofunctor on
-$RT_k$ componentwise to give a notion of $\omega$-cocontinuity for
-functors from $\relenv$ to $RT_k$.  Recalling from the start of this
-subsection that Definition~\ref{def:rel-sem} is given mutually
-inductively with Definition~\ref{def:set-sem} we can, at last, define:
-
-\begin{dfn}\label{def:rel-sem}
-The {\em relational interpretation} $\relsem{\cdot} : \F \to [\relenv,
- \rel]$ is defined by
-\begin{align*}
-  \relsem{\Gamma;\Phi \vdash \zerot}\rho &= 0\\
-  \relsem{\Gamma;\Phi \vdash \onet}\rho &= 1\\
-  \end{align*}
-  \begin{align*}
-  \relsem{\Gamma; \emptyset \vdash \Nat^{\ol{\alpha}} \,F\,G}\rho &= \{\eta
-  : \lambda \ol{R}.\,\relsem{\Gamma; \ol{\alpha} \vdash
-    F}\rho[\ol{\alpha := R}] \Rightarrow \lambda \ol{R}. \,\relsem{
-    \Gamma; \ol{\alpha} \vdash G}\rho[\ol{\alpha := R}]\}\\
-  &=
-  \{(t,t') \in \setsem{\Gamma; \emptyset
-    \vdash \Nat^{\ol{\alpha}}
-    \,F\,G} (\pi_1 \rho) \times \setsem{ 
-    \Gamma;\emptyset
-    \vdash \Nat^{\ol{\alpha}} \,F\,G} (\pi_2
-  \rho)~|~\\ 
-  & \hspace{0.3in} \forall {R_1 : \rel(A_1,B_1)}\,...\,{R_k : \rel(A_k,B_k)}.\\
-  & \hspace{0.4in} (t_{\ol{A}},t'_{\ol{B}}) \in
-  (\relsem{\Gamma; \ol{\alpha} \vdash G}\rho[\ol{\alpha :=
-      R}])^{\relsem{\Gamma;\ol{\alpha}\vdash F}\rho[\ol{\alpha := R}]} \}\\  
-  \relsem{\Gamma;\Phi \vdash \phi \ol{F}}\rho &=
-  (\rho\phi)\ol{\relsem{\Gamma;\Phi \vdash 
-    F}\rho}\\
-  \relsem{\Gamma;\Phi \vdash F+G}\rho &=
-  \relsem{\Gamma;\Phi \vdash F}\rho +
-  \relsem{\Gamma;\Phi \vdash G}\rho\\
-  \relsem{\Gamma;\Phi \vdash F\times G}\rho &=
-  \relsem{\Gamma;\Phi \vdash F}\rho \times
-  \relsem{\Gamma;\Phi \vdash G}\rho\\  
-   \relsem{\Gamma;\Phi,\ol\gamma \vdash (\mu \phi.\lambda
-    \ol{\alpha}. H)\ol{G}}\rho
-  &= (\mu T_{H,\rho})\ol{\relsem{\Gamma;\Phi,\ol\gamma \vdash
-     G}\rho}\\
-  \text{where }	T_{H,\rho}
-    &= (T^\set_{H,\pi_1\rho}, T^\set_{H,\pi_2\rho}, T^\rel_{H,\rho}) \\
-  \text{and } T^\rel_{H,\rho}\,F
-    &= \lambda \ol{R}. \relsem{
-      \Gamma;\ol\gamma,\phi,\ol{\alpha} \vdash H}\rho[\phi :=
-    F][\ol{\alpha := R}]\\
-  \text{and } T^\rel_{H,\rho}\,\delta
-    &= \lambda \ol{R}. \relsem{
-      \Gamma;\ol\gamma,\phi,\ol{\alpha} \vdash H}\id_\rho[\phi :=
-    \delta][\ol{\alpha := \id_{\ol{R}}}]
-\end{align*}
-\end{dfn}
-
-The interpretations in Definitions~\ref{def:rel-sem}
-and~\ref{def:rel-sem-funcs} below respect weakening, and also ensure
-that $\relsem{\Gamma \vdash F \to G}$ and $\relsem{\Gamma \vdash
-  \forall \ol \alpha. F}$ are as expected in any parametric model.
-%$\relsem{\Gamma \vdash
-%  \sigma \to \tau}\rho = \relsem{\Gamma \vdash \sigma} \rho \to
-%\relsem{\Gamma \vdash \tau} \rho$.
-As for set interpretations, $\relsem{\Gamma; \emptyset \vdash
-  \Nat^{\ol\alpha} F\,G}$ is $\omega$-cocontinuous because it is a constant
-functor.
-%follows from $\omega$-cocontinuity of
-%$\relsem{\Gamma;\ol\alpha \vdash F}$ and
-%$\relsem{\Gamma;\ol\alpha \vdash G}$.
-If $\rho \in \relenv$
-and $\vdash F$, then we write $\relsem{\vdash F}$ instead of
-$\relsem{\vdash F}\rho$.  For the last clause in
-Definition~\ref{def:rel-sem} to be well-defined we need $T_{H,\rho}$
-to be an $\omega$-cocontinuous endofunctor on $RT$ so that, by
-Lemma~\ref{lem:fp}, it admits a fixpoint. Since $T_{H,\rho}$ is
-defined in terms of $\relsem{\Gamma;\ol\gamma,\phi^k, \ol{\alpha}
-  \vdash H}$, this means that relational interpretations of types must
-be $\omega$-cocontinuous functors from $\relenv$ to $RT_0$, which in
-turn entails that the actions of relational interpretations of types
-on objects and on morphisms in $\relenv$ are intertwined. As for set
-interpretations, we know from~\cite{jp19} that, for every $\Gamma;
-\ol{\alpha} \vdash F$, $\relsem{\Gamma; \ol{\alpha} \vdash F}$ is
-actually in $[\rel^k,\rel]$ where $k = |\ol \alpha|$.
-%In fact, we already know from~\cite{jp19} that, for every $\Gamma;
-%\ol{\alpha} \vdash \tau : \F$, $\relsem{\Gamma; \ol{\alpha} \vdash
-%  \tau}$ is actually functorial in $\ol{\alpha}$ and
-%$\omega$-cocontinuous.
-We first define the actions of each of these functors on morphisms
-between environments in Definition~\ref{def:rel-sem-funcs}, and then
-argue that the functors given by Definitions~\ref{def:rel-sem}
-and~\ref{def:rel-sem-funcs} are well-defined and have the required
-properties. To do this, we extend $T_H$ to a {\em functor} from
-$\relenv$ to $[[\rel^k,\rel],[\rel^k,\rel]]$. Its action on an object
-$\rho \in \relenv$ is given by the higher-order functor $T_{H,\rho}$
-whose actions on objects and morphisms are given in
-Definition~\ref{def:rel-sem-funcs}. Its action on a morphism $f : \rho
-\to \rho'$ is the higher-order natural transformation $T_{H,f} :
-T_{H,\rho} \to T_{H,\rho'}$ whose action on any $F : [\rel^k,\rel]$ is
-the natural transformation $T_{H,f}\, F : T_{H,\rho}\, F \to
-T_{H,\rho'}\, F$ whose component at $\ol{R}$ is $(T_{H,f}\,
-F)_{\ol{R}} = \relsem{\Gamma; \ol\gamma,\phi,\ol{\alpha} \vdash
-  H}f[\phi := \id_F][\ol{\alpha := \id_R}]$. The next definition uses
-$T_H$ to define the actions of functors interpreting types on
-morphisms between relation environments.
-
-\begin{dfn}\label{def:rel-sem-funcs}
-Let $f: \rho \to \rho'$ for relation environments $\rho$ and $\rho'$
-(so that $\rho|_\tvars = \rho'|_\tvars$). The action
-$\relsem{\Gamma;\Phi \vdash F}f$ of $\relsem{\Gamma;\Phi \vdash F}$ on
-the morphism $f$ is given exactly as in
-Definition~\ref{def:set-sem-funcs}, except that all interpretations
-are relational interpretations and all occurrences of $T^\set_{H,f}$
-are replaced by $T_{H,f}$.
-\begin{comment}
-as follows:
-\begin{itemize}
-\item If \,$\Gamma;\Phi \vdash \zerot$ then $\relsem{\Gamma;\Phi \vdash
-  \zerot}f = \id_0$
-\item If\, $\Gamma;\Phi \vdash \onet$ then $\relsem{\Gamma;\Phi \vdash
-  \onet}f = \id_1$
-\item If \,$\Gamma; \Phi
-  \vdash \Nat^{\ol{\alpha}}\,F\,G$ then
-  $\relsem{\Gamma; \Phi
-    \vdash \Nat^{\ol{\alpha}}\,F\,G} f =$\\
-\hspace*{0.2in} $\lambda t : \relsem{\Gamma;\Phi \vdash
-    \Nat^{\ol{\alpha}}\,F\,G}\rho.\, 
-  (\lambda \ol{A}. \relsem{\Gamma; \Phi,\ol\alpha \vdash G}f[\ol{\alpha
-      := id_A}]) \circ t$  
-\item If \,$\Gamma;\Phi \vdash \phi \ol{\tau}$, then
-  $\relsem{\Gamma;\Phi \vdash \phi \ol{\tau}} f : \relsem{\Gamma;\Phi
-  \vdash \phi \ol{\tau}}\rho \to \relsem{\Gamma;\Phi \vdash \phi
-  \ol{\tau}}\rho' = (\rho\phi) \ol{\relsem{\Gamma;\Phi \vdash
-    \tau}\rho} \to (\rho'\phi) \ol{\relsem{\Gamma;\Phi \vdash
-    \tau}\rho'}$ is defined by $\relsem{\Gamma;\Phi \vdash \phi
-  \tau{A}} f = (f\phi)_{\ol{\relsem{\Gamma;\Phi \vdash \tau}\rho'}}
-  \,\circ\, (\rho\phi) \ol{\relsem{\Gamma;\Phi \vdash \tau}f} =
-  (\rho'\phi) \ol{\relsem{\Gamma;\Phi \vdash \tau}f} \,\circ\, (f
-  \phi)_{\ol{\relsem{\Gamma;\Phi \vdash \tau}\rho}}$
-\item If\, $\Gamma;\Phi\vdash \sigma + \tau$ then $\relsem{\Gamma;\Phi
-  \vdash \sigma + \tau}f$ is defined by $\relsem{\Gamma;\Phi \vdash
-  \sigma + \tau}f(\inl\,x) = \inl\,(\relsem{\Gamma;\Phi \vdash
-  \sigma}f x)$ and $\relsem{\Gamma;\Phi \vdash \sigma +
-  \tau}f(\inr\,y) = \inr\,(\relsem{\Gamma;\Phi \vdash \tau}f y)$
-\item If\, $\Gamma;\Phi\vdash \sigma \times \tau$ then
-  $\relsem{\Gamma;\Phi \vdash \sigma \times \tau}f =
-  \relsem{\Gamma;\Phi \vdash \sigma}f \times \relsem{\Gamma;\Phi
-    \vdash \tau}f$
-\item If\, $\Gamma;\Phi \vdash (\mu \phi^k.\lambda
-  \ol{\alpha}. H)\ol{\tau}$ then $\relsem{\Gamma;\Phi \vdash (\mu
-    \phi.\lambda \ol{\alpha}. H)\ol{\tau}} f = (\mu
-  T_{H,f})\ol{\relsem{\Gamma;\Phi \vdash \tau}\rho'} \circ (\mu
-  T_{H,\rho})\ol{\relsem{\Gamma;\Phi \vdash \tau}f} = (\mu
-  T_{H,\rho'})\ol{\relsem{\Gamma;\Phi \vdash \tau}f} \circ (\mu
-  T_{H,f})\ol{\relsem{\Gamma;\Phi \vdash \tau}\rho}$
-\end{itemize}
-\end{comment}
-\end{dfn}
-
-To see that the functors given by Definitions~\ref{def:rel-sem}
-and~\ref{def:rel-sem-funcs} are well-defined we must show that, for
-every $H$, $T_{H,\rho}\,F$ is a relation transformer for any relation
-transformer $F$, and that $T_{H,f}\, F : T_{H,\rho}\, F \to
-T_{H,\rho'}\, F$ is a morphism of relation transformers for every
-relation transformer $F$ and every morphism $f : \rho \to \rho'$ in
-$\relenv$. This is an immediate consequence of 
-\begin{lemma}\label{lem:rel-transf-morph}
-%The interpretations in Definitions~\ref{def:rel-sem}
-%and~\ref{def:rel-sem-funcs} are well-defined and,
-For every $\Gamma;\Phi \vdash F$, $\sem{\Gamma;\Phi \vdash F} =
-(\setsem{\Gamma;\Phi \vdash F}, \setsem{\Gamma;\Phi \vdash
-  F},\relsem{\Gamma;\Phi \vdash F}) \in
-%is an $\omega$-cocontinuous functor from $\relenv$ to $RT_0$, i.e., is an
-%element of
-[\relenv,RT_0]$.
-\end{lemma}
-\noindent
-The proof is a straightforward induction on the structure of $F$,
-%%each case of which first shows that $(\setsem{\Gamma;\Phi \vdash
-%  \tau}, \setsem{\Gamma;\Phi \vdash \tau},\relsem{\Gamma;\Phi \vdash
-%  \tau})$ is indeed a functor from $\relenv$ to $RT_0$, and then uses
-using an appropriate result from~\cite{jp19} to deduce
-$\omega$-cocontinuity of $\sem{\Gamma;\Phi \vdash F}$ in each case,
-together with Lemma~\ref{lem:fp} and Equation~\ref{eq:mu} for
-$\mu$-types.
-
-\begin{comment}
-\begin{proof}
-By induction on the structure of $\tau$. The only interesting cases
-are when $\tau = \phi\ol{\tau}$ and when $\tau = (\mu \phi^k. \lambda
-\overline{\alpha}. H)\overline{\tau}$. We consider each in turn.
-
-\begin{itemize}
-\item When $\tau = \Gamma; \Phi \vdash \phi\ol{\tau}$, we have    
-\[\begin{array}{ll}
-  & \pi_i (\relsem{\Gamma; \Phi \vdash \phi\ol{\tau}}\rho)\\
-= & \pi_i ((\rho \phi)\overline{\relsem{\Gamma; \Phi \vdash \tau}\rho})\\
-= & (\pi_i (\rho \phi)) (\pi_i(\overline{\relsem{\Gamma; \Phi \vdash \tau}\rho}))\\
-= & ((\pi_i \rho) \phi) (\overline{\setsem{\Gamma; \Phi \vdash
-    \tau}(\pi_i \rho)})\\
-= & \setsem{\Gamma; \Phi \vdash \phi\ol{\tau}}(\pi_i \rho)
-\end{array}\]
-and, for $f : \rho \to \rho'$ in $\relenv$,
-\[\begin{array}{ll}
-  & \pi_i (\relsem{\Gamma; \Phi \vdash \phi\ol{\tau}}f)\\
-= & \pi_i ((f\phi)_{\overline{\relsem{\Gamma; \Phi \vdash \tau}\rho'}})
-\circ \pi_i((\rho\phi)({\overline{\relsem{\Gamma; \Phi \vdash \tau}f}}))\\
-= & (\pi_i (f\phi))_{\overline{\pi_i (\relsem{\Gamma; \Phi \vdash \tau}\rho')}}
-\circ (\pi_i(\rho\phi))({\overline{\pi_i (\relsem{\Gamma; \Phi \vdash \tau}f}}))\\
-= & ((\pi_i f)\phi)_{\overline{\setsem{\Gamma; \Phi \vdash \tau}(\pi_i\rho')}}
-\circ ((\pi_i\rho)\phi)({\overline{\setsem{\Gamma; \Phi \vdash \tau}(\pi_i f)}})\\
-= & \setsem{\Gamma; \Phi \vdash \phi\ol{\tau}}(\pi_i f)
-\end{array}\]
-The third equalities of each of the above derivations are by the
-induction hypothesis. That $\sem{\Gamma; \Phi \vdash
-  \phi\ol{\tau}}$ is $\omega$-cocontinuous is an immediate
-consequence of the facts that $\set$ and $\rel$ are locally finitely
-presentable, together with Corollary~12 of~\cite{jp19}.
-\item When $\tau = (\mu \phi. \lambda
-  \overline{\alpha}. H)\overline{\tau}$ we first show that $\sem{ (\mu
-    \phi. \lambda \overline{\alpha}. H)\overline{\tau}}$ is
-  well-defined.
-\begin{itemize}
-\item \underline{$T_\rho : [RT_k,RT_k]$}:\/ We must show that, for any
-  relation transformer $F = 
-  (F^1, F^2, F^*)$, the triple $T_{\rho} F = (T^\set_{\pi_1 \rho}F^1,
-  T^\set_{\pi_2 \rho}F^2, T^\rel_{\rho}F)$ is also a relation
-  transformer.  Let $\overline{R : \rel(A, B)}$. Then for
-  $i = 1, 2$, we have
-\[\begin{split}
-\pi_i(T^\rel_{\rho}\,F\,\overline{R})
-&= \pi_i(\relsem{\Gamma;\Phi,\phi,\overline{\alpha} \vdash H}\rho[\phi := F]\overline{[\alpha := R]}) \\
-&= \setsem{\Gamma;\Phi,\phi,\overline{\alpha} \vdash H} (\pi_i (\rho[\phi := F]\overline{[\alpha := R]})) \\
-&= \setsem{\Gamma;\Phi,\phi,\overline{\alpha} \vdash H} (\pi_i \rho)[\phi := \pi_i F]\overline{[\alpha := \pi_i R]}) \\
-&= T^\set_{\pi_i \rho} (\pi_i F) (\overline{\pi_i R})
-\end{split}\]
-and 
-\[\begin{split}
-\pi_i(T^\rel_{\rho}\,F\,\overline{\gamma})
-&= \pi_i(\relsem{\Gamma;\Phi,\phi,\overline{\alpha} \vdash H}\id_\rho[\phi
-  := \id_F]\overline{[\alpha := \gamma]}) \\
-&= \setsem{\Gamma;\Phi,\phi,\overline{\alpha} \vdash H} (\pi_i (\id_\rho[\phi := \id_F]\overline{[\alpha := \gamma]})) \\
-&= \setsem{\Gamma;\Phi,\phi,\overline{\alpha} \vdash H} \id_{\pi_i \rho}[\phi := \id_{\pi_i F}]\overline{[\alpha := \pi_i \gamma]} \\
-&= T^\set_{\pi_i \rho} (\pi_i F) (\overline{\pi_i \gamma})
-\end{split}\]
-Here, the second equality in each of the above chains of equalities is
-by the induction hypothesis.
-
-We also have that, for every morphism $\delta = (\delta^1, \delta^2) :
-F \to G$ in $RT_k$ and all $\overline{R : \rel(A, B)}$,
-\[\begin{array}{ll}
-  & \pi_i((T^\rel_\rho \delta)_{\overline{R}})\\
-= & \pi_i(\relsem{\Gamma;\Phi,\phi,\overline{\alpha} \vdash H}\id_\rho[\phi :=
-  \delta]\overline{[\alpha := \id_R]})\\
-= & \setsem{\Gamma;\Phi,\phi,\overline{\alpha} \vdash H}\id_{\pi_i\rho}[\phi :=
-  \pi_i \delta]\overline{[\alpha := \id_{\pi_i R}]}\\
-= & (T^\set_{\pi_i \rho} (\pi_i \delta))_{\overline{\pi_i R}}
-\end{array}\]
-
-Here, the second equality is by the induction hypothesis.  That
-$T_\rho$ is $\omega$-cocontinuous follows immediately from the
-induction hypothesis on $\sem{\Gamma;\Phi,\phi,\ol{\alpha} \vdash H}$
-and the fact that colimts are computed componentwise in $RT$.
-\item \underline{$\sigma_f = (\sigma^\set_{\pi_1 f},
-  \sigma^\set_{\pi_2 f})$ is a natural transformation from $T_{\rho}$
-  to $T_{\rho'}$}:\/ We must show that $(\sigma_f)_F =
-  ((\sigma^\set_{\pi_1 f})_{F^1}, (\sigma^\set_{\pi_2 f})_{F^2})$ is a
-  morphism in $RT_k$ for all relation transformers $F = (F^1, F^2,
-  F^*)$, i.e., that $((\sigma_f)_F)_{\overline{R}}
-  =(((\sigma^\set_{\pi_1 f})_{F^1})_{\overline{A}},
-  ((\sigma^\set_{\pi_2 f})_{F_2})_{\overline{B}})$ is a morphism in
-  $\rel$ for all relations $\overline{R : \rel(A, B)}$. Indeed, we
-  have that
-\[
-((\sigma_f)_F)_{\overline{R}} =
-\relsem{\Gamma;\Phi,\phi,\overline{\alpha} \vdash H}f[\phi :=
-  \id_F]\overline{[\alpha := \id_R]}
-\]
-is a morphism in $RT_0$ (and thus in $\rel$) by the induction hypothesis.
-\end{itemize}
-
-\vspace*{0.05in}
-
-The relation transformer $\mu T_\rho$ is therefore a fixpoint of
-$T_\rho$ by Lemma~\ref{lem:fp}, and $\mu \sigma_f$ is a morphism in
-$RT_k$ from $\mu T_\rho$ to $\mu T_{\rho'}$. ($\mu$ is shown to be a
-functor in~\cite{jp19}.)  So $\relsem{\Gamma; \Phi \vdash (\mu
-  \phi. \lambda \overline{\alpha}. H)\overline{\tau}}$, and thus
-$\sem{\Gamma; \Phi \vdash (\mu \phi. \lambda
-  \overline{\alpha}. H)\overline{\tau}}$, is well-defined.
-
-\vspace*{0.1in}
-
-To see that $\sem{\Gamma; \Phi \vdash (\mu \phi. \lambda
-  \overline{\alpha}. H)\overline{\tau}} : [\relenv,RT_0]$, we must
-verify three conditions:
-\begin{itemize}
-\item Condition (1) after Definition~\ref{def:relenv-functor} is
-  satisfied since
-\[\begin{split}
-\pi_i(\relsem{\Gamma;\Phi \vdash (\mu \phi. \lambda
-  \overline{\alpha}. H) \overline{\tau}}\rho) 
-&= \pi_i( (\mu T_\rho) (\overline{\relsem{\Gamma;\Phi \vdash
-    \tau}\rho})) \\ 
-&= \pi_i(\mu T_{\rho}) (\overline{\pi_i(\relsem{\Gamma;\Phi \vdash
-    \tau}\rho})) \\ 
-&= \mu T^\set_{\pi_i\rho} (\overline{\setsem{\Gamma;\Phi \vdash
-    \tau}(\pi_i\rho)}) \\ 
-&= \setsem{\Gamma;\Phi \vdash (\mu \phi. \lambda \overline\alpha. H)
-  \overline{\tau}}(\pi_i\rho) 
-\end{split}\]
-The third equality is by Equation ~\ref{eq:mu} and the induction
-hypothesis.
-\item Condition (2) after Definition~\ref{def:relenv-functor} is
-  satisfied since it is subsumed by the previous condition because $k
-  = 0$.
-\item The third bullet point of Definition~\ref{def:relenv-functor} is
-  satisfied because
-\[\begin{split}
-& \pi_i(\relsem{\Gamma;\Phi \vdash (\mu \phi. \lambda
-  \overline\alpha. H)\overline{\tau}}f)\\
-&= \pi_i((\mu T_{\rho'})(\overline{\relsem{\Gamma;\Phi \vdash
-    \tau}f}) \circ (\mu \sigma_f)_{\overline{\relsem{\Gamma;\Phi
-      \vdash \tau}\rho}}) \\ 
-&= \pi_i((\mu T_{\rho'})(\overline{\relsem{\Gamma;\Phi \vdash
-    \tau}f})) \circ \pi_i((\mu
-\sigma_f)_{\overline{\relsem{\Gamma;\Phi \vdash \tau}\rho}}) \\  
-&= \pi_i(\mu T_{\rho'})(\overline{\pi_i(\relsem{\Gamma;\Phi
-    \vdash \tau}f})) \circ \pi_i(\mu
-\sigma_f)_{\overline{\pi_i(\relsem{\Gamma;\Phi \vdash 
-      \tau}\rho})} \\ 
-&= (\mu T^\set_{\pi_i \rho'})(\overline{\setsem{\Gamma;\Phi \vdash
-    \tau}(\pi_i f)}) \circ (\mu \sigma^\set_{\pi_i
-  f})_{\overline{\setsem{\Gamma;\Phi \vdash \tau}(\pi_i\rho)}} \\ 
-&= \setsem{\Gamma;\Phi \vdash (\mu \phi. \lambda
-  \overline\alpha. H)\overline{\tau}}(\pi_i f). 
-\end{split}\]
-The fourth equality is by~\ref{eq:mu} and the induction hypothesis.
-\end{itemize}
-As before, that $\sem{\Gamma; \Phi \vdash (\mu \phi. \lambda
-  \overline{\alpha}. H)\overline{\tau}}$ is $\omega$-cocontinuous
-follows from the facts that $\set$ and $\rel$ are locally finitely
-presentable, and that colimits in $\relenv$ are computed
-componentwise, together with Corollary~12 of~\cite{jp19}.
-\end{itemize}
-\end{proof}
-\end{comment}
-
-%We need
-%$\relsem{\Gamma; \Phi, \phi \vdash \tau}\rho$ $=
-%\relsem{\Gamma, \psi; \Phi \vdash \tau[\phi :==
-%    \psi]}\rho[\psi:=\rho\phi]$, and that $\relsem{\Gamma \vdash
-%  \sigma \to \tau}$ and
-%$\setsem{\Gamma; \Phi, \phi \vdash \tau}\rho$ $=
-%\setsem{\Gamma, \psi; \Phi \vdash \tau[\phi :==
-%    \psi]}\rho[\psi:=\rho\phi]$. 
-We can also prove by simultaneous induction that our interpretations
-of types interact well with demotion of functorial variables. Indeed,
-we have that, if $\rho, \rho' : \setenv$, \,$f : \rho \to \rho'$,
-\,$\rho \phi = \rho \psi = \rho' \phi = \rho' \psi$, \, $f \phi = f
-\psi = \id_{\rho \phi}$,\, $\Gamma; \Phi, \phi^k \vdash F$,\,
-$\Gamma;\Phi,\ol{\alpha} \vdash G$,\, $\Gamma;\Phi,\alpha_1...\alpha_k
-\vdash H$, and $\ol{\Gamma;\Phi \vdash K}$, then
-%for any $\rho,\rho'$ and $f: \rho \to \rho'$ in
-%$\setenv$, 
-\begin{gather}
-\label{thm:demotion-objects}
-\setsem{\Gamma; \Phi, \phi \vdash F} \rho = \setsem{\Gamma, \psi; \Phi
-  \vdash F[\phi :== \psi] } \rho\\
-\label{thm:demotion-morphisms}
-\setsem{\Gamma; \Phi, \phi \vdash F} f = \setsem{\Gamma, \psi; \Phi
-  \vdash F[\phi :== \psi]} f\\
-%\end{gather}
-%Moreover,
-%\begin{lemma}\label{lem:demotion-objects}
-%\begin{gather}
-%\label{thm:demotion-objects}
-%\setsem{\Gamma; \Phi, \phi \vdash F} \rho = \setsem{\Gamma, \psi; \Phi
-%  \vdash F[\phi :== \psi] } \rho{\color{blue}[\psi := \rho \phi]}\\
-%\label{thm:demotion-morphisms}
-%\setsem{\Gamma; \Phi, \phi \vdash F} f = \setsem{\Gamma, \psi; \Phi
-%  \vdash F[\phi :== \psi]} f{\color{blue}[\psi \mapsto f\phi]}\\
-\label{eq:subs-var}
-\setsem{\Gamma;\Phi \vdash G[\ol{\alpha := K}]}\rho =
-\setsem{\Gamma;\Phi,\ol{\alpha} \vdash G}\rho[\ol{\alpha := 
-\setsem{\Gamma;\Phi \vdash K}\rho}]\\
-\label{eq:subs-var-morph}
-\setsem{\Gamma;\Phi \vdash G[\ol{\alpha := K}]}f =
-\setsem{\Gamma;\Phi,\ol{\alpha} \vdash G}f[\ol{\alpha :=
-\setsem{\Gamma;\Phi \vdash K}f}]\\
-\label{eq:subs-const}
-\setsem{\Gamma; \Phi \vdash F[\phi := H]}\rho
-= \setsem{\Gamma; \Phi, \phi \vdash F}\rho
-[\phi := \lambda \ol{A}.\, \setsem{\Gamma;\Phi,\overline{\alpha}\vdash
-    H}\rho[\overline{\alpha := A}]] \\ 
-\label{eq:subs-const-morph}
-\setsem{\Gamma; \Phi \vdash F[\phi := H]}f
-= \setsem{\Gamma; \Phi, \phi \vdash F}f
-[\phi := \lambda \ol{A}.\,\setsem{\Gamma;\Phi,\overline{\alpha}\vdash
-    H}f[\overline{\alpha := \id_{\ol{A}}}]] 
-\end{gather}
-Identities analogous to (\ref{thm:demotion-objects}) through
-(\ref{eq:subs-const-morph}) hold for relational interpretations as well.
-%\end{lemma}
-
-\begin{comment}
-\begin{lemma}\label{lem:substitution}
-Let $\rho$ and $\rho'$ be set environments and let $f : \rho \to \rho'$ be a
-morphism of set environments. 
-\begin{itemize}
-\item If $\Gamma;\Phi,\ol{\alpha} \vdash F$ and $\Gamma;\Phi \vdash \tau$,
-  then  
-\begin{gather}\label{eq:subs-var}
-\setsem{\Gamma;\Phi \vdash F[\ol{\alpha := \tau}]}\rho =
-\setsem{\Gamma;\Phi,\ol{\alpha} \vdash F}\rho[\ol{\alpha :=
-\setsem{\Gamma;\Phi \vdash \tau}\rho}]\\
-\intertext{and}\label{eq:subs-var-morph}
-\setsem{\Gamma;\Phi \vdash F[\ol{\alpha := \tau}]}f =
-\setsem{\Gamma;\Phi,\ol{\alpha} \vdash F}f[\ol{\alpha :=
-\setsem{\Gamma;\Phi \vdash \tau}f}]
-\end{gather}
-\item If $\Gamma;\Phi,\phi^k\vdash F$ and
-  $\Gamma;\Phi,\alpha_1...\alpha_k \vdash H$, then 
-\begin{gather}\label{eq:subs-const}
-\setsem{\Gamma; \Phi \vdash F[\phi := H]}\rho
-= \setsem{\Gamma; \Phi, \phi \vdash F}\rho
-[\phi := \lambda \ol{A}.\, \setsem{\Gamma;\Phi,\overline{\alpha}\vdash
-    H}\rho[\overline{\alpha := A}]] \\ 
-\intertext{and}\label{eq:subs-const-morph}
-\setsem{\Gamma; \Phi \vdash F[\phi := H]}f
-= \setsem{\Gamma; \Phi, \phi \vdash F}f
-[\phi := \lambda \ol{A}.\,\setsem{\Gamma;\Phi,\overline{\alpha}\vdash
-    H}f[\overline{\alpha := \id_{\ol{A}}}]] 
-\end{gather}
-\end{itemize}
-Analogous identities hold for relation environments and morphisms
-between them.
-\end{lemma}
-
-\end{comment}
-
-\begin{comment}
-
-\begin{thm}\label{thm:demotion-objects}
-Let \, $\Gamma; \Phi, \phi \vdash \tau : \F$. If $\rho, \rho' : \setenv$
-are such that $\rho \phi = \rho \psi = \rho' \phi = \rho' \psi$, and
-if $f : \rho \to \rho'$ is a morphism of set environments such that $f
-\phi = f \psi = \id_{\rho \phi}$, then
-\[\setsem{\Gamma; \Phi, \phi \vdash \tau} \rho = \setsem{\Gamma, \psi;
-  \Phi \vdash \tau[\phi :== \psi] } \rho\]
-and
-\[\setsem{\Gamma; \Phi, \phi \vdash \tau} f = \setsem{\Gamma, \psi;
-  \Phi \vdash \tau[\phi :== \psi]} f\]
-
-\vspace*{0.1in}
-
-\noindent
-Analogously, if $\rho, \rho' : \relenv$ are such that $\rho \phi =
-\rho \psi = \rho' \phi = \rho' \psi$, and if $f : \rho \to \rho'$ is a
-morphism of relation environments such that $f \phi = f \psi =
-\id_{\rho \phi}$, then
-\[\relsem{\Gamma; \Phi, \phi \vdash \tau} \rho = \relsem{\Gamma, \psi;
-  \Phi \vdash \tau[\phi :== \psi] } \rho\]
-and
-\[\relsem{\Gamma; \Phi, \phi \vdash \tau} f = \relsem{\Gamma, \psi;
-  \Phi \vdash \tau[\phi :== \psi]} f\]
-\end{thm}
-
-\end{comment}
-
-\begin{comment}
----
-The next two theorems are proven by simultaneous induction. We are
-actually only interested in using Theorem~\ref{thm:demotion-objects},
-but in order to prove the $\mu$ case for this theorem, we need
-Theorem~\ref{thm:demotion-morph} to show that two functors have equal
-actions on morphisms.
-
-\begin{thm}\label{thm:demotion-objects}
-Let $\Gamma; \Phi, \phi \vdash \tau : \F$. If $\rho, \rho' : \setenv$
-are such that $\rho \phi = \rho \psi = \rho' \phi = \rho' \psi$, and
-if $f : \rho \to \rho'$ is a morphism of set environments such that $f
-\phi = f \psi = \id_{\rho \phi}$, then
-\[\setsem{\Gamma; \Phi, \phi \vdash \tau} \rho = \setsem{\Gamma, \psi;
-  \Phi \vdash \tau[\phi :== \psi] } \rho\]
-and
-\[\setsem{\Gamma; \Phi, \phi \vdash \tau} f = \setsem{\Gamma, \psi;
-  \Phi \vdash \tau[\phi :== \psi]} f\]
-
-\vspace*{0.1in}
-
-\noindent
-Analogously, if $\rho, \rho' : \relenv$ are such that $\rho \phi =
-\rho \psi = \rho' \phi = \rho' \psi$, and if $f : \rho \to \rho'$ is a
-morphism of relation environments such that $f \phi = f \psi =
-\id_{\rho \phi}$, then
-\[\relsem{\Gamma; \Phi, \phi \vdash \tau} \rho = \relsem{\Gamma, \psi;
-  \Phi \vdash \tau[\phi :== \psi] } \rho\]
-and
-\[\relsem{\Gamma; \Phi, \phi \vdash \tau} f = \relsem{\Gamma, \psi;
-  \Phi \vdash \tau[\phi :== \psi]} f\]
-\end{thm}
-\begin{proof}
-We prove the result for set interpretations by induction on the
-structure of $\tau$.  The case for relational interpretations proceeds
-analogously. Since the only interesting cases are the application case
-and the $\mu$-case, we elide the others.
-\begin{itemize}
-\item If $\Gamma; \Phi, \phi \vdash \phi \ol\tau : \F$, then the
-  induction hypothesis gives that  
-\[\setsem{\Gamma; \Phi, \phi \vdash \tau } \rho = \setsem{\Gamma,
-  \psi; \Phi \vdash \tau[\phi :== \psi] } \rho\] 
-and
-\[\setsem{\Gamma; \Phi, \phi \vdash \tau} f = \setsem{\Gamma, \psi;
-  \Phi \vdash \tau[\phi :== \psi]} f\]
-for each $\tau$. Then
-\begin{align*}
- & \setsem{\Gamma; \Phi, \phi \vdash \phi \ol\tau } \rho \\
-= \; & (\rho \phi) \ol{\setsem{\Gamma; \Phi, \phi \vdash \tau } \rho} \\
-= \; & (\rho \phi) \ol{\setsem{\Gamma, \psi; \Phi \vdash \tau[\phi :==
-      \psi] } \rho} \\ 
-= \; & (\rho \psi) \ol{\setsem{\Gamma, \psi; \Phi \vdash \tau[\phi :==
-      \psi] } \rho} \\ 
-= \; & \setsem{\Gamma, \psi; \Phi \vdash \psi \ol{\tau[\phi :== \psi]}
-} \rho \\ 
-= \; & \setsem{\Gamma, \psi; \Phi \vdash (\phi \ol\tau)[\phi :== \psi]
-} \rho 
-\end{align*}
-Here, the first and fifth equalities are by
-Definition~\ref{def:set-sem}, and the fourth equality is by equality
-of the functors $\rho \phi$ and $\rho \psi$. We also have that
-\begin{align*}
-& \setsem{\Gamma; \Phi, \phi \vdash \phi \ol\tau} f \\
-= \; & (f \phi)_{\ol{\setsem{\Gamma; \Phi, \phi \vdash \tau} \rho'}} 
-    \circ (\rho \phi) \ol{\setsem{\Gamma; \Phi, \phi \vdash \tau} f} \\
-= \; & (\id_{\rho \phi})_{\ol{\setsem{\Gamma; \Phi, \phi \vdash \tau} \rho'}} 
-    \circ (\rho \phi) \ol{\setsem{\Gamma; \Phi, \phi \vdash \tau} f} \\
-= \; & (\rho \phi) \ol{\setsem{\Gamma; \Phi, \phi \vdash \tau} f} \\
-= \; & (\rho \psi) \ol{\setsem{\Gamma, \psi; \Phi \vdash \tau[\phi :==
-      \psi]} f} \\ 
-= \; & (\id_{\rho \psi})_{\ol{\setsem{\Gamma, \psi; \Phi \vdash
-      \tau[\phi :== \psi]} \rho'}} \circ (\rho \psi)
-\ol{\setsem{\Gamma, \psi; \Phi \vdash \tau[\phi :== \psi]} f} \\ 
-= \; & (f \psi)_{\ol{\setsem{\Gamma, \psi; \Phi \vdash \tau[\phi :==
-        \psi]} \rho'}} \circ (\rho \psi) \ol{\setsem{\Gamma, \psi;
-    \Phi \vdash \tau[\phi :== \psi]} f} \\ 
-= \; & \setsem{\Gamma, \psi; \Phi \vdash \psi \ol{\tau[\phi :== \psi]}} f \\
-= \; & \setsem{\Gamma, \psi; \Phi \vdash (\phi \ol\tau) [\phi :== \psi]} f
-\end{align*}
-\item If $\Gamma; \Phi, \phi \vdash (\mu \phi'. \lambda
-  \ol\alpha. H)\ol\tau : \F$, then the induction hypothesis gives that
-\[\setsem{\Gamma; \Phi, \phi', \ol{\alpha}, \phi \vdash H } \rho =
-\setsem{\Gamma, \psi; \Phi, \phi', \ol{\alpha} \vdash H[\phi :== \psi]
-} \rho\]
-and 
-\[\setsem{\Gamma; \Phi, \phi \vdash \tau } \rho = \setsem{\Gamma,
-  \psi; \Phi \vdash \tau[\phi :== \psi] } \rho\]
-as well as that
-\[\setsem{\Gamma; \Phi, \phi', \ol{\alpha}, \phi \vdash H } f =
-\setsem{\Gamma, \psi; \Phi, \phi', \ol{\alpha} \vdash H[\phi :== \psi]
-} f\]
-and 
-\[\setsem{\Gamma; \Phi, \phi \vdash \tau } f = \setsem{\Gamma, \psi;
-  \Phi \vdash \tau[\phi :== \psi] } f\]
-for each $\tau$. Then
-\begin{align*}
-& \setsem{\Gamma; \Phi, \phi \vdash (\mu \phi'. \lambda
-    \ol\alpha. H)\ol\tau } \rho \\ 
-= \; & (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma; \Phi, \phi',
-  \ol{\alpha}, \phi \vdash H} \rho[\phi' := F][\ol{\alpha := A}]))
-\ol{\setsem{ \Gamma; \Phi, \phi \vdash \tau} \rho} \\ 
-= \; & (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma, \psi; \Phi,
-  \phi', \ol{\alpha} \vdash H[\phi :== \psi]} \rho[\phi' :=
-  F][\ol{\alpha := A}])) \ol{\setsem{ \Gamma; \Phi, \phi \vdash \tau}
-  \rho} \\ 
-= \; &  (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma, \psi; \Phi,
-  \phi', \ol{\alpha} \vdash H[\phi :== \psi]} \rho[\phi' :=
-  F][\ol{\alpha := A}])\ol{\setsem{ \Gamma, \psi; \Phi \vdash
-    \tau[\phi :== \psi]} \rho} \\ 
-= \; & \setsem{\Gamma, \psi; \Phi \vdash (\mu \phi'. \lambda
-  \ol\alpha. H[\phi :== \psi]) \ol{\tau[\phi :== \psi]} } \rho \\ 
-= \; & \setsem{\Gamma, \psi; \Phi \vdash ((\mu \phi'. \lambda
-  \ol\alpha. H) \ol\tau)[\phi :== \psi] } \rho   
-\end{align*} 
-The first and fifth equalities are by Definition~\ref{def:set-sem}.
-The second equality follows from the following equality:
-\begin{align*}
-&\lambda F. \lambda \ol{A}. \setsem{\Gamma; \Phi, \phi', \ol{\alpha},
-    \phi \vdash H} \rho[\phi' := F][\ol{\alpha := A}] \\ 
-& = \lambda F. \lambda \ol{A}. \setsem{\Gamma, \psi; \Phi, \phi',
-    \ol{\alpha} \vdash H[\phi :== \psi]} \rho[\phi' := F][\ol{\alpha
-      := A}] 
-\end{align*}
-These two maps have the same actions on objects and morphisms by the
-induction hypothesis on $H$, and the fact that the extended
-environment $\rho[\phi' := F][\ol{\alpha := A}]$ satisfies the
-required hypothesis. They are thus equal as functors and so have the
-same fixpoint. We also have that 
-\begin{align*}
-& \setsem{\Gamma; \Phi, \phi \vdash (\mu \phi'. \lambda
-    \ol\alpha. H)\ol\tau } f \\ 
-= \; & (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma; \Phi, \phi',
-  \ol{\alpha}, \phi \vdash H} f [\phi' := \id_F][\ol{\alpha :=
-    \id_A}]))_{\ol{\setsem{\Gamma; \Phi, \phi \vdash \tau} \rho'}} \\ 
-\; & \circ (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma; \Phi,
-  \phi', \ol{\alpha}, \phi \vdash H} \rho[\phi' := F][\ol{\alpha :=
-    A}])) \ol{\setsem{ \Gamma; \Phi, \phi \vdash \tau} f} \\ 
-= \; & (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma; \Phi, \phi',
-  \ol{\alpha}, \phi \vdash H} f [\phi' := \id_F][\ol{\alpha :=
-    \id_A}]))_{\ol{\setsem{\Gamma, \psi; \Phi \vdash \tau[\phi :==
-        \psi]} \rho'}} \\ 
-\; & \circ (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma; \Phi,
-  \phi', \ol{\alpha}, \phi \vdash H} \rho[\phi' := F][\ol{\alpha :=
-    A}])) \ol{\setsem{ \Gamma, \psi; \Phi \vdash \tau[\phi :== \psi]}
-  f} \\ 
-= \; & (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma, \psi; \Phi,
-  \phi', \ol{\alpha} \vdash H[\phi :== \psi]} f [\phi' :=
-  \id_F][\ol{\alpha := \id_A}]))_{\ol{\setsem{\Gamma, \psi; \Phi
-      \vdash \tau[\phi :== \psi]} \rho'}} \\ 
-\; & \circ (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma; \Phi,
-  \phi', \ol{\alpha}, \phi \vdash H} \rho[\phi' := F][\ol{\alpha :=
-    A}])) \ol{\setsem{ \Gamma, \psi; \Phi \vdash \tau[\phi :== \psi]}
-  f} \\ 
-= \; & (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma, \psi; \Phi,
-  \phi', \ol{\alpha} \vdash H[\phi :== \psi]} f [\phi' :=
-  \id_F][\ol{\alpha := \id_A}]))_{\ol{\setsem{\Gamma, \psi; \Phi
-      \vdash \tau[\phi :== \psi]} \rho'}} \\ 
-\; & \circ (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma, \psi;
-  \Phi, \phi', \ol{\alpha} \vdash H[\phi :== \psi]} \rho[\phi' :=
-  F][\ol{\alpha := A}])) \ol{\setsem{ \Gamma, \psi; \Phi \vdash
-    \tau[\phi :== \psi]} f} \\ 
-= \; & \setsem{\Gamma, \psi; \Phi \vdash (\mu \phi'. \lambda
-  \ol\alpha. H[\phi :== \psi]) \ol{\tau[\phi :== \psi]} } f \\ 
-= \; & \setsem{\Gamma, \psi; \Phi \vdash ((\mu \phi'. \lambda
-  \ol\alpha. H) \ol\tau)[\phi :== \psi] } f 
-\end{align*} 
-The first and fifth equalities are by Definition~\ref{def:set-sem}.
-The third equality is by the equality of the arguments to the first
-$\mu$ operator:
-\begin{align*}
-& \lambda F. \lambda \ol{A}. \setsem{\Gamma; \Phi, \phi', \ol{\alpha},
-    \phi \vdash H} f [\phi' := \id_F][\ol{\alpha := \id_A}] \\ 
-& = \lambda F. \lambda \ol{A}. \setsem{\Gamma, \psi; \Phi, \phi',
-    \ol{\alpha} \vdash H[\phi :== \psi]} f [\phi' := \id_F][\ol{\alpha
-      := \id_A}] 
-\end{align*}
-By the induction hypothesis on $H$ and the fact that the morphism
-$f[\phi' := \id_F][\ol{\alpha := \id_A}] : \rho[\phi' := F][\ol{\alpha
-    := A}] \to \rho'[\phi' := F][\ol{\alpha := A}]$ still satisfies
-the required hypotheses.  The fourth equality is by the equality of
-the arguments to the second $\mu$ operator:
-\begin{align*}
-& \lambda F. \lambda \ol{A}. \setsem{\Gamma; \Phi, \phi', \ol{\alpha},
-    \phi \vdash H} \rho[\phi' := F][\ol{\alpha := A}] \\ 
-& = \lambda F. \lambda \ol{A}. \setsem{\Gamma, \psi; \Phi, \phi',
-    \ol{\alpha} \vdash H[\phi :== \psi]} \rho[\phi' := F][\ol{\alpha
-      := A}] 
-\end{align*}
-By the same reasoning as above, these two maps are equal as functors,
-and thus have the same fixpoint.
-\end{itemize}
-\end{proof}
-
----
-
-The following lemma ensures that substitution interacts well with type
-interpretations.  It is a consequence of
-Definitions~\ref{def:second-order-subst},~\ref{def:set-interp},
-and~\ref{def:rel-interp}.
-
-\begin{lemma}\label{lem:substitution}
-Let $\rho$ and $\rho'$ be set environments and let $f : \rho \to \rho'$ be a
-morphism of set environments. 
-\begin{itemize}
-\item If $\Gamma;\Phi,\ol{\alpha} \vdash F$ and $\Gamma;\Phi \vdash \tau$,
-  then  
-\begin{gather}\label{eq:subs-var}
-\setsem{\Gamma;\Phi \vdash F[\ol{\alpha := \tau}]}\rho =
-\setsem{\Gamma;\Phi,\ol{\alpha} \vdash F}\rho[\ol{\alpha :=
-\setsem{\Gamma;\Phi \vdash \tau}\rho}]\\
-\intertext{and}\label{eq:subs-var-morph}
-\setsem{\Gamma;\Phi \vdash F[\ol{\alpha := \tau}]}f =
-\setsem{\Gamma;\Phi,\ol{\alpha} \vdash F}f[\ol{\alpha :=
-\setsem{\Gamma;\Phi \vdash \tau}f}]
-\end{gather}
-\item If $\Gamma;\Phi,\phi^k\vdash F$ and
-  $\Gamma;\Phi,\alpha_1...\alpha_k \vdash H$, then 
-\begin{gather}\label{eq:subs-const}
-\setsem{\Gamma; \Phi \vdash F[\phi := H]}\rho
-= \setsem{\Gamma; \Phi, \phi \vdash F}\rho
-[\phi := \lambda \ol{A}.\, \setsem{\Gamma;\Phi,\overline{\alpha}\vdash
-    H}\rho[\overline{\alpha := A}]] \\ 
-\intertext{and}\label{eq:subs-const-morph}
-\setsem{\Gamma; \Phi \vdash F[\phi := H]}f
-= \setsem{\Gamma; \Phi, \phi \vdash F}f
-[\phi := \lambda \ol{A}.\,\setsem{\Gamma;\Phi,\overline{\alpha}\vdash
-    H}f[\overline{\alpha := \id_{\ol{A}}}]] 
-\end{gather}
-\end{itemize}
-Analogous identities hold for relation environments and morphisms
-between them.
-\end{lemma}
-\end{comment}
-
-\begin{comment}
-\begin{proof}
-The proofs for the set and relational interpretations are completely
-analogous, so we just prove the former. Likewise, we only prove
-Equations~\ref{eq:subs-var} and~\ref{eq:subs-const}, since the proofs
-for Equations~\ref{eq:subs-var-morph} and~\ref{eq:subs-const-morph}
-are again analogous. Finally, we prove Equation~\ref{eq:subs-var} for
-substitution for just a single type variable since the proof for
-multiple simultaneous substitutions proceeds similarly.
-
-Although Equation~\ref{eq:subs-var} is a special case of
-Equation~\ref{eq:subs-const}, it is convenient to prove
-Equation~\ref{eq:subs-var} first, and then use it to prove
-Equation~\ref{eq:subs-const}. We prove Equation~\ref{eq:subs-var} by
-induction on the structure of $F$ as follows:
-\begin{itemize}
-\item If $\Gamma; \emptyset \vdash F : \T$, or if $F$ is $\onet$ or
-  $\zerot$, then $F$ does not contain any functorial variables to
-  replace, so there is nothing to prove.
-\item If $F$ is $F_1 \times F_2$ or $F_1 + F_2$, then the substitution
-  distributes over the product or coproduct as appropriate, so the
-  result follows immediately from the induction hypothesis.
-\item If $F = \beta$ with $\beta \neq \alpha$, then there is nothing to
-  prove.
-\item If $F = \alpha$, then
-\[\setsem{\Gamma;\Phi\vdash\alpha[\alpha := \tau]}\rho \, = \,
-\setsem{\Gamma;\Phi\vdash\tau}\rho \, = \,
-\setsem{\Gamma;\Phi,\alpha\vdash\alpha}\rho[\alpha :=
-  \setsem{\Gamma;\Phi\vdash\tau}\rho]\]
-\item If $F = \phi \overline{\sigma}$ with $\phi \neq \alpha$, then
-\[\begin{array}{ll}
- &\setsem{\Gamma; \Phi \vdash (\phi \overline{\sigma})[\alpha := \tau]}\rho\\
-=&\setsem{\Gamma; \Phi \vdash \phi (\overline{\sigma[\alpha := \tau]})}\rho \\
-=&(\rho\phi)\overline{\setsem{\Gamma;\Phi\vdash\sigma[\alpha := \tau]}\rho} \\
-=&(\rho\phi)\overline{\setsem{\Gamma;\Phi,\alpha\vdash\sigma}
-      \rho [\alpha := \setsem{\Gamma;\Phi\vdash\tau}] } \\
-=&\setsem{\Gamma; \Phi, \alpha \vdash \phi \overline{\sigma}}
-      \rho [\alpha := \setsem{\Gamma;\Phi\vdash\tau}]
-\end{array}\]
-Here, the third equality is by the induction hypothesis.
-\item If $F = (\mu \phi. \lambda \overline{\beta}. G)
-  \overline{\sigma}$, then
-\[\begin{array}{ll}
-  & \setsem{\Gamma; \Phi \vdash ((\mu \phi. \lambda
-  \overline{\beta}. G) \overline{\sigma}) [\alpha := \tau]}\rho \\
-=&\setsem{\Gamma; \Phi \vdash (\mu \phi. \lambda
-  \overline{\beta}. G[\alpha := \tau]) (\overline{\sigma [\alpha :=
-      \tau]})}\rho \\ 
-=&\mu(\setsem{\Gamma; \Phi, \phi, \overline{\beta} \vdash G[\alpha := \tau]}
-      \rho[\phi := \text{--}][\overline{\beta := \text{--}}])
-      (\overline{\setsem{\Gamma;\Phi \vdash \sigma [\alpha := \tau]}\rho}) \\
-=&\mu(\setsem{\Gamma; \Phi, \phi, \overline{\beta}, \alpha \vdash G}
-\rho [\alpha := \setsem{\Gamma;\Phi \vdash \tau}\rho] [\phi := \text{--}]
-[\overline{\beta := \text{--}}]) \\
- &\hspace{0.5in} (\overline{\setsem{\Gamma;\Phi,\alpha \vdash \sigma}
-   \rho [\alpha := \setsem{\Gamma;\Phi\vdash \tau}\rho]}) \\
-=&\setsem{\Gamma;\Phi,\alpha\vdash(\mu\phi.\lambda\overline{\beta}.G)
-  \overline{\sigma}}\rho  [\alpha := \setsem{\Gamma;\Phi \vdash
-   \tau}\rho] 
-\end{array}  \]
-Here, the third equality is by the induction hypothesis and weakening.
-\end{itemize}
-
-We now prove Equation~\ref{eq:subs-const}, again by induction on the
-structure of $F$.
-\begin{itemize}
-\item If $\Gamma; \emptyset \vdash F : \T$, or if $F$ is $\onet$ or
-  $\zerot$, then $F$ does not contain any functorial variables to
-  replace, so there is nothing to prove.
-\item If $F$ is $F_1 \times F_2$ or $F_1 + F_2$, then the substitution
-  distributes over the product or coproduct as appropriate, so the
-  result follows immediately from the induction hypothesis.
-\item If $F = \phi \overline{\tau}$, then
-\[\begin{array}{ll}
- &\setsem{\Gamma; \Phi \vdash (\phi \overline{\tau})[\phi := H]}\rho \\
-=&\setsem{\Gamma; \Phi \vdash H[\overline{\alpha:= \tau[\phi := H]}]}\rho \\
-=&\setsem{\Gamma; \Phi \vdash H}\rho [\overline{\alpha:=
-    \setsem{\Gamma;\Phi\vdash\tau[\phi := H]}\rho}] \\ 
-=&\setsem{\Gamma; \Phi \vdash H}\rho [\overline{\alpha:=
-    \setsem{\Gamma;\Phi,\phi\vdash\tau} \rho [\phi :=
-      \setsem{\Gamma;\Phi,\overline{\alpha}\vdash
-        H}\rho[\overline{\alpha := \text{--}}]]}] \\
-=&\setsem{\Gamma; \Phi, \phi \vdash \phi \overline{\tau}}\rho [\phi :=
-  \setsem{\Gamma;\Phi,\overline{\alpha}\vdash H}\rho[\overline{\alpha
-      := \text{--}}]] 
-  \end{array}\]
-Here, the first equality is by
-Definition~\ref{def:second-order-subst}, the second is by
-Equation~\ref{eq:subs-var}, the third is by the induction hypothesis,
-and the fourth is by Definition~\ref{def:set-sem}.
-\item If $F = \psi \overline{\tau}$ with $\psi \neq \phi$, then the
-  proof is similar to that for the previous case, but simpler, because
-  $\phi$ only needs to be substituted in the arguments $\ol{\tau}$ of
-  $\psi$.
-\item If $F = (\mu \psi. \lambda \overline{\beta}. G)
-  \overline{\tau}$, then
-\[\begin{array}{ll}
- &\setsem{\Gamma; \Phi \vdash ((\mu \psi. \lambda \overline{\beta}. G)
- \overline{\tau})[\phi := H]}\rho \\ 
-=&\setsem{\Gamma; \Phi \vdash (\mu \psi. \lambda
-  \overline{\beta}. G[\phi := H]) (\overline{\tau[\phi := H]})}\rho \\
-=&\mu(\setsem{\Gamma; \Phi, \psi, \overline{\beta} \vdash G[\phi := H]}
-    \rho[\psi := \text{--}][\overline{\beta := \text{--}}])
-    (\overline{\setsem{\Gamma;\Phi \vdash \tau[\phi := H]}\rho}) \\
-=&\mu(\setsem{\Gamma; \Phi, \psi, \overline{\beta}, \phi \vdash G}
-\rho [\phi := \setsem{\Gamma;\Phi,\overline{\alpha}\vdash
-    H}\rho[\overline{\alpha := \text{--}}]][\psi := \text{--}]
-     [\overline{\beta := \text{--}}]) \\
-&\hspace{0.5in} (\overline{\setsem{\Gamma;\Phi,\phi \vdash \tau} \rho
-       [\phi := \setsem{\Gamma;\Phi,\overline{\alpha}\vdash H}
-         \rho[\overline{\alpha := \text{--}}]]}) \\
-=&\setsem{\Gamma;\Phi,\phi\vdash(\mu\psi.\lambda\overline{\beta}.G)
-  \overline{\tau}}\rho [\phi :=
-  \setsem{\Gamma;\Phi,\overline{\alpha}\vdash H}\rho[\overline{\alpha
-      := \text{--}}]] 
-\end{array}\]
-Here, the first equality is by
-Definition~\ref{def:second-order-subst}, the second and fourth are by
-Definition~\ref{def:set-sem}, and the third is by the induction
-hypothesis and weakening.
-\end{itemize}
-\end{proof}
-\end{comment}
-
-\section{The Identity Extension Lemma}\label{sec:iel}
-
-In most treatments of parametricity, equality relations on sets are
-taken as {\em given} --- either directly as diagonal relations, or
-perhaps via reflexive graphs if kinds are also being tracked --- and
-the graph relations used to validate existence of initial algebras are
-defined in terms of them. We take a different approach here, giving a
-categorical definition of graph relations for morphisms (i.e., natural
-transformations) between functors and {\em constructing} equality
-relations as particular graph relations. Our definitions specialize to
-the usual ones for the graph relation for a morphism between sets and
-the equality relation for a set. In light of its novelty, we spell out
-our construction in detail.
-
-The standard definition of the graph for a morphism $f : A \to B$ in
-$\set$ is the relation $\graph{f} : \rel(A,B)$ defined by $(x,y) \in
-\graph{f}$ iff $fx = y$. This definition naturally generalizes to
-associate to each natural transformation between $k$-ary functors on
-$\set$ a $k$-ary relation transformer as follows:
-
-\begin{dfn}\label{dfn:graph-nat-transf}
-If $F, G: \Set^k \to \Set$
-%are $k$-ary functors
-and $\alpha : F \to G$ is a natural transformation, then the functor
-$\graph{\alpha}^*: \rel^k \to \rel$ is defined as follows. Given $R_1
-: \rel(A_1, B_1),...,R_k : \rel(A_k,B_k)$, let $\iota_{R_i} : R_i
-\hookrightarrow A_i \times B_i$, for $i = 1,...,k$, be the inclusion
-of $R_i$ as a subset of $A_i \times B_i$,
-%. By the universal property of the product, there exists a
-let $h_{\overline{A \times B}}$ be the unique morphism making the diagram
-{\footnotesize\[\begin{tikzcd}[row sep = large]
-        F\overline{A}
-        &F(\overline{A \times B})
-        \ar[l, "{F\overline{\pi_1}}"']
-        \ar[r, "{F\overline{\pi_2}}"]
-        \ar[d, dashed, "{h_{\overline{A \times B}}}"]
-        &F\overline{B}
-        \ar[r, "{\alpha_{\ol{B}}}"]
-        &G\overline{B} \\
-        &F\overline{A} \times G\overline{B}
-        \ar[ul, "{\pi_1}"] \ar[urr, "{\pi_2}"']
-\end{tikzcd}\]}
-
-\noindent
-commute, and let $h_{\overline{R}} : F\overline{R} \to F\overline{A}
-\times G\overline{B}$ be $h_{\overline{A \times B}} \circ
-F\overline{\iota_R}$. Further, let $\alpha^\wedge\overline{R}$ be the
-subobject through which $h_{\overline{R}}$ is factorized by the
-mono-epi factorization system in $\set$, as shown in the
-following diagram:
-{\footnotesize\[\begin{tikzcd}
-        F\overline{R}
-        \ar[rr, "{h_{\overline{R}}}"]
-        \ar[dr, twoheadrightarrow, "{q_{\alpha^\wedge\overline{R}}}"']
-        &&F\overline{A} \times G\overline{B} \\
-        &\alpha^\wedge\overline{R}
-        \ar[ur, hookrightarrow, "{\iota_{\alpha^\wedge\overline{R}}}"']
-\end{tikzcd}\]}
-
-\noindent
-Then $\alpha^\wedge\overline{R} : \rel(F\overline{A}, G\overline{B})$
-by construction, so the action of $\langle \alpha \rangle^*$ on
-objects can be given by $\langle \alpha \rangle^* \overline{(A,B,R)} =
-(F\overline{A}, G\overline{B}, \iota_{\alpha^\wedge
-  \overline{R}}\alpha^\wedge\overline{R})$. Its action on morphisms is
-given by $\graph{\alpha}^*\overline{(\beta, \beta')} =
-(F\overline\beta, G\overline\beta')$.
-\end{dfn}
-
-The data in Definition~\ref{dfn:graph-nat-transf} yield the {\em graph relation
-transformer for $\alpha$}, denoted $\graph{\alpha} = (F, G,
-\graph{\alpha}^*)$.
-% yield a relation transformer $\graph{\alpha} = (F, G,
-%\graph{\alpha}^*)$, called the {\em graph relation transformer for
-%$\alpha$}.
-
-\begin{lemma}\label{lem:graph-reln-functors}
-If $F,G : [\set^k,\set]$, and if $\alpha : F \to G$ is a natural
-transformation, then $\graph{\alpha}$ is in $RT_k$.
-\end{lemma}
-\begin{proof}
-Clearly, $\graph{\alpha}^*$ is $\omega$-cocontinuous, so
-$\graph{\alpha}^* : [\rel^k,\rel]$. Now, suppose $\overline{R :
-  \rel(A, B)}$, $\overline{S : \rel(C, D)}$, and $\overline{(\beta,
-  \beta') : R \to S}$. We want to show that there exists a morphism
-$\epsilon : \alpha^\wedge\overline{R} \to \alpha^\wedge\overline{S}$
-such that the diagram on the left below commutes. Since
-$\ol{(\beta,\beta') : R \to S}$, there exist $\overline{\gamma : R \to
-  S}$ such that each diagram in the middle commutes.
-Moreover, since both $h_{\overline{C \times D}} \circ
-F(\overline{\beta \times \beta'})$ and $(F\overline{\beta} \times
-G\overline{\beta'}) \circ h_{\overline{A \times B}}$ make the diagram
-on the right commute, they must be equal.
-\begin{figure*}[ht]
-\vspace*{-0.1in}
-  \hspace*{-0.85in}
-  \begin{minipage}[b]{0.25\linewidth}
- {\small    \[
-    \begin{tikzcd}
-        \alpha^\wedge\overline{R}
-        \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{R}}}"]
-        \ar[d, "{\epsilon}"']
-        & F\overline{A} \times G\overline{B}
-        \ar[d, "{F\overline{\beta} \times G\overline{\beta'}}"] \\
-        \alpha^\wedge\overline{S}
-        \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{S}}}"']
-        & F\overline{C} \times G\overline{D}
-    \end{tikzcd}
-    \]}
-\end{minipage}
-\begin{minipage}[b]{0.25\linewidth}
-{\small    \[
-    \begin{tikzcd}
-        R_i
-        \ar[d, "{\gamma_i}"']
-        \ar[r, hookrightarrow, "{\iota_{R_i}}"]
-        &A_i \times B_i
-        \ar[d, "{\beta_i \times \beta'_i}"] \\
-        S_i
-        \ar[r, hookrightarrow, "{\iota_{S_i}}"]
-        &C_i \times D_i
-    \end{tikzcd}
-    \]}
-\end{minipage}
-\begin{minipage}[b]{0.25\linewidth}
-{\footnotesize \[
-      \begin{tikzcd}[row sep = large]
-          F\overline{C}
-          &F\overline{C} \times F\overline{D}
-          \ar[l, "{\pi_1}"'] \ar[r, "{\pi_2}"]
-          &F\overline{D}
-          \ar[r, "{\alpha_{\ol{D}}}"]
-          &G\overline{D}\\
-          &F(\overline{A \times B})
-          \ar[u, dashed, "{\exists !}"]
-          \ar[ul, "{F\ol{\pi_1} \circ F(\overline{\beta \times \beta'})}"]
-          \ar[urr, "{\alpha_{\ol{D}} \circ F\ol{\pi_2} \circ F(\overline{\beta \times \beta'})}"']
-      \end{tikzcd}
-      \]}
-\end{minipage}
-\end{figure*}
-%{\footnotesize    \[
-%    \begin{tikzcd}
-%        \alpha^\wedge\overline{R}
-%        \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{R}}}"]
-%        \ar[d, "{\epsilon}"']
-%        & F\overline{A} \times G\overline{B}
-%        \ar[d, "{F\overline{\beta} \times G\overline{\beta'}}"] \\
-%        \alpha^\wedge\overline{S}
-%        \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{S}}}"']
-%        & F\overline{C} \times G\overline{D}
-%    \end{tikzcd}
-%    \]}
-%    commutes.
-%    Since $\ol{(\beta,\beta') : R \to S}$, there exist $\overline{\gamma : R \to S}$
-%    such that each diagram
-%{\footnotesize    \[
-%    \begin{tikzcd}
-%        R_i
-%        \ar[d, "{\gamma_i}"']
-%        \ar[r, hookrightarrow, "{\iota_{R_i}}"]
-%        &A_i \times B_i
-%        \ar[d, "{\beta_i \times \beta'_i}"] \\
-%        S_i
-%        \ar[r, hookrightarrow, "{\iota_{S_i}}"]
-%        &C_i \times D_i
-%    \end{tikzcd}
-%    \]}
-%    commutes.
-%Moreover, since both $h_{\overline{C \times D}} \circ
-%F(\overline{\beta \times \beta'})$ and $(F\overline{\beta} \times
-%G\overline{\beta'}) \circ h_{\overline{A \times B}}$ make
-%{\footnotesize \[
-%      \begin{tikzcd}[row sep = large]
-%          F\overline{C}
-%          &F\overline{C} \times F\overline{D}
-%          \ar[l, "{\pi_1}"'] \ar[r, "{\pi_2}"]
-%          &F\overline{D}
-%          \ar[r, "{\alpha_{\ol{D}}}"]
-%          &G\overline{D}\\
-%          &F(\overline{A \times B})
-%          \ar[u, dashed, "{\exists !}"]
-%          \ar[ul, "{F\ol{\pi_1} \circ F(\overline{\beta \times \beta'})}"]
-%          \ar[urr, "{\alpha_{\ol{D}} \circ F\ol{\pi_2} \circ F(\overline{\beta \t%imes \beta'})}"']
-%      \end{tikzcd}
-%      \]}
-%      commute, they must be equal.
-
-\vspace*{-0.1in}
-
-\noindent
-We therefore get that the right-hand square in the diagram on the left
-below commutes, and thus that the entire diagram does as well.
-Finally, by the left-lifting property of $q_{F^\wedge\overline{R}}$
-with respect to $\iota_{F^\wedge\overline{S}}$ given by the epi-mono
-factorization system, there exists an $\epsilon$ such that the diagram
-on the right below commutes.
-\begin{figure*}[ht]
-  \vspace*{-0.1in}
-  \hspace*{-0.5in}
-  \begin{minipage}[b]{0.45\linewidth}
-{\footnotesize \[
-      \begin{tikzcd}
-          F\overline{R}
-          \ar[d, "{F\overline{\gamma}}"']
-          \ar[r, hookrightarrow, "{F\overline{\iota_R}}"]
-          \ar[rr, bend left, "{h_{\overline{R}}}"]
-          &F(\overline{A \times B})
-          \ar[d, "{F(\overline{\beta \times \beta'})}"]
-          \ar[r, "{h_{\overline{A \times B}}}"]
-          &F\overline{A} \times G\overline{B}
-          \ar[d, "{F\overline{\beta} \times F\overline{\beta'}}"] \\
-          F\overline{S}
-          \ar[r, hookrightarrow, "{F\overline{\iota_S}}"']
-          \ar[rr, bend right, "{h_{\overline{S}}}"']
-          &F(\overline{C \times D})
-          \ar[r, "{h_{\overline{C \times D}}}"']
-          &F\overline{C} \times G\overline{D}
-      \end{tikzcd}
-      \]}
-\end{minipage}
-  \vspace*{-0.5in}
-  \begin{minipage}[b]{0.45\linewidth}
-      {\footnotesize
-        \[  \vspace*{0.4in}
-      \begin{tikzcd}
-          F\overline{R}
-          \ar[d, "{F\overline{\gamma}}"']
-          \ar[r, twoheadrightarrow, "{q_{\alpha^\wedge\overline{R}}}"]
-          &\alpha^\wedge\overline{R}
-          \ar[d, dashed, "{\epsilon}"]
-          \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{R}}}"]
-          &F\overline{A} \times G\overline{B}
-          \ar[d, "{F\overline{\beta} \times G\overline{\beta'}}"] \\
-          F\overline{S}
-          \ar[r, twoheadrightarrow, "{q_{\alpha^\wedge\overline{S}}}"']
-          &\alpha^\wedge\overline{S}
-          \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{S}}}"']
-          &F\overline{C} \times G\overline{D}
-      \end{tikzcd}
-      \]}
-\end{minipage}
-\end{figure*}
-
-\begin{comment}
-
-Finally, by the left-lifting property of $q_{F^\wedge\overline{R}}$
-      with respect to $\iota_{F^\wedge\overline{S}}$ given by the epi-mono
-      factorization system, there exists an $\epsilon$ such that the
-      following diagram commutes:
-      {\footnotesize
-        \[
-      \begin{tikzcd}
-          F\overline{R}
-          \ar[d, "{F\overline{\gamma}}"']
-          \ar[r, twoheadrightarrow, "{q_{\alpha^\wedge\overline{R}}}"]
-          &\alpha^\wedge\overline{R}
-          \ar[d, dashed, "{\epsilon}"]
-          \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{R}}}"]
-          &F\overline{A} \times G\overline{B}
-          \ar[d, "{F\overline{\beta} \times G\overline{\beta'}}"] \\
-          F\overline{S}
-          \ar[r, twoheadrightarrow, "{q_{\alpha^\wedge\overline{S}}}"']
-          &\alpha^\wedge\overline{S}
-          \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{S}}}"']
-          &F\overline{C} \times G\overline{D}
-      \end{tikzcd}
-      \]}
-
-      \vspace*{-0.1in}
-\end{comment}
-\end{proof}
-
-\vspace*{0.1in}
-
-If $f : A \to B$ is a morphism in $\set$ then the definition of the
-graph relation transformer $\langle f \rangle$ for $f$ as a natural
-transformation between $0$-ary functors $A$ and $B$ coincides with its
-standard definition.  Graph relation transformers are thus a
-reasonable extension of graph relations to functors.
-\begin{comment}
-
-\vspace*{0.2in}
-
-If $f : A \to B$ is a function with graph relation $\graph{f} = (A, B,
-\graph{f}^*)$, then $\langle \id_{A}, f \rangle : A \to A \times B$
-and $\langle \id_{A}, f \rangle\, A = \graph{f}^*$.  Moreover, if
-$\iota_{\graph{f}} : \graph{f}^* \hookrightarrow A \times B$ is the
-inclusion of $\graph{f}^*$ into $A \times B$ then there is an
-isomorphism of subobjects
-\[\begin{tikzcd}
-A \ar[rr, "{\cong}"] \ar[dr, "{\langle \id_{A}, f \rangle}"']
-&&{\graph{f}^*} \ar[dl, "{\iota_{\graph{f}}}"]\\
-&A \times B
-\end{tikzcd}\]
-
-We also note that if $f : A \to B$ is a function seen as a natural
-transformation between 0-ary functors, then $\graph{f}$ is (the 0-ary
-relation transformer associated with) the graph relation of $f$.
-Indeed, we need to apply Definition~\ref{dfn:graph-nat-transf} with $k
-= 0$, i.e., to the degenerate relation $\ast : \rel(\ast, \ast)$.  As
-degenerate $0$-ary functors, $A$ and $B$ are constant functors, i.e.,
-$A\, \ast = A$ and $B\, \ast = B$.  By the universal property of the
-product, there exists a unique $h$ making the diagram
-\[ \begin{tikzcd}[row sep = large]
-        A
-        &A
-        \ar[l, equal]
-        \ar[r, equal]
-        \ar[d, dashed, "{h}"]
-        &A
-        \ar[r, "{f}"]
-        &B \\
-        &A \times B
-        \ar[ul, "{\pi_1}"] \ar[urr, "{\pi_2}"']
-\end{tikzcd}\]
-commute. Since $\iota_\ast : \ast \to \ast$ is the identity on $\ast$,
-and $A\, \id_{\ast} = \id_{A}$, we have $h_{\ast} = h$.  Moreover,
-$h_{\overline{A \times B}} = \langle \id_{A}, f \rangle$ is a
-monomorphism in $\set$ because $\id_{A}$ is.  Then,
-$\iota_{f^\wedge\ast} = \langle \id_{A}, f \rangle$ and $f^\wedge\ast
-= A$, from which we deduce that $\iota_{f^\wedge\ast} f^\wedge\ast =
-\langle \id_{A}, f \rangle\, A = \graph{f}^*$. This ensures that the
-graph of $f$ as a 0-ary natural transformation coincides with the
-graph of $f$ as a morphism in $\set$, and so that
-Definition~\ref{dfn:graph-nat-transf} is a reasonable generalization
-of Definition~\ref{def:graph}.
-
-Just as the equality relation $\Eq_B$ on a set $B$ coincides with
-$\graph{\id_B}$, the graph of the identity on the set, so we can
-define the equality relation transformer to be the graph of the
-identity natural transformation. This gives
-
-\begin{dfn}
-Let $F : [\set^k, \set]$.  The equality relation transformer on $F$ is
-defined to be $\Eq_F = \graph{\id_{F}}$. This entails that $Eq_F = (F,
-F, \Eq_F^*)$ with $\Eq_F^* = \graph{\id_{F}}^*$.
-\end{dfn}
-\end{comment}
-
-To prove the IEL, we will need to know that the equality relation
-transformer preserves equality relations; see
-Equation~\ref{eq:eq-pres-eq} below. This will follow from the next
-lemma, which shows how to compute the action of a graph relation
-transformer on any graph relation.
-
-\begin{lemma}\label{lem:eq-reln-equalities}
-If $\alpha : F \to G$ is a morphism in $[\Set^k, \Set]$
-and $f_1: A_1 \to B_1, ..., f_k : A_k \to B_k$,
-then $\graph{\alpha}^* \graph{\overline{f}}
-= \langle G \ol{f} \circ \alpha_{\ol{A}} \rangle
-= \langle \alpha_{\ol{B}} \circ F \ol{f} \rangle$.
-\end{lemma}
-\begin{proof}
-Since $h_{\overline{A \times B}}$ is the unique morphism making the
-bottom triangle of the diagram on the left below commute, and since
-$h_{\graph{\overline{f}}} = h_{\overline{A \times B}} \circ F
-\,\ol{\iota_{\graph{f}}} = h_{\overline{A \times B}} \circ F
-\overline{\langle \id_A, f \rangle}$, the universal property of the
-product depicted in the diagram on the right gives
-$h_{\graph{\overline{f}}} = \langle \id_{F \ol{A}}, \alpha_{\ol{B}}
-\circ F\ol{f} \rangle : F \ol{A} \to F \ol{A} \times G \ol{B}$.
-
-\begin{figure*}[ht]
-  \vspace*{-0.15in}
-%  \hspace*{-0.5in}
-  \begin{minipage}[b]{0.45\linewidth}
-{\footnotesize
-\[\begin{tikzcd}[row sep = large]
-        &F\overline{A}
-        \ar[d, "{F \overline{\langle \id_A, f \rangle}}" description]
-        \ar[dl, equal]
-        \ar[dr, "{F\ol{f}}"]\\
-        F\overline{A}
-        &F(\overline{A \times B})
-        \ar[l, "{F\overline{\pi_1}}"']
-        \ar[r, "{F\overline{\pi_2}}"]
-        \ar[d, "{h_{\overline{A \times B}}}"]
-        &F\overline{B}
-        \ar[r, "{\alpha_{\ol{B}}}"]
-        &G\overline{B}\\
-        &F\overline{A} \times G\overline{B}
-        \ar[ul, "{\pi_1}"] \ar[urr, "{\pi_2}"']
-\end{tikzcd}\]}
-\end{minipage}
-%  \vspace*{-0.5in}
-  \begin{minipage}[b]{0.45\linewidth}
-{\footnotesize
-\[
-      \begin{tikzcd}[row sep = large]
-          F\overline{A}
-          &F\overline{A} \times G\overline{B}
-          \ar[l, "{\pi_1}"'] \ar[r, "{\pi_2}"]
-          &G\overline{B}\\
-          &F\overline{A}
-          \ar[u, dashed, "{\exists !}"]
-          \ar[ul, equal]
-          \ar[r, "{F\ol{f}}"']
-          &F{\ol{B}}
-          \ar[u, "{\alpha_{\ol{B}}}"']
-      \end{tikzcd}
-      \]}
-\end{minipage}
-\end{figure*}
-
-\vspace*{-0.1in}
-
-\noindent
-Moreover, $\langle \id_{F \ol{A}}, \alpha_{\ol{B}} \circ
-F\ol{f} \rangle$ is a monomorphism in $\set$ because $\id_{F \ol{A}}$
-is, so its epi-mono factorization gives $\iota_{\alpha^\wedge
-  \graph{\overline{f}}} = \langle \id_{F \ol{A}}, \alpha_{\ol{B}}
-\circ F\ol{f} \rangle$, and thus $\alpha^\wedge \graph{\overline{f}} =
-%the domain of $\iota_{\alpha^\wedge \graph{\overline{f}}}$, is equal
-%to
-F\overline{A}$.  Then $\iota_{\alpha^\wedge
-  \graph{\overline{f}}} \alpha^\wedge \graph{\overline{f}} = \langle
-\id_{F \ol{A}}, \alpha_{\ol{B}} \circ F\ol{f} \rangle (F \ol{A}) =
-\graph{ \alpha_{\ol{B}} \circ F\ol{f} }^*$,
-%(where the last equality is by Remark~\ref{rmk:graph-fn}).
-%We therefore conclude that
-so that $\graph{ \alpha }^* \graph{ \overline{f} } = (F\overline{A},
-G\overline{B}, \iota_{\alpha^\wedge \graph{\overline{f}}}\,
-\alpha^\wedge \graph{\overline{f}}) = (F\overline{A}, G\overline{B},
-\graph{ \alpha_{\ol{B}} \circ F\ol{f} }^*) = \graph{ \alpha_{\ol{B}}
-  \circ F\ol{f} }$.  Finally, $\alpha_{\ol{B}} \circ F\ol{f} = G\ol{f}
-\circ \alpha_{\ol{A}}$ by naturality of $\alpha$.
-\end{proof}
-
-The {\em equality relation transformer} on $F : [\set^k,\set]$ is
-defined to be $\Eq_F = \graph{\id_{F}}$. Specifically, $\Eq_F = (F, F,
-\Eq_F^*)$ with $\Eq_F^* = \graph{\id_{F}}^*$, and
-Lemma~\ref{lem:eq-reln-equalities} indeed ensures that
-\begin{equation}\label{eq:eq-pres-eq}
-\Eq^*_F \ol{\Eq_A}
-= \graph{\id_F}^* \graph{\id_{\ol{A}}}
-= \graph{F \id_{\ol{A}} \circ (\id_F)_{\ol{A}}}
-= \graph{\id_{F\ol{A}} \circ \id_{F\ol{A}}}
-= \graph{\id_{F\ol{A}}}
-= \Eq_{F\ol{A}}
-\end{equation}
-%$\Eq^*_F \,\ol{\Eq_A} = \Eq_{F\ol{A}}$
-for all $\ol{A : \set}$.
-\begin{comment}
-  \begin{lemma}
-$\Eq^*_F \,\ol{\Eq_A}
-= \Eq_{F\ol{A}}$ for all $\ol{A : \set}$.
-  \end{lemma}
-\begin{proof}
-We have that\
-\begin{equation}\label{eq:eq-pres-eq}
-\Eq^*_F \ol{\Eq_A}
-= \graph{\id_F}^* \graph{\id_{\ol{A}}}
-= \graph{F \id_{\ol{A}} \circ (\id_F)_{\ol{A}}}  %by your Lemma 5
-= \graph{\id_{F\ol{A}} \circ \id_{F\ol{A}}}
-%= \graph{\id_F \ol{A} \circ \id_{\ol{A}}}
-= \graph{\id_{F\ol{A}}}
-= \Eq_{F\ol{A}}
-\end{equation}
-The second identity here is by Lemma~\ref{lem:eq-reln-equalities}.
-\end{proof}
-\end{comment}
-Graph relation transformers in general, and equality relation
-transformers in particular, extend to relation environments in the
-obvious ways.
-%\begin{definition}
-Indeed, if $\rho, \rho' : \setenv$ and $f : \rho \to \rho'$, then the
-{\em graph relation environment} $\graph{f}$ is defined pointwise by
-$\graph{f} \phi = \graph{f \phi}$ for every $\phi$, which entails that
-$\pi_1 \graph{f} = \rho$ and $\pi_2 \graph{f} = \rho'$. In particular,
-the {\em equality relation environment} $\Eq_\rho$ is defined to be
-$\graph{\id_{\rho}}$, which entails that $\Eq_\rho \phi = \Eq_{\rho
-  \phi}$ for every $\phi$.
-%\end{definition}
-With these definitions in hand, we can state and prove both an
-Identity Extension Lemma and a Graph Lemma for our calculus.
-\begin{thm}[IEL]\label{thm:iel}
-  If $\rho : \setenv$ and $\Gamma; \Phi \vdash F$ then
-  $\relsem{\Gamma; \Phi \vdash F} \Eq_\rho = \Eq_{\setsem{\Gamma;
-      \Phi \vdash F}\rho}$.
-\end{thm}
-The proof is by induction on the structure of $F$. Only the $\Nat$,
-application, and fixpoint cases are non-routine. The latter two use
-Lemma~\ref{lem:eq-reln-equalities}. The fixpoint case also uses the
-observation that, for every $n \in \nat$, the following intermediate
-results can be proved by simultaneous induction with
-Theorem~\ref{thm:iel}: for any $H$, $\rho$, $A$, and any subformula
-$J$ of $H$, both
-%\begin{equation}\label{eq:iel-fix-point-intermediate1}
-$T^n_{H,\Eq_{\rho}} K_0\, \ol{\Eq_A}
-%{\setsem{\Gamma; \Phi \vdash F}\rho}}
-= (\Eq_{(T^\set_{H,\rho})^n K_0})^*
-\ol{\Eq_A}$\;
-  %{\setsem{\Gamma; \Phi \vdash F}\rho}}$\;
-%\end{equation}
-and\;
-%\begin{equation}\label{eq:iel-fix-point-intermediate2}
-%\begin{split}
-\[\begin{array}{ll}
-&  \relsem{\Gamma; \Phi, \phi, \ol{\alpha} \vdash J} \Eq_{\rho} [\phi
-  := T^{n}_{H,\Eq_{\rho}} K_0] \overline{[\alpha := \Eq_A]}\\
-%{\setsem{\Gamma; \Phi \vdash F}\rho}]}\\
-=  &
-\relsem{\Gamma; \Phi, \phi,
-  \ol{\alpha} \vdash J} \Eq_{\rho} [\phi := \Eq_{(T^\set_{H,\rho})^n K_0}]
-\overline{[\alpha := \Eq_A]}
-%         {\setsem{\Gamma; \Phi \vdash F}\rho}]}.
-\end{array}\]
-%\end{split}
-%\end{equation}
-hold. With these results in hand, the proof follows easily. It is
-given in detail in the accompanying anonymous supplementary
-material. As noted there, if functorial variables of arity greater
-than $0$ were allowed to appear in the bodies of $\mu$-types, then the
-IEL would fail.
-
-\begin{comment}
-\begin{proof}
-By induction on the structure of $\tau$.
-\begin{itemize}
-\item $\relsem{\Gamma; \emptyset \vdash v}\Eq_{\rho} = \Eq_{\rho} v =
-  \Eq_{\rho v} = \Eq_{\setsem{\Gamma; \emptyset \vdash v}\rho}$ where
-  $v \in \Gamma$.
-\item By definition, $\relsem{\Gamma; \emptyset \vdash
-  \Nat^{\overline\alpha} \,F\,G} \Eq_{\rho}$ is the relation on
-  $\setsem{\Gamma; \emptyset \vdash \Nat^{\overline\alpha} \,F\,G}
-  \rho$ relating $t$ and $t'$ if, for all ${R_1 :
-    \rel(A_1,B_1)},...,{R_k : \rel(A_k,B_k)}$, $(t_{\overline{A}},
-  t'_{\overline{B}})$ is a morphism from $\relsem{\Gamma;
-    \overline\alpha \vdash F} \Eq_{\rho}\overline{[\alpha := R]}$ to
-  $\relsem{\Gamma ; \overline\alpha \vdash G}
-  \Eq_{\rho}\overline{[\alpha := R]}$ in $\rel$.  To prove that this
-  is equal to $\Eq_{\setsem{\Gamma; \emptyset \vdash
-      \Nat^{\overline\alpha} \,F\,G} \rho}$ we need to show that
-  $(t_{\overline{A}}, t'_{\overline{B}})$ is a morphism from
-  $\relsem{\Gamma; \overline\alpha \vdash F}
-  \Eq_{\rho}\overline{[\alpha := R]}$ to $\relsem{\Gamma ;
-    \overline\alpha \vdash G} \Eq_{\rho}\overline{[\alpha := R]}$ in
-  $\rel$ for all ${R_1 : \rel(A_1,B_1)},...,{R_k : \rel(A_k,B_k)}$ if
-  and only if $t = t'$ and $(t_{\overline{A}}, t_{\overline{B}})$ is a
-  morphism from $\relsem{\Gamma; \overline\alpha \vdash F}
-  \Eq_{\rho}\overline{[\alpha := R]}$ to $\relsem{\Gamma ; \overline
-    \alpha \vdash G} \Eq_{\rho}\overline{[\alpha := R]}$ in $\rel$ for
-  all ${R_1 : \rel(A_1,B_1)}, ...,$ ${R_k : \rel(A_k,B_k)}$. The only
-  interesting part of this equivalence is to show that if
-  $(t_{\overline{A}}, t'_{\overline{B}})$ is a morphism from
-  $\relsem{\Gamma; \overline\alpha \vdash F}
-  \Eq_{\rho}\overline{[\alpha := R]}$ to $\relsem{\Gamma ;
-    \overline\alpha \vdash G} \Eq_{\rho}\overline{[\alpha := R]}$ in
-  $\rel$ for all ${R_1 : \rel(A_1,B_1),}$ $...,{R_k : \rel(A_k,B_k)}$,
-  then $t = t'$.  By hypothesis, $(t_{\overline{A}},
-  t'_{\overline{A}})$ is a morphism from $\relsem{\Gamma;
-    \overline\alpha \vdash F} \Eq_{\rho}\overline{[\alpha :=
-      \Eq_{A}]}$ to $\relsem{\Gamma ; \overline\alpha \vdash G}
-  \Eq_{\rho}\overline{[\alpha := \Eq_{A}]}$ in $\rel$ for all
-  $A_1\,...\,A_k : \set$. By the induction hypothesis, it is therefore
-  a morphism from $\Eq_{\setsem{\Gamma; \overline\alpha \vdash F}
-    \rho\overline{[\alpha := A]}}$ to $\Eq_{\setsem{\Gamma ;
-      \overline\alpha \vdash G} \rho\overline{[\alpha := A]}}$ in
-  $\rel$. This means that, for every $x : \Eq_{\setsem{\Gamma;
-      \overline\alpha \vdash F} \rho\overline{[\alpha := A]}}$,
-  $t_{\overline{A}}x = t'_{\overline{A}}x$.  Then, by extensionality,
-  $t = t'$.
-\item $\relsem{\Gamma; \Phi \vdash \zerot} \Eq_{\rho} = 0_\rel =
-  \Eq_{0_\set} = \Eq_{\setsem{\Gamma; \Phi \vdash \zerot}\rho}$
-\item $\relsem{\Gamma; \Phi \vdash \onet} \Eq_{\rho} = 1_\rel =
-  \Eq_{1_\set} = \Eq_{\setsem{\Gamma; \Phi \vdash \onet}\rho}$
-\item The application case is proved by the following sequence of
-  equalities, where the second equality is by the induction hypothesis
-  and the definition of the relation environment $\Eq_\rho$, the third
-  is by the definition of application of relation transformers, and
-  the fourth is by Lemma~\ref{lem:eq-reln-equalities}:
-\[
-\begin{split}
-\relsem{\Gamma; \Phi \vdash \phi\ol{\tau}}\Eq_{\rho} &=
-(\Eq_{\rho}\phi)\ol{\relsem{\Gamma; \Phi \vdash \tau}
-\Eq_{\rho}}\\
-&= \Eq_{\rho \phi}\, \ol{\Eq_{\setsem{\Gamma; \Phi \vdash \tau}
-  \rho}}\\
-&= (\Eq_{\rho \phi})^* \,\ol{\Eq_{\setsem{\Gamma; \Phi \vdash \tau}
-  \rho}}\\
-&= \Eq_{(\rho \phi) \,\ol{\setsem{\Gamma; \Phi \vdash \tau} \rho}}\\
-&= \Eq_{\setsem{\Gamma; \Phi \vdash \phi\ol{\tau}}\rho}
-\end{split}
-\]
-\item The fixpoint case is proven by the sequence of equalities
-\[
-\begin{split}
-\relsem{\Gamma; \Phi \vdash (\mu \phi.\lambda
-  \ol{\alpha}. H)\ol{\tau}}\Eq_{\rho} 
-&=(\mu {T_{\Eq_{\rho}}}) \,\ol{\relsem{\Gamma; \Phi \vdash \tau}\Eq_{\rho}}\\ 
-&= \colim{n \in \nat}{T^n_{\Eq_{\rho}} K_0}\, \ol{\relsem{\Gamma; \Phi
-  \vdash \tau}\Eq_{\rho}}\\
-&= \colim{n \in \nat}{ T^n_{\Eq_{\rho}} K_0 \,\ol{\Eq_{\setsem{\Gamma;
-    \Phi \vdash \tau}\rho}}}\\
-&= \colim{n \in \nat}{(\Eq_{(T^\set_\rho)^n K_0})^*
-  \ol{\Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}}}\\
-&= \colim{n \in \nat}{\Eq_{(T^\set_\rho)^n K_0 \,\ol{\setsem{\Gamma;
-        \Phi \vdash \tau}\rho}}}\\ 
-&= \Eq_{\colim{n \in \nat}{ (T^\set_{\rho})^n K_0\,
-    \ol{\setsem{\Gamma; \Phi \vdash \tau}\rho}}}\\
-&= \Eq_{\setsem{\Gamma; \Phi \vdash (\mu \phi.\lambda
-      \ol{\alpha}. H)\ol{\tau}}\rho}
-\end{split}
-\]
-Here, the third equality is by induction hypothesis, the fifth is by
-Lemma~\ref{lem:eq-reln-equalities} and the fourth equality is because,
-for every $n \in \nat$, the following two statements can be proved by
-simultaneous induction:
-\begin{equation}\label{eq:iel-fix-point-intermediate1}
-T^n_{\Eq_{\rho}} K_0\, \ol{\Eq_{\setsem{\Gamma; \Phi \vdash
-      \tau}\rho}} = (\Eq_{(T^\set_\rho)^n K_0})^*
-\ol{\Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}}
-\end{equation}
-and
-\begin{equation}\label{eq:iel-fix-point-intermediate2}
-\begin{split}
-  \relsem{\Gamma; \Phi, \phi, \ol{\alpha} \vdash H}
-\Eq_{\rho} [\phi := 
- & T^{n}_{\Eq_{\rho}} K_0] \overline{[\alpha :=
-    \Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}]} \\
-=\;\; & \relsem{\Gamma; \Phi, \phi, \ol{\alpha} \vdash H} \Eq_{\rho} [\phi
-  := \Eq_{(T^\set_\rho)^n K_0}] \overline{[\alpha :=
-    \Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}]}
-\end{split}
-\end{equation}
-We prove~\eqref{eq:iel-fix-point-intermediate1}.  The case $n=0$ is
-trivial, because $T^0_{\Eq_{\rho}} K_0 = K_0$ and
-$(T^\set_\rho)^0 K_0 = K_0$; the inductive step is
-proved by the following sequence of equalities:
-\[
-\begin{split}
-T^{n+1}_{\Eq_{\rho}} K_0\, \overline{\Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}}
-&= T^\rel_{\Eq_{\rho}} (T^{n}_{\Eq_{\rho}} K_0)
-\overline{\Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}} \\ 
-&= \relsem{\Gamma; \Phi, \phi, \ol{\alpha} \vdash H} \Eq_{\rho} [\phi
-  := T^{n}_{\Eq_{\rho}} K_0] \overline{[\alpha :=
-    \Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}]} \\ 
-&= \relsem{\Gamma; \Phi, \phi, \ol{\alpha} \vdash H} \Eq_{\rho} [\phi
-  := \Eq_{(T^\set_\rho)^n K_0}] \overline{[\alpha :=
-    \Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}]} \\ 
-&= \relsem{\Gamma; \Phi, \phi, \ol{\alpha} \vdash H} \Eq_{\rho [\phi
-    := (T^\set_\rho)^n K_0] \overline{[\alpha :=
-      \setsem{\Gamma; \Phi \vdash \tau}\rho]}} \\ 
-&= \Eq_{\setsem{\Gamma; \Phi, \phi, \ol{\alpha} \vdash H} \rho [\phi
-    := (T^\set_\rho)^n K_0] \overline{[\alpha :=
-      \setsem{\Gamma; \Phi \vdash \tau}\rho]}} \\ 
-&= \Eq_{(T^\set_\rho)^{n+1} K_0 \overline{\setsem{\Gamma; \Phi
-      \vdash \tau}\rho}} \\ 
-&= (\Eq_{(T^\set_\rho)^{n+1} K_0})^*\, \overline{\Eq_{\setsem{\Gamma;
-      \Phi \vdash \tau}\rho}} 
-\end{split}
-\]
-Here, the third equality is by~\eqref{eq:iel-fix-point-intermediate2},
-the fifth by the induction hypothesis on $H$, and the last by
-Lemma~\ref{lem:eq-reln-equalities}.  We prove the induction step
-of~\eqref{eq:iel-fix-point-intermediate2} by structural induction on
-$H$: the only interesting case, though, is when $\phi$ is applied,
-i.e., for $H = \phi \ol{\sigma}$, which is proved by the sequence of
-equalities:
-\[
-\begin{split}
-& \relsem{\Gamma; \Phi, \phi, \ol{\alpha} \vdash \phi
-    \ol{\sigma}} \Eq_{\rho} [\phi := T^{n}_{\Eq_{\rho}} K_0]
-  \overline{[\alpha := \Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}]}
-  \\
-&= T^{n}_{\Eq_{\rho}} K_0\, \overline{\relsem{\Gamma; \Phi,
-      \phi, \ol{\alpha} \vdash \sigma} \Eq_{\rho} [\phi :=
-      T^{n}_{\Eq_{\rho}} K_0] \overline{[\alpha :=
-        \Eq_{\setsem{\Gamma; \Phi \vdash \tau} \rho}]}} \\ 
-&= T^{n}_{\Eq_{\rho}} K_0\, \overline{\relsem{\Gamma; \Phi,
-      \phi, \ol{\alpha} \vdash \sigma} \Eq_{\rho} [\phi :=
-      \Eq_{(T^\set_\rho)^{n} K_0}] \overline{[\alpha :=
-        \Eq_{\setsem{\Gamma; \Phi \vdash \tau} \rho}]}} \\ 
-&= T^{n}_{\Eq_{\rho}} K_0\, \overline{\relsem{\Gamma; \Phi,
-      \phi, \ol{\alpha} \vdash \sigma} \Eq_{\rho [\phi := (T^\set_\rho)^{n}
-        K_0] \overline{[\alpha := \setsem{\Gamma; \Phi \vdash
-            \tau} \rho]}}} \\ 
-&= T^{n}_{\Eq_{\rho}} K_0\, \overline{\Eq_{\setsem{\Gamma;
-        \Phi, \phi, \ol{\alpha} \vdash \sigma} \rho [\phi :=
-        (T^\set_\rho)^{n} K_0] \overline{[\alpha :=
-          \setsem{\Gamma; \Phi \vdash \tau} \rho]}}} \\ 
-&= (\Eq_{(T^\set_\rho)^{n} K_0})^* \,\overline{\Eq_{\setsem{\Gamma;
-        \Phi, \phi, \ol{\alpha} \vdash \sigma} \rho [\phi :=
-        (T^\set_\rho)^{n} K_0] \overline{[\alpha :=
-          \setsem{\Gamma; \Phi \vdash \tau} \rho]}}} \\ 
-&= (\Eq_{(T^\set_{\rho})^{n} K_0})^* \overline{\relsem{\Gamma;
-      \Phi, \phi, \ol{\alpha} \vdash \sigma} \Eq_{\rho} [\phi :=
-      \Eq_{(T^\set_{\rho})^{n} K_0}] \overline{[\alpha :=
-        \Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}]}} \\ 
-&= \relsem{\Gamma; \Phi, \phi, \ol{\alpha} \vdash \phi \ol{\sigma}}
-  \Eq_{\rho} [\phi := \Eq_{(T^\set_{\rho})^{n} K_0}]
-  \overline{[\alpha := \Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}]} 
-\end{split}
-\]
-Here, the second equality is by the induction hypothesis
-for~\eqref{eq:iel-fix-point-intermediate2} on the $\sigma$s, the
-fourth is by the induction hypothesis for Theorem~\ref{thm:iel} on the
-$\sigma$s, and the fifth is by the induction hypothesis on $n$
-for~\eqref{eq:iel-fix-point-intermediate1}.
-\item $\relsem{\Gamma; \Phi \vdash \sigma + \tau} \Eq_{\rho} =
-  \relsem{\Gamma; \Phi \vdash \sigma} \Eq_{\rho} + \relsem{\Gamma;
-    \Phi \vdash \tau} \Eq_{\rho} = \Eq_{\setsem{\Gamma; \Phi \vdash
-      \sigma}\rho} + \Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho} =
-  \Eq_{\setsem{\Gamma; \Phi \vdash \sigma}\rho + \setsem{\Gamma; \Phi
-      \vdash \tau}\rho} = \Eq_{\setsem{\Gamma; \Phi \vdash \sigma +
-      \tau}\rho}$
-\item $\relsem{\Gamma; \Phi \vdash \sigma \times \tau} \Eq_{\rho} =
-  \relsem{\Gamma; \Phi \vdash \sigma}\Eq_{\rho} \times \relsem{\Gamma;
-    \Phi \vdash \tau}\Eq_{\rho} = \Eq_{\setsem{\Gamma; \Phi \vdash
-      \sigma}\rho} \times \Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}
-  = \Eq_{\setsem{\Gamma; \Phi \vdash \sigma}\rho \times
-    \setsem{\Gamma; \Phi \vdash \tau}\rho} = \Eq_{\setsem{\Gamma;
-      \Phi \vdash \sigma \times \tau}\rho}$
-\end{itemize}
-\end{proof}
-\end{comment}
-
-%It follows from Theorem~\ref{thm:iel} that $\relsem{\Gamma \vdash
-%  \sigma \to \tau} \Eq_{\rho} = \Eq_{\setsem{\Gamma \vdash \sigma \to
-%    \tau}\rho}$, as expected.  Moreover,
-
-%\vspace*{0.05in}
-%
-%The Graph Lemma appropriate to our setting is now easily obtained:
-\begin{lemma}[Graph Lemma]\label{lem:graph}
-If $\rho, \rho' : \setenv$, $f : \rho \to \rho'$, and $\Gamma; \Phi
-\vdash F$, then $\graph{\setsem{\Gamma; \Phi \vdash F} f} =
-\relsem{\Gamma; \Phi \vdash F}\graph{f}$.
-\end{lemma}
-\begin{proof}
-Applying Lemma~\ref{lem:rel-transf-morph} to the morphisms $(f,
-\id_{\rho'}) : \graph{f} \to \Eq_{\rho'}$ and $(\id_{\rho}, f) :
-\Eq_{\rho} \to \graph{f}$ of relation environments gives that
-$(\setsem{\Gamma; \Phi \vdash F}f, \setsem{\Gamma; \Phi \vdash
-  F}\id_{\rho'}) = \relsem{\Gamma; \Phi \vdash F} (f, \id_{\rho'}) :
-\relsem{\Gamma; \Phi \vdash F}\graph{f} \to \relsem{\Gamma; \Phi
-  \vdash F}\Eq_{\rho'}$ and $(\setsem{\Gamma; \Phi \vdash
-  F}\id_{\rho}, \setsem{\Gamma; \Phi \vdash F}f) = \relsem{\Gamma;
-  \Phi \vdash F} (\id_{\rho}, f) : \relsem{\Gamma; \Phi \vdash
-  F}\Eq_{\rho} \to \relsem{\Gamma; \Phi \vdash F}\graph{f}$.
-Expanding the first equation gives that if $(x,y) \in \relsem{\Gamma;
-  \Phi \vdash F}\graph{f}$ then $(\setsem{\Gamma; \Phi \vdash F} f\,
-x, \setsem{\Gamma; \Phi \vdash F}\id_{\rho'}\, y) \in \relsem{\Gamma;
-  \Phi \vdash F}\Eq_{\rho'}$. So $\setsem{\Gamma; \Phi \vdash
-  F}\id_{\rho'}\, y = \id_{\setsem{\Gamma; \Phi \vdash F}\rho'}\, y =
-y$ and $\relsem{\Gamma; \Phi \vdash F}\Eq_{\rho'} =
-\Eq_{\setsem{\Gamma; \Phi \vdash F}\rho'}$, and if $(x,y) \in
-\relsem{\Gamma; \Phi \vdash F}\graph{f}$ then $(\setsem{\Gamma; \Phi
-  \vdash F} f\, x, y) \in \Eq_{\setsem{\Gamma; \Phi \vdash F}\rho'}$,
-i.e., $\setsem{\Gamma; \Phi \vdash F} f\, x = y$, i.e., $(x, y) \in
-\graph{\setsem{\Gamma; \Phi \vdash F} f}$.  So, we have that
-$\relsem{\Gamma; \Phi \vdash F}\graph{f} \subseteq
-\graph{\setsem{\Gamma; \Phi \vdash F}f}$.  Expanding the second
-equation gives that if $x \in \setsem{\Gamma; \Phi \vdash F}\rho$ then
-$(\setsem{\Gamma; \Phi \vdash F}\id_{\rho}\, x,$\\
-$\setsem{\Gamma; \Phi
-  \vdash F} f\, x) \in \relsem{\Gamma; \Phi \vdash F}\graph{f}$.  Then
-$\setsem{\Gamma; \Phi \vdash F}\id_{\rho}\, x = \id_{\setsem{\Gamma;
-    \Phi \vdash F}\rho} x = x$, so for any $x \in \setsem{\Gamma; \Phi
-  \vdash F}\rho$ we have that $(x, \setsem{\Gamma; \Phi \vdash F}f\,
-x) \in \relsem{\Gamma; \Phi \vdash F}\graph{f}$.  Moreover, $x \in
-\setsem{\Gamma; \Phi \vdash F}\rho$ if and only if $(x,
-\setsem{\Gamma; \Phi \vdash F} f\, x) \in \graph{\setsem{\Gamma; \Phi
-    \vdash F}f}$ and, if $x \in \setsem{\Gamma; \Phi \vdash F}\rho$
-then $(x, \setsem{\Gamma; \Phi \vdash F} f\, x) \in \relsem{\Gamma;
-  \Phi \vdash F} \graph{f}$, so if $(x, \setsem{\Gamma; \Phi \vdash F}
-f\, x) \in \graph{\setsem{\Gamma; \Phi \vdash F}f}$ then $(x,
-\setsem{\Gamma; \Phi \vdash F} f\, x) \in \relsem{\Gamma; \Phi \vdash
-  F} \graph{f}$, i.e., $\graph{\setsem{\Gamma; \Phi \vdash F}f}
-\subseteq \relsem{\Gamma; \Phi \vdash F} \graph{f}$.
-
-\end{proof}
-
-\begin{comment}
-
-\begin{proof}
-First observe that $(f, \id_{\rho'}) : \graph{f} \to \Eq_{\rho'}$ and
-$(\id_{\rho}, f) : \Eq_{\rho} \to \graph{f}$ are morphisms of relation
-environments.  Applying Lemma~\ref{lem:rel-transf-morph} to each of
-these observations gives that
-\begin{equation}\label{eq:graph-one}
-(\setsem{\Gamma; \Phi \vdash F}f, \setsem{\Gamma; \Phi \vdash
-    F}\id_{\rho'}) = \relsem{\Gamma; \Phi \vdash F} (f, \id_{\rho'}) :
-  \relsem{\Gamma; \Phi \vdash F}\graph{f} \to \relsem{\Gamma; \Phi
-    \vdash F}\Eq_{\rho'}
-\end{equation}
-and
-\begin{equation}\label{eq:graph-two}
-(\setsem{\Gamma; \Phi \vdash F}\id_{\rho}, \setsem{\Gamma; \Phi \vdash F}f)
-= \relsem{\Gamma; \Phi \vdash F} (\id_{\rho}, f)
-: \relsem{\Gamma; \Phi \vdash F}\Eq_{\rho} \to \relsem{\Gamma; \Phi \vdash F}\graph{f}
-\end{equation}
-Expanding Equation~\ref{eq:graph-one} gives that if
-$(x,y) \in \relsem{\Gamma; \Phi \vdash F}\graph{f}$
-then
-\[(\setsem{\Gamma; \Phi \vdash F} f\, x, \setsem{\Gamma; \Phi \vdash
-  F}\id_{\rho'}\, y) \in \relsem{\Gamma; \Phi \vdash F}\Eq_{\rho'}\]
-Observe that $\setsem{\Gamma; \Phi \vdash F}\id_{\rho'}\, y =
-\id_{\setsem{\Gamma; \Phi \vdash F}\rho'}\, y = y$ and
-$\relsem{\Gamma; \Phi \vdash F}\Eq_{\rho'} = \Eq_{\setsem{\Gamma; \Phi
-    \vdash F}\rho'}$. So, if $(x,y) \in \relsem{\Gamma; \Phi \vdash
-  F}\graph{f}$ then $(\setsem{\Gamma; \Phi \vdash F} f\, x, y) \in
-\Eq_{\setsem{\Gamma; \Phi \vdash F}\rho'}$, i.e., $\setsem{\Gamma;
-  \Phi \vdash F} f\, x = y$, i.e., $(x, y) \in \graph{\setsem{\Gamma;
-    \Phi \vdash F} f}$.  So, we have that $\relsem{\Gamma; \Phi \vdash
-  F}\graph{f} \subseteq \graph{\setsem{\Gamma; \Phi \vdash F}f}$
-
-Expanding Equation~\ref{eq:graph-two} gives that, for any
-$x \in \setsem{\Gamma; \Phi \vdash F}\rho$,
-then
-\[
-(\setsem{\Gamma; \Phi \vdash F}\id_{\rho}\, x, \setsem{\Gamma; \Phi
-  \vdash F} f\, x) \in \relsem{\Gamma; \Phi \vdash F}\graph{f} 
-\]
-Observe that $\setsem{\Gamma; \Phi \vdash F}\id_{\rho}\, x =
-\id_{\setsem{\Gamma; \Phi \vdash F}\rho} x = x$ so, for any $x \in
-\setsem{\Gamma; \Phi \vdash F}\rho$, we have that $(x, \setsem{\Gamma;
-  \Phi \vdash F}f\, x) \in \relsem{\Gamma; \Phi \vdash F}\graph{f}$.
-Moreover, $x \in \setsem{\Gamma; \Phi \vdash F}\rho$ if and only if
-$(x, \setsem{\Gamma; \Phi \vdash F} f\, x) \in \graph{\setsem{\Gamma;
-    \Phi \vdash F}f}$ and, if $x \in \setsem{\Gamma; \Phi \vdash
-  F}\rho$ then $(x, \setsem{\Gamma; \Phi \vdash F} f\, x) \in
-\relsem{\Gamma; \Phi \vdash F} \graph{f}$, so if $(x, \setsem{\Gamma;
-  \Phi \vdash F} f\, x) \in \graph{\setsem{\Gamma; \Phi \vdash F}f}$
-then $(x, \setsem{\Gamma; \Phi \vdash F} f\, x) \in \relsem{\Gamma;
-  \Phi \vdash F} \graph{f}$, i.e., $\graph{\setsem{\Gamma; \Phi \vdash
-    F}f} \subseteq \relsem{\Gamma; \Phi \vdash F} \graph{f}$. We
-conclude that $\relsem{\Gamma; \Phi \vdash F}\graph{f} =
-\graph{\setsem{\Gamma; \Phi \vdash F}f}$ as desired.
-\end{proof}
-\end{comment}
-
-\vspace*{-.2in}
-
-\section{Interpreting Terms}\label{sec:term-interp}
-
-If $\Delta = x_1 : F_1,...,x_n : F_n$ is a term context for $\Gamma$
-and $\Phi$, then the interpretations $\setsem{\Gamma;\Phi \vdash \Delta}$ and
-$\relsem{\Gamma;\Phi \vdash \Delta}$ are defined by
-\[\begin{array}{lll}
-\setsem{\Gamma;\Phi \vdash \Delta} & = & \setsem{\Gamma;\Phi \vdash
-  F_1} \times ... \times \setsem{\Gamma;\Phi \vdash F_n}\\ 
-\relsem{\Gamma;\Phi \vdash \Delta} & = & \relsem{\Gamma;\Phi \vdash
-  F_1} \times ... \times \relsem{\Gamma;\Phi \vdash F_n}\\ 
-\end{array}\]
-Every well-formed term $\Gamma;\Phi~|~\Delta \vdash t : F$ then
-has, for every $\rho \in \setenv$, set interpretations
-$\setsem{\Gamma;\Phi~|~\Delta \vdash t : F}\rho$ as natural
-transformations from $\setsem{\Gamma; \Phi \vdash \Delta}\rho$ to
-$\setsem{\Gamma; \Phi \vdash F}\rho$, and, for every $\rho \in
-\relenv$, relational interpretations $\relsem{\Gamma;\Phi~|~\Delta
-  \vdash t : F}\rho$ as natural transformations from
-$\relsem{\Gamma; \Phi \vdash \Delta}\rho$ to $\relsem{\Gamma; \Phi
-  \vdash F}\rho$. These are given in Definition~\ref{def:set-interp}.
-
-\begin{dfn}\label{def:set-interp}
-If $\rho$ is a set (resp., relation) environment and
-$\Gamma;\Phi~|~\Delta \vdash t : F$ then
-$\setsem{\Gamma;\Phi~|~\Delta \vdash t : F}\rho$ (resp.,
-$\relsem{\Gamma;\Phi~|~\Delta \vdash t : F}\rho$) is defined as in
-Figure~\ref{fig:term-sem}, where $\mathsf D$ is either $\set$ or
-$\rel$ as appropriate.
-\end{dfn}
-
-\subsection{Basic Properties of Term Interpretations}
-
-The interpretations in Definition~\ref{def:set-interp} respect
-weakening, i.e., a term and its weakenings all have the same set and
-relational interpretations. Specifically, for any $\rho \in \setenv$,
-$\setsem{\Gamma;\Phi \,|\, \Delta, x : F \vdash t : G}\rho =
-(\setsem{\Gamma;\Phi \,|\, \Delta \vdash t : G}\rho) \circ
-\pi_{\Delta}$, where $\pi_{\Delta}$ is the projection
-$\setsem{\Gamma;\Phi \vdash \Delta, x : F} \to
-\setsem{\Gamma;\Phi \vdash \Delta}$. A similar result holds for
-relational interpretations.
-
-\begin{figure*}
-\hspace*{-3.1in}
-\resizebox{0.46\linewidth}{!}{
-\begin{minipage}[t]{0.5\textwidth}
-\[\begin{array}{lll}
-\dsem{\Gamma;\Phi \,|\, \Delta,x :F \vdash x : F} \rho& = &
-\pi_{|\Delta|+1}\\
-%\dsem{\Gamma;\emptyset \,|\, \Delta \vdash \lambda x.t : \sigma \to \tau}\rho &
-%= & \curry (\dsem{\Gamma;\emptyset \,|\, \Delta, x : \sigma \vdash t :
-%  \tau}\rho)\\ 
-%\dsem{\Gamma;\emptyset \,|\, \Delta \vdash st: \tau} \rho & = &
-%\eval \circ
-% \langle \dsem{\Gamma;\emptyset \,|\, \Delta \vdash s: \sigma \to
-%  \tau}\rho, \dsem{\Gamma;\emptyset \,|\, \Delta \vdash t: \sigma}\rho
-%\rangle\\
-\dsem{\Gamma;\emptyset \,|\, \Delta \vdash L_{\overline \alpha} x.t : \Nat^{\overline
-    \alpha} \,F \,G}\rho & = &  \curry (\dsem{\Gamma;\overline \alpha
-  \,|\, \Delta, x : F \vdash t: G}\rho[\overline{\alpha := \_}])\\
-\dsem{\Gamma;\Phi \,|\, \Delta \vdash t_{\overline K} s:
-  G [\overline{\alpha := K}]}\rho & = & \eval \circ \langle
-  \lambda d.\,(\dsem{\Gamma;\emptyset \,|\, \Delta \vdash t :
-  \Nat^{\overline{\alpha}} \,F \,G}\rho\; d)_{\overline{\dsem{\Gamma;\Phi
-      \vdash K}\rho}},\\ 
- & & \hspace*{0.5in} \dsem{\Gamma;\Phi \,|\,
-    \Delta \vdash s: F [\overline{\alpha := K}]}\rho \rangle\\ 
-& & \\
-%% \color{red} \mbox{Add rules for } \forall \mbox{ if we include it} & & \\
-%\dsem{\Gamma;\Phi \,|\, \Delta,x :F \vdash x : F} \rho& = &
-%\pi_{|\Delta|+1}\\
-\dsem{\Gamma;\Phi \,|\, \Delta \vdash \bot_F t : F} \rho& = &
-!^0_{\dsem{\Gamma;\Phi \vdash F}\rho} \circ
-  \dsem{\Gamma;\Phi~|~\Delta \vdash t : \zerot}\rho, \mbox{ where } \\
- & & \hspace*{0.1in} !^0_{\dsem{\Gamma;\Phi \vdash F}\rho}
-\mbox{ is the unique morphism from } 0\\
- & & \hspace*{0.1in} \mbox{ to } \dsem{\Gamma;\Phi \vdash F}\rho\\
-\dsem{\Gamma;\Phi \,|\, \Delta \vdash \top : \onet}\rho & = &
-!^{\dsem{\Gamma;\Phi\vdash \Delta}\rho}_1, \mbox{ where }
-!^{\dsem{\Gamma;\Phi\vdash \Delta}\rho}_1\\ 
-& & \hspace*{0.1in} \mbox{ is the unique morphism from }
-\dsem{\Gamma;\Phi\vdash \Delta}\rho \mbox{ to } 1\\ 
-\dsem{\Gamma;\Phi \,|\, \Delta \vdash (s,t) : F \times G} \rho& = &
-\dsem{\Gamma;\Phi \,|\, \Delta \vdash s: F} \rho\times
-\dsem{\Gamma;\Phi \,|\, \Delta \vdash t : G} \rho\\
-\dsem{\Gamma;\Phi \,|\, \Delta \vdash \pi_1 t : F} \rho& = &
-\pi_1 \circ \dsem{\Gamma;\Phi \,|\, \Delta \vdash t : F \times G}\rho\\
-\dsem{\Gamma;\Phi \,|\, \Delta \vdash \pi_2 t : G}\rho & = &
-\pi_2 \circ \dsem{\Gamma;\Phi \,|\, \Delta \vdash t : F \times
-  G} \rho\\
-\dsem{\Gamma;\Phi~|~\Delta \vdash \case{t}{x \mapsto l}{y \mapsto r} :
-  K}\rho & = & \eval \circ \langle \curry \,[\dsem{\Gamma;\Phi
-    \,|\, \Delta, x : F \vdash l : K}\rho,\\
-   & & \hspace*{0.79in} \dsem{\Gamma;\Phi \,|\, \Delta, y
-    : G \vdash r : K}\rho],\\
-   & &  \hspace*{0.5in} \dsem{\Gamma;\Phi \,|\, \Delta \vdash t :
-  F + G} \rho\rangle\\   
-\dsem{\Gamma;\Phi \,|\, \Delta \vdash \inl \,s: F + G} \rho& = &
-\inl \circ \dsem{\Gamma;\Phi \,|\, \Delta \vdash s: F}\rho\\
-\dsem{\Gamma;\Phi \,|\, \Delta \vdash \inr \,t: F + G}\rho & = & 
-\inr \circ \dsem{\Gamma;\Phi \,|\, \Delta \vdash t : G}\rho\\
-\llbracket \Gamma;\emptyset \,|\, \emptyset \vdash \map^{\ol{F},\ol{G}}_H
-  : \Nat^\emptyset (\Nat^{\ol\beta,\ol\gamma} F\,G)
-& = & \lambda d\, \ol\eta\,\ol{C}.\,
-\dsem{\Gamma; \ol{\phi},\ol{\gamma}\vdash H}\id_{\rho[\ol{\gamma:=
-      C}]}[\ol{\phi := \lambda \ol B. \eta_{\ol B\, \ol C}}]\\ 
-\hspace*{0.5in}
-  (\Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\,H[\ol{\phi
-      :=_{\ol{\beta}} G}]) \rrbracket^{\mathsf D} \rho & &\\
-%\hspace*{0.79in}  & & \hspace*{0.1in}\id_{\rho[\ol{\gamma :=
-%      C}]}[\ol{\phi := \lambda
-%    \ol{B}. (\dsem{\Gamma;\ol\alpha\,|\,\Delta \vdash t : 
-%    \Nat^{\ol{\beta},\ol{\gamma}}\,F\,G}\rho\,d)_{\ol{B}\,\ol{C}}}]\\
-\llbracket \Gamma;\emptyset \,|\, \emptyset \vdash \tin_H :
-Nat^{\ol{\beta},\ol{\gamma}} \, H[\phi := (\mu \phi.\lambda {\overline
-    \alpha}.H){\overline \beta}][\ol{\alpha := \beta}] & = &
-\lambda d\,\ol{B}\,
-\ol{C}.\,(\mathit{in}_{{T}^X_{H,\rho[\ol{\gamma :=
-        C}]}})_{\ol{B}} \;\;\mbox{ where } X \mbox{ is } \set \mbox{ when } \\ 
-\hspace*{0.79in}(\mu \phi.\lambda {\overline \alpha}.H){\overline
-  \beta} \rrbracket^{\mathsf{D}} \rho & & \hspace*{0.2in}  
-\mathsf{D} = \set \mbox{ and not present when }
-\mathsf{D} = \rel\\  
-\llbracket \Gamma;\emptyset \,|\, \emptyset \vdash
-  \fold^F_H : \Nat^\emptyset\;(\Nat^{\ol{\beta}, \ol{\gamma}}\,H[\phi
-    :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F) & = &  
-\lambda d\,\eta\,\ol{B}\,\ol{C}.\,
-(\mathit{fold}_{T^X_{H,\rho[\ol{\gamma := C}]}} \, (\lambda
-\ol{A}.\,\eta_{\ol{A}\,\ol{C}}))_{\ol{B}}\\ 
-\hspace*{0.79in}(\Nat^{{\ol{\beta},\ol{\gamma}} }\,(\mu
-  \phi.\lambda \overline \alpha.H)\overline \beta\;F)
-\rrbracket^{\mathsf{D}} \rho & & \hspace*{0.2in} \mbox{ where } X \mbox{ is as above
-}\vspace*{-0.1in} 
-\end{array}\]
-\caption{Term semantics}\label{fig:term-sem} 
-\end{minipage}}\vspace*{-0.05in}
-\end{figure*}
-
-\begin{comment}
-\begin{dfn}\label{def:rel-interp}
-If $\rho$ is a relation environment and $\Gamma;\Phi~|~\Delta \vdash t :
-\tau$ then $\relsem{\Gamma;\Phi~|~\Delta \vdash t : \tau}\rho$ is
-defined as follows:
-\[\begin{array}{lll}
-\relsem{\Gamma;\emptyset \,|\, \Delta,x :\tau \vdash x : \tau} \rho& = &
-\pi_{|\Delta|+1}\\
-%\relsem{\Gamma;\emptyset \,|\, \Delta \vdash \lambda x.t : \sigma \to \tau}\rho &
-%= & \curry (\relsem{\Gamma;\emptyset \,|\, \Delta, x : \sigma \vdash t :
-%  \tau}\rho)\\ 
-%\relsem{\Gamma;\emptyset \,|\, \Delta \vdash st: \tau} \rho & = & \eval
-%\circ \langle \relsem{\Gamma;\emptyset \,|\, \Delta \vdash s: \sigma \to
-%  \tau}\rho, \relsem{\Gamma;\emptyset \,|\, \Delta \vdash t: \sigma}\rho
-%\rangle\\
-\relsem{\Gamma;\emptyset \,|\, \Delta \vdash L_{\overline \alpha} x.t : \Nat^{\overline
-    \alpha} \,F \,G}\rho & = &  \curry (\relsem{\Gamma;\overline \alpha
-  \,|\, \Delta, x : F \vdash t: G}\rho[\overline{\alpha := \_}])\\
-\relsem{\Gamma;\Phi \,|\, \Delta \vdash t_{\overline{\tau}} s:
-  G [\overline{\alpha := \tau}]}\rho & = & \eval \circ \langle
-  \lambda e. \,(\relsem{\Gamma;\emptyset \,|\, \Delta \vdash t :
-  \Nat^{\overline{\alpha}} \,F \,G}\rho\; e)_{\overline{\relsem{\Gamma;\Phi
-      \vdash \tau}\rho}},\\ 
- & & \hspace*{0.5in} \relsem{\Gamma;\Phi \,|\,
-    \Delta \vdash s: F [\overline{\alpha := \tau}]}\rho \rangle\\ 
-& & \\
-%% \color{red} \mbox{Add rules for } \forall \mbox{ if we include it} & & \\
-\relsem{\Gamma;\Phi \,|\, \Delta,x :\tau \vdash x : \tau} \rho& = &
-\pi_{|\Delta|+1}\\
-\relsem{\Gamma;\Phi \,|\, \Delta \vdash \bot_\tau t : \tau} \rho& = &
-!^0_{\relsem{\Gamma;\Phi \vdash \tau}\rho} \circ
-  \relsem{\Gamma;\Phi~|~\Delta \vdash t : \zerot}\rho, \mbox{ where } \\
- & & \hspace*{0.1in} !^0_{\relsem{\Gamma;\Phi \vdash \tau}\rho}
-\mbox{ is the unique morphism from } 0\\
- & & \hspace*{0.1in} \mbox{ to } \relsem{\Gamma;\Phi \vdash \tau}\rho\\
-\relsem{\Gamma;\Phi \,|\, \Delta \vdash \top : \onet}\rho & = &
-!^{\relsem{\Gamma;\Phi\vdash \Delta}\rho}_1, \mbox{ where }
-!^{\relsem{\Gamma;\Phi\vdash \Delta}\rho}_1\\ 
-& & \hspace*{0.1in} \mbox{ is the unique morphism from }
-\relsem{\Gamma;\Phi\vdash \Delta}\rho \mbox{ to } 1\\ 
-\relsem{\Gamma;\Phi \,|\, \Delta \vdash (s,t) : \sigma \times \tau} \rho& = &
-\relsem{\Gamma;\Phi \,|\, \Delta \vdash s: \sigma} \rho\times
-\relsem{\Gamma;\Phi \,|\, \Delta \vdash t: \tau} \rho\\
-\relsem{\Gamma;\Phi \,|\, \Delta \vdash \pi_1 t : \sigma} \rho& = &
-\pi_1 \circ \relsem{\Gamma;\Phi \,|\, \Delta \vdash t : \sigma \times \tau}\rho\\
-\relsem{\Gamma;\Phi \,|\, \Delta \vdash \pi_2 t : \sigma}\rho & = &
-\pi_2 \circ \relsem{\Gamma;\Phi \,|\, \Delta \vdash t : \sigma \times
-  \tau} \rho\\
-\relsem{\Gamma;\Phi~|~\Delta \vdash \case{t}{x \mapsto l}{y \mapsto r} :
-  \gamma}\rho & = & \eval \circ \langle \curry \,[\relsem{\Gamma;\Phi
-    \,|\, \Delta, x : \sigma \vdash l : \gamma}\rho,\\
-   & & \hspace*{0.79in} \relsem{\Gamma;\Phi \,|\, \Delta, y
-    : \tau \vdash r : \gamma}\rho],\\
-   & &  \hspace*{0.5in} \relsem{\Gamma;\Phi \,|\, \Delta \vdash t :
-  \sigma + \tau} \rho\rangle \\
-\relsem{\Gamma;\Phi \,|\, \Delta \vdash \inl \,s: \sigma + \tau} \rho& = &
-\inl \circ \relsem{\Gamma;\Phi \,|\, \Delta \vdash s: \sigma}\rho\\
-\relsem{\Gamma;\Phi \,|\, \Delta \vdash \inr \,t: \sigma + \tau}\rho & = & 
-\inr \circ \relsem{\Gamma;\Phi \,|\, \Delta \vdash t : \tau}\rho\\
-\llbracket \Gamma;\emptyset \,|\, \emptyset \vdash \map^{\ol{F},\ol{G}}_H :
-\Nat^\emptyset\;(\ol{\Nat^{\ol{\beta},\ol{\gamma}}\,F\,G})\;&
-= & \lambda d\, \ol{\eta}\,\ol{R}.\,
-\relsem{\Gamma; \ol{\phi},\ol{\gamma}\vdash H}\id_{\rho[\ol{\gamma :=
-      R}]}[\ol{\phi := \lambda \ol{S}.\eta_{\ol{S}\,\ol{R}}}]\\
-\hspace*{0.79in} (\Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\,H[\ol{\phi
-      :=_{\ol{\beta}} G}]) \rrbracket^\rel \rho & & \\
-\llbracket \Gamma;\emptyset \,|\, \emptyset \vdash \tin_H :
-Nat^{\ol{\beta},\ol{\gamma}} \, H[\phi := (\mu \phi.\lambda {\overline
-    \alpha}.H){\overline \beta}][\ol{\alpha := \beta}] & = &
-\lambda d\,\ol{R}\, \ol{S}.\,(\mathit{in}_{T_{\rho[\ol{\gamma := S}]}})_{\ol{R}}\\
-\hspace*{0.79in}(\mu \phi.\lambda {\overline \alpha}.H){\overline
-  \beta} \rrbracket^\rel \rho & & \\
-\llbracket \Gamma;\emptyset \,|\, \emptyset \vdash
-  \fold^F_H : \Nat^\emptyset\;(\Nat^{\ol{\beta}, \ol{\gamma}}\,H[\phi
-    :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F) & = &  
-\lambda d\,\eta\,\ol{R}\,\ol{S}.\, (\mathit{fold}_{T_{\rho[\ol{\gamma
-        := S}]}} \, (\lambda
-\ol{Z}.\,\eta_{\ol{Z}\,\ol{S}}))_{\ol{R}}\\
-\hspace*{0.79in}(\Nat^{{\ol{\beta},\ol{\gamma}} }\,(\mu
-  \phi.\lambda \overline \alpha.H)\overline \beta\;F)
-\rrbracket^\rel \rho & & 
-\end{array}\]
-\end{dfn}
-{\color{blue} Add return type for fold in last clause? Should be
-  $\relsem{\Gamma;\ol\beta,\ol\gamma \vdash F}\rho[\ol{\gamma :=
-  C}]$.}
-
-This interpretation gives that $\relsem{\Gamma;\emptyset \,|\, \Delta
-  \vdash \lambda x.t : \sigma \to \tau}\rho = \curry
-(\relsem{\Gamma;\emptyset \,|\, \Delta, x : \sigma \vdash t :
-  \tau}\rho)$ and $\relsem{\Gamma;\emptyset \,|\, \Delta \vdash st:
-  \tau} \rho = \eval \circ \langle \relsem{\Gamma;\emptyset \,|\,
-  \Delta \vdash s: \sigma \to \tau}\rho, \relsem{\Gamma;\emptyset
-  \,|\, \Delta \vdash t: \sigma}\rho \rangle$, as expected.
-
-If $t$ is closed, i.e., if $\emptyset; \emptyset~|~\emptyset \vdash t
-: \tau$, then we write $\setsem{\vdash t : \tau}$ instead of
-$\setsem{\emptyset; \emptyset~|~\emptyset \vdash t : \tau}$.
-\end{comment}
-
-\begin{comment}
-  , and for any $\rho \in \relenv$, 
-\[ \relsem{\Gamma;\Phi \,|\, \Delta, x : \sigma \vdash t : \tau}\rho
-  = (\relsem{\Gamma;\Phi \,|\, \Delta \vdash t : \tau}\rho) \circ
-  \pi_{\Delta}\] where $\pi_{\Delta}$ is the projection
-  $\relsem{\Gamma;\Phi \vdash \Delta, x : \sigma} \to
-  \relsem{\Gamma;\Phi \vdash \Delta}$.
-\end{comment}
-
-The return type for the semantic fold is
-  $\dsem{\Gamma;\ol\beta,\ol\gamma \vdash F}\rho[\ol{\gamma :=
-    C}][\ol{\beta := B}]$.
-  This interpretation gives $\dsem{\Gamma;\emptyset \,|\, \Delta
-    \vdash \lambda x.t : F \to G}\rho = \curry
-  (\dsem{\Gamma;\emptyset \,|\, \Delta, x : F \vdash t :
-    G}\rho)$ and $\dsem{\Gamma;\emptyset \,|\, \Delta \vdash st:
-    G} \rho = \eval \circ \langle \dsem{\Gamma;\emptyset \,|\,
-    \Delta \vdash s: F \to G}\rho, \dsem{\Gamma;\emptyset
-    \,|\, \Delta \vdash t: F}\rho \rangle$, so it specializes to
-  the standard interpretations for System F terms.  If $t$ is closed,
-  i.e., if $\emptyset; \emptyset~|~\emptyset \vdash t : F$, then we
-  write $\dsem{\vdash t : F}$ instead of $\dsem{\emptyset;
-    \emptyset~|~\emptyset \vdash t : F}$.
-%, and similarly for $\relsem{\emptyset; \emptyset~|~\emptyset \vdash
-%t : \tau}$.
-Moreover, if
-  $\Gamma,\alpha;\Phi \,|\, \Delta \vdash t : F$ and
-  $\Gamma;\Phi,\alpha \,|\, \Delta \vdash t' : F$ and $\Gamma;\Phi
-  \vdash G$ then
-\begin{itemize}
-\item $\setsem{\Gamma;\Phi \,|\, \Delta[\alpha := G] \vdash
-  t[\alpha := G] : F[\alpha := G]}\rho =
-  \setsem{\Gamma,\alpha;\Phi \,|\, \Delta \vdash t : F }\rho [
-    \alpha := \setsem{\Gamma;\Phi\vdash G}\rho ]$
-%\item $\relsem{\Gamma;\Phi \,|\, \Delta[\alpha := \sigma] \vdash
-%  t[\alpha := \sigma] : \tau[\alpha := \sigma]}\rho =
-%  \relsem{\Gamma,\alpha;\Phi \,|\, \Delta \vdash t : \tau }\rho [
-%    \alpha := \relsem{\Gamma;\Phi\vdash\sigma}\rho ]$
-\item $\setsem{\Gamma;\Phi \,|\, \Delta[\alpha := G] \vdash
-  t'[\alpha := G] : F[\alpha := G]}\rho =
-  \setsem{\Gamma;\Phi,\alpha \,|\, \Delta \vdash t' : F }\rho [
-    \alpha := \setsem{\Gamma;\Phi\vdash G}\rho ]$
-%\item $\relsem{\Gamma;\Phi \,|\, \Delta[\alpha := \sigma] \vdash
-%  t'[\alpha := \sigma] : \tau[\alpha := \sigma]}\rho =
-%  \relsem{\Gamma;\Phi,\alpha \,|\, \Delta \vdash t' : \tau }\rho [
-%    \alpha := \relsem{\Gamma;\Phi\vdash\sigma}\rho ]$
-\end{itemize}
-and if $\Gamma;\Phi \,|\, \Delta, x: G \vdash t : F$ and
-$\Gamma;\Phi \,|\, \Delta \vdash s : G$ then
-\begin{itemize}
-\item $\lambda A.\, \setsem{\Gamma;\Phi \,|\, \Delta \vdash t[x := s]
-  : F }\rho \, A = \lambda A.\,\setsem{\Gamma;\Phi \,|\, \Delta, x:
-  G \vdash t : F}\rho \,(A, \setsem{\Gamma;\Phi \,|\,
-  \Delta\vdash s: G}\rho\, A)$
-%\item $\lambda R.\,\relsem{\Gamma;\Phi \,|\, \Delta \vdash t[x := s] :
-%  \tau }\rho\,R = \lambda R.\,\relsem{\Gamma;\Phi \,|\, \Delta, x:
-%  \sigma \vdash t : \tau}\rho \,(R, \relsem{\Gamma;\Phi \,|\,
-%  \Delta\vdash s: \sigma}\rho\,R)$
-\end{itemize}
-Direct calculation reveals that the set interpretations of terms also
-satisfy
-\begin{itemize}
-\item $\setsem{\Gamma; \Phi~|~\Delta \vdash
-  (L_{\ol{\alpha}}x.t)_{\ol{K}}s} = \setsem{\Gamma; \Phi~|~\Delta
-  \vdash t [\ol{\alpha := K}][x := s]}$
-\end{itemize}
-Term extensionality with respect to types and terms --- i.e.,
-$\setsem{\Gamma;\Phi\vdash (L_\alpha x.t)_\alpha \top : F} =
-\setsem{\Gamma;\Phi \vdash t : F}$ and $\setsem{\Gamma;\Phi\vdash
-  (L_\alpha x.t)_\alpha x : F} = \setsem{\Gamma;\Phi \vdash t : F}$
-%$\relsem{\Gamma;\Phi\vdash (L_\alpha x.t)_\alpha t} =
-%\relsem{\Gamma;\Phi \vdash t}$, as well as
-%term extensionality
-%and $\relsem{\Gamma;\Phi\vdash
-%  (L_\alpha x.t)_\alpha \top} = \relsem{\Gamma;\Phi \vdash t}$
---- follow (when both sides of the equations are defined). Similar
-properties hold for relational interpretations.
-
-\subsection{Properties of Terms of $\Nat$-Type}\label{sec:Nat-type-terms}
-
-Define, for $\Gamma; \ol{\alpha} \vdash F$, the term $\id_F$ to be
-$\Gamma;\emptyset~|~\emptyset \vdash L_{\ol{\alpha}}x.x :
-\Nat^{\ol{\alpha}} F\,F$ and, for terms $\Gamma; \emptyset \,|\,
-\Delta \vdash t: \Nat^{\overline{\alpha}} F\,G$ and $\Gamma; \emptyset
-\,|\, \Delta \vdash s: \Nat^{\overline{\alpha}} G\,H$, the {\em
-  composition} $s \circ t$ of $t$ and $s$ to be $\Gamma;
-\emptyset\,|\, \Delta \vdash L_{\overline{\alpha}}
-x. s_{\overline{\alpha}}(t_{\overline{\alpha}}x):
-\Nat^{\overline{\alpha}} F\,H$. Then $\setsem{\Gamma; \emptyset \,|\,
-  \emptyset \vdash \id_{F} : \Nat^{\ol{\alpha}} F\,F} \rho\, \ast =
-\id_{\lambda \ol{A}. \setsem{\Gamma; \ol{\alpha} \vdash F} \rho
-  [\ol{\alpha := A}]}$ for any set environment $\rho$, and
-$\setsem{\Gamma; \emptyset \,|\, \Delta \vdash s \circ t:
-  \Nat^{\overline{\alpha}} F\,H} = \setsem{\Gamma; \emptyset \,|\,
-  \Delta \vdash s: \Nat^{\overline{\alpha}} G\,H} \circ
-\setsem{\Gamma; \emptyset \,|\, \Delta \vdash t:
-  \Nat^{\overline{\alpha}} F\,G}$. Straightforward calculation shows
-that terms of $\Nat$ type behave as natural transformations with
-respect to their source and target functorial types, i.e., we have
-\begin{thm}\label{eq:ft-from-nat} 
-  If\, $\Gamma; \emptyset \,|\, \Delta \vdash s : \Nat^{\overline{\alpha},
-  \overline{\gamma}} F\,G$ and $\overline{\Gamma; \emptyset \,|\,
-  \Delta \vdash t : \Nat^{\overline{\gamma}} K\, H}$, then
-  \[\begin{array}{l}
-  \hspace*{0.135in}
-\setsem{\Gamma; \emptyset\,|\, \Delta
-  \vdash 
-  ((\map_G^{\overline{K}, \overline{H}})_\emptyset \,\overline{t}) \circ
-(L_{\overline{\gamma}} z. s_{\overline{K}, \overline{\gamma}}
-  z)
-  : \Nat^{\overline{\gamma}} F[\overline{\alpha := K}]\,
-  G[\overline{\alpha := H}]}\\
-= \setsem{ \Gamma; \emptyset \,|\, \Delta \vdash
-(L_{\overline{\gamma}} z.
-  s_{\overline{H}, \overline{\gamma}} z)\circ
-  ((\map_F^{\overline{K}, \overline{H}})_\emptyset \,
-  \overline{t})  : \Nat^{\overline{\gamma}} F[\overline{\alpha :=
-      K}]\, G[\overline{\alpha := H}]}
-\end{array}\]
-\end{thm}
-\noindent
-When $s = \mathsf{in}_H$, and
-$\overline{\Gamma; \emptyset \,|\, t :
- \Nat^{\overline{\gamma}} F\, G}$,
-and $\xi = \Nat^{\overline{\gamma}} H[\phi :=
-  (\mu \phi. \lambda \overline{\alpha}. H)
-  \overline{\beta}][\overline{\alpha := F}]\, (\mu \phi. \lambda
-\overline{\alpha}. H)\overline{G}$, Theorem~\ref{eq:ft-from-nat}
-specializes to
-\begin{equation}\label{thm:subst}
-\begin{array}{l}
-\hspace*{0.135in}\setsem{\Gamma; \emptyset \,|\, \Delta \vdash
-((\map_{(\mu \phi. \lambda
-    \overline{\alpha}. H) \overline{\beta}}^{\overline{F},
-    \overline{G}})_\emptyset\, \overline{t}) \circ (L_{\overline{\gamma}} z.
-  (\tin_{H})_{\overline{F},
-    \overline{\gamma}} z) : \xi}\\
-= \setsem{\Gamma; \emptyset \,|\, \Delta
-\vdash (L_{\overline{\gamma}} z. (\tin_H)_{\overline{G},
-  \overline{\gamma}} z) \circ
-  ((\map_{H[\phi := (\mu \phi. \lambda \overline{\alpha}. H)
-      \overline{\beta}]}^{\overline{F},
-    \overline{G}})_\emptyset\, \overline{t})
-  : \xi}
-\end{array}
-\end{equation}
-\noindent
-Theorem~\ref{eq:ft-from-nat} gives a family of free theorems that are
-consequences of naturality, and thus do not require the full power of
-parametricity. Other specific instances include the three
-$\mathtt{map}$/$\mathtt{reverse}$ theorems from
-Section~\ref{sec:intro}. The $\mathtt{map}$/$\mathtt{reverse}$ theorem
-for bushes, e.g., is
-\begin{align*}
-  &\setsem{\emptyset; \emptyset~|~ y : \Nat^\emptyset F \, G
-    \vdash ((\map_{\mathit{Bush}\,\alpha}^{{F},
-      {G}}P)_\emptyset\, y) \circ (L_{{\emptyset}}
-    z. \mathit{reverseBush}_{F}\, z) : \Nat^{{\emptyset}}
-    (\mathit{Bush} \, F) \, (\mathit{Bush} \, G) }\\ =\;
-  &\setsem{\emptyset; \emptyset~|~ y : \Nat^\emptyset F\, G
-    \vdash (L_{\emptyset} z. \mathit{reverseBush}_{G}\,z) \circ
-    ((\map_{\mathit{Bush} \, \alpha}^{F, G})_\emptyset\,
-           {y}) : \Nat^{\emptyset} (\mathit{Bush} \, F ) \,
-           (\mathit{Bush} \, G)}
-\end{align*}
-for any term $\vdash \mathit{reverseBush} : \Nat^\alpha (\mathit{Bush}
-\,\alpha) \, (\mathit{Bush} \, \alpha)$. We can also prove a theorem
-for the data type $\emptyset;\alpha,\gamma \vdash
-\mathit{Tree}\,\alpha\,\gamma$ from Section~\ref{sec:calculus} that is
-similar but only maps along some of its functorial variables:
-%  \begin{align*}
-%  &\setsem{\emptyset; \alpha~|~ y : \Nat^\emptyset \sigma \, \tau
-%    \vdash ((\map_{\mathit{PTree}\,\alpha}^{{\sigma},
-%      {\tau}})_{\emptyset} {y}) \circ (L_{{\emptyset}}
-%    z. \mathit{reverse}_{\sigma} z) : \Nat^{{\emptyset}}
-%    (\mathit{PTree} \, \sigma) \, (\mathit{PTree} \, \tau) }\\ =\;
-%  &\setsem{\emptyset; \alpha~|~ y : \Nat^\emptyset \sigma \, \tau
-%    \vdash (L_{\emptyset} z. \mathit{reverse}_{\tau} z) \circ
-%    ((\map_{\mathit{PTree} \, \alpha}^{{\sigma}, {\tau}})_{\emptyset} {y}) :
-%    \Nat^{\emptyset} (\mathit{PTree} \, \sigma ) \, (\mathit{PTree} \,
-%    \tau)}
-%\end{align*}
-%for any term $\vdash \mathit{reverse} : \Nat^\alpha (\mathit{PTree}
-%\,\alpha) \, (\mathit{PTree} \, \alpha)$,
-if $\vdash \mathit{reverseTree} : \Nat^{\alpha, \gamma} (\mathit{Tree}
-\,\alpha \, \gamma) \, (\mathit{Tree} \, \alpha \, \gamma)$ then
-\begin{align*}
-  &\setsem{\emptyset; \emptyset~|~y : \Nat^\gamma F \, G 
-  \vdash  ((\map_{\mathit{Tree}\,\alpha \, \gamma}^{F,
-      G})_\emptyset\, y) \circ (L_{{\gamma}} z. \mathit{reverseTree}_{F,
-      \gamma}\, z) : \Nat^{{\gamma}} 
-    (\mathit{Tree} \, F \, \gamma) \, (\mathit{Tree} \, G \,
-    \gamma) }\\ =\; &\setsem{\emptyset; \emptyset~|~y :
-    \Nat^\gamma F \, G \vdash (L_{\gamma}
-    z. \mathit{reverseTree}_{G, \gamma}\,z) \circ
-    ((\map_{\mathit{Tree} \, \alpha\, \gamma}^{{F},
-      {G}})_\emptyset\, y) : \Nat^{\gamma} (\mathit{Tree} \,
-    F \, \gamma) \, (\mathit{Tree} \, G\, \gamma)}
-\end{align*}
-\noindent
-%Note this instance uses the following version of map:
-%$$\map_{\mathit{Tree} \,\alpha \, \gamma}^{\sigma, \tau} : 
-%\Nat^{\emptyset} (\Nat^\gamma \, \sigma \,\tau)\,
-%(\Nat^\gamma(\mathit{Tree}\,\sigma\,\gamma)\,(\mathit{Tree}\,\tau\,\gamma))$$
-%as opposed to 
-%$$\map_{\mathit{Tree} \,\alpha \, \gamma}^{\sigma, \sigma', \tau, \tau'} : 
-%\Nat^{\emptyset} ((\Nat^\emptyset \, \sigma \,\tau) \times
-%(\Nat^\emptyset\,\sigma'\,\tau')) \,
-%(\Nat^\emptyset\,(\mathit{Tree}\,\sigma\,\sigma')\,(\mathit{Tree}\,\tau\,\tau'))$$
-We observed in Section~\ref{sec:calculus} that $ \phi; \alpha \vdash
-\mathit{GRose}\,\phi\,\alpha$ cannot appear as the (co)domain of a
-$\Nat$-type binding both $\phi$ and $\alpha$. As a result, there are
-no instances of, say, a $\mathtt{map}$/$\mathtt{reverse}$ free theorem
-for this data type that maps replacements for both $\phi$ and
-$\alpha$. On the other hand, $\phi; \alpha \vdash
-\mathit{GRose}\,\phi\,\alpha$ does support a
-$\mathtt{map}$/$\mathtt{reverse}$ theorem that maps only a replacement
-for $\alpha$.
-
-As usual, analogous results hold for relation environments and
-relational interpretations.
-
-\subsection{Properties of Initial Algebraic Constructs}\label{sec:iaps}
-
-Our semantics interprets applications of $\map$ as applications of
-semantic $\mathit{map}$ functions.
-\begin{comment}
-Indeed, if \(\Gamma; \ol{\phi},
-\ol{\gamma} \vdash H\), \(\ol{\Gamma; \emptyset, \ol{\gamma} \vdash
-  F}\), and \(\ol{\Gamma; \Phi, \emptyset, \ol{\gamma} \vdash G}\),
-%and $\Gamma;\Phi\,|\,\Delta \vdash t :
-%\Nat^{\ol{\beta},\ol{\gamma}}\,F\,G$
-then 
-Definition~\ref{def:set-interp} gives
-%\begin{multline}\label{eq:map-sem-def}
-\begin{align}
-& \setsem{
-\Gamma;\Phi \,|\, \Delta
-\vdash \map^{\ol{F},\ol{G}}_H \,\ol t
-: %\Nat^\emptyset\;(\ol{\Nat^{\ol{\beta},\ol{\gamma}}\,F\,G})\; {}
-\Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\,H[\ol{\phi :=_{\ol{\beta}} G}]
-} \rho \nonumber \\ 
-=\;\; & \lambda d\, \ol{\eta}\,\ol{B}.\,
-\setsem{\Gamma; \ol{\phi},\ol{\gamma}\vdash H}
-\id_{\rho[\ol{\gamma := B}]}[\ol{\phi := \lambda
-    \ol{A}.\eta_{\ol{A}\,\ol{B}}}] \label{eq:map-sem-def}
-\end{align}
-%\end{multline}
-\end{comment}
-In particular, if $\Gamma; \ol{\alpha} \vdash H$, $\ol{\Gamma;
-  \emptyset \vdash F}$, and $\ol{\Gamma; \emptyset \vdash G}$, then
-Definition~\ref{def:set-interp} gives, for all $\ol{f :
-  \setsem{\Gamma;\emptyset \vdash \Nat^\emptyset F\,G}\rho}$, that
-\[\begin{array}{rl}
-&\setsem{ \Gamma;\emptyset\,|\,\ol{x : \Nat^\emptyset F \, G}
-  \vdash (\map^{\ol{F},\ol{G}}_H)_\emptyset\,\ol x :
-  \Nat^{\emptyset}\,H[\ol{\alpha := F}]\,H[\ol{\alpha := G}]}
-\rho\,\ol f\\ =& \setsem{\Gamma; \ol{\alpha}\vdash H}
-\id_{\rho}[\ol{\alpha := f}] \\ =&
-%\lambda \ol{f : \setsem{\Gamma; \emptyset \vdash \sigma}\rho \to \setsem{\Gamma%; \emptyset \vdash \tau}\rho}.\,
-\mathit{map}_{\lambda \ol{A}.\,\setsem{\Gamma; \ol{\alpha}\vdash H} \rho [\ol{\alpha := A}]} \ol{f} \\
-%=& \mathit{map}_{\lambda \ol{A}.\,\setsem{\Gamma; \ol{\alpha}\vdash H} \rho [\ol{\alpha := A}]}
-\end{array}
-\]
-\begin{comment}
-\[\begin{array}{rl}
-&\setsem{
-\Gamma;\emptyset \,|\, \emptyset
-\vdash \map^{\ol{\sigma},\ol{\tau}}_H
-: \Nat^\emptyset\;(\ol{\Nat^{\emptyset}\,\sigma\,\tau})\; {}
-(\Nat^{\emptyset}\,H[\ol{\alpha := \sigma}]\,H[\ol{\alpha := \tau}])
-} \rho\, \ast \\
-=& \lambda \ol{f : \setsem{\Gamma; \emptyset \vdash \sigma}\rho \to \setsem{\Gamma; \emptyset \vdash \tau}\rho}.\,
-\setsem{\Gamma; \ol{\alpha}\vdash H}
-\id_{\rho}[\ol{\alpha := f}] \\
-=& \lambda \ol{f : \setsem{\Gamma; \emptyset \vdash \sigma}\rho \to \setsem{\Gamma; \emptyset \vdash \tau}\rho}.\,
-\mathit{map}_{\lambda \ol{A}.\,\setsem{\Gamma; \ol{\alpha}\vdash H} \rho [\ol{\alpha := A}]} \ol{f} \\
-=& \mathit{map}_{\lambda \ol{A}.\,\setsem{\Gamma; \ol{\alpha}\vdash H} \rho [\ol{\alpha := A}]}
-\end{array}
-\]
-\end{comment}
-
-\vspace*{0.05in}
-
-\noindent
-Here, we obtain the first equality from the appropriate instance of
-Definition~\ref{def:set-interp}, and the second one by noting that
-$\lambda \ol{A}.\,\setsem{\Gamma; \ol{\alpha}\vdash H} \rho
-[\ol{\alpha := A}]$ is a functor in $\alpha$ and using
-$\mathit{map}_G\,\ol f$ to denote the action of the semantic functor
-$G$ on morphisms $\ol f$.
-
-\vspace*{0.1in}
-
-We have the expected properties of the interpretations of $\map$,
-$\tin$, and $\fold$, namely:
-
-\vspace*{0.1in}
-
-\noindent
-$\bullet$\;  If
-$\Gamma; \ol{\psi}, \ol{\gamma} \vdash H$,\;
-$\ol{\Gamma; \ol{\alpha}, \ol{\gamma}, \ol{\phi} \vdash K}$,\;
-$\ol{\Gamma; \ol{\beta}, \ol{\gamma} \vdash F}$,\; and
-$\ol{\Gamma; \ol{\beta}, \ol{\gamma} \vdash G}$,
-then
-\[\setsem{\Gamma; \emptyset \,|\, \emptyset \vdash
-\map_{H[\ol{\psi := K}]}^{\ol{F}, \ol{G}} : \xi} = \setsem{\Gamma;
-  \emptyset \,|\, \emptyset \vdash L_\emptyset \ol
-  z. (\map_H^{\ol{K[\ol{\phi := F}]}, 
-    \ol{K[\ol{\phi := G}]}})_\emptyset \,(\ol{(\map_K^{\ol{F}, \ol{G}})_\emptyset \, \ol z}) :
-  \xi}\]
-for $\xi = \Nat^{\emptyset} (\ol{\Nat^{\ol{\alpha}, \ol{\beta},
-    \ol{\gamma}} F\, G})\, (\Nat^{\ol{\gamma}} H[\ol{\psi :=
-    K}][\ol{\phi := F}]\, H[\ol{\psi := K}][\ol{\phi := G}])$
-
-\vspace*{0.18in}
-
-\noindent
-$\bullet$\; If $\Gamma; \overline{\beta}, \overline{\gamma} \vdash
-H$,\; $\Gamma; \overline{\beta}, \overline{\gamma} \vdash K$,\;
-$\overline{\Gamma; \overline{\alpha}, \overline{\gamma} \vdash F}$,\;
-$\overline{\Gamma; \emptyset, \overline{\alpha}, \overline{\gamma}
-  \vdash G}$,\; 
-$\overline{\Gamma; \phi, \overline{\psi}, \overline{\gamma} \vdash
-  M}$,\; $\overline{I}$ is the sequence $\overline{F}, H$,\;
-$\overline{J}$ is the sequence $\overline{G}, K$, and $\Gamma;
-\emptyset\,|\,\Delta \vdash s : \Nat^{\overline{\beta},
-  \overline{\gamma}} H\, K$ and $\ol{\Gamma; \emptyset \,|\,\Delta \vdash
-  t : \Nat^{\overline{\alpha}, \overline{\gamma}} F\, G}$, then
-\[\hspace*{0.13in}\setsem{\Gamma; \emptyset \,|\, \Delta
-    \vdash L_{\overline{\gamma}} z.\,
-    s_{\overline{M [\overline{\psi := G}] [\phi := K]},
-      \overline{\gamma}} \Big(\big( (\map_{H}^{\overline{M
-        [\overline{\psi := F}] [\phi := H]}, \overline{M
-        [\overline{\psi := G}] [\phi := K]}})_\emptyset
-    (\overline{(\map_{M}^{\overline{I}, \overline{J}})_\emptyset
-      (s, \overline{t})}) \big)_{\overline{\gamma}} z \Big) : \xi}\]
-\[= \setsem{\Gamma; \emptyset\,|\, \Delta \vdash
-    L_{\overline{\gamma}} z.\, \big( (\map_{K}^{\overline{M
-      [\overline{\psi := F}] [\phi := H]}, \overline{M [\overline{\psi
-          := G}] [\phi := K]}})_\emptyset (\overline{(\map_{M}^{\overline{I},
-    \overline{J}})_\emptyset (s, \overline{t})}) \big)_{\overline{\gamma}} \Big(
-s_{\overline{M [\overline{\psi := F}] [\phi := H]}, \overline{\gamma}}
-z \Big) : \xi}\] for $\xi = \Nat^{\overline{\gamma}}\,
-  H[\overline{\beta:= M}] [\overline{\psi := F}] [\phi := H]\;
-  K[\overline{\beta := M}] [\overline{\psi := G}] [\phi := K]$
-
-\vspace*{0.18in}
-
-\noindent
-$\bullet$\; If $\xi = \Nat^{\overline{\beta}, \overline{\gamma}}\,
-H[\phi := (\mu \phi. \lambda
-  \overline{\alpha}. H)\overline{\beta}][\overline{\alpha := \beta}]\;
-F$, then
-\[\hspace*{-0.72in}\setsem{\Gamma; \emptyset \,|\, x:
-  \Nat^{\overline{\beta}, \overline{\gamma}} H[\phi :=
-    F][\overline{\alpha := \beta}]\, F \vdash ((\fold_{H,
-    F})_{\emptyset} x) \circ \tin_{H} : \xi}\]
-
-\vspace*{-0.15in}
-
-\[= \setsem{ \Gamma; \emptyset \,|\, x: \Nat^{\overline{\beta},
-    \overline{\gamma}} H[\phi := F][\overline{\alpha := \beta}]\, F
-  \vdash x \circ \big( (\map_{H [\ol{\alpha := \beta}]}^{(\mu
-    \phi. \lambda \overline{\alpha} H) \overline{\beta},
-    F})_\emptyset ((\fold_{H, F})_{\emptyset} x) \big) : \xi}\]
-
-\vspace*{0.1in}
-
-\noindent
-$\bullet$\; If $\xi = \Nat^{\overline{\beta}, \overline{\gamma}}\, (\mu
-\phi. \lambda \overline{\alpha}. H)\overline{\beta}\; (\mu
-\phi. \lambda \overline{\alpha}. H)\overline{\beta}$, then
-\[\setsem{\Gamma; \emptyset \,|\, \emptyset
-  \vdash \tin_H \circ (\fold_{H, H[\phi := (\mu \phi. \lambda
-      \overline{\alpha}. H)\overline{\beta}]})_{\emptyset}
-  ((\map_{H}^{H[\phi := (\mu \phi. \lambda
-      \overline{\alpha}. H)\overline{\beta}][\overline{\alpha :=
-        \beta}], (\mu \phi. \lambda
-    \overline{\alpha}. H)\overline{\beta}})_\emptyset \tin_H) :
-  \xi}\]
-
-\vspace*{-0.15in}
-
-\[\hspace*{-3.46in} = \setsem{\Gamma; \emptyset \,|\, \emptyset \vdash \mathit{id}_{(\mu
-    \phi. \lambda \overline{\alpha}. H)\overline{\beta}} : \xi}\]
-
-\vspace*{0.15in}
-
-\noindent
-$\bullet$\; If $\xi = \Nat^{\overline{\beta}, \overline{\gamma}}
-H[\phi := (\mu \phi. \lambda \overline{\alpha}. H)\overline{\beta}]\,
-H[\phi := (\mu \phi. \lambda \overline{\alpha}. H)\overline{\beta}]$,
-then
-\[\setsem{\Gamma; \emptyset \,|\, \emptyset \vdash
-(\fold_{H, H[\phi := (\mu \phi. \lambda
-      \overline{\alpha}. H)\overline{\beta}]})_{\emptyset}
-  ((\map_{H}^{H[\phi := (\mu \phi. \lambda
-      \overline{\alpha}. H)\overline{\beta}][\overline{\alpha :=
-        \beta}], (\mu \phi. \lambda
-    \overline{\alpha}. H)\overline{\beta}})_\emptyset \tin_H) \circ
-  \tin_H : \xi}\]
-
-\vspace*{-0.15in}
-
-\[\hspace*{-3.15in} =\setsem{ \Gamma; \emptyset \,|\, \emptyset \vdash
-  \mathit{id}_{H[\phi := (\mu \phi. \lambda
-      \overline{\alpha}. H)\overline{\beta}]} : \xi}\]
-
-\vspace*{0.1in}
-
-\noindent
-Analogous results hold for relational interpretations of terms and
-relational environments.
-
-\begin{comment}
-\subsection{Free Theorems Derived from Naturality}\label{sec:ft-nat}
-
-In our setting, free theorems of a certain familiar form are derivable
-directly from the semantic properties of $\Nat$-types. Indeed, the
-following theorem is immediate from Equation~\ref{thm:subst}, by taking
-$\gamma$ to be empty, expanding the interpretations on the left- and
-right-hand sides, and using the fact from Section~\ref{sec:iaps} that
-$\map$ terms are interpreted as semantic $\mathit{map}$ functions:
-\begin{thm}\label{thm:ft-nat}
-If\, $\Gamma; \ol\alpha \vdash F : \F$,\, $\Gamma; \ol\alpha \vdash G :
-\F$, $\rho : \setenv$, $x \in \setsem{\Gamma; \emptyset \vdash
-  \Nat^{\ol\alpha} F\, G } \rho$, and, for vectors of length
-$|\ol\alpha|$, \, $\ol{\Gamma; \emptyset \vdash \sigma : \F}$,\,
-$\ol{\Gamma; \emptyset \vdash \tau: \F}$, and $\ol{y \in
-  \setsem{\Gamma; \emptyset \vdash \Nat^{\emptyset} \sigma\, \tau }
-  \rho}$, then
-\[\mathit{map}_{\setsem{\Gamma; \ol\alpha \vdash G} \rho[\ol{\alpha := \_}]} \ol{y}
-  \circ x_{\ol{{\setsem{\Gamma; \emptyset \vdash \sigma} \rho}}} =
-  x_{\ol{\setsem{\Gamma; \emptyset \vdash \tau} \rho} } \circ
-  \mathit{map}_{\setsem{\Gamma; \ol\alpha \vdash F} \rho[\ol{\alpha :=
-        \_}]} \, \ol{y}\]
-\end{thm}
-\noindent
-%INCORRECT
-%In particular, for $\ol{A, B: \set}$, variables $\ol{v}$ and $\ol{w}$
-%with $\ol{\rho v = A}$ and $\ol{\rho w = B}$, and functions $\ol{f : A
-%  \to B}$, the following holds: $\mathit{map}_{\setsem{\Gamma; \ol\alpha \vdash G}
-%  \rho[\ol{\alpha := \_}]} \ol{f} \circ x_{\ol{A}} = x_{\ol{B}} \circ
-%\mathit{map}_{\setsem{\Gamma; \ol\alpha \vdash F} \rho[\ol{\alpha :=
-%      \_}]} \, \ol{f}$.
-A similar result holds for relational interpretations.
-
-Specific instances of Theorem~\ref{thm:ft-nat} include {\color{blue}
-  Wadler free theorems, map-reverse theorems from intro, ???}
-\end{comment}
-
-\begin{comment}
-We already know from Equation~\ref{thm:subst} that 
-\begin{multline}
-\setsem{\Gamma; \emptyset \,|\, x :
-  \Nat^{\overline{\alpha}, 
-    \overline{\gamma}} F\,G, \overline{y : \Nat^{\overline{\gamma}}
-    \sigma\, \tau} \vdash ((\map_G^{\overline{\sigma},
-    \overline{\tau}})_{\emptyset} \overline{y}) \circ
-  (L_{\overline{\gamma}} z. x_{\overline{\sigma}, \overline{\gamma}}
-  z) : \Nat^{\overline{\gamma}} F[\overline{\alpha := \sigma}]\,
-  G[\overline{\alpha := \tau}]} \\
-= \setsem{ \Gamma; \emptyset \,|\, x
-  : \Nat^{\overline{\alpha}, \overline{\gamma}} F\, G, \overline{y :
-    \Nat^{\overline{\gamma}} \sigma\, \tau} \vdash
-  (L_{\overline{\gamma}} z. x_{\overline{\tau}, \overline{\gamma}} z)
-  \circ ((\map_F^{\overline{\sigma}, \overline{\tau}})_{\emptyset}
-  \overline{y}) : \Nat^{\overline{\gamma}} F[\overline{\alpha :=
-      \sigma}]\, G[\overline{\alpha := \tau}]}
-\end{multline}
-
-In particular, for any set environment $\rho$, natural transformation 
-$x \in \setsem{\Gamma; \emptyset \vdash \Nat^{\ol\alpha} F\, G } \rho$, and functions
-$\ol{y \in \setsem{\Gamma; \emptyset \vdash \Nat^{\emptyset} \sigma\, \tau } \rho}$
-(thus specializing to the case where there are no $\gamma$’s), we have
-\begin{multline}
-\setsem{\Gamma; \emptyset \,|\, x :
-  \Nat^{\overline{\alpha}} F\,G, \overline{y : \Nat^{\emptyset}
-    \sigma\, \tau} \vdash ((\map_G^{\overline{\sigma},
-    \overline{\tau}})_{\emptyset} \overline{y}) \circ
-  (L_{\emptyset} z. x_{\overline{\sigma}}
-  z) : \Nat^{\emptyset} F[\overline{\alpha := \sigma}]\,
-  G[\overline{\alpha := \tau}]} \rho (x, \ol{y}) \\
-= \setsem{ \Gamma; \emptyset \,|\, x
-  : \Nat^{\overline{\alpha}} F\, G, \overline{y :
-    \Nat^{\emptyset} \sigma\, \tau} \vdash
-  (L_{\emptyset} z. x_{\overline{\tau}} z)
-  \circ ((\map_F^{\overline{\sigma}, \overline{\tau}})_{\emptyset}
-  \overline{y}) : \Nat^{\emptyset} F[\overline{\alpha :=
-      \sigma}]\, G[\overline{\alpha := \tau}]} \rho (x, \ol{y})
-\end{multline}
-
-Unwinding these interpretations yields
-\begin{multline}
-\setsem{\Gamma; \emptyset \,|\, x :
-  \Nat^{\overline{\alpha}} F\,G, \overline{y : \Nat^{\emptyset}
-    \sigma\, \tau} \vdash ((\map_G^{\overline{\sigma},
-    \overline{\tau}})_{\emptyset} \overline{y}) \circ
-  (L_\emptyset z. x_{\overline{\sigma}}
-  z) : \Nat^{\emptyset} F[\overline{\alpha := \sigma}]\,
-  G[\overline{\alpha := \tau}]} \rho \, (x , \ol{y}) \\
-= 
-\setsem{\Gamma; \emptyset \,|\, x : \Nat^{\overline{\alpha}} F\,G, 
-  \overline{y : \Nat^{\emptyset}
-    \sigma\, \tau} \vdash (\map_G^{\overline{\sigma},
-    \overline{\tau}})_{\emptyset} \overline{y} } \rho \, (x, \ol{y})
-   \circ
-  \setsem{\Gamma; \emptyset \,|\, x : \Nat^{\overline{\alpha}} F\,G 
-   , \overline{y : \Nat^{\emptyset}
-      \sigma\, \tau} \vdash
-   L_{\emptyset} z. x_{\overline{\sigma}}
-  z } \rho \, (x, \ol{y}) \\
-= 
-\setsem{\Gamma; \emptyset \,|\, \overline{y : \Nat^{\emptyset}
-    \sigma\, \tau} \vdash (\map_G^{\overline{\sigma},
-    \overline{\tau}})_{\emptyset} \overline{y} } \rho \, \ol{y}
-   \circ
-  \setsem{\Gamma; \emptyset \,|\, x : \Nat^{\overline{\alpha}} F\,G \vdash
-    L_{\emptyset} z. x_{\overline{\sigma}}
-  z } \rho \, x \\
-= \map_{\setsem{\Gamma; \ol\alpha \vdash G} \rho[\ol{\alpha := \_}]} \ol{y}
-  \circ 
-  x_{\ol{{\setsem{\Gamma; \emptyset \vdash \sigma} \rho}}}
-\end{multline}
-
-and 
-\begin{multline}
-\setsem{ \Gamma; \emptyset \,|\, x
-  : \Nat^{\overline{\alpha}} F\, G, \overline{y :
-    \Nat^{\emptyset} \sigma\, \tau} \vdash
-  (L_{\emptyset} z. x_{\overline{\tau}} z)
-  \circ ((\map_F^{\overline{\sigma}, \overline{\tau}})_{\emptyset}
-  \overline{y}) : \Nat^{\emptyset} F[\overline{\alpha :=
-      \sigma}]\, G[\overline{\alpha := \tau}]} \rho (x, \ol{y}) \\
-=
-\setsem{ \Gamma; \emptyset \,|\, x
-  : \Nat^{\overline{\alpha}} F\, G, \overline{y :
-    \Nat^{\emptyset} \sigma\, \tau} \vdash
-  L_{\emptyset} z. x_{\overline{\tau}} z)} \rho (x, \ol{y})
-  \circ 
-  \setsem{\Gamma; \emptyset
-  \,|\, x : \Nat^{\overline{\alpha}} F\, G, \overline{y :
-    \Nat^{\emptyset} \sigma\, \tau} \vdash 
-  (\map_F^{\overline{\sigma}, \overline{\tau}})_{\emptyset}
-  \overline{y} } \rho (x, \ol{y}) \\
-= \setsem{\Gamma; \emptyset \,|\, x
-  : \Nat^{\overline{\alpha}} F\, G \vdash 
-  L_{\emptyset} z. x_{\overline{\tau}} z} \rho \, x
-  \circ 
-  \setsem{\Gamma;\emptyset \,|\, \overline{y :
-    \Nat^{\emptyset} \sigma\, \tau} \vdash
-    (\map_F^{\overline{\sigma}, \overline{\tau}})_{\emptyset}
-  \overline{y} } \rho \, \ol{y}  \\
-  = x_{\ol{\setsem{\Gamma; \emptyset \vdash \tau} \rho} }
-  \circ \map_{\setsem{\Gamma; \ol\alpha \vdash F} 
-    \rho[\ol{\alpha := \_}]} \, \ol{y}
-\end{multline}
-
-So, we can conclude that
-\begin{equation}
-\map_{\setsem{\Gamma; \ol\alpha \vdash G} \rho[\ol{\alpha := \_}]} \ol{y}
-  \circ 
-  x_{\ol{{\setsem{\Gamma; \emptyset \vdash \sigma} \rho}}}
-= x_{\ol{\setsem{\Gamma; \emptyset \vdash \tau} \rho} }
-\circ \map_{\setsem{\Gamma; \ol\alpha \vdash F} 
-  \rho[\ol{\alpha := \_}]} \, \ol{y}
-\end{equation}
-
-Moreover, for any $\ol{A, B: \set}$,
-we can choose $\ol{\sigma = v}$ and $\ol{\tau = w}$ to be variables such that
-$\ol{\rho v = A}$ and $\ol{\rho w = B}$.
-Then for any functions $\ol{f : A \to B}$ we have that
-\begin{equation}
-\map_{\setsem{\Gamma; \ol\alpha \vdash G} \rho[\ol{\alpha := \_}]} \ol{f}
-  \circ 
-  x_{\ol{A}}
-  = x_{\ol{B}}
-\circ \map_{\setsem{\Gamma; \ol\alpha \vdash F} 
-  \rho[\ol{\alpha := \_}]} \, \ol{f}
-\end{equation}
-\end{proof}
-\end{comment}
-
-\subsection{The Abstraction Theorem}\label{sec:thms}
-
-To go beyond naturality and get {\em all} consequences of
-parametricity, we prove an Abstraction Theorem for our calculus. As
-usual for such theorems, we prove a more general result in
-Theorem~\ref{thm:at-gen} that handles possibly open terms. We then
-recover the Abstraction Theorem (Theorem~\ref{thm:abstraction}) as the
-special case Theorem~\ref{thm:at-gen} for closed terms of closed type.
-
-\begin{thm}\label{thm:at-gen}
-Every well-formed term $\Gamma;\Phi~|~\Delta \vdash t : F$ induces
-a natural transformation from $\sem{\Gamma; \Phi \vdash \Delta}$ to
-$\sem{\Gamma; \Phi \vdash F}$, i.e., a triple of natural
-transformations 
-\[(\setsem{\Gamma;\Phi~|~\Delta \vdash t : F},\,
-\setsem{\Gamma;\Phi~|~\Delta \vdash t : F},\,
-\relsem{\Gamma;\Phi~|~\Delta \vdash t : F})\]
-where
-\[\begin{array}{lll}
-\setsem{\Gamma;\Phi~|~\Delta \vdash t : F} & : & \setsem{\Gamma;
-  \Phi \vdash \Delta} \to \setsem{\Gamma; \Phi \vdash F}
-\end{array}\]
-has as its component at $\rho : \setenv$ a morphism
-\[\begin{array}{lll}
-\setsem{\Gamma;\Phi~|~\Delta \vdash t : F}\rho & : & \setsem{\Gamma;
-  \Phi \vdash \Delta}\rho \to \setsem{\Gamma; \Phi \vdash F}\rho
-\end{array}\]
-in $\set$,
-\[\begin{array}{lll}
-\relsem{\Gamma;\Phi~|~\Delta \vdash t : F} & : & \relsem{\Gamma;
-  \Phi \vdash \Delta} \to \relsem{\Gamma; \Phi \vdash F}
-\end{array}\]
-has as its component at $\rho : \relenv$ a morphism
-\[\begin{array}{lll}
-\relsem{\Gamma;\Phi~|~\Delta \vdash t : F}\rho & : & \relsem{\Gamma;
-  \Phi \vdash \Delta}\rho \to \relsem{\Gamma; \Phi \vdash F}\rho
-\end{array}\]
-in $\rel$,
-and, for all $\rho : \relenv$,
-\begin{equation}\label{eq:projs}
-\relsem{\Gamma;\Phi~|~\Delta \vdash t : F}\rho =
-(\setsem{\Gamma;\Phi~|~\Delta \vdash t : F}(\pi_1 \rho),\,
-\setsem{\Gamma;\Phi~|~\Delta \vdash t : F}(\pi_2 \rho))
-\end{equation}
-\end{thm}
-\begin{proof}
-By structural induction on the type judgment for $t$. The overall
-challenge of our development manifests in this proof. It lies in
-showing that the set and relational interpretations of each term
-judgment are natural transformations, and that all set interpretations
-of terms of $\Nat$-types satisfy the appropriate equality preservation
-conditions from Definition~\ref{def:set-sem}.  For the interesting
-cases of abstraction, application, $\map$, $\tin$, and $\fold$ terms,
-propagating the naturality conditions is quite involved; the latter
-two cases especially require some rather delicate diagram
-chasing. That this is possible provides strong evidence that our
-definitions and development are sensible, natural, and at an
-appropriate level of abstraction. Detailed proofs for the set
-interpretations for the five interesting forms of terms are included
-in the accompanying anonymous supplementary material. Even in the
-supplementary material we omit the other, straightforward, cases.  The
-proofs for the relational interpretations of all term judgments are
-entirely analogous to the proofs for their set interpretations. The
-proof that Equation~\ref{eq:projs} holds in each case is by direct
-calculation, using the facts that projections are surjective and that
-the set and relational interpretations are defined ``in parallel'',
-i.e., are fibred.
-\end{proof}
-
-The following theorem is an immediate consequence of
-Theorem~\ref{thm:at-gen}:
-\begin{thm}\label{thm:at-gen-rel}
-If \,$\Gamma; \Phi \,|\, \Delta \vdash t : F$ and \,$\rho \in \relenv$,
-  and if \,$(a, b) \in \relsem{\Gamma; \Phi \vdash \Delta} \rho$,
-  then \\
-  $(\setsem{\Gamma; \Phi \,|\, \Delta \vdash t : F} (\pi_1 \rho) \,a \, ,
-      \setsem{\Gamma; \Phi \,|\, \Delta \vdash t : F } (\pi_2 \rho) \,b) \in 
-    \relsem{\Gamma; \Phi \vdash F} \rho$
-\end{thm}
-\noindent
-Finally, the Abstraction Theorem is the instantiation of
-Theorem~\ref{thm:at-gen-rel} to closed terms of closed type:
-\begin{thm}[Abstraction Theorem]\label{thm:abstraction}
-If $\vdash t : F$, then $(\setsem{\vdash t : F},\setsem{\vdash t
-  : F}) \in \relsem{\vdash F}$.
-\end{thm}
-
-\section{Free Theorems for Nested Types}\label{sec:ftnt}
-
-In this section we show how Theorem~\ref{thm:abstraction} can be used
-to prove more advanced free theorems than those of
-Section~\ref{sec:Nat-type-terms} in the presence of directly
-constructed nested types.
-
-\subsection{Free Theorem for Type of Polymorphic
-  Bottom}\label{sec:bottom}
-
-Suppose $ \vdash g : \Nat^\alpha \,\onet\,\alpha$, let $G^\set =
-\setsem{\vdash g : \Nat^\alpha \,\onet\,\alpha}$, and let $G^\rel =
-\relsem{\vdash g : \Nat^\alpha \,\onet\,\alpha}$.  By
-Theorem~\ref{thm:at-gen}, $(G^\set(\pi_1\rho),G^\set(\pi_2\rho))
-= G^\rel\rho$. Thus, for all $\rho \in \relenv$ and any $(a, b) \in
-\relsem{\vdash \emptyset}\rho = 1$, eliding the only possible
-instantiations of $a$ and $b$ gives that
-%\[\begin{array}{lll}
-$(G^\set,G^\set) \;= \; (G^\set(\pi_1 \rho), G^\set (\pi_2 \rho))$
-$ \in  \relsem{\vdash \Nat^\alpha \,\onet\,\alpha}\rho$
-$ =  \{\eta : K_1 \Rightarrow \id\}$
-$ =  \{(\eta_1 : K_1 \Rightarrow \id, \eta_2 : K_1 \Rightarrow
-\id)\}$ 
-%\end{array}\]
-That is, $G^\set$ is a natural transformation from the constantly
-$1$-valued functor to the identity functor in $\set$. In particular,
-for every $S : \set$, $G^\set_S : 1 \to S$. Note, however, that if $S
-= \emptyset$, then there can be no such morphism, so no such natural
-transformation,
-%in $\set$,
-and thus no term $\vdash g : \Nat^\alpha \onet \,\alpha$, can exist.
-%in our calculus.
-That is, our calculus admits no non-terminating terms, i.e., terms
-with the closed type $\Nat^\alpha \onet \,\alpha$ of the polymorphic
-bottom.
-
-\subsection{Free Theorem for Type of Polymorphic
-  Identity}\label{sec:identity}
-
-Suppose $ \vdash g : \Nat^\alpha \,\alpha\,\alpha$, let $G^\set =
-\setsem{\vdash g : \Nat^\alpha \,\alpha\,\alpha}$, and let $G^\rel =
-\relsem{\vdash g : \Nat^\alpha \,\alpha\,\alpha}$.  By
-Theorem~\ref{thm:at-gen}, $(G^\set(\pi_1\rho),G^\set(\pi_2\rho))
-= G^\rel\rho$. Thus, for all $\rho \in \relenv$ and any $(a, b) \in
-\relsem{\vdash \emptyset}\rho = 1$, eliding the only possible
-instantiations of $a$ and $b$ gives that
-%\[\begin{array}{lll}
-$(G^\set, G^\set) \; = \; (G^\set(\pi_1 \rho), G^\set (\pi_2 \rho))$
-$ \in \relsem{\vdash \Nat^\alpha \,\alpha\,\alpha}\rho$
-$ =  \{\eta : \id \Rightarrow \id\}$
-$ =  \{(\eta_1 : \id \Rightarrow \id, \eta_2 : \id \Rightarrow
-\id)\}$
-%\end{array}\]
-That is, $G^\set$ is a natural transformation from the identity
-functor on $\set$ to itself. Now let $S$ be any set.  If $S =
-\emptyset$, then there is exactly one morphism $\id_S: S \to S$, so
-$G^\set_S : S \to S$ must be $\id_S$. If $S \not = \emptyset$, if
-$a$ is any element of $S$, and if $K_a :S \to S$ is the constantly
-$a$-valued morphism on $S$, then instantiating the naturality square
-implied by the above equality gives that $G^\set_S \circ K_a = K_a
-\circ G^\set_S$, i.e., $G^\set_S \, a = a$, i.e., $G^\set_S = \id_S$.
-Putting these two cases together we have that for every $S : \set$,
-$G^\set_S = \id_S$, i.e., $G^\set$ is the identity natural
-transformation for the identity functor on $\set$. So every closed
-term $g$ of closed type $\Nat^\alpha\alpha\,\alpha$ always denotes the
-identity natural transformation for the identity functor on $\set$,
-i.e., every closed term $g$ of type $\Nat^\alpha\alpha\,\alpha$
-denotes the polymorphic identity function.
-
-\subsection{Standard Free Theorems for ADTs and Their Analogues for
-  Nested types}\label{sec:ft-adt}
-
-We can derive in our calculus even those free theorems for polymorphic
-functions over ADTs that are not consequences of naturality.  We can,
-e.g., prove the free theorem for $\mathit{filter}$'s type as follows:
-
-%\begin{lemma}\label{lem:list-graph}
-%If $g : A \to B$, $\rho : \relenv$, and $\rho\alpha = (A, B,
-%\graph{g})$, then $\relsem{\alpha; \emptyset \vdash List \, \alpha}
-%\rho = \graph{map_{\mathit{List}} \, g}$.
-%\end{lemma}
-%\begin{proof}
-%The result follows by straightforward calculation using
-%Definition~\ref{def:rel-sem} and Equation~\ref{eq:mu}, together with
-%the facts that $\mathit{List} \,(A, B, \graph{g}) =
-%\graph{\mathit{map}_{\mathit{List}} \,g}$ and, if $H = \onet + \beta
-%\times\phi \beta$ {\color{blue} Why not $H = \onet + \alpha \times
-%  \beta$? Does it matter?}, then $(T_{H,\rho}^n K_0)^* R =
-%\Sigma_{i=0}^n R^i$ for all $n \in \nat$ and $R \in \rel$ by induction
-%on $n$.
-%\end{proof}
-
-\begin{thm} 
-If $g : A \to B$, $\rho : \relenv$, $\rho \alpha = (A, B, \graph{g})$,
-$(a, b) \in \relsem{\alpha ;\emptyset \vdash \Delta} \rho$, $(s \circ
-g, s) \in \relsem{\alpha; \emptyset \vdash \Nat^\emptyset \alpha \,
-  \mathit{Bool}} \rho$, and % for $s : A \to \mathit{Bool}$, and
-%$\alpha; \emptyset \,|\, \Delta \vdash t :
-%\filtype$, if
-$\mathit{filter} = \setsem{\alpha; \emptyset \,|\, \Delta \vdash t :
-  \filtype}$ for some $t$, then
-\[  \mathit{map}_{\mathit{List}} \,g \circ \mathit{filter} \, (\pi_1
-\rho) \, a \, (s \circ g) = \mathit{filter}\, (\pi_2\rho) \, b \, s
-\circ \mathit{map}_{\mathit{List}} \,g\]
-\end{thm}
-\begin{proof}
-By Theorem~\ref{thm:at-gen-rel}, \[(\mathit{filter}\, (\pi_1 \rho)\, a,
-\mathit{filter}\, (\pi_2 \rho)\, b) \in \relsem{\alpha; \emptyset
-  \vdash \filtype} \rho\] so if $(s', s) \in \relsem{\alpha; \emptyset
-  \vdash \Nat^\emptyset \alpha \, \mathit{Bool}} \rho = \rho\alpha \to
-\Eq_{\mathit{Bool}}$ and $(xs', xs) \in \relsem{\alpha; \emptyset
-  \vdash List \, \alpha} \rho$ then
-% then
-%  \begin{align*}
-%    (t (\pi_1 \rho) \,a \,s, t (\pi_2 \rho) \,b \,s') &\in \relsem{\alpha; \emptys%et \vdash \Nat^\emptyset (List \, \alpha) \, (List \, \alpha)} \rho \\
-%   &= \relsem{\alpha; \emptyset \vdash List \, \alpha} \rho \to \relsem{\alpha; \e%mptyset \vdash List \, \alpha} \rho \\
-%  \end{align*}
-%  So if $(xs, xs') \in \relsem{\alpha; \emptyset \vdash List \, \alpha} \rho$ then,
-\begin{equation}\label{eq:filter-thm-list}
-  (\mathit{filter}\, (\pi_1\rho) \,a \,s' \,xs', \mathit{filter} \,
-  (\pi_2\rho) \,b \,s \,xs) \in \relsem{\alpha; \emptyset \vdash
-    \mathit{List} \, \alpha} \rho
-\end{equation}
-If $\rho\alpha = (A, B, \graph{g})$, then $\relsem{\alpha; \emptyset
-  \vdash \mathit{List} \, \alpha} \rho =
-\graph{\mathit{map}_{\mathit{List}} \, g}$ by Lemma~\ref{lem:graph}
-and Equation~\ref{thm:demotion-objects}.  Moreover, $xs =
-\mathit{map}_{\mathit{List}} \,g \,xs'$ and $(s', s) \in \graph{g} \to
-\Eq_{\mathit{Bool}}$, so $s' = s \circ g$. The result
-follows from % Instantiating
-Equation~\ref{eq:filter-thm-list}.\end{proof}
-%gives the result.
-% \begin{align*}
-%    &(t (\pi_1\rho) \,a \,(s' \circ g) \,xs, t (\pi_2\rho) \,b \,s' \,(\mathit{map%}_{\mathit{List}} \,g \,xs)) \in \graph{\mathit{map}_{\mathit{List}} \,g}, \\ 
-%    & i.e., \\ 
-%    &\mathit{map}_{\mathit{List}} \,g \, (t (\pi_1\rho) \,a \,(s' \circ g) \,xs) =% t (\pi_2\rho) \,b \,s' \,(\mathit{map}_{\mathit{List}} \,g \,xs), \\ 
-%    & i.e., \\
-%    &\mathit{map}_{\mathit{List}} \,g \circ t (\pi_1 \rho) \, a \, (s' \circ g) = %t (\pi_2\rho) \, b \, s' \circ \mathit{map}_{\mathit{List}} \,g
-%  \end{align*}
-%  as desired.
-%\end{proof}
-
-\begin{comment}
-\begin{thm}\label{thm:at-gen-rel}
-If \,$\Gamma; \Phi \,|\, \Delta \vdash t : \tau$ and \,$\rho \in \relenv$,
-  and if \,$(a, b) \in \relsem{\Gamma; \Phi \vdash \Delta} \rho$,
-  then \\
-  $(\setsem{\Gamma; \Phi \,|\, \Delta \vdash t : \tau} (\pi_1 \rho) \,a \, ,
-      \setsem{\Gamma; \Phi \,|\, \Delta \vdash t : \tau } (\pi_2 \rho) \,b) \in 
-    \relsem{\Gamma; \Phi \vdash \tau} \rho$
-\end{thm}
-\begin{proof}
- Immediate from Theorem~\ref{thm:at-gen}.
-\end{proof}
-\end{comment}
-
-\begin{comment}
-\begin{thm} 
-  If $g : A \to B$, 
-  $\rho : \relenv$, 
-  $\rho \alpha = (A, B, \graph{g})$,
-  $(a, b) \in \relsem{\alpha ;\emptyset \vdash \Delta} \rho$, 
-  $(s \circ g, s) \in \relsem{\alpha; \emptyset \vdash \Nat^\emptyset \alpha \, \mathit{Bool}} \rho$,
-  and, for some well-formed term $\mathit{filter}$, \\
-  $$t = \setsem{\alpha; \emptyset \,|\, \Delta \vdash \mathit{filter} : \filtype} \emph{, then }$$
-  \begin{align*}
-  \mathit{map}_{\mathit{List}} \,g \circ t (\pi_1 \rho) \, a \, (s \circ g) = t (\pi_2\rho) \, b \, s \circ \mathit{map}_{\mathit{List}} \,g
-  \end{align*}
-\end{thm}
-\begin{proof}
-  By Theorem~\ref{thm:at-gen-rel},
-  $(t (\pi_1 \rho) a, t (\pi_2 \rho) b) \in \relsem{\alpha; \emptyset \vdash \filtype} \rho$.
-  Thus if 
-  $(s, s') \in \relsem{\alpha; \emptyset \vdash \Nat^\emptyset \alpha \, \mathit{Bool}} \rho = \rho\alpha \to \Eq_{\mathit{Bool}}$,
-  then 
-  \begin{align*}
-    (t (\pi_1 \rho) \,a \,s, t (\pi_2 \rho) \,b \,s') &\in \relsem{\alpha; \emptyset \vdash \Nat^\emptyset (List \, \alpha) \, (List \, \alpha)} \rho \\
-   &= \relsem{\alpha; \emptyset \vdash List \, \alpha} \rho \to \relsem{\alpha; \emptyset \vdash List \, \alpha} \rho \\
-  \end{align*}
-  So if $(xs, xs') \in \relsem{\alpha; \emptyset \vdash List \, \alpha} \rho$ then,
-  \begin{equation}\label{eq:filter-thm}
-  (t (\pi_1\rho) \,a \,s \,xs, t (\pi_2\rho) \,b \,s' \,xs') \in \relsem{\alpha; \emptyset \vdash List \, \alpha} \rho
-  \end{equation}
-
-  Consider the case in which $\rho\alpha = (A, B, \graph{g})$. Then $\relsem{\alpha; \emptyset \vdash List \, \alpha} \rho
-  = \graph{\mathit{map}_{\mathit{List}} \, g}$, by Lemma~\ref{lem:list-graph}, and $(xs, xs') \in \graph{\mathit{map}_{\mathit{List}} \,g}$ 
-  implies $xs' = \mathit{map}_{\mathit{List}} \,g \,xs$. We also have that 
-  $(s, s') \in \graph{g} \to \Eq_{\mathit{Bool}}$ implies 
-  $\forall (x, g x) \in \graph{g}. \,\, s x = s' (g x)$ and thus
-  $s = s' \circ g$ due to the definition of morphisms between relations.
-  %% TODO show (s, s') \in <g> -> \mathit{Bool} implies s = s' \circ g %% 
-  With these instantiations, Equation~\ref{eq:filter-thm} becomes
-  \begin{align*}
-    &(t (\pi_1\rho) \,a \,(s' \circ g) \,xs, t (\pi_2\rho) \,b \,s' \,(\mathit{map}_{\mathit{List}} \,g \,xs)) \in \graph{\mathit{map}_{\mathit{List}} \,g}, \\ 
-    & i.e., \\ 
-    &\mathit{map}_{\mathit{List}} \,g \, (t (\pi_1\rho) \,a \,(s' \circ g) \,xs) = t (\pi_2\rho) \,b \,s' \,(\mathit{map}_{\mathit{List}} \,g \,xs), \\ 
-    & i.e., \\
-    &\mathit{map}_{\mathit{List}} \,g \circ t (\pi_1 \rho) \, a \, (s' \circ g) = t (\pi_2\rho) \, b \, s' \circ \mathit{map}_{\mathit{List}} \,g
-  \end{align*}
-  as desired.
-\end{proof}
-\end{comment}
-
-%\subsection{Free Theorem for Type of $\mathit{filter}$ for
-%  $\GRose$}\label{sec:filter-grose}
-
-A similar proof establishes the analogous result for, say, generalized
-rose trees. 
-\begin{thm} 
-  If $g : A \to B$,
-$F, G : \set \to \set$,
-  $\eta : F \to G$ in $\set$, $\rho : \relenv$, $\rho \alpha =
- (A, B, \graph{g})$, $\rho \psi = (F, G, \graph{\eta})$, $(a, b) \in
- \relsem{\alpha, \psi ;\emptyset \vdash \Delta} \rho$, $(s \circ
- g, s) \in \relsem{\alpha; \emptyset \vdash \Nat^\emptyset \alpha \,
-   \mathit{Bool}} \rho$, and
- \[ \mathit{filter} = \setsem{\alpha, \psi; \emptyset \,|\, \Delta
-      \vdash t : \filtypeGRose}  \]
-for some $t$, then
-\[ \semmap_{\mathit{GRose}}\, \eta\, (g + 1) \circ \mathit{filter} \,
-(\pi_1 \rho) \, a \, (s \circ g) = \mathit{filter} \, (\pi_2\rho) \, b
-\, s \circ 
-\semmap_{\mathit{GRose}}\, \eta\, g\]
-\end{thm}
-
-\noindent
-This is not surprising since rose trees are essentially
-ADT-like. However, as noted in Section~\ref{sec:terms}, our calculus
-cannot express the type of a polymorphic filter function for a proper
-nested type.
-
-\subsection{Short Cut Fusion for Lists}\label{sec:short-cut}
-
-We can recover standard short cut fusion for lists~\cite{glp93} in our
-calculus: 
-\begin{thm}
-If \,$\vdash F$, $\vdash H$, and 
-%$\beta; \emptyset \,|\, \emptyset \vdash g : \Nat^{\emptyset}
-%(\Nat^{\emptyset} (\onet + \tau \times \beta)\, \beta)\, \beta$. 
-%If
-$G \, = \, \setsem{\beta; \emptyset \,|\, \emptyset \vdash g :
-  \Nat^{\emptyset} (\Nat^{\emptyset} (\onet + F \times \beta)\,
-  \beta)\, \beta}$ for some $g$, then
-\[\mathit{fold}_{1 + \setsem{\vdash F} \times \_}\, n\, c\; (G\; (\mathit{List}\,
-\setsem{\vdash F})\,\mathit{nil} \,\mathit{cons}) = G \,\setsem{\vdash
-  H}\, n\, c \]
-\end{thm}
-\begin{proof}
-%Let $\vdash \tau : \F$ and $\vdash \tau' : \F$, let
-%\[\beta; \emptyset \,|\, \emptyset \vdash g : \Nat^{\emptyset}
-%(\Nat^{\emptyset} (\onet + \tau \times \beta)\, \beta)\, \beta\]
-%and let
-%\[G = \setsem{\beta; \emptyset \,|\, \emptyset
-%  \vdash g : \Nat^{\emptyset} (\Nat^{\emptyset} (\onet + \tau \times
-  %  \beta)\, \beta)\, \beta}\] Then
-Theorem~\ref{thm:abstraction} gives
-that, for any $\rho : \relenv$,
-%and any $(a, b) \in
-%\relsem{\beta; \emptyset \vdash \emptyset}\rho = 1$, then (eliding the
-%only possible instantiations of $a$ and $b$) we have
-\[(G \,(\pi_1 \rho), G\, (\pi_2 \rho))
- \in  \relsem{\beta; \emptyset \vdash \Nat^{\emptyset} (\Nat^{\emptyset}
-  (\onet + F \times \beta)\, \beta)\, \beta} \rho\\
-\cong (((\relsem{\vdash F} \rho \times
-\rho\beta) \to \rho\beta) \times \rho\beta) \to \rho\beta\] 
-%\[\begin{array}{lll}
-%(G \,(\pi_1 \rho), G\, (\pi_2 \rho))
-% & \in & \relsem{\beta; \emptyset \vdash \Nat^{\emptyset} (\Nat^{\emptyset}
-%  (\onet + \tau \times \beta)\, \beta)\, \beta} \rho\\
-%& \cong& (((\relsem{\vdash \tau} \rho \times
-%\rho\beta) \to \rho\beta) \times \rho\beta) \to \rho\beta 
-%\end{array}\]
-%
-%Since
-%\[\begin{array}{rl}
-% & \relsem{\beta; \emptyset \vdash \Nat^{\emptyset}
-%  (\Nat^{\emptyset} (\onet + \tau \times \beta)\, \beta)\, \beta} \rho\\  
-%=& \relsem{\beta; \emptyset \vdash \Nat^{\emptyset} (\onet +
-%  \tau \times \beta)\, \beta} \rho \to \rho\beta \\ 
-%=& (\relsem{\beta; \emptyset \vdash \onet + \tau \times \beta}
-%\rho \to \rho\beta) \to \rho\beta \\ 
-%=& ((\onet + \relsem{\vdash \tau} \rho \times
-%\rho\beta) \to \rho\beta) \to \rho\beta \\ 
-%\cong& (((\relsem{\vdash \tau} \rho \times
-%\rho\beta) \to \rho\beta) \times \rho\beta) \to \rho\beta 
-%\end{array}\]
-%we have that
-so if $(c', c) \in \relsem{\vdash F} \rho \times
-\rho\beta \to \rho\beta$ and $(n', n) \in \rho\beta$ then
-$(G \,(\pi_1 \rho)\, n'\, c', G\, (\pi_2 \rho)\, n\, c)
-\in \rho \beta$.
-In addition,
-\[\setsem{\vdash \fold_{\onet + F
-    \times \beta}^{H} : \Nat^{\emptyset} (\Nat^{\emptyset} (\onet
-  + F \times H)\, H)\, (\Nat^{\emptyset} (\mu \alpha. \onet
-  + F \times \alpha)\, H)} =
-\mathit{fold}_{1 + \setsem{\vdash F} \times \_}\] so that if $c \in
-\setsem{\vdash F} \,\times\, \setsem{\vdash H} \to
-\setsem{\vdash H}$ and $n \in \setsem{\vdash H}$, then
-$(n, c) \in \setsem{\vdash \Nat^{\emptyset} (\onet + F \times H)\,
-  H}$. The instantiation
-$\pi_1 \rho \beta = \setsem{\vdash \mu \alpha. \onet + F \times
-  \alpha}=\mathit{List}\,\setsem{\vdash F}$,\, 
-$\pi_2 \rho \beta = \setsem{\vdash H}$,\,
-$\rho \beta = \graph{\mathit{fold}_{1 + \setsem{\vdash F} \times \_}\, n\, c} :
-\rel(\pi_1 \rho \beta, \pi_2 \rho \beta)$,\,
-$c' = \mathit{cons}$,\, and 
-$n' = \mathit{nil}$
-%Consider the instantiation:
-%\[\begin{array}{rcl}
-%\pi_1 \rho \beta &=& \setsem{\vdash \mu \alpha. \onet + \tau \times
-%  \alpha} \;=\;\List\,\setsem{\vdash \tau}\\   
-%\pi_2 \rho \beta &=& \setsem{\vdash \tau'}\\
-%\rho \beta &=& \graph{\mathit{fold}_{1 + \tau \times \_}\, n\, c} :
-%\rel(\pi_1 \rho \beta, \pi_2 \rho \beta) \\ 
-%c' &=& \mathit{cons}\\
-%n' &=& \mathit{nil}
-%\end{array}\]
-%Clearly, $(\mathit{nil}, n) \in \rho\beta =
-%\graph{\mathit{fold}_{1 + \tau \times \_}\, n\, c}$ because $
-%\mathit{fold}_{1 + \tau \times \_}\, n\, c \,\mathit{nil} = n$.  Moreover,
-%$(\mathit{cons}, c) \in \relsem{\vdash \tau} \,\times\, \rho\beta \to
-%\rho\beta$ since if $(x, x') \in \relsem{\vdash \tau}$, i.e., $x =
-%x'$, and if $(y, y') \in \rho\beta =
-%\graph{\mathit{fold}_{1 + \tau \times \_} n\, c}$, i.e., $y' =
-%\mathit{fold}_{1 + \tau \times \_} n\, c\, y$, then
-%\[(\mathit{cons}\, x\, y, c\, x\, (\mathit{fold}_{1 + \tau \times \_}\, n\,
-%c\, y)) \in \graph{\mathit{fold}_{1 + \tau \times \_}\, n\, c}\]
-%i.e.,
-%\[c\, x\, (\mathit{fold}_{1 + \tau \times \_}\, n\, c\, y)
-%= \mathit{fold}_{1 + \tau \times \_}\, n\, c\, (\mathit{cons} \,x\, y)\]
-%holds by definition of $\mathit{fold}_{1 + \tau \times \_}$.  We
-%therefore conclude that
-thus gives that $(G \;(\mathit{List}\,\setsem{\vdash
-  F})\,\mathit{nil} \,\mathit{cons}, G \,\setsem{\vdash H} \,
-n\, c) \in \graph{\mathit{fold}_{1 + \setsem{\vdash F} \times \_}\, n\, c}$, i.e.,
-that $\mathit{fold}_{1 + \setsem{\vdash F} \times \_}\, n\, c\; (G\;
-(\mathit{List}\,\setsem{\vdash F})\, \mathit{nil}\, \mathit{cons})
-= G \,\setsem{\vdash H}\, n\, c$.
-\end{proof}
-
-We can extend short cut fusion results to arbitrary ADTs, as
-in~\cite{joh02,pit98}.
-
-\begin{comment}
-\subsection{Short Cut Fusion for Arbitrary ADTs}\label{sec:short-cut-adt}
-
-\begin{thm}
-Let $\ol{\vdash \tau}$, let $\vdash \tau'$, let $\ol{\alpha}; \beta
-\vdash F$, and let $\beta; \emptyset  \,|\,
-\emptyset \vdash g : \Nat^{\emptyset} (\Nat^{\emptyset} F[\ol{\alpha
-    := \tau}]\, \beta)\, \beta$.  If we regard
-\[\begin{array}{lll}
-H & = & \setsem{\emptyset;\beta \vdash F[\ol{\alpha :=
-      \tau}]}\\
-G & = & \setsem{\beta; \emptyset \,|\, \emptyset \vdash g :
-  \Nat^{\emptyset} (\Nat^{\emptyset}\, F[\ol{\alpha := \tau}]\,
-  \beta)\, \beta}
-\end{array}\]
-as functors in $\beta$, then for every $B \in H \setsem{\vdash \tau'}
-\rightarrow \setsem{\vdash \tau'}$ we have
-\[\mathit{fold}_H\, B \; (G\; \mu H \; in_H) = G \,\setsem{\vdash \tau'}\, B\]
-\end{thm}
-
-\vspace*{0.2in}
-
-\begin{proof}
-  We first note that the type of $g$ is well-formed, since
-  $\emptyset;\beta \vdash F[\ol{\alpha := \tau}]$ so our
-  promotion theorem gives that $\beta;\emptyset\vdash F[\ol{\alpha :=
-      \tau}]$, and $\emptyset;\beta\vdash\beta$ so that our
-  promotion theorem gives $\beta;\emptyset\vdash\beta$. From
-  these facts we deduce that $\beta;\emptyset \vdash
-  \Nat^\emptyset\,F[\ol{\alpha := \tau}]\,\beta : \T$, and thus that
-  $\beta; \emptyset \vdash \Nat^{\emptyset} (\Nat^{\emptyset}
-  F[\ol{\alpha := \tau}]\, \beta)\, \beta : \T$.
-
-  Theorem~\ref{thm:abstraction} gives that, for any $\rho : \relenv$
-  and any $(a, b) \in \relsem{\beta; \emptyset \vdash \emptyset}\rho =
-  1$, eliding the only possible instantiations of $a$ and $b$ gives
-  that
-\[(G \,(\pi_1 \rho), G\, (\pi_2 \rho))
-\in \relsem{\beta; \emptyset \vdash \Nat^{\emptyset} (\Nat^{\emptyset}
-  F[\ol{\alpha := \tau}]\, \beta)\, \beta} \rho\]
-Since
-\[\begin{array}{rl}
- & \relsem{\beta; \emptyset \vdash \Nat^{\emptyset}
-  (\Nat^{\emptyset} F[\ol{\alpha := \tau}]\, \beta)\, \beta} \rho\\  
-=& \relsem{\beta; \emptyset \vdash \Nat^{\emptyset} F[\ol{\alpha :=
-      \tau}] \, \beta} \rho \to \rho\beta \\ 
-\end{array}\]
-we have that if $(A, B) \in \relsem{\beta; \emptyset \vdash
-  \Nat^{\emptyset} F[\ol{\alpha := \tau}] \, \beta} \rho$
-then
-\[(G \,(\pi_1 \rho)\, A, G\, (\pi_2 \rho)\, B)
-\in \rho \beta\]
-Now note that
-\[\setsem{\vdash \fold_{F[\ol{\alpha := \tau}]}^{\tau'} :
-  \Nat^{\emptyset}\, (\Nat^{\emptyset}\, F[\ol{\alpha := \tau}][\beta
-    := \tau'] \, \tau')\, (\Nat^{\emptyset} \,(\mu
-  \beta. F[\ol{\alpha := \tau}] \, \tau')} =
-\mathit{fold}_H\]
-and consider the instantiation
-\[\begin{array}{lll}
-A &=& \mathit{in}_H :  H (\mu H) \to \mu H\\
-B & : & H\setsem{\vdash \tau'} \to \setsem{\vdash \tau'}\\
-\rho \beta &=& \graph{\mathit{fold}_H\, B}
-\end{array}\]
-(Note that all the types here are well-formed.)  This gives
-\[\begin{array}{lll}
-\pi_1 \rho \beta &=& \setsem{\vdash \mu \beta. F[\ol{\alpha := \tau}]}
-\; = \; \mu H\\
-\pi_2 \rho \beta &=& \setsem{\vdash \tau'}\\
-\rho \beta &:& \rel(\pi_1 \rho \beta, \pi_2 \rho \beta)\\
-A & : & \setsem{\beta; \emptyset \vdash \Nat^{\emptyset}
-  F[\ol{\alpha := \tau}] \, \beta} (\pi_1 \rho)\\
-B & : & \setsem{\beta; \emptyset \vdash \Nat^{\emptyset} F[\ol{\alpha
-      := \tau}] \, \beta}(\pi_2 \rho)\\
-\end{array}\]
-since
-\[\begin{array}{lll}
-A \; = \; \mathit{in}_H & : & H(\mu H) \to \mu H\\
- & = & \setsem{\emptyset;\beta \vdash F[\ol{\alpha := \tau}]}(\mu
-\setsem{\emptyset;\beta \vdash F[\ol{\alpha := \tau}]}) \to \mu
-\setsem{\emptyset;\beta \vdash F[\ol{\alpha := \tau}]}\\
-& = & \setsem{\emptyset;\beta \vdash F[\ol{\alpha :=
-      \tau}]}(\pi_1\rho) \to \setsem{\emptyset;\beta \vdash
-  \beta}(\pi_1 \rho)\\
-& = & \setsem{\beta;\emptyset \vdash F[\ol{\alpha :=
-      \tau}]}(\pi_1\rho) \to \setsem{\beta; \emptyset \vdash
-  \beta}(\pi_1 \rho) \hspace*{0.25in} {\color{red} \mbox{Daniel's
-    trick; now a theorem}} \\ 
-& = & \setsem{\beta;\emptyset \vdash \Nat^\emptyset F[\ol{\alpha :=
-      \tau}]\,\beta}(\pi_1 \rho) 
-\end{array}\]
-{\color{red} where ``Daniel's trick'' is the observation that a
-  functor can be seen as non-functorial when we only care about its
-  action on objects. This is now a theorem.}  We also have
-\[\begin{array}{lll}
-(A,B) \,=\, (\mathit{in}_H,B) & \in & \relsem{\beta; \emptyset \vdash
-  \Nat^{\emptyset} F[\ol{\alpha := \tau}] \, \beta} \rho \\
- & = & \relsem{\beta; \emptyset \vdash F[\ol{\alpha :=
-      \tau}]}\rho[\beta := \graph{\mathit{fold}_H\, B}] \to
-\graph{\mathit{fold}_H\, B}\\
- & = & \relsem{\beta; \emptyset \vdash F[\ol{\alpha :=
-      \tau}]}\graph{\mathit{fold}_H\, B} \to
-\graph{\mathit{fold}_H\, B}\\
- & = & \relsem{\emptyset; \beta\vdash F[\ol{\alpha :=
-      \tau}]}\graph{\mathit{fold}_H\, B} \to
-\graph{\mathit{fold}_H\, B} \hspace*{0.35in} {\color{red}
-  \mbox{Daniel's trick; now a theorem}}\\  
-& = & \graph{\setsem{\emptyset; \beta \vdash F[\ol{\alpha :=
-      \tau}]}\,(\mathit{fold}_H\, B)} \to
-\graph{\mathit{fold}_H\, B} \hspace*{0.25in} {\color{red} \mbox{by the
-    graph lemma}}\\ 
-& = & \graph{\mathit{map}_H \,(\mathit{fold}_H\,B)} \to
-\graph{\mathit{fold}_H\, B}\\
-\end{array}\]
-since if $(x,y) \in \graph{\mathit{map}_H \,(\mathit{fold}_H\,B)}$,
-i.e., if $\mathit{map}_H \,(\mathit{fold}_H\,B) \, x = y$, then
-$\mathit{fold}_H\, B\, (\mathit{in}_H\,x) = B\,y = B\, (\mathit{map}_H
-\,(\mathit{fold}_H\,B) \, x)$ by the definition of $\mathit{fold}_H$
-as a (indeed, the unique) morphism from $\mathit{in}_H$ to $B$. Thus,
-\[(G \,(\pi_1 \rho)\, A, G\, (\pi_2 \rho)\, B) \in
-\graph{\mathit{fold}_H\, B}\]
-i.e.,
-\[\mathit{fold}_H \, B \, (G\, (\pi_1 \rho) \, \mathit{in}_H) =
-G\,(\pi_2 \rho)\,B\]
-
-\vspace*{0.1in}
-
-\noindent
-Since $\beta$ is the only free variable in $G$, this simplifies to
-\[\mathit{fold}_H\, B \, (G\, \mu H\, \mathit{in}_H) =
-G\,\setsem{\vdash \tau'}\,B\]  
-\end{proof}
-\end{comment}
-
-\subsection{Short Cut Fusion for Arbitrary Nested
-  Types}\label{sec:short-cut-nested}
-
-%fold/build rule shows that Church encodings for nested types are iso
-%(have the same interp) as the nested types themselves. Just like Ex
-%2.8 in parpoe, but categorical rather than operational semantics. But
-%built-in data types are more versatile in implementations. They can be
-%inducted on, e.g. And they are stored in the heap, not the run-time
-%stack, and are therefore more efficient.
-%``Most higher order type languages meant for human programmers eschew
-%fully impredicative polymorphism'' says Andy on p322.
-
-We can also, finally, formally prove correctness of the categorically
-inspired short cut fusion for nested types from~\cite{jg10}. Indeed,
-we have:
-\begin{thm}\label{thm:short-cut-nested}
-If $\emptyset;\phi,\alpha \vdash F$, \,$\emptyset; \alpha
-\vdash K$, \,
-%let $\phi ;\emptyset\,|\,\emptyset\vdash g :
-%\Nat^\emptyset\,(\Nat^\alpha\,F\,(\phi\alpha)) \,
-%(\Nat^\alpha\,\onet \, (\phi\alpha))$. If we let
-$H : [\set,\set] \to [\set,\set]$ is defined by
-\[\begin{array}{lll}
-H\,f\,x & = & \setsem{\emptyset; \phi, \alpha \vdash F}[\phi :=
-  f][\alpha := x]\\
-\end{array}\]
-and 
-\[G = \setsem{\phi;\emptyset\,|\,\emptyset \vdash g :
-\Nat^\emptyset\,(\Nat^\alpha\,F\,(\phi\alpha))\,(\Nat^\alpha\,\onet \,
-(\phi\alpha))}\] for some $g$, then, for every $B \in H
-\setsem{\emptyset;\alpha \vdash K} \rightarrow \setsem{\emptyset;
-  \alpha \vdash K}$,
-$\mathit{fold}_{H}\, B \, (G\; \mu H \; \mathit{in}_{H}) = G
-\,\setsem{\emptyset;\alpha \vdash K}\, B$.
-\end{thm}
-
-\begin{proof}
-%We first note that the type of $g$ is well-formed since
-%$\emptyset;\phi,\alpha \vdash F : \F$ so our promotion theorem gives
-%that $\phi;\alpha \vdash F : \F$, and $\phi;\alpha \vdash \phi\alpha :
-%\F$, so that $\phi;\emptyset \vdash \Nat^\alpha F\,(\phi\alpha) : \T$
-%and $\phi;\emptyset \vdash \Nat^\alpha \onet \,(\phi\alpha) : \T$.
-%Then $\phi;\emptyset \vdash \Nat^\alpha F\,(\phi\alpha) : \F$ and
-%$\phi;\emptyset \vdash \Nat^\alpha \onet \,(\phi\alpha) : \F$ also
-%%%hold, and, finally, 
-%$\phi ;\emptyset\vdash 
-%\Nat^\emptyset\,(\Nat^\alpha\,F\,(\phi\alpha)) \,
-%(\Nat^\alpha\,\onet \, (\phi\alpha)) : \T$
-%  
-Theorem~\ref{thm:abstraction} gives that, for any 
-$\rho : \relenv$,
-%and any $(a, b) \in \relsem{\phi,\alpha;\emptyset \vdash
-%  \emptyset}\rho = 1$, eliding the only possible instantiations
-%of $a$ and $b$ gives that
-\[\begin{array}{lll}
-(G \,(\pi_1 \rho), G\, (\pi_2 \rho))
-& \in & \relsem{\phi;\emptyset\vdash \Nat^{\emptyset} (\Nat^\alpha
-  F\, (\phi\alpha))\, (\Nat^\alpha\,\onet \, (\phi\alpha))}
-\rho\\ 
-& = & \relsem{\phi;\emptyset\vdash \Nat^\alpha F\,
-  (\phi\alpha)}\rho \to \relsem{\phi;\emptyset\vdash
-  \Nat^\alpha\,\onet \, (\phi\alpha)}\rho\\ 
-%& = & \relsem{\phi;\emptyset\vdash \Nat^\alpha F\,
-%  (\phi\alpha)}\rho \to (\,\lambda A. 1 \Rightarrow \lambda A.\,(\rho
-%\phi) A\,)\\   
-%& = & \relsem{\phi;\emptyset\vdash \Nat^\alpha F\,
-%  (\phi\alpha)}\rho \to (1 \Rightarrow \rho \phi)\\ 
-& = & \relsem{\phi;\emptyset\vdash \Nat^\alpha F\,
-  (\phi\alpha)}\rho \to \rho \phi
-\end{array}\]
-\noindent
-so if $(A, B) \in \relsem{\phi;\emptyset\vdash \Nat^\alpha F\,
-  (\phi\alpha)}\rho$ then $(G \,(\pi_1 \rho)\, A, G\, (\pi_2 \rho)\,
-B) \in \rho \phi$.
-In addition,
-%$\mathit{fold}_H = \setsem{\vdash \fold_F^K :
-%  \Nat^{\emptyset}\, (\Nat^\alpha F[\phi := K]\,K)\, (\Nat^\alpha
-%  ((\mu \phi.\lambda\alpha.F)\alpha)\,K)}$.
-\[\setsem{\vdash \fold_F^K :
-  \Nat^{\emptyset}\, (\Nat^\alpha F[\phi := K]\,K)\, (\Nat^\alpha
-  ((\mu \phi.\lambda\alpha.F)\alpha)\,K)} = \mathit{fold}_H\]
-Consider the instantiation $A = \mathit{in}_H : H (\mu H) \Rightarrow
-\mu H$,\, $B : H\setsem{\emptyset;\alpha\vdash K} \Rightarrow
-\setsem{\emptyset;\alpha \vdash K}$,\, $\rho \phi =
-\graph{\mathit{fold}_H\, B}$,\, $\pi_1 \rho \phi = \mu H$,\, $\pi_2
-\rho \phi = \setsem{\emptyset;\alpha\vdash K}$,\, $\rho \phi :
-\rel(\pi_1 \rho \phi, \pi_2 \rho \phi)$,\, $A : \setsem{\phi;
-  \emptyset \vdash \Nat^\alpha F \, (\phi \alpha)} (\pi_1 \rho)$,\,
-and $B : \setsem{\phi; \emptyset \vdash \Nat^\alpha F \, (\phi
-  \alpha)} (\pi_2 \rho)$.
-%\[\begin{array}{lll}
-%A &=& \mathit{in}_H :  H (\mu H) \Rightarrow \mu H\\
-%B & : & H\setsem{\emptyset;\alpha\vdash K} \Rightarrow
-%\setsem{\emptyset;\alpha \vdash K}\\
-%\rho \phi &=& \graph{\mathit{fold}_H\, B} \hspace*{0.2in} {
-%  \color{red} \mbox{a graph of a natural transformation, defined in
-%    Enrico's notes}} 
-%\end{array}\]
-%(Note that all the types here are well-formed.) This gives
-%\[\begin{array}{lll}
-%\pi_1 \rho \phi &=& \mu H\\
-%\pi_2 \rho \phi &=& \setsem{\emptyset;\alpha\vdash K}\\
-%\rho \phi &:& \rel(\pi_1 \rho \phi, \pi_2 \rho \phi)\\
-%A & : & \setsem{\phi; \emptyset \vdash \Nat^\alpha
-%  F \, (\phi \alpha)} (\pi_1 \rho)\\
-%B & : & \setsem{\phi; \emptyset \vdash \Nat^\alpha
-%  F \, (\phi \alpha)} (\pi_2 \rho)\\
-%\end{array}\]
-%since
-Equation~\ref{thm:demotion-objects} ensures that $A = \mathit{in}_H :
-H(\mu H) \Rightarrow \mu H = \setsem{\phi;\emptyset \vdash \Nat^\alpha
-  F \,(\phi\alpha)}(\pi_1\rho)$,
-%\[\begin{array}{lll}
-%A \; = \; \mathit{in}_H & : & H(\mu H) \Rightarrow \mu H\\
-%& = & \setsem{\emptyset; \phi, \alpha \vdash F}[\phi := \mu
-%  \setsem{\emptyset; \phi, \alpha \vdash F}] \Rightarrow \mu
-%  \setsem{\emptyset; \phi, \alpha \vdash F}\\  
-%& = & \setsem{\emptyset; \phi, \alpha \vdash F}(\pi_1\rho)
-%  \Rightarrow \setsem{\emptyset; \phi,\alpha \vdash \phi\alpha}(\pi_1 \rho)\\
-%& = & \setsem{\phi; \alpha \vdash F}(\pi_1\rho)
-%  \Rightarrow \setsem{\phi;\alpha \vdash \phi\alpha}(\pi_1
-%  \rho) \hspace*{0.25in} {\color{red} \mbox{Daniel's trick; now a theorem}}\\
-%& = & \setsem{\phi;\emptyset \vdash \Nat^\alpha F
-%    \,(\phi\alpha)}(\pi_1\rho) 
-%\end{array}\]
-and Equation~\ref{thm:demotion-objects} and Lemma~\ref{lem:graph}
-together give that
-\[\begin{array}{lll}
-(A,B) \,=\, (\mathit{in}_H,B) & \in & \relsem{\phi;\emptyset\vdash
-  \Nat^\alpha F\, (\phi\alpha)}\rho\\
-% & = & \lambda A. \relsem{\phi;\alpha\vdash F}\rho[\alpha := A]
-%\Rightarrow \lambda A. (\rho\phi) A\\
-& = & \lambda A. \relsem{\phi;\alpha\vdash F}[\phi :=
-  \graph{\mathit{fold}_H\, B}][\alpha := A] \Rightarrow 
- \graph{\mathit{fold}_H\, B} \\ 
-%& = & \lambda A. \relsem{\emptyset;\phi,\alpha\vdash F}[\phi :=
-%  \graph{\mathit{fold}_H\, B}][\alpha := A] \Rightarrow 
-%\graph{\mathit{fold}_H\, B}  \hspace*{0.25in} {\color{red}
-%  \mbox{Daniel's trick; now a theorem}}\\  
-& = & \relsem{\emptyset;\phi,\alpha\vdash F}
-  \graph{\mathit{fold}_H\, B} \Rightarrow \graph{\mathit{fold}_H\,
-    B}\\
-  & = & \graph{\setsem{\emptyset;\phi,\alpha\vdash F}
-    \,(\mathit{fold}_H\,B)} \Rightarrow \graph{\mathit{fold}_H\, B}\\
-%\hspace*{0.25in} {\color{red} \mbox{Graph Lemma}}\\
-  & = & \graph{\mathit{map}_H \,(\mathit{fold}_H\,B)} \Rightarrow
-\graph{\mathit{fold}_H\, B}\\
-\end{array}\]
-since if $(x,y) \in \graph{\mathit{map}_H \,(\mathit{fold}_H\,B)}$,
-%i.e., if $\mathit{map}_H \,(\mathit{fold}_H\,B) \, x = y$,
-then $\mathit{fold}_H\, B\, (\mathit{in}_H\,x) = B\,y = B\,
-(\mathit{map}_H \,(\mathit{fold}_H\,B) \, x)$ by the definition of
-$\mathit{fold}_H$ as a (indeed, the unique) morphism from
-$\mathit{in}_H$ to $B$.  Thus, $(G \,(\pi_1 \rho)\, A, G\, (\pi_2
-\rho)\, B) \in \graph{\mathit{fold}_H\, B}$, i.e., $\mathit{fold}_H \,
-B \, (G\, (\pi_1 \rho) \, \mathit{in}_H) = G\,(\pi_2 \rho)\,B$.  But
-since $\phi$ is the only free variable in $G$, this simplifies to
-$\mathit{fold}_H\, B \, (G\, \mu H\, \mathit{in}_H) =
-G\,\setsem{\emptyset;\alpha\vdash K}\,B$. 
-\end{proof}
-
-\vspace*{-0.08in}
-
-As in~\cite{jg10}, replacing $\onet$ with any type $\emptyset;\alpha
-\vdash C$ generalizes
-%with $\setsem{\emptyset;\alpha \vdash c}\rho = C$
-Theorem~\ref{thm:short-cut-nested} to deliver more general a free
-theorem whose conclusion is $\mathit{fold}_{H}\, B \; \circ \; G\; \mu
-H \; \mathit{in}_{H} = G \,\setsem{\emptyset;\alpha \vdash K}\, B$.
-Although it is standard to prove that the parametric model constructed
-verifies the existence of initial algebras, this is unnecessary here
-since initial algebras are built directly into our model.
-
-\section{Conclusion and Directions for Future Work}\label{sec:conclusion}
-
-We have constructed a parametric model for a calculus
-providing primitives for constructing nested types directly as
-fixpoints.
-%rather than representing them via their Church encodings. We
-We have also used the Abstraction Theorem for this model to derive
-free theorems in the presence of nested types. This was not possible
-before~\cite{jp19} because the nested types in our calculus were not
-previously known to have well-defined interpretations in locally
-finitely presentable categories like $\set$ and $\rel$. No calculus
-for terms of these types existed, either. The key to obtaining our
-parametric model is
-%{\color{blue} adjusting our type syntax to respect the accessibility
-%  cardinal for the semantic category and}
-the delicate threading of functoriality and its accompanying
-naturality conditions throughout the model construction.
-
-Even when nested types have well-defined interpretations in a semantic
-category, the types of standard functions over them may not. This is
-because all recursion in our calculus must be coded using standard
-folds.  Since folds for nested types can only express computations
-that return natural transformations, their expressivity has long been
-a vexing issue, and this is naturally inherited by our calculus.
-%indeed, such folds
-%Our calculus naturally inherits these constraints.
-%Of course our calculus inherits the constraints
-%standard folds for nested types impose, and thus is perhaps not as
-%expressive as we would ideally like.
-In addition, while there are many papers on higher-kinded types, we
-know of none that considers expressibility issues arising from the
-local $\lambda$-presentability cardinal $\lambda$ of the categories in
-which higher-kinded types are interpreted. Because $\lambda = \omega$
-for $\set$ and $\rel$, the expressivity of our
-%$\set$/$\rel$-based parametric model
-calculus is, unfortunately, not easily improved there. Construction of
-a $\set$/$\rel$-based parametric model for our calculus is
-nevertheless worthwhile.  For one thing, the expressivity issue is
-orthogonal to constructing parametric models,
-%in the presence of nested types built from primitives,
-and trying to address both in a single paper would
-detract from the conceptual clarity of our model construction.
-%our current focus on deriving principled parametricity
-%results for practical programming languages, such as Haskell and Agda,
-%that directly construct nested types as fixpoints.
-%
-%conceptual clarity of this paper.
-For another, if the construction in this paper
-%we couldn't
-%construct a $\set$/$\rel$-based parametric model for the calculus
-%presented here,
-weren't possible, then trying to construct a parametric model from
-richer semantic categories for fancier data types and/or recursion
-schemes would be futile. We naturally expect the construction reported
-here to guide its generalization to the locally $\lambda$-presentable
-categories for $\lambda > \omega$ that will be needed to accommodate
-more general data types and recursion schemes for them.
-
-% As is, the results reported here provide a
-%blueprint for deriving principled parametricity results for practical
-%programming languages, such as Haskell and Agda, that directly
-%construct nested types as fixpoints.
-
-%{\color{blue} Perhaps surprisingly, the accessibility cardinal
-%$\omega$ for the semantic categories $\set$ and $\rel$ also impacts
-%the types of functions (and thus the functions themselves), even over
-%expressible data types, that can be expressed. Although there are
-%many papers on higher-kinded types, no other work that we know of
-%even considers expressibility issues resulting from the semantic
-%settings in which they are interpreted.}
-%{\color{blue} In light of the aforementioned
-%  restrictions imposed by the accessibility cardinal of $\set$ and
-%  $\rel$, this paper is perhaps best read as a proof-of-concept for
-%  constructing parametric models in the locally finitely presentable
-%  categories $\set$ and $\rel$ for calculi that construct nested types
-%  and other advanced data types as fixed points (rather than
-%  representing them via their Church encodings) and that support
-%  structured recursion over them. If we couldn't construct such a
-%  model even in this restricted setting, then there'd be no point in
-%  trying to do so for more robust categories, data types, and/or
-%  recursion schemes. Moreover, our constructions in this simplest
-%  possible setting will guide their generalizations to the locally
-%  $\lambda$-presentable categories fopr $\lambda > \omega$ required to
-%  handle more general data types, structured recursive functions over
-%  then, and more general fixpoints.}
-
-%But since this expressivity issue is orthogonal to constructing
-%parametric models in the presence of nested types built from
-%primitives, and since it would detract from the conceptual clarity of
-%our presentation, we have deliberately chosen not to address it in
-%this paper. (However, we discuss some ideas for increasing
-%expressivity in the next paragraph.) Despite the limitations of
-%standard folds, our results provide a blueprint for deriving
-%principled parametricity results for practical programming languages,
-%such as Haskell and Agda, that directly construct nested types as
-%fixpoints.
-
-%All recursion in our calculus must be coded using folds. Expressivity
-%of standard folds for nested types has long been a vexing issue,
-%%indeed, such folds
-%since they can only express computations that return natural
-%transformations; our calculus naturally inherits these constraints.
-%%Of course our calculus inherits the constraints
-%%standard folds for nested types impose, and thus is perhaps not as
-%%expressive as we would ideally like.
-%But since this expressivity issue
-%is orthogonal to constructing parametric models in the presence of
-%nested types built from primitives, and since it would detract from
-%the conceptual clarity of our presentation, we have deliberately
-%chosen not to address it in this paper. (However, we discuss some
-%ideas for increasing expressivity in the next paragraph.) Despite the
-%limitations of standard folds, our results provide a blueprint for
-%deriving principled parametricity results for practical programming
-%languages, such as Haskell and Agda, that directly construct nested
-%types as fixpoints.
-
-%Many extensions of standard folds have been
-%proposed. Perhaps the most significant of these are
-Generalized folds~\cite{bp99} are an extension of standard folds that
-%and Mendler iterators~\cite{amu05}.
-%which
-can express computations whose results are not natural
-transformations. These were shown in~\cite{jg10} to be equivalent to
-standard folds in the presence of object-level {\em right} Kan
-extensions.
-%; interpreting such
-%a construct in $\set$ and $\rel$ should be unproblematic {\color{blue}
-%  since these categories have all(?) limits}.
-It was also shown in~\cite{jg08} that GADTs can be represented using
-object-level {\em left} Kan extensions, and
-%
-%This suggests
-%extending our term calculus with such object-level constructs.
-%
-%{\color{blue} We may need to impose restrictions inspired
-%  by those for $\mu$-types on LKEs.}
-adding a carefully designed left Kan extension construct to our
-calculus was shown in~\cite{jp19} to preserve the cocontinuity needed
-for GADTs to have well-defined interpretations. (Note, however, that
-%not necessarily in
-%However, this will not, in general, be
-%the case in locally finitely presentable categories. Indeed, as shown
-%in~\cite{jp19},
-locally $\lambda$-presentable categories for $\lambda > \aleph_1$ are
-required to interpret even commonly used GADTs.) This suggests
-extending our calculus with Kan extension constructs and
-%augmenting our type formation rules with such a construct.
-%The constructions of this paper should all be possible
-carrying out our model construction in a locally $\lambda$-presentable
-cartesian closed category (lpccc) $\cal C$ whose category of
-(abstract) relations, obtained by pullback as in~\cite{jac99},
-is also a lpccc and appropriately fibred over $\cal C$.  This would
-%extend the results presented here along the lines of~\cite{gjfor15},
-%thus giving
-give a framework for constructing parametric models for calculi
-with nested types built from primitives that is based on locally
-$\lambda$-presentable fibrations, for some appropriate definition
-thereof.  Another option is to add term-level fixpoints
-%, as in~\cite{pit00},
-to the calculus presented here. Of course, this would require the
-categories interpreting types to be not just locally
-$\lambda$-presentable, but to support some kind of domain structure as
-well.
-%Unfortunately, $\omega$-CPO, the most obvious category
-%of domains to replace $\set$, is not locally finitely presentable.
-
-\pagebreak
-
-\noindent
-
-%Won't print in anonymous mode.
-\begin{acks}
-We thank Andrew Polonsky for valuable feedback and fruitful
-discussions. Supported by NSF awards CCR-1906388 and 1420175.
-\end{acks}
 
@@
 \end{document}
-
-fixpoints at term level ala Pitts:
-
-I'm not sure if we actually want to do this or not (because not all
-fixpoints will necessarily exist, and this will have consequences, and
-because of the restriction to strict functions it will entail and the
-fact that I haven't yet thought about how those will thread through
-our calculus), but another idea we might consider is adding fixpoints
-at the term level. This is done for System F in Wadler's section
-7. [It is also done in Pitts' Parametric Polymorphism and Operational
-  Equivalence paper, where the term introduction rule is SEE SOURCE
-
-%Gamma \vdash F : \tau \to \tau
-%-------------------------------
-%Gamma \vdash fix (F) : tau
-
-But of course the model there is entirely operational rather than
-categorical.]
-
-On the other hand, check out the last paragraph of Wadler's paper, which
-suggests that (at least in 1989) there was some interest in calculi like
-ours:
-
-"The desire to derive theorems from types therefore suggests that it would be
-valuable to explore programming languages that prohibit recursion or only
-allow its restricted use. In theory, this is well understood; we have already
-noted that any computable function that is provably total in second-order
-Peano arithmetic can be defined in the pure polymorphic lambda calculus,
-without using the fixpoint as a primitive. However, practical languages based
-on this notion remain terra incognita."
-
-Of course we are not working in System F, but perhaps the same desire applies
-here, and we do get the same kinds of theorems as Wadler, many just as a
-consequence of our interpretation, without invoking parametricity.
-
-In any case, I guess Wadler could be cited as justifying restricting
-our attention to our calculus since it could be seen as a way to have
-a practical language that ensures that all programs are
-terminating. This has been considered in the past, e.g., in ADL,
-Charity, PolyP.
-
-TERMINATING RECURSION
-Another possible solution to the expressivity issue is extend folds to
-track of deep structure of nested types as in~\cite{jp19}, so that
-specialization of fixpoint can be computed as fixpoint of
-specialization.
-
-Add more polymorphism (all foralls), even though most free theorems
-only use one level (or maybe two, like short cut).
-
-%test

### 2020-07-10 20-gadts/inf-rules.tex
modified Sections 1 and 2

@@
 \renewcommand{\greyout}[1]{} %{{\color{gray} {#1}}} -- toggle to remove greyed text
-\newcommand{\new}[1]{\color{blue}#1}
+
+\newcommand{\new}[1]{{\color{blue} {#1}}}
 
@@
 
-\section{Typing Rules}\label{sec:types}
+\new{Maybe develop our theory for {\em any} $\lambda \geq \omega_1$,
+  and then specialize to $\omega_1$ when discussing GADTs? Can we do
+  that? It seems we really use properties of $\oCPO$ to get that
+  interpretations of $\Nat$-types are well-defined.}
 
+\section{The Calculus}
+
+\subsection{Types}\label{sec:types}
+
+For each $k \ge 0$, we assume countable sets $\tvars^k$ of \emph{type
+  constructor variables of arity $k$} and $\fvars^k$ of
+\emph{functorial variables of arity $k$}, all mutually disjoint.
+%disjoint for distinct $k$ and disjoint from each other.
+The sets of all type constructor variables and functorial variables
+are $\tvars = \bigcup_{k \ge 0} \tvars^k$ and $\fvars = \bigcup_{k \ge
+  0} \fvars^k$, respectively, and a \emph{type variable} is any
+element of $\tvars \cup \fvars$.  We use lower case Greek letters for
+type variables, writing $\phi^k$ to indicate that $\phi \in \tvars^k
+\cup \fvars^k$, and omitting the arity indicator $k$ when convenient,
+unimportant, or clear from context. We reserve letters from the
+beginning of the alphabet to denote type variables of arity $0$, i.e.,
+elements of $\tvars^0 \cup \fvars^0$. We write $\overline{\zeta}$ for
+either a set $\{\zeta_1,...,\zeta_n\}$ of type constructor variables
+or a set of functorial variables when the cardinality $n$ of the set
+is unimportant or clear from context. If $\Pos\,$ is a set of type
+variables we write $\Pos, \overline{\phi}$ for $\Pos\, \cup
+\overline{\phi}$ when $\Pos\, \cap \overline{\phi} = \emptyset$.  We
+omit the vector notation for a singleton set, thus writing $\phi$,
+instead of $\overline{\phi}$, for $\{\phi\}$.
 \begin{dfn}
+Let $V$ be a finite subset of\, $\tvars$, $\Pos$ be a finite
+subset of\, $\fvars$, $\overline{\alpha}$ be a finite subset of\,
+$\fvars^0$ disjoint from $\Pos$, and $\phi^k \in \fvars^k
+\setminus \Pos$.  The set $\mcF^\Pos(V)$ of {\em functorial
+  expressions} over $\Pos$ and $V$ are given by
+\begin{align*}
+  \mcF^\Pos(V) \; ::= \; 
+  & \hspace*{0.15in}
+\zerot \mid \onet 
+\mid \Nat^\Pos \, \mcF^\Pos (V) \;
+\mcF^\Pos (V) 
+\mid \Pos\; \ol{\mcF^\Pos(V)}  \,
+\mid V\, \ol{\mcF^\Pos(V)}  
+\mid \mcF^{\Pos}(V) + \mcF^\Pos(V)\\
+&\mid \mcF^{\Pos}(V) \times \mcF^\Pos(V) \mid \left(\mu
+\phi^{~k}. \lambda \alpha_1...\alpha_k.  
+\mcF^{\Pos,\alpha_1,...,\alpha_k,\phi}(V)\right)
+\ol{\mcF^{\Pos}(V)}\\
+&\mid (\Lan^{\ol\alpha}_{\ol{\F^{\ol\alpha}}} \F^{\Pos,\ol\alpha})\ol{\F^\Pos}
+\end{align*}
+\end{dfn}
+\noindent
+A \emph{type} over $\Pos$ and $V$ is any element of $\F^\Pos(V)$. The
+difference with~\cite{jgj20} here lies solely in the incorporation of
+functorial expressions constructed from $\Lan$.
+
+The notation for types entails that an application $F F_1...F_k$ is
+allowed only when $F$ is a type variable of arity $k$, or $F$ is a
+subexpression of the form $\mu \phi^{k}.\lambda
+\alpha_1...\alpha_k.F'$ or $\Lan^{\ol\alpha}_{\ol K}\, F'$. Moreover,
+if $F$ has arity $k$ then $F$ must be applied to exactly $k$
+arguments.  Accordingly, an overbar indicates a sequence of
+subexpressions whose length matches the arity of the type applied to
+it.  The fact that types are always in \emph{$\eta$-long normal form}
+avoids having to consider $\beta$-conversion at the level of types. In
+a subexpression $\Nat^\Phi F\,G$, the $\Nat$ operator binds all
+occurrences of the variables in $\Phi$ in $F$ and $G$. Note that, by
+contrast with~\cite{jgj20}, variables of arity greater than $0$ are
+allowed in $\Phi$; this is necessary to construct well-typed terms of
+$\Lan$ types. In a subexpression $\mu \phi^k.\lambda \ol{\alpha}.F$,
+the $\mu$ operator binds all occurrences of the variable $\phi$, and
+the $\lambda$ operator binds all occurrences of the variables in
+$\ol{\alpha}$, in the body $F$. And in a subexpression
+$(\Lan^{\ol\alpha}_{\ol K}\, F)\ol A$, the $\Lan$ operator binds all
+occurrences of the variables in $\ol\alpha$ in every element of $\ol
+K$, as well as in $F$.
+
+A {\em type constructor context} is a finite set $\Gamma$ of type
+constructor variables, and a {\em functorial context} is a finite set
+$\Phi$ of functorial variables. In Definition~\ref{def:wftypes}, a
+judgment of the form $\Gamma;\Phi \vdash F$ indicates that the
+type $F$ is intended to be functorial in the variables in $\Phi$
+but not necessarily in those in $\Gamma$.
+
+\begin{dfn}\label{def:wftypes}
 The formation rules for the set $\F \subseteq \bigcup_{V \subseteq
@@
 \[\begin{array}{c}
-\new{\AXC{$\Gamma;\Phi \vdash F$}
+\AXC{$\Gamma;\Phi \vdash F$}
 \AXC{$\Gamma;\Phi  \vdash G$}
 \BIC{$\Gamma;\emptyset \vdash \Nat^\Phi F \,G$}
-\DisplayProof}
+\DisplayProof
 \\[4ex]
@@
 \\[5ex]
-\new{
 \AXC{$\Gamma; \Phi,\ol\alpha^0 \vdash F$}
@@
 \TIC{$\Gamma;\Phi \vdash (\Lan^{\ol{\alpha^0}}_{\ol K} F)\, \ol A$}
-\DisplayProof}
+\DisplayProof
 \end{array}\]
 \end{dfn}
+
+In addition to textual replacement, we also have a proper substitution
+operation on types. If $F$ is a type over $P$ and $V$, if $\Pos$
+and $V$ contain only type variables of arity $0$, and if $k=0$ for
+every occurrence of $\phi^k$ bound by $\mu$ in $F$, then we say
+that $F$ is {\em first-order}; otherwise we say that $F$ is {\em
+  second-order}.  Substitution for first-order types is the usual
+capture-avoiding textual substitution. We write $F[\alpha :=
+  \sigma]$ for the result of substituting $\sigma$ for $\alpha$ in
+$F$, and $F[\alpha_1 := F_1,...,\alpha_k := F_k]$, or
+$F[\ol{\alpha := F}]$ when convenient, for $F[\alpha_1 :=
+  F_1][\alpha_2 := F_2,...,\alpha_k := F_k]$. Substitution
+for second-order types is defined below, where we adopt a similar
+notational convention for vectors of types. Note that it is not
+correct to substitute along non-functorial variables.
+
+\begin{dfn}\label{def:second-order-subst}
+If \,$\Gamma; \Phi,\phi^k \vdash H$
+%with $k \geq 1$,
+and if\, $\Gamma;\Phi, \ol{\alpha} \vdash F$ with
+$|\ol\alpha| = k$, then $\Gamma;\Phi \vdash H[\phi :=_{\ol{\alpha}}
+  F]$.  Similarly, if \,$\Gamma, \phi^k; \Phi \vdash H$,
+%with $k \geq 1$,
+and if\, $\Gamma; \ol\psi,\ol{\alpha} \vdash F$ with $|\ol\alpha| = k$
+and $\Phi \cap \ol\psi = \emptyset$, then $\Gamma,\ol\psi';\Phi \vdash
+H[\phi :=_{\ol{\alpha}} F[\ol{\psi :== \psi'}]]$. Here, the operation
+  $(\cdot)[\phi :=_{\ol \alpha} F]$ of {\em second-order type
+    substitution along $\ol\alpha$} is defined by:
+\[\begin{array}{lll}
+\zerot[\phi :=_{\ol{\alpha}} F] & = & \zerot\\[0.5ex]
+\onet[\phi :=_{\ol{\alpha}} F] & = & \onet\\[0.25ex]
+(\Nat^{\ol\beta} G \,K)[\phi :=_{\ol{\alpha}} F]
+& = & \Nat^{\ol\beta}\, (G[\phi :=_{\ol{\alpha}} F]) \,(K[\phi
+  :=_{\ol{\alpha}} F])\\
+(\psi\ol{G})[\phi :=_{\ol{\alpha}} F] & = &
+\left\{\begin{array}{ll}
+\psi \,\ol{G[\phi :=_{\ol{\alpha}} F]} & \mbox{if } \psi \not = \phi\\
+  F[\ol{\alpha  := G[\phi :=_{\ol{\alpha}} F]}] 
+  & \mbox{if } \psi = \phi
+\end{array}\right.\\[2.8ex]
+(G + K)[\phi :=_{\ol{\alpha}} F] & = & G[\phi
+  :=_{\ol{\alpha}} F] + K[\phi :=_{\ol{\alpha}} F]\\[0.5ex] 
+(G \times K)[\phi :=_{\ol{\alpha}} F] & = &
+G[\phi :=_{\ol{\alpha}} F] \times K[\phi
+  :=_{\ol{\alpha}} F]\\[0.5ex]   
+((\mu \psi. \lambda \ol{\beta}.\, G)\ol{K})[\phi :=_{\ol{\alpha}}
+  F] & = & (\mu \psi. \lambda \ol{\beta}. \,G[\phi :=_{\ol{\alpha}}
+  F])\, \ol{K[\phi :=_{\ol{\alpha}} F]}\\[0.5ex]
+((\Lan^{\ol{\beta}}_{\ol H}\, G)\ol{K})[\phi :=_{\ol{\alpha}}
+  F] & = & (\Lan^{\ol\beta}_{\ol H} \,G[\phi :=_{\ol{\alpha}}
+  F])\, \ol{K[\phi :=_{\ol{\alpha}} F]}
+\end{array}\]
+\end{dfn}
+\noindent
+We note that $(\cdot)[\phi^0 :=_\emptyset F]$ coincides with
+first-order substitution. We also omit $\ol\alpha$ when convenient.
+
+\subsection{Terms}\label{sec:terms}
+
+We now define our term calculus. To do so we assume an infinite set
+$\cal V$ of term variables disjoint from $\tvars$ and $\fvars$. If
+$\Gamma$ is a type constructor context and $\Phi$ is a functorial
+context, then a {\em term context for $\Gamma$ and $\Phi$} is a finite
+set of bindings of the form $x : F$, where $x \in {\cal V}$ and
+$\Gamma; \Phi \vdash F$. We adopt the same conventions for denoting
+disjoint unions and for vectors in term contexts as for type
+constructor contexts and functorial contexts.
 
@@
 \[\begin{array}{c}
-\new{
 \AXC{$\Gamma; \Phi \vdash F$}
@@
   \,|\, \Delta \vdash L_\Phi x.t : \Nat^\Phi \,F \,G$}
-\DisplayProof}
+\DisplayProof
 \\\\
-\new{\AXC{$\ol{\Gamma;\Phi,\ol\beta \vdash K}$}
+\AXC{$\ol{\Gamma;\Phi,\ol\beta \vdash K}$}
 \AXC{$\Gamma; \emptyset
@@
   G[\overline{\psi :=_{\ol\beta} K}]$}
-\DisplayProof}
+\DisplayProof
 \\\\
-\new{
 \AXC{$\Gamma; \Phi, \ol{\phi} \vdash H$}
@@
       :=_{\ol{\beta}} G}])$} 
-\DisplayProof}
+\DisplayProof
 \\\\
-\new{\AXC{$\Gamma; \Phi, \phi, \ol{\alpha} \vdash H$}
+\AXC{$\Gamma; \Phi, \phi, \ol{\alpha} \vdash H$}
 \UIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \tin_H :
@@
   \phi.\lambda \ol{\alpha}.H)\ol{\beta}$}
-\DisplayProof}
+\DisplayProof
 \\\\
-\new{\AXC{$\Gamma; \phi,\Phi,\ol{\alpha} \vdash H$}
+\AXC{$\Gamma; \phi,\Phi,\ol{\alpha} \vdash H$}
 \AXC{$\Gamma; \Phi, \ol{\beta} \vdash F$}
@@
     \,(\mu \phi.\lambda \ol{\alpha}.H)\ol{\beta}\,F)$}
-\DisplayProof}
+\DisplayProof
 \\\\
-\new{
 \AXC{$\Gamma; \Phi,\ol{\alpha}\vdash F$}
@@
 \AXC{$\ol{\Gamma;\Phi \vdash A} \hspace*{0.3in} \Gamma;\Phi~|~\Delta \vdash t : F[\ol{\alpha :=A}]$}
-\TIC{$\Gamma;\Phi~|~\Delta \vdash \int_{K,F} t :
+\TIC{$\Gamma;\Phi~|~\Delta \vdash \int_{\ol K,F} t :
   (\Lan^{\ol\alpha}_{\ol K}\,  F)\,\ol{K[\ol{\alpha := A}]}$}  
-\DisplayProof}
+\DisplayProof
 \\\\
-\new{
 \AXC{$\Gamma;\emptyset~|~\Delta \vdash \eta : \Nat^{\Phi,\ol\alpha}\,
@@
   G[\ol{\beta := A}]$}
-\DisplayProof}
+\DisplayProof
 \end{array}\]
@@
 
-Things to check:
-\begin{itemize}
-\item \new{Are these $\Nat$-types $\omega$CPOs? An argument similar to the
-  one in the POPL submission should hold.}
-\item \new{Does IEL hold if we allow functoriality in variables of arity $> 0$?}
-\item \new{Does AT hold if we allow functoriality in variables of arity $> 0$?}
-\end{itemize}
+In the rule for $L_{\ol{\alpha}}x.t$, the $L$ operator binds all
+occurrences of the type variables in $\ol{\alpha}$ in the type of the
+term variable $x$ and in the body $t$, as well as all occurrences of
+$x$ in $t$. In the rule for $t_{\ol K} s$ there is one functorial
+expression in $\ol K$ for every functorial variable in $\ol
+\alpha$. In the rule for $\map^{\ol{F},\ol{G}}_H$ there is one
+functorial expression $F$ and one functorial expression $G$ for each
+functorial variable in $\ol\phi$. Moreover, for each $\phi^k$ in
+$\ol\phi$ the number of functorial variables in $\ol\beta$ in the
+judgments for its corresponding functorial expresssions $F$ and $G$ is
+$k$. In the rules for $\tin_H$ and $\fold^F_H$, the functorial
+variables in $\ol{\beta}$ are fresh with respect to $H$, and there is
+one $\beta$ for every $\alpha$. (Recall from above that, in order for
+the types of $\tin_H$ and $\fold^F_H$ to be well-formed, the length of
+$\alpha$ must equal the arity of $\phi$.) In the rule for $\int_{\ol
+  K,H}\,t$, there is one functorial expression $A$ for every
+functorial variable in $\ol \alpha$, and in the rule for
+$\partial^{G,\ol K}_F\,\eta\,t$, there is one functorial expression
+$A$ for every functorial expression in $\ol K$ (and hence for every
+functorial variable in $\ol \beta$).
 
-Sum and product intro and elim rules should be annotated with
-  constituent types for consistency.
+Substitution for terms is the obvious extension
+of the usual capture-avoiding textual substitution, and
+Definition~\ref{def:well-formed-terms} ensures that the expected
+weakening rules for well-formed terms hold.
+
+\vspace*{0.2in}
+
+\new{Sum and product intro and elim rules should be annotated with
+  constituent types for consistency?
 
@@
 \end{array}\]
+This will appear as a computational property of the term
+interpretations.}
 
+\section{Interpreting Types}\label{sec:type-interp}
+
+The fundamental idea underlying Reynolds' parametricity is to give
+each type $F(\alpha)$ with one free variable $\alpha$ both an {\em
+  object interpretation} $F_0$ taking sets to sets and a
+\emph{relational interpretation} $F_1$ taking relations $R :
+\rel(A,B)$ to relations $F_1 (R) : \rel(F_0 (A), F_0 (B))$, and to
+interpret each term $t(\alpha,x) : F(\alpha)$ with one free term
+variable $x : G(\alpha)$ as a map $t_0$ associating to each set $A$ a
+homomorphism $t_0(A) : G_0(A) \to F_0(A)$, and to each relation $R$ a
+morphism $t_1(R) : G_1(R) \to F_1(R)$.  These interpretations are to
+be given inductively on the structures of $F$ and $t$ in such a way
+that they imply two fundamental theorems. The first is an
+\emph{Identity Extension Lemma}, which states that $F_1(\Eq_A) =
+\Eq_{F_0(A)}$, and is the essential property that makes a model
+relationally parametric rather than just induced by a logical
+relation. The second is an \emph{Abstraction Theorem}, which states
+that, for any $R :\rel(A, B)$, $(t_0(A),t_0(B))$ is a morphism in
+$\rel$ from $(G_0(A),G_0(B),G_1(R))$ to $(F_0(A),F_0(B),F_1(R))$. The
+Identity Extension Lemma is similar to the Abstraction Theorem except
+that it holds for {\em all} elements of a type's interpretation, not
+just those that are interpretations of terms.  Similar theorems are
+expected to hold for types and terms with any number of free
+variables.
+
+To accommodate GADTs, we will need to transition Reynolds' approach
+from a $\set$-based semantics to a semantics based on
+$\omega$-complete partial orders. We denote the category of
+$\omega$-complete partial orders ($\oCPO$s) and their sup-preserving
+morphisms by $\omega$CPO. The underlying set of an $\omega$-CPO $A$ is
+denoted $|A|$.  The category $\oCPOR$ of relations on $\omega$CPOs has
+as its objects triples $(A,B,R)$, where $A, B: \oCPO$ and $R:
+\rel(|A|, |B|)$, and has as its morphisms from $(A,B,R)$ to
+$(A',B',R')$ pairs $(f : A \to A',g : B \to B')$ of morphisms in
+$\oCPO$ such that $(f\,a, g\,b) \in R'$ whenever $(a,b) \in R$. We
+write $R : \oCPOR(A,B)$ in place of $(A,B,R): \oCPOR$ when
+convenient. If $R : \oCPOR(A,B)$ then we write $\pi_1 R$ and $\pi_2 R$
+for the {\em domain} $A$ of $R$ and the {\em codomain} $B$ of $R$,
+respectively. If $A : \oCPO$, then we write $\Eq_A = (A,A,\Eq_{|A|})$
+for the {\em equality relation} on $A$.
+
+To adapt Reynolds' approach, we first inductively define, for each
+type, an object interpretation in $\oCPO$ and a relational
+interpretation in $\oCPOR$.  Next, we show that these interpretations
+satisfy both an Identity Extension Lemma (Theorem~\ref{thm:iel}) and
+an Abstraction Theorem (Theorem~\ref{thm:gen-at}) appropriate to the
+$\oCPO$ setting. The key to proving our Identity Extension Lemma is a
+familiar ``cutting down'' of the interpretations of universally
+quantified types to include only the ``parametric'' elements; as
+in~\cite{jgj20}, the relevant types of the calculus defined above are
+the (now richer) $\Nat$-types.  The requisite cutting down requires
+that the object interpretations of our types in $\omega$CPO are
+defined simultaneously with their relational interpretations in
+$\oCPOR$.  We give the object interpretations for our types in
+Section~\ref{sec:obj-interp} and give their relational interpretations
+in Section~\ref{sec:rel-interp}.  While the former are relatively
+straightforward, the latter are less so, mainly because of the
+cocontinuity conditions, adapted from the $\set$-based setting
+of~\cite{jgj20}, that must hold if they are to be well-defined. We
+develop these conditions in Section~\ref{sec:rel-interp}, which
+separates Definitions~\ref{def:set-sem} and~\ref{def:rel-sem} in
+space, but otherwise has no impact on the fact that they are given by
+mutual induction.
+
+\subsection{Object Interpretations of Types}\label{sec:obj-interp}
+
+The object interpretations of the types in our calculus will be
+$\omega_1$-cocontinuous functors between categories of
+$\omega_1$-cocontinuous functors on categories constructed from the
+locally $\omega_1$-presentable category $\oCPO$.  We therefore begin
+by recording some important facts about locally $\omega_1$-presentable
+categories and functors on them, and verifying verify the properties
+needed to interpret our syntax.
+
+\subsubsection{Preliminaries}
+
+\new{Perhaps have as preliminaries to entire paper. Do everything for
+  $\lambda CPOs$? Define these; investigate their properties.}
+
+A category is \emph{small} if its collection of morphisms is a set.
+It is \emph{locally small} if, for any two objects $A$ and $B$, the
+collection of morphisms from $A$ to $B$ is a set. A {\em small
+  (co)limit} in a category $\mcC$ is a (co)limit of a diagram $F :
+\mcA \to \mcC$, where $\mcA$ is a small category. A category $\mcC$
+is {\em (co)complete} if it has all small (co)limits.
+
+A poset $\mcD = (D,\leq)$ is \emph{$\omega_1$-directed} if every
+countable subset of $D$ has a supremum. When $\mcD$ is considered
+as a category, we write $d \in \mcD$ to indicate that $d$ is an object
+of $\mcD$ (i.e., $d \in D$). An {\em $\omega_1$-directed colimit} in a
+category $\mcC$ is a colimit of a diagram $F : {\mathcal D} \to \mcC$,
+where $\mathcal D$ is an $\omega_1$-directed poset. A category $\mcC$
+is {\em $\omega_1$-cocomplete} if it has all $\omega_1$-directed
+colimits; a {\em cocomplete} category is one that has all colimits.
+
+If $\mcA$ and $\mcC$ are $\omega_1$-cocomplete, then the functor $F :
+\mcA \to \mcC$ is {\em $\omega_1$-cocontinuous} if it preserves
+$\omega_1$-directed colimits. We write $[\A,\C]_{\omega_1}$ for 
+the category of $\omega_1$-cocontinuous functors from $\A$ to $\C$,
+and $\C^\A$ for the category of {\em all} functors from $\A$ to $\C$.
+Since (co)limits of functors are computed pointwise, $C^\A$ has all
+(co)limits that $\C$ has, and (co)limits of (co)continuous functors
+are again (co)continuous. It follows that $[\A,\C]_{\omega_1}$ is
+($\omega_1$1)(co)complete whenever $\C$ is.
+
+If $\mcA$ is locally small, then an object $A$ of $\mcA$ is {\em
+  $\omega_1$-presentable} if the functor $\mathsf{Hom}_\mcA(A, -) :
+\mcA \to \Set$ preserves $\omega_1$-directed colimits, i.e., if for
+every $\omega_1$-directed poset $\mathcal D$ and every functor $F :
+{\mathcal D} \to \mcC$, there is a canonical isomorphism $\colim{d \in
+  \mcD}{\mathsf{Hom}_\mcA(A,Fd)} \simeq \mathsf{Hom}_{\mcA}(A,
+\colim{d \in \mcD}{Fd})$. A locally small category $\mcA$ is {\em
+  $\omega_1$-accessible} if it is $\omega_1$-cocomplete and has a set
+$\mcA_0$ of $\omega_1$-presentable objects such that every object is
+an $\omega_1$-directed colimit of objects in $\mcA_0$; a locally small
+category is {\em locally $\omega_1$-presentable} if it is
+$\omega_1$-accessible and cocomplete.
+
+The category $\oCPO$ is locally $\omega_1$-presentable (but not
+locally finitely presentable); see Examples~1.18(2)
+of~\cite{ar94}). Its $\omega_1$-presentable objects are precisely the
+$\omega$CPOs cardinality less than $\omega_1$, i.e., the countable
+$\omega$CPOs. In the next subsection we will interpret type variables
+in $\tvars^k \cup \fvars^k$ as elements of
+$[\oCPO^k,\oCPO]_{\omega_1}$; the following special cases of standard
+results (see, e.g.,~\cite{ar94}) will therefore be critical to
+deducing important properties of our object interpretations of types:
+\begin{prop}\label{prop:prelims}
+\begin{enumerate}
+\item If $\C_1,...,\C_n$ are locally $\omega_1$-presentable categories
+  then so is $\C_1 \times ... \times \C_n$. Moreover, the presentable
+  objects of $\C_1 \times ... \times \C_n$ are exactly the tuples of
+  the form $(P_1,...,P_n)$, where, for each $i = 1,...,n$, the object
+  $P_i$ is presentable in $\C_i$.
+\item If $\mathcal{A}$ is $\omega_1$-accessible and $\mathcal{C}$ is
+  $\lambda$-cocomplete, then the category $[\mathcal{A},
+  \mathcal{C}]_{\omega_1}$ is naturally equivalent to the category
+  $\mathcal{C}^{\mathcal{A}_0}$.
+\item If $\mathcal{C}$ is locally $\omega_1$-presentable and
+  $\mathcal{A}_0$ is essentially small, then
+  $\mathcal{C}^{\mathcal{A}_0}$ is locally $\omega_1$-presentable.
+\end{enumerate}
+\end{prop}
+\noindent
+Together, the statements in Proposition~\ref{prop:prelims} give that
+if $\mathcal{A}$ and $\mathcal{C}$ are locally $\omega_1$-presentable,
+then $[\mathcal{A}, \mathcal{C}]_{\omega_1}$ is naturally equivalent
+to $\mathcal{C}^{\mathcal{A}_0}$, and thus is
+$\omega_1$-presentable. Thus, for all $k_1,...,k_n \in \nat^n$,
+$[\mathcal{A}, \mathcal{C}]_{\omega_1}^{k_1} \times ... \times
+[\mathcal{A}, \mathcal{C}]_{\omega_1}^{k_n}$ is locally
+$\omega_1$-presentable, and therefore $[[\mathcal{A},
+    \mathcal{C}]_{\omega_1}^{k_1} \times ... \times [\mathcal{A},
+    \mathcal{C}]_{\omega_1}^{k_n}, \mathcal{C}]_{\omega_1}$ is as
+well. Taking both $\A$ and $C$ to be $\oCPO$ --- as we will to ensure
+that the fixpoints interpreting $\mu$-types in $\oCPO$ exist --- we
+have
+\begin{prop}\label{prop:nat-interps-sets}
+  For all $k_1,...,k_n \in \nat^n$,
+  \[[[\oCPO, \oCPO]_{\omega_1}^{k_1}
+  \times ... \times [\oCPO, \oCPO]_{\omega_1}^{k_n},
+  \oCPO]_{\omega_1}\] is locally $\omega_1$-presentable.
+\end{prop}
+
+\subsubsection{Object Interpretations}
+
+To define the object interpretations of the types in
+Definition~\ref{def:wftypes} we must first interpret their
+variables. We have:
+
+\begin{dfn}\label{def:set-env}
+A {\em $\oCPO$ environment} maps each type variable in $\tvars^k \cup
+\fvars^k$ to an element of $[\oCPO^k,\oCPO]_{\omega_1}$.  A morphism
+$f : \rho \to \rho'$ for set environments $\rho$ and $\rho'$ with
+$\rho|_\tvars = \rho'|_\tvars$ maps each type constructor variable
+$\psi^k \in \tvars$ to the identity natural transformation on $\rho
+\psi^k = \rho'\psi^k$ and each functorial variable $\phi^k \in \fvars$
+to a natural transformation from the $k$-ary functor $\rho \phi^k$ on
+$\oCPO$ to the $k$-ary functor $\rho' \phi^k$ on $\oCPO$.  Composition
+of morphisms on set environments is given componentwise, with the
+identity morphism mapping each set environment to itself. This gives a
+category of set environments and morphisms between them, which we
+denote $\oCPOenv$.
+\end{dfn}
+
+When convenient we identify a functor in $[\oCPO^0, \oCPO]_{\omega_1}$
+with its value on $\ast$ and consider a $\oCPO$ environment to map a
+type variable of arity $0$ to an $\omega_1$-cocontinuous functor from
+$\oCPO^0$ to $\oCPO$, i.e., to an $\oCPO$. If $\Phi =
+\{\phi_1^{k_1},...,\phi_n^{k_n}\}$ and $\ol K = \{K_1,...,K_n\}$ where
+$K_i : [\oCPO^{k_i},\oCPO]_{\omega_1}$ for $i = 1,...,n$, then we
+write $\rho[\ol{\Phi := K}]$ for the $\oCPO$ environment $\rho'$ such
+that $\rho' \phi_i = K_i$ for $i = 1,...,n$ and $\rho' \phi = \rho
+\phi$ if $\phi \not \in \Phi$.  If $\rho$ is an $\oCPO$ environment,
+we write $\Eq_\rho$ for the $\oCPOR$ environment (see
+Definition~\ref{def:reln-env}) such that $\Eq_\rho v = \Eq_{\rho v}$
+for every type variable $v$. The categories $\oCPORT_k$ and relational
+interpretations appearing in the third clause of
+Definition~\ref{def:set-sem} are given in full in Section~\ref{?}.
+
+\begin{dfn}\label{def:set-sem}
+The {\em object interpretation} $\oCPOsem{\cdot} : \F \to [\oCPOenv,
+  \oCPO]_{\omega_1}$ is defined by
+\begin{align*}
+  \oCPOsem{\Gamma;\Phi \vdash \zerot}\rho &= 0\\
+  \oCPOsem{\Gamma;\Phi \vdash \onet}\rho &= 1\\
+  \oCPOsem{\Gamma; \emptyset
+    \vdash \Nat^\Phi
+    \,F\,G}\rho &= \{\eta : \lambda \ol{K}. \,\oCPOsem{\Gamma;
+    \Phi \vdash
+    F}\rho[\ol{\Phi := K}] 
+      \Rightarrow \lambda \ol{K}.\,\oCPOsem{\Gamma; 
+        \Phi \vdash G}\rho[\ol{\Phi := K}] \\ 
+      &\hspace{0.3in}|~\forall \ol{K = (K_0, K_1, K^*) : \oCPORT_k}.\\
+      &\hspace{0.4in}(\eta_{\overline{K_0}}, \eta_{\overline{K_1}})
+      : \oCPORsem{\Gamma; \Phi \vdash F}\Eq_{\rho}[\ol{\Phi := K}]
+      \rightarrow \oCPORsem{\Gamma; \Phi \vdash
+        G}\Eq_{\rho}[\ol{\Phi := K}] \} \\
+  \oCPOsem{\Gamma;\Phi \vdash \phi\ol{F}}\rho &=
+  (\rho\phi)\,\ol{\oCPOsem{\Gamma;\Phi \vdash
+    F}\rho}\\
+  \oCPOsem{\Gamma;\Phi \vdash F+G}\rho &=
+  \oCPOsem{\Gamma;\Phi \vdash F}\rho +
+  \oCPOsem{\Gamma;\Phi \vdash G}\rho\\
+  \oCPOsem{\Gamma;\Phi \vdash F \times G}\rho &=
+  \oCPOsem{\Gamma;\Phi \vdash F}\rho \times
+  \oCPOsem{\Gamma;\Phi \vdash G}\rho\\ 
+  \oCPOsem{\Gamma;\Phi,\ol\gamma \vdash (\mu \phi.\lambda
+    \ol{\alpha}. H)\ol{G}}\rho &= (\mu
+    T^\oCPO_{H,\rho})\ol{\oCPOsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho}\\
+    \text{where } T^\oCPO_{H,\rho}\,F & = \lambda
+  \ol{A}. \oCPOsem{\Gamma;\ol\gamma,\phi, \ol{\alpha} \vdash
+    H}\rho[\phi :=  F][\ol{\alpha := A}]\\
+  \text{and } T^\oCPO_{H,\rho}\,\eta &= \lambda
+  \ol{A}. \oCPOsem{\Gamma;\ol\gamma,\phi, \ol{\alpha} \vdash
+    H}\id_\rho[\phi := \eta][\ol{\alpha := \id_{A}}] \\
+  \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}\rho &=
+    (\textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}
+      \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := \_}])
+    \ol{\oCPOsem{\Gamma; \Phi \vdash A}\rho}
+\end{align*}
+\end{dfn}
+If $\rho \in \oCPOenv$ and $\vdash F$ then we write $\oCPOsem{\vdash F}$
+instead of $\oCPOsem{\vdash F}\rho$ since the environment is
+immaterial. The third clause of Definition~\ref{def:set-sem} does
+indeed define an $\oCPO$. First, we have
+
+\begin{lemma}
+The collection of all natural transformations 
+\[
+\eta : \lambda \ol K. \setsem{\Gamma;\Phi \vdash F}\rho[\ol{\Phi := K}]
+\Rightarrow \lambda \ol K. \setsem{\Gamma;\Phi \vdash G}\rho[\ol{\Phi :=
+  K}]
+\]
+defines a set.
+\end{lemma}
+\begin{proof}
+We first note that $\lambda \ol K. \setsem{\Gamma;\Phi \vdash
+  F}\rho[\ol{\Phi := K}]$ and $\lambda \ol K. \setsem{\Gamma;\Phi
+  \vdash G}\rho[\ol{\Phi := K}]$ are both in $[\ol{[\oCPO^k ,
+      \oCPO]_{\omega_1}}, \oCPO]_{\omega_1}$. By
+Proposition~\ref{prop:nat-interps-sets}, $[\ol{[\oCPO^k ,
+      \oCPO]_{\omega_1}}, \oCPO]_{\omega_1}$ is locally
+$\omega_1$-presentable. It is therefore locally small, so there are
+only $\set$-many morphisms (i.e., natural transformations) between any
+two functors in $[\ol{[\oCPO^k , \oCPO]_{\omega_1}},
+  \oCPO]_{\omega_1}$. In particular, there are only $\set$-many
+natural transformations from $\lambda \ol K. \setsem{\Gamma;\Phi
+  \vdash F}\rho[\ol{\Phi := K}]$ to $\lambda \ol
+K. \setsem{\Gamma;\Phi \vdash G}\rho[\ol{\Phi := K}]$.
+\end{proof}
+
+\new{Need to modify to take equality conditions into account.}
+  If $\C$ is any category, then given functors $F,G : \C \to
+  \oCPO$ and natural transformations $\eta, \eta' : F \to G$, we
+  define $\eta \leq \eta'$ iff $\eta_c \leq \eta'_c$ for all $c \in
+  \C$, i.e., iff $\eta_c\,x \leq \eta'_c\,x$ in $Gc$ for all $c \in
+  \C$ and $x \in Fc$.  If $(\eta_i)_{i < \omega}$ is a chain in
+  $\{\eta : F \to G\}$, then the family of morphisms $(\bigvee_{i <
+    \omega} \eta_i)_c = \lambda x. \bigvee_{i < \omega} (\eta_i)_c\,x
+  : Fc \to Gc$ defines a natural transformation $\bigvee_{i < \omega}
+  \eta_i : F \to G$, and this natural transformation is clearly the
+  supremum of $(\eta_i)_{i < \omega}$ in $\{\eta : F \to G\}$. We
+  therefore have that $\{\eta : F \to G\}$ is itself an $\oCPO$.
+  Letting $F = \lambda \ol K. \setsem{\Gamma;\Phi \vdash
+    F}\rho[\ol{\Phi := K}]$ and $G = \lambda \ol
+  K. \setsem{\Gamma;\Phi \vdash G}\rho[\ol{\Phi := K}]$, we have that
+  $\oCPOsem{\Gamma; \emptyset \vdash \Nat^\Phi F\,G}$ is indeed an
+  $\oCPO$, i.e., is a constantly-valued functor into $\oCPO$.
+
+Interpretations of $\Nat$-types ensure that $\oCPOsem{\Gamma \vdash F
+  \to G}$ and $\oCPOsem{\Gamma \vdash \forall \ol\alpha. F}$ are as
+expected in any parametric model.
+
+To make sense of the last clause in Definition~\ref{def:set-sem}, we
+need to know that, for each $\rho \in \oCPOenv$, $T^\oCPO_{H,\rho}$ is
+an $\omega_1$-cocontinuous endofunctor on $[\oCPO^k,
+  \oCPO]_{\omega_1}$, and thus admits a fixpoint.  Since
+$T_{H,\rho}^\oCPO$ is defined in terms of
+$\oCPOsem{\Gamma;\ol\gamma,\phi, \ol{\alpha} \vdash H}$, this means
+that interpretations of types must be such functors, which in turn
+means that the actions of set interpretations of types on objects and
+on morphisms in $\oCPOenv$ are intertwined. Fortunately, we know
+from~\cite{jp19} that, for every $\Gamma; \Phi \vdash G$,
+$\oCPOsem{\Gamma; \Phi \vdash G}$ is actually in
+$[\ol{[\oCPO^k,\oCPO]_{\omega_1}},\oCPO]_{\omega_1}$, or,
+equivalently, $[\oCPOenv,\oCPO]_{\omega_1}$.  Therefore, for each
+$\oCPOsem{\Gamma; \ol\gamma, \phi^k, \ol{\alpha} \vdash H}$, the
+corresponding operator $T^\oCPO_{H}$ can be extended to a {\em
+  functor} from $\oCPOenv$ to
+$[[\oCPO^k,\oCPO]_{\omega_1},[\oCPO^k,\oCPO]_{\omega_1}]_{\omega_1}$.
+The action of $T^\oCPO_H$ on an object $\rho \in \oCPOenv$ is given by
+the higher-order functor $T_{H,\rho}^\oCPO$, whose actions on objects
+(functors in $[\oCPO^k, \oCPO]_{\omega_1}$) and on morphisms (natural
+transformations between such functors) are given in
+Definition~\ref{def:set-sem}. The action of $T^\oCPO_H$ on a morphism
+$f : \rho \to \rho'$ is the higher-order natural transformation
+$T^\oCPO_{H,f} : T^\oCPO_{H,\rho} \to T^\oCPO_{H,\rho'}$ whose action
+on $F : [\oCPO^k,\oCPO]_{\omega_1}$ is the natural transformation
+$T^\oCPO_{H,f}\, F : T^\oCPO_{H,\rho}\,F \to T^\oCPO_{H,\rho'}\,F$
+whose component at $\ol A$ is $(T^\oCPO_{H,f}\, F)_{\ol A} =
+\oCPOsem{\Gamma; \ol\gamma,\phi,\ol{\alpha} \vdash H}f[\phi :=
+  \id_F][\ol{\alpha := \id_A}]$.
+
+In addition, for each $\ol K$, we have that $\textit{Lan}_{\ol K}$ is
+itself a (higher-order) functor. Specifically, given functors $F, F' :
+\C \to \D$, a sequence of functors $\ol K = K_1, ..., K_n$ with $K_i :
+\C \to \C_i$ for $i = 1,...,n$, and a natural transformation $\alpha :
+F \to G$, the functorial action $\textit{Lan}_{\ol K} \alpha :
+\textit{Lan}_{\ol K} F \to \textit{Lan}_{\ol K} F'$ of
+$\mathit{Lan}_{\ol K}$ on $\alpha$ is defined to be the unique natural
+transformation such that $((\textit{Lan}_{\ol K} \alpha) \,\circ
+\,\langle K_1,...,K_n \rangle) \circ \eta_F = \eta_{F'} \,\circ\,
+\alpha$. Here, $\eta_F : F \to (\textit{Lan}_{\ol K} F) \circ \langle
+K_1,...,K_n \rangle$ and $\eta_{F'} : F' \to (\textit{Lan}_{\ol K} F')
+\circ \langle K_1,...,K_n \rangle$ are the natural transformations
+associated with the functors $\textit{Lan}_{\ol K} F$ and
+$\textit{Lan}_{\ol K} F'$ from $\Pi_{i \in \{1,...,n\}} \C_i$ to $\D$,
+respectively. It is not hard to see that $\mathit{Lan}_{\ol K}$ is a
+(higher-order) functor under this definition.
+
+The next definition uses the functors $T^\oCPO_H$ and $\mathit{Lan}_K$
+to define the actions of functors interpreting types on morphisms
+between set environments.
+
+\begin{dfn}\label{def:set-sem-funcs}
+Let $f: \rho \to \rho'$ be a morphism between $\omega$-CPO environments $\rho$
+and $\rho'$ (so that $\rho|_\tvars = \rho'|_\tvars$). The action
+$\oCPOsem{\Gamma;\Phi \vdash F}f$ of\, $\oCPOsem{\Gamma;\Phi \vdash F}$
+on the morphism $f$ is given as follows:
+\begin{itemize}
+\item If \,$\Gamma;\Phi \vdash \zerot$ then $\oCPOsem{\Gamma;\Phi \vdash
+  \zerot}f = \id_0$
+\item If \,$\Gamma;\Phi \vdash \onet$ then $\oCPOsem{\Gamma;\Phi \vdash
+  \onet}f = \id_1$
+\item If \,$\Gamma; \emptyset
+  \vdash \Nat^\Phi F\,G$ then
+  $\oCPOsem{\Gamma; \emptyset
+    \vdash \Nat^\Phi F\,G} f =
+  \id_{\oCPOsem{\Gamma; \emptyset
+      \vdash \Nat^\Phi F\,G}\rho}$
+\item If \,$\Gamma;\Phi \vdash \phi \ol{F}$ then
+ \begin{multline*}
+ \oCPOsem{\Gamma;\Phi \vdash \phi \ol{F}} f : \oCPOsem{\Gamma;\Phi
+  \vdash \phi \ol{F}}\rho \to \oCPOsem{\Gamma;\Phi \vdash
+  \phi\ol{F}}\rho' \\
+  = (\rho\phi) \ol{\oCPOsem{\Gamma;\Phi \vdash
+  F}\rho} \to (\rho'\phi) \ol{\oCPOsem{\Gamma;\Phi \vdash
+  F}\rho'}
+ \end{multline*}
+  is defined by
+  \[
+  \begin{split}
+  \oCPOsem{\Gamma;\Phi \vdash \phi \ol{F}} f
+  &= (f\phi)_{\ol{\oCPOsem{\Gamma;\Phi \vdash
+      F}\rho'}}\, \circ\, (\rho\phi) {\ol{\oCPOsem{\Gamma;\Phi
+      \vdash F}f}} \\
+  &= (\rho'\phi) {\ol{\oCPOsem{\Gamma;\Phi \vdash
+      F}f}}\, \circ\, (f \phi)_{\ol{\oCPOsem{\Gamma;\Phi \vdash
+      F}\rho}}
+  \end{split}
+  \]
+  The latter equality holds because $\rho\phi$ and
+  $\rho'\phi$ are functors and $f\phi : \rho\phi \to \rho'\phi$ is a
+  natural transformation, so the following naturality square commutes:
+{\footnotesize\begin{equation}\label{eq:cd2}
+\begin{CD}
+  (\rho\phi) \ol{\oCPOsem{\Gamma;\Phi \vdash F}\rho} @> (f\phi)_{
+    \ol{\oCPOsem{\Gamma;\Phi \vdash F}\rho}} >> (\rho'\phi)
+  \ol{\oCPOsem{\Gamma;\Phi \vdash F}\rho} \\ @V(\rho\phi)
+  \ol{\oCPOsem{\Gamma;\Phi \vdash F}f}VV @V (\rho'\phi)
+  \ol{\oCPOsem{\Gamma;\Phi \vdash F}f} VV \\ (\rho\phi)
+  \ol{\oCPOsem{\Gamma;\Phi \vdash F}\rho'} @>(f\phi)_{
+    \ol{\oCPOsem{\Gamma;\Phi \vdash F}\rho'}}>> (\rho'\phi)
+  \ol{\oCPOsem{\Gamma;\Phi \vdash F}\rho'}
+\end{CD}
+\end{equation}}
+\item If\, $\Gamma;\Phi \vdash F + G$ then $\oCPOsem{\Gamma;\Phi
+  \vdash F + G}f$ is defined by
+  \begin{align*}
+  \oCPOsem{\Gamma;\Phi \vdash F + G}f(\inl\,x) 
+  &= \inl\,(\oCPOsem{\Gamma;\Phi \vdash
+  F}f x) \\
+  \oCPOsem{\Gamma;\Phi \vdash F + G}f(\inr\,y)
+  &= \inr\,(\oCPOsem{\Gamma;\Phi \vdash G}f y)
+  \end{align*}
+\item If \,$\Gamma;\Phi\vdash F \times G$ then
+  $\oCPOsem{\Gamma;\Phi \vdash F \times G}f = 
+  \oCPOsem{\Gamma;\Phi \vdash F}f \times \oCPOsem{\Gamma;\Phi \vdash
+    G}f$
+\item If \,$\Gamma;\Phi,\ol\gamma \vdash (\mu \phi.\lambda
+  \ol{\alpha}. H)\ol{G}$ then
+  \[\begin{array}{lll}
+  \oCPOsem{\Gamma;\Phi,\ol\gamma \vdash (\mu  \phi.\lambda
+    \ol{\alpha}. H)\ol{G}} f &: &\oCPOsem{\Gamma;\Phi,\ol\gamma 
+    \vdash (\mu \phi.\lambda \ol{\alpha}. H)\ol{G}} \rho \to
+  \oCPOsem{\Gamma;\Phi,\ol\gamma \vdash (\mu
+    \phi.\lambda\ol{\alpha}. H)\ol{G}} \rho'\\
+  &= &(\mu
+  T^\oCPO_{H,\rho})\ol{\oCPOsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho} \to (\mu
+  T^\oCPO_{H,\rho'})\ol{\oCPOsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho'}
+  \end{array}\]
+  is
+  defined by
+  \[\begin{array}{ll}
+  & (\mu T^\oCPO_{H,f})_{\ol{\oCPOsem{\Gamma;\Phi,\ol\gamma \vdash
+      G}\rho'}} \circ (\mu T^\oCPO_{H,\rho})\ol{\oCPOsem{\Gamma;\Phi,\ol\gamma
+      \vdash G}f}\\[2ex] = & (\mu
+  T^\oCPO_{H,\rho'})\ol{\oCPOsem{\Gamma;\Phi,\ol\gamma 
+      \vdash G}f} \circ (\mu T^\oCPO_{H,f})_{\ol{\oCPOsem{\Gamma;\Phi,\ol\gamma
+      \vdash G}\rho}}\end{array}\]  The latter equality holds because $\mu
+  T^\oCPO_{H,\rho}$ and $\mu T^\oCPO_{H,\rho'}$ are functors and $\mu
+  T_{H,f}^\oCPO : \mu T_{H,\rho}^\oCPO \to \mu T_{H,\rho'}^\oCPO$ is a
+  natural transformation, so the following naturality square commutes:
+{\footnotesize\begin{equation}\label{eq:cd3}
+\begin{CD}
+ (\mu T^\oCPO_{H,\rho}) \ol{\oCPOsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho} @> (\mu
+  T^\oCPO_{H,f})_{\ol{\oCPOsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho}} >> (\mu
+  T^\oCPO_{H,\rho'}) \ol{\oCPOsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho} \\ 
+ @V(\mu T^\oCPO_{H,\rho}) \ol{\oCPOsem{\Gamma;\Phi,\ol\gamma \vdash G}f}VV @V  (\mu
+ T^\oCPO_{H,\rho'}) \ol{\oCPOsem{\Gamma;\Phi,\ol\gamma \vdash G}f} VV \\ 
+(\mu T^\oCPO_{H,\rho}) \ol{\oCPOsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho'} @>(\mu
+ T^\oCPO_{H,f})_{\ol{\oCPOsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho'}}>> (\mu
+ T^\oCPO_{H,\rho'}) \ol{\oCPOsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho'}
+\end{CD}
+\end{equation}}
+\item If\, $\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}$ then
+  \[\oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f
+: \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}}
+    F)\ol{A}}\rho \to \oCPOsem{\Gamma;\Phi \vdash
+    (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}\rho'\] is defined by
+\[\begin{array}{ll}
+ & \big(\textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}\,
+      \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha := \id_{\_}}]\big)
+    \ol{\oCPOsem{\Gamma; \Phi \vdash A}\rho'} \\
+ &  \circ\;
+  \big(\textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}\,
+      \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := \_}]\big)
+    \ol{\oCPOsem{\Gamma; \Phi \vdash A}f} \\
+= & \big(\textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}\,
+      \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := \_}]\big)
+    \ol{\oCPOsem{\Gamma; \Phi \vdash A}f} \\
+ &  \circ\;
+  \big(\textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}\,
+      \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha := \id_{\_}}]\big)
+    \ol{\oCPOsem{\Gamma; \Phi \vdash A}\rho}
+\end{array}\]
+where the above equality holds by naturality of
+$\textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash
+        K}\rho[\ol{\alpha := \_}]}}\,\oCPOsem{\Gamma; \Phi, \ol\alpha
+    \vdash F}f[\ol{\alpha := \id_{\_}}]$.
+%(and functoriality of
+%  $\textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash
+%        K}\rho[\ol{\alpha := \_}]}}$,
+%  $\textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash
+%        K}\rho'[\ol{\alpha := \_}]}}$, and $\ol{\oCPOsem{\Gamma; \Phi
+%      \vdash A}}$).
+\end{itemize}
+\end{dfn}
+
+Definitions~\ref{def:set-sem} and~\ref{def:set-sem-funcs} respect
+weakening, i.e., ensure that a type and its weakenings have the same
+set interpretations.
+
+\subsection{Relational Interpretations of Types}\label{sec:rel-interp}
+
+\vspace*{1in}
+
+\hrule
+
+\vspace*{0.1in}
+
+\new{
 Some questions/issues:
@@
   implementations.
-\item What kind of category can handle $\Lan$ types? Can $\omega$CPO?
-  This caregory is locally $\omega_1$-presentable (but not locally
-  finitely presentable); see 1.18(2) of~\cite{ar94}). The
-  $\omega_1$-presentable objects are the $\omega$CPOs cardinality
-  less than $\omega_1$, i.e., the countable $\omega$CPOs.
 \item $\omega$CPO is a natural choice for modeling general
@@
 \end{itemize}
+}
 

### 2020-07-10 20-gadts/inf-rules.tex
added Section 2.2 text and am mofidying

@@
 
-\new{Need to modify to take equality conditions into account.}
-  If $\C$ is any category, then given functors $F,G : \C \to
-  \oCPO$ and natural transformations $\eta, \eta' : F \to G$, we
-  define $\eta \leq \eta'$ iff $\eta_c \leq \eta'_c$ for all $c \in
-  \C$, i.e., iff $\eta_c\,x \leq \eta'_c\,x$ in $Gc$ for all $c \in
-  \C$ and $x \in Fc$.  If $(\eta_i)_{i < \omega}$ is a chain in
-  $\{\eta : F \to G\}$, then the family of morphisms $(\bigvee_{i <
-    \omega} \eta_i)_c = \lambda x. \bigvee_{i < \omega} (\eta_i)_c\,x
-  : Fc \to Gc$ defines a natural transformation $\bigvee_{i < \omega}
-  \eta_i : F \to G$, and this natural transformation is clearly the
-  supremum of $(\eta_i)_{i < \omega}$ in $\{\eta : F \to G\}$. We
-  therefore have that $\{\eta : F \to G\}$ is itself an $\oCPO$.
-  Letting $F = \lambda \ol K. \setsem{\Gamma;\Phi \vdash
-    F}\rho[\ol{\Phi := K}]$ and $G = \lambda \ol
-  K. \setsem{\Gamma;\Phi \vdash G}\rho[\ol{\Phi := K}]$, we have that
-  $\oCPOsem{\Gamma; \emptyset \vdash \Nat^\Phi F\,G}$ is indeed an
-  $\oCPO$, i.e., is a constantly-valued functor into $\oCPO$.
+\new{Need to modify to take equality conditions into account.}  If
+$\C$ is any category, then given functors $F,G : \C \to \oCPO$ and
+natural transformations $\eta, \eta' : F \to G$, we define $\eta \leq
+\eta'$ iff $\eta_c \leq \eta'_c$ for all $c \in \C$, i.e., iff
+$\eta_c\,x \leq \eta'_c\,x$ in $Gc$ for all $c \in \C$ and $x \in Fc$.
+If $(\eta_i)_{i < \omega}$ is a chain in $\{\eta : F \to G\}$, then
+the family of morphisms $(\bigvee_{i < \omega} \eta_i)_c = \lambda
+x. \bigvee_{i < \omega} (\eta_i)_c\,x : Fc \to Gc$ defines a natural
+transformation $\bigvee_{i < \omega} \eta_i : F \to G$, and this
+natural transformation is clearly the supremum of $(\eta_i)_{i <
+  \omega}$ in $\{\eta : F \to G\}$. We therefore have that $\{\eta : F
+\to G\}$ is itself an $\oCPO$.  Letting $F = \lambda \ol
+K. \setsem{\Gamma;\Phi \vdash F}\rho[\ol{\Phi := K}]$ and $G = \lambda
+\ol K. \setsem{\Gamma;\Phi \vdash G}\rho[\ol{\Phi := K}]$, we have
+that $\oCPOsem{\Gamma; \emptyset \vdash \Nat^\Phi F\,G}$ is a
+{\color{red} discrete?} functor.
+%  indeed an
+%  $\oCPO$, i.e., is a constantly-valued functor into $\oCPO$.
 
@@
 
+\begin{dfn}\label{def:rel-transf}
+A {\em $k$-ary $\oCPO$ relation transformer} $F$ is a triple $(F^1,
+F^2,F^*)$, where $F^1,F^2 : [\oCPO^k,\oCPO]_{\omega_1}$ and $F^* :
+[\oCPOR^k, \oCPOR]_{\omega_1}$ are such that
+$R_1:\oCPOR(A_1,B_1),...,R_k:\oCPOR(A_k,B_k)$ then $F^* \ol{R} :
+\oCPOR(F^1 \ol{A}, F^2 \ol{B})$, and if $(\alpha_1, \beta_1) \in
+\HomoCPOR(R_1,S_1),..., (\alpha_k, \beta_k) \in \HomoCPOR(R_k,S_k)$
+then $F^* \ol{(\alpha, \beta)} = (F^1 \ol{\alpha}, F^2 \ol{\beta})$.
+We define $F\ol{R}$ to be $F^*\overline{R}$ and
+$F\overline{(\alpha,\beta)}$ to be $F^*\overline{(\alpha,\beta)}$.
+\end{dfn}
+The last condition of the first sentence of
+Definition~\ref{def:rel-transf} expands to: if $\ol{(a,b) \in R}$
+implies $\ol{(\alpha\,a,\beta\,b) \in S}$ then $(c,d) \in F^*\ol{R}$
+implies $(F^1 \ol{\alpha}\,c,F^2 \ol{\beta}\,d) \in F^*\ol{S}$. When
+convenient we identify a $0$-ary $\oCPO$ relation transformer
+$(A,B,R)$ with $R : \oCPOR(A,B)$. We may also write $\pi_1 F$ for
+$F^1$ and $\pi_2 F$ for $F^2$. We extend these conventions to $\oCPO$
+relation environments, introduced in Definition~\ref{def:reln-env}
+below, in the obvious way.
+
+\begin{dfn}
+The category $\oCPORT_k$ of $k$-ary $\oCPO$ relation transformers is
+given by the following data:
+\begin{itemize}
+\item An object of $\oCPORT_k$ is a $k$-ary $\oCPO$ relation transformer.
+\item A morphism $\delta : (G^1,G^2,G^*) \to (H^1,H^2,H^*)$ in
+  $\oCPORT_k$ is a pair of natural transformations $(\delta^1,
+  \delta^2)$, where $\delta^1 : G^1 \to H^1$ and $\delta^2 : G^2 \to
+  H^2$ are such that, for all $\ol{R : \oCPOR(A, B)}$, if $(x, y) \in
+  G^*\ol{R}$ then $(\delta^1_{\ol{A}}x, \delta^2_{\ol{B}}y) \in
+  H^*\ol{R}$.
+\item Identity morphisms and composition are inherited from the
+  category of functors on $\oCPO$.
+\end{itemize}
+\end{dfn}
+
+\begin{dfn}\label{def:RT-functor}
+An endofunctor $H$ on $\oCPORT_k$ is a triple $H = (H^1,H^2,H^*)$, where
+\begin{itemize}
+\item $H^1$ and $H^2$ are functors from $[\oCPO^k,\oCPO]_{\omega_1}$ to $[\oCPO^k,\oCPO]_{\omega_1}$
+\item $H^*$ is a functor from $\oCPORT_k$ to $[\oCPOR^k,\oCPOR]_{\omega_1}$
+\item for all $\overline{R : \oCPOR(A,B)}$,
+  $\pi_1((H^*(\delta^1,\delta^2))_{\overline{R}}) = (H^1
+  \delta^1)_{\overline{A}}$ and
+  $\pi_2((H^*(\delta^1,\delta^2))_{\overline{R}}) = (H^2
+  \delta^2)_{\overline{B}}$
+\item The action of $H$ on objects is given by $H\,(F^1,F^2,F^*) =
+  (H^1F^1,\,H^2F^2,\,H^*(F^1,F^2,F^*))$
+\item The action of $H$ on morphisms is given by
+  $H\,(\delta^1,\delta^2) = (H^1\delta^1,H^2\delta^2)$ for
+  $(\delta^1,\delta^2) : (F^1,F^2,F^*)\to (G^1,G^2,G^*)$
+\end{itemize}
+\end{dfn}
+Since the results of applying an endofunctor $H$ to $k$-ary $\oCPO$
+relation transformers and morphisms between them must again be $k$-ary
+$\oCPO$ relation transformers and morphisms between them,
+respectively, Definition~\ref{def:RT-functor} implicitly requires that
+the following three conditions hold:\,{\em i}) if
+$R_1:\oCPOR(A_1,B_1),...,R_k:\oCPOR(A_k,B_k)$, then $H^*(F^1,F^2,F^*)
+\ol{R} : \oCPOR(H^1F^1 \ol{A}, H^2F^2 \ol{B})$; {\em ii}) if
+$(\alpha_1, \beta_1) \in \HomoCPOR(R_1,S_1),..., (\alpha_k, \beta_k)
+\in \HomoCPOR(R_k,S_k)$, then $H^*(F^1,F^2,F^*)\, \ol{(\alpha, \beta)}
+= (H^1F^1\ol{\alpha}, H^2F^2 \ol{\beta})$; and {\em iii}) if
+$(\delta^1,\delta^2) : (F^1,F^2,F^*)\to (G^1,G^2,G^*)$ and
+$R_1:\oCPOR(A_1,B_1),...,R_k:\oCPOR(A_k,B_k)$, then
+$((H^1\delta^1)_{\ol{A}}x, (H^2\delta^2)_{\ol{B}}y) \in
+H^*(G^1,G^2,G^*)\ol{R}$ whenever $(x, y) \in
+H^*(F^1,F^2,F^*)\ol{R}$. Note, however, that this last condition is
+automatically satisfied because it is implied by the third bullet
+point of Definition~\ref{def:RT-functor}.
+
+\begin{dfn}\label{def:RT-nat-trans}
+If $H$ and $K$ are endofunctors on $\oCPORT_k$, then a {\em natural
+  transformation} $\sigma : H \to K$ is a pair $\sigma = (\sigma^1,
+\sigma^2)$, where $\sigma^1 : H^1 \to K^1$ and $\sigma^2 : H^2 \to
+K^2$ are natural transformations between endofunctors on
+$[\oCPO^k,\oCPO]_{\omega_1}$ and the component of $\sigma$ at $F
+%= (F^1,F^2,F^*)
+\in \oCPORT_k$ is given by $\sigma_F = (\sigma^1_{F^1}, \sigma^2_{F^2})$.
+\end{dfn}
+
+Definition~\ref{def:RT-nat-trans} entails that $\sigma^i_{F^i}$ must
+be natural in $F^i : [\oCPO^k,\oCPO]_{\omega_1}$, and, for every $F$,
+$(\sigma^1_{F^1})_{\overline{A}}$ and
+$(\sigma^2_{F^2})_{\overline{B}}$ must be natural in $\overline{A}$
+and $\ol B$, respectively.  Moreover, since the results of applying
+$\sigma$ to $k$-ary $\oCPO$ relation transformers must be morphisms of
+$k$-ary relation transformers, Definition~\ref{def:RT-nat-trans}
+implicitly requires that $(\sigma_F)_{\overline{R}} = (
+(\sigma^1_{F^1})_{\overline{A}}, (\sigma^2_{F^2})_{\overline{B}})$ is
+a morphism in $\oCPOR$ for any $k$-tuple of relations $\overline{R :
+  \rel(A, B)}$, i.e., that if $(x, y) \in H^*F\overline{R}$, then
+$((\sigma^1_{F^1})_{\overline{A}} x, (\sigma^2_{F^2})_{\overline{B}}
+y) \in K^*F\overline{R}$.
+
+Critically, we can compute $\omega_1$-directed colimits in
+$\oCPORT_k$: it is not hard to see that if $\cal D$ is an
+$\omega_1$-directed set then $\colim{d \in {\cal D}}{(F^1_d,
+  F^2_d,F^*_d)} = (\colim{d \in {\cal D}}{F^1_d}, \colim{d \in {\cal
+    D}}{F^2_d}, \colim{d \in {\cal D}}{F^*_d})$.  We define an
+endofunctor $T = (T^1,T^2,T^*)$ on $\oCPORT_k$ to be {\em
+  $\omega_1$-cocontinuous} if $T^1$ and $T^2$ are
+$\omega_1$-cocontinuous endofunctors on $[\oCPO^k,\oCPO]_{\omega_1}$
+and $T^*$ is an $\omega_1$-cocontinuous functor from $\oCPORT_k$ to
+$[\oCPOR^k,\oCPOR]_{\omega_1}$, i.e., is in
+$[\oCPORT_k,[\oCPOR^k,\oCPOR]_{\omega_1}]_{\omega_1}$.
+
+Now, for any $k$, any $A : \oCPO$, and any $R : \oCPOR(A, B)$, let
+$K^\oCPO_A$ be the constantly $A$-valued functor from $\oCPO^k$ to
+$\oCPO$ and $K^\oCPOR_R$ be the constantly $R$-valued functor from
+$\oCPOR^k$ to $\oCPOR$.  Also let $0$ denote either the initial object
+of $\oCPO$ or the initial object of $\oCPOR$, as appropriate.
+Observing that, for every $k$, $K^\oCPO_0$ is initial in
+$[\oCPO^k,\oCPO]_{\omega_1}$, and $K^\oCPOR_0$ is initial in
+$[\oCPOR^k,\oCPOR]_{\omega_1}$, we have that, for each $k$, $K_0 =
+(K^\oCPO_0,K^\oCPO_0,K^\oCPOR_0)$ is initial in $\oCPORT_k$. Thus, if
+$T = (T^1,T^2,T^*) : \oCPORT_k \to \oCPORT_k$ is an endofunctor on
+$\oCPORT_k$ then we can define the $\omega$-CPO relation transformer
+$\mu T$ to be $\colim{n \in \nat}{T^n K_0}$.
+It is not hard to see that $\mu T$ is given explicitly as 
+\begin{equation}\label{eq:mu}
+  \mu T = (\mu T^1,\mu T^2,
+\colim{n \in \nat}{(T^nK_0)^*})
+\end{equation}
+and that, as our notation suggests, it really is a fixpoint for $T$ if
+$T$ is $\omega_1$-cocontinuous:
+\begin{lemma}\label{lem:fp}
+For any $T : [\oCPORT_k,\oCPORT_k]_{\omega_1}$, $\mu T \cong T(\mu T)$.
+\end{lemma}
+\noindent
+The isomorphism is given by the morphisms $(\mathit{in}_1,
+\mathit{in}_2) : T(\mu T) \to \mu T$ and $(in_1^{-1}, in_2^{-1}) : \mu
+T \to T(\mu T)$ in $\oCPORT_k$. The latter is always a morphism in $\oCPORT_k$,
+but the former need not be if $T$ is not $\omega_1$-cocontinuous.
+
+It is worth noting that the third component in Equation~(\ref{eq:mu})
+is the colimit in $[\oCPOR^k,\oCPOR]_{\omega_1}$ of third components
+of $\oCPO$ relation transformers, rather than a fixpoint of an
+endofunctor on $[\oCPO^k,\oCPO]_{\omega_1}$. That there is an
+asymmetry between the first two components of $\mu T$ and its third
+reflects the important conceptual observation that the third component
+of an endofunctor on $\oCPORT_k$ need not be a functor on all of
+$[\oCPOR^k,\oCPOR]_{\omega_1}$. In particular, although we can define
+$T_{H,\rho}\, F$ for an $\oCPO$ relation transformer $F$ in
+Definition~\ref{def:rel-sem} below, it is not clear how we could
+define it for an arbitrary $F : [\oCPOR^k,\oCPOR]_{\omega_1}$.
+
+\begin{dfn}\label{def:reln-env}
+An {\em $\oCPO$ relation environment} maps each type variable in
+$\tvars^k \cup \fvars^k$ to a $k$-ary $\oCPO$ relation
+transformer.  A morphism $f : \rho \to \rho'$ between $\oCPO$
+relation environments $\rho$ and $\rho'$ with $\rho|_\tvars =
+\rho'|_\tvars$ maps each type constructor variable $\psi^k \in \tvars$
+to the identity morphism on $\rho \psi^k = \rho' \psi^k$ and each
+functorial variable $\phi^k \in \fvars$ to a morphism from the $k$-ary
+$\oCPO$ relation transformer $\rho \phi$ to the $k$-ary
+$\oCPO$ relation transformer $\rho' \phi$. Composition of
+morphisms on $\oCPO$ relation environments is given
+componentwise, with the identity morphism mapping each $\oCPO$
+relation environment to itself. This gives a category of $\oCPO$
+relation environments and morphisms between them, which we denote
+$\oCPORenv$.
+\end{dfn}
+
+When convenient we identify a $0$-ary $\oCPO$ relation transformer
+with the $\oCPO$ relation (transformer) that is its codomain and
+consider an $\oCPO$ relation environment to map a type variable of
+arity $0$ to an $\oCPO$ relation.  We write $\rho[\ol{\alpha := R}]$
+for the $\oCPO$ relation environment $\rho'$ such that $\rho' \alpha_i
+\, = R_i$ for $i = 1,...,k$ and $\rho' \alpha = \rho\alpha$ if $\alpha
+\not \in \{\alpha_1,...,\alpha_k\}$.  If $\rho$ is an $\oCPO$ relation
+environment, we write $\pi_1 \rho$ and $\pi_2 \rho$ for the $\oCPO$
+environments mapping each type variable $\phi$ to the functors
+$(\rho\phi)^1$ and $(\rho\phi)^2$, respectively.
+
+{\color{red} HERE!!!!!}
+
+We define, for each $k$, the notion of an $\omega_1$-cocontinuous
+functor from $\oCPORenv$ to $\oCPORT_k$:
+\begin{dfn}\label{def:relenv-functor}
+A functor $H : [\oCPORenv, \oCPORT_k]_{\omega_1}$ is a triple $H =
+(H^1,H^2,H^*)$, where
+\begin{itemize}
+\item $H^1$ and $H^2$ are objects in
+  $[\oCPOenv,[\oCPO^k,\oCPO]_{\omega_1}]_{\omega_1}$
+\item $H^*$ is a an object in
+  $[\oCPORenv,[\oCPOR^k,\oCPOR]_{\omega_1}]_{\omega_1}$
+\item for all $\overline{R : \oCPOR(A,B)}$ and morphisms $f$ in
+  $\oCPORenv$, $\pi_1(H^*f \,{\overline{R}}) = H^1 (\pi_1
+  f)\,{\overline{A}}$ and $\pi_2(H^*f \,{\overline{R}}) = H^2 (\pi_2
+  f)\,{\overline{B}}$
+\item The action of $H$ on $\rho$ in $\oCPORenv$ is given by $H \rho = (H^1
+  (\pi_1 \rho),\,H^2 (\pi_2 \rho),\,H^*\rho)$
+\item The action of $H$ on morphisms $f : \rho \to \rho'$ in $\oCPORenv$
+  is given by $Hf = (H^1 (\pi_1 f),H^2 (\pi_2 f))$
+\end{itemize}
+\end{dfn}
+\noindent Spelling out the last two bullet points above gives the
+following analogues of the three conditions immediately following
+Definition~\ref{def:RT-functor}: {\em i}) if $R_1 :
+\oCPOR(A_1,B_1),...,R_k : \oCPOR(A_k,B_k)$, then $H^*\rho\, \ol{R} :
+\oCPOR(H^1(\pi_1 \rho)\, \ol{A}, H^2(\pi_2 \rho)\, \ol{B})$; {\em ii})
+if $(\alpha_1, \beta_1) \in \HomoCPOR(R_1,S_1),..., (\alpha_k,
+\beta_k) \in \HomoCPOR(R_k,S_k)$, then $H^*\rho\, \ol{(\alpha, \beta)}
+= (H^1(\pi_1 \rho)\,\ol{\alpha}, H^2(\pi_2 \rho)\, \ol{\beta})$; and
+{\em iii}) if $f : \rho \to \rho'$ and
+$R_1:\oCPOR(A_1,B_1),...,R_k:\oCPOR(A_k,B_k)$, then $(H^1(\pi_1
+f)\,{\ol{A}}\,x, H^2(\pi_2 f)\,{\ol{B}}\,y) \in H^*\rho'\,\ol{R}$
+whenever $(x, y) \in H^*\rho\,\ol{R}$. As before, the last condition
+is automatically satisfied because it is implied by the third bullet
+point of Definition~\ref{def:relenv-functor}.
+
+Considering $\oCPORenv$ as a product $\Pi_{\phi^k \in \tvars \cup
+  \fvars} \oCPORT_k$, we extend the computation of $\omega_1$-directed
+colimits in $\oCPORT_k$ to compute colimits in $\oCPORenv$ componentwise. We
+similarly extend the notion of an $\omega_1$-cocontinuous endofunctor on
+$\oCPORT_k$ componentwise to give a notion of $\omega_1$-cocontinuity for
+functors from $\oCPORenv$ to $\oCPORT_k$.  Recalling from the start of this
+subsection that Definition~\ref{def:rel-sem} is given mutually
+inductively with Definition~\ref{def:set-sem} we can, at last, define:
+
+\begin{dfn}\label{def:rel-sem}
+The {\em relational interpretation}
+$\oCPORsem{\cdot} : \F \to [\oCPORenv, \oCPOR]_{\omega_1}$ is defined by
+\begin{align*}
+  \oCPORsem{\Gamma;\Phi \vdash \zerot}\rho &= 0\\
+  \oCPORsem{\Gamma;\Phi \vdash \onet}\rho &= 1\\
+  \oCPORsem{\Gamma; \emptyset \vdash \Nat^{\ol\psi} \,F\,G}\rho &= \{\eta
+  : \lambda \ol{K}.\,\oCPORsem{\Gamma; \ol{\psi} \vdash
+    F}\rho[\ol{\psi := K}] \Rightarrow \lambda \ol{K}. \,\oCPORsem{
+    \Gamma; \ol{\psi} \vdash G}\rho[\ol{\psi := K}]\}\\
+  &=
+  \{(t,t') \in \oCPOsem{\Gamma; \emptyset
+    \vdash \Nat^{\ol{\psi}}
+    \,F\,G} (\pi_1 \rho) \times \oCPOsem{ 
+    \Gamma;\emptyset
+    \vdash \Nat^{\ol{\psi}} \,F\,G} (\pi_2
+  \rho)~|~\\ 
+  & \hspace{0.3in} \forall \ol{K = (K^0, K^1, K^*) : \oCPORT_{k}}.\\
+  & \hspace{0.4in} (t_{\ol{K^0}},t'_{\ol{K^1}}) \in
+  (\oCPORsem{\Gamma; \ol{\psi} \vdash G}\rho[\ol{\psi :=
+      K}])^{\oCPORsem{\Gamma;\ol{\psi}\vdash F}\rho[\ol{\psi := K}]} \}\\  
+  \oCPORsem{\Gamma;\Phi \vdash \phi \ol{F}}\rho &=
+  (\rho\phi)\ol{\oCPORsem{\Gamma;\Phi \vdash 
+    F}\rho}\\
+  \oCPORsem{\Gamma;\Phi \vdash F+G}\rho &=
+  \oCPORsem{\Gamma;\Phi \vdash F}\rho +
+  \oCPORsem{\Gamma;\Phi \vdash G}\rho\\
+  \oCPORsem{\Gamma;\Phi \vdash F\times G}\rho &=
+  \oCPORsem{\Gamma;\Phi \vdash F}\rho \times
+  \oCPORsem{\Gamma;\Phi \vdash G}\rho\\  
+   \oCPORsem{\Gamma;\Phi,\ol\gamma \vdash (\mu \phi.\lambda
+    \ol{\alpha}. H)\ol{G}}\rho
+  &= (\mu T_{H,\rho})\ol{\oCPORsem{\Gamma;\Phi,\ol\gamma \vdash
+     G}\rho}\\
+  \text{where }	T_{H,\rho}
+    &= (T^\set_{H,\pi_1\rho}, T^\set_{H,\pi_2\rho}, T^\rel_{H,\rho}) \\
+  \text{and } T^\oCPOR_{H,\rho}\,F
+    &= \lambda \ol{R}. \oCPORsem{
+      \Gamma;\ol\gamma,\phi,\ol{\alpha} \vdash H}\rho[\phi :=
+    F][\ol{\alpha := R}]\\
+  \text{and } T^\oCPOR_{H,\rho}\,\delta
+    &= \lambda \ol{R}. \oCPORsem{
+      \Gamma;\ol\gamma,\phi,\ol{\alpha} \vdash H}\id_\rho[\phi :=
+    \delta][\ol{\alpha := \id_{\ol{R}}}] \\
+  \oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}\rho &=
+    (\textit{Lan}_{\ol{\oCPORsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}
+      \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := \_}])
+    \ol{\oCPORsem{\Gamma; \Phi \vdash A}\rho}
+\end{align*}
+\end{dfn}
+
+The interpretations in Definitions~\ref{def:rel-sem}
+and~\ref{def:rel-sem-funcs} below respect weakening, and also ensure
+that $\oCPORsem{\Gamma \vdash F \to G}$ and $\oCPORsem{\Gamma \vdash
+  \forall \ol \alpha. F}$ are as expected in any parametric model.
+Observe that
+$\oCPORsem{\Gamma; \emptyset \vdash \Nat^{\ol\psi} F\,G} \rho$
+is an $\oCPO$ relation.
+%on $\oCPOsem{\Gamma; \emptyset \vdash \Nat^{\ol\psi} F\,G} (\pi_1 \rho)$
+%and $\oCPOsem{\Gamma; \emptyset \vdash \Nat^{\ol\psi} F\,G} (\pi_2 \rho)$.
+Indeed, its relation preserves sups of $\omega$-chains of natural transformations
+because the sups are computed pointwise:
+if $(t_d, t'_d) \colon \oCPORsem{\Gamma; \emptyset \vdash \Nat^{\ol\psi} F\,G} \rho$
+for all $d$ in an $\omega$-chain $\cal D$, then
+for all $\ol{K = (K^0, K^1, K^*) : \oCPORT_{k}}$
+we have that $(t_d)_{\ol{K^0}}$ and $(t'_d)_{\ol{K^1}}$
+are related for every $d : \cal D$ and they form $\omega$-chains in a $\oCPO$, and thus
+\[
+  \Big((\bigvee_d t_d)_{\ol{K^0}}, (\bigvee_d t'_d)_{\ol{K^1}} \Big)
+= \Big(\bigvee_d (t_d)_{\ol{K^0}}, \bigvee_d (t'_d)_{\ol{K^1}} \Big) \in
+  (\oCPORsem{\Gamma; \ol{\psi} \vdash G}\rho[\ol{\psi :=
+  K}])^{\oCPORsem{\Gamma;\ol{\psi}\vdash F}\rho[\ol{\psi := K}]}
+\]
+Moreover, as for $\oCPO$ interpretations,
+$\oCPORsem{\Gamma; \emptyset \vdash \Nat^{\ol\psi} F\,G}$
+is an $\oCPO$ relation
+ is $\omega_1$-cocontinuous because it is a constant
+{\color{red} discrete?} functor.
+If $\rho \in \oCPORenv$
+and $\vdash F$, then we write $\oCPORsem{\vdash F}$ instead of
+$\oCPORsem{\vdash F}\rho$.  For the second-to-last clause in
+Definition~\ref{def:rel-sem} to be well-defined we need $T_{H,\rho}$
+to be an $\omega_1$-cocontinuous endofunctor on $\oCPORT$ so that, by
+Lemma~\ref{lem:fp}, it admits a fixpoint. Since $T_{H,\rho}$ is
+defined in terms of $\oCPORsem{\Gamma;\ol\gamma,\phi^k, \ol{\alpha}
+  \vdash H}$, this means that relational interpretations of types must
+be $\omega_1$-cocontinuous functors from $\oCPORenv$ to $\oCPORT_0$, which in
+turn entails that the actions of relational interpretations of types
+on objects and on morphisms in $\oCPORenv$ are intertwined. 
+Moreover, for the ast clause in
+Definition~\ref{def:rel-sem} to be well-defined we need
+$\oCPORsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]$
+and
+$\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := \_}]$
+to be $\omega_1$-cocontinuous functors.
+As for $\oCPO$ interpretations,
+we know from~\cite{jp19} that, for every $\Gamma;
+\ol{\alpha} \vdash F$, $\oCPORsem{\Gamma; \ol{\alpha} \vdash F}$ is
+actually in $[\oCPOR^k,\oCPOR]_{\omega_1}$ where $k = |\ol \alpha|$.
+%In fact, we already know from~\cite{jp19} that, for every $\Gamma;
+%\ol{\alpha} \vdash \tau : \F$, $\relsem{\Gamma; \ol{\alpha} \vdash
+%  \tau}$ is actually functorial in $\ol{\alpha}$ and
+%$\omega$-cocontinuous.
+We first define the actions of each of these functors on morphisms
+between environments in Definition~\ref{def:rel-sem-funcs}, and then
+argue that the functors given by Definitions~\ref{def:rel-sem}
+and~\ref{def:rel-sem-funcs} are well-defined and have the required
+properties. To do this, we extend $T_H$ to a {\em functor} from
+$\oCPORenv$ to $[[\oCPOR^k,\oCPOR]_{\omega_1},[\oCPOR^k,\oCPOR]_{\omega_1}]_{\omega_1}$.
+Its action on an object
+$\rho \in \oCPORenv$ is given by the higher-order functor $T_{H,\rho}$
+whose actions on objects and morphisms are given in
+Definition~\ref{def:rel-sem-funcs}. Its action on a morphism $f : \rho
+\to \rho'$ is the higher-order natural transformation $T_{H,f} :
+T_{H,\rho} \to T_{H,\rho'}$ whose action on any $F : [\oCPOR^k,\oCPOR]_{\omega_1}$ is
+the natural transformation $T_{H,f}\, F : T_{H,\rho}\, F \to
+T_{H,\rho'}\, F$ whose component at $\ol{R}$ is $(T_{H,f}\,
+F)_{\ol{R}} = \oCPORsem{\Gamma; \ol\gamma,\phi,\ol{\alpha} \vdash
+  H}f[\phi := \id_F][\ol{\alpha := \id_R}]$. The next definition uses
+$T_H$ to define the actions of functors interpreting types on
+morphisms between $\oCPO$ relation environments.
+
+\begin{dfn}\label{def:rel-sem-funcs}
+Let $f: \rho \to \rho'$ for $\oCPO$ relation environments $\rho$ and $\rho'$
+(so that $\rho|_\tvars = \rho'|_\tvars$). The action
+$\oCPORsem{\Gamma;\Phi \vdash F}f$ of $\oCPORsem{\Gamma;\Phi \vdash F}$ on
+the morphism $f$ is given exactly as in
+Definition~\ref{def:set-sem-funcs}, except that all interpretations
+are relational interpretations and all occurrences of $T^\oCPO_{H,f}$
+are replaced by $T_{H,f}$.
+\end{dfn}
+
+To see that the functors given by Definitions~\ref{def:rel-sem}
+and~\ref{def:rel-sem-funcs} are well-defined we must show that, for
+every $H$, $T_{H,\rho}\,F$ is an $\oCPO$ relation transformer for any $\oCPO$ relation
+transformer $F$, and that $T_{H,f}\, F : T_{H,\rho}\, F \to
+T_{H,\rho'}\, F$ is a morphism of $\oCPO$ relation transformers for every
+$\oCPO$ relation transformer $F$ and every morphism $f : \rho \to \rho'$ in
+$\oCPORenv$. This is an immediate consequence of the following Lemma.
+
+\begin{lemma}\label{lem:rel-transf-morph}
+For every $\Gamma;\Phi \vdash F$, $\sem{\Gamma;\Phi \vdash F} =
+(\oCPOsem{\Gamma;\Phi \vdash F}, \oCPOsem{\Gamma;\Phi \vdash
+  F},\oCPORsem{\Gamma;\Phi \vdash F}) \in
+[\oCPORenv,\oCPORT_0]_{\omega_1}$.
+\end{lemma}
+\noindent
+The proof is a straightforward induction on the structure of $F$,
+%%each case of which first shows that $(\setsem{\Gamma;\Phi \vdash
+%  \tau}, \setsem{\Gamma;\Phi \vdash \tau},\relsem{\Gamma;\Phi \vdash
+%  \tau})$ is indeed a functor from $\relenv$ to $RT_0$, and then uses
+using an appropriate result from~\cite{jp19} to deduce
+$\omega_1$-cocontinuity of $\sem{\Gamma;\Phi \vdash F}$ in each case,
+together with Lemma~\ref{lem:fp} and Equation~\ref{eq:mu} for
+$\mu$-types.
+
+%We need
+%$\relsem{\Gamma; \Phi, \phi \vdash \tau}\rho$ $=
+%\relsem{\Gamma, \psi; \Phi \vdash \tau[\phi :==
+%    \psi]}\rho[\psi:=\rho\phi]$, and that $\relsem{\Gamma \vdash
+%  \sigma \to \tau}$ and
+%$\setsem{\Gamma; \Phi, \phi \vdash \tau}\rho$ $=
+%\setsem{\Gamma, \psi; \Phi \vdash \tau[\phi :==
+%    \psi]}\rho[\psi:=\rho\phi]$. 
+We can also prove by simultaneous induction that our interpretations
+of types interact well with demotion of functorial variables. Indeed,
+we have that, if $\rho, \rho' : \oCPOenv$, \,$f : \rho \to \rho'$,
+\,$\rho \phi = \rho \psi = \rho' \phi = \rho' \psi$, \, $f \phi = f
+\psi = \id_{\rho \phi}$,\, $\Gamma; \Phi, \phi^k \vdash F$,\,
+$\Gamma;\Phi,\ol{\alpha} \vdash G$,\, $\Gamma;\Phi,\alpha_1...\alpha_k
+\vdash H$, and $\ol{\Gamma;\Phi \vdash K}$, then
+%for any $\rho,\rho'$ and $f: \rho \to \rho'$ in
+%$\setenv$, 
+\begin{gather}
+\label{thm:demotion-objects}
+\oCPOsem{\Gamma; \Phi, \phi \vdash F} \rho = \oCPOsem{\Gamma, \psi; \Phi
+  \vdash F[\phi :== \psi] } \rho\\
+\label{thm:demotion-morphisms}
+\oCPOsem{\Gamma; \Phi, \phi \vdash F} f = \oCPOsem{\Gamma, \psi; \Phi
+  \vdash F[\phi :== \psi]} f\\
+%\end{gather}
+%Moreover,
+%\begin{lemma}\label{lem:demotion-objects}
+%\begin{gather}
+%\label{thm:demotion-objects}
+%\setsem{\Gamma; \Phi, \phi \vdash F} \rho = \setsem{\Gamma, \psi; \Phi
+%  \vdash F[\phi :== \psi] } \rho{\color{blue}[\psi := \rho \phi]}\\
+%\label{thm:demotion-morphisms}
+%\setsem{\Gamma; \Phi, \phi \vdash F} f = \setsem{\Gamma, \psi; \Phi
+%  \vdash F[\phi :== \psi]} f{\color{blue}[\psi \mapsto f\phi]}\\
+\label{eq:subs-var}
+\oCPOsem{\Gamma;\Phi \vdash G[\ol{\alpha := K}]}\rho =
+\oCPOsem{\Gamma;\Phi,\ol{\alpha} \vdash G}\rho[\ol{\alpha := 
+\oCPOsem{\Gamma;\Phi \vdash K}\rho}]\\
+\label{eq:subs-var-morph}
+\oCPOsem{\Gamma;\Phi \vdash G[\ol{\alpha := K}]}f =
+\oCPOsem{\Gamma;\Phi,\ol{\alpha} \vdash G}f[\ol{\alpha :=
+\oCPOsem{\Gamma;\Phi \vdash K}f}]\\
+\label{eq:subs-const}
+\oCPOsem{\Gamma; \Phi \vdash F[\phi := H]}\rho
+= \oCPOsem{\Gamma; \Phi, \phi \vdash F}\rho
+[\phi := \lambda \ol{A}.\, \oCPOsem{\Gamma;\Phi,\overline{\alpha}\vdash
+    H}\rho[\overline{\alpha := A}]] \\ 
+\label{eq:subs-const-morph}
+\oCPOsem{\Gamma; \Phi \vdash F[\phi := H]}f
+= \oCPOsem{\Gamma; \Phi, \phi \vdash F}f
+[\phi := \lambda \ol{A}.\,\oCPOsem{\Gamma;\Phi,\overline{\alpha}\vdash
+    H}f[\overline{\alpha := \id_{\ol{A}}}]] 
+\end{gather}
+Identities analogous to (\ref{thm:demotion-objects}) through
+(\ref{eq:subs-const-morph}) hold for relational interpretations as well.
+
 \vspace*{1in}

### 2020-07-13 20-gadts/inf-rules.tex
latest version in progress before laptop repair

@@
 $\omega$-complete partial orders ($\oCPO$s) and their sup-preserving
-morphisms by $\omega$CPO. The underlying set of an $\omega$-CPO $A$ is
+morphisms by $\omega$CPO. The underlying set of an $\oCPO$ $A$ is
 denoted $|A|$.  The category $\oCPOR$ of relations on $\omega$CPOs has
@@
 interpretations appearing in the third clause of
-Definition~\ref{def:set-sem} are given in full in Section~\ref{?}.
+Definition~\ref{def:set-sem} are given in full in
+Section~\ref{sec:iel}.
+
+{\color{red} Need to describe $\omega_1$-directed colimits in
+  $\oCPOenv$ for the type of $\oCPOsem{\cdot}$ to make sense. We do it
+  pointwise.}
 
@@
 \ol K. \setsem{\Gamma;\Phi \vdash G}\rho[\ol{\Phi := K}]$, we have
-that $\oCPOsem{\Gamma; \emptyset \vdash \Nat^\Phi F\,G}$ is a
-{\color{red} discrete?} functor.
-%  indeed an
-%  $\oCPO$, i.e., is a constantly-valued functor into $\oCPO$.
-
-Interpretations of $\Nat$-types ensure that $\oCPOsem{\Gamma \vdash F
-  \to G}$ and $\oCPOsem{\Gamma \vdash \forall \ol\alpha. F}$ are as
-expected in any parametric model.
-
-To make sense of the last clause in Definition~\ref{def:set-sem}, we
-need to know that, for each $\rho \in \oCPOenv$, $T^\oCPO_{H,\rho}$ is
-an $\omega_1$-cocontinuous endofunctor on $[\oCPO^k,
-  \oCPO]_{\omega_1}$, and thus admits a fixpoint.  Since
-$T_{H,\rho}^\oCPO$ is defined in terms of
+that $\oCPOsem{\Gamma; \emptyset \vdash \Nat^\Phi F\,G}\rho$ is indeed
+an $\oCPO$. \new{Does this use functorial actions of interps? I don't
+  think so, but double-check with Enrico.} {\color{red} That each of
+  the above interpretations is $\omega_1$-cocontinuous follows from
+  Corollary~12 of~\cite{jp19} if we APPROPRIATELY RESTRICT THE
+  SUBSCRIPTS OF $\Lan$s.  For $\Nat$-types, we note that
+  $\oCPOsem{\Gamma; \emptyset \vdash \Nat^\Phi F\,G}$ is an
+  $\omega_1$-cocontinuous functor because it is constant on
+  $\omega_1$-directed sets.} Interpretations of $\Nat$-types ensure
+that $\oCPOsem{\Gamma \vdash F \to G}$ and $\oCPOsem{\Gamma \vdash
+  \forall \ol\alpha. F}$ are as expected in any parametric model.
+
+To make sense of the next-to-last clause in
+Definition~\ref{def:set-sem}, we need to know that, for each $\rho \in
+\oCPOenv$, $T^\oCPO_{H,\rho}$ is an $\omega_1$-cocontinuous
+endofunctor on $[\oCPO^k, \oCPO]_{\omega_1}$, and thus admits a
+fixpoint.  Since $T_{H,\rho}^\oCPO$ is defined in terms of
 $\oCPOsem{\Gamma;\ol\gamma,\phi, \ol{\alpha} \vdash H}$, this means
@@
 on morphisms in $\oCPOenv$ are intertwined. Fortunately, we know
-from~\cite{jp19} that, for every $\Gamma; \Phi \vdash G$,
-$\oCPOsem{\Gamma; \Phi \vdash G}$ is actually in
-$[\ol{[\oCPO^k,\oCPO]_{\omega_1}},\oCPO]_{\omega_1}$, or,
-equivalently, $[\oCPOenv,\oCPO]_{\omega_1}$.  Therefore, for each
-$\oCPOsem{\Gamma; \ol\gamma, \phi^k, \ol{\alpha} \vdash H}$, the
+from~\cite{jp19} that, for every $\Gamma; \ol\alpha \vdash F$,
+$\oCPOsem{\Gamma; \Phi \vdash F}$ is actually in
+$[\oCPO^k,\oCPO]_{\omega_1}$, where $k = |\ol\alpha|$.  Therefore, for
+each $\oCPOsem{\Gamma; \ol\gamma, \phi^k, \ol{\alpha} \vdash H}$, the
 corresponding operator $T^\oCPO_{H}$ can be extended to a {\em
@@
 
+%For the last clause in Definition~\ref{def:set-sem} to be well-defined
+%we need $\oCPOsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]$
+%and $\oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha :=
+%    \_}]$ to be $\omega_1$-cocontinuous functors.
 In addition, for each $\ol K$, we have that $\textit{Lan}_{\ol K}$ is
@@
 \begin{dfn}\label{def:set-sem-funcs}
-Let $f: \rho \to \rho'$ be a morphism between $\omega$-CPO environments $\rho$
+Let $f: \rho \to \rho'$ be a morphism between $\oCPO$ environments $\rho$
 and $\rho'$ (so that $\rho|_\tvars = \rho'|_\tvars$). The action
@@
 F^2,F^*)$, where $F^1,F^2 : [\oCPO^k,\oCPO]_{\omega_1}$ and $F^* :
-[\oCPOR^k, \oCPOR]_{\omega_1}$ are such that
+[\oCPOR^k, \oCPOR]_{\omega_1}$ are such that if
 $R_1:\oCPOR(A_1,B_1),...,R_k:\oCPOR(A_k,B_k)$ then $F^* \ol{R} :
@@
 $T = (T^1,T^2,T^*) : \oCPORT_k \to \oCPORT_k$ is an endofunctor on
-$\oCPORT_k$ then we can define the $\omega$-CPO relation transformer
+$\oCPORT_k$ then we can define the $\oCPO$ relation transformer
 $\mu T$ to be $\colim{n \in \nat}{T^n K_0}$.
@@
 $(\rho\phi)^1$ and $(\rho\phi)^2$, respectively.
-
-{\color{red} HERE!!!!!}
 
@@
   \oCPORsem{\Gamma;\Phi \vdash \onet}\rho &= 1\\
-  \oCPORsem{\Gamma; \emptyset \vdash \Nat^{\ol\psi} \,F\,G}\rho &= \{\eta
-  : \lambda \ol{K}.\,\oCPORsem{\Gamma; \ol{\psi} \vdash
-    F}\rho[\ol{\psi := K}] \Rightarrow \lambda \ol{K}. \,\oCPORsem{
-    \Gamma; \ol{\psi} \vdash G}\rho[\ol{\psi := K}]\}\\
+  \oCPORsem{\Gamma; \emptyset \vdash \Nat^\Phi \,F\,G}\rho &= \{\eta
+  : \lambda \ol{K}.\,\oCPORsem{\Gamma; \Phi \vdash
+    F}\rho[\ol{\Phi := K}] \Rightarrow \lambda \ol{K}. \,\oCPORsem{
+    \Gamma; \Phi \vdash G}\rho[\ol{\phi := K}]\}\\
   &=
   \{(t,t') \in \oCPOsem{\Gamma; \emptyset
-    \vdash \Nat^{\ol{\psi}}
+    \vdash \Nat^\Phi
     \,F\,G} (\pi_1 \rho) \times \oCPOsem{ 
     \Gamma;\emptyset
-    \vdash \Nat^{\ol{\psi}} \,F\,G} (\pi_2
+    \vdash \Nat^\Phi \,F\,G} (\pi_2
   \rho)~|~\\ 
@@
   & \hspace{0.4in} (t_{\ol{K^0}},t'_{\ol{K^1}}) \in
-  (\oCPORsem{\Gamma; \ol{\psi} \vdash G}\rho[\ol{\psi :=
-      K}])^{\oCPORsem{\Gamma;\ol{\psi}\vdash F}\rho[\ol{\psi := K}]} \}\\  
+  (\oCPORsem{\Gamma; \Phi \vdash G}\rho[\ol{\Phi :=
+      K}])^{\oCPORsem{\Gamma;\Phi\vdash F}\rho[\ol{\Phi := K}]} \}\\  
   \oCPORsem{\Gamma;\Phi \vdash \phi \ol{F}}\rho &=
@@
 \end{dfn}
-
 The interpretations in Definitions~\ref{def:rel-sem}
@@
   \forall \ol \alpha. F}$ are as expected in any parametric model.
-Observe that
-$\oCPORsem{\Gamma; \emptyset \vdash \Nat^{\ol\psi} F\,G} \rho$
-is an $\oCPO$ relation.
-%on $\oCPOsem{\Gamma; \emptyset \vdash \Nat^{\ol\psi} F\,G} (\pi_1 \rho)$
-%and $\oCPOsem{\Gamma; \emptyset \vdash \Nat^{\ol\psi} F\,G} (\pi_2 \rho)$.
-Indeed, its relation preserves sups of $\omega$-chains of natural transformations
-because the sups are computed pointwise:
-if $(t_d, t'_d) \colon \oCPORsem{\Gamma; \emptyset \vdash \Nat^{\ol\psi} F\,G} \rho$
-for all $d$ in an $\omega$-chain $\cal D$, then
-for all $\ol{K = (K^0, K^1, K^*) : \oCPORT_{k}}$
-we have that $(t_d)_{\ol{K^0}}$ and $(t'_d)_{\ol{K^1}}$
-are related for every $d : \cal D$ and they form $\omega$-chains in a $\oCPO$, and thus
+Observe that $\oCPORsem{\Gamma; \emptyset \vdash \Nat^\Phi F\,G} \rho$
+is in $\oCPOR$. \new{Same comments here as above.} {\color{red} It is
+  indeed a relation between $\oCPO$s, and it preserves sups of
+  $\omega$-chains of natural transformations because their sups are
+  computed pointwise. Concretely, if $(t_i, t_i') \colon
+  \oCPORsem{\Gamma; \emptyset \vdash \Nat^\Phi F\,G} \rho$ for all $i
+  < \omega$, then for all $\ol{K = (K^0, K^1, K^*) : \oCPORT_{k}}$ we
+  have that
+\[((t_i)_{\ol{K^0}},\, (t'_i)_{\ol{K^1}}) \in (\oCPORsem{\Gamma; \Phi
+  \vdash G}\rho[\ol{\Phi := K}])^{\oCPORsem{\Gamma;\Phi\vdash
+    F}\rho[\ol{\Phi := K}]}\] for every $i < \omega$. In addition,
+$((t_i)_{\ol{K^0}})_{i < \omega}$ is an $\omega$-chain in
+$(\oCPOsem{\Gamma; \Phi \vdash G}\rho[\ol{\Phi :=
+    K^0}])^{\oCPOsem{\Gamma;\Phi\vdash F}\rho[\ol{\Phi := K^0}]}$ and
+$((t'_i)_{\ol{K^1}})_{i < \omega}$ is an $\omega$-chain in
+$(\oCPOsem{\Gamma; \Phi \vdash G}\rho[\ol{\Phi :=
+    K^1}])^{\oCPOsem{\Gamma;\Phi\vdash F}\rho[\ol{\Phi := K^1}]}$, so
+their sups are not only well-defined but must also be related
+in
+$(\oCPORsem{\Gamma;
+    \Phi \vdash G}\rho[\ol{\Phi := K}])^{\oCPORsem{\Gamma;\Phi\vdash
+      F}\rho[\ol{\Phi := K}]}$. This gives that
 \[
-  \Big((\bigvee_d t_d)_{\ol{K^0}}, (\bigvee_d t'_d)_{\ol{K^1}} \Big)
-= \Big(\bigvee_d (t_d)_{\ol{K^0}}, \bigvee_d (t'_d)_{\ol{K^1}} \Big) \in
-  (\oCPORsem{\Gamma; \ol{\psi} \vdash G}\rho[\ol{\psi :=
-  K}])^{\oCPORsem{\Gamma;\ol{\psi}\vdash F}\rho[\ol{\psi := K}]}
+  \Big((\bigvee_{i < \omega} t_i)_{\ol{K^0}}, (\bigvee_{i < \omega}
+  t'_i)_{\ol{K^1}} \Big) = \Big(\bigvee_{i < \omega} (t_i)_{\ol{K^0}},
+  \bigvee_{i < \omega} (t'_i)_{\ol{K^1}} \Big) \in (\oCPORsem{\Gamma;
+    \Phi \vdash G}\rho[\ol{\Phi := K}])^{\oCPORsem{\Gamma;\Phi\vdash
+      F}\rho[\ol{\Phi := K}]}
 \]
-Moreover, as for $\oCPO$ interpretations,
-$\oCPORsem{\Gamma; \emptyset \vdash \Nat^{\ol\psi} F\,G}$
-is an $\oCPO$ relation
- is $\omega_1$-cocontinuous because it is a constant
-{\color{red} discrete?} functor.
-If $\rho \in \oCPORenv$
-and $\vdash F$, then we write $\oCPORsem{\vdash F}$ instead of
+Moreover, as for $\oCPO$ interpretations, $\omega_1$-cocontinuity of
+each of the above interpretations follows from Corollary~12
+of~\cite{jp19} if we APPROPRIATELY RESTRICT THE SUBSCRIPTS of the
+$\Lan$s. For $\Nat$-types, $\oCPORsem{\Gamma; \emptyset \vdash
+  \Nat^\Phi F\,G}$ is an $\omega_1$-cocontinuous functor because it is
+constant on $\omega_1$-directed sets.} If $\rho \in \oCPORenv$ and
+$\vdash F$, then we write $\oCPORsem{\vdash F}$ instead of
 $\oCPORsem{\vdash F}\rho$.  For the second-to-last clause in
@@
   \vdash H}$, this means that relational interpretations of types must
-be $\omega_1$-cocontinuous functors from $\oCPORenv$ to $\oCPORT_0$, which in
-turn entails that the actions of relational interpretations of types
-on objects and on morphisms in $\oCPORenv$ are intertwined. 
-Moreover, for the ast clause in
-Definition~\ref{def:rel-sem} to be well-defined we need
-$\oCPORsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]$
-and
-$\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := \_}]$
-to be $\omega_1$-cocontinuous functors.
-As for $\oCPO$ interpretations,
-we know from~\cite{jp19} that, for every $\Gamma;
-\ol{\alpha} \vdash F$, $\oCPORsem{\Gamma; \ol{\alpha} \vdash F}$ is
-actually in $[\oCPOR^k,\oCPOR]_{\omega_1}$ where $k = |\ol \alpha|$.
-%In fact, we already know from~\cite{jp19} that, for every $\Gamma;
-%\ol{\alpha} \vdash \tau : \F$, $\relsem{\Gamma; \ol{\alpha} \vdash
-%  \tau}$ is actually functorial in $\ol{\alpha}$ and
-%$\omega$-cocontinuous.
-We first define the actions of each of these functors on morphisms
-between environments in Definition~\ref{def:rel-sem-funcs}, and then
-argue that the functors given by Definitions~\ref{def:rel-sem}
-and~\ref{def:rel-sem-funcs} are well-defined and have the required
-properties. To do this, we extend $T_H$ to a {\em functor} from
-$\oCPORenv$ to $[[\oCPOR^k,\oCPOR]_{\omega_1},[\oCPOR^k,\oCPOR]_{\omega_1}]_{\omega_1}$.
-Its action on an object
+be $\omega_1$-cocontinuous functors from $\oCPORenv$ to $\oCPORT_0$,
+which in turn entails that the actions of relational interpretations
+of types on objects and on morphisms in $\oCPORenv$ are intertwined.
+%Moreover, for the last clause in Definition~\ref{def:rel-sem} to be
+%well-defined we need $\oCPORsem{\Gamma; \ol\alpha \vdash
+%  K}\rho[\ol{\alpha := \_}]$ and $\oCPORsem{\Gamma; \Phi, \ol\alpha
+%  \vdash F}\rho[\ol{\alpha := \_}]$ to be $\omega_1$-cocontinuous
+%functors.
+As for $\oCPO$ interpretations, we know from~\cite{jp19} that, for
+every $\Gamma; \ol\alpha \vdash F$, $\oCPORsem{\Gamma; \ol\alpha
+  \vdash F}$ is actually in $[\oCPOR^k,\oCPOR]_{\omega_1}$, where $k =
+|\ol\alpha|$. We first define the actions of each of these functors
+on morphisms between $\oCPO$ relation environments in
+Definition~\ref{def:rel-sem-funcs}, and then argue that the functors
+given by Definitions~\ref{def:rel-sem} and~\ref{def:rel-sem-funcs} are
+well-defined and have the required properties. To do this, we extend
+$T_H$ to a {\em functor} from $\oCPORenv$ to
+$[[\oCPOR^k,\oCPOR]_{\omega_1},
+  [\oCPOR^k,\oCPOR]_{\omega_1}]_{\omega_1}$.  Its action on an object
 $\rho \in \oCPORenv$ is given by the higher-order functor $T_{H,\rho}$
@@
 \to \rho'$ is the higher-order natural transformation $T_{H,f} :
-T_{H,\rho} \to T_{H,\rho'}$ whose action on any $F : [\oCPOR^k,\oCPOR]_{\omega_1}$ is
-the natural transformation $T_{H,f}\, F : T_{H,\rho}\, F \to
-T_{H,\rho'}\, F$ whose component at $\ol{R}$ is $(T_{H,f}\,
-F)_{\ol{R}} = \oCPORsem{\Gamma; \ol\gamma,\phi,\ol{\alpha} \vdash
-  H}f[\phi := \id_F][\ol{\alpha := \id_R}]$. The next definition uses
-$T_H$ to define the actions of functors interpreting types on
-morphisms between $\oCPO$ relation environments.
+T_{H,\rho} \to T_{H,\rho'}$ whose action on any $F :
+[\oCPOR^k,\oCPOR]_{\omega_1}$ is the higher-order natural
+transformation $T_{H,f}\, F : T_{H,\rho}\, F \to T_{H,\rho'}\, F$
+whose component at $\ol{R}$ is $(T_{H,f}\, F)_{\ol{R}} =
+\oCPORsem{\Gamma; \ol\gamma,\phi,\ol{\alpha} \vdash H}f[\phi :=
+  \id_F][\ol{\alpha := \id_R}]$. The next definition uses $T_H$ to
+define the actions of functors interpreting types on morphisms between
+$\oCPO$ relation environments.
 
 \begin{dfn}\label{def:rel-sem-funcs}
-Let $f: \rho \to \rho'$ for $\oCPO$ relation environments $\rho$ and $\rho'$
-(so that $\rho|_\tvars = \rho'|_\tvars$). The action
-$\oCPORsem{\Gamma;\Phi \vdash F}f$ of $\oCPORsem{\Gamma;\Phi \vdash F}$ on
-the morphism $f$ is given exactly as in
+Let $f: \rho \to \rho'$ for $\oCPO$ relation environments $\rho$ and
+$\rho'$ (so that $\rho|_\tvars = \rho'|_\tvars$). The action
+$\oCPORsem{\Gamma;\Phi \vdash F}f$ of $\oCPORsem{\Gamma;\Phi \vdash
+  F}$ on the morphism $f$ is given exactly as in
 Definition~\ref{def:set-sem-funcs}, except that all interpretations
-are relational interpretations and all occurrences of $T^\oCPO_{H,f}$
-are replaced by $T_{H,f}$.
+are $\oCPO$ relational interpretations and all occurrences of
+$T^\oCPO_{H,f}$ are replaced by $T_{H,f}$.
 \end{dfn}
@@
 and~\ref{def:rel-sem-funcs} are well-defined we must show that, for
-every $H$, $T_{H,\rho}\,F$ is an $\oCPO$ relation transformer for any $\oCPO$ relation
-transformer $F$, and that $T_{H,f}\, F : T_{H,\rho}\, F \to
-T_{H,\rho'}\, F$ is a morphism of $\oCPO$ relation transformers for every
-$\oCPO$ relation transformer $F$ and every morphism $f : \rho \to \rho'$ in
-$\oCPORenv$. This is an immediate consequence of the following Lemma.
+every $H$, $T_{H,\rho}\,F$ is an $\oCPO$ relation transformer for any
+$\oCPO$ relation transformer $F$, and that $T_{H,f}\, F : T_{H,\rho}\,
+F \to T_{H,\rho'}\, F$ is a morphism of $\oCPO$ relation transformers
+for every $\oCPO$ relation transformer $F$ and every morphism $f :
+\rho \to \rho'$ in $\oCPORenv$. This is an immediate consequence of
+the following Lemma.
 
@@
 The proof is a straightforward induction on the structure of $F$,
-%%each case of which first shows that $(\setsem{\Gamma;\Phi \vdash
-%  \tau}, \setsem{\Gamma;\Phi \vdash \tau},\relsem{\Gamma;\Phi \vdash
-%  \tau})$ is indeed a functor from $\relenv$ to $RT_0$, and then uses
 using an appropriate result from~\cite{jp19} to deduce
@@
 together with Lemma~\ref{lem:fp} and Equation~\ref{eq:mu} for
-$\mu$-types.
-
-%We need
-%$\relsem{\Gamma; \Phi, \phi \vdash \tau}\rho$ $=
-%\relsem{\Gamma, \psi; \Phi \vdash \tau[\phi :==
-%    \psi]}\rho[\psi:=\rho\phi]$, and that $\relsem{\Gamma \vdash
-%  \sigma \to \tau}$ and
-%$\setsem{\Gamma; \Phi, \phi \vdash \tau}\rho$ $=
-%\setsem{\Gamma, \psi; \Phi \vdash \tau[\phi :==
-%    \psi]}\rho[\psi:=\rho\phi]$. 
+$\mu$-types. {\color{red} Anything needed for $\Lan$ types??}
+
 We can also prove by simultaneous induction that our interpretations
@@
 \vdash H$, and $\ol{\Gamma;\Phi \vdash K}$, then
-%for any $\rho,\rho'$ and $f: \rho \to \rho'$ in
-%$\setenv$, 
 \begin{gather}
@@
 Identities analogous to (\ref{thm:demotion-objects}) through
-(\ref{eq:subs-const-morph}) hold for relational interpretations as well.
+(\ref{eq:subs-const-morph}) hold for $\oCPO$ relational
+interpretations as well.
 

### 2020-07-13 20-gadts/inf-rules.tex
Sec 2 complete

@@
 as its objects triples $(A,B,R)$, where $A, B: \oCPO$ and $R:
-\rel(|A|, |B|)$, and has as its morphisms from $(A,B,R)$ to
+\rel(|A|, |B|)$ and $(\bigvee_{i < \omega} a_i, \bigvee_{i < \omega}
+b_i) \in R$ whenever $(a_i)_{i < \omega}$ and $(b_i)_{i < \omega}$ are
+chains in $A$ and $B$, respectively, such that $(a_i,b_i) \in R$ for
+all $i$. The category $\oCPOR$ has as its morphisms from $(A,B,R)$ to
 $(A',B',R')$ pairs $(f : A \to A',g : B \to B')$ of morphisms in
 $\oCPO$ such that $(f\,a, g\,b) \in R'$ whenever $(a,b) \in R$. We
-write $R : \oCPOR(A,B)$ in place of $(A,B,R): \oCPOR$ when
-convenient. If $R : \oCPOR(A,B)$ then we write $\pi_1 R$ and $\pi_2 R$
-for the {\em domain} $A$ of $R$ and the {\em codomain} $B$ of $R$,
-respectively. If $A : \oCPO$, then we write $\Eq_A = (A,A,\Eq_{|A|})$
-for the {\em equality relation} on $A$.
+note that if $(f,g) : (A,B,R) \to (A',B',R')$ and $(a_i)_{i < \omega}$
+and $(b_i)_{i < \omega}$ are chains in $A$ and $B$, respectively, then
+$ (f \, (\bigvee_{i < \omega} a_i), g \, (\bigvee_{i < \omega} b_i)) =
+(\bigvee_{i < \omega} (f a_i), \bigvee_{i < \omega} (g b_i)) \in R'$
+necessarily holds.  We write $R : \oCPOR(A,B)$ in place of $(A,B,R):
+\oCPOR$ when convenient. If $R : \oCPOR(A,B)$ then we write $\pi_1 R$
+and $\pi_2 R$ for the {\em domain} $A$ of $R$ and the {\em codomain}
+$B$ of $R$, respectively. If $A : \oCPO$, then we write $\Eq_A =
+(A,A,\Eq_{|A|})$ for the {\em equality relation} on $A$.
 
@@
 Definition~\ref{def:set-sem} are given in full in
-Section~\ref{sec:iel}.
-
-{\color{red} Need to describe $\omega_1$-directed colimits in
-  $\oCPOenv$ for the type of $\oCPOsem{\cdot}$ to make sense. We do it
-  pointwise.}
+Section~\ref{sec:rel-interp}. We note that $\omega_1$-directed
+colimits in $\oCPOenv$ are taken pointwise.
 
@@
 instead of $\oCPOsem{\vdash F}\rho$ since the environment is
-immaterial. The third clause of Definition~\ref{def:set-sem} does
-indeed define an $\oCPO$. First, we have
-
-\begin{lemma}
+immaterial.
+
+For Definition~\ref{def:set-sem} to be well-defined, we have to check
+that each object interpretation is in $\oCPO$ and, in particular, that
+each contains sups of all $\omega$-chains. This will be proved by
+induction on types, and in most cases existence of sups of
+$\omega$-chains will follow from the induction hypotheses. However,
+well-definedness needs to be proved directly for object
+interpretations of $\Nat$-types. First, we have
+\begin{lemma}\label{lem:nat-types-sets}
 The collection of all natural transformations 
@@
 
-\new{Need to modify to take equality conditions into account.}  If
-$\C$ is any category, then given functors $F,G : \C \to \oCPO$ and
+\new{Need to modify to take equality conditions into account.}  Now,
+if $\C$ is any category, then given functors $F,G : \C \to \oCPO$ and
 natural transformations $\eta, \eta' : F \to G$, we define $\eta \leq
@@
 the family of morphisms $(\bigvee_{i < \omega} \eta_i)_c = \lambda
-x. \bigvee_{i < \omega} (\eta_i)_c\,x : Fc \to Gc$ defines a natural
+x. \bigvee_{i < \omega} ((\eta_i)_c\,x) : Fc \to Gc$ defines a natural
 transformation $\bigvee_{i < \omega} \eta_i : F \to G$, and this
@@
 that $\oCPOsem{\Gamma; \emptyset \vdash \Nat^\Phi F\,G}\rho$ is indeed
-an $\oCPO$. \new{Does this use functorial actions of interps? I don't
-  think so, but double-check with Enrico.} {\color{red} That each of
-  the above interpretations is $\omega_1$-cocontinuous follows from
-  Corollary~12 of~\cite{jp19} if we APPROPRIATELY RESTRICT THE
-  SUBSCRIPTS OF $\Lan$s.  For $\Nat$-types, we note that
+an $\oCPO$.
+
+That each of the above interpretations is $\omega_1$-cocontinuous
+follows from Corollary~12 of~\cite{jp19} if we \new{APPROPRIATELY
+  RESTRICT THE SUBSCRIPTS OF $\Lan$s}.  For $\Nat$-types, we note that
 $\oCPOsem{\Gamma; \emptyset \vdash \Nat^\Phi F\,G}$ is an
-  $\omega_1$-cocontinuous functor because it is constant on
-  $\omega_1$-directed sets.} Interpretations of $\Nat$-types ensure
-that $\oCPOsem{\Gamma \vdash F \to G}$ and $\oCPOsem{\Gamma \vdash
-  \forall \ol\alpha. F}$ are as expected in any parametric model.
+$\omega_1$-cocontinuous functor because, in accordance with
+Definition~\ref{def:set-sem-funcs}, it is constant on
+$\omega_1$-directed sets. Interpretations of $\Nat$-types ensure that
+$\oCPOsem{\Gamma \vdash F \to G}$ and $\oCPOsem{\Gamma \vdash \forall
+  \ol\alpha. F}$ are as expected in any parametric model.
 
@@
 \end{dfn}
-The last condition of the first sentence of
+The first condition of the first sentence of
+Definition~\ref{def:rel-transf} entails that $F^*\ol R$ relates sups
+of chains of pairwise related elements in $F^1 \ol A$ and $F^2 \ol
+B$. The last condition of the first sentence of
 Definition~\ref{def:rel-transf} expands to: if $\ol{(a,b) \in R}$
@@
 $\oCPORT_k$ then we can define the $\oCPO$ relation transformer
-$\mu T$ to be $\colim{n \in \nat}{T^n K_0}$.
+$\mu T$ to be $\colim{n < \omega_1}{T^n K_0}$.
 It is not hard to see that $\mu T$ is given explicitly as 
@@
   \mu T = (\mu T^1,\mu T^2,
-\colim{n \in \nat}{(T^nK_0)^*})
+\colim{n < \omega_1}{(T^nK_0)^*})
 \end{equation}
@@
 \end{dfn}
-The interpretations in Definitions~\ref{def:rel-sem}
-and~\ref{def:rel-sem-funcs} below respect weakening, and also ensure
-that $\oCPORsem{\Gamma \vdash F \to G}$ and $\oCPORsem{\Gamma \vdash
-  \forall \ol \alpha. F}$ are as expected in any parametric model.
-Observe that $\oCPORsem{\Gamma; \emptyset \vdash \Nat^\Phi F\,G} \rho$
-is in $\oCPOR$. \new{Same comments here as above.} {\color{red} It is
-  indeed a relation between $\oCPO$s, and it preserves sups of
-  $\omega$-chains of natural transformations because their sups are
-  computed pointwise. Concretely, if $(t_i, t_i') \colon
-  \oCPORsem{\Gamma; \emptyset \vdash \Nat^\Phi F\,G} \rho$ for all $i
-  < \omega$, then for all $\ol{K = (K^0, K^1, K^*) : \oCPORT_{k}}$ we
-  have that
-\[((t_i)_{\ol{K^0}},\, (t'_i)_{\ol{K^1}}) \in (\oCPORsem{\Gamma; \Phi
-  \vdash G}\rho[\ol{\Phi := K}])^{\oCPORsem{\Gamma;\Phi\vdash
-    F}\rho[\ol{\Phi := K}]}\] for every $i < \omega$. In addition,
-$((t_i)_{\ol{K^0}})_{i < \omega}$ is an $\omega$-chain in
-$(\oCPOsem{\Gamma; \Phi \vdash G}\rho[\ol{\Phi :=
-    K^0}])^{\oCPOsem{\Gamma;\Phi\vdash F}\rho[\ol{\Phi := K^0}]}$ and
-$((t'_i)_{\ol{K^1}})_{i < \omega}$ is an $\omega$-chain in
-$(\oCPOsem{\Gamma; \Phi \vdash G}\rho[\ol{\Phi :=
-    K^1}])^{\oCPOsem{\Gamma;\Phi\vdash F}\rho[\ol{\Phi := K^1}]}$, so
-their sups are not only well-defined but must also be related
-in
-$(\oCPORsem{\Gamma;
-    \Phi \vdash G}\rho[\ol{\Phi := K}])^{\oCPORsem{\Gamma;\Phi\vdash
-      F}\rho[\ol{\Phi := K}]}$. This gives that
+
+For Definition~\ref{def:rel-sem} to be well-defined, we have to check
+that each relational interpretation is in $\oCPOR$ and, in particular,
+that each relates sups of pairwise related $\omega$-chains. This will
+be proved by induction on types, and in most cases it will follow from
+the induction hypotheses. However, well-definedness needs to be proved
+directly for relational interpretations of $\Nat$-types.
+
+The proof that relational interpretations of $\Nat$-types define sets
+is analogous to the proof of Lemma~\ref{lem:nat-types-sets}.  Next, we
+observe that $\oCPORsem{\Gamma; \emptyset \vdash \Nat^\Phi F\,G} \rho$
+is in $\oCPOR$. It is indeed a relation between $\oCPO$s, and it
+relates sups of pairwise related $\omega$-chains of natural
+transformations because their sups are computed pointwise. More
+specifically, if $(t_i, t_i') \colon \oCPORsem{\Gamma; \emptyset
+  \vdash \Nat^\Phi F\,G} \rho$ for all $i < \omega$, then for every
+$\ol{K = (K^1, K^2, K^*) : \oCPORT_{k}}$ and every $i < \omega$ we
+have
+  \[\big((t_i)_{\ol{K^1}},\, (t'_i)_{\ol{K^2}}\big) \in
+  (\oCPORsem{\Gamma; \Phi \vdash G}\rho[\ol{\Phi :=
+      K}])^{\oCPORsem{\Gamma;\Phi\vdash F}\rho[\ol{\Phi := K}]}\] So
+  if $(a,b) \in \oCPORsem{\Gamma;\Phi\vdash F}\rho[\ol{\Phi := K}]$,
+  then $\big((t_i)_{\ol{K^1}}\,a,\, (t'_i)_{\ol{K^2}}\,b\big) \in
+  \oCPORsem{\Gamma;\Phi\vdash G}\rho[\ol{\Phi := K}]$ for all $i <
+  \omega$. But then since $\big((t_i)_{\ol{K^1}} a\big)_{i < \omega}$
+  is an $\omega$-chain in $\oCPOsem{\Gamma; \Phi \vdash
+    G}\rho[\ol{\Phi := K^1}]$ and $\big((t'_i)_{\ol{K^2}}\big)_{i <
+    \omega} b$ is an $\omega$-chain in $\oCPOsem{\Gamma; \Phi \vdash
+    G}\rho[\ol{\Phi := K^2}]$, not only are $\bigvee_{i < \omega}
+  ((t_i)_{\ol{K^1}}\,a)$ and $\bigvee_{i < \omega}
+  ((t'_i)_{\ol{K^2}}\,b)$ well-defined, but, since
+\[\oCPORsem{\Gamma;\Phi\vdash G}\rho[\ol{\Phi := K}] : \oCPOR(
+  \oCPOsem{\Gamma;\Phi\vdash G}\rho[\ol{\Phi := K^1}],
+  \oCPOsem{\Gamma;\Phi\vdash G}\rho[\ol{\Phi := K^2}])\]
+we have
   \[
-  \Big((\bigvee_{i < \omega} t_i)_{\ol{K^0}}, (\bigvee_{i < \omega}
-  t'_i)_{\ol{K^1}} \Big) = \Big(\bigvee_{i < \omega} (t_i)_{\ol{K^0}},
-  \bigvee_{i < \omega} (t'_i)_{\ol{K^1}} \Big) \in (\oCPORsem{\Gamma;
+  \Big(\big(\bigvee_{i < \omega} t_i\big)_{\ol{K^1}} \,a,
+  \big(\bigvee_{i < \omega} t'_i\big)_{\ol{K^2}} \,b\Big)
+  = \Big(\bigvee_{i < \omega} \big((t_i)_{\ol{K^1}}\,a\big),\,
+  \bigvee_{i < \omega} \big((t'_i)_{\ol{K^2}}\,b\big)\Big) \in
+          \oCPORsem{\Gamma;\Phi\vdash G}\rho[\ol{\Phi
+             := K}]\] as well. That is,
+\[
+  \Big(\big(\bigvee_{i < \omega} t_i\big)_{\ol{K^1}}, \big(\bigvee_{i
+    < \omega} t'_i\big)_{\ol{K^2}} \Big)
+  \in (\oCPORsem{\Gamma;
     \Phi \vdash G}\rho[\ol{\Phi := K}])^{\oCPORsem{\Gamma;\Phi\vdash
@@
   \]
+i.e.,
+$( \bigvee_{i < \omega} t_i, \bigvee_{i < \omega} t'_i ) \in
+  \oCPORsem{\Gamma; \emptyset \vdash \Nat^\Phi F\,G} \rho$.
+
 Moreover, as for $\oCPO$ interpretations, $\omega_1$-cocontinuity of
 each of the above interpretations follows from Corollary~12
-of~\cite{jp19} if we APPROPRIATELY RESTRICT THE SUBSCRIPTS of the
-$\Lan$s. For $\Nat$-types, $\oCPORsem{\Gamma; \emptyset \vdash
+of~\cite{jp19} if we \new{APPROPRIATELY RESTRICT THE SUBSCRIPTS of
+$\Lan$s}. For $\Nat$-types, $\oCPORsem{\Gamma; \emptyset \vdash
   \Nat^\Phi F\,G}$ is an $\omega_1$-cocontinuous functor because it is
-constant on $\omega_1$-directed sets.} If $\rho \in \oCPORenv$ and
-$\vdash F$, then we write $\oCPORsem{\vdash F}$ instead of
-$\oCPORsem{\vdash F}\rho$.  For the second-to-last clause in
-Definition~\ref{def:rel-sem} to be well-defined we need $T_{H,\rho}$
-to be an $\omega_1$-cocontinuous endofunctor on $\oCPORT$ so that, by
-Lemma~\ref{lem:fp}, it admits a fixpoint.  Since $T_{H,\rho}$ is
-defined in terms of $\oCPORsem{\Gamma;\ol\gamma,\phi^k, \ol{\alpha}
-  \vdash H}$, this means that relational interpretations of types must
-be $\omega_1$-cocontinuous functors from $\oCPORenv$ to $\oCPORT_0$,
+constant on $\omega_1$-directed sets.  Interpretations of $\Nat$-types
+ensure that $\oCPORsem{\Gamma \vdash F \to G}$ and $\oCPORsem{\Gamma
+  \vdash \forall \ol \alpha. F}$ are as expected in any parametric
+model.
+
+For the next-to-last clause in Definition~\ref{def:rel-sem} to be
+well-defined we need $T_{H,\rho}$ to be an $\omega_1$-cocontinuous
+endofunctor on $\oCPORT$ so that, by Lemma~\ref{lem:fp}, it admits a
+fixpoint.  Since $T_{H,\rho}$ is defined in terms of
+$\oCPORsem{\Gamma;\ol\gamma,\phi^k, \ol{\alpha} \vdash H}$, this means
+that relational interpretations of types must be
+$\omega_1$-cocontinuous functors from $\oCPORenv$ to $\oCPORT_0$,
 which in turn entails that the actions of relational interpretations
@@
 
+If $\rho \in \oCPORenv$ and $\vdash F$, then we write
+$\oCPORsem{\vdash F}$ instead of $\oCPORsem{\vdash F}\rho$. The
+interpretations in Definitions~\ref{def:rel-sem} and in
+Definition~\ref{def:rel-sem-funcs} below respect weakening.
+
 \begin{dfn}\label{def:rel-sem-funcs}
@@
 together with Lemma~\ref{lem:fp} and Equation~\ref{eq:mu} for
-$\mu$-types. {\color{red} Anything needed for $\Lan$ types??}
+$\mu$-types. \new{$\Lan$ types need restriction on their
+  subscripts.}
 

### 2020-07-14 20-gadts/inf-rules.tex
small repairs

@@
   F\;G[\ol{\beta := K}]$}
-\AXC{$\ol{\Gamma;\Phi \vdash A}$}
-\AXC{$\Gamma;\Phi~|~\Delta \vdash t : (\Lan^{\ol\alpha}_{\ol K} F)\,\ol A $}
+\AXC{$\ol{\Gamma;\Phi \vdash B}$}
+\AXC{$\Gamma;\Phi~|~\Delta \vdash t : (\Lan^{\ol\alpha}_{\ol K} F)\,\ol B $}
 \TIC{$\Gamma;\Phi~|~\Delta \vdash \partial^{G, \ol K}_F \eta\,t :
-  G[\ol{\beta := A}]$}
+ G[\ol{\beta := B}]$}
 \DisplayProof
@@
 variable $x : G(\alpha)$ as a map $t_0$ associating to each set $A$ a
-homomorphism $t_0(A) : G_0(A) \to F_0(A)$, and to each relation $R$ a
+function $t_0(A) : G_0(A) \to F_0(A)$, and to each relation $R$ a
 morphism $t_1(R) : G_1(R) \to F_1(R)$.  These interpretations are to
@@
 morphisms by $\omega$CPO. The underlying set of an $\oCPO$ $A$ is
-denoted $|A|$.  The category $\oCPOR$ of relations on $\omega$CPOs has
+denoted $|A|$.  The category $\oCPOR$ of $\oCPO$ relations has
 as its objects triples $(A,B,R)$, where $A, B: \oCPO$ and $R:
@@
 
-If $\mcA$ and $\mcC$ are $\omega_1$-cocomplete, then the functor $F :
+If $\mcA$ and $\mcC$ are $\omega_1$-cocomplete, then a functor $F :
 \mcA \to \mcC$ is {\em $\omega_1$-cocontinuous} if it preserves
@@
 are again (co)continuous. It follows that $[\A,\C]_{\omega_1}$ is
-($\omega_1$1)(co)complete whenever $\C$ is.
+($\omega_1$-)(co)complete whenever $\C$ is.
 
@@
 every $\omega_1$-directed poset $\mathcal D$ and every functor $F :
-{\mathcal D} \to \mcC$, there is a canonical isomorphism $\colim{d \in
+{\mathcal D} \to \A$, there is a canonical isomorphism $\colim{d \in
   \mcD}{\mathsf{Hom}_\mcA(A,Fd)} \simeq \mathsf{Hom}_{\mcA}(A,
@@
 $\omega_1$-presentable. Thus, for all $k_1,...,k_n \in \nat^n$,
-$[\mathcal{A}, \mathcal{C}]_{\omega_1}^{k_1} \times ... \times
-[\mathcal{A}, \mathcal{C}]_{\omega_1}^{k_n}$ is locally
-$\omega_1$-presentable, and therefore $[[\mathcal{A},
-    \mathcal{C}]_{\omega_1}^{k_1} \times ... \times [\mathcal{A},
-    \mathcal{C}]_{\omega_1}^{k_n}, \mathcal{C}]_{\omega_1}$ is as
+$[\mathcal{A}^{k_1}, \mathcal{C}]_{\omega_1} \times ... \times
+[\mathcal{A}^{k_n}, \mathcal{C}]_{\omega_1}$ is locally
+$\omega_1$-presentable, and therefore $[[\mathcal{A}^{k_1},
+    \mathcal{C}]_{\omega_1} \times ... \times [\mathcal{A}^{k_n},
+    \mathcal{C}]_{\omega_1}, \mathcal{C}]_{\omega_1}$ is as
 well. Taking both $\A$ and $C$ to be $\oCPO$ --- as we will to ensure
@@
   For all $k_1,...,k_n \in \nat^n$,
-  \[[[\oCPO, \oCPO]_{\omega_1}^{k_1}
-  \times ... \times [\oCPO, \oCPO]_{\omega_1}^{k_n},
+  \[[[\oCPO^{k_1}, \oCPO]_{\omega_1}
+  \times ... \times [\oCPO^{k_n}, \oCPO]_{\omega_1},
   \oCPO]_{\omega_1}\] is locally $\omega_1$-presentable.
@@
 $K_i : [\oCPO^{k_i},\oCPO]_{\omega_1}$ for $i = 1,...,n$, then we
-write $\rho[\ol{\Phi := K}]$ for the $\oCPO$ environment $\rho'$ such
-that $\rho' \phi_i = K_i$ for $i = 1,...,n$ and $\rho' \phi = \rho
-\phi$ if $\phi \not \in \Phi$.  If $\rho$ is an $\oCPO$ environment,
-we write $\Eq_\rho$ for the $\oCPOR$ environment (see
-Definition~\ref{def:reln-env}) such that $\Eq_\rho v = \Eq_{\rho v}$
-for every type variable $v$. The categories $\oCPORT_k$ and relational
-interpretations appearing in the third clause of
-Definition~\ref{def:set-sem} are given in full in
+write either $\rho[\ol{\Phi := K}]$ or $\rho[\ol{\phi := K}]$ for the
+$\oCPO$ environment $\rho'$ such that $\rho' \phi_i = K_i$ for $i =
+1,...,n$ and $\rho' \phi = \rho \phi$ if $\phi \not \in \Phi$.  If
+$\rho$ is an $\oCPO$ environment, we write $\Eq_\rho$ for the $\oCPOR$
+environment (see Definition~\ref{def:reln-env}) such that $\Eq_\rho v
+= \Eq_{\rho v}$ for every type variable $v$. The categories
+$\oCPORT_k$ and relational interpretations appearing in the third
+clause of Definition~\ref{def:set-sem} are given in full in
 Section~\ref{sec:rel-interp}. We note that $\omega_1$-directed
@@
         \Phi \vdash G}\rho[\ol{\Phi := K}] \\ 
-      &\hspace{0.3in}|~\forall \ol{K = (K_0, K_1, K^*) : \oCPORT_k}.\\
-      &\hspace{0.4in}(\eta_{\overline{K_0}}, \eta_{\overline{K_1}})
+      &\hspace{0.3in}|~\forall \ol{K = (K^1, K^2, K^*) : \oCPORT_k}.\\
+      &\hspace{0.4in}(\eta_{\overline{K^1}}, \eta_{\overline{K^2}})
       : \oCPORsem{\Gamma; \Phi \vdash F}\Eq_{\rho}[\ol{\Phi := K}]
@@
 \C \to \C_i$ for $i = 1,...,n$, and a natural transformation $\alpha :
-F \to G$, the functorial action $\textit{Lan}_{\ol K} \alpha :
+F \to F'$, the functorial action $\textit{Lan}_{\ol K} \alpha :
 \textit{Lan}_{\ol K} F \to \textit{Lan}_{\ol K} F'$ of
@@
 consider an $\oCPO$ relation environment to map a type variable of
-arity $0$ to an $\oCPO$ relation.  We write $\rho[\ol{\alpha := R}]$
-for the $\oCPO$ relation environment $\rho'$ such that $\rho' \alpha_i
-\, = R_i$ for $i = 1,...,k$ and $\rho' \alpha = \rho\alpha$ if $\alpha
-\not \in \{\alpha_1,...,\alpha_k\}$.  If $\rho$ is an $\oCPO$ relation
-environment, we write $\pi_1 \rho$ and $\pi_2 \rho$ for the $\oCPO$
-environments mapping each type variable $\phi$ to the functors
-$(\rho\phi)^1$ and $(\rho\phi)^2$, respectively.
+arity $0$ to an $\oCPO$ relation. If $\Phi =
+\{\phi_1^{k_1},...,\phi_n^{k_n}\}$ and $\ol K = \{K_1,...,K_n\}$ where
+$K_i : [\oCPOR^{k_i},\oCPOR]_{\omega_1}$ for $i = 1,...,n$, then we
+write either $\rho[\ol{\Phi := K}]$ or $\rho[\ol{\phi := K}]$ for the
+$\oCPO$ relation environment $\rho'$ such that $\rho' \phi_i = K_i$
+for $i = 1,...,n$ and $\rho' \phi = \rho \phi$ if $\phi \not \in
+\Phi$. If $\rho$ is an $\oCPO$ relation environment, we write $\pi_1
+\rho$ and $\pi_2 \rho$ for the $\oCPO$ relation environments mapping
+each type variable $\phi$ to the functors $(\rho\phi)^1$ and
+$(\rho\phi)^2$, respectively.
 
@@
 \item for all $\overline{R : \oCPOR(A,B)}$ and morphisms $f$ in
-  $\oCPORenv$, $\pi_1(H^*f \,{\overline{R}}) = H^1 (\pi_1
-  f)\,{\overline{A}}$ and $\pi_2(H^*f \,{\overline{R}}) = H^2 (\pi_2
-  f)\,{\overline{B}}$
+  $\oCPORenv$, $\pi_1((H^*f)_{\overline{R}}) = (H^1 (\pi_1
+  f))_{\overline{A}}$ and $\pi_2((H^*f)_{\overline{R}}) = (H^2 (\pi_2
+  f))_{\overline{B}}$
 \item The action of $H$ on $\rho$ in $\oCPORenv$ is given by $H \rho = (H^1
@@
 {\em iii}) if $f : \rho \to \rho'$ and
-$R_1:\oCPOR(A_1,B_1),...,R_k:\oCPOR(A_k,B_k)$, then $(H^1(\pi_1
-f)\,{\ol{A}}\,x, H^2(\pi_2 f)\,{\ol{B}}\,y) \in H^*\rho'\,\ol{R}$
+$R_1:\oCPOR(A_1,B_1),...,R_k:\oCPOR(A_k,B_k)$, then $((H^1(\pi_1
+f))_{\ol{A}}\,x, (H^2(\pi_2 f))_{\ol{B}}\,y) \in H^*\rho'\,\ol{R}$
 whenever $(x, y) \in H^*\rho\,\ol{R}$. As before, the last condition
@@
   \rho)~|~\\ 
-  & \hspace{0.3in} \forall \ol{K = (K^0, K^1, K^*) : \oCPORT_{k}}.\\
-  & \hspace{0.4in} (t_{\ol{K^0}},t'_{\ol{K^1}}) \in
+  & \hspace{0.3in} \forall \ol{K = (K^1, K^2, K^*) : \oCPORT_{k}}.\\
+  & \hspace{0.4in} (t_{\ol{K^1}},t'_{\ol{K^2}}) \in
   (\oCPORsem{\Gamma; \Phi \vdash G}\rho[\ol{\Phi :=
@@
 whose actions on objects and morphisms are given in
-Definition~\ref{def:rel-sem-funcs}. Its action on a morphism $f : \rho
+Definition~\ref{def:rel-sem}. Its action on a morphism $f : \rho
 \to \rho'$ is the higher-order natural transformation $T_{H,f} :

### 2020-07-15 20-gadts/inf-rules.tex
added IEL section and started supp.tex

@@
   interpretations of $\Nat$-types are well-defined.}
+
+\new{Locally presentable categories are the closest approximation to
+  small complete categories. It's remarkable that local presentability
+  is sufficient not just to give semantics to GADTs, but is also all
+  that is needed to ensure that the model induced by that semantics is
+  actually parametric.}
 
@@
 \mcA \to \mcC$ is {\em $\omega_1$-cocontinuous} if it preserves
-$\omega_1$-directed colimits. We write $[\A,\C]_{\omega_1}$ for 
-the category of $\omega_1$-cocontinuous functors from $\A$ to $\C$,
-and $\C^\A$ for the category of {\em all} functors from $\A$ to $\C$.
+$\omega_1$-directed colimits. We write $[\A,\C]_{\omega_1}$ for the
+category of $\omega_1$-cocontinuous functors from $\A$ to $\C$, and
+$\C^\A$ for the category of {\em all} functors from $\A$ to $\C$.
 Since (co)limits of functors are computed pointwise, $C^\A$ has all
 (co)limits that $\C$ has, and (co)limits of (co)continuous functors
-are again (co)continuous. It follows that $[\A,\C]_{\omega_1}$ is
-($\omega_1$-)(co)complete whenever $\C$ is.
+are again (co)continuous. It follows that $[\A,\C]_{(\omega_1)}$ is
+($\omega_1$-)(co)complete whenever $\C$ is.  {\color{red} In
+  particular, if $F : [\C, \C]_{(\omega_1)}$ for a(n)
+  ($\omega_1$-)cocomplete category $\C$, then we can define $F^0 =
+  \Id$, $F^{\alpha + 1} = F \circ F^\alpha$ at successor ordinals, and
+  $F^{\bigcup_{\beta < \alpha} \beta} = \colim{\beta < \alpha}
+  F^\beta$ at limit ordinals.}
 
@@
 $T = (T^1,T^2,T^*) : \oCPORT_k \to \oCPORT_k$ is an endofunctor on
-$\oCPORT_k$ then we can define the $\oCPO$ relation transformer
-$\mu T$ to be $\colim{n < \omega_1}{T^n K_0}$.
-It is not hard to see that $\mu T$ is given explicitly as 
+$\oCPORT_k$ then we can define the $\oCPO$ relation transformer $\mu
+T$ to be $\colim{i < \omega_1}{T^i K_0}$. It is not hard to see that
+$\mu T$ is given explicitly as
 \begin{equation}\label{eq:mu}
   \mu T = (\mu T^1,\mu T^2,
-\colim{n < \omega_1}{(T^nK_0)^*})
+\colim{i < \omega_1}{(T^iK_0)^*})
 \end{equation}
@@
 
+\section{The Identity Extension Lemma}\label{sec:iel}
+
+The standard definition of the graph for a morphism $f : A \to B$ in
+$\set$ is the relation $\graph{f} : \rel(A,B)$ defined by $(x,y) \in
+\graph{f}$ iff $fx = y$. This definition naturally generalizes to
+associate to each natural transformation between $k$-ary functors on
+$\oCPO$ a $k$-ary $\oCPO$ relation transformer as follows.
+
+First, let $f : A \to B$ be a morphism in $\oCPO$. Then the graph of
+the underlying function $|f| : |A| \to |B|$ in $\set$ is an $\oCPO$
+relation.  Indeed, if $(a_i, b_i) : \graph{|f|}$ for all $i < \omega$
+is a chain, then $|f| a_i = b_i$ for all $i$, and consequently
+$|f|(\bigvee_{i < \omega} a_i) = \bigvee_{i < \omega} (|f| a_i) =
+\bigvee_{i < \omega} b_i$, i.e., $(\bigvee_{i < \omega} a_i,
+\bigvee_{i < \omega} b_i) : \graph{|f|}$. We therefore define the {\em
+  $\oCPO$ graph of $f$} in $\oCPO$ to be $\graph{f} =
+\graph{|f|}$. Note in particular that if $A : \oCPO$ then
+$\graph{\id_A}$ is an $\oCPO$ relation. We denote this $\oCPO$
+relation, called the {\em equality relation on $A$}, by $\Eq_A$. It
+coincides exactly with the equality relation on the underlying set
+$|A|$ of $A$.
+
+The notion of an $\oCPO$ graph relation naturally generalizes to
+associate to each natural transformation between $k$-ary functors on
+$\oCPO$ a $k$-ary $\oCPO$ relation transformer as follows. Recall
+that, since $\oCPO$ is a locally $\omega_1$-presentable category,
+Proposition 1.6.1 of~\cite{ar94} ensures that it has a (strong epi,
+mono) factorization system. We then have:
+
+\begin{dfn}\label{dfn:graph-nat-transf}
+If $F, G: \oCPO^k \to \oCPO$ and $\alpha : F \to G$ is a natural
+transformation, then the functor $\graph{\alpha}^*: \oCPOR^k \to
+\oCPOR$ is defined as follows. Given $R_1 : \oCPOR(A_1, B_1),...,R_k :
+\oCPOR(A_k,B_k)$, let $\iota_{R_i} : R_i \hookrightarrow A_i \times
+B_i$, for $i = 1,...,k$, be the inclusion of $R_i$ as a
+sub$\oCPO$ of $A_i \times B_i$, let $h_{\overline{A \times B}}$
+be the unique morphism making the diagram
+{\footnotesize\[\begin{tikzcd}[row sep = large] F\overline{A}
+  &F(\overline{A \times B}) \ar[l, "{F\overline{\pi_1}}"'] \ar[r,
+    "{F\overline{\pi_2}}"] \ar[d, dashed, "{h_{\overline{A \times
+          B}}}"] &F\overline{B} \ar[r, "{\alpha_{\ol{B}}}"]
+  &G\overline{B} \\ &F\overline{A} \times G\overline{B} \ar[ul,
+    "{\pi_1}"] \ar[urr, "{\pi_2}"']
+\end{tikzcd}\]}
+\noindent
+\!\!commute, and let $h_{\overline{R}} : F\overline{R} \to
+F\overline{A} \times G\overline{B}$ be $h_{\overline{A \times B}}
+\circ F\overline{\iota_R}$. Further, let $\alpha^\wedge\overline{R}$
+be the
+%If $f = e \circ m = e' \circ m'$ then $cod e = dom m$ is the same as
+%$cod e' = dom m'$ up to isomorphism. This follows easily from the
+%definition of a factorization system.
+subobject through which $h_{\overline{R}}$ is factorized by the
+(strong epi, mono) factorization system of $\oCPO$, as shown in the
+following diagram: {\footnotesize\[\begin{tikzcd} F\overline{R}
+  \ar[rr, "{h_{\overline{R}}}"] \ar[dr, twoheadrightarrow,
+    "{q_{\alpha^\wedge\overline{R}}}"'] &&F\overline{A} \times
+  G\overline{B} \\ &\alpha^\wedge\overline{R} \ar[ur, hookrightarrow,
+    "{\iota_{\alpha^\wedge\overline{R}}}"']
+\end{tikzcd}\]}
+
+\noindent
+Then $\alpha^\wedge\overline{R} : \oCPOR(F\overline{A},
+G\overline{B})$ by construction, so the action of $\graph{\alpha}^*$
+on objects of $\oCPOR$ can be given by $\langle \alpha \rangle^*
+\overline{(A,B,R)} = (F\overline{A}, G\overline{B},
+\iota_{\alpha^\wedge \overline{R}}\alpha^\wedge\overline{R})$. The
+action of $\graph{\alpha}^*$ on morphisms of $\oCPOR$ is given by
+$\graph{\alpha}^*\overline{(\beta, \beta')} = (F\overline\beta,
+G\overline\beta')$.
+\end{dfn}
+
+The next lemma shows that the data in
+Definition~\ref{dfn:graph-nat-transf} actually yield a relation
+transformer $\graph{\alpha} = (F, G, \graph{\alpha}^*)$. We call this
+the {\em $\oCPO$ graph relation transformer} for $\alpha$.
+
+\begin{lemma}\label{lem:graph-reln-functors}
+If $F,G : [\oCPO^k,\oCPO]_{\omega_1}$, and if $\alpha : F \to G$ is a natural
+transformation, then $\graph{\alpha}$ is in $\oCPORT_k$.
+\end{lemma}
+\begin{proof}
+Clearly, $\graph{\alpha}^*$ is $\omega_1$-cocontinuous, so
+$\graph{\alpha}^* : [\oCPOR^k,\oCPOR]_{\omega_1}$. Now, suppose
+$\overline{R : \oCPOR(A, B)}$, $\overline{S : \oCPOR(C, D)}$, and
+$\overline{(\beta, \beta') : R \to S}$. We want to show that there
+exists a morphism $\epsilon : \alpha^\wedge\overline{R} \to
+\alpha^\wedge\overline{S}$ such that the diagram on the left below
+commutes. Since $\ol{(\beta,\beta') : R \to S}$, there exist
+$\overline{\gamma : R \to S}$ such that each diagram in the middle
+commutes.  Moreover, since both $h_{\overline{C \times D}} \circ
+F(\overline{\beta \times \beta'})$ and $(F\overline{\beta} \times
+G\overline{\beta'}) \circ h_{\overline{A \times B}}$ make the diagram
+on the right commute, they must be equal.
+\begin{figure*}[ht]
+\vspace*{-0.1in}
+  \hspace*{-0.85in}
+  \begin{minipage}[b]{0.25\linewidth}
+ {\small    \[
+    \begin{tikzcd}
+        \alpha^\wedge\overline{R}
+        \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{R}}}"]
+        \ar[d, "{\epsilon}"']
+        & F\overline{A} \times G\overline{B}
+        \ar[d, "{F\overline{\beta} \times G\overline{\beta'}}"] \\
+        \alpha^\wedge\overline{S}
+        \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{S}}}"']
+        & F\overline{C} \times G\overline{D}
+    \end{tikzcd}
+    \]}
+\end{minipage}
+\begin{minipage}[b]{0.25\linewidth}
+{\small    \[
+    \begin{tikzcd}
+        R_i
+        \ar[d, "{\gamma_i}"']
+        \ar[r, hookrightarrow, "{\iota_{R_i}}"]
+        &A_i \times B_i
+        \ar[d, "{\beta_i \times \beta'_i}"] \\
+        S_i
+        \ar[r, hookrightarrow, "{\iota_{S_i}}"]
+        &C_i \times D_i
+    \end{tikzcd}
+    \]}
+\end{minipage}
+\begin{minipage}[b]{0.25\linewidth}
+{\footnotesize \[
+      \begin{tikzcd}[row sep = large]
+          F\overline{C}
+          &F\overline{C} \times F\overline{D}
+          \ar[l, "{\pi_1}"'] \ar[r, "{\pi_2}"]
+          &F\overline{D}
+          \ar[r, "{\alpha_{\ol{D}}}"]
+          &G\overline{D}\\
+          &F(\overline{A \times B})
+          \ar[u, dashed, "{\exists !}"]
+          \ar[ul, "{F\ol{\pi_1} \circ F(\overline{\beta \times \beta'})}"]
+          \ar[urr, "{\alpha_{\ol{D}} \circ F\ol{\pi_2} \circ F(\overline{\beta \times \beta'})}"']
+      \end{tikzcd}
+      \]}
+\end{minipage}
+\end{figure*}
+We therefore get that the right-hand square in the diagram on the left
+below commutes, and thus that the entire diagram does as well.
+Finally, by the left-lifting property of
+$q_{\alpha^\wedge\overline{R}}$ with respect to
+$\iota_{\alpha^\wedge\overline{S}}$ given by the (strong epi, mono)
+factorization system there exists an $\epsilon$ such that the diagram
+on the right below commutes.
+\[
+      \begin{tikzcd}
+          F\overline{R}
+          \ar[d, "{F\overline{\gamma}}"']
+          \ar[r, hookrightarrow, "{F\overline{\iota_R}}"]
+          \ar[rr, bend left, "{h_{\overline{R}}}"]
+          &F(\overline{A \times B})
+          \ar[d, "{F(\overline{\beta \times \beta'})}"]
+          \ar[r, "{h_{\overline{A \times B}}}"]
+          &F\overline{A} \times G\overline{B}
+          \ar[d, "{F\overline{\beta} \times F\overline{\beta'}}"] \\
+          F\overline{S}
+          \ar[r, hookrightarrow, "{F\overline{\iota_S}}"']
+          \ar[rr, bend right, "{h_{\overline{S}}}"']
+          &F(\overline{C \times D})
+          \ar[r, "{h_{\overline{C \times D}}}"']
+          &F\overline{C} \times G\overline{D}
+      \end{tikzcd}
+\qquad
+      \begin{tikzcd}
+          F\overline{R}
+          \ar[d, "{F\overline{\gamma}}"']
+          \ar[r, twoheadrightarrow, "{q_{\alpha^\wedge\overline{R}}}"]
+          &\alpha^\wedge\overline{R}
+          \ar[d, dashed, "{\epsilon}"]
+          \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{R}}}"]
+          &F\overline{A} \times G\overline{B}
+          \ar[d, "{F\overline{\beta} \times G\overline{\beta'}}"] \\
+          F\overline{S}
+          \ar[r, twoheadrightarrow, "{q_{\alpha^\wedge\overline{S}}}"']
+          &\alpha^\wedge\overline{S}
+          \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{S}}}"']
+          &F\overline{C} \times G\overline{D}
+      \end{tikzcd}
+\]
+\end{proof}
+
+If $f : A \to B$ is a morphism in $\oCPO$ then the definition of the
+$\oCPO$ graph relation transformer $\graph{f}$ for $f$ as a natural
+transformation between $0$-ary functors $A$ and $B$ coincides with the
+definition of $\graph{f}$ for $f$ as a morphism in $\oCPO$ given in
+the second paragraph of this section.  As a result, $\oCPO$ graph
+relation transformers are a reasonable extension of $\oCPO$ graph
+relations to functors.
+
+To prove the IEL, we will need to know that the equality $\oCPO$
+relation transformer preserves equality relations in $\oCPOR$; see
+Equation~\ref{eq:eq-pres-eq} below. This will follow from the next
+lemma, which shows how to compute the action of an $\oCPO$ graph
+relation transformer on any $\oCPO$ graph relation.
+
+\begin{lemma}\label{lem:eq-reln-equalities}
+If $\alpha : F \to G$ is a morphism in $[\oCPO^k, \oCPO]_{\omega_1}$
+and $f_1: A_1 \to B_1, ..., f_k : A_k \to B_k$, then $\graph{\alpha}^*
+\graph{\overline{f}} = \graph{G \ol{f} \circ \alpha_{\ol{A}}} =
+\graph{\alpha_{\ol{B}} \circ F \ol{f}}$.
+\end{lemma}
+\begin{proof}
+Since $h_{\overline{A \times B}}$ is the unique morphism making the
+bottom triangle of the diagram on the left below commute, and since
+$h_{\graph{\overline{f}}} = h_{\overline{A \times B}} \circ F
+\,\ol{\iota_{\graph{f}}} = h_{\overline{A \times B}} \circ F
+\overline{\graph{\id_A, f}}$, the universal property of the
+product depicted in the diagram on the right gives
+$h_{\graph{\overline{f}}} = \graph{\id_{F \ol{A}}, \alpha_{\ol{B}}
+\circ F\ol{f}} : F \ol{A} \to F \ol{A} \times G \ol{B}$.
+\begin{figure*}[ht]
+  \vspace*{-0.15in}
+  \begin{minipage}[b]{0.45\linewidth}
+{\footnotesize
+\[\begin{tikzcd}[row sep = large]
+        &F\overline{A}
+        \ar[d, "{F \overline{\graph{\id_A, f}}}" description]
+        \ar[dl, equal]
+        \ar[dr, "{F\ol{f}}"]\\
+        F\overline{A}
+        &F(\overline{A \times B})
+        \ar[l, "{F\overline{\pi_1}}"']
+        \ar[r, "{F\overline{\pi_2}}"]
+        \ar[d, "{h_{\overline{A \times B}}}"]
+        &F\overline{B}
+        \ar[r, "{\alpha_{\ol{B}}}"]
+        &G\overline{B}\\
+        &F\overline{A} \times G\overline{B}
+        \ar[ul, "{\pi_1}"] \ar[urr, "{\pi_2}"']
+\end{tikzcd}\]}
+\end{minipage}
+  \begin{minipage}[b]{0.45\linewidth}
+{\footnotesize
+\[
+      \begin{tikzcd}[row sep = large]
+          F\overline{A}
+          &F\overline{A} \times G\overline{B}
+          \ar[l, "{\pi_1}"'] \ar[r, "{\pi_2}"]
+          &G\overline{B}\\
+          &F\overline{A}
+          \ar[u, dashed, "{\exists !}"]
+          \ar[ul, equal]
+          \ar[r, "{F\ol{f}}"']
+          &F{\ol{B}}
+          \ar[u, "{\alpha_{\ol{B}}}"']
+      \end{tikzcd}
+      \]}
+\end{minipage}
+\end{figure*}
+
+\vspace*{-0.1in}
+
+\noindent
+Moreover, $\graph{\id_{F \ol{A}}, \alpha_{\ol{B}} \circ F\ol{f}}$ is a
+monomorphism in $\oCPO$ because $\id_{F \ol{A}}$ is, so its (strong
+epi, mono) factorization gives $\iota_{\alpha^\wedge
+  \graph{\overline{f}}} = \langle \id_{F \ol{A}}, \alpha_{\ol{B}}
+\circ F\ol{f} \rangle$, and thus that $\alpha^\wedge
+\graph{\overline{f}} = F\overline{A}$.
+%If $e \circ m = e \circ m'$ then $m = m'$ is trivial from the
+%definition of a factorization system.
+Then $\iota_{\alpha^\wedge \graph{\overline{f}}} \alpha^\wedge
+\graph{\overline{f}} = \graph{\id_{F \ol{A}}, \alpha_{\ol{B}} \circ
+  F\ol{f}} (F \ol{A}) = \graph{ \alpha_{\ol{B}} \circ F\ol{f}
+}^*$. Then $\graph{ \alpha }^* \graph{ \overline{f} } =
+(F\overline{A}, G\overline{B}, \iota_{\alpha^\wedge
+  \graph{\overline{f}}}\, \alpha^\wedge \graph{\overline{f}}) =
+(F\overline{A}, G\overline{B}, \graph{ \alpha_{\ol{B}} \circ F\ol{f}
+}^*) = \graph{ \alpha_{\ol{B}} \circ F\ol{f} }$, and, finally,
+$\alpha_{\ol{B}} \circ F\ol{f} = G\ol{f} \circ \alpha_{\ol{A}}$ by
+naturality of $\alpha$.
+\end{proof}
+
+The {\em equality $\oCPO$ relation transformer} on $F :
+[\oCPO^k,\oCPO]_{\omega_1}$ is defined to be $\Eq_F =
+\graph{\id_{F}}$. Specifically, $\Eq_F = (F, F, \Eq_F^*)$ with
+$\Eq_F^* = \graph{\id_{F}}^*$, and Lemma~\ref{lem:eq-reln-equalities}
+indeed ensures that
+\begin{equation}\label{eq:eq-pres-eq}
+\Eq^*_F \ol{\Eq_A}
+= \graph{\id_F}^* \graph{\id_{\ol{A}}}
+= \graph{F \id_{\ol{A}} \circ (\id_F)_{\ol{A}}}
+= \graph{\id_{F\ol{A}} \circ \id_{F\ol{A}}}
+= \graph{\id_{F\ol{A}}}
+= \Eq_{F\ol{A}}
+\end{equation}
+for all $\ol{A : \oCPO}$.  Graph $\oCPO$ relation transformers in
+general, and equality $\oCPO$ relation transformers in particular,
+extend to $\oCPO$ relation environments in the obvious ways.  Indeed,
+if $\rho, \rho' : \oCPOenv$ and $f : \rho \to \rho'$, then the {\em
+  graph $\oCPO$ relation environment} $\graph{f}$ is defined pointwise
+by $\graph{f} \phi = \graph{f \phi}$ for every $\phi$, which entails
+that $\pi_1 \graph{f} = \rho$ and $\pi_2 \graph{f} = \rho'$. In
+particular, the {\em equality $\oCPO$ relation environment} $\Eq_\rho$
+is defined to be $\graph{\id_{\rho}}$, which entails that $\Eq_\rho
+\phi = \Eq_{\rho \phi}$ for every $\phi$.  With these definitions in
+hand, we can state and prove both an Identity Extension Lemma and a
+Graph Lemma for our calculus.
+
+\begin{thm}[IEL]\label{thm:iel}
+  If $\rho : \oCPOenv$ and $\Gamma; \Phi \vdash F$ then
+  $\oCPORsem{\Gamma; \Phi \vdash F} \Eq_\rho = \Eq_{\oCPOsem{\Gamma;
+      \Phi \vdash F}\rho}$.
+\end{thm}
+The proof is by induction on the structure of $F$. Only the $\Nat$,
+application, and fixpoint cases are non-routine. The latter two use
+Lemma~\ref{lem:eq-reln-equalities}. The fixpoint case also uses the
+observation that, for every $i < \omega_1$, the following intermediate
+results can be proved by simultaneous induction with
+Theorem~\ref{thm:iel}: for any $H$, $\rho$, $A$, and any subformula
+$J$ of $H$, both
+$T^i_{H,\Eq_{\rho}} K_0\, \ol{\Eq_A}
+= (\Eq_{(T^\set_{H,\rho})^i K_0})^*
+\ol{\Eq_A}$\;
+and\;
+\[\begin{array}{ll}
+&  \oCPORsem{\Gamma; \Phi, \phi, \ol{\alpha} \vdash J} \Eq_{\rho} [\phi
+  := T^i_{H,\Eq_{\rho}} K_0] \overline{[\alpha := \Eq_A]}\\
+=  &
+\oCPORsem{\Gamma; \Phi, \phi,
+  \ol{\alpha} \vdash J} \Eq_{\rho} [\phi := \Eq_{(T^\oCPO_{H,\rho})^i K_0}]
+\overline{[\alpha := \Eq_A]}
+\end{array}\]
+hold. With these results in hand, the proof follows easily. It is
+given in detail in the accompanying anonymous supplementary
+material. As noted there, if functorial variables of arity greater
+than $0$ were allowed to appear in the bodies of $\mu$-types, then the
+IEL would fail.
+
+\begin{lemma}[Graph Lemma]\label{lem:graph}
+If $\rho, \rho' : \oCPOenv$, $f : \rho \to \rho'$, and $\Gamma; \Phi
+\vdash F$, then $\graph{\oCPOsem{\Gamma; \Phi \vdash F} f} =
+\oCPORsem{\Gamma; \Phi \vdash F}\graph{f}$.
+\end{lemma}
+\begin{proof}
+Applying Lemma~\ref{lem:rel-transf-morph} to the morphisms $(f,
+\id_{\rho'}) : \graph{f} \to \Eq_{\rho'}$ and $(\id_{\rho}, f) :
+\Eq_{\rho} \to \graph{f}$ of relation environments gives that
+\begin{multline*}
+(\oCPOsem{\Gamma; \Phi \vdash F}f, \oCPOsem{\Gamma; \Phi \vdash F}\id_{\rho'})
+  = \oCPORsem{\Gamma; \Phi \vdash F} (f, \id_{\rho'}) \\
+:\oCPORsem{\Gamma; \Phi \vdash F}\graph{f} \to 
+  \oCPORsem{\Gamma; \Phi \vdash F}\Eq_{\rho'}
+\end{multline*}
+and
+\begin{multline*}
+(\oCPOsem{\Gamma; \Phi \vdash F}\id_{\rho}, \oCPOsem{\Gamma; \Phi \vdash F}f)
+  = \oCPORsem{\Gamma; \Phi \vdash F} (\id_{\rho}, f) \\
+: \oCPORsem{\Gamma; \Phi \vdash F}\Eq_{\rho} 
+  \to \oCPORsem{\Gamma; \Phi \vdash F}\graph{f}
+\end{multline*}
+Expanding the first equation gives that if $(x,y) \in \oCPORsem{\Gamma;
+  \Phi \vdash F}\graph{f}$ then \[(\oCPOsem{\Gamma; \Phi \vdash F} f\,
+x, \oCPOsem{\Gamma; \Phi \vdash F}\id_{\rho'}\, y) \in \oCPORsem{\Gamma;
+  \Phi \vdash F}\Eq_{\rho'}\] So $\oCPOsem{\Gamma; \Phi \vdash
+  F}\id_{\rho'}\, y = \id_{\oCPOsem{\Gamma; \Phi \vdash F}\rho'}\, y =
+y$ and $\oCPORsem{\Gamma; \Phi \vdash F}\Eq_{\rho'} =
+\Eq_{\oCPOsem{\Gamma; \Phi \vdash F}\rho'}$, and if $(x,y) \in
+\oCPORsem{\Gamma; \Phi \vdash F}\graph{f}$ then $(\oCPOsem{\Gamma; \Phi
+  \vdash F} f\, x, y) \in \Eq_{\oCPOsem{\Gamma; \Phi \vdash F}\rho'}$,
+i.e., $\oCPOsem{\Gamma; \Phi \vdash F} f\, x = y$, i.e., $(x, y) \in
+\graph{\oCPOsem{\Gamma; \Phi \vdash F} f}$.  So, we have that
+$\oCPORsem{\Gamma; \Phi \vdash F}\graph{f} \subseteq
+\graph{\oCPOsem{\Gamma; \Phi \vdash F}f}$.  Expanding the second
+equation gives that if $x \in \oCPOsem{\Gamma; \Phi \vdash F}\rho$ then
+$(\oCPOsem{\Gamma; \Phi \vdash F}\id_{\rho}\, x,$\\
+$\oCPOsem{\Gamma; \Phi
+  \vdash F} f\, x) \in \oCPORsem{\Gamma; \Phi \vdash F}\graph{f}$.  Then
+$\oCPOsem{\Gamma; \Phi \vdash F}\id_{\rho}\, x = \id_{\oCPOsem{\Gamma;
+    \Phi \vdash F}\rho} x = x$, so for any $x \in \oCPOsem{\Gamma; \Phi
+  \vdash F}\rho$ we have that $(x, \oCPOsem{\Gamma; \Phi \vdash F}f\,
+x) \in \oCPORsem{\Gamma; \Phi \vdash F}\graph{f}$.  Moreover, $x \in
+\oCPOsem{\Gamma; \Phi \vdash F}\rho$ if and only if $(x,
+\oCPOsem{\Gamma; \Phi \vdash F} f\, x) \in \graph{\oCPOsem{\Gamma; \Phi
+    \vdash F}f}$ and, if $x \in \oCPOsem{\Gamma; \Phi \vdash F}\rho$
+then $(x, \oCPOsem{\Gamma; \Phi \vdash F} f\, x) \in \oCPORsem{\Gamma;
+  \Phi \vdash F} \graph{f}$, so if $(x, \oCPOsem{\Gamma; \Phi \vdash F}
+f\, x) \in \graph{\oCPOsem{\Gamma; \Phi \vdash F}f}$ then $(x,
+\oCPOsem{\Gamma; \Phi \vdash F} f\, x) \in \oCPORsem{\Gamma; \Phi \vdash
+  F} \graph{f}$, i.e., $\graph{\oCPOsem{\Gamma; \Phi \vdash F}f}
+\subseteq \oCPORsem{\Gamma; \Phi \vdash F} \graph{f}$.
+\end{proof}
+
 \vspace*{1in}

### 2020-07-16 20-gadts/inf-rules.tex
added limit ordinal cases to supp.tex

@@
 The proof is by induction on the structure of $F$. Only the $\Nat$,
-application, and fixpoint cases are non-routine. The latter two use
+application, and fixpoint {\color{red} and $\Lan$?} cases are
+non-routine. The latter two use
 Lemma~\ref{lem:eq-reln-equalities}. The fixpoint case also uses the
@@
 Theorem~\ref{thm:iel}: for any $H$, $\rho$, $A$, and any subformula
-$J$ of $H$, both
-$T^i_{H,\Eq_{\rho}} K_0\, \ol{\Eq_A}
-= (\Eq_{(T^\set_{H,\rho})^i K_0})^*
-\ol{\Eq_A}$\;
-and\;
+$J$ of $H$, both $T^i_{H,\Eq_{\rho}} K_0\, \ol{\Eq_A} =
+(\Eq_{(T^\set_{H,\rho})^i K_0})^* \ol{\Eq_A}$\; and\;
 \[\begin{array}{ll}

### 2020-07-17 20-gadts/inf-rules.tex
completed proof that obj interps of Nat types are wCPOs

@@
 are again (co)continuous. It follows that $[\A,\C]_{(\omega_1)}$ is
-($\omega_1$-)(co)complete whenever $\C$ is.  {\color{red} In
-  particular, if $F : [\C, \C]_{(\omega_1)}$ for a(n)
-  ($\omega_1$-)cocomplete category $\C$, then we can define $F^0 =
-  \Id$, $F^{\alpha + 1} = F \circ F^\alpha$ at successor ordinals, and
-  $F^{\bigcup_{\beta < \alpha} \beta} = \colim{\beta < \alpha}
-  F^\beta$ at limit ordinals.}
+($\omega_1$-)(co)complete whenever $\C$ is. In particular, if $F :
+[\C, \C]_{(\omega_1)}$ for a(n) ($\omega_1$-)cocomplete category $\C$,
+then we can define $F^0 = \Id$, $F^{\alpha + 1} = F \circ F^\alpha$ at
+successor ordinals, and $F^{\bigcup_{\beta < \alpha} \beta} =
+\colim{\beta < \alpha} F^\beta$ at limit ordinals.
 
@@
 
-\new{Need to modify to take equality conditions into account.}  Now,
-if $\C$ is any category, then given functors $F,G : \C \to \oCPO$ and
-natural transformations $\eta, \eta' : F \to G$, we define $\eta \leq
-\eta'$ iff $\eta_c \leq \eta'_c$ for all $c \in \C$, i.e., iff
+%Now, if $\C$ is any category, then given functors $F,G : \C \to \oCPO$
+%and natural transformations $\eta, \eta' : F \to G$, we define $\eta
+%\leq \eta'$ iff $\eta_c \leq \eta'_c$ for all $c \in \C$, i.e., iff
+%$\eta_c\,x \leq \eta'_c\,x$ in $Gc$ for all $c \in \C$ and $x \in Fc$.
+%If $(\eta_i)_{i < \omega}$ is a chain in $\{\eta : F \to G\}$, then
+%the family of morphisms $(\bigvee_{i < \omega} \eta_i)_c = \lambda
+%x. \bigvee_{i < \omega} ((\eta_i)_c\,x) : Fc \to Gc$ defines a natural
+%transformation $\bigvee_{i < \omega} \eta_i : F \to G$, and this
+%natural transformation is clearly the supremum of $(\eta_i)_{i <
+%  \omega}$ in $\{\eta : F \to G\}$. We therefore have that $\{\eta : F
+%\to G\}$ is itself an $\oCPO$.
+%Letting $F = \lambda \ol
+%K. \setsem{\Gamma;\Phi \vdash F}\rho[\ol{\Phi := K}]$ and $G = \lambda
+%\ol K. \setsem{\Gamma;\Phi \vdash G}\rho[\ol{\Phi := K}]$, we have
+%that $\oCPOsem{\Gamma; \emptyset \vdash \Nat^\Phi F\,G}\rho$ is indeed
+%an $\oCPO$.
+
+Now, if $\C$ is any category, then given functors $F,G : \C \to \oCPO$
+and natural transformations $\eta, \eta' : F \to G$, we define $\eta
+\leq \eta'$ iff $\eta_c \leq \eta'_c$ for all $c \in \C$, i.e., iff
 $\eta_c\,x \leq \eta'_c\,x$ in $Gc$ for all $c \in \C$ and $x \in Fc$.
@@
   \omega}$ in $\{\eta : F \to G\}$. We therefore have that $\{\eta : F
-\to G\}$ is itself an $\oCPO$.  Letting $F = \lambda \ol
-K. \setsem{\Gamma;\Phi \vdash F}\rho[\ol{\Phi := K}]$ and $G = \lambda
-\ol K. \setsem{\Gamma;\Phi \vdash G}\rho[\ol{\Phi := K}]$, we have
-that $\oCPOsem{\Gamma; \emptyset \vdash \Nat^\Phi F\,G}\rho$ is indeed
-an $\oCPO$.
+\to G\}$ is itself an $\oCPO$.  To show that $\oCPOsem{\Gamma;
+  \emptyset \vdash \Nat^\Phi \,F\,G}\rho$ is an $\oCPO$, we then still
+need to prove that, for any chain of natural transformations
+$(\eta_i)_{i < \omega}$ in $\oCPOsem{\Gamma; \emptyset \vdash
+  \Nat^\Phi \,F\,G}\rho$, $\bigvee_{i < \omega} \eta_i$ is again in
+$\oCPOsem{\Gamma; \emptyset \vdash \Nat^\Phi \,F\,G}\rho$.  By
+hypothesis, each $\eta_i$ is such that, for all $\ol{K = (K^1, K^2,
+  K^*) : \oCPORT_k}$,
+\[((\eta_i)_{\overline{K^1}},
+(\eta_i)_{\overline{K^2}}) : \oCPORsem{\Gamma; \Phi \vdash
+  F}\Eq_{\rho}[\ol{\Phi := K}] \rightarrow \oCPORsem{\Gamma; \Phi
+  \vdash G}\Eq_{\rho}[\ol{\Phi := K}]\] is a morphism in
+$\oCPOR$. To see that, for all $\ol{K = (K^1, K^2, K^*) : \oCPORT_k}$,
+%\[((\bigvee_{i < \omega} \eta_i)_{\overline{K^1}}, (\bigvee_{i <
+%%  \omega} \eta_i)_{\overline{K^2}}) : \oCPORsem{\Gamma; \Phi \vdash
+%  F}\Eq_{\rho}[\ol{\Phi := K}] \rightarrow \oCPORsem{\Gamma; \Phi
+%  \vdash G}\Eq_{\rho}[\ol{\Phi := K}]\] is again a morphism in
+%$\oCPOR$, we observe that
+$((\bigvee_{i < \omega} \eta_i)_{\overline{K^1}}, (\bigvee_{i <
+  \omega} \eta_i)_{\overline{K^2}})$ is a morphism from
+$\oCPORsem{\Gamma; \Phi \vdash F}\Eq_{\rho}[\ol{\Phi := K}]$ to
+$\oCPORsem{\Gamma; \Phi \vdash G}\Eq_{\rho}[\ol{\Phi := K}]$ in
+$\oCPOR$, we simply observe that
+\[((\bigvee_{i < \omega} \eta_i)_{\overline{K^1}}, (\bigvee_{i < \omega}
+\eta_i)_{\overline{K^2}}) = (\bigvee_{i < \omega}
+(\eta_i)_{\overline{K^1}}, \bigvee_{i < \omega}
+(\eta_i)_{\overline{K^2}}) = \bigvee_{i < \omega}
+((\eta_i)_{\overline{K^1}}, (\eta_i)_{\overline{K^2}})\] so
+$((\bigvee_{i < \omega} \eta_i)_{\overline{K^1}}, (\bigvee_{i <
+  \omega} \eta_i)_{\overline{K^2}})$ is the sup of a chain of
+morphisms from $\oCPORsem{\Gamma; \Phi \vdash F}\Eq_{\rho}[\ol{\Phi :=
+    K}]$ to $\oCPORsem{\Gamma; \Phi \vdash G}\Eq_{\rho}[\ol{\Phi :=
+    K}]$ in $\oCPOR$ and thus is itself such a morphism. Taken
+together, this all shows that $\oCPOsem{\Gamma; \emptyset \vdash
+  \Nat^\Phi \,F\,G}\rho$ is indeed an $\oCPO$.
 

### 2020-07-23 20-gadts/inf-rules.tex
relations on wCPO?

@@
 $\omega$-complete partial orders. We denote the category of
-$\omega$-complete partial orders ($\oCPO$s) and their sup-preserving
-morphisms by $\omega$CPO. The underlying set of an $\oCPO$ $A$ is
-denoted $|A|$.  The category $\oCPOR$ of $\oCPO$ relations has
-as its objects triples $(A,B,R)$, where $A, B: \oCPO$ and $R:
-\rel(|A|, |B|)$ and $(\bigvee_{i < \omega} a_i, \bigvee_{i < \omega}
-b_i) \in R$ whenever $(a_i)_{i < \omega}$ and $(b_i)_{i < \omega}$ are
-chains in $A$ and $B$, respectively, such that $(a_i,b_i) \in R$ for
-all $i$. The category $\oCPOR$ has as its morphisms from $(A,B,R)$ to
-$(A',B',R')$ pairs $(f : A \to A',g : B \to B')$ of morphisms in
-$\oCPO$ such that $(f\,a, g\,b) \in R'$ whenever $(a,b) \in R$. We
-note that if $(f,g) : (A,B,R) \to (A',B',R')$ and $(a_i)_{i < \omega}$
-and $(b_i)_{i < \omega}$ are chains in $A$ and $B$, respectively, then
-$ (f \, (\bigvee_{i < \omega} a_i), g \, (\bigvee_{i < \omega} b_i)) =
-(\bigvee_{i < \omega} (f a_i), \bigvee_{i < \omega} (g b_i)) \in R'$
-necessarily holds.  We write $R : \oCPOR(A,B)$ in place of $(A,B,R):
-\oCPOR$ when convenient. If $R : \oCPOR(A,B)$ then we write $\pi_1 R$
-and $\pi_2 R$ for the {\em domain} $A$ of $R$ and the {\em codomain}
-$B$ of $R$, respectively. If $A : \oCPO$, then we write $\Eq_A =
-(A,A,\Eq_{|A|})$ for the {\em equality relation} on $A$.
+$\omega$-complete partial orders ($\oCPO$s) and their {\color{blue}
+  monotone?} sup-preserving morphisms by $\omega$CPO. The underlying
+set of an $\oCPO$ $A$ is denoted $|A|$.  The category $\oCPOR$ of
+$\oCPO$ relations has as its objects triples $(A,B,R)$, where $A, B:
+\oCPO$; $R: \rel(|A|, |B|)$; {\color{blue} $(a,b) \in R, (a',b') \in
+  R$, and $a \leq a'$ imply $b \leq b'$;} and $(\bigvee_{i < \omega}
+a_i, \bigvee_{i < \omega} b_i) \in R$ whenever $(a_i)_{i < \omega}$
+and $(b_i)_{i < \omega}$ are chains in $A$ and $B$, respectively, such
+that $(a_i,b_i) \in R$ for all $i$. The category $\oCPOR$ has as its
+morphisms from $(A,B,R)$ to $(A',B',R')$ pairs $(f : A \to A',g : B
+\to B')$ of morphisms in $\oCPO$ such that $(f\,a, g\,b) \in R'$
+whenever $(a,b) \in R$. We note that if $(f,g) : (A,B,R) \to
+(A',B',R')$ and $(a_i)_{i < \omega}$ and $(b_i)_{i < \omega}$ are
+chains in $A$ and $B$, respectively, then $ (f \, (\bigvee_{i <
+  \omega} a_i), g \, (\bigvee_{i < \omega} b_i)) = (\bigvee_{i <
+  \omega} (f a_i), \bigvee_{i < \omega} (g b_i)) \in R'$ necessarily
+holds.  We write $R : \oCPOR(A,B)$ in place of $(A,B,R): \oCPOR$ when
+convenient. If $R : \oCPOR(A,B)$ then we write $\pi_1 R$ and $\pi_2 R$
+for the {\em domain} $A$ of $R$ and the {\em codomain} $B$ of $R$,
+respectively. If $A : \oCPO$, then we write $\Eq_A = (A,A,\Eq_{|A|})$
+for the {\em equality relation} on $A$.
 

### 2020-07-23 20-gadts/inf-rules.tex
relations on wCPO again?

--- 2020-07-17 20-gadts/inf-rules.tex
@@
 $\omega$-complete partial orders. We denote the category of
-$\omega$-complete partial orders ($\oCPO$s) and their sup-preserving
-morphisms by $\omega$CPO. The underlying set of an $\oCPO$ $A$ is
-denoted $|A|$.  The category $\oCPOR$ of $\oCPO$ relations has
-as its objects triples $(A,B,R)$, where $A, B: \oCPO$ and $R:
-\rel(|A|, |B|)$ and $(\bigvee_{i < \omega} a_i, \bigvee_{i < \omega}
-b_i) \in R$ whenever $(a_i)_{i < \omega}$ and $(b_i)_{i < \omega}$ are
-chains in $A$ and $B$, respectively, such that $(a_i,b_i) \in R$ for
-all $i$. The category $\oCPOR$ has as its morphisms from $(A,B,R)$ to
+$\omega$-complete partial orders ($\oCPO$s) and their {\color{blue}
+  monotone?} sup-preserving morphisms by $\omega$CPO. The underlying
+set of an $\oCPO$ $A$ is denoted $|A|$. The category $\oCPOR$ of
+$\oCPO$ relations has as its objects triples $(A,B,R)$, where $A, B:
+\oCPO$; $R: \rel(|A|, |B|)$; {\color{blue} $(a,b) \in R$, $a \leq a'$,
+  and $b \leq b'$ imply $(a',b') \in R$; so next part is redundant?}
+and $(\bigvee_{i < \omega} a_i, \bigvee_{i < \omega} b_i) \in R$
+whenever $(a_i)_{i < \omega}$ and $(b_i)_{i < \omega}$ are chains in
+$A$ and $B$, respectively, such that $(a_i,b_i) \in R$ for all
+$i$. The category $\oCPOR$ has as its morphisms from $(A,B,R)$ to
 $(A',B',R')$ pairs $(f : A \to A',g : B \to B')$ of morphisms in

### 2020-07-24 20-gadts/lan-prog.tex
lan programming note

% double-blind review submission, w/o CCS and ACM Reference (max
%
% submission space)
\documentclass[acmsmall,review,anonymous]{acmart}
\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

%% Journal information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{POPL} % CONF = POPL or ICFP or OOPSLA
\acmArticle{1}
\acmYear{2020}
\acmMonth{1}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2018}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations
%\citestyle{acmnumeric}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Note: Authors migrating a paper from PACMPL format to traditional
%% SIGPLAN proceedings format must update the '\documentclass' and
%% topmatter commands above; see 'acmart-sigplanproc-template.tex'.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage[utf8]{inputenc}
\usepackage{ccicons}
\usepackage{verbatim}

\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amscd}
%\usepackage{MnSymbol}
\usepackage{xcolor}

\usepackage{bbold}
\usepackage{url}
\usepackage{upgreek}
%\usepackage{stmaryrd}

\usepackage{lipsum}
\usepackage{tikz-cd}
\usetikzlibrary{cd}
\usetikzlibrary{calc}
\usetikzlibrary{arrows}

\usepackage{bussproofs}
\EnableBpAbbreviations

\input{macros}
\input{mytheorem}

\theoremstyle{definition}
\newtheorem{exmpl}{Example}

\renewcommand{\greyout}[1]{} %{{\color{gray} {#1}}} -- toggle to remove greyed text

\newcommand{\ininv}[2]{(\tin^{-1}_{\onet +
  \beta \times \phi (\phi\beta)})_{#1}\, #2}
\newcommand{\emptyfun}{{[]}}
\newcommand{\cal}{\mathcal}
%\newcommand{\fold}{\mathit{fold}}
\newcommand{\F}{\mathcal{F}}
\renewcommand{\G}{\mathcal{G}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\B}{\mathcal{B}}
\renewcommand{\P}{\mathcal{A}}
\newcommand{\pred}{\mathsf{Fam}}
\newcommand{\env}{\mathsf{Env}}
\newcommand{\set}{\mathsf{Set}}
\renewcommand{\S}{\mathcal S}
\renewcommand{\C}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\A}{\mathcal{A}}
\renewcommand{\id}{\mathit{id}}
\newcommand{\map}{\mathsf{map}}
\newcommand{\pid}{\underline{\mathit{id}}}
\newcommand{\pcirc}{\,\underline{\circ}\,}
\newcommand{\pzero}{\underline{0}}
\newcommand{\pone}{\underline{1}}
\newcommand{\psum}{\,\underline{+}\,}
\newcommand{\pinl}{\underline{\mathit{inL}}\,}
\newcommand{\pinr}{\underline{\mathit{inR}}\,}
\newcommand{\ptimes}{\,\underline{\times}\,}
\newcommand{\ppi}{\underline{\pi_1}}
\newcommand{\pppi}{\underline{\pi_2}}
\newcommand{\pmu}{\underline{\mu}}
\newcommand{\semmap}{\mathit{map}}
\newcommand{\subst}{\mathit{subst}}

\newcommand{\tb}[1]{~~ \mbox{#1} ~~}
\newcommand{\listt}[1]{(\mu \phi. \lambda \beta . \onet + \beta \times
  \phi \beta) #1} 
\newcommand{\filtype}{\Nat^\emptyset 
 (\Nat^\emptyset \, \alpha \, \mathit{Bool})\, (\Nat^\emptyset 
  (List \, \alpha) \, (List \, \alpha))} 
\newcommand{\filtypeGRose}{\Nat^\emptyset 
 (\Nat^\emptyset \, \alpha \, \mathit{Bool})\, (\Nat^\emptyset 
  (\mathit{GRose}\,\psi \, \alpha) \, (\mathit{GRose}\,\psi \, (\alpha
  + \onet)))} 
\newcommand{\maplist}{\mathit{map}_{\lambda A. \setsem{\emptyset; \alpha
      \vdash \mathit{List} \, \alpha} \rho[\alpha := A]}} 
\newcommand{\PLeaves}{\mathsf{PLeaves}}
\newcommand{\swap}{\mathsf{swap}}
\newcommand{\reverse}{\mathsf{reverse}}
\newcommand{\Bcons}{\mathit{Bcons}}
\newcommand{\Bnil}{\mathit{Bnil}}

\title[Free Theorems for Nested Types]{Free Theorems for
  %Primitive
  Nested Types} %% [Short Title] is optional; when present,
                         %% will be used in header instead of Full
                         %% Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
%\subtitle{Subtitle}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

%% Author information
%% Contents and number of authors suppressed with 'anonymous'.
%% Each author should be introduced by \author, followed by
%% \authornote (optional), \orcid (optional), \affiliation, and
%% \email.
%% An author may have multiple affiliations and/or emails; repeat the
%% appropriate command.
%% Many elements are not rendered, but should be provided for metadata
%% extraction tools.

%% Author with single affiliation.
\author{Patricia Johann, Enrico Ghiorzi, and Daniel Jeffries}
%\authornote{with author1 note}          %% \authornote is optional;
%                                        %% can be repeated if necessary
%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
%  \position{Position1}
%  \department{Department1}              %% \department is recommended
  \institution{Appalachian State University}            %% \institution is required
%  \streetaddress{Street1 Address1}
%  \city{City1}
%  \state{State1}
%  \postcode{Post-Code1}
%  \country{Country1}                    %% \country is recommended
}
\email{johannp@appstate.edu, ghiorzie@appstate.edu, jeffriesd@appstate.edu}          %% \email is recommended

%% macros
\newcommand{\nats}{\mu \beta. \onet + \beta}

\begin{document}

% Equality type GADT

\begin{comment}

    \[\begin{array}{l}
      \mathtt{data\; Eql \;(A : Set)\;(B : Set)\;:\;Set\;where}\\
    \hspace*{0.4in}\mathtt{refl\;:\; \;Eql \;A\;A}
    \end{array}\]

    % singleton Z - not a GADT
    \[\begin{array}{l}
    \mathtt{data\; Z \;:\;Set\;where}\\
    \hspace*{0.4in}\mathtt{szero\;:\; \;Z}
    \end{array}\]

    % singleton S - not a GADT
    \[\begin{array}{l}
      \mathtt{data\; S \;(A : Set)\;:\;Set\;where}\\
    \hspace*{0.4in}\mathtt{ssucc\;:\; A \rightarrow S\;A}
    \end{array}\]

    % GNat  - GADT
    \[\begin{array}{l}
    \mathtt{data\; GNat\;(A : Set)\;:\;Set\;where}\\
    \hspace*{0.4in}\mathtt{gzero\;:\; GNat\;Z}\\
      \hspace*{0.4in}\mathtt{gsucc\;:\;GNat\;A \rightarrow GNat\;(S\;A)}
    \end{array}\]

    % GVec  - GADT
    % --- uses GADT (GNat) in subscript!
    \[\begin{array}{l}
      \mathtt{data\; GVec\;(A : Set)\;(N : Set)\;:\;Set\;where}\\
      \hspace*{0.4in}\mathtt{gnil\;:\; GVec\;A\;(GNat\;Z)}\\
      \hspace*{0.4in}\mathtt{gcons\;:\; A \rightarrow 
          GVec\;A\;(GNat\;N) \rightarrow GVec\;A\;(GNat\;(S\;N))} 
    \end{array}\]

    % PTree 
    \[\begin{array}{l}
      \mathtt{data\; PTree\;(A : Set)\;:\;Set\;where}\\
      \hspace*{0.4in}\mathtt{pleaf\;:\; A \rightarrow PTree\;A}\\
      \hspace*{0.4in}\mathtt{pnode\;:\; PTree\;(A \times A) \rightarrow PTree\;A} \\
    \end{array}\]

    % GTree
    \[\begin{array}{l}
      \mathtt{data\; GTree\;(A : Set)\;(N : Set)\;:\;Set\;where}\\
      \hspace*{0.4in}\mathtt{gleaf\;:\; A \rightarrow GTree\;A\;Z}\\
      \hspace*{0.4in}\mathtt{gnode\;:\; GTree\;A\;N \rightarrow GTree\;A\;N 
            \rightarrow GTree\;A\;(S\;N)} \\
    \end{array}\]

    % ETree
    \[\begin{array}{l}
      \mathtt{data\; ETree\;(A : Set)\;(N : Set)\;:\;Set\;where}\\
      \hspace*{0.4in}\mathtt{eleaf\;:\; A \rightarrow Eql\;N\;Z \rightarrow ETree\;A\;N}\\
      \hspace*{0.4in}\mathtt{enode\;:\; Eql\;N\;(S\;M) \rightarrow ETree\;A\;M \rightarrow ETree\;A\;M 
            \rightarrow ETree\;A\;N} \\
    \end{array}\]

    % lambda terms from wobbly types (minus a few constructors)
    \[\begin{array}{l}
      \mathtt{data\; Term\;(A : Set)\;:\;Set\;where}\\
      \hspace*{0.4in}\mathtt{lit\;:\; \nat \rightarrow Term\;\nat}\\
      \hspace*{0.4in}\mathtt{inc\;:\; Term\;\nat \rightarrow Term\;\nat}\\
      \hspace*{0.4in}\mathtt{isZ\;:\; Term\;\nat \rightarrow Term\;Bool}\\
      \hspace*{0.4in}\mathtt{pair\;:\; Term\;A \rightarrow Term\;B \rightarrow Term\;(A \times B)}
    \end{array}\]

    % 
    \[\begin{array}{l}
    % Natural numbers in our calculus
    \nat = (\mu \beta. \onet + \beta) \\
    % Z in our calculus
    Z = \onet \\
    % S in our calculus
    S \, \beta = (\mu \phi. \lambda \alpha. \alpha) \beta \\
    % GNat in our calculus
    \mathit{GNat} \, \beta = 
    (\mu \phi. \lambda \alpha. 
      (\Lan^\emptyset_Z \, 1)\alpha 
      + (\Lan^\gamma_{S\,\gamma} \, \phi \gamma)\alpha)\beta \\
    % GVec in our calculus
    \mathit{GVec} \, \beta_1 \, \beta_2 = 
      (\mu \phi. \lambda \alpha_1, \alpha_2. 
        (\Lan^\gamma_{\gamma, \mathit{GNat}\,Z} \, 1)(\alpha_1, \alpha_2)
      + (\Lan^{\gamma_1,\gamma_2}_{\gamma_1, \mathit{GNat} \, (S\,\gamma_2)}
            \,\gamma_1 \times \phi (\gamma_1, \mathit{GNat} \, \gamma_2))(\alpha_1, \alpha_2)
            )(\beta_1, \beta_2) \\
      \color{red} \mbox{Note: GVec uses GNat as a subscript}\\
    % GTree in our calculus
      \mathit{GTree} \, \beta_1 \, \beta_2 =
        (\mu \phi. \lambda \alpha_1, \alpha_2.
          (\Lan^\gamma_{\gamma, Z} \, \gamma)(\alpha_1, \alpha_2)
        + (\Lan^{\gamma_1, \gamma_2}_{\gamma, S\,\gamma_2} \, \phi\,(\gamma_1,\gamma_2) \times \phi\,(\gamma_1,\gamma_2))(\alpha_1, \alpha_2)
        )(\beta_1, \beta_2) \\
    % PTree in our calculus
      \mathit{PTree} \, \beta = 
        (\mu \phi. \lambda \alpha. \alpha + \phi\,(\alpha \times \alpha))\beta \\

    % Term data type in our calculus
    \mathit{Term} \, \beta = (\mu \phi. \lambda \alpha. 
      (\Lan^\emptyset_\nat \, \nat)\alpha
    + (\Lan^\emptyset_\nat \, \phi \nat)\alpha
    + (\Lan^\emptyset_{\mathit{Bool}} \, \phi \nat)\alpha
    + (\Lan^{\gamma_1, \gamma_2}_{\gamma_1 \times \gamma_2} \, \phi \gamma_1 \times \phi \gamma_2)\alpha)\beta\\
    % Eql data type 
      \mathit{Eql} \, \beta_1 \, \beta_2 = (\Lan^\gamma_{\gamma, \gamma} 1)(\beta_1, \beta_2)
    \end{array}\]
  \end{comment}

\section{Programming with $\Lan$ types}
We can use $\Lan$ types in several different ways in our calculus. In this note
we will explore how to use $\Lan$ types to encode GADTs in our calculus. 
These $\Lan$ types / GADTs have several useful applications in our calculus.
In addition to allowing for more general return types in data type constructors,
$\Lan$ can be combined with singleton types to mimick some dependent types.

We can also define existential types and an equality type using $\Lan$,
but the latter is not really useful in the absence of type refinement(?).

The usefulness of existential types is also somewhat limited depending on 
what is allowed in the context of $G$ in the $\partial$ rule.

\subsection{Using Lan in a constructor}
Our calculus does not include "GADT-syntax" (like Agda, Haskell + GADTs) in which 
one can make the "return type" explicit.  Instead,
we write a data type as the fixed point of the sum of its constructors. 
To produce a term of some data type with $n$ constructors $(\mu \phi. \lambda \ol{\alpha}. F_1 + ... + F_n)\ol{G}$,
we first construct a term 
$$t : F_i [\phi :=_\beta (\mu \phi. \lambda \ol{\alpha}. F_1 + ... + F_n)\ol{\beta}][\ol{\alpha := G}]$$
and invoke $\tin$ and some combination of $\inl/\inr$ to get 
$$(\tin_{F_1 + ... + F_n})_{\ol{G}} \, (..\inl/\inr..  \,t) : (\mu \phi. \lambda \ol{\alpha}. F_1 + ... + F_n)\ol{G}$$

So how are
these restrictions on the return types enforced in our calculus?
Although $\tin_{F_1 + ... + F_n}$ is a natural transformation from 
$$(F_1 + ... + F_n)[\phi :=_\beta (\mu \phi. \lambda \ol{\alpha}. F_1 + ... + F_n)\ol{\beta}][\ol{\alpha := \beta}]$$
to 
$$(\mu \phi.\ol\alpha. F_1 + ... + F_n)\ol\beta$$
we can only use the component $(\tin_{F_1 + ... + F_n})_{\ol{G}}$ when
$F_i [\phi :=_\beta (\mu \phi. \lambda \ol{\alpha}. F_1 + ... + F_n)\ol{\beta}][\ol{\alpha := G}]$
is inhabited for some $1 \le i \le n$.

So when $F_i = (\Lan^{\ol\gamma}_{\ol K} \,H)\ol\alpha$, we know from the $\int$ rule
that we can only introduce terms of type $(\Lan^{\ol\gamma}_{\ol K} \,H)\ol{K[\ol{\gamma := A}]}$
for some types $\ol{A}$, so we can never use a component of $\tin_{F_1 + ... F_n}$ unless it has the form
$(\tin_{F_1 + ... F_n})_{\ol{K[\ol{\gamma := A}]}}$

% $\alpha_j = K_j[\ol{\gamma := A}]$ for $j = 1,...,|\ol{K}|$ and types $\ol{A}$.
For example, consider the third constructor in
the type of simply-typed lambda terms:
\[\begin{array}{l}
  \mathtt{data\; Term\;(A : Set)\;:\;Set\;where}\\
  \hspace*{0.4in}\mathtt{lit\;:\; \nat \rightarrow Term\;\nat}\\
  \hspace*{0.4in}\mathtt{inc\;:\; Term\;\nat \rightarrow Term\;\nat}\\
  \hspace*{0.4in}\mathtt{isZ\;:\; Term\;\nat \rightarrow Term\;Bool}\\
  \hspace*{0.4in}\mathtt{pair\;:\; Term\;A \rightarrow Term\;B \rightarrow Term\;(A \times B)} \\ \\

\mathit{Term} \, \beta = (\mu \phi. \lambda \alpha. 
  (\Lan^\emptyset_\nat \, \nat)\alpha
+ (\Lan^\emptyset_\nat \, \phi \nat)\alpha
+ (\Lan^\emptyset_{\mathit{Bool}} \, \phi \nat)\alpha
+ (\Lan^{\gamma_1, \gamma_2}_{\gamma_1 \times \gamma_2} \, \phi \gamma_1 \times \phi \gamma_2)\alpha)\beta
\end{array}\]

In this case, $F_i = (\Lan^{\emptyset}_{\mathit{Bool}} \, \phi\,\nat)\alpha$, which is only inhabited when $\alpha = \mathit{Bool}$.
The only component of $\tin_{F_1 + ... F_n}$ we can hope to use with this constructor is $(\tin_{F_1 + ... F_n})_\mathit{Bool}$,
and this restriction corresponds exactly with the return type of \texttt{isZ} in the Agda definition.

% This restriction in the $\int$ rule also restricts the choice of $\ol{B}$ in the $\partial$ rule,
% in practice.

\section{Singleton Types and Fake Dependent Types}
Since the return types of constructors in a GADT can vary, 
the type of a term can now potentially give us information about its value.

We have already seen the $\mathit{Term}$ datatype, which ensures that 
the lambda terms which inhabit the datatype are well-typed. This additional
structure can also make programs more efficient in a practical setting
by eliminating some pattern matching cases.

It turns out that we can mimick dependent types by using GADTs with "singleton types", 
which are type-level representations of terms. For instance, we can 
represent type-level natural numbers using $Z = \onet$ and 
$S \, \beta = (\mu \phi. \lambda \alpha. \alpha) \beta$.

Using $Z$ and $S$ we can define a type of length-indexed lists:
\[\begin{array}{l}
  \mathtt{data\; Vec\;(A : Set)\;(N : Set)\;:\;Set\;where}\\
  \hspace*{0.4in}\mathtt{gnil\;:\; Vec\;A\;\;Z}\\
  \hspace*{0.4in}\mathtt{gcons\;:\; A \rightarrow 
      Vec\;A\;N \rightarrow Vec\;A\;(S\;N)}  \\ \\

\mathit{Vec} \, \beta_1 \beta_2 = 
  (\mu \phi. \lambda \alpha_1, \alpha_2. 
    (\Lan^\gamma_{\gamma, Z} \, 1)(\alpha_1, \alpha_2)
  + (\Lan^{\gamma_1,\gamma_2}_{\gamma_1, S\,\gamma_2}
        \,\gamma_1 \times \phi (\gamma_1, \gamma_2))(\alpha_1, \alpha_2)
        )(\beta_1, \beta_2)
\end{array}\]
and for convenience we can define some constructors \\(abusing notation to write
$\tin_{\mathit{Vec}}$ for $\tin_{
    (\Lan^\gamma_{\gamma, Z} \, 1)(\alpha_1, \alpha_2)
  + (\Lan^{\gamma_1,\gamma_2}_{\gamma_1, S\,\gamma_2}
        \,\gamma_1 \times \phi (\gamma_1, \gamma_2))(\alpha_1, \alpha_2)}$ and similar below):

\[\begin{array}{l}
  \mathit{nil}: \Nat^\alpha \, \onet \, (\mathit{List}\,\alpha) \\
  \vspace{0.05in}
  \mathit{nil} = L_\alpha \ast. (\tin_{\onet + \alpha \times \phi\alpha})_\alpha (\inl \, \ast) \\
  \mathit{cons} : \Nat^\alpha \, (\alpha \times \mathit{List}\,\alpha)\, (\mathit{List} \,\alpha) \\
  \vspace{0.05in}
  \mathit{cons} = L_\alpha t. (\tin_{\onet + \alpha \times \phi\alpha})_\alpha (\inr \, t) \\
  \mathit{vnil} : \Nat^\alpha \, \onet \, (\mathit{Vec} \,\alpha \,Z) \\
  \vspace{0.05in}
  \mathit{vnil} = L_\alpha \ast. (\tin_{\mathit{Vec}})_{\alpha, Z} (\inl \,(\int_{(\gamma, Z), \onet} \ast)) \\
  \mathit{vcons} : \Nat^{\alpha, n} \, (\alpha \times \mathit{Vec} \, \alpha \, n)\, (\mathit{Vec} \, \alpha\,(S\,n)) \\
  \mathit{vcons} = L_{\alpha, n} t. (\tin_{\mathit{Vec}})_{\alpha, (S \,n)} (\inr\, (\int_{(\gamma_1, S\,\gamma_2), \gamma_1 \times \mathit{Vec} \, \gamma_1\,\gamma_2} t))
\end{array}\]

We can write a function to convert a vector into a list: 
\[\begin{array}{l}
  \mathit{vecToList} : \Nat^{\alpha, n} \, (\mathit{Vec} \, \alpha \, n) \, (\mathit{List} \, \alpha) \\
  \vspace{0.05in}
  \mathit{vecToList} = (\fold_{\mathit{Vec}})_\emptyset \mathit{alg} \\
  \mathit{alg} : \Nat^{\alpha, n} \, 
  ((\Lan^\gamma_{\gamma, Z} \, \onet)(\alpha, n) + (\Lan^{\gamma_1, \gamma_2}_{\gamma_1, S\,\gamma_2} \, \gamma_1 \times \mathit{List}\, \gamma_1)(\alpha, n))
  \; (\mathit{List}\,\alpha)\\
  \mathit{alg} = L_{\alpha, n} \, t. 
    \mathsf{case} \; t \; \mathsf{of} \; \{ \\ 
    \hspace{0.5in} vn \mapsto \mathit{nil}_\alpha \ast; \\
    \hspace{0.5in} vc \mapsto \mathit{cons}_\alpha 
    (\partial^{\beta \times \mathit{List}\,\beta, (\gamma_1, S\,\gamma_2)}_{\gamma_1 \times \mathit{List}\,\gamma_1} \, (\id_{\gamma_1\times \mathit{List}\,\gamma_1}) \, vc) \\
    \}
\end{array}\]

In $alg$, we use the following instance of the $\Lan$-elimination rule:
\[\begin{array}{c}
  \AXC{$\vdash \id_{\gamma_1 \times \mathit{List}\, \gamma_1} : \Nat^{\alpha,n,\gamma_1,\gamma_2}\,
  (\gamma_1 \times \mathit{List} \, \gamma_1)\;(\gamma_1 \times \mathit{List} \, \gamma_1)$}
  \AXC{$\emptyset;\alpha,n~|~\emptyset \vdash vc : (\Lan^{\gamma_1,\gamma_2}_{\gamma_1, S\,\gamma_2} \gamma_1 \times \mathit{List} \,\gamma_1)\,(\alpha, n) $}
  \BIC{$\emptyset;\alpha,n~|~\emptyset \vdash \partial^{\beta_1 \times \mathit{List}\,\beta_1, (\gamma_1, S\,\gamma_2)}_{\gamma_1 \times \mathit{List}\,\gamma_1} \eta\,vc :
  (\beta_1 \times \mathit{List}\, \beta_1)[\beta_1 := \alpha][\beta_2 := n]$}
\DisplayProof
\end{array}\]
to produce a term of type $\alpha \times \mathit{List}\, \alpha$.
We can also convert lists to vectors, but this is tackled in the next section 
with existential types.

\subsection{Safe(?) head}
Given a the type of length-indexed vectors, one might ask whether we can 
write a $head$ function of type $\Nat^{\alpha, n} (\mathit{Vec}\,\alpha\,(S\,n))\;(\alpha)$ 
in which we only need to give the case for the $\mathit{vcons}$ constructor.

\[\begin{array}{l}
  \mathit{head} : \Nat^{\alpha, n} (\mathit{Vec}\,\alpha\,(S\,n))\;(\alpha)\\
  \mathit{head} = L_{\alpha, n} t. \mathsf{case} \, ((\tin^{-1}_{\mathit{Vec}})_{\alpha, S\,n} t) \, \mathsf{of} \{ \\
  \hspace{0.5in} vn : (\Lan^\gamma_{\gamma, Z} \, 1)(\alpha, S\,n) \mapsto {\color{red}???}; \\
  \hspace{0.5in} vc : (\Lan^{\gamma_1, \gamma_2}_{\gamma_1, S\,\gamma_2} \, 
  \gamma_1 \times \mathit{Vec}\,\gamma_1\,\gamma_2)(\alpha, S\,n) 
  \mapsto  \partial^{\beta_1, (\gamma_1, \gamma_2)}_{\gamma_1 \times \mathit{Vec}\,\gamma_1\,\gamma_2} \, (L_{\alpha, n, \gamma_1, \gamma_2}\, x. \pi_1 x) \, vc \\
  \}
\end{array}\]

In the second case above, we can use the following instance of the $\Lan$-elimination rule
to extract a term of type $\alpha$ from a term of type $(\Lan^{\gamma_1, \gamma_2}_{\gamma_1, S\,\gamma_2} \, 
  \gamma_1 \times \mathit{Vec}\,\gamma_1\,\gamma_2)(\alpha, n)$:
\[\begin{array}{c}
  \AXC{$\vdash \eta : \Nat^{\alpha,n,\gamma_1,\gamma_2}\,
  (\gamma_1 \times \mathit{Vec} \, \gamma_1 \,\gamma_2)\;(\gamma_1)$}
  \AXC{$\emptyset;\alpha,n~|~ vc \vdash vc : (\Lan^{\gamma_1,\gamma_2}_{\gamma_1, S\,\gamma_2} \gamma_1 \times \mathit{Vec} \,\gamma_1\,\gamma_2)\,(\alpha, S\,n) $}
  \BIC{$\emptyset;\alpha,n~|~ vc \vdash \partial^{\beta_1, (\gamma_1, S\,\gamma_2)}_{\gamma_1 \times \mathit{Vec}\,\gamma_1\,\gamma_2}\, \eta\,vc :
  (\beta_1)[\beta_1 := \alpha][\beta_2 := S\,n]$}
\DisplayProof
\end{array}\]

Unfortunately, we have no way of producing a term of type $\alpha$ in the first case. 
In a practical setting, the type unification algorithm would determine at this point
that the first constructor pattern need not be matched, because $Z$ cannot be unified 
with $S\,n$ for any $n$. It is also true in our calculus that, due to the $\Lan$-introduction rule,
it is impossible to construct a term of type $(\Lan^{\ol{\gamma}}_{\ol{K}}\, F)\ol{B}$ unless
$B_i = K_{i}[\ol{\gamma := A}]$ for $i = 1,...,|\ol{K}|$ and some types $\ol{A}$.
In particular, it is impossible to construct a term of type 
$(\Lan^\gamma_{\gamma, Z} \, 1)(\alpha, S\,n)$, so assuming its existence should
allow us to do something akin to $\bot$-eliimination in the first case above.
{\color{blue} I don't know how we could do this because it seems we would need type-equality
at the meta-level in order to have a rule that can check the arguments of a $\Lan$ type in this way.
This concern is probably orthogonal to the goals of this project, but without it, the
"fake dependent types" are not incredibly useful.}

{\color{red} Note about S and Z }

Note that we cannot write a function 
from the type-level natural numbers to 
the natural numbers data type. We 
can write each instance, e.g. $Zto0 : \Nat^{\emptyset} (Z) \, (\nats)$,
$SZto1 : \Nat^{\emptyset} (S\,Z) \, (\nats)$, but we cannot 
write them all in a single function because S and Z are separate types. 
(What would the type be? $\Nat^{\alpha} (Z + S \,\alpha) (\nats)$? This is too 
generic because it must be defined for all $\alpha$. )
We address this issue with the following GADT:

\[\begin{array}{l}
\mathtt{data\; GNat\;(A : Set)\;:\;Set\;where}\\
\hspace*{0.4in}\mathtt{gzero\;:\; GNat\;Z}\\
  \hspace*{0.4in}\mathtt{gsucc\;:\;GNat\;A \rightarrow GNat\;(S\;A)}
\end{array}\]
Note that using $\mathit{GNat}$ in the definition of $\mathit{Vec}$ 
would required having a $\Lan$ type ($\mathit{GNat}$) in the subscript of another $\Lan$
(a constructor for $\mathit{Vec}$).

\section{Existential Types} 
The following defines an existential type in Agda:
\[\begin{array}{l}
\mathtt{data\; Exists \;(F : Set \rightarrow Set)\;:\;Set_1\;where}\\
\hspace*{0.4in}\mathtt{e\;:\; F \,A \rightarrow Exists\;F}\\
\end{array}\]
The aspect that makes it existential is that the type $\mathtt{A}$ used in the constructor argument is hidden in the return type.
To provide a witness to the fact that there exists a type $\mathtt{A}$ such that $\mathtt{List\, A}$
is inhabited, we just need to provide an inhabitant of $\mathtt{List \, A}$ for some type $\mathtt{A}$.

Unfortunately, we cannot define an exact analogue (with closed type) 
of the $\mathtt{Exists}$ datatype in our calculus because it is parameterized over a $1$-ary variable ($\mathtt{F}$),
and we can only bind 0-ary variables in our types (we can bind other variables with $\mu$, but that isn't what we want here).

However, we can define an existential type for a variable $\psi$ as $\emptyset; \psi \vdash (\Lan^{\ol\alpha}_\emptyset\,\psi\,\ol\alpha)$,
or given some type $\Gamma; \Phi, \ol\gamma \vdash F$ we can existentially quantify $\ol\gamma$ 
in $F$ with $\Gamma; \Phi \vdash (\Lan^{\ol\alpha}_\emptyset \, F[\ol{\gamma := \alpha}])$.

The introduction and elimination rules for $\Lan$ give these types their "existential" character:

\[\begin{array}{c}
\AXC{$\Gamma; \Phi,\ol{\alpha}\vdash F$}
  \AXC{$\ol{\Gamma;\Phi \vdash A}$} 
  \AXC{$\Gamma;\Phi~|~\Delta \vdash t : F[\ol{\alpha :=A}]$}
\TIC{$\Gamma;\Phi~|~\Delta \vdash \int_{\emptyset,F} t :
  (\Lan^{\ol\alpha}_{\emptyset}\,  F)$}
\DisplayProof
\\\\
\AXC{$\Gamma;\emptyset~|~\Delta \vdash \eta : \Nat^{\Phi,\ol\alpha}\,
  F\;G$}
\AXC{$\Gamma;\Phi~|~\Delta \vdash t : (\Lan^{\ol\alpha}_{\emptyset} F) $}
\BIC{$\Gamma;\Phi~|~\Delta \vdash \partial^{G, \emptyset}_F \eta\,t :
 G$}
\DisplayProof
\end{array}\]

The introduction rule is precisely as we expect:
to introduce an existential type, we just need a term of type $F[\ol{\alpha := A}]$
for some types $\ol{A}$. The elimination rule also behaves as expected: 
if we have a term $t$ of existential type and a natural transformation from $F$ to $G$, 
we don't need to know which instance of $F$ produced $t$ in order to reach $G$, because 
the natural transformation is defined for all instances of $F$.
For example, if we have a list whose elements are of an unknown type, we can still
compute its length given a natural transformation of type $\Nat^\alpha (\mathit{List}\,\alpha) (\nat)$.

\subsection{From List to Vec (PTree to GTree)}
In order to write a natural transformation from lists to length-indexed vectors,
we must existentially quantify the length variable in the return type. Otherwise,
we must define the natural transformation for all possible values of $n$.
The function is defined as follows:

{\color{red} I am having issues defining the list to vector function at the moment, but I will define an analogous
embedding function for $\mathit{PTree}$ and $\mathit{GTree}$:}
\[\begin{array}{l}
  \mathtt{data\; PTree\;(A : Set)\;:\;Set\;where}\\
  \hspace*{0.4in}\mathtt{pleaf\;:\; A \rightarrow PTree\;A}\\
  \hspace*{0.4in}\mathtt{pnode\;:\; PTree\;(A \times A) \rightarrow PTree\;A} \\ \\
  \mathit{PTree} \, \beta = 
    (\mu \phi. \lambda \alpha. \alpha + \phi\,(\alpha \times \alpha))\beta \\ \\ \\

  \mathtt{data\; GTree\;(A : Set)\;(N : Set)\;:\;Set\;where}\\
  \hspace*{0.4in}\mathtt{gleaf\;:\; A \rightarrow GTree\;A\;Z}\\
  \hspace*{0.4in}\mathtt{gnode\;:\; GTree\;A\;N \rightarrow GTree\;A\;N 
        \rightarrow GTree\;A\;(S\;N)} \\ \\

  \mathit{GTree} \, \beta_1 \, \beta_2 =
    (\mu \phi. \lambda \alpha_1, \alpha_2.
      (\Lan^\gamma_{\gamma, Z} \, \gamma)(\alpha_1, \alpha_2)
    + (\Lan^{\gamma_1, \gamma_2}_{\gamma, S\,\gamma_2} \, \phi\,(\gamma_1,\gamma_2) \times \phi\,(\gamma_1,\gamma_2))(\alpha_1, \alpha_2)
    )(\beta_1, \beta_2) 
\end{array}\]

%%%%% haven't quite figured this out 
%%%%%  
%% \[\begin{array}{l}
%%   \mathit{listToVec} : \Nat^\alpha \, (\mathit{List}\,\alpha)\,(\Lan^n_\emptyset \, \mathit{Vec}\,\alpha\,n) \\
%%   \vspace{0.05in}
%%   \mathit{listToVec} = (\fold_{\onet + \beta \times \phi\,\beta})_\emptyset \, \mathit{alg} \\
%%   \mathit{alg} : \Nat^\alpha (\onet + \alpha \times (\Lan^n_\emptyset \, \mathit{Vec}\,\alpha\,n)) 
%%     \; (\Lan^n_\emptyset \,\mathit{Vec}\,\alpha\,n) \\
%%   \mathit{alg} = L_\alpha \,s. \mathsf{case} \;s \; \mathsf{of} \; \{ \\
%%     \hspace{0.5in} \ast \mapsto \int_{\emptyset, \mathit{Vec}\,\alpha\,n} (\mathit{vnil}_\alpha \ast)  ; \\
%%     \hspace{0.5in} t 
%%     : \alpha \times (\Lan^n_\emptyset \,\mathit{Vec}\,\alpha\,n) 
%%     \mapsto ??? \\
%%     \}
%% \end{array}\]

We first define some constructors for $\mathit{GTree}$ and some helper functions for 
$\mathit{ptreeToGTree}$. There is some trickiness with variable names in the $\Lan$ types, and so we define
$\mathit{alphaRename}$ to deal with this.

The origin of this issue is in the typing of the 
$\Lan$-elimination rule used in the second case of $alg$ below. In particular, it does
not seem possible to let the codomain of $\eta$ be typed as $\emptyset; \alpha, n \vdash (\Lan^n_\emptyset \, \mathit{GTree}\,\alpha\,n)$,
even though it is legal to form $\emptyset; \alpha \vdash (\Lan^n_\emptyset \,\mathit{GTree}\,\alpha\,n)$ and 
then weaken to get  $\emptyset; \alpha, n \vdash (\Lan^n_\emptyset \, \mathit{GTree}\,\alpha\,n)$. The issue
is that a natural transformation of type 
$\Nat^{\alpha,n}\,
  (\mathit{GTree}\,(\alpha\times\alpha)\,n) \;(\Lan^n_\emptyset\,\mathit{GTree}\,\alpha\,n)
  $ 
apparently cannot be defined directly.
To define such a natural transformation, we would need to change $\mathit{helper}$'s type and definition:  \\ \\
  $\mathit{helper} : \Nat^{\alpha,n}\,
  (\mathit{GTree}\,(\alpha\times\alpha)\,n) \;(\Lan^{\color{red}n}_\emptyset\,\mathit{GTree}\,\alpha\,{\color{red}n})$ \\
  $\mathit{helper} = L_{\alpha, n} \, t. \, \int_{\emptyset, \mathit{GTree}\,\alpha\,{\color{red}n}}  \,  (\mathit{gnode}_{\alpha, n}  \,  (\mathit{unzipGTree}_{\alpha, n} \, t)) $ \\

It turns out that the use of $\int_{\emptyset, \mathit{GTree}\,\alpha\,n}$ is not well-typed in this 
version (specifically because it requires $n$ to be in both $\Phi$ and $\ol\alpha$ in the context of $F$
in the $\Lan$-introduction rule), and so we must define $\mathit{alphaRename}$.

\[\begin{array}{l}
  % constructors for GTree
  \mathit{gleaf} : \Nat^{\alpha}\, (\alpha) \, (\mathit{GTree}\,\alpha\,Z) \\
  \mathit{gleaf} = L_\alpha \, t. (\tin_{\mathit{GTree}})_{\alpha,Z} \, (\inl \, (\int_{(\gamma, Z), \gamma} \, t)) \\ \\

  \mathit{gnode} : \Nat^{\alpha, n}\, (\mathit{GTree}\,\alpha\,n \times \mathit{GTree}\,\alpha\,n) \, (\mathit{GTree}\,\alpha\, (S\,n)) \\
  \mathit{gnode} = L_{\alpha, n} \, t. (\tin_{\mathit{GTree}})_{\alpha,(S\,n)} \, (\inr \, (\int_{(\gamma_1, S\,\gamma_2), \mathit{GTree}\,\gamma_1\,\gamma_2\times\mathit{GTree}\,\gamma_1\,\gamma_2} \, t)) \\ \\

  \mathit{ptreeToGTree} : \Nat^\alpha \, (\mathit{PTree}\,\alpha)\,(\Lan^n_\emptyset \, \mathit{GTree}\,\alpha\,n) \\
  \vspace{0.05in}
  \mathit{ptreeToGTree} = (\fold_{\beta + \phi\,(\beta \times \beta)})_\emptyset \, \mathit{alg} \\ \\ 
  \mathit{alg} : \Nat^\alpha (\alpha + (\Lan^n_\emptyset \, \mathit{GTree}\,(\alpha \times \alpha) \,n)) 
    \; (\Lan^n_\emptyset \,\mathit{GTree}\,\alpha\,n) \\
  \mathit{alg} = L_\alpha \,s. \mathsf{case} \;s \; \mathsf{of} \; \{ \\
    \hspace{0.5in} \mathit{lf} : \alpha \mapsto \int_{\emptyset, \mathit{GTree}\,\alpha\,n} (\mathit{gleaf}_\alpha \mathit{lf})  ; \\
    \hspace{0.5in} \mathit{nd}
    : (\Lan^n_\emptyset \,\mathit{GTree}\,(\alpha \times \alpha)\,n) 
    \mapsto \mathit{alphaRename}_{\mathit{GTree}\,\alpha\,\beta}\, ( \partial^{(\Lan^m_\emptyset \,\mathit{GTree}\,\alpha\,m),\emptyset}_{\mathit{GTree}\,(\alpha \times \alpha)\,n} \mathit{helper} \,t ) \\
    \}\\ \\

  \mathit{helper} : \Nat^{\alpha,n}\,
  (\mathit{GTree}\,(\alpha\times\alpha)\,n) \;(\Lan^m_\emptyset\,\mathit{GTree}\,\alpha\,m) \\
  \mathit{helper} = L_{\alpha, n} \, t. \, \int_{\emptyset, \mathit{GTree}\,\alpha\,m}  \,  (\mathit{gnode}_{\alpha, n}  \,  (\mathit{unzipGTree}_{\alpha, n} \, t)) \\ \\

  \mathit{unzipGTree} : \Nat^{\alpha, n} \, (\mathit{GTree}\,(\alpha \times \alpha)\,n) \, (\mathit{GTree}\,\alpha\,n \times \mathit{GTree}\,\alpha\,n) \\
  \vspace{0.05in}
  \mathit{unzipGTree} = L_{\alpha, n}\, t. (     
    ((\map^{\alpha \times \alpha, \alpha}_{\mathit{GTree}\,\beta\,n})_\emptyset \, (L_{\alpha, n} x. \pi_1 x))_{\alpha, n} \, t ,\\
    \hspace{1.3in}       ((\map^{\alpha \times \alpha, \alpha}_{\mathit{GTree}\,\beta\,n})_\emptyset \, (L_{\alpha, n} x. \pi_2 x))_{\alpha, n} \, t)\\ \\

  \mathit{existsGTree} : \Nat^{\alpha, n} (\mathit{GTree}\,\alpha\,n) \; (\Lan^m_\emptyset \,\mathit{GTree}\,\alpha\,m) \\
    \vspace{0.15in}
    % F = GTree alpha n 
    % t : (GTree alpha n)[n := S n] -- ! cant do this because \Phi and \ol\alpha 
    % must be disjoint for F in \int rule 
    % -- but we can let Phi := alpha, n ; \ol\alpha := m ; A := S n
    % -- F := \alpha, n, m \vdash GTree \alpha m 
    % -- F[\alpha := A] := GTree \alpha (S n)
    \mathit{existsGTree} = L_{\alpha, n} \, t. \int_{\emptyset,\mathit{GTree}\,\alpha\,m} t \\ 
    \vspace{0.05in}
    % we cant replace along m, n here can we ? yes we can , e.g. 
    % dfn 3 
    % "H" - phi \vdash (Lan^m phi m ) 
    % "F" - alpha, n \vdash Vec alpha n 
    % --------
    % \alpha \vdash (Lan^m phi m)[phi :=_n Vec alpha n]
    % =
    % \alpha \vdash (Lan^m Vec alpha m)
    %
    %
  \mathit{alphaRename} : \Nat^{\phi} (\Lan^m_\emptyset \,\phi\,m) \; (\Lan^n_\emptyset \,\phi\,n)\\
    \vspace{0.05in}
    \mathit{alphaRename} = L_{\phi}\, t.  \partial^{(\Lan^n_\emptyset \phi\,n), \emptyset}_{(\phi\,m)} \mathit{exists}  \;t \\
    \\
    \mathit{exists} : \Nat^{\phi, m} (\phi\,m) \; (\Lan^n_\emptyset \, \phi \,n) \\
    \mathit{exists} = L_{\phi, m}\, t . \int_{\emptyset, \phi\,n} t \\ \\

\end{array}\]

{\color{red} Let $\emptyset; \alpha \vdash G = (\Lan^n_\emptyset\,\mathit{GTree}\,\alpha\,n)$.
Is this allowed?
}

\[\begin{array}{c}
  \vspace{0.05in}
  \mbox{\textbf{Typing of $\Lan$-elimination in \textit{nd} case of \textit{alg} } }\\
\AXC{$\emptyset;\emptyset~|~\Delta \vdash \eta : \Nat^{\alpha,n}\,
  (\mathit{GTree}\,(\alpha\times\alpha)\,n) \;(\Lan^m_\emptyset\,\mathit{GTree}\,\alpha\,m)
  $}
  \AXC{$\emptyset;\alpha~|~\Delta \vdash t : (\Lan^n_\emptyset \,\mathit{GTree}\,(\alpha \times \alpha) \,n) $}
  \BIC{$\emptyset;\alpha~|~\Delta \vdash \partial^{(\Lan^m_\emptyset \,\mathit{GTree}\,\alpha\,m),\emptyset}_{\mathit{GTree}\,(\alpha \times \alpha)\,n} \eta\,t :
  (\Lan^m_\emptyset \,\mathit{GTree}\,\alpha\,m)  
  $}
\DisplayProof \\ \\

  \vspace{0.05in}
  \mbox{\textbf{Typing of $\Lan$-elimination in \textit{alphaRename} } }\\
  % typing of alpha conversion for Lan
  \AXC{$\emptyset;\emptyset~|~t\vdash \eta : \Nat^{\phi, m}\,
    (\phi\,m)\; (\Lan^n_\emptyset \,\phi\,n)$}
  \AXC{$\emptyset;\phi~|~ t \vdash t : (\Lan^{m}_{\emptyset} (\phi\,m))$}
  \BIC{$\emptyset;\phi~|~ t \vdash \partial^{(\Lan^n_\emptyset \phi\,n), \emptyset}_{(\phi\,m)} \eta\,t :
   (\Lan^n_\emptyset \,\phi\,n)$}
  \DP \\ \\

  \vspace{0.05in}
  \mbox{\textbf{Typing of $\Lan$-introduction in \textit{exists} } }\\
  % typing of alpha conversion helper function for Lan (exists function)
  \AXC{$\emptyset; \phi, m, n \vdash \phi \,n$}
  \AXC{$\emptyset; \phi, m \vdash m $}
  \AXC{$\emptyset; \phi, m \,|\, t \vdash t : \phi n [n := m]$}
  \TIC{$\emptyset; \phi, m \,|\, t \vdash \int_{\emptyset, \phi \,n} t : (\Lan^n_\emptyset \phi \,n)$}
  \DP
\end{array}\]

\subsection{Varying $K$ in Existential Types}
We can also represent existential types with $(\Lan^{\ol\alpha}_\onet \, F)\onet$, or
by using any other constant functor as a subscript. This is because using
a constant functor in the subscript restricts the introduction rule 
so that the instantiating types $\ol{A}$ are hidden in the resulting $\Lan$-type.

\[\begin{array}{c}
  \mbox{\textbf{$\Lan$-introduction rule with constant subscript} }\\
\AXC{$\Gamma; \Phi,\ol{\alpha}\vdash F$}
\AXC{$\ol{\Gamma; \ol{\alpha} \vdash K}$}
\AXC{$\ol{\Gamma;\Phi \vdash A} \hspace*{0.3in} \Gamma;\Phi~|~\Delta \vdash t : F[\ol{\alpha :=A}]$}
\TIC{$\Gamma;\Phi~|~\Delta \vdash \int_{\ol K,F} t :
  (\Lan^{\ol\alpha}_{\ol K}\,  F)\,\ol{K}$}  
\DisplayProof
\end{array}\]

Varying $K$ changes the elimination rule as well. In particular,
the types $\ol{B}$ must be $\ol{K}$,

\[\begin{array}{c}
  \mbox{\textbf{$\Lan$-elimination rule with constant subscript} }\\
\AXC{$\Gamma;\emptyset~|~\Delta \vdash \eta : \Nat^{\Phi,\ol\alpha}\,
  F\;G[\ol{\beta := K}]$}
\AXC{$\Gamma;\Phi~|~\Delta \vdash t : (\Lan^{\ol\alpha}_{\ol K} F)\,\ol K $}
\BIC{$\Gamma;\Phi~|~\Delta \vdash \partial^{G, \ol K}_F \eta\,t :
 G[\ol{\beta := K}]$}
\DisplayProof 
\end{array}\]

The most general existential type is $(\Lan^{\ol\alpha}_\emptyset \,F)$,
or equivalently $(\Lan^{\ol\alpha}_\onet\, F)\onet$, but if we 
change the subscript to $\nat$, for example, then we are somewhat
restricted in what we can compute with this type due to the 
elimination rule.

% % \subsection{Computing with Existential Types}
% % {\color{blue} ($\Lan^\alpha_{\nat}\,\mathit{List}\,\alpha$) example}

\section{Equality Types}
A common use of GADTs is to define a datatype representing 
type equality:

\[\begin{array}{l}
  \mathtt{data\; Eql \;(A : Set)\;(B : Set)\;:\;Set\;where}\\
\hspace*{0.4in}\mathtt{refl\;:\; \;Eql \;A\;A}
\end{array}\]
We can define this in our calculus as 
  $\mathit{Eql} \, \beta_1 \, \beta_2 = (\Lan^\gamma_{\gamma, \gamma} 1)(\beta_1, \beta_2)$.
Looking at the $\Lan$-introduction rule tells us that we can only produce a term of 
this type when $\beta_1$ and $\beta_2$ are the same type:

\[\begin{array}{c}
\AXC{$\Gamma; \Phi, \gamma\vdash \onet$}
\AXC{$\ol{\Gamma; \gamma \vdash \gamma}$}
  \AXC{$\Gamma;\Phi \vdash A $}
  \AXC{$\vdash t : \onet$}
\QIC{$\Gamma;\Phi~|~\emptyset \vdash \int_{\ol K,F} t :
  (\Lan^{\gamma}_{\gamma, \gamma}\,  \onet)\,(A, A)$}
\DisplayProof
\end{array}\]

We can use the elimination rule to show that this equality
is symmetric, but so far, I have not found many useful applications
of this type. It seems these types are mostly useful in practical settings
with type refinement/unification.

\[\begin{array}{c}
  \mbox{\textbf{Generic $\Lan$-elimination rule for Eql } }\\
\AXC{$\Gamma;\emptyset~|~\Delta \vdash \eta : \Nat^{\Phi,\gamma}\,
  \onet\;G[\beta_1 := \gamma][\beta_2 := \gamma]$}
\AXC{$\Gamma;\Phi \vdash A$}
\AXC{$\Gamma;\Phi \vdash B$}
  \AXC{$\Gamma;\Phi~|~\Delta \vdash t : (\Lan^{\gamma}_{\gamma, \gamma} \onet)\,(A, B) $}
\QIC{$\Gamma;\Phi~|~\Delta \vdash \partial^{G, \ol K}_F \eta\,t :
  G[\beta_1 := A][\beta_2 := B] $}
\DisplayProof \\ \\

\mbox{\textbf{$\Lan$-elimination rule for Eql with G = ($\mathit{Eql} \, \beta_2 \, \beta_1$)} }\\
\AXC{$\Gamma;\emptyset~|~\Delta \vdash \eta : \Nat^{\Phi,\gamma}\,
  \onet\;(\mathit{Eql} \, \gamma \, \gamma)$}

  % G[\beta_1 := \gamma][\beta_2 := \gamma]$}
\AXC{$\Gamma;\Phi \vdash A$}
\AXC{$\Gamma;\Phi \vdash B$}
  \AXC{$\Gamma;\Phi~|~\Delta \vdash t : \mathit{Eql}\,A\,B
  % : (\Lan^{\gamma}_{\gamma, \gamma} \onet)\,(A, B)
  $}
\QIC{$\Gamma;\Phi~|~\Delta \vdash \partial^{G, \ol K}_F \eta\,t :
  \mathit{Eql}\,B\,A
  % G[\beta_1 := A][\beta_2 := B] 
  $}
\DisplayProof \\
\end{array}\]

We can also use the $\mathit{Eql}$ type to define GADTs, e.g., an analogue of
$\mathit{GTree}$:

\[\begin{array}{l}
  \mathtt{data\; ETree\;(A : Set)\;(N : Set)\;:\;Set\;where}\\
  \hspace*{0.4in}\mathtt{eleaf\;:\; A \rightarrow Eql\;N\;Z \rightarrow ETree\;A\;N}\\
  \hspace*{0.4in}\mathtt{enode\;:\; Eql\;N\;(S\;M) \rightarrow ETree\;A\;M \rightarrow ETree\;A\;M 
        \rightarrow ETree\;A\;N} \\

  \mathit{ETree} \, \beta_1 \, \beta_2 = \\
     (\mu \phi. \lambda \alpha, n. 
      \alpha \times (\Lan^\gamma_{\gamma, \gamma} \, \onet)(n, Z)
    + (\Lan^{\alpha', n', m}_{\alpha', n'} \,(\Lan^\gamma_{\gamma, \gamma}\,\onet)(n', S\, m) \times \phi\,(\alpha',m) \times \phi\,(\alpha',m))(\alpha, n) 
    )(\beta_1, \beta_2) \\
\end{array}\]

\noindent
I have not yet explored $\mathit{ETree}$ or similar types thoroughly.

% % \section{Other}
% % \subsection{Case Elimination}
% % {\color{red} head: Vec a Sn -> a example}
% % 
% % \subsection{Properties of Lan}
% % 
% % {\color{red} What is allowed for the context of G in $\partial$ rule??}
% % 
% % 
% % \subsection{Balanced tree variants}

\end{document}

### 2020-08-25 20-gadts/inf-rules.tex
progress on proof of IEL

@@
   \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}\rho &=
-    (\textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}
-      \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := \_}])
-    \ol{\oCPOsem{\Gamma; \Phi \vdash A}\rho}
+    \{ t : (\textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash
+          K}\rho[\ol{\alpha := \_}]}} \oCPOsem{\Gamma; \Phi, \ol\alpha
+      \vdash F}\rho[\ol{\alpha := \_}]) \ol{\oCPOsem{\Gamma; \Phi
+        \vdash A}\rho}\\
+    &\hspace{0.3in}|~ (t,t) \in \oCPORsem{\Gamma;\Phi \vdash
+      (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}\Eq_\rho\}
 \end{align*}
@@
 
+{\color{blue} Need to check that the extra condition in the interp of
+  Lans is preserved for the last clause to be well-defined.}
+
 Definitions~\ref{def:set-sem} and~\ref{def:set-sem-funcs} respect
@@
       \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := \_}])
-    \ol{\oCPORsem{\Gamma; \Phi \vdash A}\rho}
+    \ol{\oCPORsem{\Gamma; \Phi \vdash A}\rho}\\
+  \oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}}
+    F)\ol{A}}\rho &= \{ (t_1,t_2) : \oCPOsem{\Gamma;\Phi \vdash
+    (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}\pi_1 \rho \,\times\,
+  \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}}
+    F)\ol{A}}\pi_2 \rho\\
+  &\hspace{0.3in}|~( (\eta_1)_{\ol{\oCPORsem{\Gamma;\Phi \vdash A}\pi_1
+    \rho}} t_1,  (\eta_2)_{\ol{\oCPORsem{\Gamma;\Phi \vdash A}\pi_2
+    \rho}} t_2) \in\\
+  &\hspace{0.5in} (\textit{Lan}_{\ol{\oCPORsem{\Gamma; \ol\alpha \vdash
+          K}\rho[\ol{\alpha := \_}]}} \oCPORsem{\Gamma; \Phi, \ol\alpha
+      \vdash F}\rho[\ol{\alpha := \_}]) \ol{\oCPORsem{\Gamma; \Phi
+        \vdash A}\rho} \}\\
 \end{align*}
 \end{dfn}
+{\color{blue} In the final clause of Definition~\ref{def:rel-sem},
+  $\eta_1$ and $\eta_2$ are...}
 
@@
 $T^\oCPO_{H,f}$ are replaced by $T_{H,f}$.
+
+{\color{blue} This is not true for the relational interps of
+  Lans. Define explicitly.}
+
 \end{dfn}
@@
 The proof is by induction on the structure of $F$. Only the $\Nat$,
-application, and fixpoint {\color{red} and $\Lan$?} cases are
-non-routine. The latter two use
-Lemma~\ref{lem:eq-reln-equalities}. The fixpoint case also uses the
+application, fixpoint, and $\Lan$ cases are non-routine. The latter
+two use Equation~\ref{eq:eq-pres-eq}. The fixpoint case also uses the
 observation that, for every $i < \omega_1$, the following intermediate
 results can be proved by simultaneous induction with
-Theorem~\ref{thm:iel}: for any $H$, $\rho$, $A$, and any subformula
-$J$ of $H$, both $T^i_{H,\Eq_{\rho}} K_0\, \ol{\Eq_A} =
-(\Eq_{(T^\set_{H,\rho})^i K_0})^* \ol{\Eq_A}$\; and\;
+Theorem~\ref{thm:iel}: {\color{red} CHECK THIS!} for any $H$, $\rho$,
+$A$, and any subformula $J$ of $H$, both $T^i_{H,\Eq_{\rho}} K_0\,
+\ol{\Eq_A} = (\Eq_{(T^\set_{H,\rho})^i K_0})^* \ol{\Eq_A}$\; and\;
 \[\begin{array}{ll}

### 2020-08-30 20-gadts/inf-rules.tex
IEL for gadts complete?
another change to condition in rel interp of Lans
add lines 669-671 to inf-rules.tex

@@
     F)\ol{A}}\pi_2 \rho\\
-  &\hspace{0.3in}|~( (\eta_1)_{\ol{\oCPORsem{\Gamma;\Phi \vdash A}\pi_1
-    \rho}} t_1,  (\eta_2)_{\ol{\oCPORsem{\Gamma;\Phi \vdash A}\pi_2
-    \rho}} t_2) \in\\
-  &\hspace{0.5in} (\textit{Lan}_{\ol{\oCPORsem{\Gamma; \ol\alpha \vdash
-          K}\rho[\ol{\alpha := \_}]}} \oCPORsem{\Gamma; \Phi, \ol\alpha
-      \vdash F}\rho[\ol{\alpha := \_}]) \ol{\oCPORsem{\Gamma; \Phi
-        \vdash A}\rho} \}\\
+  &\hspace{0.3in}|~\forall \sigma : \setenv, (
+  (\eta^\sigma_1)_{\ol{\oCPOsem{\Gamma;\Phi \vdash A}\sigma }} t_1,
+  (\eta^\sigma_2)_{\ol{\oCPOsem{\Gamma;\Phi \vdash A}\sigma }} t_2) \in\\
+  &\hspace{0.5in} (\textit{Lan}_{\ol{\oCPORsem{\Gamma; \ol\alpha
+        \vdash K}\Eq_\sigma[\ol{\alpha := \Eq_{\_}}]}}
+  \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\Eq_\sigma[\ol{\alpha :=
+      \Eq_{\_}}]) \ol{\oCPORsem{\Gamma; \Phi \vdash A}\Eq_\sigma} \}\\
 \end{align*}
@@
 {\color{blue} In the final clause of Definition~\ref{def:rel-sem},
-  $\eta_1$ and $\eta_2$ are...}
+  $\eta^\sigma_i : \textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash
+        K}\sigma[\ol{\alpha := \_}]}} \oCPOsem{\Gamma; \Phi, \ol\alpha
+    \vdash F}\sigma[\ol{\alpha := \_}] \; \to \;
+  \pi_i\,(\textit{Lan}_{\ol{\oCPORsem{\Gamma; \ol\alpha \vdash
+        K}\Eq_\sigma[\ol{\alpha := \Eq_{\_}}]}} \oCPORsem{\Gamma;
+    \Phi, \ol\alpha \vdash F}\Eq_\sigma[\ol{\alpha := \Eq_{\_}}])$ for
+  each $\sigma : \setenv$ and $i = 1,2$.}
 

### 2020-08-31 20-gadts/inf-rules.tex
new conditions for Lan interps

--- 2020-08-25 20-gadts/inf-rules.tex
@@
         \vdash A}\rho}\\
-    &\hspace{0.3in}|~ (t,t) \in \oCPORsem{\Gamma;\Phi \vdash
-      (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}\Eq_\rho\}
+    &\hspace{0.3in}|~
+%    (t,t) \in \oCPORsem{\Gamma;\Phi \vdash
+%      (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}\Eq_\rho\}\\
+{\color{blue} ( (\eta_1)_{\ol{\oCPOsem{\Gamma;\Phi \vdash A}\rho }} t,
+  (\eta_2)_{\ol{\oCPOsem{\Gamma;\Phi \vdash A}\rho }} t) \in}\\
+  &{\color{blue} \hspace{0.5in} (\textit{Lan}_{\ol{\oCPORsem{\Gamma; \ol\alpha
+        \vdash K}\Eq_\rho[\ol{\alpha := \Eq_{\_}}]}}
+  \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\Eq_\rho[\ol{\alpha :=
+      \Eq_{\_}}]) \ol{\oCPORsem{\Gamma; \Phi \vdash A}\Eq_\rho} \}}\\
 \end{align*}
 \end{dfn}
+{\color{blue} Here, $\eta_i : \textit{Lan}_{\ol{\oCPOsem{\Gamma;
+        \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}} \oCPOsem{\Gamma;
+    \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := \_}] \; \to \;
+  \pi_i\,(\textit{Lan}_{\ol{\oCPORsem{\Gamma; \ol\alpha \vdash
+        K}\Eq_\rho[\ol{\alpha := \Eq_{\_}}]}} \oCPORsem{\Gamma; \Phi,
+    \ol\alpha \vdash F}\Eq_\rho[\ol{\alpha := \Eq_{\_}}])$ for $i =
+  1,2$, i.e., $\eta_i$ is for the (mixed $\set/\rel$) diagram top p2
+  of Daniel's notes for relational environment $\Eq_\rho[\alpha :=
+    \Eq_{\_}]$. (Note that the codomains of $\eta_1$ and $\eta_2$
+  should be the same so we should have $\eta_1=\eta_2$). The condition
+  in the final clause should therefore simplify by writing $\eta$ for
+  $\eta_1 = \eta_2$.)}
+
 If $\rho \in \oCPOenv$ and $\vdash F$ then we write $\oCPOsem{\vdash F}$
@@
   &=
-  \{(t,t') \in \oCPOsem{\Gamma; \emptyset
+  \{(\eta_1,\eta_2) \in \oCPOsem{\Gamma; \emptyset
     \vdash \Nat^\Phi
@@
   & \hspace{0.3in} \forall \ol{K = (K^1, K^2, K^*) : \oCPORT_{k}}.\\
-  & \hspace{0.4in} (t_{\ol{K^1}},t'_{\ol{K^2}}) \in
+  & \hspace{0.4in} ((\eta_1)_{\ol{K^1}},(\eta_2)_{\ol{K^2}}) \in
   (\oCPORsem{\Gamma; \Phi \vdash G}\rho[\ol{\Phi :=
@@
     \delta][\ol{\alpha := \id_{\ol{R}}}] \\
-  \oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}\rho &=
-    (\textit{Lan}_{\ol{\oCPORsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}
-      \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := \_}])
-    \ol{\oCPORsem{\Gamma; \Phi \vdash A}\rho}\\
   \oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}}
-    F)\ol{A}}\rho &= \{ (t_1,t_2) : \oCPOsem{\Gamma;\Phi \vdash
+    F)\ol{A}}\rho &= \{ (t_1, t_2) : \oCPOsem{\Gamma;\Phi \vdash
     (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}\pi_1 \rho \,\times\,
@@
     F)\ol{A}}\pi_2 \rho\\
-  &\hspace{0.3in}|~( (\eta_1)_{\ol{\oCPORsem{\Gamma;\Phi \vdash A}\pi_1
-    \rho}} t_1,  (\eta_2)_{\ol{\oCPORsem{\Gamma;\Phi \vdash A}\pi_2
-    \rho}} t_2) \in\\
-  &\hspace{0.5in} (\textit{Lan}_{\ol{\oCPORsem{\Gamma; \ol\alpha \vdash
-          K}\rho[\ol{\alpha := \_}]}} \oCPORsem{\Gamma; \Phi, \ol\alpha
-      \vdash F}\rho[\ol{\alpha := \_}]) \ol{\oCPORsem{\Gamma; \Phi
-        \vdash A}\rho} \}\\
+  &\hspace{0.3in}|~{\color{blue}
+    ((\eta_1)_{\ol{\oCPOsem{\Gamma; \Phi \vdash A}\pi_1 \rho}} t_1,
+(\eta_2)_{\ol{\oCPOsem{\Gamma; \Phi \vdash A}\pi_2 \rho}} t_2) \in}\\
+  &\hspace{0.3in}{\color{blue}
+(\textit{Lan}_{\ol{\oCPORsem{\Gamma;
+          \ol\alpha \vdash K}\rho[\ol{\alpha := \Eq_{\_}}]}}
+    \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha :=
+        \Eq_{\_}}])
+    \ol{\oCPORsem{\Gamma; \Phi \vdash A}\rho}} \}  
+  %\forall \sigma : \setenv, (
+  %(\eta^\sigma_1)_{\ol{\oCPOsem{\Gamma;\Phi \vdash A}\sigma }} t_1,
+  %(\eta^\sigma_2)_{\ol{\oCPOsem{\Gamma;\Phi \vdash A}\sigma }} t_2) \in\\
+  %&\hspace{0.5in} (\textit{Lan}_{\ol{\oCPORsem{\Gamma; \ol\alpha
+  %      \vdash K}\Eq_\sigma[\ol{\alpha := \Eq_{\_}}]}}
+  %\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\Eq_\sigma[\ol{\alpha :=
+  %    \Eq_{\_}}]) \ol{\oCPORsem{\Gamma; \Phi \vdash A}\Eq_\sigma} \}\\
 \end{align*}
 \end{dfn}
-{\color{blue} In the final clause of Definition~\ref{def:rel-sem},
-  $\eta_1$ and $\eta_2$ are...}
+{\color{blue} where $\eta_i$ is for the (mixed $\set/\rel$) diagram
+  top p2 of Daniel's notes for relational environment $\Eq_{\pi_i
+    \rho}[\alpha := \Eq_{\_}]$.}
+
+%{\color{blue} In the final clause of Definition~\ref{def:rel-sem},
+%  $\eta^\sigma_i : \textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash
+%        K}\sigma[\ol{\alpha := \_}]}} \oCPOsem{\Gamma; \Phi, \ol\alpha
+%    \vdash F}\sigma[\ol{\alpha := \_}] \; \to \;
+%  \pi_i\,(\textit{Lan}_{\ol{\oCPORsem{\Gamma; \ol\alpha \vdash
+%        K}\Eq_\sigma[\ol{\alpha := \Eq_{\_}}]}} \oCPORsem{\Gamma;
+%    \Phi, \ol\alpha \vdash F}\Eq_\sigma[\ol{\alpha := \Eq_{\_}}])$ for
+%  each $\sigma : \setenv$ and $i = 1,2$.}
 

### 2020-09-02 20-gadts/inf-rules.tex
tweak condition for rel interp of Lans

@@
 $(A,B,R)$ with $R : \oCPOR(A,B)$. We may also write $\pi_1 F$ for
-$F^1$ and $\pi_2 F$ for $F^2$. We extend these conventions to $\oCPO$
-relation environments, introduced in Definition~\ref{def:reln-env}
-below, in the obvious way.
+$F^1$ and $\pi_2 F$ for $F^2$. {\color{blue} Without loss of
+  generality, we may assume that $\pi_1$ and $\pi_2$ are surjective on
+  relations.} We extend these conventions to $\oCPO$ relation
+environments, introduced in Definition~\ref{def:reln-env} below, in
+the obvious way.
 
@@
   &\hspace{0.3in}|~{\color{blue}
-    ((\eta_1)_{\ol{\oCPOsem{\Gamma; \Phi \vdash A}\pi_1 \rho}} t_1,
-(\eta_2)_{\ol{\oCPOsem{\Gamma; \Phi \vdash A}\pi_2 \rho}} t_2) \in}\\
+    (\nu_1 ((\eta_1)_{\ol{\oCPOsem{\Gamma; \Phi \vdash A}\pi_1 \rho}} t_1),
+\nu_2 ((\eta_2)_{\ol{\oCPOsem{\Gamma; \Phi \vdash A}\pi_2 \rho}} t_2)) \in}\\
   &\hspace{0.3in}{\color{blue}
@@
   top p2 of Daniel's notes for relational environment $\Eq_{\pi_i
-    \rho}[\alpha := \Eq_{\_}]$.}
+    \rho}[\alpha := \Eq_{\_}]$ and $\nu_i$ is the unique morphism in
+  $\set$ from \[(\textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash
+        K}\pi_i\rho[\ol{\alpha := \_}]}} \oCPOsem{\Gamma; \Phi,
+    \ol\alpha \vdash F}\pi_i\rho[\ol{\alpha := \_}])
+  \ol{\oCPOsem{\Gamma; \Phi \vdash A}\pi_i\rho}\] to
+\[\pi_i(\textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash
+      K}\rho[\ol{\alpha := \Eq_{\_}}]}} \oCPOsem{\Gamma; \Phi,
+  \ol\alpha \vdash F}\rho[\ol{\alpha := \Eq_{\_}}])
+\ol{\oCPOsem{\Gamma; \Phi \vdash A}\pi_i\rho}\]
+given by the colimit formulation of left Kan extensions.}
 

### 2020-09-03 20-gadts/inf-rules.tex
notes about parametricity for GADTs; corresponding restrictions on K in Lans

--- 2020-08-31 20-gadts/inf-rules.tex
@@
   actually parametric.}
+
+{\color{red} We should ask what, intuitively, it means to a GADT to be
+  ``parametric'', i.e., uniform in its type indices. For ADTs and
+  nested types it means that the datatype elements are constructed in
+  exactly the same ways at each type. For GADTs, it means that as
+  well, but here we have to take into account the inhabited instances
+  as well. [Of course we did not have to do that ADTs and nested types
+    because {\em all} instances of ADTs and nested types are
+    inhabited.] So, intuitively, we consider a GADT to be
+  ``parametric'' if we can construct the inhabitants of all of the
+  inhabited type instances in a type-independent way. We capture this
+  in our calculus by requiring that $\emptyset;\ol\beta \vdash K$ (and
+  that $K$ does not involve any constants?) in $\Gamma;\Phi \vdash
+  (\Lan^{\ol{\alpha^0}}_{\ol K} F)\, \ol A$. Then to get that the real
+  live Lans interpreting $\Lan$-constructs are $\omega$-cocontinuous
+  we also require that $K$ is a polynomial functor in the $\alpha$s
+  that does not involve any constant functors. This means that our
+  constructors have types of the form $F\ol\beta \to F(K\ol\beta)$
+  with no constants in $K$, so that exactly the corresponding
+  instances of the GADT are inhabited for every instantiation of the
+  $\alpha$s, and the objects for each instantiation of the $\alpha$s
+  are constructed in exactly the same way. That is, the inhabited
+  instances and the construction of the data in those instances are
+  type independent.}
 
@@
 \AXC{$\Gamma; \Phi,\ol\alpha^0 \vdash F$}
-\AXC{$\ol{\Gamma;\ol\alpha^0 \vdash K}$}
+\AXC{$\ol{{\color{blue} \emptyset};\ol\alpha^0 \vdash K}$}
 \AXC{$\ol{\Gamma;\Phi \vdash A}$}
@@
 \AXC{$\Gamma; \Phi,\ol{\alpha}\vdash F$}
-\AXC{$\ol{\Gamma; \ol{\alpha} \vdash K}$}
+\AXC{$\ol{{\color{blue} \emptyset}; \ol{\alpha} \vdash K}$}
 \AXC{$\ol{\Gamma;\Phi \vdash A} \hspace*{0.3in} \Gamma;\Phi~|~\Delta \vdash t : F[\ol{\alpha :=A}]$}
@@
   \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}\rho &=
-    \{ t : (\textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash
-          K}\rho[\ol{\alpha := \_}]}} \oCPOsem{\Gamma; \Phi, \ol\alpha
+    \{ t : (\textit{Lan}_{\ol{\oCPOsem{\emptyset; \ol\alpha \vdash
+          K}[\ol{\alpha := \_}]}} \oCPOsem{\Gamma; \Phi, \ol\alpha
       \vdash F}\rho[\ol{\alpha := \_}]) \ol{\oCPOsem{\Gamma; \Phi
@@
   (\eta_2)_{\ol{\oCPOsem{\Gamma;\Phi \vdash A}\rho }} t) \in}\\
-  &{\color{blue} \hspace{0.5in} (\textit{Lan}_{\ol{\oCPORsem{\Gamma; \ol\alpha
-        \vdash K}\Eq_\rho[\ol{\alpha := \Eq_{\_}}]}}
+  &{\color{blue} \hspace{0.5in}
+     (\textit{Lan}_{\ol{\oCPORsem{\emptyset; \ol\alpha 
+        \vdash K}[\ol{\alpha := \Eq_{\_}}]}}
   \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\Eq_\rho[\ol{\alpha :=
@@
 \end{dfn}
-{\color{blue} Here, $\eta_i : \textit{Lan}_{\ol{\oCPOsem{\Gamma;
-        \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}} \oCPOsem{\Gamma;
+{\color{blue} Here, $\eta_i : \textit{Lan}_{\ol{\oCPOsem{\emptyset;
+        \ol\alpha \vdash K}[\ol{\alpha := \_}]}} \oCPOsem{\Gamma;
     \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := \_}] \; \to \;
-  \pi_i\,(\textit{Lan}_{\ol{\oCPORsem{\Gamma; \ol\alpha \vdash
-        K}\Eq_\rho[\ol{\alpha := \Eq_{\_}}]}} \oCPORsem{\Gamma; \Phi,
+  \pi_i\,(\textit{Lan}_{\ol{\oCPORsem{\emptyset; \ol\alpha \vdash
+        K}[\ol{\alpha := \Eq_{\_}}]}} \oCPORsem{\Gamma; \Phi,
     \ol\alpha \vdash F}\Eq_\rho[\ol{\alpha := \Eq_{\_}}])$ for $i =
@@
 follows from Corollary~12 of~\cite{jp19} if we \new{APPROPRIATELY
-  RESTRICT THE SUBSCRIPTS OF $\Lan$s}.  For $\Nat$-types, we note that
-$\oCPOsem{\Gamma; \emptyset \vdash \Nat^\Phi F\,G}$ is an
-$\omega_1$-cocontinuous functor because, in accordance with
+  RESTRICT THE SUBSCRIPTS OF $\Lan$s; we will require that $K$ is a
+  polynomial functor in the $\alpha$s that does not involve any
+  constant functors}. For $\Nat$-types, we note that $\oCPOsem{\Gamma;
+  \emptyset \vdash \Nat^\Phi F\,G}$ is an $\omega_1$-cocontinuous
+functor because, in accordance with
 Definition~\ref{def:set-sem-funcs}, it is constant on
@@
 \[\begin{array}{ll}
- & \big(\textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}\,
+ & \big(\textit{Lan}_{\ol{\oCPOsem{\emptyset; \ol\alpha \vdash K}[\ol{\alpha := \_}]}}\,
       \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha := \id_{\_}}]\big)
@@
  &  \circ\;
-  \big(\textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}\,
+  \big(\textit{Lan}_{\ol{\oCPOsem{\emptyset; \ol\alpha \vdash K}[\ol{\alpha := \_}]}}\,
       \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := \_}]\big)
     \ol{\oCPOsem{\Gamma; \Phi \vdash A}f} \\
-= & \big(\textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}\,
+= & \big(\textit{Lan}_{\ol{\oCPOsem{\emptyset; \ol\alpha \vdash K}[\ol{\alpha := \_}]}}\,
       \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := \_}]\big)
@@
  &  \circ\;
-  \big(\textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}\,
+  \big(\textit{Lan}_{\ol{\oCPOsem{\emptyset; \ol\alpha \vdash K}[\ol{\alpha := \_}]}}\,
       \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha := \id_{\_}}]\big)
@@
 where the above equality holds by naturality of
-$\textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash
-        K}\rho[\ol{\alpha := \_}]}}\,\oCPOsem{\Gamma; \Phi, \ol\alpha
+$\textit{Lan}_{\ol{\oCPOsem{\emptyset; \ol\alpha \vdash
+        K}[\ol{\alpha := \_}]}}\,\oCPOsem{\Gamma; \Phi, \ol\alpha
     \vdash F}f[\ol{\alpha := \id_{\_}}]$.
@@
   &\hspace{0.3in}{\color{blue}
-(\textit{Lan}_{\ol{\oCPORsem{\Gamma;
-          \ol\alpha \vdash K}\rho[\ol{\alpha := \Eq_{\_}}]}}
+(\textit{Lan}_{\ol{\oCPORsem{\emptyset;
+          \ol\alpha \vdash K}[\ol{\alpha := \Eq_{\_}}]}}
     \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha :=

### 2020-09-03 20-gadts/inf-rules.tex
added more intution about par for GADTs at start of main paper

@@
 
-{\color{red} We should ask what, intuitively, it means to a GADT to be
-  ``parametric'', i.e., uniform in its type indices. For ADTs and
-  nested types it means that the datatype elements are constructed in
-  exactly the same ways at each type. For GADTs, it means that as
-  well, but here we have to take into account the inhabited instances
-  as well. [Of course we did not have to do that ADTs and nested types
+{\color{red} We should ask what, intuitively, it means to have
+  parametricity for a calculus with GADTs. In particular it should
+  mean that any polymorphic function that takes a GADT as input should
+  operate uniformly over all type instances of the GADT. That means
+  that the GADT must itself be ``parametric'', i.e., uniform in its
+  type indices. But what does this mean? For ADTs and nested types it
+  means that the datatype elements are constructed in exactly the same
+  ways at each type instance. For GADTs, it means that as well, but
+  here we also have to take into account which instances are actually
+  inhabited, since not all type instances of a GADT are inhabited.
+  [Of course we did not have to do this for ADTs and nested types
     because {\em all} instances of ADTs and nested types are
     inhabited.] So, intuitively, we consider a GADT to be
-  ``parametric'' if we can construct the inhabitants of all of the
-  inhabited type instances in a type-independent way. We capture this
-  in our calculus by requiring that $\emptyset;\ol\beta \vdash K$ (and
-  that $K$ does not involve any constants?) in $\Gamma;\Phi \vdash
-  (\Lan^{\ol{\alpha^0}}_{\ol K} F)\, \ol A$. Then to get that the real
-  live Lans interpreting $\Lan$-constructs are $\omega$-cocontinuous
-  we also require that $K$ is a polynomial functor in the $\alpha$s
-  that does not involve any constant functors. This means that our
-  constructors have types of the form $F\ol\beta \to F(K\ol\beta)$
-  with no constants in $K$, so that exactly the corresponding
-  instances of the GADT are inhabited for every instantiation of the
-  $\alpha$s, and the objects for each instantiation of the $\alpha$s
-  are constructed in exactly the same way. That is, the inhabited
-  instances and the construction of the data in those instances are
-  type independent.}
+  ``parametric'' if 1) the corresponding instances of the GADT are
+  inhabited for for every instantiation of its type parameters
+  $\ol\alpha$, and 2) the objects for each instantiation of the
+  $\alpha$s are constructed in exactly the same way. That is, which
+  instances are inhabited, as well as how the data in each of those
+  instances are constructed, is type independent. We capture this in
+  our calculus by requiring that, for $\Gamma;\Phi \vdash
+  (\Lan^{\ol{\alpha^0}}_{\ol K} F)\, \ol A$, we have that
+  $\emptyset;\ol\beta \vdash K$, and that $K$ does not involve any
+  constants. (The latter ensures that a GADT with a constructor of
+  type $\alpha \to Int$ is not parametric, which makes sense since
+  there will be more ways to construct Int instances of the GADT than
+  ways to construct its other type instances.)  Further, to ensure
+  that the categorical Lans interpreting $\Lan$-constructs are
+  $\omega$-cocontinuous we also require that $K$ is a polynomial
+  functor in the $\alpha$s that does not involve any constant
+  functors. This means that our constructors have types of the form
+  $F\ol\beta \to F(K\ol\beta)$ with no constants in $K$, so that
+  exactly the corresponding instances of the GADT are inhabited for
+  every instantiation of the $\alpha$s, and the objects for each
+  instantiation of the $\alpha$s are constructed in exactly the same
+  way. That is, the inhabited instances and the construction of the
+  data in those instances are type independent.}
 

### 2020-09-17 20-gadts/inf-rules.tex
actions of interps of Lans defined on objects and IEL proved(?); actions of interps on morphisms still to be defined

@@
   there will be more ways to construct Int instances of the GADT than
-  ways to construct its other type instances.)  Further, to ensure
-  that the categorical Lans interpreting $\Lan$-constructs are
+  ways to construct its other type instances.) Further, to ensure that
+  the categorical Lans interpreting $\Lan$-constructs are
   $\omega$-cocontinuous we also require that $K$ is a polynomial
-  functor in the $\alpha$s that does not involve any constant
-  functors. This means that our constructors have types of the form
-  $F\ol\beta \to F(K\ol\beta)$ with no constants in $K$, so that
-  exactly the corresponding instances of the GADT are inhabited for
-  every instantiation of the $\alpha$s, and the objects for each
+  functor in the $\alpha$s (that necessarily will not involve any
+  constant functors). This means that our constructors have types of
+  the form $F\ol\beta \to F(K\ol\beta)$ with no constants in $K$, so
+  that exactly the corresponding instances of the GADT are inhabited
+  for every instantiation of the $\alpha$s, and the objects for each
   instantiation of the $\alpha$s are constructed in exactly the same
@@
 \AXC{$\Gamma; \Phi,\ol\alpha^0 \vdash F$}
-\AXC{$\ol{{\color{blue} \emptyset};\ol\alpha^0 \vdash K}$}
+\AXC{$\ol{\Gamma;\ol\alpha^0 \vdash K}$}
 \AXC{$\ol{\Gamma;\Phi \vdash A}$}
@@
 \AXC{$\Gamma; \Phi,\ol{\alpha}\vdash F$}
-\AXC{$\ol{{\color{blue} \emptyset}; \ol{\alpha} \vdash K}$}
+\AXC{$\ol{\Gamma; \ol{\alpha} \vdash K}$}
 \AXC{$\ol{\Gamma;\Phi \vdash A} \hspace*{0.3in} \Gamma;\Phi~|~\Delta \vdash t : F[\ol{\alpha :=A}]$}
@@
 
+{\color{red} NOT NEEDED ANYMORE??
 To accommodate GADTs, we will need to transition Reynolds' approach
@@
 (\bigvee_{i < \omega} (f a_i), \bigvee_{i < \omega} (g b_i)) \in R'$
-necessarily holds.  We write $R : \oCPOR(A,B)$ in place of $(A,B,R):
+necessarily holds.}  We write $R : \oCPOR(A,B)$ in place of $(A,B,R):
 \oCPOR$ when convenient. If $R : \oCPOR(A,B)$ then we write $\pi_1 R$
@@
 The object interpretations of the types in our calculus will be
-$\omega_1$-cocontinuous functors between categories of
-$\omega_1$-cocontinuous functors on categories constructed from the
-locally $\omega_1$-presentable category $\oCPO$.  We therefore begin
-by recording some important facts about locally $\omega_1$-presentable
+$\omega$-cocontinuous functors between categories of
+$\omega$-cocontinuous functors on categories constructed from the
+locally $\omega$-presentable category $\oCPO$.  We therefore begin
+by recording some important facts about locally $\omega$-presentable
 categories and functors on them, and verifying verify the properties
@@
 
-\new{Perhaps have as preliminaries to entire paper. Do everything for
-  $\lambda CPOs$? Define these; investigate their properties.}
+\new{Perhaps have as preliminaries to entire paper.}
 
@@
 
-A poset $\mcD = (D,\leq)$ is \emph{$\omega_1$-directed} if every
+A poset $\mcD = (D,\leq)$ is \emph{$\omega$-directed} if every
 countable subset of $D$ has a supremum. When $\mcD$ is considered
 as a category, we write $d \in \mcD$ to indicate that $d$ is an object
-of $\mcD$ (i.e., $d \in D$). An {\em $\omega_1$-directed colimit} in a
+of $\mcD$ (i.e., $d \in D$). An {\em $\omega$-directed colimit} in a
 category $\mcC$ is a colimit of a diagram $F : {\mathcal D} \to \mcC$,
-where $\mathcal D$ is an $\omega_1$-directed poset. A category $\mcC$
-is {\em $\omega_1$-cocomplete} if it has all $\omega_1$-directed
+where $\mathcal D$ is an $\omega$-directed poset. A category $\mcC$
+is {\em $\omega$-cocomplete} if it has all $\omega$-directed
 colimits; a {\em cocomplete} category is one that has all colimits.
 
-If $\mcA$ and $\mcC$ are $\omega_1$-cocomplete, then a functor $F :
-\mcA \to \mcC$ is {\em $\omega_1$-cocontinuous} if it preserves
-$\omega_1$-directed colimits. We write $[\A,\C]_{\omega_1}$ for the
-category of $\omega_1$-cocontinuous functors from $\A$ to $\C$, and
+If $\mcA$ and $\mcC$ are $\omega$-cocomplete, then a functor $F :
+\mcA \to \mcC$ is {\em $\omega$-cocontinuous} if it preserves
+$\omega$-directed colimits. We write $[\A,\C]_{\omega}$ for the
+category of $\omega$-cocontinuous functors from $\A$ to $\C$, and
 $\C^\A$ for the category of {\em all} functors from $\A$ to $\C$.
@@
 (co)limits that $\C$ has, and (co)limits of (co)continuous functors
-are again (co)continuous. It follows that $[\A,\C]_{(\omega_1)}$ is
-($\omega_1$-)(co)complete whenever $\C$ is. In particular, if $F :
-[\C, \C]_{(\omega_1)}$ for a(n) ($\omega_1$-)cocomplete category $\C$,
+are again (co)continuous. It follows that $[\A,\C]_{(\omega)}$ is
+($\omega$-)(co)complete whenever $\C$ is. In particular, if $F :
+[\C, \C]_{(\omega)}$ for a(n) ($\omega$-)cocomplete category $\C$,
 then we can define $F^0 = \Id$, $F^{\alpha + 1} = F \circ F^\alpha$ at
@@
 If $\mcA$ is locally small, then an object $A$ of $\mcA$ is {\em
-  $\omega_1$-presentable} if the functor $\mathsf{Hom}_\mcA(A, -) :
-\mcA \to \Set$ preserves $\omega_1$-directed colimits, i.e., if for
-every $\omega_1$-directed poset $\mathcal D$ and every functor $F :
+  $\omega$-presentable} if the functor $\mathsf{Hom}_\mcA(A, -) :
+\mcA \to \Set$ preserves $\omega$-directed colimits, i.e., if for
+every $\omega$-directed poset $\mathcal D$ and every functor $F :
 {\mathcal D} \to \A$, there is a canonical isomorphism $\colim{d \in
@@
 \colim{d \in \mcD}{Fd})$. A locally small category $\mcA$ is {\em
-  $\omega_1$-accessible} if it is $\omega_1$-cocomplete and has a set
-$\mcA_0$ of $\omega_1$-presentable objects such that every object is
-an $\omega_1$-directed colimit of objects in $\mcA_0$; a locally small
-category is {\em locally $\omega_1$-presentable} if it is
-$\omega_1$-accessible and cocomplete.
-
-The category $\oCPO$ is locally $\omega_1$-presentable (but not
-locally finitely presentable); see Examples~1.18(2)
-of~\cite{ar94}). Its $\omega_1$-presentable objects are precisely the
-$\omega$CPOs cardinality less than $\omega_1$, i.e., the countable
-$\omega$CPOs. In the next subsection we will interpret type variables
-in $\tvars^k \cup \fvars^k$ as elements of
-$[\oCPO^k,\oCPO]_{\omega_1}$; the following special cases of standard
-results (see, e.g.,~\cite{ar94}) will therefore be critical to
-deducing important properties of our object interpretations of types:
+  $\omega$-accessible} if it is $\omega$-cocomplete and has a set
+$\mcA_0$ of $\omega$-presentable objects such that every object is
+an $\omega$-directed colimit of objects in $\mcA_0$; a locally small
+category is {\em locally $\omega$-presentable} if it is
+$\omega$-accessible and cocomplete.
+
+The category $\oCPO$ is locally presentable; its presentable objects
+are precisely the finite sets.
+%The category $\oCPO$ is locally $\omega$-presentable (but not
+%locally finitely presentable); see Examples~1.18(2)
+%of~\cite{ar94}). Its $\omega$-presentable objects are precisely the
+%$\omega$CPOs cardinality less than $\omega$, i.e., the countable
+%$\omega$CPOs.
+In the next subsection we will interpret type variables in $\tvars^k
+\cup \fvars^k$ as elements of $[\oCPO^k,\oCPO]_{\omega}$; the
+following special cases of standard results (see, e.g.,~\cite{ar94})
+will therefore be critical to deducing important properties of our
+object interpretations of types:
 \begin{prop}\label{prop:prelims}
 \begin{enumerate}
-\item If $\C_1,...,\C_n$ are locally $\omega_1$-presentable categories
+\item If $\C_1,...,\C_n$ are locally $\omega$-presentable categories
   then so is $\C_1 \times ... \times \C_n$. Moreover, the presentable
@@
   $P_i$ is presentable in $\C_i$.
-\item If $\mathcal{A}$ is $\omega_1$-accessible and $\mathcal{C}$ is
+\item If $\mathcal{A}$ is $\omega$-accessible and $\mathcal{C}$ is
   $\lambda$-cocomplete, then the category $[\mathcal{A},
-  \mathcal{C}]_{\omega_1}$ is naturally equivalent to the category
+  \mathcal{C}]_{\omega}$ is naturally equivalent to the category
   $\mathcal{C}^{\mathcal{A}_0}$.
-\item If $\mathcal{C}$ is locally $\omega_1$-presentable and
+\item If $\mathcal{C}$ is locally $\omega$-presentable and
   $\mathcal{A}_0$ is essentially small, then
-  $\mathcal{C}^{\mathcal{A}_0}$ is locally $\omega_1$-presentable.
+  $\mathcal{C}^{\mathcal{A}_0}$ is locally $\omega$-presentable.
 \end{enumerate}
@@
 Together, the statements in Proposition~\ref{prop:prelims} give that
-if $\mathcal{A}$ and $\mathcal{C}$ are locally $\omega_1$-presentable,
-then $[\mathcal{A}, \mathcal{C}]_{\omega_1}$ is naturally equivalent
+if $\mathcal{A}$ and $\mathcal{C}$ are locally $\omega$-presentable,
+then $[\mathcal{A}, \mathcal{C}]_{\omega}$ is naturally equivalent
 to $\mathcal{C}^{\mathcal{A}_0}$, and thus is
-$\omega_1$-presentable. Thus, for all $k_1,...,k_n \in \nat^n$,
-$[\mathcal{A}^{k_1}, \mathcal{C}]_{\omega_1} \times ... \times
-[\mathcal{A}^{k_n}, \mathcal{C}]_{\omega_1}$ is locally
-$\omega_1$-presentable, and therefore $[[\mathcal{A}^{k_1},
-    \mathcal{C}]_{\omega_1} \times ... \times [\mathcal{A}^{k_n},
-    \mathcal{C}]_{\omega_1}, \mathcal{C}]_{\omega_1}$ is as
+$\omega$-presentable. Thus, for all $k_1,...,k_n \in \nat^n$,
+$[\mathcal{A}^{k_1}, \mathcal{C}]_{\omega} \times ... \times
+[\mathcal{A}^{k_n}, \mathcal{C}]_{\omega}$ is locally
+$\omega$-presentable, and therefore $[[\mathcal{A}^{k_1},
+    \mathcal{C}]_{\omega} \times ... \times [\mathcal{A}^{k_n},
+    \mathcal{C}]_{\omega}, \mathcal{C}]_{\omega}$ is as
 well. Taking both $\A$ and $C$ to be $\oCPO$ --- as we will to ensure
@@
   For all $k_1,...,k_n \in \nat^n$,
-  \[[[\oCPO^{k_1}, \oCPO]_{\omega_1}
-  \times ... \times [\oCPO^{k_n}, \oCPO]_{\omega_1},
-  \oCPO]_{\omega_1}\] is locally $\omega_1$-presentable.
+  \[[[\oCPO^{k_1}, \oCPO]_{\omega}
+  \times ... \times [\oCPO^{k_n}, \oCPO]_{\omega},
+  \oCPO]_{\omega}\] is locally $\omega$-presentable.
 \end{prop}
@@
 A {\em $\oCPO$ environment} maps each type variable in $\tvars^k \cup
-\fvars^k$ to an element of $[\oCPO^k,\oCPO]_{\omega_1}$.  A morphism
+\fvars^k$ to an element of $[\oCPO^k,\oCPO]_{\omega}$.  A morphism
 $f : \rho \to \rho'$ for set environments $\rho$ and $\rho'$ with
@@
 
-When convenient we identify a functor in $[\oCPO^0, \oCPO]_{\omega_1}$
+When convenient we identify a functor in $[\oCPO^0, \oCPO]_{\omega}$
 with its value on $\ast$ and consider a $\oCPO$ environment to map a
-type variable of arity $0$ to an $\omega_1$-cocontinuous functor from
+type variable of arity $0$ to an $\omega$-cocontinuous functor from
 $\oCPO^0$ to $\oCPO$, i.e., to an $\oCPO$. If $\Phi =
 \{\phi_1^{k_1},...,\phi_n^{k_n}\}$ and $\ol K = \{K_1,...,K_n\}$ where
-$K_i : [\oCPO^{k_i},\oCPO]_{\omega_1}$ for $i = 1,...,n$, then we
+$K_i : [\oCPO^{k_i},\oCPO]_{\omega}$ for $i = 1,...,n$, then we
 write either $\rho[\ol{\Phi := K}]$ or $\rho[\ol{\phi := K}]$ for the
@@
 clause of Definition~\ref{def:set-sem} are given in full in
-Section~\ref{sec:rel-interp}. We note that $\omega_1$-directed
+Section~\ref{sec:rel-interp}. We note that $\omega$-directed
 colimits in $\oCPOenv$ are taken pointwise.
@@
 The {\em object interpretation} $\oCPOsem{\cdot} : \F \to [\oCPOenv,
-  \oCPO]_{\omega_1}$ is defined by
+  \oCPO]_{\omega}$ is defined by
 \begin{align*}
@@
   \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}\rho &=
-    \{ t : (\textit{Lan}_{\ol{\oCPOsem{\emptyset; \ol\alpha \vdash
-          K}[\ol{\alpha := \_}]}} \oCPOsem{\Gamma; \Phi, \ol\alpha
+    \{ t : (\textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash
+          K}\rho[\ol{\alpha := \_}]}} \oCPOsem{\Gamma; \Phi, \ol\alpha
       \vdash F}\rho[\ol{\alpha := \_}]) \ol{\oCPOsem{\Gamma; \Phi
         \vdash A}\rho}\\
-    &\hspace{0.3in}|~
-%    (t,t) \in \oCPORsem{\Gamma;\Phi \vdash
-%      (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}\Eq_\rho\}\\
-{\color{blue} ( (\eta_1)_{\ol{\oCPOsem{\Gamma;\Phi \vdash A}\rho }} t,
-  (\eta_2)_{\ol{\oCPOsem{\Gamma;\Phi \vdash A}\rho }} t) \in}\\
-  &{\color{blue} \hspace{0.5in}
-     (\textit{Lan}_{\ol{\oCPORsem{\emptyset; \ol\alpha 
-        \vdash K}[\ol{\alpha := \Eq_{\_}}]}}
-  \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\Eq_\rho[\ol{\alpha :=
-      \Eq_{\_}}]) \ol{\oCPORsem{\Gamma; \Phi \vdash A}\Eq_\rho} \}}\\
+    &\hspace{0.3in}|~(t,t) \in \oCPORsem{\Gamma;\Phi \vdash
+      (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}\Eq_\rho\}\\
 \end{align*}
 \end{dfn}
-{\color{blue} Here, $\eta_i : \textit{Lan}_{\ol{\oCPOsem{\emptyset;
-        \ol\alpha \vdash K}[\ol{\alpha := \_}]}} \oCPOsem{\Gamma;
-    \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := \_}] \; \to \;
-  \pi_i\,(\textit{Lan}_{\ol{\oCPORsem{\emptyset; \ol\alpha \vdash
-        K}[\ol{\alpha := \Eq_{\_}}]}} \oCPORsem{\Gamma; \Phi,
-    \ol\alpha \vdash F}\Eq_\rho[\ol{\alpha := \Eq_{\_}}])$ for $i =
-  1,2$, i.e., $\eta_i$ is for the (mixed $\set/\rel$) diagram top p2
-  of Daniel's notes for relational environment $\Eq_\rho[\alpha :=
-    \Eq_{\_}]$. (Note that the codomains of $\eta_1$ and $\eta_2$
-  should be the same so we should have $\eta_1=\eta_2$). The condition
-  in the final clause should therefore simplify by writing $\eta$ for
-  $\eta_1 = \eta_2$.)}
 
@@
   \vdash G}\rho[\ol{\Phi := K}]$ are both in $[\ol{[\oCPO^k ,
-      \oCPO]_{\omega_1}}, \oCPO]_{\omega_1}$. By
+      \oCPO]_{\omega}}, \oCPO]_{\omega}$. By
 Proposition~\ref{prop:nat-interps-sets}, $[\ol{[\oCPO^k ,
-      \oCPO]_{\omega_1}}, \oCPO]_{\omega_1}$ is locally
-$\omega_1$-presentable. It is therefore locally small, so there are
+      \oCPO]_{\omega}}, \oCPO]_{\omega}$ is locally
+$\omega$-presentable. It is therefore locally small, so there are
 only $\set$-many morphisms (i.e., natural transformations) between any
-two functors in $[\ol{[\oCPO^k , \oCPO]_{\omega_1}},
-  \oCPO]_{\omega_1}$. In particular, there are only $\set$-many
+two functors in $[\ol{[\oCPO^k , \oCPO]_{\omega}},
+  \oCPO]_{\omega}$. In particular, there are only $\set$-many
 natural transformations from $\lambda \ol K. \setsem{\Gamma;\Phi
@@
 
-Now, if $\C$ is any category, then given functors $F,G : \C \to \oCPO$
-and natural transformations $\eta, \eta' : F \to G$, we define $\eta
-\leq \eta'$ iff $\eta_c \leq \eta'_c$ for all $c \in \C$, i.e., iff
-$\eta_c\,x \leq \eta'_c\,x$ in $Gc$ for all $c \in \C$ and $x \in Fc$.
-If $(\eta_i)_{i < \omega}$ is a chain in $\{\eta : F \to G\}$, then
-the family of morphisms $(\bigvee_{i < \omega} \eta_i)_c = \lambda
-x. \bigvee_{i < \omega} ((\eta_i)_c\,x) : Fc \to Gc$ defines a natural
-transformation $\bigvee_{i < \omega} \eta_i : F \to G$, and this
-natural transformation is clearly the supremum of $(\eta_i)_{i <  
-  \omega}$ in $\{\eta : F \to G\}$. We therefore have that $\{\eta : F
-\to G\}$ is itself an $\oCPO$.  To show that $\oCPOsem{\Gamma;
-  \emptyset \vdash \Nat^\Phi \,F\,G}\rho$ is an $\oCPO$, we then still
-need to prove that, for any chain of natural transformations
-$(\eta_i)_{i < \omega}$ in $\oCPOsem{\Gamma; \emptyset \vdash
-  \Nat^\Phi \,F\,G}\rho$, $\bigvee_{i < \omega} \eta_i$ is again in
-$\oCPOsem{\Gamma; \emptyset \vdash \Nat^\Phi \,F\,G}\rho$.  By
-hypothesis, each $\eta_i$ is such that, for all $\ol{K = (K^1, K^2,
-  K^*) : \oCPORT_k}$,
-\[((\eta_i)_{\overline{K^1}},
-(\eta_i)_{\overline{K^2}}) : \oCPORsem{\Gamma; \Phi \vdash
-  F}\Eq_{\rho}[\ol{\Phi := K}] \rightarrow \oCPORsem{\Gamma; \Phi
-  \vdash G}\Eq_{\rho}[\ol{\Phi := K}]\] is a morphism in
-$\oCPOR$. To see that, for all $\ol{K = (K^1, K^2, K^*) : \oCPORT_k}$,
-%\[((\bigvee_{i < \omega} \eta_i)_{\overline{K^1}}, (\bigvee_{i <
-%%  \omega} \eta_i)_{\overline{K^2}}) : \oCPORsem{\Gamma; \Phi \vdash
+%Now, if $\C$ is any category, then given functors $F,G : \C \to \oCPO$
+%and natural transformations $\eta, \eta' : F \to G$, we define $\eta
+%\leq \eta'$ iff $\eta_c \leq \eta'_c$ for all $c \in \C$, i.e., iff
+%$\eta_c\,x \leq \eta'_c\,x$ in $Gc$ for all $c \in \C$ and $x \in Fc$.
+%If $(\eta_i)_{i < \omega}$ is a chain in $\{\eta : F \to G\}$, then
+%the family of morphisms $(\bigvee_{i < \omega} \eta_i)_c = \lambda
+%x. \bigvee_{i < \omega} ((\eta_i)_c\,x) : Fc \to Gc$ defines a natural
+%transformation $\bigvee_{i < \omega} \eta_i : F \to G$, and this
+%natural transformation is clearly the supremum of $(\eta_i)_{i <  
+%  \omega}$ in $\{\eta : F \to G\}$. We therefore have that $\{\eta : F
+%\to G\}$ is itself an $\oCPO$.  To show that $\oCPOsem{\Gamma;
+%  \emptyset \vdash \Nat^\Phi \,F\,G}\rho$ is an $\oCPO$, we then still
+%need to prove that, for any chain of natural transformations
+%$(\eta_i)_{i < \omega}$ in $\oCPOsem{\Gamma; \emptyset \vdash
+%  \Nat^\Phi \,F\,G}\rho$, $\bigvee_{i < \omega} \eta_i$ is again in
+%$\oCPOsem{\Gamma; \emptyset \vdash \Nat^\Phi \,F\,G}\rho$.  By
+%hypothesis, each $\eta_i$ is such that, for all $\ol{K = (K^1, K^2,
+%  K^*) : \oCPORT_k}$,
+%\[((\eta_i)_{\overline{K^1}},
+%(\eta_i)_{\overline{K^2}}) : \oCPORsem{\Gamma; \Phi \vdash
 %  F}\Eq_{\rho}[\ol{\Phi := K}] \rightarrow \oCPORsem{\Gamma; \Phi
-%  \vdash G}\Eq_{\rho}[\ol{\Phi := K}]\] is again a morphism in
-%$\oCPOR$, we observe that
-$((\bigvee_{i < \omega} \eta_i)_{\overline{K^1}}, (\bigvee_{i <
-  \omega} \eta_i)_{\overline{K^2}})$ is a morphism from
-$\oCPORsem{\Gamma; \Phi \vdash F}\Eq_{\rho}[\ol{\Phi := K}]$ to
-$\oCPORsem{\Gamma; \Phi \vdash G}\Eq_{\rho}[\ol{\Phi := K}]$ in
-$\oCPOR$, we simply observe that
-\[((\bigvee_{i < \omega} \eta_i)_{\overline{K^1}}, (\bigvee_{i < \omega}
-\eta_i)_{\overline{K^2}}) = (\bigvee_{i < \omega}
-(\eta_i)_{\overline{K^1}}, \bigvee_{i < \omega}
-(\eta_i)_{\overline{K^2}}) = \bigvee_{i < \omega}
-((\eta_i)_{\overline{K^1}}, (\eta_i)_{\overline{K^2}})\] so
-$((\bigvee_{i < \omega} \eta_i)_{\overline{K^1}}, (\bigvee_{i <
-  \omega} \eta_i)_{\overline{K^2}})$ is the sup of a chain of
-morphisms from $\oCPORsem{\Gamma; \Phi \vdash F}\Eq_{\rho}[\ol{\Phi :=
-    K}]$ to $\oCPORsem{\Gamma; \Phi \vdash G}\Eq_{\rho}[\ol{\Phi :=
-    K}]$ in $\oCPOR$ and thus is itself such a morphism. Taken
-together, this all shows that $\oCPOsem{\Gamma; \emptyset \vdash
-  \Nat^\Phi \,F\,G}\rho$ is indeed an $\oCPO$.
-
-That each of the above interpretations is $\omega_1$-cocontinuous
+%  \vdash G}\Eq_{\rho}[\ol{\Phi := K}]\] is a morphism in
+%$\oCPOR$. To see that, for all $\ol{K = (K^1, K^2, K^*) : \oCPORT_k}$,
+%%\[((\bigvee_{i < \omega} \eta_i)_{\overline{K^1}}, (\bigvee_{i <
+%%%  \omega} \eta_i)_{\overline{K^2}}) : \oCPORsem{\Gamma; \Phi \vdash
+%%  F}\Eq_{\rho}[\ol{\Phi := K}] \rightarrow \oCPORsem{\Gamma; \Phi
+%%  \vdash G}\Eq_{\rho}[\ol{\Phi := K}]\] is again a morphism in
+%%$\oCPOR$, we observe that
+%$((\bigvee_{i < \omega} \eta_i)_{\overline{K^1}}, (\bigvee_{i <
+%  \omega} \eta_i)_{\overline{K^2}})$ is a morphism from
+%$\oCPORsem{\Gamma; \Phi \vdash F}\Eq_{\rho}[\ol{\Phi := K}]$ to
+%$\oCPORsem{\Gamma; \Phi \vdash G}\Eq_{\rho}[\ol{\Phi := K}]$ in
+%$\oCPOR$, we simply observe that
+%\[((\bigvee_{i < \omega} \eta_i)_{\overline{K^1}}, (\bigvee_{i < \omega}
+%\eta_i)_{\overline{K^2}}) = (\bigvee_{i < \omega}
+%(\eta_i)_{\overline{K^1}}, \bigvee_{i < \omega}
+%(\eta_i)_{\overline{K^2}}) = \bigvee_{i < \omega}
+%((\eta_i)_{\overline{K^1}}, (\eta_i)_{\overline{K^2}})\] so
+%$((\bigvee_{i < \omega} \eta_i)_{\overline{K^1}}, (\bigvee_{i <
+%  \omega} \eta_i)_{\overline{K^2}})$ is the sup of a chain of
+%morphisms from $\oCPORsem{\Gamma; \Phi \vdash F}\Eq_{\rho}[\ol{\Phi :=
+%    K}]$ to $\oCPORsem{\Gamma; \Phi \vdash G}\Eq_{\rho}[\ol{\Phi :=
+%    K}]$ in $\oCPOR$ and thus is itself such a morphism. Taken
+%together, this all shows that $\oCPOsem{\Gamma; \emptyset \vdash
+%  \Nat^\Phi \,F\,G}\rho$ is indeed an $\oCPO$.
+
+That each of the above interpretations is $\omega$-cocontinuous
 follows from Corollary~12 of~\cite{jp19} if we \new{APPROPRIATELY
@@
   constant functors}. For $\Nat$-types, we note that $\oCPOsem{\Gamma;
-  \emptyset \vdash \Nat^\Phi F\,G}$ is an $\omega_1$-cocontinuous
+  \emptyset \vdash \Nat^\Phi F\,G}$ is an $\omega$-cocontinuous
 functor because, in accordance with
 Definition~\ref{def:set-sem-funcs}, it is constant on
-$\omega_1$-directed sets. Interpretations of $\Nat$-types ensure that
+$\omega$-directed sets. Interpretations of $\Nat$-types ensure that
 $\oCPOsem{\Gamma \vdash F \to G}$ and $\oCPOsem{\Gamma \vdash \forall
@@
 Definition~\ref{def:set-sem}, we need to know that, for each $\rho \in
-\oCPOenv$, $T^\oCPO_{H,\rho}$ is an $\omega_1$-cocontinuous
-endofunctor on $[\oCPO^k, \oCPO]_{\omega_1}$, and thus admits a
+\oCPOenv$, $T^\oCPO_{H,\rho}$ is an $\omega$-cocontinuous
+endofunctor on $[\oCPO^k, \oCPO]_{\omega}$, and thus admits a
 fixpoint.  Since $T_{H,\rho}^\oCPO$ is defined in terms of
@@
 $\oCPOsem{\Gamma; \Phi \vdash F}$ is actually in
-$[\oCPO^k,\oCPO]_{\omega_1}$, where $k = |\ol\alpha|$.  Therefore, for
+$[\oCPO^k,\oCPO]_{\omega}$, where $k = |\ol\alpha|$.  Therefore, for
 each $\oCPOsem{\Gamma; \ol\gamma, \phi^k, \ol{\alpha} \vdash H}$, the
@@
   functor} from $\oCPOenv$ to
-$[[\oCPO^k,\oCPO]_{\omega_1},[\oCPO^k,\oCPO]_{\omega_1}]_{\omega_1}$.
+$[[\oCPO^k,\oCPO]_{\omega},[\oCPO^k,\oCPO]_{\omega}]_{\omega}$.
 The action of $T^\oCPO_H$ on an object $\rho \in \oCPOenv$ is given by
 the higher-order functor $T_{H,\rho}^\oCPO$, whose actions on objects
-(functors in $[\oCPO^k, \oCPO]_{\omega_1}$) and on morphisms (natural
+(functors in $[\oCPO^k, \oCPO]_{\omega}$) and on morphisms (natural
 transformations between such functors) are given in
@@
 $T^\oCPO_{H,f} : T^\oCPO_{H,\rho} \to T^\oCPO_{H,\rho'}$ whose action
-on $F : [\oCPO^k,\oCPO]_{\omega_1}$ is the natural transformation
+on $F : [\oCPO^k,\oCPO]_{\omega}$ is the natural transformation
 $T^\oCPO_{H,f}\, F : T^\oCPO_{H,\rho}\,F \to T^\oCPO_{H,\rho'}\,F$
@@
 %and $\oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha :=
-%    \_}]$ to be $\omega_1$-cocontinuous functors.
+%    \_}]$ to be $\omega$-cocontinuous functors.
 In addition, for each $\ol K$, we have that $\textit{Lan}_{\ol K}$ is
@@
 \[\begin{array}{ll}
- & \big(\textit{Lan}_{\ol{\oCPOsem{\emptyset; \ol\alpha \vdash K}[\ol{\alpha := \_}]}}\,
+ & \big(\textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}\,
       \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha := \id_{\_}}]\big)
@@
  &  \circ\;
-  \big(\textit{Lan}_{\ol{\oCPOsem{\emptyset; \ol\alpha \vdash K}[\ol{\alpha := \_}]}}\,
+  \big(\textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}\,
       \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := \_}]\big)
     \ol{\oCPOsem{\Gamma; \Phi \vdash A}f} \\
-= & \big(\textit{Lan}_{\ol{\oCPOsem{\emptyset; \ol\alpha \vdash K}[\ol{\alpha := \_}]}}\,
+= & \big(\textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}\,
       \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := \_}]\big)
@@
  &  \circ\;
-  \big(\textit{Lan}_{\ol{\oCPOsem{\emptyset; \ol\alpha \vdash K}[\ol{\alpha := \_}]}}\,
+  \big(\textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}\,
       \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha := \id_{\_}}]\big)
@@
 where the above equality holds by naturality of
-$\textit{Lan}_{\ol{\oCPOsem{\emptyset; \ol\alpha \vdash
-        K}[\ol{\alpha := \_}]}}\,\oCPOsem{\Gamma; \Phi, \ol\alpha
+$\textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash
+        K}\rho[\ol{\alpha := \_}]}}\,\oCPOsem{\Gamma; \Phi, \ol\alpha
     \vdash F}f[\ol{\alpha := \id_{\_}}]$.
@@
 A {\em $k$-ary $\oCPO$ relation transformer} $F$ is a triple $(F^1,
-F^2,F^*)$, where $F^1,F^2 : [\oCPO^k,\oCPO]_{\omega_1}$ and $F^* :
-[\oCPOR^k, \oCPOR]_{\omega_1}$ are such that if
+F^2,F^*)$, where $F^1,F^2 : [\oCPO^k,\oCPO]_{\omega}$ and $F^* :
+[\oCPOR^k, \oCPOR]_{\omega}$ are such that if
 $R_1:\oCPOR(A_1,B_1),...,R_k:\oCPOR(A_k,B_k)$ then $F^* \ol{R} :
@@
 \begin{itemize}
-\item $H^1$ and $H^2$ are functors from $[\oCPO^k,\oCPO]_{\omega_1}$ to $[\oCPO^k,\oCPO]_{\omega_1}$
-\item $H^*$ is a functor from $\oCPORT_k$ to $[\oCPOR^k,\oCPOR]_{\omega_1}$
+\item $H^1$ and $H^2$ are functors from $[\oCPO^k,\oCPO]_{\omega}$ to $[\oCPO^k,\oCPO]_{\omega}$
+\item $H^*$ is a functor from $\oCPORT_k$ to $[\oCPOR^k,\oCPOR]_{\omega}$
 \item for all $\overline{R : \oCPOR(A,B)}$,
@@
 K^2$ are natural transformations between endofunctors on
-$[\oCPO^k,\oCPO]_{\omega_1}$ and the component of $\sigma$ at $F
+$[\oCPO^k,\oCPO]_{\omega}$ and the component of $\sigma$ at $F
 %= (F^1,F^2,F^*)
@@
 Definition~\ref{def:RT-nat-trans} entails that $\sigma^i_{F^i}$ must
-be natural in $F^i : [\oCPO^k,\oCPO]_{\omega_1}$, and, for every $F$,
+be natural in $F^i : [\oCPO^k,\oCPO]_{\omega}$, and, for every $F$,
 $(\sigma^1_{F^1})_{\overline{A}}$ and
@@
 
-Critically, we can compute $\omega_1$-directed colimits in
+Critically, we can compute $\omega$-directed colimits in
 $\oCPORT_k$: it is not hard to see that if $\cal D$ is an
-$\omega_1$-directed set then $\colim{d \in {\cal D}}{(F^1_d,
+$\omega$-directed set then $\colim{d \in {\cal D}}{(F^1_d,
   F^2_d,F^*_d)} = (\colim{d \in {\cal D}}{F^1_d}, \colim{d \in {\cal
@@
 endofunctor $T = (T^1,T^2,T^*)$ on $\oCPORT_k$ to be {\em
-  $\omega_1$-cocontinuous} if $T^1$ and $T^2$ are
-$\omega_1$-cocontinuous endofunctors on $[\oCPO^k,\oCPO]_{\omega_1}$
-and $T^*$ is an $\omega_1$-cocontinuous functor from $\oCPORT_k$ to
-$[\oCPOR^k,\oCPOR]_{\omega_1}$, i.e., is in
-$[\oCPORT_k,[\oCPOR^k,\oCPOR]_{\omega_1}]_{\omega_1}$.
+  $\omega$-cocontinuous} if $T^1$ and $T^2$ are
+$\omega$-cocontinuous endofunctors on $[\oCPO^k,\oCPO]_{\omega}$
+and $T^*$ is an $\omega$-cocontinuous functor from $\oCPORT_k$ to
+$[\oCPOR^k,\oCPOR]_{\omega}$, i.e., is in
+$[\oCPORT_k,[\oCPOR^k,\oCPOR]_{\omega}]_{\omega}$.
 
@@
 Observing that, for every $k$, $K^\oCPO_0$ is initial in
-$[\oCPO^k,\oCPO]_{\omega_1}$, and $K^\oCPOR_0$ is initial in
-$[\oCPOR^k,\oCPOR]_{\omega_1}$, we have that, for each $k$, $K_0 =
+$[\oCPO^k,\oCPO]_{\omega}$, and $K^\oCPOR_0$ is initial in
+$[\oCPOR^k,\oCPOR]_{\omega}$, we have that, for each $k$, $K_0 =
 (K^\oCPO_0,K^\oCPO_0,K^\oCPOR_0)$ is initial in $\oCPORT_k$. Thus, if
@@
 $\oCPORT_k$ then we can define the $\oCPO$ relation transformer $\mu
-T$ to be $\colim{i < \omega_1}{T^i K_0}$. It is not hard to see that
+T$ to be $\colim{i < \omega}{T^i K_0}$. It is not hard to see that
 $\mu T$ is given explicitly as
@@
   \mu T = (\mu T^1,\mu T^2,
-\colim{i < \omega_1}{(T^iK_0)^*})
+\colim{i < \omega}{(T^iK_0)^*})
 \end{equation}
 and that, as our notation suggests, it really is a fixpoint for $T$ if
-$T$ is $\omega_1$-cocontinuous:
+$T$ is $\omega$-cocontinuous:
 \begin{lemma}\label{lem:fp}
-For any $T : [\oCPORT_k,\oCPORT_k]_{\omega_1}$, $\mu T \cong T(\mu T)$.
+For any $T : [\oCPORT_k,\oCPORT_k]_{\omega}$, $\mu T \cong T(\mu T)$.
 \end{lemma}
@@
 T \to T(\mu T)$ in $\oCPORT_k$. The latter is always a morphism in $\oCPORT_k$,
-but the former need not be if $T$ is not $\omega_1$-cocontinuous.
+but the former need not be if $T$ is not $\omega$-cocontinuous.
 
 It is worth noting that the third component in Equation~(\ref{eq:mu})
-is the colimit in $[\oCPOR^k,\oCPOR]_{\omega_1}$ of third components
+is the colimit in $[\oCPOR^k,\oCPOR]_{\omega}$ of third components
 of $\oCPO$ relation transformers, rather than a fixpoint of an
-endofunctor on $[\oCPO^k,\oCPO]_{\omega_1}$. That there is an
+endofunctor on $[\oCPO^k,\oCPO]_{\omega}$. That there is an
 asymmetry between the first two components of $\mu T$ and its third
@@
 of an endofunctor on $\oCPORT_k$ need not be a functor on all of
-$[\oCPOR^k,\oCPOR]_{\omega_1}$. In particular, although we can define
+$[\oCPOR^k,\oCPOR]_{\omega}$. In particular, although we can define
 $T_{H,\rho}\, F$ for an $\oCPO$ relation transformer $F$ in
 Definition~\ref{def:rel-sem} below, it is not clear how we could
-define it for an arbitrary $F : [\oCPOR^k,\oCPOR]_{\omega_1}$.
+define it for an arbitrary $F : [\oCPOR^k,\oCPOR]_{\omega}$.
 
@@
 \{\phi_1^{k_1},...,\phi_n^{k_n}\}$ and $\ol K = \{K_1,...,K_n\}$ where
-$K_i : [\oCPOR^{k_i},\oCPOR]_{\omega_1}$ for $i = 1,...,n$, then we
+$K_i : [\oCPOR^{k_i},\oCPOR]_{\omega}$ for $i = 1,...,n$, then we
 write either $\rho[\ol{\Phi := K}]$ or $\rho[\ol{\phi := K}]$ for the
@@
 
-We define, for each $k$, the notion of an $\omega_1$-cocontinuous
+We define, for each $k$, the notion of an $\omega$-cocontinuous
 functor from $\oCPORenv$ to $\oCPORT_k$:
 \begin{dfn}\label{def:relenv-functor}
-A functor $H : [\oCPORenv, \oCPORT_k]_{\omega_1}$ is a triple $H =
+A functor $H : [\oCPORenv, \oCPORT_k]_{\omega}$ is a triple $H =
 (H^1,H^2,H^*)$, where
@@
 \item $H^1$ and $H^2$ are objects in
-  $[\oCPOenv,[\oCPO^k,\oCPO]_{\omega_1}]_{\omega_1}$
+  $[\oCPOenv,[\oCPO^k,\oCPO]_{\omega}]_{\omega}$
 \item $H^*$ is a an object in
-  $[\oCPORenv,[\oCPOR^k,\oCPOR]_{\omega_1}]_{\omega_1}$
+  $[\oCPORenv,[\oCPOR^k,\oCPOR]_{\omega}]_{\omega}$
 \item for all $\overline{R : \oCPOR(A,B)}$ and morphisms $f$ in
@@
 Considering $\oCPORenv$ as a product $\Pi_{\phi^k \in \tvars \cup
-  \fvars} \oCPORT_k$, we extend the computation of $\omega_1$-directed
+  \fvars} \oCPORT_k$, we extend the computation of $\omega$-directed
 colimits in $\oCPORT_k$ to compute colimits in $\oCPORenv$ componentwise. We
-similarly extend the notion of an $\omega_1$-cocontinuous endofunctor on
-$\oCPORT_k$ componentwise to give a notion of $\omega_1$-cocontinuity for
+similarly extend the notion of an $\omega$-cocontinuous endofunctor on
+$\oCPORT_k$ componentwise to give a notion of $\omega$-cocontinuity for
 functors from $\oCPORenv$ to $\oCPORT_k$.  Recalling from the start of this
@@
 The {\em relational interpretation}
-$\oCPORsem{\cdot} : \F \to [\oCPORenv, \oCPOR]_{\omega_1}$ is defined by
+$\oCPORsem{\cdot} : \F \to [\oCPORenv, \oCPOR]_{\omega}$ is defined by
 \begin{align*}
@@
     F)\ol{A}}\pi_2 \rho\\
-  &\hspace{0.3in}|~{\color{blue}
-    ((\eta_1)_{\ol{\oCPOsem{\Gamma; \Phi \vdash A}\pi_1 \rho}} t_1,
-(\eta_2)_{\ol{\oCPOsem{\Gamma; \Phi \vdash A}\pi_2 \rho}} t_2) \in}\\
-  &\hspace{0.3in}{\color{blue}
-(\textit{Lan}_{\ol{\oCPORsem{\emptyset;
-          \ol\alpha \vdash K}[\ol{\alpha := \Eq_{\_}}]}}
-    \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha :=
-        \Eq_{\_}}])
-    \ol{\oCPORsem{\Gamma; \Phi \vdash A}\rho}} \}  
-  %\forall \sigma : \setenv, (
-  %(\eta^\sigma_1)_{\ol{\oCPOsem{\Gamma;\Phi \vdash A}\sigma }} t_1,
-  %(\eta^\sigma_2)_{\ol{\oCPOsem{\Gamma;\Phi \vdash A}\sigma }} t_2) \in\\
-  %&\hspace{0.5in} (\textit{Lan}_{\ol{\oCPORsem{\Gamma; \ol\alpha
-  %      \vdash K}\Eq_\sigma[\ol{\alpha := \Eq_{\_}}]}}
-  %\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\Eq_\sigma[\ol{\alpha :=
-  %    \Eq_{\_}}]) \ol{\oCPORsem{\Gamma; \Phi \vdash A}\Eq_\sigma} \}\\
+  &\hspace{0.2in}|~\exists\, \ol{Z : \oCPO_0}, t_1' : \oCPOsem{\Gamma;
+    \ol{\alpha} \vdash K}\pi_1\rho[\ol{\alpha := Z}],\, t_2' :
+  \oCPOsem{\Gamma; \ol{\alpha} \vdash K}\pi_2\rho[\ol{\alpha := Z}],\\
+  & \hspace{0.2in} f :
+  \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho[\ol{\alpha := \Eq_Z}] \to
+  \oCPORsem{\Gamma; \Phi \vdash A}\rho\\
+  &\hspace{0.2in}\mbox{such that } \iota_{Z,\pi_1 f}t_1' = t_1,\,
+  \kappa_{Z,\pi_2 f}t_2' = t_2,\\
+  &  \hspace*{0.2in}\mbox{and } (t_1',t_2') :
+  \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := \Eq_Z}]\,\}
 \end{align*}
 \end{dfn}
-{\color{blue} where $\eta_i$ is for the (mixed $\set/\rel$) diagram
-  top p2 of Daniel's notes for relational environment $\Eq_{\pi_i
-    \rho}[\alpha := \Eq_{\_}]$.}
-
-%{\color{blue} In the final clause of Definition~\ref{def:rel-sem},
-%  $\eta^\sigma_i : \textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash
-%        K}\sigma[\ol{\alpha := \_}]}} \oCPOsem{\Gamma; \Phi, \ol\alpha
-%    \vdash F}\sigma[\ol{\alpha := \_}] \; \to \;
-%  \pi_i\,(\textit{Lan}_{\ol{\oCPORsem{\Gamma; \ol\alpha \vdash
-%        K}\Eq_\sigma[\ol{\alpha := \Eq_{\_}}]}} \oCPORsem{\Gamma;
-%    \Phi, \ol\alpha \vdash F}\Eq_\sigma[\ol{\alpha := \Eq_{\_}}])$ for
-%  each $\sigma : \setenv$ and $i = 1,2$.}
+\noindent
+{\color{blue} In the final clause of Definition~\ref{def:rel-sem},
+  $\iota_{Z,h}$ is the injection into
+\[\colim{Z \,: \,\oCPO_0,\; h \,:\, \oCPOsem{\Gamma; \ol{\alpha} \vdash
+      K}\pi_1\rho[\ol{\alpha := Z}] \to \oCPOsem{\Gamma; \Phi \vdash
+      A}\pi_1\rho}{\oCPOsem{\Gamma; \ol{\alpha} \vdash
+    F}\pi_1\rho[\ol{\alpha := Z}]}\]
+  and 
+  $\kappa_{Z,h}$ is the injection into
+\[\colim{Z\, :\, \oCPO_0, \;h \,:\, \oCPOsem{\Gamma; \ol{\alpha} \vdash
+      K}\pi_2\rho[\ol{\alpha := Z}] \to \oCPOsem{\Gamma; \Phi \vdash
+    A}\pi_2\rho}{\oCPOsem{\Gamma; \ol{\alpha} \vdash
+    F}\pi_2\rho[\ol{\alpha := Z}]}\]}
 
@@
 The proof that relational interpretations of $\Nat$-types define sets
-is analogous to the proof of Lemma~\ref{lem:nat-types-sets}.  Next, we
-observe that $\oCPORsem{\Gamma; \emptyset \vdash \Nat^\Phi F\,G} \rho$
-is in $\oCPOR$. It is indeed a relation between $\oCPO$s, and it
-relates sups of pairwise related $\omega$-chains of natural
-transformations because their sups are computed pointwise. More
-specifically, if $(t_i, t_i') \colon \oCPORsem{\Gamma; \emptyset
-  \vdash \Nat^\Phi F\,G} \rho$ for all $i < \omega$, then for every
-$\ol{K = (K^1, K^2, K^*) : \oCPORT_{k}}$ and every $i < \omega$ we
-have
-  \[\big((t_i)_{\ol{K^1}},\, (t'_i)_{\ol{K^2}}\big) \in
-  (\oCPORsem{\Gamma; \Phi \vdash G}\rho[\ol{\Phi :=
-      K}])^{\oCPORsem{\Gamma;\Phi\vdash F}\rho[\ol{\Phi := K}]}\] So
-  if $(a,b) \in \oCPORsem{\Gamma;\Phi\vdash F}\rho[\ol{\Phi := K}]$,
-  then $\big((t_i)_{\ol{K^1}}\,a,\, (t'_i)_{\ol{K^2}}\,b\big) \in
-  \oCPORsem{\Gamma;\Phi\vdash G}\rho[\ol{\Phi := K}]$ for all $i <
-  \omega$. But then since $\big((t_i)_{\ol{K^1}} a\big)_{i < \omega}$
-  is an $\omega$-chain in $\oCPOsem{\Gamma; \Phi \vdash
-    G}\rho[\ol{\Phi := K^1}]$ and $\big((t'_i)_{\ol{K^2}}\big)_{i <
-    \omega} b$ is an $\omega$-chain in $\oCPOsem{\Gamma; \Phi \vdash
-    G}\rho[\ol{\Phi := K^2}]$, not only are $\bigvee_{i < \omega}
-  ((t_i)_{\ol{K^1}}\,a)$ and $\bigvee_{i < \omega}
-  ((t'_i)_{\ol{K^2}}\,b)$ well-defined, but, since
-\[\oCPORsem{\Gamma;\Phi\vdash G}\rho[\ol{\Phi := K}] : \oCPOR(
-  \oCPOsem{\Gamma;\Phi\vdash G}\rho[\ol{\Phi := K^1}],
-  \oCPOsem{\Gamma;\Phi\vdash G}\rho[\ol{\Phi := K^2}])\]
-we have
-  \[
-  \Big(\big(\bigvee_{i < \omega} t_i\big)_{\ol{K^1}} \,a,
-  \big(\bigvee_{i < \omega} t'_i\big)_{\ol{K^2}} \,b\Big)
-  = \Big(\bigvee_{i < \omega} \big((t_i)_{\ol{K^1}}\,a\big),\,
-  \bigvee_{i < \omega} \big((t'_i)_{\ol{K^2}}\,b\big)\Big) \in
-          \oCPORsem{\Gamma;\Phi\vdash G}\rho[\ol{\Phi
-             := K}]\] as well. That is,
-\[
-  \Big(\big(\bigvee_{i < \omega} t_i\big)_{\ol{K^1}}, \big(\bigvee_{i
-    < \omega} t'_i\big)_{\ol{K^2}} \Big)
-  \in (\oCPORsem{\Gamma;
-    \Phi \vdash G}\rho[\ol{\Phi := K}])^{\oCPORsem{\Gamma;\Phi\vdash
-      F}\rho[\ol{\Phi := K}]}
-  \]
-i.e.,
-$( \bigvee_{i < \omega} t_i, \bigvee_{i < \omega} t'_i ) \in
-  \oCPORsem{\Gamma; \emptyset \vdash \Nat^\Phi F\,G} \rho$.
-
-Moreover, as for $\oCPO$ interpretations, $\omega_1$-cocontinuity of
+is analogous to the proof of Lemma~\ref{lem:nat-types-sets}.
+
+%Next, we
+%observe that $\oCPORsem{\Gamma; \emptyset \vdash \Nat^\Phi F\,G} \rho$
+%is in $\oCPOR$. It is indeed a relation between $\oCPO$s, and it
+%relates sups of pairwise related $\omega$-chains of natural
+%transformations because their sups are computed pointwise. More
+%specifically, if $(t_i, t_i') \colon \oCPORsem{\Gamma; \emptyset
+%  \vdash \Nat^\Phi F\,G} \rho$ for all $i < \omega$, then for every
+%$\ol{K = (K^1, K^2, K^*) : \oCPORT_{k}}$ and every $i < \omega$ we
+%have
+%  \[\big((t_i)_{\ol{K^1}},\, (t'_i)_{\ol{K^2}}\big) \in
+%  (\oCPORsem{\Gamma; \Phi \vdash G}\rho[\ol{\Phi :=
+%      K}])^{\oCPORsem{\Gamma;\Phi\vdash F}\rho[\ol{\Phi := K}]}\] So
+%  if $(a,b) \in \oCPORsem{\Gamma;\Phi\vdash F}\rho[\ol{\Phi := K}]$,
+%  then $\big((t_i)_{\ol{K^1}}\,a,\, (t'_i)_{\ol{K^2}}\,b\big) \in
+%  \oCPORsem{\Gamma;\Phi\vdash G}\rho[\ol{\Phi := K}]$ for all $i <
+%  \omega$. But then since $\big((t_i)_{\ol{K^1}} a\big)_{i < \omega}$
+%  is an $\omega$-chain in $\oCPOsem{\Gamma; \Phi \vdash
+%    G}\rho[\ol{\Phi := K^1}]$ and $\big((t'_i)_{\ol{K^2}}\big)_{i <
+%    \omega} b$ is an $\omega$-chain in $\oCPOsem{\Gamma; \Phi \vdash
+%    G}\rho[\ol{\Phi := K^2}]$, not only are $\bigvee_{i < \omega}
+%  ((t_i)_{\ol{K^1}}\,a)$ and $\bigvee_{i < \omega}
+%  ((t'_i)_{\ol{K^2}}\,b)$ well-defined, but, since
+%\[\oCPORsem{\Gamma;\Phi\vdash G}\rho[\ol{\Phi := K}] : \oCPOR(
+%  \oCPOsem{\Gamma;\Phi\vdash G}\rho[\ol{\Phi := K^1}],
+%  \oCPOsem{\Gamma;\Phi\vdash G}\rho[\ol{\Phi := K^2}])\]
+%we have
+%  \[
+%  \Big(\big(\bigvee_{i < \omega} t_i\big)_{\ol{K^1}} \,a,
+%  \big(\bigvee_{i < \omega} t'_i\big)_{\ol{K^2}} \,b\Big)
+%  = \Big(\bigvee_{i < \omega} \big((t_i)_{\ol{K^1}}\,a\big),\,
+%  \bigvee_{i < \omega} \big((t'_i)_{\ol{K^2}}\,b\big)\Big) \in
+%          \oCPORsem{\Gamma;\Phi\vdash G}\rho[\ol{\Phi
+%             := K}]\] as well. That is,
+%\[
+%  \Big(\big(\bigvee_{i < \omega} t_i\big)_{\ol{K^1}}, \big(\bigvee_{i
+%    < \omega} t'_i\big)_{\ol{K^2}} \Big)
+%  \in (\oCPORsem{\Gamma;
+%    \Phi \vdash G}\rho[\ol{\Phi := K}])^{\oCPORsem{\Gamma;\Phi\vdash
+%      F}\rho[\ol{\Phi := K}]}
+%  \]
+%i.e.,
+%$( \bigvee_{i < \omega} t_i, \bigvee_{i < \omega} t'_i ) \in
+%  \oCPORsem{\Gamma; \emptyset \vdash \Nat^\Phi F\,G} \rho$.
+
+Moreover,
+%as for $\oCPO$ interpretations,
+$\omega$-cocontinuity of
 each of the above interpretations follows from Corollary~12
@@
 $\Lan$s}. For $\Nat$-types, $\oCPORsem{\Gamma; \emptyset \vdash
-  \Nat^\Phi F\,G}$ is an $\omega_1$-cocontinuous functor because it is
-constant on $\omega_1$-directed sets.  Interpretations of $\Nat$-types
+  \Nat^\Phi F\,G}$ is an $\omega$-cocontinuous functor because it is
+constant on $\omega$-directed sets.  Interpretations of $\Nat$-types
 ensure that $\oCPORsem{\Gamma \vdash F \to G}$ and $\oCPORsem{\Gamma
@@
 For the next-to-last clause in Definition~\ref{def:rel-sem} to be
-well-defined we need $T_{H,\rho}$ to be an $\omega_1$-cocontinuous
+well-defined we need $T_{H,\rho}$ to be an $\omega$-cocontinuous
 endofunctor on $\oCPORT$ so that, by Lemma~\ref{lem:fp}, it admits a
@@
 that relational interpretations of types must be
-$\omega_1$-cocontinuous functors from $\oCPORenv$ to $\oCPORT_0$,
+$\omega$-cocontinuous functors from $\oCPORenv$ to $\oCPORT_0$,
 which in turn entails that the actions of relational interpretations
@@
 %  K}\rho[\ol{\alpha := \_}]$ and $\oCPORsem{\Gamma; \Phi, \ol\alpha
-%  \vdash F}\rho[\ol{\alpha := \_}]$ to be $\omega_1$-cocontinuous
+%  \vdash F}\rho[\ol{\alpha := \_}]$ to be $\omega$-cocontinuous
 %functors.
@@
 every $\Gamma; \ol\alpha \vdash F$, $\oCPORsem{\Gamma; \ol\alpha
-  \vdash F}$ is actually in $[\oCPOR^k,\oCPOR]_{\omega_1}$, where $k =
+  \vdash F}$ is actually in $[\oCPOR^k,\oCPOR]_{\omega}$, where $k =
 |\ol\alpha|$. We first define the actions of each of these functors
@@
 $T_H$ to a {\em functor} from $\oCPORenv$ to
-$[[\oCPOR^k,\oCPOR]_{\omega_1},
-  [\oCPOR^k,\oCPOR]_{\omega_1}]_{\omega_1}$.  Its action on an object
+$[[\oCPOR^k,\oCPOR]_{\omega},
+  [\oCPOR^k,\oCPOR]_{\omega}]_{\omega}$.  Its action on an object
 $\rho \in \oCPORenv$ is given by the higher-order functor $T_{H,\rho}$
@@
 T_{H,\rho} \to T_{H,\rho'}$ whose action on any $F :
-[\oCPOR^k,\oCPOR]_{\omega_1}$ is the higher-order natural
+[\oCPOR^k,\oCPOR]_{\omega}$ is the higher-order natural
 transformation $T_{H,f}\, F : T_{H,\rho}\, F \to T_{H,\rho'}\, F$
@@
   F},\oCPORsem{\Gamma;\Phi \vdash F}) \in
-[\oCPORenv,\oCPORT_0]_{\omega_1}$.
+[\oCPORenv,\oCPORT_0]_{\omega}$.
 \end{lemma}
@@
 using an appropriate result from~\cite{jp19} to deduce
-$\omega_1$-cocontinuity of $\sem{\Gamma;\Phi \vdash F}$ in each case,
+$\omega$-cocontinuity of $\sem{\Gamma;\Phi \vdash F}$ in each case,
 together with Lemma~\ref{lem:fp} and Equation~\ref{eq:mu} for
@@
 $\oCPO$ a $k$-ary $\oCPO$ relation transformer as follows. Recall
-that, since $\oCPO$ is a locally $\omega_1$-presentable category,
+that, since $\oCPO$ is a locally $\omega$-presentable category,
 Proposition 1.6.1 of~\cite{ar94} ensures that it has a (strong epi,
@@
 \begin{lemma}\label{lem:graph-reln-functors}
-If $F,G : [\oCPO^k,\oCPO]_{\omega_1}$, and if $\alpha : F \to G$ is a natural
+If $F,G : [\oCPO^k,\oCPO]_{\omega}$, and if $\alpha : F \to G$ is a natural
 transformation, then $\graph{\alpha}$ is in $\oCPORT_k$.
@@
 \begin{proof}
-Clearly, $\graph{\alpha}^*$ is $\omega_1$-cocontinuous, so
-$\graph{\alpha}^* : [\oCPOR^k,\oCPOR]_{\omega_1}$. Now, suppose
+Clearly, $\graph{\alpha}^*$ is $\omega$-cocontinuous, so
+$\graph{\alpha}^* : [\oCPOR^k,\oCPOR]_{\omega}$. Now, suppose
 $\overline{R : \oCPOR(A, B)}$, $\overline{S : \oCPOR(C, D)}$, and
@@
 \begin{lemma}\label{lem:eq-reln-equalities}
-If $\alpha : F \to G$ is a morphism in $[\oCPO^k, \oCPO]_{\omega_1}$
+If $\alpha : F \to G$ is a morphism in $[\oCPO^k, \oCPO]_{\omega}$
 and $f_1: A_1 \to B_1, ..., f_k : A_k \to B_k$, then $\graph{\alpha}^*
@@
 The {\em equality $\oCPO$ relation transformer} on $F :
-[\oCPO^k,\oCPO]_{\omega_1}$ is defined to be $\Eq_F =
+[\oCPO^k,\oCPO]_{\omega}$ is defined to be $\Eq_F =
 \graph{\id_{F}}$. Specifically, $\Eq_F = (F, F, \Eq_F^*)$ with
@@
 \end{thm}
-The proof is by induction on the structure of $F$. Only the $\Nat$,
-application, fixpoint, and $\Lan$ cases are non-routine. The latter
-two use Equation~\ref{eq:eq-pres-eq}. The fixpoint case also uses the
-observation that, for every $i < \omega_1$, the following intermediate
+\begin{proof}
+By induction on the structure of $F$. Only the $\Nat$, application,
+fixpoint, and $\Lan$ cases are non-routine.
+
+The latter two use
+Equation~\ref{eq:eq-pres-eq}. The fixpoint case also uses the
+observation that, for every $i < \omega$, the following intermediate
 results can be proved by simultaneous induction with
@@
 \end{array}\]
-hold. With these results in hand, the proof follows easily. It is
-given in detail in the accompanying anonymous supplementary
-material. As noted there, if functorial variables of arity greater
-than $0$ were allowed to appear in the bodies of $\mu$-types, then the
-IEL would fail.
+hold. With these results in hand, the proof follows easily
+{\color{red} for all but $\Lan$-types}. The entire proof is given in
+detail in the accompanying anonymous supplementary material. As noted
+there, if functorial variables of arity greater than $0$ were allowed
+to appear in the bodies of $\mu$-types, then the IEL would fail.
+
+{\color{red} For $\Lan$-types,....}
+
+\end{proof}
 
@@
   $\omega$-cocontinuous. It also doesn't preserve
-  $\omega_1$-presentable objects, i.e., countable $\omega$CPOs since
+  $\omega$-presentable objects, i.e., countable $\omega$CPOs since
   $\Nat \to \Nat$ is not countable. So we cannot have a functor like
   $N$ as the subscript to $\Lan$ and expect the resulting $\Lan$ to be
-  $\omega_1$-cocontinuous.
+  $\omega$-cocontinuous.
 \item What functors can be subscripts to $\Lan$ and produce
-  $\omega_1$-cocontinuous functors? We can use functors that preserve
+  $\omega$-cocontinuous functors? We can use functors that preserve
   presentable objects by theorem in~\cite{jp19}, and possibly others

### 2020-09-23 20-gadts/inf-rules.tex
added functorial actions of set and rel interps of types

@@
 
-{\color{blue} Need to check that the extra condition in the interp of
-  Lans is preserved for the last clause to be well-defined.}
+We need to check that the functorial action on $\Lan$ types is well-defined.
+That amounts to showing that the extra condition in the interpretation of
+$\Lan$ types is preserved by the functorial action, so that,
+if $t : \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \rho$,
+then $\oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f t : \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \rho'$.
+By definition, if $t : \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \rho$,
+then $(t, t) \in \oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \Eq_{\rho}$,
+i.e., there exist
+$\ol{Z : \oCPO_0}$,
+$t_1 : \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := Z}]$,
+$t_2 : \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := Z}]$,
+and $\ol{g : \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\Eq_{\rho}[\ol{\alpha := \Eq_Z}] \to \oCPORsem{\Gamma; \Phi \vdash A}\Eq_{\rho}}$
+such that
+$\iota_{\ol{Z}, \ol{\pi_1 g}}t_1 = t$,
+$\kappa_{\ol{Z}, \ol{\pi_2 g}}t_2 = t$,
+and
+$(t_1,t_2) : \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\Eq_{\rho}[\ol{\alpha := \Eq_Z}]$.
+To prove that
+$\oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f t
+: \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \rho'$,
+let $\ol{Z' = Z}$, $t'_i = \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F} f [\ol{\alpha := \id_{Z}}] t_i$ for $i = 1, 2$,
+and $\ol{g' = \oCPORsem{\Gamma; \Phi \vdash A}f \circ g}$.
+We thus have to show that
+$\iota'_{\ol{Z'}, \ol{\pi_1 g'}}t'_1 = \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f t$,
+$\kappa'_{\ol{Z'}, \ol{\pi_2 g'}}t'_2 = \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f t$,
+and
+$(t_1',t_2') : \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\Eq_{\rho'}[\ol{\alpha := \Eq_{Z'}}]$.
+That $(t_1',t_2') : \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\Eq_{\rho'}[\ol{\alpha := \Eq_{Z'}}]$
+follows from the fact that
+$(t_1',t_2') = \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F} Eq_{f} [\ol{\alpha := \id_{\Eq_{Z}}}] (t_1, t_2)$
+and $(t_1,t_2) : \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\Eq_{\rho}[\ol{\alpha := \Eq_Z}]$.
+That $\iota'_{\ol{Z'}, \ol{\pi_1 g'}}t'_1 = \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f t$
+is because
+\[
+\begin{array}{rl}
+\iota'_{\ol{Z'}, \ol{\pi_1 g'}}t'_1
+&= \iota'_{\ol{Z'}, \ol{\pi_1 g'}} (\oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F} f [\ol{\alpha := \id_{Z}}] t_1) \\
+&= \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f (\iota_{\ol{Z}, \ol{\pi_1 g}} t_1) \\
+&= \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f t
+\end{array}
+\]
+where the second equality is justified by the commutativity of the diagram
+\begin{equation}\label{cd:lan-comm}
+\begin{tikzcd}[row sep = huge]
+\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := Z}]
+  \ar[r, "{\iota_{\ol{Z}, \ol{\pi_1 g}}}"]
+  \ar[d, "{\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F} f [\ol{\alpha := \id_Z}]}" description]
+&\colim{\ol{S}, \ol{h : \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho[\ol{\alpha := S}] \to \oCPORsem{\Gamma; \Phi \vdash A}\rho}}
+{\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := S}]}
+  \ar[d, "{(\Lan_{\ol{\oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho[\ol{\alpha := \_}]}} \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha := \id_{\_}}]) \ol{\oCPORsem{\Gamma; \Phi \vdash A}\rho}}" description] \\
+\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := Z}]
+  \ar[r, "{\iota''_{\ol{Z}, \ol{\pi_1 g}}}"]
+  \ar[d, equal]
+& \colim{\ol{S}, \ol{h : \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho'[\ol{\alpha := S}] \to \oCPORsem{\Gamma; \Phi \vdash A}\rho'}}
+{\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := S}]}
+  \ar[d, "{(\Lan_{\ol{\oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho[\ol{\alpha := \_}]}} \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := \_}]) \ol{\oCPORsem{\Gamma; \Phi \vdash A}f}}" description] \\
+\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := Z}]
+  \ar[r, shift right = 1, "{\iota'_{\ol{Z'}, \ol{\pi_1 (\oCPORsem{\Gamma; \Phi \vdash A}f \circ g)}}}"']
+& \colim{\ol{S}, \ol{h : \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho'[\ol{\alpha := S}] \to \oCPORsem{\Gamma; \Phi \vdash A}\rho'}}
+{\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := S}]}
+\end{tikzcd}
+\end{equation}
+%\[
+%\begin{tikzcd}
+%\colim{\ol{S}, \ol{h : \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho[\ol{\alpha := S}] \to \oCPORsem{\Gamma; \Phi \vdash A}\rho}}
+%{\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := S}]}
+%  \ar[r, bend left = 10, "{(\Lan_{\ol{\oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho[\ol{\alpha := \_}]}} \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha := \id_{\_}}]) \ol{\oCPORsem{\Gamma; \Phi \vdash A}\rho}}"]
+%& \colim{\ol{S}, \ol{h : \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho'[\ol{\alpha := S}] \to \oCPORsem{\Gamma; \Phi \vdash A}\rho'}}
+%{\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := S}]} \\
+%\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := Z}]
+%  \ar[u, "{\iota_{\ol{Z}, \ol{\pi_1 g}}}"]
+%  \ar[r, "{\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F} f [\ol{\alpha := \id_Z}]}"]
+%& \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := Z}]
+%  \ar[u, "{\iota''_{\ol{Z}, \ol{\pi_1 g}}}"]
+%\end{tikzcd}
+%\]
+%and
+%\[
+%\begin{tikzcd}
+%\colim{\ol{S}, \ol{h : \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho[\ol{\alpha := S}] \to \oCPORsem{\Gamma; \Phi \vdash A}\rho}}
+%{\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := S}]}
+%  \ar[r, bend left = 10, "{(\Lan_{\ol{\oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho[\ol{\alpha := \_}]}} \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := \_}]) \ol{\oCPORsem{\Gamma; \Phi \vdash A}f}}"]
+%& \colim{\ol{S}, \ol{h : \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho'[\ol{\alpha := S}] \to \oCPORsem{\Gamma; \Phi \vdash A}\rho'}}
+%{\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := S}]} \\
+%\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := Z}]
+%  \ar[u, "{\iota''_{\ol{Z}, \ol{\pi_1 g}}}"]
+%  \ar[r, equal]
+%& \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := Z}]
+%  \ar[u, "{\iota'_{\ol{Z'}, \ol{\pi_1 (\oCPORsem{\Gamma; \Phi \vdash A}f \circ g)}}}"]
+%\end{tikzcd}
+%\]
+Analogously, we show that $\kappa_{Z,\pi_2 g'}t'_2 = \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f t$.
 
@@
     F)\ol{A}}\pi_2 \rho\\
-  &\hspace{0.2in}|~\exists\, \ol{Z : \oCPO_0}, t_1' : \oCPOsem{\Gamma;
-    \ol{\alpha} \vdash K}\pi_1\rho[\ol{\alpha := Z}],\, t_2' :
-  \oCPOsem{\Gamma; \ol{\alpha} \vdash K}\pi_2\rho[\ol{\alpha := Z}],\\
-  & \hspace{0.2in} f :
+  &\hspace{0.2in}|~\exists\, \ol{Z : \oCPO_0}, t_1' : \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\pi_1\rho[\ol{\alpha := Z}],\, t_2' :
+  \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\pi_2\rho[\ol{\alpha := Z}],\\
+  & \hspace{0.2in} \ol{f :
   \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho[\ol{\alpha := \Eq_Z}] \to
-  \oCPORsem{\Gamma; \Phi \vdash A}\rho\\
-  &\hspace{0.2in}\mbox{such that } \iota_{Z,\pi_1 f}t_1' = t_1,\,
-  \kappa_{Z,\pi_2 f}t_2' = t_2,\\
+  \oCPORsem{\Gamma; \Phi \vdash A}\rho}\\
+  &\hspace{0.2in}\mbox{such that } \iota_{\ol{Z},\ol{\pi_1 f}}t_1' = t_1,\,
+  \kappa_{\ol{Z}, \ol{\pi_2 f}}t_2' = t_2,\\
   &  \hspace*{0.2in}\mbox{and } (t_1',t_2') :
@@
 $\rho'$ (so that $\rho|_\tvars = \rho'|_\tvars$). The action
-$\oCPORsem{\Gamma;\Phi \vdash F}f$ of $\oCPORsem{\Gamma;\Phi \vdash
-  F}$ on the morphism $f$ is given exactly as in
+$\oCPORsem{\Gamma;\Phi \vdash F}f$ of $\oCPORsem{\Gamma;\Phi \vdash F}$
+on the morphism $f$ for all types but $\Lan$ is given exactly as in
 Definition~\ref{def:set-sem-funcs}, except that all interpretations
@@
 $T^\oCPO_{H,f}$ are replaced by $T_{H,f}$.
-
-{\color{blue} This is not true for the relational interps of
-  Lans. Define explicitly.}
-
+In the $\Lan$ case, instead, define
+\[
+\oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} f
+%: \oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \rho
+%\to \oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \rho'
+%\]
+%as
+%\[
+= (
+\oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \pi_1 f,
+\oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \pi_2 f
+)
+\]
 \end{dfn}
 
+We need to show that the functorial action for the $\Lan$ case is well-defined, i.e.,
+that for every $(t_1, t_2) : \oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}$
+we have that
+\[
+(
+\oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \pi_1 f t_1,
+\oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \pi_2 f t_2
+)
+: \oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \rho'
+\]
+From $(t_1, t_2) : \oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}$
+we have, by definition, that there exist
+$\ol{Z : \oCPO_0}$,
+$t_1' : \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\pi_1\rho[\ol{\alpha := Z}]$,
+$t_2' : \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\pi_2\rho[\ol{\alpha := Z}]$,
+$\ol{g : \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho[\ol{\alpha := \Eq_Z}] \to \oCPORsem{\Gamma; \Phi \vdash A}\rho}$
+such that
+$\iota_{\ol{Z},\ol{\pi_1 g}}t_1' = t_1$,
+$\kappa_{\ol{Z}, \ol{\pi_2 g}}t_2' = t_2$,
+and
+$(t_1',t_2') : \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := \Eq_Z}]$.
+To show that
+\[
+(
+\oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \pi_1 f t_1,
+\oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \pi_2 f t_2
+)
+: \oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \rho'
+\]
+let
+$\ol{S = Z : \oCPO_0}$,
+$s_1 = \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F} (\pi_1 f) [\ol{\alpha := Z}] t'_1$,
+$s_2 = \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F} (\pi_2 f) [\ol{\alpha := Z}] t'_2$,
+and
+$\ol{h = \oCPORsem{\Gamma; \Phi \vdash A} f \circ g}$.
+The first thing we need to show is that $(s_1, s_2) : \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := \Eq_S}]$,
+and that is because $(s_1, s_2) = \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F} f [\ol{\alpha := \Eq_Z}] (t'_1, t'_2)$.
+The second thing we need to show is that $\iota'_{\ol{S},\ol{\pi_1 h}} s_1' = s_1$ and $\kappa'_{\ol{S},\ol{\pi_2 h}} s_2' = s_2$.
+That is because
+\[
+\begin{array}{rl}
+\iota'_{\ol{S},\ol{\pi_1 h}} s_1'
+&= \iota'_{\ol{S},\ol{\oCPOsem{\Gamma; \Phi \vdash A} \pi_1 f \circ \pi_1 g}} ( \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F} (\pi_1 f) [\ol{\alpha := Z}] t'_1 ) \\
+&= (\oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} (\pi_1 f) \circ \iota_{\ol{S},\ol{\pi_1 g}}) t'_1 \\
+&= \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} (\pi_1 f) (\iota_{\ol{S},\ol{\pi_1 g}} t'_1) \\
+&= \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} (\pi_1 f) t_1 \\
+&= s_1
+\end{array}
+\]
+where the second equality is by the commutativity of diagram~\ref{cd:lan-comm}.
+Analogously, we show that $\kappa'_{\ol{S},\ol{\pi_2 h}} s_2' = s_2$.
+
 To see that the functors given by Definitions~\ref{def:rel-sem}
-and~\ref{def:rel-sem-funcs} are well-defined we must show that, for
+and~\ref{def:rel-sem-funcs} are well-defined we must also show that, for
 every $H$, $T_{H,\rho}\,F$ is an $\oCPO$ relation transformer for any

### 2020-09-24 20-gadts/inf-rules.tex
repair functorial actions, change Lan intro and elim rules

--- 2020-09-17 20-gadts/inf-rules.tex
@@
 \AXC{$\ol{\Gamma; \ol{\alpha} \vdash K}$}
-\AXC{$\ol{\Gamma;\Phi \vdash A} \hspace*{0.3in} \Gamma;\Phi~|~\Delta \vdash t : F[\ol{\alpha :=A}]$}
-\TIC{$\Gamma;\Phi~|~\Delta \vdash \int_{\ol K,F} t :
-  (\Lan^{\ol\alpha}_{\ol K}\,  F)\,\ol{K[\ol{\alpha := A}]}$}  
+%\AXC{$\ol{\Gamma;\Phi \vdash A} \hspace*{0.3in} \Gamma;\Phi~|~\Delta \vdash t : F[\ol{\alpha :=A}]$}
+\BIC{$\Gamma;\emptyset~|~\Delta \vdash \int_{\ol K,F} :
+\Nat^{\Phi,\ol\alpha}\,F\, (\Lan^{\ol\alpha}_{\ol K}\,  F)\ol{K}$}  
 \DisplayProof
@@
   F\;G[\ol{\beta := K}]$}
-\AXC{$\ol{\Gamma;\Phi \vdash B}$}
-\AXC{$\Gamma;\Phi~|~\Delta \vdash t : (\Lan^{\ol\alpha}_{\ol K} F)\,\ol B $}
-\TIC{$\Gamma;\Phi~|~\Delta \vdash \partial^{G, \ol K}_F \eta\,t :
- G[\ol{\beta := B}]$}
+%\AXC{$\ol{\Gamma;\Phi \vdash B}$}
+%\AXC{$\Gamma;\Phi~|~\Delta \vdash t : (\Lan^{\ol\alpha}_{\ol K} F)\,\ol B $}
+\UIC{$\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol K}_F \eta :
+\Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K} F)\ol \beta\; G$}
 \DisplayProof
@@
 
-{\color{blue} Need to check that the extra condition in the interp of
-  Lans is preserved for the last clause to be well-defined.}
+We need to check that the functorial action on $\Lan$ types is
+well-defined, i.e., that the extra condition in the interpretation of
+$\Lan$ types is preserved by the functorial action, i.e., if $t :
+\oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}
+\rho$, then $\oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}}
+  F)\ol{A}}f t : \oCPOsem{\Gamma;\Phi \vdash
+  (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \rho'$.  By definition, if $t
+: \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}
+\rho$, then $(t, t) \in \oCPORsem{\Gamma;\Phi \vdash
+  (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \Eq_{\rho}$, i.e., there
+exist $\ol{Z : \oCPO_0}$, $t_1 : \oCPOsem{\Gamma; \Phi, \ol\alpha
+  \vdash F}\rho[\ol{\alpha := Z}]$, $t_2 : \oCPOsem{\Gamma; \Phi,
+  \ol\alpha \vdash F}\rho[\ol{\alpha := Z}]$, and $\ol{g :
+  \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\Eq_{\rho}[\ol{\alpha :=
+      \Eq_Z}] \to \oCPORsem{\Gamma; \Phi \vdash A}\Eq_{\rho}}$ such
+that $\iota_{\ol{Z}, \ol{\pi_1 g}}t_1 = t$, $\kappa_{\ol{Z}, \ol{\pi_2
+    g}}t_2 = t$, and $(t_1,t_2) : \oCPORsem{\Gamma; \Phi, \ol\alpha
+  \vdash F}\Eq_{\rho}[\ol{\alpha := \Eq_Z}]$.  To prove that
+$\oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f t
+: \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}
+\rho'$, let $\ol{Z' = Z}$, $t'_i = \oCPOsem{\Gamma; \Phi, \ol\alpha
+  \vdash F} f [\ol{\alpha := \id_{Z}}] t_i$ for $i = 1, 2$, and
+$\ol{g' = \oCPORsem{\Gamma; \Phi \vdash A}\Eq_f \circ g}$.  We have
+to show that $\iota'_{\ol{Z'}, \ol{\pi_1 g'}}t'_1 =
+\oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f
+t$, $\kappa'_{\ol{Z'}, \ol{\pi_2 g'}}t'_2 = \oCPOsem{\Gamma;\Phi
+  \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f t$, and $(t_1',t_2')
+: \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\Eq_{\rho'}[\ol{\alpha :=
+    \Eq_{Z'}}]$.  That $(t_1',t_2') : \oCPORsem{\Gamma; \Phi,
+  \ol\alpha \vdash F}\Eq_{\rho'}[\ol{\alpha := \Eq_{Z'}}]$ follows
+from the fact that $(t_1',t_2') = \oCPORsem{\Gamma; \Phi, \ol\alpha
+  \vdash F} \Eq_f [\ol{\alpha := \id_{\Eq_{Z}}}] (t_1, t_2)$ and
+$(t_1,t_2) : \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash
+  F}\Eq_{\rho}[\ol{\alpha := \Eq_Z}]$.  That $\iota'_{\ol{Z'},
+  \ol{\pi_1 g'}}t'_1 = \oCPOsem{\Gamma;\Phi \vdash
+  (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f t$ is because
+\[
+\begin{array}{rl}
+\iota'_{\ol{Z'}, \ol{\pi_1 g'}}t'_1
+&= \iota'_{\ol{Z'}, \ol{\pi_1 g'}} (\oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F} f [\ol{\alpha := \id_{Z}}] t_1) \\
+&= \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f (\iota_{\ol{Z}, \ol{\pi_1 g}} t_1) \\
+&= \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f t
+\end{array}
+\]
+where the second equality is justified by the commutativity of the diagram
+\begin{equation}\label{cd:lan-comm}
+\begin{tikzcd}[row sep = huge]
+\oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := Z}]
+  \ar[r, "{\iota_{\ol{Z}, \ol{\pi_1 g}}}"]
+  \ar[d, "{\oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F} f [\ol{\alpha := \id_Z}]}" description]
+&\colim{\ol{S}, \ol{h : \oCPOsem{\Gamma; \ol{\alpha} \vdash K}\rho[\ol{\alpha := S}] \to \oCPOsem{\Gamma; \Phi \vdash A}\rho}}
+{\oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := S}]}
+  \ar[d, "{(\Lan_{\ol{\oCPOsem{\Gamma; \ol{\alpha} \vdash K}\rho[\ol{\alpha := \_}]}} \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha := \id_{\_}}]) \ol{\oCPOsem{\Gamma; \Phi \vdash A}\rho}}" description] \\
+\oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := Z}]
+  \ar[r, "{\iota''_{\ol{Z}, \ol{\pi_1 g}}}"]
+  \ar[d, equal]
+& \colim{\ol{S}, \ol{h : \oCPOsem{\Gamma; \ol{\alpha} \vdash K}\rho'[\ol{\alpha := S}] \to \oCPOsem{\Gamma; \Phi \vdash A}\rho}}
+{\oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := S}]}
+  \ar[d, "{(\Lan_{\ol{\oCPOsem{\Gamma; \ol{\alpha} \vdash K}\rho[\ol{\alpha := \_}]}} \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := \_}]) \ol{\oCPOsem{\Gamma; \Phi \vdash A}f}}" description] \\
+\oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := Z}]
+  \ar[r, shift right = 1, "{\iota'_{\ol{Z'}, \ol{\pi_1 (\oCPORsem{\Gamma; \Phi \vdash A}\Eq_f \circ g)}}}"']
+& \colim{\ol{S}, \ol{h : \oCPOsem{\Gamma; \ol{\alpha} \vdash K}\rho'[\ol{\alpha := S}] \to \oCPOsem{\Gamma; \Phi \vdash A}\rho'}}
+{\oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := S}]}
+\end{tikzcd}
+\end{equation}
+%\[
+%\begin{tikzcd}
+%\colim{\ol{S}, \ol{h : \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho[\ol{\alpha := S}] \to \oCPORsem{\Gamma; \Phi \vdash A}\rho}}
+%{\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := S}]}
+%  \ar[r, bend left = 10, "{(\Lan_{\ol{\oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho[\ol{\alpha := \_}]}} \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha := \id_{\_}}]) \ol{\oCPORsem{\Gamma; \Phi \vdash A}\rho}}"]
+%& \colim{\ol{S}, \ol{h : \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho'[\ol{\alpha := S}] \to \oCPORsem{\Gamma; \Phi \vdash A}\rho'}}
+%{\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := S}]} \\
+%\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := Z}]
+%  \ar[u, "{\iota_{\ol{Z}, \ol{\pi_1 g}}}"]
+%  \ar[r, "{\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F} f [\ol{\alpha := \id_Z}]}"]
+%& \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := Z}]
+%  \ar[u, "{\iota''_{\ol{Z}, \ol{\pi_1 g}}}"]
+%\end{tikzcd}
+%\]
+%and
+%\[
+%\begin{tikzcd}
+%\colim{\ol{S}, \ol{h : \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho[\ol{\alpha := S}] \to \oCPORsem{\Gamma; \Phi \vdash A}\rho}}
+%{\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := S}]}
+%  \ar[r, bend left = 10, "{(\Lan_{\ol{\oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho[\ol{\alpha := \_}]}} \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := \_}]) \ol{\oCPORsem{\Gamma; \Phi \vdash A}f}}"]
+%& \colim{\ol{S}, \ol{h : \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho'[\ol{\alpha := S}] \to \oCPORsem{\Gamma; \Phi \vdash A}\rho'}}
+%{\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := S}]} \\
+%\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := Z}]
+%  \ar[u, "{\iota''_{\ol{Z}, \ol{\pi_1 g}}}"]
+%  \ar[r, equal]
+%& \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := Z}]
+%  \ar[u, "{\iota'_{\ol{Z'}, \ol{\pi_1 (\oCPORsem{\Gamma; \Phi \vdash A}f \circ g)}}}"]
+%\end{tikzcd}
+%\]
+Analogously, we show that $\kappa_{Z,\pi_2 g'}t'_2 = \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f t$.
 
@@
     A}\pi_2\rho}{\oCPOsem{\Gamma; \ol{\alpha} \vdash
-    F}\pi_2\rho[\ol{\alpha := Z}]}\]}
+    F}\pi_2\rho[\ol{\alpha := Z}]}\]
+We note that the pair $(\iota_{Z,h},\kappa_{Z,h})$ need not coincide
+with the injection $j_{Z,\Eq_h}$ into 
+\[\colim{Z\, :\, \oCPO_0, \;g \,:\, \oCPORsem{\Gamma; \ol{\alpha} \vdash
+      K}\rho[\ol{\alpha := \Eq_Z}] \to \oCPORsem{\Gamma; \Phi \vdash
+    A}\rho}{\oCPORsem{\Gamma; \ol{\alpha} \vdash
+    F}\rho[\ol{\alpha := \Eq_Z}]}\]}
 
@@
 $T^\oCPO_{H,f}$ are replaced by $T_{H,f}$.
-
-{\color{blue} This is not true for the relational interps of
-  Lans. Define explicitly.}
-
+In the $\Lan$ case, instead, define
+\[
+\oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} f
+%: \oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \rho
+%\to \oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \rho'
+%\]
+%as
+%\[
+= (
+\oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \pi_1 f,
+\oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \pi_2 f
+)
+\]
 \end{dfn}
 
+We need to show that the functorial action for the $\Lan$ case is
+well-defined, i.e., that for every $(t_1, t_2) : \oCPORsem{\Gamma;\Phi
+  \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}$ we have that
+\[
+(
+\oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} (\pi_1 f) t_1,
+\oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} (\pi_2 f) t_2
+)
+: \oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \rho'
+\]
+From $(t_1, t_2) : \oCPORsem{\Gamma;\Phi \vdash
+  (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}$ we have, by definition, that
+there exist $\ol{Z : \oCPO_0}$, $t_1' : \oCPOsem{\Gamma; \Phi,
+  \ol\alpha \vdash F}(\pi_1\rho)[\ol{\alpha := Z}]$, $t_2' :
+\oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}(\pi_2\rho)[\ol{\alpha :=
+    Z}]$, $\ol{g : \oCPORsem{\Gamma; \ol{\alpha} \vdash
+    K}\rho[\ol{\alpha := \Eq_Z}] \to \oCPORsem{\Gamma; \Phi \vdash
+    A}\rho}$ such that $\iota_{\ol{Z},\ol{\pi_1 g}}t_1' = t_1$,
+$\kappa_{\ol{Z}, \ol{\pi_2 g}}t_2' = t_2$, and $(t_1',t_2') :
+\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha :=
+    \Eq_Z}]$.  To show that
+\[
+(
+\oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} (\pi_1 f) t_1,
+\oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} (\pi_2 f) t_2
+)
+: \oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \rho'
+\]
+let $\ol{S = Z}$, $s_1' = \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}
+(\pi_1 f) [\ol{\alpha := Z}] t'_1$, $s_2' = \oCPOsem{\Gamma; \Phi,
+  \ol\alpha \vdash F} (\pi_2 f) [\ol{\alpha := Z}] t'_2$, and $\ol{h =
+  \oCPORsem{\Gamma; \Phi \vdash A} f \circ g}$.  The first thing we
+need to show is that $(s'_1, s'_2) : \oCPORsem{\Gamma; \Phi, \ol\alpha
+  \vdash F}\rho'[\ol{\alpha := \Eq_Z}]$, and that is because $(s'_1,
+s'_2) = \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F} f [\ol{\alpha :=
+    \Eq_Z}] (t'_1, t'_2)$.  The second thing we need to show is that
+\[\iota'_{\ol{Z},\ol{\pi_1 h}} s_1' = \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} (\pi_1 f) t_1\] and
+\[\kappa'_{\ol{Z},\ol{\pi_2 h}} s_2' = \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} (\pi_2 f) t_2\]  That is because
+\[
+\begin{array}{rl}
+\iota'_{\ol{Z},\ol{\pi_1 h}} s_1' &=
+\iota'_{\ol{Z},\ol{\oCPOsem{\Gamma; \Phi \vdash A} (\pi_1 f) \circ \pi_1
+    g}} ( \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F} (\pi_1 f)
+      [\ol{\alpha := \id_Z}] t'_1 ) \\
+%      &= (\oCPOsem{\Gamma;\Phi \vdash
+%        (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} (\pi_1 f) \circ
+      %      \iota_{\ol{Z},\ol{\pi_1 g}}) t'_1
+       &= \oCPOsem{\Gamma;\Phi
+        \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} (\pi_1 f)
+      (\iota_{\ol{Z},\ol{\pi_1 g}} t'_1) \\ &= \oCPOsem{\Gamma;\Phi
+        \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} (\pi_1 f)
+      t_1 \end{array}
+\]
+where the second equality is by the commutativity of diagram~\ref{cd:lan-comm}.
+Analogously, we show that $\kappa'_{\ol{Z},\ol{\pi_2 h}} s_2' = \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} (\pi_2 f) t_2$.
+
 To see that the functors given by Definitions~\ref{def:rel-sem}
-and~\ref{def:rel-sem-funcs} are well-defined we must show that, for
+and~\ref{def:rel-sem-funcs} are well-defined we must also show that, for
 every $H$, $T_{H,\rho}\,F$ is an $\oCPO$ relation transformer for any

### 2020-09-27 20-gadts/inf-rules.tex
revised Sections 1 and 2

--- 2020-09-17 20-gadts/inf-rules.tex
@@
 
-\new{Maybe develop our theory for {\em any} $\lambda \geq \omega_1$,
-  and then specialize to $\omega_1$ when discussing GADTs? Can we do
-  that? It seems we really use properties of $\oCPO$ to get that
-  interpretations of $\Nat$-types are well-defined.}
-
-\new{Locally presentable categories are the closest approximation to
-  small complete categories. It's remarkable that local presentability
-  is sufficient not just to give semantics to GADTs, but is also all
-  that is needed to ensure that the model induced by that semantics is
-  actually parametric.}
-
-{\color{red} We should ask what, intuitively, it means to have
+{\color{blue} We should ask what, intuitively, it means to have
   parametricity for a calculus with GADTs. In particular it should
@@
   means that the datatype elements are constructed in exactly the same
-  ways at each type instance. For GADTs, it means that as well, but
-  here we also have to take into account which instances are actually
-  inhabited, since not all type instances of a GADT are inhabited.
-  [Of course we did not have to do this for ADTs and nested types
-    because {\em all} instances of ADTs and nested types are
-    inhabited.] So, intuitively, we consider a GADT to be
-  ``parametric'' if 1) the corresponding instances of the GADT are
-  inhabited for for every instantiation of its type parameters
-  $\ol\alpha$, and 2) the objects for each instantiation of the
-  $\alpha$s are constructed in exactly the same way. That is, which
-  instances are inhabited, as well as how the data in each of those
-  instances are constructed, is type independent. We capture this in
-  our calculus by requiring that, for $\Gamma;\Phi \vdash
-  (\Lan^{\ol{\alpha^0}}_{\ol K} F)\, \ol A$, we have that
-  $\emptyset;\ol\beta \vdash K$, and that $K$ does not involve any
-  constants. (The latter ensures that a GADT with a constructor of
-  type $\alpha \to Int$ is not parametric, which makes sense since
-  there will be more ways to construct Int instances of the GADT than
-  ways to construct its other type instances.) Further, to ensure that
-  the categorical Lans interpreting $\Lan$-constructs are
-  $\omega$-cocontinuous we also require that $K$ is a polynomial
-  functor in the $\alpha$s (that necessarily will not involve any
-  constant functors). This means that our constructors have types of
-  the form $F\ol\beta \to F(K\ol\beta)$ with no constants in $K$, so
-  that exactly the corresponding instances of the GADT are inhabited
-  for every instantiation of the $\alpha$s, and the objects for each
-  instantiation of the $\alpha$s are constructed in exactly the same
-  way. That is, the inhabited instances and the construction of the
-  data in those instances are type independent.}
+  ways at each type instance. For GADTs, it means that as well. That
+  is, we require related elements of a GADT to be constructed with the
+  same sequence of constructors acting on related data. This entails
+  that which instances are inhabited, as well as how the data in each
+  of those instances are constructed, is type independent.}
+
+\new{Locally presentable categories are the closest approximation to
+  small complete categories. It's remarkable that local presentability
+  is sufficient not just to give semantics to GADTs, but is also all
+  that is needed to ensure that the model induced by that semantics is
+  actually parametric.}
+
+\new{Maybe develop our theory for {\em any} $\lambda \geq \omega$, and
+  then specialize to $\omega_1$ (for $\omega$CPO) when discussing
+  GADTs in future work? Can we do that? It seems we really use
+  properties of $\oCPO$ to get that interpretations of $\Nat$-types
+  are well-defined. It's not clear how to put $\omega$CPO structures
+  on the interpretations of $\Lan$-types.}
 
@@
 difference with~\cite{jgj20} here lies solely in the incorporation of
-functorial expressions constructed from $\Lan$.
+functorial expressions constructed using $\Lan$.
 
@@
 \vspace*{0.2in}
+
+\new{Sum and product intro and elim rules should be annotated with
+  constituent types for consistency?}
 
@@
 the types of $\tin_H$ and $\fold^F_H$ to be well-formed, the length of
-$\alpha$ must equal the arity of $\phi$.) In the rule for $\int_{\ol
-  K,H}\,t$, there is one functorial expression $A$ for every
-functorial variable in $\ol \alpha$, and in the rule for
-$\partial^{G,\ol K}_F\,\eta\,t$, there is one functorial expression
-$A$ for every functorial expression in $\ol K$ (and hence for every
-functorial variable in $\ol \beta$).
-
-Substitution for terms is the obvious extension
-of the usual capture-avoiding textual substitution, and
+$\alpha$ must equal the arity of $\phi$.)  {\color{red} In the rule
+  for $\int_{\ol K,H}$, ANYTHING NEEDED HERE?}
+
+Substitution for terms is the obvious extension of the usual
+capture-avoiding textual substitution, and
 Definition~\ref{def:well-formed-terms} ensures that the expected
@@
 
-\vspace*{0.2in}
-
-\new{Sum and product intro and elim rules should be annotated with
-  constituent types for consistency?
-
-\vspace*{0.2in}
-
-We should have a computation rule along the lines of: If $\eta :
+\vspace*{0.1in}
+
+\new{We should have a computation rule along the lines of: If $\eta :
 \Nat^{\ol\alpha} F \,G[\ol{\beta := K}]$ then
@@
 \section{Interpreting Types}\label{sec:type-interp}
+
+We denote the category of sets and functions by $\set$. The category
+$\rel$ has as its objects triples $(A,B,R)$ where $R$ is a relation
+between the objects $A$ and $B$ in $\set$, i.e., a subset of $A \times
+B$, and has as its morphisms from $(A,B,R)$ to $(A',B',R')$ pairs $(f
+: A \to A',g : B \to B')$ of morphisms in $\set$ such that $(f a,g\,b)
+\in R'$ whenever $(a,b) \in R$. We write $R : \rel(A,B)$ in place of
+$(A,B,R)$ when convenient.  If $R : \rel(A,B)$ we write $\pi_1 R$ and
+$\pi_2 R$ for the {\em domain} $A$ of $R$ and the {\em codomain} $B$
+of $R$, respectively.  If $A : \set$, then we write $\Eq_A =
+(A,A,\{(x,x)~|~ x \in A\})$ for the {\em equality relation} on $A$.
 
@@
 
-{\color{red} NOT NEEDED ANYMORE??
-To accommodate GADTs, we will need to transition Reynolds' approach
-from a $\set$-based semantics to a semantics based on
-$\omega$-complete partial orders. We denote the category of
-$\omega$-complete partial orders ($\oCPO$s) and their {\color{blue}
-  monotone?} sup-preserving morphisms by $\omega$CPO. The underlying
-set of an $\oCPO$ $A$ is denoted $|A|$. The category $\oCPOR$ of
-$\oCPO$ relations has as its objects triples $(A,B,R)$, where $A, B:
-\oCPO$; $R: \rel(|A|, |B|)$; {\color{blue} $(a,b) \in R$, $a \leq a'$,
-  and $b \leq b'$ imply $(a',b') \in R$; so next part is redundant?}
-and $(\bigvee_{i < \omega} a_i, \bigvee_{i < \omega} b_i) \in R$
-whenever $(a_i)_{i < \omega}$ and $(b_i)_{i < \omega}$ are chains in
-$A$ and $B$, respectively, such that $(a_i,b_i) \in R$ for all
-$i$. The category $\oCPOR$ has as its morphisms from $(A,B,R)$ to
-$(A',B',R')$ pairs $(f : A \to A',g : B \to B')$ of morphisms in
-$\oCPO$ such that $(f\,a, g\,b) \in R'$ whenever $(a,b) \in R$. We
-note that if $(f,g) : (A,B,R) \to (A',B',R')$ and $(a_i)_{i < \omega}$
-and $(b_i)_{i < \omega}$ are chains in $A$ and $B$, respectively, then
-$ (f \, (\bigvee_{i < \omega} a_i), g \, (\bigvee_{i < \omega} b_i)) =
-(\bigvee_{i < \omega} (f a_i), \bigvee_{i < \omega} (g b_i)) \in R'$
-necessarily holds.}  We write $R : \oCPOR(A,B)$ in place of $(A,B,R):
-\oCPOR$ when convenient. If $R : \oCPOR(A,B)$ then we write $\pi_1 R$
-and $\pi_2 R$ for the {\em domain} $A$ of $R$ and the {\em codomain}
-$B$ of $R$, respectively. If $A : \oCPO$, then we write $\Eq_A =
-(A,A,\Eq_{|A|})$ for the {\em equality relation} on $A$.
-
-To adapt Reynolds' approach, we first inductively define, for each
-type, an object interpretation in $\oCPO$ and a relational
-interpretation in $\oCPOR$.  Next, we show that these interpretations
-satisfy both an Identity Extension Lemma (Theorem~\ref{thm:iel}) and
-an Abstraction Theorem (Theorem~\ref{thm:gen-at}) appropriate to the
-$\oCPO$ setting. The key to proving our Identity Extension Lemma is a
-familiar ``cutting down'' of the interpretations of universally
-quantified types to include only the ``parametric'' elements; as
-in~\cite{jgj20}, the relevant types of the calculus defined above are
-the (now richer) $\Nat$-types.  The requisite cutting down requires
-that the object interpretations of our types in $\omega$CPO are
+In the remainder of this section we first inductively define, for each
+type, an object interpretation in $\set$ and a relational
+interpretation in $\rel$. In Section~\ref{sec:iel} we show that these
+interpretations satisfy both an Identity Extension Lemma
+(Theorem~\ref{thm:iel}) and an Abstraction Theorem
+(Theorem~\ref{thm:gen-at}) appropriate to the GADT setting. The key to
+proving our Identity Extension Lemma is a familiar ``cutting down'' of
+the interpretations of universally quantified types to include only
+the ``parametric'' elements; the relevant types of the calculus
+defined above are the $\Nat$-types (which are now richer than those
+of~\cite{jgj20}) and the $\Lan$-types. The requisite cutting down
+requires that the object interpretations of our types in $\set$ are
 defined simultaneously with their relational interpretations in
-$\oCPOR$.  We give the object interpretations for our types in
+$\rel$. We give the object interpretations for our types in
 Section~\ref{sec:obj-interp} and give their relational interpretations
 in Section~\ref{sec:rel-interp}.  While the former are relatively
-straightforward, the latter are less so, mainly because of the
-cocontinuity conditions, adapted from the $\set$-based setting
-of~\cite{jgj20}, that must hold if they are to be well-defined. We
-develop these conditions in Section~\ref{sec:rel-interp}, which
-separates Definitions~\ref{def:set-sem} and~\ref{def:rel-sem} in
-space, but otherwise has no impact on the fact that they are given by
-mutual induction.
+straightforward, the latter are less so because of the delicacy of
+interpreting the $\Lan$-types. (The object and relational
+interpretations of $\Lan$-types are not defined in the same
+``parallel'' way, as are the two interpretations of the other types.)
+Another subtle point is guaranteeing the cocontinuity conditions,
+adapted to our richer setting from~\cite{jgj20}, that must hold if the
+type interpretations are to be well-defined. We develop these
+conditions in Section~\ref{sec:rel-interp}, which separates
+Definitions~\ref{def:set-sem} and~\ref{def:rel-sem} in space, but
+otherwise has no impact on the fact that these definitions are given
+by mutual induction.
 
@@
 $\omega$-cocontinuous functors between categories of
-$\omega$-cocontinuous functors on categories constructed from the
-locally $\omega$-presentable category $\oCPO$.  We therefore begin
-by recording some important facts about locally $\omega$-presentable
-categories and functors on them, and verifying verify the properties
-needed to interpret our syntax.
-
-\subsubsection{Preliminaries}
-
-\new{Perhaps have as preliminaries to entire paper.}
-
-A category is \emph{small} if its collection of morphisms is a set.
-It is \emph{locally small} if, for any two objects $A$ and $B$, the
-collection of morphisms from $A$ to $B$ is a set. A {\em small
-  (co)limit} in a category $\mcC$ is a (co)limit of a diagram $F :
-\mcA \to \mcC$, where $\mcA$ is a small category. A category $\mcC$
-is {\em (co)complete} if it has all small (co)limits.
-
-A poset $\mcD = (D,\leq)$ is \emph{$\omega$-directed} if every
-countable subset of $D$ has a supremum. When $\mcD$ is considered
-as a category, we write $d \in \mcD$ to indicate that $d$ is an object
-of $\mcD$ (i.e., $d \in D$). An {\em $\omega$-directed colimit} in a
-category $\mcC$ is a colimit of a diagram $F : {\mathcal D} \to \mcC$,
-where $\mathcal D$ is an $\omega$-directed poset. A category $\mcC$
-is {\em $\omega$-cocomplete} if it has all $\omega$-directed
-colimits; a {\em cocomplete} category is one that has all colimits.
-
-If $\mcA$ and $\mcC$ are $\omega$-cocomplete, then a functor $F :
-\mcA \to \mcC$ is {\em $\omega$-cocontinuous} if it preserves
-$\omega$-directed colimits. We write $[\A,\C]_{\omega}$ for the
-category of $\omega$-cocontinuous functors from $\A$ to $\C$, and
-$\C^\A$ for the category of {\em all} functors from $\A$ to $\C$.
-Since (co)limits of functors are computed pointwise, $C^\A$ has all
-(co)limits that $\C$ has, and (co)limits of (co)continuous functors
-are again (co)continuous. It follows that $[\A,\C]_{(\omega)}$ is
-($\omega$-)(co)complete whenever $\C$ is. In particular, if $F :
-[\C, \C]_{(\omega)}$ for a(n) ($\omega$-)cocomplete category $\C$,
-then we can define $F^0 = \Id$, $F^{\alpha + 1} = F \circ F^\alpha$ at
-successor ordinals, and $F^{\bigcup_{\beta < \alpha} \beta} =
-\colim{\beta < \alpha} F^\beta$ at limit ordinals.
-
-If $\mcA$ is locally small, then an object $A$ of $\mcA$ is {\em
-  $\omega$-presentable} if the functor $\mathsf{Hom}_\mcA(A, -) :
-\mcA \to \Set$ preserves $\omega$-directed colimits, i.e., if for
-every $\omega$-directed poset $\mathcal D$ and every functor $F :
-{\mathcal D} \to \A$, there is a canonical isomorphism $\colim{d \in
-  \mcD}{\mathsf{Hom}_\mcA(A,Fd)} \simeq \mathsf{Hom}_{\mcA}(A,
-\colim{d \in \mcD}{Fd})$. A locally small category $\mcA$ is {\em
-  $\omega$-accessible} if it is $\omega$-cocomplete and has a set
-$\mcA_0$ of $\omega$-presentable objects such that every object is
-an $\omega$-directed colimit of objects in $\mcA_0$; a locally small
-category is {\em locally $\omega$-presentable} if it is
-$\omega$-accessible and cocomplete.
-
-The category $\oCPO$ is locally presentable; its presentable objects
-are precisely the finite sets.
-%The category $\oCPO$ is locally $\omega$-presentable (but not
-%locally finitely presentable); see Examples~1.18(2)
-%of~\cite{ar94}). Its $\omega$-presentable objects are precisely the
-%$\omega$CPOs cardinality less than $\omega$, i.e., the countable
-%$\omega$CPOs.
-In the next subsection we will interpret type variables in $\tvars^k
-\cup \fvars^k$ as elements of $[\oCPO^k,\oCPO]_{\omega}$; the
-following special cases of standard results (see, e.g.,~\cite{ar94})
-will therefore be critical to deducing important properties of our
-object interpretations of types:
-\begin{prop}\label{prop:prelims}
-\begin{enumerate}
-\item If $\C_1,...,\C_n$ are locally $\omega$-presentable categories
-  then so is $\C_1 \times ... \times \C_n$. Moreover, the presentable
-  objects of $\C_1 \times ... \times \C_n$ are exactly the tuples of
-  the form $(P_1,...,P_n)$, where, for each $i = 1,...,n$, the object
-  $P_i$ is presentable in $\C_i$.
-\item If $\mathcal{A}$ is $\omega$-accessible and $\mathcal{C}$ is
-  $\lambda$-cocomplete, then the category $[\mathcal{A},
-  \mathcal{C}]_{\omega}$ is naturally equivalent to the category
-  $\mathcal{C}^{\mathcal{A}_0}$.
-\item If $\mathcal{C}$ is locally $\omega$-presentable and
-  $\mathcal{A}_0$ is essentially small, then
-  $\mathcal{C}^{\mathcal{A}_0}$ is locally $\omega$-presentable.
-\end{enumerate}
-\end{prop}
-\noindent
-Together, the statements in Proposition~\ref{prop:prelims} give that
-if $\mathcal{A}$ and $\mathcal{C}$ are locally $\omega$-presentable,
-then $[\mathcal{A}, \mathcal{C}]_{\omega}$ is naturally equivalent
-to $\mathcal{C}^{\mathcal{A}_0}$, and thus is
-$\omega$-presentable. Thus, for all $k_1,...,k_n \in \nat^n$,
-$[\mathcal{A}^{k_1}, \mathcal{C}]_{\omega} \times ... \times
-[\mathcal{A}^{k_n}, \mathcal{C}]_{\omega}$ is locally
-$\omega$-presentable, and therefore $[[\mathcal{A}^{k_1},
-    \mathcal{C}]_{\omega} \times ... \times [\mathcal{A}^{k_n},
-    \mathcal{C}]_{\omega}, \mathcal{C}]_{\omega}$ is as
-well. Taking both $\A$ and $C$ to be $\oCPO$ --- as we will to ensure
-that the fixpoints interpreting $\mu$-types in $\oCPO$ exist --- we
-have
-\begin{prop}\label{prop:nat-interps-sets}
-  For all $k_1,...,k_n \in \nat^n$,
-  \[[[\oCPO^{k_1}, \oCPO]_{\omega}
-  \times ... \times [\oCPO^{k_n}, \oCPO]_{\omega},
-  \oCPO]_{\omega}\] is locally $\omega$-presentable.
-\end{prop}
-
-\subsubsection{Object Interpretations}
-
-To define the object interpretations of the types in
-Definition~\ref{def:wftypes} we must first interpret their
-variables. We have:
+$\omega$-cocontinuous functors between categories constructed from the
+locally finitely presentable category $\set$~\cite{ar94}; the
+relational interpretations will be similar for the locally finitely
+presentable category $\rel$. The fact that functor categories of
+locally finitely presentable categories are again locally finitely
+presentable ensures that the fixpoints interpreting $\mu$-types in
+$\set$ and $\rel$ exist, and thus that both the set and relational
+interpretations of all of the types in Definition~\ref{def:wftypes}
+are well-defined~\cite{jp19}.
+
+To define the interpretations of the types in
+Definition~\ref{def:wftypes} we must first interpret their variables.
+Writing $[\C,\D]$ for the category of $\omega$-cocontinuous functors
+from $\C$ to $\D$ for the locally finitely presentable categories $\C$
+and $\D$, for our $\set$ interpretations we have:
 
 \begin{dfn}\label{def:set-env}
-A {\em $\oCPO$ environment} maps each type variable in $\tvars^k \cup
-\fvars^k$ to an element of $[\oCPO^k,\oCPO]_{\omega}$.  A morphism
-$f : \rho \to \rho'$ for set environments $\rho$ and $\rho'$ with
-$\rho|_\tvars = \rho'|_\tvars$ maps each type constructor variable
-$\psi^k \in \tvars$ to the identity natural transformation on $\rho
-\psi^k = \rho'\psi^k$ and each functorial variable $\phi^k \in \fvars$
-to a natural transformation from the $k$-ary functor $\rho \phi^k$ on
+A {\em $\set$ environment} maps each type variable in $\tvars^k \cup
+\fvars^k$ to an element of\, $[\oCPO^k,\oCPO]$. A morphism $f : \rho
+\to \rho'$ for set environments $\rho$ and $\rho'$ with $\rho|_\tvars
+= \rho'|_\tvars$ maps each type constructor variable $\psi^k \in
+\tvars$ to the identity natural transformation on $\rho \psi^k =
+\rho'\psi^k$ and each functorial variable $\phi^k \in \fvars$ to a
+natural transformation from the $k$-ary functor $\rho \phi^k$ on
 $\oCPO$ to the $k$-ary functor $\rho' \phi^k$ on $\oCPO$.  Composition
@@
 
-When convenient we identify a functor in $[\oCPO^0, \oCPO]_{\omega}$
-with its value on $\ast$ and consider a $\oCPO$ environment to map a
-type variable of arity $0$ to an $\omega$-cocontinuous functor from
+When convenient we identify a functor in $[\oCPO^0, \oCPO]$ with its
+value on $\ast$ and consider a $\oCPO$ environment to map a type
+variable of arity $0$ to an $\omega$-cocontinuous functor from
 $\oCPO^0$ to $\oCPO$, i.e., to an $\oCPO$. If $\Phi =
 \{\phi_1^{k_1},...,\phi_n^{k_n}\}$ and $\ol K = \{K_1,...,K_n\}$ where
-$K_i : [\oCPO^{k_i},\oCPO]_{\omega}$ for $i = 1,...,n$, then we
-write either $\rho[\ol{\Phi := K}]$ or $\rho[\ol{\phi := K}]$ for the
-$\oCPO$ environment $\rho'$ such that $\rho' \phi_i = K_i$ for $i =
-1,...,n$ and $\rho' \phi = \rho \phi$ if $\phi \not \in \Phi$.  If
-$\rho$ is an $\oCPO$ environment, we write $\Eq_\rho$ for the $\oCPOR$
-environment (see Definition~\ref{def:reln-env}) such that $\Eq_\rho v
-= \Eq_{\rho v}$ for every type variable $v$. The categories
-$\oCPORT_k$ and relational interpretations appearing in the third
-clause of Definition~\ref{def:set-sem} are given in full in
-Section~\ref{sec:rel-interp}. We note that $\omega$-directed
-colimits in $\oCPOenv$ are taken pointwise.
+$K_i : [\oCPO^{k_i},\oCPO]_{\omega}$ for $i = 1,...,n$, then we write
+$\rho[\Phi := \ol K]$ for the $\oCPO$ environment $\rho'$ such that
+$\rho' \phi_i = K_i$ for $i = 1,...,n$ and $\rho' \phi = \rho \phi$ if
+$\phi \not \in \Phi$.  If $\rho$ is an $\oCPO$ environment, we write
+$\Eq_\rho$ for the $\oCPOR$ environment (see
+Definition~\ref{def:reln-env}) such that $\Eq_\rho v = \Eq_{\rho v}$
+for every type variable $v$. The categories $\oCPORT_k$ and relational
+interpretations appearing in the third clause of
+Definition~\ref{def:set-sem} are given in full in
+Section~\ref{sec:rel-interp}.
 
@@
 The {\em object interpretation} $\oCPOsem{\cdot} : \F \to [\oCPOenv,
-  \oCPO]_{\omega}$ is defined by
+  \oCPO]$ is defined by
 \begin{align*}
@@
     \Phi \vdash
-    F}\rho[\ol{\Phi := K}] 
+    F}\rho[\Phi := \ol K] 
       \Rightarrow \lambda \ol{K}.\,\oCPOsem{\Gamma; 
-        \Phi \vdash G}\rho[\ol{\Phi := K}] \\ 
+        \Phi \vdash G}\rho[\Phi := \ol K] \\ 
       &\hspace{0.3in}|~\forall \ol{K = (K^1, K^2, K^*) : \oCPORT_k}.\\
       &\hspace{0.4in}(\eta_{\overline{K^1}}, \eta_{\overline{K^2}})
-      : \oCPORsem{\Gamma; \Phi \vdash F}\Eq_{\rho}[\ol{\Phi := K}]
+      : \oCPORsem{\Gamma; \Phi \vdash F}\Eq_{\rho}[\Phi := \ol K]
       \rightarrow \oCPORsem{\Gamma; \Phi \vdash
-        G}\Eq_{\rho}[\ol{\Phi := K}] \} \\
+        G}\Eq_{\rho}[\Phi := \ol K] \} \\
   \oCPOsem{\Gamma;\Phi \vdash \phi\ol{F}}\rho &=
@@
 \end{dfn}
-
-If $\rho \in \oCPOenv$ and $\vdash F$ then we write $\oCPOsem{\vdash F}$
-instead of $\oCPOsem{\vdash F}\rho$ since the environment is
-immaterial.
-
-For Definition~\ref{def:set-sem} to be well-defined, we have to check
-that each object interpretation is in $\oCPO$ and, in particular, that
-each contains sups of all $\omega$-chains. This will be proved by
-induction on types, and in most cases existence of sups of
-$\omega$-chains will follow from the induction hypotheses. However,
-well-definedness needs to be proved directly for object
-interpretations of $\Nat$-types. First, we have
+\new{WHERE DOES THIS ALL GO? Must explain $\Lan$s, say why they are
+  functors, say they can be computed using colimits in $\set$ and
+  $\rel$, define injections, etc. If $f : B \to B'$ in $\set$, then
+  the functorial action of $\mathit{Lan}_{\ol K}\,F$ on $f$ is the
+  unique morphism $(\mathit{Lan}_{\ol K}\,F)f$ such that
+  \begin{equation}\label{def:functorial-action-lan}
+(\mathit{Lan}_{\ol K}\,F)f \circ \iota_{X,g} = \iota'_{X, f \circ
+      g}~~~MISSING~SOME~BARS
+\end{equation}
+Here, $\iota_{X,g}$ is the injection of $FX$ into $(\mathit{Lan}_{\ol
+  K}\,F)\,B = \colim{X : \set_0,\, g : KX \to B}{FX}$ and
+$\iota'_{X,g'}$ is the injection of $FX$ into $(\mathit{Lan}_{\ol
+  K}\,F)\,B' = \colim{X : \set_0,\, g' : KX \to B'}{FX}$.}
+
+If $\rho : \oCPOenv$ and $\vdash F$ then we write $\oCPOsem{\vdash
+  F}$ instead of $\oCPOsem{\vdash F}\rho$ since the environment is
+immaterial. To know that Definition~\ref{def:set-sem} is
+well-defined we have to check that each object interpretation is in
+$\set$. First, we have
 \begin{lemma}\label{lem:nat-types-sets}
@@
 \[
-\eta : \lambda \ol K. \setsem{\Gamma;\Phi \vdash F}\rho[\ol{\Phi := K}]
-\Rightarrow \lambda \ol K. \setsem{\Gamma;\Phi \vdash G}\rho[\ol{\Phi :=
-  K}]
+\eta : \lambda \ol K. \setsem{\Gamma;\Phi \vdash F}\rho[\Phi := \ol{K}]
+\Rightarrow \lambda \ol K. \setsem{\Gamma;\Phi \vdash G}\rho[\Phi :=
+  \ol{K}]
 \]
@@
 We first note that $\lambda \ol K. \setsem{\Gamma;\Phi \vdash
-  F}\rho[\ol{\Phi := K}]$ and $\lambda \ol K. \setsem{\Gamma;\Phi
-  \vdash G}\rho[\ol{\Phi := K}]$ are both in $[\ol{[\oCPO^k ,
-      \oCPO]_{\omega}}, \oCPO]_{\omega}$. By
-Proposition~\ref{prop:nat-interps-sets}, $[\ol{[\oCPO^k ,
-      \oCPO]_{\omega}}, \oCPO]_{\omega}$ is locally
-$\omega$-presentable. It is therefore locally small, so there are
-only $\set$-many morphisms (i.e., natural transformations) between any
-two functors in $[\ol{[\oCPO^k , \oCPO]_{\omega}},
-  \oCPO]_{\omega}$. In particular, there are only $\set$-many
-natural transformations from $\lambda \ol K. \setsem{\Gamma;\Phi
-  \vdash F}\rho[\ol{\Phi := K}]$ to $\lambda \ol
-K. \setsem{\Gamma;\Phi \vdash G}\rho[\ol{\Phi := K}]$.
+  F}\rho[\Phi := \ol{K}]$ and $\lambda \ol K. \setsem{\Gamma;\Phi
+  \vdash G}\rho[\Phi := \ol{K}]$ are both in $[\ol{[\oCPO^k , \oCPO]},
+  \oCPO]$. Since $[\ol{[\oCPO^k , \oCPO]}, \oCPO]$ is locally finitely
+presentable it is locally small. There are therefore only $\set$-many
+morphisms (i.e., natural transformations) between any two functors in
+$[\ol{[\oCPO^k , \oCPO]}, \oCPO]$. In particular, there are only
+$\set$-many natural transformations from $\lambda \ol
+K. \setsem{\Gamma;\Phi \vdash F}\rho[\Phi := \ol{K}]$ to $\lambda \ol
+K. \setsem{\Gamma;\Phi \vdash G}\rho[\Phi := \ol{K}]$.
 \end{proof}
-
-%Now, if $\C$ is any category, then given functors $F,G : \C \to \oCPO$
-%and natural transformations $\eta, \eta' : F \to G$, we define $\eta
-%\leq \eta'$ iff $\eta_c \leq \eta'_c$ for all $c \in \C$, i.e., iff
-%$\eta_c\,x \leq \eta'_c\,x$ in $Gc$ for all $c \in \C$ and $x \in Fc$.
-%If $(\eta_i)_{i < \omega}$ is a chain in $\{\eta : F \to G\}$, then
-%the family of morphisms $(\bigvee_{i < \omega} \eta_i)_c = \lambda
-%x. \bigvee_{i < \omega} ((\eta_i)_c\,x) : Fc \to Gc$ defines a natural
-%transformation $\bigvee_{i < \omega} \eta_i : F \to G$, and this
-%natural transformation is clearly the supremum of $(\eta_i)_{i <
-%  \omega}$ in $\{\eta : F \to G\}$. We therefore have that $\{\eta : F
-%\to G\}$ is itself an $\oCPO$.
-%Letting $F = \lambda \ol
-%K. \setsem{\Gamma;\Phi \vdash F}\rho[\ol{\Phi := K}]$ and $G = \lambda
-%\ol K. \setsem{\Gamma;\Phi \vdash G}\rho[\ol{\Phi := K}]$, we have
-%that $\oCPOsem{\Gamma; \emptyset \vdash \Nat^\Phi F\,G}\rho$ is indeed
-%an $\oCPO$.
-
-%Now, if $\C$ is any category, then given functors $F,G : \C \to \oCPO$
-%and natural transformations $\eta, \eta' : F \to G$, we define $\eta
-%\leq \eta'$ iff $\eta_c \leq \eta'_c$ for all $c \in \C$, i.e., iff
-%$\eta_c\,x \leq \eta'_c\,x$ in $Gc$ for all $c \in \C$ and $x \in Fc$.
-%If $(\eta_i)_{i < \omega}$ is a chain in $\{\eta : F \to G\}$, then
-%the family of morphisms $(\bigvee_{i < \omega} \eta_i)_c = \lambda
-%x. \bigvee_{i < \omega} ((\eta_i)_c\,x) : Fc \to Gc$ defines a natural
-%transformation $\bigvee_{i < \omega} \eta_i : F \to G$, and this
-%natural transformation is clearly the supremum of $(\eta_i)_{i <  
-%  \omega}$ in $\{\eta : F \to G\}$. We therefore have that $\{\eta : F
-%\to G\}$ is itself an $\oCPO$.  To show that $\oCPOsem{\Gamma;
-%  \emptyset \vdash \Nat^\Phi \,F\,G}\rho$ is an $\oCPO$, we then still
-%need to prove that, for any chain of natural transformations
-%$(\eta_i)_{i < \omega}$ in $\oCPOsem{\Gamma; \emptyset \vdash
-%  \Nat^\Phi \,F\,G}\rho$, $\bigvee_{i < \omega} \eta_i$ is again in
-%$\oCPOsem{\Gamma; \emptyset \vdash \Nat^\Phi \,F\,G}\rho$.  By
-%hypothesis, each $\eta_i$ is such that, for all $\ol{K = (K^1, K^2,
-%  K^*) : \oCPORT_k}$,
-%\[((\eta_i)_{\overline{K^1}},
-%(\eta_i)_{\overline{K^2}}) : \oCPORsem{\Gamma; \Phi \vdash
-%  F}\Eq_{\rho}[\ol{\Phi := K}] \rightarrow \oCPORsem{\Gamma; \Phi
-%  \vdash G}\Eq_{\rho}[\ol{\Phi := K}]\] is a morphism in
-%$\oCPOR$. To see that, for all $\ol{K = (K^1, K^2, K^*) : \oCPORT_k}$,
-%%\[((\bigvee_{i < \omega} \eta_i)_{\overline{K^1}}, (\bigvee_{i <
-%%%  \omega} \eta_i)_{\overline{K^2}}) : \oCPORsem{\Gamma; \Phi \vdash
-%%  F}\Eq_{\rho}[\ol{\Phi := K}] \rightarrow \oCPORsem{\Gamma; \Phi
-%%  \vdash G}\Eq_{\rho}[\ol{\Phi := K}]\] is again a morphism in
-%%$\oCPOR$, we observe that
-%$((\bigvee_{i < \omega} \eta_i)_{\overline{K^1}}, (\bigvee_{i <
-%  \omega} \eta_i)_{\overline{K^2}})$ is a morphism from
-%$\oCPORsem{\Gamma; \Phi \vdash F}\Eq_{\rho}[\ol{\Phi := K}]$ to
-%$\oCPORsem{\Gamma; \Phi \vdash G}\Eq_{\rho}[\ol{\Phi := K}]$ in
-%$\oCPOR$, we simply observe that
-%\[((\bigvee_{i < \omega} \eta_i)_{\overline{K^1}}, (\bigvee_{i < \omega}
-%\eta_i)_{\overline{K^2}}) = (\bigvee_{i < \omega}
-%(\eta_i)_{\overline{K^1}}, \bigvee_{i < \omega}
-%(\eta_i)_{\overline{K^2}}) = \bigvee_{i < \omega}
-%((\eta_i)_{\overline{K^1}}, (\eta_i)_{\overline{K^2}})\] so
-%$((\bigvee_{i < \omega} \eta_i)_{\overline{K^1}}, (\bigvee_{i <
-%  \omega} \eta_i)_{\overline{K^2}})$ is the sup of a chain of
-%morphisms from $\oCPORsem{\Gamma; \Phi \vdash F}\Eq_{\rho}[\ol{\Phi :=
-%    K}]$ to $\oCPORsem{\Gamma; \Phi \vdash G}\Eq_{\rho}[\ol{\Phi :=
-%    K}]$ in $\oCPOR$ and thus is itself such a morphism. Taken
-%together, this all shows that $\oCPOsem{\Gamma; \emptyset \vdash
-%  \Nat^\Phi \,F\,G}\rho$ is indeed an $\oCPO$.
-
-That each of the above interpretations is $\omega$-cocontinuous
-follows from Corollary~12 of~\cite{jp19} if we \new{APPROPRIATELY
-  RESTRICT THE SUBSCRIPTS OF $\Lan$s; we will require that $K$ is a
-  polynomial functor in the $\alpha$s that does not involve any
-  constant functors}. For $\Nat$-types, we note that $\oCPOsem{\Gamma;
-  \emptyset \vdash \Nat^\Phi F\,G}$ is an $\omega$-cocontinuous
-functor because, in accordance with
-Definition~\ref{def:set-sem-funcs}, it is constant on
-$\omega$-directed sets. Interpretations of $\Nat$-types ensure that
-$\oCPOsem{\Gamma \vdash F \to G}$ and $\oCPOsem{\Gamma \vdash \forall
-  \ol\alpha. F}$ are as expected in any parametric model.
-
-To make sense of the next-to-last clause in
-Definition~\ref{def:set-sem}, we need to know that, for each $\rho \in
-\oCPOenv$, $T^\oCPO_{H,\rho}$ is an $\omega$-cocontinuous
-endofunctor on $[\oCPO^k, \oCPO]_{\omega}$, and thus admits a
-fixpoint.  Since $T_{H,\rho}^\oCPO$ is defined in terms of
+\noindent
+Moreover, for any $\rho : \setenv$ we have that $\oCPOsem{\Gamma;
+  \emptyset \vdash \Nat^\Phi F\,G}\rho$ is an $\omega$-cocontinuous
+functor because it is a constant functor.  Interpretations of
+$\Nat$-types ensure that $\oCPOsem{\Gamma \vdash F \to G}\rho$ and
+$\oCPOsem{\Gamma \vdash \forall \ol\alpha. F}\rho$ are as expected in
+any parametric model.
+
+To make sense of the penultimate clause of
+Definition~\ref{def:set-sem}, we need to know that, for each $\rho :
+\oCPOenv$, $T^\oCPO_{H,\rho}$ is an $\omega$-cocontinuous endofunctor
+on $[\oCPO^k, \oCPO]$, and thus admits a fixpoint.  Since
+$T_{H,\rho}^\oCPO$ is defined in terms of
 $\oCPOsem{\Gamma;\ol\gamma,\phi, \ol{\alpha} \vdash H}$, this means
 that interpretations of types must be such functors, which in turn
-means that the actions of set interpretations of types on objects and
-on morphisms in $\oCPOenv$ are intertwined. Fortunately, we know
+means that the actions of set interpretations of types on the objects
+and morphisms in $\oCPOenv$ are intertwined. Fortunately, we know
 from~\cite{jp19} that, for every $\Gamma; \ol\alpha \vdash F$,
-$\oCPOsem{\Gamma; \Phi \vdash F}$ is actually in
-$[\oCPO^k,\oCPO]_{\omega}$, where $k = |\ol\alpha|$.  Therefore, for
-each $\oCPOsem{\Gamma; \ol\gamma, \phi^k, \ol{\alpha} \vdash H}$, the
-corresponding operator $T^\oCPO_{H}$ can be extended to a {\em
-  functor} from $\oCPOenv$ to
-$[[\oCPO^k,\oCPO]_{\omega},[\oCPO^k,\oCPO]_{\omega}]_{\omega}$.
-The action of $T^\oCPO_H$ on an object $\rho \in \oCPOenv$ is given by
-the higher-order functor $T_{H,\rho}^\oCPO$, whose actions on objects
-(functors in $[\oCPO^k, \oCPO]_{\omega}$) and on morphisms (natural
-transformations between such functors) are given in
-Definition~\ref{def:set-sem}. The action of $T^\oCPO_H$ on a morphism
-$f : \rho \to \rho'$ is the higher-order natural transformation
-$T^\oCPO_{H,f} : T^\oCPO_{H,\rho} \to T^\oCPO_{H,\rho'}$ whose action
-on $F : [\oCPO^k,\oCPO]_{\omega}$ is the natural transformation
-$T^\oCPO_{H,f}\, F : T^\oCPO_{H,\rho}\,F \to T^\oCPO_{H,\rho'}\,F$
-whose component at $\ol A$ is $(T^\oCPO_{H,f}\, F)_{\ol A} =
-\oCPOsem{\Gamma; \ol\gamma,\phi,\ol{\alpha} \vdash H}f[\phi :=
-  \id_F][\ol{\alpha := \id_A}]$.
-
-%For the last clause in Definition~\ref{def:set-sem} to be well-defined
-%we need $\oCPOsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]$
-%and $\oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha :=
-%    \_}]$ to be $\omega$-cocontinuous functors.
-In addition, for each $\ol K$, we have that $\textit{Lan}_{\ol K}$ is
-itself a (higher-order) functor. Specifically, given functors $F, F' :
-\C \to \D$, a sequence of functors $\ol K = K_1, ..., K_n$ with $K_i :
-\C \to \C_i$ for $i = 1,...,n$, and a natural transformation $\alpha :
-F \to F'$, the functorial action $\textit{Lan}_{\ol K} \alpha :
-\textit{Lan}_{\ol K} F \to \textit{Lan}_{\ol K} F'$ of
-$\mathit{Lan}_{\ol K}$ on $\alpha$ is defined to be the unique natural
-transformation such that $((\textit{Lan}_{\ol K} \alpha) \,\circ
-\,\langle K_1,...,K_n \rangle) \circ \eta_F = \eta_{F'} \,\circ\,
-\alpha$. Here, $\eta_F : F \to (\textit{Lan}_{\ol K} F) \circ \langle
-K_1,...,K_n \rangle$ and $\eta_{F'} : F' \to (\textit{Lan}_{\ol K} F')
-\circ \langle K_1,...,K_n \rangle$ are the natural transformations
-associated with the functors $\textit{Lan}_{\ol K} F$ and
-$\textit{Lan}_{\ol K} F'$ from $\Pi_{i \in \{1,...,n\}} \C_i$ to $\D$,
-respectively. It is not hard to see that $\mathit{Lan}_{\ol K}$ is a
-(higher-order) functor under this definition.
-
-The next definition uses the functors $T^\oCPO_H$ and $\mathit{Lan}_K$
-to define the actions of functors interpreting types on morphisms
-between set environments.
+$\oCPOsem{\Gamma; \Phi \vdash F}$ is actually in $[\oCPO^k,\oCPO]$,
+where $k = |\ol\alpha|$.  Therefore, for each $\oCPOsem{\Gamma;
+  \ol\gamma, \phi^k, \ol{\alpha} \vdash H}$, the corresponding
+operator $T^\oCPO_{H}$ can be extended to a {\em functor} from
+$\oCPOenv$ to $[[\oCPO^k,\oCPO],[\oCPO^k,\oCPO]]$.  The action of
+$T^\oCPO_H$ on an object $\rho : \oCPOenv$ is given by the
+higher-order functor $T_{H,\rho}^\oCPO$, whose actions on objects
+(i.e., functors in $[\oCPO^k, \oCPO]$) and on morphisms (natural
+transformations between such functors) are given in the penultimate
+clause of Definition~\ref{def:set-sem}. The action of $T^\oCPO_H$ on a
+morphism $f : \rho \to \rho'$ is the higher-order natural
+transformation $T^\oCPO_{H,f} : T^\oCPO_{H,\rho} \to
+T^\oCPO_{H,\rho'}$ whose action on $F : [\oCPO^k,\oCPO]$ is the
+natural transformation $T^\oCPO_{H,f}\, F : T^\oCPO_{H,\rho}\,F \to
+T^\oCPO_{H,\rho'}\,F$ whose component at $\ol A$ is given by
+$(T^\oCPO_{H,f}\, F)_{\ol A} = \oCPOsem{\Gamma;
+  \ol\gamma,\phi,\ol{\alpha} \vdash H}f[\phi := \id_F][\ol{\alpha :=
+    \id_A}]$.
+
+Finally, to make sense of the final clause
+Definition~\ref{def:set-sem}, we first observe that, for any $\rho :
+\setenv$, $\textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash
+      K}\rho[\ol{\alpha := \_}]}} \oCPOsem{\Gamma; \Phi, \ol\alpha
+  \vdash F}\rho[\ol{\alpha := \_}]$ is a functor from $\set^m$ to
+$\set$ for some $m$. This ensures that
+$(\textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash
+      K}\rho[\ol{\alpha := \_}]}} \oCPOsem{\Gamma; \Phi, \ol\alpha
+  \vdash F}\rho[\ol{\alpha := \_}]) \ol{\oCPOsem{\Gamma; \Phi \vdash
+    A}\rho}$ is a set.  Moreover, for any $\rho : \setenv$, the fact
+that $\oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}}
+  F)\ol{A}}\rho$ is an $\omega$-cocontinuous functor follows from
+Corollary~12 of~\cite{jp19} provided each functor in
+$\ol{\setsem{\Gamma; \ol\alpha \vdash K}}$ is polynomial in the
+variables in $\ol\alpha$. We also note that, for each vector of
+functors $\ol K$, $\textit{Lan}_{\ol K}$ is itself a (higher-order)
+functor. Specifically, given functors $F, F' : \C \to \D$, a sequence
+of functors $\ol K = K_1, ..., K_n$ with $K_i : \C \to \C_i$ for $i =
+1,...,n$, and a natural transformation $\alpha : F \to F'$, the
+functorial action $\textit{Lan}_{\ol K} \alpha : \textit{Lan}_{\ol K}
+F \to \textit{Lan}_{\ol K} F'$ of $\mathit{Lan}_{\ol K}$ on $\alpha$
+is defined to be the unique natural transformation such that
+$((\textit{Lan}_{\ol K} \alpha) \,\circ \,\langle K_1,...,K_n \rangle)
+\circ \eta_F = \eta_{F'} \,\circ\, \alpha$. Here, $\eta_F : F \to
+(\textit{Lan}_{\ol K} F) \circ \langle K_1,...,K_n \rangle$ and
+$\eta_{F'} : F' \to (\textit{Lan}_{\ol K} F') \circ \langle
+K_1,...,K_n \rangle$ are the natural transformations associated with
+the functors $\textit{Lan}_{\ol K} F$ and $\textit{Lan}_{\ol K} F'$
+from $\Pi_{i \in \{1,...,n\}} \C_i$ to $\D$, respectively. In $\set$,
+we have the slightly more general property
+\begin{equation}\label{eq:ho-functorial-action-lan}
+(\textit{Lan}_{\ol K}\, \alpha)\ol B \circ 
+  \iota_{X,g} = \iota'_{X,g} \circ \alpha_X
+\end{equation}
+Here, $\iota_{X,g}$ is the injection of $FX$ into $\colim{X :
+  \set_0,\,g : KX \to B}{FX}$ and $\iota'_{X,g}$ is the injection of
+$F'X$ into $\colim{X : \set_0, \,g : KX \to B}{F'X}$.
+
+The next definition uses the functors $T^\oCPO_H$ and
+$\mathit{Lan}_{\ol K}$ to define the actions of functors interpreting
+types on morphisms between set environments.
 
@@
 
-{\color{blue} Need to check that the extra condition in the interp of
-  Lans is preserved for the last clause to be well-defined.}
+To see that the functorial action on $\Lan$ types is well-defined, we
+need to check that it preserves the extra condition in the
+interpretation of $\Lan$ types. That is, we need to show that if $t :
+\oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}
+\rho$, then $\oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}}
+  F)\ol{A}}f t : \oCPOsem{\Gamma;\Phi \vdash
+  (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \rho'$.  By definition, if $t
+: \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}
+\rho$, then $(t, t) \in \oCPORsem{\Gamma;\Phi \vdash
+  (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \Eq_{\rho}$, i.e., there
+exist $\ol{Z : \oCPO_0}$, $t_1 : \oCPOsem{\Gamma; \Phi, \ol\alpha
+  \vdash F}\rho[\ol{\alpha := Z}]$, $t_2 : \oCPOsem{\Gamma; \Phi,
+  \ol\alpha \vdash F}\rho[\ol{\alpha := Z}]$, and $\ol{g :
+  \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\Eq_{\rho}[\ol{\alpha :=
+      \Eq_Z}] \to \oCPORsem{\Gamma; \Phi \vdash A}\Eq_{\rho}}$ such
+that $\iota_{\ol{Z}, \ol{\pi_1 g}}t_1 = t$, $\kappa_{\ol{Z}, \ol{\pi_2
+    g}}t_2 = t$, and $(t_1,t_2) : \oCPORsem{\Gamma; \Phi, \ol\alpha
+  \vdash F}\Eq_{\rho}[\ol{\alpha := \Eq_Z}]$, where $\iota$ and
+$\kappa$ are as in the final clause of Definition~\ref{def:rel-sem}
+below.  Now, to prove that $\oCPOsem{\Gamma;\Phi \vdash
+  (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f t : \oCPOsem{\Gamma;\Phi
+  \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \rho'$, let $\ol{Z' =
+  Z}$, let $t'_i = \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F} f
+[\ol{\alpha := \id_{Z}}] t_i$ for $i = 1, 2$, and let $\ol{g' =
+  \oCPORsem{\Gamma; \Phi \vdash A}\Eq_f \circ g}$.  We have to show
+that $\iota'_{\ol{Z'}, \ol{\pi_1 g'}}t'_1 = \oCPOsem{\Gamma;\Phi
+  \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f t$,
+$\kappa'_{\ol{Z'}, \ol{\pi_2 g'}}t'_2 = \oCPOsem{\Gamma;\Phi \vdash
+  (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f t$, and $(t_1',t_2') :
+\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\Eq_{\rho'}[\ol{\alpha :=
+    \Eq_{Z'}}]$, where $\iota'$ and $\kappa'$ are the injections for
+$\rho'$ corresponding to $\iota$ and $\kappa$ for $\rho$. The latter
+is immediate from the facts that $(t_1',t_2') = \oCPORsem{\Gamma;
+  \Phi, \ol\alpha \vdash F} \Eq_f [\ol{\alpha := \id_{\Eq_{Z}}}] (t_1,
+t_2)$ and $(t_1,t_2) : \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash
+  F}\Eq_{\rho}[\ol{\alpha := \Eq_Z}]$.  That $\iota'_{\ol{Z'},
+  \ol{\pi_1 g'}}t'_1 = \oCPOsem{\Gamma;\Phi \vdash
+  (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f t$ is because
+\[
+\begin{array}{rl}
+\iota'_{\ol{Z'}, \ol{\pi_1 g'}}t'_1
+&= \iota'_{\ol{Z'}, \ol{\pi_1 g'}} (\oCPOsem{\Gamma; \Phi, \ol\alpha
+  \vdash F} f [\ol{\alpha := \id_{Z}}] t_1) \\ 
+&= \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f
+(\iota_{\ol{Z}, \ol{\pi_1 g}} t_1) \\ 
+&= \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f
+t 
+\end{array}
+\]
+where the second equality here is justified by the commutativity of
+the following diagram, and $\iota''$ is the obvious injection
+indicated:
+\begin{equation}\label{cd:lan-comm}
+\begin{tikzcd}[row sep = huge]
+\oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := Z}]
+  \ar[r, "{\iota_{\ol{Z}, \ol{\pi_1 g}}}"]
+  \ar[d, "{\oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F} f [\ol{\alpha :=
+          \id_Z}]}" description] 
+&\colim{\ol{S}, \ol{h : \oCPOsem{\Gamma; \ol{\alpha} \vdash
+        K}\rho[\ol{\alpha := S}] \to \oCPOsem{\Gamma; \Phi \vdash
+        A}\rho}} {\oCPOsem{\Gamma; \Phi, \ol\alpha \vdash
+      F}\rho[\ol{\alpha := S}]} 
+  \ar[d, "{(\mathit{Lan}_{\ol{\oCPOsem{\Gamma; \ol{\alpha} \vdash
+            K}\rho[\ol{\alpha := \_}]}} \oCPOsem{\Gamma; \Phi,
+        \ol\alpha \vdash F}f[\ol{\alpha := \id_{\_}}])
+      \ol{\oCPOsem{\Gamma; \Phi \vdash A}\rho}}" description] \\ 
+\oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := Z}]
+  \ar[r, "{\iota''_{\ol{Z}, \ol{\pi_1 g}}}"]
+  \ar[d, equal]
+& \colim{\ol{S}, \ol{h : \oCPOsem{\Gamma; \ol{\alpha} \vdash
+        K}\rho'[\ol{\alpha := S}] \to \oCPOsem{\Gamma; \Phi \vdash
+        A}\rho}} 
+{\oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := S}]} 
+  \ar[d, "{ (\mathit{Lan}_{\ol{\oCPOsem{\Gamma; \ol{\alpha} \vdash
+            K}\rho[\ol{\alpha := \_}]}} \oCPOsem{\Gamma; \Phi,
+        \ol\alpha \vdash F}\rho'[\ol{\alpha := \_}])
+      \ol{\oCPOsem{\Gamma; \Phi \vdash A}f}}" description] \\ 
+\oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := Z}]
+  \ar[r, shift right = 1, "{\iota'_{\ol{Z'}, \ol{\pi_1
+          (\oCPORsem{\Gamma; \Phi \vdash A}\Eq_f \circ g)}}}"'] 
+& \colim{\ol{S}, \ol{h : \oCPOsem{\Gamma; \ol{\alpha} \vdash
+        K}\rho'[\ol{\alpha := S}] \to \oCPOsem{\Gamma; \Phi \vdash
+        A}\rho'}} 
+{\oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := S}]} 
+\end{tikzcd}
+\end{equation}
+Here, the top diagram commutes by the definition of the ``higher''
+functorial action of $\mathit{Lan}_{\ol{\oCPOsem{\Gamma; \ol{\alpha}
+      \vdash K}\rho[\ol{\alpha := \_}]}}$ on $\oCPOsem{\Gamma; \Phi,
+  \ol\alpha \vdash F}f[\ol{\alpha := \id_{\_}}]$ as given
+in~\eqref{eq:ho-functorial-action-lan}, and the bottom one commutes by
+the definition of the functorial action of
+$\mathit{Lan}_{\ol{\oCPOsem{\Gamma; \ol{\alpha} \vdash
+      K}\rho[\ol{\alpha := \_}]}} \oCPOsem{\Gamma; \Phi, \ol\alpha
+  \vdash F}\rho'[\ol{\alpha := \_}]$ on $\ol{\oCPOsem{\Gamma; \Phi
+    \vdash A}f}$ as given in~\eqref{def:functorial-action-lan}.  The
+proof that $\kappa_{Z,\pi_2 g'}t'_2 = \oCPOsem{\Gamma;\Phi \vdash
+  (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f t$ is analogous.
 
 Definitions~\ref{def:set-sem} and~\ref{def:set-sem-funcs} respect
-weakening, i.e., ensure that a type and its weakenings have the same
-set interpretations.
+weakening. That is, they ensure that a type and its weakenings all
+have the same set interpretations.
 
@@
 \begin{dfn}\label{def:rel-transf}
-A {\em $k$-ary $\oCPO$ relation transformer} $F$ is a triple $(F^1,
+A {\em $k$-ary relation transformer} $F$ is a triple $(F^1,
 F^2,F^*)$, where $F^1,F^2 : [\oCPO^k,\oCPO]_{\omega}$ and $F^* :
@@
 implies $(F^1 \ol{\alpha}\,c,F^2 \ol{\beta}\,d) \in F^*\ol{S}$. When
-convenient we identify a $0$-ary $\oCPO$ relation transformer
-$(A,B,R)$ with $R : \oCPOR(A,B)$. We may also write $\pi_1 F$ for
-$F^1$ and $\pi_2 F$ for $F^2$. {\color{blue} Without loss of
-  generality, we may assume that $\pi_1$ and $\pi_2$ are surjective on
-  relations.} We extend these conventions to $\oCPO$ relation
-environments, introduced in Definition~\ref{def:reln-env} below, in
-the obvious way.
+convenient we identify a $0$-ary relation transformer $(A,B,R)$ with
+$R : \oCPOR(A,B)$. We may also write $\pi_1 F$ for $F^1$ and $\pi_2 F$
+for $F^2$. Without loss of generality, we assume that $\pi_1$ and
+$\pi_2$ are surjective on relations. We extend these conventions to
+$\oCPO$ relation environments, introduced in
+Definition~\ref{def:reln-env} below, in the obvious way.
 
 \begin{dfn}
-The category $\oCPORT_k$ of $k$-ary $\oCPO$ relation transformers is
+The category $\oCPORT_k$ of $k$-ary relation transformers is
 given by the following data:
 \begin{itemize}
-\item An object of $\oCPORT_k$ is a $k$-ary $\oCPO$ relation transformer.
+\item An object of $\oCPORT_k$ is a $k$-ary relation transformer.
 \item A morphism $\delta : (G^1,G^2,G^*) \to (H^1,H^2,H^*)$ in
@@
 \begin{itemize}
-\item $H^1$ and $H^2$ are functors from $[\oCPO^k,\oCPO]_{\omega}$ to $[\oCPO^k,\oCPO]_{\omega}$
-\item $H^*$ is a functor from $\oCPORT_k$ to $[\oCPOR^k,\oCPOR]_{\omega}$
+\item $H^1$ and $H^2$ are functors from $[\oCPO^k,\oCPO]$ to $[\oCPO^k,\oCPO]$
+\item $H^*$ is a functor from $\oCPORT_k$ to $[\oCPOR^k,\oCPOR]$
 \item for all $\overline{R : \oCPOR(A,B)}$,
@@
 \end{dfn}
-Since the results of applying an endofunctor $H$ to $k$-ary $\oCPO$
+Since the results of applying an endofunctor $H$ to $k$-ary 
 relation transformers and morphisms between them must again be $k$-ary
-$\oCPO$ relation transformers and morphisms between them,
+relation transformers and morphisms between them,
 respectively, Definition~\ref{def:RT-functor} implicitly requires that
@@
 K^2$ are natural transformations between endofunctors on
-$[\oCPO^k,\oCPO]_{\omega}$ and the component of $\sigma$ at $F
-%= (F^1,F^2,F^*)
-\in \oCPORT_k$ is given by $\sigma_F = (\sigma^1_{F^1}, \sigma^2_{F^2})$.
+$[\oCPO^k,\oCPO]$ and the component of $\sigma$ at $F = (F^1,F^2,F^*)
+\in \oCPORT_k$ is $\sigma_F = (\sigma^1_{F^1}, \sigma^2_{F^2})$.
 \end{dfn}
@@
 Definition~\ref{def:RT-nat-trans} entails that $\sigma^i_{F^i}$ must
-be natural in $F^i : [\oCPO^k,\oCPO]_{\omega}$, and, for every $F$,
+be natural in $F^i : [\oCPO^k,\oCPO]$, and, for every $F$,
 $(\sigma^1_{F^1})_{\overline{A}}$ and
@@
 and $\ol B$, respectively.  Moreover, since the results of applying
-$\sigma$ to $k$-ary $\oCPO$ relation transformers must be morphisms of
-$k$-ary relation transformers, Definition~\ref{def:RT-nat-trans}
-implicitly requires that $(\sigma_F)_{\overline{R}} = (
+$\sigma$ to $k$-ary relation transformers must be morphisms of $k$-ary
+relation transformers, Definition~\ref{def:RT-nat-trans} implicitly
+requires that $(\sigma_F)_{\overline{R}} = (
 (\sigma^1_{F^1})_{\overline{A}}, (\sigma^2_{F^2})_{\overline{B}})$ is
@@
 
-Critically, we can compute $\omega$-directed colimits in
-$\oCPORT_k$: it is not hard to see that if $\cal D$ is an
-$\omega$-directed set then $\colim{d \in {\cal D}}{(F^1_d,
-  F^2_d,F^*_d)} = (\colim{d \in {\cal D}}{F^1_d}, \colim{d \in {\cal
-    D}}{F^2_d}, \colim{d \in {\cal D}}{F^*_d})$.  We define an
-endofunctor $T = (T^1,T^2,T^*)$ on $\oCPORT_k$ to be {\em
-  $\omega$-cocontinuous} if $T^1$ and $T^2$ are
-$\omega$-cocontinuous endofunctors on $[\oCPO^k,\oCPO]_{\omega}$
-and $T^*$ is an $\omega$-cocontinuous functor from $\oCPORT_k$ to
-$[\oCPOR^k,\oCPOR]_{\omega}$, i.e., is in
-$[\oCPORT_k,[\oCPOR^k,\oCPOR]_{\omega}]_{\omega}$.
+Critically, we can compute $\omega$-directed colimits in $\oCPORT_k$:
+it is not hard to see that if $\cal D$ is an $\omega$-directed set
+then $\colim{d \in {\cal D}}{(F^1_d, F^2_d,F^*_d)} = (\colim{d \in
+  {\cal D}}{F^1_d}, \colim{d \in {\cal D}}{F^2_d}, \colim{d \in {\cal
+    D}}{F^*_d})$.  We define an endofunctor $T = (T^1,T^2,T^*)$ on
+$\oCPORT_k$ to be {\em $\omega$-cocontinuous} if $T^1$ and $T^2$ are
+$\omega$-cocontinuous endofunctors on $[\oCPO^k,\oCPO]$ and $T^*$ is
+an $\omega$-cocontinuous functor from $\oCPORT_k$ to
+$[\oCPOR^k,\oCPOR]$, i.e., is in $[\oCPORT_k,[\oCPOR^k,\oCPOR]]$.
 
@@
 Observing that, for every $k$, $K^\oCPO_0$ is initial in
-$[\oCPO^k,\oCPO]_{\omega}$, and $K^\oCPOR_0$ is initial in
-$[\oCPOR^k,\oCPOR]_{\omega}$, we have that, for each $k$, $K_0 =
-(K^\oCPO_0,K^\oCPO_0,K^\oCPOR_0)$ is initial in $\oCPORT_k$. Thus, if
-$T = (T^1,T^2,T^*) : \oCPORT_k \to \oCPORT_k$ is an endofunctor on
-$\oCPORT_k$ then we can define the $\oCPO$ relation transformer $\mu
-T$ to be $\colim{i < \omega}{T^i K_0}$. It is not hard to see that
-$\mu T$ is given explicitly as
+$[\oCPO^k,\oCPO]$, and $K^\oCPOR_0$ is initial in $[\oCPOR^k,\oCPOR]$,
+we have that, for each $k$, $K_0 = (K^\oCPO_0,K^\oCPO_0,K^\oCPOR_0)$
+is initial in $\oCPORT_k$. Thus, if $T = (T^1,T^2,T^*) : \oCPORT_k \to
+\oCPORT_k$ is an endofunctor on $\oCPORT_k$ then we can define the
+relation transformer $\mu T$ to be $\colim{i < \omega}{T^i K_0}$. It
+is not hard to see that $\mu T$ is given explicitly as
 \begin{equation}\label{eq:mu}
@@
 \begin{lemma}\label{lem:fp}
-For any $T : [\oCPORT_k,\oCPORT_k]_{\omega}$, $\mu T \cong T(\mu T)$.
+For any $T : [\oCPORT_k,\oCPORT_k]$, $\mu T \cong T(\mu T)$.
 \end{lemma}
@@
 It is worth noting that the third component in Equation~(\ref{eq:mu})
-is the colimit in $[\oCPOR^k,\oCPOR]_{\omega}$ of third components
-of $\oCPO$ relation transformers, rather than a fixpoint of an
-endofunctor on $[\oCPO^k,\oCPO]_{\omega}$. That there is an
-asymmetry between the first two components of $\mu T$ and its third
-reflects the important conceptual observation that the third component
-of an endofunctor on $\oCPORT_k$ need not be a functor on all of
-$[\oCPOR^k,\oCPOR]_{\omega}$. In particular, although we can define
-$T_{H,\rho}\, F$ for an $\oCPO$ relation transformer $F$ in
-Definition~\ref{def:rel-sem} below, it is not clear how we could
-define it for an arbitrary $F : [\oCPOR^k,\oCPOR]_{\omega}$.
+is the colimit in $[\oCPOR^k,\oCPOR]$ of third components of relation
+transformers, rather than a fixpoint of an endofunctor on
+$[\oCPO^k,\oCPO]$. That there is an asymmetry between the first two
+components of $\mu T$ and its third reflects the important conceptual
+observation that the third component of an endofunctor on $\oCPORT_k$
+need not be a functor on all of $[\oCPOR^k,\oCPOR]$. In particular,
+although we can define $T_{H,\rho}\, F$ for a relation transformer $F$
+in Definition~\ref{def:rel-sem} below, it is not clear how we could
+define it for an arbitrary $F : [\oCPOR^k,\oCPOR]$.
 
@@
 An {\em $\oCPO$ relation environment} maps each type variable in
-$\tvars^k \cup \fvars^k$ to a $k$-ary $\oCPO$ relation
+$\tvars^k \cup \fvars^k$ to a $k$-ary relation
 transformer.  A morphism $f : \rho \to \rho'$ between $\oCPO$
@@
 functorial variable $\phi^k \in \fvars$ to a morphism from the $k$-ary
-$\oCPO$ relation transformer $\rho \phi$ to the $k$-ary
-$\oCPO$ relation transformer $\rho' \phi$. Composition of
+relation transformer $\rho \phi$ to the $k$-ary
+relation transformer $\rho' \phi$. Composition of
 morphisms on $\oCPO$ relation environments is given
@@
 
-When convenient we identify a $0$-ary $\oCPO$ relation transformer
-with the $\oCPO$ relation (transformer) that is its codomain and
-consider an $\oCPO$ relation environment to map a type variable of
-arity $0$ to an $\oCPO$ relation. If $\Phi =
-\{\phi_1^{k_1},...,\phi_n^{k_n}\}$ and $\ol K = \{K_1,...,K_n\}$ where
-$K_i : [\oCPOR^{k_i},\oCPOR]_{\omega}$ for $i = 1,...,n$, then we
-write either $\rho[\ol{\Phi := K}]$ or $\rho[\ol{\phi := K}]$ for the
-$\oCPO$ relation environment $\rho'$ such that $\rho' \phi_i = K_i$
-for $i = 1,...,n$ and $\rho' \phi = \rho \phi$ if $\phi \not \in
-\Phi$. If $\rho$ is an $\oCPO$ relation environment, we write $\pi_1
-\rho$ and $\pi_2 \rho$ for the $\oCPO$ relation environments mapping
-each type variable $\phi$ to the functors $(\rho\phi)^1$ and
-$(\rho\phi)^2$, respectively.
+When convenient we identify a $0$-ary relation transformer with the
+$\oCPO$ relation (transformer) that is its codomain and consider an
+$\oCPO$ relation environment to map a type variable of arity $0$ to an
+$\oCPO$ relation. If $\Phi = \{\phi_1^{k_1},...,\phi_n^{k_n}\}$ and
+$\ol K = \{K_1,...,K_n\}$ where $K_i : [\oCPOR^{k_i},\oCPOR]$ for $i =
+1,...,n$, then we write either $\rho[\ol{\Phi := K}]$ or
+$\rho[\ol{\phi := K}]$ for the $\oCPO$ relation environment $\rho'$
+such that $\rho' \phi_i = K_i$ for $i = 1,...,n$ and $\rho' \phi =
+\rho \phi$ if $\phi \not \in \Phi$. If $\rho$ is an $\oCPO$ relation
+environment, we write $\pi_1 \rho$ and $\pi_2 \rho$ for the $\oCPO$
+relation environments mapping each type variable $\phi$ to the
+functors $(\rho\phi)^1$ and $(\rho\phi)^2$, respectively.
 
@@
 \begin{dfn}\label{def:relenv-functor}
-A functor $H : [\oCPORenv, \oCPORT_k]_{\omega}$ is a triple $H =
+A functor $H : [\oCPORenv, \oCPORT_k]$ is a triple $H =
 (H^1,H^2,H^*)$, where
@@
 \item $H^1$ and $H^2$ are objects in
-  $[\oCPOenv,[\oCPO^k,\oCPO]_{\omega}]_{\omega}$
-\item $H^*$ is a an object in
-  $[\oCPORenv,[\oCPOR^k,\oCPOR]_{\omega}]_{\omega}$
+  $[\oCPOenv,[\oCPO^k,\oCPO]]$
+\item $H^*$ is a an object in $[\oCPORenv,[\oCPOR^k,\oCPOR]]$
 \item for all $\overline{R : \oCPOR(A,B)}$ and morphisms $f$ in
@@
 \begin{dfn}\label{def:rel-sem}
-The {\em relational interpretation}
-$\oCPORsem{\cdot} : \F \to [\oCPORenv, \oCPOR]_{\omega}$ is defined by
+The {\em relational interpretation} $\oCPORsem{\cdot} : \F \to
+[\oCPORenv, \oCPOR]$ is defined by
 \begin{align*}
@@
     F)\ol{A}}\rho &= \{ (t_1, t_2) : \oCPOsem{\Gamma;\Phi \vdash
-    (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}\pi_1 \rho \,\times\,
+    (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}(\pi_1 \rho) \,\times\,
   \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}}
-    F)\ol{A}}\pi_2 \rho\\
+    F)\ol{A}}(\pi_2 \rho)\\
   &\hspace{0.2in}|~\exists\, \ol{Z : \oCPO_0}, t_1' : \oCPOsem{\Gamma;
-    \ol{\alpha} \vdash K}\pi_1\rho[\ol{\alpha := Z}],\, t_2' :
-  \oCPOsem{\Gamma; \ol{\alpha} \vdash K}\pi_2\rho[\ol{\alpha := Z}],\\
-  & \hspace{0.2in} f :
+    \Phi, \ol\alpha \vdash F}(\pi_1\rho)[\ol{\alpha := Z}],\, t_2' : 
+  \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}(\pi_2\rho)[\ol{\alpha := Z}],\\
+  & \hspace{0.2in} \ol{f :
   \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho[\ol{\alpha := \Eq_Z}] \to
-  \oCPORsem{\Gamma; \Phi \vdash A}\rho\\
-  &\hspace{0.2in}\mbox{such that } \iota_{Z,\pi_1 f}t_1' = t_1,\,
-  \kappa_{Z,\pi_2 f}t_2' = t_2,\\
-  &  \hspace*{0.2in}\mbox{and } (t_1',t_2') :
+  \oCPORsem{\Gamma; \Phi \vdash A}\rho} \mbox{ such that }\\
+  &\hspace{0.2in}\iota_{\ol{Z},\ol{\pi_1 f}}t_1' = t_1,\,
+  \kappa_{\ol{Z}, \ol{\pi_2 f}}t_2' = t_2, \mbox{ and } (t_1',t_2') :
   \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := \Eq_Z}]\,\}
@@
 \noindent
-{\color{blue} In the final clause of Definition~\ref{def:rel-sem},
+In the final clause of Definition~\ref{def:rel-sem},
   $\iota_{Z,h}$ is the injection into
 \[\colim{Z \,: \,\oCPO_0,\; h \,:\, \oCPOsem{\Gamma; \ol{\alpha} \vdash
-      K}\pi_1\rho[\ol{\alpha := Z}] \to \oCPOsem{\Gamma; \Phi \vdash
-      A}\pi_1\rho}{\oCPOsem{\Gamma; \ol{\alpha} \vdash
-    F}\pi_1\rho[\ol{\alpha := Z}]}\]
+      K}(\pi_1\rho)[\ol{\alpha := Z}] \to \oCPOsem{\Gamma; \Phi \vdash
+      A}(\pi_1\rho)}{\oCPOsem{\Gamma; \Phi, \ol{\alpha} \vdash
+    F}(\pi_1\rho)[\ol{\alpha := Z}]}\]
   and 
@@
 \[\colim{Z\, :\, \oCPO_0, \;h \,:\, \oCPOsem{\Gamma; \ol{\alpha} \vdash
-      K}\pi_2\rho[\ol{\alpha := Z}] \to \oCPOsem{\Gamma; \Phi \vdash
-    A}\pi_2\rho}{\oCPOsem{\Gamma; \ol{\alpha} \vdash
-    F}\pi_2\rho[\ol{\alpha := Z}]}\]}
-
-For Definition~\ref{def:rel-sem} to be well-defined, we have to check
-that each relational interpretation is in $\oCPOR$ and, in particular,
-that each relates sups of pairwise related $\omega$-chains. This will
-be proved by induction on types, and in most cases it will follow from
-the induction hypotheses. However, well-definedness needs to be proved
-directly for relational interpretations of $\Nat$-types.
-
-The proof that relational interpretations of $\Nat$-types define sets
-is analogous to the proof of Lemma~\ref{lem:nat-types-sets}.
-
-%Next, we
-%observe that $\oCPORsem{\Gamma; \emptyset \vdash \Nat^\Phi F\,G} \rho$
-%is in $\oCPOR$. It is indeed a relation between $\oCPO$s, and it
-%relates sups of pairwise related $\omega$-chains of natural
-%transformations because their sups are computed pointwise. More
-%specifically, if $(t_i, t_i') \colon \oCPORsem{\Gamma; \emptyset
-%  \vdash \Nat^\Phi F\,G} \rho$ for all $i < \omega$, then for every
-%$\ol{K = (K^1, K^2, K^*) : \oCPORT_{k}}$ and every $i < \omega$ we
-%have
-%  \[\big((t_i)_{\ol{K^1}},\, (t'_i)_{\ol{K^2}}\big) \in
-%  (\oCPORsem{\Gamma; \Phi \vdash G}\rho[\ol{\Phi :=
-%      K}])^{\oCPORsem{\Gamma;\Phi\vdash F}\rho[\ol{\Phi := K}]}\] So
-%  if $(a,b) \in \oCPORsem{\Gamma;\Phi\vdash F}\rho[\ol{\Phi := K}]$,
-%  then $\big((t_i)_{\ol{K^1}}\,a,\, (t'_i)_{\ol{K^2}}\,b\big) \in
-%  \oCPORsem{\Gamma;\Phi\vdash G}\rho[\ol{\Phi := K}]$ for all $i <
-%  \omega$. But then since $\big((t_i)_{\ol{K^1}} a\big)_{i < \omega}$
-%  is an $\omega$-chain in $\oCPOsem{\Gamma; \Phi \vdash
-%    G}\rho[\ol{\Phi := K^1}]$ and $\big((t'_i)_{\ol{K^2}}\big)_{i <
-%    \omega} b$ is an $\omega$-chain in $\oCPOsem{\Gamma; \Phi \vdash
-%    G}\rho[\ol{\Phi := K^2}]$, not only are $\bigvee_{i < \omega}
-%  ((t_i)_{\ol{K^1}}\,a)$ and $\bigvee_{i < \omega}
-%  ((t'_i)_{\ol{K^2}}\,b)$ well-defined, but, since
-%\[\oCPORsem{\Gamma;\Phi\vdash G}\rho[\ol{\Phi := K}] : \oCPOR(
-%  \oCPOsem{\Gamma;\Phi\vdash G}\rho[\ol{\Phi := K^1}],
-%  \oCPOsem{\Gamma;\Phi\vdash G}\rho[\ol{\Phi := K^2}])\]
-%we have
-%  \[
-%  \Big(\big(\bigvee_{i < \omega} t_i\big)_{\ol{K^1}} \,a,
-%  \big(\bigvee_{i < \omega} t'_i\big)_{\ol{K^2}} \,b\Big)
-%  = \Big(\bigvee_{i < \omega} \big((t_i)_{\ol{K^1}}\,a\big),\,
-%  \bigvee_{i < \omega} \big((t'_i)_{\ol{K^2}}\,b\big)\Big) \in
-%          \oCPORsem{\Gamma;\Phi\vdash G}\rho[\ol{\Phi
-%             := K}]\] as well. That is,
-%\[
-%  \Big(\big(\bigvee_{i < \omega} t_i\big)_{\ol{K^1}}, \big(\bigvee_{i
-%    < \omega} t'_i\big)_{\ol{K^2}} \Big)
-%  \in (\oCPORsem{\Gamma;
-%    \Phi \vdash G}\rho[\ol{\Phi := K}])^{\oCPORsem{\Gamma;\Phi\vdash
-%      F}\rho[\ol{\Phi := K}]}
-%  \]
-%i.e.,
-%$( \bigvee_{i < \omega} t_i, \bigvee_{i < \omega} t'_i ) \in
-%  \oCPORsem{\Gamma; \emptyset \vdash \Nat^\Phi F\,G} \rho$.
-
-Moreover,
-%as for $\oCPO$ interpretations,
-$\omega$-cocontinuity of
-each of the above interpretations follows from Corollary~12
-of~\cite{jp19} if we \new{APPROPRIATELY RESTRICT THE SUBSCRIPTS of
-$\Lan$s}. For $\Nat$-types, $\oCPORsem{\Gamma; \emptyset \vdash
-  \Nat^\Phi F\,G}$ is an $\omega$-cocontinuous functor because it is
-constant on $\omega$-directed sets.  Interpretations of $\Nat$-types
-ensure that $\oCPORsem{\Gamma \vdash F \to G}$ and $\oCPORsem{\Gamma
-  \vdash \forall \ol \alpha. F}$ are as expected in any parametric
-model.
-
-For the next-to-last clause in Definition~\ref{def:rel-sem} to be
-well-defined we need $T_{H,\rho}$ to be an $\omega$-cocontinuous
-endofunctor on $\oCPORT$ so that, by Lemma~\ref{lem:fp}, it admits a
-fixpoint.  Since $T_{H,\rho}$ is defined in terms of
-$\oCPORsem{\Gamma;\ol\gamma,\phi^k, \ol{\alpha} \vdash H}$, this means
-that relational interpretations of types must be
-$\omega$-cocontinuous functors from $\oCPORenv$ to $\oCPORT_0$,
-which in turn entails that the actions of relational interpretations
-of types on objects and on morphisms in $\oCPORenv$ are intertwined.
+      K}(\pi_2\rho)[\ol{\alpha := Z}] \to \oCPOsem{\Gamma; \Phi \vdash
+    A}(\pi_2\rho)}{\oCPOsem{\Gamma; \Phi, \ol{\alpha} \vdash
+    F}(\pi_2\rho)[\ol{\alpha := Z}]}\]
+We note that the pair $(\iota_{Z,h},\kappa_{Z,h})$ need not coincide
+with the injection $j_{Z,\Eq_h}$ into 
+\[\colim{Z\, :\, \oCPO_0, \;g \,:\, \oCPORsem{\Gamma; \ol{\alpha} \vdash
+      K}\rho[\ol{\alpha := \Eq_Z}] \to \oCPORsem{\Gamma; \Phi \vdash
+    A}\rho}{\oCPORsem{\Gamma; \Phi, \ol{\alpha} \vdash
+    F}\rho[\ol{\alpha := \Eq_Z}]}\] In fact, it is precisely this that
+keeps the object and relational interpretations of $\Lan$ types in
+from being ``parallel''. {\color{blue} Of course $\iota$ and $\kappa$
+  depend on $\rho$, too. Perhaps the notation should reflect this.}
+
+If $\rho : \oCPORenv$ and $\vdash F$, then we write
+$\oCPORsem{\vdash F}$ instead of $\oCPORsem{\vdash F}\rho$. The
+interpretations in Definitions~\ref{def:rel-sem} and in
+Definition~\ref{def:rel-sem-funcs} below respect weakening.
+
+As for Definition~\ref{def:set-sem}, to know that
+Definition~\ref{def:rel-sem} is well-defined we have to check that,
+for every $\rho : \relenv$, each relational interpretation is in
+$\oCPOR$. The proof that relational interpretations of $\Nat$-types
+define relations is analogous to the proof of
+Lemma~\ref{lem:nat-types-sets}.  Moreover, for any $\rho : \relenv$,
+we have that $\oCPORsem{\Gamma; \emptyset \vdash \Nat^\Phi F\,G}\rho$
+is an $\omega$-cocontinuous functor because it is a constant functor.
+Interpretations of $\Nat$-types ensure that $\oCPORsem{\Gamma \vdash F
+  \to G}\rho$ and $\oCPORsem{\Gamma \vdash \forall \ol \alpha. F}\rho$
+are as expected in any parametric model.
+
+To make sense of the penultimate clause of
+Definition~\ref{def:rel-sem}, we need to know that, for each $\rho :
+\relenv$, $T_{H,\rho}$ is an $\omega$-cocontinuous endofunctor on
+$\oCPORT$, and thus admits a fixpoint. Since $T_{H,\rho}$ is defined
+in terms of $\oCPORsem{\Gamma;\ol\gamma,\phi^k, \ol{\alpha} \vdash
+  H}$, this means that relational interpretations of types must be
+$\omega$-cocontinuous functors from $\oCPORenv$ to $\oCPORT_0$, which
+in turn entails that the actions of relational interpretations of
+types on objects and on morphisms in $\oCPORenv$ are intertwined.
 %Moreover, for the last clause in Definition~\ref{def:rel-sem} to be
@@
   \vdash F}$ is actually in $[\oCPOR^k,\oCPOR]_{\omega}$, where $k =
-|\ol\alpha|$. We first define the actions of each of these functors
-on morphisms between $\oCPO$ relation environments in
-Definition~\ref{def:rel-sem-funcs}, and then argue that the functors
-given by Definitions~\ref{def:rel-sem} and~\ref{def:rel-sem-funcs} are
-well-defined and have the required properties. To do this, we extend
-$T_H$ to a {\em functor} from $\oCPORenv$ to
-$[[\oCPOR^k,\oCPOR]_{\omega},
-  [\oCPOR^k,\oCPOR]_{\omega}]_{\omega}$.  Its action on an object
-$\rho \in \oCPORenv$ is given by the higher-order functor $T_{H,\rho}$
-whose actions on objects and morphisms are given in
-Definition~\ref{def:rel-sem}. Its action on a morphism $f : \rho
-\to \rho'$ is the higher-order natural transformation $T_{H,f} :
-T_{H,\rho} \to T_{H,\rho'}$ whose action on any $F :
-[\oCPOR^k,\oCPOR]_{\omega}$ is the higher-order natural
-transformation $T_{H,f}\, F : T_{H,\rho}\, F \to T_{H,\rho'}\, F$
-whose component at $\ol{R}$ is $(T_{H,f}\, F)_{\ol{R}} =
-\oCPORsem{\Gamma; \ol\gamma,\phi,\ol{\alpha} \vdash H}f[\phi :=
-  \id_F][\ol{\alpha := \id_R}]$. The next definition uses $T_H$ to
-define the actions of functors interpreting types on morphisms between
-$\oCPO$ relation environments.
-
-If $\rho \in \oCPORenv$ and $\vdash F$, then we write
-$\oCPORsem{\vdash F}$ instead of $\oCPORsem{\vdash F}\rho$. The
-interpretations in Definitions~\ref{def:rel-sem} and in
-Definition~\ref{def:rel-sem-funcs} below respect weakening.
+|\ol\alpha|$. Therefore, for each $\oCPORsem{\Gamma;\ol\gamma,\phi^k,
+  \ol{\alpha} \vdash H}$, the corresponding operator $T_H$ can be
+extended to a {\em functor} from $\oCPORenv$ to $[[\oCPOR^k,\oCPOR],
+  [\oCPOR^k,\oCPOR]]$. The action of $T_H$ on an object $\rho :
+\oCPORenv$ is given by the higher-order functor $T_{H,\rho}$ whose
+actions on objects and morphisms are given in the penultimate clause
+of Definition~\ref{def:rel-sem}. The action of $T_{H}$ on a morphism
+$f : \rho \to \rho'$ is the higher-order natural transformation
+$T_{H,f} : T_{H,\rho} \to T_{H,\rho'}$ whose action on any $F :
+[\oCPOR^k,\oCPOR]_{\omega}$ is the higher-order natural transformation
+$T_{H,f}\, F : T_{H,\rho}\, F \to T_{H,\rho'}\, F$ whose component at
+$\ol{R}$ is $(T_{H,f}\, F)_{\ol{R}} = \oCPORsem{\Gamma;
+  \ol\gamma,\phi,\ol{\alpha} \vdash H}f[\phi := \id_F][\ol{\alpha :=
+    \id_R}]$. Finally, to see that the functor given by the
+penultimate clause of Definition~\ref{def:rel-sem} is well-defined we
+must also show that, for every $H$, $T_{H,\rho}\,F$ is a relation
+transformer for any relation transformer $F$. To know that the functor
+given by Definition~\ref{def:rel-sem-funcs} is well-defined we will
+similarly need to show and that $T_{H,f}\, F : T_{H,\rho}\, F \to
+T_{H,\rho'}\, F$ is a morphism of relation transformers for every
+relation transformer $F$ and every morphism $f : \rho \to \rho'$ in
+$\oCPORenv$. These results will be immediate consequences of
+Lemma~\ref{lem:rel-transf-morph} below.
+
+To make sense of the final clause of
+Definition~\ref{def:rel-sem}, we first observe that, for any $\rho :
+\relenv$, $\oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}}
+  F)\ol{A}}\rho$ is a set because $\oCPOsem{\Gamma;\Phi \vdash
+  (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}(\pi_1 \rho)$ and
+$\oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}}
+  F)\ol{A}}(\pi_2 \rho)$ are.  Moreover, for any $\rho : \relenv$, the
+fact that $\oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}}
+  F)\ol{A}}\rho$ is an $\omega$-cocontinuous functor follows from
+Corollary~12 of~\cite{jp19} provided each functor in
+$\ol{\setsem{\Gamma; \ol\alpha \vdash K}}$ is polynomial in the
+variables in $\ol\alpha$.
+
+The next definition gives the actions of functors interpreting types
+on morphisms between relation environments.
 
 \begin{dfn}\label{def:rel-sem-funcs}
-Let $f: \rho \to \rho'$ for $\oCPO$ relation environments $\rho$ and
-$\rho'$ (so that $\rho|_\tvars = \rho'|_\tvars$). The action
+Let $f: \rho \to \rho'$ for relation environments $\rho$ and $\rho'$
+(so that $\rho|_\tvars = \rho'|_\tvars$). The actions
 $\oCPORsem{\Gamma;\Phi \vdash F}f$ of $\oCPORsem{\Gamma;\Phi \vdash
-  F}$ on the morphism $f$ is given exactly as in
-Definition~\ref{def:set-sem-funcs}, except that all interpretations
-are $\oCPO$ relational interpretations and all occurrences of
-$T^\oCPO_{H,f}$ are replaced by $T_{H,f}$.
-
-{\color{blue} This is not true for the relational interps of
-  Lans. Define explicitly.}
-
+  F}$ on the morphism $f$ for all types other than $\Lan$-types are
+given exactly as in Definition~\ref{def:set-sem-funcs}, except that
+all interpretations involved are now relational interpretations and
+all occurrences of $T^\oCPO_{H,f}$ are replaced by $T_{H,f}$. For
+$\Lan$-types, we define
+\[ \oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} f
+= \big( \oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}}
+  F)\ol{A}} (\pi_1 f),\, \oCPORsem{\Gamma;\Phi \vdash
+  (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} (\pi_2 f) \big)\]
 \end{dfn}
 
-To see that the functors given by Definitions~\ref{def:rel-sem}
-and~\ref{def:rel-sem-funcs} are well-defined we must show that, for
-every $H$, $T_{H,\rho}\,F$ is an $\oCPO$ relation transformer for any
-$\oCPO$ relation transformer $F$, and that $T_{H,f}\, F : T_{H,\rho}\,
-F \to T_{H,\rho'}\, F$ is a morphism of $\oCPO$ relation transformers
-for every $\oCPO$ relation transformer $F$ and every morphism $f :
-\rho \to \rho'$ in $\oCPORenv$. This is an immediate consequence of
-the following Lemma.
+We now need to show that the functorial actions of $\Lan$-types are
+well-defined, i.e., that for every $(t_1, t_2) : \oCPORsem{\Gamma;\Phi
+  \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}\rho$ we have that
+\begin{equation}\label{eq:lan-rel-interp-well-defined}
+  \big( \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}
+(\pi_1 f) t_1, \oCPOsem{\Gamma;\Phi \vdash
+  (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} (\pi_2 f) t_2 \big) :
+\oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}
+\rho'
+\end{equation}
+Since $(t_1, t_2) : \oCPORsem{\Gamma;\Phi \vdash
+  (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}$ there exist $\ol{Z :
+  \oCPO_0}$, $t_1' : \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash
+  F}(\pi_1\rho)[\ol{\alpha := Z}]$, $t_2' : \oCPOsem{\Gamma; \Phi,
+  \ol\alpha \vdash F}(\pi_2\rho)[\ol{\alpha := Z}]$, $\ol{g :
+  \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho[\ol{\alpha := \Eq_Z}]
+  \to \oCPORsem{\Gamma; \Phi \vdash A}\rho}$ such that
+$\iota_{\ol{Z},\ol{\pi_1 g}}t_1' = t_1$, $\kappa_{\ol{Z}, \ol{\pi_2
+    g}}t_2' = t_2$, and $(t_1',t_2') : \oCPORsem{\Gamma; \Phi,
+  \ol\alpha \vdash F}\rho[\ol{\alpha := \Eq_Z}]$.  To show that
+\eqref{eq:lan-rel-interp-well-defined} holds, let $\ol{S = Z}$, $s_1'
+= \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F} (\pi_1 f) [\ol{\alpha :=
+    Z}] t'_1$, $s_2' = \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}
+(\pi_2 f) [\ol{\alpha := Z}] t'_2$, and $\ol{h = \oCPORsem{\Gamma;
+    \Phi \vdash A} f \circ g}$. Then $(s'_1, s'_2) : \oCPORsem{\Gamma;
+  \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := \Eq_Z}]$ because
+$(t_1',t_2') : \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash
+  F}\rho[\ol{\alpha := \Eq_Z}]$ and $(s'_1, s'_2) = \oCPORsem{\Gamma;
+  \Phi, \ol\alpha \vdash F} f [\ol{\alpha := \Eq_Z}] (t'_1, t'_2)$.
+In addition, $\iota'_{\ol{Z},\ol{\pi_1 h}} s_1' = \oCPOsem{\Gamma;\Phi
+  \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} (\pi_1 f) t_1$ and
+$\kappa'_{\ol{Z},\ol{\pi_2 h}} s_2' = \oCPOsem{\Gamma;\Phi \vdash
+  (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} (\pi_2 f) t_2$ hold because
+\[
+\begin{array}{rl}
+\iota'_{\ol{Z},\ol{\pi_1 h}} s_1' &=
+\iota'_{\ol{Z},\ol{\oCPOsem{\Gamma; \Phi \vdash A} (\pi_1 f) \circ \pi_1
+    g}} ( \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F} (\pi_1 f)
+      [\ol{\alpha := \id_Z}] t'_1 ) \\
+%      &= (\oCPOsem{\Gamma;\Phi \vdash
+%        (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} (\pi_1 f) \circ
+      %      \iota_{\ol{Z},\ol{\pi_1 g}}) t'_1
+       &= \oCPOsem{\Gamma;\Phi
+        \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} (\pi_1 f)
+      (\iota_{\ol{Z},\ol{\pi_1 g}} t'_1) \\ &= \oCPOsem{\Gamma;\Phi
+        \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} (\pi_1 f)
+      t_1 \end{array}
+\]
+where the second equality is by~\eqref{cd:lan-comm}, and similarly
+for $\kappa'_{\ol{Z},\ol{\pi_2 h}} s_2' = \oCPOsem{\Gamma;\Phi \vdash
+  (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} (\pi_2 f) t_2$.
+
+With the relational interpretations of types laid out we can now
+prove:
 
 \begin{lemma}\label{lem:rel-transf-morph}
-For every $\Gamma;\Phi \vdash F$, $\sem{\Gamma;\Phi \vdash F} =
+For every $\Gamma;\Phi \vdash F$, \[\sem{\Gamma;\Phi \vdash F} =
 (\oCPOsem{\Gamma;\Phi \vdash F}, \oCPOsem{\Gamma;\Phi \vdash
   F},\oCPORsem{\Gamma;\Phi \vdash F}) \in
-[\oCPORenv,\oCPORT_0]_{\omega}$.
+[\oCPORenv,\oCPORT_0]\]
 \end{lemma}
@@
 together with Lemma~\ref{lem:fp} and Equation~\ref{eq:mu} for
-$\mu$-types. \new{$\Lan$ types need restriction on their
-  subscripts.}
+$\mu$-types. The result holds by construction for $\Lan$-types.
 
@@
 \vdash H$, and $\ol{\Gamma;\Phi \vdash K}$, then
+{\color{red} WE NEED TO CHECK LAN CASES HERE!}
 \begin{gather}
@@
   \vdash F[\phi :== \psi]} f\\
-%\end{gather}
-%Moreover,
-%\begin{lemma}\label{lem:demotion-objects}
-%\begin{gather}
-%\label{thm:demotion-objects}
-%\setsem{\Gamma; \Phi, \phi \vdash F} \rho = \setsem{\Gamma, \psi; \Phi
-%  \vdash F[\phi :== \psi] } \rho{\color{blue}[\psi := \rho \phi]}\\
-%\label{thm:demotion-morphisms}
-%\setsem{\Gamma; \Phi, \phi \vdash F} f = \setsem{\Gamma, \psi; \Phi
-%  \vdash F[\phi :== \psi]} f{\color{blue}[\psi \mapsto f\phi]}\\
 \label{eq:subs-var}
@@
 Identities analogous to (\ref{thm:demotion-objects}) through
-(\ref{eq:subs-const-morph}) hold for $\oCPO$ relational
-interpretations as well.
+(\ref{eq:subs-const-morph}) hold for relational interpretations as
+well.
 
@@
 associate to each natural transformation between $k$-ary functors on
-$\oCPO$ a $k$-ary $\oCPO$ relation transformer as follows.
-
-First, let $f : A \to B$ be a morphism in $\oCPO$. Then the graph of
-the underlying function $|f| : |A| \to |B|$ in $\set$ is an $\oCPO$
-relation.  Indeed, if $(a_i, b_i) : \graph{|f|}$ for all $i < \omega$
-is a chain, then $|f| a_i = b_i$ for all $i$, and consequently
-$|f|(\bigvee_{i < \omega} a_i) = \bigvee_{i < \omega} (|f| a_i) =
-\bigvee_{i < \omega} b_i$, i.e., $(\bigvee_{i < \omega} a_i,
-\bigvee_{i < \omega} b_i) : \graph{|f|}$. We therefore define the {\em
-  $\oCPO$ graph of $f$} in $\oCPO$ to be $\graph{f} =
-\graph{|f|}$. Note in particular that if $A : \oCPO$ then
-$\graph{\id_A}$ is an $\oCPO$ relation. We denote this $\oCPO$
-relation, called the {\em equality relation on $A$}, by $\Eq_A$. It
-coincides exactly with the equality relation on the underlying set
-$|A|$ of $A$.
-
-The notion of an $\oCPO$ graph relation naturally generalizes to
-associate to each natural transformation between $k$-ary functors on
-$\oCPO$ a $k$-ary $\oCPO$ relation transformer as follows. Recall
-that, since $\oCPO$ is a locally $\omega$-presentable category,
-Proposition 1.6.1 of~\cite{ar94} ensures that it has a (strong epi,
-mono) factorization system. We then have:
+$\oCPO$ a $k$-ary relation transformer as follows.
+
+The notion of a graph relation naturally generalizes to associate to
+each natural transformation between $k$-ary functors on $\oCPO$ a
+$k$-ary relation transformer as follows. Recall that, since $\oCPO$ is
+a locally finitely presentable category, Proposition 1.6.1
+of~\cite{ar94} ensures that it has a (strong epi, mono) factorization
+system. We then have:
 
@@
 transformer $\graph{\alpha} = (F, G, \graph{\alpha}^*)$. We call this
-the {\em $\oCPO$ graph relation transformer} for $\alpha$.
+the {\em graph relation transformer} for $\alpha$.
 
 \begin{lemma}\label{lem:graph-reln-functors}
-If $F,G : [\oCPO^k,\oCPO]_{\omega}$, and if $\alpha : F \to G$ is a natural
+If $F,G : [\oCPO^k,\oCPO]$, and if $\alpha : F \to G$ is a natural
 transformation, then $\graph{\alpha}$ is in $\oCPORT_k$.
@@
 Clearly, $\graph{\alpha}^*$ is $\omega$-cocontinuous, so
-$\graph{\alpha}^* : [\oCPOR^k,\oCPOR]_{\omega}$. Now, suppose
-$\overline{R : \oCPOR(A, B)}$, $\overline{S : \oCPOR(C, D)}$, and
+$\graph{\alpha}^* : [\oCPOR^k,\oCPOR]$. Now, let $\overline{R :
+  \oCPOR(A, B)}$,\, $\overline{S : \oCPOR(C, D)}$, and
 $\overline{(\beta, \beta') : R \to S}$. We want to show that there
@@
 If $f : A \to B$ is a morphism in $\oCPO$ then the definition of the
-$\oCPO$ graph relation transformer $\graph{f}$ for $f$ as a natural
+graph relation transformer $\graph{f}$ for $f$ as a natural
 transformation between $0$-ary functors $A$ and $B$ coincides with the
 definition of $\graph{f}$ for $f$ as a morphism in $\oCPO$ given in
-the second paragraph of this section.  As a result, $\oCPO$ graph
-relation transformers are a reasonable extension of $\oCPO$ graph
-relations to functors.
-
-To prove the IEL, we will need to know that the equality $\oCPO$
-relation transformer preserves equality relations in $\oCPOR$; see
+the second paragraph of this section.  As a result, graph relation
+transformers are a reasonable extension of graph relations to
+functors.
+
+To prove the IEL, we will need to know that the equality relation
+transformer preserves equality relations in $\oCPOR$; see
 Equation~\ref{eq:eq-pres-eq} below. This will follow from the next
-lemma, which shows how to compute the action of an $\oCPO$ graph
-relation transformer on any $\oCPO$ graph relation.
+lemma, which shows how to compute the action of a graph relation
+transformer on any graph relation.
 
 \begin{lemma}\label{lem:eq-reln-equalities}
-If $\alpha : F \to G$ is a morphism in $[\oCPO^k, \oCPO]_{\omega}$
-and $f_1: A_1 \to B_1, ..., f_k : A_k \to B_k$, then $\graph{\alpha}^*
+If $\alpha : F \to G$ is a morphism in $[\oCPO^k, \oCPO]$ and $f_1:
+A_1 \to B_1, ..., f_k : A_k \to B_k$, then $\graph{\alpha}^*
 \graph{\overline{f}} = \graph{G \ol{f} \circ \alpha_{\ol{A}}} =
@@
 
-The {\em equality $\oCPO$ relation transformer} on $F :
-[\oCPO^k,\oCPO]_{\omega}$ is defined to be $\Eq_F =
-\graph{\id_{F}}$. Specifically, $\Eq_F = (F, F, \Eq_F^*)$ with
-$\Eq_F^* = \graph{\id_{F}}^*$, and Lemma~\ref{lem:eq-reln-equalities}
-indeed ensures that
+The {\em equality relation transformer} on $F : [\oCPO^k,\oCPO]$ is
+defined to be $\Eq_F = \graph{\id_{F}}$. Specifically, $\Eq_F = (F, F,
+\Eq_F^*)$ with $\Eq_F^* = \graph{\id_{F}}^*$, and
+Lemma~\ref{lem:eq-reln-equalities} indeed ensures that
 \begin{equation}\label{eq:eq-pres-eq}
@@
 \end{equation}
-for all $\ol{A : \oCPO}$.  Graph $\oCPO$ relation transformers in
-general, and equality $\oCPO$ relation transformers in particular,
-extend to $\oCPO$ relation environments in the obvious ways.  Indeed,
+for all $\ol{A : \oCPO}$.  Graph relation transformers in
+general, and equality relation transformers in particular,
+extend to relation environments in the obvious ways.  Indeed,
 if $\rho, \rho' : \oCPOenv$ and $f : \rho \to \rho'$, then the {\em
-  graph $\oCPO$ relation environment} $\graph{f}$ is defined pointwise
+  graph relation environment} $\graph{f}$ is defined pointwise
 by $\graph{f} \phi = \graph{f \phi}$ for every $\phi$, which entails
 that $\pi_1 \graph{f} = \rho$ and $\pi_2 \graph{f} = \rho'$. In
-particular, the {\em equality $\oCPO$ relation environment} $\Eq_\rho$
+particular, the {\em equality relation environment} $\Eq_\rho$
 is defined to be $\graph{\id_{\rho}}$, which entails that $\Eq_\rho
@@
 Theorem~\ref{thm:iel}: {\color{red} CHECK THIS!} for any $H$, $\rho$,
-$A$, and any subformula $J$ of $H$, both $T^i_{H,\Eq_{\rho}} K_0\,
+$\ol A$, and any subformula $J$ of $H$, both $T^i_{H,\Eq_{\rho}} K_0\,
 \ol{\Eq_A} = (\Eq_{(T^\set_{H,\rho})^i K_0})^* \ol{\Eq_A}$\; and\;

### 2020-09-28 20-gadts/inf-rules.tex
sections 1 through 3 of inf-rules.tex and IEL part of supp.tex done

@@
 %\AXC{$\ol{\Gamma;\Phi \vdash A} \hspace*{0.3in} \Gamma;\Phi~|~\Delta \vdash t : F[\ol{\alpha :=A}]$}
-\BIC{$\Gamma;\emptyset~|~\Delta \vdash \int_{\ol K,F} :
+\BIC{$\Gamma;\emptyset~|~\emptyset \vdash \int_{\ol K,F} :
 \Nat^{\Phi,\ol\alpha}\,F\, (\Lan^{\ol\alpha}_{\ol K}\,  F)\ol{K}$}  
@@
 the types of $\tin_H$ and $\fold^F_H$ to be well-formed, the length of
-$\alpha$ must equal the arity of $\phi$.)  {\color{red} In the rule
-  for $\int_{\ol K,H}$, ANYTHING NEEDED HERE?}
+$\alpha$ must equal the arity of $\phi$.)
 
@@
 $\iota'_{X,g'}$ is the injection of $FX$ into $(\mathit{Lan}_{\ol
-  K}\,F)\,B' = \colim{X : \set_0,\, g' : KX \to B'}{FX}$.}
+  K}\,F)\,B' = \colim{X : \set_0,\, g' : KX \to B'}{FX}$.
+
+\vspace*{0.1in}
+
+**Say somewhere that recognizing the non-parallel nature of the Lan
+interps, and yet still being able to define the set and rel interps so
+that everything is intertwined in exactly the right way to make the
+proofs (e.g., of the IEL) work, is what's hard here. This is where the
+insight is required.**}
 
@@
 By induction on the structure of $F$. Only the $\Nat$, application,
-fixpoint, and $\Lan$ cases are non-routine.
-
-The latter two use
-Equation~\ref{eq:eq-pres-eq}. The fixpoint case also uses the
-observation that, for every $i < \omega$, the following intermediate
-results can be proved by simultaneous induction with
-Theorem~\ref{thm:iel}: {\color{red} CHECK THIS!} for any $H$, $\rho$,
-$\ol A$, and any subformula $J$ of $H$, both $T^i_{H,\Eq_{\rho}} K_0\,
-\ol{\Eq_A} = (\Eq_{(T^\set_{H,\rho})^i K_0})^* \ol{\Eq_A}$\; and\;
+fixpoint, and $\Lan$ cases are non-routine. The application and
+fixpoint cases use Equation~\ref{eq:eq-pres-eq}. The fixpoint case
+$\Gamma;\Phi,\ol\gamma \vdash (\mu \phi.\lambda \ol\alpha.H)\ol F$
+also uses the observation that, for every $i < \omega$, the following
+intermediate results can be proved by simultaneous induction with
+Theorem~\ref{thm:iel}: for any $H$, $\rho$, $\ol A$, and any
+subformula $J$ of $H$, $T^i_{H,\Eq_{\rho}} K_0\, \ol{\Eq_A} =
+(\Eq_{(T^\set_{H,\rho})^i K_0})^* \ol{\Eq_A}$\; and\;
 \[\begin{array}{ll}
@@
 \end{array}\]
-hold. With these results in hand, the proof follows easily
-{\color{red} for all but $\Lan$-types}. The entire proof is given in
-detail in the accompanying anonymous supplementary material. As noted
-there, if functorial variables of arity greater than $0$ were allowed
-to appear in the bodies of $\mu$-types, then the IEL would fail.
-
-{\color{red} For $\Lan$-types,....}
-
+The case of the latter equation for $J = (\Lan_{\ol K}^{\ol
+  \beta}\,F)\ol B$ uses the facts that $\pi_1 (T^i_{H,\Eq_\rho} K_0) =
+(T^\set_{H,\rho})^i K_0$ for all $i < \omega$, $\pi_2
+(T^i_{H,\Eq_\rho} K_0) = (T^\set_{H,\rho})^i K_0$ for all $i <
+\omega$, and that, without loss of generality, the variables in $\ol
+\beta$ do not appear free in $H$. With these results in hand, the
+proof of Theorem~\ref{thm:iel} follows. The simultaneous induction
+required to make the various parts of the proof hang together
+appropriately is somewhat delicate, so we give the proof in its
+entirety in the {\color{blue} appendix}. As noted there, if functorial
+variables of arity greater than $0$ were allowed in the bodies of
+$\mu$-types then the IEL would fail.
 \end{proof}
@@
 
-\vspace*{1in}
-
-\hrule
-
-\vspace*{0.1in}
-
-\new{
-Some questions/issues:
-\begin{itemize}
-\item Can we write zipBush and appendBush with $\partial$ and $\int$?
-  We could already represent the uncurried type of appendBush
-  (although not its curried type), but couldn't recurse over both
-  input bushes because folds take natural transformations as inputs.
-\item More generally, how do we compute with $\partial$ and $\int?$
-  Can we use the colimit formulation of Lans (see Lemma 6.3.7
-  of~\cite{rie16}) to get a handle on this?
-\item What is the connection between exponentials and natural
-  transformations? (Should we assume only small objects are
-  exponentiable?) Do we want the former or the latter for
-  computational purposes? (I suspect the latter.)
-
-  [From nlab:
-  In a functor category $D^C$ , a natural transformation $\alpha : F \to
-  G$ is exponentiable if (though probably not ``only if'') it is
-  cartesian and each component $\alpha_c : F c \to G c$ is
-  exponentiable in $D$.  Given $H \to F$ we define $(\Pi_\alpha H) c =
-  \Pi_{\alpha_c}(Hc)$; then for $u : c \to c'$ to obtain a map
-  $\Pi_{\alpha_c} (Hc) \to \Pi_{\alpha_{c'}} (Hc')$ we need a map
-  $\alpha^*_{c'}(\Pi_{\alpha_c}(Hc)) \to Hc'$. But since $\alpha$ is
-  cartesian, $\alpha^*_{c'}(\Pi_{\alpha_c}(Hc)) \cong
-  \alpha^*_{c}(\Pi_{\alpha_c}(Hc))$, so we have the counit
-  $\alpha^*_{c}(\Pi_{\alpha_c}(Hc)) \to Hc$ that we can compose with
-  $Hu$.]
-\item After we understand what we can do with Lans and folds on GADTs
-  we might want to try to extend calculus with term-level
-  fixpoints. This would give a categorical analogue for GADTs
-  of~\cite{pit98,pit00} for ADTs. Would it also more accurately
-  reflect how GADTs are used in practice, or are functions over GADTs
-  usually folds? Investigate applications in the literature and/or in
-  implementations.
-\item $\omega$CPO is a natural choice for modeling general
-  recursion. We know $(Lan^\gamma_C \onet)D$ is $C \to D$ for any
-  closed type $C$. (Also for select classes of open types?) So can
-  model $\Nat \to \gamma$. But the functor $NX = \Nat \to X$ isn't
-  $\omega$-cocontinuous. It also doesn't preserve
-  $\omega$-presentable objects, i.e., countable $\omega$CPOs since
-  $\Nat \to \Nat$ is not countable. So we cannot have a functor like
-  $N$ as the subscript to $\Lan$ and expect the resulting $\Lan$ to be
-  $\omega$-cocontinuous.
-\item What functors can be subscripts to $\Lan$ and produce
-  $\omega$-cocontinuous functors? We can use functors that preserve
-  presentable objects by theorem in~\cite{jp19}, and possibly others
-  as well. These include polynomial functors, ADTs and nested types
-  seen as functors, certain (which?) GADTs seen as functors? How big
-  can GADTs get? 
-\end{itemize}
-}
+\section{Interpreting Terms}
 

### 2020-09-29 20-gadts/inf-rules.tex
adding well-definedness of term interps to inf.rules

@@
 
+\subsection{Preliminaries}
+
+\new{WHERE DOES THIS ALL GO? Must explain $\Lan$s, say why they are
+  functors, say they can be computed using colimits in $\set$ and
+  $\rel$, define injections, etc. If $f : B \to B'$ in $\set$, then
+  the functorial action of $\mathit{Lan}_{\ol K}\,F$ on $f$ is the
+  unique morphism $(\mathit{Lan}_{\ol K}\,F)f$ such that
+  \begin{equation}\label{def:functorial-action-lan}
+(\mathit{Lan}_{\ol K}\,F)f \circ \iota_{X,g} = \iota'_{X, f \circ
+      g}~~~MISSING~SOME~BARS
+\end{equation}
+Here, $\iota_{X,g}$ is the injection of $FX$ into $(\mathit{Lan}_{\ol
+  K}\,F)\,B = \colim{X : \set_0,\, g : KX \to B}{FX}$ and
+$\iota'_{X,g'}$ is the injection of $FX$ into $(\mathit{Lan}_{\ol
+  K}\,F)\,B' = \colim{X : \set_0,\, g' : KX \to B'}{FX}$.
+**Say somewhere that recognizing the non-parallel nature of the Lan
+interps, and yet still being able to define the set and rel interps so
+that everything is intertwined in exactly the right way to make the
+proofs (e.g., of the IEL) work, is what's hard here. This is where the
+insight is required.**}
+
 \subsection{Object Interpretations of Types}\label{sec:obj-interp}
@@
 and $\D$, for our $\set$ interpretations we have:
+
+{\color{red} We need to restrict set and relational environments to
+  $\set_0$ and $\rel_0$ for interps of $\Lan$-types and -terms, and
+  note that this is OK because these are our presentable objects, all
+  (other) objects can be written as $\omega$-directed colimits of
+  presentables, and our interpretations are $\omega$-cocontinuous and
+  therefore preserve $\omega$-directed colimits. Say once here, and
+  then just use throughout.}
 
@@
 \end{dfn}
-\new{WHERE DOES THIS ALL GO? Must explain $\Lan$s, say why they are
-  functors, say they can be computed using colimits in $\set$ and
-  $\rel$, define injections, etc. If $f : B \to B'$ in $\set$, then
-  the functorial action of $\mathit{Lan}_{\ol K}\,F$ on $f$ is the
-  unique morphism $(\mathit{Lan}_{\ol K}\,F)f$ such that
-  \begin{equation}\label{def:functorial-action-lan}
-(\mathit{Lan}_{\ol K}\,F)f \circ \iota_{X,g} = \iota'_{X, f \circ
-      g}~~~MISSING~SOME~BARS
-\end{equation}
-Here, $\iota_{X,g}$ is the injection of $FX$ into $(\mathit{Lan}_{\ol
-  K}\,F)\,B = \colim{X : \set_0,\, g : KX \to B}{FX}$ and
-$\iota'_{X,g'}$ is the injection of $FX$ into $(\mathit{Lan}_{\ol
-  K}\,F)\,B' = \colim{X : \set_0,\, g' : KX \to B'}{FX}$.
 
 \vspace*{0.1in}
-
-**Say somewhere that recognizing the non-parallel nature of the Lan
-interps, and yet still being able to define the set and rel interps so
-that everything is intertwined in exactly the right way to make the
-proofs (e.g., of the IEL) work, is what's hard here. This is where the
-insight is required.**}
 
@@
 
+If $\Delta = x_1 : F_1,...,x_n : F_n$ is a term context for $\Gamma$
+and $\Phi$, then the interpretations $\setsem{\Gamma;\Phi \vdash \Delta}$ and
+$\relsem{\Gamma;\Phi \vdash \Delta}$ are defined by
+\[\begin{array}{lll}
+\setsem{\Gamma;\Phi \vdash \Delta} & = & \setsem{\Gamma;\Phi \vdash
+  F_1} \times ... \times \setsem{\Gamma;\Phi \vdash F_n}\\ 
+\relsem{\Gamma;\Phi \vdash \Delta} & = & \relsem{\Gamma;\Phi \vdash
+  F_1} \times ... \times \relsem{\Gamma;\Phi \vdash F_n}\\ 
+\end{array}\]
+Every well-formed term $\Gamma;\Phi~|~\Delta \vdash t : F$ then
+has, for every $\rho \in \setenv$, set interpretations
+$\setsem{\Gamma;\Phi~|~\Delta \vdash t : F}\rho$ as natural
+transformations from $\setsem{\Gamma; \Phi \vdash \Delta}\rho$ to
+$\setsem{\Gamma; \Phi \vdash F}\rho$, and, for every $\rho \in
+\relenv$, relational interpretations $\relsem{\Gamma;\Phi~|~\Delta
+  \vdash t : F}\rho$ as natural transformations from
+$\relsem{\Gamma; \Phi \vdash \Delta}\rho$ to $\relsem{\Gamma; \Phi
+  \vdash F}\rho$. These are given in Definition~\ref{def:set-interp}.
+
+\begin{dfn}\label{def:set-interp}
+If $\rho$ is a set (resp., relation) environment and
+$\Gamma;\Phi~|~\Delta \vdash t : F$ then
+$\setsem{\Gamma;\Phi~|~\Delta \vdash t : F}\rho$ (resp.,
+$\relsem{\Gamma;\Phi~|~\Delta \vdash t : F}\rho$) is defined as in
+Figure~\ref{fig:term-sem}, where $\mathsf D$ is either $\set$ or
+$\rel$ as appropriate.
+\end{dfn}
+
+\begin{figure*}
+\hspace*{-3.1in}
+\resizebox{0.46\linewidth}{!}{
+\begin{minipage}[t]{0.5\textwidth}
+\[\begin{array}{lll}
+\dsem{\Gamma;\Phi \,|\, \Delta,x :F \vdash x : F} \rho& = &
+\pi_{|\Delta|+1}\\
+%\dsem{\Gamma;\emptyset \,|\, \Delta \vdash \lambda x.t : \sigma \to \tau}\rho &
+%= & \curry (\dsem{\Gamma;\emptyset \,|\, \Delta, x : \sigma \vdash t :
+%  \tau}\rho)\\ 
+%\dsem{\Gamma;\emptyset \,|\, \Delta \vdash st: \tau} \rho & = &
+%\eval \circ
+% \langle \dsem{\Gamma;\emptyset \,|\, \Delta \vdash s: \sigma \to
+%  \tau}\rho, \dsem{\Gamma;\emptyset \,|\, \Delta \vdash t: \sigma}\rho
+%\rangle\\
+\dsem{\Gamma;\emptyset \,|\, \Delta \vdash L_{\overline \alpha} x.t : \Nat^{\overline
+    \alpha} \,F \,G}\rho & = &  \curry (\dsem{\Gamma;\overline \alpha
+  \,|\, \Delta, x : F \vdash t: G}\rho[\overline{\alpha := \_}])\\
+\dsem{\Gamma;\Phi \,|\, \Delta \vdash t_{\overline K} s:
+  G [\overline{\alpha := K}]}\rho & = & \eval \circ \langle
+  \lambda d.\,(\dsem{\Gamma;\emptyset \,|\, \Delta \vdash t :
+  \Nat^{\overline{\alpha}} \,F \,G}\rho\; d)_{\overline{\dsem{\Gamma;\Phi
+      \vdash K}\rho}},\\ 
+ & & \hspace*{0.5in} \dsem{\Gamma;\Phi \,|\,
+    \Delta \vdash s: F [\overline{\alpha := K}]}\rho \rangle\\ 
+& & \\
+%% \color{red} \mbox{Add rules for } \forall \mbox{ if we include it} & & \\
+%\dsem{\Gamma;\Phi \,|\, \Delta,x :F \vdash x : F} \rho& = &
+%\pi_{|\Delta|+1}\\
+\dsem{\Gamma;\Phi \,|\, \Delta \vdash \bot_F t : F} \rho& = &
+!^0_{\dsem{\Gamma;\Phi \vdash F}\rho} \circ
+  \dsem{\Gamma;\Phi~|~\Delta \vdash t : \zerot}\rho, \mbox{ where } \\
+ & & \hspace*{0.1in} !^0_{\dsem{\Gamma;\Phi \vdash F}\rho}
+\mbox{ is the unique morphism from } 0\\
+ & & \hspace*{0.1in} \mbox{ to } \dsem{\Gamma;\Phi \vdash F}\rho\\
+\dsem{\Gamma;\Phi \,|\, \Delta \vdash \top : \onet}\rho & = &
+!^{\dsem{\Gamma;\Phi\vdash \Delta}\rho}_1, \mbox{ where }
+!^{\dsem{\Gamma;\Phi\vdash \Delta}\rho}_1\\ 
+& & \hspace*{0.1in} \mbox{ is the unique morphism from }
+\dsem{\Gamma;\Phi\vdash \Delta}\rho \mbox{ to } 1\vspace*{0.05in}\\ 
+\dsem{\Gamma;\Phi \,|\, \Delta \vdash (s,t) : F \times G} \rho& = &
+\dsem{\Gamma;\Phi \,|\, \Delta \vdash s: F} \rho\times
+\dsem{\Gamma;\Phi \,|\, \Delta \vdash t : G} \rho\\
+\dsem{\Gamma;\Phi \,|\, \Delta \vdash \pi_1 t : F} \rho& = &
+\pi_1 \circ \dsem{\Gamma;\Phi \,|\, \Delta \vdash t : F \times G}\rho\\
+\dsem{\Gamma;\Phi \,|\, \Delta \vdash \pi_2 t : G}\rho & = &
+\pi_2 \circ \dsem{\Gamma;\Phi \,|\, \Delta \vdash t : F \times
+  G} \rho\\
+\dsem{\Gamma;\Phi~|~\Delta \vdash \case{t}{x \mapsto l}{y \mapsto r} :
+  K}\rho & = & \eval \circ \langle \curry \,[\dsem{\Gamma;\Phi
+    \,|\, \Delta, x : F \vdash l : K}\rho,\\
+   & & \hspace*{0.79in} \dsem{\Gamma;\Phi \,|\, \Delta, y
+    : G \vdash r : K}\rho],\\
+   & &  \hspace*{0.5in} \dsem{\Gamma;\Phi \,|\, \Delta \vdash t :
+  F + G} \rho\rangle\vspace*{0.05in}\\   
+\dsem{\Gamma;\Phi \,|\, \Delta \vdash \inl \,s: F + G} \rho& = &
+\inl \circ \dsem{\Gamma;\Phi \,|\, \Delta \vdash s: F}\rho\\
+\dsem{\Gamma;\Phi \,|\, \Delta \vdash \inr \,t: F + G}\rho & = & 
+\inr \circ \dsem{\Gamma;\Phi \,|\, \Delta \vdash t : G}\rho\\
+\llbracket \Gamma;\emptyset \,|\, \emptyset \vdash \map^{\ol{F},\ol{G}}_H
+  : \Nat^\emptyset (\Nat^{\ol\beta,\ol\gamma} F\,G)
+& = & \lambda d\, \ol\eta\,\ol{C}.\,
+\dsem{\Gamma; \ol{\phi},\ol{\gamma}\vdash H}\id_{\rho[\ol{\gamma:=
+      C}]}[\ol{\phi := \lambda \ol B. \eta_{\ol B\, \ol C}}]\\ 
+\hspace*{0.5in}
+  (\Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\,H[\ol{\phi
+      :=_{\ol{\beta}} G}]) \rrbracket^{\mathsf D} \rho & &\\
+%\hspace*{0.79in}  & & \hspace*{0.1in}\id_{\rho[\ol{\gamma :=
+%      C}]}[\ol{\phi := \lambda
+%    \ol{B}. (\dsem{\Gamma;\ol\alpha\,|\,\Delta \vdash t : 
+%    \Nat^{\ol{\beta},\ol{\gamma}}\,F\,G}\rho\,d)_{\ol{B}\,\ol{C}}}]\\
+\llbracket \Gamma;\emptyset \,|\, \emptyset \vdash \tin_H :
+Nat^{\ol{\beta},\ol{\gamma}} \, H[\phi := (\mu \phi.\lambda {\overline
+    \alpha}.H){\overline \beta}][\ol{\alpha := \beta}] & = &
+\lambda d\,\ol{B}\,
+\ol{C}.\,(\mathit{in}_{{T}^X_{H,\rho[\ol{\gamma :=
+        C}]}})_{\ol{B}} \;\;\mbox{ where } X \mbox{ is } \set \mbox{ when } \\ 
+\hspace*{0.79in}(\mu \phi.\lambda {\overline \alpha}.H){\overline
+  \beta} \rrbracket^{\mathsf{D}} \rho & & \hspace*{0.2in}  
+\mathsf{D} = \set \mbox{ and not present when }
+\mathsf{D} = \rel\vspace*{0.05in}\\  
+\llbracket \Gamma;\emptyset \,|\, \emptyset \vdash
+  \fold^F_H : \Nat^\emptyset\;(\Nat^{\ol{\beta}, \ol{\gamma}}\,H[\phi
+    :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F) & = &  
+\lambda d\,\eta\,\ol{B}\,\ol{C}.\,
+(\mathit{fold}_{T^X_{H,\rho[\ol{\gamma := C}]}} \, (\lambda
+\ol{A}.\,\eta_{\ol{A}\,\ol{C}}))_{\ol{B}}\\ 
+\hspace*{0.79in}(\Nat^{{\ol{\beta},\ol{\gamma}} }\,(\mu
+  \phi.\lambda \overline \alpha.H)\overline \beta\;F)
+\rrbracket^{\mathsf{D}} \rho & & \hspace*{0.2in} \mbox{ where } X
+\mbox{ is as above }\vspace*{0.05in} \\
+\setsem{\Gamma;\emptyset~|~\emptyset \vdash \int_{\ol K,F}
+: \Nat^{\Phi,\ol\alpha}\,F\, (\Lan^{\ol\alpha}_{\ol K}\, F)\ol{K}}\rho
+& = & \vspace*{0.05in}\\
+\relsem{\Gamma;\emptyset~|~\emptyset \vdash \int_{\ol K,F} :
+  \Nat^{\Phi,\ol\alpha}\,F\, (\Lan^{\ol\alpha}_{\ol K}\,
+  F)\ol{K}}\rho\,d & = & \vspace*{0.05in} \\
+\setsem{\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol
+      K}_F \eta : Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}
+  F)\ol \beta\; G}\rho & = &\lambda \ol{N} \, \ol{B}.\,
+((\mu_{\ol{N}})_{\ol{B}})|_{\setsem{\Gamma; \Phi, \ol\beta \vdash
+    (\Lan^{\ol\alpha}_{\ol K} F)\ol \beta } \rho[\Phi, \ol \beta :=
+    \ol N, \ol B]}\\
+& & \hspace*{0.2in} \mbox{where } \mu_{\ol{N}} \mbox{ is as described below}
+\vspace*{0.05in}\\
+\relsem{\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol
+      K}_F \eta : Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}
+    F)\ol \beta\; G}\rho & = & \\
+\vspace*{-0.1in} 
+\end{array}\]
+\caption{Term semantics}\label{fig:term-sem} 
+\end{minipage}}\vspace*{-0.05in}
+\end{figure*}
+
+\new{In the fourth-to-last clause of Definition~\ref{def:set-interp},...}
+
+\new{In the third-to-last clause of Definition~\ref{def:set-interp},...}
+
+In the next-to-last clause of Definition~\ref{def:set-interp},
+  \[\mu_{\ol{N}} : \Lan_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho
+  {[\ol{\alpha := \_} ]}}} \setsem{\Gamma; \Phi, \ol\alpha \vdash
+    F}\rho {[\Phi, \ol \alpha := \ol N, \_ ]} \to \setsem{\Gamma;
+    \Phi, \ol\beta \vdash G} \rho[\Phi, \ol \beta := \ol N, \_]\]
+  is the unique natural transformation such that
+\begin{equation}\label{eq:defn-of-mu-sub-N}
+    (\setsem{\Gamma; \emptyset \,|\, \Delta \vdash \eta : \Nat^{\Phi,
+    \ol\alpha} \, F \,\, G[\ol{\beta := K}]}\rho \, d)_{\ol{N}} =
+\mu_{\ol{N}} \, \ol{\setsem{\Gamma; \ol\alpha \vdash K}
+  \rho[\ol{\alpha := \_}]} \circ \epsilon_{\ol{N}}
+  \end{equation}
+Here, $\epsilon_{\ol{N}}$ is the natural transformation associated
+with the left Kan extension
+\[\begin{tikzcd} [column sep = large, row sep = large]
+\set^{|\ol \alpha|}
+\ar[rr, "\setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho {[\Phi, \ol
+      \alpha := \ol N, \_ ]}"{name=F}] 
+\ar[dr, "\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho {[\ol{\alpha := \_} ]}}"{left}] && \set \\ 
+&\set^{|\ol K|} \ar[ur, "\Lan_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho {[\ol{\alpha := \_} ]}}}
+\setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho {[\Phi, \ol \alpha :=
+    \ol N, \_ ]}"{right}]
+\ar[d, Rightarrow, shorten <= 2mm, shorten >= 1mm, 
+          from=F, to=2-2,  "\epsilon_{\ol{N}}"{}]
+\end{tikzcd}\]
+
+\new{In the final last clause of Definition~\ref{def:set-interp},...}
+
+\vspace*{1in}
+
+If $t$ is closed, i.e., if $\emptyset; \emptyset~|~\emptyset \vdash t
+: F$, then we write $\dsem{\vdash t : F}$ instead of $\dsem{\emptyset;
+  \emptyset~|~\emptyset \vdash t : F}$.  The return type for the
+semantic fold is $\dsem{\Gamma;\ol\beta,\ol\gamma \vdash
+  F}\rho[\ol{\gamma := C}][\ol{\beta := B}]$.  This interpretation
+gives $\dsem{\Gamma;\emptyset \,|\, \Delta \vdash \lambda x.t : F \to
+  G}\rho = \curry (\dsem{\Gamma;\emptyset \,|\, \Delta, x : F \vdash t
+  : G}\rho)$ and $\dsem{\Gamma;\emptyset \,|\, \Delta \vdash st: G}
+\rho = \eval \circ \langle \dsem{\Gamma;\emptyset \,|\, \Delta \vdash
+  s: F \to G}\rho, \dsem{\Gamma;\emptyset \,|\, \Delta \vdash t:
+  F}\rho \rangle$, so it specializes to the standard interpretations
+for System F terms.  To see that the object and relational
+interpretations of $\Gamma;\emptyset~|~\emptyset \vdash \int_{\ol K,F}
+: \Nat^{\Phi,\ol\alpha}\,F\, (\Lan^{\ol\alpha}_{\ol K}\, F)\ol{K}$ and
+$\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol K}_F \eta :
+\Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K} F)\ol \beta\; G$ are
+well-defined is significantly more involved, as we see in the next
+four subsections.
+
+\subsection{$\setsem{\Gamma;\emptyset~|~\emptyset \vdash \int_{\ol K,F}
+: \Nat^{\Phi,\ol\alpha}\,F\, (\Lan^{\ol\alpha}_{\ol K}\, F)\ol{K}}$ is
+  well-defined}
+
+\subsection{$\relsem{\Gamma;\emptyset~|~\emptyset \vdash \int_{\ol K,F}
+: \Nat^{\Phi,\ol\alpha}\,F\, (\Lan^{\ol\alpha}_{\ol K}\, F)\ol{K}}$ is
+  well-defined}
+
+\subsection{$\setsem{\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol
+      K}_F \eta : Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}
+    F)\ol \beta\; G}$ is well-defined} We must show that, for every
+$\rho : \setenv$ and $d : \setsem{\Gamma; \emptyset \vdash \Delta}
+\rho$, \[\setsem{\Gamma; \emptyset \,|\, \Delta \vdash \partial^{G,
+    \ol K}_F \eta : \Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}
+  F)\ol \beta\; G} \rho \, d\] has the correct type and satisfies the
+condition required for it to be in $\setsem{\Gamma; \emptyset \vdash
+  \Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K} F)\ol \beta\; G}
+\rho$. For the former we observe that each component
+$(\mu_{\ol{N}})_{\ol{B}}$ of $\mu_{\ol{N}}$ has type
+  \[\Lan_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho
+  {[\ol{\alpha := B} ]}}} \setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho
+        {[\Phi, \ol \alpha := \ol N, \ol B]} \to \setsem{\Gamma; \Phi,
+          \ol\beta \vdash G} \rho[\Phi, \ol \beta := \ol N, \ol B]\]
+        Each component of $(\mu_{\ol{N}})_{\ol{B}} |_{\setsem{\Gamma;
+            \Phi, \ol\beta \vdash (\Lan^{\ol\alpha}_{\ol K} F)\ol
+            \beta } \rho[\Phi, \ol \beta := \ol N, \ol B]}$ therefore
+        has the correct type to be the component at $\ol N$ and $\ol
+        B$ of a natural transformation in $\setsem{\Gamma; \emptyset
+          \vdash \Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}
+          F)\ol \beta\; G} \rho$.
+
+The next two lemmas show that $ \lambda \ol{N} \,
+\ol{B}. (\mu_{\ol{N}})_{\ol{B}} |_{\setsem{\Gamma; \Phi, \ol\beta
+    \vdash (\Lan^{\ol\alpha}_{\ol K} F)\ol \beta } \rho[\ol{\Phi,
+      \beta := N, B}]}$ is natural in both $\ol{N}$ and $\ol{B}$ and
+that it satisfies the condition required for it to be in
+$\setsem{\Gamma; \emptyset \vdash \Nat^{\Phi, \ol\beta}\,
+  (\Lan^{\ol\alpha}_{\ol K} F)\ol \beta\; G} \rho$, namely, if $\Phi =
+{\phi_1,...,\phi_n}$ and $\phi_i$ has arity $k_i$ for $i = 1,...,n$,
+then 
+\begin{equation}\label{eq:P2}
+    \begin{split}
+      &\forall \ol{N = (N^1, N^2, N^*) : RT_{k_i}}, \forall \ol{B =
+        (B^1, B^2, B^*) :
+        RT_0}.\\ &((\mu_{\ol{N^1}})_{\ol{B^1}}|_{\setsem{\Gamma; \Phi,
+          \ol\beta \vdash (\Lan^{\ol\beta}_{\ol K} F)\ol \beta }
+        \rho[\Phi, \ol \beta := \ol{N^1}, \ol{B^1}]} ,
+      (\mu_{\ol{N^2}})_{\ol{B^2}}|_{\setsem{\Gamma; \Phi, \ol\beta
+          \vdash (\Lan^{\ol\beta}_{\ol K} F)\ol \beta } \rho[\Phi, \ol
+          \beta := \ol{N^2}, \ol{B^2}]} )\\ &: \relsem{\Gamma; \Phi,
+        \ol\beta \vdash (\Lan^{\ol\beta}_{\ol K} F)\ol \beta
+      }\Eq_{\rho}[\Phi, \ol \beta := \ol N, \ol B] \rightarrow
+      \relsem{\Gamma; \Phi, \ol\beta \vdash G}\Eq_{\rho}[\Phi, \ol
+        \beta := \ol N, \ol B]
+    \end{split}
+    \end{equation}
+
+\begin{lemma}
+For every $\rho : \setenv$, $d : \setsem{\Gamma; \emptyset \vdash
+  \Delta} \rho$, $\ol{N = (N^1, N^2, N^*) : RT_{k_i}}$, $\ol{B = (B^1,
+  B^2, B^*) : RT_0}$, and $(t_1, t_2) \in \relsem{\Gamma; \Phi,
+  \ol\beta \vdash (\Lan^{\ol\beta}_{\ol K} F)\ol \beta
+}\Eq_{\rho}[\Phi, \ol \beta := \ol N, \ol B]$, it holds that
+\[((\mu_{\ol{N^1}})_{\ol{B^1}} \,\, t_1, (\mu_{\ol{N^2}})_{\ol{B^2}} \,\, t_2) \in
+\relsem{\Gamma; \Phi, \ol\beta \vdash G}\Eq_{\rho}[\Phi, \ol \beta :=
+  \ol N, \ol B]\]
+\end{lemma}
+
+\begin{proof}
+Unfolding the definition of relational interpretation for $\Lan$-types
+in the hypothesis gives that $(t_1, t_2) : \setsem{\Gamma;\Phi,
+  \ol\beta \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{\beta}}\rho[\Phi,
+  \ol \beta := \ol{N^1}, \ol{B^1}] \,\times\, \setsem{\Gamma;\Phi,
+  \ol\beta \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{\beta}}\rho[\Phi,
+  \ol \beta := \ol{N^2}, \ol{B^2}]$ and there exist $\ol{Z : \set_0},
+t_1' : \setsem{\Gamma; \Phi, \ol\alpha \vdash F} \rho[\ol{\Phi :=
+    N^1}][\ol{\alpha := Z}],\, t_2' : \setsem{\Gamma; \Phi, \ol\alpha
+  \vdash F}\rho[\ol{\Phi := N^2 }][\ol{\alpha := Z}]$,\\ $\ol{f :
+  \relsem{\Gamma; \ol{\alpha} \vdash K}\Eq_\rho[\ol{\alpha := \Eq_Z}]
+  \to B}$ \, such that $\iota_{\ol{Z},\ol{\pi_1 f}}t_1' = t_1,\,
+\kappa_{\ol{Z}, \ol{\pi_2 f}}t_2' = t_2,$ and $(t_1',t_2') \in
+\\ \relsem{\Gamma; \Phi, \ol\alpha \vdash F}\Eq_\rho[\ol{\Phi :=
+    N}][\ol{\alpha := \Eq_Z}].$ Inlining $t_1 =
+\iota_{\ol{Z},\ol{\pi_1 f}}t_1'$ and $t_2 = \kappa_{\ol{Z}, \ol{\pi_2
+    f}}t_2'$ in in the goal in the statement of the lemma gives us the
+following rewritten goal to prove:
+\begin{equation}\label{eq:new-goal}
+((\mu_{\ol{N^1}})_{\ol{B^1}} \,\, \iota_{\ol{Z},\ol{\pi_1 f}}\, t_1' ,
+  (\mu_{\ol{N^2}})_{\ol{B^2}} \,\,  \kappa_{\ol{Z}, \ol{\pi_2 f}}\, t_2'  )
+\in \relsem{\Gamma; \Phi, \ol\beta \vdash
+G}\Eq_{\rho}[\ol{\Phi, \beta := N, B}]
+\end{equation}
+
+To this end, consider the commuting diagram\\
+
+\begin{tikzcd}[column sep = large, row sep = large]
+    & \setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\Phi, \ol \alpha
+    := \ol{N^1}, \ol Z]
+    \ar[dr, "\;\;\;\;\;\iota_{\ol{Z}, \ol{\pi_1 f}}"{right}] 
+    \ar[dl, "\iota'_{\ol Z, 
+                \ol{\id_{\setsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := Z}]}}}\;\;\;\;\;\;\;\;\;\;\;\;"{left}] 
+        & \\
+    LKZ % (\Lan_{\setsem{\Gamma; \ol\alpha \vdash K}\rho{[\ol{\alpha := \_}]}}
+    %               \setsem{\Gamma; \Phi, \ol\alpha \vdash F}
+    %                 \rho{[\ol{\Phi, \alpha := N^1, \_}]}) \ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := Z}]}
+    \ar[rr, dashed, "(\Lan_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho{[\ol{\alpha := \_}]}}}
+                  \setsem{\Gamma; \Phi, \ol\alpha \vdash F}
+                    \rho{[\Phi, \ol \alpha := \ol{N^1}, \_]}) \ol{\pi_1 f}"{below}]
+       & & LB
+  \end{tikzcd}
+
+\noindent
+where $LKZ = (\Lan_{\ol{\setsem{\Gamma; \ol\alpha \vdash
+      K}\rho{[\ol{\alpha := \_}]}}} \setsem{\Gamma; \Phi, \ol\alpha
+  \vdash F} \rho{[\Phi, \ol \alpha := \ol{N^1}, \_]})
+\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := Z}]}$ and
+\\ $LB = (\Lan_{\ol{\setsem{\Gamma; \ol\alpha \vdash
+      K}\rho{[\ol{\alpha := \_}]}}} \setsem{\Gamma; \Phi, \ol\alpha
+  \vdash F} \rho{[\Phi, \ol \alpha := \ol{N^1}, \_]}) \ol{B^1}$. This
+diagram commutes by the universal property of $LKZ$ regarded as a
+colimit in $\set$.  Theorem 6.2.2 of~\cite{rie16} also gives that
+$\iota'_{\ol Z, \ol{\id_{\setsem{\Gamma; \ol\alpha \vdash
+        K}\rho[\ol{\alpha := Z}]}}} = (\epsilon_{\ol{N^1}})_{\ol{Z}}$.
+Together these two facts give
+\[  \iota_{\ol{Z}, \ol{\pi_1 f}}
+  = (\Lan_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho{[\ol{\alpha := \_}]}}}
+                  \setsem{\Gamma; \Phi, \ol\alpha \vdash F}
+                    \rho{[\ol{\Phi, \alpha := N^1, \_}]}) \ol{\pi_1 f}
+  \circ (\epsilon_{\ol{N^1}})_{\ol{Z}}
+\]
+\noindent
+An analogous proof gives 
+\[  \kappa_{\ol{Z}, \ol{\pi_2 f}}
+  = (\Lan_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho{[\ol{\alpha := \_}]}}}
+                  \setsem{\Gamma; \Phi, \ol\alpha \vdash F}
+                    \rho{[\ol{\Phi, \alpha := N^2, \_}]}) \ol{\pi_2 f}
+  \circ (\epsilon_{\ol{N^2}})_{\ol{Z}}
+\]
+Using these equalities, we can rewrite our goal~\eqref{eq:new-goal}
+again, this time as
+\begin{align}\label{eq:new-new-goal} 
+  ((\mu_{\ol{N^1}})_{\ol{B^1}} \,\, 
+  ((\Lan_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho{[\ol{\alpha := \_}]}}}
+                  \setsem{\Gamma; \Phi, \ol\alpha \vdash F}
+                    \rho{[\ol{\Phi, \alpha := N^1, \_}]}) \ol{\pi_1 f}
+  ((\epsilon_{\ol{N^1}})_{\ol{Z}}
+        \, t_1' )) , \\\nonumber
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+(\mu_{\ol{N^2}})_{\ol{B^2}} \,\,  
+  ((\Lan_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho{[\ol{\alpha := \_}]}}}
+                  \setsem{\Gamma; \Phi, \ol\alpha \vdash F}
+                    \rho{[\ol{\Phi, \alpha := N^2, \_}]}) \ol{\pi_2 f}
+  ((\epsilon_{\ol{N^2}})_{\ol{Z}}
+        \, t_2'  ))) \\\nonumber
+        \in \relsem{\Gamma; \Phi, \ol\beta \vdash
+    G}\Eq_{\rho}[\Phi, \ol \beta := \ol N, \ol B]
+\end{align}
+
+Now, by naturality of $\mu_{\ol{N^i}}$ for the morphisms $\ol{\pi_i
+  f}$ we have
+\begin{align*}
+(\mu_{\ol{N^i}})_{\ol{B^i}} \circ 
+  ((\Lan_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho{[\ol{\alpha := \_}]}}}
+                  \setsem{\Gamma; \Phi, \ol\alpha \vdash F}
+                    \rho{[\ol{\Phi, \alpha := N^i, \_}]}) \ol{\pi_i f} \\
+= \setsem{\Gamma; \Phi, \ol\beta \vdash G}\id_\rho 
+    [\ol{\Phi, \beta := \id_{N^i}, \pi_i f}]
+  \circ 
+  (\mu_{\ol{N^i}})_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}
+      \rho[\ol{\alpha := Z}]}}
+\end{align*}
+Using this, we can rewrite our goal~\eqref{eq:new-new-goal} a final
+time as
+\begin{align}\label{eq:new-new-new-goal} 
+  \big(\setsem{\Gamma; \Phi, \ol\beta \vdash G}\id_\rho 
+    [\Phi, \ol \beta := \ol{\id_{N^1}}, \ol{\pi_1 f}]
+  ((\mu_{\ol{N^1}})_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}
+      \rho[\ol{\alpha := Z}]}}
+  ((\epsilon_{\ol{N^1}})_{\ol{Z}}
+        \, t_1' )) , \\\nonumber
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+  \setsem{\Gamma; \Phi, \ol\beta \vdash G}\id_\rho 
+    [\Phi, \ol \beta := \ol{\id_{N^2}}, \ol{\pi_2 f}]
+  ((\mu_{\ol{N^2}})_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}
+      \rho[\ol{\alpha := Z}]}}
+  ((\epsilon_{\ol{N^2}})_{\ol{Z}}
+        \, t_2'  ))\big) \\\nonumber
+        \in \relsem{\Gamma; \Phi, \ol\beta \vdash
+    G}\Eq_{\rho}[\Phi, \ol \beta := \ol N, \ol B]
+\end{align}
+
+  {\color{red} START HERE!}
+
+To finish the proof, we would like to use that fact that
+$\relsem{\Gamma; \Phi,\ol\beta \vdash G}\id_{\Eq_\rho} [\Phi,
+   \ol \beta := \ol{\id_N}, \ol{f}]$ is a morphism in $\rel$ and thus preserves
+related elements. But in order to apply this morphism, we must first show
+that
+\begin{align*}
+((\mu_{\ol{N^1}})_{\ol{\setsem{\Gamma; \ol\alpha \vdash K} \rho[\ol{\alpha := Z}]}} 
+  ((\epsilon_{\ol{N^1}})_{\ol{Z}} \,\, t_1') &, 
+  %%%
+  (\mu_{\ol{N^2}})_{\ol{\setsem{\Gamma; \ol\alpha \vdash K} \rho[\ol{\alpha := Z}]}} 
+  ((\epsilon_{\ol{N^2}})_{\ol{Z}} \,\, t_2')) \\
+  & \in 
+  \relsem{\Gamma; \Phi, \ol\alpha \vdash
+        G[\ol{\beta := K}]}\Eq_{\rho}[\Phi, \ol \alpha := \ol N,
+    \ol{\Eq_Z}]
+\end{align*}
+To show this, we use the condition imposed on $\setsem{\Gamma;
+  \emptyset \,|\, \Delta \vdash \eta : \Nat^{\Phi, \ol\alpha} \, F
+  \,\, G[\ol{\beta := K}]} \rho \, d$ to be in $\setsem{\Gamma;
+  \emptyset \vdash \Nat^{\Phi, \ol\alpha} \, F \,\, G[\ol{\beta :=
+      K}]} \rho$, i.e.,
+
+    \begin{equation}\label{eq:P1}
+    \begin{split}
+      &\forall \ol{N = (N^1, N^2, N^*) : RT_k}, 
+          \forall \ol{M = (M^1, M^2, M^*) : RT_k}.\\
+      &((\setsem{\Gamma; \emptyset \,|\, \Delta \vdash \eta : \Nat^{\Phi, \ol\alpha} \, F \,\, G[\ol{\beta := K}]} \rho \, d)_{\ol{N^1}, \ol{M^1}}, 
+      (\setsem{\Gamma; \emptyset \,|\, \Delta \vdash \eta} : \Nat^{\Phi, \ol\alpha} \, F \,\, G[\ol{\beta := K}] \rho \, d)_{\ol{N^2}, \ol{M^2}}) \\
+      &: \relsem{\Gamma; \Phi, \ol\alpha \vdash F}\Eq_{\rho}[\ol{\Phi, \alpha := N, M}]
+      \rightarrow \relsem{\Gamma; \Phi, \ol\alpha \vdash
+        G[\ol{\beta := K}]}\Eq_{\rho}[\ol{\Phi, \alpha := N, M}]
+    \end{split}
+    \end{equation}
+along with the universal property of 
+$\Lan_{\setsem{\Gamma; \ol\alpha \vdash K}\rho {[\ol{\alpha := \_} ]}}
+      \setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho {[\ol{\Phi, \alpha := N^i, \_} ]}$
+      for $i = 1, 2$
+and the fact that $t_1'$ and $t_2'$ are related in 
+$\relsem{\Gamma; \Phi, \ol\alpha \vdash F}\Eq_{\rho}[\ol{\Phi, \alpha := N, \Eq_Z}]$.
+Applying Equation \eqref{eq:P1} with $\ol{M} = \ol{Z}$ 
+gives (eliding the type of $\eta$)
+    \begin{equation*}
+    \begin{split}
+      &((\setsem{\Gamma; \emptyset \,|\, \Delta \vdash \eta} \rho \, d)_{\ol{N^1}, \ol{Z}}, 
+      (\setsem{\Gamma; \emptyset \,|\, \Delta \vdash \eta} \rho \, d)_{\ol{N^2}, \ol{Z}}) \\
+      &: \relsem{\Gamma; \Phi, \ol\alpha \vdash F}\Eq_{\rho}[\ol{\Phi, \alpha := N, \Eq_Z}]
+      \rightarrow \relsem{\Gamma; \Phi, \ol\alpha \vdash
+        G[\ol{\beta := K}]}\Eq_{\rho}[\ol{\Phi, \alpha := N, \Eq_Z}]
+    \end{split}
+    \end{equation*}
+and since we know 
+$(t_1',t_2') \in
+  \relsem{\Gamma; \Phi, \ol\alpha \vdash F}\Eq_\rho[\ol{\Phi := N}][\ol{\alpha := \Eq_Z}]$, 
+we can apply this morphism
+%%
+to $(t_1', t_2')$:
+\begin{align*}
+  ((\setsem{\Gamma; \emptyset \,|\, \Delta \vdash \eta} \rho \, d)_{\ol{N^1}, \ol{Z}} t_1', 
+  (\setsem{\Gamma; \emptyset \,|\, \Delta \vdash \eta} \rho \, d)_{\ol{N^2}, \ol{Z}} t_2') 
+  \in 
+\relsem{\Gamma; \Phi, \ol\alpha \vdash
+        G[\ol{\beta := K}]}\Eq_{\rho}[\ol{\Phi, \alpha := N, \Eq_Z}]
+\end{align*}
+%%
+Applying the universal property at component $\ol{Z}$
+$$
+  ((\setsem{\Gamma; \emptyset \,|\, \Delta \vdash \eta}\rho \, d)_{\ol{N^i}})_{\ol{Z}}
+  = (\mu_{\ol{N^i}})_{\ol{\setsem{\Gamma; \ol\alpha \vdash K} \rho[\ol{\alpha := Z}]}} \circ 
+  (\epsilon_{\ol{N^i}})_{\ol{Z}}
+$$
+yields 
+\begin{align*}
+((\mu_{\ol{N^1}})_{\ol{\setsem{\Gamma; \ol\alpha \vdash K} \rho[\ol{\alpha := Z}]}} 
+  ((\epsilon_{\ol{N^1}})_{\ol{Z}} \,\, t_1') &, 
+  %%%
+  (\mu_{\ol{N^2}})_{\ol{\setsem{\Gamma; \ol\alpha \vdash K} \rho[\ol{\alpha := Z}]}} 
+  ((\epsilon_{\ol{N^2}})_{\ol{Z}} \,\, t_2')) \\
+  & \in 
+  \relsem{\Gamma; \Phi, \ol\alpha \vdash
+        G[\ol{\beta := K}]}\Eq_{\rho}[\ol{\Phi, \alpha := N, \Eq_Z}] \\
+  & = \relsem{\Gamma; \Phi,\ol\beta \vdash G}\Eq_\rho
+  [\ol{\Phi, \beta := N, \relsem{\Gamma; \ol\alpha \vdash K}\Eq_\rho[\ol{\alpha := \Eq_Z}]}]
+\end{align*}
+
+Finally, we can apply the morphism 
+\begin{align*}
+  \relsem{\Gamma; \Phi,\ol\beta \vdash G}\id_{\Eq_\rho} [\ol{\Phi, \beta := \id_N, f}] 
+  : \,\,
+  &\relsem{\Gamma; \Phi,\ol\beta \vdash G}\Eq_\rho
+  [\ol{\Phi, \beta := N, \relsem{\Gamma; \ol\alpha \vdash K}\Eq_\rho[\ol{\alpha := \Eq_Z}]}] \\
+  \rightarrow \,\,
+  &\relsem{\Gamma; \Phi,\ol\beta \vdash G}\Eq_\rho [\ol{\Phi, \beta := N, B}]
+\end{align*}
+to the pair
+$((\mu_{\ol{N^1}})_{\ol{\setsem{\Gamma; \ol\alpha \vdash K} \rho[\ol{\alpha := Z}]}} 
+  ((\epsilon_{\ol{N^1}})_{\ol{Z}} \,\, t_1') , 
+  %%%
+  (\mu_{\ol{N^2}})_{\ol{\setsem{\Gamma; \ol\alpha \vdash K} \rho[\ol{\alpha := Z}]}} 
+  ((\epsilon_{\ol{N^2}})_{\ol{Z}} \,\, t_2'))$
+, which gives
+
+\begin{align*}
+( 
+  &\setsem{\Gamma; \Phi,\ol\beta \vdash G}\id_{\rho} [\ol{\Phi, \beta := \id_{N^1}, \pi_1 f}] 
+  ((\mu_{\ol{N^1}})_{\ol{\setsem{\Gamma; \ol\alpha \vdash K} \rho[\ol{\alpha := Z}]}} 
+  ((\epsilon_{\ol{N^1}})_{\ol{Z}} \,\, t_1')) ,  \\
+  %%%
+  &\setsem{\Gamma; \Phi,\ol\beta \vdash G}\id_{\rho} [\ol{\Phi, \beta := \id_{N^2}, \pi_2 f}] 
+  ((\mu_{\ol{N^2}})_{\ol{\setsem{\Gamma; \ol\alpha \vdash K} \rho[\ol{\alpha := Z}]}} 
+  ((\epsilon_{\ol{N^2}})_{\ol{Z}} \,\, t_2'))) \\
+  & \in 
+  \relsem{\Gamma; \Phi, \ol\alpha \vdash
+        G}\Eq_{\rho}[\ol{\Phi, \alpha := N,B}] 
+\end{align*}
+and thus 
+$$((\mu_{\ol{N^1}})_{\ol{B^1}} \,\, t_1, 
+                (\mu_{\ol{N^2}})_{\ol{B^2}} \,\, t_2)
+                \in \relsem{\Gamma; \Phi, \ol\beta \vdash
+            G}\Eq_{\rho}[\ol{\Phi, \beta := N, B}]$$
+
+\end{proof}
+
+\begin{lemma}
+  For all $\rho : \setenv$ and $d : \setsem{\Gamma; \emptyset \vdash \Delta} \rho$, 
+  the family of morphisms 
+  $$
+  \lambda \ol{N} \, \ol{B}. 
+  ((\mu_{\ol{N}})_{\ol{B}})|_{\setsem{\Gamma; \Phi, \ol\beta 
+                  \vdash (\Lan^{\ol\alpha}_{\ol K} F)\ol \beta } \rho[\ol{\Phi, \beta := N, B}]} 
+  $$ 
+  is natural in $\ol{N}$ and $\ol{B}$, where $\mu_{\ol{N}}$ is the unique natural transformation such that
+  $$
+  (\setsem{\Gamma; \emptyset \,|\, \Delta \vdash \eta : \Nat^{\Phi, \ol\alpha} \, F \,\, G[\ol{\beta := K}]}\rho \, d)_{\ol{N}}
+  = \mu_{\ol{N}} \, \ol{\setsem{\Gamma; \ol\alpha \vdash K} \rho[\ol{\alpha := \_}]} \circ 
+  \epsilon_{\ol{N}}
+  $$
+
+\end{lemma}
+\begin{proof}
+To prove that this family of morphisms is natural, it suffices to show that the family of unrestricted morphisms 
+  $\lambda \ol{N} \, \ol{B}. (\mu_{\ol{N}})_{\ol{B}}$ is natural in $\ol{N}$ and $\ol{B}$.
+Let us give convenient shorthand names to the following functors and natural transformations.
+\begin{align*}
+\ol{K} &= \ol{\lambda \ol M. \oCPOsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := M}]} \\
+F_{\ol N} &= \lambda \ol M. \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\Phi := \ol N][\ol{\alpha := M}] \\
+F_{\ol{N'}} &= \lambda \ol M. \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\Phi := \ol{N'}][\ol{\alpha := M}] \\
+F_{\ol h} &= \lambda \ol M. \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\id_{\rho}[\Phi := \ol{h}][\ol{\alpha := \id_{M}}] \colon F_{\ol{N}} \to F_{\ol{N'}} \\
+G_{\ol N} &= \lambda \ol B. \oCPOsem{\Gamma; \Phi, \ol\beta \vdash G}\rho[\Phi := \ol N][\ol{\beta := B}] \\
+G_{\ol{N'}} &= \lambda \ol B. \oCPOsem{\Gamma; \Phi, \ol\beta \vdash G}\rho[\Phi := \ol{N'}][\ol{\beta := B}] \\
+G_{\ol h} &= \lambda \ol B. \oCPOsem{\Gamma; \Phi, \ol\beta \vdash G}\id_{\rho}[\Phi := \ol{h}][\ol{\beta := \id_{B}}] \colon G_{\ol{N}} \to G_{\ol{N'}} \\
+\eta &= \oCPOsem{\Gamma; \emptyset \,|\, \Delta \vdash \eta : \Nat^{\Phi, \ol\alpha} F\; G[\ol{\beta := K}]} \rho\, d
+\end{align*}
+Moreover, let $|\ol \alpha| = j$ and $|\ol \beta| = |\ol K| = k$.
+
+Notice that \(\eta\) is natural, so that
+\begin{equation*}\label{eq:lan-elim-eta-nat}
+	\eta_{\ol{N'}} \circ F_{\ol h} = G_{\ol h} \ol{K} \circ \eta_{\ol N} : F_{\ol N} \to G_{\ol{N'}} \ol{K}
+\end{equation*}
+The situation is illustrated in the following diagram.
+\[
+\begin{tikzcd}[column sep = huge, row sep = huge]
+\Set^j
+\ar[r, bend left, "{F_{\ol N}}", ""{name=F, below}]
+\ar[r, bend right, "{F_{\ol{N'}}}"{name=F'b, below}, ""{name=F'a, above}]
+\ar[Rightarrow, from=F, to=F'a, "{F_{\ol{h}}}"']
+\ar[dr, bend right, "{\ol K}"']
+&\Set \\
+&\Set^k
+\ar[Rightarrow, from=F'b, "{\eta_{\ol{N'}}}"']
+\ar[Rightarrow, from=F, "{\eta_{\ol N}}" description]
+\ar[u, "{G_{\ol N}}"{name=L, description}]
+\ar[u, bend right = 90, "{G_{\ol{N'}}}"{name=L', right}]
+\ar[Rightarrow, from=L, to=L', "{G_{\ol h}}"]
+\end{tikzcd}
+\]
+
+Let $\epsilon_{\ol N} \colon F_{\ol N} \to (\Lan_{\ol K} F_{\ol N}) K$ and $\epsilon_{\ol{N'}} \colon F_{\ol{N'}} \to (\Lan_{\ol K} F_{\ol{N'}}) \ol{K}$ be the natural transformations associated to the left Kan extensions, and notice that $\epsilon$ is natural, so that
+\begin{equation}\label{eq:lan-elim-epsilon-nat}
+	(\Lan_{\ol K} F_{\ol h}) \ol{K} \circ \epsilon_{\ol N} = \epsilon_{\ol{N'}} \circ F_{\ol h}
+\end{equation}
+The situation is illustrated in the following diagram.
+\[
+\begin{tikzcd}[column sep = huge, row sep = huge]
+\Set^j
+\ar[r, bend left, "{F_{\ol N}}", ""{name=F, below}]
+\ar[r, bend right, "{F_{\ol{N'}}}"{name=F'b, below}, ""{name=F'a, above}]
+\ar[Rightarrow, from=F, to=F'a, "{F_{\ol{h}}}"']
+\ar[dr, bend right, "{\ol K}"']
+&\Set \\
+&\Set^k
+\ar[Rightarrow, from=F'b, "{\epsilon_{\ol{N'}}}"']
+\ar[Rightarrow, from=F, "{\epsilon_{\ol N}}" description]
+\ar[u, "{\Lan_{\ol K} F_{\ol N}}"{name=L, description}]
+\ar[u, bend right = 90, "{\Lan_{\ol K} F_{\ol{N'}}}"{name=L', right}]
+\ar[Rightarrow, from=L, to=L', "{\Lan_K F_{\ol h}}"]
+\end{tikzcd}
+\]
+
+Let $\mu_{\ol N} : \Lan_{\ol K} F_{\ol N} \to G_{\ol{N}}$ be the unique natural transformation such that
+\begin{equation}\label{eq:lan-elim-mu-univ}
+	\mu_{\ol N} \ol{K} \circ \epsilon_{\ol N} = \eta_{\ol N}
+\end{equation}
+and $\mu_{\ol{N'}} : \Lan_{\ol K} F_{\ol{N'}} \to G_{\ol{N'}}$ be the unique natural transformation such that
+\begin{equation}\label{eq:lan-elim-mu'-univ}
+	\mu_{\ol{N'}} \ol{K} \circ \epsilon_{\ol{N'}} = \eta_{\ol{N'}}
+\end{equation}
+The situation is illustrated in the following diagrams.
+\[
+\begin{tikzcd}[column sep = huge, row sep = huge]
+\Set^j
+\ar[r, "{F_{\ol N}}"{name=Fa}, ""{name=F, below}]
+\ar[dr, "{\ol K}"']
+&\Set \\
+&\Set^k
+\ar[Rightarrow, from=F, "{\epsilon_{\ol N}}" description]
+\ar[u, "{\Lan_{\ol K} F_{\ol N}}"{name=L, description}]
+\ar[u, bend right = 90, "{G_{\ol N}}"{name=G, right}, ""'{name=Gr, right}]
+\ar[Rightarrow, bend left = 90, from=Fa, to=Gr, "{\eta_{\ol{N}}}"]
+\ar[Rightarrow, from=L, to=G, "{\mu_{\ol N}}"]
+\end{tikzcd}
+%%%
+\begin{tikzcd}[column sep = huge, row sep = huge]
+\Set^j
+\ar[r, "{F_{\ol{N'}}}"{name=Fa}, ""{name=F, below}]
+\ar[dr, "{\ol K}"']
+&\Set \\
+&\Set^k
+\ar[Rightarrow, from=F, "{\epsilon_{\ol{N'}}}" description]
+\ar[u, "{\Lan_{\ol K} F_{\ol{N'}}}"{name=L, description}]
+\ar[u, bend right = 90, "{G_{\ol{N'}}}"{name=G, right}, ""'{name=Gr, right}]
+\ar[Rightarrow, bend left = 90, from=Fa, to=Gr, "{\eta_{\ol{N'}}}"]
+\ar[Rightarrow, from=L, to=G, "{\mu_{\ol{N'}}}"]
+\end{tikzcd}
+\]
+
+Notice that, by the universal property of the left Kan extension, there exists a unique $\theta : \Lan_K F_{\ol{N}} \to G_{\ol{N'}}$ such that
+\[
+	\theta \ol{K} \circ \epsilon_{\ol N} = \eta_{\ol{N'}} \circ F_{\ol h} = G_{\ol h} \ol{K} \circ \eta_{\ol N}
+\]
+We shall now show that both $G_{\ol h} \circ \mu_{\ol N}$ and $\mu_{\ol{N'}} \circ \Lan_{\ol K} F_{\ol{h}}$ satisfy this universal property, and thus they are equal to each other.
+\[
+	(G_{\ol h} \circ \mu_{\ol N}) \ol{K} \circ \epsilon_{\ol N}
+	= G_{\ol h}\ol{K} \circ \mu_{\ol N} \ol{K} \circ \epsilon_{\ol N}
+	= G_{\ol h}\ol{K} \circ \eta_{\ol N}
+\]
+where the second equality is by Equation~\ref{eq:lan-elim-mu-univ}
+and
+\[
+	(\mu_{\ol{N'}} \circ \Lan_{\ol K} F_{\ol{h}}) \ol{K} \circ \epsilon_{\ol N}
+	= \mu_{\ol{N'}}\ol{K} \circ \Lan_{\ol K} F_{\ol{h}}\ol{K} \circ \epsilon_{\ol N}
+	= \mu_{\ol{N'}}\ol{K} \circ \epsilon_{\ol{N'}} \circ F_{\ol h}
+	= \eta_{\ol{N'}} \circ F_{\ol h}
+\]
+where the second equality is by Equation~\ref{eq:lan-elim-epsilon-nat} and the third equality is by Equation~\ref{eq:lan-elim-mu'-univ}.
+
+\end{proof}
+
+\subsection{$\relsem{\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol
+      K}_F \eta : Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}
+    F)\ol \beta\; G}$ is well-defined}
+
+\section{Basic Properties of Term Interpretations}
+
+The interpretations in Definition~\ref{def:set-interp} respect
+weakening, i.e., a term and its weakenings all have the same set and
+relational interpretations. Specifically, for any $\rho \in \setenv$,
+$\setsem{\Gamma;\Phi \,|\, \Delta, x : F \vdash t : G}\rho =
+(\setsem{\Gamma;\Phi \,|\, \Delta \vdash t : G}\rho) \circ
+\pi_{\Delta}$, where $\pi_{\Delta}$ is the projection
+$\setsem{\Gamma;\Phi \vdash \Delta, x : F} \to
+\setsem{\Gamma;\Phi \vdash \Delta}$. A similar result holds for
+relational interpretations.
+
+Moreover, if
+  $\Gamma,\alpha;\Phi \,|\, \Delta \vdash t : F$ and
+  $\Gamma;\Phi,\alpha \,|\, \Delta \vdash t' : F$ and $\Gamma;\Phi
+  \vdash G$ then
+\begin{itemize}
+\item $\setsem{\Gamma;\Phi \,|\, \Delta[\alpha := G] \vdash
+  t[\alpha := G] : F[\alpha := G]}\rho =
+  \setsem{\Gamma,\alpha;\Phi \,|\, \Delta \vdash t : F }\rho [
+    \alpha := \setsem{\Gamma;\Phi\vdash G}\rho ]$
+%\item $\relsem{\Gamma;\Phi \,|\, \Delta[\alpha := \sigma] \vdash
+%  t[\alpha := \sigma] : \tau[\alpha := \sigma]}\rho =
+%  \relsem{\Gamma,\alpha;\Phi \,|\, \Delta \vdash t : \tau }\rho [
+%    \alpha := \relsem{\Gamma;\Phi\vdash\sigma}\rho ]$
+\item $\setsem{\Gamma;\Phi \,|\, \Delta[\alpha := G] \vdash
+  t'[\alpha := G] : F[\alpha := G]}\rho =
+  \setsem{\Gamma;\Phi,\alpha \,|\, \Delta \vdash t' : F }\rho [
+    \alpha := \setsem{\Gamma;\Phi\vdash G}\rho ]$
+%\item $\relsem{\Gamma;\Phi \,|\, \Delta[\alpha := \sigma] \vdash
+%  t'[\alpha := \sigma] : \tau[\alpha := \sigma]}\rho =
+%  \relsem{\Gamma;\Phi,\alpha \,|\, \Delta \vdash t' : \tau }\rho [
+%    \alpha := \relsem{\Gamma;\Phi\vdash\sigma}\rho ]$
+\end{itemize}
+and if $\Gamma;\Phi \,|\, \Delta, x: G \vdash t : F$ and
+$\Gamma;\Phi \,|\, \Delta \vdash s : G$ then
+\begin{itemize}
+\item $\lambda A.\, \setsem{\Gamma;\Phi \,|\, \Delta \vdash t[x := s]
+  : F }\rho \, A = \lambda A.\,\setsem{\Gamma;\Phi \,|\, \Delta, x:
+  G \vdash t : F}\rho \,(A, \setsem{\Gamma;\Phi \,|\,
+  \Delta\vdash s: G}\rho\, A)$
+%\item $\lambda R.\,\relsem{\Gamma;\Phi \,|\, \Delta \vdash t[x := s] :
+%  \tau }\rho\,R = \lambda R.\,\relsem{\Gamma;\Phi \,|\, \Delta, x:
+%  \sigma \vdash t : \tau}\rho \,(R, \relsem{\Gamma;\Phi \,|\,
+%  \Delta\vdash s: \sigma}\rho\,R)$
+\end{itemize}
+Direct calculation reveals that the set interpretations of terms also
+satisfy
+\begin{itemize}
+\item $\setsem{\Gamma; \Phi~|~\Delta \vdash
+  (L_{\ol{\alpha}}x.t)_{\ol{K}}s} = \setsem{\Gamma; \Phi~|~\Delta
+  \vdash t [\ol{\alpha := K}][x := s]}$
+\end{itemize}
+Term extensionality with respect to types and terms --- i.e.,
+$\setsem{\Gamma;\Phi\vdash (L_\alpha x.t)_\alpha \top : F} =
+\setsem{\Gamma;\Phi \vdash t : F}$ and $\setsem{\Gamma;\Phi\vdash
+  (L_\alpha x.t)_\alpha x : F} = \setsem{\Gamma;\Phi \vdash t : F}$
+%$\relsem{\Gamma;\Phi\vdash (L_\alpha x.t)_\alpha t} =
+%\relsem{\Gamma;\Phi \vdash t}$, as well as
+%term extensionality
+%and $\relsem{\Gamma;\Phi\vdash
+%  (L_\alpha x.t)_\alpha \top} = \relsem{\Gamma;\Phi \vdash t}$
+--- follow (when both sides of the equations are defined). Similar
+properties hold for relational interpretations.
+
 \bibliography{references}

### 2020-09-30 20-gadts/inf-rules.tex
set sem for Lan elim added

@@
 is defined to be the unique natural transformation such that
-$((\textit{Lan}_{\ol K} \alpha) \,\circ \,\langle K_1,...,K_n \rangle)
-\circ \eta_F = \eta_{F'} \,\circ\, \alpha$. Here, $\eta_F : F \to
-(\textit{Lan}_{\ol K} F) \circ \langle K_1,...,K_n \rangle$ and
-$\eta_{F'} : F' \to (\textit{Lan}_{\ol K} F') \circ \langle
-K_1,...,K_n \rangle$ are the natural transformations associated with
-the functors $\textit{Lan}_{\ol K} F$ and $\textit{Lan}_{\ol K} F'$
-from $\Pi_{i \in \{1,...,n\}} \C_i$ to $\D$, respectively. In $\set$,
-we have the slightly more general property
+\begin{equation}\label{eq:functorial-action-of-lan}
+((\textit{Lan}_{\ol K} \alpha) \,\circ \,\langle K_1,...,K_n \rangle)
+  \circ \eta_F = \eta_{F'} \,\circ\, \alpha
+\end{equation}
+\noindent
+Here, $\eta_F : F \to (\textit{Lan}_{\ol K} F) \circ \langle
+K_1,...,K_n \rangle$ and $\eta_{F'} : F' \to (\textit{Lan}_{\ol K} F')
+\circ \langle K_1,...,K_n \rangle$ are the natural transformations
+associated with the functors $\textit{Lan}_{\ol K} F$ and
+$\textit{Lan}_{\ol K} F'$ from $\Pi_{i \in \{1,...,n\}} \C_i$ to $\D$,
+respectively. In $\set$, we have the slightly more general property
 \begin{equation}\label{eq:ho-functorial-action-lan}
@@
 \ol{B}. (\mu_{\ol{N}})_{\ol{B}} |_{\setsem{\Gamma; \Phi, \ol\beta
-    \vdash (\Lan^{\ol\alpha}_{\ol K} F)\ol \beta } \rho[\ol{\Phi,
-      \beta := N, B}]}$ is natural in both $\ol{N}$ and $\ol{B}$ and
-that it satisfies the condition required for it to be in
-$\setsem{\Gamma; \emptyset \vdash \Nat^{\Phi, \ol\beta}\,
+    \vdash (\Lan^{\ol\alpha}_{\ol K} F)\ol \beta } \rho[\Phi, \ol
+    \beta := \ol N, \ol B]}$ is natural in both $\ol{N}$ and $\ol{B}$,
+as well as that it satisfies the additional condition required for it
+to be in $\setsem{\Gamma; \emptyset \vdash \Nat^{\Phi, \ol\beta}\,
   (\Lan^{\ol\alpha}_{\ol K} F)\ol \beta\; G} \rho$, namely, if $\Phi =
@@
     \end{equation}
+
+\begin{lemma}
+For all $\rho : \setenv$ and $d : \setsem{\Gamma; \emptyset \vdash
+  \Delta} \rho$, the family of morphisms
+\[ \lambda \ol{N} \, \ol{B}. 
+  ((\mu_{\ol{N}})_{\ol{B}})|_{\setsem{\Gamma; \Phi, \ol\beta \vdash
+    (\Lan^{\ol\alpha}_{\ol K} F)\ol \beta } \rho[\Phi, \ol \beta :=
+    \ol N, \ol B]} \]
+is natural in $\ol{N}$ and $\ol{B}$.
+\end{lemma}
+\begin{proof}
+To prove that this family of morphisms is natural, it suffices to show
+that the family of unrestricted morphisms $\lambda \ol{N} \,
+\ol{B}. (\mu_{\ol{N}})_{\ol{B}}$ is natural in $\ol{N}$ and $\ol{B}$.
+Since each $\mu_{\ol N}$ is already natural in $\ol B$ we need only
+show that $\mu$ is natural in $\ol N$. To this end, let $\ol{h : N
+  \to N'}$, and define functors and natural transformations as
+follows:
+\begin{align*}
+\ol{K} &= \ol{\lambda \ol M. \oCPOsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := M}]} \\
+F_{\ol N} &= \lambda \ol M. \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\Phi := \ol N][\ol{\alpha := M}] \\
+F_{\ol{N'}} &= \lambda \ol M. \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\Phi := \ol{N'}][\ol{\alpha := M}] \\
+F_{\ol h} &= \lambda \ol M. \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\id_{\rho}[\Phi := \ol{h}][\ol{\alpha := \id_{M}}] \colon F_{\ol{N}} \to F_{\ol{N'}} \\
+G_{\ol N} &= \lambda \ol B. \oCPOsem{\Gamma; \Phi, \ol\beta \vdash G}\rho[\Phi := \ol N][\ol{\beta := B}] \\
+G_{\ol{N'}} &= \lambda \ol B. \oCPOsem{\Gamma; \Phi, \ol\beta \vdash G}\rho[\Phi := \ol{N'}][\ol{\beta := B}] \\
+G_{\ol h} &= \lambda \ol B. \oCPOsem{\Gamma; \Phi, \ol\beta \vdash G}\id_{\rho}[\Phi := \ol{h}][\ol{\beta := \id_{B}}] \colon G_{\ol{N}} \to G_{\ol{N'}} \\
+\eta &= \oCPOsem{\Gamma; \emptyset \,|\, \Delta \vdash \eta : \Nat^{\Phi, \ol\alpha} F\; G[\ol{\beta := K}]} \rho\, d
+\end{align*}
+First note that \(\eta\) is natural, so that
+\begin{equation}\label{eq:lan-elim-eta-nat}
+\eta_{\ol{N'}} \circ F_{\ol h} = G_{\ol h} \ol{K} \circ \eta_{\ol N} :
+F_{\ol N} \to G_{\ol{N'}} \ol{K}
+\end{equation}
+The situation is illustrated in the following diagram:
+\[
+\begin{tikzcd}[column sep = huge, row sep = huge]
+\Set^{|\ol \alpha|}
+\ar[r, bend left, "{F_{\ol N}}", ""{name=F, below}]
+\ar[r, bend right, "{F_{\ol{N'}}}"{name=F'b, below}, ""{name=F'a, above}]
+\ar[Rightarrow, from=F, to=F'a, "{F_{\ol{h}}}"']
+\ar[dr, bend right, "{\ol K}"']
+&\Set \\
+&\Set^{|\ol K|}
+\ar[Rightarrow, from=F'b, "{\eta_{\ol{N'}}}"']
+\ar[Rightarrow, from=F, "{\eta_{\ol N}}" description]
+\ar[u, "{G_{\ol N}}"{name=L, description}]
+\ar[u, bend right = 90, "{G_{\ol{N'}}}"{name=L', right}]
+\ar[Rightarrow, from=L, to=L', "{G_{\ol h}}"]
+\end{tikzcd}
+\]
+Now, note that the collection $\{\epsilon_{\ol N}\}$ actually defines
+a natural transformation in $\ol N$ since
+\begin{equation}\label{eq:lan-elim-epsilon-nat}
+	(\Lan_{\ol K} F_{\ol h}) \ol{K} \circ \epsilon_{\ol N} = \epsilon_{\ol{N'}} \circ F_{\ol h}
+\end{equation}
+holds by~\eqref{eq:functorial-action-of-lan}. That is, we have
+\[
+\begin{tikzcd}[column sep = huge, row sep = huge]
+\Set^j
+\ar[r, bend left, "{F_{\ol N}}", ""{name=F, below}]
+\ar[r, bend right, "{F_{\ol{N'}}}"{name=F'b, below}, ""{name=F'a, above}]
+\ar[Rightarrow, from=F, to=F'a, "{F_{\ol{h}}}"']
+\ar[dr, bend right, "{\ol K}"']
+&\Set \\
+&\Set^k
+\ar[Rightarrow, from=F'b, "{\epsilon_{\ol{N'}}}"']
+\ar[Rightarrow, from=F, "{\epsilon_{\ol N}}" description]
+\ar[u, "{\Lan_{\ol K} F_{\ol N}}"{name=L, description}]
+\ar[u, bend right = 90, "{\Lan_{\ol K} F_{\ol{N'}}}"{name=L', right}]
+\ar[Rightarrow, from=L, to=L', "{\Lan_K F_{\ol h}}"]
+\end{tikzcd}
+\]
+Since $\mu_{\ol N} : \Lan_{\ol K}\, F_{\ol N} \to G_{\ol{N}}$ is the unique
+natural transformation such that
+\begin{equation}\label{eq:lan-elim-mu-univ}
+\mu_{\ol N} \ol{K} \circ \epsilon_{\ol N} = \eta_{\ol N}
+\end{equation}
+and since $\mu_{\ol{N'}} : \Lan_{\ol K}\, F_{\ol{N'}} \to G_{\ol{N'}}$ is
+the unique natural transformation such that
+\begin{equation}\label{eq:lan-elim-mu'-univ}
+\mu_{\ol{N'}} \ol{K} \circ \epsilon_{\ol{N'}} = \eta_{\ol{N'}}
+\end{equation}
+we have the following commuting diagrams:
+\[
+\begin{tikzcd}[column sep = huge, row sep = huge]
+\Set^{|\ol \alpha|}
+\ar[r, "{F_{\ol N}}"{name=Fa}, ""{name=F, below}]
+\ar[dr, "{\ol K}"']
+&\Set \\
+&\Set^{|\ol K|}
+\ar[Rightarrow, from=F, "{\epsilon_{\ol N}}" description]
+\ar[u, "{\Lan_{\ol K} F_{\ol N}}"{name=L, description}]
+\ar[u, bend right = 90, "{G_{\ol N}}"{name=G, right}, ""'{name=Gr, right}]
+\ar[Rightarrow, bend left = 90, from=Fa, to=Gr, "{\eta_{\ol{N}}}"]
+\ar[Rightarrow, from=L, to=G, "{\mu_{\ol N}}"]
+\end{tikzcd}
+%%%
+\begin{tikzcd}[column sep = huge, row sep = huge]
+\Set^{|\ol \alpha|}
+\ar[r, "{F_{\ol{N'}}}"{name=Fa}, ""{name=F, below}]
+\ar[dr, "{\ol K}"']
+&\Set \\
+&\Set^{|\ol K|}
+\ar[Rightarrow, from=F, "{\epsilon_{\ol{N'}}}" description]
+\ar[u, "{\Lan_{\ol K} F_{\ol{N'}}}"{name=L, description}]
+\ar[u, bend right = 90, "{G_{\ol{N'}}}"{name=G, right}, ""'{name=Gr, right}]
+\ar[Rightarrow, bend left = 90, from=Fa, to=Gr, "{\eta_{\ol{N'}}}"]
+\ar[Rightarrow, from=L, to=G, "{\mu_{\ol{N'}}}"]
+\end{tikzcd}
+\]
+By the universal property of $\Lan_K \, F_{\ol{N}}$ there exists a
+unique $\theta : \Lan_K \,F_{\ol{N}} \to G_{\ol{N'}}$ such that
+\[\theta \ol{K} \circ \epsilon_{\ol N} = \eta_{\ol{N'}} \circ F_{\ol h} = G_{\ol h} \ol{K} \circ \eta_{\ol N}
+\]
+Here, the last equality is by~\eqref{eq:lan-elim-eta-nat}.
+But $G_{\ol h} \circ \mu_{\ol N}$ satisfies this universal property because
+\[
+	(G_{\ol h} \circ \mu_{\ol N}) \ol{K} \circ \epsilon_{\ol N}
+	= G_{\ol h}\ol{K} \circ \mu_{\ol N} \ol{K} \circ \epsilon_{\ol N}
+	= G_{\ol h}\ol{K} \circ \eta_{\ol N}
+\]
+where the second equality is by~\eqref{eq:lan-elim-mu-univ}, and 
+$\mu_{\ol{N'}} \circ \Lan_{\ol K} F_{\ol{h}}$ satisfies it as well
+because
+\[
+	(\mu_{\ol{N'}} \circ \Lan_{\ol K} F_{\ol{h}}) \ol{K} \circ \epsilon_{\ol N}
+	= \mu_{\ol{N'}}\ol{K} \circ \Lan_{\ol K} F_{\ol{h}}\ol{K} \circ \epsilon_{\ol N}
+	= \mu_{\ol{N'}}\ol{K} \circ \epsilon_{\ol{N'}} \circ F_{\ol h}
+	= \eta_{\ol{N'}} \circ F_{\ol h}
+\]
+Here, the second equality is by~\eqref{eq:lan-elim-epsilon-nat} and
+the third equality is by~\eqref{eq:lan-elim-mu'-univ}. We must
+therefore have that $G_{\ol h} \circ \mu_{\ol N} = \mu_{\ol{N'}}
+\circ \Lan_{\ol K} F_{\ol{h}}$ as desired.
+\end{proof}
 
@@
 
-  {\color{red} START HERE!}
-
 To finish the proof, we would like to use that fact that
@@
   (\mu_{\ol{N^2}})_{\ol{\setsem{\Gamma; \ol\alpha \vdash K} \rho[\ol{\alpha := Z}]}} 
-  ((\epsilon_{\ol{N^2}})_{\ol{Z}} \,\, t_2')) \\
-  & \in 
-  \relsem{\Gamma; \Phi, \ol\alpha \vdash
-        G[\ol{\beta := K}]}\Eq_{\rho}[\Phi, \ol \alpha := \ol N,
-    \ol{\Eq_Z}]
+  ((\epsilon_{\ol{N^2}})_{\ol{Z}} \,\, t_2'))
 \end{align*}
-To show this, we use the condition imposed on $\setsem{\Gamma;
-  \emptyset \,|\, \Delta \vdash \eta : \Nat^{\Phi, \ol\alpha} \, F
-  \,\, G[\ol{\beta := K}]} \rho \, d$ to be in $\setsem{\Gamma;
-  \emptyset \vdash \Nat^{\Phi, \ol\alpha} \, F \,\, G[\ol{\beta :=
-      K}]} \rho$, i.e.,
-
+is in its domain
+\begin{align*}
+&\relsem{\Gamma; \Phi,\ol\beta \vdash G}\Eq_\rho [\Phi, \ol \beta :=
+     \ol N, \ol{\relsem{\Gamma; \ol\alpha \vdash K}\Eq_\rho[\ol{\alpha :=
+          \Eq_Z}]}]\\
+=~~~~&\relsem{\Gamma; \Phi, \ol\alpha \vdash G[\ol{\beta :=
+        K}]}\Eq_{\rho}[\Phi, \ol \alpha := \ol N, \ol{\Eq_Z}]
+\end{align*}
+To do so we note that since $\setsem{\Gamma; \emptyset \,|\, \Delta
+  \vdash \eta : \Nat^{\Phi, \ol\alpha} \, F \,\, G[\ol{\beta := K}]}
+\rho \, d$ is in $\setsem{\Gamma; \emptyset \vdash \Nat^{\Phi,
+    \ol\alpha} \, F \,\, G[\ol{\beta := K}]} \rho$ it must satisfy
     \begin{equation}\label{eq:P1}
     \begin{split}
-      &\forall \ol{N = (N^1, N^2, N^*) : RT_k}, 
-          \forall \ol{M = (M^1, M^2, M^*) : RT_k}.\\
-      &((\setsem{\Gamma; \emptyset \,|\, \Delta \vdash \eta : \Nat^{\Phi, \ol\alpha} \, F \,\, G[\ol{\beta := K}]} \rho \, d)_{\ol{N^1}, \ol{M^1}}, 
-      (\setsem{\Gamma; \emptyset \,|\, \Delta \vdash \eta} : \Nat^{\Phi, \ol\alpha} \, F \,\, G[\ol{\beta := K}] \rho \, d)_{\ol{N^2}, \ol{M^2}}) \\
-      &: \relsem{\Gamma; \Phi, \ol\alpha \vdash F}\Eq_{\rho}[\ol{\Phi, \alpha := N, M}]
-      \rightarrow \relsem{\Gamma; \Phi, \ol\alpha \vdash
-        G[\ol{\beta := K}]}\Eq_{\rho}[\ol{\Phi, \alpha := N, M}]
+      &\forall \ol{N = (N^1, N^2, N^*) : RT_{k_i}}, \forall \ol{M =
+        (M^1, M^2, M^*) : RT_0}.\\ &((\setsem{\Gamma; \emptyset \,|\,
+        \Delta \vdash \eta : \Nat^{\Phi, \ol\alpha} \, F \,\,
+        G[\ol{\beta := K}]} \rho \, d)_{\ol{N^1}, \ol{M^1}},
+      (\setsem{\Gamma; \emptyset \,|\, \Delta \vdash \eta} :
+      \Nat^{\Phi, \ol\alpha} \, F \,\, G[\ol{\beta := K}] \rho \,
+      d)_{\ol{N^2}, \ol{M^2}}) \\ :~~~~ & \relsem{\Gamma; \Phi,
+        \ol\alpha \vdash F}\Eq_{\rho}[\Phi, \ol \alpha := \ol N, \ol
+        M] \rightarrow \relsem{\Gamma; \Phi, \ol\alpha \vdash
+        G[\ol{\beta := K}]}\Eq_{\rho}[\Phi, \ol \alpha := \ol N, \ol
+        M]
     \end{split}
     \end{equation}
-along with the universal property of 
-$\Lan_{\setsem{\Gamma; \ol\alpha \vdash K}\rho {[\ol{\alpha := \_} ]}}
-      \setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho {[\ol{\Phi, \alpha := N^i, \_} ]}$
-      for $i = 1, 2$
-and the fact that $t_1'$ and $t_2'$ are related in 
-$\relsem{\Gamma; \Phi, \ol\alpha \vdash F}\Eq_{\rho}[\ol{\Phi, \alpha := N, \Eq_Z}]$.
-Applying Equation \eqref{eq:P1} with $\ol{M} = \ol{Z}$ 
+together with the universal properties of $\Lan_{\ol{\setsem{\Gamma;
+      \ol\alpha \vdash K}\rho {[\ol{\alpha := \_} ]}}} \setsem{\Gamma;
+  \Phi, \ol\alpha \vdash F}\rho {[\Phi, \ol \alpha := \ol{N^i}, \_ ]}$
+for $i = 1, 2$, and the fact that $(t_1',t_2') \in \relsem{\Gamma;
+  \Phi, \ol\alpha \vdash F}\Eq_{\rho}[\Phi, \ol\alpha := \ol N,
+  \ol{\Eq_Z}]$. Indeed, applying~\eqref{eq:P1} with $\ol{M} = \ol{\Eq_Z}$
 gives (eliding the type of $\eta$)
-    \begin{equation*}
+    \begin{equation}\label{eq:morph}
     \begin{split}
@@
       (\setsem{\Gamma; \emptyset \,|\, \Delta \vdash \eta} \rho \, d)_{\ol{N^2}, \ol{Z}}) \\
-      &: \relsem{\Gamma; \Phi, \ol\alpha \vdash F}\Eq_{\rho}[\ol{\Phi, \alpha := N, \Eq_Z}]
+      &: \relsem{\Gamma; \Phi, \ol\alpha \vdash F}\Eq_{\rho}[\Phi, \ol
+        \alpha := \ol N, \ol{\Eq_Z}]
       \rightarrow \relsem{\Gamma; \Phi, \ol\alpha \vdash
-        G[\ol{\beta := K}]}\Eq_{\rho}[\ol{\Phi, \alpha := N, \Eq_Z}]
+        G[\ol{\beta := K}]}\Eq_{\rho}[\Phi, \ol \alpha := \ol N, \ol{\Eq_Z}]
     \end{split}
-    \end{equation*}
-and since we know 
-$(t_1',t_2') \in
-  \relsem{\Gamma; \Phi, \ol\alpha \vdash F}\Eq_\rho[\ol{\Phi := N}][\ol{\alpha := \Eq_Z}]$, 
-we can apply this morphism
-%%
-to $(t_1', t_2')$:
+    \end{equation}
+Then since $(t_1',t_2') \in \relsem{\Gamma; \Phi, \ol\alpha \vdash
+  F}\Eq_\rho[\Phi := \ol{N}][\ol{\alpha := \Eq_Z}]$, we can apply
+~\eqref{eq:morph} to it to get
 \begin{align*}
@@
 \relsem{\Gamma; \Phi, \ol\alpha \vdash
-        G[\ol{\beta := K}]}\Eq_{\rho}[\ol{\Phi, \alpha := N, \Eq_Z}]
+        G[\ol{\beta := K}]}\Eq_{\rho}[\Phi, \ol \alpha := \ol N, \ol{\Eq_Z}]
 \end{align*}
 %%
-Applying the universal property at component $\ol{Z}$
-$$
+Applying the universal property~\eqref{eq:defn-of-mu-sub-N} at component $\ol{Z}$ gives
+\[
  ((\setsem{\Gamma; \emptyset \,|\, \Delta \vdash \eta}\rho \, d)_{\ol{N^i}})_{\ol{Z}}
@@
   (\epsilon_{\ol{N^i}})_{\ol{Z}}
-$$
-yields 
+\]
+so that
 \begin{align*}
-((\mu_{\ol{N^1}})_{\ol{\setsem{\Gamma; \ol\alpha \vdash K} \rho[\ol{\alpha := Z}]}} 
+\big((\mu_{\ol{N^1}})_{\ol{\setsem{\Gamma; \ol\alpha \vdash K} \rho[\ol{\alpha := Z}]}} 
   ((\epsilon_{\ol{N^1}})_{\ol{Z}} \,\, t_1') &, 
@@
   (\mu_{\ol{N^2}})_{\ol{\setsem{\Gamma; \ol\alpha \vdash K} \rho[\ol{\alpha := Z}]}} 
-  ((\epsilon_{\ol{N^2}})_{\ol{Z}} \,\, t_2')) \\
+  ((\epsilon_{\ol{N^2}})_{\ol{Z}} \,\, t_2')\big) \\
   & \in 
   \relsem{\Gamma; \Phi, \ol\alpha \vdash
-        G[\ol{\beta := K}]}\Eq_{\rho}[\ol{\Phi, \alpha := N, \Eq_Z}] \\
-  & = \relsem{\Gamma; \Phi,\ol\beta \vdash G}\Eq_\rho
-  [\ol{\Phi, \beta := N, \relsem{\Gamma; \ol\alpha \vdash K}\Eq_\rho[\ol{\alpha := \Eq_Z}]}]
+        G[\ol{\beta := K}]}\Eq_{\rho}[\Phi, \ol \alpha := \ol N,\ol{\Eq_Z}] 
+%  & = \relsem{\Gamma; \Phi,\ol\beta \vdash G}\Eq_\rho
+%  [\ol{\Phi, \beta := N, \relsem{\Gamma; \ol\alpha \vdash K}\Eq_\rho[\ol{\alpha := \Eq_Z}]}]
 \end{align*}
-
-Finally, we can apply the morphism 
+\noindent
+Finally, then, we can apply the morphism 
 \begin{align*}
-  \relsem{\Gamma; \Phi,\ol\beta \vdash G}\id_{\Eq_\rho} [\ol{\Phi, \beta := \id_N, f}] 
+  \relsem{\Gamma; \Phi,\ol\beta \vdash G}\id_{\Eq_\rho} [\Phi, \ol
+    \beta := \ol{\id_N}, \ol f] 
   : \,\,
   &\relsem{\Gamma; \Phi,\ol\beta \vdash G}\Eq_\rho
-  [\ol{\Phi, \beta := N, \relsem{\Gamma; \ol\alpha \vdash K}\Eq_\rho[\ol{\alpha := \Eq_Z}]}] \\
+  [\Phi, \ol \beta := \ol N, \ol{\relsem{\Gamma; \ol\alpha \vdash K}\Eq_\rho[\ol{\alpha := \Eq_Z}]}] \\
   \rightarrow \,\,
-  &\relsem{\Gamma; \Phi,\ol\beta \vdash G}\Eq_\rho [\ol{\Phi, \beta := N, B}]
+  &\relsem{\Gamma; \Phi,\ol\beta \vdash G}\Eq_\rho [\Phi, \ol \beta :=
+    \ol N, \ol B]
 \end{align*}
@@
   (\mu_{\ol{N^2}})_{\ol{\setsem{\Gamma; \ol\alpha \vdash K} \rho[\ol{\alpha := Z}]}} 
-  ((\epsilon_{\ol{N^2}})_{\ol{Z}} \,\, t_2'))$
-, which gives
+  ((\epsilon_{\ol{N^2}})_{\ol{Z}} \,\, t_2'))$.
+This gives that
 
 \begin{align*}
-( 
-  &\setsem{\Gamma; \Phi,\ol\beta \vdash G}\id_{\rho} [\ol{\Phi, \beta := \id_{N^1}, \pi_1 f}] 
+\big( 
+  &\setsem{\Gamma; \Phi,\ol\beta \vdash G}\id_{\rho} [\Phi, \ol \beta := \ol{\id_{N^1}}, \ol{\pi_1 f}] 
   ((\mu_{\ol{N^1}})_{\ol{\setsem{\Gamma; \ol\alpha \vdash K} \rho[\ol{\alpha := Z}]}} 
@@
   %%%
-  &\setsem{\Gamma; \Phi,\ol\beta \vdash G}\id_{\rho} [\ol{\Phi, \beta := \id_{N^2}, \pi_2 f}] 
+  &\setsem{\Gamma; \Phi,\ol\beta \vdash G}\id_{\rho} [\Phi, \ol \beta := \o;{\id_{N^2}}, \ol{\pi_2 f}] 
   ((\mu_{\ol{N^2}})_{\ol{\setsem{\Gamma; \ol\alpha \vdash K} \rho[\ol{\alpha := Z}]}} 
-  ((\epsilon_{\ol{N^2}})_{\ol{Z}} \,\, t_2'))) \\
+  ((\epsilon_{\ol{N^2}})_{\ol{Z}} \,\, t_2'))\big) \\
   & \in 
-  \relsem{\Gamma; \Phi, \ol\alpha \vdash
-        G}\Eq_{\rho}[\ol{\Phi, \alpha := N,B}] 
+  \relsem{\Gamma; \Phi, \ol \beta \vdash
+        G}\Eq_{\rho}[\Phi, \ol \beta := \ol N, \ol B] 
 \end{align*}
-and thus 
-$$((\mu_{\ol{N^1}})_{\ol{B^1}} \,\, t_1, 
+i.e., that 
+\[((\mu_{\ol{N^1}})_{\ol{B^1}} \,\, t_1, 
                 (\mu_{\ol{N^2}})_{\ol{B^2}} \,\, t_2)
                 \in \relsem{\Gamma; \Phi, \ol\beta \vdash
-            G}\Eq_{\rho}[\ol{\Phi, \beta := N, B}]$$
-
-\end{proof}
-
-\begin{lemma}
-  For all $\rho : \setenv$ and $d : \setsem{\Gamma; \emptyset \vdash \Delta} \rho$, 
-  the family of morphisms 
-  $$
-  \lambda \ol{N} \, \ol{B}. 
-  ((\mu_{\ol{N}})_{\ol{B}})|_{\setsem{\Gamma; \Phi, \ol\beta 
-                  \vdash (\Lan^{\ol\alpha}_{\ol K} F)\ol \beta } \rho[\ol{\Phi, \beta := N, B}]} 
-  $$ 
-  is natural in $\ol{N}$ and $\ol{B}$, where $\mu_{\ol{N}}$ is the unique natural transformation such that
-  $$
-  (\setsem{\Gamma; \emptyset \,|\, \Delta \vdash \eta : \Nat^{\Phi, \ol\alpha} \, F \,\, G[\ol{\beta := K}]}\rho \, d)_{\ol{N}}
-  = \mu_{\ol{N}} \, \ol{\setsem{\Gamma; \ol\alpha \vdash K} \rho[\ol{\alpha := \_}]} \circ 
-  \epsilon_{\ol{N}}
-  $$
-
-\end{lemma}
-\begin{proof}
-To prove that this family of morphisms is natural, it suffices to show that the family of unrestricted morphisms 
-  $\lambda \ol{N} \, \ol{B}. (\mu_{\ol{N}})_{\ol{B}}$ is natural in $\ol{N}$ and $\ol{B}$.
-Let us give convenient shorthand names to the following functors and natural transformations.
-\begin{align*}
-\ol{K} &= \ol{\lambda \ol M. \oCPOsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := M}]} \\
-F_{\ol N} &= \lambda \ol M. \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\Phi := \ol N][\ol{\alpha := M}] \\
-F_{\ol{N'}} &= \lambda \ol M. \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\Phi := \ol{N'}][\ol{\alpha := M}] \\
-F_{\ol h} &= \lambda \ol M. \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\id_{\rho}[\Phi := \ol{h}][\ol{\alpha := \id_{M}}] \colon F_{\ol{N}} \to F_{\ol{N'}} \\
-G_{\ol N} &= \lambda \ol B. \oCPOsem{\Gamma; \Phi, \ol\beta \vdash G}\rho[\Phi := \ol N][\ol{\beta := B}] \\
-G_{\ol{N'}} &= \lambda \ol B. \oCPOsem{\Gamma; \Phi, \ol\beta \vdash G}\rho[\Phi := \ol{N'}][\ol{\beta := B}] \\
-G_{\ol h} &= \lambda \ol B. \oCPOsem{\Gamma; \Phi, \ol\beta \vdash G}\id_{\rho}[\Phi := \ol{h}][\ol{\beta := \id_{B}}] \colon G_{\ol{N}} \to G_{\ol{N'}} \\
-\eta &= \oCPOsem{\Gamma; \emptyset \,|\, \Delta \vdash \eta : \Nat^{\Phi, \ol\alpha} F\; G[\ol{\beta := K}]} \rho\, d
-\end{align*}
-Moreover, let $|\ol \alpha| = j$ and $|\ol \beta| = |\ol K| = k$.
-
-Notice that \(\eta\) is natural, so that
-\begin{equation*}\label{eq:lan-elim-eta-nat}
-	\eta_{\ol{N'}} \circ F_{\ol h} = G_{\ol h} \ol{K} \circ \eta_{\ol N} : F_{\ol N} \to G_{\ol{N'}} \ol{K}
-\end{equation*}
-The situation is illustrated in the following diagram.
-\[
-\begin{tikzcd}[column sep = huge, row sep = huge]
-\Set^j
-\ar[r, bend left, "{F_{\ol N}}", ""{name=F, below}]
-\ar[r, bend right, "{F_{\ol{N'}}}"{name=F'b, below}, ""{name=F'a, above}]
-\ar[Rightarrow, from=F, to=F'a, "{F_{\ol{h}}}"']
-\ar[dr, bend right, "{\ol K}"']
-&\Set \\
-&\Set^k
-\ar[Rightarrow, from=F'b, "{\eta_{\ol{N'}}}"']
-\ar[Rightarrow, from=F, "{\eta_{\ol N}}" description]
-\ar[u, "{G_{\ol N}}"{name=L, description}]
-\ar[u, bend right = 90, "{G_{\ol{N'}}}"{name=L', right}]
-\ar[Rightarrow, from=L, to=L', "{G_{\ol h}}"]
-\end{tikzcd}
-\]
-
-Let $\epsilon_{\ol N} \colon F_{\ol N} \to (\Lan_{\ol K} F_{\ol N}) K$ and $\epsilon_{\ol{N'}} \colon F_{\ol{N'}} \to (\Lan_{\ol K} F_{\ol{N'}}) \ol{K}$ be the natural transformations associated to the left Kan extensions, and notice that $\epsilon$ is natural, so that
-\begin{equation}\label{eq:lan-elim-epsilon-nat}
-	(\Lan_{\ol K} F_{\ol h}) \ol{K} \circ \epsilon_{\ol N} = \epsilon_{\ol{N'}} \circ F_{\ol h}
-\end{equation}
-The situation is illustrated in the following diagram.
-\[
-\begin{tikzcd}[column sep = huge, row sep = huge]
-\Set^j
-\ar[r, bend left, "{F_{\ol N}}", ""{name=F, below}]
-\ar[r, bend right, "{F_{\ol{N'}}}"{name=F'b, below}, ""{name=F'a, above}]
-\ar[Rightarrow, from=F, to=F'a, "{F_{\ol{h}}}"']
-\ar[dr, bend right, "{\ol K}"']
-&\Set \\
-&\Set^k
-\ar[Rightarrow, from=F'b, "{\epsilon_{\ol{N'}}}"']
-\ar[Rightarrow, from=F, "{\epsilon_{\ol N}}" description]
-\ar[u, "{\Lan_{\ol K} F_{\ol N}}"{name=L, description}]
-\ar[u, bend right = 90, "{\Lan_{\ol K} F_{\ol{N'}}}"{name=L', right}]
-\ar[Rightarrow, from=L, to=L', "{\Lan_K F_{\ol h}}"]
-\end{tikzcd}
-\]
-
-Let $\mu_{\ol N} : \Lan_{\ol K} F_{\ol N} \to G_{\ol{N}}$ be the unique natural transformation such that
-\begin{equation}\label{eq:lan-elim-mu-univ}
-	\mu_{\ol N} \ol{K} \circ \epsilon_{\ol N} = \eta_{\ol N}
-\end{equation}
-and $\mu_{\ol{N'}} : \Lan_{\ol K} F_{\ol{N'}} \to G_{\ol{N'}}$ be the unique natural transformation such that
-\begin{equation}\label{eq:lan-elim-mu'-univ}
-	\mu_{\ol{N'}} \ol{K} \circ \epsilon_{\ol{N'}} = \eta_{\ol{N'}}
-\end{equation}
-The situation is illustrated in the following diagrams.
-\[
-\begin{tikzcd}[column sep = huge, row sep = huge]
-\Set^j
-\ar[r, "{F_{\ol N}}"{name=Fa}, ""{name=F, below}]
-\ar[dr, "{\ol K}"']
-&\Set \\
-&\Set^k
-\ar[Rightarrow, from=F, "{\epsilon_{\ol N}}" description]
-\ar[u, "{\Lan_{\ol K} F_{\ol N}}"{name=L, description}]
-\ar[u, bend right = 90, "{G_{\ol N}}"{name=G, right}, ""'{name=Gr, right}]
-\ar[Rightarrow, bend left = 90, from=Fa, to=Gr, "{\eta_{\ol{N}}}"]
-\ar[Rightarrow, from=L, to=G, "{\mu_{\ol N}}"]
-\end{tikzcd}
-%%%
-\begin{tikzcd}[column sep = huge, row sep = huge]
-\Set^j
-\ar[r, "{F_{\ol{N'}}}"{name=Fa}, ""{name=F, below}]
-\ar[dr, "{\ol K}"']
-&\Set \\
-&\Set^k
-\ar[Rightarrow, from=F, "{\epsilon_{\ol{N'}}}" description]
-\ar[u, "{\Lan_{\ol K} F_{\ol{N'}}}"{name=L, description}]
-\ar[u, bend right = 90, "{G_{\ol{N'}}}"{name=G, right}, ""'{name=Gr, right}]
-\ar[Rightarrow, bend left = 90, from=Fa, to=Gr, "{\eta_{\ol{N'}}}"]
-\ar[Rightarrow, from=L, to=G, "{\mu_{\ol{N'}}}"]
-\end{tikzcd}
-\]
-
-Notice that, by the universal property of the left Kan extension, there exists a unique $\theta : \Lan_K F_{\ol{N}} \to G_{\ol{N'}}$ such that
-\[
-	\theta \ol{K} \circ \epsilon_{\ol N} = \eta_{\ol{N'}} \circ F_{\ol h} = G_{\ol h} \ol{K} \circ \eta_{\ol N}
-\]
-We shall now show that both $G_{\ol h} \circ \mu_{\ol N}$ and $\mu_{\ol{N'}} \circ \Lan_{\ol K} F_{\ol{h}}$ satisfy this universal property, and thus they are equal to each other.
-\[
-	(G_{\ol h} \circ \mu_{\ol N}) \ol{K} \circ \epsilon_{\ol N}
-	= G_{\ol h}\ol{K} \circ \mu_{\ol N} \ol{K} \circ \epsilon_{\ol N}
-	= G_{\ol h}\ol{K} \circ \eta_{\ol N}
-\]
-where the second equality is by Equation~\ref{eq:lan-elim-mu-univ}
-and
-\[
-	(\mu_{\ol{N'}} \circ \Lan_{\ol K} F_{\ol{h}}) \ol{K} \circ \epsilon_{\ol N}
-	= \mu_{\ol{N'}}\ol{K} \circ \Lan_{\ol K} F_{\ol{h}}\ol{K} \circ \epsilon_{\ol N}
-	= \mu_{\ol{N'}}\ol{K} \circ \epsilon_{\ol{N'}} \circ F_{\ol h}
-	= \eta_{\ol{N'}} \circ F_{\ol h}
-\]
-where the second equality is by Equation~\ref{eq:lan-elim-epsilon-nat} and the third equality is by Equation~\ref{eq:lan-elim-mu'-univ}.
-
+            G}\Eq_{\rho}[\Phi, \ol \beta := \ol N, \ol B]\]
 \end{proof}

### 2020-10-01 20-gadts/inf-rules.tex
added Lan-intro-interp to inf-rules.tex

@@
 \AXC{$\ol{\Gamma; \ol{\alpha} \vdash K}$}
-%\AXC{$\ol{\Gamma;\Phi \vdash A} \hspace*{0.3in} \Gamma;\Phi~|~\Delta \vdash t : F[\ol{\alpha :=A}]$}
-\BIC{$\Gamma;\emptyset~|~\emptyset \vdash \int_{\ol K,F} :
-\Nat^{\Phi,\ol\alpha}\,F\, (\Lan^{\ol\alpha}_{\ol K}\,  F)\ol{K}$}  
+\AXC{$\ol{\Gamma; \Phi \vdash A}$}
+\AXC{$\Gamma; \Phi~|~\Delta \vdash t : F[\ol{\alpha := A}]$}
+\QIC{$\Gamma;\Phi~|~\Delta \vdash \int_{\ol K,F} t : (\Lan^{\ol\alpha}_{\ol{K}}\,  F)\, \ol{K[\ol{\alpha := A}]}$}
 \DisplayProof
@@
 \vspace*{0.2in}
+
+\new{Note that we cannot have
+\[
+\AXC{$\Gamma; \Phi,\ol{\alpha}\vdash F$} \AXC{$\ol{\Gamma; \ol{\alpha}
+    \vdash K}$} \BIC{$\Gamma;\emptyset~|~\emptyset \vdash \int_{\ol
+    K,F} : \Nat^{\Phi,\ol\alpha}\,F\, (\Lan^{\ol\alpha}_{\ol K}\,
+  F)\ol{K}$} \DisplayProof\] We cannot prove the additional property
+needed to define an object semantics for this rule because we can only
+introduce components of the natural transformation at representable
+types, as in the rule above. We cannot produce the entire natural
+transformation required in the conclusion of the alternative rule
+here. This makes sense because not all parts of the functors
+underlying GADTs are representable in syntax, which in turn means that
+not all components of the associated natural transformations are
+representable in syntax.}
 
@@
 \mbox{ is as above }\vspace*{0.05in} \\
-\setsem{\Gamma;\emptyset~|~\emptyset \vdash \int_{\ol K,F}
-: \Nat^{\Phi,\ol\alpha}\,F\, (\Lan^{\ol\alpha}_{\ol K}\, F)\ol{K}}\rho
-& = & \vspace*{0.05in}\\
-\relsem{\Gamma;\emptyset~|~\emptyset \vdash \int_{\ol K,F} :
-  \Nat^{\Phi,\ol\alpha}\,F\, (\Lan^{\ol\alpha}_{\ol K}\,
-  F)\ol{K}}\rho\,d & = & \vspace*{0.05in} \\
+\setsem{\Gamma;\Phi~|~\Delta \vdash \int_{\ol K,F} t :
+  (\Lan^{\ol\alpha}_{\ol{K}}\,  F)\, \ol{K[\ol{\alpha := A}]}}\rho\;d
+& = & = \iota_{\ol{\oCPOsem{\Gamma;\Phi \vdash A}\rho},
+  \ol{\id_{\oCPOsem{\Gamma; \Phi \vdash K[\ol{\alpha := A}]}\rho}}}
+\oCPOsem{\Gamma; \Phi~|~\Delta \vdash t : F[\ol{\alpha := A}]} \rho\, 
+d \vspace*{0.05in}\\
+\relsem{\Gamma;\Phi~|~\Delta \vdash \int_{\ol K,F} t :
+  (\Lan^{\ol\alpha}_{\ol{K}}\,  F)\, \ol{K[\ol{\alpha := A}]}}\rho\;d
+& = & \vspace*{0.05in} \\
 \setsem{\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol
@@
 
-\new{In the fourth-to-last clause of Definition~\ref{def:set-interp},...}
+\new{In the fourth-to-last clause of Definition~\ref{def:set-interp},
+  $\iota_{Z, f}$ is the injection into
+\[\colim{\ol{Z : \Set_0},\, \ol{f : \oCPOsem{\Gamma; \ol\alpha \vdash K}
+    \rho[\ol{\alpha := Z}] \to \oCPOsem{\Gamma; \Phi \vdash
+      K[\ol{\alpha := A}]} \rho}}{\,\oCPOsem{\Gamma; \Phi, \ol\alpha
+    \vdash F} \rho[\ol{\alpha := Z}]}\]
+
+}
 
@@
 
-\new{In the final last clause of Definition~\ref{def:set-interp},...}
+\new{In the final clause of Definition~\ref{def:set-interp},...}
 
@@
 
-\subsection{$\setsem{\Gamma;\emptyset~|~\emptyset \vdash \int_{\ol K,F}
-: \Nat^{\Phi,\ol\alpha}\,F\, (\Lan^{\ol\alpha}_{\ol K}\, F)\ol{K}}$ is
+\subsection{Object interpretations of $\Lan$-introduction
+  are well-defined}
+
+We must show that, for every $\rho : \setenv$ and $d :
+\oCPOsem{\Gamma;\Phi \vdash \Delta}\rho$,
+\[\oCPOsem{\Gamma;\Phi~|~\Delta \vdash \int_{\ol K,F} t :
+  (\Lan^{\ol\alpha}_{\ol{K}}\, F)\, \ol{K[\ol{\alpha := A}]}}\rho\, d
+\in \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol\alpha}_{\ol{K}}\, F)\,
+  \ol{K[\ol{\alpha := A}]}}\rho\] i.e., that there exist $\ol{Z :
+  \oCPO_0}$, $t_1', t_2' : \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash
+  F}\rho[\ol{\alpha := Z}]$, and
+\[\ol{f : \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\Eq_{\rho}[\ol{\alpha
+      := \Eq_Z}] \to \oCPORsem{\Gamma; \Phi \vdash K[\ol{\alpha :=
+        A}]}\Eq_{\rho}} \]
+such that
+\[\begin{array}{rcl}
+\iota_{\ol{Z},\ol{\pi_1 f}}t_1' &=& \iota_{\ol{\oCPOsem{\Gamma;\Phi \vdash A}\rho}, \ol{\id_{\oCPOsem{\Gamma; \Phi \vdash K[\ol{\alpha := A}]}\rho}}}\,
+\oCPOsem{\Gamma; \Phi~|~\Delta \vdash t : F[\ol{\alpha := A}]} \rho\, d \\[2ex]
+\iota_{\ol{Z}, \ol{\pi_2 f}}t_2' &=& \iota_{\ol{\oCPOsem{\Gamma;\Phi \vdash A}\rho}, \ol{\id_{\oCPOsem{\Gamma; \Phi \vdash K[\ol{\alpha := A}]}\rho}}}\,
+\oCPOsem{\Gamma; \Phi~|~\Delta \vdash t : F[\ol{\alpha := A}]} \rho\,
+d\\[2ex]
+(t_1',t_2') &\in& \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash
+  F}\Eq_{\rho}[\ol{\alpha := \Eq_Z}] \;=\; \oCPORsem{\Gamma;
+  \Phi, \ol\alpha \vdash F}\Eq_{\rho[\ol{\alpha := Z}]}
+\end{array}\]
+Let $\ol{Z = \oCPOsem{\Gamma; \Phi \vdash A} \rho}$, $t_1' = t_2' =
+\oCPOsem{\Gamma; \Phi~|~\Delta \vdash t : F[\ol{\alpha := A}]} \rho\,
+d$, and $\ol{f = \id_{\oCPORsem{\Gamma; \ol\alpha \vdash K[\ol{\alpha
+          := A}]}\Eq_{\rho}}}$.  Then~\eqref{eq:subs-var} gives that
+$t_1', t_2' : \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash
+  F}\rho[\ol{\alpha := Z}]$ as required, as well as that
+$\oCPORsem{\Gamma; \ol{\alpha} \vdash K}\Eq_{\rho}[\ol{\alpha :=
+    \Eq_{\oCPOsem{\Gamma; \Phi \vdash A} \rho}}] = \oCPORsem{\Gamma;
+  \Phi \vdash K[\ol{\alpha := A}]}\Eq_{\rho}$. In light of the latter
+we take $f$ to be
+\[\ol{ \id_{\oCPORsem{\Gamma; \ol\alpha \vdash K[\ol{\alpha :=
+          A}]}\Eq_{\rho}} : \oCPORsem{\Gamma; \ol{\alpha} \vdash
+    K}\Eq_{\rho}[\ol{\alpha := \Eq_{\oCPOsem{\Gamma; \Phi \vdash A}
+        \rho}}] \to \oCPORsem{\Gamma; \Phi \vdash K[\ol{\alpha :=
+        A}]}\Eq_{\rho} }\] Then the first two properties above hold by
+construction, and the third holds because $t_1' = t_2'$ and because
+$\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\Eq_{\rho[\ol{\alpha :=
+      Z}]} = \Eq_{\setsem{\Gamma; \Phi, \ol\alpha \vdash
+    F}\rho[\ol{\alpha := Z}]}$ by the IEL.
+
+\subsection{Relational interpretations of $\Lan$-introduction are
   well-defined}
 
-\subsection{$\relsem{\Gamma;\emptyset~|~\emptyset \vdash \int_{\ol K,F}
-: \Nat^{\Phi,\ol\alpha}\,F\, (\Lan^{\ol\alpha}_{\ol K}\, F)\ol{K}}$ is
-  well-defined}
-
-\subsection{$\setsem{\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol
-      K}_F \eta : Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}
-    F)\ol \beta\; G}$ is well-defined} We must show that, for every
+\subsection{Object interpretations of $\Lan$-elimination are
+  well-defined} We must show that, for every 
 $\rho : \setenv$ and $d : \setsem{\Gamma; \emptyset \vdash \Delta}
@@
 
-\subsection{$\relsem{\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol
-      K}_F \eta : Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}
-    F)\ol \beta\; G}$ is well-defined}
+\subsection{Relational interpretations of $\Lan$-elimination are
+  well-defined}
 

### 2020-10-01 20-gadts/inf-rules.tex
repair small error
repaired small errors

@@
 \vdash H$, and $\ol{\Gamma;\Phi \vdash K}$, then
-{\color{red} WE NEED TO CHECK LAN CASES HERE!}
 \begin{gather}
@@
 
-\new{In the fourth-to-last clause of Definition~\ref{def:set-interp},
+In the fourth-to-last clause of Definition~\ref{def:set-interp},
   $\iota_{Z, f}$ is the injection into
@@
     \vdash F} \rho[\ol{\alpha := Z}]}\]
-
-}
 

### 2020-12-09 free-theorems-nesteds/20-mscs.tex
added nested types journal paper

--- 2020-07-01 free-theorems-nesteds/paper5.tex
@@
 for nested data types. Rather than representing nested types via their
-Church encodings in a higher-kinded extension of System F, we adopt
-a functional programming perspective and design a polymorphic calculus
-that provides primitives for constructing a robust class of nested
-types directly as fixpoints.
-%\`{a} la Pitts' PolyPCF.
-At the term level, our calculus supports
-%the usual
-pattern matching, map functions, and stylized recursion via folds for
-all such types.
-%Using a variation on Atkey's 2012 construction for $F_\omega$,
-%
-%of a parametric model
-%that explicitly constructs equality operators rather than merely
-%postulating them,
-Our main technical contribution is the construction of a parametric
-model for this calculus. This is both delicate and challenging.  To
-establish a suitable Identity Extension Lemma for our calculus, extra
-care must be taken in the formation of fixpoint types; all ADTs and
-all (truly) nested types appearing in the literature can nevertheless
-be expressed in our calculus. In addition, to ensure the existence of
-semantic fixpoints interpreting nested types, cocontinuity conditions
-on the functors interpreting all of the types in our calculus must
-be threaded throughout our model construction.  Overall, this paper
-provides a blueprint for deriving principled parametricity results for
-practical programming languages, such as Haskell and Agda, that
-directly construct nested types as fixpoints rather than representing
-them via Church encodings.
-%encoding them.
-It also shows how parametric models can derive useful free
-theorems for programs over nested types. Like their counterparts for
-ADTs, these free theorems
-%our model derives
-provide
-%, in particular,
-a principled alternative to proving program properties via induction,
-e.g., using the deep induction rules for nested types recently
-developed by Johann and Polonsky.
-%We derive a number of free theorems for nested types.
-%To illustrate, we derive a number of free theorems.
-Our approach is illustrated with several examples. 
+Church encodings in a higher-kinded or dependently typed extension of
+System F, we adopt a functional programming perspective and design a
+Hindley-Milner-style calculus with primitives for constructing nested
+types directly as fixpoints. Our calculus can express all nested types
+appearing in the literature, including truly nested types. At the
+level of terms, it supports primitive pattern matching, map functions,
+and fold combinators for nested types. Our main contribution is the
+construction of a parametric model for our calculus. This is both
+delicate and challenging. In particular, to ensure the existence of
+semantic fixpoints interpreting nested types, and thus to establish a
+suitable Identity Extension Lemma for our calculus, our type system
+must explicitly track functoriality of types, and cocontinuity
+conditions on the functors interpreting them must be appropriately
+threaded throughout the model construction. We also prove that our
+model satisfies an appropriate Abstraction Theorem, as well as that it
+verifies all standard consequences of parametricity in the presence of
+primitive nested types. In particular, we give several examples
+illustrating how how our model can be used to derive useful free
+theorems for programs over nested types.
 \end{abstract}
 
-%\begin{CCSXML}
-%<ccs2012>
-%<concept>
-%<concept_id>10011007.10011006.10011008</concept_id>
-%<concept_desc>Software and its engineering~General programming languages</concept_desc>
-%<concept_significance>500</concept_significance>
-%</concept>
-%<concept>
-%<concept_id>10003456.10003457.10003521.10003525</concept_id>
-%<concept_desc>Social and professional topics~History of programming languages</concept_desc>
-%<concept_significance>300</concept_significance>
-%</concept>
-%</ccs2012>
-%\end{CCSXML}
-%
-%\ccsdesc[500]{Software and its engineering~General programming languages}
-%\ccsdesc[300]{Social and professional topics~History of programming languages}
-%% End of generated code
-
-%% Keywords
-%% comma separated list
-%\keywords{keyword1, keyword2, keyword3}  %% \keywords is optional
-
 \maketitle
@@
 
-Suppose we want to prove some property of programs over an algebraic
-data type (ADT) such as that of lists, coded in Agda
-%\footnote{All code appearing in this paper are written in Agda.}
-as
-
-\vspace*{-0.15in}
+{\em Algebraic data types} (ADTs), both built-in and user-defined,
+have long been at the core of functional languages such as Haskell,
+ML, Agda, Epigram, and Idris. ADTs, such as that of natural numbers,
+can be unindexed. But they can also be indexed over other types. For
+example, the ADT of lists (here coded in Agda)
 
@@
 \end{array}\]}
-A natural approach to the problem uses structural induction on the
-input data structure in question. This requires knowing not just the
-definition of the ADT of which the input data structure is an
-instance, but also the program text for the functions involved in the
-properties to be proved. For example, to prove by induction that
-mapping a polymorphic function over a list and then reversing the
-resulting list is the same as reversing the original list and then
-mapping the function over the result, we unwind the (recursive)
-definitions of the $\mathtt{reverse}$ and $\mathtt{map}$ functions
-over lists according to the inductive structure of the input
-list. Such data-driven induction proofs over ADTs are routine enough
-to be included in, say, an undergraduate functional programming
-course.
-
-An alternative approach to proving results like the above
-$\mathtt{map}$-$\mathtt{reverse}$ property for lists is to use
-parametricity, a formalization of extensional type-uniformity in
-polymorphic languages. Parametricity captures the intuition that a
-polymorphic program must act uniformly on all of its possible type
-instantiations; it is formalized as the requirement that every
-polymorphic program preserves all relations between any pair of types
-that it is instantiated with.  Parametricity was originally put forth
-by Reynolds~\cite{rey83} for System F~\cite{gr89}, the formal calculus
-at the core of all polymorphic functional languages. It was later
-popularized as Wadler's ``theorems for free''~\cite{wad89} because it
-allows the deduction of many properties of programs in such languages
-solely from their types, i.e., with no knowledge whatsoever of the
-text of the programs involved. To get interesting free theorems,
-Wadler's calculus included built-in ADTs; indeed, most of the free
-theorems in~\cite{wad89} are consequences of naturality for
-polymorphic list-processing functions. However, the techniques
-of~\cite{wad89} show that parametricity can also be used to derive 
-naturality properties for non-list ADTs, as well as results, like
-correctness of program optimizations like {\em short cut
-  fusion}~\cite{glp93},
-%,joh02,joh03},
-that go beyond simple naturality.
-%In addition, parametricity in the presence of ADTs
-%{\color{blue} check!}  has further been used to prove i) {\em
-%  properties of programs}, e.g., that they perform no illegal
-%operations, satisfy certain security criteria~\cite{ep03,rp10}, or are
-%observationally equivalent to their compiled forms~\cite{bh09,hd11};
-%ii) {\em whole-language properties}, e.g., that they support data
-%abstraction and modularity via representation independence,
-%\cite{bm05,dr04,jac99,kat11,mr92}, enforce information flow
-%policies~\cite{ss01,ts04}, or guarantee privacy~\cite{rp10}; and iii)
-%{\em properties of implementations}, e.g., compiler
-%correctness~\cite{ab08,bh09,hd11,glpj93}. Parametricity is thus an
-%important tool for reasoning about programs in polymorphic languages
-%that natively support ADTs. {\color{blue} Be clear about what this
-%  means. Re-word?}
-
-This paper is about parametricity and free theorems for a polymorphic
-calculus with explicit syntax not just for ADTs, but for the
-generalization of ADTs known as {\em nested types}~\cite{bm98}, as
-well. An ADT defines a {\em family of inductive data types}, one for
-each input type. For example, the $\mathtt{List}$ data type definition
-above defines a collection of data types $\mathtt{List\; A}$,
-$\mathtt{List\; B}$, $\mathtt{List\; (A \times B)}$, $\mathtt{List\;
-  (\List\;A)}$, etc., each independent of all the others. By contrast,
-a nested type is an {\em inductive family of data types} that is
-defined over, or is defined mutually recursively with, (other) such
-data types. Since the structures of the data type at one input type
-can depend on those at other input types, the entire family of types
-must be defined at once. Examples of nested types include, trivially,
-ordinary ADTs, such as list and tree types; simple nested types, such
-as the data type {\small
+
+\noindent
+is indexed over its element type $\mathtt{A}$.  The instance of
+$\mathtt{List}$ at index $\mathtt{A}$ depends only on itself, and so is
+independent of $\mathtt{List\,B}$ for any other index $\mathtt{B}$.
+That is, $\mathtt{List}$, like all other ADTs, defines a {\em family
+  of inductive types}, one for each index type.
+
+Over time, there has been a notable trend toward data types whose
+non-regular indexing can capture invariants and other sophisticated
+properties that can be used for program verification and other
+applications.  A simple example of such a type is given by Bird and
+Meertens'~\cite{bm98} prototypical nested type
 \[\begin{array}{l}
@@
 \hspace*{0.4in}\mathtt{pnode\;:\;PTree\;(A \times A) \rightarrow PTree\;A}
-\end{array}\]}
-\hspace{-0.04in}of perfect trees, whose recursive occurrences never
-appear below other type constructors; ``deep'' nested
-types~\cite{jp20}, such as the data type {\small
+\end{array}\]
+\noindent
+of perfect trees, which can be thought of as constraining lists to
+have lengths that are powers of 2.  The above code makes clear that
+perfect trees at index type $\mathtt{A}$ are defined in terms of
+perfect trees at index type $\mathtt{A \times A}$. This is typical of
+nested types, one type instance of which can depend on others, so that
+the entire family of types must actually be defined at once. A nested
+type thus defines not a family of inductive types, but rather an {\em
+  inductive family of types}.  Nested types include simple nested
+types, like perfect trees, none of whose recursive occurrences occur
+below another type constructor; ``deep'' nested types~\cite{jp20},
+such as the nested type {\small
 \[\begin{array}{l}
-\mathtt{data\; Forest\;(A : Set)\;:\;Set\;where}\\
-\hspace*{0.4in}\mathtt{fempty\;:\;Forest\;A}\\
-\hspace*{0.4in}\mathtt{fnode\;:\; A \rightarrow PTree\;(Forest\;A) \to
-Forest\;A}
+\mathtt{data\; PForest\;(A : Set)\;:\;Set\;where}\\
+\hspace*{0.4in}\mathtt{fempty\;:\;PForest\;A}\\
+\hspace*{0.4in}\mathtt{fnode\;:\; A \rightarrow PTree\;(PForest\;A) \to
+PForest\;A}
 \end{array}\]}
@@
 below type constructors for other nested types; and truly nested
-types\footnote{Nested types that are defined over themselves are known
-  as {\em truly nested types}.}, such as the data type {\small
+types,
+%\footnote{Nested types that are defined over themselves are known
+%  as {\em truly nested types}.}
+such as the nested type {\small
 \[\begin{array}{l}
@@
 \end{array}\]}
-\hspace{-0.04in}of bushes (also called {\em bootstrapped heaps}
-in~\cite{oka99}), whose recursive occurrences appear below their own
-type constructors.
-
-Suppose we now want to prove properties of functions over nested
-types. We might, for example, want to prove a
-$\mathtt{map}$-$\mathtt{reverse}$ property for the functions on
-perfect trees in Figure~\ref{fig:funs}, or for those on
-bushes\footnote{To define the $\mathtt{foldBush}$ and
-  $\mathtt{mapBush}$ functions in Figure~\ref{fig:funs2} it is
-  necessary to turn off Agda's termination checker.} in
-Figure~\ref{fig:funs2}. A few well-chosen examples quickly convince us
-that such a property should indeed hold for perfect trees, and,
-drawing inspiration from the situation for ADTs, we easily construct a
-proof by induction on the input perfect tree. To formally establish
-this result, we could even prove it in Coq or Agda: each of these
-provers actually generates an induction rule for perfect trees, and
-the rule it generates gives the expected result because proving
-properties of perfect trees requires only that we induct over the
-top-level perfect tree in the recursive position. That is, any data
-internal to the input tree is left untouched.
-
-%want to prove that $\mathtt{reversePTree}$ commutes with
-%$\mathtt{mapPTree}$ or that $\mathtt{reverseBush}$ commutes with
-%$\mathtt{mapBush}$. It is not hard to convince oneself of the former:
-%a few well-chosen examples show clearly why mapping a polymorphic
-%%function over a perfect tree and then reversing the data at the leaves
-%of the resulting perfect tree is indeed the same result as mapping
-%that same function over the result of reversing the data at the leaves
-%of the original perfect tree.
-
-Unfortunately, it is nowhere near as clear that analogous intuitive or
-formal inductive arguments can be made for the
-$\mathtt{map}$-$\mathtt{reverse}$ property for bushes. Indeed, a proof
-by induction on the input bush must recursively induct over the bushes
-that are internal to the top-level bush in the recursive position.
-This is sufficiently delicate that no induction rule for bushes or
-other truly nested types was known until very recently, when {\em deep
-  induction}~\cite{jp20} was developed as a way to induct over {\em
-  all} of the structured data present in an input. Deep induction thus
-not only gave the first principled and practically useful structural
-induction rules for bushes and other truly nested types, but has also
-opened the way for incorporating automatic generation of such rules
-for (truly) nested data types --- and, eventually, even GADTs --- into
-modern proof assistants.
-\begin{figure*}
-\hspace*{-0.7in}
-\resizebox{0.35\linewidth}{!}{
-\begin{minipage}[t]{0.5\textwidth}
-\[\begin{array}{l}
-\mathtt{reversePTree : \forall \{A : Set\}  \rightarrow PTree\; A
-  \rightarrow PTree \;A}\\  
-\mathtt{reversePTree \; \{A\} = foldPTree \; \{A\}\; \{PTree\}}\\
-\hspace*{1.3in} \mathtt{pleaf}\\
-\hspace*{1.3in} \mathtt{( \lambda p \rightarrow pnode\; (mapPTree\;
-  swap\; p) )}\\  
-\\
-\mathtt{foldPTree : \forall \{A : Set\} \rightarrow \{F : Set \rightarrow Set \}
-  \rightarrow}\\
-\hspace*{0.8in} \mathtt{( \{B : Set\} \rightarrow B \rightarrow F\;B) \rightarrow}\\
-\hspace*{0.8in} \mathtt{(\{B : Set\} \rightarrow F\;(B \times B)
-  \rightarrow F \;B) \rightarrow}\\
-\hspace*{0.8in}\mathtt{PTree \;A \rightarrow F\;A}\\
-\mathtt{foldPTree\; n\; c\; (pleaf\; x) = n \;x}\\
-\mathtt{foldPTree\; n\; c \;(pnode\; p) = c\; (foldPTree \;n \;c \;p)}\\
-\\
-\mathtt{mapPTree : \forall \{A\, B : Set\} \rightarrow (A \rightarrow B)
-  \rightarrow PTree\; A \rightarrow PTree\; B}\\ 
-\mathtt{mapPTree \;f\; (pleaf \;x) = pleaf\; (f\; x)}\\
-\mathtt{mapPTree\; f \;(pnode\; p) =}\\
-\hspace*{0.7in}  \mathtt{pnode\; (mapPTree \;(\lambda p
-  \rightarrow (f (\pi_1 \,p), f (\pi_2\,p)))\; p)}\\ 
-\\
-\mathtt{swap : \forall \{A : Set\} \rightarrow (A \times A)
-  \rightarrow (A \times A)}\\ 
-\mathtt{swap \;(x,y) = (y,x)}
-\end{array}\]
-\caption{$\mathtt{reversePTree}$ and auxiliary functions in Agda}\label{fig:funs} 
-\end{minipage}}
-\quad\quad\quad\quad\quad%\quad\quad\quad
-\resizebox{0.35\linewidth}{!}{
-\begin{minipage}[t]{0.5\textwidth}
-\[\begin{array}{l}
-  \mathtt{reverseBush : \forall \{A : Set\} \rightarrow Bush\; A
-  \rightarrow Bush\; A}\\ 
-\mathtt{reverseBush \;\{A\} = foldBush\; \{A\}\; \{Bush\}\; bnil \;balg}\\
-\\
-\mathtt{foldBush : \forall \{A : Set\} \rightarrow \{F : Set \rightarrow
-  Set\} \rightarrow}\\
-\hspace*{0.5in} \mathtt{(\{B : Set\} \rightarrow F \;B) \rightarrow}\\
-\hspace*{0.5in} \mathtt{(\{B : Set\} \rightarrow B \rightarrow F\;
-  (F\; B) \rightarrow F\; B) \rightarrow}\\ 
-\hspace*{0.5in} \mathtt{Bush\; A \rightarrow F \;A}\\
-\mathtt{foldBush\; bn\; bc\; bnil = bn}\\
-\mathtt{foldBush \;bn \;bc \;(bcons\; x \;bbx) =}\\
-\hspace*{0.5in} \mathtt{bc\; x \;(foldBush \;bn\; bc
-  \;(mapBush \;(foldBush\; bn\; bc) \;bbx))}\\
-\\
-\mathtt{mapBush : \forall \{A \,B : Set\} \to (A \to B) \to Bush\, A \to
-  Bush\, B}\\ 
-\mathtt{mapBush\; \_ \; bnil = bnil}\\
-\mathtt{mapBush\; f\; (bcons\; x \;bbx) = bcons \;(f\, x) \; (mapBush\;
-  (mapBush\, f) \; bbx)}\\ 
-\\
-\mathtt{balg : \forall \{B : Set\} \rightarrow B \rightarrow Bush
-  \;(Bush\; B) \rightarrow Bush\; B}\\
-\mathtt{balg \;x \;bnil = bcons\; x \;bnil}\\
-\mathtt{balg \;x\; (bcons \;bnil \;bbbx) = bcons\; x \;(bcons \;bnil\; bbbx)}\\
-\mathtt{balg\; x \;(bcons \;(bcons\; y \;bbx)\; bbbx) =}\\
-\hspace*{0.5in} \mathtt{bcons\; y\; (bcons\; (bcons \;x \;bbx) \;bbbx)}\\
-\end{array}\]
-\caption{$\mathtt{reverseBush}$ and auxiliary
-  functions in Agda}\label{fig:funs2} 
-\end{minipage}}
-\end{figure*}
-
-\vspace*{-0.2in}
-
-Of course it is great to know that we {\em can}, at last, prove
-properties of programs over (truly) nested types by induction.  But
-recalling that parametricity often provides a viable alternative to
-inductive proofs for ADTs,
-%inductive proofs over ADTs can sometimes be circumvented in the
-%presence of parametricity,
-we might naturally ask:
-\begin{verse}
-{\em Can we derive properties of functions over (truly) nested types
-  from parametricity?}
-\end{verse}
-This paper answers the above question in the affirmative by
-constructing a parametric model for a polymorphic calculus providing
-primitives for {\em constructing} nested types
-%(including ADTs)
-directly via type-level recursion --- rather than representing them
-indirectly by Church encodings as in, e.g., System F.
-
-We introduce our calculus in Section~\ref{sec:calculus}.  At the type
-level, it is the level-2-truncation of the higher-kinded calculus
-from~\cite{jp19}, augmented with a primitive type of natural
-transformations.  To construct nested types, it carefully builds up
-type expressions not just from standard type variables, but also from
-type constructor variables of various arities, and includes an
-explicit $\mu$-construct for type-level recursion with respect to
-these variables. The class of nested types it constructs is very
-robust and includes all (truly) nested types known from the
-literature.  In Section~\ref{sec:type-interp} we give set and
-relational interpretations
-%{\color{blue} locally finitely presentable}
-for these types. As is standard when modeling parametricity, types are
-interpreted as functors from environments interpreting their type
-variable contexts to sets or relations, as appropriate. But to ensure
+\hspace{-0.04in}of bushes, whose recursive occurrences appear below
+their own type constructors.
+
+{\em Relational parametricity} encodes a powerful notion of
+type-uniformity, or representation independence, for data types in
+polymorphic languages. It formalizes the intuition that a polymorphic
+program must act uniformly on all of its possible type instantiations
+by requiring that every such program preserves all relations between
+pairs of types at which it is instantiated. Parametricity was
+originally put forth by Reynolds~\cite{rey83} for System
+F~\cite{gir72}, the calculus at the core of all polymorphic functional
+languages. It was later popularized as Wadler's ``theorems for
+free''~\cite{wad89}, so called because it can deduce properties of
+programs in such languages solely from their types, i.e., with no
+knowledge whatsoever of the text of the programs involved.  Most of
+Wadler's free theorems are consequences of naturality for polymorphic
+list-processing functions. However, parametricity can also derive
+results that go beyond just naturality, such as correctness for ADTs
+of the program optimization known as {\em short cut
+  fusion}~\cite{glp93,joh02}.
+
+But what about nested types? Does parametricity still hold if such
+types are added to polymorphic calculi? More practically, can we
+justifiably reason type-independently about (functions over) nested
+types in functional languages?
+
+Type-independent reasoning about ADTs in functional languages is
+usually justified by first representing ADTs by their Church
+encodings, and then reasoning type-independently about these
+encodings. This is typically justified by constructing a parametric
+model --- i.e, a model in which polymorphic functions preserve
+relations {\em \'a la} Reynolds --- for a suitable fragment of System
+F, demonstrating that an initial algebra exists for the positive type
+constructor corresponding to the functor underlying an ADT of
+interest, and showing that each such initial algebra is suitably
+isomorphic to its corresponding Church encoding. In fact, this
+isomorphism of initial algebras and their Church encodings is one of
+the ``litmus tests'' for the goodness of a parametric model.
+
+This approach works well for ADTs, which are always fixpoints of {\em
+  first-order} functors, and whose Church encodings, which involve
+quantification over only type variables, are always expressible in
+System F. For example, $\mathtt{List\,A}$ is the fixpoint of the
+first-order functor $F\,X = 1 + A \times X$ and has Church encoding
+$\forall \alpha. \, \alpha \to (\mathtt{A} \to \alpha \to \alpha) \to
+\alpha$. But despite Cardelli's~\cite{car97} claim that ``virtually
+any basic type of interest can be encoded within F$_2$'' --- i.e.,
+within System F --- non-ADT nested types cannot.  Not even our
+prototypical nested type of perfect trees has a Church encoding
+expressible in System F!  Indeed, $\mathtt{PTree\,A}$ cannot be
+represented as the fixpoint of any {\em first-order} functor. However,
+it can be seen as the instance at index $\mathtt{A}$ of the fixpoint
+of the {\em higher-order} functor $H\,F\,A\,=\, (A \to F\,A) \to (F
+\,(A \times A) \to F\,A) \to F\,A$. It thus has Church encoding
+$\forall f.\, (\forall \alpha.\,\alpha \to f\alpha) \to (\forall
+\alpha. \,f (\alpha \times \alpha) \to f\alpha) \to \forall \alpha.\,
+f\alpha$, which requires quantification at the higher kind $* \to *$
+for $f$. A similar situation obtains for any (non-ADT) nested
+type. Unfortunately, higher-kinded quantification is not available in
+System F, so if we want to reason type-independently about nested
+types in a language based on it we have only two options: {\em
+  i})\,move to an extension of System F, such as the higher-kinded
+calculus F$_\omega$ or a dependent type theory, and reason via their
+Church encodings in a known parametric model for that extension, or
+{\em ii})\, add nested types to System F as primitives --- i.e., as
+primitive type-level fixpoints --- and construct a parametric model
+for the result.
+
+Since the type systems of F$_\omega$ and dependent type theories are
+designed to extend System F with far more than non-ADT data types, it
+seems like serious overkill to pass to their parametric models to
+reason about nested types in System F. Indeed, such calculi support
+fundamentally new features that add complexity to their models that is
+entirely unnecessary for reasoning about nested types. This paper
+therefore pursues the second option above.  We first design a
+Hindley-Milner-style calculus supporting primitive nested types,
+together with primitive types of natural transformations representing
+morphisms between them. Our calculus can express all nested types
+appearing in the literature, including truly nested types.  At the
+term-level, it supports primitive pattern matching, map functions, and
+fold combinators for nested types.\footnote{We leave incorporating
+  general term-level recursion to future work because, as
+  Pitts~\cite{pit00} reminds us, ``it is hard to construct models of
+  both impredicative polymorphism and fixpoint recursion''. In fact,
+  as the development in this paper shows, constructing a parametric
+  model even for our predicative calculus with primitive nested types
+  --- and even without term-level fixpoints --- is already rather
+  involved.
+%  in part because, as
+%  Pitts~\cite{pit00} reminds us, ``it is hard to construct models of
+%  both impredicative polymorphism and fixpoint recursion'', and in
+%  part because constructing a parametric model even for a predicative
+%  calculus with primitive nested types is already rather involved.
+  On
+  the other hand, our calculus is strongly normalizing, so it perhaps
+  edges us toward the kind of provably total practical programming
+  language proposed in~\cite{wad89}.}  Our main contribution is the
+construction of a parametric model for our calculus. This is both
+delicate and challenging. To ensure the existence of semantic
+fixpoints interpreting nested types, and thus to establish a suitable
+Identity Extension Lemma, our type system must explicitly track
+functoriality of types, and cocontinuity conditions on the functors
+interpreting them must be appropriately threaded throughout the model
+construction. Our model validates all standard consequences of
+parametricity in the presence of primitive nested types, including the
+isomorphism of primitive ADTs and their Church encodings, and
+correctness of short cut fusion for nested types. The relationship
+between naturality and parametricity has long been of interest, and
+our inclusion of a primitive type of natural transformations allows us
+to clearly delineate those consequences of parametricity that follow
+from naturality, from those, such as short cut fusion for nested
+types, that require the full power of parametricity.
+
+\vspace*{0.1in}
+
+\noindent
+{\bf Structure of this Paper}\/ We introduce our calculus in
+Section~\ref{sec:calculus}.  Its type system is based on the
+level-2-truncation of the higher-kinded grammar from~\cite{jp19},
+augmented with a primitive type of natural
+transformations. (Since~\cite{jp19} contains no term calculus, the
+issue of parametricity could not even be raised there.)  In
+Section~\ref{sec:type-interp} we give set and relational
+interpretations of our types. Set interpretations are possible
+precisely because our calculus is predicative --- as ensured by our
+primitive natural transformation types --- and~\cite{jp19} guarantees
+that local finite presentability of $\set$ makes it suitable for
+interpreting nested types.  As is standard in categorical models,
+types are interpreted as functors from environments interpreting their
+type variable contexts to sets or relations, as appropriate. To ensure
 that these functors satisfy the cocontinuity properties needed for the
-semantic fixpoints interpreting $\mu$-types to exist, set environments
-must map each $k$-ary type constructor variable to an appropriately
-cocontinuous $k$-ary functor on sets, relation environments must map
-each $k$-ary type constructor variable to an appropriately
-cocontinuous $k$-ary relation transformer, and these cocontinuity
-conditions must be threaded throughout the type interpretations in
-such a way that the resulting model is guaranteed to satisfy an
-appropriate Identity Extension Lemma (Theorem~\ref{thm:iel}).
-%{\color{blue} Syntactic restrictions on
-%  $\mu$-types are essential to proving IEL.}
-Properly propagating the cocontinuity conditions turns out to be both
-subtle and challenging, and Section~\ref{sec:iel}, where it is done,
-is where the bulk of the work in constructing our model lies.  At the
-term level, our calculus includes primitive constructs for the actions
-on morphisms of the functors interpreting types, initial algebras for
-fixpoints of these functors, and structured recursion over elements of
-these initial algebras (i.e., $\map$, $\tin$, and $\fold$ constructs,
-respectively). While our calculus does not support general recursion
-at the term level, it is strongly normalizing, so does perhaps edge us
-toward the kind of provably total practical programming language
-proposed at the end of~\cite{wad89}.  In
-Section~\ref{sec:term-interp}, we give set and relational
-interpretations for the terms of our calculus. As usual in parametric
+semantic fixpoints interpreting nested types to exist, set environments
+must map $k$-ary type constructor variables to appropriately
+cocontinuous $k$-ary functors on sets, relation environments must map
+$k$-ary type constructor variables to appropriately cocontinuous
+$k$-ary relation transformers, and these cocontinuity conditions must
+be threaded through our type interpretations in such a way that an
+Identity Extension Lemma (Theorem~\ref{thm:iel}) can be
+proved. Properly propagating the cocontinuity conditions requires
+considerable care, and Section~\ref{sec:iel}, where it is done, is
+(apart from tracking functoriality in the calculus so that it is
+actually possible) where the bulk of the work in constructing our
+model lies.
+
+In Section~\ref{sec:term-interp}, we give set and relational
+interpretations for the terms of our calculus. As usual in categorical
 models, terms are interpreted as natural transformations from
-interpretations of the term contexts in which they are formed to the
-interpretations of their types, and these must cohere in what is
-essentially a fibrational way. Immediately from the definitions of our
-interpretations we prove in Section~\ref{sec:Nat-type-terms} a scheme
-deriving free theorems that are consequences of naturality of
-functions that are polymorphic over nested types. This scheme is very
-general, is parameterized over both the data type and the type of the
-polymorphic function at hand, and has each of the above
-$\mathtt{map}$-$\mathtt{reverse}$ theorems as instances.  The
-relationship between naturality and parametricity has long been of
-interest, and our inclusion of a primitive type of natural
-transformations allows us to clearly delineate those free theorems
-that are consequences of naturality, and may thus hold even in
-non-parametric models, from those that require the full power of
-parametricity.  In Section~\ref{sec:thms} we prove that our model
-satisfies an Abstraction Theorem (Theorem~\ref{thm:abstraction}), and
-we derive several of this latter kind of free theorem from it in
-Section~\ref{sec:ftnt}.  Specifically, we state and prove
-(non-)inhabitation results in Sections~\ref{sec:bottom}
-and~\ref{sec:identity}, a free theorem for the type of a filter
-function on generalized rose trees in Section~\ref{sec:ft-adt}, the
-correctness of short cut fusion for lists in
-Section~\ref{sec:short-cut}, and its generalization to nested types in
-Section~\ref{sec:short-cut-nested}.
-
-There is a long line of work on categorical models of parametricity
-for System F; see,
+interpretations of their term contexts to interpretations of their
+types, and these must cohere in what is essentially a fibred way.  In
+Section~\ref{sec:Nat-type-terms} we prove a scheme deriving free
+theorems that are consequences of naturality of polymorphic functions
+over nested types. This scheme is very general, and is parameterized
+over both the data type and the type of the polymorphic function at
+hand. It has, for example, analogues for nested types of Wadler's
+map-rearrangement free theorems as instances. In
+Section~\ref{sec:thms} we prove that our model satisfies an
+Abstraction Theorem (Theorem~\ref{thm:abstraction}), which we use to
+derive other parametricity results that go beyond naturality. We
+conclude in Section~\ref{sec:conclusion}.
+
+\vspace*{0.05in}
+
+\noindent
+{\bf Related Work}\/ There is a long line of work on categorical
+models of parametricity for System F; see,
 e.g.,~\cite{bfss90,bm05,dr04,gjfor15,has94,jac99,mr92,rr94}.  To our
-knowledge, all categorical models that treat (algebraic) data types do
-so via their Church encodings, verifying in the
-%Much of this work does ultimately extend parametricity to data types
-%that are modeled as fixpoints of first-order functors, but to our
-%knowledge this has never been done by building the data types directly
-%into the calculus whose parametricity is to be modeled
-%categorically. Indeed, all extensions we know of use the categorical
-%equivalent of Atkey's approach. That is, they verify in the
-just-constructed parametric model that
-%the Church encoding constructed
-%from a type constructor is interpreted as the least fixpoint of the
-%functor interpreting that type constructor.
-the Church encoding of each such type is interpreted as the least
-fixpoint of the (first-order) functor interpreting the type
-constructor from which its Church encoding was constructed.
-%This is the categorical analogue of the approach in~\cite{atk12}
-%described above.
-The present paper draws on this rich tradition of categorical models
-of parametricity for System F, but treats nested types as well as
-ADTs, and is the first to treat these types by direct construction via
-primitives rather than by Church encodings. This requires that we
-%modify the type calculus to ensure that
-%System F's full impredicative polymorphism interacts reasonably with
-%functoriality,
-design our calculus in such a way that its types all have
-interpretations as appropriately cocontinuous functors, so that the
-existence of the fixpoints by which nested types are to be interpreted
-is ensured.
-%
-%functoriality is guaranteed syntactically, and that it is reflected in
-%our model construction so that the existence of the fixpoints by which
-%nested types are to be interpreted is ensured.
-%
-%parametricity to primitive nested types {\em as the model is being
-%constructed} by taking care to ensure that System F's full
-%impredicative polymorphism does not interfere with the functoriality
-%needed to support primitive nested types.
-
-Like our calculus, Pitts' PolyPCF~\cite{pit98,pit00}
-%extends parametricity from pure System F to System F augmented with
-%to a calculus that
-includes primitives for constructing data types directly. Only list
-types are added in~\cite{pit00}, but other polynomial ADTs are easily
-included as in~\cite{pit98}. Part of Pitts' motivation is to show that
-ADTs and their Church encodings have the same operational behavior in
-his calculus. We cannot even ask this question about our calculus,
-which cannot express Church encodings of even simple ADTs such as
-list, or pair, or sum types, since these are not functorial;
-nevertheless, we do prove the correctness of short cut fusion for
-nested types, whose operational analogue is valid in the parametric
-model Pitts constructs and is the means by which he proves his
-equivalence result. It would be interesting to know what operational
-analogues of functoriality and cocontinuity are needed to extend
-Pitts' parametricity results from a calculus with primitives for
-constructing data types modeled as fixpoints of first-order functors
-to one that also provides such primitives for data types modeled as
-fixpoints of higher-order functors.
-
-We are not the first to consider parametricity at higher
-kinds. Atkey~\cite{atk12} constructs a parametric model for full
-System $F_\omega$, but within the impredicative Calculus of Inductive
-Constructions (iCIC) rather than in a semantic category. His
-construction is in some ways similar to ours, but it represents data
-types using Church encodings rather than constructing them via
-primitives. However, the $\mathit{fmap}$ functions associated to 
-Atkey's functors must be {\em given} together with their
-underlying type constructors, and he need not 
-impose cocontinuity conditions on his model to ensure that 
-fixpoints of his functors exist. Unfortunately, Atkey does not 
-indicate which
-type constructors support the $\mathit{fmap}$ functions
-needed for them to be functors, but we suspect spelling this
-out explicitly would result in a full higher-kinded extension of the
-calculus presented here.
+knowledge, all such models treat ADTs via their Church encodings,
+verifying in the just-constructed parametric model that each Church
+encoding is isomorphic to the data type it encodes.  The present paper
+draws on this rich tradition of categorical models of parametricity
+for System F, but modifies them to treat nested types (and therefore
+ADTs) as primitive data types.
+
+The only other extensions we know of System F with primitive data
+types are those in~\cite{mat11,mg01,pit98,pit00,wad89}.
+Wadler~\cite{wad89} treats full System F, and sketches parametricity
+for its extension with lists. Martin and Gibbons~\cite{mg01} outline a
+semantics for a grammar of primitive nested types similar to that
+in~\cite{jp19}, but treat only polynomial nested types, i.e., nested
+types that are fixpoints of polynomial higher-order
+functors. Unfortunately, the model suggested in~\cite{mg01} is not
+entirely correct (see~\cite{jp19}), and parametricity is nowhere
+mentioned.  Matthes~\cite{mat11} treats System F with non-polynomial
+ADTs and nested types, but his focus is on expressivity of generalized
+Mendler iteration for them. He gives no semantics.
+
+In~\cite{pit00}, Pitts adds list ADTs to full System F with a
+term-level fixpoint primitive. Other ADTs are included
+in~\cite{pit98}, but nested types are not expressible in either
+syntax. Pitts constructs parametric models for his calculi based on
+operational, rather than categorical, semantics. A benefit of using
+operational semantics to build parametric models is that it avoids
+needing to work in a suitable metatheory to accommodate System F's
+impredicativity. It is well-known that there are no set-based
+parametric models of System F~\cite{rey84}, so parametric models for
+it and its extensions are often constructed in a syntactic metatheory
+such as the impredicative Calculus of Inductive Constructions (iCIC).
+By adding primitive nested types to a Hindley-Milner-style calculus
+and working in a categorical setting we side-step such metatheoretic
+distractions. It is important to note that different consequences of
+parametricity are available in syntactic and semantic
+metatheories. Consequences of parametricity are possible for both
+closed and open System F terms in a syntactic metatheory --- although
+not all that can be formulated can be always proved; see, e.g., the
+end of Section~7 of~\cite{bm98}. By contrast, in a categorical
+metatheory consequences of parametricity are expressible only for {\em
+  closed} terms. For this reason, validating the standard consequences
+of parametricity for closed terms is --- going all the way back to
+Reynolds~\cite{rey83} --- all that is required for a model of
+parametricity to be considered good.
+
+Atkey~\cite{atk12} treats parametricity for arbitrary higher kinds,
+constructing a parametric model for System F$_\omega$ within iCIC,
+rather than in a semantic category. His construction is in some ways
+similar to ours, but he represents (now higher-kinded) data types
+using Church encodings rather than as primitives. Moreover, the
+$\mathit{fmap}$ functions associated to Atkey's functors must be {\em
+  given}, presumably by the programmer, together with their underlying
+type constructors. This absolves him of imposing cocontinuity
+conditions on his model to ensure that fixpoints of his functors
+exist, but, unfortunately, he does not indicate which type
+constructors support $\mathit{fmap}$ functions. We suspect explicitly
+spelling out which types can be interpreted as strictly positive
+functors would result in a full higher-kinded extension of a calculus
+akin to that presented here.
+
 \section{The Calculus}\label{sec:calculus}
@@
   constructor variables of arity $k$} and $\fvars^k$ of
-\emph{functorial variables of arity $k$}, all mutually disjoint.
-%disjoint for distinct $k$ and disjoint from each other.
-The sets of all type constructor variables and functorial variables
-are $\tvars = \bigcup_{k \ge 0} \tvars^k$ and $\fvars = \bigcup_{k \ge
-  0} \fvars^k$, respectively, and a \emph{type variable} is any
-element of $\tvars \cup \fvars$.  We use lower case Greek letters for
-type variables, writing $\phi^k$ to indicate that $\phi \in \tvars^k
-\cup \fvars^k$, and omitting the arity indicator $k$ when convenient,
-unimportant, or clear from context. We reserve letters from the
-beginning of the alphabet to denote type variables of arity $0$, i.e.,
-elements of $\tvars^0 \cup \fvars^0$. We write $\overline{\zeta}$ for
-either a set $\{\zeta_1,...,\zeta_n\}$ of type constructor variables
-or a set of functorial variables when the cardinality $n$ of the set
-is unimportant or clear from context. If $\Pos\,$ is a set of type
-variables we write $\Pos, \overline{\phi}$ for $\Pos\, \cup
-\overline{\phi}$ when $\Pos\, \cap \overline{\phi} = \emptyset$.  We
-omit the vector notation for a singleton set, thus writing $\phi$,
-instead of $\overline{\phi}$, for $\{\phi\}$.
-\begin{dfn}
-Let $V$ be a finite subset of\, $\tvars$, $\Pos$ be a finite
-subset of\, $\fvars$, $\overline{\alpha}$ be a finite subset of\,
-$\fvars^0$ disjoint from $\Pos$, and $\phi^k \in \fvars^k
-\setminus \Pos$.  The set $\mcF^\Pos(V)$ of {\em functorial
-  expressions} over $\Pos$ and $V$ are given by
-\begin{align*}
-  \mcF^\Pos(V) \; ::= \; %\T(V)
-  & \hspace*{0.15in}
-\zerot \mid \onet 
-\mid \Nat^{\ol{\alpha}} \, \mcF^{\overline{\alpha}}(V) \;
-\mcF^{\ol{\alpha}}(V) %\mid V \ol{\T(V)}
-\mid \Pos\; \ol{\mcF^\Pos(V)}  \,
-\mid V\, \ol{\mcF^\Pos(V)}  
-\mid \mcF^{\Pos}(V) + \mcF^\Pos(V)\\
-&\mid \mcF^{\Pos}(V) \times \mcF^\Pos(V) \mid \left(\mu
-\phi^{~k}. \lambda \alpha_1...\alpha_k.  
-\mcF^{\Pos,\alpha_1,...,\alpha_k,\phi}(V)\right)
-  \ol{\mcF^{\Pos}(V)}
-\end{align*}
-\end{dfn}
-\noindent
-A \emph{type} over $\Pos$ and $V$ is any element of % $\T(V) \cup
-$\F^\Pos(V)$.
-
-The notation for types entails that an application $F F_1...F_k$ is
-allowed only when $F$ is a type variable of arity $k$, or $F$ is a
-subexpression of the form $\mu \phi^{k}.\lambda
-\alpha_1...\alpha_k.F'$. Moreover, if $F$ has arity $k$ then $F$ must
-be applied to exactly $k$ arguments.  Accordingly, an overbar
-indicates a sequence of subexpressions whose length matches the arity
-of the type applied to it.  The fact that types are always in
-\emph{$\eta$-long normal form} avoids having to consider
+\emph{functorial variables of arity $k$}, all mutually disjoint.  The
+sets of all type constructor variables and functorial variables are
+$\tvars = \bigcup_{k \ge 0} \tvars^k$ and $\fvars = \bigcup_{k \ge 0}
+\fvars^k$, respectively, and a \emph{type variable} is any element of
+$\tvars \cup \fvars$.  We use lower case Greek letters for type
+variables, writing $\phi^k$ to indicate that $\phi \in \tvars^k \cup
+\fvars^k$, and omitting the arity indicator $k$ when convenient,
+unimportant, or clear from context. Letters from the beginning of the
+alphabet denote type variables of arity $0$, i.e., elements of
+$\tvars^0 \cup \fvars^0$. We write $\overline{\phi}$ for either a set
+$\{\phi_1,...,\phi_n\}$ of type constructor variables or a set of
+functorial variables when the cardinality $n$ of the set is
+unimportant or clear from context. If $V$ is a set of type variables
+we write $V, \overline{\phi}$ for $V \cup \overline{\phi}$ when $V
+\cap \overline{\phi} = \emptyset$.  We omit the vector notation for a
+singleton set, thus writing $\phi$, instead of $\overline{\phi}$, for
+$\{\phi\}$.
+
+If $\Gamma$ is a finite subset of\, $\tvars$, $\Phi$ is a finite
+subset of\, $\fvars$, $\overline{\alpha}$ is a finite subset of\,
+$\fvars^0$ disjoint from $\Phi$, and $\phi^k \in \fvars^k \setminus
+\Phi$, then the set $\mcF^\Pos(V)$ of {\em functorial expressions}
+over $\Gamma$ and $\Phi$ are given in
+Definition~\ref{def:wftypes}. The notation there entails that an
+application $F F_1...F_k$ is allowed only when $F$ is a type variable
+of arity $k$, or $F$ is a subexpression of the form $\mu
+\phi^{k}.\lambda \alpha_1...\alpha_k.F'$. Moreover, if $F$ has arity
+$k$ then $F$ must be applied to exactly $k$ arguments.  Accordingly,
+an overbar indicates a sequence of subexpressions whose length matches
+the arity of the type applied to it.  Requiring that types are always
+in such \emph{$\eta$-long normal form} avoids having to consider
 $\beta$-conversion at the level of types. In a subexpression
 $\Nat^{\ol{\alpha}}F\,G$, the $\Nat$ operator binds all occurrences of
-the variables in $\ol{\alpha}$ in $F$ and $G$. Similarly, in a
-subexpression $\mu \phi^k.\lambda \ol{\alpha}.F$, the $\mu$ operator
-binds all occurrences of the variable $\phi$, and the $\lambda$
-operator binds all occurrences of the variables in $\ol{\alpha}$, in
-the body $F$.
-
-A {\em type constructor context} is a finite set $\Gamma$ of type
-constructor variables, and a {\em functorial context} is a finite set
-$\Phi$ of functorial variables. In Definition~\ref{def:wftypes}, a
-judgment of the form $\Gamma;\Phi \vdash F$ indicates that the
-type $F$ is intended to be functorial in the variables in $\Phi$
-but not necessarily in those in $\Gamma$.
+the variables in $\ol{\alpha}$ in $F$ and $G$; intuitively,
+$\Nat^{\ol{\alpha}}F\,G$ represents the type of a natural
+transformation in $\ol{\alpha}$ from the functor $F$ to the functor
+$G$.  In a subexpression $\mu \phi^k.\lambda \ol{\alpha}.F$, the $\mu$
+operator binds all occurrences of the variable $\phi$, and the
+$\lambda$ operator binds all occurrences of the variables in
+$\ol{\alpha}$, in the body $F$.
+
+A {\em type constructor context}, or {\em non-functorial context}, is
+a finite set $\Gamma$ of type constructor variables, and a {\em
+  functorial context} is a finite set $\Phi$ of functorial
+variables. In Definition~\ref{def:wftypes}, a judgment of the form
+$\Gamma;\Phi \vdash F$ indicates that the type $F$ is intended to be
+functorial in the variables in $\Phi$ but not necessarily in those in
+$\Gamma$.
+
 \begin{dfn}\label{def:wftypes}
-The formation rules for the set $\F \subseteq \bigcup_{V \subseteq
-  \tvars, \Pos\, \subseteq \fvars}\F^\Pos(V)$ of\, {\em well-formed
-  types} are
+The formation rules for the set $\F$ of\, {\em well-formed types
+  {\color{red} over $\Gamma$ and $\Phi$?}} are
 
@@
 \AXC{$\Gamma;\ol{\gamma^0},\ol{\alpha^0},\phi^k \vdash F$}
-\AXC{$\quad\quad\ol{\Gamma;\Phi,\ol{\gamma^0} \vdash G}$}
-\BIC{$\Gamma;\Phi,\ol{\gamma^0} \vdash (\mu \phi^k.\lambda
-  \ol{\alpha^0}. \,F)\,\ol{G}$} 
+\AXC{$\quad\quad\ol{\Gamma;\Phi \vdash G}$}
+\AXC{$\quad\quad\ol{\Gamma;\Phi \vdash H}$}
+\TIC{$\Gamma;\Phi \vdash (\mu \phi^k.\lambda
+  \ol{\alpha^0}. \,F[\ol{\gamma := H}])\,\ol{G}$}  
 \DisplayProof
@@
 \end{array}\]
-%\noindent
-%A type $\tau$ is {\em well-formed} if it is
-%either a well-formed type constructor expression or a
-%well-formed functorial expression.
 \end{dfn}
-If $F$ is a closed type we may write $\vdash F$, rather than
-$\emptyset;\emptyset \vdash F$, for the judgment that it is
-well-formed.  Definition~\ref{def:wftypes} ensures that the expected
-weakening rules for well-formed types hold, although weakening does
-not change the contexts in which types can be formed. If
-$\Gamma;\emptyset \vdash F$ and $\Gamma;\emptyset \vdash G$, then our
-rules allow formation of the type $\Gamma;\emptyset \vdash
-\Nat^\emptyset F \,G$. Since $\Gamma; \emptyset \vdash \Nat^{\ol
-  \alpha}F\,G$ represents a natural transformation in $\ol \alpha$
-from $F$ to $G$, the type $\Gamma;\emptyset \vdash \Nat^\emptyset F
-\,G$ represents the arrow type $\Gamma \vdash F \to G$ in our
-calculus. The type $\Gamma; \emptyset \vdash \Nat^{\ol\alpha} \,\onet
-\,F$ similarly represents the $\forall$-type $\Gamma; \emptyset \vdash
-\forall \ol\alpha . F$.  Note, however, that
-%not even all
-%Hindley-Milner types of System F are representable in our calculus.
-not all System F types are 
-%only those System F types that can be interpreted as functors are
-representable in our calculus.
-%
-%However, if $\ol\alpha$ is non-empty then $\tau$ cannot be of the
-%form $\Nat^{\ol\beta} H \, K$ since $\Gamma; \ol\alpha \vdash
-%\Nat^{\ol\beta} H \, K$ is not a valid type judgment in our calculus
-%(except by weakening).
+We write $\vdash F$ for $\emptyset;\emptyset \vdash F$.
+Definition~\ref{def:wftypes} ensures that the expected weakening rules
+for well-formed types hold, although weakening does not change the
+contexts in which types can be formed. If $\Gamma;\emptyset \vdash F$
+and $\Gamma;\emptyset \vdash G$, then our rules allow formation of the
+types $\Gamma;\emptyset \vdash \Nat^\emptyset F \,G$ and $\Gamma;
+\emptyset \vdash \Nat^{\ol\alpha} \,\onet \,F$, which represent the
+System F types $\Gamma \vdash F \to G$ and $\forall$-type $\Gamma;
+\emptyset \vdash \forall \ol\alpha . F$, respectively. However, some
+System F types, such as $\forall \alpha. (\alpha \to \alpha) \to
+\alpha$, are not representable in our calculus.
 
@@
 can be functorial in variables of arity greater than $0$, the body $F$
-of a type $(\mu \phi.\lambda \ol\alpha.F)\,\ol G$ cannot. For example,
+of a type $(\mu \phi.\lambda \ol\alpha.F)\ol G$ cannot. For example,
 the type $\mathit{GRose}\, \phi \, \alpha = \mu\beta . \alpha \times
@@
 functors. In Section~\ref{sec:conclusion} we discuss how
-%GADTs and other
 more expressive types might be accommodated.
-%consider ways to lessen the expressivity constraints they impose.
-%
-%These restrictions on higher-arity functoriality are necessary
-%for constructing our parametric model. Fortunately, they are entirely
-%natural, and {\color{blue} preclude only truly bizarre nested types}.
-%do not preclude any nested
-%types {\color{blue} occurring in practice}.
-
-%If $\mathit{Tree}\,\alpha\,\gamma = \mu \,\beta.\, \alpha + \beta \times
-%\gamma \times \beta$, then Definition~\ref{def:wftypes} also allows
-%the derivation of, e.g., the type $\gamma; \emptyset \vdash
-%\Nat \alpha (\mathit{List}\, \alpha) \, (\mathit{Tree} \, \alpha\,
-%\gamma)$ representing a natural transformation from lists to trees
-%that is natural in $\alpha$ but not necessarily in $\gamma$. We
-%emphasize that types can be functorial in variables of arity greater
-%than $0$. For example, the type $\mathit{GRose}\, \phi \, \alpha =
-%\mu\beta . \alpha \times \phi \beta$ can be functorial in $\phi$ if
-%$\phi \in \Phi$. As usual, whether $\phi \in \Gamma$ or $\phi \in
-%\Phi$ determines whether types such as $\Nat^\alpha ( \mathit{GRose}
-%\, \phi \, \alpha)\, (List\, \alpha)$ are well-formed. But even if
-%$\mathit{GRose}$ is functorial in $\phi$, it still cannot be the
-%domain of a $\Nat$ type representing a natural transformation in
-%$\phi$. This is because our calculus does not allow naturality in
-%variables of arity greater than $0$.
 
 Definition~\ref{def:wftypes} allows well-formed types to be 
-%%considers types in $\T$ to be types in $\F$ that are
-functorial in no variables.
-%This allows the formation of types such as
-%$\mathit{List}\,(\sigma \to \tau)$ and $\mathit{PTree}\, (\forall
-%\alpha.\tau)$.
+functorial in no variables
 Functorial variables
 %in a well-formed type $\tau$
-can also be demoted to non-functorial status. The proof is by
-induction on the structure of $F$.
-
-\begin{lemma}
-If\, $\Gamma; \Phi, \phi^k \vdash F$, then $\Gamma, \psi^k; \Phi
-\vdash F[\phi^k :== \psi^k]$ is also derivable. Here, $F[\phi :==
-  \psi]$ is the textual replacement of $\phi$ in $F$, meaning that all
-occurrences of $\phi\ol\sigma$ in $F$ become $\psi\ol\sigma$.
-\end{lemma}
-
-In addition to textual replacement, we also have a proper substitution
-operation on types. If $F$ is a type over $P$ and $V$, if $\Pos$
-and $V$ contain only type variables of arity $0$, and if $k=0$ for
-every occurrence of $\phi^k$ bound by $\mu$ in $F$, then we say
-that $F$ is {\em first-order}; otherwise we say that $F$ is {\em
+can also be demoted to non-functorial status: if\,$F[\phi :== \psi]$
+is the textual replacement of $\phi$ in $F$, then $\Gamma, \psi^k;
+\Phi \vdash F[\phi^k :== \psi^k]$ is derivable whenever $\Gamma; \Phi,
+\phi^k \vdash F$ is. The proof is by induction on the structure of
+$F$. In addition to textual replacement, we also have a
+substitution operation on types.
+
+If $\Gamma; \Phi \vdash F$ is a type, if $\Gamma$ and $\Phi$ contain
+only type variables of arity $0$, and if $k=0$ for every occurrence of
+$\phi^k$ bound by $\mu$ in $F$, then we say that $F$ is {\em
+  first-order}; otherwise we say that $F$ is {\em
   second-order}. Substitution for first-order types is the usual
-capture-avoiding textual substitution. We write $F[\alpha :=
-  \sigma]$ for the result of substituting $\sigma$ for $\alpha$ in
-$F$, and $F[\alpha_1 := F_1,...,\alpha_k := F_k]$, or
-$F[\ol{\alpha := F}]$ when convenient, for $F[\alpha_1 :=
-  F_1][\alpha_2 := F_2,...,\alpha_k := F_k]$. Substitution
-for second-order types is defined below, where we adopt a similar
-notational convention for vectors of types. Note that it is not
-correct to substitute along non-functorial variables.
-
+capture-avoiding textual substitution. We write $F[\alpha := \sigma]$
+for the result of substituting $\sigma$ for $\alpha$ in $F$, and
+$F[\alpha_1 := F_1,...,\alpha_k := F_k]$, or $F[\ol{\alpha := F}]$
+when convenient, for $F[\alpha_1 := F_1][\alpha_2 := F_2,...,\alpha_k
+  := F_k]$. The operation of {\em second-order type substitution along
+  $\ol\alpha$} is given in Definition~\ref{def:second-order-subst},
+where we adopt a similar notational convention for vectors of types.
+Of course, $(\cdot)[\phi^0 :=_\emptyset F]$ coincides with first-order
+substitution. We omit $\ol\alpha$ when convenient, but note that it
+is not correct to substitute along non-functorial variables.
+
+{\color{red} Check no change needed for new $\mu$-rule in next def.}
 \begin{dfn}\label{def:second-order-subst}
-If \,$\Gamma; \Phi,\phi^k \vdash H$
-%with $k \geq 1$,
-and if\, $\Gamma;\Phi, \ol{\alpha} \vdash F$ with
-$|\ol\alpha| = k$, then $\Gamma;\Phi \vdash H[\phi :=_{\ol{\alpha}}
-  F]$.  Similarly, if \,$\Gamma, \phi^k; \Phi \vdash H$,
-%with $k \geq 1$,
-and if\, $\Gamma; \ol\psi,\ol{\alpha} \vdash F$ with $|\ol\alpha| = k$
-and $\Phi \cap \ol\psi = \emptyset$, then $\Gamma,\ol\psi';\Phi \vdash
-H[\phi :=_{\ol{\alpha}} F[\ol{\psi :== \psi'}]]$. Here, the operation
-  $(\cdot)[\phi :=_{\ol \alpha} F]$ of {\em second-order type
+If \,$\Gamma; \Phi,\phi^k \vdash H$ and if\, $\Gamma;\Phi, \ol{\alpha}
+\vdash F$ with $|\ol\alpha| = k$, then $\Gamma;\Phi \vdash H[\phi
+  :=_{\ol{\alpha}} F]$.  Similarly, if \,$\Gamma, \phi^k; \Phi \vdash
+H$, and if\, $\Gamma; \ol\psi,\ol{\alpha} \vdash F$ with $|\ol\alpha|
+= k$ and $\Phi \cap \ol\psi = \emptyset$, then $\Gamma,\ol\psi';\Phi
+\vdash H[\phi :=_{\ol{\alpha}} F[\ol{\psi :== \psi'}]]$. Here, the
+operation $(\cdot)[\phi :=_{\ol \alpha} F]$ of {\em second-order type
   substitution along $\ol\alpha$} is defined by:
@@
   :=_{\ol{\alpha}} F]\\[0.5ex]   
-((\mu \psi. \lambda \ol{\beta}.\, G)\ol{K})[\phi :=_{\ol{\alpha}}
+((\mu \psi. \lambda \ol{\beta}. G)\ol{K})[\phi :=_{\ol{\alpha}}
   F] & = & (\mu \psi. \lambda \ol{\beta}. \,G[\phi :=_{\ol{\alpha}}
@@
 \end{dfn}
-\noindent
-We note that $(\cdot)[\phi^0 :=_\emptyset F]$ coincides with
-first-order substitution. We also omit $\ol\alpha$ when convenient.
 
@@
 
-We now define our term calculus. To do so we assume an infinite set
-$\cal V$ of term variables disjoint from $\tvars$ and $\fvars$. If
-$\Gamma$ is a type constructor context and $\Phi$ is a functorial
-context, then a {\em term context for $\Gamma$ and $\Phi$} is a finite
-set of bindings of the form $x : F$, where $x \in {\cal V}$ and
-$\Gamma; \Phi \vdash F$. We adopt the same conventions for denoting
-disjoint unions and for vectors in term contexts as for type
-constructor contexts and functorial contexts.
+To define our term calculus we assume an infinite set $\cal V$ of term
+variables disjoint from $\tvars$ and $\fvars$. If $\Gamma$ is a type
+constructor context and $\Phi$ is a functorial context, then a {\em
+  term context for $\Gamma$ and $\Phi$} is a finite set of bindings of
+the form $x : F$, where $x \in {\cal V}$ and $\Gamma; \Phi \vdash
+F$. We adopt the above conventions for denoting disjoint unions and
+vectors in term contexts.
 
@@
 In the rule for $L_{\ol{\alpha}}x.t$, the $L$ operator binds all
-occurrences of the type variables in $\ol{\alpha}$ in the type of the
-term variable $x$ and in the body $t$, as well as all occurrences of
-$x$ in $t$. In the rule for $t_{\ol K} s$ there is one functorial
-expression in $\ol K$ for every functorial variable in $\ol \alpha$. In
-the rule for $\map^{\ol{F},\ol{G}}_H$ there is one functorial
-expression $F$ and one functorial expression $G$ for each functorial
-variable in $\ol\phi$. Moreover, for each $\phi^k$ in $\ol\phi$ the
-number of functorial variables in $\ol\beta$ in the judgments for its
+occurrences of the type variables in $\ol{\alpha}$ in the types of $x$
+and $t$, as well as all occurrences of $x$ in $t$. In the rule for
+$t_{\ol K} s$ there is one functorial expression in $\ol K$ for every
+functorial variable in $\ol \alpha$. In the rule for
+$\map^{\ol{F},\ol{G}}_H$ there is one functorial expression $F$ and
+one functorial expression $G$ for each functorial variable in
+$\ol\phi$. Moreover, for each $\phi^k$ in $\ol\phi$ the number of
+functorial variables in $\ol\beta$ in the judgments for its
 corresponding functorial expresssions $F$ and $G$ is $k$. In the rules
@@
 of the usual capture-avoiding textual substitution, and
-Definition~\ref{def:well-formed-terms} ensures that the expected
-weakening rules for well-formed terms hold.
-
-Using Definition~\ref{def:well-formed-terms} we can represent the
-$\mathtt{reversePTree}$ function from Figure~\ref{fig:funs} in our
-calculus as
+Definition~\ref{def:well-formed-terms} ensures that weakening is
+respected.
+
+Our calculus is expressive enough to define, e.g., a function
+$\mathit{reversePTree :
+  \Nat^\alpha\,(\mathit{PTree}\,\alpha)\,(\mathit{PTree}\,\alpha)}$
+that reverses the order of the leaves in a perfect tree as
 \[\vdash (\fold_{\beta + \phi(\beta \times \beta)}^{\mathit{PTree}\,
-  \alpha})_\emptyset \,
-%(\tin_{\beta + \phi(\beta \times \beta)} \circ
-s :
-\Nat^{\alpha} (\mathit{PTree}\,\alpha)\,(\mathit{PTree}\,\alpha)\] 
-where
+  \alpha})_\emptyset \, s : \Nat^{\alpha}
+(\mathit{PTree}\,\alpha)\,(\mathit{PTree}\,\alpha)\] where
 \[\begin{array}{lll}
@@
 \alpha))\, (\mathit{PTree}\,(\alpha \times \alpha)))
-%\[\begin{array}{lll}
-%\vdash \fold_{\beta + \phi(\beta \times \beta)}^{\mathit{PTree}\,\alpha} &
-%: & \Nat^{\emptyset} (\Nat^{\alpha} (\alpha + \mathit{PTree}\,(\alpha \times
-%\alpha)) \; (\mathit{PTree}\,\alpha))\; (\Nat^{\alpha}
-%(\mathit{PTree}\,\alpha) \; (\mathit{PTree}\,\alpha))\\
-%\vdash \tin_{\beta + \phi(\beta \times \beta)} & : & \Nat^{\alpha}
-%(\alpha + \mathit{PTree}\,(\alpha \times \alpha)) \; (\mathit{PTree}\,
-%\alpha)\\ 
-%\vdash \map_{\mathit{PTree}\,\alpha}^{\alpha \times \alpha, \alpha \times
-%  \alpha} & : & \Nat^{\emptyset} (\Nat^{\alpha} (\alpha \times \alpha)\,
-%(\alpha \times \alpha))\; (\Nat^{\alpha} (\mathit{PTree}\,(\alpha \times
-%\alpha))\, (\mathit{PTree}\,(\alpha \times \alpha)))
 \end{array}\]
@@
 \vdash \tin_{\beta + \phi (\beta \times \beta)} \circ (
-L_{\alpha}\, x.\, \inl\, x) \; : \; \Nat^{\alpha}\, \alpha\,
+L_{\alpha} x.\, \inl\, x) \; : \; \Nat^{\alpha}\, \alpha\,
 (\mathit{PTree}\,\alpha)\\ 
 \vdash \tin_{\beta + \phi (\beta \times \beta)} \circ (
-L_{\alpha}\, x.\, \inr\, x) \; : \; \Nat^{\alpha}\, \mathit{PTree}
+L_{\alpha} x.\, \inr\, x) \; : \; \Nat^{\alpha}\, \mathit{PTree}
 (\alpha \times \alpha)\,(\mathit{PTree}\,\alpha)\\ 
@@
 \end{array}\]
-respectively. We can similarly represent the $\mathtt{reverseBush}$
-function from Figure~\ref{fig:funs2} as
+respectively. For example, $\mathit{reversePTree}$ maps the perfect
+tree $((1,2),(3,4))$ to $((4,3),(2,1))$. Our calculus can similarly
+define a $\mathtt{reverseBush}$ function that {\color{red} reverses
+  the data in a bush???} as
 \[\vdash (\fold^{\mathit{Bush}\,\alpha}_{\onet + \beta \times \phi (\phi \beta)})_\emptyset\, balg
 : \Nat^\alpha (\mathit{Bush}\,\alpha)\,(\mathit{Bush}\,\alpha)\]
-%\[\vdash \fold_{\onet + \beta \times
-%\phi (\phi\beta)}^{\mathit{Bush}\,\alpha}\, ( \tin_{\onet + \beta \times
-%  \phi (\phi\beta)} \circ (\onet + t \circ i \circ i') ) :
-%\Nat^{\alpha} (\mathit{Bush}\,\alpha)\,(\mathit{Bush}\,\alpha)\] 
 where
@@
 \alpha \times \mathit{Bush}\,(\mathit{Bush}\,\alpha))\\ 
-%\vdash L_{\alpha}\, (b, s). \,\mathsf{case}\;s\,\mathsf{of}\, \{ \hspace*{0.21in}
-%        \ast \mapsto
-%\mathit{bcons}_{\alpha}\, b\, (\mathit{bnil}_{\alpha}\, \ast); \\
-%\hspace*{1.15in}\,(s', u) \mapsto \mathsf{case}\,{s'}\,\mathsf{of}\,
-%\{\hspace*{0.3in} \ast
-%\mapsto \mathit{bcons}_{\alpha}\, b \,(\mathit{bcons}_{\mathit{Bush}\,
-%  \alpha}\, (\mathit{bnil}_{\alpha}\, \ast)\, u); \\
-%\hspace*{2.3in}(b', u') \mapsto \mathit{bcons}_{\alpha}\, b'
-%(\mathit{bcons}_{\mathit{Bush}\, \alpha}\, (\mathit{bcons}_{\alpha}\,
-%b\, u )\, u')\}\}\\
-%\hspace{0.2in} : \Nat^{\alpha} (\alpha \times (\onet + (\onet + \alpha
-%\times \mathit{Bush}\, (\mathit{Bush} \, \alpha))) \times
-%\mathit{Bush}\, (\mathit{Bush} \, (\mathit{Bush}\, \alpha))) \;
-%(\alpha \times \mathit{Bush}\, (\mathit{Bush} \, \alpha))\\
-%\vdash \alpha \times (\onet + \tin^{-1}_{\onet + \beta \times \phi
-%  (\phi\beta)} \times \mathit{Bush}\, (\mathit{Bush} \,
-%(\mathit{Bush}\, \alpha)))\\
-%\hspace{0.2in} : \Nat^{\alpha} (\alpha \times (\onet + \mathit{Bush}
-%\, \alpha \times \mathit{Bush}\,( \mathit{Bush} \, (\mathit{Bush}\,
-%\alpha))))\\
-%\hspace*{0.57in}(\alpha \times (\onet + (\onet + \alpha \times
-%\mathit{Bush}\, (\mathit{Bush} \, \alpha)) \times \mathit{Bush}\,
-%(\mathit{Bush} \, (\mathit{Bush}\, \alpha))))\\
-%\vdash \alpha \times (L_{\alpha}\,x.\, (\tin^{-1}_{\onet +
-%  \beta \times \phi (\phi\beta)})_{\mathit{Bush}\,\alpha}\, x)\\
-%\hspace{0.2in} : \Nat^{\alpha} (\alpha
-%\times \mathit{Bush} \, (\mathit{Bush}\, \alpha)) \; (\alpha \times (\onet
-%+ \mathit{Bush} \, (\alpha \times \mathit{Bush}\, (\mathit{Bush} \,
-%(\mathit{Bush}\, \alpha)))))
-%\[\begin{array}{l}
   \vdash L_{\alpha} \, s. 
@@
   (\mathit{bcons}_\alpha (a, bba'), bbba)) \} \! \}\\    
-%  \quad \}  
 \hspace*{0.2in} : \Nat^\alpha \, (\alpha \times \mathit{Bush} (\mathit{Bush} \,\alpha)) \, (\mathit{Bush}\,\alpha) \\
 \end{array}\]
-respectively.
-%Here, $\Gamma; \emptyset \,|\, \Delta \vdash H+
-%\eta : \Nat^{\ol{\alpha}} (H+ F)\; (H+ G)$ and $\Gamma;
-%\emptyset \,|\, \Delta \vdash H\times \eta : \Nat^{\ol{\alpha}}
-%(H\times F)\; (H\times G)$ for $H+ \eta :=
-%L_{\ol{\alpha}}\, x.\, \case{x}{s \mapsto s}{t
-%  \mapsto  
-%  (\eta_{\ol{\alpha}} t)}$ and $H\times \eta := L_{\ol{\alpha}}\,
-%x.\, (\pi_1 x, \eta_{\ol{\alpha}} (\pi_2 x))$ for $\Gamma; \emptyset
-%\,|\, \Delta \vdash \eta : \Nat^{\ol{\alpha}} F\; G$ and $\Gamma;
-%\ol{\alpha} \vdash H$.
+respectively. {\color{red} For example, $\mathit{reverseBush}$ maps
+  the bush $?$ to $??$.}
+
+Unfortunately, however, we cannot define recursive functions ---
+
+HERE!!
 

### 2020-12-10 free-theorems-nesteds/20-mscs.tex
updates to mscs.tex

@@
 % submission space)
-\documentclass[acmsmall,review,anonymous]{acmart}
+\documentclass[acmsmall,review]{acmart}
 \settopmatter{printfolios=true,printccs=false,printacmref=false}
@@
 
+\usepackage{adjustbox}
+\usepackage{marvosym}
+\usepackage{graphicx}
 \usepackage{amsmath}
@@
 
-\title[Free Theorems for Nested Types]{Free Theorems for
-  %Primitive
-  Nested Types} %% [Short Title] is optional; when present,
+\title[Paramrtricity for Primitive Nested Types and
+  GADTs]{Parametricity for Primitive
+  Nested Types and GADTs} %% [Short Title] is optional; when present,
                          %% will be used in header instead of Full
@@
 
+{\color{red} Say extended version of FoSSaCS paper. Say what's
+  new/different. GADTs, implementation.
+
+  Send to LMCS? Editor?
+
+Can bind variables of arity > 0 in Nat-types? Improves expressibilty?
+Worth it?}
+
 \section{The Calculus}\label{sec:calculus}
@@
 \end{dfn}
+
 We write $\vdash F$ for $\emptyset;\emptyset \vdash F$.
@@
 functorial in $\phi$. Moreover, $\mathit{GRose}\,\phi\,\alpha$ cannot
-be the (co)domain of a $\Nat$ type representing a natural
+be the (co)domain of a $\Nat$-type representing a natural
 transformation functorial in $\phi$, and the type $\mu \phi. \lambda
@@
 \begin{dfn}\label{def:second-order-subst}
-If \,$\Gamma; \Phi,\phi^k \vdash H$ and if\, $\Gamma;\Phi, \ol{\alpha}
-\vdash F$ with $|\ol\alpha| = k$, then $\Gamma;\Phi \vdash H[\phi
-  :=_{\ol{\alpha}} F]$.  Similarly, if \,$\Gamma, \phi^k; \Phi \vdash
-H$, and if\, $\Gamma; \ol\psi,\ol{\alpha} \vdash F$ with $|\ol\alpha|
-= k$ and $\Phi \cap \ol\psi = \emptyset$, then $\Gamma,\ol\psi';\Phi
-\vdash H[\phi :=_{\ol{\alpha}} F[\ol{\psi :== \psi'}]]$. Here, the
-operation $(\cdot)[\phi :=_{\ol \alpha} F]$ of {\em second-order type
-  substitution along $\ol\alpha$} is defined by:
+If \,$\Gamma; \Phi,\phi^k \vdash H$ and $\Gamma;\Phi, \ol{\alpha}
+\vdash F$ with $|\ol\alpha| = k$, then the operation $H[\phi :=_{\ol
+    \alpha} F]$ of {\em second-order type substitution along
+  $\ol\alpha$} is defined by induction on $H$ as follows:
 \[\begin{array}{lll}
@@
 \end{dfn}
+\noindent
+It is not hard to see that $\Gamma;\Phi \vdash H[\phi :=_{\ol{\alpha}}
+  F]$. In addition, if \,$\Gamma, \phi^k; \Phi \vdash H$,\, $\Gamma;
+\ol\psi,\ol{\alpha} \vdash F$, $|\ol\alpha| = k$, and $\Phi \cap
+\ol\psi = \emptyset$, then $\Gamma,\ol\psi';\Phi \vdash H[\phi
+  :=_{\ol{\alpha}} F[\ol{\psi :== \psi'}]]$.
 
@@
 
-To define our term calculus we assume an infinite set $\cal V$ of term
-variables disjoint from $\tvars$ and $\fvars$. If $\Gamma$ is a type
-constructor context and $\Phi$ is a functorial context, then a {\em
-  term context for $\Gamma$ and $\Phi$} is a finite set of bindings of
-the form $x : F$, where $x \in {\cal V}$ and $\Gamma; \Phi \vdash
-F$. We adopt the above conventions for denoting disjoint unions and
-vectors in term contexts.
-
-\begin{dfn}\label{def:well-formed-terms}
-Let $\Delta$ be a term context for $\Gamma$ and $\Phi$.  The formation
-rules for the set of\, {\em well-formed terms over $\Delta$} are
+\begin{figure*}
+
+  \begin{adjustbox}{varwidth=5.4in, max width=5in, fbox, center}
        \[\begin{array}{ccc}
-%\AXC{$\Gamma;\emptyset \vdash \tau : \T$}
-%\UIC{$\Gamma;\emptyset\,|\, \Delta,x :\tau \vdash x : \tau$}
-%\DisplayProof
        \AXC{$\Gamma;\Phi \vdash F$}
        \UIC{$\Gamma;\Phi \,|\, \Delta,x :F \vdash x : F$}
-\DisplayProof
+       \hspace*{0.2in}\DisplayProof\hspace*{0.05in}
        &
@@
        \BIC{$\Gamma;\Phi \,|\, \Delta \vdash \bot_F t  : F$}
-\DisplayProof
+       \DisplayProof\hspace*{0.05in}
        &
@@
        \end{array}\]
+
+       \vspace*{-0.15in}
+
        \[\begin{array}{cc}
@@
        \UIC{$\Gamma;\Phi \,|\, \Delta \vdash \inl \,s: F + G$}
-\DisplayProof
+       \hspace*{0.8in}\DisplayProof\hspace*{0.05in}
        &
@@
        \end{array}\]
+
+       \vspace*{-0.1in}
+
        \[\begin{array}{c}
@@
        \AXC{$\Gamma;\Phi \,|\, \Delta \vdash t : F+G$}
-\AXC{$\Gamma;\Phi \,|\, \Delta, x : F \vdash l : K \hspace{0.3in} \Gamma;\Phi \,|\, \Delta, y : G \vdash r : K$}
-\TIC{$\Gamma;\Phi~|~\Delta \vdash \case{t}{x \mapsto l}{y \mapsto r} : K$}
-\DisplayProof
+       \AXC{$\Gamma;\Phi \,|\, \Delta, x : F \vdash l : K \hspace{0.2in} \Gamma;\Phi \,|\, \Delta, y : G \vdash r : K$}
+       \TIC{$\Gamma;\Phi~|~\Delta \vdash \cse{t}{x \mapsto l}{y \mapsto r} : K$}
+       \hspace*{-0.2in}\DisplayProof
        \end{array}\]
@@
        \BIC{$\Gamma;\Phi \,|\, \Delta \vdash (s,t) : F \times G$}
-\DisplayProof
+       \DisplayProof\hspace*{0.05in}
        &
@@
        \UIC{$\Gamma;\Phi \,|\, \Delta \vdash \pi_1 t : F$}
-\DisplayProof
+       \DisplayProof\hspace*{0.05in}
        &
@@
 
-\vspace*{0.05in}
-
        \[\begin{array}{c}
-%\AXC{$\Gamma; \emptyset \vdash \Nat^{\ol{\alpha}} \,F\,G : \T$}
        \AXC{$\Gamma; \ol{\alpha} \vdash F$}
@@
          \,|\, \Delta \vdash L_{\ol{\alpha}} x.t : \Nat^{\ol{\alpha}} \,F \,G$}
-\DisplayProof
+       \DisplayProof\vspace*{-0.05in}
        \\\\
@@
          G[\overline{\alpha := K}]$}
-\DisplayProof
+       \DisplayProof\vspace*{-0.1in}
        \\\\
@@
          ~|~\emptyset
-%  \ol{ x : \Nat^{\ol{\beta},\ol{\gamma}}\,F\,G}
-  %  \emptyset
          \vdash \map^{\ol{F},\ol{G}}_H :
@@
              :=_{\ol{\beta}} G}])$} 
-\DisplayProof
+       \DisplayProof\vspace*{-0.1in}
        \\\\
-%\AXC{$\Gamma;\Phi \,|\, \Delta \vdash t : H[\phi := (\mu \phi.\lambda 
-%  \ol{\alpha}.H)\ol{\beta}][\ol{\alpha := \tau}]$}
-%\AXC{$\ol{\Gamma;\Phi \vdash \tau : \F}$}
        \AXC{$\Gamma; \phi, \ol{\alpha},\ol{\gamma} \vdash H$}
@@
          \phi.\lambda \ol{\alpha}.H)\ol{\beta}$}
-\DisplayProof
+       \hspace*{0.2in}\DisplayProof\vspace*{-0.1in}
        \\\\
@@
        \AXC{$\Gamma; \ol{\beta}, \ol{\gamma} \vdash F$}
-%\AXC{$\Gamma;\emptyset \,|\, \Delta \vdash t :
-%  \Nat^{\ol{\beta}, \ol{\gamma}}\,H[\phi :=_{\ol{\beta}} F][\ol{\alpha
-%      := \beta}]\,F$} 
-%\TIC{$\Gamma;\emptyset \,|\, \Delta \vdash
-%  \fold_{H,F}\, t : \Nat^{\ol{\beta}, \ol{\gamma}}\,((\mu
-%  \phi.\lambda \ol{\alpha}.H)\ol{\beta})\,F$} 
        \BIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \fold^F_H :
@@
            \ol{\gamma}}\,(\mu \phi.\lambda \ol{\alpha}.H)\ol{\beta}\,F)$}
+       \hspace*{0.2in}\DisplayProof\vspace*{-0.1in}
+       \end{array}\]
+
+       \vspace*{0.05in}
+
+       \caption{Well-formed terms {\color{red} Change last three
+           judgements for new $\mu$ rule???}}\label{fig:terms} \vspace*{-0.00in}
+\end{adjustbox}
+       \vspace*{-0.25in}
+\end{figure*}
+
+To define our term calculus we assume an infinite set $\cal V$ of term
+variables disjoint from $\tvars$ and $\fvars$. If $\Gamma$ is a type
+constructor context and $\Phi$ is a functorial context, then a {\em
+  term context for $\Gamma$ and $\Phi$} is a finite set of bindings of
+the form $x : F$, where $x \in {\cal V}$ and $\Gamma; \Phi \vdash
+F$. We adopt the above conventions for denoting disjoint unions and
+vectors in term contexts. If $\Delta$ is a term context for $\Gamma$
+and $\Phi$, then the formation rules for the set of {\em well-formed
+  terms over $\Delta$} are as in Figure~\ref{fig:terms}. In the rule
+there for $L_{\ol{\alpha}}x.t$, the $L$ operator binds all occurrences
+of the type variables in $\ol{\alpha}$ in the types of $x$ and $t$, as
+well as all occurrences of $x$ in $t$. In the rule for $t_{\ol K} s$
+there is one functorial expression in $\ol K$ for every functorial
+variable in $\ol \alpha$. In the rule for $\map^{\ol{F},\ol{G}}_H$
+there is one functorial expression $F$ and one functorial expression
+$G$ for each functorial variable in $\ol\phi$. Moreover, for each
+$\phi^k$ in $\ol\phi$ the number of functorial variables in $\ol\beta$
+in the judgments for its corresponding functorial expresssions $F$ and
+$G$ is $k$. In the rules for $\tin_H$ and $\fold^F_H$, the functorial
+variables in $\ol{\beta}$ are fresh with respect to $H$, and there is
+one $\beta$ for every $\alpha$. Substitution for terms is the obvious
+extension of the usual capture-avoiding textual substitution, and the
+rules of Figure~\ref{fig:terms} ensure that weakening is respected.
+
+{\color{red} Change the rest of the section to agree with new term
+  formation rules.}  The ``extra'' functorial variables $\ol{\gamma}$
+in the rules for $\map^{\ol{F},\ol{G}}_H$, $\tin_H$, and $\fold^F_H$
+(i.e., those variables not affected by the substitution of $\phi$)
+deserve comment. They allow us to map or fold polymorphic functions
+over nested types. Suppose, for example, we want to map the
+polymorphic function $\mathit{flatten} : \Nat^\beta
+(\mathit{PTree}\,\beta)\,(\mathit{List}\,\beta)$ over a list.  Even in
+the absence of extra variables the instance of $\map$ required to map
+each non-functorial monomorphic instantiation of $\mathit{flatten}$
+over a list of perfect trees is well-typed:
+\[\begin{array}{l}
+\AXC{$\Gamma;\alpha \vdash \mathit{List} \, \alpha$}
+\AXC{$\Gamma;\emptyset \vdash \mathit{PTree}\,F$}
+\AXC{$\Gamma;\emptyset \vdash \mathit{List}\,G$}
+\TIC{$\Gamma;\emptyset \,|\, \emptyset \vdash \map^{\mathit{PTree}\,F,
+    \,\mathit{List}\,G}_{\mathit{List}\,\alpha} :
+  \Nat^\emptyset (\Nat^\emptyset\,(\mathit{PTree}\, F)\,(\mathit{List}\, G))\,
+  (\Nat^\emptyset\,(\mathit{List}\, (\mathit{PTree}\,
+  F))\,(\mathit{List}\, (\mathit{List}\, G)))$} \DisplayProof
+\end{array}\]
+But in the absence of $\ol \gamma$, the instance
+\[\Gamma;\emptyset~|~\emptyset \vdash
+\map^{\mathit{PTree}\,\beta,\mathit{List}\,\beta}_{\mathit{List}\,\alpha}
+: \Nat^\emptyset ( \Nat^\beta(\mathit{PTree}\,
+\beta)\,(\mathit{List}\, \beta))\, (\Nat^\beta\,(\mathit{List}\,
+(\mathit{PTree}\, \beta))\,(\mathit{List}\, (\mathit{List}\,
+\beta)))\] required to map the {\em polymorphic} $\mathit{flatten}$
+function over a list of perfect trees is not: indeed, the functorial
+contexts for $F$ and $G$ in the rule for $\map^{F,G}_H$ would have to
+be empty, but because the polymorphic $\mathit{flatten}$ function is
+natural in some variable, say $\delta$, it cannot possibly have a type
+of the form $\Nat^\emptyset F\, G$ as would be required for it to be
+the function input to $\map$. Untypeability of this instance of $\map$
+is unsatisfactory in a polymorphic calculus, where we naturally expect
+to be able to manipulate entire polymorphic functions rather than just
+their monomorphic instances, but the ``extra'' variables $\ol \gamma$
+remedy the situation, ensuring that the instance of $\map$ needed to
+map the polymorphic $\mathit{flatten}$ function is typeable as
+follows:
+
+\vspace*{0.1in}
+
+\begin{adjustbox}{varwidth=6.4in, max width=\linewidth, center}
+\[\begin{array}{l}
+\AXC{$\Gamma;\alpha,\gamma \vdash \mathit{List} \, \alpha$}
+\AXC{$\Gamma;\gamma \vdash \mathit{PTree}\,\gamma \hspace*{0.3in}
+  \Gamma;\gamma \vdash \mathit{List}\,\gamma$}
+\BIC{$\Gamma;\emptyset~|~\emptyset \vdash
+  \map^{\mathit{PTree}\,\gamma,\mathit{List}\,\gamma}_{\mathit{List}\,\alpha}  : \Nat^\emptyset
+  (\Nat^\gamma(\mathit{PTree}\, 
+  \gamma)\,(\mathit{List}\, \gamma))\,
+ (\Nat^\gamma\,(\mathit{List}\,
+  (\mathit{PTree}\, \gamma))\,(\mathit{List}\, (\mathit{List}\,
+  \gamma)))$}
 \DisplayProof
   \end{array}\]
-\end{dfn}
-
-In the rule for $L_{\ol{\alpha}}x.t$, the $L$ operator binds all
-occurrences of the type variables in $\ol{\alpha}$ in the types of $x$
-and $t$, as well as all occurrences of $x$ in $t$. In the rule for
-$t_{\ol K} s$ there is one functorial expression in $\ol K$ for every
-functorial variable in $\ol \alpha$. In the rule for
-$\map^{\ol{F},\ol{G}}_H$ there is one functorial expression $F$ and
-one functorial expression $G$ for each functorial variable in
-$\ol\phi$. Moreover, for each $\phi^k$ in $\ol\phi$ the number of
-functorial variables in $\ol\beta$ in the judgments for its
-corresponding functorial expresssions $F$ and $G$ is $k$. In the rules
-for $\tin_H$ and $\fold^F_H$, the functorial variables in $\ol{\beta}$
-are fresh with respect to $H$, and there is one $\beta$ for every
-$\alpha$. (Recall from above that, in order for the types of $\tin_H$
-and $\fold^F_H$ to be well-formed, the length of $\alpha$ must equal
-the arity of $\phi$.) Substitution for terms is the obvious extension
-of the usual capture-avoiding textual substitution, and
-Definition~\ref{def:well-formed-terms} ensures that weakening is
-respected.
-
-Our calculus is expressive enough to define, e.g., a function
+\end{adjustbox}
+
+\vspace*{0.1in}
+
+\noindent
+Similar remarks justify the appearance of $\ol \gamma$ in the typing
+rules for $\tin$ and $\mathsf{fold}$.
+
+Our calculus is expressive enough to define a function
 $\mathit{reversePTree :
   \Nat^\alpha\,(\mathit{PTree}\,\alpha)\,(\mathit{PTree}\,\alpha)}$
-that reverses the order of the leaves in a perfect tree as
+that reverses the order of the leaves in a perfect tree. This function
+maps, e.g., the perfect tree $((1,2),(3,4))$ to $((4,3),(2,1))$, and
+can be defined as
 \[\vdash (\fold_{\beta + \phi(\beta \times \beta)}^{\mathit{PTree}\,
@@
 \Nat^{\alpha} (\alpha \times \alpha)\, (\alpha \times \alpha)\\
-\vdash L_{\alpha} t. \,\case{t}{b \mapsto
+\vdash L_{\alpha} t. \,\cse{t}{b \mapsto
  \mathit{pleaf}\, b}{t' \mapsto \mathit{pnode}\, 
@@
 \end{array}\]
-respectively. For example, $\mathit{reversePTree}$ maps the perfect
-tree $((1,2),(3,4))$ to $((4,3),(2,1))$. Our calculus can similarly
-define a $\mathtt{reverseBush}$ function that {\color{red} reverses
-  the data in a bush???} as
+respectively. Our calculus can similarly define a
+$\mathtt{reverseBush}$ function that reverses the data in a bush. This
+function maps, e.g., {\color{red} the bush $?$ to $??$}, and can be
+defined as
 \[\vdash (\fold^{\mathit{Bush}\,\alpha}_{\onet + \beta \times \phi (\phi \beta)})_\emptyset\, balg
@@
   \vdash L_{\alpha} \, s. 
-  \case{s}{* \mapsto \mathit{bnil}_\alpha *}{(a, bba) \mapsto consalg_\alpha (a, bba)}
+  \cse{s}{* \mapsto \mathit{bnil}_\alpha *}{(a, bba) \mapsto consalg_\alpha (a, bba)}
   : \Nat^\alpha \, (\onet + \alpha \times \mathit{Bush} (\mathit{Bush} \,\alpha)) \, (\mathit{Bush}\,\alpha) \\
@@
 \end{array}\]
-respectively. {\color{red} For example, $\mathit{reverseBush}$ maps
-  the bush $?$ to $??$.}
-
-Unfortunately, however, we cannot define recursive functions ---
-
-HERE!!
-
-Our rule for forming $\Nat$-types means that 
-%Because our calculus is designed to ensure functoriality, and
-%functoriality does not interact with polymorphism as well as might be
-%hoped,
-our calculus cannot express types of recursive functions, such as a
-concatenation function for perfect trees, that take as inputs a nested
-type and an argument of another type, both of which are parameterized
-over the same variable $\alpha$.
-%not just one nested type parameterized over a variable $\alpha$ as
-%input, but other inputs parameterized over $\alpha$ as well.
-For the same reason, our calculus cannot express, e.g., the type of a
-zip function for bushes. The fundamental
-issue is that recursion is expressible only via folds, and a fold over
-a nested type must take a natural transformation as its argument and
-return a natural transformation as its result.
-%and, unfortunately, the polymorphism natural transformations entail
-%leaves no nontrivial way to incorporate data from the second argument
-%to a function with a type like that above into a fold over its first
-%argument.
-In fact, even for ADTs there is a difference between which folds over
-them we can type when ADTs are viewed as ADTs --- i.e., as fixpoints
-of types interpreted as first-order functors --- and which folds we
-can type when ADTs are viewed as proper nested types --- i.e., as
-fixpoints of types that are interpreted as higher-order functors ---
-because the arity of the recursion variable $\phi$ in the fixpoint in
-the return type of $\fold$ must equal the number of variables bound by
-its $\Nat$ construct. For ADTs this number can be $0$, making it
-possible to write types such as $\alpha; \emptyset \vdash
-\Nat^\emptyset \ (\List\,\alpha)\,(\Nat^\emptyset \alpha\,(\List (\mu
-\beta. 1 + \beta)))$, all of whose argument types are parameterized
-over the same type $\alpha$, which are not possible for nested types
-even when they are semantically equivalent to ADTs.
-%Currying the type of the function does not help either, since the
-%codomain of a fold applied to an algebra must be a functor, and this
-%clearly is not the case for a function on nested types that takes
-%more than one input involving the same type variable.
-%
-%For example, we can represent the type of a zip-like function for bushes.
-%
-%A well-known deeper limitation on folds is that a fold over a nested
-%type must take a natural transformation as an argument and return a
-%natural transformation; this issue is discussed in more detail in
-%Section~\ref{sec:conclusion},
-%Unfortunately, the polymorphism natural transformations
-%entail leaves no nontrivial way to incorporate data from the second
-%argument to a function with a type like that above into a fold over
-%its first argument.
-Our calculus can, however, express types of recursive functions
-that take
-%    as arguments
-%one proper nested type and as many other (possibly proper)
-%as many proper nested types as we wish
+respectively.
+
+Unfortunately, our calculus cannot express types of recursive
+functions --- such as a concatenation function for perfect trees or a
+zip function for bushes --- that take as inputs a nested type and an
+argument of another type, both of which are parameterized over the
+same variable. The fundamental issue is that recursion is expressible
+only via $\fold$, which produces natural transformations in some
+variables $\ol\alpha$ from $\mu$-types to other functors $F$. The
+restrictions on $\Nat$-types entail that $F$ cannot itself be a
+$\Nat$-type containing $\ol{\alpha}$, so, e.g., $\Nat^\alpha
+(\mathit{PTree}\,\alpha) \,(\Nat^\emptyset (\mathit{PTree}\,\alpha)\,
+(\mathit{PTree}\,(\alpha \times \alpha)))$ is not well-typed.
+Uncurrying gives $\Nat^\alpha (\mathit{PTree}\,\alpha \times
+\mathit{PTree}\,\alpha)\, (\mathit{PTree}\,(\alpha \times \alpha))$,
+which is well-typed, but $\mathsf{fold}$ cannot produce a term of this
+type because $\mathit{PTree}\,\alpha \times \mathit{PTree}\,\alpha$ is
+not a $\mu$-type.  Our calculus can, however, express types of
+recursive functions that take multiple nested types as arguments,
+provided they are parameterized over disjoint sets of type variables
+and the return type of the function is parameterized over only the
+variables occurring in the type of its final argument.  Even for ADTs
+there is a difference between which folds over them we can type when
+they are viewed as ADTs (i.e., as fixpoints of first-order functors)
+versus as proper nested types (i.e., as fixpoints of higher-order
+functors). This is because, in the return type of $\mathsf{fold}$, the
+arguments of the $\mu$-type must be variables bound by $\Nat$.  For
+ADTs, the $\mu$-type takes no arguments, making it possible to write
+recursive functions, such as a concatenation function for lists of
+type $\alpha ; \emptyset \vdash \Nat^\emptyset \ (\mu \beta. \onet +
+\alpha \times \beta)\, (\Nat^\emptyset (\mu \beta. \onet + \alpha
+\times \beta) \, (\mu \beta. \onet + \alpha \times \beta))$.  This is
+not possible for nested types --- even when they are semantically
+equivalent to ADTs.
+
+Interestingly, even some recursive functions of a single proper nested
+type --- e.g., a reverse function for bushes --- cannot be expressed
+as folds because the algebra arguments needed to define them are again
+recursive functions with types of the same problematic form as the
+type of, e.g., a zip function for perfect trees.  Expressivity of
+folds for nested types has long been a vexing issue, and this is
+naturally inherited by our calculus.  Adding more expressive recursion
+combinators could help, but since this is orthogonal to the issue of
+parametricity in the presence of primitive nested types we do not
+consider it further here.
+
+{\color{red} Integerate next paragraph into above discussion?} Our
+calculus can, however, express types of recursive functions that take
 multiple nested types as arguments, provided those nested types are
-parameterized over disjoint sets of type variables and the return type of
-the function is parameterized over only
-%any or all of
-the variables occurring in the type of its final argument.
-Interestingly, even some recursive functions of a single proper
-nested type --- e.g., a $\mathtt{reverseBush}$ function that is a true
-involution --- cannot be expressed as folds because the algebra
-arguments needed to define them are again recursive functions
+parameterized over disjoint sets of type variables and the return type
+of the function is parameterized over only the variables occurring in
+the type of its final argument.  Interestingly, even some recursive
+functions of a single proper nested type --- e.g., a
+$\mathtt{reverseBush}$ function that is a true involution --- cannot
+be expressed as folds because the algebra arguments needed to define
+them are again recursive functions
 %  cannot be defined. This happens, for
@@
 %
-Adding more expressive recursion combinators could help, too.
-%enhance the expressivity of our calculus
-But since the expressivity issue for folds for nested types long
-pre-dates, and is somewhat orthogonal to, the issue of parametricity
-for languages that construct nested types as fixpoints, we do not
-consider it further in this paper other than to identify, in
-Section~\ref{sec:conclusion}, some ideas for addressing it in future
-work.
-%These and other extensions are discussed
-
-The presence of the ``extra'' functorial variables in $\ol{\gamma}$ in
-the rules for $\map^{\ol{F},\ol{G}}_H$, $\tin_H$, and $\fold^F_H$
-merit special mention. They allow us to map or fold polymorphic
-functions over nested types. Consider, for example, the polymorphic
-function $\mathit{flatten} : \Nat^\beta
-(\mathit{PTree}\,\beta)\,(\mathit{List}\,\beta)$ that uniformly maps
-perfect trees to lists. Even in the absence of extra variables the
-instance of $\map$ required to map each non-functorial monomorphic
-instantiation of $\mathit{flatten}$ over a list of perfect trees is
-well-typed:
-\[\begin{array}{l}
-\AXC{$\Gamma;\alpha \vdash \mathit{List} \, \alpha$}
-\AXC{$\Gamma;\emptyset \vdash \mathit{PTree}\,F$}
-\AXC{$\Gamma;\emptyset \vdash \mathit{List}\,G$}
-\TIC{$\Gamma;\emptyset \,|\, \emptyset \vdash \map^{\mathit{PTree}\,F,
-    \,\mathit{List}\,G}_{\mathit{List}\,\alpha} :
-  \Nat^\emptyset (\Nat^\emptyset\,(\mathit{PTree}\, F)\,(\mathit{List}\, G))\,
-  (\Nat^\emptyset\,(\mathit{List}\, (\mathit{PTree}\,
-  F))\,(\mathit{List}\, (\mathit{List}\, G)))$} \DisplayProof
-\end{array}\]
-But in the absence of $\ol \gamma$, the instance
-\[\Gamma;\emptyset~|~\emptyset \vdash
-\map^{\mathit{PTree}\,\beta,\mathit{List}\,\beta}_{\mathit{List}\,\alpha}
-: \Nat^\emptyset ( \Nat^\beta(\mathit{PTree}\,
-\beta)\,(\mathit{List}\, \beta))\, (\Nat^\beta\,(\mathit{List}\,
-(\mathit{PTree}\, 
-\beta))\,(\mathit{List}\, (\mathit{List}\, \beta)))\]
-%\[\Gamma;\emptyset \,|\, x : \Nat^\beta(\mathit{PTree}\,
-%\beta)\,(\mathit{List}\, \beta) \vdash
-%\map^{\mathit{PTree}\,\beta,\mathit{List}\,\beta}_{\mathit{List}\,\alpha}\,x
-%: \Nat^\beta\,(\mathit{List}\,
-%(\mathit{PTree}\, \beta))\,(\mathit{List}\, (\mathit{List}\,
-%\beta))\] of $\mathtt{map}$
-required to map the {\em polymorphic} $\mathit{flatten}$ function over
-a list of perfect trees is not: in that setting the functorial
-contexts for $F$ and $G$ in the rule for $\map^{F,G}_H$ would have to
-be empty, but because the polymorphic $\mathit{flatten}$ function is
-natural in some variable, say $\delta$, it cannot possibly have a
-type of the form $\Nat^\emptyset F\, G$ that would be required for it
-to be the function input to $\map$. Since untypeability of this
-instance of $\map$ is unsatisfactory in a polymorphic calculus, where
-we naturally expect to be able to manipulate entire polymorphic
-functions rather than just their monomorphic instances, we use the
-``extra'' variables in $\ol \gamma$ to remedy the
-situation. Specifically, Definition~\ref{def:well-formed-terms}
-ensures that the instance of $\map$ needed to map the polymorphic
-$\mathit{flatten}$ function is typeable as follows:
-\[\begin{array}{l}
-\AXC{$\Gamma;\alpha,\beta \vdash \mathit{List} \, \alpha$}
-\AXC{$\Gamma;\beta \vdash \mathit{PTree}\,\beta \hspace*{0.3in}
-  \Gamma;\beta \vdash \mathit{List}\,\beta$}
-\BIC{$\Gamma;\emptyset~|~\emptyset \vdash
-  \map^{F,G}_{\mathit{List}\,\alpha}  : \Nat^\emptyset
-  (\Nat^\beta(\mathit{PTree}\, 
-  \beta)\,(\mathit{List}\, \beta))\,
- (\Nat^\beta\,(\mathit{List}\,
-  (\mathit{PTree}\, \beta))\,(\mathit{List}\, (\mathit{List}\,
-  \beta)))$}
-\DisplayProof
-\end{array}\]
-Similar remarks explain the appearance of $\ol \gamma$ in the typing
-rules for $\tin$ and $\fold$.
-
-\vspace*{0.05in}
 
@@
 holds for {\em all} elements of a type's interpretation, not just
-those that are interpretations of terms.
-%i.e., $t_0(A)$ and $t_0(B)$ map related arguments to
-%related results.
-Similar theorems are expected to hold for types and terms with any
-number of free variables.
-
-The key to proving the Identity Extension Lemma
-(Theorem~\ref{thm:iel}) in our setting is a familiar ``cutting down''
-of the interpretations of universally quantified types to include only
-the ``parametric'' elements; the relevant types in our calculus are
-the $\Nat$-types.
-%(See, e.g.,~\cite{atk12,bfss90,gjfor15,rey83}).
-This cutting down requires, as usual, that the set interpretations of
-our types are defined simultaneously with their relational
-interpretations. We give the set interpretations for our types in
-Section~\ref{sec:set-interp} and give their relational interpretations
-in Section~\ref{sec:rel-interp}.  While the set interpretations are
+those that are interpretations of terms. Similar theorems are
+expected to hold for types and terms with any number of free
+variables.
+
+The key to proving the Identity Extension Lemma in our setting
+(Theorem~\ref{thm:iel}) is a familiar ``cutting down'' of the
+interpretations of universally quantified types to include only the
+``parametric'' elements; the relevant types in our calculus are the
+$\Nat$-types.  This cutting down requires, as usual, that the set
+interpretations of types (Section~\ref{sec:set-interp}) are defined
+simultaneously with their relational interpretations
+(Section~\ref{sec:rel-interp}). While the set interpretations are
 relatively straightforward, their relation interpretations are less
@@
 that they are well-defined. We develop these conditions in
-Section~\ref{sec:rel-interp}, which separates
-Definitions~\ref{def:set-sem} and~\ref{def:rel-sem} in space, but
-otherwise has no impact on the fact that they are given by mutual
-induction.
+Sections~\ref{sec:set-interp} and~\ref{sec:rel-interp}. This separates
+our set and relational interpretations in space, but otherwise has no
+impact on the fact that they are given by mutual induction.
 
@@
 
-%A poset $\mcD = (D,\leq)$ is \emph{$\omega$-directed} if every finite
-%subset of $D$ has an upper bound. When $\mcD$ is considered as a
-%category, we write $d \in \mcD$ to indicate that $d$ is an object of
-%$\mcD$ (i.e., $d \in D$). An {\em $\omega$-directed colimit} in a
-%category $\mcC$ is a colimit of a diagram $F : {\mathcal D} \to \mcC$,
-%where $\mathcal D$ is an $\omega$-directed poset. A category $\mcC$ is
-%{\em $\omega$-cocomplete} if it has all $\omega$-directed colimits. A
-%cocomplete category is one that has all colimits.
-%
-%If $\mcA$ and $\mcC$ are $\omega$-cocomplete, then the functor $F :
-%\mcA \to \mcC$ is {\em $\omega$-cocontinuous} if it preserves
-%$\omega$-directed colimits.  If $\mcA$ is locally small, then an
-%object $A$ of $\mcA$ is {\em finitely presentable} if the functor
-%$\mathsf{Hom}_\mcA(A, -) : \mcA \to \Set$ preserves $\omega$-directed
-%colimits, i.e., if for every $\omega$-directed poset $\mathcal D$ and
-%every functor $F : {\mathcal D} \to \mcC$, there is a canonical
-%isomorphism $\colim{d \in \mcD}{\mathsf{Hom}_\mcA(A,Fd)} \simeq
-%\mathsf{Hom}_{\mcA}(A, \colim{d \in \mcD}{Fd})$. A category $\mcA$ is
-%       {\em finitely accessible} if it is $\omega$-cocomplete and has
-%       a set $\mcA_0$ of finitely presentable objects such that every
-%%       object is an $omega$-directed colimit of objects in $\mcA_0$;
-%       it is {\em locally finitely presentable} if it is finitely
-%       accessible and cocomplete.
 We interpret types in our calculus as $\omega$-cocontinuous functors
@@
 When convenient we identify a functor in $[\set^0, \set]$ with its
-value on $\ast$ and consider a set environment to map a type
-variable of arity $0$ to
-%an $\omega$-cocontinuous functor from $\set^0$ to $\set$, i.e., to
-a set.  If $\ol{\alpha} = \{\alpha_1,...,\alpha_k\}$ and $\ol{A} =
-\{A_1,...,A_k\}$, then we write $\rho[\ol{\alpha := A}]$ for the set
-environment $\rho'$ such that $\rho' \alpha_i = A_i$ for $i = 1,...,k$
-and $\rho' \alpha = \rho \alpha$ if $\alpha \not \in
+value on $\ast$ and consider a set environment to map a type variable
+of arity $0$ to a set.  If $\ol{\alpha} = \{\alpha_1,...,\alpha_k\}$
+and $\ol{A} = \{A_1,...,A_k\}$, then we write $\rho[\ol{\alpha := A}]$
+for the set environment $\rho'$ such that $\rho' \alpha_i = A_i$ for
+$i = 1,...,k$ and $\rho' \alpha = \rho \alpha$ if $\alpha \not \in
 \{\alpha_1,...,\alpha_k\}$.  If $\rho$ is a set environment we write
@@
 for every type variable $v$.
-%; see Definition~\ref{def:reln-env} for the complete definition of a
-%relation environment.
-The relational interpretations appearing in the second clause of
-Definition~\ref{def:set-sem} are given in full in
+
+We can now define our set interpretations.  The relational
+interpretations appearing in the second clause of
+Definition~\ref{def:set-sem} are given in
 Definition~\ref{def:rel-sem}.
@@
 \begin{dfn}\label{def:set-sem}
-%Let $\rho$ be a set environment.
-The {\em set interpretation} $\setsem{\cdot} : \F \to [\setenv, \set]$
-is defined by
+{\color{red} Check clause for $\mu$-types.} The {\em set
+  interpretation} $\setsem{\cdot} : \F \to [\setenv, \set]$ is defined
+by:
 \begin{align*}
@@
   \setsem{\Gamma;\Phi \vdash \onet}\rho &= 1\\
-%  \setsem{\Gamma;\emptyset \vdash v} \rho &= \rho v \mbox{
-%    if } v \in \tvars^0 \\ 
   \setsem{\Gamma; \emptyset
@@
 ${\setsem{\Gamma;\ol{\alpha} \vdash F}\rho[ \ol{\alpha := S}]}$ to
-${\setsem{\Gamma;\ol{\alpha} \vdash G}\rho[\ol{\alpha := S}]}$. But
-there are only $\set$-many choices of morphisms between any two
-objects since $\set$ is locally small.  We note that $\setsem{\Gamma;
-  \emptyset \vdash \Nat^{\ol\alpha} F\,G}$ is $\omega$-cocontinuous because
-it is a constant functor (and that allowing functorial variables not
-in $\ol \alpha$ in $G$ fails here, as shown in
-Section~\ref{sec:terms}).
-%follows from $\omega$-cocontinuity of $\setsem{\Gamma;\ol\alpha
-%\vdash F}$, $\setsem{\Gamma;\ol\alpha \vdash G}$, and the equality
-%relation environment $\Eq$.
-Interpretations of $\Nat$-types ensure
-that $\setsem{\Gamma \vdash F \to G}$ and $\setsem{\Gamma \vdash
-  \forall \ol\alpha. F}$ are as expected in any parametric model.
-
-To make sense of the last clause in
-Definition~\ref{def:set-sem}, we need to know that, for each $\rho \in
-\setenv$, $T^\set_{H,\rho}$ is an $\omega$-cocontinuous endofunctor on
-$[\set^k, \set]$, and thus admits a fixpoint.  Since
-$T_{H,\rho}^\set$ is defined in terms of $\setsem{\Gamma;\ol\gamma,\phi,
-  \ol{\alpha} \vdash H}$, this means that interpretations of types
-must be such functors, which in turn means that the actions of set
-interpretations of types on objects and on morphisms in $\setenv$ are
-intertwined. Fortunately, we know from~\cite{jp19} that, for every
-$\Gamma; \ol{\alpha} \vdash G$, $\setsem{\Gamma; \ol{\alpha} \vdash
-  G}$ is actually in $[\set^k,\set]$ where $k = |\ol \alpha|$. This
-means that for each $\setsem{\Gamma; \ol\gamma, \phi^k, \ol{\alpha} \vdash
-  H}$, the corresponding operator $T^\set_{H}$ can be extended to a
-{\em functor} from $\setenv$ to $[[\set^k,\set],[\set^k,\set]]$. The
-action of $T^\set_H$ on an object $\rho \in \setenv$ is given by the
-higher-order functor $T_{H,\rho}^\set$, whose actions on objects
-(functors in $[\set^k, \set]$) and morphisms (natural transformations
-between such functors) are given in Definition~\ref{def:set-sem}. Its
-action on a morphism $f : \rho \to \rho'$ is the higher-order natural
-transformation $T^\set_{H,f} : T^\set_{H,\rho} \to T^\set_{H,\rho'}$
-whose action on $F : [\set^k,\set]$ is the natural transformation
-$T^\set_{H,f}\, F : T^\set_{H,\rho}\,F \to T^\set_{H,\rho'}\,F$ whose
-component at $\ol{A}$ is $(T^\set_{H,f}\, F)_{\ol{A}} =
-\setsem{\Gamma; \ol\gamma,\phi,\ol{\alpha} \vdash H}f[\phi :=
-  \id_F][\ol{\alpha := \id_A}]$. The next definition uses the functor
-$T^\set_H$ to define the actions of functors interpreting types on
-morphisms between set environments.
-
-\begin{comment}
-To make sense of the last clause in
-Definition~\ref{def:set-sem}, we need to know that, for each $\rho \in
-\setenv$, $T^\set_{H,\rho}$ is an $\omega$-cocontinuous endofunctor on
-$[\set^k, \set]$, and thus admits a fixpoint.  Since
-$T_{H,\rho}^\set$ is defined in terms of $\setsem{\Gamma;\ol\gamma,\phi,
-  \ol{\alpha} \vdash H}$, this means that interpretations of types
-must be such functors, which in turn means that the actions of set
-interpretations of types on objects and on morphisms in $\setenv$ are
-intertwined. Fortunately, we know from~\cite{jp19} that, for every
-$\Gamma; \ol{\alpha} \vdash G$, $\setsem{\Gamma; \ol{\alpha} \vdash
-  G}$ is actually in $[\set^k,\set]$ where $k = |\ol \alpha|$. This
-means that for each $\setsem{\Gamma; \ol\gamma, \phi^k, \ol{\alpha} \vdash
-  H}$, the corresponding operator $T^\set_{H}$ can be extended to a
-{\em functor} from $\setenv$ to $[[\set^k,\set],[\set^k,\set]]$. The
-action of $T^\set_H$ on an object $\rho \in \setenv$ is given by the
-higher-order functor $T_{H,\rho}^\set$, whose actions on objects
-(functors in $[\set^k, \set]$) and morphisms (natural transformations
-between such functors) are given in Definition~\ref{def:set-sem}. Its
-action on a morphism $f : \rho \to \rho'$ is the higher-order natural
-transformation $T^\set_{H,f} : T^\set_{H,\rho} \to T^\set_{H,\rho'}$
-whose action on $F : [\set^k,\set]$ is the natural transformation
-$T^\set_{H,f}\, F : T^\set_{H,\rho}\,F \to T^\set_{H,\rho'}\,F$ whose
-component at $\ol{A}$ is $(T^\set_{H,f}\, F)_{\ol{A}} =
-\setsem{\Gamma; \ol\gamma,\phi,\ol{\alpha} \vdash H}f[\phi :=
-  \id_F][\ol{\alpha := \id_A}]$. The next definition uses the functor
-$T^\set_H$ to define the actions of functors interpreting types on
-morphisms between set environments.
-\end{comment}
-
+${\setsem{\Gamma;\ol{\alpha} \vdash G}\rho[\ol{\alpha := S}]}$, but
+only $\set$-many choices of morphisms between any two objects since
+$\set$ is locally small. Also, $\setsem{\Gamma; \emptyset \vdash
+  \Nat^{\ol\alpha} F\,G}$ is $\omega$-cocontinuous since it is
+constant on $\omega$-directed sets. Interpretations of $\Nat$-types
+ensure that $\setsem{\Gamma \vdash F \to G}$ and $\setsem{\Gamma
+  \vdash \forall \ol\alpha. F}$ are as expected in any parametric
+model.
+
+To make sense of the last clause in Definition~\ref{def:set-sem}, we
+need to know that, for each $\rho \in \setenv$, $T^\set_{H,\rho}$ is
+an $\omega$-cocontinuous endofunctor on $[\set^k, \set]$, and thus
+admits a fixpoint.  Since $T_{H,\rho}^\set$ is defined in terms of
+$\setsem{\Gamma;\ol\gamma,\phi, \ol{\alpha} \vdash H}$, this means
+that interpretations of types must be such functors, which in turn
+means that the actions of set interpretations of types on objects and
+on morphisms in $\setenv$ are intertwined. Fortunately, we know
+from~\cite{jp19} that, for every $\Gamma; \ol{\alpha} \vdash G$,
+$\setsem{\Gamma; \ol{\alpha} \vdash G}$ is actually in $[\set^k,\set]$
+where $k = |\ol \alpha|$. This means that for each $\setsem{\Gamma;
+  \ol\gamma, \phi^k, \ol{\alpha} \vdash H}$, the corresponding
+operator $T^\set_{H}$ can be extended to a {\em functor} from
+$\setenv$ to $[[\set^k,\set],[\set^k,\set]]$. The action of $T^\set_H$
+on an object $\rho \in \setenv$ is given by the higher-order functor
+$T_{H,\rho}^\set$, whose actions on objects (functors in $[\set^k,
+  \set]$) and morphisms (natural transformations) between them are
+given in Definition~\ref{def:set-sem}. Its action on a morphism $f :
+\rho \to \rho'$ is the higher-order natural transformation
+$T^\set_{H,f} : T^\set_{H,\rho} \to T^\set_{H,\rho'}$ whose action on
+$F : [\set^k,\set]$ is the natural transformation $T^\set_{H,f}\, F :
+T^\set_{H,\rho}\,F \to T^\set_{H,\rho'}\,F$ whose component at
+$\ol{A}$ is $(T^\set_{H,f}\, F)_{\ol{A}} = \setsem{\Gamma;
+  \ol\gamma,\phi,\ol{\alpha} \vdash H}f[\phi := \id_F][\ol{\alpha :=
+    \id_A}]$.
+
+Using $T^\set_H$, we can define the functorial action of set
+interpretation.
 \begin{dfn}\label{def:set-sem-funcs}
@@
 $\setsem{\Gamma;\Phi \vdash F}f$ of\, $\setsem{\Gamma;\Phi \vdash F}$
-on the morphism $f$ is given as follows:
+on $f$ is given by:
 \begin{itemize}
@@
   \onet}f = \id_1$
-%\item If \,$\Gamma,v;\emptyset \vdash v$ then
-%  $\setsem{\Gamma,v;\emptyset \vdash v}f = \id_{\rho v}$
 \item If \,$\Gamma; \emptyset
@@
       \vdash \Nat^{\ol{\alpha}}\,F\,G}\rho}$
-%\hspace*{0.2in} $\lambda t : \setsem{\Gamma;\emptyset\vdash
-%    \Nat^{\ol{\alpha}}\,F\,G}\rho.\, 
-%  (\lambda \ol{A}. \setsem{\Gamma; \ol\alpha \vdash G}f[\ol{\alpha
-%      := id_A}]) \circ t$  
 \item If \,$\Gamma;\Phi \vdash \phi \ol{F}$ then
@@
 A {\em $k$-ary relation transformer} $F$ is a triple $(F^1, F^2,F^*)$,
-where $F^1,F^2 : [\set^k,\set]$ are functors, $F^* : [\rel^k, \rel]$
-is a functor, if $R_1:\rel(A_1,B_1),...,R_k:\rel(A_k,B_k)$, then $F^*
-\ol{R} : \rel(F^1 \ol{A}, F^2 \ol{B})$, and if $(\alpha_1, \beta_1)
-\in \Homrel(R_1,S_1),..., (\alpha_k, \beta_k) \in \Homrel(R_k,S_k)$
-then $F^* \ol{(\alpha, \beta)} = (F^1 \ol{\alpha}, F^2 \ol{\beta})$.
+where
+\begin{itemize}
+\item $F^1,F^2 : [\set^k,\set]$ and $F^* : [\rel^k, \rel]$ are
+  functors
+\item If $R_1:\rel(A_1,B_1),...,R_k:\rel(A_k,B_k)$, then $F^* \ol{R} :
+  \rel(F^1 \ol{A}, F^2 \ol{B})$
+\item If $(\alpha_1, \beta_1) \in \Homrel(R_1,S_1),..., (\alpha_k,
+  \beta_k) \in \Homrel(R_k,S_k)$ then $F^* \ol{(\alpha, \beta)} = (F^1
+  \ol{\alpha}, F^2 \ol{\beta})$
+\end{itemize}
 We define $F\ol{R}$ to be $F^*\overline{R}$ and
@@
 \in F^*\ol{S}$. When convenient we identify a $0$-ary relation
-transformer $(A,B,R)$ with $R : \rel(A,B)$. We may also write $\pi_1
-F$ for $F^1$ and $\pi_2 F$ for $F^2$. We extend these conventions to
-relation environments, introduced in Definition~\ref{def:reln-env}
-below, in the obvious way.
+transformer $(A,B,R)$ with $R : \rel(A,B)$, and write $\pi_1 F$ for
+$F^1$ and $\pi_2 F$ for $F^2$. Below we extend these conventions to
+relation environments in the obvious ways.
 
@@
 \begin{itemize}
-\item An object of $RT_k$ is a relation transformer.
+\item An object of $RT_k$ is a relation transformer
 \item A morphism $\delta : (G^1,G^2,G^*) \to (H^1,H^2,H^*)$ in $RT_k$
-  is a pair of natural transformations $(\delta^1, \delta^2)$ where
-  $\delta^1 : G^1 \to H^1$, $\delta^2 : G^2 \to H^2$ such that, for
-  all $\ol{R : \rel(A, B)}$, if $(x, y) \in G^*\ol{R}$ then
-  $(\delta^1_{\ol{A}}x, \delta^2_{\ol{B}}y) \in H^*\ol{R}$.
+  is a pair of natural transformations $(\delta^1, \delta^2)$, where
+  $\delta^1 : G^1 \to H^1$ and $\delta^2 : G^2 \to H^2$, such that,
+  for all $\ol{R : \rel(A, B)}$, if $(x, y) \in G^*\ol{R}$ then
+  $(\delta^1_{\ol{A}}x, \delta^2_{\ol{B}}y) \in H^*\ol{R}$
 \item Identity morphisms and composition are inherited from the
-  category of functors on $\set$.
+  category of functors on $\set$
 \end{itemize}
@@
 \item $H^*$ is a functor from $RT_k$ to $[\rel^k,\rel]$
-\item for all $\overline{R : \rel(A,B)}$,
+\item For all $\overline{R : \rel(A,B)}$,
   $\pi_1((H^*(\delta^1,\delta^2))_{\overline{R}}) = (H^1
@@
 Definition~\ref{def:RT-functor} implicitly requires that the following
-three conditions hold:\,{\em i})
-if $R_1:\rel(A_1,B_1),...,R_k:\rel(A_k,B_k)$, then
-  $H^*(F^1,F^2,F^*) \ol{R} : \rel(H^1F^1 \ol{A}, H^2F^2 
-  \ol{B})$;
-  %In other words, $\pi_1 (H^*(F^1,F^2,F^*) \ol{R}) = H^1F^1
-  %\ol{A}$ and $\pi_2 (H^*(F^1,F^2,F^*) \ol{R}) = H^2F^2 \ol{B}$.
-{\em ii}) if $(\alpha_1, \beta_1) \in \Homrel(R_1,S_1),..., (\alpha_k,
-  \beta_k) \in \Homrel(R_k,S_k)$, then
-  $H^*(F^1,F^2,F^*)\, \ol{(\alpha, \beta)} = (H^1F^1\ol{\alpha}, H^2F^2
-  \ol{\beta})$; and
-  %In other words, $\pi_1 (H^*(F^1,F^2,F^*) \ol{(\alpha,
-  %  \beta)}) = H^1F^1 \ol{\alpha}$ and $\pi_2 (H^*(F^1,F^2,F^*) 
-  % \ol{(\alpha, \beta)}) = H^2F^2 \ol{\beta}$.  {\em iii}) if
-  $(\delta^1,\delta^2) : (F^1,F^2,F^*)\to (G^1,G^2,G^*)$ and
+three conditions hold:
+\begin{itemize}
+\item $H^*(F^1,F^2,F^*) \ol{R} : \rel(H^1F^1 \ol{A}, H^2F^2 \ol{B})$
+  whenever $R_1:\rel(A_1,B_1),...,R_k:\rel(A_k,B_k)$
+\item $H^*(F^1,F^2,F^*)\, \ol{(\alpha, \beta)} = (H^1F^1\ol{\alpha},
+  H^2F^2 \ol{\beta})$ whenever $(\alpha_1, \beta_1) \in
+  \Homrel(R_1,S_1),..., (\alpha_k, \beta_k) \in \Homrel(R_k,S_k)$
+\item If $(\delta^1,\delta^2) : (F^1,F^2,F^*)\to (G^1,G^2,G^*)$ and
   $R_1:\rel(A_1,B_1),...,R_k:\rel(A_k,B_k)$, then
   $((H^1\delta^1)_{\ol{A}}x, (H^2\delta^2)_{\ol{B}}y) \in
-  H^*(G^1,G^2,G^*)\ol{R}$ whenever $(x, y) \in
-  H^*(F^1,F^2,F^*)\ol{R}$. Note, however, that this last condition is
+  H^*(G^1,G^2,G^*)\ol{R}$ whenever $(x, y) \in H^*(F^1,F^2,F^*)\ol{R}$
+\end{itemize}
+  Note, however, that this last condition is
   automatically satisfied because it is implied by the third bullet
@@
 \begin{dfn}\label{def:RT-nat-trans}
-If $H$ and $K$ are endofunctors on $RT_k$, then a {\em natural
-  transformation} $\sigma : H \to K$ is a pair $\sigma = (\sigma^1,
-\sigma^2)$, where $\sigma^1 : H^1 \to K^1$ and $\sigma^2 : H^2 \to
-K^2$ are natural transformations between endofunctors on
-$[\set^k,\set]$ and the component of $\sigma$ at $F
-%= (F^1,F^2,F^*)
-\in RT_k$ is given by $\sigma_F = (\sigma^1_{F^1}, \sigma^2_{F^2})$.
+For endofunctors $H$ and $K$ on $RT_k$, a {\em natural transformation}
+$\sigma : H \to K$ is a pair $\sigma = (\sigma^1, \sigma^2)$, where
+$\sigma^1 : H^1 \to K^1$ and $\sigma^2 : H^2 \to K^2$ are natural
+transformations between endofunctors on $[\set^k,\set]$ and the
+component of $\sigma$ at $F = (F^1,F^2,F^*) \in RT_k$ is given by
+$\sigma_F = (\sigma^1_{F^1}, \sigma^2_{F^2})$.
 \end{dfn}
-Definition~\ref{def:RT-nat-trans} entails that $\sigma^i_{F^i}$ must
-be natural in $F^i : [\set^k,\set]$, and, for every $F$, both
+Definition~\ref{def:RT-nat-trans} entails that $\sigma^i_{F^i}$ is
+natural in $F^i : [\set^k,\set]$, and, for every $F$, both
 $(\sigma^1_{F^1})_{\overline{A}}$ and
-$(\sigma^2_{F^2})_{\overline{A}}$ must be natural in $\overline{A}$.
+$(\sigma^2_{F^2})_{\overline{A}}$ are natural in $\overline{A}$.
 Moreover, since the results of applying $\sigma$ to $k$-ary relation
@@
 \vspace*{0.1in}
-
-%{\color{blue} More context? Weave cocontinuity requirement for fixed
-%  points through text better.}
-Critically, we can compute $\omega$-directed colimits in $RT_k$: it is
-not hard to see that if $\cal D$ is an $\omega$-directed set then
+Critically, we can compute $\omega$-directed colimits in
+$RT_k$. Indeed, if $\cal D$ is an $\omega$-directed set then
 $\colim{d \in {\cal D}}{(F^1_d, F^2_d,F^*_d)} = (\colim{d \in {\cal
@@
 $\rel^k$ to $\rel$.  Also let $0$ denote either the initial object of
-$\set$ or the initial object of $\rel$, as appropriate.  Observing
-that, for every $k$, $K^\set_0$ is initial in $[\set^k,\set]$, and
-$K^\rel_0$ is initial in $[\rel^k,\rel]$, we have that, for each $k$,
-$K_0 = (K^\set_0,K^\set_0,K^\rel_0)$ is initial in $RT_k$. Thus, if $T
-= (T^1,T^2,T^*) : RT_k \to RT_k$ is an endofunctor on $RT_k$ then we
-can define the relation transformer $\mu T$ to be $\colim{n \in
-  \nat}{T^n K_0}$.
-%Then Lemma~\ref{lem:colimits} shows $\mu T$ is indeed a relation
-%transformer, and that it is given explicitly by
-It is not hard to see that $\mu T$ is given explicitly as 
+either $\set$ or $\rel$, as appropriate.  Observing that, for every
+$k$, $K^\set_0$ is initial in $[\set^k,\set]$, and $K^\rel_0$ is
+initial in $[\rel^k,\rel]$, we have that, for each $k$, $K_0 =
+(K^\set_0,K^\set_0,K^\rel_0)$ is initial in $RT_k$. Thus, if $T =
+(T^1,T^2,T^*) : RT_k \to RT_k$ is an endofunctor on $RT_k$ then we can
+define the relation transformer $\mu T$ to be $\colim{n \in \nat}{T^n
+  K_0}$. It is not hard to see that $\mu T$ is given explicitly as
 \begin{equation}\label{eq:mu}
-  %\colim{n \in \nat}{T^n K_0}
-  \mu T = (\mu T^1,\mu T^2,
-\colim{n \in \nat}{(T^nK_0)^*})
+\mu T = (\mu T^1,\mu T^2, \colim{n \in \nat}{(T^nK_0)^*})
 \end{equation}
-and that, as our notation suggests, it really is a fixpoint for $T$ if
-$T$ is $\omega$-cocontinuous:
+Moreover, $\mu T$ really is a fixpoint for $T$ if $T$ is
+$\omega$-cocontinuous:
 \begin{lemma}\label{lem:fp}
@@
 
-\begin{comment}
-\begin{proof}
-We have $T(\mu T) = T(\colim{n \in \nat}{(T^nK_0)}) \cong \colim{n \in
-\nat}{T(T^nK_0)} =
-\mu T$.
-\end{proof}
-In fact, the isomorphism in Lemma~\ref{lem:fp} is given by the
-morphisms $(in_1, in_2) : T(\mu T) \to \mu T$ and $(in_1^{-1},
-in_2^{-1}) : \mu T \to T(\mu T)$ in $RT_k$. It is worth noting that the
-latter is always a morphism in $RT_k$, but the former isn't necessarily
-a morphism in $RT_k$ unless $T$ is $\omega$-cocontinuous.
-\end{comment}
-
 It is worth noting that the third component in Equation~(\ref{eq:mu})
@@
 transformers, rather than a fixpoint of an endofunctor on
-$[\rel^k,\rel]$. That there is an asymmetry between the first two
-components of $\mu T$ and its third reflects the important conceptual
-observation that the third component of an endofunctor on $RT_k$ need
-not be a functor on all of $[\rel^k,\rel]$. In particular, although we
-can define $T_{H,\rho}\, F$ for a relation transformer $F$ in
+$[\rel^k,\rel]$. There is thus an asymmetry between the first two
+components of $\mu T$ and its third component, which reflects the
+important conceptual observation that the third component of an
+endofunctor on $RT_k$ need not be a functor on all of
+$[\rel^k,\rel]$. {\color{red} KEEP? In particular, although we can
+  define $T_{H,\rho}\, F$ for a relation transformer $F$ in
   Definition~\ref{def:rel-sem} below, it is not clear how we could
-define it for $F : [\rel^k,\rel]$.
+  define it for an arbitrary $F : [\rel^k,\rel]$. This observation
+  will also be critical in Section~\ref{sec:GADTs}.}
 
@@
 \end{dfn}
-When convenient we identify a $0$-ary relation transformer with the
-relation (transformer) that is its codomain and consider
-% With this convention,
-a relation environment to map a type variable of arity $0$
-%to a $0$-ary relation transformer, i.e.,
-to a relation.  We write $\rho[\ol{\alpha := R}]$ for the relation
-environment $\rho'$ such that $\rho' \alpha_i \, = R_i$ for $i =
-1,...,k$ and $\rho' \alpha = \rho\alpha$ if $\alpha \not \in
-\{\alpha_1,...,\alpha_k\}$.  If $\rho$ is a relation environment, we
-write $\pi_1 \rho$ and $\pi_2 \rho$ for the set environments mapping
-each type variable $\phi$ to the functors $(\rho\phi)^1$ and
-$(\rho\phi)^2$, respectively.
-
-We define, for each $k$, the notion of an $\omega$-cocontinuous
-functor from $\relenv$ to $RT_k$:
+We identify a $0$-ary relation transformer with the relation
+(transformer) that is its codomain and consider a relation environment
+to map a type variable of arity $0$ to a relation.  We write
+$\rho[\ol{\alpha := R}]$ for the relation environment $\rho'$ such
+that $\rho' \alpha_i \, = R_i$ for $i = 1,...,k$ and $\rho' \alpha =
+\rho\alpha$ if $\alpha \not \in \{\alpha_1,...,\alpha_k\}$.  If $\rho$
+is a relation environment, we write $\pi_1 \rho$ and $\pi_2 \rho$ for
+the set environments mapping each type variable $\phi$ to the functors
+$(\rho\phi)^1$ and $(\rho\phi)^2$, respectively.
+
 \begin{dfn}\label{def:relenv-functor}
-A functor $H : [\relenv, RT_k]$ is a triple $H = (H^1,H^2,H^*)$,
-where
+For each $k$, an $\omega$-cocontinuous functor $H : [\relenv, RT_k]$
+is a triple $H = (H^1,H^2,H^*)$, where
 \begin{itemize}
@@
 \item $H^*$ is a an object in $[\relenv,[\rel^k,\rel]]$
-\item for all $\overline{R : \rel(A,B)}$ and morphisms $f$ in
+\item For all $\overline{R : \rel(A,B)}$ and morphisms $f$ in
   $\relenv$, $\pi_1(H^*f \,{\overline{R}}) = H^1 (\pi_1
@@
 following analogues of the three conditions immediately following
-Definition~\ref{def:RT-functor}: {\em i}) if $R_1 :
-\rel(A_1,B_1),...,R_k : \rel(A_k,B_k)$, then $H^*\rho\, \ol{R} :
-\rel(H^1(\pi_1 \rho)\, \ol{A}, H^2(\pi_2 \rho)\, \ol{B})$;
-%In other words, $\pi_1 (H^*\rho \ol{R}) = H^1(\pi_1 \rho)
-%  \,\ol{A}$ and $\pi_2 (H^*\rho \ol{R}) = H^2 (\pi_2 \rho) \,\ol{B}$;
-{\em ii}) if $(\alpha_1, \beta_1) \in \Homrel(R_1,S_1),..., (\alpha_k,
-\beta_k) \in \Homrel(R_k,S_k)$, then $H^*\rho\, \ol{(\alpha, \beta)} =
-(H^1(\pi_1 \rho)\,\ol{\alpha}, H^2(\pi_2 \rho)\, \ol{\beta})$;
-%  In other words, $\pi_1 (H^*\rho\, \ol{(\alpha, \beta)}) = H^1(\pi_1
-%  \rho)\,\ol{\alpha}$ and $\pi_2 (H^*\rho\, \ol{(\alpha, \beta)}) =
-%  H^2(\pi_2 \rho)\,\ol{\beta}$;
-and {\em iii}) if $f : \rho \to \rho'$ and
+Definition~\ref{def:RT-functor}:
+\begin{itemize}
+\item $H^*\rho\, \ol{R} : \rel(H^1(\pi_1 \rho)\, \ol{A}, H^2(\pi_2
+  \rho)\, \ol{B})$ whenever $R_1 : \rel(A_1,B_1),...,R_k :
+  \rel(A_k,B_k)$
+\item $H^*\rho\, \ol{(\alpha, \beta)} = (H^1(\pi_1 \rho)\,\ol{\alpha},
+  H^2(\pi_2 \rho)\, \ol{\beta})$ whenever $(\alpha_1, \beta_1) \in
+  \Homrel(R_1,S_1),..., (\alpha_k, \beta_k) \in \Homrel(R_k,S_k)$
+\item If $f : \rho \to \rho'$ and
   $R_1:\rel(A_1,B_1),...,R_k:\rel(A_k,B_k)$, then $(H^1(\pi_1
   f)\,{\ol{A}}\,x, H^2(\pi_2 f)\,{\ol{B}}\,y) \in H^*\rho'\,\ol{R}$
-whenever $(x, y) \in H^*\rho\,\ol{R}$. As before, the last
-condition is automatically satisfied because it is implied by the
-third bullet point of Definition~\ref{def:relenv-functor}.
+  whenever $(x, y) \in H^*\rho\,\ol{R}$.
+\end{itemize}
+As before, the last condition is automatically satisfied because it is
+implied by the third bullet point of
+Definition~\ref{def:relenv-functor}.
 
 Considering $\relenv$ as a product $\Pi_{\phi^k \in \tvars \cup
-  \fvars} RT_k$, we extend the computation of $\omega$-directed
-colimits in $RT_k$ to compute colimits in $\relenv$ componentwise. We
-similarly extend the notion of an $\omega$-cocontinuous endofunctor on
-$RT_k$ componentwise to give a notion of $\omega$-cocontinuity for
-functors from $\relenv$ to $RT_k$.  Recalling from the start of this
-subsection that Definition~\ref{def:rel-sem} is given mutually
-inductively with Definition~\ref{def:set-sem} we can, at last, define:
+  \fvars} RT_k$, computation of $\omega$-directed colimits in $RT_k$
+extends componentwise to $\relenv$. Similarly, $\omega$-cocontinuity
+for endofunctors on $RT_k$ extends componentwise to functors from
+$\relenv$ to $RT_k$.  Recalling from the start of this subsection that
+Definition~\ref{def:rel-sem} is given mutually inductively with
+Definition~\ref{def:set-sem} we can now define our relational
+interpretations.
 
@@
   \end{align*}
+
+\vspace*{-0.5in}
+
 \begin{align*}
@@
 that $\relsem{\Gamma \vdash F \to G}$ and $\relsem{\Gamma \vdash
-  \forall \ol \alpha. F}$ are as expected in any parametric model.
-%$\relsem{\Gamma \vdash
-%  \sigma \to \tau}\rho = \relsem{\Gamma \vdash \sigma} \rho \to
-%\relsem{\Gamma \vdash \tau} \rho$.
-As for set interpretations, $\relsem{\Gamma; \emptyset \vdash
-  \Nat^{\ol\alpha} F\,G}$ is $\omega$-cocontinuous because it is a constant
-functor.
-%follows from $\omega$-cocontinuity of
-%$\relsem{\Gamma;\ol\alpha \vdash F}$ and
-%$\relsem{\Gamma;\ol\alpha \vdash G}$.
-If $\rho \in \relenv$
-and $\vdash F$, then we write $\relsem{\vdash F}$ instead of
-$\relsem{\vdash F}\rho$.  For the last clause in
-Definition~\ref{def:rel-sem} to be well-defined we need $T_{H,\rho}$
-to be an $\omega$-cocontinuous endofunctor on $RT$ so that, by
-Lemma~\ref{lem:fp}, it admits a fixpoint. Since $T_{H,\rho}$ is
-defined in terms of $\relsem{\Gamma;\ol\gamma,\phi^k, \ol{\alpha}
-  \vdash H}$, this means that relational interpretations of types must
-be $\omega$-cocontinuous functors from $\relenv$ to $RT_0$, which in
-turn entails that the actions of relational interpretations of types
-on objects and on morphisms in $\relenv$ are intertwined. As for set
-interpretations, we know from~\cite{jp19} that, for every $\Gamma;
-\ol{\alpha} \vdash F$, $\relsem{\Gamma; \ol{\alpha} \vdash F}$ is
-actually in $[\rel^k,\rel]$ where $k = |\ol \alpha|$.
-%In fact, we already know from~\cite{jp19} that, for every $\Gamma;
-%\ol{\alpha} \vdash \tau : \F$, $\relsem{\Gamma; \ol{\alpha} \vdash
-%  \tau}$ is actually functorial in $\ol{\alpha}$ and
-%$\omega$-cocontinuous.
-We first define the actions of each of these functors on morphisms
-between environments in Definition~\ref{def:rel-sem-funcs}, and then
-argue that the functors given by Definitions~\ref{def:rel-sem}
-and~\ref{def:rel-sem-funcs} are well-defined and have the required
-properties. To do this, we extend $T_H$ to a {\em functor} from
-$\relenv$ to $[[\rel^k,\rel],[\rel^k,\rel]]$. Its action on an object
-$\rho \in \relenv$ is given by the higher-order functor $T_{H,\rho}$
-whose actions on objects and morphisms are given in
+  \forall \ol \alpha. F}$ are as expected in any parametric model.  As
+for set interpretations, $\relsem{\Gamma; \emptyset \vdash
+  \Nat^{\ol\alpha} F\,G}$ is $\omega$-cocontinuous because it is
+constant on $\omega$-directed sets. If $\rho \in \relenv$ and $\vdash
+F$, then we write $\relsem{\vdash F}$ instead of $\relsem{\vdash
+  F}\rho$.  For the last clause in Definition~\ref{def:rel-sem} to be
+well-defined we need $T_{H,\rho}$ to be an $\omega$-cocontinuous
+endofunctor on $RT$ so that, by Lemma~\ref{lem:fp}, it admits a
+fixpoint. Since $T_{H,\rho}$ is defined in terms of
+$\relsem{\Gamma;\ol\gamma,\phi^k, \ol{\alpha} \vdash H}$, this means
+that relational interpretations of types must be $\omega$-cocontinuous
+functors from $\relenv$ to $RT_0$, which in turn entails that the
+actions of relational interpretations of types on objects and on
+morphisms in $\relenv$ are intertwined. As for set interpretations, we
+know from~\cite{jp19} that, for every $\Gamma; \ol{\alpha} \vdash F$,
+$\relsem{\Gamma; \ol{\alpha} \vdash F}$ is actually in $[\rel^k,\rel]$
+where $k = |\ol \alpha|$. We first define the actions of each of these
+functors on morphisms between environments, and then argue that they
+are well-defined and have the required properties. To do this, we
+extend $T_H$ to a {\em functor} from $\relenv$ to
+$[[\rel^k,\rel],[\rel^k,\rel]]$. Its action on an object $\rho \in
+\relenv$ is given by the higher-order functor $T_{H,\rho}$ whose
+actions on objects and morphisms are given in
 Definition~\ref{def:rel-sem-funcs}. Its action on a morphism $f : \rho
@@
 F)_{\ol{R}} = \relsem{\Gamma; \ol\gamma,\phi,\ol{\alpha} \vdash
-  H}f[\phi := \id_F][\ol{\alpha := \id_R}]$. The next definition uses
-$T_H$ to define the actions of functors interpreting types on
-morphisms between relation environments.
-
+  H}f[\phi := \id_F][\ol{\alpha := \id_R}]$.
+
+Using $T_H$, we can define the functorial action of relational
+interpretation.
 \begin{dfn}\label{def:rel-sem-funcs}
@@
 
-To see that the functors given by Definitions~\ref{def:rel-sem}
-and~\ref{def:rel-sem-funcs} are well-defined we must show that, for
-every $H$, $T_{H,\rho}\,F$ is a relation transformer for any relation
+For the functors given in Definitions~\ref{def:rel-sem}
+and~\ref{def:rel-sem-funcs} to be well-defined we need that, for every
+$H$, $T_{H,\rho}\,F$ is a relation transformer for any relation
 transformer $F$, and that $T_{H,f}\, F : T_{H,\rho}\, F \to
@@
 \begin{lemma}\label{lem:rel-transf-morph}
-%The interpretations in Definitions~\ref{def:rel-sem}
-%and~\ref{def:rel-sem-funcs} are well-defined and,
-For every $\Gamma;\Phi \vdash F$, $\sem{\Gamma;\Phi \vdash F} =
-(\setsem{\Gamma;\Phi \vdash F}, \setsem{\Gamma;\Phi \vdash
-  F},\relsem{\Gamma;\Phi \vdash F}) \in
-%is an $\omega$-cocontinuous functor from $\relenv$ to $RT_0$, i.e., is an
-%element of
-[\relenv,RT_0]$.
+For every $\Gamma;\Phi \vdash F$,
+\[\sem{\Gamma;\Phi \vdash F} = (\setsem{\Gamma;\Phi \vdash F},
+\setsem{\Gamma;\Phi \vdash F},\relsem{\Gamma;\Phi \vdash F})
+\in[\relenv,RT_0]\] 
 \end{lemma}
@@
 The proof is a straightforward induction on the structure of $F$,
-%%each case of which first shows that $(\setsem{\Gamma;\Phi \vdash
-%  \tau}, \setsem{\Gamma;\Phi \vdash \tau},\relsem{\Gamma;\Phi \vdash
-%  \tau})$ is indeed a functor from $\relenv$ to $RT_0$, and then uses
 using an appropriate result from~\cite{jp19} to deduce
 $\omega$-cocontinuity of $\sem{\Gamma;\Phi \vdash F}$ in each case,
-together with Lemma~\ref{lem:fp} and Equation~\ref{eq:mu} for
-$\mu$-types.
-
-\begin{comment}
-\begin{proof}
-By induction on the structure of $\tau$. The only interesting cases
-are when $\tau = \phi\ol{\tau}$ and when $\tau = (\mu \phi^k. \lambda
-\overline{\alpha}. H)\overline{\tau}$. We consider each in turn.
-
-\begin{itemize}
-\item When $\tau = \Gamma; \Phi \vdash \phi\ol{\tau}$, we have    
-\[\begin{array}{ll}
-  & \pi_i (\relsem{\Gamma; \Phi \vdash \phi\ol{\tau}}\rho)\\
-= & \pi_i ((\rho \phi)\overline{\relsem{\Gamma; \Phi \vdash \tau}\rho})\\
-= & (\pi_i (\rho \phi)) (\pi_i(\overline{\relsem{\Gamma; \Phi \vdash \tau}\rho}))\\
-= & ((\pi_i \rho) \phi) (\overline{\setsem{\Gamma; \Phi \vdash
-    \tau}(\pi_i \rho)})\\
-= & \setsem{\Gamma; \Phi \vdash \phi\ol{\tau}}(\pi_i \rho)
-\end{array}\]
-and, for $f : \rho \to \rho'$ in $\relenv$,
-\[\begin{array}{ll}
-  & \pi_i (\relsem{\Gamma; \Phi \vdash \phi\ol{\tau}}f)\\
-= & \pi_i ((f\phi)_{\overline{\relsem{\Gamma; \Phi \vdash \tau}\rho'}})
-\circ \pi_i((\rho\phi)({\overline{\relsem{\Gamma; \Phi \vdash \tau}f}}))\\
-= & (\pi_i (f\phi))_{\overline{\pi_i (\relsem{\Gamma; \Phi \vdash \tau}\rho')}}
-\circ (\pi_i(\rho\phi))({\overline{\pi_i (\relsem{\Gamma; \Phi \vdash \tau}f}}))\\
-= & ((\pi_i f)\phi)_{\overline{\setsem{\Gamma; \Phi \vdash \tau}(\pi_i\rho')}}
-\circ ((\pi_i\rho)\phi)({\overline{\setsem{\Gamma; \Phi \vdash \tau}(\pi_i f)}})\\
-= & \setsem{\Gamma; \Phi \vdash \phi\ol{\tau}}(\pi_i f)
-\end{array}\]
-The third equalities of each of the above derivations are by the
-induction hypothesis. That $\sem{\Gamma; \Phi \vdash
-  \phi\ol{\tau}}$ is $\omega$-cocontinuous is an immediate
-consequence of the facts that $\set$ and $\rel$ are locally finitely
-presentable, together with Corollary~12 of~\cite{jp19}.
-\item When $\tau = (\mu \phi. \lambda
-  \overline{\alpha}. H)\overline{\tau}$ we first show that $\sem{ (\mu
-    \phi. \lambda \overline{\alpha}. H)\overline{\tau}}$ is
-  well-defined.
-\begin{itemize}
-\item \underline{$T_\rho : [RT_k,RT_k]$}:\/ We must show that, for any
-  relation transformer $F = 
-  (F^1, F^2, F^*)$, the triple $T_{\rho} F = (T^\set_{\pi_1 \rho}F^1,
-  T^\set_{\pi_2 \rho}F^2, T^\rel_{\rho}F)$ is also a relation
-  transformer.  Let $\overline{R : \rel(A, B)}$. Then for
-  $i = 1, 2$, we have
-\[\begin{split}
-\pi_i(T^\rel_{\rho}\,F\,\overline{R})
-&= \pi_i(\relsem{\Gamma;\Phi,\phi,\overline{\alpha} \vdash H}\rho[\phi := F]\overline{[\alpha := R]}) \\
-&= \setsem{\Gamma;\Phi,\phi,\overline{\alpha} \vdash H} (\pi_i (\rho[\phi := F]\overline{[\alpha := R]})) \\
-&= \setsem{\Gamma;\Phi,\phi,\overline{\alpha} \vdash H} (\pi_i \rho)[\phi := \pi_i F]\overline{[\alpha := \pi_i R]}) \\
-&= T^\set_{\pi_i \rho} (\pi_i F) (\overline{\pi_i R})
-\end{split}\]
-and 
-\[\begin{split}
-\pi_i(T^\rel_{\rho}\,F\,\overline{\gamma})
-&= \pi_i(\relsem{\Gamma;\Phi,\phi,\overline{\alpha} \vdash H}\id_\rho[\phi
-  := \id_F]\overline{[\alpha := \gamma]}) \\
-&= \setsem{\Gamma;\Phi,\phi,\overline{\alpha} \vdash H} (\pi_i (\id_\rho[\phi := \id_F]\overline{[\alpha := \gamma]})) \\
-&= \setsem{\Gamma;\Phi,\phi,\overline{\alpha} \vdash H} \id_{\pi_i \rho}[\phi := \id_{\pi_i F}]\overline{[\alpha := \pi_i \gamma]} \\
-&= T^\set_{\pi_i \rho} (\pi_i F) (\overline{\pi_i \gamma})
-\end{split}\]
-Here, the second equality in each of the above chains of equalities is
-by the induction hypothesis.
-
-We also have that, for every morphism $\delta = (\delta^1, \delta^2) :
-F \to G$ in $RT_k$ and all $\overline{R : \rel(A, B)}$,
-\[\begin{array}{ll}
-  & \pi_i((T^\rel_\rho \delta)_{\overline{R}})\\
-= & \pi_i(\relsem{\Gamma;\Phi,\phi,\overline{\alpha} \vdash H}\id_\rho[\phi :=
-  \delta]\overline{[\alpha := \id_R]})\\
-= & \setsem{\Gamma;\Phi,\phi,\overline{\alpha} \vdash H}\id_{\pi_i\rho}[\phi :=
-  \pi_i \delta]\overline{[\alpha := \id_{\pi_i R}]}\\
-= & (T^\set_{\pi_i \rho} (\pi_i \delta))_{\overline{\pi_i R}}
-\end{array}\]
-
-Here, the second equality is by the induction hypothesis.  That
-$T_\rho$ is $\omega$-cocontinuous follows immediately from the
-induction hypothesis on $\sem{\Gamma;\Phi,\phi,\ol{\alpha} \vdash H}$
-and the fact that colimts are computed componentwise in $RT$.
-\item \underline{$\sigma_f = (\sigma^\set_{\pi_1 f},
-  \sigma^\set_{\pi_2 f})$ is a natural transformation from $T_{\rho}$
-  to $T_{\rho'}$}:\/ We must show that $(\sigma_f)_F =
-  ((\sigma^\set_{\pi_1 f})_{F^1}, (\sigma^\set_{\pi_2 f})_{F^2})$ is a
-  morphism in $RT_k$ for all relation transformers $F = (F^1, F^2,
-  F^*)$, i.e., that $((\sigma_f)_F)_{\overline{R}}
-  =(((\sigma^\set_{\pi_1 f})_{F^1})_{\overline{A}},
-  ((\sigma^\set_{\pi_2 f})_{F_2})_{\overline{B}})$ is a morphism in
-  $\rel$ for all relations $\overline{R : \rel(A, B)}$. Indeed, we
-  have that
-\[
-((\sigma_f)_F)_{\overline{R}} =
-\relsem{\Gamma;\Phi,\phi,\overline{\alpha} \vdash H}f[\phi :=
-  \id_F]\overline{[\alpha := \id_R]}
-\]
-is a morphism in $RT_0$ (and thus in $\rel$) by the induction hypothesis.
-\end{itemize}
-
-\vspace*{0.05in}
-
-The relation transformer $\mu T_\rho$ is therefore a fixpoint of
-$T_\rho$ by Lemma~\ref{lem:fp}, and $\mu \sigma_f$ is a morphism in
-$RT_k$ from $\mu T_\rho$ to $\mu T_{\rho'}$. ($\mu$ is shown to be a
-functor in~\cite{jp19}.)  So $\relsem{\Gamma; \Phi \vdash (\mu
-  \phi. \lambda \overline{\alpha}. H)\overline{\tau}}$, and thus
-$\sem{\Gamma; \Phi \vdash (\mu \phi. \lambda
-  \overline{\alpha}. H)\overline{\tau}}$, is well-defined.
-
-\vspace*{0.1in}
-
-To see that $\sem{\Gamma; \Phi \vdash (\mu \phi. \lambda
-  \overline{\alpha}. H)\overline{\tau}} : [\relenv,RT_0]$, we must
-verify three conditions:
-\begin{itemize}
-\item Condition (1) after Definition~\ref{def:relenv-functor} is
-  satisfied since
-\[\begin{split}
-\pi_i(\relsem{\Gamma;\Phi \vdash (\mu \phi. \lambda
-  \overline{\alpha}. H) \overline{\tau}}\rho) 
-&= \pi_i( (\mu T_\rho) (\overline{\relsem{\Gamma;\Phi \vdash
-    \tau}\rho})) \\ 
-&= \pi_i(\mu T_{\rho}) (\overline{\pi_i(\relsem{\Gamma;\Phi \vdash
-    \tau}\rho})) \\ 
-&= \mu T^\set_{\pi_i\rho} (\overline{\setsem{\Gamma;\Phi \vdash
-    \tau}(\pi_i\rho)}) \\ 
-&= \setsem{\Gamma;\Phi \vdash (\mu \phi. \lambda \overline\alpha. H)
-  \overline{\tau}}(\pi_i\rho) 
-\end{split}\]
-The third equality is by Equation ~\ref{eq:mu} and the induction
-hypothesis.
-\item Condition (2) after Definition~\ref{def:relenv-functor} is
-  satisfied since it is subsumed by the previous condition because $k
-  = 0$.
-\item The third bullet point of Definition~\ref{def:relenv-functor} is
-  satisfied because
-\[\begin{split}
-& \pi_i(\relsem{\Gamma;\Phi \vdash (\mu \phi. \lambda
-  \overline\alpha. H)\overline{\tau}}f)\\
-&= \pi_i((\mu T_{\rho'})(\overline{\relsem{\Gamma;\Phi \vdash
-    \tau}f}) \circ (\mu \sigma_f)_{\overline{\relsem{\Gamma;\Phi
-      \vdash \tau}\rho}}) \\ 
-&= \pi_i((\mu T_{\rho'})(\overline{\relsem{\Gamma;\Phi \vdash
-    \tau}f})) \circ \pi_i((\mu
-\sigma_f)_{\overline{\relsem{\Gamma;\Phi \vdash \tau}\rho}}) \\  
-&= \pi_i(\mu T_{\rho'})(\overline{\pi_i(\relsem{\Gamma;\Phi
-    \vdash \tau}f})) \circ \pi_i(\mu
-\sigma_f)_{\overline{\pi_i(\relsem{\Gamma;\Phi \vdash 
-      \tau}\rho})} \\ 
-&= (\mu T^\set_{\pi_i \rho'})(\overline{\setsem{\Gamma;\Phi \vdash
-    \tau}(\pi_i f)}) \circ (\mu \sigma^\set_{\pi_i
-  f})_{\overline{\setsem{\Gamma;\Phi \vdash \tau}(\pi_i\rho)}} \\ 
-&= \setsem{\Gamma;\Phi \vdash (\mu \phi. \lambda
-  \overline\alpha. H)\overline{\tau}}(\pi_i f). 
-\end{split}\]
-The fourth equality is by~\ref{eq:mu} and the induction hypothesis.
-\end{itemize}
-As before, that $\sem{\Gamma; \Phi \vdash (\mu \phi. \lambda
-  \overline{\alpha}. H)\overline{\tau}}$ is $\omega$-cocontinuous
-follows from the facts that $\set$ and $\rel$ are locally finitely
-presentable, and that colimits in $\relenv$ are computed
-componentwise, together with Corollary~12 of~\cite{jp19}.
-\end{itemize}
-\end{proof}
-\end{comment}
-
-%We need
-%$\relsem{\Gamma; \Phi, \phi \vdash \tau}\rho$ $=
-%\relsem{\Gamma, \psi; \Phi \vdash \tau[\phi :==
-%    \psi]}\rho[\psi:=\rho\phi]$, and that $\relsem{\Gamma \vdash
-%  \sigma \to \tau}$ and
-%$\setsem{\Gamma; \Phi, \phi \vdash \tau}\rho$ $=
-%\setsem{\Gamma, \psi; \Phi \vdash \tau[\phi :==
-%    \psi]}\rho[\psi:=\rho\phi]$. 
-We can also prove by simultaneous induction that our interpretations
-of types interact well with demotion of functorial variables. Indeed,
-we have that, if $\rho, \rho' : \setenv$, \,$f : \rho \to \rho'$,
-\,$\rho \phi = \rho \psi = \rho' \phi = \rho' \psi$, \, $f \phi = f
-\psi = \id_{\rho \phi}$,\, $\Gamma; \Phi, \phi^k \vdash F$,\,
+{\color{red} KEEP? together with Lemma~\ref{lem:fp} and
+  Equation~\ref{eq:mu} for $\mu$-types.}
+
+We can prove by simultaneous induction that our interpretations of
+types interact well with demotion of functorial variables to
+non-functorial ones, along with other useful identities. Indeed, if
+$\rho, \rho' : \setenv$, \,$f : \rho \to \rho'$, \,$\rho \phi = \rho
+\psi = \rho' \phi = \rho' \psi$, \, $f \phi = f \psi = \id_{\rho
+  \phi}$,\, $\Gamma; \Phi, \phi^k \vdash F$,\,
 $\Gamma;\Phi,\ol{\alpha} \vdash G$,\, $\Gamma;\Phi,\alpha_1...\alpha_k
 \vdash H$, and $\ol{\Gamma;\Phi \vdash K}$, then
-%for any $\rho,\rho'$ and $f: \rho \to \rho'$ in
-%$\setenv$, 
 \begin{gather}
@@
 (\ref{eq:subs-const-morph}) hold for relational interpretations as well.
-%\end{lemma}
-
-\begin{comment}
-\begin{lemma}\label{lem:substitution}
-Let $\rho$ and $\rho'$ be set environments and let $f : \rho \to \rho'$ be a
-morphism of set environments. 
-\begin{itemize}
-\item If $\Gamma;\Phi,\ol{\alpha} \vdash F$ and $\Gamma;\Phi \vdash \tau$,
-  then  
-\begin{gather}\label{eq:subs-var}
-\setsem{\Gamma;\Phi \vdash F[\ol{\alpha := \tau}]}\rho =
-\setsem{\Gamma;\Phi,\ol{\alpha} \vdash F}\rho[\ol{\alpha :=
-\setsem{\Gamma;\Phi \vdash \tau}\rho}]\\
-\intertext{and}\label{eq:subs-var-morph}
-\setsem{\Gamma;\Phi \vdash F[\ol{\alpha := \tau}]}f =
-\setsem{\Gamma;\Phi,\ol{\alpha} \vdash F}f[\ol{\alpha :=
-\setsem{\Gamma;\Phi \vdash \tau}f}]
-\end{gather}
-\item If $\Gamma;\Phi,\phi^k\vdash F$ and
-  $\Gamma;\Phi,\alpha_1...\alpha_k \vdash H$, then 
-\begin{gather}\label{eq:subs-const}
-\setsem{\Gamma; \Phi \vdash F[\phi := H]}\rho
-= \setsem{\Gamma; \Phi, \phi \vdash F}\rho
-[\phi := \lambda \ol{A}.\, \setsem{\Gamma;\Phi,\overline{\alpha}\vdash
-    H}\rho[\overline{\alpha := A}]] \\ 
-\intertext{and}\label{eq:subs-const-morph}
-\setsem{\Gamma; \Phi \vdash F[\phi := H]}f
-= \setsem{\Gamma; \Phi, \phi \vdash F}f
-[\phi := \lambda \ol{A}.\,\setsem{\Gamma;\Phi,\overline{\alpha}\vdash
-    H}f[\overline{\alpha := \id_{\ol{A}}}]] 
-\end{gather}
-\end{itemize}
-Analogous identities hold for relation environments and morphisms
-between them.
-\end{lemma}
-
-\end{comment}
-
-\begin{comment}
-
-\begin{thm}\label{thm:demotion-objects}
-Let \, $\Gamma; \Phi, \phi \vdash \tau : \F$. If $\rho, \rho' : \setenv$
-are such that $\rho \phi = \rho \psi = \rho' \phi = \rho' \psi$, and
-if $f : \rho \to \rho'$ is a morphism of set environments such that $f
-\phi = f \psi = \id_{\rho \phi}$, then
-\[\setsem{\Gamma; \Phi, \phi \vdash \tau} \rho = \setsem{\Gamma, \psi;
-  \Phi \vdash \tau[\phi :== \psi] } \rho\]
-and
-\[\setsem{\Gamma; \Phi, \phi \vdash \tau} f = \setsem{\Gamma, \psi;
-  \Phi \vdash \tau[\phi :== \psi]} f\]
-
-\vspace*{0.1in}
-
-\noindent
-Analogously, if $\rho, \rho' : \relenv$ are such that $\rho \phi =
-\rho \psi = \rho' \phi = \rho' \psi$, and if $f : \rho \to \rho'$ is a
-morphism of relation environments such that $f \phi = f \psi =
-\id_{\rho \phi}$, then
-\[\relsem{\Gamma; \Phi, \phi \vdash \tau} \rho = \relsem{\Gamma, \psi;
-  \Phi \vdash \tau[\phi :== \psi] } \rho\]
-and
-\[\relsem{\Gamma; \Phi, \phi \vdash \tau} f = \relsem{\Gamma, \psi;
-  \Phi \vdash \tau[\phi :== \psi]} f\]
-\end{thm}
-
-\end{comment}
-
-\begin{comment}
----
-The next two theorems are proven by simultaneous induction. We are
-actually only interested in using Theorem~\ref{thm:demotion-objects},
-but in order to prove the $\mu$ case for this theorem, we need
-Theorem~\ref{thm:demotion-morph} to show that two functors have equal
-actions on morphisms.
-
-\begin{thm}\label{thm:demotion-objects}
-Let $\Gamma; \Phi, \phi \vdash \tau : \F$. If $\rho, \rho' : \setenv$
-are such that $\rho \phi = \rho \psi = \rho' \phi = \rho' \psi$, and
-if $f : \rho \to \rho'$ is a morphism of set environments such that $f
-\phi = f \psi = \id_{\rho \phi}$, then
-\[\setsem{\Gamma; \Phi, \phi \vdash \tau} \rho = \setsem{\Gamma, \psi;
-  \Phi \vdash \tau[\phi :== \psi] } \rho\]
-and
-\[\setsem{\Gamma; \Phi, \phi \vdash \tau} f = \setsem{\Gamma, \psi;
-  \Phi \vdash \tau[\phi :== \psi]} f\]
-
-\vspace*{0.1in}
-
-\noindent
-Analogously, if $\rho, \rho' : \relenv$ are such that $\rho \phi =
-\rho \psi = \rho' \phi = \rho' \psi$, and if $f : \rho \to \rho'$ is a
-morphism of relation environments such that $f \phi = f \psi =
-\id_{\rho \phi}$, then
-\[\relsem{\Gamma; \Phi, \phi \vdash \tau} \rho = \relsem{\Gamma, \psi;
-  \Phi \vdash \tau[\phi :== \psi] } \rho\]
-and
-\[\relsem{\Gamma; \Phi, \phi \vdash \tau} f = \relsem{\Gamma, \psi;
-  \Phi \vdash \tau[\phi :== \psi]} f\]
-\end{thm}
-\begin{proof}
-We prove the result for set interpretations by induction on the
-structure of $\tau$.  The case for relational interpretations proceeds
-analogously. Since the only interesting cases are the application case
-and the $\mu$-case, we elide the others.
-\begin{itemize}
-\item If $\Gamma; \Phi, \phi \vdash \phi \ol\tau : \F$, then the
-  induction hypothesis gives that  
-\[\setsem{\Gamma; \Phi, \phi \vdash \tau } \rho = \setsem{\Gamma,
-  \psi; \Phi \vdash \tau[\phi :== \psi] } \rho\] 
-and
-\[\setsem{\Gamma; \Phi, \phi \vdash \tau} f = \setsem{\Gamma, \psi;
-  \Phi \vdash \tau[\phi :== \psi]} f\]
-for each $\tau$. Then
-\begin{align*}
- & \setsem{\Gamma; \Phi, \phi \vdash \phi \ol\tau } \rho \\
-= \; & (\rho \phi) \ol{\setsem{\Gamma; \Phi, \phi \vdash \tau } \rho} \\
-= \; & (\rho \phi) \ol{\setsem{\Gamma, \psi; \Phi \vdash \tau[\phi :==
-      \psi] } \rho} \\ 
-= \; & (\rho \psi) \ol{\setsem{\Gamma, \psi; \Phi \vdash \tau[\phi :==
-      \psi] } \rho} \\ 
-= \; & \setsem{\Gamma, \psi; \Phi \vdash \psi \ol{\tau[\phi :== \psi]}
-} \rho \\ 
-= \; & \setsem{\Gamma, \psi; \Phi \vdash (\phi \ol\tau)[\phi :== \psi]
-} \rho 
-\end{align*}
-Here, the first and fifth equalities are by
-Definition~\ref{def:set-sem}, and the fourth equality is by equality
-of the functors $\rho \phi$ and $\rho \psi$. We also have that
-\begin{align*}
-& \setsem{\Gamma; \Phi, \phi \vdash \phi \ol\tau} f \\
-= \; & (f \phi)_{\ol{\setsem{\Gamma; \Phi, \phi \vdash \tau} \rho'}} 
-    \circ (\rho \phi) \ol{\setsem{\Gamma; \Phi, \phi \vdash \tau} f} \\
-= \; & (\id_{\rho \phi})_{\ol{\setsem{\Gamma; \Phi, \phi \vdash \tau} \rho'}} 
-    \circ (\rho \phi) \ol{\setsem{\Gamma; \Phi, \phi \vdash \tau} f} \\
-= \; & (\rho \phi) \ol{\setsem{\Gamma; \Phi, \phi \vdash \tau} f} \\
-= \; & (\rho \psi) \ol{\setsem{\Gamma, \psi; \Phi \vdash \tau[\phi :==
-      \psi]} f} \\ 
-= \; & (\id_{\rho \psi})_{\ol{\setsem{\Gamma, \psi; \Phi \vdash
-      \tau[\phi :== \psi]} \rho'}} \circ (\rho \psi)
-\ol{\setsem{\Gamma, \psi; \Phi \vdash \tau[\phi :== \psi]} f} \\ 
-= \; & (f \psi)_{\ol{\setsem{\Gamma, \psi; \Phi \vdash \tau[\phi :==
-        \psi]} \rho'}} \circ (\rho \psi) \ol{\setsem{\Gamma, \psi;
-    \Phi \vdash \tau[\phi :== \psi]} f} \\ 
-= \; & \setsem{\Gamma, \psi; \Phi \vdash \psi \ol{\tau[\phi :== \psi]}} f \\
-= \; & \setsem{\Gamma, \psi; \Phi \vdash (\phi \ol\tau) [\phi :== \psi]} f
-\end{align*}
-\item If $\Gamma; \Phi, \phi \vdash (\mu \phi'. \lambda
-  \ol\alpha. H)\ol\tau : \F$, then the induction hypothesis gives that
-\[\setsem{\Gamma; \Phi, \phi', \ol{\alpha}, \phi \vdash H } \rho =
-\setsem{\Gamma, \psi; \Phi, \phi', \ol{\alpha} \vdash H[\phi :== \psi]
-} \rho\]
-and 
-\[\setsem{\Gamma; \Phi, \phi \vdash \tau } \rho = \setsem{\Gamma,
-  \psi; \Phi \vdash \tau[\phi :== \psi] } \rho\]
-as well as that
-\[\setsem{\Gamma; \Phi, \phi', \ol{\alpha}, \phi \vdash H } f =
-\setsem{\Gamma, \psi; \Phi, \phi', \ol{\alpha} \vdash H[\phi :== \psi]
-} f\]
-and 
-\[\setsem{\Gamma; \Phi, \phi \vdash \tau } f = \setsem{\Gamma, \psi;
-  \Phi \vdash \tau[\phi :== \psi] } f\]
-for each $\tau$. Then
-\begin{align*}
-& \setsem{\Gamma; \Phi, \phi \vdash (\mu \phi'. \lambda
-    \ol\alpha. H)\ol\tau } \rho \\ 
-= \; & (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma; \Phi, \phi',
-  \ol{\alpha}, \phi \vdash H} \rho[\phi' := F][\ol{\alpha := A}]))
-\ol{\setsem{ \Gamma; \Phi, \phi \vdash \tau} \rho} \\ 
-= \; & (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma, \psi; \Phi,
-  \phi', \ol{\alpha} \vdash H[\phi :== \psi]} \rho[\phi' :=
-  F][\ol{\alpha := A}])) \ol{\setsem{ \Gamma; \Phi, \phi \vdash \tau}
-  \rho} \\ 
-= \; &  (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma, \psi; \Phi,
-  \phi', \ol{\alpha} \vdash H[\phi :== \psi]} \rho[\phi' :=
-  F][\ol{\alpha := A}])\ol{\setsem{ \Gamma, \psi; \Phi \vdash
-    \tau[\phi :== \psi]} \rho} \\ 
-= \; & \setsem{\Gamma, \psi; \Phi \vdash (\mu \phi'. \lambda
-  \ol\alpha. H[\phi :== \psi]) \ol{\tau[\phi :== \psi]} } \rho \\ 
-= \; & \setsem{\Gamma, \psi; \Phi \vdash ((\mu \phi'. \lambda
-  \ol\alpha. H) \ol\tau)[\phi :== \psi] } \rho   
-\end{align*} 
-The first and fifth equalities are by Definition~\ref{def:set-sem}.
-The second equality follows from the following equality:
-\begin{align*}
-&\lambda F. \lambda \ol{A}. \setsem{\Gamma; \Phi, \phi', \ol{\alpha},
-    \phi \vdash H} \rho[\phi' := F][\ol{\alpha := A}] \\ 
-& = \lambda F. \lambda \ol{A}. \setsem{\Gamma, \psi; \Phi, \phi',
-    \ol{\alpha} \vdash H[\phi :== \psi]} \rho[\phi' := F][\ol{\alpha
-      := A}] 
-\end{align*}
-These two maps have the same actions on objects and morphisms by the
-induction hypothesis on $H$, and the fact that the extended
-environment $\rho[\phi' := F][\ol{\alpha := A}]$ satisfies the
-required hypothesis. They are thus equal as functors and so have the
-same fixpoint. We also have that 
-\begin{align*}
-& \setsem{\Gamma; \Phi, \phi \vdash (\mu \phi'. \lambda
-    \ol\alpha. H)\ol\tau } f \\ 
-= \; & (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma; \Phi, \phi',
-  \ol{\alpha}, \phi \vdash H} f [\phi' := \id_F][\ol{\alpha :=
-    \id_A}]))_{\ol{\setsem{\Gamma; \Phi, \phi \vdash \tau} \rho'}} \\ 
-\; & \circ (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma; \Phi,
-  \phi', \ol{\alpha}, \phi \vdash H} \rho[\phi' := F][\ol{\alpha :=
-    A}])) \ol{\setsem{ \Gamma; \Phi, \phi \vdash \tau} f} \\ 
-= \; & (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma; \Phi, \phi',
-  \ol{\alpha}, \phi \vdash H} f [\phi' := \id_F][\ol{\alpha :=
-    \id_A}]))_{\ol{\setsem{\Gamma, \psi; \Phi \vdash \tau[\phi :==
-        \psi]} \rho'}} \\ 
-\; & \circ (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma; \Phi,
-  \phi', \ol{\alpha}, \phi \vdash H} \rho[\phi' := F][\ol{\alpha :=
-    A}])) \ol{\setsem{ \Gamma, \psi; \Phi \vdash \tau[\phi :== \psi]}
-  f} \\ 
-= \; & (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma, \psi; \Phi,
-  \phi', \ol{\alpha} \vdash H[\phi :== \psi]} f [\phi' :=
-  \id_F][\ol{\alpha := \id_A}]))_{\ol{\setsem{\Gamma, \psi; \Phi
-      \vdash \tau[\phi :== \psi]} \rho'}} \\ 
-\; & \circ (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma; \Phi,
-  \phi', \ol{\alpha}, \phi \vdash H} \rho[\phi' := F][\ol{\alpha :=
-    A}])) \ol{\setsem{ \Gamma, \psi; \Phi \vdash \tau[\phi :== \psi]}
-  f} \\ 
-= \; & (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma, \psi; \Phi,
-  \phi', \ol{\alpha} \vdash H[\phi :== \psi]} f [\phi' :=
-  \id_F][\ol{\alpha := \id_A}]))_{\ol{\setsem{\Gamma, \psi; \Phi
-      \vdash \tau[\phi :== \psi]} \rho'}} \\ 
-\; & \circ (\mu (\lambda F. \lambda \ol{A}. \setsem{\Gamma, \psi;
-  \Phi, \phi', \ol{\alpha} \vdash H[\phi :== \psi]} \rho[\phi' :=
-  F][\ol{\alpha := A}])) \ol{\setsem{ \Gamma, \psi; \Phi \vdash
-    \tau[\phi :== \psi]} f} \\ 
-= \; & \setsem{\Gamma, \psi; \Phi \vdash (\mu \phi'. \lambda
-  \ol\alpha. H[\phi :== \psi]) \ol{\tau[\phi :== \psi]} } f \\ 
-= \; & \setsem{\Gamma, \psi; \Phi \vdash ((\mu \phi'. \lambda
-  \ol\alpha. H) \ol\tau)[\phi :== \psi] } f 
-\end{align*} 
-The first and fifth equalities are by Definition~\ref{def:set-sem}.
-The third equality is by the equality of the arguments to the first
-$\mu$ operator:
-\begin{align*}
-& \lambda F. \lambda \ol{A}. \setsem{\Gamma; \Phi, \phi', \ol{\alpha},
-    \phi \vdash H} f [\phi' := \id_F][\ol{\alpha := \id_A}] \\ 
-& = \lambda F. \lambda \ol{A}. \setsem{\Gamma, \psi; \Phi, \phi',
-    \ol{\alpha} \vdash H[\phi :== \psi]} f [\phi' := \id_F][\ol{\alpha
-      := \id_A}] 
-\end{align*}
-By the induction hypothesis on $H$ and the fact that the morphism
-$f[\phi' := \id_F][\ol{\alpha := \id_A}] : \rho[\phi' := F][\ol{\alpha
-    := A}] \to \rho'[\phi' := F][\ol{\alpha := A}]$ still satisfies
-the required hypotheses.  The fourth equality is by the equality of
-the arguments to the second $\mu$ operator:
-\begin{align*}
-& \lambda F. \lambda \ol{A}. \setsem{\Gamma; \Phi, \phi', \ol{\alpha},
-    \phi \vdash H} \rho[\phi' := F][\ol{\alpha := A}] \\ 
-& = \lambda F. \lambda \ol{A}. \setsem{\Gamma, \psi; \Phi, \phi',
-    \ol{\alpha} \vdash H[\phi :== \psi]} \rho[\phi' := F][\ol{\alpha
-      := A}] 
-\end{align*}
-By the same reasoning as above, these two maps are equal as functors,
-and thus have the same fixpoint.
-\end{itemize}
-\end{proof}
-
----
-
-The following lemma ensures that substitution interacts well with type
-interpretations.  It is a consequence of
-Definitions~\ref{def:second-order-subst},~\ref{def:set-interp},
-and~\ref{def:rel-interp}.
-
-\begin{lemma}\label{lem:substitution}
-Let $\rho$ and $\rho'$ be set environments and let $f : \rho \to \rho'$ be a
-morphism of set environments. 
-\begin{itemize}
-\item If $\Gamma;\Phi,\ol{\alpha} \vdash F$ and $\Gamma;\Phi \vdash \tau$,
-  then  
-\begin{gather}\label{eq:subs-var}
-\setsem{\Gamma;\Phi \vdash F[\ol{\alpha := \tau}]}\rho =
-\setsem{\Gamma;\Phi,\ol{\alpha} \vdash F}\rho[\ol{\alpha :=
-\setsem{\Gamma;\Phi \vdash \tau}\rho}]\\
-\intertext{and}\label{eq:subs-var-morph}
-\setsem{\Gamma;\Phi \vdash F[\ol{\alpha := \tau}]}f =
-\setsem{\Gamma;\Phi,\ol{\alpha} \vdash F}f[\ol{\alpha :=
-\setsem{\Gamma;\Phi \vdash \tau}f}]
-\end{gather}
-\item If $\Gamma;\Phi,\phi^k\vdash F$ and
-  $\Gamma;\Phi,\alpha_1...\alpha_k \vdash H$, then 
-\begin{gather}\label{eq:subs-const}
-\setsem{\Gamma; \Phi \vdash F[\phi := H]}\rho
-= \setsem{\Gamma; \Phi, \phi \vdash F}\rho
-[\phi := \lambda \ol{A}.\, \setsem{\Gamma;\Phi,\overline{\alpha}\vdash
-    H}\rho[\overline{\alpha := A}]] \\ 
-\intertext{and}\label{eq:subs-const-morph}
-\setsem{\Gamma; \Phi \vdash F[\phi := H]}f
-= \setsem{\Gamma; \Phi, \phi \vdash F}f
-[\phi := \lambda \ol{A}.\,\setsem{\Gamma;\Phi,\overline{\alpha}\vdash
-    H}f[\overline{\alpha := \id_{\ol{A}}}]] 
-\end{gather}
-\end{itemize}
-Analogous identities hold for relation environments and morphisms
-between them.
-\end{lemma}
-\end{comment}
-
-\begin{comment}
-\begin{proof}
-The proofs for the set and relational interpretations are completely
-analogous, so we just prove the former. Likewise, we only prove
-Equations~\ref{eq:subs-var} and~\ref{eq:subs-const}, since the proofs
-for Equations~\ref{eq:subs-var-morph} and~\ref{eq:subs-const-morph}
-are again analogous. Finally, we prove Equation~\ref{eq:subs-var} for
-substitution for just a single type variable since the proof for
-multiple simultaneous substitutions proceeds similarly.
-
-Although Equation~\ref{eq:subs-var} is a special case of
-Equation~\ref{eq:subs-const}, it is convenient to prove
-Equation~\ref{eq:subs-var} first, and then use it to prove
-Equation~\ref{eq:subs-const}. We prove Equation~\ref{eq:subs-var} by
-induction on the structure of $F$ as follows:
-\begin{itemize}
-\item If $\Gamma; \emptyset \vdash F : \T$, or if $F$ is $\onet$ or
-  $\zerot$, then $F$ does not contain any functorial variables to
-  replace, so there is nothing to prove.
-\item If $F$ is $F_1 \times F_2$ or $F_1 + F_2$, then the substitution
-  distributes over the product or coproduct as appropriate, so the
-  result follows immediately from the induction hypothesis.
-\item If $F = \beta$ with $\beta \neq \alpha$, then there is nothing to
-  prove.
-\item If $F = \alpha$, then
-\[\setsem{\Gamma;\Phi\vdash\alpha[\alpha := \tau]}\rho \, = \,
-\setsem{\Gamma;\Phi\vdash\tau}\rho \, = \,
-\setsem{\Gamma;\Phi,\alpha\vdash\alpha}\rho[\alpha :=
-  \setsem{\Gamma;\Phi\vdash\tau}\rho]\]
-\item If $F = \phi \overline{\sigma}$ with $\phi \neq \alpha$, then
-\[\begin{array}{ll}
- &\setsem{\Gamma; \Phi \vdash (\phi \overline{\sigma})[\alpha := \tau]}\rho\\
-=&\setsem{\Gamma; \Phi \vdash \phi (\overline{\sigma[\alpha := \tau]})}\rho \\
-=&(\rho\phi)\overline{\setsem{\Gamma;\Phi\vdash\sigma[\alpha := \tau]}\rho} \\
-=&(\rho\phi)\overline{\setsem{\Gamma;\Phi,\alpha\vdash\sigma}
-      \rho [\alpha := \setsem{\Gamma;\Phi\vdash\tau}] } \\
-=&\setsem{\Gamma; \Phi, \alpha \vdash \phi \overline{\sigma}}
-      \rho [\alpha := \setsem{\Gamma;\Phi\vdash\tau}]
-\end{array}\]
-Here, the third equality is by the induction hypothesis.
-\item If $F = (\mu \phi. \lambda \overline{\beta}. G)
-  \overline{\sigma}$, then
-\[\begin{array}{ll}
-  & \setsem{\Gamma; \Phi \vdash ((\mu \phi. \lambda
-  \overline{\beta}. G) \overline{\sigma}) [\alpha := \tau]}\rho \\
-=&\setsem{\Gamma; \Phi \vdash (\mu \phi. \lambda
-  \overline{\beta}. G[\alpha := \tau]) (\overline{\sigma [\alpha :=
-      \tau]})}\rho \\ 
-=&\mu(\setsem{\Gamma; \Phi, \phi, \overline{\beta} \vdash G[\alpha := \tau]}
-      \rho[\phi := \text{--}][\overline{\beta := \text{--}}])
-      (\overline{\setsem{\Gamma;\Phi \vdash \sigma [\alpha := \tau]}\rho}) \\
-=&\mu(\setsem{\Gamma; \Phi, \phi, \overline{\beta}, \alpha \vdash G}
-\rho [\alpha := \setsem{\Gamma;\Phi \vdash \tau}\rho] [\phi := \text{--}]
-[\overline{\beta := \text{--}}]) \\
- &\hspace{0.5in} (\overline{\setsem{\Gamma;\Phi,\alpha \vdash \sigma}
-   \rho [\alpha := \setsem{\Gamma;\Phi\vdash \tau}\rho]}) \\
-=&\setsem{\Gamma;\Phi,\alpha\vdash(\mu\phi.\lambda\overline{\beta}.G)
-  \overline{\sigma}}\rho  [\alpha := \setsem{\Gamma;\Phi \vdash
-   \tau}\rho] 
-\end{array}  \]
-Here, the third equality is by the induction hypothesis and weakening.
-\end{itemize}
-
-We now prove Equation~\ref{eq:subs-const}, again by induction on the
-structure of $F$.
-\begin{itemize}
-\item If $\Gamma; \emptyset \vdash F : \T$, or if $F$ is $\onet$ or
-  $\zerot$, then $F$ does not contain any functorial variables to
-  replace, so there is nothing to prove.
-\item If $F$ is $F_1 \times F_2$ or $F_1 + F_2$, then the substitution
-  distributes over the product or coproduct as appropriate, so the
-  result follows immediately from the induction hypothesis.
-\item If $F = \phi \overline{\tau}$, then
-\[\begin{array}{ll}
- &\setsem{\Gamma; \Phi \vdash (\phi \overline{\tau})[\phi := H]}\rho \\
-=&\setsem{\Gamma; \Phi \vdash H[\overline{\alpha:= \tau[\phi := H]}]}\rho \\
-=&\setsem{\Gamma; \Phi \vdash H}\rho [\overline{\alpha:=
-    \setsem{\Gamma;\Phi\vdash\tau[\phi := H]}\rho}] \\ 
-=&\setsem{\Gamma; \Phi \vdash H}\rho [\overline{\alpha:=
-    \setsem{\Gamma;\Phi,\phi\vdash\tau} \rho [\phi :=
-      \setsem{\Gamma;\Phi,\overline{\alpha}\vdash
-        H}\rho[\overline{\alpha := \text{--}}]]}] \\
-=&\setsem{\Gamma; \Phi, \phi \vdash \phi \overline{\tau}}\rho [\phi :=
-  \setsem{\Gamma;\Phi,\overline{\alpha}\vdash H}\rho[\overline{\alpha
-      := \text{--}}]] 
-  \end{array}\]
-Here, the first equality is by
-Definition~\ref{def:second-order-subst}, the second is by
-Equation~\ref{eq:subs-var}, the third is by the induction hypothesis,
-and the fourth is by Definition~\ref{def:set-sem}.
-\item If $F = \psi \overline{\tau}$ with $\psi \neq \phi$, then the
-  proof is similar to that for the previous case, but simpler, because
-  $\phi$ only needs to be substituted in the arguments $\ol{\tau}$ of
-  $\psi$.
-\item If $F = (\mu \psi. \lambda \overline{\beta}. G)
-  \overline{\tau}$, then
-\[\begin{array}{ll}
- &\setsem{\Gamma; \Phi \vdash ((\mu \psi. \lambda \overline{\beta}. G)
- \overline{\tau})[\phi := H]}\rho \\ 
-=&\setsem{\Gamma; \Phi \vdash (\mu \psi. \lambda
-  \overline{\beta}. G[\phi := H]) (\overline{\tau[\phi := H]})}\rho \\
-=&\mu(\setsem{\Gamma; \Phi, \psi, \overline{\beta} \vdash G[\phi := H]}
-    \rho[\psi := \text{--}][\overline{\beta := \text{--}}])
-    (\overline{\setsem{\Gamma;\Phi \vdash \tau[\phi := H]}\rho}) \\
-=&\mu(\setsem{\Gamma; \Phi, \psi, \overline{\beta}, \phi \vdash G}
-\rho [\phi := \setsem{\Gamma;\Phi,\overline{\alpha}\vdash
-    H}\rho[\overline{\alpha := \text{--}}]][\psi := \text{--}]
-     [\overline{\beta := \text{--}}]) \\
-&\hspace{0.5in} (\overline{\setsem{\Gamma;\Phi,\phi \vdash \tau} \rho
-       [\phi := \setsem{\Gamma;\Phi,\overline{\alpha}\vdash H}
-         \rho[\overline{\alpha := \text{--}}]]}) \\
-=&\setsem{\Gamma;\Phi,\phi\vdash(\mu\psi.\lambda\overline{\beta}.G)
-  \overline{\tau}}\rho [\phi :=
-  \setsem{\Gamma;\Phi,\overline{\alpha}\vdash H}\rho[\overline{\alpha
-      := \text{--}}]] 
-\end{array}\]
-Here, the first equality is by
-Definition~\ref{def:second-order-subst}, the second and fourth are by
-Definition~\ref{def:set-sem}, and the third is by the induction
-hypothesis and weakening.
-\end{itemize}
-\end{proof}
-\end{comment}
 
@@
 relations as particular graph relations. Our definitions specialize to
-the usual ones for the graph relation for a morphism between sets and
-the equality relation for a set. In light of its novelty, we spell out
+the usual ones for the graph relation for morphisms between sets and
+equality relations on sets. In light of its novelty, we spell out
 our construction in detail.
@@
 \begin{dfn}\label{dfn:graph-nat-transf}
-If $F, G: \Set^k \to \Set$
-%are $k$-ary functors
-and $\alpha : F \to G$ is a natural transformation, then the functor
-$\graph{\alpha}^*: \rel^k \to \rel$ is defined as follows. Given $R_1
-: \rel(A_1, B_1),...,R_k : \rel(A_k,B_k)$, let $\iota_{R_i} : R_i
-\hookrightarrow A_i \times B_i$, for $i = 1,...,k$, be the inclusion
-of $R_i$ as a subset of $A_i \times B_i$,
+If $F, G: \Set^k \to \Set$ and $\alpha : F \to G$ is a natural
+transformation, then the functor $\graph{\alpha}^*: \rel^k \to \rel$
+is defined as follows. Given $R_1 : \rel(A_1, B_1),...,R_k :
+\rel(A_k,B_k)$, let $\iota_{R_i} : R_i \hookrightarrow A_i \times
+B_i$, for $i = 1,...,k$, be the inclusion of $R_i$ as a subset of $A_i
+\times B_i$,
 %. By the universal property of the product, there exists a
@@
 
-The data in Definition~\ref{dfn:graph-nat-transf} yield the {\em graph relation
-transformer for $\alpha$}, denoted $\graph{\alpha} = (F, G,
+The data in Definition~\ref{dfn:graph-nat-transf} yield the {\em graph
+  relation transformer for $\alpha$}, denoted $\graph{\alpha} = (F, G,
 \graph{\alpha}^*)$.
-% yield a relation transformer $\graph{\alpha} = (F, G,
-%\graph{\alpha}^*)$, called the {\em graph relation transformer for
-%$\alpha$}.
 
@@
 \end{figure*}
-%{\footnotesize    \[
-%    \begin{tikzcd}
-%        \alpha^\wedge\overline{R}
-%        \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{R}}}"]
-%        \ar[d, "{\epsilon}"']
-%        & F\overline{A} \times G\overline{B}
-%        \ar[d, "{F\overline{\beta} \times G\overline{\beta'}}"] \\
-%        \alpha^\wedge\overline{S}
-%        \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{S}}}"']
-%        & F\overline{C} \times G\overline{D}
-%    \end{tikzcd}
-%    \]}
-%    commutes.
-%    Since $\ol{(\beta,\beta') : R \to S}$, there exist $\overline{\gamma : R \to S}$
-%    such that each diagram
-%{\footnotesize    \[
-%    \begin{tikzcd}
-%        R_i
-%        \ar[d, "{\gamma_i}"']
-%        \ar[r, hookrightarrow, "{\iota_{R_i}}"]
-%        &A_i \times B_i
-%        \ar[d, "{\beta_i \times \beta'_i}"] \\
-%        S_i
-%        \ar[r, hookrightarrow, "{\iota_{S_i}}"]
-%        &C_i \times D_i
-%    \end{tikzcd}
-%    \]}
-%    commutes.
-%Moreover, since both $h_{\overline{C \times D}} \circ
-%F(\overline{\beta \times \beta'})$ and $(F\overline{\beta} \times
-%G\overline{\beta'}) \circ h_{\overline{A \times B}}$ make
-%{\footnotesize \[
-%      \begin{tikzcd}[row sep = large]
-%          F\overline{C}
-%          &F\overline{C} \times F\overline{D}
-%          \ar[l, "{\pi_1}"'] \ar[r, "{\pi_2}"]
-%          &F\overline{D}
-%          \ar[r, "{\alpha_{\ol{D}}}"]
-%          &G\overline{D}\\
-%          &F(\overline{A \times B})
-%          \ar[u, dashed, "{\exists !}"]
-%          \ar[ul, "{F\ol{\pi_1} \circ F(\overline{\beta \times \beta'})}"]
-%          \ar[urr, "{\alpha_{\ol{D}} \circ F\ol{\pi_2} \circ F(\overline{\beta \t%imes \beta'})}"']
-%      \end{tikzcd}
-%      \]}
-%      commute, they must be equal.
 
@@
 Finally, by the left-lifting property of $q_{F^\wedge\overline{R}}$
-with respect to $\iota_{F^\wedge\overline{S}}$ given by the epi-mono
+with respect to $\iota_{F^\wedge\overline{S}}$ given by the mono-epi
 factorization system, there exists an $\epsilon$ such that the diagram
-on the right below commutes.
+on the right below commutes as desired.
 \begin{figure*}[ht]
@@
 Finally, by the left-lifting property of $q_{F^\wedge\overline{R}}$
-      with respect to $\iota_{F^\wedge\overline{S}}$ given by the epi-mono
+      with respect to $\iota_{F^\wedge\overline{S}}$ given by the mono-epi
       factorization system, there exists an $\epsilon$ such that the
@@
 
-To prove the IEL, we will need to know that the equality relation
-transformer preserves equality relations; see
-Equation~\ref{eq:eq-pres-eq} below. This will follow from the next
-lemma, which shows how to compute the action of a graph relation
-transformer on any graph relation.
+The action of a graph relation transformer on a graph relation can be
+computed explicitly:
 
@@
   \vspace*{-0.15in}
-%  \hspace*{-0.5in}
   \begin{minipage}[b]{0.45\linewidth}
@@
 \end{minipage}
-%  \vspace*{-0.5in}
   \begin{minipage}[b]{0.45\linewidth}
@@
 \noindent
-Moreover, $\langle \id_{F \ol{A}}, \alpha_{\ol{B}} \circ
-F\ol{f} \rangle$ is a monomorphism in $\set$ because $\id_{F \ol{A}}$
-is, so its epi-mono factorization gives $\iota_{\alpha^\wedge
+Moreover, $\langle \id_{F \ol{A}}, \alpha_{\ol{B}} \circ F\ol{f}
+\rangle$ is a monomorphism in $\set$ because $\id_{F \ol{A}}$ is, so
+its mono-epi factorization gives $\iota_{\alpha^\wedge
   \graph{\overline{f}}} = \langle \id_{F \ol{A}}, \alpha_{\ol{B}}
 \circ F\ol{f} \rangle$, and thus $\alpha^\wedge \graph{\overline{f}} =
-%the domain of $\iota_{\alpha^\wedge \graph{\overline{f}}}$, is equal
-%to
-F\overline{A}$.  Then $\iota_{\alpha^\wedge
-  \graph{\overline{f}}} \alpha^\wedge \graph{\overline{f}} = \langle
-\id_{F \ol{A}}, \alpha_{\ol{B}} \circ F\ol{f} \rangle (F \ol{A}) =
-\graph{ \alpha_{\ol{B}} \circ F\ol{f} }^*$,
-%(where the last equality is by Remark~\ref{rmk:graph-fn}).
-%We therefore conclude that
-so that $\graph{ \alpha }^* \graph{ \overline{f} } = (F\overline{A},
-G\overline{B}, \iota_{\alpha^\wedge \graph{\overline{f}}}\,
-\alpha^\wedge \graph{\overline{f}}) = (F\overline{A}, G\overline{B},
-\graph{ \alpha_{\ol{B}} \circ F\ol{f} }^*) = \graph{ \alpha_{\ol{B}}
-  \circ F\ol{f} }$.  Finally, $\alpha_{\ol{B}} \circ F\ol{f} = G\ol{f}
-\circ \alpha_{\ol{A}}$ by naturality of $\alpha$.
+F\overline{A}$.  Then $\iota_{\alpha^\wedge \graph{\overline{f}}}
+\alpha^\wedge \graph{\overline{f}} = \langle \id_{F \ol{A}},
+\alpha_{\ol{B}} \circ F\ol{f} \rangle (F \ol{A}) = \graph{
+  \alpha_{\ol{B}} \circ F\ol{f} }^*$, so that $\graph{ \alpha }^*
+\graph{ \overline{f} } = (F\overline{A}, G\overline{B},
+\iota_{\alpha^\wedge \graph{\overline{f}}}\, \alpha^\wedge
+\graph{\overline{f}}) = (F\overline{A}, G\overline{B}, \graph{
+  \alpha_{\ol{B}} \circ F\ol{f} }^*) = \graph{ \alpha_{\ol{B}} \circ
+  F\ol{f} }$.  Finally, $\alpha_{\ol{B}} \circ F\ol{f} = G\ol{f} \circ
+\alpha_{\ol{A}}$ by naturality of $\alpha$.
 \end{proof}
 
-The {\em equality relation transformer} on $F : [\set^k,\set]$ is
-defined to be $\Eq_F = \graph{\id_{F}}$. Specifically, $\Eq_F = (F, F,
-\Eq_F^*)$ with $\Eq_F^* = \graph{\id_{F}}^*$, and
-Lemma~\ref{lem:eq-reln-equalities} indeed ensures that
+To prove the IEL, we also need to know that the equality relation
+transformer preserves equality relations.  The {\em equality relation
+  transformer} on $F : [\set^k,\set]$ is defined to be $\Eq_F =
+\graph{\id_{F}} = (F, F, \graph{\id_{F}}^*)$.
+Lemma~\ref{lem:eq-reln-equalities} then gives that, for all $\ol{A :
+\set}$,
 \begin{equation}\label{eq:eq-pres-eq}
@@
 \end{equation}
-%$\Eq^*_F \,\ol{\Eq_A} = \Eq_{F\ol{A}}$
-for all $\ol{A : \set}$.
-\begin{comment}
-  \begin{lemma}
-$\Eq^*_F \,\ol{\Eq_A}
-= \Eq_{F\ol{A}}$ for all $\ol{A : \set}$.
-  \end{lemma}
-\begin{proof}
-We have that\
-\begin{equation}\label{eq:eq-pres-eq}
-\Eq^*_F \ol{\Eq_A}
-= \graph{\id_F}^* \graph{\id_{\ol{A}}}
-= \graph{F \id_{\ol{A}} \circ (\id_F)_{\ol{A}}}  %by your Lemma 5
-= \graph{\id_{F\ol{A}} \circ \id_{F\ol{A}}}
-%= \graph{\id_F \ol{A} \circ \id_{\ol{A}}}
-= \graph{\id_{F\ol{A}}}
-= \Eq_{F\ol{A}}
-\end{equation}
-The second identity here is by Lemma~\ref{lem:eq-reln-equalities}.
-\end{proof}
-\end{comment}
+
 Graph relation transformers in general, and equality relation
-transformers in particular, extend to relation environments in the
-obvious ways.
-%\begin{definition}
+transformers in particular, naturally extend to relation environments.
 Indeed, if $\rho, \rho' : \setenv$ and $f : \rho \to \rho'$, then the
@@
   \phi}$ for every $\phi$.
-%\end{definition}
+
 With these definitions in hand, we can state and prove both an
@@
 \begin{thm}[IEL]\label{thm:iel}
-  If $\rho : \setenv$ and $\Gamma; \Phi \vdash F$ then
-  $\relsem{\Gamma; \Phi \vdash F} \Eq_\rho = \Eq_{\setsem{\Gamma;
-      \Phi \vdash F}\rho}$.
+If $\rho : \setenv$ and $\Gamma; \Phi \vdash F$ then $\relsem{\Gamma;
+  \Phi \vdash F} \Eq_\rho = \Eq_{\setsem{\Gamma; \Phi \vdash F}\rho}$.
 \end{thm}
+{\color{red} INSERT ENTIRE PROOF. CHECK PROOF FOR $\mu$-TYPES AND
+  CHECK NOTATION.}
 The proof is by induction on the structure of $F$. Only the $\Nat$,
@@
 
-%It follows from Theorem~\ref{thm:iel} that $\relsem{\Gamma \vdash
-%  \sigma \to \tau} \Eq_{\rho} = \Eq_{\setsem{\Gamma \vdash \sigma \to
-%    \tau}\rho}$, as expected.  Moreover,
-
-%\vspace*{0.05in}
-%
-%The Graph Lemma appropriate to our setting is now easily obtained:
+With the IEL in hand we can prove a Graph Lemma appropriate to our
+setting:
 \begin{lemma}[Graph Lemma]\label{lem:graph}
-If $\rho, \rho' : \setenv$, $f : \rho \to \rho'$, and $\Gamma; \Phi
-\vdash F$, then $\graph{\setsem{\Gamma; \Phi \vdash F} f} =
-\relsem{\Gamma; \Phi \vdash F}\graph{f}$.
+If $\rho, \rho' : \setenv$ and $f : \rho \to \rho'$ then
+$\graph{\setsem{\Gamma; \Phi \vdash F} f} = \relsem{\Gamma; \Phi
+  \vdash F}\graph{f}$.
 \end{lemma}
@@
 \end{comment}
+
+{\color{red} STOPPED HERE!!!!!}
 
@@
   G} \rho\\
-\dsem{\Gamma;\Phi~|~\Delta \vdash \case{t}{x \mapsto l}{y \mapsto r} :
+\dsem{\Gamma;\Phi~|~\Delta \vdash \cse{t}{x \mapsto l}{y \mapsto r} :
   K}\rho & = & \eval \circ \langle \curry \,[\dsem{\Gamma;\Phi
@@
   \tau} \rho\\
-\relsem{\Gamma;\Phi~|~\Delta \vdash \case{t}{x \mapsto l}{y \mapsto r} :
+\relsem{\Gamma;\Phi~|~\Delta \vdash \cse{t}{x \mapsto l}{y \mapsto r} :
   \gamma}\rho & = & \eval \circ \langle \curry \,[\relsem{\Gamma;\Phi
@@
   \Nat^{\overline{\alpha}} F\,G}$. Straightforward calculation shows
-that terms of $\Nat$ type behave as natural transformations with
+that terms of $\Nat$-type behave as natural transformations with
 respect to their source and target functorial types, i.e., we have

### 2020-12-11 free-theorems-nesteds/20-mscs.tex
more updates to mscs.tex

@@
-% double-blind review submission, w/o CCS and ACM Reference (max
-% submission space)
-\documentclass[acmsmall,review]{acmart}
-\settopmatter{printfolios=true,printccs=false,printacmref=false}
-%% For double-blind review submission, w/ CCS and ACM Reference
-%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
-%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
-%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-%% For single-blind review submission, w/ CCS and ACM Reference
-%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
-%% For final camera-ready submission, w/ required CCS and ACM Reference
-%\documentclass[acmsmall]{acmart}\settopmatter{}
-
-%% Journal information
-%% Supplied to authors by publisher for camera-ready submission;
-%% use defaults for review submission.
-\acmJournal{PACMPL}
-\acmVolume{1}
-\acmNumber{POPL} % CONF = POPL or ICFP or OOPSLA
-\acmArticle{1}
-\acmYear{2020}
-\acmMonth{1}
-\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
-
-%% Copyright information
-%% Supplied to authors (based on authors' rights management selection;
-%% see authors.acm.org) by publisher for camera-ready submission;
-%% use 'none' for review submission.
-\setcopyright{none}
-%\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
-%\copyrightyear{2018}           %% If different from \acmYear
-
-%% Bibliography style
-\bibliographystyle{ACM-Reference-Format}
-%% Citation style
-%% Note: author/year citations are required for papers published as an
-%% issue of PACMPL.
-\citestyle{acmauthoryear}   %% For author/year citations
-%\citestyle{acmnumeric}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Note: Authors migrating a paper from PACMPL format to traditional
-%% SIGPLAN proceedings format must update the '\documentclass' and
-%% topmatter commands above; see 'acmart-sigplanproc-template.tex'.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\documentclass{lmcs}
+
+\keywords{MANDATORY list of keywords}
 
 \usepackage[utf8]{inputenc}
-\usepackage{ccicons}
+%\usepackage{ccicons}
 \usepackage{verbatim}
-
 \usepackage{adjustbox}
@@
 \usepackage{upgreek}
-%\usepackage{stmaryrd}
+\usepackage{stmaryrd}
 
@@
 
+\usepackage{hyperref}
+
 \input{macros}
-\input{mytheorem}
-
-\theoremstyle{definition}
-\newtheorem{exmpl}{Example}
-
-\renewcommand{\greyout}[1]{} %{{\color{gray} {#1}}} -- toggle to remove greyed text
-
+%\input{mytheorem}
+
+\theoremstyle{plain}\newtheorem{satz}[thm]{Satz}
+
+\def\eg{{\em e.g.}}
+\def\cf{{\em cf.}}
+
+%\theoremstyle{definition}
+
+\newcommand{\inl}{\mathsf{inl}}
+\newcommand{\inr}{\mathsf{inr}}
+\newcommand{\fold}{\mathsf{fold}}
 \newcommand{\ininv}[2]{(\tin^{-1}_{\onet +
@@
 \newcommand{\F}{\mathcal{F}}
-\renewcommand{\G}{\mathcal{G}}
+\newcommand{\G}{\mathcal{G}}
 \newcommand{\N}{\mathcal{N}}
@@
 \renewcommand{\S}{\mathcal S}
-\renewcommand{\C}{\mathcal{C}}
+\newcommand{\C}{\mathcal{C}}
 \newcommand{\D}{\mathcal{D}}
@@
 
-\title[Paramrtricity for Primitive Nested Types and
-  GADTs]{Parametricity for Primitive
-  Nested Types and GADTs} %% [Short Title] is optional; when present,
-                         %% will be used in header instead of Full
-                         %% Title.
-%\titlenote{with title note}             %% \titlenote is optional;
-                                        %% can be repeated if necessary;
-                                        %% contents suppressed with 'anonymous'
-%\subtitle{Subtitle}                     %% \subtitle is optional
-%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
-                                        %% can be repeated if necessary;
-                                        %% contents suppressed with 'anonymous'
-
-%% Author information
-%% Contents and number of authors suppressed with 'anonymous'.
-%% Each author should be introduced by \author, followed by
-%% \authornote (optional), \orcid (optional), \affiliation, and
-%% \email.
-%% An author may have multiple affiliations and/or emails; repeat the
-%% appropriate command.
-%% Many elements are not rendered, but should be provided for metadata
-%% extraction tools.
-
-%% Author with single affiliation.
-\author{Patricia Johann, Enrico Ghiorzi, and Daniel Jeffries}
-%\authornote{with author1 note}          %% \authornote is optional;
-%                                        %% can be repeated if necessary
-%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
-\affiliation{
-%  \position{Position1}
-%  \department{Department1}              %% \department is recommended
-  \institution{Appalachian State University}            %% \institution is required
-%  \streetaddress{Street1 Address1}
-%  \city{City1}
-%  \state{State1}
-%  \postcode{Post-Code1}
-%  \country{Country1}                    %% \country is recommended
-}
-\email{johannp@appstate.edu, ghiorzie@appstate.edu, jeffriesd@appstate.edu}          %% \email is recommended
+%% due to the dependence on amsart.cls, \begin{document} has to occur
+%% BEFORE the title and author information:
 
 \begin{document}
+
+\title[Parametricity for Primitive
+  Nested Types and GADTs]{Parametricity for Primitive
+  Nested Types and GADTs}
+\titlecomment{{\lsuper*}OPTIONAL comment concerning the title, \eg,
+  if a variant or an extended abstract of the paper has appeared elsewhere.}
+
+\author[P.~Johann]{Patricia Johann}	%required
+\address{Appalachian State University}	%required
+\email{johannp@appstate.edu}  %optional
+%\thanks{thanks 1, optional.}	%optional
+
+\author[E.~Ghiorzi]{Enrico Ghiorzi}	%optional
+\address{Appalachian State University}	%optional
+\email{ghiorzie@appstate.edu}  %optional
+%\thanks{thanks 2, optional.}	%optional
+
+\author[D.~Jeffries]{Daniel Jeffries}	%optional
+\address{Appalachian State University}	%optional
+\email{jeffriesd@appstate.edu}
+%\thanks{thanks 3, optional.}	%optional
 
@@
 
-\begin{dfn}\label{def:wftypes}
+\begin{defi}\label{def:wftypes}
 The formation rules for the set $\F$ of\, {\em well-formed types
@@
 \end{array}\]
-\end{dfn}
+\end{defi}
 
@@
 {\color{red} Check no change needed for new $\mu$-rule in next def.}
-\begin{dfn}\label{def:second-order-subst}
+\begin{defi}\label{def:second-order-subst}
 If \,$\Gamma; \Phi,\phi^k \vdash H$ and $\Gamma;\Phi, \ol{\alpha}
@@
 \end{array}\]
-\end{dfn}
+\end{defi}
 \noindent
@@
 
-\begin{dfn}\label{def:set-env}
+\begin{defi}\label{def:set-env}
 A {\em set environment} maps each type variable in $\tvars^k \cup
@@
 environments and morphisms between them, which we denote $\setenv$.
-\end{dfn}
+\end{defi}
 When convenient we identify a functor in $[\set^0, \set]$ with its
@@
 
-\begin{dfn}\label{def:set-sem}
+\begin{defi}\label{def:set-sem}
 {\color{red} Check clause for $\mu$-types.} The {\em set
@@
 \end{align*}
-\end{dfn}
+\end{defi}
 %The interpretations in Definition~\ref{def:set-sem} respect weakening,
@@
 interpretation.
-\begin{dfn}\label{def:set-sem-funcs}
+\begin{defi}\label{def:set-sem-funcs}
 Let $f: \rho \to \rho'$ be a morphism between set environments $\rho$
@@
 \end{itemize}
-\end{dfn}
+\end{defi}
 
@@
 
-\begin{dfn}\label{def:rel-transf}
+\begin{defi}\label{def:rel-transf}
 A {\em $k$-ary relation transformer} $F$ is a triple $(F^1, F^2,F^*)$,
@@
 $F\overline{(\alpha,\beta)}$ to be $F^*\overline{(\alpha,\beta)}$.
-\end{dfn}
+\end{defi}
 The last clause of Definition~\ref{def:rel-transf} expands to: if
@@
 
-\begin{dfn}
+\begin{defi}
 The category $RT_k$ of $k$-ary relation transformers is given by the
@@
 \end{itemize}
-\end{dfn}
-
-\begin{dfn}\label{def:RT-functor}
+\end{defi}
+
+\begin{defi}\label{def:RT-functor}
 An endofunctor $H$ on $RT_k$ is a triple $H = (H^1,H^2,H^*)$, where
@@
 \end{itemize}
-\end{dfn}
+\end{defi}
 Since the results of applying an endofunctor $H$ to $k$-ary relation
@@
 
-\begin{dfn}\label{def:RT-nat-trans}
+\begin{defi}\label{def:RT-nat-trans}
 For endofunctors $H$ and $K$ on $RT_k$, a {\em natural transformation}
@@
 $\sigma_F = (\sigma^1_{F^1}, \sigma^2_{F^2})$.
-\end{dfn}
+\end{defi}
 Definition~\ref{def:RT-nat-trans} entails that $\sigma^i_{F^i}$ is
@@
 \begin{comment}
-\begin{dfn}\label{def:omega-cocont}
+\begin{defi}\label{def:omega-cocont}
 An endofunctor $T = (T^1,T^2,T^*)$ on $RT_k$ is {\em
@@
 $[RT_k,[\rel^k,\rel]]$.
-\end{dfn}
+\end{defi}
 \end{comment}
@@
 $\omega$-cocontinuous:
-\begin{lemma}\label{lem:fp}
+\begin{lem}\label{lem:fp}
 For any $T : [RT_k,RT_k]$, $\mu T \cong T(\mu T)$.
-\end{lemma}
+\end{lem}
 \noindent
@@
 
-\begin{dfn}\label{def:reln-env}
+\begin{defi}\label{def:reln-env}
 A {\em relation environment} maps each type variable in $\tvars^k \cup
@@
 denote $\relenv$.
-\end{dfn}
+\end{defi}
 We identify a $0$-ary relation transformer with the relation
@@
 
-\begin{dfn}\label{def:relenv-functor}
+\begin{defi}\label{def:relenv-functor}
 For each $k$, an $\omega$-cocontinuous functor $H : [\relenv, RT_k]$
@@
 \end{itemize}
-\end{dfn}
+\end{defi}
 \noindent Spelling out the last two bullet points above gives the
@@
 
-\begin{dfn}\label{def:rel-sem}
+\begin{defi}\label{def:rel-sem}
 The {\em relational interpretation} $\relsem{\cdot} : \F \to [\relenv,
@@
 \end{align*}
-\end{dfn}
+\end{defi}
 
@@
 interpretation.
-\begin{dfn}\label{def:rel-sem-funcs}
+\begin{defi}\label{def:rel-sem-funcs}
 Let $f: \rho \to \rho'$ for relation environments $\rho$ and $\rho'$
@@
 \end{comment}
-\end{dfn}
+\end{defi}
 
@@
 $\relenv$. This is an immediate consequence of
-\begin{lemma}\label{lem:rel-transf-morph}
+\begin{lem}\label{lem:rel-transf-morph}
 For every $\Gamma;\Phi \vdash F$,
@@
 \in[\relenv,RT_0]\] 
-\end{lemma}
+\end{lem}
 \noindent
@@
 
-\begin{dfn}\label{dfn:graph-nat-transf}
+\begin{defi}\label{dfn:graph-nat-transf}
 If $F, G: \Set^k \to \Set$ and $\alpha : F \to G$ is a natural
@@
 (F\overline\beta, G\overline\beta')$.
-\end{dfn}
+\end{defi}
 
@@
 
-\begin{lemma}\label{lem:graph-reln-functors}
+\begin{lem}\label{lem:graph-reln-functors}
 If $F,G : [\set^k,\set]$, and if $\alpha : F \to G$ is a natural
 transformation, then $\graph{\alpha}$ is in $RT_k$.
-\end{lemma}
-\begin{proof}
+\end{lem}
+\proof
 Clearly, $\graph{\alpha}^*$ is $\omega$-cocontinuous, so
@@
 \end{comment}
-\end{proof}
+\qed
 
@@
 
-\begin{dfn}
+\begin{defi}
 Let $F : [\set^k, \set]$.  The equality relation transformer on $F$ is
@@
 F, \Eq_F^*)$ with $\Eq_F^* = \graph{\id_{F}}^*$.
-\end{dfn}
+\end{defi}
 \end{comment}
@@
 
-\begin{lemma}\label{lem:eq-reln-equalities}
+\begin{lem}\label{lem:eq-reln-equalities}
 If $\alpha : F \to G$ is a morphism in $[\Set^k, \Set]$
@@
 = \langle \alpha_{\ol{B}} \circ F \ol{f} \rangle$.
-\end{lemma}
-\begin{proof}
+\end{lem}
+\proof
 Since $h_{\overline{A \times B}}$ is the unique morphism making the
@@
 \alpha_{\ol{A}}$ by naturality of $\alpha$.
-\end{proof}
+\qed
 
@@
 \begin{comment}
-\begin{proof}
+\proof
 By induction on the structure of $\tau$.
@@
 \end{itemize}
-\end{proof}
+\qed
 \end{comment}
@@
 setting:
-\begin{lemma}[Graph Lemma]\label{lem:graph}
+\begin{lem}[Graph Lemma]\label{lem:graph}
 If $\rho, \rho' : \setenv$ and $f : \rho \to \rho'$ then
@@
   \vdash F}\graph{f}$.
-\end{lemma}
-\begin{proof}
+\end{lem}
+\proof
 Applying Lemma~\ref{lem:rel-transf-morph} to the morphisms $(f,
@@
 \subseteq \relsem{\Gamma; \Phi \vdash F} \graph{f}$.
-\end{proof}
+\qed
 
@@
 
-\begin{dfn}\label{def:set-interp}
+\begin{defi}\label{def:set-interp}
 If $\rho$ is a set (resp., relation) environment and
@@
 $\rel$ as appropriate.
-\end{dfn}
+\end{defi}
 
@@
 \begin{comment}
-\begin{dfn}\label{def:rel-interp}
+\begin{defi}\label{def:rel-interp}
 If $\rho$ is a relation environment and $\Gamma;\Phi~|~\Delta \vdash t :
@@
 \end{array}\]
-\end{dfn}
+\end{defi}
 {\color{blue} Add return type for fold in last clause? Should be
@@
 \end{thm}
-\begin{proof}
+\proof
 By structural induction on the type judgment for $t$. The overall
@@
 i.e., are fibred.
-\end{proof}
+\qed
 
@@
 \end{thm}
-\begin{proof}
+\proof
 By Theorem~\ref{thm:at-gen-rel}, \[(\mathit{filter}\, (\pi_1 \rho)\, a,
@@
 follows from % Instantiating
-Equation~\ref{eq:filter-thm-list}.\end{proof}
+Equation~\ref{eq:filter-thm-list}.
+\qed
 %gives the result.
@@
 \end{thm}
-\begin{proof}
+\proof
  Immediate from Theorem~\ref{thm:at-gen}.
-\end{proof}
+\qed
 \end{comment}
@@
 \end{thm}
-\begin{proof}
+\proof
   By Theorem~\ref{thm:at-gen-rel},
@@
   as desired.
-\end{proof}
+\qed
 \end{comment}
@@
 \end{thm}
-\begin{proof}
+\proof
 %Let $\vdash \tau : \F$ and $\vdash \tau' : \F$, let
@@
 = G \,\setsem{\vdash H}\, n\, c$.
-\end{proof}
+\qed
 
@@
 
-\begin{proof}
+\proof
   We first note that the type of $g$ is well-formed, since
@@
 G\,\setsem{\vdash \tau'}\,B\]  
-\end{proof}
+\qed
 \end{comment}
@@
 
-\begin{proof}
+\proof
 %We first note that the type of $g$ is well-formed since
@@
 G\,\setsem{\emptyset;\alpha\vdash K}\,B$. 
-\end{proof}
+\qed
 
@@
 
-%Won't print in anonymous mode.
-\begin{acks}
-We thank Andrew Polonsky for valuable feedback and fruitful
-discussions. Supported by NSF awards CCR-1906388 and 1420175.
-\end{acks}
-
-\bibliography{references}
+\section*{Acknowledgement}
+Supported by NSF awards CCR-1906388 and 1420175.
+
+\begin{thebibliography}{key}
+%\bibliography{references}
+\end{thebibliography}
 
 \end{document}
-
-fixpoints at term level ala Pitts:
-
-I'm not sure if we actually want to do this or not (because not all
-fixpoints will necessarily exist, and this will have consequences, and
-because of the restriction to strict functions it will entail and the
-fact that I haven't yet thought about how those will thread through
-our calculus), but another idea we might consider is adding fixpoints
-at the term level. This is done for System F in Wadler's section
-7. [It is also done in Pitts' Parametric Polymorphism and Operational
-  Equivalence paper, where the term introduction rule is SEE SOURCE
-
-%Gamma \vdash F : \tau \to \tau
-%-------------------------------
-%Gamma \vdash fix (F) : tau
-
-But of course the model there is entirely operational rather than
-categorical.]
-
-On the other hand, check out the last paragraph of Wadler's paper, which
-suggests that (at least in 1989) there was some interest in calculi like
-ours:
-
-"The desire to derive theorems from types therefore suggests that it would be
-valuable to explore programming languages that prohibit recursion or only
-allow its restricted use. In theory, this is well understood; we have already
-noted that any computable function that is provably total in second-order
-Peano arithmetic can be defined in the pure polymorphic lambda calculus,
-without using the fixpoint as a primitive. However, practical languages based
-on this notion remain terra incognita."
-
-Of course we are not working in System F, but perhaps the same desire applies
-here, and we do get the same kinds of theorems as Wadler, many just as a
-consequence of our interpretation, without invoking parametricity.
-
-In any case, I guess Wadler could be cited as justifying restricting
-our attention to our calculus since it could be seen as a way to have
-a practical language that ensures that all programs are
-terminating. This has been considered in the past, e.g., in ADL,
-Charity, PolyP.
-
-TERMINATING RECURSION
-Another possible solution to the expressivity issue is extend folds to
-track of deep structure of nested types as in~\cite{jp19}, so that
-specialization of fixpoint can be computed as fixpoint of
-specialization.
-
-Add more polymorphism (all foralls), even though most free theorems
-only use one level (or maybe two, like short cut).
-
-%test

### 2020-12-11 free-theorems-nesteds/20-mscs.tex
more updates to mscs.tex

@@
 
-%% due to the dependence on amsart.cls, \begin{document} has to occur
-%% BEFORE the title and author information:
-
 \begin{document}
 
-\title[Parametricity for Primitive
-  Nested Types and GADTs]{Parametricity for Primitive
-  Nested Types and GADTs}
-\titlecomment{{\lsuper*}OPTIONAL comment concerning the title, \eg,
-  if a variant or an extended abstract of the paper has appeared elsewhere.}
-
-\author[P.~Johann]{Patricia Johann}	%required
-\address{Appalachian State University}	%required
-\email{johannp@appstate.edu}  %optional
-%\thanks{thanks 1, optional.}	%optional
-
-\author[E.~Ghiorzi]{Enrico Ghiorzi}	%optional
-\address{Appalachian State University}	%optional
-\email{ghiorzie@appstate.edu}  %optional
-%\thanks{thanks 2, optional.}	%optional
-
-\author[D.~Jeffries]{Daniel Jeffries}	%optional
-\address{Appalachian State University}	%optional
+\title[Parametricity for Primitive Nested Types and
+  GADTs]{Parametricity for Primitive Nested Types and GADTs}
+
+\author[P.~Johann]{Patricia Johann}	
+\address{Appalachian State University}	
+\email{johannp@appstate.edu}  
+\thanks{Supported by NSF awards CCR-1906388 and 1420175.}
+
+\author[E.~Ghiorzi]{Enrico Ghiorzi}	
+\address{Appalachian State University}	
+\email{ghiorzie@appstate.edu}
+
+\author[D.~Jeffries]{Daniel Jeffries}	
+\address{Appalachian State University}	
 \email{jeffriesd@appstate.edu}
-%\thanks{thanks 3, optional.}	%optional
 
@@
 illustrating how how our model can be used to derive useful free
-theorems for programs over nested types.
+theorems for programs over nested types. {\color{red} Add ainfo about
+  GADTs and implementation}
 \end{abstract}
@@
 
-We write $\vdash F$ for $\emptyset;\emptyset \vdash F$.
-Definition~\ref{def:wftypes} ensures that the expected weakening rules
-for well-formed types hold, although weakening does not change the
-contexts in which types can be formed. If $\Gamma;\emptyset \vdash F$
-and $\Gamma;\emptyset \vdash G$, then our rules allow formation of the
-types $\Gamma;\emptyset \vdash \Nat^\emptyset F \,G$ and $\Gamma;
-\emptyset \vdash \Nat^{\ol\alpha} \,\onet \,F$, which represent the
-System F types $\Gamma \vdash F \to G$ and $\forall$-type $\Gamma;
-\emptyset \vdash \forall \ol\alpha . F$, respectively. However, some
-System F types, such as $\forall \alpha. (\alpha \to \alpha) \to
-\alpha$, are not representable in our calculus.
+{\color{red} Note that we use first-order substitution in the
+  $\mu$-clause even though it isn't defined until the next
+  page. Mutually inductive definition? The purpose of the substitution
+  in the $\mu$-type clause of Def 2.1 is to ensure that variables in
+  $\Phi$ are not applied to types involving variables from $\ol\alpha$
+  in $F$; indeed, the first-order substitution $[\ol{\gamma := H}]$
+  constrains the types in $\ol H$ to have arity $0$, and constrains
+  the variables in $\Phi$ to already be fully applied to types not
+  involving $\ol\alpha$ in $\ol H$, making such applications
+  impossible. Of course, rather than intertwining the definitions of
+  well-formed types of our calculus and the operation of first-order
+  substitution on them we could instead prohibit such applications in
+  a side condition to the rule for well-formed $\mu$-types, but we
+  favor the precision of the current presentation.} We write $\vdash
+F$ for $\emptyset;\emptyset \vdash F$.  Definition~\ref{def:wftypes}
+ensures that the expected weakening rules for well-formed types hold,
+although weakening does not change the contexts in which types can be
+formed. If $\Gamma;\emptyset \vdash F$ and $\Gamma;\emptyset \vdash
+G$, then our rules allow formation of the types $\Gamma;\emptyset
+\vdash \Nat^\emptyset F \,G$ and $\Gamma; \emptyset \vdash
+\Nat^{\ol\alpha} \,\onet \,F$, which represent the System F types
+$\Gamma \vdash F \to G$ and $\forall$-type $\Gamma; \emptyset \vdash
+\forall \ol\alpha . F$, respectively. However, some System F types,
+such as $\forall \alpha. (\alpha \to \alpha) \to \alpha$, are not
+representable in our calculus.
 
@@
 \end{adjustbox}
-       \vspace*{-0.25in}
+%       \vspace*{-0.25in}
 \end{figure*}
@@
 
-{\color{red} Integerate next paragraph into above discussion?} Our
-calculus can, however, express types of recursive functions that take
-multiple nested types as arguments, provided those nested types are
-parameterized over disjoint sets of type variables and the return type
-of the function is parameterized over only the variables occurring in
-the type of its final argument.  Interestingly, even some recursive
-functions of a single proper nested type --- e.g., a
-$\mathtt{reverseBush}$ function that is a true involution --- cannot
-be expressed as folds because the algebra arguments needed to define
-them are again recursive functions
-%  cannot be defined. This happens, for
-%  example, when the algebra argument to a fold must be a recursive
-%  function
-%, and thus must itself be defined as a fold
-with types of the same problematic forms.
-%More sophisticated combinators for stylized recursion (e.g.,
-%generalized folds)
-%%  ~\cite{bp99})
-%don't mitigate these difficulties, either. Indeed, they appear to be a
-%fundamental consequence of the interaction of functoriality and
-%polymorphism, not just some peculiarity of our particular calculus.
-Allowing extra functorial variables in the codomains of $\Nat$-types
-would recover typeability of, say, zip for bushes, but such extended
-$\Nat$-types need not have semantics as $\omega$-cocontinuous functors
-in $\set$, as will be required in Section~\ref{sec:type-interp}. For
-example,
-%as shown in Example D, Section VI of~\cite{jp19},
-the type $\emptyset; \gamma \vdash \Nat^\emptyset\, (\mu \beta. \onet + \beta)\,
-\gamma$ has no such semantics. To remedy this, we would need to
-construct our models from semantic categories that are locally
-$\lambda$-presentable for $\lambda > \omega$.
-%It is tempting to try to recover typeability of, say, a zip function
-%for bushes by allowing extra functorial variables in the codomain type
-%of $\Nat$-types. But such extended $\Nat$-types need not have
-%semantics as $\omega$-cocontinuous functors, as is required in
-%Section~\ref{sec:type-interp}. For example, as shown in Example D,
-%Section VI of~\cite{jp19}, the type $\emptyset; \gamma \vdash
-%\Nat^\emptyset (\onet + \onet)\, \gamma$ does not.
-%
-
 \section{Interpreting Types}\label{sec:type-interp}
@@
 We interpret types in our calculus as $\omega$-cocontinuous functors
-on locally finitely presentable categories~\cite{ar94}. Since functor
+on locally finitely presentable categories~\cite{ar94}. The categories
+$\set$ and $\rel$ are both locally finitely presentable. Since functor
 categories of locally finitely presentable categories are again
@@
 $\omega$-cocontinuous functors from $\C$ to $\D$.
-%Note that the categories $\set$ and $\rel$ are both locally finitely
-%presentable.
 
@@
 \end{thm}
-{\color{red} INSERT ENTIRE PROOF. CHECK PROOF FOR $\mu$-TYPES AND
-  CHECK NOTATION.}
+\proof
 The proof is by induction on the structure of $F$. Only the $\Nat$,
-application, and fixpoint cases are non-routine. The latter two use
-Lemma~\ref{lem:eq-reln-equalities}. The fixpoint case also uses the
-observation that, for every $n \in \nat$, the following intermediate
-results can be proved by simultaneous induction with
-Theorem~\ref{thm:iel}: for any $H$, $\rho$, $A$, and any subformula
-$J$ of $H$, both
-%\begin{equation}\label{eq:iel-fix-point-intermediate1}
-$T^n_{H,\Eq_{\rho}} K_0\, \ol{\Eq_A}
-%{\setsem{\Gamma; \Phi \vdash F}\rho}}
-= (\Eq_{(T^\set_{H,\rho})^n K_0})^*
-\ol{\Eq_A}$\;
-  %{\setsem{\Gamma; \Phi \vdash F}\rho}}$\;
-%\end{equation}
-and\;
-%\begin{equation}\label{eq:iel-fix-point-intermediate2}
-%\begin{split}
-\[\begin{array}{ll}
-&  \relsem{\Gamma; \Phi, \phi, \ol{\alpha} \vdash J} \Eq_{\rho} [\phi
-  := T^{n}_{H,\Eq_{\rho}} K_0] \overline{[\alpha := \Eq_A]}\\
-%{\setsem{\Gamma; \Phi \vdash F}\rho}]}\\
-=  &
-\relsem{\Gamma; \Phi, \phi,
-  \ol{\alpha} \vdash J} \Eq_{\rho} [\phi := \Eq_{(T^\set_{H,\rho})^n K_0}]
-\overline{[\alpha := \Eq_A]}
-%         {\setsem{\Gamma; \Phi \vdash F}\rho}]}.
-\end{array}\]
-%\end{split}
-%\end{equation}
-hold. With these results in hand, the proof follows easily. It is
-given in detail in the accompanying anonymous supplementary
-material. As noted there, if functorial variables of arity greater
-than $0$ were allowed to appear in the bodies of $\mu$-types, then the
-IEL would fail.
-
-\begin{comment}
-\proof
-By induction on the structure of $\tau$.
+application, and fixpoint cases are non-routine.
 \begin{itemize}
-\item $\relsem{\Gamma; \emptyset \vdash v}\Eq_{\rho} = \Eq_{\rho} v =
-  \Eq_{\rho v} = \Eq_{\setsem{\Gamma; \emptyset \vdash v}\rho}$ where
-  $v \in \Gamma$.
-\item By definition, $\relsem{\Gamma; \emptyset \vdash
-  \Nat^{\overline\alpha} \,F\,G} \Eq_{\rho}$ is the relation on
-  $\setsem{\Gamma; \emptyset \vdash \Nat^{\overline\alpha} \,F\,G}
-  \rho$ relating $t$ and $t'$ if, for all ${R_1 :
-    \rel(A_1,B_1)},...,{R_k : \rel(A_k,B_k)}$, $(t_{\overline{A}},
-  t'_{\overline{B}})$ is a morphism from $\relsem{\Gamma;
-    \overline\alpha \vdash F} \Eq_{\rho}\overline{[\alpha := R]}$ to
-  $\relsem{\Gamma ; \overline\alpha \vdash G}
-  \Eq_{\rho}\overline{[\alpha := R]}$ in $\rel$.  To prove that this
-  is equal to $\Eq_{\setsem{\Gamma; \emptyset \vdash
-      \Nat^{\overline\alpha} \,F\,G} \rho}$ we need to show that
+\item $\relsem{\Gamma; \Phi \vdash \zerot} \Eq_{\rho} = 0_\rel =
+  \Eq_{0_\set} = \Eq_{\setsem{\Gamma; \Phi \vdash \zerot}\rho}$
+\item $\relsem{\Gamma; \Phi \vdash \onet} \Eq_{\rho} = 1_\rel =
+  \Eq_{1_\set} = \Eq_{\setsem{\Gamma; \Phi \vdash \onet}\rho}$
+\item {\color{red} Use vector notation?} By definition,
+  $\relsem{\Gamma; \emptyset \vdash \Nat^{\overline\alpha} \,F\,G}
+  \Eq_{\rho}$ is the relation on $\setsem{\Gamma; \emptyset \vdash
+    \Nat^{\overline\alpha} \,F\,G} \rho$ relating $t$ and $t'$ if, for
+  all ${R_1 : \rel(A_1,B_1)},...,{R_k : \rel(A_k,B_k)}$,
   $(t_{\overline{A}}, t'_{\overline{B}})$ is a morphism from
   $\relsem{\Gamma; \overline\alpha \vdash F}
+  \Eq_{\rho}\overline{[\alpha := R]}$ to $\relsem{\Gamma ;
+    \overline\alpha \vdash G} \Eq_{\rho}\overline{[\alpha := R]}$ in
+  $\rel$.  To prove that this relation is $\Eq_{\setsem{\Gamma;
+      \emptyset \vdash \Nat^{\overline\alpha} \,F\,G} \rho}$ we need
+  to show that $(t_{\overline{A}}, t'_{\overline{B}})$ is a morphism
+  from $\relsem{\Gamma; \overline\alpha \vdash F}
   \Eq_{\rho}\overline{[\alpha := R]}$ to $\relsem{\Gamma ;
@@
   \Eq_{\rho}\overline{[\alpha := R]}$ to $\relsem{\Gamma ;
-    \overline\alpha \vdash G} \Eq_{\rho}\overline{[\alpha := R]}$ in
-  $\rel$ for all ${R_1 : \rel(A_1,B_1),}$ $...,{R_k : \rel(A_k,B_k)}$,
-  then $t = t'$.  By hypothesis, $(t_{\overline{A}},
-  t'_{\overline{A}})$ is a morphism from $\relsem{\Gamma;
-    \overline\alpha \vdash F} \Eq_{\rho}\overline{[\alpha :=
-      \Eq_{A}]}$ to $\relsem{\Gamma ; \overline\alpha \vdash G}
-  \Eq_{\rho}\overline{[\alpha := \Eq_{A}]}$ in $\rel$ for all
-  $A_1\,...\,A_k : \set$. By the induction hypothesis, it is therefore
-  a morphism from $\Eq_{\setsem{\Gamma; \overline\alpha \vdash F}
-    \rho\overline{[\alpha := A]}}$ to $\Eq_{\setsem{\Gamma ;
+    \overline\alpha \vdash G} \Eq_{\rho}\overline{[\alpha := R]}$ for
+  all ${R_1 : \rel(A_1,B_1),}$ $...,{R_k : \rel(A_k,B_k)}$ then $t =
+  t'$.  By hypothesis, for all $A_1\,...\,A_k : \set$,
+  $(t_{\overline{A}}, t'_{\overline{A}})$ is a morphism from
+  $\relsem{\Gamma; \overline\alpha \vdash F}
+  \Eq_{\rho}\overline{[\alpha := \Eq_{A}]}$ to $\relsem{\Gamma ;
+    \overline\alpha \vdash G} \Eq_{\rho}\overline{[\alpha :=
+      \Eq_{A}]}$. By the induction hypothesis, it is therefore a
+  morphism from $\Eq_{\setsem{\Gamma; \overline\alpha \vdash F}
+    \rho\overline{[\alpha := A]}}$ to $\Eq_{\setsem{\Gamma;
       \overline\alpha \vdash G} \rho\overline{[\alpha := A]}}$ in
-  $\rel$. This means that, for every $x : \Eq_{\setsem{\Gamma;
+  $\rel$. This means that, for all $x : \Eq_{\setsem{\Gamma;
       \overline\alpha \vdash F} \rho\overline{[\alpha := A]}}$,
-  $t_{\overline{A}}x = t'_{\overline{A}}x$.  Then, by extensionality,
-  $t = t'$.
-\item $\relsem{\Gamma; \Phi \vdash \zerot} \Eq_{\rho} = 0_\rel =
-  \Eq_{0_\set} = \Eq_{\setsem{\Gamma; \Phi \vdash \zerot}\rho}$
-\item $\relsem{\Gamma; \Phi \vdash \onet} \Eq_{\rho} = 1_\rel =
-  \Eq_{1_\set} = \Eq_{\setsem{\Gamma; \Phi \vdash \onet}\rho}$
+  $t_{\overline{A}}x = t'_{\overline{A}}x$, so $t = t'$ by
+  extensionality.
 \item The application case is proved by the following sequence of
@@
   and the definition of the relation environment $\Eq_\rho$, the third
-  is by the definition of application of relation transformers, and
-  the fourth is by Lemma~\ref{lem:eq-reln-equalities}:
+  is by the definition of application of relation transformers from
+  Definition~9, and the fourth is by
+  Lemma~21:
 \[
 \begin{split}
-\relsem{\Gamma; \Phi \vdash \phi\ol{\tau}}\Eq_{\rho} &=
-(\Eq_{\rho}\phi)\ol{\relsem{\Gamma; \Phi \vdash \tau}
+\relsem{\Gamma; \Phi \vdash \phi\ol{F}}\Eq_{\rho} &=
+(\Eq_{\rho}\phi)\ol{\relsem{\Gamma; \Phi \vdash F}
 \Eq_{\rho}}\\
-&= \Eq_{\rho \phi}\, \ol{\Eq_{\setsem{\Gamma; \Phi \vdash \tau}
+&= \Eq_{\rho \phi}\, \ol{\Eq_{\setsem{\Gamma; \Phi \vdash F}
   \rho}}\\
-&= (\Eq_{\rho \phi})^* \,\ol{\Eq_{\setsem{\Gamma; \Phi \vdash \tau}
+&= (\Eq_{\rho \phi})^* \,\ol{\Eq_{\setsem{\Gamma; \Phi \vdash F}
   \rho}}\\
-&= \Eq_{(\rho \phi) \,\ol{\setsem{\Gamma; \Phi \vdash \tau} \rho}}\\
-&= \Eq_{\setsem{\Gamma; \Phi \vdash \phi\ol{\tau}}\rho}
+&= \Eq_{(\rho \phi) \,\ol{\setsem{\Gamma; \Phi \vdash F} \rho}}\\
+&= \Eq_{\setsem{\Gamma; \Phi \vdash \phi\ol{F}}\rho}
 \end{split}
@@
 \begin{split}
-\relsem{\Gamma; \Phi \vdash (\mu \phi.\lambda
-  \ol{\alpha}. H)\ol{\tau}}\Eq_{\rho} 
-&=(\mu {T_{\Eq_{\rho}}}) \,\ol{\relsem{\Gamma; \Phi \vdash \tau}\Eq_{\rho}}\\ 
-&= \colim{n \in \nat}{T^n_{\Eq_{\rho}} K_0}\, \ol{\relsem{\Gamma; \Phi
-  \vdash \tau}\Eq_{\rho}}\\
-&= \colim{n \in \nat}{ T^n_{\Eq_{\rho}} K_0 \,\ol{\Eq_{\setsem{\Gamma;
-    \Phi \vdash \tau}\rho}}}\\
-&= \colim{n \in \nat}{(\Eq_{(T^\set_\rho)^n K_0})^*
-  \ol{\Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}}}\\
-&= \colim{n \in \nat}{\Eq_{(T^\set_\rho)^n K_0 \,\ol{\setsem{\Gamma;
-        \Phi \vdash \tau}\rho}}}\\ 
-&= \Eq_{\colim{n \in \nat}{ (T^\set_{\rho})^n K_0\,
-    \ol{\setsem{\Gamma; \Phi \vdash \tau}\rho}}}\\
-&= \Eq_{\setsem{\Gamma; \Phi \vdash (\mu \phi.\lambda
-      \ol{\alpha}. H)\ol{\tau}}\rho}
+\relsem{\Gamma; \Phi, \ol\gamma \vdash (\mu \phi.\lambda
+  \ol{\alpha}. H)\ol{F}}\Eq_{\rho} 
+&=(\mu {T_{H, \Eq_{\rho}}}) \,\ol{\relsem{\Gamma; \Phi, \ol\gamma  \vdash F}\Eq_{\rho}}\\ 
+&= \colim{n \in \nat}{T^n_{H, \Eq_{\rho}} K_0}\, \ol{\relsem{\Gamma; \Phi, \ol\gamma 
+  \vdash F}\Eq_{\rho}}\\
+&= \colim{n \in \nat}{ T^n_{H, \Eq_{\rho}} K_0 \,\ol{\Eq_{\setsem{\Gamma;
+    \Phi, \ol\gamma  \vdash F}\rho}}}\\
+&= \colim{n \in \nat}{(\Eq_{(T^\set_{H,\rho})^n K_0})^*
+  \ol{\Eq_{\setsem{\Gamma; \Phi, \ol\gamma  \vdash F}\rho}}}\\
+&= \colim{n \in \nat}{\Eq_{(T^\set_{H,\rho})^n K_0 \,\ol{\setsem{\Gamma;
+        \Phi, \ol\gamma  \vdash F}\rho}}}\\ 
+&= \Eq_{\colim{n \in \nat}{ (T^\set_{H,\rho})^n K_0\,
+    \ol{\setsem{\Gamma; \Phi, \ol\gamma  \vdash F}\rho}}}\\
+&= \Eq_{\setsem{\Gamma; \Phi, \ol\gamma  \vdash (\mu \phi.\lambda
+      \ol{\alpha}. H)\ol{F}}\rho}
 \end{split}
@@
 Here, the third equality is by induction hypothesis, the fifth is by
-Lemma~\ref{lem:eq-reln-equalities} and the fourth equality is because,
-for every $n \in \nat$, the following two statements can be proved by
-simultaneous induction:
+Lemma~21 and the fourth equality is because, for every $n \in \nat$,
+the following two statements can be proved by simultaneous induction:
+and for any $H$, $\rho$, $A$, and subformula $J$ of $H$,
 \begin{equation}\label{eq:iel-fix-point-intermediate1}
-T^n_{\Eq_{\rho}} K_0\, \ol{\Eq_{\setsem{\Gamma; \Phi \vdash
-      \tau}\rho}} = (\Eq_{(T^\set_\rho)^n K_0})^*
-\ol{\Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}}
+T^n_{H, \Eq_{\rho}} K_0\, \ol{\Eq_A} = (\Eq_{(T^\set_{H, \rho})^n K_0})^*
+\ol{\Eq_A}
 \end{equation}
@@
 \begin{split}
-  \relsem{\Gamma; \Phi, \phi, \ol{\alpha} \vdash H}
-\Eq_{\rho} [\phi := 
- & T^{n}_{\Eq_{\rho}} K_0] \overline{[\alpha :=
-    \Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}]} \\
-=\;\; & \relsem{\Gamma; \Phi, \phi, \ol{\alpha} \vdash H} \Eq_{\rho} [\phi
-  := \Eq_{(T^\set_\rho)^n K_0}] \overline{[\alpha :=
-    \Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}]}
+& \relsem{\Gamma; \Phi', \phi, \ol{\alpha} \vdash J} \Eq_{\rho} [\phi
+    := T^{n}_{H,\Eq_{\rho}} K_0] \overline{[\alpha := \Eq_A]} \\
+=\;\; & \relsem{\Gamma; \Phi', \phi, \ol{\alpha} \vdash J} \Eq_{\rho}
+[\phi := \Eq_{(T^\set_{H,\rho})^n K_0}] \overline{[\alpha := \Eq_A]}
 \end{split}
 \end{equation}
-We prove~\eqref{eq:iel-fix-point-intermediate1}.  The case $n=0$ is
-trivial, because $T^0_{\Eq_{\rho}} K_0 = K_0$ and
-$(T^\set_\rho)^0 K_0 = K_0$; the inductive step is
-proved by the following sequence of equalities:
+(Notice that we don't know what's in the context $\Phi'$.)
+We prove~\eqref{eq:iel-fix-point-intermediate1} by induction on $n$.
+The case $n=0$ is trivial, because $T^0_{H,\Eq_{\rho}} K_0 = K_0$ and
+$(T^\set_{H,\rho})^0 K_0 = K_0$; the inductive step is proved by the
+following sequence of equalities:
 \[
 \begin{split}
-T^{n+1}_{\Eq_{\rho}} K_0\, \overline{\Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}}
-&= T^\rel_{\Eq_{\rho}} (T^{n}_{\Eq_{\rho}} K_0)
-\overline{\Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}} \\ 
-&= \relsem{\Gamma; \Phi, \phi, \ol{\alpha} \vdash H} \Eq_{\rho} [\phi
-  := T^{n}_{\Eq_{\rho}} K_0] \overline{[\alpha :=
-    \Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}]} \\ 
-&= \relsem{\Gamma; \Phi, \phi, \ol{\alpha} \vdash H} \Eq_{\rho} [\phi
-  := \Eq_{(T^\set_\rho)^n K_0}] \overline{[\alpha :=
-    \Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}]} \\ 
-&= \relsem{\Gamma; \Phi, \phi, \ol{\alpha} \vdash H} \Eq_{\rho [\phi
-    := (T^\set_\rho)^n K_0] \overline{[\alpha :=
-      \setsem{\Gamma; \Phi \vdash \tau}\rho]}} \\ 
-&= \Eq_{\setsem{\Gamma; \Phi, \phi, \ol{\alpha} \vdash H} \rho [\phi
-    := (T^\set_\rho)^n K_0] \overline{[\alpha :=
-      \setsem{\Gamma; \Phi \vdash \tau}\rho]}} \\ 
-&= \Eq_{(T^\set_\rho)^{n+1} K_0 \overline{\setsem{\Gamma; \Phi
-      \vdash \tau}\rho}} \\ 
-&= (\Eq_{(T^\set_\rho)^{n+1} K_0})^*\, \overline{\Eq_{\setsem{\Gamma;
-      \Phi \vdash \tau}\rho}} 
+T^{n+1}_{H,\Eq_{\rho}} K_0\, \overline{\Eq_A}
+&= T^\rel_{H,\Eq_{\rho}} (T^{n}_{H,\Eq_{\rho}} K_0)
+\overline{\Eq_A} \\ 
+&= \relsem{\Gamma; \phi, \ol{\alpha}, \ol\gamma \vdash H} \Eq_{\rho} [\phi
+  := T^{n}_{H, \Eq_{\rho}} K_0] \overline{[\alpha :=
+    \Eq_A]} \\ 
+&= \relsem{\Gamma; \phi, \ol{\alpha}, \ol\gamma \vdash H} \Eq_{\rho} [\phi
+  := \Eq_{(T^\set_{H,\rho})^n K_0}] \overline{[\alpha :=
+    \Eq_A]} \\ 
+&= \relsem{\Gamma; \phi, \ol{\alpha}, \ol\gamma \vdash H} \Eq_{\rho [\phi
+    := (T^\set_{H,\rho})^n K_0] \overline{[\alpha :=
+      A]}} \\ 
+&= \Eq_{\setsem{\Gamma; \phi, \ol{\alpha}, \ol\gamma \vdash H} \rho [\phi
+    := (T^\set_{H,\rho})^n K_0] \overline{[\alpha :=
+      A]}} \\ 
+&= \Eq_{(T^\set_{H,\rho})^{n+1} K_0 \overline{A}} \\ 
+&= (\Eq_{(T^\set_{H,\rho})^{n+1} K_0})^*\, \overline{\Eq_A} 
 \end{split}
 \]
-Here, the third equality is by~\eqref{eq:iel-fix-point-intermediate2},
-the fifth by the induction hypothesis on $H$, and the last by
-Lemma~\ref{lem:eq-reln-equalities}.  We prove the induction step
-of~\eqref{eq:iel-fix-point-intermediate2} by structural induction on
-$H$: the only interesting case, though, is when $\phi$ is applied,
-i.e., for $H = \phi \ol{\sigma}$, which is proved by the sequence of
-equalities:
+Here, the third equality is by~\eqref{eq:iel-fix-point-intermediate2}
+for $J = H$, the fifth by the induction hypothesis of the IEL on $H$,
+and the last is by Lemma~21.
+
+\vspace*{0.1in}
+
+We prove~\eqref{eq:iel-fix-point-intermediate2} by structural
+induction on $J$. The only interesting cases, though, are when
+$J = \phi \ol{G}$ and when $J = (\mu \psi.\lambda \ol\beta. G)\ol K$.
+\begin{itemize}
+\item The case $J = \phi \ol G$ is proved by the sequence of equalities:
 \[
 \begin{split}
-& \relsem{\Gamma; \Phi, \phi, \ol{\alpha} \vdash \phi
-    \ol{\sigma}} \Eq_{\rho} [\phi := T^{n}_{\Eq_{\rho}} K_0]
-  \overline{[\alpha := \Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}]}
+& \relsem{\Gamma; \Phi', \phi, \ol{\alpha} \vdash \phi
+    \ol{G}} \Eq_{\rho} [\phi := T^{n}_{H,\Eq_{\rho}} K_0]
+  \overline{[\alpha := \Eq_A]}
   \\
-&= T^{n}_{\Eq_{\rho}} K_0\, \overline{\relsem{\Gamma; \Phi,
-      \phi, \ol{\alpha} \vdash \sigma} \Eq_{\rho} [\phi :=
-      T^{n}_{\Eq_{\rho}} K_0] \overline{[\alpha :=
-        \Eq_{\setsem{\Gamma; \Phi \vdash \tau} \rho}]}} \\ 
-&= T^{n}_{\Eq_{\rho}} K_0\, \overline{\relsem{\Gamma; \Phi,
-      \phi, \ol{\alpha} \vdash \sigma} \Eq_{\rho} [\phi :=
-      \Eq_{(T^\set_\rho)^{n} K_0}] \overline{[\alpha :=
-        \Eq_{\setsem{\Gamma; \Phi \vdash \tau} \rho}]}} \\ 
-&= T^{n}_{\Eq_{\rho}} K_0\, \overline{\relsem{\Gamma; \Phi,
-      \phi, \ol{\alpha} \vdash \sigma} \Eq_{\rho [\phi := (T^\set_\rho)^{n}
-        K_0] \overline{[\alpha := \setsem{\Gamma; \Phi \vdash
-            \tau} \rho]}}} \\ 
-&= T^{n}_{\Eq_{\rho}} K_0\, \overline{\Eq_{\setsem{\Gamma;
-        \Phi, \phi, \ol{\alpha} \vdash \sigma} \rho [\phi :=
-        (T^\set_\rho)^{n} K_0] \overline{[\alpha :=
-          \setsem{\Gamma; \Phi \vdash \tau} \rho]}}} \\ 
-&= (\Eq_{(T^\set_\rho)^{n} K_0})^* \,\overline{\Eq_{\setsem{\Gamma;
-        \Phi, \phi, \ol{\alpha} \vdash \sigma} \rho [\phi :=
-        (T^\set_\rho)^{n} K_0] \overline{[\alpha :=
-          \setsem{\Gamma; \Phi \vdash \tau} \rho]}}} \\ 
-&= (\Eq_{(T^\set_{\rho})^{n} K_0})^* \overline{\relsem{\Gamma;
-      \Phi, \phi, \ol{\alpha} \vdash \sigma} \Eq_{\rho} [\phi :=
-      \Eq_{(T^\set_{\rho})^{n} K_0}] \overline{[\alpha :=
-        \Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}]}} \\ 
-&= \relsem{\Gamma; \Phi, \phi, \ol{\alpha} \vdash \phi \ol{\sigma}}
-  \Eq_{\rho} [\phi := \Eq_{(T^\set_{\rho})^{n} K_0}]
-  \overline{[\alpha := \Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}]} 
+&= T^{n}_{H,\Eq_{\rho}} K_0\, \overline{\relsem{\Gamma; \Phi',
+      \phi, \ol{\alpha} \vdash G} \Eq_{\rho} [\phi :=
+      T^{n}_{H,\Eq_{\rho}} K_0] \overline{[\alpha :=
+        \Eq_A]}} \\ 
+&= T^{n}_{H,\Eq_{\rho}} K_0\, \overline{\relsem{\Gamma; \Phi',
+      \phi, \ol{\alpha} \vdash G} \Eq_{\rho} [\phi :=
+      \Eq_{(T^\set_{H,\rho})^{n} K_0}] \overline{[\alpha :=
+        \Eq_A]}} \\ 
+&= T^{n}_{H,\Eq_{\rho}} K_0\, \overline{\relsem{\Gamma; \Phi',
+      \phi, \ol{\alpha} \vdash G} \Eq_{\rho [\phi := (T^\set_{H,\rho})^{n}
+        K_0] \overline{[\alpha := A]}}} \\ 
+&= T^{n}_{H,\Eq_{\rho}} K_0\, \overline{\Eq_{\setsem{\Gamma;
+        \Phi', \phi, \ol{\alpha} \vdash G} \rho [\phi :=
+        (T^\set_{H,\rho})^{n} K_0] \overline{[\alpha :=
+          A]}}} \\ 
+&= (\Eq_{(T^\set_{H,\rho})^{n} K_0})^* \,\overline{\Eq_{\setsem{\Gamma;
+        \Phi', \phi, \ol{\alpha} \vdash G} \rho [\phi :=
+        (T^\set_{H,\rho})^{n} K_0] \overline{[\alpha :=
+          A]}}} \\ 
+&= (\Eq_{(T^\set_{H,\rho})^{n} K_0})^* \overline{\relsem{\Gamma;
+      \Phi', \phi, \ol{\alpha} \vdash G} \Eq_{\rho} [\phi :=
+      \Eq_{(T^\set_{H,\rho})^{n} K_0}] \overline{[\alpha :=
+        \Eq_A]}} \\ 
+&= \relsem{\Gamma; \Phi', \phi, \ol{\alpha} \vdash \phi \ol{G}}
+  \Eq_{\rho} [\phi := \Eq_{(T^\set_{H,\rho})^{n} K_0}]
+  \overline{[\alpha := \Eq_A]} 
 \end{split}
@@
 Here, the second equality is by the induction hypothesis
-for~\eqref{eq:iel-fix-point-intermediate2} on the $\sigma$s, the
-fourth is by the induction hypothesis for Theorem~\ref{thm:iel} on the
-$\sigma$s, and the fifth is by the induction hypothesis on $n$
+for~\eqref{eq:iel-fix-point-intermediate2} on the $G$s, the fourth is
+by the induction hypothesis for the IEL on the $G$s, and the fifth is
+by the induction hypothesis on $n$
 for~\eqref{eq:iel-fix-point-intermediate1}.
-\item $\relsem{\Gamma; \Phi \vdash \sigma + \tau} \Eq_{\rho} =
-  \relsem{\Gamma; \Phi \vdash \sigma} \Eq_{\rho} + \relsem{\Gamma;
-    \Phi \vdash \tau} \Eq_{\rho} = \Eq_{\setsem{\Gamma; \Phi \vdash
-      \sigma}\rho} + \Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho} =
-  \Eq_{\setsem{\Gamma; \Phi \vdash \sigma}\rho + \setsem{\Gamma; \Phi
-      \vdash \tau}\rho} = \Eq_{\setsem{\Gamma; \Phi \vdash \sigma +
-      \tau}\rho}$
-\item $\relsem{\Gamma; \Phi \vdash \sigma \times \tau} \Eq_{\rho} =
-  \relsem{\Gamma; \Phi \vdash \sigma}\Eq_{\rho} \times \relsem{\Gamma;
-    \Phi \vdash \tau}\Eq_{\rho} = \Eq_{\setsem{\Gamma; \Phi \vdash
-      \sigma}\rho} \times \Eq_{\setsem{\Gamma; \Phi \vdash \tau}\rho}
-  = \Eq_{\setsem{\Gamma; \Phi \vdash \sigma}\rho \times
-    \setsem{\Gamma; \Phi \vdash \tau}\rho} = \Eq_{\setsem{\Gamma;
-      \Phi \vdash \sigma \times \tau}\rho}$
+
+\item 
+  The case $J = (\mu \psi.\, \lambda \ol{\beta}.\, G) \ol{K}$
+is proved by the sequence of equalities (where $\Phi' = \Phi'', \ol\gamma$):
+\[
+\begin{split}
+& \relsem{\Gamma; \Phi'', \ol\gamma, \phi, \ol{\alpha} \vdash (\mu \psi.\, \lambda \ol{\beta}.\, G) \ol{K}}
+  \Eq_{\rho} [\phi := T^{n}_{H,\Eq_{\rho}} K_0]
+  [\overline{\alpha := \Eq_A}] \\
+&= (\mu {T_{G, \Eq_{\rho}[\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]}})
+  \,\ol{\relsem{\Gamma; \Phi'', \ol\gamma, \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
+  [\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]} \\ 
+&= \colim{m \in \nat} {T^m_{G, \Eq_{\rho}[\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]}}\,K_0
+  \,(\ol{\relsem{\Gamma; \Phi'', \ol\gamma, \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
+  [\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]}) \\ 
+&= \colim{m \in \nat} {T^m_{G, \Eq_{\rho}[\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]}}\,K_0
+  \,(\ol{\relsem{\Gamma; \Phi'', \ol\gamma, \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
+  [\phi := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha := \Eq_A}]}) \\  %%
+%% &= \colim{m \in \nat} {T^m_{G, \Eq_{\rho[\overline{\alpha := A}]}}}\,K_0
+%%   \,(\ol{\relsem{\Gamma; \Phi, \ol\gamma, \phi, \ol{\alpha} \vdash K}
+%%   \Eq_{\rho[\phi := (T^{\set}_{H,\rho})^n K_0][\overline{\alpha := A}]}}) \\ 
+%% &= \colim{m \in \nat} {T^m_{G, \Eq_{\rho[\overline{\alpha := A}]}}}\,K_0
+%%   \,\ol{\Eq_{\setsem{\Gamma; \Phi, \ol\gamma, \phi, \ol{\alpha} \vdash K}
+%%   \rho[\phi := (T^{\set}_{H,\rho})^n K_0][\overline{\alpha := A}]}} \\ 
+%% &= \colim{m \in \nat} (\Eq_{{(T^{\set}_{G, \rho[\overline{\alpha := A}]}})^m\,K_0})^*
+%%   \,\ol{\Eq_{\setsem{\Gamma; \Phi, \ol\gamma, \phi, \ol{\alpha} \vdash K}
+%%   \rho[\phi := (T^{\set}_{H,\rho})^n K_0][\overline{\alpha := A}]}} \\ 
+%% &= (\Eq_{\colim{m \in \nat} {(T^{\set}_{G, \rho[\overline{\alpha := A}]}})^m\,K_0})^*
+%%   \,\ol{\Eq_{\setsem{\Gamma; \Phi, \ol\gamma, \phi, \ol{\alpha} \vdash K}
+%%   \rho[\phi := (T^{\set}_{H,\rho})^n K_0][\overline{\alpha := A}]}} \\ 
+%% &= (\Eq_{{\mu T^{\set}_{G, \rho[\overline{\alpha := A}]}}})^*
+%%   \,\ol{\Eq_{\setsem{\Gamma; \Phi, \ol\gamma, \phi, \ol{\alpha} \vdash K}
+%%   \rho[\phi := (T^{\set}_{H,\rho})^n K_0][\overline{\alpha := A}]}} \\
+%% &= \Eq_{{\mu T^{\set}_{G, \rho[\overline{\alpha := A}]}}
+%%   \,\ol{\setsem{\Gamma; \Phi, \ol\gamma, \phi, \ol{\alpha} \vdash K}
+%%   \rho[\phi := (T^{\set}_{H,\rho})^n K_0][\overline{\alpha := A}]}} \\
+%=& \Eq_{\setsem{\Gamma; \Phi, \ol\gamma, \ol{\alpha}, \ol{\gamma} \vdash (\mu \psi.\, \lambda \beta.\, G) \ol{\gamma}}
+%  \rho[\overline{\alpha := A}][\ol{\gamma := \setsem{\Gamma; \Phi, \ol\gamma, \phi, \ol{\alpha} \vdash K}
+%  \rho[\phi := (T^{\set}_{H,\rho})^n K_0][\overline{\alpha := A}]}]} \\
+&= \colim{m \in \nat} {T^m_{G, \Eq_{\rho}[\phi := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha := \Eq_A}]}}\,K_0
+  \,(\ol{\relsem{\Gamma; \Phi'', \ol\gamma, \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
+  [\phi := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha := \Eq_A}]}) \\ 
+&= (\mu {T_{G, \Eq_{\rho}[\phi := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha := \Eq_A}]}})
+  \,\ol{\relsem{\Gamma; \Phi'', \ol\gamma, \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
+  [\phi := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha := \Eq_A}]} \\ 
+&= \relsem{\Gamma; \Phi'', \ol\gamma, \phi, \ol{\alpha} \vdash (\mu \psi.\, \lambda \beta.\, G) \ol{K}}
+  \Eq_{\rho}[\phi := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha := \Eq_{A}}]
+\end{split}
+\]
+Here, the third equality is by the induction hypothesis
+for~\eqref{eq:iel-fix-point-intermediate2} on the $K$s, and the fourth
+equality holds because we can prove that, for all $m \in \nat$,
+\begin{equation}\label{eq:helper}
+T^m_{G,
+\Eq_{\rho}[\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha :=
+    \Eq_A}]}\,K_0 \ = T^m_{G, \Eq_{\rho}[\phi :=
+    \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha :=
+      \Eq_A}]}\,K_0
+\end{equation}
+Indeed, the base case of~\eqref{eq:helper} is trivial because
+\[T^0_{G,
+  \Eq_{\rho}[\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha :=
+      \Eq_A}]}\,K_0 \ = K_0 = T^0_{G, \Eq_{\rho}[\phi :=
+    \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha :=
+      \Eq_A}]}\,K_0\]
+and the inductive case is proved by: \\
+\begin{align*}
+& T^{m+1}_{G, \Eq_{\rho}[\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]}\,K_0 \\
+&= T_{G, \Eq_{\rho}[\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha
+        := \Eq_A}]} (T^{m}_{G, \Eq_{\rho}[\phi := T^{n}_{H,\Eq_{\rho}}
+      K_0][\overline{\alpha := \Eq_A}]}\,K_0) \\
+&= T_{G, \Eq_{\rho}[\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha
+        := \Eq_A}]} (T^{m}_{G, \Eq_{\rho}[\phi :=
+      \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha :=
+        \Eq_A}]}\,K_0) \\
+&= \lambda \ol{R}. \relsem{\Gamma; \ol\gamma, \phi, \ol\alpha, \psi,
+    \ol\beta \vdash G} \Eq_{\rho} [\phi := T^{n}_{H,\Eq_{\rho}}
+    K_0][\overline{\alpha := \Eq_A}][\psi := T^{m}_{G, \Eq_{\rho}[\phi
+        := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha :=
+          \Eq_A}]}\,K_0][\overline{\beta := R}] \\
+&= \lambda \ol{R}. \relsem{\Gamma; \ol\gamma, \phi, \ol\alpha, \psi,
+    \ol\beta \vdash G} \Eq_{\rho} [\phi := \Eq_{(T^{\set}_{H,\rho})^n
+      K_0}][\overline{\alpha := \Eq_A}][\psi := T^{m}_{G,
+      \Eq_{\rho}[\phi := \Eq_{(T^{\set}_{H,\rho})^n
+          K_0}][\overline{\alpha := \Eq_A}]}\,K_0][\overline{\beta :=
+      R}] \\
+&= T_{G, \Eq_{\rho}[\phi := \Eq_{(T^{\set}_{H,\rho})^n
+        K_0}][\overline{\alpha := \Eq_A}]} (T^{m}_{G, \Eq_{\rho}[\phi
+      := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha :=
+        \Eq_A}]}\,K_0) \\
+&= T^{m+1}_{G, \Eq_{\rho}[\phi := \Eq_{(T^{\set}_{H,\rho})^n
+        K_0}][\overline{\alpha := \Eq_A}]}\,K_0
+\end{align*}
+
+Here, the second equality holds by the induction hypothesis
+for~\eqref{eq:helper} on $m$. The fourth equality holds because, due
+to typing rule restrictions for the $\mu$ types, $\phi$ either does
+not appear in $G$, or must have arity $0$, in which case $\ol\alpha$
+must be empty, if $\phi$ appears in $G$, and
+uses~\eqref{eq:iel-fix-point-intermediate2} for $G$ when $\phi$ has
+arity $0$. {\color{red} Note that if functorial variables of arity
+  greater than $0$ were allowed to appear in the bodies of
+  $\mu$-types, then the aboe clause of the IEL would fail.}
+\end{itemize}
+\item $\relsem{\Gamma; \Phi \vdash F + G} \Eq_{\rho} =
+  \relsem{\Gamma; \Phi \vdash F} \Eq_{\rho} + \relsem{\Gamma;
+    \Phi \vdash G} \Eq_{\rho} = \Eq_{\setsem{\Gamma; \Phi \vdash
+      F}\rho} + \Eq_{\setsem{\Gamma; \Phi \vdash G}\rho} =
+  \Eq_{\setsem{\Gamma; \Phi \vdash F}\rho + \setsem{\Gamma; \Phi
+      \vdash G}\rho} = \Eq_{\setsem{\Gamma; \Phi \vdash F +
+      G}\rho}$
+\item $\relsem{\Gamma; \Phi \vdash F \times G} \Eq_{\rho} =
+  \relsem{\Gamma; \Phi \vdash F}\Eq_{\rho} \times \relsem{\Gamma;
+    \Phi \vdash G}\Eq_{\rho} = \Eq_{\setsem{\Gamma; \Phi \vdash
+      F}\rho} \times \Eq_{\setsem{\Gamma; \Phi \vdash G}\rho}
+  = \Eq_{\setsem{\Gamma; \Phi \vdash F}\rho \times
+    \setsem{\Gamma; \Phi \vdash G}\rho} = \Eq_{\setsem{\Gamma;
+      \Phi \vdash F \times G}\rho}$
 \end{itemize}
 \qed
-\end{comment}
 
@@
 
-\pagebreak
-
-\noindent
-
-\section*{Acknowledgement}
-Supported by NSF awards CCR-1906388 and 1420175.
-
-\begin{thebibliography}{key}
+\begin{thebibliography}{9}
 %\bibliography{references}

### 2020-12-13 free-theorems-nesteds/20-mscs.tex
add biblio and new mu rule

@@
 illustrating how how our model can be used to derive useful free
-theorems for programs over nested types. {\color{red} Add ainfo about
-  GADTs and implementation}
+theorems for programs over nested types. {\color{red} Add info about
+  GADTs and implementation.}
 \end{abstract}
@@
 
-  Send to LMCS? Editor?
-
-Can bind variables of arity > 0 in Nat-types? Improves expressibilty?
-Worth it?}
+Can bind variables of arity greater than 0 in Nat-types? Improves
+expressibility?  Worth it?}
 
@@
 \\[3ex]
-\AXC{$\Gamma;\ol{\gamma^0},\ol{\alpha^0},\phi^k \vdash F$}
+\AXC{$\Gamma;{\color{red}\ol{\gamma^0}},\ol{\alpha^0},\phi^k \vdash F$}
 \AXC{$\quad\quad\ol{\Gamma;\Phi \vdash G}$}
@@
 \TIC{$\Gamma;\Phi \vdash (\mu \phi^k.\lambda
-  \ol{\alpha^0}. \,F[\ol{\gamma := H}])\,\ol{G}$}  
+  \ol{\alpha^0}. \,F {\color{red} [\ol{\gamma := H}]})\,\ol{G}$}  
 \DisplayProof
@@
 
+{\color{red} CHANGE TEXT FOR NEW $\mu$-RULE.}
 Definition~\ref{def:wftypes} allows the formation of all of the
-(closed) nested types from the introduction:
+(closed {\color{red} NOT!}) nested types from the introduction:
 \[\begin{array}{lll}
@@
 \end{array}\]
-Each of these types can be considered either functorial in $\alpha$ or
-not, according to whether $\alpha$ is in the functorial or
-non-functorial context in which it is formed. In particular, if
-$\emptyset;\alpha \vdash \mathit{List} \,\alpha$, then $\vdash
-\Nat^\alpha \onet\, (\mathit{List}\,\alpha)$ is well-formed; if
-$\alpha;\emptyset \vdash \List \,\alpha$, then it is not.  Similarly,
-if $\mathit{Tree}\,\alpha\,\gamma = \mu \,\beta.\, \alpha + \beta
-\times \gamma \times \beta$, then $\gamma;\emptyset \vdash \Nat^\alpha
+{\color{red} The representation of $\mathit{List}\,\alpha$ as the ADT
+  $\mu \beta. \onet + \alpha \times \beta$ cannot be functorial in
+  $\alpha$. By contrast, the representation of $\mathit{List}\,\alpha$
+  as the nested type $(\mu \phi. \lambda \beta.\,\onet + \beta \times
+  \phi \beta)\,\alpha$
+%Each of these types
+  can be considered either functorial in $\alpha$ or not, according to
+  whether $\alpha$ is in the functorial or non-functorial context in
+  which this type is formed.} In particular, if $\emptyset;\alpha
+\vdash \mathit{List} \,\alpha$, then $\vdash \Nat^\alpha \onet\,
+(\mathit{List}\,\alpha)$ is well-formed; if $\alpha;\emptyset \vdash
+\List \,\alpha$, then it is not.  Similarly, if
+$\mathit{Tree}\,\alpha\,\gamma = \mu \,\beta.\, \alpha + \beta \times
+\gamma \times \beta$, then $\gamma;\emptyset \vdash \Nat^\alpha
 (\mathit{List}\, \alpha) \, (\mathit{Tree} \, \alpha\, \gamma)$ is
 well-formed, but $\emptyset;\gamma \vdash \Nat^\alpha (\mathit{List}\,
-\alpha) \, (\mathit{Tree} \, \alpha\, \gamma)$ is not. Although types
-can be functorial in variables of arity greater than $0$, the body $F$
-of a type $(\mu \phi.\lambda \ol\alpha.F)\ol G$ cannot. For example,
-the type $\mathit{GRose}\, \phi \, \alpha = \mu\beta . \alpha \times
-\phi \beta$ can be functorial or not in $\alpha$, but cannot be
-functorial in $\phi$. Moreover, $\mathit{GRose}\,\phi\,\alpha$ cannot
-be the (co)domain of a $\Nat$-type representing a natural
-transformation functorial in $\phi$, and the type $\mu \phi. \lambda
+\alpha) \, (\mathit{Tree} \, \alpha\, \gamma)$ is not. {\color{red}
+  KEEP? Although types can be functorial in variables of arity greater
+  than $0$, the body $F$ of a type $(\mu \phi.\lambda \ol\alpha.F)\ol
+  G$ cannot.} For example, the type $\mathit{GRose}\, \phi \, \alpha =
+\mu\beta . \alpha \times \phi \beta$ {\color{red} can be functorial
+  --- NOT!} or not in $\alpha$, but cannot be functorial in
+$\phi$. Moreover, $\mathit{GRose}\,\phi\,\alpha$ cannot be the
+(co)domain of a $\Nat$-type representing a natural transformation
+functorial in $\phi$, and the type $\mu \phi. \lambda
 \alpha. \mathit{GRose}\,\phi\,\alpha$ is not well-formed in our
@@
 interpretations of types in Section~\ref{sec:type-interp} are
-functors. In Section~\ref{sec:conclusion} we discuss how
-more expressive types might be accommodated.
-
-Definition~\ref{def:wftypes} allows well-formed types to be 
-functorial in no variables
-Functorial variables
+functors. In Section~\ref{sec:conclusion} we discuss how more
+expressive types might be accommodated.
+
+Definition~\ref{def:wftypes} allows well-formed types to be functorial
+in no variables. Functorial variables
 %in a well-formed type $\tau$
@@
 
-{\color{red} Check no change needed for new $\mu$-rule in next def.}
+{\color{red} Check no change needed for new $\mu$-rule in next
+  def. Just note that in application case the args to $\phi$ go in for
+  the $\alpha$s in $F$s replacing $\phi$.}
 \begin{defi}\label{def:second-order-subst}
@@
        \\\\
-       \AXC{$\Gamma; \phi, \ol{\alpha},\ol{\gamma} \vdash H$}
+       \AXC{$\Gamma; \phi, \ol{\alpha}, {\color{red}\ol{\gamma}} \vdash H$}
        \UIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \tin_H :
-         \Nat^{\ol{\beta},\ol{\gamma}} H[\phi :=_{\ol{\beta}} (\mu
+         \Nat^{\ol{\beta},{\color{red}\ol{\gamma}}} H[\phi :=_{\ol{\beta}} (\mu
            \phi.\lambda \ol{\alpha}.H)\ol{\beta}][\ol{\alpha := \beta}]\,(\mu
@@
        \\\\
-       \AXC{$\Gamma; \phi,\ol{\alpha}, \ol{\gamma} \vdash H$}
-       \AXC{$\Gamma; \ol{\beta}, \ol{\gamma} \vdash F$}
+       \AXC{$\Gamma; \phi,\ol{\alpha}, {\color{red} \ol{\gamma}} \vdash H$}
+       \AXC{$\Gamma; \ol{\beta}, {\color{red} \ol{\gamma}} \vdash F$}
        \BIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \fold^F_H :
-         \Nat^\emptyset\; (\Nat^{\ol{\beta}, \ol{\gamma}}\,H[\phi
+         \Nat^\emptyset\; (\Nat^{\ol{\beta}, {\color{red} \ol{\gamma}}}\,H[\phi
            :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\; (\Nat^{\ol{\beta},
-           \ol{\gamma}}\,(\mu \phi.\lambda \ol{\alpha}.H)\ol{\beta}\,F)$}
+          {\color{red} \ol{\gamma}}}\,(\mu \phi.\lambda \ol{\alpha}.H)\ol{\beta}\,F)$}
        \hspace*{0.2in}\DisplayProof\vspace*{-0.1in}
@@
 
-       \caption{Well-formed terms {\color{red} Change last three
-           judgements for new $\mu$ rule???}}\label{fig:terms} \vspace*{-0.00in}
+       \caption{Well-formed terms {\color{red} Remove red $\gamma$s?
+           Leave rule for map as
+           is???}}\label{fig:terms} \vspace*{-0.00in}
 \end{adjustbox}
@@
   formation rules.}  The ``extra'' functorial variables $\ol{\gamma}$
-in the rules for $\map^{\ol{F},\ol{G}}_H$, $\tin_H$, and $\fold^F_H$
-(i.e., those variables not affected by the substitution of $\phi$)
-deserve comment. They allow us to map or fold polymorphic functions
-over nested types. Suppose, for example, we want to map the
-polymorphic function $\mathit{flatten} : \Nat^\beta
+in the rules for $\map^{\ol{F},\ol{G}}_H$ {\color{red} , $\tin_H$, and
+  $\fold^F_H$} (i.e., those variables not affected by the substitution
+of $\phi$) deserve comment. They allow us to map {\color{red} or fold}
+polymorphic functions over nested types. Suppose, for example, we want
+to map the polymorphic function $\mathit{flatten} : \Nat^\beta
 (\mathit{PTree}\,\beta)\,(\mathit{List}\,\beta)$ over a list.  Even in
@@
 
+{\color{red} The next discussion should still be OK.}
 Our calculus is expressive enough to define a function
@@
 and the return type of the function is parameterized over only the
-variables occurring in the type of its final argument.  Even for ADTs
-there is a difference between which folds over them we can type when
-they are viewed as ADTs (i.e., as fixpoints of first-order functors)
-versus as proper nested types (i.e., as fixpoints of higher-order
-functors). This is because, in the return type of $\mathsf{fold}$, the
-arguments of the $\mu$-type must be variables bound by $\Nat$.  For
-ADTs, the $\mu$-type takes no arguments, making it possible to write
-recursive functions, such as a concatenation function for lists of
-type $\alpha ; \emptyset \vdash \Nat^\emptyset \ (\mu \beta. \onet +
-\alpha \times \beta)\, (\Nat^\emptyset (\mu \beta. \onet + \alpha
-\times \beta) \, (\mu \beta. \onet + \alpha \times \beta))$.  This is
-not possible for nested types --- even when they are semantically
-equivalent to ADTs.
+variables occurring in the type of its final argument.  {\color{red}
+  CHECK THE FOLLOWING! SHOULD BE OK.}  Even for ADTs there is a
+difference between which folds over them we can type when they are
+viewed as ADTs {\color{red} (i.e., as fixpoints of first-order
+  functors) EARLIER?}  versus as proper nested types {\color{red}
+  (i.e., as fixpoints of higher-order functors) EARLIER?}. This is
+because, in the return type of $\mathsf{fold}$, the arguments of the
+$\mu$-type must be variables bound by $\Nat$.  For ADTs, the
+$\mu$-type takes no arguments, making it possible to write recursive
+functions, such as a concatenation function for lists of type $\alpha
+; \emptyset \vdash \Nat^\emptyset \ (\mu \beta. \onet + \alpha \times
+\beta)\, (\Nat^\emptyset (\mu \beta. \onet + \alpha \times \beta) \,
+(\mu \beta. \onet + \alpha \times \beta))$.  This is not possible for
+nested types --- even when they are semantically equivalent to ADTs.
 
@@
   \setsem{\Gamma;\Phi \vdash G}\rho\\ 
-  \setsem{\Gamma;\Phi,\ol\gamma \vdash (\mu \phi.\lambda
+  \setsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash (\mu \phi.\lambda
     \ol{\alpha}. H)\ol{G}}\rho &= (\mu
-    T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho}\\
+    T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi,{\color{red} \ol\gamma} \vdash G}\rho}\\
     \text{where } T^\set_{H,\rho}\,F & = \lambda
-  \ol{A}. \setsem{\Gamma;\ol\gamma,\phi, \ol{\alpha} \vdash
+  \ol{A}. \setsem{\Gamma;{\color{red}\ol\gamma},\phi, \ol{\alpha} \vdash
     H}\rho[\phi :=  F][\ol{\alpha := A}]\\
   \text{and } T^\set_{H,\rho}\,\eta &= \lambda
-  \ol{A}. \setsem{\Gamma;\ol\gamma,\phi, \ol{\alpha} \vdash
+  \ol{A}. \setsem{\Gamma;{\color{red}\ol\gamma},\phi, \ol{\alpha} \vdash
     H}\id_\rho[\phi := \eta][\ol{\alpha := \id_{A}}]
@@
 admits a fixpoint.  Since $T_{H,\rho}^\set$ is defined in terms of
-$\setsem{\Gamma;\ol\gamma,\phi, \ol{\alpha} \vdash H}$, this means
+$\setsem{\Gamma;{\color{red} \ol\gamma},\phi, \ol{\alpha} \vdash H}$, this means
 that interpretations of types must be such functors, which in turn
@@
 where $k = |\ol \alpha|$. This means that for each $\setsem{\Gamma;
-  \ol\gamma, \phi^k, \ol{\alpha} \vdash H}$, the corresponding
+ {\color{red} \ol\gamma}, \phi^k, \ol{\alpha} \vdash H}$, the corresponding
 operator $T^\set_{H}$ can be extended to a {\em functor} from
@@
 $\ol{A}$ is $(T^\set_{H,f}\, F)_{\ol{A}} = \setsem{\Gamma;
-  \ol\gamma,\phi,\ol{\alpha} \vdash H}f[\phi := \id_F][\ol{\alpha :=
+  {\color{red}\ol\gamma},\phi,\ol{\alpha} \vdash H}f[\phi := \id_F][\ol{\alpha :=
     \id_A}]$.
@@
     G}f$
-\item If \,$\Gamma;\Phi,\ol\gamma \vdash (\mu \phi.\lambda
+\item If \,$\Gamma;\Phi,{\color{red}\ol\gamma} \vdash (\mu \phi.\lambda
   \ol{\alpha}. H)\ol{G}$ then
   \[\begin{array}{lll}
-  \setsem{\Gamma;\Phi,\ol\gamma \vdash (\mu  \phi.\lambda
-    \ol{\alpha}. H)\ol{G}} f &: &\setsem{\Gamma;\Phi,\ol\gamma 
+  \setsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash (\mu  \phi.\lambda
+    \ol{\alpha}. H)\ol{G}} f &: &\setsem{\Gamma;\Phi,{\color{red}\ol\gamma} 
     \vdash (\mu \phi.\lambda \ol{\alpha}. H)\ol{G}} \rho \to
-  \setsem{\Gamma;\Phi,\ol\gamma \vdash (\mu
+  \setsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash (\mu
     \phi.\lambda\ol{\alpha}. H)\ol{G}} \rho'\\
   &= &(\mu
-  T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho} \to (\mu
-  T^\set_{H,\rho'})\ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho'}
+  T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash G}\rho} \to (\mu
+  T^\set_{H,\rho'})\ol{\setsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash G}\rho'}
   \end{array}\]
@@
   defined by
-  \[\begin{array}{ll} & (\mu T^\set_{H,f})\ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash
-      G}\rho'} \circ (\mu T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi,\ol\gamma
-      \vdash G}f}\\ = & (\mu T^\set_{H,\rho'})\ol{\setsem{\Gamma;\Phi,\ol\gamma
-      \vdash G}f} \circ (\mu T^\set_{H,f})\ol{\setsem{\Gamma;\Phi,\ol\gamma
+  \[\begin{array}{ll} & (\mu T^\set_{H,f})\ol{\setsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash
+      G}\rho'} \circ (\mu T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi,{\color{red}\ol\gamma}
+      \vdash G}f}\\ = & (\mu T^\set_{H,\rho'})\ol{\setsem{\Gamma;\Phi,{\color{red}\ol\gamma}
+      \vdash G}f} \circ (\mu T^\set_{H,f})\ol{\setsem{\Gamma;\Phi,{\color{red}\ol\gamma}
       \vdash G}\rho}\end{array}\]  The latter equality holds because $\mu
@@
 \begin{CD}
- (\mu T^\set_{H,\rho}) \ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho} @> (\mu
-  T^\set_{H,f})_{\ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho}} >> (\mu
-  T^\set_{H,\rho'}) \ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho} \\ 
- @V(\mu T^\set_{H,\rho}) \ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}f}VV @V  (\mu
- T^\set_{H,\rho'}) \ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}f} VV \\ 
-(\mu T^\set_{H,\rho}) \ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho'} @>(\mu
- T^\set_{H,f})_{\ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho'}}>> (\mu
- T^\set_{H,\rho'}) \ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho'}
+ (\mu T^\set_{H,\rho}) \ol{\setsem{\Gamma;\Phi,{\color{red} \ol\gamma} \vdash G}\rho} @> (\mu
+  T^\set_{H,f})_{\ol{\setsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash G}\rho}} >> (\mu
+  T^\set_{H,\rho'}) \ol{\setsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash G}\rho} \\ 
+ @V(\mu T^\set_{H,\rho}) \ol{\setsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash G}f}VV @V  (\mu
+ T^\set_{H,\rho'}) \ol{\setsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash G}f} VV \\ 
+(\mu T^\set_{H,\rho}) \ol{\setsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash G}\rho'} @>(\mu
+ T^\set_{H,f})_{\ol{\setsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash G}\rho'}}>> (\mu
+ T^\set_{H,\rho'}) \ol{\setsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash G}\rho'}
 \end{CD}
@@
   \relsem{\Gamma;\Phi \vdash G}\rho\\  
-   \relsem{\Gamma;\Phi,\ol\gamma \vdash (\mu \phi.\lambda
+   \relsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash (\mu \phi.\lambda
     \ol{\alpha}. H)\ol{G}}\rho
-  &= (\mu T_{H,\rho})\ol{\relsem{\Gamma;\Phi,\ol\gamma \vdash
+  &= (\mu T_{H,\rho})\ol{\relsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash
      G}\rho}\\
@@
     &= \lambda \ol{R}. \relsem{
-      \Gamma;\ol\gamma,\phi,\ol{\alpha} \vdash H}\rho[\phi :=
+      \Gamma;{\color{red}\ol\gamma},\phi,\ol{\alpha} \vdash H}\rho[\phi :=
     F][\ol{\alpha := R}]\\
@@
     &= \lambda \ol{R}. \relsem{
-      \Gamma;\ol\gamma,\phi,\ol{\alpha} \vdash H}\id_\rho[\phi :=
+      \Gamma;{\color{red}\ol\gamma},\phi,\ol{\alpha} \vdash H}\id_\rho[\phi :=
     \delta][\ol{\alpha := \id_{\ol{R}}}]
@@
 fixpoint. Since $T_{H,\rho}$ is defined in terms of
-$\relsem{\Gamma;\ol\gamma,\phi^k, \ol{\alpha} \vdash H}$, this means
+$\relsem{\Gamma;{\color{red}\ol\gamma},\phi^k, \ol{\alpha} \vdash H}$, this means
 that relational interpretations of types must be $\omega$-cocontinuous
@@
 T_{H,\rho'}\, F$ whose component at $\ol{R}$ is $(T_{H,f}\,
-F)_{\ol{R}} = \relsem{\Gamma; \ol\gamma,\phi,\ol{\alpha} \vdash
+F)_{\ol{R}} = \relsem{\Gamma; {\color{red}\ol\gamma},\phi,\ol{\alpha} \vdash
   H}f[\phi := \id_F][\ol{\alpha := \id_R}]$.
@@
   is by the definition of application of relation transformers from
-  Definition~9, and the fourth is by
-  Lemma~21:
+  Definition~9, and the fourth is by Equation~\ref{eq:eq-pres-eq}:
 \[
@@
 \begin{split}
-\relsem{\Gamma; \Phi, \ol\gamma \vdash (\mu \phi.\lambda
+\relsem{\Gamma; \Phi, {\color{red}\ol\gamma} \vdash (\mu \phi.\lambda
   \ol{\alpha}. H)\ol{F}}\Eq_{\rho} 
-&=(\mu {T_{H, \Eq_{\rho}}}) \,\ol{\relsem{\Gamma; \Phi, \ol\gamma  \vdash F}\Eq_{\rho}}\\ 
-&= \colim{n \in \nat}{T^n_{H, \Eq_{\rho}} K_0}\, \ol{\relsem{\Gamma; \Phi, \ol\gamma 
+&=(\mu {T_{H, \Eq_{\rho}}}) \,\ol{\relsem{\Gamma; \Phi, {\color{red} \ol\gamma}  \vdash F}\Eq_{\rho}}\\ 
+&= \colim{n \in \nat}{T^n_{H, \Eq_{\rho}} K_0}\, \ol{\relsem{\Gamma; \Phi, {\color{red}\ol\gamma} 
   \vdash F}\Eq_{\rho}}\\
 &= \colim{n \in \nat}{ T^n_{H, \Eq_{\rho}} K_0 \,\ol{\Eq_{\setsem{\Gamma;
-    \Phi, \ol\gamma  \vdash F}\rho}}}\\
+    \Phi, {\color{red}\ol\gamma}  \vdash F}\rho}}}\\
 &= \colim{n \in \nat}{(\Eq_{(T^\set_{H,\rho})^n K_0})^*
-  \ol{\Eq_{\setsem{\Gamma; \Phi, \ol\gamma  \vdash F}\rho}}}\\
+  \ol{\Eq_{\setsem{\Gamma; \Phi, {\color{red}\ol\gamma}  \vdash F}\rho}}}\\
 &= \colim{n \in \nat}{\Eq_{(T^\set_{H,\rho})^n K_0 \,\ol{\setsem{\Gamma;
-        \Phi, \ol\gamma  \vdash F}\rho}}}\\ 
+        \Phi, {\color{red}\ol\gamma}  \vdash F}\rho}}}\\ 
 &= \Eq_{\colim{n \in \nat}{ (T^\set_{H,\rho})^n K_0\,
-    \ol{\setsem{\Gamma; \Phi, \ol\gamma  \vdash F}\rho}}}\\
-&= \Eq_{\setsem{\Gamma; \Phi, \ol\gamma  \vdash (\mu \phi.\lambda
+    \ol{\setsem{\Gamma; \Phi, {\color{red}\ol\gamma}  \vdash F}\rho}}}\\
+&= \Eq_{\setsem{\Gamma; \Phi, {\color{red}\ol\gamma}  \vdash (\mu \phi.\lambda
       \ol{\alpha}. H)\ol{F}}\rho}
@@
 Here, the third equality is by induction hypothesis, the fifth is by
-Lemma~21 and the fourth equality is because, for every $n \in \nat$,
-the following two statements can be proved by simultaneous induction:
-and for any $H$, $\rho$, $A$, and subformula $J$ of $H$,
+Equation~\ref{eq:eq-pres-eq}, and the fourth equality is because, for
+every $n \in \nat$, the following two statements can be proved by
+simultaneous induction: and for any $H$, $\rho$, $A$, and subformula
+$J$ of $H$,
 \begin{equation}\label{eq:iel-fix-point-intermediate1}
@@
 \end{equation}
-(Notice that we don't know what's in the context $\Phi'$.)
-We prove~\eqref{eq:iel-fix-point-intermediate1} by induction on $n$.
+(Notice that we don't know what's in the context $\Phi'$.)  We prove
+Equation~\ref{eq:iel-fix-point-intermediate1} by induction on $n$.
 The case $n=0$ is trivial, because $T^0_{H,\Eq_{\rho}} K_0 = K_0$ and
@@
 \overline{\Eq_A} \\ 
-&= \relsem{\Gamma; \phi, \ol{\alpha}, \ol\gamma \vdash H} \Eq_{\rho} [\phi
+&= \relsem{\Gamma; \phi, \ol{\alpha}, {\color{red}\ol\gamma} \vdash H} \Eq_{\rho} [\phi
   := T^{n}_{H, \Eq_{\rho}} K_0] \overline{[\alpha :=
     \Eq_A]} \\ 
-&= \relsem{\Gamma; \phi, \ol{\alpha}, \ol\gamma \vdash H} \Eq_{\rho} [\phi
+&= \relsem{\Gamma; \phi, \ol{\alpha}, {\color{red}\ol\gamma} \vdash H} \Eq_{\rho} [\phi
   := \Eq_{(T^\set_{H,\rho})^n K_0}] \overline{[\alpha :=
     \Eq_A]} \\ 
-&= \relsem{\Gamma; \phi, \ol{\alpha}, \ol\gamma \vdash H} \Eq_{\rho [\phi
+&= \relsem{\Gamma; \phi, \ol{\alpha}, {\color{red}\ol\gamma} \vdash H} \Eq_{\rho [\phi
     := (T^\set_{H,\rho})^n K_0] \overline{[\alpha :=
       A]}} \\ 
-&= \Eq_{\setsem{\Gamma; \phi, \ol{\alpha}, \ol\gamma \vdash H} \rho [\phi
+&= \Eq_{\setsem{\Gamma; \phi, \ol{\alpha}, {\color{red}\ol\gamma} \vdash H} \rho [\phi
     := (T^\set_{H,\rho})^n K_0] \overline{[\alpha :=
@@
 \]
-Here, the third equality is by~\eqref{eq:iel-fix-point-intermediate2}
-for $J = H$, the fifth by the induction hypothesis of the IEL on $H$,
-and the last is by Lemma~21.
+Here, the third equality is by
+Equation~\ref{eq:iel-fix-point-intermediate2} for $J = H$, the fifth
+by the induction hypothesis of the IEL on $H$, and the last is by
+Equation~\ref{eq:eq-pres-eq}.
 
@@
 
-We prove~\eqref{eq:iel-fix-point-intermediate2} by structural
-induction on $J$. The only interesting cases, though, are when
-$J = \phi \ol{G}$ and when $J = (\mu \psi.\lambda \ol\beta. G)\ol K$.
+We prove Equation~\ref{eq:iel-fix-point-intermediate2} by structural
+induction on $J$. The only interesting cases, though, are when $J =
+\phi \ol{G}$ and when $J = (\mu \psi.\lambda \ol\beta. G)\ol K$.
 \begin{itemize}
@@
 &= T^{n}_{H,\Eq_{\rho}} K_0\, \overline{\Eq_{\setsem{\Gamma;
-        \Phi', \phi, \ol{\alpha} \vdash G} \rho [\phi :=
+        \Phi', \phi, \ol{\alpha} \,\vdash G} \rho [\phi :=
         (T^\set_{H,\rho})^{n} K_0] \overline{[\alpha :=
@@
 &= (\Eq_{(T^\set_{H,\rho})^{n} K_0})^* \,\overline{\Eq_{\setsem{\Gamma;
-        \Phi', \phi, \ol{\alpha} \vdash G} \rho [\phi :=
+        \Phi', \phi, \ol{\alpha} \,\vdash G} \rho [\phi :=
         (T^\set_{H,\rho})^{n} K_0] \overline{[\alpha :=
@@
 \]
-Here, the second equality is by the induction hypothesis
-for~\eqref{eq:iel-fix-point-intermediate2} on the $G$s, the fourth is
-by the induction hypothesis for the IEL on the $G$s, and the fifth is
-by the induction hypothesis on $n$
-for~\eqref{eq:iel-fix-point-intermediate1}.
+Here, the second equality is by the induction hypothesis for
+Equation~\ref{eq:iel-fix-point-intermediate2} on the $G$s, the fourth
+is by the induction hypothesis for the IEL on the $G$s, and the fifth
+is by the induction hypothesis on $n$ for
+Equation~\ref{eq:iel-fix-point-intermediate1}.
 
@@
   The case $J = (\mu \psi.\, \lambda \ol{\beta}.\, G) \ol{K}$
-is proved by the sequence of equalities (where $\Phi' = \Phi'', \ol\gamma$):
+is proved by the sequence of equalities {\color{red} (where $\Phi' = \Phi'', \ol\gamma$)}:
 \[
 \begin{split}
-& \relsem{\Gamma; \Phi'', \ol\gamma, \phi, \ol{\alpha} \vdash (\mu \psi.\, \lambda \ol{\beta}.\, G) \ol{K}}
+& \relsem{\Gamma; {\color{red} \Phi'', \ol\gamma}, \phi, \ol{\alpha} \vdash (\mu \psi.\, \lambda \ol{\beta}.\, G) \ol{K}}
   \Eq_{\rho} [\phi := T^{n}_{H,\Eq_{\rho}} K_0]
@@
 &= (\mu {T_{G, \Eq_{\rho}[\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]}})
-  \,\ol{\relsem{\Gamma; \Phi'', \ol\gamma, \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
+  \,\ol{\relsem{\Gamma; {\color{red}\Phi'', \ol\gamma}, \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
   [\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]} \\ 
 &= \colim{m \in \nat} {T^m_{G, \Eq_{\rho}[\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]}}\,K_0
-  \,(\ol{\relsem{\Gamma; \Phi'', \ol\gamma, \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
+  \,(\ol{\relsem{\Gamma; {\color{red} \Phi'', \ol\gamma}, \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
   [\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]}) \\ 
 &= \colim{m \in \nat} {T^m_{G, \Eq_{\rho}[\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]}}\,K_0
-  \,(\ol{\relsem{\Gamma; \Phi'', \ol\gamma, \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
+  \,(\ol{\relsem{\Gamma; {\color{red}\Phi'', \ol\gamma}, \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
   [\phi := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha := \Eq_A}]}) \\  %%
-%% &= \colim{m \in \nat} {T^m_{G, \Eq_{\rho[\overline{\alpha := A}]}}}\,K_0
-%%   \,(\ol{\relsem{\Gamma; \Phi, \ol\gamma, \phi, \ol{\alpha} \vdash K}
-%%   \Eq_{\rho[\phi := (T^{\set}_{H,\rho})^n K_0][\overline{\alpha := A}]}}) \\ 
-%% &= \colim{m \in \nat} {T^m_{G, \Eq_{\rho[\overline{\alpha := A}]}}}\,K_0
-%%   \,\ol{\Eq_{\setsem{\Gamma; \Phi, \ol\gamma, \phi, \ol{\alpha} \vdash K}
-%%   \rho[\phi := (T^{\set}_{H,\rho})^n K_0][\overline{\alpha := A}]}} \\ 
-%% &= \colim{m \in \nat} (\Eq_{{(T^{\set}_{G, \rho[\overline{\alpha := A}]}})^m\,K_0})^*
-%%   \,\ol{\Eq_{\setsem{\Gamma; \Phi, \ol\gamma, \phi, \ol{\alpha} \vdash K}
-%%   \rho[\phi := (T^{\set}_{H,\rho})^n K_0][\overline{\alpha := A}]}} \\ 
-%% &= (\Eq_{\colim{m \in \nat} {(T^{\set}_{G, \rho[\overline{\alpha := A}]}})^m\,K_0})^*
-%%   \,\ol{\Eq_{\setsem{\Gamma; \Phi, \ol\gamma, \phi, \ol{\alpha} \vdash K}
-%%   \rho[\phi := (T^{\set}_{H,\rho})^n K_0][\overline{\alpha := A}]}} \\ 
-%% &= (\Eq_{{\mu T^{\set}_{G, \rho[\overline{\alpha := A}]}}})^*
-%%   \,\ol{\Eq_{\setsem{\Gamma; \Phi, \ol\gamma, \phi, \ol{\alpha} \vdash K}
-%%   \rho[\phi := (T^{\set}_{H,\rho})^n K_0][\overline{\alpha := A}]}} \\
-%% &= \Eq_{{\mu T^{\set}_{G, \rho[\overline{\alpha := A}]}}
-%%   \,\ol{\setsem{\Gamma; \Phi, \ol\gamma, \phi, \ol{\alpha} \vdash K}
-%%   \rho[\phi := (T^{\set}_{H,\rho})^n K_0][\overline{\alpha := A}]}} \\
-%=& \Eq_{\setsem{\Gamma; \Phi, \ol\gamma, \ol{\alpha}, \ol{\gamma} \vdash (\mu \psi.\, \lambda \beta.\, G) \ol{\gamma}}
-%  \rho[\overline{\alpha := A}][\ol{\gamma := \setsem{\Gamma; \Phi, \ol\gamma, \phi, \ol{\alpha} \vdash K}
-%  \rho[\phi := (T^{\set}_{H,\rho})^n K_0][\overline{\alpha := A}]}]} \\
 &= \colim{m \in \nat} {T^m_{G, \Eq_{\rho}[\phi := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha := \Eq_A}]}}\,K_0
-  \,(\ol{\relsem{\Gamma; \Phi'', \ol\gamma, \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
+  \,(\ol{\relsem{\Gamma; {\color{red}\Phi'', \ol\gamma}, \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
   [\phi := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha := \Eq_A}]}) \\ 
 &= (\mu {T_{G, \Eq_{\rho}[\phi := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha := \Eq_A}]}})
-  \,\ol{\relsem{\Gamma; \Phi'', \ol\gamma, \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
+  \,\ol{\relsem{\Gamma; {\color{red}\Phi'', \ol\gamma}, \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
   [\phi := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha := \Eq_A}]} \\ 
-&= \relsem{\Gamma; \Phi'', \ol\gamma, \phi, \ol{\alpha} \vdash (\mu \psi.\, \lambda \beta.\, G) \ol{K}}
+&= \relsem{\Gamma; {\color{red}\Phi'', \ol\gamma}, \phi, \ol{\alpha} \vdash (\mu \psi.\, \lambda \beta.\, G) \ol{K}}
   \Eq_{\rho}[\phi := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha := \Eq_{A}}]
@@
 \]
-Here, the third equality is by the induction hypothesis
-for~\eqref{eq:iel-fix-point-intermediate2} on the $K$s, and the fourth
-equality holds because we can prove that, for all $m \in \nat$,
+Here, the third equality is by the induction hypothesis for
+Equation~\ref{eq:iel-fix-point-intermediate2} on the $K$s, and the
+fourth equality holds because we can prove that, for all $m \in \nat$,
 \begin{equation}\label{eq:helper}
@@
 \end{equation}
-Indeed, the base case of~\eqref{eq:helper} is trivial because
+Indeed, the base case of Equation~\ref{eq:helper} is trivial because
 \[T^0_{G,
@@
         \Eq_A}]}\,K_0) \\
-&= \lambda \ol{R}. \relsem{\Gamma; \ol\gamma, \phi, \ol\alpha, \psi,
+&= \lambda \ol{R}. \relsem{\Gamma; {\color{red}\ol\gamma}, \phi, \ol\alpha, \psi,
     \ol\beta \vdash G} \Eq_{\rho} [\phi := T^{n}_{H,\Eq_{\rho}}
@@
           \Eq_A}]}\,K_0][\overline{\beta := R}] \\
-&= \lambda \ol{R}. \relsem{\Gamma; \ol\gamma, \phi, \ol\alpha, \psi,
+&= \lambda \ol{R}. \relsem{\Gamma; {\color{red} \ol\gamma}, \phi, \ol\alpha, \psi,
     \ol\beta \vdash G} \Eq_{\rho} [\phi := \Eq_{(T^{\set}_{H,\rho})^n
@@
 
-Here, the second equality holds by the induction hypothesis
-for~\eqref{eq:helper} on $m$. The fourth equality holds because, due
-to typing rule restrictions for the $\mu$ types, $\phi$ either does
-not appear in $G$, or must have arity $0$, in which case $\ol\alpha$
-must be empty, if $\phi$ appears in $G$, and
-uses~\eqref{eq:iel-fix-point-intermediate2} for $G$ when $\phi$ has
+Here, the second equality holds by the induction hypothesis for
+Equation~\ref{eq:helper} on $m$. The fourth equality holds because,
+due to typing rule restrictions for the $\mu$ types, $\phi$ either
+does not appear in $G$, or must have arity $0$, in which case
+$\ol\alpha$ must be empty, if $\phi$ appears in $G$, and uses
+Equation~\ref{eq:iel-fix-point-intermediate2} for $G$ when $\phi$ has
 arity $0$. {\color{red} Note that if functorial variables of arity
@@
 
-\begin{thebibliography}{9}
-%\bibliography{references}
-\end{thebibliography}
+\bibliographystyle{alpha}
+\bibliography{references}
 

### 2020-12-14 free-theorems-nesteds/20-mscs.tex
started on second half of journal version

@@
 \newcommand{\reverse}{\mathsf{reverse}}
-\newcommand{\Bcons}{\mathit{Bcons}}
-\newcommand{\Bnil}{\mathit{Bnil}}
+\newcommand{\bcons}{\mathit{bcons}}
+\newcommand{\bnil}{\mathit{bnil}}
 
@@
 \begin{defi}\label{def:wftypes}
-The formation rules for the set $\F$ of\, {\em well-formed types
-  {\color{red} over $\Gamma$ and $\Phi$?}} are
+The formation rules for the set $\F$ of\, {\em well-formed types} are
 
@@
 \\[3ex]
-\AXC{$\Gamma;{\color{red}\ol{\gamma^0}},\ol{\alpha^0},\phi^k \vdash F$}
+\AXC{$\Gamma;\ol{\alpha^0},\phi^k \vdash F$}
 \AXC{$\quad\quad\ol{\Gamma;\Phi \vdash G}$}
-\AXC{$\quad\quad\ol{\Gamma;\Phi \vdash H}$}
-\TIC{$\Gamma;\Phi \vdash (\mu \phi^k.\lambda
-  \ol{\alpha^0}. \,F {\color{red} [\ol{\gamma := H}]})\,\ol{G}$}  
+\BIC{$\Gamma;\Phi \vdash (\mu \phi^k.\lambda
+  \ol{\alpha^0}. \,F)\,\ol{G}$}  
 \DisplayProof
-\end{array}\]
-
-\vspace*{0.04in}
-
-\[\begin{array}{cc}
+\\[4ex]
 \AXC{$\Gamma;\Phi \vdash F$}
@@
 \DisplayProof
-&
+\\[3ex]
 \AXC{$\Gamma;\Phi \vdash F$}
@@
 \end{defi}
-
-{\color{red} Note that we use first-order substitution in the
-  $\mu$-clause even though it isn't defined until the next
-  page. Mutually inductive definition? The purpose of the substitution
-  in the $\mu$-type clause of Def 2.1 is to ensure that variables in
-  $\Phi$ are not applied to types involving variables from $\ol\alpha$
-  in $F$; indeed, the first-order substitution $[\ol{\gamma := H}]$
-  constrains the types in $\ol H$ to have arity $0$, and constrains
-  the variables in $\Phi$ to already be fully applied to types not
-  involving $\ol\alpha$ in $\ol H$, making such applications
-  impossible. Of course, rather than intertwining the definitions of
-  well-formed types of our calculus and the operation of first-order
-  substitution on them we could instead prohibit such applications in
-  a side condition to the rule for well-formed $\mu$-types, but we
-  favor the precision of the current presentation.} We write $\vdash
-F$ for $\emptyset;\emptyset \vdash F$.  Definition~\ref{def:wftypes}
-ensures that the expected weakening rules for well-formed types hold,
-although weakening does not change the contexts in which types can be
-formed. If $\Gamma;\emptyset \vdash F$ and $\Gamma;\emptyset \vdash
-G$, then our rules allow formation of the types $\Gamma;\emptyset
-\vdash \Nat^\emptyset F \,G$ and $\Gamma; \emptyset \vdash
-\Nat^{\ol\alpha} \,\onet \,F$, which represent the System F types
-$\Gamma \vdash F \to G$ and $\forall$-type $\Gamma; \emptyset \vdash
-\forall \ol\alpha . F$, respectively. However, some System F types,
-such as $\forall \alpha. (\alpha \to \alpha) \to \alpha$, are not
-representable in our calculus.
-
-{\color{red} CHANGE TEXT FOR NEW $\mu$-RULE.}
-Definition~\ref{def:wftypes} allows the formation of all of the
-(closed {\color{red} NOT!}) nested types from the introduction:
+We write $\vdash F$ for $\emptyset;\emptyset \vdash F$.
+Definition~\ref{def:wftypes} ensures that the expected weakening rules
+for well-formed types hold, although weakening does not change the
+contexts in which types can be formed. If $\Gamma;\emptyset \vdash F$
+and $\Gamma;\emptyset \vdash G$, then our rules allow formation of the
+types $\Gamma;\emptyset \vdash \Nat^\emptyset F \,G$ and $\Gamma;
+\emptyset \vdash \Nat^{\ol\alpha} \,\onet \,F$, which represent the
+System F types $\Gamma \vdash F \to G$ and $\forall$-type $\Gamma;
+\emptyset \vdash \forall \ol\alpha . F$, respectively. However, some
+System F types, such as $\forall \alpha. (\alpha \to \alpha) \to
+\alpha$, are not representable in our calculus. Note that, in the
+rule for $\mu$-types, no variables in $\Phi$ are applied to types
+involving variables from $\ol\alpha$ in the body $F$ of $\mu
+\phi. \lambda \ol\alpha.F$. This will be critical to proving the
+Identity Extension Lemma for our calculus.
+
+Definition~\ref{def:wftypes} allows the formation of all of the types
+from Section~\ref{sec:intro}:
 \[\begin{array}{lll}
@@
 \end{array}\]
-{\color{red} The representation of $\mathit{List}\,\alpha$ as the ADT
-  $\mu \beta. \onet + \alpha \times \beta$ cannot be functorial in
-  $\alpha$. By contrast, the representation of $\mathit{List}\,\alpha$
-  as the nested type $(\mu \phi. \lambda \beta.\,\onet + \beta \times
-  \phi \beta)\,\alpha$
-%Each of these types
-  can be considered either functorial in $\alpha$ or not, according to
-  whether $\alpha$ is in the functorial or non-functorial context in
-  which this type is formed.} In particular, if $\emptyset;\alpha
-\vdash \mathit{List} \,\alpha$, then $\vdash \Nat^\alpha \onet\,
-(\mathit{List}\,\alpha)$ is well-formed; if $\alpha;\emptyset \vdash
-\List \,\alpha$, then it is not.  Similarly, if
-$\mathit{Tree}\,\alpha\,\gamma = \mu \,\beta.\, \alpha + \beta \times
-\gamma \times \beta$, then $\gamma;\emptyset \vdash \Nat^\alpha
-(\mathit{List}\, \alpha) \, (\mathit{Tree} \, \alpha\, \gamma)$ is
-well-formed, but $\emptyset;\gamma \vdash \Nat^\alpha (\mathit{List}\,
-\alpha) \, (\mathit{Tree} \, \alpha\, \gamma)$ is not. {\color{red}
-  KEEP? Although types can be functorial in variables of arity greater
-  than $0$, the body $F$ of a type $(\mu \phi.\lambda \ol\alpha.F)\ol
-  G$ cannot.} For example, the type $\mathit{GRose}\, \phi \, \alpha =
-\mu\beta . \alpha \times \phi \beta$ {\color{red} can be functorial
-  --- NOT!} or not in $\alpha$, but cannot be functorial in
-$\phi$. Moreover, $\mathit{GRose}\,\phi\,\alpha$ cannot be the
-(co)domain of a $\Nat$-type representing a natural transformation
-functorial in $\phi$, and the type $\mu \phi. \lambda
-\alpha. \mathit{GRose}\,\phi\,\alpha$ is not well-formed in our
+The representation of $\mathit{List}\,\alpha$ as the ADT $\mu
+\beta. \onet + \alpha \times \beta$ cannot be functorial in
+$\alpha$. By contrast, its representation as the nested type $(\mu
+\phi. \lambda \beta.\,\onet + \beta \times \phi \beta)\,\alpha$ can be
+considered either functorial in $\alpha$ or not, according to whether
+$\alpha$ is in the functorial or non-functorial context in which this
+type is formed. If $\emptyset;\alpha \vdash \mathit{List} \,\alpha$,
+then $\vdash \Nat^\alpha \onet\, (\mathit{List}\,\alpha)$ is
+well-formed; if $\alpha;\emptyset \vdash \List \,\alpha$, then it is
+not.  Similarly, if $\mathit{Tree}\,\alpha\,\gamma = \mu \,\beta.\,
+\alpha + \beta \times \gamma \times \beta$, then $\gamma;\emptyset
+\vdash \Nat^\alpha (\mathit{List}\, \alpha) \, (\mathit{Tree} \,
+\alpha\, \gamma)$ is well-formed, but $\emptyset;\gamma \vdash
+\Nat^\alpha (\mathit{List}\, \alpha) \, (\mathit{Tree} \, \alpha\,
+\gamma)$ is not. The body $F$ of a type $(\mu \phi.\lambda
+\ol\alpha.F)\ol G$ can only be functorial in $\phi$ and the variables
+in $\ol\alpha$. As a result, both $\phi$ and the variables in
+$\ol\alpha$ must be non-functorial in $\mathit{GRose}\, \phi \, \alpha
+= \mu\beta . \alpha \times \phi \beta$, $\mathit{GRose}\,\phi\,\alpha$
+cannot be the (co)domain of a $\Nat$-type binding any variables, and
+no type of the form $(\mu \phi. \lambda
+\alpha. \mathit{GRose}\,\phi\,\alpha)\tau$ is well-formed in our
 system. These restrictions ensure that the set and relational
@@
 functors. In Section~\ref{sec:conclusion} we discuss how more
-expressive types might be accommodated.
+expressive types might be accommodated. {\color{red} Daniel: Check the
+  above carefully?}
 
@@
 
-{\color{red} Check no change needed for new $\mu$-rule in next
-  def. Just note that in application case the args to $\phi$ go in for
-  the $\alpha$s in $F$s replacing $\phi$.}
 \begin{defi}\label{def:second-order-subst}
@@
 (\Nat^{\ol\beta} G \,K)[\phi :=_{\ol{\alpha}} F]
-& = & \Nat^{\ol\beta}\, (G[\phi :=_{\ol{\alpha}} F]) \,(K[\phi
-  :=_{\ol{\alpha}} F])\\
+& = & \Nat^{\ol\beta}\, G \,K\\
 (\psi\ol{G})[\phi :=_{\ol{\alpha}} F] & = &
@@
 ((\mu \psi. \lambda \ol{\beta}. G)\ol{K})[\phi :=_{\ol{\alpha}}
-  F] & = & (\mu \psi. \lambda \ol{\beta}. \,G[\phi :=_{\ol{\alpha}}
-  F])\, \ol{K[\phi :=_{\ol{\alpha}} F]}
+  F] & = & (\mu \psi. \lambda \ol{\beta}. \,G)\, \ol{K[\phi :=_{\ol{\alpha}} F]}
 \end{array}\]
@@
 \noindent
-It is not hard to see that $\Gamma;\Phi \vdash H[\phi :=_{\ol{\alpha}}
-  F]$. In addition, if \,$\Gamma, \phi^k; \Phi \vdash H$,\, $\Gamma;
-\ol\psi,\ol{\alpha} \vdash F$, $|\ol\alpha| = k$, and $\Phi \cap
-\ol\psi = \emptyset$, then $\Gamma,\ol\psi';\Phi \vdash H[\phi
-  :=_{\ol{\alpha}} F[\ol{\psi :== \psi'}]]$.
+{\color{red} Daniel: Agree?} The idea is that the arguments to $\phi$
+get substituted for the variables in $\ol\alpha$ in each $F$ replacing
+an occurrence of $\phi$. It is not hard to see that $\Gamma;\Phi
+\vdash H[\phi :=_{\ol{\alpha}} F]$. In addition, if \,$\Gamma, \phi^k;
+\Phi \vdash H$,\, $\Gamma; \ol\psi,\ol{\alpha} \vdash F$, $|\ol\alpha|
+= k$, and $\Phi \cap \ol\psi = \emptyset$, then $\Gamma,\ol\psi';\Phi
+\vdash H[\phi :=_{\ol{\alpha}} F[\ol{\psi :== \psi'}]]$.
 
@@
 
-  \begin{adjustbox}{varwidth=5.4in, max width=5in, fbox, center}
+  \begin{adjustbox}{varwidth=5.8in, max width=5.5in, fbox, center}
        \[\begin{array}{ccc}
@@
        \[\begin{array}{c}
-       \AXC{$\Gamma; \Phi \vdash F,G$}
+       \;\;\AXC{$\Gamma; \Phi \vdash F,G$}
        \AXC{$\Gamma;\Phi \,|\, \Delta \vdash t : F+G$}
@@
        \\\\
-       \AXC{$\Gamma; \phi, \ol{\alpha}, {\color{red}\ol{\gamma}} \vdash H$}
+       \AXC{$\Gamma; \phi, \ol{\alpha} \vdash H$}
        \UIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \tin_H :
-         \Nat^{\ol{\beta},{\color{red}\ol{\gamma}}} H[\phi :=_{\ol{\beta}} (\mu
+         \Nat^{\ol{\beta}} H[\phi :=_{\ol{\beta}} (\mu
            \phi.\lambda \ol{\alpha}.H)\ol{\beta}][\ol{\alpha := \beta}]\,(\mu
@@
        \\\\
-       \AXC{$\Gamma; \phi,\ol{\alpha}, {\color{red} \ol{\gamma}} \vdash H$}
-       \AXC{$\Gamma; \ol{\beta}, {\color{red} \ol{\gamma}} \vdash F$}
+       \AXC{$\Gamma; \phi,\ol{\alpha} \vdash H$}
+       \AXC{$\Gamma; \ol\beta \vdash F$}
        \BIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \fold^F_H :
-         \Nat^\emptyset\; (\Nat^{\ol{\beta}, {\color{red} \ol{\gamma}}}\,H[\phi
-           :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\; (\Nat^{\ol{\beta},
-          {\color{red} \ol{\gamma}}}\,(\mu \phi.\lambda \ol{\alpha}.H)\ol{\beta}\,F)$}
+         \Nat^\emptyset\; (\Nat^{\ol{\beta}}\,H[\phi
+           :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\;
+         (\Nat^{\ol{\beta}}\,(\mu \phi.\lambda
+         \ol{\alpha}.H)\ol{\beta}\,F)$} 
        \hspace*{0.2in}\DisplayProof\vspace*{-0.1in}
@@
 
-       \caption{Well-formed terms {\color{red} Remove red $\gamma$s?
-           Leave rule for map as
-           is???}}\label{fig:terms} \vspace*{-0.00in}
+       \caption{Well-formed terms}\label{fig:terms} \vspace*{-0.00in}
 \end{adjustbox}
@@
 
-{\color{red} Change the rest of the section to agree with new term
-  formation rules.}  The ``extra'' functorial variables $\ol{\gamma}$
-in the rules for $\map^{\ol{F},\ol{G}}_H$ {\color{red} , $\tin_H$, and
-  $\fold^F_H$} (i.e., those variables not affected by the substitution
-of $\phi$) deserve comment. They allow us to map {\color{red} or fold}
+The ``extra'' functorial variables $\ol{\gamma}$ in the rules for
+$\map^{\ol{F},\ol{G}}_H$ (i.e., those variables not affected by the
+substitution of $\phi$) deserve comment. They allow us to map
 polymorphic functions over nested types. Suppose, for example, we want
@@
 
-\noindent
-Similar remarks justify the appearance of $\ol \gamma$ in the typing
-rules for $\tin$ and $\mathsf{fold}$.
-
-{\color{red} The next discussion should still be OK.}
 Our calculus is expressive enough to define a function
@@
 that reverses the order of the leaves in a perfect tree. This function
-maps, e.g., the perfect tree $((1,2),(3,4))$ to $((4,3),(2,1))$, and
+maps, e.g., the perfect tree $((1,2),(3,4))$ to $((4,3),(2,1))$
+{\color{red} CHANGE TO $pnil/pnode$ NOTATION?}, and
 can be defined as
@@
 $\mathtt{reverseBush}$ function that reverses the data in a bush. This
-function maps, e.g., {\color{red} the bush $?$ to $??$}, and can be
-defined as
+function maps, e.g., the bush
+\[\begin{array}{l}
+\mathit{bcons}\,0 \,(\mathit{bcons}\, (\mathit{bcons}\, 1\,
+(\mathit{bcons}\, (\mathit{bcons}\, 2\,\mathit{bnil}) \,
+\mathit{bnil}))\\ \hspace*{0.93in}(\mathit{bcons}\, (\mathit{bcons}\,
+(\mathit{bcons}\, 2 \, (\mathit{bcons}\, (\mathit{bcons}\, 3\,
+\mathit{bnil})\,  \mathit{bnil}))\, \mathit{bnil})\, \mathit{bnil}))
+\end{array}\]
+to
+\[\begin{array}{c}
+\hspace*{-1.64in}\mathit{bcons}\,3\, (\mathit{bcons}\, (\mathit{bcons}\, 0
+\,(\mathit{bcons}\, (\mathit{bcons}\, 2 \, \mathit{bnil}) \,
+\mathit{bnil}))\\ 
+\hspace*{0.93in}(\mathit{bcons}\, (\mathit{bcons}\, (\mathit{bcons}\, 2
+\,(\mathit{bcons}\, (\mathit{bcons}\, 1 \, \mathit{bnil}) \,
+\mathit{bnil})) \, \mathit{bnil})\, \mathit{bnil})) 
+\end{array}\]
+It can be defined as
 \[\vdash (\fold^{\mathit{Bush}\,\alpha}_{\onet + \beta \times \phi (\phi \beta)})_\emptyset\, balg
@@
 and the return type of the function is parameterized over only the
-variables occurring in the type of its final argument.  {\color{red}
-  CHECK THE FOLLOWING! SHOULD BE OK.}  Even for ADTs there is a
-difference between which folds over them we can type when they are
-viewed as ADTs {\color{red} (i.e., as fixpoints of first-order
-  functors) EARLIER?}  versus as proper nested types {\color{red}
-  (i.e., as fixpoints of higher-order functors) EARLIER?}. This is
-because, in the return type of $\mathsf{fold}$, the arguments of the
-$\mu$-type must be variables bound by $\Nat$.  For ADTs, the
-$\mu$-type takes no arguments, making it possible to write recursive
-functions, such as a concatenation function for lists of type $\alpha
-; \emptyset \vdash \Nat^\emptyset \ (\mu \beta. \onet + \alpha \times
-\beta)\, (\Nat^\emptyset (\mu \beta. \onet + \alpha \times \beta) \,
-(\mu \beta. \onet + \alpha \times \beta))$.  This is not possible for
-nested types --- even when they are semantically equivalent to ADTs.
+variables occurring in the type of its final argument. Even for ADTs
+there is a difference between which folds over them we can type when
+they are viewed as ADTs (i.e., as fixpoints of first-order functors)
+versus as proper nested types (i.e., as fixpoints of higher-order
+functors). This is because, in the return type of $\mathsf{fold}$, the
+arguments of the $\mu$-type must be variables bound by $\Nat$.  For
+ADTs, the $\mu$-type takes no arguments, making it possible to write
+recursive functions, such as a concatenation function for lists of
+type $\alpha ; \emptyset \vdash \Nat^\emptyset \ (\mu \beta. \onet +
+\alpha \times \beta)\, (\Nat^\emptyset (\mu \beta. \onet + \alpha
+\times \beta) \, (\mu \beta. \onet + \alpha \times \beta))$.  This is
+not possible for nested types --- even when they are semantically
+equivalent to ADTs.
 
@@
 \begin{defi}\label{def:set-sem}
-{\color{red} Check clause for $\mu$-types.} The {\em set
-  interpretation} $\setsem{\cdot} : \F \to [\setenv, \set]$ is defined
-by:
+The {\em set interpretation} $\setsem{\cdot} : \F \to [\setenv, \set]$
+is defined by:
 \begin{align*}
@@
   \setsem{\Gamma;\Phi \vdash G}\rho\\ 
-  \setsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash (\mu \phi.\lambda
+  \setsem{\Gamma;\Phi \vdash (\mu \phi.\lambda
     \ol{\alpha}. H)\ol{G}}\rho &= (\mu
-    T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi,{\color{red} \ol\gamma} \vdash G}\rho}\\
+    T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi \vdash G}\rho}\\
     \text{where } T^\set_{H,\rho}\,F & = \lambda
-  \ol{A}. \setsem{\Gamma;{\color{red}\ol\gamma},\phi, \ol{\alpha} \vdash
+  \ol{A}. \setsem{\Gamma;\phi, \ol{\alpha} \vdash
     H}\rho[\phi :=  F][\ol{\alpha := A}]\\
   \text{and } T^\set_{H,\rho}\,\eta &= \lambda
-  \ol{A}. \setsem{\Gamma;{\color{red}\ol\gamma},\phi, \ol{\alpha} \vdash
+  \ol{A}. \setsem{\Gamma;\phi, \ol{\alpha} \vdash
     H}\id_\rho[\phi := \eta][\ol{\alpha := \id_{A}}]
@@
 admits a fixpoint.  Since $T_{H,\rho}^\set$ is defined in terms of
-$\setsem{\Gamma;{\color{red} \ol\gamma},\phi, \ol{\alpha} \vdash H}$, this means
-that interpretations of types must be such functors, which in turn
-means that the actions of set interpretations of types on objects and
-on morphisms in $\setenv$ are intertwined. Fortunately, we know
+$\setsem{\Gamma;\phi, \ol{\alpha} \vdash H}$, this means that
+interpretations of types must be such functors, which in turn means
+that the actions of set interpretations of types on objects and on
+morphisms in $\setenv$ are intertwined. Fortunately, we know
 from~\cite{jp19} that, for every $\Gamma; \ol{\alpha} \vdash G$,
@@
 where $k = |\ol \alpha|$. This means that for each $\setsem{\Gamma;
- {\color{red} \ol\gamma}, \phi^k, \ol{\alpha} \vdash H}$, the corresponding
-operator $T^\set_{H}$ can be extended to a {\em functor} from
-$\setenv$ to $[[\set^k,\set],[\set^k,\set]]$. The action of $T^\set_H$
-on an object $\rho \in \setenv$ is given by the higher-order functor
+  \phi^k, \ol{\alpha} \vdash H}$, the corresponding operator
+$T^\set_{H}$ can be extended to a {\em functor} from $\setenv$ to
+$[[\set^k,\set],[\set^k,\set]]$. The action of $T^\set_H$ on an object
+$\rho \in \setenv$ is given by the higher-order functor
 $T_{H,\rho}^\set$, whose actions on objects (functors in $[\set^k,
@@
 $\ol{A}$ is $(T^\set_{H,f}\, F)_{\ol{A}} = \setsem{\Gamma;
-  {\color{red}\ol\gamma},\phi,\ol{\alpha} \vdash H}f[\phi := \id_F][\ol{\alpha :=
-    \id_A}]$.
+  \phi,\ol{\alpha} \vdash H}f[\phi := \id_F][\ol{\alpha := \id_A}]$.
 
@@
     G}f$
-\item If \,$\Gamma;\Phi,{\color{red}\ol\gamma} \vdash (\mu \phi.\lambda
+\item If \,$\Gamma;\Phi \vdash (\mu \phi.\lambda
   \ol{\alpha}. H)\ol{G}$ then
   \[\begin{array}{lll}
-  \setsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash (\mu  \phi.\lambda
-    \ol{\alpha}. H)\ol{G}} f &: &\setsem{\Gamma;\Phi,{\color{red}\ol\gamma} 
+  \setsem{\Gamma;\Phi \vdash (\mu  \phi.\lambda
+    \ol{\alpha}. H)\ol{G}} f &: &\setsem{\Gamma;\Phi
     \vdash (\mu \phi.\lambda \ol{\alpha}. H)\ol{G}} \rho \to
-  \setsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash (\mu
+  \setsem{\Gamma;\Phi \vdash (\mu
     \phi.\lambda\ol{\alpha}. H)\ol{G}} \rho'\\
   &= &(\mu
-  T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash G}\rho} \to (\mu
-  T^\set_{H,\rho'})\ol{\setsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash G}\rho'}
+  T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi \vdash G}\rho} \to (\mu
+  T^\set_{H,\rho'})\ol{\setsem{\Gamma;\Phi \vdash G}\rho'}
   \end{array}\]
@@
   defined by
-  \[\begin{array}{ll} & (\mu T^\set_{H,f})\ol{\setsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash
-      G}\rho'} \circ (\mu T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi,{\color{red}\ol\gamma}
-      \vdash G}f}\\ = & (\mu T^\set_{H,\rho'})\ol{\setsem{\Gamma;\Phi,{\color{red}\ol\gamma}
-      \vdash G}f} \circ (\mu T^\set_{H,f})\ol{\setsem{\Gamma;\Phi,{\color{red}\ol\gamma}
+  \[\begin{array}{ll} & (\mu T^\set_{H,f})\ol{\setsem{\Gamma;\Phi \vdash
+      G}\rho'} \circ (\mu T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi \vdash G}f}\\ = & (\mu T^\set_{H,\rho'})\ol{\setsem{\Gamma;\Phi
+      \vdash G}f} \circ (\mu T^\set_{H,f})\ol{\setsem{\Gamma;\Phi
       \vdash G}\rho}\end{array}\]  The latter equality holds because $\mu
@@
 \begin{CD}
- (\mu T^\set_{H,\rho}) \ol{\setsem{\Gamma;\Phi,{\color{red} \ol\gamma} \vdash G}\rho} @> (\mu
-  T^\set_{H,f})_{\ol{\setsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash G}\rho}} >> (\mu
-  T^\set_{H,\rho'}) \ol{\setsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash G}\rho} \\ 
- @V(\mu T^\set_{H,\rho}) \ol{\setsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash G}f}VV @V  (\mu
- T^\set_{H,\rho'}) \ol{\setsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash G}f} VV \\ 
-(\mu T^\set_{H,\rho}) \ol{\setsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash G}\rho'} @>(\mu
- T^\set_{H,f})_{\ol{\setsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash G}\rho'}}>> (\mu
- T^\set_{H,\rho'}) \ol{\setsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash G}\rho'}
+ (\mu T^\set_{H,\rho}) \ol{\setsem{\Gamma;\Phi \vdash G}\rho} @> (\mu
+  T^\set_{H,f})_{\ol{\setsem{\Gamma;\Phi \vdash G}\rho}} >> (\mu
+  T^\set_{H,\rho'}) \ol{\setsem{\Gamma;\Phi \vdash G}\rho} \\ 
+ @V(\mu T^\set_{H,\rho}) \ol{\setsem{\Gamma;\Phi \vdash G}f}VV @V  (\mu
+ T^\set_{H,\rho'}) \ol{\setsem{\Gamma;\Phi \vdash G}f} VV \\ 
+(\mu T^\set_{H,\rho}) \ol{\setsem{\Gamma;\Phi \vdash G}\rho'} @>(\mu
+ T^\set_{H,f})_{\ol{\setsem{\Gamma;\Phi \vdash G}\rho'}}>> (\mu
+ T^\set_{H,\rho'}) \ol{\setsem{\Gamma;\Phi \vdash G}\rho'}
 \end{CD}
@@
 endofunctor on $RT_k$ need not be a functor on all of
-$[\rel^k,\rel]$. {\color{red} KEEP? In particular, although we can
-  define $T_{H,\rho}\, F$ for a relation transformer $F$ in
-  Definition~\ref{def:rel-sem} below, it is not clear how we could
-  define it for an arbitrary $F : [\rel^k,\rel]$. This observation
-  will also be critical in Section~\ref{sec:GADTs}.}
+$[\rel^k,\rel]$. In particular, although we can define $T_{H,\rho}\,
+F$ for a relation transformer $F$ in Definition~\ref{def:rel-sem}
+below, it is not clear how we could define it for an arbitrary $F :
+[\rel^k,\rel]$. This observation will be especially critical in
+Section~\ref{sec:GADTs}.
 
@@
   \relsem{\Gamma;\Phi \vdash G}\rho\\  
-   \relsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash (\mu \phi.\lambda
+   \relsem{\Gamma;\Phi \vdash (\mu \phi.\lambda
     \ol{\alpha}. H)\ol{G}}\rho
-  &= (\mu T_{H,\rho})\ol{\relsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash
-     G}\rho}\\
+  &= (\mu T_{H,\rho})\ol{\relsem{\Gamma;\Phi \vdash G}\rho}\\
   \text{where }	T_{H,\rho}
@@
     &= \lambda \ol{R}. \relsem{
-      \Gamma;{\color{red}\ol\gamma},\phi,\ol{\alpha} \vdash H}\rho[\phi :=
+      \Gamma;\phi,\ol{\alpha} \vdash H}\rho[\phi :=
     F][\ol{\alpha := R}]\\
@@
     &= \lambda \ol{R}. \relsem{
-      \Gamma;{\color{red}\ol\gamma},\phi,\ol{\alpha} \vdash H}\id_\rho[\phi :=
+      \Gamma;\phi,\ol{\alpha} \vdash H}\id_\rho[\phi :=
     \delta][\ol{\alpha := \id_{\ol{R}}}]
@@
 fixpoint. Since $T_{H,\rho}$ is defined in terms of
-$\relsem{\Gamma;{\color{red}\ol\gamma},\phi^k, \ol{\alpha} \vdash H}$, this means
+$\relsem{\Gamma;\phi, \ol{\alpha} \vdash H}$, this means
 that relational interpretations of types must be $\omega$-cocontinuous
@@
 T_{H,\rho'}\, F$ whose component at $\ol{R}$ is $(T_{H,f}\,
-F)_{\ol{R}} = \relsem{\Gamma; {\color{red}\ol\gamma},\phi,\ol{\alpha} \vdash
+F)_{\ol{R}} = \relsem{\Gamma; \phi,\ol{\alpha} \vdash
   H}f[\phi := \id_F][\ol{\alpha := \id_R}]$.
@@
 $\omega$-cocontinuity of $\sem{\Gamma;\Phi \vdash F}$ in each case,
-{\color{red} KEEP? together with Lemma~\ref{lem:fp} and
-  Equation~\ref{eq:mu} for $\mu$-types.}
+together with Lemma~\ref{lem:fp} and Equation~\ref{eq:mu} for
+$\mu$-types.
 
@@
 on the right commute, they must be equal.
-\begin{figure*}[ht]
-\vspace*{-0.1in}
-  \hspace*{-0.85in}
+%\begin{figure*}[ht]
+%\vspace*{-0.1in}
+%  \hspace*{-0.85in}
   \begin{minipage}[b]{0.25\linewidth}
@@
 \end{minipage}
-\end{figure*}
-
-\vspace*{-0.1in}
+%\end{figure*}
+
+%\vspace*{-0.1in}
 
@@
 \alpha_{\ol{B}} \circ F\ol{f} \rangle (F \ol{A}) = \graph{
-  \alpha_{\ol{B}} \circ F\ol{f} }^*$, so that $\graph{ \alpha }^*
+  \alpha_{\ol{B}} \circ F\ol{f} }^*$, so that
+\[\graph{ \alpha }^*
 \graph{ \overline{f} } = (F\overline{A}, G\overline{B},
@@
   \alpha_{\ol{B}} \circ F\ol{f} }^*) = \graph{ \alpha_{\ol{B}} \circ
-  F\ol{f} }$.  Finally, $\alpha_{\ol{B}} \circ F\ol{f} = G\ol{f} \circ
+  F\ol{f} }\]  Finally, $\alpha_{\ol{B}} \circ F\ol{f} = G\ol{f} \circ
 \alpha_{\ol{A}}$ by naturality of $\alpha$.
@@
   \Eq_{1_\set} = \Eq_{\setsem{\Gamma; \Phi \vdash \onet}\rho}$
-\item {\color{red} Use vector notation?} By definition,
-  $\relsem{\Gamma; \emptyset \vdash \Nat^{\overline\alpha} \,F\,G}
-  \Eq_{\rho}$ is the relation on $\setsem{\Gamma; \emptyset \vdash
-    \Nat^{\overline\alpha} \,F\,G} \rho$ relating $t$ and $t'$ if, for
-  all ${R_1 : \rel(A_1,B_1)},...,{R_k : \rel(A_k,B_k)}$,
+\item By definition, $\relsem{\Gamma; \emptyset \vdash
+  \Nat^{\overline\alpha} \,F\,G} \Eq_{\rho}$ is the relation on
+  $\setsem{\Gamma; \emptyset \vdash \Nat^{\overline\alpha} \,F\,G}
+  \rho$ relating $t$ and $t'$ if, for all ${R_1 :
+    \rel(A_1,B_1)},...,{R_k : \rel(A_k,B_k)}$, $(t_{\overline{A}},
+  t'_{\overline{B}})$ is a morphism from $\relsem{\Gamma;
+    \overline\alpha \vdash F} \Eq_{\rho}\overline{[\alpha := R]}$ to
+  $\relsem{\Gamma ; \overline\alpha \vdash G}
+  \Eq_{\rho}\overline{[\alpha := R]}$ in $\rel$.  To prove that this
+  relation is $\Eq_{\setsem{\Gamma; \emptyset \vdash
+      \Nat^{\overline\alpha} \,F\,G} \rho}$ we need to show that
   $(t_{\overline{A}}, t'_{\overline{B}})$ is a morphism from
   $\relsem{\Gamma; \overline\alpha \vdash F}
-  \Eq_{\rho}\overline{[\alpha := R]}$ to $\relsem{\Gamma ;
-    \overline\alpha \vdash G} \Eq_{\rho}\overline{[\alpha := R]}$ in
-  $\rel$.  To prove that this relation is $\Eq_{\setsem{\Gamma;
-      \emptyset \vdash \Nat^{\overline\alpha} \,F\,G} \rho}$ we need
-  to show that $(t_{\overline{A}}, t'_{\overline{B}})$ is a morphism
-  from $\relsem{\Gamma; \overline\alpha \vdash F}
   \Eq_{\rho}\overline{[\alpha := R]}$ to $\relsem{\Gamma ;
@@
 \begin{split}
-\relsem{\Gamma; \Phi, {\color{red}\ol\gamma} \vdash (\mu \phi.\lambda
+\relsem{\Gamma; \Phi \vdash (\mu \phi.\lambda
   \ol{\alpha}. H)\ol{F}}\Eq_{\rho} 
-&=(\mu {T_{H, \Eq_{\rho}}}) \,\ol{\relsem{\Gamma; \Phi, {\color{red} \ol\gamma}  \vdash F}\Eq_{\rho}}\\ 
-&= \colim{n \in \nat}{T^n_{H, \Eq_{\rho}} K_0}\, \ol{\relsem{\Gamma; \Phi, {\color{red}\ol\gamma} 
+&=(\mu {T_{H, \Eq_{\rho}}}) \,\ol{\relsem{\Gamma; \Phi \vdash F}\Eq_{\rho}}\\ 
+&= \colim{n \in \nat}{T^n_{H, \Eq_{\rho}} K_0}\, \ol{\relsem{\Gamma; \Phi 
   \vdash F}\Eq_{\rho}}\\
 &= \colim{n \in \nat}{ T^n_{H, \Eq_{\rho}} K_0 \,\ol{\Eq_{\setsem{\Gamma;
-    \Phi, {\color{red}\ol\gamma}  \vdash F}\rho}}}\\
+    \Phi \vdash F}\rho}}}\\
 &= \colim{n \in \nat}{(\Eq_{(T^\set_{H,\rho})^n K_0})^*
-  \ol{\Eq_{\setsem{\Gamma; \Phi, {\color{red}\ol\gamma}  \vdash F}\rho}}}\\
+  \ol{\Eq_{\setsem{\Gamma; \Phi \vdash F}\rho}}}\\
 &= \colim{n \in \nat}{\Eq_{(T^\set_{H,\rho})^n K_0 \,\ol{\setsem{\Gamma;
-        \Phi, {\color{red}\ol\gamma}  \vdash F}\rho}}}\\ 
+        \Phi \vdash F}\rho}}}\\ 
 &= \Eq_{\colim{n \in \nat}{ (T^\set_{H,\rho})^n K_0\,
-    \ol{\setsem{\Gamma; \Phi, {\color{red}\ol\gamma}  \vdash F}\rho}}}\\
-&= \Eq_{\setsem{\Gamma; \Phi, {\color{red}\ol\gamma}  \vdash (\mu \phi.\lambda
+    \ol{\setsem{\Gamma; \Phi \vdash F}\rho}}}\\
+&= \Eq_{\setsem{\Gamma; \Phi \vdash (\mu \phi.\lambda
       \ol{\alpha}. H)\ol{F}}\rho}
@@
 \begin{split}
-& \relsem{\Gamma; \Phi', \phi, \ol{\alpha} \vdash J} \Eq_{\rho} [\phi
+& \relsem{\Gamma; {\color{red}\Phi',} \phi, \ol{\alpha} \vdash J} \Eq_{\rho} [\phi
     := T^{n}_{H,\Eq_{\rho}} K_0] \overline{[\alpha := \Eq_A]} \\
-=\;\; & \relsem{\Gamma; \Phi', \phi, \ol{\alpha} \vdash J} \Eq_{\rho}
+=\;\; & \relsem{\Gamma; {\color{red}\Phi',} \phi, \ol{\alpha} \vdash J} \Eq_{\rho}
 [\phi := \Eq_{(T^\set_{H,\rho})^n K_0}] \overline{[\alpha := \Eq_A]}
@@
 \end{equation}
-(Notice that we don't know what's in the context $\Phi'$.)  We prove
-Equation~\ref{eq:iel-fix-point-intermediate1} by induction on $n$.
-The case $n=0$ is trivial, because $T^0_{H,\Eq_{\rho}} K_0 = K_0$ and
-$(T^\set_{H,\rho})^0 K_0 = K_0$; the inductive step is proved by the
-following sequence of equalities:
+{\color{red}(Notice that we don't know what's in the context
+  $\Phi'$.)}  We prove Equation~\ref{eq:iel-fix-point-intermediate1}
+by induction on $n$.  The case $n=0$ is trivial, because
+$T^0_{H,\Eq_{\rho}} K_0 = K_0$ and $(T^\set_{H,\rho})^0 K_0 = K_0$;
+the inductive step is proved by the following sequence of equalities:
 \[
@@
 \overline{\Eq_A} \\ 
-&= \relsem{\Gamma; \phi, \ol{\alpha}, {\color{red}\ol\gamma} \vdash H} \Eq_{\rho} [\phi
+&= \relsem{\Gamma; \phi, \ol{\alpha} \vdash H} \Eq_{\rho} [\phi
   := T^{n}_{H, \Eq_{\rho}} K_0] \overline{[\alpha :=
     \Eq_A]} \\ 
-&= \relsem{\Gamma; \phi, \ol{\alpha}, {\color{red}\ol\gamma} \vdash H} \Eq_{\rho} [\phi
+&= \relsem{\Gamma; \phi, \ol{\alpha} \vdash H} \Eq_{\rho} [\phi
   := \Eq_{(T^\set_{H,\rho})^n K_0}] \overline{[\alpha :=
     \Eq_A]} \\ 
-&= \relsem{\Gamma; \phi, \ol{\alpha}, {\color{red}\ol\gamma} \vdash H} \Eq_{\rho [\phi
+&= \relsem{\Gamma; \phi, \ol{\alpha} \vdash H} \Eq_{\rho [\phi
     := (T^\set_{H,\rho})^n K_0] \overline{[\alpha :=
       A]}} \\ 
-&= \Eq_{\setsem{\Gamma; \phi, \ol{\alpha}, {\color{red}\ol\gamma} \vdash H} \rho [\phi
+&= \Eq_{\setsem{\Gamma; \phi, \ol{\alpha} \vdash H} \rho [\phi
     := (T^\set_{H,\rho})^n K_0] \overline{[\alpha :=
@@
 \begin{split}
-& \relsem{\Gamma; \Phi', \phi, \ol{\alpha} \vdash \phi
+& \relsem{\Gamma; {\color{red} \Phi',} \phi, \ol{\alpha} \vdash \phi
     \ol{G}} \Eq_{\rho} [\phi := T^{n}_{H,\Eq_{\rho}} K_0]
@@
   \\
-&= T^{n}_{H,\Eq_{\rho}} K_0\, \overline{\relsem{\Gamma; \Phi',
+&= T^{n}_{H,\Eq_{\rho}} K_0\, \overline{\relsem{\Gamma; {\color{red} \Phi',}
       \phi, \ol{\alpha} \vdash G} \Eq_{\rho} [\phi :=
@@
         \Eq_A]}} \\ 
-&= T^{n}_{H,\Eq_{\rho}} K_0\, \overline{\relsem{\Gamma; \Phi',
+&= T^{n}_{H,\Eq_{\rho}} K_0\, \overline{\relsem{\Gamma; {\color{red} \Phi',}
       \phi, \ol{\alpha} \vdash G} \Eq_{\rho} [\phi :=
@@
         \Eq_A]}} \\ 
-&= T^{n}_{H,\Eq_{\rho}} K_0\, \overline{\relsem{\Gamma; \Phi',
+&= T^{n}_{H,\Eq_{\rho}} K_0\, \overline{\relsem{\Gamma; {\color{red}\Phi',}
       \phi, \ol{\alpha} \vdash G} \Eq_{\rho [\phi := (T^\set_{H,\rho})^{n}
@@
 &= T^{n}_{H,\Eq_{\rho}} K_0\, \overline{\Eq_{\setsem{\Gamma;
-        \Phi', \phi, \ol{\alpha} \,\vdash G} \rho [\phi :=
+        {\color{red}\Phi',} \phi, \ol{\alpha} \,\vdash G} \rho [\phi :=
         (T^\set_{H,\rho})^{n} K_0] \overline{[\alpha :=
@@
 &= (\Eq_{(T^\set_{H,\rho})^{n} K_0})^* \,\overline{\Eq_{\setsem{\Gamma;
-        \Phi', \phi, \ol{\alpha} \,\vdash G} \rho [\phi :=
+        {\color{red}\Phi',} \phi, \ol{\alpha} \,\vdash G} \rho [\phi :=
         (T^\set_{H,\rho})^{n} K_0] \overline{[\alpha :=
@@
 &= (\Eq_{(T^\set_{H,\rho})^{n} K_0})^* \overline{\relsem{\Gamma;
-      \Phi', \phi, \ol{\alpha} \vdash G} \Eq_{\rho} [\phi :=
+      {\color{red}\Phi',} \phi, \ol{\alpha} \vdash G} \Eq_{\rho} [\phi :=
       \Eq_{(T^\set_{H,\rho})^{n} K_0}] \overline{[\alpha :=
         \Eq_A]}} \\ 
-&= \relsem{\Gamma; \Phi', \phi, \ol{\alpha} \vdash \phi \ol{G}}
+&= \relsem{\Gamma; {\color{red}\Phi',} \phi, \ol{\alpha} \vdash \phi \ol{G}}
   \Eq_{\rho} [\phi := \Eq_{(T^\set_{H,\rho})^{n} K_0}]
@@
   The case $J = (\mu \psi.\, \lambda \ol{\beta}.\, G) \ol{K}$
-is proved by the sequence of equalities {\color{red} (where $\Phi' = \Phi'', \ol\gamma$)}:
+is proved by the sequence of equalities 
 \[
 \begin{split}
-& \relsem{\Gamma; {\color{red} \Phi'', \ol\gamma}, \phi, \ol{\alpha} \vdash (\mu \psi.\, \lambda \ol{\beta}.\, G) \ol{K}}
+& \relsem{\Gamma; {\color{red}\Phi',} \phi, \ol{\alpha} \vdash (\mu \psi.\, \lambda
+    \ol{\beta}.\, G) \ol{K}} 
   \Eq_{\rho} [\phi := T^{n}_{H,\Eq_{\rho}} K_0]
@@
 &= (\mu {T_{G, \Eq_{\rho}[\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]}})
-  \,\ol{\relsem{\Gamma; {\color{red}\Phi'', \ol\gamma}, \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
+  \,\ol{\relsem{\Gamma; {\color{red}\Phi',} \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
   [\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]} \\ 
 &= \colim{m \in \nat} {T^m_{G, \Eq_{\rho}[\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]}}\,K_0
-  \,(\ol{\relsem{\Gamma; {\color{red} \Phi'', \ol\gamma}, \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
+  \,(\ol{\relsem{\Gamma; {\color{red}\Phi',} \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
   [\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]}) \\ 
 &= \colim{m \in \nat} {T^m_{G, \Eq_{\rho}[\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]}}\,K_0
-  \,(\ol{\relsem{\Gamma; {\color{red}\Phi'', \ol\gamma}, \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
+  \,(\ol{\relsem{\Gamma; {\color{red}\Phi',} \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
   [\phi := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha := \Eq_A}]}) \\  %%
 &= \colim{m \in \nat} {T^m_{G, \Eq_{\rho}[\phi := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha := \Eq_A}]}}\,K_0
-  \,(\ol{\relsem{\Gamma; {\color{red}\Phi'', \ol\gamma}, \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
+  \,(\ol{\relsem{\Gamma; {\color{red}\Phi',} \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
   [\phi := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha := \Eq_A}]}) \\ 
 &= (\mu {T_{G, \Eq_{\rho}[\phi := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha := \Eq_A}]}})
-  \,\ol{\relsem{\Gamma; {\color{red}\Phi'', \ol\gamma}, \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
+  \,\ol{\relsem{\Gamma; {\color{red}\Phi',} \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
   [\phi := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha := \Eq_A}]} \\ 
-&= \relsem{\Gamma; {\color{red}\Phi'', \ol\gamma}, \phi, \ol{\alpha} \vdash (\mu \psi.\, \lambda \beta.\, G) \ol{K}}
+&= \relsem{\Gamma; {\color{red}\Phi',} \phi, \ol{\alpha} \vdash (\mu \psi.\, \lambda \beta.\, G) \ol{K}}
   \Eq_{\rho}[\phi := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha := \Eq_{A}}]
@@
         \Eq_A}]}\,K_0) \\
-&= \lambda \ol{R}. \relsem{\Gamma; {\color{red}\ol\gamma}, \phi, \ol\alpha, \psi,
+&= \lambda \ol{R}. \relsem{\Gamma; {\color{red}\phi, \ol\alpha,} \psi,
     \ol\beta \vdash G} \Eq_{\rho} [\phi := T^{n}_{H,\Eq_{\rho}}
@@
           \Eq_A}]}\,K_0][\overline{\beta := R}] \\
-&= \lambda \ol{R}. \relsem{\Gamma; {\color{red} \ol\gamma}, \phi, \ol\alpha, \psi,
+&= \lambda \ol{R}. \relsem{\Gamma; {\color{red}\phi, \ol\alpha,} \psi,
     \ol\beta \vdash G} \Eq_{\rho} [\phi := \Eq_{(T^{\set}_{H,\rho})^n
@@
 \end{align*}
-
 Here, the second equality holds by the induction hypothesis for
-Equation~\ref{eq:helper} on $m$. The fourth equality holds because,
-due to typing rule restrictions for the $\mu$ types, $\phi$ either
-does not appear in $G$, or must have arity $0$, in which case
-$\ol\alpha$ must be empty, if $\phi$ appears in $G$, and uses
-Equation~\ref{eq:iel-fix-point-intermediate2} for $G$ when $\phi$ has
-arity $0$. {\color{red} Note that if functorial variables of arity
-  greater than $0$ were allowed to appear in the bodies of
-  $\mu$-types, then the aboe clause of the IEL would fail.}
+Equation~\ref{eq:helper} on $m$. The fourth equality holds because
+$\phi$ and the variables in $\ol\alpha$ do not appear in
+$G$. 
+%due to typing rule restrictions for the $\mu$ types, $\phi$ either
+%does not appear in $G$, or must have arity $0$, in which case
+%$\ol\alpha$ must be empty, if $\phi$ appears in $G$, and uses
+%Equation~\ref{eq:iel-fix-point-intermediate2} for $G$ when $\phi$ has
+%arity $0$. {\color{red} Note that if functorial variables of arity
+%  greater than $0$ were allowed to appear in the bodies of
+%  $\mu$-types, then the aboe clause of the IEL would fail.}
 \end{itemize}
@@
 
-{\color{red} STOPPED HERE!!!!!}
-
 \section{Interpreting Terms}\label{sec:term-interp}
@@
 If $\Delta = x_1 : F_1,...,x_n : F_n$ is a term context for $\Gamma$
-and $\Phi$, then the interpretations $\setsem{\Gamma;\Phi \vdash \Delta}$ and
-$\relsem{\Gamma;\Phi \vdash \Delta}$ are defined by
+and $\Phi$, then the interpretations $\setsem{\Gamma;\Phi \vdash
+  \Delta}$ and $\relsem{\Gamma;\Phi \vdash \Delta}$ are defined by
 \[\begin{array}{lll}
@@
 Every well-formed term $\Gamma;\Phi~|~\Delta \vdash t : F$ then
-has, for every $\rho \in \setenv$, set interpretations
-$\setsem{\Gamma;\Phi~|~\Delta \vdash t : F}\rho$ as natural
-transformations from $\setsem{\Gamma; \Phi \vdash \Delta}\rho$ to
+has, for every $\rho \in \setenv$, a set interpretation
+$\setsem{\Gamma;\Phi~|~\Delta \vdash t : F}\rho$ as a natural
+transformation from $\setsem{\Gamma; \Phi \vdash \Delta}\rho$ to
 $\setsem{\Gamma; \Phi \vdash F}\rho$, and, for every $\rho \in
-\relenv$, relational interpretations $\relsem{\Gamma;\Phi~|~\Delta
-  \vdash t : F}\rho$ as natural transformations from
+\relenv$, a relational interpretation $\relsem{\Gamma;\Phi~|~\Delta
+  \vdash t : F}\rho$ as a natural transformation from
 $\relsem{\Gamma; \Phi \vdash \Delta}\rho$ to $\relsem{\Gamma; \Phi
-  \vdash F}\rho$. These are given in Definition~\ref{def:set-interp}.
+  \vdash F}\rho$.
 
@@
 relational interpretations. Specifically, for any $\rho \in \setenv$,
-$\setsem{\Gamma;\Phi \,|\, \Delta, x : F \vdash t : G}\rho =
+\[\setsem{\Gamma;\Phi \,|\, \Delta, x : F \vdash t : G}\rho =
 (\setsem{\Gamma;\Phi \,|\, \Delta \vdash t : G}\rho) \circ
-\pi_{\Delta}$, where $\pi_{\Delta}$ is the projection
+\pi_{\Delta}\] where $\pi_{\Delta}$ is the projection
 $\setsem{\Gamma;\Phi \vdash \Delta, x : F} \to
@@
       :=_{\ol{\beta}} G}]) \rrbracket^{\mathsf D} \rho & &\\
-%\hspace*{0.79in}  & & \hspace*{0.1in}\id_{\rho[\ol{\gamma :=
-%      C}]}[\ol{\phi := \lambda
-%    \ol{B}. (\dsem{\Gamma;\ol\alpha\,|\,\Delta \vdash t : 
-%    \Nat^{\ol{\beta},\ol{\gamma}}\,F\,G}\rho\,d)_{\ol{B}\,\ol{C}}}]\\
 \llbracket \Gamma;\emptyset \,|\, \emptyset \vdash \tin_H :
-Nat^{\ol{\beta},\ol{\gamma}} \, H[\phi := (\mu \phi.\lambda {\overline
+Nat^{\ol{\beta}{\color{red}, \ol{\gamma}}} \, H[\phi := (\mu \phi.\lambda {\overline
     \alpha}.H){\overline \beta}][\ol{\alpha := \beta}] & = &
 \lambda d\,\ol{B}\,
-\ol{C}.\,(\mathit{in}_{{T}^X_{H,\rho[\ol{\gamma :=
-        C}]}})_{\ol{B}} \;\;\mbox{ where } X \mbox{ is } \set \mbox{ when } \\ 
+{\color{red}\ol{C}}.\,(\mathit{in}_{{T}^X_{H,\rho{\color{red}[\ol{\gamma :=
+        C}]}}})_{\ol{B}} \;\;\mbox{ where } X \mbox{ is } \set \mbox{ when } \\ 
 \hspace*{0.79in}(\mu \phi.\lambda {\overline \alpha}.H){\overline
@@
 \llbracket \Gamma;\emptyset \,|\, \emptyset \vdash
-  \fold^F_H : \Nat^\emptyset\;(\Nat^{\ol{\beta}, \ol{\gamma}}\,H[\phi
+  \fold^F_H : \Nat^\emptyset\;(\Nat^{\ol{\beta}{\color{red}, \ol{\gamma}}}\,H[\phi
     :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F) & = &  
-\lambda d\,\eta\,\ol{B}\,\ol{C}.\,
-(\mathit{fold}_{T^X_{H,\rho[\ol{\gamma := C}]}} \, (\lambda
+\lambda d\,\eta\,\ol{B}{\color{red}\,\ol{C}}.\,
+(\mathit{fold}_{T^X_{H,\rho{\color{red}[\ol{\gamma := C}]}}} \, (\lambda
 \ol{A}.\,\eta_{\ol{A}\,\ol{C}}))_{\ol{B}}\\ 
-\hspace*{0.79in}(\Nat^{{\ol{\beta},\ol{\gamma}} }\,(\mu
+\hspace*{0.79in}(\Nat^{{\ol{\beta}{\color{red} ,\ol{\gamma}}} }\,(\mu
   \phi.\lambda \overline \alpha.H)\overline \beta\;F)
@@
 
-\begin{comment}
-\begin{defi}\label{def:rel-interp}
-If $\rho$ is a relation environment and $\Gamma;\Phi~|~\Delta \vdash t :
-\tau$ then $\relsem{\Gamma;\Phi~|~\Delta \vdash t : \tau}\rho$ is
-defined as follows:
-\[\begin{array}{lll}
-\relsem{\Gamma;\emptyset \,|\, \Delta,x :\tau \vdash x : \tau} \rho& = &
-\pi_{|\Delta|+1}\\
-%\relsem{\Gamma;\emptyset \,|\, \Delta \vdash \lambda x.t : \sigma \to \tau}\rho &
-%= & \curry (\relsem{\Gamma;\emptyset \,|\, \Delta, x : \sigma \vdash t :
-%  \tau}\rho)\\ 
-%\relsem{\Gamma;\emptyset \,|\, \Delta \vdash st: \tau} \rho & = & \eval
-%\circ \langle \relsem{\Gamma;\emptyset \,|\, \Delta \vdash s: \sigma \to
-%  \tau}\rho, \relsem{\Gamma;\emptyset \,|\, \Delta \vdash t: \sigma}\rho
-%\rangle\\
-\relsem{\Gamma;\emptyset \,|\, \Delta \vdash L_{\overline \alpha} x.t : \Nat^{\overline
-    \alpha} \,F \,G}\rho & = &  \curry (\relsem{\Gamma;\overline \alpha
-  \,|\, \Delta, x : F \vdash t: G}\rho[\overline{\alpha := \_}])\\
-\relsem{\Gamma;\Phi \,|\, \Delta \vdash t_{\overline{\tau}} s:
-  G [\overline{\alpha := \tau}]}\rho & = & \eval \circ \langle
-  \lambda e. \,(\relsem{\Gamma;\emptyset \,|\, \Delta \vdash t :
-  \Nat^{\overline{\alpha}} \,F \,G}\rho\; e)_{\overline{\relsem{\Gamma;\Phi
-      \vdash \tau}\rho}},\\ 
- & & \hspace*{0.5in} \relsem{\Gamma;\Phi \,|\,
-    \Delta \vdash s: F [\overline{\alpha := \tau}]}\rho \rangle\\ 
-& & \\
-%% \color{red} \mbox{Add rules for } \forall \mbox{ if we include it} & & \\
-\relsem{\Gamma;\Phi \,|\, \Delta,x :\tau \vdash x : \tau} \rho& = &
-\pi_{|\Delta|+1}\\
-\relsem{\Gamma;\Phi \,|\, \Delta \vdash \bot_\tau t : \tau} \rho& = &
-!^0_{\relsem{\Gamma;\Phi \vdash \tau}\rho} \circ
-  \relsem{\Gamma;\Phi~|~\Delta \vdash t : \zerot}\rho, \mbox{ where } \\
- & & \hspace*{0.1in} !^0_{\relsem{\Gamma;\Phi \vdash \tau}\rho}
-\mbox{ is the unique morphism from } 0\\
- & & \hspace*{0.1in} \mbox{ to } \relsem{\Gamma;\Phi \vdash \tau}\rho\\
-\relsem{\Gamma;\Phi \,|\, \Delta \vdash \top : \onet}\rho & = &
-!^{\relsem{\Gamma;\Phi\vdash \Delta}\rho}_1, \mbox{ where }
-!^{\relsem{\Gamma;\Phi\vdash \Delta}\rho}_1\\ 
-& & \hspace*{0.1in} \mbox{ is the unique morphism from }
-\relsem{\Gamma;\Phi\vdash \Delta}\rho \mbox{ to } 1\\ 
-\relsem{\Gamma;\Phi \,|\, \Delta \vdash (s,t) : \sigma \times \tau} \rho& = &
-\relsem{\Gamma;\Phi \,|\, \Delta \vdash s: \sigma} \rho\times
-\relsem{\Gamma;\Phi \,|\, \Delta \vdash t: \tau} \rho\\
-\relsem{\Gamma;\Phi \,|\, \Delta \vdash \pi_1 t : \sigma} \rho& = &
-\pi_1 \circ \relsem{\Gamma;\Phi \,|\, \Delta \vdash t : \sigma \times \tau}\rho\\
-\relsem{\Gamma;\Phi \,|\, \Delta \vdash \pi_2 t : \sigma}\rho & = &
-\pi_2 \circ \relsem{\Gamma;\Phi \,|\, \Delta \vdash t : \sigma \times
-  \tau} \rho\\
-\relsem{\Gamma;\Phi~|~\Delta \vdash \cse{t}{x \mapsto l}{y \mapsto r} :
-  \gamma}\rho & = & \eval \circ \langle \curry \,[\relsem{\Gamma;\Phi
-    \,|\, \Delta, x : \sigma \vdash l : \gamma}\rho,\\
-   & & \hspace*{0.79in} \relsem{\Gamma;\Phi \,|\, \Delta, y
-    : \tau \vdash r : \gamma}\rho],\\
-   & &  \hspace*{0.5in} \relsem{\Gamma;\Phi \,|\, \Delta \vdash t :
-  \sigma + \tau} \rho\rangle \\
-\relsem{\Gamma;\Phi \,|\, \Delta \vdash \inl \,s: \sigma + \tau} \rho& = &
-\inl \circ \relsem{\Gamma;\Phi \,|\, \Delta \vdash s: \sigma}\rho\\
-\relsem{\Gamma;\Phi \,|\, \Delta \vdash \inr \,t: \sigma + \tau}\rho & = & 
-\inr \circ \relsem{\Gamma;\Phi \,|\, \Delta \vdash t : \tau}\rho\\
-\llbracket \Gamma;\emptyset \,|\, \emptyset \vdash \map^{\ol{F},\ol{G}}_H :
-\Nat^\emptyset\;(\ol{\Nat^{\ol{\beta},\ol{\gamma}}\,F\,G})\;&
-= & \lambda d\, \ol{\eta}\,\ol{R}.\,
-\relsem{\Gamma; \ol{\phi},\ol{\gamma}\vdash H}\id_{\rho[\ol{\gamma :=
-      R}]}[\ol{\phi := \lambda \ol{S}.\eta_{\ol{S}\,\ol{R}}}]\\
-\hspace*{0.79in} (\Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\,H[\ol{\phi
-      :=_{\ol{\beta}} G}]) \rrbracket^\rel \rho & & \\
-\llbracket \Gamma;\emptyset \,|\, \emptyset \vdash \tin_H :
-Nat^{\ol{\beta},\ol{\gamma}} \, H[\phi := (\mu \phi.\lambda {\overline
-    \alpha}.H){\overline \beta}][\ol{\alpha := \beta}] & = &
-\lambda d\,\ol{R}\, \ol{S}.\,(\mathit{in}_{T_{\rho[\ol{\gamma := S}]}})_{\ol{R}}\\
-\hspace*{0.79in}(\mu \phi.\lambda {\overline \alpha}.H){\overline
-  \beta} \rrbracket^\rel \rho & & \\
-\llbracket \Gamma;\emptyset \,|\, \emptyset \vdash
-  \fold^F_H : \Nat^\emptyset\;(\Nat^{\ol{\beta}, \ol{\gamma}}\,H[\phi
-    :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F) & = &  
-\lambda d\,\eta\,\ol{R}\,\ol{S}.\, (\mathit{fold}_{T_{\rho[\ol{\gamma
-        := S}]}} \, (\lambda
-\ol{Z}.\,\eta_{\ol{Z}\,\ol{S}}))_{\ol{R}}\\
-\hspace*{0.79in}(\Nat^{{\ol{\beta},\ol{\gamma}} }\,(\mu
-  \phi.\lambda \overline \alpha.H)\overline \beta\;F)
-\rrbracket^\rel \rho & & 
-\end{array}\]
-\end{defi}
-{\color{blue} Add return type for fold in last clause? Should be
-  $\relsem{\Gamma;\ol\beta,\ol\gamma \vdash F}\rho[\ol{\gamma :=
-  C}]$.}
-
-This interpretation gives that $\relsem{\Gamma;\emptyset \,|\, \Delta
-  \vdash \lambda x.t : \sigma \to \tau}\rho = \curry
-(\relsem{\Gamma;\emptyset \,|\, \Delta, x : \sigma \vdash t :
-  \tau}\rho)$ and $\relsem{\Gamma;\emptyset \,|\, \Delta \vdash st:
-  \tau} \rho = \eval \circ \langle \relsem{\Gamma;\emptyset \,|\,
-  \Delta \vdash s: \sigma \to \tau}\rho, \relsem{\Gamma;\emptyset
-  \,|\, \Delta \vdash t: \sigma}\rho \rangle$, as expected.
-
-If $t$ is closed, i.e., if $\emptyset; \emptyset~|~\emptyset \vdash t
-: \tau$, then we write $\setsem{\vdash t : \tau}$ instead of
-$\setsem{\emptyset; \emptyset~|~\emptyset \vdash t : \tau}$.
-\end{comment}
-
-\begin{comment}
-  , and for any $\rho \in \relenv$, 
-\[ \relsem{\Gamma;\Phi \,|\, \Delta, x : \sigma \vdash t : \tau}\rho
-  = (\relsem{\Gamma;\Phi \,|\, \Delta \vdash t : \tau}\rho) \circ
-  \pi_{\Delta}\] where $\pi_{\Delta}$ is the projection
-  $\relsem{\Gamma;\Phi \vdash \Delta, x : \sigma} \to
-  \relsem{\Gamma;\Phi \vdash \Delta}$.
-\end{comment}
-
 The return type for the semantic fold is
-  $\dsem{\Gamma;\ol\beta,\ol\gamma \vdash F}\rho[\ol{\gamma :=
-    C}][\ol{\beta := B}]$.
-  This interpretation gives $\dsem{\Gamma;\emptyset \,|\, \Delta
+  $\dsem{\Gamma;\ol\beta,{\color{red}\ol\gamma} \vdash F}\rho{\color{red}[\ol{\gamma :=
+    C}]}[\ol{\beta := B}]$.
+This interpretation gives
+\[\dsem{\Gamma;\emptyset \,|\, \Delta
     \vdash \lambda x.t : F \to G}\rho = \curry
   (\dsem{\Gamma;\emptyset \,|\, \Delta, x : F \vdash t :
-    G}\rho)$ and $\dsem{\Gamma;\emptyset \,|\, \Delta \vdash st:
-    G} \rho = \eval \circ \langle \dsem{\Gamma;\emptyset \,|\,
-    \Delta \vdash s: F \to G}\rho, \dsem{\Gamma;\emptyset
-    \,|\, \Delta \vdash t: F}\rho \rangle$, so it specializes to
-  the standard interpretations for System F terms.  If $t$ is closed,
-  i.e., if $\emptyset; \emptyset~|~\emptyset \vdash t : F$, then we
-  write $\dsem{\vdash t : F}$ instead of $\dsem{\emptyset;
-    \emptyset~|~\emptyset \vdash t : F}$.
-%, and similarly for $\relsem{\emptyset; \emptyset~|~\emptyset \vdash
-%t : \tau}$.
-Moreover, if
-  $\Gamma,\alpha;\Phi \,|\, \Delta \vdash t : F$ and
-  $\Gamma;\Phi,\alpha \,|\, \Delta \vdash t' : F$ and $\Gamma;\Phi
-  \vdash G$ then
-\begin{itemize}
-\item $\setsem{\Gamma;\Phi \,|\, \Delta[\alpha := G] \vdash
-  t[\alpha := G] : F[\alpha := G]}\rho =
-  \setsem{\Gamma,\alpha;\Phi \,|\, \Delta \vdash t : F }\rho [
-    \alpha := \setsem{\Gamma;\Phi\vdash G}\rho ]$
-%\item $\relsem{\Gamma;\Phi \,|\, \Delta[\alpha := \sigma] \vdash
-%  t[\alpha := \sigma] : \tau[\alpha := \sigma]}\rho =
-%  \relsem{\Gamma,\alpha;\Phi \,|\, \Delta \vdash t : \tau }\rho [
-%    \alpha := \relsem{\Gamma;\Phi\vdash\sigma}\rho ]$
-\item $\setsem{\Gamma;\Phi \,|\, \Delta[\alpha := G] \vdash
-  t'[\alpha := G] : F[\alpha := G]}\rho =
-  \setsem{\Gamma;\Phi,\alpha \,|\, \Delta \vdash t' : F }\rho [
-    \alpha := \setsem{\Gamma;\Phi\vdash G}\rho ]$
-%\item $\relsem{\Gamma;\Phi \,|\, \Delta[\alpha := \sigma] \vdash
-%  t'[\alpha := \sigma] : \tau[\alpha := \sigma]}\rho =
-%  \relsem{\Gamma;\Phi,\alpha \,|\, \Delta \vdash t' : \tau }\rho [
-%    \alpha := \relsem{\Gamma;\Phi\vdash\sigma}\rho ]$
-\end{itemize}
-and if $\Gamma;\Phi \,|\, \Delta, x: G \vdash t : F$ and
-$\Gamma;\Phi \,|\, \Delta \vdash s : G$ then
-\begin{itemize}
-\item $\lambda A.\, \setsem{\Gamma;\Phi \,|\, \Delta \vdash t[x := s]
-  : F }\rho \, A = \lambda A.\,\setsem{\Gamma;\Phi \,|\, \Delta, x:
-  G \vdash t : F}\rho \,(A, \setsem{\Gamma;\Phi \,|\,
-  \Delta\vdash s: G}\rho\, A)$
-%\item $\lambda R.\,\relsem{\Gamma;\Phi \,|\, \Delta \vdash t[x := s] :
-%  \tau }\rho\,R = \lambda R.\,\relsem{\Gamma;\Phi \,|\, \Delta, x:
-%  \sigma \vdash t : \tau}\rho \,(R, \relsem{\Gamma;\Phi \,|\,
-%  \Delta\vdash s: \sigma}\rho\,R)$
-\end{itemize}
-Direct calculation reveals that the set interpretations of terms also
-satisfy
-\begin{itemize}
-\item $\setsem{\Gamma; \Phi~|~\Delta \vdash
-  (L_{\ol{\alpha}}x.t)_{\ol{K}}s} = \setsem{\Gamma; \Phi~|~\Delta
-  \vdash t [\ol{\alpha := K}][x := s]}$
-\end{itemize}
-Term extensionality with respect to types and terms --- i.e.,
-$\setsem{\Gamma;\Phi\vdash (L_\alpha x.t)_\alpha \top : F} =
-\setsem{\Gamma;\Phi \vdash t : F}$ and $\setsem{\Gamma;\Phi\vdash
-  (L_\alpha x.t)_\alpha x : F} = \setsem{\Gamma;\Phi \vdash t : F}$
-%$\relsem{\Gamma;\Phi\vdash (L_\alpha x.t)_\alpha t} =
-%\relsem{\Gamma;\Phi \vdash t}$, as well as
-%term extensionality
-%and $\relsem{\Gamma;\Phi\vdash
-%  (L_\alpha x.t)_\alpha \top} = \relsem{\Gamma;\Phi \vdash t}$
---- follow (when both sides of the equations are defined). Similar
-properties hold for relational interpretations.
-
-\subsection{Properties of Terms of $\Nat$-Type}\label{sec:Nat-type-terms}
-
-Define, for $\Gamma; \ol{\alpha} \vdash F$, the term $\id_F$ to be
-$\Gamma;\emptyset~|~\emptyset \vdash L_{\ol{\alpha}}x.x :
-\Nat^{\ol{\alpha}} F\,F$ and, for terms $\Gamma; \emptyset \,|\,
+  G}\rho)\]
+and
+\[\dsem{\Gamma;\emptyset \,|\, \Delta \vdash st:
+    G} \rho = \eval \circ \langle \dsem{\Gamma;\emptyset \,|\, \Delta
+  \vdash s: F \to G}\rho, \dsem{\Gamma;\emptyset \,|\, \Delta \vdash
+  t: F}\rho \rangle\]
+
+\vspace*{0.05in}
+
+\noindent
+so it specializes to the standard interpretations for System F terms.
+If $t$ is closed, i.e., if $\emptyset; \emptyset~|~\emptyset \vdash t
+: F$, then we write $\dsem{\vdash t : F}$ instead of $\dsem{\emptyset;
+  \emptyset~|~\emptyset \vdash t : F}$.  Moreover, term interpretation
+respects substitution for both functorial and non-functorial type
+variables, as well as term substitution, i.e., if $\Gamma,\alpha;\Phi
+\,|\, \Delta \vdash t : F$ and $\Gamma;\Phi,\alpha \,|\, \Delta \vdash
+t' : F$ and $\Gamma;\Phi \vdash G$ then
+\[\dsem{\Gamma;\Phi \,|\, \Delta[\alpha := G] \vdash t[\alpha :=
+    G] : F[\alpha := G]}\rho = \dsem{\Gamma,\alpha;\Phi \,|\, \Delta
+  \vdash t : F }\rho [ \alpha := \dsem{\Gamma;\Phi\vdash G}\rho ]\]
+and
+\[\dsem{\Gamma;\Phi \,|\, \Delta[\alpha := G] \vdash t'[\alpha :=
+    G] : F[\alpha := G]}\rho = \dsem{\Gamma;\Phi,\alpha \,|\, \Delta
+  \vdash t' : F }\rho [ \alpha := \dsem{\Gamma;\Phi\vdash G}\rho ]\]
+
+\vspace*{0.05in}
+
+\noindent
+Moreover, if $\Gamma;\Phi \,|\, \Delta, x: G \vdash t : F$ and $\Gamma;\Phi
+\,|\, \Delta \vdash s : G$ then
+\[\lambda A.\, \dsem{\Gamma;\Phi \,|\, \Delta \vdash t[x := s] :
+  F }\rho \, A = \lambda A.\,\dsem{\Gamma;\Phi \,|\, \Delta, x: G
+  \vdash t : F}\rho \,(A, \dsem{\Gamma;\Phi \,|\, \Delta\vdash s:
+  G}\rho\, A)\]
+
+\vspace*{0.05in}
+
+\noindent
+Direct calculation reveals that term interpretations also satisfy
+\[\dsem{\Gamma; \Phi~|~\Delta \vdash
+  (L_{\ol{\alpha}}x.t)_{\ol{K}}s} = \dsem{\Gamma; \Phi~|~\Delta \vdash
+  t [\ol{\alpha := K}][x := s]}\] Term extensionality for both types
+and terms --- i.e., $\dsem{\Gamma;\Phi\vdash (L_\alpha x.t)_\alpha
+  \top : F} = \dsem{\Gamma;\Phi \vdash t : F}$ and
+$\dsem{\Gamma;\Phi\vdash (L_\alpha x.t)_\alpha x : F} =
+\dsem{\Gamma;\Phi \vdash t : F}$ --- follow (when both sides of these
+equations are defined).
+
+\section{Free Theorems for Nested Types}\label{sec:free-theorems}
+
+\subsection{Consequences of Naturality}\label{sec:Nat-type-terms}
+
+Define, for $\Gamma; \ol{\alpha} \vdash F$, the {\em identity} $\id_F$
+on $F$ to be
+\[\id_F = \Gamma;\emptyset~|~\emptyset \vdash L_{\ol{\alpha}}x.x :
+\Nat^{\ol{\alpha}} F\,F\] and, for terms $\Gamma; \emptyset \,|\,
 \Delta \vdash t: \Nat^{\overline{\alpha}} F\,G$ and $\Gamma; \emptyset
 \,|\, \Delta \vdash s: \Nat^{\overline{\alpha}} G\,H$, the {\em
-  composition} $s \circ t$ of $t$ and $s$ to be $\Gamma;
+  composition} $s \circ t$ of $t$ and $s$ to be \[s \circ t = \Gamma;
 \emptyset\,|\, \Delta \vdash L_{\overline{\alpha}}
 x. s_{\overline{\alpha}}(t_{\overline{\alpha}}x):
-\Nat^{\overline{\alpha}} F\,H$. Then $\setsem{\Gamma; \emptyset \,|\,
+\Nat^{\overline{\alpha}} F\,H\] Then \[\setsem{\Gamma; \emptyset \,|\,
   \emptyset \vdash \id_{F} : \Nat^{\ol{\alpha}} F\,F} \rho\, \ast =
 \id_{\lambda \ol{A}. \setsem{\Gamma; \ol{\alpha} \vdash F} \rho
-  [\ol{\alpha := A}]}$ for any set environment $\rho$, and
-$\setsem{\Gamma; \emptyset \,|\, \Delta \vdash s \circ t:
+  [\ol{\alpha := A}]}\] for any set environment $\rho$, and
+\[\setsem{\Gamma; \emptyset \,|\, \Delta \vdash s \circ t:
   \Nat^{\overline{\alpha}} F\,H} = \setsem{\Gamma; \emptyset \,|\,
@@
 \setsem{\Gamma; \emptyset \,|\, \Delta \vdash t:
-  \Nat^{\overline{\alpha}} F\,G}$. Straightforward calculation shows
+  \Nat^{\overline{\alpha}} F\,G}\] Straightforward calculation shows
 that terms of $\Nat$-type behave as natural transformations with
@@
 \noindent
-When $s = \mathsf{in}_H$, and
-$\overline{\Gamma; \emptyset \,|\, t :
- \Nat^{\overline{\gamma}} F\, G}$,
-and $\xi = \Nat^{\overline{\gamma}} H[\phi :=
-  (\mu \phi. \lambda \overline{\alpha}. H)
-  \overline{\beta}][\overline{\alpha := F}]\, (\mu \phi. \lambda
-\overline{\alpha}. H)\overline{G}$, Theorem~\ref{eq:ft-from-nat}
-specializes to
-\begin{equation}\label{thm:subst}
-\begin{array}{l}
-\hspace*{0.135in}\setsem{\Gamma; \emptyset \,|\, \Delta \vdash
-((\map_{(\mu \phi. \lambda
-    \overline{\alpha}. H) \overline{\beta}}^{\overline{F},
-    \overline{G}})_\emptyset\, \overline{t}) \circ (L_{\overline{\gamma}} z.
-  (\tin_{H})_{\overline{F},
-    \overline{\gamma}} z) : \xi}\\
-= \setsem{\Gamma; \emptyset \,|\, \Delta
-\vdash (L_{\overline{\gamma}} z. (\tin_H)_{\overline{G},
-  \overline{\gamma}} z) \circ
-  ((\map_{H[\phi := (\mu \phi. \lambda \overline{\alpha}. H)
-      \overline{\beta}]}^{\overline{F},
-    \overline{G}})_\emptyset\, \overline{t})
-  : \xi}
-\end{array}
-\end{equation}
-\noindent
+%{\color{red} When $s = \mathsf{in}_H$, and
+%$\overline{\Gamma; \emptyset \,|\, t :
+% \Nat^{\overline{\gamma}} F\, G}$,
+%and $\xi = \Nat^{\overline{\gamma}} H[\phi :=
+%  (\mu \phi. \lambda \overline{\alpha}. H)
+%  \overline{\beta}][\overline{\alpha := F}]\, (\mu \phi. \lambda
+%\overline{\alpha}. H)\overline{G}$, Theorem~\ref{eq:ft-from-nat}
+%specializes to
+%\begin{equation}\label{thm:subst}
+%\begin{array}{l}
+%\hspace*{0.135in}\setsem{\Gamma; \emptyset \,|\, \Delta \vdash
+%((\map_{(\mu \phi. \lambda
+%    \overline{\alpha}. H) \overline{\beta}}^{\overline{F},
+%    \overline{G}})_\emptyset\, \overline{t}) \circ (L_{\overline{\gamma}} z.
+%  (\tin_{H})_{\overline{F},
+%  \overline{\gamma}} z) : \xi}\\
+%= \setsem{\Gamma; \emptyset \,|\, \Delta
+%\vdash (L_{\overline{\gamma}} z. (\tin_H)_{\overline{G},
+%  \overline{\gamma}} z) \circ
+%  ((\map_{H[\phi := (\mu \phi. \lambda \overline{\alpha}. H)
+%      \overline{\beta}]}^{\overline{F},
+%    \overline{G}})_\emptyset\, \overline{t})
+%  : \xi}
+%\end{array}
+%\end{equation}
+%\noindent
+%stating that $\map$ and $\tin$ commute.}  More generally,
 Theorem~\ref{eq:ft-from-nat} gives a family of free theorems that are
 consequences of naturality, and thus do not require the full power of
-parametricity. Other specific instances include the three
-$\mathtt{map}$/$\mathtt{reverse}$ theorems from
-Section~\ref{sec:intro}. The $\mathtt{map}$/$\mathtt{reverse}$ theorem
-for bushes, e.g., is
+parametricity. All of Wadler's free theorems apart from short cut
+fusion are in this family.  We can also prove that the interpretation
+of every $\map_H$ is a functor, and that $\map$ is itself a
+higher-order functor. For example, the former property can be stated
+as: if $\Gamma; \ol\alpha, \ol\gamma \vdash H$, $\ol{\Gamma; \emptyset
+  \,|\, \Delta \vdash g : \Nat^{\ol\gamma} F \, G}$, and $\ol{\Gamma;
+  \emptyset \,|\, \Delta \vdash f : \Nat^{\ol\gamma} G \, K}$, then
   \begin{align*}
-  &\setsem{\emptyset; \emptyset~|~ y : \Nat^\emptyset F \, G
-    \vdash ((\map_{\mathit{Bush}\,\alpha}^{{F},
-      {G}}P)_\emptyset\, y) \circ (L_{{\emptyset}}
-    z. \mathit{reverseBush}_{F}\, z) : \Nat^{{\emptyset}}
-    (\mathit{Bush} \, F) \, (\mathit{Bush} \, G) }\\ =\;
-  &\setsem{\emptyset; \emptyset~|~ y : \Nat^\emptyset F\, G
-    \vdash (L_{\emptyset} z. \mathit{reverseBush}_{G}\,z) \circ
-    ((\map_{\mathit{Bush} \, \alpha}^{F, G})_\emptyset\,
-           {y}) : \Nat^{\emptyset} (\mathit{Bush} \, F ) \,
-           (\mathit{Bush} \, G)}
-\end{align*}
-for any term $\vdash \mathit{reverseBush} : \Nat^\alpha (\mathit{Bush}
-\,\alpha) \, (\mathit{Bush} \, \alpha)$. We can also prove a theorem
-for the data type $\emptyset;\alpha,\gamma \vdash
-\mathit{Tree}\,\alpha\,\gamma$ from Section~\ref{sec:calculus} that is
-similar but only maps along some of its functorial variables:
-%  \begin{align*}
-%  &\setsem{\emptyset; \alpha~|~ y : \Nat^\emptyset \sigma \, \tau
-%    \vdash ((\map_{\mathit{PTree}\,\alpha}^{{\sigma},
-%      {\tau}})_{\emptyset} {y}) \circ (L_{{\emptyset}}
-%    z. \mathit{reverse}_{\sigma} z) : \Nat^{{\emptyset}}
-%    (\mathit{PTree} \, \sigma) \, (\mathit{PTree} \, \tau) }\\ =\;
-%  &\setsem{\emptyset; \alpha~|~ y : \Nat^\emptyset \sigma \, \tau
-%    \vdash (L_{\emptyset} z. \mathit{reverse}_{\tau} z) \circ
-%    ((\map_{\mathit{PTree} \, \alpha}^{{\sigma}, {\tau}})_{\emptyset} {y}) :
-%    \Nat^{\emptyset} (\mathit{PTree} \, \sigma ) \, (\mathit{PTree} \,
-%    \tau)}
-%\end{align*}
-%for any term $\vdash \mathit{reverse} : \Nat^\alpha (\mathit{PTree}
-%\,\alpha) \, (\mathit{PTree} \, \alpha)$,
-if $\vdash \mathit{reverseTree} : \Nat^{\alpha, \gamma} (\mathit{Tree}
-\,\alpha \, \gamma) \, (\mathit{Tree} \, \alpha \, \gamma)$ then
-\begin{align*}
-  &\setsem{\emptyset; \emptyset~|~y : \Nat^\gamma F \, G 
-  \vdash  ((\map_{\mathit{Tree}\,\alpha \, \gamma}^{F,
-      G})_\emptyset\, y) \circ (L_{{\gamma}} z. \mathit{reverseTree}_{F,
-      \gamma}\, z) : \Nat^{{\gamma}} 
-    (\mathit{Tree} \, F \, \gamma) \, (\mathit{Tree} \, G \,
-    \gamma) }\\ =\; &\setsem{\emptyset; \emptyset~|~y :
-    \Nat^\gamma F \, G \vdash (L_{\gamma}
-    z. \mathit{reverseTree}_{G, \gamma}\,z) \circ
-    ((\map_{\mathit{Tree} \, \alpha\, \gamma}^{{F},
-      {G}})_\emptyset\, y) : \Nat^{\gamma} (\mathit{Tree} \,
-    F \, \gamma) \, (\mathit{Tree} \, G\, \gamma)}
+&\setsem{
+\Gamma; \emptyset \,|\, \Delta \vdash 
+  (\map^{\ol{F}, \ol{K}}_H \,)_{\emptyset} \, \ol{(f \circ g)}
+  : \Nat^{\ol\gamma} H[\ol{\alpha := F}] \, H[\ol{\alpha := K}]} \\ 
+= \hspace{0.03in} 
+  &\setsem{
+  \Gamma; \emptyset \,|\, \Delta \vdash 
+  (\map^{\ol{G}, \ol{K}}_H \,)_{\emptyset} \, \ol{f} \circ 
+  (\map^{\ol{F}, \ol{G}}_H \,)_{\emptyset} \, \ol{g}
+  : \Nat^{\ol\gamma} H[\ol{\alpha := F}] \, H[\ol{\alpha := K}]}
 \end{align*}
 \noindent
-%Note this instance uses the following version of map:
-%$$\map_{\mathit{Tree} \,\alpha \, \gamma}^{\sigma, \tau} : 
-%\Nat^{\emptyset} (\Nat^\gamma \, \sigma \,\tau)\,
-%(\Nat^\gamma(\mathit{Tree}\,\sigma\,\gamma)\,(\mathit{Tree}\,\tau\,\gamma))$$
-%as opposed to 
-%$$\map_{\mathit{Tree} \,\alpha \, \gamma}^{\sigma, \sigma', \tau, \tau'} : 
-%\Nat^{\emptyset} ((\Nat^\emptyset \, \sigma \,\tau) \times
-%(\Nat^\emptyset\,\sigma'\,\tau')) \,
-%(\Nat^\emptyset\,(\mathit{Tree}\,\sigma\,\sigma')\,(\mathit{Tree}\,\tau\,\tau'))$$
-We observed in Section~\ref{sec:calculus} that $ \phi; \alpha \vdash
-\mathit{GRose}\,\phi\,\alpha$ cannot appear as the (co)domain of a
-$\Nat$-type binding both $\phi$ and $\alpha$. As a result, there are
-no instances of, say, a $\mathtt{map}$/$\mathtt{reverse}$ free theorem
-for this data type that maps replacements for both $\phi$ and
-$\alpha$. On the other hand, $\phi; \alpha \vdash
-\mathit{GRose}\,\phi\,\alpha$ does support a
-$\mathtt{map}$/$\mathtt{reverse}$ theorem that maps only a replacement
-for $\alpha$.
-
-As usual, analogous results hold for relation environments and
-relational interpretations.
+We can also prove the expected properties of $\map$, $\tin$, and
+$\fold$, and their interpretations, e.g., that uniqueness and the
+universal property of the interpretation of $\fold$, and that the
+interpretation of $\tin$ is an isomorphism.
+
+{\color{red} INCLUDE THEOREMS AND PROOFS?}
+
+{\color{red} WAS HERE, BUT CONSEQ OF NAT OR MORE GENERAL?} Further,
+our calculus admits no terms with the type $\Nat^\alpha \onet
+\,\alpha$ of the polymorphic bottom, and every closed term $g$ of type
+$\Nat^\alpha\alpha\,\alpha$ denotes the polymorphic identity
+function.
 
@@
 \proof
+{\color{red} ADD IN PROOF HERE}
 By structural induction on the type judgment for $t$. The overall
@@
 In this section we show how Theorem~\ref{thm:abstraction} can be used
-to prove more advanced free theorems than those of
-Section~\ref{sec:Nat-type-terms} in the presence of directly
-constructed nested types.
+to prove free theorems, such as that for the type of the standard
+$\mathit{filter}$ function for lists, that go beyond mere naturality,
+and extend them to those nested types for which analogous functions
+can be defined. In particular, we can extend short cut short cut
+fusion for lists~\cite{glp93} to nested types, thereby formally
+proving correctness of the categorically inspired theorem
+from~\cite{jg10}.
 
@@
 
-\vspace*{-0.08in}
-
 As in~\cite{jg10}, replacing $\onet$ with any type $\emptyset;\alpha
@@
 
+\section{Parametricity for GADTs}\label{sec:GADTs}
+
 \section{Conclusion and Directions for Future Work}\label{sec:conclusion}
+
+{\color{red} CHANGE? COMPARE AGAINST INTRO.}
 

### 2020-12-15 free-theorems-nesteds/20-mscs.tex
adding proofs in

@@
 below type constructors for other nested types; and truly nested
-types,
-%\footnote{Nested types that are defined over themselves are known
-%  as {\em truly nested types}.}
-such as the nested type {\small
+types, such as the nested type {\small
 \[\begin{array}{l}
@@
   --- and even without term-level fixpoints --- is already rather
-  involved.
-%  in part because, as
-%  Pitts~\cite{pit00} reminds us, ``it is hard to construct models of
-%  both impredicative polymorphism and fixpoint recursion'', and in
-%  part because constructing a parametric model even for a predicative
-%  calculus with primitive nested types is already rather involved.
-  On
-  the other hand, our calculus is strongly normalizing, so it perhaps
-  edges us toward the kind of provably total practical programming
-  language proposed in~\cite{wad89}.}  Our main contribution is the
-construction of a parametric model for our calculus. This is both
-delicate and challenging. To ensure the existence of semantic
-fixpoints interpreting nested types, and thus to establish a suitable
-Identity Extension Lemma, our type system must explicitly track
-functoriality of types, and cocontinuity conditions on the functors
-interpreting them must be appropriately threaded throughout the model
-construction. Our model validates all standard consequences of
-parametricity in the presence of primitive nested types, including the
-isomorphism of primitive ADTs and their Church encodings, and
-correctness of short cut fusion for nested types. The relationship
-between naturality and parametricity has long been of interest, and
-our inclusion of a primitive type of natural transformations allows us
-to clearly delineate those consequences of parametricity that follow
-from naturality, from those, such as short cut fusion for nested
-types, that require the full power of parametricity.
+  involved. On the other hand, our calculus is strongly normalizing,
+  so it perhaps edges us toward the kind of provably total practical
+  programming language proposed in~\cite{wad89}.}  Our main
+contribution is the construction of a parametric model for our
+calculus. This is both delicate and challenging. To ensure the
+existence of semantic fixpoints interpreting nested types, and thus to
+establish a suitable Identity Extension Lemma, our type system must
+explicitly track functoriality of types, and cocontinuity conditions
+on the functors interpreting them must be appropriately threaded
+throughout the model construction. Our model validates all standard
+consequences of parametricity in the presence of primitive nested
+types, including the isomorphism of primitive ADTs and their Church
+encodings, and correctness of short cut fusion for nested types. The
+relationship between naturality and parametricity has long been of
+interest, and our inclusion of a primitive type of natural
+transformations allows us to clearly delineate those consequences of
+parametricity that follow from naturality, from those, such as short
+cut fusion for nested types, that require the full power of
+parametricity.
 
@@
 $\fvars^0$ disjoint from $\Phi$, and $\phi^k \in \fvars^k \setminus
-\Phi$, then the set $\mcF^\Pos(V)$ of {\em functorial expressions}
-over $\Gamma$ and $\Phi$ are given in
-Definition~\ref{def:wftypes}. The notation there entails that an
-application $F F_1...F_k$ is allowed only when $F$ is a type variable
-of arity $k$, or $F$ is a subexpression of the form $\mu
-\phi^{k}.\lambda \alpha_1...\alpha_k.F'$. Moreover, if $F$ has arity
-$k$ then $F$ must be applied to exactly $k$ arguments.  Accordingly,
-an overbar indicates a sequence of subexpressions whose length matches
-the arity of the type applied to it.  Requiring that types are always
-in such \emph{$\eta$-long normal form} avoids having to consider
+\Phi$, then the set $\mcF$ of {\em functorial expressions} over
+$\Gamma$ and $\Phi$ are given in Definition~\ref{def:wftypes}. The
+notation there entails that an application $F F_1...F_k$ is allowed
+only when $F$ is a type variable of arity $k$, or $F$ is a
+subexpression of the form $\mu \phi^{k}.\lambda
+\alpha_1...\alpha_k.F'$. Moreover, if $F$ has arity $k$ then $F$ must
+be applied to exactly $k$ arguments.  Accordingly, an overbar
+indicates a sequence of subexpressions whose length matches the arity
+of the type applied to it.  Requiring that types are always in such
+\emph{$\eta$-long normal form} avoids having to consider
 $\beta$-conversion at the level of types. In a subexpression
@@
 functors. In Section~\ref{sec:conclusion} we discuss how more
-expressive types might be accommodated. {\color{red} Daniel: Check the
-  above carefully?}
+expressive types might be accommodated.
 
 Definition~\ref{def:wftypes} allows well-formed types to be functorial
-in no variables. Functorial variables
-%in a well-formed type $\tau$
-can also be demoted to non-functorial status: if\,$F[\phi :== \psi]$
-is the textual replacement of $\phi$ in $F$, then $\Gamma, \psi^k;
-\Phi \vdash F[\phi^k :== \psi^k]$ is derivable whenever $\Gamma; \Phi,
-\phi^k \vdash F$ is. The proof is by induction on the structure of
-$F$. In addition to textual replacement, we also have a
-substitution operation on types.
+in no variables. Functorial variables can also be demoted to
+non-functorial status: if\,$F[\phi :== \psi]$ is the textual
+replacement of $\phi$ in $F$, then $\Gamma, \psi^k; \Phi \vdash
+F[\phi^k :== \psi^k]$ is derivable whenever $\Gamma; \Phi, \phi^k
+\vdash F$ is. The proof is by induction on the structure of $F$. In
+addition to textual replacement, we also have a substitution operation
+on types.
 
@@
 \noindent
-{\color{red} Daniel: Agree?} The idea is that the arguments to $\phi$
-get substituted for the variables in $\ol\alpha$ in each $F$ replacing
-an occurrence of $\phi$. It is not hard to see that $\Gamma;\Phi
-\vdash H[\phi :=_{\ol{\alpha}} F]$. In addition, if \,$\Gamma, \phi^k;
-\Phi \vdash H$,\, $\Gamma; \ol\psi,\ol{\alpha} \vdash F$, $|\ol\alpha|
-= k$, and $\Phi \cap \ol\psi = \emptyset$, then $\Gamma,\ol\psi';\Phi
-\vdash H[\phi :=_{\ol{\alpha}} F[\ol{\psi :== \psi'}]]$.
+The idea is that the arguments to $\phi$ get substituted for the
+variables in $\ol\alpha$ in each $F$ replacing an occurrence of
+$\phi$. It is not hard to see that $\Gamma;\Phi \vdash H[\phi
+  :=_{\ol{\alpha}} F]$. In addition, if \,$\Gamma, \phi^k; \Phi \vdash
+H$,\, $\Gamma; \ol\psi,\ol{\alpha} \vdash F$, $|\ol\alpha| = k$, and
+$\Phi \cap \ol\psi = \emptyset$, then $\Gamma,\ol\psi';\Phi \vdash
+H[\phi :=_{\ol{\alpha}} F[\ol{\psi :== \psi'}]]$.
 
@@
 \end{adjustbox}
-%       \vspace*{-0.25in}
 \end{figure*}
@@
 \[\begin{array}{l}
-\AXC{$\Gamma;\alpha \vdash \mathit{List} \, \alpha$}
+\hspace*{-0.1in}\AXC{$\Gamma;\alpha \vdash \mathit{List} \, \alpha$}
 \AXC{$\Gamma;\emptyset \vdash \mathit{PTree}\,F$}
@@
 be empty, but because the polymorphic $\mathit{flatten}$ function is
-natural in some variable, say $\delta$, it cannot possibly have a type
-of the form $\Nat^\emptyset F\, G$ as would be required for it to be
-the function input to $\map$. Untypeability of this instance of $\map$
-is unsatisfactory in a polymorphic calculus, where we naturally expect
-to be able to manipulate entire polymorphic functions rather than just
+natural in $\beta$ it cannot possibly have a type of the form
+$\Nat^\emptyset F\, G$ as would be required for it to be the function
+input to $\map$. Untypeability of this instance of $\map$ is
+unsatisfactory in a polymorphic calculus, where we naturally expect to
+be able to manipulate entire polymorphic functions rather than just
 their monomorphic instances, but the ``extra'' variables $\ol \gamma$
@@
 that reverses the order of the leaves in a perfect tree. This function
-maps, e.g., the perfect tree $((1,2),(3,4))$ to $((4,3),(2,1))$
-{\color{red} CHANGE TO $pnil/pnode$ NOTATION?}, and
-can be defined as
+maps, e.g., the perfect tree
+\[\mathit{pnode}\,(\mathit{pnode}\,
+(\mathit{pleaf}\,1,\mathit{pleaf}\,2))\,(\mathit{pnode}\,
+(\mathit{pleaf}\,3,\mathit{pleaf}\,4) )\]
+to the perfect tree
+\[\mathit{pnode}\,(\mathit{pnode}\,
+(\mathit{pleaf}\,4,\mathit{pleaf}\,3))\,(\mathit{pnode}\,
+(\mathit{pleaf}\,2,\mathit{pleaf}\,1) )\] i.e., maps $((1,2),(3,4))$
+to $((4,3),(2,1))$.  It can be defined as
 \[\vdash (\fold_{\beta + \phi(\beta \times \beta)}^{\mathit{PTree}\,
@@
 \end{array}\]
-to
+to the bush
 \[\begin{array}{c}
@@
 \end{defi}
-%The interpretations in Definition~\ref{def:set-sem} respect weakening,
-%i.e., a type and its weakenings have the same set interpretations.
-%
-%= \setsem{\Gamma \vdash \sigma} \rho \to
-%\setsem{\Gamma \vdash \tau} \rho$, as expected {\color{blue} by IEL}.
 If $\rho \in \setenv$ and $\vdash F$ then we write $\setsem{\vdash F}$
@@
 \begin{split}
-& \relsem{\Gamma; {\color{red}\Phi',} \phi, \ol{\alpha} \vdash J} \Eq_{\rho} [\phi
+& \relsem{\Gamma; \phi, \ol{\alpha} \vdash J} \Eq_{\rho} [\phi
     := T^{n}_{H,\Eq_{\rho}} K_0] \overline{[\alpha := \Eq_A]} \\
-=\;\; & \relsem{\Gamma; {\color{red}\Phi',} \phi, \ol{\alpha} \vdash J} \Eq_{\rho}
+=\;\; & \relsem{\Gamma; \phi, \ol{\alpha} \vdash J} \Eq_{\rho}
 [\phi := \Eq_{(T^\set_{H,\rho})^n K_0}] \overline{[\alpha := \Eq_A]}
@@
 \end{equation}
-{\color{red}(Notice that we don't know what's in the context
-  $\Phi'$.)}  We prove Equation~\ref{eq:iel-fix-point-intermediate1}
-by induction on $n$.  The case $n=0$ is trivial, because
-$T^0_{H,\Eq_{\rho}} K_0 = K_0$ and $(T^\set_{H,\rho})^0 K_0 = K_0$;
-the inductive step is proved by the following sequence of equalities:
+We prove Equation~\ref{eq:iel-fix-point-intermediate1} by induction on
+$n$.  The case $n=0$ is trivial, because $T^0_{H,\Eq_{\rho}} K_0 =
+K_0$ and $(T^\set_{H,\rho})^0 K_0 = K_0$; the inductive step is proved
+by the following sequence of equalities:
 \[
@@
 \begin{split}
-& \relsem{\Gamma; {\color{red} \Phi',} \phi, \ol{\alpha} \vdash \phi
+& \relsem{\Gamma; \phi, \ol{\alpha} \vdash \phi
     \ol{G}} \Eq_{\rho} [\phi := T^{n}_{H,\Eq_{\rho}} K_0]
@@
   \\
-&= T^{n}_{H,\Eq_{\rho}} K_0\, \overline{\relsem{\Gamma; {\color{red} \Phi',}
+&= T^{n}_{H,\Eq_{\rho}} K_0\, \overline{\relsem{\Gamma; 
       \phi, \ol{\alpha} \vdash G} \Eq_{\rho} [\phi :=
@@
         \Eq_A]}} \\ 
-&= T^{n}_{H,\Eq_{\rho}} K_0\, \overline{\relsem{\Gamma; {\color{red} \Phi',}
+&= T^{n}_{H,\Eq_{\rho}} K_0\, \overline{\relsem{\Gamma; 
       \phi, \ol{\alpha} \vdash G} \Eq_{\rho} [\phi :=
@@
         \Eq_A]}} \\ 
-&= T^{n}_{H,\Eq_{\rho}} K_0\, \overline{\relsem{\Gamma; {\color{red}\Phi',}
+&= T^{n}_{H,\Eq_{\rho}} K_0\, \overline{\relsem{\Gamma; 
       \phi, \ol{\alpha} \vdash G} \Eq_{\rho [\phi := (T^\set_{H,\rho})^{n}
@@
 &= T^{n}_{H,\Eq_{\rho}} K_0\, \overline{\Eq_{\setsem{\Gamma;
-        {\color{red}\Phi',} \phi, \ol{\alpha} \,\vdash G} \rho [\phi :=
+        \phi, \ol{\alpha} \,\vdash G} \rho [\phi :=
         (T^\set_{H,\rho})^{n} K_0] \overline{[\alpha :=
@@
 &= (\Eq_{(T^\set_{H,\rho})^{n} K_0})^* \,\overline{\Eq_{\setsem{\Gamma;
-        {\color{red}\Phi',} \phi, \ol{\alpha} \,\vdash G} \rho [\phi :=
+        \phi, \ol{\alpha} \,\vdash G} \rho [\phi :=
         (T^\set_{H,\rho})^{n} K_0] \overline{[\alpha :=
@@
 &= (\Eq_{(T^\set_{H,\rho})^{n} K_0})^* \overline{\relsem{\Gamma;
-      {\color{red}\Phi',} \phi, \ol{\alpha} \vdash G} \Eq_{\rho} [\phi :=
+      \phi, \ol{\alpha} \vdash G} \Eq_{\rho} [\phi :=
       \Eq_{(T^\set_{H,\rho})^{n} K_0}] \overline{[\alpha :=
         \Eq_A]}} \\ 
-&= \relsem{\Gamma; {\color{red}\Phi',} \phi, \ol{\alpha} \vdash \phi \ol{G}}
+&= \relsem{\Gamma; \phi, \ol{\alpha} \vdash \phi \ol{G}}
   \Eq_{\rho} [\phi := \Eq_{(T^\set_{H,\rho})^{n} K_0}]
@@
 \begin{split}
-& \relsem{\Gamma; {\color{red}\Phi',} \phi, \ol{\alpha} \vdash (\mu \psi.\, \lambda
+& \relsem{\Gamma; \phi, \ol{\alpha} \vdash (\mu \psi.\, \lambda
     \ol{\beta}.\, G) \ol{K}} 
@@
 &= (\mu {T_{G, \Eq_{\rho}[\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]}})
-  \,\ol{\relsem{\Gamma; {\color{red}\Phi',} \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
+  \,\ol{\relsem{\Gamma; \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
   [\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]} \\ 
 &= \colim{m \in \nat} {T^m_{G, \Eq_{\rho}[\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]}}\,K_0
-  \,(\ol{\relsem{\Gamma; {\color{red}\Phi',} \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
+  \,(\ol{\relsem{\Gamma; \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
   [\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]}) \\ 
 &= \colim{m \in \nat} {T^m_{G, \Eq_{\rho}[\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]}}\,K_0
-  \,(\ol{\relsem{\Gamma; {\color{red}\Phi',} \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
+  \,(\ol{\relsem{\Gamma; \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
   [\phi := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha := \Eq_A}]}) \\  %%
 &= \colim{m \in \nat} {T^m_{G, \Eq_{\rho}[\phi := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha := \Eq_A}]}}\,K_0
-  \,(\ol{\relsem{\Gamma; {\color{red}\Phi',} \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
+  \,(\ol{\relsem{\Gamma; \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
   [\phi := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha := \Eq_A}]}) \\ 
 &= (\mu {T_{G, \Eq_{\rho}[\phi := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha := \Eq_A}]}})
-  \,\ol{\relsem{\Gamma; {\color{red}\Phi',} \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
+  \,\ol{\relsem{\Gamma; \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
   [\phi := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha := \Eq_A}]} \\ 
-&= \relsem{\Gamma; {\color{red}\Phi',} \phi, \ol{\alpha} \vdash (\mu \psi.\, \lambda \beta.\, G) \ol{K}}
+&= \relsem{\Gamma; \phi, \ol{\alpha} \vdash (\mu \psi.\, \lambda \beta.\, G) \ol{K}}
   \Eq_{\rho}[\phi := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha := \Eq_{A}}]
@@
         \Eq_A}]}\,K_0) \\
-&= \lambda \ol{R}. \relsem{\Gamma; {\color{red}\phi, \ol\alpha,} \psi,
+&= \lambda \ol{R}. \relsem{\Gamma; \psi,
     \ol\beta \vdash G} \Eq_{\rho} [\phi := T^{n}_{H,\Eq_{\rho}}
@@
           \Eq_A}]}\,K_0][\overline{\beta := R}] \\
-&= \lambda \ol{R}. \relsem{\Gamma; {\color{red}\phi, \ol\alpha,} \psi,
+&= \lambda \ol{R}. \relsem{\Gamma; \psi,
     \ol\beta \vdash G} \Eq_{\rho} [\phi := \Eq_{(T^{\set}_{H,\rho})^n
@@
 \id_{\rho'}) : \graph{f} \to \Eq_{\rho'}$ and $(\id_{\rho}, f) :
-\Eq_{\rho} \to \graph{f}$ of relation environments gives that
-$(\setsem{\Gamma; \Phi \vdash F}f, \setsem{\Gamma; \Phi \vdash
-  F}\id_{\rho'}) = \relsem{\Gamma; \Phi \vdash F} (f, \id_{\rho'}) :
+\Eq_{\rho} \to \graph{f}$ of relation environments gives 
+\[\begin{array}{ll}
+  & (\setsem{\Gamma; \Phi \vdash F}f, \setsem{\Gamma; \Phi \vdash
+  F}\id_{\rho'})\\
+= & \relsem{\Gamma; \Phi \vdash F} (f, \id_{\rho'}) :
 \relsem{\Gamma; \Phi \vdash F}\graph{f} \to \relsem{\Gamma; \Phi
-  \vdash F}\Eq_{\rho'}$ and $(\setsem{\Gamma; \Phi \vdash
-  F}\id_{\rho}, \setsem{\Gamma; \Phi \vdash F}f) = \relsem{\Gamma;
-  \Phi \vdash F} (\id_{\rho}, f) : \relsem{\Gamma; \Phi \vdash
-  F}\Eq_{\rho} \to \relsem{\Gamma; \Phi \vdash F}\graph{f}$.
+  \vdash F}\Eq_{\rho'}
+\end{array}\] and
+\[\begin{array}{ll}
+  & (\setsem{\Gamma; \Phi \vdash F}\id_{\rho}, \setsem{\Gamma; \Phi
+  \vdash F}f)\\
+= & \relsem{\Gamma; \Phi \vdash F} (\id_{\rho}, f) : \relsem{\Gamma;
+  \Phi \vdash F}\Eq_{\rho} \to \relsem{\Gamma; \Phi \vdash
+  F}\graph{f}
+\end{array}\]
+
 Expanding the first equation gives that if $(x,y) \in \relsem{\Gamma;
-  \Phi \vdash F}\graph{f}$ then $(\setsem{\Gamma; \Phi \vdash F} f\,
+  \Phi \vdash F}\graph{f}$ then
+\[(\setsem{\Gamma; \Phi \vdash F} f\,
 x, \setsem{\Gamma; \Phi \vdash F}\id_{\rho'}\, y) \in \relsem{\Gamma;
-  \Phi \vdash F}\Eq_{\rho'}$. So $\setsem{\Gamma; \Phi \vdash
+  \Phi \vdash F}\Eq_{\rho'}\] So $\setsem{\Gamma; \Phi \vdash
   F}\id_{\rho'}\, y = \id_{\setsem{\Gamma; \Phi \vdash F}\rho'}\, y =
@@
 i.e., $\setsem{\Gamma; \Phi \vdash F} f\, x = y$, i.e., $(x, y) \in
-\graph{\setsem{\Gamma; \Phi \vdash F} f}$.  So, we have that
+\graph{\setsem{\Gamma; \Phi \vdash F} f}$.  So, we have
 $\relsem{\Gamma; \Phi \vdash F}\graph{f} \subseteq
-\graph{\setsem{\Gamma; \Phi \vdash F}f}$.  Expanding the second
-equation gives that if $x \in \setsem{\Gamma; \Phi \vdash F}\rho$ then
-$(\setsem{\Gamma; \Phi \vdash F}\id_{\rho}\, x,$\\
-$\setsem{\Gamma; \Phi
-  \vdash F} f\, x) \in \relsem{\Gamma; \Phi \vdash F}\graph{f}$.  Then
-$\setsem{\Gamma; \Phi \vdash F}\id_{\rho}\, x = \id_{\setsem{\Gamma;
-    \Phi \vdash F}\rho} x = x$, so for any $x \in \setsem{\Gamma; \Phi
-  \vdash F}\rho$ we have that $(x, \setsem{\Gamma; \Phi \vdash F}f\,
-x) \in \relsem{\Gamma; \Phi \vdash F}\graph{f}$.  Moreover, $x \in
-\setsem{\Gamma; \Phi \vdash F}\rho$ if and only if $(x,
-\setsem{\Gamma; \Phi \vdash F} f\, x) \in \graph{\setsem{\Gamma; \Phi
-    \vdash F}f}$ and, if $x \in \setsem{\Gamma; \Phi \vdash F}\rho$
+\graph{\setsem{\Gamma; \Phi \vdash F}f}$.
+
+Expanding the second equation gives that if $x \in \setsem{\Gamma;
+  \Phi \vdash F}\rho$ then
+\[(\setsem{\Gamma; \Phi \vdash F}\id_{\rho}\, x, \setsem{\Gamma;
+  \Phi \vdash F} f\, x) \in \relsem{\Gamma; \Phi \vdash F}\graph{f}\]
+Then $\setsem{\Gamma; \Phi \vdash F}\id_{\rho}\, x =
+\id_{\setsem{\Gamma; \Phi \vdash F}\rho} x = x$, so for any $x \in
+\setsem{\Gamma; \Phi \vdash F}\rho$ we have that $(x, \setsem{\Gamma;
+  \Phi \vdash F}f\, x) \in \relsem{\Gamma; \Phi \vdash F}\graph{f}$.
+Moreover, $x \in \setsem{\Gamma; \Phi \vdash F}\rho$ if and only if
+$(x, \setsem{\Gamma; \Phi \vdash F} f\, x) \in \graph{\setsem{\Gamma;
+    \Phi \vdash F}f}$ and, if $x \in \setsem{\Gamma; \Phi \vdash
+  F}\rho$ then $(x, \setsem{\Gamma; \Phi \vdash F} f\, x) \in
+\relsem{\Gamma; \Phi \vdash F} \graph{f}$. Thus, if $(x, \setsem{\Gamma;
+  \Phi \vdash F} f\, x) \in \graph{\setsem{\Gamma; \Phi \vdash F}f}$
 then $(x, \setsem{\Gamma; \Phi \vdash F} f\, x) \in \relsem{\Gamma;
-  \Phi \vdash F} \graph{f}$, so if $(x, \setsem{\Gamma; \Phi \vdash F}
-f\, x) \in \graph{\setsem{\Gamma; \Phi \vdash F}f}$ then $(x,
-\setsem{\Gamma; \Phi \vdash F} f\, x) \in \relsem{\Gamma; \Phi \vdash
-  F} \graph{f}$, i.e., $\graph{\setsem{\Gamma; \Phi \vdash F}f}
-\subseteq \relsem{\Gamma; \Phi \vdash F} \graph{f}$.
-\qed
+  \Phi \vdash F} \graph{f}$, i.e., $\graph{\setsem{\Gamma; \Phi \vdash
+    F}f} \subseteq \relsem{\Gamma; \Phi \vdash F} \graph{f}$.  \qed
 
@@
 
-\subsection{Basic Properties of Term Interpretations}
-
-The interpretations in Definition~\ref{def:set-interp} respect
-weakening, i.e., a term and its weakenings all have the same set and
-relational interpretations. Specifically, for any $\rho \in \setenv$,
+If $t$ is closed, i.e., if $\emptyset; \emptyset~|~\emptyset \vdash t
+: F$, then we write $\dsem{\vdash t : F}$ instead of $\dsem{\emptyset;
+  \emptyset~|~\emptyset \vdash t : F}$.  The interpretations in
+Definition~\ref{def:set-interp} respect weakening, i.e., a term and
+its weakenings all have the same set and relational
+interpretations. Specifically, for any $\rho \in \setenv$,
 \[\setsem{\Gamma;\Phi \,|\, \Delta, x : F \vdash t : G}\rho =
@@
 so it specializes to the standard interpretations for System F terms.
-If $t$ is closed, i.e., if $\emptyset; \emptyset~|~\emptyset \vdash t
-: F$, then we write $\dsem{\vdash t : F}$ instead of $\dsem{\emptyset;
-  \emptyset~|~\emptyset \vdash t : F}$.  Moreover, term interpretation
-respects substitution for both functorial and non-functorial type
-variables, as well as term substitution, i.e., if $\Gamma,\alpha;\Phi
-\,|\, \Delta \vdash t : F$ and $\Gamma;\Phi,\alpha \,|\, \Delta \vdash
-t' : F$ and $\Gamma;\Phi \vdash G$ then
+Term interpretation also respects substitution for both functorial and
+non-functorial type variables, as well as term substitution. That is,
+if $\Gamma,\alpha;\Phi \,|\, \Delta \vdash t : F$ and
+$\Gamma;\Phi,\alpha \,|\, \Delta \vdash t' : F$ and $\Gamma;\Phi
+\vdash G$ then
 \[\dsem{\Gamma;\Phi \,|\, \Delta[\alpha := G] \vdash t[\alpha :=
@@
 \noindent
-Moreover, if $\Gamma;\Phi \,|\, \Delta, x: G \vdash t : F$ and $\Gamma;\Phi
+and if $\Gamma;\Phi \,|\, \Delta, x: G \vdash t : F$ and $\Gamma;\Phi
 \,|\, \Delta \vdash s : G$ then
@@
 \noindent
-Direct calculation reveals that term interpretations also satisfy
+Direct calculation further reveals that term interpretations also
+satisfy
 \[\dsem{\Gamma; \Phi~|~\Delta \vdash
@@
   t [\ol{\alpha := K}][x := s]}\] Term extensionality for both types
-and terms --- i.e., $\dsem{\Gamma;\Phi\vdash (L_\alpha x.t)_\alpha
-  \top : F} = \dsem{\Gamma;\Phi \vdash t : F}$ and
-$\dsem{\Gamma;\Phi\vdash (L_\alpha x.t)_\alpha x : F} =
-\dsem{\Gamma;\Phi \vdash t : F}$ --- follow (when both sides of these
-equations are defined).
+and terms --- i.e., \[\dsem{\Gamma;\Phi\vdash (L_\alpha x.t)_\alpha
+  \top : F} = \dsem{\Gamma;\Phi \vdash t : F}\] and
+\[\dsem{\Gamma;\Phi\vdash (L_\alpha x.t)_\alpha x : F} =
+\dsem{\Gamma;\Phi \vdash t : F}\] --- then follow (assuming both sides
+of these equations are defined).
 
@@
 
-\subsection{Consequences of Naturality}\label{sec:Nat-type-terms}
-
-Define, for $\Gamma; \ol{\alpha} \vdash F$, the {\em identity} $\id_F$
-on $F$ to be
-\[\id_F = \Gamma;\emptyset~|~\emptyset \vdash L_{\ol{\alpha}}x.x :
-\Nat^{\ol{\alpha}} F\,F\] and, for terms $\Gamma; \emptyset \,|\,
-\Delta \vdash t: \Nat^{\overline{\alpha}} F\,G$ and $\Gamma; \emptyset
-\,|\, \Delta \vdash s: \Nat^{\overline{\alpha}} G\,H$, the {\em
-  composition} $s \circ t$ of $t$ and $s$ to be \[s \circ t = \Gamma;
-\emptyset\,|\, \Delta \vdash L_{\overline{\alpha}}
-x. s_{\overline{\alpha}}(t_{\overline{\alpha}}x):
-\Nat^{\overline{\alpha}} F\,H\] Then \[\setsem{\Gamma; \emptyset \,|\,
+\subsection{Naturality and Its Consequences}\label{sec:Nat-type-terms}
+
+If $\Gamma; \ol{\alpha} \vdash F$ then the {\em identity} $\id_F$ on
+$F$ is defined by $\id_F = \Gamma;\emptyset~|~\emptyset \vdash
+L_{\ol{\alpha}}x.x : \Nat^{\ol{\alpha}} F\,F$, and if $\Gamma;
+\emptyset \,|\, \Delta \vdash t: \Nat^{\overline{\alpha}} F\,G$ and
+$\Gamma; \emptyset \,|\, \Delta \vdash s: \Nat^{\overline{\alpha}}
+G\,H$ are terms then the {\em composition} $s \circ t$ of $t$ and $s$
+is defined by $s \circ t = \Gamma; \emptyset\,|\, \Delta \vdash
+L_{\overline{\alpha}} x. s_{\overline{\alpha}}(t_{\overline{\alpha}}x):
+\Nat^{\overline{\alpha}} F\,H$. Then \[\setsem{\Gamma; \emptyset \,|\,
   \emptyset \vdash \id_{F} : \Nat^{\ol{\alpha}} F\,F} \rho\, \ast =
@@
 \setsem{\Gamma; \emptyset \,|\, \Delta \vdash t:
-  \Nat^{\overline{\alpha}} F\,G}\] Straightforward calculation shows
-that terms of $\Nat$-type behave as natural transformations with
-respect to their source and target functorial types, i.e., we have
+  \Nat^{\overline{\alpha}} F\,G}\]
+Straightforward calculation shows that terms of $\Nat$-type behave as
+natural transformations with respect to their source and target
+functorial types, i.e., that
 \begin{thm}\label{eq:ft-from-nat} 
@@
 \noindent
-%{\color{red} When $s = \mathsf{in}_H$, and
-%$\overline{\Gamma; \emptyset \,|\, t :
-% \Nat^{\overline{\gamma}} F\, G}$,
-%and $\xi = \Nat^{\overline{\gamma}} H[\phi :=
-%  (\mu \phi. \lambda \overline{\alpha}. H)
-%  \overline{\beta}][\overline{\alpha := F}]\, (\mu \phi. \lambda
-%\overline{\alpha}. H)\overline{G}$, Theorem~\ref{eq:ft-from-nat}
-%specializes to
-%\begin{equation}\label{thm:subst}
-%\begin{array}{l}
-%\hspace*{0.135in}\setsem{\Gamma; \emptyset \,|\, \Delta \vdash
-%((\map_{(\mu \phi. \lambda
-%    \overline{\alpha}. H) \overline{\beta}}^{\overline{F},
-%    \overline{G}})_\emptyset\, \overline{t}) \circ (L_{\overline{\gamma}} z.
-%  (\tin_{H})_{\overline{F},
-%  \overline{\gamma}} z) : \xi}\\
-%= \setsem{\Gamma; \emptyset \,|\, \Delta
-%\vdash (L_{\overline{\gamma}} z. (\tin_H)_{\overline{G},
-%  \overline{\gamma}} z) \circ
-%  ((\map_{H[\phi := (\mu \phi. \lambda \overline{\alpha}. H)
-%      \overline{\beta}]}^{\overline{F},
-%    \overline{G}})_\emptyset\, \overline{t})
-%  : \xi}
-%\end{array}
-%\end{equation}
-%\noindent
-%stating that $\map$ and $\tin$ commute.}  More generally,
-Theorem~\ref{eq:ft-from-nat} gives a family of free theorems that are
-consequences of naturality, and thus do not require the full power of
-parametricity. All of Wadler's free theorems apart from short cut
-fusion are in this family.  We can also prove that the interpretation
-of every $\map_H$ is a functor, and that $\map$ is itself a
-higher-order functor. For example, the former property can be stated
-as: if $\Gamma; \ol\alpha, \ol\gamma \vdash H$, $\ol{\Gamma; \emptyset
-  \,|\, \Delta \vdash g : \Nat^{\ol\gamma} F \, G}$, and $\ol{\Gamma;
-  \emptyset \,|\, \Delta \vdash f : \Nat^{\ol\gamma} G \, K}$, then
+
+Theorem~\ref{eq:ft-from-nat} gives rise to an entire family of free
+theorems that are consequences of naturality, and thus do not require
+the full power of parametricity. In particular, we can prove all of
+the standard properties of the initial algebraic constructs $\map$,
+$\tin$, and $\fold$. We have, for example, that the interpretation of
+every $\map_H$ is a functor, i.e., if $\Gamma; \ol\alpha, \ol\gamma
+\vdash H$, $\ol{\Gamma; \emptyset \,|\, \Delta \vdash g :
+  \Nat^{\ol\gamma} F \, G}$, and $\ol{\Gamma; \emptyset \,|\, \Delta
+  \vdash f : \Nat^{\ol\gamma} G \, K}$, then
   \begin{align*}
@@
 \end{align*}
-\noindent
-We can also prove the expected properties of $\map$, $\tin$, and
-$\fold$, and their interpretations, e.g., that uniqueness and the
-universal property of the interpretation of $\fold$, and that the
-interpretation of $\tin$ is an isomorphism.
-
-{\color{red} INCLUDE THEOREMS AND PROOFS?}
-
-{\color{red} WAS HERE, BUT CONSEQ OF NAT OR MORE GENERAL?} Further,
-our calculus admits no terms with the type $\Nat^\alpha \onet
-\,\alpha$ of the polymorphic bottom, and every closed term $g$ of type
-$\Nat^\alpha\alpha\,\alpha$ denotes the polymorphic identity
-function.
-
-\subsection{Properties of Initial Algebraic Constructs}\label{sec:iaps}
-
-Our semantics interprets applications of $\map$ as applications of
-semantic $\mathit{map}$ functions.
-\begin{comment}
-Indeed, if \(\Gamma; \ol{\phi},
-\ol{\gamma} \vdash H\), \(\ol{\Gamma; \emptyset, \ol{\gamma} \vdash
-  F}\), and \(\ol{\Gamma; \Phi, \emptyset, \ol{\gamma} \vdash G}\),
-%and $\Gamma;\Phi\,|\,\Delta \vdash t :
-%\Nat^{\ol{\beta},\ol{\gamma}}\,F\,G$
-then 
-Definition~\ref{def:set-interp} gives
-%\begin{multline}\label{eq:map-sem-def}
-\begin{align}
-& \setsem{
-\Gamma;\Phi \,|\, \Delta
-\vdash \map^{\ol{F},\ol{G}}_H \,\ol t
-: %\Nat^\emptyset\;(\ol{\Nat^{\ol{\beta},\ol{\gamma}}\,F\,G})\; {}
-\Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\,H[\ol{\phi :=_{\ol{\beta}} G}]
-} \rho \nonumber \\ 
-=\;\; & \lambda d\, \ol{\eta}\,\ol{B}.\,
-\setsem{\Gamma; \ol{\phi},\ol{\gamma}\vdash H}
-\id_{\rho[\ol{\gamma := B}]}[\ol{\phi := \lambda
-    \ol{A}.\eta_{\ol{A}\,\ol{B}}}] \label{eq:map-sem-def}
-\end{align}
-%\end{multline}
-\end{comment}
-In particular, if $\Gamma; \ol{\alpha} \vdash H$, $\ol{\Gamma;
-  \emptyset \vdash F}$, and $\ol{\Gamma; \emptyset \vdash G}$, then
-Definition~\ref{def:set-interp} gives, for all $\ol{f :
-  \setsem{\Gamma;\emptyset \vdash \Nat^\emptyset F\,G}\rho}$, that
+In fact, we have that if $\Gamma; \ol{\alpha} \vdash H$, $\ol{\Gamma;
+  \emptyset \vdash F}$, and $\ol{\Gamma; \emptyset \vdash G}$, then,
+for all $\ol{f : \setsem{\Gamma;\emptyset \vdash \Nat^\emptyset
+    F\,G}\rho}$,
 \[\begin{array}{rl}
@@
 \id_{\rho}[\ol{\alpha := f}] \\ =&
-%\lambda \ol{f : \setsem{\Gamma; \emptyset \vdash \sigma}\rho \to \setsem{\Gamma%; \emptyset \vdash \tau}\rho}.\,
 \mathit{map}_{\lambda \ol{A}.\,\setsem{\Gamma; \ol{\alpha}\vdash H} \rho [\ol{\alpha := A}]} \ol{f} \\
-%=& \mathit{map}_{\lambda \ol{A}.\,\setsem{\Gamma; \ol{\alpha}\vdash H} \rho [\ol{\alpha := A}]}
-\end{array}
-\]
-\begin{comment}
-\[\begin{array}{rl}
-&\setsem{
-\Gamma;\emptyset \,|\, \emptyset
-\vdash \map^{\ol{\sigma},\ol{\tau}}_H
-: \Nat^\emptyset\;(\ol{\Nat^{\emptyset}\,\sigma\,\tau})\; {}
-(\Nat^{\emptyset}\,H[\ol{\alpha := \sigma}]\,H[\ol{\alpha := \tau}])
-} \rho\, \ast \\
-=& \lambda \ol{f : \setsem{\Gamma; \emptyset \vdash \sigma}\rho \to \setsem{\Gamma; \emptyset \vdash \tau}\rho}.\,
-\setsem{\Gamma; \ol{\alpha}\vdash H}
-\id_{\rho}[\ol{\alpha := f}] \\
-=& \lambda \ol{f : \setsem{\Gamma; \emptyset \vdash \sigma}\rho \to \setsem{\Gamma; \emptyset \vdash \tau}\rho}.\,
-\mathit{map}_{\lambda \ol{A}.\,\setsem{\Gamma; \ol{\alpha}\vdash H} \rho [\ol{\alpha := A}]} \ol{f} \\
-=& \mathit{map}_{\lambda \ol{A}.\,\setsem{\Gamma; \ol{\alpha}\vdash H} \rho [\ol{\alpha := A}]}
-\end{array}
-\]
-\end{comment}
-
-\vspace*{0.05in}
-
-\noindent
+\end{array}\]
 Here, we obtain the first equality from the appropriate instance of
@@
 $\mathit{map}_G\,\ol f$ to denote the action of the semantic functor
-$G$ on morphisms $\ol f$.
+$G$ on morphisms $\ol f$. We also have, e.g., that:
 
@@
 
-We have the expected properties of the interpretations of $\map$,
-$\tin$, and $\fold$, namely:
-
-\vspace*{0.1in}
-
 \noindent
-$\bullet$\;  If
-$\Gamma; \ol{\psi}, \ol{\gamma} \vdash H$,\;
-$\ol{\Gamma; \ol{\alpha}, \ol{\gamma}, \ol{\phi} \vdash K}$,\;
-$\ol{\Gamma; \ol{\beta}, \ol{\gamma} \vdash F}$,\; and
-$\ol{\Gamma; \ol{\beta}, \ol{\gamma} \vdash G}$,
-then
+$\bullet$\; $\map$ is a higher-order functor, i.e., if $\Gamma;
+\ol{\psi}, \ol{\gamma} \vdash H$,\; $\ol{\Gamma; \ol{\alpha},
+  \ol{\gamma}, \ol{\phi} \vdash K}$,\; $\ol{\Gamma; \ol{\beta},
+  \ol{\gamma} \vdash F}$,\; $\ol{\Gamma; \ol{\beta}, \ol{\gamma}
+  \vdash G}$, and $\xi = \Nat^{\emptyset} (\ol{\Nat^{\ol{\alpha},
+    \ol{\beta}, \ol{\gamma}} F\, G})\, (\Nat^{\ol{\gamma}} H[\ol{\psi
+    := K}][\ol{\phi := F}]\, H[\ol{\psi := K}][\ol{\phi := G}])$, then
+
+\vspace*{-0.08in}
+
 \[\setsem{\Gamma; \emptyset \,|\, \emptyset \vdash
@@
   \emptyset \,|\, \emptyset \vdash L_\emptyset \ol
-  z. (\map_H^{\ol{K[\ol{\phi := F}]}, 
-    \ol{K[\ol{\phi := G}]}})_\emptyset \,(\ol{(\map_K^{\ol{F}, \ol{G}})_\emptyset \, \ol z}) :
-  \xi}\]
-for $\xi = \Nat^{\emptyset} (\ol{\Nat^{\ol{\alpha}, \ol{\beta},
-    \ol{\gamma}} F\, G})\, (\Nat^{\ol{\gamma}} H[\ol{\psi :=
-    K}][\ol{\phi := F}]\, H[\ol{\psi := K}][\ol{\phi := G}])$
-
-\vspace*{0.18in}
+  z. (\map_H^{\ol{K[\ol{\phi := F}]}, \ol{K[\ol{\phi :=
+          G}]}})_\emptyset \,(\ol{(\map_K^{\ol{F}, \ol{G}})_\emptyset
+    \, \ol z}) : \xi}\]
+
+\vspace*{0.1in}
 
 \noindent
-$\bullet$\; If $\Gamma; \overline{\beta}, \overline{\gamma} \vdash
-H$,\; $\Gamma; \overline{\beta}, \overline{\gamma} \vdash K$,\;
-$\overline{\Gamma; \overline{\alpha}, \overline{\gamma} \vdash F}$,\;
-$\overline{\Gamma; \emptyset, \overline{\alpha}, \overline{\gamma}
-  \vdash G}$,\; 
-$\overline{\Gamma; \phi, \overline{\psi}, \overline{\gamma} \vdash
-  M}$,\; $\overline{I}$ is the sequence $\overline{F}, H$,\;
-$\overline{J}$ is the sequence $\overline{G}, K$, and $\Gamma;
-\emptyset\,|\,\Delta \vdash s : \Nat^{\overline{\beta},
-  \overline{\gamma}} H\, K$ and $\ol{\Gamma; \emptyset \,|\,\Delta \vdash
-  t : \Nat^{\overline{\alpha}, \overline{\gamma}} F\, G}$, then
-\[\hspace*{0.13in}\setsem{\Gamma; \emptyset \,|\, \Delta
-    \vdash L_{\overline{\gamma}} z.\,
-    s_{\overline{M [\overline{\psi := G}] [\phi := K]},
-      \overline{\gamma}} \Big(\big( (\map_{H}^{\overline{M
-        [\overline{\psi := F}] [\phi := H]}, \overline{M
-        [\overline{\psi := G}] [\phi := K]}})_\emptyset
-    (\overline{(\map_{M}^{\overline{I}, \overline{J}})_\emptyset
-      (s, \overline{t})}) \big)_{\overline{\gamma}} z \Big) : \xi}\]
-\[= \setsem{\Gamma; \emptyset\,|\, \Delta \vdash
-    L_{\overline{\gamma}} z.\, \big( (\map_{K}^{\overline{M
-      [\overline{\psi := F}] [\phi := H]}, \overline{M [\overline{\psi
-          := G}] [\phi := K]}})_\emptyset (\overline{(\map_{M}^{\overline{I},
-    \overline{J}})_\emptyset (s, \overline{t})}) \big)_{\overline{\gamma}} \Big(
-s_{\overline{M [\overline{\psi := F}] [\phi := H]}, \overline{\gamma}}
-z \Big) : \xi}\] for $\xi = \Nat^{\overline{\gamma}}\,
-  H[\overline{\beta:= M}] [\overline{\psi := F}] [\phi := H]\;
-  K[\overline{\beta := M}] [\overline{\psi := G}] [\phi := K]$
-
-\vspace*{0.18in}
-
-\noindent
-$\bullet$\; If $\xi = \Nat^{\overline{\beta}, \overline{\gamma}}\,
-H[\phi := (\mu \phi. \lambda
+$\bullet$\; $\fold$ satisfies its universal property, i.e., if $\xi =
+\Nat^{\overline{\beta}{\color{red}, \overline{\gamma}}}\, H[\phi :=
+  (\mu \phi. \lambda
   \overline{\alpha}. H)\overline{\beta}][\overline{\alpha := \beta}]\;
 F$, then
-\[\hspace*{-0.72in}\setsem{\Gamma; \emptyset \,|\, x:
-  \Nat^{\overline{\beta}, \overline{\gamma}} H[\phi :=
+
+\vspace*{-0.05in}
+
+\[\hspace*{-1in}\setsem{\Gamma; \emptyset \,|\, x:
+  \Nat^{\overline{\beta}{\color{red}, \overline{\gamma}}} H[\phi :=
     F][\overline{\alpha := \beta}]\, F \vdash ((\fold_{H,
@@
 
-\[= \setsem{ \Gamma; \emptyset \,|\, x: \Nat^{\overline{\beta},
-    \overline{\gamma}} H[\phi := F][\overline{\alpha := \beta}]\, F
+\[= \setsem{ \Gamma; \emptyset \,|\, x: \Nat^{\overline{\beta}{\color{red},
+    \overline{\gamma}}} H[\phi := F][\overline{\alpha := \beta}]\, F
   \vdash x \circ \big( (\map_{H [\ol{\alpha := \beta}]}^{(\mu
@@
 \noindent
-$\bullet$\; If $\xi = \Nat^{\overline{\beta}, \overline{\gamma}}\, (\mu
-\phi. \lambda \overline{\alpha}. H)\overline{\beta}\; (\mu
-\phi. \lambda \overline{\alpha}. H)\overline{\beta}$, then
-\[\setsem{\Gamma; \emptyset \,|\, \emptyset
+$\bullet$\; $\tin$ has a right inverse, i.e., if $\xi =
+\Nat^{\overline{\beta}{\color{red}, \overline{\gamma}}}\, (\mu \phi. \lambda
+\overline{\alpha}. H)\overline{\beta}\; (\mu \phi. \lambda
+\overline{\alpha}. H)\overline{\beta}$, then
+
+\vspace*{-0.05in}
+
+\[\hspace*{0.43in}\setsem{\Gamma; \emptyset \,|\, \emptyset
   \vdash \tin_H \circ (\fold_{H, H[\phi := (\mu \phi. \lambda
@@
 
-\[\hspace*{-3.46in} = \setsem{\Gamma; \emptyset \,|\, \emptyset \vdash \mathit{id}_{(\mu
-    \phi. \lambda \overline{\alpha}. H)\overline{\beta}} : \xi}\]
+\[\hspace*{-3.46in} = \setsem{\Gamma; \emptyset \,|\, \emptyset \vdash \mathit{id}_{(\mu  \phi. \lambda \overline{\alpha}. H)\overline{\beta}} : \xi}\]
 
@@
 \noindent
-$\bullet$\; If $\xi = \Nat^{\overline{\beta}, \overline{\gamma}}
-H[\phi := (\mu \phi. \lambda \overline{\alpha}. H)\overline{\beta}]\,
-H[\phi := (\mu \phi. \lambda \overline{\alpha}. H)\overline{\beta}]$,
-then
-\[\setsem{\Gamma; \emptyset \,|\, \emptyset \vdash
+$\bullet$\; $\tin$ has a left inverse, i.e., if $\xi =
+\Nat^{\overline{\beta}{\color{red}, \overline{\gamma}}} H[\phi := (\mu
+  \phi. \lambda \overline{\alpha}. H)\overline{\beta}]\, H[\phi :=
+  (\mu \phi. \lambda \overline{\alpha}. H)\overline{\beta}]$, then
+
+\vspace*{-0.05in}
+
+\[\hspace*{0.41in}\setsem{\Gamma; \emptyset \,|\, \emptyset \vdash
 (\fold_{H, H[\phi := (\mu \phi. \lambda
@@
 Analogous results hold for relational interpretations of terms and
-relational environments.
-
-\begin{comment}
-\subsection{Free Theorems Derived from Naturality}\label{sec:ft-nat}
-
-In our setting, free theorems of a certain familiar form are derivable
-directly from the semantic properties of $\Nat$-types. Indeed, the
-following theorem is immediate from Equation~\ref{thm:subst}, by taking
-$\gamma$ to be empty, expanding the interpretations on the left- and
-right-hand sides, and using the fact from Section~\ref{sec:iaps} that
-$\map$ terms are interpreted as semantic $\mathit{map}$ functions:
-\begin{thm}\label{thm:ft-nat}
-If\, $\Gamma; \ol\alpha \vdash F : \F$,\, $\Gamma; \ol\alpha \vdash G :
-\F$, $\rho : \setenv$, $x \in \setsem{\Gamma; \emptyset \vdash
-  \Nat^{\ol\alpha} F\, G } \rho$, and, for vectors of length
-$|\ol\alpha|$, \, $\ol{\Gamma; \emptyset \vdash \sigma : \F}$,\,
-$\ol{\Gamma; \emptyset \vdash \tau: \F}$, and $\ol{y \in
-  \setsem{\Gamma; \emptyset \vdash \Nat^{\emptyset} \sigma\, \tau }
-  \rho}$, then
-\[\mathit{map}_{\setsem{\Gamma; \ol\alpha \vdash G} \rho[\ol{\alpha := \_}]} \ol{y}
-  \circ x_{\ol{{\setsem{\Gamma; \emptyset \vdash \sigma} \rho}}} =
-  x_{\ol{\setsem{\Gamma; \emptyset \vdash \tau} \rho} } \circ
-  \mathit{map}_{\setsem{\Gamma; \ol\alpha \vdash F} \rho[\ol{\alpha :=
-        \_}]} \, \ol{y}\]
-\end{thm}
-\noindent
-%INCORRECT
-%In particular, for $\ol{A, B: \set}$, variables $\ol{v}$ and $\ol{w}$
-%with $\ol{\rho v = A}$ and $\ol{\rho w = B}$, and functions $\ol{f : A
-%  \to B}$, the following holds: $\mathit{map}_{\setsem{\Gamma; \ol\alpha \vdash G}
-%  \rho[\ol{\alpha := \_}]} \ol{f} \circ x_{\ol{A}} = x_{\ol{B}} \circ
-%\mathit{map}_{\setsem{\Gamma; \ol\alpha \vdash F} \rho[\ol{\alpha :=
-%      \_}]} \, \ol{f}$.
-A similar result holds for relational interpretations.
-
-Specific instances of Theorem~\ref{thm:ft-nat} include {\color{blue}
-  Wadler free theorems, map-reverse theorems from intro, ???}
-\end{comment}
-
-\begin{comment}
-We already know from Equation~\ref{thm:subst} that 
-\begin{multline}
-\setsem{\Gamma; \emptyset \,|\, x :
-  \Nat^{\overline{\alpha}, 
-    \overline{\gamma}} F\,G, \overline{y : \Nat^{\overline{\gamma}}
-    \sigma\, \tau} \vdash ((\map_G^{\overline{\sigma},
-    \overline{\tau}})_{\emptyset} \overline{y}) \circ
-  (L_{\overline{\gamma}} z. x_{\overline{\sigma}, \overline{\gamma}}
-  z) : \Nat^{\overline{\gamma}} F[\overline{\alpha := \sigma}]\,
-  G[\overline{\alpha := \tau}]} \\
-= \setsem{ \Gamma; \emptyset \,|\, x
-  : \Nat^{\overline{\alpha}, \overline{\gamma}} F\, G, \overline{y :
-    \Nat^{\overline{\gamma}} \sigma\, \tau} \vdash
-  (L_{\overline{\gamma}} z. x_{\overline{\tau}, \overline{\gamma}} z)
-  \circ ((\map_F^{\overline{\sigma}, \overline{\tau}})_{\emptyset}
-  \overline{y}) : \Nat^{\overline{\gamma}} F[\overline{\alpha :=
-      \sigma}]\, G[\overline{\alpha := \tau}]}
-\end{multline}
-
-In particular, for any set environment $\rho$, natural transformation 
-$x \in \setsem{\Gamma; \emptyset \vdash \Nat^{\ol\alpha} F\, G } \rho$, and functions
-$\ol{y \in \setsem{\Gamma; \emptyset \vdash \Nat^{\emptyset} \sigma\, \tau } \rho}$
-(thus specializing to the case where there are no $\gamma$’s), we have
-\begin{multline}
-\setsem{\Gamma; \emptyset \,|\, x :
-  \Nat^{\overline{\alpha}} F\,G, \overline{y : \Nat^{\emptyset}
-    \sigma\, \tau} \vdash ((\map_G^{\overline{\sigma},
-    \overline{\tau}})_{\emptyset} \overline{y}) \circ
-  (L_{\emptyset} z. x_{\overline{\sigma}}
-  z) : \Nat^{\emptyset} F[\overline{\alpha := \sigma}]\,
-  G[\overline{\alpha := \tau}]} \rho (x, \ol{y}) \\
-= \setsem{ \Gamma; \emptyset \,|\, x
-  : \Nat^{\overline{\alpha}} F\, G, \overline{y :
-    \Nat^{\emptyset} \sigma\, \tau} \vdash
-  (L_{\emptyset} z. x_{\overline{\tau}} z)
-  \circ ((\map_F^{\overline{\sigma}, \overline{\tau}})_{\emptyset}
-  \overline{y}) : \Nat^{\emptyset} F[\overline{\alpha :=
-      \sigma}]\, G[\overline{\alpha := \tau}]} \rho (x, \ol{y})
-\end{multline}
-
-Unwinding these interpretations yields
-\begin{multline}
-\setsem{\Gamma; \emptyset \,|\, x :
-  \Nat^{\overline{\alpha}} F\,G, \overline{y : \Nat^{\emptyset}
-    \sigma\, \tau} \vdash ((\map_G^{\overline{\sigma},
-    \overline{\tau}})_{\emptyset} \overline{y}) \circ
-  (L_\emptyset z. x_{\overline{\sigma}}
-  z) : \Nat^{\emptyset} F[\overline{\alpha := \sigma}]\,
-  G[\overline{\alpha := \tau}]} \rho \, (x , \ol{y}) \\
-= 
-\setsem{\Gamma; \emptyset \,|\, x : \Nat^{\overline{\alpha}} F\,G, 
-  \overline{y : \Nat^{\emptyset}
-    \sigma\, \tau} \vdash (\map_G^{\overline{\sigma},
-    \overline{\tau}})_{\emptyset} \overline{y} } \rho \, (x, \ol{y})
-   \circ
-  \setsem{\Gamma; \emptyset \,|\, x : \Nat^{\overline{\alpha}} F\,G 
-   , \overline{y : \Nat^{\emptyset}
-      \sigma\, \tau} \vdash
-   L_{\emptyset} z. x_{\overline{\sigma}}
-  z } \rho \, (x, \ol{y}) \\
-= 
-\setsem{\Gamma; \emptyset \,|\, \overline{y : \Nat^{\emptyset}
-    \sigma\, \tau} \vdash (\map_G^{\overline{\sigma},
-    \overline{\tau}})_{\emptyset} \overline{y} } \rho \, \ol{y}
-   \circ
-  \setsem{\Gamma; \emptyset \,|\, x : \Nat^{\overline{\alpha}} F\,G \vdash
-    L_{\emptyset} z. x_{\overline{\sigma}}
-  z } \rho \, x \\
-= \map_{\setsem{\Gamma; \ol\alpha \vdash G} \rho[\ol{\alpha := \_}]} \ol{y}
-  \circ 
-  x_{\ol{{\setsem{\Gamma; \emptyset \vdash \sigma} \rho}}}
-\end{multline}
-
-and 
-\begin{multline}
-\setsem{ \Gamma; \emptyset \,|\, x
-  : \Nat^{\overline{\alpha}} F\, G, \overline{y :
-    \Nat^{\emptyset} \sigma\, \tau} \vdash
-  (L_{\emptyset} z. x_{\overline{\tau}} z)
-  \circ ((\map_F^{\overline{\sigma}, \overline{\tau}})_{\emptyset}
-  \overline{y}) : \Nat^{\emptyset} F[\overline{\alpha :=
-      \sigma}]\, G[\overline{\alpha := \tau}]} \rho (x, \ol{y}) \\
-=
-\setsem{ \Gamma; \emptyset \,|\, x
-  : \Nat^{\overline{\alpha}} F\, G, \overline{y :
-    \Nat^{\emptyset} \sigma\, \tau} \vdash
-  L_{\emptyset} z. x_{\overline{\tau}} z)} \rho (x, \ol{y})
-  \circ 
-  \setsem{\Gamma; \emptyset
-  \,|\, x : \Nat^{\overline{\alpha}} F\, G, \overline{y :
-    \Nat^{\emptyset} \sigma\, \tau} \vdash 
-  (\map_F^{\overline{\sigma}, \overline{\tau}})_{\emptyset}
-  \overline{y} } \rho (x, \ol{y}) \\
-= \setsem{\Gamma; \emptyset \,|\, x
-  : \Nat^{\overline{\alpha}} F\, G \vdash 
-  L_{\emptyset} z. x_{\overline{\tau}} z} \rho \, x
-  \circ 
-  \setsem{\Gamma;\emptyset \,|\, \overline{y :
-    \Nat^{\emptyset} \sigma\, \tau} \vdash
-    (\map_F^{\overline{\sigma}, \overline{\tau}})_{\emptyset}
-  \overline{y} } \rho \, \ol{y}  \\
-  = x_{\ol{\setsem{\Gamma; \emptyset \vdash \tau} \rho} }
-  \circ \map_{\setsem{\Gamma; \ol\alpha \vdash F} 
-    \rho[\ol{\alpha := \_}]} \, \ol{y}
-\end{multline}
-
-So, we can conclude that
-\begin{equation}
-\map_{\setsem{\Gamma; \ol\alpha \vdash G} \rho[\ol{\alpha := \_}]} \ol{y}
-  \circ 
-  x_{\ol{{\setsem{\Gamma; \emptyset \vdash \sigma} \rho}}}
-= x_{\ol{\setsem{\Gamma; \emptyset \vdash \tau} \rho} }
-\circ \map_{\setsem{\Gamma; \ol\alpha \vdash F} 
-  \rho[\ol{\alpha := \_}]} \, \ol{y}
-\end{equation}
-
-Moreover, for any $\ol{A, B: \set}$,
-we can choose $\ol{\sigma = v}$ and $\ol{\tau = w}$ to be variables such that
-$\ol{\rho v = A}$ and $\ol{\rho w = B}$.
-Then for any functions $\ol{f : A \to B}$ we have that
-\begin{equation}
-\map_{\setsem{\Gamma; \ol\alpha \vdash G} \rho[\ol{\alpha := \_}]} \ol{f}
-  \circ 
-  x_{\ol{A}}
-  = x_{\ol{B}}
-\circ \map_{\setsem{\Gamma; \ol\alpha \vdash F} 
-  \rho[\ol{\alpha := \_}]} \, \ol{f}
-\end{equation}
-\end{proof}
-\end{comment}
+relation environments.
 
@@
 
-To go beyond naturality and get {\em all} consequences of
-parametricity, we prove an Abstraction Theorem for our calculus. As
-usual for such theorems, we prove a more general result in
-Theorem~\ref{thm:at-gen} that handles possibly open terms. We then
-recover the Abstraction Theorem (Theorem~\ref{thm:abstraction}) as the
-special case Theorem~\ref{thm:at-gen} for closed terms of closed type.
+To get consequences of parametricity that are not merely consequences
+of naturality, we prove an Abstraction Theorem
+(Theorem~\ref{thm:abstraction}) for our calculus. As is usual for such
+theorems, we first prove a more general result
+(Theorem~\ref{thm:at-gen}) for open terms, and recover our Abstraction
+Theorem as a special case of it for closed terms of closed type.
 
@@
 \relsem{\Gamma;\Phi~|~\Delta \vdash t : F})\]
-where
+where, for $\mathsf{D} \in \{\set,\rel\}$ and for $\rho \in \setenv$
+or $\rho \in \relenv$ as appropriate,
 \[\begin{array}{lll}
-\setsem{\Gamma;\Phi~|~\Delta \vdash t : F} & : & \setsem{\Gamma;
-  \Phi \vdash \Delta} \to \setsem{\Gamma; \Phi \vdash F}
+\dsem{\Gamma;\Phi~|~\Delta \vdash t : F} & : & \dsem{\Gamma;
+  \Phi \vdash \Delta} \to \dsem{\Gamma; \Phi \vdash F}
 \end{array}\]
@@
 \[\begin{array}{lll}
-\setsem{\Gamma;\Phi~|~\Delta \vdash t : F}\rho & : & \setsem{\Gamma;
-  \Phi \vdash \Delta}\rho \to \setsem{\Gamma; \Phi \vdash F}\rho
-\end{array}\]
-in $\set$,
-\[\begin{array}{lll}
-\relsem{\Gamma;\Phi~|~\Delta \vdash t : F} & : & \relsem{\Gamma;
-  \Phi \vdash \Delta} \to \relsem{\Gamma; \Phi \vdash F}
-\end{array}\]
-has as its component at $\rho : \relenv$ a morphism
-\[\begin{array}{lll}
-\relsem{\Gamma;\Phi~|~\Delta \vdash t : F}\rho & : & \relsem{\Gamma;
-  \Phi \vdash \Delta}\rho \to \relsem{\Gamma; \Phi \vdash F}\rho
-\end{array}\]
-in $\rel$,
-and, for all $\rho : \relenv$,
+\dsem{\Gamma;\Phi~|~\Delta \vdash t : F}\rho & : & \dsem{\Gamma;
+  \Phi \vdash \Delta}\rho \to \dsem{\Gamma; \Phi \vdash F}\rho
+\end{array}\]
+Moreover, for all $\rho : \relenv$, 
 \begin{equation}\label{eq:projs}
@@
 \end{thm}
-\proof
-{\color{red} ADD IN PROOF HERE}
-By structural induction on the type judgment for $t$. The overall
-challenge of our development manifests in this proof. It lies in
-showing that the set and relational interpretations of each term
-judgment are natural transformations, and that all set interpretations
-of terms of $\Nat$-types satisfy the appropriate equality preservation
-conditions from Definition~\ref{def:set-sem}.  For the interesting
-cases of abstraction, application, $\map$, $\tin$, and $\fold$ terms,
+\proof The proof is by induction on $t$. It requires showing that set
+and relational interpretations of term judgments are natural
+transformations, and that all set interpretations of terms of $\Nat$
+types satisfy the appropriate equality preservation conditions from
+Definition~\ref{def:set-sem}.  For the interesting cases of
+abstraction, application, $\map$, $\tin$, and $\mathsf{fold}$ terms,
 propagating the naturality conditions is quite involved; the latter
-two cases especially require some rather delicate diagram
-chasing. That this is possible provides strong evidence that our
-definitions and development are sensible, natural, and at an
-appropriate level of abstraction. Detailed proofs for the set
-interpretations for the five interesting forms of terms are included
-in the accompanying anonymous supplementary material. Even in the
-supplementary material we omit the other, straightforward, cases.  The
-proofs for the relational interpretations of all term judgments are
-entirely analogous to the proofs for their set interpretations. The
-proof that Equation~\ref{eq:projs} holds in each case is by direct
-calculation, using the facts that projections are surjective and that
-the set and relational interpretations are defined ``in parallel'',
-i.e., are fibred.
+two especially require some rather delicate diagram chasing. That it
+is possible provides strong evidence that our development is sensible,
+natural, and at an appropriate level of abstraction.
+
+The only interesting cases are the cases for abstraction, application,
+$\map$, $\tin$, and $\fold$. We omit the others.
+
+\begin{itemize}
+\item $\underline{\Gamma; \emptyset \,|\, \Delta \vdash
+  L_{\overline{\alpha}} x.t : \Nat^{\overline{\alpha}} \,F \,G}$ \; To
+  see that $\setsem{\Gamma; \emptyset \,|\, \Delta \vdash
+    L_{\overline{\alpha}} x.t : \Nat^{\overline{\alpha}} \,F \,G}$ is
+  a natural transformation from $\setsem{\Gamma; \emptyset \vdash
+    \Delta}$ to $\setsem{\Gamma; \emptyset \vdash
+    \Nat^{\overline{\alpha}} \,F \,G}$ we need to show that, for every
+  $\rho : \setenv$, $\setsem{\Gamma; \emptyset \,|\, \Delta \vdash
+    L_{\overline{\alpha}} x.t : \Nat^{\overline{\alpha}} \,F \,G}\rho$
+  is a morphism in $\set$ from $\setsem{\Gamma; \emptyset \vdash
+    \Delta}\rho$ to $\setsem{\Gamma; \emptyset \vdash
+    \Nat^{\overline{\alpha}} \,F \,G}\rho$, and that such a family of
+  morphisms is natural.  First, we need to show that, for all $\ol{A :
+    \set}$ and all $d : \setsem{\Gamma; \emptyset \vdash \Delta}\rho =
+  \setsem{\Gamma; \ol{\alpha} \vdash \Delta}\rho[\overline{\alpha :=
+      A}]$, we have that
+  \begin{align*}
+      &(\setsem{\Gamma; \emptyset \,|\, \Delta \vdash L_{\overline{\alpha}}
+      x.t : \Nat^{\overline{\alpha}} \,F \,G}\rho\,d)_{\ol{A}} :  \\
+      &\hspace{1in} \setsem{\Gamma; \ol{\alpha} \vdash F}\rho[\overline{\alpha := A}]
+    \to \setsem{\Gamma; \ol{\alpha} \vdash G}\rho[\overline{\alpha :=
+        A}]
+  \end{align*}
+      but this follows easily from the induction hypothesis.
+That these maps comprise a natural transformation $\eta :
+\setsem{\Gamma; \ol{\alpha} \vdash F}\rho[\overline{\alpha := \_}] \to
+\setsem{\Gamma; \ol{\alpha} \vdash G}\rho[\overline{\alpha := \_}]$ is
+clear since $\eta_{\ol{A}} \, = \, \curry\,
+(\setsem{\Gamma; \overline{\alpha} \,|\, \Delta, x : F \vdash t:
+  G}\rho[\overline{\alpha := A}])\,d$ is the component at $\ol{A}$ of
+the partial specialization to $d$ of the natural transformation
+$$\setsem{\Gamma; \overline{\alpha} \,|\, \Delta, x : F \vdash t:
+  G}\rho[\overline{\alpha := \_}]$$  To see that the components of
+$\eta$ also satisfy the additional condition needed for $\eta$ to
+be in $\setsem{\Gamma; \emptyset \vdash \Nat^{\overline{\alpha}} \,F
+  \,G}\rho$, let $\overline{R : \rel(A, B)}$ and suppose
+\[\begin{array}{lll}
+(u, v) &  \in & \relsem{\Gamma;\overline{\alpha} \vdash F}
+\Eq_{\rho}[\overline{\alpha := R}]\\
+&  = & (\setsem{\Gamma;\overline{\alpha} \vdash F}
+\rho[\overline{\alpha := A}], \\ 
+  & &\,\,\setsem{\Gamma;\overline{\alpha} \vdash
+  F} \rho[\overline{\alpha := B}],
+(\relsem{\Gamma;\overline{\alpha} \vdash F}
+\Eq_{\rho}[\overline{\alpha := R}])^*)
+\end{array}\]
+Then the induction hypothesis and 
+$(d,d) \in \relsem{\Gamma; \emptyset \vdash \Delta} \Eq_\rho =
+\relsem{\Gamma; \emptyset \vdash \Delta} \Eq_\rho[\ol{\alpha := R}]$
+ensure that
+\[\begin{array}{ll}
+& (\eta_{\ol{A}}u,\eta_{\ol{B}}v)\\
+= & (\curry\, (\setsem{\Gamma; \overline{\alpha} \,|\, \Delta, x : F
+  \vdash t: G}\rho[\overline{\alpha := A}])\,d\,u, \\
+  &\hspace{0.5in} \curry\,
+(\setsem{\Gamma; \overline{\alpha} \,|\, \Delta, x : F \vdash t:
+  G}\rho[\overline{\alpha := B}])\,d\,v)\\
+= & \curry\, (\relsem{\Gamma; \overline{\alpha} \,|\, \Delta, x : F
+  \vdash t: G}\Eq_\rho[\overline{\alpha := R}])\,(d,d)\,(u,v)\\
+: & \relsem{\Gamma; \overline{\alpha} \vdash G}
+\Eq_{\rho}[\overline{\alpha := R}]  
+\end{array}\]
+Moreover,
+$\setsem{\Gamma; \emptyset \,|\, \Delta \vdash L_{\overline{\alpha}} x.t
+: \Nat^{\overline{\alpha}} \,F \,G} \rho$
+is trivially natural in $\rho$,
+as the functorial action of
+$\setsem{\Gamma; \emptyset \vdash \Delta}$
+and $\setsem{\Gamma; \emptyset \vdash \Nat^{\overline{\alpha}} \,F \,G}$
+on morphisms is the identity.
+
+\item 
+$\underline{\Gamma;\Phi \,|\, \Delta \vdash t_{\overline K} s: G
+  [\overline{\alpha := K}]}$\; 
+  To see that $\setsem{\Gamma;\Phi \,|\,
+  \Delta \vdash t_{\overline K} s: G [\overline{\alpha := K}]}$ is a
+  natural transformation from $\setsem{\Gamma; \Phi \vdash \Delta}$ to
+  $\setsem{\Gamma;\Phi \vdash G [\overline{\alpha := K}]}$ we 
+  show that, for every $\rho : \setenv$, $\setsem{\Gamma;\Phi \,|\,
+    \Delta \vdash t_{\overline K} s: G [\overline{\alpha := K}]}\rho$
+  is a morphism from $\setsem{\Gamma; \Phi \vdash \Delta}\rho$ to
+  $\setsem{\Gamma;\Phi \vdash G [\overline{\alpha := K}]}\rho$, and
+  that this family of morphisms is natural in $\rho$. Let $d :
+  \setsem{\Gamma; \Phi \vdash \Delta}\rho$. Then
+  \[\begin{array}{ll}
+  & \setsem{\Gamma;\Phi \,|\, \Delta \vdash t_{\overline K} s: G
+  [\overline{\alpha := K}]}\,\rho\,d\\
+= & (\eval \circ \langle (\setsem{\Gamma; \emptyset \,|\, \Delta \vdash
+  t : \Nat^{\overline{\alpha}} \,F \,G}\rho\;
+\_)_{\overline{\setsem{\Gamma;\Phi \vdash K}\rho}},\,
+\setsem{\Gamma;\Phi \,|\, \Delta \vdash s: F [\overline{\alpha :=
+      K}]}\rho \rangle)\,d\\
+= & \eval ((\setsem{\Gamma; \emptyset \,|\, \Delta \vdash t :
+  \Nat^{\overline{\alpha}} \,F \,G}\rho\;
+\_)_{\overline{\setsem{\Gamma;\Phi \vdash K}\rho}} \,d,\,
+\setsem{\Gamma;\Phi \,|\, \Delta \vdash s: F [\overline{\alpha :=
+      K}]}\rho\, d)\\
+= & \eval ((\setsem{\Gamma; \emptyset \,|\, \Delta \vdash t :
+  \Nat^{\overline{\alpha}} \,F \,G}\rho\;
+d)_{\overline{\setsem{\Gamma;\Phi \vdash K}\rho}},\,
+\setsem{\Gamma;\Phi \,|\, \Delta \vdash s: F [\overline{\alpha :=
+      K}]}\rho\, d)\\
+\end{array}\]
+The induction hypothesis ensures that $(\setsem{\Gamma; \emptyset \,|\,
+  \Delta \vdash t : \Nat^{\overline{\alpha}} \,F \,G}\rho\,
+d)_{\overline{\setsem{\Gamma;\Phi \vdash K}\rho}}$ has type
+$\setsem{\Gamma; \ol{\alpha} \vdash F}\rho[\ol{\alpha :=
+    \setsem{\Gamma;\Phi \vdash K}\rho}] \to \setsem{\Gamma;
+  \ol{\alpha} \vdash G}\rho[\ol{\alpha := \setsem{\Gamma;\Phi \vdash
+      K}\rho}]$.  Since, in addition, 
+\[\begin{array}{ll}
+  &\setsem{\Gamma;\Phi \,|\,
+  \Delta \vdash s: F [\overline{\alpha := K}]}\rho\, d :
+\setsem{\Gamma; \Phi \vdash F[\ol{\alpha := K}]}\rho \\ 
+  =&\setsem{\Gamma; \Phi, \ol{\alpha} \vdash F}\rho[\ol{\alpha :=
+    \setsem{\Gamma;\Phi \vdash K}\rho}] \\
+    =& \setsem{\Gamma;
+  \ol{\alpha} \vdash F}\rho[\ol{\alpha := \setsem{\Gamma;\Phi \vdash
+      K}\rho}] 
+\end{array}\]
+      we have that 
+\[\begin{array}{ll}
+  &\setsem{\Gamma;\Phi \,|\, \Delta
+  \vdash t_{\overline K} s: G [\overline{\alpha :=
+      K}]}\,\rho\,d : \setsem{\Gamma; \Phi ,\ol{\alpha} \vdash
+  G}\rho[\ol{\alpha := \setsem{\Gamma;\Phi \vdash K}\rho}]  \\
+  =&\setsem{\Gamma; \Phi \vdash G[\ol{\alpha := K}]}\rho
+\end{array}\]
+as desired.
+
+\vspace*{0.1in}
+
+To see that the family of maps comprising $\setsem{\Gamma;\Phi \,|\,
+  \Delta \vdash t_{\overline K} s: G [\overline{\alpha := K}]}$
+is natural in $\rho$
+we need to show that, if $f : \rho \to \rho'$ in $\setenv$, then the
+following diagram commutes, where $g = \setsem{\Gamma; \emptyset \,|\,
+  \Delta \vdash t : \Nat^{\overline{\alpha}} \,F \,G}$ and $h =
+\setsem{\Gamma;\Phi \,|\, \Delta \vdash s: F [\overline{\alpha :=
+      K}]}$:
+{\footnotesize
+\[\hspace*{-0.6in}
+\begin{tikzcd}
+\setsem{\Gamma;\Phi \vdash \Delta}\rho \ar[r, "{\setsem{\Gamma;\Phi
+  \vdash \Delta}f}"] \ar[d, "{\langle g \rho, h \rho\rangle}"']
+& \setsem{\Gamma;\Phi \vdash 
+  \Delta}\rho' \ar[d, "{\langle g \rho', h \rho' \rangle}"]\\
+\setsem{\Gamma;\emptyset \vdash \Nat^{\overline{\alpha}} \,F \,G}\rho
+\times \setsem{\Gamma;\Phi \vdash F [\overline{\alpha := K}]}\rho
+\ar[d, "{\eval \circ ((-)_{\overline{\sem{\Gamma;\Phi \vdash K}\rho}} \times
+    \id)}"']
+\ar[r, bend left = 5, "{\setsem{\Gamma;\emptyset\vdash
+      \Nat^{\overline{\alpha}} \,F \,G}f\, \times\, \setsem{\Gamma;\Phi
+      \vdash F [\overline{\alpha := K}]}f}"] &
+\setsem{\Gamma;\emptyset \vdash \Nat^{\overline{\alpha}} \,F \,G}\rho'
+\times \setsem{\Gamma;\Phi \vdash F [\overline{\alpha := K}]}\rho'
+\ar[d, "{\eval \circ ((-)_{\overline{\sem{\Gamma;\Phi \vdash
+          K}\rho'}} \times \id)}"] \\
+\setsem{\Gamma;\Phi \vdash G [\overline{\alpha := K}]}\rho
+\ar[r, "{\setsem{\Gamma;\Phi \vdash G [\overline{\alpha := K}]}f}"']
+&
+\setsem{\Gamma;\Phi \vdash G [\overline{\alpha := K}]}\rho'
+\end{tikzcd}\]}
+
+\noindent
+The top diagram commutes because $g$ and $h$ are natural in $\rho$ by
+the induction hypothesis.
+To see that the bottom diagram commutes,
+we need to show that
+\[\begin{array}{ll}
+& \setsem{\Gamma;\Phi \vdash G [\overline{\alpha := K}]}f
+(\eta_{\overline{\sem{\Gamma;\Phi \vdash K}\rho}} x)\\
+= &
+(\setsem{\Gamma; \emptyset \vdash \Nat^{\overline{\alpha}} \,F \,G} f\, \eta
+)_{\overline{\sem{\Gamma;\Phi \vdash K}\rho'}}
+(\setsem{\Gamma;\Phi \vdash F [\overline{\alpha := K}]}f x)
+\end{array}\]
+holds for all $\eta \in \setsem{\Gamma; \emptyset \vdash
+  \Nat^{\overline{\alpha}} \,F \,G}\rho$ and $x \in
+\setsem{\Gamma;\Phi \vdash F [\overline{\alpha := K}]}\rho$,
+i.e.,
+%{\color{red} by remembering the following facts,
+%\begin{align*}
+%\setsem{\Gamma;\Phi \vdash F[\ol{\alpha := K}]}\rho
+%&= \setsem{\Gamma; \ol{\alpha} \vdash F}\rho[\ol{\alpha :=
+%    \setsem{\Gamma;\Phi \vdash K}\rho}] \\
+%\setsem{\Gamma;\Phi \vdash F[\ol{\alpha := K}]}f
+%&= \setsem{\Gamma; \ol{\alpha} \vdash F}
+%  f [\ol{\alpha := \setsem{\Gamma;\Phi \vdash K}f}] \\
+%\setsem{\Gamma;\Phi \vdash G[\ol{\alpha := K}]}\rho
+%&= \setsem{\Gamma; \ol{\alpha} \vdash G}\rho[\ol{\alpha :=
+%    \setsem{\Gamma;\Phi \vdash K}\rho}] \\
+%\setsem{\Gamma;\Phi \vdash G[\ol{\alpha := K}]}f
+%&= \setsem{\Gamma; \ol{\alpha} \vdash
+%  G} f [\ol{\alpha := \setsem{\Gamma;\Phi \vdash K}f}]
+%\end{align*}}
+%we need to show
+that
+\[\begin{array}{ll}
+ & \setsem{\Gamma; \ol{\alpha} \vdash G} f[\overline{\alpha :=
+    \setsem{\Gamma;\Phi \vdash K} f }] \circ
+\eta_{\overline{\setsem{\Gamma;\Phi \vdash K}\rho}} \\ 
+=\;&
+\eta_{\overline{\setsem{\Gamma;\Phi \vdash K}\rho'}}
+\circ
+\setsem{\Gamma; \ol{\alpha} \vdash F} f [\ol{\alpha := \setsem{\Gamma;\Phi \vdash K}f}]
+\end{array}\]
+for all $\eta \in \setsem{\Gamma;\emptyset \vdash
+  \Nat^{\overline{\alpha}} \,F \,G}\rho$.
+But this follows from the naturality of $\eta$, which indeed ensures
+the commutativity of 
+{\footnotesize
+\[\hspace*{-0.3in}\begin{tikzcd}[column sep = large]
+\setsem{\Gamma; \ol{\alpha} \vdash F}\rho[\ol{\alpha :=
+    \setsem{\Gamma;\Phi \vdash K}\rho}] \ar[r,
+  "{\;\;\;\eta_{\ol{\setsem{\Gamma;\Phi \vdash K}\rho}}\;\;\; }"]
+\ar[d, "{\setsem{\Gamma; \ol{\alpha} \vdash F} f [\ol{\alpha := 
+        \setsem{\Gamma;\Phi \vdash K}f}]}"']
+& \setsem{\Gamma;
+  \ol{\alpha} \vdash G}\rho[\ol{\alpha := \setsem{\Gamma;\Phi \vdash
+      K}\rho}]
+\ar[d, "{\setsem{\Gamma; \ol{\alpha} \vdash G} f [\ol{\alpha := 
+        \setsem{\Gamma;\Phi \vdash K}f}]}"]\\
+\setsem{\Gamma; \ol{\alpha} \vdash F}\rho'[\ol{\alpha :=
+    \setsem{\Gamma;\Phi \vdash K}\rho'}] \ar[r,
+  "{\eta_{\ol{\setsem{\Gamma;\Phi \vdash K}\rho'}} }"]
+& \setsem{\Gamma; \ol{\alpha} \vdash G}\rho'[\ol{\alpha :=
+    \setsem{\Gamma;\Phi \vdash K}\rho'}]
+\end{tikzcd}\]}
+
+\item 
+  $\underline{\Gamma;\emptyset~|~\emptyset \vdash
+  \map^{\ol{F},\ol{G}}_H :
+  \Nat^\emptyset\;(\ol{\Nat^{\ol{\beta},\ol{\gamma}}\,F\,G})\;
+  (\Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\,H[\ol{\phi
+      :=_{\ol{\beta}} G}])}$\;
+To see that
+\[
+\setsem{\Gamma; \emptyset~|~\emptyset \vdash \map^{\ol{F},\ol{G}}_H
+    : \Nat^\emptyset\;(\ol{\Nat^{\ol{\beta},\ol{\gamma}}\,F\,G})\;
+    (\Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\,H[\ol{\phi
+        :=_{\ol{\beta}} G}])}\,\rho\,d\, \ol{\eta}
+\]
+is in $\setsem{\Gamma; \emptyset \vdash
+    \Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\,H[\ol{\phi
+        :=_{\ol{\beta}} G}]} \rho$
+for all $\rho : \setenv$, $d : \setsem{\Gamma;\emptyset \vdash \emptyset} \rho$,
+and $\ol{\eta : \setsem{\Gamma; \emptyset
+  \vdash\Nat^{\ol{\beta},\ol{\gamma}}\,F\,G} \rho}$,
+  we first note that
+$\setsem{\Gamma ;\ol{\phi}, \ol{\gamma} \vdash H}$ is a functor from
+  $\setenv$ to $\set$ and, for any $\ol C$, $\id_{\rho[\ol{\gamma :=
+        C}]}[\ol{\phi := \lambda \ol{B}. \eta_{\ol{B}\,\ol{C}}}]$ is a
+  morphism in $\setenv$ from \[\rho[\ol{\gamma := C}][\ol{\phi :=
+      \lambda \ol{B}.\setsem{\Gamma; \ol{\gamma},\ol{\beta} \vdash
+        F}\rho[\ol{\gamma := C}][\ol{\beta := B}]}]\] to
+\[\rho[\ol{\gamma := C}][\ol{\phi := \lambda \ol{B}.\setsem{\Gamma;
+\ol{\gamma},\ol{\beta} \vdash G}\rho[\ol{\gamma := C}][\ol{\beta := B}]}]\]
+so that
+\begin{align*}
+&(\setsem{\Gamma; \emptyset~|~\emptyset \vdash
+  \map^{\ol{F},\ol{G}}_H :
+\Nat^\emptyset\;(\ol{\Nat^{\ol{\beta},\ol{\gamma}}\,F\,G})\;
+(\Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\,H[\ol{\phi
+    :=_{\ol{\beta}} G}])}\,\rho\,d\, \ol{\eta})_{\ol{C}} \\
+  &=  \setsem{\Gamma; \ol{\phi},\ol{\gamma} \vdash H}\id_{\rho[\ol{\gamma
+      := C}]}[\ol{\phi := \lambda \ol{B}. \eta_{\ol{B}\,\ol{C}}}]
+\end{align*}
+is indeed a morphism of type
+$$\setsem{\Gamma ;\ol{\gamma} \vdash H[\ol{\phi := F}]}\rho[\ol{\gamma
+      := C}]% $
+\to 
+% $ 
+\setsem{\Gamma ;\ol{\gamma} \vdash H[\ol{\phi := G}]}\rho[\ol{\gamma
+      := C}]$$
+This family of morphisms is natural in $\ol C$: if $\ol{f : C \to C'}$
+then, writing $\xi$ for
+\[
+\setsem{\Gamma; \emptyset~|~\emptyset \vdash \map^{\ol{F},\ol{G}}_H
+    : \Nat^\emptyset\;(\ol{\Nat^{\ol{\beta},\ol{\gamma}}\,F\,G})\;
+    (\Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\,H[\ol{\phi
+        :=_{\ol{\beta}} G}])}\,\rho\,d\, \ol{\eta}
+\]
+the naturality of $\eta$,
+together with the fact that composition of environments is
+computed componentwise, ensure that the following naturality diagram
+for $\xi$ commutes:
+{\footnotesize
+\[\begin{tikzcd}
+\setsem{\Gamma ;\ol{\gamma} \vdash H[\ol{\phi := F}]}\rho[\ol{\gamma
+      := C}] \ar[r, "{\xi_{\ol{C}}}"]
+\ar[d, "{\setsem{\Gamma ;\ol{\gamma} \vdash H[\ol{\phi :=
+          F}]}\id_{\rho}[\ol{\gamma := f}]}"']
+& \setsem{\Gamma ; \ol{\gamma} \vdash H[\ol{\phi := G}]}\rho[\ol{\gamma
+      := C}]
+\ar[d, "{\setsem{\Gamma ; \ol{\gamma} \vdash H[\ol{\phi :=
+          G}]}\id_{\rho}[\ol{\gamma := f}]}"]\\
+\setsem{\Gamma ;\ol{\gamma} \vdash H[\ol{\phi := F}]}\rho[\ol{\gamma
+      := C'}] \ar[r, "{\xi_{\ol{C'}}}"]
+& \setsem{\Gamma ; \ol{\gamma} \vdash H[\ol{\phi := G}]}\rho[\ol{\gamma
+      := C'}] 
+\end{tikzcd}\]}
+That, for all $\rho : \setenv$ and $d : \setsem{\Gamma; \emptyset \vdash
+  \emptyset}\rho$, $\xi$ satisfies the additional condition needed for
+it to be in $\setsem{\Gamma; \emptyset \vdash
+  \Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\,H[\ol{\phi
+      :=_{\ol{\beta}} G}]}\rho$ follows from the fact
+that $\eta$ satisfies the extra
+condition needed for it to be in its corresponding
+$\setsem{\Gamma; \emptyset \vdash \Nat^{\ol{\beta},\ol{\gamma}}\,F\,G} \rho$.
+Finally, since $\Phi = \emptyset$, the naturality of
+\[
+\setsem{\Gamma; \emptyset~|~\emptyset \vdash \map^{\ol{F},\ol{G}}_H
+    : \Nat^\emptyset\;(\ol{\Nat^{\ol{\beta},\ol{\gamma}}\,F\,G})\;
+    (\Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\,H[\ol{\phi
+        :=_{\ol{\beta}} G}])}\rho
+\]
+in $\rho$ is trivial.
+
+\item
+$\underline{\Gamma;\emptyset \,|\, \emptyset \vdash \tin_H :
+  \Nat^{\ol{\beta}{\color{red},\ol{\gamma}}} \, H[\phi := (\mu \phi.\lambda
+    {\overline \alpha}.H){\overline \beta}][\ol{\alpha := \beta}]
+  \;(\mu \phi.\lambda {\overline \alpha}.H){\overline \beta}}$\; To
+  see that if $d : \setsem{\Gamma;\emptyset \vdash \emptyset} \rho$
+  then $$\setsem{\Gamma;\emptyset \,|\, \emptyset \vdash \tin_H :
+    \Nat^{\ol{\beta}{\color{red},\ol{\gamma}}} \, H[\phi := (\mu \phi.\lambda
+      {\overline \alpha}.H){\overline \beta}][\ol{\alpha := \beta}]
+    \;(\mu \phi.\lambda {\overline \alpha}.H){\overline \beta}}\,
+  \rho\,d$$ 
+  is in $\setsem{\Gamma;\emptyset \vdash
+    \Nat^{\ol{\beta}{\color{red},\ol{\gamma}}} \, H[\phi := (\mu \phi.\lambda
+      {\overline \alpha}.H){\overline \beta}][\ol{\alpha := \beta}]
+    \;(\mu \phi.\lambda {\overline \alpha}.H){\overline \beta}}\,
+  \rho$, we first note that, for all $\ol{B}$ {\color{red} and $\ol{C}$},
+  \begin{align*}
+    &(\setsem{\Gamma;\emptyset \,|\, \emptyset \vdash \tin_H :
+    \Nat^{\ol{\beta}{\color{red},\ol{\gamma}}} \, H[\phi := (\mu \phi.\lambda
+      {\overline \alpha}.H){\overline \beta}][\ol{\alpha := \beta}]
+    \;(\mu \phi.\lambda {\overline \alpha}.H){\overline \beta}}\,
+  \rho\,d)_{\ol{B}{\color{red}\,\ol{C}}}\, \\
+    =\,
+    &(\mathit{in}_{T^\set_{H,\rho{\color{red}[\ol{\gamma := C}]}}})_{\ol{B}}
+  \end{align*}
+  maps
+  \begin{align*}
+    &\setsem{\Gamma;\ol{\beta}{\color{red},\ol{\gamma}} \vdash H[\phi := (\mu
+      \phi.\lambda {\overline \alpha}.H){\overline \beta}][\ol{\alpha
+        := \beta}]}\rho[\ol{\beta := B}]{\color{red}[\ol{\gamma := C}]} 
+        \\ = \,
+    &T^\set_{H,\rho{\color{red}[\ol{\gamma := C}]}}\, (\mu T^\set_{H,\rho{\color{red}[\ol{\gamma
+        := C}]}}) \, \ol{B}
+  \end{align*}
+        to $\setsem{\Gamma;\ol{\beta},{\color{red}\ol{\gamma}}
+    \vdash (\mu \phi.\lambda {\overline \alpha}.H){\overline \beta}}
+  \rho[\ol{\beta := B}]{\color{red}[\ol{\gamma := C}]} = (\mu
+  T^\set_{H,\rho{\color{red}[\ol{\gamma := C}]}}) \, \ol{B}$.  \\
+  Secondly, we observe
+  that
+  \begin{align*}
+    &\setsem{\Gamma;\emptyset \,|\, \emptyset \vdash \tin_H :
+    \Nat^{\ol{\beta}{\color{red},\ol{\gamma}}} \, H[\phi := (\mu \phi.\lambda
+      {\overline \alpha}.H){\overline \beta}][\ol{\alpha := \beta}]
+    \;(\mu \phi.\lambda {\overline \alpha}.H){\overline
+      \beta}}\,\rho\,d \\ 
+      = \, &\lambda
+  \ol{B}{\color{red}\,\ol{C}}.\,(\mathit{in}_{T^\set_{H,\rho{\color{red}[\ol{\gamma :=
+          C}]}}})_{\ol{B}}
+  \end{align*}
+  is natural in $\ol{B}$ {\color{red} and $\ol{C}$}, since
+  naturality of $\mathit{in}$ with respect to its functor argument and
+  naturality of $\mathit{in}_{T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}}}$ ensure
+  that the following diagram commutes for all $\ol{f : B \to B'}$
+  {\color{red} and
+  $\ol{g : C \to C'}$}:
+
+  {\tiny
+  \[\hspace*{-0.15in}\begin{tikzcd}[column sep=2.5in, row sep=0.75in]
+%T^\set_{H,\rho[\ol{\gamma := C}]}\, (\mu T^\set_{H,\rho[\ol{\gamma := C}]})\, \ol{B}
+%\ar[d, "{T^\set_{H,\id_\rho[\ol{\gamma := g}]}\,(\mu
+%    T^\set_{H,\id_\rho[\ol{\gamma := g}]}) \, \ol{B}}"] \ar[r,
+%  "{(\mathit{in}_{T^\set_{H,\rho[\ol{\gamma := C}]}})_{\ol{B}}}" ]
+%& (\mu T^\set_{H,\rho[\ol{\gamma := C}]})\, \ol{B} \ar[d, "{(\mu
+%    T^\set_{H,\id_\rho[\ol{\gamma := g}]}) \, \ol{B}}"]\\
+T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}}\, (\mu
+T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}})\, \ol{B} \ar[d, "{T^\set_{H,\rho{\color{red}[\ol{\gamma :=
+          C'}]}}\, (\mu T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}})\, \ol{f}}"]
+\ar[r,"{(\mathit{in}_{T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}})_{\ol{B}}}}" ] & 
+  (\mu T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}})\, \ol{B}
+\ar[d,"{\mu {T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}}}\, {\ol{f}}}" ] 
+\\
+T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}}\, (\mu T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}})\,
+\ol{B'} \ar[r, "{(\mathit{in}_{T^\set_{H,\rho{\color{red}[\ol{\gamma :=
+            C'}]}}})_{\ol{B'}}}"] & (\mu T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}})\,
+\ol{B'}
+\end{tikzcd}\]
+}
+  That, for all $\rho : \setenv$ and $d :
+\setsem{\Gamma;\emptyset \vdash \emptyset}\rho$,
+\[\setsem{\Gamma;\emptyset \,|\, \emptyset \vdash
+  \tin_H : \Nat^{\ol{\beta}{\color{red},\ol{\gamma}}} \, H[\phi := (\mu \phi.\lambda
+    {\overline \alpha}.H){\overline \beta}][\ol{\alpha := \beta}]
+  \;(\mu \phi.\lambda {\overline \alpha}.H){\overline
+    \beta}}\,\rho\,d\] satisfies the additional property needed for
+it to be in 
+\[\setsem{\Gamma;\emptyset \vdash
+  \Nat^{\ol{\beta}{\color{red},\ol{\gamma}}} \, H[\phi := (\mu \phi.\lambda
+    {\overline \alpha}.H){\overline \beta}][\ol{\alpha := \beta}]
+  \;(\mu \phi.\lambda {\overline \alpha}.H){\overline \beta}}\,\rho\]
+follows from the fact that, for every $\ol{R : \rel(B,B')}$
+{\color{red} and $\ol{S : \rel(C,C')}$},
+\[\hspace*{-0.2in}
+  \begin{array}{ll}
+ & (\,(\setsem{\Gamma;\emptyset \,|\, \emptyset \vdash \tin_H :
+      \Nat^{\ol{\beta}{\color{red},\ol{\gamma}}} \, H[\phi := (\mu \phi.\lambda
+        {\overline \alpha}.H){\overline \beta}][\ol{\alpha := \beta}]
+      \;(\mu \phi.\lambda {\overline \alpha}.H){\overline
+        \beta}}\,\rho\,d)_{\ol{B}{\color{red},\ol{C}}},\,\\ & \hspace*{0.5in}(\setsem{\Gamma;\emptyset
+      \,|\, \emptyset \vdash \tin_H : \Nat^{\ol{\beta}{\color{red},\ol{\gamma}}} \,
+      H[\phi := (\mu \phi.\lambda {\overline \alpha}.H){\overline
+          \beta}][\ol{\alpha := \beta}] \;(\mu \phi.\lambda {\overline
+        \alpha}.H){\overline
+        \beta}}\,\rho\,d)_{\ol{B'}{\color{red},\ol{C'}}}\,)\\ &= (\,
+    (\mathit{in}_{T^\set_{H,\rho{\color{red}[\ol{\gamma := C}]}}})_{\ol{B}},
+    (\mathit{in}_{T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}}})_{\ol{B'}}\,)
+\end{array}\]
+has type
+\[\begin{array}{ll}
+& (\, T^\set_{H,\rho{\color{red}[\ol{\gamma := C}]}}\, (\mu
+T^\set_{H,\rho{\color{red}[\ol{\gamma := C}]}}) \, 
+\ol{B} \to (\mu T^\set_{H,\rho{\color{red}[\ol{\gamma := C}]}}) \, \ol{B}, \, \\
+& \hspace*{0.5in}T^\set_{H,\rho{\color{red} [\ol{\gamma := C'}]}}\, (\mu
+T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}}) \, 
+\ol{B'} \to (\mu T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}}) \,\ol{B'} \, )\\
+= &
+ \relsem{\Gamma;\ol{\beta},{\color{red}\ol{\gamma}} \vdash H[\phi := (\mu
+    \phi.\lambda {\overline \alpha}.H){\overline \beta}][\ol{\alpha :=
+      \beta}]}\Eq_\rho[\ol{\beta := R}]{\color{red}[\ol{\gamma := S}]} \to\\
+ & \hspace*{0.5in} \relsem{\Gamma;\ol{\beta}{\color{red},\ol{\gamma}} \vdash (\mu
+  \phi.\lambda \ol{\alpha}.H)\ol{\beta}} \Eq_\rho[\ol{\beta:=
+    R}]{\color{red}[\ol{\gamma :=S}]}
+\end{array}\]
+
+Finally, since $\Phi = \emptyset$, the naturality of
+\[
+\setsem{\Gamma;\emptyset \,|\, \emptyset \vdash \tin_H :
+  \Nat^{\ol{\beta}{\color{red},\ol{\gamma}}} \, H[\phi := (\mu \phi.\lambda
+    {\overline \alpha}.H){\overline \beta}][\ol{\alpha := \beta}]
+  \;(\mu \phi.\lambda {\overline \alpha}.H){\overline \beta}}
+\]
+in $\rho$ is trivial.
+
+\item
+$\underline{\Gamma; \emptyset~|~\emptyset \vdash \fold^F_H :
+  \Nat^\emptyset\;(\Nat^{\ol{\beta}{\color{red}, \ol{\gamma}}}\,H[\phi
+    :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\;
+  (\Nat^{{\ol{\beta}{\color{red},\ol{\gamma}}} }\,(\mu \phi.\lambda \overline
+  \alpha.H)\overline \beta \;F)}$ \; Since $\Phi$ is empty, to see
+  that 
+  $$\setsem{ \Gamma; \emptyset~|~\emptyset \vdash \fold^F_H :
+    \Nat^\emptyset\;(\Nat^{\ol{\beta}{\color{red}, \ol{\gamma}}}\,H[\phi
+      :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\;
+    (\Nat^{{\ol{\beta}{\color{red},\ol{\gamma}}} }\,(\mu \phi.\lambda \overline
+    \alpha.H)\overline \beta \;F)}$$
+    is a natural transformation from
+  $\setsem{\Gamma;\emptyset \vdash \emptyset}$ to \[\setsem{\Gamma;
+    \emptyset \vdash \Nat^\emptyset\;(\Nat^{\ol{\beta}{\color{red},
+      \ol{\gamma}}}\,H[\phi :=_{\ol{\beta}} F][\ol{\alpha :=
+        \beta}]\,F)\; (\Nat^{{\ol{\beta}{\color{red},\ol{\gamma}}} }\,(\mu
+    \phi.\lambda \overline \alpha.H)\overline \beta\,F)}\] we need only
+  show that, for all $\rho : \setenv$, the unique $d :
+  \setsem{\Gamma;\emptyset \vdash \emptyset} \rho$, and all $\eta :
+  \setsem{\Gamma; \emptyset \vdash \Nat^{\ol{\beta}{\color{red},
+      \ol{\gamma}}}\,H[\phi :=_{\ol{\beta}} F][\ol{\alpha :=
+        \beta}]\,F} \rho$,
+\[ \setsem{\Gamma; \emptyset~|~\emptyset \vdash \fold^F_H :
+  \Nat^\emptyset\;(\Nat^{\ol{\beta}{\color{red}, \ol{\gamma}}}\,H[\phi
+    :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\;
+  (\Nat^{{\ol{\beta}{\color{red},\ol{\gamma}}} }\,(\mu \phi.\lambda \overline
+  \alpha.H)\overline \beta\,F)}\,\rho\,d\,\eta\] has type
+$\setsem{\Gamma; \emptyset \vdash \Nat^{{\ol{\beta}{\color{red},\ol{\gamma}}}
+  }\,(\mu \phi.\lambda \overline \alpha.H)\overline \beta\,F}\,\rho$,
+i.e., for any $\ol{B}$ {\color{red} and $\ol{C}$},
+\[(\setsem{\Gamma; \emptyset~|~\emptyset \vdash \fold^F_H :
+  \Nat^\emptyset\;(\Nat^{\ol{\beta}{\color{red}, \ol{\gamma}}}\,H[\phi
+    :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\;
+  (\Nat^{{\ol{\beta}{\color{red},\ol{\gamma}}} }\,(\mu \phi.\lambda \overline
+  \alpha.H)\overline \beta\,F)}\,\rho\,d\,\eta)_{\ol{B}{\color{red}\,\ol{C}}}\] is a
+morphism from $\setsem{\Gamma; \ol{\beta}{\color{red},\ol{\gamma}} \vdash (\mu
+  \phi.\lambda \overline \alpha.H)\overline \beta}\rho[\ol{\beta :=
+    B}]{\color{red}[\ol{\gamma := C}]} \,=\,(\mu T^\set_{H,\rho{\color{red}[\ol{\gamma := C}]}})
+\ol{B}$\\ to $\setsem{\Gamma; \ol{\beta}{\color{red},\ol{\gamma}} \vdash
+  F}\rho[\ol{\beta := B}]{\color{red}[\ol{\gamma := C}]}$.  To see this, note
+that $\eta$ is a natural transformation from
+\[\begin{array}{ll}
+ & \lambda \ol{B}{\color{red}\,\ol{C}}.\,\setsem{\Gamma; \ol{\beta}{\color{red},\ol{\gamma}}
+  \vdash H[\phi := F][\ol{\alpha := \beta}]}\rho[\ol{\beta :=
+    B}]{\color{red}[\ol{\gamma := C}]}\\
+= & \lambda \ol{B}{\color{red}\,\ol{C}}.\,T^\set_{H,\rho{\color{red}[\ol{\gamma:=
+     C}]}}\,(\lambda \ol{A}. \, \setsem{\Gamma;\ol{\beta}{\color{red},\ol{\gamma}} 
+  \vdash F}\rho[\ol{\beta := A}]{\color{red}[\ol{\gamma := C}]}) \, \ol{B}
+\end{array}\]
+to
+\[\begin{array}{ll}
+ & \lambda \ol{B}{\color{red}\,\ol{C}}.\,(\lambda
+\ol{A}.\,\setsem{\Gamma;\ol{\beta}{\color{red},\ol{\gamma}} \vdash F}\rho[\ol{\beta
+    := A}]{\color{red}[\ol{\gamma := C}]}) \ol{B}\\
+= & \lambda
+\ol{B}{\color{red}\,\ol{C}}.\,\setsem{\Gamma;\ol{\beta}{\color{red},\ol{\gamma}} \vdash
+  F}\rho[\ol{\beta := B}]{\color{red}[\ol{\gamma := C}]}
+\end{array}\]
+and thus for each $\ol{B}$ {\color{red} and $\ol{C}$},
+\[(\setsem{\Gamma; \emptyset~|~\emptyset \vdash \fold^F_H :
+  \Nat^\emptyset\;(\Nat^{\ol{\beta}{\color{red}, \ol{\gamma}}}\,H[\phi
+    :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\;
+  (\Nat^{{\ol{\beta}{\color{red},\ol{\gamma}}} }\,(\mu \phi.\lambda \overline
+  \alpha.H)\overline \beta\,F)}\,\rho\,d\,\eta)_{\ol{B}{\color{red}\,\ol{C}}}\] is a
+morphism from
+$\setsem{\Gamma; \ol{\beta}{\color{red},\ol{\gamma}} \vdash (\mu
+  \phi.\lambda \overline \alpha.H)\overline \beta}\rho[\ol{\beta :=
+    B}]{\color{red}[\ol{\gamma := C}]}\,=\, (\mu T^\set_{H,\rho{\color{red}[\ol{\gamma := C}]}})
+\ol{B}$ to $\setsem{\Gamma;\ol{\beta}{\color{red},\ol{\gamma}} \vdash
+  F}\rho[\ol{\beta := B}]{\color{red}[\ol{\gamma := C}]}$.
+To see that this family of morphisms is natural in $\ol{B}$
+{\color{red} and
+$\ol{C}$}, we observe that the following diagram commutes for all
+$\ol{f : B \to B'}$ {\color{red} and $\ol{g : C \to C'}$}:
+{\tiny
+\[\hspace*{-0.3in}\begin{tikzcd}[column sep=2.5in, row sep=0.75in]
+%(\mu T^\set_{H,\rho[\ol{\gamma := C}]})\, \ol{B}
+%\ar[d, "{(\mu T^\set_{H,\id_\rho[\ol{\gamma := g}]}) \, \ol{B}}"'] \ar[r, 
+%  "{(\mathit{fold}_{T^\set_{H,\rho[\ol{\gamma := C}]}}\,(\lambda
+%    \ol{A}.\,\eta_{\ol{A}\,\ol{C}}))_{\ol{B}}}"] 
+%& \setsem{\Gamma; \ol{\beta},\ol{\gamma} \vdash F}\rho[\ol{\gamma :=
+%    C}][\ol{\beta := B}]\ar[d, "{\setsem{\Gamma;
+%      \ol{\beta},\ol{\gamma} \vdash F}\id_\rho[\ol{\gamma := 
+%    g}][\ol{\beta := \id_B}]}"]\\
+(\mu T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}})\, \ol{B} 
+\ar[r,"{(\mathit{fold}_{T^\set_{H,\rho{\color{red}[\ol{\gamma :=
+            C'}]}}}\,(\lambda \ol{A}.\,\eta_{\ol{A}{\color{red}\,\ol{C'}}}))_{\ol{B}}\,}" ] 
+\ar[d, "{(\mu T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}}) \, \ol{f}}"'] & 
+ \setsem{\Gamma; \ol{\beta}{\color{red},\ol{\gamma}} \vdash F}\rho{\color{red}[\ol{\gamma :=
+    C'}]}[\ol{\beta := B}]\ar[d, "{\setsem{\Gamma;
+      \ol{\beta}{\color{red},\ol{\gamma}} \vdash F}\id_\rho{\color{red}[\ol{\gamma := 
+    \id_{C'}}]}[\ol{\beta := f}]}"]\\
+ (\mu T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}})\, \ol{B'} 
+\ar[r,"{(\mathit{fold}_{T^\set_{H,\rho{\color{red}[\ol{\gamma :=
+            C'}]}}}\,(\lambda \ol{A}.\,\eta_{\ol{A}{\color{red}\,\ol{C'}}}))_{\ol{B'}}\,}"
+] &  
+ \setsem{\Gamma; \ol{\beta}{\color{red},\ol{\gamma}} \vdash F}\rho{\color{red}[\ol{\gamma :=
+    C'}]}[\ol{\beta := B'}]
+\end{tikzcd}\]}
+%Indeed,
+by naturality of $\mathit{fold}_{T^\set_{H,\rho[{\color{red}\ol{\gamma
+          := C'}]}}}\,(\lambda
+\ol{A}.\,\eta_{\ol{A}{\color{red}\,\ol{C'}}})$.
+%
+%ensures that the bottom diagram commutes. {\color{red} BOTTOM!}
+%To see that the top one commutes we first observe that, given a
+%natural transformation $\Theta : H \to K : [\set^k, \set] \to [\set^k,
+%  \set]$, the fixpoint natural transformation $\mu \Theta : \mu H \to
+%\mu K : \set^k \to \set$ is defined to be
+%$\mathit{fold}_{H}(\Theta\,(\mu K) \circ \mathit{in}_{K})$, i.e., the
+%unique morphism making the following diagram commute:\label{page:dia1}
+%{\footnotesize
+%\[\begin{tikzcd}[column sep = large]
+%H(\mu H)
+%	\ar[dd, "{\mathit{in}_H}"']
+%	\ar[r, "{H(\mu \Theta)}"]
+%& H(\mu K)
+%	\ar[d, "{\Theta (\mu K)}"] \\
+%& K(\mu K)
+%	\ar[d, "{\mathit{in}_K}"] \\
+%\mu H
+%	\ar[r, "{\mu \Theta}"']
+%& \mu K
+%\end{tikzcd}\]}
+%Taking $\Theta = T^{\set}_{H,f}: T^{\set}_{H,\rho} \to
+%T^{\set}_{H,\rho'}$ thus gives that, for any $f : \rho \to \rho'$ in
+%$\setenv$,
+%\begin{equation}\label{eq:mu-sigma-def}
+%\mathit{in}_{T^{\set}_{H,\rho'}} \circ 
+%T^{\set}_{H,f} (\mu T^{\set}_{H,\rho'}) \circ 
+%T^{\set}_{H,\rho}(\mu T^{\set}_{H,f}) \,=\, 
+%\mu T^{\set}_{H,f} \circ \mathit{in}_{T^{\set}_{H,\rho}}
+%\end{equation}
+Next, note that the action of the functor
+\[\lambda \ol{B}.{\color{red} \lambda \ol{C}.} \setsem{\Gamma; \ol{\beta}{\color{red},
+  \ol{\gamma}} \vdash H[\phi := F][\ol{\alpha := \beta}]}\rho
+        [\ol{\beta := B}] {\color{red}[\ol{\gamma := C}]}\]
+on the morphisms $\ol{f : B \to B'}{\color{red}, \ol{g : C \to C'}}$ is given by
+\[\begin{array}{ll}
+ & \setsem{\Gamma; \ol{\beta{\color{red}, \ol{\gamma}}} \vdash H[\phi :=
+      F][\ol{\alpha := \beta}]} \id_{\rho} [\ol{\beta := f}]
+           {\color{red}[\ol{\gamma := g}]}\\
+= & \setsem{\Gamma; \phi, \ol{\alpha}{\color{red}, \ol{\gamma}} \vdash H}
+\id_{\rho}[\ol{\alpha := f}] {\color{red}[\ol{\gamma := g}]}[\phi := \lambda
+  \ol{A}. \setsem{\Gamma; \ol{\beta}{\color{red}, \ol{\gamma}} \vdash F} \id_{\rho
+    [\ol{\beta := A}]}{\color{red} [\ol{\gamma := g}]}] \\ 
+=& \setsem{\Gamma; \phi, \ol{\alpha}{\color{red}, \ol{\gamma}} \vdash H} \id_{\rho
+  {\color{red}[\ol{\gamma := C'}]}[\phi := \lambda \ol{A}. \setsem{\Gamma;
+      \ol{\beta} {\color{red}, \ol{\gamma}} \vdash F} \rho [\ol{\beta := A}]
+   {\color{red} [\ol{\gamma := C'}]}]} [\ol{\alpha := f}] \\
+&\hspace{3em} \circ \setsem{\Gamma; \phi, \ol{\alpha}{\color{red}, \ol{\gamma}}
+  \vdash H} \id_{\rho [\ol{\alpha := B}][\phi := \lambda
+    \ol{A}. \setsem{\Gamma; \ol{\beta}{\color{red}, \ol{\gamma}} \vdash F} \rho
+       [\ol{\beta := A}]{\color{red} [\ol{\gamma := C'}]}]} {\color{red}[\ol{\gamma := g}]} \\
+&\hspace{3em} \circ \setsem{\Gamma; \phi, \ol{\alpha} {\color{red}, \ol{\gamma}}
+  \vdash H} \id_{\rho [\ol{\alpha := B}] {\color{red}[\ol{\gamma := C}]}} [\phi :=
+  \lambda \ol{A}. \setsem{\Gamma; \ol{\beta}{\color{red}, \ol{\gamma}} \vdash F}
+  \id_{\rho [\ol{\beta := A}]}{\color{red} [\ol{\gamma := g}]}] \\
+= & T^{\set}_{H,\rho {\color{red}[\ol{\gamma := C'}]}} (\lambda
+\ol{A}. \setsem{\Gamma; \ol{\beta} {\color{red}, \ol{\gamma}} \vdash F} \rho
+   [\ol{\beta := A}]{\color{red} [\ol{\gamma := C'}]}) \ol{f} \\
+&\hspace{3em} \circ \big( T^{\set}_{H,\id_{\rho}{\color{red}[\ol{\gamma := g}]}}
+   (\lambda \ol{A}. \setsem{\Gamma; \ol{\beta} {\color{red}, \ol{\gamma}} \vdash F}
+   \rho [\ol{\beta := A}] {\color{red}[\ol{\gamma := C'}]}) \big)_{\ol{B}} \\
+&\hspace{3em} \circ \big( T^{\set}_{H,\rho {\color{red}[\ol{\gamma := C}]}} (\lambda
+   \ol{A}. \setsem{\Gamma; \ol{\beta}{\color{red}, \ol{\gamma} }\vdash F} \id_{\rho
+     [\ol{\beta := A}]} {\color{red}[\ol{\gamma := g}]}) \big)_{\ol{B}}
+\end{array}\]
+So if $\eta$ is a natural transformation such that $\eta_{\ol B {\color{red}, \ol
+  C}}$ has type
+\[
+\setsem{\Gamma; \ol{\beta},
+  {\color{red}\ol{\gamma}} \vdash H[\phi := F][\ol{\alpha := \beta}]}\rho
+[\ol{\beta := B}] {\color{red}[\ol{\gamma := C}]}
+\to
+\setsem{\Gamma; \ol{\beta},
+  {\color{red} \ol{\gamma}} \vdash F}\rho [\ol{\beta := B}]
+   {\color{red}    [\ol{\gamma := C}]}\] 
+then, by naturality,
+\[\begin{array}{ll}
+ & \setsem{\Gamma; \ol{\beta}{\color{red}, \ol{\gamma}} \vdash F} \id_{\rho}
+       [\ol{\beta := f}] {\color{red}[\ol{\gamma := g}]} \circ \eta_{\ol{B}{\color{red},
+         \ol{C}}} \\ 
+= & \eta_{\ol{B'}{\color{red}, \ol{C'}}} \circ T^{\set}_{H,\rho{\color{red} [\ol{\gamma := C'}]}}
+(\lambda \ol{A}. \setsem{\Gamma; \ol{\beta} {\color{red}, \ol{\gamma}} \vdash F}
+\rho [\ol{\beta := A}]{\color{red} [\ol{\gamma := C'}]}) \ol{f} \\
+& \circ \big( T^{\set}_{H,\id_{\rho}{\color{red}[\ol{\gamma := g}]}} (\lambda
+\ol{A}. \setsem{\Gamma; \ol{\beta}{\color{red}, \ol{\gamma}} \vdash F} \rho
+   [\ol{\beta := A}]{\color{red} [\ol{\gamma := C'}]}) \big)_{\ol{B}} \\
+& \circ \big( T^{\set}_{H,\rho {\color{red}[\ol{\gamma := C}]}} (\lambda
+   \ol{A}. \setsem{\Gamma; \ol{\beta}{\color{red}, \ol{\gamma}} \vdash F} \id_{\rho
+     [\ol{\beta := A}]} {\color{red}[\ol{\gamma := g}]}) \big)_{\ol{B}}
+\end{array}\]
+As a special case when $\ol{f = \id_B}$ we have
+\begin{equation}\label{eq:T-sigma-functor}
+\begin{split}
+  & \lambda \ol{B}. \setsem{\Gamma; \ol{\beta} {\color{red}, \ol{\gamma}} \vdash F}
+\id_{\rho [\ol{\beta := B}]} {\color{red}[\ol{\gamma := g}]} \circ \lambda
+\ol{B}.\eta_{\ol{B}{\color{red}, \ol{C}}} \\  
+=\;\;\; & \lambda \ol{B}.\eta_{\ol{B}{\color{red}, \ol{C'}}} \circ
+T^{\set}_{H,\id_{\rho}{\color{red}[\ol{\gamma := g}]}} (\lambda
+\ol{A}. \setsem{\Gamma; \ol{\beta}{\color{red}, \ol{\gamma}} \vdash F} \rho
+   [\ol{\beta := A}]{\color{red} [\ol{\gamma := C'}]}) \\
+ & \hspace*{0.5in} \circ
+   T^{\set}_{H,\rho {\color{red}[\ol{\gamma := C}]}} (\lambda \ol{A}. \setsem{\Gamma;
+     \ol{\beta}{\color{red}, \ol{\gamma}} \vdash F} \id_{\rho [\ol{\beta := A}]}
+ {\color{red}  [\ol{\gamma := g}]}) 
+\end{split}
+\end{equation}
+
+HERE!!!!
+
+Finally, to see that the top diagram in the diagram on
+page~\pageref{page:dia1} commutes we first note that functoriality of
+$T^{\set}_{H,\rho [\ol{\gamma := C}]}$, naturality of
+$T^{\set}_{H,\id_{\rho}[\ol{\gamma := g}]}$, the universal property of
+$\mathit{fold}_{T^{\set}_{H,\rho [\ol{\gamma := C'}]}} (\lambda
+\ol{A}. \eta_{\ol{A}, \ol{C'}})$ and Equation~\ref{eq:mu-sigma-def}
+ensure that the following diagram commutes: {\footnotesize
+\begin{equation}\label{eq:one-side}
+\hspace*{-0.7in}  \begin{tikzcd}[column sep = huge, row sep = huge]
+T^{\set}_{H,\rho [\ol{\gamma := C}]} (\mu T^{\set}_{H,\rho [\ol{\gamma :=
+      C}]}) \ar[rr, "{T^{\set}_{H,\rho [\ol{\gamma := C}]} (
+    \mathit{fold}_{T^{\set}_{H,\rho [\ol{\gamma := C'}]}} (\lambda
+    \ol{A}. \eta_{\ol{A}, \ol{C'}}) \circ \mu
+    T^{\set}_{H,\id_{\rho}[\ol{\gamma := g}]} )}"] \ar[dd,
+    "{\mathit{in}_{T^{\set}_{H,\rho [\ol{\gamma := C}]}}}"']
+&& T^{\set}_{H,\rho [\ol{\gamma := C}]} (\lambda \ol{B}. \setsem{\Gamma;
+  \ol{\beta}, \ol{\gamma} \vdash F} \rho [\ol{\beta := B}] [\ol{\gamma
+    := C'}]) \ar[d, "{ T^{\set}_{H,\id_{\rho}[\ol{\gamma := g}]}
+    (\lambda \ol{B}. \setsem{\Gamma; \ol{\beta}, \ol{\gamma} \vdash F}
+    \rho [\ol{\beta := B}] [\ol{\gamma := C'}]) }" description] \\
+&& T^{\set}_{H,\rho [\ol{\gamma := C'}]} (\lambda
+\ol{B}. \setsem{\Gamma; \ol{\beta}, \ol{\gamma} \vdash F} \rho
+   [\ol{\beta := B}][\ol{\gamma := C'}]) \ar[d, "{ \lambda \ol{A}. \eta_{\ol{A}, \ol{C'}}
+     }" description] \\
+\mu T^{\set}_{H,\rho [\ol{\gamma := C}]}\ar[r, "{\mu
+    T^{\set}_{H,\id_{\rho}[\ol{\gamma := g}]}}"'] 
+&\mu T^{\set}_{H,\rho [\ol{\gamma := C'}]} \ar[r, bend right = 10,
+  "{\mathit{fold}_{T^{\set}_{H,\rho [\ol{\gamma := C'}]}} (\lambda
+    \ol{A}. \eta_{\ol{A}, \ol{C'}})}"']
+& \lambda \ol{B}. \setsem{\Gamma; \ol{\beta}, \ol{\gamma} \vdash F}
+\rho [\ol{\beta := B}] [\ol{\gamma := C'}]
+  \end{tikzcd}
+  \end{equation}}
+Next, we note that functoriality of $T^{\set}_{H,\rho [\ol{\gamma :=
+      C}]}$, Equation~\ref{eq:T-sigma-functor}, and the universal
+property of $\mathit{fold}_{T^{\set}_{H,\rho [\ol{\gamma := C}]}} (\lambda
+\ol{A}. \eta_{\ol{A}, \ol{C}})$ ensure that the following diagram
+commutes: {\footnotesize
+\begin{equation}\label{eq:other-side}
+\hspace*{-1in}\begin{tikzcd}[row sep = huge]
+T^{\set}_{H,\rho [\ol{\gamma := C}]} (\mu T^{\set}_{H,\rho [\ol{\gamma :=
+      C}]}) \ar[rr, "{T^{\set}_{H,\rho [\ol{\gamma := C}]} (
+    \lambda \ol{B}.\setsem{\Gamma; \ol{\beta}, \ol{\gamma} \vdash F} \id_{\rho
+      [\ol{\beta := B}]} [\ol{\gamma := g}] \circ
+    \mathit{fold}_{T^{\set}_{H,\rho [\ol{\gamma := C}]}} (\lambda
+    \ol{A}. \eta_{\ol{A}, \ol{C}}) )}"] \ar[dd, "{\mathit{in}_{T^{\set}_{H,\rho
+        [\ol{\gamma := C}]}}}"']
+&& T^{\set}_{H,\rho [\ol{\gamma := C}]} (\lambda \ol{B}.\setsem{\Gamma; \ol{\beta},
+  \ol{\gamma} \vdash F} \rho [\ol{\beta := B}] [\ol{\gamma := C'}])
+\ar[d, "{ T^{\set}_{H,\id_{\rho}[\ol{\gamma := g}]}
+    (\lambda \ol{B}.\setsem{\Gamma; \ol{\beta}, \ol{\gamma} \vdash F} \rho [\ol{\beta
+        := B}] [\ol{\gamma := C'}]) }" description]\\
+&& T^{\set}_{H,\rho [\ol{\gamma := C'}]} (\lambda \ol{B}.\setsem{\Gamma; \ol{\beta},
+  \ol{\gamma} \vdash F} \rho [\ol{\beta := B}][\ol{\gamma := C'}]) \ar[d, "{ \lambda
+    \ol{A}. \eta_{\ol{A}, \ol{C'}} }" description] \\
+\mu T^{\set}_{H,\rho [\ol{\gamma := C}]} \ar[r, bend right = 10,
+  "{\mathit{fold}_{T^{\set}_{H,\rho [\ol{\gamma := C}]}} (\lambda
+    \ol{A}. \eta_{\ol{A}, \ol{C}})}"'] & \lambda \ol{B}.\setsem{\Gamma; \ol{\beta},
+  \ol{\gamma} \vdash F} \rho [\ol{\beta := B}] [\ol{\gamma := C}]
+\ar[r, bend right = 10, "{\lambda \ol{B}.\setsem{\Gamma; \ol{\beta}, \ol{\gamma}
+      \vdash F} \id_{\rho [\ol{\beta := B}]} [\ol{\gamma := g}]}"'] &
+\lambda \ol{B}.\setsem{\Gamma; \ol{\beta}, \ol{\gamma} \vdash F} \rho [\ol{\beta :=
+    B}] [\ol{\gamma := C'}]
+\end{tikzcd}
+\end{equation}}
+Combining the equations entailed by~\ref{eq:one-side}
+and~\ref{eq:other-side}, we get that
+the top diagram in the diagram on
+  page~\pageref{page:dia1} commutes, as desired.
+To see that, for all $\rho : \setenv$, $d \in
+\setsem{\Gamma; \emptyset \vdash \emptyset}\rho$, and $\eta :
+\setsem{\Gamma; \emptyset \vdash \Nat^{\ol{\beta},
+    \ol{\gamma}}\,H[\phi :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F}
+\rho$,
+\[\setsem{\Gamma; \emptyset~|~\emptyset
+  \vdash \fold^F_H : \Nat^\emptyset\;(\Nat^{\ol{\beta},
+    \ol{\gamma}}\,H[\phi :=_{\ol{\beta}} F][\ol{\alpha :=
+      \beta}]\,F)\; (\Nat^{{\ol{\beta},\ol{\gamma}} }\,(\mu
+  \phi.\lambda \overline \alpha.H)\overline \beta\,F)}\,\rho\,d\,\eta\]
+satisfies the additional condition needed for it to be in
+$\setsem{\Gamma;\emptyset \vdash \Nat^{{\ol{\beta},\ol{\gamma}}
+  }\,(\mu \phi.\lambda \overline \alpha.H)\overline \beta\;F}\,\rho$,
+let $\ol{R : \rel(B,B')}$ and $\ol{S : \rel(C,C')}$.  Since $\eta$
+satisfies the additional condition needed for it to be in
+$\setsem{\Gamma; \emptyset \vdash \Nat^{\ol{\beta},
+    \ol{\gamma}}\,(H[\phi := F][\ol{\alpha := \beta}])\,F} \rho$,
+\[\begin{array}{ll}
+ & (\,(\mathit{fold}_{T^\set_{H,\rho[\ol{\gamma :=
+        C}]}}\,(\lambda \ol{A}.\,\eta_{\ol{A}\,\ol{C}}))_{\ol{B}},\,
+(\mathit{fold}_{T^\set_{H,\rho[\ol{\gamma :=
+        C'}]}}\,(\lambda \ol{A}.\eta_{\ol{A}\,\ol{C'}}))_{\ol{B'}}\,) 
+\end{array}\]
+has type
+\[\begin{array}{ll}
+  & (\mu T_{H,\Eq_\rho[\ol{\gamma := S}]}) \,\ol{R} \to
+\relsem{\Gamma;\ol{\gamma},\ol{\beta} \vdash F}\Eq_\rho[\ol{\gamma := 
+    S}][\ol{\beta:= R}]\\ 
+= & (\mu T_{H,\Eq_\rho[\ol{\gamma :=
+      S}]})\,\ol{\relsem{\Gamma;\ol{\gamma},\ol{\beta} 
+  \vdash \beta}\Eq_\rho[\ol{\gamma := S}][\ol{\beta := R}]} \\ 
+  &\hspace{1.5in} \to
+\relsem{\Gamma;\ol{\gamma},\ol{\beta} \vdash F}\Eq_\rho[\ol{\gamma := 
+    S}][\ol{\beta:= R}]\\ 
+= & \relsem{\Gamma; \ol{\gamma},\ol{\beta} \vdash (\mu \phi. \lambda
+  \ol{\alpha}. H)\ol{\beta}}\Eq_\rho[\ol{\gamma := S}][\ol{\beta := R}] \\ 
+  &\hspace{1.5in}\to
+\relsem{\Gamma;\ol{\gamma},\ol{\beta} \vdash F}\Eq_\rho[\ol{\gamma := 
+    S}][\ol{\beta:= R}]
+\end{array}\]
+\end{itemize}
 \qed
@@
 
-In this section we show how Theorem~\ref{thm:abstraction} can be used
-to prove free theorems, such as that for the type of the standard
-$\mathit{filter}$ function for lists, that go beyond mere naturality,
-and extend them to those nested types for which analogous functions
-can be defined. In particular, we can extend short cut short cut
+In this section we show how Theorem~\ref{thm:at-gen} and its
+consequences can be used to prove free theorems that go beyond mere
+naturality.  We also show that we can extend short cut short cut
 fusion for lists~\cite{glp93} to nested types, thereby formally
@@
 
-Suppose $ \vdash g : \Nat^\alpha \,\onet\,\alpha$, let $G^\set =
-\setsem{\vdash g : \Nat^\alpha \,\onet\,\alpha}$, and let $G^\rel =
+Let $ \vdash g : \Nat^\alpha \,\onet\,\alpha$, let $G^\set =
+\setsem{\vdash g : \Nat^\alpha \,\onet\,\alpha}$ and $G^\rel =
 \relsem{\vdash g : \Nat^\alpha \,\onet\,\alpha}$.  By
-Theorem~\ref{thm:at-gen}, $(G^\set(\pi_1\rho),G^\set(\pi_2\rho))
-= G^\rel\rho$. Thus, for all $\rho \in \relenv$ and any $(a, b) \in
+Theorem~\ref{thm:at-gen}, $(G^\set(\pi_1\rho),G^\set(\pi_2\rho)) =
+G^\rel\rho$. Thus, for all $\rho \in \relenv$ and any $(a, b) \in
 \relsem{\vdash \emptyset}\rho = 1$, eliding the only possible
-instantiations of $a$ and $b$ gives that
-%\[\begin{array}{lll}
-$(G^\set,G^\set) \;= \; (G^\set(\pi_1 \rho), G^\set (\pi_2 \rho))$
-$ \in  \relsem{\vdash \Nat^\alpha \,\onet\,\alpha}\rho$
-$ =  \{\eta : K_1 \Rightarrow \id\}$
-$ =  \{(\eta_1 : K_1 \Rightarrow \id, \eta_2 : K_1 \Rightarrow
-\id)\}$ 
-%\end{array}\]
-That is, $G^\set$ is a natural transformation from the constantly
-$1$-valued functor to the identity functor in $\set$. In particular,
-for every $S : \set$, $G^\set_S : 1 \to S$. Note, however, that if $S
-= \emptyset$, then there can be no such morphism, so no such natural
-transformation,
-%in $\set$,
-and thus no term $\vdash g : \Nat^\alpha \onet \,\alpha$, can exist.
-%in our calculus.
-That is, our calculus admits no non-terminating terms, i.e., terms
-with the closed type $\Nat^\alpha \onet \,\alpha$ of the polymorphic
-bottom.
+instantiations of $a$ and $b$ gives that $(G^\set,G^\set) \;= \;
+(G^\set(\pi_1 \rho), G^\set (\pi_2 \rho))$ $ \in \relsem{\vdash
+  \Nat^\alpha \,\onet\,\alpha}\rho$ $ = \{\eta : K_1 \Rightarrow
+\id\}$ $ = \{(\eta_1 : K_1 \Rightarrow \id, \eta_2 : K_1 \Rightarrow
+\id)\}$ That is, $G^\set$ is a natural transformation from the
+constantly $1$-valued functor to the identity functor in $\set$. In
+particular, for every $S : \set$, $G^\set_S : 1 \to S$. Note, however,
+that if $S = \emptyset$, then there can be no such morphism, so no
+such natural transformation, and thus no term $\vdash g : \Nat^\alpha
+\onet \,\alpha$, can exist.  That is, our calculus admits no
+non-terminating terms, i.e., terms with the closed type $\Nat^\alpha
+\onet \,\alpha$ of the polymorphic bottom.
 
@@
 
-Suppose $ \vdash g : \Nat^\alpha \,\alpha\,\alpha$, let $G^\set =
-\setsem{\vdash g : \Nat^\alpha \,\alpha\,\alpha}$, and let $G^\rel =
+Let $ \vdash g : \Nat^\alpha \,\alpha\,\alpha$, \/$G^\set =
+\setsem{\vdash g : \Nat^\alpha \,\alpha\,\alpha}$, and $G^\rel =
 \relsem{\vdash g : \Nat^\alpha \,\alpha\,\alpha}$.  By
-Theorem~\ref{thm:at-gen}, $(G^\set(\pi_1\rho),G^\set(\pi_2\rho))
-= G^\rel\rho$. Thus, for all $\rho \in \relenv$ and any $(a, b) \in
+Theorem~\ref{thm:at-gen}, $(G^\set(\pi_1\rho),G^\set(\pi_2\rho)) =
+G^\rel\rho$. Thus, for all $\rho \in \relenv$ and any $(a, b) \in
 \relsem{\vdash \emptyset}\rho = 1$, eliding the only possible
-instantiations of $a$ and $b$ gives that
-%\[\begin{array}{lll}
-$(G^\set, G^\set) \; = \; (G^\set(\pi_1 \rho), G^\set (\pi_2 \rho))$
-$ \in \relsem{\vdash \Nat^\alpha \,\alpha\,\alpha}\rho$
-$ =  \{\eta : \id \Rightarrow \id\}$
-$ =  \{(\eta_1 : \id \Rightarrow \id, \eta_2 : \id \Rightarrow
-\id)\}$
-%\end{array}\]
-That is, $G^\set$ is a natural transformation from the identity
-functor on $\set$ to itself. Now let $S$ be any set.  If $S =
+instantiations of $a$ and $b$ gives that $(G^\set, G^\set) \; = \;
+(G^\set(\pi_1 \rho), G^\set (\pi_2 \rho))$ $ \in \relsem{\vdash
+  \Nat^\alpha \,\alpha\,\alpha}\rho$ $ = \{\eta : \id \Rightarrow
+\id\}$ $ = \{(\eta_1 : \id \Rightarrow \id, \eta_2 : \id \Rightarrow
+\id)\}$ That is, $G^\set$ is a natural transformation from the
+identity functor on $\set$ to itself. Now let $S$ be any set.  If $S =
 \emptyset$, then there is exactly one morphism $\id_S: S \to S$, so
-$G^\set_S : S \to S$ must be $\id_S$. If $S \not = \emptyset$, if
-$a$ is any element of $S$, and if $K_a :S \to S$ is the constantly
+$G^\set_S : S \to S$ must be $\id_S$. If $S \not = \emptyset$, if $a$
+is any element of $S$, and if $K_a :S \to S$ is the constantly
 $a$-valued morphism on $S$, then instantiating the naturality square
@@
 
-\subsection{Standard Free Theorems for ADTs and Their Analogues for
-  Nested types}\label{sec:ft-adt}
+\subsection{Standard Free Theorems for ADTs and Nested
+  types}\label{sec:ft-adt}
 
@@
 
-%\begin{lemma}\label{lem:list-graph}
-%If $g : A \to B$, $\rho : \relenv$, and $\rho\alpha = (A, B,
-%\graph{g})$, then $\relsem{\alpha; \emptyset \vdash List \, \alpha}
-%\rho = \graph{map_{\mathit{List}} \, g}$.
-%\end{lemma}
-%\begin{proof}
-%The result follows by straightforward calculation using
-%Definition~\ref{def:rel-sem} and Equation~\ref{eq:mu}, together with
-%the facts that $\mathit{List} \,(A, B, \graph{g}) =
-%\graph{\mathit{map}_{\mathit{List}} \,g}$ and, if $H = \onet + \beta
-%\times\phi \beta$ {\color{blue} Why not $H = \onet + \alpha \times
-%  \beta$? Does it matter?}, then $(T_{H,\rho}^n K_0)^* R =
-%\Sigma_{i=0}^n R^i$ for all $n \in \nat$ and $R \in \rel$ by induction
-%on $n$.
-%\end{proof}
-
 \begin{thm} 
-If $g : A \to B$, $\rho : \relenv$, $\rho \alpha = (A, B, \graph{g})$,
+If $g : A \to B$, $\rho \in \relenv$, $\rho \alpha = (A, B, \graph{g})$,
 $(a, b) \in \relsem{\alpha ;\emptyset \vdash \Delta} \rho$, $(s \circ
 g, s) \in \relsem{\alpha; \emptyset \vdash \Nat^\emptyset \alpha \,
-  \mathit{Bool}} \rho$, and % for $s : A \to \mathit{Bool}$, and
-%$\alpha; \emptyset \,|\, \Delta \vdash t :
-%\filtype$, if
-$\mathit{filter} = \setsem{\alpha; \emptyset \,|\, \Delta \vdash t :
-  \filtype}$ for some $t$, then
+  \mathit{Bool}} \rho$, and
+\[\mathit{filter} = \setsem{\alpha; \emptyset \,|\, \Delta \vdash t :
+  \filtype}\]
+\noindent
+for some $t$, then
 \[  \mathit{map}_{\mathit{List}} \,g \circ \mathit{filter} \, (\pi_1
@@
 \proof
-By Theorem~\ref{thm:at-gen-rel}, \[(\mathit{filter}\, (\pi_1 \rho)\, a,
-\mathit{filter}\, (\pi_2 \rho)\, b) \in \relsem{\alpha; \emptyset
+By Theorem~\ref{thm:abstraction}, \[(\mathit{filter}\, (\pi_1 \rho)\,
+a, \mathit{filter}\, (\pi_2 \rho)\, b) \in \relsem{\alpha; \emptyset
   \vdash \filtype} \rho\] so if $(s', s) \in \relsem{\alpha; \emptyset
@@
   \vdash List \, \alpha} \rho$ then
-% then
-%  \begin{align*}
-%    (t (\pi_1 \rho) \,a \,s, t (\pi_2 \rho) \,b \,s') &\in \relsem{\alpha; \emptys%et \vdash \Nat^\emptyset (List \, \alpha) \, (List \, \alpha)} \rho \\
-%   &= \relsem{\alpha; \emptyset \vdash List \, \alpha} \rho \to \relsem{\alpha; \e%mptyset \vdash List \, \alpha} \rho \\
-%  \end{align*}
-%  So if $(xs, xs') \in \relsem{\alpha; \emptyset \vdash List \, \alpha} \rho$ then,
 \begin{equation}\label{eq:filter-thm-list}
@@
 \graph{\mathit{map}_{\mathit{List}} \, g}$ by Lemma~\ref{lem:graph}
-and Equation~\ref{thm:demotion-objects}.  Moreover, $xs =
+and demotion.  Moreover, $xs =
 \mathit{map}_{\mathit{List}} \,g \,xs'$ and $(s', s) \in \graph{g} \to
 \Eq_{\mathit{Bool}}$, so $s' = s \circ g$. The result
-follows from % Instantiating
-Equation~\ref{eq:filter-thm-list}.
-\qed
-%gives the result.
-% \begin{align*}
-%    &(t (\pi_1\rho) \,a \,(s' \circ g) \,xs, t (\pi_2\rho) \,b \,s' \,(\mathit{map%}_{\mathit{List}} \,g \,xs)) \in \graph{\mathit{map}_{\mathit{List}} \,g}, \\ 
-%    & i.e., \\ 
-%    &\mathit{map}_{\mathit{List}} \,g \, (t (\pi_1\rho) \,a \,(s' \circ g) \,xs) =% t (\pi_2\rho) \,b \,s' \,(\mathit{map}_{\mathit{List}} \,g \,xs), \\ 
-%    & i.e., \\
-%    &\mathit{map}_{\mathit{List}} \,g \circ t (\pi_1 \rho) \, a \, (s' \circ g) = %t (\pi_2\rho) \, b \, s' \circ \mathit{map}_{\mathit{List}} \,g
-%  \end{align*}
-%  as desired.
-%\end{proof}
-
-\begin{comment}
-\begin{thm}\label{thm:at-gen-rel}
-If \,$\Gamma; \Phi \,|\, \Delta \vdash t : \tau$ and \,$\rho \in \relenv$,
-  and if \,$(a, b) \in \relsem{\Gamma; \Phi \vdash \Delta} \rho$,
-  then \\
-  $(\setsem{\Gamma; \Phi \,|\, \Delta \vdash t : \tau} (\pi_1 \rho) \,a \, ,
-      \setsem{\Gamma; \Phi \,|\, \Delta \vdash t : \tau } (\pi_2 \rho) \,b) \in 
-    \relsem{\Gamma; \Phi \vdash \tau} \rho$
-\end{thm}
-\proof
- Immediate from Theorem~\ref{thm:at-gen}.
-\qed
-\end{comment}
-
-\begin{comment}
-\begin{thm} 
-  If $g : A \to B$, 
-  $\rho : \relenv$, 
-  $\rho \alpha = (A, B, \graph{g})$,
-  $(a, b) \in \relsem{\alpha ;\emptyset \vdash \Delta} \rho$, 
-  $(s \circ g, s) \in \relsem{\alpha; \emptyset \vdash \Nat^\emptyset \alpha \, \mathit{Bool}} \rho$,
-  and, for some well-formed term $\mathit{filter}$, \\
-  $$t = \setsem{\alpha; \emptyset \,|\, \Delta \vdash \mathit{filter} : \filtype} \emph{, then }$$
-  \begin{align*}
-  \mathit{map}_{\mathit{List}} \,g \circ t (\pi_1 \rho) \, a \, (s \circ g) = t (\pi_2\rho) \, b \, s \circ \mathit{map}_{\mathit{List}} \,g
-  \end{align*}
-\end{thm}
-\proof
-  By Theorem~\ref{thm:at-gen-rel},
-  $(t (\pi_1 \rho) a, t (\pi_2 \rho) b) \in \relsem{\alpha; \emptyset \vdash \filtype} \rho$.
-  Thus if 
-  $(s, s') \in \relsem{\alpha; \emptyset \vdash \Nat^\emptyset \alpha \, \mathit{Bool}} \rho = \rho\alpha \to \Eq_{\mathit{Bool}}$,
-  then 
-  \begin{align*}
-    (t (\pi_1 \rho) \,a \,s, t (\pi_2 \rho) \,b \,s') &\in \relsem{\alpha; \emptyset \vdash \Nat^\emptyset (List \, \alpha) \, (List \, \alpha)} \rho \\
-   &= \relsem{\alpha; \emptyset \vdash List \, \alpha} \rho \to \relsem{\alpha; \emptyset \vdash List \, \alpha} \rho \\
-  \end{align*}
-  So if $(xs, xs') \in \relsem{\alpha; \emptyset \vdash List \, \alpha} \rho$ then,
-  \begin{equation}\label{eq:filter-thm}
-  (t (\pi_1\rho) \,a \,s \,xs, t (\pi_2\rho) \,b \,s' \,xs') \in \relsem{\alpha; \emptyset \vdash List \, \alpha} \rho
-  \end{equation}
-
-  Consider the case in which $\rho\alpha = (A, B, \graph{g})$. Then $\relsem{\alpha; \emptyset \vdash List \, \alpha} \rho
-  = \graph{\mathit{map}_{\mathit{List}} \, g}$, by Lemma~\ref{lem:list-graph}, and $(xs, xs') \in \graph{\mathit{map}_{\mathit{List}} \,g}$ 
-  implies $xs' = \mathit{map}_{\mathit{List}} \,g \,xs$. We also have that 
-  $(s, s') \in \graph{g} \to \Eq_{\mathit{Bool}}$ implies 
-  $\forall (x, g x) \in \graph{g}. \,\, s x = s' (g x)$ and thus
-  $s = s' \circ g$ due to the definition of morphisms between relations.
-  %% TODO show (s, s') \in <g> -> \mathit{Bool} implies s = s' \circ g %% 
-  With these instantiations, Equation~\ref{eq:filter-thm} becomes
-  \begin{align*}
-    &(t (\pi_1\rho) \,a \,(s' \circ g) \,xs, t (\pi_2\rho) \,b \,s' \,(\mathit{map}_{\mathit{List}} \,g \,xs)) \in \graph{\mathit{map}_{\mathit{List}} \,g}, \\ 
-    & i.e., \\ 
-    &\mathit{map}_{\mathit{List}} \,g \, (t (\pi_1\rho) \,a \,(s' \circ g) \,xs) = t (\pi_2\rho) \,b \,s' \,(\mathit{map}_{\mathit{List}} \,g \,xs), \\ 
-    & i.e., \\
-    &\mathit{map}_{\mathit{List}} \,g \circ t (\pi_1 \rho) \, a \, (s' \circ g) = t (\pi_2\rho) \, b \, s' \circ \mathit{map}_{\mathit{List}} \,g
-  \end{align*}
-  as desired.
-\qed
-\end{comment}
-
-%\subsection{Free Theorem for Type of $\mathit{filter}$ for
-%  $\GRose$}\label{sec:filter-grose}
+follows from 
+Equation~\ref{eq:filter-thm-list}.\qed
 
@@
 $F, G : \set \to \set$,
-  $\eta : F \to G$ in $\set$, $\rho : \relenv$, $\rho \alpha =
+  $\eta : F \to G$ in $\set$, $\rho \in \relenv$, $\rho \alpha =
  (A, B, \graph{g})$, $\rho \psi = (F, G, \graph{\eta})$, $(a, b) \in
@@
 If \,$\vdash F$, $\vdash H$, and 
-%$\beta; \emptyset \,|\, \emptyset \vdash g : \Nat^{\emptyset}
-%(\Nat^{\emptyset} (\onet + \tau \times \beta)\, \beta)\, \beta$. 
-%If
 $G \, = \, \setsem{\beta; \emptyset \,|\, \emptyset \vdash g :
   \Nat^{\emptyset} (\Nat^{\emptyset} (\onet + F \times \beta)\,
-  \beta)\, \beta}$ for some $g$, then
+  \beta)\, \beta}$ for some $g$, and if $c \in \setsem{\vdash F}
+\,\times\, \setsem{\vdash H} \to \setsem{\vdash H}$ and $n \in
+\setsem{\vdash H}$, then
 \[\mathit{fold}_{1 + \setsem{\vdash F} \times \_}\, n\, c\; (G\; (\mathit{List}\,
@@
 \proof
-%Let $\vdash \tau : \F$ and $\vdash \tau' : \F$, let
-%\[\beta; \emptyset \,|\, \emptyset \vdash g : \Nat^{\emptyset}
-%(\Nat^{\emptyset} (\onet + \tau \times \beta)\, \beta)\, \beta\]
-%and let
-%\[G = \setsem{\beta; \emptyset \,|\, \emptyset
-%  \vdash g : \Nat^{\emptyset} (\Nat^{\emptyset} (\onet + \tau \times
-  %  \beta)\, \beta)\, \beta}\] Then
 Theorem~\ref{thm:abstraction} gives
-that, for any $\rho : \relenv$,
-%and any $(a, b) \in
-%\relsem{\beta; \emptyset \vdash \emptyset}\rho = 1$, then (eliding the
-%only possible instantiations of $a$ and $b$) we have
-\[(G \,(\pi_1 \rho), G\, (\pi_2 \rho))
- \in  \relsem{\beta; \emptyset \vdash \Nat^{\emptyset} (\Nat^{\emptyset}
-  (\onet + F \times \beta)\, \beta)\, \beta} \rho\\
-\cong (((\relsem{\vdash F} \rho \times
-\rho\beta) \to \rho\beta) \times \rho\beta) \to \rho\beta\] 
-%\[\begin{array}{lll}
-%(G \,(\pi_1 \rho), G\, (\pi_2 \rho))
-% & \in & \relsem{\beta; \emptyset \vdash \Nat^{\emptyset} (\Nat^{\emptyset}
-%  (\onet + \tau \times \beta)\, \beta)\, \beta} \rho\\
-%& \cong& (((\relsem{\vdash \tau} \rho \times
-%\rho\beta) \to \rho\beta) \times \rho\beta) \to \rho\beta 
-%\end{array}\]
-%
-%Since
-%\[\begin{array}{rl}
-% & \relsem{\beta; \emptyset \vdash \Nat^{\emptyset}
-%  (\Nat^{\emptyset} (\onet + \tau \times \beta)\, \beta)\, \beta} \rho\\  
-%=& \relsem{\beta; \emptyset \vdash \Nat^{\emptyset} (\onet +
-%  \tau \times \beta)\, \beta} \rho \to \rho\beta \\ 
-%=& (\relsem{\beta; \emptyset \vdash \onet + \tau \times \beta}
-%\rho \to \rho\beta) \to \rho\beta \\ 
-%=& ((\onet + \relsem{\vdash \tau} \rho \times
-%\rho\beta) \to \rho\beta) \to \rho\beta \\ 
-%\cong& (((\relsem{\vdash \tau} \rho \times
-%\rho\beta) \to \rho\beta) \times \rho\beta) \to \rho\beta 
-%\end{array}\]
-%we have that
+that, for any $\rho \in \relenv$,
+\[\begin{array}{lll}
+(G \,(\pi_1 \rho), \,G\, (\pi_2 \rho)) & \in 
+& \relsem{\beta; \emptyset \vdash \Nat^{\emptyset} (\Nat^{\emptyset}
+  (\onet + F \times \beta)\, \beta)\, \beta} \rho \\
+& \cong & (((\relsem{\vdash F} \rho \times
+\rho\beta) \to \rho\beta) \times \rho\beta) \to \rho\beta
+\end{array}\]
 so if $(c', c) \in \relsem{\vdash F} \rho \times
@@
 In addition,
-\[\setsem{\vdash \fold_{\onet + F
+\begin{align*}
+  &\setsem{\vdash \fold_{\onet + F
     \times \beta}^{H} : \Nat^{\emptyset} (\Nat^{\emptyset} (\onet
   + F \times H)\, H)\, (\Nat^{\emptyset} (\mu \alpha. \onet
-  + F \times \alpha)\, H)} =
-\mathit{fold}_{1 + \setsem{\vdash F} \times \_}\] so that if $c \in
+  + F \times \alpha)\, H)} \\ =\;
+& \mathit{fold}_{1 + \setsem{\vdash F} \times \_}
+\end{align*}
+so that if $c \in
 \setsem{\vdash F} \,\times\, \setsem{\vdash H} \to
-\setsem{\vdash H}$ and $n \in \setsem{\vdash H}$, then
-$(n, c) \in \setsem{\vdash \Nat^{\emptyset} (\onet + F \times H)\,
-  H}$. The instantiation
-$\pi_1 \rho \beta = \setsem{\vdash \mu \alpha. \onet + F \times
-  \alpha}=\mathit{List}\,\setsem{\vdash F}$,\, 
-$\pi_2 \rho \beta = \setsem{\vdash H}$,\,
-$\rho \beta = \graph{\mathit{fold}_{1 + \setsem{\vdash F} \times \_}\, n\, c} :
-\rel(\pi_1 \rho \beta, \pi_2 \rho \beta)$,\,
-$c' = \mathit{cons}$,\, and 
-$n' = \mathit{nil}$
-%Consider the instantiation:
-%\[\begin{array}{rcl}
-%\pi_1 \rho \beta &=& \setsem{\vdash \mu \alpha. \onet + \tau \times
-%  \alpha} \;=\;\List\,\setsem{\vdash \tau}\\   
-%\pi_2 \rho \beta &=& \setsem{\vdash \tau'}\\
-%\rho \beta &=& \graph{\mathit{fold}_{1 + \tau \times \_}\, n\, c} :
-%\rel(\pi_1 \rho \beta, \pi_2 \rho \beta) \\ 
-%c' &=& \mathit{cons}\\
-%n' &=& \mathit{nil}
-%\end{array}\]
-%Clearly, $(\mathit{nil}, n) \in \rho\beta =
-%\graph{\mathit{fold}_{1 + \tau \times \_}\, n\, c}$ because $
-%\mathit{fold}_{1 + \tau \times \_}\, n\, c \,\mathit{nil} = n$.  Moreover,
-%$(\mathit{cons}, c) \in \relsem{\vdash \tau} \,\times\, \rho\beta \to
-%\rho\beta$ since if $(x, x') \in \relsem{\vdash \tau}$, i.e., $x =
-%x'$, and if $(y, y') \in \rho\beta =
-%\graph{\mathit{fold}_{1 + \tau \times \_} n\, c}$, i.e., $y' =
-%\mathit{fold}_{1 + \tau \times \_} n\, c\, y$, then
-%\[(\mathit{cons}\, x\, y, c\, x\, (\mathit{fold}_{1 + \tau \times \_}\, n\,
-%c\, y)) \in \graph{\mathit{fold}_{1 + \tau \times \_}\, n\, c}\]
-%i.e.,
-%\[c\, x\, (\mathit{fold}_{1 + \tau \times \_}\, n\, c\, y)
-%= \mathit{fold}_{1 + \tau \times \_}\, n\, c\, (\mathit{cons} \,x\, y)\]
-%holds by definition of $\mathit{fold}_{1 + \tau \times \_}$.  We
-%therefore conclude that
+\setsem{\vdash H}$ and $n \in \setsem{\vdash H}$,
+ then
+\[(n, c) \in \setsem{\vdash \Nat^{\emptyset} (\onet + F \times H)\,
+  H}\] The instantiation
+\[\begin{array}{lll}
+\pi_1 \rho \beta & = & \setsem{\vdash \mu \alpha. \onet + F \times
+  \alpha}=\mathit{List}\,\setsem{\vdash F}\\
+\pi_2 \rho \beta & = & \setsem{\vdash H}\\
+\rho \beta & = & \graph{\mathit{fold}_{1 + \setsem{\vdash F} \times \_}\, n\, c} :
+\rel(\pi_1 \rho \beta, \pi_2 \rho \beta)\\
+c' & = & \mathit{cons}\\
+n' & = & \mathit{nil}
+\end{array}\]
 thus gives that $(G \;(\mathit{List}\,\setsem{\vdash
@@
 
-\begin{comment}
-\subsection{Short Cut Fusion for Arbitrary ADTs}\label{sec:short-cut-adt}
-
-\begin{thm}
-Let $\ol{\vdash \tau}$, let $\vdash \tau'$, let $\ol{\alpha}; \beta
-\vdash F$, and let $\beta; \emptyset  \,|\,
-\emptyset \vdash g : \Nat^{\emptyset} (\Nat^{\emptyset} F[\ol{\alpha
-    := \tau}]\, \beta)\, \beta$.  If we regard
-\[\begin{array}{lll}
-H & = & \setsem{\emptyset;\beta \vdash F[\ol{\alpha :=
-      \tau}]}\\
-G & = & \setsem{\beta; \emptyset \,|\, \emptyset \vdash g :
-  \Nat^{\emptyset} (\Nat^{\emptyset}\, F[\ol{\alpha := \tau}]\,
-  \beta)\, \beta}
-\end{array}\]
-as functors in $\beta$, then for every $B \in H \setsem{\vdash \tau'}
-\rightarrow \setsem{\vdash \tau'}$ we have
-\[\mathit{fold}_H\, B \; (G\; \mu H \; in_H) = G \,\setsem{\vdash \tau'}\, B\]
-\end{thm}
-
-\vspace*{0.2in}
-
-\proof
-  We first note that the type of $g$ is well-formed, since
-  $\emptyset;\beta \vdash F[\ol{\alpha := \tau}]$ so our
-  promotion theorem gives that $\beta;\emptyset\vdash F[\ol{\alpha :=
-      \tau}]$, and $\emptyset;\beta\vdash\beta$ so that our
-  promotion theorem gives $\beta;\emptyset\vdash\beta$. From
-  these facts we deduce that $\beta;\emptyset \vdash
-  \Nat^\emptyset\,F[\ol{\alpha := \tau}]\,\beta : \T$, and thus that
-  $\beta; \emptyset \vdash \Nat^{\emptyset} (\Nat^{\emptyset}
-  F[\ol{\alpha := \tau}]\, \beta)\, \beta : \T$.
-
-  Theorem~\ref{thm:abstraction} gives that, for any $\rho : \relenv$
-  and any $(a, b) \in \relsem{\beta; \emptyset \vdash \emptyset}\rho =
-  1$, eliding the only possible instantiations of $a$ and $b$ gives
-  that
-\[(G \,(\pi_1 \rho), G\, (\pi_2 \rho))
-\in \relsem{\beta; \emptyset \vdash \Nat^{\emptyset} (\Nat^{\emptyset}
-  F[\ol{\alpha := \tau}]\, \beta)\, \beta} \rho\]
-Since
-\[\begin{array}{rl}
- & \relsem{\beta; \emptyset \vdash \Nat^{\emptyset}
-  (\Nat^{\emptyset} F[\ol{\alpha := \tau}]\, \beta)\, \beta} \rho\\  
-=& \relsem{\beta; \emptyset \vdash \Nat^{\emptyset} F[\ol{\alpha :=
-      \tau}] \, \beta} \rho \to \rho\beta \\ 
-\end{array}\]
-we have that if $(A, B) \in \relsem{\beta; \emptyset \vdash
-  \Nat^{\emptyset} F[\ol{\alpha := \tau}] \, \beta} \rho$
-then
-\[(G \,(\pi_1 \rho)\, A, G\, (\pi_2 \rho)\, B)
-\in \rho \beta\]
-Now note that
-\[\setsem{\vdash \fold_{F[\ol{\alpha := \tau}]}^{\tau'} :
-  \Nat^{\emptyset}\, (\Nat^{\emptyset}\, F[\ol{\alpha := \tau}][\beta
-    := \tau'] \, \tau')\, (\Nat^{\emptyset} \,(\mu
-  \beta. F[\ol{\alpha := \tau}] \, \tau')} =
-\mathit{fold}_H\]
-and consider the instantiation
-\[\begin{array}{lll}
-A &=& \mathit{in}_H :  H (\mu H) \to \mu H\\
-B & : & H\setsem{\vdash \tau'} \to \setsem{\vdash \tau'}\\
-\rho \beta &=& \graph{\mathit{fold}_H\, B}
-\end{array}\]
-(Note that all the types here are well-formed.)  This gives
-\[\begin{array}{lll}
-\pi_1 \rho \beta &=& \setsem{\vdash \mu \beta. F[\ol{\alpha := \tau}]}
-\; = \; \mu H\\
-\pi_2 \rho \beta &=& \setsem{\vdash \tau'}\\
-\rho \beta &:& \rel(\pi_1 \rho \beta, \pi_2 \rho \beta)\\
-A & : & \setsem{\beta; \emptyset \vdash \Nat^{\emptyset}
-  F[\ol{\alpha := \tau}] \, \beta} (\pi_1 \rho)\\
-B & : & \setsem{\beta; \emptyset \vdash \Nat^{\emptyset} F[\ol{\alpha
-      := \tau}] \, \beta}(\pi_2 \rho)\\
-\end{array}\]
-since
-\[\begin{array}{lll}
-A \; = \; \mathit{in}_H & : & H(\mu H) \to \mu H\\
- & = & \setsem{\emptyset;\beta \vdash F[\ol{\alpha := \tau}]}(\mu
-\setsem{\emptyset;\beta \vdash F[\ol{\alpha := \tau}]}) \to \mu
-\setsem{\emptyset;\beta \vdash F[\ol{\alpha := \tau}]}\\
-& = & \setsem{\emptyset;\beta \vdash F[\ol{\alpha :=
-      \tau}]}(\pi_1\rho) \to \setsem{\emptyset;\beta \vdash
-  \beta}(\pi_1 \rho)\\
-& = & \setsem{\beta;\emptyset \vdash F[\ol{\alpha :=
-      \tau}]}(\pi_1\rho) \to \setsem{\beta; \emptyset \vdash
-  \beta}(\pi_1 \rho) \hspace*{0.25in} {\color{red} \mbox{Daniel's
-    trick; now a theorem}} \\ 
-& = & \setsem{\beta;\emptyset \vdash \Nat^\emptyset F[\ol{\alpha :=
-      \tau}]\,\beta}(\pi_1 \rho) 
-\end{array}\]
-{\color{red} where ``Daniel's trick'' is the observation that a
-  functor can be seen as non-functorial when we only care about its
-  action on objects. This is now a theorem.}  We also have
-\[\begin{array}{lll}
-(A,B) \,=\, (\mathit{in}_H,B) & \in & \relsem{\beta; \emptyset \vdash
-  \Nat^{\emptyset} F[\ol{\alpha := \tau}] \, \beta} \rho \\
- & = & \relsem{\beta; \emptyset \vdash F[\ol{\alpha :=
-      \tau}]}\rho[\beta := \graph{\mathit{fold}_H\, B}] \to
-\graph{\mathit{fold}_H\, B}\\
- & = & \relsem{\beta; \emptyset \vdash F[\ol{\alpha :=
-      \tau}]}\graph{\mathit{fold}_H\, B} \to
-\graph{\mathit{fold}_H\, B}\\
- & = & \relsem{\emptyset; \beta\vdash F[\ol{\alpha :=
-      \tau}]}\graph{\mathit{fold}_H\, B} \to
-\graph{\mathit{fold}_H\, B} \hspace*{0.35in} {\color{red}
-  \mbox{Daniel's trick; now a theorem}}\\  
-& = & \graph{\setsem{\emptyset; \beta \vdash F[\ol{\alpha :=
-      \tau}]}\,(\mathit{fold}_H\, B)} \to
-\graph{\mathit{fold}_H\, B} \hspace*{0.25in} {\color{red} \mbox{by the
-    graph lemma}}\\ 
-& = & \graph{\mathit{map}_H \,(\mathit{fold}_H\,B)} \to
-\graph{\mathit{fold}_H\, B}\\
-\end{array}\]
-since if $(x,y) \in \graph{\mathit{map}_H \,(\mathit{fold}_H\,B)}$,
-i.e., if $\mathit{map}_H \,(\mathit{fold}_H\,B) \, x = y$, then
-$\mathit{fold}_H\, B\, (\mathit{in}_H\,x) = B\,y = B\, (\mathit{map}_H
-\,(\mathit{fold}_H\,B) \, x)$ by the definition of $\mathit{fold}_H$
-as a (indeed, the unique) morphism from $\mathit{in}_H$ to $B$. Thus,
-\[(G \,(\pi_1 \rho)\, A, G\, (\pi_2 \rho)\, B) \in
-\graph{\mathit{fold}_H\, B}\]
-i.e.,
-\[\mathit{fold}_H \, B \, (G\, (\pi_1 \rho) \, \mathit{in}_H) =
-G\,(\pi_2 \rho)\,B\]
-
-\vspace*{0.1in}
-
-\noindent
-Since $\beta$ is the only free variable in $G$, this simplifies to
-\[\mathit{fold}_H\, B \, (G\, \mu H\, \mathit{in}_H) =
-G\,\setsem{\vdash \tau'}\,B\]  
-\qed
-\end{comment}
-
 \subsection{Short Cut Fusion for Arbitrary Nested
@@
 
-%fold/build rule shows that Church encodings for nested types are iso
-%(have the same interp) as the nested types themselves. Just like Ex
-%2.8 in parpoe, but categorical rather than operational semantics. But
-%built-in data types are more versatile in implementations. They can be
-%inducted on, e.g. And they are stored in the heap, not the run-time
-%stack, and are therefore more efficient.
-%``Most higher order type languages meant for human programmers eschew
-%fully impredicative polymorphism'' says Andy on p322.
-
-We can also, finally, formally prove correctness of the categorically
-inspired short cut fusion for nested types from~\cite{jg10}. Indeed,
-we have:
+We can extend short cut fusion for lists~\cite{glp93} to nested types,
+thereby formally prove correctness of the categorically inspired
+theorem from~\cite{jg10}.  We have:
 \begin{thm}\label{thm:short-cut-nested}
@@
 \vdash K$, \,
-%let $\phi ;\emptyset\,|\,\emptyset\vdash g :
-%\Nat^\emptyset\,(\Nat^\alpha\,F\,(\phi\alpha)) \,
-%(\Nat^\alpha\,\onet \, (\phi\alpha))$. If we let
 $H : [\set,\set] \to [\set,\set]$ is defined by
@@
   \alpha \vdash K}$,
-$\mathit{fold}_{H}\, B \, (G\; \mu H \; \mathit{in}_{H}) = G
-\,\setsem{\emptyset;\alpha \vdash K}\, B$.
+$$\mathit{fold}_{H}\, B \, (G\; \mu H \; \mathit{in}_{H}) = G
+\,\setsem{\emptyset;\alpha \vdash K}\, B$$
 \end{thm}
@@
 \proof
-%We first note that the type of $g$ is well-formed since
-%$\emptyset;\phi,\alpha \vdash F : \F$ so our promotion theorem gives
-%that $\phi;\alpha \vdash F : \F$, and $\phi;\alpha \vdash \phi\alpha :
-%\F$, so that $\phi;\emptyset \vdash \Nat^\alpha F\,(\phi\alpha) : \T$
-%and $\phi;\emptyset \vdash \Nat^\alpha \onet \,(\phi\alpha) : \T$.
-%Then $\phi;\emptyset \vdash \Nat^\alpha F\,(\phi\alpha) : \F$ and
-%$\phi;\emptyset \vdash \Nat^\alpha \onet \,(\phi\alpha) : \F$ also
-%%%hold, and, finally, 
-%$\phi ;\emptyset\vdash 
-%\Nat^\emptyset\,(\Nat^\alpha\,F\,(\phi\alpha)) \,
-%(\Nat^\alpha\,\onet \, (\phi\alpha)) : \T$
-%  
 Theorem~\ref{thm:abstraction} gives that, for any 
-$\rho : \relenv$,
-%and any $(a, b) \in \relsem{\phi,\alpha;\emptyset \vdash
-%  \emptyset}\rho = 1$, eliding the only possible instantiations
-%of $a$ and $b$ gives that
+$\rho \in \relenv$,
 \[\begin{array}{lll}
-(G \,(\pi_1 \rho), G\, (\pi_2 \rho))
-& \in & \relsem{\phi;\emptyset\vdash \Nat^{\emptyset} (\Nat^\alpha
+(G \,(\pi_1 \rho), G\, (\pi_2 \rho)) & \in &
+\relsem{\phi;\emptyset\vdash \Nat^{\emptyset} (\Nat^\alpha 
   F\, (\phi\alpha))\, (\Nat^\alpha\,\onet \, (\phi\alpha))}
@@
   \Nat^\alpha\,\onet \, (\phi\alpha)}\rho\\ 
-%& = & \relsem{\phi;\emptyset\vdash \Nat^\alpha F\,
-%  (\phi\alpha)}\rho \to (\,\lambda A. 1 \Rightarrow \lambda A.\,(\rho
-%\phi) A\,)\\   
-%& = & \relsem{\phi;\emptyset\vdash \Nat^\alpha F\,
-%  (\phi\alpha)}\rho \to (1 \Rightarrow \rho \phi)\\ 
 & = & \relsem{\phi;\emptyset\vdash \Nat^\alpha F\,
@@
 B) \in \rho \phi$.
-In addition,
-%$\mathit{fold}_H = \setsem{\vdash \fold_F^K :
-%  \Nat^{\emptyset}\, (\Nat^\alpha F[\phi := K]\,K)\, (\Nat^\alpha
-%  ((\mu \phi.\lambda\alpha.F)\alpha)\,K)}$.
+Also,
 \[\setsem{\vdash \fold_F^K :
@@
   ((\mu \phi.\lambda\alpha.F)\alpha)\,K)} = \mathit{fold}_H\]
-Consider the instantiation $A = \mathit{in}_H : H (\mu H) \Rightarrow
+Now let $A = \mathit{in}_H : H (\mu H) \Rightarrow
 \mu H$,\, $B : H\setsem{\emptyset;\alpha\vdash K} \Rightarrow
@@
   \alpha)} (\pi_2 \rho)$.
-%\[\begin{array}{lll}
-%A &=& \mathit{in}_H :  H (\mu H) \Rightarrow \mu H\\
-%B & : & H\setsem{\emptyset;\alpha\vdash K} \Rightarrow
-%\setsem{\emptyset;\alpha \vdash K}\\
-%\rho \phi &=& \graph{\mathit{fold}_H\, B} \hspace*{0.2in} {
-%  \color{red} \mbox{a graph of a natural transformation, defined in
-%    Enrico's notes}} 
-%\end{array}\]
-%(Note that all the types here are well-formed.) This gives
-%\[\begin{array}{lll}
-%\pi_1 \rho \phi &=& \mu H\\
-%\pi_2 \rho \phi &=& \setsem{\emptyset;\alpha\vdash K}\\
-%\rho \phi &:& \rel(\pi_1 \rho \phi, \pi_2 \rho \phi)\\
-%A & : & \setsem{\phi; \emptyset \vdash \Nat^\alpha
-%  F \, (\phi \alpha)} (\pi_1 \rho)\\
-%B & : & \setsem{\phi; \emptyset \vdash \Nat^\alpha
-%  F \, (\phi \alpha)} (\pi_2 \rho)\\
-%\end{array}\]
-%since
-Equation~\ref{thm:demotion-objects} ensures that $A = \mathit{in}_H :
-H(\mu H) \Rightarrow \mu H = \setsem{\phi;\emptyset \vdash \Nat^\alpha
-  F \,(\phi\alpha)}(\pi_1\rho)$,
-%\[\begin{array}{lll}
-%A \; = \; \mathit{in}_H & : & H(\mu H) \Rightarrow \mu H\\
-%& = & \setsem{\emptyset; \phi, \alpha \vdash F}[\phi := \mu
-%  \setsem{\emptyset; \phi, \alpha \vdash F}] \Rightarrow \mu
-%  \setsem{\emptyset; \phi, \alpha \vdash F}\\  
-%& = & \setsem{\emptyset; \phi, \alpha \vdash F}(\pi_1\rho)
-%  \Rightarrow \setsem{\emptyset; \phi,\alpha \vdash \phi\alpha}(\pi_1 \rho)\\
-%& = & \setsem{\phi; \alpha \vdash F}(\pi_1\rho)
-%  \Rightarrow \setsem{\phi;\alpha \vdash \phi\alpha}(\pi_1
-%  \rho) \hspace*{0.25in} {\color{red} \mbox{Daniel's trick; now a theorem}}\\
-%& = & \setsem{\phi;\emptyset \vdash \Nat^\alpha F
-%    \,(\phi\alpha)}(\pi_1\rho) 
-%\end{array}\]
-and Equation~\ref{thm:demotion-objects} and Lemma~\ref{lem:graph}
-together give that
+Demotion ensures that $A = \mathit{in}_H : H(\mu H) \Rightarrow \mu H
+= \setsem{\phi;\emptyset \vdash \Nat^\alpha F
+  \,(\phi\alpha)}(\pi_1\rho)$,
+and demotion and Lemma~\ref{lem:graph} together give that
 \[\begin{array}{lll}
@@
   \Nat^\alpha F\, (\phi\alpha)}\rho\\
-% & = & \lambda A. \relsem{\phi;\alpha\vdash F}\rho[\alpha := A]
-%\Rightarrow \lambda A. (\rho\phi) A\\
 & = & \lambda A. \relsem{\phi;\alpha\vdash F}[\phi :=
@@
  \graph{\mathit{fold}_H\, B} \\ 
-%& = & \lambda A. \relsem{\emptyset;\phi,\alpha\vdash F}[\phi :=
-%  \graph{\mathit{fold}_H\, B}][\alpha := A] \Rightarrow 
-%\graph{\mathit{fold}_H\, B}  \hspace*{0.25in} {\color{red}
-%  \mbox{Daniel's trick; now a theorem}}\\  
 & = & \relsem{\emptyset;\phi,\alpha\vdash F}
@@
     \,(\mathit{fold}_H\,B)} \Rightarrow \graph{\mathit{fold}_H\, B}\\
-%\hspace*{0.25in} {\color{red} \mbox{Graph Lemma}}\\
   & = & \graph{\mathit{map}_H \,(\mathit{fold}_H\,B)} \Rightarrow
@@
 since if $(x,y) \in \graph{\mathit{map}_H \,(\mathit{fold}_H\,B)}$,
-%i.e., if $\mathit{map}_H \,(\mathit{fold}_H\,B) \, x = y$,
-then $\mathit{fold}_H\, B\, (\mathit{in}_H\,x) = B\,y = B\,
-(\mathit{map}_H \,(\mathit{fold}_H\,B) \, x)$ by the definition of
+then $$\mathit{fold}_H\, B\, (\mathit{in}_H\,x) = B\,y = B\,
+(\mathit{map}_H \,(\mathit{fold}_H\,B) \, x)$$ by the definition of
 $\mathit{fold}_H$ as a (indeed, the unique) morphism from
@@
 $\mathit{fold}_H\, B \, (G\, \mu H\, \mathit{in}_H) =
-G\,\setsem{\emptyset;\alpha\vdash K}\,B$. 
-\qed
+G\,\setsem{\emptyset;\alpha\vdash K}\,B$. \qed
+
+\vspace*{0.05in}
 
 As in~\cite{jg10}, replacing $\onet$ with any type $\emptyset;\alpha
-\vdash C$ generalizes
-%with $\setsem{\emptyset;\alpha \vdash c}\rho = C$
-Theorem~\ref{thm:short-cut-nested} to deliver more general a free
-theorem whose conclusion is $\mathit{fold}_{H}\, B \; \circ \; G\; \mu
-H \; \mathit{in}_{H} = G \,\setsem{\emptyset;\alpha \vdash K}\, B$.
+\vdash C$ generalizes Theorem~\ref{thm:short-cut-nested} to deliver a
+more general a free theorem whose conclusion is $\mathit{fold}_{H}\, B
+\; \circ \; G\; \mu H \; \mathit{in}_{H} = G
+\,\setsem{\emptyset;\alpha \vdash K}\, B$.
+
+\vspace*{0.05in}
+
 Although it is standard to prove that the parametric model constructed
@@
 
-{\color{red} CHANGE? COMPARE AGAINST INTRO.}
+We have constructed a parametric model for a calculus supporting
+primitive nested types, and used its Abstraction Theorem to derive
+free theorems for nested types.  This was not possible
+before~\cite{jp19} because these types were not previously known to
+have well-defined interpretations in locally finitely presentable
+categories (here, $\set$ and $\rel$), and, to our knowledge, no term
+calculus for them existed.  We naturally expect the construction
+elaborated here to guide its generalization to more advanced data
+types.  For example, GADTs can be represented using left Kan
+extensions, and it was shown in~\cite{jp19} that adding a
+$\mathsf{Lan}$ construct to a calculus such as ours preserves the
+$\lambda$-cocontinuity needed for the data types it defines to have
+well-defined interpretations in locally $\lambda$-presentable
+categories. (Interestingly, $\lambda > \aleph_1$ is required to
+interpret even common GADTs.) This suggests carrying out our model
+construction in locally $\lambda$-presentable cartesian closed
+categories (lpcccs) $\cal C$ whose categories of (abstract) relations,
+obtained by pullback as in~\cite{jac99}, are also lpcccs and are
+appropriately fibred over $\cal C$.  Adding term-level fixpoints
+further requires our semantic categories not just to be locally
+$\lambda$-presentable, but also to support some kind of domain
+structure.
+
+{\color{red} KEEP ANYTHING FROM THIS? ADD GADTS, AT LEAST TO SUMMARY.
 
@@
 %of domains to replace $\set$, is not locally finitely presentable.
+}
 

### 2020-12-16 free-theorems-nesteds/20-mscs.tex
finish body of paper draft except GADTs

@@
 \end{array}\]
-The representation of $\mathit{List}\,\alpha$ as the ADT $\mu
-\beta. \onet + \alpha \times \beta$ cannot be functorial in
-$\alpha$. By contrast, its representation as the nested type $(\mu
-\phi. \lambda \beta.\,\onet + \beta \times \phi \beta)\,\alpha$ can be
-considered either functorial in $\alpha$ or not, according to whether
-$\alpha$ is in the functorial or non-functorial context in which this
-type is formed. If $\emptyset;\alpha \vdash \mathit{List} \,\alpha$,
-then $\vdash \Nat^\alpha \onet\, (\mathit{List}\,\alpha)$ is
-well-formed; if $\alpha;\emptyset \vdash \List \,\alpha$, then it is
-not.  Similarly, if $\mathit{Tree}\,\alpha\,\gamma = \mu \,\beta.\,
-\alpha + \beta \times \gamma \times \beta$, then $\gamma;\emptyset
-\vdash \Nat^\alpha (\mathit{List}\, \alpha) \, (\mathit{Tree} \,
-\alpha\, \gamma)$ is well-formed, but $\emptyset;\gamma \vdash
-\Nat^\alpha (\mathit{List}\, \alpha) \, (\mathit{Tree} \, \alpha\,
-\gamma)$ is not. The body $F$ of a type $(\mu \phi.\lambda
-\ol\alpha.F)\ol G$ can only be functorial in $\phi$ and the variables
-in $\ol\alpha$. As a result, both $\phi$ and the variables in
-$\ol\alpha$ must be non-functorial in $\mathit{GRose}\, \phi \, \alpha
-= \mu\beta . \alpha \times \phi \beta$, $\mathit{GRose}\,\phi\,\alpha$
-cannot be the (co)domain of a $\Nat$-type binding any variables, and
-no type of the form $(\mu \phi. \lambda
-\alpha. \mathit{GRose}\,\phi\,\alpha)\tau$ is well-formed in our
-system. These restrictions ensure that the set and relational
-interpretations of types in Section~\ref{sec:type-interp} are
-functors. In Section~\ref{sec:conclusion} we discuss how more
-expressive types might be accommodated.
+
+Note that since the body F of a type $(\mu \phi. \lambda
+\ol\alpha. F)\ol{G}$ can only be functorial in $\phi$ and the
+variables in $\ol{\alpha}$, the representation of
+$\mathit{List}\,\alpha$ as the ADT $\mu \beta. \,\onet + \alpha \times
+\beta$ cannot be functorial in $\alpha$. By contrast, if
+$\mathit{List}\,\alpha$ is represented as the nested type $(\mu
+\phi. \lambda \beta.\,\onet + \beta \times \phi \beta)\,\alpha$ then
+we can choose $\alpha$ to be a functorial variable or not when forming
+the type. This observation holds for other ADTs as well; for example,
+if $\mathit{Tree}\,\alpha\,\gamma = \mu \beta. \alpha + \beta \times
+\gamma \times \beta$, then $\alpha, \gamma; \emptyset \vdash
+\mathit{Tree}\,\alpha\,\gamma$ is well-formed, but $\emptyset; \alpha,
+\gamma \vdash \mathit{Tree}\,\alpha\,\gamma$ is not. And it also
+applies to some non-ADT types, such as $\mathit{GRose}\,\phi\,\alpha =
+\mu \beta. \onet + \alpha \times \phi\beta$, in which $\phi$ and
+$\alpha$ must both be non-functorial variables. {\color{red} It is in
+  fact possible to allow ``extra'' $0$-ary functorial variables in the
+  body of $\mu$-types (functorial variables of higher arity are the
+  real problem), which would allow the first-order representations of
+  ADTs to be functorial.  However, it simplifies the formation rule
+  for $\mu$-types considerably to disallow even these $0$-ary
+  functorial variables. Since we can always use an ADT's (semantically
+  equivalent) second-order representation when functoriality is
+  needed, this restriction allows us to simplify the type syntax
+  without sacrificing the expressivity of the calculus.}
+%The representation of $\mathit{List}\,\alpha$ as the ADT $\mu
+%\beta. \onet + \alpha \times \beta$ cannot be functorial in
+%$\alpha$. By contrast, its representation as the nested type $(\mu
+%\phi. \lambda \beta.\,\onet + \beta \times \phi \beta)\,\alpha$ can be
+%considered either functorial in $\alpha$ or not, according to whether
+%$\alpha$ is in the functorial or non-functorial context in which this
+%type is formed. If $\emptyset;\alpha \vdash \mathit{List} \,\alpha$,
+%then $\vdash \Nat^\alpha \onet\, (\mathit{List}\,\alpha)$ is
+%well-formed; if $\alpha;\emptyset \vdash \List \,\alpha$, then it is
+%not.  Similarly, if $\mathit{Tree}\,\alpha\,\gamma = \mu \,\beta.\,
+%\alpha + \beta \times \gamma \times \beta$, then $\gamma;\emptyset
+%\vdash \Nat^\alpha (\mathit{List}\, \alpha) \, (\mathit{Tree} \,
+%\alpha\, \gamma)$ is well-formed, but $\emptyset;\gamma \vdash
+%\Nat^\alpha (\mathit{List}\, \alpha) \, (\mathit{Tree} \, \alpha\,
+%\gamma)$ is not. The body $F$ of a type $(\mu \phi.\lambda
+%\ol\alpha.F)\ol G$ can only be functorial in $\phi$ and the variables
+%in $\ol\alpha$. As a result, both $\phi$ and the variables in
+%$\ol\alpha$ must be non-functorial in $\mathit{GRose}\, \phi \, \alpha
+%= \mu\beta . \alpha \times \phi \beta$, $\mathit{GRose}\,\phi\,\alpha$
+%cannot be the (co)domain of a $\Nat$-type binding any variables, and
+%no type of the form $(\mu \phi. \lambda
+%\alpha. \mathit{GRose}\,\phi\,\alpha)\tau$ is well-formed in our
+%system.
+These restrictions ensure that the set and relational interpretations
+of types in Section~\ref{sec:type-interp} are functors. In
+Section~\ref{sec:conclusion} we discuss how more expressive types
+might be accommodated.
 
@@
     \alpha}.H){\overline \beta}][\ol{\alpha := \beta}] & = &
-\lambda d\,\ol{B}\,
-{\color{red}\ol{C}}.\,(\mathit{in}_{{T}^X_{H,\rho{\color{red}[\ol{\gamma :=
-        C}]}}})_{\ol{B}} \;\;\mbox{ where } X \mbox{ is } \set \mbox{ when } \\ 
+\lambda
+d{\color{red}\,\ol{B}\,\ol{C}.\,(}\mathit{in}_{{T}^X_{H,\rho{\color{red}[\ol{\gamma
+          := 
+        C}]}}}{\color{red})_{\ol{B}}} \;\;\mbox{ where } X \mbox{ is } \set \mbox{ when } \\ 
 \hspace*{0.79in}(\mu \phi.\lambda {\overline \alpha}.H){\overline
@@
     :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F) & = &  
-\lambda d\,\eta\,\ol{B}{\color{red}\,\ol{C}}.\,
-(\mathit{fold}_{T^X_{H,\rho{\color{red}[\ol{\gamma := C}]}}} \, (\lambda
-\ol{A}.\,\eta_{\ol{A}\,\ol{C}}))_{\ol{B}}\\ 
+\lambda d\,\eta{\color{red}\,\ol{B}\,\ol{C}.\,
+(}\mathit{fold}_{T^X_{H,\rho{\color{red}[\ol{\gamma := C}]}}} \, (\lambda
+\ol{A}.\,\eta_{\ol{A}\,\ol{C}}){\color{red})_{\ol{B}}}\\ 
 \hspace*{0.79in}(\Nat^{{\ol{\beta}{\color{red} ,\ol{\gamma}}} }\,(\mu
@@
 
-\section{Free Theorems for Nested Types}\label{sec:free-theorems}
+\section{Naturality and the Abstraction Theorem}\label{sec:free-theorems}
 
@@
 
-If $\Gamma; \ol{\alpha} \vdash F$ then the {\em identity} $\id_F$ on
-$F$ is defined by $\id_F = \Gamma;\emptyset~|~\emptyset \vdash
-L_{\ol{\alpha}}x.x : \Nat^{\ol{\alpha}} F\,F$, and if $\Gamma;
+We first show that terms of $\Nat$-type behave as natural
+transformations with respect to their source and target functorial
+types. If $\Gamma; \ol{\alpha} \vdash F$ then define the {\em
+  identity} $\id_F$ on $F$ by $\id_F = \Gamma;\emptyset~|~\emptyset
+\vdash L_{\ol{\alpha}}x.x : \Nat^{\ol{\alpha}} F\,F$, and if $\Gamma;
 \emptyset \,|\, \Delta \vdash t: \Nat^{\overline{\alpha}} F\,G$ and
 $\Gamma; \emptyset \,|\, \Delta \vdash s: \Nat^{\overline{\alpha}}
-G\,H$ are terms then the {\em composition} $s \circ t$ of $t$ and $s$
-is defined by $s \circ t = \Gamma; \emptyset\,|\, \Delta \vdash
-L_{\overline{\alpha}} x. s_{\overline{\alpha}}(t_{\overline{\alpha}}x):
+G\,H$ are terms then define the {\em composition} $s \circ t$ of $t$
+and $s$ by $s \circ t = \Gamma; \emptyset\,|\, \Delta \vdash
+L_{\overline{\alpha}}
+x. s_{\overline{\alpha}}(t_{\overline{\alpha}}x):
 \Nat^{\overline{\alpha}} F\,H$. Then \[\setsem{\Gamma; \emptyset \,|\,
@@
   \Nat^{\overline{\alpha}} F\,G}\]
-Straightforward calculation shows that terms of $\Nat$-type behave as
-natural transformations with respect to their source and target
-functorial types, i.e., that
+Naturality of term interpretations is then easily verified:
 \begin{thm}\label{eq:ft-from-nat} 
@@
   \rho[\ol{\beta := B}]{\color{red}[\ol{\gamma := C}]} = (\mu
-  T^\set_{H,\rho{\color{red}[\ol{\gamma := C}]}}) \, \ol{B}$.  \\
+  T^\set_{H,\rho{\color{red}[\ol{\gamma := C}]}}) \, \ol{B}$. 
   Secondly, we observe
@@
   is natural in $\ol{B}$ {\color{red} and $\ol{C}$}, since
-  naturality of $\mathit{in}$ with respect to its functor argument and
-  naturality of $\mathit{in}_{T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}}}$ ensure
+{\color{red}  naturality of $\mathit{in}$ with respect to its functor argument and}
+  naturality of $\mathit{in}_{T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}}}$ ensures
   that the following diagram commutes for all $\ol{f : B \to B'}$
@@
 \end{array}\]
-
-Finally, since $\Phi = \emptyset$, the naturality of
+Finally, since $\Phi = \emptyset$, naturality of
 \[
@@
   \phi.\lambda \overline \alpha.H)\overline \beta}\rho[\ol{\beta :=
-    B}]{\color{red}[\ol{\gamma := C}]} \,=\,(\mu T^\set_{H,\rho{\color{red}[\ol{\gamma := C}]}})
-\ol{B}$\\ to $\setsem{\Gamma; \ol{\beta}{\color{red},\ol{\gamma}} \vdash
+    B}]{\color{red}[\ol{\gamma := C}]} \,=\,(\mu T^\set_{H,\rho{\color{red}[\ol{\gamma := C}]}})_{\ol{B}}$ to $\setsem{\Gamma; \ol{\beta}{\color{red},\ol{\gamma}} \vdash
   F}\rho[\ol{\beta := B}]{\color{red}[\ol{\gamma := C}]}$.  To see this, note
@@
     :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\;
-  (\Nat^{{\ol{\beta}{\color{red},\ol{\gamma}}} }\,(\mu \phi.\lambda \overline
-  \alpha.H)\overline \beta\,F)}\,\rho\,d\,\eta)_{\ol{B}{\color{red}\,\ol{C}}}\] is a
-morphism from
-$\setsem{\Gamma; \ol{\beta}{\color{red},\ol{\gamma}} \vdash (\mu
-  \phi.\lambda \overline \alpha.H)\overline \beta}\rho[\ol{\beta :=
-    B}]{\color{red}[\ol{\gamma := C}]}\,=\, (\mu T^\set_{H,\rho{\color{red}[\ol{\gamma := C}]}})
-\ol{B}$ to $\setsem{\Gamma;\ol{\beta}{\color{red},\ol{\gamma}} \vdash
-  F}\rho[\ol{\beta := B}]{\color{red}[\ol{\gamma := C}]}$.
-To see that this family of morphisms is natural in $\ol{B}$
-{\color{red} and
+  (\Nat^{{\ol{\beta}{\color{red},\ol{\gamma}}} }\,(\mu \phi.\lambda
+  \overline \alpha.H)\overline
+  \beta\,F)}\,\rho\,d\,\eta)_{\ol{B}{\color{red}\,\ol{C}}}\] is a
+morphism from $\setsem{\Gamma; \ol{\beta}{\color{red},\ol{\gamma}}
+  \vdash (\mu \phi.\lambda \overline \alpha.H)\overline
+  \beta}\rho[\ol{\beta := B}]{\color{red}[\ol{\gamma := C}]}\,=\, (\mu
+T^\set_{H,\rho{\color{red}[\ol{\gamma := C}]}})_{\ol{B}}$ to
+$\setsem{\Gamma;\ol{\beta}{\color{red},\ol{\gamma}} \vdash
+  F}\rho[\ol{\beta := B}]{\color{red}[\ol{\gamma := C}]}$.  To see
+that this family of morphisms is natural in $\ol{B}$ {\color{red} and
   $\ol{C}$}, we observe that the following diagram commutes for all
-$\ol{f : B \to B'}$ {\color{red} and $\ol{g : C \to C'}$}:
-{\tiny
+$\ol{f : B \to B'}$ {\color{red} and $\ol{g : C \to C'}$}: {\tiny
 \[\hspace*{-0.3in}\begin{tikzcd}[column sep=2.5in, row sep=0.75in]
@@
 \ar[r,"{(\mathit{fold}_{T^\set_{H,\rho{\color{red}[\ol{\gamma :=
-            C'}]}}}\,(\lambda \ol{A}.\,\eta_{\ol{A}{\color{red}\,\ol{C'}}}))_{\ol{B}}\,}" ] 
+            C'}]}}}\,{\color{red}(\lambda \ol{A}.\,}\eta{\color{red}_{\ol{A}\,\ol{C'}})})_{\ol{B}}\,}" ] 
 \ar[d, "{(\mu T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}}) \, \ol{f}}"'] & 
@@
 \ar[r,"{(\mathit{fold}_{T^\set_{H,\rho{\color{red}[\ol{\gamma :=
-            C'}]}}}\,(\lambda \ol{A}.\,\eta_{\ol{A}{\color{red}\,\ol{C'}}}))_{\ol{B'}}\,}"
+            C'}]}}}{\color{red}\,(\lambda \ol{A}.\,}\eta{\color{red}_{\ol{A}\,\ol{C'}})})_{\ol{B'}}\,}"
 ] &  
@@
 \end{tikzcd}\]}
-%Indeed,
+{\color{red} Indeed,}
 by naturality of $\mathit{fold}_{T^\set_{H,\rho[{\color{red}\ol{\gamma
-          := C'}]}}}\,(\lambda
-\ol{A}.\,\eta_{\ol{A}{\color{red}\,\ol{C'}}})$.
-%
-%ensures that the bottom diagram commutes. {\color{red} BOTTOM!}
-%To see that the top one commutes we first observe that, given a
-%natural transformation $\Theta : H \to K : [\set^k, \set] \to [\set^k,
-%  \set]$, the fixpoint natural transformation $\mu \Theta : \mu H \to
-%\mu K : \set^k \to \set$ is defined to be
-%$\mathit{fold}_{H}(\Theta\,(\mu K) \circ \mathit{in}_{K})$, i.e., the
-%unique morphism making the following diagram commute:\label{page:dia1}
-%{\footnotesize
-%\[\begin{tikzcd}[column sep = large]
-%H(\mu H)
-%	\ar[dd, "{\mathit{in}_H}"']
-%	\ar[r, "{H(\mu \Theta)}"]
-%& H(\mu K)
-%	\ar[d, "{\Theta (\mu K)}"] \\
-%& K(\mu K)
-%	\ar[d, "{\mathit{in}_K}"] \\
-%\mu H
-%	\ar[r, "{\mu \Theta}"']
-%& \mu K
-%\end{tikzcd}\]}
-%Taking $\Theta = T^{\set}_{H,f}: T^{\set}_{H,\rho} \to
-%T^{\set}_{H,\rho'}$ thus gives that, for any $f : \rho \to \rho'$ in
-%$\setenv$,
-%\begin{equation}\label{eq:mu-sigma-def}
-%\mathit{in}_{T^{\set}_{H,\rho'}} \circ 
-%T^{\set}_{H,f} (\mu T^{\set}_{H,\rho'}) \circ 
-%T^{\set}_{H,\rho}(\mu T^{\set}_{H,f}) \,=\, 
-%\mu T^{\set}_{H,f} \circ \mathit{in}_{T^{\set}_{H,\rho}}
-%\end{equation}
-Next, note that the action of the functor
+          := C'}]}}}{\color{red}\,(\lambda
+\ol{A}.\,}\eta{\color{red}_{\ol{A}\,\ol{C'}}})$.
+{\color{red} ensures that the bottom diagram commutes. 
+To see that the top one commutes we first observe that, given a
+natural transformation $\Theta : H \to K : [\set^k, \set] \to [\set^k,
+  \set]$, the fixpoint natural transformation $\mu \Theta : \mu H \to
+\mu K : \set^k \to \set$ is defined to be
+$\mathit{fold}_{H}(\Theta\,(\mu K) \circ \mathit{in}_{K})$, i.e., the
+unique morphism making the following diagram commute:\label{page:dia1}
+{\footnotesize
+\[\begin{tikzcd}[column sep = large]
+H(\mu H)
+	\ar[dd, "{\mathit{in}_H}"']
+	\ar[r, "{H(\mu \Theta)}"]
+& H(\mu K)
+	\ar[d, "{\Theta (\mu K)}"] \\
+& K(\mu K)
+	\ar[d, "{\mathit{in}_K}"] \\
+\mu H
+	\ar[r, "{\mu \Theta}"']
+& \mu K
+\end{tikzcd}\]}
+Taking $\Theta = T^{\set}_{H,f}: T^{\set}_{H,\rho} \to
+T^{\set}_{H,\rho'}$ thus gives that, for any $f : \rho \to \rho'$ in
+$\setenv$,
+\begin{equation}\label{eq:mu-sigma-def}
+\mathit{in}_{T^{\set}_{H,\rho'}} \circ 
+T^{\set}_{H,f} (\mu T^{\set}_{H,\rho'}) \circ 
+T^{\set}_{H,\rho}(\mu T^{\set}_{H,f}) \,=\, 
+\mu T^{\set}_{H,f} \circ \mathit{in}_{T^{\set}_{H,\rho}}
+\end{equation}}
+{\color{red} Next, note that the action of the functor
 \[\lambda \ol{B}.{\color{red} \lambda \ol{C}.} \setsem{\Gamma; \ol{\beta}{\color{red},
@@
 \end{equation}
-
-HERE!!!!
-
 Finally, to see that the top diagram in the diagram on
@@
 the top diagram in the diagram on
-  page~\pageref{page:dia1} commutes, as desired.
+  page~\pageref{page:dia1} commutes, as desired.}
 To see that, for all $\rho : \setenv$, $d \in
 \setsem{\Gamma; \emptyset \vdash \emptyset}\rho$, and $\eta :
-\setsem{\Gamma; \emptyset \vdash \Nat^{\ol{\beta},
-    \ol{\gamma}}\,H[\phi :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F}
+\setsem{\Gamma; \emptyset \vdash \Nat^{\ol{\beta}{\color{red},
+    \ol{\gamma}}}\,H[\phi :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F}
 \rho$,
 \[\setsem{\Gamma; \emptyset~|~\emptyset
-  \vdash \fold^F_H : \Nat^\emptyset\;(\Nat^{\ol{\beta},
-    \ol{\gamma}}\,H[\phi :=_{\ol{\beta}} F][\ol{\alpha :=
-      \beta}]\,F)\; (\Nat^{{\ol{\beta},\ol{\gamma}} }\,(\mu
-  \phi.\lambda \overline \alpha.H)\overline \beta\,F)}\,\rho\,d\,\eta\]
-satisfies the additional condition needed for it to be in
-$\setsem{\Gamma;\emptyset \vdash \Nat^{{\ol{\beta},\ol{\gamma}}
-  }\,(\mu \phi.\lambda \overline \alpha.H)\overline \beta\;F}\,\rho$,
-let $\ol{R : \rel(B,B')}$ and $\ol{S : \rel(C,C')}$.  Since $\eta$
-satisfies the additional condition needed for it to be in
-$\setsem{\Gamma; \emptyset \vdash \Nat^{\ol{\beta},
-    \ol{\gamma}}\,(H[\phi := F][\ol{\alpha := \beta}])\,F} \rho$,
+  \vdash \fold^F_H : \Nat^\emptyset\;(\Nat^{\ol{\beta}{\color{red},
+    \ol{\gamma}}}\,H[\phi :=_{\ol{\beta}} F][\ol{\alpha :=
+      \beta}]\,F)\; (\Nat^{{\ol{\beta}{\color{red},\ol{\gamma}}} }\,(\mu
+  \phi.\lambda \overline \alpha.H)\overline
+  \beta\,F)}\,\rho\,d\,\eta\] satisfies the additional condition
+needed for it to be in $\setsem{\Gamma;\emptyset \vdash
+  \Nat^{{\ol{\beta}{\color{red},\ol{\gamma}}} }\,(\mu \phi.\lambda \overline
+  \alpha.H)\overline \beta\;F}\,\rho$, let $\ol{R : \rel(B,B')}$
+{\color{red} and $\ol{S : \rel(C,C')}$}.  Since $\eta$ satisfies the
+additional condition needed for it to be in $\setsem{\Gamma; \emptyset
+  \vdash \Nat^{\ol{\beta}{\color{red}, \ol{\gamma}}}\,(H[\phi := F][\ol{\alpha :=
+      \beta}])\,F} \rho$,
 \[\begin{array}{ll}
- & (\,(\mathit{fold}_{T^\set_{H,\rho[\ol{\gamma :=
-        C}]}}\,(\lambda \ol{A}.\,\eta_{\ol{A}\,\ol{C}}))_{\ol{B}},\,
-(\mathit{fold}_{T^\set_{H,\rho[\ol{\gamma :=
-        C'}]}}\,(\lambda \ol{A}.\eta_{\ol{A}\,\ol{C'}}))_{\ol{B'}}\,) 
+ & (\,(\mathit{fold}_{T^\set_{H,\rho{\color{red}[\ol{\gamma :=
+        C}]}}}{\color{red}\,(\lambda \ol{A}.\,}\eta{\color{red}_{\ol{A}\,\ol{C}})})_{\ol{B}},\,
+(\mathit{fold}_{T^\set_{H,\rho{\color{red}[\ol{\gamma :=
+        C'}]}}}{\color{red}\,(\lambda \ol{A}.}\eta{\color{red}_{\ol{A}\,\ol{C'}})})_{\ol{B'}}\,) 
 \end{array}\]
@@
 \[\begin{array}{ll}
-  & (\mu T_{H,\Eq_\rho[\ol{\gamma := S}]}) \,\ol{R} \to
-\relsem{\Gamma;\ol{\gamma},\ol{\beta} \vdash F}\Eq_\rho[\ol{\gamma := 
-    S}][\ol{\beta:= R}]\\ 
-= & (\mu T_{H,\Eq_\rho[\ol{\gamma :=
-      S}]})\,\ol{\relsem{\Gamma;\ol{\gamma},\ol{\beta} 
-  \vdash \beta}\Eq_\rho[\ol{\gamma := S}][\ol{\beta := R}]} \\ 
+  & (\mu T_{H,\Eq_\rho{\color{red}[\ol{\gamma := S}]}}) \,\ol{R} \to
+\relsem{\Gamma;{\color{red}\ol{\gamma},}\ol{\beta} \vdash F}\Eq_\rho{\color{red}[\ol{\gamma := 
+    S}]}[\ol{\beta:= R}]\\ 
+= & (\mu T_{H,\Eq_\rho{\color{red}[\ol{\gamma :=
+      S}]}})\,\ol{\relsem{\Gamma;{\color{red}\ol{\gamma},}\ol{\beta} 
+  \vdash \beta}\Eq_\rho{\color{red}[\ol{\gamma := S}]}[\ol{\beta := R}]} \\ 
   &\hspace{1.5in} \to
-\relsem{\Gamma;\ol{\gamma},\ol{\beta} \vdash F}\Eq_\rho[\ol{\gamma := 
-    S}][\ol{\beta:= R}]\\ 
-= & \relsem{\Gamma; \ol{\gamma},\ol{\beta} \vdash (\mu \phi. \lambda
-  \ol{\alpha}. H)\ol{\beta}}\Eq_\rho[\ol{\gamma := S}][\ol{\beta := R}] \\ 
+\relsem{\Gamma;{\color{red}\ol{\gamma},}\ol{\beta} \vdash F}\Eq_\rho{\color{red}[\ol{\gamma := 
+    S}]}[\ol{\beta:= R}]\\ 
+= & \relsem{\Gamma; {\color{red}\ol{\gamma},}\ol{\beta} \vdash (\mu \phi. \lambda
+  \ol{\alpha}. H)\ol{\beta}}\Eq_\rho{\color{red}[\ol{\gamma := S}]}[\ol{\beta := R}] \\ 
   &\hspace{1.5in}\to
-\relsem{\Gamma;\ol{\gamma},\ol{\beta} \vdash F}\Eq_\rho[\ol{\gamma := 
-    S}][\ol{\beta:= R}]
-\end{array}\]
+\relsem{\Gamma;{\color{red}\ol{\gamma},}\ol{\beta} \vdash F}\Eq_\rho{\color{red}[\ol{\gamma := 
+    S}]}[\ol{\beta:= R}]
+\end{array}\]
+Finally, since $\Phi = \emptyset$, naturality of
+\[\setsem{\Gamma; \emptyset~|~\emptyset \vdash \fold^F_H :
+  \Nat^\emptyset\;(\Nat^{\ol{\beta}{\color{red}, \ol{\gamma}}}\,H[\phi
+    :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\;
+  (\Nat^{{\ol{\beta}{\color{red},\ol{\gamma}}} }\,(\mu \phi.\lambda
+  \overline \alpha.H)\overline \beta \;F)}\] in $\rho$ is trivial.
 \end{itemize}

### 2020-12-16 free-theorems-nesteds/20-mscs.tex
Sections 5-8 for proofreading

@@
 \mu \beta. \onet + \alpha \times \phi\beta$, in which $\phi$ and
-$\alpha$ must both be non-functorial variables. {\color{red} It is in
-  fact possible to allow ``extra'' $0$-ary functorial variables in the
-  body of $\mu$-types (functorial variables of higher arity are the
-  real problem), which would allow the first-order representations of
-  ADTs to be functorial.  However, it simplifies the formation rule
-  for $\mu$-types considerably to disallow even these $0$-ary
-  functorial variables. Since we can always use an ADT's (semantically
-  equivalent) second-order representation when functoriality is
-  needed, this restriction allows us to simplify the type syntax
-  without sacrificing the expressivity of the calculus.}
+$\alpha$ must both be non-functorial variables.  It is in fact
+possible to allow ``extra'' $0$-ary functorial variables in the body
+of $\mu$-types (functorial variables of higher arity are the real
+problem), which would allow the first-order representations of ADTs to
+be functorial. However, doing this requires some changes to the
+formation rule for $\mu$-types, as well as the delicate threading of
+some additional conditions throughout the model construction.  But
+since we can always use an ADT's (semantically equivalent)
+second-order representation when functoriality is needed, disallowing
+such ``extra'' variables does not negatively impact the expressivity
+of our calculus. We therefore pursue the simpler syntax here.
 %The representation of $\mathit{List}\,\alpha$ as the ADT $\mu
@@
 %system.
-These restrictions ensure that the set and relational interpretations
-of types in Section~\ref{sec:type-interp} are functors. In
-Section~\ref{sec:conclusion} we discuss how more expressive types
-might be accommodated.
+%These restrictions ensure that the set and relational interpretations
+%of types in Section~\ref{sec:type-interp} are functors. In
+%Section~\ref{sec:conclusion} we discuss how more expressive types
+%might be accommodated.
 
@@
 transformations with respect to their source and target functorial
-types. If $\Gamma; \ol{\alpha} \vdash F$ then define the {\em
-  identity} $\id_F$ on $F$ by $\id_F = \Gamma;\emptyset~|~\emptyset
-\vdash L_{\ol{\alpha}}x.x : \Nat^{\ol{\alpha}} F\,F$, and if $\Gamma;
-\emptyset \,|\, \Delta \vdash t: \Nat^{\overline{\alpha}} F\,G$ and
-$\Gamma; \emptyset \,|\, \Delta \vdash s: \Nat^{\overline{\alpha}}
-G\,H$ are terms then define the {\em composition} $s \circ t$ of $t$
-and $s$ by $s \circ t = \Gamma; \emptyset\,|\, \Delta \vdash
-L_{\overline{\alpha}}
+types, and derive some conseqences of this observation. If $\Gamma;
+\ol{\alpha} \vdash F$ then define the {\em identity} $\id_F$ on $F$ by
+$\id_F = \Gamma;\emptyset~|~\emptyset \vdash L_{\ol{\alpha}}x.x :
+\Nat^{\ol{\alpha}} F\,F$, and if $\Gamma; \emptyset \,|\, \Delta
+\vdash t: \Nat^{\overline{\alpha}} F\,G$ and $\Gamma; \emptyset \,|\,
+\Delta \vdash s: \Nat^{\overline{\alpha}} G\,H$ are terms then define
+the {\em composition} $s \circ t$ of $t$ and $s$ by $s \circ t =
+\Gamma; \emptyset\,|\, \Delta \vdash L_{\overline{\alpha}}
 x. s_{\overline{\alpha}}(t_{\overline{\alpha}}x):

### 2020-12-16 free-theorems-nesteds/20-mscs.tex
Sections 5-8 for proofreading

@@
 of our calculus. We therefore pursue the simpler syntax here.
-%The representation of $\mathit{List}\,\alpha$ as the ADT $\mu
-%\beta. \onet + \alpha \times \beta$ cannot be functorial in
-%$\alpha$. By contrast, its representation as the nested type $(\mu
-%\phi. \lambda \beta.\,\onet + \beta \times \phi \beta)\,\alpha$ can be
-%considered either functorial in $\alpha$ or not, according to whether
-%$\alpha$ is in the functorial or non-functorial context in which this
-%type is formed. If $\emptyset;\alpha \vdash \mathit{List} \,\alpha$,
-%then $\vdash \Nat^\alpha \onet\, (\mathit{List}\,\alpha)$ is
-%well-formed; if $\alpha;\emptyset \vdash \List \,\alpha$, then it is
-%not.  Similarly, if $\mathit{Tree}\,\alpha\,\gamma = \mu \,\beta.\,
-%\alpha + \beta \times \gamma \times \beta$, then $\gamma;\emptyset
-%\vdash \Nat^\alpha (\mathit{List}\, \alpha) \, (\mathit{Tree} \,
-%\alpha\, \gamma)$ is well-formed, but $\emptyset;\gamma \vdash
-%\Nat^\alpha (\mathit{List}\, \alpha) \, (\mathit{Tree} \, \alpha\,
-%\gamma)$ is not. The body $F$ of a type $(\mu \phi.\lambda
-%\ol\alpha.F)\ol G$ can only be functorial in $\phi$ and the variables
-%in $\ol\alpha$. As a result, both $\phi$ and the variables in
-%$\ol\alpha$ must be non-functorial in $\mathit{GRose}\, \phi \, \alpha
-%= \mu\beta . \alpha \times \phi \beta$, $\mathit{GRose}\,\phi\,\alpha$
-%cannot be the (co)domain of a $\Nat$-type binding any variables, and
-%no type of the form $(\mu \phi. \lambda
-%\alpha. \mathit{GRose}\,\phi\,\alpha)\tau$ is well-formed in our
-%system.
-%These restrictions ensure that the set and relational interpretations
-%of types in Section~\ref{sec:type-interp} are functors. In
-%Section~\ref{sec:conclusion} we discuss how more expressive types
-%might be accommodated.
 
@@
 Interestingly, even some recursive functions of a single proper nested
-type --- e.g., a reverse function for bushes --- cannot be expressed
-as folds because the algebra arguments needed to define them are again
-recursive functions with types of the same problematic form as the
-type of, e.g., a zip function for perfect trees.  Expressivity of
-folds for nested types has long been a vexing issue, and this is
-naturally inherited by our calculus. Adding more expressive recursion
-combinators could help, but since this is orthogonal to the issue of
-parametricity in the presence of primitive nested types we do not
-consider it further here.
+type --- e.g., a reverse function for bushes that is a true involution
+--- cannot be expressed as folds because the algebra arguments needed
+to define them are again recursive functions with types of the same
+problematic form as the type of, e.g., a zip function for perfect
+trees.  Expressivity of folds for nested types has long been a vexing
+issue, and this is naturally inherited by our calculus. Adding more
+expressive recursion combinators could help, but since this is
+orthogonal to the issue of parametricity in the presence of primitive
+nested types we do not consider it further here.
 

### 2020-12-16 free-theorems-nesteds/20-mscs.tex
Sections 2-8 done

@@
 $G$. 
-%due to typing rule restrictions for the $\mu$ types, $\phi$ either
-%does not appear in $G$, or must have arity $0$, in which case
-%$\ol\alpha$ must be empty, if $\phi$ appears in $G$, and uses
-%Equation~\ref{eq:iel-fix-point-intermediate2} for $G$ when $\phi$ has
-%arity $0$. {\color{red} Note that if functorial variables of arity
-%  greater than $0$ were allowed to appear in the bodies of
-%  $\mu$-types, then the aboe clause of the IEL would fail.}
 \end{itemize}
@@
 \end{array}\]
-Every well-formed term $\Gamma;\Phi~|~\Delta \vdash t : F$ then
-has, for every $\rho \in \setenv$, a set interpretation
-$\setsem{\Gamma;\Phi~|~\Delta \vdash t : F}\rho$ as a natural
-transformation from $\setsem{\Gamma; \Phi \vdash \Delta}\rho$ to
-$\setsem{\Gamma; \Phi \vdash F}\rho$, and, for every $\rho \in
-\relenv$, a relational interpretation $\relsem{\Gamma;\Phi~|~\Delta
-  \vdash t : F}\rho$ as a natural transformation from
-$\relsem{\Gamma; \Phi \vdash \Delta}\rho$ to $\relsem{\Gamma; \Phi
-  \vdash F}\rho$.
+Every well-formed term $\Gamma;\Phi~|~\Delta \vdash t : F$ then has,
+for every $\rho \in \setenv$, a set interpretation
+$\setsem{\Gamma;\Phi~|~\Delta \vdash t : F}\rho$ as the component at
+$\rho$ of a natural transformation from $\setsem{\Gamma; \Phi \vdash
+  \Delta}$ to $\setsem{\Gamma; \Phi \vdash F}$, and, for every $\rho
+\in \relenv$, a relational interpretation
+$\relsem{\Gamma;\Phi~|~\Delta \vdash t : F}\rho$ as the component at
+$\rho$ of a natural transformation from $\relsem{\Gamma; \Phi \vdash
+  \Delta}$ to $\relsem{\Gamma; \Phi \vdash F}$.
 
@@
 \pi_{|\Delta|+1}\\
-%\dsem{\Gamma;\emptyset \,|\, \Delta \vdash \lambda x.t : \sigma \to \tau}\rho &
-%= & \curry (\dsem{\Gamma;\emptyset \,|\, \Delta, x : \sigma \vdash t :
-%  \tau}\rho)\\ 
-%\dsem{\Gamma;\emptyset \,|\, \Delta \vdash st: \tau} \rho & = &
-%\eval \circ
-% \langle \dsem{\Gamma;\emptyset \,|\, \Delta \vdash s: \sigma \to
-%  \tau}\rho, \dsem{\Gamma;\emptyset \,|\, \Delta \vdash t: \sigma}\rho
-%\rangle\\
-\dsem{\Gamma;\emptyset \,|\, \Delta \vdash L_{\overline \alpha} x.t : \Nat^{\overline
+\dsem{\Gamma;\emptyset \,|\, \Delta \vdash L_{\overline \alpha} x.t :
+  \Nat^{\overline 
     \alpha} \,F \,G}\rho & = &  \curry (\dsem{\Gamma;\overline \alpha
@@
 & & \\
-%% \color{red} \mbox{Add rules for } \forall \mbox{ if we include it} & & \\
-%\dsem{\Gamma;\Phi \,|\, \Delta,x :F \vdash x : F} \rho& = &
-%\pi_{|\Delta|+1}\\
 \dsem{\Gamma;\Phi \,|\, \Delta \vdash \bot_F t : F} \rho& = &
@@
 \llbracket \Gamma;\emptyset \,|\, \emptyset \vdash \tin_H :
-Nat^{\ol{\beta}{\color{red}, \ol{\gamma}}} \, H[\phi := (\mu \phi.\lambda {\overline
+Nat^{\ol{\beta}} \, H[\phi := (\mu \phi.\lambda {\overline
     \alpha}.H){\overline \beta}][\ol{\alpha := \beta}] & = &
 \lambda
-d{\color{red}\,\ol{B}\,\ol{C}.\,(}\mathit{in}_{{T}^X_{H,\rho{\color{red}[\ol{\gamma
-          := 
-        C}]}}}{\color{red})_{\ol{B}}} \;\;\mbox{ where } X \mbox{ is } \set \mbox{ when } \\ 
+d.\,\mathit{in}_{{T}^X_{H,\rho}} \;\;\mbox{ where } X \mbox{ is } \set \mbox{ when } \\ 
 \hspace*{0.79in}(\mu \phi.\lambda {\overline \alpha}.H){\overline
@@
 \llbracket \Gamma;\emptyset \,|\, \emptyset \vdash
-  \fold^F_H : \Nat^\emptyset\;(\Nat^{\ol{\beta}{\color{red}, \ol{\gamma}}}\,H[\phi
+  \fold^F_H : \Nat^\emptyset\;(\Nat^{\ol{\beta}}\,H[\phi
     :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F) & = &  
-\lambda d\,\eta{\color{red}\,\ol{B}\,\ol{C}.\,
-(}\mathit{fold}_{T^X_{H,\rho{\color{red}[\ol{\gamma := C}]}}} \, (\lambda
-\ol{A}.\,\eta_{\ol{A}\,\ol{C}}){\color{red})_{\ol{B}}}\\ 
-\hspace*{0.79in}(\Nat^{{\ol{\beta}{\color{red} ,\ol{\gamma}}} }\,(\mu
+\lambda d\,\eta. \,\mathit{fold}_{T^X_{H,\rho}} \, \eta\\ 
+\hspace*{0.79in}(\Nat^{{\ol{\beta}} }\,(\mu
   \phi.\lambda \overline \alpha.H)\overline \beta\;F)
@@
 The return type for the semantic fold is
-  $\dsem{\Gamma;\ol\beta,{\color{red}\ol\gamma} \vdash F}\rho{\color{red}[\ol{\gamma :=
-    C}]}[\ol{\beta := B}]$.
+  $\dsem{\Gamma;\ol\beta \vdash F}\rho[\ol{\beta := B}]$.
 This interpretation gives
@@
   t [\ol{\alpha := K}][x := s]}\] Term extensionality for both types
-and terms --- i.e., \[\dsem{\Gamma;\Phi\vdash (L_\alpha x.t)_\alpha
-  \top : F} = \dsem{\Gamma;\Phi \vdash t : F}\] and
-\[\dsem{\Gamma;\Phi\vdash (L_\alpha x.t)_\alpha x : F} =
-\dsem{\Gamma;\Phi \vdash t : F}\] --- then follow (assuming both sides
-of these equations are defined).
+and terms --- i.e., \[\dsem{\Gamma;\Phi~|~\Delta\vdash (L_\alpha x.t)_\alpha
+  \top : F} = \dsem{\Gamma;\Phi~|~\Delta \vdash t : F}\] and
+\[\dsem{\Gamma;\Phi~|~\Delta\vdash (L_\alpha x.t)_\alpha x : F} =
+\dsem{\Gamma;\Phi~|~\Delta \vdash t : F}\] --- then follow (assuming
+both sides of these equations are defined).
 
@@
 $\bullet$\; $\fold$ satisfies its universal property, i.e., if $\xi =
-\Nat^{\overline{\beta}{\color{red}, \overline{\gamma}}}\, H[\phi :=
+\Nat^{\overline{\beta}}\, H[\phi :=
   (\mu \phi. \lambda
@@
 \[\hspace*{-1in}\setsem{\Gamma; \emptyset \,|\, x:
-  \Nat^{\overline{\beta}{\color{red}, \overline{\gamma}}} H[\phi :=
+  \Nat^{\overline{\beta}} H[\phi :=
     F][\overline{\alpha := \beta}]\, F \vdash ((\fold_{H,
@@
 
-\[= \setsem{ \Gamma; \emptyset \,|\, x: \Nat^{\overline{\beta}{\color{red},
-    \overline{\gamma}}} H[\phi := F][\overline{\alpha := \beta}]\, F
+\[= \setsem{ \Gamma; \emptyset \,|\, x: \Nat^{\overline{\beta}} H[\phi
+    := F][\overline{\alpha := \beta}]\, F 
   \vdash x \circ \big( (\map_{H [\ol{\alpha := \beta}]}^{(\mu
@@
 $\bullet$\; $\tin$ has a right inverse, i.e., if $\xi =
-\Nat^{\overline{\beta}{\color{red}, \overline{\gamma}}}\, (\mu \phi. \lambda
+\Nat^{\overline{\beta}}\, (\mu \phi. \lambda
 \overline{\alpha}. H)\overline{\beta}\; (\mu \phi. \lambda
@@
 $\bullet$\; $\tin$ has a left inverse, i.e., if $\xi =
-\Nat^{\overline{\beta}{\color{red}, \overline{\gamma}}} H[\phi := (\mu
+\Nat^{\overline{\beta}} H[\phi := (\mu
   \phi. \lambda \overline{\alpha}. H)\overline{\beta}]\, H[\phi :=
@@
 i.e.,
-%{\color{red} by remembering the following facts,
-%\begin{align*}
-%\setsem{\Gamma;\Phi \vdash F[\ol{\alpha := K}]}\rho
-%&= \setsem{\Gamma; \ol{\alpha} \vdash F}\rho[\ol{\alpha :=
-%    \setsem{\Gamma;\Phi \vdash K}\rho}] \\
-%\setsem{\Gamma;\Phi \vdash F[\ol{\alpha := K}]}f
-%&= \setsem{\Gamma; \ol{\alpha} \vdash F}
-%  f [\ol{\alpha := \setsem{\Gamma;\Phi \vdash K}f}] \\
-%\setsem{\Gamma;\Phi \vdash G[\ol{\alpha := K}]}\rho
-%&= \setsem{\Gamma; \ol{\alpha} \vdash G}\rho[\ol{\alpha :=
-%    \setsem{\Gamma;\Phi \vdash K}\rho}] \\
-%\setsem{\Gamma;\Phi \vdash G[\ol{\alpha := K}]}f
-%&= \setsem{\Gamma; \ol{\alpha} \vdash
-%  G} f [\ol{\alpha := \setsem{\Gamma;\Phi \vdash K}f}]
-%\end{align*}}
-%we need to show
 that
@@
 $\underline{\Gamma;\emptyset \,|\, \emptyset \vdash \tin_H :
-  \Nat^{\ol{\beta}{\color{red},\ol{\gamma}}} \, H[\phi := (\mu \phi.\lambda
+  \Nat^{\ol{\beta}} \, H[\phi := (\mu \phi.\lambda
     {\overline \alpha}.H){\overline \beta}][\ol{\alpha := \beta}]
@@
   then $$\setsem{\Gamma;\emptyset \,|\, \emptyset \vdash \tin_H :
-    \Nat^{\ol{\beta}{\color{red},\ol{\gamma}}} \, H[\phi := (\mu \phi.\lambda
+    \Nat^{\ol{\beta}} \, H[\phi := (\mu \phi.\lambda
       {\overline \alpha}.H){\overline \beta}][\ol{\alpha := \beta}]
@@
   is in $\setsem{\Gamma;\emptyset \vdash
-    \Nat^{\ol{\beta}{\color{red},\ol{\gamma}}} \, H[\phi := (\mu \phi.\lambda
+    \Nat^{\ol{\beta}} \, H[\phi := (\mu \phi.\lambda
       {\overline \alpha}.H){\overline \beta}][\ol{\alpha := \beta}]
     \;(\mu \phi.\lambda {\overline \alpha}.H){\overline \beta}}\,
-  \rho$, we first note that, for all $\ol{B}$ {\color{red} and $\ol{C}$},
+  \rho$, we first note that, for all $\ol{B}$,
   \begin{align*}
     &(\setsem{\Gamma;\emptyset \,|\, \emptyset \vdash \tin_H :
-    \Nat^{\ol{\beta}{\color{red},\ol{\gamma}}} \, H[\phi := (\mu \phi.\lambda
+    \Nat^{\ol{\beta}} \, H[\phi := (\mu \phi.\lambda
       {\overline \alpha}.H){\overline \beta}][\ol{\alpha := \beta}]
     \;(\mu \phi.\lambda {\overline \alpha}.H){\overline \beta}}\,
-  \rho\,d)_{\ol{B}{\color{red}\,\ol{C}}}\, \\
+  \rho\,d)_{\ol{B}}\, \\
     =\,
-    &(\mathit{in}_{T^\set_{H,\rho{\color{red}[\ol{\gamma := C}]}}})_{\ol{B}}
+    &(\mathit{in}_{T^\set_{H,\rho}})_{\ol{B}}
   \end{align*}
@@
   \begin{align*}
-    &\setsem{\Gamma;\ol{\beta}{\color{red},\ol{\gamma}} \vdash H[\phi := (\mu
+    &\setsem{\Gamma;\ol{\beta} \vdash H[\phi := (\mu
       \phi.\lambda {\overline \alpha}.H){\overline \beta}][\ol{\alpha
-        := \beta}]}\rho[\ol{\beta := B}]{\color{red}[\ol{\gamma := C}]} 
-        \\ = \,
-    &T^\set_{H,\rho{\color{red}[\ol{\gamma := C}]}}\, (\mu T^\set_{H,\rho{\color{red}[\ol{\gamma
-        := C}]}}) \, \ol{B}
+        := \beta}]}\rho[\ol{\beta := B}]  \\ = \,
+    &T^\set_{H,\rho}\, (\mu T^\set_{H,\rho}) \, \ol{B}
   \end{align*}
-        to $\setsem{\Gamma;\ol{\beta},{\color{red}\ol{\gamma}}
+        to $\setsem{\Gamma;\ol{\beta}
     \vdash (\mu \phi.\lambda {\overline \alpha}.H){\overline \beta}}
-  \rho[\ol{\beta := B}]{\color{red}[\ol{\gamma := C}]} = (\mu
-  T^\set_{H,\rho{\color{red}[\ol{\gamma := C}]}}) \, \ol{B}$. 
+  \rho[\ol{\beta := B}] = (\mu
+  T^\set_{H,\rho}) \, \ol{B}$. 
   Secondly, we observe
@@
     &\setsem{\Gamma;\emptyset \,|\, \emptyset \vdash \tin_H :
-    \Nat^{\ol{\beta}{\color{red},\ol{\gamma}}} \, H[\phi := (\mu \phi.\lambda
+    \Nat^{\ol{\beta}} \, H[\phi := (\mu \phi.\lambda
       {\overline \alpha}.H){\overline \beta}][\ol{\alpha := \beta}]
@@
       \beta}}\,\rho\,d \\ 
-      = \, &\lambda
-  \ol{B}{\color{red}\,\ol{C}}.\,(\mathit{in}_{T^\set_{H,\rho{\color{red}[\ol{\gamma :=
-          C}]}}})_{\ol{B}}
+      = \, &\mathit{in}_{T^\set_{H,\rho}}
   \end{align*}
-  is natural in $\ol{B}$ {\color{red} and $\ol{C}$}, since
-{\color{red}  naturality of $\mathit{in}$ with respect to its functor argument and}
-  naturality of $\mathit{in}_{T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}}}$ ensures
-  that the following diagram commutes for all $\ol{f : B \to B'}$
-  {\color{red} and
-  $\ol{g : C \to C'}$}:
+  is natural in $\ol{B}$, since naturality of
+  $\mathit{in}_{T^\set_{H,\rho}}$
+  ensures that the following diagram commutes for all $\ol{f : B \to
+    B'}$:
 
@@
   \[\hspace*{-0.15in}\begin{tikzcd}[column sep=2.5in, row sep=0.75in]
-%T^\set_{H,\rho[\ol{\gamma := C}]}\, (\mu T^\set_{H,\rho[\ol{\gamma := C}]})\, \ol{B}
-%\ar[d, "{T^\set_{H,\id_\rho[\ol{\gamma := g}]}\,(\mu
-%    T^\set_{H,\id_\rho[\ol{\gamma := g}]}) \, \ol{B}}"] \ar[r,
-%  "{(\mathit{in}_{T^\set_{H,\rho[\ol{\gamma := C}]}})_{\ol{B}}}" ]
-%& (\mu T^\set_{H,\rho[\ol{\gamma := C}]})\, \ol{B} \ar[d, "{(\mu
-%    T^\set_{H,\id_\rho[\ol{\gamma := g}]}) \, \ol{B}}"]\\
-T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}}\, (\mu
-T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}})\, \ol{B} \ar[d, "{T^\set_{H,\rho{\color{red}[\ol{\gamma :=
-          C'}]}}\, (\mu T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}})\, \ol{f}}"]
-\ar[r,"{(\mathit{in}_{T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}})_{\ol{B}}}}" ] & 
-  (\mu T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}})\, \ol{B}
-\ar[d,"{\mu {T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}}}\, {\ol{f}}}" ] 
+T^\set_{H,\rho}\, (\mu
+T^\set_{H,\rho})\, \ol{B} \ar[d, "{T^\set_{H,\rho}\, (\mu T^\set_{H,\rho})\, \ol{f}}"]
+\ar[r,"{(\mathit{in}_{T^\set_{H,\rho})_{\ol{B}}}}" ] & 
+  (\mu T^\set_{H,\rho})\, \ol{B}
+\ar[d,"{\mu {T^\set_{H,\rho}}\, {\ol{f}}}" ] 
 \\
-T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}}\, (\mu T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}})\,
-\ol{B'} \ar[r, "{(\mathit{in}_{T^\set_{H,\rho{\color{red}[\ol{\gamma :=
-            C'}]}}})_{\ol{B'}}}"] & (\mu T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}})\,
+T^\set_{H,\rho}\, (\mu T^\set_{H,\rho})\,
+\ol{B'} \ar[r, "{(\mathit{in}_{T^\set_{H,\rho}})_{\ol{B'}}}"] & (\mu
+T^\set_{H,\rho})\, 
 \ol{B'}
@@
 \[\setsem{\Gamma;\emptyset \,|\, \emptyset \vdash
-  \tin_H : \Nat^{\ol{\beta}{\color{red},\ol{\gamma}}} \, H[\phi := (\mu \phi.\lambda
+  \tin_H : \Nat^{\ol{\beta}} \, H[\phi := (\mu \phi.\lambda
     {\overline \alpha}.H){\overline \beta}][\ol{\alpha := \beta}]
@@
 \[\setsem{\Gamma;\emptyset \vdash
-  \Nat^{\ol{\beta}{\color{red},\ol{\gamma}}} \, H[\phi := (\mu \phi.\lambda
+  \Nat^{\ol{\beta}} \, H[\phi := (\mu \phi.\lambda
     {\overline \alpha}.H){\overline \beta}][\ol{\alpha := \beta}]
   \;(\mu \phi.\lambda {\overline \alpha}.H){\overline \beta}}\,\rho\]
-follows from the fact that, for every $\ol{R : \rel(B,B')}$
-{\color{red} and $\ol{S : \rel(C,C')}$},
+follows from the fact that, for every $\ol{R : \rel(B,B')}$,
 \[\hspace*{-0.2in}
@@
  & (\,(\setsem{\Gamma;\emptyset \,|\, \emptyset \vdash \tin_H :
-      \Nat^{\ol{\beta}{\color{red},\ol{\gamma}}} \, H[\phi := (\mu \phi.\lambda
+      \Nat^{\ol{\beta}} \, H[\phi := (\mu \phi.\lambda
         {\overline \alpha}.H){\overline \beta}][\ol{\alpha := \beta}]
       \;(\mu \phi.\lambda {\overline \alpha}.H){\overline
-        \beta}}\,\rho\,d)_{\ol{B}{\color{red},\ol{C}}},\,\\ & \hspace*{0.5in}(\setsem{\Gamma;\emptyset
-      \,|\, \emptyset \vdash \tin_H : \Nat^{\ol{\beta}{\color{red},\ol{\gamma}}} \,
+        \beta}}\,\rho\,d)_{\ol{B}},\,\\ & \hspace*{0.5in}(\setsem{\Gamma;\emptyset
+      \,|\, \emptyset \vdash \tin_H : \Nat^{\ol{\beta}} \,
       H[\phi := (\mu \phi.\lambda {\overline \alpha}.H){\overline
@@
         \alpha}.H){\overline
-        \beta}}\,\rho\,d)_{\ol{B'}{\color{red},\ol{C'}}}\,)\\ &= (\,
-    (\mathit{in}_{T^\set_{H,\rho{\color{red}[\ol{\gamma := C}]}}})_{\ol{B}},
-    (\mathit{in}_{T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}}})_{\ol{B'}}\,)
+        \beta}}\,\rho\,d)_{\ol{B'}}\,)\\ &= (\,
+    (\mathit{in}_{T^\set_{H,\rho}})_{\ol{B}},
+    (\mathit{in}_{T^\set_{H,\rho}})_{\ol{B'}}\,)
 \end{array}\]
@@
 \[\begin{array}{ll}
-& (\, T^\set_{H,\rho{\color{red}[\ol{\gamma := C}]}}\, (\mu
-T^\set_{H,\rho{\color{red}[\ol{\gamma := C}]}}) \, 
-\ol{B} \to (\mu T^\set_{H,\rho{\color{red}[\ol{\gamma := C}]}}) \, \ol{B}, \, \\
-& \hspace*{0.5in}T^\set_{H,\rho{\color{red} [\ol{\gamma := C'}]}}\, (\mu
-T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}}) \, 
-\ol{B'} \to (\mu T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}}) \,\ol{B'} \, )\\
+& (\, T^\set_{H,\rho}\, (\mu T^\set_{H,\rho}) \, \ol{B} \to (\mu
+T^\set_{H,\rho}) \, \ol{B}, \, T^\set_{H,\rho}\, (\mu T^\set_{H,\rho})
+\, \ol{B'} \to (\mu T^\set_{H,\rho}) \,\ol{B'} \, )\\[1ex]
 = &
- \relsem{\Gamma;\ol{\beta},{\color{red}\ol{\gamma}} \vdash H[\phi := (\mu
+ \relsem{\Gamma;\ol{\beta} \vdash H[\phi := (\mu
     \phi.\lambda {\overline \alpha}.H){\overline \beta}][\ol{\alpha :=
-      \beta}]}\Eq_\rho[\ol{\beta := R}]{\color{red}[\ol{\gamma := S}]} \to\\
- & \hspace*{0.5in} \relsem{\Gamma;\ol{\beta}{\color{red},\ol{\gamma}} \vdash (\mu
+      \beta}]}\Eq_\rho[\ol{\beta := R}]\to\\
+ & \hspace*{0.5in} \relsem{\Gamma;\ol{\beta} \vdash (\mu
   \phi.\lambda \ol{\alpha}.H)\ol{\beta}} \Eq_\rho[\ol{\beta:=
-    R}]{\color{red}[\ol{\gamma :=S}]}
+    R}]
 \end{array}\]
@@
 \setsem{\Gamma;\emptyset \,|\, \emptyset \vdash \tin_H :
-  \Nat^{\ol{\beta}{\color{red},\ol{\gamma}}} \, H[\phi := (\mu \phi.\lambda
+  \Nat^{\ol{\beta}} \, H[\phi := (\mu \phi.\lambda
     {\overline \alpha}.H){\overline \beta}][\ol{\alpha := \beta}]
@@
 $\underline{\Gamma; \emptyset~|~\emptyset \vdash \fold^F_H :
-  \Nat^\emptyset\;(\Nat^{\ol{\beta}{\color{red}, \ol{\gamma}}}\,H[\phi
+  \Nat^\emptyset\;(\Nat^{\ol{\beta}}\,H[\phi
     :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\;
-  (\Nat^{{\ol{\beta}{\color{red},\ol{\gamma}}} }\,(\mu \phi.\lambda \overline
+  (\Nat^{{\ol{\beta}} }\,(\mu \phi.\lambda \overline
   \alpha.H)\overline \beta \;F)}$ \; Since $\Phi$ is empty, to see
@@
   $$\setsem{ \Gamma; \emptyset~|~\emptyset \vdash \fold^F_H :
-    \Nat^\emptyset\;(\Nat^{\ol{\beta}{\color{red}, \ol{\gamma}}}\,H[\phi
+    \Nat^\emptyset\;(\Nat^{\ol{\beta}}\,H[\phi
       :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\;
-    (\Nat^{{\ol{\beta}{\color{red},\ol{\gamma}}} }\,(\mu \phi.\lambda \overline
+    (\Nat^{{\ol{\beta}} }\,(\mu \phi.\lambda \overline
     \alpha.H)\overline \beta \;F)}$$
@@
   $\setsem{\Gamma;\emptyset \vdash \emptyset}$ to \[\setsem{\Gamma;
-    \emptyset \vdash \Nat^\emptyset\;(\Nat^{\ol{\beta}{\color{red},
-      \ol{\gamma}}}\,H[\phi :=_{\ol{\beta}} F][\ol{\alpha :=
-        \beta}]\,F)\; (\Nat^{{\ol{\beta}{\color{red},\ol{\gamma}}} }\,(\mu
+    \emptyset \vdash \Nat^\emptyset\;(\Nat^{\ol{\beta}}\,H[\phi
+      :=_{\ol{\beta}} F][\ol{\alpha := 
+        \beta}]\,F)\; (\Nat^{{\ol{\beta}} }\,(\mu
     \phi.\lambda \overline \alpha.H)\overline \beta\,F)}\] we need only
@@
   \setsem{\Gamma;\emptyset \vdash \emptyset} \rho$, and all $\eta :
-  \setsem{\Gamma; \emptyset \vdash \Nat^{\ol{\beta}{\color{red},
-      \ol{\gamma}}}\,H[\phi :=_{\ol{\beta}} F][\ol{\alpha :=
+  \setsem{\Gamma; \emptyset \vdash \Nat^{\ol{\beta}}\,H[\phi
+      :=_{\ol{\beta}} F][\ol{\alpha := 
         \beta}]\,F} \rho$,
 \[ \setsem{\Gamma; \emptyset~|~\emptyset \vdash \fold^F_H :
-  \Nat^\emptyset\;(\Nat^{\ol{\beta}{\color{red}, \ol{\gamma}}}\,H[\phi
+  \Nat^\emptyset\;(\Nat^{\ol{\beta}}\,H[\phi
     :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\;
-  (\Nat^{{\ol{\beta}{\color{red},\ol{\gamma}}} }\,(\mu \phi.\lambda \overline
+  (\Nat^{{\ol{\beta}} }\,(\mu \phi.\lambda \overline
   \alpha.H)\overline \beta\,F)}\,\rho\,d\,\eta\] has type
-$\setsem{\Gamma; \emptyset \vdash \Nat^{{\ol{\beta}{\color{red},\ol{\gamma}}}
+$\setsem{\Gamma; \emptyset \vdash \Nat^{{\ol{\beta}}
   }\,(\mu \phi.\lambda \overline \alpha.H)\overline \beta\,F}\,\rho$,
-i.e., for any $\ol{B}$ {\color{red} and $\ol{C}$},
+i.e., for any $\ol{B}$,
 \[(\setsem{\Gamma; \emptyset~|~\emptyset \vdash \fold^F_H :
-  \Nat^\emptyset\;(\Nat^{\ol{\beta}{\color{red}, \ol{\gamma}}}\,H[\phi
-    :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\;
-  (\Nat^{{\ol{\beta}{\color{red},\ol{\gamma}}} }\,(\mu \phi.\lambda \overline
-  \alpha.H)\overline \beta\,F)}\,\rho\,d\,\eta)_{\ol{B}{\color{red}\,\ol{C}}}\] is a
-morphism from $\setsem{\Gamma; \ol{\beta}{\color{red},\ol{\gamma}} \vdash (\mu
-  \phi.\lambda \overline \alpha.H)\overline \beta}\rho[\ol{\beta :=
-    B}]{\color{red}[\ol{\gamma := C}]} \,=\,(\mu T^\set_{H,\rho{\color{red}[\ol{\gamma := C}]}})_{\ol{B}}$ to $\setsem{\Gamma; \ol{\beta}{\color{red},\ol{\gamma}} \vdash
-  F}\rho[\ol{\beta := B}]{\color{red}[\ol{\gamma := C}]}$.  To see this, note
-that $\eta$ is a natural transformation from
+  \Nat^\emptyset\;(\Nat^{\ol{\beta}}\,H[\phi :=_{\ol{\beta}}
+    F][\ol{\alpha := \beta}]\,F)\; (\Nat^{{\ol{\beta}} }\,(\mu
+  \phi.\lambda \overline \alpha.H)\overline
+  \beta\,F)}\,\rho\,d\,\eta)_{\ol{B}}\] is a morphism from
+$\setsem{\Gamma; \ol{\beta}\vdash (\mu \phi.\lambda \overline
+  \alpha.H)\overline \beta}\rho[\ol{\beta := B}] \,=\,(\mu
+T^\set_{H,\rho})_{\ol{B}}$ to $\setsem{\Gamma; \ol{\beta} \vdash
+  F}\rho[\ol{\beta := B}]$.  To see this, note that $\eta$ is a
+natural transformation from
 \[\begin{array}{ll}
- & \lambda \ol{B}{\color{red}\,\ol{C}}.\,\setsem{\Gamma; \ol{\beta}{\color{red},\ol{\gamma}}
+ & \lambda \ol{B}.\,\setsem{\Gamma; \ol{\beta}
   \vdash H[\phi := F][\ol{\alpha := \beta}]}\rho[\ol{\beta :=
-    B}]{\color{red}[\ol{\gamma := C}]}\\
-= & \lambda \ol{B}{\color{red}\,\ol{C}}.\,T^\set_{H,\rho{\color{red}[\ol{\gamma:=
-     C}]}}\,(\lambda \ol{A}. \, \setsem{\Gamma;\ol{\beta}{\color{red},\ol{\gamma}} 
-  \vdash F}\rho[\ol{\beta := A}]{\color{red}[\ol{\gamma := C}]}) \, \ol{B}
+    B}]\\
+= & \lambda \ol{B}.\,T^\set_{H,\rho}\,(\lambda \ol{A}. \,
+\setsem{\Gamma;\ol{\beta}
+  \vdash F}\rho[\ol{\beta := A}]) \, \ol{B}
 \end{array}\]
@@
 \[\begin{array}{ll}
- & \lambda \ol{B}{\color{red}\,\ol{C}}.\,(\lambda
-\ol{A}.\,\setsem{\Gamma;\ol{\beta}{\color{red},\ol{\gamma}} \vdash F}\rho[\ol{\beta
-    := A}]{\color{red}[\ol{\gamma := C}]}) \ol{B}\\
+ & \lambda \ol{B}.\,(\lambda
+\ol{A}.\,\setsem{\Gamma;\ol{\beta}\vdash F}\rho[\ol{\beta
+    := A}]) \ol{B}\\
 = & \lambda
-\ol{B}{\color{red}\,\ol{C}}.\,\setsem{\Gamma;\ol{\beta}{\color{red},\ol{\gamma}} \vdash
-  F}\rho[\ol{\beta := B}]{\color{red}[\ol{\gamma := C}]}
-\end{array}\]
-and thus for each $\ol{B}$ {\color{red} and $\ol{C}$},
+\ol{B}.\,\setsem{\Gamma;\ol{\beta} \vdash
+  F}\rho[\ol{\beta := B}]
+\end{array}\]
+and thus for each $\ol{B}$,
 \[(\setsem{\Gamma; \emptyset~|~\emptyset \vdash \fold^F_H :
-  \Nat^\emptyset\;(\Nat^{\ol{\beta}{\color{red}, \ol{\gamma}}}\,H[\phi
-    :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\;
-  (\Nat^{{\ol{\beta}{\color{red},\ol{\gamma}}} }\,(\mu \phi.\lambda
-  \overline \alpha.H)\overline
-  \beta\,F)}\,\rho\,d\,\eta)_{\ol{B}{\color{red}\,\ol{C}}}\] is a
-morphism from $\setsem{\Gamma; \ol{\beta}{\color{red},\ol{\gamma}}
-  \vdash (\mu \phi.\lambda \overline \alpha.H)\overline
-  \beta}\rho[\ol{\beta := B}]{\color{red}[\ol{\gamma := C}]}\,=\, (\mu
-T^\set_{H,\rho{\color{red}[\ol{\gamma := C}]}})_{\ol{B}}$ to
-$\setsem{\Gamma;\ol{\beta}{\color{red},\ol{\gamma}} \vdash
-  F}\rho[\ol{\beta := B}]{\color{red}[\ol{\gamma := C}]}$.  To see
-that this family of morphisms is natural in $\ol{B}$ {\color{red} and
-  $\ol{C}$}, we observe that the following diagram commutes for all
-$\ol{f : B \to B'}$ {\color{red} and $\ol{g : C \to C'}$}: {\tiny
+  \Nat^\emptyset\;(\Nat^{\ol{\beta}}\,H[\phi :=_{\ol{\beta}}
+    F][\ol{\alpha := \beta}]\,F)\; (\Nat^{{\ol{\beta}} }\,(\mu
+  \phi.\lambda \overline \alpha.H)\overline
+  \beta\,F)}\,\rho\,d\,\eta)_{\ol{B}}\] is a morphism from
+$\setsem{\Gamma; \ol{\beta} \vdash (\mu \phi.\lambda \overline
+  \alpha.H)\overline \beta}\rho[\ol{\beta := B}]\,=\, (\mu
+T^\set_{H,\rho})_{\ol{B}}$ to $\setsem{\Gamma;\ol{\beta}\vdash
+  F}\rho[\ol{\beta := B}]$.  To see that this family of morphisms is
+natural in $\ol{B}$, we observe that the following diagram commutes
+for all $\ol{f : B \to B'}$':
+{\tiny
 \[\hspace*{-0.3in}\begin{tikzcd}[column sep=2.5in, row sep=0.75in]
-%(\mu T^\set_{H,\rho[\ol{\gamma := C}]})\, \ol{B}
-%\ar[d, "{(\mu T^\set_{H,\id_\rho[\ol{\gamma := g}]}) \, \ol{B}}"'] \ar[r, 
-%  "{(\mathit{fold}_{T^\set_{H,\rho[\ol{\gamma := C}]}}\,(\lambda
-%    \ol{A}.\,\eta_{\ol{A}\,\ol{C}}))_{\ol{B}}}"] 
-%& \setsem{\Gamma; \ol{\beta},\ol{\gamma} \vdash F}\rho[\ol{\gamma :=
-%    C}][\ol{\beta := B}]\ar[d, "{\setsem{\Gamma;
-%      \ol{\beta},\ol{\gamma} \vdash F}\id_\rho[\ol{\gamma := 
-%    g}][\ol{\beta := \id_B}]}"]\\
-(\mu T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}})\, \ol{B} 
-\ar[r,"{(\mathit{fold}_{T^\set_{H,\rho{\color{red}[\ol{\gamma :=
-            C'}]}}}\,{\color{red}(\lambda \ol{A}.\,}\eta{\color{red}_{\ol{A}\,\ol{C'}})})_{\ol{B}}\,}" ] 
-\ar[d, "{(\mu T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}}) \, \ol{f}}"'] & 
- \setsem{\Gamma; \ol{\beta}{\color{red},\ol{\gamma}} \vdash F}\rho{\color{red}[\ol{\gamma :=
-    C'}]}[\ol{\beta := B}]\ar[d, "{\setsem{\Gamma;
-      \ol{\beta}{\color{red},\ol{\gamma}} \vdash F}\id_\rho{\color{red}[\ol{\gamma := 
-    \id_{C'}}]}[\ol{\beta := f}]}"]\\
- (\mu T^\set_{H,\rho{\color{red}[\ol{\gamma := C'}]}})\, \ol{B'} 
-\ar[r,"{(\mathit{fold}_{T^\set_{H,\rho{\color{red}[\ol{\gamma :=
-            C'}]}}}{\color{red}\,(\lambda \ol{A}.\,}\eta{\color{red}_{\ol{A}\,\ol{C'}})})_{\ol{B'}}\,}"
+(\mu T^\set_{H,\rho})\, \ol{B} 
+\ar[r,"{(\mathit{fold}_{T^\set_{H,\rho}}\,\eta)_{\ol{B}}\,}" ] 
+\ar[d, "{(\mu T^\set_{H,\rho}) \, \ol{f}}"'] & 
+ \setsem{\Gamma; \ol{\beta} \vdash F}\rho[\ol{\beta := B}]\ar[d,
+   "{\setsem{\Gamma; 
+      \ol{\beta} \vdash F}\id_\rho[\ol{\beta := f}]}"]\\
+ (\mu T^\set_{H,\rho})\, \ol{B'} 
+\ar[r,"{(\mathit{fold}_{T^\set_{H,\rho}}\eta)_{\ol{B'}}\,}" 
 ] &  
- \setsem{\Gamma; \ol{\beta}{\color{red},\ol{\gamma}} \vdash F}\rho{\color{red}[\ol{\gamma :=
-    C'}]}[\ol{\beta := B'}]
+ \setsem{\Gamma; \ol{\beta} \vdash F}\rho[\ol{\beta := B'}]
 \end{tikzcd}\]}
-{\color{red} Indeed,}
-by naturality of $\mathit{fold}_{T^\set_{H,\rho[{\color{red}\ol{\gamma
-          := C'}]}}}{\color{red}\,(\lambda
-\ol{A}.\,}\eta{\color{red}_{\ol{A}\,\ol{C'}}})$.
-{\color{red} ensures that the bottom diagram commutes. 
-To see that the top one commutes we first observe that, given a
-natural transformation $\Theta : H \to K : [\set^k, \set] \to [\set^k,
-  \set]$, the fixpoint natural transformation $\mu \Theta : \mu H \to
-\mu K : \set^k \to \set$ is defined to be
-$\mathit{fold}_{H}(\Theta\,(\mu K) \circ \mathit{in}_{K})$, i.e., the
-unique morphism making the following diagram commute:\label{page:dia1}
-{\footnotesize
-\[\begin{tikzcd}[column sep = large]
-H(\mu H)
-	\ar[dd, "{\mathit{in}_H}"']
-	\ar[r, "{H(\mu \Theta)}"]
-& H(\mu K)
-	\ar[d, "{\Theta (\mu K)}"] \\
-& K(\mu K)
-	\ar[d, "{\mathit{in}_K}"] \\
-\mu H
-	\ar[r, "{\mu \Theta}"']
-& \mu K
-\end{tikzcd}\]}
-Taking $\Theta = T^{\set}_{H,f}: T^{\set}_{H,\rho} \to
-T^{\set}_{H,\rho'}$ thus gives that, for any $f : \rho \to \rho'$ in
-$\setenv$,
-\begin{equation}\label{eq:mu-sigma-def}
-\mathit{in}_{T^{\set}_{H,\rho'}} \circ 
-T^{\set}_{H,f} (\mu T^{\set}_{H,\rho'}) \circ 
-T^{\set}_{H,\rho}(\mu T^{\set}_{H,f}) \,=\, 
-\mu T^{\set}_{H,f} \circ \mathit{in}_{T^{\set}_{H,\rho}}
-\end{equation}}
-{\color{red} Next, note that the action of the functor
-\[\lambda \ol{B}.{\color{red} \lambda \ol{C}.} \setsem{\Gamma; \ol{\beta}{\color{red},
-  \ol{\gamma}} \vdash H[\phi := F][\ol{\alpha := \beta}]}\rho
-        [\ol{\beta := B}] {\color{red}[\ol{\gamma := C}]}\]
-on the morphisms $\ol{f : B \to B'}{\color{red}, \ol{g : C \to C'}}$ is given by
-\[\begin{array}{ll}
- & \setsem{\Gamma; \ol{\beta{\color{red}, \ol{\gamma}}} \vdash H[\phi :=
-      F][\ol{\alpha := \beta}]} \id_{\rho} [\ol{\beta := f}]
-           {\color{red}[\ol{\gamma := g}]}\\
-= & \setsem{\Gamma; \phi, \ol{\alpha}{\color{red}, \ol{\gamma}} \vdash H}
-\id_{\rho}[\ol{\alpha := f}] {\color{red}[\ol{\gamma := g}]}[\phi := \lambda
-  \ol{A}. \setsem{\Gamma; \ol{\beta}{\color{red}, \ol{\gamma}} \vdash F} \id_{\rho
-    [\ol{\beta := A}]}{\color{red} [\ol{\gamma := g}]}] \\ 
-=& \setsem{\Gamma; \phi, \ol{\alpha}{\color{red}, \ol{\gamma}} \vdash H} \id_{\rho
-  {\color{red}[\ol{\gamma := C'}]}[\phi := \lambda \ol{A}. \setsem{\Gamma;
-      \ol{\beta} {\color{red}, \ol{\gamma}} \vdash F} \rho [\ol{\beta := A}]
-   {\color{red} [\ol{\gamma := C'}]}]} [\ol{\alpha := f}] \\
-&\hspace{3em} \circ \setsem{\Gamma; \phi, \ol{\alpha}{\color{red}, \ol{\gamma}}
-  \vdash H} \id_{\rho [\ol{\alpha := B}][\phi := \lambda
-    \ol{A}. \setsem{\Gamma; \ol{\beta}{\color{red}, \ol{\gamma}} \vdash F} \rho
-       [\ol{\beta := A}]{\color{red} [\ol{\gamma := C'}]}]} {\color{red}[\ol{\gamma := g}]} \\
-&\hspace{3em} \circ \setsem{\Gamma; \phi, \ol{\alpha} {\color{red}, \ol{\gamma}}
-  \vdash H} \id_{\rho [\ol{\alpha := B}] {\color{red}[\ol{\gamma := C}]}} [\phi :=
-  \lambda \ol{A}. \setsem{\Gamma; \ol{\beta}{\color{red}, \ol{\gamma}} \vdash F}
-  \id_{\rho [\ol{\beta := A}]}{\color{red} [\ol{\gamma := g}]}] \\
-= & T^{\set}_{H,\rho {\color{red}[\ol{\gamma := C'}]}} (\lambda
-\ol{A}. \setsem{\Gamma; \ol{\beta} {\color{red}, \ol{\gamma}} \vdash F} \rho
-   [\ol{\beta := A}]{\color{red} [\ol{\gamma := C'}]}) \ol{f} \\
-&\hspace{3em} \circ \big( T^{\set}_{H,\id_{\rho}{\color{red}[\ol{\gamma := g}]}}
-   (\lambda \ol{A}. \setsem{\Gamma; \ol{\beta} {\color{red}, \ol{\gamma}} \vdash F}
-   \rho [\ol{\beta := A}] {\color{red}[\ol{\gamma := C'}]}) \big)_{\ol{B}} \\
-&\hspace{3em} \circ \big( T^{\set}_{H,\rho {\color{red}[\ol{\gamma := C}]}} (\lambda
-   \ol{A}. \setsem{\Gamma; \ol{\beta}{\color{red}, \ol{\gamma} }\vdash F} \id_{\rho
-     [\ol{\beta := A}]} {\color{red}[\ol{\gamma := g}]}) \big)_{\ol{B}}
-\end{array}\]
-So if $\eta$ is a natural transformation such that $\eta_{\ol B {\color{red}, \ol
-  C}}$ has type
-\[
-\setsem{\Gamma; \ol{\beta},
-  {\color{red}\ol{\gamma}} \vdash H[\phi := F][\ol{\alpha := \beta}]}\rho
-[\ol{\beta := B}] {\color{red}[\ol{\gamma := C}]}
-\to
-\setsem{\Gamma; \ol{\beta},
-  {\color{red} \ol{\gamma}} \vdash F}\rho [\ol{\beta := B}]
-   {\color{red}    [\ol{\gamma := C}]}\] 
-then, by naturality,
-\[\begin{array}{ll}
- & \setsem{\Gamma; \ol{\beta}{\color{red}, \ol{\gamma}} \vdash F} \id_{\rho}
-       [\ol{\beta := f}] {\color{red}[\ol{\gamma := g}]} \circ \eta_{\ol{B}{\color{red},
-         \ol{C}}} \\ 
-= & \eta_{\ol{B'}{\color{red}, \ol{C'}}} \circ T^{\set}_{H,\rho{\color{red} [\ol{\gamma := C'}]}}
-(\lambda \ol{A}. \setsem{\Gamma; \ol{\beta} {\color{red}, \ol{\gamma}} \vdash F}
-\rho [\ol{\beta := A}]{\color{red} [\ol{\gamma := C'}]}) \ol{f} \\
-& \circ \big( T^{\set}_{H,\id_{\rho}{\color{red}[\ol{\gamma := g}]}} (\lambda
-\ol{A}. \setsem{\Gamma; \ol{\beta}{\color{red}, \ol{\gamma}} \vdash F} \rho
-   [\ol{\beta := A}]{\color{red} [\ol{\gamma := C'}]}) \big)_{\ol{B}} \\
-& \circ \big( T^{\set}_{H,\rho {\color{red}[\ol{\gamma := C}]}} (\lambda
-   \ol{A}. \setsem{\Gamma; \ol{\beta}{\color{red}, \ol{\gamma}} \vdash F} \id_{\rho
-     [\ol{\beta := A}]} {\color{red}[\ol{\gamma := g}]}) \big)_{\ol{B}}
-\end{array}\]
-As a special case when $\ol{f = \id_B}$ we have
-\begin{equation}\label{eq:T-sigma-functor}
-\begin{split}
-  & \lambda \ol{B}. \setsem{\Gamma; \ol{\beta} {\color{red}, \ol{\gamma}} \vdash F}
-\id_{\rho [\ol{\beta := B}]} {\color{red}[\ol{\gamma := g}]} \circ \lambda
-\ol{B}.\eta_{\ol{B}{\color{red}, \ol{C}}} \\  
-=\;\;\; & \lambda \ol{B}.\eta_{\ol{B}{\color{red}, \ol{C'}}} \circ
-T^{\set}_{H,\id_{\rho}{\color{red}[\ol{\gamma := g}]}} (\lambda
-\ol{A}. \setsem{\Gamma; \ol{\beta}{\color{red}, \ol{\gamma}} \vdash F} \rho
-   [\ol{\beta := A}]{\color{red} [\ol{\gamma := C'}]}) \\
- & \hspace*{0.5in} \circ
-   T^{\set}_{H,\rho {\color{red}[\ol{\gamma := C}]}} (\lambda \ol{A}. \setsem{\Gamma;
-     \ol{\beta}{\color{red}, \ol{\gamma}} \vdash F} \id_{\rho [\ol{\beta := A}]}
- {\color{red}  [\ol{\gamma := g}]}) 
-\end{split}
-\end{equation}
-Finally, to see that the top diagram in the diagram on
-page~\pageref{page:dia1} commutes we first note that functoriality of
-$T^{\set}_{H,\rho [\ol{\gamma := C}]}$, naturality of
-$T^{\set}_{H,\id_{\rho}[\ol{\gamma := g}]}$, the universal property of
-$\mathit{fold}_{T^{\set}_{H,\rho [\ol{\gamma := C'}]}} (\lambda
-\ol{A}. \eta_{\ol{A}, \ol{C'}})$ and Equation~\ref{eq:mu-sigma-def}
-ensure that the following diagram commutes: {\footnotesize
-\begin{equation}\label{eq:one-side}
-\hspace*{-0.7in}  \begin{tikzcd}[column sep = huge, row sep = huge]
-T^{\set}_{H,\rho [\ol{\gamma := C}]} (\mu T^{\set}_{H,\rho [\ol{\gamma :=
-      C}]}) \ar[rr, "{T^{\set}_{H,\rho [\ol{\gamma := C}]} (
-    \mathit{fold}_{T^{\set}_{H,\rho [\ol{\gamma := C'}]}} (\lambda
-    \ol{A}. \eta_{\ol{A}, \ol{C'}}) \circ \mu
-    T^{\set}_{H,\id_{\rho}[\ol{\gamma := g}]} )}"] \ar[dd,
-    "{\mathit{in}_{T^{\set}_{H,\rho [\ol{\gamma := C}]}}}"']
-&& T^{\set}_{H,\rho [\ol{\gamma := C}]} (\lambda \ol{B}. \setsem{\Gamma;
-  \ol{\beta}, \ol{\gamma} \vdash F} \rho [\ol{\beta := B}] [\ol{\gamma
-    := C'}]) \ar[d, "{ T^{\set}_{H,\id_{\rho}[\ol{\gamma := g}]}
-    (\lambda \ol{B}. \setsem{\Gamma; \ol{\beta}, \ol{\gamma} \vdash F}
-    \rho [\ol{\beta := B}] [\ol{\gamma := C'}]) }" description] \\
-&& T^{\set}_{H,\rho [\ol{\gamma := C'}]} (\lambda
-\ol{B}. \setsem{\Gamma; \ol{\beta}, \ol{\gamma} \vdash F} \rho
-   [\ol{\beta := B}][\ol{\gamma := C'}]) \ar[d, "{ \lambda \ol{A}. \eta_{\ol{A}, \ol{C'}}
-     }" description] \\
-\mu T^{\set}_{H,\rho [\ol{\gamma := C}]}\ar[r, "{\mu
-    T^{\set}_{H,\id_{\rho}[\ol{\gamma := g}]}}"'] 
-&\mu T^{\set}_{H,\rho [\ol{\gamma := C'}]} \ar[r, bend right = 10,
-  "{\mathit{fold}_{T^{\set}_{H,\rho [\ol{\gamma := C'}]}} (\lambda
-    \ol{A}. \eta_{\ol{A}, \ol{C'}})}"']
-& \lambda \ol{B}. \setsem{\Gamma; \ol{\beta}, \ol{\gamma} \vdash F}
-\rho [\ol{\beta := B}] [\ol{\gamma := C'}]
-  \end{tikzcd}
-  \end{equation}}
-Next, we note that functoriality of $T^{\set}_{H,\rho [\ol{\gamma :=
-      C}]}$, Equation~\ref{eq:T-sigma-functor}, and the universal
-property of $\mathit{fold}_{T^{\set}_{H,\rho [\ol{\gamma := C}]}} (\lambda
-\ol{A}. \eta_{\ol{A}, \ol{C}})$ ensure that the following diagram
-commutes: {\footnotesize
-\begin{equation}\label{eq:other-side}
-\hspace*{-1in}\begin{tikzcd}[row sep = huge]
-T^{\set}_{H,\rho [\ol{\gamma := C}]} (\mu T^{\set}_{H,\rho [\ol{\gamma :=
-      C}]}) \ar[rr, "{T^{\set}_{H,\rho [\ol{\gamma := C}]} (
-    \lambda \ol{B}.\setsem{\Gamma; \ol{\beta}, \ol{\gamma} \vdash F} \id_{\rho
-      [\ol{\beta := B}]} [\ol{\gamma := g}] \circ
-    \mathit{fold}_{T^{\set}_{H,\rho [\ol{\gamma := C}]}} (\lambda
-    \ol{A}. \eta_{\ol{A}, \ol{C}}) )}"] \ar[dd, "{\mathit{in}_{T^{\set}_{H,\rho
-        [\ol{\gamma := C}]}}}"']
-&& T^{\set}_{H,\rho [\ol{\gamma := C}]} (\lambda \ol{B}.\setsem{\Gamma; \ol{\beta},
-  \ol{\gamma} \vdash F} \rho [\ol{\beta := B}] [\ol{\gamma := C'}])
-\ar[d, "{ T^{\set}_{H,\id_{\rho}[\ol{\gamma := g}]}
-    (\lambda \ol{B}.\setsem{\Gamma; \ol{\beta}, \ol{\gamma} \vdash F} \rho [\ol{\beta
-        := B}] [\ol{\gamma := C'}]) }" description]\\
-&& T^{\set}_{H,\rho [\ol{\gamma := C'}]} (\lambda \ol{B}.\setsem{\Gamma; \ol{\beta},
-  \ol{\gamma} \vdash F} \rho [\ol{\beta := B}][\ol{\gamma := C'}]) \ar[d, "{ \lambda
-    \ol{A}. \eta_{\ol{A}, \ol{C'}} }" description] \\
-\mu T^{\set}_{H,\rho [\ol{\gamma := C}]} \ar[r, bend right = 10,
-  "{\mathit{fold}_{T^{\set}_{H,\rho [\ol{\gamma := C}]}} (\lambda
-    \ol{A}. \eta_{\ol{A}, \ol{C}})}"'] & \lambda \ol{B}.\setsem{\Gamma; \ol{\beta},
-  \ol{\gamma} \vdash F} \rho [\ol{\beta := B}] [\ol{\gamma := C}]
-\ar[r, bend right = 10, "{\lambda \ol{B}.\setsem{\Gamma; \ol{\beta}, \ol{\gamma}
-      \vdash F} \id_{\rho [\ol{\beta := B}]} [\ol{\gamma := g}]}"'] &
-\lambda \ol{B}.\setsem{\Gamma; \ol{\beta}, \ol{\gamma} \vdash F} \rho [\ol{\beta :=
-    B}] [\ol{\gamma := C'}]
-\end{tikzcd}
-\end{equation}}
-Combining the equations entailed by~\ref{eq:one-side}
-and~\ref{eq:other-side}, we get that
-the top diagram in the diagram on
-  page~\pageref{page:dia1} commutes, as desired.}
+by naturality of $\mathit{fold}_{T^\set_{H,\rho}}\eta$.
 To see that, for all $\rho : \setenv$, $d \in
 \setsem{\Gamma; \emptyset \vdash \emptyset}\rho$, and $\eta :
-\setsem{\Gamma; \emptyset \vdash \Nat^{\ol{\beta}{\color{red},
-    \ol{\gamma}}}\,H[\phi :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F}
+\setsem{\Gamma; \emptyset \vdash \Nat^{\ol{\beta}}\,H[\phi :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F}
 \rho$,
 \[\setsem{\Gamma; \emptyset~|~\emptyset
-  \vdash \fold^F_H : \Nat^\emptyset\;(\Nat^{\ol{\beta}{\color{red},
-    \ol{\gamma}}}\,H[\phi :=_{\ol{\beta}} F][\ol{\alpha :=
-      \beta}]\,F)\; (\Nat^{{\ol{\beta}{\color{red},\ol{\gamma}}} }\,(\mu
+  \vdash \fold^F_H : \Nat^\emptyset\;(\Nat^{\ol{\beta}}\,H[\phi :=_{\ol{\beta}} F][\ol{\alpha :=
+      \beta}]\,F)\; (\Nat^{{\ol{\beta}} }\,(\mu
   \phi.\lambda \overline \alpha.H)\overline
@@
 needed for it to be in $\setsem{\Gamma;\emptyset \vdash
-  \Nat^{{\ol{\beta}{\color{red},\ol{\gamma}}} }\,(\mu \phi.\lambda \overline
-  \alpha.H)\overline \beta\;F}\,\rho$, let $\ol{R : \rel(B,B')}$
-{\color{red} and $\ol{S : \rel(C,C')}$}.  Since $\eta$ satisfies the
+  \Nat^{{\ol{\beta}} }\,(\mu \phi.\lambda \overline
+  \alpha.H)\overline \beta\;F}\,\rho$, let $\ol{R : \rel(B,B')}$.
+Since $\eta$ satisfies the 
 additional condition needed for it to be in $\setsem{\Gamma; \emptyset
-  \vdash \Nat^{\ol{\beta}{\color{red}, \ol{\gamma}}}\,(H[\phi := F][\ol{\alpha :=
+  \vdash \Nat^{\ol{\beta}}\,(H[\phi := F][\ol{\alpha :=
       \beta}])\,F} \rho$,
 \[\begin{array}{ll}
- & (\,(\mathit{fold}_{T^\set_{H,\rho{\color{red}[\ol{\gamma :=
-        C}]}}}{\color{red}\,(\lambda \ol{A}.\,}\eta{\color{red}_{\ol{A}\,\ol{C}})})_{\ol{B}},\,
-(\mathit{fold}_{T^\set_{H,\rho{\color{red}[\ol{\gamma :=
-        C'}]}}}{\color{red}\,(\lambda \ol{A}.}\eta{\color{red}_{\ol{A}\,\ol{C'}})})_{\ol{B'}}\,) 
+ & (\,(\mathit{fold}_{T^\set_{H,\rho}}\eta)_{\ol{B}},\,
+(\mathit{fold}_{T^\set_{H,\rho}}\eta)_{\ol{B'}}\,) 
 \end{array}\]
@@
 \[\begin{array}{ll}
-  & (\mu T_{H,\Eq_\rho{\color{red}[\ol{\gamma := S}]}}) \,\ol{R} \to
-\relsem{\Gamma;{\color{red}\ol{\gamma},}\ol{\beta} \vdash F}\Eq_\rho{\color{red}[\ol{\gamma := 
-    S}]}[\ol{\beta:= R}]\\ 
-= & (\mu T_{H,\Eq_\rho{\color{red}[\ol{\gamma :=
-      S}]}})\,\ol{\relsem{\Gamma;{\color{red}\ol{\gamma},}\ol{\beta} 
-  \vdash \beta}\Eq_\rho{\color{red}[\ol{\gamma := S}]}[\ol{\beta := R}]} \\ 
-  &\hspace{1.5in} \to
-\relsem{\Gamma;{\color{red}\ol{\gamma},}\ol{\beta} \vdash F}\Eq_\rho{\color{red}[\ol{\gamma := 
-    S}]}[\ol{\beta:= R}]\\ 
-= & \relsem{\Gamma; {\color{red}\ol{\gamma},}\ol{\beta} \vdash (\mu \phi. \lambda
-  \ol{\alpha}. H)\ol{\beta}}\Eq_\rho{\color{red}[\ol{\gamma := S}]}[\ol{\beta := R}] \\ 
-  &\hspace{1.5in}\to
-\relsem{\Gamma;{\color{red}\ol{\gamma},}\ol{\beta} \vdash F}\Eq_\rho{\color{red}[\ol{\gamma := 
-    S}]}[\ol{\beta:= R}]
+  & (\mu T_{H,\Eq_\rho}) \,\ol{R} \to
+\relsem{\Gamma;\ol{\beta} \vdash F}\Eq_\rho[\ol{\beta:= R}]\\ 
+= & (\mu T_{H,\Eq_\rho})\,\ol{\relsem{\Gamma;\ol{\beta} 
+  \vdash \beta}\Eq_\rho[\ol{\beta := R}]} \to
+\relsem{\Gamma;\ol{\beta} \vdash F}\Eq_\rho[\ol{\beta:= R}]\\ 
+= & \relsem{\Gamma; \ol{\beta} \vdash (\mu \phi. \lambda
+  \ol{\alpha}. H)\ol{\beta}}\Eq_\rho[\ol{\beta := R}] \to
+\relsem{\Gamma;\ol{\beta} \vdash F}\Eq_\rho[\ol{\beta:= R}]
 \end{array}\]
@@
 \[\setsem{\Gamma; \emptyset~|~\emptyset \vdash \fold^F_H :
-  \Nat^\emptyset\;(\Nat^{\ol{\beta}{\color{red}, \ol{\gamma}}}\,H[\phi
+  \Nat^\emptyset\;(\Nat^{\ol{\beta}}\,H[\phi
     :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\;
-  (\Nat^{{\ol{\beta}{\color{red},\ol{\gamma}}} }\,(\mu \phi.\lambda
+  (\Nat^{{\ol{\beta}} }\,(\mu \phi.\lambda
   \overline \alpha.H)\overline \beta \;F)}\] in $\rho$ is trivial.

### 2020-12-20 free-theorems-nesteds/20-mscs.tex
added GADT section to be revised

@@
   (\Nat^{{\ol{\beta}} }\,(\mu \phi.\lambda
-  \overline \alpha.H)\overline \beta \;F)}\] in $\rho$ is trivial.
+  \overline \alpha.H)\overline \beta \;F)}\] in $\rho$ is trivial.\qed
 \end{itemize}
-\qed
+
+\vspace*{0.1in}
 
@@
 
+Generalized algebraic datatypes (GADTs) generalize nested types by...
+
+For example, ...
+
+A natural next step in this line of work is to extend our
+parametricity results to GADTs.  It was shown in~\cite{jg08} that (the
+data of) GADTs can be represented using object-level left Kan
+extensions (over discrete categories), and further shown
+in~\cite{jp19} that adding a carefully designed left Kan extension
+construct to a calculus supporting primitive nested types preserves
+the cocontinuity needed for primitive GADTs to have well-defined,
+properly functorial interpretations.
+
+This suggests adding GADTs to our calculus by extending
+Definition~\ref{def:wftypes} with such object-level constructs, and
+extending Figure~\ref{fig:terms} with categorically inspired
+constructs to introduce and eliminate terms of these types, along the
+same entirely standard lines as we did there for products, coproducts,
+and fixpoints.  In this section we outline how our model from
+Section~\ref{sec:term-interp} can be extended to a parametric model
+when some classes of GADTs are incorporated as primitives in this
+way. However, as we argue in~???? below, we suspect this is not
+possible in general.
+
+HERE!!
+
+To show when this can be done, as well as to give a sense of what goes
+wrong in the general case, we begin by recalling that GADTs are
+naturally interpreted as initial algebras of left Kan extensions, as
+suggested in discrete~\cite{jg08} and
+
+Functorial semantics in~\cite{jp19}.
+
+introducing a new
+type, $\Lan$, to represent left Kan extensions.  Indeed,
+
+To begin with, we add the following type formation rule for $\Lan$
+types to our type system.
+\[
+\AXC{$\Gamma; \Phi,\ol\alpha^0 \vdash F$}
+\AXC{$\ol{\Gamma;\ol\alpha^0 \vdash K}$}
+\AXC{$\ol{\Gamma;\Phi \vdash A}$}
+\TIC{$\Gamma;\Phi \vdash (\Lan^{\ol{\alpha^0}}_{\ol K} F) \ol A$}
+\DisplayProof
+\]
+Notice that the length of $\ol K$ coincides with that of $\ol A$.
+Introducing the new type also requires the following additional clause in the inductive definition of higher-order substitution.
+\[
+((\Lan^{\ol{\alpha}}_{\ol K}\,F) \ol A)[\phi :=_{\ol{\beta}} G]
+  = (\Lan^{\ol\alpha}_{\ol K}\,F[\phi :=_{\ol{\beta}} G])\, \ol{A[\phi :=_{\ol{\beta}} G]}
+\]
+
+Before we extend the term calculus with the term introduction and elimination rules for $\Lan$ types,
+it is necessary to generalize $\Nat$ types to bind type constructor variables of arbitrary arity.
+So, the formation rule for $\Nat$ types will be modified as
+\[
+\AXC{$\Gamma;\Phi \vdash F$}
+\AXC{$\Gamma;\Phi  \vdash G$}
+\BIC{$\Gamma;\emptyset \vdash \Nat^\Phi F \,G$}
+\DisplayProof
+\]
+and the term introduction and elimination rules will be
+\[
+\AXC{$\Gamma; \Phi \vdash F$}
+\AXC{$\Gamma; \Phi \vdash G$}
+\AXC{$\Gamma; \Phi \,|\, \Delta, x : F \vdash t: G$} 
+\TIC{$\Gamma; \emptyset
+  \,|\, \Delta \vdash L_\Phi x.t : \Nat^\Phi \,F \,G$}
+\DisplayProof
+\]
+and
+\[
+\AXC{$\ol{\Gamma;\Phi,\ol\beta \vdash K}$}
+\AXC{$\Gamma; \emptyset
+  \,|\, \Delta \vdash t : \Nat^{\ol\psi} \,F \,G$}
+\AXC{$\Gamma;\Phi \,|\, \Delta \vdash s: F[\overline{\psi :=_{\ol\beta} K}]$}
+\TIC{$\Gamma;\Phi\,|\, \Delta \vdash t_{\ol K} s:
+  G[\overline{\psi :=_{\ol\beta} K}]$}
+\DisplayProof
+\]
+respectively.
+It is necessary, of course, to suitably extend the semantic interpretations of $\Nat$ types and terms.
+The type constructor variables bound by $\Nat$ will be interpreted as either functors
+or relation transformers.
+The set and relational interpretations for $\Nat$ types are defined as follows.
+\begin{align*}
+\setsem{\Gamma; \emptyset \vdash \Nat^\Phi\,F\,G}\rho
+&= \{\eta : \lambda \ol{K}. \,\setsem{\Gamma; \Phi \vdash F}\rho[\Phi := \ol K] 
+  \Rightarrow \lambda \ol{K}.\,\setsem{\Gamma; \Phi \vdash G}\rho[\Phi := \ol K]~|~\\ 
+&\hspace{0.3in} \forall \ol{K = (K^1, K^2, K^*) : RT_k}.\\
+&\hspace{0.4in} (\eta_{\overline{K^1}}, \eta_{\overline{K^2}})
+: \setsem{\Gamma; \Phi \vdash F}\Eq_{\rho}[\Phi := \ol K]
+\rightarrow \setsem{\Gamma; \Phi \vdash G}\Eq_{\rho}[\Phi := \ol K] \} \\
+%%%
+\relsem{\Gamma; \emptyset \vdash \Nat^\Phi \,F\,G}\rho
+&= \{\eta : \lambda \ol{K}.\,\relsem{\Gamma; \Phi \vdash F}\rho[\ol{\Phi := K}]
+\Rightarrow \lambda \ol{K}. \,\relsem{\Gamma; \Phi \vdash G}\rho[\ol{\phi := K}]\}\\
+&=
+  \{(\eta_1,\eta_2) \in \setsem{\Gamma; \emptyset
+    \vdash \Nat^\Phi
+    \,F\,G} (\pi_1 \rho) \times \setsem{ 
+    \Gamma;\emptyset
+    \vdash \Nat^\Phi \,F\,G} (\pi_2
+  \rho)~|~\\
+& \hspace{0.3in} \forall \ol{K = (K^1, K^2, K^*) : RT_{k}}.\\
+& \hspace{0.4in} ((\eta_1)_{\ol{K^1}},(\eta_2)_{\ol{K^2}}) \in
+  (\relsem{\Gamma; \Phi \vdash G}\rho[\ol{\Phi := K}])^{\relsem{\Gamma;\Phi\vdash F}\rho[\ol{\Phi := K}]} \}\\  
+\end{align*}
+
+We can now give the term introduction and elimination rules for $\Lan$, which are
+\[
+\AXC{$\Gamma; \Phi,\ol{\alpha}\vdash F$}
+\AXC{$\ol{\Gamma; \ol{\alpha} \vdash K}$}
+\BIC{$\Gamma;\emptyset~|~\Delta \vdash \int_{\ol K,F} : \Nat^{\Phi}\, F\, (\Lan^{\ol\alpha}_{\ol{K}}\,  F) \ol{K}$}
+\DisplayProof
+\]
+and
+\[
+\AXC{$\Gamma;\emptyset~|~\Delta \vdash \eta : \Nat^{\Phi,\ol\alpha}\,
+  F\;G[\ol{\beta := K}]$}
+\UIC{$\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol K}_F \eta :
+\Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}\,F)\ol\beta\, G$}
+\DisplayProof
+\]
+respectively.
+Notice that, in these rules, we make essential use of the extended $\Nat$ types.
+
+To give the semantics for $\Lan$ types we follow~\cite{jp19}
+and use the fundamental idea that these have to be interpreted as left Kan extensions.
+We give the definition of left Kan extension and some useful notation in the following definition.
+
+\begin{defi}
+Let $F : \set^k \to \set$ and $K : \set^k \to \set^h$.
+Then, the left Kan extension of $F$ along $\ol K$ is given by a functor $\Lan_K\,F : \set^h \to \set$
+and a natural transformation $\epsilon : F \to (\Lan_K\,F) \circ K$
+which are initial among such pairs of functor and natural transformations, i.e.,
+such that for every $L : \set^h \to \set$ and $\eta :  F \to L \circ K$
+there exists a unique natural transformation $\mu : \Lan_K\,F \to L$
+such that $(\mu K) \circ \epsilon = \eta$, as shown in the following diagram.
+\[
+\begin{tikzcd}[row sep = huge]
+\set^k
+\ar[rr, "{F}"{name=Fa}, ""{name=F, below}]
+\ar[rd, "{K}"']
+&& \set \\
+& \set^h
+\ar[Rightarrow, bend right = 10, from=F, "{\epsilon}" description]
+\ar[ur, bend left, "{\Lan_K\,F}"{name=Lan, description}]
+\ar[ur, bend right, "{L}"'{name=L, right}, ""'{name=Lr, right}]
+\ar[Rightarrow, bend left = 90, from=Fa, to=Lr, "{\eta}"]
+\ar[Rightarrow, from=Lan, to=L, "{\mu}"]
+\end{tikzcd}
+\]
+Of course, we can similarly define the left Kan extension for functors on $\rel$.
+\end{defi}
+
+One might try to restrict the semantic interpretation of $\Lan$ types to a subset of the set produced by the left Kan extension.
+We show in the following proposition that this is not possible, if we want such interpretation to include the interpretation of the terms given by the term-introduction rule.
+
+\begin{prop}\label{prop:full-lan}
+Let $F : \set^h \to \set$ and $\ol{K : \set^h \to \set}$ be functors.
+Assume that $L : \set^k \to \set$ is a functor such that the following properties hold:
+\begin{enumerate}
+\item for all $\ol{A : \set}$, we have that $L \ol{A} \subseteq (\textit{Lan}_{\ol K}\, F)\ol{A}$
+\item for all $\ol{f : A \to B}$ and $\ol{x \in A}$, we have that $(\textit{Lan}_{\ol K}\, F)\ol{f}|_{L \ol{A}}\, x \in L \ol{B}$
+\item for all $\ol{A : \set}$ and $y \in F \ol{A}$, we have that $\epsilon_{\ol{A}} y \in L ( \ol{K \ol{A}} )$
+\end{enumerate}
+Then $L = \textit{Lan}_{\ol K}\, F$.
+\end{prop}
+\begin{proof}
+For every $\ol{A : \set}$ and $z \in (\textit{Lan}_{\ol K}\, F)\ol{A}$
+there exist $\ol{S : \set_0}$, $\ol{f : K \ol{S} \to A}$ and $w \in F \ol{S}$ such that $z = \iota_{\ol{S}, \ol{f}} w$.
+Thus, $z = \iota_{\ol{S}, \ol{f}} w = (\textit{Lan}_{\ol K}\, F) \ol{f} (\epsilon_{\ol{S}} w)$;
+by the third assumption we have that $\epsilon_{\ol{S}} w \in L ( \ol{K \ol{S}} )$
+and by the second assumption we have that $z \in L \ol{A}$.
+We can than conclude that $L \ol{A} = (\textit{Lan}_{\ol K}\, F)\ol{A}$.
+\end{proof}
+
+Thus, we are forced to interpret $\Lan$ types as whole left Kan extensions, without any cutting down.
+The set interpretation of the $\Lan$ type then is
+\[
+  \setsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}}\,F)\ol{A}}\rho
+  = (\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}
+  \rho[\ol{\alpha := \_}]}}\,
+  \setsem{\Gamma; \Phi, \ol\alpha \vdash F}
+  \rho[\ol{\alpha := \_}])\,
+  \ol{\setsem{\Gamma; \Phi \vdash A}\rho}
+\]
+and the relational interpretation is
+\[
+  \relsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}}\,F)\ol{A}}\rho
+  = (\textit{Lan}_{\ol{\relsem{\Gamma; \ol\alpha \vdash
+          K}\rho[\ol{\alpha := \_}]}}\, \relsem{\Gamma; \Phi, \ol\alpha
+      \vdash F}\rho[\ol{\alpha := \_}])\, \ol{\relsem{\Gamma; \Phi
+        \vdash A}\rho}
+\]
+We also need to define the functorial action of the interpretation on morphisms.
+Let $f : \rho \to \rho'$ be a morphism of set environments.
+Then
+\[
+\setsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}}\,F)\ol{A}}f
+: \setsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}}\,F)\ol{A}}\rho
+\to \setsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}}\,F)\ol{A}}\rho'
+\]
+is defined as
+\begin{multline*}
+\big(\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}\,
+      \setsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha := \id_{\_}}]\big)\,
+    \ol{\setsem{\Gamma; \Phi \vdash A}\rho'} \\
+\circ\;
+  \big(\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}\,
+      \setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := \_}]\big)\,
+    \ol{\setsem{\Gamma; \Phi \vdash A}f}
+\end{multline*}
+or, equivalently by naturality,
+\begin{multline*}
+\big(\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}\,
+      \setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := \_}]\big)\,
+    \ol{\setsem{\Gamma; \Phi \vdash A}f} \\
+\circ\;
+  \big(\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}\,
+      \setsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha := \id_{\_}}]\big)\,
+    \ol{\setsem{\Gamma; \Phi \vdash A}\rho}
+\end{multline*}
+The functorial action for the relational interpretation is defined analogously.
+
+The left Kan extension can be expressed via a colimit formula~\cite{rie16} as
+\[
+	(\Lan_K\,F) \ol{A} = \colim{\ol{S} : \set^k_0, \ol{f : K\ol{S} \to A}}{F \ol{S}}
+\]
+for any $\ol{A} : \set^h$, and where $\set_0$ is the category finite sets
+(i.e., the full subcategory of $\omega$-presentable objects in $\set$).
+For any $\ol{S} : \set^k_0$ and $\ol{f : K\ol{S} \to A}$, let
+$j_{\ol{S}, \ol{f}} : F \ol{S} \to (\Lan_K\,F) \ol{A}$
+be the morphism indexed by $\ol{S}$ and $\ol{f}$ of the cocone into the colimit.
+Moreover, for any $h$-tuple of functions $\ol{g : A \to B}$,
+the functorial action $(\Lan_K\,F) \ol{g}$ is the unique function $(\Lan_K\,F) \ol{A} \to (\Lan_K\,F) \ol{B}$
+such that, for every $\ol{S} : \set^k_0$ and $\ol{f : K\ol{S} \to A}$,
+we have that $((\Lan_K\,F) \ol{g}) \circ j_{\ol{S}, \ol{f}} = j_{\ol{S}, \ol{g \circ f}}$.
+We can similarly express via a colimit formula the left Kan extension of functors on $\rel$,
+and in that case we call the cocone over it $\iota$.
+
+We define the semantics for the terms from the introduction and elimination rules for $\Lan$ types as, respectively,
+\begin{multline*}
+\setsem{\Gamma;\Phi~|~\Delta \vdash \int_{\ol K,F} t :
+  (\Lan^{\ol\alpha}_{\ol{K}}\, F)\, \ol{K[\ol{\alpha := A}]}}\rho\;d \\
+= j_{\ol{\setsem{\Gamma;\Phi \vdash A}\rho},
+  \ol{\id_{\setsem{\Gamma; \Phi \vdash K[\ol{\alpha := A}]}\rho}}}\,
+(\setsem{\Gamma; \Phi~|~\Delta \vdash t : F[\ol{\alpha := A}]} \rho\, d)
+\end{multline*}
+and
+\[
+\setsem{\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol K}_F \eta 
+: \Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}\,F) \ol \beta\; G}\rho
+= \lambda \ol{N} \, \ol{B}.\, (\mu_{\ol{N}})_{\ol{B}}
+\]
+where
+  \[\mu_{\ol{N}} : \Lan_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho
+  {[\ol{\alpha := \_} ]}}} \setsem{\Gamma; \Phi, \ol\alpha \vdash
+    F}\rho {[\Phi, \ol \alpha := \ol N, \_ ]} \to \setsem{\Gamma;
+    \Phi, \ol\beta \vdash G} \rho[\Phi, \ol \beta := \ol N, \_]\]
+  is the unique natural transformation such that
+\begin{equation}\label{eq:defn-of-mu-sub-N}
+    (\setsem{\Gamma; \emptyset \,|\, \Delta \vdash \eta : \Nat^{\Phi,
+    \ol\alpha} \, F \,\, G[\ol{\beta := K}]}\rho \, d)_{\ol{N}} =
+(\mu_{\ol{N}} \, \ol{\setsem{\Gamma; \ol\alpha \vdash K}
+  \rho[\ol{\alpha := \_}]}) \circ \epsilon_{\ol{N}}
+  \end{equation}
+as given by the universal property of the left Kan extension.
+Here, $\epsilon_{\ol{N}}$ is the natural transformation associated
+with the left Kan extension, as shown in the following diagram.
+\[\begin{tikzcd} [column sep = large, row sep = large]
+\set^{|\ol \alpha|}
+\ar[rr, "\setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho {[\Phi, \ol
+      \alpha := \ol N, \_ ]}"{name=F}] 
+\ar[dr, "\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho {[\ol{\alpha := \_} ]}}"{left}] && \set \\ 
+&\set^{|\ol K|} \ar[ur, "\Lan_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho {[\ol{\alpha := \_} ]}}}\,
+\setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho {[\Phi, \ol \alpha :=
+    \ol N, \_ ]}"{right}]
+\ar[d, Rightarrow, shorten <= 2mm, shorten >= 1mm, 
+          from=F, to=2-2,  "\epsilon_{\ol{N}}"{}]
+\end{tikzcd}\]
+The relational interpretations are define analogously.
+
+Having defined the syntax and both the set and relational semantics for $\Lan$ types and terms,
+we need to check that these definitions yield a parametric model.
+The first step would be to extend Lemma~\ref{lem:rel-transf-morph} to $\Lan$ types,
+by proving a new clause of the inductive step of the proof.
+Unfortunately, it turns out that, in general, this is not possible, as the following example shows.
+
+\begin{exa}
+We provide a counterexample to Lemma~\ref{lem:rel-transf-morph} for $\Lan$ types.
+Consider the type $\emptyset; \alpha \vdash (\Lan^{\emptyset}_{\onet}\, \onet) \alpha$.
+Then, using the colimit formula for left Kan extensions, we have
+\[
+\begin{array}{rl}
+\relsem{ \emptyset; \alpha \vdash (\Lan^{\emptyset}_{\onet}\, \onet) \alpha } \rho [\alpha := (1, 0, 0)]
+&= (\textit{Lan}_{\Eq_1}\, \Eq_1)(1, 0, 0) \\
+&= \colim{f : \Eq_1 \to (1, 0, 0)}{\Eq_1} \\
+&= (0, 0, 0)
+\end{array}
+\]
+because there exists no morphism $f : \Eq_1 \to (1, 0, 0)$ in $\rel$.
+But, considering the set interpretation on the first projection of the relation environment, we have
+\[
+\begin{array}{rl}
+\setsem{ \emptyset; \alpha \vdash (\Lan^{\emptyset}_{\onet}\, \onet) \alpha } ( \pi_1 ( \rho [\alpha := (1, 0, 0)] ) )
+&= \setsem{ \emptyset; \alpha \vdash (\Lan^{\emptyset}_{\onet}\, \onet) \alpha } (\pi_1 \rho) [\alpha := 1] \\
+&= (\textit{Lan}_{1}\, 1) 1 \\
+&= \colim{1 \to 1}{1} \\
+&= 1
+\end{array}
+\]
+because there exist a unique arrow $1 \to 1$ in $\set$, which is the identity on $1$.
+\end{exa}
+
+The issue lies in the fact that the functor along which we extend contains constants (such as $\onet$).
+Indeed, if we assume that $\ol{\emptyset; \overline{\alpha} \vdash K}$ are polynomial functors not containing constants,
+then we can prove that the relational interpretation of
+$\Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A}$
+is fibered over its set interpretation, as proven in the following proposition.
+
+\begin{prop}
+Assume that the types in $\ol{K}$ are polynomial and contain no constants
+(i.e., they are formed only by the functorial type variables $\ol\alpha$, $+$ and $\times$).
+Then
+\[
+\pi_i (
+\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } \rho
+)
+=
+\setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } (\pi_i \rho)
+\]
+for every relation environment $\rho$ and $i = 1, 2$.
+\end{prop}
+\begin{proof}
+Notice that, by the colimit formula,
+\[
+\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } \rho
+=
+\colim{\ol{R}, \ol{m : \relsem{ \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha := R}] \to \relsem{ \Gamma; \Phi \vdash A } \rho}}{
+\relsem{ \Gamma; \Phi, \ol\alpha \vdash F } \rho[\ol{\alpha := R}]
+}
+\]
+(where we call $\iota$ the cocone of the colimit)
+and thus
+\[
+\pi_i (
+\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } \rho
+)
+=
+\colim{\ol{R : \rel_0}, \ol{m : \relsem{ \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha := R}] \to \relsem{ \Gamma; \Phi \vdash A } \rho}}{
+\relsem{ \Gamma; \Phi, \ol\alpha \vdash F } (\pi_i \rho) [\ol{\alpha := \pi_i R}]}
+\]
+(where the cocone of the colimit is $\pi_i \iota$),
+whereas, again by the colimit formula,
+\[
+\setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } (\pi_i \rho)
+=
+\colim{\ol{S : \set_0}, \ol{f : \setsem{ \emptyset; \ol\alpha \vdash K } (\pi_i \rho)[\ol{\alpha := S}] \to \setsem{ \Gamma; \Phi \vdash A } (\pi_i \rho)}}{
+\setsem{ \Gamma; \Phi, \ol\alpha \vdash F } (\pi_i \rho) [\ol{\alpha := S}]}
+\]
+(where we call $j$ the cocone of the colimit).
+
+We now prove that
+\[
+\pi_1 (
+\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } \rho
+)
+=
+\setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } (\pi_1 \rho)
+\]
+by providing a pair of inverse functions
+(the proof for the second projection is analogous).
+Define
+\[
+h :
+\pi_1 (
+\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } \rho
+)
+\to
+\setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } (\pi_1 \rho)
+\]
+as the unique function given by the universal property of the colimit such that, for any $\ol{R : \rel_0}$ and $\ol{m : \relsem{ \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha := R}] \to \relsem{ \Gamma; \Phi \vdash A } \rho}$, we have that $h \circ (\pi_1 \iota_{\ol{R}, \ol{m}}) = j_{\ol{\pi_1 R}, \ol{\pi_1 m}}$.
+
+Observe that, since $K$ does not contain constants, then $\setsem{ \emptyset; \ol\alpha \vdash K } (\pi_2 \rho) [\ol{\alpha := 0}] = 0$.
+For any
+\[
+\ol{S : \set_0}
+\qquad \text{and} \qquad
+\ol{f : \setsem{ \emptyset; \ol\alpha \vdash K } (\pi_1 \rho)[\ol{\alpha := S}] \to \setsem{ \Gamma; \Phi \vdash A } (\pi_1 \rho)}
+\]
+consider the relations $\ol{(S, 0, 0)}$ and the morphisms $\ol{(f, !) : \relsem{ \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha := (S, 0, 0)}] \to \relsem{ \Gamma; \Phi \vdash A } \rho}$ (where $!$ is the unique function from the empty set).
+Then, define
+\[
+k :
+\setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } (\pi_1 \rho)
+\to
+\pi_1 (
+\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } \rho
+)
+\]
+as the unique function given by the universal property of the colimit such that, for any $\ol{S : \set_0}$ and $\ol{f : \setsem{ \emptyset; \ol\alpha \vdash K } (\pi_1 \rho)[\ol{\alpha := S}] \to \setsem{ \Gamma; \Phi \vdash A } (\pi_1 \rho)}$, we have that $k \circ j_{\ol{S}, \ol{f}} = \pi_1 \iota_{\ol{(S, 0, 0)}, \ol{(f, !)}}$.
+
+We now have to show that $h$ and $k$ are mutually inverse.
+First, we prove that $h \circ k = \id$.
+Indeed, observe that
+\[
+h \circ k \circ j_{\ol{S}, \ol{f}}
+= h \circ (\pi_1 \iota_{\ol{(S, 0, 0)}, \ol{(f, !)}})
+= j_{\ol{\pi_1 (S, 0, 0)}, \ol{\pi_1 (f, !)}}
+= j_{\ol{S}, \ol{f}}
+\]
+for every $\ol{S : \set_0}$ and
+$\ol{f : \setsem{ \emptyset; \ol\alpha \vdash K } (\pi_1 \rho)[\ol{\alpha := S}] \to \setsem{ \Gamma; \Phi \vdash A } (\pi_1 \rho)}$
+and thus $h \circ k = \id$.
+Secondly, we prove that $k \circ h = \id$.
+Indeed, observe that
+\[
+k \circ h \circ (\pi_1 \iota_{\ol{R}, \ol{m}})
+= k \circ j_{\ol{\pi_1 R}, \ol{\pi_1 m}}
+= \pi_1 \iota_{\ol{(\pi_1 R, 0, 0)}, \ol{(\pi_1 m, !)}}
+\]
+for every $\ol{R : \rel_0}$ and $\ol{m : \relsem{ \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha := R}] \to \relsem{ \Gamma; \Phi \vdash A } \rho}$.
+Moreover, consider the morphisms $\ol{(\id_{\pi_1 R}, !) : (\pi_1 R, 0, 0) \to R}$ and observe that $m \circ \relsem{ \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha := (\id_{\pi_1 R}, !)}] = (\pi_1 m, !)$, implying that $\iota_{\ol{R}, \ol{m}} = \iota_{\ol{(\pi_1 R, 0, 0)}, \ol{(\pi_1 m, !)}}$.
+Then $k \circ h \circ (\pi_1 \iota_{\ol{R}, \ol{m}}) = \pi_1 \iota_{\ol{R}, \ol{m}}$ for every $R$ and $m$, and thus $h \circ k = \id$.
+
+We can also prove that
+\[
+\pi_i (
+\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } f
+)
+=
+\setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } (\pi_i f)
+\]
+for every morphism of relation environments $f : \rho \to \rho'$ and $i = 1, 2$.
+Let $\iota$ be the cocone for the colimit $\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } \rho$,
+$j$ the cocone for the colimit $\setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } (\pi_i\rho)$
+and $h$ and $k$ the isomorphisms showing that
+$\pi_i(\setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } \rho) = \setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } (\pi_i\rho)$;
+analogously, let $\iota'$, $j'$, $h'$ and $k'$ be as above but for interpretations in $\rho'$ in place of $\rho$.
+Then $\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } f$
+is the unique morphism of relations such that, for any $\ol{R : \rel_0}$ and
+$\ol{m : \relsem{ \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha := R}] \to \relsem{ \Gamma; \Phi \vdash A } \rho}$,
+we have that
+\[
+(\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } f) \circ \iota_{\ol{R}, \ol{m}}
+= \iota'_{\ol{R}, \ol{\relsem{ \Gamma; \Phi \vdash A } f \circ m}}
+\]
+but then, by projecting the $i$-th component, we have that
+\[
+\pi_i (\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } f) \circ \pi_i \iota_{\ol{R}, \ol{m}}
+= \pi_i \iota'_{\ol{R}, \ol{\relsem{ \Gamma; \Phi \vdash A } f \circ m}}
+\]
+and thus, using the facts that $h \circ (\pi_i \iota_{\ol{R}, \ol{m}}) = j_{\ol{\pi_i R}, \ol{\pi_i m}}$
+and $h' \circ (\pi_i \iota'_{\ol{R}, \ol{m}}) = j'_{\ol{\pi_i R}, \ol{\pi_i m}}$ from above,
+we get
+\[
+\pi_i (\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } f) \circ k \circ j_{\ol{\pi_i R}, \ol{\pi_i m}}
+= k' \circ j'_{\ol{\pi_i R}, \ol{\relsem{ \Gamma; \Phi \vdash A } (\pi_i f) \circ (\pi_i m)}}
+\]
+i.e., $\pi_i (\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } f)$
+satisfy the universal property of $\setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } (\pi_i f)$,
+and thus they are equal (up to natural isomorphism).
+\end{proof}
+
+The requirement that the $\ol{K}$s be polynomial functors without constants is quite restrictive,
+as it precludes the possibility of expressing GADTs used in practice, such as the following type $\mathtt{Expr}$ of typed expressions.
+\[
+\begin{array}{l}
+\mathtt{data\; Expr \;(A : Set)\;:\;Set\;where}\\
+\hspace*{0.4in}\mathtt{const\;:\; Int \to Expr\; Int}\\
+\hspace*{0.4in}\mathtt{is\_zero\;:\; Expr\; Int \rightarrow Expr\; Bool}\\
+\hspace*{0.4in}\mathtt{if\;:\; Expr\; Bool \to Expr\; A \to Expr\; A \to Expr\; A}
+\end{array}
+\]
+
+{\color{red} This is not quite right.}
+Even when extending along polynomials containing no constants, it is unclear how to prove the Identity Extension Lemma for $\Lan$ types, or if it holds at all.
+In order to satisfy the identity extension lemma in the case of $\Nat$ types, for example, it is necessary to cut down the semantic interpretation by requiring that the natural transformations in the set interpretation preserve equalities.
+The impossibility of restricting the semantic interpretation of $\Lan$, as proved by Proposition~\ref{prop:full-lan}, suggests that the Identity Extension Lemma might indeed not hold for a type system that features such types.
+
+We must then conclude that it is not possible to use functorial semantics to provide a parametric model for $\Lan$ types.
+
 \section{Conclusion and Directions for Future Work}\label{sec:conclusion}

### 2020-12-21 free-theorems-nesteds/20-mscs.tex
revising GADT section - in progress

@@
 
-Generalized algebraic datatypes (GADTs) generalize nested types by...
-
-For example, ...
-
-A natural next step in this line of work is to extend our
-parametricity results to GADTs.  It was shown in~\cite{jg08} that (the
-data of) GADTs can be represented using object-level left Kan
-extensions (over discrete categories), and further shown
-in~\cite{jp19} that adding a carefully designed left Kan extension
-construct to a calculus supporting primitive nested types preserves
-the cocontinuity needed for primitive GADTs to have well-defined,
-properly functorial interpretations.
-
-This suggests adding GADTs to our calculus by extending
-Definition~\ref{def:wftypes} with such object-level constructs, and
-extending Figure~\ref{fig:terms} with categorically inspired
-constructs to introduce and eliminate terms of these types, along the
-same entirely standard lines as we did there for products, coproducts,
-and fixpoints.  In this section we outline how our model from
-Section~\ref{sec:term-interp} can be extended to a parametric model
-when some classes of GADTs are incorporated as primitives in this
-way. However, as we argue in~???? below, we suspect this is not
-possible in general.
-
-HERE!!
-
-To show when this can be done, as well as to give a sense of what goes
-wrong in the general case, we begin by recalling that GADTs are
-naturally interpreted as initial algebras of left Kan extensions, as
-suggested in discrete~\cite{jg08} and
-
-Functorial semantics in~\cite{jp19}.
-
-introducing a new
-type, $\Lan$, to represent left Kan extensions.  Indeed,
-
-To begin with, we add the following type formation rule for $\Lan$
-types to our type system.
+As discussed in Section~\ref{sec:intro}, type indices for nested types
+can be any types, including, in the case of truly nested types like
+that of bushes, types involving the very same nested type that is
+being defined. But every data constructor for a nested type must still
+have as its return type exactly the instance being defined. For
+example, the data constructors for the instance $\mathtt{PTree\; A}$ of the
+nested type $\mathtt{PTree}$ are $\mathtt{pleaf :: PTree\; A}$ and
+$\mathtt{pnode :: PTree\; (A \times A) \to PTree\; A}$, and
+the data constructors for the instance $\mathtt{Bush \;A}$ of the truly
+nested type $\mathtt{Bush}$ are $\mathtt{bnil :: Bush\; A}$ and
+$\mathtt{bcons :: A \to Bush\; (Bush\; A) \to Bush\; A}$.
+
+Generalized algebraic datatypes (GADTs){\color{red}~\cite{???}} generalize nested
+types by relaxing the above restriction to allow the return types of
+data constructors to be different instances of the datatype than the
+one being defined. For example, the GADT
+{\small
+\[\begin{array}{l}
+\mathtt{data\; Seq \;(A : Set)\;:\;Set\;where}\\
+\hspace*{0.4in}\mathtt{sconst\;:\; A \rightarrow List\;A}\\
+\hspace*{0.4in}\mathtt{spair\;:\;Seq\;A \rightarrow Seq\;B \rightarrow
+  Seq\;(A\times B)}\\
+\hspace*{0.4in}\mathtt{sseq\;:\;(Nat \rightarrow Seq\;A) \rightarrow
+  Seq\;(Nat \to A)}
+\end{array}\]}
+\noindent
+\!\!has data constructors $\mathtt{spair}$ and $\mathtt{sseq}$ with
+return types $\mathtt{Seq\; (A \times B)}$ and $\mathtt{Seq\;(Nat \to
+  A)}$. These types are not only at different instances of \verb|Seq|
+from the instance $\mathtt{Seq\; A}$ being defined, but also at
+different instances from one another. The resulting interdependence of
+different instances of GADTs means that they can express more
+constraints than ADTs and nested types. For example, the ADT
+$\mathtt{List}$ expresses the invariant that all of the data in the
+lists it defines is of the same type, while the nested type
+$\mathtt{PTree}$ expresses this invariant {\em as well as} the
+invariant that all of the lists it defines have lengths that are
+powers of 2. The GADT $\mathtt{Seq}$ enforces {\em even more general}
+well-formedness conditions for sequences of values that simply cannot
+be expressed with ADTs and nested types alone.
+
+GADTs are widely used in modern functional languages, such as Haskell,
+as well as in proof assistants, such as Agda, that are based on
+dependent type theories. A natural next step in the line of work
+reported in this paper is therefore to extend our parametricity
+results to GADTs. A promising starting point for this endeavor is the
+observation from~\cite{jg08} that the data objects of GADTs can be
+represented using object-level left Kan extensions over discrete
+categories. The more recent results of~\cite{jp19} further show that
+adding a carefully designed object-level left Kan extension construct
+to a calculus supporting primitive nested types preserves the
+cocontinuity needed for primitive GADTs to have well-defined, properly
+functorial interpretations. Together this suggests extending the type
+system in Definition~\ref{def:wftypes} with such a left Kan extension
+construct, and extending the calculus in Figure~\ref{fig:terms} with
+corresponding categorically inspired constructs to introduce and
+eliminate terms of these richer types. This approach exactly mirrors
+the (entirely standard) approach taken above for product, coproduct,
+and fixpoint types. In this section we outline this approach and show
+how our model from Section~\ref{sec:term-interp} can be extended to a
+parametric model when some classes of primitive GADTs are incorporated
+in this manner. However, as we argue in{\color{red}~????} below, such
+extension is not possible in general.
+
+\subsection{Left Kan Extensions}\label{sec:lke}
+
+We begin by recalling the definition of a left Kan extension and
+establishing some useful notation surounding them.
+
+\begin{defi}\label{def:lke}
+If $F : \set^k \to \set$ and $\ol{K : \set^k \to \set^h}$ are functors
+over $\set$, then the {\em left Kan extension of $F$ along $\ol K$} is
+a functor $\mathit{Lan}_{\ol K}\,F : \set^h \to \set$ together with a
+natural transformation $\epsilon : F \to (\mathit{Lan}_{\ol K}\,F)
+\circ \ol K$
+%are initial among such pairs of functor and natural transformations,
+%i.e.,
+such that, for every functor $L : \set^h \to \set$ and natural
+transformation $\eta : F \to L \circ \ol K$, there exists a unique
+natural transformation $\mu : \mathit{Lan}_{\ol K}\,F \to L$ such that
+$(\mu {\ol K}) \circ \epsilon = \eta$. This is depicted in the
+following diagram:
+
+\vspace*{-0.15in}
+
+\[
+\begin{tikzcd}[row sep = huge]
+\set^k
+\ar[rr, "{F}"{name=Fa}, ""{name=F, below}]
+\ar[rd, "{\ol K}"']
+&& \set \\
+& \set^h
+\ar[Rightarrow, bend right = 10, from=F, "{\epsilon}" description]
+\ar[ur, bend left, "{\mathit{Lan}_K\,F}"{name=Lan, description}]
+\ar[ur, bend right, "{L}"'{name=L, right}, ""'{name=Lr, right}]
+\ar[Rightarrow, bend left = 90, from=Fa, to=Lr, "{\eta}"]
+\ar[Rightarrow, dashed, from=Lan, to=L, "{\mu}"]
+\end{tikzcd}
+\]
+
+\vspace*{0.05in}
+
+\noindent
+Replacing $\set$ by $\rel$ everywhere in Definition~\ref{def:lke} we
+can similarly define the left Kan extension of $F$ along $\ol K$ for
+functors $F$ and $\ol K$ over $\rel$.
+\end{defi}
+
+{\color{red}
+
+An alternative presentation of left Kan extensions characterizes them
+in terms of an adjunction.
+
+Using this adjunction we can represent the types of the data
+constructors for $\mathtt{Seq}$ as:
+}
+
+\subsection{Extending the Calculus}
+
+In light of the above, we incorporate GADTs into our calculus by first
+adding to Definition~\ref{def:wftypes} the following type formation
+rule for $\Lan$-types:
+
+\vspace*{0.05in}
+
 \[
@@
 \]
-Notice that the length of $\ol K$ coincides with that of $\ol A$.
-Introducing the new type also requires the following additional clause in the inductive definition of higher-order substitution.
-\[
-((\Lan^{\ol{\alpha}}_{\ol K}\,F) \ol A)[\phi :=_{\ol{\beta}} G]
-  = (\Lan^{\ol\alpha}_{\ol K}\,F[\phi :=_{\ol{\beta}} G])\, \ol{A[\phi :=_{\ol{\beta}} G]}
-\]
-
-Before we extend the term calculus with the term introduction and elimination rules for $\Lan$ types,
-it is necessary to generalize $\Nat$ types to bind type constructor variables of arbitrary arity.
-So, the formation rule for $\Nat$ types will be modified as
+
+\vspace*{0.05in}
+
+\noindent
+Here, $\Lan$ binds the variables in $\ol\alpha$, and these variables
+must always have arity $0$. In addition, the vectors $\ol K$ and $\ol
+A$ must have the same length.
+
+Intuitively, $\Lan^{\ol{\alpha^0}}_K F$ is a syntactic representation
+of the left Kan extension of the functor in the variables in
+$\ol\alpha$ denoted by $F$ along the functor in the variables in
+$\ol\alpha$ denoted by $K$. Using $\Lan$-types, we can therefore
+represent the GADT $\mathtt{Seq}$ in our calculus as {\color{red}
+  CHECK!!
+ \begin{align*}
+\mathtt{Seq}\,\alpha &\df \left( \mu \phi.\lambda \beta.
+\beta + \left(\Lan^{\gamma_1,\gamma_2}_{\gamma_1 \times \gamma_2}
+(\phi \gamma_1 \times \phi \gamma_2) \right) \beta \right. \\ &\qquad
+\left. + \left(\Lan^{\gamma}_{\mathtt{Nat} \to \gamma} (\mathtt{Nat} \to \phi
+\gamma) \right) \beta + \phi (\mathtt{Nat} \to \beta) \times
+\mathtt{Nat} \right) \alpha
+\end{align*}
+\noindent
+where $\mathtt{Nat} = \mu \alpha. \bone + \alpha$ and $\mathtt{Nat}
+\to D$ abbreviates $(\Lan^\beta_{\mathtt{Nat}} \bone)D$.} This
+representation corresponds exactly to the rewriting of $\mathit{Seq}$
+in Section~\ref{sec:lke}. As explained in Section~IV.D of~\cite{jp19},
+the more general construct $\Lan^{\ol{\alpha^0}}_{\ol K} F$ allowing
+extensions along vectors of functors makes it possible to represent
+GADTs with two or more type arguments that depend on one another. Such
+GADTs cannot be represented using just unary $\Lan$-types, i.e,
+$\Lan$-types of the form $(\Lan^{\ol{\alpha^0}}_K F)\ol A$.
+{\color{red} $\Lan$ vs $\Lan$-type.}
+
+To define the term-formation rules for terms of $\Lan$-types we also
+need to generalize $\Nat$-types to bind type constructor variables not
+just of arity $0$, but of arbitrary arity. {\color{red} This is
+  because...WHY????}  Accordingly, we replace the rule for
+$\Nat$-types in Definition~\ref{def:wftypes} with
+
+\vspace*{0.05in}
+
 \[
@@
 \]
-and the term introduction and elimination rules will be
+
+\vspace*{0.05in}
+
+Finally, adding $\Lan$-types to the types of our calculus also
+requires the extension of Definition~\ref{def:second-order-subst} with
+the following new clause:
+\[((\Lan^{\ol{\alpha}}_{\ol K}\,F) \ol A)[\phi :=_{\ol{\beta}} G] =
+(\Lan^{\ol\alpha}_{\ol K}\,F[\phi :=_{\ol{\beta}} G])\, \ol{A[\phi
+    :=_{\ol{\beta}} G]}\]
+
+\vspace*{0.1in}
+
+To also extend our term calculus to accommodate GADTs we must first
+give introduction and elimination rules appropriate to our generalized
+$\Nat$-types. We therefore replace the tenth and eleventh rules in
+Figure~\ref{fig:terms} with
+
+\vspace*{0.05in}
+
 \[
@@
 and
+
+\vspace*{0.05in}
+
 \[
@@
 \]
-respectively.
-It is necessary, of course, to suitably extend the semantic interpretations of $\Nat$ types and terms.
-The type constructor variables bound by $\Nat$ will be interpreted as either functors
-or relation transformers.
-The set and relational interpretations for $\Nat$ types are defined as follows.
+respectively. Finally, we add to the term calculus in
+Figure~\ref{fig:terms} the following introduction and elimination
+rules for terms of $\Lan$-types, respectively:
+
+\vspace*{0.05in}
+
+\[
+\AXC{$\Gamma; \Phi,\ol{\alpha}\vdash F$}
+\AXC{$\ol{\Gamma; \ol{\alpha} \vdash K}$}
+\BIC{$\Gamma;\emptyset~|~\Delta \vdash \int_{\ol K,F} : \Nat^{\Phi}\, F\, (\Lan^{\ol\alpha}_{\ol{K}}\,  F) \ol{K}$}
+\DisplayProof
+\]
+
+\vspace*{0.05in}
+
+\[
+\AXC{$\Gamma;\emptyset~|~\Delta \vdash \eta : \Nat^{\Phi,\ol\alpha}\,
+  F\;G[\ol{\beta := K}]$}
+\UIC{$\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol K}_F \eta :
+\Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}\,F)\ol\beta\, G$}
+\DisplayProof
+\]
+
+\vspace*{0.05in}
+
+\noindent
+Note that these rules do indeed make essential use of our new,
+generalized $\Nat$-types.
+
+{\color{red} Say how these rules mirror real LKEs just as for
+  products, coproducts, etc.}
+
+\subsection{Extending the Semantics}
+
+It is necessary, of course, to suitably extend the semantic
+interpretations of $\Nat$ types and terms.  The type constructor
+variables bound by $\Nat$ will be interpreted as either functors or
+relation transformers.  The set and relational interpretations for
+$\Nat$ types are defined as follows.
 \begin{align*}
@@
 
-We can now give the term introduction and elimination rules for $\Lan$, which are
-\[
-\AXC{$\Gamma; \Phi,\ol{\alpha}\vdash F$}
-\AXC{$\ol{\Gamma; \ol{\alpha} \vdash K}$}
-\BIC{$\Gamma;\emptyset~|~\Delta \vdash \int_{\ol K,F} : \Nat^{\Phi}\, F\, (\Lan^{\ol\alpha}_{\ol{K}}\,  F) \ol{K}$}
-\DisplayProof
-\]
-and
-\[
-\AXC{$\Gamma;\emptyset~|~\Delta \vdash \eta : \Nat^{\Phi,\ol\alpha}\,
-  F\;G[\ol{\beta := K}]$}
-\UIC{$\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol K}_F \eta :
-\Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}\,F)\ol\beta\, G$}
-\DisplayProof
-\]
-respectively.
-Notice that, in these rules, we make essential use of the extended $\Nat$ types.
-
-To give the semantics for $\Lan$ types we follow~\cite{jp19}
-and use the fundamental idea that these have to be interpreted as left Kan extensions.
-We give the definition of left Kan extension and some useful notation in the following definition.
-
-\begin{defi}
-Let $F : \set^k \to \set$ and $K : \set^k \to \set^h$.
-Then, the left Kan extension of $F$ along $\ol K$ is given by a functor $\Lan_K\,F : \set^h \to \set$
-and a natural transformation $\epsilon : F \to (\Lan_K\,F) \circ K$
-which are initial among such pairs of functor and natural transformations, i.e.,
-such that for every $L : \set^h \to \set$ and $\eta :  F \to L \circ K$
-there exists a unique natural transformation $\mu : \Lan_K\,F \to L$
-such that $(\mu K) \circ \epsilon = \eta$, as shown in the following diagram.
-\[
-\begin{tikzcd}[row sep = huge]
-\set^k
-\ar[rr, "{F}"{name=Fa}, ""{name=F, below}]
-\ar[rd, "{K}"']
-&& \set \\
-& \set^h
-\ar[Rightarrow, bend right = 10, from=F, "{\epsilon}" description]
-\ar[ur, bend left, "{\Lan_K\,F}"{name=Lan, description}]
-\ar[ur, bend right, "{L}"'{name=L, right}, ""'{name=Lr, right}]
-\ar[Rightarrow, bend left = 90, from=Fa, to=Lr, "{\eta}"]
-\ar[Rightarrow, from=Lan, to=L, "{\mu}"]
-\end{tikzcd}
-\]
-Of course, we can similarly define the left Kan extension for functors on $\rel$.
-\end{defi}
-
-One might try to restrict the semantic interpretation of $\Lan$ types to a subset of the set produced by the left Kan extension.
-We show in the following proposition that this is not possible, if we want such interpretation to include the interpretation of the terms given by the term-introduction rule.
+We intend to interpret $\Lan$-types as actual left Kan extensions. But
+in order to guarantee parametricity we might think we have to cut down
+their set interpretations by restricting to a well-behaved collection
+of the sets produced by such interpretations, just as we did both in
+our original interpretations of $\Nat$-types and the above
+interpretations of our generalized $\Nat$-types.
+
+{\color{red} REORDER!!!!}
+
+Along similar lines, we might try to restrict the semantic
+interpretation of $\Lan$ types to a subset of the set produced by the
+left Kan extension. {\color{red} WHY?????} We show in the following
+proposition that this is not possible, if we want such interpretation
+to include the interpretation of the terms given by the
+term-introduction rule.
 

### 2020-12-22 free-theorems-nesteds/20-mscs.tex
more on Sec 8

@@
 \mathtt{data\; Seq \;(A : Set)\;:\;Set\;where}\\
-\hspace*{0.4in}\mathtt{sconst\;:\; A \rightarrow List\;A}\\
+\hspace*{0.4in}\mathtt{sconst\;:\; A \rightarrow Seq\;A}\\
 \hspace*{0.4in}\mathtt{spair\;:\;Seq\;A \rightarrow Seq\;B \rightarrow
@@
 We begin by recalling the definition of a left Kan extension and
-establishing some useful notation surounding them.
+establishing some useful notation and results for them.
 
 \begin{defi}\label{def:lke}
-If $F : \set^k \to \set$ and $\ol{K : \set^k \to \set^h}$ are functors
+If $F : \set^k \to \set$ and $\ol{K} : \set^k \to \set^h$ are functors
 over $\set$, then the {\em left Kan extension of $F$ along $\ol K$} is
 a functor $\mathit{Lan}_{\ol K}\,F : \set^h \to \set$ together with a
-natural transformation $\epsilon : F \to (\mathit{Lan}_{\ol K}\,F)
+natural transformation $\eta : F \to (\mathit{Lan}_{\ol K}\,F)
 \circ \ol K$
@@
 such that, for every functor $L : \set^h \to \set$ and natural
-transformation $\eta : F \to L \circ \ol K$, there exists a unique
+transformation $\epsilon : F \to L \circ \ol K$, there exists a unique
 natural transformation $\mu : \mathit{Lan}_{\ol K}\,F \to L$ such that
-$(\mu {\ol K}) \circ \epsilon = \eta$. This is depicted in the
+$(\mu {\ol K}) \circ \eta = \epsilon$. This is depicted in the
 following diagram:
@@
 & \set^h
-\ar[Rightarrow, bend right = 10, from=F, "{\epsilon}" description]
+\ar[Rightarrow, bend right = 10, from=F, "{\eta}" description]
 \ar[ur, bend left, "{\mathit{Lan}_K\,F}"{name=Lan, description}]
 \ar[ur, bend right, "{L}"'{name=L, right}, ""'{name=Lr, right}]
-\ar[Rightarrow, bend left = 90, from=Fa, to=Lr, "{\eta}"]
+\ar[Rightarrow, bend left = 90, from=Fa, to=Lr, "{\epsilon}"]
 \ar[Rightarrow, dashed, from=Lan, to=L, "{\mu}"]
@@
 
-{\color{red}
-
-An alternative presentation of left Kan extensions characterizes them
-in terms of an adjunction.
-
-Using this adjunction we can represent the types of the data
-constructors for $\mathtt{Seq}$ as:
-}
-
-\subsection{Extending the Calculus}
-
-In light of the above, we incorporate GADTs into our calculus by first
-adding to Definition~\ref{def:wftypes} the following type formation
-rule for $\Lan$-types:
+An alternative presentation of characterizes the left Kan extension
+$(\mathit{Lan}_{\ol K} F, \eta)$ in terms of the adjunction between
+natural transformations from $F$ to $G \circ \ol{K}$ and natural
+transformations from $\mathit{Lan}_{\ol K} F$ to $G$, for which
+$\eta$ is the unit. If Agda were to support a primitive
+$\mathsf{Lan}$ for left Kan extensions, we could use this adjunction
+to rewrite the type of each data constructor for the GADT
+$\mathtt{Seq}$ to arrive at the following equivalent representation:
+{\small
+\[\begin{array}{l}
+\mathtt{data\; Seq \;(A : Set)\;:\;Set\;where}\\
+\hspace*{0.4in}\mathtt{sconst\;:\; A \rightarrow Seq\;A}\\
+\hspace*{0.4in}\mathtt{spair\;:\;(\mathsf{Lan}_{\lambda C\,D.\,C \times D}
+  (\lambda C\,D.\,Seq\;C\times Seq\;D))\;A \rightarrow Seq\;A}\\ 
+\hspace*{0.4in}\mathtt{sseq\;:\;(\mathsf{Lan}_{\lambda C.\,Nat \to C}
+  (\lambda C.\,Nat \rightarrow Seq\;C))\;A \rightarrow
+  Seq\;A}
+\end{array}\]}
+\noindent
+\!\!Our calculus will represent $\mathtt{Seq}$ and other GADTs in
+precisely this way.
+
+\vspace*{0.05in}
+
+A third representation of left Kan extensions in locally presentable
+categories is given in terms of colimits. Writing ${\mathcal C}_0$ for
+the full subcategory of finitely presentable objects in the locally
+presentable category $\mathcal C$, the left Kan extension can be
+expressed as
+\[(\mathit{Lan}_{\ol K} F)\ol A \;=\; \colim{(\ol{X : \,{\mathcal C}_0},\,
+  \ol{f : K{\ol X} \to A})}{F \ol X}\]
+
+Thus, if $F : \set^k \to \set$, $\ol{K} :
+  \set^k \to \set^h$, $\ol{A} : \set^h$, and $\set_0$ is the full
+  subcategory of finitely presentable objects in $\set$, --- i.e., is
+  the category finite sets --- we  have that
+\[(\mathit{Lan}_K F) \ol{A} = \colim{(\ol{S} : \set_0^k, \,\ol{f : K\ol{S} \to
+    A})}{F \ol{S}}\]
+If $\ol{S} : \set_0^k$ and $\ol{f : K\ol{S} \to A}$, let $j_{\ol{S},
+  \ol{f}} : F \ol{S} \to (\mathit{Lan}_KF) \ol{A}$ be the morphism
+indexed by $\ol{S}$ and $\ol{f}$ mapping the cocone into the colimit.
+Then for any $h$-tuple of functions $\ol{g : A \to B}$, the functorial
+action $(\mathit{Lan}_K F) \ol{g}$ is the unique function from
+$(\mathit{Lan}_K F) \ol{A}$ to $(\mathit{Lan}_K F) \ol{B}$ such that,
+for every $\ol{S} : \set^k_0$ and $\ol{f : K\ol{S} \to A}$,\, 
+$(\mathit{Lan}_KF) \ol{g} \circ j_{\ol{S},\, \ol{f}} = j_{\ol{S},\,
+  \ol{g\, \circ f}}$ holds.
+
+We can similarly represent left Kan extensions of functors over $\rel$
+in terms of colimits. In that setting we will denote the morphism
+mapping the cocone into the colimit $\iota$. We will make good use of
+both of these representations in Section~\ref{sec:term-sem} below.
+
+\vspace*{0.05in}
+
+More about each of the above three representations of left Kan
+extensions and the connections between them can be found in,
+e.g.,~\cite{rie16}.
+
+\subsection{Extending the Calculus}\label{sec:ext-calc}
+
+We incorporate GADTs into our calculus by first adding to
+Definition~\ref{def:wftypes} the following type formation rule for
+$\Lan$-types:
 
@@
 \noindent
-Here, $\Lan$ binds the variables in $\ol\alpha$, and these variables
-must always have arity $0$. In addition, the vectors $\ol K$ and $\ol
-A$ must have the same length.
+Here, the type constructor $\Lan$ binds the variables in $\ol\alpha$,
+and these variables must always have arity $0$. In addition, the
+vectors $\ol K$ and $\ol A$ must have the same length.
 
@@
 represent the GADT $\mathtt{Seq}$ in our calculus as {\color{red}
-  CHECK!!
+  CHECK!!}
  \begin{align*}
@@
 \beta + \left(\Lan^{\gamma_1,\gamma_2}_{\gamma_1 \times \gamma_2}
-(\phi \gamma_1 \times \phi \gamma_2) \right) \beta \right. \\ &\qquad
+(\phi \gamma_1 \times \phi \gamma_2) \right) \beta \right. 
 \left. + \left(\Lan^{\gamma}_{\mathtt{Nat} \to \gamma} (\mathtt{Nat} \to \phi
-\gamma) \right) \beta + \phi (\mathtt{Nat} \to \beta) \times
-\mathtt{Nat} \right) \alpha
+\gamma) \right) \beta \right) \alpha
 \end{align*}
 \noindent
-where $\mathtt{Nat} = \mu \alpha. \bone + \alpha$ and $\mathtt{Nat}
-\to D$ abbreviates $(\Lan^\beta_{\mathtt{Nat}} \bone)D$.} This
-representation corresponds exactly to the rewriting of $\mathit{Seq}$
-in Section~\ref{sec:lke}. As explained in Section~IV.D of~\cite{jp19},
-the more general construct $\Lan^{\ol{\alpha^0}}_{\ol K} F$ allowing
-extensions along vectors of functors makes it possible to represent
-GADTs with two or more type arguments that depend on one another. Such
-GADTs cannot be represented using just unary $\Lan$-types, i.e,
-$\Lan$-types of the form $(\Lan^{\ol{\alpha^0}}_K F)\ol A$.
-{\color{red} $\Lan$ vs $\Lan$-type.}
-
-To define the term-formation rules for terms of $\Lan$-types we also
-need to generalize $\Nat$-types to bind type constructor variables not
-just of arity $0$, but of arbitrary arity. {\color{red} This is
-  because...WHY????}  Accordingly, we replace the rule for
-$\Nat$-types in Definition~\ref{def:wftypes} with
+where $\mathtt{Nat} = \mu \alpha. 1 + \alpha$ and $\mathtt{Nat}
+\to C$ abbreviates $(\Lan^\alpha_{\mathtt{Nat}} 1)\,C$.  This
+representation of $\mathit{Seq}$ corresponds exactly to the rewriting
+in Section~\ref{sec:lke}.  As explained in Section~IV.\,D
+of~\cite{jp19}, the more general construct $\Lan^{\ol{\alpha^0}}_{\ol
+  K} F$ allowing extensions along vectors of functors as depicted
+above makes it possible to represent GADTs with two or more type
+arguments that depend on one another. Such GADTs cannot be represented
+using just unary $\Lan$-types, i.e, $\Lan$-types of the form
+$(\Lan^{\ol{\alpha^0}}_K F)\ol A$.
+
+As we will see below, in order to define the term-formation rules for
+terms of $\Lan$-types we also need to generalize $\Nat$-types to bind
+type constructor variables not just of arity $0$, but of arbitrary
+arity. Accordingly, we replace the rule for $\Nat$-types in
+Definition~\ref{def:wftypes} with
 
@@
 the following new clause:
+
 \[((\Lan^{\ol{\alpha}}_{\ol K}\,F) \ol A)[\phi :=_{\ol{\beta}} G] =
-(\Lan^{\ol\alpha}_{\ol K}\,F[\phi :=_{\ol{\beta}} G])\, \ol{A[\phi
+(\Lan^{\ol\alpha}_{\ol K}\,F[\phi :=_{\ol{\beta}} G]) \ol{A[\phi
     :=_{\ol{\beta}} G]}\]
@@
 
-To also extend our term calculus to accommodate GADTs we must first
-give introduction and elimination rules appropriate to our generalized
-$\Nat$-types. We therefore replace the tenth and eleventh rules in
-Figure~\ref{fig:terms} with
+We must also extend our term calculus to accommodate GADTs. To this
+end, we first give introduction and elimination rules appropriate to
+our generalized $\Nat$-types, replacing the tenth and eleventh rules
+in Figure~\ref{fig:terms} with
 
@@
 \]
-respectively. Finally, we add to the term calculus in
-Figure~\ref{fig:terms} the following introduction and elimination
-rules for terms of $\Lan$-types, respectively:
+respectively. Next, we add to the term calculus in
+Figure~\ref{fig:terms} introduction and elimination rules for terms of
+$\Lan$-types according to Definition~\ref{def:lke}. This gives the
+rules
 
@@
 \]
-
-\vspace*{0.05in}
-
+and
 \[
@@
 \noindent
-Note that these rules do indeed make essential use of our new,
+respectively. Note that both of these rules make essential use of our
 generalized $\Nat$-types.
 
-{\color{red} Say how these rules mirror real LKEs just as for
-  products, coproducts, etc.}
-
-\subsection{Extending the Semantics}
-
-It is necessary, of course, to suitably extend the semantic
-interpretations of $\Nat$ types and terms.  The type constructor
-variables bound by $\Nat$ will be interpreted as either functors or
-relation transformers.  The set and relational interpretations for
-$\Nat$ types are defined as follows.
+\subsection{Extending the Type Semantics}
+
+To construct a parametric model of the kind we seek, we must give
+suitable interpretations of our generalized $\Nat$-types and
+$\Lan$-types, as well as of terms of both such types, in both $\set$
+and $\rel$. Recalling that the type constructor variables bound by
+$\Nat$ in the formation rule for generalized $\Nat$-types are
+interpreted in $\set$ as functors and in $\rel$ as relation
+transformers, we extend the set and relational interpretations for
+$\Nat$ types from Section~\ref{sec:type-interp} as follows:
 \begin{align*}
@@
 
-We intend to interpret $\Lan$-types as actual left Kan extensions. But
-in order to guarantee parametricity we might think we have to cut down
-their set interpretations by restricting to a well-behaved collection
-of the sets produced by such interpretations, just as we did both in
-our original interpretations of $\Nat$-types and the above
-interpretations of our generalized $\Nat$-types.
-
-{\color{red} REORDER!!!!}
-
-Along similar lines, we might try to restrict the semantic
-interpretation of $\Lan$ types to a subset of the set produced by the
-left Kan extension. {\color{red} WHY?????} We show in the following
-proposition that this is not possible, if we want such interpretation
-to include the interpretation of the terms given by the
-term-introduction rule.
-
-\begin{prop}\label{prop:full-lan}
-Let $F : \set^h \to \set$ and $\ol{K : \set^h \to \set}$ be functors.
-Assume that $L : \set^k \to \set$ is a functor such that the following properties hold:
-\begin{enumerate}
-\item for all $\ol{A : \set}$, we have that $L \ol{A} \subseteq (\textit{Lan}_{\ol K}\, F)\ol{A}$
-\item for all $\ol{f : A \to B}$ and $\ol{x \in A}$, we have that $(\textit{Lan}_{\ol K}\, F)\ol{f}|_{L \ol{A}}\, x \in L \ol{B}$
-\item for all $\ol{A : \set}$ and $y \in F \ol{A}$, we have that $\epsilon_{\ol{A}} y \in L ( \ol{K \ol{A}} )$
-\end{enumerate}
-Then $L = \textit{Lan}_{\ol K}\, F$.
-\end{prop}
-\begin{proof}
-For every $\ol{A : \set}$ and $z \in (\textit{Lan}_{\ol K}\, F)\ol{A}$
-there exist $\ol{S : \set_0}$, $\ol{f : K \ol{S} \to A}$ and $w \in F \ol{S}$ such that $z = \iota_{\ol{S}, \ol{f}} w$.
-Thus, $z = \iota_{\ol{S}, \ol{f}} w = (\textit{Lan}_{\ol K}\, F) \ol{f} (\epsilon_{\ol{S}} w)$;
-by the third assumption we have that $\epsilon_{\ol{S}} w \in L ( \ol{K \ol{S}} )$
-and by the second assumption we have that $z \in L \ol{A}$.
-We can than conclude that $L \ol{A} = (\textit{Lan}_{\ol K}\, F)\ol{A}$.
-\end{proof}
-
-Thus, we are forced to interpret $\Lan$ types as whole left Kan extensions, without any cutting down.
-The set interpretation of the $\Lan$ type then is
+Of course, we intend to interpret $\Lan$-types as actual left Kan
+extensions. To get started, we define their set and relational
+interpretations to be
 \[
-  \setsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}}\,F)\ol{A}}\rho
-  = (\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}
-  \rho[\ol{\alpha := \_}]}}\,
-  \setsem{\Gamma; \Phi, \ol\alpha \vdash F}
-  \rho[\ol{\alpha := \_}])\,
-  \ol{\setsem{\Gamma; \Phi \vdash A}\rho}
-\]
-and the relational interpretation is
-\[
-  \relsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}}\,F)\ol{A}}\rho
-  = (\textit{Lan}_{\ol{\relsem{\Gamma; \ol\alpha \vdash
-          K}\rho[\ol{\alpha := \_}]}}\, \relsem{\Gamma; \Phi, \ol\alpha
-      \vdash F}\rho[\ol{\alpha := \_}])\, \ol{\relsem{\Gamma; \Phi
+  \setsem{\Gamma;\Phi \vdash
+    (\Lan^{\ol{\alpha}}_{\ol{K}}\,F)\ol{A}}\rho = (\lambda \ol{S}.\,
+  \textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}
+      \rho[\ol{\alpha := S}]}}\, \setsem{\Gamma; \Phi, \ol\alpha
+    \vdash F} \rho[\ol{\alpha := S}])\, \ol{\setsem{\Gamma; \Phi
       \vdash A}\rho}
 \]
-We also need to define the functorial action of the interpretation on morphisms.
-Let $f : \rho \to \rho'$ be a morphism of set environments.
-Then
+and
+
+\[
+  \relsem{\Gamma;\Phi \vdash
+    (\Lan^{\ol{\alpha}}_{\ol{K}}\,F)\ol{A}}\rho = (\lambda \ol R.\,
+    \textit{Lan}_{\ol{\relsem{\Gamma; \ol\alpha \vdash
+          K}\rho[\ol{\alpha := R}]}}\, \relsem{\Gamma; \Phi, \ol\alpha
+      \vdash F}\rho[\ol{\alpha := R}])\, \ol{\relsem{\Gamma; \Phi
+        \vdash A}\rho}
+\]
+
+\vspace*{0.05in}
+
+\noindent
+respectively. But in order to guarantee that the IEL, and thus
+parametricity, continues to hold for the model we are constructing for
+our extended calculus, we might think we have to cut down the set
+interpretations of $\Lan$-types by restricting to the subcollection of
+those interpretations that are well-behaved with respect to the
+IEL. This would mirror the cutting down we have already seen in the
+set interpretations of both our original $\Nat$-types and our of
+generalized $\Nat$-types. However, as we will see in
+Proposition~\ref{prop:full-lan} below, this will not be possible if we
+want the set and relational interpretations of our $\Lan$-types to
+include the set and relational interpretations of the terms given by
+the $\Lan$-introduction rule. Since we do indeed want this, the upshot
+is that the above interpretations of $\Lan$-types are, in fact, the
+only possibility.
+
+We must also define the functorial actions of the above set and
+relational interpretations of $\Lan$-types on morphisms. If $f : \rho
+\to \rho'$ is a morphism of set environments, then
 \[
@@
 \]
-is defined as
+
+\vspace*{0.05in}
+
+\noindent
+is defined to be
 \begin{multline*}
@@
 \end{multline*}
-The functorial action for the relational interpretation is defined analogously.
-
-The left Kan extension can be expressed via a colimit formula~\cite{rie16} as
-\[
-	(\Lan_K\,F) \ol{A} = \colim{\ol{S} : \set^k_0, \ol{f : K\ol{S} \to A}}{F \ol{S}}
-\]
-for any $\ol{A} : \set^h$, and where $\set_0$ is the category finite sets
-(i.e., the full subcategory of $\omega$-presentable objects in $\set$).
-For any $\ol{S} : \set^k_0$ and $\ol{f : K\ol{S} \to A}$, let
-$j_{\ol{S}, \ol{f}} : F \ol{S} \to (\Lan_K\,F) \ol{A}$
-be the morphism indexed by $\ol{S}$ and $\ol{f}$ of the cocone into the colimit.
-Moreover, for any $h$-tuple of functions $\ol{g : A \to B}$,
-the functorial action $(\Lan_K\,F) \ol{g}$ is the unique function $(\Lan_K\,F) \ol{A} \to (\Lan_K\,F) \ol{B}$
-such that, for every $\ol{S} : \set^k_0$ and $\ol{f : K\ol{S} \to A}$,
-we have that $((\Lan_K\,F) \ol{g}) \circ j_{\ol{S}, \ol{f}} = j_{\ol{S}, \ol{g \circ f}}$.
-We can similarly express via a colimit formula the left Kan extension of functors on $\rel$,
-and in that case we call the cocone over it $\iota$.
-
-We define the semantics for the terms from the introduction and elimination rules for $\Lan$ types as, respectively,
+
+\vspace*{0.05in}
+
+\noindent
+The functorial action for the relational interpretations of
+$\Lan$-types is defined analogously.
+
+\subsection{Extending the Term Semantics}\label{sec:term-sem}
+
+Recalling the notation for the colimit representation of left Kan
+extensions from Section~\ref{sec:lke}, we define the set and
+relational interpretations of the terms from the introduction and
+elimination rules for $\Lan$ types from Section~\ref{sec:ext-calc} by 
+{\color{red} CAUTION: semantics don't match the rules!}
 \begin{multline*}
@@
 \]
-where
-  \[\mu_{\ol{N}} : \Lan_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho
+respectively, where
+\[\mu_{\ol{N}} : \mathit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho
   {[\ol{\alpha := \_} ]}}} \setsem{\Gamma; \Phi, \ol\alpha \vdash
@@
   \end{equation}
+{\color{red} $\eta$ vs $\epsilon$}
 as given by the universal property of the left Kan extension.
@@
 \ar[dr, "\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho {[\ol{\alpha := \_} ]}}"{left}] && \set \\ 
-&\set^{|\ol K|} \ar[ur, "\Lan_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho {[\ol{\alpha := \_} ]}}}\,
+&\set^{|\ol K|} \ar[ur, "\mathit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho {[\ol{\alpha := \_} ]}}}\,
 \setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho {[\Phi, \ol \alpha :=
@@
 Unfortunately, it turns out that, in general, this is not possible, as the following example shows.
+
+\begin{prop}\label{prop:full-lan}
+Let $F : \set^h \to \set$ and $\ol{K : \set^h \to \set}$ be functors.
+Assume that $L : \set^k \to \set$ is a functor such that the following properties hold:
+\begin{enumerate}
+\item for all $\ol{A : \set}$, we have that $L \ol{A} \subseteq (\textit{Lan}_{\ol K}\, F)\ol{A}$
+\item for all $\ol{f : A \to B}$ and $\ol{x \in A}$, we have that $(\textit{Lan}_{\ol K}\, F)\ol{f}|_{L \ol{A}}\, x \in L \ol{B}$
+\item for all $\ol{A : \set}$ and $y \in F \ol{A}$, we have that $\epsilon_{\ol{A}} y \in L ( \ol{K \ol{A}} )$
+\end{enumerate}
+Then $L = \textit{Lan}_{\ol K}\, F$.
+\end{prop}
+\begin{proof}
+For every $\ol{A : \set}$ and $z \in (\textit{Lan}_{\ol K}\, F)\ol{A}$
+there exist $\ol{S : \set_0}$, $\ol{f : K \ol{S} \to A}$ and $w \in F \ol{S}$ such that $z = \iota_{\ol{S}, \ol{f}} w$.
+Thus, $z = \iota_{\ol{S}, \ol{f}} w = (\textit{Lan}_{\ol K}\, F) \ol{f} (\epsilon_{\ol{S}} w)$;
+by the third assumption we have that $\epsilon_{\ol{S}} w \in L ( \ol{K \ol{S}} )$
+and by the second assumption we have that $z \in L \ol{A}$.
+We can than conclude that $L \ol{A} = (\textit{Lan}_{\ol K}\, F)\ol{A}$.
+\end{proof}
 

### 2020-12-24 free-theorems-nesteds/20-mscs.tex
more work on Sec 8

@@
 $\fvars^0$ disjoint from $\Phi$, and $\phi^k \in \fvars^k \setminus
-\Phi$, then the set $\mcF$ of {\em functorial expressions} over
+\Phi$, then the set $\mcF$ of {\color{red} {\em functorial expressions}} over
 $\Gamma$ and $\Phi$ are given in Definition~\ref{def:wftypes}. The
@@
 \begin{defi}\label{def:wftypes}
-The formation rules for the set $\F$ of\, {\em well-formed types} are
+The formation rules for the set $\F$ of\, {\color{red} {\em
+    well-formed types}} are
 
@@
 
-Generalized algebraic datatypes (GADTs){\color{red}~\cite{???}} generalize nested
-types by relaxing the above restriction to allow the return types of
-data constructors to be different instances of the datatype than the
-one being defined. For example, the GADT
-{\small
+Generalized algebraic datatypes (GADTs){\color{red}~\cite{???}}
+generalize nested types by relaxing the above restriction to allow the
+return types of data constructors to be different instances of the
+datatype than the one being defined. For example, the GADT {\small
 \[\begin{array}{l}
@@
 \circ \ol K$
-%are initial among such pairs of functor and natural transformations,
-%i.e.,
-such that, for every functor $L : \set^h \to \set$ and natural
-transformation $\epsilon : F \to L \circ \ol K$, there exists a unique
-natural transformation $\mu : \mathit{Lan}_{\ol K}\,F \to L$ such that
-$(\mu {\ol K}) \circ \eta = \epsilon$. This is depicted in the
+such that, for every functor $G : \set^h \to \set$ and natural
+transformation $\gamma : F \to G \circ \ol K$, there exists a unique
+natural transformation $\mu : \mathit{Lan}_{\ol K}\,F \to G$ such that
+$(\mu {\ol K}) \circ \eta = \gamma$. This is depicted in the
 following diagram:
@@
 \ar[ur, bend left, "{\mathit{Lan}_K\,F}"{name=Lan, description}]
-\ar[ur, bend right, "{L}"'{name=L, right}, ""'{name=Lr, right}]
-\ar[Rightarrow, bend left = 90, from=Fa, to=Lr, "{\epsilon}"]
+\ar[ur, bend right, "{G}"'{name=L, right}, ""'{name=Lr, right}]
+\ar[Rightarrow, bend left = 90, from=Fa, to=Lr, "{\gamma}"]
 \ar[Rightarrow, dashed, from=Lan, to=L, "{\mu}"]
@@
 expressed as
-\[(\mathit{Lan}_{\ol K} F)\ol A \;=\; \colim{(\ol{X : \,{\mathcal C}_0},\,
-  \ol{f : K{\ol X} \to A})}{F \ol X}\]
-
+\[
+(\mathit{Lan}_{\ol K} F)\ol A \;=\; \colim{\ol{X : \,{\mathcal
+        C}_0},\, \ol{f : K{\ol X} \to A}}{F \ol X}\]
 Thus, if $F : \set^k \to \set$, $\ol{K} :
@@
   the category finite sets --- we  have that
-\[(\mathit{Lan}_K F) \ol{A} = \colim{(\ol{S} : \set_0^k, \,\ol{f : K\ol{S} \to
-    A})}{F \ol{S}}\]
+\begin{equation}\label{eq:colim-form}
+(\mathit{Lan}_K F) \ol{A} = \colim{\ol{S} : \set_0^k, \,\ol{f :
+      K\ol{S} \to  A}}{F \ol{S}}
+\end{equation}
 If $\ol{S} : \set_0^k$ and $\ol{f : K\ol{S} \to A}$, let $j_{\ol{S},
   \ol{f}} : F \ol{S} \to (\mathit{Lan}_KF) \ol{A}$ be the morphism
-indexed by $\ol{S}$ and $\ol{f}$ mapping the cocone into the colimit.
-Then for any $h$-tuple of functions $\ol{g : A \to B}$, the functorial
-action $(\mathit{Lan}_K F) \ol{g}$ is the unique function from
-$(\mathit{Lan}_K F) \ol{A}$ to $(\mathit{Lan}_K F) \ol{B}$ such that,
-for every $\ol{S} : \set^k_0$ and $\ol{f : K\ol{S} \to A}$,\, 
-$(\mathit{Lan}_KF) \ol{g} \circ j_{\ol{S},\, \ol{f}} = j_{\ol{S},\,
-  \ol{g\, \circ f}}$ holds.
-
-We can similarly represent left Kan extensions of functors over $\rel$
-in terms of colimits. In that setting we will denote the morphism
-mapping the cocone into the colimit $\iota$. We will make good use of
-both of these representations in Section~\ref{sec:term-sem} below.
+indexed by $\ol{S}$ and $\ol{f}$ mapping the cocone into the colimit
+in Equation~\ref{eq:colim-form}. Then for any $h$-tuple of functions
+$\ol{g : A \to B}$, the functorial action $(\mathit{Lan}_K F) \ol{g}$
+is the unique function from $(\mathit{Lan}_K F) \ol{A}$ to
+$(\mathit{Lan}_K F) \ol{B}$ such that, for all $\ol{S} : \set^k_0$ and
+$\ol{f : K\ol{S} \to A}$,\,
+\begin{equation}\label{eq:cocone-def}
+(\mathit{Lan}_KF) \ol{g} \circ j_{\ol{S},\, \ol{f}} = j_{\ol{S},\,
+    \ol{g\, \circ f}}
+  \end{equation}
+holds.  We can similarly represent left Kan extensions of functors
+over $\rel$ in terms of colimits. In that setting we will denote the
+morphism mapping the cocone into the colimit $\iota$. We will make
+good use of both of these representations in
+Section~\ref{sec:term-sem} below.
 
@@
 \AXC{$\ol{\Gamma; \ol{\alpha} \vdash K}$}
-\BIC{$\Gamma;\emptyset~|~\Delta \vdash \int_{\ol K,F} : \Nat^{\Phi}\, F\, (\Lan^{\ol\alpha}_{\ol{K}}\,  F) \ol{K}$}
+\BIC{$\Gamma;\emptyset~|~\emptyset \vdash \int_{\ol K,F} : \Nat^{\Phi,\ol\alpha}\, F\, (\Lan^{\ol\alpha}_{\ol{K}}\,  F) \ol{K}$}
 \DisplayProof
@@
 \[
-\AXC{$\Gamma;\emptyset~|~\Delta \vdash \eta : \Nat^{\Phi,\ol\alpha}\,
+\AXC{$\Gamma;\emptyset~|~\Delta \vdash t : \Nat^{\Phi,\ol\alpha}\,
   F\;G[\ol{\beta := K}]$}
-\UIC{$\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol K}_F \eta :
+\UIC{$\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol K}_F t :
 \Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}\,F)\ol\beta\, G$}
@@
 \begin{multline*}
-\big(\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}\,
-      \setsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha := \id_{\_}}]\big)\,
+\big(\lambda \ol{S}.\,\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha
+      \vdash K}\rho[\ol{\alpha := S}]}}\,  
+      \setsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha := \id_S}]\big)\,
     \ol{\setsem{\Gamma; \Phi \vdash A}\rho'} \\
 \circ\;
-  \big(\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}\,
-      \setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := \_}]\big)\,
+  \big(\lambda \ol{S}.\,\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha
+        \vdash K}\rho[\ol{\alpha := S}]}}\,  
+      \setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := S}]\big)\,
     \ol{\setsem{\Gamma; \Phi \vdash A}f}
@@
 \begin{multline*}
-\big(\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}\,
-      \setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := \_}]\big)\,
+\big(\lambda \ol{S}.\,\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha
+      \vdash K}\rho[\ol{\alpha := S}]}}\, 
+      \setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := S}]\big)\,
     \ol{\setsem{\Gamma; \Phi \vdash A}f} \\
 \circ\;
-  \big(\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}\,
-      \setsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha := \id_{\_}}]\big)\,
+  \big(\lambda \ol{S}.\,\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha
+        \vdash K}\rho[\ol{\alpha := S}]}}\, 
+      \setsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha := \id_S}]\big)\,
     \ol{\setsem{\Gamma; \Phi \vdash A}\rho}
@@
 elimination rules for $\Lan$ types from Section~\ref{sec:ext-calc} by 
-{\color{red} CAUTION: semantics don't match the rules!}
-\begin{multline*}
-\setsem{\Gamma;\Phi~|~\Delta \vdash \int_{\ol K,F} t :
-  (\Lan^{\ol\alpha}_{\ol{K}}\, F)\, \ol{K[\ol{\alpha := A}]}}\rho\;d \\
-= j_{\ol{\setsem{\Gamma;\Phi \vdash A}\rho},
-  \ol{\id_{\setsem{\Gamma; \Phi \vdash K[\ol{\alpha := A}]}\rho}}}\,
-(\setsem{\Gamma; \Phi~|~\Delta \vdash t : F[\ol{\alpha := A}]} \rho\, d)
-\end{multline*}
+\[
+\setsem{\Gamma;\emptyset~|~\emptyset \vdash \int_{\ol K,F}
+: \Nat^{\Phi, \ol\alpha}\, F\, (\Lan^{\ol\alpha}_{\ol{K}}\,  F) \ol{K}}\rho\,d
+= \eta
+\]
 and
 \[
-\setsem{\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol K}_F \eta 
+\setsem{\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol K}_F t 
 : \Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}\,F) \ol \beta\; G}\rho
-= \lambda \ol{N} \, \ol{B}.\, (\mu_{\ol{N}})_{\ol{B}}
+= \mu
 \]
-respectively, where
-\[\mu_{\ol{N}} : \mathit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho
-  {[\ol{\alpha := \_} ]}}} \setsem{\Gamma; \Phi, \ol\alpha \vdash
-    F}\rho {[\Phi, \ol \alpha := \ol N, \_ ]} \to \setsem{\Gamma;
-    \Phi, \ol\beta \vdash G} \rho[\Phi, \ol \beta := \ol N, \_]\]
+respectively, where, for each $\ol{N} {\color{red} : ???}$,
+$\eta_{\ol{N}}$ is the natural transformation associated with the left
+Kan extension
+\[\begin{tikzcd} [column sep = large, row sep = large]
+\set^{|\ol \alpha|}
+\ar[rr, "\lambda \ol{A}.\,\setsem{\Gamma; \Phi, \ol\alpha
+    \vdash F}\rho {[\Phi, \ol \alpha := \ol{N}, \ol{A} ]}"{name=F}] 
+\ar[dr, "\ol{\lambda \ol{A}.\,\setsem{\Gamma; \ol\alpha \vdash
+      K}\rho {[\ol{\alpha := A} ]}}"{left}] && \set \\  
+&\set^{|\ol K|} \ar[ur, "\mathit{Lan}_{\ol{\lambda \ol A.\,
+      \setsem{\Gamma; \ol\alpha \vdash K}\rho {[\ol{\alpha := A}
+  ]}}}\, \lambda \ol A.\, \setsem{\Gamma; \Phi, \ol\alpha \vdash
+    F}\rho {[\Phi, \ol \alpha :=  \ol N, \ol A ]}"{right}]
+\ar[d, Rightarrow, shorten <= 2mm, shorten >= 1mm, 
+          from=F, to=2-2,  "\eta_{\ol{N}}"{}]
+\end{tikzcd}\]
+and, for all $\ol N {\color{red} : ???}$ and $d {\color{red} : ???}$,
+\[\begin{array}{l}
+\mu_{\ol{N}\, d} : \lambda \ol{A}.\,
+\mathit{Lan}_{\ol{\lambda \ol A.\,\setsem{\Gamma; \ol\alpha \vdash K}\rho 
+  {[\ol{\alpha := A} ]}}} \lambda \ol A.\,\setsem{\Gamma; \Phi,
+  \ol\alpha \vdash F}\rho {[\Phi, \ol \alpha := \ol N, \ol A ]}\\[2ex]
+\hspace*{0.8in}\to \lambda \ol B.\, \setsem{\Gamma; 
+  \Phi, \ol\beta \vdash G} \rho[\Phi, \ol \beta := \ol N, \ol B]
+\end{array}\]
 is the unique natural transformation such that
 \begin{equation}\label{eq:defn-of-mu-sub-N}
-    (\setsem{\Gamma; \emptyset \,|\, \Delta \vdash \eta : \Nat^{\Phi,
-    \ol\alpha} \, F \,\, G[\ol{\beta := K}]}\rho \, d)_{\ol{N}} =
-(\mu_{\ol{N}} \, \ol{\setsem{\Gamma; \ol\alpha \vdash K}
-  \rho[\ol{\alpha := \_}]}) \circ \epsilon_{\ol{N}}
+(\mu_{\ol{N}\, d} \, \ol{\lambda \ol A.\,\setsem{\Gamma; \ol\alpha \vdash K}
+  \rho[\ol{\alpha := A}]}) \circ \eta_{\ol{N}}
+= (\setsem{\Gamma; \emptyset \,|\, \Delta \vdash t : \Nat^{\Phi,
+  \ol\alpha} \, F \,\, G[\ol{\beta := K}]}\rho \, d)_{\ol{N}}
 \end{equation}
-{\color{red} $\eta$ vs $\epsilon$}
-as given by the universal property of the left Kan extension.
-Here, $\epsilon_{\ol{N}}$ is the natural transformation associated
-with the left Kan extension, as shown in the following diagram.
-\[\begin{tikzcd} [column sep = large, row sep = large]
-\set^{|\ol \alpha|}
-\ar[rr, "\setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho {[\Phi, \ol
-      \alpha := \ol N, \_ ]}"{name=F}] 
-\ar[dr, "\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho {[\ol{\alpha := \_} ]}}"{left}] && \set \\ 
-&\set^{|\ol K|} \ar[ur, "\mathit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho {[\ol{\alpha := \_} ]}}}\,
-\setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho {[\Phi, \ol \alpha :=
-    \ol N, \_ ]}"{right}]
-\ar[d, Rightarrow, shorten <= 2mm, shorten >= 1mm, 
-          from=F, to=2-2,  "\epsilon_{\ol{N}}"{}]
-\end{tikzcd}\]
-The relational interpretations are define analogously.
-
-Having defined the syntax and both the set and relational semantics for $\Lan$ types and terms,
-we need to check that these definitions yield a parametric model.
-The first step would be to extend Lemma~\ref{lem:rel-transf-morph} to $\Lan$ types,
-by proving a new clause of the inductive step of the proof.
-Unfortunately, it turns out that, in general, this is not possible, as the following example shows.
+as given by the universal property of the left Kan extension.  The
+relational interpretations $\relsem{\Gamma;\emptyset~|~\emptyset
+  \vdash \int_{\ol K,F} : \Nat^{\Phi, \ol\alpha}\, F\,
+  (\Lan^{\ol\alpha}_{\ol{K}}\, F) \ol{K}}$ and
+$\setsem{\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol K}_F t :
+  \Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}\,F) \ol \beta\;
+  G}$ are defined entirely analogously.
+
+\vspace*{0.05in}
+
+We conclude this subsection by showing, as promised above, that no
+cutting down of set and relational interpretations of $\Lan$-types is
+possible. We have:
 
@@
 Let $F : \set^h \to \set$ and $\ol{K : \set^h \to \set}$ be functors.
-Assume that $L : \set^k \to \set$ is a functor such that the following properties hold:
-\begin{enumerate}
-\item for all $\ol{A : \set}$, we have that $L \ol{A} \subseteq (\textit{Lan}_{\ol K}\, F)\ol{A}$
-\item for all $\ol{f : A \to B}$ and $\ol{x \in A}$, we have that $(\textit{Lan}_{\ol K}\, F)\ol{f}|_{L \ol{A}}\, x \in L \ol{B}$
-\item for all $\ol{A : \set}$ and $y \in F \ol{A}$, we have that $\epsilon_{\ol{A}} y \in L ( \ol{K \ol{A}} )$
-\end{enumerate}
-Then $L = \textit{Lan}_{\ol K}\, F$.
+If $L : \set^k \to \set$ is a functor such that 
+\begin{itemize}
+\item $L \ol{A} \subseteq
+  (\textit{Lan}_{\ol K} F)\ol{A}$ for all $\ol{A : \set}$,
+\item $(\textit{Lan}_{\ol K} F)\ol{f}\,|_{L \ol A}\; x \in L
+  \ol{B}$ for all $\ol{f : A \to B}$ and $x \in L \ol A$, and 
+\item $\eta_{\ol{A}}\, y \in L ( \ol{K \ol{A}} )$ for all $\ol{A :
+  \set}$ and $y \in F \ol{A}$, 
+\end{itemize}
+then $L = \textit{Lan}_{\ol K} F$.
 \end{prop}
 \begin{proof}
-For every $\ol{A : \set}$ and $z \in (\textit{Lan}_{\ol K}\, F)\ol{A}$
-there exist $\ol{S : \set_0}$, $\ol{f : K \ol{S} \to A}$ and $w \in F \ol{S}$ such that $z = \iota_{\ol{S}, \ol{f}} w$.
-Thus, $z = \iota_{\ol{S}, \ol{f}} w = (\textit{Lan}_{\ol K}\, F) \ol{f} (\epsilon_{\ol{S}} w)$;
-by the third assumption we have that $\epsilon_{\ol{S}} w \in L ( \ol{K \ol{S}} )$
-and by the second assumption we have that $z \in L \ol{A}$.
-We can than conclude that $L \ol{A} = (\textit{Lan}_{\ol K}\, F)\ol{A}$.
+For all $\ol{A : \set}$ and $z \in (\textit{Lan}_{\ol K} F)\ol{A}$
+there exist $\ol{S : \set_0}$, $\ol{f : K \ol{S} \to A}$, and $w \in F
+\ol{S}$ such that $z = \iota_{\ol{S}, \ol{f}}\, w$.  Thus, $z =
+\iota_{\ol{S}, \ol{f}}\, w = (\textit{Lan}_{\ol K} F) \ol{f}
+(\eta_{\ol{S}} \,w)$ by Equation~\ref{eq:cocone-def}. Then by the
+third assumption above we have that $\eta_{\ol{S}}\, w \in L (\ol{K
+  \ol{S}})$, and by the second assumption above we have that $z \in L
+\ol{A}$. This gives $(\textit{Lan}_{\ol K} F)\ol{A} \subseteq L \ol
+A$. Finally, by the first assumption above, we therefore have that $L
+\ol{A} = (\textit{Lan}_{\ol K} F)\ol{A}$.
 \end{proof}
-
-\begin{exa}
-We provide a counterexample to Lemma~\ref{lem:rel-transf-morph} for $\Lan$ types.
-Consider the type $\emptyset; \alpha \vdash (\Lan^{\emptyset}_{\onet}\, \onet) \alpha$.
-Then, using the colimit formula for left Kan extensions, we have
+{\color{red} Thus, if $L \subseteq \mathit{Lan}_{\ol K}F$ were the
+  interpretation of $\Lan$-types, and if $L$ were to contain the set
+  interpretation $\eta$ of ?????, then ????????.
+
+An analogous result holds for relational interpretations of
+$\Lan$-types.}
+
+\subsection{Parametricity Results for GADTs}
+
+Having extended our calculus with both $\Lan$-types and terms of such
+types, and having given the only sensible set and relational
+interpretations for these types and terms, we now need to verify that
+these interpretations give rise to a parametric model. The first step
+in this process is to extend Lemma~\ref{lem:rel-transf-morph} to
+$\Lan$-types by adding a clause for $\Lan$-types to the proof.
+Unfortunately, however, Lemma~\ref{lem:rel-transf-morph} does not
+extend to arbitrary $\Lan$-types, as the following example shows.
+
+\begin{exa}\label{ex:no-rel-transf}
+Consider the type $\emptyset; \alpha \vdash
+(\Lan^{\emptyset}_{\onet}\, \onet) \alpha$. The analogue for $\rel$ of
+Equation~\ref{eq:colim-form} gives
 \[
 \begin{array}{rl}
-\relsem{ \emptyset; \alpha \vdash (\Lan^{\emptyset}_{\onet}\, \onet) \alpha } \rho [\alpha := (1, 0, 0)]
-&= (\textit{Lan}_{\Eq_1}\, \Eq_1)(1, 0, 0) \\
+\relsem{ \emptyset; \alpha \vdash (\Lan^{\emptyset}_{\onet} \onet)
+  \alpha } \rho [\alpha := (1, 0, 0)] 
+&= (\textit{Lan}_{\Eq_1} \Eq_1)(1, 0, 0) \\
 &= \colim{f : \Eq_1 \to (1, 0, 0)}{\Eq_1} \\
@@
 \]
-because there exists no morphism $f : \Eq_1 \to (1, 0, 0)$ in $\rel$.
-But, considering the set interpretation on the first projection of the relation environment, we have
+Here, the last equality holds because there are no morphisms in $\rel$
+from $\Eq_1$ to $(1, 0, 0)$, and because $(0,0,0)$ is the initial
+object in that category. On the other hand, for the set interpretation
+with respect to the first projection of the relation environment
+$\rho[\alpha := (1,0,0)]$, Equation~\ref{eq:colim-form} gives
 \[
 \begin{array}{rl}
-\setsem{ \emptyset; \alpha \vdash (\Lan^{\emptyset}_{\onet}\, \onet) \alpha } ( \pi_1 ( \rho [\alpha := (1, 0, 0)] ) )
-&= \setsem{ \emptyset; \alpha \vdash (\Lan^{\emptyset}_{\onet}\, \onet) \alpha } (\pi_1 \rho) [\alpha := 1] \\
-&= (\textit{Lan}_{1}\, 1) 1 \\
+\setsem{ \emptyset; \alpha \vdash (\Lan^{\emptyset}_{\onet} \onet)
+  \alpha } ( \pi_1 ( \rho [\alpha := (1, 0, 0)] ) ) 
+&= \setsem{ \emptyset; \alpha \vdash (\Lan^{\emptyset}_{\onet} \onet)
+  \alpha } (\pi_1 \rho) [\alpha := 1] \\ 
+&= (\textit{Lan}_{1} 1) 1 \\
 &= \colim{1 \to 1}{1} \\
@@
 \]
-because there exist a unique arrow $1 \to 1$ in $\set$, which is the identity on $1$.
+because $\id_1$ is the unique arrow from $1$ to $1$ in $\set$.
+Lemma~\ref{lem:rel-transf-morph} therefore cannot hold.
 \end{exa}
 
-The issue lies in the fact that the functor along which we extend contains constants (such as $\onet$).
-Indeed, if we assume that $\ol{\emptyset; \overline{\alpha} \vdash K}$ are polynomial functors not containing constants,
-then we can prove that the relational interpretation of
-$\Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A}$
-is fibered over its set interpretation, as proven in the following proposition.
+The problem in Example~\ref{ex:no-rel-transf} lies in the fact that
+the {\color{red} functorial expression} along whose intepretation we
+extend contains constants, i.e., {\color{red} subexpressions}
+constructed from $\onet$. Indeed, if $\ol{\emptyset; \overline{\alpha}
+  \vdash K}$ consists only of polynomial (i.e., sum-of-products)
+expressions not containing constants (i.e., formed only from $+$,
+$\times$, and the variables in $\alpha$), then
+Lemma~\ref{lem:rel-transf-morph} actually does holds for $\Gamma; \Phi
+\vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A}$. This is proved in
 
 \begin{prop}
-Assume that the types in $\ol{K}$ are polynomial and contain no constants
-(i.e., they are formed only by the functorial type variables $\ol\alpha$, $+$ and $\times$).
-Then
-\[
-\pi_i (
-\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } \rho
-)
-=
-\setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } (\pi_i \rho)
-\]
-for every relation environment $\rho$ and $i = 1, 2$.
+If $\ol{\emptyset; \overline{\alpha} \vdash K}$ consists only of
+polynomial expressions not containing constants, then
+\[(\setsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A}
+}, \setsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A} },
+\relsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A} }) \in
+[\relenv,RT_0]\]
 \end{prop}
 \begin{proof}
-Notice that, by the colimit formula,
+We need only show that
+\begin{equation}\label{eq:former}
+\pi_i \,( \relsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A}
+} \rho ) = \setsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}}
+  F)\ol{A} } (\pi_i \rho)
+\end{equation}
+and
+\begin{equation}\label{eq:latter}
+\pi_i \,( \relsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A}
+} f ) = \setsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}}
+  F)\ol{A} } (\pi_i f)
+\end{equation}
+for all $\rho$, $\rho'$, $f : \rho \to \rho'$, and $i \in \{1, 2\}$.
+
+To prove Equation~\ref{eq:former} we first observe that, by (the
+analogue for $\rel$ of) Equation~\ref{eq:colim-form}, we have
 \[
-\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } \rho
-=
-\colim{\ol{R}, \ol{m : \relsem{ \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha := R}] \to \relsem{ \Gamma; \Phi \vdash A } \rho}}{
-\relsem{ \Gamma; \Phi, \ol\alpha \vdash F } \rho[\ol{\alpha := R}]
-}
+\relsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A} } \rho =
+\colim{\ol{R : \rel_0},\, \ol{m : \relsem{ \emptyset; \ol\alpha \vdash K }
+    \rho[\ol{\alpha := R}] \to \relsem{ \Gamma; \Phi \vdash A }
+    \rho}}{ \relsem{ \Gamma; \Phi, \ol\alpha \vdash F }
+  \rho[\ol{\alpha := R}] }
 \]
-(where we call $\iota$ the cocone of the colimit)
-and thus
+where $\iota$ is the morphism mapping the cocone into the colimit.
+Since each projection $\pi_i$ is cocontinuous, we therefore have that 
 \[
-\pi_i (
-\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } \rho
-)
-=
-\colim{\ol{R : \rel_0}, \ol{m : \relsem{ \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha := R}] \to \relsem{ \Gamma; \Phi \vdash A } \rho}}{
-\relsem{ \Gamma; \Phi, \ol\alpha \vdash F } (\pi_i \rho) [\ol{\alpha := \pi_i R}]}
+\pi_i\, ( \relsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}}
+  F)\ol{A} } \rho ) = \colim{\ol{R : \rel_0}, \ol{m : \relsem{
+      \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha := R}] \to
+    \relsem{ \Gamma; \Phi \vdash A } \rho}}{ \relsem{ \Gamma; \Phi,
+    \ol\alpha \vdash F } (\pi_i \rho) [\ol{\alpha := \pi_i R}]}
 \]
-(where the cocone of the colimit is $\pi_i \iota$),
-whereas, again by the colimit formula,
+with $\pi_i \iota$ mapping this cocone into this colimit.  On the
+other hand, Equation~\ref{eq:colim-form} also gives
 \[
-\setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } (\pi_i \rho)
-=
-\colim{\ol{S : \set_0}, \ol{f : \setsem{ \emptyset; \ol\alpha \vdash K } (\pi_i \rho)[\ol{\alpha := S}] \to \setsem{ \Gamma; \Phi \vdash A } (\pi_i \rho)}}{
-\setsem{ \Gamma; \Phi, \ol\alpha \vdash F } (\pi_i \rho) [\ol{\alpha := S}]}
+\setsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A} }
+(\pi_i \rho) = \colim{\ol{S : \set_0}, \ol{f : \setsem{ \emptyset;
+      \ol\alpha \vdash K } (\pi_i \rho)[\ol{\alpha := S}] \to \setsem{
+      \Gamma; \Phi \vdash A } (\pi_i \rho)}}{ \setsem{ \Gamma; \Phi,
+    \ol\alpha \vdash F } (\pi_i \rho) [\ol{\alpha := S}]}
 \]
-(where we call $j$ the cocone of the colimit).
-
-We now prove that
+where $j$ is the morphism mapping the cocone into the colimit.
+
+When $i = 1$ we prove that
 \[
-\pi_1 (
-\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } \rho
-)
-=
-\setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } (\pi_1 \rho)
+\pi_1 \,( \relsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}}
+  F)\ol{A} } \rho ) = \setsem{ \Gamma; \Phi \vdash (\Lan^{\ol
+    \alpha}_{\ol{K}} F)\ol{A} } (\pi_1 \rho)
 \]
-by providing a pair of inverse functions
-(the proof for the second projection is analogous).
-Define
+by
+providing a pair of inverse functions; the proof when $i=2$ is
+entirely analogous. To this end, we define
 \[
-h :
-\pi_1 (
-\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } \rho
-)
-\to
-\setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } (\pi_1 \rho)
+h : \pi_1\, ( \relsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}}
+  F)\ol{A} } \rho ) \to \setsem{ \Gamma; \Phi \vdash (\Lan^{\ol
+    \alpha}_{\ol{K}} F)\ol{A} } (\pi_1 \rho)
 \]
-as the unique function given by the universal property of the colimit such that, for any $\ol{R : \rel_0}$ and $\ol{m : \relsem{ \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha := R}] \to \relsem{ \Gamma; \Phi \vdash A } \rho}$, we have that $h \circ (\pi_1 \iota_{\ol{R}, \ol{m}}) = j_{\ol{\pi_1 R}, \ol{\pi_1 m}}$.
-
-Observe that, since $K$ does not contain constants, then $\setsem{ \emptyset; \ol\alpha \vdash K } (\pi_2 \rho) [\ol{\alpha := 0}] = 0$.
-For any
+to be the unique function given by the universal property of the
+colimit representation of $h$'s domain. That is, we take $h$ to be the
+unique morphism such that, for any $\ol{R : \rel_0}$ and $\ol{m :
+  \relsem{ \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha := R}] \to
+  \relsem{ \Gamma; \Phi \vdash A } \rho}$, we have $h \circ (\pi_1
+\iota_{\ol{R},\, \ol{m}}) = j_{\ol{\pi_1 R}, \,\ol{\pi_1 m}}$.
+
+Now, since $K$ does not contain constants, $\setsem{ \emptyset;
+  \ol\alpha \vdash K } (\pi_2 \rho) [\ol{\alpha := 0}] = 0$. Thus, for
+any $\ol{S : \set_0}$ and $\ol{f : \setsem{ \emptyset; \ol\alpha
+    \vdash K } (\pi_1 \rho)[\ol{\alpha := S}] \to \setsem{ \Gamma;
+    \Phi \vdash A } (\pi_1 \rho)} $ we have relations $\ol{(S, 0, 0)}$
+and morphisms $\ol{(f, !) : \relsem{ \emptyset; \ol\alpha \vdash K }
+  \rho[\ol{\alpha := (S, 0, 0)}] \to \relsem{ \Gamma; \Phi \vdash A }
+  \rho}$, where, for any set $X$, we suppress the sub- and
+superscripts and write $!$ for the unique morphism $!^0_X$ from $0$ to
+$X$. We then define
 \[
-\ol{S : \set_0}
-\qquad \text{and} \qquad
-\ol{f : \setsem{ \emptyset; \ol\alpha \vdash K } (\pi_1 \rho)[\ol{\alpha := S}] \to \setsem{ \Gamma; \Phi \vdash A } (\pi_1 \rho)}
+k : \setsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A}
+} (\pi_1 \rho) \to \pi_1 \,( \relsem{ \Gamma; \Phi \vdash (\Lan^{\ol
+    \alpha}_{\ol{K}} F)\ol{A} } \rho )
 \]
-consider the relations $\ol{(S, 0, 0)}$ and the morphisms $\ol{(f, !) : \relsem{ \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha := (S, 0, 0)}] \to \relsem{ \Gamma; \Phi \vdash A } \rho}$ (where $!$ is the unique function from the empty set).
-Then, define
+to be the unique function given by the universal property of the
+colimit representation of $k$'s domain. That is, we take $k$ to be the
+unique morphism such that, for any $\ol{S : \set_0}$ and $\ol{f :
+  \setsem{ \emptyset; \ol\alpha \vdash K } (\pi_1 \rho)[\ol{\alpha :=
+      S}] \to \setsem{ \Gamma; \Phi \vdash A } (\pi_1 \rho)}$, we have
+$k \circ j_{\ol{S},\, \ol{f}} = \pi_1 \iota_{\ol{(S, 0, 0)},\, \ol{(f,
+    !)}}$.
+
+To see that $h$ and $k$ are mutually inverse we first observe that
 \[
-k :
-\setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } (\pi_1 \rho)
-\to
-\pi_1 (
-\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } \rho
-)
+h \circ k \circ j_{\ol{S},\, \ol{f}}
+= h \circ (\pi_1 \iota_{\ol{(S, 0, 0)}, \,\ol{(f, !)}})
+= j_{\ol{\pi_1 (S, 0, 0)}, \,\ol{\pi_1 (f, !)}}
+= j_{\ol{S},\, \ol{f}}
 \]
-as the unique function given by the universal property of the colimit such that, for any $\ol{S : \set_0}$ and $\ol{f : \setsem{ \emptyset; \ol\alpha \vdash K } (\pi_1 \rho)[\ol{\alpha := S}] \to \setsem{ \Gamma; \Phi \vdash A } (\pi_1 \rho)}$, we have that $k \circ j_{\ol{S}, \ol{f}} = \pi_1 \iota_{\ol{(S, 0, 0)}, \ol{(f, !)}}$.
-
-We now have to show that $h$ and $k$ are mutually inverse.
-First, we prove that $h \circ k = \id$.
-Indeed, observe that
+for all $\ol{S : \set_0}$ and $\ol{f : \setsem{ \emptyset; \ol\alpha
+    \vdash K } (\pi_1 \rho)[\ol{\alpha := S}] \to \setsem{ \Gamma;
+    \Phi \vdash A } (\pi_1 \rho)}$, so that $h \circ k = \id$.
+To see that  $k \circ h  = \id$ we first observe that
 \[
-h \circ k \circ j_{\ol{S}, \ol{f}}
-= h \circ (\pi_1 \iota_{\ol{(S, 0, 0)}, \ol{(f, !)}})
-= j_{\ol{\pi_1 (S, 0, 0)}, \ol{\pi_1 (f, !)}}
-= j_{\ol{S}, \ol{f}}
+k \circ h \circ (\pi_1 \iota_{\ol{R},\, \ol{m}}) = k \circ
+j_{\ol{\pi_1 R}, \,\ol{\pi_1 m}} = \pi_1 \iota_{\ol{(\pi_1 R, 0,
+    0)},\, \ol{(\pi_1 m, !)}}
 \]
-for every $\ol{S : \set_0}$ and
-$\ol{f : \setsem{ \emptyset; \ol\alpha \vdash K } (\pi_1 \rho)[\ol{\alpha := S}] \to \setsem{ \Gamma; \Phi \vdash A } (\pi_1 \rho)}$
-and thus $h \circ k = \id$.
-Secondly, we prove that $k \circ h = \id$.
-Indeed, observe that
+for all $\ol{R : \rel_0}$ and $\ol{m : \relsem{ \emptyset; \ol\alpha
+    \vdash K } \rho[\ol{\alpha := R}] \to \relsem{ \Gamma; \Phi \vdash
+    A } \rho}$. Then note that, for each sequence of morphisms
+$\ol{(\id_{\pi_1 R}, !) : (\pi_1 R, 0, 0) \to R}$, we have that $\ol m
+\,\circ\, \relsem{ \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha :=
+    (\id_{\pi_1 R}, !)}] = (\ol{\pi_1 m}, !)$ in the indexing category
+for the colimit representation of the codomain of $k$. This in turn
+implies that {\color{red} $\iota_{\ol{R},\, \ol{m}} =
+  \iota_{\ol{(\pi_1 R, 0, 0)},\, \ol{(\pi_1 m, !)}}$}, so that $k
+\circ h \circ (\pi_1 \iota_{\ol{R},\, \ol{m}}) = \pi_1
+\iota_{\ol{R},\, \ol{m}}$ for all $\ol R$ and $\ol m$, and thus $h
+\circ k = \id$.
+
+To prove Equation~\ref{eq:latter}, let $\iota$ be the cocone for the
+colimit $\relsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}}
+  F)\ol{A} } \rho$, let $j$ be the cocone for the colimit $\setsem{
+  \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A} }
+(\pi_i\rho)$, and let $h$ and $k$ be the isomorphisms showing that
+$\pi_i(\setsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}}
+  F)\ol{A} } \rho) = \setsem{ \Gamma; \Phi \vdash (\Lan^{\ol
+    \alpha}_{\ol{K}} F)\ol{A} } (\pi_i\rho)$. Analogously, let
+$\iota'$, $j'$, $h'$ and $k'$ be as above but for interpretations with
+respect to $\rho'$ rather than $\rho$.  By definition, $\relsem{
+  \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A} } f$ is the
+unique morphism of relations such that, for any $\ol{R : \rel_0}$ and
+$\ol{m : \relsem{ \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha :=
+      R}] \to \relsem{ \Gamma; \Phi \vdash A } \rho}$,
 \[
-k \circ h \circ (\pi_1 \iota_{\ol{R}, \ol{m}})
-= k \circ j_{\ol{\pi_1 R}, \ol{\pi_1 m}}
-= \pi_1 \iota_{\ol{(\pi_1 R, 0, 0)}, \ol{(\pi_1 m, !)}}
+(\relsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A} }
+f) \circ \iota_{\ol{R},\, \ol{m}} = \iota'_{\ol{R},\, \ol{\relsem{ \Gamma;
+      \Phi \vdash A } f \, \circ\, m}}
 \]
-for every $\ol{R : \rel_0}$ and $\ol{m : \relsem{ \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha := R}] \to \relsem{ \Gamma; \Phi \vdash A } \rho}$.
-Moreover, consider the morphisms $\ol{(\id_{\pi_1 R}, !) : (\pi_1 R, 0, 0) \to R}$ and observe that $m \circ \relsem{ \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha := (\id_{\pi_1 R}, !)}] = (\pi_1 m, !)$, implying that $\iota_{\ol{R}, \ol{m}} = \iota_{\ol{(\pi_1 R, 0, 0)}, \ol{(\pi_1 m, !)}}$.
-Then $k \circ h \circ (\pi_1 \iota_{\ol{R}, \ol{m}}) = \pi_1 \iota_{\ol{R}, \ol{m}}$ for every $R$ and $m$, and thus $h \circ k = \id$.
-
-We can also prove that
+holds. Projecting onto the $i^{th}$ component of both sides thus gives
+that
 \[
-\pi_i (
-\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } f
-)
-=
-\setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } (\pi_i f)
+\pi_i (\relsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}}
+  F)\ol{A} } f) \circ \pi_i \iota_{\ol{R}, \,\ol{m}} = \pi_i
+\iota'_{\ol{R},\, \ol{\relsem{ \Gamma; \Phi \vdash A } f \,\circ\, m}}
 \]
-for every morphism of relation environments $f : \rho \to \rho'$ and $i = 1, 2$.
-Let $\iota$ be the cocone for the colimit $\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } \rho$,
-$j$ the cocone for the colimit $\setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } (\pi_i\rho)$
-and $h$ and $k$ the isomorphisms showing that
-$\pi_i(\setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } \rho) = \setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } (\pi_i\rho)$;
-analogously, let $\iota'$, $j'$, $h'$ and $k'$ be as above but for interpretations in $\rho'$ in place of $\rho$.
-Then $\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } f$
-is the unique morphism of relations such that, for any $\ol{R : \rel_0}$ and
-$\ol{m : \relsem{ \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha := R}] \to \relsem{ \Gamma; \Phi \vdash A } \rho}$,
-we have that
+Now, using the facts from above that $h \circ (\pi_i \iota_{\ol{R},
+  \,\ol{m}}) = j_{\ol{\pi_i R},\, \ol{\pi_i m}}$ and, analogously,
+that $h' \circ (\pi_i \iota'_{\ol{R},\, \ol{m}}) = j'_{\ol{\pi_i R},
+  \,\ol{\pi_i m}}$, we get
+{\color{red}
 \[
-(\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } f) \circ \iota_{\ol{R}, \ol{m}}
-= \iota'_{\ol{R}, \ol{\relsem{ \Gamma; \Phi \vdash A } f \circ m}}
+\pi_i (\relsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}}
+  F)\ol{A} } f) \circ k \circ j_{\ol{\pi_i R},\, \ol{\pi_i m}} = k'
+\circ j'_{\ol{\pi_i R},\, \ol{\relsem{ \Gamma; \Phi \vdash A } (\pi_i f)
+  \, \circ\, (\pi_i m)}}
 \]
-but then, by projecting the $i$-th component, we have that
-\[
-\pi_i (\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } f) \circ \pi_i \iota_{\ol{R}, \ol{m}}
-= \pi_i \iota'_{\ol{R}, \ol{\relsem{ \Gamma; \Phi \vdash A } f \circ m}}
-\]
-and thus, using the facts that $h \circ (\pi_i \iota_{\ol{R}, \ol{m}}) = j_{\ol{\pi_i R}, \ol{\pi_i m}}$
-and $h' \circ (\pi_i \iota'_{\ol{R}, \ol{m}}) = j'_{\ol{\pi_i R}, \ol{\pi_i m}}$ from above,
-we get
-\[
-\pi_i (\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } f) \circ k \circ j_{\ol{\pi_i R}, \ol{\pi_i m}}
-= k' \circ j'_{\ol{\pi_i R}, \ol{\relsem{ \Gamma; \Phi \vdash A } (\pi_i f) \circ (\pi_i m)}}
-\]
-i.e., $\pi_i (\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } f)$
-satisfy the universal property of $\setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } (\pi_i f)$,
-and thus they are equal (up to natural isomorphism).
+i.e., $\pi_i (\relsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}}
+  F)\ol{A} } f)$ satisfies the universal property for $\setsem{
+  \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A} } (\pi_i
+f)$.} We therefore have that $\pi_i (\relsem{ \Gamma; \Phi \vdash
+  (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A} } f) = \setsem{ \Gamma; \Phi
+  \vdash (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A} } (\pi_i f)$ (up to
+natural isomorphism) as desired.
 \end{proof}
 
-The requirement that the $\ol{K}$s be polynomial functors without constants is quite restrictive,
-as it precludes the possibility of expressing GADTs used in practice, such as the following type $\mathtt{Expr}$ of typed expressions.
+The requirement that the {\color{red} expressions} in $\ol{K}$ be
+constant-free polynomials is quite restrictive. In fact, it precludes
+the expression of many GADTs commonly used in practice, such as the
+following GADT $\mathtt{Expr}$ of typed expressions:
 \[
@@
 
-{\color{red} This is not quite right.}
-Even when extending along polynomials containing no constants, it is unclear how to prove the Identity Extension Lemma for $\Lan$ types, or if it holds at all.
-In order to satisfy the identity extension lemma in the case of $\Nat$ types, for example, it is necessary to cut down the semantic interpretation by requiring that the natural transformations in the set interpretation preserve equalities.
-The impossibility of restricting the semantic interpretation of $\Lan$, as proved by Proposition~\ref{prop:full-lan}, suggests that the Identity Extension Lemma might indeed not hold for a type system that features such types.
-
-We must then conclude that it is not possible to use functorial semantics to provide a parametric model for $\Lan$ types.
+{\color{red} CHECK!} Even when extending along constant-free
+polynomials, it is unclear how to prove the Identity Extension Lemma
+for $\Lan$-types or, indeed, whether it holds at all in their
+presence. To prove the IEL for $\Nat$-types, for example, it is
+necessary to cut down the semantic interpretation by requiring that
+the natural transformations in the set interpretation preserve
+equalities. The impossibility of restricting the semantic
+interpretations of $\Lan$-types, as shown in
+Proposition~\ref{prop:full-lan}, suggests that the IEL may fail for
+them. We have been able neither to prove nor disprove it thus far, but
+our substantial and lengthy investigations into the issue lead us to
+conjecture that it is not possible to define a functorial semantics
+for $\Lan$-types that gives rise to parametric models for languages
+supporting even constant-free polynomial primitive GADTs.
 

### 2020-12-26 free-theorems-nesteds/20-mscs.tex
Repair proof of Prop 8.4

@@
 indexed by $\ol{S}$ and $\ol{f}$ mapping the cocone into the colimit
-in Equation~\ref{eq:colim-form}. Then for any $h$-tuple of functions
+in Equation~\ref{eq:colim-form}.
+%Then for any $h$-tuple of functions
+%$\ol{g : A \to B}$, the functorial action $(\mathit{Lan}_K F) \ol{g}$
+%is the unique function from $(\mathit{Lan}_K F) \ol{A}$ to
+%$(\mathit{Lan}_K F) \ol{B}$ such that, for all $\ol{S} : \set^k_0$ and
+%$\ol{f : K\ol{S} \to A}$,\,
+%\begin{equation}\label{eq:cocone-def}
+%(\mathit{Lan}_KF) \ol{g} \circ j_{\ol{S},\, \ol{f}} = j_{\ol{S},\,
+%    \ol{g\, \circ f}}
+%  \end{equation}
+%holds, where
+For any $h$-tuple of functions
 $\ol{g : A \to B}$, the functorial action $(\mathit{Lan}_K F) \ol{g}$
@@
 \begin{equation}\label{eq:cocone-def}
-(\mathit{Lan}_KF) \ol{g} \circ j_{\ol{S},\, \ol{f}} = j_{\ol{S},\,
-    \ol{g\, \circ f}}
+(\mathit{Lan}_KF) \ol{g} \circ j_{\ol{S},\, \ol{f}}
+= j'_{\ol{S},\, \ol{g\, \circ f}}
+: F \ol{S} \to (\mathit{Lan}_KF) \ol{B}
 \end{equation}
-holds.  We can similarly represent left Kan extensions of functors
-over $\rel$ in terms of colimits. In that setting we will denote the
-morphism mapping the cocone into the colimit $\iota$. We will make
-good use of both of these representations in
-Section~\ref{sec:term-sem} below.
+holds, where $j$ is the morphism mapping the cocone into
+$(\mathit{Lan}_K F) \ol{A} = \colim{\ol{S} : \set_0^k, \,\ol{f :
+    K\ol{S} \to A}}{F \ol{S}}$ and $j'$ is the morphism mapping the
+cocone into $(\mathit{Lan}_K F) \ol{B} = \colim{\ol{S} : \set_0^k,
+  \,\ol{f : K\ol{S} \to B}}{F \ol{S}}$.  Moreover, if $\alpha : F \to
+F'$ is a natural transformation, then $\mathit{Lan}_K \alpha :
+(\mathit{Lan}_K F) \to (\mathit{Lan}_K F')$ is defined to be the
+induced natural transformation whose component $(\mathit{Lan}_K
+\alpha) \ol{A} : (\mathit{Lan}_K F) \ol{A} \to (\mathit{Lan}_K F')
+\ol{A}$ is the unique function such that $(\mathit{Lan}_K \alpha)
+\ol{A} \,\circ\, j_{\ol{S}, \ol{f}} = j'_{\ol{S}, \ol{f}} \,\circ\,
+\alpha_{\ol{S}}$. {\color{red} where $j$ is the cocone into
+  $(\mathit{Lan}_K F) \ol{A}$ and $j'$ is the cocone into
+  $(\mathit{Lan}_K F') \ol{A}$. REDUNDANT?} We can similarly
+represent left Kan extensions of functors over $\rel$ in terms of
+colimits. In that setting we will denote the morphism mapping the
+cocone into the colimit $\iota$. We will make good use of both of
+these representations in Section~\ref{sec:term-sem} below.
 
@@
   \setsem{\Gamma;\Phi \vdash
-    (\Lan^{\ol{\alpha}}_{\ol{K}}\,F)\ol{A}}\rho = (\lambda \ol{S}.\,
-  \textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}
-      \rho[\ol{\alpha := S}]}}\, \setsem{\Gamma; \Phi, \ol\alpha
+    (\Lan^{\ol{\alpha}}_{\ol{K}}\,F)\ol{A}}\rho = (
+  \textit{Lan}_{\lambda \ol{S}.\,\ol{\setsem{\Gamma; \ol\alpha \vdash K}
+      \rho[\ol{\alpha := S}]}}\, \lambda \ol{S}.\,\setsem{\Gamma; \Phi, \ol\alpha
     \vdash F} \rho[\ol{\alpha := S}])\, \ol{\setsem{\Gamma; \Phi
@@
   \relsem{\Gamma;\Phi \vdash
-    (\Lan^{\ol{\alpha}}_{\ol{K}}\,F)\ol{A}}\rho = (\lambda \ol R.\,
-    \textit{Lan}_{\ol{\relsem{\Gamma; \ol\alpha \vdash
-          K}\rho[\ol{\alpha := R}]}}\, \relsem{\Gamma; \Phi, \ol\alpha
-      \vdash F}\rho[\ol{\alpha := R}])\, \ol{\relsem{\Gamma; \Phi
-        \vdash A}\rho}
+    (\Lan^{\ol{\alpha}}_{\ol{K}}\,F)\ol{A}}\rho = (
+  \textit{Lan}_{\lambda \ol R.\,\ol{\relsem{\Gamma; \ol\alpha \vdash
+        K}\rho[\ol{\alpha := R}]}}\, \lambda \ol R.\,\relsem{\Gamma;
+    \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := R}])\,
+  \ol{\relsem{\Gamma; \Phi \vdash A}\rho}
 \]
@@
 \begin{multline*}
-\big(\lambda \ol{S}.\,\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha
+\big(\textit{Lan}_{\lambda \ol{S}.\,\ol{\setsem{\Gamma; \ol\alpha
       \vdash K}\rho[\ol{\alpha := S}]}}\,  
-      \setsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha := \id_S}]\big)\,
+      \lambda \ol{S}.\,\setsem{\Gamma; \Phi, \ol\alpha \vdash
+        F}f[\ol{\alpha := \id_S}]\big)\, 
     \ol{\setsem{\Gamma; \Phi \vdash A}\rho'} \\
 \circ\;
-  \big(\lambda \ol{S}.\,\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha
+  \big(\textit{Lan}_{\lambda \ol{S}.\,\ol{\setsem{\Gamma; \ol\alpha
         \vdash K}\rho[\ol{\alpha := S}]}}\,  
-      \setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := S}]\big)\,
+      \lambda \ol{S}.\,\setsem{\Gamma; \Phi, \ol\alpha \vdash
+        F}\rho[\ol{\alpha := S}]\big)\, 
     \ol{\setsem{\Gamma; \Phi \vdash A}f}
@@
 \begin{multline*}
-\big(\lambda \ol{S}.\,\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha
+\big(\textit{Lan}_{\lambda \ol{S}.\,\ol{\setsem{\Gamma; \ol\alpha
       \vdash K}\rho[\ol{\alpha := S}]}}\, 
-      \setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := S}]\big)\,
+      \lambda \ol{S}.\,\setsem{\Gamma; \Phi, \ol\alpha \vdash
+        F}\rho'[\ol{\alpha := S}]\big)\, 
     \ol{\setsem{\Gamma; \Phi \vdash A}f} \\
 \circ\;
-  \big(\lambda \ol{S}.\,\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha
+  \big(\textit{Lan}_{\lambda \ol{S}.\,\ol{\setsem{\Gamma; \ol\alpha
         \vdash K}\rho[\ol{\alpha := S}]}}\, 
-      \setsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha := \id_S}]\big)\,
+     \lambda \ol{S}.\, \setsem{\Gamma; \Phi, \ol\alpha \vdash
+       F}f[\ol{\alpha := \id_S}]\big)\, 
     \ol{\setsem{\Gamma; \Phi \vdash A}\rho}
@@
 Lemma~\ref{lem:rel-transf-morph} actually does holds for $\Gamma; \Phi
-\vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A}$. This is proved in
+\vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A}$. This is proved in the
+following proposition, which covers the case of
+Lemma~\ref{lem:rel-transf-morph} for such $\Lan$-types.
 
@@
   \relsem{ \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha := R}] \to
-  \relsem{ \Gamma; \Phi \vdash A } \rho}$, we have $h \circ (\pi_1
-\iota_{\ol{R},\, \ol{m}}) = j_{\ol{\pi_1 R}, \,\ol{\pi_1 m}}$.
+  \relsem{ \Gamma; \Phi \vdash A } \rho}$, we have
+\begin{equation}\label{eq:cocone-proj}
+h \circ (\pi_1 \iota_{\ol{R},\, \ol{m}}) = j_{\ol{\pi_1 R},
+  \,\ol{\pi_1 m}}
+\end{equation}
 
@@
 To see that  $k \circ h  = \id$ we first observe that
-\[
+\begin{equation}\label{eq:k-comp-h}
 k \circ h \circ (\pi_1 \iota_{\ol{R},\, \ol{m}}) = k \circ
@@
     0)},\, \ol{(\pi_1 m, !)}}
-\]
+\end{equation}
 for all $\ol{R : \rel_0}$ and $\ol{m : \relsem{ \emptyset; \ol\alpha
@@
     A } \rho}$. Then note that, for each sequence of morphisms
-$\ol{(\id_{\pi_1 R}, !) : (\pi_1 R, 0, 0) \to R}$, we have that $\ol m
-\,\circ\, \relsem{ \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha :=
-    (\id_{\pi_1 R}, !)}] = (\ol{\pi_1 m}, !)$ in the indexing category
-for the colimit representation of the codomain of $k$. This in turn
-implies that {\color{red} $\iota_{\ol{R},\, \ol{m}} =
-  \iota_{\ol{(\pi_1 R, 0, 0)},\, \ol{(\pi_1 m, !)}}$}, so that $k
-\circ h \circ (\pi_1 \iota_{\ol{R},\, \ol{m}}) = \pi_1
-\iota_{\ol{R},\, \ol{m}}$ for all $\ol R$ and $\ol m$, and thus $h
-\circ k = \id$.
-
-To prove Equation~\ref{eq:latter}, let $\iota$ be the cocone for the
-colimit $\relsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}}
-  F)\ol{A} } \rho$, let $j$ be the cocone for the colimit $\setsem{
-  \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A} }
-(\pi_i\rho)$, and let $h$ and $k$ be the isomorphisms showing that
-$\pi_i(\setsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}}
-  F)\ol{A} } \rho) = \setsem{ \Gamma; \Phi \vdash (\Lan^{\ol
-    \alpha}_{\ol{K}} F)\ol{A} } (\pi_i\rho)$. Analogously, let
-$\iota'$, $j'$, $h'$ and $k'$ be as above but for interpretations with
-respect to $\rho'$ rather than $\rho$.  By definition, $\relsem{
-  \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A} } f$ is the
-unique morphism of relations such that, for any $\ol{R : \rel_0}$ and
-$\ol{m : \relsem{ \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha :=
-      R}] \to \relsem{ \Gamma; \Phi \vdash A } \rho}$,
+$\ol{(\id_{\pi_1 R}, !) : (\pi_1 R, 0, 0) \to R}$, we have that
+\[\ol{m
+\,\circ\, \relsem{ \emptyset; \ol\alpha \vdash K } \id_\rho[\ol{\alpha
+    := (\id_{\pi_1 R}, !)}] = (\pi_1 m, !)}\]
+in the indexing category for the colimit representation of the
+codomain of $k$. This implies that
+\[\iota_{\ol{R},\, \ol{m}} \,\circ\, \relsem{ \Gamma; \Phi,
+  \ol\alpha \vdash F } \id_{\rho}[\ol{\alpha := (\id_{\pi_1 R}, !)}]
+= \iota_{\ol{(\pi_1 R, 0, 0)},\, \ol{(\pi_1 m, !)}}\] Projecting the
+first component thus gives
+\[(\pi_1 \iota_{\ol{R},\,
+  \ol{m}}) \, \circ\, \pi_1 ( \relsem{ \Gamma; \Phi, \ol\alpha \vdash
+  F } \id_{\rho}[\ol{\alpha := (\id_{\pi_1 R}, !)}]) = \pi_1
+\iota_{\ol{(\pi_1 R, 0, 0)},\, \ol{(\pi_1 m, !)}}\]
+Now, the induction hypothesis on $\Gamma;\Phi,\ol\alpha
+\vdash F$ gives that
+\[\begin{array}{lll}
+ &  & \pi_1 ( \relsem{ \Gamma; \Phi, \ol\alpha \vdash F}
+\id_{\rho}[\ol{\alpha := (\id_{\pi_1 R}, !)}])\\
+& = & \setsem{ \Gamma; \Phi, \ol\alpha \vdash F } \id_{\pi_1
+  \rho}[\ol{\alpha := \id_{\pi_1 R}}]\\
+& = & \id_{\setsem{ \Gamma; \Phi, \ol\alpha \vdash F } (\pi_1
+  \rho)[\ol{\alpha := \pi_1 R}]}
+\end{array}\]
+so that, in fact, $\pi_1 \iota_{\ol R,\,\ol m} =
+\pi_1\iota_{\ol{(\pi_1 R, 0, 0)},\, \ol{(\pi_1 m, !)}}$.  Finally, by
+Equation~\ref{eq:k-comp-h} we have $k \circ h \circ (\pi_1
+\iota_{\ol{R},\, \ol{m}}) = \pi_1 \iota_{\ol{(\pi_1 R, 0, 0)},\,
+  \ol{(\pi_1 m, !)}}  = \pi_1 \iota_{\ol{R},\, \ol{m}}$ for all $\ol
+R$ and $\ol m$, and thus $k \circ h = \id$.
+
+To prove Equation~\ref{eq:latter}, first recall that
+$\relsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A} }
+f$ is defined to be
+\begin{multline*}
+\big(\textit{Lan}_{\ol{\lambda \ol{R}.\,\relsem{\Gamma; \ol\alpha
+      \vdash K}\rho[\ol{\alpha := R}]}}\, \lambda
+\ol{R}.\,\relsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha :=
+    \id_R}]\big)\, \ol{\relsem{\Gamma; \Phi \vdash A}\rho'} \\ \circ\;
+\big(\textit{Lan}_{\ol{\lambda \ol{R}.\,\relsem{\Gamma; \ol\alpha
+      \vdash K}\rho[\ol{\alpha := R}]}}\, \lambda
+\ol{R}.\,\relsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha :=
+    R}]\big)\, \ol{\relsem{\Gamma; \Phi \vdash A}f}
+\end{multline*}
+and $\setsem{\Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A}
+} (\pi_i f)$ is defined to be
+\begin{multline*}
+\big(\textit{Lan}_{\ol{\lambda \ol{S}.\,\setsem{\Gamma; \ol\alpha
+      \vdash K}(\pi_i \rho)[\ol{\alpha := S}]}}\, \lambda
+\ol{S}.\,\setsem{\Gamma; \Phi, \ol\alpha \vdash F}(\pi_i f)[\ol{\alpha
+    := \id_S}]\big)\, \ol{\setsem{\Gamma; \Phi \vdash A}(\pi_i
+  \rho')}\\ \circ\; \big(\textit{Lan}_{\ol{\lambda
+    \ol{S}.\,\setsem{\Gamma; \ol\alpha \vdash K}(\pi_i
+    \rho)[\ol{\alpha := S}]}}\, \lambda \ol{S}.\,\setsem{\Gamma; \Phi,
+  \ol\alpha \vdash F}(\pi_i \rho)[\ol{\alpha := S}]\big)\,
+\ol{\relsem{\Gamma; \Phi \vdash A}(\pi_i f)}
+\end{multline*}
+It therefore suffices to show that
+\begin{multline}\label{eq:part-one}
+\pi_i \Big( \big(\textit{Lan}_{\ol{\lambda \ol{R}.\,\relsem{\Gamma; \ol\alpha
+      \vdash K}\rho[\ol{\alpha := R}]}}\,  
+      \lambda \ol{R}.\,\relsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha := \id_R}]\big)\,
+    \ol{\relsem{\Gamma; \Phi \vdash A}\rho'} \Big) \\
+= \big(\textit{Lan}_{\ol{\lambda \ol{S}.\,\setsem{\Gamma; \ol\alpha
+      \vdash K} (\pi_1 \rho) [\ol{\alpha := S}]}}\,  
+      \lambda \ol{S}.\,\setsem{\Gamma; \Phi, \ol\alpha \vdash F}(\pi_i f)[\ol{\alpha := \id_S}]\big)\,
+    \ol{\setsem{\Gamma; \Phi \vdash A}(\pi_i \rho')}
+\end{multline}
+and that
+\begin{multline}\label{eq:part-two}
+\pi_i \Big( \big(\textit{Lan}_{\ol{\lambda \ol{R}.\,\relsem{\Gamma; \ol\alpha
+        \vdash K}\rho[\ol{\alpha := R}]}}\,  
+      \lambda \ol{R}.\,\relsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := R}]\big)\,
+    \ol{\relsem{\Gamma; \Phi \vdash A}f} \Big) \\
+= \big(\textit{Lan}_{\ol{\lambda \ol{S}.\,\setsem{\Gamma; \ol\alpha
+        \vdash K} (\pi_i \rho) [\ol{\alpha := S}]}}\,  
+      \lambda \ol{S}.\,\setsem{\Gamma; \Phi, \ol\alpha \vdash F} (\pi_i \rho) [\ol{\alpha := S}]\big)\,
+    \ol{\setsem{\Gamma; \Phi \vdash A} (\pi_i f)}
+\end{multline}
+
+\vspace*{0.1in}
+
+To prove Equation~\ref{eq:part-one} we first observe that
+  the universal property (Equation~\ref{eq:cocone-def}) of
 \[
-(\relsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A} }
-f) \circ \iota_{\ol{R},\, \ol{m}} = \iota'_{\ol{R},\, \ol{\relsem{ \Gamma;
-      \Phi \vdash A } f \, \circ\, m}}
+\big(\textit{Lan}_{\ol{\lambda \ol{R}.\,\relsem{\Gamma; \ol\alpha
+      \vdash K}\rho[\ol{\alpha := R}]}}\, \lambda
+\ol{R}.\,\relsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha :=
+    \id_R}]\big)\, \ol{\relsem{\Gamma; \Phi \vdash A}\rho'}
 \]
-holds. Projecting onto the $i^{th}$ component of both sides thus gives
-that
+gives that, for all $\ol{R : \rel_0}$ and $\ol{m : \relsem{\Gamma;
+    \ol\alpha \vdash K}\rho[\ol{\alpha := R}] \to \relsem{\Gamma; \Phi
+    \vdash A}\rho'}$,
+\[\begin{array}{lll}
+& & \big(\textit{Lan}_{\ol{\lambda \ol{R}.\,\relsem{\Gamma; \ol\alpha
+      \vdash K}\rho[\ol{\alpha := R}]}}\, \lambda
+\ol{R}.\,\relsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha :=
+    \id_R}]\big)\, \ol{\relsem{\Gamma; \Phi \vdash A}\rho'} \,\circ\,
+\iota_{\ol{R}, \ol{m}}\\[2ex]
+& = & \iota'_{\ol{R}, \ol{m}} \,\circ\, \relsem{\Gamma; \Phi,
+  \ol\alpha \vdash F}f[\ol{\alpha := \id_R}]
+\end{array}\]
+Here, $\iota$ is the morphism mapping the cocone into the colimit
 \[
-\pi_i (\relsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}}
-  F)\ol{A} } f) \circ \pi_i \iota_{\ol{R}, \,\ol{m}} = \pi_i
-\iota'_{\ol{R},\, \ol{\relsem{ \Gamma; \Phi \vdash A } f \,\circ\, m}}
+\big(\textit{Lan}_{\ol{\lambda \ol{R}.\,\relsem{\Gamma; \ol\alpha
+      \vdash K}\rho[\ol{\alpha := R}]}}\, \lambda
+\ol{R}.\,\relsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha :=
+    R}]\big)\, \ol{\relsem{\Gamma; \Phi \vdash A}\rho'}
 \]
-Now, using the facts from above that $h \circ (\pi_i \iota_{\ol{R},
-  \,\ol{m}}) = j_{\ol{\pi_i R},\, \ol{\pi_i m}}$ and, analogously,
-that $h' \circ (\pi_i \iota'_{\ol{R},\, \ol{m}}) = j'_{\ol{\pi_i R},
-  \,\ol{\pi_i m}}$, we get
-{\color{red}
+and $\iota'$ is the morphism mapping the cocone into the colimit
 \[
-\pi_i (\relsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}}
-  F)\ol{A} } f) \circ k \circ j_{\ol{\pi_i R},\, \ol{\pi_i m}} = k'
-\circ j'_{\ol{\pi_i R},\, \ol{\relsem{ \Gamma; \Phi \vdash A } (\pi_i f)
-  \, \circ\, (\pi_i m)}}
+\big(\textit{Lan}_{\ol{\lambda \ol{R}.\,\relsem{\Gamma; \ol\alpha
+      \vdash K}\rho[\ol{\alpha := R}]}}\, \lambda
+\ol{R}.\,\relsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha :=
+    R}]\big)\, \ol{\relsem{\Gamma; \Phi \vdash A}\rho'}
 \]
-i.e., $\pi_i (\relsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}}
-  F)\ol{A} } f)$ satisfies the universal property for $\setsem{
-  \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A} } (\pi_i
-f)$.} We therefore have that $\pi_i (\relsem{ \Gamma; \Phi \vdash
-  (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A} } f) = \setsem{ \Gamma; \Phi
-  \vdash (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A} } (\pi_i f)$ (up to
-natural isomorphism) as desired.
+Projecting, together with the induction hypothesis for $\Gamma; \Phi,
+\ol\alpha\vdash F$, thus gives
+\[\begin{array}{lll}
+ & & \pi_i \Big( \big(\textit{Lan}_{\ol{\lambda \ol{R}.\,\relsem{\Gamma;
+      \ol\alpha \vdash K}\rho[\ol{\alpha := R}]}}\, \lambda
+\ol{R}.\,\relsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha :=
+    \id_R}]\big)\, \ol{\relsem{\Gamma; \Phi \vdash A}\rho'} \Big)
+\,\circ\, \pi_i \iota_{\ol{R}, \ol{m}} \\[2ex]
+& = & \pi_i \iota'_{\ol{R},
+  \ol{m}} \,\circ\, \setsem{\Gamma; \Phi, \ol\alpha \vdash F} (\pi_i
+f)[\ol{\alpha := \id_{\pi_i R}}]
+\end{array}\]
+Equation~\ref{eq:cocone-proj} {\color{red} and its analogue for
+  $\iota'$ and $j'$ then give --- HOW?????}
+\[\begin{array}{lll}
+ & & \pi_i \Big( \big(\textit{Lan}_{\ol{\lambda
+    \ol{R}.\,\relsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha :=
+        R}]}}\, \lambda \ol{R}.\,\relsem{\Gamma; \Phi, \ol\alpha
+  \vdash F}f[\ol{\alpha := \id_R}]\big)\, \ol{\relsem{\Gamma; \Phi
+    \vdash A}\rho'} \Big) \,\circ\, {\color{red} j?}f_{\ol{\pi_i R},
+  \ol{\pi_i m}} \\[2ex]
+& = & j'_{\ol{\pi_i R}, \ol{\pi_i m}} \,\circ\, \setsem{\Gamma; \Phi,
+      \ol\alpha \vdash F} (\pi_i f)[\ol{\alpha := \id_{\pi_i R}}]
+\end{array}\]
+where $j$ is the morphism mapping the cocone into the colimit
+\[
+\big(\textit{Lan}_{\ol{\lambda \ol{S}.\,\setsem{\Gamma; \ol\alpha
+      \vdash K}(\pi_i \rho)[\ol{\alpha := S}]}}\, \lambda
+\ol{S}.\,\setsem{\Gamma; \Phi, \ol\alpha \vdash F}(\pi_i
+\rho)[\ol{\alpha := S}]\big)\, \ol{\setsem{\Gamma; \Phi \vdash
+    A}(\pi_i \rho')}
+\]
+and $j'$ is the morphism mapping the cocone into the colimit
+\[
+\big(\textit{Lan}_{\ol{\lambda \ol{S}.\,\setsem{\Gamma; \ol\alpha
+      \vdash K}\rho[\ol{\alpha := S}]}}\, \lambda
+\ol{R}.\,\setsem{\Gamma; \Phi, \ol\alpha \vdash F}(\pi_i
+\rho')[\ol{\alpha := S}]\big)\, \ol{\setsem{\Gamma; \Phi \vdash
+    A}(\pi_i \rho')}
+\]
+By the surjectivity of each $\pi_i : \rel \to \set$ we have that
+\[\begin{array}{lll}
+& & \pi_i \Big( \big(\textit{Lan}_{\ol{\lambda
+    \ol{R}.\,\relsem{\Gamma; 
+      \ol\alpha \vdash K}\rho[\ol{\alpha := R}]}}\, \lambda
+\ol{R}.\,\relsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha :=
+    \id_R}]\big)\, \ol{\relsem{\Gamma; \Phi \vdash A}\rho'} \Big)
+\,\circ\, j_{\ol{S}, \ol{g}} \\[2ex]
+& = & j'_{\ol{S}, \ol{g}} \,\circ\, \setsem{\Gamma; \Phi, \ol\alpha
+  \vdash F} (\pi_i f)[\ol{\alpha := \id_{S}}]
+\end{array}\]
+for all $\ol{S: \set_0}$ and $\ol{g : \setsem{\Gamma; \ol\alpha \vdash
+    K}(\pi_i \rho)[\ol{\alpha := S}] \to \setsem{\Gamma; \Phi \vdash
+    A}(\pi_i \rho')}$.  That is, \[\pi_i \Big(
+\big(\textit{Lan}_{\ol{\lambda \ol{R}.\,\relsem{\Gamma; \ol\alpha
+      \vdash K}\rho[\ol{\alpha := R}]}}\, \lambda
+\ol{R}.\,\relsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha :=
+    \id_R}]\big)\, \ol{\relsem{\Gamma; \Phi \vdash A}\rho'} \Big)\]
+satisfies the universal property (Equation~\ref{eq:cocone-def}) of 
+\[
+\big(\textit{Lan}_{\ol{\lambda \ol{S}.\,\setsem{\Gamma; \ol\alpha
+      \vdash K} (\pi_1 \rho) [\ol{\alpha := S}]}}\, \lambda
+\ol{S}.\,\setsem{\Gamma; \Phi, \ol\alpha \vdash F}(\pi_i f)[\ol{\alpha
+    := \id_S}]\big)\, \ol{\setsem{\Gamma; \Phi \vdash A}(\pi_i \rho')}
+\]
+The two expressions must therefore be equal, and thus
+Equation~\ref{eq:part-one} holds.
+
+\vspace*{0.1in}
+
+To prove Equation~\ref{eq:part-two} we first observe that 
+  the universal property (Equation~\ref{eq:cocone-def}) of 
+\[
+\big(\textit{Lan}_{\ol{\lambda \ol{R}.\,\relsem{\Gamma; \ol\alpha
+      \vdash K}\rho[\ol{\alpha := R}]}}\, \lambda
+\ol{R}.\,\relsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha :=
+    R}]\big)\, \ol{\relsem{\Gamma; \Phi \vdash A}f}
+\] gives that,
+for every $\ol{R : \rel_0}$ and $\ol{m : \relsem{\Gamma; \ol\alpha
+    \vdash K}\rho[\ol{\alpha := R}] \to \relsem{\Gamma; \Phi \vdash
+    A}\rho}$, 
+\[\begin{array}{lll}
+ & & \big(\textit{Lan}_{\ol{\lambda \ol{R}.\,\relsem{\Gamma; \ol\alpha
+      \vdash K}\rho[\ol{\alpha := R}]}}\, \lambda
+\ol{R}.\,\relsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha :=
+    R}]\big)\, \ol{\relsem{\Gamma; \Phi \vdash A}f} \,\circ\,
+\iota_{\ol{R}, \ol{m}}\\[2ex]
+& = & \iota'_{\ol{R}, \ol{\relsem{\Gamma; \Phi \vdash A}f \,\circ\,
+    m}}
+\end{array}\]
+where $\iota$ is the cocone into the colimit
+\[
+\big(\textit{Lan}_{\ol{\lambda \ol{R}.\,\relsem{\Gamma; \ol\alpha
+      \vdash K}\rho[\ol{\alpha := R}]}}\, \lambda
+\ol{R}.\,\relsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha :=
+    R}]\big)\, \ol{\relsem{\Gamma; \Phi \vdash A}\rho}
+\]
+and $\iota'$ is the cocone into the colimit
+\[\big(\textit{Lan}_{\ol{\lambda \ol{R}.\,\relsem{\Gamma; \ol\alpha
+ \vdash K}\rho[\ol{\alpha := R}]}}\, \lambda \ol{R}.\,\relsem{\Gamma;
+  \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := R}\big)\,
+  \ol{\relsem{\Gamma; \Phi \vdash A}\rho'}
+\]
+Projecting, together with the induction hypothesis for
+$\Gamma;\Phi,\ol\alpha\vdash F$, thus gives
+\[\begin{array}{lll}
+ & & \pi_i \Big( \big(\textit{Lan}_{\ol{\lambda
+    \ol{R}.\,\relsem{\Gamma; \ol\alpha 
+    \vdash K}\rho[\ol{\alpha := R}]}}\,  
+    \lambda \ol{R}.\,\relsem{\Gamma; \Phi, \ol\alpha \vdash
+      F}\rho[\ol{\alpha := R}]\big)\, 
+    \ol{\relsem{\Gamma; \Phi \vdash A}f} \Big)
+\,\circ\, \pi_i \iota_{\ol{R}, \ol{m}} \\[2ex]
+& = & \pi_i \iota'_{\ol{R}, \ol{\relsem{\Gamma; \Phi \vdash A}f \,\circ\,
+    m}} 
+\end{array}\]
+{\color{red} Equation~\ref{eq:cocone-proj} and its analogue for
+  $\iota'$ and $j'$ then give}
+\[\begin{array}{lll}
+ & & \pi_i \Big( \big(\textit{Lan}_{\ol{\lambda \ol{R}.\,\relsem{\Gamma;
+      \ol\alpha 
+    \vdash K}\rho[\ol{\alpha := R}]}}\,   
+    \lambda \ol{R}.\,\relsem{\Gamma; \Phi, \ol\alpha \vdash
+      F}\rho[\ol{\alpha := R}]\big)\, 
+    \ol{\relsem{\Gamma; \Phi \vdash A}f} \Big)
+\,\circ\, j_{\ol{\pi_i R}, \ol{\pi_i m}} \\[2ex]
+& = & j'_{\ol{\pi_i R}, \ol{\relsem{\Gamma; \Phi \vdash A}(\pi_i f)
+    \,\circ\, \pi_i m}} 
+\end{array}\]
+where $j$ is the cocone into the colimit
+\[
+\big(\textit{Lan}_{\ol{\lambda \ol{S}.\,\setsem{\Gamma; \ol\alpha
+      \vdash K}(\pi_i \rho)[\ol{\alpha := S}]}}\, \lambda
+\ol{S}.\,\setsem{\Gamma; \Phi, \ol\alpha \vdash F}(\pi_i
+\rho)[\ol{\alpha := S}]\big)\, \ol{\setsem{\Gamma; \Phi \vdash
+    A}(\pi_i \rho)}
+\]
+and $j'$ is the cocone into the colimit
+\[
+\big(\textit{Lan}_{\ol{\lambda \ol{S}.\,\setsem{\Gamma; \ol\alpha
+      \vdash K}(\pi_i \rho)[\ol{\alpha := S}]}}\, \lambda
+\ol{S}.\,\setsem{\Gamma; \Phi, \ol\alpha \vdash F}(\pi_i
+\rho)[\ol{\alpha := S}]\big)\, \ol{\relsem{\Gamma; \Phi \vdash
+    A}(\pi_i \rho')}
+\]
+By the surjectivity of each $\pi_i : \rel \to \set$ we have that
+\[\begin{array}{lll}
+ & & \pi_i \Big( \big(\textit{Lan}_{\ol{\lambda
+    \ol{R}.\,\relsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha :=
+        R}]}}\, \lambda \ol{R}.\,\relsem{\Gamma; \Phi, \ol\alpha
+  \vdash F}\rho[\ol{\alpha := R}]\big)\, \ol{\relsem{\Gamma; \Phi
+    \vdash A}f} \Big) \,\circ\, j_{\ol{S}, \ol{g}} \\[2ex]
+& = & j'_{\ol{S}, \ol{\relsem{\Gamma; \Phi \vdash A}(\pi_i f) \,\circ\, g}}
+\end{array}\]
+for every $\ol{S: \set_0}$ and $\ol{g : \setsem{\Gamma; \ol\alpha
+    \vdash K}(\pi_i \rho)[\ol{\alpha := S}] \to \setsem{\Gamma; \Phi
+    \vdash A}(\pi_i \rho)}$. That is,
+\[\pi_i \Big( \big(\textit{Lan}_{\ol{\lambda
+    \ol{R}.\,\relsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha :=
+        R}]}}\, \lambda \ol{R}.\,\relsem{\Gamma; \Phi, \ol\alpha
+  \vdash F}\rho[\ol{\alpha := R}]\big)\, \ol{\relsem{\Gamma; \Phi
+    \vdash A}f} \Big)\]
+satisfies the universal property (Equation~\ref{eq:cocone-def}) of
+\[
+\big(\textit{Lan}_{\ol{\lambda \ol{S}.\,\setsem{\Gamma; \ol\alpha
+      \vdash K} (\pi_i \rho) [\ol{\alpha := S}]}}\, \lambda
+\ol{S}.\,\setsem{\Gamma; \Phi, \ol\alpha \vdash F} (\pi_i \rho)
+   [\ol{\alpha := S}]\big)\, \ol{\setsem{\Gamma; \Phi \vdash A} (\pi_i
+     f)}
+\]
+The two expressions must therefore be equal, and thus
+Equation~\ref{eq:part-two} holds.
 \end{proof}

### 2020-12-26 free-theorems-nesteds/20-mscs.tex
finished Sec 8

@@
 
-To prove Equation~\ref{eq:part-one} we first observe that
-  the universal property (Equation~\ref{eq:cocone-def}) of
+To prove Equation~\ref{eq:part-one} we first observe that the
+morphisms $\iota'_{\ol{R}, \ol{m}} \,\circ\, \relsem{\Gamma; \Phi,
+  \ol\alpha \vdash F}f[\ol{\alpha := \id_R}]$ form a cocone for
+\[
+\big(\textit{Lan}_{\ol{\lambda \ol{R}.\,\relsem{\Gamma; \ol\alpha
+      \vdash K}\rho[\ol{\alpha := R}]}}\, \lambda
+\ol{R}.\,\relsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha :=
+    R}]\big)\, \ol{\relsem{\Gamma; \Phi \vdash A}\rho'}
+\]
+with vertex 
+\[
+\big(\textit{Lan}_{\ol{\lambda \ol{R}.\,\relsem{\Gamma; \ol\alpha
+      \vdash K}\rho[\ol{\alpha := R}]}}\, \lambda
+\ol{R}.\,\relsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha :=
+    R}]\big)\, \ol{\relsem{\Gamma; \Phi \vdash A}\rho'}
+\]
+The universal property (Equation~\ref{eq:cocone-def}) of
 \[
@@
 \]
-gives that, for all $\ol{R : \rel_0}$ and $\ol{m : \relsem{\Gamma;
-    \ol\alpha \vdash K}\rho[\ol{\alpha := R}] \to \relsem{\Gamma; \Phi
-    \vdash A}\rho'}$,
+therefore gives that, for all $\ol{R : \rel_0}$ and $\ol{m :
+  \relsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := R}] \to
+  \relsem{\Gamma; \Phi \vdash A}\rho'}$,
 \[\begin{array}{lll}
@@
 \end{array}\]
-Equation~\ref{eq:cocone-proj} {\color{red} and its analogue for
-  $\iota'$ and $j'$ then give --- HOW?????}
+Equation~\ref{eq:cocone-proj} and its analogue for
+  $\iota'$ and $j'$ then give 
 \[\begin{array}{lll}
@@
   \vdash F}f[\ol{\alpha := \id_R}]\big)\, \ol{\relsem{\Gamma; \Phi
-    \vdash A}\rho'} \Big) \,\circ\, {\color{red} j?}f_{\ol{\pi_i R},
+    \vdash A}\rho'} \Big) \,\circ \, k \,\circ\, j_{\ol{\pi_i R},
   \ol{\pi_i m}} \\[2ex]
-& = & j'_{\ol{\pi_i R}, \ol{\pi_i m}} \,\circ\, \setsem{\Gamma; \Phi,
+&  = & \pi_i \Big( \big(\textit{Lan}_{\ol{\lambda \ol{R}.\,\relsem{\Gamma;
+      \ol\alpha \vdash K}\rho[\ol{\alpha := R}]}}\, \lambda
+\ol{R}.\,\relsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha :=
+    \id_R}]\big)\, \ol{\relsem{\Gamma; \Phi \vdash A}\rho'} \Big)
+\,\circ\, k \,\circ\, h \,\circ \,\pi_i \iota_{\ol{R}, \ol{m}} \\[2ex]
+& = & \pi_i \Big( \big(\textit{Lan}_{\ol{\lambda \ol{R}.\,\relsem{\Gamma;
+      \ol\alpha \vdash K}\rho[\ol{\alpha := R}]}}\, \lambda
+\ol{R}.\,\relsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha :=
+    \id_R}]\big)\, \ol{\relsem{\Gamma; \Phi \vdash A}\rho'} \Big)
+\,\circ\, \pi_i \iota_{\ol{R}, \ol{m}} \\[2ex]
+& = & \pi_i \iota'_{\ol{R},
+  \ol{m}} \,\circ\, \setsem{\Gamma; \Phi, \ol\alpha \vdash F} (\pi_i
+f)[\ol{\alpha := \id_{\pi_i R}}]\\[2ex]
+& = & k' \, \circ \, h'\,\circ\,\pi_i \iota'_{\ol{R},
+  \ol{m}} \,\circ\, \setsem{\Gamma; \Phi, \ol\alpha \vdash F} (\pi_i
+f)[\ol{\alpha := \id_{\pi_i R}}]\\[2ex]
+& = & k' \,\circ \, j'_{\ol{\pi_i R}, \ol{\pi_i m}} \,\circ\, \setsem{\Gamma; \Phi,
       \ol\alpha \vdash F} (\pi_i f)[\ol{\alpha := \id_{\pi_i R}}]
@@
 \]
+Since both $k$ and $k'$ are isomorphisms, we have that, up to
+isomorphism, 
+\[\begin{array}{lll}
+& & \pi_i \Big( \big(\textit{Lan}_{\ol{\lambda
+    \ol{R}.\,\relsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha :=
+        R}]}}\, \lambda \ol{R}.\,\relsem{\Gamma; \Phi, \ol\alpha
+  \vdash F}f[\ol{\alpha := \id_R}]\big)\, \ol{\relsem{\Gamma; \Phi
+    \vdash A}\rho'} \Big) \circ\, j_{\ol{\pi_i R},
+  \ol{\pi_i m}} \\[2ex]
+& = & j'_{\ol{\pi_i R}, \ol{\pi_i m}} \,\circ\, \setsem{\Gamma; \Phi,
+      \ol\alpha \vdash F} (\pi_i f)[\ol{\alpha := \id_{\pi_i R}}]
+\end{array}\]
 By the surjectivity of each $\pi_i : \rel \to \set$ we have that
@@
 \end{array}\]
-{\color{red} Equation~\ref{eq:cocone-proj} and its analogue for
-  $\iota'$ and $j'$ then give}
+Equation~\ref{eq:cocone-proj} and its analogue for
+  $\iota'$ and $j'$ then give that, up to isomorphism,
 \[\begin{array}{lll}

### 2020-12-26 free-theorems-nesteds/20-mscs.tex
finished Sec 8 except for text

@@
   interpretation of $\Lan$-types, and if $L$ were to contain the set
-  interpretation $\eta$ of ?????, then ????????.
+  interpretation $\eta$ of ?????, then ????????.}
 
 An analogous result holds for relational interpretations of
-$\Lan$-types.}
+$\Lan$-types.
 

### 2020-12-29 free-theorems-nesteds/20-mscs.tex
finished draft journal paper

@@
 verifies all standard consequences of parametricity in the presence of
-primitive nested types. In particular, we give several examples
-illustrating how how our model can be used to derive useful free
-theorems for programs over nested types. {\color{red} Add info about
-  GADTs and implementation.}
+primitive nested types. We give several concrete examples illustrating
+how our model can be used to derive useful free theorems, including a
+short cut fusion transformation, for programs over nested
+types. Finally, we consider generalizing our results to GADTs, and
+argue that no extension of our parametric model for nested types can
+give a functorial interpretation of GADTs in terms of left Kan
+extensions and still be parametric.
 \end{abstract}
@@
 map-rearrangement free theorems as instances. In
-Section~\ref{sec:thms} we prove that our model satisfies an
+Section~\ref{sec:abs-and-nat} we prove that our model satisfies an
 Abstraction Theorem (Theorem~\ref{thm:abstraction}), which we use to
-derive other parametricity results that go beyond naturality. We
-conclude in Section~\ref{sec:conclusion}.
+derive, in Section~\ref{sec:ftnt}, other parametricity results that go
+beyond mere naturality. In Section~\ref{sec:GADTs} we show that the
+results of this paper do not extend to arbitrary GADTs, and argue that
+no extension of the parametric model we construct here for nested
+types can give a functorial interpretation of GADTs in terms of left
+Kan extensions as proposed in~\cite{jp19} and still be parametric. We
+conclude and offer some directions for future work in
+Section~\ref{sec:conclusion}.
+
+This paper is an extended version of the conference
+paper~\cite{jgj21}. Compared to that paper, this one includes all
+proofs, as well as more examples. The discussion of GADTs in
+Section~\ref{sec:GADTs} is also entirely new.
 
@@
 
-{\color{red} Say extended version of FoSSaCS paper. Say what's
-  new/different. GADTs, implementation.
-
-Can bind variables of arity greater than 0 in Nat-types? Improves
-expressibility?  Worth it?}
-
 \section{The Calculus}\label{sec:calculus}
@@
 $\fvars^0$ disjoint from $\Phi$, and $\phi^k \in \fvars^k \setminus
-\Phi$, then the set $\mcF$ of {\color{red} {\em functorial expressions}} over
-$\Gamma$ and $\Phi$ are given in Definition~\ref{def:wftypes}. The
-notation there entails that an application $F F_1...F_k$ is allowed
-only when $F$ is a type variable of arity $k$, or $F$ is a
-subexpression of the form $\mu \phi^{k}.\lambda
-\alpha_1...\alpha_k.F'$. Moreover, if $F$ has arity $k$ then $F$ must
-be applied to exactly $k$ arguments.  Accordingly, an overbar
-indicates a sequence of subexpressions whose length matches the arity
-of the type applied to it.  Requiring that types are always in such
-\emph{$\eta$-long normal form} avoids having to consider
-$\beta$-conversion at the level of types. In a subexpression
+\Phi$, then the set $\mcF$ of well-formed types is given in
+Definition~\ref{def:wftypes}. The notation there entails that an
+application $F F_1...F_k$ is allowed only when $F$ is a type variable
+of arity $k$, or $F$ is a subexpression of the form $\mu
+\phi^{k}.\lambda \alpha_1...\alpha_k.F'$. Moreover, if $F$ has arity
+$k$ then $F$ must be applied to exactly $k$ arguments.  Accordingly,
+an overbar indicates a sequence of types whose length matches the
+arity of the type applied to it.  Requiring that types are always in
+such \emph{$\eta$-long normal form} avoids having to consider
+$\beta$-conversion at the level of types. In a type
 $\Nat^{\ol{\alpha}}F\,G$, the $\Nat$ operator binds all occurrences of
@@
 $\lambda$ operator binds all occurrences of the variables in
-$\ol{\alpha}$, in the body $F$.
+$\ol{\alpha}$, in the body $F$. {\color{red} Can bind variables of
+  arity greater than 0 in Nat-types even just for nesteds? Improves
+  expressibility? Worth it? Worth mentioning?}
 
@@
 \begin{defi}\label{def:wftypes}
-The formation rules for the set $\F$ of\, {\color{red} {\em
-    well-formed types}} are
+The formation rules for the set $\F$ of\, {\em (well-formed) types}
+are
 
@@
 well as all occurrences of $x$ in $t$. In the rule for $t_{\ol K} s$
-there is one functorial expression in $\ol K$ for every functorial
+there is one type in $\ol K$ for every functorial
 variable in $\ol \alpha$. In the rule for $\map^{\ol{F},\ol{G}}_H$
-there is one functorial expression $F$ and one functorial expression
+there is one type $F$ and one type
 $G$ for each functorial variable in $\ol\phi$. Moreover, for each
 $\phi^k$ in $\ol\phi$ the number of functorial variables in $\ol\beta$
-in the judgments for its corresponding functorial expresssions $F$ and
+in the judgments for its corresponding type $F$ and
 $G$ is $k$. In the rules for $\tin_H$ and $\fold^F_H$, the functorial
@@
 We interpret types in our calculus as $\omega$-cocontinuous functors
-on locally finitely presentable categories~\cite{ar94}. The categories
-$\set$ and $\rel$ are both locally finitely presentable. Since functor
+on locally finitely presentable categories~\cite{ar94}. Both $\set$
+and $\rel$ are locally finitely presentable categories. Since functor
 categories of locally finitely presentable categories are again
-locally finitely presentable, this ensures that the fixpoints
-interpreting $\mu$-types in $\set$ and $\rel$ exist, and thus that
-both the set and relational interpretations of all of the types in
-Definition~\ref{def:wftypes} are well-defined~\cite{jp19}. To
-bootstrap this process, we interpret type variables as
-$\omega$-cocontinuous functors in Definitions~\ref{def:set-env}
-and~\ref{def:reln-env}. If $\C$ and $\D$ are locally finitely
-presentable categories, we write $[\C,\D]$ for the category of
-$\omega$-cocontinuous functors from $\C$ to $\D$.
+locally finitely presentable, the fixpoints interpreting $\mu$-types
+in $\set$ and $\rel$ must all exist, and thus the set and relational
+interpretations of all of the types in Definition~\ref{def:wftypes},
+are well-defined~\cite{jp19}. To bootstrap this process, we interpret
+type variables as $\omega$-cocontinuous functors in
+Definitions~\ref{def:set-env} and~\ref{def:reln-env}. If $\C$ and $\D$
+are locally finitely presentable categories, we write $[\C,\D]$ for
+the category of $\omega$-cocontinuous functors from $\C$ to $\D$.
 
@@
 mapping each set environment to itself. This gives a category of set
-environments and morphisms between them, which we denote $\setenv$.
+environments and morphisms between them, denoted $\setenv$.
 \end{defi}
@@
 morphism mapping each relation environment to itself. This gives a
-category of relation environments and morphisms between them, which we
-denote $\relenv$.
+category of relation environments and morphisms between them, denoted
+$\relenv$.
 \end{defi}
@@
 
-\section{Naturality and the Abstraction Theorem}\label{sec:free-theorems}
+\section{Naturality and the Abstraction Theorem}\label{sec:abs-and-nat}
 
@@
 
-Generalized algebraic datatypes (GADTs){\color{red}~\cite{???}}
-generalize nested types by relaxing the above restriction to allow the
-return types of data constructors to be different instances of the
-datatype than the one being defined. For example, the GADT {\small
+Generalized algebraic datatypes (GADTs) --- also known as guarded
+recursive data types or first-class phantom types --- generalize
+nested types by relaxing the above restriction to allow the return
+types of data constructors to be different instances of the datatype
+than the one being defined. For example, the GADT {\small
 \[\begin{array}{l}
@@
 parametric model when some classes of primitive GADTs are incorporated
-in this manner. However, as we argue in{\color{red}~????} below, such
-extension is not possible in general.
+in this manner. However, as we argue at the end of this section, such
+extension is likely not possible.
 
@@
 \ol{A} \,\circ\, j_{\ol{S}, \ol{f}} = j'_{\ol{S}, \ol{f}} \,\circ\,
-\alpha_{\ol{S}}$. {\color{red} where $j$ is the cocone into
-  $(\mathit{Lan}_K F) \ol{A}$ and $j'$ is the cocone into
-  $(\mathit{Lan}_K F') \ol{A}$. REDUNDANT?} We can similarly
-represent left Kan extensions of functors over $\rel$ in terms of
-colimits. In that setting we will denote the morphism mapping the
-cocone into the colimit $\iota$. We will make good use of both of
+\alpha_{\ol{S}}$, where $j$ is the cocone into $(\mathit{Lan}_K F)
+\ol{A}$ and $j'$ is the cocone into $(\mathit{Lan}_K F') \ol{A}$. We
+can similarly represent left Kan extensions of functors over $\rel$ in
+terms of colimits. In that setting we will denote the morphism mapping
+the cocone into the colimit $\iota$. We will make good use of both of
 these representations in Section~\ref{sec:term-sem} below.
@@
 $\Nat$ types from Section~\ref{sec:type-interp} as follows:
+{\color{red} Relate $k$ to arities in $\Phi$.}
 \begin{align*}
@@
 \]
-respectively, where, for each $\ol{N} {\color{red} : ???}$,
-$\eta_{\ol{N}}$ is the natural transformation associated with the left
-Kan extension
+respectively, where, for each $\ol{N : RT_k}$ {\color{red} Relate
+  $k$ to arities in $\Phi$.}, $\eta_{\ol{N}}$ is the natural
+transformation associated with the left Kan extension
+
+\vspace*{-0.1in}
+
 \[\begin{tikzcd} [column sep = large, row sep = large]
@@
 \end{tikzcd}\]
-and, for all $\ol N {\color{red} : ???}$ and $d {\color{red} : ???}$,
+and, for all $\ol{N : RT_k}$ {\color{red} Relate $k$ to arities in
+  $\Phi$.} and $d : \setsem{\Gamma;\emptyset \vdash \Delta}$, 
 \[\begin{array}{l}
@@
 \begin{prop}\label{prop:full-lan}
-Let $F : \set^h \to \set$ and $\ol{K : \set^h \to \set}$ be functors.
+Let $F : \set^h \to \set$ and $\ol{K : \set^h \to \set^k}$ be functors.
 If $L : \set^k \to \set$ is a functor such that 
@@
 \end{proof}
-{\color{red} Thus, if $L \subseteq \mathit{Lan}_{\ol K}F$ were the
-  interpretation of $\Lan$-types, and if $L$ were to contain the set
-  interpretation $\eta$ of ?????, then ????????.}
-
-An analogous result holds for relational interpretations of
-$\Lan$-types.
+Thus, {\color{red} CHECK!} if $L$ were a restriction of
+$\setsem{\Gamma; \Phi,\ol\alpha \vdash (\Lan^{\ol\alpha}_{\ol K}
+  F){\ol K}}\rho$, if the functorial action of $L$ were a restriction
+of that of $\setsem{\Gamma; \Phi,\ol\alpha \vdash
+  (\Lan^{\ol\alpha}_{\ol K} F){\ol K}}\rho$, and if
+$L(\ol{\setsem{\Gamma;\ol\alpha \vdash K}\rho \ol N \, \ol A})$ were
+to contain $\setsem{\Gamma;\emptyset~|~\emptyset \vdash \int_{\ol K,F}
+  : \Nat^{\Phi, \ol\alpha}\, F\, (\Lan^{\ol\alpha}_{\ol{K}}\, F)
+  \ol{K}}\rho\,d\,\ol N\,\ol A\,y$ for all $\ol N$, $\ol A$, and $y$,
+then $L$ would have to be the entirety of $\mathit{Lan}_{\ol{\lambda
+    \ol A.\, \setsem{\Gamma; \ol\alpha \vdash K}\rho {[\ol{\alpha :=
+          A} ]}}}\, \lambda \ol A.\, \setsem{\Gamma; \Phi, \ol\alpha
+  \vdash F}\rho {[\Phi, \ol \alpha := \ol N, \ol A ]}$. An analogous
+result holds for relational interpretations of $\Lan$-types.
 
@@
 The problem in Example~\ref{ex:no-rel-transf} lies in the fact that
-the {\color{red} functorial expression} along whose intepretation we
-extend contains constants, i.e., {\color{red} subexpressions}
-constructed from $\onet$. Indeed, if $\ol{\emptyset; \overline{\alpha}
-  \vdash K}$ consists only of polynomial (i.e., sum-of-products)
-expressions not containing constants (i.e., formed only from $+$,
-$\times$, and the variables in $\alpha$), then
+the type along whose intepretation we extend contains constants, i.e.,
+subtypes constructed from $\onet$. Indeed, if $\ol{\emptyset;
+  \overline{\alpha} \vdash K}$ consists only of polynomial (i.e.,
+sum-of-products) types not containing constants (i.e., formed only
+from $+$, $\times$, and the variables in $\alpha$), then
 Lemma~\ref{lem:rel-transf-morph} actually does holds for $\Gamma; \Phi
@@
 If $\ol{\emptyset; \overline{\alpha} \vdash K}$ consists only of
-polynomial expressions not containing constants, then
+polynomial types not containing constants, then
 \[(\setsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A}
@@
 
-The requirement that the {\color{red} expressions} in $\ol{K}$ be
-constant-free polynomials is quite restrictive. In fact, it precludes
-the expression of many GADTs commonly used in practice, such as the
-following GADT $\mathtt{Expr}$ of typed expressions:
+The requirement that the types in $\ol{K}$ be constant-free
+polynomials is quite restrictive. In fact, it precludes the expression
+of many GADTs commonly used in practice, such as the following GADT
+$\mathtt{Expr}$ of typed expressions:
 \[
@@
 
-We have constructed a parametric model for a calculus supporting
-primitive nested types, and used its Abstraction Theorem to derive
-free theorems for nested types.  This was not possible
-before~\cite{jp19} because these types were not previously known to
-have well-defined interpretations in locally finitely presentable
-categories (here, $\set$ and $\rel$), and, to our knowledge, no term
-calculus for them existed.  We naturally expect the construction
-elaborated here to guide its generalization to more advanced data
-types.  For example, GADTs can be represented using left Kan
-extensions, and it was shown in~\cite{jp19} that adding a
-$\mathsf{Lan}$ construct to a calculus such as ours preserves the
-$\lambda$-cocontinuity needed for the data types it defines to have
-well-defined interpretations in locally $\lambda$-presentable
-categories. (Interestingly, $\lambda > \aleph_1$ is required to
-interpret even common GADTs.) This suggests carrying out our model
+We have constructed a parametric model for a calculus providing
+primitives for constructing nested types directly as fixpoints, rather
+than representing them via their Church encodings. We have also used
+the Abstraction Theorem for this model to derive free theorems for
+nested types. This was not possible before~\cite{jp19} because such
+types were not previously known to have well-defined interpretations
+in locally finitely presentable categories (here, $\set$ and $\rel$),
+and, to our knowledge, no term calculus for them existed either. The
+key to obtaining our parametric model is the delicate threading of
+functoriality and its accompanying naturality conditions throughout
+the model construction.
+
+We were surprised to find that, although GADTs were shown
+in~\cite{jp19} to have appropriately cocontinuous functorial semantics
+when interpreted using left Kan extensions, our model construction
+does not extend to give a parametric model for them. It may be
+possible to modify the categories of relations and relation
+transformers so that these are always surjective on their domans and
+codomains; this would prohibit situations like that in
+Example~\ref{ex:no-rel-transf}, and might therefore allow us to
+recover our IEL and, ultimately, our Abstraction Theorem for GADTs.
+If this turns out to be possible, then extending this new model
+construction to locally $\lambda$-presentable categories for $\lambda
+> \omega$ would make it possible to handle broader classes of
+GADTs. (Interestingly, $\lambda > \aleph_1$ is required to interpret
+even common GADTs.) We could even attempt to carry out our
 construction in locally $\lambda$-presentable cartesian closed
@@
 obtained by pullback as in~\cite{jac99}, are also lpcccs and are
-appropriately fibred over $\cal C$.  Adding term-level fixpoints
-further requires our semantic categories not just to be locally
-$\lambda$-presentable, but also to support some kind of domain
-structure.
-
-{\color{red} KEEP ANYTHING FROM THIS? ADD GADTS, AT LEAST TO SUMMARY.
-
-We have constructed a parametric model for a calculus
-providing primitives for constructing nested types directly as
-fixpoints.
-%rather than representing them via their Church encodings. We
-We have also used the Abstraction Theorem for this model to derive
-free theorems in the presence of nested types. This was not possible
-before~\cite{jp19} because the nested types in our calculus were not
-previously known to have well-defined interpretations in locally
-finitely presentable categories like $\set$ and $\rel$. No calculus
-for terms of these types existed, either. The key to obtaining our
-parametric model is
-%{\color{blue} adjusting our type syntax to respect the accessibility
-%  cardinal for the semantic category and}
-the delicate threading of functoriality and its accompanying
-naturality conditions throughout the model construction.
-
-Even when nested types have well-defined interpretations in a semantic
-category, the types of standard functions over them may not. This is
-because all recursion in our calculus must be coded using standard
-folds.  Since folds for nested types can only express computations
-that return natural transformations, their expressivity has long been
-a vexing issue, and this is naturally inherited by our calculus.
-%indeed, such folds
-%Our calculus naturally inherits these constraints.
-%Of course our calculus inherits the constraints
-%standard folds for nested types impose, and thus is perhaps not as
-%expressive as we would ideally like.
-In addition, while there are many papers on higher-kinded types, we
-know of none that considers expressibility issues arising from the
-local $\lambda$-presentability cardinal $\lambda$ of the categories in
-which higher-kinded types are interpreted. Because $\lambda = \omega$
-for $\set$ and $\rel$, the expressivity of our
-%$\set$/$\rel$-based parametric model
-calculus is, unfortunately, not easily improved there. Construction of
-a $\set$/$\rel$-based parametric model for our calculus is
-nevertheless worthwhile.  For one thing, the expressivity issue is
-orthogonal to constructing parametric models,
-%in the presence of nested types built from primitives,
-and trying to address both in a single paper would
-detract from the conceptual clarity of our model construction.
-%our current focus on deriving principled parametricity
-%results for practical programming languages, such as Haskell and Agda,
-%that directly construct nested types as fixpoints.
-%
-%conceptual clarity of this paper.
-For another, if the construction in this paper
-%we couldn't
-%construct a $\set$/$\rel$-based parametric model for the calculus
-%presented here,
-weren't possible, then trying to construct a parametric model from
-richer semantic categories for fancier data types and/or recursion
-schemes would be futile. We naturally expect the construction reported
-here to guide its generalization to the locally $\lambda$-presentable
-categories for $\lambda > \omega$ that will be needed to accommodate
-more general data types and recursion schemes for them.
-
-% As is, the results reported here provide a
-%blueprint for deriving principled parametricity results for practical
-%programming languages, such as Haskell and Agda, that directly
-%construct nested types as fixpoints.
-
-%{\color{blue} Perhaps surprisingly, the accessibility cardinal
-%$\omega$ for the semantic categories $\set$ and $\rel$ also impacts
-%the types of functions (and thus the functions themselves), even over
-%expressible data types, that can be expressed. Although there are
-%many papers on higher-kinded types, no other work that we know of
-%even considers expressibility issues resulting from the semantic
-%settings in which they are interpreted.}
-%{\color{blue} In light of the aforementioned
-%  restrictions imposed by the accessibility cardinal of $\set$ and
-%  $\rel$, this paper is perhaps best read as a proof-of-concept for
-%  constructing parametric models in the locally finitely presentable
-%  categories $\set$ and $\rel$ for calculi that construct nested types
-%  and other advanced data types as fixed points (rather than
-%  representing them via their Church encodings) and that support
-%  structured recursion over them. If we couldn't construct such a
-%  model even in this restricted setting, then there'd be no point in
-%  trying to do so for more robust categories, data types, and/or
-%  recursion schemes. Moreover, our constructions in this simplest
-%  possible setting will guide their generalizations to the locally
-%  $\lambda$-presentable categories fopr $\lambda > \omega$ required to
-%  handle more general data types, structured recursive functions over
-%  then, and more general fixpoints.}
-
-%But since this expressivity issue is orthogonal to constructing
-%parametric models in the presence of nested types built from
-%primitives, and since it would detract from the conceptual clarity of
-%our presentation, we have deliberately chosen not to address it in
-%this paper. (However, we discuss some ideas for increasing
-%expressivity in the next paragraph.) Despite the limitations of
-%standard folds, our results provide a blueprint for deriving
-%principled parametricity results for practical programming languages,
-%such as Haskell and Agda, that directly construct nested types as
-%fixpoints.
-
-%All recursion in our calculus must be coded using folds. Expressivity
-%of standard folds for nested types has long been a vexing issue,
-%%indeed, such folds
-%since they can only express computations that return natural
-%transformations; our calculus naturally inherits these constraints.
-%%Of course our calculus inherits the constraints
-%%standard folds for nested types impose, and thus is perhaps not as
-%%expressive as we would ideally like.
-%But since this expressivity issue
-%is orthogonal to constructing parametric models in the presence of
-%nested types built from primitives, and since it would detract from
-%the conceptual clarity of our presentation, we have deliberately
-%chosen not to address it in this paper. (However, we discuss some
-%ideas for increasing expressivity in the next paragraph.) Despite the
-%limitations of standard folds, our results provide a blueprint for
-%deriving principled parametricity results for practical programming
-%languages, such as Haskell and Agda, that directly construct nested
-%types as fixpoints.
-
-%Many extensions of standard folds have been
-%proposed. Perhaps the most significant of these are
-Generalized folds~\cite{bp99} are an extension of standard folds that
-%and Mendler iterators~\cite{amu05}.
-%which
-can express computations whose results are not natural
-transformations. These were shown in~\cite{jg10} to be equivalent to
-standard folds in the presence of object-level {\em right} Kan
-extensions.
-%; interpreting such
-%a construct in $\set$ and $\rel$ should be unproblematic {\color{blue}
-%  since these categories have all(?) limits}.
-It was also shown in~\cite{jg08} that GADTs can be represented using
-object-level {\em left} Kan extensions, and
-%
-%This suggests
-%extending our term calculus with such object-level constructs.
-%
-%{\color{blue} We may need to impose restrictions inspired
-%  by those for $\mu$-types on LKEs.}
-adding a carefully designed left Kan extension construct to our
-calculus was shown in~\cite{jp19} to preserve the cocontinuity needed
-for GADTs to have well-defined interpretations. (Note, however, that
-%not necessarily in
-%However, this will not, in general, be
-%the case in locally finitely presentable categories. Indeed, as shown
-%in~\cite{jp19},
-locally $\lambda$-presentable categories for $\lambda > \aleph_1$ are
-required to interpret even commonly used GADTs.) This suggests
-extending our calculus with Kan extension constructs and
-%augmenting our type formation rules with such a construct.
-%The constructions of this paper should all be possible
-carrying out our model construction in a locally $\lambda$-presentable
-cartesian closed category (lpccc) $\cal C$ whose category of
-(abstract) relations, obtained by pullback as in~\cite{jac99},
-is also a lpccc and appropriately fibred over $\cal C$.  This would
-%extend the results presented here along the lines of~\cite{gjfor15},
-%thus giving
-give a framework for constructing parametric models for calculi
-with nested types built from primitives that is based on locally
-$\lambda$-presentable fibrations, for some appropriate definition
-thereof.  Another option is to add term-level fixpoints
-%, as in~\cite{pit00},
-to the calculus presented here. Of course, this would require the
+appropriately fibred over $\cal C$. This would give a framework for
+constructing parametric models for calculi with primitive GADTs that
+is based on locally $\lambda$-presentable fibrations, for some
+appropriate definition thereof.
+
+The expressivity of folds for nested types has long been a vexing
+issue (see, e.g.,~\cite{bm98}), and this is naturally inherited by the
+calculus presented here. Since it codes all recursion using standard
+folds, and since folds for nested types must return natural
+transformations, many standard functions over nested types cannot be
+represented in our calculus. Another important direction for future
+work would therefore be to improve its expressivity by adding, say,
+generalized folds~\cite{bp99}, Mendler iterators~\cite{amu05}, or
+otherwise extend standard folds to express computations whose results
+are not natural transformations. In particular, we may wish to add
+term-level fixpoints as, e.g., in~\cite{pit00}. This would require the
 categories interpreting types to be not just locally
 $\lambda$-presentable, but to support some kind of domain structure as
-well.
-%Unfortunately, $\omega$-CPO, the most obvious category
-%of domains to replace $\set$, is not locally finitely presentable.
-}
+well. At the moment it seems that such an endeavor will have to wait
+for a generalization of the results presented here to at least locally
+$\omega_1$-presentable categories: $\omega$-CPO, the most natural
+category of domains to replace $\set$, is not locally finitely
+presentable.
 

### 2020-12-30 free-theorems-nesteds/20-mscs.tex
fine-tuning journal draft

@@
 
-\keywords{MANDATORY list of keywords}
+\keywords{Nested types, GADTs, categorical semantics, parametricity}
 
@@
 
-This paper is an extended version of the conference
-paper~\cite{jgj21}. Compared to that paper, this one includes all
-proofs, as well as more examples. The discussion of GADTs in
-Section~\ref{sec:GADTs} is also entirely new.
+This paper extends the conference paper~\cite{jgj21} by including
+proofs of all theorems, as well as more examples. The discussion of
+GADTs in Section~\ref{sec:GADTs} is also entirely new.
 
@@
 formation rule for $\mu$-types, as well as the delicate threading of
-some additional conditions throughout the model construction.  But
+some additional conditions throughout our model construction.  But
 since we can always use an ADT's (semantically equivalent)
@@
 functoriality and its accompanying naturality conditions throughout
-the model construction.
+our model construction.
 
@@
 in~\cite{jp19} to have appropriately cocontinuous functorial semantics
-when interpreted using left Kan extensions, our model construction
-does not extend to give a parametric model for them. It may be
-possible to modify the categories of relations and relation
-transformers so that these are always surjective on their domans and
-codomains; this would prohibit situations like that in
-Example~\ref{ex:no-rel-transf}, and might therefore allow us to
-recover our IEL and, ultimately, our Abstraction Theorem for GADTs.
-If this turns out to be possible, then extending this new model
-construction to locally $\lambda$-presentable categories for $\lambda
-> \omega$ would make it possible to handle broader classes of
-GADTs. (Interestingly, $\lambda > \aleph_1$ is required to interpret
-even common GADTs.) We could even attempt to carry out our
-construction in locally $\lambda$-presentable cartesian closed
-categories (lpcccs) $\cal C$ whose categories of (abstract) relations,
-obtained by pullback as in~\cite{jac99}, are also lpcccs and are
-appropriately fibred over $\cal C$. This would give a framework for
-constructing parametric models for calculi with primitive GADTs that
-is based on locally $\lambda$-presentable fibrations, for some
-appropriate definition thereof.
+in terms of left Kan extensions, our model construction does not
+extend to give a parametric model for them. It may be possible to
+modify the categories of relations and relation transformers so that
+these are always surjective on their domans and codomains; this would
+prohibit situations like that in Example~\ref{ex:no-rel-transf}, and
+might therefore allow us to recover our IEL and, ultimately, an
+Abstraction Theorem appropriate to GADTs.  If this turns out to be
+possible, then generalizing the resulting model construction to
+locally $\lambda$-presentable categories for $\lambda > \omega$ would
+make it possible to handle broader classes of GADTs. (As shown
+in~\cite{jp19}, $\lambda > \omega_1$ is required to interpret even
+common GADTs.) We could even attempt to carry out our construction in
+locally $\lambda$-presentable cartesian closed categories (lpcccs)
+$\cal C$ whose categories of (abstract) relations, obtained by
+pullback as in~\cite{jac99}, are also lpcccs and are appropriately
+fibred over $\cal C$. This would give a framework for constructing
+parametric models for calculi with primitive GADTs that is based on
+locally $\lambda$-presentable fibrations, for some appropriate
+definition thereof.
 
@@
 transformations, many standard functions over nested types cannot be
-represented in our calculus. Another important direction for future
-work would therefore be to improve its expressivity by adding, say,
-generalized folds~\cite{bp99}, Mendler iterators~\cite{amu05}, or
-otherwise extend standard folds to express computations whose results
-are not natural transformations. In particular, we may wish to add
-term-level fixpoints as, e.g., in~\cite{pit00}. This would require the
-categories interpreting types to be not just locally
-$\lambda$-presentable, but to support some kind of domain structure as
-well. At the moment it seems that such an endeavor will have to wait
-for a generalization of the results presented here to at least locally
-$\omega_1$-presentable categories: $\omega$-CPO, the most natural
-category of domains to replace $\set$, is not locally finitely
-presentable.
+represented in this calculus. Another important direction for future
+work is therefore to improve the expressivity of our calculus by
+adding, say, generalized folds~\cite{bp99}, or Mendler
+iterators~\cite{amu05}, or otherwise extending standard folds to
+express computations whose results are not natural transformations. In
+particular, we may wish to add term-level fixpoints as, e.g.,
+in~\cite{pit00}. This would require the categories interpreting types
+to be not just locally $\lambda$-presentable, but also to support some
+kind of domain structure. At the moment it seems that such an endeavor
+will have to wait for a generalization of the results presented here
+to at least locally $\omega_1$-presentable categories, however:
+$\omega$-CPO, the most natural category of domains to replace $\set$,
+is not locally finitely presentable.
 

### 2021-01-04 free-theorems-nesteds/20-mscs.tex
journal version ready for proofreading

@@
 \title[Parametricity for Primitive Nested Types and
-  GADTs]{Parametricity for Primitive Nested Types and GADTs}
+  GADTs]{Parametricity for Primitive Nested Types}
 
@@
 
-This paper extends the conference paper~\cite{jgj21} by including
-proofs of all theorems, as well as more examples. The discussion of
-GADTs in Section~\ref{sec:GADTs} is also entirely new.
+This paper extends the conference paper~\cite{jgj21} of the same name
+by including more exposition, proofs of all theorems, and more
+examples. In addition, the discussion of parametricity for GADTs in
+Section~\ref{sec:GADTs} is entirely new.
 
@@
 ADTs and nested types, but his focus is on expressivity of generalized
-Mendler iteration for them. He gives no semantics.
+Mendler iteration for them. He gives no semantics whatsoever.
 
@@
 $\lambda$ operator binds all occurrences of the variables in
-$\ol{\alpha}$, in the body $F$. {\color{red} Can bind variables of
-  arity greater than 0 in Nat-types even just for nesteds? Improves
-  expressibility? Worth it? Worth mentioning?}
+$\ol{\alpha}$, in the body $F$.
 
@@
   \relsem{\Gamma;\Phi \vdash \onet}\rho &= 1\\
-  \end{align*}
-
-\vspace*{-0.5in}
-
-\begin{align*}
   \relsem{\Gamma; \emptyset \vdash \Nat^{\ol{\alpha}} \,F\,G}\rho &= \{\eta
@@
   \relsem{\Gamma;\Phi \vdash G}\rho\\  
+\end{align*}
+
+\vspace*{-0.5in}
+
+\begin{align*}
    \relsem{\Gamma;\Phi \vdash (\mu \phi.\lambda
@@
 \subsection{Standard Free Theorems for ADTs and Nested
-  types}\label{sec:ft-adt}
+  Types}\label{sec:ft-adt}
 
@@
 $\ol\alpha$ denoted by $K$. Using $\Lan$-types, we can therefore
-represent the GADT $\mathtt{Seq}$ in our calculus as {\color{red}
-  CHECK!!}
+represent the GADT $\mathtt{Seq}$ in our calculus as 
  \begin{align*}
@@
 interpreted in $\set$ as functors and in $\rel$ as relation
-transformers, we extend the set and relational interpretations for
+transformers, we let $k_i$ be the arity of the $i^{th}$ element of the
+sequence $\Phi$ and extend the set and relational interpretations for
 $\Nat$ types from Section~\ref{sec:type-interp} as follows:
-{\color{red} Relate $k$ to arities in $\Phi$.}
 \begin{align*}
 \setsem{\Gamma; \emptyset \vdash \Nat^\Phi\,F\,G}\rho
-&= \{\eta : \lambda \ol{K}. \,\setsem{\Gamma; \Phi \vdash F}\rho[\Phi := \ol K] 
-  \Rightarrow \lambda \ol{K}.\,\setsem{\Gamma; \Phi \vdash G}\rho[\Phi := \ol K]~|~\\ 
+&= \{\eta : \lambda \ol{K}. \,\setsem{\Gamma; \Phi \vdash F}\rho[\Phi := \ol{K}] 
+  \Rightarrow \lambda \ol{K}.\,\setsem{\Gamma; \Phi \vdash G}\rho[\Phi := \ol{K}]~|~\\ 
 &\hspace{0.3in} \forall \ol{K = (K^1, K^2, K^*) : RT_k}.\\
 &\hspace{0.4in} (\eta_{\overline{K^1}}, \eta_{\overline{K^2}})
-: \setsem{\Gamma; \Phi \vdash F}\Eq_{\rho}[\Phi := \ol K]
-\rightarrow \setsem{\Gamma; \Phi \vdash G}\Eq_{\rho}[\Phi := \ol K] \} \\
+: \setsem{\Gamma; \Phi \vdash F}\Eq_{\rho}[\Phi := \ol{K}]
+\rightarrow \setsem{\Gamma; \Phi \vdash G}\Eq_{\rho}[\Phi := \ol{K}] \} \\
 %%%
 \relsem{\Gamma; \emptyset \vdash \Nat^\Phi \,F\,G}\rho
-&= \{\eta : \lambda \ol{K}.\,\relsem{\Gamma; \Phi \vdash F}\rho[\ol{\Phi := K}]
-\Rightarrow \lambda \ol{K}. \,\relsem{\Gamma; \Phi \vdash G}\rho[\ol{\phi := K}]\}\\
+&= \{\eta : \lambda \ol{K}.\,\relsem{\Gamma; \Phi \vdash F}\rho[\Phi := \ol{K}]
+\Rightarrow \lambda \ol{K}. \,\relsem{\Gamma; \Phi \vdash G}\rho[\Phi := \ol{K}]\}\\
 &=
@@
   \rho)~|~\\
-& \hspace{0.3in} \forall \ol{K = (K^1, K^2, K^*) : RT_{k}}.\\
+& \hspace{0.3in} \forall \ol{K = (K^1, K^2, K^*) : RT_k}.\\
 & \hspace{0.4in} ((\eta_1)_{\ol{K^1}},(\eta_2)_{\ol{K^2}}) \in
-  (\relsem{\Gamma; \Phi \vdash G}\rho[\ol{\Phi := K}])^{\relsem{\Gamma;\Phi\vdash F}\rho[\ol{\Phi := K}]} \}\\  
+  (\relsem{\Gamma; \Phi \vdash G}\rho[\Phi := \ol{K}])^{\relsem{\Gamma;\Phi\vdash F}\rho[\Phi := \ol{K}]} \}\\  
 \end{align*}
@@
 \]
-respectively, where, for each $\ol{N : RT_k}$ {\color{red} Relate
-  $k$ to arities in $\Phi$.}, $\eta_{\ol{N}}$ is the natural
-transformation associated with the left Kan extension
+respectively, where, for each $\ol{N : RT_k}$, $\eta_{\ol{N}}$
+is the natural transformation associated with the left Kan extension
 
@@
 \end{tikzcd}\]
-and, for all $\ol{N : RT_k}$ {\color{red} Relate $k$ to arities in
-  $\Phi$.} and $d : \setsem{\Gamma;\emptyset \vdash \Delta}$, 
+and, for all $\ol{N : RT_k}$ and $d : \setsem{\Gamma;\emptyset \vdash
+  \Delta}$,
 \[\begin{array}{l}
@@
 \end{proof}
-Thus, {\color{red} CHECK!} if $L$ were a restriction of
-$\setsem{\Gamma; \Phi,\ol\alpha \vdash (\Lan^{\ol\alpha}_{\ol K}
-  F){\ol K}}\rho$, if the functorial action of $L$ were a restriction
-of that of $\setsem{\Gamma; \Phi,\ol\alpha \vdash
-  (\Lan^{\ol\alpha}_{\ol K} F){\ol K}}\rho$, and if
-$L(\ol{\setsem{\Gamma;\ol\alpha \vdash K}\rho \ol N \, \ol A})$ were
-to contain $\setsem{\Gamma;\emptyset~|~\emptyset \vdash \int_{\ol K,F}
-  : \Nat^{\Phi, \ol\alpha}\, F\, (\Lan^{\ol\alpha}_{\ol{K}}\, F)
-  \ol{K}}\rho\,d\,\ol N\,\ol A\,y$ for all $\ol N$, $\ol A$, and $y$,
-then $L$ would have to be the entirety of $\mathit{Lan}_{\ol{\lambda
-    \ol A.\, \setsem{\Gamma; \ol\alpha \vdash K}\rho {[\ol{\alpha :=
-          A} ]}}}\, \lambda \ol A.\, \setsem{\Gamma; \Phi, \ol\alpha
-  \vdash F}\rho {[\Phi, \ol \alpha := \ol N, \ol A ]}$. An analogous
-result holds for relational interpretations of $\Lan$-types.
-
-\subsection{Parametricity Results for GADTs}
+Thus, if $L$ were a restriction of $\setsem{\Gamma; \Phi,\ol\alpha
+  \vdash (\Lan^{\ol\alpha}_{\ol K} F){\ol K}}\rho$, if the functorial
+action of $L$ were a restriction of that of $\setsem{\Gamma;
+  \Phi,\ol\alpha \vdash (\Lan^{\ol\alpha}_{\ol K} F){\ol K}}\rho$, and
+if $L(\ol{\setsem{\Gamma;\ol\alpha \vdash K}\rho \ol N \, \ol A})$
+were to contain $\setsem{\Gamma;\emptyset~|~\emptyset \vdash \int_{\ol
+    K,F} : \Nat^{\Phi, \ol\alpha}\, F\, (\Lan^{\ol\alpha}_{\ol{K}}\,
+  F) \ol{K}}\rho\,d\,\ol N\,\ol A\,y$ for all $\ol N$, $\ol A$, and
+$y$, then $L$ would have to be the entirety of
+$\mathit{Lan}_{\ol{\lambda \ol A.\, \setsem{\Gamma; \ol\alpha \vdash
+      K}\rho {[\ol{\alpha := A} ]}}}\, \lambda \ol A.\,
+\setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho {[\Phi, \ol \alpha :=
+    \ol N, \ol A ]}$. An analogous result holds for relational
+interpretations of $\Lan$-types.
+
+\subsection{Parametricity and GADTs}
 
@@
 
-{\color{red} CHECK!} Even when extending along constant-free
-polynomials, it is unclear how to prove the Identity Extension Lemma
-for $\Lan$-types or, indeed, whether it holds at all in their
-presence. To prove the IEL for $\Nat$-types, for example, it is
-necessary to cut down the semantic interpretation by requiring that
-the natural transformations in the set interpretation preserve
-equalities. The impossibility of restricting the semantic
-interpretations of $\Lan$-types, as shown in
+Even when extending along constant-free polynomials, it is unclear how
+to prove the Identity Extension Lemma for $\Lan$-types or, indeed,
+whether it holds at all in their presence. To prove the IEL for
+$\Nat$-types, for example, it is necessary to cut down the semantic
+interpretation by requiring that the natural transformations in the
+set interpretation preserve equalities. The impossibility of
+restricting the semantic interpretations of $\Lan$-types, as shown in
 Proposition~\ref{prop:full-lan}, suggests that the IEL may fail for

### 2021-01-11 free-theorems-nesteds/20-mscs.tex
updated LMCS paper

@@
 [\phi := \lambda \ol{A}.\,\setsem{\Gamma;\Phi,\overline{\alpha}\vdash
-    H}f[\overline{\alpha := \id_{\ol{A}}}]] 
+    H}f[\overline{\alpha := \id_A}]] 
 \end{gather}
@@
     :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F) & = &  
-\lambda d\,\eta. \,\mathit{fold}_{T^X_{H,\rho}} \, \eta\\ 
+\lambda d. \,\mathit{fold}_{T^X_{H,\rho}} \\ 
 \hspace*{0.79in}(\Nat^{{\ol{\beta}} }\,(\mu
@@
 \ar[Rightarrow, bend right = 10, from=F, "{\eta}" description]
-\ar[ur, bend left, "{\mathit{Lan}_K\,F}"{name=Lan, description}]
+\ar[ur, bend left, "{\mathit{Lan}_{\ol K}\,F}"{name=Lan, description}]
 \ar[ur, bend right, "{G}"'{name=L, right}, ""'{name=Lr, right}]
@@
 \begin{equation}\label{eq:colim-form}
-(\mathit{Lan}_K F) \ol{A} = \colim{\ol{S} : \set_0^k, \,\ol{f :
+(\mathit{Lan}_{\ol K} F) \ol{A} = \colim{\ol{S} : \set_0^k, \,\ol{f :
       K\ol{S} \to  A}}{F \ol{S}}
@@
 If $\ol{S} : \set_0^k$ and $\ol{f : K\ol{S} \to A}$, let $j_{\ol{S},
-  \ol{f}} : F \ol{S} \to (\mathit{Lan}_KF) \ol{A}$ be the morphism
+  \ol{f}} : F \ol{S} \to (\mathit{Lan}_{\ol K} F) \ol{A}$ be the morphism
 indexed by $\ol{S}$ and $\ol{f}$ mapping the cocone into the colimit
@@
 For any $h$-tuple of functions
-$\ol{g : A \to B}$, the functorial action $(\mathit{Lan}_K F) \ol{g}$
-is the unique function from $(\mathit{Lan}_K F) \ol{A}$ to
-$(\mathit{Lan}_K F) \ol{B}$ such that, for all $\ol{S} : \set^k_0$ and
+$\ol{g : A \to B}$, the functorial action $(\mathit{Lan}_{\ol K
+  }F) \ol{g}$
+is the unique function from $(\mathit{Lan}_{\ol K} F) \ol{A}$ to
+$(\mathit{Lan}_{\ol K} F) \ol{B}$ such that, for all $\ol{S} : \set^k_0$ and
 $\ol{f : K\ol{S} \to A}$,\,
 \begin{equation}\label{eq:cocone-def}
-(\mathit{Lan}_KF) \ol{g} \circ j_{\ol{S},\, \ol{f}}
+(\mathit{Lan}_{\ol K}F) \ol{g} \circ j_{\ol{S},\, \ol{f}}
 = j'_{\ol{S},\, \ol{g\, \circ f}}
-: F \ol{S} \to (\mathit{Lan}_KF) \ol{B}
+: F \ol{S} \to (\mathit{Lan}_{\ol K}F) \ol{B}
 \end{equation}
 holds, where $j$ is the morphism mapping the cocone into
-$(\mathit{Lan}_K F) \ol{A} = \colim{\ol{S} : \set_0^k, \,\ol{f :
+$(\mathit{Lan}_{\ol K} F) \ol{A} = \colim{\ol{S} : \set_0^k, \,\ol{f :
     K\ol{S} \to A}}{F \ol{S}}$ and $j'$ is the morphism mapping the
-cocone into $(\mathit{Lan}_K F) \ol{B} = \colim{\ol{S} : \set_0^k,
+cocone into $(\mathit{Lan}_{\ol K} F) \ol{B} = \colim{\ol{S} : \set_0^k,
   \,\ol{f : K\ol{S} \to B}}{F \ol{S}}$.  Moreover, if $\alpha : F \to
-F'$ is a natural transformation, then $\mathit{Lan}_K \alpha :
-(\mathit{Lan}_K F) \to (\mathit{Lan}_K F')$ is defined to be the
-induced natural transformation whose component $(\mathit{Lan}_K
-\alpha) \ol{A} : (\mathit{Lan}_K F) \ol{A} \to (\mathit{Lan}_K F')
-\ol{A}$ is the unique function such that $(\mathit{Lan}_K \alpha)
+F'$ is a natural transformation, then $\mathit{Lan}_{\ol K} \alpha :
+(\mathit{Lan}_{\ol K} F) \to (\mathit{Lan}_{\ol K} F')$ is defined to be the
+induced natural transformation whose component $(\mathit{Lan}_{\ol K}
+\alpha) \ol{A} : (\mathit{Lan}_{\ol K} F) \ol{A} \to (\mathit{Lan}_{\ol K} F')
+\ol{A}$ is the unique function such that $(\mathit{Lan}_{\ol K} \alpha)
 \ol{A} \,\circ\, j_{\ol{S}, \ol{f}} = j'_{\ol{S}, \ol{f}} \,\circ\,
-\alpha_{\ol{S}}$, where $j$ is the cocone into $(\mathit{Lan}_K F)
-\ol{A}$ and $j'$ is the cocone into $(\mathit{Lan}_K F') \ol{A}$. We
+\alpha_{\ol{S}}$, where $j$ is the cocone into $(\mathit{Lan}_{\ol K} F)
+\ol{A}$ and $j'$ is the cocone into $(\mathit{Lan}_{\ol K} F') \ol{A}$. We
 can similarly represent left Kan extensions of functors over $\rel$ in
@@
 
-Intuitively, $\Lan^{\ol{\alpha^0}}_K F$ is a syntactic representation
+Intuitively, $\Lan^{\ol{\alpha^0}}_{\ol K} F$ is a syntactic representation
 of the left Kan extension of the functor in the variables in
 $\ol\alpha$ denoted by $F$ along the functor in the variables in
-$\ol\alpha$ denoted by $K$. Using $\Lan$-types, we can therefore
+$\ol\alpha$ denoted by $\ol K$. Using $\Lan$-types, we can therefore
 represent the GADT $\mathtt{Seq}$ in our calculus as 
@@
 using just unary $\Lan$-types, i.e, $\Lan$-types of the form
-$(\Lan^{\ol{\alpha^0}}_K F)\ol A$.
+$(\Lan^{\ol{\alpha^0}}_{\ol K} F)\ol A$.
 
@@
 &\hspace{0.4in} (\eta_{\overline{K^1}}, \eta_{\overline{K^2}})
-: \setsem{\Gamma; \Phi \vdash F}\Eq_{\rho}[\Phi := \ol{K}]
-\rightarrow \setsem{\Gamma; \Phi \vdash G}\Eq_{\rho}[\Phi := \ol{K}] \} \\
+: \relsem{\Gamma; \Phi \vdash F}\Eq_{\rho}[\Phi := \ol{K}]
+\rightarrow \relsem{\Gamma; \Phi \vdash G}\Eq_{\rho}[\Phi := \ol{K}] \} \\
 %%%
@@
 \begin{prop}\label{prop:full-lan}
-Let $F : \set^h \to \set$ and $\ol{K : \set^h \to \set^k}$ be functors.
-If $L : \set^k \to \set$ is a functor such that 
+Let $F : \set^k \to \set$ and $\ol K : \set^k \to \set^h$ be functors.
+If $L : \set^h \to \set$ is a functor such that 
 \begin{itemize}

### 2021-01-12 free-theorems-nesteds/20-mscs.tex
final LMCS submission

@@
 \title[Parametricity for Primitive Nested Types and
-  GADTs]{Parametricity for Primitive Nested Types}
+  GADTs]{Parametricity for Primitive Nested Types and GADTs}
 
@@
 
-This paper extends the conference paper~\cite{jgj21} of the same name
-by including more exposition, proofs of all theorems, and more
-examples. In addition, the discussion of parametricity for GADTs in
-Section~\ref{sec:GADTs} is entirely new.
+This paper extends the conference paper~\cite{jgj21} by including more
+exposition, proofs of all theorems, and more examples. In addition,
+the discussion of parametricity for GADTs in Section~\ref{sec:GADTs}
+is entirely new.
 
@@
   H^2F^2 \ol{\beta})$ whenever $(\alpha_1, \beta_1) \in
-  \Homrel(R_1,S_1),..., (\alpha_k, \beta_k) \in \Homrel(R_k,S_k)$
+  \Homrel(R_1,S_1),..., (\alpha_k, \beta_k)$ $\in \Homrel(R_k,S_k)$
 \item If $(\delta^1,\delta^2) : (F^1,F^2,F^*)\to (G^1,G^2,G^*)$ and
@@
 Clearly, $\graph{\alpha}^*$ is $\omega$-cocontinuous, so
-$\graph{\alpha}^* : [\rel^k,\rel]$. Now, let $\overline{R :
+$\graph{\alpha}^* : [\rel^k,\rel]$. Let $\overline{R :
   \rel(A, B)}$, $\overline{S : \rel(C, D)}$, and $\overline{(\beta,
@@
 
-\vspace*{-0.15in}
+\vspace*{-0.05in}
 
@@
 \set^k
-\ar[rr, "{F}"{name=Fa}, ""{name=F, below}]
+\ar[rr, "{F}"{name=Fa, above}, ""{name=F, below}]
 \ar[rd, "{\ol K}"']
@@
 & \set^h
-\ar[Rightarrow, bend right = 10, from=F, "{\eta}" description]
+\ar[Rightarrow, bend right = 25, from=F, "{\eta}"']
 \ar[ur, bend left, "{\mathit{Lan}_{\ol K}\,F}"{name=Lan, description}]
 \ar[ur, bend right, "{G}"'{name=L, right}, ""'{name=Lr, right}]
-\ar[Rightarrow, bend left = 90, from=Fa, to=Lr, "{\gamma}"]
-\ar[Rightarrow, dashed, from=Lan, to=L, "{\mu}"]
+\ar[Rightarrow, bend left = 45, from=Fa, to=Lr, "{\gamma}" near start]
+\ar[Rightarrow, dashed, from=Lan, to=L, "{\mu}"']
 \end{tikzcd}
@@
 
-An alternative presentation of characterizes the left Kan extension
+An alternative presentation characterizes the left Kan extension
 $(\mathit{Lan}_{\ol K} F, \eta)$ in terms of the adjunction between
@@
   subcategory of finitely presentable objects in $\set$, --- i.e., is
-  the category finite sets --- we  have that
+  the category of finite sets --- we  have that
 \begin{equation}\label{eq:colim-form}
@@
 using just unary $\Lan$-types, i.e, $\Lan$-types of the form
-$(\Lan^{\ol{\alpha^0}}_{\ol K} F)\ol A$.
+$(\Lan^{\ol{\alpha^0}}_K F)\ol A$.
 
@@
 
+\vspace*{-0.1in}
+
 \[
@@
 \ar[dr, "\ol{\lambda \ol{A}.\,\setsem{\Gamma; \ol\alpha \vdash
-      K}\rho {[\ol{\alpha := A} ]}}"{left}] && \set \\  
-&\set^{|\ol K|} \ar[ur, "\mathit{Lan}_{\ol{\lambda \ol A.\,
+      K}\rho {[\ol{\alpha := A} ]}}\;\;\;\;"{left}] && \set \\  
+&\set^{|\ol K|} \ar[ur, "\;\;\mathit{Lan}_{\ol{\lambda \ol A.\,
       \setsem{\Gamma; \ol\alpha \vdash K}\rho {[\ol{\alpha := A}
@@
 \end{proof}
-Thus, if $L$ were a restriction of $\setsem{\Gamma; \Phi,\ol\alpha
-  \vdash (\Lan^{\ol\alpha}_{\ol K} F){\ol K}}\rho$, if the functorial
-action of $L$ were a restriction of that of $\setsem{\Gamma;
-  \Phi,\ol\alpha \vdash (\Lan^{\ol\alpha}_{\ol K} F){\ol K}}\rho$, and
-if $L(\ol{\setsem{\Gamma;\ol\alpha \vdash K}\rho \ol N \, \ol A})$
-were to contain $\setsem{\Gamma;\emptyset~|~\emptyset \vdash \int_{\ol
+Thus, if $L$ were a restriction of $\lambda \ol B. \setsem{\Gamma; \Phi,\ol\beta
+  \vdash (\Lan^{\ol\alpha}_{\ol K} F){\ol \beta}}\rho[\Phi := \ol
+  N][\ol{\beta := B}]$,
+if the functorial 
+action of $L$ were a restriction of that of $\lambda \ol B. \setsem{\Gamma;
+  \Phi,\ol\beta \vdash (\Lan^{\ol\alpha}_{\ol K} F){\ol
+    \beta}}\rho[\Phi := \ol N][\ol{\beta := B}]$, and if
+$L(\ol{\setsem{\Gamma;\ol\alpha \vdash K}\rho[\ol{\alpha := A}]})$ 
+contained $\setsem{\Gamma;\emptyset~|~\emptyset \vdash \int_{\ol
     K,F} : \Nat^{\Phi, \ol\alpha}\, F\, (\Lan^{\ol\alpha}_{\ol{K}}\,
@@
 $y$, then $L$ would have to be the entirety of
-$\mathit{Lan}_{\ol{\lambda \ol A.\, \setsem{\Gamma; \ol\alpha \vdash
-      K}\rho {[\ol{\alpha := A} ]}}}\, \lambda \ol A.\,
-\setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho {[\Phi, \ol \alpha :=
-    \ol N, \ol A ]}$. An analogous result holds for relational
+ $\lambda \ol B. \setsem{\Gamma; \Phi,\ol\beta
+  \vdash (\Lan^{\ol\alpha}_{\ol K} F){\ol \beta}}\rho[\Phi := \ol
+  N][\ol{\beta := B}]$.
+%$\mathit{Lan}_{\ol{\lambda \ol A.\, \setsem{\Gamma; \ol\alpha \vdash
+%      K}\rho {[\ol{\alpha := A} ]}}}\, \lambda \ol A.\,
+%\setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho {[\Phi, \ol \alpha :=
+%    \ol N, \ol A ]}$.
+An analogous result holds for relational
 interpretations of $\Lan$-types.

### 2021-01-21 20-gadts/GADTs-negative-result.tex
added GADTs-negative-result.tex

\documentclass{lmcs}

\keywords{Nested types, GADTs, categorical semantics, parametricity}

\usepackage[utf8]{inputenc}
%\usepackage{ccicons}
\usepackage{verbatim}
\usepackage{adjustbox}
\usepackage{marvosym}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amscd}
%\usepackage{MnSymbol}
\usepackage{xcolor}

\usepackage{bbold}
\usepackage{url}
\usepackage{upgreek}
\usepackage{stmaryrd}

\usepackage{lipsum}
\usepackage{tikz-cd}
\usetikzlibrary{cd}
\usetikzlibrary{calc}
\usetikzlibrary{arrows}

\usepackage{bussproofs}
\EnableBpAbbreviations

\usepackage{hyperref}

\input{macros}
%\input{mytheorem}

\theoremstyle{plain}\newtheorem{satz}[thm]{Satz}

\def\eg{{\em e.g.}}
\def\cf{{\em cf.}}

%\theoremstyle{definition}

\newcommand{\inl}{\mathsf{inl}}
\newcommand{\inr}{\mathsf{inr}}
\newcommand{\fold}{\mathsf{fold}}
\newcommand{\ininv}[2]{(\tin^{-1}_{\onet +
  \beta \times \phi (\phi\beta)})_{#1}\, #2}
\newcommand{\emptyfun}{{[]}}
\newcommand{\cal}{\mathcal}
%\newcommand{\fold}{\mathit{fold}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\B}{\mathcal{B}}
\renewcommand{\P}{\mathcal{A}}
\newcommand{\pred}{\mathsf{Fam}}
\newcommand{\env}{\mathsf{Env}}
\newcommand{\set}{\mathsf{Set}}
\renewcommand{\S}{\mathcal S}
\newcommand{\C}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\A}{\mathcal{A}}
\renewcommand{\id}{\mathit{id}}
\newcommand{\map}{\mathsf{map}}
\newcommand{\pid}{\underline{\mathit{id}}}
\newcommand{\pcirc}{\,\underline{\circ}\,}
\newcommand{\pzero}{\underline{0}}
\newcommand{\pone}{\underline{1}}
\newcommand{\psum}{\,\underline{+}\,}
\newcommand{\pinl}{\underline{\mathit{inL}}\,}
\newcommand{\pinr}{\underline{\mathit{inR}}\,}
\newcommand{\ptimes}{\,\underline{\times}\,}
\newcommand{\ppi}{\underline{\pi_1}}
\newcommand{\pppi}{\underline{\pi_2}}
\newcommand{\pmu}{\underline{\mu}}
\newcommand{\semmap}{\mathit{map}}
\newcommand{\subst}{\mathit{subst}}

\newcommand{\tb}[1]{~~ \mbox{#1} ~~}
\newcommand{\listt}[1]{(\mu \phi. \lambda \beta . \onet + \beta \times
  \phi \beta) #1} 
\newcommand{\filtype}{\Nat^\emptyset 
 (\Nat^\emptyset \, \alpha \, \mathit{Bool})\, (\Nat^\emptyset 
  (List \, \alpha) \, (List \, \alpha))} 
\newcommand{\filtypeGRose}{\Nat^\emptyset 
 (\Nat^\emptyset \, \alpha \, \mathit{Bool})\, (\Nat^\emptyset 
  (\mathit{GRose}\,\psi \, \alpha) \, (\mathit{GRose}\,\psi \, (\alpha
  + \onet)))} 
\newcommand{\maplist}{\mathit{map}_{\lambda A. \setsem{\emptyset; \alpha
      \vdash \mathit{List} \, \alpha} \rho[\alpha := A]}} 
\newcommand{\PLeaves}{\mathsf{PLeaves}}
\newcommand{\swap}{\mathsf{swap}}
\newcommand{\reverse}{\mathsf{reverse}}
\newcommand{\bcons}{\mathit{bcons}}
\newcommand{\bnil}{\mathit{bnil}}

\DeclareMathOperator{\SumIn}{in}

\begin{document}

\title[Negative result on parametricity for GADTs]{Negative result on parametricity for GADTs}

\author[P.~Johann]{Patricia Johann}	
\address{Appalachian State University}	
\email{johannp@appstate.edu}

\author[E.~Ghiorzi]{Enrico Ghiorzi}	
\address{Appalachian State University}	
\email{ghiorzie@appstate.edu}

\author[D.~Jeffries]{Daniel Jeffries}	
\address{Appalachian State University}	
\email{jeffriesd@appstate.edu}

\maketitle

We augmented our calculus with $\Lan$ types to allow for primitive GADTs.
We then found out that $\Lan$ types do not support relational semantics for the usual category of relations, and thus the calculus does not have a parametric model.
Given that a counterexample was provided by relations of the form $(S, \emptyset, \emptyset)$, we consider restricting the category of relations to those relations that are onto their domain and codomain, i.e., those relations $R: \rel(A, B)$ such that for all $a \in A$ there exists $b \in B$ such that $(a, b) \in R$, and for all $b \in B$ there exists $a \in A$ such that $(a, b) \in R$.
Each onto relation $R: \rel(A, B)$ also comes equipped with two choice functions: $c_R : A \to B$ and $d_R : B \to A$, such that for every $a : A$ we have $(a, c_R a) \in R$, and for every $b : B$ we have $(d_R b, b) \in R$.
By abuse of notation, we still call $\rel$ this subcategory.

In the following, we will discuss how restricting $\rel$ to onto relations is still insufficient to provide a relational semantics for $\Lan$ types.

\section{Fibered interpretation of $\Lan$}

For every $\Gamma; \Phi, \ol\alpha \vdash F$, $\ol{\emptyset; \alpha \vdash K}$, $\ol{\Gamma; \Phi \vdash B}$ and relation environment $\rho$, we define the set interpretation of $\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}}\,F)\ol{B}$ as
\[
  \setsem{\Gamma;\Phi \vdash
    (\Lan^{\ol{\alpha}}_{\ol{K}}\,F)\ol{B}}\rho = (
  \textit{Lan}_{\lambda \ol{S}.\,\ol{\setsem{\Gamma; \ol\alpha \vdash K}
      \rho[\ol{\alpha := S}]}}\, \lambda \ol{S}.\,\setsem{\Gamma; \Phi, \ol\alpha
    \vdash F} \rho[\ol{\alpha := S}])\, \ol{\setsem{\Gamma; \Phi
      \vdash B}\rho}
\]
and its relational interpretation as
\[
  \relsem{\Gamma;\Phi \vdash
    (\Lan^{\ol{\alpha}}_{\ol{K}}\,F)\ol{B}}\rho = (
  \textit{Lan}_{\lambda \ol R.\,\ol{\relsem{\Gamma; \ol\alpha \vdash
        K}\rho[\ol{\alpha := R}]}}\, \lambda \ol R.\,\relsem{\Gamma;
    \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := R}])\,
  \ol{\relsem{\Gamma; \Phi \vdash B}\rho}
\]

We want to show that the relational interpretation for $\Lan$ types is fibered over the set interpretation, i.e., that there is an isomorphism between
\begin{equation}\label{eq:LHS}
\colim{\ol{S : \set_0}, \ol{g : \setsem{\emptyset; \ol{\alpha} \vdash K}[\ol{\alpha := S}] \to \setsem{\Gamma; \Phi \vdash B}(\pi_1 \rho)}}
{\setsem{\Gamma; \Phi, \ol{\alpha} \vdash F}(\pi_1 \rho)[\ol{\alpha := S}]}
\end{equation}
where $\iota_{\ol{S}, \ol{g}} : \setsem{\Gamma; \Phi, \ol{\alpha} \vdash F}(\pi_1 \rho)[\ol{\alpha := S}] \to \eqref{eq:LHS}$ is the cocone into the colimit~\eqref{eq:LHS}, and
\begin{equation}\label{eq:RHS}
\colim{\ol{R : \rel_0}, \ol{f : \relsem{\emptyset; \ol{\alpha} \vdash K}[\ol{\alpha := R}] \to \relsem{\Gamma; \Phi \vdash B}\rho}}
{\setsem{\Gamma; \Phi, \ol{\alpha} \vdash F}(\pi_1 \rho)[\ol{\alpha := \pi_1 R}]}
\end{equation}
where $j_{\ol{R}, \ol{f}} : \setsem{\Gamma; \Phi, \ol{\alpha} \vdash F}(\pi_1 \rho)[\ol{\alpha := \pi_1 R}] \to \eqref{eq:RHS}$ is the cocone into the colimit~\eqref{eq:RHS}.

We define $k : \eqref{eq:LHS} \to \eqref{eq:RHS}$ as the unique function such that $k \circ \iota_{\ol{S}, \ol{g}} = j_{\ol{\Eq_S}, \ol{(\id, \setsem{\Gamma; \Phi \vdash B} c_\rho) \circ \Eq_g}}$ for every $\ol{S : \set_0}$ and $\ol{g : \setsem{\emptyset; \ol{\alpha} \vdash K}[\ol{\alpha := S}] \to \setsem{\Gamma; \Phi \vdash B}(\pi_1 \rho)}$.

We define $h : \eqref{eq:RHS} \to \eqref{eq:LHS}$ as the unique function such that $h \circ j_{\ol{R}, \ol{f}} = \iota_{\ol{\pi_1 R}, \ol{\pi_1 f}}$ for every $\ol{R : \rel_0}$ and $\ol{f : \relsem{\emptyset; \ol{\alpha} \vdash K}[\ol{\alpha := R}] \to \relsem{\Gamma; \Phi \vdash B}\rho}$.

We want to show that $h$ and $k$ are inverses.

Consider the composition $h \circ k$.
We have that
\[
h \circ k \circ \iota_{\ol{S}, \ol{g}}
= h \circ j_{\ol{\Eq_S}, \ol{(\id, \setsem{\Gamma; \Phi \vdash B} c_\rho) \circ \Eq_g}}
= \iota_{\ol{\pi_1 \Eq_S}, \ol{\pi_1 ((\id, \setsem{\Gamma; \Phi \vdash B} c_\rho) \circ \Eq_g)}}
= \iota_{\ol{S}, \ol{g}}
\]
and so, $h \circ k = \id_{\eqref{eq:LHS}}$.

Consider the composition $k \circ h$.
We have that
\[
k \circ h \circ j_{\ol{R}, \ol{f}}
= k \circ \iota_{\ol{\pi_1 R}, \ol{\pi_1 f}}
= j_{\ol{\Eq_{\pi_1 R}}, \ol{(\id, \setsem{\Gamma; \Phi \vdash B} c_\rho) \circ \Eq_{\pi_1 f}}}
\]
To prove that $j_{\ol{R}, \ol{f}} = j_{\ol{\Eq_{\pi_1 R}}, \ol{(\id, \setsem{\Gamma; \Phi \vdash B} c_\rho) \circ \Eq_{\pi_1 f}}}$,
it suffices to show that, in the indexing category of the colimit~\eqref{eq:RHS}, either the triangle
\[
\begin{tikzcd}[row sep = huge]
\relsem{\emptyset; \ol{\alpha} \vdash K}[\ol{\alpha := R}]
	\ar[rr, "{\relsem{\emptyset; \ol{\alpha} \vdash K}[\ol{\alpha := (\id, d_{R})}]}"]
	\ar[dr, "{f}"']
&& \relsem{\emptyset; \ol{\alpha} \vdash K}[\ol{\alpha := \Eq_{\pi_1 R}}]
	\ar[dl, "{(\id, \setsem{\Gamma; \Phi \vdash B} c_\rho) \circ \Eq_{\pi_1 f}}"] \\
&\relsem{\Gamma; \Phi \vdash B}\rho
\end{tikzcd}
\]
or the triangle
\[
\begin{tikzcd}[row sep = huge]
\relsem{\emptyset; \ol{\alpha} \vdash K}[\ol{\alpha := R}]
	\ar[dr, "{f}"']
&& \relsem{\emptyset; \ol{\alpha} \vdash K}[\ol{\alpha := \Eq_{\pi_1 R}}]
	\ar[dl, "{(\id, \setsem{\Gamma; \Phi \vdash B} c_\rho) \circ \Eq_{\pi_1 f}}"]
	\ar[ll, "{\relsem{\emptyset; \ol{\alpha} \vdash K}[\ol{\alpha := (\id, c_{R})}]}"'] \\
&\relsem{\Gamma; \Phi \vdash B}\rho
\end{tikzcd}
\]
commutes, but that does not seem to be the case.
Indeed, there is no reason why it should be true that $\setsem{\Gamma; \Phi \vdash B} c_\rho \circ (\pi_1 f) \circ \setsem{\emptyset; \ol{\alpha} \vdash K}[\ol{\alpha := d_{R}}] = (\pi_1 f)$
or $(\pi_1 f) \circ \setsem{\emptyset; \ol{\alpha} \vdash K}[\ol{\alpha := c_{R}}] = \setsem{\Gamma; \Phi \vdash B} c_\rho \circ (\pi_1 f)$.

Since our first approach did not work, we shall instead try to show that \eqref{eq:RHS} is isomorphic to
\begin{equation}\label{eq:RHS_EQ}
\colim{\ol{S : \set_0}, \ol{f : \relsem{\emptyset; \ol{\alpha} \vdash K}[\ol{\alpha := \Eq_S}] \to \relsem{\Gamma; \Phi \vdash B}\rho}}
{\setsem{\Gamma; \Phi, \ol{\alpha} \vdash F}(\pi_1 \rho)[\ol{\alpha := S}]}
\end{equation}
which should hold true if equality relations are cofinal in $\rel$.
Moreover, the fact that the colimit \eqref{eq:RHS_EQ} is indexed over sets suggests that it is also isomorphic to \eqref{eq:LHS}.
Before proceeding, though, we need to introduce a way to associate a set to an arbitrary relation, which we shall do in the following section.

\section{Sets from relations}

Given a set $S$, there are a few canonical ways to produce a relation that represents it.
\begin{itemize}
\item The most obvious one is the equality relation, $\Eq_S = (S, S, \Delta_S)$.
\item The relations $(S, \emptyset, \emptyset)$ and $(\emptyset, S, \emptyset)$, although they are not onto relations in the sense defined in the previous section.
\item The relations $(S, 1, S \times 1)$ and $(1, S, 1 \times S)$, although they are more of a set seen as a relation than a relation representing $S$.
\end{itemize}
Overall, the most natural option is the equality relation over a set.
Moreover, for the purposes of the Identity Extension Lemma, we really need to consider equality relations.

There is an equality functor $\Eq : \set \to \rel$ sending a set to its equality relation.
We want to find a left adjoint $\Sigma : \rel \to \set$, i.e., a way to construct a set $\Sigma{R}$ out of a relation $R$ together with a canonical arrow $R \to \Eq_{\Sigma{R}}$.

\begin{defi}
Given a relation $R : \rel(A, B)$, we define an equivalence relation $\sim$ on $A + B$ as the smallest equivalence relation generated by the pairs $(\SumIn_1 a, \SumIn_2 b)$ such that $(a, b) \in R$.
Explicitly, $\SumIn_1 a \sim \SumIn_2 b$ if and only if there is a finite sequence $a_1, b_1, a_2, b_2, \dots, a_n, b_n$ such that $a_1 = a$, $b_n = b$, $(a_i, b_i) \in R$ for all $i = 1 \dots n$ and $(a_{i + 1}, b_i) \in R$ for all $i = 1 \dots n - 1$.
We define $\sim$ analogously for the remaining three cases: $\SumIn_1 a \sim \SumIn_1 a'$, $\SumIn_2 b \sim \SumIn_2 b'$, and $\SumIn_2 b \sim \SumIn_1 a$.
Finally, we define $\Sigma{R}$ as the quotient $A + B / \sim$.

Given a morphism of relations $m = (m_1, m_2) : R \to Q$ we define the function $\Sigma{m} : \Sigma{R} \to \Sigma{Q}$ by $\Sigma{m}[\SumIn_1 a] = [ \SumIn_1 (m_1 a) ]$ and $\Sigma{m}[\SumIn_2 b] = [\SumIn_2 (m_2 b)]$.
To verify that $\Sigma{m}$ is well-defined, consider two representatives of the same class of equivalence, $\SumIn_1 a \sim \SumIn_2 b$.
Then there is a finite sequence $a_1, b_1, a_2, b_2, \dots, a_n, b_n$ such that $a_1 = a$, $b_n = b$, $(a_i, b_i) \in R$ for all $i = 1 \dots n$ and $(a_{i + 1}, b_{i}) \in R$ for all $i = 1 \dots n - 1$.
The sequence $m_1 a_1, m_2 b_1, m_1 a_2, m_2 b_2, \dots, m_1 a_n, m_2 b_n$ is such that $(m_1 a_i, m_2 b_i) \in Q$ for all $i = 1 \dots n$ and $(m_1 a_{i + 1}, m_2 b_{i}) \in Q$ for all $i = 1 \dots n - 1$, thus proving that $\Sigma{m}[\SumIn_1 a] = [\SumIn_1 (m_1 a)] = [\SumIn_2 (m_2 b)] = \Sigma{m}[\SumIn_2 b]$.

These data yield a functor $\Sigma : \rel \to \set$.
\end{defi}

We want to show that the above construction is the best way to represent a relation by a set, i.e., that there is an adjunction between the functors $\Sigma$ and $\Eq$.
The intuition that this is the case is reinforced by noticing that, for every relation $R : \rel(A, B)$, there is a morphism $(q \circ \SumIn_1, q \circ \SumIn_2) : R \to \Eq_{\Sigma{R}}$, where $q : A + B \to A + B/\sim$ is the projection onto the quotient.
The next proposition gives a complete proof of this fact.

\begin{prop}
There is an adjunction $\Sigma \dashv \Eq$.
\end{prop}
\begin{proof}
Let $R : \rel(A, B)$ and $S: \set$.
We want to show that there is a bijective correspondence between functions $\Sigma{R} \to S$ and morphisms of relations $R \to \Eq_S$.

Let $f : \Sigma{R} \to S$ be a function.
Define a morphism of relations $\tilde{f} = (\tilde{f}_1, \tilde{f}_2) : R \to \Eq_S$ as $\tilde{f}_1 a = f [ \SumIn_1 a ]$ and $\tilde{f}_2 b = f [ \SumIn_2 b ]$.
This is a well-defined morphism of relations: indeed, if $(a, b) \in R$, then $\SumIn_1 a \sim \SumIn_2 b$ and thus $\tilde{f}_1 a = f [ \SumIn_1 a ] = f [ \SumIn_2 b ] = \tilde{f}_2 b$.

Let $m = (m_1, m_2) : R \to \Eq_S$ be a morphism of relations.
Define a function $\hat{m} : \Sigma{R} \to S$ as $\hat{m} [ \SumIn_1 a ] = m_1 a$ and $\hat{m} [ \SumIn_2 b ] = m_2 b$.
This is well-defined: indeed, if $\SumIn_1 a \sim \SumIn_2 b$ (one out of four possible cases, the other three being $\SumIn_1 a \sim \SumIn_1 a'$, $\SumIn_2 b \sim \SumIn_1 a$, and $\SumIn_2 b \sim \SumIn_2 b'$), there is a finite sequence $a_1, b_1, a_2, b_2, \dots, a_n, b_n$ such that $a_1 = a$, $b_n = b$, $(a_i, b_i) \in R$ for all $i = 1 \dots n$ and $(a_{i + 1}, b_{i}) \in R$ for all $i = 1 \dots n - 1$.
Then $m_1 a_1 = m_2 b_1 = m_1 a_2 = m_2 b_2 = \dots = m_1 a_n = m_2 b_n$, thus proving that $\hat{m}[\SumIn_1 a] = m_1 a = m_2 b = \hat{m}[\SumIn_2 b]$.
We have analogous proofs for the other three cases.

We now want to show that the two constructions are inverse to each other.
In one direction, $\hat{\tilde{f}} [\SumIn_1 a] = \tilde{f}_1 a = f [\SumIn_1 a]$ and $\hat{\tilde{f}} [\SumIn_2 b] = \tilde{f}_2 b = f [\SumIn_2 b]$, and thus $\hat{\tilde{f}} = f$.
In the other direction, $\tilde{\hat{m}}_1 a = \hat{m} [\SumIn_1 a] = m_1 a$ and $\tilde{\hat{m}}_2 b = \hat{m} [\SumIn_2 b] = m_2 b$, and thus $\tilde{\hat{m}} = m$.
\end{proof}

Observe that, for every relation $R : \rel(A, B)$, we can use the axiom of choice to define choice functions $\Xi_1 : \Sigma R \to A$ and $\Xi_2 : \Sigma R \to B$ such that $\SumIn_1 ( \Xi_1 x ) \in x$, $\SumIn_2 (\Xi_2 x ) \in x$, and $(\Xi_1 x, \Xi_2 x) \in R$ for every $x : \Sigma R$.
In other words, $(\Xi_1, \Xi_2)$ chooses a pair of related elements out of each equivalence class of $\Sigma R$.
Moreover, since $(\Xi_1 x, \Xi_2 x) \in R$ for every $x : \Sigma R$, it's immediate that $(\Xi_1, \Xi_2) : \Eq_{\Sigma R} \to R$ is a morphism of relations.

\section{Counterexample to fibered interpretation of $\Lan$}

We want to show that there is an isomorphism between \eqref{eq:RHS} and \eqref{eq:RHS_EQ}.
The situation is represented in the following diagram
\[
\begin{tikzcd}[column sep = huge]
&\setsem{\Gamma; \Phi, \ol\alpha \vdash F} (\pi_1 \rho) [\ol{\alpha := S}]
\ar[d, "{l_{\ol{S}, \ol{g}}}"]
\ar[dl, bend right = 15, "{j_{\ol{\Eq_{S}}, \ol{g}}}"'] \\
\eqref{eq:RHS}
\ar[r, bend right = 5, "{\Phi}"']
&\eqref{eq:RHS_EQ}
\ar[l, bend right = 5, "{\Psi}"'] \\
\setsem{\Gamma; \Phi, \ol\alpha \vdash F} (\pi_1 \rho) [\ol{\alpha := \pi_1 R}]
\ar[r, bend right = 5, "{\setsem{\Gamma; \Phi, \ol\alpha \vdash F} \id_{\pi_1 \rho} [\ol{\alpha := q \circ \SumIn_1}]}"']
\ar[u, "{j_{\ol{R}, \ol{f}}}"]
&\setsem{\Gamma; \Phi, \ol\alpha \vdash F} (\pi_1 \rho) [\ol{\alpha := \Sigma{R}}]
\ar[u, "{l_{\ol{\Sigma{R}}, \ol{f \circ \relsem{\emptyset; \ol\alpha \vdash K}[\ol{\alpha := (\Xi_1, \Xi_2)}]}}}"']
\end{tikzcd}
\]
where $l_{\ol{S}, \ol{g}} : \setsem{\Gamma; \Phi, \ol\alpha \vdash F} (\pi_1 \rho) [\ol{\alpha := S}] \to \eqref{eq:RHS_EQ}$ is the cocone into the colimit~\eqref{eq:RHS_EQ}.

We define $\Phi : \eqref{eq:RHS} \to \eqref{eq:RHS_EQ}$ as the unique function such that $\Phi \circ j_{\ol{R}, \ol{f}} = l_{\ol{\Sigma{R}}, \ol{f \circ \relsem{\emptyset; \ol\alpha \vdash K}[\ol{\alpha := (\Xi_1, \Xi_2)}]}} \circ \setsem{\Gamma; \Phi, \ol\alpha \vdash F} \id_{\pi_1 \rho}[\ol{\alpha := q \circ \SumIn_1}]$ for every $\ol{R : \rel_0}$ and $\ol{f : \relsem{\emptyset; \ol{\alpha} \vdash K}[\ol{\alpha := R}] \to \relsem{\Gamma; \Phi \vdash B}\rho}$.

We define $\Psi : \eqref{eq:RHS_EQ} \to \eqref{eq:RHS}$ as the unique function such that $\Psi \circ l_{\ol{S}, \ol{g}} = j_{\ol{\Eq_S}, \ol{g}}$ for every $\ol{S : \set_0}$ and $\ol{g : \relsem{\emptyset; \ol{\alpha} \vdash K}[\ol{\alpha := \Eq_S}] \to \relsem{\Gamma; \Phi \vdash B}\rho}$.

We want to show that $\Psi$ and $\Phi$ are inverses.
For the composition $\Phi \circ \Psi$, we have that
\[
\Phi \circ \Psi \circ l_{\ol{S}, \ol{g}}
= \Phi \circ j_{\ol{\Eq_S}, \ol{g}}
= l_{\ol{\Sigma{\Eq_S}}, \ol{g \circ \relsem{\emptyset; \ol\alpha \vdash K}[\ol{\alpha := (\Xi_1, \Xi_2)}]}} \circ \setsem{\Gamma; \Phi, \ol\alpha \vdash F} \id_{\pi_1 \rho}[\ol{\alpha := q \circ \SumIn_1}]
\]
To show that $l_{\ol{S}, \ol{g}} = l_{\ol{\Sigma{\Eq_S}}, \ol{g \circ \relsem{\emptyset; \ol\alpha \vdash K}[\ol{\alpha := (\Xi_1, \Xi_2)}]}} \circ \setsem{\Gamma; \Phi, \ol\alpha \vdash F} \id_{\pi_1 \rho}[\ol{\alpha := q \circ \SumIn_1}]$, consider the following diagram in the indexing category:
\[
\begin{tikzcd}[column sep = huge]
\relsem{\emptyset; \ol{\alpha} \vdash K}[\ol{\alpha := \Eq_S}]
\ar[r, bend left = 5, "{\relsem{\emptyset; \ol{\alpha} \vdash K}[\ol{\alpha := (q \circ \SumIn_1, q \circ \SumIn_2)}]}"]
\ar[d, "{g}"']
& \relsem{\emptyset; \ol{\alpha} \vdash K}[\ol{\alpha := \Eq_{\Sigma{\Eq_S}}}]
\ar[d, "{\relsem{\emptyset; \ol{\alpha} \vdash K}[\ol{\alpha := (\Xi_1, \Xi_2)}]}"] \\
\relsem{\Gamma; \Phi \vdash B}\rho
& \relsem{\emptyset; \ol{\alpha} \vdash K}[\ol{\alpha := \Eq_{S}}]
\ar[l, "{g}"]
\end{tikzcd}
\]
The diagram commutes because $(\Xi_1 \circ q \circ \SumIn_1, \Xi_2 \circ q \circ \SumIn_2) = \id_{\Eq_S}$, i.e., $\Xi_1 \circ q \circ \SumIn_1 = \id_S$ and $\Xi_2 \circ q \circ \SumIn_2 = \id_S$.
Indeed, $q$ is the projection into the quotient for the equivalence relation $\sim_{\Eq_S}$, for which $\SumIn_i x \sim_{\Eq_S} \SumIn_{i'} x'$ if and only if $x = x'$ for $i, i' = 1, 2$, so that $\Xi_1 ( q ( \SumIn_1 x ) ) = x$ and $\Xi_2 ( q ( \SumIn_2 x ) ) = x$.
Thus, $l_{\ol{S}, \ol{g}} = l_{\ol{\Sigma{\Eq_S}}, \ol{g \circ \relsem{\emptyset; \ol\alpha \vdash K}[\ol{\alpha := (\Xi_1, \Xi_1)}]}} \circ \setsem{\Gamma; \Phi, \ol\alpha \vdash F} \id_{\pi_1 \rho}[\ol{\alpha := q \circ \SumIn_1}]$
and we can conclude that $\Phi \circ \Psi = \id$.

In the other direction, for the composition $\Psi \circ \Phi$, we have that
\begin{align*}
\Psi \circ \Phi \circ j_{\ol{R}, \ol{f}}
&= \Psi \circ l_{\ol{\Sigma{R}}, \ol{f \circ \relsem{\emptyset; \ol\alpha \vdash K}[\ol{\alpha := (\Xi_1, \Xi_2)}]}} \circ \setsem{\Gamma; \Phi, \ol\alpha \vdash F} \id_{\pi_1 \rho}[\ol{\alpha := q \circ \SumIn_1}] \\
&= j_{\ol{\Eq_{\Sigma{R}}}, \ol{f \circ \relsem{\emptyset; \ol\alpha \vdash K}[\ol{\alpha := (\Xi_1, \Xi_2)}]}} \circ \setsem{\Gamma; \Phi, \ol\alpha \vdash F} \id_{\pi_1 \rho}[\ol{\alpha := q \circ \SumIn_1}]
\end{align*}
To show that
\begin{equation}\label{eq:PsiPhi}
j_{\ol{R}, \ol{f}} = j_{\ol{\Eq_{\Sigma{R}}}, \ol{f \circ \relsem{\emptyset; \ol\alpha \vdash K}[\ol{\alpha := (\Xi_1, \Xi_2)}]}} \circ \setsem{\Gamma; \Phi, \ol\alpha \vdash F} \id_{\pi_1 \rho}[\ol{\alpha := q \circ \SumIn_1}]
\end{equation}
it would suffice the following diagram in the indexing category to commute:
\begin{equation}\label{eq:index-comm}
\begin{tikzcd}[column sep = huge]
\relsem{\emptyset; \ol{\alpha} \vdash K}[\ol{\alpha := R}]
\ar[r, bend left = 5, "{\relsem{\emptyset; \ol{\alpha} \vdash K}[\ol{\alpha := (q \circ \SumIn_1, q \circ \SumIn_2)}]}"]
\ar[d, "{f}"']
& \relsem{\emptyset; \ol{\alpha} \vdash K}[\ol{\alpha := \Eq_{\Sigma{R}}}]
\ar[d, "{\relsem{\emptyset; \ol{\alpha} \vdash K}[\ol{\alpha := (\Xi_1, \Xi_2)}]}"] \\
\relsem{\Gamma; \Phi \vdash B}\rho
& \relsem{\emptyset; \ol{\alpha} \vdash K}[\ol{\alpha := R}]
\ar[l, "{f}"]
\end{tikzcd}
\end{equation}
As before, this requires $(\Xi_1 \circ q \circ \SumIn_1, \Xi_2 \circ q \circ \SumIn_2) = \id_{R}$, i.e., that $\Xi_1 \circ q \circ \SumIn_1 = \id_{\pi_1 R}$ and $\Xi_2 \circ q \circ \SumIn_2 = \id_{\pi_2 R}$, where $q$ is the projection into the quotient for the equivalence relation $\sim_{R}$.
But that was true in the previous direction because we were considering an equality relation, for which the equivalence classes are singletons.
For a generic relation $R$, as in this case, there seem to be no reason why this should be true, and indeed we shall provide a counterexample.

It is worth noting that we have not definitively disproved Equation~\ref{eq:PsiPhi}, but we have just shown how this way of proving the equation does not work.

\begin{exa}
Let $R$ be the relation $(\{ 1, 2, 3\}, \{a, b\}, \{(1, a), (2, b), (3, b)\}, d_R, c_R)$ where $d_R$ maps $a$ into $1$ and $b$ into $2$, and $c_R$ maps $1$ into $a$, and $2$ and $3$ into $b$.
Thus, $\Sigma{R}$ is the set $\{ \{1, a\}, \{2, 3, b\} \}$.
We also have the morphism of relations $(q \circ \SumIn_1, q \circ \SumIn_2) : R \to \Eq_{\Sigma{R}}$ where $q \circ \SumIn_1$ sends $1$ into $\{1, a\}$, and $2$ and $3$ into $\{2, 3, b\}$; $q \circ \SumIn_2$ sends $a$ into $\{1, a\}$, and $b$ into $\{2, 3, b\}$.
Moreover, we have the choice functions $\Xi_1 : \Sigma R \to \{ 1, 2, 3\}$ sending $\{1, a\}$ to $1$ and $\{2, 3, b\}$ to $2$, and $\Xi_2 : \Sigma_R \to \{a, b\}$ sending $\{1, a\}$ to $a$ and $\{2, 3, b\}$ to $b$.

Let $K = \alpha$, $B = \beta$ with $\rho \beta = R$ and $f = \id_R$.
Instantiating the above diagram, we get that it commutes if and only if $(\Xi_1, \Xi_2) \circ (q \circ \SumIn_1, q \circ \SumIn_2) = \id_R$.
But $\Xi_1 \circ q \circ \SumIn_1$ is not the identity (as it is not injective, because $\Xi_1 ( q ( \SumIn_1 2 ) ) = \Xi_1 \{ 2, 3, b \} = \Xi_1 ( q ( \SumIn_1 3 ) )$) and thus the diagram does not commute.

This still does not exclude that Equation~\eqref{eq:PsiPhi} might hold.
Let $F = \alpha$.
Then Equation~\eqref{eq:PsiPhi} becomes $j_{\ol{\Eq_{\Sigma{R}}}, \ol{(\Xi_1, \Xi_2)}} \circ q \circ \SumIn_1 = j_{\ol{R}, \ol{\id}}$, as shown in the following diagram:
\[
\begin{tikzcd}
\pi_1 R
\ar[r, "{j_{\ol{R}, \ol{\id}}}"]
\ar[d, "{q \circ \SumIn_1}"']
&\colim{Q: \rel_0, f: Q \to R}{\pi_1 Q} \\
\Sigma{R}
\ar[ur, "{j_{\ol{\Eq_{\Sigma{R}}}, \ol{(\Xi_1, \Xi_2)}}}"']
\end{tikzcd}
\]
Notice that $\colim{Q: \rel_0, f: Q \to R}{\pi_1 Q} = \pi_1 \colim{Q: \rel_0, f: Q \to R}{Q} = \pi_1 R$, and thus $j_{\ol{R}, \ol{\id}} = \id_{\pi_1 R}$ and $j_{\ol{\Eq_{\Sigma{R}}}, \ol{(\Xi_1, \Xi_2)}} = \Xi_1$.
Finally, Equation~\eqref{eq:PsiPhi} becomes $\Xi_1 \circ q \circ \SumIn_1 = \id_{\pi_1 R}$, which we already showed being false.
\end{exa}

Observe that, for the choice of $K$, $F$ and $B$ as in the example above, it is actually the case that $\eqref{eq:LHS} \cong \eqref{eq:RHS}$, as they are both equal to $\pi_1 R$.
Indeed, by instantiating, we have that $\eqref{eq:LHS} = \colim{\ol{S : \set_0}, \ol{g : S \to \pi_1 R}}{S} = \pi_1 R$ and
$\eqref{eq:RHS} = \colim{\ol{Q : \rel_0}, \ol{f : Q \to R}}{\pi_1 Q} = \pi_1 \colim{\ol{Q : \rel_0}, \ol{f : Q \to R}}{Q} = \pi_1 R$.
So, the above is not a counterexample to $\eqref{eq:LHS} \cong \eqref{eq:RHS}$, but just to the commutativity of Diagram~\ref{eq:index-comm}.

\bibliographystyle{alpha}
\bibliography{references}

\end{document}

### 2021-02-11 20-gadts/GADT-story.tex
added GADT-story.tex

% For double-blind review submission, w/o CCS and ACM Reference (max
% submission space)
\documentclass[acmsmall,screen,review,anonymous]{acmart}
\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

%% Journal information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{POPL} % CONF = POPL or ICFP or OOPSLA
\acmArticle{1}
\acmYear{2021}
\acmMonth{1}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2018}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations
%\citestyle{acmnumeric}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Note: Authors migrating a paper from PACMPL format to traditional
%% SIGPLAN proceedings format must update the '\documentclass' and
%% topmatter commands above; see 'acmart-sigplanproc-template.tex'.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage[utf8]{inputenc}
\usepackage{ccicons}
\usepackage{verbatim}

\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amscd}
%\usepackage{MnSymbol}
\usepackage{xcolor}

\usepackage{bbold}
\usepackage{url}
\usepackage{upgreek}
%\usepackage{stmaryrd}

\usepackage{lipsum}
\usepackage{tikz-cd}
\usetikzlibrary{cd}
\usetikzlibrary{calc}
\usetikzlibrary{arrows}

\usepackage{bussproofs}
\EnableBpAbbreviations

\input{macros}
\input{mytheorem}

\theoremstyle{definition}
\newtheorem{exmpl}{Example}

\renewcommand{\greyout}[1]{}

\newcommand{\new}[1]{{\color{blue} {#1}}}

\newcommand{\emptyfun}{{[]}}
\newcommand{\cal}{\mathcal}
%\newcommand{\fold}{\mathit{fold}}
\newcommand{\F}{\mathcal{F}}
\renewcommand{\G}{\mathcal{G}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\B}{\mathcal{B}}
\renewcommand{\P}{\mathcal{A}}
\newcommand{\pred}{\mathsf{Fam}}
\newcommand{\env}{\mathsf{Env}}
\newcommand{\set}{\mathsf{Set}}
\renewcommand{\S}{\mathcal S}
\renewcommand{\C}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\A}{\mathcal{A}}
\renewcommand{\id}{\mathit{id}}
\newcommand{\map}{\mathsf{map}}
\newcommand{\pid}{\underline{\mathit{id}}}
\newcommand{\pcirc}{\,\underline{\circ}\,}
\newcommand{\pzero}{\underline{0}}
\newcommand{\pone}{\underline{1}}
\newcommand{\psum}{\,\underline{+}\,}
\newcommand{\pinl}{\underline{\mathit{inL}}\,}
\newcommand{\pinr}{\underline{\mathit{inR}}\,}
\newcommand{\ptimes}{\,\underline{\times}\,}
\newcommand{\ppi}{\underline{\pi_1}}
\newcommand{\pppi}{\underline{\pi_2}}
\newcommand{\pmu}{\underline{\mu}}
\newcommand{\semmap}{\mathit{map}}
\newcommand{\subst}{\mathit{subst}}

\newcommand{\tb}[1]{~~ \mbox{#1} ~~}
\newcommand{\listt}[1]{(\mu \phi. \lambda \beta . \onet + \beta \times
  \phi \beta) #1} 
\newcommand{\filtype}{\Nat^\emptyset 
 (\Nat^\emptyset \, \alpha \, \mathit{Bool})\, (\Nat^\emptyset 
  (List \, \alpha) \, (List \, \alpha))} 
\newcommand{\filtypeGRose}{\Nat^\emptyset 
 (\Nat^\emptyset \, \alpha \, \mathit{Bool})\, (\Nat^\emptyset 
  (\mathit{GRose}\,\psi \, \alpha) \, (\mathit{GRose}\,\psi \, (\alpha
  + \onet)))} 
\newcommand{\maplist}{\mathit{map}_{\lambda A. \setsem{\emptyset; \alpha
      \vdash \mathit{List} \, \alpha} \rho[\alpha := A]}} 
\newcommand{\PLeaves}{\mathsf{PLeaves}}
\newcommand{\swap}{\mathsf{swap}}
\newcommand{\reverse}{\mathsf{reverse}}
\newcommand{\Bcons}{\mathit{Bcons}}
\newcommand{\Bnil}{\mathit{Bnil}}

\title[Parametricity Consequences of GADTs
  Representations]{Parametricity Consequences of GADTs
  Representations}

\author{Patricia Johann, Enrico Ghiorzi, and Daniel Jeffries}
\affiliation{ \institution{Appalachian State University}}
\email{johannp@appstate.edu, ghiorzie@appstate.edu,
  jeffriesd@appstate.edu}

\begin{document}

\begin{abstract}
FIX ME!!!! GADTs are misnamed. They don't really generalize ADTs. And
they don't behave the way we expect data types to behave. In
particular, they force us to choose between the naturality properties
we expect of their fmap functions and other parametricity results for
them, causing us to question in what sense GADTs are data types, and
what the value of the parametricity results for them as determined by
their syntax really is.
\end{abstract}

\maketitle

\section{Introduction}

There are two standard ways to represent data types when studying
modern functional languages: via their Church encodings in a (possibly
higher-kinded) extension of System F~\cite{gir72}, the calculus at the
core of all such languages, and by augmenting System F
%a Hindley-Milner style calculus
with primitives for constructing them directly as fixpoints.
%{\color{blue} Something about when each of these representations is
%  used. The former perhaps more closely represents how data types are
%  represented in implementations, but the latter are how they have
%  heretofore primarily been treated theoretically.}
In models that validate the program optimization known as {\em short
  cut fusion}~\cite{glp93} the two representations are, in fact,
equivalent, in the sense that they both have the same interpretation
in the model. Models in which the Church encodings and primitive
representations of all {\color{blue} (polynomial)} algebraic data
types (ADTs) are equivalent include the operationally-based model
of~\cite{pit98,pit00} and the categorical models
of~\cite{joh02,joh03}.
%\footnote{{\color{blue} Pitts proves the isomorphism only for
%    polynomial ADTs, whereas Johann {\em et al.}  prove it for any
%    data type that is the fixpoint of a first-order functor.}}
This equivalence was recently extended to the class of what might be
called ``hereditarily polynomial'' {\color{blue} to include truly
  nested such types} nested types via the categorical model
of~\cite{jgj21}. But, as shown in Section~\ref{sec:?} below, for some
{\color{blue} (even polynomial)} GADTs,
%that are not hereditarily polynomial nested types,
no such isomorphism holds.

GADTs are traditionally represented via Church encodings. In this
view, GADTs are completely determined by their syntax. For example,
the GADT $\mathsf{G}$ defined by
\[\begin{array}{l}
\mathsf{data\,G : * \to *}\\
\mathsf{\;\;\;\;\;\;\;\;c : G\,1}
\end{array}\]
{\color{blue} Use $\top$ instead of $1$ throughout?}

\noindent
{\color{blue} data type vs data structure --- which should have the
functorial completion? The data type.}

\noindent
comprises exactly one element, namely $\mathsf{c}$. By contrast, the
primitive representation sees a GADT as the {\em functorial
  completion}~\cite{jp19} of its syntax. {\color{blue} Completing the
  syntax is necessary to have a {\em functor} to take the least fixed
  point of, so is essential to the view of data types as primitive.}
Viewed as a primitive data type, $\mathsf{G}$ comprises not just
$\mathsf{c}$, but also all data elements of the form
$\mathsf{map\;f\;c}$ for all types $\mathsf{t}$ and all functions
$\mathsf{f : 1 \to t}$ definable in the language, as well as all data
elements of the form $\mathsf{map\;g\;x}$ for each appropriately typed
function $\mathsf{g}$ and each element $\mathsf{x}$ already added to
the data type, and so on.  Although functorial completion adds no new
data elements to the syntax of a (polynomial) ADT or a (hereditarily
polynomial) nested type, so that its primitive Church encoding
representations coincide, these are rarely the same for GADTs.
Indeed, they are quite different for even the very simple GADT
$\mathsf{G}$ above.

But does it really matter which way we view GADTs?

A GADT programmer is likely to use GADTs precisely because they
exhibit different behaviors at different types, and thus to consider
the GADT to be completely specified by its syntax. When used this way,
the shape of a GADTs is actually {\em determined} by the data it
contains. As a result, it is not expected to support a $\mathsf{map}$
function --- i.e., to be an instance of Haskell's $\mathsf{Functor}$
class --- like other data types. For example, the definition for
$\mathsf{G}$ above specifies data only at instance $\mathsf{G\,1}$.
Since there is a function from $\mathsf{1}$ to any other type
$\mathsf{t}$, and yet $\mathsf{G\,t}$ is uninhabited at any instance
other than $\mathsf{1}$, $\mathsf{G}$ cannot support a $\mathsf{map}$
function. When it is intended to capture this kind of non-uniform
behavior, a GADT's behavior does not generalize that of ADTs as
advertised, and in some sense doesn't even behave like a traditional
data type.  When used this way, a GADT cannot be seen as the least
fixpoint of a functor, and must instead be represented via its Church
encoding.

A semanticist, on the other hand, is likely to follow~\cite{bfss90}
and expect GADTs to behave as their names suggests --- i.e., as
generalizations of ADTs. The semanticist is thus likely to expect a
GADT to be able to be used as a ADT is used --- i.e., as a shape
filled with data that can be changed without also changing the
shape. When used this way, it is more natural to view the GADT in
terms of its primitive representation. In fact, the functorial
completion of a GADT inherent in its primitive representation is
exactly what is needed to ensure that the GADT supports a
$\mathsf{map}$ function, i.e., is a functor both in the catgorical
sense and in the sense of being an instance of Haskell's
$\mathsf{Functor}$ class.  It is worth noting, however, that since the
functorial completion of a GADT includes only those data elements that
are derivable via repeated applications of $\mathsf{map}$ starting
from those specified by its syntax, primitive GADTs need not, in
general, be nested types. For example, the functorial completion of
our GADT $\mathsf{G}$ includes elements at any instance
$\mathsf{G\,t}$ because there is a function from $\mathsf{1}$ to any
type $\mathsf{t}$, but it does not include any terms at instance
$\mathsf{G\,0}$ because no calculus allows the definition of a
function into $\mathsf{0}$. {\color{blue} Use $\bot$ instead of $0$
  here?}

The key observation of this paper is that, while the viewpoints of
both the GADT programmer and the semanticist are both valid, the two
are irreconcilable. Unsurprisingly, which point of view is adopted has
deep consequences for the ways GADTs can be used and reasoned
about. In this paper we show that the choice of GADT representation
has deep implications for parametric reasoning about structures of
those types. Indeed, a GADT programmer who views GADTs as their Church
encodings cannot safely use program transformations or reasoning
principles for them that involve map functions for them. This may
surprise the GADT programmer by being contrary to expectation for
something supposedly generalizing a data type. The GADT programmer
may, however, be able to program with and reason about GADTs using
non-naturality consequences of parametricity, such as type
inhabitation results. On the other hand, a semanticist who views GADTs
as primitive data types will have all naturality-based program
transformations and reasoning principles for GADTs at their disposal,
but non-naturality consequences of parametric reasoning may not be
available to them.

In this paper we show how these observations can be made precise. That
is, we show that how we view GADTs has significant consequences, and
thus answer the question we posed above in the affirmative: Yes, it
really does matter.

\section{Representations of Data Types}

\subsection{Algebraic Data Types}

A (polynomial) {\em algebraic data type} (ADT) has the form
\[\mathsf{T\,a} = \mathsf{c_1 t_{11}}...\mathsf{t_{1k_1}}\, |\, ...\, |\,
\mathsf{c_n t_{n1}}...\mathsf{t_{nk_n}}\] where each $\mathsf{t_{ij}}$
is a type also depending only on $\mathsf{a}$. Such a data type can be
thought of as a ``container'' for data of type $\mathsf{a}$. The data
in an ADT are arranged at various {\em positions} in its underlying
{\em shape}, which is determined by the types of its {\em
  constructors} $\mathsf{c_1},...,\mathsf{c_n}$. An ADT's constructors
are used to build the data values of the data type, as well as to
analyze those values using {\em pattern matching}. ADTs are used
extensively in functional programming to structure computations, to
express invariants of the data over which computations are defined,
and to ensure the type safety of programs specifying those
computations.

List types are the quintessential examples of ADTs. The shape of the
container underlying the list type
\[\mathsf{List\,a} = \mathsf{Nil} \,|\, \mathsf{Cons\,a\,(List\,a)}\]
%of lists of elements of type $\mathsf{a}$
is determined by the types of its two constructors $\mathsf{Nil\, ::\,
  List\, a}$ and $\mathsf{Cons \,::\,a \to \List\,a\to
  List\,a}$. These constructors specify that the data in a list of
type $\mathsf{List\,a}$ are arranged linearly. The shape underlying
the type $\mathsf{List\,a}$ is therefore given by the set $\nat$ of
natural numbers, and the positions in a structure of this type are
given by specific natural numbers. Since the argument to every
occurrence of the type constructor $\mathsf{List}$ in the right-hand
side of above definition is the same as on its left-hand side, the
type $\mathsf{List\,a}$ enforces the invariant that all of the data in
a structure of this type have the same type $\mathsf{a}$. In a similar
way, the tree type
\[\mathsf{Tree\, a} = \mathsf{Leaf\,a}
\,|\,\mathsf{Node\,(Tree\,a)\,a\,(Tree\,a)}\] of binary trees
has as its underlying shape the type of binary trees of units, and the
positions in a structure of this type are given by sequences of L (for
``left'') and R (for ``right'') navigating a path
%to either a node or a leaf in
through the structure. The type $\mathsf{Tree\, a}$ enforces the
invariant that all of the data at the nodes and leaves in a structure
of this type have the same type $\mathsf{a}$.

{\color{blue} polynomial ADTs?} Since the shape of an ADT structure
--- i.e., a structure whose type is an instance of an ADT --- is
independent of the type of data it contains, ADTs can be defined
polymorphically. As a result, an ADT structure containing data of type
$\mathsf{a}$ can be transformed into another ADT structure of the
exact same shape containing data of another type $\mathsf{b}$ simply
by applying a given function $\mathsf{f \,:\,a \to b}$ to each of its
elements. Indeed, every ADT can be made an instance of Haskell's
$\mathsf{Functor}$ class by defining a type-and-data-uniform
type-preserving, data-changing $\mathsf{fmap}$ {\color{blue} map
  vs. fmap} function for it.  Moreover, given a type-independent way
of rearranging an ADT structure's shape $\mathsf{T\,a}$ into the shape
for another ADT structure $\mathsf{T'\,a}$, we get the same structure
of type $\mathsf{T'\,b}$ regardless of whether we first rearrange the
original structure of type $\mathsf{T\,a}$ into one of type
$\mathsf{T'\,a}$ and then use $\mathsf{fmap_{T'}}$ to convert that
resulting structure to one of type $\mathsf{T'\,b}$, or we first use
$\mathsf{fmap_T}$ to convert the original structure of type
$\mathsf{T\,a}$ to one of type $\mathsf{T\,b}$ and then rearrange that
resulting structure into one of type $\mathsf{T'\,b}$. For example, if
$\mathsf{f : a \to b}$, $\mathsf{xs \,:\,List\,a}$, and $\mathsf{g :
  List\,a\to Tree\,a}$ arranges lists into trees in a type-independent
way, then
\[\mathsf{fmap_{Tree}\,f\,(g \, xs) \,=\, g\,(fmap_{List}\,f\,xs)}\]

\subsubsection{Church Encodings of ADTs}

One way to represent ADTs is via their Church encodings. A Church
encoding is a representation of a data type as a pure function in a
lambda calculus (System F in the case of functional languages). Church
encodings can be used to represent data types in languages that do not
support primitives for sum types, product types, or recursion.
{\color{blue} Church encodings (and the related Parigot encodings)
  have recently been popularized as {\em visitor patterns} in
  object-oriented programming.} Represention of ADTs as their Church
encodings is possible in any language that supports functions.

The Church encodings of the ADTs $\mathsf{List\,a}$ and
$\mathsf{Tree\,a}$, for example, are
\[\mathsf{List\,a} = \mathsf{\forall b.\,b \to (a \to b \to b) \to b}\]
and 
\[\mathsf{Tree\,a} = \mathsf{\forall b.\,(a \to b) \to (b \to a \to b
  \to b) \to b}\] respectively. The argument types are abstractions
of the types of ADTs' constructors. For instance, $\mathsf{b}$
abstracts the type of the constructor $\mathsf{Nil}$, and $\mathsf{a
  \to b \to b}$ abstracts the type of the constructor $\mathsf{Cons}$,
for lists. Because these types are each uniform in their argument type
$\mathsf{a}$, we can also encode the type constructors $\mathsf{List}$
and $\mathsf{Tree}$ themselves as
\[\mathsf{List} = \mathsf{\forall a. \forall b.\,b \to (a \to b \to b)
  \to b}\] 
and 
\[\mathsf{Tree} = \mathsf{\forall a. \forall b.\,(a \to b) \to (b \to
  a \to b \to b) \to b}\]
respectively.

It is possible to make $\mathsf{List}$ and $\mathsf{Tree}$ instances
of Haskell's $\mathsf{Functor}$ class. Indeed, the Prelude's
$\mathsf{map}$ function does exactly that for $\mathsf{List}$. For
$\mathsf{Tree}$ we can define
\[\begin{array}{lll}
\mathsf{map_{Tree}} & \mathsf{::} & \mathsf{(a \to b) \to Tree\,a \to Tree \,b}\\
\mathsf{map_{Tree}\, f\, (Leaf\,x)} & \mathsf{=} & \mathsf{Leaf\,(f\,x)}\\
\mathsf{map_{Tree}\, f\, (Node\,t_1\,x\,t_2)} & \mathsf{=} &
\mathsf{Node\,(map_{Tree}\, f \,t_1)\,x\,(map_{Tree}\,f\,t_2)}\\
\end{array}\]
But the Church encoding of an ADT carries with it no expectation
whatsoever that such a type-and-data-uniform, type-preserving,
data-changing $\mathsf{map}$ function can be defined. Nevertheless,
for polynomial ADTs this can always be done. But even when the Church
encoding of an ADT supports a $\mathsf{map}$ function, there is no
guarantee that the rearrange-transform property will hold for it.

\subsubsection{ADTs as Primitive Data Types}

By contrast, the ability to define such a $\mathsf{map}$ function is
inherent in the view of ADTs as primitive data types. Such a view is
possible in any language that supports primitives for sum types,
product types, and recursion. In such a language, the primitive
representations of the ADTs $\mathsf{List\,a}$ and $\mathsf{Tree\,a}$
are
\[\mathsf{List\,a} = \mathsf{\mu b.\, 1 + (a \times b)}\] 
and 
\[\mathsf{Tree\,a} = \mathsf{\mu b.\, a + (b \times a \times b)}\]
respectively. Here, $\mathsf{\mu}$ is the primitive fixpoint
operator.

Primitive representations capture in syntax the fact that ADTs can be
considered as fixpoints. For example $\mathsf{List\,a}$ does indeed
satisfy the fixpoint equation indicated above, i.e., is indeed such
that
\[\mathsf{List\,a} = \mathsf{1 + (a \times List\,a)}\] 
since every element of $\mathsf{List\,a}$ is either empty or is
obtained by $\mathsf{Cons}$ing an element of type $\mathsf{a}$ onto an
already-existing structure of type $\mathsf{List\,a}$. (The fixpoint
equation is just a rewriting of the Haskell data type declaration for
$\mathsf{List\,a}$.) The ADT $\mathsf{Tree\,a}$ can similarly be seen
to be a solution to the fixpoint equation
\[\mathsf{Tree\,a} = \mathsf{a + (Tree\,a \times a \times Tree\,a)}\] 
determined by its Haskell data type declaration.

These fixpoint equations are sensible at the level of types. But to
ensure that the syntactic fixpoint actually denotes a semantic object
computed as a real-live fixpoint, the semantic fixpoint calculation
must converge. This means that the fixpoint being taken must be of
{\em functors} on the category interpreting types~\cite{tfca}, rather
than mere functions between the sets interpreting them. Said
differently, functions between sets interpreting types must have
corresponding {\em functorial actions} on functions between
interpretations of types. Reflecting this requirement back into syntax
gives that the body of a primitive data type representation of an ADT
must support a $\mathsf{map}$ function, i.e., must be an instance of
Haskell's $\mathsf{Functor}$ class.

A concrete example is given by the ADT $\mathsf{List\,a}$. This type
is is modelled as the fixpoint $\mu F_a$ of the first-order functor
whose action on objects is given by $F_a\,X = 1 + a \times X$ and,
whose action on morphisms (which models its corresponding
$\mathsf{map}$ function) is given by $F_a\,f = 1 + \id_a \times
f$. The type constructor $\mathsf{List}$ is then modelled by the
higher-order functor $H$ whose action on objects is given by $H\,F\,X
= 1 + X \times H\,F\,X$ and whose action on morphisms is givne by
$H\,F\,f = \id_1 + f \times H\,F\,f$. (A {\em higher-order functor} in
this context is a functor not directly on the category interpreting
types, but rather on the category of functors interpreting type
constructors, i.e., on the category of functors on the category of
functors interpreting types.) The ADT $\mathsf{List\,a}$ is then
modelled as the functor $(\mu H)a$. Since $H$ is a higher-order
functor, not only does it have a corresponding functorial action,
i.e., support a function $\mathsf{map_H}$, but $\mu H$ is also a
functor with its own functorial action. That is, the type constructor
$\mathsf{List}$ itself supports a function $\mathsf{fmap_{List}}$.

{\color{blue} At the level of objects/types, short cut fusion can be
  used to show that the Church encoding and the primitive
  representation of an ADT are isomorphic.} Note, however, that
whereas it is impossible even to {\em state} the rearrange-transform
property for ADTs whose Church encodings do not have corresponding
$\mathsf{map}$ functions, and although this property is guaranteed to
hold even when it can be stated for Church encodings, it is simply the
reflection back into syntax of the naturality property of the
functorial action of the interpretation of the type constructor
associated with an ADT's representation as a primitive data type.

%In the former case the ADT $\mathsf{List\,a}$ for each $\mathsf{a}$ is
%modelled as the fixpoint of its underlying functor $\mu F_a$; by
%contrast, in the latter case it is the type constructor
%$\mathsf{List}$ itself that is modelled as $\mu H$, and the ADT
%$\mathsf{List\,a}$ is actually modelled as $(\mu H)\,a$. That is, the
%ADTs $\mathsf{List\,a}$ for the various $\mathsf{a}$s are all modelled
%uniformly. (The extreme uniformity comes from the fact that the
%{\color{blue} type/second} argument $X$ is the same in each recursive
%call to $H$. This requirement will be lifted in
%Section~\ref{sec:gadts} below.)  This is precisely what makes it
%possible to define am $\mathsf{fmap}$ for list types.

%A crucial point is that while the $\mu F_a$ is not required to be a
%functor, and thus is not required to have a {\em functorial action} on
%morphisms --- i.e., to support an $\mathsf{fmap_{F_a}}$ function in
%programming terms --- the functor $\mu H$ modelling $\mathsf{List}$
%must do so. Indeed, since $H$ is a higher-order functor, $\mu H$ is
%itself a functor. Applying the functorial action $\mathsf{fmap_{\mu
%    H}}$ of this functor to a morphism $\mathsf{f}$ yields a natural
%transformation $\mathsf{fmap_{\mu H}\,f}$. This in turn guarantees the
%order-independence of type-independent rearrangement and conversion
%via $\mathsf{fmap_{\mu H}\,f}$ for lists {\color{blue} and trees}.

%{\color{blue} Extension of a container}

\begin{comment}
{\color{blue} Type-independent reasoning about ADTs in functional
  languages is usually justified by first representing ADTs by their
  Church encodings, and then reasoning type-independently about these
  encodings. This is typically justified by constructing a parametric
  model --- i.e, a model in which polymorphic functions preserve
  relations {\em \'a la} Reynolds --- for a suitable fragment of
  System F, demonstrating that an initial algebra exists for the
  positive type constructor corresponding to the functor underlying an
  ADT of interest, and showing that each such initial algebra is
  suitably isomorphic to its corresponding Church encoding. In fact,
  this isomorphism of initial algebras and their Church encodings is
  one of the ``litmus tests'' for the goodness of a parametric model.}

This approach works well for ADTs, which are always fixpoints of {\em
  first-order} functors, and whose Church encodings, which involve
quantification over only type variables, are always expressible in
System F. For example, $\mathtt{List\,A}$ is the fixpoint of the
first-order functor $F\,X = 1 + A \times X$ and has Church encoding
$\forall \alpha. \, \alpha \to (\mathtt{A} \to \alpha \to \alpha) \to
\alpha$. But despite Cardelli's~\cite{car97} claim that ``virtually
any basic type of interest can be encoded within F$_2$'' --- i.e.,
within System F --- non-ADT nested types cannot.  Not even our
prototypical nested type of perfect trees has a Church encoding
expressible in System F!  Indeed, $\mathtt{PTree\,A}$ cannot be
represented as the fixpoint of any {\em first-order} functor. However,
it can be seen as the instance at index $\mathtt{A}$ of the fixpoint
of the {\em higher-order} functor $H\,F\,A\,=\, (A \to F\,A) \to (F
\,(A \times A) \to F\,A) \to F\,A$. It thus has Church encoding
$\forall f.\, (\forall \alpha.\,\alpha \to f\alpha) \to (\forall
\alpha. \,f (\alpha \times \alpha) \to f\alpha) \to \forall \alpha.\,
f\alpha$, which requires quantification at the higher kind $* \to *$
for $f$. A similar situation obtains for any (non-ADT) nested
type. Unfortunately, higher-kinded quantification is not available in
System F, so if we want to reason type-independently about nested
types in a language based on it we have only two options: {\em
  i})\,move to an extension of System F, such as the higher-kinded
calculus F$_\omega$ or a dependent type theory, and reason via their
Church encodings in a known parametric model for that extension, or
{\em ii})\, add nested types to System F as primitives --- i.e., as
primitive type-level fixpoints --- and construct a parametric model
for the result.

Since the type systems of F$_\omega$ and dependent type theories are
designed to extend System F with far more than non-ADT data types, it
seems like serious overkill to pass to their parametric models to
reason about nested types in System F. Indeed, such calculi support
fundamentally new features that add complexity to their models that is
entirely unnecessary for reasoning about nested types. This paper
therefore pursues the second option above.  We first design a
Hindley-Milner-style calculus supporting primitive nested types,
together with primitive types of natural transformations representing
morphisms between them. Our calculus can express all nested types
appearing in the literature, including truly nested types.  At the
term-level, it supports primitive pattern matching, map functions, and
fold combinators for nested types.\footnote{We leave incorporating
  general term-level recursion to future work because, as
  Pitts~\cite{pit00} reminds us, ``it is hard to construct models of
  both impredicative polymorphism and fixpoint recursion''. In fact,
  as the development in this paper shows, constructing a parametric
  model even for our predicative calculus with primitive nested types
  --- and even without term-level fixpoints --- is already rather
  involved.  On the other hand, our calculus is strongly normalizing,
  so it perhaps edges us toward the kind of provably total practical
  programming language proposed in~\cite{wad89}.}  Our main
contribution is the construction of a parametric model for our
calculus. This is both delicate and challenging. To ensure the
existence of semantic fixpoints interpreting nested types, and thus to
establish a suitable Identity Extension Lemma, our type system must
explicitly track functoriality of types, and cocontinuity conditions
on the functors interpreting them must be appropriately threaded
throughout the model construction. Our model validates all standard
consequences of parametricity in the presence of primitive nested
types, including the isomorphism of primitive ADTs and their Church
encodings, and correctness of short cut fusion for nested types. The
relationship between naturality and parametricity has long been of
interest, and our inclusion of a primitive type of natural
transformations allows us to clearly delineate those consequences of
parametricity that follow from naturality, from those, such as short
cut fusion for nested types, that require the full power of
parametricity.
\end{comment}

\subsection{GADTs}\label{sec:gadts}

GADTs and their uses and ubiquity in modern computing. Note in
particular that GADTs are purposely defined not to behave
type-uniformly. We will explore the consequences of this in
Section~\ref{sec:GADT-semantics}.

IMPORT ENRICO

%Seeing even ADTs as fixpoints of higher-order functors opens the door
%to GADTs...

For ADTs, shape and data are independent. Choose the shape first and
then fill it. GADTs are not like ADTs: the shape of the ``container''
depends on the contained data, so that it's not possible to modify the
data while retaining the shape because shape and data are not
independent.  So if we try to add GADTs, then we have to account for
the dependency between shape and data. The relationship between (how)
the data (is constructed) and the shape is left out.  But the data in
a GADT shouldn't disappear just because you map a function over it!
This is odd! So in what sense to GADTs generalize ADTs? In what sense
are they even data types?

The representation choices are similar for GADTs. The difference is
that the two are not isomorphic.

Once we see that we are interested in fixpoints of higher-order
functors, we begin to relax the uniformity. This has naturally led to
consideration of nested types such as... and GADTs such as...

Having interpretations as fixpoints of higher-order functors, nested
types also have corresponding $\mathsf{fmap}$ functions.

Certain ``polynomial'' GADTs correspond to polynomial ADTs. So we
expect the same behaviors.

Polynomial nested types are actually the GADTs that mimic polynomial
ADT behavior.

But pattern matching fails for the extension (functorial completion)
of a GADT.

\section{Semantics of GADTs}\label{sec:GADT-semantics}

Functorial completion containerizes the GADT. Extension of a
container. The extension of a container is an endofunctor generalizing
map.

Eq semantics minimal (least commitment, most natural) for CEs; Lan
semantics minimal for primitives. Minimal way to add to any calculus
supporting them.

An analysis of the Eq A B type in Bob's systems shows that the set
interpretation is empty whenever the A and B have different set
interpretations.  Likewise, its relational interpretation is the empty
relation whenever the A and B have different relational
interpretations (even when the underlying set interpretations are the
same).  The consequences of parametricity are thus immediate: there
can be no term of type Eq A B if A and B are different types.
Moreover, this behavior is fundamentally non-functorial, so that there
is no hope of defining a map for Eq.  In particular, that means we
cannot expect to retrieve non-trivial naturality properties for the Eq
type, despite these are usually considered to be a special case of
parametricity properties.  Indeed, the common understanding that
parametricity generalizes naturality only makes sense for functorial
types.  We can specialize these considerations to the type Eq 1 B,
representing G, our example GADT defined by a unique constructor c ::
G 1.

Since the types encoding GADTs in Bob's system make essential use
of equality types, these considerations extend and possibly generalize
to all GADTs.  Still, the equality is just a shorthand for a type made
by higher-order forall and arrow types, and thus this considerations
are not limited to a specific type, but are intrinsic of Bob's
system.
%We can thus expect that not all types in Bob's system will be
%functorial and that, consequently, the consequences of parametricity
%will not include functoriality but will provide information about,
%e.g., inhabitatedness.  See
%<https://www.cis.upenn.edu/~sweirich/talks/param-gadt.pdf> and
%<https://www.seas.upenn.edu/~sweirich/papers/rtheorems.pdf> for more
%about parametricity and GADTs.

In contrast, $(Lan_1 1) B$, the type representing G in our system, is
functorial (as are all types in our systems) and inhabited for every
instance of B other than 0.  Indeed, we have a term $c : (Lan_1 1) 1$
given by the Lan introduction rule, and thus every term t : T yields a
term in $(Lan_1 1) T$ by use of the map of $(Lan_1 1) B$.  This shows
that, as expected by now, the same GADT really does have different
semantic interpretations in Bob's system and in ours.

\section{Parametricity}

{\em Relational parametricity} encodes a powerful notion of
type-uniformity, or representation independence, for data types in
functional languages. It formalizes the intuition that a polymorphic
program must act uniformly on all of its possible type instantiations
by requiring that every such program preserves all relations between
pairs of types at which it is instantiated. Parametricity was
originally put forth by Reynolds~\cite{rey83} for System F. It was
later popularized as Wadler's ``theorems for free''~\cite{wad89}, so
called because it can deduce properties of programs in such languages
solely from their types, i.e., with no knowledge whatsoever of the
text of the programs involved.  Most of Wadler's free theorems are
consequences of naturality for polymorphic list-processing
functions. However, parametricity can also derive results that go
beyond just naturality, such as establishing {\color{blue} proving}
the equivalence of the Church encodings and the primitive
representations of ADTs and {\color{blue} hereditarily polynomial}
nested types by validating the aforementioned short cut fusion
transformation for them.

One important consequence of representation choice is for parametric
reasoning about GADTs. In particular, we can use parametricity to get
our transform-rearrange property for ADTs. Do it for lists?

We can thus expect that not all types in Bob's system will
be functorial and that, consequently, the consequences of
parametricity will not include functoriality but will provide
information about, e.g., inhabitatedness.  See
<https://www.cis.upenn.edu/~sweirich/talks/param-gadt.pdf> and
<https://www.seas.upenn.edu/~sweirich/papers/rtheorems.pdf> for more
about parametricity and GADTs.

If we had a parametric model for GADTs, it could perhaps validate
short cut and this show that the two reps for them are equivalent.
(This is a sort of converse.)
The fact that GADTs don't validate any such iso has important
consequences for their parametricity properties.

In this paper we show that there is not an iso between the two reps of
even the simplest GADT {\color{blue} that isn't a hereditarily
  polynomial nested type}, and investigate the consequences of this
discovery for parametric reasoning about GADTs.

Non-uniformity renders build, therefore short cut, for GADTs
impossible. Can't even formulate! Maybe there's another way to prove
equivalence, but I don't know what it would be. In any case, we have a
counterexample, so this settles the question.

Importantly, we get the same parametricity properties for ADTs whether
we regard them as fixpoints of higher-order functors --- i.e., as
primitive data types --- or as their Church encodings. This is not the
case for GADTs that are not nested types. (What about truly nested
types?)

We do minimal thing to fix functoriality, then investigate
parametricity properties for both interpretations. We are forced to
choose.

Parametric polymorphism as a way to capture Strachey's notion of
uniform, type-independent behavior of functions. Parametric models as
formalizations of this notion of uniformity, as first given in
Reynolds' original paper.

Bob's system gives a parametric model that models GADTs. But given the
non-uniformity captured by GADTs by purposeful design, that should be
both completely unexpected and alarming! However, upon reflection it's
not really all that surprising once we realize what the parametricity
properties of GADTs in Bob's system look like.

%It turns out that such properties are essentially trivial (because of
%its essential use of Eq, which itself has a trivial parametricity
%property --- IS THIS TRUE??). So, in the end, parametricity is not
%really something I can use to reason non-trivially about GADTs in
%Bob's system. (MAY NEED TO REVISE AFTER INVESTIGATION INTO PAR
%PROPERTY OF EQ.)

Tellingly, Bob gives no parametricity properties for GADTs. If he had,
it would have illustrated the phenomenon we are describing, raised the
question: What good is a par property for a GADT or another type using
the Eq type?

GADTs Force a Representation Choice. It has consequences for how one
reasons about them.

It depends on primitive vs. Church encodings. Primitive forces
functoriality, as in LICS'19.

FOR ADTs and Nesteds, it does not depend. We get the same par property
whether they are seen as primitives or we use their CE.

\section{Conclusion}

{\color{blue} Some of this probably belongs in previous section.}

Moreover, the consequences of parametricity that one has in Bob's
system do not hold in our system, as we can construct terms of type
$(Lan_1 1)$ B for instances of B other than 1.  Thus, either our
system does not admit any parametric model, or, if it does, the
consequences of parametricity are different.  That the consequences of
parametricity depended on the model or the representation of GADTs
would come as a surprise, though!  (IT DOES. CONSIDER INTERPRETING
BASE TYPES AS ORDERED SETS. THIS GIVES DIFFERENT PARAMETRICITY
PROPERTIES THAN INTERPRETING THEM AS DIAGONAL RELATIONS.)

That parametricity properties for GADTs in Bob's system are as we
described before is interesting because, for ADTs and nested types,
which GADTs claim to generalize, we expect naturality to be a special
case of parametricity. However, we don't actually get non-trivial
naturality consequences of parametricity for GADTs in Bob's
system. Why? Because this would require GADTs to be interpreted as
functors, which, in Bob's system, requires the programmer to supply a
map function on the intepretation of every GADT defined. Moreover,
each such map function would have to behave functorially. Bob's system
does indeed deliver non-trivial naturality results for every GADT for
which the user can define a map function that behaves
functorially. The problem is that there may not be many such GADTs (in
fact it will not be possible for proper GADTs, i.e., GADTs that are
not actually nested types).

Perhaps expecting naturality as a special case of parametricity is
wrongheaded, though, since, with its arbitrary forall-types and arrow
types, System F is inherently non-functorial. Using Church encodings,
Bob's system handles arbitrary GADT syntax. However, GADTs are not
guaranteed to behave functorially in his system, and the user is
ultimately tasked with supplying the relevant map functions (and
proving their properties) for any GADTs they want to work with. That
the user must supply the map function for a GADT of interest makes
sense: Church encodings cannot say anything about functorial nature of
data types since they describe only the data (and not the action on
morphisms since they are not required to support map functions).  Our
approach is different. By contrast with Bob's system, we restrict to
GADTs generated by a specific grammar, and this grammar is designed
specifically to guarantee functoriality of all data types definable in
it. We also treat GADTs as primitive data types --- i.e., as fixpoints
of (higher-order) functors --- rather than via their Church encodings
since this more closely aligns with how ADTs are treated in
implemented programming languages.

Wrt Bob's system we can ask: what good is parametricity for GADTs in a
system that derives only trivial naturality consequences of
parametricity for them? It may be possible to get non-naturality
consequences of parametricity (such as short cut fusion or
VYTINIOTIS'?) for GADTs in Bob's system (IS IT?), but since many of
the most useful consequences of parametricity for ADTs and nested
types are actually natuality results (IS THIS TRUE? MAYBE NOT FOR
GADTs), we'd be missing out on a very large class of expected and
useful free theorems. Indeed, many of Wadler's free theorems are
precisely naturality results for functions over data types (mainly
lists, there). If GADTs really do generalize ADTs (and nested types)
then we would expect to be able to derive analogues for GADTs of the
parametricity results --- including naturality results --- that we can
derive for ADTs. Yet, by design we cannot. So are GADTs actually
misnamed? In what sense are they data types, and in what sense do they
generalize ADTs? Perhaps what are now known as nested types should
really be called GADTs since *they* are actually the (level-2)
higher-kinded generalizations of ADTs.

In our system we can show that we can recover the expected naturality
consequences of parametricity for (at least a large class of) GADTs
(that can be characterized by a grammar for which every type is
interpreted as a functor, including those characterized as least
fixpoints, i.e., as carriers of initial algebras of functors, i.e., as
primitive data types). The way our system achieves this is by
interpreting GADTs as their functorial completions. So in our system
the interpretations of GADTs are guaranteed to have fmap functions ---
i.e., to be functorial --- and to satisfy the expected naturality
properties. And our free theorems really are free for GADTs that fit
our syntax. But now the bump is elsewhere under the rug: First,
semantically we don't really have the same GADT anymore (functoriality
adds extra data to the data type), and second, our counterexample
shows that we don't get non-naturality consequences of parmetricity
even for the data types defined by our grammar.  So this isn't really
what we wanted either.

The upshot is that we can consider GADTs as data types (i.e., we can
compute the semantics of GADTs) in two ways: just as determined by
their syntax, as in Bob's system, or as the functorial completion of
their syntax, obtained by supplying a suitable map primitive, as in
ours. But, either way, GADTs fail to be well-behaved wrt
parametricity. In the first case, GADTs don't have fmaps so their
parametricity properties can't include naturality properties. In the
second case, we still can't get non-naturality consequences of
parametricity. That is, we can't get BOTH the naturality consequences
of parametricity AND the non-naturality consequences of parametricity
at the same time.

One can read our counterexample as showing that Bob's parametricity
results cannot be extended to one that gives a functorial semantics to
GADTs. So the user is forced to make a choice between naturality
consequences of parametricity and other consequences of
parametricity. If GADTs really did generalize ADTs, and really were
data types in a normal sense, then the user would not have to.

\bibliography{gadt-story-references}

\end{document}

Daniel:

I have tried reconstructing my example using our simpler GADT with a single
constructor c :: G 1. I don't think it is possible to modify my example for a
GADT with only a single constructor. In particular, I don't see a way to
exploit the map function unless we have constructors for G X and G Y where X
!= Y.
If we assume a map function exists, then given f :: 1 -> A we can produce
(map f c) :: G A. But I don't know what to do from there.

I was able to simplify the types in the example slightly though. The example
I gave before can be adapted to use either

data G3 b where
    G3unit :: G3 ()
    G3id    :: b -> G3 (Either b ())

or

data G5 b where
    G5unit   :: G5 ()
    G5void   :: Void -> G5 Bool

as the problematic GADT.

Enrico:

Just to make sure I understand it correctly: when we define the function

unGbool :: G c Bool -> c
unGbool (Gbool x) = x

we are doing so in Haskell, right?
In our system, it would not be possible to define unGbool (because we could
not define it only on the constructor Gbool of G). For this reason, this
example provides a contradiction in Haskell (+functoriality), but not in our
system. Is that correct?

Daniel:

The example is Haskell and relies on Haskell's rules for
pattern matching.

Enrico:

By the way, as a general remark not directly related to the introduction you
wrote, I think that the set semantics of our type G is

[[G \alpha]]^set [\alpha := A] = A

Indeed, (Lan_1 1) A = A for every set A. Intuitively this makes sense,
because for every a : A there is a fuction

k_a : 1 —> A

and thus an element

map k_a \ast : G A

In particular, [[G 0]]^set = \emptyset.

Still, it is NOT true that its relational interpretation is [[G \alpha]]^rel
[\alpha := R] = R (and indeed the relational interpretation is not fibered
over the set interpretation)!

### 2021-02-13 20-gadts/GADT-story.tex
work on gadt-story

@@
 \[\begin{array}{l}
-\mathsf{data\,G : * \to *}\\
+\mathsf{data\,G : * \to *\,where}\\
 \mathsf{\;\;\;\;\;\;\;\;c : G\,1}
@@
 natural numbers, and the positions in a structure of this type are
-given by specific natural numbers. Since the argument to every
+given by specific natural numbers. Since the type argument to every
 occurrence of the type constructor $\mathsf{List}$ in the right-hand
-side of above definition is the same as on its left-hand side, the
-type $\mathsf{List\,a}$ enforces the invariant that all of the data in
-a structure of this type have the same type $\mathsf{a}$. In a similar
-way, the tree type
+side of above definition is the same the type instance being defined
+on its left-hand side, the type $\mathsf{List\,a}$ enforces the
+invariant that all of the data in a structure of this type have the
+same type $\mathsf{a}$. In a similar way, the tree type
 \[\mathsf{Tree\, a} = \mathsf{Leaf\,a}
@@
 {\em functors} on the category interpreting types~\cite{tfca}, rather
-than mere functions between the sets interpreting them. Said
+than of mere functions between the sets interpreting them. Said
 differently, functions between sets interpreting types must have
@@
 
-\subsection{GADTs}\label{sec:gadts}
+\subsection{GADT Syntax}\label{sec:gadts}
 
-GADTs and their uses and ubiquity in modern computing. Note in
-particular that GADTs are purposely defined not to behave
-type-uniformly. We will explore the consequences of this in
-Section~\ref{sec:GADT-semantics}.
+{\color{blue} polynomial} GADTs relax the restriction on the type
+instances appearing in a data type definition. The special form of
+GADTs known as {\em nested types} allow the data constructors of a
+GADT to take as arguments data whose types involve type instances of
+the GADT other than the one being defined. For example, the 
+definition 
+\[\mathsf{PTree \, a = PLeaf \,a \,|\, PNode \,(PTree\,(a \times
+  a))}\] of the nested type $\mathsf{PTree\,a}$ of perfect trees
+introduces the data constructor(s $\mathsf{PLeaf :: a \to PTree\,a}$
+and) $\mathsf{PNode :: PTree \,(a\times a) \to PTree\,a}$. It enforces
+not only the invariant that all of the data in a structure of type
+$\mathsf{PTree\,a}$ is of the same type $\mathsf{a}$, but also the
+invariant that all perfect trees have lengths that are powers of
+2. More general GADTs allow their constructors both to take as
+arguments {\em and return as results} data whose types involve type
+instances of the GADT other than the one being defined. An example is
+the following GADT, given in Haskell syntax:
+\[\begin{array}{l}
+\mathsf{data\, Seq : * \to *\,where}\\
+\mathsf{\;\;\;\;\;\;\;\;Const :\, a \to Seq\,a}\\
+\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, :\, Seq \,a \to Seq\,b \to
+  Seq\,(a \times b)}\\ 
+\end{array}\]
+Since the return type of the data constructor $\mathsf{Pair}$ is not
+simply a variable, $\mathsf{Seq}$ is not a nested type.
 
-IMPORT ENRICO
+By contrast with the ADT $\mathsf{List\, a}$, where the type parameter
+$\mathsf{a}$ is integral to the type being defined, the type parameter
+$\mathsf{a}$ in $\mathsf{PTree\,a}$ and $\mathsf{Seq\,a}$ appears as a
+``dummy'' parameter used only to give the kind \verb|* -> *| of the
+type constructors $\mathsf{PTree}$ and $\mathsf{Seq}$. This is
+explicitly captured by the above GADT syntax. A GADT --- even a nested
+type --- thus does not define a {\em family of inductive types}, one
+for each type argument, like an ADT does, but instead defines an
+entire family of types that must be constructed simultaneously. That
+is, a GADT defines an {\em inductive family of types}.
 
-%Seeing even ADTs as fixpoints of higher-order functors opens the door
-%to GADTs...
+\subsubsection{Church Encodings of GADTs}
 
-For ADTs, shape and data are independent. Choose the shape first and
-then fill it. GADTs are not like ADTs: the shape of the ``container''
-depends on the contained data, so that it's not possible to modify the
-data while retaining the shape because shape and data are not
-independent.  So if we try to add GADTs, then we have to account for
-the dependency between shape and data. The relationship between (how)
-the data (is constructed) and the shape is left out.  But the data in
-a GADT shouldn't disappear just because you map a function over it!
-This is odd! So in what sense to GADTs generalize ADTs? In what sense
-are they even data types?
+The syntax of GADTs is consonant with programmers' frequent use of
+them in situations where type uniformity is specifically to be
+eschewed.  {\color{blue} Typical applications of GADTs include generic
+  programming, modeling programming languages via higher-order
+  abstract syntax, maintaining invariants in data structures, and
+  expressing constraints in embedded domain-specific languages. GADTs
+  have also been used to implement tagless
+  interpreters~\cite{pl04,pr06,pvww06} by trading the definition of a
+  universal value domain for a direct specification of the property of
+  being a value. Other applications are described in,
+  e.g.,~\cite{js15,rou06,MORE,UPDATED}. OTHER SITUATIONS NEEDING
+  TYPE-BASED DISPATCH.}
 
+GADTs are useful in such situations precisely because generalizing the
+type arguments in the instances of a GADT appearing as the return
+types of its data constructors entails a strong connection between a
+GADTs shape and the data it contains.  With ADTs, we first choose the
+shape of the container and then fill that container with data of
+whatever type we like. Critically, the choice of shape is independent
+of the data to be stored. With GADTs, however, the shape of the
+``container'' may actually {\em depend} on (the type of) the data to
+be contained. On other words, the shape and the data are not
+independent! As a result, it may not always be possible to modify a
+GADT's data while retaining its shape because changing (the type of)
+the data may actually change the shape.
+
+To determine the possible shapes of a GADT's container we have to
+pattern-match on the type of the data to be contained. For this, our
+GADT calculus must support an equality type $\mathsf{Eq}$ that is
+inhabited when its two type arguments are the same and is uninhabited
+otherwise, i.e., the syntactic reflection of semantic equality. Such a
+type can either be taken as primitive or
+%, in a higher-kinded calculus such as $F\omega$,
+can be represented by its own Church encoding
+\[\mathsf{Eq\,t_1\,t_2} = \mathsf{\forall f.\, (\forall a.\,
+  f\,a\,a) \to f\,t_1\,t_2}\] 
+Using $\mathsf{Eq}$ the GADT $\mathsf{Seq}$, e.g., can be written as 
+{\color{red} START HERE}
+\[ {\color{blue} WHAT GOES HERE??} \]
+
+Note that Seq has no syntactic map (because Pair doesn't), i.e., is
+not an instance of the Functor class, and is not required to in this
+representation.
+
+\subsubsection{GADTs as Primitive Data Types}
+
+Moreover, because its type argument varies in the instances of
+$\mathsf{Seq}$ appearing in the types of its data constructors,
+$\mathsf{Seq}$ cannot be characterized by any first-order fixpoint
+equation. It can, be viewed as a solution to the higher-order fixpoint
+equation
+\[\mathsf{H\,f\,a = a \,+\,Lan_{\lambda a\,b.\,a \times b}\,
+  \lambda a\,b.\,f\,a \times f\,b}\] where $\mathsf{Lan_k\, g}$ is the
+syntactic reflection of the {\em left Kan extension} $\Lan_K \,G$ of
+the functor $G$ interpreting $\mathsf{g}$ along the functor $K$
+interpreting $\mathsf{k}$~\cite{mac71}.
+
+Other GADTs
+are similarly characterizable by higher-order functors.  The
+difference between ADTs and GADTs can thus be seen as a difference in
+{\em orders} of their underlying functors. {\color{blue} The role of
+  higher-order functors here is essential.}
+
+{\color{blue} GADTs make essential use of higher-order functors.}
+
+Need syntax functors (Lan)
 The representation choices are similar for GADTs. The difference is
 that the two are not isomorphic.
-
-Once we see that we are interested in fixpoints of higher-order
-functors, we begin to relax the uniformity. This has naturally led to
-consideration of nested types such as... and GADTs such as...
 
@@
 \section{Semantics of GADTs}\label{sec:GADT-semantics}
+
+But the data in a GADT shouldn't disappear just because you map a
+function over it!  This is odd! So in what sense to GADTs generalize
+ADTs? In what sense are they even data types?
+
+The semantics of GADTs is not so clear.
 

### 2021-02-14 20-gadts/GADT-story.tex
work on gadt-story Sec 2.2

@@
 \[\begin{array}{l}
-\mathsf{data\, Seq : * \to *\,where}\\
-\mathsf{\;\;\;\;\;\;\;\;Const :\, a \to Seq\,a}\\
-\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, :\, Seq \,a \to Seq\,b \to
+\mathsf{data\, Seq :: * \to *\,where}\\
+\mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Seq\,a}\\
+\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, ::\, Seq \,a \to Seq\,b \to
   Seq\,(a \times b)}\\ 
@@
 otherwise, i.e., the syntactic reflection of semantic equality. Such a
-type can either be taken as primitive or
+type can either be taken as primitive, as in
+\[\begin{array}{l}
+\mathsf{data\, Eq :: * \to * \to *\,where}\\
+\mathsf{\;\;\;\;\;\;\;\;Refl ::\, Eq\,a\,a}\\
+\end{array}\]
+\noindent
+or
 %, in a higher-kinded calculus such as $F\omega$,
-can be represented by its own Church encoding
+represented by its own Church encoding
 \[\mathsf{Eq\,t_1\,t_2} = \mathsf{\forall f.\, (\forall a.\,
   f\,a\,a) \to f\,t_1\,t_2}\] 
-Using $\mathsf{Eq}$ the GADT $\mathsf{Seq}$, e.g., can be written as 
+Using $\mathsf{Eq}$, the GADT $\mathsf{Seq}$, e.g., can be written as 
 {\color{red} START HERE}
@@
 
-Note that Seq has no syntactic map (because Pair doesn't), i.e., is
-not an instance of the Functor class, and is not required to in this
-representation.
+Note that neither Eq nor Seq has a syntactic map (because one for Eq
+leads to a contradition as in Daniel's exampe below, and because one
+for Seq can't exist because of Pair), i.e., is not an instance of the
+Functor class, and is not required to in this representation.
+
+data Equal a b where
+  Refl :: Equal a a
+
+eqElim :: Equal a b -> a -> b
+eqElim Refl x = x
+
+emptyFromMapEq :: (forall a b c. (a -> b) -> Equal a c -> Equal b c) -> Void
+emptyFromMapEq fmap = eqElim (fmap absurd Refl) ()
 

### 2021-02-14 20-gadts/GADT-story.tex
work on gadt-story Sec 2.2

@@
 higher-order functor $H$ whose action on objects is given by $H\,F\,X
-= 1 + X \times H\,F\,X$ and whose action on morphisms is givne by
-$H\,F\,f = \id_1 + f \times H\,F\,f$. (A {\em higher-order functor} in
+= 1 + X \times F\,X$ and whose action on morphisms is given by
+$H\,F\,f = \id_1 + f \times F\,f$. (A {\em higher-order functor} in
 this context is a functor not directly on the category interpreting
@@
 inhabited when its two type arguments are the same and is uninhabited
-otherwise, i.e., the syntactic reflection of semantic equality. Such a
-type can either be taken as primitive, as in
+otherwise, i.e., is the syntactic reflection of semantic equality
+$\Eq$. Such a type can either be taken as primitive, as in
 \[\begin{array}{l}
@@
   f\,a\,a) \to f\,t_1\,t_2}\]
-Using $\mathsf{Eq}$, the GADT $\mathsf{Seq}$, e.g., can be written as 
-{\color{red} START HERE}
-\[ {\color{blue} WHAT GOES HERE??} \]
-
-Note that neither Eq nor Seq has a syntactic map (because one for Eq
-leads to a contradition as in Daniel's exampe below, and because one
-for Seq can't exist because of Pair), i.e., is not an instance of the
-Functor class, and is not required to in this representation.
-
-data Equal a b where
-  Refl :: Equal a a
-
-eqElim :: Equal a b -> a -> b
-eqElim Refl x = x
-
-emptyFromMapEq :: (forall a b c. (a -> b) -> Equal a c -> Equal b c) -> Void
-emptyFromMapEq fmap = eqElim (fmap absurd Refl) ()
+Using $\mathsf{Eq}$, the GADT $\mathsf{Seq}$ can be written as
+%\[\mathsf{Seq\,a} = \mathsf{a + Eq\,a\,(b \times c) \times Seq\,b
+%  \times Seq\,c} \] 
+\[\begin{array}{l}
+\mathsf{data\, Seq :: * \to *\,where}\\
+\mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Seq\,a}\\
+\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, ::\, Eq\,a\,(b \times c) \to
+  Seq \,b \to Seq\,c \to Seq\,a}\\ 
+\end{array}\]
+
+This rewriting of $\mathsf{Seq}$'s definition shows that it can be
+seen as a fixpoint $\mu H$ of the higher-order function $H$ defined on
+objects by $H\,F\,a = a + Eq\,a\,(b \times c) \times F\,b \times
+F\,c$. We note, however, that $Eq\,a$ cannot be made a functor, or,
+equivalently, that $\mathsf{Eq\,a}$ cannot be an instance of Haskell's
+$\mathsf{Functor}$ class. Indeed, if $\mathsf{Eq\,a}$ supported a
+function $\mathsf{map_{Eq\,a} :: (b \to c) \to Eq\,a\,b \to Eq\,a\,c}$
+then defining
+\[\begin{array}{l}
+\mathsf{eqElim :: Eq\, a\, b \to b \to a}\,{\color{red} wasEq\,a\,b\to b\to a}\\
+\mathsf{eqElim\, Refl\, x = x} 
+\end{array}\]
+would allow us to construct the following element of the type
+$\mathsf{Void}$:
+\[\begin{array}{l}
+\mathsf{emptyFromMapEqa :: (\forall b\, c.\, (b \to c) \to Eq\, a\,
+  b \to Eq\, a \,c) \to Void}\\ 
+\mathsf{emptyFromMapEqa \,map_{Eq\,a} = eqElim\, (map_{\Eq\,a}\,
+  absurd\, Refl)\, ()}  
+\end{array}\]
+Of course this is impossible.  It follows immediately that
+$\mathsf{Seq}$ cannot support a map function either. This is already
+evident from the Haskell definition of $\mathsf{Seq}$, though. The
+clause of $\mathsf{map}$ for the $\mathsf{Pair}$ constructor should
+feed $\mathsf{map}$ a function $\mathsf{f :: (b \times c) \to e}$ and
+a term of the form $\mathsf{Pair \,t_1\,t_2}$ for $\mathsf{x :: b}$
+and $\mathsf{y :: c}$, and produce a term $\mathsf{Pair\,t_1'\,t_2'}$
+for some appropriately typed terms $\mathsf{t_1'}$ and
+$\mathsf{t_2'}$. But it is not clear how to achieve this since
+$\mathsf{e}$ need not necessarily be a product type. And even if
+$\mathsf{e}$ {\em were} known to be of the form $\mathsf{w \times z}$,
+then we still wouldn't necessarily have a way to produce data of type
+$\mathsf{w \times z}$ from only $\mathsf{f :: b \times c \to w \times
+  z}$ and $\mathsf{x}$ and $\mathsf{y}$ unless we knew, e.g., that
+$\mathsf{f}$ was a pair of functions.
 
@@
 
-Moreover, because its type argument varies in the instances of
-$\mathsf{Seq}$ appearing in the types of its data constructors,
-$\mathsf{Seq}$ cannot be characterized by any first-order fixpoint
-equation. It can, be viewed as a solution to the higher-order fixpoint
-equation
+The primitive representations of {\color{blue} polynomial} GADTs
+capture in syntax the fact that GADTs can be considered as fixpoints
+of higher-order functors. The use of {\em higher-order functions} is
+essential, since the type arguments to the GADT being defined are not
+uniform across all of its instances in the types of its data
+constructors; the use of higher-order funct{\em ors} is essential, as
+for ADTs, to the very existence of the fixpoints being
+computed~\cite{tfca}.
+
+Consider, for example, the GADT $\mathsf{Seq}$ from above.  Because
+its type argument varies in the instances of $\mathsf{Seq}$ appearing
+in the types of its data constructor $\mathsf{Pair}$, $\mathsf{Seq}$
+cannot be characterized by any first-order fixpoint equation. It can,
+however, be viewed as a solution to the higher-order fixpoint equation
+{\color{blue} syntax or semantics?}
 \[\mathsf{H\,f\,a = a \,+\,Lan_{\lambda a\,b.\,a \times b}\,
@@
 
-Other GADTs
-are similarly characterizable by higher-order functors.  The
-difference between ADTs and GADTs can thus be seen as a difference in
-{\em orders} of their underlying functors. {\color{blue} The role of
-  higher-order functors here is essential.}
-
-{\color{blue} GADTs make essential use of higher-order functors.}
+Other GADTs are similarly characterizable by higher-order functors.
+The difference between ADTs and GADTs can thus be seen as a difference
+in {\em orders} of their underlying functors. {\color{blue} GADTs
+  make essential use of higher-order functors.}
 
@@
 
-Having interpretations as fixpoints of higher-order functors, nested
-types also have corresponding $\mathsf{fmap}$ functions.
-
 Certain ``polynomial'' GADTs correspond to polynomial ADTs. So we
@@
 The semantics of GADTs is not so clear.
+
+Functorial comnpletion doesn't give nested types
 

### 2021-02-15 20-gadts/GADT-story.tex
work on gadt-story Sec 2.2

@@
 to be a solution to the fixpoint equation
-\[\mathsf{Tree\,a} = \mathsf{a + (Tree\,a \times a \times Tree\,a)}\] 
+\[\mathsf{Tree\,a} = \mathsf{a + Tree\,a \times a \times Tree\,a}\] 
 determined by its Haskell data type declaration.
@@
 of the data to be stored. With GADTs, however, the shape of the
-``container'' may actually {\em depend} on (the type of) the data to
-be contained. On other words, the shape and the data are not
-independent! As a result, it may not always be possible to modify a
-GADT's data while retaining its shape because changing (the type of)
-the data may actually change the shape.
+container may actually {\em depend} on (the type of) the data to be
+contained. On other words, the shape and the data are not independent!
+For example, while $\mathsf{Const}$ can create data of any shape
+$\mathsf{Seq\,a}$, $\mathsf{Pair}$ can produce data of shape
+$\mathsf{Seq\,a}$ where $\mathsf{a}$ is a pair type. As a result, it
+may not always be possible to modify a GADT's data while retaining its
+shape because changing (the type of) the data may actually change the
+shape.
 
@@
 otherwise, i.e., is the syntactic reflection of semantic equality
-$\Eq$. Such a type can either be taken as primitive, as in
+function $Eq$. Such a type can either be taken as primitive, as in
 \[\begin{array}{l}
@@
 or
-%, in a higher-kinded calculus such as $F\omega$,
 represented by its own Church encoding
 \[\mathsf{Eq\,t_1\,t_2} = \mathsf{\forall f.\, (\forall a.\,
-  f\,a\,a) \to f\,t_1\,t_2}\]
-Using $\mathsf{Eq}$, the GADT $\mathsf{Seq}$ can be written as
-%\[\mathsf{Seq\,a} = \mathsf{a + Eq\,a\,(b \times c) \times Seq\,b
-%  \times Seq\,c} \] 
+  f\,a\,a) \to f\,t_1\,t_2}\] Using $\mathsf{Eq}$, the GADT
+$\mathsf{Seq}$ can then be represented as its Church encoding
+         {\color{red}
+\[\mathsf{Seq\,a} = \mathsf{\forall f.\, (\forall a.\,a \to f\, a)
+  \to (\forall a\,b\,c.\,Eq\,a\,(b\times c) \to f\,b \to f\,c \to
+  f\,a) \to f\,a}\]}
+in a higher-kinded calculus such as $F_\omega$.
+
+Importantly, $Eq\,a$ cannot be made into a functor, or, equivalently,
+$\mathsf{Eq\,a}$ cannot be made an instance of Haskell's
+$\mathsf{Functor}$ class: if $\mathsf{Eq\,a}$ supported a function
+$\mathsf{map_{Eq\,a} :: (b \to c) \to Eq\,a\,b \to Eq\,a\,c}$ then
+defining
 \[\begin{array}{l}
-\mathsf{data\, Seq :: * \to *\,where}\\
-\mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Seq\,a}\\
-\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, ::\, Eq\,a\,(b \times c) \to
-  Seq \,b \to Seq\,c \to Seq\,a}\\ 
-\end{array}\]
-
-This rewriting of $\mathsf{Seq}$'s definition shows that it can be
-seen as a fixpoint $\mu H$ of the higher-order function $H$ defined on
-objects by $H\,F\,a = a + Eq\,a\,(b \times c) \times F\,b \times
-F\,c$. We note, however, that $Eq\,a$ cannot be made a functor, or,
-equivalently, that $\mathsf{Eq\,a}$ cannot be an instance of Haskell's
-$\mathsf{Functor}$ class. Indeed, if $\mathsf{Eq\,a}$ supported a
-function $\mathsf{map_{Eq\,a} :: (b \to c) \to Eq\,a\,b \to Eq\,a\,c}$
-then defining
-\[\begin{array}{l}
-\mathsf{eqElim :: Eq\, a\, b \to b \to a}\,{\color{red} wasEq\,a\,b\to b\to a}\\
+\mathsf{eqElim :: Eq\, a\, b \to b \to a}\,{\color{red} was Eq\,a\,b\to a\to b}\\
 \mathsf{eqElim\, Refl\, x = x} 
 \end{array}\]
-would allow us to construct the following element of the type
-$\mathsf{Void}$:
+would allow us to construct the element 
 \[\begin{array}{l}
@@
 \end{array}\]
-Of course this is impossible.  It follows immediately that
-$\mathsf{Seq}$ cannot support a map function either. This is already
-evident from the Haskell definition of $\mathsf{Seq}$, though. The
-clause of $\mathsf{map}$ for the $\mathsf{Pair}$ constructor should
-feed $\mathsf{map}$ a function $\mathsf{f :: (b \times c) \to e}$ and
-a term of the form $\mathsf{Pair \,t_1\,t_2}$ for $\mathsf{x :: b}$
-and $\mathsf{y :: c}$, and produce a term $\mathsf{Pair\,t_1'\,t_2'}$
-for some appropriately typed terms $\mathsf{t_1'}$ and
+of type $\mathsf{Void}$, which is impossible. It follows immediately
+from this contradiction that $\mathsf{Seq}$ cannot support a map
+function either. This is already intuitively evident from the Haskell
+definition of $\mathsf{Seq}$, though. Indeed, the clause of
+$\mathsf{map}$ for the $\mathsf{Pair}$ constructor should feed
+$\mathsf{map}$ a function $\mathsf{f :: (b \times c) \to e}$ and a
+term of the form $\mathsf{Pair \,t_1\,t_2}$ for $\mathsf{x :: b}$ and
+$\mathsf{y :: c}$, and produce a term $\mathsf{Pair\,t_1'\,t_2'}$ for
+some appropriately typed terms $\mathsf{t_1'}$ and
 $\mathsf{t_2'}$. But it is not clear how to achieve this since
@@
 capture in syntax the fact that GADTs can be considered as fixpoints
-of higher-order functors. The use of {\em higher-order functions} is
-essential, since the type arguments to the GADT being defined are not
-uniform across all of its instances in the types of its data
-constructors; the use of higher-order funct{\em ors} is essential, as
-for ADTs, to the very existence of the fixpoints being
-computed~\cite{tfca}.
-
-Consider, for example, the GADT $\mathsf{Seq}$ from above.  Because
-its type argument varies in the instances of $\mathsf{Seq}$ appearing
-in the types of its data constructor $\mathsf{Pair}$, $\mathsf{Seq}$
-cannot be characterized by any first-order fixpoint equation. It can,
-however, be viewed as a solution to the higher-order fixpoint equation
-{\color{blue} syntax or semantics?}
-\[\mathsf{H\,f\,a = a \,+\,Lan_{\lambda a\,b.\,a \times b}\,
-  \lambda a\,b.\,f\,a \times f\,b}\] where $\mathsf{Lan_k\, g}$ is the
-syntactic reflection of the {\em left Kan extension} $\Lan_K \,G$ of
-the functor $G$ interpreting $\mathsf{g}$ along the functor $K$
-interpreting $\mathsf{k}$~\cite{mac71}.
-
-Other GADTs are similarly characterizable by higher-order functors.
-The difference between ADTs and GADTs can thus be seen as a difference
-in {\em orders} of their underlying functors. {\color{blue} GADTs
-  make essential use of higher-order functors.}
+of higher-order functors. Syntactically, the use of higher-order
+funct{\em ions} is essential, since the type arguments to the GADT
+being defined are not uniform across all of its instances in the types
+of its data constructors. Semantically, the use of higher-order
+funct{\em ors} is essential, as for ADTs, to the very existence of the
+fixpoints being computed~\cite{tfca}.
+%We will consider GADTs'
+%semantics in more detail in Section~\ref{sec:GADT-semantics}.
+
+Consider, again, the GADT $\mathsf{Seq}$ from the start of
+Section~\ref{sec:gadts}. Because its type argument varies in the
+instances of $\mathsf{Seq}$ appearing in the types of its data
+constructor $\mathsf{Pair}$, $\mathsf{Seq}$ cannot be characterized by
+any first-order fixpoint equation. It can, however, be viewed as a
+solution to the higher-order fixpoint equation
+\[H\,f\,a = a \,+\,Lan_{\lambda a b.\,a \times b}\, \lambda
+a b.\,f\,a \times f\,b\] 
+where $\Lan_K \,G$ is the 
+%$Lan_k\, g$ is the syntactic reflection of the
+{\em left Kan extension}~\cite{mac71}
+%$\Lan_K \,G$
+of the functor $G$
+%interpreting $\mathsf{g}$
+along the functor $K$.
+%interpreting $\mathsf{k}$~\cite{mac71}.
+The left Kan extension $\Lan_K \,G$ is the smallest functor
+approximating/extending $K$ and coinciding with $G$ {\color{blue}
+  better explanation}.
+
+LKE gives the functorial completion of the GADT.
+
+%written as
+%%\[\mathsf{Seq\,a} = \mathsf{a + Eq\,a\,(b \times c) \times Seq\,b
+%%  \times Seq\,c} \] 
+%\[\begin{array}{l}
+%\mathsf{data\, Seq :: * \to *\,where}\\
+%\mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Seq\,a}\\
+%\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, ::\, Eq\,a\,(b \times c) \to
+%  Seq \,b \to Seq\,c \to Seq\,a}\\ 
+%\end{array}\]
+%
+%The above rewriting of $\mathsf{Seq}$'s definition shows that it can
+%be seen as a fixpoint $\mu H$ of the higher-order function $H$ defined
+%on objects by $H\,F\,a = a + Eq\,a\,(b \times c) \times F\,b \times
+%F\,c$.
+
+{\color{blue} Need syntax that reflects semantic Lan.}
+
+%Other GADTs are similarly characterizable by higher-order functors.
+%{\color{blue} GADTs make essential use of higher-order functors.}
 
@@
 
-Certain ``polynomial'' GADTs correspond to polynomial ADTs. So we
-expect the same behaviors.
-
-Polynomial nested types are actually the GADTs that mimic polynomial
-ADT behavior.
+Certain polynomial GADTs correspond to polynomial ADTs. So we expect
+the same behaviors.  Polynomial nested types are actually the GADTs
+that mimic polynomial ADT behavior.
 
@@
 \section{Semantics of GADTs}\label{sec:GADT-semantics}
+
+{\color{blue} OMIT THIS SECTION?}
 
@@
 
-Functorial comnpletion doesn't give nested types
+Functorial completion doesn't give nested types
 
@@
 supporting them.
+
+Bob requires map functions for fixpoint but then doesn't use them.
 

### 2021-02-15 20-gadts/GADT-story.tex
work on gadt-story Sec 2.2

@@
 pattern-match on the type of the data to be contained. For this, our
-GADT calculus must support an equality type $\mathsf{Eq}$ that is
+GADT calculus must support an equality type $\mathsf{Equal}$ that is
 inhabited when its two type arguments are the same and is uninhabited
 otherwise, i.e., is the syntactic reflection of semantic equality
-function $Eq$. Such a type can either be taken as primitive, as in
+function $\mathit{Equal}$. Such a type can either be taken as primitive, as in
 \[\begin{array}{l}
-\mathsf{data\, Eq :: * \to * \to *\,where}\\
-\mathsf{\;\;\;\;\;\;\;\;Refl ::\, Eq\,a\,a}\\
+\mathsf{data\, Equal :: * \to * \to *\,where}\\
+\mathsf{\;\;\;\;\;\;\;\;Refl ::\, Equal\,a\,a}\\
 \end{array}\]
@@
 represented by its own Church encoding
-\[\mathsf{Eq\,t_1\,t_2} = \mathsf{\forall f.\, (\forall a.\,
-  f\,a\,a) \to f\,t_1\,t_2}\] Using $\mathsf{Eq}$, the GADT
+\[\mathsf{Equal\,t_1\,t_2} = \mathsf{\forall f.\, (\forall a.\,
+  f\,a\,a) \to f\,t_1\,t_2}\] Using $\mathsf{Equal}$, the GADT
 $\mathsf{Seq}$ can then be represented as its Church encoding
-         {\color{red}
 \[\mathsf{Seq\,a} = \mathsf{\forall f.\, (\forall a.\,a \to f\, a)
-  \to (\forall a\,b\,c.\,Eq\,a\,(b\times c) \to f\,b \to f\,c \to
-  f\,a) \to f\,a}\]}
+  \to (\forall a\,b\,c.\,Equal\,a\,(b\times c) \to f\,b \to f\,c \to
+  f\,a) \to f\,a}\]
 in a higher-kinded calculus such as $F_\omega$.
 
-Importantly, $Eq\,a$ cannot be made into a functor, or, equivalently,
-$\mathsf{Eq\,a}$ cannot be made an instance of Haskell's
-$\mathsf{Functor}$ class: if $\mathsf{Eq\,a}$ supported a function
-$\mathsf{map_{Eq\,a} :: (b \to c) \to Eq\,a\,b \to Eq\,a\,c}$ then
-defining
+Importantly, $\mathit{Equal}\,a$ cannot be made into a functor, or,
+equivalently, $\mathsf{Equal\,a}$ cannot be made an instance of
+Haskell's $\mathsf{Functor}$ class: if $\mathsf{Equal\,a}$ supported a
+function
+\[\mathsf{map_{Equal\,a} :: (b \to c) \to Equal\,a\,b \to
+  Equal\,a\,c}\]
+then defining
 \[\begin{array}{l}
-\mathsf{eqElim :: Eq\, a\, b \to b \to a}\,{\color{red} was Eq\,a\,b\to a\to b}\\
+\mathsf{eqElim :: Equal\, a\, b \to b \to a}\\
 \mathsf{eqElim\, Refl\, x = x} 
@@
 \[\begin{array}{l}
-\mathsf{emptyFromMapEqa :: (\forall b\, c.\, (b \to c) \to Eq\, a\,
-  b \to Eq\, a \,c) \to Void}\\ 
-\mathsf{emptyFromMapEqa \,map_{Eq\,a} = eqElim\, (map_{\Eq\,a}\,
+\mathsf{emptyFromMapEquala :: (\forall b\, c.\, (b \to c) \to Equal\, a\,
+  b \to Equal\, a \,c) \to Void}\\ 
+\mathsf{emptyFromMapEquala \,map_{Equal\,a} = eqElim\, (map_{Equal\,a}\,
   absurd\, Refl)\, ()}  
@@
 fixpoints being computed~\cite{tfca}.
-%We will consider GADTs'
-%semantics in more detail in Section~\ref{sec:GADT-semantics}.
+%We will consider GADTs' semantics in more detail in
+%Section~\ref{sec:GADT-semantics}.
 
@@
 \[H\,f\,a = a \,+\,Lan_{\lambda a b.\,a \times b}\, \lambda
-a b.\,f\,a \times f\,b\] 
-where $\Lan_K \,G$ is the 
-%$Lan_k\, g$ is the syntactic reflection of the
-{\em left Kan extension}~\cite{mac71}
-%$\Lan_K \,G$
-of the functor $G$
-%interpreting $\mathsf{g}$
-along the functor $K$.
-%interpreting $\mathsf{k}$~\cite{mac71}.
-The left Kan extension $\Lan_K \,G$ is the smallest functor
-approximating/extending $K$ and coinciding with $G$ {\color{blue}
-  better explanation}.
-
-LKE gives the functorial completion of the GADT.
-
-%written as
-%%\[\mathsf{Seq\,a} = \mathsf{a + Eq\,a\,(b \times c) \times Seq\,b
-%%  \times Seq\,c} \] 
-%\[\begin{array}{l}
-%\mathsf{data\, Seq :: * \to *\,where}\\
-%\mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Seq\,a}\\
-%\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, ::\, Eq\,a\,(b \times c) \to
-%  Seq \,b \to Seq\,c \to Seq\,a}\\ 
-%\end{array}\]
-%
-%The above rewriting of $\mathsf{Seq}$'s definition shows that it can
-%be seen as a fixpoint $\mu H$ of the higher-order function $H$ defined
-%on objects by $H\,F\,a = a + Eq\,a\,(b \times c) \times F\,b \times
-%F\,c$.
-
-{\color{blue} Need syntax that reflects semantic Lan.}
-
-%Other GADTs are similarly characterizable by higher-order functors.
-%{\color{blue} GADTs make essential use of higher-order functors.}
-
-Need syntax functors (Lan)
-The representation choices are similar for GADTs. The difference is
-that the two are not isomorphic.
-
-Certain polynomial GADTs correspond to polynomial ADTs. So we expect
-the same behaviors.  Polynomial nested types are actually the GADTs
-that mimic polynomial ADT behavior.
-
-But pattern matching fails for the extension (functorial completion)
-of a GADT.
+a b.\,f a \times f b\] where $\Lan_K \,G$ is the left Kan extension of
+$G$ along $K$.  The left Kan extension $\Lan_K \,G : {\cal E} \to
+{\cal D}$ of the functor $G : {\cal C} \to {\cal D}$ along the functor
+$K : {\cal C} \to {\cal E}$ is the best functorial approximation to
+$G$ that factors through $K$. Here, ``best functorial approximation''
+means that $\Lan_K \,G$ is the smallest functor that both extends the
+image of $K$ to $\cal D$ and coincides with $G$ on $\cal C$, in the
+sense that, for any other such functor $H$, there is a morphism of
+functors (i.e., a natural transformation) from $\Lan_K \,G$ to
+$H$. This is captured formally by the following
+definition~\cite{mac71}:
+\begin{definition}\label{def:lke}
+If $G : {\cal C} \to {\cal D}$ and $K : {\cal C} \to {\cal E}$ are
+functors, then the {\em left Kan extension of $G$ along $K$} is a
+functor $\mathit{Lan}_K\,G : {\cal E} \to {\cal D}$ together with a
+natural transformation $\eta : G \to (\mathit{Lan}_K\,G) \circ K$
+such that, for every functor $H : {\cal E} \to {\cal D}$ and natural
+transformation $\gamma : G \to H \circ K$, there exists a unique
+natural transformation $\mu : \mathit{Lan}_K\,G \to H$ such that $(\mu
+K) \circ \eta = \gamma$. This is depicted in the diagram
+\[\begin{tikzcd}[row sep = huge]
+{\cal C}
+\ar[rr, "{G}"{name=Fa, above}, ""{name=F, below}]
+\ar[rd, "{K}"']
+&& {\cal D} \\
+& {\cal E}
+\ar[Rightarrow, bend right = 25, from=F, "{\eta}"']
+\ar[ur, bend left, "{\mathit{Lan}_{K}\,G}"{name=Lan, description}]
+\ar[ur, bend right, "{H}"'{name=L, right}, ""'{name=Lr, right}]
+\ar[Rightarrow, bend left = 45, from=Fa, to=Lr, "{\gamma}" near start]
+\ar[Rightarrow, dashed, from=Lan, to=L, "{\mu}"']
+\end{tikzcd}\]
+\end{definition}
+
+To represent GADTs as primitive data types a calculus must support a
+primitive fixpoint constructor $\mathsf{\mu}$ that is the syntactic
+reflection of the higher-order fixpoint operator on the category of
+functors on the cagegory interpretating types, as well as a primitive
+type constructor $\mathsf{Lan}$ such that $\mathsf{Lan_k\,g}$ is the
+syntactic reflection of the left Kan extension $\Lan_K \,G$ of the
+functor $G$ interpreting $\mathsf{g}$ along the functor $K$
+interpreting $\mathsf{k}$. {\color{blue} This was elided in ADT
+  case. Why? Tie together better.}  The type constructor
+$\mathsf{Lan}$ thus computes the smallest data type that is generated
+by a GADT's {\color{blue} GADT or constructor?} syntax and also
+supports a $\mathsf{map}$ function. That is, $\mathsf{Lan}$ computes
+the functorial completion of the data type determined by a GADT's
+syntax.
+
+Note, however, that the functorial completion of a GADT need not be a
+nested type. As mentioned in the introduction, only those instances of
+the GADT reachable from the syntax of the GADT are in the functorial
+completion.
+
+The representation choices are the same for GADTs as for ADTs. The
+difference is that, although these two representations are isomorphic
+for ADTs, they are not isomorphic for GADTs.  We will see this in
+Section~\ref{??} below.
 

### 2021-02-15 20-gadts/GADT-story.tex
work on gadt-story Sec 2

@@
 To determine the possible shapes of a GADT's container we have to
-pattern-match on the type of the data to be contained. For this, our
+pattern-match on the type of the data to be contained. For this, a
 GADT calculus must support an equality type $\mathsf{Equal}$ that is
@@
 otherwise, i.e., is the syntactic reflection of semantic equality
-function $\mathit{Equal}$. Such a type can either be taken as primitive, as in
+function $\mathit{Equal}$. {\color{blue} This is the least commitment
+  thing Chirch encodings require.} Such a type can either be taken as
+primitive, as in
 \[\begin{array}{l}
@@
 solution to the higher-order fixpoint equation
-\[H\,f\,a = a \,+\,Lan_{\lambda a b.\,a \times b}\, \lambda
-a b.\,f a \times f b\] where $\Lan_K \,G$ is the left Kan extension of
+\[H\,f\,a = a \,+\,Lan_{\lambda a b. a \times b}\, \lambda
+a b. f a \times f b\] where $\Lan_K \,G$ is the left Kan extension of
 $G$ along $K$.  The left Kan extension $\Lan_K \,G : {\cal E} \to
@@
 transformation $\gamma : G \to H \circ K$, there exists a unique
-natural transformation $\mu : \mathit{Lan}_K\,G \to H$ such that $(\mu
+natural transformation $\delta : \mathit{Lan}_K\,G \to H$ such that $(\mu
 K) \circ \eta = \gamma$. This is depicted in the diagram
@@
 \ar[Rightarrow, bend left = 45, from=Fa, to=Lr, "{\gamma}" near start]
-\ar[Rightarrow, dashed, from=Lan, to=L, "{\mu}"']
+\ar[Rightarrow, dashed, from=Lan, to=L, "{\delta}"']
 \end{tikzcd}\]
@@
 functor $G$ interpreting $\mathsf{g}$ along the functor $K$
-interpreting $\mathsf{k}$. {\color{blue} This was elided in ADT
-  case. Why? Tie together better.}  The type constructor
+interpreting $\mathsf{k}$. {\color{blue} This is the least commitment
+  thing primitive representations can require. Fixpoints were elided
+  in ADT case. Why? Tie together better.}  The type constructor
 $\mathsf{Lan}$ thus computes the smallest data type that is generated
@@
 the functorial completion of the data type determined by a GADT's
-syntax.
-
-Note, however, that the functorial completion of a GADT need not be a
-nested type. As mentioned in the introduction, only those instances of
-the GADT reachable from the syntax of the GADT are in the functorial
-completion.
-
-The representation choices are the same for GADTs as for ADTs. The
-difference is that, although these two representations are isomorphic
-for ADTs, they are not isomorphic for GADTs.  We will see this in
-Section~\ref{??} below.
-
-\section{Semantics of GADTs}\label{sec:GADT-semantics}
-
-{\color{blue} OMIT THIS SECTION?}
-
-But the data in a GADT shouldn't disappear just because you map a
-function over it!  This is odd! So in what sense to GADTs generalize
-ADTs? In what sense are they even data types?
-
-The semantics of GADTs is not so clear.
-
-Functorial completion doesn't give nested types
-
-Functorial completion containerizes the GADT. Extension of a
-container. The extension of a container is an endofunctor generalizing
-map.
-
-Eq semantics minimal (least commitment, most natural) for CEs; Lan
-semantics minimal for primitives. Minimal way to add to any calculus
-supporting them.
-
-Bob requires map functions for fixpoint but then doesn't use them.
-
-An analysis of the Eq A B type in Bob's systems shows that the set
-interpretation is empty whenever the A and B have different set
-interpretations.  Likewise, its relational interpretation is the empty
-relation whenever the A and B have different relational
-interpretations (even when the underlying set interpretations are the
-same).  The consequences of parametricity are thus immediate: there
-can be no term of type Eq A B if A and B are different types.
-Moreover, this behavior is fundamentally non-functorial, so that there
-is no hope of defining a map for Eq.  In particular, that means we
-cannot expect to retrieve non-trivial naturality properties for the Eq
-type, despite these are usually considered to be a special case of
-parametricity properties.  Indeed, the common understanding that
-parametricity generalizes naturality only makes sense for functorial
-types.  We can specialize these considerations to the type Eq 1 B,
-representing G, our example GADT defined by a unique constructor c ::
-G 1.
-
-Since the types encoding GADTs in Bob's system make essential use
-of equality types, these considerations extend and possibly generalize
-to all GADTs.  Still, the equality is just a shorthand for a type made
-by higher-order forall and arrow types, and thus this considerations
-are not limited to a specific type, but are intrinsic of Bob's
-system.
-%We can thus expect that not all types in Bob's system will be
-%functorial and that, consequently, the consequences of parametricity
-%will not include functoriality but will provide information about,
-%e.g., inhabitatedness.  See
-%<https://www.cis.upenn.edu/~sweirich/talks/param-gadt.pdf> and
-%<https://www.seas.upenn.edu/~sweirich/papers/rtheorems.pdf> for more
-%about parametricity and GADTs.
-
-In contrast, $(Lan_1 1) B$, the type representing G in our system, is
-functorial (as are all types in our systems) and inhabited for every
-instance of B other than 0.  Indeed, we have a term $c : (Lan_1 1) 1$
-given by the Lan introduction rule, and thus every term t : T yields a
-term in $(Lan_1 1) T$ by use of the map of $(Lan_1 1) B$.  This shows
-that, as expected by now, the same GADT really does have different
-semantic interpretations in Bob's system and in ours.
-
-\section{Parametricity}
+syntax, filling it out with data elements that are derivable via
+repeated applications of $\mathsf{map}$ starting from those elements
+specified by its syntax. In the primitive view, GADTs are, in general,
+underspecified by their syntax.
+
+Just as for ADTs, we can represent GADTs as either Church encodings or
+primitive data types. But whereas the two representations are always
+isomorphic for ADTs, they not, in general, for GADTs. We will see this
+in Section~\ref{??} below. {\color{blue} But we get
+  rearrange-transform property for primitive rep since this is a
+  consequence of functoriality.}
+
+{\color{blue}
+Forced to choose, a programmer would likely find the idea that a GADT
+contains data that are not specified by its syntax disturbing. What,
+they might ask, should a data type contain other than data that are
+constructed using its data constuctors? Why should a GADT contain
+``secret'' elements that are only attainable by mapping? On the other
+hand, a semanticist would likely find the primitive representation of
+GADTs more reasonable. After all, they would likely find the
+nonfunctorial nature of the Church encoding unnerving at best. After
+all, they would argue, the data in a GADT shouldn't disappear just
+because you map a function over it! The fact that this happens for
+Church encodings of GADTs --- i.e., for GADTs as specified by their
+syntax --- highlights how GADTs {\em do not} generalize the essential
+container-ish nature of ADTs at all. A semanticist might therefore
+conclude that GADTs are seriously misnamed. They might even go on to
+wonder: in what sense are GADTs data types at all?}
+
+\section{Parametricity in the Presence of GADTs}
 
@@
 
-One important consequence of representation choice is for parametric
-reasoning about GADTs. In particular, we can use parametricity to get
-our transform-rearrange property for ADTs. Do it for lists?
+1. Parametricity requires a fibred semantics. This doesn't happen for
+primitive rep.
+
+2. Non-uniformity renders build, therefore short cut, for GADTs
+impossible. Can't even formulate! It is possible there could be
+another way to prove equivalence, but we know the two representations
+are not equivalent because we have a counterexample.
+
+The fact that GADTs don't validate any such iso has important
+consequences for their parametricity properties.
+
+Bob requires map functions for fixpoint but then doesn't use them.
+
+An analysis of the Eq A B type in Bob's systems shows that the set
+interpretation is empty whenever the A and B have different set
+interpretations.  Likewise, its relational interpretation is the empty
+relation whenever the A and B have different relational
+interpretations (even when the underlying set interpretations are the
+same).  The consequences of parametricity are thus immediate: there
+can be no term of type Eq A B if A and B are different types.
+Moreover, this behavior is fundamentally non-functorial, so that there
+is no hope of defining a map for Eq.  In particular, that means we
+cannot expect to retrieve non-trivial naturality properties for the Eq
+type, despite these are usually considered to be a special case of
+parametricity properties.  Indeed, the common understanding that
+parametricity generalizes naturality only makes sense for functorial
+types.  We can specialize these considerations to the type Eq 1 B,
+representing G, our example GADT defined by a unique constructor c ::
+G 1.
+
+Since the types encoding GADTs in Bob's system make essential use
+of equality types, these considerations extend and possibly generalize
+to all GADTs.  Still, the equality is just a shorthand for a type made
+by higher-order forall and arrow types, and thus this considerations
+are not limited to a specific type, but are intrinsic of Bob's
+system.
+
+In contrast, $(Lan_1 1) B$, the type representing G in our system, is
+functorial (as are all types in our systems) and inhabited for every
+instance of B other than 0.  Indeed, we have a term $c : (Lan_1 1) 1$
+given by the Lan introduction rule, and thus every term t : T yields a
+term in $(Lan_1 1) T$ by use of the map of $(Lan_1 1) B$.  This shows
+that, as expected by now, the same GADT really does have different
+semantic interpretations in Bob's system and in ours.
 
@@
 parametricity will not include functoriality but will provide
-information about, e.g., inhabitatedness.  See
-<https://www.cis.upenn.edu/~sweirich/talks/param-gadt.pdf> and
-<https://www.seas.upenn.edu/~sweirich/papers/rtheorems.pdf> for more
-about parametricity and GADTs.
-
-If we had a parametric model for GADTs, it could perhaps validate
-short cut and this show that the two reps for them are equivalent.
-(This is a sort of converse.)
-The fact that GADTs don't validate any such iso has important
-consequences for their parametricity properties.
-
-In this paper we show that there is not an iso between the two reps of
-even the simplest GADT {\color{blue} that isn't a hereditarily
-  polynomial nested type}, and investigate the consequences of this
-discovery for parametric reasoning about GADTs.
-
-Non-uniformity renders build, therefore short cut, for GADTs
-impossible. Can't even formulate! Maybe there's another way to prove
-equivalence, but I don't know what it would be. In any case, we have a
-counterexample, so this settles the question.
+information about, e.g., inhabitatedness.
+%See
+%<https://www.cis.upenn.edu/~sweirich/talks/param-gadt.pdf> and
+%<https://www.seas.upenn.edu/~sweirich/papers/rtheorems.pdf> for more
+%about parametricity and GADTs.
+
+{\color{blue} Our favorite GADT isn't a hereditarily
+  polynomial nested type.}
 
@@
 choose.
-
-Parametric polymorphism as a way to capture Strachey's notion of
-uniform, type-independent behavior of functions. Parametric models as
-formalizations of this notion of uniformity, as first given in
-Reynolds' original paper.
 
@@
 reasons about them.
-
-It depends on primitive vs. Church encodings. Primitive forces
-functoriality, as in LICS'19.
 

### 2021-02-16 20-gadts/GADT-story.tex
setting up Par section for Enrico

@@
 
+{\color{blue} THIS TRANSITION FROM EQUIVALENCE OF REPRESENTATION TO
+  PARAMETRICITY GOES SOMEWHERE:
+Non-uniformity renders build, therefore short cut, for GADTs
+impossible. Can't even formulate! It is possible there could be
+another way to prove equivalence, but we know the two representations
+are not equivalent because we have a counterexample.
+
+The fact that GADTs don't validate any such iso has important
+consequences for their parametricity properties.}
+
 \section{Parametricity in the Presence of GADTs}
@@
 
+{\color{blue} Say somewhere: A relation is a triple $(A,B,R)$, where
+  $R$ is a relation between sets $A$ and $B$.  A function from
+  $(A,B,R)$ to $(A',B',R')$ is a pair $(f : A \to A',g : B \to B')$ of
+  functions on sets such that $(f a,g\,b) \in R'$ if $(a,b) \in R$. We
+  may write $R : \rel(A,B)$ for $(A,B,R)$.  If $R : \rel(A,B)$ we
+  write $\pi_1 R$ and $\pi_2 R$ for the {\em domain} $A$ of $R$ and
+  the {\em codomain} $B$ of $R$, respectively, and assume $\pi_1$ and
+  $\pi_2$ are surjective. We write $\Eq_A = (A,A,\{(x,x)~|~ x \in
+  A\})$ for the {\em equality relation} on the set $A$.}
+
+The key idea underlying Reynolds' parametricity is to give each type
+$F(\alpha)$ with one free variable $\alpha$ a {\em set interpretation}
+$F_0$ taking sets to sets and a \emph{relational interpretation} $F_1$
+taking relations $R : \rel(A,B)$ to relations $F_1 (R) : \rel(F_0 (A),
+F_0 (B))$, and to interpret each term $t(\alpha,x) : F(\alpha)$ with
+one free term variable $x : G(\alpha)$ as a map $t_0$ associating to
+each set $A$ a function $t_0(A) : G_0(A) \to F_0(A)$. These
+interpretations are given inductively on the structures of $F$ and $t$
+in such a way that they imply two fundamental theorems. The first is
+an \emph{Identity Extension Lemma}, which states that $F_1(\Eq_A) =
+\Eq_{F_0(A)}$, and is the essential property that makes a model
+relationally parametric rather than just induced by a logical
+relation. The second is an \emph{Abstraction Theorem}, which states
+that, for any $R :\rel(A, B)$, $(t_0(A),t_0(B))$ is a morphism in
+$\rel$ from $(G_0(A),G_0(B),G_1(R))$ to $(F_0(A),F_0(B),F_1(R))$. The
+Identity Extension Lemma is similar to the Abstraction Theorem except
+that it holds for {\em all} elements of a type's interpretation, not
+just those that interpret terms.  Similar theorems are required for
+types and terms with any number of free variables.
+
 1. Parametricity requires a fibred semantics. This doesn't happen for
@@
 
-2. Non-uniformity renders build, therefore short cut, for GADTs
-impossible. Can't even formulate! It is possible there could be
-another way to prove equivalence, but we know the two representations
-are not equivalent because we have a counterexample.
-
-The fact that GADTs don't validate any such iso has important
-consequences for their parametricity properties.
+2. What are the parametricity properties for GADTs in each setting?
+(Naturality only with primitives, inhabitation etc. with CEs)
+
+{\color{red} ENRICO: Just write a section that can go here about
+  1. for our favorite GADT.}
+
+\vspace*{1in}
+
+NOTES/POINTS:
 

### 2021-02-16 20-gadts/GADT-story.tex
setting up Par section for Enrico

@@
 
+{\color{red} This might not be relevant, since we only care about the
+  fibredness.} 
 The key idea underlying Reynolds' parametricity is to give each type

### 2021-02-16 20-gadts/GADT-story.tex
setting up Par section

@@
 isomorphic for ADTs, they not, in general, for GADTs. We will see this
-in Section~\ref{??} below. {\color{blue} But we get
-  rearrange-transform property for primitive rep since this is a
-  consequence of functoriality.}
-
-{\color{blue}
+in the next section.
+
+\section{Non-Equivalence of Church Encodings and Primitive
+  Representations of GADTs}\label{sec:non-equiv}
+
+Intuitively it seems the two representations might be different. One
+is functorial, one is not. Non-uniform nature of GADTs renders build,
+therefore short cut, for GADTs impossible. Can't even formulate! Of
+course, there could be another way to prove equivalence.  But, in
+fact, CEs and primitive representations are not equivalent for
+arbitrary GADTs, as the following example shows.
+
+\begin{example}
+Use our favorite GADT to establish non-equivalence of the two reps
+\end{example}
+
 Forced to choose, a programmer would likely find the idea that a GADT
@@
 conclude that GADTs are seriously misnamed. They might even go on to
-wonder: in what sense are GADTs data types at all?}
-
-{\color{blue} THIS TRANSITION FROM EQUIVALENCE OF REPRESENTATION TO
-  PARAMETRICITY GOES SOMEWHERE:
-Non-uniformity renders build, therefore short cut, for GADTs
-impossible. Can't even formulate! It is possible there could be
-another way to prove equivalence, but we know the two representations
-are not equivalent because we have a counterexample.
-
-The fact that GADTs don't validate any such iso has important
-consequences for their parametricity properties.}
+wonder: in what sense are GADTs data types at all?
+
+The disappointment that GADTs don't validate any such equivalence is
+further compounded by the fact that the the two representations have
+very different parametricity properties.  {\color{blue} For example,
+  we get rearrange-transform property for primitive rep since this is
+  a consequence of functoriality, but don't get it for Church
+  Encodings.} We explore the differences in parametricity results for
+the two representations in the next section.
 
@@
   $\pi_2$ are surjective. We write $\Eq_A = (A,A,\{(x,x)~|~ x \in
-  A\})$ for the {\em equality relation} on the set $A$.}
+  A\})$ for the {\em equality relation} on the set $A$. ALSO, WANT OUR
+RELATIONS TO BE ONTO.}
 

### 2021-02-16 20-gadts/GADT-story.tex
added gadt-story-references.bib

@@
   nested such types} nested types via the categorical model
-of~\cite{jgj21}. But, as shown in Section~\ref{sec:?} below, for some
-{\color{blue} (even polynomial)} GADTs,
+of~\cite{jgj21}. But, as shown in Section~\ref{sec:par} below, for
+some {\color{blue} (even polynomial)} GADTs,
 %that are not hereditarily polynomial nested types,
@@
 
-GADTs are traditionally represented via Church encodings. In this
-view, GADTs are completely determined by their syntax. For example,
-the GADT $\mathsf{G}$ defined by
+{\color{red} Maybe use a less degenerate example.}  GADTs are
+traditionally represented via Church encodings. In this view, GADTs
+are completely determined by their syntax. For example, the GADT
+$\mathsf{G}$ defined by
 \[\begin{array}{l}
@@
 
+{\color{red} This will probably be more convincing if we use, e.g.,
+  an expression GADT from Stephanie above.}
 A GADT programmer is likely to use GADTs precisely because they
@@
 \[H\,f\,a = a \,+\,Lan_{\lambda a b. a \times b}\, \lambda
-a b. f a \times f b\] where $\Lan_K \,G$ is the left Kan extension of
-$G$ along $K$.  The left Kan extension $\Lan_K \,G : {\cal E} \to
+a b. f a \times f b\] where $Lan_K \,G$ is the left Kan extension of
+$G$ along $K$.  The left Kan extension $Lan_K \,G : {\cal E} \to
 {\cal D}$ of the functor $G : {\cal C} \to {\cal D}$ along the functor
@@
 $G$ that factors through $K$. Here, ``best functorial approximation''
-means that $\Lan_K \,G$ is the smallest functor that both extends the
+means that $Lan_K \,G$ is the smallest functor that both extends the
 image of $K$ to $\cal D$ and coincides with $G$ on $\cal C$, in the
 sense that, for any other such functor $H$, there is a morphism of
-functors (i.e., a natural transformation) from $\Lan_K \,G$ to
+functors (i.e., a natural transformation) from $Lan_K \,G$ to
 $H$. This is captured formally by the following
@@
 reflection of the higher-order fixpoint operator on the category of
-functors on the cagegory interpretating types, as well as a primitive
+functors on the category interpretating types, as well as a primitive
 type constructor $\mathsf{Lan}$ such that $\mathsf{Lan_k\,g}$ is the
@@
 functor $G$ interpreting $\mathsf{g}$ along the functor $K$
-interpreting $\mathsf{k}$. {\color{blue} This is the least commitment
+interpreting $\mathsf{k}$. Using the $\mu$ and $\Lan$ type
+constructors we can represent, e.g., the GADT $\mathsf{Seq}$ as
+\[\mathsf{Seq\,a} = \mathsf{\mu \phi.\, a + Lan_{\lambda c d. c \times d}
+  \lambda c d. \phi c \times \phi d}\]
+
+{\color{blue} This is the least commitment
   thing primitive representations can require. Fixpoints were elided
@@
 primitive data types. But whereas the two representations are always
-isomorphic for ADTs, they not, in general, for GADTs. We will see this
-in the next section.
+isomorphic for ADTs, they are not, in general, for GADTs. We will see
+this in the next section.
 
@@
 Intuitively it seems the two representations might be different. One
-is functorial, one is not. Non-uniform nature of GADTs renders build,
-therefore short cut, for GADTs impossible. Can't even formulate! Of
-course, there could be another way to prove equivalence.  But, in
-fact, CEs and primitive representations are not equivalent for
-arbitrary GADTs, as the following example shows.
-
-\begin{example}
+is functorial, one is not. It is possible they could always have the
+same data elements even if not the same morphisms. But as we see in
+Example~\ref{ex:g1}, this turns out not to be the case.
+
+Non-uniform nature of GADTs renders build, therefore short cut, for
+GADTs impossible. Can't even formulate! Of course, there could be
+another way to prove equivalence.  But, in fact, CEs and primitive
+representations are not equivalent for arbitrary GADTs, as the
+following example shows.
+
+\begin{example}\label{ex:g1}
 Use our favorite GADT to establish non-equivalence of the two reps
@@
 Forced to choose, a programmer would likely find the idea that a GADT
-contains data that are not specified by its syntax disturbing. What,
-they might ask, should a data type contain other than data that are
-constructed using its data constuctors? Why should a GADT contain
-``secret'' elements that are only attainable by mapping? On the other
-hand, a semanticist would likely find the primitive representation of
-GADTs more reasonable. After all, they would likely find the
-nonfunctorial nature of the Church encoding unnerving at best. After
-all, they would argue, the data in a GADT shouldn't disappear just
-because you map a function over it! The fact that this happens for
-Church encodings of GADTs --- i.e., for GADTs as specified by their
-syntax --- highlights how GADTs {\em do not} generalize the essential
+contains data that are not specified by its syntax more than a little
+disturbing. What, they might ask, should a data type contain other
+than data that are constructed using its data constuctors? Why should
+a GADT contain ``hidden'' elements that are only accessible by
+mapping? On the other hand, a semanticist would likely find the
+primitive representation of GADTs more reasonable. Indeed, they would
+likely find the nonfunctorial nature of the Church encoding unnerving
+at best. After all, they would likely argue, the data in a GADT
+shouldn't disappear just because you map a function over it! The fact
+that this happens when GADTs are represented by their Church encodings
+actually highlights how GADTs {\em do not} generalize the essential,
 container-ish nature of ADTs at all. A semanticist might therefore
@@
 
-\section{Parametricity in the Presence of GADTs}
+\section{Parametricity in the Presence of GADTs}\label{sec:par}
 
@@
 1. Parametricity requires a fibred semantics. This doesn't happen for
-primitive rep.
-
-2. What are the parametricity properties for GADTs in each setting?
-(Naturality only with primitives, inhabitation etc. with CEs)
+primitive rep so can't have a par model with this rep. Can still get
+naturality properties, however.
+
+2. For CEs can get par model and par properties. But what do these say
+for GADTs? Give examples?
 
@@
 
-We do minimal thing to fix functoriality, then investigate
-parametricity properties for both interpretations. We are forced to
-choose.
-
 Bob's system gives a parametric model that models GADTs. But given the
@@
 the Eq type?
-
-GADTs Force a Representation Choice. It has consequences for how one
-reasons about them.
 

### 2021-02-17 20-gadts/GADT-story.tex
added Enrico's text

@@
 
-\subsection{Algebraic Data Types}
+\subsection{Algebraic Data Types}\label{sec:adts}
 
@@
 \begin{example}\label{ex:g1}
-Use our favorite GADT to establish non-equivalence of the two reps
+{\color{red} THIS EXAMPLE IS ENRICO'S TEXT}
+Consider the following GADT, defined in a Haskell-like syntax:
+\begin{equation}\label{eq:example-GADT}
+\begin{array}{l}
+\mathsf{data }\; G : \ast \to \ast\;\mathsf{where}\\
+\hspace*{0.4in}c\;:\;G\;1
+\end{array}
+\end{equation}
+This GADT has a unique constructor, providing a data of type $G\;1$.
+While we might say this is the simplest GADT possible, it still
+provides an interesting case-study, showing the differences between
+the primitive data type representation and the Church encoding
+representation.
+
+We now represent $G$ via its Church encoding.  Since $G$ has a single
+constructor, which provides a data of type $G\,1$ without requiring
+any input, its Church encoding is
+\[
+G\,\alpha = \mathsf{\forall f.\, (\forall a.\,
+  f\,a) \to f\,\onet}
+\]
+meaning that $G\,A = \mathsf{Equal}\,\onet\,A$.  Thus, $G$ tests its
+argument for equality against the unit type, $\onet$.
+
+Recall from the previous section {\color{blue} use reference} that the
+semantic interpretation of $\mathsf{Equal}\,\onet\,A$ is given by a
+characteristic function, in the sense that it is the singleton 1 when
+$A$ is 1, and the empty set otherwise.  Thus, as a data type
+represented by Church encoding, G will be the data type containing one
+(and only one) piece of data when instantiated on $\onet$, and
+uninhabited otherwise.  This perfectly mirrors the syntax of $G$,
+which indeed provides just one constructor when the type is
+instantiated on 1, and none otherwise.  Thus, given a set $A$, the set
+interpretation $G_0(A)$ is 1 if $A = 1$, and $\emptyset$ otherwise.
+
+We now want a primitive representation of $G$.  Recall that $G$ has a
+unique constructor $c : G\,\onet$, which can be equivalently expressed
+as $c: \onet \to G\,\onet$.  By using a $\Lan$ type, we can express
+$c$ as a constructor $(\Lan_{\onet} \onet)\alpha \to G \alpha$.  Thus,
+as a primitive data type, we represent $G$ as $\mu \phi. \lambda
+\alpha. (\Lan^{\emptyset}_{\onet} \onet)\alpha$.  {\color{blue}
+  (WHERE?) notice that the constructors of a GADT have to be functors,
+  because the GADT is the fixpoint of a higher order functor.}
+
+The semantic interpretation of $\mu \phi. \lambda
+\alpha. (\Lan^{\emptyset}_{\onet} \onet)\alpha$ is given by the
+fixpoint of the functor yielded by the semantic interpretation of
+$\Lan^{\emptyset}_{\onet} \onet$, the body of the $\mu$ type.  Notice,
+though, that the recursive variable $\phi$ does not appear in the body
+of the $\mu$ type.  Thus, the semantic interpretation is actually just
+the semantic interpretation of the body itself.  Therefore, the
+semantic interpretation is $\mathsf{\Lan}_1 1$, where $1$ is the
+constantly $1$-valued functor $\ast \to \set$ from the category with a
+single object $\ast$ to $\set$.
+
+It turns out that $\mathsf{\Lan}_1 1$ is, in fact, none other than the
+identity functor on $\set$.  Indeed, for any set $A$,
+$(\mathsf{\Lan}_1 1) A$ is the set of all functions $a : 1 \to A$
+modulo the smallest equivalence relation $\sim$ such that, for any $f
+: 1 \to 1$, we have $a \sim a \circ f$.  But this is isomorphic to
+$A$, because the only function $1 \to 1$ is the identity.
+{\color{blue} Reference Computing left Kan extensions - M.R.Bush
+  M.Leeming R.F.C.Walters}
+
+The Church encoding representation of $G$ and the primitive
+representation of $G$ are thus non-equivalent: while the former only
+holds one data element in the instance $G\,1$ and none in other
+instances, the latter is inhabited in all instances with the exception
+of $G\,\emptyset$.  This makes sense if one remembers that the left
+Kan extension, reflected into the syntactic type $\Lan$, is a functor,
+and thus it comes with a functorial map.  This means that, for every
+element $a \in A$, we can map the associated function $\bar{a} : 1 \to
+A$ along $G$ to get a function $\map_{G}\,\bar{a} : G\,1 \to G\,A$,
+and thus we have an element in $G\,A$ given by $\map_{G}\,\bar{a}\, c$
+where $c \in G\,1$ is the element provided by the constructor.
 \end{example}
@@
 
-{\color{red} ENRICO: Just write a section that can go here about
-  1. for our favorite GADT.}
+{\color{red} EVERYTHING BEFORE THE NOTES BELOW IS ENRICO'S TEXT:} To
+even consider the question of whether our semantics is parametric we
+first need to define a relational semantics for our types, i.e., a
+relational interpretation for every type $G : \ast \to \ast$ such
+that, for every relation $R : \rel(A, B)$, we have $G_1(R) :
+\rel(G_0(A), G_0(B))$.
+
+We shall show that this is not generally possible for the primitive
+representation of data types, by considering the previously defined
+type $G$ in Equation~\ref{eq:example-GADT}.  Recall that the semantic
+interpretation of $G$ is given by $\mathsf{\Lan}_1 1$ which, in fact,
+is none other than the identity functor on $\set$.
+
+Analogously to the set interpretation, the relational interpretation
+for $G$ is given by $\mathsf{\Lan}_{\Eq_1} \Eq_1$.  Notice that the
+interpretation still essentially relies on a left Kan extension, but
+this time it is an extension of $\rel$-valued functors.  We would need
+to prove that, for every relation $R : \rel(A, B)$, we have that
+$(\mathsf{\Lan}_{\Eq_1} \Eq_1) R$ is a relation on $(\mathsf{\Lan}_1
+1) A$ and $(\mathsf{\Lan}_1 1)B$, i.e., $A$ and $B$.  Instead, we will
+provide a counterexample showing that this is not true.
+
+Consider the relation $R = (1, \emptyset, \emptyset)$.  We would thus
+expect $(\mathsf{\Lan}_{\Eq_1} \Eq_1) R$ to be a relation with domain
+$1$.  Because the left Kan extension preserves
+projections~\cite{rie16} we can compute the domain as
+$(\mathsf{\Lan}_{\Eq_1} 1) R$ (notice that the left Kan extension of
+$1 : \ast \to \set$ along $\Eq_1 : \ast \to \rel$ is a functor $\rel
+\to \set$).
+
+Following {\color{blue} Reference Computing left Kan extensions -
+  M.R.Bush M.Leeming R.F.C.Walters} we know that
+$(\mathsf{\Lan}_{\Eq_1} 1) R$ is the set of all morphisms of relations
+$f : \Eq_1 \to R$ modulo an equivalence relation; but there can be no
+morphism of relations $\Eq_1 \to R$ because its second component would
+have to be a function $1 \to \emptyset$, and there can be no such
+function.  Thus, $(\mathsf{\Lan}_{\Eq_1} 1) R = \emptyset$, which is
+not the domain we would have expected.
+
+The above example shows that the set interpretation and the relational
+interpretation for the primitive representation of a data type are
+inconsistent with each other, and thus it is not even possible to
+consider the question of whether the semantics is parametric.
+
+One could regard the above example as artificial, as it employs an
+empty relation on a non-empty domain.  Though, it is possible to apply
+the same argument, with a slightly more complex calculation, even to a
+total relation such as $(1, 2, 1 \times 2)$.
+
+\vspace*{0.2in}
+
+%%% CE are non functorial but parametric
+As we have seen in the previous section {\color{blue} ref}, GADTs
+represented as Church encodings do not necessarily have a map or a
+functorial semantics.  In particular, this means that we cannot
+possibly derive naturality results for those types, event though their
+semantics is parametric.  Indeed, even though naturality is sometimes
+seen as a consequence of parametricity, that is only true in case we
+have a functorial semantics and types supporting a map.  Still, GADTs
+represented as Church encodings can still exhibit consequences of
+paramentricity other than naturality, as they do admit parametric
+semantics {\color{blue} cite Bob's paper?}.
+
+%%% Example with G
+{\color{blue} In this paragraph, $\mathsf{G}$ and $\mathsf{t}$
+  represent syntax, so should be sans-serif, but in the notes, the
+  paragraph about set/relational interpretation and
+  parametricity/IEL/AT uses serif font.}  For example, consider the
+Church encoding of the GADT $\mathsf{G}$ defined
+in~\ref{eq:example-GADT}.  If $\mathsf{t}$ is a term of type
+$\mathsf{G}$ and $R$ is a relation on the sets $A$ and $B$, then
+$t_0(A) \in G_0(A)$ and $t_0(B) \in G_0(B)$, and by the abstraction
+theorem we have that $t_0(A)$ and $t_0(B)$ are related in $G_1(R)$.
+We also know that $G_1(R)$ is the empty relation whenever $R$ is not
+$\mathit{Equal}\,1$.  We thus deduce that there can be no term of type
+$\mathsf{G\,a}$ unless $\mathsf{a}$ is semantically equivalent to
+$\mathsf{1}$.  This result is the syntactic counterpart of the
+inhabitation result for $\mathsf{G}$ {\color{blue} are inhabitation
+  results about semantics or syntax?}  showing that all of the
+instances of $\mathsf{G}$ are uninhabited, with the exception of the
+instance for $\mathsf{1}$.  Of course, we could not possibly have a
+term of an uninhabited type.
+
+%%% Primitive representation of GADT is non parametric, but functorial
+On the contrary, as we have shown previously {\color{blue} ref}, a
+GADT represented as a primitive data type does not necessarily admit a
+parametric semantics, but, as a fixpoint of a higher-order functor, it
+does admit a functorial semantics and, consequently, a functorial map.
+In particular, every type that has a functorial map also benefits from
+naturality properties.  In other words, there is a suitable
+rearrange-transform property as explained in Section~\ref{sec:adts}.
+
+%%% Example with G
+For example, $\mathsf{G}$ is associated to a functorial map,
+$\map_{\mathsf{G}}$, such that, for every function $\mathsf{f : a \to
+  b}$, we get a function $\map_{\mathsf{G}}\,\mathsf{f} : \mathsf{G\,a
+  \to G\,b}$.  Then we have a term $\map_{\mathsf{G}}\,\mathsf{f\,c} :
+\mathsf{G\,a}$ for every function $\mathsf{f : \mathsf{1} \to a}$,
+where $\mathsf{c : G\,\mathsf{1}}$ is the constructor provided by the
+definition of $\mathsf{G}$.
 

### 2021-02-18 20-gadts/GADT-story.tex
work on gadt-story

@@
 \begin{abstract}
-FIX ME!!!! GADTs are misnamed. They don't really generalize ADTs. And
-they don't behave the way we expect data types to behave. In
+{\color{red} FIX ME!} GADTs are misnamed. They don't really generalize
+ADTs. And they don't behave the way we expect data types to behave. In
 particular, they force us to choose between the naturality properties
@@
   cut fusion}~\cite{glp93} the two representations are, in fact,
-equivalent, in the sense that they both have the same interpretation
-in the model. Models in which the Church encodings and primitive
+semantically equivalent, i.e., both have the same interpretation in
+the model. Models in which the Church encodings and primitive
 representations of all {\color{blue} (polynomial)} algebraic data
-types (ADTs) are equivalent include the operationally-based model
-of~\cite{pit98,pit00} and the categorical models
-of~\cite{joh02,joh03}.
+types (ADTs) are semantically equivalent include the
+operationally-based model of~\cite{pit98,pit00} and the categorical
+models of~\cite{joh02,joh03}.
 %\footnote{{\color{blue} Pitts proves the isomorphism only for
@@
 %    data type that is the fixpoint of a first-order functor.}}
-This equivalence was recently extended to the class of what might be
-called ``hereditarily polynomial'' {\color{blue} to include truly
-  nested such types} nested types via the categorical model
-of~\cite{jgj21}. But, as shown in Section~\ref{sec:par} below, for
-some {\color{blue} (even polynomial)} GADTs,
+Correctness of short cut fusion, and thus semantic equivalence of
+Church encodings and primitive representations, was recently extended
+to the class of what might be called ``hereditarily polynomial''
+{\color{blue} to include truly nested such types} nested types via the
+categorical model of~\cite{jgj21}. But, as shown in
+Section~\ref{sec:par} below, for some {\color{blue} (even polynomial)}
+GADTs,
 %that are not hereditarily polynomial nested types,
@@
 
-{\color{red} Maybe use a less degenerate example.}  GADTs are
-traditionally represented via Church encodings. In this view, GADTs
-are completely determined by their syntax. For example, the GADT
-$\mathsf{G}$ defined by
-\[\begin{array}{l}
-\mathsf{data\,G : * \to *\,where}\\
-\mathsf{\;\;\;\;\;\;\;\;c : G\,1}
-\end{array}\]
-{\color{blue} Use $\top$ instead of $1$ throughout?}
-
+GADTs are traditionally represented via Church encodings. In this
+view, GADTs are completely determined by their syntax; for example,
+the GADT $\mathsf{G}$ defined in Haskell by
+\begin{equation}\label{eq:example-GADT}
+\begin{array}{l}
+\mathsf{data\, Expr\,a\,where}\\
+\mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Expr\,a}\\
+\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, ::\, Expr \,a \to Expr\,b \to
+  Expr\,(a \times b)}\\ 
+\end{array}
+\end{equation}
+%\[\begin{array}{l}
+%\mathsf{data\,G : * \to *\,where}\\
+%\mathsf{\;\;\;\;\;\;\;\;c : G\,1}
+%\end{array}\]
+%{\color{blue} Use $\top$ instead of $1$ throughout?}
 \noindent
-{\color{blue} data type vs data structure --- which should have the
-functorial completion? The data type.}
-
-\noindent
-comprises exactly one element, namely $\mathsf{c}$. By contrast, the
-primitive representation sees a GADT as the {\em functorial
-  completion}~\cite{jp19} of its syntax. {\color{blue} Completing the
-  syntax is necessary to have a {\em functor} to take the least fixed
-  point of, so is essential to the view of data types as primitive.}
-Viewed as a primitive data type, $\mathsf{G}$ comprises not just
-$\mathsf{c}$, but also all data elements of the form
-$\mathsf{map\;f\;c}$ for all types $\mathsf{t}$ and all functions
-$\mathsf{f : 1 \to t}$ definable in the language, as well as all data
-elements of the form $\mathsf{map\;g\;x}$ for each appropriately typed
-function $\mathsf{g}$ and each element $\mathsf{x}$ already added to
-the data type, and so on.  Although functorial completion adds no new
-data elements to the syntax of a (polynomial) ADT or a (hereditarily
-polynomial) nested type, so that its primitive Church encoding
-representations coincide, these are rarely the same for GADTs.
-Indeed, they are quite different for even the very simple GADT
-$\mathsf{G}$ above.
+comprises expressions of any type $\mathsf{a}$ and expressions
+obtained by pairing other expressions. By contrast, the primitive
+representation sees a GADT as the {\em functorial
+  completion}~\cite{jp19} of its syntax. Completing a GADT's syntax to
+a {\em functor} is necessary in order to be able to take the least
+fixed point~\cite{tfca}, so it is inherent in the view of data types
+as their primitive representations.  Viewed as its primitive
+representation, $\mathsf{Expr}$ comprises not just those data elements
+representable in syntax, but also all data elements of the form
+$\mathsf{map\,f\,e}$ for all types $\mathsf{t_1}$ and $\mathsf{t_1}$,
+all functions $\mathsf{f : t_1 \to t_2}$ definable in the language,
+and all $\mathsf{e :: Expr\,t_1}$, as well as all data elements of the
+form $\mathsf{map\,g\,e'}$ for each appropriately typed function
+$\mathsf{g}$ and each element $\mathsf{e'}$ already added to the data
+type, and so on. Functorial completion adds no new data elements to
+the syntax of a GADT that is actually a (polynomial) ADT or a
+(hereditarily polynomial) nested type, so its Church encoding and
+primitive representation coincide. However, these representations are
+rarely the same for GADTs.
+%Indeed, they are quite different for even the very simple GADT
+%$\mathsf{G}$ above.
 
@@
 
-A semanticist, on the other hand, is likely to follow~\cite{bfss90}
-and expect GADTs to behave as their names suggests --- i.e., as
-generalizations of ADTs. The semanticist is thus likely to expect a
-GADT to be able to be used as a ADT is used --- i.e., as a shape
-filled with data that can be changed without also changing the
-shape. When used this way, it is more natural to view the GADT in
-terms of its primitive representation. In fact, the functorial
-completion of a GADT inherent in its primitive representation is
-exactly what is needed to ensure that the GADT supports a
-$\mathsf{map}$ function, i.e., is a functor both in the catgorical
-sense and in the sense of being an instance of Haskell's
-$\mathsf{Functor}$ class.  It is worth noting, however, that since the
-functorial completion of a GADT includes only those data elements that
-are derivable via repeated applications of $\mathsf{map}$ starting
-from those specified by its syntax, primitive GADTs need not, in
-general, be nested types. For example, the functorial completion of
-our GADT $\mathsf{G}$ includes elements at any instance
-$\mathsf{G\,t}$ because there is a function from $\mathsf{1}$ to any
-type $\mathsf{t}$, but it does not include any terms at instance
-$\mathsf{G\,0}$ because no calculus allows the definition of a
-function into $\mathsf{0}$. {\color{blue} Use $\bot$ instead of $0$
-  here?}
+A person interested in the semantics of functional languages, on the
+other hand, is likely to follow~\cite{bfss90} and expect GADTs to
+behave as their names suggests --- i.e., as generalizations of
+ADTs. The semanticist is thus likely to expect a GADT to be able to be
+used as a ADT is used --- i.e., as a shape filled with data that can
+be changed without also changing the shape. When used this way, it is
+more natural to view the GADT in terms of its primitive
+representation. In fact, the functorial completion of a GADT inherent
+in its primitive representation is exactly what is needed to ensure
+that the GADT supports a $\mathsf{map}$ function, i.e., is a functor
+both in the catgorical sense and in the sense of being an instance of
+Haskell's $\mathsf{Functor}$ class.  It is worth noting, however, that
+since the functorial completion of a GADT includes only those data
+elements that are derivable via repeated applications of
+$\mathsf{map}$ starting from those specified by its syntax, primitive
+GADTs need not, in general, be nested types. For example, the
+functorial completion of our GADT $\mathsf{G}$ includes elements at
+any instance $\mathsf{G\,t}$ because there is a function from
+$\mathsf{1}$ to any type $\mathsf{t}$, but it does not include any
+terms at instance $\mathsf{G\,0}$ because no calculus allows the
+definition of a function into $\mathsf{0}$. {\color{blue} Use $\bot$
+  instead of $0$ here?}
 
@@
 
-\section{Representations of Data Types}
+\section{Representations of Data Types}\label{sec:reps}
 
@@
 \[\mathsf{fmap_{Tree}\,f\,(g \, xs) \,=\, g\,(fmap_{List}\,f\,xs)}\]
+{\color{violet} mention naturality?}
 
@@
 for polynomial ADTs this can always be done. But even when the Church
-encoding of an ADT supports a $\mathsf{map}$ function, there is no
-guarantee that the rearrange-transform property will hold for it.
+encoding of an ADT supports function with the same type as its actual
+$\mathsf{map}$, there is no guarantee that the rearrange-transform
+property will hold for it. For example, if {\color{violet} Are there
+  examples of such non-functorial maps?}
 
@@
 
-\subsubsection{Church Encodings of GADTs}
+\subsubsection{Church Encodings of GADTs}\label{sec:CEs}
 
@@
 function $\mathit{Equal}$. {\color{blue} This is the least commitment
-  thing Chirch encodings require.} Such a type can either be taken as
+  thing Church encodings require.} Such a type can either be taken as
 primitive, as in
@@
 
-\subsubsection{GADTs as Primitive Data Types}
+\subsubsection{GADTs as Primitive Data Types}\label{sec:prim-gadts}
 
@@
 
+{\color{blue} rearrange-transform property for GADTs?}
+
 \section{Non-Equivalence of Church Encodings and Primitive
@@
 
-Intuitively it seems the two representations might be different. One
-is functorial, one is not. It is possible they could always have the
-same data elements even if not the same morphisms. But as we see in
+{\color{red} Check all fonts!! Syntax vs. semantics!}  Intuitively it
+seems the two representations might be different. One is functorial,
+one is not. It is possible they could always have the same data
+elements even if not the same morphisms. But as we will see in
 Example~\ref{ex:g1}, this turns out not to be the case.
@@
 following example shows.
+{\color{blue} Our favorite GADT isn't a hereditarily
+  polynomial nested type.}
 
 \begin{example}\label{ex:g1}
-{\color{red} THIS EXAMPLE IS ENRICO'S TEXT}
 Consider the following GADT, defined in a Haskell-like syntax:
@@
 \begin{array}{l}
-\mathsf{data }\; G : \ast \to \ast\;\mathsf{where}\\
-\hspace*{0.4in}c\;:\;G\;1
+\mathsf{data\,G : * \to *\,where}\\
+\mathsf{\;\;\;\;\;\;\;\;c : G\,1}
 \end{array}
 \end{equation}
-This GADT has a unique constructor, providing a data of type $G\;1$.
-While we might say this is the simplest GADT possible, it still
-provides an interesting case-study, showing the differences between
-the primitive data type representation and the Church encoding
-representation.
-
-We now represent $G$ via its Church encoding.  Since $G$ has a single
-constructor, which provides a data of type $G\,1$ without requiring
-any input, its Church encoding is
-\[
-G\,\alpha = \mathsf{\forall f.\, (\forall a.\,
-  f\,a) \to f\,\onet}
-\]
-meaning that $G\,A = \mathsf{Equal}\,\onet\,A$.  Thus, $G$ tests its
-argument for equality against the unit type, $\onet$.
-
-Recall from the previous section {\color{blue} use reference} that the
-semantic interpretation of $\mathsf{Equal}\,\onet\,A$ is given by a
-characteristic function, in the sense that it is the singleton 1 when
-$A$ is 1, and the empty set otherwise.  Thus, as a data type
-represented by Church encoding, G will be the data type containing one
-(and only one) piece of data when instantiated on $\onet$, and
-uninhabited otherwise.  This perfectly mirrors the syntax of $G$,
-which indeed provides just one constructor when the type is
-instantiated on 1, and none otherwise.  Thus, given a set $A$, the set
-interpretation $G_0(A)$ is 1 if $A = 1$, and $\emptyset$ otherwise.
-
-We now want a primitive representation of $G$.  Recall that $G$ has a
-unique constructor $c : G\,\onet$, which can be equivalently expressed
-as $c: \onet \to G\,\onet$.  By using a $\Lan$ type, we can express
-$c$ as a constructor $(\Lan_{\onet} \onet)\alpha \to G \alpha$.  Thus,
-as a primitive data type, we represent $G$ as $\mu \phi. \lambda
-\alpha. (\Lan^{\emptyset}_{\onet} \onet)\alpha$.  {\color{blue}
-  (WHERE?) notice that the constructors of a GADT have to be functors,
-  because the GADT is the fixpoint of a higher order functor.}
-
-The semantic interpretation of $\mu \phi. \lambda
-\alpha. (\Lan^{\emptyset}_{\onet} \onet)\alpha$ is given by the
-fixpoint of the functor yielded by the semantic interpretation of
-$\Lan^{\emptyset}_{\onet} \onet$, the body of the $\mu$ type.  Notice,
-though, that the recursive variable $\phi$ does not appear in the body
-of the $\mu$ type.  Thus, the semantic interpretation is actually just
-the semantic interpretation of the body itself.  Therefore, the
-semantic interpretation is $\mathsf{\Lan}_1 1$, where $1$ is the
-constantly $1$-valued functor $\ast \to \set$ from the category with a
-single object $\ast$ to $\set$.
-
+Syntactically, this GADT comprises a single data element, namely
+$\mathsf{c :: G \, 1}$. Despite the simplicity of this GADT, it serves
+as an informative case study highlighting the differences between the
+Church encoding of a GADT and that GADT's representation as a
+primitive data type.
+
+The Church encoding of $\mathsf{G}$ is
+\[\mathsf{G\,a} = \mathsf{\forall f.\, (\forall b.\,
+  f\,b) \to f\,1} \] In other words, $\mathsf{G\,a =
+  \mathsf{Equal}\,1\,a}$, so $\mathsf{G}$'s effect is simply to test
+its argument for equality against the unit type $\mathsf{1}$.
+
+Recall from the Section~\ref{sec:CEs} that the semantic interpretation
+of $\mathsf{Equal\,1\,a}$ is given by a characteristic function, in
+the sense that it is the singleton $1$ when the interpretation of
+$\mathsf{a}$ is $1$, and is the empty set otherwise.  Thus, as a data
+type represented by Church encoding, $\mathsf{G}$ will be the data
+type containing one (and only one) piece of data when instantiated on
+$\mathsf{1}$, and uninhabited otherwise.  This perfectly mirrors the
+syntax of $\mathsf{G}$, which indeed provides just one constructor
+when the type is instantiated on $\mathsf{1}$, and none otherwise.
+Thus, given a set $A$ interpreting $\mathsf{a}$, the set
+interpretation $G_0(A)$ is $1$ if $A = 1$ and $\emptyset$ otherwise.
+
+We now want a primitive representation of $\mathsf{G}$.  Recall that
+$\mathsf{G}$ has a unique constructor $\mathsf{c : G\,1}$, which can
+be equivalently expressed as $\mathsf{c: 1 \to G\,1}$.  By using a
+a$\Lan$ type, we can express $\mathsf{c}$ as a constructor
+$\mathsf{(Lan_1\, 1)\alpha \to G \alpha}$.  Thus, as a primitive data
+type, we represent $\mathsf{G}$ as $\mu \phi. \lambda
+\alpha. (\Lan^{\emptyset}_1 \,1)\alpha$.  {\color{blue} (WHERE?)
+  Notice that the constructors of a GADT have to be functors, because
+  the GADT is the fixpoint of a higher order functor. Also, we're not
+  using the superscript notation on $\Lan$.}
+
+The semantic interpretation of $\mathsf{\mu \phi. \lambda
+  \alpha. (\Lan^{\emptyset}_1\, 1)\alpha}$ is given by the fixpoint of
+the functor yielded by the semantic interpretation of
+$\mathsf{Lan^{\emptyset}_1\, 1}$, the body of the $\mathsf{\mu}$-type.
+Notice, though, that the recursive variable $\mathsf{\phi}$ does not
+appear in the body of the $\mathsf{\mu}$-type.  Thus, the semantic
+interpretation is actually just the semantic interpretation of the
+body itself.  Therefore, the semantic interpretation is
+$\mathsf{Lan_1 1}$, where $1$ is the constantly $1$-valued functor
+$\ast \to \set$ from the category with a single object $\ast$ to
+$\set$.
+
+{\color{red} Syntax? Semantics? Which are we talking here?}
 It turns out that $\mathsf{\Lan}_1 1$ is, in fact, none other than the
@@
 : 1 \to 1$, we have $a \sim a \circ f$.  But this is isomorphic to
-$A$, because the only function $1 \to 1$ is the identity.
-{\color{blue} Reference Computing left Kan extensions - M.R.Bush
-  M.Leeming R.F.C.Walters}
+$A$, because the only function $1 \to 1$ is the identity~\cite{blw03}.
 
@@
 
-1. Parametricity requires a fibred semantics. This doesn't happen for
-primitive rep so can't have a par model with this rep. Can still get
-naturality properties, however.
-
-2. For CEs can get par model and par properties. But what do these say
-for GADTs? Give examples?
-
-{\color{red} EVERYTHING BEFORE THE NOTES BELOW IS ENRICO'S TEXT:} To
+\begin{example}
+{\color{red} PRIMITIVE REP DOESN'T HAVE FIBRED SEMANTICS, AS REQUIRED
+  FOR PAR} To
 even consider the question of whether our semantics is parametric we
@@
 
-Following {\color{blue} Reference Computing left Kan extensions -
-  M.R.Bush M.Leeming R.F.C.Walters} we know that
-$(\mathsf{\Lan}_{\Eq_1} 1) R$ is the set of all morphisms of relations
-$f : \Eq_1 \to R$ modulo an equivalence relation; but there can be no
-morphism of relations $\Eq_1 \to R$ because its second component would
-have to be a function $1 \to \emptyset$, and there can be no such
-function.  Thus, $(\mathsf{\Lan}_{\Eq_1} 1) R = \emptyset$, which is
-not the domain we would have expected.
+Following~\cite{blw03} we know that $(\mathsf{\Lan}_{\Eq_1} 1) R$ is
+the set of all morphisms of relations $f : \Eq_1 \to R$ modulo an
+equivalence relation; but there can be no morphism of relations $\Eq_1
+\to R$ because its second component would have to be a function $1 \to
+\emptyset$, and there can be no such function.  Thus,
+$(\mathsf{\Lan}_{\Eq_1} 1) R = \emptyset$, which is not the domain we
+would have expected.
+\end{example}
 
@@
 
+We still get naturality properties, though.
+
 \vspace*{0.2in}
 
-%%% CE are non functorial but parametric
-As we have seen in the previous section {\color{blue} ref}, GADTs
+As we have seen in the previous Section~\ref{sec:reps}, GADTs
 represented as Church encodings do not necessarily have a map or a
@@
 paramentricity other than naturality, as they do admit parametric
-semantics {\color{blue} cite Bob's paper?}.
-
-%%% Example with G
+semantics~\cite{atk12}.
+
+\begin{example}
 {\color{blue} In this paragraph, $\mathsf{G}$ and $\mathsf{t}$
@@
   parametricity/IEL/AT uses serif font.} For example, consider the
-Church encoding of the GADT $\mathsf{G}$ defined
-in~\ref{eq:example-GADT}.  If $\mathsf{t}$ is a term of type
+Church encoding of the GADT $\mathsf{G}$ defined in
+Equation~\ref{eq:example-GADT}.  If $\mathsf{t}$ is a term of type
 $\mathsf{G}$ and $R$ is a relation on the sets $A$ and $B$, then
@@
 theorem we have that $t_0(A)$ and $t_0(B)$ are related in $G_1(R)$.
-We also know that $G_1(R)$ is the empty relation whenever $R$ is not
-$\mathit{Equal}\,1$.  We thus deduce that there can be no term of type
-$\mathsf{G\,a}$ unless $\mathsf{a}$ is semantically equivalent to
-$\mathsf{1}$.  This result is the syntactic counterpart of the
-inhabitation result for $\mathsf{G}$ {\color{blue} are inhabitation
-  results about semantics or syntax?}  showing that all of the
-instances of $\mathsf{G}$ are uninhabited, with the exception of the
-instance for $\mathsf{1}$.  Of course, we could not possibly have a
-term of an uninhabited type.
+{\color{blue} Here, $G = (G_0,G_1)$ is the interpretation of
+  $\mathsf{G}$.}  We also know that $G_1(R)$ is the empty relation
+whenever $R$ is not $\mathit{Equal}\,1$.  We thus deduce that there
+can be no term of type $\mathsf{G\,a}$ unless $\mathsf{a}$ is
+semantically equivalent to $\mathsf{1}$.  This result is the syntactic
+counterpart of the inhabitation result for $\mathsf{G}$ {\color{blue}
+  are inhabitation results about semantics or syntax?}  showing that
+all of the instances of $\mathsf{G}$ are uninhabited, with the
+exception of the instance for $\mathsf{1}$.  Of course, we could not
+possibly have a term of an uninhabited type.
+\end{example}
 
 %%% Primitive representation of GADT is non parametric, but functorial
-On the contrary, as we have shown previously {\color{blue} ref}, a
+On the contrary, as we have shown in Section~\ref{sec:gadts}, a
 GADT represented as a primitive data type does not necessarily admit a
@@
 naturality properties.  In other words, there is a suitable
-rearrange-transform property as explained in Section~\ref{sec:adts}.
+rearrange-transform property as explained in Section~\ref{sec:gadts}.
 
@@
 
-\vspace*{1in}
-
-NOTES/POINTS:
-
-Bob requires map functions for fixpoint but then doesn't use them.
-
+We can read our counterexample as showing that parametricity results
+for calculi representing GADTs as their Church encodings cannot be
+extended to calculi that represent GADTs as primitive data types. And
+functoriality results for calucli representing GADTs as primitive
+cannot be extended to calculi that represent GADTs as their Church
+encodings.  So the user is forced to make a choice between naturality
+consequences of parametricity and other consequences of
+parametricity. If GADTs really did generalize ADTs, and really were
+data types in a normal sense, then these would coincide and the user
+would not have to choose.
+
+\begin{comment}
+{\color{violet} This is already in the paper, split in multiple places.}
 An analysis of the Eq A B type in Bob's systems shows that the set
@@
 G 1.
-
-Since the types encoding GADTs in Bob's system make essential use
-of equality types, these considerations extend and possibly generalize
-to all GADTs.  Still, the equality is just a shorthand for a type made
-by higher-order forall and arrow types, and thus this considerations
-are not limited to a specific type, but are intrinsic of Bob's
-system.
-
-In contrast, $(Lan_1 1) B$, the type representing G in our system, is
-functorial (as are all types in our systems) and inhabited for every
-instance of B other than 0.  Indeed, we have a term $c : (Lan_1 1) 1$
-given by the Lan introduction rule, and thus every term t : T yields a
-term in $(Lan_1 1) T$ by use of the map of $(Lan_1 1) B$.  This shows
-that, as expected by now, the same GADT really does have different
-semantic interpretations in Bob's system and in ours.
-
-We can thus expect that not all types in Bob's system will
-be functorial and that, consequently, the consequences of
-parametricity will not include functoriality but will provide
-information about, e.g., inhabitatedness.
-%See
-%<https://www.cis.upenn.edu/~sweirich/talks/param-gadt.pdf> and
-%<https://www.seas.upenn.edu/~sweirich/papers/rtheorems.pdf> for more
-%about parametricity and GADTs.
-
-{\color{blue} Our favorite GADT isn't a hereditarily
-  polynomial nested type.}
-
-Importantly, we get the same parametricity properties for ADTs whether
-we regard them as fixpoints of higher-order functors --- i.e., as
-primitive data types --- or as their Church encodings. This is not the
-case for GADTs that are not nested types. (What about truly nested
-types?)
-
-Bob's system gives a parametric model that models GADTs. But given the
-non-uniformity captured by GADTs by purposeful design, that should be
-both completely unexpected and alarming! However, upon reflection it's
-not really all that surprising once we realize what the parametricity
-properties of GADTs in Bob's system look like.
-
-%It turns out that such properties are essentially trivial (because of
-%its essential use of Eq, which itself has a trivial parametricity
-%property --- IS THIS TRUE??). So, in the end, parametricity is not
-%really something I can use to reason non-trivially about GADTs in
-%Bob's system. (MAY NEED TO REVISE AFTER INVESTIGATION INTO PAR
-%PROPERTY OF EQ.)
-
-Tellingly, Bob gives no parametricity properties for GADTs. If he had,
-it would have illustrated the phenomenon we are describing, raised the
-question: What good is a par property for a GADT or another type using
-the Eq type?
-
+\end{comment}
+
+\vspace*{1in}
+
+NOTES:
+
+{\color{violet} TRUE????}
 FOR ADTs and Nesteds, it does not depend. We get the same par property
@@
 
-\section{Conclusion}
-
-{\color{blue} Some of this probably belongs in previous section.}
+\subsection{Comparison with Bob's Parametric Model}
 
@@
 consequences of parametricity are different.  That the consequences of
-parametricity depended on the model or the representation of GADTs
-would come as a surprise, though!  (IT DOES. CONSIDER INTERPRETING
-BASE TYPES AS ORDERED SETS. THIS GIVES DIFFERENT PARAMETRICITY
-PROPERTIES THAN INTERPRETING THEM AS DIAGONAL RELATIONS.)
+parametricity depended on the model or the representation of GADTs is
+to be expected, however. (IT IS. CONSIDER INTERPRETING BASE TYPES AS
+ORDERED SETS. THIS GIVES DIFFERENT PARAMETRICITY PROPERTIES THAN
+INTERPRETING THEM AS DIAGONAL RELATIONS.)
+
+It is possible to give a parametric model that models GADTs
+represented as Church encodings~\cite{atk12}.  But given the
+non-uniformity captured by GADTs by purposeful design, that should be
+both completely unexpected and alarming! However, upon reflection it's
+not really all that surprising once we realize what the parametricity
+properties of GADTs represented as Church encodings look like. Bob
+requires map functions for fixpoint but then doesn't use them.
+
+Tellingly, Bob gives no parametricity properties for GADTs. If he had,
+it would have illustrated the phenomenon we are describing, raised the
+question: What good is a par property for a GADT or another type using
+the Eq type?
 
@@
 
+{\color{violet} We say "naturality consequences of parametricity"
+when in fact we have naturality without having any parametricity at all.
+Change the wording to indicate that naturality follows from
+functoriality of the interp.}
 In our system we can show that we can recover the expected naturality
@@
 
-The upshot is that we can consider GADTs as data types (i.e., we can
-compute the semantics of GADTs) in two ways: just as determined by
-their syntax, as in Bob's system, or as the functorial completion of
-their syntax, obtained by supplying a suitable map primitive, as in
-ours. But, either way, GADTs fail to be well-behaved wrt
-parametricity. In the first case, GADTs don't have fmaps so their
-parametricity properties can't include naturality properties. In the
-second case, we still can't get non-naturality consequences of
-parametricity. That is, we can't get BOTH the naturality consequences
-of parametricity AND the non-naturality consequences of parametricity
-at the same time.
-
-One can read our counterexample as showing that Bob's parametricity
-results cannot be extended to one that gives a functorial semantics to
-GADTs. So the user is forced to make a choice between naturality
-consequences of parametricity and other consequences of
-parametricity. If GADTs really did generalize ADTs, and really were
-data types in a normal sense, then the user would not have to.
-
-\bibliography{gadt-story-references}
-
-\end{document}
+\pagebreak
+
+\begin{verbatim}
 
@@
 over the set interpretation)!
+\end{verbatim}
+
+\section{Conclusion}
+
+The upshot is that we can consider GADTs as data types (i.e., we can
+compute the semantics of GADTs) in two ways: just as determined by
+their syntax, as when they are represented by their Church encodings,
+or as the functorial completion of their syntax, as obtained by
+supplying a suitable map construct, when they are represented as
+primitive data types. But, either way, GADTs fail to be well-behaved
+wrt parametricity. In the first case, GADTs don't have fmaps so their
+parametricity properties can't include naturality properties. In the
+second case, we still can't get non-naturality consequences of
+parametricity. That is, we can't get BOTH the naturality consequences
+of parametricity AND the non-naturality consequences of parametricity
+at the same time.
+
+\bibliography{gadt-story-references}
+
+\end{document}

### 2021-02-20 20-gadts/GADT-story.tex
work on gadt-story

@@
 \begin{array}{l}
-\mathsf{data\, Expr\,a\,where}\\
-\mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Expr\,a}\\
-\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, ::\, Expr \,a \to Expr\,b \to
-  Expr\,(a \times b)}\\ 
+\mathsf{data\, Seq\,a\,where}\\
+\mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Seq\,a}\\
+\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, ::\, Seq \,a \to Seq\,b \to
+  Seq\,(a \times b)}\\ 
 \end{array}
@@
 \noindent
-comprises expressions of any type $\mathsf{a}$ and expressions
-obtained by pairing other expressions. By contrast, the primitive
-representation sees a GADT as the {\em functorial
+comprises sequences of any type $\mathsf{a}$ and sequences obtained by
+pairing the data in two already existing sequences. By contrast, the
+primitive representation sees a GADT as the {\em functorial
   completion}~\cite{jp19} of its syntax. Completing a GADT's syntax to
@@
 as their primitive representations. Viewed as its primitive
-representation, $\mathsf{Expr}$ comprises not just those data elements
+representation, $\mathsf{Seq}$ comprises not just those data elements
 representable in syntax, but also all data elements of the form
-$\mathsf{map\,f\,e}$ for all types $\mathsf{t_1}$ and $\mathsf{t_1}$,
+$\mathsf{map\,f\,s}$ for all types $\mathsf{t_1}$ and $\mathsf{t_2}$,
 all functions $\mathsf{f : t_1 \to t_2}$ definable in the language,
-and all $\mathsf{e :: Expr\,t_1}$, as well as all data elements of the
-form $\mathsf{map\,g\,e'}$ for each appropriately typed function
-$\mathsf{g}$ and each element $\mathsf{e'}$ already added to the data
-type, and so on. Functorial completion adds no new data elements to
-the syntax of a GADT that is actually a (polynomial) ADT or a
-(hereditarily polynomial) nested type, so its Church encoding and
-primitive representation coincide. However, these representations are
-rarely the same for GADTs.
-%Indeed, they are quite different for even the very simple GADT
-%$\mathsf{G}$ above.
+and all $\mathsf{s :: Seq\,t_1}$, as well as all data elements of the
+form $\mathsf{map\,g\,s'}$ for each appropriately typed function
+$\mathsf{g}$ and each element $\mathsf{s'}$ already added to the data
+type, and so on. Functorial completion for $\mathsf{Seq}$ adds, in
+particular, data elements of the form
+$\mathsf{map\,g\,(Pair\,u_1\,u_2)}$ even though these may not
+themselves be of the form $\mathsf{Pair\,v_1\,v_2}$ for any terms
+$\mathsf{v_1}$ and $\mathsf{v_2}$. By contrast, functorial completion
+adds no new data elements to the syntax of a GADT that is actually a
+(polynomial) ADT or a (hereditarily polynomial) nested type, so its
+Church encoding and primitive representation coincide. However, these
+representations are rarely the same for GADTs.
 
@@
 
-{\color{red} This will probably be more convincing if we use, e.g.,
-  an expression GADT from Stephanie above.}
 A GADT programmer is likely to use GADTs precisely because they
-exhibit different behaviors at different types, and thus to consider
-the GADT to be completely specified by its syntax. When used this way,
-the shape of a GADTs is actually {\em determined} by the data it
-contains. As a result, it is not expected to support a $\mathsf{map}$
-function --- i.e., to be an instance of Haskell's $\mathsf{Functor}$
-class --- like other data types. For example, the definition for
-$\mathsf{G}$ above specifies data only at instance $\mathsf{G\,1}$.
-Since there is a function from $\mathsf{1}$ to any other type
-$\mathsf{t}$, and yet $\mathsf{G\,t}$ is uninhabited at any instance
-other than $\mathsf{1}$, $\mathsf{G}$ cannot support a $\mathsf{map}$
-function. When it is intended to capture this kind of non-uniform
-behavior, a GADT's behavior does not generalize that of ADTs as
-advertised, and in some sense doesn't even behave like a traditional
-data type.  When used this way, a GADT cannot be seen as the least
-fixpoint of a functor, and must instead be represented via its Church
-encoding.
+exhibit different behaviors at different types, and thus to consider a
+GADT to be completely specified by its syntax. When used this way, the
+shape of a particular element of a GADTs is actually {\em determined}
+by the data it contains. As a result, it is not expected to support a
+$\mathsf{map}$ function --- i.e., to be an instance of Haskell's
+$\mathsf{Functor}$ class --- like other data types. Indeed, the
+definition for $\mathsf{Seq}$ above specifies that an element of the
+form $\mathsf{Pair\,u_1\,u_2}$ must have the shape of a sequence of
+data of pair type rather than a sequence of data of arbitrary type
+$\mathsf{e}$.  The clause of $\mathsf{map}$ for the $\mathsf{Pair}$
+constructor should therefore feed $\mathsf{map}$ a function $\mathsf{f
+  :: (a \times b) \to e}$ and a term of the form $\mathsf{Pair
+  \,u_1\,u_2}$ for $\mathsf{u_1 :: Seq\,a}$ and $\mathsf{_2 ::
+  Seq\,b}$, and produce a term $\mathsf{Pair\,v_1\,v_2'}$ for some
+appropriately typed terms $\mathsf{v_1}$ and $\mathsf{v_2}$. However,
+it is not clear how to achieve this since $\mathsf{e}$ need not
+necessarily be a product type. {\color{blue} And even if $\mathsf{e}$
+  {\em were} known to be of the form $\mathsf{w \times z}$, then we
+  still wouldn't necessarily have a way to produce data of type
+  $\mathsf{w \times z}$ from only $\mathsf{f :: a \times b \to w
+    \times z}$ and $\mathsf{u_1}$ and $\mathsf{u_2}$ unless we knew,
+  e.g., that $\mathsf{f}$ was a pair of functions.}
+
+{\color{red} HERE!}
+
+data only at instance
+$\mathsf{G\,1}$.  Since there is a function from $\mathsf{1}$ to any
+other type $\mathsf{t}$, and yet $\mathsf{G\,t}$ is uninhabited at any
+instance other than $\mathsf{1}$, $\mathsf{G}$ cannot support a
+$\mathsf{map}$ function. When it is intended to capture this kind of
+non-uniform behavior, a GADT's behavior does not generalize that of
+ADTs as advertised, and in some sense doesn't even behave like a
+traditional data type.  When used this way, a GADT cannot be seen as
+the least fixpoint of a functor, and must instead be represented via
+its Church encoding.
 
@@
 function either. This is already intuitively evident from the Haskell
-definition of $\mathsf{Seq}$, though. Indeed, the clause of
+definition of $\mathsf{Seq}$, though. ALREADY USED??? Indeed, the clause of
 $\mathsf{map}$ for the $\mathsf{Pair}$ constructor should feed

### 2021-02-21 20-gadts/GADT-story.tex
work on gadt-story

@@
 \newcommand{\cal}{\mathcal}
-%\newcommand{\fold}{\mathit{fold}}
 \newcommand{\F}{\mathcal{F}}
@@
 \section{Introduction}
+{\color{blue} map vs. fmap throughout}
 
@@
 the model. Models in which the Church encodings and primitive
-representations of all {\color{blue} (polynomial)} algebraic data
-types (ADTs) are semantically equivalent include the
-operationally-based model of~\cite{pit98,pit00} and the categorical
-models of~\cite{joh02,joh03}.
+representations of all algebraic data types (ADTs) are semantically
+equivalent include the operationally-based model of~\cite{pit98,pit00}
+and the categorical models of~\cite{joh02,joh03}.
 %\footnote{{\color{blue} Pitts proves the isomorphism only for
@@
 categorical model of~\cite{jgj21}. But, as shown in
-Section~\ref{sec:par} below, for some {\color{blue} (even polynomial)}
-GADTs,
-%that are not hereditarily polynomial nested types,
-no such isomorphism holds.
+Section~\ref{sec:par} below, for some GADTs --- even polynomial ones
+--- no such isomorphism holds.
 
@@
 the GADT $\mathsf{G}$ defined in Haskell by
-\begin{equation}\label{eq:example-GADT}
+\begin{equation}\label{eq:seq}
 \begin{array}{l}
@@
 \end{equation}
-%\[\begin{array}{l}
-%\mathsf{data\,G : * \to *\,where}\\
-%\mathsf{\;\;\;\;\;\;\;\;c : G\,1}
-%\end{array}\]
-%{\color{blue} Use $\top$ instead of $1$ throughout?}
 \noindent
@@
   :: (a \times b) \to e}$ and a term of the form $\mathsf{Pair
-  \,u_1\,u_2}$ for $\mathsf{u_1 :: Seq\,a}$ and $\mathsf{_2 ::
-  Seq\,b}$, and produce a term $\mathsf{Pair\,v_1\,v_2'}$ for some
+  \,u_1\,u_2}$ for $\mathsf{u_1 :: Seq\,a}$ and $\mathsf{u_2 ::
+  Seq\,b}$, and produce a term $\mathsf{Pair\,v_1\,v_2}$ for some
 appropriately typed terms $\mathsf{v_1}$ and $\mathsf{v_2}$. However,
 it is not clear how to achieve this since $\mathsf{e}$ need not
-necessarily be a product type. {\color{blue} And even if $\mathsf{e}$
-  {\em were} known to be of the form $\mathsf{w \times z}$, then we
-  still wouldn't necessarily have a way to produce data of type
-  $\mathsf{w \times z}$ from only $\mathsf{f :: a \times b \to w
-    \times z}$ and $\mathsf{u_1}$ and $\mathsf{u_2}$ unless we knew,
-  e.g., that $\mathsf{f}$ was a pair of functions.}
-
-{\color{red} HERE!}
-
-data only at instance
-$\mathsf{G\,1}$.  Since there is a function from $\mathsf{1}$ to any
-other type $\mathsf{t}$, and yet $\mathsf{G\,t}$ is uninhabited at any
-instance other than $\mathsf{1}$, $\mathsf{G}$ cannot support a
-$\mathsf{map}$ function. When it is intended to capture this kind of
-non-uniform behavior, a GADT's behavior does not generalize that of
-ADTs as advertised, and in some sense doesn't even behave like a
-traditional data type.  When used this way, a GADT cannot be seen as
-the least fixpoint of a functor, and must instead be represented via
-its Church encoding.
-
-A person interested in the semantics of functional languages, on the
-other hand, is likely to follow~\cite{bfss90} and expect GADTs to
-behave as their names suggests --- i.e., as generalizations of
-ADTs. The semanticist is thus likely to expect a GADT to be able to be
-used as a ADT is used --- i.e., as a shape filled with data that can
-be changed without also changing the shape. When used this way, it is
-more natural to view the GADT in terms of its primitive
-representation. In fact, the functorial completion of a GADT inherent
-in its primitive representation is exactly what is needed to ensure
-that the GADT supports a $\mathsf{map}$ function, i.e., is a functor
-both in the catgorical sense and in the sense of being an instance of
-Haskell's $\mathsf{Functor}$ class.  It is worth noting, however, that
-since the functorial completion of a GADT includes only those data
-elements that are derivable via repeated applications of
+necessarily be a product type. And even if $\mathsf{e}$ {\em were}
+known to be of the form $\mathsf{w \times z}$, then we still wouldn't
+necessarily have a way to produce data of type $\mathsf{w \times z}$
+from only $\mathsf{f :: a \times b \to w \times z}$ and $\mathsf{u_1}$
+and $\mathsf{u_2}$ unless we knew, e.g., that $\mathsf{f}$ was a pair
+of functions $\mathsf{(f_1 :: a \to w, f_2 :: b \to z)}$. When it is
+intended to capture this kind of non-uniform behavior, a GADT's
+behavior does not generalize that of ADTs as advertised, and the GADT
+doesn't even really behave like a traditional data type. In this
+situation, the GADT cannot be seen as the least fixpoint of a functor,
+and must instead be represented via its Church encoding.
+
+A computer scientist interested in the semantics of functional
+languages, on the other hand, is likely to follow~\cite{bfss90} and
+expect GADTs to behave as their names suggests --- i.e., as
+generalizations of ADTs. The semanticist is thus likely to expect a
+GADT to be able to be used as a ADT is used --- i.e., as a shape
+filled with data that can be changed without also changing the
+shape. To be used in this way, a GADT must be viewed in terms of its
+primitive representation. In fact, the functorial completion of a GADT
+inherent in its primitive representation is exactly what is needed to
+ensure that the GADT supports a $\mathsf{map}$ function, i.e., is
+interpreted as a true categorical functor, and can therefore be made
+an instance of Haskell's $\mathsf{Functor}$ class. It is worth noting,
+however, that since the functorial completion of a GADT includes only
+those data elements that are derivable via repeated applications of
 $\mathsf{map}$ starting from those specified by its syntax, primitive
 GADTs need not, in general, be nested types. For example, the
-functorial completion of our GADT $\mathsf{G}$ includes elements at
-any instance $\mathsf{G\,t}$ because there is a function from
-$\mathsf{1}$ to any type $\mathsf{t}$, but it does not include any
-terms at instance $\mathsf{G\,0}$ because no calculus allows the
-definition of a function into $\mathsf{0}$. {\color{blue} Use $\bot$
+functorial completion of the GADT $\mathsf{G}$ defined by
+\begin{equation}\label{eq:G}
+\begin{array}{l}
+\mathsf{data\,G\,a\,where}\\
+\mathsf{\;\;\;\;\;\;\;\;c : G\,1}
+\end{array}
+\end{equation}
+{\color{blue} Use $\top$ instead of $1$ throughout?} includes elements
+at any instance $\mathsf{G\,a}$ for any type $\mathsf{a} \not =
+\mathsf{0}$ since there is always a function from $\mathsf{1}$ to such
+an $\mathsf{a}$. But $\mathsf{G}$ does not include any terms at
+instance $\mathsf{G\,0}$ because $\mathsf{G\,0}$ is not inhabited via
+the syntactic specifiation of $\mathsf{G}$, and it is not inhabited
+via functorial completion because it is not possible to define a
+function from $\mathsf{1}$ into $\mathsf{0}$. {\color{blue} Use $\bot$
   instead of $0$ here?}
@@
 both the GADT programmer and the semanticist are both valid, the two
-are irreconcilable. Unsurprisingly, which point of view is adopted has
-deep consequences for the ways GADTs can be used and reasoned
-about. In this paper we show that the choice of GADT representation
-has deep implications for parametric reasoning about structures of
-those types. Indeed, a GADT programmer who views GADTs as their Church
-encodings cannot safely use program transformations or reasoning
-principles for them that involve map functions for them. This may
-surprise the GADT programmer by being contrary to expectation for
-something supposedly generalizing a data type. The GADT programmer
-may, however, be able to program with and reason about GADTs using
-non-naturality consequences of parametricity, such as type
-inhabitation results. On the other hand, a semanticist who views GADTs
-as primitive data types will have all naturality-based program
-transformations and reasoning principles for GADTs at their disposal,
-but non-naturality consequences of parametric reasoning may not be
-available to them.
-
-In this paper we show how these observations can be made precise. That
-is, we show that how we view GADTs has significant consequences, and
-thus answer the question we posed above in the affirmative: Yes, it
-really does matter.
+are irreconcilable. Importantly, which point of view is adopted in any
+particular setting has deep consequences for the ways GADTs can be
+used and reasoned about there. In this paper we show that the choice
+of GADT representation has deep implications for parametric reasoning
+about structures of those types. In particular, a GADT programmer who
+views GADTs as their Church encodings cannot safely use program
+transformations or reasoning principles for them that involve
+$\mathsf{map}$ functions for them, although they may be able to
+program with and reason about GADTs using other consequences of
+parametricity, such as type inhabitation results. On the other hand, a
+semanticist who views GADTs as primitive data types will have all
+naturality-based program transformations and reasoning principles for
+GADTs at their disposal, but, since no parametric model can be
+constructed for primitive GADTs, non-naturality consequences of
+parametric reasoning may not be available to them.
+
+The goal of this paper we show how these observations can be made
+precise, and therefore to show that how GADTs are viewed has
+significant consequences. This paper thus answers the question we
+posed above in the affirmative: Yes, it really does matter.
 
@@
 \[\mathsf{List\,a} = \mathsf{Nil} \,|\, \mathsf{Cons\,a\,(List\,a)}\]
-%of lists of elements of type $\mathsf{a}$
 is determined by the types of its two constructors $\mathsf{Nil\, ::\,
@@
 the type $\mathsf{List\,a}$ is therefore given by the set $\nat$ of
-natural numbers, and the positions in a structure of this type are
-given by specific natural numbers. Since the type argument to every
-occurrence of the type constructor $\mathsf{List}$ in the right-hand
-side of above definition is the same the type instance being defined
-on its left-hand side, the type $\mathsf{List\,a}$ enforces the
-invariant that all of the data in a structure of this type have the
-same type $\mathsf{a}$. In a similar way, the tree type
+natural numbers, with each natural number representing a choice of
+length for a list structure, and the positions in a structure of shape
+$\mathsf{n}$ are given by natural numbers ranging from $\mathsf{0}$ to
+$\mathsf{n-1}$. Since the type argument to every occurrence of the
+type constructor $\mathsf{List}$ in the right-hand side of above
+definition is the same the type instance being defined on its
+left-hand side, the type $\mathsf{List\,a}$ enforces the invariant
+that all of the data in a structure of this type have the same type
+$\mathsf{a}$. In a similar way, the tree type
 \[\mathsf{Tree\, a} = \mathsf{Leaf\,a}
-\,|\,\mathsf{Node\,(Tree\,a)\,a\,(Tree\,a)}\] of binary trees
-has as its underlying shape the type of binary trees of units, and the
+\,|\,\mathsf{Node\,(Tree\,a)\,a\,(Tree\,a)}\] of binary trees has as
+its underlying shape the type of binary trees of units, and the
 positions in a structure of this type are given by sequences of L (for
-``left'') and R (for ``right'') navigating a path
-%to either a node or a leaf in
-through the structure. The type $\mathsf{Tree\, a}$ enforces the
-invariant that all of the data at the nodes and leaves in a structure
-of this type have the same type $\mathsf{a}$.
-
-{\color{blue} polynomial ADTs?} Since the shape of an ADT structure
---- i.e., a structure whose type is an instance of an ADT --- is
-independent of the type of data it contains, ADTs can be defined
-polymorphically. As a result, an ADT structure containing data of type
-$\mathsf{a}$ can be transformed into another ADT structure of the
-exact same shape containing data of another type $\mathsf{b}$ simply
-by applying a given function $\mathsf{f \,:\,a \to b}$ to each of its
-elements. Indeed, every ADT can be made an instance of Haskell's
-$\mathsf{Functor}$ class by defining a type-and-data-uniform
-type-preserving, data-changing $\mathsf{fmap}$ {\color{blue} map
-  vs. fmap} function for it.  Moreover, given a type-independent way
-of rearranging an ADT structure's shape $\mathsf{T\,a}$ into the shape
-for another ADT structure $\mathsf{T'\,a}$, we get the same structure
-of type $\mathsf{T'\,b}$ regardless of whether we first rearrange the
-original structure of type $\mathsf{T\,a}$ into one of type
-$\mathsf{T'\,a}$ and then use $\mathsf{fmap_{T'}}$ to convert that
-resulting structure to one of type $\mathsf{T'\,b}$, or we first use
-$\mathsf{fmap_T}$ to convert the original structure of type
-$\mathsf{T\,a}$ to one of type $\mathsf{T\,b}$ and then rearrange that
-resulting structure into one of type $\mathsf{T'\,b}$. For example, if
-$\mathsf{f : a \to b}$, $\mathsf{xs \,:\,List\,a}$, and $\mathsf{g :
-  List\,a\to Tree\,a}$ arranges lists into trees in a type-independent
-way, then
+``left'') and R (for ``right'') navigating a path through the
+structure. The type $\mathsf{Tree\, a}$ enforces the invariant that
+all of the data at the nodes and leaves in a structure of this type
+have the same type $\mathsf{a}$.
+
+Since the shape of an ADT structure --- i.e., a structure whose type
+is an instance of an ADT --- is independent of the type of data it
+contains, ADTs can be defined polymorphically. As a result, an ADT
+structure containing data of type $\mathsf{a}$ can be transformed into
+another ADT structure of the exact same shape containing data of
+another type $\mathsf{b}$ simply by applying a given function
+$\mathsf{f \,:\,a \to b}$ to each of its elements. Indeed, every ADT
+$\mathsf{T}$ can be made an instance of Haskell's $\mathsf{Functor}$
+class by defining a type-and-data-uniform, structure-preserving,
+data-changing $\mathsf{fmap_T}$ function for it.  Moreover, given a
+type-independent way of rearranging an ADT structure's shape
+$\mathsf{T\,a}$ into the shape for another ADT structure
+$\mathsf{T'\,a}$, we get the same structure of type $\mathsf{T'\,b}$
+regardless of whether we first rearrange the original structure of
+type $\mathsf{T\,a}$ into one of type $\mathsf{T'\,a}$ and then use
+$\mathsf{fmap_{T'}}$ to convert that resulting structure to one of
+type $\mathsf{T'\,b}$, or we first use $\mathsf{fmap_T}$ to convert
+the original structure of type $\mathsf{T\,a}$ to one of type
+$\mathsf{T\,b}$ and then rearrange that resulting structure into one
+of type $\mathsf{T'\,b}$. For example, if $\mathsf{f : a \to b}$,
+$\mathsf{xs \,:\,List\,a}$, and $\mathsf{g : List\,a\to Tree\,a}$
+arranges lists into trees in a type-independent way, then
 \[\mathsf{fmap_{Tree}\,f\,(g \, xs) \,=\, g\,(fmap_{List}\,f\,xs)}\]
-{\color{violet} mention naturality?}
 
@@
 One way to represent ADTs is via their Church encodings. A Church
-encoding is a representation of a data type as a pure function in a
-lambda calculus (System F in the case of functional languages). Church
-encodings can be used to represent data types in languages that do not
-support primitives for sum types, product types, or recursion.
-{\color{blue} Church encodings (and the related Parigot encodings)
-  have recently been popularized as {\em visitor patterns} in
-  object-oriented programming.} Represention of ADTs as their Church
-encodings is possible in any language that supports functions.
-
-The Church encodings of the ADTs $\mathsf{List\,a}$ and
+encoding is a representation of a data type as a function in a pure
+lambda calculus, such as System F and its higher-kinded
+extensions. They, together with other related encodings, have recently
+been popularized as various {\em visitor patterns} in object-oriented
+programming~\cite{owg08,gon21}.
+
+Church encodings of ADTs can be defined in any language that supports
+functions. They can therefore be used to represent ADTs in languages
+that do not support primitives for sum types, product types, or
+recursion. The Church encodings of the ADTs $\mathsf{List\,a}$ and
 $\mathsf{Tree\,a}$, for example, are
@@
 \[\mathsf{Tree\,a} = \mathsf{\forall b.\,(a \to b) \to (b \to a \to b
-  \to b) \to b}\] respectively. The argument types are abstractions
-of the types of ADTs' constructors. For instance, $\mathsf{b}$
-abstracts the type of the constructor $\mathsf{Nil}$, and $\mathsf{a
-  \to b \to b}$ abstracts the type of the constructor $\mathsf{Cons}$,
-for lists. Because these types are each uniform in their argument type
-$\mathsf{a}$, we can also encode the type constructors $\mathsf{List}$
-and $\mathsf{Tree}$ themselves as
+  \to b) \to b}\] respectively. The argument types are abstractions of
+the types of ADTs' constructors. For instance, $\mathsf{b}$ abstracts
+the type $\mathsf{List\,a}$ of the constructor $\mathsf{Nil}$ for
+lists, and $\mathsf{a \to b \to b}$ abstracts the type $\mathsf{a \to
+  List\,a \to List\,a}$ of the constructor $\mathsf{Cons}$.
+
+Because the types of the ``abstract constructors'' for an ADT are
+uniform in their argument types, it is always possible to Church
+encode the type constructors themselves as well. For example, the
+Church encodings of the type constructors $\mathsf{List}$ and
+$\mathsf{Tree}$ are
 \[\mathsf{List} = \mathsf{\forall a. \forall b.\,b \to (a \to b \to b)
@@
 \[\mathsf{Tree} = \mathsf{\forall a. \forall b.\,(a \to b) \to (b \to
-  a \to b \to b) \to b}\]
-respectively.
-
-It is possible to make $\mathsf{List}$ and $\mathsf{Tree}$ instances
-of Haskell's $\mathsf{Functor}$ class. Indeed, the Prelude's
-$\mathsf{map}$ function does exactly that for $\mathsf{List}$. For
-$\mathsf{Tree}$ we can define
+  a \to b \to b) \to b}\] respectively. This is what allows an ADT's
+associated type constructor to be made an instance of Haskell's
+$\mathsf{Functor}$ class. For example, Haskell's built-in
+$\mathsf{map}$ function makes the type constructor $\mathsf{List}$ an
+instance of the $\mathsf{Functor}$ class, and for $\mathsf{Tree}$ we
+can define
 \[\begin{array}{lll}
@@
 \end{array}\]
-But the Church encoding of an ADT carries with it no expectation
-whatsoever that such a type-and-data-uniform, type-preserving,
-data-changing $\mathsf{map}$ function can be defined. Nevertheless,
-for polynomial ADTs this can always be done. But even when the Church
-encoding of an ADT supports function with the same type as its actual
-$\mathsf{map}$, there is no guarantee that the rearrange-transform
-property will hold for it. For example, if {\color{violet} Are there
-  examples of such non-functorial maps?}
+Note, however, that the Church encoding of an ADT carries with it no
+expectation whatsoever that such a type-and-data-uniform,
+structure-preserving, data-changing $\mathsf{map}$ function can be
+defined.
 
@@
 are
-\[\mathsf{List\,a} = \mathsf{\mu b.\, 1 + (a \times b)}\] 
+\begin{equation}\label{eq:list}
+  \mathsf{List\,a} = \mathsf{\mu X.\, 1 + a \times X}
+\end{equation}
 and 
-\[\mathsf{Tree\,a} = \mathsf{\mu b.\, a + (b \times a \times b)}\]
-respectively. Here, $\mathsf{\mu}$ is the primitive fixpoint
-operator.
+\begin{equation}\label{eq:tree}
+  \mathsf{Tree\,a} = \mathsf{\mu X.\, a + X \times a \times X}
+\end{equation}
+respectively, where $\mathsf{\mu}$ is a primitive fixpoint operator.
 
@@
 considered as fixpoints. For example $\mathsf{List\,a}$ does indeed
-satisfy the fixpoint equation indicated above, i.e., is indeed such
-that
-\[\mathsf{List\,a} = \mathsf{1 + (a \times List\,a)}\] 
+satisfy the fixpoint equation $\mathsf{F\,X} = \mathsf{1 + a \times
+  X}$ determined by~\eqref{eq:list}, i.e., is indeed such that
+\[\mathsf{List\,a} = \mathsf{1 + a \times List\,a}\] 
 since every element of $\mathsf{List\,a}$ is either empty or is
 obtained by $\mathsf{Cons}$ing an element of type $\mathsf{a}$ onto an
-already-existing structure of type $\mathsf{List\,a}$. (The fixpoint
-equation is just a rewriting of the Haskell data type declaration for
-$\mathsf{List\,a}$.) The ADT $\mathsf{Tree\,a}$ can similarly be seen
-to be a solution to the fixpoint equation
-\[\mathsf{Tree\,a} = \mathsf{a + Tree\,a \times a \times Tree\,a}\] 
-determined by its Haskell data type declaration.
-
-These fixpoint equations are sensible at the level of types. But to
-ensure that the syntactic fixpoint actually denotes a semantic object
-computed as a real-live fixpoint, the semantic fixpoint calculation
-must converge. This means that the fixpoint being taken must be of
-{\em functors} on the category interpreting types~\cite{tfca}, rather
-than of mere functions between the sets interpreting them. Said
-differently, functions between sets interpreting types must have
-corresponding {\em functorial actions} on functions between
-interpretations of types. Reflecting this requirement back into syntax
-gives that the body of a primitive data type representation of an ADT
-must support a $\mathsf{map}$ function, i.e., must be an instance of
-Haskell's $\mathsf{Functor}$ class.
+already-existing structure of type $\mathsf{List\,a}$. In fact, this
+fixpoint equation is just a rewriting of the Haskell data type
+declaration for $\mathsf{List\,a}$. Similarly, the ADT
+$\mathsf{Tree\,a}$ can be seen to be a solution to the fixpoint
+equation
+\[\mathsf{F\,X} = \mathsf{a + X \times a \times X}\] 
+determined by~\eqref{eq:tree}.
+%its Haskell data type declaration.
+
+These fixpoint equations are entirely sensible at the level of
+types. But to ensure that the syntactic fixpoint representing an ADT
+actually denotes a semantic object computed as a real-live fixpoint,
+the semantic fixpoint calculation must converge. This means that the
+fixpoint being taken must be of {\em functors} on the category
+interpreting types~\cite{tfca}, rather than of mere functions between
+the sets interpreting types. Said differently, the function $F$
+interpreting the type constructor $\mathsf{F}$ constructing the body
+of a syntactic fixpoint must actually be a {\em functor} on the
+category $\mathit{Set}$ of sets and functions. That is $F$ must not
+only have an action on sets interpreting different type arguments to
+$\mathsf{F}$, but must also have a {\em functorial action} on
+functions between such sets. Reflecting this requirement back into
+syntax gives that $\mathsf{F}$ must support a $\mathsf{map}$ function,
+i.e., must be an instance of Haskell's $\mathsf{Functor}$ class.
+
+Requiring $F$ to be a functor will ensure that the interpretation $\mu
+F$ of the ADT $\mathsf{T\,a} = \mathsf{\mu F}$ exists. But to ensure
+that $\mu F$ is itself a functor, so that the type constructor
+$\mathsf{T}$ associated with the ADT $\mathsf{T\,a}$ supports a
+$\mathsf{map}$ function, $F$ must itself be the fixpoint $\mu H$ of a
+functor $H$. But the functor $H$ must be a functor on the category of
+functors on the category $\mathit{Set}$ interpreting types. In other
+words, $H$ must be a {\em higher-order} functor on
+$\mathit{Set}$. Reflecting this requirement back into syntax gives
+that $\mathsf{T\,a} = \mathsf{(\mu F)\,a} = \mathsf{(\mu\,(\mu
+  H))\,a}$ for the ``type constructor constructor'' $\mathsf{H}$ whose
+interpretation is $H$.
+
+{\color{red} HERE!!!!}
+
+\vspace*{1in}
 
@@
 associated with an ADT's representation as a primitive data type.
+
+{\color{violet} Mention naturality? Semantics? Uniformity because we
+  actually view ADTs as fixpoints of higher-order functors --- not
+  first-order ones --- to get maps.}
 
@@
 representation of data types, by considering the previously defined
-type $G$ in Equation~\ref{eq:example-GADT}.  Recall that the semantic
+type $G$ in Equation~\ref{eq:G}.  Recall that the semantic
 interpretation of $G$ is given by $\mathsf{\Lan}_1 1$ which, in fact,
@@
 Church encoding of the GADT $\mathsf{G}$ defined in
-Equation~\ref{eq:example-GADT}.  If $\mathsf{t}$ is a term of type
+Equation~\ref{eq:G}.  If $\mathsf{t}$ is a term of type
 $\mathsf{G}$ and $R$ is a relation on the sets $A$ and $B$, then

### 2021-02-22 20-gadts/GADT-story.tex
work on gadt-story

@@
 
-\section{Representations of Data Types}\label{sec:reps}
-
-\subsection{Algebraic Data Types}\label{sec:adts}
+\section{Representations of Algebraic Data Types}\label{sec:adts}
 
@@
 
-\subsubsection{Church Encodings of ADTs}
+\subsection{Church Encodings of ADTs}
 
@@
 
-\subsubsection{ADTs as Primitive Data Types}
+\subsection{ADTs as Primitive Data Types}
 
@@
 Primitive representations capture in syntax the fact that ADTs can be
-considered as fixpoints. For example $\mathsf{List\,a}$ does indeed
-satisfy the fixpoint equation $\mathsf{F\,X} = \mathsf{1 + a \times
-  X}$ determined by~\eqref{eq:list}, i.e., is indeed such that
+considered as fixpoints. For example $\mathsf{List\,a}$ is indeed a
+fixpoint of $\mathsf{F_{List\,a}}$, where $\mathsf{F_{List\,a}\,X} =
+\mathsf{1 + a \times X}$ as determined by~\eqref{eq:list}, i.e., is
+indeed such that
 \[\mathsf{List\,a} = \mathsf{1 + a \times List\,a}\] 
@@
 fixpoint equation is just a rewriting of the Haskell data type
-declaration for $\mathsf{List\,a}$. Similarly, the ADT
-$\mathsf{Tree\,a}$ can be seen to be a solution to the fixpoint
-equation
-\[\mathsf{F\,X} = \mathsf{a + X \times a \times X}\] 
-determined by~\eqref{eq:tree}.
-%its Haskell data type declaration.
-
-These fixpoint equations are entirely sensible at the level of
-types. But to ensure that the syntactic fixpoint representing an ADT
-actually denotes a semantic object computed as a real-live fixpoint,
-the semantic fixpoint calculation must converge. This means that the
-fixpoint being taken must be of {\em functors} on the category
-interpreting types~\cite{tfca}, rather than of mere functions between
-the sets interpreting types. Said differently, the function $F$
-interpreting the type constructor $\mathsf{F}$ constructing the body
-of a syntactic fixpoint must actually be a {\em functor} on the
-category $\mathit{Set}$ of sets and functions. That is $F$ must not
-only have an action on sets interpreting different type arguments to
-$\mathsf{F}$, but must also have a {\em functorial action} on
-functions between such sets. Reflecting this requirement back into
-syntax gives that $\mathsf{F}$ must support a $\mathsf{map}$ function,
-i.e., must be an instance of Haskell's $\mathsf{Functor}$ class.
+declaration for $\mathsf{List\,a}$. We therefore have that
+$\mathsf{List\, a} = \mathsf{\mu F_{List\,a}}$ is modelled by $\mu
+F_{\mathit{List}\,a}$, where the functor $F_{\mathit{List}\,a}$ models
+the type constructor $\mathsf{F_{List\,a}}$. Similarly, the ADT
+$\mathsf{Tree\,a}$ can be seen to be a fixpoint of
+$\mathsf{F_{Tree\,a}}$, where $\mathsf{F_{Tree\,a}\,X} = \mathsf{a + X
+  \times a \times X}$ as determined by~\eqref{eq:tree}, so that
+$\mathsf{Tree\, a} = \mathsf{ \mu F_{Tree\,a}}$ is modelled by $\mu
+F_{\mathit{Tree}\,a}$, where $F_{\mathit{Tree}\,a} $ models
+$\mathsf{F_{Tree\,a}}$.
+
+These kinds of fixpoint equations are entirely sensible at the level
+of types. But to ensure that the syntactic fixpoint representing an
+ADT actually denotes a semantic object computed as a real-live
+fixpoint, the semantic fixpoint calculation must converge.  If we
+interpret our types as sets, then this means that the fixpoint being
+taken must be of a {\em functor} on the category $\mathit{Set}$ of
+sets and functions between them, rather than of a mere function
+between sets~\cite{tfca}.  That is, the function $F$ interpreting the
+type constructor $\mathsf{F}$ constructing the body of a syntactic
+fixpoint must not only have an action on sets, but must also have a
+{\em functorial action} on functions between sets. Reflecting this
+requirement back into syntax gives that $\mathsf{F}$ must support a
+suitable\footnote{The $\mathsf{map}$ function for $\mathsf{F}$ should
+  satisfy syntactic reflections of the functoriality properties ---
+  i.e., should preserve identities and compositions --- in
+  $\mathit{Set}$ even though there is no mechanism in Haskell for
+  enforcing this.} $\mathsf{map}$ function, i.e., must be an instance
+of Haskell's $\mathsf{Functor}$ class.
 
@@
 $\mathsf{T}$ associated with the ADT $\mathsf{T\,a}$ supports a
-$\mathsf{map}$ function, $F$ must itself be the fixpoint $\mu H$ of a
-functor $H$. But the functor $H$ must be a functor on the category of
-functors on the category $\mathit{Set}$ interpreting types. In other
-words, $H$ must be a {\em higher-order} functor on
-$\mathit{Set}$. Reflecting this requirement back into syntax gives
-that $\mathsf{T\,a} = \mathsf{(\mu F)\,a} = \mathsf{(\mu\,(\mu
-  H))\,a}$ for the ``type constructor constructor'' $\mathsf{H}$ whose
-interpretation is $H$.
-
-{\color{red} HERE!!!!}
-
-\vspace*{1in}
+$\mathsf{map}$ function as well, $F$ must again be the fixpoint $\mu
+H$ of a functor $H$. In this case, however, $H$ must be a functor on
+the category $\mathit{Func(Set)}$ of functors and natural
+transformations on $\mathit{Set}$, i.e., $H$ must be a {\em
+  higher-order} functor on $\mathit{Set}$. Reflecting this requirement
+back into syntax gives that $\mathsf{T\,a} = \mathsf{(\mu F)\,a} =
+\mathsf{(\mu\,(\mu H))\,a}$ for the ``type constructor constructor''
+$\mathsf{H}$ that supports suitable\footnote{{\color{blue} The
+    $\mathsf{map}$ function for $\mathsf{H}$ should satisfy syntactic
+    reflections of the functoriality properties in
+    $\mathit{Func(Set)}$ --- i.e., should preserve identity natural
+    transformations and composition of natural transformations --- and
+    the $\mathsf{map}$ function for $\mathsf{H\,F}$ should satisfy
+    syntactic reflections of the functoriality properties in
+    $\mathit{Set}$ as in the previous footnote. Again, there is no
+    mechanism in Haskell for enforcing this.}} $\mathsf{map}$
+functions.  {\color{blue} Mention naturality?}
+
+%class HFunctor f where
+%  ffmap :: Functor g => (a -> b) -> f g a -> f g b
+%  hfmap :: (g :~> h) -> f g :~> f h
+%
+%  type (~>) f g = forall x. f x -> g x
 
 A concrete example is given by the ADT $\mathsf{List\,a}$. This type
-is is modelled as the fixpoint $\mu F_a$ of the first-order functor
-whose action on objects is given by $F_a\,X = 1 + a \times X$ and,
-whose action on morphisms (which models its corresponding
-$\mathsf{map}$ function) is given by $F_a\,f = 1 + \id_a \times
-f$. The type constructor $\mathsf{List}$ is then modelled by the
-higher-order functor $H$ whose action on objects is given by $H\,F\,X
-= 1 + X \times F\,X$ and whose action on morphisms is given by
-$H\,F\,f = \id_1 + f \times F\,f$. (A {\em higher-order functor} in
-this context is a functor not directly on the category interpreting
-types, but rather on the category of functors interpreting type
-constructors, i.e., on the category of functors on the category of
-functors interpreting types.) The ADT $\mathsf{List\,a}$ is then
-modelled as the functor $(\mu H)a$. Since $H$ is a higher-order
-functor, not only does it have a corresponding functorial action,
-i.e., support a function $\mathsf{map_H}$, but $\mu H$ is also a
-functor with its own functorial action. That is, the type constructor
-$\mathsf{List}$ itself supports a function $\mathsf{fmap_{List}}$.
-
-{\color{blue} At the level of objects/types, short cut fusion can be
-  used to show that the Church encoding and the primitive
-  representation of an ADT are isomorphic.} Note, however, that
-whereas it is impossible even to {\em state} the rearrange-transform
-property for ADTs whose Church encodings do not have corresponding
-$\mathsf{map}$ functions, and although this property is guaranteed to
-hold even when it can be stated for Church encodings, it is simply the
-reflection back into syntax of the naturality property of the
-functorial action of the interpretation of the type constructor
-associated with an ADT's representation as a primitive data type.
-
-{\color{violet} Mention naturality? Semantics? Uniformity because we
-  actually view ADTs as fixpoints of higher-order functors --- not
-  first-order ones --- to get maps.}
-
-%In the former case the ADT $\mathsf{List\,a}$ for each $\mathsf{a}$ is
-%modelled as the fixpoint of its underlying functor $\mu F_a$; by
-%contrast, in the latter case it is the type constructor
-%$\mathsf{List}$ itself that is modelled as $\mu H$, and the ADT
-%$\mathsf{List\,a}$ is actually modelled as $(\mu H)\,a$. That is, the
-%ADTs $\mathsf{List\,a}$ for the various $\mathsf{a}$s are all modelled
-%uniformly. (The extreme uniformity comes from the fact that the
-%{\color{blue} type/second} argument $X$ is the same in each recursive
-%call to $H$. This requirement will be lifted in
-%Section~\ref{sec:gadts} below.)  This is precisely what makes it
-%possible to define am $\mathsf{fmap}$ for list types.
-
-%A crucial point is that while the $\mu F_a$ is not required to be a
-%functor, and thus is not required to have a {\em functorial action} on
-%morphisms --- i.e., to support an $\mathsf{fmap_{F_a}}$ function in
-%programming terms --- the functor $\mu H$ modelling $\mathsf{List}$
-%must do so. Indeed, since $H$ is a higher-order functor, $\mu H$ is
-%itself a functor. Applying the functorial action $\mathsf{fmap_{\mu
-%    H}}$ of this functor to a morphism $\mathsf{f}$ yields a natural
-%transformation $\mathsf{fmap_{\mu H}\,f}$. This in turn guarantees the
-%order-independence of type-independent rearrangement and conversion
-%via $\mathsf{fmap_{\mu H}\,f}$ for lists {\color{blue} and trees}.
-
-%{\color{blue} Extension of a container}
-
-\begin{comment}
-{\color{blue} Type-independent reasoning about ADTs in functional
-  languages is usually justified by first representing ADTs by their
-  Church encodings, and then reasoning type-independently about these
-  encodings. This is typically justified by constructing a parametric
-  model --- i.e, a model in which polymorphic functions preserve
-  relations {\em \'a la} Reynolds --- for a suitable fragment of
-  System F, demonstrating that an initial algebra exists for the
-  positive type constructor corresponding to the functor underlying an
-  ADT of interest, and showing that each such initial algebra is
-  suitably isomorphic to its corresponding Church encoding. In fact,
-  this isomorphism of initial algebras and their Church encodings is
-  one of the ``litmus tests'' for the goodness of a parametric model.}
-
-This approach works well for ADTs, which are always fixpoints of {\em
-  first-order} functors, and whose Church encodings, which involve
-quantification over only type variables, are always expressible in
-System F. For example, $\mathtt{List\,A}$ is the fixpoint of the
-first-order functor $F\,X = 1 + A \times X$ and has Church encoding
-$\forall \alpha. \, \alpha \to (\mathtt{A} \to \alpha \to \alpha) \to
-\alpha$. But despite Cardelli's~\cite{car97} claim that ``virtually
-any basic type of interest can be encoded within F$_2$'' --- i.e.,
-within System F --- non-ADT nested types cannot.  Not even our
-prototypical nested type of perfect trees has a Church encoding
-expressible in System F!  Indeed, $\mathtt{PTree\,A}$ cannot be
-represented as the fixpoint of any {\em first-order} functor. However,
-it can be seen as the instance at index $\mathtt{A}$ of the fixpoint
-of the {\em higher-order} functor $H\,F\,A\,=\, (A \to F\,A) \to (F
-\,(A \times A) \to F\,A) \to F\,A$. It thus has Church encoding
-$\forall f.\, (\forall \alpha.\,\alpha \to f\alpha) \to (\forall
-\alpha. \,f (\alpha \times \alpha) \to f\alpha) \to \forall \alpha.\,
-f\alpha$, which requires quantification at the higher kind $* \to *$
-for $f$. A similar situation obtains for any (non-ADT) nested
-type. Unfortunately, higher-kinded quantification is not available in
-System F, so if we want to reason type-independently about nested
-types in a language based on it we have only two options: {\em
-  i})\,move to an extension of System F, such as the higher-kinded
-calculus F$_\omega$ or a dependent type theory, and reason via their
-Church encodings in a known parametric model for that extension, or
-{\em ii})\, add nested types to System F as primitives --- i.e., as
-primitive type-level fixpoints --- and construct a parametric model
-for the result.
-
-Since the type systems of F$_\omega$ and dependent type theories are
-designed to extend System F with far more than non-ADT data types, it
-seems like serious overkill to pass to their parametric models to
-reason about nested types in System F. Indeed, such calculi support
-fundamentally new features that add complexity to their models that is
-entirely unnecessary for reasoning about nested types. This paper
-therefore pursues the second option above.  We first design a
-Hindley-Milner-style calculus supporting primitive nested types,
-together with primitive types of natural transformations representing
-morphisms between them. Our calculus can express all nested types
-appearing in the literature, including truly nested types.  At the
-term-level, it supports primitive pattern matching, map functions, and
-fold combinators for nested types.\footnote{We leave incorporating
-  general term-level recursion to future work because, as
-  Pitts~\cite{pit00} reminds us, ``it is hard to construct models of
-  both impredicative polymorphism and fixpoint recursion''. In fact,
-  as the development in this paper shows, constructing a parametric
-  model even for our predicative calculus with primitive nested types
-  --- and even without term-level fixpoints --- is already rather
-  involved.  On the other hand, our calculus is strongly normalizing,
-  so it perhaps edges us toward the kind of provably total practical
-  programming language proposed in~\cite{wad89}.}  Our main
-contribution is the construction of a parametric model for our
-calculus. This is both delicate and challenging. To ensure the
-existence of semantic fixpoints interpreting nested types, and thus to
-establish a suitable Identity Extension Lemma, our type system must
-explicitly track functoriality of types, and cocontinuity conditions
-on the functors interpreting them must be appropriately threaded
-throughout the model construction. Our model validates all standard
-consequences of parametricity in the presence of primitive nested
-types, including the isomorphism of primitive ADTs and their Church
-encodings, and correctness of short cut fusion for nested types. The
-relationship between naturality and parametricity has long been of
-interest, and our inclusion of a primitive type of natural
-transformations allows us to clearly delineate those consequences of
-parametricity that follow from naturality, from those, such as short
-cut fusion for nested types, that require the full power of
-parametricity.
-\end{comment}
-
-\subsection{GADT Syntax}\label{sec:gadts}
-
-{\color{blue} polynomial} GADTs relax the restriction on the type
-instances appearing in a data type definition. The special form of
-GADTs known as {\em nested types} allow the data constructors of a
-GADT to take as arguments data whose types involve type instances of
-the GADT other than the one being defined. For example, the 
-definition 
+is is modelled as the fixpoint $\mu F_{\mathit{List\,a}}$ of the
+first-order functor whose action on sets is given by
+$F_{\mathit{List}\,a}\,X = 1 + a \times X$ and whose action on
+functions is given by $F_{\mathit{List}\,a}\,f = 1 + \id_a \times
+f$. The type constructor $\mathsf{List}$ is modelled by the
+higher-order functor $H$ whose action on a functor $F$ is given by the
+functor $H\,F$ whose actions on sets and functions between them are
+given by $H\,F\,X = 1 + X \times F\,X$ and $H\,F\,f = \id_1 + f \times
+F\,f$, respectively, and whose action on a natural transformation
+$\eta$ is the natural transformation given by $H\,\eta = \id_1 + \eta
+\times F\,\eta$. The type constructor $\mathsf{List}$ is then
+modelled as the functor $\mu H$. Reflecting the functorial action of
+$\mu H$ back into syntax gives Haskell's built-in $\mathsf{map}$
+function as the type-and-data-uniform, structure-preserving,
+data-changing function associated with $\mathsf{List}$. {\color{blue}
+  HOW?}
+
+In a parametric model, short cut fusion can be used to show that the
+Church encoding and the primitive representation of an ADT are
+isomorphic as sets. Note, however, that whereas it is impossible even
+to {\em state} the rearrange-transform property for Church encodings
+of ADTs, it is simply the reflection back into syntax of the
+{\color{blue} naturality property - right?} of the functorial action
+of the interpretation of the type constructor associated with an ADT's
+primitive representation.
+
+\section{Representations of GADTs}\label{sec:gadts}
+
+{\color{red} HERE!}
+
+GADTs relax the restriction on the type instances appearing in a data
+type definition. The special form of GADTs known as {\em nested types}
+allow the data constructors of a GADT to take as arguments data whose
+types involve type instances of the GADT other than the one being
+defined. For example, the definition
 \[\mathsf{PTree \, a = PLeaf \,a \,|\, PNode \,(PTree\,(a \times
@@
 
-\subsubsection{Church Encodings of GADTs}\label{sec:CEs}
+\subsection{Church Encodings of GADTs}\label{sec:CEs}
 
@@
 
-\subsubsection{GADTs as Primitive Data Types}\label{sec:prim-gadts}
+\subsection{GADTs as Primitive Data Types}\label{sec:prim-gadts}
 
@@
 
-As we have seen in the previous Section~\ref{sec:reps}, GADTs
+As we have seen in the previous Section~\ref{sec:adts}, GADTs
 represented as Church encodings do not necessarily have a map or a

### 2021-02-23 20-gadts/GADT-story.tex
work on gadt-story

@@
 suitable\footnote{The $\mathsf{map}$ function for $\mathsf{F}$ should
-  satisfy syntactic reflections of the functoriality properties ---
-  i.e., should preserve identities and compositions --- in
-  $\mathit{Set}$ even though there is no mechanism in Haskell for
-  enforcing this.} $\mathsf{map}$ function, i.e., must be an instance
-of Haskell's $\mathsf{Functor}$ class.
+  satisfy syntactic reflections of the functoriality properties in
+  $\mathit{Set}$ --- i.e., preservation of identity functions and
+  composition of functions --- even though there is no mechanism in
+  Haskell for enforcing this.} $\mathsf{map}$ function, i.e., must be
+an instance of Haskell's $\mathsf{Functor}$ class.
 
@@
 $\mathsf{T}$ associated with the ADT $\mathsf{T\,a}$ supports a
-$\mathsf{map}$ function as well, $F$ must again be the fixpoint $\mu
-H$ of a functor $H$. In this case, however, $H$ must be a functor on
-the category $\mathit{Func(Set)}$ of functors and natural
-transformations on $\mathit{Set}$, i.e., $H$ must be a {\em
+$\mathsf{map}$ function as well, we can require that $F$ also be the
+fixpoint $\mu H$ of a functor $H$. In this case, however, $H$ must be
+a functor on the category $\mathit{Set}^\mathit{Set}$ of functors and
+natural transformations on $\mathit{Set}$. That is, $H$ must be a {\em
   higher-order} functor on $\mathit{Set}$. Reflecting this requirement
@@
 \mathsf{(\mu\,(\mu H))\,a}$ for the ``type constructor constructor''
-$\mathsf{H}$ that supports suitable\footnote{{\color{blue} The
-    $\mathsf{map}$ function for $\mathsf{H}$ should satisfy syntactic
-    reflections of the functoriality properties in
-    $\mathit{Func(Set)}$ --- i.e., should preserve identity natural
-    transformations and composition of natural transformations --- and
-    the $\mathsf{map}$ function for $\mathsf{H\,F}$ should satisfy
-    syntactic reflections of the functoriality properties in
-    $\mathit{Set}$ as in the previous footnote. Again, there is no
-    mechanism in Haskell for enforcing this.}} $\mathsf{map}$
-functions.  {\color{blue} Mention naturality?}
+$\mathsf{H}$ that supports suitable\footnote{The $\mathsf{map}$
+  function for $\mathsf{H}$ should satisfy syntactic reflections of
+  the functoriality properties in $\mathit{Set}^\mathit{Set}$ ---
+  i.e., preservation of identity natural transformations and
+  composition of natural transformations --- and the $\mathsf{map}$
+  function for $\mathsf{H\,F}$ should satisfy syntactic reflections of
+  the functoriality properties in $\mathit{Set}$ as in the previous
+  footnote. Again, there is no mechanism in Haskell for enforcing
+  this.} $\mathsf{map}$ functions.
 
@@
 A concrete example is given by the ADT $\mathsf{List\,a}$. This type
-is is modelled as the fixpoint $\mu F_{\mathit{List\,a}}$ of the
+is modelled as the fixpoint $\mu F_{\mathit{List\,a}}$ of the
 first-order functor whose action on sets is given by
@@
 functions is given by $F_{\mathit{List}\,a}\,f = 1 + \id_a \times
-f$. The type constructor $\mathsf{List}$ is modelled by the
-higher-order functor $H$ whose action on a functor $F$ is given by the
-functor $H\,F$ whose actions on sets and functions between them are
-given by $H\,F\,X = 1 + X \times F\,X$ and $H\,F\,f = \id_1 + f \times
-F\,f$, respectively, and whose action on a natural transformation
-$\eta$ is the natural transformation given by $H\,\eta = \id_1 + \eta
-\times F\,\eta$. The type constructor $\mathsf{List}$ is then
-modelled as the functor $\mu H$. Reflecting the functorial action of
-$\mu H$ back into syntax gives Haskell's built-in $\mathsf{map}$
-function as the type-and-data-uniform, structure-preserving,
-data-changing function associated with $\mathsf{List}$. {\color{blue}
-  HOW?}
+f$. The type constructor $\mathsf{List}$ is modelled by the functor
+that is the fixpoint $\mu H$ of the higher-order functor $H$ whose
+action on a functor $F$ is given by the functor $H\,F$ whose actions
+on sets and functions between them are given by $H\,F\,X = 1 + X
+\times F\,X$ and $H\,F\,f = \id_1 + f \times F\,f$, respectively, and
+whose action on a natural transformation $\eta$ is the natural
+transformation given by $H\,\eta = \id + \eta \times
+F\,\eta$. Reflecting the functorial action of $\mu H$ back into syntax
+gives exactly Haskell's built-in $\mathsf{map}$ function as the
+type-and-data-uniform, structure-preserving, data-changing function
+associated with $\mathsf{List}$. {\color{blue} Double-check this
+  paragraph.}
 
@@
 to {\em state} the rearrange-transform property for Church encodings
-of ADTs, it is simply the reflection back into syntax of the
-{\color{blue} naturality property - right?} of the functorial action
-of the interpretation of the type constructor associated with an ADT's
-primitive representation.
+of ADTs, this property is simply the reflection back into syntax of
+the instance of naturality for the type-independent function that
+rearranges structures of type $\mathsf{T\,a}$ into ones of type
+$\mathsf{T'\,a}$ and the functions $\mathsf{map_T\,f}$ and
+$\mathsf{map_{T'}\,f}$ for a function $\mathsf{f :: a \to b}$ and the
+type constructors $\mathsf{T}$ and $\mathsf{T'}$ associated with these
+ADTs, respectively. {\color{blue} Double-check this paragraph.}
 
 \section{Representations of GADTs}\label{sec:gadts}
-
-{\color{red} HERE!}
 
@@
 instances of the GADT other than the one being defined. An example is
-the following GADT, given in Haskell syntax:
+the GADT $\mathsf{Seq}$ given in~\eqref{eq:seq}. Note that since the
+return type of the data constructor $\mathsf{Pair}$ is not of the form
+$\mathsf{Seq\,a}$ for any variable $\mathsf{a}$, $\mathsf{Seq}$ is a
+GADT that is not a nested type.
+
+By contrast with the ADT $\mathsf{List\, a}$, where the type parameter
+$\mathsf{a}$ is integral to the type being defined, the type parameter
+$\mathsf{a}$ appears in both $\mathsf{PTree\,a}$ and $\mathsf{Seq\,a}$
+as a ``dummy'' parameter used only to give the kind \verb|* -> *| of
+the type constructors $\mathsf{PTree}$ and $\mathsf{Seq}$. This is
+explicitly captured in the alternative Haskell syntax {\color{blue}
+    Correct?}
 \[\begin{array}{l}
@@
 \end{array}\]
-Since the return type of the data constructor $\mathsf{Pair}$ is not
-simply a variable, $\mathsf{Seq}$ is not a nested type.
-
-By contrast with the ADT $\mathsf{List\, a}$, where the type parameter
-$\mathsf{a}$ is integral to the type being defined, the type parameter
-$\mathsf{a}$ in $\mathsf{PTree\,a}$ and $\mathsf{Seq\,a}$ appears as a
-``dummy'' parameter used only to give the kind \verb|* -> *| of the
-type constructors $\mathsf{PTree}$ and $\mathsf{Seq}$. This is
-explicitly captured by the above GADT syntax. A GADT --- even a nested
-type --- thus does not define a {\em family of inductive types}, one
-for each type argument, like an ADT does, but instead defines an
-entire family of types that must be constructed simultaneously. That
-is, a GADT defines an {\em inductive family of types}.
+A GADT --- even a nested type --- thus does not define a {\em family
+  of inductive types}, one for each type argument, like an ADT does,
+but instead defines an entire family of types that must be constructed
+simultaneously. That is, a GADT defines an {\em inductive family of
+  types}.
 
 \subsection{Church Encodings of GADTs}\label{sec:CEs}
+
+{\color{red} HERE!}
 

### 2021-02-23 20-gadts/GADT-story.tex
work on gadt-story

@@
 
-\section{Introduction}
-{\color{blue} map vs. fmap throughout}
+\section{Introduction}\label{sec:intro}
+{\color{blue} map vs. fmap throughout.} {\color{blue} Add polynomial
+  throughout?}
 
@@
 the type constructors $\mathsf{PTree}$ and $\mathsf{Seq}$. This is
-explicitly captured in the alternative Haskell syntax {\color{blue}
-    Correct?}
+explicitly captured in the alternative ``kind signature'' Haskell
+syntax
 \[\begin{array}{l}
@@
 
-{\color{red} HERE!}
-
-The syntax of GADTs is consonant with programmers' frequent use of
-them in situations where type uniformity is specifically to be
-eschewed.  {\color{blue} Typical applications of GADTs include generic
-  programming, modeling programming languages via higher-order
-  abstract syntax, maintaining invariants in data structures, and
-  expressing constraints in embedded domain-specific languages. GADTs
-  have also been used to implement tagless
-  interpreters~\cite{pl04,pr06,pvww06} by trading the definition of a
-  universal value domain for a direct specification of the property of
-  being a value. Other applications are described in,
-  e.g.,~\cite{js15,rou06,MORE,UPDATED}. OTHER SITUATIONS NEEDING
-  TYPE-BASED DISPATCH.}
-
-GADTs are useful in such situations precisely because generalizing the
-type arguments in the instances of a GADT appearing as the return
-types of its data constructors entails a strong connection between a
-GADTs shape and the data it contains.  With ADTs, we first choose the
-shape of the container and then fill that container with data of
-whatever type we like. Critically, the choice of shape is independent
-of the data to be stored. With GADTs, however, the shape of the
-container may actually {\em depend} on (the type of) the data to be
-contained. On other words, the shape and the data are not independent!
-For example, while $\mathsf{Const}$ can create data of any shape
-$\mathsf{Seq\,a}$, $\mathsf{Pair}$ can produce data of shape
-$\mathsf{Seq\,a}$ where $\mathsf{a}$ is a pair type. As a result, it
-may not always be possible to modify a GADT's data while retaining its
-shape because changing (the type of) the data may actually change the
-shape.
+The syntax of GADTs is consistent with programmers' use of them in
+situations where different behaviors at different type instances is
+desired. This is achieved by allowing the progammer to give the type
+signatures of the different data constructors independently, and then
+using pattern matching to force the desired type refinement.  Typical
+applications of GADTs include generic programming, modeling
+programming languages via higher-order abstract syntax, maintaining
+invariants in data structures, and expressing constraints in embedded
+domain-specific languages. They have also been used, e.g., to
+implement tagless interpreters~\cite{pl04,pr06,pvww06}, to improve
+memory performance~\cite{min15}, and to design APIs~\cite{pen20}.
+
+Allowing non-variable type arguments in the return types of a GADT's
+data constructors establishes a strong connection between a GADT's
+shape and the data it contains. With ADTs, we first choose the shape
+of the container and then fill that container with data of whatever
+type we like. Critically, the choice of shape is independent of the
+data to be stored. With GADTs, however, the shape of the container may
+actually {\em depend} on (the type of) the data to be contained. For
+example, $\mathsf{Const}$ can create data of any shape
+$\mathsf{Seq\,a}$, but $\mathsf{Pair}$ can produce data of shape
+$\mathsf{Seq\,a}$ only if $\mathsf{a}$ is a pair type. As a result,
+modifying the data in a GADT may actually change the shape of --- or
+may even destroy! --- the GADT.
 
 To determine the possible shapes of a GADT's container we have to
-pattern-match on the type of the data to be contained. For this, a
-GADT calculus must support an equality type $\mathsf{Equal}$ that is
-inhabited when its two type arguments are the same and is uninhabited
-otherwise, i.e., is the syntactic reflection of semantic equality
-function $\mathit{Equal}$. {\color{blue} This is the least commitment
-  thing Church encodings require.} Such a type can either be taken as
-primitive, as in
+pattern-match on the type of the data to be contained. For this, it is
+essential that a GADT calculus support an equality type
+$\mathsf{Equal}$ that is inhabited when its two type arguments are the
+same and is uninhabited otherwise. That is, the type $\mathsf{Equal}$
+must be the syntactic reflection of semantic equality function
+$\mathit{Equal}$. The type $\mathsf{Equal}$ can either be defined via
+GADT syntax, as in
 \[\begin{array}{l}
-\mathsf{data\, Equal :: * \to * \to *\,where}\\
-\mathsf{\;\;\;\;\;\;\;\;Refl ::\, Equal\,a\,a}\\
+\mathsf{data\, Equal\,a\,b\,where}\\
+\mathsf{\;\;\;\;\;\;\;\;Refl ::\, Equal\,c\,c}\\
 \end{array}\]
 \noindent
-or
-represented by its own Church encoding
-\[\mathsf{Equal\,t_1\,t_2} = \mathsf{\forall f.\, (\forall a.\,
-  f\,a\,a) \to f\,t_1\,t_2}\] Using $\mathsf{Equal}$, the GADT
-$\mathsf{Seq}$ can then be represented as its Church encoding
+or, equivalently, represented by its own Church encoding
+\[\mathsf{Equal\,a\,b} = \mathsf{\forall f.\, (\forall c.\,
+  f\,c\,c) \to f\,a\,b}\]
+The GADT $\mathsf{Seq}$ can then be represented as its Church encoding
 \[\mathsf{Seq\,a} = \mathsf{\forall f.\, (\forall a.\,a \to f\, a)
@@
 
-Importantly, $\mathit{Equal}\,a$ cannot be made into a functor, or,
-equivalently, $\mathsf{Equal\,a}$ cannot be made an instance of
-Haskell's $\mathsf{Functor}$ class: if $\mathsf{Equal\,a}$ supported a
-function
+Importantly, the function $\mathit{Equal}\,a$ cannot be made into a
+functor. Equivalently, $\mathsf{Equal\,a}$ (in either representation)
+cannot be made an instance of Haskell's $\mathsf{Functor}$
+class. Indeed, if $\mathsf{Equal\,a}$ supported a function
 \[\mathsf{map_{Equal\,a} :: (b \to c) \to Equal\,a\,b \to
@@
 \end{array}\]
-would allow us to construct the element 
+would allow us to construct an element 
 \[\begin{array}{l}
@@
 \end{array}\]
-of type $\mathsf{Void}$, which is impossible. It follows immediately
-from this contradiction that $\mathsf{Seq}$ cannot support a map
-function either. This is already intuitively evident from the Haskell
-definition of $\mathsf{Seq}$, though. ALREADY USED??? Indeed, the clause of
-$\mathsf{map}$ for the $\mathsf{Pair}$ constructor should feed
-$\mathsf{map}$ a function $\mathsf{f :: (b \times c) \to e}$ and a
-term of the form $\mathsf{Pair \,t_1\,t_2}$ for $\mathsf{x :: b}$ and
-$\mathsf{y :: c}$, and produce a term $\mathsf{Pair\,t_1'\,t_2'}$ for
-some appropriately typed terms $\mathsf{t_1'}$ and
-$\mathsf{t_2'}$. But it is not clear how to achieve this since
-$\mathsf{e}$ need not necessarily be a product type. And even if
-$\mathsf{e}$ {\em were} known to be of the form $\mathsf{w \times z}$,
-then we still wouldn't necessarily have a way to produce data of type
-$\mathsf{w \times z}$ from only $\mathsf{f :: b \times c \to w \times
-  z}$ and $\mathsf{x}$ and $\mathsf{y}$ unless we knew, e.g., that
-$\mathsf{f}$ was a pair of functions.
+of type $\mathsf{Void}$. {\color{blue} Use same type conventions as in
+  Intro.} But this should be impossible. It is immediate that no GADT
+that is not actually a nested type can support a $\mathsf{map}$
+function either. In the case of $\mathsf{Seq}$, this is already
+intuitively evident from its Haskell definition as discussed in
+Section~\ref{sec:intro}.
 
@@
 
-The primitive representations of {\color{blue} polynomial} GADTs
-capture in syntax the fact that GADTs can be considered as fixpoints
-of higher-order functors. Syntactically, the use of higher-order
-funct{\em ions} is essential, since the type arguments to the GADT
-being defined are not uniform across all of its instances in the types
-of its data constructors. Semantically, the use of higher-order
-funct{\em ors} is essential, as for ADTs, to the very existence of the
-fixpoints being computed~\cite{tfca}.
+{\color{red} HERE!!!}
+
+Primitive representations of GADTs capture in syntax the fact that
+GADTs can, like ADTs, be considered as fixpoints of higher-order
+functors. Syntactically, the use of higher-order funct{\em ions} is
+essential, since the type arguments to the GADT being defined are not
+uniform across all of its instances in the types of its data
+constructors. Semantically, the use of higher-order funct{\em ors} is
+essential, as for ADTs, to the very existence of the fixpoints being
+computed~\cite{tfca}.
 %We will consider GADTs' semantics in more detail in
@@
 
-Consider, again, the GADT $\mathsf{Seq}$ from the start of
-Section~\ref{sec:gadts}. Because its type argument varies in the
-instances of $\mathsf{Seq}$ appearing in the types of its data
-constructor $\mathsf{Pair}$, $\mathsf{Seq}$ cannot be characterized by
-any first-order fixpoint equation. It can, however, be viewed as a
+Consider, again, the GADT $\mathsf{Seq}$. Because its type argument
+varies in the instances of $\mathsf{Seq}$ appearing in the types of
+its data constructor $\mathsf{Pair}$, $\mathsf{Seq}$ simply cannot be
+characterized by any first-order fixpoint equation.  {\color{blue}
+  Compare with ADT situation.}  It can, however, be viewed as a
 solution to the higher-order fixpoint equation
@@
 a b. f a \times f b\] where $Lan_K \,G$ is the left Kan extension of
-$G$ along $K$.  The left Kan extension $Lan_K \,G : {\cal E} \to
-{\cal D}$ of the functor $G : {\cal C} \to {\cal D}$ along the functor
-$K : {\cal C} \to {\cal E}$ is the best functorial approximation to
-$G$ that factors through $K$. Here, ``best functorial approximation''
-means that $Lan_K \,G$ is the smallest functor that both extends the
-image of $K$ to $\cal D$ and coincides with $G$ on $\cal C$, in the
-sense that, for any other such functor $H$, there is a morphism of
-functors (i.e., a natural transformation) from $Lan_K \,G$ to
-$H$. This is captured formally by the following
+$G$ along $K$.  In general, the left Kan extension $Lan_K \,G : {\cal
+  E} \to {\cal D}$ of the functor $G : {\cal C} \to {\cal D}$ along
+the functor $K : {\cal C} \to {\cal E}$ is the best functorial
+approximation to $G$ that factors through $K$. Here, ``best functorial
+approximation'' means that $Lan_K \,G$ is the smallest functor that
+both extends the image of $K$ to $\cal D$ and coincides with $G$ on
+$\cal C$, in the sense that, for any other such functor $H$, there is
+a morphism of functors (i.e., a natural transformation) from $Lan_K
+\,G$ to $H$. This is captured formally by the following
 definition~\cite{mac71}:
@@
 
+\subsection{Comparison with Stephanie and Dmitrios}
+
 \subsection{Comparison with Bob's Parametric Model}
+
+GADTs don't really fit into his scheme because the two reps of GADTs
+aren't the same. Bob's stuff only works for functorial data types of
+any kind -- so not GADTs. So last section is misleading.
 

### 2021-02-23 20-gadts/GADT-story.tex
work on gadt-story

@@
 view, GADTs are completely determined by their syntax; for example,
-the GADT $\mathsf{G}$ defined in Haskell by
+the GADT $\mathsf{Seq}$ defined in Haskell by
 \begin{equation}\label{eq:seq}
@@
   higher-order} functor on $\mathit{Set}$. Reflecting this requirement
-back into syntax gives that $\mathsf{T\,a} = \mathsf{(\mu F)\,a} =
-\mathsf{(\mu\,(\mu H))\,a}$ for the ``type constructor constructor''
+back into syntax gives that
+$\mathsf{T\,a} = \mathsf{(\mu H)\,a}$
+for the ``type constructor constructor''
 $\mathsf{H}$ that supports suitable\footnote{The $\mathsf{map}$
@@
 whose action on a natural transformation $\eta$ is the natural
-transformation given by $H\,\eta = \id + \eta \times
-F\,\eta$. Reflecting the functorial action of $\mu H$ back into syntax
-gives exactly Haskell's built-in $\mathsf{map}$ function as the
-type-and-data-uniform, structure-preserving, data-changing function
-associated with $\mathsf{List}$. {\color{blue} Double-check this
-  paragraph.}
+transformation whose component at $X$ is given by $(H\,\eta)_X = \id_1
++ \id_X \times \eta_X$. Reflecting the functorial action of $\mu H$
+back into syntax gives exactly Haskell's built-in $\mathsf{map}$
+function as the type-and-data-uniform, structure-preserving,
+data-changing function associated with $\mathsf{List}$.
 
@@
 rearranges structures of type $\mathsf{T\,a}$ into ones of type
-$\mathsf{T'\,a}$ and the functions $\mathsf{map_T\,f}$ and
-$\mathsf{map_{T'}\,f}$ for a function $\mathsf{f :: a \to b}$ and the
-type constructors $\mathsf{T}$ and $\mathsf{T'}$ associated with these
-ADTs, respectively. {\color{blue} Double-check this paragraph.}
+$\mathsf{T'\,a}$ and the content-transforming functions
+$\mathsf{map_T\,f}$ and $\mathsf{map_{T'}\,f}$ for a function
+$\mathsf{f :: a \to b}$, where $\mathsf{T}$ and $\mathsf{T'}$ are the
+type constructors associated with these ADTs, respectively.
 

### 2021-02-27 20-gadts/GADT-story.tex
work on Sections 3.2 and 4

@@
 
-A GADT programmer is likely to use GADTs precisely because they
+A GADT programmer is likely to use GADTs precisely {\em because} they
 exhibit different behaviors at different types, and thus to consider a
@@
 
-A computer scientist interested in the semantics of functional
-languages, on the other hand, is likely to follow~\cite{bfss90} and
-expect GADTs to behave as their names suggests --- i.e., as
+A semanticist, on the other hand, is likely to follow~\cite{bfss90}
+and expect GADTs to behave as their names suggests --- i.e., as
 generalizations of ADTs. The semanticist is thus likely to expect a
@@
 \mathsf{0}$ since there is always a function from $\mathsf{1}$ to such
-an $\mathsf{a}$. But $\mathsf{G}$ does not include any terms at
-instance $\mathsf{G\,0}$ because $\mathsf{G\,0}$ is not inhabited via
-the syntactic specifiation of $\mathsf{G}$, and it is not inhabited
-via functorial completion because it is not possible to define a
-function from $\mathsf{1}$ into $\mathsf{0}$. {\color{blue} Use $\bot$
-  instead of $0$ here?}
+an $\mathsf{a}$, but does not include any terms at instance
+$\mathsf{G\,0}$. Indeed, $\mathsf{G\,0}$ is not inhabited via the
+syntactic specifiation of $\mathsf{G}$, and is not inhabited via
+functorial completion because it is not possible to define a function
+from $\mathsf{1}$ into $\mathsf{0}$. {\color{blue} Use $\bot$ instead
+  of $0$ here?}
 
@@
 
-The goal of this paper we show how these observations can be made
-precise, and therefore to show that how GADTs are viewed has
-significant consequences. This paper thus answers the question we
-posed above in the affirmative: Yes, it really does matter.
+Below we show how the above observations can be made precise, thereby
+answering the question we posed above in the affirmative: Yes, it
+really does matter which way we view GADTs.
 
@@
 $\mathsf{xs \,:\,List\,a}$, and $\mathsf{g : List\,a\to Tree\,a}$
-arranges lists into trees in a type-independent way, then
+arranges lists into trees in a type-independent way, then we have the
+following rearrange-transform property:
 \[\mathsf{fmap_{Tree}\,f\,(g \, xs) \,=\, g\,(fmap_{List}\,f\,xs)}\]
@@
 structure-preserving, data-changing $\mathsf{map}$ function can be
-defined.
+defined. It therefore need not satisfy any rearrange-transform
+property either.
 
@@
 ADT actually denotes a semantic object computed as a real-live
-fixpoint, the semantic fixpoint calculation must converge.  If we
-interpret our types as sets, then this means that the fixpoint being
-taken must be of a {\em functor} on the category $\mathit{Set}$ of
-sets and functions between them, rather than of a mere function
-between sets~\cite{tfca}.  That is, the function $F$ interpreting the
-type constructor $\mathsf{F}$ constructing the body of a syntactic
-fixpoint must not only have an action on sets, but must also have a
-{\em functorial action} on functions between sets. Reflecting this
-requirement back into syntax gives that $\mathsf{F}$ must support a
+fixpoint, the semantic fixpoint calculation must converge.  If
+{\color{blue} Since? Establish this earlier?} we interpret our types
+as sets, then this means that the fixpoint being taken must be of a
+{\em functor} on the category $\mathit{Set}$ of sets and functions
+between them, rather than of a mere function between sets~\cite{tfca}.
+That is, the function $F$ interpreting the type constructor
+$\mathsf{F}$ constructing the body of a syntactic fixpoint must not
+only have an action on sets, but must also have a {\em functorial
+  action} on functions between sets. Reflecting this requirement back
+into syntax gives that $\mathsf{F}$ must support a
 suitable\footnote{The $\mathsf{map}$ function for $\mathsf{F}$ should
@@
 to {\em state} the rearrange-transform property for Church encodings
-of ADTs, this property is simply the reflection back into syntax of
-the instance of naturality for the type-independent function that
+of ADTs {\color{blue} unless we first define $\mathsf{map}$ functions
+  for them}, this property is simply the reflection back into syntax
+of the instance of naturality for the type-independent function that
 rearranges structures of type $\mathsf{T\,a}$ into ones of type
@@
 essential that a GADT calculus support an equality type
-$\mathsf{Equal}$ that is inhabited when its two type arguments are the
-same and is uninhabited otherwise. That is, the type $\mathsf{Equal}$
-must be the syntactic reflection of semantic equality function
-$\mathit{Equal}$. The type $\mathsf{Equal}$ can either be defined via
-GADT syntax, as in
+$\mathsf{Equal}$ that is a singleton set when its two type arguments
+are the same and is the empty set otherwise. That is, the type
+$\mathsf{Equal}$ must be the syntactic reflection of semantic equality
+function $\mathit{Equal}$. The type $\mathsf{Equal}$ can either be
+defined via GADT syntax, as in
 \[\begin{array}{l}
@@
   f\,a) \to f\,a}\]
-in a higher-kinded calculus such as $F_\omega$.
+in a higher-kinded calculus such as $F_\omega$~\cite{atk12}.
 
@@
 of type $\mathsf{Void}$. {\color{blue} Use same type conventions as in
-  Intro.} But this should be impossible. It is immediate that no GADT
-that is not actually a nested type can support a $\mathsf{map}$
-function either. In the case of $\mathsf{Seq}$, this is already
-intuitively evident from its Haskell definition as discussed in
-Section~\ref{sec:intro}.
+  Intro.} But this should be impossible. It is immediate that a GADT
+that is not actually a nested type cannot support a $\mathsf{map}$
+function (or a rearrange-transform property) either. In the case of
+$\mathsf{Seq}$, this is already intuitively evident from its Haskell
+definition as discussed in Section~\ref{sec:intro}.
 
@@
 
-{\color{red} HERE!!!}
-
-Primitive representations of GADTs capture in syntax the fact that
-GADTs can, like ADTs, be considered as fixpoints of higher-order
-functors. Syntactically, the use of higher-order funct{\em ions} is
-essential, since the type arguments to the GADT being defined are not
-uniform across all of its instances in the types of its data
-constructors. Semantically, the use of higher-order funct{\em ors} is
-essential, as for ADTs, to the very existence of the fixpoints being
-computed~\cite{tfca}.
-%We will consider GADTs' semantics in more detail in
-%Section~\ref{sec:GADT-semantics}.
+The Church encoding of a GADT corresponds to the data type comprising
+just those elements specified by the GADT's syntax. By contrast, the
+primitive representation of a GADT corresponds to the data type
+comprising all data elements in the functorial completion of the
+GADT's syntax. In this alternate reading, GADTs can, like ADTs, be
+modelled as fixpoints of higher-order functors. Syntactically, the use
+of higher-order funct{\em ions} is essential; since the type arguments
+to the GADT being defined are not necessarily uniform across all of
+its instances in the types of its data constructors, GADTs cannot be
+seen as fixpoints of first-order functors the way ADTs
+can. Semantically, the use of higher-order funct{\em ors} is
+essential, as in the case of ADTs, to guarantee the existence of the
+fixpoints being computed~\cite{tfca}.
 
@@
 varies in the instances of $\mathsf{Seq}$ appearing in the types of
-its data constructor $\mathsf{Pair}$, $\mathsf{Seq}$ simply cannot be
-characterized by any first-order fixpoint equation.  {\color{blue}
-  Compare with ADT situation.}  It can, however, be viewed as a
-solution to the higher-order fixpoint equation
-\[H\,f\,a = a \,+\,Lan_{\lambda a b. a \times b}\, \lambda
-a b. f a \times f b\] where $Lan_K \,G$ is the left Kan extension of
-$G$ along $K$.  In general, the left Kan extension $Lan_K \,G : {\cal
-  E} \to {\cal D}$ of the functor $G : {\cal C} \to {\cal D}$ along
-the functor $K : {\cal C} \to {\cal E}$ is the best functorial
-approximation to $G$ that factors through $K$. Here, ``best functorial
-approximation'' means that $Lan_K \,G$ is the smallest functor that
-both extends the image of $K$ to $\cal D$ and coincides with $G$ on
+its data constructor $\mathsf{Pair}$, $\mathsf{Seq}$ cannot be
+modelled as the fixpoint of any first-order functor. As shown
+in~\cite{jp19}, it can, however, be modelled as a solution to the
+higher-order fixpoint equation
+\[H\,f\,a = a \,+\,(Lan_{\lambda c d. c \times d}\, \lambda
+c d. f c \times f d)\,a\]
+{\color{red} Check this equation against next displayed one below.}
+where $Lan_K \,F$ is the left Kan extension of
+the functor $F$ along the functor $K$. In general, the left Kan
+extension $Lan_K \,F : {\cal E} \to {\cal D}$ of $F : {\cal C} \to
+{\cal D}$ along $K : {\cal C} \to {\cal E}$ is the best functorial
+approximation to $F$ that factors through $K$. Here, ``best functorial
+approximation'' means that $Lan_K \,F$ is the smallest functor that
+both extends the image of $K$ to $\cal D$ and coincides with $F$ on
 $\cal C$, in the sense that, for any other such functor $H$, there is
 a morphism of functors (i.e., a natural transformation) from $Lan_K
-\,G$ to $H$. This is captured formally by the following
+\,F$ to $H$. This is captured formally by the following
 definition~\cite{mac71}:
 \begin{definition}\label{def:lke}
-If $G : {\cal C} \to {\cal D}$ and $K : {\cal C} \to {\cal E}$ are
-functors, then the {\em left Kan extension of $G$ along $K$} is a
-functor $\mathit{Lan}_K\,G : {\cal E} \to {\cal D}$ together with a
-natural transformation $\eta : G \to (\mathit{Lan}_K\,G) \circ K$
-such that, for every functor $H : {\cal E} \to {\cal D}$ and natural
-transformation $\gamma : G \to H \circ K$, there exists a unique
-natural transformation $\delta : \mathit{Lan}_K\,G \to H$ such that $(\mu
-K) \circ \eta = \gamma$. This is depicted in the diagram
+If $F : {\cal C} \to {\cal D}$ and $K : {\cal C} \to {\cal E}$ are
+functors, then the {\em left Kan extension of $F$ along $K$} is a
+functor $\mathit{Lan}_K\,F : {\cal E} \to {\cal D}$ together with a
+natural transformation $\eta : F \to (\mathit{Lan}_K\,F) \circ K$ such
+that, for every functor $H : {\cal E} \to {\cal D}$ and natural
+transformation $\gamma : F \to H \circ K$, there exists a unique
+natural transformation $\delta : \mathit{Lan}_K\,F \to H$ such that
+$(\delta K) \circ \eta = \gamma$. This is depicted in the diagram
 \[\begin{tikzcd}[row sep = huge]
 {\cal C}
-\ar[rr, "{G}"{name=Fa, above}, ""{name=F, below}]
+\ar[rr, "{F}"{name=Fa, above}, ""{name=F, below}]
 \ar[rd, "{K}"']
@@
 \ar[Rightarrow, bend right = 25, from=F, "{\eta}"']
-\ar[ur, bend left, "{\mathit{Lan}_{K}\,G}"{name=Lan, description}]
+\ar[ur, bend left, "{\mathit{Lan}_{K}\,F}"{name=Lan, description}]
 \ar[ur, bend right, "{H}"'{name=L, right}, ""'{name=Lr, right}]
@@
 
-To represent GADTs as primitive data types a calculus must support a
-primitive fixpoint constructor $\mathsf{\mu}$ that is the syntactic
-reflection of the higher-order fixpoint operator on the category of
-functors on the category interpretating types, as well as a primitive
-type constructor $\mathsf{Lan}$ such that $\mathsf{Lan_k\,g}$ is the
-syntactic reflection of the left Kan extension $\Lan_K \,G$ of the
-functor $G$ interpreting $\mathsf{g}$ along the functor $K$
-interpreting $\mathsf{k}$. Using the $\mu$ and $\Lan$ type
-constructors we can represent, e.g., the GADT $\mathsf{Seq}$ as
-\[\mathsf{Seq\,a} = \mathsf{\mu \phi.\, a + Lan_{\lambda c d. c \times d}
-  \lambda c d. \phi c \times \phi d}\]
-
-{\color{blue} This is the least commitment
-  thing primitive representations can require. Fixpoints were elided
-  in ADT case. Why? Tie together better.}  The type constructor
-$\mathsf{Lan}$ thus computes the smallest data type that is generated
-by a GADT's {\color{blue} GADT or constructor?} syntax and also
-supports a $\mathsf{map}$ function. That is, $\mathsf{Lan}$ computes
-the functorial completion of the data type determined by a GADT's
-syntax, filling it out with data elements that are derivable via
-repeated applications of $\mathsf{map}$ starting from those elements
-specified by its syntax. In the primitive view, GADTs are, in general,
-underspecified by their syntax.
-
-Just as for ADTs, we can represent GADTs as either Church encodings or
-primitive data types. But whereas the two representations are always
-isomorphic for ADTs, they are not, in general, for GADTs. We will see
-this in the next section.
-
-{\color{blue} rearrange-transform property for GADTs?}
+To represent GADTs as primitive data types in a setting in which types
+are interpreted as sets, a calculus must support a primitive construct
+$\mathsf{Lan}$ such that the type constructor $\mathsf{Lan_k\,f}$ is
+the syntactic reflection of the left Kan extension $\Lan_K \,F$ of the
+functor $F$ interpreting $\mathsf{f}$ along the functor $K$
+interpreting $\mathsf{k}$. In this setting, the categories
+$\mathcal{C}$, $\mathcal{D}$, and $\mathcal{E}$ must all be
+$\mathit{Set}$. Using $\mathsf{Lan}$ we can then rewrite the type of a
+constructor $\mathsf{c :: f a \to g(k\,a)}$ as $\mathsf{c ::
+  (Lan_k\,f)\,a \to g\,a}$ since morphisms (i.e., natural
+transformations) from $F$ to $G \circ K$ are in one-to-one
+corresondence with those from $\mathit{Lan}_K\,F$ to $G$ (see,
+e.g.,~\cite{rie16}). That is, writing $F \Rightarrow G$ for the set of
+natural transformations from a functor $F$ to a functor $G$, we have
+\begin{equation}\label{eq:nat-transfs}
+F \Rightarrow G \circ K\; \simeq \; Lan_K\,F \Rightarrow G
+\end{equation}
+The calculus must also support a primitive type
+constructor $\mathsf{\mu}$ that is the syntactic reflection of the
+(now higher-order) fixpoint operator on
+$\mathit{Set}^\mathit{Set}$. Using $\mu$ and $\Lan$ we can then
+represent a GADT as a higher-order higher-order fixpoint. For example,
+we can represent the GADT $\mathsf{Seq}$ as
+\[\mathsf{Seq\,a} = \mathsf{(\mu \phi.\lambda b.\, b + (Lan_{\lambda c
+    d. c \times d} \lambda c d. \phi c \times \phi d)\,b)\,a}\] The
+fact that $Lan_K\,G$ is the best functorial approximation to $G$
+factoring through $K$ means that the type constructor
+$\mathsf{Lan_k\,g}$ computes the smallest collection of data that is
+generated by the corresponding GADT data constructor's syntax and also
+supports a $\mathsf{map}$ function. The primitive representation of
+any GADT thus comprises the smallest data type that both includes the
+data specified by that GADT's syntax and also supports a
+$\mathsf{map}$ function (and thus a rearrange-transform property). In
+the primitive view, then, GADTs are generally underspecified by their
+syntax.
+
+In this section we have seen that GADTs can be represented either as
+Church encodings or as primitive data types. This exactly mirrors the
+situation for ADTs described in Section~\ref{sec:adts}. But whereas
+the two representations are always isomorphic for ADTs, they are not,
+in general, the same for GADTs. This will be proved in the next
+section.
 
@@
 
-{\color{red} Check all fonts!! Syntax vs. semantics!}  Intuitively it
-seems the two representations might be different. One is functorial,
-one is not. It is possible they could always have the same data
-elements even if not the same morphisms. But as we will see in
-Example~\ref{ex:g1}, this turns out not to be the case.
+{\color{blue} Is G or G\,a the GADT?}
+
+To see that the Church encoding of a GADT and its primitive
+representation need not be the same, consider again the GADT
+$\mathsf{G\,a}$ defined in~\eqref{eq:G}. Despite its simplicity, this
+GADT serves as an informative case study highlighting the differences
+--- even if we only consider the data elements a GADT contains ---
+between a GADT's Church encoding its representation as a primitive
+data type.
+
+\begin{example}\label{ex:g1}
+Syntactically, the GADT $\mathsf{G\,a}$ defined in~\eqref{eq:G}
+comprises a single data element, namely $\mathsf{c :: G \, 1}$. This
+is captured by $\mathsf{G}$'s Church encoding
+\[\mathsf{G\,a} = \mathsf{Equal\,1\,a}\]
+from~\cite{atk12}, so $\mathsf{G}$'s effect is simply to test its
+argument for equality against the unit type
+$\mathsf{1}$. {\color{blue} Use same notation for unit type as in
+  intro.}
+%Recalling from Section~\ref{sec:CEs} that the
+The interpretation of $\mathsf{G\,a}$ represented as the Chirch
+encoding $\mathsf{Equal\,1\,a}$ is thus the function whose value is a
+singleton set when the interpretation of $\mathsf{a}$ is $1$ and the
+empty set otherwise. This perfectly mirrors $\mathsf{G}$'s syntax,
+which indeed delivers just one data element when $\mathsf{a}$ is
+$\mathsf{1}$ and no elements otherwise.
+
+We now want to compute the interpretation of $\mathsf{G}$'s primitive
+representation. The type of $\mathsf{G}$'s unique constructor
+$\mathsf{c :: G\,1}$ can be equivalently expressed as $\mathsf{c :: 1
+  \to G\,1}$ or, using~\eqref{eq:nat-transfs}, as $\mathsf{c ::
+  (Lan_{\lambda *. 1}\,\lambda *. 1)\,a \to G\, a}$, where
+$\mathsf{\lambda *. 1}$ is the syntactic reflection of the constantly
+$1$-valued functor from the category with a single object $*$ to
+$\mathit{Set}$. We can therefore represent $\mathsf{G}$ as
+\[\mathsf{G\,a} = (\mathsf{\mu \phi. \lambda b.
+ (Lan_{\lambda *. 1} \,\lambda *.  1)\,b)\,a}\] The interpretation of
+$\mathsf{G\, a}$ is therefore just the application to $\mathsf{a}$ of
+the fixpoint of the interpretation of the body $\mathsf{(Lan_{\lambda
+    *. 1} \,\lambda *. 1)\,b}$ of the syntactic fixpoint.  But since
+the recursion variable $\mathsf{\phi}$ does not appear in this body,
+the interpretation of the fixpoint is just the interpretation of the
+body itself.  The interpretation of $\mathsf{G\,a}$ is therefore
+$(\mathit{Lan}_{\lambda *. 1} \lambda *. 1)\, A$, where $A$ interprets
+$\mathsf{a}$.
+%and $1$ is the {\color{blue} constantly $1$-valued
+%  functor $\ast \to \set$ from the category with a single object
+%  $\ast$ to $\set$.}
+It turns out, however, that, for any set $A$, $(\mathit{Lan}_{\lambda
+  *. 1} \lambda *. 1)\,A$ is, in fact, exactly $A$. Indeed,
+Proposition~7.1 of~\cite{blw03} gives that $(\mathit{Lan}_{\lambda
+  *. 1} \lambda *. 1)\,A$ can be computed as the set of all functions
+$a : 1 \to A$ modulo the smallest equivalence relation $\sim$ such
+that, for any $f : 1 \to 1$, we have $a \sim a \circ f$.  But this set
+is isomorphic to $A$, because the only function $f : 1 \to 1$ is the
+identity function.
+
+Putting it all together, we see that the Church encoding of
+$\mathsf{G}$ and its primitive representation are not equivalent: the
+interpretation of the former has exactly one data element at instance
+$\mathsf{G\,1}$ and no elements at any other instances, whereas the
+interpretation of the latter has data elements at every instance other
+than $\mathsf{G\,0}$.
+%Indeed,
+These additional data elements can be obtained by reflecting back into
+syntax the elements $\mathit{map}_G\,f_a\,c : G\,A$ resulting from
+applying the functorial action $\mathit{map}_G$ of $\mathsf{G}$'s
+interpretation $G$ to the functions $f_a : 1 \to A$ determined by the
+elements $a$ of $A \not = \emptyset$ and the interpretation $c$ of
+$\mathsf{c}$.
+\end{example}
+
+{\color{red} STOPPED HERE}
 
@@
   polynomial nested type.}
-
-\begin{example}\label{ex:g1}
-Consider the following GADT, defined in a Haskell-like syntax:
-\begin{equation}\label{eq:example-GADT}
-\begin{array}{l}
-\mathsf{data\,G : * \to *\,where}\\
-\mathsf{\;\;\;\;\;\;\;\;c : G\,1}
-\end{array}
-\end{equation}
-Syntactically, this GADT comprises a single data element, namely
-$\mathsf{c :: G \, 1}$. Despite the simplicity of this GADT, it serves
-as an informative case study highlighting the differences between the
-Church encoding of a GADT and that GADT's representation as a
-primitive data type.
-
-The Church encoding of $\mathsf{G}$ is
-\[\mathsf{G\,a} = \mathsf{\forall f.\, (\forall b.\,
-  f\,b) \to f\,1} \] In other words, $\mathsf{G\,a =
-  \mathsf{Equal}\,1\,a}$, so $\mathsf{G}$'s effect is simply to test
-its argument for equality against the unit type $\mathsf{1}$.
-
-Recall from the Section~\ref{sec:CEs} that the semantic interpretation
-of $\mathsf{Equal\,1\,a}$ is given by a characteristic function, in
-the sense that it is the singleton $1$ when the interpretation of
-$\mathsf{a}$ is $1$, and is the empty set otherwise.  Thus, as a data
-type represented by Church encoding, $\mathsf{G}$ will be the data
-type containing one (and only one) piece of data when instantiated on
-$\mathsf{1}$, and uninhabited otherwise.  This perfectly mirrors the
-syntax of $\mathsf{G}$, which indeed provides just one constructor
-when the type is instantiated on $\mathsf{1}$, and none otherwise.
-Thus, given a set $A$ interpreting $\mathsf{a}$, the set
-interpretation $G_0(A)$ is $1$ if $A = 1$ and $\emptyset$ otherwise.
-
-We now want a primitive representation of $\mathsf{G}$.  Recall that
-$\mathsf{G}$ has a unique constructor $\mathsf{c : G\,1}$, which can
-be equivalently expressed as $\mathsf{c: 1 \to G\,1}$.  By using a
-a$\Lan$ type, we can express $\mathsf{c}$ as a constructor
-$\mathsf{(Lan_1\, 1)\alpha \to G \alpha}$.  Thus, as a primitive data
-type, we represent $\mathsf{G}$ as $\mu \phi. \lambda
-\alpha. (\Lan^{\emptyset}_1 \,1)\alpha$.  {\color{blue} (WHERE?)
-  Notice that the constructors of a GADT have to be functors, because
-  the GADT is the fixpoint of a higher order functor. Also, we're not
-  using the superscript notation on $\Lan$.}
-
-The semantic interpretation of $\mathsf{\mu \phi. \lambda
-  \alpha. (\Lan^{\emptyset}_1\, 1)\alpha}$ is given by the fixpoint of
-the functor yielded by the semantic interpretation of
-$\mathsf{Lan^{\emptyset}_1\, 1}$, the body of the $\mathsf{\mu}$-type.
-Notice, though, that the recursive variable $\mathsf{\phi}$ does not
-appear in the body of the $\mathsf{\mu}$-type.  Thus, the semantic
-interpretation is actually just the semantic interpretation of the
-body itself.  Therefore, the semantic interpretation is
-$\mathsf{Lan_1 1}$, where $1$ is the constantly $1$-valued functor
-$\ast \to \set$ from the category with a single object $\ast$ to
-$\set$.
-
-{\color{red} Syntax? Semantics? Which are we talking here?}
-It turns out that $\mathsf{\Lan}_1 1$ is, in fact, none other than the
-identity functor on $\set$.  Indeed, for any set $A$,
-$(\mathsf{\Lan}_1 1) A$ is the set of all functions $a : 1 \to A$
-modulo the smallest equivalence relation $\sim$ such that, for any $f
-: 1 \to 1$, we have $a \sim a \circ f$.  But this is isomorphic to
-$A$, because the only function $1 \to 1$ is the identity~\cite{blw03}.
-
-The Church encoding representation of $G$ and the primitive
-representation of $G$ are thus non-equivalent: while the former only
-holds one data element in the instance $G\,1$ and none in other
-instances, the latter is inhabited in all instances with the exception
-of $G\,\emptyset$.  This makes sense if one remembers that the left
-Kan extension, reflected into the syntactic type $\Lan$, is a functor,
-and thus it comes with a functorial map.  This means that, for every
-element $a \in A$, we can map the associated function $\bar{a} : 1 \to
-A$ along $G$ to get a function $\map_{G}\,\bar{a} : G\,1 \to G\,A$,
-and thus we have an element in $G\,A$ given by $\map_{G}\,\bar{a}\, c$
-where $c \in G\,1$ is the element provided by the constructor.
-\end{example}
 
@@
 further compounded by the fact that the the two representations have
-very different parametricity properties.  {\color{blue} For example,
-  we get rearrange-transform property for primitive rep since this is
-  a consequence of functoriality, but don't get it for Church
-  Encodings.} We explore the differences in parametricity results for
-the two representations in the next section.
+very different parametricity properties. We explore the differences in
+parametricity results for the two representations in the next section.
 

### 2021-02-28 20-gadts/GADT-story.tex
finish draft of Section 4

@@
 representation need not be the same, consider again the GADT
-$\mathsf{G\,a}$ defined in~\eqref{eq:G}. Despite its simplicity, this
+$\mathsf{G}$ defined in~\eqref{eq:G}. Despite its simplicity, this
 GADT serves as an informative case study highlighting the differences
---- even if we only consider the data elements a GADT contains ---
-between a GADT's Church encoding its representation as a primitive
-data type.
+between a GADT's Church encoding its primitive representation --- even
+if we consider only the data elements it contains, and ignore whether
+or not it supports a $\mathsf{map}$ function.
 
 \begin{example}\label{ex:g1}
-Syntactically, the GADT $\mathsf{G\,a}$ defined in~\eqref{eq:G}
-comprises a single data element, namely $\mathsf{c :: G \, 1}$. This
-is captured by $\mathsf{G}$'s Church encoding
+Syntactically, the GADT $\mathsf{G}$ defined in~\eqref{eq:G} comprises
+a single data element, namely $\mathsf{c :: G \, 1}$. This is captured
+by $\mathsf{G}$'s Church encoding
 \[\mathsf{G\,a} = \mathsf{Equal\,1\,a}\]
-from~\cite{atk12}, so $\mathsf{G}$'s effect is simply to test its
-argument for equality against the unit type
-$\mathsf{1}$. {\color{blue} Use same notation for unit type as in
-  intro.}
-%Recalling from Section~\ref{sec:CEs} that the
-The interpretation of $\mathsf{G\,a}$ represented as the Chirch
-encoding $\mathsf{Equal\,1\,a}$ is thus the function whose value is a
-singleton set when the interpretation of $\mathsf{a}$ is $1$ and the
-empty set otherwise. This perfectly mirrors $\mathsf{G}$'s syntax,
-which indeed delivers just one data element when $\mathsf{a}$ is
-$\mathsf{1}$ and no elements otherwise.
-
-We now want to compute the interpretation of $\mathsf{G}$'s primitive
-representation. The type of $\mathsf{G}$'s unique constructor
-$\mathsf{c :: G\,1}$ can be equivalently expressed as $\mathsf{c :: 1
-  \to G\,1}$ or, using~\eqref{eq:nat-transfs}, as $\mathsf{c ::
-  (Lan_{\lambda *. 1}\,\lambda *. 1)\,a \to G\, a}$, where
+from~\cite{atk12}. As its Church encoding makes clear, $\mathsf{G}$'s
+effect is simply to test its argument for equality against the unit
+type $\mathsf{1}$. {\color{blue} Use same notation for unit type as in
+  intro.} The interpretation of the Church encoding of $\mathsf{G}$
+is thus, in accordance with Section~\ref{sec:CEs}, the function whose
+value is a singleton set when the interpretation of $\mathsf{a}$ is
+$1$ and the empty set otherwise. This perfectly accords with
+$\mathsf{G}$'s syntax, which indeed delivers just one data element
+when $\mathsf{a}$ is $\mathsf{1}$ and no elements otherwise.
+
+To compute the interpretation of $\mathsf{G}$'s primitive
+representation we first note that the type of $\mathsf{G}$'s solitary
+constructor $\mathsf{c :: G\,1}$ can be equivalently expressed as
+$\mathsf{c :: 1 \to G\,1}$ or, using~\eqref{eq:nat-transfs}, as
+$\mathsf{c :: (Lan_{\lambda *. 1}\,\lambda *. 1)\,a \to G\, a}$, where
 $\mathsf{\lambda *. 1}$ is the syntactic reflection of the constantly
@@
 $\mathit{Set}$. We can therefore represent $\mathsf{G}$ as
-\[\mathsf{G\,a} = (\mathsf{\mu \phi. \lambda b.
+\[\mathsf{G\,a} = \mathsf{(\mu \phi. \lambda b.
  (Lan_{\lambda *. 1} \,\lambda *.  1)\,b)\,a}\] The interpretation of
-$\mathsf{G\, a}$ is therefore just the application to $\mathsf{a}$ of
-the fixpoint of the interpretation of the body $\mathsf{(Lan_{\lambda
-    *. 1} \,\lambda *. 1)\,b}$ of the syntactic fixpoint.  But since
-the recursion variable $\mathsf{\phi}$ does not appear in this body,
-the interpretation of the fixpoint is just the interpretation of the
-body itself.  The interpretation of $\mathsf{G\,a}$ is therefore
-$(\mathit{Lan}_{\lambda *. 1} \lambda *. 1)\, A$, where $A$ interprets
-$\mathsf{a}$.
+$\mathsf{G}$ is therefore the application to $\mathsf{a}$ of the
+fixpoint of the interpretation of the body $\mathsf{\lambda
+  b. (Lan_{\lambda *. 1} \,\lambda *. 1)\,b}$ of the syntactic
+fixpoint $\mathsf{\mu \phi. \lambda b.  (Lan_{\lambda *. 1} \,\lambda
+  *.  1)\,b}$. But since the recursion variable $\mathsf{\phi}$ does
+not appear in this body, the interpretation of the fixpoint is just
+the interpretation of the body itself.  The interpretation of
+$\mathsf{G}$ is therefore $(\mathit{Lan}_{\lambda *. 1} \lambda
+*. 1)\, A$, where $A$ interprets $\mathsf{a}$.
 %and $1$ is the {\color{blue} constantly $1$-valued
@@
 
-{\color{red} STOPPED HERE}
-
-Non-uniform nature of GADTs renders build, therefore short cut, for
-GADTs impossible. Can't even formulate! Of course, there could be
-another way to prove equivalence.  But, in fact, CEs and primitive
-representations are not equivalent for arbitrary GADTs, as the
-following example shows.
-{\color{blue} Our favorite GADT isn't a hereditarily
-  polynomial nested type.}
-
 Forced to choose, a programmer would likely find the idea that a GADT
-contains data that are not specified by its syntax more than a little
+contains data not specified by its syntax more than a little
 disturbing. What, they might ask, should a data type contain other
 than data that are constructed using its data constuctors? Why should
-a GADT contain ``hidden'' elements that are only accessible by
-mapping? On the other hand, a semanticist would likely find the
-primitive representation of GADTs more reasonable. Indeed, they would
-likely find the nonfunctorial nature of the Church encoding unnerving
-at best. After all, they would likely argue, the data in a GADT
-shouldn't disappear just because you map a function over it! The fact
-that this happens when GADTs are represented by their Church encodings
-actually highlights how GADTs {\em do not} generalize the essential,
-container-ish nature of ADTs at all. A semanticist might therefore
-conclude that GADTs are seriously misnamed. They might even go on to
-wonder: in what sense are GADTs data types at all?
-
-The disappointment that GADTs don't validate any such equivalence is
-further compounded by the fact that the the two representations have
-very different parametricity properties. We explore the differences in
-parametricity results for the two representations in the next section.
+a GADT contain ``hidden'' elements that are not specified by the
+GADT's syntax and are only accessible via applications of
+$\mathsf{map}$? A semanticist, on the other hand, would likely find
+the primitive representation of GADTs entirely reasonable. Indeed,
+they would likely find the nonfunctorial nature of a GADT's Church
+encoding unnerving at best. After all, they would likely argue, the
+data in a GADT shouldn't disappear just because you map a function
+over it! The fact that this happens when GADTs are represented by
+their Church encodings actually highlights how GADTs {\em do not}
+generalize the essential, container-ish nature of ADTs at all. A
+semanticist might therefore conclude that GADTs are seriously
+misnamed. They might even go on to wonder: in what sense do GADTs
+really deserve to be considered {\em data types} at all?
+
+The reader familiar with short cut fusion will quickly see that the
+non-uniform nature of GADTs that are not nested types entails that
+short cut fusion cannot be validated for their Church encodings. In
+fact, it is not even possible to define a $\mathsf{build}$ function,
+which is a crucial ingredient in short cut fusion, for the Church
+encoding of a GADT that is not a nested type. But the fact that not
+all of the elements in the functorial completion of a GADT that is not
+a nested type are directly representable as applications of data
+constructors means that it is also impossible to define a
+$\mathsf{build}$ function for the primitive representation of such a
+GADT.  This is not asurorising, however: if short cut fusion were to
+hold for a GADT then its Church encoding and primitive representation
+would necessarily be the equivalent.
+
+In addition to not being equivalent, the two representations of GADTs
+have very different implications for parametricity for GADTs. We
+explore the differences in parametricity results for the two
+representations in the next section.
 

### 2021-03-01 20-gadts/GADT-story.tex
Work on Section 5 up to Example 5.4

@@
 the interpretation of the body itself.  The interpretation of
-$\mathsf{G}$ is therefore $(\mathit{Lan}_{\lambda *. 1} \lambda
+$\mathsf{G\,a}$ is therefore $(\mathit{Lan}_{\lambda *. 1} \lambda
 *. 1)\, A$, where $A$ interprets $\mathsf{a}$.
@@
 Putting it all together, we see that the Church encoding of
-$\mathsf{G}$ and its primitive representation are not equivalent: the
-interpretation of the former has exactly one data element at instance
-$\mathsf{G\,1}$ and no elements at any other instances, whereas the
-interpretation of the latter has data elements at every instance other
-than $\mathsf{G\,0}$.
+$\mathsf{G}$ and its primitive representation are not semantically
+equivalent: the interpretation of the former has exactly one data
+element at instance $\mathsf{G\,1}$ and no elements at any other
+instances, whereas the interpretation of the latter has data elements
+at every instance other than $\mathsf{G\,0}$.
 %Indeed,
@@
 $\mathsf{build}$ function for the primitive representation of such a
-GADT.  This is not asurorising, however: if short cut fusion were to
+GADT.  This is not surprising, however: if short cut fusion were to
 hold for a GADT then its Church encoding and primitive representation
-would necessarily be the equivalent.
-
-In addition to not being equivalent, the two representations of GADTs
-have very different implications for parametricity for GADTs. We
-explore the differences in parametricity results for the two
-representations in the next section.
-
-\section{Parametricity in the Presence of GADTs}\label{sec:par}
+would necessarily be semantically equivalent.
+
+In addition to not being semantically equivalent, the two
+representations of GADTs have very different implications for
+parametricity. We explore the differences in parametricity results for
+GADTs with respect to the two representations in the next section.
+
+\section{Parametricity for GADTs}\label{sec:par}
 
@@
 functions. However, parametricity can also derive results that go
-beyond just naturality, such as establishing {\color{blue} proving}
-the equivalence of the Church encodings and the primitive
-representations of ADTs and {\color{blue} hereditarily polynomial}
-nested types by validating the aforementioned short cut fusion
-transformation for them.
-
-{\color{blue} Say somewhere: A relation is a triple $(A,B,R)$, where
-  $R$ is a relation between sets $A$ and $B$.  A function from
-  $(A,B,R)$ to $(A',B',R')$ is a pair $(f : A \to A',g : B \to B')$ of
-  functions on sets such that $(f a,g\,b) \in R'$ if $(a,b) \in R$. We
-  may write $R : \rel(A,B)$ for $(A,B,R)$.  If $R : \rel(A,B)$ we
-  write $\pi_1 R$ and $\pi_2 R$ for the {\em domain} $A$ of $R$ and
-  the {\em codomain} $B$ of $R$, respectively, and assume $\pi_1$ and
-  $\pi_2$ are surjective. We write $\Eq_A = (A,A,\{(x,x)~|~ x \in
-  A\})$ for the {\em equality relation} on the set $A$. ALSO, WANT OUR
-RELATIONS TO BE ONTO.}
-
-{\color{red} This might not be relevant, since we only care about the
-  fibredness.} 
-The key idea underlying Reynolds' parametricity is to give each type
-$F(\alpha)$ with one free variable $\alpha$ a {\em set interpretation}
-$F_0$ taking sets to sets and a \emph{relational interpretation} $F_1$
-taking relations $R : \rel(A,B)$ to relations $F_1 (R) : \rel(F_0 (A),
-F_0 (B))$, and to interpret each term $t(\alpha,x) : F(\alpha)$ with
-one free term variable $x : G(\alpha)$ as a map $t_0$ associating to
-each set $A$ a function $t_0(A) : G_0(A) \to F_0(A)$. These
-interpretations are given inductively on the structures of $F$ and $t$
-in such a way that they imply two fundamental theorems. The first is
-an \emph{Identity Extension Lemma}, which states that $F_1(\Eq_A) =
-\Eq_{F_0(A)}$, and is the essential property that makes a model
-relationally parametric rather than just induced by a logical
-relation. The second is an \emph{Abstraction Theorem}, which states
-that, for any $R :\rel(A, B)$, $(t_0(A),t_0(B))$ is a morphism in
-$\rel$ from $(G_0(A),G_0(B),G_1(R))$ to $(F_0(A),F_0(B),F_1(R))$. The
-Identity Extension Lemma is similar to the Abstraction Theorem except
-that it holds for {\em all} elements of a type's interpretation, not
-just those that interpret terms.  Similar theorems are required for
-types and terms with any number of free variables.
+beyond just naturality, such as proving equivalence of Church
+encodings and primitive representations of ADTs and {\color{blue}
+  hereditarily polynomial} nested types by validating short cut fusion
+for them.
+
+To discuss relational parametricity will need to interpret data types
+not just in $\mathit{Set}$, but in a suitable category of relations as
+well. The following definition is standard:
+\begin{definition}
+  The category $\mathit{Rel}$ has:
+  \begin{itemize}
+\item objects: A relation is a triple $(A,B,R)$, where $R$ is a
+  relation between sets $A$ and $B$.
+\item morphisms: A morphism from $(A,B,R)$ to $(A',B',R')$ is a pair
+  $(f : A \to A',g : B \to B')$ of functions in $\mathit{Set}$ such
+  that $(f a,g\,b) \in R'$ if $(a,b) \in R$.
+\item identities: The identity morphism on $(A,B,R)$ is the pair
+  $(\mathit{id}_A : A \to A, \mathit{id}_B : B \to B)$.
+\item composition: Composition is the componentwise composition in
+  $\mathit{Set}$. 
+\end{itemize}
+\end{definition}
+\noindent
+We write $R : \mathit{Rel}\,(A,B)$ for $(A,B,R) \in \mathit{Rel}$.  If
+$R : \mathit{Rel}\,(A,B)$ then we write $\pi_1 R$ and $\pi_2 R$ for
+the {\em domain} $A$ of $R$ and the {\em codomain} $B$ of $R$,
+respectively. Below, we assume $\pi_1$ and $\pi_2$ are surjective.  We
+write $\mathit{Eq}_A = (A,A,\{(x,x)~|~ x \in A\})$ for the {\em
+  equality relation} on the set $A$.
+
+The key idea underlying parametricity is to give each type
+$\mathsf{G[a]}$ with one free variable $\mathsf{a}$ a {\em set
+  interpretation} $G_0$ taking sets to sets and a \emph{relational
+  interpretation} $G_1$ taking relations $R : \mathit{Rel}\,(A,B)$ to
+relations $G_1 \,R : \mathit{Rel}\,(G_0 \,A, G_0 \,B)$, and to
+interpret each term $\mathsf{t\,(a,x) :: G[a]}$ with one free term
+variable $\mathsf{x :: F[a]}$ as a function $t$ associating to each
+set $A$ a morphism $t \,A : F_0\,A \to G_0\,A$ in $\mathit{Set}$.
+These interpretations are given inductively on the structures of
+$\mathsf{G}$ and $\mathsf{t}$ in such a way that they imply two
+fundamental theorems. The first is an \emph{Identity Extension Lemma},
+which states that $G_1\,\mathit{Eq}_A = \mathit{Eq}_{G_0 A}$, and is
+the essential property that makes a model relationally parametric
+rather than just induced by a logical relation.  The second is an
+\emph{Abstraction Theorem}, which states that, for any $R
+:\mathit{Rel}\,(A, B)$, $(t\, A, t\,B)$ is a morphism in
+$\mathit{Rel}$ from $(F_0\,A,F_0\,B,F_1\,R)$ to
+$(G_0\,A,G_0\,B,G_1\,R)$. The Identity Extension Lemma is similar to
+the Abstraction Theorem except that it holds for {\em all} elements of
+a type's interpretation, not just those that interpret terms.  Similar
+theorems are required for types and terms with any number of free
+variables.
+
+The existence of parametric models that interpret GADTs as the
+interpretations of their Church encodings follows from, e.g., the
+existence of the parametric model of $F_\omega$ constructed
+in~\cite{atk12}.  In that model, {\color{blue} ``parallel
+  interpretations'' of $Rel$ over $Set$...}.  The parametricity
+property for a GADT is therefore...  {\color{blue} SAY MORE!} In
+particular, the parametricity property for the GADT $\mathsf{G}$
+from~\eqref{eq:G} is exactly the syntactic counterpart of the
+inhabitation result for the interpretation of $\mathsf{G}$'s Church
+encoding, as the following example shows.
 
 \begin{example}
-{\color{red} PRIMITIVE REP DOESN'T HAVE FIBRED SEMANTICS, AS REQUIRED
-  FOR PAR} To
-even consider the question of whether our semantics is parametric we
-first need to define a relational semantics for our types, i.e., a
-relational interpretation for every type $G : \ast \to \ast$ such
-that, for every relation $R : \rel(A, B)$, we have $G_1(R) :
-\rel(G_0(A), G_0(B))$.
-
-We shall show that this is not generally possible for the primitive
-representation of data types, by considering the previously defined
-type $G$ in Equation~\ref{eq:G}.  Recall that the semantic
-interpretation of $G$ is given by $\mathsf{\Lan}_1 1$ which, in fact,
-is none other than the identity functor on $\set$.
-
-Analogously to the set interpretation, the relational interpretation
-for $G$ is given by $\mathsf{\Lan}_{\Eq_1} \Eq_1$.  Notice that the
-interpretation still essentially relies on a left Kan extension, but
-this time it is an extension of $\rel$-valued functors.  We would need
-to prove that, for every relation $R : \rel(A, B)$, we have that
-$(\mathsf{\Lan}_{\Eq_1} \Eq_1) R$ is a relation on $(\mathsf{\Lan}_1
-1) A$ and $(\mathsf{\Lan}_1 1)B$, i.e., $A$ and $B$.  Instead, we will
-provide a counterexample showing that this is not true.
+Let $\mathsf{t}$ be a term of type $\mathsf{G}$ for the GADT
+$\mathsf{G}$ defined in~\eqref{eq:G}, let $G = (G_0,G_1)$ be the
+interpretation of the Church encoding of $\mathsf{G}$, let $t$ be the
+interpretation of $\mathsf{t}$, and let $R :
+\mathit{Rel}\,(A,B)$. Then $t\, A \in G_0\,A$ and $t\,B \in G_0\,B$
+and, by the Abstraction Theorem, $t\,A$ and $t\,B$ are related in
+$G_1\,R$. However, {\color{blue} using the semantic interpretations
+  in~\cite{atk12} --- EXPLAIN!}, we have that the relational
+interpretation $G_1\,R$ of $\mathsf{G}$ is the empty relation whenever
+$R$ differs from the relational interpretation $\mathit{Eq}\,1$ of
+$\mathsf{1}$.  {\color{blue} $Eq$ vs. $Equal$?}  We thus deduce that
+there can be no term of type $\mathsf{G\,a}$ unless $\mathsf{a}$ is
+semantically equivalent to $\mathsf{1}$.
+\end{example}
+
+For primitive representations of GADTs the story is completely
+different. First off, we have that {\color{blue} the interp of Lan
+  must be full Lan}. Indeed,... {\color{blue} MORE!!}
+\begin{proposition}\label{prop:full-lans}
+Full Lan prop goes here. Explain cutting down.
+\end{proposition}
+\noindent
+In particular, both the set interpretation and the relational
+interpetation of a GADT must involve the entireties of any Lans in
+their fixpoint representations. We therefore have:
+
+\begin{example}
+In any parametric model we must give both a set interpretation and a
+relational interpretion for every type as described at the start of
+this section. That is, for every GADT $\mathsf{G}$ we must give an
+interpretation $G = (G_0,G_1)$ such that, for every relation $R :
+\mathit{Rel}\,(A, B)$, we have $G_1\,R : \mathit{Rel}\,(G_0\,A,
+G_0\,B)$. But, as we now show, this is not generally possible for in
+models that interpret GADTs as the interpretations of their primitive
+representations.
+
+Consider again the GADT $\mathsf{G}$ given by~\eqref{eq:G}.  Applying
+Proposition~\ref{prop:full-lans} first in $\mathit{Set}$ and then in
+$\mathit{Rel}$, we have that the set interpretation of $\mathsf{G}$ is
+$\mathit{Lan}_{\lambda *.1}\,\lambda *.1$, i.e., is exactly the
+identity functor on $\mathit{Set}$ {\color{blue} Tie back to previous
+  calculation}. By the exact same reasoning, this time in
+$\mathit{Rel}$ rather than in $\mathit{Set}$, the relational
+interpretation of $\mathsf{G}$ is $\mathit{Lan}_{\lambda
+  *.\mathit{Eq}_1} \,\lambda *\!.\,\mathit{Eq}_1$, where $\lambda
+*\!.\,\mathit{Eq}_1$ is the constantly $\mathit{Eq}_1$-valued functor
+from the category with a single object $*$ to $\mathit{Rel}$.  Indeed,
+the interpretation is still a left Kan extension, but now it is the
+left Kan extension determined by the functor interpreting the type
+$\mathsf{\lambda *. 1}$ in $\mathit{Rel}$. To prove that the Identity
+Extension Lemma holds we need to show that, for every relation $R :
+\mathit{Rel}\,(A, B)$, we have that $(\mathit{\Lan}_{\lambda
+  *\!.\,\mathit{Eq}_1} \,\lambda *\!.\,\mathit{Eq}_1)\, R$ is a
+relation between the sets $(\mathit{Lan}_{\lambda *. 1}\,\lambda
+*\!. 1)\, A$ and $(\mathit{Lan}_{\lambda *. 1}\,\lambda *\!. 1)\,B$,
+i.e., between the sets $A$ and $B$.  Instead, we will give a
+counterexample showing that this is not the case.
+
+Consider the relation $R = (1, 2, 1 \times 2)$. We expect
+$(\mathit{\Lan}_{\lambda *\!.\,\mathit{Eq}_1} \lambda
+*\!.\,\mathit{Eq}_1)\, R$ to be a relation with domain {\color{blue}
+  $1$}. Since left Kan extensions preserve projections~\cite{rie16},
+we can compute the domain as
+\[\pi_1\,\big( (\mathit{\Lan}_{\lambda *\!.\,\mathit{Eq}_1} \lambda
+*\!.\,\mathit{Eq}_1) \, R \big)\;=\; (\mathsf{\Lan}_{\lambda
+  *\!.\,\mathit{Eq}_1} \lambda *\!. \,1)\, 1\]
+
+(notice that the left
+Kan extension of $1 : \ast \to \set$ along $\Eq_1 : \ast \to \rel$ is
+a functor $\rel \to \set$).
 
@@
 
-We still get naturality properties, though.
-
 \vspace*{0.2in}
 
-As we have seen in the previous Section~\ref{sec:adts}, GADTs
-represented as Church encodings do not necessarily have a map or a
-functorial semantics.  In particular, this means that we cannot
-possibly derive naturality results for those types, event though their
-semantics is parametric.  Indeed, even though naturality is sometimes
-seen as a consequence of parametricity, that is only true in case we
-have a functorial semantics and types supporting a map.  Still, GADTs
-represented as Church encodings can still exhibit consequences of
-paramentricity other than naturality, as they do admit parametric
-semantics~\cite{atk12}.
-
-\begin{example}
-{\color{blue} In this paragraph, $\mathsf{G}$ and $\mathsf{t}$
-  represent syntax, so should be sans-serif, but in the notes, the
-  paragraph about set/relational interpretation and
-  parametricity/IEL/AT uses serif font.} For example, consider the
-Church encoding of the GADT $\mathsf{G}$ defined in
-Equation~\ref{eq:G}.  If $\mathsf{t}$ is a term of type
-$\mathsf{G}$ and $R$ is a relation on the sets $A$ and $B$, then
-$t_0(A) \in G_0(A)$ and $t_0(B) \in G_0(B)$, and by the abstraction
-theorem we have that $t_0(A)$ and $t_0(B)$ are related in $G_1(R)$.
-{\color{blue} Here, $G = (G_0,G_1)$ is the interpretation of
-  $\mathsf{G}$.}  We also know that $G_1(R)$ is the empty relation
-whenever $R$ is not $\mathit{Equal}\,1$.  We thus deduce that there
-can be no term of type $\mathsf{G\,a}$ unless $\mathsf{a}$ is
-semantically equivalent to $\mathsf{1}$.  This result is the syntactic
-counterpart of the inhabitation result for $\mathsf{G}$ {\color{blue}
-  are inhabitation results about semantics or syntax?}  showing that
-all of the instances of $\mathsf{G}$ are uninhabited, with the
-exception of the instance for $\mathsf{1}$.  Of course, we could not
-possibly have a term of an uninhabited type.
-\end{example}
+Although there can exist no parametric models interpreting GADTs as
+their primitive representations, we still get naturality properties
+for prim reps. On the other hand, although a GADT represented as a
+Church encoding does not necessarily have a map or a functorial
+semantics, so that we cannot possibly derive naturality results for
+them, parametric models interpreting GADTs as their Church encodings
+do exist.  This shows that, even though naturality is generally
+regarded as a consequence of parametricity, this is only true when we
+have a functorial semantics and types supporting a map.
 
 %%% Primitive representation of GADT is non parametric, but functorial
-On the contrary, as we have shown in Section~\ref{sec:gadts}, a
-GADT represented as a primitive data type does not necessarily admit a
-parametric semantics, but, as a fixpoint of a higher-order functor, it
-does admit a functorial semantics and, consequently, a functorial map.
-In particular, every type that has a functorial map also benefits from
-naturality properties.  In other words, there is a suitable
-rearrange-transform property as explained in Section~\ref{sec:gadts}.
-
-%%% Example with G
-For example, $\mathsf{G}$ is associated to a functorial map,
-$\map_{\mathsf{G}}$, such that, for every function $\mathsf{f : a \to
-  b}$, we get a function $\map_{\mathsf{G}}\,\mathsf{f} : \mathsf{G\,a
-  \to G\,b}$.  Then we have a term $\map_{\mathsf{G}}\,\mathsf{f\,c} :
-\mathsf{G\,a}$ for every function $\mathsf{f : \mathsf{1} \to a}$,
-where $\mathsf{c : G\,\mathsf{1}}$ is the constructor provided by the
-definition of $\mathsf{G}$.
-
-We can read our counterexample as showing that parametricity results
-for calculi representing GADTs as their Church encodings cannot be
-extended to calculi that represent GADTs as primitive data types. And
-functoriality results for calucli representing GADTs as primitive
-cannot be extended to calculi that represent GADTs as their Church
-encodings.  So the user is forced to make a choice between naturality
-consequences of parametricity and other consequences of
+%On the contrary, as we have shown in Section~\ref{sec:gadts}, a
+%GADT represented as a primitive data type does not necessarily admit a
+%parametric semantics, but, as a fixpoint of a higher-order functor, it
+%does admit a functorial semantics and, consequently, a functorial map.
+%In particular, every type that has a functorial map also benefits from
+%naturality properties.  In other words, there is a suitable
+%rearrange-transform property as explained in Section~\ref{sec:gadts}.
+
+We can read the results of this section as showing that parametricity
+results for calculi representing GADTs as their Church encodings
+cannot be extended to calculi that represent GADTs as primitive data
+types. And functoriality results for calucli representing GADTs as
+primitive cannot be extended to calculi that represent GADTs as their
+Church encodings.  So the user is forced to make a choice between
+naturality consequences of parametricity and other consequences of
 parametricity. If GADTs really did generalize ADTs, and really were
@@
 would not have to choose.
+
+\vspace*{0.2in}
+
+NOTES:
+
+{\color{violet} TRUE????}
+FOR ADTs and Nesteds, it does not depend. We get the same par property
+whether they are seen as primitives or we use their CE.
 
@@
 
-\vspace*{1in}
-
-NOTES:
-
-{\color{violet} TRUE????}
-FOR ADTs and Nesteds, it does not depend. We get the same par property
-whether they are seen as primitives or we use their CE.
+\pagebreak
 

### 2021-03-01 20-gadts/GADT-story.tex
small revisions

@@
 \mathsf{data\,G\,a\,where}\\
-\mathsf{\;\;\;\;\;\;\;\;c : G\,1}
+\mathsf{\;\;\;\;\;\;\;\;c :: G\,1}
 \end{array}
 \end{equation}
-{\color{blue} Use $\top$ instead of $1$ throughout?} includes elements
-at any instance $\mathsf{G\,a}$ for any type $\mathsf{a} \not =
-\mathsf{0}$ since there is always a function from $\mathsf{1}$ to such
-an $\mathsf{a}$, but does not include any terms at instance
+Here, $\mathsf{1}$ is the unit type, whose single element is also
+denoted by $\mathsf{1}$, includes elements at any instance
+$\mathsf{G\,a}$ for any type $\mathsf{a}$ that is not the empty type
+$\mathsf{0}$ since there is always a function from $\mathsf{1}$ to
+such an $\mathsf{a}$, but does not include any terms at instance
 $\mathsf{G\,0}$. Indeed, $\mathsf{G\,0}$ is not inhabited via the
@@
 functorial completion because it is not possible to define a function
-from $\mathsf{1}$ into $\mathsf{0}$. {\color{blue} Use $\bot$ instead
-  of $0$ here?}
+from $\mathsf{1}$ into $\mathsf{0}$.
 
@@
 \end{array}\]
-would allow us to construct an element 
-\[\begin{array}{l}
-\mathsf{emptyFromMapEquala :: (\forall b\, c.\, (b \to c) \to Equal\, a\,
-  b \to Equal\, a \,c) \to Void}\\ 
-\mathsf{emptyFromMapEquala \,map_{Equal\,a} = eqElim\, (map_{Equal\,a}\,
-  absurd\, Refl)\, ()}  
-\end{array}\]
-of type $\mathsf{Void}$. {\color{blue} Use same type conventions as in
-  Intro.} But this should be impossible. It is immediate that a GADT
-that is not actually a nested type cannot support a $\mathsf{map}$
-function (or a rearrange-transform property) either. In the case of
-$\mathsf{Seq}$, this is already intuitively evident from its Haskell
-definition as discussed in Section~\ref{sec:intro}.
+would allow us to construct
+an element
+\[\mathsf{eqElim\, (map_{Equal\,Void}\, absurd\, Refl)\, 1}\]
+of type $\mathsf{Void}$, where $\mathsf{absurd :: 0 \to c}$ is the
+syntactic reflection of the empty function from the empty set into the
+interpretation of $\mathsf{c}$. {\color{blue} Is there a way to say
+  this without passing to semantics?} But this should be
+impossible. It is immediate that a GADT that is not actually a nested
+type cannot support a $\mathsf{map}$ function (or a
+rearrange-transform property) either. In the case of $\mathsf{Seq}$,
+this is already intuitively evident from its Haskell definition as
+discussed in Section~\ref{sec:intro}.
 
@@
 \[H\,f\,a = a \,+\,(Lan_{\lambda c d. c \times d}\, \lambda
-c d. f c \times f d)\,a\]
-{\color{red} Check this equation against next displayed one below.}
-where $Lan_K \,F$ is the left Kan extension of
-the functor $F$ along the functor $K$. In general, the left Kan
-extension $Lan_K \,F : {\cal E} \to {\cal D}$ of $F : {\cal C} \to
-{\cal D}$ along $K : {\cal C} \to {\cal E}$ is the best functorial
-approximation to $F$ that factors through $K$. Here, ``best functorial
-approximation'' means that $Lan_K \,F$ is the smallest functor that
-both extends the image of $K$ to $\cal D$ and coincides with $F$ on
-$\cal C$, in the sense that, for any other such functor $H$, there is
-a morphism of functors (i.e., a natural transformation) from $Lan_K
-\,F$ to $H$. This is captured formally by the following
+c d. f c \times f d)\,a\] {\color{red} Check this equation against
+  next displayed one below.}  where $Lan_K \,F$ is the left Kan
+extension of the functor $F$ along the functor $K$. In general, the
+left Kan extension $Lan_K \,F : {\cal E} \to {\cal D}$ of $F : {\cal
+  C} \to {\cal D}$ along $K : {\cal C} \to {\cal E}$ is the best
+functorial approximation to $F$ that factors through $K$. Here, ``best
+functorial approximation'' means that $Lan_K \,F$ is the smallest
+functor that both extends the image of $K$ to $\cal D$ and agrees with
+$F$ on $\cal C$, in the sense that, for any other such functor $H$,
+there is a morphism of functors (i.e., a natural transformation) from
+$Lan_K \,F$ to $H$. This is captured formally by the following
 definition~\cite{mac71}:
@@
 interpreting $\mathsf{k}$. In this setting, the categories
-$\mathcal{C}$, $\mathcal{D}$, and $\mathcal{E}$ must all be
-$\mathit{Set}$. Using $\mathsf{Lan}$ we can then rewrite the type of a
-constructor $\mathsf{c :: f a \to g(k\,a)}$ as $\mathsf{c ::
-  (Lan_k\,f)\,a \to g\,a}$ since morphisms (i.e., natural
+$\mathcal{C}$, $\mathcal{D}$, and $\mathcal{E}$ must all be of theform $\mathit{Set}^n$ for some $n$. Using $\mathsf{Lan}$ we can then
+rewrite the type of a constructor $\mathsf{c :: f a \to g(k\,a)}$ as
+$\mathsf{c :: (Lan_k\,f)\,a \to g\,a}$ since morphisms (i.e., natural
 transformations) from $F$ to $G \circ K$ are in one-to-one
-corresondence with those from $\mathit{Lan}_K\,F$ to $G$ (see,
+corresondence with those from $\mathit{Lan}_K\,F$ to $G$ (see
 e.g.,~\cite{rie16}). That is, writing $F \Rightarrow G$ for the set of
@@
 effect is simply to test its argument for equality against the unit
-type $\mathsf{1}$. {\color{blue} Use same notation for unit type as in
-  intro.} The interpretation of the Church encoding of $\mathsf{G}$
-is thus, in accordance with Section~\ref{sec:CEs}, the function whose
-value is a singleton set when the interpretation of $\mathsf{a}$ is
-$1$ and the empty set otherwise. This perfectly accords with
-$\mathsf{G}$'s syntax, which indeed delivers just one data element
-when $\mathsf{a}$ is $\mathsf{1}$ and no elements otherwise.
+type $\mathsf{1}$. The interpretation of the Church encoding of
+$\mathsf{G}$ is thus, in accordance with Section~\ref{sec:CEs}, the
+function whose value is a singleton set when the interpretation of
+$\mathsf{a}$ is $1$ and the empty set otherwise. This perfectly
+accords with $\mathsf{G}$'s syntax, which indeed delivers just one
+data element when $\mathsf{a}$ is $\mathsf{1}$ and no elements
+otherwise.
 
@@
 \begin{proposition}\label{prop:full-lans}
-Full Lan prop goes here. Explain cutting down.
+Full Lan prop goes here. Explain cutting down. Or just argue intuitively?
 \end{proposition}
@@
 Consider the relation $R = (1, 2, 1 \times 2)$. We expect
-$(\mathit{\Lan}_{\lambda *\!.\,\mathit{Eq}_1} \lambda
-*\!.\,\mathit{Eq}_1)\, R$ to be a relation with domain {\color{blue}
-  $1$}. Since left Kan extensions preserve projections~\cite{rie16},
-we can compute the domain as
-\[\pi_1\,\big( (\mathit{\Lan}_{\lambda *\!.\,\mathit{Eq}_1} \lambda
-*\!.\,\mathit{Eq}_1) \, R \big)\;=\; (\mathsf{\Lan}_{\lambda
-  *\!.\,\mathit{Eq}_1} \lambda *\!. \,1)\, 1\]
-
-(notice that the left
-Kan extension of $1 : \ast \to \set$ along $\Eq_1 : \ast \to \rel$ is
-a functor $\rel \to \set$).
-
-Consider the relation $R = (1, \emptyset, \emptyset)$.  We would thus
-expect $(\mathsf{\Lan}_{\Eq_1} \Eq_1) R$ to be a relation with domain
-$1$.  Because the left Kan extension preserves
-projections~\cite{rie16} we can compute the domain as
-$(\mathsf{\Lan}_{\Eq_1} 1) R$ (notice that the left Kan extension of
-$1 : \ast \to \set$ along $\Eq_1 : \ast \to \rel$ is a functor $\rel
-\to \set$).
+$(\mathit{Lan}_{\lambda *\!.\,\mathit{Eq}_1} \lambda
+*\!.\,\mathit{Eq}_1)\, R$ to be a relation with domain $1$. Since left
+Kan extensions preserve projections~\cite{rie16}, we can compute the
+domain as
+\[\pi_1\,\big( (\mathit{Lan}_{\lambda *\!.\,\mathit{Eq}_1} \lambda
+*\!.\,\mathit{Eq}_1) \, R \big)\;=\; (\mathit{Lan}_{\lambda
+  *\!.\,\mathit{Eq}_1} \lambda *\!. \,1)\, 1\] (Note that the left Kan
+  extension of $\lambda *\!. \,1$ along $\lambda \!.\,\mathit{Eq}_1$ 
+ is a functor from $\mathit{Rel}$ to $\mathit{Set}$.)
+
+HERE!!!!!
 

### 2021-03-02 20-gadts/GADT-story.tex
finished draft

@@
 
-\title[Parametricity Consequences of GADTs
-  Representations]{Parametricity Consequences of GADTs
-  Representations}
+\title[Parametricity in the Presence of GADTs]{Parametricity in the
+  Presence of GADTs}
 
@@
 \begin{abstract}
-{\color{red} FIX ME!} GADTs are misnamed. They don't really generalize
-ADTs. And they don't behave the way we expect data types to behave. In
-particular, they force us to choose between the naturality properties
-we expect of their fmap functions and other parametricity results for
-them, causing us to question in what sense GADTs are data types, and
-what the value of the parametricity results for them as determined by
-their syntax really is.
+{\color{red} FIX ME!} There are two ways to represent/interpret
+GADTs. They are not the same, and how we view GADTs has profound
+consequences for how we program with, and reason about, them.
 \end{abstract}
@@
   throughout?}
+{\color{blue} We say "naturality consequences of parametricity"
+when in fact we have naturality without having any parametricity at all.
+Change the wording to indicate that naturality follows from
+functoriality of the interp.}
 
@@
 $\mathsf{G\,0}$. Indeed, $\mathsf{G\,0}$ is not inhabited via the
-syntactic specifiation of $\mathsf{G}$, and is not inhabited via
+syntactic specification of $\mathsf{G}$, and is not inhabited via
 functorial completion because it is not possible to define a function
@@
 declaration for $\mathsf{List\,a}$. We therefore have that
-$\mathsf{List\, a} = \mathsf{\mu F_{List\,a}}$ is modelled by $\mu
+$\mathsf{List\, a} = \mathsf{\mu F_{List\,a}}$ is modeled by $\mu
 F_{\mathit{List}\,a}$, where the functor $F_{\mathit{List}\,a}$ models
@@
   \times a \times X}$ as determined by~\eqref{eq:tree}, so that
-$\mathsf{Tree\, a} = \mathsf{ \mu F_{Tree\,a}}$ is modelled by $\mu
+$\mathsf{Tree\, a} = \mathsf{ \mu F_{Tree\,a}}$ is modeled by $\mu
 F_{\mathit{Tree}\,a}$, where $F_{\mathit{Tree}\,a} $ models
@@
 ADT actually denotes a semantic object computed as a real-live
-fixpoint, the semantic fixpoint calculation must converge.  If
-{\color{blue} Since? Establish this earlier?} we interpret our types
-as sets, then this means that the fixpoint being taken must be of a
-{\em functor} on the category $\mathit{Set}$ of sets and functions
-between them, rather than of a mere function between sets~\cite{tfca}.
-That is, the function $F$ interpreting the type constructor
-$\mathsf{F}$ constructing the body of a syntactic fixpoint must not
-only have an action on sets, but must also have a {\em functorial
-  action} on functions between sets. Reflecting this requirement back
-into syntax gives that $\mathsf{F}$ must support a
-suitable\footnote{The $\mathsf{map}$ function for $\mathsf{F}$ should
-  satisfy syntactic reflections of the functoriality properties in
-  $\mathit{Set}$ --- i.e., preservation of identity functions and
-  composition of functions --- even though there is no mechanism in
-  Haskell for enforcing this.} $\mathsf{map}$ function, i.e., must be
-an instance of Haskell's $\mathsf{Functor}$ class.
+fixpoint, the semantic fixpoint calculation must converge.  If, as is
+typical, we interpret our types as sets, then this means that the
+fixpoint being taken must be of a {\em functor} on the category
+$\mathit{Set}$ of sets and functions between them, rather than of a
+mere function between sets~\cite{tfca}.  That is, the function $F$
+interpreting the type constructor $\mathsf{F}$ constructing the body
+of a syntactic fixpoint must not only have an action on sets, but must
+also have a {\em functorial action} on functions between
+sets. Reflecting this requirement back into syntax gives that
+$\mathsf{F}$ must support a suitable\footnote{The $\mathsf{map}$
+  function for $\mathsf{F}$ should satisfy syntactic reflections of
+  the functoriality properties in $\mathit{Set}$ --- i.e.,
+  preservation of identity functions and composition of functions ---
+  even though there is no mechanism in Haskell for enforcing this.}
+$\mathsf{map}$ function, i.e., must be an instance of Haskell's
+$\mathsf{Functor}$ class.
 
@@
 A concrete example is given by the ADT $\mathsf{List\,a}$. This type
-is modelled as the fixpoint $\mu F_{\mathit{List\,a}}$ of the
+is modeled as the fixpoint $\mu F_{\mathit{List\,a}}$ of the
 first-order functor whose action on sets is given by
@@
 functions is given by $F_{\mathit{List}\,a}\,f = 1 + \id_a \times
-f$. The type constructor $\mathsf{List}$ is modelled by the functor
+f$. The type constructor $\mathsf{List}$ is modeled by the functor
 that is the fixpoint $\mu H$ of the higher-order functor $H$ whose
@@
 situations where different behaviors at different type instances is
-desired. This is achieved by allowing the progammer to give the type
+desired. This is achieved by allowing the programmer to give the type
 signatures of the different data constructors independently, and then
@@
 \noindent
-or, equivalently, represented by its own Church encoding
+or, equivalently, represented by its Church encoding
 \[\mathsf{Equal\,a\,b} = \mathsf{\forall f.\, (\forall c.\,
-  f\,c\,c) \to f\,a\,b}\]
-The GADT $\mathsf{Seq}$ can then be represented as its Church encoding
-\[\mathsf{Seq\,a} = \mathsf{\forall f.\, (\forall a.\,a \to f\, a)
-  \to (\forall a\,b\,c.\,Equal\,a\,(b\times c) \to f\,b \to f\,c \to
-  f\,a) \to f\,a}\]
-in a higher-kinded calculus such as $F_\omega$~\cite{atk12}.
+  f\,c\,c) \to f\,a\,b}\] in a higher-kinded calculus such as
+$F_\omega$~\cite{atk12}. Following the technique in Section~4.5
+of~\cite{atk12}, we can represent the GADT $\mathsf{Seq}$ as its
+Church encoding
+\[\mathsf{Seq\,a} = \mathsf{\forall f.\,(\forall d. \,(f\,d + \exists b
+  c.\, f\, b \times f \,c \times \mathsf{Equal}\, d \,(b \times c))
+  \to f \,d) \to f\, a}\]
+This is logically equivalent to the more intuitive encoding
+\[\mathsf{Seq\,a} = \mathsf{\forall f.\, (\forall d.\,d \to f\, d)
+  \to (\forall b\,c\,d.\,Equal\,d\,(b\times c) \to f\,b \to f\,c \to
+  f\,d) \to f\,a}\]
 
@@
 an element
-\[\mathsf{eqElim\, (map_{Equal\,Void}\, absurd\, Refl)\, 1}\]
-of type $\mathsf{Void}$, where $\mathsf{absurd :: 0 \to c}$ is the
-syntactic reflection of the empty function from the empty set into the
-interpretation of $\mathsf{c}$. {\color{blue} Is there a way to say
-  this without passing to semantics?} But this should be
-impossible. It is immediate that a GADT that is not actually a nested
-type cannot support a $\mathsf{map}$ function (or a
+\[\mathsf{eqElim\, (map_{Equal\,0}\, absurd\, Refl)\, 1}\]
+of the empty type $\mathsf{0}$, where $\mathsf{absurd :: 0 \to c}$ is
+the
+%syntactic reflection of the
+empty function from  $\mathsf{0}$
+%the empty type set
+into
+%the interpretation of
+$\mathsf{c}$.
+%{\color{blue} Is there a way to say this without passing to
+%semantics?}
+But this should be impossible. It is immediate that a GADT that is not
+actually a nested type cannot support a $\mathsf{map}$ function (or a
 rearrange-transform property) either. In the case of $\mathsf{Seq}$,
@@
 GADT's syntax. In this alternate reading, GADTs can, like ADTs, be
-modelled as fixpoints of higher-order functors. Syntactically, the use
+modeled as fixpoints of higher-order functors. Syntactically, the use
 of higher-order funct{\em ions} is essential; since the type arguments
@@
 its data constructor $\mathsf{Pair}$, $\mathsf{Seq}$ cannot be
-modelled as the fixpoint of any first-order functor. As shown
-in~\cite{jp19}, it can, however, be modelled as a solution to the
+modeled as the fixpoint of any first-order functor. As shown
+in~\cite{jp19}, it can, however, be modeled as a solution to the
 higher-order fixpoint equation
@@
 interpreting $\mathsf{k}$. In this setting, the categories
-$\mathcal{C}$, $\mathcal{D}$, and $\mathcal{E}$ must all be of theform $\mathit{Set}^n$ for some $n$. Using $\mathsf{Lan}$ we can then
+$\mathcal{C}$, $\mathcal{D}$, and $\mathcal{E}$ must all be of the form $\mathit{Set}^n$ for some $n$. Using $\mathsf{Lan}$ we can then
 rewrite the type of a constructor $\mathsf{c :: f a \to g(k\,a)}$ as
@@
 transformations) from $F$ to $G \circ K$ are in one-to-one
-corresondence with those from $\mathit{Lan}_K\,F$ to $G$ (see
+correspondence with those from $\mathit{Lan}_K\,F$ to $G$ (see
 e.g.,~\cite{rie16}). That is, writing $F \Rightarrow G$ for the set of
@@
 
+{\color{blue} Use $\mathsf{1}$ in place of $\mathsf{*}$ here?}
 To compute the interpretation of $\mathsf{G}$'s primitive
@@
 disturbing. What, they might ask, should a data type contain other
-than data that are constructed using its data constuctors? Why should
+than data that are constructed using its data constructors? Why should
 a GADT contain ``hidden'' elements that are not specified by the
@@
 
-\section{Parametricity for GADTs}\label{sec:par}
+\section{Parametricity in the Presence of GADTs}\label{sec:par}
 
@@
 $R : \mathit{Rel}\,(A,B)$ then we write $\pi_1 R$ and $\pi_2 R$ for
-the {\em domain} $A$ of $R$ and the {\em codomain} $B$ of $R$,
-respectively. Below, we assume $\pi_1$ and $\pi_2$ are surjective.  We
-write $\mathit{Eq}_A = (A,A,\{(x,x)~|~ x \in A\})$ for the {\em
-  equality relation} on the set $A$.
+the {\em domain} $A$ and {\em codomain} $B$ of $R$,
+respectively. Below, we assume $\pi_1$ and $\pi_2$ are surjective, so
+that the relations in $\mathit{Rel}$ are {\em onto}. {\color{blue} Say
+  why we care about onto?} We write $\mathit{Eq}_A = (A,A,\{(x,x)~|~ x
+\in A\})$ for the {\em equality relation} on the set $A$.
 
@@
 existence of the parametric model of $F_\omega$ constructed
-in~\cite{atk12}.  In that model, {\color{blue} ``parallel
-  interpretations'' of $Rel$ over $Set$...}.  The parametricity
-property for a GADT is therefore...  {\color{blue} SAY MORE!} In
+in~\cite{atk12}.  In that model, types are interpreted ``in parallel''
+in $\mathit{Set}$ and $\mathit{Rel}$ in the usual way, including the
+familiar ``cutting down'' of the interpretations of $\forall$-types to
+just those elements that are ``parametric''~\cite{rey83,wad89} to
+ensure that the Identity Extension Lemma holds. If {\color{blue} the
+  set interpretation of $\mathsf{Equal}$ is the function
+  $\mathit{Equal}$ and its relational interpretation is $\mathit{Eq}$}
+as intended, then the parametricity property for a GADT is an
+inhabitation result saying that the set interpreting any instance of
+that GADT contains exactly the interpretations of the data elements
+that can be formed using its data constructors and whose type is that
+instance.
+%Reflecting this property back into syntax gives that
+In terms of syntax the parametricity property for a GADT's Church
+encoding gives that the GADT is inhabited at any instance for which
+data elements can be formed using its data constructors. In
 particular, the parametricity property for the GADT $\mathsf{G}$
-from~\eqref{eq:G} is exactly the syntactic counterpart of the
-inhabitation result for the interpretation of $\mathsf{G}$'s Church
-encoding, as the following example shows.
-
-\begin{example}
+from~\eqref{eq:G} gives that $\mathsf{G\,a}$ contains a single data
+element if $\mathsf{a}$ is semantically equivalent to $\mathsf{1}$ and
+none otherwise:
+%gives exactly the expected
+%%syntactic counterpart to the
+%inhabitation result for the interpretation of $\mathsf{G}$'s Church
+%encoding:
+
+\begin{example}\label{ex:CE-par}
 Let $\mathsf{t}$ be a term of type $\mathsf{G}$ for the GADT
@@
 \mathit{Rel}\,(A,B)$. Then $t\, A \in G_0\,A$ and $t\,B \in G_0\,B$
-and, by the Abstraction Theorem, $t\,A$ and $t\,B$ are related in
-$G_1\,R$. However, {\color{blue} using the semantic interpretations
-  in~\cite{atk12} --- EXPLAIN!}, we have that the relational
-interpretation $G_1\,R$ of $\mathsf{G}$ is the empty relation whenever
-$R$ differs from the relational interpretation $\mathit{Eq}\,1$ of
-$\mathsf{1}$.  {\color{blue} $Eq$ vs. $Equal$?}  We thus deduce that
-there can be no term of type $\mathsf{G\,a}$ unless $\mathsf{a}$ is
-semantically equivalent to $\mathsf{1}$.
+and, by the Abstraction Theorem (Theorem~3) in~\cite{atk12}, $t\,A$
+and $t\,B$ must be related in $G_1\,R$. However, under the semantics
+given in~\cite{atk12}, {\color{blue} which includes the aforementioned
+  interpretations of $\mathsf{Equal}$}, the relational interpretation
+$G_1\,R$ of $\mathsf{G}$ is {\color{blue} $\mathit{Eq}_1$ when $R$ is
+  the relational interpretation $\mathit{Eq}_1$ of $\mathsf{1}$, and
+  the empty relation whenever $R$ differs from
+  %the relational interpretation
+  $\mathit{Eq}_1$
+  %of $\mathsf{1}$
+  }. Reflecting back into syntax we
+deduce that there can be no term in the type that is the Church
+encoding of $\mathsf{G\,a}$ unless $\mathsf{a}$ is semantically
+equivalent to $\mathsf{1}$.
 \end{example}
@@
 For primitive representations of GADTs the story is completely
-different. First off, we have that {\color{blue} the interp of Lan
-  must be full Lan}. Indeed,... {\color{blue} MORE!!}
-\begin{proposition}\label{prop:full-lans}
-Full Lan prop goes here. Explain cutting down. Or just argue intuitively?
-\end{proposition}
-\noindent
-In particular, both the set interpretation and the relational
-interpetation of a GADT must involve the entireties of any Lans in
-their fixpoint representations. We therefore have:
-
-\begin{example}
+different. If, as intended, the set interpretation of
+$\mathsf{Lan_k\,f}$ is $\mathit{Lan}_K\,F$, where $K$ is the set
+interpretation of $\mathsf{k}$ and $F$ is the set interpretation of
+$\mathsf{f}$, then the exact same reasoning give that the relational
+interpretation of $\mathsf{Lan_k\,f}$ is $\mathit{Lan}_K\,F$, where
+$K$ is the relational interpretation of $\mathsf{k}$ and $F$ is the
+relational interpretation of $\mathsf{f}$. But under these
+interpretations there can be no parametric model, as the following
+example shows.
+
+\begin{example}\label{ex:prim-par}
 In any parametric model we must give both a set interpretation and a
-relational interpretion for every type as described at the start of
-this section. That is, for every GADT $\mathsf{G}$ we must give an
-interpretation $G = (G_0,G_1)$ such that, for every relation $R :
+relational interpretation for every type as described at the start of
+this section. In particular, for every GADT $\mathsf{G}$ we must give
+an interpretation $G = (G_0,G_1)$ such that, for every relation $R :
 \mathit{Rel}\,(A, B)$, we have $G_1\,R : \mathit{Rel}\,(G_0\,A,
-G_0\,B)$. But, as we now show, this is not generally possible for in
-models that interpret GADTs as the interpretations of their primitive
+G_0\,B)$. But, as we now show, this is not generally possible when
+GADTs are interpreted as the interpretations of their primitive
 representations.
 
-Consider again the GADT $\mathsf{G}$ given by~\eqref{eq:G}.  Applying
-Proposition~\ref{prop:full-lans} first in $\mathit{Set}$ and then in
-$\mathit{Rel}$, we have that the set interpretation of $\mathsf{G}$ is
-$\mathit{Lan}_{\lambda *.1}\,\lambda *.1$, i.e., is exactly the
-identity functor on $\mathit{Set}$ {\color{blue} Tie back to previous
-  calculation}. By the exact same reasoning, this time in
-$\mathit{Rel}$ rather than in $\mathit{Set}$, the relational
+Consider again the GADT $\mathsf{G}$ given by~\eqref{eq:G}. The set
 interpretation of $\mathsf{G}$ is $\mathit{Lan}_{\lambda
-  *.\mathit{Eq}_1} \,\lambda *\!.\,\mathit{Eq}_1$, where $\lambda
-*\!.\,\mathit{Eq}_1$ is the constantly $\mathit{Eq}_1$-valued functor
-from the category with a single object $*$ to $\mathit{Rel}$.  Indeed,
-the interpretation is still a left Kan extension, but now it is the
-left Kan extension determined by the functor interpreting the type
-$\mathsf{\lambda *. 1}$ in $\mathit{Rel}$. To prove that the Identity
-Extension Lemma holds we need to show that, for every relation $R :
-\mathit{Rel}\,(A, B)$, we have that $(\mathit{\Lan}_{\lambda
-  *\!.\,\mathit{Eq}_1} \,\lambda *\!.\,\mathit{Eq}_1)\, R$ is a
-relation between the sets $(\mathit{Lan}_{\lambda *. 1}\,\lambda
-*\!. 1)\, A$ and $(\mathit{Lan}_{\lambda *. 1}\,\lambda *\!. 1)\,B$,
-i.e., between the sets $A$ and $B$.  Instead, we will give a
-counterexample showing that this is not the case.
-
-Consider the relation $R = (1, 2, 1 \times 2)$. We expect
+  *\!. 1}\,\lambda *\!. 1$, i.e., is, by the reasoning of
+Example~\ref{ex:g1}, the identity functor on $\mathit{Set}$. By the
+exact same reasoning, this time in $\mathit{Rel}$ rather than in
+$\mathit{Set}$, the relational interpretation of $\mathsf{G}$ is
+$\mathit{Lan}_{\lambda *\!.\,\mathit{Eq}_1} \,\lambda
+*\!.\,\mathit{Eq}_1$, where $\lambda *\!.\,\mathit{Eq}_1$ is the
+constantly $\mathit{Eq}_1$-valued functor from the category with a
+single object $*$ to $\mathit{Rel}$.  Indeed, the interpretation is
+still a left Kan extension, but now it is the left Kan extension
+determined by the functor interpreting the type $\mathsf{\lambda
+  *\!. 1}$ in $\mathit{Rel}$. For the Identity Extension Lemma to hold
+we would need that, for every relation $R : \mathit{Rel}\,(A, B)$, we
+have that $(\mathit{\Lan}_{\lambda *\!.\,\mathit{Eq}_1} \,\lambda
+*\!.\,\mathit{Eq}_1)\, R$ is a relation between the sets
+$(\mathit{Lan}_{\lambda *. 1}\,\lambda *\!. 1)\, A$ and
+$(\mathit{Lan}_{\lambda *. 1}\,\lambda *\!. 1)\,B$, i.e., between the
+sets $A$ and $B$. However, we now give a counterexample showing that
+this need not be the case.
+
+Consider the relation $R = (1, 2, 1 \times 2)$, where $1 \times 2$ is
+the total relation on $1$ and $2$, and thus is onto. We expect
 $(\mathit{Lan}_{\lambda *\!.\,\mathit{Eq}_1} \lambda
@@
 *\!.\,\mathit{Eq}_1) \, R \big)\;=\; (\mathit{Lan}_{\lambda
-  *\!.\,\mathit{Eq}_1} \lambda *\!. \,1)\, 1\] (Note that the left Kan
-  extension of $\lambda *\!. \,1$ along $\lambda \!.\,\mathit{Eq}_1$ 
- is a functor from $\mathit{Rel}$ to $\mathit{Set}$.)
-
-HERE!!!!!
-
-Following~\cite{blw03} we know that $(\mathsf{\Lan}_{\Eq_1} 1) R$ is
-the set of all morphisms of relations $f : \Eq_1 \to R$ modulo an
-equivalence relation; but there can be no morphism of relations $\Eq_1
-\to R$ because its second component would have to be a function $1 \to
-\emptyset$, and there can be no such function.  Thus,
-$(\mathsf{\Lan}_{\Eq_1} 1) R = \emptyset$, which is not the domain we
-would have expected.
+ *\!.\,\mathit{Eq}_1} \lambda *\!. \,1)\, R\] (Note that the left Kan
+ extension of $\lambda *\!. \,1$ along $\lambda *\!.\,\mathit{Eq}_1$
+ is a functor from $\mathit{Rel}$ to $\mathit{Set}$.) By
+ Proposition~7.1 of~\cite{blw03} the left Kan extension on the
+ right-hand side can be computed as
+\[ \big( \bigcup_{m\, :\, \mathit{Eq}_1 \to R} 1 \big) \; /\; \approx\]
+where $\approx$ is the smallest equivalence relation such that $(\ast,
+m) \approx (\ast, m')$ for $m, m' : \mathit{Eq}_1 \to R$ if there
+exists $f : \mathit{Eq}_1 \to \mathit{Eq}_1$ such that $f (\ast, \ast)
+= (\ast, \ast)$ and the following diagram commutes:
+\[
+\begin{tikzcd}
+\mathit{Eq}_1 \ar[rr, "{f}"] \ar[dr, "m"']
+&&\mathit{Eq}_1 \ar[dl, "m'"] \\
+&R
+\end{tikzcd}
+\]
+The only morphism $f : \mathit{Eq}_1 \to \mathit{Eq}_1$ is $f = (!,
+!)$. Since this morphism does satisfy the condition $f (\ast, \ast) =
+(\ast, \ast)$, we have that $(\ast, m) \approx (\ast, m')$ if and only
+if $m = m'$. Thus
+\[ \big( \bigcup_{m\, :\,
+  \mathit{Eq}_1 \to R} 1 \big) \; /\; \approx \;\;\; = \;\; \{m :
+\mathit{Eq}_1 \to R\} \;=\; \{(!, k_0), (!, k_1)\}\] where $k_0, k_1 :
+1 \to 2$ are the constantly $0$-valued and $1$-valued functions in
+$\mathit{Set}$, respectively, and therefore $\pi_1\,\big(
+(\mathit{Lan}_{\lambda *\!.\,\mathit{Eq}_1} \lambda
+*\!.\,\mathit{Eq}_1) \, R \big)$ is not $1$, as would be needed for
+the Identity Extension Lemma to hold.  Since the Identity Extension
+Lemma does not hold for models in which GADTs are interpreted by the
+interpretations of their primitive representations, such models cannot
+possibly be parametric.
 \end{example}
 
-The above example shows that the set interpretation and the relational
-interpretation for the primitive representation of a data type are
-inconsistent with each other, and thus it is not even possible to
-consider the question of whether the semantics is parametric.
-
-One could regard the above example as artificial, as it employs an
-empty relation on a non-empty domain.  Though, it is possible to apply
-the same argument, with a slightly more complex calculation, even to a
-total relation such as $(1, 2, 1 \times 2)$.
-
-\vspace*{0.2in}
-
-Although there can exist no parametric models interpreting GADTs as
-their primitive representations, we still get naturality properties
-for prim reps. On the other hand, although a GADT represented as a
-Church encoding does not necessarily have a map or a functorial
-semantics, so that we cannot possibly derive naturality results for
-them, parametric models interpreting GADTs as their Church encodings
-do exist.  This shows that, even though naturality is generally
-regarded as a consequence of parametricity, this is only true when we
-have a functorial semantics and types supporting a map.
-
-%%% Primitive representation of GADT is non parametric, but functorial
-%On the contrary, as we have shown in Section~\ref{sec:gadts}, a
-%GADT represented as a primitive data type does not necessarily admit a
-%parametric semantics, but, as a fixpoint of a higher-order functor, it
-%does admit a functorial semantics and, consequently, a functorial map.
-%In particular, every type that has a functorial map also benefits from
-%naturality properties.  In other words, there is a suitable
-%rearrange-transform property as explained in Section~\ref{sec:gadts}.
-
-We can read the results of this section as showing that parametricity
-results for calculi representing GADTs as their Church encodings
-cannot be extended to calculi that represent GADTs as primitive data
-types. And functoriality results for calucli representing GADTs as
-primitive cannot be extended to calculi that represent GADTs as their
-Church encodings.  So the user is forced to make a choice between
-naturality consequences of parametricity and other consequences of
-parametricity. If GADTs really did generalize ADTs, and really were
-data types in a normal sense, then these would coincide and the user
-would not have to choose.
-
-\vspace*{0.2in}
-
-NOTES:
-
-{\color{violet} TRUE????}
-FOR ADTs and Nesteds, it does not depend. We get the same par property
-whether they are seen as primitives or we use their CE.
-
-\begin{comment}
-{\color{violet} This is already in the paper, split in multiple places.}
-An analysis of the Eq A B type in Bob's systems shows that the set
-interpretation is empty whenever the A and B have different set
-interpretations.  Likewise, its relational interpretation is the empty
-relation whenever the A and B have different relational
-interpretations (even when the underlying set interpretations are the
-same).  The consequences of parametricity are thus immediate: there
-can be no term of type Eq A B if A and B are different types.
-Moreover, this behavior is fundamentally non-functorial, so that there
-is no hope of defining a map for Eq.  In particular, that means we
-cannot expect to retrieve non-trivial naturality properties for the Eq
-type, despite these are usually considered to be a special case of
-parametricity properties.  Indeed, the common understanding that
-parametricity generalizes naturality only makes sense for functorial
-types.  We can specialize these considerations to the type Eq 1 B,
-representing G, our example GADT defined by a unique constructor c ::
-G 1.
-\end{comment}
-
-\pagebreak
-
-\subsection{Comparison with Stephanie and Dmitrios}
-
-\subsection{Comparison with Bob's Parametric Model}
-
-GADTs don't really fit into his scheme because the two reps of GADTs
-aren't the same. Bob's stuff only works for functorial data types of
-any kind -- so not GADTs. So last section is misleading.
-
-Moreover, the consequences of parametricity that one has in Bob's
-system do not hold in our system, as we can construct terms of type
-$(Lan_1 1)$ B for instances of B other than 1.  Thus, either our
-system does not admit any parametric model, or, if it does, the
-consequences of parametricity are different.  That the consequences of
-parametricity depended on the model or the representation of GADTs is
-to be expected, however. (IT IS. CONSIDER INTERPRETING BASE TYPES AS
-ORDERED SETS. THIS GIVES DIFFERENT PARAMETRICITY PROPERTIES THAN
-INTERPRETING THEM AS DIAGONAL RELATIONS.)
-
-It is possible to give a parametric model that models GADTs
-represented as Church encodings~\cite{atk12}.  But given the
-non-uniformity captured by GADTs by purposeful design, that should be
-both completely unexpected and alarming! However, upon reflection it's
-not really all that surprising once we realize what the parametricity
-properties of GADTs represented as Church encodings look like. Bob
-requires map functions for fixpoint but then doesn't use them.
-
-Tellingly, Bob gives no parametricity properties for GADTs. If he had,
-it would have illustrated the phenomenon we are describing, raised the
-question: What good is a par property for a GADT or another type using
-the Eq type?
+Taken together, Examples~\ref{ex:CE-par} and~\ref{ex:prim-par} show
+that the parametricity properties of GADTs differ greatly depending on
+whether they are represented as their Church encodings or represented
+as primitive data types. This contrasts sharply with the fact that
+both ADTs and nested types have the same parametricity properties
+regardless of whether they are represented by their Church encodings
+or their primitive representations. {\color{blue} Double-check that
+  this last part is true.}
+
+\section{Conclusion}
+
+{\color{blue} Combine these two paragraphs.}
+
+We have shown that GADTs can be considered as data types in two
+contrasting ways: as completely determined by their syntax, in which
+case they are represented by their Church encodings, or as the
+functorial completion of their syntax, in which case they are
+represented as primitive data types.  But, regardless of which
+representation we choose, GADTs fail to be have the expected
+parametricity properties. Indeed, we can read the results of
+Section~\ref{sec:par} as showing that i) some standard parametricity
+results (e.g., inhabitation results) for calculi representing GADTs as
+their Church encodings cannot be extended to calculi that represent
+GADTs as primitive data types may not be possible, and ii) some
+naturality results (e.g., rearrange-transform properties) for calculi
+representing GADTs as primitive cannot be extended to calculi that
+represent GADTs as their Church encodings. So, with neither
+representation can we obtain {\em both} the naturality consequences of
+parametricity {\em and} the non-naturality consequences of
+parametricity. Intuitively speaking, this is because a GADT is not a
+data type in the usual sense of being a container, and completing it
+so that it becomes one destroys the relationally uniform behavior it
+must exhibit in a parametric model. Which representation of GADTs the
+user chooses can therefore have profound consequences for how they are
+able to program with, and reason about, them.
+
+All is not lost, however. Despite there being no parametric models for
+calculi incorporating GADTs as primitive data types, we still get all
+possible rearrange-transform properties for these types. This means
+that analogues of most of Wadler's free theorems for ADTs hold for
+GADTs, since almost all of the specific consequences of parametricity
+given in~\cite{wad89} (see especially Figure~1) are actually of this
+form for list types. In fact, even though such naturality properties
+are generally regarded as consequences of {\em parametricity}, they
+are actually consequences of data types' {\em functorial semantics}.
+On the other hand, despite there being no models interpreting type
+constructors as functors for calculi incorporating GADTs as Church
+encodings, and thus there being no models validating all
+rearrange-transform properties for these types, parametric models for
+such calculi do exist, and these can be used, e.g., to determine which
+instances of a GADT are inhabited. {\color{blue} Is this all?}
+
+{\color{blue} Comparison with Stephanie and Dmitrios? Comparison with
+  Bob's Parametric Model? Other conclusions????}
+
+\bibliography{gadt-story-references}
+
+\end{document}
 
@@
 functors, which, in Bob's system, requires the programmer to supply a
-map function on the intepretation of every GADT defined. Moreover,
-each such map function would have to behave functorially. Bob's system
-does indeed deliver non-trivial naturality results for every GADT for
-which the user can define a map function that behaves
-functorially. The problem is that there may not be many such GADTs (in
-fact it will not be possible for proper GADTs, i.e., GADTs that are
-not actually nested types).
+map function on the interpretation of every GADT defined.  (Bob
+requires map functions for fixpoint but then doesn't use them.)
+Moreover, each such map function would have to behave
+functorially. Bob's system does indeed deliver non-trivial naturality
+results for every GADT for which the user can define a map function
+that behaves functorially. The problem is that there may not be many
+such GADTs (in fact it will not be possible for proper GADTs, i.e.,
+GADTs that are not actually nested types).
 
@@
 data types since they describe only the data (and not the action on
-morphisms since they are not required to support map functions).  Our
-approach is different. By contrast with Bob's system, we restrict to
-GADTs generated by a specific grammar, and this grammar is designed
-specifically to guarantee functoriality of all data types definable in
-it. We also treat GADTs as primitive data types --- i.e., as fixpoints
-of (higher-order) functors --- rather than via their Church encodings
-since this more closely aligns with how ADTs are treated in
-implemented programming languages.
+morphisms since they are not required to support map functions).
+But Bob doesn't use the map function, and his Church encodings don't
+produce data types that support map functions.
 
@@
 the most useful consequences of parametricity for ADTs and nested
-types are actually natuality results (IS THIS TRUE? MAYBE NOT FOR
+types are actually naturality results (IS THIS TRUE? MAYBE NOT FOR
 GADTs), we'd be missing out on a very large class of expected and
@@
 parametricity results --- including naturality results --- that we can
-derive for ADTs. Yet, by design we cannot. So are GADTs actually
-misnamed? In what sense are they data types, and in what sense do they
-generalize ADTs? Perhaps what are now known as nested types should
-really be called GADTs since *they* are actually the (level-2)
-higher-kinded generalizations of ADTs.
-
-{\color{violet} We say "naturality consequences of parametricity"
-when in fact we have naturality without having any parametricity at all.
-Change the wording to indicate that naturality follows from
-functoriality of the interp.}
+derive for ADTs. Yet, by design we cannot.
+
+Tellingly, Bob gives no parametricity properties for GADTs.
+
 In our system we can show that we can recover the expected naturality
@@
 adds extra data to the data type), and second, our counterexample
-shows that we don't get non-naturality consequences of parmetricity
+shows that we don't get non-naturality consequences of parametricity
 even for the data types defined by our grammar.  So this isn't really
 what we wanted either.
-
-\pagebreak
-
-\begin{verbatim}
-
-Daniel:
-
-I have tried reconstructing my example using our simpler GADT with a single
-constructor c :: G 1. I don't think it is possible to modify my example for a
-GADT with only a single constructor. In particular, I don't see a way to
-exploit the map function unless we have constructors for G X and G Y where X
-!= Y.
-If we assume a map function exists, then given f :: 1 -> A we can produce
-(map f c) :: G A. But I don't know what to do from there.
-
-I was able to simplify the types in the example slightly though. The example
-I gave before can be adapted to use either
-
-data G3 b where
-    G3unit :: G3 ()
-    G3id    :: b -> G3 (Either b ())
-
-or
-
-data G5 b where
-    G5unit   :: G5 ()
-    G5void   :: Void -> G5 Bool
-
-as the problematic GADT.
-
-Enrico:
-
-Just to make sure I understand it correctly: when we define the function
-
-unGbool :: G c Bool -> c
-unGbool (Gbool x) = x
-
-we are doing so in Haskell, right?
-In our system, it would not be possible to define unGbool (because we could
-not define it only on the constructor Gbool of G). For this reason, this
-example provides a contradiction in Haskell (+functoriality), but not in our
-system. Is that correct?
-
-Daniel:
-
-The example is Haskell and relies on Haskell's rules for
-pattern matching.
-
-Enrico:
-
-By the way, as a general remark not directly related to the introduction you
-wrote, I think that the set semantics of our type G is
-
-[[G \alpha]]^set [\alpha := A] = A
-
-Indeed, (Lan_1 1) A = A for every set A. Intuitively this makes sense,
-because for every a : A there is a fuction
-
-k_a : 1 —> A
-
-and thus an element
-
-map k_a \ast : G A
-
-In particular, [[G 0]]^set = \emptyset.
-
-Still, it is NOT true that its relational interpretation is [[G \alpha]]^rel
-[\alpha := R] = R (and indeed the relational interpretation is not fibered
-over the set interpretation)!
-\end{verbatim}
-
-\section{Conclusion}
-
-The upshot is that we can consider GADTs as data types (i.e., we can
-compute the semantics of GADTs) in two ways: just as determined by
-their syntax, as when they are represented by their Church encodings,
-or as the functorial completion of their syntax, as obtained by
-supplying a suitable map construct, when they are represented as
-primitive data types. But, either way, GADTs fail to be well-behaved
-wrt parametricity. In the first case, GADTs don't have fmaps so their
-parametricity properties can't include naturality properties. In the
-second case, we still can't get non-naturality consequences of
-parametricity. That is, we can't get BOTH the naturality consequences
-of parametricity AND the non-naturality consequences of parametricity
-at the same time.
-
-\bibliography{gadt-story-references}
-
-\end{document}

### 2021-03-02 20-gadts/GADT-story.tex
updated draft

@@
 an instance of Haskell's $\mathsf{Functor}$ class. It is worth noting,
-however, that since the functorial completion of a GADT includes only
-those data elements that are derivable via repeated applications of
-$\mathsf{map}$ starting from those specified by its syntax, primitive
-GADTs need not, in general, be nested types. For example, the
-functorial completion of the GADT $\mathsf{G}$ defined by
+however, that since the functorial completion of a GADT includes
+{\color{blue} only?} those data elements that are derivable via
+repeated applications of $\mathsf{map}$ starting from those specified
+by its syntax, primitive GADTs need not, in general, be nested
+types. For example, the functorial completion of the GADT $\mathsf{G}$
+defined by
 \begin{equation}\label{eq:G}
@@
 GADT's syntax and are only accessible via applications of
-$\mathsf{map}$? A semanticist, on the other hand, would likely find
-the primitive representation of GADTs entirely reasonable. Indeed,
-they would likely find the nonfunctorial nature of a GADT's Church
-encoding unnerving at best. After all, they would likely argue, the
-data in a GADT shouldn't disappear just because you map a function
-over it! The fact that this happens when GADTs are represented by
-their Church encodings actually highlights how GADTs {\em do not}
-generalize the essential, container-ish nature of ADTs at all. A
-semanticist might therefore conclude that GADTs are seriously
-misnamed. They might even go on to wonder: in what sense do GADTs
-really deserve to be considered {\em data types} at all?
+$\mathsf{map}$ {\color{blue} and other elements, too?}? A semanticist,
+on the other hand, would likely find the primitive representation of
+GADTs entirely reasonable. Indeed, they would likely find the
+nonfunctorial nature of a GADT's Church encoding unnerving at
+best. After all, they would likely argue, the data in a GADT shouldn't
+disappear just because you map a function over it! The fact that this
+happens when GADTs are represented by their Church encodings actually
+highlights how GADTs {\em do not} generalize the essential,
+container-ish nature of ADTs at all. A semanticist might therefore
+conclude that GADTs are seriously misnamed. They might even go on to
+wonder: in what sense do GADTs really deserve to be considered {\em
+  data types} at all?
 
@@
 that can be formed using its data constructors and whose type is that
-instance.
-%Reflecting this property back into syntax gives that
-In terms of syntax the parametricity property for a GADT's Church
-encoding gives that the GADT is inhabited at any instance for which
-data elements can be formed using its data constructors. In
+instance. In terms of syntax the parametricity property for a GADT's
+Church encoding gives that the GADT is inhabited at any instance for
+which data elements can be formed using its data constructors. In
 particular, the parametricity property for the GADT $\mathsf{G}$
@@
 none otherwise:
-%gives exactly the expected
-%%syntactic counterpart to the
-%inhabitation result for the interpretation of $\mathsf{G}$'s Church
-%encoding:
 
@@
   the empty relation whenever $R$ differs from
-  %the relational interpretation
-  $\mathit{Eq}_1$
-  %of $\mathsf{1}$
-  }. Reflecting back into syntax we
-deduce that there can be no term in the type that is the Church
-encoding of $\mathsf{G\,a}$ unless $\mathsf{a}$ is semantically
-equivalent to $\mathsf{1}$.
+  $\mathit{Eq}_1$}. Reflecting back into syntax we deduce that there
+can be no term in the type that is the Church encoding of
+$\mathsf{G\,a}$ unless $\mathsf{a}$ is semantically equivalent to
+$\mathsf{1}$.
 \end{example}

### 2021-03-04 20-gadts/GADT-story.tex
almost final gadt-story

@@
 \acmVolume{1}
-\acmNumber{POPL} % CONF = POPL or ICFP or OOPSLA
+\acmNumber{Haskell Symposium} % CONF = POPL or ICFP or OOPSLA
 \acmArticle{1}
 \acmYear{2021}
-\acmMonth{1}
+\acmMonth{8}
 \acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
@@
 \begin{abstract}
-{\color{red} FIX ME!} There are two ways to represent/interpret
-GADTs. They are not the same, and how we view GADTs has profound
-consequences for how we program with, and reason about, them.
+GADTs can be represented either by their Church encodings {\em \`a la}
+Atkey or as primitive data types {\em \`a la} Johann and Polonsky.  A
+GADT represented as a Church encoding need not have a functorial
+semantics, i.e., need not support a $\mathsf{map}$ function. By
+contrast, the primitive prepresentation of a GADT must have a
+functorial semantics even to be well-defined. The two representations
+of a GADT thus need not be the same in general. In this paper we show
+that the choice of representation for GADTs has profound consequences
+for their parametricity properties, and thus for how we program with,
+and reason about, these advanced data types.
 \end{abstract}
@@
 \section{Introduction}\label{sec:intro}
-{\color{blue} map vs. fmap throughout.} {\color{blue} Add polynomial
-  throughout?}
-{\color{blue} We say "naturality consequences of parametricity"
-when in fact we have naturality without having any parametricity at all.
-Change the wording to indicate that naturality follows from
-functoriality of the interp.}
 
@@
 to the class of what might be called ``hereditarily polynomial''
-{\color{blue} to include truly nested such types} nested types via the
-categorical model of~\cite{jgj21}. But, as shown in
-Section~\ref{sec:par} below, for some GADTs --- even polynomial ones
---- no such isomorphism holds.
+nested types via the categorical model of~\cite{jgj21}. But, as shown
+in Section~\ref{sec:par} below, for some GADTs --- even polynomial
+ones --- no such isomorphism holds.
 
@@
 an instance of Haskell's $\mathsf{Functor}$ class. It is worth noting,
-however, that since the functorial completion of a GADT includes
-{\color{blue} only?} those data elements that are derivable via
-repeated applications of $\mathsf{map}$ starting from those specified
-by its syntax, primitive GADTs need not, in general, be nested
-types. For example, the functorial completion of the GADT $\mathsf{G}$
-defined by
+however, that since the functorial completion of a GADT includes those
+data elements that are derivable via repeated applications of
+$\mathsf{map}$ starting from those specified by its syntax, primitive
+GADTs need not, in general, be nested types. For example, let
+$\mathsf{1}$ be the unit type, whose single element is also denoted by
+$\mathsf{1}$, and let $\mathsf{G}$ be the GADT defined by
 \begin{equation}\label{eq:G}
@@
 \end{equation}
-Here, $\mathsf{1}$ is the unit type, whose single element is also
-denoted by $\mathsf{1}$, includes elements at any instance
-$\mathsf{G\,a}$ for any type $\mathsf{a}$ that is not the empty type
-$\mathsf{0}$ since there is always a function from $\mathsf{1}$ to
-such an $\mathsf{a}$, but does not include any terms at instance
+Then the functorial completion of $\mathsf{G}$ includes elements at
+any instance $\mathsf{G\,a}$ for any type $\mathsf{a}$ that is not the
+empty type $\mathsf{0}$ (there is always a function from $\mathsf{1}$
+to such an $\mathsf{a}$), but does not include any terms at instance
 $\mathsf{G\,0}$. Indeed, $\mathsf{G\,0}$ is not inhabited via the
@@
 class by defining a type-and-data-uniform, structure-preserving,
-data-changing $\mathsf{fmap_T}$ function for it.  Moreover, given a
-type-independent way of rearranging an ADT structure's shape
-$\mathsf{T\,a}$ into the shape for another ADT structure
-$\mathsf{T'\,a}$, we get the same structure of type $\mathsf{T'\,b}$
-regardless of whether we first rearrange the original structure of
-type $\mathsf{T\,a}$ into one of type $\mathsf{T'\,a}$ and then use
-$\mathsf{fmap_{T'}}$ to convert that resulting structure to one of
-type $\mathsf{T'\,b}$, or we first use $\mathsf{fmap_T}$ to convert
-the original structure of type $\mathsf{T\,a}$ to one of type
+data-changing function $\mathsf{map_T}$ for it.\footnote{We write
+  $\mathsf{map_f}$, or simply $\mathsf{map}$ when $\mathsf{f}$ is
+  clear from context, for the function $\mathsf{fmap :: (a \to b) \to
+    (f\,a \to f\,b)}$ witnessing that a type constructor $\mathsf{f}$
+  is an instance of Haskell's $\mathsf{Functor}$ class. This
+  emphasizes that $\mathsf{fmap}$ functions in Haskell are intended to
+  {\color{blue} satisfy syntactic reflections of the functoriality
+    properties --- i.e., preservation of identity functions and
+    composition of functions --- even though}
+%    there is no mechanism in  Haskell for enforcing this}.}
+%
+%  be functorial --- i.e., to preserve identity functions and function
+%  composition --- even though satisfaction of these properties
+this is not
+  enforced by the compiler and is instead left to the good intentions
+  of the programmer.}
+Moreover, given a type-independent way of rearranging an ADT
+structure's shape $\mathsf{T\,a}$ into the shape for another ADT
+structure $\mathsf{T'\,a}$, we get the same structure of type
+$\mathsf{T'\,b}$ regardless of whether we first rearrange the original
+structure of type $\mathsf{T\,a}$ into one of type $\mathsf{T'\,a}$
+and then use $\mathsf{map_{T'}}$ to convert that resulting structure
+to one of type $\mathsf{T'\,b}$, or we first use $\mathsf{map_T}$ to
+convert the original structure of type $\mathsf{T\,a}$ to one of type
 $\mathsf{T\,b}$ and then rearrange that resulting structure into one
@@
 following rearrange-transform property:
-\[\mathsf{fmap_{Tree}\,f\,(g \, xs) \,=\, g\,(fmap_{List}\,f\,xs)}\]
+\[\mathsf{map_{Tree}\,f\,(g \, xs) \,=\, g\,(map_{List}\,f\,xs)}\]
 
@@
 structure-preserving, data-changing $\mathsf{map}$ function can be
-defined. It therefore need not satisfy any rearrange-transform
-property either.
+defined (although one always can). It therefore need not satisfy any
+rearrange-transform property either (although it always does if such a
+$\mathsf{map}$ function is defined for it).
 
@@
 sets. Reflecting this requirement back into syntax gives that
-$\mathsf{F}$ must support a suitable\footnote{The $\mathsf{map}$
-  function for $\mathsf{F}$ should satisfy syntactic reflections of
-  the functoriality properties in $\mathit{Set}$ --- i.e.,
-  preservation of identity functions and composition of functions ---
-  even though there is no mechanism in Haskell for enforcing this.}
+$\mathsf{F}$ must support a suitable
+%\footnote{The $\mathsf{map}$
+%  function for $\mathsf{F}$ should satisfy syntactic reflections of
+%  the functoriality properties in $\mathit{Set}$ --- i.e.,
+%  preservation of identity functions and composition of functions ---
+%  even though there is no mechanism in Haskell for enforcing this.}
 $\mathsf{map}$ function, i.e., must be an instance of Haskell's
@@
 
-Requiring $F$ to be a functor will ensure that the interpretation $\mu
-F$ of the ADT $\mathsf{T\,a} = \mathsf{\mu F}$ exists. But to ensure
-that $\mu F$ is itself a functor, so that the type constructor
-$\mathsf{T}$ associated with the ADT $\mathsf{T\,a}$ supports a
-$\mathsf{map}$ function as well, we can require that $F$ also be the
-fixpoint $\mu H$ of a functor $H$. In this case, however, $H$ must be
-a functor on the category $\mathit{Set}^\mathit{Set}$ of functors and
-natural transformations on $\mathit{Set}$. That is, $H$ must be a {\em
+Requiring $F$ to be a functor ensures that the interpretation $\mu F$
+of the ADT $\mathsf{T\,a} = \mathsf{\mu F}$ exists. But to ensure that
+$\mu F$ is itself a functor, so that the type constructor $\mathsf{T}$
+associated with the ADT $\mathsf{T\,a}$ also supports a $\mathsf{map}$
+function, we can require that $F$ also be the fixpoint $\mu H$ of a
+functor $H$. In this case, however, $H$ must be a functor on the
+category $\mathit{Set}^\mathit{Set}$ of functors and natural
+transformations on $\mathit{Set}$. That is, $H$ must be a {\em
   higher-order} functor on $\mathit{Set}$. Reflecting this requirement
-back into syntax gives that
-$\mathsf{T\,a} = \mathsf{(\mu H)\,a}$
-for the ``type constructor constructor''
-$\mathsf{H}$ that supports suitable\footnote{The $\mathsf{map}$
-  function for $\mathsf{H}$ should satisfy syntactic reflections of
-  the functoriality properties in $\mathit{Set}^\mathit{Set}$ ---
-  i.e., preservation of identity natural transformations and
-  composition of natural transformations --- and the $\mathsf{map}$
-  function for $\mathsf{H\,F}$ should satisfy syntactic reflections of
-  the functoriality properties in $\mathit{Set}$ as in the previous
-  footnote. Again, there is no mechanism in Haskell for enforcing
-  this.} $\mathsf{map}$ functions.
-
-%class HFunctor f where
-%  ffmap :: Functor g => (a -> b) -> f g a -> f g b
-%  hfmap :: (g :~> h) -> f g :~> f h
-%
-%  type (~>) f g = forall x. f x -> g x
+back into syntax gives that $\mathsf{T\,a} = \mathsf{(\mu H)\,a}$ for
+the ``type constructor constructor'' $\mathsf{H}$ that supports
+suitable\footnote{The $\mathsf{map}$ function for $\mathsf{H}$ should
+  satisfy syntactic reflections of the functoriality properties in
+  $\mathit{Set}^\mathit{Set}$ --- i.e., preservation of identity
+  natural transformations and composition of natural transformations
+  --- and the $\mathsf{map}$ function for $\mathsf{H\,F}$ should
+  satisfy syntactic reflections of the functoriality properties in
+  $\mathit{Set}$ as in the previous footnote. Again, there is no
+  mechanism in Haskell for enforcing this.} $\mathsf{map}$ functions.
 
@@
 
-In a parametric model, short cut fusion can be used to show that the
-Church encoding and the primitive representation of an ADT are
-isomorphic as sets. Note, however, that whereas it is impossible even
-to {\em state} the rearrange-transform property for Church encodings
-of ADTs {\color{blue} unless we first define $\mathsf{map}$ functions
-  for them}, this property is simply the reflection back into syntax
-of the instance of naturality for the type-independent function that
+\vspace*{0.05in}
+
+In models that validate short cut fusion for ADTs and nested types,
+the Church encoding and the primitive representation of an ADT or
+nested type are isomorphic as sets. But whereas it is impossible even
+to {\em state} the rearrange-transform property for {\color{blue} raw
+  --- i.e., $\mathsf{map}$less ---} Church encodings of such data
+types, such a property is simply the reflection back into syntax of
+the instance of naturality for the type-independent function that
 rearranges structures of type $\mathsf{T\,a}$ into ones of type
@@
   a))}\] of the nested type $\mathsf{PTree\,a}$ of perfect trees
-introduces the data constructor(s $\mathsf{PLeaf :: a \to PTree\,a}$
-and) $\mathsf{PNode :: PTree \,(a\times a) \to PTree\,a}$. It enforces
+introduces the data constructors $\mathsf{PLeaf :: a \to PTree\,a}$
+and $\mathsf{PNode :: PTree \,(a\times a) \to PTree\,a}$. It enforces
 not only the invariant that all of the data in a structure of type
@@
 
+GADTs are used in precisely those situations in which different
+behaviors at different instances of a data type are desired. This is
+achieved by allowing the programmer to give the type signatures of the
+GADT's data constructors independently, and then using pattern
+matching to force the desired type refinement. Applications of GADTs
+include generic programming, modeling programming languages via
+higher-order abstract syntax, maintaining invariants in data
+structures, and expressing constraints in embedded domain-specific
+languages. GADTs have also been used, e.g., to implement tagless
+interpreters~\cite{pl04,pr06,pvww06}, to improve memory
+performance~\cite{min15}, and to design APIs~\cite{pen20}.
+
 \subsection{Church Encodings of GADTs}\label{sec:CEs}
 
-The syntax of GADTs is consistent with programmers' use of them in
-situations where different behaviors at different type instances is
-desired. This is achieved by allowing the programmer to give the type
-signatures of the different data constructors independently, and then
-using pattern matching to force the desired type refinement.  Typical
-applications of GADTs include generic programming, modeling
-programming languages via higher-order abstract syntax, maintaining
-invariants in data structures, and expressing constraints in embedded
-domain-specific languages. They have also been used, e.g., to
-implement tagless interpreters~\cite{pl04,pr06,pvww06}, to improve
-memory performance~\cite{min15}, and to design APIs~\cite{pen20}.
-
-Allowing non-variable type arguments in the return types of a GADT's
-data constructors establishes a strong connection between a GADT's
-shape and the data it contains. With ADTs, we first choose the shape
-of the container and then fill that container with data of whatever
-type we like. Critically, the choice of shape is independent of the
-data to be stored. With GADTs, however, the shape of the container may
-actually {\em depend} on (the type of) the data to be contained. For
-example, $\mathsf{Const}$ can create data of any shape
-$\mathsf{Seq\,a}$, but $\mathsf{Pair}$ can produce data of shape
+The syntax of GADTs allows non-variable type arguments in the return
+types of their data constructors. This establishes a strong connection
+between a GADT's shape and the data it contains. With ADTs, we first
+choose the shape of the container and then fill that container with
+data of whatever type we like; critically, the choice of shape is
+independent of the data to be stored. With GADTs, however, the shape
+of the container may actually {\em depend} on (the type of) the data
+to be contained. For example, $\mathsf{Const}$ can create data of any
+shape $\mathsf{Seq\,a}$, but $\mathsf{Pair}$ can produce data of shape
 $\mathsf{Seq\,a}$ only if $\mathsf{a}$ is a pair type. As a result,
-modifying the data in a GADT may actually change the shape of --- or
-may even destroy! --- the GADT.
-
-To determine the possible shapes of a GADT's container we have to
+modifying the data in a GADT's container may actually change the shape
+of --- or may even destroy! --- that container.
+
+To determine the possible shapes of a GADT's container we must
 pattern-match on the type of the data to be contained. For this, it is
@@
 higher-order fixpoint equation
-\[H\,f\,a = a \,+\,(Lan_{\lambda c d. c \times d}\, \lambda
-c d. f c \times f d)\,a\] {\color{red} Check this equation against
-  next displayed one below.}  where $Lan_K \,F$ is the left Kan
-extension of the functor $F$ along the functor $K$. In general, the
-left Kan extension $Lan_K \,F : {\cal E} \to {\cal D}$ of $F : {\cal
-  C} \to {\cal D}$ along $K : {\cal C} \to {\cal E}$ is the best
-functorial approximation to $F$ that factors through $K$. Here, ``best
-functorial approximation'' means that $Lan_K \,F$ is the smallest
-functor that both extends the image of $K$ to $\cal D$ and agrees with
-$F$ on $\cal C$, in the sense that, for any other such functor $H$,
-there is a morphism of functors (i.e., a natural transformation) from
-$Lan_K \,F$ to $H$. This is captured formally by the following
+\[H\,f\,b = b \,+\,(Lan_{\lambda c d. c \times d}\, \lambda
+c d. f c \times f d)\,b\] where $Lan_K \,F$ is the left Kan extension
+of the functor $F$ along the functor $K$. In general, the left Kan
+extension $Lan_K \,F : {\cal E} \to {\cal D}$ of $F : {\cal C} \to
+{\cal D}$ along $K : {\cal C} \to {\cal E}$ is the best functorial
+approximation to $F$ that factors through $K$. Here, ``best functorial
+approximation'' means that $Lan_K \,F$ is the smallest functor that
+both extends the image of $K$ to $\cal D$ and agrees with $F$ on $\cal
+C$, in the sense that, for any other such functor $H$, there is a
+morphism of functors (i.e., a natural transformation) from $Lan_K \,F$
+to $H$. This is captured formally by the following
 definition~\cite{mac71}:
@@
 interpreting $\mathsf{k}$. In this setting, the categories
-$\mathcal{C}$, $\mathcal{D}$, and $\mathcal{E}$ must all be of the form $\mathit{Set}^n$ for some $n$. Using $\mathsf{Lan}$ we can then
+$\mathcal{C}$, $\mathcal{D}$, and $\mathcal{E}$ must all be of the
+form $\mathit{Set}^n$ for some $n$. Using $\mathsf{Lan}$ we can then
 rewrite the type of a constructor $\mathsf{c :: f a \to g(k\,a)}$ as
@@
 
+\vspace*{0.05in}
+
 In this section we have seen that GADTs can be represented either as
@@
   Representations of GADTs}\label{sec:non-equiv}
-
-{\color{blue} Is G or G\,a the GADT?}
 
@@
 
-{\color{blue} Use $\mathsf{1}$ in place of $\mathsf{*}$ here?}
 To compute the interpretation of $\mathsf{G}$'s primitive
@@
 a GADT contain ``hidden'' elements that are not specified by the
-GADT's syntax and are only accessible via applications of
-$\mathsf{map}$ {\color{blue} and other elements, too?}? A semanticist,
-on the other hand, would likely find the primitive representation of
-GADTs entirely reasonable. Indeed, they would likely find the
-nonfunctorial nature of a GADT's Church encoding unnerving at
-best. After all, they would likely argue, the data in a GADT shouldn't
-disappear just because you map a function over it! The fact that this
-happens when GADTs are represented by their Church encodings actually
-highlights how GADTs {\em do not} generalize the essential,
-container-ish nature of ADTs at all. A semanticist might therefore
-conclude that GADTs are seriously misnamed. They might even go on to
-wonder: in what sense do GADTs really deserve to be considered {\em
-  data types} at all?
+GADT's syntax and are only accessible, e.g., via applications of
+$\mathsf{map}$? A semanticist, on the other hand, would likely find
+the primitive representation of GADTs entirely reasonable. Indeed,
+they would likely find the nonfunctorial nature of a GADT's Church
+encoding unnerving at best. After all, they would likely argue, the
+data in a GADT shouldn't disappear just because you map a function
+over it! The fact that this happens when GADTs are represented by
+their Church encodings actually highlights how GADTs {\em do not}
+generalize the essential, container-ish nature of ADTs at all. A
+semanticist might therefore conclude that GADTs are seriously
+misnamed. They might even go on to wonder: in what sense do GADTs
+really deserve to be considered {\em data types} at all?
 
@@
 beyond just naturality, such as proving equivalence of Church
-encodings and primitive representations of ADTs and {\color{blue}
-  hereditarily polynomial} nested types by validating short cut fusion
-for them.
+encodings and primitive representations of ADTs and hereditarily
+polynomial nested types by validating short cut fusion for them.
 
@@
 \item objects: A relation is a triple $(A,B,R)$, where $R$ is a
-  relation between sets $A$ and $B$.
+  subset of $A \times B$.
 \item morphisms: A morphism from $(A,B,R)$ to $(A',B',R')$ is a pair
@@
 $R : \mathit{Rel}\,(A,B)$ then we write $\pi_1 R$ and $\pi_2 R$ for
-the {\em domain} $A$ and {\em codomain} $B$ of $R$,
-respectively. Below, we assume $\pi_1$ and $\pi_2$ are surjective, so
-that the relations in $\mathit{Rel}$ are {\em onto}. {\color{blue} Say
-  why we care about onto?} We write $\mathit{Eq}_A = (A,A,\{(x,x)~|~ x
-\in A\})$ for the {\em equality relation} on the set $A$.
+the {\em domain} $A$ and {\em codomain} $B$ of $R$, respectively. We
+write $\mathit{Eq}_A = (A,A,\{(x,x)~|~ x \in A\})$ for the {\em
+  equality relation} on the set $A$.
 
@@
 just those elements that are ``parametric''~\cite{rey83,wad89} to
-ensure that the Identity Extension Lemma holds. If {\color{blue} the
-  set interpretation of $\mathsf{Equal}$ is the function
-  $\mathit{Equal}$ and its relational interpretation is $\mathit{Eq}$}
-as intended, then the parametricity property for a GADT is an
-inhabitation result saying that the set interpreting any instance of
-that GADT contains exactly the interpretations of the data elements
-that can be formed using its data constructors and whose type is that
-instance. In terms of syntax the parametricity property for a GADT's
-Church encoding gives that the GADT is inhabited at any instance for
-which data elements can be formed using its data constructors. In
-particular, the parametricity property for the GADT $\mathsf{G}$
-from~\eqref{eq:G} gives that $\mathsf{G\,a}$ contains a single data
-element if $\mathsf{a}$ is semantically equivalent to $\mathsf{1}$ and
-none otherwise:
+ensure that the Identity Extension Lemma holds. If the set
+interpretation of $\mathsf{Equal}$ is the function $\mathit{Equal}$
+and its relational interpretation is $\mathit{Eq}$ as intended, then
+the parametricity property for a GADT is an inhabitation result saying
+that the set interpreting any instance of that GADT contains exactly
+the interpretations of the data elements that can be formed using its
+data constructors and whose type is that instance. In terms of syntax
+the parametricity property for a GADT's Church encoding gives that the
+GADT is inhabited at any instance for which data elements can be
+formed using its data constructors. In particular, the parametricity
+property for the GADT $\mathsf{G}$ from~\eqref{eq:G} gives that
+$\mathsf{G\,a}$ contains a single data element if $\mathsf{a}$ is
+semantically equivalent to $\mathsf{1}$ and none otherwise:
 
@@
 and $t\,B$ must be related in $G_1\,R$. However, under the semantics
-given in~\cite{atk12}, {\color{blue} which includes the aforementioned
-  interpretations of $\mathsf{Equal}$}, the relational interpretation
-$G_1\,R$ of $\mathsf{G}$ is {\color{blue} $\mathit{Eq}_1$ when $R$ is
-  the relational interpretation $\mathit{Eq}_1$ of $\mathsf{1}$, and
-  the empty relation whenever $R$ differs from
-  $\mathit{Eq}_1$}. Reflecting back into syntax we deduce that there
-can be no term in the type that is the Church encoding of
-$\mathsf{G\,a}$ unless $\mathsf{a}$ is semantically equivalent to
-$\mathsf{1}$.
+given in~\cite{atk12}, which includes the aforementioned
+interpretations of $\mathsf{Equal}$, the relational interpretation
+$G_1\,R$ of $\mathsf{G}$ is $\mathit{Eq}_1$ when $R$ is the relational
+interpretation $\mathit{Eq}_1$ of $\mathsf{1}$, and the empty relation
+whenever $R$ differs from $\mathit{Eq}_1$. Reflecting back into syntax
+we deduce that there can be no term in the type that is the Church
+encoding of $\mathsf{G\,a}$ unless $\mathsf{a}$ is semantically
+equivalent to $\mathsf{1}$.
 \end{example}
@@
 
+We note that is actually possible to construct a simpler
+counterexample to the Identity Extension Lemma for primitive GADTs
+using the relation $R = (1,\emptyset, \emptyset)$. However, this
+relation is somewhat artificial, in the sense that its domain is
+larger than is strictly necessary to define an empty relation. We
+therefore give the above example using the {\color{blue} total}
+relation $(1,2,1 \times 2)$ instead.
+
 Taken together, Examples~\ref{ex:CE-par} and~\ref{ex:prim-par} show
-that the parametricity properties of GADTs differ greatly depending on
-whether they are represented as their Church encodings or represented
-as primitive data types. This contrasts sharply with the fact that
-both ADTs and nested types have the same parametricity properties
-regardless of whether they are represented by their Church encodings
-or their primitive representations. {\color{blue} Double-check that
-  this last part is true.}
+that Church encodings and primitive representations of GADTs behave
+very differently with respect to parametricity.  This contrasts
+sharply with the fact that both ADTs and nested types have the same
+parametricity properties regardless of whether they are represented by
+their Church encodings or their primitive
+representations.
 
@@
 represent GADTs as their Church encodings. So, with neither
-representation can we obtain {\em both} the naturality consequences of
-parametricity {\em and} the non-naturality consequences of
-parametricity. Intuitively speaking, this is because a GADT is not a
-data type in the usual sense of being a container, and completing it
-so that it becomes one destroys the relationally uniform behavior it
-must exhibit in a parametric model. Which representation of GADTs the
-user chooses can therefore have profound consequences for how they are
-able to program with, and reason about, them.
+representation can we obtain {\em both} naturality properties {\em
+  and} non-naturality consequences of parametricity. Intuitively
+speaking, this is because a GADT is not a data type in the usual sense
+of being a container, and completing it so that it becomes one
+destroys the relationally uniform behavior it must exhibit in a
+parametric model. Which representation of GADTs the user chooses can
+therefore have profound consequences for how they are able to program
+with, and reason about, them.
 
@@
 such calculi do exist, and these can be used, e.g., to determine which
-instances of a GADT are inhabited. {\color{blue} Is this all?}
+instances of a GADT are inhabited.
 
@@
 case of parametricity. However, we don't actually get non-trivial
-naturality consequences of parametricity for GADTs in Bob's
-system. Why? Because this would require GADTs to be interpreted as
-functors, which, in Bob's system, requires the programmer to supply a
-map function on the interpretation of every GADT defined.  (Bob
-requires map functions for fixpoint but then doesn't use them.)
-Moreover, each such map function would have to behave
-functorially. Bob's system does indeed deliver non-trivial naturality
-results for every GADT for which the user can define a map function
-that behaves functorially. The problem is that there may not be many
-such GADTs (in fact it will not be possible for proper GADTs, i.e.,
-GADTs that are not actually nested types).
+naturality properties for GADTs in Bob's system. Why? Because this
+would require GADTs to be interpreted as functors, which, in Bob's
+system, requires the programmer to supply a map function on the
+interpretation of every GADT defined.  (Bob requires map functions for
+fixpoint but then doesn't use them.)  Moreover, each such map function
+would have to behave functorially. Bob's system does indeed deliver
+non-trivial naturality results for every GADT for which the user can
+define a map function that behaves functorially. The problem is that
+there may not be many such GADTs (in fact it will not be possible for
+proper GADTs, i.e., GADTs that are not actually nested types).
 

### 2021-03-04 20-gadts/GADT-story.tex
all but Section 6 done

@@
 data-changing function $\mathsf{map_T}$ for it.\footnote{We write
-  $\mathsf{map_f}$, or simply $\mathsf{map}$ when $\mathsf{f}$ is
+  $\mathsf{map_T}$, or simply $\mathsf{map}$ when $\mathsf{T}$ is
   clear from context, for the function $\mathsf{fmap :: (a \to b) \to
-    (f\,a \to f\,b)}$ witnessing that a type constructor $\mathsf{f}$
+    (T\,a \to T\,b)}$ witnessing that a type constructor $\mathsf{T}$
   is an instance of Haskell's $\mathsf{Functor}$ class. This
   emphasizes that $\mathsf{fmap}$ functions in Haskell are intended to
-  {\color{blue} satisfy syntactic reflections of the functoriality
-    properties --- i.e., preservation of identity functions and
-    composition of functions --- even though}
-%    there is no mechanism in  Haskell for enforcing this}.}
-%
-%  be functorial --- i.e., to preserve identity functions and function
-%  composition --- even though satisfaction of these properties
-this is not
-  enforced by the compiler and is instead left to the good intentions
-  of the programmer.}
+  satisfy syntactic reflections of the functoriality properties ---
+  i.e., preservation of identity functions and composition of
+  functions --- even though this is not enforced by the compiler and
+  is instead left to the good intentions of the programmer.}
 Moreover, given a type-independent way of rearranging an ADT
@@
 nested type are isomorphic as sets. But whereas it is impossible even
-to {\em state} the rearrange-transform property for {\color{blue} raw
-  --- i.e., $\mathsf{map}$less ---} Church encodings of such data
-types, such a property is simply the reflection back into syntax of
+to {\em state} the rearrange-transform property for Church encodings
+of such data types unless $\mathsf{map}$ functions have been defined
+for them, such a property is simply the reflection back into syntax of
 the instance of naturality for the type-independent function that
@@
 
-Consider the relation $R = (1, 2, 1 \times 2)$, where $1 \times 2$ is
-the total relation on $1$ and $2$, and thus is onto. We expect
-$(\mathit{Lan}_{\lambda *\!.\,\mathit{Eq}_1} \lambda
-*\!.\,\mathit{Eq}_1)\, R$ to be a relation with domain $1$. Since left
-Kan extensions preserve projections~\cite{rie16}, we can compute the
-domain as
+Consider the relation $R = (1, 2, 1 \times 2)$, where $1 \times 2$
+relates the single element of $1$ to both elements of the two-element
+set $2$. We expect $(\mathit{Lan}_{\lambda *\!.\,\mathit{Eq}_1}
+\lambda *\!.\,\mathit{Eq}_1)\, R$ to be a relation with domain
+$1$. Since left Kan extensions preserve projections~\cite{rie16}, we
+can compute the domain as
 \[\pi_1\,\big( (\mathit{Lan}_{\lambda *\!.\,\mathit{Eq}_1} \lambda
@@
 larger than is strictly necessary to define an empty relation. We
-therefore give the above example using the {\color{blue} total}
-relation $(1,2,1 \times 2)$ instead.
+therefore give the above example using the relation $(1,2,1 \times 2)$
+instead.
 

### 2021-03-08 20-gadts/GADT-story.tex
final version of paper?

@@
 
-\title[Parametricity in the Presence of GADTs]{Parametricity in the
-  Presence of GADTs}
+\title[Functional Pearl: Parametricity in the Presence of
+  GADTs]{Functional Pearl: Parametricity in the Presence of GADTs}
 
@@
 
+\vspace*{-0.05in}
+
 We note that is actually possible to construct a simpler
@@
 
-{\color{blue} Combine these two paragraphs.}
-
 We have shown that GADTs can be considered as data types in two
@@
 functorial completion of their syntax, in which case they are
-represented as primitive data types.  But, regardless of which
-representation we choose, GADTs fail to be have the expected
+represented as primitive data types.  Significantly, regardless of
+which representation we choose, GADTs fail to have {\em all} expected
 parametricity properties. Indeed, we can read the results of
-Section~\ref{sec:par} as showing that i) some standard parametricity
-results (e.g., inhabitation results) for calculi representing GADTs as
-their Church encodings cannot be extended to calculi that represent
-GADTs as primitive data types may not be possible, and ii) some
-naturality results (e.g., rearrange-transform properties) for calculi
-representing GADTs as primitive cannot be extended to calculi that
-represent GADTs as their Church encodings. So, with neither
-representation can we obtain {\em both} naturality properties {\em
-  and} non-naturality consequences of parametricity. Intuitively
-speaking, this is because a GADT is not a data type in the usual sense
-of being a container, and completing it so that it becomes one
-destroys the relationally uniform behavior it must exhibit in a
-parametric model. Which representation of GADTs the user chooses can
-therefore have profound consequences for how they are able to program
-with, and reason about, them.
-
-All is not lost, however. Despite there being no parametric models for
-calculi incorporating GADTs as primitive data types, we still get all
-possible rearrange-transform properties for these types. This means
-that analogues of most of Wadler's free theorems for ADTs hold for
-GADTs, since almost all of the specific consequences of parametricity
-given in~\cite{wad89} (see especially Figure~1) are actually of this
-form for list types. In fact, even though such naturality properties
-are generally regarded as consequences of {\em parametricity}, they
-are actually consequences of data types' {\em functorial semantics}.
-On the other hand, despite there being no models interpreting type
-constructors as functors for calculi incorporating GADTs as Church
-encodings, and thus there being no models validating all
-rearrange-transform properties for these types, parametric models for
-such calculi do exist, and these can be used, e.g., to determine which
-instances of a GADT are inhabited.
-
-{\color{blue} Comparison with Stephanie and Dmitrios? Comparison with
-  Bob's Parametric Model? Other conclusions????}
+Section~\ref{sec:par} as showing that i)
+%some standard parametricity results (e.g.,
+inhabitation results for calculi representing GADTs as Church
+encodings cannot be extended to calculi representing them as primitive
+data types, and ii)
+%some
+naturality results
+%(e.g., rearrange-transform properties)
+for calculi representing GADTs as primitive data types cannot be
+extended to calculi representing them as Church encodings. So, with
+neither representation can we obtain {\em both} naturality
+consequences {\em and} non-naturality consequences of
+parametricity. Intuitively, this is because a GADT is not a data type
+in the usual ``container'' sense, and completing it so that it becomes
+one destroys its relationally uniform --- i.e., parametric ---
+behavior.
+%required in a parametric model for parametricity.
+Which representation of GADTs a user chooses can therefore have
+profound consequences for programming with, and reasoning about, them.
+
+Although there are no parametric models for calculi representing GADTs
+as primitive data types, GADT analogues of most of Wadler's free
+theorems for ADTs still hold.
+%all possible rearrange-transform properties
+%in such calculi since
+%This means that analogues of most of Wadler's free theorems for ADTs
+%hold for GADTs,
+Indeed, most of the parametricity results in Figure~1 of~\cite{wad89}
+are actually naturality properties.
+%of this form for list types. In fact, even though such
+Naturality properties are often regarded as consequences of {\em
+  parametricity}, but are, in fact, derivable directly from data
+types' {\em functorial semantics}. Even for data types whose Church
+encodings and primitive representations coincide, the analysis in this
+paper clearly distinguishes those free theorems that are consequences
+of naturality from those that are consequences of parametricity {\em
+  per se}.
+
+\begin{comment}
+On the other hand, despite there
+being no models interpreting type constructors as functors for calculi
+incorporating GADTs as Church encodings,
+%and thus there being no models validating all
+%rearrange-transform properties for these types,
+parametric models for such calculi do exist, and these can be used,
+e.g., to determine which instances of a GADT are inhabited.
+\end{comment}
+
+Atkey's parametric model for $F_\omega$ represents data types as
+Church encodings. It requires the user to supply a $\mathsf{map}$
+function for the type constructor whose fixpoint characterizes the
+data type. But, importantly, functoriality of an underlying type
+constructor does not imply functoriality of its fixpoint, so the data
+type itself still need not necessarily support a
+$\mathsf{map}$. Similarly,~\cite{vw10} presents a parametric model for
+an extension of $F_\omega$ that supports type equality, and can encode
+GADTs. Functoriality is not guaranteed here, either; in fact, the
+parametric properties of GADTs described in the preliminary
+work~\cite{vw06} are all inhabitation results, rather than naturality
+results. Finally,~\cite{ms09} encodes GADTs as Scott encodings rather
+than Church encodings but, again, cites only inhabitation results for
+them.
 

### 2021-03-09 20-gadts/GADT-story.tex
small updates

@@
 semantics, i.e., need not support a $\mathsf{map}$ function. By
-contrast, the primitive prepresentation of a GADT must have a
+contrast, the primitive representation of a GADT must have a
 functorial semantics even to be well-defined. The two representations
 of a GADT thus need not be the same in general. In this paper we show
-that the choice of representation for GADTs has profound consequences
-for their parametricity properties, and thus for how we program with,
-and reason about, these advanced data types.
+that the choice of representation for GADTs has important consequences
+for parametricity, and thus for how we program with, and reason about,
+these advanced data types.
 \end{abstract}
@@
   cut fusion}~\cite{glp93} the two representations are, in fact,
-semantically equivalent, i.e., both have the same interpretation in
-the model. Models in which the Church encodings and primitive
+semantically equivalent, i.e., have the same interpretation in the
+model. Models in which the Church encodings and primitive
 representations of all algebraic data types (ADTs) are semantically
@@
 Correctness of short cut fusion, and thus semantic equivalence of
-Church encodings and primitive representations, was recently extended
+Church encodings and primitive representations, can also be extended
 to the class of what might be called ``hereditarily polynomial''
-nested types via the categorical model of~\cite{jgj21}. But, as shown
-in Section~\ref{sec:par} below, for some GADTs --- even polynomial
+nested types in the categorical setting of~\cite{jgj21}. But, as argued
+in Section~\ref{sec:non-equiv} below, for some GADTs --- even polynomial
 ones --- no such isomorphism holds.
+%Correctness of short cut fusion, and thus semantic equivalence of
+%Church encodings and primitive representations, was recently extended
+%to the class of what might be called ``hereditarily polynomial''
+%nested types via the categorical model of~\cite{jgj21}. But, as shown
+%in Section~\ref{sec:par} below, for some GADTs --- even polynomial
+%ones --- no such isomorphism holds.
 
 GADTs are traditionally represented via Church encodings. In this
-view, GADTs are completely determined by their syntax; for example,
+view, GADTs are completely determined by their syntax. For example,
 the GADT $\mathsf{Seq}$ defined in Haskell by
@@
 a {\em functor} is necessary in order to be able to take the least
-fixed point~\cite{tfca}, so it is inherent in the view of data types
-as their primitive representations. Viewed as its primitive
-representation, $\mathsf{Seq}$ comprises not just those data elements
+fixed point~\cite{tfca}, so functoriality is inherent in the view of
+data types as their primitive representations. Viewed as its primitive
+representation, $\mathsf{Seq}$ contains not just those data elements
 representable in syntax, but also all data elements of the form
 $\mathsf{map\,f\,s}$ for all types $\mathsf{t_1}$ and $\mathsf{t_2}$,
-all functions $\mathsf{f : t_1 \to t_2}$ definable in the language,
-and all $\mathsf{s :: Seq\,t_1}$, as well as all data elements of the
-form $\mathsf{map\,g\,s'}$ for each appropriately typed function
+all functions $\mathsf{f : t_1 \to t_2}$ definable in the language and
+all $\mathsf{s :: Seq\,t_1}$, as well as all data elements of the form
+$\mathsf{map\,g\,s'}$ for each appropriately typed function
 $\mathsf{g}$ and each element $\mathsf{s'}$ already added to the data
@@
 themselves be of the form $\mathsf{Pair\,v_1\,v_2}$ for any terms
-$\mathsf{v_1}$ and $\mathsf{v_2}$. By contrast, functorial completion
+$\mathsf{v_1}$ and $\mathsf{v_2}$. Importantly, functorial completion
 adds no new data elements to the syntax of a GADT that is actually a
 (polynomial) ADT or a (hereditarily polynomial) nested type, so its
-Church encoding and primitive representation coincide. However, these
-representations are rarely the same for GADTs.
+Church encoding and primitive representation coincide. However, the
+two representations are rarely the same for more general GADTs.
 
@@
 GADT to be completely specified by its syntax. When used this way, the
-shape of a particular element of a GADTs is actually {\em determined}
+shape of a particular element of a GADT is actually {\em determined}
 by the data it contains. As a result, it is not expected to support a
-$\mathsf{map}$ function --- i.e., to be an instance of Haskell's
-$\mathsf{Functor}$ class --- like other data types. Indeed, the
-definition for $\mathsf{Seq}$ above specifies that an element of the
-form $\mathsf{Pair\,u_1\,u_2}$ must have the shape of a sequence of
-data of pair type rather than a sequence of data of arbitrary type
+$\mathsf{map}$ function --- equivalently, to be an instance of
+Haskell's $\mathsf{Functor}$ class --- like other data types. Indeed,
+the definition for $\mathsf{Seq}$ above specifies that an element of
+the form $\mathsf{Pair\,u_1\,u_2}$ must have the shape of a sequence
+of data of pair type rather than a sequence of data of arbitrary type
 $\mathsf{e}$.  The clause of $\mathsf{map}$ for the $\mathsf{Pair}$
@@
 doesn't even really behave like a traditional data type. In this
-situation, the GADT cannot be seen as the least fixpoint of a functor,
-and must instead be represented via its Church encoding.
+situation, the GADT cannot be seen as a functor, and must instead be
+represented via its Church encoding.
 
@@
 generalizations of ADTs. The semanticist is thus likely to expect a
-GADT to be able to be used as a ADT is used --- i.e., as a shape
+GADT to be able to be used as an ADT is used --- i.e., as a shape
 filled with data that can be changed without also changing the
@@
 ensure that the GADT supports a $\mathsf{map}$ function, i.e., is
-interpreted as a true categorical functor, and can therefore be made
-an instance of Haskell's $\mathsf{Functor}$ class. It is worth noting,
-however, that since the functorial completion of a GADT includes those
-data elements that are derivable via repeated applications of
-$\mathsf{map}$ starting from those specified by its syntax, primitive
-GADTs need not, in general, be nested types. For example, let
-$\mathsf{1}$ be the unit type, whose single element is also denoted by
-$\mathsf{1}$, and let $\mathsf{G}$ be the GADT defined by
+interpreted as a true categorical functor, and thus can reasonably be
+made an instance of Haskell's $\mathsf{Functor}$ class. It is worth
+noting, however, that since the functorial completion of a GADT
+includes those data elements that are derivable via repeated
+applications of $\mathsf{map}$ starting from those specified by its
+syntax, primitive GADTs need not, in general, be nested types. For
+example, let $\mathsf{1}$ be the unit type, whose single element is
+also denoted by $\mathsf{1}$, and let $\mathsf{G}$ be the GADT defined
+by
 \begin{equation}\label{eq:G}
@@
 
-The key observation of this paper is that, while the viewpoints of
-both the GADT programmer and the semanticist are both valid, the two
-are irreconcilable. Importantly, which point of view is adopted in any
-particular setting has deep consequences for the ways GADTs can be
-used and reasoned about there. In this paper we show that the choice
-of GADT representation has deep implications for parametric reasoning
-about structures of those types. In particular, a GADT programmer who
-views GADTs as their Church encodings cannot safely use program
-transformations or reasoning principles for them that involve
-$\mathsf{map}$ functions for them, although they may be able to
-program with and reason about GADTs using other consequences of
-parametricity, such as type inhabitation results. On the other hand, a
-semanticist who views GADTs as primitive data types will have all
-naturality-based program transformations and reasoning principles for
-GADTs at their disposal, but, since no parametric model can be
-constructed for primitive GADTs, non-naturality consequences of
-parametric reasoning may not be available to them.
-
-Below we show how the above observations can be made precise, thereby
-answering the question we posed above in the affirmative: Yes, it
-really does matter which way we view GADTs.
+The key observation of this paper is that, while the viewpoints of the
+GADT programmer and the semanticist are both valid, the two are
+irreconcilable. Importantly, which point of view is adopted in any
+particular setting has significant consequences for the ways GADTs can
+be used and reasoned about there. In particular, the choice of GADT
+representation has deep implications for parametric reasoning about
+structures of those types. Indeed, a GADT programmer who views GADTs
+as their Church encodings cannot safely use program transformations or
+reasoning principles that involve $\mathsf{map}$ functions for them,
+although they may be able to program with and reason about GADTs using
+other consequences of parametricity, such as type inhabitation
+results. On the other hand, a semanticist who views GADTs as primitive
+data types will have all naturality-based program transformations and
+reasoning principles for GADTs at their disposal, but, since no
+parametric model can be constructed for primitive GADTs,
+non-naturality consequences of parametric reasoning may not be
+available to them.
+
+In this paper we show how these observations can be made precise,
+thereby answering the question we posed above in the affirmative: Yes,
+it really does matter which way we view GADTs.
 
@@
 \mathsf{c_n t_{n1}}...\mathsf{t_{nk_n}}\] where each $\mathsf{t_{ij}}$
-is a type also depending only on $\mathsf{a}$. Such a data type can be
+is a type depending only on $\mathsf{a}$. Such a data type can be
 thought of as a ``container'' for data of type $\mathsf{a}$. The data
@@
 Primitive representations capture in syntax the fact that ADTs can be
-considered as fixpoints. For example $\mathsf{List\,a}$ is indeed a
-fixpoint of $\mathsf{F_{List\,a}}$, where $\mathsf{F_{List\,a}\,X} =
-\mathsf{1 + a \times X}$ as determined by~\eqref{eq:list}, i.e., is
-indeed such that
+considered as fixpoints of functors. For example $\mathsf{List\,a}$ is
+indeed a fixpoint of $\mathsf{F_{List\,a}}$, where
+$\mathsf{F_{List\,a}\,X} = \mathsf{1 + a \times X}$ as determined
+by~\eqref{eq:list}, i.e., is indeed such that
 \[\mathsf{List\,a} = \mathsf{1 + a \times List\,a}\] 
 since every element of $\mathsf{List\,a}$ is either empty or is
-obtained by $\mathsf{Cons}$ing an element of type $\mathsf{a}$ onto an
+obtained by consing an element of type $\mathsf{a}$ onto an
 already-existing structure of type $\mathsf{List\,a}$. In fact, this
@@
 to {\em state} the rearrange-transform property for Church encodings
-of such data types unless $\mathsf{map}$ functions have been defined
-for them, such a property is simply the reflection back into syntax of
-the instance of naturality for the type-independent function that
-rearranges structures of type $\mathsf{T\,a}$ into ones of type
-$\mathsf{T'\,a}$ and the content-transforming functions
-$\mathsf{map_T\,f}$ and $\mathsf{map_{T'}\,f}$ for a function
-$\mathsf{f :: a \to b}$, where $\mathsf{T}$ and $\mathsf{T'}$ are the
-type constructors associated with these ADTs, respectively.
+of such data types (unless $\mathsf{map}$ functions have for some
+reason been defined for them), such a property is simply the
+reflection back into syntax of the instance of naturality for the
+type-independent function that rearranges structures of type
+$\mathsf{T\,a}$ into ones of type $\mathsf{T'\,a}$ and the
+content-transforming functions $\mathsf{map_T\,f}$ and
+$\mathsf{map_{T'}\,f}$ for a function $\mathsf{f :: a \to b}$, where
+$\mathsf{T}$ and $\mathsf{T'}$ are the type constructors associated
+with these ADTs, respectively.
 
@@
 Importantly, the function $\mathit{Equal}\,a$ cannot be made into a
-functor. Equivalently, $\mathsf{Equal\,a}$ (in either representation)
-cannot be made an instance of Haskell's $\mathsf{Functor}$
-class. Indeed, if $\mathsf{Equal\,a}$ supported a function
+functor. Equivalently, $\mathsf{Equal\,a}$ cannot be made an instance
+of Haskell's $\mathsf{Functor}$ class. Indeed, if $\mathsf{Equal\,a}$
+supported a function
 \[\mathsf{map_{Equal\,a} :: (b \to c) \to Equal\,a\,b \to
@@
 GADT serves as an informative case study highlighting the differences
-between a GADT's Church encoding its primitive representation --- even
-if we consider only the data elements it contains, and ignore whether
-or not it supports a $\mathsf{map}$ function.
+between a GADT's Church encoding and its primitive representation ---
+even if we consider only the data elements it contains, and ignore
+whether or not it supports a $\mathsf{map}$ function.
 
 \begin{example}\label{ex:g1}
+
 Syntactically, the GADT $\mathsf{G}$ defined in~\eqref{eq:G} comprises
 a single data element, namely $\mathsf{c :: G \, 1}$. This is captured
-by $\mathsf{G}$'s Church encoding
-\[\mathsf{G\,a} = \mathsf{Equal\,1\,a}\]
-from~\cite{atk12}. As its Church encoding makes clear, $\mathsf{G}$'s
+by $\mathsf{G}$'s Church encoding $\mathsf{G\,a} = \mathsf{\forall\,f.
+  \, (\forall\,c.\, Equal\,1\,c \to f\,c) \to f\,a}$ from
+~\cite{atk12}, which is equivalent in the model given there to
+\[\mathsf{G\, a = Equal\,1\,a}\]
+since the definition of $\mathsf{G}$ is not recursive. As its Church
+encoding makes clear, 
+%Syntactically, the GADT $\mathsf{G}$ defined in~\eqref{eq:G} comprises
+%a single data element, namely $\mathsf{c :: G \, 1}$. This is captured
+%by $\mathsf{G}$'s Church encoding
+%\[{\color{blue}\mathsf{G\,a} = \mathsf{Equal\,1\,a}}\]
+%from~\cite{atk12}. As its Church encoding makes clear,
+$\mathsf{G}$'s
 effect is simply to test its argument for equality against the unit
@@
 $\mathsf{\lambda *. 1}$ is the syntactic reflection of the constantly
-$1$-valued functor from the category with a single object $*$ to
-$\mathit{Set}$. We can therefore represent $\mathsf{G}$ as
+$1$-valued functor from the category $\mathit{Set}^0$ with a single
+object $*$ to $\mathit{Set}$. We can therefore represent $\mathsf{G}$
+as
 \[\mathsf{G\,a} = \mathsf{(\mu \phi. \lambda b.
@@
 $\mathsf{build}$ function for the primitive representation of such a
-GADT.  This is not surprising, however: if short cut fusion were to
-hold for a GADT then its Church encoding and primitive representation
-would necessarily be semantically equivalent.
+GADT. Ultimately this is not surprising, though: if short cut fusion
+were to hold for a GADT then its Church encoding and primitive
+representation would necessarily be semantically equivalent.
 
@@
 functions. However, parametricity can also derive results that go
-beyond just naturality, such as proving equivalence of Church
-encodings and primitive representations of ADTs and hereditarily
-polynomial nested types by validating short cut fusion for them.
-
-To discuss relational parametricity will need to interpret data types
-not just in $\mathit{Set}$, but in a suitable category of relations as
-well. The following definition is standard:
+beyond just naturality, such as inhabitation results, and prove the
+equivalence of Church encodings and primitive representations of ADTs
+and hereditarily polynomial nested types by validating short cut
+fusion for them.
+
+To discuss relational parametricity we will need to interpret data
+types not just in $\mathit{Set}$, but in a suitable category of
+relations as well. The following definition is standard:
 \begin{definition}
@@
 in~\cite{atk12}.  In that model, types are interpreted ``in parallel''
-in $\mathit{Set}$ and $\mathit{Rel}$ in the usual way, including the
-familiar ``cutting down'' of the interpretations of $\forall$-types to
-just those elements that are ``parametric''~\cite{rey83,wad89} to
-ensure that the Identity Extension Lemma holds. If the set
-interpretation of $\mathsf{Equal}$ is the function $\mathit{Equal}$
-and its relational interpretation is $\mathit{Eq}$ as intended, then
-the parametricity property for a GADT is an inhabitation result saying
-that the set interpreting any instance of that GADT contains exactly
-the interpretations of the data elements that can be formed using its
-data constructors and whose type is that instance. In terms of syntax
-the parametricity property for a GADT's Church encoding gives that the
-GADT is inhabited at any instance for which data elements can be
-formed using its data constructors. In particular, the parametricity
-property for the GADT $\mathsf{G}$ from~\eqref{eq:G} gives that
-$\mathsf{G\,a}$ contains a single data element if $\mathsf{a}$ is
-semantically equivalent to $\mathsf{1}$ and none otherwise:
+in (types corresponding to) $\mathit{Set}$ and $\mathit{Rel}$ in the
+usual way, including the familiar ``cutting down'' of the
+interpretations of $\forall$-types to just those elements that are
+``parametric''~\cite{rey83,wad89} to ensure that the Identity
+Extension Lemma holds. If the set interpretation of $\mathsf{Equal}$
+is the function $\mathit{Equal}$ and if the relational interpretation
+of a closed type $\mathsf{a}$ with interpretation $A$ is
+$\mathit{Eq_A}$ as intended, then the parametricity property for a
+GADT is an inhabitation result saying that the set interpreting any
+instance of that GADT contains exactly the interpretations of the data
+elements that can be formed using its data constructors and whose type
+is that instance. In terms of syntax the parametricity property for a
+GADT's Church encoding gives that the GADT is inhabited at any
+instance for which data elements can be formed using its data
+constructors. In particular, the parametricity property for the GADT
+$\mathsf{G}$ from~\eqref{eq:G} gives that $\mathsf{G\,a}$ contains a
+single data element if $\mathsf{a}$ is semantically equivalent to
+$\mathsf{1}$ and none otherwise:
 
@@
 given in~\cite{atk12}, which includes the aforementioned
-interpretations of $\mathsf{Equal}$, the relational interpretation
-$G_1\,R$ of $\mathsf{G}$ is $\mathit{Eq}_1$ when $R$ is the relational
-interpretation $\mathit{Eq}_1$ of $\mathsf{1}$, and the empty relation
-whenever $R$ differs from $\mathit{Eq}_1$. Reflecting back into syntax
-we deduce that there can be no term in the type that is the Church
-encoding of $\mathsf{G\,a}$ unless $\mathsf{a}$ is semantically
-equivalent to $\mathsf{1}$.
+interpretations of $\mathsf{Equal}$ and closed types, the relational
+interpretation $G_1\,R$ of $\mathsf{G}$ is itself $\mathit{Eq}_1$ when
+$R$ is the relational interpretation $\mathit{Eq}_1$ of $\mathsf{1}$,
+and the empty relation whenever $R$ differs from
+$\mathit{Eq}_1$. Reflecting back into syntax we deduce that there can
+be no term in the type that is the Church encoding of $\mathsf{G\,a}$
+unless $\mathsf{a}$ is semantically equivalent to $\mathsf{1}$.
 \end{example}
@@
 *\!.\,\mathit{Eq}_1$, where $\lambda *\!.\,\mathit{Eq}_1$ is the
-constantly $\mathit{Eq}_1$-valued functor from the category with a
-single object $*$ to $\mathit{Rel}$.  Indeed, the interpretation is
-still a left Kan extension, but now it is the left Kan extension
-determined by the functor interpreting the type $\mathsf{\lambda
-  *\!. 1}$ in $\mathit{Rel}$. For the Identity Extension Lemma to hold
-we would need that, for every relation $R : \mathit{Rel}\,(A, B)$, we
-have that $(\mathit{\Lan}_{\lambda *\!.\,\mathit{Eq}_1} \,\lambda
+constantly $\mathit{Eq}_1$-valued functor from the category
+$\mathit{Rel}\,^0$ with a single object (also called $*$) to
+$\mathit{Rel}$.  Indeed, the interpretation is still a left Kan
+extension, but now it is the left Kan extension determined by the
+functor interpreting the type $\mathsf{\lambda *\!. 1}$ in
+$\mathit{Rel}$. For the Identity Extension Lemma to hold we would need
+that, for every relation $R : \mathit{Rel}\,(A, B)$, we have that
+$(\mathit{\Lan}_{\lambda *\!.\,\mathit{Eq}_1} \,\lambda
 *\!.\,\mathit{Eq}_1)\, R$ is a relation between the sets
@@
 
-\section{Conclusion}
+\section{Conclusion and Related Work}
 
@@
 functorial completion of their syntax, in which case they are
-represented as primitive data types.  Significantly, regardless of
-which representation we choose, GADTs fail to have {\em all} expected
+represented as primitive data types.  But regardless of which
+representation we choose, GADTs fail to have {\em all} expected
 parametricity properties. Indeed, we can read the results of
@@
 parametricity. Intuitively, this is because a GADT is not a data type
-in the usual ``container'' sense, and completing it so that it becomes
+in the usual container-ish sense, and completing it so that it becomes
 one destroys its relationally uniform --- i.e., parametric ---
@@
 encodings and primitive representations coincide, the analysis in this
-paper clearly distinguishes those free theorems that are consequences
-of naturality from those that are consequences of parametricity {\em
-  per se}.
+paper clearly distinguishes those free theorems that are actually
+consequences of naturality from those that are true consequences of
+parametricity.
 
@@
 $\mathsf{map}$. Similarly,~\cite{vw10} presents a parametric model for
-an extension of $F_\omega$ that supports type equality, and can encode
-GADTs. Functoriality is not guaranteed here, either; in fact, the
-parametric properties of GADTs described in the preliminary
+an extension of $F_\omega$ that supports type equality, and thus can
+encode GADTs, but still does not guarantee functoriality. In fact,
+%encode GADTs. But functoriality is not guaranteed here, either; in fact,
+the parametric properties of GADTs described in the preliminary
 work~\cite{vw06} are all inhabitation results, rather than naturality

### 2021-03-18 20-gadts/GADT-story.tex

@@
 of type $\mathsf{T'\,b}$. For example, if $\mathsf{f : a \to b}$,
-$\mathsf{xs \,:\,List\,a}$, and $\mathsf{g : List\,a\to Tree\,a}$
+$\mathsf{t \,:\,Tree\,a}$, and $\mathsf{g : Tree\,a\to List\,a}$
 arranges lists into trees in a type-independent way, then we have the
 following rearrange-transform property:
-\[\mathsf{map_{Tree}\,f\,(g \, xs) \,=\, g\,(map_{List}\,f\,xs)}\]
+\[\mathsf{map_{List}\,f\,(g \, t) \,=\, g\,(map_{Tree}\,f\,t)}\]
 
@@
   \, (\forall\,c.\, Equal\,1\,c \to f\,c) \to f\,a}$ from
-~\cite{atk12}, which is equivalent in the model given there to
+~\cite{atk12}, which is equivalent in the parametric model given there
+to
 \[\mathsf{G\, a = Equal\,1\,a}\]
-since the definition of $\mathsf{G}$ is not recursive. As its Church
-encoding makes clear, 
+As this makes clear,
 %Syntactically, the GADT $\mathsf{G}$ defined in~\eqref{eq:G} comprises

### 2021-03-26 induction-paper/GADT-induction.tex
Induction paper, first commit

% For double-blind review submission, w/o CCS and ACM Reference (max
% submission space)
\documentclass[acmsmall,screen,review,anonymous]{acmart}
\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

%% Journal information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{Haskell Symposium} % CONF = POPL or ICFP or OOPSLA
\acmArticle{1}
\acmYear{2021}
\acmMonth{8}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2018}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations
%\citestyle{acmnumeric}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Note: Authors migrating a paper from PACMPL format to traditional
%% SIGPLAN proceedings format must update the '\documentclass' and
%% topmatter commands above; see 'acmart-sigplanproc-template.tex'.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage[utf8]{inputenc}
\usepackage{ccicons}
\usepackage{verbatim}

\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amscd}
%\usepackage{MnSymbol}
\usepackage{xcolor}

\usepackage{bbold}
\usepackage{url}
\usepackage{upgreek}
%\usepackage{stmaryrd}

\usepackage{lipsum}
\usepackage{tikz-cd}
\usetikzlibrary{cd}
\usetikzlibrary{calc}
\usetikzlibrary{arrows}

\usepackage{bussproofs}
\EnableBpAbbreviations

\input{macros}
\input{mytheorem}

\theoremstyle{definition}
\newtheorem{exmpl}{Example}

\renewcommand{\greyout}[1]{}

\newcommand{\new}[1]{{\color{blue} {#1}}}

\newcommand{\emptyfun}{{[]}}
\newcommand{\cal}{\mathcal}
\newcommand{\F}{\mathcal{F}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\B}{\mathcal{B}}
\renewcommand{\P}{\mathcal{A}}
\newcommand{\pred}{\mathsf{Fam}}
\newcommand{\env}{\mathsf{Env}}
\newcommand{\set}{\mathsf{Set}}
\renewcommand{\S}{\mathcal S}
\newcommand{\C}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\A}{\mathcal{A}}
\renewcommand{\id}{\mathit{id}}
\newcommand{\map}{\mathsf{map}}
\newcommand{\pid}{\underline{\mathit{id}}}
\newcommand{\pcirc}{\,\underline{\circ}\,}
\newcommand{\pzero}{\underline{0}}
\newcommand{\pone}{\underline{1}}
\newcommand{\psum}{\,\underline{+}\,}
\newcommand{\pinl}{\underline{\mathit{inL}}\,}
\newcommand{\pinr}{\underline{\mathit{inR}}\,}
\newcommand{\ptimes}{\,\underline{\times}\,}
\newcommand{\ppi}{\underline{\pi_1}}
\newcommand{\pppi}{\underline{\pi_2}}
\newcommand{\pmu}{\underline{\mu}}
\newcommand{\semmap}{\mathit{map}}
\newcommand{\subst}{\mathit{subst}}

\newcommand{\tb}[1]{~~ \mbox{#1} ~~}
\newcommand{\listt}[1]{(\mu \phi. \lambda \beta . \onet + \beta \times
  \phi \beta) #1} 
\newcommand{\filtype}{\Nat^\emptyset 
 (\Nat^\emptyset \, \alpha \, \mathit{Bool})\, (\Nat^\emptyset 
  (List \, \alpha) \, (List \, \alpha))} 
\newcommand{\filtypeGRose}{\Nat^\emptyset 
 (\Nat^\emptyset \, \alpha \, \mathit{Bool})\, (\Nat^\emptyset 
  (\mathit{GRose}\,\psi \, \alpha) \, (\mathit{GRose}\,\psi \, (\alpha
  + \onet)))} 
\newcommand{\maplist}{\mathit{map}_{\lambda A. \setsem{\emptyset; \alpha
      \vdash \mathit{List} \, \alpha} \rho[\alpha := A]}} 
\newcommand{\PLeaves}{\mathsf{PLeaves}}
\newcommand{\swap}{\mathsf{swap}}
\newcommand{\reverse}{\mathsf{reverse}}
\newcommand{\Bcons}{\mathit{Bcons}}
\newcommand{\Bnil}{\mathit{Bnil}}

\title[(Deep) Induction for GADTs]{(Deep) Induction for GADTs}

\author{Patricia Johann, Enrico Ghiorzi, and Daniel Jeffries}
\affiliation{ \institution{Appalachian State University}}
\email{johannp@appstate.edu, ghiorzie@appstate.edu,
  jeffriesd@appstate.edu}

\begin{document}

\begin{abstract}
Abstract
\end{abstract}

\maketitle

\section{Introduction}\label{sec:intro}

\section{Deep induction for ADTs and nested types}

%\begin{itemize}
%\item Say what ADTs and nested types are (look~\cite{jp20,jgj21})
%\item Examples: $\mathsf{List}$ and $\mathsf{PTree}$ and $\mathsf{Bush}$
%\item Explain induction for $\mathsf{List}$ and $\mathsf{PTree}$ (refer paper Patricia Neil Clement 2010) notice it does not extend to truly nested types like $\mathsf{Bush}$
%\item To get induction rule for truly nested types such as $\mathsf{Bush}$ requires deep induction (refer to~\cite{jp20}) (explain various pieces of deep induction, including lifting, informally)
%\item Deep induction makes sense for other structures as well (not just truly nested types)
%\end{itemize}

\subsection{Syntax of ADTs and nested types}\label{ssec:syntax-adts}

\emph{(Polynomial) algebraic data types} (ADTs), both built-in and user-defined,
have long been at the core of functional languages such as Haskell, ML, Agda, Epigram, and Idris.
ADTs are used extensively in functional programming to structure computations,
to express invariants of the data over which computations are defined,
and to ensure the type safety of programs specifying those computations.
ADTs include unindexed types, such as the type of natural numbers,
and types indexed over other types, such as the quintessential example of an ADT,
the type of lists (here coded in Agda)
{\color{blue}(Ask Daniel which flavor of syntax, paper as literate Agda, naming conventions?)}
\begin{equation}\label{eq:list}
\begin{array}{l}
%\mathsf{data\, List : Set \to Set\,where}\\
\mathsf{data\, List\,(a : Set) : Set\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Nil\,\,\,\,\,\, :\, List\,a}\\
\mathsf{\;\;\;\;\;\;\;\;Cons\, :\, a \to List\,a \to List\,a} 
\end{array}
\end{equation}
Notice that all occurrences of $\mathsf{List}$ in the above encoding are instantiated at the same index $\mathsf{a}$.
Thus, the instances of $\mathsf{List}$ at various indices are defined independently from one another.
That is a defining feature of ADTs:
an ADT defines a \emph{family of inductive types}, one for each index type.

Over time, there has been a notable trend toward data types
whose non-regular indexing can capture invariants and other sophisticated properties
that can be used for program verification and other applications.
A simple example of such a type is given by Bird and Meertens’~\cite{bm98} prototypical \emph{nested type}
\begin{equation}\label{eq:ptree}
\begin{array}{l}
\mathsf{data\, PTree\,(a : Set) : Set\,where}\\
\mathsf{\;\;\;\;\;\;\;\;PLeaf\,\,\, :\, a \to PTree\,a}\\
\mathsf{\;\;\;\;\;\;\;\;PNode\, :\, PTree\,(a \times a) \to PTree\,a} 
\end{array}
\end{equation}
of perfect trees, which can be thought of as constraining lists to have lengths that are powers of 2.
In the above code, the constructor $\mathsf{PNode}$ uses data of type $\mathsf{PNode\,(a \times a)}$ to construct data of type $\mathsf{PNode\,a}$.
Thus, it is clear that the instantiations of $\mathsf{PNode}$ at various indices cannot be defined independently, so that the entire family of types must actually be defined at once.
A nested type thus defines not a family of inductive types, but rather an \emph{inductive family of types}.

Nested types include simple nested types, like perfect trees,
none of whose recursive occurrences occur below another type constructor,
and \emph{truly} nested types, such as the nested type
\begin{equation}\label{eq:ptree}
\begin{array}{l}
\mathsf{data\, Bush\,(a : Set) : Set\,where}\\
\mathsf{\;\;\;\;\;\;\;\;BNil\,\,\,\,\,\, :\, Bush\,a}\\
\mathsf{\;\;\;\;\;\;\;\;BCons\, :\, a \to Bush\,(Bush\,a) \to Bush\,a} 
\end{array}
\end{equation}
of bushes, whose recursive occurrences appear below their own type constructors.
Note that, while the constructors of a nested type can contain occurrences of the type instantiated at any index,
the return types of its constructors still have to be the same type instance of the type being defined.
In other words, all constructors of $\mathsf{PTree\,a}$ have to return an element of type $\mathsf{PTree\,a}$,
and all constructors of $\mathsf{Bush\,a}$ have to return an element of type $\mathsf{Bush\,a}$.
%the return types of the constructors are still required to be instantiated at the same variable $\mathsf{a}$.}
%As previously noted, the return types of the constructors of a nested
%type have to be the same type instance of the type being defined.

\subsection{Induction principles for ADTs and nested types}\label{ssec:induction-nested}

An induction principle for a data type allows proving
that a predicate holds for every element of that data type,
provided that it holds for every element inductively produced by the type's constructors.
In this paper, we are interested in induction principles for proof-relevant predicates.
A proof-relevant predicate on a type $\mathsf{a}$ is a function $\mathsf{a \to Set}$
(where $\mathsf{Set}$ is the type of sets)
mapping each $\mathsf{x : a}$ to the set of proofs that the predicate holds for $\mathsf{x}$.
For example, the induction principle for $\mathsf{List}$ is
\begin{multline}
\mathsf{\forall (a : Set) (P : List\,a \to Set) \to P\,Nil \to \big( \forall (x : a) (xs: List\,a) \to P\,xs \to P\,(Cons\,x\,xs)\big)} \\
\mathsf{\to \forall (xs : List\,a) \to P\, xs }
\end{multline}
Note that the data inside a structure of type $\mathsf{List}$ is treated monolithically (i.e., ignored) by this induction rule.
Indeed, the induction rule inducts over only the top-level structures of data types,
leaving any data internal to the top-level structure untouched.
Since this kind of induction principle is only concerned with the structure of the type,
and unconcerned with the contained data,
we will then refer to it as \emph{structural induction}.

We can extend such a structural induction principle to some nested types, such as $\mathsf{PTree}$.
The only difference from the induction principle for ADTs is that,
since a nested type is defined as a whole inductive family of types at once,
its induction rule has to necessary involve a polymorphic predicate.
Thus, the induction rule for $\mathsf{PTree}$ is
\begin{multline}
\mathsf{\forall (P : \forall (a : Set) \to PTree\, a \to Set)
\to \big( \forall (a : Set) (x : a) \to P\,a\,(PLeaf\, x) \big)} \\
\mathsf{\to \big( \forall (a : Set) (x : PTree\,(a \times a)) \to P\,(a \times a)\,x \to P\,a\,(PNode\,x)\big)} \\
\mathsf{\to \forall (a : Set) (x : PTree\,a) \to P\,a\,x }
\end{multline}

%This kind of induction principles, which we will call \emph{structural induction},
Structural induction principles cannot be extended to truly nested types, such as $\mathsf{Bush}$.
%To induce over a truly nested type, we need to use a \emph{deep induction} principle~\cite{jp20}.
Instead, for such data types it is necessary to use a \emph{deep induction} principle~\cite{jp20}.
Such a principle, unlike structural induction, inducts over all of the structured data present,
by traversing not just the outer structure with a predicate $\mathsf{P}$,
but also each data element contained in the data type with a custom predicate $\mathsf{Q}$.
This additional predicate is lifted to predicates on any internal structure containing these data, and the resulting predicates on these internal structures are lifted to predicates on any internal structures containing structures at the previous level, and so on, until the internal structures at all levels of the data type definition, including the top level, have been so processed.
Satisfaction of a predicate by the data at one level of a structure is then conditioned upon satisfaction of the appropriate predicates by all of the data at the preceding level.

For example, the deep induction rule for $\mathsf{Bush}$ is
\begin{multline}
\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to Bush\, a \to Set)
\to \big( \forall (a : Set) \to P\,a\,BNil \big)} \\
\mathsf{\to \big( \forall (a : Set) (Q : a \to Set) (x : a) (y : Bush\,(Bush\,a))} \\
\mathsf{\to Q\,x \to P\,(Bush\,a)\,(Bush^{\wedge}\,a\,Q)\,y \to P\,a\,Q\,(BCons\,x\,y) \big)} \\
\mathsf{\to \forall (a : Set) (Q : a \to Set) (x : Bush\,a) \to Bush^{\wedge}\,a\,Q\,x \to P\,a\,Q\,x }
\end{multline}
where $\mathsf{Bush^{\wedge} : \forall (a : Set) \to (a \to Set) \to Bush\,a \to Set}$ lifts a predicate $\mathsf{Q}$ on data of type $\mathsf{a}$ to a predicate on data of type $\mathsf{Bush\,a}$ asserting that $\mathsf{Q}$ holds for every element of type $\mathsf{a}$ contained in its argument bush.
%where $\mathsf{Bush^{\wedge} : \forall (a : Set) \to ( a \to Set ) \to Bush\,a \to Set}$
%is the polymorphic function that lifts predicates to $\mathsf{Bush}$,
It is defined as
\begin{align*}
\mathsf{Bush^{\wedge}\,a\,Q\,BNil} &= \mathsf{1} \\
\mathsf{Bush^{\wedge}\,a\,Q\,(BCons\,x\,y)} &= \mathsf{Q\,x \times Bush^{\wedge}\,(Bush\,a)\,(Bush^{\wedge}\,a\,Q)\,y}
\end{align*}

Despite deep induction being motivated by the need to produce an induction principle for truly nested types,
it can equally be applied to all other ADTs and nested types.
For example, the deep induction principle for $\mathsf{List}$ is
\begin{multline}
\mathsf{\forall (a : Set) (P : List\, a \to Set) (Q : a \to Set)} \\
\mathsf{\to P\,Nil \to \big( \forall (x : a) (xs: List\,a) \to Q\,x \to P\,xs \to P\,(Cons\,x\,xs)\big)} \\
\mathsf{\to \forall (xs : List\,a) \to List^{\wedge}\,a\,Q\,xs \to P\, xs }
\end{multline}
where $\mathsf{List^{\wedge} : \forall (a : Set) \to (a \to Set) \to List\,a \to Set}$ lifts a predicate $\mathsf{Q}$ on data of type $\mathsf{a}$ to a predicate on data of type $\mathsf{List\,a}$ asserting that $\mathsf{Q}$ holds for every element of its argument list.
Finally, the deep induction rule for $\mathsf{PTree}$ is
\begin{multline}
\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to PTree\, a \to Set)} \\
\mathsf{\to \big( \forall (a : Set) (Q : a \to Set) (x : a) \to Q\,x \to P\,a\,Q\,(PLeaf\, x) \big)} \\
\mathsf{\to \big( \forall (a : Set) (Q : a \to Set) (x : PTree\,(a \times a)) \to P\,(a \times a)\,(Pair^{\wedge}\,a\,Q)\,x \to P\,a\,Q\,(PNode\,x)\big)} \\
\mathsf{\to \forall (a : Set) (Q : a \to Set) (x : PTree\,a) \to PTree^{\wedge}\,a\,Q\,x \to P\,a\,Q\,x }
\end{multline}
where $\mathsf{Pair^{\wedge} : \forall (a : Set) \to (a \to Set) \to a \times a \to Set}$ lifts a predicate $\mathsf{Q}$ on $\mathsf{a}$ to a predicate on pairs of type $\mathsf{a \times a}$,
so that $\mathsf{Pair^{\wedge}\,a\,Q\,(x,y) = Q\,x \times Q\,y}$,
and $\mathsf{PTree^{\wedge} : \forall (a : Set) \to (a \to Set) \to PTree\,a \to Set}$ lifts a predicate $\mathsf{Q}$ on data of type $\mathsf{a}$ to a predicate on data of type $\mathsf{PTree\,a}$ asserting that $\mathsf{Q}$ holds for every element of type $\mathsf{a}$ contained in its argument perfect tree.

\section{Introducing GADTs}

%\begin{itemize}
%\item Say what GADTs are
%\item Example: $\mathsf{Equal}$
%\[\begin{array}{l}
%\mathsf{data\, Equal\,a\,b\,where}\\
%\mathsf{\;\;\;\;\;\;\;\;Refl ::\, Equal\,c\,c}
%\end{array}\]
%\item GADTs' encoding using $\mathsf{Equal}$
%\item Example (in Haskell syntax):
%%Haskell definition of $\mathsf{Seq}$ %(this encoding might have been introduced after the other one, see phantom types)
%\begin{equation}\label{eq:seq}
%\begin{array}{l}
%\mathsf{data\, Seq\,a\,where}\\
%\mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Seq\,a}\\
%\mathsf{\;\;\;\;\;\;\;\;SPair\,\, ::\, Seq \,a \to Seq\,b \to
%  Seq\,(a \times b)}
%\end{array}
%\end{equation}
%\item Traditionally GADTs are understood in terms of $\mathsf{Equal}$ data type (see Cheney Hinze or Tim Sheard)
%For example
%\begin{equation}\label{eq:eq_seq}
%\begin{array}{l}
%\mathsf{data\, Seq\,a\,where}\\
%\mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Seq\,a}\\
%\mathsf{\;\;\;\;\;\;\;\;SPair\,\, ::\, \exists (b : Set) (c : Set).\,Equal\,a\,(b \times c) \to Seq\,b \to Seq\,c \to
%  Seq\,a}\\ 
%\end{array}
%\end{equation}
%(This gives the kind of Church encoding that is in Bob's paper) (might omit)
%\item Add example for GADT with deep structure (Ask Daniel which one)
%\end{itemize}
%%Equality version of $\mathsf{Seq}$ (see Cheney Hinze or Tim Sheard)

As noted in Subsection~\ref{ssec:syntax-adts},
the return types of the constructors of a nested
type have to be the same type instance of the type being defined.
As a further generalization of ADTs and nested types,
\emph{generalized algebraic data types} (GADTs)~\cite{ch03,xcc03,sp04}
relax the restriction on the type instances appearing in a data
type definition by allowing their constructors both to take as
arguments \emph{and return as results} data whose types involve type
instances of the GADT other than the one being defined.

GADTs are used in precisely those situations in which different
behaviors at different instances of a data type are desired. This is
achieved by allowing the programmer to give the type signatures of the
GADT's data constructors independently, and then using pattern
matching to force the desired type refinement. Applications of GADTs
include generic programming, modeling programming languages via
higher-order abstract syntax, maintaining invariants in data
structures, and expressing constraints in embedded domain-specific
languages. GADTs have also been used, e.g., to implement tagless
interpreters~\cite{pl04,pr06,pvww06}, to improve memory
performance~\cite{min15}, and to design APIs~\cite{pen20}.

%\subsection{Agda syntax (bad title, revise)}

As a first and notable example of GADT, we consider the the $\mathsf{Equal}$ type.
This GADT is parametrized by two type indices,
but it is only possible to construct a data element if the two indices are instantiated at the same type.
In Agda, we code it as
\begin{equation}\label{eq:eq}
\begin{array}{l}
\mathsf{data\, Equal\,(a\,b : Set) : Set\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Refl :\, Equal\,c\,c}
\end{array}
\end{equation}
$\mathsf{Equal}$ has thus a single data element when its two type arguments
are the same and no data elements otherwise.

A more complex example for a GADT is
\begin{equation}\label{eq:eq-seq}
\begin{array}{l}
\mathsf{data\, Seq\,(a : Set) : Set\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Const :\, a \to Seq\,a}\\
\mathsf{\;\;\;\;\;\;\;\;SPair\,\, :\, Seq \,a \to Seq\,b \to
  Seq\,(a \times b)}
\end{array}
\end{equation}
which comprises sequences of any type $\mathsf{a}$
and sequences obtained by pairing the data in two already existing sequences.
Such GADTs can be understood in terms of the $\mathsf{Equal}$ data type~\cite{ch03,sp04}.
For example, we can rewrite the $\mathsf{Seq}$ type as
\begin{equation}\label{eq:eq_seq}
\begin{array}{l}
\mathsf{data\, Seq\,(a : Set) : Set\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Const :\, a \to Seq\,a}\\
\mathsf{\;\;\;\;\;\;\;\;SPair\,\, :\, \exists (b : Set) (c : Set).\,Equal\,a\,(b \times c) \to Seq\,b \to Seq\,c \to
  Seq\,a}\\ 
\end{array}
\end{equation}
where the requirement that the $\mathsf{SPair}$ constructor produce an instance of $\mathsf{Seq}$ at a product type has been replaced with the requirement that the instance of $\mathsf{Seq}$ returned by $\mathsf{SPair}$ is \emph{equal} to a product.
This encoding is particularly convenient when representing GADTs as Church encodings~\cite{vw10,atk12}.

Add third example here.

%{\color{blue} Do we really need the existential quantifiers?}

%To determine the possible shapes of a GADT's container we must
%pattern-match on the type of the data to be contained.
%For this, it is essential that a GADT calculus support an equality type
%$\mathsf{Equal}$ that is a singleton set when its two type arguments
%are the same and is the empty set otherwise.
%That is, the type $\mathsf{Equal}$ must be the syntactic reflection
%of semantic equality function $\mathit{Equal}$.
%The type $\mathsf{Equal}$ can either be
%defined via GADT syntax, as in
%\[\begin{array}{l}
%\mathsf{data\, Equal\,a\,b\,where}\\
%\mathsf{\;\;\;\;\;\;\;\;Refl ::\, Equal\,c\,c}
%\end{array}\]

\section{(Deep) induction for GADTs}

%\begin{itemize}
%    \item Recall structural induction does not work for GADTs because it does not work for truly nested types.
%    \item Induction rules are traditionally derived using functoriality
%    \item This cannot be extended to GADTs in this setting because they have no functorial semantics
%    \item This paper nevertheless shows how to extend (deep) induction to GADTs.
%\end{itemize}

As we have seen in Subsection~\ref{ssec:induction-nested},
truly nested types do not support a structural induction rule,
which is the reason why it was necessary to introduce a deep induction rule supporting them.
Consequently, GADTs do not support a structural induction rule either,
as they generalize nested types.
Still, there is hope for GADTs to support a deep induction rule,
like nested types do.

Induction rules, and specifically deep induction rules for nested types,
are traditionally derived using the functorial semantics of data types
in the setting of a parametric model~\cite{jp20}.
In particular, relational parametricity is used to validate the induction principle because induction is, itself,
a form of unary parametricity, where binary relations have been replaced with predicates,
which are essentially unary relations.
%in the sense that induction involves predicates, which are unary relations,
%as opposed to the binary relations involved in relational parametricity.

Unfortunately, this approach cannot possibly be employed
to prove a deep induction rule for GADTs,
as these types do not allow for a functorial interpretation,
at least in a parametric model~\cite{Haskell-paper}.

Nevertheless, this paper shows how to extend deep induction to GADTs.
We will first demonstrate how to derive the deep induction rule for some example GADTs,
and then provide a general principle that works for all GADTs.

\subsection{(Deep) induction for $\mathsf{Equal}$}

%\begin{itemize}
%\item Define $\mathsf{Equal}^{\wedge}$
%\[
%\mathsf{Equal^{\wedge} : \forall (a : Set) (b : Set) \to (a \to Set) \to (b \to Set) \to Equal\,a\,b \to Set}
%\]
%is defined by
%\[
%\mathsf{Equal^{\wedge}\,a\,a\,Q\,Q'\,Refl = \forall (x : a) \to Equal\,(Q\,x)(Q'\,x)}
%\]
%\item (Deep) induction rule for $\mathsf{Equal}$
%\item Prove deep induction for $\mathsf{Equal}$ (define term)
%\end{itemize}

We first derive the induction rule for the $\mathsf{Equal}$ type.
This will provide a simple case study that will help with the derivation of the induction rule of more complex GADTs.
More importantly, since we define GADTs using the $\mathsf{Equal}$ type,
as for example in Equation~\ref{eq:eq-seq},
this example will be instrumental in the induction rule of other GADTs.

%We first need to represent $\mathsf{Equal}$ as 
%\begin{equation}\label{eq:alt-eq}
%\begin{array}{l}
%\mathsf{data\, Equal\,(a\,b : Set) : Set\,where}\\
%\mathsf{\;\;\;\;\;\;\;\;Refl :\, \exists (c : Set) \to Equal\,a\,c \to Equal\,b\,c \to Equal\,a\,b}
%\end{array}
%\end{equation}
%\begin{multline}
%\mathsf{\forall (P : \forall (a\,b : Set) \to (a \to Set) \to (b \to Set) \to Equal\,a\,b \to Set)} \\
%\mathsf{\to \big( \forall (a : Set) (b : Set) (c : Set) (Q_a : a \to Set) (Q_b : b \to Set) (Q_c : c \to Set) (e_a: Equal\,a\,c) (e_b: Equal\,b\,c)}\\
%	\mathsf{\to Equal^{\wedge}\,a\,c\,Q_a\,Q_c\,e_a \to Equal^{\wedge}\,b\,c\,Q_b\,Q_c\,e_b \to P\,c\,c\,Q_c\,Q_c\,Refl \big)} \\
%\mathsf{\to \big( \forall (a : Set) (b : Set) (Q_a : a \to Set) (Q_b : b \to Set) (e: Equal\,a\,b)
%	\to Equal^{\wedge}\,a\,b\,Q_a\,Q_b\,e \to P\,a\,b\,Q_a\,Q_b\,e \big)}
%\end{multline}

To define the induction rule for $\mathsf{Equal}$ we first need a lifting operation
\[
\mathsf{Equal^{\wedge} : \forall (a : Set) (b : Set) \to (a \to Set) \to (b \to Set) \to Equal\,a\,b \to Set}
\]
defined as
\[
\mathsf{Equal^{\wedge}\,a\,a\,Q\,Q'\,Refl = \forall (x : a) \to Equal\,(Q\,x)(Q'\,x)}
\]
The induction rule for $\mathsf{Equal}$ has type
\begin{multline}
\mathsf{DIEqual : \forall (P : \forall (a\,b : Set) \to (a \to Set) \to (b \to Set) \to Equal\,a\,b \to Set)} \\
\mathsf{\to \big( \forall (c : Set) (Q : c \to Set) (Q' : c \to Set) \to Equal^{\wedge}\,c\,c\,Q\,Q'\,Refl \to P\,c\,c\,Q\,Q'\,Refl \big)} \\
\mathsf{\to \big( \forall (a : Set) (b : Set) (Q_a : a \to Set) (Q_b : b \to Set) (e: Equal\,a\,b)} \\
	\mathsf{\to Equal^{\wedge}\,a\,b\,Q_a\,Q_b\,e \to P\,a\,b\,Q_a\,Q_b\,e \big)}
\end{multline}
and it is defined as
\[
\mathsf{DIEqual\,P\,PRefl\,a\,a\,Q\,Q'\,Refl\,l = PRefl\,a\,Q\,Q'\,l}
\]
where
$\mathsf{PRefl : \forall (c : Set) (Q : c \to Set) (Q' : c \to Set) \to Equal^{\wedge}\,c\,c\,Q\,Q'\,Refl \to P\,c\,c\,Q\,Q'\,Refl}$,
$\mathsf{Q : a \to Set}$, $\mathsf{Q' : a \to Set}$
and $\mathsf{l : Equal^{\wedge}\,a\,a\,Q\,Q'\,Refl}$.

\subsection{(Deep) induction for $\mathsf{Seq}$}

Next we shall consider the $\mathsf{Seq}$ type defined in Equation~\ref{eq:eq_seq}.
Again, the first step in deriving its induction rule consists in defining the lifting
\[
\mathsf{Seq^{\wedge} : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set}
\]
which is given by
\[
\mathsf{Seq^{\wedge}\,a\,Q\,(Const\,x) = Q\,x}
\]
and
\begin{multline*}
\mathsf{Seq^{\wedge}\,a\,Q_a\,(SPair\,b\,c\,e\,s_b\,s_c)} \\
\mathsf{=
\exists (Q_b : b \to Set) (Q_c : c \to Set)
\to Equal^{\wedge}\,a\, (b \times c)\, Q_a\, (Q_b \times Q_c) \, e \times Seq^{\wedge}\,b\,Q_b\,s_b \times Seq^{\wedge}\,c\,Q_c\,s_c
}
\end{multline*}
We also need the lifting of the type $\mathsf{Pair = \forall (b : Set)(c : Set) \to b \times c}$, which is 
\[
\mathsf{Pair^{\wedge} : \forall (b : Set)(c : Set) \to (b \to Set) \to (c \to Set) \to b \times c \to Set}
\]
defined as
\[
\mathsf{Pair^{\wedge}\,b\,c\,Q_b\,Q_c\,(y, z) = Q_b\,y \times Q_c\,z}
\]

With these tools, we have an induction rule for $\mathsf{Seq}$
\begin{multline}
\mathsf{DISeq : \forall (P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set)} \\
\mathsf{\to \big( \forall (a : Set) (Q_a : a \to Set) (x : a) \to Q_a\,x \to P\,a\,Q_a\,(Const\,x) \big)} \\
\mathsf{\to \big( \forall (a : Set) (b : Set) (c : Set)
	(Q_a : a \to Set) (Q_b : b \to Set) (Q_c : c \to Set)} \\
	\mathsf{(s_b : Seq\,b) (s_c : Seq\,c) (e : Equal\,a\,(b \times c))
	\to Equal^{\wedge} a\, (b \times c)\, Q_a\, (Pair^{\wedge}\,b\,c\,Q_b\,Q_c)\, e} \\
	\mathsf{\to P\,b\,Q_b\,s_b \to P\,c\,Q_c\,s_c
	\to P a Q_a ( SPair\,b\,c\,e\,s_b\,s_c ) \big)} \\
\mathsf{\to \forall (a : Set)(Q_a : a \to Set)(s_a : Seq\,a) \to Seq^{\wedge}\,a\,Q_a\,s_a \to P\,a\,Q_a\,s_a}
\end{multline}
We define $\mathsf{DISeq\,P\,PConst\,PSPair\,a\,Q_a\,s_a\,l_a}$,
where
$\mathsf{P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set}$,
$\mathsf{Q_a : a \to Set}$, $\mathsf{s_a : Seq\,a}$
and $\mathsf{l_a : Seq^{\wedge}\,a\,Q\,s_a}$,
by pattern-matching on $\mathsf{s_a}$.
Let $\mathsf{s_a = Const\,x}$ for $\mathsf{x : a}$, and define
\[
\mathsf{DISeq\,P\,PConst\,PSPair\,a\,Q_a\,(Const\,x)\,l_a = PConst\,a\,Q_a\,x\,l_a}
\]
Notice that $\mathsf{Seq^{\wedge}\,a\,Q_a\,(Const\,x) = Q_a\,x}$, and thus $\mathsf{l_a : Q_a\,x}$,
making the right-hand-side in the above expression type-check.
Now, let $\mathsf{s_a = SPair\,b\,c\,e\,s_b\,s_c}$
for $\mathsf{e : Equal\,a\,(b \times c)}$, $\mathsf{s_b : Seq\,b}$ and $\mathsf{s_c : Seq\,c}$,
and define
\[
\mathsf{DISeq\,P\,PConst\,PSPair\,a\,Q_a\,(SPair\,b\,c\,e\,s_b\,s_c)\,(Q_b, Q_c, l_e, l_b, l_c)
	= PSPair\,a\,b\,c\,Q_a\,Q_b\,Q_c\,s_b\,s_c\,e\,l_e\,p_b\,p_c}
\]
where
$\mathsf{p_b : P\,b\,Q_b\,s_b}$, $\mathsf{p_c : P\,c\,Q_c\,s_c}$,
$\mathsf{l_e : Equal^{\wedge}\,a\, (b \times c)\, Q_a\, (Q_b \times Q_c) \, e}$,
$\mathsf{l_b : Seq^{\wedge}\,b\,Q_b\,s_b}$,
$\mathsf{l_c : Seq^{\wedge}\,c\,Q_c\,s_c}$
(notice that $\mathsf{(Q_b, Q_c, l_e, l_b, l_c) : Seq^{\wedge}\,a\,Q\,(SPair\,b\,c\,e\,x\,y)}$,
making the right-hand-side in the above expression type-check)
and $\mathsf{p_b}$ and $\mathsf{p_c}$ are defined as follows:
\begin{align*}
\mathsf{p_b} &=\mathsf{DISeq\,P\,PConst\,PSPair\,b\,Q_b\,s_b\,l_b : P\,b\,Q_b\,s_b} \\
\mathsf{p_c} &=\mathsf{DISeq\,P\,PConst\,PSPair\,c\,Q_c\,s_c\,l_c : P\,c\,Q_c\,s_c}
\end{align*}

%\subsection{(Deep) induction for $\mathsf{Seq}$ (general style)}
%
%We need the lifting of the type $\mathsf{Pair = \forall (b : Set)(c : Set) \to b \times c}$, which is 
%\[
%\mathsf{Pair^{\wedge} : \forall (b : Set)(c : Set) \to (b \to Set) \to (c \to Set) \to b \times c \to Set}
%\]
%defined as
%\[
%\mathsf{Pair^{\wedge}\,b\,c\,Q_b\,Q_c\,(y, z) = Q_b\,y \times Q_c\,z}
%\]
%
%%Let 
%%$\mathsf{F, F_1, F_2 : (Set \to Set) \to Set \to Set \to Set}$
%%be defined as
%%$\mathsf{F_1\,G\,b\,c = G\,b}$,
%%$\mathsf{F_2\,G\,b\,c = G\,c}$,
%%$\mathsf{F\,G\,b\,c = F_1\,G\,b\,c \times F_2\,G\,b\,c}$,
%%and $\mathsf{K = Pair}$.
%
%
%Let 
%$\mathsf{F : (Set \to Set) \to Set \to Set \to Set}$
%be defined as
%$\mathsf{F\,G\,b\,c = G\,b \times G\,c}$
%and $\mathsf{K = Pair}$.
%
%%The type of $\mathsf{F_i^{\wedge}}$ is
%%\begin{multline*}
%%\mathsf{\forall (G : Set \to Set) (b : Set) (c : Set)
%%\to (\forall (a : Set) \to (a \to Set) \to G\,a \to Set)} \\
%%\mathsf{\to (b \to Set) \to (c \to Set) \to (F_i\,G\,b\,c) \to Set}
%%\end{multline*}
%The type of $\mathsf{F^{\wedge}}$ is
%\begin{multline*}
%\mathsf{\forall (G : Set \to Set) (b : Set) (c : Set)
%\to (\forall (a : Set) \to (a \to Set) \to G\,a \to Set)} \\
%\mathsf{\to (b \to Set) \to (c \to Set) \to (F_1\,G\,b\,c \times F_2\,G\,b\,c) \to Set}
%\end{multline*}
%%and $\mathsf{F^{\wedge}\,G\,b\,c\,P\,Q_b\,Q_c\,(s_b, s_c)
%%= F_1^{\wedge}\,G\,b\,c\,P\,Q_b\,Q_c\,s_b \times F_2^{\wedge}\,G\,b\,c\,P\,Q_b\,Q_c\,s_c}$
%
%Next we shall consider the $\mathsf{Seq}$ type defined in Equation~\ref{eq:eq_seq}.
%Again, the first step in deriving its induction rule consists in defining the lifting
%\[
%\mathsf{Seq^{\wedge} : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set}
%\]
%which is given by
%\[
%\mathsf{Seq^{\wedge}\,a\,Q\,(Const\,x) = Q\,x}
%\]
%and
%\begin{multline*}
%\mathsf{Seq^{\wedge}\,a\,Q_a\,(SPair\,b\,c\,e\,(s_b,s_c))
%=
%\exists (Q_b : b \to Set) (Q_c : c \to Set)} \\
%\mathsf{
%\to Equal^{\wedge}\,a\, (Pair\,b\,c)\, Q_a\, (Pair^{\wedge}\,b\,c\,Q_b\,Q_c) \, e
%\times F^{\wedge}\,Seq\,b\,c\,Seq^{\wedge}\,Q_b\,Q_c\,(s_b, s_c)
%}
%\end{multline*}
%%\begin{multline*}
%%\mathsf{Seq^{\wedge}\,a\,Q_a\,(SPair\,b\,c\,e\,(s_b,s_c))
%%=
%%\exists (Q_b : b \to Set) (Q_c : c \to Set)} \\
%%\mathsf{
%%\to Equal^{\wedge}\,a\, (Pair\,b\,c)\, Q_a\, (Pair^{\wedge}\,b\,c\,Q_b\,Q_c) \, e
%%\times F_1^{\wedge}\,Seq\,b\,c\,Seq^{\wedge}\,Q_b\,Q_c\,s_b
%%\times F_2^{\wedge}\,Seq\,b\,c\,Seq^{\wedge}\,Q_b\,Q_c\,s_c
%%}
%%\end{multline*}
%
%Define
%\[
%\mathsf{PredMap : \forall (a : Set) \to (a \to Set) \to (a \to Set) \to Set}
%\]
%as $\mathsf{PredMap\,a\,Q_1\,Q_2 = \forall (x : a) \to Q_1\,x \to Q_2\,x}$
%
%%  -- Lifting also acts on morphisms between predicates with the same carrier
%%  {-# TERMINATING #-}
%%  Bush^-map : ∀ {l} {X : Set} {Q1 Q2 : X → Set l}
%%              → Pred-map {l} Q1 Q2 → Pred-map (Bush^ Q1) (Bush^ Q2)
%%  Bush^-map Q12 Bnil q = bigtt
%%  Bush^-map Q12 (Bcons x b) (y , bby) =
%%                Q12 x y , Bush^-map (Bush^-map Q12) b bby
%Define
%\[
%\mathsf{SeqLMap : \forall (a : Set) (Q_1\,Q_2 : a \to Set) \to PredMap\,a\,Q_1\,Q_2 \to PredMap\,(G\,a)\,(Seq^{\wedge}\,a\,Q_1)\,(Seq^{\wedge}\,a\,Q_2)}
%\]
%as
%\begin{gather*}
%\mathsf{
%SeqLMap\,a\,Q_1\,Q_2\,M\,(Const\,x)\,z = M\,x\,z
%} \\
%\mathsf{
%SeqLMap\,a\,Q_1\,Q_2\,M\,(SPair\,b\,c\,e\,s_b\,s_c)\,(Q_b, Q_c, l_E, l_F) =
%}
%\end{gather*}
%
%
%We have an induction rule for $\mathsf{Seq}$
%\begin{multline}
%\mathsf{DISeq : \forall (P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set)} \\
%\mathsf{\to \big( \forall (a : Set) (Q_a : a \to Set) (x : a) \to Q_a\,x \to P\,a\,Q_a\,(Const\,x) \big)} \\
%\mathsf{\to \big( \forall (a : Set) (b : Set) (c : Set)
%	(Q_a : a \to Set) (Q_b : b \to Set) (Q_c : c \to Set)} \\
%	\mathsf{( (s_b, s_c) : F\,Seq\,b\,c) (e : Equal\,a\,(Pair\,b\,c))
%	\to Equal^{\wedge} a\, (Pair\,b\,c)\, Q_a\, (Pair^{\wedge}\,b\,c\,Q_b\,Q_c)\, e} \\
%	\mathsf{\to F^{\wedge}\,G\,b\,c\,P\,Q_b\,Q_c\,(s_b,s_c)
%	\to P a Q_a ( SPair\,b\,c\,e\,(s_b,s_c) ) \big)} \\
%\mathsf{\to \forall (a : Set)(Q_a : a \to Set)(s_a : Seq\,a) \to Seq^{\wedge}\,a\,Q_a\,s_a \to P\,a\,Q_a\,s_a}
%\end{multline}
%We define $\mathsf{DISeq\,P\,PConst\,PSPair\,a\,Q_a\,s_a\,l_a}$,
%where
%$\mathsf{P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set}$,
%$\mathsf{Q_a : a \to Set}$, $\mathsf{s_a : Seq\,a}$
%and $\mathsf{l_a : Seq^{\wedge}\,a\,Q\,s_a}$,
%by pattern-matching on $\mathsf{s_a}$.
%Let $\mathsf{s_a = Const\,x}$ for $\mathsf{x : a}$, and define
%\[
%\mathsf{DISeq\,P\,PConst\,PSPair\,a\,Q_a\,(Const\,x)\,l_a = PConst\,a\,Q_a\,x\,l_a}
%\]
%Notice that $\mathsf{Seq^{\wedge}\,a\,Q_a\,(Const\,x) = Q_a\,x}$, and thus $\mathsf{l_a : Q_a\,x}$,
%making the right-hand-side in the above expression type-check.
%Now, let $\mathsf{s_a = SPair\,b\,c\,e\,(s_b,s_c)}$
%for $\mathsf{e : Equal\,a\,(Pair\,b\,c)}$ and $\mathsf{(s_b, s_c) : F\,Seq\,b\,c}$,
%and define
%\[
%\mathsf{DISeq\,P\,PConst\,PSPair\,a\,Q_a\,(SPair\,b\,c\,e\,(s_b,s_c))\,(Q_b, Q_c, l_e, l_{F_1}, l_{F_2})
%	= PSPair\,a\,b\,c\,Q_a\,Q_b\,Q_c\,(s_b,s_c)\,e\,l_e\,p}
%\]
%where
%$\mathsf{p : F^{\wedge}\,G\,b\,c\,P\,Q_b\,Q_c\,(s_b,s_c)}$,
%$\mathsf{l_e : Equal^{\wedge}\,a\, (Pair\,b\,c)\, Q_a\, (Pair^{\wedge}\,Q_b\,Q_c) \, e}$,
%$\mathsf{l_{F_1} : F_1^{\wedge}\,Seq\,b\,c\,Seq^{\wedge}\,Q_b\,Q_c\,s_b}$,
%$\mathsf{l_{F_2} : F_2^{\wedge}\,Seq\,b\,c\,Seq^{\wedge}\,Q_b\,Q_c\,s_c}$
%(notice that $\mathsf{(Q_b, Q_c, l_e, l_{F_1}, l_{F_2}) : Seq^{\wedge}\,a\,Q\,(SPair\,b\,c\,e\,x\,y)}$,
%making the right-hand-side in the above expression type-check)
%and $\mathsf{p}$ is defined as follows:
%\[
%\mathsf{p = (???, ???) 
%: F_1^{\wedge}\,G\,b\,c\,P\,Q_b\,Q_c\,s_b \times F_2^{\wedge}\,G\,b\,c\,P\,Q_b\,Q_c\,s_c} \\
%\]

%\begin{align*}
%\mathsf{DISeq\,P\,PConst\,PSPair\,a\,Q\,(Const\,x)\,l}
%	&=\mathsf{PConst\,a\,Q\,x\,l} \\
%\mathsf{DISeq\,P\,PConst\,PSPair\,a\,Q_a\,(SPair\,b\,c\,e\,x\,y)\,(Q_b\,Q_c\,e', l_b, l_c)}
%	&=\mathsf{PSpair\,a\,b\,c\,Q_a\,Q_b\,Q_c\,x\,y\,e\,p_b\,p_c\,e'}
%\end{align*}
%where
%$\mathsf{Q_a : a \to Set}$, $\mathsf{Q_b : b \to Set}$, $\mathsf{Q_c : c \to Set}$,
%$\mathsf{x : Seq\,b}$, $\mathsf{y : Seq\,c}$,
%$\mathsf{e : Equal\,a\,(b \times c)}$,
%$\mathsf{p_b : P\,b\,Q_b\,x}$, $\mathsf{p_c : P\,c\,Q_c\,y}$,
%$\mathsf{e' : Equal^{\wedge}\,a\, (b \times c)\, Q\, (Q_b \times Q_c) \, e}$,
%$\mathsf{l_b : Seq^{\wedge}\,b\,Q_b\,x}$,
%$\mathsf{l_c : Seq^{\wedge}\,c\,Q_c\,y}$
%and
%\begin{align*}
%\mathsf{p_b} &=\mathsf{DISeq\,P\,PConst\,PSPair\,b\,Q_b\,x\,l_b : P\,b\,Q_b\,x} \\
%\mathsf{p_c} &=\mathsf{DISeq\,P\,PConst\,PSPair\,c\,Q_c\,y\,l_c : P\,c\,Q_c\,y}
%\end{align*}
%
%\begin{align*}
%\mathsf{Q_a} &: \mathsf{a \to Set} \\
%\mathsf{Q_b} &: \mathsf{b \to Set} \\
%\mathsf{Q_c} &: \mathsf{c \to Set} \\
%\mathsf{x} &: \mathsf{Seq\,b} \\
%\mathsf{y} &: \mathsf{Seq\,c} \\
%\mathsf{e} &: \mathsf{Equal\,a\,(b \times c)} \\
%\mathsf{p_b} &: \mathsf{P\,b\,Q_b\,x} \\
%\mathsf{p_c} &: \mathsf{P\,c\,Q_c\,y} \\
%\mathsf{e'} &: \mathsf{Equal^{\wedge}\,a\, (b \times c)\, Q\, (Q_b \times Q_c) \, e} \\
%\mathsf{l_b} &: \mathsf{Seq^{\wedge}\,b\,Q_b\,x} \\
%\mathsf{l_c} &: \mathsf{Seq^{\wedge}\,c\,Q_c\,y} \\
%\mathsf{p_b = DISeq\,P\,PConst\,PSPair\,b\,Q_b\,x\,l_b} &: \mathsf{P\,b\,Q_b\,x} \\
%\mathsf{p_c = DISeq\,P\,PConst\,PSPair\,c\,Q_c\,y\,l_c} &: \mathsf{P\,c\,Q_c\,y}
%\end{align*}

\subsection{Third example introduced above}

%Same structure as previous two.
%Runs through the paper.
%Still to decide?

\subsection{General case}

%Outline general heuristics for this process, generalizing process in deep induction paper.
%\begin{itemize}
%    \item Lifting for GADT (possibly a restricted class?)
%    \item Deep induction rule
%    \item Prove deep induction (define term)
%\end{itemize}

TODO:
\begin{itemize}
\item explain overline is vector
\end{itemize}

In this section we will generalize the approach taken in the previous examples
and provide a general framework to derive induction rules for GADTs.
We restrict our scope to GADTs not featuring nesting,
in the sense that their constructors contain no nested calls of the recursive variable,
as truly nested types do.
The reason for this restriction will be discussed in Section~\ref{sec:GADT-nested}.

Consider a generic GADT
\begin{equation}\label{eq:gadt}
\begin{array}{l}
\mathsf{data\, G\,(\ol{a : Set}) : Set\,where}\\
\mathsf{\;\;\;\;\;\;\;\;C_i\, :\, F_i\,G\,\ol{b} \to G (\ol{K_i\,\ol{b}})}
\end{array}
\end{equation}
with a finite number of constructors $\mathsf{C_i}$,
and where $\mathsf{F_i}$ has type $\mathsf{(Set^{\lvert\ol{a}\rvert} \to Set) \to Set^{\lvert\ol{b}\rvert} \to Set}$
and each $\mathsf{K_i}$ has type $\mathsf{Set^{\lvert\ol{b}\rvert} \to Set}$.
Each constructor $\mathsf{C_i}$ can be rewritten as
\[
\mathsf{C_i : \exists (\ol{b : Set}) \to \ol{Equal\,a\,(K_i\,\ol{b})} \to F_i\,G\,\ol{b} \to G\,\ol{a}}
\]
which is the form we shall use from now on.

We define the lifting of $\mathsf{G}$
\[
\mathsf{G^{\wedge} : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set}
\]
as
\[
\mathsf{G^{\wedge}\,\ol{a}\,\ol{Q_a}\,(C_i\,\ol{b}\,\ol{e_a}\,x)
= \exists (\ol{Q_b : b \to Set})
\to \ol{Equal^{\wedge}\,a\,(K_i\,\ol{b})\,Q_a\,(K_i^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
\times F_i^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,{\ol{Q_b}}\,x}
\]
where $\mathsf{Q_a : a \to Set}$, $\mathsf{e_a : Equal\,a\,(K_i\,\ol{b})}$ and $\mathsf{x : F_i\,G\,\ol{b}}$,
and the liftings
\begin{multline*}
\mathsf{F_i^{\wedge} : \forall (G : Set^{\lvert\ol{a}\rvert} \to Set) (\ol{b : Set})
\to (\forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set)} \\
\mathsf{\to (\ol{b \to Set})
\to F_i\,G\,\ol{b} \to Set}
\end{multline*}
and
\[
\mathsf{K_i^{\wedge} : \forall (\ol{b : Set}) \to (\ol{b \to Set}) \to K_i\,\ol{b} \to Set}
\]
are defined by induction on the structure of the type.

Define
\[
\mathsf{PredMap : \forall (a : Set) \to (a \to Set) \to (a \to Set) \to Set}
\]
as $\mathsf{PredMap\,a\,Q_1\,Q_2 = \forall (x : a) \to Q_1\,x \to Q_2\,x}$

%  -- Lifting also acts on morphisms between predicates with the same carrier
%  {-# TERMINATING #-}
%  Bush^-map : ∀ {l} {X : Set} {Q1 Q2 : X → Set l}
%              → Pred-map {l} Q1 Q2 → Pred-map (Bush^ Q1) (Bush^ Q2)
%  Bush^-map Q12 Bnil q = bigtt
%  Bush^-map Q12 (Bcons x b) (y , bby) =
%                Q12 x y , Bush^-map (Bush^-map Q12) b bby
Define
\[
\mathsf{GLMap : \forall (a : Set) (Q_1\,Q_2 : a \to Set) \to PredMap\,a\,Q_1\,Q_2 \to PredMap\,(G\,a)\,(G^{\wedge}\,a\,Q_1)\,(G^{\wedge}\,a\,Q_2)}
\]
as
\[
\mathsf{
GLMap\,\ol{a}\,\ol{Q_1}\,\ol{Q_2}\,\ol{M}\,(C_i\,\ol{b}\,\ol{e_a}\,x)(\ol{Q_b}, \ol{e'_a}, l)
= \ol{M}
}
\]
%\begin{gather*}
%\end{gather*}

Associate the type
\begin{multline}
\mathsf{PC_i = \forall (\ol{a : Set}) (\ol{b : Set}) (\ol{Q_a : a \to Set}) (\ol{Q_b : b \to Set}) (\ol{e_a : Equal\,a\,(K_i\,\ol{b})}) (x : F_i\,G\,\ol{b})} \\
\mathsf{\to \ol{Equal^{\wedge}\,a\,(K_i\,\ol{b})\,Q_a\,(K_i^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
	\to F_i^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x
	\to P\,\ol{a}\,\ol{Q_a}\,(C_i\,\ol{b}\,\ol{e_a}\,x)}
\end{multline}
to each constructor $\mathsf{C_i}$.

Define the induction rule for $\mathsf{G}$
\begin{multline}
\mathsf{DIG : \forall (P : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set)} \\
\mathsf{\to \bigtimes_i PC_i
\to \forall (\ol{a : Set})(\ol{Q_a : a \to Set})(x : G\,\ol{a}) \to G^{\wedge}\,\ol{a}\,\ol{Q_a}\,x \to P\,\ol{a}\,\ol{Q_a}\,x}
\end{multline}
as
\[
\mathsf{DIG\,P\,\ol{pc_i}\,\ol{a}\,\ol{Q_a}\,(C_i\,\ol{b}\,\ol{e_a}\,x)\,(\ol{Q_b}, \ol{e'_a}, l)
= pc_i\,\ol{a}\,\ol{b}\,\ol{Q_a}\,\ol{Q_b}\,\ol{e_a}\,x\,\ol{e'_a}\,\ol{p}}
\]
where
\[
\mathsf{p = ??? : F_i^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}
\]

\begin{lemma}
For every $\mathsf{F, G, \ol{b}, P, \ol{Q_b}, x}$, there exists a function
\[
\mathsf{p : F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x \to F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}
\]
\end{lemma}
\begin{proof}
By structural induction on $\mathsf{F}$:
\begin{itemize}
\item Case $\mathsf{F = F_1 \times F_2}$.
Thus, $\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x = F_1^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x \times F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}$.
By inductive hypothesis, there exist functions
$\mathsf{p_1 : F_1^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x_1 \to ??? \to F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x_1}$ and
$\mathsf{p_2 : F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x_2 \to ??? \to F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x_2}$.
%%%
\item Case $\mathsf{F\,G\,\ol{b} = G(\ol{F_a\,\ol{b}})}$.
Thus, $\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b} = P (\ol{F_a\,\ol{b}}) (\ol{F_a^{\wedge}\,\ol{b}\,\ol{Q_b}})}$.
So, $\mathsf{p : P (\ol{F_a\,\ol{b}}) (\ol{F_a^{\wedge}\,\ol{b}\,\ol{Q_b}}) x}$ is
\[
\mathsf{p = DIG\,P\,\ol{pc_i}\,(\ol{F_a\,\ol{b}})\,(\ol{F_a^{\wedge}\,\ol{b}\,\ol{Q_b}})\,x\,l}
\]
Note: No nesting!
\end{itemize}
\end{proof}

%Define $\mathsf{p_b}$ by structural induction on $F_i$.
%
%Case $\mathsf{F_i\,G\,\ol{b} = F'_i\,G\,\ol{b} \times F''_i\,G\,\ol{b}}$.
%We have $\mathsf{F_i^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,(x', x'') = (F'_i)^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x' \times (F''_i)^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x''}$.
%Thus define
%\[
%\mathsf{
%p_b\,P\,PC_i\,\ol{a}\,\ol{Q_a}\,\ol{b}\,\ol{e_a}\,(x', x'')\,\ol{e'_a}\,(l', l'')
%=
%(
%p'_b\,P\,PC_i\,\ol{a}\,\ol{Q_a}\,\ol{b}\,\ol{e_a}\,x'\,\ol{e'_a}\,l',
%p''_b\,P\,PC_i\,\ol{a}\,\ol{Q_a}\,\ol{b}\,\ol{e_a}\,x''\,\ol{e'_a}\,l''
%)
%} 
%\]
%where $\mathsf{p'_b}$ is the function associated to $\mathsf{F'_i}$
%and $\mathsf{p''_b}$ is the function associated to $\mathsf{F''_i}$.
%
%
%Case $\mathsf{F_i\,G\,\ol{b} = H(\ol{F'_i\,G\,\ol{b}})}$ with $\mathsf{H}$ data type with constructors $\mathsf{D_i}$.
%We have
%\[
%\mathsf{F_i^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x = H^{\wedge}\,(\ol{F'_i\,G\,\ol{b}}) (\ol{(F'_i)^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}}) x}
%\]
%Thus define
%\[
%\mathsf{
%p_b\,P\,PC_i\,\ol{a}\,\ol{Q_a}\,\ol{b}\,\ol{e_a}\,x\,\ol{e'_a}\,l
%=
%} 
%\]

%\subsection{General case (with product)}
%
%Note: case with one constructor to make things simpler
%
%Consider a generic GADT
%\begin{equation}\label{eq:gadt}
%\begin{array}{l}
%\mathsf{data\, G\,(\ol{a : Set}) : Set\,where}\\
%\mathsf{\;\;\;\;\;\;\;\;C\, :\, F_1\,G\,\ol{b} \times \dots \times F_n\,G\,\ol{b} \to G (\ol{K\,\ol{b}})}
%\end{array}
%\end{equation}
%where each $\mathsf{F_i}$ has type $\mathsf{(Set^{\lvert\ol{a}\rvert} \to Set) \to Set^{\lvert\ol{b}\rvert} \to Set}$
%and $\mathsf{K}$ has type $\mathsf{Set^{\lvert\ol{b}\rvert} \to Set}$.
%The constructor $\mathsf{C}$ can be rewritten as
%\[
%\mathsf{C : \exists (\ol{b : Set}) \to \ol{Equal\,a\,(K\,\ol{b})} \to F_1\,G\,\ol{b} \to F_n\,G\,\ol{b} \to G\,\ol{a}}
%\]
%which is the form we shall use from now on.
%
%We define the lifting of $\mathsf{G}$
%\[
%\mathsf{G^{\wedge} : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set}
%\]
%as
%\begin{multline*}
%\mathsf{G^{\wedge}\,\ol{a}\,\ol{Q_a}\,(C\,\ol{b}\,\ol{e_a}\,x_1 \dots x_n)} \\
%\mathsf{= \exists (\ol{Q_b : b \to Set})
%\to \ol{Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
%\times F_1^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,{\ol{Q_b}}\,x_1
%\dots
%\times F_n^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,{\ol{Q_b}}\,x_n}
%\end{multline*}
%where $\mathsf{Q_a : a \to Set}$, $\mathsf{e_a : Equal\,a\,(K\,\ol{b})}$ and $\mathsf{x_i : F_i\,G\,\ol{b}}$,
%and the liftings
%\begin{multline*}
%\mathsf{F_i^{\wedge} : \forall (G : Set^{\lvert\ol{a}\rvert} \to Set) (\ol{b : Set})
%\to (\forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set)} \\
%\mathsf{\to (\ol{b \to Set})
%\to F_i\,G\,\ol{b} \to Set}
%\end{multline*}
%and
%\[
%\mathsf{K^{\wedge} : \forall (\ol{b : Set}) \to (\ol{b \to Set}) \to K\,\ol{b} \to Set}
%\]
%are defined by induction on the structure of the type.
%
%Define the induction rule for $\mathsf{G}$
%\begin{multline}
%\mathsf{DIG : \forall (P : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set)} \\
%\mathsf{\to \bigtimes_i \big( \forall (\ol{a : Set}) (\ol{b : Set}) (\ol{Q_a : a \to Set}) (\ol{Q_b : b \to Set}) (\ol{e_a : Equal\,a\,(K\,\ol{b})}) (x_i : F_i\,G\,\ol{b})} \\
%\mathsf{\to \ol{Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
%	\to F_1^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x_1 \to \dots \to F_n^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x_n
%	\to P\,\ol{a}\,\ol{Q_a}\,(C\,\ol{b}\,\ol{e_a}\,x_1 \dots x_n) \big)} \\
%\mathsf{\to \forall (\ol{a : Set})(\ol{Q_a : a \to Set})(x : G\,\ol{a}) \to G^{\wedge}\,\ol{a}\,\ol{Q_a}\,x \to P\,\ol{a}\,\ol{Q_a}\,x}
%\end{multline}
%as
%\[
%\mathsf{DIG\,P\,PC\,\ol{a}\,\ol{Q_a}\,(C\,\ol{b}\,\ol{e_a}\,x_1 \dots x_n)\,(\ol{Q_b}, \ol{e'_a}, l_{F_1}, \dots, l_{F_n})
%= PC\,\ol{a}\,\ol{b}\,\ol{Q_a}\,\ol{Q_b}\,\ol{e_a}\,x_1 \dots x_n\,\ol{e'_a}\,p_1 \dots p_n}
%\]
%where
%\[
%\mathsf{p_i = DIG\,P\,PC\,\ol{a}\,\ol{Q_a} ? : F_i^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x_i}
%\]

%Consider a generic GADT
%\begin{equation}\label{eq:gadt}
%\begin{array}{l}
%\mathsf{data\, G\,(\ol{a : Set}) : Set\,where}\\
%\mathsf{\;\;\;\;\;\;\;\;C_i\, :\, f_i[G/\phi,\ol{b/\beta}] \to G (\ol{k_i[\ol{b/\beta}]})}
%\end{array}
%\end{equation}
%with a finite number of constructors $\mathsf{C_i}$.
%Each constructor $\mathsf{C_i}$ is defined by type expressions $\mathsf{f_i}$,
%containing a type constructor variable $\mathsf{\phi}$ of arity matching the number of type arguments of $G$,
%and type variables $\mathsf{\ol{\beta}}$;
%and $\mathsf{\ol{g_i}}$ containing the type variables $\mathsf{\ol{\beta}}$.
%Each constructor $\mathsf{C_i}$ can be rewritten as
%\[
%\mathsf{C_i : \exists (\ol{b : Set}) \to \ol{Equal\,a\,(k_i[\ol{b/\beta}])} \to f_i[G/\phi,\ol{b/\beta}] \to G\,\ol{a}}
%\]
%which is the form we shall use from now on.
%
%We define the types
%\begin{gather*}
%	\mathsf{F_i : (Set^{\lvert\ol{a}\rvert} \to Set) \to Set^{\lvert\ol{b}\rvert} \to Set} \\
%	\mathsf{F_i }
%\end{gather*}
%
%
%Define the lifting of $\mathsf{G}$
%\[
%\mathsf{G^{\wedge} : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set}
%\]
%as
%\[
%\mathsf{G^{\wedge}\,\ol{a}\,\ol{P_a}\,(C_i\,\ol{b}\,\ol{e_a}\,x)
%= \exists (\ol{Q_b : b \to Set})
%\to \ol{Equal^{\wedge}\,a\,k_i[\ol{b/\beta}]\,P_a\,(k_i^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
%\times f_i[G/\phi]^{\wedge}\,\ol{b}\,{\ol{Q_b}}\,x}
%\]
%where $\mathsf{P_a : a \to Set}$, $\mathsf{e_a : Equal\,a\,(k_i\,\ol{b})}$ and $\mathsf{x : f_i\,\ol{b}}$,
%and the liftings $\mathsf{f_i^{\wedge}}$ and $\mathsf{k_i^{\wedge}}$
%are defined the same way by induction on the structure of the type.

\section{Induction for GADTs with nesting}\label{sec:GADT-nested}

In the previous sections,
we derive induction rules for GADTs that do not feature nesting,
in the sense that their constructors contain no nested calls of the recursive variable,
as truly nested types do.
Since both nested types~\cite{jp20}
and GADTs without nesting admit induction rules,
it is just natural to expect that GADTs with nesting would as well.
Surprisingly, that is not the case:
indeed, induction rules for nested types rely on functorial semantics,
but GADTs cannot admit both functorial and parametric semantics at the same time~\cite{HaskellPaper}.
In this section we will show how induction for GADTs featuring nesting goes wrong
by analyzing a concrete example.

Consider the type
\begin{equation}
\begin{array}{l}
\mathsf{data\, G\,(a : Set) : Set\,where}\\
\mathsf{\;\;\;\;\;\;C :\, G(G\,a) \to G(a \times a)}
\end{array}
\end{equation}
whose constructor $\mathsf{C}$ can be rewritten as
\[
\mathsf{C : \exists (b : Set) \to Equal\,a\,(b \times b) \to G(G\,b) \to G\,a}
\]
which is the form we shall use from now on.
Its predicate-lifting function
\[
\mathsf{G^{\wedge} : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set}
\]
is defined as
\[
\mathsf{G^{\wedge}\,a\,Q_a\,(C\,b\,e\,x)
= \exists (Q_b : b \to Set)
\to Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e
\times G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,x}
\]
where $\mathsf{Q_a : a \to Set}$, $\mathsf{e : Equal\,a\,(b \times b)}$ and $\mathsf{x : G(G\,b)}$.

We need to define an induction rule for $\mathsf{G}$.
That has to be a term
\begin{multline}
\mathsf{DIG : \forall (P : \forall (a : Set) \to (a \to Set) \to G\,a \to Set)} \\
\mathsf{\to \Big( \forall (a\;b : Set) (Q_a : a \to Set) (Q_b : b \to Set) (e : Equal\,a\,(b \times b)) (x : G\,(G\,b))} \\
\mathsf{\to Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e
	\to P\,(G\,b)\,(P\,b\,Q_b)\,x
	\to P\,a\,Q_a\,(C\,b\,e\,x) \Big)}\\
\mathsf{\to \forall (a : Set)(Q_a : a \to Set)(x : G\,a) \to G^{\wedge}\,a\,Q_a\,x \to P\,a\,Q_a\,x}
\end{multline}
Consistently with the examples in the previous sections, we define $\mathsf{DIG}$ as
\[
\mathsf{DIG\,P\,pc\,a\,Q_a\,(C\,b\,e\,x)\,(Q_b, L_E, L_G)
= pc\,a\,b\,Q_a\,Q_b\,e\,x\,L_E\,p}
\]
where $\mathsf{(Q_b, L_E, L_G)}$, the last term applied to $\mathsf{DIG}$, has type $\mathsf{G^{\wedge}\,a\,Q_a\,(C\,b\,e\,x)}$,
i.e., $\mathsf{Q_b : b \to Set}$, $\mathsf{L_E : Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e}$,
and $\mathsf{L_G : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,x}$.
%\begin{gather*}
%\mathsf{Q_b : b \to Set} \\
%\mathsf{L_E : Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e} \\
%\mathsf{L_G : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,x}
%%\mathsf{p = DIG\,P\,pc\,(G\,b)\,(P\,b\,Q_b)\,x\,q : P\,(G\,b)\,(P\,b\,Q_b)\,x}
%\end{gather*}
We still need to define $\mathsf{p : P\,(G\,b)\,(P\,b\,Q_b)\,x}$.
We do so by using the induction rule, letting $\mathsf{p = DIG\,P\,pc\,(G\,b)\,(P\,b\,Q_b)\,x\,q}$
where we still need to provide $\mathsf{q : G^{\wedge}\,(G\,b)\,(P\,b\,Q_b)\,x}$.
To produce such $\mathsf{q}$, we need the map function of $\mathsf{G^{\wedge}}$, with type
\[
\mathsf{GLMap : \forall (a : Set) (Q_a\;Q_a' : a \to Set) \to PredMap\,a\,Q_a\,Q_a' \to PredMap\,(G\,a)\,(G^{\wedge}\,a\,Q_a)\,(G^{\wedge}\,a\,Q_a')}
\]
where $\mathsf{PredMap : \forall (a : Set) \to (a \to Set) \to (a \to Set) \to Set}$
is defined as
\[
\mathsf{PredMap\,a\,Q_a\,Q_a' = \forall (x : a) \to Q_a\,x \to Q_a'\,x}
\]
and represents the type of morphisms between predicates.
If we had $\mathsf{GLMap}$, then we would be able to define
\[
\mathsf{q = GLMap\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,(P\,b\,Q_b)\,(DIG\,P\,pc\,b\,Q_b)\,x\,L_G}
\]
Unfortunately, we cannot define such a $\mathsf{GLMap}$.
Indeed, its definition would have to be
\[
\mathsf{GLMap\,a\,Q_a\,Q_a'\,M\,(C\,b\,e\,x)\,(Q_b, L_E, L_G) = (Q_b', L_E', L_G')}
\]
where $\mathsf{(Q_b, L_E, L_G)}$ has type $\mathsf{G^{\wedge}\,a\,Q_a\,(C\,b\,e\,x)}$
and $\mathsf{(Q_b', L_E', L_G')}$ has type $\mathsf{G^{\wedge}\,a\,Q_a'\,(C\,b\,e\,x)}$, i.e.,
\begin{gather*}
\mathsf{Q_a\;Q_a' : a \to Set} \\
\mathsf{Q_b\;Q_b' : b \to Set} \\
%\mathsf{M : PredMap\,b\,Q_a\,Q_a'} \\
\mathsf{L_E : Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e} \\
\mathsf{L_G\ : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,x} \\
\mathsf{L_E' : Equal^{\wedge}\,a\,(b \times b)\,Q_a'\,(Pair^{\wedge}\,b\,b\,Q'_b\,Q'_b)\,e} \\
\mathsf{L_G' : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b')\,x}
\end{gather*}
In other words, we have a proof $\mathsf{L_E}$ of the (extensional) equality
of the predicates $\mathsf{Q_a}$ and $\mathsf{Pair^{\wedge}\,b\,b\,Q_b\,Q_b}$
and a morphism of predicates $\mathsf{M}$ from $\mathsf{Q_a}$ to $\mathsf{Q_a'}$,
and we need to use those to deduce a proof of the (extensional) equality
of the predicates $\mathsf{Q_a'}$ and $\mathsf{Pair^{\wedge}\,b\,b\,Q_b'\,Q_b'}$.
But that is not generally possible:
the facts that $\mathsf{Q_a}$ is equal to $\mathsf{Pair^{\wedge}\,b\,b\,Q_b\,Q_b}$
and that there is a morphism of predicates $\mathsf{M}$ from $\mathsf{Q_a}$ to $\mathsf{Q_a'}$
do not guarantee that $\mathsf{Q_a'}$ is equal to $\mathsf{Pair^{\wedge}\,bb\,Q_b'\,Q_b'}$ for some $\mathsf{Q_b'}$.

At a deeper level, the fundamental issue at play here is that
the $\mathsf{Equal}$ type does not have functorial semantics,
so that having morphisms $\mathsf{A \to A'}$ and $\mathsf{B \to B'}$
and a proof that $\mathsf{A}$ is equal to $\mathsf{A'}$
does not provide a proof that $\mathsf{B}$ is equal to $\mathsf{B'}$.
This is because GADTs can either have a syntax-only semantics or a functorial-completion semantics.
Since we are interested in induction rules, we considered GADTs with their syntax-only semantics,
which is parametric but it is not functorial.
Had we considered the functorial-completion semantics, instead,
we would have forfeited parametricity~\cite{HaskellPaper}.
In both cases, thus, we cannot derive an induction rule for generic GADTs when they feature nesting.

\section{Applications}

Lambda normal form example.

%Still don't know which applications we might have

\section{Primitive representation for GADTs}

No induction with primitive representation (reference Haskell Symposium paper and~\cite{jp19} and paper Patricia Neil Clement 2010)

\section{Conclusion}

Mention Patricia/Neil2008 paper

\section{TODO}

\begin{itemize}
\item reference
\end{itemize}

%%% Added to bibliography:

%@inproceedings{jp20,
%  author = {P. Johann and A. Polonsky},
%  title = {Deep Induction: Induction Rules for (Truly) Nested Types},
%  year = {2020},
%  booktitle = {Foundations of Software Science and Computation Structures},
%  pages = {339-358}
%}
%
%@inproceedings{jgj21,
%  author = {P. Johann and E. Ghiorzi and D. Jeffries},
%  title = {Parametricity for Primitive Nested Types},
%  year = {2021},
%  booktitle = {Foundations of Software Science and Computation Structures},
%  pages = {324-343}
%}
%
%@inproceedings{bm98,
% title={Nested datatypes},
% author={R. Bird, and L. Meertens},
% booktitle={Mathematics of Program Construction},
% year={1998}
% pages = {52–67}
%}

\bibliography{gadt-story-references}

\end{document}

### 2021-03-30 induction-paper/GADT-induction.tex
Section: General case

@@
 
-%Outline general heuristics for this process, generalizing process in deep induction paper.
-%\begin{itemize}
-%    \item Lifting for GADT (possibly a restricted class?)
-%    \item Deep induction rule
-%    \item Prove deep induction (define term)
-%\end{itemize}
-
-TODO:
-\begin{itemize}
-\item explain overline is vector
-\end{itemize}
-
 In this section we will generalize the approach taken in the previous examples
@@
 \mathsf{data\, G\,(\ol{a : Set}) : Set\,where}\\
-\mathsf{\;\;\;\;\;\;\;\;C_i\, :\, F_i\,G\,\ol{b} \to G (\ol{K_i\,\ol{b}})}
+\mathsf{\;\;\;\;\;\;\;\;C\, :\, F\,G\,\ol{b} \to G (\ol{K\,\ol{b}})}
 \end{array}
 \end{equation}
-with a finite number of constructors $\mathsf{C_i}$,
-and where $\mathsf{F_i}$ has type $\mathsf{(Set^{\lvert\ol{a}\rvert} \to Set) \to Set^{\lvert\ol{b}\rvert} \to Set}$
-and each $\mathsf{K_i}$ has type $\mathsf{Set^{\lvert\ol{b}\rvert} \to Set}$.
-Each constructor $\mathsf{C_i}$ can be rewritten as
-\[
-\mathsf{C_i : \exists (\ol{b : Set}) \to \ol{Equal\,a\,(K_i\,\ol{b})} \to F_i\,G\,\ol{b} \to G\,\ol{a}}
+which, for ease of notation,
+we assume to have only one constructor $\mathsf{C}$.
+Extending the argument to the case of a GADT with multiple constructors presents no difficulty.
+In the definition above, $\mathsf{F}$ has type $\mathsf{(Set^{\alpha} \to Set) \to Set^{\beta} \to Set}$
+and each $\mathsf{K}$ has type $\mathsf{Set^{\beta} \to Set}$.
+The overline notation denotes a finite list:
+%and the vertical bars notation denotes the length of a list:
+$\mathsf{G}$ depends on a finite list of types $\mathsf{\ol{a}}$ of length $\mathsf{\alpha}$,
+each $\mathsf{K}$ depends on a finite list of types $\mathsf{\ol{b}}$ of length $\mathsf{\beta}$,
+and notice that the length of the list $\mathsf{\ol{a}}$ matches that of the list $\mathsf{\ol{K}}$.
+With that in mind, the constructor $\mathsf{C}$ can be rewritten as
+\[
+\mathsf{C : \exists (\ol{b : Set}) \to \ol{Equal\,a\,(K\,\ol{b})} \to F\,G\,\ol{b} \to G\,\ol{a}}
 \]
@@
 
-We define the lifting of $\mathsf{G}$
+In order to state the induction rule for $\mathsf{G}$,
+we first need to define its associated predicate-lifting function 
 \[
@@
 \[
-\mathsf{G^{\wedge}\,\ol{a}\,\ol{Q_a}\,(C_i\,\ol{b}\,\ol{e_a}\,x)
+\mathsf{G^{\wedge}\,\ol{a}\,\ol{Q_a}\,(C\,\ol{b}\,\ol{e_a}\,x)
 = \exists (\ol{Q_b : b \to Set})
-\to \ol{Equal^{\wedge}\,a\,(K_i\,\ol{b})\,Q_a\,(K_i^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
-\times F_i^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,{\ol{Q_b}}\,x}
-\]
-where $\mathsf{Q_a : a \to Set}$, $\mathsf{e_a : Equal\,a\,(K_i\,\ol{b})}$ and $\mathsf{x : F_i\,G\,\ol{b}}$,
-and the liftings
+\to \ol{Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
+\times F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,{\ol{Q_b}}\,x}
+\]
+where $\mathsf{Q_a : a \to Set}$, $\mathsf{e_a : Equal\,a\,(K\,\ol{b})}$ and $\mathsf{x : F\,G\,\ol{b}}$.
+Notice that we need a type $\mathsf{Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}$
+for each type $\mathsf{a}$.
+We also assume to have liftings for $\mathsf{F}$,
 \begin{multline*}
-\mathsf{F_i^{\wedge} : \forall (G : Set^{\lvert\ol{a}\rvert} \to Set) (\ol{b : Set})
+\mathsf{F^{\wedge} : \forall (G : Set^{\alpha} \to Set) (\ol{b : Set})
 \to (\forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set)} \\
 \mathsf{\to (\ol{b \to Set})
-\to F_i\,G\,\ol{b} \to Set}
+\to F\,G\,\ol{b} \to Set}
 \end{multline*}
-and
-\[
-\mathsf{K_i^{\wedge} : \forall (\ol{b : Set}) \to (\ol{b \to Set}) \to K_i\,\ol{b} \to Set}
-\]
-are defined by induction on the structure of the type.
-
-Define
-\[
-\mathsf{PredMap : \forall (a : Set) \to (a \to Set) \to (a \to Set) \to Set}
-\]
-as $\mathsf{PredMap\,a\,Q_1\,Q_2 = \forall (x : a) \to Q_1\,x \to Q_2\,x}$
-
-%  -- Lifting also acts on morphisms between predicates with the same carrier
-%  {-# TERMINATING #-}
-%  Bush^-map : ∀ {l} {X : Set} {Q1 Q2 : X → Set l}
-%              → Pred-map {l} Q1 Q2 → Pred-map (Bush^ Q1) (Bush^ Q2)
-%  Bush^-map Q12 Bnil q = bigtt
-%  Bush^-map Q12 (Bcons x b) (y , bby) =
-%                Q12 x y , Bush^-map (Bush^-map Q12) b bby
-Define
-\[
-\mathsf{GLMap : \forall (a : Set) (Q_1\,Q_2 : a \to Set) \to PredMap\,a\,Q_1\,Q_2 \to PredMap\,(G\,a)\,(G^{\wedge}\,a\,Q_1)\,(G^{\wedge}\,a\,Q_2)}
-\]
-as
-\[
-\mathsf{
-GLMap\,\ol{a}\,\ol{Q_1}\,\ol{Q_2}\,\ol{M}\,(C_i\,\ol{b}\,\ol{e_a}\,x)(\ol{Q_b}, \ol{e'_a}, l)
-= \ol{M}
-}
-\]
-%\begin{gather*}
-%\end{gather*}
-
-Associate the type
-\begin{multline}
-\mathsf{PC_i = \forall (\ol{a : Set}) (\ol{b : Set}) (\ol{Q_a : a \to Set}) (\ol{Q_b : b \to Set}) (\ol{e_a : Equal\,a\,(K_i\,\ol{b})}) (x : F_i\,G\,\ol{b})} \\
-\mathsf{\to \ol{Equal^{\wedge}\,a\,(K_i\,\ol{b})\,Q_a\,(K_i^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
-	\to F_i^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x
-	\to P\,\ol{a}\,\ol{Q_a}\,(C_i\,\ol{b}\,\ol{e_a}\,x)}
-\end{multline}
-to each constructor $\mathsf{C_i}$.
+and for each  $\mathsf{K}$,
+\[
+\mathsf{K^{\wedge} : \forall (\ol{b : Set}) \to (\ol{b \to Set}) \to K\,\ol{b} \to Set}
+\]
+as we could not possibly define an induction rule otherwise.
+A way to concretely define the predicate-lifting function for a type is to proceed by induction on the structure of the type.
+% on the structure of $\mathsf{F}$ and $\mathsf{K}$ respectively.
+
+%Associate the type
+%\begin{multline}
+%\mathsf{PC = \forall (\ol{a : Set}) (\ol{b : Set}) (\ol{Q_a : a \to Set}) (\ol{Q_b : b \to Set}) (\ol{e_a : Equal\,a\,(K\,\ol{b})}) (x : F\,G\,\ol{b})} \\
+%\mathsf{\to \ol{Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
+%	\to F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x
+%	\to P\,\ol{a}\,\ol{Q_a}\,(C\,\ol{b}\,\ol{e_a}\,x)}
+%\end{multline}
+%to the constructor $\mathsf{C}$.
 
@@
 \mathsf{DIG : \forall (P : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set)} \\
-\mathsf{\to \bigtimes_i PC_i
-\to \forall (\ol{a : Set})(\ol{Q_a : a \to Set})(x : G\,\ol{a}) \to G^{\wedge}\,\ol{a}\,\ol{Q_a}\,x \to P\,\ol{a}\,\ol{Q_a}\,x}
+\mathsf{\to \Big(
+\forall (\ol{a : Set}) (\ol{b : Set}) (\ol{Q_a : a \to Set}) (\ol{Q_b : b \to Set}) (\ol{e_a : Equal\,a\,(K\,\ol{b})}) (x : F\,G\,\ol{b})} \\
+\mathsf{\to \ol{Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
+	\to F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x
+	\to P\,\ol{a}\,\ol{Q_a}\,(C\,\ol{b}\,\ol{e_a}\,x)
+\Big)} \\
+\mathsf{\to \forall (\ol{a : Set})(\ol{Q_a : a \to Set})(x : G\,\ol{a}) \to G^{\wedge}\,\ol{a}\,\ol{Q_a}\,x \to P\,\ol{a}\,\ol{Q_a}\,x}
 \end{multline}
@@
 \[
-\mathsf{DIG\,P\,\ol{pc_i}\,\ol{a}\,\ol{Q_a}\,(C_i\,\ol{b}\,\ol{e_a}\,x)\,(\ol{Q_b}, \ol{e'_a}, l)
-= pc_i\,\ol{a}\,\ol{b}\,\ol{Q_a}\,\ol{Q_b}\,\ol{e_a}\,x\,\ol{e'_a}\,\ol{p}}
+\mathsf{DIG\,P\,PC\,\ol{a}\,\ol{Q_a}\,(C\,\ol{b}\,\ol{e_a}\,x)\,(\ol{Q_b}, \ol{E_a}, L_F)
+= PC\,\ol{a}\,\ol{b}\,\ol{Q_a}\,\ol{Q_b}\,\ol{e_a}\,x\,\ol{E_a}\,(p\,x\,L_F)}
 \]
 where
-\[
-\mathsf{p = ??? : F_i^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}
-\]
-
-\begin{lemma}
-For every $\mathsf{F, G, \ol{b}, P, \ol{Q_b}, x}$, there exists a function
-\[
-\mathsf{p : F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x \to F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}
-\]
-\end{lemma}
-\begin{proof}
-By structural induction on $\mathsf{F}$:
+%$\mathsf{\ol{pc_i : PC_i}}$,
+%$\mathsf{\ol{Q_a : a \to Set}}$,
+we have $\mathsf{e_a : Equal\,a\,(K\,\ol{b})}$ for each $\mathsf{a}$,
+$\mathsf{x : F\,G\,\ol{b}}$,
+$\mathsf{Q_b : b \to Set}$ for each $\mathsf{b}$ ,
+$\mathsf{E_a : Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}$  for each $\mathsf{a}$,
+$\mathsf{L_F : F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,{\ol{Q_b}}\,x}$,
+and
+\[
+\mathsf{p : PredMap\,(F\,G\,\ol{b})\,(F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}) (F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})}
+\]
+is defined by structural induction on $\mathsf{F}$ as follows:
 \begin{itemize}
-\item Case $\mathsf{F = F_1 \times F_2}$.
-Thus, $\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x = F_1^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x \times F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}$.
-By inductive hypothesis, there exist functions
-$\mathsf{p_1 : F_1^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x_1 \to ??? \to F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x_1}$ and
-$\mathsf{p_2 : F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x_2 \to ??? \to F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x_2}$.
+\item Case $\mathsf{F = F_1 \times F_2}$ where $\mathsf{F_1}$ and $\mathsf{F_2}$ have the same type as $\mathsf{F}$.
+We have that
+\[
+\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}
+= Pair^{\wedge}(F_1\,G\,\ol{b})(F_2\,G\,\ol{b})(F_1^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}) (F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})}
+\]
+By inductive hypothesis, there exist morphisms of predicates
+\begin{gather*}
+\mathsf{p_1 : PredMap\,(F_1\,G\,\ol{b})\,((F_1)^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}) ((F_1)^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})} \\
+\mathsf{p_2 : PredMap\,(F_2\,G\,\ol{b})\,((F_2)^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}) ((F_2)^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})}
+\end{gather*}
+Thus, we define $\mathsf{p (x_1, x_2) (L_1, L_2) = (p_1\,x_1\,L_1, p_2\,x_2\,L_2)}$ 
+for 
+$\mathsf{x_1 : F_1\,G\,\ol{b}}$,
+$\mathsf{x_2 : F_2\,G\,\ol{b}}$,
+$\mathsf{L_1 : F_1^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x_1}$
+and $\mathsf{L_2 : F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x_2}$.
 %%%
-\item Case $\mathsf{F\,G\,\ol{b} = G(\ol{F_a\,\ol{b}})}$.
+\item Case $\mathsf{F = F_1 + F_2}$ where $\mathsf{F_1}$ and $\mathsf{F_2}$ have the same type as $\mathsf{F}$.
+Analogous to case $\mathsf{F = F_1 \times F_2}$.
+%%%
+\item Case $\mathsf{F = F_1 \to F_2}$
+where $\mathsf{F_1}$ and $\mathsf{F_2}$ have the same type as $\mathsf{F}$.
+%where $\mathsf{F_1}$ does not contain the recursive variable,
+%i.e., $\mathsf{F_a : Set^\beta \to Set}$,
+%and $\mathsf{F_2}$ has the same type as $\mathsf{F}$.
+We have that
+\begin{gather*}
+\mathsf{F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x = \forall (y : F_1\,G\,\ol{b}) \to F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,(x\,y)} \\
+%F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,x \to F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x} \\
+\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x = \forall (y : F_1\,G\,\ol{b}) \to F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,(x\,y)}
+%\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x = F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,x \to F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}
+\end{gather*}
+By inductive hypothesis, there exist a morphism of predicates
+\[
+\mathsf{p_2 : PredMap\,(F_2\,G\,\ol{b}) (F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}) (F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})}
+\]
+Thus, we define $\mathsf{p\,x\,L_F\,y = p_2 (x\,y) (L_F\,y)}$.
+{\color{blue} Revise this case!}
+%Thus, we define $\mathsf{p\,x\,L_F\,L_1 = p_2\,(L_F\,L_1)}$
+%for $\mathsf{L_1 : F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,x}$.
+%Notice that $\mathsf{F_1}$ does not contain the recursive variable.
+%That is a necessary restriction, as the proof relies on
+%$\mathsf{F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x}$
+%and $\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}$
+%having the same domain.
+%%%
+\item Case $\mathsf{F\,G\,\ol{b} = G(\ol{F_a\,\ol{b}})}$
+where none of the $\mathsf{F_a}$ contains the recursive variable,
+i.e., $\mathsf{F_a : Set^\beta \to Set}$
+for each $\mathsf{a}$.
 Thus, $\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b} = P (\ol{F_a\,\ol{b}}) (\ol{F_a^{\wedge}\,\ol{b}\,\ol{Q_b}})}$.
-So, $\mathsf{p : P (\ol{F_a\,\ol{b}}) (\ol{F_a^{\wedge}\,\ol{b}\,\ol{Q_b}}) x}$ is
-\[
-\mathsf{p = DIG\,P\,\ol{pc_i}\,(\ol{F_a\,\ol{b}})\,(\ol{F_a^{\wedge}\,\ol{b}\,\ol{Q_b}})\,x\,l}
-\]
-Note: No nesting!
+So, $\mathsf{p}$ is defined as
+\[
+\mathsf{p = DIG\,P\,PC\,(\ol{F_a\,\ol{b}})\,(\ol{F_a^{\wedge}\,\ol{b}\,\ol{Q_b}})}
+\]
+%for $\mathsf{L_F : F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x}$.
+Notice that the fact that the types $\mathsf{F_a}$ do not contain the recursive variable
+means that $\mathsf{G}$ does not feature nesting.
+As mentioned before, that is a fundamental assumption to make the argument work.
+Why the argument fails in presence of truly nested GADTs will be shown in the next section.
+%%%
+\item Case $\mathsf{F\,G\,\ol{b} = H\,\ol{b}}$
+where $\mathsf{H}$ is a type constructor not containing the recursive variable,
+i.e., $\mathsf{H : Set^\beta \to Set}$.
+In such case, $\mathsf{p : PredMap (H\,\ol{b}) (H^{\wedge}\,\ol{b}\,\ol{Q_b}) (H^{\wedge}\,\ol{b}\,\ol{Q_b})}$
+is just the identity morphism of predicates.
+%%%
+\item Case $\mathsf{F\,G\,\ol{b} = H\,(\ol{F_c\,G\,\ol{b}})}$
+where $\mathsf{H}$ is a type constructor not containing the recursive variable,
+i.e., $\mathsf{H : Set^\gamma \to Set}$,
+and $\mathsf{F_c}$ has the same type as $\mathsf{F}$, for every $\mathsf{c = 1 \dots \gamma}$.
+Moreover, we assume that $\mathsf{H}$ has an associated predicate-lifting function,
+\[
+\mathsf{H^{\wedge} : \forall (\ol{c : Set}) \to (\ol{c \to Set}) \to H\,\ol{c} \to Set}
+\]
+and that this predicate-lifting function has a map function
+\[
+\mathsf{HLMap : \forall (\ol{c : Set}) (\ol{Q_c\;Q_c' : c \to Set}) \to PredMap\,\ol{c}\,\ol{Q_c}\,\ol{Q_c'}
+\to PredMap\,(H\,\ol{c})\,(H^{\wedge}\,\ol{c}\,\ol{Q_c})\,(H^{\wedge}\,\ol{c}\,\ol{Q_c'})}
+\]
+That means that $\mathsf{H}$ cannot be a GADT, as those do not have functorial semantics~\cite{HaskellPaper},
+but it can be an ADT or even a nested type~\cite{jp19,jgj21}.
+Thus,
+\[
+\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b} = H^{\wedge} (\ol{F_c\,G\,\ol{b}}) (\ol{F_c^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}})}
+\]
+By induction hypothesis, there is a morphism of predicates
+\[
+\mathsf{p_c : PredMap\,(F_c\,G\,\ol{b}) (F_c^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}) (F_c^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})}
+\]
+for every $\mathsf{c = 1 \dots \gamma}$.
+So, $\mathsf{p}$ is defined as
+\[
+\mathsf{p
+= HLMap\,(\ol{F_c\,G\,\ol{b}}) (\ol{F_c^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}}) (\ol{F_c^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}})\,\ol{p_c}}
+\]
+
 \end{itemize}
-\end{proof}
+Notice that, by proceeding by structural induction as above,
+we have implicitly defined a grammar for GADTs.
+In particular, we allow $\mathsf{F}$ to feature sums, products, arrow types and (limited) type application.
+
+%\subsection{General case}
+%
+%%Outline general heuristics for this process, generalizing process in deep induction paper.
+%%\begin{itemize}
+%%    \item Lifting for GADT (possibly a restricted class?)
+%%    \item Deep induction rule
+%%    \item Prove deep induction (define term)
+%%\end{itemize}
+%
+%TODO:
+%\begin{itemize}
+%\item explain overline is vector
+%\end{itemize}
+%
+%In this section we will generalize the approach taken in the previous examples
+%and provide a general framework to derive induction rules for GADTs.
+%We restrict our scope to GADTs not featuring nesting,
+%in the sense that their constructors contain no nested calls of the recursive variable,
+%as truly nested types do.
+%The reason for this restriction will be discussed in Section~\ref{sec:GADT-nested}.
+%
+%Consider a generic GADT
+%\begin{equation}\label{eq:gadt}
+%\begin{array}{l}
+%\mathsf{data\, G\,(\ol{a : Set}) : Set\,where}\\
+%\mathsf{\;\;\;\;\;\;\;\;C_i\, :\, F_i\,G\,\ol{b} \to G (\ol{K_i\,\ol{b}})}
+%\end{array}
+%\end{equation}
+%with a finite number of constructors $\mathsf{C_i}$,
+%and where $\mathsf{F_i}$ has type $\mathsf{(Set^{\lvert\ol{a}\rvert} \to Set) \to Set^{\lvert\ol{b}\rvert} \to Set}$
+%and each $\mathsf{K_i}$ has type $\mathsf{Set^{\lvert\ol{b}\rvert} \to Set}$.
+%The overline notation denotes a finite list:
+%$\mathsf{G}$ depends on a finite list of types $\mathsf{a}$,
+%each $\mathsf{K_i}$ depends on a finite list of types $\mathsf{b}$,
+%and notice that the length of the list of $\mathsf{a}$ matches that of the list of $\mathsf{K_i}$.
+%With that in mind, each constructor $\mathsf{C_i}$ can be rewritten as
+%\[
+%\mathsf{C_i : \exists (\ol{b : Set}) \to \ol{Equal\,a\,(K_i\,\ol{b})} \to F_i\,G\,\ol{b} \to G\,\ol{a}}
+%\]
+%which is the form we shall use from now on.
+%
+%We define the lifting of $\mathsf{G}$
+%\[
+%\mathsf{G^{\wedge} : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set}
+%\]
+%as
+%\[
+%\mathsf{G^{\wedge}\,\ol{a}\,\ol{Q_a}\,(C_i\,\ol{b}\,\ol{e_a}\,x)
+%= \exists (\ol{Q_b : b \to Set})
+%\to \ol{Equal^{\wedge}\,a\,(K_i\,\ol{b})\,Q_a\,(K_i^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
+%\times F_i^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,{\ol{Q_b}}\,x}
+%\]
+%where $\mathsf{Q_a : a \to Set}$, $\mathsf{e_a : Equal\,a\,(K_i\,\ol{b})}$ and $\mathsf{x : F_i\,G\,\ol{b}}$,
+%and the liftings
+%\begin{multline*}
+%\mathsf{F_i^{\wedge} : \forall (G : Set^{\lvert\ol{a}\rvert} \to Set) (\ol{b : Set})
+%\to (\forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set)} \\
+%\mathsf{\to (\ol{b \to Set})
+%\to F_i\,G\,\ol{b} \to Set}
+%\end{multline*}
+%and
+%\[
+%\mathsf{K_i^{\wedge} : \forall (\ol{b : Set}) \to (\ol{b \to Set}) \to K_i\,\ol{b} \to Set}
+%\]
+%are defined by induction on the structure of the type.
+%
+%Associate the type
+%\begin{multline}
+%\mathsf{PC_i = \forall (\ol{a : Set}) (\ol{b : Set}) (\ol{Q_a : a \to Set}) (\ol{Q_b : b \to Set}) (\ol{e_a : Equal\,a\,(K_i\,\ol{b})}) (x : F_i\,G\,\ol{b})} \\
+%\mathsf{\to \ol{Equal^{\wedge}\,a\,(K_i\,\ol{b})\,Q_a\,(K_i^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
+%	\to F_i^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x
+%	\to P\,\ol{a}\,\ol{Q_a}\,(C_i\,\ol{b}\,\ol{e_a}\,x)}
+%\end{multline}
+%to each constructor $\mathsf{C_i}$.
+%
+%Define the induction rule for $\mathsf{G}$
+%\begin{multline}
+%\mathsf{DIG : \forall (P : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set)} \\
+%\mathsf{\to \bigtimes_i PC_i
+%\to \forall (\ol{a : Set})(\ol{Q_a : a \to Set})(x : G\,\ol{a}) \to G^{\wedge}\,\ol{a}\,\ol{Q_a}\,x \to P\,\ol{a}\,\ol{Q_a}\,x}
+%\end{multline}
+%as
+%\[
+%\mathsf{DIG\,P\,\ol{pc_i}\,\ol{a}\,\ol{Q_a}\,(C_i\,\ol{b}\,\ol{e_a}\,x)\,(\ol{Q_b}, \ol{E_a}, L_F)
+%= pc_i\,\ol{a}\,\ol{b}\,\ol{Q_a}\,\ol{Q_b}\,\ol{e_a}\,x\,\ol{E_a}\,(p\,L_F)}
+%\]
+%where $\mathsf{\ol{pc_i : PC_i}}$,
+%$\mathsf{\ol{Q_a : a \to Set}}$,
+%$\mathsf{\ol{e_a : Equal\,a\,(K_i\,\ol{b})}}$,
+%$\mathsf{x : F_i\,G\,\ol{b}}$,
+%$\mathsf{\ol{Q_b : b \to Set}}$,
+%$\mathsf{\ol{E_a : Equal^{\wedge}\,a\,(K_i\,\ol{b})\,Q_a\,(K_i^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}}$,
+%$\mathsf{L_F : F_i^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,{\ol{Q_b}}\,x}$,
+%and $\mathsf{p}$ is defined in the following lemma.
+%
+%\begin{lemma}
+%For every $\mathsf{F, G, \ol{b}, P, \ol{Q_b}, x}$, there exists a function
+%\[
+%\mathsf{p : F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x \to F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}
+%\]
+%\end{lemma}
+%\begin{proof}
+%By structural induction on $\mathsf{F}$:
+%\begin{itemize}
+%\item Case $\mathsf{F = F_1 \times F_2}$.
+%We have that $\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x = F_1^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x \times F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}$.
+%By inductive hypothesis, there exist functions
+%$\mathsf{p_1 : F_1^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x_1 \to F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x_1}$ and
+%$\mathsf{p_2 : F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x_2 \to F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x_2}$.
+%Thus, we define $\mathsf{p (L_1, L_2) = (p_1\,L_1) \times (p_2\,L_2)}$
+%%%%
+%\item Case $\mathsf{F = F_1 + F_2}$. Analogous to case $\mathsf{F = F_1 \times F_2}$.
+%%%%
+%\item Case $\mathsf{F = F_1 \to F_2}$ where $\mathsf{F_1}$ does not contain the recursive variable.
+%We have that
+%\begin{gather*}
+%\mathsf{F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x = F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,x \to F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x} \\
+%\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x = F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,x \to F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}
+%\end{gather*}
+%By inductive hypothesis, there exist a function
+%$\mathsf{p_2 : F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x_2 \to F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x_2}$.
+%Thus, we define $\mathsf{p\,L\,l = p_2\,(L\,l)}$ for
+%$\mathsf{L : F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}$
+%and $\mathsf{l : F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,x}$.
+%
+%{\color{blue} TODO: explain why $\mathsf{F_1}$ does not contain the recursive variable.}
+%%%%
+%\item Case $\mathsf{F\,G\,\ol{b} = G(\ol{F_a\,\ol{b}})}$.
+%Thus, $\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b} = P (\ol{F_a\,\ol{b}}) (\ol{F_a^{\wedge}\,\ol{b}\,\ol{Q_b}})}$.
+%So, $\mathsf{p}$ is defined as
+%\[
+%\mathsf{p\,L_F = DIG\,P\,\ol{pc_i}\,(\ol{F_a\,\ol{b}})\,(\ol{F_a^{\wedge}\,\ol{b}\,\ol{Q_b}})\,x\,L_F}
+%\]
+%Note: No nesting!
+%\end{itemize}
+%\end{proof}
 
@@
 \begin{itemize}
-\item reference
+\item reference (correctly) Haskell Symposium paper
 \end{itemize}

### 2021-03-30 induction-paper/GADT-induction.tex
Rearranged material in section 4

@@
 
-\subsection{General case}
-
-In this section we will generalize the approach taken in the previous examples
-and provide a general framework to derive induction rules for GADTs.
-We restrict our scope to GADTs not featuring nesting,
+\subsection{Induction for GADTs with nesting}\label{sec:GADT-nested}
+
+In the previous sections,
+we derive induction rules for examples of GADTs that do not feature nesting,
 in the sense that their constructors contain no nested calls of the recursive variable,
 as truly nested types do.
-The reason for this restriction will be discussed in Section~\ref{sec:GADT-nested}.
-
-Consider a generic GADT
+Since both nested types~\cite{jp20}
+and GADTs without nesting admit induction rules,
+it is just natural to expect that GADTs with nesting would as well.
+Surprisingly, that is not the case:
+indeed, induction rules for nested types rely on functorial semantics,
+but GADTs cannot admit both functorial and parametric semantics at the same time~\cite{HaskellPaper}.
+In this section we show how induction for GADTs featuring nesting goes wrong
+by analyzing the following concrete example.
+\begin{equation}
+\begin{array}{l}
+\mathsf{data\, G\,(a : Set) : Set\,where}\\
+\mathsf{\;\;\;\;\;\;C :\, G(G\,a) \to G(a \times a)}
+\end{array}
+\end{equation}
+The constructor $\mathsf{C}$ can be rewritten as
+\[
+\mathsf{C : \exists (b : Set) \to Equal\,a\,(b \times b) \to G(G\,b) \to G\,a}
+\]
+which is the form we shall use from now on.
+Its predicate-lifting function
+\[
+\mathsf{G^{\wedge} : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set}
+\]
+is defined as
+\[
+\mathsf{G^{\wedge}\,a\,Q_a\,(C\,b\,e\,x)
+= \exists (Q_b : b \to Set)
+\to Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e
+\times G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,x}
+\]
+where $\mathsf{Q_a : a \to Set}$, $\mathsf{e : Equal\,a\,(b \times b)}$ and $\mathsf{x : G(G\,b)}$.
+
+An induction rule for $\mathsf{G}$ is a term
+\begin{multline}
+\mathsf{DIG : \forall (P : \forall (a : Set) \to (a \to Set) \to G\,a \to Set)} \\
+\mathsf{\to \Big( \forall (a\;b : Set) (Q_a : a \to Set) (Q_b : b \to Set) (e : Equal\,a\,(b \times b)) (x : G\,(G\,b))} \\
+\mathsf{\to Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e
+	\to P\,(G\,b)\,(P\,b\,Q_b)\,x
+	\to P\,a\,Q_a\,(C\,b\,e\,x) \Big)}\\
+\mathsf{\to \forall (a : Set)(Q_a : a \to Set)(x : G\,a) \to G^{\wedge}\,a\,Q_a\,x \to P\,a\,Q_a\,x}
+\end{multline}
+Consistently with the examples in the previous sections, we define $\mathsf{DIG}$ as
+\[
+\mathsf{DIG\,P\,pc\,a\,Q_a\,(C\,b\,e\,x)\,(Q_b, L_E, L_G)
+= pc\,a\,b\,Q_a\,Q_b\,e\,x\,L_E\,p}
+\]
+where $\mathsf{C\,b\,e\,x : G\,a}$,
+%, the second to last term applied to $\mathsf{DIG}$,
+%has type $\mathsf{G\,a}$,
+i.e., $\mathsf{e : Equal\,a\,(b \times b)}$ and $\mathsf{x : G(G\,b)}$,
+and $\mathsf{(Q_b, L_E, L_G) : G^{\wedge}\,a\,Q_a\,(C\,b\,e\,x)}$,
+%, the last term applied to $\mathsf{DIG}$,
+%has type $\mathsf{G^{\wedge}\,a\,Q_a\,(C\,b\,e\,x)}$,
+i.e., $\mathsf{Q_b : b \to Set}$, $\mathsf{L_E : Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e}$,
+and $\mathsf{L_G : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,x}$.
+%\begin{gather*}
+%\mathsf{Q_b : b \to Set} \\
+%\mathsf{L_E : Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e} \\
+%\mathsf{L_G : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,x}
+%%\mathsf{p = DIG\,P\,pc\,(G\,b)\,(P\,b\,Q_b)\,x\,q : P\,(G\,b)\,(P\,b\,Q_b)\,x}
+%\end{gather*}
+We still need to define $\mathsf{p : P\,(G\,b)\,(P\,b\,Q_b)\,x}$.
+We do so by using the induction rule and letting
+\[
+\mathsf{p = DIG\,P\,pc\,(G\,b)\,(P\,b\,Q_b)\,x\,q}
+\]
+where we still need to provide $\mathsf{q : G^{\wedge}\,(G\,b)\,(P\,b\,Q_b)\,x}$.
+To produce such $\mathsf{q}$, we need the map function of $\mathsf{G^{\wedge}}$, with type
+\[
+\mathsf{GLMap : \forall (a : Set) (Q_a\;Q_a' : a \to Set) \to PredMap\,a\,Q_a\,Q_a' \to PredMap\,(G\,a)\,(G^{\wedge}\,a\,Q_a)\,(G^{\wedge}\,a\,Q_a')}
+\]
+where $\mathsf{PredMap : \forall (a : Set) \to (a \to Set) \to (a \to Set) \to Set}$
+is defined as
+\[
+\mathsf{PredMap\,a\,Q_a\,Q_a' = \forall (x : a) \to Q_a\,x \to Q_a'\,x}
+\]
+and represents the type of morphisms between predicates.
+If we had $\mathsf{GLMap}$, then we would be able to define
+\[
+\mathsf{q = GLMap\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,(P\,b\,Q_b)\,(DIG\,P\,pc\,b\,Q_b)\,x\,L_G}
+\]
+Unfortunately, we cannot define such a $\mathsf{GLMap}$.
+Indeed, its definition would have to be
+\[
+\mathsf{GLMap\,a\,Q_a\,Q_a'\,M\,(C\,b\,e\,x)\,(Q_b, L_E, L_G) = (Q_b', L_E', L_G')}
+\]
+where $\mathsf{(Q_b, L_E, L_G)}$ has type $\mathsf{G^{\wedge}\,a\,Q_a\,(C\,b\,e\,x)}$
+and $\mathsf{(Q_b', L_E', L_G')}$ has type $\mathsf{G^{\wedge}\,a\,Q_a'\,(C\,b\,e\,x)}$, i.e.,
+\begin{gather*}
+\mathsf{Q_a\;Q_a' : a \to Set} \\
+\mathsf{Q_b\;Q_b' : b \to Set} \\
+%\mathsf{M : PredMap\,b\,Q_a\,Q_a'} \\
+\mathsf{L_E : Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e} \\
+\mathsf{L_G\ : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,x} \\
+\mathsf{L_E' : Equal^{\wedge}\,a\,(b \times b)\,Q_a'\,(Pair^{\wedge}\,b\,b\,Q'_b\,Q'_b)\,e} \\
+\mathsf{L_G' : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b')\,x}
+\end{gather*}
+In other words, we have a proof $\mathsf{L_E}$ of the (extensional) equality
+of the predicates $\mathsf{Q_a}$ and $\mathsf{Pair^{\wedge}\,b\,b\,Q_b\,Q_b}$
+and a morphism of predicates $\mathsf{M}$ from $\mathsf{Q_a}$ to $\mathsf{Q_a'}$,
+and we need to use those to deduce a proof of the (extensional) equality
+of the predicates $\mathsf{Q_a'}$ and $\mathsf{Pair^{\wedge}\,b\,b\,Q_b'\,Q_b'}$.
+But that is not generally possible:
+the facts that $\mathsf{Q_a}$ is equal to $\mathsf{Pair^{\wedge}\,b\,b\,Q_b\,Q_b}$
+and that there is a morphism of predicates $\mathsf{M}$ from $\mathsf{Q_a}$ to $\mathsf{Q_a'}$
+do not guarantee that $\mathsf{Q_a'}$ is equal to $\mathsf{Pair^{\wedge}\,bb\,Q_b'\,Q_b'}$ for some $\mathsf{Q_b'}$.
+
+At a deeper level, the fundamental issue at play here is that
+the $\mathsf{Equal}$ type does not have functorial semantics,
+so that having morphisms $\mathsf{A \to A'}$ and $\mathsf{B \to B'}$
+and a proof that $\mathsf{A}$ is equal to $\mathsf{A'}$
+does not provide a proof that $\mathsf{B}$ is equal to $\mathsf{B'}$.
+This is because GADTs can either have a syntax-only semantics or a functorial-completion semantics.
+Since we are interested in induction rules, we considered GADTs with their syntax-only semantics,
+which is parametric but it is not functorial.
+Had we considered the functorial-completion semantics, instead,
+we would have forfeited parametricity~\cite{HaskellPaper}.
+In both cases, thus, we cannot derive an induction rule for generic GADTs when they feature nesting.
+
+\subsection{General case}
+
+%We restrict our scope to GADTs not featuring nesting,
+%in the sense that their constructors contain no nested calls of the recursive variable,
+%as truly nested types do.
+%This way we can avoid the issue described in Section~\ref{sec:GADT-nested}.
+%The reason for this restriction will be discussed in Section~\ref{sec:GADT-nested}.
+
+In this section we generalize the approach taken in the previous examples
+and provide a general framework to derive induction rules for GADTs.
+Since, as seen in Section~\ref{sec:GADT-nested},
+GADTs featuring nesting do not admit an induction rule,
+we restrict our scope to GADTs not featuring nesting.
+Consider such a generic GADT
 \begin{equation}\label{eq:gadt}
@@
 we assume to have only one constructor $\mathsf{C}$.
-Extending the argument to the case of a GADT with multiple constructors presents no difficulty.
+Extending the argument to GADTs with multiple constructors presents no difficulty.
 In the definition above, $\mathsf{F}$ has type $\mathsf{(Set^{\alpha} \to Set) \to Set^{\beta} \to Set}$
@@
 $\mathsf{G}$ depends on a finite list of types $\mathsf{\ol{a}}$ of length $\mathsf{\alpha}$,
-each $\mathsf{K}$ depends on a finite list of types $\mathsf{\ol{b}}$ of length $\mathsf{\beta}$,
-and notice that the length of the list $\mathsf{\ol{a}}$ matches that of the list $\mathsf{\ol{K}}$.
+each of the $\mathsf{\alpha}$-many $\mathsf{K}$ depends on a finite list of types $\mathsf{\ol{b}}$ of length $\mathsf{\beta}$,
+and notice that the length of the list of types $\mathsf{\ol{a}}$ matches that of the list of type constructors $\mathsf{\ol{K}}$.
 With that in mind, the constructor $\mathsf{C}$ can be rewritten as
@@
 which is the form we shall use from now on.
+Notice that, in the above constructor, we have a type
+$\mathsf{Equal\,a\,(K\,\ol{b})}$ for each $\mathsf{a}$ and its corresponding type constructor $\mathsf{K}$.
 
 In order to state the induction rule for $\mathsf{G}$,
-we first need to define its associated predicate-lifting function 
+we first need to define $\mathsf{G}$'s associated predicate-lifting function 
 \[
@@
 \]
-where $\mathsf{Q_a : a \to Set}$, $\mathsf{e_a : Equal\,a\,(K\,\ol{b})}$ and $\mathsf{x : F\,G\,\ol{b}}$.
+where $\mathsf{Q_a : a \to Set}$
+and $\mathsf{C\,\ol{b}\,\ol{e_a}\,x : G\,\ol{a}}$,
+i.e., $\mathsf{e_a : Equal\,a\,(K\,\ol{b})}$ and $\mathsf{x : F\,G\,\ol{b}}$.
 Notice that we need a type $\mathsf{Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}$
-for each type $\mathsf{a}$.
+for each type $\mathsf{a}$ and its corresponding type constructor $\mathsf{K}$.
 We also assume to have liftings for $\mathsf{F}$,
@@
 as we could not possibly define an induction rule otherwise.
-A way to concretely define the predicate-lifting function for a type is to proceed by induction on the structure of the type.
-% on the structure of $\mathsf{F}$ and $\mathsf{K}$ respectively.
+A way to concretely define the predicate-lifting function for a type is to proceed by induction on the structure of the type,
+and we have seen in the previous sections examples of how to do so for products and type application.
+We do not give here the general definition of lifting,
+as that would require to first present a full type calculus,
+and that is beyond the scope of the paper.
 
@@
 
-Define the induction rule for $\mathsf{G}$
+We can now define the induction rule for $\mathsf{G}$
 \begin{multline}
@@
 \Big)} \\
-\mathsf{\to \forall (\ol{a : Set})(\ol{Q_a : a \to Set})(x : G\,\ol{a}) \to G^{\wedge}\,\ol{a}\,\ol{Q_a}\,x \to P\,\ol{a}\,\ol{Q_a}\,x}
+\mathsf{\to \forall (\ol{a : Set})(\ol{Q_a : a \to Set})(y : G\,\ol{a}) \to G^{\wedge}\,\ol{a}\,\ol{Q_a}\,y \to P\,\ol{a}\,\ol{Q_a}\,y}
 \end{multline}
@@
 where
-%$\mathsf{\ol{pc_i : PC_i}}$,
-%$\mathsf{\ol{Q_a : a \to Set}}$,
-we have $\mathsf{e_a : Equal\,a\,(K\,\ol{b})}$ for each $\mathsf{a}$,
-$\mathsf{x : F\,G\,\ol{b}}$,
-$\mathsf{Q_b : b \to Set}$ for each $\mathsf{b}$ ,
+\begin{itemize}
+\item $\mathsf{C\,\ol{b}\,\ol{e_a}\,x : G\,\ol{a}}$,
+i.e., $\mathsf{e_a : Equal\,a\,(K\,\ol{b})}$ for each $\mathsf{a}$
+and $\mathsf{x : F\,G\,\ol{b}}$;
+\item $\mathsf{(\ol{Q_b}, \ol{E_a}, L_F) : G^{\wedge}\,\ol{a}\,\ol{Q_a} (C\,\ol{b}\,\ol{e_a}\,x)}$,
+i.e.,
+$\mathsf{Q_b : b \to Set}$ for each $\mathsf{b}$,
 $\mathsf{E_a : Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}$  for each $\mathsf{a}$,
-$\mathsf{L_F : F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,{\ol{Q_b}}\,x}$,
-and
+and $\mathsf{L_F : F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,{\ol{Q_b}}\,x}$.
+\end{itemize}
+Finally, the morphism of predicates
 \[
@@
 \begin{gather*}
-\mathsf{F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x = \forall (y : F_1\,G\,\ol{b}) \to F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,(x\,y)} \\
+\mathsf{F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x = \forall (z : F_1\,G\,\ol{b}) \to F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,(x\,y)} \\
 %F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,x \to F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x} \\
-\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x = \forall (y : F_1\,G\,\ol{b}) \to F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,(x\,y)}
+\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x = \forall (z : F_1\,G\,\ol{b}) \to F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,(x\,y)}
 %\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x = F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,x \to F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}
 \end{gather*}
+where $\mathsf{x : F\,G\,\ol{b} = F_1\,G\,\ol{b} \to F_2\,G\,\ol{b}}$.
 By inductive hypothesis, there exist a morphism of predicates
@@
 \]
-Thus, we define $\mathsf{p\,x\,L_F\,y = p_2 (x\,y) (L_F\,y)}$.
+Thus, we define $\mathsf{p\,x\,L_F\,z = p_2 (x\,z) (L_F\,z)}$.
 {\color{blue} Revise this case!}
@@
 %for $\mathsf{L_F : F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x}$.
-Notice that the fact that the types $\mathsf{F_a}$ do not contain the recursive variable
+Note that the fact that the type constructors $\mathsf{F_a}$ do not contain the recursive variable
 means that $\mathsf{G}$ does not feature nesting.
-As mentioned before, that is a fundamental assumption to make the argument work.
-Why the argument fails in presence of truly nested GADTs will be shown in the next section.
+As explained and exemplified in Section~\ref{sec:GADT-nested},
+that assumption is required to make induction work.
 %%%
@@
 %%%
-\item Case $\mathsf{F\,G\,\ol{b} = H\,(\ol{F_c\,G\,\ol{b}})}$
+\item Case $\mathsf{F\,G\,\ol{b} = H (\ol{F_c\,G\,\ol{b}})}$
 where $\mathsf{H}$ is a type constructor not containing the recursive variable,
@@
 \]
-That means that $\mathsf{H}$ cannot be a GADT, as those do not have functorial semantics~\cite{HaskellPaper},
+That means that $\mathsf{H}$ cannot be a GADT,
+as GADTs do not have functorial semantics~\cite{HaskellPaper}
+and we would incur again in the same issue encountered in Section~\ref{sec:GADT-nested},
 but it can be an ADT or even a nested type~\cite{jp19,jgj21}.
@@
 In particular, we allow $\mathsf{F}$ to feature sums, products, arrow types and (limited) type application.
-
-%\subsection{General case}
-%
-%%Outline general heuristics for this process, generalizing process in deep induction paper.
-%%\begin{itemize}
-%%    \item Lifting for GADT (possibly a restricted class?)
-%%    \item Deep induction rule
-%%    \item Prove deep induction (define term)
-%%\end{itemize}
-%
-%TODO:
-%\begin{itemize}
-%\item explain overline is vector
-%\end{itemize}
-%
-%In this section we will generalize the approach taken in the previous examples
-%and provide a general framework to derive induction rules for GADTs.
-%We restrict our scope to GADTs not featuring nesting,
-%in the sense that their constructors contain no nested calls of the recursive variable,
-%as truly nested types do.
-%The reason for this restriction will be discussed in Section~\ref{sec:GADT-nested}.
-%
-%Consider a generic GADT
-%\begin{equation}\label{eq:gadt}
-%\begin{array}{l}
-%\mathsf{data\, G\,(\ol{a : Set}) : Set\,where}\\
-%\mathsf{\;\;\;\;\;\;\;\;C_i\, :\, F_i\,G\,\ol{b} \to G (\ol{K_i\,\ol{b}})}
-%\end{array}
-%\end{equation}
-%with a finite number of constructors $\mathsf{C_i}$,
-%and where $\mathsf{F_i}$ has type $\mathsf{(Set^{\lvert\ol{a}\rvert} \to Set) \to Set^{\lvert\ol{b}\rvert} \to Set}$
-%and each $\mathsf{K_i}$ has type $\mathsf{Set^{\lvert\ol{b}\rvert} \to Set}$.
-%The overline notation denotes a finite list:
-%$\mathsf{G}$ depends on a finite list of types $\mathsf{a}$,
-%each $\mathsf{K_i}$ depends on a finite list of types $\mathsf{b}$,
-%and notice that the length of the list of $\mathsf{a}$ matches that of the list of $\mathsf{K_i}$.
-%With that in mind, each constructor $\mathsf{C_i}$ can be rewritten as
-%\[
-%\mathsf{C_i : \exists (\ol{b : Set}) \to \ol{Equal\,a\,(K_i\,\ol{b})} \to F_i\,G\,\ol{b} \to G\,\ol{a}}
-%\]
-%which is the form we shall use from now on.
-%
-%We define the lifting of $\mathsf{G}$
-%\[
-%\mathsf{G^{\wedge} : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set}
-%\]
-%as
-%\[
-%\mathsf{G^{\wedge}\,\ol{a}\,\ol{Q_a}\,(C_i\,\ol{b}\,\ol{e_a}\,x)
-%= \exists (\ol{Q_b : b \to Set})
-%\to \ol{Equal^{\wedge}\,a\,(K_i\,\ol{b})\,Q_a\,(K_i^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
-%\times F_i^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,{\ol{Q_b}}\,x}
-%\]
-%where $\mathsf{Q_a : a \to Set}$, $\mathsf{e_a : Equal\,a\,(K_i\,\ol{b})}$ and $\mathsf{x : F_i\,G\,\ol{b}}$,
-%and the liftings
-%\begin{multline*}
-%\mathsf{F_i^{\wedge} : \forall (G : Set^{\lvert\ol{a}\rvert} \to Set) (\ol{b : Set})
-%\to (\forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set)} \\
-%\mathsf{\to (\ol{b \to Set})
-%\to F_i\,G\,\ol{b} \to Set}
-%\end{multline*}
-%and
-%\[
-%\mathsf{K_i^{\wedge} : \forall (\ol{b : Set}) \to (\ol{b \to Set}) \to K_i\,\ol{b} \to Set}
-%\]
-%are defined by induction on the structure of the type.
-%
-%Associate the type
-%\begin{multline}
-%\mathsf{PC_i = \forall (\ol{a : Set}) (\ol{b : Set}) (\ol{Q_a : a \to Set}) (\ol{Q_b : b \to Set}) (\ol{e_a : Equal\,a\,(K_i\,\ol{b})}) (x : F_i\,G\,\ol{b})} \\
-%\mathsf{\to \ol{Equal^{\wedge}\,a\,(K_i\,\ol{b})\,Q_a\,(K_i^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
-%	\to F_i^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x
-%	\to P\,\ol{a}\,\ol{Q_a}\,(C_i\,\ol{b}\,\ol{e_a}\,x)}
-%\end{multline}
-%to each constructor $\mathsf{C_i}$.
-%
-%Define the induction rule for $\mathsf{G}$
-%\begin{multline}
-%\mathsf{DIG : \forall (P : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set)} \\
-%\mathsf{\to \bigtimes_i PC_i
-%\to \forall (\ol{a : Set})(\ol{Q_a : a \to Set})(x : G\,\ol{a}) \to G^{\wedge}\,\ol{a}\,\ol{Q_a}\,x \to P\,\ol{a}\,\ol{Q_a}\,x}
-%\end{multline}
-%as
-%\[
-%\mathsf{DIG\,P\,\ol{pc_i}\,\ol{a}\,\ol{Q_a}\,(C_i\,\ol{b}\,\ol{e_a}\,x)\,(\ol{Q_b}, \ol{E_a}, L_F)
-%= pc_i\,\ol{a}\,\ol{b}\,\ol{Q_a}\,\ol{Q_b}\,\ol{e_a}\,x\,\ol{E_a}\,(p\,L_F)}
-%\]
-%where $\mathsf{\ol{pc_i : PC_i}}$,
-%$\mathsf{\ol{Q_a : a \to Set}}$,
-%$\mathsf{\ol{e_a : Equal\,a\,(K_i\,\ol{b})}}$,
-%$\mathsf{x : F_i\,G\,\ol{b}}$,
-%$\mathsf{\ol{Q_b : b \to Set}}$,
-%$\mathsf{\ol{E_a : Equal^{\wedge}\,a\,(K_i\,\ol{b})\,Q_a\,(K_i^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}}$,
-%$\mathsf{L_F : F_i^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,{\ol{Q_b}}\,x}$,
-%and $\mathsf{p}$ is defined in the following lemma.
-%
-%\begin{lemma}
-%For every $\mathsf{F, G, \ol{b}, P, \ol{Q_b}, x}$, there exists a function
-%\[
-%\mathsf{p : F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x \to F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}
-%\]
-%\end{lemma}
-%\begin{proof}
-%By structural induction on $\mathsf{F}$:
-%\begin{itemize}
-%\item Case $\mathsf{F = F_1 \times F_2}$.
-%We have that $\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x = F_1^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x \times F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}$.
-%By inductive hypothesis, there exist functions
-%$\mathsf{p_1 : F_1^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x_1 \to F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x_1}$ and
-%$\mathsf{p_2 : F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x_2 \to F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x_2}$.
-%Thus, we define $\mathsf{p (L_1, L_2) = (p_1\,L_1) \times (p_2\,L_2)}$
-%%%%
-%\item Case $\mathsf{F = F_1 + F_2}$. Analogous to case $\mathsf{F = F_1 \times F_2}$.
-%%%%
-%\item Case $\mathsf{F = F_1 \to F_2}$ where $\mathsf{F_1}$ does not contain the recursive variable.
-%We have that
-%\begin{gather*}
-%\mathsf{F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x = F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,x \to F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x} \\
-%\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x = F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,x \to F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}
-%\end{gather*}
-%By inductive hypothesis, there exist a function
-%$\mathsf{p_2 : F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x_2 \to F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x_2}$.
-%Thus, we define $\mathsf{p\,L\,l = p_2\,(L\,l)}$ for
-%$\mathsf{L : F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}$
-%and $\mathsf{l : F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,x}$.
-%
-%{\color{blue} TODO: explain why $\mathsf{F_1}$ does not contain the recursive variable.}
-%%%%
-%\item Case $\mathsf{F\,G\,\ol{b} = G(\ol{F_a\,\ol{b}})}$.
-%Thus, $\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b} = P (\ol{F_a\,\ol{b}}) (\ol{F_a^{\wedge}\,\ol{b}\,\ol{Q_b}})}$.
-%So, $\mathsf{p}$ is defined as
-%\[
-%\mathsf{p\,L_F = DIG\,P\,\ol{pc_i}\,(\ol{F_a\,\ol{b}})\,(\ol{F_a^{\wedge}\,\ol{b}\,\ol{Q_b}})\,x\,L_F}
-%\]
-%Note: No nesting!
-%\end{itemize}
-%\end{proof}
-
-%Define $\mathsf{p_b}$ by structural induction on $F_i$.
-%
-%Case $\mathsf{F_i\,G\,\ol{b} = F'_i\,G\,\ol{b} \times F''_i\,G\,\ol{b}}$.
-%We have $\mathsf{F_i^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,(x', x'') = (F'_i)^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x' \times (F''_i)^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x''}$.
-%Thus define
-%\[
-%\mathsf{
-%p_b\,P\,PC_i\,\ol{a}\,\ol{Q_a}\,\ol{b}\,\ol{e_a}\,(x', x'')\,\ol{e'_a}\,(l', l'')
-%=
-%(
-%p'_b\,P\,PC_i\,\ol{a}\,\ol{Q_a}\,\ol{b}\,\ol{e_a}\,x'\,\ol{e'_a}\,l',
-%p''_b\,P\,PC_i\,\ol{a}\,\ol{Q_a}\,\ol{b}\,\ol{e_a}\,x''\,\ol{e'_a}\,l''
-%)
-%} 
-%\]
-%where $\mathsf{p'_b}$ is the function associated to $\mathsf{F'_i}$
-%and $\mathsf{p''_b}$ is the function associated to $\mathsf{F''_i}$.
-%
-%
-%Case $\mathsf{F_i\,G\,\ol{b} = H(\ol{F'_i\,G\,\ol{b}})}$ with $\mathsf{H}$ data type with constructors $\mathsf{D_i}$.
-%We have
-%\[
-%\mathsf{F_i^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x = H^{\wedge}\,(\ol{F'_i\,G\,\ol{b}}) (\ol{(F'_i)^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}}) x}
-%\]
-%Thus define
-%\[
-%\mathsf{
-%p_b\,P\,PC_i\,\ol{a}\,\ol{Q_a}\,\ol{b}\,\ol{e_a}\,x\,\ol{e'_a}\,l
-%=
-%} 
-%\]
-
-%\subsection{General case (with product)}
-%
-%Note: case with one constructor to make things simpler
-%
-%Consider a generic GADT
-%\begin{equation}\label{eq:gadt}
-%\begin{array}{l}
-%\mathsf{data\, G\,(\ol{a : Set}) : Set\,where}\\
-%\mathsf{\;\;\;\;\;\;\;\;C\, :\, F_1\,G\,\ol{b} \times \dots \times F_n\,G\,\ol{b} \to G (\ol{K\,\ol{b}})}
-%\end{array}
-%\end{equation}
-%where each $\mathsf{F_i}$ has type $\mathsf{(Set^{\lvert\ol{a}\rvert} \to Set) \to Set^{\lvert\ol{b}\rvert} \to Set}$
-%and $\mathsf{K}$ has type $\mathsf{Set^{\lvert\ol{b}\rvert} \to Set}$.
-%The constructor $\mathsf{C}$ can be rewritten as
-%\[
-%\mathsf{C : \exists (\ol{b : Set}) \to \ol{Equal\,a\,(K\,\ol{b})} \to F_1\,G\,\ol{b} \to F_n\,G\,\ol{b} \to G\,\ol{a}}
-%\]
-%which is the form we shall use from now on.
-%
-%We define the lifting of $\mathsf{G}$
-%\[
-%\mathsf{G^{\wedge} : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set}
-%\]
-%as
-%\begin{multline*}
-%\mathsf{G^{\wedge}\,\ol{a}\,\ol{Q_a}\,(C\,\ol{b}\,\ol{e_a}\,x_1 \dots x_n)} \\
-%\mathsf{= \exists (\ol{Q_b : b \to Set})
-%\to \ol{Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
-%\times F_1^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,{\ol{Q_b}}\,x_1
-%\dots
-%\times F_n^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,{\ol{Q_b}}\,x_n}
-%\end{multline*}
-%where $\mathsf{Q_a : a \to Set}$, $\mathsf{e_a : Equal\,a\,(K\,\ol{b})}$ and $\mathsf{x_i : F_i\,G\,\ol{b}}$,
-%and the liftings
-%\begin{multline*}
-%\mathsf{F_i^{\wedge} : \forall (G : Set^{\lvert\ol{a}\rvert} \to Set) (\ol{b : Set})
-%\to (\forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set)} \\
-%\mathsf{\to (\ol{b \to Set})
-%\to F_i\,G\,\ol{b} \to Set}
-%\end{multline*}
-%and
-%\[
-%\mathsf{K^{\wedge} : \forall (\ol{b : Set}) \to (\ol{b \to Set}) \to K\,\ol{b} \to Set}
-%\]
-%are defined by induction on the structure of the type.
-%
-%Define the induction rule for $\mathsf{G}$
-%\begin{multline}
-%\mathsf{DIG : \forall (P : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set)} \\
-%\mathsf{\to \bigtimes_i \big( \forall (\ol{a : Set}) (\ol{b : Set}) (\ol{Q_a : a \to Set}) (\ol{Q_b : b \to Set}) (\ol{e_a : Equal\,a\,(K\,\ol{b})}) (x_i : F_i\,G\,\ol{b})} \\
-%\mathsf{\to \ol{Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
-%	\to F_1^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x_1 \to \dots \to F_n^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x_n
-%	\to P\,\ol{a}\,\ol{Q_a}\,(C\,\ol{b}\,\ol{e_a}\,x_1 \dots x_n) \big)} \\
-%\mathsf{\to \forall (\ol{a : Set})(\ol{Q_a : a \to Set})(x : G\,\ol{a}) \to G^{\wedge}\,\ol{a}\,\ol{Q_a}\,x \to P\,\ol{a}\,\ol{Q_a}\,x}
-%\end{multline}
-%as
-%\[
-%\mathsf{DIG\,P\,PC\,\ol{a}\,\ol{Q_a}\,(C\,\ol{b}\,\ol{e_a}\,x_1 \dots x_n)\,(\ol{Q_b}, \ol{e'_a}, l_{F_1}, \dots, l_{F_n})
-%= PC\,\ol{a}\,\ol{b}\,\ol{Q_a}\,\ol{Q_b}\,\ol{e_a}\,x_1 \dots x_n\,\ol{e'_a}\,p_1 \dots p_n}
-%\]
-%where
-%\[
-%\mathsf{p_i = DIG\,P\,PC\,\ol{a}\,\ol{Q_a} ? : F_i^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x_i}
-%\]
-
-%Consider a generic GADT
-%\begin{equation}\label{eq:gadt}
-%\begin{array}{l}
-%\mathsf{data\, G\,(\ol{a : Set}) : Set\,where}\\
-%\mathsf{\;\;\;\;\;\;\;\;C_i\, :\, f_i[G/\phi,\ol{b/\beta}] \to G (\ol{k_i[\ol{b/\beta}]})}
-%\end{array}
-%\end{equation}
-%with a finite number of constructors $\mathsf{C_i}$.
-%Each constructor $\mathsf{C_i}$ is defined by type expressions $\mathsf{f_i}$,
-%containing a type constructor variable $\mathsf{\phi}$ of arity matching the number of type arguments of $G$,
-%and type variables $\mathsf{\ol{\beta}}$;
-%and $\mathsf{\ol{g_i}}$ containing the type variables $\mathsf{\ol{\beta}}$.
-%Each constructor $\mathsf{C_i}$ can be rewritten as
-%\[
-%\mathsf{C_i : \exists (\ol{b : Set}) \to \ol{Equal\,a\,(k_i[\ol{b/\beta}])} \to f_i[G/\phi,\ol{b/\beta}] \to G\,\ol{a}}
-%\]
-%which is the form we shall use from now on.
-%
-%We define the types
-%\begin{gather*}
-%	\mathsf{F_i : (Set^{\lvert\ol{a}\rvert} \to Set) \to Set^{\lvert\ol{b}\rvert} \to Set} \\
-%	\mathsf{F_i }
-%\end{gather*}
-%
-%
-%Define the lifting of $\mathsf{G}$
-%\[
-%\mathsf{G^{\wedge} : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set}
-%\]
-%as
-%\[
-%\mathsf{G^{\wedge}\,\ol{a}\,\ol{P_a}\,(C_i\,\ol{b}\,\ol{e_a}\,x)
-%= \exists (\ol{Q_b : b \to Set})
-%\to \ol{Equal^{\wedge}\,a\,k_i[\ol{b/\beta}]\,P_a\,(k_i^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
-%\times f_i[G/\phi]^{\wedge}\,\ol{b}\,{\ol{Q_b}}\,x}
-%\]
-%where $\mathsf{P_a : a \to Set}$, $\mathsf{e_a : Equal\,a\,(k_i\,\ol{b})}$ and $\mathsf{x : f_i\,\ol{b}}$,
-%and the liftings $\mathsf{f_i^{\wedge}}$ and $\mathsf{k_i^{\wedge}}$
-%are defined the same way by induction on the structure of the type.
-
-\section{Induction for GADTs with nesting}\label{sec:GADT-nested}
-
-In the previous sections,
-we derive induction rules for GADTs that do not feature nesting,
-in the sense that their constructors contain no nested calls of the recursive variable,
-as truly nested types do.
-Since both nested types~\cite{jp20}
-and GADTs without nesting admit induction rules,
-it is just natural to expect that GADTs with nesting would as well.
-Surprisingly, that is not the case:
-indeed, induction rules for nested types rely on functorial semantics,
-but GADTs cannot admit both functorial and parametric semantics at the same time~\cite{HaskellPaper}.
-In this section we will show how induction for GADTs featuring nesting goes wrong
-by analyzing a concrete example.
-
-Consider the type
-\begin{equation}
-\begin{array}{l}
-\mathsf{data\, G\,(a : Set) : Set\,where}\\
-\mathsf{\;\;\;\;\;\;C :\, G(G\,a) \to G(a \times a)}
-\end{array}
-\end{equation}
-whose constructor $\mathsf{C}$ can be rewritten as
-\[
-\mathsf{C : \exists (b : Set) \to Equal\,a\,(b \times b) \to G(G\,b) \to G\,a}
-\]
-which is the form we shall use from now on.
-Its predicate-lifting function
-\[
-\mathsf{G^{\wedge} : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set}
-\]
-is defined as
-\[
-\mathsf{G^{\wedge}\,a\,Q_a\,(C\,b\,e\,x)
-= \exists (Q_b : b \to Set)
-\to Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e
-\times G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,x}
-\]
-where $\mathsf{Q_a : a \to Set}$, $\mathsf{e : Equal\,a\,(b \times b)}$ and $\mathsf{x : G(G\,b)}$.
-
-We need to define an induction rule for $\mathsf{G}$.
-That has to be a term
-\begin{multline}
-\mathsf{DIG : \forall (P : \forall (a : Set) \to (a \to Set) \to G\,a \to Set)} \\
-\mathsf{\to \Big( \forall (a\;b : Set) (Q_a : a \to Set) (Q_b : b \to Set) (e : Equal\,a\,(b \times b)) (x : G\,(G\,b))} \\
-\mathsf{\to Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e
-	\to P\,(G\,b)\,(P\,b\,Q_b)\,x
-	\to P\,a\,Q_a\,(C\,b\,e\,x) \Big)}\\
-\mathsf{\to \forall (a : Set)(Q_a : a \to Set)(x : G\,a) \to G^{\wedge}\,a\,Q_a\,x \to P\,a\,Q_a\,x}
-\end{multline}
-Consistently with the examples in the previous sections, we define $\mathsf{DIG}$ as
-\[
-\mathsf{DIG\,P\,pc\,a\,Q_a\,(C\,b\,e\,x)\,(Q_b, L_E, L_G)
-= pc\,a\,b\,Q_a\,Q_b\,e\,x\,L_E\,p}
-\]
-where $\mathsf{(Q_b, L_E, L_G)}$, the last term applied to $\mathsf{DIG}$, has type $\mathsf{G^{\wedge}\,a\,Q_a\,(C\,b\,e\,x)}$,
-i.e., $\mathsf{Q_b : b \to Set}$, $\mathsf{L_E : Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e}$,
-and $\mathsf{L_G : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,x}$.
-%\begin{gather*}
-%\mathsf{Q_b : b \to Set} \\
-%\mathsf{L_E : Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e} \\
-%\mathsf{L_G : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,x}
-%%\mathsf{p = DIG\,P\,pc\,(G\,b)\,(P\,b\,Q_b)\,x\,q : P\,(G\,b)\,(P\,b\,Q_b)\,x}
-%\end{gather*}
-We still need to define $\mathsf{p : P\,(G\,b)\,(P\,b\,Q_b)\,x}$.
-We do so by using the induction rule, letting $\mathsf{p = DIG\,P\,pc\,(G\,b)\,(P\,b\,Q_b)\,x\,q}$
-where we still need to provide $\mathsf{q : G^{\wedge}\,(G\,b)\,(P\,b\,Q_b)\,x}$.
-To produce such $\mathsf{q}$, we need the map function of $\mathsf{G^{\wedge}}$, with type
-\[
-\mathsf{GLMap : \forall (a : Set) (Q_a\;Q_a' : a \to Set) \to PredMap\,a\,Q_a\,Q_a' \to PredMap\,(G\,a)\,(G^{\wedge}\,a\,Q_a)\,(G^{\wedge}\,a\,Q_a')}
-\]
-where $\mathsf{PredMap : \forall (a : Set) \to (a \to Set) \to (a \to Set) \to Set}$
-is defined as
-\[
-\mathsf{PredMap\,a\,Q_a\,Q_a' = \forall (x : a) \to Q_a\,x \to Q_a'\,x}
-\]
-and represents the type of morphisms between predicates.
-If we had $\mathsf{GLMap}$, then we would be able to define
-\[
-\mathsf{q = GLMap\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,(P\,b\,Q_b)\,(DIG\,P\,pc\,b\,Q_b)\,x\,L_G}
-\]
-Unfortunately, we cannot define such a $\mathsf{GLMap}$.
-Indeed, its definition would have to be
-\[
-\mathsf{GLMap\,a\,Q_a\,Q_a'\,M\,(C\,b\,e\,x)\,(Q_b, L_E, L_G) = (Q_b', L_E', L_G')}
-\]
-where $\mathsf{(Q_b, L_E, L_G)}$ has type $\mathsf{G^{\wedge}\,a\,Q_a\,(C\,b\,e\,x)}$
-and $\mathsf{(Q_b', L_E', L_G')}$ has type $\mathsf{G^{\wedge}\,a\,Q_a'\,(C\,b\,e\,x)}$, i.e.,
-\begin{gather*}
-\mathsf{Q_a\;Q_a' : a \to Set} \\
-\mathsf{Q_b\;Q_b' : b \to Set} \\
-%\mathsf{M : PredMap\,b\,Q_a\,Q_a'} \\
-\mathsf{L_E : Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e} \\
-\mathsf{L_G\ : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,x} \\
-\mathsf{L_E' : Equal^{\wedge}\,a\,(b \times b)\,Q_a'\,(Pair^{\wedge}\,b\,b\,Q'_b\,Q'_b)\,e} \\
-\mathsf{L_G' : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b')\,x}
-\end{gather*}
-In other words, we have a proof $\mathsf{L_E}$ of the (extensional) equality
-of the predicates $\mathsf{Q_a}$ and $\mathsf{Pair^{\wedge}\,b\,b\,Q_b\,Q_b}$
-and a morphism of predicates $\mathsf{M}$ from $\mathsf{Q_a}$ to $\mathsf{Q_a'}$,
-and we need to use those to deduce a proof of the (extensional) equality
-of the predicates $\mathsf{Q_a'}$ and $\mathsf{Pair^{\wedge}\,b\,b\,Q_b'\,Q_b'}$.
-But that is not generally possible:
-the facts that $\mathsf{Q_a}$ is equal to $\mathsf{Pair^{\wedge}\,b\,b\,Q_b\,Q_b}$
-and that there is a morphism of predicates $\mathsf{M}$ from $\mathsf{Q_a}$ to $\mathsf{Q_a'}$
-do not guarantee that $\mathsf{Q_a'}$ is equal to $\mathsf{Pair^{\wedge}\,bb\,Q_b'\,Q_b'}$ for some $\mathsf{Q_b'}$.
-
-At a deeper level, the fundamental issue at play here is that
-the $\mathsf{Equal}$ type does not have functorial semantics,
-so that having morphisms $\mathsf{A \to A'}$ and $\mathsf{B \to B'}$
-and a proof that $\mathsf{A}$ is equal to $\mathsf{A'}$
-does not provide a proof that $\mathsf{B}$ is equal to $\mathsf{B'}$.
-This is because GADTs can either have a syntax-only semantics or a functorial-completion semantics.
-Since we are interested in induction rules, we considered GADTs with their syntax-only semantics,
-which is parametric but it is not functorial.
-Had we considered the functorial-completion semantics, instead,
-we would have forfeited parametricity~\cite{HaskellPaper}.
-In both cases, thus, we cannot derive an induction rule for generic GADTs when they feature nesting.
 

### 2021-03-31 induction-paper/GADT-induction.tex
General revision of Section 4

@@
 
-Nevertheless, this paper shows how to extend deep induction to GADTs.
+Nevertheless, this paper shows how to extend deep induction to some GADTs.
 We will first demonstrate how to derive the deep induction rule for some example GADTs,
-and then provide a general principle that works for all GADTs.
+and then provide a general principle that works for GADTs not featuring nesting in their definition.
 
@@
 
-We first derive the induction rule for the $\mathsf{Equal}$ type.
-This will provide a simple case study that will help with the derivation of the induction rule of more complex GADTs.
-More importantly, since we define GADTs using the $\mathsf{Equal}$ type,
+As a first example, we derive the induction rule for the $\mathsf{Equal}$ type.
+This will provide a simple case study that will help with the derivation of the induction rule for more complex GADTs.
+Moreover, since we define GADTs using the $\mathsf{Equal}$ type,
 as for example in Equation~\ref{eq:eq-seq},
@@
 
-To define the induction rule for $\mathsf{Equal}$ we first need a lifting operation
+To define the induction rule for $\mathsf{Equal}$ we first need a predicate-lifting operation
 \[
@@
 \]
-The induction rule for $\mathsf{Equal}$ has type
-\begin{multline}
-\mathsf{DIEqual : \forall (P : \forall (a\,b : Set) \to (a \to Set) \to (b \to Set) \to Equal\,a\,b \to Set)} \\
-\mathsf{\to \big( \forall (c : Set) (Q : c \to Set) (Q' : c \to Set) \to Equal^{\wedge}\,c\,c\,Q\,Q'\,Refl \to P\,c\,c\,Q\,Q'\,Refl \big)} \\
-\mathsf{\to \big( \forall (a : Set) (b : Set) (Q_a : a \to Set) (Q_b : b \to Set) (e: Equal\,a\,b)} \\
-	\mathsf{\to Equal^{\wedge}\,a\,b\,Q_a\,Q_b\,e \to P\,a\,b\,Q_a\,Q_b\,e \big)}
-\end{multline}
-and it is defined as
-\[
-\mathsf{DIEqual\,P\,PRefl\,a\,a\,Q\,Q'\,Refl\,l = PRefl\,a\,Q\,Q'\,l}
+
+Let $\mathsf{CRefl}$ be the following type associated to the $\mathsf{Refl}$ constructor:
+\begin{multline*}
+\mathsf{\forall (P : \forall (a\,b : Set) \to (a \to Set) \to (b \to Set) \to Equal\,a\,b \to Set)} \\
+\mathsf{\to \forall (c : Set) (Q : c \to Set) (Q' : c \to Set) \to Equal^{\wedge}\,c\,c\,Q\,Q'\,Refl \to P\,c\,c\,Q\,Q'\,Refl}
+\end{multline*}
+
+The induction rule for $\mathsf{Equal}$ is the type
+\begin{multline*}
+\mathsf{\forall (P : \forall (a\,b : Set) \to (a \to Set) \to (b \to Set) \to Equal\,a\,b \to Set)} \\
+%\mathsf{\to \big( \forall (c : Set) (Q : c \to Set) (Q' : c \to Set) \to Equal^{\wedge}\,c\,c\,Q\,Q'\,Refl \to P\,c\,c\,Q\,Q'\,Refl \big)} \\
+\mathsf{\to CRefl\,P \to \forall (a : Set) (b : Set) (Q_a : a \to Set) (Q_b : b \to Set) (e: Equal\,a\,b)} \\
+	\mathsf{\to Equal^{\wedge}\,a\,b\,Q_a\,Q_b\,e \to P\,a\,b\,Q_a\,Q_b\,e}
+\end{multline*}
+We define a term $\mathsf{DIEqual}$ of such type as
+\[
+\mathsf{DIEqual\,P\,crefl\,a\,a\,Q_a\,Q_a'\,Refl\,L_E = crefl\,a\,Q_a\,Q_a'\,L_E}
 \]
 where
-$\mathsf{PRefl : \forall (c : Set) (Q : c \to Set) (Q' : c \to Set) \to Equal^{\wedge}\,c\,c\,Q\,Q'\,Refl \to P\,c\,c\,Q\,Q'\,Refl}$,
-$\mathsf{Q : a \to Set}$, $\mathsf{Q' : a \to Set}$
-and $\mathsf{l : Equal^{\wedge}\,a\,a\,Q\,Q'\,Refl}$.
+$\mathsf{crefl : CRefl\,P}$,
+%$\mathsf{PRefl : \forall (c : Set) (Q : c \to Set) (Q' : c \to Set) \to Equal^{\wedge}\,c\,c\,Q\,Q'\,Refl \to P\,c\,c\,Q\,Q'\,Refl}$
+$\mathsf{Q_a : a \to Set}$, $\mathsf{Q_a' : a \to Set}$
+and $\mathsf{L_E : Equal^{\wedge}\,a\,a\,Q_a\,Q_a'\,Refl}$.
+Having provided a well-defined term for it,
+we have shown that the induction rule for $\mathsf{Equal}$ holds.
 
@@
 
-Next we shall consider the $\mathsf{Seq}$ type defined in Equation~\ref{eq:eq_seq}.
-Again, the first step in deriving its induction rule consists in defining the lifting
+Next, we shall provide an induction rule for the $\mathsf{Seq}$ type defined in Equation~\ref{eq:eq_seq}.
+Again, the first step in deriving the induction rule for $\mathsf{Seq}$ consists in defining the predicate-lifting function over it, 
 \[
@@
 \]
-which is given by
-\[
-\mathsf{Seq^{\wedge}\,a\,Q\,(Const\,x) = Q\,x}
-\]
+which is given by pattern-matching as
+\[
+\mathsf{Seq^{\wedge}\,a\,Q_a\,(Const\,x) = Q_a\,x}
+\]
+where $\mathsf{Q_a : a \to Set}$ and $\mathsf{x : a}$,
 and
@@
 \end{multline*}
-We also need the lifting of the type $\mathsf{Pair = \forall (b : Set)(c : Set) \to b \times c}$, which is 
-\[
-\mathsf{Pair^{\wedge} : \forall (b : Set)(c : Set) \to (b \to Set) \to (c \to Set) \to b \times c \to Set}
-\]
-defined as
+where $\mathsf{e : Equal\,a\,(b \times c)}$,
+$\mathsf{s_b : Seq\,b}$ and $\mathsf{s_c : Seq\,c}$.
+We also need the lifting of predicates over the polymorphic type of pairs, $\mathsf{Pair = \forall (b : Set)(c : Set) \to b \times c}$, which is 
+\[
+\mathsf{Pair^{\wedge} : \forall (b : Set)(c : Set) \to (b \to Set) \to (c \to Set) \to Pair\,b\,c \to Set}
+\]
+and it is defined as
 \[
@@
 \]
-
-With these tools, we have an induction rule for $\mathsf{Seq}$
-\begin{multline}
-\mathsf{DISeq : \forall (P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set)} \\
-\mathsf{\to \big( \forall (a : Set) (Q_a : a \to Set) (x : a) \to Q_a\,x \to P\,a\,Q_a\,(Const\,x) \big)} \\
-\mathsf{\to \big( \forall (a : Set) (b : Set) (c : Set)
+where $\mathsf{Q_b : b \to Set}$, $\mathsf{Q_c : c \to Set}$, $\mathsf{y : b}$ and $\mathsf{z : c}$.
+
+Finally, let $\mathsf{CConst}$ be the type
+\begin{multline*}
+\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set)} \\
+\mathsf{\to \forall (a : Set) (Q_a : a \to Set) (x : a) \to Q_a\,x \to P\,a\,Q_a\,(Const\,x)}
+\end{multline*}
+associated to the $\mathsf{Const}$ constructor,
+and let $\mathsf{CSPair}$ be the type
+\begin{multline*}
+\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set)} \\
+\mathsf{\to \forall (a : Set) (b : Set) (c : Set)
 	(Q_a : a \to Set) (Q_b : b \to Set) (Q_c : c \to Set)} \\
@@
 	\mathsf{\to P\,b\,Q_b\,s_b \to P\,c\,Q_c\,s_c
-	\to P a Q_a ( SPair\,b\,c\,e\,s_b\,s_c ) \big)} \\
+	\to P a Q_a ( SPair\,b\,c\,e\,s_b\,s_c )}
+\end{multline*}
+associated to the $\mathsf{SPair}$ constructor,
+
+With these tools, we have an induction rule for $\mathsf{Seq}$,
+\begin{multline*}
+\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set)}
+\mathsf{\to CConst\,P \to CSPair\,P} \\
+%\mathsf{\to \big( \forall (a : Set) (Q_a : a \to Set) (x : a) \to Q_a\,x \to P\,a\,Q_a\,(Const\,x) \big)} \\
+%\mathsf{\to \big( \forall (a : Set) (b : Set) (c : Set)
+%	(Q_a : a \to Set) (Q_b : b \to Set) (Q_c : c \to Set)} \\
+%	\mathsf{(s_b : Seq\,b) (s_c : Seq\,c) (e : Equal\,a\,(b \times c))
+%	\to Equal^{\wedge} a\, (b \times c)\, Q_a\, (Pair^{\wedge}\,b\,c\,Q_b\,Q_c)\, e} \\
+%	\mathsf{\to P\,b\,Q_b\,s_b \to P\,c\,Q_c\,s_c
+%	\to P a Q_a ( SPair\,b\,c\,e\,s_b\,s_c ) \big)} \\
 \mathsf{\to \forall (a : Set)(Q_a : a \to Set)(s_a : Seq\,a) \to Seq^{\wedge}\,a\,Q_a\,s_a \to P\,a\,Q_a\,s_a}
-\end{multline}
-We define $\mathsf{DISeq\,P\,PConst\,PSPair\,a\,Q_a\,s_a\,l_a}$,
+\end{multline*}
+To validate the induction rule, we define a term $\mathsf{DISeq}$ for it.
+We have to define
+\[
+\mathsf{DISeq\,P\,cconst\,cspair\,a\,Q_a\,s_a\,L_a : P\,a\,Q_a\,s_a}
+\]
 where
 $\mathsf{P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set}$,
+$\mathsf{cconst : CConst\,P}$,
+$\mathsf{cspair : CSPair\,P}$,
+% are terms for the clauses
+%associated to the constructors $\mathsf{Const}$ and $\mathsf{SPair}$,
+%respectively, in the induction rule,
 $\mathsf{Q_a : a \to Set}$, $\mathsf{s_a : Seq\,a}$
-and $\mathsf{l_a : Seq^{\wedge}\,a\,Q\,s_a}$,
-by pattern-matching on $\mathsf{s_a}$.
+and $\mathsf{L_a : Seq^{\wedge}\,a\,Q\,s_a}$,
+and we proceed by pattern-matching on $\mathsf{s_a}$.
 Let $\mathsf{s_a = Const\,x}$ for $\mathsf{x : a}$, and define
 \[
-\mathsf{DISeq\,P\,PConst\,PSPair\,a\,Q_a\,(Const\,x)\,l_a = PConst\,a\,Q_a\,x\,l_a}
-\]
-Notice that $\mathsf{Seq^{\wedge}\,a\,Q_a\,(Const\,x) = Q_a\,x}$, and thus $\mathsf{l_a : Q_a\,x}$,
+\mathsf{DISeq\,P\,cconst\,cspair\,a\,Q_a\,(Const\,x)\,L_a = cconst\,a\,Q_a\,x\,L_a}
+\]
+Notice that $\mathsf{Seq^{\wedge}\,a\,Q_a\,(Const\,x) = Q_a\,x}$, and thus $\mathsf{L_a : Q_a\,x}$,
 making the right-hand-side in the above expression type-check.
@@
 \[
-\mathsf{DISeq\,P\,PConst\,PSPair\,a\,Q_a\,(SPair\,b\,c\,e\,s_b\,s_c)\,(Q_b, Q_c, l_e, l_b, l_c)
-	= PSPair\,a\,b\,c\,Q_a\,Q_b\,Q_c\,s_b\,s_c\,e\,l_e\,p_b\,p_c}
-\]
-where
-$\mathsf{p_b : P\,b\,Q_b\,s_b}$, $\mathsf{p_c : P\,c\,Q_c\,s_c}$,
-$\mathsf{l_e : Equal^{\wedge}\,a\, (b \times c)\, Q_a\, (Q_b \times Q_c) \, e}$,
-$\mathsf{l_b : Seq^{\wedge}\,b\,Q_b\,s_b}$,
-$\mathsf{l_c : Seq^{\wedge}\,c\,Q_c\,s_c}$
-(notice that $\mathsf{(Q_b, Q_c, l_e, l_b, l_c) : Seq^{\wedge}\,a\,Q\,(SPair\,b\,c\,e\,x\,y)}$,
-making the right-hand-side in the above expression type-check)
+\mathsf{DISeq\,P\,cconst\,cspair\,a\,Q_a\,(SPair\,b\,c\,e\,s_b\,s_c)\,(Q_b, Q_c, L_e, L_b, L_c)
+	= cspair\,a\,b\,c\,Q_a\,Q_b\,Q_c\,s_b\,s_c\,e\,L_e\,p_b\,p_c}
+\]
+where $\mathsf{(Q_b, Q_c, L_e, L_b, L_c) : Seq^{\wedge}\,a\,Q\,(SPair\,b\,c\,e\,x\,y)}$, i.e.,
+\begin{itemize}
+\item $\mathsf{Q_b : b \to Set}$ and $\mathsf{Q_c : c \to Set}$;
+%\item $\mathsf{p_b : P\,b\,Q_b\,s_b}$ and $\mathsf{p_c : P\,c\,Q_c\,s_c}$;
+\item $\mathsf{L_e : Equal^{\wedge}\,a\, (b \times c)\, Q_a\, (Q_b \times Q_c) \, e}$;
+\item $\mathsf{L_b : Seq^{\wedge}\,b\,Q_b\,s_b}$ and $\mathsf{L_c : Seq^{\wedge}\,c\,Q_c\,s_c}$;
+%(notice that $\mathsf{(Q_b, Q_c, L_e, L_b, L_c) : Seq^{\wedge}\,a\,Q\,(SPair\,b\,c\,e\,x\,y)}$,
+%making the right-hand-side in the above expression type-check)
+\end{itemize}
 and $\mathsf{p_b}$ and $\mathsf{p_c}$ are defined as follows:
 \begin{align*}
-\mathsf{p_b} &=\mathsf{DISeq\,P\,PConst\,PSPair\,b\,Q_b\,s_b\,l_b : P\,b\,Q_b\,s_b} \\
-\mathsf{p_c} &=\mathsf{DISeq\,P\,PConst\,PSPair\,c\,Q_c\,s_c\,l_c : P\,c\,Q_c\,s_c}
+\mathsf{p_b} &=\mathsf{DISeq\,P\,cconst\,cspair\,b\,Q_b\,s_b\,L_b : P\,b\,Q_b\,s_b} \\
+\mathsf{p_c} &=\mathsf{DISeq\,P\,cconst\,cspair\,c\,Q_c\,s_c\,L_c : P\,c\,Q_c\,s_c}
 \end{align*}
@@
 In this section we show how induction for GADTs featuring nesting goes wrong
-by analyzing the following concrete example.
+by analyzing the following concrete example of such a type.
 \begin{equation}
@@
 which is the form we shall use from now on.
-Its predicate-lifting function
+The predicate-lifting function of $\mathsf{G}$,
 \[
@@
 
-An induction rule for $\mathsf{G}$ is a term
-\begin{multline}
-\mathsf{DIG : \forall (P : \forall (a : Set) \to (a \to Set) \to G\,a \to Set)} \\
-\mathsf{\to \Big( \forall (a\;b : Set) (Q_a : a \to Set) (Q_b : b \to Set) (e : Equal\,a\,(b \times b)) (x : G\,(G\,b))} \\
+Finally, let $\mathsf{CC}$ be the type
+\begin{multline*}
+\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to G\,a \to Set)} \\
+\mathsf{\to \forall (a\;b : Set) (Q_a : a \to Set) (Q_b : b \to Set) (e : Equal\,a\,(b \times b)) (x : G\,(G\,b))} \\
 \mathsf{\to Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e
 	\to P\,(G\,b)\,(P\,b\,Q_b)\,x
-	\to P\,a\,Q_a\,(C\,b\,e\,x) \Big)}\\
-\mathsf{\to \forall (a : Set)(Q_a : a \to Set)(x : G\,a) \to G^{\wedge}\,a\,Q_a\,x \to P\,a\,Q_a\,x}
-\end{multline}
-Consistently with the examples in the previous sections, we define $\mathsf{DIG}$ as
-\[
-\mathsf{DIG\,P\,pc\,a\,Q_a\,(C\,b\,e\,x)\,(Q_b, L_E, L_G)
-= pc\,a\,b\,Q_a\,Q_b\,e\,x\,L_E\,p}
-\]
-where $\mathsf{C\,b\,e\,x : G\,a}$,
+	\to P\,a\,Q_a\,(C\,b\,e\,x)}
+\end{multline*}
+associated to the $\mathsf{C}$ constructor.
+
+The induction rule for $\mathsf{G}$ is
+\begin{multline*}
+\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to G\,a \to Set) \to CC\,P} \\
+%\mathsf{\to \Big( \forall (a\;b : Set) (Q_a : a \to Set) (Q_b : b \to Set) (e : Equal\,a\,(b \times b)) (x : G\,(G\,b))} \\
+%\mathsf{\to Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e
+%	\to P\,(G\,b)\,(P\,b\,Q_b)\,x
+%	\to P\,a\,Q_a\,(C\,b\,e\,x) \Big)}\\
+\mathsf{\to \forall (a : Set)(Q_a : a \to Set)(y : G\,a) \to G^{\wedge}\,a\,Q_a\,y \to P\,a\,Q_a\,y}
+\end{multline*}
+Consistently with the previous examples,
+we define a term validating the induction rule, $\mathsf{DIG}$, as
+\[
+\mathsf{DIG\,P\,cc\,a\,Q_a\,(C\,b\,e\,x)\,(Q_b, L_E, L_G)
+= cc\,a\,b\,Q_a\,Q_b\,e\,x\,L_E\,p}
+\]
+where $\mathsf{cc : CC\,P}$,
+$\mathsf{C\,b\,e\,x : G\,a}$,
 %, the second to last term applied to $\mathsf{DIG}$,
@@
 \[
-\mathsf{p = DIG\,P\,pc\,(G\,b)\,(P\,b\,Q_b)\,x\,q}
+\mathsf{p = DIG\,P\,cc\,(G\,b)\,(P\,b\,Q_b)\,x\,q}
 \]
 where we still need to provide $\mathsf{q : G^{\wedge}\,(G\,b)\,(P\,b\,Q_b)\,x}$.
-To produce such $\mathsf{q}$, we need the map function of $\mathsf{G^{\wedge}}$, with type
+To produce such $\mathsf{q}$, we need the map function of $\mathsf{G^{\wedge}}$,
 \[
@@
 \[
-\mathsf{q = GLMap\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,(P\,b\,Q_b)\,(DIG\,P\,pc\,b\,Q_b)\,x\,L_G}
+\mathsf{q = GLMap\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,(P\,b\,Q_b)\,(DIG\,P\,cc\,b\,Q_b)\,x\,L_G}
 \]
@@
 \]
-where $\mathsf{(Q_b, L_E, L_G)}$ has type $\mathsf{G^{\wedge}\,a\,Q_a\,(C\,b\,e\,x)}$
+where $\mathsf{Q_a : a \to Set}$, $\mathsf{Q_a' : a \to Set}$,
+$\mathsf{M : PredMap\,a\,Q_a\,Q_a'}$,
+$\mathsf{C\,b\,e\,x : G\,a}$, i.e.,
+\begin{itemize}
+\item $\mathsf{e : Equal\,a\,(b \times b)}$;
+\item $\mathsf{x : G\,(G\,b)}$;
+\end{itemize}
+$\mathsf{(Q_b, L_E, L_G)}$ has type $\mathsf{G^{\wedge}\,a\,Q_a\,(C\,b\,e\,x)}$, i.e.,
+\begin{itemize}
+\item $\mathsf{Q_b : b \to Set}$;
+\item $\mathsf{L_E : Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e}$;
+\item $\mathsf{L_G : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,x}$;
+\end{itemize}
 and $\mathsf{(Q_b', L_E', L_G')}$ has type $\mathsf{G^{\wedge}\,a\,Q_a'\,(C\,b\,e\,x)}$, i.e.,
-\begin{gather*}
-\mathsf{Q_a\;Q_a' : a \to Set} \\
-\mathsf{Q_b\;Q_b' : b \to Set} \\
-%\mathsf{M : PredMap\,b\,Q_a\,Q_a'} \\
-\mathsf{L_E : Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e} \\
-\mathsf{L_G\ : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,x} \\
-\mathsf{L_E' : Equal^{\wedge}\,a\,(b \times b)\,Q_a'\,(Pair^{\wedge}\,b\,b\,Q'_b\,Q'_b)\,e} \\
-\mathsf{L_G' : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b')\,x}
-\end{gather*}
+\begin{itemize}
+\item $\mathsf{Q_b' : b \to Set}$;
+\item $\mathsf{L_E' : Equal^{\wedge}\,a\,(b \times b)\,Q_a'\,(Pair^{\wedge}\,b\,b\,Q'_b\,Q'_b)\,e}$;
+\item $\mathsf{L_G' : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b')\,x}$;
+\end{itemize}
+%\begin{gather*}
+%\mathsf{Q_a\;Q_a' : a \to Set} \\
+%\mathsf{Q_b\;Q_b' : b \to Set} \\
+%%\mathsf{M : PredMap\,b\,Q_a\,Q_a'} \\
+%\mathsf{L_E : Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e} \\
+%\mathsf{L_G\ : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,x} \\
+%\mathsf{L_E' : Equal^{\wedge}\,a\,(b \times b)\,Q_a'\,(Pair^{\wedge}\,b\,b\,Q'_b\,Q'_b)\,e} \\
+%\mathsf{L_G' : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b')\,x}
+%\end{gather*}
 In other words, we have a proof $\mathsf{L_E}$ of the (extensional) equality
@@
 and we need to use those to deduce a proof of the (extensional) equality
-of the predicates $\mathsf{Q_a'}$ and $\mathsf{Pair^{\wedge}\,b\,b\,Q_b'\,Q_b'}$.
+of the predicates $\mathsf{Q_a'}$ and $\mathsf{Pair^{\wedge}\,b\,b\,Q_b'\,Q_b'}$,
+for some  for some predicate $\mathsf{Q_b'}$ on $\mathsf{b}$.
 But that is not generally possible:
@@
 
-At a deeper level, the fundamental issue at play here is that
+At a deeper level, the fundamental issue is that
 the $\mathsf{Equal}$ type does not have functorial semantics,
@@
 
-In this section we generalize the approach taken in the previous examples
+Finally, we generalize the approach taken in the previous examples
 and provide a general framework to derive induction rules for GADTs.
@@
 
-%Associate the type
-%\begin{multline}
-%\mathsf{PC = \forall (\ol{a : Set}) (\ol{b : Set}) (\ol{Q_a : a \to Set}) (\ol{Q_b : b \to Set}) (\ol{e_a : Equal\,a\,(K\,\ol{b})}) (x : F\,G\,\ol{b})} \\
+Finally, associate the type
+\begin{multline*}
+\mathsf{CC = \forall (P : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set)} \\
+\mathsf{\to \forall (\ol{a : Set}) (\ol{b : Set}) (\ol{Q_a : a \to Set}) (\ol{Q_b : b \to Set}) (\ol{e_a : Equal\,a\,(K\,\ol{b})}) (x : F\,G\,\ol{b})} \\
+\mathsf{\to \ol{Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
+	\to F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x
+	\to P\,\ol{a}\,\ol{Q_a}\,(C\,\ol{b}\,\ol{e_a}\,x)}
+\end{multline*}
+to the constructor $\mathsf{C}$.
+
+%We can now define the induction rule for $\mathsf{G}$
+The induction rule for $\mathsf{G}$ is
+\begin{multline*}
+\mathsf{\forall (P : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set) \to CC\,P} \\
+%\mathsf{\to \Big(
+%\forall (\ol{a : Set}) (\ol{b : Set}) (\ol{Q_a : a \to Set}) (\ol{Q_b : b \to Set}) (\ol{e_a : Equal\,a\,(K\,\ol{b})}) (x : F\,G\,\ol{b})} \\
 %\mathsf{\to \ol{Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
 %	\to F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x
-%	\to P\,\ol{a}\,\ol{Q_a}\,(C\,\ol{b}\,\ol{e_a}\,x)}
-%\end{multline}
-%to the constructor $\mathsf{C}$.
-
-We can now define the induction rule for $\mathsf{G}$
-\begin{multline}
-\mathsf{DIG : \forall (P : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set)} \\
-\mathsf{\to \Big(
-\forall (\ol{a : Set}) (\ol{b : Set}) (\ol{Q_a : a \to Set}) (\ol{Q_b : b \to Set}) (\ol{e_a : Equal\,a\,(K\,\ol{b})}) (x : F\,G\,\ol{b})} \\
-\mathsf{\to \ol{Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
-	\to F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x
-	\to P\,\ol{a}\,\ol{Q_a}\,(C\,\ol{b}\,\ol{e_a}\,x)
-\Big)} \\
+%	\to P\,\ol{a}\,\ol{Q_a}\,(C\,\ol{b}\,\ol{e_a}\,x)
+%\Big)} \\
 \mathsf{\to \forall (\ol{a : Set})(\ol{Q_a : a \to Set})(y : G\,\ol{a}) \to G^{\wedge}\,\ol{a}\,\ol{Q_a}\,y \to P\,\ol{a}\,\ol{Q_a}\,y}
-\end{multline}
-as
-\[
-\mathsf{DIG\,P\,PC\,\ol{a}\,\ol{Q_a}\,(C\,\ol{b}\,\ol{e_a}\,x)\,(\ol{Q_b}, \ol{E_a}, L_F)
-= PC\,\ol{a}\,\ol{b}\,\ol{Q_a}\,\ol{Q_b}\,\ol{e_a}\,x\,\ol{E_a}\,(p\,x\,L_F)}
+\end{multline*}
+As always, we now define a term $\mathsf{DIG}$ for the induction rule.
+Define
+\[
+\mathsf{DIG\,P\,cc\,\ol{a}\,\ol{Q_a}\,(C\,\ol{b}\,\ol{e_a}\,x)\,(\ol{Q_b}, \ol{E_a}, L_F)
+= cc\,\ol{a}\,\ol{b}\,\ol{Q_a}\,\ol{Q_b}\,\ol{e_a}\,x\,\ol{E_a}\,(p\,x\,L_F)}
 \]
@@
 \begin{itemize}
+\item $\mathsf{cc : CC\,P}$;
 \item $\mathsf{C\,\ol{b}\,\ol{e_a}\,x : G\,\ol{a}}$,
-i.e., $\mathsf{e_a : Equal\,a\,(K\,\ol{b})}$ for each $\mathsf{a}$
+i.e., $\mathsf{e_a : Equal\,a\,(K\,\ol{b})}$ for each $\mathsf{a}$,
 and $\mathsf{x : F\,G\,\ol{b}}$;
@@
 Thus, we define $\mathsf{p\,x\,L_F\,z = p_2 (x\,z) (L_F\,z)}$.
-{\color{blue} Revise this case!}
+%{\color{blue} Revise this case!}
 %Thus, we define $\mathsf{p\,x\,L_F\,L_1 = p_2\,(L_F\,L_1)}$
@@
 \[
-\mathsf{p = DIG\,P\,PC\,(\ol{F_a\,\ol{b}})\,(\ol{F_a^{\wedge}\,\ol{b}\,\ol{Q_b}})}
+\mathsf{p = DIG\,P\,cc\,(\ol{F_a\,\ol{b}})\,(\ol{F_a^{\wedge}\,\ol{b}\,\ol{Q_b}})}
 \]
@@
 \]
-and that this predicate-lifting function has a map function
-\[
-\mathsf{HLMap : \forall (\ol{c : Set}) (\ol{Q_c\;Q_c' : c \to Set}) \to PredMap\,\ol{c}\,\ol{Q_c}\,\ol{Q_c'}
+and that this predicate-lifting function has a map function $\mathsf{HLMap}$ of type
+\[
+\mathsf{\forall (\ol{c : Set}) (\ol{Q_c\;Q_c' : c \to Set}) \to PredMap\,\ol{c}\,\ol{Q_c}\,\ol{Q_c'}
 \to PredMap\,(H\,\ol{c})\,(H^{\wedge}\,\ol{c}\,\ol{Q_c})\,(H^{\wedge}\,\ol{c}\,\ol{Q_c'})}
@@
 Notice that, by proceeding by structural induction as above,
-we have implicitly defined a grammar for GADTs.
+we have implicitly defined a grammar for GADTs admitting an induction rule.
 In particular, we allow $\mathsf{F}$ to feature sums, products, arrow types and (limited) type application.

### 2021-04-02 induction-paper/GADT-induction.tex
Revised Sections 4 and 5 according to discussion with Patty and Daniel.

@@
 Structural induction principles cannot be extended to truly nested types, such as $\mathsf{Bush}$.
+{\color{red} THIS IS NOT TRUE!!!}
 %To induce over a truly nested type, we need to use a \emph{deep induction} principle~\cite{jp20}.
@@
 as they generalize nested types.
+{\color{red} THIS IS NOT TRUE!!!}
 Still, there is hope for GADTs to support a deep induction rule,
@@
 
-\subsection{Induction for GADTs with nesting}\label{sec:GADT-nested}
+\subsection{General case}
+
+%We restrict our scope to GADTs not featuring nesting,
+%in the sense that their constructors contain no nested calls of the recursive variable,
+%as truly nested types do.
+%This way we can avoid the issue described in Section~\ref{sec:GADT-nested}.
+%The reason for this restriction will be discussed in Section~\ref{sec:GADT-nested}.
+
+Finally, we generalize the approach taken in the previous examples
+and provide a general framework to derive induction rules for GADTs.
+%Since, as seen in Section~\ref{sec:GADT-nested},
+%GADTs featuring nesting do not admit an induction rule,
+%we restrict our scope to GADTs not featuring nesting.
+For that, we need to give a grammar for the types we will be considering.
+A generic GADT
+\begin{equation*}
+\begin{array}{l}
+\mathsf{data\, G\,(\ol{a : Set}) : Set\,where}\\
+\mathsf{\;\;\;\;\;\;\;\;C_i\, :\, F_i\,G\,\ol{b} \to G (\ol{K_i\,\ol{b}})}
+\end{array}
+\end{equation*}
+is defined by a finite number of constructors $\mathsf{C_i}$.
+In the definition above, $\mathsf{F_i}$ is a type constructor with signature $\mathsf{(Set^{\alpha} \to Set) \to Set^{\beta} \to Set}$
+and each $\mathsf{K_i}$ is a type constructor with signature $\mathsf{Set^{\beta} \to Set}$
+(i.e. a type constructor of arity $\mathsf{\beta}$).
+The overline notation denotes a finite list:
+$\mathsf{\ol{a}}$ is a list of types of length $\mathsf{\alpha}$,
+so that it can be applied to the type constructor $\mathsf{G}$ of arity $\mathsf{\alpha}$.
+%and the vertical bars notation denotes the length of a list:
+%$\mathsf{G}$ depends on a finite list of types $\mathsf{\ol{a}}$ of length $\mathsf{\alpha}$,
+Each of the $\mathsf{\alpha}$-many $\mathsf{K_i}$ is a type constructor of arity $\mathsf{\beta}$
+so that it can be applied to the list of types $\mathsf{\ol{b}}$ of length $\mathsf{\beta}$.
+%depends on a finite list of types $\mathsf{\ol{b}}$ of length $\mathsf{\beta}$,
+Moreover, notice that the arity of $\mathsf{G}$ matches the number of type constructors $\mathsf{\ol{K_i}}$.
+We allow each $\mathsf{F_i}$ to be inductively built in the following ways (and with the following restrictions):
+\begin{itemize}
+\item $\mathsf{F_i = F_i' \times F_i''}$ where $\mathsf{F_i'}$ and $\mathsf{F_i''}$ have the same signature as $\mathsf{F_i}$ and are built recursively from the same induction rules.
+\item $\mathsf{F_i = F_i' + F_i''}$ where $\mathsf{F_i'}$ and $\mathsf{F_i''}$ have the same signature as $\mathsf{F_i}$ and are built recursively from the same induction rules.
+\item $\mathsf{F_i = F_i' \to F_i''}$
+where $\mathsf{F_i'}$ does not contain the recursive variable,
+i.e., $\mathsf{F_i' : Set^\beta \to Set}$,
+and $\mathsf{F_i''}$ has the same signature as $\mathsf{F_i}$ and is built recursively from the same induction rules.
+\item $\mathsf{F_i\,G\,\ol{b} = G(\ol{F_a\,\ol{b}})}$
+where none of the $\mathsf{F_a}$ contains the recursive variable,
+i.e., $\mathsf{F_a : Set^\beta \to Set}$
+for each $\mathsf{a}$.
+Such restriction is necessary to prevent nesting,
+as that would break the induction rule as discussed in Section~\ref{sec:GADT-nested}.
+\item $\mathsf{F_i\,G\,\ol{b} = H\,\ol{b}}$
+where $\mathsf{H}$ is a type constructor not containing the recursive variable,
+i.e., $\mathsf{H : Set^\beta \to Set}$.
+Notice that this covers the case in which $\mathsf{F_i}$ is a closed type,
+so, in particular, the unit and empty types, $\mathsf{1}$ and $\mathsf{0}$.
+%The only requirement on $\mathsf{H}$ is that it needs to have a predicate-lifting function
+%(and any type made of constant types, sums, products, arrow types and other GADTs satisfies this condition).
+\item $\mathsf{F_i\,G\,\ol{b} = H (\ol{F_c\,G\,\ol{b}})}$
+where $\mathsf{H}$ is a $\mathsf{\gamma}$-ary type constructor not containing the recursive variable,
+i.e., $\mathsf{H : Set^\gamma \to Set}$,
+and $\mathsf{F_c}$ has the same signature as $\mathsf{F_i}$  and is built recursively from the same induction rules, for every $\mathsf{c = 1 \dots \gamma}$.
+Moreover, we require that $\mathsf{H}$ is not a GADT itself (but we allow it to be an ADT or even a nested type).
+\end{itemize}
+We can summarize the above inductive definition with the following grammar
+(but beware that the above restrictions and requirements still apply):
+\[
+\mathsf{F_i\,G\,\ol{b} :=
+F_i'\,G\,\ol{b} \times F_i''\,G\,\ol{b}
+\ \vert\ F_i'\,G\,\ol{b} + F_i''\,G\,\ol{b}
+\ \vert\ F_i'\,\ol{b} \to F_i''\,G\,\ol{b}
+\ \vert\ G(\ol{F_a\,\ol{b}})
+\ \vert\ H\,\ol{b}
+\ \vert\ H (\ol{F_c\,G\,\ol{b}})
+}
+\]
+%The only requirement on the $\mathsf{K_i}$ is that they need to have a predicate-lifting function.
+
+Consider such a generic GADT
+\begin{equation}\label{eq:gadt}
+\begin{array}{l}
+\mathsf{data\, G\,(a : Set) : Set\,where}\\
+\mathsf{\;\;\;\;\;\;\;\;C\, :\, F\,G\,\ol{b} \to G (\ol{K\,\ol{b}})}
+\end{array}
+\end{equation}
+which, for ease of notation,
+we assume to be a unary type constructor (i.e. it depends on a single type parameter $\mathsf{a}$)
+and to have only one constructor $\mathsf{C}$.
+Extending the argument to GADTs of arbitrary arity and with multiple constructors presents no difficulty.
+In the definition above, $\mathsf{F}$ has type $\mathsf{(Set \to Set) \to Set^{\beta} \to Set}$
+and each $\mathsf{K}$ has type $\mathsf{Set^{\beta} \to Set}$.
+%The overline notation denotes a finite list:
+%%and the vertical bars notation denotes the length of a list:
+%%$\mathsf{G}$ depends on a finite list of types $\mathsf{\ol{a}}$ of length $\mathsf{\alpha}$,
+%each of the $\mathsf{\alpha}$-many $\mathsf{K}$ depends on a finite list of types $\mathsf{\ol{b}}$ of length $\mathsf{\beta}$,
+%and notice that the length of the list of types $\mathsf{\ol{a}}$ matches that of the list of type constructors $\mathsf{\ol{K}}$.
+The constructor $\mathsf{C}$ can be rewritten as
+\[
+\mathsf{C : \exists (\ol{b : Set}) \to Equal\,a\,(K\,\ol{b}) \to F\,G\,\ol{b} \to G\,a}
+\]
+which is the form we shall use from now on.
+%Notice that, in the above constructor, we have a type
+%$\mathsf{Equal\,a\,(K\,\ol{b})}$ for each $\mathsf{a}$ and its corresponding type constructor $\mathsf{K}$.
+
+In order to state the induction rule for $\mathsf{G}$,
+we first need to define $\mathsf{G}$'s associated predicate-lifting function 
+\[
+\mathsf{G^{\wedge} : \forall (a : Set) \to (a \to Set) \to G\,a \to Set}
+\]
+as
+\[
+\mathsf{G^{\wedge}\,a\,Q_a\,(C\,\ol{b}\,e\,x)
+= \exists (\ol{Q_b : b \to Set})
+\to Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e
+\times F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,{\ol{Q_b}}\,x}
+\]
+where $\mathsf{Q_a : a \to Set}$
+and $\mathsf{C\,\ol{b}\,e\,x : G\,a}$,
+i.e., $\mathsf{e : Equal\,a\,(K\,\ol{b})}$ and $\mathsf{x : F\,G\,\ol{b}}$.
+%Notice that we need a type $\mathsf{Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}$
+%for each type $\mathsf{a}$ and its corresponding type constructor $\mathsf{K}$.
+We also assume to have liftings for $\mathsf{F}$,
+\begin{multline*}
+\mathsf{F^{\wedge} : \forall (G : Set^{\alpha} \to Set) (\ol{b : Set})
+\to (\forall (a : Set) \to (a \to Set) \to G\,a \to Set)} \\
+\mathsf{\to (\ol{b \to Set})
+\to F\,G\,\ol{b} \to Set}
+\end{multline*}
+and for  $\mathsf{K}$,
+\[
+\mathsf{K^{\wedge} : \forall (\ol{b : Set}) \to (\ol{b \to Set}) \to K\,\ol{b} \to Set}
+\]
+as we could not possibly define an induction rule otherwise.
+A way to concretely define the predicate-lifting function for a type is to proceed by induction on the structure of the type,
+and we have seen in the previous sections examples of how to do so for products and type application.
+We do not give here the general definition of lifting,
+as that would require to first present a full type calculus,
+and that is beyond the scope of the paper.
+
+Finally, associate the type
+\begin{multline*}
+\mathsf{CC = \forall (P : \forall (a : Set) \to (a \to Set) \to G\,a \to Set)} \\
+\mathsf{\to \forall (a : Set) (\ol{b : Set}) (Q_a : a \to Set) (\ol{Q_b : b \to Set}) (e : Equal\,a\,(K\,\ol{b})) (x : F\,G\,\ol{b})} \\
+\mathsf{\to Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e
+	\to F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x
+	\to P\,a\,Q_a\,(C\,\ol{b}\,e\,x)}
+\end{multline*}
+to the constructor $\mathsf{C}$.
+
+%We can now define the induction rule for $\mathsf{G}$
+The induction rule for $\mathsf{G}$ is
+\begin{multline*}
+\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to G\,a \to Set) \to CC\,P} \\
+%\mathsf{\to \Big(
+%\forall (\ol{a : Set}) (\ol{b : Set}) (\ol{Q_a : a \to Set}) (\ol{Q_b : b \to Set}) (\ol{e_a : Equal\,a\,(K\,\ol{b})}) (x : F\,G\,\ol{b})} \\
+%\mathsf{\to \ol{Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
+%	\to F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x
+%	\to P\,\ol{a}\,\ol{Q_a}\,(C\,\ol{b}\,\ol{e_a}\,x)
+%\Big)} \\
+\mathsf{\to \forall (a : Set)(Q_a : a \to Set)(y : G\,a) \to G^{\wedge}\,a\,Q_a\,y \to P\,a\,Q_a\,y}
+\end{multline*}
+As always, we now define a term $\mathsf{DIG}$ for the induction rule.
+Define
+\[
+\mathsf{DIG\,P\,cc\,a\,Q_a\,(C\,\ol{b}\,e\,x)\,(\ol{Q_b}, L_E, L_F)
+= cc\,a\,\ol{b}\,Q_a\,\ol{Q_b}\,e\,x\,L_E\,(p\,x\,L_F)}
+\]
+where
+\begin{itemize}
+\item $\mathsf{cc : CC\,P}$;
+\item $\mathsf{C\,\ol{b}\,e\,x : G\,a}$,
+i.e., $\mathsf{e : Equal\,a\,(K\,\ol{b})}$,
+and $\mathsf{x : F\,G\,\ol{b}}$;
+\item $\mathsf{(\ol{Q_b}, L_E, L_F) : G^{\wedge}\,a\,Q_a (C\,\ol{b}\,e\,x)}$,
+i.e.,
+$\mathsf{Q_b : b \to Set}$ for each $\mathsf{b}$,
+$\mathsf{L_E : Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e}$,
+and $\mathsf{L_F : F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,{\ol{Q_b}}\,x}$.
+\end{itemize}
+Finally, the morphism of predicates
+\[
+\mathsf{p : PredMap\,(F\,G\,\ol{b})\,(F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}) (F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})}
+\]
+is defined by structural induction on $\mathsf{F}$ as follows:
+\begin{itemize}
+\item Case $\mathsf{F = F_1 \times F_2}$ where $\mathsf{F_1}$ and $\mathsf{F_2}$ have the same signature as $\mathsf{F}$.
+We have that
+\[
+\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}
+= Pair^{\wedge}(F_1\,G\,\ol{b})(F_2\,G\,\ol{b})(F_1^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}) (F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})}
+\]
+By inductive hypothesis, there exist morphisms of predicates
+\begin{gather*}
+\mathsf{p_1 : PredMap\,(F_1\,G\,\ol{b})\,((F_1)^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}) ((F_1)^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})} \\
+\mathsf{p_2 : PredMap\,(F_2\,G\,\ol{b})\,((F_2)^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}) ((F_2)^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})}
+\end{gather*}
+Thus, we define $\mathsf{p (x_1, x_2) (L_1, L_2) = (p_1\,x_1\,L_1, p_2\,x_2\,L_2)}$ 
+for 
+$\mathsf{x_1 : F_1\,G\,\ol{b}}$,
+$\mathsf{x_2 : F_2\,G\,\ol{b}}$,
+$\mathsf{L_1 : F_1^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x_1}$
+and $\mathsf{L_2 : F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x_2}$.
+%%%
+\item Case $\mathsf{F = F_1 + F_2}$ where $\mathsf{F_1}$ and $\mathsf{F_2}$ have the same signature as $\mathsf{F}$.
+Analogous to case $\mathsf{F = F_1 \times F_2}$.
+%%%
+\item Case $\mathsf{F = F_1 \to F_2}$
+%where $\mathsf{F_1}$ and $\mathsf{F_2}$ have the same type as $\mathsf{F}$.
+where $\mathsf{F_1}$ does not contain the recursive variable,
+i.e., $\mathsf{F_1 : Set^\beta \to Set}$,
+and $\mathsf{F_2}$ has the same signature as $\mathsf{F}$.
+We have that
+\[
+\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x = \forall (z : F_1\,\ol{b}) \to F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,z \to F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,(x\,z)}
+\]
+%\begin{gather*}
+%\mathsf{F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x = \forall (z : F_1\,G\,\ol{b}) \to F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,(x\,y)} \\
+%%F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,x \to F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x} \\
+%\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x = \forall (z : F_1\,G\,\ol{b}) \to F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,(x\,y)}
+%%\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x = F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,x \to F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}
+%\end{gather*}
+where $\mathsf{x : F\,G\,\ol{b} = F_1\,\ol{b} \to F_2\,G\,\ol{b}}$.
+By inductive hypothesis, there exist a morphism of predicates
+\[
+\mathsf{p_2 : PredMap\,(F_2\,G\,\ol{b}) (F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}) (F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})}
+\]
+Thus, we define $\mathsf{p\,x\,L_F : F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}$
+for $\mathsf{L_F : F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x}$
+as $\mathsf{p\,x\,L_F\,z\,L_1 = p_2 (x\,z) (L_F\,z\,L_1)}$
+for $\mathsf{z : F_1\,\ol{b}}$
+and $\mathsf{L_1 : F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,z}$.
+%{\color{blue} Revise this case!}
+%Thus, we define $\mathsf{p\,x\,L_F\,L_1 = p_2\,(L_F\,L_1)}$
+%for $\mathsf{L_1 : F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,x}$.
+Notice that $\mathsf{F_1}$ not containing the recursive variable
+is a necessary restriction, as the proof relies on
+$\mathsf{F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x}$
+and $\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}$
+having the same domain.
+%%%
+\item Case $\mathsf{F\,G\,\ol{b} = G(F'\,\ol{b})}$
+where $\mathsf{F'}$ does not contain the recursive variable,
+i.e., $\mathsf{F' : Set^\beta \to Set}$.
+Thus, $\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b} = P (F'\,\ol{b}) (F'^{\wedge}\,\ol{b}\,\ol{Q_b})}$.
+So, $\mathsf{p}$ is defined as
+\[
+\mathsf{p = DIG\,P\,cc\,(F'\,\ol{b})\,(F'^{\wedge}\,\ol{b}\,\ol{Q_b})}
+\]
+%for $\mathsf{L_F : F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x}$.
+%Note that the fact that the type constructor $\mathsf{F'}$ does not contain the recursive variable
+%means that $\mathsf{G}$ does not feature nesting.
+%As explained and exemplified in Section~\ref{sec:GADT-nested},
+%that assumption is required to make induction work.
+%%%
+\item Case $\mathsf{F\,G\,\ol{b} = H\,\ol{b}}$
+where $\mathsf{H}$ is a $\mathsf{\beta}$-ary type constructor not containing the recursive variable,
+i.e., $\mathsf{H : Set^\beta \to Set}$.
+In such case, $\mathsf{p : PredMap (H\,\ol{b}) (H^{\wedge}\,\ol{b}\,\ol{Q_b}) (H^{\wedge}\,\ol{b}\,\ol{Q_b})}$
+is just the identity morphism of predicates.
+%%%
+\item Case $\mathsf{F\,G\,\ol{b} = H (\ol{F_c\,G\,\ol{b}})}$
+where $\mathsf{H}$ is a $\mathsf{\gamma}$-ary type constructor not containing the recursive variable,
+i.e., $\mathsf{H : Set^\gamma \to Set}$,
+and $\mathsf{F_c}$ has the same signature as $\mathsf{F}$, for every $\mathsf{c = 1 \dots \gamma}$.
+Moreover, we assume that $\mathsf{H}$ has an associated predicate-lifting function,
+\[
+\mathsf{H^{\wedge} : \forall (\ol{c : Set}) \to (\ol{c \to Set}) \to H\,\ol{c} \to Set}
+\]
+and that this predicate-lifting function has a map function $\mathsf{HLMap}$ of type
+\[
+\mathsf{\forall (\ol{c : Set}) (\ol{Q_c\;Q_c' : c \to Set}) \to PredMap\,\ol{c}\,\ol{Q_c}\,\ol{Q_c'}
+\to PredMap\,(H\,\ol{c})\,(H^{\wedge}\,\ol{c}\,\ol{Q_c})\,(H^{\wedge}\,\ol{c}\,\ol{Q_c'})}
+\]
+That means that $\mathsf{H}$ cannot be a GADT,
+as GADTs do not have functorial semantics~\cite{HaskellPaper}
+and we would incur again in the same issue encountered in Section~\ref{sec:GADT-nested},
+but it can be an ADT or even a nested type~\cite{jp19,jgj21}.
+Thus,
+\[
+\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b} = H^{\wedge} (\ol{F_c\,G\,\ol{b}}) (\ol{F_c^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}})}
+\]
+By induction hypothesis, there is a morphism of predicates
+\[
+\mathsf{p_c : PredMap\,(F_c\,G\,\ol{b}) (F_c^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}) (F_c^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})}
+\]
+for every $\mathsf{c = 1 \dots \gamma}$.
+So, $\mathsf{p}$ is defined as
+\[
+\mathsf{p
+= HLMap\,(\ol{F_c\,G\,\ol{b}}) (\ol{F_c^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}}) (\ol{F_c^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}})\,\ol{p_c}}
+\]
+\end{itemize}
+Notice that, by proceeding by structural induction as above,
+we have implicitly defined a grammar for GADTs admitting an induction rule.
+In particular, we allow $\mathsf{F}$ to feature sums, products, arrow types and (limited) type application.
+
+%\subsection{General case (Daniel's revision)}
+%
+%%We restrict our scope to GADTs not featuring nesting,
+%%in the sense that their constructors contain no nested calls of the recursive variable,
+%%as truly nested types do.
+%%This way we can avoid the issue described in Section~\ref{sec:GADT-nested}.
+%%The reason for this restriction will be discussed in Section~\ref{sec:GADT-nested}.
+%
+%Finally, we generalize the approach taken in the previous examples
+%and provide a general framework to derive induction rules for GADTs.
+%%Since, as seen in Section~\ref{sec:GADT-nested},
+%%GADTs featuring nesting do not admit an induction rule,
+%%we restrict our scope to GADTs not featuring nesting.
+%For that, we need to give a grammar for the types we will be considering.
+%A generic GADT
+%\begin{equation}\label{eq:gadt}
+%\begin{array}{l}
+%\mathsf{data\, G\,(\ol{a : Set}) : Set\,where}\\
+%\mathsf{\;\;\;\;\;\;\;\;C_i\, :\, F_i\,G\,\ol{b} \to G (\ol{K_i\,\ol{b}})}
+%\end{array}
+%\end{equation}
+%is defined by a finite number of constructors $\mathsf{C_i}$.
+%In the definition above, $\mathsf{F_i}$ has type $\mathsf{(Set^{\alpha} \to Set) \to Set^{\beta} \to Set}$
+%and each $\mathsf{K_i}$ has type $\mathsf{Set^{\beta} \to Set}$.
+%The overline notation denotes a finite list:
+%%and the vertical bars notation denotes the length of a list:
+%$\mathsf{G}$ depends on a finite list of types $\mathsf{\ol{a}}$ of length $\mathsf{\alpha}$,
+%each of the $\mathsf{\alpha}$-many $\mathsf{K_i}$ depends on a finite list of types $\mathsf{\ol{b}}$ of length $\mathsf{\beta}$,
+%and notice that the length of the list of types $\mathsf{\ol{a}}$ matches that of the list of type constructors $\mathsf{\ol{K_i}}$.
+%We allow each $\mathsf{F_i}$ to be inductively built in the following ways (and with the following restrictions):
+%\begin{itemize}
+%\item $\mathsf{F_i = F_i' \times F_i''}$ where $\mathsf{F_i'}$ and $\mathsf{F_i''}$ have the same type as $\mathsf{F_i}$.
+%\item $\mathsf{F_i = F_i' + F_i''}$ where $\mathsf{F_i'}$ and $\mathsf{F_i''}$ have the same type as $\mathsf{F_i}$.
+%\item $\mathsf{F_i = F_i' \to F_i''}$
+%where $\mathsf{F_i'}$ does not contain the recursive variable,
+%i.e., $\mathsf{F_i' : Set^\beta \to Set}$,
+%and $\mathsf{F_i''}$ has the same type as $\mathsf{F_i}$.
+%\item $\mathsf{F_i\,G\,\ol{b} = G(\ol{F_a\,\ol{b}})}$
+%where none of the $\mathsf{F_a}$ contains the recursive variable,
+%i.e., $\mathsf{F_a : Set^\beta \to Set}$
+%for each $\mathsf{a}$.
+%Such restriction is necessary to prevent nesting,
+%as that would break the induction rule as discussed in Section~\ref{sec:GADT-nested}.
+%\item $\mathsf{F_i\,G\,\ol{b} = H\,\ol{b}}$
+%where $\mathsf{H}$ is a type constructor not containing the recursive variable,
+%i.e., $\mathsf{H : Set^\beta \to Set}$.
+%Notice that this covers the case in which $\mathsf{F_i}$ is a closed type,
+%so, in particular, the unit and empty types, $\mathsf{1}$ and $\mathsf{0}$.
+%The only requirement on $\mathsf{H}$ is that it needs to have a predicate-lifting function
+%(and any type made of constant types, sums, products, arrow types and other GADTs satisfies this condition).
+%\item $\mathsf{F_i\,G\,\ol{b} = H (\ol{F_c\,G\,\ol{b}})}$
+%where $\mathsf{H}$ is a type constructor not containing the recursive variable,
+%i.e., $\mathsf{H : Set^\gamma \to Set}$,
+%and $\mathsf{F_c}$ has the same type as $\mathsf{F_i}$, for every $\mathsf{c = 1 \dots \gamma}$.
+%Moreover, we assume that $\mathsf{H}$ is not a GADT itself (but we allow it to be an ADT or even a nested type).
+%\end{itemize}
+%The only requirement on the $\mathsf{K_i}$ is that they need to have a predicate-lifting function.
+%
+%Consider such a generic GADT
+%\begin{equation}\label{eq:gadt}
+%\begin{array}{l}
+%\mathsf{data\, G\,(\ol{a : Set}) : Set\,where}\\
+%\mathsf{\;\;\;\;\;\;\;\;C\, :\, F\,G\,\ol{b} \to G (\ol{K\,\ol{b}})}
+%\end{array}
+%\end{equation}
+%which, for ease of notation,
+%we assume to have only one constructor $\mathsf{C}$.
+%Extending the argument to GADTs with multiple constructors presents no difficulty.
+%In the definition above, $\mathsf{F}$ has type $\mathsf{(Set^{\alpha} \to Set) \to Set^{\beta} \to Set}$
+%and each $\mathsf{K}$ has type $\mathsf{Set^{\beta} \to Set}$.
+%The overline notation denotes a finite list:
+%%and the vertical bars notation denotes the length of a list:
+%$\mathsf{G}$ depends on a finite list of types $\mathsf{\ol{a}}$ of length $\mathsf{\alpha}$,
+%each of the $\mathsf{\alpha}$-many $\mathsf{K}$ depends on a finite list of types $\mathsf{\ol{b}}$ of length $\mathsf{\beta}$,
+%and notice that the length of the list of types $\mathsf{\ol{a}}$ matches that of the list of type constructors $\mathsf{\ol{K}}$.
+%With that in mind, the constructor $\mathsf{C}$ can be rewritten as
+%\[
+%\mathsf{C : \exists (\ol{b : Set}) \to \ol{Equal\,a\,(K\,\ol{b})} \to F\,G\,\ol{b} \to G\,\ol{a}}
+%\]
+%which is the form we shall use from now on.
+%Notice that, in the above constructor, we have a type
+%$\mathsf{Equal\,a\,(K\,\ol{b})}$ for each $\mathsf{a}$ and its corresponding type constructor $\mathsf{K}$.
+%
+%In order to state the induction rule for $\mathsf{G}$,
+%we first need to define $\mathsf{G}$'s associated predicate-lifting function 
+%\[
+%\mathsf{G^{\wedge} : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set}
+%\]
+%as
+%\[
+%\mathsf{G^{\wedge}\,\ol{a}\,\ol{Q_a}\,(C\,\ol{b}\,\ol{e_a}\,x)
+%= \exists (\ol{Q_b : b \to Set})
+%\to F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,{\ol{Q_b}}\,x}
+%\]
+%where $\mathsf{Q_a : a \to Set}$
+%and $\mathsf{C\,\ol{b}\,\ol{e_a}\,x : G\,\ol{a}}$,
+%i.e., $\mathsf{e_a : Equal\,a\,(K\,\ol{b})}$ and $\mathsf{x : F\,G\,\ol{b}}$.
+%We also assume to have liftings for $\mathsf{F}$,
+%\begin{multline*}
+%\mathsf{F^{\wedge} : \forall (G : Set^{\alpha} \to Set) (\ol{b : Set})
+%\to (\forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set)} \\
+%\mathsf{\to (\ol{b \to Set})
+%\to F\,G\,\ol{b} \to Set}
+%\end{multline*}
+%and for each  $\mathsf{K}$,
+%\[
+%\mathsf{K^{\wedge} : \forall (\ol{b : Set}) \to (\ol{b \to Set}) \to K\,\ol{b} \to Set}
+%\]
+%as we could not possibly define an induction rule otherwise.
+%A way to concretely define the predicate-lifting function for a type is to proceed by induction on the structure of the type,
+%and we have seen in the previous sections examples of how to do so for products and type application.
+%We do not give here the general definition of lifting,
+%as that would require to first present a full type calculus,
+%and that is beyond the scope of the paper.
+%
+%Finally, associate the type
+%\begin{multline*}
+%\mathsf{CC = \forall (P : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set)} \\
+%\mathsf{\to \forall (\ol{a : Set}) (\ol{b : Set}) (\ol{Q_a : a \to Set}) (\ol{Q_b : b \to Set}) (\ol{e_a : Equal\,a\,(K\,\ol{b})}) (x : F\,G\,\ol{b})} \\
+%\mathsf{\to F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x
+%	\to P\,\ol{a}\,\ol{Q_a}\,(C\,\ol{b}\,\ol{e_a}\,x)}
+%\end{multline*}
+%to the constructor $\mathsf{C}$.
+%
+%%We can now define the induction rule for $\mathsf{G}$
+%The induction rule for $\mathsf{G}$ is
+%\begin{multline*}
+%\mathsf{\forall (P : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set) \to CC\,P} \\
+%%\mathsf{\to \Big(
+%%\forall (\ol{a : Set}) (\ol{b : Set}) (\ol{Q_a : a \to Set}) (\ol{Q_b : b \to Set}) (\ol{e_a : Equal\,a\,(K\,\ol{b})}) (x : F\,G\,\ol{b})} \\
+%%\mathsf{\to \ol{Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
+%%	\to F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x
+%%	\to P\,\ol{a}\,\ol{Q_a}\,(C\,\ol{b}\,\ol{e_a}\,x)
+%%\Big)} \\
+%\mathsf{\to \forall (\ol{a : Set})(\ol{Q_a : a \to Set})(y : G\,\ol{a}) \to G^{\wedge}\,\ol{a}\,\ol{Q_a}\,y \to P\,\ol{a}\,\ol{Q_a}\,y}
+%\end{multline*}
+%As always, we now define a term $\mathsf{DIG}$ for the induction rule.
+%Define
+%\[
+%\mathsf{DIG\,P\,cc\,\ol{a}\,\ol{Q_a}\,(C\,\ol{b}\,\ol{e_a}\,x)\,(\ol{Q_b}, L_F)
+%= cc\,\ol{a}\,\ol{b}\,\ol{Q_a}\,\ol{Q_b}\,\ol{e_a}\,x\,(p\,x\,L_F)}
+%\]
+%where
+%\begin{itemize}
+%\item $\mathsf{cc : CC\,P}$;
+%\item $\mathsf{C\,\ol{b}\,\ol{e_a}\,x : G\,\ol{a}}$,
+%i.e., $\mathsf{e_a : Equal\,a\,(K\,\ol{b})}$ for each $\mathsf{a}$,
+%and $\mathsf{x : F\,G\,\ol{b}}$;
+%\item $\mathsf{(\ol{Q_b}, L_F) : G^{\wedge}\,\ol{a}\,\ol{Q_a} (C\,\ol{b}\,\ol{e_a}\,x)}$,
+%i.e.,
+%$\mathsf{Q_b : b \to Set}$ for each $\mathsf{b}$,
+%and $\mathsf{L_F : F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,{\ol{Q_b}}\,x}$.
+%\end{itemize}
+%Finally, the morphism of predicates
+%\[
+%\mathsf{p : PredMap\,(F\,G\,\ol{b})\,(F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}) (F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})}
+%\]
+%is defined by structural induction on $\mathsf{F}$ as follows:
+%\begin{itemize}
+%\item Case $\mathsf{F = F_1 \times F_2}$ where $\mathsf{F_1}$ and $\mathsf{F_2}$ have the same type as $\mathsf{F}$.
+%We have that
+%\[
+%\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}
+%= Pair^{\wedge}(F_1\,G\,\ol{b})(F_2\,G\,\ol{b})(F_1^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}) (F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})}
+%\]
+%By inductive hypothesis, there exist morphisms of predicates
+%\begin{gather*}
+%\mathsf{p_1 : PredMap\,(F_1\,G\,\ol{b})\,((F_1)^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}) ((F_1)^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})} \\
+%\mathsf{p_2 : PredMap\,(F_2\,G\,\ol{b})\,((F_2)^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}) ((F_2)^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})}
+%\end{gather*}
+%Thus, we define $\mathsf{p (x_1, x_2) (L_1, L_2) = (p_1\,x_1\,L_1, p_2\,x_2\,L_2)}$ 
+%for 
+%$\mathsf{x_1 : F_1\,G\,\ol{b}}$,
+%$\mathsf{x_2 : F_2\,G\,\ol{b}}$,
+%$\mathsf{L_1 : F_1^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x_1}$
+%and $\mathsf{L_2 : F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x_2}$.
+%%%%
+%\item Case $\mathsf{F = F_1 + F_2}$ where $\mathsf{F_1}$ and $\mathsf{F_2}$ have the same type as $\mathsf{F}$.
+%Analogous to case $\mathsf{F = F_1 \times F_2}$.
+%%%%
+%\item Case $\mathsf{F = F_1 \to F_2}$
+%%where $\mathsf{F_1}$ and $\mathsf{F_2}$ have the same type as $\mathsf{F}$.
+%where $\mathsf{F_1}$ does not contain the recursive variable,
+%i.e., $\mathsf{F_1 : Set^\beta \to Set}$,
+%and $\mathsf{F_2}$ has the same type as $\mathsf{F}$.
+%We have that
+%\[
+%\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x = \forall (z : F_1\,\ol{b}) \to F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,z \to F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,(x\,z)}
+%\]
+%%\begin{gather*}
+%%\mathsf{F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x = \forall (z : F_1\,G\,\ol{b}) \to F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,(x\,y)} \\
+%%%F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,x \to F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x} \\
+%%\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x = \forall (z : F_1\,G\,\ol{b}) \to F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,(x\,y)}
+%%%\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x = F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,x \to F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}
+%%\end{gather*}
+%where $\mathsf{x : F\,G\,\ol{b} = F_1\,\ol{b} \to F_2\,G\,\ol{b}}$.
+%By inductive hypothesis, there exist a morphism of predicates
+%\[
+%\mathsf{p_2 : PredMap\,(F_2\,G\,\ol{b}) (F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}) (F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})}
+%\]
+%Thus, we define $\mathsf{p\,x\,L_F : F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}$
+%for $\mathsf{L_F : F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x}$
+%as $\mathsf{p\,x\,L_F\,z\,L_1 = p_2 (x\,z) (L_F\,z\,L_1)}$
+%for $\mathsf{z : F_1\,\ol{b}}$
+%and $\mathsf{L_1 : F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,z}$.
+%%{\color{blue} Revise this case!}
+%%Thus, we define $\mathsf{p\,x\,L_F\,L_1 = p_2\,(L_F\,L_1)}$
+%%for $\mathsf{L_1 : F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,x}$.
+%Notice that $\mathsf{F_1}$ does not containing the recursive variable
+%is a necessary restriction, as the proof relies on
+%$\mathsf{F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x}$
+%and $\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}$
+%having the same domain.
+%%%%
+%\item Case $\mathsf{F\,G\,\ol{b} = G(\ol{F_a\,\ol{b}})}$
+%where none of the $\mathsf{F_a}$ contains the recursive variable,
+%i.e., $\mathsf{F_a : Set^\beta \to Set}$
+%for each $\mathsf{a}$.
+%Thus, $\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b} = P (\ol{F_a\,\ol{b}}) (\ol{F_a^{\wedge}\,\ol{b}\,\ol{Q_b}})}$.
+%So, $\mathsf{p}$ is defined as
+%\[
+%\mathsf{p = DIG\,P\,cc\,(\ol{F_a\,\ol{b}})\,(\ol{F_a^{\wedge}\,\ol{b}\,\ol{Q_b}})}
+%\]
+%%for $\mathsf{L_F : F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x}$.
+%Note that the fact that the type constructors $\mathsf{F_a}$ do not contain the recursive variable
+%means that $\mathsf{G}$ does not feature nesting.
+%As explained and exemplified in Section~\ref{sec:GADT-nested},
+%that assumption is required to make induction work.
+%%%%
+%\item Case $\mathsf{F\,G\,\ol{b} = H\,\ol{b}}$
+%where $\mathsf{H}$ is a type constructor not containing the recursive variable,
+%i.e., $\mathsf{H : Set^\beta \to Set}$.
+%In such case, $\mathsf{p : PredMap (H\,\ol{b}) (H^{\wedge}\,\ol{b}\,\ol{Q_b}) (H^{\wedge}\,\ol{b}\,\ol{Q_b})}$
+%is just the identity morphism of predicates.
+%%%%
+%\item Case $\mathsf{F\,G\,\ol{b} = H (\ol{F_c\,G\,\ol{b}})}$
+%where $\mathsf{H}$ is a type constructor not containing the recursive variable,
+%i.e., $\mathsf{H : Set^\gamma \to Set}$,
+%and $\mathsf{F_c}$ has the same type as $\mathsf{F}$, for every $\mathsf{c = 1 \dots \gamma}$.
+%Moreover, we assume that $\mathsf{H}$ has an associated predicate-lifting function,
+%\[
+%\mathsf{H^{\wedge} : \forall (\ol{c : Set}) \to (\ol{c \to Set}) \to H\,\ol{c} \to Set}
+%\]
+%and that this predicate-lifting function has a map function $\mathsf{HLMap}$ of type
+%\[
+%\mathsf{\forall (\ol{c : Set}) (\ol{Q_c\;Q_c' : c \to Set}) \to PredMap\,\ol{c}\,\ol{Q_c}\,\ol{Q_c'}
+%\to PredMap\,(H\,\ol{c})\,(H^{\wedge}\,\ol{c}\,\ol{Q_c})\,(H^{\wedge}\,\ol{c}\,\ol{Q_c'})}
+%\]
+%That means that $\mathsf{H}$ cannot be a GADT,
+%as GADTs do not have functorial semantics~\cite{HaskellPaper}
+%and we would incur again in the same issue encountered in Section~\ref{sec:GADT-nested},
+%but it can be an ADT or even a nested type~\cite{jp19,jgj21}.
+%Thus,
+%\[
+%\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b} = H^{\wedge} (\ol{F_c\,G\,\ol{b}}) (\ol{F_c^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}})}
+%\]
+%By induction hypothesis, there is a morphism of predicates
+%\[
+%\mathsf{p_c : PredMap\,(F_c\,G\,\ol{b}) (F_c^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}) (F_c^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})}
+%\]
+%for every $\mathsf{c = 1 \dots \gamma}$.
+%So, $\mathsf{p}$ is defined as
+%\[
+%\mathsf{p
+%= HLMap\,(\ol{F_c\,G\,\ol{b}}) (\ol{F_c^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}}) (\ol{F_c^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}})\,\ol{p_c}}
+%\]
+%\end{itemize}
+%Notice that, by proceeding by structural induction as above,
+%we have implicitly defined a grammar for GADTs admitting an induction rule.
+%In particular, we allow $\mathsf{F}$ to feature sums, products, arrow types and (limited) type application.
+
+\section{Induction for GADTs with nesting}\label{sec:GADT-nested}
 
@@
 
-\subsection{General case}
-
-%We restrict our scope to GADTs not featuring nesting,
-%in the sense that their constructors contain no nested calls of the recursive variable,
-%as truly nested types do.
-%This way we can avoid the issue described in Section~\ref{sec:GADT-nested}.
-%The reason for this restriction will be discussed in Section~\ref{sec:GADT-nested}.
-
-Finally, we generalize the approach taken in the previous examples
-and provide a general framework to derive induction rules for GADTs.
-Since, as seen in Section~\ref{sec:GADT-nested},
-GADTs featuring nesting do not admit an induction rule,
-we restrict our scope to GADTs not featuring nesting.
-Consider such a generic GADT
-\begin{equation}\label{eq:gadt}
-\begin{array}{l}
-\mathsf{data\, G\,(\ol{a : Set}) : Set\,where}\\
-\mathsf{\;\;\;\;\;\;\;\;C\, :\, F\,G\,\ol{b} \to G (\ol{K\,\ol{b}})}
-\end{array}
-\end{equation}
-which, for ease of notation,
-we assume to have only one constructor $\mathsf{C}$.
-Extending the argument to GADTs with multiple constructors presents no difficulty.
-In the definition above, $\mathsf{F}$ has type $\mathsf{(Set^{\alpha} \to Set) \to Set^{\beta} \to Set}$
-and each $\mathsf{K}$ has type $\mathsf{Set^{\beta} \to Set}$.
-The overline notation denotes a finite list:
-%and the vertical bars notation denotes the length of a list:
-$\mathsf{G}$ depends on a finite list of types $\mathsf{\ol{a}}$ of length $\mathsf{\alpha}$,
-each of the $\mathsf{\alpha}$-many $\mathsf{K}$ depends on a finite list of types $\mathsf{\ol{b}}$ of length $\mathsf{\beta}$,
-and notice that the length of the list of types $\mathsf{\ol{a}}$ matches that of the list of type constructors $\mathsf{\ol{K}}$.
-With that in mind, the constructor $\mathsf{C}$ can be rewritten as
-\[
-\mathsf{C : \exists (\ol{b : Set}) \to \ol{Equal\,a\,(K\,\ol{b})} \to F\,G\,\ol{b} \to G\,\ol{a}}
-\]
-which is the form we shall use from now on.
-Notice that, in the above constructor, we have a type
-$\mathsf{Equal\,a\,(K\,\ol{b})}$ for each $\mathsf{a}$ and its corresponding type constructor $\mathsf{K}$.
-
-In order to state the induction rule for $\mathsf{G}$,
-we first need to define $\mathsf{G}$'s associated predicate-lifting function 
-\[
-\mathsf{G^{\wedge} : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set}
-\]
-as
-\[
-\mathsf{G^{\wedge}\,\ol{a}\,\ol{Q_a}\,(C\,\ol{b}\,\ol{e_a}\,x)
-= \exists (\ol{Q_b : b \to Set})
-\to \ol{Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
-\times F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,{\ol{Q_b}}\,x}
-\]
-where $\mathsf{Q_a : a \to Set}$
-and $\mathsf{C\,\ol{b}\,\ol{e_a}\,x : G\,\ol{a}}$,
-i.e., $\mathsf{e_a : Equal\,a\,(K\,\ol{b})}$ and $\mathsf{x : F\,G\,\ol{b}}$.
-Notice that we need a type $\mathsf{Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}$
-for each type $\mathsf{a}$ and its corresponding type constructor $\mathsf{K}$.
-We also assume to have liftings for $\mathsf{F}$,
-\begin{multline*}
-\mathsf{F^{\wedge} : \forall (G : Set^{\alpha} \to Set) (\ol{b : Set})
-\to (\forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set)} \\
-\mathsf{\to (\ol{b \to Set})
-\to F\,G\,\ol{b} \to Set}
-\end{multline*}
-and for each  $\mathsf{K}$,
-\[
-\mathsf{K^{\wedge} : \forall (\ol{b : Set}) \to (\ol{b \to Set}) \to K\,\ol{b} \to Set}
-\]
-as we could not possibly define an induction rule otherwise.
-A way to concretely define the predicate-lifting function for a type is to proceed by induction on the structure of the type,
-and we have seen in the previous sections examples of how to do so for products and type application.
-We do not give here the general definition of lifting,
-as that would require to first present a full type calculus,
-and that is beyond the scope of the paper.
-
-Finally, associate the type
-\begin{multline*}
-\mathsf{CC = \forall (P : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set)} \\
-\mathsf{\to \forall (\ol{a : Set}) (\ol{b : Set}) (\ol{Q_a : a \to Set}) (\ol{Q_b : b \to Set}) (\ol{e_a : Equal\,a\,(K\,\ol{b})}) (x : F\,G\,\ol{b})} \\
-\mathsf{\to \ol{Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
-	\to F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x
-	\to P\,\ol{a}\,\ol{Q_a}\,(C\,\ol{b}\,\ol{e_a}\,x)}
-\end{multline*}
-to the constructor $\mathsf{C}$.
-
-%We can now define the induction rule for $\mathsf{G}$
-The induction rule for $\mathsf{G}$ is
-\begin{multline*}
-\mathsf{\forall (P : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set) \to CC\,P} \\
-%\mathsf{\to \Big(
-%\forall (\ol{a : Set}) (\ol{b : Set}) (\ol{Q_a : a \to Set}) (\ol{Q_b : b \to Set}) (\ol{e_a : Equal\,a\,(K\,\ol{b})}) (x : F\,G\,\ol{b})} \\
-%\mathsf{\to \ol{Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
-%	\to F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x
-%	\to P\,\ol{a}\,\ol{Q_a}\,(C\,\ol{b}\,\ol{e_a}\,x)
-%\Big)} \\
-\mathsf{\to \forall (\ol{a : Set})(\ol{Q_a : a \to Set})(y : G\,\ol{a}) \to G^{\wedge}\,\ol{a}\,\ol{Q_a}\,y \to P\,\ol{a}\,\ol{Q_a}\,y}
-\end{multline*}
-As always, we now define a term $\mathsf{DIG}$ for the induction rule.
-Define
-\[
-\mathsf{DIG\,P\,cc\,\ol{a}\,\ol{Q_a}\,(C\,\ol{b}\,\ol{e_a}\,x)\,(\ol{Q_b}, \ol{E_a}, L_F)
-= cc\,\ol{a}\,\ol{b}\,\ol{Q_a}\,\ol{Q_b}\,\ol{e_a}\,x\,\ol{E_a}\,(p\,x\,L_F)}
-\]
-where
-\begin{itemize}
-\item $\mathsf{cc : CC\,P}$;
-\item $\mathsf{C\,\ol{b}\,\ol{e_a}\,x : G\,\ol{a}}$,
-i.e., $\mathsf{e_a : Equal\,a\,(K\,\ol{b})}$ for each $\mathsf{a}$,
-and $\mathsf{x : F\,G\,\ol{b}}$;
-\item $\mathsf{(\ol{Q_b}, \ol{E_a}, L_F) : G^{\wedge}\,\ol{a}\,\ol{Q_a} (C\,\ol{b}\,\ol{e_a}\,x)}$,
-i.e.,
-$\mathsf{Q_b : b \to Set}$ for each $\mathsf{b}$,
-$\mathsf{E_a : Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}$ for each $\mathsf{a}$,
-and $\mathsf{L_F : F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,{\ol{Q_b}}\,x}$.
-\end{itemize}
-Finally, the morphism of predicates
-\[
-\mathsf{p : PredMap\,(F\,G\,\ol{b})\,(F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}) (F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})}
-\]
-is defined by structural induction on $\mathsf{F}$ as follows:
-\begin{itemize}
-\item Case $\mathsf{F = F_1 \times F_2}$ where $\mathsf{F_1}$ and $\mathsf{F_2}$ have the same type as $\mathsf{F}$.
-We have that
-\[
-\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}
-= Pair^{\wedge}(F_1\,G\,\ol{b})(F_2\,G\,\ol{b})(F_1^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}) (F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})}
-\]
-By inductive hypothesis, there exist morphisms of predicates
-\begin{gather*}
-\mathsf{p_1 : PredMap\,(F_1\,G\,\ol{b})\,((F_1)^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}) ((F_1)^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})} \\
-\mathsf{p_2 : PredMap\,(F_2\,G\,\ol{b})\,((F_2)^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}) ((F_2)^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})}
-\end{gather*}
-Thus, we define $\mathsf{p (x_1, x_2) (L_1, L_2) = (p_1\,x_1\,L_1, p_2\,x_2\,L_2)}$ 
-for 
-$\mathsf{x_1 : F_1\,G\,\ol{b}}$,
-$\mathsf{x_2 : F_2\,G\,\ol{b}}$,
-$\mathsf{L_1 : F_1^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x_1}$
-and $\mathsf{L_2 : F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x_2}$.
-%%%
-\item Case $\mathsf{F = F_1 + F_2}$ where $\mathsf{F_1}$ and $\mathsf{F_2}$ have the same type as $\mathsf{F}$.
-Analogous to case $\mathsf{F = F_1 \times F_2}$.
-%%%
-\item Case $\mathsf{F = F_1 \to F_2}$
-where $\mathsf{F_1}$ and $\mathsf{F_2}$ have the same type as $\mathsf{F}$.
-%where $\mathsf{F_1}$ does not contain the recursive variable,
-%i.e., $\mathsf{F_a : Set^\beta \to Set}$,
-%and $\mathsf{F_2}$ has the same type as $\mathsf{F}$.
-We have that
-\begin{gather*}
-\mathsf{F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x = \forall (z : F_1\,G\,\ol{b}) \to F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,(x\,y)} \\
-%F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,x \to F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x} \\
-\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x = \forall (z : F_1\,G\,\ol{b}) \to F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,(x\,y)}
-%\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x = F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,x \to F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}
-\end{gather*}
-where $\mathsf{x : F\,G\,\ol{b} = F_1\,G\,\ol{b} \to F_2\,G\,\ol{b}}$.
-By inductive hypothesis, there exist a morphism of predicates
-\[
-\mathsf{p_2 : PredMap\,(F_2\,G\,\ol{b}) (F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}) (F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})}
-\]
-Thus, we define $\mathsf{p\,x\,L_F\,z = p_2 (x\,z) (L_F\,z)}$.
-%{\color{blue} Revise this case!}
-%Thus, we define $\mathsf{p\,x\,L_F\,L_1 = p_2\,(L_F\,L_1)}$
-%for $\mathsf{L_1 : F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,x}$.
-%Notice that $\mathsf{F_1}$ does not contain the recursive variable.
-%That is a necessary restriction, as the proof relies on
-%$\mathsf{F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x}$
-%and $\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}$
-%having the same domain.
-%%%
-\item Case $\mathsf{F\,G\,\ol{b} = G(\ol{F_a\,\ol{b}})}$
-where none of the $\mathsf{F_a}$ contains the recursive variable,
-i.e., $\mathsf{F_a : Set^\beta \to Set}$
-for each $\mathsf{a}$.
-Thus, $\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b} = P (\ol{F_a\,\ol{b}}) (\ol{F_a^{\wedge}\,\ol{b}\,\ol{Q_b}})}$.
-So, $\mathsf{p}$ is defined as
-\[
-\mathsf{p = DIG\,P\,cc\,(\ol{F_a\,\ol{b}})\,(\ol{F_a^{\wedge}\,\ol{b}\,\ol{Q_b}})}
-\]
-%for $\mathsf{L_F : F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x}$.
-Note that the fact that the type constructors $\mathsf{F_a}$ do not contain the recursive variable
-means that $\mathsf{G}$ does not feature nesting.
-As explained and exemplified in Section~\ref{sec:GADT-nested},
-that assumption is required to make induction work.
-%%%
-\item Case $\mathsf{F\,G\,\ol{b} = H\,\ol{b}}$
-where $\mathsf{H}$ is a type constructor not containing the recursive variable,
-i.e., $\mathsf{H : Set^\beta \to Set}$.
-In such case, $\mathsf{p : PredMap (H\,\ol{b}) (H^{\wedge}\,\ol{b}\,\ol{Q_b}) (H^{\wedge}\,\ol{b}\,\ol{Q_b})}$
-is just the identity morphism of predicates.
-%%%
-\item Case $\mathsf{F\,G\,\ol{b} = H (\ol{F_c\,G\,\ol{b}})}$
-where $\mathsf{H}$ is a type constructor not containing the recursive variable,
-i.e., $\mathsf{H : Set^\gamma \to Set}$,
-and $\mathsf{F_c}$ has the same type as $\mathsf{F}$, for every $\mathsf{c = 1 \dots \gamma}$.
-Moreover, we assume that $\mathsf{H}$ has an associated predicate-lifting function,
-\[
-\mathsf{H^{\wedge} : \forall (\ol{c : Set}) \to (\ol{c \to Set}) \to H\,\ol{c} \to Set}
-\]
-and that this predicate-lifting function has a map function $\mathsf{HLMap}$ of type
-\[
-\mathsf{\forall (\ol{c : Set}) (\ol{Q_c\;Q_c' : c \to Set}) \to PredMap\,\ol{c}\,\ol{Q_c}\,\ol{Q_c'}
-\to PredMap\,(H\,\ol{c})\,(H^{\wedge}\,\ol{c}\,\ol{Q_c})\,(H^{\wedge}\,\ol{c}\,\ol{Q_c'})}
-\]
-That means that $\mathsf{H}$ cannot be a GADT,
-as GADTs do not have functorial semantics~\cite{HaskellPaper}
-and we would incur again in the same issue encountered in Section~\ref{sec:GADT-nested},
-but it can be an ADT or even a nested type~\cite{jp19,jgj21}.
-Thus,
-\[
-\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b} = H^{\wedge} (\ol{F_c\,G\,\ol{b}}) (\ol{F_c^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}})}
-\]
-By induction hypothesis, there is a morphism of predicates
-\[
-\mathsf{p_c : PredMap\,(F_c\,G\,\ol{b}) (F_c^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}) (F_c^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})}
-\]
-for every $\mathsf{c = 1 \dots \gamma}$.
-So, $\mathsf{p}$ is defined as
-\[
-\mathsf{p
-= HLMap\,(\ol{F_c\,G\,\ol{b}}) (\ol{F_c^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}}) (\ol{F_c^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}})\,\ol{p_c}}
-\]
-
-\end{itemize}
-Notice that, by proceeding by structural induction as above,
-we have implicitly defined a grammar for GADTs admitting an induction rule.
-In particular, we allow $\mathsf{F}$ to feature sums, products, arrow types and (limited) type application.
-
 \section{Applications}

### 2021-04-05 induction-paper/GADT-induction.tex
Fix extra overline.
Comment on structural induction as consequence of deep induction.

@@
 and $\mathsf{PTree^{\wedge} : \forall (a : Set) \to (a \to Set) \to PTree\,a \to Set}$ lifts a predicate $\mathsf{Q}$ on data of type $\mathsf{a}$ to a predicate on data of type $\mathsf{PTree\,a}$ asserting that $\mathsf{Q}$ holds for every element of type $\mathsf{a}$ contained in its argument perfect tree.
+
+{\color{blue} Moreover, for types admitting both deep induction and structural induction, the former generalizes the latter.
+Indeed, structural induction rules can be derived from deep induction rules by choosing the constantly true predicate as the custom predicate traversing each data element contained in the data type.
+This way, deep induction only inspects the structure of the data type and not its content, just like structural induction does.
+A concrete example of this technique will be demonstrated in Section~\ref{sec:ind-equal}.}
 
@@
 
-\subsection{(Deep) induction for $\mathsf{Equal}$}
+\subsection{(Deep) induction for $\mathsf{Equal}$}\label{sec:ind-equal}
 
@@
 %\mathsf{\to \big( \forall (c : Set) (Q : c \to Set) (Q' : c \to Set) \to Equal^{\wedge}\,c\,c\,Q\,Q'\,Refl \to P\,c\,c\,Q\,Q'\,Refl \big)} \\
-\mathsf{\to CRefl\,P \to \forall (a : Set) (b : Set) (Q_a : a \to Set) (Q_b : b \to Set) (e: Equal\,a\,b)} \\
+\mathsf{\to CRefl\,P \to \forall (a\,b : Set) (Q_a : a \to Set) (Q_b : b \to Set) (e: Equal\,a\,b)} \\
 	\mathsf{\to Equal^{\wedge}\,a\,b\,Q_a\,Q_b\,e \to P\,a\,b\,Q_a\,Q_b\,e}
@@
 we have shown that the induction rule for $\mathsf{Equal}$ holds.
+
+The type $\mathsf{Equal}$ also has a standard structural induction rule $\mathsf{SIEqual}$,
+\begin{multline*}
+\mathsf{\forall (Q : \forall (a\,b : Set) \to Equal\,a\,b \to Set)} \\
+\mathsf{\to \big( \forall (c : Set) \to P\,c\,c\,Refl \big)
+	\to \forall (a\,b : Set) (e: Equal\,a\,b) \to P\,a\,b\,e}
+\end{multline*}
+As is the case for ADTs and nested types,
+the structural induction rule for $\mathsf{Equal}$ is a consequence of the deep induction rule.
+Indeed, we can define $\mathsf{SIEqual}$ as
+\[
+\mathsf{SIEqual\,Q\,srefl\,a\,b\,e = DIEqual\,P\,srefl\,a\,b\,K_1\,K_1\,e\,L_E}
+\]
+where $\mathsf{srefl: \forall (c : Set) \to P\,c\,c\,Refl}$
+and $\mathsf{e : Equal\,a\,b}$, and
+\begin{itemize}
+\item $\mathsf{P : \forall (a\,b : Set) \to (a \to Set) \to (b \to Set) \to Equal\,a\,b \to Set}$ is defined as
+$\mathsf{P\,a\,b\,Q_a\,Q_b\,e = Q\,a\,b\,e}$;
+\item the left-most $\mathsf{K_1}$ is the constantly $\mathsf{1}$-valued predicate on $\mathsf{a}$;
+\item the right-most $\mathsf{K_1}$ is the constantly $\mathsf{1}$-valued predicate on $\mathsf{b}$;
+\item $\mathsf{L_E : Equal^{\wedge}\,a\,b\,K_1\,K_1\,e}$ is defined by pattern matching,
+i.e., in case $\mathsf{a = b}$ and $\mathsf{e}$ is $\mathsf{Refl}$,
+it is defined as $\mathsf{L_E\,x = Refl : Equal\,a\,a}$.
+\end{itemize}
+
+That the structural induction rule is a consequence of the deep induction one is also true for all the examples below,
+even though we will not remark it every time.
 
@@
 $\mathsf{s_b : Seq\,b}$ and $\mathsf{s_c : Seq\,c}$.
-We also need the lifting of predicates over the polymorphic type of pairs, $\mathsf{Pair = \forall (b : Set)(c : Set) \to b \times c}$, which is 
-\[
-\mathsf{Pair^{\wedge} : \forall (b : Set)(c : Set) \to (b \to Set) \to (c \to Set) \to Pair\,b\,c \to Set}
+We also need the lifting of predicates over the polymorphic type of pairs, $\mathsf{Pair = \forall (b\,c : Set) \to b \times c}$, which is 
+\[
+\mathsf{Pair^{\wedge} : \forall (b\,c : Set) \to (b \to Set) \to (c \to Set) \to Pair\,b\,c \to Set}
 \]
@@
 \mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set)} \\
-\mathsf{\to \forall (a : Set) (b : Set) (c : Set)
+\mathsf{\to \forall (a\,b\,c : Set)
 	(Q_a : a \to Set) (Q_b : b \to Set) (Q_c : c \to Set)} \\
@@
 \mathsf{data\, G\,(a : Set) : Set\,where}\\
-\mathsf{\;\;\;\;\;\;\;\;C\, :\, F\,G\,\ol{b} \to G (\ol{K\,\ol{b}})}
+\mathsf{\;\;\;\;\;\;\;\;C\, :\, F\,G\,\ol{b} \to G (K\,\ol{b})}
 \end{array}

### 2021-04-06 induction-paper/GADT-induction.tex
Revise Section 4

--- 2021-03-31 induction-paper/GADT-induction.tex
@@
 and $\mathsf{PTree^{\wedge} : \forall (a : Set) \to (a \to Set) \to PTree\,a \to Set}$ lifts a predicate $\mathsf{Q}$ on data of type $\mathsf{a}$ to a predicate on data of type $\mathsf{PTree\,a}$ asserting that $\mathsf{Q}$ holds for every element of type $\mathsf{a}$ contained in its argument perfect tree.
+
+{\color{blue} Moreover, for types admitting both deep induction and structural induction, the former generalizes the latter.
+Indeed, structural induction rules can be derived from deep induction rules by choosing the constantly true predicate as the custom predicate traversing each data element contained in the data type.
+That way, deep induction only inspects the structure of the data type and not its content, just like structural induction does.
+A concrete example of this technique will be demonstrated in Section~\ref{sec:ind-equal}.}
 
@@
 In Agda, we code it as
-\begin{equation}\label{eq:eq}
+\begin{equation}\label{eq:equal}
 \begin{array}{l}
@@
 \mathsf{\;\;\;\;\;\;\;\;Const :\, a \to Seq\,a}\\
-\mathsf{\;\;\;\;\;\;\;\;SPair\,\, :\, \exists (b : Set) (c : Set).\,Equal\,a\,(b \times c) \to Seq\,b \to Seq\,c \to
+\mathsf{\;\;\;\;\;\;\;\;SPair\,\, :\, \exists (b\,c : Set).\,Equal\,a\,(b \times c) \to Seq\,b \to Seq\,c \to
   Seq\,a}\\ 
@@
 \end{equation}
-where the requirement that the $\mathsf{SPair}$ constructor produce an instance of $\mathsf{Seq}$ at a product type has been replaced with the requirement that the instance of $\mathsf{Seq}$ returned by $\mathsf{SPair}$ is \emph{equal} to a product.
+where the requirement that the $\mathsf{SPair}$ constructor produces an instance of $\mathsf{Seq}$ at a product type has been replaced with the requirement that the instance of $\mathsf{Seq}$ returned by $\mathsf{SPair}$ is \emph{equal} to some product type.
 This encoding is particularly convenient when representing GADTs as Church encodings~\cite{vw10,atk12}.
 
-Add third example here.
+{\color{red} Add third example here.}
 
@@
 
-As we have seen in Subsection~\ref{ssec:induction-nested},
+As we have seen in Section~\ref{ssec:induction-nested},
 truly nested types do not support a structural induction rule,
@@
 are traditionally derived using the functorial semantics of data types
-in the setting of a parametric model~\cite{jp20}.
+in the setting of a parametric model~\cite{jp19}.
 In particular, relational parametricity is used to validate the induction principle because induction is, itself,
@@
 
-As a first example, we derive the induction rule for the $\mathsf{Equal}$ type.
-This will provide a simple case study that will help with the derivation of the induction rule for more complex GADTs.
+As a first example, we derive the induction rule for the $\mathsf{Equal}$ type from Equation~\ref{eq:equal}.
+This will provide a simple case study that will inform the investigation of more complex GADTs.
 Moreover, since we define GADTs using the $\mathsf{Equal}$ type,
 as for example in Equation~\ref{eq:eq-seq},
-this example will be instrumental in the induction rule of other GADTs.
-
-%We first need to represent $\mathsf{Equal}$ as 
-%\begin{equation}\label{eq:alt-eq}
-%\begin{array}{l}
-%\mathsf{data\, Equal\,(a\,b : Set) : Set\,where}\\
-%\mathsf{\;\;\;\;\;\;\;\;Refl :\, \exists (c : Set) \to Equal\,a\,c \to Equal\,b\,c \to Equal\,a\,b}
-%\end{array}
-%\end{equation}
-%\begin{multline}
-%\mathsf{\forall (P : \forall (a\,b : Set) \to (a \to Set) \to (b \to Set) \to Equal\,a\,b \to Set)} \\
-%\mathsf{\to \big( \forall (a : Set) (b : Set) (c : Set) (Q_a : a \to Set) (Q_b : b \to Set) (Q_c : c \to Set) (e_a: Equal\,a\,c) (e_b: Equal\,b\,c)}\\
-%	\mathsf{\to Equal^{\wedge}\,a\,c\,Q_a\,Q_c\,e_a \to Equal^{\wedge}\,b\,c\,Q_b\,Q_c\,e_b \to P\,c\,c\,Q_c\,Q_c\,Refl \big)} \\
-%\mathsf{\to \big( \forall (a : Set) (b : Set) (Q_a : a \to Set) (Q_b : b \to Set) (e: Equal\,a\,b)
-%	\to Equal^{\wedge}\,a\,b\,Q_a\,Q_b\,e \to P\,a\,b\,Q_a\,Q_b\,e \big)}
-%\end{multline}
-
-To define the induction rule for $\mathsf{Equal}$ we first need a predicate-lifting operation
-\[
-\mathsf{Equal^{\wedge} : \forall (a : Set) (b : Set) \to (a \to Set) \to (b \to Set) \to Equal\,a\,b \to Set}
+this example will be instrumental in stating and deriving the induction rule of other GADTs.
+
+To define an induction rule for a type $\mathsf{G}$ we first need a predicate-lifting operation
+which takes predicates on a type $\mathsf{a}$ and lifts them to predicates on $\mathsf{G\,a}$.
+the predicate-lifting function for $\mathsf{Equal}$ is the function
+\[
+\mathsf{Equal^{\wedge} : \forall (a\,b : Set) \to (a \to Set) \to (b \to Set) \to Equal\,a\,b \to Set}
 \]
@@
 \]
-
-Let $\mathsf{CRefl}$ be the following type associated to the $\mathsf{Refl}$ constructor:
+i.e., the function that takes two predicates on the same type and tests them for extensional equality.
+
+Next, we need to associate each constructor of the GADT under consideration
+to the expression that a given predicate is preserved by such constructor.
+Let $\mathsf{CRefl}$ be the following function associated to the $\mathsf{Refl}$ constructor:
 \begin{multline*}
-\mathsf{\forall (P : \forall (a\,b : Set) \to (a \to Set) \to (b \to Set) \to Equal\,a\,b \to Set)} \\
+\mathsf{\lambda (P : \forall (a\,b : Set) \to (a \to Set) \to (b \to Set) \to Equal\,a\,b \to Set)} \\
 \mathsf{\to \forall (c : Set) (Q : c \to Set) (Q' : c \to Set) \to Equal^{\wedge}\,c\,c\,Q\,Q'\,Refl \to P\,c\,c\,Q\,Q'\,Refl}
@@
 
-The induction rule for $\mathsf{Equal}$ is the type
+The induction rule states that,
+if a predicate is preserved by all of the constructors of the GADT under consideration,
+then the predicate is satisfied by any element of the GADT.
+The induction rule for $\mathsf{Equal}$ is thus the type
 \begin{multline*}
@@
 %\mathsf{\to \big( \forall (c : Set) (Q : c \to Set) (Q' : c \to Set) \to Equal^{\wedge}\,c\,c\,Q\,Q'\,Refl \to P\,c\,c\,Q\,Q'\,Refl \big)} \\
-\mathsf{\to CRefl\,P \to \forall (a : Set) (b : Set) (Q_a : a \to Set) (Q_b : b \to Set) (e: Equal\,a\,b)} \\
+\mathsf{\to CRefl\,P \to \forall (a\,b : Set) (Q_a : a \to Set) (Q_b : b \to Set) (e: Equal\,a\,b)} \\
 	\mathsf{\to Equal^{\wedge}\,a\,b\,Q_a\,Q_b\,e \to P\,a\,b\,Q_a\,Q_b\,e}
 \end{multline*}
-We define a term $\mathsf{DIEqual}$ of such type as
+To validate the induction rule we need to provide it with a witness, i.e.,
+we need to show that the associated type is inhabited.
+We define a term $\mathsf{DIEqual}$ of the above type as
 \[
@@
 $\mathsf{crefl : CRefl\,P}$,
-%$\mathsf{PRefl : \forall (c : Set) (Q : c \to Set) (Q' : c \to Set) \to Equal^{\wedge}\,c\,c\,Q\,Q'\,Refl \to P\,c\,c\,Q\,Q'\,Refl}$
 $\mathsf{Q_a : a \to Set}$, $\mathsf{Q_a' : a \to Set}$
@@
 Having provided a well-defined term for it,
-we have shown that the induction rule for $\mathsf{Equal}$ holds.
+we have shown that the induction rule for $\mathsf{Equal}$ is sound.
+
+The type $\mathsf{Equal}$ also has a standard structural induction rule $\mathsf{SIEqual}$,
+\begin{multline*}
+\mathsf{\forall (Q : \forall (a\,b : Set) \to Equal\,a\,b \to Set)} \\
+\mathsf{\to \big( \forall (c : Set) \to P\,c\,c\,Refl \big)
+	\to \forall (a\,b : Set) (e: Equal\,a\,b) \to P\,a\,b\,e}
+\end{multline*}
+As is the case for ADTs and nested types,
+the structural induction rule for $\mathsf{Equal}$ is a consequence of the deep induction rule.
+Indeed, we can define $\mathsf{SIEqual}$ as
+\[
+\mathsf{SIEqual\,Q\,srefl\,a\,b\,e = DIEqual\,P\,srefl\,a\,b\,K^a_1\,K^b_1\,e\,L_E}
+\]
+where
+$\mathsf{Q : \forall (a\,b : Set) \to Equal\,a\,b \to Set}$,
+$\mathsf{srefl: \forall (c : Set) \to P\,c\,c\,Refl}$
+and $\mathsf{e : Equal\,a\,b}$, and
+\begin{itemize}
+\item $\mathsf{P : \forall (a\,b : Set) \to (a \to Set) \to (b \to Set) \to Equal\,a\,b \to Set}$ is defined as
+$\mathsf{P\,a\,b\,Q_a\,Q_b\,e = Q\,a\,b\,e}$;
+\item $\mathsf{K^a_1}$ and $\mathsf{K^b_1}$ are the constantly $\mathsf{1}$-valued predicates on, respectively, $\mathsf{a}$ and $\mathsf{b}$;
+%\item $\mathsf{K^b_1}$ is the constantly $\mathsf{1}$-valued predicate on $\mathsf{b}$;
+\item $\mathsf{L_E : Equal^{\wedge}\,a\,b\,K^a_1\,K^b_1\,e}$ is defined by pattern matching,
+i.e., in case $\mathsf{a = b}$ and $\mathsf{e = Refl}$,
+it is defined as $\mathsf{L_E\,x = Refl : Equal\,a\,a}$.
+\end{itemize}
+
+That the structural induction rule is a consequence of the deep induction one is also true for all the examples below,
+even though we will not remark it every time.
 
@@
 $\mathsf{s_b : Seq\,b}$ and $\mathsf{s_c : Seq\,c}$.
-We also need the lifting of predicates over the polymorphic type of pairs, $\mathsf{Pair = \forall (b : Set)(c : Set) \to b \times c}$, which is 
-\[
-\mathsf{Pair^{\wedge} : \forall (b : Set)(c : Set) \to (b \to Set) \to (c \to Set) \to Pair\,b\,c \to Set}
+We also need to define the lifting of predicates over the polymorphic type of pairs, $\mathsf{Pair = \forall (b\,c : Set) \to b \times c}$, which is 
+\[
+\mathsf{Pair^{\wedge} : \forall (b\,c : Set) \to (b \to Set) \to (c \to Set) \to Pair\,b\,c \to Set}
 \]
@@
 
-Finally, let $\mathsf{CConst}$ be the type
+Finally, let $\mathsf{CConst}$ be the function
 \begin{multline*}
-\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set)} \\
+\mathsf{\lambda (P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set)} \\
 \mathsf{\to \forall (a : Set) (Q_a : a \to Set) (x : a) \to Q_a\,x \to P\,a\,Q_a\,(Const\,x)}
@@
 associated to the $\mathsf{Const}$ constructor,
-and let $\mathsf{CSPair}$ be the type
+and let $\mathsf{CSPair}$ be the function
 \begin{multline*}
-\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set)} \\
-\mathsf{\to \forall (a : Set) (b : Set) (c : Set)
+\mathsf{\lambda (P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set)} \\
+\mathsf{\to \forall (a\,b\,c : Set)
 	(Q_a : a \to Set) (Q_b : b \to Set) (Q_c : c \to Set)} \\
@@
 
-With these tools, we have an induction rule for $\mathsf{Seq}$,
+With these tools we can formulate an induction rule for $\mathsf{Seq}$,
 \begin{multline*}
@@
 \mathsf{\to CConst\,P \to CSPair\,P} \\
-%\mathsf{\to \big( \forall (a : Set) (Q_a : a \to Set) (x : a) \to Q_a\,x \to P\,a\,Q_a\,(Const\,x) \big)} \\
-%\mathsf{\to \big( \forall (a : Set) (b : Set) (c : Set)
-%	(Q_a : a \to Set) (Q_b : b \to Set) (Q_c : c \to Set)} \\
-%	\mathsf{(s_b : Seq\,b) (s_c : Seq\,c) (e : Equal\,a\,(b \times c))
-%	\to Equal^{\wedge} a\, (b \times c)\, Q_a\, (Pair^{\wedge}\,b\,c\,Q_b\,Q_c)\, e} \\
-%	\mathsf{\to P\,b\,Q_b\,s_b \to P\,c\,Q_c\,s_c
-%	\to P a Q_a ( SPair\,b\,c\,e\,s_b\,s_c ) \big)} \\
 \mathsf{\to \forall (a : Set)(Q_a : a \to Set)(s_a : Seq\,a) \to Seq^{\wedge}\,a\,Q_a\,s_a \to P\,a\,Q_a\,s_a}
 \end{multline*}
-To validate the induction rule, we define a term $\mathsf{DISeq}$ for it.
+To validate the induction rule, we define a term $\mathsf{DISeq}$ for the above type.
 We have to define
@@
 $\mathsf{cspair : CSPair\,P}$,
-% are terms for the clauses
-%associated to the constructors $\mathsf{Const}$ and $\mathsf{SPair}$,
-%respectively, in the induction rule,
 $\mathsf{Q_a : a \to Set}$, $\mathsf{s_a : Seq\,a}$
@@
 \item $\mathsf{Q_b : b \to Set}$ and $\mathsf{Q_c : c \to Set}$;
-%\item $\mathsf{p_b : P\,b\,Q_b\,s_b}$ and $\mathsf{p_c : P\,c\,Q_c\,s_c}$;
 \item $\mathsf{L_e : Equal^{\wedge}\,a\, (b \times c)\, Q_a\, (Q_b \times Q_c) \, e}$;
 \item $\mathsf{L_b : Seq^{\wedge}\,b\,Q_b\,s_b}$ and $\mathsf{L_c : Seq^{\wedge}\,c\,Q_c\,s_c}$;
-%(notice that $\mathsf{(Q_b, Q_c, L_e, L_b, L_c) : Seq^{\wedge}\,a\,Q\,(SPair\,b\,c\,e\,x\,y)}$,
-%making the right-hand-side in the above expression type-check)
 \end{itemize}
@@
 
-%\subsection{(Deep) induction for $\mathsf{Seq}$ (general style)}
-%
-%We need the lifting of the type $\mathsf{Pair = \forall (b : Set)(c : Set) \to b \times c}$, which is 
-%\[
-%\mathsf{Pair^{\wedge} : \forall (b : Set)(c : Set) \to (b \to Set) \to (c \to Set) \to b \times c \to Set}
-%\]
-%defined as
-%\[
-%\mathsf{Pair^{\wedge}\,b\,c\,Q_b\,Q_c\,(y, z) = Q_b\,y \times Q_c\,z}
-%\]
-%
-%%Let 
-%%$\mathsf{F, F_1, F_2 : (Set \to Set) \to Set \to Set \to Set}$
-%%be defined as
-%%$\mathsf{F_1\,G\,b\,c = G\,b}$,
-%%$\mathsf{F_2\,G\,b\,c = G\,c}$,
-%%$\mathsf{F\,G\,b\,c = F_1\,G\,b\,c \times F_2\,G\,b\,c}$,
-%%and $\mathsf{K = Pair}$.
-%
-%
-%Let 
-%$\mathsf{F : (Set \to Set) \to Set \to Set \to Set}$
-%be defined as
-%$\mathsf{F\,G\,b\,c = G\,b \times G\,c}$
-%and $\mathsf{K = Pair}$.
-%
-%%The type of $\mathsf{F_i^{\wedge}}$ is
-%%\begin{multline*}
-%%\mathsf{\forall (G : Set \to Set) (b : Set) (c : Set)
-%%\to (\forall (a : Set) \to (a \to Set) \to G\,a \to Set)} \\
-%%\mathsf{\to (b \to Set) \to (c \to Set) \to (F_i\,G\,b\,c) \to Set}
-%%\end{multline*}
-%The type of $\mathsf{F^{\wedge}}$ is
-%\begin{multline*}
-%\mathsf{\forall (G : Set \to Set) (b : Set) (c : Set)
-%\to (\forall (a : Set) \to (a \to Set) \to G\,a \to Set)} \\
-%\mathsf{\to (b \to Set) \to (c \to Set) \to (F_1\,G\,b\,c \times F_2\,G\,b\,c) \to Set}
-%\end{multline*}
-%%and $\mathsf{F^{\wedge}\,G\,b\,c\,P\,Q_b\,Q_c\,(s_b, s_c)
-%%= F_1^{\wedge}\,G\,b\,c\,P\,Q_b\,Q_c\,s_b \times F_2^{\wedge}\,G\,b\,c\,P\,Q_b\,Q_c\,s_c}$
-%
-%Next we shall consider the $\mathsf{Seq}$ type defined in Equation~\ref{eq:eq_seq}.
-%Again, the first step in deriving its induction rule consists in defining the lifting
-%\[
-%\mathsf{Seq^{\wedge} : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set}
-%\]
-%which is given by
-%\[
-%\mathsf{Seq^{\wedge}\,a\,Q\,(Const\,x) = Q\,x}
-%\]
-%and
-%\begin{multline*}
-%\mathsf{Seq^{\wedge}\,a\,Q_a\,(SPair\,b\,c\,e\,(s_b,s_c))
-%=
-%\exists (Q_b : b \to Set) (Q_c : c \to Set)} \\
-%\mathsf{
-%\to Equal^{\wedge}\,a\, (Pair\,b\,c)\, Q_a\, (Pair^{\wedge}\,b\,c\,Q_b\,Q_c) \, e
-%\times F^{\wedge}\,Seq\,b\,c\,Seq^{\wedge}\,Q_b\,Q_c\,(s_b, s_c)
-%}
-%\end{multline*}
-%%\begin{multline*}
-%%\mathsf{Seq^{\wedge}\,a\,Q_a\,(SPair\,b\,c\,e\,(s_b,s_c))
-%%=
-%%\exists (Q_b : b \to Set) (Q_c : c \to Set)} \\
-%%\mathsf{
-%%\to Equal^{\wedge}\,a\, (Pair\,b\,c)\, Q_a\, (Pair^{\wedge}\,b\,c\,Q_b\,Q_c) \, e
-%%\times F_1^{\wedge}\,Seq\,b\,c\,Seq^{\wedge}\,Q_b\,Q_c\,s_b
-%%\times F_2^{\wedge}\,Seq\,b\,c\,Seq^{\wedge}\,Q_b\,Q_c\,s_c
-%%}
-%%\end{multline*}
-%
-%Define
-%\[
-%\mathsf{PredMap : \forall (a : Set) \to (a \to Set) \to (a \to Set) \to Set}
-%\]
-%as $\mathsf{PredMap\,a\,Q_1\,Q_2 = \forall (x : a) \to Q_1\,x \to Q_2\,x}$
-%
-%%  -- Lifting also acts on morphisms between predicates with the same carrier
-%%  {-# TERMINATING #-}
-%%  Bush^-map : ∀ {l} {X : Set} {Q1 Q2 : X → Set l}
-%%              → Pred-map {l} Q1 Q2 → Pred-map (Bush^ Q1) (Bush^ Q2)
-%%  Bush^-map Q12 Bnil q = bigtt
-%%  Bush^-map Q12 (Bcons x b) (y , bby) =
-%%                Q12 x y , Bush^-map (Bush^-map Q12) b bby
-%Define
-%\[
-%\mathsf{SeqLMap : \forall (a : Set) (Q_1\,Q_2 : a \to Set) \to PredMap\,a\,Q_1\,Q_2 \to PredMap\,(G\,a)\,(Seq^{\wedge}\,a\,Q_1)\,(Seq^{\wedge}\,a\,Q_2)}
-%\]
-%as
-%\begin{gather*}
-%\mathsf{
-%SeqLMap\,a\,Q_1\,Q_2\,M\,(Const\,x)\,z = M\,x\,z
-%} \\
-%\mathsf{
-%SeqLMap\,a\,Q_1\,Q_2\,M\,(SPair\,b\,c\,e\,s_b\,s_c)\,(Q_b, Q_c, l_E, l_F) =
-%}
-%\end{gather*}
-%
-%
-%We have an induction rule for $\mathsf{Seq}$
-%\begin{multline}
-%\mathsf{DISeq : \forall (P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set)} \\
-%\mathsf{\to \big( \forall (a : Set) (Q_a : a \to Set) (x : a) \to Q_a\,x \to P\,a\,Q_a\,(Const\,x) \big)} \\
-%\mathsf{\to \big( \forall (a : Set) (b : Set) (c : Set)
-%	(Q_a : a \to Set) (Q_b : b \to Set) (Q_c : c \to Set)} \\
-%	\mathsf{( (s_b, s_c) : F\,Seq\,b\,c) (e : Equal\,a\,(Pair\,b\,c))
-%	\to Equal^{\wedge} a\, (Pair\,b\,c)\, Q_a\, (Pair^{\wedge}\,b\,c\,Q_b\,Q_c)\, e} \\
-%	\mathsf{\to F^{\wedge}\,G\,b\,c\,P\,Q_b\,Q_c\,(s_b,s_c)
-%	\to P a Q_a ( SPair\,b\,c\,e\,(s_b,s_c) ) \big)} \\
-%\mathsf{\to \forall (a : Set)(Q_a : a \to Set)(s_a : Seq\,a) \to Seq^{\wedge}\,a\,Q_a\,s_a \to P\,a\,Q_a\,s_a}
-%\end{multline}
-%We define $\mathsf{DISeq\,P\,PConst\,PSPair\,a\,Q_a\,s_a\,l_a}$,
-%where
-%$\mathsf{P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set}$,
-%$\mathsf{Q_a : a \to Set}$, $\mathsf{s_a : Seq\,a}$
-%and $\mathsf{l_a : Seq^{\wedge}\,a\,Q\,s_a}$,
-%by pattern-matching on $\mathsf{s_a}$.
-%Let $\mathsf{s_a = Const\,x}$ for $\mathsf{x : a}$, and define
-%\[
-%\mathsf{DISeq\,P\,PConst\,PSPair\,a\,Q_a\,(Const\,x)\,l_a = PConst\,a\,Q_a\,x\,l_a}
-%\]
-%Notice that $\mathsf{Seq^{\wedge}\,a\,Q_a\,(Const\,x) = Q_a\,x}$, and thus $\mathsf{l_a : Q_a\,x}$,
-%making the right-hand-side in the above expression type-check.
-%Now, let $\mathsf{s_a = SPair\,b\,c\,e\,(s_b,s_c)}$
-%for $\mathsf{e : Equal\,a\,(Pair\,b\,c)}$ and $\mathsf{(s_b, s_c) : F\,Seq\,b\,c}$,
-%and define
-%\[
-%\mathsf{DISeq\,P\,PConst\,PSPair\,a\,Q_a\,(SPair\,b\,c\,e\,(s_b,s_c))\,(Q_b, Q_c, l_e, l_{F_1}, l_{F_2})
-%	= PSPair\,a\,b\,c\,Q_a\,Q_b\,Q_c\,(s_b,s_c)\,e\,l_e\,p}
-%\]
-%where
-%$\mathsf{p : F^{\wedge}\,G\,b\,c\,P\,Q_b\,Q_c\,(s_b,s_c)}$,
-%$\mathsf{l_e : Equal^{\wedge}\,a\, (Pair\,b\,c)\, Q_a\, (Pair^{\wedge}\,Q_b\,Q_c) \, e}$,
-%$\mathsf{l_{F_1} : F_1^{\wedge}\,Seq\,b\,c\,Seq^{\wedge}\,Q_b\,Q_c\,s_b}$,
-%$\mathsf{l_{F_2} : F_2^{\wedge}\,Seq\,b\,c\,Seq^{\wedge}\,Q_b\,Q_c\,s_c}$
-%(notice that $\mathsf{(Q_b, Q_c, l_e, l_{F_1}, l_{F_2}) : Seq^{\wedge}\,a\,Q\,(SPair\,b\,c\,e\,x\,y)}$,
-%making the right-hand-side in the above expression type-check)
-%and $\mathsf{p}$ is defined as follows:
-%\[
-%\mathsf{p = (???, ???) 
-%: F_1^{\wedge}\,G\,b\,c\,P\,Q_b\,Q_c\,s_b \times F_2^{\wedge}\,G\,b\,c\,P\,Q_b\,Q_c\,s_c} \\
-%\]
-
-%\begin{align*}
-%\mathsf{DISeq\,P\,PConst\,PSPair\,a\,Q\,(Const\,x)\,l}
-%	&=\mathsf{PConst\,a\,Q\,x\,l} \\
-%\mathsf{DISeq\,P\,PConst\,PSPair\,a\,Q_a\,(SPair\,b\,c\,e\,x\,y)\,(Q_b\,Q_c\,e', l_b, l_c)}
-%	&=\mathsf{PSpair\,a\,b\,c\,Q_a\,Q_b\,Q_c\,x\,y\,e\,p_b\,p_c\,e'}
-%\end{align*}
-%where
-%$\mathsf{Q_a : a \to Set}$, $\mathsf{Q_b : b \to Set}$, $\mathsf{Q_c : c \to Set}$,
-%$\mathsf{x : Seq\,b}$, $\mathsf{y : Seq\,c}$,
-%$\mathsf{e : Equal\,a\,(b \times c)}$,
-%$\mathsf{p_b : P\,b\,Q_b\,x}$, $\mathsf{p_c : P\,c\,Q_c\,y}$,
-%$\mathsf{e' : Equal^{\wedge}\,a\, (b \times c)\, Q\, (Q_b \times Q_c) \, e}$,
-%$\mathsf{l_b : Seq^{\wedge}\,b\,Q_b\,x}$,
-%$\mathsf{l_c : Seq^{\wedge}\,c\,Q_c\,y}$
-%and
-%\begin{align*}
-%\mathsf{p_b} &=\mathsf{DISeq\,P\,PConst\,PSPair\,b\,Q_b\,x\,l_b : P\,b\,Q_b\,x} \\
-%\mathsf{p_c} &=\mathsf{DISeq\,P\,PConst\,PSPair\,c\,Q_c\,y\,l_c : P\,c\,Q_c\,y}
-%\end{align*}
-%
-%\begin{align*}
-%\mathsf{Q_a} &: \mathsf{a \to Set} \\
-%\mathsf{Q_b} &: \mathsf{b \to Set} \\
-%\mathsf{Q_c} &: \mathsf{c \to Set} \\
-%\mathsf{x} &: \mathsf{Seq\,b} \\
-%\mathsf{y} &: \mathsf{Seq\,c} \\
-%\mathsf{e} &: \mathsf{Equal\,a\,(b \times c)} \\
-%\mathsf{p_b} &: \mathsf{P\,b\,Q_b\,x} \\
-%\mathsf{p_c} &: \mathsf{P\,c\,Q_c\,y} \\
-%\mathsf{e'} &: \mathsf{Equal^{\wedge}\,a\, (b \times c)\, Q\, (Q_b \times Q_c) \, e} \\
-%\mathsf{l_b} &: \mathsf{Seq^{\wedge}\,b\,Q_b\,x} \\
-%\mathsf{l_c} &: \mathsf{Seq^{\wedge}\,c\,Q_c\,y} \\
-%\mathsf{p_b = DISeq\,P\,PConst\,PSPair\,b\,Q_b\,x\,l_b} &: \mathsf{P\,b\,Q_b\,x} \\
-%\mathsf{p_c = DISeq\,P\,PConst\,PSPair\,c\,Q_c\,y\,l_c} &: \mathsf{P\,c\,Q_c\,y}
-%\end{align*}
-
 \subsection{Third example introduced above}
@@
 
-\subsection{Induction for GADTs with nesting}\label{sec:GADT-nested}
+\subsection{General case}
+
+%We restrict our scope to GADTs not featuring nesting,
+%in the sense that their constructors contain no nested calls of the recursive variable,
+%as truly nested types do.
+%This way we can avoid the issue described in Section~\ref{sec:GADT-nested}.
+%The reason for this restriction will be discussed in Section~\ref{sec:GADT-nested}.
+
+Finally, we generalize the approach taken in the previous examples
+and provide a general framework to derive induction rules for arbitrary GADTs.
+%Since, as seen in Section~\ref{sec:GADT-nested},
+%GADTs featuring nesting do not admit an induction rule,
+%we restrict our scope to GADTs not featuring nesting.
+For that, we need to give a grammar for the types we will be considering.
+A generic GADT
+\begin{equation*}
+\begin{array}{l}
+\mathsf{data\, G\,(\ol{a : Set}) : Set\,where}\\
+\mathsf{\;\;\;\;\;\;\;\;C_i\, :\, F_i\,G\,\ol{b} \to G (\ol{K_i\,\ol{b}})}
+\end{array}
+\end{equation*}
+is defined by a finite number of constructors $\mathsf{C_i}$.
+In the definition above, $\mathsf{F_i}$ is a type constructor with signature $\mathsf{(Set^{\alpha} \to Set) \to Set^{\beta} \to Set}$
+and each $\mathsf{K_i}$ is a type constructor with signature $\mathsf{Set^{\beta} \to Set}$
+(i.e. a type constructor of arity $\mathsf{\beta}$).
+The overline notation denotes a finite list:
+$\mathsf{\ol{a}}$ is a list of types of length $\mathsf{\alpha}$,
+so that it can be applied to the type constructor $\mathsf{G}$ of arity $\mathsf{\alpha}$.
+%and the vertical bars notation denotes the length of a list:
+%$\mathsf{G}$ depends on a finite list of types $\mathsf{\ol{a}}$ of length $\mathsf{\alpha}$,
+Each of the $\mathsf{\alpha}$-many $\mathsf{K_i}$ is a type constructor of arity $\mathsf{\beta}$
+so that it can be applied to the list of types $\mathsf{\ol{b}}$ of length $\mathsf{\beta}$.
+%depends on a finite list of types $\mathsf{\ol{b}}$ of length $\mathsf{\beta}$,
+Moreover, notice that the arity of $\mathsf{G}$ matches the number of type constructors $\mathsf{\ol{K_i}}$.
+We allow each $\mathsf{F_i}$ to be inductively built in the following ways (and with the following restrictions):
+\begin{itemize}
+\item $\mathsf{F_i = F_i' \times F_i''}$ where $\mathsf{F_i'}$ and $\mathsf{F_i''}$ have the same signature as $\mathsf{F_i}$ and are built recursively from the same induction rules.
+\item $\mathsf{F_i = F_i' + F_i''}$ where $\mathsf{F_i'}$ and $\mathsf{F_i''}$ have the same signature as $\mathsf{F_i}$ and are built recursively from the same induction rules.
+\item $\mathsf{F_i = F_i' \to F_i''}$
+where $\mathsf{F_i'}$ does not contain the recursive variable,
+i.e., $\mathsf{F_i' : Set^\beta \to Set}$ is a type constructor of arity $\mathsf{\beta}$,
+and $\mathsf{F_i''}$ has the same signature as $\mathsf{F_i}$ and is built recursively from the same induction rules.
+\item $\mathsf{F_i\,G\,\ol{b} = G(\ol{F_a\,\ol{b}})}$
+where none of the $\mathsf{F_a}$ contains the recursive variable,
+i.e., $\mathsf{F_a : Set^\beta \to Set}$
+is a type constructor of arity $\mathsf{\beta}$
+for each $\mathsf{a}$.
+Such restriction is necessary to prevent nesting,
+as that would break the induction rule as discussed in Section~\ref{sec:GADT-nested}.
+\item $\mathsf{F_i\,G\,\ol{b} = H\,\ol{b}}$
+where $\mathsf{H}$ is a type constructor of arity $\mathsf{\beta}$ not containing the recursive variable,
+i.e., $\mathsf{H : Set^\beta \to Set}$.
+Notice that this covers the case in which $\mathsf{F_i}$ is a closed type,
+so, in particular, the unit and empty types, $\mathsf{1}$ and $\mathsf{0}$.
+%The only requirement on $\mathsf{H}$ is that it needs to have a predicate-lifting function
+%(and any type made of constant types, sums, products, arrow types and other GADTs satisfies this condition).
+\item $\mathsf{F_i\,G\,\ol{b} = H (\ol{F_c\,G\,\ol{b}})}$
+where $\mathsf{H}$ is a $\mathsf{\gamma}$-ary type constructor not containing the recursive variable,
+i.e., $\mathsf{H : Set^\gamma \to Set}$,
+and $\mathsf{F_c}$ has the same signature as $\mathsf{F_i}$  and is built recursively from the same induction rules, for every $\mathsf{c = 1 \dots \gamma}$.
+Moreover, we require that $\mathsf{H}$ is not a GADT itself (but we allow it to be an ADT or even a nested type).
+This way $\mathsf{H}$ admits functorial semantics, and thus we have the map function of $\mathsf{H^{\wedge}}$,
+\[
+\mathsf{HLMap : \forall (\ol{c : Set}) (\ol{Q_c\;Q_c' : c \to Set})
+	\to \ol{PredMap\,c\,Q_c\,Q_c'}
+	\to PredMap\,(H\,\ol{c})\,(H^{\wedge}\,\ol{c}\,\ol{Q_c})\,(H^{\wedge}\,\ol{c}\,\ol{Q_c'})}
+\]
+where $\mathsf{PredMap : \forall (c : Set) \to (c \to Set) \to (c \to Set) \to Set}$
+is defined as
+\[
+\mathsf{PredMap\,c\,Q_c\,Q_c' = \forall (x : c) \to Q_c\,x \to Q_c'\,x}
+\]
+and represents the type of morphisms between predicates.
+\end{itemize}
+We can summarize the above inductive definition with the following grammar
+(but beware that the above restrictions and requirements still apply):
+\[
+\mathsf{F_i\,G\,\ol{b} :=
+F_i'\,G\,\ol{b} \times F_i''\,G\,\ol{b}
+\ \vert\ F_i'\,G\,\ol{b} + F_i''\,G\,\ol{b}
+\ \vert\ F_i'\,\ol{b} \to F_i''\,G\,\ol{b}
+\ \vert\ G(\ol{F_a\,\ol{b}})
+\ \vert\ H\,\ol{b}
+\ \vert\ H (\ol{F_c\,G\,\ol{b}})
+}
+\]
+A further requirement that applies to all of the types appearing above,
+including the types $\mathsf{K_i}$,
+is that every type needs to have a predicate-lifting function.
+This is not an overly restrictive condition, though:
+all types made by sums, products, arrow types and type application do, and so do GADTs as defined above.
+A way to concretely define the predicate-lifting function for a type is to proceed by induction on the structure of the type,
+and we have seen in the previous sections examples of how to do so for products and type application.
+We do not give here the general definition of lifting,
+as that would require to first present a full type calculus,
+and that is beyond the scope of the paper.
+
+%Before proceeding with the derivation of the induction rule for a generic GADT,
+%we need to define the type
+%\[
+%\mathsf{PredMap\,a\,Q_a\,Q_a' = \forall (x : a) \to Q_a\,x \to Q_a'\,x}
+%\]
+%where $\mathsf{Q_a}$ and  $\mathsf{Q_a'}$ are predicates on  $\mathsf{a : Set}$,
+%which represents the type of morphisms between predicates $\mathsf{Q_a}$ and  $\mathsf{Q_a'}$.
+%Moreover, if $\mathsf{H}$ is a type constructor having an associated predicate-lifting function $\mathsf{H^{\wedge}}$,
+%we define a map function $\mathsf{HLMap}$ for the lifting $\mathsf{H^{\wedge}}$ as
+%\[
+%\mathsf{HLMap : \forall (a : Set) (Q_a\;Q_a' : a \to Set) \to PredMap\,a\,Q_a\,Q_a' \to PredMap\,(G\,a)\,(G^{\wedge}\,a\,Q_a)\,(G^{\wedge}\,a\,Q_a')}
+%\]
+
+Consider a generic GADT as defined above,
+\begin{equation}\label{eq:gadt}
+\begin{array}{l}
+\mathsf{data\, G\,(a : Set) : Set\,where}\\
+\mathsf{\;\;\;\;\;\;\;\;C\, :\, F\,G\,\ol{b} \to G (K\,\ol{b})}
+\end{array}
+\end{equation}
+which, for ease of notation,
+we assume to be a unary type constructor (i.e. it depends on a single type parameter $\mathsf{a}$)
+and to have only one constructor $\mathsf{C}$.
+Extending the argument to GADTs of arbitrary arity and with multiple constructors presents no difficulty
+other than heavier notation.
+In the definition above, $\mathsf{F}$ has signature $\mathsf{(Set \to Set) \to Set^{\beta} \to Set}$
+and each $\mathsf{K}$ has signature $\mathsf{Set^{\beta} \to Set}$.
+The constructor $\mathsf{C}$ can be rewritten using the $\mathsf{Equal}$ type as
+\[
+\mathsf{C : \exists (\ol{b : Set}) \to Equal\,a\,(K\,\ol{b}) \to F\,G\,\ol{b} \to G\,a}
+\]
+which is the form we shall use from now on.
+
+In order to state the induction rule for $\mathsf{G}$,
+we first need to define $\mathsf{G}$'s associated predicate-lifting function 
+\[
+\mathsf{G^{\wedge} : \forall (a : Set) \to (a \to Set) \to G\,a \to Set}
+\]
+as
+\[
+\mathsf{G^{\wedge}\,a\,Q_a\,(C\,\ol{b}\,e\,x)
+= \exists (\ol{Q_b : b \to Set})
+\to Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e
+\times F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,{\ol{Q_b}}\,x}
+\]
+where $\mathsf{Q_a : a \to Set}$
+and $\mathsf{C\,\ol{b}\,e\,x : G\,a}$,
+i.e., $\mathsf{e : Equal\,a\,(K\,\ol{b})}$ and $\mathsf{x : F\,G\,\ol{b}}$.
+As already mentioned before, we also assume to have liftings for $\mathsf{F}$,
+\begin{multline*}
+\mathsf{F^{\wedge} : \forall (G : Set^{\alpha} \to Set) (\ol{b : Set})
+\to (\forall (a : Set) \to (a \to Set) \to G\,a \to Set)} \\
+\mathsf{\to (\ol{b \to Set})
+\to F\,G\,\ol{b} \to Set}
+\end{multline*}
+and for  $\mathsf{K}$,
+\[
+\mathsf{K^{\wedge} : \forall (\ol{b : Set}) \to (\ol{b \to Set}) \to K\,\ol{b} \to Set}
+\]
+Finally, associate the function
+\begin{multline*}
+\mathsf{CC = \lambda (P : \forall (a : Set) \to (a \to Set) \to G\,a \to Set)} \\
+\mathsf{\to \forall (a : Set) (\ol{b : Set}) (Q_a : a \to Set) (\ol{Q_b : b \to Set}) (e : Equal\,a\,(K\,\ol{b})) (x : F\,G\,\ol{b})} \\
+\mathsf{\to Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e
+	\to F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x
+	\to P\,a\,Q_a\,(C\,\ol{b}\,e\,x)}
+\end{multline*}
+to the constructor $\mathsf{C}$.
+
+%We can now define the induction rule for $\mathsf{G}$
+The induction rule for $\mathsf{G}$ is
+\begin{multline*}
+\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to G\,a \to Set) \to CC\,P} \\
+%\mathsf{\to \Big(
+%\forall (\ol{a : Set}) (\ol{b : Set}) (\ol{Q_a : a \to Set}) (\ol{Q_b : b \to Set}) (\ol{e_a : Equal\,a\,(K\,\ol{b})}) (x : F\,G\,\ol{b})} \\
+%\mathsf{\to \ol{Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
+%	\to F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x
+%	\to P\,\ol{a}\,\ol{Q_a}\,(C\,\ol{b}\,\ol{e_a}\,x)
+%\Big)} \\
+\mathsf{\to \forall (a : Set)(Q_a : a \to Set)(y : G\,a) \to G^{\wedge}\,a\,Q_a\,y \to P\,a\,Q_a\,y}
+\end{multline*}
+As we already did in the previous examples,
+we validate the induction rule by providing a term $\mathsf{DIG}$ for the type above.
+Define
+\[
+\mathsf{DIG\,P\,cc\,a\,Q_a\,(C\,\ol{b}\,e\,x)\,(\ol{Q_b}, L_E, L_F)
+= cc\,a\,\ol{b}\,Q_a\,\ol{Q_b}\,e\,x\,L_E\,(p\,x\,L_F)}
+\]
+where $\mathsf{cc : CC\,P}$ and
+\begin{itemize}
+\item $\mathsf{C\,\ol{b}\,e\,x : G\,a}$,
+i.e., $\mathsf{e : Equal\,a\,(K\,\ol{b})}$,
+and $\mathsf{x : F\,G\,\ol{b}}$;
+\item $\mathsf{(\ol{Q_b}, L_E, L_F) : G^{\wedge}\,a\,Q_a (C\,\ol{b}\,e\,x)}$,
+i.e.,
+$\mathsf{Q_b : b \to Set}$ for each $\mathsf{b}$,
+$\mathsf{L_E : Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e}$,
+and $\mathsf{L_F : F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,{\ol{Q_b}}\,x}$.
+\end{itemize}
+Finally, the morphism of predicates
+\[
+\mathsf{p : PredMap\,(F\,G\,\ol{b})\,(F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}) (F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})}
+\]
+is defined by structural induction on $\mathsf{F}$ as follows:
+\begin{itemize}
+\item Case $\mathsf{F = F_1 \times F_2}$ where $\mathsf{F_1}$ and $\mathsf{F_2}$ have the same signature as $\mathsf{F}$.
+We have that
+\[
+\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}
+= Pair^{\wedge}(F_1\,G\,\ol{b})(F_2\,G\,\ol{b})(F_1^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}) (F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})}
+\]
+By inductive hypothesis, there exist morphisms of predicates
+\begin{gather*}
+\mathsf{p_1 : PredMap\,(F_1\,G\,\ol{b})\,((F_1)^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}) ((F_1)^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})} \\
+\mathsf{p_2 : PredMap\,(F_2\,G\,\ol{b})\,((F_2)^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}) ((F_2)^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})}
+\end{gather*}
+Thus, we define $\mathsf{p (x_1, x_2) (L_1, L_2) = (p_1\,x_1\,L_1, p_2\,x_2\,L_2)}$ 
+for 
+$\mathsf{x_1 : F_1\,G\,\ol{b}}$,
+$\mathsf{L_1 : F_1^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x_1}$,
+$\mathsf{x_2 : F_2\,G\,\ol{b}}$
+and $\mathsf{L_2 : F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x_2}$.
+%%%
+\item Case $\mathsf{F = F_1 + F_2}$ where $\mathsf{F_1}$ and $\mathsf{F_2}$ have the same signature as $\mathsf{F}$.
+Analogous to case $\mathsf{F = F_1 \times F_2}$.
+%%%
+\item Case $\mathsf{F = F_1 \to F_2}$
+where $\mathsf{F_1}$ does not contain the recursive variable,
+i.e., $\mathsf{F_1 : Set^\beta \to Set}$,
+and $\mathsf{F_2}$ has the same signature as $\mathsf{F}$.
+We have that
+\[
+\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x = \forall (z : F_1\,\ol{b}) \to F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,z \to F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,(x\,z)}
+\]
+where $\mathsf{x : F\,G\,\ol{b} = F_1\,\ol{b} \to F_2\,G\,\ol{b}}$.
+By inductive hypothesis, there exist a morphism of predicates
+\[
+\mathsf{p_2 : PredMap\,(F_2\,G\,\ol{b}) (F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}) (F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})}
+\]
+Thus, we define $\mathsf{p\,x\,L_F : F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}$
+for $\mathsf{L_F : F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x}$
+as $\mathsf{p\,x\,L_F\,z\,L_1 = p_2 (x\,z) (L_F\,z\,L_1)}$
+for $\mathsf{z : F_1\,\ol{b}}$
+and $\mathsf{L_1 : F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,z}$.
+Notice that $\mathsf{F_1}$ not containing the recursive variable
+is a necessary restriction, as the proof relies on
+$\mathsf{F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x}$
+and $\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}$
+having the same domain $\mathsf{F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,z}$.
+%%%
+\item Case $\mathsf{F\,G\,\ol{b} = G(F'\,\ol{b})}$
+where $\mathsf{F'}$ does not contain the recursive variable,
+i.e., $\mathsf{F' : Set^\beta \to Set}$.
+Thus, $\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b} = P (F'\,\ol{b}) (F'^{\wedge}\,\ol{b}\,\ol{Q_b})}$.
+So, $\mathsf{p}$ is defined as
+\[
+\mathsf{p = DIG\,P\,cc\,(F'\,\ol{b})\,(F'^{\wedge}\,\ol{b}\,\ol{Q_b})}
+\]
+%for $\mathsf{L_F : F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x}$.
+%Note that the fact that the type constructor $\mathsf{F'}$ does not contain the recursive variable
+%means that $\mathsf{G}$ does not feature nesting.
+%As explained and exemplified in Section~\ref{sec:GADT-nested},
+%that assumption is required to make induction work.
+%%%
+\item Case $\mathsf{F\,G\,\ol{b} = H\,\ol{b}}$
+where $\mathsf{H}$ is a $\mathsf{\beta}$-ary type constructor not containing the recursive variable,
+i.e., $\mathsf{H : Set^\beta \to Set}$.
+In such case, $\mathsf{p : PredMap (H\,\ol{b}) (H^{\wedge}\,\ol{b}\,\ol{Q_b}) (H^{\wedge}\,\ol{b}\,\ol{Q_b})}$
+is just the identity morphism of predicates.
+%%%
+\item Case $\mathsf{F\,G\,\ol{b} = H (\ol{F_c\,G\,\ol{b}})}$
+where $\mathsf{H}$ is a $\mathsf{\gamma}$-ary type constructor not containing the recursive variable,
+i.e., $\mathsf{H : Set^\gamma \to Set}$,
+and $\mathsf{F_c}$ has the same signature as $\mathsf{F}$, for every $\mathsf{c = 1 \dots \gamma}$.
+Moreover, we assume that $\mathsf{H}$ has an associated predicate-lifting function,
+\[
+\mathsf{H^{\wedge} : \forall (\ol{c : Set}) \to (\ol{c \to Set}) \to H\,\ol{c} \to Set}
+\]
+and that this predicate-lifting function has a map function $\mathsf{HLMap}$ of type
+\[
+\mathsf{\forall (\ol{c : Set}) (\ol{Q_c\;Q_c' : c \to Set}) \to PredMap\,\ol{c}\,\ol{Q_c}\,\ol{Q_c'}
+\to PredMap\,(H\,\ol{c})\,(H^{\wedge}\,\ol{c}\,\ol{Q_c})\,(H^{\wedge}\,\ol{c}\,\ol{Q_c'})}
+\]
+That means that $\mathsf{H}$ cannot be a GADT,
+as GADTs have no functorial semantics~\cite{HaskellPaper}
+and incur in the issue exposed in Section~\ref{sec:GADT-nested},
+but it can be an ADT or even a nested type as those types have functorial semantics~\cite{jp19,jgj21}.
+Thus,
+\[
+\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b} = H^{\wedge} (\ol{F_c\,G\,\ol{b}}) (\ol{F_c^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}})}
+\]
+By induction hypothesis, there is a morphism of predicates
+\[
+\mathsf{p_c : PredMap\,(F_c\,G\,\ol{b}) (F_c^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}) (F_c^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})}
+\]
+for every $\mathsf{c = 1 \dots \gamma}$.
+So, $\mathsf{p}$ is defined as
+\[
+\mathsf{p
+= HLMap\,(\ol{F_c\,G\,\ol{b}}) (\ol{F_c^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}}) (\ol{F_c^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}})\,\ol{p_c}}
+\]
+\end{itemize}
+%Notice that, by proceeding by structural induction as above,
+%we have implicitly defined a grammar for GADTs admitting an induction rule.
+%In particular, we allow $\mathsf{F}$ to feature sums, products, arrow types and (limited) type application.
+
+\section{Induction for GADTs with nesting}\label{sec:GADT-nested}
 

### 2021-04-06 induction-paper/GADT-induction.tex
Revise Section 5

@@
 and \emph{truly} nested types, such as the nested type
-\begin{equation}\label{eq:ptree}
+\begin{equation}\label{eq:bush}
 \begin{array}{l}
@@
 For example, the induction principle for $\mathsf{List}$ is
-\begin{multline}
+\begin{multline*}
 \mathsf{\forall (a : Set) (P : List\,a \to Set) \to P\,Nil \to \big( \forall (x : a) (xs: List\,a) \to P\,xs \to P\,(Cons\,x\,xs)\big)} \\
 \mathsf{\to \forall (xs : List\,a) \to P\, xs }
-\end{multline}
+\end{multline*}
 Note that the data inside a structure of type $\mathsf{List}$ is treated monolithically (i.e., ignored) by this induction rule.
@@
 Thus, the induction rule for $\mathsf{PTree}$ is
-\begin{multline}
+\begin{multline*}
 \mathsf{\forall (P : \forall (a : Set) \to PTree\, a \to Set)
@@
 \mathsf{\to \forall (a : Set) (x : PTree\,a) \to P\,a\,x }
-\end{multline}
+\end{multline*}
 
@@
 For example, the deep induction rule for $\mathsf{Bush}$ is
-\begin{multline}
+\begin{multline*}
 \mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to Bush\, a \to Set)
@@
 \mathsf{\to \forall (a : Set) (Q : a \to Set) (x : Bush\,a) \to Bush^{\wedge}\,a\,Q\,x \to P\,a\,Q\,x }
-\end{multline}
+\end{multline*}
 where $\mathsf{Bush^{\wedge} : \forall (a : Set) \to (a \to Set) \to Bush\,a \to Set}$ lifts a predicate $\mathsf{Q}$ on data of type $\mathsf{a}$ to a predicate on data of type $\mathsf{Bush\,a}$ asserting that $\mathsf{Q}$ holds for every element of type $\mathsf{a}$ contained in its argument bush.
@@
 For example, the deep induction principle for $\mathsf{List}$ is
-\begin{multline}
+\begin{multline*}
 \mathsf{\forall (a : Set) (P : List\, a \to Set) (Q : a \to Set)} \\
@@
 \mathsf{\to \forall (xs : List\,a) \to List^{\wedge}\,a\,Q\,xs \to P\, xs }
-\end{multline}
+\end{multline*}
 where $\mathsf{List^{\wedge} : \forall (a : Set) \to (a \to Set) \to List\,a \to Set}$ lifts a predicate $\mathsf{Q}$ on data of type $\mathsf{a}$ to a predicate on data of type $\mathsf{List\,a}$ asserting that $\mathsf{Q}$ holds for every element of its argument list.
 Finally, the deep induction rule for $\mathsf{PTree}$ is
-\begin{multline}
+\begin{multline*}
 \mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to PTree\, a \to Set)} \\
@@
 \mathsf{\to \forall (a : Set) (Q : a \to Set) (x : PTree\,a) \to PTree^{\wedge}\,a\,Q\,x \to P\,a\,Q\,x }
-\end{multline}
+\end{multline*}
 where $\mathsf{Pair^{\wedge} : \forall (a : Set) \to (a \to Set) \to a \times a \to Set}$ lifts a predicate $\mathsf{Q}$ on $\mathsf{a}$ to a predicate on pairs of type $\mathsf{a \times a}$,
@@
 Moreover, we require that $\mathsf{H}$ is not a GADT itself (but we allow it to be an ADT or even a nested type).
-This way $\mathsf{H}$ admits functorial semantics, and thus we have the map function of $\mathsf{H^{\wedge}}$,
+This way $\mathsf{H}$ admits functorial semantics~\cite{jp20}, and thus we have a map function for $\mathsf{H^{\wedge}}$,
 \[
@@
 and represents the type of morphisms between predicates.
+A concrete way to define $\mathsf{HLMap}$ is to proceed by induction on the structure of the type $\mathsf{H}$,
+and give an inductive definition when $\mathsf{H}$ is an ADT or a nested type.
+Such details are not essential to the present discussion, and thus we omit them.
 \end{itemize}
@@
 all types made by sums, products, arrow types and type application do, and so do GADTs as defined above.
-A way to concretely define the predicate-lifting function for a type is to proceed by induction on the structure of the type,
+A concrete way to define the predicate-lifting function for a type is to proceed by induction on the structure of the type,
 and we have seen in the previous sections examples of how to do so for products and type application.
@@
 in the sense that their constructors contain no nested calls of the recursive variable,
-as truly nested types do.
-Since both nested types~\cite{jp20}
+as truly nested types (such as $\mathsf{Bush}$, Equation~\ref{eq:bush}) do.
+Since both nested types
 and GADTs without nesting admit induction rules,
+as seen in the previous sections,
 it is just natural to expect that GADTs with nesting would as well.
 Surprisingly, that is not the case:
-indeed, induction rules for nested types rely on functorial semantics,
+indeed, the induction principle generally relies on (unary) parametricity of the semantic interpretation,
+and in the case of nested types it also relies on functorial semantics~\cite{jp20},
 but GADTs cannot admit both functorial and parametric semantics at the same time~\cite{HaskellPaper}.
@@
 by analyzing the following concrete example of such a type.
-\begin{equation}
+\begin{equation}\label{gadt-nested}
 \begin{array}{l}
@@
 \[
-\mathsf{G^{\wedge} : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set}
+\mathsf{G^{\wedge} : \forall (a : Set) \to (a \to Set) \to G\,\ol{a} \to Set}
 \]
@@
 where $\mathsf{Q_a : a \to Set}$, $\mathsf{e : Equal\,a\,(b \times b)}$ and $\mathsf{x : G(G\,b)}$.
-
-Finally, let $\mathsf{CC}$ be the type
-\begin{multline*}
-\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to G\,a \to Set)} \\
+Finally, let $\mathsf{CC}$ be the function
+\begin{multline*}
+\mathsf{\lambda (P : \forall (a : Set) \to (a \to Set) \to G\,a \to Set)} \\
 \mathsf{\to \forall (a\;b : Set) (Q_a : a \to Set) (Q_b : b \to Set) (e : Equal\,a\,(b \times b)) (x : G\,(G\,b))} \\
@@
 \mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to G\,a \to Set) \to CC\,P} \\
-%\mathsf{\to \Big( \forall (a\;b : Set) (Q_a : a \to Set) (Q_b : b \to Set) (e : Equal\,a\,(b \times b)) (x : G\,(G\,b))} \\
-%\mathsf{\to Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e
-%	\to P\,(G\,b)\,(P\,b\,Q_b)\,x
-%	\to P\,a\,Q_a\,(C\,b\,e\,x) \Big)}\\
 \mathsf{\to \forall (a : Set)(Q_a : a \to Set)(y : G\,a) \to G^{\wedge}\,a\,Q_a\,y \to P\,a\,Q_a\,y}
@@
 Consistently with the previous examples,
-we define a term validating the induction rule, $\mathsf{DIG}$, as
+to validate the induction rule
+we try to define a term of the above type, $\mathsf{DIG}$, as
 \[
@@
 \]
-where $\mathsf{cc : CC\,P}$,
-$\mathsf{C\,b\,e\,x : G\,a}$,
-%, the second to last term applied to $\mathsf{DIG}$,
-%has type $\mathsf{G\,a}$,
-i.e., $\mathsf{e : Equal\,a\,(b \times b)}$ and $\mathsf{x : G(G\,b)}$,
-and $\mathsf{(Q_b, L_E, L_G) : G^{\wedge}\,a\,Q_a\,(C\,b\,e\,x)}$,
-%, the last term applied to $\mathsf{DIG}$,
-%has type $\mathsf{G^{\wedge}\,a\,Q_a\,(C\,b\,e\,x)}$,
+where $\mathsf{cc : CC\,P}$ and
+\begin{itemize}
+\item $\mathsf{C\,b\,e\,x : G\,a}$,
+i.e., $\mathsf{e : Equal\,a\,(b \times b)}$ and $\mathsf{x : G(G\,b)}$;
+\item $\mathsf{(Q_b, L_E, L_G) : G^{\wedge}\,a\,Q_a\,(C\,b\,e\,x)}$,
 i.e., $\mathsf{Q_b : b \to Set}$, $\mathsf{L_E : Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e}$,
 and $\mathsf{L_G : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,x}$.
-%\begin{gather*}
-%\mathsf{Q_b : b \to Set} \\
-%\mathsf{L_E : Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e} \\
-%\mathsf{L_G : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,x}
-%%\mathsf{p = DIG\,P\,pc\,(G\,b)\,(P\,b\,Q_b)\,x\,q : P\,(G\,b)\,(P\,b\,Q_b)\,x}
-%\end{gather*}
+\end{itemize}
 We still need to define $\mathsf{p : P\,(G\,b)\,(P\,b\,Q_b)\,x}$.
@@
 where we still need to provide $\mathsf{q : G^{\wedge}\,(G\,b)\,(P\,b\,Q_b)\,x}$.
-To produce such $\mathsf{q}$, we need the map function of $\mathsf{G^{\wedge}}$,
+If we had the map function of $\mathsf{G^{\wedge}}$,
 \[
@@
 \]
-where $\mathsf{PredMap : \forall (a : Set) \to (a \to Set) \to (a \to Set) \to Set}$
-is defined as
-\[
-\mathsf{PredMap\,a\,Q_a\,Q_a' = \forall (x : a) \to Q_a\,x \to Q_a'\,x}
-\]
-and represents the type of morphisms between predicates.
-If we had $\mathsf{GLMap}$, then we would be able to define
+then we would be able to define
 \[
@@
 \end{itemize}
-$\mathsf{(Q_b, L_E, L_G)}$ has type $\mathsf{G^{\wedge}\,a\,Q_a\,(C\,b\,e\,x)}$, i.e.,
+$\mathsf{(Q_b, L_E, L_G) :G^{\wedge}\,a\,Q_a\,(C\,b\,e\,x)}$, i.e.,
 \begin{itemize}
@@
 \end{itemize}
-and $\mathsf{(Q_b', L_E', L_G')}$ has type $\mathsf{G^{\wedge}\,a\,Q_a'\,(C\,b\,e\,x)}$, i.e.,
+and $\mathsf{(Q_b', L_E', L_G') : G^{\wedge}\,a\,Q_a'\,(C\,b\,e\,x)}$, i.e.,
 \begin{itemize}
@@
 This is because GADTs can either have a syntax-only semantics or a functorial-completion semantics.
-Since we are interested in induction rules, we considered GADTs with their syntax-only semantics,
-which is parametric but it is not functorial.
-Had we considered the functorial-completion semantics, instead,
-we would have forfeited parametricity~\cite{HaskellPaper}.
-In both cases, thus, we cannot derive an induction rule for generic GADTs when they feature nesting.
+Since we are interested in induction rules, we considered the syntax-only semantics,
+which is parametric but not functorial.
+Had we considered the functorial-completion semantics, which is functorial,
+we would have forfeited parametricity instead.
+In both cases, thus, we cannot derive an induction rule for GADTs featuring nesting.
+Unlike nestes types, indeed, GADTs do not admit a semantic interpretation that is both parametric and functorial~\cite{HaskellPaper}.
 

### 2021-04-06 induction-paper/GADT-induction.tex
add lambda terms example

--- 2021-04-05 induction-paper/GADT-induction.tex
@@
 
-Add third example here.
-
 %{\color{blue} Do we really need the existential quantifiers?}
@@
 %\end{array}\]
+
+{\color{red} STLC example}
+
+GADTs are expressive enough to represent, for example, the types and terms of the simply typed lambda calculus. 
+We will use a variant of the STLC including a base type of Bool, and type constructors for arrow types and list types:
+
+\begin{equation}\label{eq:eq_ltype}
+\begin{array}{l}
+\mathsf{data\, LType\,(a : Set) : Set\,where}\\
+\mathsf{\;\;\;\;\;\;\;\;TBool :\, \exists (b : Set). \,Equal\,a\,(KBool\, b) \to LType\,a}\\
+\mathsf{\;\;\;\;\;\;\;\;TArr\,\,\, :\, \exists (b : Set) (c : Set).\,Equal\,a\,(b \to c) \to LType\,b \to LType\,c \to LType\,a}\\
+  \mathsf{\;\;\;\;\;\;\;\;TList\,\, :\, \exists (b : Set).\,Equal\,a\,(\List\,b) \to LType\,b \to LType\,a}
+\end{array}
+\end{equation}
+
+The term calculus for this type system includes variables, abstraction, application, 
+and also includes an introduction rule that takes a list of terms of type $\mathsf{b}$ and produces a term of type $\mathsf{List\,b}$.
+
+\begin{equation}\label{eq:eq_lterm}
+\begin{array}{l}
+\mathsf{data\, LTerm\,(a : Set) : Set\,where}\\
+\mathsf{\;\;\;\;\;\;\;\;Var\,\,\,\,\,\,:\, String \to LType\,a \to LTerm\,a} \\
+\mathsf{\;\;\;\;\;\;\;\;Abs\,\,\,\,\,\, :\, \exists (b : Set) (c : Set).\,Equal\,a\,(b \to c) \to String 
+      \to LType\,b \to LTerm\,c \to LTerm\,a}\\
+  \mathsf{\;\;\;\;\;\;\;\;App\,\,\,\,\, :\, \exists (b : Set).\,LTerm (b \to a) \to LTerm\,b \to LTerm\,a}\\
+  \mathsf{\;\;\;\;\;\;\;\;ListC\,\, :\, \exists (b : Set).\,Equal\,a\,(\List\,b) \to List\,(LTerm\,b) \to LTerm\,a}
+\end{array}
+\end{equation}
+
+The type parameter for $\mathsf{LTerm}$ tracks the types of STLC terms. So the type $\mathsf{LTerm\,A}$ 
+contains lambda terms of type $\mathsf{A}$. The $\mathsf{LType}$ GADT is used to encode the 
+grammar of permissible types for this calculus.
+
+Variables are represented as strings and are tagged with their type. 
+The $\mathsf{Abs}$ constructor takes a name (string) and a type for the bound variable,
+and a term of type $\mathsf{c}$ for the body of the abstraction.
+The $\mathsf{App}$ constructor applies a term of type $\mathsf{b \to a}$ to a term of type $\mathsf{b}$, producing
+a term of type $\mathsf{a}$. 
+The $\mathsf{ListC}$ constructor takes a list of terms of type $\mathsf{b}$ and produces a term of type $\mathsf{List\,b}$.
 
@@
 
-\subsection{Third example introduced above}
+\subsection{(Deep) induction for $\mathsf{LTerm}$ }
 
@@
 %Still to decide?
+
+{\color{red} Use $\mathsf{Pair^{\wedge}}$ everywhere}
+
+{\color{blue} Some notational conventions: I am using uppercase letters for elements of LType and 
+lowercase letters for elements of LTerm. For example, $\mathsf{(T_b : LType \, b)}$ or $\mathsf{(t_c : LTerm \,c)}$.}
+
+Now we will define an induction rule for the $\mathsf{LTerm}$ type defined in 
+Equation~\ref{eq:eq_lterm}. To define this, we need liftings for both $\mathsf{LType}$ and
+$\mathsf{LTerm}$. We will also need a lifting of predicates for arrow types, since arrow types appear in 
+$\mathsf{LType}$ and $\mathsf{LTerm}$. The lifting for arrow types 
+\[
+  \mathsf{Arr^{\wedge} : \forall (a\, b : Set) \to (a \to Set) \to (b \to Set) \to (a \to b) \to Set}
+\]
+is defined as: 
+\[
+  \mathsf{Arr^{\wedge}\, a\, b\, Q_a\, Q_b\, f = \forall (x : a) \to Q_a\, x \to Q_b\, (f\, x)}
+\]
+
+\noindent
+Now we can define the lifting for $\mathsf{LType}$, which has type 
+\[
+\mathsf{LType^{\wedge} : \forall (a : Set) \to (a \to Set) \to LType\,a \to Set}
+\]
+and is defined by pattern-matching as
+\[
+  \mathsf{LType^{\wedge}\,a\,Q_a\,(TBool\,b\,e) = \exists (Q_b : b \to Set) 
+  \to Equal^{\wedge}\, a\, b\, Q_a\, (KBool^{\wedge} \,b \,Q_b) \,e}
+\]
+where $\mathsf{Q_a : a \to Set}$ and $\mathsf{e : Equal \,a \,(KBool\,b)}$,
+and
+\begin{multline*}
+  \mathsf{LType^{\wedge}\,a\,Q_a\,(TArr\, b\, c\, e\, T_b\, T_c)} \\
+\mathsf{=
+\exists (Q_b : b \to Set) (Q_c : c \to Set)
+  \to Equal^{\wedge}\,a\, (b \to c)\, Q_a\, (Arr^{\wedge} \, b\, c\, Q_b \, Q_c) \, e} \\ 
+\mathsf{
+  \times \, LType^{\wedge}\,b\,Q_b\,T_b \times LType^{\wedge}\,c\,Q_c\,T_c
+}
+\end{multline*}
+
+where $\mathsf{Q_a : a \to Set}$, $\mathsf{e : Equal\, a\, (b \to c)}$, $\mathsf{T_b : LType \,b}$, and 
+$\mathsf{T_c : LType \,c}$,
+and 
+\begin{multline*}
+  \mathsf{LType^{\wedge}\,a\,Q_a\,(TList\, b\, e\, T_b)} \\
+\mathsf{=
+\exists (Q_b : b \to Set) 
+  \to Equal^{\wedge}\,a\, (List\, b)\, Q_a\, (List^{\wedge} \, b\, Q_b) \, e 
+  \times LType^{\wedge}\,b\,Q_b\,T_b 
+}
+\end{multline*}
+
+Given the lifting for $\mathsf{LType}$, we can define the lifting for $\mathsf{LTerm}$. Again, 
+the lifting 
+\[
+\mathsf{LTerm^{\wedge} : \forall (a : Set) \to (a \to Set) \to LTerm\,a \to Set}
+\]
+is defined by pattern-matching as
+\[
+  \mathsf{LTerm^{\wedge}\,a\,Q_a\,(Var\,s\,T_a) = LType^{\wedge}\, a\, Q_a\, T_a}
+\]
+where $\mathsf{s : String}$ and $\mathsf{T_a : LType\, a}$,  and 
+\begin{multline*}
+  \mathsf{LTerm^{\wedge}\,a\,Q_a\,
+  (Abs \,b \,c \,e \,s \,T_b \,t_c)} \\ 
+\mathsf{
+  = \exists (Q_b : b \to Set) (Q_c : c \to Set) \to
+      Equal^{\wedge} \, a\, (b \to c)\, Q_a\, (Arr^{\wedge} \, b\, c\, Q_b \, Q_c)\, e} \\ 
+\mathsf{
+      \times \, LType^{\wedge}\, b\, Q_b\, T_b
+      \times \, LTerm^{\wedge}\, c\, Q_c\, t_c }
+\end{multline*}
+where $\mathsf{e : Equal \, a \, (b \to c)}$, $\mathsf{s : String}$ , $\mathsf{T_b : LType\,b}$, and $\mathsf{t_c : LTerm \, c}$,
+
+and 
+\begin{multline*}
+  \mathsf{LTerm^{\wedge}\,a\,Q_a\, (App\, b\, t_{ba}\, t_b)} \\ 
+\mathsf{
+  = \exists (Q_b : b \to Set) \to 
+      LTerm^{\wedge}\, (b \to a)\, (Arr^{\wedge} \, b\, a\, Q_b \, Q_a)\, t_{ba}
+      \times LTerm^{\wedge}\, b\, Q_b\, t_b }
+\end{multline*}
+
+where $\mathsf{t_{ba} : LTerm \, (B \to A)}$ and $\mathsf{t_b : LTerm \, B}$, and
+\begin{multline*}
+  \mathsf{LTerm^{\wedge}\,a\,Q_a\, (ListC\, b\, e\, ts)} \\ 
+\mathsf{
+  = \exists (Q_b : b \to Set) \to 
+      Equal^{\wedge} \, a\, (List\,b)\, Q_a\, (List^{\wedge} \, b\, Q_b) \, e
+      \times List^{\wedge}\, (LTerm\,b) \, (LTerm^{\wedge} \, b\, Q_b) \, ts
+    }
+\end{multline*}
+
+where $\mathsf{e : Equal\, a\, (List \,b)}$ and $\mathsf{ts : List\, (LTerm b)}$. 
+Notice we use the lifting for $\mathsf{List}$ in the case for the $\mathsf{ListC}$ constructor.
+
+With these liftings, we can define the deep induction princple for $\mathsf{LTerm}$.
+
+% Var case 
+Let CVar be the type 
+\begin{multline*}
+\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to LTerm\,a \to Set)} \\
+\mathsf{\to 
+  \forall (a : Set) (Q_a : a \to Set) (s : String) (T_a : LType\, a) 
+  \to LType^{\wedge} \, a\, Q_a\, T_a
+  \to P \, a\, Q_a\, (Var \, s\, T_a)
+  }
+\end{multline*}
+associated to the $\mathsf{Var}$ constructor.
+
+% Abs case 
+Let CAbs be the type 
+\begin{align*}
+  &\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to LTerm\,a \to Set)} \\
+  &\mathsf{\to 
+  \forall (a \,b\, c: Set) (Q_a : a \to Set)  (Q_b : b \to Set) (Q_c : c \to Set)
+  (e : Equal\, a\, (b \to c)) (s : String) } \\ 
+  &\mathsf{ \to (T_b : LType\, b) \to (t_c : LTerm\, c)
+  \to Equal^{\wedge}\,a\,(b \to c)\, Q_a \, (Arr^{\wedge} \, b\, c\, Q_b \, Q_c) \, e
+  } \\
+  &\mathsf{
+  \to LType^{\wedge}\, b\, Q_b\, T_b
+  \to P\, c\, Q_c\, t_c\, 
+  \to P \, a\, Q_a\, (Abs \,b \,c \, e \,s \,T_b \, t_c)
+  }
+\end{align*}
+associated to the $\mathsf{Abs}$ constructor.
+
+% App case 
+Let CApp be the type 
+\begin{align*}
+  &\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to LTerm\,a \to Set)} \\
+  &\mathsf{\to 
+  \forall (a \,b : Set) (Q_a : a \to Set)  (Q_b : b \to Set) 
+   (t_{ba} : LTerm\, (b \to a)) (t_b : LTerm\, b)} \\
+  &\mathsf{
+  \to P\, (b \to a)\, (Arr^{\wedge} \, b\, a\, Q_b \, Q_a) \, t_{ba} \, 
+  \to P\, b\, Q_b\, t_b\, 
+  \to P \, a\, Q_a\, (App \,b \,t_{ba} \, t_b) }
+\end{align*}
+associated to the $\mathsf{App}$ constructor.
+
+% ListC case 
+Let CListC be the type 
+\begin{align*}
+  &\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to LTerm\,a \to Set)} \\
+  &\mathsf{\to 
+  \forall (a \,b : Set) (Q_a : a \to Set)  (Q_b : b \to Set) 
+    (e : Equal\, a\, (List\, b)) (ts : List\, (LTerm\, b))} \\ 
+  &\mathsf{
+    \to Equal^{\wedge}\, a\, (List\,b)\, Q_a\, (List^{\wedge}\, b\, Q_b)\, e 
+  \to List^{\wedge}\, (LTerm\,b) (P\, b\, Q_b)\, ts} \\
+&\mathsf{
+  \to P \, a\, Q_a\, (ListC \,b \,e \, ts) }
+\end{align*}
+associated to the $\mathsf{ListC}$ constructor.
+
+Then the type of the deep induction principle for $\mathsf{LTerm}$ is,
+\begin{multline*}
+\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to LTerm\,a \to Set)}
+\mathsf{\to CVar\,P \to CAbs\,P \to CApp\,P \to CListC\,P} \\
+\mathsf{\to \forall (a : Set)(Q_a : a \to Set)(t_a : LTerm\,a) \to LTerm^{\wedge}\,a\,Q_a\,t_a \to P\,a\,Q_a\,t_a}
+\end{multline*}
+
+To prove the induction principle, we define a term $\mathsf{DILTerm}$ for it. We have to define 
+\[ 
+\mathsf{DILTerm \, P\, cvar \, cabs\, capp\, clistc \, a\, Q_a\, t_a\, L_a : P\, a\, Q_a\, t_a}
+\]
+% where $\mathsf{P : \forall (a : Set) \to (a \to Set) \to LTerm\,a \to Set}$,
+
+As before, we prove the induction principle by pattern matching on $\mathsf{t_a}$.
+
+For the $\mathsf{Var}$ case, let $\mathsf{t_a = Var\, s\, T_a}$ and define
+\[
+  \mathsf{DILTerm \, P\, cvar \, cabs\, capp\, clistc \, a\, Q_a\, (Var\,s\,T_a) \, L_a = cvar \, a\, Q_a\, s\, T_a\, L_a }
+\]
+% where $\mathsf{L_a : LTerm^{\wedge}\, a\, Q_a\, (Var\,s\,T_a)}$, i.e., $\mathsf{L_a : LType^{\wedge}\,a\,Q_a\,T_a}$ 
+Notice that $\mathsf{LTerm^{\wedge}\, a\, Q_a\, (Var\,s\,T_a) = LType^{\wedge}\,a\,Q_a\,T_a}$, so 
+$\mathsf{L_a : LType^{\wedge}\,a\,Q_a\,T_a}$.
+
+% Abs case 
+For the $\mathsf{Abs}$ case, let $\mathsf{t_a = (Abs \,b \,c \, e \,s \,T_b \, t_c)} $ and define
+\begin{multline*}
+  \mathsf{DILTerm \, P\, cvar \, cabs\, capp\, clistc \, a\, Q_a\, (Abs \,b \,c \,e \,s \,T_b \, t_c) \, (Q_b , Q_c , L_e, L_{T_b}, L_{t_c})} \\
+  \mathsf{= cabs\,a\,b\,c\, Q_a\, Q_b\, Q_c\, e\, s\, T_b\, t_c\, L_e\, L_{T_b}\, p_c}
+\end{multline*}
+
+where $\mathsf{(Q_b , Q_c , L_e, L_{T_b}, L_{t_c}) : LTerm^{\wedge}\, a\, Q_a\, (Abs \,b \,c \, e \,s \,T_b \, t_c)}$
+\begin{itemize}
+  \item $\mathsf{Q_b : b \to Set}$
+  \item $\mathsf{Q_c : c \to Set}$
+  \item $\mathsf{L_e : Equal^{\wedge} \, a\, (b \to c)\, Q_a\, (Arr^{\wedge} \, b\, c\, Q_b \, Q_c) \, e}$
+  \item $\mathsf{L_{T_b}: LType^{\wedge} \, b\, Q_b\, T_b}$
+  \item $\mathsf{L_{t_c}: LTerm^{\wedge} \, c\, Q_c\, T_c}$
+\end{itemize}
+
+and $\mathsf{pc}$ is defined as:
+\[
+  \mathsf{pc =  DILTerm\,P\,cvar\,cabs \,capp \,clistc\, c\, Q_c\, t_c\, L_{t_c}  : P \, c\, Q_c \, t_c }
+\]
+
+% App case - (App \,b \,t_{ba} \, t_b)
+For the $\mathsf{App}$ case, let $\mathsf{t_a = (App \,b \,\,t_{ba} \, t_b)} $ and define
+\begin{multline*}
+  \mathsf{DILTerm \, P\, cvar \, cabs\, capp\, clistc \, a\, Q_a\,  (App \,b \,\,t_{ba} \, t_b)\, (Q_b , L_{t_{ba}}, L_{t_b})} \\
+  \mathsf{= capp\,a\,b\,Q_a\, Q_b\, t_{ba}\, t_b\, p_{ba} \, p_b}
+\end{multline*}
+
+where $\mathsf{(Q_b , L_{t_{ba}}, L_{t_b}) : LTerm^{\wedge}\, a\, Q_a\, (App \,b \,t_{ba} \, t_b)}$
+\begin{itemize}
+  \item $\mathsf{Q_b : b \to Set}$
+  \item $\mathsf{L_{t_{ba}}: LTerm^{\wedge} \, (b \to a)\, (Arr^{\wedge} \, b\, a\, Q_b \, Q_a)\, t_{ba}}$
+  \item $\mathsf{L_{t_b}: LTerm^{\wedge} \, b\, Q_b\, t_b}$
+\end{itemize}
+
+and $\mathsf{p_{ba}}$
+and 
+$\mathsf{p_b}$
+and are defined as:
+\begin{align*}
+  \mathsf{p_{ba}} &=  \mathsf{DILTerm\,P\,cvar\,cabs \,capp \,clistc\, (b \to a)\,(Arr^{\wedge} \, b\, a\, Q_b \, Q_a) \, t_{ba}\, L_{t_{ba}}   : P \, (b \to a)\, (Arr^{\wedge} \, b\, a\, Q_b \, Q_a) \, t_{ba} } \\
+  \mathsf{p_b   } &=  \mathsf{DILTerm\,P\,cvar\,cabs \,capp \,clistc\, b\, Q_b\, t_b\, L_{t_b}  : P \, b\, Q_b \, t_b } 
+\end{align*}
+
+% ListC case - (ListC \,b \,e \, ts)
+For the $\mathsf{ListC}$ case, let $\mathsf{t_a = (ListC \,b \,e \, ts)} $ and define
+\begin{multline*}
+  \mathsf{DILTerm \, P\, cvar \, cabs\, capp\, clistc \, a\, Q_a\, (ListC \,b \,e \, ts) \, (Q_b , L_e, L_{List})
+  = clistc \,a\,b\,Q_a\, Q_b\, e\, ts\, L_e\, p_{List} }
+\end{multline*}
+
+where $\mathsf{(Q_b , L_e, L_{List}): LTerm^{\wedge}\, a\, Q_a\, (ListC \,b \,e \, ts)}$
+\begin{itemize}
+  \item $\mathsf{Q_b : b \to Set}$
+  \item $\mathsf{L_e : Equal^{\wedge}\, a\, (List\,b)\, Q_a\, (List^{\wedge}\, b\, Q_b)\, e}$
+  \item $\mathsf{L_{List}: List^{\wedge} \, (LTerm\, b) \, (LTerm^{\wedge}\, b\, Q_b) \, ts}$
+\end{itemize}
+
+and $\mathsf{p_{List}}$ is defined as:
+\begin{multline*}
+  \mathsf{p_{List} = List^{\wedge}map \, (LTerm\, b) \, (LTerm^{\wedge} \, b \, Q_b)\, 
+    (P\,b\,Q_b)\, \,p_{ts} \,\, ts\,\, L_{List}  : List^{\wedge}\, (LTerm\,b) \, (P\,b\,Q_b) \, ts}
+\end{multline*}
+
+where $\mathsf{p_{ts}}$ is defined as:
+
+{\color{red} PredMap not defined yet} 
+\begin{multline*}
+  \mathsf{
+  p_{ts} = DILTerm\, P\, cvar\, cabs\, capp\, clistc\, b\, Q_b :  PredMap \,(LTerm^{\wedge}\, b\, Q_b) \, (P\,b\,Q_b)
+  }
+\end{multline*}
 
@@
 
-Lambda normal form example.
-
-%Still don't know which applications we might have
+{\color{red} Compare with structural induction? or at least state why structural induction fails }
+
+In this section we use deep induction for the $\mathsf{LTerm}$ GADT to 
+extract the type from a lambda term. 
+We have a predicate 
+\begin{align*}
+  &\mathsf{getType : \forall \, (a : Set) \to (t : LTerm\,a) \to Set} \\
+  &\mathsf{getType \,a \,t = Maybe \, (LType \, a)}
+\end{align*}
+
+and we want to show this predicate is satisfied for every element of $\mathsf{LTerm\,a}$.
+Because of the $\mathsf{ListC}$ constructor, this cannot be achieved without 
+deep induction. In particular, deep induction is required to apply the induction to the 
+individual terms in a list of terms.
+
+So, using deep induction, we want to prove: 
+\[
+  \mathsf{getTypeProof : \forall \, (a : Set) \to (t : LTerm\,a) \to getType \,a \,t} \\
+\]
+which we prove by 
+\[
+  \mathsf{getTypeProof \,a \,t = 
+    DILTerm\, (\lambda \, b\, Q_b\, t\, \to getType \,b\, t) \,
+    gtVar\,  gtABs\,  gtApp\,  gtListC\,  x\,\, K1\,\, t\, (LTerm^{\wedge}K1\, x\, t)
+    }
+\]
+where $\mathsf{K1 : x \to Set}$ and $\mathsf{LTerm^{\wedge}K1\, x\, t : LTerm^{\wedge}\, x\, K1\, t}$.
+Notice that there is no space in $\mathsf{LTerm^{\wedge}K1}$, because
+\[
+  \mathsf{LTerm^{\wedge}K1 : \forall (a : Set) (t : LTerm A) \to LTerm^{\wedge}\, a\, K1\,t}
+\]
+is a function that we will define. In addition to defining $\mathsf{LTerm^{\wedge}K1}$,
+we also have to give a proof for each constructor $\mathsf{Var, Abs, App, ListC}$:
+
+\begin{multline*}
+  \mathsf{gtVar : \forall (a : Set) (Q_a : a \to Set) (s : String) (T_a : LType\, a) 
+      \to LType^{\wedge} \, a\, Q_a\, T_a\, \to Maybe\, (LType\, a)}
+\end{multline*}
+
+% abs 
+\begin{multline*}
+    \mathsf{gtAbs : \forall (a \, b\, c: Set) (Q_a : a \to Set) (Q_b : b \to Set) (Q_c : c \to Set) 
+      (e : Equal\, a\, (b \to c)) (s : String)} \\ 
+    %
+    \mathsf{ (T_b : LType\, b) (t_c : LTerm \, c)
+      \to Equal^{\wedge}\, a\, (b \to c)\, Q_a\, (Arr^{\wedge}\, b\, c\, Q_b\, Q_c)\, e
+      \to LType^{\wedge} \, b\, Q_b\, T_b} \\ 
+    %
+    \mathsf{
+      \to Maybe\, (LType\,c) \to Maybe\, (LType\, a)}
+\end{multline*}
+
+% app 
+\begin{multline*}
+    \mathsf{gtApp : \forall (a \, b : Set) (Q_a : a \to Set) (Q_b : b \to Set) 
+      (t_{ba} : LTerm\, (b \to a)) (t_b : LTerm\, b)} \\ 
+    \mathsf{
+      \to Maybe\, (LType\, (b \to a))
+      \to Maybe\, (LType\, b)
+      \to Maybe\, (LType\, a)
+    }
+\end{multline*}
+
+% list 
+\begin{multline*}
+    % abs
+    \mathsf{gtListC : \forall (a \, b : Set) (Q_a : a \to Set) (Q_b : b \to Set) 
+      (e : Equal\, a\, (List\, b)) (ts : List\, (LTerm\, b))} \\ 
+    \mathsf{
+      \to Equal^{\wedge}\, a\, (List\,b) \, Q_a\, (List^{\wedge}\, b\, Q_b)\, e
+      \to List^{\wedge}\, (LTerm\, b)\, (getType\, b)\, ts
+      \to Maybe\, (LType\, a)
+    }
+\end{multline*}
+
+For variables we simply return the type $\mathsf{T_a}$ given in $\mathsf{Var\, s\, T_a}$, 
+and the cases for abstraction and application are similar.
+The interesting case is $\mathsf{gtListC}$, in which we have to use 
+the results of ($\mathsf{List^{\wedge}\, (LTerm\, b)\, (getType\, b)\, ts}$)
+in order to extract the type of one of the terms in the list. 
+To define $\mathsf{gtListC}$ we pattern-match on the list of terms $\mathsf{ts}$.
+
+If $\mathsf{ts}$ is the empty list (denoted by $\mathsf{[]}$), we cannot extract a type, 
+so we return $\mathsf{nothing}$. \\
+{\color{red} Maybe handle this case differently, by using non-empty lists, for example}
+\[
+  \mathsf{gtListC\, a\, b\, Q_a\, Q_b\, e\, [] \, L_e L_{ts} = nothing}
+\]
+
+If $\mathsf{ts}$ is a non-empty list, we pattern match on $\mathsf{L_{ts}}$ and use the result
+to construct the type we need:
+\begin{align*}
+  &\mathsf{gtListC\, a\, b\, Q_a\, Q_b\, e\, (t :: ts) \, L_e (nothing , L_{ts}) = nothing} \\
+  &\mathsf{gtListC\, a\, b\, Q_a\, Q_b\, e\, (t :: ts) \, L_e (just\, T_b , L_{ts}) = just \, (TList\, b\, e\, T_b)}
+\end{align*}
+
+where $\mathsf{e : Equal\, a\, (List\,b)}$ and $\mathsf{T_b : LType\, b}$.
+
+The last piece of infrastructure we need to define $\mathsf{getTypeProof}$ is a function 
+\[
+  \mathsf{LTerm^{\wedge}K1 : \forall (a : Set) (t : LTerm A) \to LTerm^{\wedge}\, a\, K1\,t}
+\]
+that provides a 
+proof of $\mathsf{LTerm^{\wedge}\,a\,K1\,t}$ for any term $\mathsf{t : LTerm\, a}$. 
+Because $\mathsf{LTerm^{\wedge}}$ is defined in terms of $\mathsf{LType^{\wedge}}$, $\mathsf{Arr^{\wedge}}$, 
+and $\mathsf{List^{\wedge}}$, we will need analogous functions for these liftings as well.
+
+{\color{red} Finish example by defining $\mathsf{LTerm^{\wedge}K1}$}
 

### 2021-04-06 induction-paper/GADT-induction.tex

--- 2021-04-06 induction-paper/GADT-induction.tex
@@
 
-{\color{red} Add third example here.}
-
 %{\color{blue} Do we really need the existential quantifiers?}
@@
 
+{\color{red} STLC example}
+
+GADTs are expressive enough to represent, for example, the types and terms of the simply typed lambda calculus. 
+We will use a variant of the STLC including a base type of Bool, and type constructors for arrow types and list types:
+
+\begin{equation}\label{eq:eq_ltype}
+\begin{array}{l}
+\mathsf{data\, LType\,(a : Set) : Set\,where}\\
+\mathsf{\;\;\;\;\;\;\;\;TBool :\, \exists (b : Set). \,Equal\,a\,(KBool\, b) \to LType\,a}\\
+\mathsf{\;\;\;\;\;\;\;\;TArr\,\,\, :\, \exists (b : Set) (c : Set).\,Equal\,a\,(b \to c) \to LType\,b \to LType\,c \to LType\,a}\\
+  \mathsf{\;\;\;\;\;\;\;\;TList\,\, :\, \exists (b : Set).\,Equal\,a\,(\List\,b) \to LType\,b \to LType\,a}
+\end{array}
+\end{equation}
+
+The term calculus for this type system includes variables, abstraction, application, 
+and also includes an introduction rule that takes a list of terms of type $\mathsf{b}$ and produces a term of type $\mathsf{List\,b}$.
+
+\begin{equation}\label{eq:eq_lterm}
+\begin{array}{l}
+\mathsf{data\, LTerm\,(a : Set) : Set\,where}\\
+\mathsf{\;\;\;\;\;\;\;\;Var\,\,\,\,\,\,:\, String \to LType\,a \to LTerm\,a} \\
+\mathsf{\;\;\;\;\;\;\;\;Abs\,\,\,\,\,\, :\, \exists (b : Set) (c : Set).\,Equal\,a\,(b \to c) \to String 
+      \to LType\,b \to LTerm\,c \to LTerm\,a}\\
+  \mathsf{\;\;\;\;\;\;\;\;App\,\,\,\,\, :\, \exists (b : Set).\,LTerm (b \to a) \to LTerm\,b \to LTerm\,a}\\
+  \mathsf{\;\;\;\;\;\;\;\;ListC\,\, :\, \exists (b : Set).\,Equal\,a\,(\List\,b) \to List\,(LTerm\,b) \to LTerm\,a}
+\end{array}
+\end{equation}
+
+The type parameter for $\mathsf{LTerm}$ tracks the types of STLC terms. So the type $\mathsf{LTerm\,A}$ 
+contains lambda terms of type $\mathsf{A}$. The $\mathsf{LType}$ GADT is used to encode the 
+grammar of permissible types for this calculus.
+
+Variables are represented as strings and are tagged with their type. 
+The $\mathsf{Abs}$ constructor takes a name (string) and a type for the bound variable,
+and a term of type $\mathsf{c}$ for the body of the abstraction.
+The $\mathsf{App}$ constructor applies a term of type $\mathsf{b \to a}$ to a term of type $\mathsf{b}$, producing
+a term of type $\mathsf{a}$. 
+The $\mathsf{ListC}$ constructor takes a list of terms of type $\mathsf{b}$ and produces a term of type $\mathsf{List\,b}$.
+
 \section{(Deep) induction for GADTs}
@@
 
-\subsection{Third example introduced above}
+%\subsection{(Deep) induction for $\mathsf{Seq}$ (general style)}
+%
+%We need the lifting of the type $\mathsf{Pair = \forall (b : Set)(c : Set) \to b \times c}$, which is 
+%\[
+%\mathsf{Pair^{\wedge} : \forall (b : Set)(c : Set) \to (b \to Set) \to (c \to Set) \to b \times c \to Set}
+%\]
+%defined as
+%\[
+%\mathsf{Pair^{\wedge}\,b\,c\,Q_b\,Q_c\,(y, z) = Q_b\,y \times Q_c\,z}
+%\]
+%
+%%Let 
+%%$\mathsf{F, F_1, F_2 : (Set \to Set) \to Set \to Set \to Set}$
+%%be defined as
+%%$\mathsf{F_1\,G\,b\,c = G\,b}$,
+%%$\mathsf{F_2\,G\,b\,c = G\,c}$,
+%%$\mathsf{F\,G\,b\,c = F_1\,G\,b\,c \times F_2\,G\,b\,c}$,
+%%and $\mathsf{K = Pair}$.
+%
+%
+%Let 
+%$\mathsf{F : (Set \to Set) \to Set \to Set \to Set}$
+%be defined as
+%$\mathsf{F\,G\,b\,c = G\,b \times G\,c}$
+%and $\mathsf{K = Pair}$.
+%
+%%The type of $\mathsf{F_i^{\wedge}}$ is
+%%\begin{multline*}
+%%\mathsf{\forall (G : Set \to Set) (b : Set) (c : Set)
+%%\to (\forall (a : Set) \to (a \to Set) \to G\,a \to Set)} \\
+%%\mathsf{\to (b \to Set) \to (c \to Set) \to (F_i\,G\,b\,c) \to Set}
+%%\end{multline*}
+%The type of $\mathsf{F^{\wedge}}$ is
+%\begin{multline*}
+%\mathsf{\forall (G : Set \to Set) (b : Set) (c : Set)
+%\to (\forall (a : Set) \to (a \to Set) \to G\,a \to Set)} \\
+%\mathsf{\to (b \to Set) \to (c \to Set) \to (F_1\,G\,b\,c \times F_2\,G\,b\,c) \to Set}
+%\end{multline*}
+%%and $\mathsf{F^{\wedge}\,G\,b\,c\,P\,Q_b\,Q_c\,(s_b, s_c)
+%%= F_1^{\wedge}\,G\,b\,c\,P\,Q_b\,Q_c\,s_b \times F_2^{\wedge}\,G\,b\,c\,P\,Q_b\,Q_c\,s_c}$
+%
+%Next we shall consider the $\mathsf{Seq}$ type defined in Equation~\ref{eq:eq_seq}.
+%Again, the first step in deriving its induction rule consists in defining the lifting
+%\[
+%\mathsf{Seq^{\wedge} : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set}
+%\]
+%which is given by
+%\[
+%\mathsf{Seq^{\wedge}\,a\,Q\,(Const\,x) = Q\,x}
+%\]
+%and
+%\begin{multline*}
+%\mathsf{Seq^{\wedge}\,a\,Q_a\,(SPair\,b\,c\,e\,(s_b,s_c))
+%=
+%\exists (Q_b : b \to Set) (Q_c : c \to Set)} \\
+%\mathsf{
+%\to Equal^{\wedge}\,a\, (Pair\,b\,c)\, Q_a\, (Pair^{\wedge}\,b\,c\,Q_b\,Q_c) \, e
+%\times F^{\wedge}\,Seq\,b\,c\,Seq^{\wedge}\,Q_b\,Q_c\,(s_b, s_c)
+%}
+%\end{multline*}
+%%\begin{multline*}
+%%\mathsf{Seq^{\wedge}\,a\,Q_a\,(SPair\,b\,c\,e\,(s_b,s_c))
+%%=
+%%\exists (Q_b : b \to Set) (Q_c : c \to Set)} \\
+%%\mathsf{
+%%\to Equal^{\wedge}\,a\, (Pair\,b\,c)\, Q_a\, (Pair^{\wedge}\,b\,c\,Q_b\,Q_c) \, e
+%%\times F_1^{\wedge}\,Seq\,b\,c\,Seq^{\wedge}\,Q_b\,Q_c\,s_b
+%%\times F_2^{\wedge}\,Seq\,b\,c\,Seq^{\wedge}\,Q_b\,Q_c\,s_c
+%%}
+%%\end{multline*}
+%
+%Define
+%\[
+%\mathsf{PredMap : \forall (a : Set) \to (a \to Set) \to (a \to Set) \to Set}
+%\]
+%as $\mathsf{PredMap\,a\,Q_1\,Q_2 = \forall (x : a) \to Q_1\,x \to Q_2\,x}$
+%
+%%  -- Lifting also acts on morphisms between predicates with the same carrier
+%%  {-# TERMINATING #-}
+%%  Bush^-map : ∀ {l} {X : Set} {Q1 Q2 : X → Set l}
+%%              → Pred-map {l} Q1 Q2 → Pred-map (Bush^ Q1) (Bush^ Q2)
+%%  Bush^-map Q12 Bnil q = bigtt
+%%  Bush^-map Q12 (Bcons x b) (y , bby) =
+%%                Q12 x y , Bush^-map (Bush^-map Q12) b bby
+%Define
+%\[
+%\mathsf{SeqLMap : \forall (a : Set) (Q_1\,Q_2 : a \to Set) \to PredMap\,a\,Q_1\,Q_2 \to PredMap\,(G\,a)\,(Seq^{\wedge}\,a\,Q_1)\,(Seq^{\wedge}\,a\,Q_2)}
+%\]
+%as
+%\begin{gather*}
+%\mathsf{
+%SeqLMap\,a\,Q_1\,Q_2\,M\,(Const\,x)\,z = M\,x\,z
+%} \\
+%\mathsf{
+%SeqLMap\,a\,Q_1\,Q_2\,M\,(SPair\,b\,c\,e\,s_b\,s_c)\,(Q_b, Q_c, l_E, l_F) =
+%}
+%\end{gather*}
+%
+%
+%We have an induction rule for $\mathsf{Seq}$
+%\begin{multline}
+%\mathsf{DISeq : \forall (P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set)} \\
+%\mathsf{\to \big( \forall (a : Set) (Q_a : a \to Set) (x : a) \to Q_a\,x \to P\,a\,Q_a\,(Const\,x) \big)} \\
+%\mathsf{\to \big( \forall (a : Set) (b : Set) (c : Set)
+%	(Q_a : a \to Set) (Q_b : b \to Set) (Q_c : c \to Set)} \\
+%	\mathsf{( (s_b, s_c) : F\,Seq\,b\,c) (e : Equal\,a\,(Pair\,b\,c))
+%	\to Equal^{\wedge} a\, (Pair\,b\,c)\, Q_a\, (Pair^{\wedge}\,b\,c\,Q_b\,Q_c)\, e} \\
+%	\mathsf{\to F^{\wedge}\,G\,b\,c\,P\,Q_b\,Q_c\,(s_b,s_c)
+%	\to P a Q_a ( SPair\,b\,c\,e\,(s_b,s_c) ) \big)} \\
+%\mathsf{\to \forall (a : Set)(Q_a : a \to Set)(s_a : Seq\,a) \to Seq^{\wedge}\,a\,Q_a\,s_a \to P\,a\,Q_a\,s_a}
+%\end{multline}
+%We define $\mathsf{DISeq\,P\,PConst\,PSPair\,a\,Q_a\,s_a\,l_a}$,
+%where
+%$\mathsf{P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set}$,
+%$\mathsf{Q_a : a \to Set}$, $\mathsf{s_a : Seq\,a}$
+%and $\mathsf{l_a : Seq^{\wedge}\,a\,Q\,s_a}$,
+%by pattern-matching on $\mathsf{s_a}$.
+%Let $\mathsf{s_a = Const\,x}$ for $\mathsf{x : a}$, and define
+%\[
+%\mathsf{DISeq\,P\,PConst\,PSPair\,a\,Q_a\,(Const\,x)\,l_a = PConst\,a\,Q_a\,x\,l_a}
+%\]
+%Notice that $\mathsf{Seq^{\wedge}\,a\,Q_a\,(Const\,x) = Q_a\,x}$, and thus $\mathsf{l_a : Q_a\,x}$,
+%making the right-hand-side in the above expression type-check.
+%Now, let $\mathsf{s_a = SPair\,b\,c\,e\,(s_b,s_c)}$
+%for $\mathsf{e : Equal\,a\,(Pair\,b\,c)}$ and $\mathsf{(s_b, s_c) : F\,Seq\,b\,c}$,
+%and define
+%\[
+%\mathsf{DISeq\,P\,PConst\,PSPair\,a\,Q_a\,(SPair\,b\,c\,e\,(s_b,s_c))\,(Q_b, Q_c, l_e, l_{F_1}, l_{F_2})
+%	= PSPair\,a\,b\,c\,Q_a\,Q_b\,Q_c\,(s_b,s_c)\,e\,l_e\,p}
+%\]
+%where
+%$\mathsf{p : F^{\wedge}\,G\,b\,c\,P\,Q_b\,Q_c\,(s_b,s_c)}$,
+%$\mathsf{l_e : Equal^{\wedge}\,a\, (Pair\,b\,c)\, Q_a\, (Pair^{\wedge}\,Q_b\,Q_c) \, e}$,
+%$\mathsf{l_{F_1} : F_1^{\wedge}\,Seq\,b\,c\,Seq^{\wedge}\,Q_b\,Q_c\,s_b}$,
+%$\mathsf{l_{F_2} : F_2^{\wedge}\,Seq\,b\,c\,Seq^{\wedge}\,Q_b\,Q_c\,s_c}$
+%(notice that $\mathsf{(Q_b, Q_c, l_e, l_{F_1}, l_{F_2}) : Seq^{\wedge}\,a\,Q\,(SPair\,b\,c\,e\,x\,y)}$,
+%making the right-hand-side in the above expression type-check)
+%and $\mathsf{p}$ is defined as follows:
+%\[
+%\mathsf{p = (???, ???) 
+%: F_1^{\wedge}\,G\,b\,c\,P\,Q_b\,Q_c\,s_b \times F_2^{\wedge}\,G\,b\,c\,P\,Q_b\,Q_c\,s_c} \\
+%\]
+
+%\begin{align*}
+%\mathsf{DISeq\,P\,PConst\,PSPair\,a\,Q\,(Const\,x)\,l}
+%	&=\mathsf{PConst\,a\,Q\,x\,l} \\
+%\mathsf{DISeq\,P\,PConst\,PSPair\,a\,Q_a\,(SPair\,b\,c\,e\,x\,y)\,(Q_b\,Q_c\,e', l_b, l_c)}
+%	&=\mathsf{PSpair\,a\,b\,c\,Q_a\,Q_b\,Q_c\,x\,y\,e\,p_b\,p_c\,e'}
+%\end{align*}
+%where
+%$\mathsf{Q_a : a \to Set}$, $\mathsf{Q_b : b \to Set}$, $\mathsf{Q_c : c \to Set}$,
+%$\mathsf{x : Seq\,b}$, $\mathsf{y : Seq\,c}$,
+%$\mathsf{e : Equal\,a\,(b \times c)}$,
+%$\mathsf{p_b : P\,b\,Q_b\,x}$, $\mathsf{p_c : P\,c\,Q_c\,y}$,
+%$\mathsf{e' : Equal^{\wedge}\,a\, (b \times c)\, Q\, (Q_b \times Q_c) \, e}$,
+%$\mathsf{l_b : Seq^{\wedge}\,b\,Q_b\,x}$,
+%$\mathsf{l_c : Seq^{\wedge}\,c\,Q_c\,y}$
+%and
+%\begin{align*}
+%\mathsf{p_b} &=\mathsf{DISeq\,P\,PConst\,PSPair\,b\,Q_b\,x\,l_b : P\,b\,Q_b\,x} \\
+%\mathsf{p_c} &=\mathsf{DISeq\,P\,PConst\,PSPair\,c\,Q_c\,y\,l_c : P\,c\,Q_c\,y}
+%\end{align*}
+%
+%\begin{align*}
+%\mathsf{Q_a} &: \mathsf{a \to Set} \\
+%\mathsf{Q_b} &: \mathsf{b \to Set} \\
+%\mathsf{Q_c} &: \mathsf{c \to Set} \\
+%\mathsf{x} &: \mathsf{Seq\,b} \\
+%\mathsf{y} &: \mathsf{Seq\,c} \\
+%\mathsf{e} &: \mathsf{Equal\,a\,(b \times c)} \\
+%\mathsf{p_b} &: \mathsf{P\,b\,Q_b\,x} \\
+%\mathsf{p_c} &: \mathsf{P\,c\,Q_c\,y} \\
+%\mathsf{e'} &: \mathsf{Equal^{\wedge}\,a\, (b \times c)\, Q\, (Q_b \times Q_c) \, e} \\
+%\mathsf{l_b} &: \mathsf{Seq^{\wedge}\,b\,Q_b\,x} \\
+%\mathsf{l_c} &: \mathsf{Seq^{\wedge}\,c\,Q_c\,y} \\
+%\mathsf{p_b = DISeq\,P\,PConst\,PSPair\,b\,Q_b\,x\,l_b} &: \mathsf{P\,b\,Q_b\,x} \\
+%\mathsf{p_c = DISeq\,P\,PConst\,PSPair\,c\,Q_c\,y\,l_c} &: \mathsf{P\,c\,Q_c\,y}
+%\end{align*}
+
+\subsection{(Deep) induction for $\mathsf{LTerm}$ }
 

### 2021-04-07 induction-paper/GADT-induction.tex
completed section 6 (lambda term example)

@@
 The term calculus for this type system includes variables, abstraction, application, 
-and also includes an introduction rule that takes a list of terms of type $\mathsf{b}$ and produces a term of type $\mathsf{List\,b}$.
+and also includes an introduction rule that takes a list of lambda terms of type $\mathsf{b}$ and produces a lambda term of type $\mathsf{List\,b}$.
 
@@
 The $\mathsf{Abs}$ constructor takes a name (string) and a type for the bound variable,
-and a term of type $\mathsf{c}$ for the body of the abstraction.
-The $\mathsf{App}$ constructor applies a term of type $\mathsf{b \to a}$ to a term of type $\mathsf{b}$, producing
-a term of type $\mathsf{a}$. 
-The $\mathsf{ListC}$ constructor takes a list of terms of type $\mathsf{b}$ and produces a term of type $\mathsf{List\,b}$.
+and a lambda term of type $\mathsf{c}$ for the body of the abstraction.
+The $\mathsf{App}$ constructor applies a lambda term of type $\mathsf{b \to a}$ to a lambda term of type $\mathsf{b}$, producing
+a lambda term of type $\mathsf{a}$. 
+The $\mathsf{ListC}$ constructor takes a list of lambda terms of type $\mathsf{b}$ and produces a lambda term of type $\mathsf{List\,b}$.
 
@@
 
-{\color{red} Use $\mathsf{Pair^{\wedge}}$ everywhere}
-
 {\color{blue} Some notational conventions: I am using uppercase letters for elements of LType and 
@@
 \end{multline*}
+where $\mathsf{Q_a : a \to Set}$, $\mathsf{e : Equal\, a\, (List\, b)}$, and $\mathsf{T_b : LType \,b}$.
 
@@
 where $\mathsf{e : Equal \, a \, (b \to c)}$, $\mathsf{s : String}$ , $\mathsf{T_b : LType\,b}$, and $\mathsf{t_c : LTerm \, c}$,
-
 and 
@@
 \end{multline*}
-
 where $\mathsf{t_{ba} : LTerm \, (B \to A)}$ and $\mathsf{t_b : LTerm \, B}$, and
@@
 \end{multline*}
-
 where $\mathsf{e : Equal\, a\, (List \,b)}$ and $\mathsf{ts : List\, (LTerm b)}$. 
@@
 With these liftings, we can define the deep induction princple for $\mathsf{LTerm}$.
-
 % Var case 
@@
 associated to the $\mathsf{Var}$ constructor. 
-
 % Abs case 
@@
 associated to the $\mathsf{Abs}$ constructor. 
-
 % App case 
@@
 associated to the $\mathsf{App}$ constructor. 
-
 % ListC case 
@@
 associated to the $\mathsf{ListC}$ constructor. 
-
 Then the type of the deep induction principle for $\mathsf{LTerm}$ is,
@@
 
+\noindent
 As before, we prove the induction principle by pattern matching on $\mathsf{t_a}$. 
-
-For the $\mathsf{Var}$ case, let $\mathsf{t_a = Var\, s\, T_a}$ and define
+For the $\mathsf{Var}$ case, let $\mathsf{t_a = (Var\, s\, T_a)}$ and define
 \[
@@
 $\mathsf{L_a : LType^{\wedge}\,a\,Q_a\,T_a}$.
-
+%
 % Abs case 
@@
 
+\noindent
 % App case - (App \,b \,t_{ba} \, t_b)
@@
 
+\noindent
 where $\mathsf{(Q_b , L_{t_{ba}}, L_{t_b}) : LTerm^{\wedge}\, a\, Q_a\, (App \,b \,t_{ba} \, t_b)}$
@@
 \end{itemize}
-
-and $\mathsf{p_{ba}}$
-and 
-$\mathsf{p_b}$
+and $\mathsf{p_{ba}}$ and $\mathsf{p_b}$
 and are defined as:
@@
 
+\noindent
 % ListC case - (ListC \,b \,e \, ts)
@@
 
+\noindent
 where $\mathsf{(Q_b , L_e, L_{List}): LTerm^{\wedge}\, a\, Q_a\, (ListC \,b \,e \, ts)}$
@@
 
+\noindent
 and $\mathsf{p_{List}}$ is defined as:
@@
 \end{multline*}
-
-where $\mathsf{p_{ts}}$ is defined as:
-
-{\color{red} PredMap not defined yet} 
+and $\mathsf{p_{ts}}$ is defined as:
 \begin{multline*}
@@
 \end{multline*}
+where 
+\[\mathsf{List^{\wedge}map : \forall\, (a : Set) \to (Q_a \, Q'_a : a \to Set) 
+  \to PredMap\,Q_a\,Q'_a \to PredMap \,(List^{\wedge}\, a\, Q_a)\, (List^{\wedge}\, a\, Q'_a)}\]
+takes a morphism of predicates and produces a morphism of lifted predicates. 
+$\mathsf{PredMap}$ gives the type of morphisms of predicates and is defined as: 
+\begin{align*}
+  &\mathsf{PredMap : \forall\, \{a : Set\} \to (Q_a \, Q'_a : a \to Set) \to Set } \\
+  &\mathsf{PredMap \, Q_a\,Q'_a\, = \forall\, (x : A) \to Q_a\,x \to Q'_a\,x}
+\end{align*}
 
@@
 
-{\color{red} Compare with structural induction? or at least state why structural induction fails }
+{\color{red} Compare with structural induction? or at least state why structural induction fails.} \\
+{\color{red} Can get rid of Maybe using non-empty lists and postulates}
 
@@
 \end{align*}
-
-and we want to show this predicate is satisfied for every element of $\mathsf{LTerm\,a}$.
+that takes a lambda term and produces its type (using $\mathsf{Maybe}$ to represent
+potential failure).
+We want to show this predicate is satisfied for every element of $\mathsf{LTerm\,a}$.
 Because of the $\mathsf{ListC}$ constructor, this cannot be achieved without 
@@
     DILTerm\, (\lambda \, b\, Q_b\, t\, \to getType \,b\, t) \,
-    gtVar\,  gtABs\,  gtApp\,  gtListC\,  x\,\, K1\,\, t\, (LTerm^{\wedge}K1\, x\, t)
+    gtVar\,  gtABs\,  gtApp\,  gtListC\,  a\,\, K1\,\, t\, (LTerm^{\wedge}K1\, a\, t)
     }
 \]
-where $\mathsf{K1 : x \to Set}$ and $\mathsf{LTerm^{\wedge}K1\, x\, t : LTerm^{\wedge}\, x\, K1\, t}$.
+where $\mathsf{K1 : a \to Set}$ and $\mathsf{LTerm^{\wedge}K1\, a\, t : LTerm^{\wedge}\, a\, K1\, t}$.
 Notice that there is no space in $\mathsf{LTerm^{\wedge}K1}$, because
@@
 we also have to give a proof for each constructor $\mathsf{Var, Abs, App, ListC}$:
-
 \begin{multline*}
@@
 \]
-
 If $\mathsf{ts}$ is a non-empty list, we pattern match on $\mathsf{L_{ts}}$ and use the result
@@
 \end{align*}
-
 where $\mathsf{e : Equal\, a\, (List\,b)}$ and $\mathsf{T_b : LType\, b}$.
 
+\subsection{Defining $\mathsf{LTerm^{\wedge}K1}$}
 The last piece of infrastructure we need to define $\mathsf{getTypeProof}$ is a function 
 \[
-  \mathsf{LTerm^{\wedge}K1 : \forall (a : Set) (t : LTerm A) \to LTerm^{\wedge}\, a\, K1\,t}
+  \mathsf{LTerm^{\wedge}K1 : \forall (a : Set) \to (t : LTerm A) \to LTerm^{\wedge}\, a\, K1\,t}
 \]
@@
 and $\mathsf{List^{\wedge}}$, we will need analogous functions for these liftings as well. 
-
-{\color{red} Finish example by defining $\mathsf{LTerm^{\wedge}K1}$}
+We only give the definition of $\mathsf{LTerm^{\wedge}K1}$, but the definitions for 
+$\mathsf{LType^{\wedge}}$, $\mathsf{Arr^{\wedge}}$, and $\mathsf{List^{\wedge}}$ are analogous.
+
+$\mathsf{LTerm^{\wedge}K1}$ is defined by pattern matching on the lambda term $\mathsf{t}$. 
+For the $\mathsf{Var}$ case, let $\mathsf{t = (Var\, s\, T_a)}$ and define
+\[
+  \mathsf{LTerm^{\wedge}K1\,a\,(Var\,s\,T_a) = LType^{\wedge}K1\,a\,T_a}
+\]
+For the $\mathsf{Abs}$ case, let $\mathsf{t = (Abs \,b \,c \, e \,s \,T_b \, t_c)} $ and recall the 
+definition of $\mathsf{LTerm^{\wedge}}$ for the $\mathsf{Abs}$ constructor, 
+instantiating the predicate $\mathsf{Q_a}$ to $\mathsf{K1}$: 
+% \,a\,Q_a\, (Abs \,b \,c \,e \,s \,T_b \,t_c)}$, :
+\begin{multline*}
+  \mathsf{LTerm^{\wedge}\,a\,K1\,
+  (Abs \,b \,c \,e \,s \,T_b \,t_c)} \\ 
+\mathsf{
+  = \exists (Q_b : b \to Set) (Q_c : c \to Set) \to
+      Equal^{\wedge} \, a\, (b \to c)\, K1\, (Arr^{\wedge} \, b\, c\, Q_b \, Q_c)\, e} \\ 
+\mathsf{
+      \times \, LType^{\wedge}\, b\, Q_b\, T_b
+      \times \, LTerm^{\wedge}\, c\, Q_c\, t_c }
+\end{multline*}
+so to define the $\mathsf{Abs}$ case of $\mathsf{LTerm^{\wedge}K1}$, we need a proof of 
+\[
+  \mathsf{Equal^{\wedge} \, a\, (b \to c)\, K1\, (Arr^{\wedge} \, b\, c\, Q_b \, Q_c)\, e}
+\]
+i.e., that $\mathsf{K1}$ is (extensionally) equal to the lifting $\mathsf{(Arr^{\wedge} \, b\, c\, Q_b \, Q_c)}$
+for some predicates $\mathsf{Q_b, Q_c}$. The only reasonable choice for $\mathsf{Q_b}$ and $\mathsf{Q_c}$ 
+is to let both be $\mathsf{K1}$, which means we need a proof of:
+\[
+  \mathsf{Equal^{\wedge} \, a\, (b \to c)\, K1\, (Arr^{\wedge} \, b\, c\, K1 \, K1)\, e}
+\]
+Since we are working with proof-relevant predicates (i.e., functions into $\mathsf{Set}$ rather than 
+functions into $\mathsf{Bool}$), 
+the lifting $\mathsf{(Arr^{\wedge} \, b\, c\, K1 \, K1)}$ of $\mathsf{K1}$ to arrow types
+is not identical to $\mathsf{K1}$ on arrow types, but the predicates are (extensionally) isomorphic. 
+We discuss this issue in more detail at the end of the section. 
+For now, we assume a proof 
+\[
+  \mathsf{Equal^{\wedge}ArrK1 : Equal^{\wedge} \, a\, (b \to c)\, K1\, (Arr^{\wedge} \, b\, c\, K1 \, K1)\, e}
+\]
+and define the $\mathsf{Abs}$ case of $\mathsf{LTerm^{\wedge}K1}$ as 
+\[
+  \mathsf{LTerm^{\wedge}K1\,a\, (Abs \,b \,c \, e \,s \,T_b \, t_c) = 
+    (K1 , K1 , Equal^{\wedge}ArrK1 , LType^{\wedge}K1 \,b\,T_b\, , LTerm^{\wedge}K1\,c\,t_c)
+  }
+\]
+For the $\mathsf{App}$ case, let $\mathsf{t = (App \,b \,\,t_{ba} \, t_b)}$ and just as we did for the $\mathsf{Abs}$ case, 
+recall the definition of $\mathsf{LTerm^{\wedge}\,a\,(App \,b \,\,t_{ba} \, t_b)}$ with all of the predicates
+instantiated with $\mathsf{K1}$: 
+\[
+\mathsf{
+      LTerm^{\wedge}\, (b \to a)\, (Arr^{\wedge} \, b\, a\, K1 \, K1)\, t_{ba}
+      \times LTerm^{\wedge}\, b\, K1\, t_b }
+\]
+The second component can be given using $\mathsf{LTerm^{\wedge}K1}$, and we can 
+define the first component using a proof of
+\[
+\mathsf{LTerm^{\wedge}\, (b \to a)\, K1\,\, t_{ba}}
+\]
+and a map-like function 
+\begin{multline*}
+  \mathsf{LTerm^{\wedge}EqualMap : \forall\, \{a : Set\} \to (Q_a\,Q'_a : a \to Set) 
+    \to (Equal^{\wedge}\,a\,a\,Q_a\,Q'_a\,rfl)} \\
+    \mathsf{
+    \to PredMap\, (LTerm^{\wedge}\,a\,Q_a)\,(LTerm^{\wedge}\,a\,Q'_a)}
+\end{multline*}
+that takes two (extensionally) equal predicates with the same carrier 
+and produces a morphism of predicates between their liftings. The definition is 
+straightforward enough, so we omit the details. 
+But it is worth noting that while a true $\mathsf{HLMap}$ function, 
+which takes a \emph{morphism} of predicates instead of a proof of equality, cannot be defined 
+for GADTs in general, an analogue of $\mathsf{LTerm^{\wedge}EqualMap}$ should be definable 
+for every GADT. These analogues of $\mathsf{LTerm^{\wedge}EqualMap}$ will be required to define 
+$\mathsf{G^{\wedge}K1}$ whenever $\mathsf{G}$ has a constructor of the form
+  ($\mathsf{c : G \, (F\, b) \to G\,(K\,b)}$).
+
+Using $\mathsf{LTerm^{\wedge}EqualMap}$, we can define the $\mathsf{App}$ case of $\mathsf{LTerm^{\wedge}K1}$ as 
+\[
+  \mathsf{LTerm^{\wedge}K1\,a\, (App \,b \,\,t_{ba} \, t_b) = 
+    (K1 , L_{Arr^{\wedge}\,K1} , LTerm^{\wedge}K1\,b\,t_b)}
+\]
+where $\mathsf{ L_{Arr^{\wedge}\,K1} : LTerm^{\wedge}\, (b \to a)\, (Arr^{\wedge} \, b\, a\, K1 \, K1)\, t_{ba}}$ is defined as
+\[
+  \mathsf{L_{Arr^{\wedge}\,K1} =  LTerm^{\wedge}EqualMap\,K1\,(Arr^{\wedge}\,b\,a\,K1\,K1)\,\, Equal^{\wedge}ArrK1\, \,\,t_{ba}\, \,L_{K1}}
+\]
+where $\mathsf{L_{K1} = LTerm^{\wedge}K1\, (b \to a)\, t_{ba} : LTerm^{\wedge}\, (b \to a) \, K1\, t_{ba}}$.
+
+Finally, we define the $\mathsf{ListC}$ case for $\mathsf{LTerm^{\wedge}K1}$. 
+Let $\mathsf{t = (ListC\, b\, e\, ts)}$ and recall the definition of $\mathsf{LTerm^{\wedge}\,a\,(ListC\, b\, e\, ts)}$ with
+all of the predicates instantiated to $\mathsf{K1}$: 
+\[
+\mathsf{ Equal^{\wedge} \, a\, (List\,b)\, K1\, (List^{\wedge} \, b\, K1) \, e
+      \times List^{\wedge}\, (LTerm\,b) \, (LTerm^{\wedge} \, b\, K1) \, ts }
+\]
+We can give the first component by assuming a proof \,
+$\mathsf{Equal^{\wedge}ListK1 : Equal^{\wedge} \, a\, (List\,b)\, K1\, (List^{\wedge} \, b\, K1) \, e}$, 
+but for the second component we again have multiple liftings nested together. In this case, we can 
+get a proof of
+\[
+\mathsf{ List^{\wedge}\, (LTerm\,b) \, (LTerm^{\wedge} \, b\, K1) \, ts }
+\]
+using 
+\[\mathsf{List^{\wedge}map : \forall\, (a : Set) \to (Q_a \, Q'_a : a \to Set) 
+  \to PredMap\,Q_a\,Q'_a \to PredMap \,(List^{\wedge}\, a\, Q_a)\, (List^{\wedge}\, a\, Q'_a)}\]
+to map a morphism of predicates 
+\[
+  \mathsf{PredMap \, (K1)\, (LTerm^{\wedge}\,b\,K1)}
+\]
+to a morphism of lifted predicates
+\[
+  \mathsf{PredMap \,(List^{\wedge}\,(LTerm\,b)\,K1) \, (List^{\wedge}\,(LTerm\,b)\, (LTerm^{\wedge}\,b\,K1))}
+\]
+
+We define the $\mathsf{ListC}$ case of $\mathsf{LTerm^{\wedge}K1}$ as 
+\[
+  \mathsf{LTerm^{\wedge}K1\,a\, (ListC\, b\, e\, ts) =
+    (K1 , \, Equal^{\wedge}ListK1 , \,  L_{List^{\wedge}LTerm^{\wedge}K1} ) }
+\]
+where
+ $\mathsf{L_{List^{\wedge}LTerm^{\wedge}K1} : List^{\wedge}\, (LTerm\,b) \, (LTerm^{\wedge} \, b\, K1) \, ts}$
+\[
+  \mathsf{ 
+ L_{List^{\wedge}LTerm^{\wedge}K1} =
+  List^{\wedge}map \, (LTerm\,b) \, K1 \, (LTerm^{\wedge}\, b\, K1) \, m_{K1} \, ts \, (List^{\wedge}K1\, (LTerm\, b)\, ts)
+    }
+\]
+and 
+ $\mathsf{m_{K1} : PredMap \, (K1)\, (LTerm^{\wedge}\,b\,K1)}$
+\[
+  \mathsf{m_{K1} \, t\, \ast\, = LTerm^{\wedge}K1\, b\, t}
+\]
+where $\mathsf{\ast}$ is the single element of $\mathsf{(K1\, t)}$. 
+The use of $\mathsf{List^{\wedge}map}$ is required in the $\mathsf{ListC}$ case because 
+$\mathsf{ListC}$ takes an argument of type $\mathsf{List\, (LTerm\, b)}$. 
+The same technique can be used to define $\mathsf{G^{\wedge}K1}$ whenever $\mathsf{G}$ has 
+a constructor of the form ($\mathsf{c : F \, (G\, a) \to G\, (K\,b)}$). 
+We only allow constructors of this form when 
+$\mathsf{F}$ is a nested type or ADT, so we are guaranteed to have a $\mathsf{F^{\wedge}map}$ function.
+
+\subsection{Liftings of $\mathsf{K1}$}
+To provide a proof of $\mathsf{G^{\wedge}\, a\, K1 \, t}$ for every term $\mathsf{t : G\, a}$, 
+we need to know that the lifting of $\mathsf{K1}$ by a type $\mathsf{H}$ is extensionally equal to $\mathsf{K1}$ on $\mathsf{H}$. 
+For example, we might need a proof that $\mathsf{Pair^{\wedge}\,a\,b\,K1\,K1}$ is equal to the predicate $\mathsf{K1}$ on pairs. 
+Given a pair $\mathsf{(x , y) : a \times b}$, we have
+
+\begin{align*}
+  &\mathsf{Pair^{\wedge}\,a\,b\,K1\,K1 (x, y)}  \\
+  = \, \, &\mathsf{K1 \, x \times K1\, y} \\ 
+  = \, \, &\mathsf{\top \times \top}
+\end{align*}
+while 
+\begin{align*}
+ \mathsf{K1\, (x, y) = \top}
+\end{align*}
+and while these types are not equal they are clearly isomorphic. So for simplicity of presentation, we 
+assume ($\mathsf{F^{\wedge}\,a\,K1}$) is equal to $\mathsf{K1}$ for every nested type and ADT $\mathsf{F}$.
 

### 2021-04-07 induction-paper/GADT-induction.tex
change foralls to lambdas in section 4.3
Add some notes
define KBool, K1, etc. to make lambda term examples and Section 6 complete

@@
 GADTs are expressive enough to represent, for example, the types and terms of the simply typed lambda calculus. 
-We will use a variant of the STLC including a base type of Bool, and type constructors for arrow types and list types:
+We will use a variant of the STLC including a base type of booleans, and type constructors for arrow types and list types. 
+We define a GADT $\mathsf{LType}$ to encode the set of permissible types for this calculus:
 
@@
 
+The $\mathsf{TBool}$ constructor represents the type of booleans in the calculus. It is 
+defined in terms of $\mathsf{KBool : \forall\, (a : Set) \to Set}$, which simply returns $\mathsf{Bool}$ on any input:
+\[
+  \mathsf{KBool \, a = Bool}
+\]
+The lifting $\mathsf{KBool^{\wedge} : \forall (a : Set) \to (a \to Set) \to (KBool \, a) \to Set}$ lifts every predicate to the constantly true predicate: 
+\[
+  \mathsf{KBool^{\wedge}\, a\, Q_a\, b = \top}
+\]
+
+% The $\mathsf{Bool}$ data type is defined as:
+% \begin{equation}\label{eq:bool}
+% \begin{array}{l}
+% \mathsf{data\, Bool\,: Set\,where}\\
+%   \mathsf{\;\;\;\;\;\;\;\;true :\, Bool} \\
+%   \mathsf{\;\;\;\;\;\;\;\;false:\, Bool} \\
+% \end{array}
+% \end{equation}
+The type system also includes an arrow type constructor, $\mathsf{TArr}$ , which takes two types $\mathsf{b}$ and $\mathsf{c}$
+and produces the arrow type $\mathsf{b \to c}$. Similarly, given a type $\mathsf{b}$, we can construct 
+the type $\mathsf{List\,b}$ using the $\mathsf{TList}$ constructor.
+
 The term calculus for this type system includes variables, abstraction, application, 
 and also includes an introduction rule that takes a list of lambda terms of type $\mathsf{b}$ and produces a lambda term of type $\mathsf{List\,b}$.
-
 \begin{equation}\label{eq:eq_lterm}
@@
 The type parameter for $\mathsf{LTerm}$ tracks the types of STLC terms. So the type $\mathsf{LTerm\,A}$ 
-contains lambda terms of type $\mathsf{A}$. The $\mathsf{LType}$ GADT is used to encode the 
-grammar of permissible types for this calculus.
-
+contains lambda terms of type $\mathsf{A}$. 
 Variables are represented as strings and are tagged with their type. 
-The $\mathsf{Abs}$ constructor takes a name (string) and a type for the bound variable,
-and a lambda term of type $\mathsf{c}$ for the body of the abstraction.
+The $\mathsf{Var}$ constructor takes a variable name and a type for the variable and produces 
+a term of the given type. 
+The $\mathsf{Abs}$ constructor takes a variable name, a type $\mathsf{b}$ for the variable,
+and a lambda term of type $\mathsf{c}$ for the body of the abstraction, and 
+it produces a term of type $\mathsf{b \to c}$. 
 The $\mathsf{App}$ constructor applies a lambda term of type $\mathsf{b \to a}$ to a lambda term of type $\mathsf{b}$, producing
@@
 The $\mathsf{ListC}$ constructor takes a list of lambda terms of type $\mathsf{b}$ and produces a lambda term of type $\mathsf{List\,b}$.
+
+Variables are tagged with their types in the $\mathsf{Var}$ (variable introduction) and 
+$\mathsf{Abs}$ (abstraction) constructors. The role of $\mathsf{LType}$ in the $\mathsf{Var}$ and $\mathsf{Abs}$ constructors
+is to enforce that variables can only be tagged with a legal type, e.g., $\mathsf{Bool}$, 
+$\mathsf{Bool \to Bool}$, $\mathsf{List\, (Bool \to Bool)}$, etc. 
+Therefore the presence of $\mathsf{LType}$ in these constructors ensures that all lambda terms produced by 
+$\mathsf{Var}$, $\mathsf{Abs}$, $\mathsf{App}$, and $\mathsf{ListC}$ are well-typed.
 
@@
 \begin{multline*}
-\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to LTerm\,a \to Set)} \\
+\mathsf{\lambda (P : \forall (a : Set) \to (a \to Set) \to LTerm\,a \to Set)} \\
 \mathsf{\to 
@@
 \begin{align*}
-  &\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to LTerm\,a \to Set)} \\
+  &\mathsf{\lambda (P : \forall (a : Set) \to (a \to Set) \to LTerm\,a \to Set)} \\
   &\mathsf{\to 
@@
 
-{\color{red} Compare with structural induction? or at least state why structural induction fails.} \\
 {\color{red} Can get rid of Maybe using non-empty lists and postulates}
@@
 \]
-where $\mathsf{K1 : a \to Set}$ and $\mathsf{LTerm^{\wedge}K1\, a\, t : LTerm^{\wedge}\, a\, K1\, t}$.
+where $\mathsf{K1 : a \to Set}$ is the constantly true predicate:
+\[
+  \mathsf{K1 \, x = \top}
+\]
+and $\mathsf{LTerm^{\wedge}K1\, a\, t : LTerm^{\wedge}\, a\, K1\, t}$.
 Notice that there is no space in $\mathsf{LTerm^{\wedge}K1}$, because
@@
 \item reference (correctly) Haskell Symposium paper
+\item reference inspiration for STLC GADT : https://www.seas.upenn.edu/~cis194/spring15/lectures/11-stlc.html
+
 \end{itemize}

### 2021-04-09 induction-paper/ENTCS-GADT-induction.tex
ENTCS version with revisions up to Section 4.2

--- 2021-04-07 induction-paper/GADT-induction.tex
@@
-% For double-blind review submission, w/o CCS and ACM Reference (max
-% submission space)
-\documentclass[acmsmall,screen,review,anonymous]{acmart}
-\settopmatter{printfolios=true,printccs=false,printacmref=false}
-%% For double-blind review submission, w/ CCS and ACM Reference
-%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
-%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
-%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-%% For single-blind review submission, w/ CCS and ACM Reference
-%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
-%% For final camera-ready submission, w/ required CCS and ACM Reference
-%\documentclass[acmsmall]{acmart}\settopmatter{}
-
-%% Journal information
-%% Supplied to authors by publisher for camera-ready submission;
-%% use defaults for review submission.
-\acmJournal{PACMPL}
-\acmVolume{1}
-\acmNumber{Haskell Symposium} % CONF = POPL or ICFP or OOPSLA
-\acmArticle{1}
-\acmYear{2021}
-\acmMonth{8}
-\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
-
-%% Copyright information
-%% Supplied to authors (based on authors' rights management selection;
-%% see authors.acm.org) by publisher for camera-ready submission;
-%% use 'none' for review submission.
-\setcopyright{none}
-%\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
-%\copyrightyear{2018}           %% If different from \acmYear
-
-%% Bibliography style
-\bibliographystyle{ACM-Reference-Format}
-%% Citation style
-%% Note: author/year citations are required for papers published as an
-%% issue of PACMPL.
-\citestyle{acmauthoryear}   %% For author/year citations
-%\citestyle{acmnumeric}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Note: Authors migrating a paper from PACMPL format to traditional
-%% SIGPLAN proceedings format must update the '\documentclass' and
-%% topmatter commands above; see 'acmart-sigplanproc-template.tex'.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\usepackage[utf8]{inputenc}
-\usepackage{ccicons}
-\usepackage{verbatim}
+\documentclass[9pt]{entcs} \usepackage{entcsmacro}
+\usepackage{graphicx}
+\sloppy
+% The following is enclosed to allow easy detection of differences in
+% ascii coding.
+% Upper-case    A B C D E F G H I J K L M N O P Q R S T U V W X Y Z
+% Lower-case    a b c d e f g h i j k l m n o p q r s t u v w x y z
+% Digits        0 1 2 3 4 5 6 7 8 9
+% Exclamation   !           Double quote "          Hash (number) #
+% Dollar        $           Percent      %          Ampersand     &
+% Acute accent  '           Left paren   (          Right paren   )
+% Asterisk      *           Plus         +          Comma         ,
+% Minus         -           Point        .          Solidus       /
+% Colon         :           Semicolon    ;          Less than     <
+% Equals        =3D           Greater than >          Question mark ?
+% At            @           Left bracket [          Backslash     \
+% Right bracket ]           Circumflex   ^          Underscore    _
+% Grave accent  `           Left brace   {          Vertical bar  |
+% Right brace   }           Tilde        ~
+
+\bibliographystyle{entcs}
+
+% A couple of exemplary definitions:
+
+\newcommand{\Nat}{{\mathbb N}}
+\newcommand{\Real}{{\mathbb R}}
+\def\lastname{Johann, Ghiorzi and Jeffries}
 
 \usepackage{amsmath}
-\usepackage{amsthm}
-\usepackage{amscd}
-%\usepackage{MnSymbol}
-\usepackage{xcolor}
-
-\usepackage{bbold}
-\usepackage{url}
-\usepackage{upgreek}
-%\usepackage{stmaryrd}
-
-\usepackage{lipsum}
-\usepackage{tikz-cd}
-\usetikzlibrary{cd}
-\usetikzlibrary{calc}
-\usetikzlibrary{arrows}
-
-\usepackage{bussproofs}
-\EnableBpAbbreviations
 
 \input{macros}
-\input{mytheorem}
-
-\theoremstyle{definition}
-\newtheorem{exmpl}{Example}
-
-\renewcommand{\greyout}[1]{}
-
-\newcommand{\new}[1]{{\color{blue} {#1}}}
-
-\newcommand{\emptyfun}{{[]}}
-\newcommand{\cal}{\mathcal}
-\newcommand{\F}{\mathcal{F}}
-\newcommand{\G}{\mathcal{G}}
-\newcommand{\N}{\mathcal{N}}
-\newcommand{\E}{\mathcal{E}}
-\newcommand{\B}{\mathcal{B}}
-\renewcommand{\P}{\mathcal{A}}
-\newcommand{\pred}{\mathsf{Fam}}
-\newcommand{\env}{\mathsf{Env}}
-\newcommand{\set}{\mathsf{Set}}
-\renewcommand{\S}{\mathcal S}
-\newcommand{\C}{\mathcal{C}}
-\newcommand{\D}{\mathcal{D}}
-\newcommand{\A}{\mathcal{A}}
-\renewcommand{\id}{\mathit{id}}
-\newcommand{\map}{\mathsf{map}}
-\newcommand{\pid}{\underline{\mathit{id}}}
-\newcommand{\pcirc}{\,\underline{\circ}\,}
-\newcommand{\pzero}{\underline{0}}
-\newcommand{\pone}{\underline{1}}
-\newcommand{\psum}{\,\underline{+}\,}
-\newcommand{\pinl}{\underline{\mathit{inL}}\,}
-\newcommand{\pinr}{\underline{\mathit{inR}}\,}
-\newcommand{\ptimes}{\,\underline{\times}\,}
-\newcommand{\ppi}{\underline{\pi_1}}
-\newcommand{\pppi}{\underline{\pi_2}}
-\newcommand{\pmu}{\underline{\mu}}
-\newcommand{\semmap}{\mathit{map}}
-\newcommand{\subst}{\mathit{subst}}
-
-\newcommand{\tb}[1]{~~ \mbox{#1} ~~}
-\newcommand{\listt}[1]{(\mu \phi. \lambda \beta . \onet + \beta \times
-  \phi \beta) #1} 
-\newcommand{\filtype}{\Nat^\emptyset 
- (\Nat^\emptyset \, \alpha \, \mathit{Bool})\, (\Nat^\emptyset 
-  (List \, \alpha) \, (List \, \alpha))} 
-\newcommand{\filtypeGRose}{\Nat^\emptyset 
- (\Nat^\emptyset \, \alpha \, \mathit{Bool})\, (\Nat^\emptyset 
-  (\mathit{GRose}\,\psi \, \alpha) \, (\mathit{GRose}\,\psi \, (\alpha
-  + \onet)))} 
-\newcommand{\maplist}{\mathit{map}_{\lambda A. \setsem{\emptyset; \alpha
-      \vdash \mathit{List} \, \alpha} \rho[\alpha := A]}} 
-\newcommand{\PLeaves}{\mathsf{PLeaves}}
-\newcommand{\swap}{\mathsf{swap}}
-\newcommand{\reverse}{\mathsf{reverse}}
-\newcommand{\Bcons}{\mathit{Bcons}}
-\newcommand{\Bnil}{\mathit{Bnil}}
-
-\title[(Deep) Induction for GADTs]{(Deep) Induction for GADTs}
-
-\author{Patricia Johann, Enrico Ghiorzi, and Daniel Jeffries}
-\affiliation{ \institution{Appalachian State University}}
-\email{johannp@appstate.edu, ghiorzie@appstate.edu,
-  jeffriesd@appstate.edu}
 
 \begin{document}
+\begin{frontmatter}
+  \title{Deep Induction for GADTs}
+
+  \author{Patricia Johann\thanksref{JohannEmail}}
+  \author{Enrico Ghiorzi\thanksref{GhiorziEmail}}
+  \author{Daniel Jeffries\thanksref{JeffriesEmail}}
+  \thanks[JohannEmail]{Email: \href{mailto:johannp@appstate.edu} {\texttt{\normalshape johannp@appstate.edu}}}
+  \thanks[GhiorziEmail]{Email: \href{mailto:ghiorzie@appstate.edu} {\texttt{\normalshape ghiorzie@appstate.edu}}}
+  \thanks[JeffriesEmail]{Email: \href{mailto:jeffriesd@appstate.edu} {\texttt{\normalshape jeffriesd@appstate.edu}}}
+  \address{Department of Computer Science\\ Appalachian State University}
 
 \begin{abstract} 
-Abstract
+  Abstract goes here.
 \end{abstract}
 
-\maketitle
+\begin{keyword}
+  Please list keywords from your paper here, separated by commas.
+\end{keyword}
+
+\end{frontmatter}
 
 \section{Introduction}\label{sec:intro}
+
+Introduction goes here.
 
@@
 the type of lists (here coded in Agda)
-{\color{blue}(Ask Daniel which flavor of syntax, paper as literate Agda, naming conventions?)}
 \begin{equation}\label{eq:list}
 \begin{array}{l}
-%\mathsf{data\, List : Set \to Set\,where}\\
-\mathsf{data\, List\,(a : Set) : Set\,where}\\
-\mathsf{\;\;\;\;\;\;\;\;Nil\,\,\,\,\,\, :\, List\,a}\\
-\mathsf{\;\;\;\;\;\;\;\;Cons\, :\, a \to List\,a \to List\,a} 
+\mathsf{data\ List : Set \to Set\ where}\\
+\mathsf{\;\;\;\;\;\;\;nil\,\,\,\,\,\, :\, List\,A}\\
+\mathsf{\;\;\;\;\;\;\;cons\, :\, A \to List\,A \to List\,A} 
 \end{array}
 \end{equation}
-Notice that all occurrences of $\mathsf{List}$ in the above encoding are instantiated at the same index $\mathsf{a}$.
+Notice that all occurrences of $\mathsf{List}$ in the above encoding are instantiated at the same index $\mathsf{A}$.
 Thus, the instances of $\mathsf{List}$ at various indices are defined independently from one another.
@@
 \begin{array}{l}
-\mathsf{data\, PTree\,(a : Set) : Set\,where}\\
-\mathsf{\;\;\;\;\;\;\;\;PLeaf\,\,\, :\, a \to PTree\,a}\\
-\mathsf{\;\;\;\;\;\;\;\;PNode\, :\, PTree\,(a \times a) \to PTree\,a} 
+\mathsf{data\ PTree : Set \to Set\ where}\\
+\mathsf{\;\;\;\;\;\;\;pLeaf\,\,\, :\, A \to PTree\,A}\\
+\mathsf{\;\;\;\;\;\;\;pNode\, :\, PTree\,(A \times A) \to PTree\,A} 
 \end{array}
@@
 of perfect trees, which can be thought of as constraining lists to have lengths that are powers of 2.
-In the above code, the constructor $\mathsf{PNode}$ uses data of type $\mathsf{PNode\,(a \times a)}$ to construct data of type $\mathsf{PNode\,a}$.
+In the above code, the constructor $\mathsf{PNode}$ uses data of type $\mathsf{PNode\,(A \times A)}$ to construct data of type $\mathsf{PNode\,A}$.
 Thus, it is clear that the instantiations of $\mathsf{PNode}$ at various indices cannot be defined independently, so that the entire family of types must actually be defined at once.
@@
 \begin{array}{l}
-\mathsf{data\, Bush\,(a : Set) : Set\,where}\\
-\mathsf{\;\;\;\;\;\;\;\;BNil\,\,\,\,\,\, :\, Bush\,a}\\
-\mathsf{\;\;\;\;\;\;\;\;BCons\, :\, a \to Bush\,(Bush\,a) \to Bush\,a} 
+\mathsf{data\ Bush : Set \to Set\ where}\\
+\mathsf{\;\;\;\;\;\;bNil\,\,\,\,\,\, :\, Bush\,A}\\
+\mathsf{\;\;\;\;\;\;bCons\, :\, A \to Bush\,(Bush\,A) \to Bush\,A} 
 \end{array}
@@
 the return types of its constructors still have to be the same type instance of the type being defined.
-In other words, all constructors of $\mathsf{PTree\,a}$ have to return an element of type $\mathsf{PTree\,a}$,
-and all constructors of $\mathsf{Bush\,a}$ have to return an element of type $\mathsf{Bush\,a}$.
+In other words, all constructors of $\mathsf{PTree\,A}$ have to return an element of type $\mathsf{PTree\,A}$,
+and all constructors of $\mathsf{Bush\,A}$ have to return an element of type $\mathsf{Bush\,A}$.
 %the return types of the constructors are still required to be instantiated at the same variable $\mathsf{a}$.}
@@
 
-An induction principle for a data type allows proving
+An induction principle for a data type allows to prove
 that a predicate holds for every element of that data type,
@@
 In this paper, we are interested in induction principles for proof-relevant predicates.
-A proof-relevant predicate on a type $\mathsf{a}$ is a function $\mathsf{a \to Set}$
-(where $\mathsf{Set}$ is the type of sets)
-mapping each $\mathsf{x : a}$ to the set of proofs that the predicate holds for $\mathsf{x}$.
+A proof-relevant predicate on a type $\mathsf{A : Set}$ is a function $\mathsf{A \to Set}$
+%(where $\mathsf{Set}$ is the type of sets)
+mapping each $\mathsf{x : A}$ to the set of proofs that the predicate holds for $\mathsf{x}$.
 For example, the induction principle for $\mathsf{List}$ is
-\begin{multline*}
-\mathsf{\forall (a : Set) (P : List\,a \to Set) \to P\,Nil \to \big( \forall (x : a) (xs: List\,a) \to P\,xs \to P\,(Cons\,x\,xs)\big)} \\
-\mathsf{\to \forall (xs : List\,a) \to P\, xs }
-\end{multline*}
+\begin{equation*}
+\mathsf{
+\forall (A : Set) (P : List\,A \to Set)
+\to P\,Nil
+\to \big( \forall (a : A) (ls: List\,A)
+\to P\,ls
+\to P\,(Cons\,a\,ls)\big)
+\to \forall (ls : List\,A)
+\to P\, ls
+}
+\end{equation*}
 Note that the data inside a structure of type $\mathsf{List}$ is treated monolithically (i.e., ignored) by this induction rule.
@@
 Thus, the induction rule for $\mathsf{PTree}$ is
-\begin{multline*}
-\mathsf{\forall (P : \forall (a : Set) \to PTree\, a \to Set)
-\to \big( \forall (a : Set) (x : a) \to P\,a\,(PLeaf\, x) \big)} \\
-\mathsf{\to \big( \forall (a : Set) (x : PTree\,(a \times a)) \to P\,(a \times a)\,x \to P\,a\,(PNode\,x)\big)} \\
-\mathsf{\to \forall (a : Set) (x : PTree\,a) \to P\,a\,x }
-\end{multline*}
+\[
+\begin{array}{l}
+\mathsf{\forall (P : \forall (A : Set) \to PTree\, A \to Set)
+\to \big( \forall (A : Set) (a : A) \to P\,A\,(PLeaf\, a) \big)} \\
+\quad\mathsf{\to \big( \forall (A : Set) (pt : PTree\,(A \times A)) \to P\,(A \times A)\,pt \to P\,a\,(PNode\,pt)\big)
+\to \forall (A : Set) (pt : PTree\,A) \to P\,A\,pt }
+\end{array}
+\]
 
@@
 For example, the deep induction rule for $\mathsf{Bush}$ is
-\begin{multline*}
-\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to Bush\, a \to Set)
-\to \big( \forall (a : Set) \to P\,a\,BNil \big)} \\
-\mathsf{\to \big( \forall (a : Set) (Q : a \to Set) (x : a) (y : Bush\,(Bush\,a))} \\
-\mathsf{\to Q\,x \to P\,(Bush\,a)\,(Bush^{\wedge}\,a\,Q)\,y \to P\,a\,Q\,(BCons\,x\,y) \big)} \\
-\mathsf{\to \forall (a : Set) (Q : a \to Set) (x : Bush\,a) \to Bush^{\wedge}\,a\,Q\,x \to P\,a\,Q\,x }
-\end{multline*}
-where $\mathsf{Bush^{\wedge} : \forall (a : Set) \to (a \to Set) \to Bush\,a \to Set}$ lifts a predicate $\mathsf{Q}$ on data of type $\mathsf{a}$ to a predicate on data of type $\mathsf{Bush\,a}$ asserting that $\mathsf{Q}$ holds for every element of type $\mathsf{a}$ contained in its argument bush.
-%where $\mathsf{Bush^{\wedge} : \forall (a : Set) \to ( a \to Set ) \to Bush\,a \to Set}$
-%is the polymorphic function that lifts predicates to $\mathsf{Bush}$,
+\[
+\begin{array}{l}
+\mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to Bush\, A \to Set)
+\to \big( \forall (A : Set) \to P\,A\,BNil \big)} \\
+\quad\mathsf{\to \big( \forall (A : Set) (Q : A \to Set) (a : A) (bb : Bush\,(Bush\,A))
+\to Q\,a \to P\,(Bush\,A)\,(Bush^{\wedge}\,A\,Q)\,bb \to P\,A\,Q\,(BCons\,a\,bb) \big)} \\
+\quad\mathsf{\to \forall (A : Set) (Q : A \to Set) (b : Bush\,A) \to Bush^{\wedge}\,A\,Q\,b \to P\,A\,Q\,b }
+\end{array}
+\]
+where $\mathsf{Bush^{\wedge} : \forall (A : Set) \to (A \to Set) \to Bush\,A \to Set}$ lifts a predicate $\mathsf{Q}$ on data of type $\mathsf{A}$ to a predicate on data of type $\mathsf{Bush\,A}$ asserting that $\mathsf{Q}$ holds for every element of type $\mathsf{A}$ contained in its argument bush.
 It is defined as
 \begin{align*}
-\mathsf{Bush^{\wedge}\,a\,Q\,BNil} &= \mathsf{1} \\
-\mathsf{Bush^{\wedge}\,a\,Q\,(BCons\,x\,y)} &= \mathsf{Q\,x \times Bush^{\wedge}\,(Bush\,a)\,(Bush^{\wedge}\,a\,Q)\,y}
+\mathsf{Bush^{\wedge}\,A\,Q\,BNil} &= \mathsf{\top} \\
+\mathsf{Bush^{\wedge}\,A\,Q\,(BCons\,a\,bb)} &= \mathsf{Q\,a \times Bush^{\wedge}\,(Bush\,A)\,(Bush^{\wedge}\,A\,Q)\,bb}
 \end{align*}
+where $\mathsf{Q : A \to Set}$ and $\mathsf{bb : Bush\,(Bush\,A)}$.
 
@@
 For example, the deep induction principle for $\mathsf{List}$ is
-\begin{multline*}
-\mathsf{\forall (a : Set) (P : List\, a \to Set) (Q : a \to Set)} \\
-\mathsf{\to P\,Nil \to \big( \forall (x : a) (xs: List\,a) \to Q\,x \to P\,xs \to P\,(Cons\,x\,xs)\big)} \\
-\mathsf{\to \forall (xs : List\,a) \to List^{\wedge}\,a\,Q\,xs \to P\, xs }
-\end{multline*}
-where $\mathsf{List^{\wedge} : \forall (a : Set) \to (a \to Set) \to List\,a \to Set}$ lifts a predicate $\mathsf{Q}$ on data of type $\mathsf{a}$ to a predicate on data of type $\mathsf{List\,a}$ asserting that $\mathsf{Q}$ holds for every element of its argument list.
-Finally, the deep induction rule for $\mathsf{PTree}$ is
-\begin{multline*}
-\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to PTree\, a \to Set)} \\
-\mathsf{\to \big( \forall (a : Set) (Q : a \to Set) (x : a) \to Q\,x \to P\,a\,Q\,(PLeaf\, x) \big)} \\
-\mathsf{\to \big( \forall (a : Set) (Q : a \to Set) (x : PTree\,(a \times a)) \to P\,(a \times a)\,(Pair^{\wedge}\,a\,Q)\,x \to P\,a\,Q\,(PNode\,x)\big)} \\
-\mathsf{\to \forall (a : Set) (Q : a \to Set) (x : PTree\,a) \to PTree^{\wedge}\,a\,Q\,x \to P\,a\,Q\,x }
-\end{multline*}
-where $\mathsf{Pair^{\wedge} : \forall (a : Set) \to (a \to Set) \to a \times a \to Set}$ lifts a predicate $\mathsf{Q}$ on $\mathsf{a}$ to a predicate on pairs of type $\mathsf{a \times a}$,
-so that $\mathsf{Pair^{\wedge}\,a\,Q\,(x,y) = Q\,x \times Q\,y}$,
-and $\mathsf{PTree^{\wedge} : \forall (a : Set) \to (a \to Set) \to PTree\,a \to Set}$ lifts a predicate $\mathsf{Q}$ on data of type $\mathsf{a}$ to a predicate on data of type $\mathsf{PTree\,a}$ asserting that $\mathsf{Q}$ holds for every element of type $\mathsf{a}$ contained in its argument perfect tree.
-
-{\color{blue} Moreover, for types admitting both deep induction and structural induction, the former generalizes the latter.
-Indeed, structural induction rules can be derived from deep induction rules by choosing the constantly true predicate as the custom predicate traversing each data element contained in the data type.
+\[
+\begin{array}{l}
+\mathsf{\forall (A : Set) (P : List\, A \to Set) (Q : A \to Set)
+\to P\,Nil \to \big( \forall (a : A) (ls: List\,A) \to Q\,a \to P\,ls \to P\,(Cons\,a\,ls)\big)} \\
+\quad\mathsf{\to \forall (ls : List\,A) \to List^{\wedge}\,A\,Q\,ls \to P\,ls }
+\end{array}
+\]
+where $\mathsf{List^{\wedge} : \forall (A : Set) \to (A \to Set) \to List\,A \to Set}$ lifts a predicate $\mathsf{Q}$ on data of type $\mathsf{A}$ to a predicate on data of type $\mathsf{List\,A}$ asserting that $\mathsf{Q}$ holds for every element of its argument list.
+
+As a further example, the deep induction rule for $\mathsf{PTree}$ is
+\[
+\begin{array}{l}
+\mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to PTree\,A \to Set)
+  \to \big( \forall (A : Set) (Q : A \to Set) (a : A) \to Q\,a \to P\,A\,Q\,(PLeaf\, a) \big)} \\
+\quad \mathsf{\to \big( \forall (A : Set) (Q : A \to Set) (a : PTree\,(A \times A)) \to P\,(A \times A)\,(Pair^{\wedge}\,A\,A\,Q\,Q)\,a \to P\,A\,Q\,(PNode\,a)\big)} \\
+\quad \mathsf{\to \forall (A : Set) (Q : A \to Set) (a : PTree\,a) \to PTree^{\wedge}\,A\,Q\,a \to P\,A\,Q\,a }
+\end{array}
+\]
+where $\mathsf{Pair^{\wedge} : \forall (A : Set) (B : Set) \to (A \to Set) \to (B \to Set) \to A \times B \to Set}$ lifts predicates $\mathsf{Q_A}$ on data of type $\mathsf{A}$ and $\mathsf{Q_B}$ on data of type $\mathsf{B}$ to a predicate on pairs of type $\mathsf{A \times B}$,
+so that $\mathsf{Pair^{\wedge}\,A\,B\,Q_A\,Q_B\,(a,b) = Q_A\,a \times Q_B\,b}$,
+and $\mathsf{PTree^{\wedge} : \forall (A : Set) \to (A \to Set) \to PTree\,A \to Set}$ lifts a predicate $\mathsf{Q}$ on data of type $\mathsf{A}$ to a predicate on data of type $\mathsf{PTree\,A}$ asserting that $\mathsf{Q}$ holds for every element of type $\mathsf{A}$ contained in its argument perfect tree.
+
+Moreover, for types admitting both deep induction and structural induction, the former generalizes the latter.
+Indeed, structural induction rules can be derived from deep induction rules by always choosing the constantly true predicate as the custom predicate traversing each data element contained in the data type.
 That way, deep induction only inspects the structure of the data type and not its content, just like structural induction does.
-A concrete example of this technique will be demonstrated in Section~\ref{sec:ind-equal}.}
+A concrete example of such derivation is demonstrated in Section~\ref{sec:ind-equal}.
 
@@
 
-%\subsection{Agda syntax (bad title, revise)}
-
+\begin{example}\label{example:equal}
 As a first and notable example of GADT, we consider the the $\mathsf{Equal}$ type.
@@
 \begin{array}{l}
-\mathsf{data\, Equal\,(a\,b : Set) : Set\,where}\\
-\mathsf{\;\;\;\;\;\;\;\;Refl :\, Equal\,c\,c}
+\mathsf{data\ Equal : Set \to Set \to Set\ where}\\
+\mathsf{\;\;\;\;\;\;\;\;\;\;\;\,refl :\, Equal\,A\,A}
 \end{array}
@@
 are the same and no data elements otherwise.
-
-A more complex example for a GADT is
+The importance of the $\mathsf{Equal}$ type lies in the fact that we can understand other GADTs in terms of equality,
+as shown in the examples below.
+\end{example}
+
+\begin{example}\label{example:seq}
+A more complex example of GADT is the type of sequences,
 \begin{equation}\label{eq:eq-seq}
 \begin{array}{l}
-\mathsf{data\, Seq\,(a : Set) : Set\,where}\\
-\mathsf{\;\;\;\;\;\;\;\;Const :\, a \to Seq\,a}\\
-\mathsf{\;\;\;\;\;\;\;\;SPair\,\, :\, Seq \,a \to Seq\,b \to
-  Seq\,(a \times b)}
+\mathsf{data\ Seq : Set \to Set\ where}\\
+\mathsf{\;\;\;\;\;\;\;\;const :\, A \to Seq\,A}\\
+\mathsf{\;\;\;\;\;\;\;\;sPair\,\, :\, Seq\,A \to Seq\,B \to Seq\,(A \times B)}
 \end{array}
@@
 \begin{array}{l}
-\mathsf{data\, Seq\,(a : Set) : Set\,where}\\
-\mathsf{\;\;\;\;\;\;\;\;Const :\, a \to Seq\,a}\\
-\mathsf{\;\;\;\;\;\;\;\;SPair\,\, :\, \exists (b\,c : Set).\,Equal\,a\,(b \times c) \to Seq\,b \to Seq\,c \to
-  Seq\,a}\\ 
+\mathsf{data\ Seq : Set \to Set\ where}\\
+\mathsf{\;\;\;\;\;\;\;\;const :\, A \to Seq\,A}\\
+\mathsf{\;\;\;\;\;\;\;\;sPair\,\, :\, \forall (B\,C : Set) \to Equal\,A\,(B \times C)
+  \to Seq\,B \to Seq\,C \to Seq\,A}\\ 
 \end{array}
 \end{equation}
-where the requirement that the $\mathsf{SPair}$ constructor produces an instance of $\mathsf{Seq}$ at a product type has been replaced with the requirement that the instance of $\mathsf{Seq}$ returned by $\mathsf{SPair}$ is \emph{equal} to some product type.
+where the requirement that the $\mathsf{sPair}$ constructor produces an instance of $\mathsf{Seq}$ at a product type has been replaced with the requirement that the instance of $\mathsf{Seq}$ returned by $\mathsf{sPair}$ is \emph{equal} to some product type.
 This encoding is particularly convenient when representing GADTs as Church encodings~\cite{vw10,atk12}.
-
-%{\color{blue} Do we really need the existential quantifiers?}
+\end{example}
 
@@
 
-{\color{red} STLC example}
-
+\begin{example}\label{example:lterm}
 GADTs are expressive enough to represent, for example, the types and terms of the simply typed lambda calculus. 
-We will use a variant of the STLC including a base type of Bool, and type constructors for arrow types and list types:
-
+We will use a variant of the simply typed lambda calculus including a base type of booleans,
+and type constructors for arrow types and list types. 
+We define a GADT $\mathsf{LType}$ to encode the set of permissible types for this calculus:
 \begin{equation}\label{eq:eq_ltype}
 \begin{array}{l}
-\mathsf{data\, LType\,(a : Set) : Set\,where}\\
-\mathsf{\;\;\;\;\;\;\;\;TBool :\, \exists (b : Set). \,Equal\,a\,(KBool\, b) \to LType\,a}\\
-\mathsf{\;\;\;\;\;\;\;\;TArr\,\,\, :\, \exists (b : Set) (c : Set).\,Equal\,a\,(b \to c) \to LType\,b \to LType\,c \to LType\,a}\\
-  \mathsf{\;\;\;\;\;\;\;\;TList\,\, :\, \exists (b : Set).\,Equal\,a\,(\List\,b) \to LType\,b \to LType\,a}
+\mathsf{data\ LType : Set \to Set\ where}\\
+\mathsf{\;\;\;\;\;\;\;\;tBool :\, \forall (B : Set) \to Equal\,A\,Bool \to LType\,A}\\
+\mathsf{\;\;\;\;\;\;\;\;tArr\,\,\, :\, \forall (B\,C : Set) \to Equal\,A\,(B \to C) \to LType\,B \to LType\,C \to LType\,A}\\
+  \mathsf{\;\;\;\;\;\;\;\;tList\,\, :\, \forall (B : Set) \to Equal\,A\,(\List\,B) \to LType\,B \to LType\,A}
 \end{array}
 \end{equation}
+The type system includes a $\mathsf{tBool}$ constructor representing the type of booleans;
+an arrow type constructor, $\mathsf{tArr}$ , which takes two types $\mathsf{B}$ and $\mathsf{C}$
+and produces the arrow type $\mathsf{B \to C}$;
+and a $\mathsf{tList}$ constructor which takes a type $\mathsf{B}$ and constructs
+the type $\mathsf{List\,B}$.
+
+%The $\mathsf{tBool}$ constructor represents the type of booleans in the calculus.
+%%It is defined in terms of $\mathsf{KBool : Set \to Set}$,
+%%which simply returns $\mathsf{Bool}$ on any input: $\mathsf{KBool \, A = Bool}$.
+%%The lifting $\mathsf{KBool^{\wedge} : \forall (A : Set) \to (A \to Set) \to (KBool \, A) \to Set}$ lifts every predicate to the constantly true predicate: 
+%%\[
+%%  \mathsf{KBool^{\wedge}\, A\, Q\, b = \top}
+%%\]
+%% The $\mathsf{Bool}$ data type is defined as:
+%% \begin{equation}\label{eq:bool}
+%% \begin{array}{l}
+%% \mathsf{data\, Bool\,: Set\,where}\\
+%%   \mathsf{\;\;\;\;\;\;\;\;true :\, Bool} \\
+%%   \mathsf{\;\;\;\;\;\;\;\;false:\, Bool} \\
+%% \end{array}
+%% \end{equation}
+%The type system also includes an arrow type constructor, $\mathsf{tArr}$ , which takes two types $\mathsf{B}$ and $\mathsf{C}$
+%and produces the arrow type $\mathsf{B \to C}$.
+%Similarly, given a type $\mathsf{B}$, we can construct
+%the type $\mathsf{List\,B}$ using the $\mathsf{tList}$ constructor.
 
 The term calculus for this type system includes variables, abstraction, application, 
-and also includes an introduction rule that takes a list of lambda terms of type $\mathsf{b}$ and produces a lambda term of type $\mathsf{List\,b}$.
-
+and also includes an introduction rule that takes a list of lambda terms of type $\mathsf{B}$ and produces a lambda term of type $\mathsf{List\,B}$.
 \begin{equation}\label{eq:eq_lterm}
 \begin{array}{l}
-\mathsf{data\, LTerm\,(a : Set) : Set\,where}\\
-\mathsf{\;\;\;\;\;\;\;\;Var\,\,\,\,\,\,:\, String \to LType\,a \to LTerm\,a} \\
-\mathsf{\;\;\;\;\;\;\;\;Abs\,\,\,\,\,\, :\, \exists (b : Set) (c : Set).\,Equal\,a\,(b \to c) \to String 
-      \to LType\,b \to LTerm\,c \to LTerm\,a}\\
-  \mathsf{\;\;\;\;\;\;\;\;App\,\,\,\,\, :\, \exists (b : Set).\,LTerm (b \to a) \to LTerm\,b \to LTerm\,a}\\
-  \mathsf{\;\;\;\;\;\;\;\;ListC\,\, :\, \exists (b : Set).\,Equal\,a\,(\List\,b) \to List\,(LTerm\,b) \to LTerm\,a}
+\mathsf{data\ LTerm : Set \to Set\ where}\\
+\mathsf{\;\;\;\;\;\;\;\;var\,\,\,\,\,\,\,:\, String \to LType\,A \to LTerm\,A} \\
+\mathsf{\;\;\;\;\;\;\;\;abs\,\,\,\,\,\, :\, \forall (B\,C : Set) \to Equal\,A\,(B \to C) \to String 
+      \to LType\,B \to LTerm\,C \to LTerm\,A}\\
+  \mathsf{\;\;\;\;\;\;\;\;app\,\,\,\,\, :\, \forall (B : Set) \to LTerm (B \to A) \to LTerm\,B \to LTerm\,A} \\
+  \mathsf{\;\;\;\;\;\;\;\;listC\,\,\, :\, \forall (B : Set) \to Equal\,A\,(List\,B) \to List\,(LTerm\,B) \to LTerm\,A}
 \end{array}
 \end{equation}
-
-The type parameter for $\mathsf{LTerm}$ tracks the types of STLC terms. So the type $\mathsf{LTerm\,A}$ 
-contains lambda terms of type $\mathsf{A}$. The $\mathsf{LType}$ GADT is used to encode the 
-grammar of permissible types for this calculus.
-
+The type parameter for $\mathsf{LTerm}$ tracks the types of simply typed lambda calculus terms.
+So the type $\mathsf{LTerm\,A}$ contains lambda terms of type $\mathsf{A}$. 
 Variables are represented as strings and are tagged with their type.
-The $\mathsf{Abs}$ constructor takes a name (string) and a type for the bound variable,
-and a lambda term of type $\mathsf{c}$ for the body of the abstraction.
-The $\mathsf{App}$ constructor applies a lambda term of type $\mathsf{b \to a}$ to a lambda term of type $\mathsf{b}$, producing
-a lambda term of type $\mathsf{a}$. 
-The $\mathsf{ListC}$ constructor takes a list of lambda terms of type $\mathsf{b}$ and produces a lambda term of type $\mathsf{List\,b}$.
-
-\section{(Deep) induction for GADTs}
+The $\mathsf{var}$ constructor takes a variable name and a type for the variable and produces
+a term of the given type.
+The $\mathsf{abs}$ constructor takes a variable name, a type $\mathsf{B}$ for the variable,
+and a lambda term of type $\mathsf{C}$ for the body of the abstraction,
+and it produces a term of type $\mathsf{B \to C}$.
+The $\mathsf{app}$ constructor applies a lambda term of type $\mathsf{B \to A}$ to a lambda term of type $\mathsf{B}$,
+producing a lambda term of type $\mathsf{A}$. 
+The $\mathsf{listC}$ constructor takes a list of lambda terms of type $\mathsf{B}$ and produces a lambda term of type $\mathsf{List\,B}$.
+
+Variables are tagged with their types in the $\mathsf{var}$ (variable introduction) and 
+$\mathsf{abs}$ (abstraction) constructors.
+The role of $\mathsf{LType}$ in the $\mathsf{var}$ and $\mathsf{abs}$ constructors
+is to enforce that variables can only be tagged with a legal type, e.g., $\mathsf{Bool}$, 
+$\mathsf{Bool \to Bool}$, $\mathsf{List\, (Bool \to Bool)}$, etc. 
+Therefore the presence of $\mathsf{LType}$ in these constructors ensures that all lambda terms produced by 
+$\mathsf{var}$, $\mathsf{abs}$, $\mathsf{app}$, and $\mathsf{listC}$ are well-typed. 
+\end{example}
+
+\section{Deep induction for GADTs}
 
@@
 
-As we have seen in Section~\ref{ssec:induction-nested},
-truly nested types do not support a structural induction rule,
-which is the reason why it was necessary to introduce a deep induction rule supporting them.
-Consequently, GADTs do not support a structural induction rule either,
-as they generalize nested types.
-Still, there is hope for GADTs to support a deep induction rule,
-like nested types do.
+%As we have seen in Section~\ref{ssec:induction-nested},
+%truly nested types do not support a structural induction rule,
+%which is the reason why it was necessary to introduce a deep induction rule supporting them.
+%Consequently, GADTs do not support a structural induction rule either,
+%as they generalize nested types.
+%Still, there is hope for GADTs to support a deep induction rule,
+%like nested types do.
+
+Reasoning about GADTs can generally be quite difficult,
+even more so than for ADTs and nested types.
+To help with that,
+we want to extend the deep induction principle seen in Section~\ref{ssec:induction-nested} to GADTs.
 
@@
 which are essentially unary relations.
+{\color{blue} Is there a reference for this?}
 %in the sense that induction involves predicates, which are unary relations,
@@
 Nevertheless, this paper shows how to extend deep induction to some GADTs.
-We will first demonstrate how to derive the deep induction rule for some example GADTs,
-and then provide a general principle that works for GADTs not featuring nesting in their definition.
-
-\subsection{(Deep) induction for $\mathsf{Equal}$}\label{sec:ind-equal}
+We will first demonstrate how to derive the deep induction rule for some example cases,
+and then provide a general principle that works for generic GADTs,
+as long as these do not feature nesting in their definition.
+
+\subsection{Deep induction for $\mathsf{Equal}$}\label{sec:ind-equal}
 
@@
 
-As a first example, we derive the induction rule for the $\mathsf{Equal}$ type from Equation~\ref{eq:equal}.
+As a first example, we derive the induction rule for the $\mathsf{Equal}$ type from example~\ref{example:equal}.
 This will provide a simple case study that will inform the investigation of more complex GADTs.
@@
 To define an induction rule for a type $\mathsf{G}$ we first need a predicate-lifting operation
-which takes predicates on a type $\mathsf{a}$ and lifts them to predicates on $\mathsf{G\,a}$.
+which takes predicates on a type $\mathsf{A}$ and lifts them to predicates on $\mathsf{G\,A}$.
 the predicate-lifting function for $\mathsf{Equal}$ is the function
 \[
-\mathsf{Equal^{\wedge} : \forall (a\,b : Set) \to (a \to Set) \to (b \to Set) \to Equal\,a\,b \to Set}
+\mathsf{Equal^{\wedge} : \forall (A\,B : Set) \to (A \to Set) \to (B \to Set) \to Equal\,A\,B \to Set}
 \]
@@
 \[
-\mathsf{Equal^{\wedge}\,a\,a\,Q\,Q'\,Refl = \forall (x : a) \to Equal\,(Q\,x)(Q'\,x)}
+\mathsf{Equal^{\wedge}\,A\,A\,Q\,Q'\,Refl = \forall (a : A) \to Equal\,(Q\,a)(Q'\,a)}
 \]
@@
 to the expression that a given predicate is preserved by such constructor.
-Let $\mathsf{CRefl}$ be the following function associated to the $\mathsf{Refl}$ constructor:
-\begin{multline*}
-\mathsf{\lambda (P : \forall (a\,b : Set) \to (a \to Set) \to (b \to Set) \to Equal\,a\,b \to Set)} \\
-\mathsf{\to \forall (c : Set) (Q : c \to Set) (Q' : c \to Set) \to Equal^{\wedge}\,c\,c\,Q\,Q'\,Refl \to P\,c\,c\,Q\,Q'\,Refl}
-\end{multline*}
+Let $\mathsf{dIndRefl}$ be the following function associated to the $\mathsf{refl}$ constructor:
+\[
+\begin{array}{l}
+\mathsf{\lambda (P : \forall (A\,B : Set) \to (A \to Set) \to (B \to Set) \to Equal\,A\,B \to Set)} \\
+\quad\mathsf{\to \forall (C : Set) (Q\, Q' : C \to Set) \to Equal^{\wedge}\,C\,C\,Q\,Q'\,refl \to P\,C\,C\,Q\,Q'\,refl}
+\end{array}
+\]
 
@@
 The induction rule for $\mathsf{Equal}$ is thus the type
-\begin{multline*}
-\mathsf{\forall (P : \forall (a\,b : Set) \to (a \to Set) \to (b \to Set) \to Equal\,a\,b \to Set)} \\
-%\mathsf{\to \big( \forall (c : Set) (Q : c \to Set) (Q' : c \to Set) \to Equal^{\wedge}\,c\,c\,Q\,Q'\,Refl \to P\,c\,c\,Q\,Q'\,Refl \big)} \\
-\mathsf{\to CRefl\,P \to \forall (a\,b : Set) (Q_a : a \to Set) (Q_b : b \to Set) (e: Equal\,a\,b)} \\
-	\mathsf{\to Equal^{\wedge}\,a\,b\,Q_a\,Q_b\,e \to P\,a\,b\,Q_a\,Q_b\,e}
-\end{multline*}
+\[
+\begin{array}{l}
+\mathsf{\forall (P : \forall (A\,B : Set) \to (A \to Set) \to (B \to Set) \to Equal\,A\,B \to Set)} \\
+\quad \mathsf{\to dIndRefl\,P \to \forall (A\,B : Set) (Q_A : A \to Set) (Q_B : B \to Set) (e: Equal\,A\,B)
+  \to Equal^{\wedge}\,A\,B\,Q_A\,Q_B\,e \to P\,A\,B\,Q_A\,Q_B\,e}
+\end{array}
+\]
 To validate the induction rule we need to provide it with a witness, i.e.,
 we need to show that the associated type is inhabited.
-We define a term $\mathsf{DIEqual}$ of the above type as
-\[
-\mathsf{DIEqual\,P\,crefl\,a\,a\,Q_a\,Q_a'\,Refl\,L_E = crefl\,a\,Q_a\,Q_a'\,L_E}
-\]
-where
-$\mathsf{crefl : CRefl\,P}$,
-$\mathsf{Q_a : a \to Set}$, $\mathsf{Q_a' : a \to Set}$
-and $\mathsf{L_E : Equal^{\wedge}\,a\,a\,Q_a\,Q_a'\,Refl}$.
+We thus need to define a term $\mathsf{dIndEqual}$ of the above type,
+i.e., we need to define
+\[
+\mathsf{dIndEqual\,P\,crefl\,A\,B\,Q_A\,Q_B\,e\,liftE : P\,A\,B\,Q_A\,Q_B\,e}
+\]
+for $\mathsf{crefl : dIndRefl\,P}$,
+$\mathsf{Q_A : A \to Set}$, $\mathsf{Q_B : B \to Set}$,
+$\mathsf{e : Equal\,A\,B}$
+and $\mathsf{liftE : Equal^{\wedge}\,A\,B\,Q_A\,Q_B\,e}$.
+By pattern-matching, we only need to consider the case where $\mathsf{A = B}$ and $\mathsf{e = refl}$,
+and define
+\[
+\mathsf{dIndEqual\,P\,crefl\,A\,A\,Q_A\,Q_A'\,refl\,liftE = crefl\,A\,Q_A\,Q_A'\,liftE}
+\]
+%where
+%$\mathsf{Q_A : A \to Set}$, $\mathsf{Q_A' : A \to Set}$
+%and $\mathsf{L_E : Equal^{\wedge}\,A\,A\,Q_A\,Q_A'\,Refl}$.
 Having provided a well-defined term for it,
@@
 
-The type $\mathsf{Equal}$ also has a standard structural induction rule $\mathsf{SIEqual}$,
-\begin{multline*}
-\mathsf{\forall (Q : \forall (a\,b : Set) \to Equal\,a\,b \to Set)} \\
-\mathsf{\to \big( \forall (c : Set) \to P\,c\,c\,Refl \big)
-	\to \forall (a\,b : Set) (e: Equal\,a\,b) \to P\,a\,b\,e}
-\end{multline*}
+The type $\mathsf{Equal}$ also has a standard structural induction rule $\mathsf{indEqual}$,
+\[
+\mathsf{\forall (Q : \forall (A\,B : Set) \to Equal\,A\,B \to Set)
+  \to \big( \forall (C : Set) \to P\,C\,C\,refl \big)
+  \to \forall (A\,B : Set) (e: Equal\,A\,B) \to P\,A\,B\,e}
+\]
 As is the case for ADTs and nested types,
 the structural induction rule for $\mathsf{Equal}$ is a consequence of the deep induction rule.
-Indeed, we can define $\mathsf{SIEqual}$ as
-\[
-\mathsf{SIEqual\,Q\,srefl\,a\,b\,e = DIEqual\,P\,srefl\,a\,b\,K^a_1\,K^b_1\,e\,L_E}
+Indeed, we can define $\mathsf{indEqual}$ as
+\[
+\mathsf{indEqual\,Q\,srefl\,A\,B\,e = dIndEqual\,P\,srefl\,A\,B\,K^A_\top\,K^B_\top\,e\,liftE}
 \]
 where
-$\mathsf{Q : \forall (a\,b : Set) \to Equal\,a\,b \to Set}$,
-$\mathsf{srefl: \forall (c : Set) \to P\,c\,c\,Refl}$
-and $\mathsf{e : Equal\,a\,b}$, and
+$\mathsf{Q : \forall (A\,B : Set) \to Equal\,A\,B \to Set}$,
+$\mathsf{srefl: \forall (C : Set) \to P\,C\,C\,refl}$
+and $\mathsf{e : Equal\,A\,B}$, and
 \begin{itemize}
-\item $\mathsf{P : \forall (a\,b : Set) \to (a \to Set) \to (b \to Set) \to Equal\,a\,b \to Set}$ is defined as
-$\mathsf{P\,a\,b\,Q_a\,Q_b\,e = Q\,a\,b\,e}$;
-\item $\mathsf{K^a_1}$ and $\mathsf{K^b_1}$ are the constantly $\mathsf{1}$-valued predicates on, respectively, $\mathsf{a}$ and $\mathsf{b}$;
+\item $\mathsf{P : \forall (A\,B : Set) \to (A \to Set) \to (B \to Set) \to Equal\,A\,B \to Set}$ is defined as
+$\mathsf{P\,A\,B\,Q_A\,Q_B\,e = Q\,A\,B\,e}$;
+\item $\mathsf{K^A_\top}$ and $\mathsf{K^B_\top}$ are the constantly $\mathsf{\top}$-valued predicates on, respectively, $\mathsf{A}$ and $\mathsf{B}$;
 %\item $\mathsf{K^b_1}$ is the constantly $\mathsf{1}$-valued predicate on $\mathsf{b}$;
-\item $\mathsf{L_E : Equal^{\wedge}\,a\,b\,K^a_1\,K^b_1\,e}$ is defined by pattern matching,
-i.e., in case $\mathsf{a = b}$ and $\mathsf{e = Refl}$,
-it is defined as $\mathsf{L_E\,x = Refl : Equal\,a\,a}$.
+\item $\mathsf{liftE : Equal^{\wedge}\,A\,B\,K^A_\top\,K^B_\top\,e}$ is defined by pattern matching,
+i.e., in case $\mathsf{A = B}$ and $\mathsf{e = refl}$,
+it is defined as $\mathsf{liftE\,a = refl : Equal\,A\,A}$ for every $\mathsf{a : A}$.
 \end{itemize}
-
 That the structural induction rule is a consequence of the deep induction one is also true for all the examples below,
@@
 
-\subsection{(Deep) induction for $\mathsf{Seq}$}
-
-Next, we shall provide an induction rule for the $\mathsf{Seq}$ type defined in Equation~\ref{eq:eq_seq}.
+\subsection{Deep induction for $\mathsf{Seq}$}
+
+Next, we shall provide an induction rule for the $\mathsf{Seq}$ type defined in Example~\ref{example:seq}.
 Again, the first step in deriving the induction rule for $\mathsf{Seq}$ consists in defining the predicate-lifting function over it, 
 \[
-\mathsf{Seq^{\wedge} : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set}
+\mathsf{Seq^{\wedge} : \forall (A : Set) \to (A \to Set) \to Seq\,A \to Set}
 \]
@@
 \[
-\mathsf{Seq^{\wedge}\,a\,Q_a\,(Const\,x) = Q_a\,x}
-\]
-where $\mathsf{Q_a : a \to Set}$ and $\mathsf{x : a}$,
+\mathsf{Seq^{\wedge}\,A\,Q_A\,(const\,a) = Q_A\,a}
+\]
+where $\mathsf{Q_A : A \to Set}$ and $\mathsf{a : A}$,
 and
-\begin{multline*}
-\mathsf{Seq^{\wedge}\,a\,Q_a\,(SPair\,b\,c\,e\,s_b\,s_c)} \\
-\mathsf{=
-\exists (Q_b : b \to Set) (Q_c : c \to Set)
-\to Equal^{\wedge}\,a\, (b \times c)\, Q_a\, (Q_b \times Q_c) \, e \times Seq^{\wedge}\,b\,Q_b\,s_b \times Seq^{\wedge}\,c\,Q_c\,s_c
+\[
+\begin{array}{l}
+\mathsf{Seq^{\wedge}\,A\,Q_A\,(sPair\,B\,C\,e\,s_B\,s_C)} \\
+\quad\mathsf{=
+\sum [Q_B : B \to Set] [Q_C : C \to Set]\;
+Equal^{\wedge}\,A\, (B \times C)\, Q_A\, (Q_B \times Q_C) \, e \times Seq^{\wedge}\,B\,Q_B\,s_B \times Seq^{\wedge}\,C\,Q_C\,s_C
 }
-\end{multline*}
-where $\mathsf{e : Equal\,a\,(b \times c)}$,
-$\mathsf{s_b : Seq\,b}$ and $\mathsf{s_c : Seq\,c}$.
-We also need to define the lifting of predicates over the polymorphic type of pairs, $\mathsf{Pair = \forall (b\,c : Set) \to b \times c}$, which is 
-\[
-\mathsf{Pair^{\wedge} : \forall (b\,c : Set) \to (b \to Set) \to (c \to Set) \to Pair\,b\,c \to Set}
-\]
-and it is defined as
-\[
-\mathsf{Pair^{\wedge}\,b\,c\,Q_b\,Q_c\,(y, z) = Q_b\,y \times Q_c\,z}
-\]
-where $\mathsf{Q_b : b \to Set}$, $\mathsf{Q_c : c \to Set}$, $\mathsf{y : b}$ and $\mathsf{z : c}$.
-
-Finally, let $\mathsf{CConst}$ be the function
-\begin{multline*}
-\mathsf{\lambda (P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set)} \\
-\mathsf{\to \forall (a : Set) (Q_a : a \to Set) (x : a) \to Q_a\,x \to P\,a\,Q_a\,(Const\,x)}
-\end{multline*}
-associated to the $\mathsf{Const}$ constructor,
-and let $\mathsf{CSPair}$ be the function
-\begin{multline*}
-\mathsf{\lambda (P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set)} \\
-\mathsf{\to \forall (a\,b\,c : Set)
-	(Q_a : a \to Set) (Q_b : b \to Set) (Q_c : c \to Set)} \\
-	\mathsf{(s_b : Seq\,b) (s_c : Seq\,c) (e : Equal\,a\,(b \times c))
-	\to Equal^{\wedge} a\, (b \times c)\, Q_a\, (Pair^{\wedge}\,b\,c\,Q_b\,Q_c)\, e} \\
-	\mathsf{\to P\,b\,Q_b\,s_b \to P\,c\,Q_c\,s_c
-	\to P a Q_a ( SPair\,b\,c\,e\,s_b\,s_c )}
-\end{multline*}
-associated to the $\mathsf{SPair}$ constructor,
-
-With these tools we can formulate an induction rule for $\mathsf{Seq}$,
-\begin{multline*}
-\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set)}
-\mathsf{\to CConst\,P \to CSPair\,P} \\
-\mathsf{\to \forall (a : Set)(Q_a : a \to Set)(s_a : Seq\,a) \to Seq^{\wedge}\,a\,Q_a\,s_a \to P\,a\,Q_a\,s_a}
-\end{multline*}
-To validate the induction rule, we define a term $\mathsf{DISeq}$ for the above type.
-We have to define
-\[
-\mathsf{DISeq\,P\,cconst\,cspair\,a\,Q_a\,s_a\,L_a : P\,a\,Q_a\,s_a}
-\]
-where
-$\mathsf{P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set}$,
-$\mathsf{cconst : CConst\,P}$,
-$\mathsf{cspair : CSPair\,P}$,
-$\mathsf{Q_a : a \to Set}$, $\mathsf{s_a : Seq\,a}$
-and $\mathsf{L_a : Seq^{\wedge}\,a\,Q\,s_a}$,
-and we proceed by pattern-matching on $\mathsf{s_a}$.
-Let $\mathsf{s_a = Const\,x}$ for $\mathsf{x : a}$, and define
-\[
-\mathsf{DISeq\,P\,cconst\,cspair\,a\,Q_a\,(Const\,x)\,L_a = cconst\,a\,Q_a\,x\,L_a}
-\]
-Notice that $\mathsf{Seq^{\wedge}\,a\,Q_a\,(Const\,x) = Q_a\,x}$, and thus $\mathsf{L_a : Q_a\,x}$,
-making the right-hand-side in the above expression type-check.
-Now, let $\mathsf{s_a = SPair\,b\,c\,e\,s_b\,s_c}$
-for $\mathsf{e : Equal\,a\,(b \times c)}$, $\mathsf{s_b : Seq\,b}$ and $\mathsf{s_c : Seq\,c}$,
-and define
-\[
-\mathsf{DISeq\,P\,cconst\,cspair\,a\,Q_a\,(SPair\,b\,c\,e\,s_b\,s_c)\,(Q_b, Q_c, L_e, L_b, L_c)
-	= cspair\,a\,b\,c\,Q_a\,Q_b\,Q_c\,s_b\,s_c\,e\,L_e\,p_b\,p_c}
-\]
-where $\mathsf{(Q_b, Q_c, L_e, L_b, L_c) : Seq^{\wedge}\,a\,Q\,(SPair\,b\,c\,e\,x\,y)}$, i.e.,
-\begin{itemize}
-\item $\mathsf{Q_b : b \to Set}$ and $\mathsf{Q_c : c \to Set}$;
-\item $\mathsf{L_e : Equal^{\wedge}\,a\, (b \times c)\, Q_a\, (Q_b \times Q_c) \, e}$;
-\item $\mathsf{L_b : Seq^{\wedge}\,b\,Q_b\,s_b}$ and $\mathsf{L_c : Seq^{\wedge}\,c\,Q_c\,s_c}$;
-\end{itemize}
-and $\mathsf{p_b}$ and $\mathsf{p_c}$ are defined as follows:
-\begin{align*}
-\mathsf{p_b} &=\mathsf{DISeq\,P\,cconst\,cspair\,b\,Q_b\,s_b\,L_b : P\,b\,Q_b\,s_b} \\
-\mathsf{p_c} &=\mathsf{DISeq\,P\,cconst\,cspair\,c\,Q_c\,s_c\,L_c : P\,c\,Q_c\,s_c}
-\end{align*}
-
-%\subsection{(Deep) induction for $\mathsf{Seq}$ (general style)}
-%
-%We need the lifting of the type $\mathsf{Pair = \forall (b : Set)(c : Set) \to b \times c}$, which is 
+\end{array}
+\]
+where $\mathsf{e : Equal\,A\,(B \times C)}$,
+$\mathsf{s_B : Seq\,B}$ and $\mathsf{s_C : Seq\,C}$.
+%We also need to define the lifting of predicates over the polymorphic type of pairs,
+%$\mathsf{Pair = \forall (B\,C : Set) \to B \times C}$, which is 
 %\[
-%\mathsf{Pair^{\wedge} : \forall (b : Set)(c : Set) \to (b \to Set) \to (c \to Set) \to b \times c \to Set}
+%\mathsf{Pair^{\wedge} : \forall (b\,c : Set) \to (b \to Set) \to (c \to Set) \to Pair\,b\,c \to Set}
 %\]
-%defined as
+%and it is defined as
 %\[
@@
 %\]
-%
-%%Let 
-%%$\mathsf{F, F_1, F_2 : (Set \to Set) \to Set \to Set \to Set}$
-%%be defined as
-%%$\mathsf{F_1\,G\,b\,c = G\,b}$,
-%%$\mathsf{F_2\,G\,b\,c = G\,c}$,
-%%$\mathsf{F\,G\,b\,c = F_1\,G\,b\,c \times F_2\,G\,b\,c}$,
-%%and $\mathsf{K = Pair}$.
-%
-%
-%Let 
-%$\mathsf{F : (Set \to Set) \to Set \to Set \to Set}$
-%be defined as
-%$\mathsf{F\,G\,b\,c = G\,b \times G\,c}$
-%and $\mathsf{K = Pair}$.
-%
-%%The type of $\mathsf{F_i^{\wedge}}$ is
-%%\begin{multline*}
-%%\mathsf{\forall (G : Set \to Set) (b : Set) (c : Set)
-%%\to (\forall (a : Set) \to (a \to Set) \to G\,a \to Set)} \\
-%%\mathsf{\to (b \to Set) \to (c \to Set) \to (F_i\,G\,b\,c) \to Set}
-%%\end{multline*}
-%The type of $\mathsf{F^{\wedge}}$ is
-%\begin{multline*}
-%\mathsf{\forall (G : Set \to Set) (b : Set) (c : Set)
-%\to (\forall (a : Set) \to (a \to Set) \to G\,a \to Set)} \\
-%\mathsf{\to (b \to Set) \to (c \to Set) \to (F_1\,G\,b\,c \times F_2\,G\,b\,c) \to Set}
-%\end{multline*}
-%%and $\mathsf{F^{\wedge}\,G\,b\,c\,P\,Q_b\,Q_c\,(s_b, s_c)
-%%= F_1^{\wedge}\,G\,b\,c\,P\,Q_b\,Q_c\,s_b \times F_2^{\wedge}\,G\,b\,c\,P\,Q_b\,Q_c\,s_c}$
-%
-%Next we shall consider the $\mathsf{Seq}$ type defined in Equation~\ref{eq:eq_seq}.
-%Again, the first step in deriving its induction rule consists in defining the lifting
-%\[
-%\mathsf{Seq^{\wedge} : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set}
-%\]
-%which is given by
-%\[
-%\mathsf{Seq^{\wedge}\,a\,Q\,(Const\,x) = Q\,x}
-%\]
-%and
-%\begin{multline*}
-%\mathsf{Seq^{\wedge}\,a\,Q_a\,(SPair\,b\,c\,e\,(s_b,s_c))
-%=
-%\exists (Q_b : b \to Set) (Q_c : c \to Set)} \\
-%\mathsf{
-%\to Equal^{\wedge}\,a\, (Pair\,b\,c)\, Q_a\, (Pair^{\wedge}\,b\,c\,Q_b\,Q_c) \, e
-%\times F^{\wedge}\,Seq\,b\,c\,Seq^{\wedge}\,Q_b\,Q_c\,(s_b, s_c)
-%}
-%\end{multline*}
-%%\begin{multline*}
-%%\mathsf{Seq^{\wedge}\,a\,Q_a\,(SPair\,b\,c\,e\,(s_b,s_c))
-%%=
-%%\exists (Q_b : b \to Set) (Q_c : c \to Set)} \\
-%%\mathsf{
-%%\to Equal^{\wedge}\,a\, (Pair\,b\,c)\, Q_a\, (Pair^{\wedge}\,b\,c\,Q_b\,Q_c) \, e
-%%\times F_1^{\wedge}\,Seq\,b\,c\,Seq^{\wedge}\,Q_b\,Q_c\,s_b
-%%\times F_2^{\wedge}\,Seq\,b\,c\,Seq^{\wedge}\,Q_b\,Q_c\,s_c
-%%}
-%%\end{multline*}
-%
-%Define
-%\[
-%\mathsf{PredMap : \forall (a : Set) \to (a \to Set) \to (a \to Set) \to Set}
-%\]
-%as $\mathsf{PredMap\,a\,Q_1\,Q_2 = \forall (x : a) \to Q_1\,x \to Q_2\,x}$
-%
-%%  -- Lifting also acts on morphisms between predicates with the same carrier
-%%  {-# TERMINATING #-}
-%%  Bush^-map : ∀ {l} {X : Set} {Q1 Q2 : X → Set l}
-%%              → Pred-map {l} Q1 Q2 → Pred-map (Bush^ Q1) (Bush^ Q2)
-%%  Bush^-map Q12 Bnil q = bigtt
-%%  Bush^-map Q12 (Bcons x b) (y , bby) =
-%%                Q12 x y , Bush^-map (Bush^-map Q12) b bby
-%Define
-%\[
-%\mathsf{SeqLMap : \forall (a : Set) (Q_1\,Q_2 : a \to Set) \to PredMap\,a\,Q_1\,Q_2 \to PredMap\,(G\,a)\,(Seq^{\wedge}\,a\,Q_1)\,(Seq^{\wedge}\,a\,Q_2)}
-%\]
-%as
-%\begin{gather*}
-%\mathsf{
-%SeqLMap\,a\,Q_1\,Q_2\,M\,(Const\,x)\,z = M\,x\,z
-%} \\
-%\mathsf{
-%SeqLMap\,a\,Q_1\,Q_2\,M\,(SPair\,b\,c\,e\,s_b\,s_c)\,(Q_b, Q_c, l_E, l_F) =
-%}
-%\end{gather*}
-%
-%
-%We have an induction rule for $\mathsf{Seq}$
-%\begin{multline}
-%\mathsf{DISeq : \forall (P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set)} \\
-%\mathsf{\to \big( \forall (a : Set) (Q_a : a \to Set) (x : a) \to Q_a\,x \to P\,a\,Q_a\,(Const\,x) \big)} \\
-%\mathsf{\to \big( \forall (a : Set) (b : Set) (c : Set)
-%	(Q_a : a \to Set) (Q_b : b \to Set) (Q_c : c \to Set)} \\
-%	\mathsf{( (s_b, s_c) : F\,Seq\,b\,c) (e : Equal\,a\,(Pair\,b\,c))
-%	\to Equal^{\wedge} a\, (Pair\,b\,c)\, Q_a\, (Pair^{\wedge}\,b\,c\,Q_b\,Q_c)\, e} \\
-%	\mathsf{\to F^{\wedge}\,G\,b\,c\,P\,Q_b\,Q_c\,(s_b,s_c)
-%	\to P a Q_a ( SPair\,b\,c\,e\,(s_b,s_c) ) \big)} \\
-%\mathsf{\to \forall (a : Set)(Q_a : a \to Set)(s_a : Seq\,a) \to Seq^{\wedge}\,a\,Q_a\,s_a \to P\,a\,Q_a\,s_a}
-%\end{multline}
-%We define $\mathsf{DISeq\,P\,PConst\,PSPair\,a\,Q_a\,s_a\,l_a}$,
-%where
-%$\mathsf{P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set}$,
-%$\mathsf{Q_a : a \to Set}$, $\mathsf{s_a : Seq\,a}$
-%and $\mathsf{l_a : Seq^{\wedge}\,a\,Q\,s_a}$,
-%by pattern-matching on $\mathsf{s_a}$.
-%Let $\mathsf{s_a = Const\,x}$ for $\mathsf{x : a}$, and define
-%\[
-%\mathsf{DISeq\,P\,PConst\,PSPair\,a\,Q_a\,(Const\,x)\,l_a = PConst\,a\,Q_a\,x\,l_a}
-%\]
-%Notice that $\mathsf{Seq^{\wedge}\,a\,Q_a\,(Const\,x) = Q_a\,x}$, and thus $\mathsf{l_a : Q_a\,x}$,
-%making the right-hand-side in the above expression type-check.
-%Now, let $\mathsf{s_a = SPair\,b\,c\,e\,(s_b,s_c)}$
-%for $\mathsf{e : Equal\,a\,(Pair\,b\,c)}$ and $\mathsf{(s_b, s_c) : F\,Seq\,b\,c}$,
-%and define
-%\[
-%\mathsf{DISeq\,P\,PConst\,PSPair\,a\,Q_a\,(SPair\,b\,c\,e\,(s_b,s_c))\,(Q_b, Q_c, l_e, l_{F_1}, l_{F_2})
-%	= PSPair\,a\,b\,c\,Q_a\,Q_b\,Q_c\,(s_b,s_c)\,e\,l_e\,p}
-%\]
-%where
-%$\mathsf{p : F^{\wedge}\,G\,b\,c\,P\,Q_b\,Q_c\,(s_b,s_c)}$,
-%$\mathsf{l_e : Equal^{\wedge}\,a\, (Pair\,b\,c)\, Q_a\, (Pair^{\wedge}\,Q_b\,Q_c) \, e}$,
-%$\mathsf{l_{F_1} : F_1^{\wedge}\,Seq\,b\,c\,Seq^{\wedge}\,Q_b\,Q_c\,s_b}$,
-%$\mathsf{l_{F_2} : F_2^{\wedge}\,Seq\,b\,c\,Seq^{\wedge}\,Q_b\,Q_c\,s_c}$
-%(notice that $\mathsf{(Q_b, Q_c, l_e, l_{F_1}, l_{F_2}) : Seq^{\wedge}\,a\,Q\,(SPair\,b\,c\,e\,x\,y)}$,
-%making the right-hand-side in the above expression type-check)
-%and $\mathsf{p}$ is defined as follows:
-%\[
-%\mathsf{p = (???, ???) 
-%: F_1^{\wedge}\,G\,b\,c\,P\,Q_b\,Q_c\,s_b \times F_2^{\wedge}\,G\,b\,c\,P\,Q_b\,Q_c\,s_c} \\
-%\]
-
-%\begin{align*}
-%\mathsf{DISeq\,P\,PConst\,PSPair\,a\,Q\,(Const\,x)\,l}
-%	&=\mathsf{PConst\,a\,Q\,x\,l} \\
-%\mathsf{DISeq\,P\,PConst\,PSPair\,a\,Q_a\,(SPair\,b\,c\,e\,x\,y)\,(Q_b\,Q_c\,e', l_b, l_c)}
-%	&=\mathsf{PSpair\,a\,b\,c\,Q_a\,Q_b\,Q_c\,x\,y\,e\,p_b\,p_c\,e'}
-%\end{align*}
-%where
-%$\mathsf{Q_a : a \to Set}$, $\mathsf{Q_b : b \to Set}$, $\mathsf{Q_c : c \to Set}$,
-%$\mathsf{x : Seq\,b}$, $\mathsf{y : Seq\,c}$,
-%$\mathsf{e : Equal\,a\,(b \times c)}$,
-%$\mathsf{p_b : P\,b\,Q_b\,x}$, $\mathsf{p_c : P\,c\,Q_c\,y}$,
-%$\mathsf{e' : Equal^{\wedge}\,a\, (b \times c)\, Q\, (Q_b \times Q_c) \, e}$,
-%$\mathsf{l_b : Seq^{\wedge}\,b\,Q_b\,x}$,
-%$\mathsf{l_c : Seq^{\wedge}\,c\,Q_c\,y}$
-%and
-%\begin{align*}
-%\mathsf{p_b} &=\mathsf{DISeq\,P\,PConst\,PSPair\,b\,Q_b\,x\,l_b : P\,b\,Q_b\,x} \\
-%\mathsf{p_c} &=\mathsf{DISeq\,P\,PConst\,PSPair\,c\,Q_c\,y\,l_c : P\,c\,Q_c\,y}
-%\end{align*}
-%
-%\begin{align*}
-%\mathsf{Q_a} &: \mathsf{a \to Set} \\
-%\mathsf{Q_b} &: \mathsf{b \to Set} \\
-%\mathsf{Q_c} &: \mathsf{c \to Set} \\
-%\mathsf{x} &: \mathsf{Seq\,b} \\
-%\mathsf{y} &: \mathsf{Seq\,c} \\
-%\mathsf{e} &: \mathsf{Equal\,a\,(b \times c)} \\
-%\mathsf{p_b} &: \mathsf{P\,b\,Q_b\,x} \\
-%\mathsf{p_c} &: \mathsf{P\,c\,Q_c\,y} \\
-%\mathsf{e'} &: \mathsf{Equal^{\wedge}\,a\, (b \times c)\, Q\, (Q_b \times Q_c) \, e} \\
-%\mathsf{l_b} &: \mathsf{Seq^{\wedge}\,b\,Q_b\,x} \\
-%\mathsf{l_c} &: \mathsf{Seq^{\wedge}\,c\,Q_c\,y} \\
-%\mathsf{p_b = DISeq\,P\,PConst\,PSPair\,b\,Q_b\,x\,l_b} &: \mathsf{P\,b\,Q_b\,x} \\
-%\mathsf{p_c = DISeq\,P\,PConst\,PSPair\,c\,Q_c\,y\,l_c} &: \mathsf{P\,c\,Q_c\,y}
-%\end{align*}
+%where $\mathsf{Q_b : b \to Set}$, $\mathsf{Q_c : c \to Set}$, $\mathsf{y : b}$ and $\mathsf{z : c}$.
+
+Finally, let $\mathsf{dIndConst}$ be the function
+\[
+\mathsf{
+\lambda (P : \forall (A : Set) \to (A \to Set) \to Seq\,A \to Set)
+\to \forall (A : Set) (Q_A : A \to Set) (a : A)
+\to Q_A\,x \to P\,A\,Q_A\,(Const\,a)
+}
+\]
+associated to the $\mathsf{const}$ constructor,
+and let $\mathsf{dIndSPair}$ be the function
+\[
+\begin{array}{l}
+\mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to Seq\,A \to Set)} \\
+\quad \mathsf{\to \forall (A\,B\,C : Set) (Q_A : A \to Set) (Q_B : B \to Set) (Q_C : C \to Set)(s_B : Seq\,B) (s_C : Seq\,C) (e : Equal\,A\,(B \times C))} \\
+\quad \mathsf{\to Equal^{\wedge} A\, (B \times C)\, Q_A\, (Pair^{\wedge}\,B\,C\,Q_B\,Q_C)\, e
+	\to P\,B\,Q_B\,s_B \to P\,C\,Q_C\,s_C
+	\to P A Q_A ( SPair\,B\,C\,e\,s_B\,s_C )}
+\end{array}
+\]
+associated to the $\mathsf{sPair}$ constructor.
+
+With these tools we can formulate an induction rule for $\mathsf{Seq}$,
+\[
+\begin{array}{l}
+\mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to Seq\,A \to Set)}
+\mathsf{\to CConst\,P \to CSPair\,P} \\
+\quad \mathsf{\to \forall (A : Set)(Q_A : A \to Set)(s_A : Seq\,A) \to Seq^{\wedge}\,A\,Q_A\,s_A \to P\,A\,Q_A\,s_A}
+\end{array}
+\]
+To validate the induction rule, we define a term $\mathsf{dIndSeq}$ for the above type.
+We have to define
+\[
+\mathsf{dIndSeq\,P\,cconst\,cspair\,A\,Q_A\,s_A\,liftA : P\,A\,Q_A\,s_A}
+\]
+where
+$\mathsf{P : \forall (A : Set) \to (A \to Set) \to Seq\,A \to Set}$,
+$\mathsf{cconst : cConst\,P}$,
+$\mathsf{cspair : cSPair\,P}$,
+$\mathsf{Q_A : A \to Set}$, $\mathsf{s_A : Seq\,A}$
+and $\mathsf{liftA : Seq^{\wedge}\,A\,Q\,s_A}$,
+and we proceed by pattern-matching on $\mathsf{s_A}$.
+Let $\mathsf{s_A = const\,a}$ for $\mathsf{a : A}$, and define
+\[
+\mathsf{dIndSeq\,P\,cconst\,cspair\,A\,Q_A\,(const\,a)\,liftA = cconst\,A\,Q_A\,a\,liftA}
+\]
+Notice that $\mathsf{Seq^{\wedge}\,A\,Q_A\,(const\,a) = Q_A\,a}$, and thus $\mathsf{liftA : Q_A\,a}$,
+making the right-hand-side in the above expression type-check.
+Now, let $\mathsf{s_A = sPair\,B\,C\,e\,s_B\,s_C}$
+for $\mathsf{e : Equal\,A\,(B \times C)}$, $\mathsf{s_B : Seq\,B}$ and $\mathsf{s_C : Seq\,C}$,
+and define
+\[
+\mathsf{dIndSeq\,P\,cconst\,cspair\,A\,Q_A\,(sPair\,B\,C\,e\,s_B\,s_C)\,(Q_B, Q_C, liftE, liftB, liftC)
+	= cspair\,A\,B\,C\,Q_A\,Q_B\,Q_C\,s_B\,s_C\,e\,liftE\,p_B\,p_C}
+\]
+where $\mathsf{(Q_B, Q_C, liftE, liftB, liftC) : Seq^{\wedge}\,A\,Q\,(sPair\,B\,C\,e\,s_B\,s_C)}$, i.e.,
+\begin{itemize}
+\item $\mathsf{Q_B : B \to Set}$ and $\mathsf{Q_C : C \to Set}$;
+\item $\mathsf{liftE : Equal^{\wedge}\,A\, (B \times C)\, Q_A\, (Q_B \times Q_C) \, e}$;
+\item $\mathsf{liftB : Seq^{\wedge}\,B\,Q_B\,s_B}$ and $\mathsf{liftC : Seq^{\wedge}\,C\,Q_C\,s_C}$;
+\end{itemize}
+and $\mathsf{p_B}$ and $\mathsf{p_C}$ are defined as follows:
+\begin{align*}
+\mathsf{p_B} &=\mathsf{DISeq\,P\,cconst\,cspair\,B\,Q_B\,s_B\,liftB : P\,B\,Q_B\,s_B} \\
+\mathsf{p_C} &=\mathsf{DISeq\,P\,cconst\,cspair\,C\,Q_C\,s_C\,liftC : P\,C\,Q_C\,s_C}
+\end{align*}
 
@@
 
+%%% Change title: something about the inherent limitations of the method
+%%% we might want to extend previous result
+%%% but we cannot so we already have the best possible result
+
 In the previous sections,
@@
 
-For variables we simply return the type $\mathsf{T_a}$ given in $\mathsf{Var\, s\, T_a}$, 
+For variables we simply return the type $\mathsf{T_a}$, 
 and the cases for abstraction and application are similar.
@@
 
-\section{Primitive representation for GADTs}
+%\section{Primitive representation for GADTs}
+
+\section{Conclusion/Related work}
+
+Mention Patricia/Neil2008 paper
 
@@
 
-\section{Conclusion}
-
-Mention Patricia/Neil2008 paper
-
 \section{TODO}
@@
 \begin{itemize}
+\item find correct entcsmacro file (current one is for 2018). Maybe ask Ana Sokolova (anas@cs.uni-salzburg.at).
 \item reference (correctly) Haskell Symposium paper
+\item reference inspiration for STLC GADT : https://www.seas.upenn.edu/~cis194/spring15/lectures/11-stlc.html
 \end{itemize}

### 2021-04-09 induction-paper/ENTCS-GADT-induction.tex
Revision Section 4.3

@@
 \newcommand{\Real}{{\mathbb R}}
-\def\lastname{Johann, Ghiorzi and Jeffries}
+\def\lastname{Johann, Ghiorzi, and Jeffries}
 
@@
 \begin{frontmatter}
-  \title{Deep Induction for GADTs}
+  \title{(Deep) Induction for GADTs}
 
@@
 
-\section{Deep induction for GADTs}
+\section{(Deep) induction for GADTs}
 
@@
 
-\subsection{Deep induction for $\mathsf{Equal}$}\label{sec:ind-equal}
+\subsection{(Deep) induction for $\mathsf{Equal}$}\label{sec:ind-equal}
 
@@
 
-\subsection{Deep induction for $\mathsf{Seq}$}
+\subsection{(Deep) induction for $\mathsf{Seq}$}
 
@@
 
-{\color{blue} Some notational conventions: I am using uppercase letters for elements of LType and 
-lowercase letters for elements of LTerm. For example, $\mathsf{(T_b : LType \, b)}$ or $\mathsf{(t_c : LTerm \,c)}$.}
-
-Now we will define an induction rule for the $\mathsf{LTerm}$ type defined in 
-Equation~\ref{eq:eq_lterm}. To define this, we need liftings for both $\mathsf{LType}$ and
+%{\color{blue} Some notational conventions: I am using uppercase letters for elements of LType and 
+%lowercase letters for elements of LTerm. For example, $\mathsf{(T_b : LType \, b)}$ or $\mathsf{(t_c : LTerm \,c)}$.}
+
+We now present an induction rule for the $\mathsf{LTerm}$ type defined in Example~\ref{example:lterm}.
+For that, we need liftings for both $\mathsf{LType}$ and
 $\mathsf{LTerm}$. We will also need a lifting of predicates for arrow types, since arrow types appear in 
@@
 \[
-  \mathsf{Arr^{\wedge} : \forall (a\, b : Set) \to (a \to Set) \to (b \to Set) \to (a \to b) \to Set}
-\]
-is defined as: 
-\[
-  \mathsf{Arr^{\wedge}\, a\, b\, Q_a\, Q_b\, f = \forall (x : a) \to Q_a\, x \to Q_b\, (f\, x)}
-\]
-
-\noindent
+  \mathsf{Arr^{\wedge} : \forall (A\, B : Set) \to (A \to Set) \to (B \to Set) \to (A \to B) \to Set}
+\]
+is defined as
+\[
+  \mathsf{Arr^{\wedge}\, A\, B\, Q_A\, Q_B\, f = \forall (a : A) \to Q_A\,a \to Q_B\, (f\,a)}
+\]
 Now we can define the lifting for $\mathsf{LType}$, which has type 
 \[
-\mathsf{LType^{\wedge} : \forall (a : Set) \to (a \to Set) \to LType\,a \to Set}
+\mathsf{LType^{\wedge} : \forall (A : Set) \to (A \to Set) \to LType\,A \to Set}
 \]
@@
 \[
-  \mathsf{LType^{\wedge}\,a\,Q_a\,(TBool\,b\,e) = \exists (Q_b : b \to Set) 
-  \to Equal^{\wedge}\, a\, b\, Q_a\, (KBool^{\wedge} \,b \,Q_b) \,e}
-\]
-where $\mathsf{Q_a : a \to Set}$ and $\mathsf{e : Equal \,a \,(KBool\,b)}$,
+  \mathsf{LType^{\wedge}\,A\,Q_A\,(tBool\,B\,e)
+  = \sum [Q_B : B \to Set]\, Equal^{\wedge}\, A\, B\, Q_A\, K^{Bool}_{\top} \,e}
+\]
+where $\mathsf{Q_A : A \to Set}$,
+$\mathsf{K^{Bool}_{\top}}$ is the constantly $\mathsf{\top}$-valued predicate on $\mathsf{Bool}$,
+$\mathsf{e : Equal\,A\,Bool}$,
 and
+\[
+\begin{array}{l}
+\mathsf{LType^{\wedge}\,A\,Q_A\,(tArr\, B\, C\, e\, T_B\, T_C)} \\
+\quad
+\mathsf{= \sum [Q_B : B \to Set] [Q_C : C \to Set]\,
+  Equal^{\wedge}\,A\, (B \to C)\, Q_A\, (Arr^{\wedge} \, B\, C\, Q_B \, Q_C) \, e
+  \times \, LType^{\wedge}\,B\,Q_B\,T_B \times LType^{\wedge}\,C\,Q_C\,T_C
+}
+\end{array}
+\]
+where $\mathsf{Q_A : A \to Set}$, $\mathsf{e : Equal\, A\, (B \to C)}$, $\mathsf{T_B : LType \,B}$, and 
+$\mathsf{T_C : LType \,C}$,
+and 
+\[
+\mathsf{LType^{\wedge}\,A\,Q_A\,(tList\, B\, e\, T_B)
+=
+\sum [Q_B : B \to Set]\, 
+  Equal^{\wedge}\,A\, (List\, B)\, Q_A\, (List^{\wedge} \, B\, Q_B) \, e 
+  \times LType^{\wedge}\,B\,Q_B\,T_B
+}
+\]
+where $\mathsf{Q_A : A \to Set}$, $\mathsf{e : Equal\, A\, (List\, B)}$, and $\mathsf{T_B : LType \,B}$.
+
+Given the lifting for $\mathsf{LType}$, we can define the lifting for $\mathsf{LTerm}$. Again, 
+the lifting 
+\[
+\mathsf{LTerm^{\wedge} : \forall (A : Set) \to (A \to Set) \to LTerm\,A \to Set}
+\]
+is defined by pattern-matching as
+\[
+  \mathsf{LTerm^{\wedge}\,A\,Q_A\,(var\,s\,T_A) = LType^{\wedge}\, A\, Q_A\, T_A}
+\]
+where $\mathsf{s : String}$ and $\mathsf{T_A : LType\, A}$, and
+\[
+\begin{array}{l}
+\mathsf{LTerm^{\wedge}\,A\,Q_A\, (abs \,B \,C \,e \,s \,T_B \,t_C)} \\ 
+\quad \mathsf{
+  = \sum [Q_B : B \to Set] [Q_C : C \to Set]\,
+      Equal^{\wedge} \, A\, (B \to C)\, Q_A\, (Arr^{\wedge} \, B\, C\, Q_B \, Q_C)\, e
+      \times \, LType^{\wedge}\, B\, Q_B\, T_B
+      \times \, LTerm^{\wedge}\, C\, Q_C\, t_C }
+\end{array}
+\]
+where $\mathsf{e : Equal \, A \, (B \to C)}$,
+$\mathsf{s : String}$,
+$\mathsf{T_B : LType\,B}$,
+and $\mathsf{t_C : LTerm \, C}$,
+and
+\[
+\mathsf{LTerm^{\wedge}\,A\,Q_A\, (app\, B\, t_{BA}\, t_B)
+=
+\sum [Q_B : B \to Set]\, 
+  LTerm^{\wedge}\, (B \to A)\, (Arr^{\wedge} \, B\, A\, Q_B \, Q_A)\, t_{BA}
+  \times LTerm^{\wedge}\, B\, Q_B\, t_B}
+\]
+where $\mathsf{t_{BA} : LTerm \, (B \to A)}$ and $\mathsf{t_B : LTerm \, B}$, and
+\[
+\mathsf{LTerm^{\wedge}\,A\,Q_A\, (listC\, B\, e\, ts)
+=
+\sum [Q_B : B \to Set]\,
+ Equal^{\wedge} \, A\, (List\,B)\, Q_A\, (List^{\wedge} \, B\, Q_B) \, e
+ \times List^{\wedge}\, (LTerm\,B) \, (LTerm^{\wedge} \, B\, Q_B) \, ts}
+\]
+where $\mathsf{e : Equal\, A\, (List \,B)}$ and $\mathsf{ts : List\, (LTerm B)}$. 
+Notice we use the lifting for $\mathsf{List}$ in the case for the $\mathsf{listC}$ constructor.
+
+With these liftings, we can define the deep induction principle for $\mathsf{LTerm}$.
+% Var case 
+Let $\mathsf{dIndVar}$ be the function 
 \begin{multline*}
-  \mathsf{LType^{\wedge}\,a\,Q_a\,(TArr\, b\, c\, e\, T_b\, T_c)} \\
-\mathsf{=
-\exists (Q_b : b \to Set) (Q_c : c \to Set)
-  \to Equal^{\wedge}\,a\, (b \to c)\, Q_a\, (Arr^{\wedge} \, b\, c\, Q_b \, Q_c) \, e} \\ 
-\mathsf{
-  \times \, LType^{\wedge}\,b\,Q_b\,T_b \times LType^{\wedge}\,c\,Q_c\,T_c
+\mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A \to Set)} \\
+\mathsf{\to 
+  \forall (A : Set) (Q_A : A \to Set) (s : String) (T_A : LType\, A) 
+  \to LType^{\wedge} \, A\, Q_A\, T_A
+  \to P \, A\, Q_A\, (var \, s\, T_A)
   }
 \end{multline*}
-
-where $\mathsf{Q_a : a \to Set}$, $\mathsf{e : Equal\, a\, (b \to c)}$, $\mathsf{T_b : LType \,b}$, and 
-$\mathsf{T_c : LType \,c}$,
-and 
-\begin{multline*}
-  \mathsf{LType^{\wedge}\,a\,Q_a\,(TList\, b\, e\, T_b)} \\
-\mathsf{=
-\exists (Q_b : b \to Set) 
-  \to Equal^{\wedge}\,a\, (List\, b)\, Q_a\, (List^{\wedge} \, b\, Q_b) \, e 
-  \times LType^{\wedge}\,b\,Q_b\,T_b 
-}
-\end{multline*}
-where $\mathsf{Q_a : a \to Set}$, $\mathsf{e : Equal\, a\, (List\, b)}$, and $\mathsf{T_b : LType \,b}$.
-
-Given the lifting for $\mathsf{LType}$, we can define the lifting for $\mathsf{LTerm}$. Again, 
-the lifting 
-\[
-\mathsf{LTerm^{\wedge} : \forall (a : Set) \to (a \to Set) \to LTerm\,a \to Set}
-\]
-is defined by pattern-matching as
-\[
-  \mathsf{LTerm^{\wedge}\,a\,Q_a\,(Var\,s\,T_a) = LType^{\wedge}\, a\, Q_a\, T_a}
-\]
-where $\mathsf{s : String}$ and $\mathsf{T_a : LType\, a}$,  and 
-\begin{multline*}
-  \mathsf{LTerm^{\wedge}\,a\,Q_a\,
-  (Abs \,b \,c \,e \,s \,T_b \,t_c)} \\ 
-\mathsf{
-  = \exists (Q_b : b \to Set) (Q_c : c \to Set) \to
-      Equal^{\wedge} \, a\, (b \to c)\, Q_a\, (Arr^{\wedge} \, b\, c\, Q_b \, Q_c)\, e} \\ 
-\mathsf{
-      \times \, LType^{\wedge}\, b\, Q_b\, T_b
-      \times \, LTerm^{\wedge}\, c\, Q_c\, t_c }
-\end{multline*}
-where $\mathsf{e : Equal \, a \, (b \to c)}$, $\mathsf{s : String}$ , $\mathsf{T_b : LType\,b}$, and $\mathsf{t_c : LTerm \, c}$,
-and 
-\begin{multline*}
-  \mathsf{LTerm^{\wedge}\,a\,Q_a\, (App\, b\, t_{ba}\, t_b)} \\ 
-\mathsf{
-  = \exists (Q_b : b \to Set) \to 
-      LTerm^{\wedge}\, (b \to a)\, (Arr^{\wedge} \, b\, a\, Q_b \, Q_a)\, t_{ba}
-      \times LTerm^{\wedge}\, b\, Q_b\, t_b }
-\end{multline*}
-where $\mathsf{t_{ba} : LTerm \, (B \to A)}$ and $\mathsf{t_b : LTerm \, B}$, and
-\begin{multline*}
-  \mathsf{LTerm^{\wedge}\,a\,Q_a\, (ListC\, b\, e\, ts)} \\ 
-\mathsf{
-  = \exists (Q_b : b \to Set) \to 
-      Equal^{\wedge} \, a\, (List\,b)\, Q_a\, (List^{\wedge} \, b\, Q_b) \, e
-      \times List^{\wedge}\, (LTerm\,b) \, (LTerm^{\wedge} \, b\, Q_b) \, ts
-    }
-\end{multline*}
-where $\mathsf{e : Equal\, a\, (List \,b)}$ and $\mathsf{ts : List\, (LTerm b)}$. 
-Notice we use the lifting for $\mathsf{List}$ in the case for the $\mathsf{ListC}$ constructor.
-
-With these liftings, we can define the deep induction princple for $\mathsf{LTerm}$.
-% Var case 
-Let CVar be the type 
-\begin{multline*}
-\mathsf{\lambda (P : \forall (a : Set) \to (a \to Set) \to LTerm\,a \to Set)} \\
-\mathsf{\to 
-  \forall (a : Set) (Q_a : a \to Set) (s : String) (T_a : LType\, a) 
-  \to LType^{\wedge} \, a\, Q_a\, T_a
-  \to P \, a\, Q_a\, (Var \, s\, T_a)
-  }
-\end{multline*}
-associated to the $\mathsf{Var}$ constructor. 
+associated to the $\mathsf{var}$ constructor. 
 % Abs case 
-Let CAbs be the type 
+Let $\mathsf{dIndAbs}$ be the function 
 \begin{align*}
-  &\mathsf{\lambda (P : \forall (a : Set) \to (a \to Set) \to LTerm\,a \to Set)} \\
-  &\mathsf{\to 
-  \forall (a \,b\, c: Set) (Q_a : a \to Set)  (Q_b : b \to Set) (Q_c : c \to Set)
-  (e : Equal\, a\, (b \to c)) (s : String) } \\ 
-  &\mathsf{ \to (T_b : LType\, b) \to (t_c : LTerm\, c)
-  \to Equal^{\wedge}\,a\,(b \to c)\, Q_a \, (Arr^{\wedge} \, b\, c\, Q_b \, Q_c) \, e
+  &\mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A \to Set)} \\
+  &\quad\mathsf{\to 
+  \forall (A\,B\,C: Set) (Q_A : A \to Set)  (Q_B : B \to Set) (Q_C : C \to Set)
+  (e : Equal\, A\, (B \to C)) (s : String) } \\ 
+  &\quad\mathsf{ \to (T_B : LType\, B) \to (t_C : LTerm\, C)
+  \to Equal^{\wedge}\,A\,(B \to C)\, Q_A \, (Arr^{\wedge} \, B\, C\, Q_B \, Q_C) \, e
   } \\
-  &\mathsf{
-  \to LType^{\wedge}\, b\, Q_b\, T_b
-  \to P\, c\, Q_c\, t_c\, 
-  \to P \, a\, Q_a\, (Abs \,b \,c \, e \,s \,T_b \, t_c)
+  &\quad\mathsf{
+  \to LType^{\wedge}\, B\, Q_B\, T_B
+  \to P\, C\, Q_C\, t_C\, 
+  \to P \, A\, Q_A\, (abs \,B \,C \, e \,s \,T_B \, t_C)
   }
 \end{align*}
-associated to the $\mathsf{Abs}$ constructor. 
+associated to the $\mathsf{abs}$ constructor. 
 % App case 
-Let CApp be the type 
+Let $\mathsf{dIndApp}$ be the function 
 \begin{align*}
-  &\mathsf{\lambda (P : \forall (a : Set) \to (a \to Set) \to LTerm\,a \to Set)} \\
-  &\mathsf{\to 
-  \forall (a \,b : Set) (Q_a : a \to Set)  (Q_b : b \to Set) 
-   (t_{ba} : LTerm\, (b \to a)) (t_b : LTerm\, b)} \\
-  &\mathsf{
-  \to P\, (b \to a)\, (Arr^{\wedge} \, b\, a\, Q_b \, Q_a) \, t_{ba} \, 
-  \to P\, b\, Q_b\, t_b\, 
-  \to P \, a\, Q_a\, (App \,b \,t_{ba} \, t_b) }
+  &\mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A \to Set)} \\
+  &\quad \mathsf{\to 
+  \forall (A \,B : Set) (Q_A : A \to Set)  (Q_B : B \to Set) 
+   (t_{BA} : LTerm\, (B \to A)) (t_B : LTerm\, B)} \\
+  &\quad \mathsf{
+  \to P\, (B \to A)\, (Arr^{\wedge} \, B\, A\, Q_B \, Q_A) \, t_{BA} \, 
+  \to P\, B\, Q_B\, t_B\, 
+  \to P \, A\, Q_A\, (app \,B \,t_{BA} \, t_B) }
 \end{align*}
-associated to the $\mathsf{App}$ constructor. 
+associated to the $\mathsf{app}$ constructor. 
 % ListC case 
-Let CListC be the type 
+Let $\mathsf{dIndListC}$ be the type 
 \begin{align*}
-  &\mathsf{\lambda (P : \forall (a : Set) \to (a \to Set) \to LTerm\,a \to Set)} \\
-  &\mathsf{\to 
-  \forall (a \,b : Set) (Q_a : a \to Set)  (Q_b : b \to Set) 
-    (e : Equal\, a\, (List\, b)) (ts : List\, (LTerm\, b))} \\ 
-  &\mathsf{
-    \to Equal^{\wedge}\, a\, (List\,b)\, Q_a\, (List^{\wedge}\, b\, Q_b)\, e 
-  \to List^{\wedge}\, (LTerm\,b) (P\, b\, Q_b)\, ts} \\
-&\mathsf{
-  \to P \, a\, Q_a\, (ListC \,b \,e \, ts) }
+  &\mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A \to Set)} \\
+  &\quad \mathsf{\to 
+  \forall (A \,B : Set) (Q_A : A \to Set)  (Q_B : B \to Set) 
+    (e : Equal\, A\, (List\, B)) (ts : List\, (LTerm\, B))} \\ 
+  &\quad \mathsf{
+    \to Equal^{\wedge}\, A\, (List\,B)\, Q_A\, (List^{\wedge}\, B\, Q_B)\, e 
+  \to List^{\wedge}\, (LTerm\,B) (P\, B\, Q_B)\, ts
+  \to P \, A\, Q_A\, (listC \,B \,e \, ts) }
 \end{align*}
-associated to the $\mathsf{ListC}$ constructor. 
-Then the type of the deep induction principle for $\mathsf{LTerm}$ is,
-\begin{multline*}
-\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to LTerm\,a \to Set)}
-\mathsf{\to CVar\,P \to CAbs\,P \to CApp\,P \to CListC\,P} \\
-\mathsf{\to \forall (a : Set)(Q_a : a \to Set)(t_a : LTerm\,a) \to LTerm^{\wedge}\,a\,Q_a\,t_a \to P\,a\,Q_a\,t_a}
-\end{multline*}
-
-To prove the induction principle, we define a term $\mathsf{DILTerm}$ for it. We have to define 
-\[ 
-\mathsf{DILTerm \, P\, cvar \, cabs\, capp\, clistc \, a\, Q_a\, t_a\, L_a : P\, a\, Q_a\, t_a}
-\]
+associated to the $\mathsf{listC}$ constructor.
+
+The type of the deep induction principle for $\mathsf{LTerm}$ is
+\[
+\begin{array}{l}
+\mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A \to Set)
+  \to dIndVar\,P \to dIndAbs\,P \to dIndApp\,P \to dIndListC\,P} \\
+\quad \mathsf{\to \forall (A : Set)(Q_A : A \to Set)(t_A : LTerm\,A) \to LTerm^{\wedge}\,A\,Q_A\,t_A \to P\,A\,Q_A\,t_A}
+\end{array}
+\]
+To prove the induction principle, we define a term $\mathsf{dIndLTerm}$ for it.
+We have to define 
+\[ 
+\mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clistc \, A\, Q_A\, t_A\, liftA : P\, A\, Q_A\, t_A}
+\]
+where $\mathsf{cvar: dIndVar}$, $\mathsf{cabs: dIndAbs}$, $\mathsf{capp : dIndApp}$, $\mathsf{clistc : dIndListC}$,
+$\mathsf{Q_A : A \to Set}$, $\mathsf{t_A : LTerm\,A}$, and $\mathsf{liftA : LTerm^{\wedge}\,A\,Q_A\,t_A}$.
 % where $\mathsf{P : \forall (a : Set) \to (a \to Set) \to LTerm\,a \to Set}$, 
-
-\noindent
-As before, we prove the induction principle by pattern matching on $\mathsf{t_a}$. 
-For the $\mathsf{Var}$ case, let $\mathsf{t_a = (Var\, s\, T_a)}$ and define
-\[
-  \mathsf{DILTerm \, P\, cvar \, cabs\, capp\, clistc \, a\, Q_a\, (Var\,s\,T_a) \, L_a = cvar \, a\, Q_a\, s\, T_a\, L_a }
-\]
+As before, we prove the induction principle by pattern matching on $\mathsf{t_A}$. 
+For the $\mathsf{var}$ constructor case, let $\mathsf{t_A = var\, s\, T_A}$ and define
+\[
+  \mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clistc \, A\, Q_A\, (var\,s\,T_A) \, liftA = cvar \, A\, Q_A\, s\, T_A\, liftA }
+\]
+for $\mathsf{s : String}$ and $\mathsf{T_A : LType\,A}$.
 % where $\mathsf{L_a : LTerm^{\wedge}\, a\, Q_a\, (Var\,s\,T_a)}$, i.e., $\mathsf{L_a : LType^{\wedge}\,a\,Q_a\,T_a}$ 
-Notice that $\mathsf{LTerm^{\wedge}\, a\, Q_a\, (Var\,s\,T_a) = LType^{\wedge}\,a\,Q_a\,T_a}$, so 
-$\mathsf{L_a : LType^{\wedge}\,a\,Q_a\,T_a}$.
+Notice that $\mathsf{LTerm^{\wedge}\, A\, Q_A\, (var\,s\,T_A) = LType^{\wedge}\,A\,Q_A\,T_A}$, so 
+$\mathsf{liftA : LType^{\wedge}\,A\,Q_A\,T_A}$.
 %
 % Abs case 
-For the $\mathsf{Abs}$ case, let $\mathsf{t_a = (Abs \,b \,c \, e \,s \,T_b \, t_c)} $ and define
-\begin{multline*}
-  \mathsf{DILTerm \, P\, cvar \, cabs\, capp\, clistc \, a\, Q_a\, (Abs \,b \,c \,e \,s \,T_b \, t_c) \, (Q_b , Q_c , L_e, L_{T_b}, L_{t_c})} \\
-  \mathsf{= cabs\,a\,b\,c\, Q_a\, Q_b\, Q_c\, e\, s\, T_b\, t_c\, L_e\, L_{T_b}\, p_c}
-\end{multline*}
-
-where $\mathsf{(Q_b , Q_c , L_e, L_{T_b}, L_{t_c}) : LTerm^{\wedge}\, a\, Q_a\, (Abs \,b \,c \, e \,s \,T_b \, t_c)}$
+For the $\mathsf{abs}$ case, let $\mathsf{t_A = abs \,B \,C \, e \,s \,T_B \, t_C)} $ and define
+\[
+\begin{array}{l}
+\mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clistc \, A\, Q_A\, (abs \,B \,C \,e \,s \,T_B \, t_C) \, (Q_B , Q_C , liftE, lift_{T_B}, lift_{t_C})} \\
+\quad \mathsf{= cabs\,A\,B\,C\, Q_A\, Q_B\, Q_C\, e\, s\, T_B\, t_C\, liftE\, lift_{T_B}\, p_C}
+\end{array}
+\]
+where $\mathsf{(Q_B , Q_C , liftE, lift_{T_B}, lift_{t_C}) : LTerm^{\wedge}\, A\, Q_A\, (abs \,B \,C \, e \,s \,T_B \, t_C)}$, i.e.,
 \begin{itemize}
-  \item $\mathsf{Q_b : b \to Set}$
-  \item $\mathsf{Q_c : c \to Set}$
-  \item $\mathsf{L_e : Equal^{\wedge} \, a\, (b \to c)\, Q_a\, (Arr^{\wedge} \, b\, c\, Q_b \, Q_c) \, e}$
-  \item $\mathsf{L_{T_b}: LType^{\wedge} \, b\, Q_b\, T_b}$
-  \item $\mathsf{L_{t_c}: LTerm^{\wedge} \, c\, Q_c\, T_c}$
+  \item $\mathsf{Q_B : B \to Set}$;
+  \item $\mathsf{Q_C : C \to Set}$;
+  \item $\mathsf{liftE : Equal^{\wedge} \, A\, (B \to C)\, Q_A\, (Arr^{\wedge} \, B\, C\, Q_B \, Q_C) \, e}$;
+  \item $\mathsf{lift_{T_B}: LType^{\wedge} \, B\, Q_B\, T_B}$;
+  \item $\mathsf{lift_{t_C}: LTerm^{\wedge} \, C\, Q_C\, T_C}$;
 \end{itemize}
-
-and $\mathsf{pc}$ is defined as:
-\[
-  \mathsf{pc =  DILTerm\,P\,cvar\,cabs \,capp \,clistc\, c\, Q_c\, t_c\, L_{t_c}  : P \, c\, Q_c \, t_c }
-\]
-
-\noindent
+and $\mathsf{p_C}$ is defined as:
+\[
+  \mathsf{p_C =  dIndLTerm\,P\,cvar\,cabs \,capp \,clistc\, C\, Q_C\, t_C\, lift_{t_C}  : P \, C\, Q_C \, t_C }
+\]
 % App case - (App \,b \,t_{ba} \, t_b)
-For the $\mathsf{App}$ case, let $\mathsf{t_a = (App \,b \,\,t_{ba} \, t_b)} $ and define
-\begin{multline*}
-  \mathsf{DILTerm \, P\, cvar \, cabs\, capp\, clistc \, a\, Q_a\,  (App \,b \,\,t_{ba} \, t_b)\, (Q_b , L_{t_{ba}}, L_{t_b})} \\
-  \mathsf{= capp\,a\,b\,Q_a\, Q_b\, t_{ba}\, t_b\, p_{ba} \, p_b}
-\end{multline*}
-
-\noindent
-where $\mathsf{(Q_b , L_{t_{ba}}, L_{t_b}) : LTerm^{\wedge}\, a\, Q_a\, (App \,b \,t_{ba} \, t_b)}$
+For the $\mathsf{app}$ case, let $\mathsf{t_A = app \,B \,\,t_{BA} \, t_B} $ and define
+\[
+  \mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clistc \, A\, Q_A\,  (app \,B \,\,t_{BA} \, t_B)\, (Q_B , list_{t_{BA}}, list_{t_B})} \\
+  \mathsf{= capp\,A\,B\,Q_A\, Q_B\, t_{BA}\, t_B\, p_{BA} \, p_B}
+\]
+where $\mathsf{(Q_B , lift_{t_{BA}}, list_{t_B}) : LTerm^{\wedge}\, A\, Q_A\, (app \,B \,t_{BA} \, t_B)}$
 \begin{itemize}
-  \item $\mathsf{Q_b : b \to Set}$
-  \item $\mathsf{L_{t_{ba}}: LTerm^{\wedge} \, (b \to a)\, (Arr^{\wedge} \, b\, a\, Q_b \, Q_a)\, t_{ba}}$
-  \item $\mathsf{L_{t_b}: LTerm^{\wedge} \, b\, Q_b\, t_b}$
+  \item $\mathsf{Q_B : B \to Set}$;
+  \item $\mathsf{lift_{t_{BA}}: LTerm^{\wedge} \, (B \to A)\, (Arr^{\wedge} \, B\, A\, Q_B \, Q_A)\, t_{BA}}$;
+  \item $\mathsf{lift_{t_B}: LTerm^{\wedge} \, B\, Q_B\, t_B}$;
 \end{itemize}
-and $\mathsf{p_{ba}}$ and $\mathsf{p_b}$
+and $\mathsf{p_{BA}}$ and $\mathsf{p_B}$
 and are defined as:
 \begin{align*}
-  \mathsf{p_{ba}} &=  \mathsf{DILTerm\,P\,cvar\,cabs \,capp \,clistc\, (b \to a)\,(Arr^{\wedge} \, b\, a\, Q_b \, Q_a) \, t_{ba}\, L_{t_{ba}}   : P \, (b \to a)\, (Arr^{\wedge} \, b\, a\, Q_b \, Q_a) \, t_{ba} } \\
-  \mathsf{p_b   } &=  \mathsf{DILTerm\,P\,cvar\,cabs \,capp \,clistc\, b\, Q_b\, t_b\, L_{t_b}  : P \, b\, Q_b \, t_b } 
+  \mathsf{p_{BA}} &= \mathsf{dIndLTerm\,P\,cvar\,cabs \,capp \,clistc\, (B \to A)\,(Arr^{\wedge} \, B\, A\, Q_B \, Q_A) \, t_{BA}\, lift_{t_{BA}}
+  : P \, (B \to A)\, (Arr^{\wedge} \, B\, A\, Q_B \, Q_A) \, t_{BA} } \\
+  \mathsf{p_B} &= \mathsf{dIndLTerm\,P\,cvar\,cabs \,capp \,clistc\, B\, Q_B\, t_B\, lift_{t_B}  : P \, B\, Q_B \, t_B } 
 \end{align*}
-
-\noindent
 % ListC case - (ListC \,b \,e \, ts)
-For the $\mathsf{ListC}$ case, let $\mathsf{t_a = (ListC \,b \,e \, ts)} $ and define
+Finally, for the $\mathsf{listC}$ case, let $\mathsf{t_A = listC \,B \,e \, ts}$ and define
 \begin{multline*}
-  \mathsf{DILTerm \, P\, cvar \, cabs\, capp\, clistc \, a\, Q_a\, (ListC \,b \,e \, ts) \, (Q_b , L_e, L_{List})
-  = clistc \,a\,b\,Q_a\, Q_b\, e\, ts\, L_e\, p_{List} }
+  \mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clistc \, A\, Q_A\, (listC \,B \,e \, ts) \, (Q_B , liftE, lift_{List})
+  = clistc \,A\,B\,Q_A\, Q_B\, e\, ts\, liftE\, p_{List} }
 \end{multline*}
-
-\noindent
-where $\mathsf{(Q_b , L_e, L_{List}): LTerm^{\wedge}\, a\, Q_a\, (ListC \,b \,e \, ts)}$
+where $\mathsf{(Q_B , liftE, lift_{List}): LTerm^{\wedge}\, A\, Q_A\, (listC \,B \,e \, ts)}$, i.e.,
 \begin{itemize}
-  \item $\mathsf{Q_b : b \to Set}$
-  \item $\mathsf{L_e : Equal^{\wedge}\, a\, (List\,b)\, Q_a\, (List^{\wedge}\, b\, Q_b)\, e}$
-  \item $\mathsf{L_{List}: List^{\wedge} \, (LTerm\, b) \, (LTerm^{\wedge}\, b\, Q_b) \, ts}$
+  \item $\mathsf{Q_B : B \to Set}$;
+  \item $\mathsf{listE : Equal^{\wedge}\, A\, (List\,B)\, Q_A\, (List^{\wedge}\, B\, Q_B)\, e}$;
+  \item $\mathsf{lift_{List}: List^{\wedge} \, (LTerm\, B) \, (LTerm^{\wedge}\, B\, Q_B) \, ts}$;
 \end{itemize}
-
-\noindent
 and $\mathsf{p_{List}}$ is defined as:
-\begin{multline*}
-  \mathsf{p_{List} = List^{\wedge}map \, (LTerm\, b) \, (LTerm^{\wedge} \, b \, Q_b)\, 
-    (P\,b\,Q_b)\, \,p_{ts} \,\, ts\,\, L_{List}  : List^{\wedge}\, (LTerm\,b) \, (P\,b\,Q_b) \, ts}
-\end{multline*}
+\[
+  \mathsf{p_{List} = liftListMap \, (LTerm\, B) \, (LTerm^{\wedge} \, B \, Q_B)\, 
+    (P\,B\,Q_B)\, p_{ts} \, ts\, lift_{List}  : List^{\wedge}\, (LTerm\,B) \, (P\,B\,Q_B) \, ts}
+\]
 and $\mathsf{p_{ts}}$ is defined as:
-\begin{multline*}
+\[
 \mathsf{
-  p_{ts} = DILTerm\, P\, cvar\, cabs\, capp\, clistc\, b\, Q_b :  PredMap \,(LTerm^{\wedge}\, b\, Q_b) \, (P\,b\,Q_b)
+  p_{ts} = dIndLTerm\, P\, cvar\, cabs\, capp\, clistc\, B\, Q_B :  PredMap \,(LTerm^{\wedge}\, B\, Q_B) \, (P\,B\,Q_B)
   }
-\end{multline*}
+\]
 where 
-\[\mathsf{List^{\wedge}map : \forall\, (a : Set) \to (Q_a \, Q'_a : a \to Set) 
-  \to PredMap\,Q_a\,Q'_a \to PredMap \,(List^{\wedge}\, a\, Q_a)\, (List^{\wedge}\, a\, Q'_a)}\]
+\[
+\mathsf{liftListMap : \forall\, (A : Set) \to (Q_A \, Q'_A : A \to Set) 
+  \to PredMap\,Q_A\,Q'_A \to PredMap \,(List^{\wedge}\, A\, Q_A)\, (List^{\wedge}\, A\, Q'_A)}
+\]
 takes a morphism of predicates and produces a morphism of lifted predicates. 
@@
 \begin{align*}
-  &\mathsf{PredMap : \forall\, \{a : Set\} \to (Q_a \, Q'_a : a \to Set) \to Set } \\
-  &\mathsf{PredMap \, Q_a\,Q'_a\, = \forall\, (x : A) \to Q_a\,x \to Q'_a\,x}
+  &\mathsf{PredMap : \forall\, \{A : Set\} \to (A \to Set) \to (A \to Set) \to Set } \\
+  &\mathsf{PredMap \, Q_A\,Q'_A\, = \forall\, (a : A) \to Q_A\,a \to Q'_A\,a}
 \end{align*}
@@
 \subsection{Defining $\mathsf{LTerm^{\wedge}K1}$}
+
+{\color{red} Maybe this section can be deleted by assuming $\mathsf{LTerm^{\wedge}\, a\, K1\,t = K1}$.
+Maybe we can say this derives from parametricity.
+Currently we say that $\mathsf{K1 \times K1 = K1}$ based on the fact that products are a built-in type and so this seems to be obviously true.}
+
 The last piece of infrastructure we need to define $\mathsf{getTypeProof}$ is a function 

### 2021-04-12 induction-paper/ENTCS-GADT-induction.tex
Revise Section 4.4

@@
 where $\mathsf{Q_A : A \to Set}$,
-$\mathsf{K^{Bool}_{\top}}$ is the constantly $\mathsf{\top}$-valued predicate on $\mathsf{Bool}$,
-$\mathsf{e : Equal\,A\,Bool}$,
-and
+$\mathsf{K^{Bool}_{\top}}$ is the constantly $\mathsf{\top}$-valued predicate on $\mathsf{Bool}$
+and $\mathsf{e : Equal\,A\,Bool}$;
 \[
@@
 \]
-where $\mathsf{Q_A : A \to Set}$, $\mathsf{e : Equal\, A\, (B \to C)}$, $\mathsf{T_B : LType \,B}$, and 
-$\mathsf{T_C : LType \,C}$,
+where $\mathsf{Q_A : A \to Set}$, $\mathsf{e : Equal\, A\, (B \to C)}$, $\mathsf{T_B : LType \,B}$ and 
+$\mathsf{T_C : LType \,C}$;
 and 
@@
 where $\mathsf{Q_A : A \to Set}$, $\mathsf{e : Equal\, A\, (List\, B)}$, and $\mathsf{T_B : LType \,B}$.
-
-Given the lifting for $\mathsf{LType}$, we can define the lifting for $\mathsf{LTerm}$. Again, 
+%Given the lifting for $\mathsf{LType}$, 
+We can define the lifting for $\mathsf{LTerm}$. Again, 
 the lifting 
@@
 \]
-where $\mathsf{s : String}$ and $\mathsf{T_A : LType\, A}$, and
+where $\mathsf{s : String}$ and $\mathsf{T_A : LType\, A}$;
 \[
@@
 $\mathsf{T_B : LType\,B}$,
-and $\mathsf{t_C : LTerm \, C}$,
-and
+and $\mathsf{t_C : LTerm \, C}$;
 \[
@@
 \]
-where $\mathsf{t_{BA} : LTerm \, (B \to A)}$ and $\mathsf{t_B : LTerm \, B}$, and
+where $\mathsf{t_{BA} : LTerm \, (B \to A)}$ and $\mathsf{t_B : LTerm \, B}$; and
 \[
@@
 where $\mathsf{e : Equal\, A\, (List \,B)}$ and $\mathsf{ts : List\, (LTerm B)}$. 
-Notice we use the lifting for $\mathsf{List}$ in the case for the $\mathsf{listC}$ constructor.
+Notice that we use the lifting for $\mathsf{List}$ in the case for the $\mathsf{listC}$ constructor.
 
@@
 % ListC case 
-Let $\mathsf{dIndListC}$ be the type 
+Let $\mathsf{dIndListC}$ be the function 
 \begin{align*}
@@
 % Abs case 
-For the $\mathsf{abs}$ case, let $\mathsf{t_A = abs \,B \,C \, e \,s \,T_B \, t_C)} $ and define
+For the $\mathsf{abs}$ case, let $\mathsf{t_A = abs \,B \,C \, e \,s \,T_B \, t_C} $ and define
 \[
@@
 \]
-where $\mathsf{(Q_B , lift_{t_{BA}}, list_{t_B}) : LTerm^{\wedge}\, A\, Q_A\, (app \,B \,t_{BA} \, t_B)}$
+where $\mathsf{(Q_B , lift_{t_{BA}}, list_{t_B}) : LTerm^{\wedge}\, A\, Q_A\, (app \,B \,t_{BA} \, t_B)}$, i.e.,
 \begin{itemize}
@@
 \mathsf{
-  p_{ts} = dIndLTerm\, P\, cvar\, cabs\, capp\, clistc\, B\, Q_B :  PredMap \,(LTerm^{\wedge}\, B\, Q_B) \, (P\,B\,Q_B)
+  p_{ts} = dIndLTerm\, P\, cvar\, cabs\, capp\, clistc\, B\, Q_B :  PredMap\,(LTerm\,B) \,(LTerm^{\wedge}\, B\, Q_B) \, (P\,B\,Q_B)
   }
@@
 \mathsf{liftListMap : \forall\, (A : Set) \to (Q_A \, Q'_A : A \to Set) 
-  \to PredMap\,Q_A\,Q'_A \to PredMap \,(List^{\wedge}\, A\, Q_A)\, (List^{\wedge}\, A\, Q'_A)}
+  \to PredMap\,A\,Q_A\,Q'_A \to PredMap\,(List\,A) \,(List^{\wedge}\, A\, Q_A)\, (List^{\wedge}\, A\, Q'_A)}
 \]
 takes a morphism of predicates and produces a morphism of lifted predicates. 
-$\mathsf{PredMap}$ gives the type of morphisms of predicates and is defined as: 
+$\mathsf{PredMap}$ is the type constructor that produces the type of morphisms between predicates, and it is defined as follows: 
 \begin{align*}
-  &\mathsf{PredMap : \forall\, \{A : Set\} \to (A \to Set) \to (A \to Set) \to Set } \\
-  &\mathsf{PredMap \, Q_A\,Q'_A\, = \forall\, (a : A) \to Q_A\,a \to Q'_A\,a}
+  &\mathsf{PredMap : \forall\, (A : Set) \to (A \to Set) \to (A \to Set) \to Set } \\
+  &\mathsf{PredMap \,A\, Q_A\,Q'_A\, = \forall\, (a : A) \to Q_A\,a \to Q'_A\,a}
 \end{align*}
@@
 \begin{array}{l}
-\mathsf{data\, G\,(\ol{a : Set}) : Set\,where}\\
-\mathsf{\;\;\;\;\;\;\;\;C_i\, :\, F_i\,G\,\ol{b} \to G (\ol{K_i\,\ol{b}})}
+\mathsf{data\ G : Set^\alpha \to Set\ where}\\
+\mathsf{\;\;\;\;\;\;\;\;c_i\, :\, F_i\,G\,\ol{B} \to G (\ol{K_i\,\ol{B}})}
 \end{array}
 \end{equation*}
-is defined by a finite number of constructors $\mathsf{C_i}$.
+is defined by a finite number of constructors $\mathsf{c_i}$.
 In the definition above, $\mathsf{F_i}$ is a type constructor with signature $\mathsf{(Set^{\alpha} \to Set) \to Set^{\beta} \to Set}$
@@
 The overline notation denotes a finite list:
-$\mathsf{\ol{a}}$ is a list of types of length $\mathsf{\alpha}$,
-so that it can be applied to the type constructor $\mathsf{G}$ of arity $\mathsf{\alpha}$.
+%$\mathsf{\ol{a}}$ is a list of types of length $\mathsf{\alpha}$,
+%so that it can be applied to the type constructor $\mathsf{G}$ of arity $\mathsf{\alpha}$.
 %and the vertical bars notation denotes the length of a list:
 %$\mathsf{G}$ depends on a finite list of types $\mathsf{\ol{a}}$ of length $\mathsf{\alpha}$,
-Each of the $\mathsf{\alpha}$-many $\mathsf{K_i}$ is a type constructor of arity $\mathsf{\beta}$
-so that it can be applied to the list of types $\mathsf{\ol{b}}$ of length $\mathsf{\beta}$.
+each of the $\mathsf{\alpha}$-many $\mathsf{K_i}$ is a type constructor of arity $\mathsf{\beta}$
+so that it can be applied to the list of types $\mathsf{\ol{B}}$ of length $\mathsf{\beta}$.
 %depends on a finite list of types $\mathsf{\ol{b}}$ of length $\mathsf{\beta}$,
-Moreover, notice that the arity of $\mathsf{G}$ matches the number of type constructors $\mathsf{\ol{K_i}}$.
+Moreover, notice that the arity of $\mathsf{G}$ matches the number of type constructors $\mathsf{K_i}$.
 We allow each $\mathsf{F_i}$ to be inductively built in the following ways (and with the following restrictions):
@@
 and $\mathsf{F_i''}$ has the same signature as $\mathsf{F_i}$ and is built recursively from the same induction rules.
-\item $\mathsf{F_i\,G\,\ol{b} = G(\ol{F_a\,\ol{b}})}$
-where none of the $\mathsf{F_a}$ contains the recursive variable,
-i.e., $\mathsf{F_a : Set^\beta \to Set}$
+\item $\mathsf{F_i\,G\,\ol{B} = G(\ol{F_j\,\ol{B}})}$
+where none of the $\mathsf{\alpha}$-many $\mathsf{F_j}$ contains the recursive variable,
+i.e., $\mathsf{F_j : Set^\beta \to Set}$
 is a type constructor of arity $\mathsf{\beta}$
-for each $\mathsf{a}$.
+for each $\mathsf{j = 1, \dots, \alpha}$.
 Such restriction is necessary to prevent nesting,
 as that would break the induction rule as discussed in Section~\ref{sec:GADT-nested}.
-\item $\mathsf{F_i\,G\,\ol{b} = H\,\ol{b}}$
+\item $\mathsf{F_i\,G\,\ol{B} = H\,\ol{B}}$
 where $\mathsf{H}$ is a type constructor of arity $\mathsf{\beta}$ not containing the recursive variable,
@@
 Notice that this covers the case in which $\mathsf{F_i}$ is a closed type,
-so, in particular, the unit and empty types, $\mathsf{1}$ and $\mathsf{0}$.
+so, in particular, the unit and empty types, $\mathsf{1}$ and $\mathsf{0}$,
+and the case in which $\mathsf{F_i}$ is a variable.
 %The only requirement on $\mathsf{H}$ is that it needs to have a predicate-lifting function
 %(and any type made of constant types, sums, products, arrow types and other GADTs satisfies this condition).
-\item $\mathsf{F_i\,G\,\ol{b} = H (\ol{F_c\,G\,\ol{b}})}$
+\item $\mathsf{F_i\,G\,\ol{B} = H (\ol{F_k\,G\,\ol{B}})}$
 where $\mathsf{H}$ is a $\mathsf{\gamma}$-ary type constructor not containing the recursive variable,
 i.e., $\mathsf{H : Set^\gamma \to Set}$,
-and $\mathsf{F_c}$ has the same signature as $\mathsf{F_i}$  and is built recursively from the same induction rules, for every $\mathsf{c = 1 \dots \gamma}$.
+and $\mathsf{F_k}$ has the same signature as $\mathsf{F_i}$  and is built recursively from the same induction rules, for every $\mathsf{k = 1 \dots \gamma}$.
 Moreover, we require that $\mathsf{H}$ is not a GADT itself (but we allow it to be an ADT or even a nested type).
-This way $\mathsf{H}$ admits functorial semantics~\cite{jp20}, and thus we have a map function for $\mathsf{H^{\wedge}}$,
-\[
-\mathsf{HLMap : \forall (\ol{c : Set}) (\ol{Q_c\;Q_c' : c \to Set})
-	\to \ol{PredMap\,c\,Q_c\,Q_c'}
-	\to PredMap\,(H\,\ol{c})\,(H^{\wedge}\,\ol{c}\,\ol{Q_c})\,(H^{\wedge}\,\ol{c}\,\ol{Q_c'})}
-\]
-where $\mathsf{PredMap : \forall (c : Set) \to (c \to Set) \to (c \to Set) \to Set}$
-is defined as
-\[
-\mathsf{PredMap\,c\,Q_c\,Q_c' = \forall (x : c) \to Q_c\,x \to Q_c'\,x}
-\]
-and represents the type of morphisms between predicates.
+This way we know that $\mathsf{H}$ admits functorial semantics~\cite{jp20}, and thus there is a map function for $\mathsf{H^{\wedge}}$,
+\[
+\mathsf{HLMap : \forall (\ol{C : Set}) (\ol{Q_C\;Q_C' : C \to Set})
+	\to \ol{PredMap\,C\,Q_C\,Q_C'}
+	\to PredMap\,(H\,\ol{C})\,(H^{\wedge}\,\ol{C}\,\ol{Q_C})\,(H^{\wedge}\,\ol{C}\,\ol{Q_C'})}
+\]
+%where $\mathsf{PredMap : \forall (C : Set) \to (C \to Set) \to (C \to Set) \to Set}$
+%is defined as
+%\[
+%\mathsf{PredMap\,C\,Q_c\,Q_c' = \forall (x : c) \to Q_c\,x \to Q_c'\,x}
+%\]
+%and represents the type of morphisms between predicates.
 A concrete way to define $\mathsf{HLMap}$ is to proceed by induction on the structure of the type $\mathsf{H}$,
@@
 \[
-\mathsf{F_i\,G\,\ol{b} :=
-F_i'\,G\,\ol{b} \times F_i''\,G\,\ol{b}
-\ \vert\ F_i'\,G\,\ol{b} + F_i''\,G\,\ol{b}
-\ \vert\ F_i'\,\ol{b} \to F_i''\,G\,\ol{b}
-\ \vert\ G(\ol{F_a\,\ol{b}})
-\ \vert\ H\,\ol{b}
-\ \vert\ H (\ol{F_c\,G\,\ol{b}})
+\mathsf{F_i\,G\,\ol{B} :=
+F_i'\,G\,\ol{B} \times F_i''\,G\,\ol{B}
+\ \vert\ F_i'\,G\,\ol{B} + F_i''\,G\,\ol{B}
+\ \vert\ F_i'\,\ol{B} \to F_i''\,G\,\ol{B}
+\ \vert\ G(\ol{F_j\,\ol{B}})
+\ \vert\ H\,\ol{B}
+\ \vert\ H (\ol{F_k\,G\,\ol{B}})
 }
@@
 \begin{array}{l}
-\mathsf{data\, G\,(a : Set) : Set\,where}\\
-\mathsf{\;\;\;\;\;\;\;\;C\, :\, F\,G\,\ol{b} \to G (K\,\ol{b})}
+\mathsf{data\ G : Set \to Set\ where}\\
+\mathsf{\;\;\;\;\;\;\;\;c\, :\, F\,G\,\ol{B} \to G (K\,\ol{B})}
 \end{array}
@@
 which, for ease of notation,
-we assume to be a unary type constructor (i.e. it depends on a single type parameter $\mathsf{a}$)
-and to have only one constructor $\mathsf{C}$.
+we assume to be a unary type constructor (i.e. it depends on a single type parameter $\mathsf{A}$)
+and to have only one constructor $\mathsf{c}$.
 Extending the argument to GADTs of arbitrary arity and with multiple constructors presents no difficulty
@@
 and each $\mathsf{K}$ has signature $\mathsf{Set^{\beta} \to Set}$.
-The constructor $\mathsf{C}$ can be rewritten using the $\mathsf{Equal}$ type as
-\[
-\mathsf{C : \exists (\ol{b : Set}) \to Equal\,a\,(K\,\ol{b}) \to F\,G\,\ol{b} \to G\,a}
+The constructor $\mathsf{c}$ can be rewritten using the $\mathsf{Equal}$ type as
+\[
+\mathsf{c : \forall (\ol{B : Set}) \to Equal\,A\,(K\,\ol{B}) \to F\,G\,\ol{B} \to G\,A}
 \]
@@
 \[
-\mathsf{G^{\wedge} : \forall (a : Set) \to (a \to Set) \to G\,a \to Set}
+\mathsf{G^{\wedge} : \forall (A : Set) \to (A \to Set) \to G\,A \to Set}
 \]
@@
 \[
-\mathsf{G^{\wedge}\,a\,Q_a\,(C\,\ol{b}\,e\,x)
-= \exists (\ol{Q_b : b \to Set})
-\to Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e
-\times F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,{\ol{Q_b}}\,x}
-\]
-where $\mathsf{Q_a : a \to Set}$
-and $\mathsf{C\,\ol{b}\,e\,x : G\,a}$,
-i.e., $\mathsf{e : Equal\,a\,(K\,\ol{b})}$ and $\mathsf{x : F\,G\,\ol{b}}$.
+\mathsf{G^{\wedge}\,A\,Q_A\,(c\,\ol{B}\,e\,x)
+= \sum [\ol{Q_B : B \to Set}]\,
+Equal^{\wedge}\,A\,(K\,\ol{B})\,Q_A\,(K^{\wedge}\,\ol{B}\,\ol{Q_B})\,e
+\times F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,{\ol{Q_B}}\,x}
+\]
+where $\mathsf{Q_A : A \to Set}$
+and $\mathsf{c\,\ol{B}\,e\,x : G\,A}$,
+i.e., $\mathsf{e : Equal\,A\,(K\,\ol{B})}$ and $\mathsf{x : F\,G\,\ol{B}}$.
 As already mentioned before, we also assume to have liftings for $\mathsf{F}$,
-\begin{multline*}
-\mathsf{F^{\wedge} : \forall (G : Set^{\alpha} \to Set) (\ol{b : Set})
-\to (\forall (a : Set) \to (a \to Set) \to G\,a \to Set)} \\
-\mathsf{\to (\ol{b \to Set})
-\to F\,G\,\ol{b} \to Set}
-\end{multline*}
+\[
+\mathsf{F^{\wedge} : \forall (G : Set^{\alpha} \to Set) (\ol{B : Set})
+\to (\forall (A : Set) \to (A \to Set) \to G\,A \to Set)
+\to (\ol{B \to Set})
+\to F\,G\,\ol{B} \to Set}
+\]
 and for  $\mathsf{K}$,
 \[
-\mathsf{K^{\wedge} : \forall (\ol{b : Set}) \to (\ol{b \to Set}) \to K\,\ol{b} \to Set}
+\mathsf{K^{\wedge} : \forall (\ol{B : Set}) \to (\ol{B \to Set}) \to K\,\ol{B} \to Set}
 \]
 Finally, associate the function
-\begin{multline*}
-\mathsf{CC = \lambda (P : \forall (a : Set) \to (a \to Set) \to G\,a \to Set)} \\
-\mathsf{\to \forall (a : Set) (\ol{b : Set}) (Q_a : a \to Set) (\ol{Q_b : b \to Set}) (e : Equal\,a\,(K\,\ol{b})) (x : F\,G\,\ol{b})} \\
-\mathsf{\to Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e
-	\to F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x
-	\to P\,a\,Q_a\,(C\,\ol{b}\,e\,x)}
-\end{multline*}
-to the constructor $\mathsf{C}$.
+\[
+\begin{array}{l}
+\mathsf{dIndC = \lambda (P : \forall (A : Set) \to (A \to Set) \to G\,A \to Set)} \\
+\quad \mathsf{\to \forall (A : Set) (\ol{B : Set}) (Q_A : A \to Set) (\ol{Q_B : B \to Set}) (e : Equal\,A\,(K\,\ol{B})) (x : F\,G\,\ol{B})} \\
+\quad \mathsf{\to Equal^{\wedge}\,A\,(K\,\ol{B})\,Q_A\,(K^{\wedge}\,\ol{B}\,\ol{Q_B})\,e
+	\to F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}\,x
+	\to P\,A\,Q_A\,(c\,\ol{B}\,e\,x)}
+\end{array}
+\]
+to the constructor $\mathsf{c}$.
 
@@
 The induction rule for $\mathsf{G}$ is
-\begin{multline*}
-\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to G\,a \to Set) \to CC\,P} \\
+\[
+\mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to G\,A \to Set) \to dIndC\,P
 %\mathsf{\to \Big(
@@
 %\Big)} \\
-\mathsf{\to \forall (a : Set)(Q_a : a \to Set)(y : G\,a) \to G^{\wedge}\,a\,Q_a\,y \to P\,a\,Q_a\,y}
-\end{multline*}
+\to \forall (A : Set)(Q_A : A \to Set)(y : G\,A) \to G^{\wedge}\,A\,Q_A\,y \to P\,A\,Q_A\,y}
+\]
 As we already did in the previous examples,
-we validate the induction rule by providing a term $\mathsf{DIG}$ for the type above.
+we validate the induction rule by providing a term $\mathsf{dIndG}$ for the type above.
 Define
 \[
-\mathsf{DIG\,P\,cc\,a\,Q_a\,(C\,\ol{b}\,e\,x)\,(\ol{Q_b}, L_E, L_F)
-= cc\,a\,\ol{b}\,Q_a\,\ol{Q_b}\,e\,x\,L_E\,(p\,x\,L_F)}
-\]
-where $\mathsf{cc : CC\,P}$ and
+\mathsf{dIndG\,P\,cc\,A\,Q_A\,(c\,\ol{B}\,e\,x)\,(\ol{Q_B}, liftE, liftF)
+= cc\,A\,\ol{B}\,Q_A\,\ol{Q_B}\,e\,x\,liftE\,(p\,x\,liftF)}
+\]
+where $\mathsf{cc : dIndC\,P}$ and
 \begin{itemize}
-\item $\mathsf{C\,\ol{b}\,e\,x : G\,a}$,
-i.e., $\mathsf{e : Equal\,a\,(K\,\ol{b})}$,
-and $\mathsf{x : F\,G\,\ol{b}}$;
-\item $\mathsf{(\ol{Q_b}, L_E, L_F) : G^{\wedge}\,a\,Q_a (C\,\ol{b}\,e\,x)}$,
+\item $\mathsf{c\,\ol{B}\,e\,x : G\,A}$,
+i.e., $\mathsf{e : Equal\,A\,(K\,\ol{B})}$,
+and $\mathsf{x : F\,G\,\ol{B}}$;
+\item $\mathsf{(\ol{Q_B}, liftE, liftF) : G^{\wedge}\,A\,Q_A (c\,\ol{B}\,e\,x)}$,
 i.e.,
-$\mathsf{Q_b : b \to Set}$ for each $\mathsf{b}$,
-$\mathsf{L_E : Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e}$,
-and $\mathsf{L_F : F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,{\ol{Q_b}}\,x}$.
+$\mathsf{Q_B : B \to Set}$ for each $\mathsf{B}$,
+$\mathsf{liftE : Equal^{\wedge}\,A\,(K\,\ol{B})\,Q_A\,(K^{\wedge}\,\ol{B}\,\ol{Q_B})\,e}$,
+and $\mathsf{liftF : F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,{\ol{Q_B}}\,x}$.
 \end{itemize}
@@
 \[
-\mathsf{p : PredMap\,(F\,G\,\ol{b})\,(F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}) (F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})}
+\mathsf{p : PredMap\,(F\,G\,\ol{B})\,(F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}) (F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}
 \]
@@
 \[
-\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}
-= Pair^{\wedge}(F_1\,G\,\ol{b})(F_2\,G\,\ol{b})(F_1^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}) (F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})}
+\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}
+= Pair^{\wedge}(F_1\,G\,\ol{B})(F_2\,G\,\ol{B})(F_1^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}) (F_2^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}
 \]
@@
 \begin{gather*}
-\mathsf{p_1 : PredMap\,(F_1\,G\,\ol{b})\,((F_1)^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}) ((F_1)^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})} \\
-\mathsf{p_2 : PredMap\,(F_2\,G\,\ol{b})\,((F_2)^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}) ((F_2)^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})}
+\mathsf{p_1 : PredMap\,(F_1\,G\,\ol{B})\,((F_1)^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}) ((F_1)^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})} \\
+\mathsf{p_2 : PredMap\,(F_2\,G\,\ol{B})\,((F_2)^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}) ((F_2)^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}
 \end{gather*}
-Thus, we define $\mathsf{p (x_1, x_2) (L_1, L_2) = (p_1\,x_1\,L_1, p_2\,x_2\,L_2)}$ 
+Thus, we define $\mathsf{p (x_1, x_2) (liftF_1, liftF_2) = (p_1\,x_1\,liftF_1, p_2\,x_2\,liftF_2)}$ 
 for 
-$\mathsf{x_1 : F_1\,G\,\ol{b}}$,
-$\mathsf{L_1 : F_1^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x_1}$,
-$\mathsf{x_2 : F_2\,G\,\ol{b}}$
-and $\mathsf{L_2 : F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x_2}$.
+$\mathsf{x_1 : F_1\,G\,\ol{B}}$,
+$\mathsf{liftF_1 : F_1^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}\,x_1}$,
+$\mathsf{x_2 : F_2\,G\,\ol{B}}$
+and $\mathsf{liftF_2 : F_2^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}\,x_2}$.
 %%%
@@
 \[
-\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x = \forall (z : F_1\,\ol{b}) \to F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,z \to F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,(x\,z)}
-\]
-where $\mathsf{x : F\,G\,\ol{b} = F_1\,\ol{b} \to F_2\,G\,\ol{b}}$.
+\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}\,x = \forall (z : F_1\,\ol{B}) \to F_1^{\wedge}\,\ol{B}\,\ol{Q_B}\,z \to F_2^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}\,(x\,z)}
+\]
+where $\mathsf{x : F\,G\,\ol{B} = F_1\,\ol{B} \to F_2\,G\,\ol{B}}$.
 By inductive hypothesis, there exist a morphism of predicates
 \[
-\mathsf{p_2 : PredMap\,(F_2\,G\,\ol{b}) (F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}) (F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})}
-\]
-Thus, we define $\mathsf{p\,x\,L_F : F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}$
-for $\mathsf{L_F : F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x}$
-as $\mathsf{p\,x\,L_F\,z\,L_1 = p_2 (x\,z) (L_F\,z\,L_1)}$
-for $\mathsf{z : F_1\,\ol{b}}$
-and $\mathsf{L_1 : F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,z}$.
+\mathsf{p_2 : PredMap\,(F_2\,G\,\ol{B}) (F_2^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}) (F_2^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}
+\]
+Thus, we define $\mathsf{p\,x\,liftF : F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}\,x}$
+for $\mathsf{liftF : F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}\,x}$
+as $\mathsf{p\,x\,liftF\,z\,liftF_1 = p_2 (x\,z) (liftF\,z\,liftF_1)}$
+for $\mathsf{z : F_1\,\ol{B}}$
+and $\mathsf{liftF_1 : F_1^{\wedge}\,\ol{B}\,\ol{Q_B}\,z}$.
 Notice that $\mathsf{F_1}$ not containing the recursive variable
 is a necessary restriction, as the proof relies on
-$\mathsf{F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x}$
-and $\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}$
-having the same domain $\mathsf{F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,z}$.
+$\mathsf{F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}\,x}$
+and $\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}\,x}$
+having the same domain $\mathsf{F_1^{\wedge}\,\ol{B}\,\ol{Q_B}\,z}$.
 %%%
-\item Case $\mathsf{F\,G\,\ol{b} = G(F'\,\ol{b})}$
+\item Case $\mathsf{F\,G\,\ol{B} = G(F'\,\ol{B})}$
 where $\mathsf{F'}$ does not contain the recursive variable,
 i.e., $\mathsf{F' : Set^\beta \to Set}$.
-Thus, $\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b} = P (F'\,\ol{b}) (F'^{\wedge}\,\ol{b}\,\ol{Q_b})}$.
+Thus, $\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B} = P (F'\,\ol{B}) (F'^{\wedge}\,\ol{B}\,\ol{Q_B})}$.
 So, $\mathsf{p}$ is defined as
 \[
-\mathsf{p = DIG\,P\,cc\,(F'\,\ol{b})\,(F'^{\wedge}\,\ol{b}\,\ol{Q_b})}
+\mathsf{p = dIndG\,P\,cc\,(F'\,\ol{B})\,(F'^{\wedge}\,\ol{B}\,\ol{Q_B})}
 \]
@@
 %%%
-\item Case $\mathsf{F\,G\,\ol{b} = H\,\ol{b}}$
+\item Case $\mathsf{F\,G\,\ol{B} = H\,\ol{B}}$
 where $\mathsf{H}$ is a $\mathsf{\beta}$-ary type constructor not containing the recursive variable,
 i.e., $\mathsf{H : Set^\beta \to Set}$.
-In such case, $\mathsf{p : PredMap (H\,\ol{b}) (H^{\wedge}\,\ol{b}\,\ol{Q_b}) (H^{\wedge}\,\ol{b}\,\ol{Q_b})}$
+In such case, $\mathsf{p : PredMap (H\,\ol{B}) (H^{\wedge}\,\ol{B}\,\ol{Q_B}) (H^{\wedge}\,\ol{B}\,\ol{Q_B})}$
 is just the identity morphism of predicates.
 %%%
-\item Case $\mathsf{F\,G\,\ol{b} = H (\ol{F_c\,G\,\ol{b}})}$
+\item Case $\mathsf{F\,G\,\ol{B} = H (\ol{F_k\,G\,\ol{B}})}$
 where $\mathsf{H}$ is a $\mathsf{\gamma}$-ary type constructor not containing the recursive variable,
 i.e., $\mathsf{H : Set^\gamma \to Set}$,
-and $\mathsf{F_c}$ has the same signature as $\mathsf{F}$, for every $\mathsf{c = 1 \dots \gamma}$.
+and $\mathsf{F_k}$ has the same signature as $\mathsf{F}$, for every $\mathsf{k = 1 \dots \gamma}$.
 Moreover, we assume that $\mathsf{H}$ has an associated predicate-lifting function,
 \[
-\mathsf{H^{\wedge} : \forall (\ol{c : Set}) \to (\ol{c \to Set}) \to H\,\ol{c} \to Set}
+\mathsf{H^{\wedge} : \forall (\ol{C : Set}) \to (\ol{C \to Set}) \to H\,\ol{C} \to Set}
 \]
@@
 \[
-\mathsf{\forall (\ol{c : Set}) (\ol{Q_c\;Q_c' : c \to Set}) \to PredMap\,\ol{c}\,\ol{Q_c}\,\ol{Q_c'}
-\to PredMap\,(H\,\ol{c})\,(H^{\wedge}\,\ol{c}\,\ol{Q_c})\,(H^{\wedge}\,\ol{c}\,\ol{Q_c'})}
+\mathsf{\forall (\ol{C : Set}) (\ol{Q_C\;Q_C' : C \to Set}) \to PredMap\,\ol{C}\,\ol{Q_C}\,\ol{Q_C'}
+\to PredMap\,(H\,\ol{C})\,(H^{\wedge}\,\ol{C}\,\ol{Q_C})\,(H^{\wedge}\,\ol{C}\,\ol{Q_C'})}
 \]
@@
 \[
-\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b} = H^{\wedge} (\ol{F_c\,G\,\ol{b}}) (\ol{F_c^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}})}
+\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B} = H^{\wedge} (\ol{F_k\,G\,\ol{B}}) (\ol{F_k^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}})}
 \]
@@
 \[
-\mathsf{p_c : PredMap\,(F_c\,G\,\ol{b}) (F_c^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}) (F_c^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b})}
-\]
-for every $\mathsf{c = 1 \dots \gamma}$.
+\mathsf{p_k : PredMap\,(F_k\,G\,\ol{B}) (F_k^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}) (F_k^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}
+\]
+for every $\mathsf{k = 1 \dots \gamma}$.
 So, $\mathsf{p}$ is defined as
@@
 \mathsf{p
-= HLMap\,(\ol{F_c\,G\,\ol{b}}) (\ol{F_c^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}}) (\ol{F_c^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}})\,\ol{p_c}}
+= HLMap\,(\ol{F_k\,G\,\ol{B}}) (\ol{F_k^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}}) (\ol{F_k^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}})\,\ol{p_k}}
 \]

### 2021-04-12 induction-paper/ENTCS-GADT-induction.tex
merging intro and draft

@@
-\documentclass[9pt]{entcs} \usepackage{entcsmacro}
+\documentclass[9pt]{entcs}
+\usepackage{entcsmacro}
 \usepackage{graphicx}
@@
 
-% A couple of exemplary definitions:
-
-\newcommand{\Nat}{{\mathbb N}}
-\newcommand{\Real}{{\mathbb R}}
 \def\lastname{Johann, Ghiorzi, and Jeffries}
@@
 \usepackage{amsmath}
+\usepackage[utf8]{inputenc}
+\usepackage{ccicons}
+\usepackage{verbatim}
+
+\usepackage{amscd}
+\usepackage{MnSymbol}
+\usepackage{xcolor}
+
+\usepackage{bbold}
+\usepackage{url}
+\usepackage{upgreek}
+\usepackage{stmaryrd}
+
+\usepackage{lipsum}
+\usepackage{tikz-cd}
+\usetikzlibrary{cd}
+\usetikzlibrary{calc}
+\usetikzlibrary{arrows}
+
+\usepackage{bussproofs}
+\EnableBpAbbreviations
+
+\newcommand{\new}[1]{{\color{blue} {#1}}}
+
+\newcommand{\set}{\mathtt{Set}}
 
@@
 
-  \author{Patricia Johann\thanksref{JohannEmail}}
-  \author{Enrico Ghiorzi\thanksref{GhiorziEmail}}
-  \author{Daniel Jeffries\thanksref{JeffriesEmail}}
-  \thanks[JohannEmail]{Email: \href{mailto:johannp@appstate.edu} {\texttt{\normalshape johannp@appstate.edu}}}
-  \thanks[GhiorziEmail]{Email: \href{mailto:ghiorzie@appstate.edu} {\texttt{\normalshape ghiorzie@appstate.edu}}}
-  \thanks[JeffriesEmail]{Email: \href{mailto:jeffriesd@appstate.edu} {\texttt{\normalshape jeffriesd@appstate.edu}}}
-  \address{Department of Computer Science\\ Appalachian State University}
+  \author{Patricia Johann~~~~}%\thanksref{JohannEmail}}
+  \author{Enrico Ghiorzi~~~~}%\thanksref{GhiorziEmail}}
+  \author{Daniel Jeffries}%\thanksref{JeffriesEmail}}
+%  \thanks[JohannEmail]{Email: \href{mailto:johannp@appstate.edu} {\texttt{\normalshape johannp@appstate.edu}}}
+%  \thanks[GhiorziEmail]{Email: \href{mailto:ghiorzie@appstate.edu} {\texttt{\normalshape ghiorzie@appstate.edu}}}
+%  \thanks[JeffriesEmail]{Email: \href{mailto:jeffriesd@appstate.edu} {\texttt{\normalshape jeffriesd@appstate.edu}}}
+  \address{$\mathtt{\{johannp,ghiorzie,jeffriesd\}@appstate.edu}$\\Department of Computer Science, Appalachian State University}
 
@@
 
+\begin{itemize}
+\item Weird to code in Agda if we're talking about induction rules for
+  Coq? 
+\item Agda style conventions
+\end{itemize}
+
 \section{Introduction}\label{sec:intro}
 
-Introduction goes here.
-
-\section{Deep induction for ADTs and nested types}
-
-%\begin{itemize}
-%\item Say what ADTs and nested types are (look~\cite{jp20,jgj21})
-%\item Examples: $\mathsf{List}$ and $\mathsf{PTree}$ and $\mathsf{Bush}$
-%\item Explain induction for $\mathsf{List}$ and $\mathsf{PTree}$ (refer paper Patricia Neil Clement 2010) notice it does not extend to truly nested types like $\mathsf{Bush}$
-%\item To get induction rule for truly nested types such as $\mathsf{Bush}$ requires deep induction (refer to~\cite{jp20}) (explain various pieces of deep induction, including lifting, informally)
-%\item Deep induction makes sense for other structures as well (not just truly nested types)
-%\end{itemize}
-
+Induction is one of the most important techniques available for
+working with advanced data types, so it is both inevitable and
+unsurprising that it plays an essential role in modern proof
+assistants. In the proof assistant Coq~\cite{coq20}, for example,
+functions and predicates over advanced types are defined inductively,
+and almost all non-trivial proofs of their properties are either
+proved by induction outright or rely on lemmas that are. Every time a
+new inductive data type is declared in Coq, an induction rule is
+automatically generated for it.
+
+The data types handled by Coq are (possibly mutually inductive)
+polynomial ADTs, and the induction rules automatically generated for
+them are the expected ones for standard structural induction. It has
+long been understood, however, that these rules are too weak to be
+genuinely useful for so-called {\em deep ADTs}~\cite{jp20}, i.e., ADTs
+that are (possibly mutually inductively) defined in terms of (other)
+such ADTs.\footnote{Such data types are called nested inductive types
+  by Chlipala~\cite{chl}, reflecting the fact that ``inductive type''
+  means ``ADT''~in~Coq.} Consider, for example, the following type of
+rose trees, here coded in Agda and defined in terms of the standard
+type of lists:
+\[\begin{array}{l}
+\mathsf{data\, Rose\, : Set \to Set\,where}\\
+\mathsf{\;\;\;\;\;\;\;\;rEmpty\, :\, Rose\,A}\\
+\mathsf{\;\;\;\;\;\;\;\;rNode\,\,\,\, :\, A \to List\,(Rose\,A) \to Rose\,A} 
+\end{array}\]
+\noindent
+The induction rule Coq automatically generates for rose trees is
+\[\begin{array}{l}
+\mathsf{\forall\, (a : \set)\,(P : Rose\,a \to \set) \to P\,REmpty \to}\\
+\mathsf{\hspace*{0.15in}
+ (\forall\, (x : a)\,(ts :
+  List\,(Rose\,a)) \to P\,(RNode\,x\,ts)) \to 
+  \forall \,(x : Rose\,a) \to P\,x}
+\end{array}\]
+\noindent
+Unfortunately, this is neither the induction rule we intuitively
+expect, nor is it expressive enough to prove even basic properties of
+rose trees that ought to be amenable to inductive proof. What is
+needed here is an enhanced notion of induction that, when specialized
+to rose trees, will propagate the predicate $\mathsf{P}$ through the
+outer list structure and to the rose trees sitting inside
+$\mathsf{RNode}$'s list argument. More generally, this enhanced notion
+of induction should traverse {\em all} of the structure present in a
+data type, propagating suitable predicates to {\em all} of the data
+contained in the structure. With data types becoming ever more
+advanced, and with deeply structured such types becoming ever more
+%increasingly
+ubiquitous in formalizations, it is critically important that proof
+assistants
+%like Coq, Agda, and Lean 
+be able to automatically generate genuinely useful induction
+principles for data types that go well beyond traditional ADTs. Such
+data types include (truly) nested types~\cite{bm98}\footnote{A truly
+  nested type is a nested type that is defined over itself.},
+generalized algebraic data types
+(GADTs)~\cite{ch03,pvww06,sp04,xcc03}, more richly indexed
+families~\cite{ch88}, and deep variants of all of these.
+
+{\em Deep induction}~\cite{jp20} is a generalization of structural
+induction that fits this bill exactly. Whereas structural induction
+rules induct over only the top-level structure of data, leaving any
+data internal to the top-level structure untouched, deep induction
+rules induct over {\em all} of the structured data present. The key
+idea is to parameterize induction rules not just over a predicate over
+the top-level data type being considered, but also over additional
+custom predicates on the types of primitive data they contain. These
+custom predicates are then lifted to predicates on any internal
+structures containing these data, and the resulting predicates on
+these internal structures are lifted to predicates on any internal
+structures containing structures at the previous level, and so on,
+until the internal structures at all levels of the data type
+definition, including the top level, have been so
+processed. Satisfaction of a predicate by the data at one level of a
+structure is then conditioned upon satisfaction of the appropriate
+predicates by {\em all} of the data at the preceding level.
+
+Deep induction was shown in~\cite{jp20} to be the form of induction
+most appropriate to nested types (including ADTs) that are defined
+over, or mutually recursively with, other such types (including,
+possibly, themselves). Deep induction delivers the following genuinely
+useful induction rule for rose trees:
+
+\[\begin{array}{l}
+\mathsf{\forall \,(a : \set)\,(P : Rose\,a \to \set)\,(Q : a \to \set)
+  \to P\,REmpty \to}\\ 
+\mathsf{\hspace*{0.15in}(\forall \,(x : a)\, (ts :
+  List\,(Rose\,a))\to Q\,x \to List^\land\,P\,ts \to P\,(RNode\,x\,ts)) \to}\\
+\mathsf{\hspace*{0.15in}\forall \,(x :
+  Rose\,a) \to Rose^\land\,Q\, x \to P\,x} 
+\end{array}\]
+
+\noindent
+%Here, $\mathsf{List^\land}$ lifts its predicate argument $\mathsf{P}$
+%on data of type $\mathsf{Rose\,a}$ to a predicate on data of type
+%$\mathsf{List\,(Rose\,a)}$ asserting that $\mathsf{P}$ holds for every
+%element of its argument list, and similarly for the lifting
+%$\mathsf{Rose^\land}$ and the predicate $\mathsf{Q}$.
+Here, $\mathsf{List^\land}$ (resp., $\mathsf{Rose^\land}$) lifts its
+predicate argument $\mathsf{P}$ (resp., $\mathsf{Q}$) on data of type
+$\mathsf{Rose\,a}$ (resp., $\mathsf{a}$) to a predicate on data of
+type $\mathsf{List\,(Rose\,a)}$ (resp., $\mathsf{Rose\,a}$) asserting
+that $\mathsf{P}$ (resp., $\mathsf{Q}$) holds for every element of its
+list (resp., rose tree) argument.\footnote{Predicate liftings such as
+  $\mathsf{List^\land}$ and $\mathsf{Rose^\land}$ can either be
+  supplied as primitives or generated automatically from their
+  associated data type definitions as described in
+  Section~\ref{sec:ADTs-and-nesteds} below. The predicate lifting for
+  a container type like $\mathsf{List\,t}$ or $\mathsf{Rose\,t}$
+  simply traverses containers of that type and applies its predicate
+  argument pointwise to the constituent data of type $\mathsf{t}$.}
+Deep induction was also shown in~\cite{jp20} to
+% be the missing piece making it possible to
+%be the theretofore missing component needed to
+% be the essential ingredient making it possible to
+deliver the first-ever induction rules --- structural or otherwise ---
+for the $\mathsf{Bush}$ data type~\cite{bm98} and other truly nested
+types. Deep induction for ADTs and nested types is reviewed in
+Section~\ref{sec:ADTs-and-nesteds} below.
+
+This paper shows how to extend deep induction to proper GADTs, i.e.,
+to GADTs that are not simply nested types (and thus are not ADTs).  A
+constructor for such a GADT $\mathsf{G}$ may, like a constructor for a
+nested type, take as arguments data whose types involve instances of
+$\mathsf{G}$ other than the one being defined --- including instances
+that involve $\mathsf{G}$ itself. But if $\mathsf{G}$ is a proper GADT
+then at least one of its constructors will also have such a structured
+instance of $\mathsf{G}$ --- albeit one not involving $\mathsf{G}$
+itself --- as its codomain.
+%But while the types of the arguments to a GADT's constructor can
+%include instances of $\mathsf{G}$ that themselves involve
+%$\mathsf{G}$, its return type cannot.
+For example, the constructor $\mathsf{SPair}$ for the GADT
+{\color{red} Perhaps Lam data type illustrates better? Perhaps also
+  shows non-inbabitation?}
+\begin{equation}\label{eq:eq-seq}
+\begin{array}{l}
+\mathsf{data\, Seq\,(a : Set) : Set\,where}\\
+\mathsf{\;\;\;\;\;\;\;\;SConst :\, a \to Seq\,a}\\
+\mathsf{\;\;\;\;\;\;\;\;SPair\,\, :\, Seq \,a \to Seq\,b \to
+  Seq\,(a \times b)}
+\end{array}
+\end{equation}
+\noindent
+of sequences only constructs sequences of pairs, rather than sequences
+of arbitrary type, as does $\mathsf{SConst}$. If all of the
+constructors for a GADT $\mathsf{G}$ return structured instances of
+$\mathsf{G}$, then some of $\mathsf{G}$'s instances might not be
+inhabited. GADTs therefore have two distinct, but equally natural,
+semantics: a functorial semantics interpreting them as left Kan
+extensions~\cite{mac71}, and a parametric semantics interpreting them
+as their Church encodings~\cite{}. As explained in~\cite{jgj21}, a key
+difference in the two semantics is that the former views GADTs as
+their {\em functorial completions}~\cite{jp19}, and thus as containing
+more data than just those expressible in syntax. By contrast, the
+latter views them as what might be called {\em syntax-only}
+GADTs. Happily, these two views of GADTs coincide for those that are
+ADTs or other nested types. However, both they and their attendant
+properties differ greatly for proper GADTs. In fact, the views
+deriving from the functorial semantics and that parametric semantics
+for proper GADTs are sufficiently distinct that, by contrast with the
+situation for ADTs and other nested
+types~\cite{bfss90,gjfor15,jgj21f}, it is not actually possible
+to define a functorial parametric semantics for them~\cite{jgj21}.
+
+This observation seems, at first, to be a death knell for the prospect
+of extending deep induction to GADTs. Indeed, since induction can be
+seen as unary parametricity, we quickly realize that GADTs viewed as
+their functorial completions cannot possibly support induction rules.
+This makes sense intuitively: induction is a syntactic proof
+technique, so of course it cannot be used to prove properties of those
+elements of a GADT's functorial completion that are not expressible in
+syntax. All is not lost, however. As we show below, the Church
+encoding interpretation's syntax-only view does support induction
+rules --- including deep induction rules --- for GADTs. Perhaps
+surprisingly, ours are the first-ever induction rules --- deep or
+otherwise --- for a general class of proper GADTs. But this paper
+actually delivers more: it gives a general framework for deriving deep
+induction rules for a general class of GADTs directly from their
+syntax. This framework can serve as a basis for extending modern proof
+assistants' automatic generation of structural induction rules for
+ADTs to automatic generation of deep induction rules for GADTs. As
+for ADTs and other nested types, the structural induction rule for any
+GADT can be recovered from its deep induction rule simply by taking
+the custom predicates in its deep induction rule to be constantly
+$\mathsf{True}$-valued predicates.
+
+Deep induction rules for GADTs cannot, however, be derived by somehow
+extending the techniques of~\cite{jp20} to syntax-only GADTs. Indeed,
+the derivation of induction rules given there makes crucial use of the
+functoriality of data types' interpretations from~\cite{jp19}, and
+that is precisely what the interpretation of GADTs as their Church
+encodings fails to deliver. Instead, we first give the (deep)
+induction rule and a predicate lifting styled after those
+of~\cite{jp20} for the simplest and most important GADT, namely the
+$\mathsf{Equality}$ GADT. We then treat any other GADT by writing it
+as its so-called {\em Henry Ford
+  encoding}~~\cite{ch03,hin03,mcb99,sjsv09,sp04} and using the
+predicate liftings for $\mathsf{Equality}$ and any other constituent
+data types to appropriately thread the custom predicates through the
+structures of the GADT. {\color{red} Significantly, since we are
+  treating GADTs as syntax-only, rather than functorial, we cannot
+  handle truly nested GADTs like we handled truly nested types, but
+  this is not much of a restriction, since I've never seen GADTs
+  defined over themselves in applications.} This process yields, in
+particular, the following deep induction rule for $\mathsf{Seq}$:
+{\color{red} Order of arguments?}
+\begin{multline*}
+\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to
+  Set)} \mathsf{\to CConst\,P \to CSPair\,P} \\ \mathsf{\to \forall (a
+  : Set)(Q_a : a \to Set)(s_a : Seq\,a) \to Seq^{\wedge}\,a\,Q_a\,s_a
+  \to P\,a\,Q_a\,s_a}
+\end{multline*}
+where $\mathsf{CConst}$ is the function {\color{red} Why $\lambda$?}
+\[\mathsf{\lambda (P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set)} 
+\mathsf{\to \forall (a : Set) (Q_a : a \to Set) (x : a) \to Q_a\,x \to
+  P\,a\,Q_a\,(Const\,x)}
+\]
+associated with the $\mathsf{Const}$ constructor and $\mathsf{CSPair}$
+is the function
+\begin{multline*}
+\mathsf{\lambda (P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set)}
+\mathsf{\to \forall (a\,b\,c : Set)
+	(Q_a : a \to Set) (Q_b : b \to Set) (Q_c : c \to Set)} \\
+	\mathsf{(s_b : Seq\,b) (s_c : Seq\,c) (e : Equal\,a\,(b \times c))
+	\to Equal^{\wedge} a\, (b \times c)\, Q_a\, (Pair^{\wedge}\,b\,c\,Q_b\,Q_c)\, e} \\
+	\mathsf{\to P\,b\,Q_b\,s_b \to P\,c\,Q_c\,s_c
+	\to P a Q_a ( SPair\,b\,c\,e\,s_b\,s_c )}
+\end{multline*}
+associated with the $\mathsf{SPair}$ constructor,
+
+All of the deep induction rules appearing in this paper have been
+derived using our general framework. Our Agda code implementing them
+is available at~\cite{web-page}.
+
+\subsection{Related Work}
+
+Ullrich's plugin?
+
+Tassi's Elpi
+
+Peter Selinger and  Fu Peng
+
+\section{Deep induction for ADTs and nested types}\label{sec:ADTs-and-nesteds}
+
+\begin{comment}
 \subsection{Syntax of ADTs and nested types}\label{ssec:syntax-adts}
@@
 \end{equation}
-Notice that all occurrences of $\mathsf{List}$ in the above encoding are instantiated at the same index $\mathsf{A}$.
-Thus, the instances of $\mathsf{List}$ at various indices are defined independently from one another.
-That is a defining feature of ADTs:
-an ADT defines a \emph{family of inductive types}, one for each index type.
-
-Over time, there has been a notable trend toward data types
-whose non-regular indexing can capture invariants and other sophisticated properties
-that can be used for program verification and other applications.
-A simple example of such a type is given by Bird and Meertens’~\cite{bm98} prototypical \emph{nested type}
+Notice that all occurrences of $\mathsf{List}$ in the above encoding
+are instantiated at the same index $\mathsf{A}$.  Thus, the instances
+of $\mathsf{List}$ at various indices are defined independently from
+one another.  That is a defining feature of ADTs: an ADT defines a
+\emph{family of inductive types}, one for each index type.
+
+Over time, there has been a notable trend toward data types whose
+non-regular indexing can capture invariants and other sophisticated
+properties that can be used for program verification and other
+applications.  A simple example of such a type is given by Bird and
+Meertens’~\cite{bm98} prototypical \emph{nested type}
 \begin{equation}\label{eq:ptree}
@@
 \end{equation}
-of perfect trees, which can be thought of as constraining lists to have lengths that are powers of 2.
-In the above code, the constructor $\mathsf{PNode}$ uses data of type $\mathsf{PNode\,(A \times A)}$ to construct data of type $\mathsf{PNode\,A}$.
-Thus, it is clear that the instantiations of $\mathsf{PNode}$ at various indices cannot be defined independently, so that the entire family of types must actually be defined at once.
-A nested type thus defines not a family of inductive types, but rather an \emph{inductive family of types}.
+of perfect trees, which can be thought of as constraining lists to
+have lengths that are powers of 2.  In the above code, the constructor
+$\mathsf{PNode}$ uses data of type $\mathsf{PNode\,(A \times A)}$ to
+construct data of type $\mathsf{PNode\,A}$.  Thus, it is clear that
+the instantiations of $\mathsf{PNode}$ at various indices cannot be
+defined independently, so that the entire family of types must
+actually be defined at once.  A nested type thus defines not a family
+of inductive types, but rather an \emph{inductive family of types}.
 
@@
 %type have to be the same type instance of the type being defined.
-
-\subsection{Induction principles for ADTs and nested types}\label{ssec:induction-nested}
-
-An induction principle for a data type allows to prove
-that a predicate holds for every element of that data type,
-provided that it holds for every element inductively produced by the type's constructors.
-In this paper, we are interested in induction principles for proof-relevant predicates.
-A proof-relevant predicate on a type $\mathsf{A : Set}$ is a function $\mathsf{A \to Set}$
-%(where $\mathsf{Set}$ is the type of sets)
-mapping each $\mathsf{x : A}$ to the set of proofs that the predicate holds for $\mathsf{x}$.
-For example, the induction principle for $\mathsf{List}$ is
+\end{comment}
+
+\subsection{Induction principles for ADTs and nested
+  types}\label{ssec:induction-nested}
+
+A structural induction principle for a data type allows us to prove
+that if a predicate holds for every element inductively produced by
+the data type's constructors then it holds for every element of the
+data type.  In this paper, we are interested in induction principles
+for proof-relevant predicates.  A proof-relevant predicate on a type
+$\mathsf{A : Set}$ is a function $\mathsf{P\,:\,A \to Set}$ mapping
+each $\mathsf{x : A}$ to the set of proofs that $\mathsf{P\,x}$ holds.
+For example, the induction principle for the standard data type
+\begin{equation*}\label{eq:list}
+\begin{array}{l}
+\mathsf{data\ List : Set \to Set\ where}\\
+\mathsf{\;\;\;\;\;\;\;nil\,\,\,\,\,\, :\, List\,A}\\
+\mathsf{\;\;\;\;\;\;\;cons\, :\, A \to List\,A \to List\,A} 
+\end{array}
+\end{equation*}
+of lists is
 \begin{equation*}
@@
 \forall (A : Set) (P : List\,A \to Set)
-\to P\,Nil
-\to \big( \forall (a : A) (ls: List\,A)
-\to P\,ls
-\to P\,(Cons\,a\,ls)\big)
-\to \forall (ls : List\,A)
-\to P\, ls
+\to P\,nil
+\to \big( \forall (a : A) (as: List\,A)
+\to P\,as
+\to P\,(cons\,a\,as)\big)
+\to \forall (as : List\,A)
+\to P\, as
 }
 \end{equation*}
-Note that the data inside a structure of type $\mathsf{List}$ is treated monolithically (i.e., ignored) by this induction rule.
-Indeed, the induction rule inducts over only the top-level structures of data types,
-leaving any data internal to the top-level structure untouched.
-Since this kind of induction principle is only concerned with the structure of the type,
-and unconcerned with the contained data,
-we will then refer to it as \emph{structural induction}.
-
-We can extend such a structural induction principle to some nested types, such as $\mathsf{PTree}$.
-The only difference from the induction principle for ADTs is that,
-since a nested type is defined as a whole inductive family of types at once,
-its induction rule has to necessary involve a polymorphic predicate.
-Thus, the induction rule for $\mathsf{PTree}$ is
-\[
-\begin{array}{l}
+As mentioned above, the data inside a structure of type
+$\mathsf{List}$ is treated monolithically (i.e., ignored) by this
+structural induction rule.
+%Indeed, the structural induction rule for a data type inducts
+%over only the top-level structures of data types, leaving any data
+%internal to the top-level structure untouched.  Since this kind of
+%induction principle is only concerned with the structure of the type,
+%and unconcerned with the contained data, we will then refer to it as
+%\emph{structural induction}.
+By contrast, the deep induction rule for lists is parameterized over a
+custom predicate $\mathsf{Q}$ on $\mathsf{A}$ as described in the
+introduction. For $\mathsf{List^\wedge}$ as described in the
+introduction, it is
+\[
+\begin{array}{l}
+\mathsf{\forall (A : Set) (P : List\, A \to Set) (Q : A \to Set)
+\to P\,Nil \to \big( \forall (a : A) (as: List\,A) \to Q\,a \to P\,as
+\to P\,(Cons\,a\,as)\big)} \\ 
+\quad\mathsf{\to \forall (as : List\,A) \to List^{\wedge}\,A\,Q\,as
+  \to P\,as } 
+\end{array}
+\]
+
+Structural induction can also be extended to nested types, such as the
+following type of perfect trees~\cite{bm98}:
+\begin{equation*}\label{eq:ptree}
+\begin{array}{l}
+\mathsf{data\ PTree : Set \to Set\ where}\\
+\mathsf{\;\;\;\;\;\;\;pLeaf\,\,\, :\, A \to PTree\,A}\\
+\mathsf{\;\;\;\;\;\;\;pNode\, :\, PTree\,(A \times A) \to PTree\,A} 
+\end{array}
+\end{equation*}
+Perfect trees can be thought of as lists constrained to have lengths
+that are powers of 2. In the above code, the constructor
+$\mathsf{pNode}$ uses data of type $\mathsf{PTree\,(A \times A)}$ to
+construct data of type $\mathsf{PTree\,A}$. Thus, it is clear that the
+instances of $\mathsf{PTree}$ at various indices cannot be defined
+independently, and that the entire inductive family of types must
+therefore be defined at once. This intertwinedness of the instances
+of nested types is reflected in their structural induction rules,
+which must necessarily involve polymorphic predicates, rather than the
+monomorphic predicates appearing in structural induction rules for
+ADTs. The structural induction rule for perfect trees, for example, is
+\[\begin{array}{l}
 \mathsf{\forall (P : \forall (A : Set) \to PTree\, A \to Set)
-\to \big( \forall (A : Set) (a : A) \to P\,A\,(PLeaf\, a) \big)} \\
-\quad\mathsf{\to \big( \forall (A : Set) (pt : PTree\,(A \times A)) \to P\,(A \times A)\,pt \to P\,a\,(PNode\,pt)\big)
-\to \forall (A : Set) (pt : PTree\,A) \to P\,A\,pt }
-\end{array}
-\]
-
-%This kind of induction principles, which we will call \emph{structural induction},
-Structural induction principles cannot be extended to truly nested types, such as $\mathsf{Bush}$.
-%To induce over a truly nested type, we need to use a \emph{deep induction} principle~\cite{jp20}.
-Instead, for such data types it is necessary to use a \emph{deep induction} principle~\cite{jp20}.
-Such a principle, unlike structural induction, inducts over all of the structured data present,
-by traversing not just the outer structure with a predicate $\mathsf{P}$,
-but also each data element contained in the data type with a custom predicate $\mathsf{Q}$.
-This additional predicate is lifted to predicates on any internal structure containing these data, and the resulting predicates on these internal structures are lifted to predicates on any internal structures containing structures at the previous level, and so on, until the internal structures at all levels of the data type definition, including the top level, have been so processed.
-Satisfaction of a predicate by the data at one level of a structure is then conditioned upon satisfaction of the appropriate predicates by all of the data at the preceding level.
-
-For example, the deep induction rule for $\mathsf{Bush}$ is
-\[
-\begin{array}{l}
+\to \big( \forall (A : Set) (a : A) \to P\,A\,(pLeaf\, a) \big)} \\
+\quad\mathsf{\to \big( \forall (A : Set) (tt : PTree\,(A \times A))
+  \to P\,(A \times A)\,t \to P\,a\,(pNode\,tt)\big) \to \forall (A :
+  Set) (t : PTree\,A) \to P\,A\,t } 
+\end{array}\]
+The deep induction rule for perfect trees similarly uses polymorphic
+predicates but otherwise follows the now-familiar pattern:
+\[\begin{array}{l}
+\mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to PTree\,A
+  \to Set) \to \big( \forall (A : Set) (Q : A \to Set) (a : A) \to
+  Q\,a \to P\,A\,Q\,(PLeaf\, a) \big)} \\ \quad \mathsf{\to \big(
+  \forall (A : Set) (Q : A \to Set) (tt : PTree\,(A \times A)) \to
+  P\,(A \times A)\,(Pair^{\wedge}\,A\,A\,Q\,Q)\,tt \to
+  P\,A\,Q\,(PNode\,tt)\big)} \\ \quad \mathsf{\to \forall (A : Set) (Q
+  : A \to Set) (t : PTree\,A) \to PTree^{\wedge}\,A\,Q\,t \to
+  P\,A\,Q\,t }
+\end{array}\]
+Here, $\mathsf{Pair^{\wedge} : \forall (A\; B: Set) \to (A \to Set)
+  \to (B \to Set) \to A \times B \to Set}$ lifts predicates
+$\mathsf{Q_A}$ on data of type $\mathsf{A}$ and $\mathsf{Q_B}$ on data
+of type $\mathsf{B}$ to a predicate on pairs of type $\mathsf{A \times
+  B}$ in such a way that $\mathsf{Pair^{\wedge}\,A\,B\,Q_A\,Q_B\,(a,b)
+  = Q_A\,a \times Q_B\,b}$. Similarly, $\mathsf{PTree^{\wedge} :
+  \forall (A : Set) \to (A \to Set) \to PTree\,A \to Set}$ lifts a
+predicate $\mathsf{Q}$ on data of type $\mathsf{A}$ to a predicate on
+data of type $\mathsf{PTree\,A}$ asserting that $\mathsf{Q}$ holds for
+every element of type $\mathsf{A}$ contained in its perfect tree
+argument.
+
+It is not possible to extend structural induction to {\em truly}
+nested types, i.e., to nested types whose recursive occurrences appear
+below their own type constructors. Indeed, defining any induction rule
+for such a type requires that we can lift the predicate argument to
+that rule over the type itself. The observation that any induction
+rule for a truly nested type would thus necessarily have to be a deep
+induction rule was, in fact, the original motivation for the
+introduction of deep induction in~\cite{jp20}.
+
+The quintessential example of a truly nested type is that of
+bushes~\cite{bm98}:
+\begin{equation*}\label{eq:bush}
+\begin{array}{l}
+\mathsf{data\ Bush : Set \to Set\ where}\\
+\mathsf{\;\;\;\;\;\;bNil\,\,\,\,\,\, :\, Bush\,A}\\
+\mathsf{\;\;\;\;\;\;bCons\, :\, A \to Bush\,(Bush\,A) \to Bush\,A} 
+\end{array}
+\end{equation*}
+Its (deep) induction rule is
+\[\begin{array}{l}
 \mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to Bush\, A \to Set)
-\to \big( \forall (A : Set) \to P\,A\,BNil \big)} \\
-\quad\mathsf{\to \big( \forall (A : Set) (Q : A \to Set) (a : A) (bb : Bush\,(Bush\,A))
-\to Q\,a \to P\,(Bush\,A)\,(Bush^{\wedge}\,A\,Q)\,bb \to P\,A\,Q\,(BCons\,a\,bb) \big)} \\
-\quad\mathsf{\to \forall (A : Set) (Q : A \to Set) (b : Bush\,A) \to Bush^{\wedge}\,A\,Q\,b \to P\,A\,Q\,b }
-\end{array}
-\]
-where $\mathsf{Bush^{\wedge} : \forall (A : Set) \to (A \to Set) \to Bush\,A \to Set}$ lifts a predicate $\mathsf{Q}$ on data of type $\mathsf{A}$ to a predicate on data of type $\mathsf{Bush\,A}$ asserting that $\mathsf{Q}$ holds for every element of type $\mathsf{A}$ contained in its argument bush.
-It is defined as
+\to \big( \forall (A : Set) \to P\,A\,bNil \big)} \\
+\quad\mathsf{\to \big( \forall (A : Set) (Q : A \to Set) (a : A) (bb :
+  Bush\,(Bush\,A)) \to Q\,a \to
+  P\,(Bush\,A)\,(Bush^{\wedge}\,A\,Q)\,bb \to P\,A\,Q\,(bCons\,a\,bb)
+  \big)} \\ 
+\quad\mathsf{\to \forall (A : Set) (Q : A \to Set) (b : Bush\,A) \to
+  Bush^{\wedge}\,A\,Q\,b \to P\,A\,Q\,b } 
+\end{array}\]
+where $\mathsf{Bush^{\wedge} : \forall (A : Set) \to (A \to Set) \to
+  Bush\,A \to Set}$ is the folllowing lifting of a predicate
+$\mathsf{Q}$ on data of type $\mathsf{A}$ to a predicate on data of
+type $\mathsf{Bush\,A}$ asserting that $\mathsf{Q}$ holds for every
+element of type $\mathsf{A}$ contained in its argument bush:
 \begin{align*}
-\mathsf{Bush^{\wedge}\,A\,Q\,BNil} &= \mathsf{\top} \\
-\mathsf{Bush^{\wedge}\,A\,Q\,(BCons\,a\,bb)} &= \mathsf{Q\,a \times Bush^{\wedge}\,(Bush\,A)\,(Bush^{\wedge}\,A\,Q)\,bb}
+\mathsf{Bush^{\wedge}\,A\,Q\,Bnil} &= \mathsf{\top} \\
+\mathsf{Bush^{\wedge}\,A\,Q\,(Bcons\,a\,bb)} &= \mathsf{Q\,a \times
+  Bush^{\wedge}\,(Bush\,A)\,(Bush^{\wedge}\,A\,Q)\,bb}
 \end{align*}
-where $\mathsf{Q : A \to Set}$ and $\mathsf{bb : Bush\,(Bush\,A)}$.
-
-Despite deep induction being motivated by the need to produce an induction principle for truly nested types,
-it can equally be applied to all other ADTs and nested types.
-For example, the deep induction principle for $\mathsf{List}$ is
-\[
-\begin{array}{l}
-\mathsf{\forall (A : Set) (P : List\, A \to Set) (Q : A \to Set)
-\to P\,Nil \to \big( \forall (a : A) (ls: List\,A) \to Q\,a \to P\,ls \to P\,(Cons\,a\,ls)\big)} \\
-\quad\mathsf{\to \forall (ls : List\,A) \to List^{\wedge}\,A\,Q\,ls \to P\,ls }
-\end{array}
-\]
-where $\mathsf{List^{\wedge} : \forall (A : Set) \to (A \to Set) \to List\,A \to Set}$ lifts a predicate $\mathsf{Q}$ on data of type $\mathsf{A}$ to a predicate on data of type $\mathsf{List\,A}$ asserting that $\mathsf{Q}$ holds for every element of its argument list.
-
-As a further example, the deep induction rule for $\mathsf{PTree}$ is
-\[
-\begin{array}{l}
-\mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to PTree\,A \to Set)
-  \to \big( \forall (A : Set) (Q : A \to Set) (a : A) \to Q\,a \to P\,A\,Q\,(PLeaf\, a) \big)} \\
-\quad \mathsf{\to \big( \forall (A : Set) (Q : A \to Set) (a : PTree\,(A \times A)) \to P\,(A \times A)\,(Pair^{\wedge}\,A\,A\,Q\,Q)\,a \to P\,A\,Q\,(PNode\,a)\big)} \\
-\quad \mathsf{\to \forall (A : Set) (Q : A \to Set) (a : PTree\,a) \to PTree^{\wedge}\,A\,Q\,a \to P\,A\,Q\,a }
-\end{array}
-\]
-where $\mathsf{Pair^{\wedge} : \forall (A : Set) (B : Set) \to (A \to Set) \to (B \to Set) \to A \times B \to Set}$ lifts predicates $\mathsf{Q_A}$ on data of type $\mathsf{A}$ and $\mathsf{Q_B}$ on data of type $\mathsf{B}$ to a predicate on pairs of type $\mathsf{A \times B}$,
-so that $\mathsf{Pair^{\wedge}\,A\,B\,Q_A\,Q_B\,(a,b) = Q_A\,a \times Q_B\,b}$,
-and $\mathsf{PTree^{\wedge} : \forall (A : Set) \to (A \to Set) \to PTree\,A \to Set}$ lifts a predicate $\mathsf{Q}$ on data of type $\mathsf{A}$ to a predicate on data of type $\mathsf{PTree\,A}$ asserting that $\mathsf{Q}$ holds for every element of type $\mathsf{A}$ contained in its argument perfect tree.
-
-Moreover, for types admitting both deep induction and structural induction, the former generalizes the latter.
-Indeed, structural induction rules can be derived from deep induction rules by always choosing the constantly true predicate as the custom predicate traversing each data element contained in the data type.
-That way, deep induction only inspects the structure of the data type and not its content, just like structural induction does.
-A concrete example of such derivation is demonstrated in Section~\ref{sec:ind-equal}.
-
-\section{Introducing GADTs}
-
-%\begin{itemize}
-%\item Say what GADTs are
-%\item Example: $\mathsf{Equal}$
-%\[\begin{array}{l}
-%\mathsf{data\, Equal\,a\,b\,where}\\
-%\mathsf{\;\;\;\;\;\;\;\;Refl ::\, Equal\,c\,c}
-%\end{array}\]
-%\item GADTs' encoding using $\mathsf{Equal}$
-%\item Example (in Haskell syntax):
-%%Haskell definition of $\mathsf{Seq}$ %(this encoding might have been introduced after the other one, see phantom types)
-%\begin{equation}\label{eq:seq}
-%\begin{array}{l}
-%\mathsf{data\, Seq\,a\,where}\\
-%\mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Seq\,a}\\
-%\mathsf{\;\;\;\;\;\;\;\;SPair\,\, ::\, Seq \,a \to Seq\,b \to
-%  Seq\,(a \times b)}
-%\end{array}
-%\end{equation}
-%\item Traditionally GADTs are understood in terms of $\mathsf{Equal}$ data type (see Cheney Hinze or Tim Sheard)
-%For example
-%\begin{equation}\label{eq:eq_seq}
-%\begin{array}{l}
-%\mathsf{data\, Seq\,a\,where}\\
-%\mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Seq\,a}\\
-%\mathsf{\;\;\;\;\;\;\;\;SPair\,\, ::\, \exists (b : Set) (c : Set).\,Equal\,a\,(b \times c) \to Seq\,b \to Seq\,c \to
-%  Seq\,a}\\ 
-%\end{array}
-%\end{equation}
-%(This gives the kind of Church encoding that is in Bob's paper) (might omit)
-%\item Add example for GADT with deep structure (Ask Daniel which one)
-%\end{itemize}
-%%Equality version of $\mathsf{Seq}$ (see Cheney Hinze or Tim Sheard)
-
-As noted in Subsection~\ref{ssec:syntax-adts},
-the return types of the constructors of a nested
-type have to be the same type instance of the type being defined.
-As a further generalization of ADTs and nested types,
-\emph{generalized algebraic data types} (GADTs)~\cite{ch03,xcc03,sp04}
-relax the restriction on the type instances appearing in a data
-type definition by allowing their constructors both to take as
-arguments \emph{and return as results} data whose types involve type
-instances of the GADT other than the one being defined.
+
+It is worth noting that for those nested types admitting both deep
+induction and structural induction, the former generalizes the latter.
+Indeed, the structural induction rule for a nested type is recoverable
+from its deep induction rule by taking the custom predicates on their
+data of primitive types to be constantly $\mathsf{True}$-valued
+predicates. Such an instantiation ensures that a deep induction rule
+only inspects the top-level structure of its argument data, rather
+than the contents of that structur. {\color{red} A concrete example of
+  such a derivation is given in Section~\ref{sec:ind-equal}.}
+
+\section{GADTs and Their Encodings}
+
+While a constructor of a nested type can take {\em as arguments} data
+whose types involve occurrences of the type instantiated at indices
+other than the one being defined, its {\em return type} still has to
+be at the same instance as the one being defined. For example, all
+constructors of $\mathsf{PTree\,A}$ have to return an element of type
+$\mathsf{PTree\,A}$. By contrast, \emph{generalized algebraic data
+  types} (GADTs)~\cite{ch03,sp04,xcc03} relax this restriction by
+allowing their constructors both to take as arguments \emph{and return
+  as results} data whose types involve type instances of the GADT
+other than the one being defined.
+%Applications of GADTs include generic programming, modeling
+%programming languages via higher-order abstract syntax, maintaining
+%invariants in data structures, and expressing constraints in embedded
+%domain-specific languages. GADTs have also been used, e.g., to
+%implement tagless interpreters~\cite{pl04,pr06,pvww06}, to improve
+%memory performance~\cite{min15}, and to design APIs~\cite{pen20}.
 
@@
 GADT's data constructors independently, and then using pattern
-matching to force the desired type refinement. Applications of GADTs
-include generic programming, modeling programming languages via
-higher-order abstract syntax, maintaining invariants in data
-structures, and expressing constraints in embedded domain-specific
-languages. GADTs have also been used, e.g., to implement tagless
-interpreters~\cite{pl04,pr06,pvww06}, to improve memory
-performance~\cite{min15}, and to design APIs~\cite{pen20}.
-
-\begin{example}\label{example:equal}
-As a first and notable example of GADT, we consider the the $\mathsf{Equal}$ type.
-This GADT is parametrized by two type indices,
-but it is only possible to construct a data element if the two indices are instantiated at the same type.
-In Agda, we code it as
+matching to force the desired type refinement. For example, the
+equality GADT
 \begin{equation}\label{eq:equal}
@@
 \mathsf{data\ Equal : Set \to Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;\;\;\;\;\,refl :\, Equal\,A\,A}
+\mathsf{\;\;\;\;\;\;\;\;\;\;\;\,eRefl :\, Equal\,A\,A}
 \end{array}
 \end{equation}
-$\mathsf{Equal}$ has thus a single data element when its two type arguments
-are the same and no data elements otherwise.
-The importance of the $\mathsf{Equal}$ type lies in the fact that we can understand other GADTs in terms of equality,
-as shown in the examples below.
-\end{example}
-
-\begin{example}\label{example:seq}
-A more complex example of GADT is the type of sequences,
-\begin{equation}\label{eq:eq-seq}
+is parametrized by two type indices, but it is only possible to
+construct data elements of type $\mathsf{Equal\,a\,b}$ if $\mathsf{a}$
+and $\mathsf{b}$ are instantiated at the same type. In this case
+$\mathsf{Equal\,a\,b}$ contains a single data element; it contains no
+data elements otherwise.
+
+The importance of the equality GADT lies in the fact that we can
+understand other GADTs in terms of it. For example, the GADT of
+sequences from the introduction
+%following GADT of sequences
+%\begin{equation}\label{eq:eq-seq}
+%\begin{array}{l}
+%\mathsf{data\ Seq : Set \to Set\ where}\\
+%\mathsf{\;\;\;\;\;\;\;\;sConst :\, A \to Seq\,A}\\
+%\mathsf{\;\;\;\;\;\;\;\;sPair\,\, :\, Seq\,A \to Seq\,B \to Seq\,(A \times B)}
+%\end{array}
+%\end{equation}
+comprises constant sequences of data of any type $\mathsf{A}$ and
+sequences obtained by pairing the data in two already existing
+sequences. This GADT can be understood in terms of the equality GADT
+by rewriting it as its Henry Ford
+encoding~\cite{ch03,hin03,mcb99,sjsv09,sp04} as follows:
+\begin{equation}\label{eq:eq_seq}
 \begin{array}{l}
 \mathsf{data\ Seq : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;\;const :\, A \to Seq\,A}\\
-\mathsf{\;\;\;\;\;\;\;\;sPair\,\, :\, Seq\,A \to Seq\,B \to Seq\,(A \times B)}
+\mathsf{\;\;\;\;\;\;\;\;sConst :\, A \to Seq\,A}\\ 
+\mathsf{\;\;\;\;\;\;\;\;sPair\,\, :\, \forall (B\,C : Set) \to
+  Equal\,A\,(B \times C) \to Seq\,B \to Seq\,C \to Seq\,A}\\ 
 \end{array}
 \end{equation}
-which comprises sequences of any type $\mathsf{a}$
-and sequences obtained by pairing the data in two already existing sequences.
-Such GADTs can be understood in terms of the $\mathsf{Equal}$ data type~\cite{ch03,sp04}.
-For example, we can rewrite the $\mathsf{Seq}$ type as
-\begin{equation}\label{eq:eq_seq}
-\begin{array}{l}
-\mathsf{data\ Seq : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;\;const :\, A \to Seq\,A}\\
-\mathsf{\;\;\;\;\;\;\;\;sPair\,\, :\, \forall (B\,C : Set) \to Equal\,A\,(B \times C)
-  \to Seq\,B \to Seq\,C \to Seq\,A}\\ 
+Here, the requirement that the $\mathsf{sPair}$ constructor produces
+an instance of $\mathsf{Seq}$ at a product type has been replaced with
+the requirement that the instance of $\mathsf{Seq}$ returned by
+$\mathsf{sPair}$ is \emph{equal} to some product type. This encoding
+is particularly convenient when viewing GADTs as their Church
+encodings~\cite{atk12,vw10}.
+
+\vspace*{0.1in}
+
+An example of a {\color{red} deep GADT} is given by an encoding of
+simply typed lambda terms. More robust variations on the simply typed
+lambda calculus are, of course, possible. But since the particular
+calculus we have chosen is rich enough to illustrate all of the
+essential aspects of deep induction for GADTs, we keep the types and
+terms to a minimum here for the purposes of exposition.
+
+Types are either booleans, arrow types, or list types. We represent
+types via the following GADT: {\color{red} Wrong naming conventions
+  throughout this example!}
+\begin{equation}\label{eq:eq_ltype}
+\begin{array}{l}
+\mathsf{data\ LType : Set \to Set\ where}\\
+\mathsf{\;\;\;\;\;\;\;\;tBool :\, \forall (B : Set) \to LType\,Bool}\\
+\mathsf{\;\;\;\;\;\;\;\;tArr\,\,\, :\, \forall (B\,C : Set) \to
+  LType\,B \to LType\,C \to LType\,(B \to C)}\\
+\mathsf{\;\;\;\;\;\;\;\;tList\,\, :\, \forall (B : Set)
+  \to LType\,B \to LType\,(List\,B)}
 \end{array}
 \end{equation}
-where the requirement that the $\mathsf{sPair}$ constructor produces an instance of $\mathsf{Seq}$ at a product type has been replaced with the requirement that the instance of $\mathsf{Seq}$ returned by $\mathsf{sPair}$ is \emph{equal} to some product type.
-This encoding is particularly convenient when representing GADTs as Church encodings~\cite{vw10,atk12}.
-\end{example}
-
-%To determine the possible shapes of a GADT's container we must
-%pattern-match on the type of the data to be contained.
-%For this, it is essential that a GADT calculus support an equality type
-%$\mathsf{Equal}$ that is a singleton set when its two type arguments
-%are the same and is the empty set otherwise.
-%That is, the type $\mathsf{Equal}$ must be the syntactic reflection
-%of semantic equality function $\mathit{Equal}$.
-%The type $\mathsf{Equal}$ can either be
-%defined via GADT syntax, as in
-%\[\begin{array}{l}
-%\mathsf{data\, Equal\,a\,b\,where}\\
-%\mathsf{\;\;\;\;\;\;\;\;Refl ::\, Equal\,c\,c}
-%\end{array}\]
-
-\begin{example}\label{example:lterm}
-GADTs are expressive enough to represent, for example, the types and terms of the simply typed lambda calculus. 
-We will use a variant of the simply typed lambda calculus including a base type of booleans,
-and type constructors for arrow types and list types. 
-We define a GADT $\mathsf{LType}$ to encode the set of permissible types for this calculus:
+Terms are either variables, abstractions, applications, or lists of
+such terms. We represent terms via 
+\begin{equation}\label{eq:eq_lterm}
+\begin{array}{l}
+\mathsf{data\ LTerm : Set \to Set\ where}\\
+\mathsf{\;\;\;\;\;\;\;\;var\,\,\,\,\,\,\,:\, String \to LType\,A \to
+  LTerm\,A} \\  
+\mathsf{\;\;\;\;\;\;\;\;abs\,\,\,\,\,\, :\, \forall (B\,C : Set) \to
+  String \to LType\,B \to LTerm\,C \to
+  LTerm\,(B \to C)}\\ 
+  \mathsf{\;\;\;\;\;\;\;\;app\,\,\,\,\, :\, \forall (B : Set) \to
+    LTerm (B \to A) \to LTerm\,B \to LTerm\,A} \\ 
+  \mathsf{\;\;\;\;\;\;\;\;listC\,\,\, :\, \forall (B : Set) \to
+    List\,(LTerm\,B) \to LTerm\,(List \,B)} 
+\end{array}
+\end{equation}
+The type parameter for $\mathsf{LTerm}$ tracks the types of simply
+typed lambda calculus terms.  For example, the type
+$\mathsf{LTerm\,A}$ contains lambda terms of type $\mathsf{A}$.
+Variables are tagged with their types by the constructors
+$\mathsf{var}$ and $\mathsf{abs}$. There, $\mathsf{LType}$ ensures
+that variables can only be tagged with legal types, e.g.,
+$\mathsf{Bool}$, $\mathsf{Bool \to Bool}$, $\mathsf{List\, (Bool \to
+  Bool)}$, etc. The presence of $\mathsf{LType}$ in these constructors
+thus guarantees that all lambda terms produced by $\mathsf{var}$,
+$\mathsf{abs}$, $\mathsf{app}$, and $\mathsf{listC}$ are well-typed.
+
+The Henry Ford encodings representing these GADTs in terms of the
+equality GADT are:
 \begin{equation}\label{eq:eq_ltype}
@@
 \mathsf{data\ LType : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;\;tBool :\, \forall (B : Set) \to Equal\,A\,Bool \to LType\,A}\\
-\mathsf{\;\;\;\;\;\;\;\;tArr\,\,\, :\, \forall (B\,C : Set) \to Equal\,A\,(B \to C) \to LType\,B \to LType\,C \to LType\,A}\\
-  \mathsf{\;\;\;\;\;\;\;\;tList\,\, :\, \forall (B : Set) \to Equal\,A\,(\List\,B) \to LType\,B \to LType\,A}
+\mathsf{\;\;\;\;\;\;\;\;tBool :\, \forall (B : Set) \to Equal\,A\,Bool
+  \to LType\,A}\\ 
+\mathsf{\;\;\;\;\;\;\;\;tArr\,\,\, :\, \forall (B\,C : Set) \to
+  Equal\,A\,(B \to C) \to LType\,B \to LType\,C \to LType\,A}\\ 
+  \mathsf{\;\;\;\;\;\;\;\;tList\,\, :\, \forall (B : Set) \to
+    Equal\,A\,(\List\,B) \to LType\,B \to LType\,A} 
 \end{array}
 \end{equation}
-The type system includes a $\mathsf{tBool}$ constructor representing the type of booleans;
-an arrow type constructor, $\mathsf{tArr}$ , which takes two types $\mathsf{B}$ and $\mathsf{C}$
-and produces the arrow type $\mathsf{B \to C}$;
-and a $\mathsf{tList}$ constructor which takes a type $\mathsf{B}$ and constructs
-the type $\mathsf{List\,B}$.
-
-%The $\mathsf{tBool}$ constructor represents the type of booleans in the calculus.
-%%It is defined in terms of $\mathsf{KBool : Set \to Set}$,
-%%which simply returns $\mathsf{Bool}$ on any input: $\mathsf{KBool \, A = Bool}$.
-%%The lifting $\mathsf{KBool^{\wedge} : \forall (A : Set) \to (A \to Set) \to (KBool \, A) \to Set}$ lifts every predicate to the constantly true predicate: 
-%%\[
-%%  \mathsf{KBool^{\wedge}\, A\, Q\, b = \top}
-%%\]
-%% The $\mathsf{Bool}$ data type is defined as:
-%% \begin{equation}\label{eq:bool}
-%% \begin{array}{l}
-%% \mathsf{data\, Bool\,: Set\,where}\\
-%%   \mathsf{\;\;\;\;\;\;\;\;true :\, Bool} \\
-%%   \mathsf{\;\;\;\;\;\;\;\;false:\, Bool} \\
-%% \end{array}
-%% \end{equation}
-%The type system also includes an arrow type constructor, $\mathsf{tArr}$ , which takes two types $\mathsf{B}$ and $\mathsf{C}$
-%and produces the arrow type $\mathsf{B \to C}$.
-%Similarly, given a type $\mathsf{B}$, we can construct
-%the type $\mathsf{List\,B}$ using the $\mathsf{tList}$ constructor.
-
-The term calculus for this type system includes variables, abstraction, application, 
-and also includes an introduction rule that takes a list of lambda terms of type $\mathsf{B}$ and produces a lambda term of type $\mathsf{List\,B}$.
+and
 \begin{equation}\label{eq:eq_lterm}
@@
 \mathsf{data\ LTerm : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;\;var\,\,\,\,\,\,\,:\, String \to LType\,A \to LTerm\,A} \\
-\mathsf{\;\;\;\;\;\;\;\;abs\,\,\,\,\,\, :\, \forall (B\,C : Set) \to Equal\,A\,(B \to C) \to String 
-      \to LType\,B \to LTerm\,C \to LTerm\,A}\\
-  \mathsf{\;\;\;\;\;\;\;\;app\,\,\,\,\, :\, \forall (B : Set) \to LTerm (B \to A) \to LTerm\,B \to LTerm\,A} \\
-  \mathsf{\;\;\;\;\;\;\;\;listC\,\,\, :\, \forall (B : Set) \to Equal\,A\,(List\,B) \to List\,(LTerm\,B) \to LTerm\,A}
+\mathsf{\;\;\;\;\;\;\;\;var\,\,\,\,\,\,\,:\, String \to LType\,A \to
+  LTerm\,A} \\  
+\mathsf{\;\;\;\;\;\;\;\;abs\,\,\,\,\,\, :\, \forall (B\,C : Set) \to
+  Equal\,A\,(B \to C) \to String \to LType\,B \to LTerm\,C \to
+  LTerm\,A}\\ 
+  \mathsf{\;\;\;\;\;\;\;\;app\,\,\,\,\, :\, \forall (B : Set) \to
+    LTerm (B \to A) \to LTerm\,B \to LTerm\,A} \\ 
+  \mathsf{\;\;\;\;\;\;\;\;listC\,\,\, :\, \forall (B : Set) \to
+    Equal\,A\,(List\,B) \to List\,(LTerm\,B) \to LTerm\,A} 
 \end{array}
 \end{equation}
-The type parameter for $\mathsf{LTerm}$ tracks the types of simply typed lambda calculus terms.
-So the type $\mathsf{LTerm\,A}$ contains lambda terms of type $\mathsf{A}$. 
-Variables are represented as strings and are tagged with their type.
-The $\mathsf{var}$ constructor takes a variable name and a type for the variable and produces
-a term of the given type.
-The $\mathsf{abs}$ constructor takes a variable name, a type $\mathsf{B}$ for the variable,
-and a lambda term of type $\mathsf{C}$ for the body of the abstraction,
-and it produces a term of type $\mathsf{B \to C}$.
-The $\mathsf{app}$ constructor applies a lambda term of type $\mathsf{B \to A}$ to a lambda term of type $\mathsf{B}$,
-producing a lambda term of type $\mathsf{A}$. 
-The $\mathsf{listC}$ constructor takes a list of lambda terms of type $\mathsf{B}$ and produces a lambda term of type $\mathsf{List\,B}$.
-
-Variables are tagged with their types in the $\mathsf{var}$ (variable introduction) and 
-$\mathsf{abs}$ (abstraction) constructors.
-The role of $\mathsf{LType}$ in the $\mathsf{var}$ and $\mathsf{abs}$ constructors
-is to enforce that variables can only be tagged with a legal type, e.g., $\mathsf{Bool}$, 
-$\mathsf{Bool \to Bool}$, $\mathsf{List\, (Bool \to Bool)}$, etc. 
-Therefore the presence of $\mathsf{LType}$ in these constructors ensures that all lambda terms produced by 
-$\mathsf{var}$, $\mathsf{abs}$, $\mathsf{app}$, and $\mathsf{listC}$ are well-typed. 
-\end{example}
+
+\vspace*{0.05in}
+
+We will revisit these GADTs in {\color{red} Sections~\ref{}
+  and~\ref{}}.
+
+\vfill
+
+\pagebreak
 
@@
 as these types do not allow for a functorial interpretation,
-at least in a parametric model~\cite{Haskell-paper}.
+at least in a parametric model~\cite{jgj21}.
 
@@
 That means that $\mathsf{H}$ cannot be a GADT,
-as GADTs have no functorial semantics~\cite{HaskellPaper}
+as GADTs have no functorial semantics~\cite{jgj21}.
 and incur in the issue exposed in Section~\ref{sec:GADT-nested},
@@
 and in the case of nested types it also relies on functorial semantics~\cite{jp20},
-but GADTs cannot admit both functorial and parametric semantics at the same time~\cite{HaskellPaper}.
+but GADTs cannot admit both functorial and parametric semantics at the same time~\cite{jgj21}.
 In this section we show how induction for GADTs featuring nesting goes wrong
@@
 In both cases, thus, we cannot derive an induction rule for GADTs featuring nesting.
-Unlike nestes types, indeed, GADTs do not admit a semantic interpretation that is both parametric and functorial~\cite{HaskellPaper}.
+Unlike nestes types, indeed, GADTs do not admit a semantic interpretation that is both parametric and functorial~\cite{jgj21}.
 
@@
 
-%%% Added to bibliography:
-
-%@inproceedings{jp20,
-%  author = {P. Johann and A. Polonsky},
-%  title = {Deep Induction: Induction Rules for (Truly) Nested Types},
-%  year = {2020},
-%  booktitle = {Foundations of Software Science and Computation Structures},
-%  pages = {339-358}
-%}
-%
-%@inproceedings{jgj21,
-%  author = {P. Johann and E. Ghiorzi and D. Jeffries},
-%  title = {Parametricity for Primitive Nested Types},
-%  year = {2021},
-%  booktitle = {Foundations of Software Science and Computation Structures},
-%  pages = {324-343}
-%}
-%
-%@inproceedings{bm98,
-% title={Nested datatypes},
-% author={R. Bird, and L. Meertens},
-% booktitle={Mathematics of Program Construction},
-% year={1998}
-% pages = {52–67}
-%}
-
-\bibliography{gadt-story-references}
+\begin{thebibliography}{10}\label{bibliography}
+
+\bibitem{atk12} Atkey, R.  {\em Relational parametricity for higher
+  kinds}.  Computer Science Logic, pp.~46-61, 2012.
+
+\bibitem{bfss90} Bainbridge, E. S., Freyd, P. Scedrov, A., and Scott,
+  P. J. {\em Functorial polymorphism}. Theoretical Computer Science
+  70(1), pp. 35-64, 1990.
+
+\bibitem{bm98} Bird, R. and Meertens, L. {\em Nested
+  datatypes}. Proceedings, Mathematics of Program Construction,
+  pp. 52–67, 1998.
+
+\bibitem{ch03} Cheney, J. and Hinze, R. {\em First-class phantom
+ types}. CUCIS TR2003-1901, Cornell University, 2003.
+
+\bibitem{ch88} Coquand, T. and Huet, G. {\em The calculus of
+  constructions}. Information and Computation 76(2/3), 1988.
+
+\bibitem{chl} Chlipala, A. {\em Library Inductive
+  Types}. $\mathtt{http://adam.chlipala.net/cpdt/html/InductiveTypes.html}$
+
+\bibitem{coq20} The Coq Development Team. {\em The Coq Proof
+  Assistant}, version 8.11.0, January 2020.
+  $\mathtt{https://doi.org/10.5281/zenodo.3744225}$
+
+\bibitem{gjfor15} Ghani, N., Johann, P., Nordvall Forsberg, F.,
+  Orsanigo, F., and Revell, T. {\em Bifibrational functorial semantics
+    for parametric polymorphism}. Proceedings, Mathematical
+  Foundations of Program Semantics, pp. 165-181, 2015.
+
+\bibitem{hin03} Hinze, R. {\em Fun with phantom types}. Proceedings,
+ The Fun of Programming, pp. 245–262, 2003.
+
+\bibitem{jgj21f} Johann, P., Ghiorzi, E., and Jeffries, D. {\em
+  Parametricity for primitive nested types}. Proceedings, Foundations
+  of Software Science and Computation Structures, pp. 324-343, 2021.
+
+\bibitem{jgj21} Johann, P., Ghiorzi, E., and Jeffries, D. {\em
+  Parametricity in the presence of GADTs}. Submitted, 2021.
+
+\bibitem{jp19} Johann, P. and Polonsky, A. {\em Higher-kinded data
+  types: Syntax and semantics} Proceedings, Logic in Computer Science
+  2019. {\color{red} PAGES?}
+
+\bibitem{jp20} Johann, P. and Polonsky, A. {\em Deep induction:
+  Induction rules for (truly) nested types}.  Proceedings, Foundations
+  of Software Science and Computation Structures, pp. 339-358, 2020.
+
+\bibitem{mac71} MacLane, S. {\em Catgories for the Working
+  Mathematician}. Springer, 1971.
+
+\bibitem{mcb99} McBride, C. {\em Dependently Typed Programs and their
+  Proofs}. PhD thesis, University of Edinburgh, 1999.
+
+\bibitem{min15} Minsky, Y.  {\em Why {GADT}s matter for performance}.
+  $\mathtt{https://blog.janestreet.com/why-gadts-matter-for-performance/}$,
+  2015.
+
+\bibitem{pl04} Pasalic, E., and Linger, N.  {\em Meta-programming with
+  typed object-language representations}.  Generic Programming and
+  Component Engineering, pp.~136-167, 2004.
+
+\bibitem{pen20} Penner, C.  {\em Simpler and safer {API} design using
+  {GADT}s}.  $\mathtt{https://chrispenner.ca/posts/gadt-design}$,
+  2020.
+
+\bibitem{pvww06} Peyton Jones, S., Vytiniotis, D., Weirich, S., and
+  Washburn, G. {\em Simple unification-based type inference for
+    GADTs}. Proceedings, International Conference on Functional
+  Programming, 2006. {\color{red} PAGES?}
+
+\bibitem{pr06} Pottier, F., and R{\'e}gis-Gianas, Y.  {\em Stratified
+  type inference for generalized algebraic data types}.  Principles of
+  Programming Languages, pp.~232-244, 2006.
+
+\bibitem{sjsv09} Schrijvers, T, Peyton Jones, S. L., Sulzmann, M., and
+  Vytiniotis, D. {\em Complete and decidable type inference for
+    GADTs}. Proceedings, International Conference on Functional
+  Programming, pp. 341– 352, 2009.
+
+\bibitem{sp04} Sheard, T., and Pasalic, E. {\em Meta-programming with
+  built-in type equality}. Proceedings, Workshop on Logical Frameworks
+  and Meta-languaegs, 2004. {\color{red} PAGES?}
+
+\bibitem{xcc03} Xi, H., Chen, C. and Chen, G. {\em Guarded recursive
+  datatype constructors}. Proceedings, Principles of Programming
+  Languages, pp. 224–235, 2003.
+
+\bibitem{vw10} Vytiniotis, D., and Weirich, S.  {\em Parametricity,
+  type equality, and higher-order polymorphism}.  Journal of
+  Functional Programming 20(2), pp.~175--210, 2010.
+
+\end{thebibliography}
 

### 2021-04-13 induction-paper/ENTCS-GADT-induction.tex
revising Sections 1-4

@@
 \item Agda style conventions
+ \item spacing in data type declarations
 \end{itemize}
@@
 \mathsf{data\, Rose\, : Set \to Set\,where}\\
-\mathsf{\;\;\;\;\;\;\;\;rEmpty\, :\, Rose\,A}\\
-\mathsf{\;\;\;\;\;\;\;\;rNode\,\,\,\, :\, A \to List\,(Rose\,A) \to Rose\,A} 
+\mathsf{\;\;\;\;\;\;\;\;empty\, :\, Rose\,A}\\
+\mathsf{\;\;\;\;\;\;\;\;node\,\,\,\, :\, A \to List\,(Rose\,A) \to Rose\,A} 
 \end{array}\]
@@
 \[\begin{array}{l}
-\mathsf{\forall\, (a : \set)\,(P : Rose\,a \to \set) \to P\,REmpty \to}\\
+\mathsf{\forall\, (a : \set)\,(P : Rose\,a \to \set) \to P\,empty \to}\\
 \mathsf{\hspace*{0.15in}
  (\forall\, (x : a)\,(ts :
-  List\,(Rose\,a)) \to P\,(RNode\,x\,ts)) \to 
+  List\,(Rose\,a)) \to P\,(node\,x\,ts)) \to 
   \forall \,(x : Rose\,a) \to P\,x}
@@
 outer list structure and to the rose trees sitting inside
-$\mathsf{RNode}$'s list argument. More generally, this enhanced notion
+$\mathsf{node}$'s list argument. More generally, this enhanced notion
 of induction should traverse {\em all} of the structure present in a
@@
 useful induction rule for rose trees:
-
-\[\begin{array}{l}
+\begin{equation}\label{eq:rose}
+\begin{array}{l}
 \mathsf{\forall \,(a : \set)\,(P : Rose\,a \to \set)\,(Q : a \to \set)
-  \to P\,REmpty \to}\\ 
+  \to P\,empty \to}\\ 
 \mathsf{\hspace*{0.15in}(\forall \,(x : a)\, (ts :
-  List\,(Rose\,a))\to Q\,x \to List^\land\,P\,ts \to P\,(RNode\,x\,ts)) \to}\\
+  List\,(Rose\,a))\to Q\,x \to List^\land\,P\,ts \to P\,(node\,x\,ts)) \to}\\
 \mathsf{\hspace*{0.15in}\forall \,(x :
   Rose\,a) \to Rose^\land\,Q\, x \to P\,x} 
-\end{array}\]
+\end{array}
+\end{equation}
 
@@
 %$\mathsf{G}$, its return type cannot.
-For example, the constructor $\mathsf{SPair}$ for the GADT
-{\color{red} Perhaps Lam data type illustrates better? Perhaps also
-  shows non-inbabitation?}
-\begin{equation}\label{eq:eq-seq}
+For example, the constructor $\mathsf{pair}$ for the GADT {\color{red}
+  Perhaps also show non-inhabitation?}
+\begin{equation}\label{eq:seq}
 \begin{array}{l}
 \mathsf{data\, Seq\,(a : Set) : Set\,where}\\
-\mathsf{\;\;\;\;\;\;\;\;SConst :\, a \to Seq\,a}\\
-\mathsf{\;\;\;\;\;\;\;\;SPair\,\, :\, Seq \,a \to Seq\,b \to
+\mathsf{\;\;\;\;\;\;\;\;const :\, a \to Seq\,a}\\
+\mathsf{\;\;\;\;\;\;\;\;\;\;\,pair :\, Seq \,a \to Seq\,b \to
   Seq\,(a \times b)}
@@
 of sequences only constructs sequences of pairs, rather than sequences
-of arbitrary type, as does $\mathsf{SConst}$. If all of the
+of arbitrary type, as does $\mathsf{const}$. If all of the
 constructors for a GADT $\mathsf{G}$ return structured instances of
@@
 extensions~\cite{mac71}, and a parametric semantics interpreting them
-as their Church encodings~\cite{}. As explained in~\cite{jgj21}, a key
-difference in the two semantics is that the former views GADTs as
-their {\em functorial completions}~\cite{jp19}, and thus as containing
-more data than just those expressible in syntax. By contrast, the
-latter views them as what might be called {\em syntax-only}
-GADTs. Happily, these two views of GADTs coincide for those that are
-ADTs or other nested types. However, both they and their attendant
-properties differ greatly for proper GADTs. In fact, the views
-deriving from the functorial semantics and that parametric semantics
-for proper GADTs are sufficiently distinct that, by contrast with the
+as their Church encodings~\cite{atk12,vw10}. As explained
+in~\cite{jgj21}, a key difference in the two semantics is that the
+former views GADTs as their {\em functorial completions}~\cite{jp19},
+and thus as containing more data than just those expressible in
+syntax. By contrast, the latter views them as what might be called
+{\em syntax-only} GADTs. Happily, these two views of GADTs coincide
+for those that are ADTs or other nested types. However, both they and
+their attendant properties differ greatly for proper GADTs. In fact,
+the views deriving from the functorial and parametric semantics for
+proper GADTs are sufficiently distinct that, by contrast with the
 situation for ADTs and other nested
-types~\cite{bfss90,gjfor15,jgj21f}, it is not actually possible
-to define a functorial parametric semantics for them~\cite{jgj21}.
+types~\cite{bfss90,gjfor15,jgj21f}, it is not actually possible to
+define a functorial parametric semantics for them~\cite{jgj21}.
 
@@
   handle truly nested GADTs like we handled truly nested types, but
-  this is not much of a restriction, since I've never seen GADTs
-  defined over themselves in applications.} This process yields, in
-particular, the following deep induction rule for $\mathsf{Seq}$:
-{\color{red} Order of arguments?}
-\begin{multline*}
-\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to
-  Set)} \mathsf{\to CConst\,P \to CSPair\,P} \\ \mathsf{\to \forall (a
-  : Set)(Q_a : a \to Set)(s_a : Seq\,a) \to Seq^{\wedge}\,a\,Q_a\,s_a
-  \to P\,a\,Q_a\,s_a}
-\end{multline*}
-where $\mathsf{CConst}$ is the function {\color{red} Why $\lambda$?}
-\[\mathsf{\lambda (P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set)} 
-\mathsf{\to \forall (a : Set) (Q_a : a \to Set) (x : a) \to Q_a\,x \to
-  P\,a\,Q_a\,(Const\,x)}
-\]
-associated with the $\mathsf{Const}$ constructor and $\mathsf{CSPair}$
-is the function
-\begin{multline*}
-\mathsf{\lambda (P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set)}
-\mathsf{\to \forall (a\,b\,c : Set)
-	(Q_a : a \to Set) (Q_b : b \to Set) (Q_c : c \to Set)} \\
-	\mathsf{(s_b : Seq\,b) (s_c : Seq\,c) (e : Equal\,a\,(b \times c))
-	\to Equal^{\wedge} a\, (b \times c)\, Q_a\, (Pair^{\wedge}\,b\,c\,Q_b\,Q_c)\, e} \\
-	\mathsf{\to P\,b\,Q_b\,s_b \to P\,c\,Q_c\,s_c
-	\to P a Q_a ( SPair\,b\,c\,e\,s_b\,s_c )}
-\end{multline*}
-associated with the $\mathsf{SPair}$ constructor,
+  this is not much of a restriction, since GADTs defined over
+  themselves have not, to our knowledge, appeared in applications.}
+This process yields, in particular, the deep induction rule for
+$\mathsf{Seq}$ given in Section~\ref{sec:ind-seq}. {\color{red} MORE
+  HERE???}
 
@@
 \end{equation}
-Notice that all occurrences of $\mathsf{List}$ in the above encoding
+Note that all occurrences of $\mathsf{List}$ in the above encoding
 are instantiated at the same index $\mathsf{A}$.  Thus, the instances
@@
 \end{equation*}
-As mentioned above, the data inside a structure of type
-$\mathsf{List}$ is treated monolithically (i.e., ignored) by this
-structural induction rule.
+As in Coq's induction rule for rose trees, the data inside a
+structure of type $\mathsf{List}$ is treated monolithically (i.e.,
+ignored) by this structural induction rule.
 %Indeed, the structural induction rule for a data type inducts
@@
 \mathsf{data\ PTree : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;pLeaf\,\,\, :\, A \to PTree\,A}\\
-\mathsf{\;\;\;\;\;\;\;pNode\, :\, PTree\,(A \times A) \to PTree\,A} 
+\mathsf{\;\;\;\;\;\;\;pleaf\,\,\, :\, A \to PTree\,A}\\
+\mathsf{\;\;\;\;\;\;\;pnode\, :\, PTree\,(A \times A) \to PTree\,A} 
 \end{array}
@@
 that are powers of 2. In the above code, the constructor
-$\mathsf{pNode}$ uses data of type $\mathsf{PTree\,(A \times A)}$ to
+$\mathsf{pnode}$ uses data of type $\mathsf{PTree\,(A \times A)}$ to
 construct data of type $\mathsf{PTree\,A}$. Thus, it is clear that the
@@
 \mathsf{\forall (P : \forall (A : Set) \to PTree\, A \to Set)
-\to \big( \forall (A : Set) (a : A) \to P\,A\,(pLeaf\, a) \big)} \\
+\to \big( \forall (A : Set) (a : A) \to P\,A\,(pleaf\, a) \big)} \\
 \quad\mathsf{\to \big( \forall (A : Set) (tt : PTree\,(A \times A))
-  \to P\,(A \times A)\,t \to P\,a\,(pNode\,tt)\big) \to \forall (A :
+  \to P\,(A \times A)\,tt \to P\,a\,(pnode\,tt)\big) \to \forall (A :
   Set) (t : PTree\,A) \to P\,A\,t } 
@@
   \to Set) \to \big( \forall (A : Set) (Q : A \to Set) (a : A) \to
-  Q\,a \to P\,A\,Q\,(PLeaf\, a) \big)} \\ \quad \mathsf{\to \big(
+  Q\,a \to P\,A\,Q\,(Pleaf\, a) \big)} \\ \quad \mathsf{\to \big(
   \forall (A : Set) (Q : A \to Set) (tt : PTree\,(A \times A)) \to
   P\,(A \times A)\,(Pair^{\wedge}\,A\,A\,Q\,Q)\,tt \to
-  P\,A\,Q\,(PNode\,tt)\big)} \\ \quad \mathsf{\to \forall (A : Set) (Q
+  P\,A\,Q\,(Pnode\,tt)\big)} \\ \quad \mathsf{\to \forall (A : Set) (Q
   : A \to Set) (t : PTree\,A) \to PTree^{\wedge}\,A\,Q\,t \to
@@
 \mathsf{data\ Bush : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;bNil\,\,\,\,\,\, :\, Bush\,A}\\
-\mathsf{\;\;\;\;\;\;bCons\, :\, A \to Bush\,(Bush\,A) \to Bush\,A} 
+\mathsf{\;\;\;\;\;\;bnil\,\,\,\,\,\, :\, Bush\,A}\\
+\mathsf{\;\;\;\;\;\;bvons\, :\, A \to Bush\,(Bush\,A) \to Bush\,A} 
 \end{array}
@@
 \mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to Bush\, A \to Set)
-\to \big( \forall (A : Set) \to P\,A\,bNil \big)} \\
+\to \big( \forall (A : Set) \to P\,A\,bnil \big)} \\ 
 \quad\mathsf{\to \big( \forall (A : Set) (Q : A \to Set) (a : A) (bb :
   Bush\,(Bush\,A)) \to Q\,a \to
-  P\,(Bush\,A)\,(Bush^{\wedge}\,A\,Q)\,bb \to P\,A\,Q\,(bCons\,a\,bb)
+  P\,(Bush\,A)\,(Bush^{\wedge}\,A\,Q)\,bb \to P\,A\,Q\,(bcons\,a\,bb)
   \big)} \\ 
@@
 where $\mathsf{Bush^{\wedge} : \forall (A : Set) \to (A \to Set) \to
-  Bush\,A \to Set}$ is the folllowing lifting of a predicate
+  Bush\,A \to Set}$ is the following lifting of a predicate
 $\mathsf{Q}$ on data of type $\mathsf{A}$ to a predicate on data of
@@
 \begin{align*}
-\mathsf{Bush^{\wedge}\,A\,Q\,Bnil} &= \mathsf{\top} \\
-\mathsf{Bush^{\wedge}\,A\,Q\,(Bcons\,a\,bb)} &= \mathsf{Q\,a \times
+\mathsf{Bush^{\wedge}\,A\,Q\,bnil} &= \mathsf{\top} \\
+\mathsf{Bush^{\wedge}\,A\,Q\,(bcons\,a\,bb)} &= \mathsf{Q\,a \times
   Bush^{\wedge}\,(Bush\,A)\,(Bush^{\wedge}\,A\,Q)\,bb}
@@
 
-It is worth noting that for those nested types admitting both deep
-induction and structural induction, the former generalizes the latter.
-Indeed, the structural induction rule for a nested type is recoverable
-from its deep induction rule by taking the custom predicates on their
-data of primitive types to be constantly $\mathsf{True}$-valued
-predicates. Such an instantiation ensures that a deep induction rule
-only inspects the top-level structure of its argument data, rather
-than the contents of that structur. {\color{red} A concrete example of
-  such a derivation is given in Section~\ref{sec:ind-equal}.}
-
-\section{GADTs and Their Encodings}
+It is worth noting that for those nested types that do admit both
+structural induction and deep induction, the latter generalizes the
+former. Indeed, the structural induction rule for a nested type is
+recoverable from its deep induction rule by taking the custom
+predicates on their data of primitive types to be constantly
+$\mathsf{True}$-valued predicates. Such an instantiation ensures that
+a deep induction rule only inspects the top-level structure of its
+argument data, rather than the contents of that structure.  A concrete
+example of such a derivation is given in Section~\ref{sec:ind-equal}.
+
+\section{(Deep) GADTs}\label{sec:GADTs}
 
 While a constructor of a nested type can take {\em as arguments} data
-whose types involve occurrences of the type instantiated at indices
-other than the one being defined, its {\em return type} still has to
-be at the same instance as the one being defined. For example, all
-constructors of $\mathsf{PTree\,A}$ have to return an element of type
-$\mathsf{PTree\,A}$. By contrast, \emph{generalized algebraic data
-  types} (GADTs)~\cite{ch03,sp04,xcc03} relax this restriction by
-allowing their constructors both to take as arguments \emph{and return
-  as results} data whose types involve type instances of the GADT
-other than the one being defined.
+whose types involve instances of that type at indices other than the
+one being defined, its return type must still be at the (variable)
+type instance being defined. For example, every constructor of
+$\mathsf{PTree\,A}$ must return an element of type
+$\mathsf{PTree\,A}$. \emph{Generalized algebraic data types}
+(GADTs)~\cite{ch03,sp04,xcc03} relax this restriction, allowing their
+constructors both to take as arguments \emph{and return as results}
+data whose types involve instances of the GADT other than the one
+being defined. As with the return type of the $\mathsf{sPair}$
+constructor in~\eqref{eq:seq}, these indices can be structured.
 %Applications of GADTs include generic programming, modeling
@@
 GADTs are used in precisely those situations in which different
-behaviors at different instances of a data type are desired. This is
+behaviors at different instances of data types are desired. This is
 achieved by allowing the programmer to give the type signatures of the
-GADT's data constructors independently, and then using pattern
-matching to force the desired type refinement. For example, the
-equality GADT
+GADT's data constructors independently, and then using {\color{red}
+  pattern matching} to force the desired type refinement. For example,
+the {\em equality} GADT
 \begin{equation}\label{eq:equal}
@@
 \mathsf{data\ Equal : Set \to Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;\;\;\;\;\,eRefl :\, Equal\,A\,A}
+\mathsf{\;\;\;\;\;\;\;\;\;\;\;\,refl :\, Equal\,A\,A}
 \end{array}
@@
 construct data elements of type $\mathsf{Equal\,a\,b}$ if $\mathsf{a}$
-and $\mathsf{b}$ are instantiated at the same type. In this case
-$\mathsf{Equal\,a\,b}$ contains a single data element; it contains no
-data elements otherwise.
+and $\mathsf{b}$ are instantiated at the same type. If the types
+$\mathsf{a}$ and $\mathsf{b}$ are syntactically identical then the
+type $\mathsf{Equal\,a\,b}$ contains a single data element. It
+contains no data elements otherwise.
 
@@
 understand other GADTs in terms of it. For example, the GADT of
-sequences from the introduction
-%following GADT of sequences
-%\begin{equation}\label{eq:eq-seq}
-%\begin{array}{l}
-%\mathsf{data\ Seq : Set \to Set\ where}\\
-%\mathsf{\;\;\;\;\;\;\;\;sConst :\, A \to Seq\,A}\\
-%\mathsf{\;\;\;\;\;\;\;\;sPair\,\, :\, Seq\,A \to Seq\,B \to Seq\,(A \times B)}
-%\end{array}
-%\end{equation}
-comprises constant sequences of data of any type $\mathsf{A}$ and
-sequences obtained by pairing the data in two already existing
-sequences. This GADT can be understood in terms of the equality GADT
-by rewriting it as its Henry Ford
-encoding~\cite{ch03,hin03,mcb99,sjsv09,sp04} as follows:
+sequences in~\eqref{eq:seq} comprises constant sequences of data of
+any type $\mathsf{A}$ and sequences obtained by pairing the data in
+two already existing sequences. This GADT can be understood as its
+{\em Henry Ford encoding}~\cite{ch03,hin03,mcb99,sjsv09,sp04}, which
+makes critical use of the equality GADT, as follows:
 \begin{equation}\label{eq:eq_seq}
@@
 \mathsf{data\ Seq : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;\;sConst :\, A \to Seq\,A}\\ 
-\mathsf{\;\;\;\;\;\;\;\;sPair\,\, :\, \forall (B\,C : Set) \to
+\mathsf{\;\;\;\;\;\;\;const :\, A \to Seq\,A}\\ 
+\mathsf{\;\;\;\;\;\;\;\;pair\,\, :\, \forall (B\,C : Set) \to
   Equal\,A\,(B \times C) \to Seq\,B \to Seq\,C \to Seq\,A}\\ 
@@
 \end{equation}
-Here, the requirement that the $\mathsf{sPair}$ constructor produces
-an instance of $\mathsf{Seq}$ at a product type has been replaced with
-the requirement that the instance of $\mathsf{Seq}$ returned by
-$\mathsf{sPair}$ is \emph{equal} to some product type. This encoding
-is particularly convenient when viewing GADTs as their Church
-encodings~\cite{atk12,vw10}.
-
-\vspace*{0.1in}
-
-An example of a {\color{red} deep GADT} is given by an encoding of
-simply typed lambda terms. More robust variations on the simply typed
-lambda calculus are, of course, possible. But since the particular
-calculus we have chosen is rich enough to illustrate all of the
-essential aspects of deep induction for GADTs, we keep the types and
-terms to a minimum here for the purposes of exposition.
-
-Types are either booleans, arrow types, or list types. We represent
-types via the following GADT: {\color{red} Wrong naming conventions
-  throughout this example!}
-\begin{equation}\label{eq:eq_ltype}
+Here, the requirement that $\mathsf{pair}$ produce data at an
+instance of $\mathsf{Seq}$ that is a product type is replaced with the
+requirement that $\mathsf{pair}$ produce data at an instance of
+$\mathsf{Seq}$ that is \emph{equal} to a product type. As we will see
+below, this use of the equality GADT is key to deriving deep induction
+rules for more general GADTs.
+
+\vspace*{0.05in}
+
+Finally, we give an example of a deep GADT. The $\mathsf{LTerm}$ GADT
+encodes terms of a simply typed lambda calculus. More robust
+variations on this calculus are, of course, possible. But since
+$\mathsf{LTerm}$ is rich enough to illustrate all of the essential
+aspects of deep GADTs --- and later, in Section~\ref{sec:ind-lam},
+their deep induction rules --- while still being small enough to
+ensure clarity of exposition, we keep our calculus to a minimum.
+
+Types are either booleans, arrow types, or list types, and are
+represented by the following GADT:
+\begin{equation*}\label{eq:eq_ltype}
 \begin{array}{l}
 \mathsf{data\ LType : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;\;tBool :\, \forall (B : Set) \to LType\,Bool}\\
-\mathsf{\;\;\;\;\;\;\;\;tArr\,\,\, :\, \forall (B\,C : Set) \to
+\mathsf{\;\;\;\;\;\;\;\;bool :\, \forall (B : Set) \to LType\,Bool}\\
+\mathsf{\;\;\;\;\;\;\;\;arr\,\,\, :\, \forall (B\,C : Set) \to
   LType\,B \to LType\,C \to LType\,(B \to C)}\\
-\mathsf{\;\;\;\;\;\;\;\;tList\,\, :\, \forall (B : Set)
+\mathsf{\;\;\;\;\;\;\;\;list\,\, :\, \forall (B : Set)
   \to LType\,B \to LType\,(List\,B)}
 \end{array}
-\end{equation}
+\end{equation*}
 Terms are either variables, abstractions, applications, or lists of
-such terms. We represent terms via 
-\begin{equation}\label{eq:eq_lterm}
+such terms. They are represented by
+\begin{equation*}\label{eq:eq_lterm}
 \begin{array}{l}
@@
     LTerm (B \to A) \to LTerm\,B \to LTerm\,A} \\ 
-  \mathsf{\;\;\;\;\;\;\;\;listC\,\,\, :\, \forall (B : Set) \to
+  \mathsf{\;\;\;\;\;\;\;\;list\,\,\, :\, \forall (B : Set) \to
     List\,(LTerm\,B) \to LTerm\,(List \,B)} 
 \end{array}
-\end{equation}
-The type parameter for $\mathsf{LTerm}$ tracks the types of simply
-typed lambda calculus terms.  For example, the type
-$\mathsf{LTerm\,A}$ contains lambda terms of type $\mathsf{A}$.
-Variables are tagged with their types by the constructors
-$\mathsf{var}$ and $\mathsf{abs}$. There, $\mathsf{LType}$ ensures
-that variables can only be tagged with legal types, e.g.,
-$\mathsf{Bool}$, $\mathsf{Bool \to Bool}$, $\mathsf{List\, (Bool \to
-  Bool)}$, etc. The presence of $\mathsf{LType}$ in these constructors
-thus guarantees that all lambda terms produced by $\mathsf{var}$,
-$\mathsf{abs}$, $\mathsf{app}$, and $\mathsf{listC}$ are well-typed.
-
-The Henry Ford encodings representing these GADTs in terms of the
-equality GADT are:
+\end{equation*}
+The type parameter for the GADT $\mathsf{LTerm}$ tracks the types of
+simply typed lambda calculus terms. For example, $\mathsf{LTerm\,A}$
+is the type of lambda terms of type $\mathsf{A}$. Variables are tagged
+with their types by the data constructors $\mathsf{var}$ and
+$\mathsf{abs}$, whose $\mathsf{LType}$ arguments ensure that the
+tagging types are legal.
+%types, e.g., $\mathsf{Bool}$, $\mathsf{Bool \to Bool}$,
+%  $\mathsf{List\, (Bool \to Bool)}$, etc. That is, the presence of
+%$\mathsf{LType}$ in these constructors thus guarantees that
+As a result, all lambda terms produced by $\mathsf{var}$,
+$\mathsf{abs}$, $\mathsf{app}$, and $\mathsf{list}$ are necessarily
+well-typed.
+
+{\color{red} Maybe only give in the following form.}
+The Henry Ford encodings representing $\mathsf{LType}$ and
+$\mathsf{LTerm}$ in terms of the equality GADT are:
 \begin{equation}\label{eq:eq_ltype}
@@
 \mathsf{data\ LType : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;\;tBool :\, \forall (B : Set) \to Equal\,A\,Bool
+\mathsf{\;\;\;\;\;\;\;\;bool :\, \forall (B : Set) \to Equal\,A\,Bool
   \to LType\,A}\\ 
-\mathsf{\;\;\;\;\;\;\;\;tArr\,\,\, :\, \forall (B\,C : Set) \to
+\mathsf{\;\;\;\;\;\;\;\;arr\,\,\, :\, \forall (B\,C : Set) \to
   Equal\,A\,(B \to C) \to LType\,B \to LType\,C \to LType\,A}\\ 
-  \mathsf{\;\;\;\;\;\;\;\;tList\,\, :\, \forall (B : Set) \to
+  \mathsf{\;\;\;\;\;\;\;\;list\,\, :\, \forall (B : Set) \to
     Equal\,A\,(\List\,B) \to LType\,B \to LType\,A} 
@@
     LTerm (B \to A) \to LTerm\,B \to LTerm\,A} \\ 
-  \mathsf{\;\;\;\;\;\;\;\;listC\,\,\, :\, \forall (B : Set) \to
+  \mathsf{\;\;\;\;\;\;\;\;list\,\,\, :\, \forall (B : Set) \to
     Equal\,A\,(List\,B) \to List\,(LTerm\,B) \to LTerm\,A} 
@@
 \end{equation}
-
-\vspace*{0.05in}
-
-We will revisit these GADTs in {\color{red} Sections~\ref{}
-  and~\ref{}}.
-
-\vfill
-
-\pagebreak
-
-\section{(Deep) induction for GADTs}
-
-%\begin{itemize}
-%    \item Recall structural induction does not work for GADTs because it does not work for truly nested types.
-%    \item Induction rules are traditionally derived using functoriality
-%    \item This cannot be extended to GADTs in this setting because they have no functorial semantics
-%    \item This paper nevertheless shows how to extend (deep) induction to GADTs.
-%\end{itemize}
-
-%As we have seen in Section~\ref{ssec:induction-nested},
-%truly nested types do not support a structural induction rule,
-%which is the reason why it was necessary to introduce a deep induction rule supporting them.
-%Consequently, GADTs do not support a structural induction rule either,
-%as they generalize nested types.
-%Still, there is hope for GADTs to support a deep induction rule,
-%like nested types do.
-
-Reasoning about GADTs can generally be quite difficult,
-even more so than for ADTs and nested types.
-To help with that,
-we want to extend the deep induction principle seen in Section~\ref{ssec:induction-nested} to GADTs.
-
-Induction rules, and specifically deep induction rules for nested types,
-are traditionally derived using the functorial semantics of data types
-in the setting of a parametric model~\cite{jp19}.
-In particular, relational parametricity is used to validate the induction principle because induction is, itself,
-a form of unary parametricity, where binary relations have been replaced with predicates,
-which are essentially unary relations.
-{\color{blue} Is there a reference for this?}
-%in the sense that induction involves predicates, which are unary relations,
-%as opposed to the binary relations involved in relational parametricity.
-
-Unfortunately, this approach cannot possibly be employed
-to prove a deep induction rule for GADTs,
-as these types do not allow for a functorial interpretation,
-at least in a parametric model~\cite{jgj21}.
-
-Nevertheless, this paper shows how to extend deep induction to some GADTs.
-We will first demonstrate how to derive the deep induction rule for some example cases,
-and then provide a general principle that works for generic GADTs,
-as long as these do not feature nesting in their definition.
-
-\subsection{(Deep) induction for $\mathsf{Equal}$}\label{sec:ind-equal}
-
-%\begin{itemize}
-%\item Define $\mathsf{Equal}^{\wedge}$
-%\[
-%\mathsf{Equal^{\wedge} : \forall (a : Set) (b : Set) \to (a \to Set) \to (b \to Set) \to Equal\,a\,b \to Set}
-%\]
-%is defined by
-%\[
-%\mathsf{Equal^{\wedge}\,a\,a\,Q\,Q'\,Refl = \forall (x : a) \to Equal\,(Q\,x)(Q'\,x)}
-%\]
-%\item (Deep) induction rule for $\mathsf{Equal}$
-%\item Prove deep induction for $\mathsf{Equal}$ (define term)
-%\end{itemize}
-
-As a first example, we derive the induction rule for the $\mathsf{Equal}$ type from example~\ref{example:equal}.
-This will provide a simple case study that will inform the investigation of more complex GADTs.
-Moreover, since we define GADTs using the $\mathsf{Equal}$ type,
-as for example in Equation~\ref{eq:eq-seq},
-this example will be instrumental in stating and deriving the induction rule of other GADTs.
-
-To define an induction rule for a type $\mathsf{G}$ we first need a predicate-lifting operation
-which takes predicates on a type $\mathsf{A}$ and lifts them to predicates on $\mathsf{G\,A}$.
-the predicate-lifting function for $\mathsf{Equal}$ is the function
-\[
-\mathsf{Equal^{\wedge} : \forall (A\,B : Set) \to (A \to Set) \to (B \to Set) \to Equal\,A\,B \to Set}
-\]
-defined as
-\[
-\mathsf{Equal^{\wedge}\,A\,A\,Q\,Q'\,Refl = \forall (a : A) \to Equal\,(Q\,a)(Q'\,a)}
-\]
-i.e., the function that takes two predicates on the same type and tests them for extensional equality.
-
-Next, we need to associate each constructor of the GADT under consideration
-to the expression that a given predicate is preserved by such constructor.
-Let $\mathsf{dIndRefl}$ be the following function associated to the $\mathsf{refl}$ constructor:
-\[
-\begin{array}{l}
-\mathsf{\lambda (P : \forall (A\,B : Set) \to (A \to Set) \to (B \to Set) \to Equal\,A\,B \to Set)} \\
-\quad\mathsf{\to \forall (C : Set) (Q\, Q' : C \to Set) \to Equal^{\wedge}\,C\,C\,Q\,Q'\,refl \to P\,C\,C\,Q\,Q'\,refl}
-\end{array}
-\]
-
-The induction rule states that,
-if a predicate is preserved by all of the constructors of the GADT under consideration,
-then the predicate is satisfied by any element of the GADT.
-The induction rule for $\mathsf{Equal}$ is thus the type
-\[
-\begin{array}{l}
-\mathsf{\forall (P : \forall (A\,B : Set) \to (A \to Set) \to (B \to Set) \to Equal\,A\,B \to Set)} \\
-\quad \mathsf{\to dIndRefl\,P \to \forall (A\,B : Set) (Q_A : A \to Set) (Q_B : B \to Set) (e: Equal\,A\,B)
-  \to Equal^{\wedge}\,A\,B\,Q_A\,Q_B\,e \to P\,A\,B\,Q_A\,Q_B\,e}
-\end{array}
-\]
-To validate the induction rule we need to provide it with a witness, i.e.,
-we need to show that the associated type is inhabited.
-We thus need to define a term $\mathsf{dIndEqual}$ of the above type,
-i.e., we need to define
-\[
-\mathsf{dIndEqual\,P\,crefl\,A\,B\,Q_A\,Q_B\,e\,liftE : P\,A\,B\,Q_A\,Q_B\,e}
-\]
-for $\mathsf{crefl : dIndRefl\,P}$,
-$\mathsf{Q_A : A \to Set}$, $\mathsf{Q_B : B \to Set}$,
-$\mathsf{e : Equal\,A\,B}$
-and $\mathsf{liftE : Equal^{\wedge}\,A\,B\,Q_A\,Q_B\,e}$.
-By pattern-matching, we only need to consider the case where $\mathsf{A = B}$ and $\mathsf{e = refl}$,
-and define
-\[
-\mathsf{dIndEqual\,P\,crefl\,A\,A\,Q_A\,Q_A'\,refl\,liftE = crefl\,A\,Q_A\,Q_A'\,liftE}
-\]
-%where
-%$\mathsf{Q_A : A \to Set}$, $\mathsf{Q_A' : A \to Set}$
-%and $\mathsf{L_E : Equal^{\wedge}\,A\,A\,Q_A\,Q_A'\,Refl}$.
-Having provided a well-defined term for it,
-we have shown that the induction rule for $\mathsf{Equal}$ is sound.
-
-The type $\mathsf{Equal}$ also has a standard structural induction rule $\mathsf{indEqual}$,
-\[
-\mathsf{\forall (Q : \forall (A\,B : Set) \to Equal\,A\,B \to Set)
-  \to \big( \forall (C : Set) \to P\,C\,C\,refl \big)
-  \to \forall (A\,B : Set) (e: Equal\,A\,B) \to P\,A\,B\,e}
-\]
-As is the case for ADTs and nested types,
-the structural induction rule for $\mathsf{Equal}$ is a consequence of the deep induction rule.
-Indeed, we can define $\mathsf{indEqual}$ as
-\[
-\mathsf{indEqual\,Q\,srefl\,A\,B\,e = dIndEqual\,P\,srefl\,A\,B\,K^A_\top\,K^B_\top\,e\,liftE}
-\]
-where
-$\mathsf{Q : \forall (A\,B : Set) \to Equal\,A\,B \to Set}$,
-$\mathsf{srefl: \forall (C : Set) \to P\,C\,C\,refl}$
-and $\mathsf{e : Equal\,A\,B}$, and
+We will revisit these GADTs in Sections~\ref{sec:deep-ind-GADTs}
+and~\ref{sec:app}.
+
+\section{(Deep) induction for GADTs}\label{sec:deep-ind-GADTs}
+
+{\color{red} Combine and/or put in figures?}
+
+The equality constraints engendered by GADTs' data constructors makes
+deriving (deep) induction rules for then more involved than for ADTs
+and other nested types. Nevertheless, we show in this section how to
+do so. We first illustrate the key components of our approach by
+deriving deep induction rules for the three specific GADTs introduced
+in Section~\ref{sec:GADTs}. Then, in Section~\ref{sec:framework}, we
+abstract these to a general framework that can be applied to any deep
+GADT that is not truly nested. {\color{red} directly or indirectly ---
+  what should we call these?} As expected, the derivation for the
+equality GADT plays a central role in that for more general ones.
+%instrumental in stating and deriving the induction rules for other
+%GADTs.
+
+\subsection{(Deep) induction for equality}\label{sec:ind-equal}
+
+To define the deep induction rule for any deep GADT $\mathsf{G}$ we
+first need a predicate lifting that lifts any predicate on a type
+$\mathsf{A}$ and to a predicate on $\mathsf{G\,A}$. The predicate
+lifting 
+\[\mathsf{Equal^{\wedge} : \forall (A\,B : Set) \to (A \to Set) \to (B
+  \to Set) \to Equal\,A\,B \to Set}\] 
+for $\mathsf{Equal}$ is defined by
+\[\mathsf{Equal^{\wedge}\,A\,A\,Q\,Q'\,refl = \forall (a : A) \to
+  Equal\,(Q\,a)(Q'\,a)}\]
+As expected, it takes two predicates on the same type as input and
+tests them for extensional equality.
+
+Next, we need to associate with each data constructor $\mathsf{c}$ of
+$\mathsf{G}$ an induction hypothesis asserting that, if the custom
+predicate arguments to a predicate $\mathsf{P}$ on $\mathsf{G}$ can be
+lifted to $\mathsf{G}$ itself, then $\mathsf{c}$ {\color{red}
+  preserves} $\mathsf{P}$. The following induction hypothesis
+$\mathsf{dIndRefl}$ is thus associated with the $\mathsf{refl}$
+constructor for $\mathsf{Equal}$:
+\begin{equation*}\label{eq:ind-refl}
+\begin{array}{l}
+\mathsf{\lambda (P : \forall (A\,B : Set) \to (A \to Set) \to (B \to
+  Set) \to Equal\,A\,B \to Set)} \\ 
+\quad\mathsf{\to \forall (C : Set) (Q\, Q' : C \to Set) \to
+  Equal^{\wedge}\,C\,C\,Q\,Q'\,refl \to P\,C\,C\,Q\,Q'\,refl} 
+\end{array}
+\end{equation*}
+The deep induction rule for $\mathsf{G}$ now states that, if a
+predicate $\mathsf{P}$ is {\color{red} preserved} by all of
+$\mathsf{G}$'s constructors, then $\mathsf{P}$ is satisfied by every
+element of $\mathsf{G}$ to which the custom predicate arguments to
+$\mathsf{P}$ can be successfully lifted.  The deep induction rule for
+the GADT $\mathsf{Equal}$ is thus
+\begin{equation}\label{eq:ind-equal}
+\begin{array}{l}
+\mathsf{\forall (P : \forall (A\,B : Set) \to (A \to Set) \to (B \to
+  Set) \to Equal\,A\,B \to Set) \to dIndRefl\,P \to}\\ \quad 
+  \mathsf{\forall (A\,B : Set) (Q_A : A \to Set) (Q_B : B \to Set) (e:
+  Equal\,A\,B) \to Equal^{\wedge}\,A\,B\,Q_A\,Q_B\,e \to
+  P\,A\,B\,Q_A\,Q_B\,e}
+\end{array}
+\end{equation}
+
+To prove that the deep induction rule for $\mathsf{Equal}$ is sound we
+must provide a witness inhabiting the type in~\eqref{eq:ind-equal}.
+We first note that, by {\color{red} pattern matching}, we only need to
+consider the case where $\mathsf{A} = \mathsf{B}$ and $\mathsf{e} =
+\mathsf{refl}$.  We can therefore define a term $\mathsf{dIndEqual}$
+of the type in \eqref{eq:ind-equal} by
+\[\mathsf{dIndEqual\;P\;crefl\;A\;A\;Q_A\;Q_A'\;refl\;liftE =
+  crefl\;A\;Q_A\;Q_A'\,liftE}\] 
+We can also recover the structural induction rule
+\begin{equation}\label{eq:sind-equal}
+\mathsf{\forall (Q : \forall (A\,B : Set)
+  \to Equal\,A\,B \to Set) \to \big( \forall (C : Set) \to
+  P\,C\,C\,refl \big) \to \forall (A\,B : Set) (e: Equal\,A\,B) \to
+  P\,A\,B\,e}
+\end{equation}
+for $\mathsf{Equal}$ by defining a term $\mathsf{indEqual}$ of the
+type in~\eqref{eq:sind-equal} by
+\[\mathsf{indEqual\;Q\;srefl\;A\;B\;e =
+  dIndEqual\;P\;srefl\;A\;B\;K^A_\top\;K^B_\top\;e\,liftE}\] Here,
+$\mathsf{e : Equal\,A\,B}$,
+$\mathsf{P : \forall (A\,B : Set) \to (A \to Set) \to (B \to
+  Set) \to Equal\,A\,B \to Set}$ is defined by
+  $\mathsf{P\;A\;B\;Q_A\;Q_B\;e =}$ $\mathsf{Q\;A\;B\;e}$,
+$\mathsf{K^A_\top}$ and $\mathsf{K^B_\top}$ are the constantly
+  {\color{red} $\mathsf{\top}$}-valued predicates on $\mathsf{A}$ and
+  $\mathsf{B}$, respectively, and
+$\mathsf{liftE :}$ $\mathsf{Equal^{\wedge}\;A\;B\;K^A_\top\;K^B_\top\;e}$
+  is fully defined (by pattern matching) by $\mathsf{liftE\,a = refl :
+    Equal\,A\,A}$ for every $\mathsf{a : A}$.
+
+{\color{red} Although the structural induction rule for any GADT
+  $\mathsf{G}$ can similarly be recovered from its deep induction rule
+  by instantiating every custom predicate by the appropriate
+  constantly {\color{red}$\mathsf{\top}$}-valued predicate, we will
+  not explicitly note this for each of the examples below.}
+
+\subsection{(Deep) induction for sequences}\label{sec:ind-seq}
+
+To derive the deep induction rule for the GADT $\mathsf{Seq}$ we use
+its Henry Ford encoding from~\eqref{eq:eq_seq}.  As usual, the first
+step is to define the predicate lifting $\mathsf{Seq^\wedge : \forall
+  (A : Set) \to (A \to Set) \to Seq\,A \to Set}$. It is defined by
+\[\mathsf{Seq^{\wedge}\,A\,Q_A\,(const\,a) = Q_A\,a}\]
+for $\mathsf{a : A}$, and
+\[\begin{array}{l}
+\mathsf{Seq^{\wedge}\,A\,Q_A\,(sPair\,B\,C\,e\,s_B\,s_C)}
+\\ \quad\mathsf{= \exists [Q_B] \exists [Q_C]\, Equal^{\wedge}\,A\, (B
+  \times C)\, Q_A\, (Q_B \times Q_C) \, e \times
+  Seq^{\wedge}\,B\,Q_B\,s_B \times Seq^{\wedge}\,C\,Q_C\,s_C }
+\end{array}\]
+for $\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C : C \to Set}$, $\mathsf{e
+  : Equal\,A\,(B \times C)}$, $\mathsf{s_B : Seq\,B}$, $\mathsf{s_C :
+  Seq\,C}$, and $\mathsf{\exists [x]\, F \,x}$ is syntactic sugar for the
+type of dependent pairs $\mathsf{(x,b)}$ where $\mathsf{x : A}$ and
+$\mathsf{b : F\, x}$ and $\mathsf{F : A \to Set}$.
+
+Next, let $\mathsf{dIndConst}$ be the induction hypothesis
+\[\mathsf{
+\lambda (P : \forall (A : Set) \to (A \to Set) \to Seq\,A \to Set) \to
+\forall (A : Set) (Q_A : A \to Set) (a : A) \to Q_A\,x \to
+P\,A\,Q_A\,(const\,a)}\] associated with the constructor
+$\mathsf{const}$, and let $\mathsf{dIndPair}$ be the induction
+hypothesis
+\[\begin{array}{l}
+\mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to Seq\,A \to
+  Set)} \to \\ 
+\quad \mathsf{\forall (A\,B\,C : Set) (Q_A : A \to Set) (Q_B : B
+  \to Set) (Q_C : C \to Set)(s_B : Seq\,B) (s_C : Seq\,C) (e :
+  Equal\,A\,(B \times C)) \to} \\ 
+\quad \mathsf{Equal^{\wedge} A\, (B \times C)\, Q_A\,
+  (Pair^{\wedge}\,B\,C\,Q_B\,Q_C)\, e \to P\,B\,Q_B\,s_B \to
+  P\,C\,Q_C\,s_C \to P A Q_A ( pair\,B\,C\,e\,s_B\,s_C )}
+\end{array}\]
+associated with the constructor $\mathsf{pair}$. The deep induction
+rule for $\mathsf{Seq}$ is thus
+\begin{equation}\label{eq:ind-seq}
+\begin{array}{l}
+\mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to Seq\,A \to
+  Set)} \mathsf{\to dIndConst\,P \to dIndPair\,P \to} \\ \quad
+\mathsf{\forall (A : Set)(Q_A : A \to Set)(s_A : Seq\,A) \to
+  Seq^{\wedge}\,A\,Q_A\,s_A \to P\,A\,Q_A\,s_A}
+\end{array}
+\end{equation}
+
+To prove that the deep induction rule for $\mathsf{Seq}$ is sound we
+must provide a witness $\mathsf{dIndSeq}$ inhabiting the type
+in~\eqref{eq:ind-seq}. We proceed by pattern matching on
+$\mathsf{s_A}$.
 \begin{itemize}
-\item $\mathsf{P : \forall (A\,B : Set) \to (A \to Set) \to (B \to Set) \to Equal\,A\,B \to Set}$ is defined as
-$\mathsf{P\,A\,B\,Q_A\,Q_B\,e = Q\,A\,B\,e}$;
-\item $\mathsf{K^A_\top}$ and $\mathsf{K^B_\top}$ are the constantly $\mathsf{\top}$-valued predicates on, respectively, $\mathsf{A}$ and $\mathsf{B}$;
-%\item $\mathsf{K^b_1}$ is the constantly $\mathsf{1}$-valued predicate on $\mathsf{b}$;
-\item $\mathsf{liftE : Equal^{\wedge}\,A\,B\,K^A_\top\,K^B_\top\,e}$ is defined by pattern matching,
-i.e., in case $\mathsf{A = B}$ and $\mathsf{e = refl}$,
-it is defined as $\mathsf{liftE\,a = refl : Equal\,A\,A}$ for every $\mathsf{a : A}$.
+\item If $\mathsf{s_A = const\,a}$ for $\mathsf{a : A}$, then we
+  define \[\mathsf{dIndSeq\;P\;cconst\;cpair\;A\;Q_A\;(const\,a)\;liftA
+    = cconst\;A\;Q_A\;a\;liftA}\] Since
+  $\mathsf{Seq^{\wedge}\,A\,Q_A\,(const\,a)}$ $\mathsf{= Q_A\,a}$, and
+  thus $\mathsf{liftA : Q_A\,a}$, the definition does indeed
+  type-check.
+\item If $\mathsf{s_A = sPair\,B\,C\,e\,s_B\,s_C}$ for $\mathsf{e :
+  Equal\,A\,(B \times C)}$, $\mathsf{s_B : Seq\,B}$, and $\mathsf{s_C
+  : Seq\,C}$, then we define
+  \[\mathsf{dIndSeq\,P\,cconst\,cpair\,A\,Q_A\,(sPair\,B\,C\,e\,s_B\,s_C)\,
+    (Q_B,Q_C, liftE, liftB, liftC) =
+    cpair\,A\,B\,C\,Q_A\,Q_B\,Q_C\,s_B\,s_C\,e\,liftE\,p_B\,p_C}\]
+  Here, $\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C : C \to Set}$,
+  $\mathsf{liftE : Equal^{\wedge}\,A\, (B \times C)\, Q_A\, (Q_B
+    \times Q_C) \, e}$, $\mathsf{liftB : Seq^{\wedge}\,B\,Q_B\,s_B}$,
+  and $\mathsf{liftC :}$\\
+\noindent
+$\mathsf{Seq^{\wedge}\,C\,Q_C\,s_C}$ --- which together ensure that
+  $\mathsf{(Q_B, Q_C, liftE, liftB, liftC) :
+    Seq^{\wedge}\,A\,Q\,(sPair\,B\,C\,e\,s_B\,s_C)}$ --- and\\
+%  $\mathsf{p_B}$ and $\mathsf{p_C}$\\ are defined by
+$\mathsf{p_B}
+  =\mathsf{dIndSeq\,P\,cconst\,cpair\,B\,Q_B\,s_B\,liftB :
+    P\,B\,Q_B\,s_B}$ and $\mathsf{p_C}
+  =\mathsf{dIndSeq\,P\,cconst\,cpair\,C\,Q_C\,s_C\,liftC :
+    P\,C\,Q_C\,s_C}$.
 \end{itemize}
-That the structural induction rule is a consequence of the deep induction one is also true for all the examples below,
-even though we will not remark it every time.
-
-\subsection{(Deep) induction for $\mathsf{Seq}$}
-
-Next, we shall provide an induction rule for the $\mathsf{Seq}$ type defined in Example~\ref{example:seq}.
-Again, the first step in deriving the induction rule for $\mathsf{Seq}$ consists in defining the predicate-lifting function over it, 
-\[
-\mathsf{Seq^{\wedge} : \forall (A : Set) \to (A \to Set) \to Seq\,A \to Set}
-\]
-which is given by pattern-matching as
-\[
-\mathsf{Seq^{\wedge}\,A\,Q_A\,(const\,a) = Q_A\,a}
-\]
-where $\mathsf{Q_A : A \to Set}$ and $\mathsf{a : A}$,
-and
-\[
-\begin{array}{l}
-\mathsf{Seq^{\wedge}\,A\,Q_A\,(sPair\,B\,C\,e\,s_B\,s_C)} \\
-\quad\mathsf{=
-\sum [Q_B : B \to Set] [Q_C : C \to Set]\;
-Equal^{\wedge}\,A\, (B \times C)\, Q_A\, (Q_B \times Q_C) \, e \times Seq^{\wedge}\,B\,Q_B\,s_B \times Seq^{\wedge}\,C\,Q_C\,s_C
-}
-\end{array}
-\]
-where $\mathsf{e : Equal\,A\,(B \times C)}$,
-$\mathsf{s_B : Seq\,B}$ and $\mathsf{s_C : Seq\,C}$.
-%We also need to define the lifting of predicates over the polymorphic type of pairs,
-%$\mathsf{Pair = \forall (B\,C : Set) \to B \times C}$, which is 
-%\[
-%\mathsf{Pair^{\wedge} : \forall (b\,c : Set) \to (b \to Set) \to (c \to Set) \to Pair\,b\,c \to Set}
-%\]
-%and it is defined as
-%\[
-%\mathsf{Pair^{\wedge}\,b\,c\,Q_b\,Q_c\,(y, z) = Q_b\,y \times Q_c\,z}
-%\]
-%where $\mathsf{Q_b : b \to Set}$, $\mathsf{Q_c : c \to Set}$, $\mathsf{y : b}$ and $\mathsf{z : c}$.
-
-Finally, let $\mathsf{dIndConst}$ be the function
-\[
-\mathsf{
-\lambda (P : \forall (A : Set) \to (A \to Set) \to Seq\,A \to Set)
-\to \forall (A : Set) (Q_A : A \to Set) (a : A)
-\to Q_A\,x \to P\,A\,Q_A\,(Const\,a)
-}
-\]
-associated to the $\mathsf{const}$ constructor,
-and let $\mathsf{dIndSPair}$ be the function
-\[
-\begin{array}{l}
-\mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to Seq\,A \to Set)} \\
-\quad \mathsf{\to \forall (A\,B\,C : Set) (Q_A : A \to Set) (Q_B : B \to Set) (Q_C : C \to Set)(s_B : Seq\,B) (s_C : Seq\,C) (e : Equal\,A\,(B \times C))} \\
-\quad \mathsf{\to Equal^{\wedge} A\, (B \times C)\, Q_A\, (Pair^{\wedge}\,B\,C\,Q_B\,Q_C)\, e
-	\to P\,B\,Q_B\,s_B \to P\,C\,Q_C\,s_C
-	\to P A Q_A ( SPair\,B\,C\,e\,s_B\,s_C )}
-\end{array}
-\]
-associated to the $\mathsf{sPair}$ constructor.
-
-With these tools we can formulate an induction rule for $\mathsf{Seq}$,
-\[
-\begin{array}{l}
-\mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to Seq\,A \to Set)}
-\mathsf{\to CConst\,P \to CSPair\,P} \\
-\quad \mathsf{\to \forall (A : Set)(Q_A : A \to Set)(s_A : Seq\,A) \to Seq^{\wedge}\,A\,Q_A\,s_A \to P\,A\,Q_A\,s_A}
-\end{array}
-\]
-To validate the induction rule, we define a term $\mathsf{dIndSeq}$ for the above type.
-We have to define
-\[
-\mathsf{dIndSeq\,P\,cconst\,cspair\,A\,Q_A\,s_A\,liftA : P\,A\,Q_A\,s_A}
-\]
-where
-$\mathsf{P : \forall (A : Set) \to (A \to Set) \to Seq\,A \to Set}$,
-$\mathsf{cconst : cConst\,P}$,
-$\mathsf{cspair : cSPair\,P}$,
-$\mathsf{Q_A : A \to Set}$, $\mathsf{s_A : Seq\,A}$
-and $\mathsf{liftA : Seq^{\wedge}\,A\,Q\,s_A}$,
-and we proceed by pattern-matching on $\mathsf{s_A}$.
-Let $\mathsf{s_A = const\,a}$ for $\mathsf{a : A}$, and define
-\[
-\mathsf{dIndSeq\,P\,cconst\,cspair\,A\,Q_A\,(const\,a)\,liftA = cconst\,A\,Q_A\,a\,liftA}
-\]
-Notice that $\mathsf{Seq^{\wedge}\,A\,Q_A\,(const\,a) = Q_A\,a}$, and thus $\mathsf{liftA : Q_A\,a}$,
-making the right-hand-side in the above expression type-check.
-Now, let $\mathsf{s_A = sPair\,B\,C\,e\,s_B\,s_C}$
-for $\mathsf{e : Equal\,A\,(B \times C)}$, $\mathsf{s_B : Seq\,B}$ and $\mathsf{s_C : Seq\,C}$,
-and define
-\[
-\mathsf{dIndSeq\,P\,cconst\,cspair\,A\,Q_A\,(sPair\,B\,C\,e\,s_B\,s_C)\,(Q_B, Q_C, liftE, liftB, liftC)
-	= cspair\,A\,B\,C\,Q_A\,Q_B\,Q_C\,s_B\,s_C\,e\,liftE\,p_B\,p_C}
-\]
-where $\mathsf{(Q_B, Q_C, liftE, liftB, liftC) : Seq^{\wedge}\,A\,Q\,(sPair\,B\,C\,e\,s_B\,s_C)}$, i.e.,
-\begin{itemize}
-\item $\mathsf{Q_B : B \to Set}$ and $\mathsf{Q_C : C \to Set}$;
-\item $\mathsf{liftE : Equal^{\wedge}\,A\, (B \times C)\, Q_A\, (Q_B \times Q_C) \, e}$;
-\item $\mathsf{liftB : Seq^{\wedge}\,B\,Q_B\,s_B}$ and $\mathsf{liftC : Seq^{\wedge}\,C\,Q_C\,s_C}$;
-\end{itemize}
-and $\mathsf{p_B}$ and $\mathsf{p_C}$ are defined as follows:
-\begin{align*}
-\mathsf{p_B} &=\mathsf{DISeq\,P\,cconst\,cspair\,B\,Q_B\,s_B\,liftB : P\,B\,Q_B\,s_B} \\
-\mathsf{p_C} &=\mathsf{DISeq\,P\,cconst\,cspair\,C\,Q_C\,s_C\,liftC : P\,C\,Q_C\,s_C}
-\end{align*}
-
-\subsection{(Deep) induction for $\mathsf{LTerm}$ }
-
-%Same structure as previous two.
-%Runs through the paper.
-%Still to decide?
-
-%{\color{blue} Some notational conventions: I am using uppercase letters for elements of LType and 
-%lowercase letters for elements of LTerm. For example, $\mathsf{(T_b : LType \, b)}$ or $\mathsf{(t_c : LTerm \,c)}$.}
-
-We now present an induction rule for the $\mathsf{LTerm}$ type defined in Example~\ref{example:lterm}.
-For that, we need liftings for both $\mathsf{LType}$ and
-$\mathsf{LTerm}$. We will also need a lifting of predicates for arrow types, since arrow types appear in 
-$\mathsf{LType}$ and $\mathsf{LTerm}$. The lifting for arrow types 
-\[
-  \mathsf{Arr^{\wedge} : \forall (A\, B : Set) \to (A \to Set) \to (B \to Set) \to (A \to B) \to Set}
-\]
-is defined as
-\[
-  \mathsf{Arr^{\wedge}\, A\, B\, Q_A\, Q_B\, f = \forall (a : A) \to Q_A\,a \to Q_B\, (f\,a)}
-\]
-Now we can define the lifting for $\mathsf{LType}$, which has type 
-\[
-\mathsf{LType^{\wedge} : \forall (A : Set) \to (A \to Set) \to LType\,A \to Set}
-\]
-and is defined by pattern-matching as
-\[
-  \mathsf{LType^{\wedge}\,A\,Q_A\,(tBool\,B\,e)
-  = \sum [Q_B : B \to Set]\, Equal^{\wedge}\, A\, B\, Q_A\, K^{Bool}_{\top} \,e}
-\]
-where $\mathsf{Q_A : A \to Set}$,
-$\mathsf{K^{Bool}_{\top}}$ is the constantly $\mathsf{\top}$-valued predicate on $\mathsf{Bool}$
-and $\mathsf{e : Equal\,A\,Bool}$;
-\[
-\begin{array}{l}
-\mathsf{LType^{\wedge}\,A\,Q_A\,(tArr\, B\, C\, e\, T_B\, T_C)} \\
-\quad
-\mathsf{= \sum [Q_B : B \to Set] [Q_C : C \to Set]\,
-  Equal^{\wedge}\,A\, (B \to C)\, Q_A\, (Arr^{\wedge} \, B\, C\, Q_B \, Q_C) \, e
-  \times \, LType^{\wedge}\,B\,Q_B\,T_B \times LType^{\wedge}\,C\,Q_C\,T_C
-}
-\end{array}
-\]
-where $\mathsf{Q_A : A \to Set}$, $\mathsf{e : Equal\, A\, (B \to C)}$, $\mathsf{T_B : LType \,B}$ and 
-$\mathsf{T_C : LType \,C}$;
-and 
-\[
-\mathsf{LType^{\wedge}\,A\,Q_A\,(tList\, B\, e\, T_B)
+
+\subsection{(Deep) induction for simply typed lambda
+  terms}\label{sec:ind-lam}
+
+To derive the deep induction rule for the GADT $\mathsf{LTerm}$ we use
+its Henry Ford encoding from~\ref{eq:eq_ltype}
+and~\eqref{eq:eq_lterm}. The first step is to define the predicate
+liftings for $\mathsf{LType}$ and $\mathsf{LTerm}$. We will also need
+a predicate lifting for arrow types, since arrow types appear in
+$\mathsf{LType}$ and $\mathsf{LTerm}$.
+
+The predicate lifting $\mathsf{Arr^{\wedge} : \forall (A\, B : Set)
+  \to (A \to Set) \to (B \to Set) \to (A \to B) \to Set}$ for arrow
+types is defined by
+\[\mathsf{Arr^{\wedge}\, A\, B\, Q_A\, Q_B\, f = \forall (a : A) \to
+  Q_A\,a \to Q_B\, (f\,a)}\] The predicate lifting
+$\mathsf{LType^{\wedge} : \forall (A : Set) \to (A \to Set) \to
+  LType\,A \to Set}$ for $\mathsf{LType}$ is defined by
+pattern matching by
+\[ \mathsf{LType^{\wedge}\,A\,Q_A\,(bool\,B\,e)
+  = \exists [Q_B]\, Equal^{\wedge}\, A\, B\, Q_A\, K^{Bool}_{\top}
+  \,e}\] where $\mathsf{Q_A : A \to Set}$, $\mathsf{Q_B : B \to Set}$,
+$\mathsf{K^{Bool}_{\top}}$ is the constantly $\mathsf{\top}$-valued
+predicate on $\mathsf{Bool}$, and $\mathsf{e : Equal\,A\,Bool}$;
+\[\mathsf{LType^{\wedge}\,A\,Q_A\,(arr\, B\, C\, e\, T_B\, T_C)}
+\mathsf{= \exists [Q_B] \,\exists [Q_c]\, Equal^{\wedge}\,A\,
+  (B \to C)\, Q_A\, (Arr^{\wedge} \, B\, C\, Q_B \, Q_C) \, e \times
+  \, LType^{\wedge}\,B\,Q_B\,T_B \times LType^{\wedge}\,C\,Q_C\,T_C}\]
+where $\mathsf{Q_A : A \to Set}$, $\mathsf{Q_B : B \to Set}$,
+$\mathsf{Q_C : C \to Set}$, $\mathsf{e : Equal\, A\, (B \to C)}$,
+$\mathsf{T_B : LType \,B}$ and $\mathsf{T_C : LType \,C}$; and
+\[\mathsf{LType^{\wedge}\,A\,Q_A\,(list\, B\, e\, T_B)
 =
-\sum [Q_B : B \to Set]\, 
+\exists [Q_B]\,
   Equal^{\wedge}\,A\, (List\, B)\, Q_A\, (List^{\wedge} \, B\, Q_B) \, e 
@@
 \]
-where $\mathsf{Q_A : A \to Set}$, $\mathsf{e : Equal\, A\, (List\, B)}$, and $\mathsf{T_B : LType \,B}$.
-%Given the lifting for $\mathsf{LType}$, 
-We can define the lifting for $\mathsf{LTerm}$. Again, 
-the lifting 
-\[
-\mathsf{LTerm^{\wedge} : \forall (A : Set) \to (A \to Set) \to LTerm\,A \to Set}
-\]
-is defined by pattern-matching as
-\[
-  \mathsf{LTerm^{\wedge}\,A\,Q_A\,(var\,s\,T_A) = LType^{\wedge}\, A\, Q_A\, T_A}
-\]
-where $\mathsf{s : String}$ and $\mathsf{T_A : LType\, A}$;
-\[
-\begin{array}{l}
-\mathsf{LTerm^{\wedge}\,A\,Q_A\, (abs \,B \,C \,e \,s \,T_B \,t_C)} \\ 
-\quad \mathsf{
-  = \sum [Q_B : B \to Set] [Q_C : C \to Set]\,
-      Equal^{\wedge} \, A\, (B \to C)\, Q_A\, (Arr^{\wedge} \, B\, C\, Q_B \, Q_C)\, e
-      \times \, LType^{\wedge}\, B\, Q_B\, T_B
-      \times \, LTerm^{\wedge}\, C\, Q_C\, t_C }
-\end{array}
-\]
-where $\mathsf{e : Equal \, A \, (B \to C)}$,
-$\mathsf{s : String}$,
-$\mathsf{T_B : LType\,B}$,
-and $\mathsf{t_C : LTerm \, C}$;
-\[
-\mathsf{LTerm^{\wedge}\,A\,Q_A\, (app\, B\, t_{BA}\, t_B)
-=
-\sum [Q_B : B \to Set]\, 
-  LTerm^{\wedge}\, (B \to A)\, (Arr^{\wedge} \, B\, A\, Q_B \, Q_A)\, t_{BA}
-  \times LTerm^{\wedge}\, B\, Q_B\, t_B}
-\]
-where $\mathsf{t_{BA} : LTerm \, (B \to A)}$ and $\mathsf{t_B : LTerm \, B}$; and
-\[
-\mathsf{LTerm^{\wedge}\,A\,Q_A\, (listC\, B\, e\, ts)
-=
-\sum [Q_B : B \to Set]\,
- Equal^{\wedge} \, A\, (List\,B)\, Q_A\, (List^{\wedge} \, B\, Q_B) \, e
- \times List^{\wedge}\, (LTerm\,B) \, (LTerm^{\wedge} \, B\, Q_B) \, ts}
-\]
-where $\mathsf{e : Equal\, A\, (List \,B)}$ and $\mathsf{ts : List\, (LTerm B)}$. 
-Notice that we use the lifting for $\mathsf{List}$ in the case for the $\mathsf{listC}$ constructor.
-
-With these liftings, we can define the deep induction principle for $\mathsf{LTerm}$.
+where $\mathsf{Q_A : A \to Set}$, $\mathsf{Q_B : B \to Set}$,
+$\mathsf{e : Equal\, A\, (List\, B)}$, and $\mathsf{T_B : LType \,B}$.
+The predicate lifting $\mathsf{LTerm^{\wedge} :}$ $\mathsf{\forall (A
+  : Set) \to (A \to Set) \to LTerm\,A \to Set}$ for $\mathsf{LTerm}$
+is defined by pattern matching by
+\[\mathsf{LTerm^{\wedge}\,A\,Q_A\,(var\,s\,T_A) = LType^{\wedge}\, A\,
+  Q_A\, T_A}\] where $\mathsf{s : String}$ and $\mathsf{T_A : LType\, A}$;
+\[\mathsf{LTerm^{\wedge}\,A\,Q_A\, (abs \,B \,C \,e \,s \,T_B \,t_C)}
+\mathsf{ = \exists [Q_B]\,\exists [Q_C]\, Equal^{\wedge} \, A\, (B \to
+  C)\, Q_A\, (Arr^{\wedge} \, B\, C\, Q_B \, Q_C)\, e \times \,
+  LType^{\wedge}\, B\, Q_B\, T_B \times \, LTerm^{\wedge}\, C\, Q_C\,
+  t_C }\] where $\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C : C \to
+  Set}$, $\mathsf{e : Equal \, A \, (B \to C)}$, $\mathsf{s :
+  String}$, $\mathsf{T_B : LType\,B}$, and $\mathsf{t_C : LTerm \,
+  C}$;
+\[\mathsf{LTerm^{\wedge}\,A\,Q_A\, (app\, B\, t_{BA}\, t_B)
+= \exists [Q_B]\, LTerm^{\wedge}\, (B \to A)\, (Arr^{\wedge} \, B\,
+A\, Q_B \, Q_A)\, t_{BA} \times LTerm^{\wedge}\, B\, Q_B\, t_B}\]
+where $\mathsf{Q_B : B \to Set}$, $\mathsf{t_{BA} : LTerm \, (B \to
+  A)}$, and $\mathsf{t_B : LTerm \, B}$; and
+\[\mathsf{LTerm^{\wedge}\,A\,Q_A\, (list\, B\, e\, ts) =
+\exists [Q_B]\, Equal^{\wedge} \, A\, (List\,B)\, Q_A\, (List^{\wedge}
+\, B\, Q_B) \, e \times List^{\wedge}\, (LTerm\,B) \, (LTerm^{\wedge}
+\, B\, Q_B) \, ts} 
+\]
+where $\mathsf{Q_B : B \to Set}$, $\mathsf{e : Equal\, A\, (List
+  \,B)}$, $\mathsf{ts : List\, (LTerm B)}$, and $\mathsf{List^\wedge}$
+is the predicate lifting for lists from~\eqref{eq:rose}.
+
+With these liftings in hand, we can define the deep induction
+principle for $\mathsf{LTerm}$.
 % Var case 
-Let $\mathsf{dIndVar}$ be the function 
+Let $\mathsf{dIndVar}$ be the induction hypothesis
 \begin{multline*}
-\mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A \to Set)} \\
-\mathsf{\to 
+\mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A
+  \to Set)} \\ \mathsf{\to 
   \forall (A : Set) (Q_A : A \to Set) (s : String) (T_A : LType\, A) 
   \to LType^{\wedge} \, A\, Q_A\, T_A
-  \to P \, A\, Q_A\, (var \, s\, T_A)
+  \to P \, A\, Q_A\, (var \; s\, T_A)
   }
 \end{multline*}
-associated to the $\mathsf{var}$ constructor. 
+associated with the constructor $\mathsf{var}$,
 % Abs case 
-Let $\mathsf{dIndAbs}$ be the function 
+let $\mathsf{dIndAbs}$ be the induction hypothesis
 \begin{align*}
-  &\mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A \to Set)} \\
+  &\mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A
+    \to Set)} \\ 
   &\quad\mathsf{\to 
-  \forall (A\,B\,C: Set) (Q_A : A \to Set)  (Q_B : B \to Set) (Q_C : C \to Set)
-  (e : Equal\, A\, (B \to C)) (s : String) } \\ 
+  \forall (A\,B\,C: Set) (Q_A : A \to Set)  (Q_B : B \to Set) (Q_C : C
+  \to Set) (e : Equal\, A\, (B \to C)) (s : String) } \\ 
   &\quad\mathsf{ \to (T_B : LType\, B) \to (t_C : LTerm\, C)
-  \to Equal^{\wedge}\,A\,(B \to C)\, Q_A \, (Arr^{\wedge} \, B\, C\, Q_B \, Q_C) \, e
-  } \\
+  \to Equal^{\wedge}\,A\,(B \to C)\, Q_A \, (Arr^{\wedge} \, B\, C\,
+  Q_B \, Q_C) \, e } \\
   &\quad\mathsf{
@@
 \end{align*}
-associated to the $\mathsf{abs}$ constructor. 
+associated with the constructor $\mathsf{abs}$, let
 % App case 
-Let $\mathsf{dIndApp}$ be the function 
+Let $\mathsf{dIndApp}$ be the induction hypothesis
 \begin{align*}
-  &\mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A \to Set)} \\
+  &\mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A
+    \to Set)} \\ 
   &\quad \mathsf{\to 
@@
 \end{align*}
-associated to the $\mathsf{app}$ constructor. 
+associated with the constructor $\mathsf{app}$, and 
 % ListC case 
-Let $\mathsf{dIndListC}$ be the function 
+let $\mathsf{dIndList}$ be the induction hypothesis
 \begin{align*}
-  &\mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A \to Set)} \\
+  &\mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A
+    \to Set)} \\ 
   &\quad \mathsf{\to 
@@
   \to List^{\wedge}\, (LTerm\,B) (P\, B\, Q_B)\, ts
-  \to P \, A\, Q_A\, (listC \,B \,e \, ts) }
+  \to P \, A\, Q_A\, (list \,B \,e \, ts) }
 \end{align*}
-associated to the $\mathsf{listC}$ constructor.
-
-The type of the deep induction principle for $\mathsf{LTerm}$ is
-\[
-\begin{array}{l}
-\mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A \to Set)
-  \to dIndVar\,P \to dIndAbs\,P \to dIndApp\,P \to dIndListC\,P} \\
-\quad \mathsf{\to \forall (A : Set)(Q_A : A \to Set)(t_A : LTerm\,A) \to LTerm^{\wedge}\,A\,Q_A\,t_A \to P\,A\,Q_A\,t_A}
-\end{array}
-\]
-To prove the induction principle, we define a term $\mathsf{dIndLTerm}$ for it.
-We have to define 
-\[ 
-\mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clistc \, A\, Q_A\, t_A\, liftA : P\, A\, Q_A\, t_A}
-\]
-where $\mathsf{cvar: dIndVar}$, $\mathsf{cabs: dIndAbs}$, $\mathsf{capp : dIndApp}$, $\mathsf{clistc : dIndListC}$,
-$\mathsf{Q_A : A \to Set}$, $\mathsf{t_A : LTerm\,A}$, and $\mathsf{liftA : LTerm^{\wedge}\,A\,Q_A\,t_A}$.
-% where $\mathsf{P : \forall (a : Set) \to (a \to Set) \to LTerm\,a \to Set}$, 
-As before, we prove the induction principle by pattern matching on $\mathsf{t_A}$. 
-For the $\mathsf{var}$ constructor case, let $\mathsf{t_A = var\, s\, T_A}$ and define
-\[
-  \mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clistc \, A\, Q_A\, (var\,s\,T_A) \, liftA = cvar \, A\, Q_A\, s\, T_A\, liftA }
-\]
-for $\mathsf{s : String}$ and $\mathsf{T_A : LType\,A}$.
-% where $\mathsf{L_a : LTerm^{\wedge}\, a\, Q_a\, (Var\,s\,T_a)}$, i.e., $\mathsf{L_a : LType^{\wedge}\,a\,Q_a\,T_a}$ 
-Notice that $\mathsf{LTerm^{\wedge}\, A\, Q_A\, (var\,s\,T_A) = LType^{\wedge}\,A\,Q_A\,T_A}$, so 
-$\mathsf{liftA : LType^{\wedge}\,A\,Q_A\,T_A}$.
-%
+associated with the constructor $\mathsf{list}$. The deep induction
+principle for $\mathsf{LTerm}$ is thus
+\begin{equation}\label{eq:ind-lam}
+\begin{array}{l}
+\mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A
+  \to Set) \to dIndVar\,P \to dIndAbs\,P \to dIndApp\,P \to
+  dIndList\,P \to} \\ \quad \mathsf{\forall (A : Set)(Q_A : A \to
+  Set)(t_A : LTerm\,A) \to LTerm^{\wedge}\,A\,Q_A\,t_A \to
+  P\,A\,Q_A\,t_A}
+\end{array}
+\end{equation}
+
+To prove that the deep induction principle for $\mathsf{LTerm}$ is
+sound, we must provide a witness $\mathsf{dIndLTerm}$ inhabiting the
+type in~\eqref{eq:ind-lam}. We proceed by pattern matching on
+$\mathsf{t_A}$.
+\begin{itemize}
+\item If $\mathsf{t_A = var\; s\, T_A}$, then we define
+\[ \mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clist \, A\, Q_A\,
+  (var\;s\,T_A) \, liftA = cvar \, A\, Q_A\, s\, T_A\, liftA }\] for
+$\mathsf{s : String}$ and $\mathsf{T_A : LType\,A}$.  Note that
+$\mathsf{LTerm^{\wedge}\, A\, Q_A\, (var\;s\,T_A) =
+  LType^{\wedge}\,A\,Q_A\,T_A}$, so $\mathsf{liftA :
+  LType^{\wedge}\,A\,Q_A\,T_A}$.
 % Abs case 
-For the $\mathsf{abs}$ case, let $\mathsf{t_A = abs \,B \,C \, e \,s \,T_B \, t_C} $ and define
-\[
-\begin{array}{l}
-\mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clistc \, A\, Q_A\, (abs \,B \,C \,e \,s \,T_B \, t_C) \, (Q_B , Q_C , liftE, lift_{T_B}, lift_{t_C})} \\
-\quad \mathsf{= cabs\,A\,B\,C\, Q_A\, Q_B\, Q_C\, e\, s\, T_B\, t_C\, liftE\, lift_{T_B}\, p_C}
-\end{array}
-\]
-where $\mathsf{(Q_B , Q_C , liftE, lift_{T_B}, lift_{t_C}) : LTerm^{\wedge}\, A\, Q_A\, (abs \,B \,C \, e \,s \,T_B \, t_C)}$, i.e.,
-\begin{itemize}
-  \item $\mathsf{Q_B : B \to Set}$;
-  \item $\mathsf{Q_C : C \to Set}$;
-  \item $\mathsf{liftE : Equal^{\wedge} \, A\, (B \to C)\, Q_A\, (Arr^{\wedge} \, B\, C\, Q_B \, Q_C) \, e}$;
-  \item $\mathsf{lift_{T_B}: LType^{\wedge} \, B\, Q_B\, T_B}$;
-  \item $\mathsf{lift_{t_C}: LTerm^{\wedge} \, C\, Q_C\, T_C}$;
-\end{itemize}
-and $\mathsf{p_C}$ is defined as:
-\[
-  \mathsf{p_C =  dIndLTerm\,P\,cvar\,cabs \,capp \,clistc\, C\, Q_C\, t_C\, lift_{t_C}  : P \, C\, Q_C \, t_C }
-\]
+\item If $\mathsf{t_A = abs \,B \,C \, e \,s \,T_B \, t_C} $, then we
+  define
+\[
+\begin{array}{l}
+\mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clist \, A\, Q_A\,
+  (abs \,B \,C \,e \,s \,T_B \, t_C) \, (Q_B , Q_C , liftE,
+  lift_{T_B}, lift_{t_C})} \\ \quad \mathsf{= cabs\,A\,B\,C\, Q_A\,
+  Q_B\, Q_C\, e\, s\, T_B\, t_C\, liftE\, lift_{T_B}\, p_C}
+\end{array}
+\]
+Here, $\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C : C \to Set}$,
+$\mathsf{liftE : Equal^{\wedge} \, A\, (B \to C)\, Q_A\, (Arr^{\wedge}
+  \, B\, C\, Q_B \, Q_C) \, e}$, $\mathsf{lift_{T_B}: LType^{\wedge}
+  \, B\, Q_B\, T_B}$, and $\mathsf{lift_{t_C}: LTerm^{\wedge} \, C\,
+  Q_C\, T_C}$ --- which together ensure that $\mathsf{(Q_B , Q_C ,
+  liftE, lift_{T_B}, lift_{t_C}) : LTerm^{\wedge}\, A\, Q_A\, (abs \,B
+  \,C \, e \,s \,T_B \, t_C)}$ --- and $\mathsf{p_C =
+  dIndLTerm\,P\,cvar\,cabs \,capp \,clist\, C\, Q_C\, t_C\,
+  lift_{t_C} : P \, C\, Q_C \, t_C }$.
 % App case - (App \,b \,t_{ba} \, t_b)
-For the $\mathsf{app}$ case, let $\mathsf{t_A = app \,B \,\,t_{BA} \, t_B} $ and define
-\[
-  \mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clistc \, A\, Q_A\,  (app \,B \,\,t_{BA} \, t_B)\, (Q_B , list_{t_{BA}}, list_{t_B})} \\
-  \mathsf{= capp\,A\,B\,Q_A\, Q_B\, t_{BA}\, t_B\, p_{BA} \, p_B}
-\]
-where $\mathsf{(Q_B , lift_{t_{BA}}, list_{t_B}) : LTerm^{\wedge}\, A\, Q_A\, (app \,B \,t_{BA} \, t_B)}$, i.e.,
-\begin{itemize}
-  \item $\mathsf{Q_B : B \to Set}$;
-  \item $\mathsf{lift_{t_{BA}}: LTerm^{\wedge} \, (B \to A)\, (Arr^{\wedge} \, B\, A\, Q_B \, Q_A)\, t_{BA}}$;
-  \item $\mathsf{lift_{t_B}: LTerm^{\wedge} \, B\, Q_B\, t_B}$;
-\end{itemize}
-and $\mathsf{p_{BA}}$ and $\mathsf{p_B}$
-and are defined as:
-\begin{align*}
-  \mathsf{p_{BA}} &= \mathsf{dIndLTerm\,P\,cvar\,cabs \,capp \,clistc\, (B \to A)\,(Arr^{\wedge} \, B\, A\, Q_B \, Q_A) \, t_{BA}\, lift_{t_{BA}}
-  : P \, (B \to A)\, (Arr^{\wedge} \, B\, A\, Q_B \, Q_A) \, t_{BA} } \\
-  \mathsf{p_B} &= \mathsf{dIndLTerm\,P\,cvar\,cabs \,capp \,clistc\, B\, Q_B\, t_B\, lift_{t_B}  : P \, B\, Q_B \, t_B } 
-\end{align*}
+\item If $\mathsf{t_A = app \,B \,\,t_{BA} \, t_B} $, then we define
+\[ \mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clist \, A\, Q_A\,
+    (app \,B \,\,t_{BA} \, t_B)\, (Q_B , list_{t_{BA}}, list_{t_B})}
+  \mathsf{ = capp\,A\,B\,Q_A\, Q_B\, t_{BA}\, t_B\, p_{BA} \, p_B}
+\]
+Here, $\mathsf{Q_B : B \to Set}$, $\mathsf{lift_{t_{BA}}:
+  LTerm^{\wedge} \, (B \to A)\, (Arr^{\wedge} \, B\, A\, Q_B \, Q_A)\,
+  t_{BA}}$, and $\mathsf{lift_{t_B}: LTerm^{\wedge} \, B\, Q_B\, t_B}$
+--- so that\\ $\mathsf{(Q_B , lift_{t_{BA}}, list_{t_B}) : 
+  LTerm^{\wedge}\, A\, Q_A\, (app \,B \,t_{BA} \, t_B)}$ --- and
+$\mathsf{p_B} = \mathsf{dIndLTerm\,P\,cvar\,cabs \,capp \,clist\, B\,
+  Q_B\, t_B\, lift_{t_B} : P \, B\, Q_B \, t_B }$ and $\mathsf{p_{BA}}
+= \mathsf{dIndLTerm\,P\,cvar\,cabs \,capp \,clist\, (B \to
+  A)\,(Arr^{\wedge} \, B\, A\, Q_B \, Q_A) \, t_{BA}\, lift_{t_{BA}} :
+  P \, (B \to A)\, (Arr^{\wedge} \, B\, A\, Q_B \, Q_A) \, t_{BA} }$.
 % ListC case - (ListC \,b \,e \, ts)
-Finally, for the $\mathsf{listC}$ case, let $\mathsf{t_A = listC \,B \,e \, ts}$ and define
+\item If let $\mathsf{t_A = list \,B \,e \, ts}$, then we define
 \begin{multline*}
-  \mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clistc \, A\, Q_A\, (listC \,B \,e \, ts) \, (Q_B , liftE, lift_{List})
+  \mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clistc \, A\, Q_A\, (list \,B \,e \, ts) \, (Q_B , liftE, lift_{List})
   = clistc \,A\,B\,Q_A\, Q_B\, e\, ts\, liftE\, p_{List} }
 \end{multline*}
-where $\mathsf{(Q_B , liftE, lift_{List}): LTerm^{\wedge}\, A\, Q_A\, (listC \,B \,e \, ts)}$, i.e.,
-\begin{itemize}
-  \item $\mathsf{Q_B : B \to Set}$;
-  \item $\mathsf{listE : Equal^{\wedge}\, A\, (List\,B)\, Q_A\, (List^{\wedge}\, B\, Q_B)\, e}$;
-  \item $\mathsf{lift_{List}: List^{\wedge} \, (LTerm\, B) \, (LTerm^{\wedge}\, B\, Q_B) \, ts}$;
+Here, $\mathsf{Q_B : B \to Set}$, $\mathsf{listE : Equal^{\wedge}\,
+  A\, (List\,B)\, Q_A\, (List^{\wedge}\, B\, Q_B)\, e}$, and
+$\mathsf{lift_{List}: List^{\wedge} \, (LTerm\, B) \,
+  (LTerm^{\wedge}\, B\, Q_B) \, ts}$ --- so that $\mathsf{(Q_B ,
+  liftE, lift_{List}): LTerm^{\wedge}\, A\, Q_A\, (list \,B \,e \,
+  ts)}$ --- and
+$\mathsf{p_{List} = liftListMap \, (LTerm\, B) \, (LTerm^{\wedge} \, B
+  \, Q_B)\, (P\,B\,Q_B)\, p_{ts} \, ts\, lift_{List} : }$\\
+$\mathsf{List^{\wedge}\,
+  (LTerm\,B) \, (P\,B\,Q_B) \, ts}$ and
+$\mathsf{ p_{ts} = dIndLTerm\, P\, cvar\, cabs\, capp\, clist\, B\,
+  Q_B : PredMap\,(LTerm\,B) \,(LTerm^{\wedge}\, B\, Q_B) \,
+  (P\,B\,Q_B)}$. Moreover, $\mathsf{PredMap : \forall\, (A : Set) \to
+  (A \to Set) \to   (A \to Set) \to Set }$ is the type constructor
+producing the type of morphisms between predicates defined by
+\[\mathsf{PredMap \,A\, Q\,Q'\, = \forall\, (a : A) \to Q\,a \to
+  Q'\,a}\] and $\mathsf{liftListMap : \forall\, (A : Set) \to (Q \, Q'
+  : A \to Set) \to PredMap\,A\,Q\,Q' \to PredMap\,(List\,A)
+  \,(List^{\wedge}\, A\, Q)\, (List^{\wedge}\, A\, Q')}$ {\color{red}
+  defined by ?????} takes a morphism $\mathsf{f}$ of predicates and
+produces a morphism of lifted predicates {\color{red} by mapping
+  $\mathsf{f}$ over the input list to which $\mathsf{Q}$ is lifted ---
+  or other simple description?}
 \end{itemize}
-and $\mathsf{p_{List}}$ is defined as:
-\[
-  \mathsf{p_{List} = liftListMap \, (LTerm\, B) \, (LTerm^{\wedge} \, B \, Q_B)\, 
-    (P\,B\,Q_B)\, p_{ts} \, ts\, lift_{List}  : List^{\wedge}\, (LTerm\,B) \, (P\,B\,Q_B) \, ts}
-\]
-and $\mathsf{p_{ts}}$ is defined as:
-\[
-\mathsf{
-  p_{ts} = dIndLTerm\, P\, cvar\, cabs\, capp\, clistc\, B\, Q_B :  PredMap\,(LTerm\,B) \,(LTerm^{\wedge}\, B\, Q_B) \, (P\,B\,Q_B)
-  }
-\]
-where 
-\[
-\mathsf{liftListMap : \forall\, (A : Set) \to (Q_A \, Q'_A : A \to Set) 
-  \to PredMap\,A\,Q_A\,Q'_A \to PredMap\,(List\,A) \,(List^{\wedge}\, A\, Q_A)\, (List^{\wedge}\, A\, Q'_A)}
-\]
-takes a morphism of predicates and produces a morphism of lifted predicates. 
-$\mathsf{PredMap}$ is the type constructor that produces the type of morphisms between predicates, and it is defined as follows: 
-\begin{align*}
-  &\mathsf{PredMap : \forall\, (A : Set) \to (A \to Set) \to (A \to Set) \to Set } \\
-  &\mathsf{PredMap \,A\, Q_A\,Q'_A\, = \forall\, (a : A) \to Q_A\,a \to Q'_A\,a}
-\end{align*}
-
-\subsection{General case}
-
-%We restrict our scope to GADTs not featuring nesting,
-%in the sense that their constructors contain no nested calls of the recursive variable,
-%as truly nested types do.
-%This way we can avoid the issue described in Section~\ref{sec:GADT-nested}.
-%The reason for this restriction will be discussed in Section~\ref{sec:GADT-nested}.
-
-Finally, we generalize the approach taken in the previous examples
-and provide a general framework to derive induction rules for arbitrary GADTs.
+
+\section{The general framework}\label{sec:framework}
+
+{\color{red} HERE!!!!}
+
 %Since, as seen in Section~\ref{sec:GADT-nested},
@@
 %we restrict our scope to GADTs not featuring nesting.
-For that, we need to give a grammar for the types we will be considering.
-A generic GADT
+Finally, we generalize the approach taken in the previous examples and
+provide a general framework to derive deep induction rules for
+arbitrary GADTs.  For that, we need to give a grammar for the types we
+will be considering.  A generic GADT
 \begin{equation*}
@@
 \]
-to the constructor $\mathsf{c}$.
+with the constructor $\mathsf{c}$.
 
@@
 
-\section{Induction for GADTs with nesting}\label{sec:GADT-nested}
+\section{{\color{red}Induction for GADTs with nesting}}\label{sec:GADT-nested}
 
@@
 \end{multline*}
-associated to the $\mathsf{C}$ constructor.
+associated with the $\mathsf{C}$ constructor.
 
@@
 
-\section{Applications}
+\section{Case Study: Well-typed lambda terms}\label{sec:app}
 

### 2021-04-13 induction-paper/ENTCS-GADT-induction.tex
revising Sections 1-4

@@
   : A \to Set) \to PredMap\,A\,Q\,Q' \to PredMap\,(List\,A)
-  \,(List^{\wedge}\, A\, Q)\, (List^{\wedge}\, A\, Q')}$ {\color{red}
-  defined by ?????} takes a morphism $\mathsf{f}$ of predicates and
-produces a morphism of lifted predicates {\color{red} by mapping
-  $\mathsf{f}$ over the input list to which $\mathsf{Q}$ is lifted ---
-  or other simple description?}
+  \,(List^{\wedge}\, A\, Q)\, (List^{\wedge}\, A\, Q')}$, which takes
+a morphism $\mathsf{f}$ of predicates and produces a morphism of
+lifted predicates, is defined by $\mathsf{liftListMap\, A\, Q\, Q'\,
+  m\, nil\, tt = tt}$ (since $\mathsf{x : List^{\wedge}\, A\, Q\, nil
+  = \top}$ must necessarily be $\mathsf{tt}$), and by
+$\mathsf{liftListMap\, A\, Q\, Q'\, m\, (cons\, a\, l')\, (y, x') =
+  (m\,a\,y, liftListMap\, A\, Q\, Q'\, m\, l'\, x')}$ since $\mathsf{x
+  : List^{\wedge}\, A\, Q\, (cons\, a\, l') = \top}$ must be of the
+form $\mathsf{x = (y, x')}$ where $\mathsf{y : Q\,a}$ and $\mathsf{x'
+  : List^{\wedge}\, A\, Q\, l'}$. {\color{red} Double-check the
+  previous paragraph.}
 \end{itemize}

### 2021-04-15 induction-paper/ENTCS-GADT-induction.tex
revised sections 1 through 4, added abstract and related work

--- 2021-04-13 induction-paper/ENTCS-GADT-induction.tex
@@
 \usepackage{graphicx}
+\usepackage{adjustbox}
+\usepackage{amsmath}
+\usepackage[utf8]{inputenc}
+\usepackage{ccicons}
+\usepackage{verbatim}
+\usepackage{amscd}
+\usepackage{MnSymbol}
+\usepackage{xcolor}
+\usepackage{bbold}
+\usepackage{url}
+\usepackage{upgreek}
+\usepackage{stmaryrd}
+\usepackage{lipsum}
+\usepackage{tikz-cd}
+\usetikzlibrary{cd}
+\usetikzlibrary{calc}
+\usetikzlibrary{arrows}
+\usepackage{bussproofs}
+\EnableBpAbbreviations
+
 \sloppy
@@
 
-\usepackage{amsmath}
-\usepackage[utf8]{inputenc}
-\usepackage{ccicons}
-\usepackage{verbatim}
-
-\usepackage{amscd}
-\usepackage{MnSymbol}
-\usepackage{xcolor}
-
-\usepackage{bbold}
-\usepackage{url}
-\usepackage{upgreek}
-\usepackage{stmaryrd}
-
-\usepackage{lipsum}
-\usepackage{tikz-cd}
-\usetikzlibrary{cd}
-\usetikzlibrary{calc}
-\usetikzlibrary{arrows}
-
-\usepackage{bussproofs}
-\EnableBpAbbreviations
-
-\newcommand{\new}[1]{{\color{blue} {#1}}}
-
-\newcommand{\set}{\mathtt{Set}}
-
 \input{macros}
@@
 \begin{frontmatter}
-  \title{(Deep) Induction for GADTs}
-
+  \vspace*{-0.1in}
+  \title{(Deep) Induction for GADTs\vspace*{-0.1in}}
   \author{Patricia Johann~~~~}%\thanksref{JohannEmail}}
@@
 %  \thanks[JeffriesEmail]{Email: \href{mailto:jeffriesd@appstate.edu} {\texttt{\normalshape jeffriesd@appstate.edu}}}
-  \address{$\mathtt{\{johannp,ghiorzie,jeffriesd\}@appstate.edu}$\\Department of Computer Science, Appalachian State University}
+  \address{$\mathtt{\{johannp,ghiorzie,jeffriesd\}@appstate.edu}$\\
+    Appalachian State University}
+
+\vspace*{-0.1in}
 
   \begin{abstract} 
-  Abstract goes here.
+Deep data types are data types that are defined in terms of other such
+data types, including, in the case of truly nested types, themselves.
+Deep induction is an extension of structural induction that traverses
+{\em all} of the structure present in a structure of such a type,
+propagating suitable predicates to {\em all} of the data contained in
+that structure. Deep induction has been shown to be the form of
+induction most suitable for applications involving deep nested
+types. In this paper we show how to extend deep induction to a robust
+class of deep GADTs that are not truly nested. We also show that it
+cannot be extended to truly nested GADTs.
 \end{abstract}
 
-\begin{keyword}
-  Please list keywords from your paper here, separated by commas.
-\end{keyword}
+%\begin{keyword}
+%  Please list keywords from your paper here, separated by commas.
+%\end{keyword}
 
@@
 
-\begin{itemize}
-\item Weird to code in Agda if we're talking about induction rules for
-  Coq? 
-\item Agda style conventions
- \item spacing in data type declarations
-\end{itemize}
+\vspace*{-0.1in}
 
 \section{Introduction}\label{sec:intro}
+
+%\vspace*{-0.09in}
 
@@
 type of lists:
+
+\vspace*{-0.075in}
+
 \[\begin{array}{l}
@@
 The induction rule Coq automatically generates for rose trees is
+
+\vspace*{-0.075in}
+
 \[\begin{array}{l}
-\mathsf{\forall\, (a : \set)\,(P : Rose\,a \to \set) \to P\,empty \to}\\
+\mathsf{\forall\, (a : Set)\,(P : Rose\,a \to Set) \to P\,empty \to}\\
 \mathsf{\hspace*{0.15in}
@@
 of induction should traverse {\em all} of the structure present in a
-data type, propagating suitable predicates to {\em all} of the data
+data element, propagating suitable predicates to {\em all} of the data
 contained in the structure. With data types becoming ever more
@@
 be able to automatically generate genuinely useful induction
-principles for data types that go well beyond traditional ADTs. Such
+rules for data types that go well beyond traditional ADTs. Such
 data types include (truly) nested types~\cite{bm98}\footnote{A truly
@@
 families~\cite{ch88}, and deep variants of all of these.
+
+\pagebreak
 
@@
 useful induction rule for rose trees:
+
+\vspace*{-0.075in}
+
 \begin{equation}\label{eq:rose}
 \begin{array}{l}
-\mathsf{\forall \,(a : \set)\,(P : Rose\,a \to \set)\,(Q : a \to \set)
+\mathsf{\forall \,(a : Set)\,(P : Rose\,a \to Set)\,(Q : a \to Set)
   \to P\,empty \to}\\ 
@@
   Perhaps also show non-inhabitation?}
+
+\vspace*{-0.1in}
+
 \begin{equation}\label{eq:seq}
@@
 actually delivers more: it gives a general framework for deriving deep
-induction rules for a general class of GADTs directly from their
+induction rules for a general class of deep GADTs directly from their
 syntax. This framework can serve as a basis for extending modern proof
 assistants' automatic generation of structural induction rules for
-ADTs to automatic generation of deep induction rules for GADTs. As
-for ADTs and other nested types, the structural induction rule for any
+ADTs to automatic generation of deep induction rules for GADTs. As for
+ADTs and other nested types, the structural induction rule for any
 GADT can be recovered from its deep induction rule simply by taking
@@
 that is precisely what the interpretation of GADTs as their Church
-encodings fails to deliver. Instead, we first give the (deep)
-induction rule and a predicate lifting styled after those
-of~\cite{jp20} for the simplest and most important GADT, namely the
-$\mathsf{Equality}$ GADT. We then treat any other GADT by writing it
-as its so-called {\em Henry Ford
-  encoding}~~\cite{ch03,hin03,mcb99,sjsv09,sp04} and using the
-predicate liftings for $\mathsf{Equality}$ and any other constituent
-data types to appropriately thread the custom predicates through the
-structures of the GADT. {\color{red} Significantly, since we are
-  treating GADTs as syntax-only, rather than functorial, we cannot
-  handle truly nested GADTs like we handled truly nested types, but
-  this is not much of a restriction, since GADTs defined over
-  themselves have not, to our knowledge, appeared in applications.}
-This process yields, in particular, the deep induction rule for
-$\mathsf{Seq}$ given in Section~\ref{sec:ind-seq}. {\color{red} MORE
-  HERE???}
+encodings fails to deliver. Instead, we first give a predicate lifting
+styled after those of~\cite{jp20}, together with a (deep) induction
+rule, and for the simplest --- and arguably most important --- GADT,
+namely the equality GADT. (See Section~\ref{sec:ind-equal}.) We can
+then derive the deep induction rule for any other GADT $\mathsf{G}$ by
+{\em i}) using the equality GADT to represent $\mathsf{G}$ as its
+so-called {\em Henry Ford
+  encoding}~\cite{ch03,hin03,mcb99,sjsv09,sp04}, and {\em ii}) using
+the predicate liftings for the equality GADT and any other GADTs
+appearing in the definition of $\mathsf{G}$ to appropriately thread
+the custom predicates for the primitive types appearing in
+$\mathsf{G}$ through its structure. This two-step process delivers
+deep induction rules for a broad class of deep GADTs. In
+Section~\ref{sec:GADTs} we introduce a series of increasingly complex
+GADTs as running examples, and in Section~\ref{sec:deep-ind-GADTs} we
+derive a deep induction rule for each of them. In particular, we
+derive the deep induction rule for $\mathsf{Seq}$ in
+Section~\ref{sec:ind-seq}. We present our general framework for
+deriving (deep) induction rules for (deep) GADTs in
+Section~\ref{sec:framework}, and observe that the derivations in
+Section~\ref{sec:deep-ind-GADTs} are all instances of it.  In
+Section~\ref{sec:GADT-nested} we show that, by contrast with truly
+nested types, which do have a functorial semantics, syntax-only GADTs'
+lack of functoriality means that it is not possible to extend
+induction --- deep or otherwise --- to truly nested GADTs. This does
+not appear to be much of a restriction, however, since GADTs defined
+over themselves do not, to our knowledge, appear in applications or
+the literature.
 
@@
 
-\subsection{Related Work}
-
-Ullrich's plugin?
-
-Tassi's Elpi
-
-Peter Selinger and  Fu Peng
+%\subsection{Related Work}
+
+\vspace*{0.05in}
+
+{\bf Related Work\/}
+Various techniques for deriving induction rules for data types that go
+beyond ADTs have been studied. For example, Fu and
+Selinger~\cite{fs18} show, via examples, how to derive induction rules
+for arbitrary nested types. Unfortunately, however, their technique is
+rather {\em ad hoc}, so is unclear how to generalize it to nested
+types other than the specific ones in the examples. Moreover, it
+actually derives induction rules for data types {\em related} to the
+original nested types rather than for the original nested types
+themselves, and it is unclear whether or not the derived rules are
+sufficiently expressive to prove all results about the original nested
+types that we would expect to be provable by induction. This latter
+point echoes the issue with Coq-derived induction rule for rose trees
+raised in Section~\ref{sec:intro}, which has the unfortunate effect of
+forcing users to manually write induction (and other) rules for such
+types for use in that system. Tassi~\cite{tas19} has done exactly
+that, deriving induction rules for data type definitions in Coq using
+unary parametricity. Tassi's technique seems to be essentially
+equivalent to that of~\cite{jp19} for nested types, although he does
+not permit true nesting. More recently, Ulrich~\cite{ull20} has
+implemented a plugin in MetaCoq to generate induction rules for nested
+types. This plugin is also based on unary parametricity and, again,
+true nesting is not permitted.  {\color{red} As far as we know, no
+  attempt has (yet) been made to extend either implementation to
+  GADTs.} In fact, we know of no work other than that reported here
+that specifically addresses induction rules for (deep) GADTs.
 
@@
 
-\begin{comment}
-\subsection{Syntax of ADTs and nested types}\label{ssec:syntax-adts}
-
-\emph{(Polynomial) algebraic data types} (ADTs), both built-in and user-defined,
-have long been at the core of functional languages such as Haskell, ML, Agda, Epigram, and Idris.
-ADTs are used extensively in functional programming to structure computations,
-to express invariants of the data over which computations are defined,
-and to ensure the type safety of programs specifying those computations.
-ADTs include unindexed types, such as the type of natural numbers,
-and types indexed over other types, such as the quintessential example of an ADT,
-the type of lists (here coded in Agda)
-\begin{equation}\label{eq:list}
-\begin{array}{l}
-\mathsf{data\ List : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;nil\,\,\,\,\,\, :\, List\,A}\\
-\mathsf{\;\;\;\;\;\;\;cons\, :\, A \to List\,A \to List\,A} 
-\end{array}
-\end{equation}
-Note that all occurrences of $\mathsf{List}$ in the above encoding
-are instantiated at the same index $\mathsf{A}$.  Thus, the instances
-of $\mathsf{List}$ at various indices are defined independently from
-one another.  That is a defining feature of ADTs: an ADT defines a
-\emph{family of inductive types}, one for each index type.
-
-Over time, there has been a notable trend toward data types whose
-non-regular indexing can capture invariants and other sophisticated
-properties that can be used for program verification and other
-applications.  A simple example of such a type is given by Bird and
-Meertens’~\cite{bm98} prototypical \emph{nested type}
-\begin{equation}\label{eq:ptree}
-\begin{array}{l}
-\mathsf{data\ PTree : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;pLeaf\,\,\, :\, A \to PTree\,A}\\
-\mathsf{\;\;\;\;\;\;\;pNode\, :\, PTree\,(A \times A) \to PTree\,A} 
-\end{array}
-\end{equation}
-of perfect trees, which can be thought of as constraining lists to
-have lengths that are powers of 2.  In the above code, the constructor
-$\mathsf{PNode}$ uses data of type $\mathsf{PNode\,(A \times A)}$ to
-construct data of type $\mathsf{PNode\,A}$.  Thus, it is clear that
-the instantiations of $\mathsf{PNode}$ at various indices cannot be
-defined independently, so that the entire family of types must
-actually be defined at once.  A nested type thus defines not a family
-of inductive types, but rather an \emph{inductive family of types}.
-
-Nested types include simple nested types, like perfect trees,
-none of whose recursive occurrences occur below another type constructor,
-and \emph{truly} nested types, such as the nested type
-\begin{equation}\label{eq:bush}
-\begin{array}{l}
-\mathsf{data\ Bush : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;bNil\,\,\,\,\,\, :\, Bush\,A}\\
-\mathsf{\;\;\;\;\;\;bCons\, :\, A \to Bush\,(Bush\,A) \to Bush\,A} 
-\end{array}
-\end{equation}
-of bushes, whose recursive occurrences appear below their own type constructors.
-Note that, while the constructors of a nested type can contain occurrences of the type instantiated at any index,
-the return types of its constructors still have to be the same type instance of the type being defined.
-In other words, all constructors of $\mathsf{PTree\,A}$ have to return an element of type $\mathsf{PTree\,A}$,
-and all constructors of $\mathsf{Bush\,A}$ have to return an element of type $\mathsf{Bush\,A}$.
-%the return types of the constructors are still required to be instantiated at the same variable $\mathsf{a}$.}
-%As previously noted, the return types of the constructors of a nested
-%type have to be the same type instance of the type being defined.
-\end{comment}
-
-\subsection{Induction principles for ADTs and nested
-  types}\label{ssec:induction-nested}
-
-A structural induction principle for a data type allows us to prove
-that if a predicate holds for every element inductively produced by
-the data type's constructors then it holds for every element of the
-data type.  In this paper, we are interested in induction principles
-for proof-relevant predicates.  A proof-relevant predicate on a type
-$\mathsf{A : Set}$ is a function $\mathsf{P\,:\,A \to Set}$ mapping
-each $\mathsf{x : A}$ to the set of proofs that $\mathsf{P\,x}$ holds.
-For example, the induction principle for the standard data type
+A structural induction rule for a data type allows us to prove that if
+a predicate holds for every element inductively produced by the data
+type's constructors then it holds for every element of the data type.
+In this paper, we are interested in induction rules for proof-relevant
+predicates.  A proof-relevant predicate on a type $\mathsf{A : Set}$
+is a function $\mathsf{P\,:\,A \to Set}$ mapping each $\mathsf{a : A}$
+to the set of proofs that $\mathsf{P\,a}$ holds.  For example, the
+induction rule for the standard list type
+
+\vspace*{-0.1in}
+
 \begin{equation*}\label{eq:list}
@@
 \end{equation*}
-of lists is
+
+\vspace*{-0.1in}
+
+\noindent
+is
+
+\vspace*{-0.2in}
+
 \begin{equation*}
@@
 \end{equation*}
+
+\vspace*{-0.1in}
+
+\noindent
 As in Coq's induction rule for rose trees, the data inside a
@@
 ignored) by this structural induction rule.
-%Indeed, the structural induction rule for a data type inducts
-%over only the top-level structures of data types, leaving any data
-%internal to the top-level structure untouched.  Since this kind of
-%induction principle is only concerned with the structure of the type,
-%and unconcerned with the contained data, we will then refer to it as
-%\emph{structural induction}.
 By contrast, the deep induction rule for lists is parameterized over a
@@
 introduction. For $\mathsf{List^\wedge}$ as described in the
-introduction, it is
-\[
-\begin{array}{l}
+introduction it is
+
+\vspace*{-0.2in}
+
+\[\begin{array}{l}
 \mathsf{\forall (A : Set) (P : List\, A \to Set) (Q : A \to Set)
@@
   \to P\,as } 
-\end{array}
-\]
-
-Structural induction can also be extended to nested types, such as the
+\end{array}\]
+
+\vspace*{-0.1in}
+
+Structural induction can be extended to nested types, such as the
 following type of perfect trees~\cite{bm98}:
+
+\pagebreak
+
 \begin{equation*}\label{eq:ptree}
@@
 independently, and that the entire inductive family of types must
-therefore be defined at once. This intertwinedness of the instances
-of nested types is reflected in their structural induction rules,
-which must necessarily involve polymorphic predicates, rather than the
-monomorphic predicates appearing in structural induction rules for
-ADTs. The structural induction rule for perfect trees, for example, is
+therefore be defined at once. This intertwinedness of the instances of
+nested types is reflected in their structural induction rules, which,
+as explained in~\cite{jp20}, must necessarily involve polymorphic
+predicates rather than the monomorphic predicates appearing in
+structural induction rules for ADTs. The structural induction rule for
+perfect trees, for example, is
+
+\vspace*{-0.2in}
+
 \[\begin{array}{l}
@@
 \end{array}\]
+
+\vspace*{-0.1in}
+
+\noindent
 The deep induction rule for perfect trees similarly uses polymorphic
 predicates but otherwise follows the now-familiar pattern:
+
+\vspace*{-0.2in}
+
 \[\begin{array}{l}
@@
 \end{array}\]
+
+\vspace*{-0.1in}
+
+\noindent
 Here, $\mathsf{Pair^{\wedge} : \forall (A\; B: Set) \to (A \to Set)
@@
 nested types, i.e., to nested types whose recursive occurrences appear
-below their own type constructors. Indeed, defining any induction rule
-for such a type requires that we can lift the predicate argument to
-that rule over the type itself. The observation that any induction
-rule for a truly nested type would thus necessarily have to be a deep
-induction rule was, in fact, the original motivation for the
-introduction of deep induction in~\cite{jp20}.
-
-The quintessential example of a truly nested type is that of
+below themselves. The quintessential example of such a type is that of
 bushes~\cite{bm98}:
@@
 \mathsf{\;\;\;\;\;\;bnil\,\,\,\,\,\, :\, Bush\,A}\\
-\mathsf{\;\;\;\;\;\;bvons\, :\, A \to Bush\,(Bush\,A) \to Bush\,A} 
+\mathsf{\;\;\;\;\;\;bcons\, :\, A \to Bush\,(Bush\,A) \to Bush\,A} 
 \end{array}
 \end{equation*}
-Its (deep) induction rule is
+
+\noindent
+Even defining a structural induction rule for bushes requires that we
+be able to lift the rule's polymorphic predicate argument to
+$\mathsf{Bush}$ itself. The more general observation that an induction
+rule for any truly nested type must therefore necessarily be a deep
+induction rule was, in fact, the original motivation for the
+development of deep induction in~\cite{jp20}. The deep induction rule
+for bushes is
 \[\begin{array}{l}
@@
 \end{array}\]
-where $\mathsf{Bush^{\wedge} : \forall (A : Set) \to (A \to Set) \to
+Here, $\mathsf{Bush^{\wedge} : \forall (A : Set) \to (A \to Set) \to
   Bush\,A \to Set}$ is the following lifting of a predicate
@@
 
-It is worth noting that for those nested types that do admit both
-structural induction and deep induction, the latter generalizes the
-former. Indeed, the structural induction rule for a nested type is
-recoverable from its deep induction rule by taking the custom
-predicates on their data of primitive types to be constantly
-$\mathsf{True}$-valued predicates. Such an instantiation ensures that
-a deep induction rule only inspects the top-level structure of its
-argument data, rather than the contents of that structure.  A concrete
-example of such a derivation is given in Section~\ref{sec:ind-equal}.
+Although a truly nested type admits only a single induction rule, it
+is worth noting that for those nested types that do admit distinct
+structural induction and deep induction rules, the latter generalizes
+the former. Indeed, the structural induction rule for such a nested
+type is recoverable from its deep induction rule by taking the custom
+predicates on its data of primitive types to be constantly
+$\mathsf{True}$-valued predicates. This instantiation ensures that the
+resulting induction rule only inspects the top-level structure of its
+argument, rather than the contents of that structure, which exactly
+coindices with what structural induction should do.
+%A concrete example of such a derivation is given in
+%Section~\ref{sec:ind-equal}.
 
@@
 
-While a constructor of a nested type can take {\em as arguments} data
-whose types involve instances of that type at indices other than the
-one being defined, its return type must still be at the (variable)
-type instance being defined. For example, every constructor of
+While a data constructor for a nested type can take {\em as arguments}
+data whose types involve instances of that type at indices other than
+the one being defined, its return type must still be at the (variable)
+type instance being defined. For example, every data constructor for
 $\mathsf{PTree\,A}$ must return an element of type
-$\mathsf{PTree\,A}$. \emph{Generalized algebraic data types}
-(GADTs)~\cite{ch03,sp04,xcc03} relax this restriction, allowing their
-constructors both to take as arguments \emph{and return as results}
-data whose types involve instances of the GADT other than the one
-being defined. As with the return type of the $\mathsf{sPair}$
-constructor in~\eqref{eq:seq}, these indices can be structured.
+$\mathsf{PTree\,A}$, regardless of the instances of $\mathsf{PTree}$
+appearing in the types of its arguments. GADTs relax this restriction,
+allowing their data constructors both to take as arguments \emph{and
+  return as results} data whose types involve instances of them other
+than the one being defined. And as with the return type of
+$\mathsf{pair}$ in~\eqref{eq:seq}, these instances can be structured.
 %Applications of GADTs include generic programming, modeling
@@
 achieved by allowing the programmer to give the type signatures of the
-GADT's data constructors independently, and then using {\color{red}
-  pattern matching} to force the desired type refinement. For example,
+GADT's data constructors independently, and then taking advantage of
+pattern matching to force the desired type refinement. For example,
 the {\em equality} GADT
@@
 $\mathsf{a}$ and $\mathsf{b}$ are syntactically identical then the
-type $\mathsf{Equal\,a\,b}$ contains a single data element. It
-contains no data elements otherwise.
+type $\mathsf{Equal\,a\,b}$ contains the single data element
+$\mathsf{refl}$. It contains no data elements otherwise.
 
 The importance of the equality GADT lies in the fact that we can
-understand other GADTs in terms of it. For example, the GADT of
-sequences in~\eqref{eq:seq} comprises constant sequences of data of
-any type $\mathsf{A}$ and sequences obtained by pairing the data in
-two already existing sequences. This GADT can be understood as its
-{\em Henry Ford encoding}~\cite{ch03,hin03,mcb99,sjsv09,sp04}, which
-makes critical use of the equality GADT, as follows:
+understand other GADTs in terms of it. For example, the GADT
+$\mathsf{Seq}$ from~\eqref{eq:seq} comprises constant sequences of
+data of any type $\mathsf{A}$ and sequences obtained by pairing the
+data in two already existing sequences. This GADT can be rewritten as
+its Henry Ford encoding, which makes critical use of the equality
+GADT, as follows:
 \begin{equation}\label{eq:eq_seq}
@@
 \end{equation}
-Here, the requirement that $\mathsf{pair}$ produce data at an
-instance of $\mathsf{Seq}$ that is a product type is replaced with the
+Here, the requirement that $\mathsf{pair}$ produce data at an instance
+of $\mathsf{Seq}$ that is a product type is replaced with the
 requirement that $\mathsf{pair}$ produce data at an instance of
 $\mathsf{Seq}$ that is \emph{equal} to a product type. As we will see
-below, this use of the equality GADT is key to deriving deep induction
-rules for more general GADTs.
-
-\vspace*{0.05in}
-
-Finally, we give an example of a deep GADT. The $\mathsf{LTerm}$ GADT
-encodes terms of a simply typed lambda calculus. More robust
-variations on this calculus are, of course, possible. But since
-$\mathsf{LTerm}$ is rich enough to illustrate all of the essential
-aspects of deep GADTs --- and later, in Section~\ref{sec:ind-lam},
-their deep induction rules --- while still being small enough to
-ensure clarity of exposition, we keep our calculus to a minimum.
-
-Types are either booleans, arrow types, or list types, and are
-represented by the following GADT:
-\begin{equation*}\label{eq:eq_ltype}
-\begin{array}{l}
-\mathsf{data\ LType : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;\;bool :\, \forall (B : Set) \to LType\,Bool}\\
-\mathsf{\;\;\;\;\;\;\;\;arr\,\,\, :\, \forall (B\,C : Set) \to
-  LType\,B \to LType\,C \to LType\,(B \to C)}\\
-\mathsf{\;\;\;\;\;\;\;\;list\,\, :\, \forall (B : Set)
-  \to LType\,B \to LType\,(List\,B)}
-\end{array}
-\end{equation*}
-Terms are either variables, abstractions, applications, or lists of
-such terms. They are represented by
-\begin{equation*}\label{eq:eq_lterm}
-\begin{array}{l}
-\mathsf{data\ LTerm : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;\;var\,\,\,\,\,\,\,:\, String \to LType\,A \to
-  LTerm\,A} \\  
-\mathsf{\;\;\;\;\;\;\;\;abs\,\,\,\,\,\, :\, \forall (B\,C : Set) \to
-  String \to LType\,B \to LTerm\,C \to
-  LTerm\,(B \to C)}\\ 
-  \mathsf{\;\;\;\;\;\;\;\;app\,\,\,\,\, :\, \forall (B : Set) \to
-    LTerm (B \to A) \to LTerm\,B \to LTerm\,A} \\ 
-  \mathsf{\;\;\;\;\;\;\;\;list\,\,\, :\, \forall (B : Set) \to
-    List\,(LTerm\,B) \to LTerm\,(List \,B)} 
-\end{array}
-\end{equation*}
-The type parameter for the GADT $\mathsf{LTerm}$ tracks the types of
-simply typed lambda calculus terms. For example, $\mathsf{LTerm\,A}$
-is the type of lambda terms of type $\mathsf{A}$. Variables are tagged
-with their types by the data constructors $\mathsf{var}$ and
-$\mathsf{abs}$, whose $\mathsf{LType}$ arguments ensure that the
-tagging types are legal.
-%types, e.g., $\mathsf{Bool}$, $\mathsf{Bool \to Bool}$,
-%  $\mathsf{List\, (Bool \to Bool)}$, etc. That is, the presence of
-%$\mathsf{LType}$ in these constructors thus guarantees that
-As a result, all lambda terms produced by $\mathsf{var}$,
-$\mathsf{abs}$, $\mathsf{app}$, and $\mathsf{list}$ are necessarily
-well-typed.
-
-{\color{red} Maybe only give in the following form.}
-The Henry Ford encodings representing $\mathsf{LType}$ and
-$\mathsf{LTerm}$ in terms of the equality GADT are:
+in Section~\ref{sec:deep-ind-GADTs}, the presence of the equality GADT is
+key to deriving deep induction rules for GADTs.
+
+Neither $\mathsf{Equal}$ nor $\mathsf{Seq}$ is a deep GADT, but the
+following GADT $\mathsf{LTerm}$, which encodes terms of a simply typed
+lambda calculus, is. More robust variations on $\mathsf{LTerm}$ are,
+of course, possible. But since this variation is rich enough to
+illustrate all essential aspects of deep GADTs --- and later, in
+Section~\ref{sec:ind-lam}, their deep induction rules --- while still
+being small enough to ensure clarity of exposition, we keep it to a
+minimum.
+
+Types are either booleans, arrow types, or list types. They are
+represented by the Henry Ford GADT
 \begin{equation}\label{eq:eq_ltype}
@@
 \end{equation}
-and
+Terms are either variables, abstractions, applications, or lists of
+terms. They are represented by
 \begin{equation}\label{eq:eq_lterm}
@@
 \end{equation}
-We will revisit these GADTs in Sections~\ref{sec:deep-ind-GADTs}
-and~\ref{sec:app}.
+The type parameter for $\mathsf{LTerm}$ tracks the types of simply
+typed lambda calculus terms. For example, $\mathsf{LTerm\,A}$ is the
+type of simply typed lambda terms of type $\mathsf{A}$. Variables are
+tagged with their types by the data constructors $\mathsf{var}$ and
+$\mathsf{abs}$, whose $\mathsf{LType}$ arguments ensure that their
+type tags are legal types. This ensures that all lambda terms
+produced by $\mathsf{var}$, $\mathsf{abs}$, $\mathsf{app}$, and
+$\mathsf{list}$ are well-typed.  We will revisit these GADTs in
+Sections~\ref{sec:deep-ind-GADTs} and~\ref{sec:app}.
 
 \section{(Deep) induction for GADTs}\label{sec:deep-ind-GADTs}
-
-{\color{red} Combine and/or put in figures?}
 
@@
 abstract these to a general framework that can be applied to any deep
-GADT that is not truly nested. {\color{red} directly or indirectly ---
-  what should we call these?} As expected, the derivation for the
-equality GADT plays a central role in that for more general ones.
-%instrumental in stating and deriving the induction rules for other
-%GADTs.
-
-\subsection{(Deep) induction for equality}\label{sec:ind-equal}
-
-To define the deep induction rule for any deep GADT $\mathsf{G}$ we
-first need a predicate lifting that lifts any predicate on a type
-$\mathsf{A}$ and to a predicate on $\mathsf{G\,A}$. The predicate
-lifting 
-\[\mathsf{Equal^{\wedge} : \forall (A\,B : Set) \to (A \to Set) \to (B
-  \to Set) \to Equal\,A\,B \to Set}\] 
+GADT that is not truly nested. As hinted above, the predicate lifting
+for the equality GADT plays a central role in deriving both s
+tructural and deep induction rules for more general GADTs.
+
+\subsection{(Deep) induction for $\mathsf{Equal}$}\label{sec:ind-equal}
+
+To define the (deep) induction rule for any (deep) GADT $\mathsf{G}$
+we first need to define a predicate lifting that maps a predicate on a
+type $\mathsf{A}$ and to a predicate on $\mathsf{G\,A}$. Such a
+predicate lifting
+$\mathsf{Equal^{\wedge} : \forall (A\,B : Set) \to (A \to Set) \to (B
+  \to Set) \to Equal\,A\,B \to Set}$
 for $\mathsf{Equal}$ is defined by
@@
   Equal\,(Q\,a)(Q'\,a)}\]
-As expected, it takes two predicates on the same type as input and
-tests them for extensional equality.
+which takes two predicates on the same type as input and tests them
+for extensional equality.
 
 Next, we need to associate with each data constructor $\mathsf{c}$ of
-$\mathsf{G}$ an induction hypothesis asserting that, if the custom
-predicate arguments to a predicate $\mathsf{P}$ on $\mathsf{G}$ can be
-lifted to $\mathsf{G}$ itself, then $\mathsf{c}$ {\color{red}
-  preserves} $\mathsf{P}$. The following induction hypothesis
-$\mathsf{dIndRefl}$ is thus associated with the $\mathsf{refl}$
-constructor for $\mathsf{Equal}$:
+$\mathsf{G}$ an {\em induction hypothesis} asserting that, if the
+custom predicate arguments to a predicate $\mathsf{P}$ on $\mathsf{G}$
+can be lifted to $\mathsf{G}$ itself, then $\mathsf{c}$ {\em respects}
+$\mathsf{P}$, i.e., $\mathsf{c}$ constructs data satisfying the
+instance of $\mathsf{P}$ at those custom predicates. The following
+induction hypothesis $\mathsf{dIndRefl}$ is thus associated with the
+$\mathsf{refl}$ constructor for $\mathsf{Equal}$:
 \begin{equation*}\label{eq:ind-refl}
@@
 \end{equation*}
-The deep induction rule for $\mathsf{G}$ now states that, if a
-predicate $\mathsf{P}$ is {\color{red} preserved} by all of
-$\mathsf{G}$'s constructors, then $\mathsf{P}$ is satisfied by every
-element of $\mathsf{G}$ to which the custom predicate arguments to
-$\mathsf{P}$ can be successfully lifted.  The deep induction rule for
-the GADT $\mathsf{Equal}$ is thus
+
+The deep induction rule for $\mathsf{G}$ now states that, if all of
+$\mathsf{G}$'s data constructors respect a predicate $\mathsf{P}$,
+then $\mathsf{P}$ is satisfied by every element of $\mathsf{G}$ to
+which the custom predicate arguments to $\mathsf{P}$ can be
+successfully lifted.  The deep induction rule for $\mathsf{Equal}$ is
+thus
 \begin{equation}\label{eq:ind-equal}
@@
 \end{equation}
-
-To prove that the deep induction rule for $\mathsf{Equal}$ is sound we
-must provide a witness inhabiting the type in~\eqref{eq:ind-equal}.
-We first note that, by {\color{red} pattern matching}, we only need to
-consider the case where $\mathsf{A} = \mathsf{B}$ and $\mathsf{e} =
-\mathsf{refl}$.  We can therefore define a term $\mathsf{dIndEqual}$
-of the type in \eqref{eq:ind-equal} by
-\[\mathsf{dIndEqual\;P\;crefl\;A\;A\;Q_A\;Q_A'\;refl\;liftE =
-  crefl\;A\;Q_A\;Q_A'\,liftE}\] 
-We can also recover the structural induction rule
+To prove that this rule is sound we must provide a witness
+$\mathsf{dIndEqual}$ inhabiting the type in~\eqref{eq:ind-equal}.  By
+pattern matching, we need only consider the case where $\mathsf{A} =
+\mathsf{B}$ and $\mathsf{e} = \mathsf{refl}$, so we can define
+$\mathsf{dIndEqual}$ by
+$\mathsf{dIndEqual\;P\;crefl\;A\;A\;Q_A\;Q_A'\;refl\;liftE =
+  crefl\;A\;Q_A\;Q_A'\,liftE}$. We can recover the structural
+induction rule
 \begin{equation}\label{eq:sind-equal}
@@
 for $\mathsf{Equal}$ by defining a term $\mathsf{indEqual}$ of the
-type in~\eqref{eq:sind-equal} by
-\[\mathsf{indEqual\;Q\;srefl\;A\;B\;e =
-  dIndEqual\;P\;srefl\;A\;B\;K^A_\top\;K^B_\top\;e\,liftE}\] Here,
-$\mathsf{e : Equal\,A\,B}$,
-$\mathsf{P : \forall (A\,B : Set) \to (A \to Set) \to (B \to
-  Set) \to Equal\,A\,B \to Set}$ is defined by
+type in~\eqref{eq:sind-equal} by $\mathsf{indEqual\;Q\;srefl\;A\;B\;e
+  = dIndEqual\;P\;srefl\;A\;B\;K^A_\top\;K^B_\top\;e\,liftE}$. Here,
+$\mathsf{e : Equal\,A\,B}$, $\mathsf{P : \forall (A\,B : Set) \to (A
+  \to Set) \to (B \to Set) \to Equal\,A\,B \to Set}$ is defined by
 $\mathsf{P\;A\;B\;Q_A\;Q_B\;e =}$ $\mathsf{Q\;A\;B\;e}$,
@@
 {\color{red} $\mathsf{\top}$}-valued predicates on $\mathsf{A}$ and
-  $\mathsf{B}$, respectively, and
-$\mathsf{liftE :}$ $\mathsf{Equal^{\wedge}\;A\;B\;K^A_\top\;K^B_\top\;e}$
-  is fully defined (by pattern matching) by $\mathsf{liftE\,a = refl :
-    Equal\,A\,A}$ for every $\mathsf{a : A}$.
-
-{\color{red} Although the structural induction rule for any GADT
-  $\mathsf{G}$ can similarly be recovered from its deep induction rule
-  by instantiating every custom predicate by the appropriate
-  constantly {\color{red}$\mathsf{\top}$}-valued predicate, we will
-  not explicitly note this for each of the examples below.}
-
-\subsection{(Deep) induction for sequences}\label{sec:ind-seq}
+$\mathsf{B}$, respectively, and $\mathsf{liftE :}$
+$\mathsf{Equal^{\wedge}\;A\;B\;K^A_\top\;K^B_\top\;e}$ is
+%fully
+defined
+%(by pattern matching)
+by $\mathsf{liftE\,a = refl : Equal\,A\,A}$ for every $\mathsf{a :
+  A}$.  The structural induction rule for any GADT $\mathsf{G}$ that
+is not truly nested can similarly be recovered from its deep induction
+rule by instantiating every custom predicate by the appropriate
+constantly {\color{red}$\mathsf{\top}$}-valued predicate.
+
+\subsection{(Deep) induction for $\mathsf{Seq}$}\label{sec:ind-seq}
 
 To derive the deep induction rule for the GADT $\mathsf{Seq}$ we use
-its Henry Ford encoding from~\eqref{eq:eq_seq}.  As usual, the first
-step is to define the predicate lifting $\mathsf{Seq^\wedge : \forall
-  (A : Set) \to (A \to Set) \to Seq\,A \to Set}$. It is defined by
-\[\mathsf{Seq^{\wedge}\,A\,Q_A\,(const\,a) = Q_A\,a}\]
-for $\mathsf{a : A}$, and
-\[\begin{array}{l}
+its Henry Ford encoding from~\eqref{eq:eq_seq}.  We first define its
+predicate lifting $\mathsf{Seq^\wedge : \forall (A : Set) \to (A \to
+  Set) \to Seq\,A \to Set}$ by
+\[\begin{array}{lll}
+\mathsf{Seq^{\wedge}\,A\,Q_A\,(const\,a)} & = & \mathsf{Q_A\,a}\\
 \mathsf{Seq^{\wedge}\,A\,Q_A\,(sPair\,B\,C\,e\,s_B\,s_C)}
-\\ \quad\mathsf{= \exists [Q_B] \exists [Q_C]\, Equal^{\wedge}\,A\, (B
+&=&\mathsf{\exists [Q_B] \exists [Q_C]\, Equal^{\wedge}\,A\, (B
   \times C)\, Q_A\, (Q_B \times Q_C) \, e \times
-  Seq^{\wedge}\,B\,Q_B\,s_B \times Seq^{\wedge}\,C\,Q_C\,s_C }
+  Seq^{\wedge}\,B\,Q_B\,s_B \times Seq^{\wedge}\,C\,Q_C\,s_C}
 \end{array}\]
-for $\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C : C \to Set}$, $\mathsf{e
-  : Equal\,A\,(B \times C)}$, $\mathsf{s_B : Seq\,B}$, $\mathsf{s_C :
-  Seq\,C}$, and $\mathsf{\exists [x]\, F \,x}$ is syntactic sugar for the
-type of dependent pairs $\mathsf{(x,b)}$ where $\mathsf{x : A}$ and
-$\mathsf{b : F\, x}$ and $\mathsf{F : A \to Set}$.
+Here, $\mathsf{a : A}$, $\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C : C
+  \to Set}$, $\mathsf{e : Equal\,A\,(B \times C)}$, $\mathsf{s_B :
+  Seq\,B}$, $\mathsf{s_C : Seq\,C}$, and $\mathsf{\exists [x]\, F
+  \,x}$ is syntactic sugar for the type of dependent pairs
+$\mathsf{(x,b)}$ where $\mathsf{x : A}$ and $\mathsf{b : F\, x}$ and
+$\mathsf{F : A \to Set}$.
 
@@
 \end{array}\]
-associated with the constructor $\mathsf{pair}$. The deep induction
-rule for $\mathsf{Seq}$ is thus
+associated with the constructor $\mathsf{pair}$. Then the deep
+induction rule for $\mathsf{Seq}$ is
 \begin{equation}\label{eq:ind-seq}
@@
 
-To prove that the deep induction rule for $\mathsf{Seq}$ is sound we
-must provide a witness $\mathsf{dIndSeq}$ inhabiting the type
-in~\eqref{eq:ind-seq}. We proceed by pattern matching on
-$\mathsf{s_A}$.
-\begin{itemize}
-\item If $\mathsf{s_A = const\,a}$ for $\mathsf{a : A}$, then we
-  define \[\mathsf{dIndSeq\;P\;cconst\;cpair\;A\;Q_A\;(const\,a)\;liftA
-    = cconst\;A\;Q_A\;a\;liftA}\] Since
-  $\mathsf{Seq^{\wedge}\,A\,Q_A\,(const\,a)}$ $\mathsf{= Q_A\,a}$, and
-  thus $\mathsf{liftA : Q_A\,a}$, the definition does indeed
-  type-check.
-\item If $\mathsf{s_A = sPair\,B\,C\,e\,s_B\,s_C}$ for $\mathsf{e :
-  Equal\,A\,(B \times C)}$, $\mathsf{s_B : Seq\,B}$, and $\mathsf{s_C
-  : Seq\,C}$, then we define
-  \[\mathsf{dIndSeq\,P\,cconst\,cpair\,A\,Q_A\,(sPair\,B\,C\,e\,s_B\,s_C)\,
-    (Q_B,Q_C, liftE, liftB, liftC) =
-    cpair\,A\,B\,C\,Q_A\,Q_B\,Q_C\,s_B\,s_C\,e\,liftE\,p_B\,p_C}\]
-  Here, $\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C : C \to Set}$,
-  $\mathsf{liftE : Equal^{\wedge}\,A\, (B \times C)\, Q_A\, (Q_B
-    \times Q_C) \, e}$, $\mathsf{liftB : Seq^{\wedge}\,B\,Q_B\,s_B}$,
-  and $\mathsf{liftC :}$\\
-\noindent
+To prove that this rule is sound we provide a witness
+$\mathsf{dIndSeq}$ inhabiting the type in~\eqref{eq:ind-seq} as
+follows:
+\[\begin{array}{lll}
+\mathsf{dIndSeq\;P\;cconst\;cpair\;A\;Q_A\;(const\,a)\;liftA}  &
+    = &  \mathsf{cconst\;A\;Q_A\;a\;liftA}\\
+\mathsf{dIndSeq\,P\,cconst\,cpair\,A\,Q_A\,(sPair\,B\,C\,e\,s_B\,s_C)\,
+    (Q_B,Q_C, liftE, liftB, liftC)}  & = &
+\mathsf{cpair\,A\,B\,C\,Q_A\,Q_B\,Q_C\,s_B\,s_C\,e\,liftE\,p_B\,p_C}
+\end{array}\]
+In the first clause ahove, $\mathsf{a : A}$, $\mathsf{Q_A : A \to
+  Set}$, $\mathsf{liftA : Seq^{\wedge}\,A\,Q_A\,(const\,a) =
+  Q_A\,a}$. In the second, $\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C :
+  C \to Set}$, $\mathsf{e : Equal\,A\,(B \times C)}$, $\mathsf{s_B :
+  Seq\,B}$, $\mathsf{s_C : Seq\,C}$, $\mathsf{liftE :
+  Equal^{\wedge}\,A\, (B \times C)\, Q_A\, (Q_B \times Q_C) \, e}$,
+$\mathsf{liftB : Seq^{\wedge}\,B\,Q_B\,s_B}$, and $\mathsf{liftC :}$
 $\mathsf{Seq^{\wedge}\,C\,Q_C\,s_C}$ --- which together ensure that
 $\mathsf{(Q_B, Q_C, liftE, liftB, liftC) :
-    Seq^{\wedge}\,A\,Q\,(sPair\,B\,C\,e\,s_B\,s_C)}$ --- and\\
-%  $\mathsf{p_B}$ and $\mathsf{p_C}$\\ are defined by
-$\mathsf{p_B}
-  =\mathsf{dIndSeq\,P\,cconst\,cpair\,B\,Q_B\,s_B\,liftB :
+  Seq^{\wedge}\,A\,Q\,(sPair\,B\,C\,e\,s_B\,s_C)}$ --- and
+$\mathsf{p_B} =\mathsf{dIndSeq\,P\,cconst\,cpair\,B\,Q_B\,s_B\,liftB :
   P\,B\,Q_B\,s_B}$ and $\mathsf{p_C}
@@
   P\,C\,Q_C\,s_C}$.
-\end{itemize}
-
-\subsection{(Deep) induction for simply typed lambda
-  terms}\label{sec:ind-lam}
-
-To derive the deep induction rule for the GADT $\mathsf{LTerm}$ we use
-its Henry Ford encoding from~\ref{eq:eq_ltype}
-and~\eqref{eq:eq_lterm}. The first step is to define the predicate
-liftings for $\mathsf{LType}$ and $\mathsf{LTerm}$. We will also need
-a predicate lifting for arrow types, since arrow types appear in
-$\mathsf{LType}$ and $\mathsf{LTerm}$.
-
-The predicate lifting $\mathsf{Arr^{\wedge} : \forall (A\, B : Set)
-  \to (A \to Set) \to (B \to Set) \to (A \to B) \to Set}$ for arrow
-types is defined by
-\[\mathsf{Arr^{\wedge}\, A\, B\, Q_A\, Q_B\, f = \forall (a : A) \to
-  Q_A\,a \to Q_B\, (f\,a)}\] The predicate lifting
-$\mathsf{LType^{\wedge} : \forall (A : Set) \to (A \to Set) \to
-  LType\,A \to Set}$ for $\mathsf{LType}$ is defined by
-pattern matching by
-\[ \mathsf{LType^{\wedge}\,A\,Q_A\,(bool\,B\,e)
-  = \exists [Q_B]\, Equal^{\wedge}\, A\, B\, Q_A\, K^{Bool}_{\top}
-  \,e}\] where $\mathsf{Q_A : A \to Set}$, $\mathsf{Q_B : B \to Set}$,
-$\mathsf{K^{Bool}_{\top}}$ is the constantly $\mathsf{\top}$-valued
-predicate on $\mathsf{Bool}$, and $\mathsf{e : Equal\,A\,Bool}$;
-\[\mathsf{LType^{\wedge}\,A\,Q_A\,(arr\, B\, C\, e\, T_B\, T_C)}
-\mathsf{= \exists [Q_B] \,\exists [Q_c]\, Equal^{\wedge}\,A\,
+
+\subsection{(Deep) induction for $\mathsf{LTerm}$}\label{sec:ind-lam}
+
+\begin{figure*}[t]
+
+  \begin{adjustbox}{varwidth=6.8in, max width=6.6in, margin=-0.1in 0in
+      -0.2in 0in, fbox, center} 
+{\small
+\[\begin{array}{lll}
+\mathsf{LType^{\wedge}\,A\,Q_A\,(bool\,B\,e)} & = &\mathsf{\exists
+  [Q_B]\, Equal^{\wedge}\, A\, B\, Q_A\, K^{Bool}_{\top} \,e}\\
+\mathsf{LType^{\wedge}\,A\,Q_A\,(arr\, B\, C\, e\, T_B\, T_C)}
+&=&\mathsf{\exists [Q_B] \,\exists [Q_c]\, Equal^{\wedge}\,A\,
   (B \to C)\, Q_A\, (Arr^{\wedge} \, B\, C\, Q_B \, Q_C) \, e \times
-  \, LType^{\wedge}\,B\,Q_B\,T_B \times LType^{\wedge}\,C\,Q_C\,T_C}\]
-where $\mathsf{Q_A : A \to Set}$, $\mathsf{Q_B : B \to Set}$,
-$\mathsf{Q_C : C \to Set}$, $\mathsf{e : Equal\, A\, (B \to C)}$,
-$\mathsf{T_B : LType \,B}$ and $\mathsf{T_C : LType \,C}$; and
-\[\mathsf{LType^{\wedge}\,A\,Q_A\,(list\, B\, e\, T_B)
-=
-\exists [Q_B]\,
-  Equal^{\wedge}\,A\, (List\, B)\, Q_A\, (List^{\wedge} \, B\, Q_B) \, e 
-  \times LType^{\wedge}\,B\,Q_B\,T_B
-}
-\]
-where $\mathsf{Q_A : A \to Set}$, $\mathsf{Q_B : B \to Set}$,
-$\mathsf{e : Equal\, A\, (List\, B)}$, and $\mathsf{T_B : LType \,B}$.
-The predicate lifting $\mathsf{LTerm^{\wedge} :}$ $\mathsf{\forall (A
-  : Set) \to (A \to Set) \to LTerm\,A \to Set}$ for $\mathsf{LTerm}$
-is defined by pattern matching by
-\[\mathsf{LTerm^{\wedge}\,A\,Q_A\,(var\,s\,T_A) = LType^{\wedge}\, A\,
-  Q_A\, T_A}\] where $\mathsf{s : String}$ and $\mathsf{T_A : LType\, A}$;
-\[\mathsf{LTerm^{\wedge}\,A\,Q_A\, (abs \,B \,C \,e \,s \,T_B \,t_C)}
-\mathsf{ = \exists [Q_B]\,\exists [Q_C]\, Equal^{\wedge} \, A\, (B \to
+  \, LType^{\wedge}\,B\,Q_B\,T_B \times LType^{\wedge}\,C\,Q_C\,T_C}\\
+\mathsf{LType^{\wedge}\,A\,Q_A\,(list\, B\, e\, T_B)} & = &
+\mathsf{\exists [Q_B]\, Equal^{\wedge}\,A\, (List\, B)\, Q_A\,
+  (List^{\wedge} \, B\, Q_B) \, e \times LType^{\wedge}\,B\,Q_B\,T_B}\\[1ex]
+\mathsf{LTerm^{\wedge}\,A\,Q_A\,(var\,s\,T_A)} & = &
+\mathsf{LType^{\wedge}\, A\, Q_A\, T_A}\\
+\mathsf{LTerm^{\wedge}\,A\,Q_A\, (abs \,B \,C \,e \,s \,T_B \,t_C)} &
+= & \mathsf{\exists [Q_B]\,\exists [Q_C]\, Equal^{\wedge} \, A\, (B \to
   C)\, Q_A\, (Arr^{\wedge} \, B\, C\, Q_B \, Q_C)\, e \times \,
   LType^{\wedge}\, B\, Q_B\, T_B \times \, LTerm^{\wedge}\, C\, Q_C\,
-  t_C }\] where $\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C : C \to
-  Set}$, $\mathsf{e : Equal \, A \, (B \to C)}$, $\mathsf{s :
-  String}$, $\mathsf{T_B : LType\,B}$, and $\mathsf{t_C : LTerm \,
-  C}$;
-\[\mathsf{LTerm^{\wedge}\,A\,Q_A\, (app\, B\, t_{BA}\, t_B)
-= \exists [Q_B]\, LTerm^{\wedge}\, (B \to A)\, (Arr^{\wedge} \, B\,
-A\, Q_B \, Q_A)\, t_{BA} \times LTerm^{\wedge}\, B\, Q_B\, t_B}\]
-where $\mathsf{Q_B : B \to Set}$, $\mathsf{t_{BA} : LTerm \, (B \to
-  A)}$, and $\mathsf{t_B : LTerm \, B}$; and
-\[\mathsf{LTerm^{\wedge}\,A\,Q_A\, (list\, B\, e\, ts) =
-\exists [Q_B]\, Equal^{\wedge} \, A\, (List\,B)\, Q_A\, (List^{\wedge}
-\, B\, Q_B) \, e \times List^{\wedge}\, (LTerm\,B) \, (LTerm^{\wedge}
-\, B\, Q_B) \, ts} 
-\]
-where $\mathsf{Q_B : B \to Set}$, $\mathsf{e : Equal\, A\, (List
-  \,B)}$, $\mathsf{ts : List\, (LTerm B)}$, and $\mathsf{List^\wedge}$
-is the predicate lifting for lists from~\eqref{eq:rose}.
-
-With these liftings in hand, we can define the deep induction
-principle for $\mathsf{LTerm}$.
-% Var case 
-Let $\mathsf{dIndVar}$ be the induction hypothesis
-\begin{multline*}
-\mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A
-  \to Set)} \\ \mathsf{\to 
-  \forall (A : Set) (Q_A : A \to Set) (s : String) (T_A : LType\, A) 
-  \to LType^{\wedge} \, A\, Q_A\, T_A
-  \to P \, A\, Q_A\, (var \; s\, T_A)
-  }
-\end{multline*}
-associated with the constructor $\mathsf{var}$,
-% Abs case 
-let $\mathsf{dIndAbs}$ be the induction hypothesis
-\begin{align*}
-  &\mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A
-    \to Set)} \\ 
-  &\quad\mathsf{\to 
+  t_C }\\
+\mathsf{LTerm^{\wedge}\,A\,Q_A\, (app\, B\, t_{BA}\, t_B)} & = &
+\mathsf{\exists [Q_B]\, LTerm^{\wedge}\, (B \to A)\, (Arr^{\wedge} \,
+  B\, A\, Q_B \, Q_A)\, t_{BA} \times LTerm^{\wedge}\, B\, Q_B\,
+  t_B}\\
+\mathsf{LTerm^{\wedge}\,A\,Q_A\, (list\, B\, e\, ts)} & = &
+\mathsf{\exists [Q_B]\, Equal^{\wedge} \, A\, (List\,B)\, Q_A\,
+  (List^{\wedge} \, B\, Q_B) \, e \times List^{\wedge}\, (LTerm\,B) \,
+  (LTerm^{\wedge} \, B\, Q_B) \, ts}
+\end{array}   \]}
+
+\vspace*{-0.1in}
+
+\caption{Predicate liftings for $\mathsf{LType}$ and
+  $\mathsf{LTerm}$}\label{fig:liftings} \vspace*{0.1in} 
+\end{adjustbox}
+\end{figure*}
+
+To derive the deep induction rule for the GADT $\mathsf{LTerm}$ we use
+its Henry Ford encoding from~\eqref{eq:eq_ltype}
+and~\eqref{eq:eq_lterm}. We first define predicate lifting
+$\mathsf{Arr^{\wedge} : \forall (A\, B : Set) \to (A \to Set) \to (B
+  \to Set) \to (A \to B) \to Set}$ for arrow types, since arrow types
+appear in $\mathsf{LType}$ and $\mathsf{LTerm}$. It is given by
+$\mathsf{Arr^{\wedge}\, A\, B\, Q_A\, Q_B\, f = \forall (a : A) \to
+  Q_A\,a \to Q_B\, (f\,a)}$. The predicate liftings
+$\mathsf{LType^{\wedge} : \forall (A : Set) \to (A \to Set) \to
+  LType\,A \to Set}$ for $\mathsf{LType}$ and $\mathsf{LTerm^{\wedge}
+  :}$ $\mathsf{\forall (A : Set) \to (A \to Set) \to LTerm\,A \to
+  Set}$ for $\mathsf{LTerm}$ are defined in Figure~\ref{fig:liftings}.
+There, $\mathsf{s : String}$, $\mathsf{Q_A : A \to Set}$, $\mathsf{Q_B
+  : B \to Set}$, $\mathsf{Q_C : C \to Set}$,
+$\mathsf{K^{Bool}_{\top}}$ is the constantly {\color{red}
+  $\mathsf{\top}$}-valued predicate on $\mathsf{Bool}$, $\mathsf{T_A :
+  LType\, A}$, $\mathsf{T_B : LType \,B}$, $\mathsf{T_C : LType \,C}$,
+$\mathsf{t_B : LTerm \, B}$, $\mathsf{t_C : LTerm \, C}$, and
+$\mathsf{t_{BA} : LTerm \, (B \to A)}$.  Moreover, $\mathsf{e :
+  Equal\,A\,Bool}$ in the first clause, $\mathsf{e : Equal\, A\, (B
+  \to C)}$ in the second, $\mathsf{e : Equal\, A\, (List\, B)}$ in the
+third, $\mathsf{e : Equal \, A \, (B \to C)}$ in the fifth, and
+$\mathsf{e : Equal\, A\, (List \,B)}$, $\mathsf{ts : List\, (LTerm
+  B)}$, and $\mathsf{List^\wedge}$ is the predicate lifting for lists
+from~\eqref{eq:rose} in the seventh.
+
+With these liftings in hand we can define the induction hypotheses
+$\mathsf{dIndVar}$, $\mathsf{dIndAbs}$, $\mathsf{dIndApp}$, and
+$\mathsf{dIndList}$ associated with $\mathsf{LTerms}$'s data
+constructors. These are, respectively,
+\[\begin{array}{l}
+\mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to
+  LTerm\,A  \to Set) \to}\\
+  \quad\mathsf{\forall (A : Set) (Q_A : A \to Set) (s : String) (T_A :
+  LType\, A) \to LType^{\wedge} \, A\, Q_A\, T_A \to P \, A\, Q_A\,
+  (var \; s\, T_A)}\\[1ex]
+\mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to
+  LTerm\,A \to Set)} \\ 
+  \quad\mathsf{\to 
   \forall (A\,B\,C: Set) (Q_A : A \to Set)  (Q_B : B \to Set) (Q_C : C
   \to Set) (e : Equal\, A\, (B \to C)) (s : String) } \\ 
-  &\quad\mathsf{ \to (T_B : LType\, B) \to (t_C : LTerm\, C)
+  \quad\mathsf{ \to (T_B : LType\, B) \to (t_C : LTerm\, C)
   \to Equal^{\wedge}\,A\,(B \to C)\, Q_A \, (Arr^{\wedge} \, B\, C\,
   Q_B \, Q_C) \, e } \\
-  &\quad\mathsf{
+  \quad\mathsf{
   \to LType^{\wedge}\, B\, Q_B\, T_B
@@
   \to P \, A\, Q_A\, (abs \,B \,C \, e \,s \,T_B \, t_C)
-  }
-\end{align*}
-associated with the constructor $\mathsf{abs}$, let
-% App case 
-Let $\mathsf{dIndApp}$ be the induction hypothesis
-\begin{align*}
-  &\mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A
+  }\\[1ex]
+  \mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A
     \to Set)} \\ 
-  &\quad \mathsf{\to 
+  \quad \mathsf{\to 
   \forall (A \,B : Set) (Q_A : A \to Set)  (Q_B : B \to Set) 
    (t_{BA} : LTerm\, (B \to A)) (t_B : LTerm\, B)} \\
-  &\quad \mathsf{
+  \quad \mathsf{
   \to P\, (B \to A)\, (Arr^{\wedge} \, B\, A\, Q_B \, Q_A) \, t_{BA} \, 
   \to P\, B\, Q_B\, t_B\, 
-  \to P \, A\, Q_A\, (app \,B \,t_{BA} \, t_B) }
-\end{align*}
-associated with the constructor $\mathsf{app}$, and 
-% ListC case 
-let $\mathsf{dIndList}$ be the induction hypothesis
-\begin{align*}
-  &\mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A
+  \to P \, A\, Q_A\, (app \,B \,t_{BA} \, t_B) }\\[1ex]
+  \mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A
     \to Set)} \\ 
-  &\quad \mathsf{\to 
+  \quad \mathsf{\to 
   \forall (A \,B : Set) (Q_A : A \to Set)  (Q_B : B \to Set) 
     (e : Equal\, A\, (List\, B)) (ts : List\, (LTerm\, B))} \\ 
-  &\quad \mathsf{
+  \quad \mathsf{
     \to Equal^{\wedge}\, A\, (List\,B)\, Q_A\, (List^{\wedge}\, B\, Q_B)\, e 
@@
   \to P \, A\, Q_A\, (list \,B \,e \, ts) }
-\end{align*}
-associated with the constructor $\mathsf{list}$. The deep induction
-principle for $\mathsf{LTerm}$ is thus
+\end{array}\]
+The deep induction rule for $\mathsf{LTerm}$ is thus
 \begin{equation}\label{eq:ind-lam}
@@
 
-To prove that the deep induction principle for $\mathsf{LTerm}$ is
-sound, we must provide a witness $\mathsf{dIndLTerm}$ inhabiting the
-type in~\eqref{eq:ind-lam}. We proceed by pattern matching on
-$\mathsf{t_A}$.
-\begin{itemize}
-\item If $\mathsf{t_A = var\; s\, T_A}$, then we define
-\[ \mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clist \, A\, Q_A\,
-  (var\;s\,T_A) \, liftA = cvar \, A\, Q_A\, s\, T_A\, liftA }\] for
-$\mathsf{s : String}$ and $\mathsf{T_A : LType\,A}$.  Note that
-$\mathsf{LTerm^{\wedge}\, A\, Q_A\, (var\;s\,T_A) =
-  LType^{\wedge}\,A\,Q_A\,T_A}$, so $\mathsf{liftA :
-  LType^{\wedge}\,A\,Q_A\,T_A}$.
-% Abs case 
-\item If $\mathsf{t_A = abs \,B \,C \, e \,s \,T_B \, t_C} $, then we
-  define
-\[
-\begin{array}{l}
+\begin{figure*}[t]
+
+  \begin{adjustbox}{varwidth=6.8in, max width=6.6in, margin=-0.1in 0in
+      -0.2in 0in, fbox, center} 
+{\small
+\[\begin{array}{lll}
+\mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clist \, A\, Q_A\,
+  (var\;s\,T_A) \, liftA} & = & \mathsf{cvar \, A\, Q_A\, s\, T_A\,
+  liftA}\\ 
 \mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clist \, A\, Q_A\,
   (abs \,B \,C \,e \,s \,T_B \, t_C) \, (Q_B , Q_C , liftE,
-  lift_{T_B}, lift_{t_C})} \\ \quad \mathsf{= cabs\,A\,B\,C\, Q_A\,
-  Q_B\, Q_C\, e\, s\, T_B\, t_C\, liftE\, lift_{T_B}\, p_C}
-\end{array}
-\]
-Here, $\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C : C \to Set}$,
-$\mathsf{liftE : Equal^{\wedge} \, A\, (B \to C)\, Q_A\, (Arr^{\wedge}
-  \, B\, C\, Q_B \, Q_C) \, e}$, $\mathsf{lift_{T_B}: LType^{\wedge}
-  \, B\, Q_B\, T_B}$, and $\mathsf{lift_{t_C}: LTerm^{\wedge} \, C\,
-  Q_C\, T_C}$ --- which together ensure that $\mathsf{(Q_B , Q_C ,
-  liftE, lift_{T_B}, lift_{t_C}) : LTerm^{\wedge}\, A\, Q_A\, (abs \,B
-  \,C \, e \,s \,T_B \, t_C)}$ --- and $\mathsf{p_C =
-  dIndLTerm\,P\,cvar\,cabs \,capp \,clist\, C\, Q_C\, t_C\,
-  lift_{t_C} : P \, C\, Q_C \, t_C }$.
-% App case - (App \,b \,t_{ba} \, t_b)
-\item If $\mathsf{t_A = app \,B \,\,t_{BA} \, t_B} $, then we define
-\[ \mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clist \, A\, Q_A\,
-    (app \,B \,\,t_{BA} \, t_B)\, (Q_B , list_{t_{BA}}, list_{t_B})}
-  \mathsf{ = capp\,A\,B\,Q_A\, Q_B\, t_{BA}\, t_B\, p_{BA} \, p_B}
-\]
-Here, $\mathsf{Q_B : B \to Set}$, $\mathsf{lift_{t_{BA}}:
-  LTerm^{\wedge} \, (B \to A)\, (Arr^{\wedge} \, B\, A\, Q_B \, Q_A)\,
-  t_{BA}}$, and $\mathsf{lift_{t_B}: LTerm^{\wedge} \, B\, Q_B\, t_B}$
---- so that\\ $\mathsf{(Q_B , lift_{t_{BA}}, list_{t_B}) : 
-  LTerm^{\wedge}\, A\, Q_A\, (app \,B \,t_{BA} \, t_B)}$ --- and
-$\mathsf{p_B} = \mathsf{dIndLTerm\,P\,cvar\,cabs \,capp \,clist\, B\,
-  Q_B\, t_B\, lift_{t_B} : P \, B\, Q_B \, t_B }$ and $\mathsf{p_{BA}}
-= \mathsf{dIndLTerm\,P\,cvar\,cabs \,capp \,clist\, (B \to
-  A)\,(Arr^{\wedge} \, B\, A\, Q_B \, Q_A) \, t_{BA}\, lift_{t_{BA}} :
-  P \, (B \to A)\, (Arr^{\wedge} \, B\, A\, Q_B \, Q_A) \, t_{BA} }$.
-% ListC case - (ListC \,b \,e \, ts)
-\item If let $\mathsf{t_A = list \,B \,e \, ts}$, then we define
-\begin{multline*}
-  \mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clistc \, A\, Q_A\, (list \,B \,e \, ts) \, (Q_B , liftE, lift_{List})
-  = clistc \,A\,B\,Q_A\, Q_B\, e\, ts\, liftE\, p_{List} }
-\end{multline*}
-Here, $\mathsf{Q_B : B \to Set}$, $\mathsf{listE : Equal^{\wedge}\,
-  A\, (List\,B)\, Q_A\, (List^{\wedge}\, B\, Q_B)\, e}$, and
-$\mathsf{lift_{List}: List^{\wedge} \, (LTerm\, B) \,
-  (LTerm^{\wedge}\, B\, Q_B) \, ts}$ --- so that $\mathsf{(Q_B ,
-  liftE, lift_{List}): LTerm^{\wedge}\, A\, Q_A\, (list \,B \,e \,
-  ts)}$ --- and
-$\mathsf{p_{List} = liftListMap \, (LTerm\, B) \, (LTerm^{\wedge} \, B
-  \, Q_B)\, (P\,B\,Q_B)\, p_{ts} \, ts\, lift_{List} : }$\\
-$\mathsf{List^{\wedge}\,
-  (LTerm\,B) \, (P\,B\,Q_B) \, ts}$ and
-$\mathsf{ p_{ts} = dIndLTerm\, P\, cvar\, cabs\, capp\, clist\, B\,
-  Q_B : PredMap\,(LTerm\,B) \,(LTerm^{\wedge}\, B\, Q_B) \,
-  (P\,B\,Q_B)}$. Moreover, $\mathsf{PredMap : \forall\, (A : Set) \to
+  lift_{T_B}, lift_{t_C})} & = & \mathsf{cabs\,A\,B\,C\, Q_A\,
+  Q_B\, Q_C\, e\, s\, T_B\, t_C\, liftE\, lift_{T_B}\, p_C}\\
+\mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clist \, A\, Q_A\,
+    (app \,B \,\,t_{BA} \, t_B)\, (Q_B , list_{t_{BA}}, list_{t_B})} &
+= & \mathsf{capp\,A\,B\,Q_A\, Q_B\, t_{BA}\, t_B\, p_{BA} \, p_B}\\
+  \mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clistc \, A\, Q_A\,
+    (list \,B \,e \, ts) \, (Q_B , liftE', lift_{List})} & = & 
+  = \mathsf{clistc \,A\,B\,Q_A\, Q_B\, e\, ts\, liftE'\, p_{List} }
+\end{array}\]}
+
+\vspace*{-0.1in}
+
+\caption{$\mathsf{dIndLTerm}$}\label{fig:dindlterm} \vspace*{0.1in} 
+\end{adjustbox}
+\end{figure*}
+
+\noindent
+To prove that this rule is sound we define a witness
+$\mathsf{dIndLTerm}$ inhabiting the type in~\eqref{eq:ind-lam} as in
+Figure~\ref{fig:dindlterm}. There, $\mathsf{s : String}$, $\mathsf{Q_A
+  : A \to Set}$, $\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C : C \to
+  Set}$, $\mathsf{T_A : LType\,A}$, $\mathsf{T_B : LType\,B}$,
+$\mathsf{t_B : LTerm\,B}$, $\mathsf{t_C : LTerm\,C}$, $\mathsf{t_{BA}
+  : LTerm\,{\color{red}??}}$, $\mathsf{liftA : LTerm^{\wedge}\, A\,
+  Q_A\, (var\;s\,T_A) = LType^{\wedge}\,A\,Q_A\,T_A}$, $\mathsf{liftE
+  : Equal^{\wedge} \, A\, (B \to C)\, Q_A\, (Arr^{\wedge} \, B\, C\,
+  Q_B \, Q_C) \, e}$, $\mathsf{lift_{T_B}: LType^{\wedge} \, B\, Q_B\,
+  T_B}$,~$\mathsf{lift_{t_C}: LTerm^{\wedge} \, C\, Q_C\,
+  T_C}$,~$\mathsf{lift_{t_{BA}}: LTerm^{\wedge} \, (B \to A)\,
+  (Arr^{\wedge} \, B\, A\, Q_B \, Q_A)\,
+  t_{BA}}$,~$\mathsf{lift_{t_B}: LTerm^{\wedge} \, B\, Q_B\,
+  t_B}$,\\ $\mathsf{liftE' : Equal^{\wedge}\, A\, (List\,B)\, Q_A\,
+  (List^{\wedge}\, B\, Q_B)\, e}$, and $\mathsf{lift_{List}:
+  List^{\wedge} \, (LTerm\, B) \, (LTerm^{\wedge}\, B\, Q_B) \, ts}$.
+Moreover,
+\[\begin{array}{lll}
+\mathsf{p_C} & = & \mathsf{dIndLTerm\,P\,cvar\,cabs \,capp \,clist\,
+  C\, Q_C\, t_C\, lift_{t_C} : P \, C\, Q_C \, t_C }\\
+\mathsf{p_B} & = & \mathsf{dIndLTerm\,P\,cvar\,cabs \,capp \,clist\,
+  B\, Q_B\, t_B\, lift_{t_B} : P \, B\, Q_B \, t_B }\\
+\mathsf{p_{BA}} & = & \mathsf{dIndLTerm\,P\,cvar\,cabs \,capp
+  \,clist\, (B \to A)\,(Arr^{\wedge} \, B\, A\, Q_B \, Q_A) \,
+  t_{BA}\, lift_{t_{BA}} : P \, (B \to A)\, (Arr^{\wedge} \, B\, A\,
+  Q_B \, Q_A) \, t_{BA}}\\ 
+\mathsf{p_{List}} & = &\mathsf{liftListMap \, (LTerm\, B) \,
+  (LTerm^{\wedge} \, B \, Q_B)\, (P\,B\,Q_B)\, p_{ts} \, ts\,
+  lift_{List} : List^{\wedge}\, (LTerm\,B) \, (P\,B\,Q_B) \, ts}\\
+\mathsf{p_{ts}} & = & \mathsf{dIndLTerm\, P\, cvar\, cabs\, capp\,
+  clist\, B\, Q_B : PredMap\,(LTerm\,B) \,(LTerm^{\wedge}\, B\, Q_B)
+  \, (P\,B\,Q_B)}
+\end{array}\]
+where, in the final clause, $\mathsf{PredMap : \forall\, (A : Set) \to
   (A \to Set) \to (A \to Set) \to Set }$ is the type constructor
 producing the type of morphisms between predicates defined by
-\[\mathsf{PredMap \,A\, Q\,Q'\, = \forall\, (a : A) \to Q\,a \to
-  Q'\,a}\] and $\mathsf{liftListMap : \forall\, (A : Set) \to (Q \, Q'
+$\mathsf{PredMap \,A\, Q\,Q'\, = \forall\, (a : A) \to Q\,a \to
+  Q'\,a}$ and $\mathsf{liftListMap : \forall\, (A : Set) \to (Q \, Q'
   : A \to Set) \to PredMap\,A\,Q\,Q' \to PredMap\,(List\,A)
-  \,(List^{\wedge}\, A\, Q)\, (List^{\wedge}\, A\, Q')}$ {\color{red}
-  defined by ?????} takes a morphism $\mathsf{f}$ of predicates and
-produces a morphism of lifted predicates {\color{red} by mapping
-  $\mathsf{f}$ over the input list to which $\mathsf{Q}$ is lifted ---
-  or other simple description?}
-\end{itemize}
+  \,(List^{\wedge}\, A\, Q)\, (List^{\wedge}\, A\, Q')}$, which takes
+a morphism $\mathsf{f}$ of predicates and produces a morphism of
+lifted predicates, is defined by $\mathsf{liftListMap\, A\, Q\, Q'\,
+  m\, nil\, tt = tt}$ (since $\mathsf{x : List^{\wedge}\, A\, Q\, nil
+  = \top}$ must necessarily be $\mathsf{tt}$), and by
+$\mathsf{liftListMap\, A\, Q\, Q'\, m\, (cons\, a\, l')\, (y, x') =
+  (m\,a\,y, liftListMap\, A\, Q\, Q'\, m\, l'\, x')}$ (since $\mathsf{x
+  : List^{\wedge}\, A\, Q\, (cons\, a\, l') = \top}$ must be of the
+form $\mathsf{x = (y, x')}$ where $\mathsf{y : Q\,a}$ and $\mathsf{x'
+  : List^{\wedge}\, A\, Q\, l'}$). {\color{red} Double-check!!}
 
@@
 Surprisingly, that is not the case:
-indeed, the induction principle generally relies on (unary) parametricity of the semantic interpretation,
+indeed, the induction rule generally relies on (unary) parametricity of the semantic interpretation,
 and in the case of nested types it also relies on functorial semantics~\cite{jp20},
@@
 
-\section{Conclusion/Related work}
-
-Mention Patricia/Neil2008 paper
-
-No induction with primitive representation (reference Haskell Symposium paper and~\cite{jp19} and paper Patricia Neil Clement 2010)
+\section{Conclusion}\label{sec:conclusion}
 
@@
 \item reference inspiration for STLC GADT : https://www.seas.upenn.edu/~cis194/spring15/lectures/11-stlc.html
+\item Data type vs. data structure
+\item Weird to code in Agda if we're talking about induction rules for
+  Coq? 
+\item Agda style conventions
+\item spacing in data type declarations
+\item Mention Agda flags that need to be toggled to handle true nesting
 \end{itemize}
@@
   $\mathtt{https://doi.org/10.5281/zenodo.3744225}$
+
+\bibitem{fs18} Fu, P. and Selinger, P.  Dependently typed folds for
+  nested data types, 2018. $\mathtt{https://arxiv.org/abs/1806.05230}$
 
@@
 
+\bibitem{web-page} Johann, P., Ghiorzi, E., and Jeffries,
+  D. Accompanying Agda code for this paper.
+    $\mathtt{https://cs.appstate.edu/~johannp/FoSSaCS21Code.html}$
+
 \bibitem{jgj21f} Johann, P., Ghiorzi, E., and Jeffries, D. {\em
@@
 
+\bibitem{tas19} Tassi, E.: Deriving proved equality tests in Coq-elpi:
+  Stronger induction principles for containers in Coq. Proceedings,
+  Interactive Theorem Proving, pp. 1-18, 2019.
+
+\bibitem{vw10} Vytiniotis, D., and Weirich, S.  {\em Parametricity,
+  type equality, and higher-order polymorphism}.  Journal of
+  Functional Programming 20(2), pp.~175--210, 2010.
+
 \bibitem{xcc03} Xi, H., Chen, C. and Chen, G. {\em Guarded recursive
@@
 
-\bibitem{vw10} Vytiniotis, D., and Weirich, S.  {\em Parametricity,
-  type equality, and higher-order polymorphism}.  Journal of
-  Functional Programming 20(2), pp.~175--210, 2010.
+\bibitem{ull20} Ullrich, M. {\em Generating Induction Principles for
+Nested Induction Types in MetaCoq}. PhD thesis, Saarland University,
+  2020.
 

### 2021-04-15 induction-paper/ENTCS-GADT-induction.tex
work on Section 4, added  Enrico's work on later sections

@@
     (list \,B \,e \, ts) \, (Q_B , liftE', lift_{List})} & = & 
-  = \mathsf{clistc \,A\,B\,Q_A\, Q_B\, e\, ts\, liftE'\, p_{List} }
+  \mathsf{clistc \,A\,B\,Q_A\, Q_B\, e\, ts\, liftE'\, p_{List} }
 \end{array}\]}
@@
 
-{\color{red} HERE!!!!}
-
-%Since, as seen in Section~\ref{sec:GADT-nested},
-%GADTs featuring nesting do not admit an induction rule,
-%we restrict our scope to GADTs not featuring nesting.
-Finally, we generalize the approach taken in the previous examples and
-provide a general framework to derive deep induction rules for
-arbitrary GADTs.  For that, we need to give a grammar for the types we
-will be considering.  A generic GADT
-\begin{equation*}
-\begin{array}{l}
-\mathsf{data\ G : Set^\alpha \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;\;c_i\, :\, F_i\,G\,\ol{B} \to G (\ol{K_i\,\ol{B}})}
-\end{array}
-\end{equation*}
-is defined by a finite number of constructors $\mathsf{c_i}$.
-In the definition above, $\mathsf{F_i}$ is a type constructor with signature $\mathsf{(Set^{\alpha} \to Set) \to Set^{\beta} \to Set}$
-and each $\mathsf{K_i}$ is a type constructor with signature $\mathsf{Set^{\beta} \to Set}$
-(i.e. a type constructor of arity $\mathsf{\beta}$).
-The overline notation denotes a finite list:
-%$\mathsf{\ol{a}}$ is a list of types of length $\mathsf{\alpha}$,
-%so that it can be applied to the type constructor $\mathsf{G}$ of arity $\mathsf{\alpha}$.
-%and the vertical bars notation denotes the length of a list:
-%$\mathsf{G}$ depends on a finite list of types $\mathsf{\ol{a}}$ of length $\mathsf{\alpha}$,
-each of the $\mathsf{\alpha}$-many $\mathsf{K_i}$ is a type constructor of arity $\mathsf{\beta}$
-so that it can be applied to the list of types $\mathsf{\ol{B}}$ of length $\mathsf{\beta}$.
-%depends on a finite list of types $\mathsf{\ol{b}}$ of length $\mathsf{\beta}$,
-Moreover, notice that the arity of $\mathsf{G}$ matches the number of type constructors $\mathsf{K_i}$.
-We allow each $\mathsf{F_i}$ to be inductively built in the following ways (and with the following restrictions):
-\begin{itemize}
-\item $\mathsf{F_i = F_i' \times F_i''}$ where $\mathsf{F_i'}$ and $\mathsf{F_i''}$ have the same signature as $\mathsf{F_i}$ and are built recursively from the same induction rules.
-\item $\mathsf{F_i = F_i' + F_i''}$ where $\mathsf{F_i'}$ and $\mathsf{F_i''}$ have the same signature as $\mathsf{F_i}$ and are built recursively from the same induction rules.
-\item $\mathsf{F_i = F_i' \to F_i''}$
-where $\mathsf{F_i'}$ does not contain the recursive variable,
-i.e., $\mathsf{F_i' : Set^\beta \to Set}$ is a type constructor of arity $\mathsf{\beta}$,
-and $\mathsf{F_i''}$ has the same signature as $\mathsf{F_i}$ and is built recursively from the same induction rules.
-\item $\mathsf{F_i\,G\,\ol{B} = G(\ol{F_j\,\ol{B}})}$
-where none of the $\mathsf{\alpha}$-many $\mathsf{F_j}$ contains the recursive variable,
-i.e., $\mathsf{F_j : Set^\beta \to Set}$
-is a type constructor of arity $\mathsf{\beta}$
-for each $\mathsf{j = 1, \dots, \alpha}$.
-Such restriction is necessary to prevent nesting,
-as that would break the induction rule as discussed in Section~\ref{sec:GADT-nested}.
-\item $\mathsf{F_i\,G\,\ol{B} = H\,\ol{B}}$
-where $\mathsf{H}$ is a type constructor of arity $\mathsf{\beta}$ not containing the recursive variable,
-i.e., $\mathsf{H : Set^\beta \to Set}$.
-Notice that this covers the case in which $\mathsf{F_i}$ is a closed type,
-so, in particular, the unit and empty types, $\mathsf{1}$ and $\mathsf{0}$,
-and the case in which $\mathsf{F_i}$ is a variable.
-%The only requirement on $\mathsf{H}$ is that it needs to have a predicate-lifting function
-%(and any type made of constant types, sums, products, arrow types and other GADTs satisfies this condition).
-\item $\mathsf{F_i\,G\,\ol{B} = H (\ol{F_k\,G\,\ol{B}})}$
-where $\mathsf{H}$ is a $\mathsf{\gamma}$-ary type constructor not containing the recursive variable,
-i.e., $\mathsf{H : Set^\gamma \to Set}$,
-and $\mathsf{F_k}$ has the same signature as $\mathsf{F_i}$  and is built recursively from the same induction rules, for every $\mathsf{k = 1 \dots \gamma}$.
-Moreover, we require that $\mathsf{H}$ is not a GADT itself (but we allow it to be an ADT or even a nested type).
-This way we know that $\mathsf{H}$ admits functorial semantics~\cite{jp20}, and thus there is a map function for $\mathsf{H^{\wedge}}$,
-\[
-\mathsf{HLMap : \forall (\ol{C : Set}) (\ol{Q_C\;Q_C' : C \to Set})
-	\to \ol{PredMap\,C\,Q_C\,Q_C'}
-	\to PredMap\,(H\,\ol{C})\,(H^{\wedge}\,\ol{C}\,\ol{Q_C})\,(H^{\wedge}\,\ol{C}\,\ol{Q_C'})}
-\]
-%where $\mathsf{PredMap : \forall (C : Set) \to (C \to Set) \to (C \to Set) \to Set}$
-%is defined as
-%\[
-%\mathsf{PredMap\,C\,Q_c\,Q_c' = \forall (x : c) \to Q_c\,x \to Q_c'\,x}
-%\]
-%and represents the type of morphisms between predicates.
-A concrete way to define $\mathsf{HLMap}$ is to proceed by induction on the structure of the type $\mathsf{H}$,
-and give an inductive definition when $\mathsf{H}$ is an ADT or a nested type.
-Such details are not essential to the present discussion, and thus we omit them.
-\end{itemize}
-We can summarize the above inductive definition with the following grammar
-(but beware that the above restrictions and requirements still apply):
-\[
-\mathsf{F_i\,G\,\ol{B} :=
-F_i'\,G\,\ol{B} \times F_i''\,G\,\ol{B}
-\ \vert\ F_i'\,G\,\ol{B} + F_i''\,G\,\ol{B}
-\ \vert\ F_i'\,\ol{B} \to F_i''\,G\,\ol{B}
-\ \vert\ G(\ol{F_j\,\ol{B}})
+We can generalize the approach taken in
+Section~\ref{sec:deep-ind-GADTs} to a general framework for deriving
+deep induction rules for deep GADTs. We will treat deep GADTs of the
+form 
+\begin{equation}\label{eq:gadts}
+\begin{array}{l}
+  \mathsf{data\ G : Set^\alpha
+    \to Set\ where}\\
+\mathsf{\;\;\;\;\;\;\;\;c\, :\, F\,G\,\ol{B} \to G (\ol{K\,\ol{B}})}
+\end{array}
+\end{equation}
+For brevity and clarity we indicate only one constructor $\mathsf{c}$
+in~\eqref{eq:gadts}, even though a GADT can, in general, have any
+finite number of them, each with a type the same form as
+$\mathsf{c}$'s. In~\eqref{eq:gadts}, $\mathsf{F}$ and $\mathsf{K}$ are
+type constructors with signatures $\mathsf{(Set^{\alpha} \to Set) \to
+  Set^{\beta} \to Set}$ and $\mathsf{Set^{\beta} \to Set}$,
+respectively. If $\mathsf{T}$ has type signature $\mathsf{Set^{\gamma}
+  \to Set}$ then we say that $\mathsf{T}$ is a {\em type constructor
+  of arity $\mathsf{\gamma}$}.  The overline notation denotes a finite
+list whose length is exactly the arity of the type constructor being
+applied to it. The number of type constructors $\mathsf{K}$ must
+therefore be $\alpha$. The type constructor $\mathsf{F}$ must be
+constructed inductively according to the following grammar:
+\[\mathsf{F\,G\,\ol{B} :=
+F'\,G\,\ol{B} \times F''\,G\,\ol{B}
+\ \vert\ F'\,G\,\ol{B} + F''\,G\,\ol{B}
+\ \vert\ F'\,\ol{B} \to F''\,G\,\ol{B}
+\ \vert\ G(\ol{F'\,\ol{B}})
 \ \vert\ H\,\ol{B}
-\ \vert\ H (\ol{F_k\,G\,\ol{B}})
-}
-\]
-A further requirement that applies to all of the types appearing above,
-including the types $\mathsf{K_i}$,
-is that every type needs to have a predicate-lifting function.
-This is not an overly restrictive condition, though:
-all types made by sums, products, arrow types and type application do, and so do GADTs as defined above.
-A concrete way to define the predicate-lifting function for a type is to proceed by induction on the structure of the type,
-and we have seen in the previous sections examples of how to do so for products and type application.
-We do not give here the general definition of lifting,
-as that would require to first present a full type calculus,
-and that is beyond the scope of the paper.
-
-%Before proceeding with the derivation of the induction rule for a generic GADT,
-%we need to define the type
-%\[
-%\mathsf{PredMap\,a\,Q_a\,Q_a' = \forall (x : a) \to Q_a\,x \to Q_a'\,x}
-%\]
-%where $\mathsf{Q_a}$ and  $\mathsf{Q_a'}$ are predicates on  $\mathsf{a : Set}$,
-%which represents the type of morphisms between predicates $\mathsf{Q_a}$ and  $\mathsf{Q_a'}$.
-%Moreover, if $\mathsf{H}$ is a type constructor having an associated predicate-lifting function $\mathsf{H^{\wedge}}$,
-%we define a map function $\mathsf{HLMap}$ for the lifting $\mathsf{H^{\wedge}}$ as
-%\[
-%\mathsf{HLMap : \forall (a : Set) (Q_a\;Q_a' : a \to Set) \to PredMap\,a\,Q_a\,Q_a' \to PredMap\,(G\,a)\,(G^{\wedge}\,a\,Q_a)\,(G^{\wedge}\,a\,Q_a')}
-%\]
-
-Consider a generic GADT as defined above,
-\begin{equation}\label{eq:gadt}
-\begin{array}{l}
-\mathsf{data\ G : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;\;c\, :\, F\,G\,\ol{B} \to G (K\,\ol{B})}
-\end{array}
-\end{equation}
-which, for ease of notation,
-we assume to be a unary type constructor (i.e. it depends on a single type parameter $\mathsf{A}$)
-and to have only one constructor $\mathsf{c}$.
-Extending the argument to GADTs of arbitrary arity and with multiple constructors presents no difficulty
-other than heavier notation.
-In the definition above, $\mathsf{F}$ has signature $\mathsf{(Set \to Set) \to Set^{\beta} \to Set}$
-and each $\mathsf{K}$ has signature $\mathsf{Set^{\beta} \to Set}$.
-The constructor $\mathsf{c}$ can be rewritten using the $\mathsf{Equal}$ type as
-\[
-\mathsf{c : \forall (\ol{B : Set}) \to Equal\,A\,(K\,\ol{B}) \to F\,G\,\ol{B} \to G\,A}
-\]
+\ \vert\ H (\ol{F'\,G\,\ol{B}})
+}\]
+
+{\color{red} HERE!}  This grammar is subject to the following
+restrictions. In the third clause the type constructor $\mathsf{F'}$
+does not contain $\mathsf{G}$. In the fourth clause, none of the
+$\mathsf{\alpha}$-many type constructors in $\mathsf{\ol{F'}}$
+contains $\mathsf{G}$.  This prevents nesting, which would make it
+impossible to give an induction rule; see
+Section~\ref{sec:GADT-nested} below. In the fifth clause, $\mathsf{H}$
+does not contain $\mathsf{G}$. This subsumes the case in which
+$\mathsf{F}$ is a closed type
+%, so, in particular, the unit and empty types, $\mathsf{1}$ and
+%$\mathsf{0}$,
+{\color{red} and the case in which $\mathsf{F}$ is a variable.}  In
+the sixth clause, $\mathsf{H : Set^\gamma \to Set}$ does not contain
+$\mathsf{G}$.  Moreover, although $\mathsf{H}$ can be
+
+construct
+
+any (truly)
+nested type, it cannot
+
+construct
+
+be a GADT.  This ensures that $\mathsf{H}$
+admits functorial semantics~\cite{jp20}, and thus that
+$\mathsf{H^{\wedge}}$, admits a map function
+\[\mathsf{HLMap : \forall (\ol{A : Set}) (\ol{Q\;Q' : A \to Set}) 
+\to \ol{PredMap\,A\,Q\,Q'} \to
+PredMap\,(H\,\ol{A})\,(H^{\wedge}\,\ol{A}\,\ol{Q})\,
+(H^{\wedge}\,\ol{A}\,\ol{Q'})}\] A concrete way to define
+$\mathsf{HLMap}$ is by induction on the structure of the type
+$\mathsf{H}$.
+%, and by giving an inductive definition when $\mathsf{H}$
+%is an ADT or a nested type.
+But we omit such details since they are not essential to the present
+discussion.
+
+A further requirement that applies to all of the types appearing
+above, including the $\mathsf{K}$s, is that every type needs to have a
+predicate lifting function.  This is not an overly restrictive
+condition, though: all types made by sums, products, arrow types and
+type application do, and so do GADTs as defined above. A concrete way
+to define the predicate-lifting function for a type is to proceed by
+induction on the structure of the type, and we have seen in the
+previous sections examples of how to do so for products and type
+application.  We do not give here the general definition of lifting,
+as that would require to first present a full type calculus, and that
+is beyond the scope of the paper.
+
+For ease of notation we assume in the development below that
+$\mathsf{G}$ is a unary type constructor, i.e., that $\alpha = 1$
+in~\eqref{eq:gadts}. Extending the argument to GADTs of arbitrary
+arity presents no difficulty other than heavier notation. The
+constructor $\mathsf{c}$ there can be rewritten using the
+$\mathsf{Equal}$ type as
+\[\mathsf{c : \forall (\ol{B : Set}) \to Equal\,A\,(K\,\ol{B}) \to
+  F\,G\,\ol{B} \to G\,A}\]
 which is the form we shall use from now on.
@@
 In the previous sections,
-we derive induction rules for examples of GADTs that do not feature nesting,
+we derived induction rules for GADTs that do not feature nesting,
 in the sense that their constructors contain no nested calls of the recursive variable,
-as truly nested types (such as $\mathsf{Bush}$, Equation~\ref{eq:bush}) do.
+as truly nested types (such as $\mathsf{Bush}$, Equation~\ref{eq:bush}) do,
+or other GADTs calling the recursive variable.
 Since both nested types
 and GADTs without nesting admit induction rules,
-as seen in the previous sections,
-it is just natural to expect that GADTs with nesting would as well.
+it seems reasonable to expect that GADTs with nesting would as well.
 Surprisingly, that is not the case:
-indeed, the induction rule generally relies on (unary) parametricity of the semantic interpretation,
+indeed, the induction principle generally relies on (unary) parametricity of the semantic interpretation,
 and in the case of nested types it also relies on functorial semantics~\cite{jp20},
@@
 \begin{array}{l}
-\mathsf{data\, G\,(a : Set) : Set\,where}\\
-\mathsf{\;\;\;\;\;\;C :\, G(G\,a) \to G(a \times a)}
+\mathsf{data\ G : Set \to Set\ where}\\
+\mathsf{\;\;\;\;\;\;\;\;\;c :\, G\,(G\,A) \to G\,(A \times A)}
 \end{array}
 \end{equation}
-The constructor $\mathsf{C}$ can be rewritten as
-\[
-\mathsf{C : \exists (b : Set) \to Equal\,a\,(b \times b) \to G(G\,b) \to G\,a}
+The constructor $\mathsf{c}$ can be rewritten as
+\[
+\mathsf{c : \forall\, (B : Set) \to Equal\,A\,(B \times B) \to G\,(G\,B) \to G\,A}
 \]
@@
 \[
-\mathsf{G^{\wedge} : \forall (a : Set) \to (a \to Set) \to G\,\ol{a} \to Set}
+\mathsf{G^{\wedge} : \forall\, (A : Set) \to (A \to Set) \to G\,A \to Set}
 \]
@@
 \[
-\mathsf{G^{\wedge}\,a\,Q_a\,(C\,b\,e\,x)
-= \exists (Q_b : b \to Set)
-\to Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e
-\times G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,x}
-\]
-where $\mathsf{Q_a : a \to Set}$, $\mathsf{e : Equal\,a\,(b \times b)}$ and $\mathsf{x : G(G\,b)}$.
-Finally, let $\mathsf{CC}$ be the function
-\begin{multline*}
-\mathsf{\lambda (P : \forall (a : Set) \to (a \to Set) \to G\,a \to Set)} \\
-\mathsf{\to \forall (a\;b : Set) (Q_a : a \to Set) (Q_b : b \to Set) (e : Equal\,a\,(b \times b)) (x : G\,(G\,b))} \\
-\mathsf{\to Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e
-	\to P\,(G\,b)\,(P\,b\,Q_b)\,x
-	\to P\,a\,Q_a\,(C\,b\,e\,x)}
-\end{multline*}
-associated with the $\mathsf{C}$ constructor.
+\mathsf{G^{\wedge}\,A\,Q\,(c\,B\,e\,x)
+= \sum\, [Q' : B \to Set]\,
+Equal^{\wedge}\,A\,(B \times B)\,Q\,(Pair^{\wedge}\,B\,B\,Q'\,Q')\,e
+\times G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q')\,x}
+\]
+where $\mathsf{Q : A \to Set}$, $\mathsf{e : Equal\,A\,(B \times B)}$ and $\mathsf{x : G(G\,B)}$.
+Finally, let $\mathsf{dIndC}$ be the function
+\[
+\begin{array}{l}
+\mathsf{\lambda\, (P : \forall\, (A : Set) \to (A \to Set) \to G\,A \to Set)} \\
+\quad\mathsf{\to \forall\, (A\;B : Set)\, (Q : A \to Set)\, (Q' : B \to Set)\, (e : Equal\,A\,(B \times B))\, (x : G\,(G\,B))} \\
+\quad\mathsf{\to Equal^{\wedge}\,A\,(B \times B)\,Q\,(Pair^{\wedge}\,B\,B\,Q'\,Q')\,e
+	\to P\,(G\,B)\,(P\,B\,Q')\,x
+	\to P\,A\,Q\,(c\,B\,e\,x)}
+\end{array}
+\]
+associated to the $\mathsf{c}$ constructor.
 
 The induction rule for $\mathsf{G}$ is
-\begin{multline*}
-\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to G\,a \to Set) \to CC\,P} \\
-\mathsf{\to \forall (a : Set)(Q_a : a \to Set)(y : G\,a) \to G^{\wedge}\,a\,Q_a\,y \to P\,a\,Q_a\,y}
-\end{multline*}
+\[
+\mathsf{\forall\, (P : \forall\, (A : Set) \to (A \to Set) \to G\,A \to Set)
+\to dIndC\,P
+\to \forall\, (A : Set)\, (Q : A \to Set)\, (y : G\,a) \to G^{\wedge}\,A\,Q\,y \to P\,A\,Q\,y}
+\]
 Consistently with the previous examples,
-to validate the induction rule
-we try to define a term of the above type, $\mathsf{DIG}$, as
-\[
-\mathsf{DIG\,P\,cc\,a\,Q_a\,(C\,b\,e\,x)\,(Q_b, L_E, L_G)
-= cc\,a\,b\,Q_a\,Q_b\,e\,x\,L_E\,p}
-\]
-where $\mathsf{cc : CC\,P}$ and
+to prove the induction rule
+we try to define a term of the above type, $\mathsf{dIndG}$, as
+\[
+\mathsf{dIndG\,P\,cc\,A\,Q\,(c\,B\,e\,x)\,(Q', liftE, liftG)
+= cc\,A\,B\,Q\,Q'\,e\,x\,liftE\,p}
+\]
+where $\mathsf{cc : dIndC\,P}$,
+$\mathsf{Q : A \to Set}$ and
 \begin{itemize}
-\item $\mathsf{C\,b\,e\,x : G\,a}$,
-i.e., $\mathsf{e : Equal\,a\,(b \times b)}$ and $\mathsf{x : G(G\,b)}$;
-\item $\mathsf{(Q_b, L_E, L_G) : G^{\wedge}\,a\,Q_a\,(C\,b\,e\,x)}$,
-i.e., $\mathsf{Q_b : b \to Set}$, $\mathsf{L_E : Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e}$,
-and $\mathsf{L_G : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,x}$.
+\item $\mathsf{c\,B\,e\,x : G\,A}$,
+i.e., $\mathsf{e : Equal\,A\,(B \times B)}$ and $\mathsf{x : G\,(G\,B)}$;
+\item $\mathsf{(Q', liftE, liftG) : G^{\wedge}\,A\,Q\,(c\,B\,e\,x)}$,
+i.e., $\mathsf{Q' : B \to Set}$, $\mathsf{liftE : Equal^{\wedge}\,A\,(B \times B)\,Q\,(Pair^{\wedge}\,B\,B\,Q'\,Q')\,e}$,
+and $\mathsf{liftG : G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q')\,x}$.
 \end{itemize}
-We still need to define $\mathsf{p : P\,(G\,b)\,(P\,b\,Q_b)\,x}$.
+We still need to define $\mathsf{p : P\,(G\,B)\,(P\,B\,Q')\,x}$.
 We do so by using the induction rule and letting
 \[
-\mathsf{p = DIG\,P\,cc\,(G\,b)\,(P\,b\,Q_b)\,x\,q}
-\]
-where we still need to provide $\mathsf{q : G^{\wedge}\,(G\,b)\,(P\,b\,Q_b)\,x}$.
+\mathsf{p = dIndG\,P\,cc\,(G\,B)\,(P\,B\,Q')\,x\,q}
+\]
+where we still need to provide $\mathsf{q : G^{\wedge}\,(G\,B)\,(P\,B\,Q')\,x}$.
 If we had the map function of $\mathsf{G^{\wedge}}$,
 \[
-\mathsf{GLMap : \forall (a : Set) (Q_a\;Q_a' : a \to Set) \to PredMap\,a\,Q_a\,Q_a' \to PredMap\,(G\,a)\,(G^{\wedge}\,a\,Q_a)\,(G^{\wedge}\,a\,Q_a')}
+\mathsf{GLMap : \forall\, (A : Set)\, (Q\;Q'' : A \to Set) \to PredMap\,A\,Q\,Q'' \to PredMap\,(G\,A)\,(G^{\wedge}\,A\,Q)\,(G^{\wedge}\,A\,Q'')}
 \]
@@
 \[
-\mathsf{q = GLMap\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,(P\,b\,Q_b)\,(DIG\,P\,cc\,b\,Q_b)\,x\,L_G}
+\mathsf{q = GLMap\,(G\,B)\,(G^{\wedge}\,B\,Q')\,(P\,B\,Q')\,(dIndG\,P\,cc\,B\,Q')\,x\,liftG}
 \]
@@
 \[
-\mathsf{GLMap\,a\,Q_a\,Q_a'\,M\,(C\,b\,e\,x)\,(Q_b, L_E, L_G) = (Q_b', L_E', L_G')}
-\]
-where $\mathsf{Q_a : a \to Set}$, $\mathsf{Q_a' : a \to Set}$,
-$\mathsf{M : PredMap\,a\,Q_a\,Q_a'}$,
-$\mathsf{C\,b\,e\,x : G\,a}$, i.e.,
+\mathsf{GLMap\,A\,Q\,Q''\,m\,(c\,B\,e\,x)\,(Q', liftE, liftG) = (Q''', liftE', liftG')}
+\]
+for some $\mathsf{(Q''', liftE', liftG') : G^{\wedge}\,A\,Q''\,(c\,B\,e\,x)}$, i.e.,
 \begin{itemize}
-\item $\mathsf{e : Equal\,a\,(b \times b)}$;
-\item $\mathsf{x : G\,(G\,b)}$;
+\item $\mathsf{Q''' : B \to Set}$;
+\item $\mathsf{liftE' : Equal^{\wedge}\,A\,(B \times B)\,Q''\,(Pair^{\wedge}\,B\,B\,Q'''\,Q''')\,e}$;
+\item $\mathsf{liftG' : G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q''')\,x}$;
 \end{itemize}
-$\mathsf{(Q_b, L_E, L_G) :G^{\wedge}\,a\,Q_a\,(C\,b\,e\,x)}$, i.e.,
+where $\mathsf{Q : A \to Set}$, $\mathsf{Q'' : A \to Set}$,
+$\mathsf{m : PredMap\,A\,Q\,Q''}$,
+$\mathsf{c\,B\,e\,x : G\,A}$, i.e.,
 \begin{itemize}
-\item $\mathsf{Q_b : b \to Set}$;
-\item $\mathsf{L_E : Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e}$;
-\item $\mathsf{L_G : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,x}$;
+\item $\mathsf{e : Equal\,A\,(B \times B)}$;
+\item $\mathsf{x : G\,(G\,B)}$;
 \end{itemize}
-and $\mathsf{(Q_b', L_E', L_G') : G^{\wedge}\,a\,Q_a'\,(C\,b\,e\,x)}$, i.e.,
+and $\mathsf{(Q', liftE, liftG) :G^{\wedge}\,A\,Q\,(c\,B\,e\,x)}$, i.e.,
 \begin{itemize}
-\item $\mathsf{Q_b' : b \to Set}$;
-\item $\mathsf{L_E' : Equal^{\wedge}\,a\,(b \times b)\,Q_a'\,(Pair^{\wedge}\,b\,b\,Q'_b\,Q'_b)\,e}$;
-\item $\mathsf{L_G' : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b')\,x}$;
+\item $\mathsf{Q' : B \to Set}$;
+\item $\mathsf{liftE : Equal^{\wedge}\,A\,(B \times B)\,Q\,(Pair^{\wedge}\,B\,B\,Q'\,Q')\,e}$;
+\item $\mathsf{liftG : G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q')\,x}$.
 \end{itemize}
-%\begin{gather*}
-%\mathsf{Q_a\;Q_a' : a \to Set} \\
-%\mathsf{Q_b\;Q_b' : b \to Set} \\
-%%\mathsf{M : PredMap\,b\,Q_a\,Q_a'} \\
-%\mathsf{L_E : Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e} \\
-%\mathsf{L_G\ : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,x} \\
-%\mathsf{L_E' : Equal^{\wedge}\,a\,(b \times b)\,Q_a'\,(Pair^{\wedge}\,b\,b\,Q'_b\,Q'_b)\,e} \\
-%\mathsf{L_G' : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b')\,x}
-%\end{gather*}
-In other words, we have a proof $\mathsf{L_E}$ of the (extensional) equality
-of the predicates $\mathsf{Q_a}$ and $\mathsf{Pair^{\wedge}\,b\,b\,Q_b\,Q_b}$
-and a morphism of predicates $\mathsf{M}$ from $\mathsf{Q_a}$ to $\mathsf{Q_a'}$,
-and we need to use those to deduce a proof of the (extensional) equality
-of the predicates $\mathsf{Q_a'}$ and $\mathsf{Pair^{\wedge}\,b\,b\,Q_b'\,Q_b'}$,
-for some  for some predicate $\mathsf{Q_b'}$ on $\mathsf{b}$.
+In other words, we have a proof $\mathsf{liftE}$ of the (extensional) equality
+of the predicates $\mathsf{Q}$ and $\mathsf{Pair^{\wedge}\,B\,B\,Q'\,Q'}$
+and a morphism of predicates $\mathsf{m}$ from $\mathsf{Q}$ to $\mathsf{Q''}$,
+and we need to use those data to deduce a proof of the (extensional) equality
+of the predicates $\mathsf{Q''}$ and $\mathsf{Pair^{\wedge}\,B\,B\,Q'''\,Q'''}$,
+for some predicate $\mathsf{Q'''}$ on $\mathsf{B}$.
 But that is not generally possible:
-the facts that $\mathsf{Q_a}$ is equal to $\mathsf{Pair^{\wedge}\,b\,b\,Q_b\,Q_b}$
-and that there is a morphism of predicates $\mathsf{M}$ from $\mathsf{Q_a}$ to $\mathsf{Q_a'}$
-do not guarantee that $\mathsf{Q_a'}$ is equal to $\mathsf{Pair^{\wedge}\,bb\,Q_b'\,Q_b'}$ for some $\mathsf{Q_b'}$.
+the facts that $\mathsf{Q}$ is equal to $\mathsf{Pair^{\wedge}\,B\,B\,Q'\,Q'}$
+and that there is a morphism of predicates $\mathsf{m}$ from $\mathsf{Q}$ to $\mathsf{Q''}$
+do not guarantee that there exists a predicate $\mathsf{Q'''}$ such that $\mathsf{Q''}$ is equal to $\mathsf{Pair^{\wedge}\,B\,B\,Q'''\,Q'''}$.
 
@@
 so that having morphisms $\mathsf{A \to A'}$ and $\mathsf{B \to B'}$
+(for any type $\mathsf{A, A', B}$ and $\mathsf{B'}$)
 and a proof that $\mathsf{A}$ is equal to $\mathsf{A'}$
@@
 This is because GADTs can either have a syntax-only semantics or a functorial-completion semantics.
-Since we are interested in induction rules, we considered the syntax-only semantics,
+Since we are interested in induction rules, we consider the syntax-only semantics,
 which is parametric but not functorial.
@@
 In both cases, thus, we cannot derive an induction rule for GADTs featuring nesting.
-Unlike nestes types, indeed, GADTs do not admit a semantic interpretation that is both parametric and functorial~\cite{jgj21}.
-
-\section{Case Study: Well-typed lambda terms}\label{sec:app}
-
-{\color{red} Can get rid of Maybe using non-empty lists and postulates}
-
-In this section we use deep induction for the $\mathsf{LTerm}$ GADT to 
+Unlike nested types, indeed, GADTs do not admit a semantic interpretation that is both parametric and functorial~\cite{jgj21}.
+
+\section{Applications}
+
+%We want a proof
+%\[
+%  \mathsf{getTypeProof : \forall \, (A : Set) \to LTerm\,A \to Maybe \, (LType \, A)} \\
+%\]
+%that takes a lambda term and produces its type (using $\mathsf{Maybe}$ to represent
+%potential failure).
+%Because of the $\mathsf{listC}$ constructor of $\mathsf{LTerm}$, this cannot be achieved without 
+%deep induction. In particular, deep induction is required to apply the induction to the 
+%individual terms in a list of terms.
+
+In this section we use deep induction for the $\mathsf{LTerm}$ GADT from Example~\ref{example:lterm} to 
 extract the type from a lambda term.
-We have a predicate 
+Consider the predicate 
 \begin{align*}
-  &\mathsf{getType : \forall \, (a : Set) \to (t : LTerm\,a) \to Set} \\
-  &\mathsf{getType \,a \,t = Maybe \, (LType \, a)}
+  &\mathsf{GetType : \forall \, (A : Set) \to LTerm\,A \to Set} \\
+  &\mathsf{GetType \,A \,t = Maybe \, (LType \, A)}
 \end{align*}
-that takes a lambda term and produces its type (using $\mathsf{Maybe}$ to represent
-potential failure).
-We want to show this predicate is satisfied for every element of $\mathsf{LTerm\,a}$.
-Because of the $\mathsf{ListC}$ constructor, this cannot be achieved without 
+that takes a lambda term and produces the type of its possible types.
+This predicate uses $\mathsf{Maybe}$ to represent
+potential failure, where $\mathsf{Maybe}$ is the type defined as follows:
+\begin{equation}\label{eq:maybe}
+\begin{array}{l}
+\mathsf{data\ Maybe : Set \to Set\ where}\\
+\mathsf{\;\;\;\;\;\;nothing :\, Maybe\,A}\\
+\mathsf{\;\;\;\;\;\;\;\;\;\;\;\;just :\, A \to Maybe\,A}
+\end{array}
+\end{equation}
+We want to show this predicate is satisfied for every element of $\mathsf{LTerm\,A}$,
+i.e., we want to prove: 
+\[
+  \mathsf{getTypeProof : \forall \, (A : Set)\, (t : LTerm\,A) \to GetType \,A \,t} \\
+\]
+Because of the $\mathsf{listC}$ constructor of $\mathsf{LTerm}$, this cannot be achieved without 
 deep induction. In particular, deep induction is required to apply the induction to the 
 individual terms in a list of terms. 
-
-So, using deep induction, we want to prove: 
-\[
-  \mathsf{getTypeProof : \forall \, (a : Set) \to (t : LTerm\,a) \to getType \,a \,t} \\
-\]
-which we prove by 
-\[
-  \mathsf{getTypeProof \,a \,t = 
-    DILTerm\, (\lambda \, b\, Q_b\, t\, \to getType \,b\, t) \,
-    gtVar\,  gtABs\,  gtApp\,  gtListC\,  a\,\, K1\,\, t\, (LTerm^{\wedge}K1\, a\, t)
+So, using the deep induction rule for $\mathsf{LTerm}$ from Section~\ref{sec:dind-lterm} {\color{red} (add back-reference)},
+we define $\mathsf{getTypeProof}$ as
+\[
+  \mathsf{getTypeProof \,A \,t = 
+    dIndLTerm\, P \,
+    cvar\,  cabs\,  capp\,  clistc\,  A\, K_\top\, t\, (LTermLKT\, A\, t)
     }
 \]
-where $\mathsf{K1 : a \to Set}$ is the constantly true predicate:
-\[
-  \mathsf{K1 \, x = \top}
-\]
-and $\mathsf{LTerm^{\wedge}K1\, a\, t : LTerm^{\wedge}\, a\, K1\, t}$.
-Notice that there is no space in $\mathsf{LTerm^{\wedge}K1}$, because
-\[
-  \mathsf{LTerm^{\wedge}K1 : \forall (a : Set) (t : LTerm A) \to LTerm^{\wedge}\, a\, K1\,t}
-\]
-is a function that we will define. In addition to defining $\mathsf{LTerm^{\wedge}K1}$,
-we also have to give a proof for each constructor $\mathsf{Var, Abs, App, ListC}$:
-\begin{multline*}
-  \mathsf{gtVar : \forall (a : Set) (Q_a : a \to Set) (s : String) (T_a : LType\, a) 
-      \to LType^{\wedge} \, a\, Q_a\, T_a\, \to Maybe\, (LType\, a)}
-\end{multline*}
-
+where $\mathsf{t : LTerm\,A}$,
+$\mathsf{P}$ is the polymorphic predicate $\mathsf{\lambda \, (A: Set)\, (Q : A \to Set)\, (t : LTerm\,A)\, \to Maybe \, (LType \, A)}$
+%where $\mathsf{K : a \to Set}$ is the constantly true predicate:
+%\[
+%  \mathsf{K_\top \, x = \top}
+%\]
+and $\mathsf{LTermLKT : \forall\, (A : Set)\, (t : LTerm A) \to LTerm^{\wedge}\, A\, K_\top\,t}$
+is a function that we will define later.
+%Notice that there is no space in $\mathsf{LTermLKT}$, because
+%\[
+%  \mathsf{LTermLKT : \forall (a : Set) (t : LTerm A) \to LTerm^{\wedge}\, a\, K_\top\,t}
+%\]
+%is a function that we will define.
+In addition to defining $\mathsf{LTermLKT}$,
+we also have to give a term for each constructor of $\mathsf{LTerm}$,
+as already discussed in Section~\ref{sec:dind-lterm} {\color{red} (add back-reference)}:
+\begin{itemize}
+% var
+\item a term $\mathsf{cvar : dIndVar\,P}$ associated to the constructor $\mathsf{var}$, i.e.,
+\[
+  \mathsf{cvar : \forall\, (A : Set)\, (Q : A \to Set)\, (s : String)\, (T : LType\, A) 
+      \to LType^{\wedge} \, A\, Q\, T\, \to Maybe\, (LType\, A)}
+\]
 % abs 
-\begin{multline*}
-    \mathsf{gtAbs : \forall (a \, b\, c: Set) (Q_a : a \to Set) (Q_b : b \to Set) (Q_c : c \to Set) 
-      (e : Equal\, a\, (b \to c)) (s : String)} \\ 
-    %
-    \mathsf{ (T_b : LType\, b) (t_c : LTerm \, c)
-      \to Equal^{\wedge}\, a\, (b \to c)\, Q_a\, (Arr^{\wedge}\, b\, c\, Q_b\, Q_c)\, e
-      \to LType^{\wedge} \, b\, Q_b\, T_b} \\ 
-    %
-    \mathsf{
-      \to Maybe\, (LType\,c) \to Maybe\, (LType\, a)}
-\end{multline*}
-
+\item a term $\mathsf{cabs : dIndAbs\,P}$ associated to the constructor $\mathsf{abs}$, i.e.,
+\[
+\begin{array}{l}
+    \mathsf{cabs : \forall\, (A \, B\, C: Set)\, (Q : A \to Set)\, (Q' : B \to Set)\, (Q'' : C \to Set)} \\
+    \quad\mathsf{(e : Equal\, A\, (B \to C))\, (s : String)\, (T : LType\, B)\, (t : LTerm \, C)} \\ 
+    \quad\mathsf{ 
+      \to Equal^{\wedge}\, A\, (B \to C)\, Q\, (Arr^{\wedge}\, B\, C\, Q'\, Q'')\, e
+      \to LType^{\wedge} \, B\, Q'\, T
+      \to Maybe\, (LType\,C) \to Maybe\, (LType\, A)}
+\end{array}
+\]
 % app
-\begin{multline*}
-    \mathsf{gtApp : \forall (a \, b : Set) (Q_a : a \to Set) (Q_b : b \to Set) 
-      (t_{ba} : LTerm\, (b \to a)) (t_b : LTerm\, b)} \\ 
-    \mathsf{
-      \to Maybe\, (LType\, (b \to a))
-      \to Maybe\, (LType\, b)
-      \to Maybe\, (LType\, a)
+\item a term $\mathsf{capp : dIndApp\,P}$ associated to the constructor $\mathsf{app}$, i.e.,
+\[
+\begin{array}{l}
+    \mathsf{capp : \forall\, (A \, B : Set)\, (Q : A \to Set)\, (Q' : B \to Set)\, 
+      (t : LTerm\, (B \to A))\, (t' : LTerm\, B)} \\ 
+    \quad\mathsf{
+      \to Maybe\, (LType\, (B \to A))
+      \to Maybe\, (LType\, B)
+      \to Maybe\, (LType\, A)
     }
-\end{multline*}
-
+\end{array}
+\]
 % list 
-\begin{multline*}
-    % abs
-    \mathsf{gtListC : \forall (a \, b : Set) (Q_a : a \to Set) (Q_b : b \to Set) 
-      (e : Equal\, a\, (List\, b)) (ts : List\, (LTerm\, b))} \\ 
-    \mathsf{
-      \to Equal^{\wedge}\, a\, (List\,b) \, Q_a\, (List^{\wedge}\, b\, Q_b)\, e
-      \to List^{\wedge}\, (LTerm\, b)\, (getType\, b)\, ts
-      \to Maybe\, (LType\, a)
+\item a term $\mathsf{clistc : dIndListC\,P}$ associated to the constructor $\mathsf{listC}$, i.e.,
+\[
+\begin{array}{l}
+    \mathsf{clistc : \forall\, (A \, B : Set)\, (Q : A \to Set)\, (Q' : B \to Set)\,
+      (e : Equal\, A\, (List\, B))\, (ts : List\, (LTerm\, B))} \\ 
+    \quad\mathsf{
+      \to Equal^{\wedge}\, A\, (List\,B) \, Q\, (List^{\wedge}\, B\, Q')\, e
+      \to List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, ts
+      \to Maybe\, (LType\, A)
     }
-\end{multline*}
-
-For variables we simply return the type $\mathsf{T_a}$, 
+\end{array}
+\]
+\end{itemize}
+For variables we simply return the type $\mathsf{T : LType\, A}$ (wrapped in $\mathsf{just}$), 
 and the cases for abstraction and application are similar.
-The interesting case is $\mathsf{gtListC}$, in which we have to use 
-the results of ($\mathsf{List^{\wedge}\, (LTerm\, b)\, (getType\, b)\, ts}$)
+The interesting case is $\mathsf{clistc}$, in which we have to use 
+the results of $\mathsf{List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, ts}$
 in order to extract the type of one of the terms in the list. 
-To define $\mathsf{gtListC}$ we pattern-match on the list of terms $\mathsf{ts}$.
-
-If $\mathsf{ts}$ is the empty list (denoted by $\mathsf{[]}$), we cannot extract a type, 
-so we return $\mathsf{nothing}$. \\
-{\color{red} Maybe handle this case differently, by using non-empty lists, for example}
-\[
-  \mathsf{gtListC\, a\, b\, Q_a\, Q_b\, e\, [] \, L_e L_{ts} = nothing}
-\]
-If $\mathsf{ts}$ is a non-empty list, we pattern match on $\mathsf{L_{ts}}$ and use the result
-to construct the type we need:
+To define $\mathsf{clistc}$ we pattern-match on the list of terms $\mathsf{ts}$.
+If $\mathsf{ts}$ is the empty list $\mathsf{nil}$, we cannot extract a type, 
+so we return $\mathsf{nothing}$:
+\[
+  \mathsf{clistc\, A\, B\, Q\, Q'\, e\, nil \, liftE\, lift_{ts} = nothing}
+\]
+where $\mathsf{liftE : Equal^{\wedge}\, A\, (List\,B) \, Q\, (List^{\wedge}\, B\, Q')\, e}$
+and $\mathsf{lift_{ts} : List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, ts}$.
+If $\mathsf{ts}$ is a non-empty list $\mathsf{cons\,t\,ts'}$,
+we pattern match on $\mathsf{list_{ts}}$ and use the result to construct the type we need.
+The type of $\mathsf{list_{ts}}$ is
 \begin{align*}
-  &\mathsf{gtListC\, a\, b\, Q_a\, Q_b\, e\, (t :: ts) \, L_e (nothing , L_{ts}) = nothing} \\
-  &\mathsf{gtListC\, a\, b\, Q_a\, Q_b\, e\, (t :: ts) \, L_e (just\, T_b , L_{ts}) = just \, (TList\, b\, e\, T_b)}
+\mathsf{List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, (cons\, t\, ts')}
+&= \mathsf{GetType\, B\, t \times List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, ts'} \\
+&= \mathsf{Maybe\, (LType\, B) \times List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, ts'}
 \end{align*}
-where $\mathsf{e : Equal\, a\, (List\,b)}$ and $\mathsf{T_b : LType\, b}$.
-
-\subsection{Defining $\mathsf{LTerm^{\wedge}K1}$}
-
-{\color{red} Maybe this section can be deleted by assuming $\mathsf{LTerm^{\wedge}\, a\, K1\,t = K1}$.
+Thus, we define
+\begin{align*}
+  &\mathsf{clistc\, A\, B\, Q\, Q'\, e\, (cons\, t\, ts') \, liftE\, (nothing , lift_{ts'}) = nothing} \\
+  &\mathsf{clistc\, A\, B\, Q\, Q'\, e\, (cons\, t\, ts') \, liftE\, (just\, T' , lift_{ts'}) = just \, (TList\, B\, e\, T')}
+\end{align*}
+where $\mathsf{e : Equal\, A\, (List\,B)}$, $\mathsf{T' : LType\, B}$
+and $\mathsf{lift_{ts'} : List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, ts'}$.
+
+\subsection{Defining $\mathsf{LTermLKT}$}
+
+{\color{red} Maybe this section can be deleted by assuming $\mathsf{LTerm^{\wedge}\, A\, K_\top\,t = K_\top}$.
 Maybe we can say this derives from parametricity.
-Currently we say that $\mathsf{K1 \times K1 = K1}$ based on the fact that products are a built-in type and so this seems to be obviously true.}
-
-The last piece of infrastructure we need to define $\mathsf{getTypeProof}$ is a function 
-\[
-  \mathsf{LTerm^{\wedge}K1 : \forall (a : Set) \to (t : LTerm A) \to LTerm^{\wedge}\, a\, K1\,t}
+Currently we say that $\mathsf{K_\top \times K_\top = K_\top}$ based on the fact that products are a built-in type and so this seems to be obviously true.}
+
+The last piece of infrastructure we need in order to define $\mathsf{getTypeProof}$ is a function 
+\[
+  \mathsf{LTermLKT : \forall\, (A : Set)\, (t : LTerm\, A) \to LTerm^{\wedge}\, A\, K_\top\,t}
 \]
 that provides a 
-proof of $\mathsf{LTerm^{\wedge}\,a\,K1\,t}$ for any term $\mathsf{t : LTerm\, a}$. 
+proof of $\mathsf{LTerm^{\wedge}\,A\,K_\top\,t}$ for any term $\mathsf{t : LTerm\, A}$. 
 Because $\mathsf{LTerm^{\wedge}}$ is defined in terms of $\mathsf{LType^{\wedge}}$, $\mathsf{Arr^{\wedge}}$, 
-and $\mathsf{List^{\wedge}}$, we will need analogous functions for these liftings as well. 
-We only give the definition of $\mathsf{LTerm^{\wedge}K1}$, but the definitions for 
-$\mathsf{LType^{\wedge}}$, $\mathsf{Arr^{\wedge}}$, and $\mathsf{List^{\wedge}}$ are analogous.
-
-$\mathsf{LTerm^{\wedge}K1}$ is defined by pattern matching on the lambda term $\mathsf{t}$. 
-For the $\mathsf{Var}$ case, let $\mathsf{t = (Var\, s\, T_a)}$ and define
-\[
-  \mathsf{LTerm^{\wedge}K1\,a\,(Var\,s\,T_a) = LType^{\wedge}K1\,a\,T_a}
-\]
-For the $\mathsf{Abs}$ case, let $\mathsf{t = (Abs \,b \,c \, e \,s \,T_b \, t_c)} $ and recall the 
-definition of $\mathsf{LTerm^{\wedge}}$ for the $\mathsf{Abs}$ constructor, 
-instantiating the predicate $\mathsf{Q_a}$ to $\mathsf{K1}$: 
+and $\mathsf{List^{\wedge}}$, we need an analogous function for each of these liftings as well
+(respectively, $\mathsf{LTypeLKT}$, $\mathsf{ArrLKT}$ and $\mathsf{ListLKT}$). 
+We only give the definition of $\mathsf{LTermLKT}$, but the definitions for 
+$\mathsf{LTypeLKT}$, $\mathsf{ArrLKT}$ and $\mathsf{ListLKT}$ are analogous. 
+We define $\mathsf{LTermLKT}$ by pattern matching on the lambda term $\mathsf{t}$:
+\begin{itemize}
+\item For the $\mathsf{var}$ case, let $\mathsf{t = var\, s\, T}$ for $\mathsf{s : String}$ and $\mathsf{T : LType\,A}$,
+and define
+\[
+  \mathsf{LTermLKT\,A\,(var\,s\,T) = LTypeLKT\,A\,T}
+\]
+\item For the $\mathsf{abs}$ case, let $\mathsf{t = abs \,B \,C \, e \,s \,T \, t'} $
+for $\mathsf{e : Equal\,A\,(B \to C)}$, $\mathsf{s : String}$, $\mathsf{T : LType\,B}$ and $\mathsf{t' : LTerm\,C}$,
+and recall the definition of $\mathsf{LTerm^{\wedge}}$ for the $\mathsf{abs}$ constructor, 
+instantiating the predicate $\mathsf{Q : A \to Set}$ to $\mathsf{K_\top}$: 
 % \,a\,Q_a\, (Abs \,b \,c \,e \,s \,T_b \,t_c)}$, :
-\begin{multline*}
-  \mathsf{LTerm^{\wedge}\,a\,K1\,
-  (Abs \,b \,c \,e \,s \,T_b \,t_c)} \\ 
-\mathsf{
-  = \exists (Q_b : b \to Set) (Q_c : c \to Set) \to
-      Equal^{\wedge} \, a\, (b \to c)\, K1\, (Arr^{\wedge} \, b\, c\, Q_b \, Q_c)\, e} \\ 
-\mathsf{
-      \times \, LType^{\wedge}\, b\, Q_b\, T_b
-      \times \, LTerm^{\wedge}\, c\, Q_c\, t_c }
-\end{multline*}
-so to define the $\mathsf{Abs}$ case of $\mathsf{LTerm^{\wedge}K1}$, we need a proof of 
-\[
-  \mathsf{Equal^{\wedge} \, a\, (b \to c)\, K1\, (Arr^{\wedge} \, b\, c\, Q_b \, Q_c)\, e}
-\]
-i.e., that $\mathsf{K1}$ is (extensionally) equal to the lifting $\mathsf{(Arr^{\wedge} \, b\, c\, Q_b \, Q_c)}$
-for some predicates $\mathsf{Q_b, Q_c}$. The only reasonable choice for $\mathsf{Q_b}$ and $\mathsf{Q_c}$ 
-is to let both be $\mathsf{K1}$, which means we need a proof of:
-\[
-  \mathsf{Equal^{\wedge} \, a\, (b \to c)\, K1\, (Arr^{\wedge} \, b\, c\, K1 \, K1)\, e}
-\]
-Since we are working with proof-relevant predicates (i.e., functions into $\mathsf{Set}$ rather than 
-functions into $\mathsf{Bool}$), 
-the lifting $\mathsf{(Arr^{\wedge} \, b\, c\, K1 \, K1)}$ of $\mathsf{K1}$ to arrow types
-is not identical to $\mathsf{K1}$ on arrow types, but the predicates are (extensionally) isomorphic. 
+\[
+\begin{array}{l}
+\mathsf{LTerm^{\wedge}\,A\,K_\top\,
+  (abs \,B \,C \,e \,s \,T \,t')} \\ 
+\quad\mathsf{
+  = \sum\, [Q' : B \to Set]\, [Q'' : C \to Set]\,
+      Equal^{\wedge} \, A\, (B \to C)\, K_\top\, (Arr^{\wedge} \, B\, C\, Q' \, Q'')\, e
+      \times \, LType^{\wedge}\, B\, Q'\, T
+      \times \, LTerm^{\wedge}\, C\, Q''\, t' }
+\end{array}
+\]
+So, to define the $\mathsf{abs}$ case of $\mathsf{LTermLKT}$, we need a proof of 
+\[
+  \mathsf{Equal^{\wedge} \, A\, (B \to C)\, K_\top\, (Arr^{\wedge} \, B\, C\, Q' \, Q'')\, e}
+\]
+i.e., that $\mathsf{K_\top}$ is (extensionally) equal to the lifting $\mathsf{Arr^{\wedge} \, B\, C\, Q' \, Q''}$
+for some predicates $\mathsf{Q' : B \to Set}$ and $\mathsf{Q'' : C \to Set}$.
+The only reasonable choice for $\mathsf{Q'}$ and $\mathsf{Q''}$ 
+is to let both be $\mathsf{K_\top}$, which means we need a proof of
+\[
+  \mathsf{Equal^{\wedge} \, A\, (B \to C)\, K_\top\, (Arr^{\wedge} \, B\, C\, K_\top \, K_\top)\, e}
+\]
+Since we are working with proof-relevant predicates
+(i.e., functions into $\mathsf{Set}$ rather than functions into $\mathsf{Bool}$), 
+the lifting $\mathsf{Arr^{\wedge} \, B\, C\, K_\top \, K_\top}$ of $\mathsf{K_\top}$ to arrow types
+is not identical to $\mathsf{K_\top}$ on arrow types, but the predicates are (extensionally) isomorphic. 
 We discuss this issue in more detail at the end of the section. 
@@
 \[
-  \mathsf{Equal^{\wedge}ArrK1 : Equal^{\wedge} \, a\, (b \to c)\, K1\, (Arr^{\wedge} \, b\, c\, K1 \, K1)\, e}
-\]
-and define the $\mathsf{Abs}$ case of $\mathsf{LTerm^{\wedge}K1}$ as 
-\[
-  \mathsf{LTerm^{\wedge}K1\,a\, (Abs \,b \,c \, e \,s \,T_b \, t_c) = 
-    (K1 , K1 , Equal^{\wedge}ArrK1 , LType^{\wedge}K1 \,b\,T_b\, , LTerm^{\wedge}K1\,c\,t_c)
+  \mathsf{EqualLArrKT : Equal^{\wedge} \, A\, (B \to C)\, K_\top\, (Arr^{\wedge} \, B\, C\, K_\top \, K_\top)\, e}
+\]
+and define the $\mathsf{abs}$ case of $\mathsf{LTermLKT}$ as 
+\[
+  \mathsf{LTermLKT\,A\, (abs\, B \,C \, e \,s \,T \, t') = 
+    (K_\top , K_\top , EqualLArrKT , LTypeLKT\, B\, T , LTermLKT\, C\, t')
   }
 \]
-For the $\mathsf{App}$ case, let $\mathsf{t = (App \,b \,\,t_{ba} \, t_b)}$ and just as we did for the $\mathsf{Abs}$ case, 
-recall the definition of $\mathsf{LTerm^{\wedge}\,a\,(App \,b \,\,t_{ba} \, t_b)}$ with all of the predicates
-instantiated with $\mathsf{K1}$: 
+\item For the $\mathsf{app}$ case, let $\mathsf{t = app\, B\, t' \, t''}$
+for $\mathsf{t' : LTerm\,(B \to A)}$ and $\mathsf{t'' : LTerm\,B}$,
+and, just as we did for the $\mathsf{abs}$ case, 
+recall the definition of $\mathsf{LTerm^{\wedge}\, A\, K_\top\, (app\, B\, t'\, t'')}$
+with all of the predicates instantiated with $\mathsf{K_\top}$: 
 \[
 \mathsf{
-      LTerm^{\wedge}\, (b \to a)\, (Arr^{\wedge} \, b\, a\, K1 \, K1)\, t_{ba}
-      \times LTerm^{\wedge}\, b\, K1\, t_b }
-\]
-The second component can be given using $\mathsf{LTerm^{\wedge}K1}$, and we can 
+      LTerm^{\wedge}\, (B \to A)\, (Arr^{\wedge} \, B\, A\, K_\top \, K_\top)\, t'
+      \times LTerm^{\wedge}\, B\, K_\top\, t''}
+\]
+The second component can be given using $\mathsf{LTermLKT}$, and we can 
 define the first component using a proof of
 \[
-\mathsf{LTerm^{\wedge}\, (b \to a)\, K1\,\, t_{ba}}
+\mathsf{LTerm^{\wedge}\, (B \to A)\, K_\top\, t'}
 \]
 and a map-like function 
-\begin{multline*}
-  \mathsf{LTerm^{\wedge}EqualMap : \forall\, \{a : Set\} \to (Q_a\,Q'_a : a \to Set) 
-    \to (Equal^{\wedge}\,a\,a\,Q_a\,Q'_a\,rfl)} \\
-    \mathsf{
-    \to PredMap\, (LTerm^{\wedge}\,a\,Q_a)\,(LTerm^{\wedge}\,a\,Q'_a)}
-\end{multline*}
+\[
+  \mathsf{LTermLEqualMap : \forall\, (A : Set)\, (Q\,Q' : A \to Set) 
+    \to Equal^{\wedge}\,A\,A\,Q\,Q'\,refl
+    \to PredMap\,(LTerm\,A)\, (LTerm^{\wedge}\,A\,Q)\,(LTerm^{\wedge}\,A\,Q')}
+\]
 that takes two (extensionally) equal predicates with the same carrier 
@@
 straightforward enough, so we omit the details. 
-But it is worth noting that while a true $\mathsf{HLMap}$ function, 
-which takes a \emph{morphism} of predicates instead of a proof of equality, cannot be defined 
-for GADTs in general, an analogue of $\mathsf{LTerm^{\wedge}EqualMap}$ should be definable 
-for every GADT. These analogues of $\mathsf{LTerm^{\wedge}EqualMap}$ will be required to define 
-$\mathsf{G^{\wedge}K1}$ whenever $\mathsf{G}$ has a constructor of the form
-  ($\mathsf{c : G \, (F\, b) \to G\,(K\,b)}$).
-
-Using $\mathsf{LTerm^{\wedge}EqualMap}$, we can define the $\mathsf{App}$ case of $\mathsf{LTerm^{\wedge}K1}$ as 
-\[
-  \mathsf{LTerm^{\wedge}K1\,a\, (App \,b \,\,t_{ba} \, t_b) = 
-    (K1 , L_{Arr^{\wedge}\,K1} , LTerm^{\wedge}K1\,b\,t_b)}
-\]
-where $\mathsf{ L_{Arr^{\wedge}\,K1} : LTerm^{\wedge}\, (b \to a)\, (Arr^{\wedge} \, b\, a\, K1 \, K1)\, t_{ba}}$ is defined as
-\[
-  \mathsf{L_{Arr^{\wedge}\,K1} =  LTerm^{\wedge}EqualMap\,K1\,(Arr^{\wedge}\,b\,a\,K1\,K1)\,\, Equal^{\wedge}ArrK1\, \,\,t_{ba}\, \,L_{K1}}
-\]
-where $\mathsf{L_{K1} = LTerm^{\wedge}K1\, (b \to a)\, t_{ba} : LTerm^{\wedge}\, (b \to a) \, K1\, t_{ba}}$.
-
-Finally, we define the $\mathsf{ListC}$ case for $\mathsf{LTerm^{\wedge}K1}$. 
-Let $\mathsf{t = (ListC\, b\, e\, ts)}$ and recall the definition of $\mathsf{LTerm^{\wedge}\,a\,(ListC\, b\, e\, ts)}$ with
-all of the predicates instantiated to $\mathsf{K1}$: 
-\[
-\mathsf{ Equal^{\wedge} \, a\, (List\,b)\, K1\, (List^{\wedge} \, b\, K1) \, e
-      \times List^{\wedge}\, (LTerm\,b) \, (LTerm^{\wedge} \, b\, K1) \, ts }
+Using $\mathsf{LTermLEqualMap}$, we can define the $\mathsf{app}$ case of $\mathsf{LTermLKT}$ as 
+\[
+  \mathsf{LTermLKT\,A\, (app\, B\, t'\, t'') = 
+    (K_\top , LTermLArr , LTermLKT\,B\,t'')}
+\]
+where $\mathsf{ LTermLArr : LTerm^{\wedge}\, (B \to A)\, (Arr^{\wedge} \, B\, A\, K_\top \, K_\top)\, t'}$ is defined as
+\[
+  \mathsf{LTermLArr =  LTermLEqualMap\,K_\top\,(Arr^{\wedge}\,B\,A\,K_\top\,K_\top)\,\, EqualLArrKT\, t'\, L_{K_\top}}
+\]
+where $\mathsf{L_{K_\top} = LTermLKT\, (B \to A)\, t' : LTerm^{\wedge}\, (B \to A) \, K_\top\, t'}$.
+\item For the $\mathsf{listC}$ case, let $\mathsf{t = listC\, B\, e\, ts}$
+for $\mathsf{e : Equal\,A\,(List\,B)}$ and $\mathsf{ts : List\,(LTerm\, B)}$,
+and recall the definition of $\mathsf{LTerm^{\wedge}\,A\,(listC\, B\, e\, ts)}$
+with all of the predicates instantiated to $\mathsf{K_\top}$: 
+\[
+\mathsf{ Equal^{\wedge} \, A\, (List\,B)\, K_\top\, (List^{\wedge} \, B\, K_\top) \, e
+      \times List^{\wedge}\, (LTerm\,B) \, (LTerm^{\wedge} \, B\, K_\top) \, ts }
 \]
 We can give the first component by assuming a proof \,
-$\mathsf{Equal^{\wedge}ListK1 : Equal^{\wedge} \, a\, (List\,b)\, K1\, (List^{\wedge} \, b\, K1) \, e}$, 
-but for the second component we again have multiple liftings nested together. In this case, we can 
-get a proof of
-\[
-\mathsf{ List^{\wedge}\, (LTerm\,b) \, (LTerm^{\wedge} \, b\, K1) \, ts }
-\]
-using 
-\[\mathsf{List^{\wedge}map : \forall\, (a : Set) \to (Q_a \, Q'_a : a \to Set) 
-  \to PredMap\,Q_a\,Q'_a \to PredMap \,(List^{\wedge}\, a\, Q_a)\, (List^{\wedge}\, a\, Q'_a)}\]
+$\mathsf{EqualLListKT : Equal^{\wedge} \, A\, (List\,B)\, K_\top\, (List^{\wedge} \, B\, K_\top) \, e}$, 
+but for the second component we again have multiple liftings nested together.
+In this case, we can get a proof of
+\[
+\mathsf{ List^{\wedge}\, (LTerm\,B) \, (LTerm^{\wedge} \, B\, K_\top) \, ts }
+\]
+using $\mathsf{liftListMap}$, as seen in Section~\ref{sec:dind-lterm} {(\color{red} add back-reference)},
 to map a morphism of predicates 
 \[
-  \mathsf{PredMap \, (K1)\, (LTerm^{\wedge}\,b\,K1)}
+  \mathsf{PredMap\, (LTerm\,B)\, (K_\top)\, (LTerm^{\wedge}\,B\,K_\top)}
 \]
@@
 \[
-  \mathsf{PredMap \,(List^{\wedge}\,(LTerm\,b)\,K1) \, (List^{\wedge}\,(LTerm\,b)\, (LTerm^{\wedge}\,b\,K1))}
-\]
-
-We define the $\mathsf{ListC}$ case of $\mathsf{LTerm^{\wedge}K1}$ as 
-\[
-  \mathsf{LTerm^{\wedge}K1\,a\, (ListC\, b\, e\, ts) =
-    (K1 , \, Equal^{\wedge}ListK1 , \,  L_{List^{\wedge}LTerm^{\wedge}K1} ) }
+  \mathsf{PredMap\,(List\,(LTerm\,B)) \,(List^{\wedge}\,(LTerm\,B)\,K_\top) \, (List^{\wedge}\,(LTerm\,B)\, (LTerm^{\wedge}\,B\,K_\top))}
+\]
+We then define the $\mathsf{listC}$ case of $\mathsf{LTermLKT}$ as 
+\[
+  \mathsf{LTermLKT\,A\, (listC\, B\, e\, ts) =
+    (K_\top , \, EqualLListKT , \,  L_{ListLLTermLKT} ) }
 \]
 where
- $\mathsf{L_{List^{\wedge}LTerm^{\wedge}K1} : List^{\wedge}\, (LTerm\,b) \, (LTerm^{\wedge} \, b\, K1) \, ts}$
+$\mathsf{L_{ListLLTermLKT} : List^{\wedge}\, (LTerm\,B) \, (LTerm^{\wedge} \, B\, K_\top) \, ts}$
+is defined as
 \[
 \mathsf{ 
- L_{List^{\wedge}LTerm^{\wedge}K1} =
-  List^{\wedge}map \, (LTerm\,b) \, K1 \, (LTerm^{\wedge}\, b\, K1) \, m_{K1} \, ts \, (List^{\wedge}K1\, (LTerm\, b)\, ts)
+  L_{ListLLTermLKT} =
+  liftListMap \, (LTerm\,B) \, K_\top \, (LTerm^{\wedge}\, B\, K_\top) \, m_{K_\top} \, ts \, (ListLKT\, (LTerm\, B)\, ts)
 }
 \]
-and 
- $\mathsf{m_{K1} : PredMap \, (K1)\, (LTerm^{\wedge}\,b\,K1)}$
-\[
-  \mathsf{m_{K1} \, t\, \ast\, = LTerm^{\wedge}K1\, b\, t}
-\]
-where $\mathsf{\ast}$ is the single element of $\mathsf{(K1\, t)}$. 
-The use of $\mathsf{List^{\wedge}map}$ is required in the $\mathsf{ListC}$ case because 
-$\mathsf{ListC}$ takes an argument of type $\mathsf{List\, (LTerm\, b)}$. 
-The same technique can be used to define $\mathsf{G^{\wedge}K1}$ whenever $\mathsf{G}$ has 
-a constructor of the form ($\mathsf{c : F \, (G\, a) \to G\, (K\,b)}$). 
-We only allow constructors of this form when 
-$\mathsf{F}$ is a nested type or ADT, so we are guaranteed to have a $\mathsf{F^{\wedge}map}$ function.
-
-\subsection{Liftings of $\mathsf{K1}$}
-To provide a proof of $\mathsf{G^{\wedge}\, a\, K1 \, t}$ for every term $\mathsf{t : G\, a}$, 
-we need to know that the lifting of $\mathsf{K1}$ by a type $\mathsf{H}$ is extensionally equal to $\mathsf{K1}$ on $\mathsf{H}$. 
-For example, we might need a proof that $\mathsf{Pair^{\wedge}\,a\,b\,K1\,K1}$ is equal to the predicate $\mathsf{K1}$ on pairs. 
-Given a pair $\mathsf{(x , y) : a \times b}$, we have
-
-\begin{align*}
-  &\mathsf{Pair^{\wedge}\,a\,b\,K1\,K1 (x, y)}  \\
-  = \, \, &\mathsf{K1 \, x \times K1\, y} \\ 
-  = \, \, &\mathsf{\top \times \top}
-\end{align*}
-while 
-\begin{align*}
- \mathsf{K1\, (x, y) = \top}
-\end{align*}
-and while these types are not equal they are clearly isomorphic. So for simplicity of presentation, we 
-assume ($\mathsf{F^{\wedge}\,a\,K1}$) is equal to $\mathsf{K1}$ for every nested type and ADT $\mathsf{F}$.
-
-%\section{Primitive representation for GADTs}
+and $\mathsf{m_{K_\top} : PredMap\, (LTerm\,B) \, (K_\top)\, (LTerm^{\wedge}\,B\,K_\top)}$
+is defined as
+\[
+  \mathsf{m_{K_\top} \, t'\, tt\, = LTermLKT\, B\, t'}
+\]
+where $\mathsf{t' : LTerm\,B}$ and $\mathsf{tt}$ is the single element of $\mathsf{K_\top\, t'}$. 
+The use of $\mathsf{liftListMap}$ is required in the $\mathsf{listC}$ case because 
+$\mathsf{listC}$ takes an argument of type $\mathsf{List\, (LTerm\, B)}$.
+\end{itemize}
+
+{\color{red} These are general considerations, and are not required to conclude the above argument.}
+The above techniques can be used
+to define a function $\mathsf{GLKT : \forall\, (A : Set)\, (x : G\,A) \to G^{\wedge}\, A\, K_\top\, x}$
+for an arbitrary GADT $\mathsf{G}$,
+as defined in Section~\ref{sec:dind-generic} {\color{red} (add back-reference)}.
+To provide a proof of $\mathsf{G^{\wedge}\, A\, K_\top \, x}$ for every term $\mathsf{x : G\, A}$, 
+we need to know that the lifting of $\mathsf{K_\top}$ by any type constructor $\mathsf{F}$
+is extensionally equal to $\mathsf{K_\top}$ on $\mathsf{F}$. 
+For example, we might need a proof that $\mathsf{Pair^{\wedge}\,A\,B\,K_\top\,K_\top}$
+is equal to the predicate $\mathsf{K_\top}$ on $\mathsf{A \times B}$. 
+Given a pair $\mathsf{(a , b) : A \times B}$, we have
+$\mathsf{Pair^{\wedge}\,A\,B\,K_\top\,K_\top (a, b) = K_\top \, a \times K_\top\, b = \top \times \top}$,
+whereas $\mathsf{K_\top\, (a, b) = \top}$.
+While these types are not equal, they are clearly isomorphic.
+So, for simplicity of presentation, we assume $\mathsf{F^{\wedge}\,A\,K_\top}$
+is equal to $\mathsf{K_\top}$ for every nested type and ADT $\mathsf{F}$. 
+Moreover, remember that, whenever $\mathsf{G}$ has a constructor of the form $\mathsf{c : F \, (G\, A) \to G\, (K\,B)}$, 
+$\mathsf{F}$ is only allowed to be a nested type or an ADT,
+and we are guaranteed to have a $\mathsf{liftFMap}$ function.
 

### 2021-04-19 induction-paper/ENTCS-GADT-induction.tex
Finished through Section 6 mostly

@@
 
-{\bf Related Work\/}
-Various techniques for deriving induction rules for data types that go
-beyond ADTs have been studied. For example, Fu and
-Selinger~\cite{fs18} show, via examples, how to derive induction rules
-for arbitrary nested types. Unfortunately, however, their technique is
-rather {\em ad hoc}, so is unclear how to generalize it to nested
-types other than the specific ones in the examples. Moreover, it
-actually derives induction rules for data types {\em related} to the
-original nested types rather than for the original nested types
-themselves, and it is unclear whether or not the derived rules are
-sufficiently expressive to prove all results about the original nested
-types that we would expect to be provable by induction. This latter
-point echoes the issue with Coq-derived induction rule for rose trees
-raised in Section~\ref{sec:intro}, which has the unfortunate effect of
-forcing users to manually write induction (and other) rules for such
-types for use in that system. Tassi~\cite{tas19} has done exactly
-that, deriving induction rules for data type definitions in Coq using
-unary parametricity. Tassi's technique seems to be essentially
-equivalent to that of~\cite{jp19} for nested types, although he does
-not permit true nesting. More recently, Ulrich~\cite{ull20} has
-implemented a plugin in MetaCoq to generate induction rules for nested
-types. This plugin is also based on unary parametricity and, again,
-true nesting is not permitted.  {\color{red} As far as we know, no
-  attempt has (yet) been made to extend either implementation to
-  GADTs.} In fact, we know of no work other than that reported here
-that specifically addresses induction rules for (deep) GADTs.
+{\bf Related Work\/} Various techniques for deriving induction rules
+for data types that go beyond ADTs have been studied. For example, Fu
+and Selinger~\cite{fs18} show, via examples, how to derive induction
+rules for arbitrary nested types. Unfortunately, however, their
+technique is rather {\em ad hoc}, so is unclear how to generalize it
+to nested types other than the specific ones in the
+examples. Moreover, it actually derives induction rules for data types
+{\em related} to the original nested types rather than for the
+original nested types themselves, and it is unclear whether or not the
+derived rules are sufficiently expressive to prove all results about
+the original nested types that we would expect to be provable by
+induction. This latter point echoes the issue with Coq-derived
+induction rule for rose trees raised in Section~\ref{sec:intro}, which
+has the unfortunate effect of forcing users to manually write
+induction (and other) rules for such types for use in that
+system. Tassi~\cite{tas19} has done exactly that, deriving induction
+rules for data type definitions in Coq using unary
+parametricity. Tassi's technique seems to be essentially equivalent to
+that of~\cite{jp19} for nested types, although he does not permit true
+nesting. More recently, Ulrich~\cite{ull20} has implemented a plugin
+in MetaCoq to generate induction rules for nested types. This plugin
+is also based on unary parametricity and, again, true nesting is not
+permitted.  As far as we know, no attempt has yet been made to extend
+either implementation to GADTs. In fact, we know of no work other than
+that reported here that specifically addresses induction rules for
+(deep) GADTs.
 
@@
 type $\mathsf{A}$ and to a predicate on $\mathsf{G\,A}$. Such a
-predicate lifting
-$\mathsf{Equal^{\wedge} : \forall (A\,B : Set) \to (A \to Set) \to (B
-  \to Set) \to Equal\,A\,B \to Set}$
-for $\mathsf{Equal}$ is defined by
-\[\mathsf{Equal^{\wedge}\,A\,A\,Q\,Q'\,refl = \forall (a : A) \to
-  Equal\,(Q\,a)(Q'\,a)}\]
-which takes two predicates on the same type as input and tests them
-for extensional equality.
-
+predicate lifting $\mathsf{Equal^{\wedge} : \forall (A\,B : Set) \to
+  (A \to Set) \to (B \to Set) \to Equal\,A\,B \to Set}$ for
+$\mathsf{Equal}$ is defined by
+$\mathsf{Equal^{\wedge}\,A\,A\,Q\,Q'\,refl = \forall (a : A) \to
+  Equal\,(Q\,a)(Q'\,a)}$.  It takes two predicates on the same type as
+input and tests them for extensional equality.
+%
 Next, we need to associate with each data constructor $\mathsf{c}$ of
@@
   }\\[1ex]
+\end{array}\]
+\[\begin{array}{l}
   \mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A
@@
 
-\noindent
 To prove that this rule is sound we define a witness
@@
 form
+
+\vspace*{-0.05in}
+
 \begin{equation}\label{eq:gadts}
@@
 \end{equation}
+
+\vspace*{-0.05in}
+
+\noindent
 For brevity and clarity we indicate only one constructor $\mathsf{c}$
@@
 constructed inductively according to the following grammar:
-\[\mathsf{F\,G\,\ol{B} :=
-F'\,G\,\ol{B} \times F''\,G\,\ol{B}
-\ \vert\ F'\,G\,\ol{B} + F''\,G\,\ol{B}
-\ \vert\ F'\,\ol{B} \to F''\,G\,\ol{B}
-\ \vert\ G(\ol{F'\,\ol{B}})
-\ \vert\ H\,\ol{B}
-\ \vert\ H (\ol{F'\,G\,\ol{B}})
-}\]
-
-{\color{red} HERE!}  This grammar is subject to the following
-restrictions. In the third clause the type constructor $\mathsf{F'}$
+
+\vspace*{-0.05in}
+
+\[\mathsf{F\,G\,\ol{B}\; :=\;
+F_1\,G\,\ol{B} \times F_2\,G\,\ol{B} \ \vert\ F_1\,G\,\ol{B} +
+F_2\,G\,\ol{B} \ \vert\ F_1\,\ol{B} \to F_2\,G\,\ol{B}
+\ \vert\ G\,(\ol{F_1\,\ol{B}}) \ \vert\ H\,\ol{B} \ \vert\ H\,
+(\ol{F_1\,G\,\ol{B}}) }\]
+This grammar is subject to the following
+restrictions. In the third clause the type constructor $\mathsf{F_1}$
 does not contain $\mathsf{G}$. In the fourth clause, none of the
-$\mathsf{\alpha}$-many type constructors in $\mathsf{\ol{F'}}$
+$\mathsf{\alpha}$-many type constructors in $\mathsf{\ol{F_1}}$
 contains $\mathsf{G}$.  This prevents nesting, which would make it
@@
 Section~\ref{sec:GADT-nested} below. In the fifth clause, $\mathsf{H}$
-does not contain $\mathsf{G}$. This subsumes the case in which
-$\mathsf{F}$ is a closed type
-%, so, in particular, the unit and empty types, $\mathsf{1}$ and
-%$\mathsf{0}$,
-{\color{red} and the case in which $\mathsf{F}$ is a variable.}  In
-the sixth clause, $\mathsf{H : Set^\gamma \to Set}$ does not contain
-$\mathsf{G}$.  Moreover, although $\mathsf{H}$ can be
-
-construct
-
-any (truly)
-nested type, it cannot
-
-construct
-
-be a GADT.  This ensures that $\mathsf{H}$
-admits functorial semantics~\cite{jp20}, and thus that
-$\mathsf{H^{\wedge}}$, admits a map function
-\[\mathsf{HLMap : \forall (\ol{A : Set}) (\ol{Q\;Q' : A \to Set}) 
+does not contain $\mathsf{G}$. This clause therefore subsumes the
+cases in which $\mathsf{F\,G\,\ol{B}}$ is a closed type or one of the
+$\mathsf{B_i}$. In the sixth clause, $\mathsf{H : Set^\gamma \to Set}$
+does not contain $\mathsf{G}$ (although $\ol{\mathsf{F_1\,G\,\ol{B}}}$
+can). Moreover, although $\mathsf{H}$ can construct any (truly)
+nested type, it cannot construct a GADT. This ensures that
+$\mathsf{H}$ admits functorial semantics~\cite{jp20}, and thus has an
+associated $\mathsf{map}$ function. From this we can also construct a
+$\mathsf{map}$ function {\color{red} Say what PredMap is.}
+
+\vspace*{-0.05in}
+
+\[\mathsf{H^\wedge Map : \forall (\ol{A : Set}) (\ol{Q\;Q' : A \to Set}) 
 \to \ol{PredMap\,A\,Q\,Q'} \to
 PredMap\,(H\,\ol{A})\,(H^{\wedge}\,\ol{A}\,\ol{Q})\, 
-(H^{\wedge}\,\ol{A}\,\ol{Q'})}\] A concrete way to define
-$\mathsf{HLMap}$ is by induction on the structure of the type
-$\mathsf{H}$.
-%, and by giving an inductive definition when $\mathsf{H}$
-%is an ADT or a nested type.
-But we omit such details since they are not essential to the present
-discussion.
-
-A further requirement that applies to all of the types appearing
-above, including the $\mathsf{K}$s, is that every type needs to have a
-predicate lifting function.  This is not an overly restrictive
-condition, though: all types made by sums, products, arrow types and
-type application do, and so do GADTs as defined above. A concrete way
-to define the predicate-lifting function for a type is to proceed by
-induction on the structure of the type, and we have seen in the
-previous sections examples of how to do so for products and type
-application.  We do not give here the general definition of lifting,
-as that would require to first present a full type calculus, and that
-is beyond the scope of the paper.
-
-For ease of notation we assume in the development below that
-$\mathsf{G}$ is a unary type constructor, i.e., that $\alpha = 1$
-in~\eqref{eq:gadts}. Extending the argument to GADTs of arbitrary
-arity presents no difficulty other than heavier notation. The
-constructor $\mathsf{c}$ there can be rewritten using the
-$\mathsf{Equal}$ type as
-\[\mathsf{c : \forall (\ol{B : Set}) \to Equal\,A\,(K\,\ol{B}) \to
-  F\,G\,\ol{B} \to G\,A}\]
-which is the form we shall use from now on.
-
-In order to state the induction rule for $\mathsf{G}$,
-we first need to define $\mathsf{G}$'s associated predicate-lifting function 
-\[
-\mathsf{G^{\wedge} : \forall (A : Set) \to (A \to Set) \to G\,A \to Set}
-\]
-as
+(H^{\wedge}\,\ol{A}\,\ol{Q'})}\]
+for $\mathsf{H^{\wedge}}$.  A
+concrete way to define $\mathsf{H^\wedge Map}$ is by induction on the
+structure of the type $\mathsf{H}$, but we omit such details since
+they are not essential to the present discussion. A further
+requirement that applies to all of the types appearing above,
+including the $\mathsf{K}$s in~\eqref{eq:gadts}, is that they must all
+admit predicate liftings. This is not an overly restrictive condition,
+though: all types constructed from sums, products, arrow types and
+type application admit predicate liftings, and so do GADTs constructed
+from the above grammar; in fact, we have seen such liftings for
+products and type application in Section~\ref{sec:deep-ind-GADTs}.  A
+concrete way to define predicate liftings more generally is, again, by
+induction on the the structure of the types. We do not give a general
+definition of predicate liftings here, though, since that would
+require us to first design a full type calculus, which is beyond the
+scope of the present paper.
+
+We assume in the development below that $\mathsf{G}$ is a unary type
+constructor, i.e., that $\alpha = 1$ in~\eqref{eq:gadts}. Extending
+the argument to GADTs of arbitrary arity presents no difficulty other
+than heavier notation. The type of the single data constructor
+$\mathsf{c}$ for $\mathsf{G}$ can be rewritten as $\mathsf{c : \forall
+  (\ol{B : Set}) \to Equal\,A\,(K\,\ol{B}) \to F\,G\,\ol{B} \to
+  G\,A}$. The predicate lifting $\mathsf{G^{\wedge} : \forall (A :
+  Set) \to (A \to Set) \to G\,A \to Set}$ for $\mathsf{G}$ is
+therefore
 \[
 \mathsf{G^{\wedge}\,A\,Q_A\,(c\,\ol{B}\,e\,x)
-= \sum [\ol{Q_B : B \to Set}]\,
+= \exists [\ol{Q_B}]\,
 Equal^{\wedge}\,A\,(K\,\ol{B})\,Q_A\,(K^{\wedge}\,\ol{B}\,\ol{Q_B})\,e
@@
 \]
-where $\mathsf{Q_A : A \to Set}$
-and $\mathsf{c\,\ol{B}\,e\,x : G\,A}$,
-i.e., $\mathsf{e : Equal\,A\,(K\,\ol{B})}$ and $\mathsf{x : F\,G\,\ol{B}}$.
-As already mentioned before, we also assume to have liftings for $\mathsf{F}$,
-\[
-\mathsf{F^{\wedge} : \forall (G : Set^{\alpha} \to Set) (\ol{B : Set})
+where $\mathsf{Q_A : A \to Set}$, $\ol{\mathsf{Q_B : B \to Set}}$,
+$\mathsf{e : Equal\,A\,(K\,\ol{B})}$, and $\mathsf{x : F\,G\,\ol{B}}$.
+Assuming predicate liftings 
+\[\begin{array}{lll}
+\mathsf{F^{\wedge}} & : & \mathsf{\forall (G : Set^{\alpha} \to Set) (\ol{B : Set})
 \to (\forall (A : Set) \to (A \to Set) \to G\,A \to Set)
-\to (\ol{B \to Set})
-\to F\,G\,\ol{B} \to Set}
-\]
-and for  $\mathsf{K}$,
-\[
-\mathsf{K^{\wedge} : \forall (\ol{B : Set}) \to (\ol{B \to Set}) \to K\,\ol{B} \to Set}
-\]
-Finally, associate the function
-\[
-\begin{array}{l}
-\mathsf{dIndC = \lambda (P : \forall (A : Set) \to (A \to Set) \to G\,A \to Set)} \\
-\quad \mathsf{\to \forall (A : Set) (\ol{B : Set}) (Q_A : A \to Set) (\ol{Q_B : B \to Set}) (e : Equal\,A\,(K\,\ol{B})) (x : F\,G\,\ol{B})} \\
-\quad \mathsf{\to Equal^{\wedge}\,A\,(K\,\ol{B})\,Q_A\,(K^{\wedge}\,\ol{B}\,\ol{Q_B})\,e
-	\to F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}\,x
-	\to P\,A\,Q_A\,(c\,\ol{B}\,e\,x)}
-\end{array}
-\]
-with the constructor $\mathsf{c}$.
-
-%We can now define the induction rule for $\mathsf{G}$
-The induction rule for $\mathsf{G}$ is
-\[
-\mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to G\,A \to Set) \to dIndC\,P
-%\mathsf{\to \Big(
-%\forall (\ol{a : Set}) (\ol{b : Set}) (\ol{Q_a : a \to Set}) (\ol{Q_b : b \to Set}) (\ol{e_a : Equal\,a\,(K\,\ol{b})}) (x : F\,G\,\ol{b})} \\
-%\mathsf{\to \ol{Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
-%	\to F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x
-%	\to P\,\ol{a}\,\ol{Q_a}\,(C\,\ol{b}\,\ol{e_a}\,x)
-%\Big)} \\
-\to \forall (A : Set)(Q_A : A \to Set)(y : G\,A) \to G^{\wedge}\,A\,Q_A\,y \to P\,A\,Q_A\,y}
-\]
-As we already did in the previous examples,
-we validate the induction rule by providing a term $\mathsf{dIndG}$ for the type above.
-Define
-\[
-\mathsf{dIndG\,P\,cc\,A\,Q_A\,(c\,\ol{B}\,e\,x)\,(\ol{Q_B}, liftE, liftF)
-= cc\,A\,\ol{B}\,Q_A\,\ol{Q_B}\,e\,x\,liftE\,(p\,x\,liftF)}
-\]
-where $\mathsf{cc : dIndC\,P}$ and
-\begin{itemize}
-\item $\mathsf{c\,\ol{B}\,e\,x : G\,A}$,
-i.e., $\mathsf{e : Equal\,A\,(K\,\ol{B})}$,
-and $\mathsf{x : F\,G\,\ol{B}}$;
-\item $\mathsf{(\ol{Q_B}, liftE, liftF) : G^{\wedge}\,A\,Q_A (c\,\ol{B}\,e\,x)}$,
-i.e.,
-$\mathsf{Q_B : B \to Set}$ for each $\mathsf{B}$,
-$\mathsf{liftE : Equal^{\wedge}\,A\,(K\,\ol{B})\,Q_A\,(K^{\wedge}\,\ol{B}\,\ol{Q_B})\,e}$,
-and $\mathsf{liftF : F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,{\ol{Q_B}}\,x}$.
-\end{itemize}
+\to (\ol{B \to Set}) \to F\,G\,\ol{B} \to Set}\\ 
+\mathsf{K^{\wedge}} & : & \mathsf{\forall (\ol{B : Set}) \to (\ol{B \to Set}) \to
+  K\,\ol{B} \to Set}
+\end{array}\]
+for $\mathsf{F}$ and for $\mathsf{K}$, respectively, the induction
+hypothesis for $\mathsf{c}$ is
+\[\begin{array}{l}
+\mathsf{dIndC = \lambda (P : \forall (A : Set) \to (A \to Set) \to
+  G\,A \to Set)}\\
+\quad \mathsf{\to \forall (A : Set) (\ol{B : Set}) (Q_A : A \to Set)
+  (\ol{Q_B : B \to Set}) (e : Equal\,A\,(K\,\ol{B})) (x :
+  F\,G\,\ol{B})} \\ 
+\quad \mathsf{\to
+  Equal^{\wedge}\,A\,(K\,\ol{B})\,Q_A\,(K^{\wedge}\,\ol{B}\,\ol{Q_B})\,e
+  \to F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}\,x \to
+  P\,A\,Q_A\,(c\,\ol{B}\,e\,x)} 
+\end{array}\]
+and the induction rule for $\mathsf{G}$ is
+\[
+\mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to G\,A \to
+  Set) \to dIndC\,P \to \forall (A : Set)(Q_A : A \to Set)(y : G\,A)
+\to G^{\wedge}\,A\,Q_A\,y \to P\,A\,Q_A\,y} \]
+
+To prove that this rule is sound we define the witness
+$\mathsf{dIndG}$ inhabiting this type by
+\[\mathsf{dIndG\,P\,cc\,A\,Q_A\,(c\,\ol{B}\,e\,x)\,(\ol{Q_B}, liftE, liftF)
+= cc\,A\,\ol{B}\,Q_A\,\ol{Q_B}\,e\,x\,liftE\,(p\,x\,liftF)}\] Here,
+$\mathsf{cc : dIndC\,P}$, $\mathsf{e : Equal\,A\,(K\,\ol{B})}$,
+$\mathsf{x : F\,G\,\ol{B}}$, $\ol{\mathsf{Q_B : B \to Set}}$,
+$\mathsf{liftE : Equal^{\wedge}\,A\,(K\,\ol{B})\,Q_A\,
+  (K^{\wedge}\,\ol{B}\,\ol{Q_B})\,e}$, and $\mathsf{liftF :
+  F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,{\ol{Q_B}}\,x}$. As a result,
+$\mathsf{(\ol{Q_B}, liftE, liftF) : G^{\wedge}\,A\,Q_A
+  (c\,\ol{B}\,e\,x)}$ as expected. 
 Finally, the morphism of predicates
-\[
-\mathsf{p : PredMap\,(F\,G\,\ol{B})\,(F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}) (F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}
-\]
+$\mathsf{p :
+  PredMap\,(F\,G\,\ol{B})\,(F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B})
+  (F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}$ 
 is defined by structural induction on $\mathsf{F}$ as follows:
 \begin{itemize}
-\item Case $\mathsf{F = F_1 \times F_2}$ where $\mathsf{F_1}$ and $\mathsf{F_2}$ have the same signature as $\mathsf{F}$.
-We have that
-\[
-\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}
-= Pair^{\wedge}(F_1\,G\,\ol{B})(F_2\,G\,\ol{B})(F_1^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}) (F_2^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}
-\]
-By inductive hypothesis, there exist morphisms of predicates
-\begin{gather*}
-\mathsf{p_1 : PredMap\,(F_1\,G\,\ol{B})\,((F_1)^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}) ((F_1)^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})} \\
-\mathsf{p_2 : PredMap\,(F_2\,G\,\ol{B})\,((F_2)^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}) ((F_2)^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}
-\end{gather*}
-Thus, we define $\mathsf{p (x_1, x_2) (liftF_1, liftF_2) = (p_1\,x_1\,liftF_1, p_2\,x_2\,liftF_2)}$ 
-for 
-$\mathsf{x_1 : F_1\,G\,\ol{B}}$,
-$\mathsf{liftF_1 : F_1^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}\,x_1}$,
-$\mathsf{x_2 : F_2\,G\,\ol{B}}$
-and $\mathsf{liftF_2 : F_2^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}\,x_2}$.
-%%%
-\item Case $\mathsf{F = F_1 + F_2}$ where $\mathsf{F_1}$ and $\mathsf{F_2}$ have the same signature as $\mathsf{F}$.
-Analogous to case $\mathsf{F = F_1 \times F_2}$.
-%%%
-\item Case $\mathsf{F = F_1 \to F_2}$
-where $\mathsf{F_1}$ does not contain the recursive variable,
-i.e., $\mathsf{F_1 : Set^\beta \to Set}$,
-and $\mathsf{F_2}$ has the same signature as $\mathsf{F}$.
-We have that
-\[
-\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}\,x = \forall (z : F_1\,\ol{B}) \to F_1^{\wedge}\,\ol{B}\,\ol{Q_B}\,z \to F_2^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}\,(x\,z)}
-\]
-where $\mathsf{x : F\,G\,\ol{B} = F_1\,\ol{B} \to F_2\,G\,\ol{B}}$.
-By inductive hypothesis, there exist a morphism of predicates
-\[
-\mathsf{p_2 : PredMap\,(F_2\,G\,\ol{B}) (F_2^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}) (F_2^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}
-\]
-Thus, we define $\mathsf{p\,x\,liftF : F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}\,x}$
-for $\mathsf{liftF : F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}\,x}$
-as $\mathsf{p\,x\,liftF\,z\,liftF_1 = p_2 (x\,z) (liftF\,z\,liftF_1)}$
-for $\mathsf{z : F_1\,\ol{B}}$
-and $\mathsf{liftF_1 : F_1^{\wedge}\,\ol{B}\,\ol{Q_B}\,z}$.
-Notice that $\mathsf{F_1}$ not containing the recursive variable
-is a necessary restriction, as the proof relies on
-$\mathsf{F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}\,x}$
-and $\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}\,x}$
-having the same domain $\mathsf{F_1^{\wedge}\,\ol{B}\,\ol{Q_B}\,z}$.
-%%%
-\item Case $\mathsf{F\,G\,\ol{B} = G(F'\,\ol{B})}$
-where $\mathsf{F'}$ does not contain the recursive variable,
-i.e., $\mathsf{F' : Set^\beta \to Set}$.
-Thus, $\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B} = P (F'\,\ol{B}) (F'^{\wedge}\,\ol{B}\,\ol{Q_B})}$.
-So, $\mathsf{p}$ is defined as
-\[
-\mathsf{p = dIndG\,P\,cc\,(F'\,\ol{B})\,(F'^{\wedge}\,\ol{B}\,\ol{Q_B})}
-\]
-%for $\mathsf{L_F : F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x}$.
-%Note that the fact that the type constructor $\mathsf{F'}$ does not contain the recursive variable
-%means that $\mathsf{G}$ does not feature nesting.
-%As explained and exemplified in Section~\ref{sec:GADT-nested},
-%that assumption is required to make induction work.
-%%%
-\item Case $\mathsf{F\,G\,\ol{B} = H\,\ol{B}}$
-where $\mathsf{H}$ is a $\mathsf{\beta}$-ary type constructor not containing the recursive variable,
-i.e., $\mathsf{H : Set^\beta \to Set}$.
-In such case, $\mathsf{p : PredMap (H\,\ol{B}) (H^{\wedge}\,\ol{B}\,\ol{Q_B}) (H^{\wedge}\,\ol{B}\,\ol{Q_B})}$
-is just the identity morphism of predicates.
-%%%
-\item Case $\mathsf{F\,G\,\ol{B} = H (\ol{F_k\,G\,\ol{B}})}$
-where $\mathsf{H}$ is a $\mathsf{\gamma}$-ary type constructor not containing the recursive variable,
-i.e., $\mathsf{H : Set^\gamma \to Set}$,
-and $\mathsf{F_k}$ has the same signature as $\mathsf{F}$, for every $\mathsf{k = 1 \dots \gamma}$.
-Moreover, we assume that $\mathsf{H}$ has an associated predicate-lifting function,
-\[
-\mathsf{H^{\wedge} : \forall (\ol{C : Set}) \to (\ol{C \to Set}) \to H\,\ol{C} \to Set}
-\]
-and that this predicate-lifting function has a map function $\mathsf{HLMap}$ of type
-\[
-\mathsf{\forall (\ol{C : Set}) (\ol{Q_C\;Q_C' : C \to Set}) \to PredMap\,\ol{C}\,\ol{Q_C}\,\ol{Q_C'}
-\to PredMap\,(H\,\ol{C})\,(H^{\wedge}\,\ol{C}\,\ol{Q_C})\,(H^{\wedge}\,\ol{C}\,\ol{Q_C'})}
-\]
-That means that $\mathsf{H}$ cannot be a GADT,
-as GADTs have no functorial semantics~\cite{jgj21}.
-and incur in the issue exposed in Section~\ref{sec:GADT-nested},
-but it can be an ADT or even a nested type as those types have functorial semantics~\cite{jp19,jgj21}.
-Thus,
-\[
-\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B} = H^{\wedge} (\ol{F_k\,G\,\ol{B}}) (\ol{F_k^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}})}
-\]
-By induction hypothesis, there is a morphism of predicates
-\[
-\mathsf{p_k : PredMap\,(F_k\,G\,\ol{B}) (F_k^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}) (F_k^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}
-\]
-for every $\mathsf{k = 1 \dots \gamma}$.
-So, $\mathsf{p}$ is defined as
-\[
-\mathsf{p
-= HLMap\,(\ol{F_k\,G\,\ol{B}}) (\ol{F_k^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}}) (\ol{F_k^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}})\,\ol{p_k}}
-\]
+\item 
+If $\mathsf{F\,G\,\ol{B} = F_1\,G\,\ol{B} \times F_2\,G\,\ol{B}}$ then
+$\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B} =
+  Pair^{\wedge}\,(F_1\,G\,\ol{B})\, (F_2\,G\,\ol{B})\,
+  (F_1^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})\,
+  (F_2^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}$. The induction hypothesis
+ensures morphisms of predicates $\mathsf{p_1 :
+  PredMap\,(F_1\,G\,\ol{B})\,(F_1^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_BQ})
+  (F_1^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}$ and $\mathsf{p_2 :
+  PredMap\,(F_2\,G\,\ol{B})\,(F_2^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B})
+  (F_2^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}$.  For $\mathsf{x_1 :
+  F_1\,G\,\ol{B}}$, $\mathsf{liftF_1 :
+  F_1^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}\,x_1}$, $\mathsf{x_2 :
+  F_2\,G\,\ol{B}}$ and $\mathsf{liftF_2 :
+  F_2^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}\,x_2}$ we therefore
+define $\mathsf{p\, (x_1, x_2)\, (liftF_1, liftF_2) =
+  (p_1\,x_1\,liftF_1,\, p_2\,x_2\,liftF_2)}$.
+\item The case $\mathsf{F\,G\,\ol{B} = F_1\,G\,\ol{B} +
+  F_2\,G\,\ol{B}}$ is analogous.
+\item If $\mathsf{F\,G\,\ol{B} = F_1\,\ol{B} \to F_2\,G\,\ol{B}}$ then
+  $\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}\,x = \forall (z :
+  F_1\,\ol{B}) \to F_1^{\wedge}\,\ol{B}\,\ol{Q_B}\,z \to
+  F_2^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}\,(x\,z)}$, where $\mathsf{x :
+  F\,G\,\ol{B}}$. The induction hypothesis ensures a morphism of
+  predicates $\mathsf{p_2 : PredMap\,(F_2\,G\,\ol{B})\,
+    (F_2^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B})\,
+    (F_2^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}$.  We define
+  $\mathsf{p\,x\,liftF : F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}\,x}$,
+  where $\mathsf{liftF :
+    F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}\,x}$ is defined by
+  $\mathsf{p\,x\,liftF\,z\,liftF_1 = p_2 (x\,z) (liftF\,z\,liftF_1)}$
+  for $\mathsf{z : F_1\,\ol{B}}$ and $\mathsf{liftF_1 :
+    F_1^{\wedge}\,\ol{B}\,\ol{Q_B}\,z}$. Note that $\mathsf{F_1}$ not
+  containing the recursive variable is a necessary restriction, since
+  the proof relies on
+  $\mathsf{F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}\,x}$ and
+  $\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}\,x}$ having the same
+  domain $\mathsf{F_1^{\wedge}\,\ol{B}\,\ol{Q_B}\,z}$.
+\item If $\mathsf{F\,G\,\ol{B} = G\,(F_1\,\ol{B})}$ and $\mathsf{F_1}$
+  does not contain $\mathsf{G}$, then
+  $\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B} = P (F_1\,\ol{B})
+    (F_1^{\wedge}\,\ol{B}\,\ol{Q_B})}$. We therefore define $\mathsf{p
+    = dIndG\,P\, cc\,(F_1\,\ol{B})\,(F_1^{\wedge}\,\ol{B}\,\ol{Q_B})}$.
+\item If $\mathsf{F\,G\,\ol{B} = H\,\ol{B}}$ and $\mathsf{H}$ does not
+  contain $\mathsf{G}$, then $\mathsf{p : PredMap\, (H\,\ol{B})\,
+    (H^{\wedge}\,\ol{B}\,\ol{Q_B})\, (H^{\wedge}\,\ol{B}\,\ol{Q_B})}$
+  is defined to be the identity morphism on predicates.
+\item If $\mathsf{F\,G\,\ol{B} = H\, (\ol{F_k\,G\,\ol{B}})}$, if
+  $\mathsf{H}$ does not contain $\mathsf{G}$, if $\mathsf{H}$ has a
+  predicate lifting $\mathsf{H^{\wedge} : \forall (\ol{C : Set}) \to
+    (\ol{C \to Set})}$ $\mathsf{\to H\,\ol{C} \to Set}$, and if
+  $\mathsf{H^\wedge}$ has a $\mathsf{map}$ function $\mathsf{H^\wedge
+    Map = \forall (\ol{C : Set}) (\ol{Q_C\;Q_C' : C \to Set}) \to
+    PredMap\,\ol{C}\,\ol{Q_C}\,\ol{Q_C'} \to}$
+  $\mathsf{PredMap\,(H\,\ol{C})\,(H^{\wedge}\,\ol{C}\,\ol{Q_C})\,
+    (H^{\wedge}\,\ol{C}\,\ol{Q_C'})}$, then since $\mathsf{H}$ cannot
+  be a GADT, $\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B} =
+    H^{\wedge}\, (\ol{F_k\,G\,\ol{B}})\,
+    (\ol{F_k^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}})}$.  The induction
+  hypothesis ensures morphisms of predicates $\ol{\mathsf{p_k :
+      PredMap\,(F_k\,G\,\ol{B})
+      (F_k^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B})
+      (F_k^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}}$. We therefore define
+  $\mathsf{p = H^\wedge Map\,(\ol{F_k\,G\,\ol{B}})\,
+    (\ol{F_k^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}})\,
+    (\ol{F_k^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}})\,\ol{p_k}}$.
 \end{itemize}
-%Notice that, by proceeding by structural induction as above,
-%we have implicitly defined a grammar for GADTs admitting an induction rule.
-%In particular, we allow $\mathsf{F}$ to feature sums, products, arrow types and (limited) type application.
-
-\section{{\color{red}Induction for GADTs with nesting}}\label{sec:GADT-nested}
-
-%%% Change title: something about the inherent limitations of the method
-%%% we might want to extend previous result
-%%% but we cannot so we already have the best possible result
-
-In the previous sections,
-we derived induction rules for GADTs that do not feature nesting,
-in the sense that their constructors contain no nested calls of the recursive variable,
-as truly nested types (such as $\mathsf{Bush}$, Equation~\ref{eq:bush}) do,
-or other GADTs calling the recursive variable.
-Since both nested types
-and GADTs without nesting admit induction rules,
-it seems reasonable to expect that GADTs with nesting would as well.
-Surprisingly, that is not the case:
-indeed, the induction principle generally relies on (unary) parametricity of the semantic interpretation,
-and in the case of nested types it also relies on functorial semantics~\cite{jp20},
-but GADTs cannot admit both functorial and parametric semantics at the same time~\cite{jgj21}.
-In this section we show how induction for GADTs featuring nesting goes wrong
-by analyzing the following concrete example of such a type.
-\begin{equation}\label{gadt-nested}
+
+\section{Truly Nested GADTs Do Not Admit Induction
+    Rules}\label{sec:GADT-nested}
+
+{\color{red} HERE!!}
+
+In Sections~\ref{sec:deep-ind-GADTs} and~\ref{sec:framework} we
+derived induction rules for GADTs that are not truly nesteed. Since
+both nested types and GADTs without true nesting admit induction
+rules, we might expect that truly nested GADTs would as
+well. Surprisingly, however, they do not. {\color{red} That is, our
+  results from the previous section are the strongest possible.}
+%
+{\color{red} In fact, the induction rule for a data type generally
+  relies on (unary) parametricity of its semantic interpretation, and
+  in the case of nested types and GADTs it also relies on the data
+  types having a functorial semantics. But whereas nested types can
+  admit a functorial parametric semantics GADTs cannot admit both
+  functorial and parametric semantics at the same
+  time~\cite{jgj21}. In this section we show how induction for truly
+  nested GADTs nesting goes wrong by analyzing the following simple
+  concrete example of such a type.}
+\begin{equation*}\label{gadt-nested}
 \begin{array}{l}
@@
 \end{array}
-\end{equation}
-The constructor $\mathsf{c}$ can be rewritten as
-\[
-\mathsf{c : \forall\, (B : Set) \to Equal\,A\,(B \times B) \to G\,(G\,B) \to G\,A}
-\]
-which is the form we shall use from now on.
-The predicate-lifting function of $\mathsf{G}$,
-\[
-\mathsf{G^{\wedge} : \forall\, (A : Set) \to (A \to Set) \to G\,A \to Set}
-\]
-is defined as
+\end{equation*}
+The constructor $\mathsf{c}$ can be rewritten as $\mathsf{c :
+  \forall\, (B : Set) \to Equal\,A\,(B \times B) \to G\,(G\,B) \to
+  G\,A}$, so the predicate lifting $\mathsf{G^{\wedge} : \forall\, (A
+  : Set) \to (A \to Set) \to G\,A \to Set}$ for $\mathsf{G}$ is
 \[
 \mathsf{G^{\wedge}\,A\,Q\,(c\,B\,e\,x)
-= \sum\, [Q' : B \to Set]\,
+= \exists\, [Q']\,
 Equal^{\wedge}\,A\,(B \times B)\,Q\,(Pair^{\wedge}\,B\,B\,Q'\,Q')\,e
@@
 \]
-where $\mathsf{Q : A \to Set}$, $\mathsf{e : Equal\,A\,(B \times B)}$ and $\mathsf{x : G(G\,B)}$.
-Finally, let $\mathsf{dIndC}$ be the function
-\[
-\begin{array}{l}
-\mathsf{\lambda\, (P : \forall\, (A : Set) \to (A \to Set) \to G\,A \to Set)} \\
-\quad\mathsf{\to \forall\, (A\;B : Set)\, (Q : A \to Set)\, (Q' : B \to Set)\, (e : Equal\,A\,(B \times B))\, (x : G\,(G\,B))} \\
-\quad\mathsf{\to Equal^{\wedge}\,A\,(B \times B)\,Q\,(Pair^{\wedge}\,B\,B\,Q'\,Q')\,e
-	\to P\,(G\,B)\,(P\,B\,Q')\,x
+where $\mathsf{Q : A \to Set}$, $\mathsf{Q' : B \to Set}$, $\mathsf{e
+  : Equal\,A\,(B \times B)}$, and $\mathsf{x : G\,(G\,B)}$.
+The induction hypothesis $\mathsf{dIndC}$ for $\mathsf{c}$ is
+\[\begin{array}{l}
+\mathsf{\lambda\, (P : \forall\, (A : Set) \to (A \to Set) \to G\,A
+  \to Set)} \\ 
+\quad\mathsf{\to \forall\, (A\;B : Set)\, (Q : A \to Set)\, (Q' : B
+  \to Set)\, (e : Equal\,A\,(B \times B))\, (x : G\,(G\,B))} \\ 
+\quad\mathsf{\to Equal^{\wedge}\,A\,(B \times
+  B)\,Q\,(Pair^{\wedge}\,B\,B\,Q'\,Q')\,e \to P\,(G\,B)\,(P\,B\,Q')\,x 
 	\to P\,A\,Q\,(c\,B\,e\,x)}
-\end{array}
-\]
-associated to the $\mathsf{c}$ constructor.
-
-The induction rule for $\mathsf{G}$ is
-\[
-\mathsf{\forall\, (P : \forall\, (A : Set) \to (A \to Set) \to G\,A \to Set)
-\to dIndC\,P
-\to \forall\, (A : Set)\, (Q : A \to Set)\, (y : G\,a) \to G^{\wedge}\,A\,Q\,y \to P\,A\,Q\,y}
-\]
-Consistently with the previous examples,
-to prove the induction rule
-we try to define a term of the above type, $\mathsf{dIndG}$, as
-\[
-\mathsf{dIndG\,P\,cc\,A\,Q\,(c\,B\,e\,x)\,(Q', liftE, liftG)
-= cc\,A\,B\,Q\,Q'\,e\,x\,liftE\,p}
-\]
-where $\mathsf{cc : dIndC\,P}$,
-$\mathsf{Q : A \to Set}$ and
-\begin{itemize}
-\item $\mathsf{c\,B\,e\,x : G\,A}$,
-i.e., $\mathsf{e : Equal\,A\,(B \times B)}$ and $\mathsf{x : G\,(G\,B)}$;
-\item $\mathsf{(Q', liftE, liftG) : G^{\wedge}\,A\,Q\,(c\,B\,e\,x)}$,
-i.e., $\mathsf{Q' : B \to Set}$, $\mathsf{liftE : Equal^{\wedge}\,A\,(B \times B)\,Q\,(Pair^{\wedge}\,B\,B\,Q'\,Q')\,e}$,
-and $\mathsf{liftG : G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q')\,x}$.
-\end{itemize}
-We still need to define $\mathsf{p : P\,(G\,B)\,(P\,B\,Q')\,x}$.
-We do so by using the induction rule and letting
-\[
-\mathsf{p = dIndG\,P\,cc\,(G\,B)\,(P\,B\,Q')\,x\,q}
-\]
-where we still need to provide $\mathsf{q : G^{\wedge}\,(G\,B)\,(P\,B\,Q')\,x}$.
-If we had the map function of $\mathsf{G^{\wedge}}$,
-\[
-\mathsf{GLMap : \forall\, (A : Set)\, (Q\;Q'' : A \to Set) \to PredMap\,A\,Q\,Q'' \to PredMap\,(G\,A)\,(G^{\wedge}\,A\,Q)\,(G^{\wedge}\,A\,Q'')}
-\]
-then we would be able to define
-\[
-\mathsf{q = GLMap\,(G\,B)\,(G^{\wedge}\,B\,Q')\,(P\,B\,Q')\,(dIndG\,P\,cc\,B\,Q')\,x\,liftG}
-\]
-Unfortunately, we cannot define such a $\mathsf{GLMap}$.
+\end{array}\]
+so the induction rule for $\mathsf{G}$ is
+\[\mathsf{\forall\, (P : \forall\, (A : Set) \to (A \to Set) \to G\,A \to Set)
+\to dIndC\,P \to \forall\, (A : Set)\, (Q : A \to Set)\, (y : G\,a)
+\to G^{\wedge}\,A\,Q\,y \to P\,A\,Q\,y}\] But if we now try to show
+that this induction rule is sound by constructing a witness
+$\mathsf{dIndG}$ inhabiting this type we run into problems. We can try
+to define $\mathsf{dIndG\,P\,cc\,A\,Q\,(c\,B\,e\,x)\,(Q', liftE,
+  liftG) = cc\,A\,B\,Q\,Q'\,e\,x\,liftE\,p}$, where $\mathsf{cc :
+  dIndC\,P}$, $\mathsf{Q : A \to Set}$, $\mathsf{e : Equal\,A\,(B
+  \times B)}$, $\mathsf{x : G\,(G\,B)}$, $\mathsf{liftG :
+  G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q')\,x}$, and $\mathsf{Q' : B
+  \to Set}$, $\mathsf{liftE : Equal^{\wedge}\,A\,(B \times
+  B)\,Q\,(Pair^{\wedge}\,B\,B\,Q'\,Q')\,e}$.  However, we still need
+to define $\mathsf{p : P\,(G\,B)\,(P\,B\,Q')\,x}$.  If we try to do so
+by using the induction rule and letting $\mathsf{p =
+  dIndG\,P\,cc\,(G\,B)\,(P\,B\,Q')\,x\,q}$, then we'd still need to
+provide $\mathsf{q : G^{\wedge}\,(G\,B)\,(P\,B\,Q')\,x}$.  If we had
+the $\mathsf{map}$ function $\mathsf{G^\wedge Map : \forall\, (A :
+  Set)\, (Q\;Q'' : A \to Set) \to PredMap\,A\,Q\,Q'' \to
+  PredMap\,(G\,A)\,(G^{\wedge}\,A\,Q)\,(G^{\wedge}\,A\,Q'')}$ for
+$\mathsf{G^{\wedge}}$, then we would be able to define $\mathsf{q =
+  GLMap\,(G\,B)\,(G^{\wedge}\,B\,Q')\,(P\,B\,Q')\,
+  (dIndG\,P\,cc\,B\,Q')\,x\,liftG}$.
+Unfortunately, however, we cannot define such a $\mathsf{G^\wedge Map}$.
 Indeed, its definition would have to be
 \[
-\mathsf{GLMap\,A\,Q\,Q''\,m\,(c\,B\,e\,x)\,(Q', liftE, liftG) = (Q''', liftE', liftG')}
-\]
-for some $\mathsf{(Q''', liftE', liftG') : G^{\wedge}\,A\,Q''\,(c\,B\,e\,x)}$, i.e.,
-\begin{itemize}
-\item $\mathsf{Q''' : B \to Set}$;
-\item $\mathsf{liftE' : Equal^{\wedge}\,A\,(B \times B)\,Q''\,(Pair^{\wedge}\,B\,B\,Q'''\,Q''')\,e}$;
-\item $\mathsf{liftG' : G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q''')\,x}$;
-\end{itemize}
-where $\mathsf{Q : A \to Set}$, $\mathsf{Q'' : A \to Set}$,
-$\mathsf{m : PredMap\,A\,Q\,Q''}$,
-$\mathsf{c\,B\,e\,x : G\,A}$, i.e.,
-\begin{itemize}
-\item $\mathsf{e : Equal\,A\,(B \times B)}$;
-\item $\mathsf{x : G\,(G\,B)}$;
-\end{itemize}
-and $\mathsf{(Q', liftE, liftG) :G^{\wedge}\,A\,Q\,(c\,B\,e\,x)}$, i.e.,
-\begin{itemize}
-\item $\mathsf{Q' : B \to Set}$;
-\item $\mathsf{liftE : Equal^{\wedge}\,A\,(B \times B)\,Q\,(Pair^{\wedge}\,B\,B\,Q'\,Q')\,e}$;
-\item $\mathsf{liftG : G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q')\,x}$.
-\end{itemize}
-In other words, we have a proof $\mathsf{liftE}$ of the (extensional) equality
-of the predicates $\mathsf{Q}$ and $\mathsf{Pair^{\wedge}\,B\,B\,Q'\,Q'}$
-and a morphism of predicates $\mathsf{m}$ from $\mathsf{Q}$ to $\mathsf{Q''}$,
-and we need to use those data to deduce a proof of the (extensional) equality
-of the predicates $\mathsf{Q''}$ and $\mathsf{Pair^{\wedge}\,B\,B\,Q'''\,Q'''}$,
-for some predicate $\mathsf{Q'''}$ on $\mathsf{B}$.
-But that is not generally possible:
-the facts that $\mathsf{Q}$ is equal to $\mathsf{Pair^{\wedge}\,B\,B\,Q'\,Q'}$
-and that there is a morphism of predicates $\mathsf{m}$ from $\mathsf{Q}$ to $\mathsf{Q''}$
-do not guarantee that there exists a predicate $\mathsf{Q'''}$ such that $\mathsf{Q''}$ is equal to $\mathsf{Pair^{\wedge}\,B\,B\,Q'''\,Q'''}$.
-
-At a deeper level, the fundamental issue is that
-the $\mathsf{Equal}$ type does not have functorial semantics,
-so that having morphisms $\mathsf{A \to A'}$ and $\mathsf{B \to B'}$
-(for any type $\mathsf{A, A', B}$ and $\mathsf{B'}$)
-and a proof that $\mathsf{A}$ is equal to $\mathsf{A'}$
-does not provide a proof that $\mathsf{B}$ is equal to $\mathsf{B'}$.
-This is because GADTs can either have a syntax-only semantics or a functorial-completion semantics.
-Since we are interested in induction rules, we consider the syntax-only semantics,
-which is parametric but not functorial.
-Had we considered the functorial-completion semantics, which is functorial,
-we would have forfeited parametricity instead.
-In both cases, thus, we cannot derive an induction rule for GADTs featuring nesting.
-Unlike nested types, indeed, GADTs do not admit a semantic interpretation that is both parametric and functorial~\cite{jgj21}.
-
-\section{Applications}
-
-%We want a proof
-%\[
-%  \mathsf{getTypeProof : \forall \, (A : Set) \to LTerm\,A \to Maybe \, (LType \, A)} \\
-%\]
-%that takes a lambda term and produces its type (using $\mathsf{Maybe}$ to represent
-%potential failure).
-%Because of the $\mathsf{listC}$ constructor of $\mathsf{LTerm}$, this cannot be achieved without 
-%deep induction. In particular, deep induction is required to apply the induction to the 
-%individual terms in a list of terms.
-
-In this section we use deep induction for the $\mathsf{LTerm}$ GADT from Example~\ref{example:lterm} to 
-extract the type from a lambda term.
+\mathsf{G^\wedge Map\,A\,Q\,Q_2\,m\,(c\,B\,e\,x)\,(Q_1, liftE, liftG) =
+  (Q_3, liftE', liftG')}
+\]
+for some $\mathsf{Q_3 : B \to Set}$, $\mathsf{liftE' :
+  Equal^{\wedge}\,A\,(B \times
+  B)\,Q_2\,(Pair^{\wedge}\,B\,B\,Q_3\,Q_3)\,e}$, and $\mathsf{liftG' :
+  G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q_3)\,x}$, where $\mathsf{Q : A
+  \to Set}$, $\mathsf{Q_2 : A \to Set}$, $\mathsf{m :
+  PredMap\,A\,Q\,Q_2}$, $\mathsf{e : Equal\,A\,(B \times B)}$,
+$\mathsf{x : G\,(G\,B)}$, $\mathsf{Q_1 : B \to Set}$, $\mathsf{liftE :
+  Equal^{\wedge}\,A\,(B \times
+  B)\,Q\,(Pair^{\wedge}\,B\,B\,Q_1\,Q_1)\,e}$, and $\mathsf{liftG :
+  G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q_1)\,x}$. In other words, we
+have a proof $\mathsf{liftE}$ of the (extensional) equality of the
+predicates $\mathsf{Q}$ and $\mathsf{Pair^{\wedge}\,B\,B\,Q_1\,Q_1}$
+and a morphism of predicates $\mathsf{m}$ from $\mathsf{Q}$ to
+$\mathsf{Q_2}$, and we need to use those data to deduce a proof of the
+(extensional) equality of the predicates $\mathsf{Q_2}$ and
+$\mathsf{Pair^{\wedge}\,B\,B\,Q_3\,Q_3}$, for some predicate
+$\mathsf{Q_3}$ on $\mathsf{B}$.  But that is not generally possible:
+the facts that $\mathsf{Q}$ is equal to
+$\mathsf{Pair^{\wedge}\,B\,B\,Q_1\,Q_1}$ and that there is a morphism
+of predicates $\mathsf{m}$ from $\mathsf{Q}$ to $\mathsf{Q_2}$ do not
+guarantee that there exists a predicate $\mathsf{Q_3}$ such that
+$\mathsf{Q_2}$ is equal to $\mathsf{Pair^{\wedge}\,B\,B\,Q_3\,Q_3}$.
+
+At a deeper level, the fundamental issue is that the $\mathsf{Equal}$
+type does not have functorial semantics~\cite{jgj21}, so that having
+morphisms $\mathsf{A \to A'}$ and $\mathsf{B \to B'}$ (for any type
+$\mathsf{A, A', B}$ and $\mathsf{B'}$) and a proof that $\mathsf{A}$
+is equal to $\mathsf{A'}$ does not provide a proof that $\mathsf{B}$
+is equal to $\mathsf{B'}$. {\color{red} This is because GADTs can
+  either have a syntax-only semantics or a functorial semantics.
+  Since we are interested in induction rules, we consider the
+  syntax-only semantics, which is parametric but not functorial.  Had
+  we considered the functorial-completion semantics, which is
+  functorial, we would have forfeited parametricity instead.  In both
+  cases, thus, we cannot derive an induction rule for GADTs featuring
+  nesting.  Unlike nested types, indeed, GADTs do not admit a semantic
+  interpretation that is both parametric and functorial~\cite{jgj21}.}
+
+\section{Applications}\label{sec:app}
+
+In this section we use deep induction for the $\mathsf{LTerm}$ GADT
+from~\eqref{eq:eq_lterm} to extract the type from a lambda term.
 Consider the predicate

### 2021-04-19 induction-paper/ENTCS-GADT-induction.tex
Started section 7

@@
 over themselves do not, to our knowledge, appear in applications or
-the literature.
-
-All of the deep induction rules appearing in this paper have been
-derived using our general framework. Our Agda code implementing them
-is available at~\cite{web-page}.
+the literature. Section~\ref{sec:app} details a case study in using
+deep induction. All of the deep induction rules appearing in this
+paper have been derived using our general framework. Our Agda code
+implementing them is available at~\cite{web-page}.
 
@@
 Moreover,
+
+\vspace*{-0.2in}
+
 \[\begin{array}{lll}
@@
 \end{array}\]
+
+\vspace*{-0.2in}
+
+\noindent
 where, in the final clause, $\mathsf{PredMap : \forall\, (A : Set) \to
@@
 constructed inductively according to the following grammar:
-
-\vspace*{-0.05in}
-
 \[\mathsf{F\,G\,\ol{B}\; :=\;
@@
 (\ol{F_1\,G\,\ol{B}}) }\]
-This grammar is subject to the following
-restrictions. In the third clause the type constructor $\mathsf{F_1}$
-does not contain $\mathsf{G}$. In the fourth clause, none of the
-$\mathsf{\alpha}$-many type constructors in $\mathsf{\ol{F_1}}$
-contains $\mathsf{G}$.  This prevents nesting, which would make it
-impossible to give an induction rule; see
-Section~\ref{sec:GADT-nested} below. In the fifth clause, $\mathsf{H}$
-does not contain $\mathsf{G}$. This clause therefore subsumes the
-cases in which $\mathsf{F\,G\,\ol{B}}$ is a closed type or one of the
-$\mathsf{B_i}$. In the sixth clause, $\mathsf{H : Set^\gamma \to Set}$
-does not contain $\mathsf{G}$ (although $\ol{\mathsf{F_1\,G\,\ol{B}}}$
-can). Moreover, although $\mathsf{H}$ can construct any (truly)
-nested type, it cannot construct a GADT. This ensures that
-$\mathsf{H}$ admits functorial semantics~\cite{jp20}, and thus has an
-associated $\mathsf{map}$ function. From this we can also construct a
-$\mathsf{map}$ function {\color{red} Say what PredMap is.}
+
+\pagebreak
+
+\noindent
+This grammar is subject to the following restrictions. In the third
+clause the type constructor $\mathsf{F_1}$ does not contain
+$\mathsf{G}$. In the fourth clause, none of the $\mathsf{\alpha}$-many
+type constructors in $\mathsf{\ol{F_1}}$ contains $\mathsf{G}$.  This
+prevents nesting, which would make it impossible to give an induction
+rule; see Section~\ref{sec:GADT-nested} below. In the fifth clause,
+$\mathsf{H}$ does not contain $\mathsf{G}$. This clause therefore
+subsumes the cases in which $\mathsf{F\,G\,\ol{B}}$ is a closed type
+or one of the $\mathsf{B_i}$. In the sixth clause, $\mathsf{H :
+  Set^\gamma \to Set}$ does not contain $\mathsf{G}$ (although
+$\ol{\mathsf{F_1\,G\,\ol{B}}}$ can). Moreover, although $\mathsf{H}$
+can construct any (truly) nested type, it cannot construct a
+GADT. This ensures that $\mathsf{H}$ admits functorial
+semantics~\cite{jp20}, and thus has an associated map function. From
+this we can also construct a map function
 
@@
 (H^{\wedge}\,\ol{A}\,\ol{Q'})}\]
-for $\mathsf{H^{\wedge}}$.  A
-concrete way to define $\mathsf{H^\wedge Map}$ is by induction on the
-structure of the type $\mathsf{H}$, but we omit such details since
-they are not essential to the present discussion. A further
-requirement that applies to all of the types appearing above,
-including the $\mathsf{K}$s in~\eqref{eq:gadts}, is that they must all
-admit predicate liftings. This is not an overly restrictive condition,
-though: all types constructed from sums, products, arrow types and
-type application admit predicate liftings, and so do GADTs constructed
-from the above grammar; in fact, we have seen such liftings for
-products and type application in Section~\ref{sec:deep-ind-GADTs}.  A
-concrete way to define predicate liftings more generally is, again, by
-induction on the the structure of the types. We do not give a general
-definition of predicate liftings here, though, since that would
-require us to first design a full type calculus, which is beyond the
-scope of the present paper.
+
+\vspace*{0.05in}
+
+\noindent
+for $\mathsf{H^{\wedge}}$.  A concrete way to define $\mathsf{H^\wedge
+  Map}$ is by induction on the structure of the type $\mathsf{H}$, but
+we omit such details since they are not essential to the present
+discussion. A further requirement that applies to all of the types
+appearing above, including the $\mathsf{K}$s in~\eqref{eq:gadts}, is
+that they must all admit predicate liftings. This is not an overly
+restrictive condition, though: all types constructed from sums,
+products, arrow types and type application admit predicate liftings,
+and so do GADTs constructed from the above grammar; in fact, we have
+seen such liftings for products and type application in
+Section~\ref{sec:deep-ind-GADTs}.  A concrete way to define predicate
+liftings more generally is, again, by induction on the the structure
+of the types. We do not give a general definition of predicate
+liftings here, though, since that would require us to first design a
+full type calculus, which is beyond the scope of the present paper.
 
@@
 therefore
-\[
-\mathsf{G^{\wedge}\,A\,Q_A\,(c\,\ol{B}\,e\,x)
+
+\vspace*{-0.05in}
+
+\[\mathsf{G^{\wedge}\,A\,Q_A\,(c\,\ol{B}\,e\,x)
 = \exists [\ol{Q_B}]\,
 Equal^{\wedge}\,A\,(K\,\ol{B})\,Q_A\,(K^{\wedge}\,\ol{B}\,\ol{Q_B})\,e
-\times F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,{\ol{Q_B}}\,x}
-\]
+\times F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,{\ol{Q_B}}\,x}\]
+
+\noindent
 where $\mathsf{Q_A : A \to Set}$, $\ol{\mathsf{Q_B : B \to Set}}$,
 $\mathsf{e : Equal\,A\,(K\,\ol{B})}$, and $\mathsf{x : F\,G\,\ol{B}}$.
-Assuming predicate liftings 
-\[\begin{array}{lll}
-\mathsf{F^{\wedge}} & : & \mathsf{\forall (G : Set^{\alpha} \to Set) (\ol{B : Set})
-\to (\forall (A : Set) \to (A \to Set) \to G\,A \to Set)
-\to (\ol{B \to Set}) \to F\,G\,\ol{B} \to Set}\\ 
-\mathsf{K^{\wedge}} & : & \mathsf{\forall (\ol{B : Set}) \to (\ol{B \to Set}) \to
-  K\,\ol{B} \to Set}
-\end{array}\]
+If we have predicate liftings $\mathsf{F^{\wedge}} : \mathsf{\forall (G
+  : Set^{\alpha} \to Set) (\ol{B : Set}) \to (\forall (A : Set) \to (A
+  \to Set) \to G\,A \to Set) \to (\ol{B \to Set}) \to F\,G\,\ol{B} \to
+  Set}$ and $\mathsf{K^{\wedge}} : \mathsf{\forall (\ol{B : Set}) \to
+  (\ol{B \to Set}) \to K\,\ol{B} \to Set}$
+%\[\begin{array}{lll}
+%\mathsf{F^{\wedge}} & : & \mathsf{\forall (G : Set^{\alpha} \to Set) (\ol{B : Set})
+%\to (\forall (A : Set) \to (A \to Set) \to G\,A \to Set)
+%\to (\ol{B \to Set}) \to F\,G\,\ol{B} \to Set}\\ 
+%\mathsf{K^{\wedge}} & : & \mathsf{\forall (\ol{B : Set}) \to (\ol{B \to Set}) \to
+%  K\,\ol{B} \to Set}
+%\end{array}\]
 for $\mathsf{F}$ and for $\mathsf{K}$, respectively, the induction
@@
     (\ol{C \to Set})}$ $\mathsf{\to H\,\ol{C} \to Set}$, and if
-  $\mathsf{H^\wedge}$ has a $\mathsf{map}$ function $\mathsf{H^\wedge
-    Map = \forall (\ol{C : Set}) (\ol{Q_C\;Q_C' : C \to Set}) \to
+  $\mathsf{H^\wedge}$ has a map function $\mathsf{H^\wedge Map =
+    \forall (\ol{C : Set}) (\ol{Q_C\;Q_C' : C \to Set}) \to
     PredMap\,\ol{C}\,\ol{Q_C}\,\ol{Q_C'} \to}$
@@
 
-\section{Truly Nested GADTs Do Not Admit Induction
+\section{Truly Nested GADTs Do Not Admit Deep Induction
     Rules}\label{sec:GADT-nested}
 
-{\color{red} HERE!!}
-
 In Sections~\ref{sec:deep-ind-GADTs} and~\ref{sec:framework} we
-derived induction rules for GADTs that are not truly nesteed. Since
-both nested types and GADTs without true nesting admit induction
-rules, we might expect that truly nested GADTs would as
-well. Surprisingly, however, they do not. {\color{red} That is, our
-  results from the previous section are the strongest possible.}
-%
-{\color{red} In fact, the induction rule for a data type generally
-  relies on (unary) parametricity of its semantic interpretation, and
-  in the case of nested types and GADTs it also relies on the data
-  types having a functorial semantics. But whereas nested types can
-  admit a functorial parametric semantics GADTs cannot admit both
-  functorial and parametric semantics at the same
-  time~\cite{jgj21}. In this section we show how induction for truly
-  nested GADTs nesting goes wrong by analyzing the following simple
-  concrete example of such a type.}
+derived deep induction rules for GADTs that are not truly
+nested. Since both nested types and GADTs without true nesting admit
+deep induction rules, we might expect that truly nested GADTs would as
+well. Surprisingly, however, they do not. That is, our results from
+the previous section are the strongest possible. Indeed, the induction
+rule for a data type generally relies on (unary) parametricity of the
+model interpreting it, and deel induction for a truly nested type or a
+ruly nested GADT also relies on this interpretation being functorial.
+But, whereas nested types can admit a functorial parametric semantics,
+GADTs cannot admit both functorial and parametric semantics at the
+same time~\cite{jgj21}. In this section we show how deep induction for
+truly nested GADTs nesting goes wrong by analyzing the following
+concrete example of such a type.\looseness=-1
+
+\vspace*{-0.05in}
+
 \begin{equation*}\label{gadt-nested}
@@
 \end{array}\]
-so the induction rule for $\mathsf{G}$ is
+so the deep induction rule for $\mathsf{G}$ is
 \[\mathsf{\forall\, (P : \forall\, (A : Set) \to (A \to Set) \to G\,A \to Set)
@@
 \to G^{\wedge}\,A\,Q\,y \to P\,A\,Q\,y}\] But if we now try to show
-that this induction rule is sound by constructing a witness
-$\mathsf{dIndG}$ inhabiting this type we run into problems. We can try
-to define $\mathsf{dIndG\,P\,cc\,A\,Q\,(c\,B\,e\,x)\,(Q', liftE,
-  liftG) = cc\,A\,B\,Q\,Q'\,e\,x\,liftE\,p}$, where $\mathsf{cc :
-  dIndC\,P}$, $\mathsf{Q : A \to Set}$, $\mathsf{e : Equal\,A\,(B
-  \times B)}$, $\mathsf{x : G\,(G\,B)}$, $\mathsf{liftG :
-  G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q')\,x}$, and $\mathsf{Q' : B
-  \to Set}$, $\mathsf{liftE : Equal^{\wedge}\,A\,(B \times
-  B)\,Q\,(Pair^{\wedge}\,B\,B\,Q'\,Q')\,e}$.  However, we still need
-to define $\mathsf{p : P\,(G\,B)\,(P\,B\,Q')\,x}$.  If we try to do so
-by using the induction rule and letting $\mathsf{p =
-  dIndG\,P\,cc\,(G\,B)\,(P\,B\,Q')\,x\,q}$, then we'd still need to
-provide $\mathsf{q : G^{\wedge}\,(G\,B)\,(P\,B\,Q')\,x}$.  If we had
-the $\mathsf{map}$ function $\mathsf{G^\wedge Map : \forall\, (A :
-  Set)\, (Q\;Q'' : A \to Set) \to PredMap\,A\,Q\,Q'' \to
+that this rule is sound by constructing a witness $\mathsf{dIndG}$
+inhabiting this type we run into problems. We can define
+$\mathsf{dIndG\,P\,cc\,A\,Q\,(c\,B\,e\,x)\,(Q', liftE, liftG) =
+  cc\,A\,B\,Q\,Q'\,e\,x\,liftE\,p}$, where $\mathsf{cc : dIndC\,P}$,
+$\mathsf{Q : A \to Set}$, $\mathsf{e : Equal\,A\,(B \times B)}$,
+$\mathsf{x : G\,(G\,B)}$, $\mathsf{liftG :
+  G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q')\,x}$, $\mathsf{Q' : B \to
+  Set}$, and $\mathsf{liftE : Equal^{\wedge}\,A\,(B \times
+  B)\,Q\,(Pair^{\wedge}\,B\,B\,Q'\,Q')\,e}$, but we still need to
+define $\mathsf{p : P\,(G\,B)\,(P\,B\,Q')\,x}$.  For this we can use
+the induction rule and let $\mathsf{p =
+  dIndG\,P\,cc\,(G\,B)\,(P\,B\,Q')\,x\,q}$, but we still need to
+define $\mathsf{q : G^{\wedge}\,(G\,B)\,(P\,B\,Q')\,x}$.  If we had
+the map function $\mathsf{G^\wedge Map : \forall\, (A : Set)\, (Q\;Q''
+  : A \to Set) \to PredMap\,A\,Q\,Q'' \to
   PredMap\,(G\,A)\,(G^{\wedge}\,A\,Q)\,(G^{\wedge}\,A\,Q'')}$ for
-$\mathsf{G^{\wedge}}$, then we would be able to define $\mathsf{q =
+$\mathsf{G^{\wedge}}$, then we could define $\mathsf{q =
   GLMap\,(G\,B)\,(G^{\wedge}\,B\,Q')\,(P\,B\,Q')\,
-  (dIndG\,P\,cc\,B\,Q')\,x\,liftG}$.
-Unfortunately, however, we cannot define such a $\mathsf{G^\wedge Map}$.
-Indeed, its definition would have to be
-\[
-\mathsf{G^\wedge Map\,A\,Q\,Q_2\,m\,(c\,B\,e\,x)\,(Q_1, liftE, liftG) =
-  (Q_3, liftE', liftG')}
-\]
-for some $\mathsf{Q_3 : B \to Set}$, $\mathsf{liftE' :
-  Equal^{\wedge}\,A\,(B \times
+  (dIndG\,P\,cc\,B\,Q')\,x\,liftG}$.  Unfortunately, however, we
+cannot define $\mathsf{G^\wedge Map}$.  Indeed, its definition would
+have to be $ \mathsf{G^\wedge Map\,A\,Q\,Q_2\,m\,(c\,B\,e\,x)\,(Q_1,
+  liftE, liftG) = (Q_3, liftE', liftG')}$ for some $\mathsf{Q_3 : B
+  \to Set}$, $\mathsf{liftE' : Equal^{\wedge}\,A\,(B \times
   B)\,Q_2\,(Pair^{\wedge}\,B\,B\,Q_3\,Q_3)\,e}$, and $\mathsf{liftG' :
@@
   G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q_1)\,x}$. In other words, we
-have a proof $\mathsf{liftE}$ of the (extensional) equality of the
-predicates $\mathsf{Q}$ and $\mathsf{Pair^{\wedge}\,B\,B\,Q_1\,Q_1}$
-and a morphism of predicates $\mathsf{m}$ from $\mathsf{Q}$ to
-$\mathsf{Q_2}$, and we need to use those data to deduce a proof of the
-(extensional) equality of the predicates $\mathsf{Q_2}$ and
-$\mathsf{Pair^{\wedge}\,B\,B\,Q_3\,Q_3}$, for some predicate
-$\mathsf{Q_3}$ on $\mathsf{B}$.  But that is not generally possible:
-the facts that $\mathsf{Q}$ is equal to
-$\mathsf{Pair^{\wedge}\,B\,B\,Q_1\,Q_1}$ and that there is a morphism
-of predicates $\mathsf{m}$ from $\mathsf{Q}$ to $\mathsf{Q_2}$ do not
-guarantee that there exists a predicate $\mathsf{Q_3}$ such that
-$\mathsf{Q_2}$ is equal to $\mathsf{Pair^{\wedge}\,B\,B\,Q_3\,Q_3}$.
+would need to produce a proof of the (extensional) equality of the
+predicates $\mathsf{Q_2}$ and $\mathsf{Pair^{\wedge}\,B\,B\,Q_3\,Q_3}$
+from just a proof
+%$\mathsf{liftE}$
+of the (extensional) equality of the predicates $\mathsf{Q}$ and
+$\mathsf{Pair^{\wedge}\,B\,B\,Q_1\,Q_1}$ and a morphism of predicates
+$\mathsf{m}$ from $\mathsf{Q}$ to $\mathsf{Q_2}$.
+%a proof $\mathsf{liftE}$ of the (extensional) equality of the
+%predicates $\mathsf{Q}$ and $\mathsf{Pair^{\wedge}\,B\,B\,Q_1\,Q_1}$
+%and a morphism of predicates $\mathsf{m}$ from $\mathsf{Q}$ to
+%$\mathsf{Q_2}$, and we need to use those data to deduce a proof of the
+%(extensional) equality of the predicates $\mathsf{Q_2}$ and
+%$\mathsf{Pair^{\wedge}\,B\,B\,Q_3\,Q_3}$, for some predicate
+%$\mathsf{Q_3}$ on $\mathsf{B}$.
+But this will not be possible in general: the facts that $\mathsf{Q}$
+is equal to $\mathsf{Pair^{\wedge}\,B\,B\,Q_1\,Q_1}$ and that there is
+a morphism of predicates $\mathsf{m}$ from $\mathsf{Q}$ to
+$\mathsf{Q_2}$ do not guarantee that there exists a predicate
+$\mathsf{Q_3}$ such that $\mathsf{Q_2}$ is equal to
+$\mathsf{Pair^{\wedge}\,B\,B\,Q_3\,Q_3}$.
 
@@
 is equal to $\mathsf{A'}$ does not provide a proof that $\mathsf{B}$
-is equal to $\mathsf{B'}$. {\color{red} This is because GADTs can
-  either have a syntax-only semantics or a functorial semantics.
-  Since we are interested in induction rules, we consider the
-  syntax-only semantics, which is parametric but not functorial.  Had
-  we considered the functorial-completion semantics, which is
-  functorial, we would have forfeited parametricity instead.  In both
-  cases, thus, we cannot derive an induction rule for GADTs featuring
-  nesting.  Unlike nested types, indeed, GADTs do not admit a semantic
-  interpretation that is both parametric and functorial~\cite{jgj21}.}
-
-\section{Applications}\label{sec:app}
+is equal to $\mathsf{B'}$. This difficulty propagates to any truly
+nested GADT because it is not possible to define a map function
+$\mathsf{Equal^\wedge Map}$ for $\mathsf{Eq^\wedge}$ when one of the
+type arguments to $\mathsf{Equal^\wedge}$ is structured (i.e., not a
+variable). This is, of course, a critical component in any approach to
+deep induction for such GADTs.\looseness=-1
+
+\section{Case Study: Extracting Types of Lambda Terms}\label{sec:app}
 
 In this section we use deep induction for the $\mathsf{LTerm}$ GADT
-from~\eqref{eq:eq_lterm} to extract the type from a lambda term.
-Consider the predicate
+from~\eqref{eq:eq_lterm} to extract the type from a lambda term.  For
+this we use the following predicate, which takes a lambda term and
+produces the type of its possible types:
 \begin{align*}
@@
 \end{align*}
-that takes a lambda term and produces the type of its possible types.
-This predicate uses $\mathsf{Maybe}$ to represent
-potential failure, where $\mathsf{Maybe}$ is the type defined as follows:
+This predicate uses $\mathsf{Maybe}$ to represent potential failure,
+where $\mathsf{Maybe}$ is defined as follows:
 \begin{equation}\label{eq:maybe}
@@
 \end{equation}
-We want to show this predicate is satisfied for every element of $\mathsf{LTerm\,A}$,
-i.e., we want to prove: 
-\[
-  \mathsf{getTypeProof : \forall \, (A : Set)\, (t : LTerm\,A) \to GetType \,A \,t} \\
-\]
-Because of the $\mathsf{listC}$ constructor of $\mathsf{LTerm}$, this cannot be achieved without 
-deep induction. In particular, deep induction is required to apply the induction to the 
-individual terms in a list of terms. 
-So, using the deep induction rule for $\mathsf{LTerm}$ from Section~\ref{sec:dind-lterm} {\color{red} (add back-reference)},
-we define $\mathsf{getTypeProof}$ as
-\[
-  \mathsf{getTypeProof \,A \,t = 
-    dIndLTerm\, P \,
-    cvar\,  cabs\,  capp\,  clistc\,  A\, K_\top\, t\, (LTermLKT\, A\, t)
-    }
-\]
-where $\mathsf{t : LTerm\,A}$,
-$\mathsf{P}$ is the polymorphic predicate $\mathsf{\lambda \, (A: Set)\, (Q : A \to Set)\, (t : LTerm\,A)\, \to Maybe \, (LType \, A)}$
-%where $\mathsf{K : a \to Set}$ is the constantly true predicate:
-%\[
-%  \mathsf{K_\top \, x = \top}
-%\]
-and $\mathsf{LTermLKT : \forall\, (A : Set)\, (t : LTerm A) \to LTerm^{\wedge}\, A\, K_\top\,t}$
-is a function that we will define later.
-%Notice that there is no space in $\mathsf{LTermLKT}$, because
-%\[
-%  \mathsf{LTermLKT : \forall (a : Set) (t : LTerm A) \to LTerm^{\wedge}\, a\, K_\top\,t}
-%\]
-%is a function that we will define.
-In addition to defining $\mathsf{LTermLKT}$,
-we also have to give a term for each constructor of $\mathsf{LTerm}$,
-as already discussed in Section~\ref{sec:dind-lterm} {\color{red} (add back-reference)}:
-\begin{itemize}
-% var
-\item a term $\mathsf{cvar : dIndVar\,P}$ associated to the constructor $\mathsf{var}$, i.e.,
-\[
-  \mathsf{cvar : \forall\, (A : Set)\, (Q : A \to Set)\, (s : String)\, (T : LType\, A) 
-      \to LType^{\wedge} \, A\, Q\, T\, \to Maybe\, (LType\, A)}
-\]
-% abs 
-\item a term $\mathsf{cabs : dIndAbs\,P}$ associated to the constructor $\mathsf{abs}$, i.e.,
-\[
-\begin{array}{l}
-    \mathsf{cabs : \forall\, (A \, B\, C: Set)\, (Q : A \to Set)\, (Q' : B \to Set)\, (Q'' : C \to Set)} \\
-    \quad\mathsf{(e : Equal\, A\, (B \to C))\, (s : String)\, (T : LType\, B)\, (t : LTerm \, C)} \\ 
-    \quad\mathsf{ 
-      \to Equal^{\wedge}\, A\, (B \to C)\, Q\, (Arr^{\wedge}\, B\, C\, Q'\, Q'')\, e
-      \to LType^{\wedge} \, B\, Q'\, T
-      \to Maybe\, (LType\,C) \to Maybe\, (LType\, A)}
-\end{array}
-\]
-% app
-\item a term $\mathsf{capp : dIndApp\,P}$ associated to the constructor $\mathsf{app}$, i.e.,
-\[
-\begin{array}{l}
-    \mathsf{capp : \forall\, (A \, B : Set)\, (Q : A \to Set)\, (Q' : B \to Set)\, 
-      (t : LTerm\, (B \to A))\, (t' : LTerm\, B)} \\ 
-    \quad\mathsf{
-      \to Maybe\, (LType\, (B \to A))
-      \to Maybe\, (LType\, B)
-      \to Maybe\, (LType\, A)
-    }
-\end{array}
-\]
-% list 
-\item a term $\mathsf{clistc : dIndListC\,P}$ associated to the constructor $\mathsf{listC}$, i.e.,
-\[
-\begin{array}{l}
-    \mathsf{clistc : \forall\, (A \, B : Set)\, (Q : A \to Set)\, (Q' : B \to Set)\,
-      (e : Equal\, A\, (List\, B))\, (ts : List\, (LTerm\, B))} \\ 
-    \quad\mathsf{
-      \to Equal^{\wedge}\, A\, (List\,B) \, Q\, (List^{\wedge}\, B\, Q')\, e
-      \to List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, ts
-      \to Maybe\, (LType\, A)
-    }
-\end{array}
-\]
-\end{itemize}
+We want to show that $\mathsf{GetType}$ is satisfied by every element
+of $\mathsf{LTerm\,A}$, i.e., we want to prove:
+\[ \mathsf{getTypeProof : \forall \, (A : Set)\, (t : LTerm\,A) \to
+   GetType \,A \,t}\] This cannot be achieved without deep induction,
+which is required to to apply the induction hyppothesis to the
+individual terms in the list of terms that is an argument to the data
+constructor $\mathsf{list}$ for $\mathsf{LTerm}$.
+
+Using the deep induction rule for $\mathsf{LTerm}$ from
+Section~\ref{sec:ind-lam} we define $\mathsf{getTypeProof}$ by
+\[ \mathsf{getTypeProof \,A \,t = 
+    dIndLTerm\, P \, cvar\, cabs\, capp\, clistc\, A\, K_\top\, t\,
+    (LTerm\wedge KT\, A\, t) }\] where $\mathsf{t : LTerm\,A}$,
+$\mathsf{P}$ is the polymorphic predicate $\mathsf{\lambda \, (A:
+  Set)\, (Q : A \to Set)\, (t : LTerm\,A)\, \to Maybe \, (LType \,
+  A)}$ and $\mathsf{LTerm^\wedge KT : \forall\, (A : Set)\, (t : LTerm
+  A) \to LTerm^{\wedge}\, A\, K_\top\,t}$ is a function that we will
+define later.  We also need the applications to $\mathsf{P}$ of the
+induction hypotheses from Section~\ref{sec:ind-lam} for
+$\mathsf{LTerm}$'s data constructors. These are:
+\[\begin{array}{lll}
+\mathsf{cvar} & : & \mathsf{\forall\, (A : Set)\, (Q_A : A \to Set)\,
+  (s : String)\, (T_A : LType\, A) \to LType^{\wedge} \, A\, Q_A\,
+  T_A\, \to Maybe\, (LType\, A)}\\
+ \mathsf{cabs} & : & \mathsf{\forall\, (A \, B\, C: Set)\, (Q_A : A
+   \to Set)\, (Q_B : B \to Set)\, (Q_C : C \to Set)} \\ 
+& &    \quad\mathsf{(e : Equal\, A\, (B \to C))\, (s : String)\, (T_B
+   : LType\, B)\, (t_C : LTerm \, C)} \\  
+& & \quad\quad\mathsf{\to Equal^{\wedge}\, A\, (B \to C)\, Q_A\,
+   (Arr^{\wedge}\, B\, C\, Q_B\, Q_C)\, e  \to LType^{\wedge} \, B\,
+   Q_B\, T_B  \to Maybe\, (LType\,C) \to Maybe\, (LType\, A)}\\
+\mathsf{capp} & : & \mathsf{\forall\, (A \, B : Set)\, (Q_A : A \to
+  Set)\, (Q_B : B \to Set)\,  (t_{BA}: LTerm\, (B \to A))\, (t_B :
+  LTerm\, B)} \\  
+& & \quad\mathsf{\to Maybe\, (LType\, (B \to A))\to Maybe\, (LType\,
+      B) \to Maybe\, (LType\, A)}\\
+\mathsf{clist} & : & \mathsf{\forall\, (A \, B : Set)\, (Q_A : A \to
+  Set)\, (Q_B : B \to Set)\, (e : Equal\, A\, (List\, B))\, (ts : 
+  List\, (LTerm\, B))}\\  
+& & \quad\mathsf{\to Equal^{\wedge}\, A\, (List\,B) \, Q_A\,
+  (List^{\wedge}\, B\, Q_B)\, e \to List^{\wedge}\, (LTerm\, B)\,
+  (GetType\, B)\, ts \to Maybe\, (LType\, A)}
+\end{array}\]
+
+{\color{red} HERE!!}
+
 For variables we simply return the type $\mathsf{T : LType\, A}$ (wrapped in $\mathsf{just}$), 
 and the cases for abstraction and application are similar.
-The interesting case is $\mathsf{clistc}$, in which we have to use 
+The interesting case is $\mathsf{clist}$, in which we have to use 
 the results of $\mathsf{List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, ts}$
 in order to extract the type of one of the terms in the list. 
-To define $\mathsf{clistc}$ we pattern-match on the list of terms $\mathsf{ts}$.
+To define $\mathsf{clist}$ we pattern-match on the list of terms $\mathsf{ts}$.
 If $\mathsf{ts}$ is the empty list $\mathsf{nil}$, we cannot extract a type, 
@@
 \[
-  \mathsf{clistc\, A\, B\, Q\, Q'\, e\, nil \, liftE\, lift_{ts} = nothing}
+  \mathsf{clist\, A\, B\, Q\, Q'\, e\, nil \, liftE\, lift_{ts} = nothing}
 \]
@@
 \begin{align*}
-  &\mathsf{clistc\, A\, B\, Q\, Q'\, e\, (cons\, t\, ts') \, liftE\, (nothing , lift_{ts'}) = nothing} \\
-  &\mathsf{clistc\, A\, B\, Q\, Q'\, e\, (cons\, t\, ts') \, liftE\, (just\, T' , lift_{ts'}) = just \, (TList\, B\, e\, T')}
+  &\mathsf{clist\, A\, B\, Q\, Q'\, e\, (cons\, t\, ts') \, liftE\, (nothing , lift_{ts'}) = nothing} \\
+  &\mathsf{clist\, A\, B\, Q\, Q'\, e\, (cons\, t\, ts') \, liftE\, (just\, T' , lift_{ts'}) = just \, (TList\, B\, e\, T')}
 \end{align*}
@@
 and define the $\mathsf{abs}$ case of $\mathsf{LTermLKT}$ as 
-\[
-  \mathsf{LTermLKT\,A\, (abs\, B \,C \, e \,s \,T \, t') = 
+\[ \mathsf{LTermLKT\,A\, (abs\, B \,C \, e \,s \,T \, t') = 
     (K_\top , K_\top , EqualLArrKT , LTypeLKT\, B\, T , LTermLKT\, C\, t')

### 2021-04-20 induction-paper/ENTCS-GADT-induction.tex
finished MFPS  paper draft; still to be proofread carefully

@@
 that of~\cite{jp19} for nested types, although he does not permit true
-nesting. More recently, Ulrich~\cite{ull20} has implemented a plugin
+nesting. More recently, Ullrich~\cite{ull20} has implemented a plugin
 in MetaCoq to generate induction rules for nested types. This plugin
@@
 argument, rather than the contents of that structure, which exactly
-coindices with what structural induction should do.
+coincides with what structural induction should do.
 %A concrete example of such a derivation is given in
@@
 GADT that is not truly nested. As hinted above, the predicate lifting
-for the equality GADT plays a central role in deriving both s
-tructural and deep induction rules for more general GADTs.
+for the equality GADT plays a central role in deriving both structural
+and deep induction rules for more general GADTs.
 
@@
 \end{array}\]
-In the first clause ahove, $\mathsf{a : A}$, $\mathsf{Q_A : A \to
+In the first clause above, $\mathsf{a : A}$, $\mathsf{Q_A : A \to
   Set}$, $\mathsf{liftA : Seq^{\wedge}\,A\,Q_A\,(const\,a) =
@@
 \end{equation}
-
-\vspace*{-0.05in}
 
@@
 rule for a data type generally relies on (unary) parametricity of the
-model interpreting it, and deel induction for a truly nested type or a
-ruly nested GADT also relies on this interpretation being functorial.
+model interpreting it, and deep induction for a truly nested type or a
+truly nested GADT also relies on this interpretation being functorial.
 But, whereas nested types can admit a functorial parametric semantics,
@@
 \end{align*}
-This predicate uses $\mathsf{Maybe}$ to represent potential failure,
-where $\mathsf{Maybe}$ is defined as follows:
+The predicate $\mathsf{GetType}$ uses the standard $\mathsf{Maybe}$
+data type to represent potential failure. It is defined by:
+
+\vspace*{-0.1in}
+
 \begin{equation}\label{eq:maybe}
@@
 \end{equation}
+
 We want to show that $\mathsf{GetType}$ is satisfied by every element
@@
    GetType \,A \,t}\] This cannot be achieved without deep induction,
-which is required to to apply the induction hyppothesis to the
-individual terms in the list of terms that is an argument to the data
-constructor $\mathsf{list}$ for $\mathsf{LTerm}$.
-
-Using the deep induction rule for $\mathsf{LTerm}$ from
-Section~\ref{sec:ind-lam} we define $\mathsf{getTypeProof}$ by
+which is needed to apply the induction hypothesis to the individual
+terms in the list of terms that the data constructor $\mathsf{list}$
+takes as an argument. Using the deep induction rule
+$\mathsf{dIndLTerm}$ from Section~\ref{sec:ind-lam} we define
+$\mathsf{getTypeProof}$ by
 \[ \mathsf{getTypeProof \,A \,t = 
     dIndLTerm\, P \, cvar\, cabs\, capp\, clistc\, A\, K_\top\, t\,
-    (LTerm\wedge KT\, A\, t) }\] where $\mathsf{t : LTerm\,A}$,
+    (LTerm^\wedge KT\, A\, t) }\] where $\mathsf{t : LTerm\,A}$,
 $\mathsf{P}$ is the polymorphic predicate $\mathsf{\lambda \, (A:
   Set)\, (Q : A \to Set)\, (t : LTerm\,A)\, \to Maybe \, (LType \,
-  A)}$ and $\mathsf{LTerm^\wedge KT : \forall\, (A : Set)\, (t : LTerm
-  A) \to LTerm^{\wedge}\, A\, K_\top\,t}$ is a function that we will
-define later.  We also need the applications to $\mathsf{P}$ of the
-induction hypotheses from Section~\ref{sec:ind-lam} for
-$\mathsf{LTerm}$'s data constructors. These are:
+  A)}$, $\mathsf{K_\top}$ is the constantly {\color{red}
+  $\mathsf{\top}$}-valued predicate, and $\mathsf{LTerm^\wedge KT :
+  \forall\, (A : Set)\, (t : LTerm A) \to LTerm^{\wedge}\, A\,
+  K_\top\,t}$ is a term, to be defined below, witnessing that
+$\mathsf{K_\top}$ can be lifted to all terms. We also need the
+applications to $\mathsf{P}$ of each of the induction hypotheses from
+Section~\ref{sec:ind-lam}. These are:
 \[\begin{array}{lll}
@@
 \end{array}\]
-
-{\color{red} HERE!!}
-
-For variables we simply return the type $\mathsf{T : LType\, A}$ (wrapped in $\mathsf{just}$), 
-and the cases for abstraction and application are similar.
-The interesting case is $\mathsf{clist}$, in which we have to use 
-the results of $\mathsf{List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, ts}$
-in order to extract the type of one of the terms in the list. 
-To define $\mathsf{clist}$ we pattern-match on the list of terms $\mathsf{ts}$.
-If $\mathsf{ts}$ is the empty list $\mathsf{nil}$, we cannot extract a type, 
-so we return $\mathsf{nothing}$:
-\[
-  \mathsf{clist\, A\, B\, Q\, Q'\, e\, nil \, liftE\, lift_{ts} = nothing}
-\]
-where $\mathsf{liftE : Equal^{\wedge}\, A\, (List\,B) \, Q\, (List^{\wedge}\, B\, Q')\, e}$
-and $\mathsf{lift_{ts} : List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, ts}$.
-If $\mathsf{ts}$ is a non-empty list $\mathsf{cons\,t\,ts'}$,
-we pattern match on $\mathsf{list_{ts}}$ and use the result to construct the type we need.
-The type of $\mathsf{list_{ts}}$ is
+In the first clause, $\mathsf{cvar}$ returns $\mathsf{just\,T_A}$.  In
+the second clause, $\mathsf{cabs}$ returns $\mathsf{nothing}$ if its
+final argument is $\mathsf{nothing}$ and $\mathsf{cabs\, A \,B \,C
+  \,Q_A\, Q_B \,Q_C\, e \,s \,T_B\, t_C\, Eq^\wedge\,
+%  LLTy^\wedge B\,
+L\, (just \,T_C) = just \, (arr\, B\, C\, e\, T_B \,T_C)}$ otherwise.
+In the third clause, $\mathsf{capp\, A \,B \, Q_A \, Q_B \, t_{BA}\,
+  t_A \, (just \, (arr\, .B \,.A\, rfl\, T_B\, T_A))\, mb = just\,
+  T_A}$ and $\mathsf{capp}$ returns $\mathsf{nothing}$ otherwise.  In
+the fourth clause, we must use $\mathsf{List^{\wedge}\, (LTerm\, B)\,
+  (GetType\, B)\, ts}$ to extract the type of the head of
+$\mathsf{ts}$ (from which we can deduce the type of the list). When
+$\mathsf{ts} = \mathsf{nil}$ we define $\mathsf{clist\, A\, B\, Q\,
+  Q'\, e\, nil \, liftE\, lift_{ts} = nothing}$, where $\mathsf{liftE
+  : Equal^{\wedge}\, A\, (List\,B) \, Q\, (List^{\wedge}\, B\, Q')\,
+  e}$, and $\mathsf{lift_{ts} : List^{\wedge}\, (LTerm\, B)\,
+  (GetType\, B)\, ts}$.  When $\mathsf{ts} = \mathsf{cons\,t\,ts'}$
+% The interesting case is
+%  $\mathsf{clist}$, in which we have to use the results of
+%  $\mathsf{List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, ts}$ in order
+%  to extract the type of one of the terms in the list.  To define
+%  $\mathsf{clist}$ we pattern-match on the list of terms
+%  $\mathsf{ts}$.  If $\mathsf{ts}$ is the empty list $\mathsf{nil}$,
+%  we cannot extract a type, so we return $\mathsf{nothing}$:
+%\[
+%  \mathsf{clist\, A\, B\, Q\, Q'\, e\, nil \, liftE\, lift_{ts} = nothing}
+%\]
+%where $\mathsf{liftE : Equal^{\wedge}\, A\, (List\,B) \, Q\,
+%  (List^{\wedge}\, B\, Q')\, e}$ and $\mathsf{lift_{ts} :
+%  List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, ts}$.
+%If $\mathsf{ts}$ is a non-empty list $\mathsf{cons\,t\,ts'}$, we
+%pattern match on
+%$\mathsf{list_{ts}}$ and use the result to construct
+%the type we need.  The type of $\mathsf{list_{ts}}$ is
+the type of $\mathsf{list_{ts}}$ becomes $\mathsf{List^{\wedge}\,
+  (LTerm\, B)\, (GetType\, B)\, (cons\, t\, ts')} = \mathsf{GetType\,
+  B\, t \times List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, ts'} =
+\mathsf{Maybe\, (LType\, B) \times List^{\wedge}\, (LTerm\, B)\,
+  (GetType\, B)\, ts'}$. We pattern match on the first component
+of the pair to define\looseness=-1
+%\[\begin{array}{lll}
+%\mathsf{list_{ts}} & : &  
+%  \mathsf{List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, (cons\, t\, ts')}\\
+%& = & \mathsf{GetType\, B\, t \times List^{\wedge}\, (LTerm\, B)\,
+%    (GetType\, B)\, ts'}\\ 
+%&= & \mathsf{Maybe\, (LType\, B) \times List^{\wedge}\, (LTerm\, B)\,
+%    (GetType\, B)\, ts'} 
+%\end{array}\]
 \begin{align*}
-\mathsf{List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, (cons\, t\, ts')}
-&= \mathsf{GetType\, B\, t \times List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, ts'} \\
-&= \mathsf{Maybe\, (LType\, B) \times List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, ts'}
+  &\mathsf{clist\, A\, B\, Q\, Q'\, e\, (cons\, t\, ts') \, liftE\,
+    (nothing , lift_{ts'}) = nothing} \\ 
+  &\mathsf{clist\, A\, B\, Q\, Q'\, e\, (cons\, t\, ts') \, liftE\,
+    (just\, T' , lift_{ts'}) = just \, (TList\, B\, e\, T')} 
 \end{align*}
-Thus, we define
-\begin{align*}
-  &\mathsf{clist\, A\, B\, Q\, Q'\, e\, (cons\, t\, ts') \, liftE\, (nothing , lift_{ts'}) = nothing} \\
-  &\mathsf{clist\, A\, B\, Q\, Q'\, e\, (cons\, t\, ts') \, liftE\, (just\, T' , lift_{ts'}) = just \, (TList\, B\, e\, T')}
-\end{align*}
-where $\mathsf{e : Equal\, A\, (List\,B)}$, $\mathsf{T' : LType\, B}$
-and $\mathsf{lift_{ts'} : List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, ts'}$.
-
-\subsection{Defining $\mathsf{LTermLKT}$}
-
-{\color{red} Maybe this section can be deleted by assuming $\mathsf{LTerm^{\wedge}\, A\, K_\top\,t = K_\top}$.
-Maybe we can say this derives from parametricity.
-Currently we say that $\mathsf{K_\top \times K_\top = K_\top}$ based on the fact that products are a built-in type and so this seems to be obviously true.}
-
-The last piece of infrastructure we need in order to define $\mathsf{getTypeProof}$ is a function 
-\[
-  \mathsf{LTermLKT : \forall\, (A : Set)\, (t : LTerm\, A) \to LTerm^{\wedge}\, A\, K_\top\,t}
-\]
-that provides a 
-proof of $\mathsf{LTerm^{\wedge}\,A\,K_\top\,t}$ for any term $\mathsf{t : LTerm\, A}$. 
-Because $\mathsf{LTerm^{\wedge}}$ is defined in terms of $\mathsf{LType^{\wedge}}$, $\mathsf{Arr^{\wedge}}$, 
-and $\mathsf{List^{\wedge}}$, we need an analogous function for each of these liftings as well
-(respectively, $\mathsf{LTypeLKT}$, $\mathsf{ArrLKT}$ and $\mathsf{ListLKT}$). 
-We only give the definition of $\mathsf{LTermLKT}$, but the definitions for 
-$\mathsf{LTypeLKT}$, $\mathsf{ArrLKT}$ and $\mathsf{ListLKT}$ are analogous. 
-We define $\mathsf{LTermLKT}$ by pattern matching on the lambda term $\mathsf{t}$:
+Here $\mathsf{e : Equal\, A\, (List\,B)}$, $\mathsf{T' : LType\, B}$,
+and $\mathsf{lift_{ts'} : List^{\wedge}\, (LTerm\, B)\, (GetType\,
+  B)\, ts'}$.
+
+\pagebreak
+
+To finish defining $\mathsf{getTypeProof}$ we still need a proof
+
+\vspace*{-0.1in}
+
+\[ \mathsf{LTerm^\wedge KT : \forall\, (A : Set)\, (t : LTerm\, A) \to
+  LTerm^{\wedge}\, A\, K_\top\,t} \]
+
+\vspace*{-0.05in}
+
+\noindent
+Since $\mathsf{LTerm^\wedge}$ is
+defined in terms of $\mathsf{LType^\wedge}$ and $\mathsf{Arr^\wedge}$,
+and since $\mathsf{LType^\wedge}$ is also defined in terms of
+$\mathsf{List^\wedge}$, we need analogous functions
+$\mathsf{LType^\wedge KT}$, $\mathsf{Arr^\wedge KT}$ and
+$\mathsf{List^\wedge KT}$, respectively, for each of these liftings as
+well.  We only give the definition of $\mathsf{LTerm^\wedge KT}$ here
+since $\mathsf{LType^\wedge KT}$, $\mathsf{Arr^\wedge KT}$, and
+$\mathsf{List^\wedge KT}$ are defined analogously. We have:
 \begin{itemize}
-\item For the $\mathsf{var}$ case, let $\mathsf{t = var\, s\, T}$ for $\mathsf{s : String}$ and $\mathsf{T : LType\,A}$,
-and define
-\[
-  \mathsf{LTermLKT\,A\,(var\,s\,T) = LTypeLKT\,A\,T}
-\]
-\item For the $\mathsf{abs}$ case, let $\mathsf{t = abs \,B \,C \, e \,s \,T \, t'} $
-for $\mathsf{e : Equal\,A\,(B \to C)}$, $\mathsf{s : String}$, $\mathsf{T : LType\,B}$ and $\mathsf{t' : LTerm\,C}$,
-and recall the definition of $\mathsf{LTerm^{\wedge}}$ for the $\mathsf{abs}$ constructor, 
-instantiating the predicate $\mathsf{Q : A \to Set}$ to $\mathsf{K_\top}$: 
-% \,a\,Q_a\, (Abs \,b \,c \,e \,s \,T_b \,t_c)}$, :
-\[
-\begin{array}{l}
-\mathsf{LTerm^{\wedge}\,A\,K_\top\,
-  (abs \,B \,C \,e \,s \,T \,t')} \\ 
-\quad\mathsf{
-  = \sum\, [Q' : B \to Set]\, [Q'' : C \to Set]\,
-      Equal^{\wedge} \, A\, (B \to C)\, K_\top\, (Arr^{\wedge} \, B\, C\, Q' \, Q'')\, e
-      \times \, LType^{\wedge}\, B\, Q'\, T
-      \times \, LTerm^{\wedge}\, C\, Q''\, t' }
-\end{array}
-\]
-So, to define the $\mathsf{abs}$ case of $\mathsf{LTermLKT}$, we need a proof of 
-\[
-  \mathsf{Equal^{\wedge} \, A\, (B \to C)\, K_\top\, (Arr^{\wedge} \, B\, C\, Q' \, Q'')\, e}
-\]
-i.e., that $\mathsf{K_\top}$ is (extensionally) equal to the lifting $\mathsf{Arr^{\wedge} \, B\, C\, Q' \, Q''}$
-for some predicates $\mathsf{Q' : B \to Set}$ and $\mathsf{Q'' : C \to Set}$.
-The only reasonable choice for $\mathsf{Q'}$ and $\mathsf{Q''}$ 
-is to let both be $\mathsf{K_\top}$, which means we need a proof of
-\[
-  \mathsf{Equal^{\wedge} \, A\, (B \to C)\, K_\top\, (Arr^{\wedge} \, B\, C\, K_\top \, K_\top)\, e}
-\]
-Since we are working with proof-relevant predicates
-(i.e., functions into $\mathsf{Set}$ rather than functions into $\mathsf{Bool}$), 
-the lifting $\mathsf{Arr^{\wedge} \, B\, C\, K_\top \, K_\top}$ of $\mathsf{K_\top}$ to arrow types
-is not identical to $\mathsf{K_\top}$ on arrow types, but the predicates are (extensionally) isomorphic. 
-We discuss this issue in more detail at the end of the section. 
-For now, we assume a proof 
-\[
-  \mathsf{EqualLArrKT : Equal^{\wedge} \, A\, (B \to C)\, K_\top\, (Arr^{\wedge} \, B\, C\, K_\top \, K_\top)\, e}
-\]
-and define the $\mathsf{abs}$ case of $\mathsf{LTermLKT}$ as 
-\[ \mathsf{LTermLKT\,A\, (abs\, B \,C \, e \,s \,T \, t') = 
-    (K_\top , K_\top , EqualLArrKT , LTypeLKT\, B\, T , LTermLKT\, C\, t')
-  }
-\]
-\item For the $\mathsf{app}$ case, let $\mathsf{t = app\, B\, t' \, t''}$
-for $\mathsf{t' : LTerm\,(B \to A)}$ and $\mathsf{t'' : LTerm\,B}$,
-and, just as we did for the $\mathsf{abs}$ case, 
-recall the definition of $\mathsf{LTerm^{\wedge}\, A\, K_\top\, (app\, B\, t'\, t'')}$
-with all of the predicates instantiated with $\mathsf{K_\top}$: 
-\[
-\mathsf{
-      LTerm^{\wedge}\, (B \to A)\, (Arr^{\wedge} \, B\, A\, K_\top \, K_\top)\, t'
-      \times LTerm^{\wedge}\, B\, K_\top\, t''}
-\]
-The second component can be given using $\mathsf{LTermLKT}$, and we can 
-define the first component using a proof of
-\[
-\mathsf{LTerm^{\wedge}\, (B \to A)\, K_\top\, t'}
-\]
-and a map-like function 
-\[
-  \mathsf{LTermLEqualMap : \forall\, (A : Set)\, (Q\,Q' : A \to Set) 
-    \to Equal^{\wedge}\,A\,A\,Q\,Q'\,refl
-    \to PredMap\,(LTerm\,A)\, (LTerm^{\wedge}\,A\,Q)\,(LTerm^{\wedge}\,A\,Q')}
-\]
-that takes two (extensionally) equal predicates with the same carrier 
-and produces a morphism of predicates between their liftings. The definition is 
-straightforward enough, so we omit the details. 
-Using $\mathsf{LTermLEqualMap}$, we can define the $\mathsf{app}$ case of $\mathsf{LTermLKT}$ as 
-\[
-  \mathsf{LTermLKT\,A\, (app\, B\, t'\, t'') = 
-    (K_\top , LTermLArr , LTermLKT\,B\,t'')}
-\]
-where $\mathsf{ LTermLArr : LTerm^{\wedge}\, (B \to A)\, (Arr^{\wedge} \, B\, A\, K_\top \, K_\top)\, t'}$ is defined as
-\[
-  \mathsf{LTermLArr =  LTermLEqualMap\,K_\top\,(Arr^{\wedge}\,B\,A\,K_\top\,K_\top)\,\, EqualLArrKT\, t'\, L_{K_\top}}
-\]
-where $\mathsf{L_{K_\top} = LTermLKT\, (B \to A)\, t' : LTerm^{\wedge}\, (B \to A) \, K_\top\, t'}$.
-\item For the $\mathsf{listC}$ case, let $\mathsf{t = listC\, B\, e\, ts}$
-for $\mathsf{e : Equal\,A\,(List\,B)}$ and $\mathsf{ts : List\,(LTerm\, B)}$,
-and recall the definition of $\mathsf{LTerm^{\wedge}\,A\,(listC\, B\, e\, ts)}$
-with all of the predicates instantiated to $\mathsf{K_\top}$: 
-\[
-\mathsf{ Equal^{\wedge} \, A\, (List\,B)\, K_\top\, (List^{\wedge} \, B\, K_\top) \, e
-      \times List^{\wedge}\, (LTerm\,B) \, (LTerm^{\wedge} \, B\, K_\top) \, ts }
-\]
-We can give the first component by assuming a proof \,
-$\mathsf{EqualLListKT : Equal^{\wedge} \, A\, (List\,B)\, K_\top\, (List^{\wedge} \, B\, K_\top) \, e}$, 
-but for the second component we again have multiple liftings nested together.
-In this case, we can get a proof of
-\[
-\mathsf{ List^{\wedge}\, (LTerm\,B) \, (LTerm^{\wedge} \, B\, K_\top) \, ts }
-\]
-using $\mathsf{liftListMap}$, as seen in Section~\ref{sec:dind-lterm} {(\color{red} add back-reference)},
-to map a morphism of predicates 
-\[
-  \mathsf{PredMap\, (LTerm\,B)\, (K_\top)\, (LTerm^{\wedge}\,B\,K_\top)}
-\]
-to a morphism of lifted predicates
-\[
-  \mathsf{PredMap\,(List\,(LTerm\,B)) \,(List^{\wedge}\,(LTerm\,B)\,K_\top) \, (List^{\wedge}\,(LTerm\,B)\, (LTerm^{\wedge}\,B\,K_\top))}
-\]
-We then define the $\mathsf{listC}$ case of $\mathsf{LTermLKT}$ as 
-\[
-  \mathsf{LTermLKT\,A\, (listC\, B\, e\, ts) =
-    (K_\top , \, EqualLListKT , \,  L_{ListLLTermLKT} ) }
-\]
-where
-$\mathsf{L_{ListLLTermLKT} : List^{\wedge}\, (LTerm\,B) \, (LTerm^{\wedge} \, B\, K_\top) \, ts}$
-is defined as
-\[
-\mathsf{ 
-  L_{ListLLTermLKT} =
-  liftListMap \, (LTerm\,B) \, K_\top \, (LTerm^{\wedge}\, B\, K_\top) \, m_{K_\top} \, ts \, (ListLKT\, (LTerm\, B)\, ts)
-}
-\]
-and $\mathsf{m_{K_\top} : PredMap\, (LTerm\,B) \, (K_\top)\, (LTerm^{\wedge}\,B\,K_\top)}$
-is defined as
-\[
-  \mathsf{m_{K_\top} \, t'\, tt\, = LTermLKT\, B\, t'}
-\]
-where $\mathsf{t' : LTerm\,B}$ and $\mathsf{tt}$ is the single element of $\mathsf{K_\top\, t'}$. 
-The use of $\mathsf{liftListMap}$ is required in the $\mathsf{listC}$ case because 
-$\mathsf{listC}$ takes an argument of type $\mathsf{List\, (LTerm\, B)}$.
+\item If $\mathsf{s : String}$ and $\mathsf{T : LType\,A}$ we define
+  $\mathsf{LTerm^\wedge KT\,A\,(var\,s\,T)} = \mathsf{LType^\wedge
+  KT\,A\,T}$.
+\item If $\mathsf{e : Equal\,A\,(B \to C)}$, $\mathsf{s : String}$,
+  $\mathsf{T : LType\,B}$, and $\mathsf{t' : LTerm\,C}$ we need to
+  define $\mathsf{LTerm^\wedge KT\,A\, (abs\, B \,C \, e \,s \,T \,
+    t')}$ of type
+
+\vspace*{-0.15in}
+
+  \[\begin{array}{l}
+  \mathsf{LTerm^{\wedge}\,A\,K_\top\, (abs \,B \,C \,e \,s \,T
+  \,t')}\\
+\quad \mathsf{ = \exists [Q_B]\, [Q_C]\, Equal^{\wedge} \, A\, (B \to
+    C)\, K_\top\, (Arr^{\wedge} \, B\, C\, Q_B \, Q_C)\, e \times \,
+    LType^{\wedge}\, B\, Q_B\, T \times \, LTerm^{\wedge}\, C\, Q_C\,
+    t' }
+\end{array}\]
+where $\mathsf{K_\top : A \to Set}$, $\mathsf{Q_B : B \to Set}$, and
+$\mathsf{Q_C : C \to Set}$.  The only reasonable choice is to let both
+$\mathsf{Q_B}$ and $\mathsf{Q_C}$ be $\mathsf{K_\top}$, which means we
+need proofs of $\mathsf{Equal^{\wedge} \, A\, (B \to C)\, K_\top\,
+  (Arr^{\wedge} \, B\, C\, K_\top \, K_\top)\, e}$,
+$\mathsf{LType^{\wedge}\, B\, K_\top\, T}$ and
+$\mathsf{LTerm^{\wedge}\, C\, K_\top\, t'}$.  We take
+$\mathsf{LType^\wedge KT\, B\, T}$ and $\mathsf{LTerm^\wedge KT\, C\,
+  t')}$ for the latter two proofs. For the former we note that, since
+we are working with proof-relevant predicates, the lifting
+$\mathsf{Arr^{\wedge} \, B\, C\, K_\top \, K_\top}$ of
+$\mathsf{K_\top}$ to arrow types is not identical to $\mathsf{K_\top}$
+on arrow types but rather (extensionally) isomorphic.  We discuss this
+issue in more detail at the end of the section, but for now we simply
+assume a proof $\mathsf{Equal^\wedge ArrKT : Equal^{\wedge} \, A\, (B
+  \to C)\, K_\top\, (Arr^{\wedge} \, B\, C\, K_\top \, K_\top)\, e}$
+and define $\mathsf{LTerm^\wedge KT\,A\, (abs\, B \,C \, e \,s \,T \,
+  t') = (K_\top , K_\top , Equal^\wedge ArrKT , LType^\wedge KT\, B\,
+  T , LTerm^\wedge KT\, C\, t') }$.
+\item If $\mathsf{t_1 : LTerm\,(B \to A)}$ and $\mathsf{t_2 :
+  LTerm\,B}$ then, by the same reasoning as in the previous case, we
+  need to define $\mathsf{LTerm^\wedge KT\,A\, (app\, B \, t_1\,t_2)}
+  : \mathsf{LTerm^{\wedge}\, (B \to A)\, (Arr^{\wedge} \, B\, A\,
+    K_\top \, K_\top)\, t_1 \times LTerm^{\wedge}\, B\, K_\top\,
+    t_2}$.  We define the second component of the pair to be
+  $\mathsf{LTerm^\wedge KT\,B\,t_2}$. We can define the first
+  component from a proof of $\mathsf{LTerm^{\wedge}\, (B \to A)\,
+    K_\top\, t_1}$ and the function $\mathsf{LTerm^\wedge EqualMap :
+    \forall\, (A : Set)\, (Q\,Q' : A \to Set) \to
+    Equal^\wedge\,A\,A\,Q\,Q'\,refl \to}$
+  $\mathsf{PredMap\,(LTerm\,A)\,
+    (LTerm^{\wedge}\,A\,Q)\,(LTerm^{\wedge}\,A\,Q')}$ that takes two
+  (extensionally) equal predicates with the same carrier and produces
+  a morphism of predicates between their liftings. The definition of
+  $\mathsf{LTerm^\wedge EqualMap}$ is straightforwardly given by
+  pattern matching on the first two arguments to $\mathsf{PredMap}$ in
+  its return type, using transitivity and symmetry of the type
+  constructor $\mathsf{Equal}$, together with the analogously defined
+  functions $\mathsf{LType^\wedge EqualMap}$ and $\mathsf{Arr^\wedge
+    EqualMap}$ in the cases when the first argument to
+  $\mathsf{PredMap}$ is constructed using $\mathsf{var}$ or
+  $\mathsf{app}$, respectively. Taking $\mathsf{L_{K_\top} :
+    LTerm^{\wedge}\, (B \to A) \, K_\top\, t_1}$ to be the proof
+  $\mathsf{L_{K_\top} = LTerm^\wedge KT\, (B \to A)\, t_1}$ and taking
+  $\mathsf{ LTerm^\wedge Arr : LTerm^{\wedge}\, (B \to A)\,
+    (Arr^{\wedge} \, B\, A\, K_\top \, K_\top)\, t_1}$ to be the proof
+  $\mathsf{LTerm^\wedge Arr =}$ $\mathsf{LTerm^\wedge
+    EqualMap\,K_\top\,(Arr^{\wedge}\,B\,A\,K_\top\,K_\top)\,\,
+    Equal^\wedge ArrKT\, t_1\, L_{K_\top}}$, we define
+  $\mathsf{LTerm^\wedge KT\,A\, (app\, B\, t_1\, t_2) = (K_\top ,
+    LTerm^\wedge Arr , LTerm^\wedge KT\,B\,t_2)}$.
+\item If $\mathsf{e : Equal\,A\,(List\,B)}$ and $\mathsf{ts :
+  List\,(LTerm\, B)}$ then, as above, we need to define
+  $\mathsf{LTerm^\wedge KT\,A\, (list\, B \,e\, ts)}$ $\mathsf{:
+  Equal^{\wedge} \, A\, (List\,B)\, K_\top\, (List^{\wedge} \, B\,
+  K_\top) \, e \times List^{\wedge}\, (LTerm\,B) \, (LTerm^{\wedge} \,
+  B\, K_\top) \, ts }$.  As in that case we assume a proof
+  $\mathsf{EqualLListKT : Equal^{\wedge} \, A\, (List\,B)\, K_\top\,
+  (List^{\wedge} \, B\, K_\top) \, e}$ for the first component.  We
+  can define the second component using $\mathsf{liftListMap}$ from
+  Section~\ref{sec:ind-lam} to map a morphism $\mathsf{PredMap\,
+    (LTerm\,B)\, (K_\top)\, (LTerm^{\wedge}\,B\,K_\top)}$ of
+  predicates to a morphism $\mathsf{PredMap\,(List\,(LTerm\,B))}$
+  $\mathsf{(List^{\wedge}\,(LTerm\,B)\,K_\top) \,
+    (List^{\wedge}\,(LTerm\,B)\, (LTerm^{\wedge}\,B\,K_\top))}$ of
+  lifted predicates.  Taking $\mathsf{m_{K_\top} : PredMap\,
+    (LTerm\,B) \, (K_\top)\, (LTerm^{\wedge}\,B\,K_\top)}$ to be the
+  proof $\mathsf{m_{K_\top} \, t'\, tt\, = LTerm^\wedge KT\, B\, t'}$,
+  where $\mathsf{t' : LTerm\,B}$ and $\mathsf{tt}$ is the single
+  element of $\mathsf{K_\top\, t'}$, and taking
+  $\mathsf{L_{List^\wedge LTerm^\wedge KT} : List^{\wedge}\,
+    (LTerm\,B) \, (LTerm^{\wedge} \, B\, K_\top) \, ts}$ to be the
+  proof $\mathsf{ L_{List^\wedge LTerm^\wedge KT} =}$
+  $\mathsf{liftListMap \, (LTerm\,B) \, K_\top \, (LTerm^{\wedge}\,
+    B\, K_\top) \, m_{K_\top} \, ts \, (List^\wedge KT\, (LTerm\, B)\,
+    ts) }$, we define $\mathsf{LTerm^\wedge KT\,A\, (listC\, B\, e\,
+    ts) = (K_\top , \, Equal^\wedge ListKT , \, L_{ListL^\wedge
+      Term^\wedge KT} ) }$.
 \end{itemize}
 
-{\color{red} These are general considerations, and are not required to conclude the above argument.}
-The above techniques can be used
-to define a function $\mathsf{GLKT : \forall\, (A : Set)\, (x : G\,A) \to G^{\wedge}\, A\, K_\top\, x}$
-for an arbitrary GADT $\mathsf{G}$,
-as defined in Section~\ref{sec:dind-generic} {\color{red} (add back-reference)}.
-To provide a proof of $\mathsf{G^{\wedge}\, A\, K_\top \, x}$ for every term $\mathsf{x : G\, A}$, 
-we need to know that the lifting of $\mathsf{K_\top}$ by any type constructor $\mathsf{F}$
-is extensionally equal to $\mathsf{K_\top}$ on $\mathsf{F}$. 
-For example, we might need a proof that $\mathsf{Pair^{\wedge}\,A\,B\,K_\top\,K_\top}$
-is equal to the predicate $\mathsf{K_\top}$ on $\mathsf{A \times B}$. 
-Given a pair $\mathsf{(a , b) : A \times B}$, we have
-$\mathsf{Pair^{\wedge}\,A\,B\,K_\top\,K_\top (a, b) = K_\top \, a \times K_\top\, b = \top \times \top}$,
-whereas $\mathsf{K_\top\, (a, b) = \top}$.
-While these types are not equal, they are clearly isomorphic.
-So, for simplicity of presentation, we assume $\mathsf{F^{\wedge}\,A\,K_\top}$
-is equal to $\mathsf{K_\top}$ for every nested type and ADT $\mathsf{F}$. 
-Moreover, remember that, whenever $\mathsf{G}$ has a constructor of the form $\mathsf{c : F \, (G\, A) \to G\, (K\,B)}$, 
-$\mathsf{F}$ is only allowed to be a nested type or an ADT,
-and we are guaranteed to have a $\mathsf{liftFMap}$ function.
+The above techniques can be used to define a function
+$\mathsf{G^\wedge KT : \forall\, (A : Set)\, (x : G\,A) \to
+  G^{\wedge}\, A\, K_\top\, x}$ for an arbitrary GADT $\mathsf{G}$, as
+defined in Section~\ref{sec:framework}.  To provide a proof of
+$\mathsf{G^{\wedge}\, A\, K_\top \, x}$ for every term $\mathsf{x :
+  G\, A}$, we need to know that if $\mathsf{G}$ has a constructor
+$\mathsf{c : H \, (\ol{F\,G\, \ol{B}}) \to G\, (\ol{K\,\ol{B}})}$,
+then $\mathsf{H}$ cannot construct GADTs so the generalization
+$\mathsf{H^\wedge Map}$ of $\mathsf{listLiftMap}$ in the final bullet
+point above is guaranteed to exist. We also need to know that the
+lifting of $\mathsf{K_\top}$ to types constructed by any nested type
+constructor $\mathsf{F}$ is extensionally equal to $\mathsf{K_\top}$
+on the types it constructs. For example, we might need a proof that
+$\mathsf{Pair^{\wedge}\,A\,B\,K_\top\,K_\top}$ is equal to
+$\mathsf{K_\top}$ on $\mathsf{A \times B}$.  Given a pair $\mathsf{(a
+  , b) : A \times B}$, we have that
+$\mathsf{Pair^{\wedge}\,A\,B\,K_\top\,K_\top (a, b) = K_\top \, a
+  \times K_\top\, b = \top \times \top}$, whereas $\mathsf{K_\top\,
+  (a, b) = \top}$. While these types are not equal, they are clearly
+isomorphic. Similar isomorphisms between
+$\mathsf{F^{\wedge}\,A\,K_\top}$ and $\mathsf{K_\top}$ hold for all
+other nested type constructors $\mathsf{F}$ as well. These
+isomorphisms can either be proved on an as-needed basis or, since
+$\mathsf{F^\wedge\,A\,K_\top = K_\top}$ is the unary analogue of the
+Identity Extension Lemma, be obtained at the meta-level as a
+consequence of unary parametricity. At the object level, our Agda code
+simply postulates each such isomorphism needed since an Agda
+implementation of full parametricity for some relevant calculus is
+beyond the scope of the present paper.
 
@@
 
+This paper extends deep induction to a robust class of GADTs that are
+not truly nested. It also shows that truly nested GADTs do not admit
+deep induction rules. Our development is implemented in Agda, as is a
+case study showing how deep induction can prove properties of GADTs
+that are not provable by structural induction.\looseness=-1
+
+\begin{comment}
 \section{TODO}
@@
 \end{itemize}
+\end{comment}
+
+\pagebreak
 
@@
   D. Accompanying Agda code for this paper.
-    $\mathtt{https://cs.appstate.edu/~johannp/FoSSaCS21Code.html}$
+    $\mathtt{https://cs.appstate.edu/\tilde~johannp/FoSSaCS21Code.html}$
 
@@
 
-\bibitem{min15} Minsky, Y.  {\em Why {GADT}s matter for performance}.
-  $\mathtt{https://blog.janestreet.com/why-gadts-matter-for-performance/}$,
-  2015.
-
-\bibitem{pl04} Pasalic, E., and Linger, N.  {\em Meta-programming with
-  typed object-language representations}.  Generic Programming and
-  Component Engineering, pp.~136-167, 2004.
-
-\bibitem{pen20} Penner, C.  {\em Simpler and safer {API} design using
-  {GADT}s}.  $\mathtt{https://chrispenner.ca/posts/gadt-design}$,
-  2020.
+%\bibitem{min15} Minsky, Y.  {\em Why {GADT}s matter for performance}.
+%  $\mathtt{https://blog.janestreet.com/why-gadts-matter-for-performance/}$,
+%  2015.
+
+%\bibitem{pl04} Pasalic, E., and Linger, N.  {\em Meta-programming with
+%  typed object-language representations}.  Generic Programming and
+%  Component Engineering, pp.~136-167, 2004.
+
+%\bibitem{pen20} Penner, C.  {\em Simpler and safer {API} design using
+%  {GADT}s}.  $\mathtt{https://chrispenner.ca/posts/gadt-design}$,
+%  2020.
 
@@
 
-\bibitem{pr06} Pottier, F., and R{\'e}gis-Gianas, Y.  {\em Stratified
-  type inference for generalized algebraic data types}.  Principles of
-  Programming Languages, pp.~232-244, 2006.
-
-\bibitem{sjsv09} Schrijvers, T, Peyton Jones, S. L., Sulzmann, M., and
+%\bibitem{pr06} Pottier, F., and R{\'e}gis-Gianas, Y.  {\em Stratified
+%  type inference for generalized algebraic data types}.  Principles of
+%  Programming Languages, pp.~232-244, 2006.
+
+\bibitem{sjsv09} Schrijvers, T., Peyton Jones, S. L., Sulzmann, M., and
   Vytiniotis, D. {\em Complete and decidable type inference for

### 2021-04-21 induction-paper/ENTCS-GADT-induction.tex
test

@@
 case study showing how deep induction can prove properties of GADTs
-that are not provable by structural induction.\looseness=-1
+that are not provable just by structural induction.\looseness=-1
 

### 2021-04-30 induction-paper/ENTCS-GADT-induction.tex
minor revisions ; rename LType constructors with 't' prefix

--- 2021-04-20 induction-paper/ENTCS-GADT-induction.tex
@@
   \vspace*{-0.1in}
-  \title{(Deep) Induction for GADTs\vspace*{-0.1in}}
+  \title{(Deep) Induction Rules for GADTs\vspace*{-0.1in}}
   \author{Patricia Johann~~~~}%\thanksref{JohannEmail}}
@@
 \begin{abstract} 
-Deep data types are data types that are defined in terms of other such
-data types, including, in the case of truly nested types, themselves.
-Deep induction is an extension of structural induction that traverses
-{\em all} of the structure present in a structure of such a type,
-propagating suitable predicates to {\em all} of the data contained in
-that structure. Deep induction has been shown to be the form of
-induction most suitable for applications involving deep nested
-types. In this paper we show how to extend deep induction to a robust
-class of deep GADTs that are not truly nested. We also show that it
-cannot be extended to truly nested GADTs.
+  Deep data types those that are defined in terms of other data types,
+  including, possibly, themselves.  In that case, they are said to be
+  truly nested.  Deep induction is an extension of structural
+  induction that traverses {\em all} of the structure in a deep data
+  type, propagating predicates on its primitive data throughout the
+  entire structure.  Deep induction was recently shown to prove
+  properties of nested types, including truly nested types, that
+  cannot be proved via structural induction.  In this paper we show
+  how to extend deep induction to deep GADTs that are not truly
+  nested. We also show that deep induction cannot be extended to truly
+  nested GADTs.\looseness=-1
 \end{abstract}
@@
 and almost all non-trivial proofs of their properties are either
-proved by induction outright or rely on lemmas that are. Every time a
-new inductive data type is declared in Coq, an induction rule is
-automatically generated for it.
+proved by induction outright or rely on lemmas that are.  {\color{red}
+  Every time a new algebraic data type (ADT) is declared in
+  Coq}\footnote{{\color{red} Algebraic data types are 
+  defined using the $\mathsf{Inductive}$ keyword in Coq.}}
+  % called inductive
+  %   types in Coq.}}
+    , an induction rule is automatically generated for
+it.\looseness=-1
 
 The data types handled by Coq are (possibly mutually inductive)
-polynomial ADTs, and the induction rules automatically generated for
-them are the expected ones for standard structural induction. It has
-long been understood, however, that these rules are too weak to be
-genuinely useful for so-called {\em deep ADTs}~\cite{jp20}, i.e., ADTs
-that are (possibly mutually inductively) defined in terms of (other)
-such ADTs.\footnote{Such data types are called nested inductive types
-  by Chlipala~\cite{chl}, reflecting the fact that ``inductive type''
-  means ``ADT''~in~Coq.} Consider, for example, the following type of
-rose trees, here coded in Agda and defined in terms of the standard
-type of lists:
+polynomial ADTs, and the induction rules
+Coq generates for them are the expected ones for standard structural
+induction. It has long been understood, however, that these rules are
+too weak to be genuinely useful for so-called {\em deep
+  ADTs}~\cite{jp20}, i.e., ADTs that are (possibly mutually
+inductively) defined in terms of (other) such ADTs.
+%\footnote{Such data
+%  types are called nested inductive types by Chlipala~\cite{chl},
+%  reflecting the fact that ``inductive type'' means ``ADT''~in~Coq.}
+Consider, for example, the following type of rose trees, here coded in
+{\color{red} Agda} and defined in terms of the standard type
+$\mathsf{List}$ of lists (see Section~\ref{sec:ADTs-and-nesteds}):
 
@@
 \mathsf{\;\;\;\;\;\;\;\;empty\, :\, Rose\,A}\\
-\mathsf{\;\;\;\;\;\;\;\;node\,\,\,\, :\, A \to List\,(Rose\,A) \to Rose\,A} 
+\mathsf{\;\;\;\;\;\;\;\;\,node\,\,\,\, :\, A \to List\,(Rose\,A) \to Rose\,A} 
 \end{array}\]
 \noindent
-The induction rule Coq automatically generates for rose trees is
+The induction rule Coq automatically generates for the analogous Coq definition of rose trees is
 
@@
 \[\begin{array}{l}
-\mathsf{\forall\, (a : Set)\,(P : Rose\,a \to Set) \to P\,empty \to}\\
+\mathsf{\forall\, (A : Set)\,(P : Rose\,A \to Set) \to P\,empty \to}\\
 \mathsf{\hspace*{0.15in}
- (\forall\, (x : a)\,(ts :
-  List\,(Rose\,a)) \to P\,(node\,x\,ts)) \to 
-  \forall \,(x : Rose\,a) \to P\,x}
+ (\forall\, (a : A)\,(ts :
+  List\,(Rose\,A)) \to P\,(node\,a\,ts)) \to 
+  \forall \,(x : Rose\,A) \to P\,x}
 \end{array}\]
@@
 $\mathsf{node}$'s list argument. More generally, this enhanced notion
-of induction should traverse {\em all} of the structure present in a
-data element, propagating suitable predicates to {\em all} of the data
-contained in the structure. With data types becoming ever more
-advanced, and with deeply structured such types becoming ever more
-%increasingly
-ubiquitous in formalizations, it is critically important that proof
-assistants
-%like Coq, Agda, and Lean 
-be able to automatically generate genuinely useful induction
-rules for data types that go well beyond traditional ADTs. Such
-data types include (truly) nested types~\cite{bm98}\footnote{A truly
-  nested type is a nested type that is defined over itself.},
+of induction should traverse {\em all} of the layers present in a data
+structure, propagating suitable predicates to {\em all} of the data it
+contains. With data types becoming ever more advanced, and with deeply
+structured such types becoming increasingly ubiquitous in
+formalizations, it is critically important that proof assistants
+%{\color{red} (e.g., Coq, Agda, Lean)}
+are able to automatically generate genuinely useful induction rules
+for data types that go well beyond traditional ADTs. Such data types
+include nested types~\cite{bm98}, such as $\mathsf{Rose}$ above, as
+well as truly nested types\footnote{A truly nested type is a nested
+  type that is defined over itself. The data type $\mathsf{Bush}$ in
+  Section~\ref{sec:ADTs-and-nesteds} provides a concrete example.},
 generalized algebraic data types
@@
 \begin{array}{l}
-\mathsf{\forall \,(a : Set)\,(P : Rose\,a \to Set)\,(Q : a \to Set)
+\mathsf{\forall \,(A : Set)\,(P : Rose\,A \to Set)\,(Q : A \to Set)
   \to P\,empty \to}\\ 
-\mathsf{\hspace*{0.15in}(\forall \,(x : a)\, (ts :
-  List\,(Rose\,a))\to Q\,x \to List^\land\,P\,ts \to P\,(node\,x\,ts)) \to}\\
+\mathsf{\hspace*{0.15in}(\forall \,(a : A)\, (ts :
+  List\,(Rose\,A))\to Q\,a \to List^\land\,P\,ts \to P\,(node\,a\,ts)) \to}\\
 \mathsf{\hspace*{0.15in}\forall \,(x :
-  Rose\,a) \to Rose^\land\,Q\, x \to P\,x} 
+  Rose\,A) \to Rose^\land\,Q\, x \to P\,x} 
 \end{array}
@@
 predicate argument $\mathsf{P}$ (resp., $\mathsf{Q}$) on data of type
-$\mathsf{Rose\,a}$ (resp., $\mathsf{a}$) to a predicate on data of
-type $\mathsf{List\,(Rose\,a)}$ (resp., $\mathsf{Rose\,a}$) asserting
+$\mathsf{Rose\,A}$ (resp., $\mathsf{A}$) to a predicate on data of
+type $\mathsf{List\,(Rose\,A)}$ (resp., $\mathsf{Rose\,A}$) asserting
 that $\mathsf{P}$ (resp., $\mathsf{Q}$) holds for every element of its
@@
 This paper shows how to extend deep induction to proper GADTs, i.e.,
-to GADTs that are not simply nested types (and thus are not ADTs).  A
-constructor for such a GADT $\mathsf{G}$ may, like a constructor for a
-nested type, take as arguments data whose types involve instances of
-$\mathsf{G}$ other than the one being defined --- including instances
-that involve $\mathsf{G}$ itself. But if $\mathsf{G}$ is a proper GADT
-then at least one of its constructors will also have such a structured
-instance of $\mathsf{G}$ --- albeit one not involving $\mathsf{G}$
-itself --- as its codomain.
+to GADTs that are not simply (truly) nested types (and thus are not
+ADTs).  A constructor for such a GADT $\mathsf{G}$ may, like a
+constructor for a nested type, take as arguments data whose types
+involve instances of $\mathsf{G}$ other than the one being defined ---
+including instances that involve $\mathsf{G}$ itself. But if
+$\mathsf{G}$ is a proper GADT then at least one of its constructors
+will also have such a structured instance of $\mathsf{G}$ --- albeit
+one not involving $\mathsf{G}$ itself --- as its codomain.
 %But while the types of the arguments to a GADT's constructor can
@@
 %$\mathsf{G}$, its return type cannot.
-For example, the constructor $\mathsf{pair}$ for the GADT {\color{red}
-  Perhaps also show non-inhabitation?}
+For example, the constructor $\mathsf{pair}$ for the GADT
 
@@
 \begin{array}{l}
-\mathsf{data\, Seq\,(a : Set) : Set\,where}\\
-\mathsf{\;\;\;\;\;\;\;\;const :\, a \to Seq\,a}\\
-\mathsf{\;\;\;\;\;\;\;\;\;\;\,pair :\, Seq \,a \to Seq\,b \to
-  Seq\,(a \times b)}
+\mathsf{data\, Seq\,(A : Set) : Set\,where}\\
+\mathsf{\;\;\;\;\;\;\;\;const :\, A \to Seq\,A}\\
+\mathsf{\;\;\;\;\;\;\;\;pair\;\;\, :\, Seq \,A \to Seq\,B \to
+  Seq\,(A \times B)}
 \end{array}
@@
 extensions~\cite{mac71}, and a parametric semantics interpreting them
-as their Church encodings~\cite{atk12,vw10}. As explained
+as their Church encodings~\cite{atk12,vw10}. As detailed
 in~\cite{jgj21}, a key difference in the two semantics is that the
@@
 This observation seems, at first, to be a death knell for the prospect
-of extending deep induction to GADTs. Indeed, since induction can be
-seen as unary parametricity, we quickly realize that GADTs viewed as
-their functorial completions cannot possibly support induction rules.
-This makes sense intuitively: induction is a syntactic proof
-technique, so of course it cannot be used to prove properties of those
-elements of a GADT's functorial completion that are not expressible in
-syntax. All is not lost, however. As we show below, the Church
-encoding interpretation's syntax-only view does support induction
+of extending deep induction to GADTs. Indeed, induction can be seen as
+unary parametricity, so we quickly realize that GADTs viewed as their
+functorial completions cannot possibly support induction rules.  This
+makes sense intuitively: induction is a syntactic proof technique, so
+of course it cannot be used to prove properties of those elements of a
+GADT's functorial completion that are not expressible in syntax. All
+is not lost, however. As we show below, the syntax-only view of GADTs
+determined by their Church encodings {\em does} support induction
 rules --- including deep induction rules --- for GADTs. Perhaps
-surprisingly, ours are the first-ever induction rules --- deep or
-otherwise --- for a general class of proper GADTs. But this paper
-actually delivers more: it gives a general framework for deriving deep
-induction rules for a general class of deep GADTs directly from their
-syntax. This framework can serve as a basis for extending modern proof
-assistants' automatic generation of structural induction rules for
-ADTs to automatic generation of deep induction rules for GADTs. As for
-ADTs and other nested types, the structural induction rule for any
-GADT can be recovered from its deep induction rule simply by taking
-the custom predicates in its deep induction rule to be constantly
-$\mathsf{True}$-valued predicates.
-
-Deep induction rules for GADTs cannot, however, be derived by somehow
-extending the techniques of~\cite{jp20} to syntax-only GADTs. Indeed,
-the derivation of induction rules given there makes crucial use of the
+surprisingly, this paper gives the first-ever induction rules --- deep
+or otherwise --- for proper GADTs. But it actually delivers far more:
+it gives a general framework for deriving deep induction rules for
+deep GADTs directly from their syntax. This framework can serve as a
+basis for extending modern proof assistants' automatic generation of
+structural induction rules for ADTs to automatic generation of deep
+induction rules for GADTs. As for ADTs and other nested types, the
+structural induction rule for any GADT can be recovered from its deep
+induction rule simply by taking the custom predicates in its deep
+induction rule to be constantly $\mathsf{True}$-valued (i.e.,
+constantly $\mathsf{\top}$-valued) predicates.
+
+Significantly, deep induction rules for GADTs cannot be derived by
+somehow extending the approach of~\cite{jp20} to syntax-only
+GADTs. Indeed, the approach taken there makes crucial use of the
 functoriality of data types' interpretations from~\cite{jp19}, and
-that is precisely what the interpretation of GADTs as their Church
-encodings fails to deliver. Instead, we first give a predicate lifting
-styled after those of~\cite{jp20}, together with a (deep) induction
-rule, and for the simplest --- and arguably most important --- GADT,
-namely the equality GADT. (See Section~\ref{sec:ind-equal}.) We can
-then derive the deep induction rule for any other GADT $\mathsf{G}$ by
-{\em i}) using the equality GADT to represent $\mathsf{G}$ as its
-so-called {\em Henry Ford
-  encoding}~\cite{ch03,hin03,mcb99,sjsv09,sp04}, and {\em ii}) using
-the predicate liftings for the equality GADT and any other GADTs
-appearing in the definition of $\mathsf{G}$ to appropriately thread
+functoriality is precisely what interpreting GADTs as their Church
+encodings fails to deliver. Our approach is to instead first give a
+predicate lifting styled after those of~\cite{jp20}, together with a
+(deep) induction rule, and for the simplest --- and arguably most
+important --- GADT, namely the equality GADT. (See
+Section~\ref{sec:ind-equal}.)  We then derive the deep induction rule
+a more complex GADT $\mathsf{G}$ by {\em i}) using the equality GADT
+to represent $\mathsf{G}$ as its so-called {\color{red} {\em Henry Ford
+  encoding}}~\cite{ch03,hin03,mcb99,sjsv09,sp04}, and {\em ii}) using
+the predicate liftings for the equality GADT, and any other GADTs
+appearing in the definition of $\mathsf{G}$, to appropriately thread
 the custom predicates for the primitive types appearing in
-$\mathsf{G}$ through its structure. This two-step process delivers
-deep induction rules for a broad class of deep GADTs. In
-Section~\ref{sec:GADTs} we introduce a series of increasingly complex
-GADTs as running examples, and in Section~\ref{sec:deep-ind-GADTs} we
-derive a deep induction rule for each of them. In particular, we
-derive the deep induction rule for $\mathsf{Seq}$ in
-Section~\ref{sec:ind-seq}. We present our general framework for
-deriving (deep) induction rules for (deep) GADTs in
+$\mathsf{G}$ throughout $\mathsf{G}$'s structure. This two-step
+process delivers deep induction rules for a very general class of deep
+GADTs. In Section~\ref{sec:GADTs} we introduce a series of
+increasingly complex GADTs as running examples, and in
+Section~\ref{sec:deep-ind-GADTs} we derive a deep induction rule for
+each of them. In particular, we derive the deep induction rule for
+$\mathsf{Seq}$ in Section~\ref{sec:ind-seq}. We present our general
+framework for deriving (deep) induction rules for (deep) GADTs in
 Section~\ref{sec:framework}, and observe that the derivations in
@@
 over themselves do not, to our knowledge, appear in applications or
-the literature. Section~\ref{sec:app} details a case study in using
+the literature. Section~\ref{sec:app} comprises a case study in using
 deep induction. All of the deep induction rules appearing in this
-paper have been derived using our general framework. Our Agda code
-implementing them is available at~\cite{web-page}.
+paper have been derived by instantiating our general framework. Our
+Agda code implementing them is available at~\cite{web-page}.
 
@@
 technique is rather {\em ad hoc}, so is unclear how to generalize it
-to nested types other than the specific ones in the
-examples. Moreover, it actually derives induction rules for data types
-{\em related} to the original nested types rather than for the
-original nested types themselves, and it is unclear whether or not the
-derived rules are sufficiently expressive to prove all results about
-the original nested types that we would expect to be provable by
+to nested types other than the specific ones in their
+examples. Moreover,~\cite{fs18} actually derives induction rules for
+data types {\em related} to the original nested types rather than for
+the original nested types themselves, and it is unclear whether or not
+the derived rules are sufficiently expressive to prove all results
+about the original nested types that we would expect to be provable by
 induction. This latter point echoes the issue with Coq-derived
@@
 induction (and other) rules for such types for use in that
-system. Tassi~\cite{tas19} has done exactly that, deriving induction
-rules for data type definitions in Coq using unary
-parametricity. Tassi's technique seems to be essentially equivalent to
-that of~\cite{jp19} for nested types, although he does not permit true
-nesting. More recently, Ullrich~\cite{ull20} has implemented a plugin
-in MetaCoq to generate induction rules for nested types. This plugin
-is also based on unary parametricity and, again, true nesting is not
-permitted.  As far as we know, no attempt has yet been made to extend
-either implementation to GADTs. In fact, we know of no work other than
-that reported here that specifically addresses induction rules for
-(deep) GADTs.
+system. Tassi~\cite{tas19} derives induction rules for data type
+definitions in Coq using unary parametricity. His technique seems to
+be essentially equivalent to that of~\cite{jp19} for nested types,
+although he does not permit true nesting. More recently,
+Ullrich~\cite{ull20} has implemented a plugin in MetaCoq to generate
+induction rules for nested types. This plugin is also based on unary
+parametricity and true nesting still is not permitted.  As far as we
+know, no attempt has yet been made to extend either implementation
+truly nested types or to proper GADTs. In fact, we know of no work
+other than that reported here that specifically addresses induction
+rules for (deep) GADTs.
 
@@
 to the set of proofs that $\mathsf{P\,a}$ holds.  For example, the
-induction rule for the standard list type
+structural induction rule for the list type
 
@@
 \mathsf{data\ List : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;nil\,\,\,\,\,\, :\, List\,A}\\
+\mathsf{\;\;\;\;\;\;\;nil\,\,\,\,\,\; :\, List\,A}\\
 \mathsf{\;\;\;\;\;\;\;cons\, :\, A \to List\,A \to List\,A} 
@@
 \noindent
-As in Coq's induction rule for rose trees, the data inside a
-structure of type $\mathsf{List}$ is treated monolithically (i.e.,
-ignored) by this structural induction rule.
-By contrast, the deep induction rule for lists is parameterized over a
-custom predicate $\mathsf{Q}$ on $\mathsf{A}$ as described in the
-introduction. For $\mathsf{List^\wedge}$ as described in the
-introduction it is
+As in Coq's induction rule for rose trees, the data inside a structure
+of type $\mathsf{List}$ is treated monolithically (i.e., is ignored)
+by this structural induction rule.  By contrast, the deep induction
+rule for lists is parameterized over a custom predicate $\mathsf{Q}$
+on $\mathsf{A}$. For $\mathsf{List^\wedge}$ as described in the
+introduction the deep induction rule for lists is
 
@@
 \mathsf{\forall (A : Set) (P : List\, A \to Set) (Q : A \to Set)
-\to P\,Nil \to \big( \forall (a : A) (as: List\,A) \to Q\,a \to P\,as
-\to P\,(Cons\,a\,as)\big)} \\ 
+\to P\,nil \to \big( \forall (a : A) (as: List\,A) \to Q\,a \to P\,as
+\to P\,(cons\,a\,as)\big)} \\ 
 \quad\mathsf{\to \forall (as : List\,A) \to List^{\wedge}\,A\,Q\,as
@@
 \mathsf{data\ PTree : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;pleaf\,\,\, :\, A \to PTree\,A}\\
+\mathsf{\;\;\;\;\;\;\;pleaf\,\,\; :\, A \to PTree\,A}\\
 \mathsf{\;\;\;\;\;\;\;pnode\, :\, PTree\,(A \times A) \to PTree\,A} 
@@
 \to \big( \forall (A : Set) (a : A) \to P\,A\,(pleaf\, a) \big)} \\
-\quad\mathsf{\to \big( \forall (A : Set) (tt : PTree\,(A \times A))
-  \to P\,(A \times A)\,tt \to P\,a\,(pnode\,tt)\big) \to \forall (A :
-  Set) (t : PTree\,A) \to P\,A\,t } 
+\quad\mathsf{\to \big( \forall (A : Set) (pp : PTree\,(A \times A))
+  \to P\,(A \times A)\,pp \to P\,a\,(pnode\,pp)\big) \to \forall (A :
+  Set) (p : PTree\,A) \to P\,A\,p } 
 \end{array}\]
@@
   Q\,a \to P\,A\,Q\,(Pleaf\, a) \big)} \\ \quad \mathsf{\to \big(
-  \forall (A : Set) (Q : A \to Set) (tt : PTree\,(A \times A)) \to
-  P\,(A \times A)\,(Pair^{\wedge}\,A\,A\,Q\,Q)\,tt \to
-  P\,A\,Q\,(Pnode\,tt)\big)} \\ \quad \mathsf{\to \forall (A : Set) (Q
-  : A \to Set) (t : PTree\,A) \to PTree^{\wedge}\,A\,Q\,t \to
-  P\,A\,Q\,t }
+  \forall (A : Set) (Q : A \to Set) (pp : PTree\,(A \times A)) \to
+  P\,(A \times A)\,(Pair^{\wedge}\,A\,A\,Q\,Q)\,pp \to
+  P\,A\,Q\,(Pnode\,pp)\big)} \\ \quad \mathsf{\to \forall (A : Set) (Q
+  : A \to Set) (p : PTree\,A) \to PTree^{\wedge}\,A\,Q\,p \to
+  P\,A\,Q\,p }
 \end{array}\]
@@
 below themselves. The quintessential example of such a type is that of
-bushes~\cite{bm98}:
+bushes
+% \footnote{To define $\mathsf{Bush}$ and other truly nested types in Agda, 
+% we must use the $\mathsf{NO\_POSITIVITY\_CHECK}$
+% pragma . Similar flags are required in Coq.}
+~\cite{bm98}:
 \begin{equation*}\label{eq:bush}
@@
 \mathsf{data\ Bush : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;bnil\,\,\,\,\,\, :\, Bush\,A}\\
+\mathsf{\;\;\;\;\;\;bnil\,\,\,\,\,\; :\, Bush\,A}\\
 \mathsf{\;\;\;\;\;\;bcons\, :\, A \to Bush\,(Bush\,A) \to Bush\,A} 
@@
 \noindent
+% {\color{red} Flags to overcome positivity restrictions?}
+
 Even defining a structural induction rule for bushes requires that we
@@
 is worth noting that for those nested types that do admit distinct
-structural induction and deep induction rules, the latter generalizes
-the former. Indeed, the structural induction rule for such a nested
-type is recoverable from its deep induction rule by taking the custom
-predicates on its data of primitive types to be constantly
-$\mathsf{True}$-valued predicates. This instantiation ensures that the
+structural induction and deep induction rules, the latter always
+generalize the former. Indeed, the structural induction rule for each
+such nested type is recoverable from its deep induction rule by taking
+the custom predicates on its data of primitive types to be constantly
+$\mathsf{\top}$-valued predicates. This instantiation ensures that the
 resulting induction rule only inspects the top-level structure of its
@@
 type instance being defined. For example, every data constructor for
-$\mathsf{PTree\,A}$ must return an element of type
+$\mathsf{PTree\,A}$ returns an element of type
 $\mathsf{PTree\,A}$, regardless of the instances of $\mathsf{PTree}$
@@
 allowing their data constructors both to take as arguments \emph{and
-  return as results} data whose types involve instances of them other
-than the one being defined. And as with the return type of
-$\mathsf{pair}$ in~\eqref{eq:seq}, these instances can be structured.
+  return as results} data whose types involve instances other
+than the one being defined. That is, GADTs' constructors' return type
+instances can, like that of $\mathsf{pair}$ in~\eqref{eq:seq}, be
+structured.\looseness=-1
+%these instances can be structured.
+%
 %Applications of GADTs include generic programming, modeling
@@
 \end{equation}
-is parametrized by two type indices, but it is only possible to
+is parameterized by two type indices, but it is only possible to
 construct data elements of type $\mathsf{Equal\,a\,b}$ if $\mathsf{a}$
@@
 \mathsf{\;\;\;\;\;\;\;const :\, A \to Seq\,A}\\ 
-\mathsf{\;\;\;\;\;\;\;\;pair\,\, :\, \forall (B\,C : Set) \to
+\mathsf{\;\;\;\;\;\;\;pair\,\;\; :\, \forall (B\,C : Set) \to
   Equal\,A\,(B \times C) \to Seq\,B \to Seq\,C \to Seq\,A}\\ 
@@
 $\mathsf{Seq}$ that is \emph{equal} to a product type. As we will see
-in Section~\ref{sec:deep-ind-GADTs}, the presence of the equality GADT is
-key to deriving deep induction rules for GADTs.
+in Section~\ref{sec:deep-ind-GADTs}, this encoding in terms of the
+equality GADT is key to deriving deep induction rules for GADTs.
 
 Neither $\mathsf{Equal}$ nor $\mathsf{Seq}$ is a deep GADT, but the
-following GADT $\mathsf{LTerm}$, which encodes terms of a simply typed
-lambda calculus, is. More robust variations on $\mathsf{LTerm}$ are,
-of course, possible. But since this variation is rich enough to
-illustrate all essential aspects of deep GADTs --- and later, in
-Section~\ref{sec:ind-lam}, their deep induction rules --- while still
-being small enough to ensure clarity of exposition, we keep it to a
-minimum.
+following GADT $\mathsf{LTerm}$, inspired by~\cite{cis194}, is.  It
+encodes terms of a simply typed lambda calculus. More robust
+variations on $\mathsf{LTerm}$ are, of course, possible. But since
+this variation is rich enough to illustrate all essential aspects of
+deep GADTs --- and later, in Section~\ref{sec:ind-lam}, their deep
+induction rules --- while still being small enough to ensure clarity
+of exposition, we keep it to a minimum.
 
@@
 \mathsf{data\ LType : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;\;bool :\, \forall (B : Set) \to Equal\,A\,Bool
+\mathsf{\;\;\;\;\;\;\;\;tbool :\, \forall (B : Set) \to Equal\,A\,Bool
   \to LType\,A}\\ 
-\mathsf{\;\;\;\;\;\;\;\;arr\,\,\, :\, \forall (B\,C : Set) \to
+\mathsf{\;\;\;\;\;\;\;\;tarr\;\;\; :\, \forall (B\,C : Set) \to
   Equal\,A\,(B \to C) \to LType\,B \to LType\,C \to LType\,A}\\ 
-  \mathsf{\;\;\;\;\;\;\;\;list\,\, :\, \forall (B : Set) \to
+  \mathsf{\;\;\;\;\;\;\;\;tlist\,\;\; :\, \forall (B : Set) \to
     Equal\,A\,(\List\,B) \to LType\,B \to LType\,A} 
@@
 \mathsf{data\ LTerm : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;\;var\,\,\,\,\,\,\,:\, String \to LType\,A \to
+\mathsf{\;\;\;\;\;\;\;\;var\,\,\,:\, String \to LType\,A \to
   LTerm\,A} \\  
-\mathsf{\;\;\;\;\;\;\;\;abs\,\,\,\,\,\, :\, \forall (B\,C : Set) \to
+\mathsf{\;\;\;\;\;\;\;\;abs\,\, :\, \forall (B\,C : Set) \to
   Equal\,A\,(B \to C) \to String \to LType\,B \to LTerm\,C \to
   LTerm\,A}\\ 
-  \mathsf{\;\;\;\;\;\;\;\;app\,\,\,\,\, :\, \forall (B : Set) \to
+  \mathsf{\;\;\;\;\;\;\;\;app\; :\, \forall (B : Set) \to
     LTerm (B \to A) \to LTerm\,B \to LTerm\,A} \\ 
-  \mathsf{\;\;\;\;\;\;\;\;list\,\,\, :\, \forall (B : Set) \to
+  \mathsf{\;\;\;\;\;\;\;\;list\;\,\, :\, \forall (B : Set) \to
     Equal\,A\,(List\,B) \to List\,(LTerm\,B) \to LTerm\,A} 
@@
 we first need to define a predicate lifting that maps a predicate on a
-type $\mathsf{A}$ and to a predicate on $\mathsf{G\,A}$. Such a
-predicate lifting $\mathsf{Equal^{\wedge} : \forall (A\,B : Set) \to
-  (A \to Set) \to (B \to Set) \to Equal\,A\,B \to Set}$ for
-$\mathsf{Equal}$ is defined by
-$\mathsf{Equal^{\wedge}\,A\,A\,Q\,Q'\,refl = \forall (a : A) \to
-  Equal\,(Q\,a)(Q'\,a)}$.  It takes two predicates on the same type as
-input and tests them for extensional equality.
+type $\mathsf{A}$ to a predicate on $\mathsf{G\,A}$. Such a predicate
+lifting $\mathsf{Equal^{\wedge} : \forall (A\,B : Set) \to (A \to Set)
+  \to (B \to Set) \to Equal\,A\,B \to Set}$ for $\mathsf{Equal}$ is
+defined by $\mathsf{Equal^{\wedge}\,A\,A\,Q\,Q'\,refl = \forall (a :
+  A) \to Equal\,(Q\,a)(Q'\,a)}$.  It takes two predicates on the same
+type as input and tests them for extensional equality.
 %
@@
 \end{equation}
+
 To prove that this rule is sound we must provide a witness
@@
 $\mathsf{dIndEqual\;P\;crefl\;A\;A\;Q_A\;Q_A'\;refl\;liftE =
-  crefl\;A\;Q_A\;Q_A'\,liftE}$. We can recover the structural
+  crefl\;A\;Q_A\;Q_A'\,liftE}$. To recover $\mathsf{Equal}$'s structural
 induction rule
@@
 \end{equation}
-for $\mathsf{Equal}$ by defining a term $\mathsf{indEqual}$ of the
-type in~\eqref{eq:sind-equal} by $\mathsf{indEqual\;Q\;srefl\;A\;B\;e
-  = dIndEqual\;P\;srefl\;A\;B\;K^A_\top\;K^B_\top\;e\,liftE}$. Here,
-$\mathsf{e : Equal\,A\,B}$, $\mathsf{P : \forall (A\,B : Set) \to (A
-  \to Set) \to (B \to Set) \to Equal\,A\,B \to Set}$ is defined by
-$\mathsf{P\;A\;B\;Q_A\;Q_B\;e =}$ $\mathsf{Q\;A\;B\;e}$,
+we define a term $\mathsf{indEqual}$ of the type
+in~\eqref{eq:sind-equal} by $\mathsf{indEqual\;Q\;srefl\;A\;B\;refl =
+  dIndEqual\;P\;srefl'\;A\;B\;
+  K^A_\top\;K^B_\top\;refl\,sliftE}$. Here, $\mathsf{P : \forall (A\,B
+  : Set) \to (A \to Set) \to (B \to Set) \to Equal\,A\,B \to Set}$ is
+defined by $\mathsf{P\;A\;B\;Q_A\;Q_B\;e =}$ $\mathsf{Q\;A\;B\;e}$,
 $\mathsf{K^A_\top}$ and $\mathsf{K^B_\top}$ are the constantly
-{\color{red} $\mathsf{\top}$}-valued predicates on $\mathsf{A}$ and
-$\mathsf{B}$, respectively, and $\mathsf{liftE :}$
-$\mathsf{Equal^{\wedge}\;A\;B\;K^A_\top\;K^B_\top\;e}$ is
-%fully
-defined
-%(by pattern matching)
-by $\mathsf{liftE\,a = refl : Equal\,A\,A}$ for every $\mathsf{a :
-  A}$.  The structural induction rule for any GADT $\mathsf{G}$ that
-is not truly nested can similarly be recovered from its deep induction
-rule by instantiating every custom predicate by the appropriate
-constantly {\color{red}$\mathsf{\top}$}-valued predicate.
+$\mathsf{\top}$-valued predicates on $\mathsf{A}$ and $\mathsf{B}$,
+respectively, $\mathsf{sliftE :}$ $\mathsf{Equal^{\wedge}\;A\;B\;
+  K^A_\top\;K^B_\top\;refl}$ is defined by $\mathsf{sliftE\,a = refl :
+  Equal\, \top\,\top}$ for every $\mathsf{a : A}$, and $\mathsf{srefl'
+  : \forall (C\,: Set) (Q_c\,Q'_c\, : C \to Set) \to
+  Equal^{\wedge}\,C\,C\,Q_c\,Q'_c\,refl \to \,Q\,C\,C\,refl}$ is
+defined by $\mathsf{srefl'\,C\,Q_c\,Q'_c\,liftE'\,=srefl\, C}$.  The
+structural induction rule for any GADT $\mathsf{G}$ that is not truly
+nested can similarly be recovered from its deep induction rule by
+instantiating every custom predicate by the appropriate constantly
+$\mathsf{\top}$-valued predicate.
 
@@
 \lambda (P : \forall (A : Set) \to (A \to Set) \to Seq\,A \to Set) \to
-\forall (A : Set) (Q_A : A \to Set) (a : A) \to Q_A\,x \to
+\forall (A : Set) (Q_A : A \to Set) (a : A) \to Q_A\,a \to
 P\,A\,Q_A\,(const\,a)}\] associated with the constructor
@@
 \end{array}\]
-In the first clause above, $\mathsf{a : A}$, $\mathsf{Q_A : A \to
-  Set}$, $\mathsf{liftA : Seq^{\wedge}\,A\,Q_A\,(const\,a) =
-  Q_A\,a}$. In the second, $\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C :
-  C \to Set}$, $\mathsf{e : Equal\,A\,(B \times C)}$, $\mathsf{s_B :
-  Seq\,B}$, $\mathsf{s_C : Seq\,C}$, $\mathsf{liftE :
-  Equal^{\wedge}\,A\, (B \times C)\, Q_A\, (Q_B \times Q_C) \, e}$,
-$\mathsf{liftB : Seq^{\wedge}\,B\,Q_B\,s_B}$, and $\mathsf{liftC :}$
+In the first clause, $\mathsf{a : A}$, $\mathsf{Q_A : A \to Set}$, and
+$\mathsf{liftA : Seq^{\wedge}\,A\,Q_A\,(const\,a) = Q_A\,a}$. In the
+second, also $\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C : C \to Set}$,
+$\mathsf{e : Equal\,A\,(B \times C)}$, $\mathsf{s_B : Seq\,B}$,
+$\mathsf{s_C : Seq\,C}$, $\mathsf{liftE : Equal^{\wedge}\,A\, (B
+  \times C)\, Q_A\, (Q_B \times Q_C) \, e}$, $\mathsf{liftB :
+  Seq^{\wedge}\,B\,Q_B\,s_B}$, and $\mathsf{liftC :}$
 $\mathsf{Seq^{\wedge}\,C\,Q_C\,s_C}$ --- which together ensure that
@@
 \[\begin{array}{lll}
-\mathsf{LType^{\wedge}\,A\,Q_A\,(bool\,B\,e)} & = &\mathsf{\exists
+\mathsf{LType^{\wedge}\,A\,Q_A\,(tbool\,B\,e)} & = &\mathsf{\exists
   [Q_B]\, Equal^{\wedge}\, A\, B\, Q_A\, K^{Bool}_{\top} \,e}\\
-\mathsf{LType^{\wedge}\,A\,Q_A\,(arr\, B\, C\, e\, T_B\, T_C)}
+\mathsf{LType^{\wedge}\,A\,Q_A\,(tarr\, B\, C\, e\, T_B\, T_C)}
 &=&\mathsf{\exists [Q_B] \,\exists [Q_c]\, Equal^{\wedge}\,A\,
@@
   \, LType^{\wedge}\,B\,Q_B\,T_B \times LType^{\wedge}\,C\,Q_C\,T_C}\\
-\mathsf{LType^{\wedge}\,A\,Q_A\,(list\, B\, e\, T_B)} & = &
+\mathsf{LType^{\wedge}\,A\,Q_A\,(tlist\, B\, e\, T_B)} & = &
 \mathsf{\exists [Q_B]\, Equal^{\wedge}\,A\, (List\, B)\, Q_A\,
@@
 its Henry Ford encoding from~\eqref{eq:eq_ltype}
-and~\eqref{eq:eq_lterm}. We first define predicate lifting
+and~\eqref{eq:eq_lterm}. We first define the predicate lifting
 $\mathsf{Arr^{\wedge} : \forall (A\, B : Set) \to (A \to Set) \to (B
@@
   : B \to Set}$, $\mathsf{Q_C : C \to Set}$,
-$\mathsf{K^{Bool}_{\top}}$ is the constantly {\color{red}
-  $\mathsf{\top}$}-valued predicate on $\mathsf{Bool}$, $\mathsf{T_A :
-  LType\, A}$, $\mathsf{T_B : LType \,B}$, $\mathsf{T_C : LType \,C}$,
-$\mathsf{t_B : LTerm \, B}$, $\mathsf{t_C : LTerm \, C}$, and
-$\mathsf{t_{BA} : LTerm \, (B \to A)}$.  Moreover, $\mathsf{e :
-  Equal\,A\,Bool}$ in the first clause, $\mathsf{e : Equal\, A\, (B
-  \to C)}$ in the second, $\mathsf{e : Equal\, A\, (List\, B)}$ in the
-third, $\mathsf{e : Equal \, A \, (B \to C)}$ in the fifth, and
-$\mathsf{e : Equal\, A\, (List \,B)}$, $\mathsf{ts : List\, (LTerm
-  B)}$, and $\mathsf{List^\wedge}$ is the predicate lifting for lists
-from~\eqref{eq:rose} in the seventh.
+$\mathsf{K^{Bool}_{\top}}$ is the constantly $\mathsf{\top}$-valued
+predicate on $\mathsf{Bool}$, $\mathsf{T_A : LType\, A}$, $\mathsf{T_B
+  : LType \,B}$, $\mathsf{T_C : LType \,C}$, $\mathsf{t_B : LTerm \,
+  B}$, $\mathsf{t_C : LTerm \, C}$, and $\mathsf{t_{BA} : LTerm \, (B
+  \to A)}$.  Moreover, $\mathsf{e : Equal\,A\,Bool}$ in the first
+clause, $\mathsf{e : Equal\, A\, (B \to C)}$ in the second, $\mathsf{e
+  : Equal\, A\, (List\, B)}$ in the third, $\mathsf{e : Equal \, A \,
+  (B \to C)}$ in the fifth, and $\mathsf{e : Equal\, A\, (List \,B)}$,
+$\mathsf{ts : List\, (LTerm B)}$, and $\mathsf{List^\wedge}$ is the
+predicate lifting for lists from~\eqref{eq:rose} in the seventh.
 
@@
 \end{array}\]
+
+\noindent
 The deep induction rule for $\mathsf{LTerm}$ is thus
+
+\vspace*{-0.225in}
+
 \begin{equation}\label{eq:ind-lam}
@@
 \mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clist \, A\, Q_A\,
-    (app \,B \,\,t_{BA} \, t_B)\, (Q_B , list_{t_{BA}}, list_{t_B})} &
+    (app \,B \,\,t_{BA} \, t_B)\, (Q_B , lift_{t_{BA}}, lift_{t_B})} &
 = & \mathsf{capp\,A\,B\,Q_A\, Q_B\, t_{BA}\, t_B\, p_{BA} \, p_B}\\
-  \mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clistc \, A\, Q_A\,
+  \mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, listc \, A\, Q_A\,
     (list \,B \,e \, ts) \, (Q_B , liftE', lift_{List})} & = & 
-  \mathsf{clistc \,A\,B\,Q_A\, Q_B\, e\, ts\, liftE'\, p_{List} }
+  \mathsf{listc \,A\,B\,Q_A\, Q_B\, e\, ts\, liftE'\, p_{List} }
 \end{array}\]}
@@
 
-To prove that this rule is sound we define a witness
+\vspace*{-0.125in}
+
+To prove this rule is sound we define a witness
 $\mathsf{dIndLTerm}$ inhabiting the type in~\eqref{eq:ind-lam} as in
@@
 $\mathsf{t_B : LTerm\,B}$, $\mathsf{t_C : LTerm\,C}$, $\mathsf{t_{BA}
-  : LTerm\,{\color{red}??}}$, $\mathsf{liftA : LTerm^{\wedge}\, A\,
+  : LTerm\,(B \to A)}$, $\mathsf{liftA : LTerm^{\wedge}\, A\,
   Q_A\, (var\;s\,T_A) = LType^{\wedge}\,A\,Q_A\,T_A}$, $\mathsf{liftE
@@
   Q_B \, Q_C) \, e}$, $\mathsf{lift_{T_B}: LType^{\wedge} \, B\, Q_B\,
-  T_B}$,~$\mathsf{lift_{t_C}: LTerm^{\wedge} \, C\, Q_C\,
-  T_C}$,~$\mathsf{lift_{t_{BA}}: LTerm^{\wedge} \, (B \to A)\,
+  T_B}$, $\mathsf{lift_{t_C}: LTerm^{\wedge} \, C\, Q_C\,
+  T_C}$, $\mathsf{lift_{t_{BA}}: LTerm^{\wedge} \, (B \to A)\,
   (Arr^{\wedge} \, B\, A\, Q_B \, Q_A)\,
-  t_{BA}}$,~$\mathsf{lift_{t_B}: LTerm^{\wedge} \, B\, Q_B\,
-  t_B}$,\\ $\mathsf{liftE' : Equal^{\wedge}\, A\, (List\,B)\, Q_A\,
+  t_{BA}}$, $\mathsf{lift_{t_B}: LTerm^{\wedge} \, B\, Q_B\,
+  t_B}$, $\mathsf{liftE' : Equal^{\wedge}\, A\, (List\,B)\, Q_A\,
   (List^{\wedge}\, B\, Q_B)\, e}$, and $\mathsf{lift_{List}:
@@
 
-\vspace*{-0.2in}
+\vspace*{-0.225in}
 
@@
 
-\vspace*{-0.2in}
+\vspace*{-0.1in}
 
@@
 lifted predicates, is defined by $\mathsf{liftListMap\, A\, Q\, Q'\,
-  m\, nil\, tt = tt}$ (since $\mathsf{x : List^{\wedge}\, A\, Q\, nil
-  = \top}$ must necessarily be $\mathsf{tt}$), and by
-$\mathsf{liftListMap\, A\, Q\, Q'\, m\, (cons\, a\, l')\, (y, x') =
-  (m\,a\,y, liftListMap\, A\, Q\, Q'\, m\, l'\, x')}$ (since $\mathsf{x
-  : List^{\wedge}\, A\, Q\, (cons\, a\, l') = \top}$ must be of the
-form $\mathsf{x = (y, x')}$ where $\mathsf{y : Q\,a}$ and $\mathsf{x'
-  : List^{\wedge}\, A\, Q\, l'}$). {\color{red} Double-check!!}
+  m\, nil\, tt = tt}$ (since $\mathsf{x : List^{\wedge}\, A\, Q\,
+  nil}$ must necessarily be the sole inhabitant $\mathsf{tt}$ of
+$\mathsf{\top}$), and by $\mathsf{liftListMap\, A\, Q\, Q'\, m\,
+  (cons\, a\, l')\, (y, x') = (m\,a\,y, liftListMap\, A\, Q\, Q'\, m\,
+  l'\, x')}$ (since $\mathsf{x : List^{\wedge}\, A\, Q\, (cons\, a\,
+  l')}$ must be of the form $\mathsf{x = (y, x')}$ where
+$\mathsf{y : Q\,a}$ and $\mathsf{x' : List^{\wedge}\, A\, Q\, l'}$).
+
+\vspace*{-0.05in}
 
@@
   Set^{\beta} \to Set}$ and $\mathsf{Set^{\beta} \to Set}$,
-respectively. If $\mathsf{T}$ has type signature $\mathsf{Set^{\gamma}
-  \to Set}$ then we say that $\mathsf{T}$ is a {\em type constructor
-  of arity $\mathsf{\gamma}$}.  The overline notation denotes a finite
-list whose length is exactly the arity of the type constructor being
-applied to it. The number of type constructors $\mathsf{K}$ must
-therefore be $\alpha$. The type constructor $\mathsf{F}$ must be
-constructed inductively according to the following grammar:
+respectively. If $\mathsf{T}$ is a type constructor with signature
+$\mathsf{Set^{\gamma} \to Set}$ then we say that $\mathsf{T}$ has
+arity $\mathsf{\gamma}$.  The overline notation denotes a finite list
+whose length is exactly the arity of the type constructor being
+applied to it. The number of type constructors in $\ol{\mathsf{K}}$
+must therefore be $\alpha$. In addition, the type constructor
+$\mathsf{F}$ must be constructed inductively according to the
+following grammar:
 \[\mathsf{F\,G\,\ol{B}\; :=\;
@@
 prevents nesting, which would make it impossible to give an induction
-rule; see Section~\ref{sec:GADT-nested} below. In the fifth clause,
-$\mathsf{H}$ does not contain $\mathsf{G}$. This clause therefore
-subsumes the cases in which $\mathsf{F\,G\,\ol{B}}$ is a closed type
-or one of the $\mathsf{B_i}$. In the sixth clause, $\mathsf{H :
-  Set^\gamma \to Set}$ does not contain $\mathsf{G}$ (although
-$\ol{\mathsf{F_1\,G\,\ol{B}}}$ can). Moreover, although $\mathsf{H}$
-can construct any (truly) nested type, it cannot construct a
-GADT. This ensures that $\mathsf{H}$ admits functorial
+rule for $\mathsf{G}$; see Section~\ref{sec:GADT-nested} below. In the
+fifth clause, $\mathsf{H}$ does not contain $\mathsf{G}$. This clause
+therefore subsumes the cases in which $\mathsf{F\,G\,\ol{B}}$ is a
+closed type or one of the $\mathsf{B_i}$. In the sixth clause,
+$\mathsf{H : Set^\gamma \to Set}$ does not contain $\mathsf{G}$
+(although $\ol{\mathsf{F_1\,G\,\ol{B}}}$ can). Moreover, although
+$\mathsf{H}$ can construct any (truly) nested type, it must not
+construct a GADT. This ensures that $\mathsf{H}$ admits functorial
 semantics~\cite{jp20}, and thus has an associated map function. From
-this we can also construct a map function
+the map function for $\mathsf{H}$ we can also construct a map function
 
@@
 
-\[\mathsf{H^\wedge Map : \forall (\ol{A : Set}) (\ol{Q\;Q' : A \to Set}) 
+\begin{equation}\label{eq:hliftmap}
+\mathsf{H^\wedge Map : \forall (\ol{A : Set}) (\ol{Q\;Q' : A \to Set}) 
 \to \ol{PredMap\,A\,Q\,Q'} \to
 PredMap\,(H\,\ol{A})\,(H^{\wedge}\,\ol{A}\,\ol{Q})\, 
-(H^{\wedge}\,\ol{A}\,\ol{Q'})}\]
+(H^{\wedge}\,\ol{A}\,\ol{Q'})}
+\end{equation}
 
@@
 we omit such details since they are not essential to the present
-discussion. A further requirement that applies to all of the types
-appearing above, including the $\mathsf{K}$s in~\eqref{eq:gadts}, is
-that they must all admit predicate liftings. This is not an overly
-restrictive condition, though: all types constructed from sums,
-products, arrow types and type application admit predicate liftings,
-and so do GADTs constructed from the above grammar; in fact, we have
-seen such liftings for products and type application in
-Section~\ref{sec:deep-ind-GADTs}.  A concrete way to define predicate
-liftings more generally is, again, by induction on the the structure
-of the types. We do not give a general definition of predicate
-liftings here, though, since that would require us to first design a
-full type calculus, which is beyond the scope of the present paper.
+discussion. A further requirement that applies to all of the type
+constructors appearing in the right-hand sides of~\eqref{eq:gadts},
+including those in $\ol{\mathsf{K}}$, is that they must all admit
+predicate liftings. This is not an overly restrictive condition,
+though: all types constructed from sums, products, arrow types and
+type application admit predicate liftings, and so do GADTs constructed
+from the above grammar; in fact, we have seen such liftings for
+products and type application in Section~\ref{sec:deep-ind-GADTs}. A
+concrete way to define more general predicate liftings is, again, by
+induction on the structure of the types. We do not give a general
+definition of predicate liftings here, though, since that would
+require us to first design a full type calculus, which is beyond the
+scope of the present paper. {\color{red} Awkward paragraph? Rephrase?}
 
@@
 the argument to GADTs of arbitrary arity presents no difficulty other
-than heavier notation. The type of the single data constructor
-$\mathsf{c}$ for $\mathsf{G}$ can be rewritten as $\mathsf{c : \forall
+than heavier notation. In this case the type of $\mathsf{G}$'s single
+data constructor $\mathsf{c}$ can be rewritten as $\mathsf{c : \forall
   (\ol{B : Set}) \to Equal\,A\,(K\,\ol{B}) \to F\,G\,\ol{B} \to
@@
 $\mathsf{e : Equal\,A\,(K\,\ol{B})}$, and $\mathsf{x : F\,G\,\ol{B}}$.
-If we have predicate liftings $\mathsf{F^{\wedge}} : \mathsf{\forall (G
-  : Set^{\alpha} \to Set) (\ol{B : Set}) \to (\forall (A : Set) \to (A
-  \to Set) \to G\,A \to Set) \to (\ol{B \to Set}) \to F\,G\,\ol{B} \to
-  Set}$ and $\mathsf{K^{\wedge}} : \mathsf{\forall (\ol{B : Set}) \to
-  (\ol{B \to Set}) \to K\,\ol{B} \to Set}$
+If we have predicate liftings $\mathsf{F^{\wedge}} : \mathsf{\forall
+  (G : Set^{\alpha} \to Set) (\ol{B : Set}) \to (\forall (A : Set) \to
+  (A \to Set) \to G\,A \to Set) \to (\ol{B \to Set}) \to F\,G\,\ol{B}
+  \to Set}$ for $\mathsf{F}$ and $\mathsf{K^{\wedge}} :
+\mathsf{\forall (\ol{B : Set}) \to (\ol{B \to Set}) \to K\,\ol{B} \to
+  Set}$
 %\[\begin{array}{lll}
@@
 %\end{array}\]
-for $\mathsf{F}$ and for $\mathsf{K}$, respectively, the induction
-hypothesis for $\mathsf{c}$ is
+for $\mathsf{K}$, then the induction hypothesis associated with
+$\mathsf{c}$ is
 \[\begin{array}{l}
@@
   G\,A \to Set)}\\
-\quad \mathsf{\to \forall (A : Set) (\ol{B : Set}) (Q_A : A \to Set)
-  (\ol{Q_B : B \to Set}) (e : Equal\,A\,(K\,\ol{B})) (x :
+\quad \mathsf{\to \forall (A : Set)\, (\ol{B : Set})\, (Q_A : A \to Set)\,
+  (\ol{Q_B : B \to Set})\, (e : Equal\,A\,(K\,\ol{B}))\, (x :
   F\,G\,\ol{B})} \\ 
@@
 
-To prove that this rule is sound we define the witness
+To prove that this rule is sound we define a witness
 $\mathsf{dIndG}$ inhabiting this type by
@@
 $\mathsf{cc : dIndC\,P}$, $\mathsf{e : Equal\,A\,(K\,\ol{B})}$,
-$\mathsf{x : F\,G\,\ol{B}}$, $\ol{\mathsf{Q_B : B \to Set}}$,
-$\mathsf{liftE : Equal^{\wedge}\,A\,(K\,\ol{B})\,Q_A\,
+$\mathsf{x : F\,G\,\ol{B}}$, $\mathsf{Q_A : A \to Set}$,
+$\ol{\mathsf{Q_B : B \to Set}}$, $\mathsf{liftE :
+  Equal^{\wedge}\,A\,(K\,\ol{B})\,Q_A\,
   (K^{\wedge}\,\ol{B}\,\ol{Q_B})\,e}$, and $\mathsf{liftF :
   F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,{\ol{Q_B}}\,x}$. As a result,
-$\mathsf{(\ol{Q_B}, liftE, liftF) : G^{\wedge}\,A\,Q_A
-  (c\,\ol{B}\,e\,x)}$ as expected. 
-Finally, the morphism of predicates
-$\mathsf{p :
+$\mathsf{(\ol{Q_B}, liftE, liftF) : G^{\wedge}\,A\,Q_A (c\,\ol{B}\,
+  e\,x)}$ as expected.  Finally, the morphism of predicates $\mathsf{p
+  :
   PredMap\,(F\,G\,\ol{B})\,(F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B})
-  (F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}$ 
-is defined by structural induction on $\mathsf{F}$ as follows:
+  (F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}$ is defined by structural
+induction on $\mathsf{F}$ as follows:
+
 \begin{itemize}
@@
   (F_2^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}$. The induction hypothesis
-ensures morphisms of predicates $\mathsf{p_1 :
+  ensures
+%{\color{red} liftings $\mathsf{F_1}$ and $\mathsf{F_2}$ (with
+%  types) and}
+morphisms of predicates $\mathsf{p_1 :
   PredMap\,(F_1\,G\,\ol{B})\,(F_1^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_BQ})
@@
   F_2^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}\,(x\,z)}$, where $\mathsf{x :
-  F\,G\,\ol{B}}$. The induction hypothesis ensures a morphism of
-  predicates $\mathsf{p_2 : PredMap\,(F_2\,G\,\ol{B})\,
+  F\,G\,\ol{B}}$. The induction hypothesis ensures
+%  {\color{red}
+% liftings $\mathsf{F_1}$ and $\mathsf{F_2}$ (with types) and}
+a morphism of predicates $\mathsf{p_2 : PredMap\,(F_2\,G\,\ol{B})\,
   (F_2^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B})\,
   (F_2^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}$.  We define
-  $\mathsf{p\,x\,liftF : F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}\,x}$,
-  where $\mathsf{liftF :
-    F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}\,x}$ is defined by
-  $\mathsf{p\,x\,liftF\,z\,liftF_1 = p_2 (x\,z) (liftF\,z\,liftF_1)}$
-  for $\mathsf{z : F_1\,\ol{B}}$ and $\mathsf{liftF_1 :
-    F_1^{\wedge}\,\ol{B}\,\ol{Q_B}\,z}$. Note that $\mathsf{F_1}$ not
-  containing the recursive variable is a necessary restriction, since
-  the proof relies on
+$\mathsf{p\;x\;liftF : F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}\,x}$, where
+$\mathsf{liftF : F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}\,x}$, to
+be $\mathsf{p\;x\;liftF\;z\;liftF_1 = p_2\, (x\,z)\,
+  (liftF\,z\,liftF_1)}$ for $\mathsf{z : F_1\,\ol{B}}$ and
+$\mathsf{liftF_1 : F_1^{\wedge}\,\ol{B}\,\ol{Q_B}\,z}$. Note that
+$\mathsf{F_1}$ not containing $\mathsf{G}$ is a necessary restriction,
+since the proof relies on
 $\mathsf{F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}\,x}$ and
@@
   does not contain $\mathsf{G}$, then
-  $\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B} = P (F_1\,\ol{B})
-    (F_1^{\wedge}\,\ol{B}\,\ol{Q_B})}$. We therefore define $\mathsf{p
-    = dIndG\,P\, cc\,(F_1\,\ol{B})\,(F_1^{\wedge}\,\ol{B}\,\ol{Q_B})}$.
+  $\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B} = P \,(F_1\,\ol{B})\,
+    (F_1^{\wedge}\,\ol{B}\,\ol{Q_B})}$ for all $\mathsf{P : \forall (A
+    : Set)\to(A \to Set) \to G\,A \to Set}$.  We therefore define
+  $\mathsf{p = dIndG\,P\,
+    cc\,(F_1\,\ol{B})\,(F_1^{\wedge}\,\ol{B}\,\ol{Q_B})}$.
 \item If $\mathsf{F\,G\,\ol{B} = H\,\ol{B}}$ and $\mathsf{H}$ does not
-  contain $\mathsf{G}$, then $\mathsf{p : PredMap\, (H\,\ol{B})\,
+  contain $\mathsf{G}$, then
+  $\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B} = H^\wedge\,\ol{Q_B}}$
+  for all $\mathsf{P : \forall (A : Set)\to(A \to Set) \to G\,A \to
+    Set}$.  We therefore define $\mathsf{p : PredMap\, (H\,\ol{B})\,
     (H^{\wedge}\,\ol{B}\,\ol{Q_B})\, (H^{\wedge}\,\ol{B}\,\ol{Q_B})}$
-  is defined to be the identity morphism on predicates.
-\item If $\mathsf{F\,G\,\ol{B} = H\, (\ol{F_k\,G\,\ol{B}})}$, if
-  $\mathsf{H}$ does not contain $\mathsf{G}$, if $\mathsf{H}$ has a
-  predicate lifting $\mathsf{H^{\wedge} : \forall (\ol{C : Set}) \to
-    (\ol{C \to Set})}$ $\mathsf{\to H\,\ol{C} \to Set}$, and if
-  $\mathsf{H^\wedge}$ has a map function $\mathsf{H^\wedge Map =
-    \forall (\ol{C : Set}) (\ol{Q_C\;Q_C' : C \to Set}) \to
-    PredMap\,\ol{C}\,\ol{Q_C}\,\ol{Q_C'} \to}$
-  $\mathsf{PredMap\,(H\,\ol{C})\,(H^{\wedge}\,\ol{C}\,\ol{Q_C})\,
-    (H^{\wedge}\,\ol{C}\,\ol{Q_C'})}$, then since $\mathsf{H}$ cannot
-  be a GADT, $\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B} =
-    H^{\wedge}\, (\ol{F_k\,G\,\ol{B}})\,
-    (\ol{F_k^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}})}$.  The induction
+  to be the identity morphism on predicates.
+\item If $\mathsf{F\,G\,\ol{B} = H\, (\ol{F_k\,G\,\ol{B}})}$ and
+  $\mathsf{H}$ does not contain $\mathsf{G}$, then
+  $\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B} = H^{\wedge}\,
+  (\ol{F_k\,G\,\ol{B}})\,
+  (\ol{F_k^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}})}$ for all $\mathsf{P :
+  \forall (A : Set)\to(A \to Set) \to G\,A \to Set}$.  Since
+  $\mathsf{H}$ is not a GADT, $\mathsf{H^\wedge}$ has a map function
+  $\mathsf{H^\wedge Map}$ as in~\eqref{eq:hliftmap}.  The induction
   hypothesis ensures morphisms of predicates $\ol{\mathsf{p_k :
@@
 nested. Since both nested types and GADTs without true nesting admit
-deep induction rules, we might expect that truly nested GADTs would as
-well. Surprisingly, however, they do not. That is, our results from
+deep induction rules, we might expect truly nested GADTs to admit them
+as well. Surprisingly, however, they do not. That is, our results from
 the previous section are the strongest possible. Indeed, the induction
@@
 truly nested GADT also relies on this interpretation being functorial.
-But, whereas nested types can admit a functorial parametric semantics,
-GADTs cannot admit both functorial and parametric semantics at the
-same time~\cite{jgj21}. In this section we show how deep induction for
-truly nested GADTs nesting goes wrong by analyzing the following
-concrete example of such a type.\looseness=-1
-
-\vspace*{-0.05in}
+But, whereas ADTs and nested types both admit functorial parametric
+semantics, proper GADTs cannot admit both functorial and parametric
+semantics at the same time~\cite{jgj21}. In this section we show how
+deep induction for truly nested GADTs nesting goes wrong by analyzing
+the following very simple nested proper GADT:\looseness=-1
+
+\vspace*{-0.1in}
 
@@
 \mathsf{G^{\wedge}\,A\,Q\,(c\,B\,e\,x)
-= \exists\, [Q']\,
-Equal^{\wedge}\,A\,(B \times B)\,Q\,(Pair^{\wedge}\,B\,B\,Q'\,Q')\,e
-\times G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q')\,x}
+= \exists\, [Q_A]\,
+Equal^{\wedge}\,A\,(B \times B)\,Q_A\,(Pair^{\wedge}\,B\,B\,Q_B\,Q_B)\,e
+\times G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q_B)\,x}
 \]
-where $\mathsf{Q : A \to Set}$, $\mathsf{Q' : B \to Set}$, $\mathsf{e
+where $\mathsf{Q_A : A \to Set}$, $\mathsf{Q_B : B \to Set}$, $\mathsf{e
   : Equal\,A\,(B \times B)}$, and $\mathsf{x : G\,(G\,B)}$.
@@
   \to Set)} \\ 
-\quad\mathsf{\to \forall\, (A\;B : Set)\, (Q : A \to Set)\, (Q' : B
+\quad\mathsf{\to \forall\, (A\;B : Set)\, (Q_A : A \to Set)\, (Q_B : B
   \to Set)\, (e : Equal\,A\,(B \times B))\, (x : G\,(G\,B))} \\ 
 \quad\mathsf{\to Equal^{\wedge}\,A\,(B \times
-  B)\,Q\,(Pair^{\wedge}\,B\,B\,Q'\,Q')\,e \to P\,(G\,B)\,(P\,B\,Q')\,x 
-	\to P\,A\,Q\,(c\,B\,e\,x)}
+  B)\,Q_A\,(Pair^{\wedge}\,B\,B\,Q_B\,Q_B)\,e \to P\,(G\,B)\,(P\,B\,Q_B)\,x 
+	\to P\,A\,Q_A\,(c\,B\, e\,x)} 
 \end{array}\]
@@
   cc\,A\,B\,Q\,Q'\,e\,x\,liftE\,p}$, where $\mathsf{cc : dIndC\,P}$,
-$\mathsf{Q : A \to Set}$, $\mathsf{e : Equal\,A\,(B \times B)}$,
-$\mathsf{x : G\,(G\,B)}$, $\mathsf{liftG :
-  G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q')\,x}$, $\mathsf{Q' : B \to
-  Set}$, and $\mathsf{liftE : Equal^{\wedge}\,A\,(B \times
+$\mathsf{Q : A \to Set}$, $\mathsf{Q' : B \to Set}$, $\mathsf{e :
+  Equal\,A\,(B \times B)}$, $\mathsf{x : G\,(G\,B)}$, $\mathsf{liftG :
+  G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q')\,x}$, and $\mathsf{liftE :
+  Equal^{\wedge}\,A\,(B \times
   B)\,Q\,(Pair^{\wedge}\,B\,B\,Q'\,Q')\,e}$, but we still need to
@@
 define $\mathsf{q : G^{\wedge}\,(G\,B)\,(P\,B\,Q')\,x}$.  If we had
-the map function $\mathsf{G^\wedge Map : \forall\, (A : Set)\, (Q\;Q''
-  : A \to Set) \to PredMap\,A\,Q\,Q'' \to
-  PredMap\,(G\,A)\,(G^{\wedge}\,A\,Q)\,(G^{\wedge}\,A\,Q'')}$ for
-$\mathsf{G^{\wedge}}$, then we could define $\mathsf{q =
-  GLMap\,(G\,B)\,(G^{\wedge}\,B\,Q')\,(P\,B\,Q')\,
+the map function $\mathsf{G^\wedge Map : \forall\, (A : Set)\, (Q\;Q'
+  : A \to Set) \to PredMap\,A\,Q\,Q' \to
+  PredMap\,(G\,A)\,(G^{\wedge}\,A\,Q)\,(G^{\wedge}\,A\,Q')}$ for
+$\mathsf{G^{\wedge}}$, then we could define $\mathsf{q = G^\wedge
+  Map\,(G\,B)\,(G^{\wedge}\,B\,Q')\,(P\,B\,Q')\,
   (dIndG\,P\,cc\,B\,Q')\,x\,liftG}$.  Unfortunately, however, we
 cannot define $\mathsf{G^\wedge Map}$. Indeed, its definition would
-have to be $ \mathsf{G^\wedge Map\,A\,Q\,Q_2\,m\,(c\,B\,e\,x)\,(Q_1,
-  liftE, liftG) = (Q_3, liftE', liftG')}$ for some $\mathsf{Q_3 : B
+have to be $ \mathsf{G^\wedge Map\,A\,Q\,Q'\,m\,(c\,B\,e\,x)\,(Q_B,
+  liftE, liftG) = (Q'_B, liftE', liftG')}$ for some $\mathsf{Q'_B : B
   \to Set}$, $\mathsf{liftE' : Equal^{\wedge}\,A\,(B \times
-  B)\,Q_2\,(Pair^{\wedge}\,B\,B\,Q_3\,Q_3)\,e}$, and $\mathsf{liftG' :
-  G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q_3)\,x}$, where $\mathsf{Q : A
-  \to Set}$, $\mathsf{Q_2 : A \to Set}$, $\mathsf{m :
-  PredMap\,A\,Q\,Q_2}$, $\mathsf{e : Equal\,A\,(B \times B)}$,
-$\mathsf{x : G\,(G\,B)}$, $\mathsf{Q_1 : B \to Set}$, $\mathsf{liftE :
+  B)\,Q'\,(Pair^{\wedge}\,B\,B\,Q'_B\,Q'_B)\,e}$, and $\mathsf{liftG'
+  : G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q'_B)\,x}$, where $\mathsf{Q :
+  A \to Set}$, $\mathsf{Q' : A \to Set}$, $\mathsf{Q_B : B \to Set}$,
+$\mathsf{m : PredMap\,A\,Q\,Q'}$, $\mathsf{e : Equal\,A\,(B \times
+  B)}$, $\mathsf{x : G\,(G\,B)}$, $\mathsf{liftE :
   Equal^{\wedge}\,A\,(B \times
-  B)\,Q\,(Pair^{\wedge}\,B\,B\,Q_1\,Q_1)\,e}$, and $\mathsf{liftG :
-  G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q_1)\,x}$. In other words, we
-would need to produce a proof of the (extensional) equality of the
-predicates $\mathsf{Q_2}$ and $\mathsf{Pair^{\wedge}\,B\,B\,Q_3\,Q_3}$
-from just a proof
-%$\mathsf{liftE}$
-of the (extensional) equality of the predicates $\mathsf{Q}$ and
-$\mathsf{Pair^{\wedge}\,B\,B\,Q_1\,Q_1}$ and a morphism of predicates
-$\mathsf{m}$ from $\mathsf{Q}$ to $\mathsf{Q_2}$.
+  B)\,Q\,(Pair^{\wedge}\,B\,B\,Q_B\,Q_B)\,e}$, and $\mathsf{liftG :
+  G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q_B)\,x}$. In other words, we
+would need to produce a proof $\mathsf{liftE'}$ of the (extensional)
+equality of the predicates $\mathsf{Q'}$ and
+$\mathsf{Pair^{\wedge}\,B\,B\,Q'_B\,Q'_B}$ from just a proof
+$\mathsf{liftE}$ of the (extensional) equality of the predicates
+$\mathsf{Q}$ and $\mathsf{Pair^{\wedge}\,B\,B\,Q_B\,Q_B}$ and a
+morphism of predicates $\mathsf{m}$ from $\mathsf{Q}$ to
+$\mathsf{Q_2'}$.
 %a proof $\mathsf{liftE}$ of the (extensional) equality of the
@@
 But this will not be possible in general: the facts that $\mathsf{Q}$
-is equal to $\mathsf{Pair^{\wedge}\,B\,B\,Q_1\,Q_1}$ and that there is
+is equal to $\mathsf{Pair^{\wedge}\,B\,B\,Q_B\,Q_B}$ and that there is
 a morphism of predicates $\mathsf{m}$ from $\mathsf{Q}$ to
-$\mathsf{Q_2}$ do not guarantee that there exists a predicate
-$\mathsf{Q_3}$ such that $\mathsf{Q_2}$ is equal to
-$\mathsf{Pair^{\wedge}\,B\,B\,Q_3\,Q_3}$.
+$\mathsf{Q'}$ do not guarantee that there exists a predicate
+$\mathsf{Q'_B}$ such that $\mathsf{Q'}$ is equal to
+$\mathsf{Pair^{\wedge}\,B\,B\,Q'_B\,Q'_B}$.
 
@@
 type arguments to $\mathsf{Equal^\wedge}$ is structured (i.e., not a
-variable). This is, of course, a critical component in any approach to
-deep induction for such GADTs.\looseness=-1
+variable). And this is, of course, a key component in any approach to
+deep induction for truly nested GADTs.\looseness=-1
 
@@
 In this section we use deep induction for the $\mathsf{LTerm}$ GADT
-from~\eqref{eq:eq_lterm} to extract the type from a lambda term.  For
-this we use the following predicate, which takes a lambda term and
-produces the type of its possible types:
+from~\eqref{eq:eq_lterm} to extract the type from a lambda term.  The
+following predicate takes a lambda term as input and either returns
+its type if it is well-typed or indicates that it is not:
+
+\vspace*{-0.3in}
+
 \begin{align*}
@@
 The predicate $\mathsf{GetType}$ uses the standard $\mathsf{Maybe}$
-data type to represent potential failure. It is defined by:
+data type to represent potential ill-typedness. It is defined by:
 
@@
 \mathsf{\;\;\;\;\;\;nothing :\, Maybe\,A}\\
-\mathsf{\;\;\;\;\;\;\;\;\;\;\;\;just :\, A \to Maybe\,A}
+\mathsf{\;\;\;\;\;\;just\;\;\;\;\;\;\; :\, A \to Maybe\,A}
 \end{array}
@@
 
-We want to show that $\mathsf{GetType}$ is satisfied by every element
-of $\mathsf{LTerm\,A}$, i.e., we want to prove:
+\noindent
+We want to show that $\mathsf{GetType\,A\,t}$ is satisfied by every
+element $\mathsf{t}$ in $\mathsf{LTerm\,A}$, i.e., we want to prove:
 \[ \mathsf{getTypeProof : \forall \, (A : Set)\, (t : LTerm\,A) \to
-   GetType \,A \,t}\] This cannot be achieved without deep induction,
-which is needed to apply the induction hypothesis to the individual
-terms in the list of terms that the data constructor $\mathsf{list}$
-takes as an argument. Using the deep induction rule
-$\mathsf{dIndLTerm}$ from Section~\ref{sec:ind-lam} we define
+   GetType \,A \,t}\] This property cannot be proved without deep
+induction, which is needed to apply the induction hypothesis to the
+individual terms in the list of terms that the data constructor
+$\mathsf{list}$ takes as an argument. But using the deep induction
+rule $\mathsf{dIndLTerm}$ from Section~\ref{sec:ind-lam} we can define
 $\mathsf{getTypeProof}$ by
 \[ \mathsf{getTypeProof \,A \,t = 
-    dIndLTerm\, P \, cvar\, cabs\, capp\, clistc\, A\, K_\top\, t\,
+    dIndLTerm\, P \, cvar\, cabs\, capp\, listc\, A\, K_\top\, t\,
     (LTerm^\wedge KT\, A\, t) }\] where $\mathsf{t : LTerm\,A}$,
-$\mathsf{P}$ is the polymorphic predicate $\mathsf{\lambda \, (A:
+$\mathsf{P}$ is the polymorphic predicate $\mathsf{ \lambda \, (A:
   Set)\, (Q : A \to Set)\, (t : LTerm\,A)\, \to Maybe \, (LType \,
-  A)}$, $\mathsf{K_\top}$ is the constantly {\color{red}
-  $\mathsf{\top}$}-valued predicate, and $\mathsf{LTerm^\wedge KT :
-  \forall\, (A : Set)\, (t : LTerm A) \to LTerm^{\wedge}\, A\,
-  K_\top\,t}$ is a term, to be defined below, witnessing that
-$\mathsf{K_\top}$ can be lifted to all terms. We also need the
-applications to $\mathsf{P}$ of each of the induction hypotheses from
-Section~\ref{sec:ind-lam}. These are:
+  A)}$, $\mathsf{K_\top}$ is the constantly $\mathsf{\top}$-valued
+predicate on $\mathsf{A}$, and $\mathsf{LTerm^\wedge KT : \forall\, (A
+  : Set)\, (t : LTerm A) \to LTerm^{\wedge}\, A\, K_\top\,t}$ is a
+term, to be defined below, witnessing that $\mathsf{K_\top}$ can be
+lifted to all terms. We also need the applications to $\mathsf{P}$ of
+each of the induction hypotheses from Section~\ref{sec:ind-lam}. These
+are:
 \[\begin{array}{lll}
@@
 final argument is $\mathsf{nothing}$ and $\mathsf{cabs\, A \,B \,C
-  \,Q_A\, Q_B \,Q_C\, e \,s \,T_B\, t_C\, Eq^\wedge\,
+  \,Q_A\, Q_B \,Q_C\, e \,s \,T_B\, t_C\, liftE\,
 %  LLTy^\wedge B\,
-L\, (just \,T_C) = just \, (arr\, B\, C\, e\, T_B \,T_C)}$ otherwise.
+L\, (just \,T_C) = just \, (tarr\, B\, C\, e\, T_B \,T_C)}$ otherwise.
 In the third clause, $\mathsf{capp\, A \,B \, Q_A \, Q_B \, t_{BA}\,
-  t_A \, (just \, (arr\, .B \,.A\, rfl\, T_B\, T_A))\, mb = just\,
+  t_A \, (just \, (tarr\, B \,A\, refl\, T_B\, T_A))\, mb = just\,
   T_A}$ and $\mathsf{capp}$ returns $\mathsf{nothing}$ otherwise.  In
@@
 %the type we need.  The type of $\mathsf{list_{ts}}$ is
-the type of $\mathsf{list_{ts}}$ becomes $\mathsf{List^{\wedge}\,
+the type of $\mathsf{lift_{ts}}$ becomes $\mathsf{List^{\wedge}\,
   (LTerm\, B)\, (GetType\, B)\, (cons\, t\, ts')} = \mathsf{GetType\,
@@
     EqualMap}$ in the cases when the first argument to
-  $\mathsf{PredMap}$ is constructed using $\mathsf{var}$ or
+  $\mathsf{PredMap}$ is constructed using $\mathsf{var}$ and
   $\mathsf{app}$, respectively. Taking $\mathsf{L_{K_\top} :
@@
   B\, K_\top) \, ts }$.  As in that case we assume a proof
-  $\mathsf{EqualLListKT : Equal^{\wedge} \, A\, (List\,B)\, K_\top\,
+  $\mathsf{Equal^{\wedge}ListKT : Equal^{\wedge} \, A\, (List\,B)\, K_\top\,
   (List^{\wedge} \, B\, K_\top) \, e}$ for the first component.  We
@@
     B\, K_\top) \, m_{K_\top} \, ts \, (List^\wedge KT\, (LTerm\, B)\,
-    ts) }$, we define $\mathsf{LTerm^\wedge KT\,A\, (listC\, B\, e\,
+    ts) }$, we define $\mathsf{LTerm^\wedge KT\,A\, (list\, B\, e\,
     ts) = (K_\top , \, Equal^\wedge ListKT , \, L_{ListL^\wedge
@@
 $\mathsf{G^\wedge KT : \forall\, (A : Set)\, (x : G\,A) \to
-  G^{\wedge}\, A\, K_\top\, x}$ for an arbitrary GADT $\mathsf{G}$, as
+  G^{\wedge}\, A\, K_\top\, x}$ for an arbitrary GADT $\mathsf{G}$ as
 defined in Section~\ref{sec:framework}.  To provide a proof of
@@
 $\mathsf{c : H \, (\ol{F\,G\, \ol{B}}) \to G\, (\ol{K\,\ol{B}})}$,
-then $\mathsf{H}$ cannot construct GADTs so the generalization
+then $\mathsf{H}$ cannot construct a GADT so the generalization
 $\mathsf{H^\wedge Map}$ of $\mathsf{listLiftMap}$ in the final bullet
@@
 Identity Extension Lemma, be obtained at the meta-level as a
-consequence of unary parametricity. At the object level, our Agda code
-simply postulates each such isomorphism needed since an Agda
-implementation of full parametricity for some relevant calculus is
-beyond the scope of the present paper.
+consequence of unary parametricity. At the object level, our Agda
+code~\cite{web-page} simply postulates each isomorphism needed since
+an Agda implementation of full parametricity for some relevant
+calculus is beyond the scope of the present paper.
 
@@
 
-This paper extends deep induction to a robust class of GADTs that are
-not truly nested. It also shows that truly nested GADTs do not admit
-deep induction rules. Our development is implemented in Agda, as is a
-case study showing how deep induction can prove properties of GADTs
-that are not provable by structural induction.\looseness=-1
-
-\begin{comment}
-\section{TODO}
-
-\begin{itemize}
-\item find correct entcsmacro file (current one is for 2018). Maybe ask Ana Sokolova (anas@cs.uni-salzburg.at).
-\item reference (correctly) Haskell Symposium paper
-\item reference inspiration for STLC GADT : https://www.seas.upenn.edu/~cis194/spring15/lectures/11-stlc.html
-\item Data type vs. data structure
-\item Weird to code in Agda if we're talking about induction rules for
-  Coq? 
-\item Agda style conventions
-\item spacing in data type declarations
-\item Mention Agda flags that need to be toggled to handle true nesting
-\end{itemize}
-\end{comment}
+This paper extends deep induction to deep GADTs that are not truly
+nested. It also shows that truly nested GADTs, deep or not, do not
+admit (deep) induction rules. Our development is implemented in Agda,
+as is a case study showing how deep induction can prove properties of
+GADTs that are not provable by structural induction.\looseness=-1
 
@@
 
-\bibitem{chl} Chlipala, A. {\em Library Inductive
-  Types}. $\mathtt{http://adam.chlipala.net/cpdt/html/InductiveTypes.html}$
+\bibitem{cis194}
+Zilberstein, N. CIS194 homepage.
+$\mathtt{https://www.seas.upenn.edu/\tilde~cis194/spring15/lectures/11\!\!-\!\!stlc.html}$
+
+%\bibitem{chl} Chlipala, A. {\em Library Inductive
+%  Types}. $\mathtt{http://adam.chlipala.net/cpdt/html/InductiveTypes.html}$
 
@@
 \bibitem{jgj21} Johann, P., Ghiorzi, E., and Jeffries, D. {\em
-  Parametricity in the presence of GADTs}. Submitted, 2021.
+  Functorial GADTs are not Parametric}. Submitted, 2021.
 
 \bibitem{jp19} Johann, P. and Polonsky, A. {\em Higher-kinded data
-  types: Syntax and semantics} Proceedings, Logic in Computer Science
-  2019. {\color{red} PAGES?}
+  types: Syntax and semantics} Proceedings, Logic in Computer Science,
+  pp. 1-13, 2019.
 
@@
 
-\bibitem{mac71} MacLane, S. {\em Catgories for the Working
+\bibitem{mac71} MacLane, S. {\em Categories for the Working
   Mathematician}. Springer, 1971.
@@
     GADTs}. Proceedings, International Conference on Functional
-  Programming, 2006. {\color{red} PAGES?}
+  Programming, pp. 50-61, 2006.
 
@@
   built-in type equality}. Proceedings, Workshop on Logical Frameworks
-  and Meta-languaegs, 2004. {\color{red} PAGES?}
+  and Meta-languages, pp. 106-124, 2004.
 

### 2021-04-30 induction-paper/ENTCS-GADT-induction.tex
Revert "minor revisions ; rename LType constructors with 't' prefix"
This reverts commit 1a5da70e8ed2880f3e287a38718f77abc0541b89.

@@
   Every time a new algebraic data type (ADT) is declared in
-  Coq}\footnote{{\color{red} Algebraic data types are 
-  defined using the $\mathsf{Inductive}$ keyword in Coq.}}
-  % called inductive
-  %   types in Coq.}}
-    , an induction rule is automatically generated for
+  Coq}\footnote{{\color{red} Algebraic data types are called inductive
+    types in Coq.}}, an induction rule is automatically generated for
 it.\looseness=-1
@@
 \noindent
-The induction rule Coq automatically generates for the analogous Coq definition of rose trees is
+The induction rule Coq automatically generates for rose trees is
 
@@
 below themselves. The quintessential example of such a type is that of
-bushes
-% \footnote{To define $\mathsf{Bush}$ and other truly nested types in Agda, 
-% we must use the $\mathsf{NO\_POSITIVITY\_CHECK}$
-% pragma . Similar flags are required in Coq.}
-~\cite{bm98}:
+bushes~\cite{bm98}:
 \begin{equation*}\label{eq:bush}
@@
 \noindent
-% {\color{red} Flags to overcome positivity restrictions?}
-
+{\color{red} Flags to overcome positivity restrictions?}
 Even defining a structural induction rule for bushes requires that we
@@
 \mathsf{data\ LType : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;\;tbool :\, \forall (B : Set) \to Equal\,A\,Bool
+\mathsf{\;\;\;\;\;\;\;\;bool :\, \forall (B : Set) \to Equal\,A\,Bool
   \to LType\,A}\\ 
-\mathsf{\;\;\;\;\;\;\;\;tarr\;\;\; :\, \forall (B\,C : Set) \to
+\mathsf{\;\;\;\;\;\;\;\;arr\;\;\; :\, \forall (B\,C : Set) \to
   Equal\,A\,(B \to C) \to LType\,B \to LType\,C \to LType\,A}\\ 
-  \mathsf{\;\;\;\;\;\;\;\;tlist\,\;\; :\, \forall (B : Set) \to
+  \mathsf{\;\;\;\;\;\;\;\;list\,\;\; :\, \forall (B : Set) \to
     Equal\,A\,(\List\,B) \to LType\,B \to LType\,A} 
@@
 \[\begin{array}{lll}
-\mathsf{LType^{\wedge}\,A\,Q_A\,(tbool\,B\,e)} & = &\mathsf{\exists
+\mathsf{LType^{\wedge}\,A\,Q_A\,(bool\,B\,e)} & = &\mathsf{\exists
   [Q_B]\, Equal^{\wedge}\, A\, B\, Q_A\, K^{Bool}_{\top} \,e}\\
-\mathsf{LType^{\wedge}\,A\,Q_A\,(tarr\, B\, C\, e\, T_B\, T_C)}
+\mathsf{LType^{\wedge}\,A\,Q_A\,(arr\, B\, C\, e\, T_B\, T_C)}
 &=&\mathsf{\exists [Q_B] \,\exists [Q_c]\, Equal^{\wedge}\,A\,
@@
   \, LType^{\wedge}\,B\,Q_B\,T_B \times LType^{\wedge}\,C\,Q_C\,T_C}\\
-\mathsf{LType^{\wedge}\,A\,Q_A\,(tlist\, B\, e\, T_B)} & = &
+\mathsf{LType^{\wedge}\,A\,Q_A\,(list\, B\, e\, T_B)} & = &
 \mathsf{\exists [Q_B]\, Equal^{\wedge}\,A\, (List\, B)\, Q_A\,
@@
 \mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clist \, A\, Q_A\,
-    (app \,B \,\,t_{BA} \, t_B)\, (Q_B , lift_{t_{BA}}, lift_{t_B})} &
+    (app \,B \,\,t_{BA} \, t_B)\, (Q_B , list_{t_{BA}}, list_{t_B})} &
 = & \mathsf{capp\,A\,B\,Q_A\, Q_B\, t_{BA}\, t_B\, p_{BA} \, p_B}\\
-  \mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, listc \, A\, Q_A\,
+  \mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clistc \, A\, Q_A\,
     (list \,B \,e \, ts) \, (Q_B , liftE', lift_{List})} & = & 
-  \mathsf{listc \,A\,B\,Q_A\, Q_B\, e\, ts\, liftE'\, p_{List} }
+  \mathsf{clistc \,A\,B\,Q_A\, Q_B\, e\, ts\, liftE'\, p_{List} }
 \end{array}\]}
@@
 \[ \mathsf{getTypeProof \,A \,t = 
-    dIndLTerm\, P \, cvar\, cabs\, capp\, listc\, A\, K_\top\, t\,
+    dIndLTerm\, P \, cvar\, cabs\, capp\, clistc\, A\, K_\top\, t\,
     (LTerm^\wedge KT\, A\, t) }\] where $\mathsf{t : LTerm\,A}$,
@@
 final argument is $\mathsf{nothing}$ and $\mathsf{cabs\, A \,B \,C
-  \,Q_A\, Q_B \,Q_C\, e \,s \,T_B\, t_C\, liftE\,
+  \,Q_A\, Q_B \,Q_C\, e \,s \,T_B\, t_C\, Eq^\wedge\,
 %  LLTy^\wedge B\,
-L\, (just \,T_C) = just \, (tarr\, B\, C\, e\, T_B \,T_C)}$ otherwise.
+L\, (just \,T_C) = just \, (arr\, B\, C\, e\, T_B \,T_C)}$ otherwise.
 In the third clause, $\mathsf{capp\, A \,B \, Q_A \, Q_B \, t_{BA}\,
-  t_A \, (just \, (tarr\, B \,A\, refl\, T_B\, T_A))\, mb = just\,
+  t_A \, (just \, (arr\, B \,A\, refl\, T_B\, T_A))\, mb = just\,
   T_A}$ and $\mathsf{capp}$ returns $\mathsf{nothing}$ otherwise.  In
@@
 %the type we need.  The type of $\mathsf{list_{ts}}$ is
-the type of $\mathsf{lift_{ts}}$ becomes $\mathsf{List^{\wedge}\,
+the type of $\mathsf{list_{ts}}$ becomes $\mathsf{List^{\wedge}\,
   (LTerm\, B)\, (GetType\, B)\, (cons\, t\, ts')} = \mathsf{GetType\,
@@
   B\, K_\top) \, ts }$.  As in that case we assume a proof
-  $\mathsf{Equal^{\wedge}ListKT : Equal^{\wedge} \, A\, (List\,B)\, K_\top\,
+  $\mathsf{EqualLListKT : Equal^{\wedge} \, A\, (List\,B)\, K_\top\,
   (List^{\wedge} \, B\, K_\top) \, e}$ for the first component.  We
@@
     B\, K_\top) \, m_{K_\top} \, ts \, (List^\wedge KT\, (LTerm\, B)\,
-    ts) }$, we define $\mathsf{LTerm^\wedge KT\,A\, (list\, B\, e\,
+    ts) }$, we define $\mathsf{LTerm^\wedge KT\,A\, (listC\, B\, e\,
     ts) = (K_\top , \, Equal^\wedge ListKT , \, L_{ListL^\wedge

### 2021-05-05 20-gadts/21-LSFA/GADT-story.tex
added 21-LFSA folder

--- 2021-03-18 20-gadts/GADT-story.tex
@@
-% For double-blind review submission, w/o CCS and ACM Reference (max
-% submission space)
-\documentclass[acmsmall,screen,review,anonymous]{acmart}
-\settopmatter{printfolios=true,printccs=false,printacmref=false}
-%% For double-blind review submission, w/ CCS and ACM Reference
-%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
-%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
-%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-%% For single-blind review submission, w/ CCS and ACM Reference
-%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
-%% For final camera-ready submission, w/ required CCS and ACM Reference
-%\documentclass[acmsmall]{acmart}\settopmatter{}
-
-%% Journal information
-%% Supplied to authors by publisher for camera-ready submission;
-%% use defaults for review submission.
-\acmJournal{PACMPL}
-\acmVolume{1}
-\acmNumber{Haskell Symposium} % CONF = POPL or ICFP or OOPSLA
-\acmArticle{1}
-\acmYear{2021}
-\acmMonth{8}
-\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
-
-%% Copyright information
-%% Supplied to authors (based on authors' rights management selection;
-%% see authors.acm.org) by publisher for camera-ready submission;
-%% use 'none' for review submission.
-\setcopyright{none}
-%\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
-%\copyrightyear{2018}           %% If different from \acmYear
-
-%% Bibliography style
-\bibliographystyle{ACM-Reference-Format}
-%% Citation style
-%% Note: author/year citations are required for papers published as an
-%% issue of PACMPL.
-\citestyle{acmauthoryear}   %% For author/year citations
-%\citestyle{acmnumeric}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Note: Authors migrating a paper from PACMPL format to traditional
-%% SIGPLAN proceedings format must update the '\documentclass' and
-%% topmatter commands above; see 'acmart-sigplanproc-template.tex'.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\documentclass[submission,copyright,creativecommons]{eptcs}
+\providecommand{\event}{LSFA 2021}
+%\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\usepackage{breakurl}             % Not needed if you use pdflatex only.
+\usepackage{underscore}           % Only needed if you use pdflatex.
 
@@
 
-\theoremstyle{definition}
-\newtheorem{exmpl}{Example}
+%\theoremstyle{definition}
+%\newtheorem{exmpl}{Example}
+\newtheorem{example}[thm]{Example}
+\newtheorem{defi}[thm]{Definition}
+\newtheorem{definition}[thm]{Definition}
 
@@
 \newcommand{\emptyfun}{{[]}}
-\newcommand{\cal}{\mathcal}
+%\newcommand{\cal}{\mathcal}
 \newcommand{\F}{\mathcal{F}}
-\renewcommand{\G}{\mathcal{G}}
+%\renewcommand{\G}{\mathcal{G}}
 \newcommand{\N}{\mathcal{N}}
@@
 \renewcommand{\S}{\mathcal S}
-\renewcommand{\C}{\mathcal{C}}
+%\renewcommand{\C}{\mathcal{C}}
 \newcommand{\D}{\mathcal{D}}
@@
 
-\title[Functional Pearl: Parametricity in the Presence of
-  GADTs]{Functional Pearl: Parametricity in the Presence of GADTs}
-
-\author{Patricia Johann, Enrico Ghiorzi, and Daniel Jeffries}
-\affiliation{ \institution{Appalachian State University}}
+\title{GADTs, Functoriality,  Parametricity: Pick Two}
+\author{Patricia Johann, Enrico Ghiorzi, and Daniel Jeffries
+\institute{Appalachian State University}
 \email{johannp@appstate.edu, ghiorzie@appstate.edu,
-  jeffriesd@appstate.edu}
+  jeffriesd@appstate.edu}}
+
+\def\titlerunning{GADTs, Functoriality,  Parametricity: Pick Two}
+\def\authorrunning{P. Johann, E. Ghiorzi, and D. Jeffries}
 
 \begin{document}
+\maketitle
 
 \begin{abstract}
-GADTs can be represented either by their Church encodings {\em \`a la}
-Atkey or as primitive data types {\em \`a la} Johann and Polonsky.  A
-GADT represented as a Church encoding need not have a functorial
-semantics, i.e., need not support a $\mathsf{map}$ function. By
-contrast, the primitive representation of a GADT must have a
-functorial semantics even to be well-defined. The two representations
-of a GADT thus need not be the same in general. In this paper we show
-that the choice of representation for GADTs has important consequences
-for parametricity, and thus for how we program with, and reason about,
-these advanced data types.
+GADTs can be represented either as their Church encodings {\em \`a la}
+Atkey, or as fixpoints {\em \`a la} Johann and Polonsky. While a GADT
+represented as its Church encoding need not support a $\mathsf{map}$
+function satisfying the functor laws, the fixpoint representation of a
+GADT must support such a $\mathsf{map}$ function even to be
+well-defined. The two representations of a GADT thus need not be the
+same in general. This observation forces a choice of representation of
+data types in languages supporting GADTs.  In this paper we show that
+choosing whether to represent data types as their Church encodings or
+as fixpoints determines whether or not a language supporting GADTs can
+have parametric models. This choice thus has important consequences
+for how we can program with, and reason about, these advanced data
+types.
 \end{abstract}
@@
 There are two standard ways to represent data types when studying
-modern functional languages: via their Church encodings in a (possibly
-higher-kinded) extension of System F~\cite{gir72}, the calculus at the
-core of all such languages, and by augmenting System F
-%a Hindley-Milner style calculus
-with primitives for constructing them directly as fixpoints.
-%{\color{blue} Something about when each of these representations is
-%  used. The former perhaps more closely represents how data types are
-%  represented in implementations, but the latter are how they have
-%  heretofore primarily been treated theoretically.}
-In models that validate the program optimization known as {\em short
-  cut fusion}~\cite{glp93} the two representations are, in fact,
-semantically equivalent, i.e., have the same interpretation in the
-model. Models in which the Church encodings and primitive
-representations of all algebraic data types (ADTs) are semantically
+modern functional languages: {\em i}) as their Church encodings in a
+(possibly higher-kinded) extension of System F~\cite{gir72}, the
+calculus at the core of all such languages, and {\em ii}) by
+augmenting System F with primitives for constructing them directly as
+fixpoints.  Models in which the Church encoding and the fixpoint
+representation of every algebraic data type (ADT) are semantically
 equivalent include the operationally-based model of~\cite{pit98,pit00}
-and the categorical models of~\cite{joh02,joh03}.
-%\footnote{{\color{blue} Pitts proves the isomorphism only for
-%    polynomial ADTs, whereas Johann {\em et al.}  prove it for any
-%    data type that is the fixpoint of a first-order functor.}}
-Correctness of short cut fusion, and thus semantic equivalence of
-Church encodings and primitive representations, can also be extended
-to the class of what might be called ``hereditarily polynomial''
-nested types in the categorical setting of~\cite{jgj21}. But, as argued
-in Section~\ref{sec:non-equiv} below, for some GADTs --- even polynomial
-ones --- no such isomorphism holds.
-%Correctness of short cut fusion, and thus semantic equivalence of
-%Church encodings and primitive representations, was recently extended
-%to the class of what might be called ``hereditarily polynomial''
-%nested types via the categorical model of~\cite{jgj21}. But, as shown
-%in Section~\ref{sec:par} below, for some GADTs --- even polynomial
-%ones --- no such isomorphism holds.
-
-GADTs are traditionally represented via Church encodings. In this
-view, GADTs are completely determined by their syntax. For example,
-the GADT $\mathsf{Seq}$ defined in Haskell by
+and the categorical models of~\cite{joh02,joh03,jgj21}.  In the
+categorical model of~\cite{jgj21} this semantic equivalence also holds
+for the syntactic generalization of ADTs known as nested
+types~\cite{bm98}. Generalized algebraic data types (GADTs) generalize
+nested types~\cite{pvww06} --- and thus further generalize ADTs ---
+syntactically (see Section~\ref{sec:gadts} below).  Indeed, we have:
+
+\vspace*{-0.1in}
+
+\[\begin{tikzcd}[column sep = huge]
+\text{\fbox{$\mathsf{ADTs}$}}\;
+\ar[r,hookrightarrow,"\text{syntactically}","\text{generalized by}"']
+& \;\text{\fbox{$\mathsf{nested~types}$}}\; \ar[r,hookrightarrow,"\text{syntactically}","\text{generalized by}"']
+& \;\text{\fbox{$\mathsf{GADTs}$}}
+\end{tikzcd}\]
+
+\vspace*{-0.05in}
+
+\noindent
+This sequence of syntactic generalizations suggests that a
+corresponding sequence of semantic generalizations might also hold.
+However, this is not the case. Specifically, the semantic equivalence
+between Church encodings and fixpoint representations that holds for
+ADTs and nested types need not hold for GADTs. In a language
+supporting GADTs we must therefore choose whether to represent data
+types as their Church encodings or as fixpoints.  The main result of
+this paper shows that the representation chosen determines whether or
+not a language supporting GADTs can have parametric
+models~\cite{rey83}. It thus determines whether or not GADTs can enjoy
+consequences of parametricity such as representation
+independence~\cite{adr09,dnb12}, equivalences between
+programs~\cite{hd11}, deep induction principles~\cite{jgj21s,jp20},
+and useful (``free'') theorems about programs derived from their types
+alone~\cite{wad89}.  This result is unintuitive, novel, and
+surprising.
+
+Consider the GADT $\mathsf{Seq}$ defined by\footnote{Although our
+  development applies to GADTs in any language, we will use Haskell
+  syntax for the code in this paper.}
 \begin{equation}\label{eq:seq}
@@
 \noindent
-comprises sequences of any type $\mathsf{a}$ and sequences obtained by
-pairing the data in two already existing sequences. By contrast, the
-primitive representation sees a GADT as the {\em functorial
-  completion}~\cite{jp19} of its syntax. Completing a GADT's syntax to
-a {\em functor} is necessary in order to be able to take the least
-fixed point~\cite{tfca}, so functoriality is inherent in the view of
-data types as their primitive representations. Viewed as its primitive
-representation, $\mathsf{Seq}$ contains not just those data elements
-representable in syntax, but also all data elements of the form
+This GADT comprises sequences of any type $\mathsf{a}$ and sequences
+obtained by pairing the data in two already existing sequences. When
+represented as its Church encoding, this GADT contains no data
+elements other than these. More generally, a GADT's Church encoding
+contains exactly those data elements that are representable by its
+syntax. By contrast, the fixpoint representation of a GADT sees that
+data type as the {\em functorial completion}~\cite{jp19} of its
+syntax.  Completing a GADT's syntax from a funct{\em ion} on types to
+a funct{\em or} on types is necessary if it is to be represented as a
+fixpoint~\cite{tfca}, so functoriality is inherent in fixpoint
+representations of data types. Since being a functor entails
+supporting a map function satisfying the functor laws, the fixpoint
+representation of a data type must include the entire ``map closure''
+of its syntax.  For example, the fixpoint representation of
+$\mathsf{Seq}$ contains not just those data elements representable by
+its syntax, but also all data elements of the form
 $\mathsf{map\,f\,s}$ for all types $\mathsf{t_1}$ and $\mathsf{t_2}$,
@@
 $\mathsf{v_1}$ and $\mathsf{v_2}$. Importantly, functorial completion
-adds no new data elements to the syntax of a GADT that is actually a
-(polynomial) ADT or a (hereditarily polynomial) nested type, so its
-Church encoding and primitive representation coincide. However, the
-two representations are rarely the same for more general GADTs.
-
-But does it really matter which way we view GADTs?
+adds no new data elements to the syntax of a GADT that is an ADT or
+other nested type. This perfectly explains why Church encodings and
+fixpoint representations coincide for ADTs and other nested
+types. However, the two representations are not the same for GADTs
+that are not nested types.\looseness=-1
+
+Even so, does it really matter how we represent GADTs?
 
@@
 shape of a particular element of a GADT is actually {\em determined}
-by the data it contains. As a result, it is not expected to support a
-$\mathsf{map}$ function --- equivalently, to be an instance of
-Haskell's $\mathsf{Functor}$ class --- like other data types. Indeed,
-the definition for $\mathsf{Seq}$ above specifies that an element of
-the form $\mathsf{Pair\,u_1\,u_2}$ must have the shape of a sequence
-of data of pair type rather than a sequence of data of arbitrary type
-$\mathsf{e}$.  The clause of $\mathsf{map}$ for the $\mathsf{Pair}$
-constructor should therefore feed $\mathsf{map}$ a function $\mathsf{f
-  :: (a \times b) \to e}$ and a term of the form $\mathsf{Pair
-  \,u_1\,u_2}$ for $\mathsf{u_1 :: Seq\,a}$ and $\mathsf{u_2 ::
-  Seq\,b}$, and produce a term $\mathsf{Pair\,v_1\,v_2}$ for some
-appropriately typed terms $\mathsf{v_1}$ and $\mathsf{v_2}$. However,
-it is not clear how to achieve this since $\mathsf{e}$ need not
-necessarily be a product type. And even if $\mathsf{e}$ {\em were}
-known to be of the form $\mathsf{w \times z}$, then we still wouldn't
-necessarily have a way to produce data of type $\mathsf{w \times z}$
-from only $\mathsf{f :: a \times b \to w \times z}$ and $\mathsf{u_1}$
-and $\mathsf{u_2}$ unless we knew, e.g., that $\mathsf{f}$ was a pair
-of functions $\mathsf{(f_1 :: a \to w, f_2 :: b \to z)}$. When it is
-intended to capture this kind of non-uniform behavior, a GADT's
-behavior does not generalize that of ADTs as advertised, and the GADT
-doesn't even really behave like a traditional data type. In this
-situation, the GADT cannot be seen as a functor, and must instead be
-represented via its Church encoding.
-
-A semanticist, on the other hand, is likely to follow~\cite{bfss90}
-and expect GADTs to behave as their names suggests --- i.e., as
-generalizations of ADTs. The semanticist is thus likely to expect a
-GADT to be able to be used as an ADT is used --- i.e., as a shape
-filled with data that can be changed without also changing the
-shape. To be used in this way, a GADT must be viewed in terms of its
-primitive representation. In fact, the functorial completion of a GADT
-inherent in its primitive representation is exactly what is needed to
-ensure that the GADT supports a $\mathsf{map}$ function, i.e., is
-interpreted as a true categorical functor, and thus can reasonably be
-made an instance of Haskell's $\mathsf{Functor}$ class. It is worth
-noting, however, that since the functorial completion of a GADT
-includes those data elements that are derivable via repeated
-applications of $\mathsf{map}$ starting from those specified by its
-syntax, primitive GADTs need not, in general, be nested types. For
-example, let $\mathsf{1}$ be the unit type, whose single element is
-also denoted by $\mathsf{1}$, and let $\mathsf{G}$ be the GADT defined
-by
+by the data it contains. As a result, it cannot be expected to support
+a functorial $\mathsf{map}$ function like ADTs and nested
+types. Indeed, the definition for $\mathsf{Seq}$ above specifies that
+an element of the form $\mathsf{Pair\,u_1\,u_2}$ must have the shape
+of a sequence of data of pair type rather than a sequence of data of
+arbitrary type $\mathsf{e}$.  The clause of $\mathsf{map}$ for the
+$\mathsf{Pair}$ constructor should therefore feed $\mathsf{map}$ a
+function $\mathsf{f :: (a \times b) \to e}$ and a term of the form
+$\mathsf{Pair \,u_1\,u_2}$ for $\mathsf{u_1 :: Seq\,a}$ and
+$\mathsf{u_2 :: Seq\,b}$, and produce a term $\mathsf{Pair\,v_1\,v_2}$
+for some appropriately typed terms $\mathsf{v_1}$ and
+$\mathsf{v_2}$. However, it is not clear how to achieve this since
+$\mathsf{e}$ need not necessarily be a product type. And even if
+$\mathsf{e}$ {\em were} known to be of the form $\mathsf{w \times z}$,
+we still wouldn't necessarily have a way to produce data of type
+$\mathsf{w \times z}$ from only $\mathsf{f :: a \times b \to w \times
+  z}$ and $\mathsf{u_1}$ and $\mathsf{u_2}$ unless we knew, e.g., that
+$\mathsf{f}$ was a pair of functions $\mathsf{(f_1 :: a \to w, f_2 ::
+  b \to z)}$. When it is intended to capture this kind of
+non-uniformity, a GADT cannot be regarded as a data-independent
+container that can be filled with data of any type in the same way
+that ADTs and nested types can. In this situation, a GADT cannot be
+seen as a functor, and so must necessarily be represented as its
+Church encoding.
+
+A semanticist, on the other hand, is likely to expect GADTs to
+generalize ADTs semantically --- i.e., to have the same kind of
+functorial semantics that ADTs and nested types
+have~\cite{bm98,jg07,jp19} --- and thus to be useable as a container
+filled with data that can be changed without changing the shape of the
+container itself.  For a GADT to be used in this way, its syntax must
+reflect functoriality. Since the Church encoding of a GADT that is not
+a nested type is not a functor, such a GADT must be viewed as its
+fixpoint representation. The functorial completion inherent in the
+fixpoint representation of a GADT adds to its syntax those, and only
+those, data elements needed to support a (functorial) $\mathsf{map}$
+function.  For example, let $\mathsf{1}$ be the unit type whose single
+element is also denoted by $\mathsf{1}$, and let $\mathsf{G}$ be the
+GADT defined by\looseness=-1
 \begin{equation}\label{eq:G}
@@
 \mathsf{data\,G\,a\,where}\\
-\mathsf{\;\;\;\;\;\;\;\;c :: G\,1}
+\mathsf{\;\;\;\;\;\;\;\;C :: G\,1}
 \end{array}
@@
 any instance $\mathsf{G\,a}$ for any type $\mathsf{a}$ that is not the
-empty type $\mathsf{0}$ (there is always a function from $\mathsf{1}$
-to such an $\mathsf{a}$), but does not include any terms at instance
-$\mathsf{G\,0}$. Indeed, $\mathsf{G\,0}$ is not inhabited via the
-syntactic specification of $\mathsf{G}$, and is not inhabited via
+empty type $\mathsf{0}$ (since there is always a function from
+$\mathsf{1}$ to such an $\mathsf{a}$), but includes no elements at
+instance $\mathsf{G\,0}$. Indeed, $\mathsf{G\,0}$ is not inhabited via
+the syntactic specification of $\mathsf{G}$, and is not inhabited via
 functorial completion because it is not possible to define a function
-from $\mathsf{1}$ into $\mathsf{0}$.
+from $\mathsf{1}$ into $\mathsf{0}$. Note, however, that since not all
+instances of the fixpoint representation of a GADT need be inhabited,
+fixpoint representations of GADTs are not, in general, nested types.
 
@@
 GADT programmer and the semanticist are both valid, the two are
-irreconcilable. Importantly, which point of view is adopted in any
-particular setting has significant consequences for the ways GADTs can
-be used and reasoned about there. In particular, the choice of GADT
-representation has deep implications for parametric reasoning about
-structures of those types. Indeed, a GADT programmer who views GADTs
-as their Church encodings cannot safely use program transformations or
-reasoning principles that involve $\mathsf{map}$ functions for them,
-although they may be able to program with and reason about GADTs using
-other consequences of parametricity, such as type inhabitation
-results. On the other hand, a semanticist who views GADTs as primitive
-data types will have all naturality-based program transformations and
-reasoning principles for GADTs at their disposal, but, since no
-parametric model can be constructed for primitive GADTs,
-non-naturality consequences of parametric reasoning may not be
-available to them.
-
-In this paper we show how these observations can be made precise,
-thereby answering the question we posed above in the affirmative: Yes,
-it really does matter which way we view GADTs.
+irreconcilable. Importantly, which representation of GADTs is adopted
+in any particular setting has significant consequences for the ways
+they can be used and reasoned about there. In particular, the way that
+GADTs are represented has deep implications for parametric reasoning
+about them. Specifically, a programmer who views GADTs as their Church
+encodings cannot safely use program transformations or reasoning
+principles that involve $\mathsf{map}$ functions for them, although
+they may be able to program with and reason about GADTs using other
+consequences of parametricity, such as type inhabitation results. On
+the other hand, a semanticist who views GADTs as fixpoints will have
+all naturality-based program transformations and reasoning principles
+for GADTs at their disposal since these all derive from
+functoriality. But since, as we show in Section~\ref{sec:par}, no
+parametric model can be constructed for fixpoint representations of
+GADTs, non-naturality consequences of parametricity will not
+necessarily hold for them.  Overall, we show that, as with software
+engineering's iron triangle~\cite{iron-triangle}, we can have any two
+of {\em GADTs}, {\em functoriality}, and {\em parametricity} we like,
+but we cannot have all three.
+
+The goal of this paper is to show how the above observations can be
+made precise. We thereby answer the question we posed above in the
+affirmative: Yes, it really does matter how we represent GADTs.
 
@@
 A (polynomial) {\em algebraic data type} (ADT) has the form
-\[\mathsf{T\,a} = \mathsf{c_1 t_{11}}...\mathsf{t_{1k_1}}\, |\, ...\, |\,
-\mathsf{c_n t_{n1}}...\mathsf{t_{nk_n}}\] where each $\mathsf{t_{ij}}$
+\[\mathsf{T\,a} = \mathsf{C_1 t_{11}}...\mathsf{t_{1k_1}}\, |\, ...\, |\,
+\mathsf{C_n t_{n1}}...\mathsf{t_{nk_n}}\] where each $\mathsf{t_{ij}}$
 is a type depending only on $\mathsf{a}$. Such a data type can be
@@
 {\em shape}, which is determined by the types of its {\em
-  constructors} $\mathsf{c_1},...,\mathsf{c_n}$. An ADT's constructors
+  constructors} $\mathsf{C_1},...,\mathsf{C_n}$. An ADT's constructors
 are used to build the data values of the data type, as well as to
@@
 List types are the quintessential examples of ADTs. The shape of the
-container underlying the list type
+container underlying the
+%list
+type
 \[\mathsf{List\,a} = \mathsf{Nil} \,|\, \mathsf{Cons\,a\,(List\,a)}\]
@@
 $\mathsf{n-1}$. Since the type argument to every occurrence of the
-type constructor $\mathsf{List}$ in the right-hand side of above
-definition is the same the type instance being defined on its
+type constructor $\mathsf{List}$ in the right-hand side of the above
+definition is the same as the type instance being defined on its
 left-hand side, the type $\mathsf{List\,a}$ enforces the invariant
@@
   emphasizes that $\mathsf{fmap}$ functions in Haskell are intended to
-  satisfy syntactic reflections of the functoriality properties ---
+  satisfy syntactic reflections of the functor laws ---
+%  iality properties ---
   i.e., preservation of identity functions and composition of
@@
 
-One way to represent ADTs is via their Church encodings. A Church
+One way to represent ADTs is as their Church encodings. A Church
 encoding is a representation of a data type as a function in a pure
@@
 been popularized as various {\em visitor patterns} in object-oriented
-programming~\cite{owg08,gon21}.
+programming~\cite{gon21,owg08}.
 
@@
 associated type constructor to be made an instance of Haskell's
-$\mathsf{Functor}$ class. For example, Haskell's built-in
-$\mathsf{map}$ function makes the type constructor $\mathsf{List}$ an
+$\mathsf{Functor}$ class. For example, the
+$\mathsf{map}$ function
+from Haskell's standard library
+makes the type constructor $\mathsf{List}$ an
 instance of the $\mathsf{Functor}$ class, and for $\mathsf{Tree}$ we
@@
 \mathsf{map_{Tree}\, f\, (Node\,t_1\,x\,t_2)} & \mathsf{=} &
-\mathsf{Node\,(map_{Tree}\, f \,t_1)\,x\,(map_{Tree}\,f\,t_2)}\\
+\mathsf{Node\,(map_{Tree}\, f \,t_1)\,(f\,x)\,(map_{Tree}\,f\,t_2)}\\
 \end{array}\]
@@
 structure-preserving, data-changing $\mathsf{map}$ function can be
-defined (although one always can). It therefore need not satisfy any
-rearrange-transform property either (although it always does if such a
-$\mathsf{map}$ function is defined for it).
-
-\subsection{ADTs as Primitive Data Types}
+defined or that, if one can, it will satisfy a rearrange-transform
+property. But such a $\mathsf{map}$ function can always be defined for
+any ADT precisely because its Church encoding and its fixpoint
+representation necessarily coincide, and this $\mathsf{map}$ function
+will, indeed, satisfy a rearrange-transform property.
+
+%(although one always can). It therefore need not satisfy any
+%rearrange-transform property either (although it always does if such a
+%$\mathsf{map}$ function is defined for it).
+
+\subsection{ADTs as Fixpoints}
 
 By contrast, the ability to define such a $\mathsf{map}$ function is
-inherent in the view of ADTs as primitive data types. Such a view is
-possible in any language that supports primitives for sum types,
-product types, and recursion. In such a language, the primitive
-representations of the ADTs $\mathsf{List\,a}$ and $\mathsf{Tree\,a}$
-are
+inherent in the view of ADTs as fixpoints. Such a view is possible in
+any language that supports primitives for sum types, product types,
+and recursion. In such a language, the fixpoint representations of
+  the ADTs $\mathsf{List\,a}$ and $\mathsf{Tree\,a}$ are
 \begin{equation}\label{eq:list}
@@
 
-Primitive representations capture in syntax the fact that ADTs can be
+Fixpoint representations capture in syntax the fact that ADTs can be
 considered as fixpoints of functors. For example $\mathsf{List\,a}$ is
@@
 F_{\mathit{List}\,a}$, where the functor $F_{\mathit{List}\,a}$ models
-the type constructor $\mathsf{F_{List\,a}}$. Similarly, the ADT
+the type constructor $\mathsf{F_{List\,a}}$.\footnote{Throughout, we
+  use $\mathtt{sans\; serif}$ font for program text and $\mathit{math
+    \; italic}$ font for semantic objects.}  Similarly, the ADT
 $\mathsf{Tree\,a}$ can be seen to be a fixpoint of
 $\mathsf{F_{Tree\,a}}$, where $\mathsf{F_{Tree\,a}\,X} = \mathsf{a + X
-  \times a \times X}$ as determined by~\eqref{eq:tree}, so that
-$\mathsf{Tree\, a} = \mathsf{ \mu F_{Tree\,a}}$ is modeled by $\mu
-F_{\mathit{Tree}\,a}$, where $F_{\mathit{Tree}\,a} $ models
-$\mathsf{F_{Tree\,a}}$.
+  \times a \times X}$ by~\eqref{eq:tree}, so that $\mathsf{Tree\, a} =
+\mathsf{ \mu F_{Tree\,a}}$ is modeled by $\mu F_{\mathit{Tree}\,a}$,
+where $F_{\mathit{Tree}\,a} $ models $\mathsf{F_{Tree\,a}}$.
 
@@
 of types. But to ensure that the syntactic fixpoint representing an
-ADT actually denotes a semantic object computed as a real-live
+ADT actually denotes a semantic object computed as a semantic
 fixpoint, the semantic fixpoint calculation must converge.  If, as is
-typical, we interpret our types as sets, then this means that the
-fixpoint being taken must be of a {\em functor} on the category
-$\mathit{Set}$ of sets and functions between them, rather than of a
-mere function between sets~\cite{tfca}.  That is, the function $F$
-interpreting the type constructor $\mathsf{F}$ constructing the body
-of a syntactic fixpoint must not only have an action on sets, but must
-also have a {\em functorial action} on functions between
+typical, we interpret our types as sets, then the fixpoint being taken
+must be of a funct{\em or} on the category $\mathit{Set}$ of sets and
+functions between them, rather than of a mere funct{\em ion} between
+sets~\cite{tfca}.  That is, the function $F$ interpreting the type
+constructor $\mathsf{F}$ constructing the body of a syntactic fixpoint
+must be a functor, i.e., must not only have an action on sets, but
+must also have a {\em functorial action} on functions between
 sets. Reflecting this requirement back into syntax gives that
-$\mathsf{F}$ must support a suitable
+$\mathsf{F}$ must support a
 %\footnote{The $\mathsf{map}$
@@
 %  even though there is no mechanism in Haskell for enforcing this.}
-$\mathsf{map}$ function, i.e., must be an instance of Haskell's
-$\mathsf{Functor}$ class.
+$\mathsf{map}$ function satisfying the functor laws. That is,
+$\mathsf{F}$ must be an instance of Haskell's $\mathsf{Functor}$ class
+(with the aforementioned caveat about the functor laws).
+%  \footnote{\color{red} As noted above, the $\mathsf{fmap}$
+%    functions in Haskell are {\em intended} to satisfy syntactic
+%    reflections of the functoriality properties, even though there is
+%    no mechanism in Haskell for enforcing this.}
 
@@
 $\mu F$ is itself a functor, so that the type constructor $\mathsf{T}$
-associated with the ADT $\mathsf{T\,a}$ also supports a $\mathsf{map}$
+associated with $\mathsf{T\,a}$ also supports its own $\mathsf{map}$
 function, we can require that $F$ also be the fixpoint $\mu H$ of a
@@
 the ``type constructor constructor'' $\mathsf{H}$ that supports
-suitable\footnote{The $\mathsf{map}$ function for $\mathsf{H}$ should
-  satisfy syntactic reflections of the functoriality properties in
+suitable\footnote{The $\mathsf{map}$ function for $\mathsf{H}$ is
+  intended to satisfy syntactic reflections of the functor laws in
   $\mathit{Set}^\mathit{Set}$ --- i.e., preservation of identity
   natural transformations and composition of natural transformations
-  --- and the $\mathsf{map}$ function for $\mathsf{H\,F}$ should
-  satisfy syntactic reflections of the functoriality properties in
-  $\mathit{Set}$ as in the previous footnote. Again, there is no
-  mechanism in Haskell for enforcing this.} $\mathsf{map}$ functions.
+  --- and the $\mathsf{map}$ function for $\mathsf{H\,F}$ is intended
+  to satisfy syntactic reflections of the functor laws in
+  $\mathit{Set}$ even though there is no mechanism in Haskell for
+  enforcing this.} $\mathsf{map}$ functions.
 
@@
 
-In models that validate short cut fusion for ADTs and nested types,
-the Church encoding and the primitive representation of an ADT or
-nested type are isomorphic as sets. But whereas it is impossible even
-to {\em state} the rearrange-transform property for Church encodings
-of such data types (unless $\mathsf{map}$ functions have for some
-reason been defined for them), such a property is simply the
+In any parametric model, the interpretations of the Church encoding
+and the fixpoint representation of an ADT or a nested type will
+necessarily be isomorphic. But whereas it is impossible even to {\em
+  state} the rearrange-transform property for their Church encodings
+(unless $\mathsf{map}$ functions have somehow been defined for them),
+for their fixpoint representations such a property is simply the
 reflection back into syntax of the instance of naturality for the
@@
 GADTs relax the restriction on the type instances appearing in a data
-type definition. The special form of GADTs known as {\em nested types}
-allow the data constructors of a GADT to take as arguments data whose
-types involve type instances of the GADT other than the one being
-defined. For example, the definition
+type definition. The special form of GADTs known as {\em nested
+  types}~\cite{bm98} allow the data constructors of a GADT to take as
+arguments data whose types involve type instances of the GADT other
+than the one being defined. However, the return type of each
+constructor of a nested type must still be precisely the one being
+defined. This is illlustrated by the definition
 \[\mathsf{PTree \, a = PLeaf \,a \,|\, PNode \,(PTree\,(a \times
-  a))}\] of the nested type $\mathsf{PTree\,a}$ of perfect trees
-introduces the data constructors $\mathsf{PLeaf :: a \to PTree\,a}$
-and $\mathsf{PNode :: PTree \,(a\times a) \to PTree\,a}$. It enforces
-not only the invariant that all of the data in a structure of type
-$\mathsf{PTree\,a}$ is of the same type $\mathsf{a}$, but also the
-invariant that all perfect trees have lengths that are powers of
-2. More general GADTs allow their constructors both to take as
-arguments {\em and return as results} data whose types involve type
-instances of the GADT other than the one being defined. An example is
-the GADT $\mathsf{Seq}$ given in~\eqref{eq:seq}. Note that since the
-return type of the data constructor $\mathsf{Pair}$ is not of the form
-$\mathsf{Seq\,a}$ for any variable $\mathsf{a}$, $\mathsf{Seq}$ is a
-GADT that is not a nested type.
+  a))}\] of the nested type $\mathsf{PTree\,a}$ of perfect trees,
+which introduces the data constructors $\mathsf{PLeaf :: a \to
+  PTree\,a}$ and $\mathsf{PNode :: PTree \,(a\times a) \to
+  PTree\,a}$. It enforces not only the invariant that all of the data
+in a structure of type $\mathsf{PTree\,a}$ is of the same type
+$\mathsf{a}$, but also the invariant that all perfect trees have
+lengths that are powers of 2. GADTs that are not nested types allow
+their constructors both to take as arguments {\em and return as
+  results} data whose types involve type instances of the GADT other
+than the one being defined. An example is the GADT $\mathsf{Seq}$
+given in~\eqref{eq:seq}. Since the return type of the data constructor
+$\mathsf{Pair}$ is not of the form $\mathsf{Seq\,a}$ for any variable
+$\mathsf{a}$, $\mathsf{Seq}$ is a GADT that is not a nested type.
 
@@
 $\mathsf{Seq\,a}$ only if $\mathsf{a}$ is a pair type. As a result,
-modifying the data in a GADT's container may actually change the shape
-of --- or may even destroy! --- that container.
+modifying the data in a GADT's container may change the shape of that
+container, or even produce an ill-typed result.
 
 To determine the possible shapes of a GADT's container we must
-pattern-match on the type of the data to be contained. For this, it is
+pattern match on the type of the data to be contained. For this, it is
 essential that a GADT calculus support an equality type
@@
 %semantics?}
-But this should be impossible. It is immediate that a GADT that is not
-actually a nested type cannot support a $\mathsf{map}$ function (or a
-rearrange-transform property) either. In the case of $\mathsf{Seq}$,
-this is already intuitively evident from its Haskell definition as
-discussed in Section~\ref{sec:intro}.
-
-\subsection{GADTs as Primitive Data Types}\label{sec:prim-gadts}
+But this is not possible.
+
+Since the Church encoding of a GADT that is not a nested type involves
+the equality type, its $\mathsf{map}$ function must necessarily
+involve the equality type's $\mathsf{map}$ function. But since the
+equality type does not support a $\mathsf{map}$ function,
+%(and thus does not support a rearrange-transform property),
+it is immediate that
+a Church encoding of a GADT that is not a nested type cannot
+support a $\mathsf{map}$ function either.
+%(or a rearrange-transform property) either.
+The underlying problem is illustrated for the GADT $\mathsf{Seq}$ in
+%In the case of $\mathsf{Seq}$, this is already intuitively
+%evident from its Haskell definition as discussed in
+Section~\ref{sec:intro}.
+
+\subsection{GADTs as Fixpoints}\label{sec:prim-gadts}
 
@@
 just those elements specified by the GADT's syntax. By contrast, the
-primitive representation of a GADT corresponds to the data type
+fixpoint representation of a GADT corresponds to the data type
 comprising all data elements in the functorial completion of the
-GADT's syntax. In this alternate reading, GADTs can, like ADTs, be
+GADT's syntax. In this latter reading, GADTs can, like ADTs, be
 modeled as fixpoints of higher-order functors. Syntactically, the use
@@
 {\cal D}$ along $K : {\cal C} \to {\cal E}$ is the best functorial
-approximation to $F$ that factors through $K$. Here, ``best functorial
+approximation to $F$ that factors through $K$. Intuitively, ``best functorial
 approximation'' means that $Lan_K \,F$ is the smallest functor that
@@
 morphism of functors (i.e., a natural transformation) from $Lan_K \,F$
-to $H$. This is captured formally by the following
+to $H$. Formally, this is captured by the following
 definition~\cite{mac71}:
@@
 
-To represent GADTs as primitive data types in a setting in which types
-are interpreted as sets, a calculus must support a primitive construct
+To represent GADTs as fixpoints in a setting in which types are
+interpreted as sets, a calculus must support a primitive construct
 $\mathsf{Lan}$ such that the type constructor $\mathsf{Lan_k\,f}$ is
@@
 form $\mathit{Set}^n$ for some $n$. Using $\mathsf{Lan}$ we can then
-rewrite the type of a constructor $\mathsf{c :: f a \to g(k\,a)}$ as
-$\mathsf{c :: (Lan_k\,f)\,a \to g\,a}$ since morphisms (i.e., natural
+rewrite the type of a constructor $\mathsf{C :: f a \to g(k\,a)}$ as
+$\mathsf{C :: (Lan_k\,f)\,a \to g\,a}$ since morphisms (i.e., natural
 transformations) from $F$ to $G \circ K$ are in one-to-one
@@
 $\mathit{Set}^\mathit{Set}$. Using $\mu$ and $\Lan$ we can then
-represent a GADT as a higher-order higher-order fixpoint. For example,
+represent a GADT as a higher-order fixpoint. For example,
 we can represent the GADT $\mathsf{Seq}$ as
@@
 generated by the corresponding GADT data constructor's syntax and also
-supports a $\mathsf{map}$ function. The primitive representation of
-any GADT thus comprises the smallest data type that both includes the
-data specified by that GADT's syntax and also supports a
-$\mathsf{map}$ function (and thus a rearrange-transform property). In
-the primitive view, then, GADTs are generally underspecified by their
-syntax.
+supports a $\mathsf{map}$ function. The fixpoint representation of any
+GADT thus comprises the smallest data type that both includes the data
+specified by that GADT's syntax and also supports a $\mathsf{map}$
+function.
+%(and thus a rearrange-transform property).
+When viewed as fixpoints, then,
+GADTs are generally underspecified by
+their syntax.
 
@@
 In this section we have seen that GADTs can be represented either as
-Church encodings or as primitive data types. This exactly mirrors the
-situation for ADTs described in Section~\ref{sec:adts}. But whereas
-the two representations are always isomorphic for ADTs, they are not,
-in general, the same for GADTs. This will be proved in the next
-section.
-
-\section{Non-Equivalence of Church Encodings and Primitive
-  Representations of GADTs}\label{sec:non-equiv}
-
-To see that the Church encoding of a GADT and its primitive
+Church encodings or as fixpoints. This exactly mirrors the situation
+for ADTs and nested types described in Sections~\ref{sec:adts}
+and~\ref{sec:gadts}. But whereas the two representations are always
+isomorphic for ADTs and nested types, they are not, in general, the
+same for GADTs. This will be proved in the next section.
+
+\section{Non-Equivalence of Church Encodings and 
+    Fixpoint Representations of GADTs}\label{sec:non-equiv}
+
+To see that the Church encoding of a GADT and its fixpoint
 representation need not be the same, consider again the GADT
@@
 GADT serves as an informative case study highlighting the differences
-between a GADT's Church encoding and its primitive representation ---
+between a GADT's Church encoding and its fixpoint representation ---
 even if we consider only the data elements it contains, and ignore
@@
 Syntactically, the GADT $\mathsf{G}$ defined in~\eqref{eq:G} comprises
-a single data element, namely $\mathsf{c :: G \, 1}$. This is captured
+a single data element, namely $\mathsf{C :: G \, 1}$. This is captured
 by $\mathsf{G}$'s Church encoding $\mathsf{G\,a} = \mathsf{\forall\,f.
@@
 
-To compute the interpretation of $\mathsf{G}$'s primitive
+To compute the interpretation of $\mathsf{G}$'s fixpoint
 representation we first note that the type of $\mathsf{G}$'s solitary
-constructor $\mathsf{c :: G\,1}$ can be equivalently expressed as
-$\mathsf{c :: 1 \to G\,1}$ or, using~\eqref{eq:nat-transfs}, as
-$\mathsf{c :: (Lan_{\lambda *. 1}\,\lambda *. 1)\,a \to G\, a}$, where
+constructor $\mathsf{C :: G\,1}$ is equivalently expressed as
+$\mathsf{C :: 1 \to G\,1}$ or, using~\eqref{eq:nat-transfs}, as
+$\mathsf{C :: (Lan_{\lambda *. 1}\,\lambda *. 1)\,a \to G\, a}$, where
 $\mathsf{\lambda *. 1}$ is the syntactic reflection of the constantly
@@
  (Lan_{\lambda *. 1} \,\lambda *.  1)\,b)\,a}\] The interpretation of
-$\mathsf{G}$ is therefore the application to $\mathsf{a}$ of the
-fixpoint of the interpretation of the body $\mathsf{\lambda
-  b. (Lan_{\lambda *. 1} \,\lambda *. 1)\,b}$ of the syntactic
-fixpoint $\mathsf{\mu \phi. \lambda b.  (Lan_{\lambda *. 1} \,\lambda
-  *.  1)\,b}$. But since the recursion variable $\mathsf{\phi}$ does
-not appear in this body, the interpretation of the fixpoint is just
-the interpretation of the body itself.  The interpretation of
-$\mathsf{G\,a}$ is therefore $(\mathit{Lan}_{\lambda *. 1} \lambda
-*. 1)\, A$, where $A$ interprets $\mathsf{a}$.
+$\mathsf{G}$ is therefore obtained by computing the fixpoint of the
+interpretation of the body $\mathsf{\lambda b. (Lan_{\lambda *. 1}
+  \,\lambda *. 1)\,b}$ of the syntactic fixpoint $\mathsf{\mu
+  \phi. \lambda b.  (Lan_{\lambda *. 1} \,\lambda *.  1)\,b}$ and
+applying the result to $\mathsf{a}$. But since the recursion variable
+$\mathsf{\phi}$ does not appear in this body, the interpretation of
+the fixpoint is just the interpretation of the body itself.  The
+interpretation of $\mathsf{G\,a}$ is therefore $(\mathit{Lan}_{\lambda
+  *. 1} \lambda *. 1)\, A$, where $A$ interprets $\mathsf{a}$.
 %and $1$ is the {\color{blue} constantly $1$-valued
@@
 is isomorphic to $A$, because the only function $f : 1 \to 1$ is the
-identity function.
+identity function. This is different from the interpretation of
+$\mathsf{G}$'s Church encoding from the previous paragraph.
 
 Putting it all together, we see that the Church encoding of
-$\mathsf{G}$ and its primitive representation are not semantically
+$\mathsf{G}$ and its fixpoint representation are not semantically
 equivalent: the interpretation of the former has exactly one data
@@
 elements $a$ of $A \not = \emptyset$ and the interpretation $c$ of
-$\mathsf{c}$.
+$\mathsf{C}$.
 \end{example}
@@
 a GADT contain ``hidden'' elements that are not specified by the
-GADT's syntax and are only accessible, e.g., via applications of
-$\mathsf{map}$? A semanticist, on the other hand, would likely find
-the primitive representation of GADTs entirely reasonable. Indeed,
-they would likely find the nonfunctorial nature of a GADT's Church
-encoding unnerving at best. After all, they would likely argue, the
-data in a GADT shouldn't disappear just because you map a function
-over it! The fact that this happens when GADTs are represented by
-their Church encodings actually highlights how GADTs {\em do not}
+GADT's syntax and are only accessible via applications of
+$\mathsf{map}$? From a semanticist's point of view, however, the
+prinitive representation of GADTs is entirely reasonable. Indeed, they
+would likely find the nonfunctorial nature of a GADT's Church encoding
+unnerving at best. After all, they would likely argue, the data in a
+GADT shouldn't change or become ill-typed just because a function is
+mapped over it. The fact that this happens when GADTs are represented
+by their Church encodings actually highlights how GADTs {\em do not}
 generalize the essential, container-ish nature of ADTs at all. A
 semanticist might therefore conclude that GADTs are seriously
-misnamed. They might even go on to wonder: in what sense do GADTs
-really deserve to be considered {\em data types} at all?
-
-The reader familiar with short cut fusion will quickly see that the
-non-uniform nature of GADTs that are not nested types entails that
-short cut fusion cannot be validated for their Church encodings. In
-fact, it is not even possible to define a $\mathsf{build}$ function,
-which is a crucial ingredient in short cut fusion, for the Church
-encoding of a GADT that is not a nested type. But the fact that not
-all of the elements in the functorial completion of a GADT that is not
-a nested type are directly representable as applications of data
-constructors means that it is also impossible to define a
-$\mathsf{build}$ function for the primitive representation of such a
-GADT. Ultimately this is not surprising, though: if short cut fusion
-were to hold for a GADT then its Church encoding and primitive
-representation would necessarily be semantically equivalent.
+misnamed.
+%They might even go on to wonder: in what sense do GADTs
+%really deserve to be considered {\em data types} at all?
+
+\begin{comment}
+The reader familiar with {\color{red} short cut fusion} will quickly
+see that the non-uniform nature of GADTs that are not nested types
+entails that short cut fusion cannot be validated for their Church
+encodings. In fact, it is not even possible to define a
+$\mathsf{build}$ function, which is a crucial ingredient in short cut
+fusion, for the Church encoding of a GADT that is not a nested
+type. But the fact that not all of the elements in the functorial
+completion of a GADT that is not a nested type are directly
+representable as applications of data constructors means that it is
+also impossible to define a $\mathsf{build}$ function for the fixpoint
+representation of such a GADT. Ultimately this is not surprising,
+though: if short cut fusion were to hold for a GADT then its Church
+encoding and {\color{red} fixpoint} primitive representation would
+necessarily be semantically equivalent.
+\end{comment}
 
@@
 
-{\em Relational parametricity} encodes a powerful notion of
-type-uniformity, or representation independence, for data types in
+{\em Relational parametricity} encodes a powerful notion of type
+uniformity, or representation independence, for data types in
 functional languages. It formalizes the intuition that a polymorphic
@@
 beyond just naturality, such as inhabitation results, and prove the
-equivalence of Church encodings and primitive representations of ADTs
-and hereditarily polynomial nested types by validating short cut
-fusion for them.
+equivalence of Church encodings and fixpoint representations of ADTs
+and
+%hereditarily polynomial
+nested types by validating short cut fusion and other program
+equivalences for them.
 
@@
 \item composition: Composition is the componentwise composition in
-  $\mathit{Set}$. 
+  $\mathit{Set}$. That is, $(g_1,g_2) \circ (f_1,f_2) = (g_1 \circ
+  f_1, g_2 \circ f_2)$, where the composition being defined on the
+  left-hand side is in $\mathit{Rel}$, and the two componentwise
+  compositions on the right-hand side are in $\mathit{Set}$.
 \end{itemize}
@@
 The key idea underlying parametricity is to give each type
-$\mathsf{G[a]}$ with one free variable $\mathsf{a}$ a {\em set
+$\mathsf{G[a]}$\footnote{The notation $\mathsf{G[a]}$ indicates that
+  $\mathsf{G}$ is a type with one hole which has been filled with the
+  type $\mathsf{a}$.}  with one free variable $\mathsf{a}$ a {\em set
   interpretation} $G_0$ taking sets to sets and a \emph{relational
@@
 \begin{example}\label{ex:CE-par}
-Let $\mathsf{t}$ be a term of type $\mathsf{G}$ for the GADT
+Let $\mathsf{t}$ be a term of type $\mathsf{G\,a}$ for the GADT
 $\mathsf{G}$ defined in~\eqref{eq:G}, let $G = (G_0,G_1)$ be the
@@
 
-For primitive representations of GADTs the story is completely
+For fixpoint representations of GADTs the story is completely
 different. If, as intended, the set interpretation of
@@
 interpretation of $\mathsf{k}$ and $F$ is the set interpretation of
-$\mathsf{f}$, then the exact same reasoning give that the relational
+$\mathsf{f}$, then the exact same reasoning gives that the relational
 interpretation of $\mathsf{Lan_k\,f}$ is $\mathit{Lan}_K\,F$, where
@@
 relational interpretation of $\mathsf{f}$. But under these
-interpretations there can be no parametric model, as the following
-example shows.
+interpretations there can be no parametric model. The following
+counterexample establishes this surprising result. It is the main
+technical contribution of this paper.
 
@@
 \mathit{Rel}\,(A, B)$, we have $G_1\,R : \mathit{Rel}\,(G_0\,A,
-G_0\,B)$. But, as we now show, this is not generally possible when
-GADTs are interpreted as the interpretations of their primitive
-representations.
-
-Consider again the GADT $\mathsf{G}$ given by~\eqref{eq:G}. The set
+G_0\,B)$.
+
+Intuitively, when $\mathsf{G}$ is viewed as a fixpoint, its data
+elements include those given by functorial completion. Thus, given any
+relation $S : \mathit{Rel}\,(C, D)$ and any morphism $m : S \to R$,
+$G_1\, R$ must contain all elements of the form $G_1\,m\,x$ for $x \in
+G_1\,S$.  But the two components $m_1 : C \to A$ and $m_2 : D \to B$
+of $m$ cannot be given independently of one another, since
+%$m$ has to send elements related by $S$ to elements
+%related by $R$ in order to be a well-defined morphism of relations.
+$(m_1\,c, m_2\,d)$ must be in $R$ whenever $(c,d)$ is in $S$ in order
+to be a well-defined morphism of relations.  The domain of $G_1\, R$
+thus depends on $A$ {\em and} $B$, rather than simply on $A$.
+Likewise, the codomain of $G_1\,R$ also depends on both $A$ and $B$.
+The domain and codomain therefore cannot simply be $G_0\, A$ and
+$G_0\, B$, respectively.  This suggests that GADTs might fail to have
+relational interpretation, and might thus fail to have a parametric
+model, as described above.
+%But, as we now show, this is not generally possible when
+%GADTs are interpreted as the interpretations of their primitive
+%representations.
+
+We can make this informal argument formal by providing a concrete
+counterexample.  Consider again the GADT $\mathsf{G}$ given
+by~\eqref{eq:G}. The set interpretation of $\mathsf{G}$ is
+$\mathit{Lan}_{\lambda *\!. 1}\,\lambda *\!. 1$, i.e., is, by the
+reasoning of Example~\ref{ex:g1}, the identity functor on
+$\mathit{Set}$. By the exact same reasoning, this time in
+$\mathit{Rel}$ rather than in $\mathit{Set}$, the relational
 interpretation of $\mathsf{G}$ is $\mathit{Lan}_{\lambda
-  *\!. 1}\,\lambda *\!. 1$, i.e., is, by the reasoning of
-Example~\ref{ex:g1}, the identity functor on $\mathit{Set}$. By the
-exact same reasoning, this time in $\mathit{Rel}$ rather than in
-$\mathit{Set}$, the relational interpretation of $\mathsf{G}$ is
-$\mathit{Lan}_{\lambda *\!.\,\mathit{Eq}_1} \,\lambda
-*\!.\,\mathit{Eq}_1$, where $\lambda *\!.\,\mathit{Eq}_1$ is the
-constantly $\mathit{Eq}_1$-valued functor from the category
-$\mathit{Rel}\,^0$ with a single object (also called $*$) to
-$\mathit{Rel}$.  Indeed, the interpretation is still a left Kan
-extension, but now it is the left Kan extension determined by the
+  *\!.\,\mathit{Eq}_1} \,\lambda *\!.\,\mathit{Eq}_1$, where $\lambda
+*\!.\,\mathit{Eq}_1$ is the constantly $\mathit{Eq}_1$-valued functor
+from the category $\mathit{Rel}\,^0$ with a single object (also called
+$*$) to $\mathit{Rel}$.  Indeed, the interpretation is still a left
+Kan extension, but now it is the left Kan extension determined by the
 functor interpreting the type $\mathsf{\lambda *\!. 1}$ in
@@
 $(\mathit{Lan}_{\lambda *. 1}\,\lambda *\!. 1)\,B$, i.e., between the
-sets $A$ and $B$. However, we now give a counterexample showing that
+sets $A$ and $B$. However,
+%we now give a counterexample showing that
 this need not be the case.
@@
 The only morphism $f : \mathit{Eq}_1 \to \mathit{Eq}_1$ is $f = (!,
-!)$. Since this morphism does satisfy the condition $f (\ast, \ast) =
-(\ast, \ast)$, we have that $(\ast, m) \approx (\ast, m')$ if and only
-if $m = m'$. Thus
+!)$, where $!$ is the unique function from $1$ to $1$.  Since this
+morphism does satisfy the condition $f (\ast, \ast) = (\ast, \ast)$,
+we have that $(\ast, m) \approx (\ast, m')$ if and only if $m =
+m'$. Thus
 \[ \big( \bigcup_{m\, :\,
@@
 Lemma does not hold for models in which GADTs are interpreted by the
-interpretations of their primitive representations, such models cannot
+interpretations of their fixpoint representations, such models cannot
 possibly be parametric.
@@
 
-We note that is actually possible to construct a simpler
-counterexample to the Identity Extension Lemma for primitive GADTs
-using the relation $R = (1,\emptyset, \emptyset)$. However, this
-relation is somewhat artificial, in the sense that its domain is
-larger than is strictly necessary to define an empty relation. We
-therefore give the above example using the relation $(1,2,1 \times 2)$
-instead.
+We note that it is actually possible to construct a simpler
+counterexample to the Identity Extension Lemma for fixpoint
+representations of GADTs using the relation $R = (1,\emptyset,
+\emptyset)$. However, this relation is somewhat artificial, in the
+sense that its domain is larger than is strictly necessary to define
+an empty relation. We therefore give the above example using the
+relation $(1,2,1 \times 2)$ instead.
 
 Taken together, Examples~\ref{ex:CE-par} and~\ref{ex:prim-par} show
-that Church encodings and primitive representations of GADTs behave
+that Church encodings and fixpoint representations of GADTs behave
 very differently with respect to parametricity.  This contrasts
 sharply with the fact that both ADTs and nested types have the same
-parametricity properties regardless of whether they are represented by
-their Church encodings or their primitive
-representations.
+parametricity properties regardless of whether they are represented as
+their Church encodings or their fixpoint representations.
 
@@
 contrasting ways: as completely determined by their syntax, in which
-case they are represented by their Church encodings, or as the
+case they are represented as their Church encodings, or as the
 functorial completion of their syntax, in which case they are
-represented as primitive data types.  But regardless of which
-representation we choose, GADTs fail to have {\em all} expected
-parametricity properties. Indeed, we can read the results of
-Section~\ref{sec:par} as showing that i)
-%some standard parametricity results (e.g.,
-inhabitation results for calculi representing GADTs as Church
-encodings cannot be extended to calculi representing them as primitive
-data types, and ii)
-%some
-naturality results
-%(e.g., rearrange-transform properties)
-for calculi representing GADTs as primitive data types cannot be
-extended to calculi representing them as Church encodings. So, with
-neither representation can we obtain {\em both} naturality
-consequences {\em and} non-naturality consequences of
-parametricity. Intuitively, this is because a GADT is not a data type
-in the usual container-ish sense, and completing it so that it becomes
-one destroys its relationally uniform --- i.e., parametric ---
-behavior.
-%required in a parametric model for parametricity.
-Which representation of GADTs a user chooses can therefore have
-profound consequences for programming with, and reasoning about, them.
+represented as fixpoints.  But regardless of which representation we
+choose, GADTs fail to have {\em all} expected parametricity
+properties. Indeed, we can read the results of Section~\ref{sec:par}
+as showing that {\em i}) inhabitation and other non-naturality results
+for calculi representing GADTs as Church encodings cannot be extended
+to calculi representing them as fixpoints, and {\em ii}) naturality
+results for calculi representing GADTs as fixpoints cannot be extended
+to calculi representing them as Church encodings. So, neither
+representation guarantees {\em both} naturality consequences {\em and}
+non-naturality consequences of parametricity. Intuitively, this is
+because a GADT is not a data type in the usual container-ish sense,
+and completing it so that it becomes one destroys its relationally
+uniform --- i.e., parametric --- behavior.  Which representation of
+GADTs a user chooses can therefore have significant consequences for
+programming with, and reasoning about, them.
 
 Although there are no parametric models for calculi representing GADTs
-as primitive data types, GADT analogues of most of Wadler's free
-theorems for ADTs still hold.
+as fixpoints, GADT analogues of most of Wadler's free theorems for
+ADTs still hold.
 %all possible rearrange-transform properties
@@
 types' {\em functorial semantics}. Even for data types whose Church
-encodings and primitive representations coincide, the analysis in this
+encodings and fixpoint representations coincide, the analysis in this
 paper clearly distinguishes those free theorems that are actually
@@
 
-\begin{comment}
-On the other hand, despite there
-being no models interpreting type constructors as functors for calculi
-incorporating GADTs as Church encodings,
-%and thus there being no models validating all
-%rearrange-transform properties for these types,
-parametric models for such calculi do exist, and these can be used,
-e.g., to determine which instances of a GADT are inhabited.
-\end{comment}
-
-Atkey's parametric model for $F_\omega$ represents data types as
+There are treatments of GADTs beyond those discussed in the main body
+of this paper.  Atkey's parametric model for $F_\omega$
+from~\cite{atk12} represents data types --- including GADTs --- as
 Church encodings. It requires the user to supply a $\mathsf{map}$
@@
 constructor does not imply functoriality of its fixpoint, so the data
-type itself still need not necessarily support a
-$\mathsf{map}$. Similarly,~\cite{vw10} presents a parametric model for
+type itself still need not necessarily support a $\mathsf{map}$ in
+Atkey's model. Similarly,~\cite{vw10} presents a parametric model for
 an extension of $F_\omega$ that supports type equality, and thus can
-encode GADTs, but still does not guarantee functoriality. In fact,
-%encode GADTs. But functoriality is not guaranteed here, either; in fact,
-the parametric properties of GADTs described in the preliminary
-work~\cite{vw06} are all inhabitation results, rather than naturality
-results. Finally,~\cite{ms09} encodes GADTs as Scott encodings rather
-than Church encodings but, again, cites only inhabitation results for
-them.
-
+encode GADTs, but this model still does not guarantee functoriality;
+significantly, the parametric properties of GADTs described in the
+precursor work~\cite{vw06} to~\cite{vw10} are all inhabitation results
+rather than naturality results. In~\cite{ms09} GADTs are represented
+as Scott encodings rather than Church encodings but, again, only
+inhabitation results are cited for them. GADTs are treated explicitly
+as fixpoints of discrete functors in~\cite{jg08}, as initial algebras
+of dependent polynomial functors in~\cite{gh03,hf11}, and as indexed
+containers in~\cite{ma09}. The latter two treatments move toward
+seeing GADTs as data types in a dependent type theory. A categorical
+parametric model of dependent types has been given in~\cite{agj14},
+but, as with the models mentioned above, this model also does not
+guarantee that data types have functorial semantics.
+
+\bibliographystyle{eptcs}
 \bibliography{gadt-story-references}
@@
 (that can be characterized by a grammar for which every type is
-interpreted as a functor, including those characterized as least
+interpreted as a functor, including those characterized as
 fixpoints, i.e., as carriers of initial algebras of functors, i.e., as

### 2021-05-06 20-gadts/21-LSFA/GADT-story.tex
final LFSA paper

@@
 for the syntactic generalization of ADTs known as nested
-types~\cite{bm98}. Generalized algebraic data types (GADTs) generalize
-nested types~\cite{pvww06} --- and thus further generalize ADTs ---
-syntactically (see Section~\ref{sec:gadts} below).  Indeed, we have:
+types~\cite{bm98}. Generalized algebraic data types
+(GADTs)~\cite{pvww06} generalize nested types --- and thus further
+generalize ADTs --- syntactically (see Section~\ref{sec:gadts} below):
 
@@
 types as their Church encodings or as fixpoints.  The main result of
-this paper shows that the representation chosen determines whether or
-not a language supporting GADTs can have parametric
-models~\cite{rey83}. It thus determines whether or not GADTs can enjoy
-consequences of parametricity such as representation
+this paper shows that the choice of representation of data types in a
+language supporting GADTs determines whether or not that language can
+have parametric models~\cite{rey83}. It thus determines whether or not
+GADTs can enjoy consequences of parametricity such as representation
 independence~\cite{adr09,dnb12}, equivalences between
@@
 syntax.  Completing a GADT's syntax from a funct{\em ion} on types to
-a funct{\em or} on types is necessary if it is to be represented as a
+a funct{\em or} on types is necessary if its interpretation is to be a
 fixpoint~\cite{tfca}, so functoriality is inherent in fixpoint
@@
 $\mathsf{map\,f\,s}$ for all types $\mathsf{t_1}$ and $\mathsf{t_2}$,
-all functions $\mathsf{f : t_1 \to t_2}$ definable in the language and
+all functions $\mathsf{f :: t_1 \to t_2}$ definable in the language and
 all $\mathsf{s :: Seq\,t_1}$, as well as all data elements of the form
@@
 adds no new data elements to the syntax of a GADT that is an ADT or
-other nested type. This perfectly explains why Church encodings and
+other nested type, which perfectly explains why Church encodings and
 fixpoint representations coincide for ADTs and other nested
@@
 container that can be filled with data of any type in the same way
-that ADTs and nested types can. In this situation, a GADT cannot be
-seen as a functor, and so must necessarily be represented as its
+that ADTs and nested types can. In this situation, a GADT cannot
+denote a functor, and so must necessarily be represented as its
 Church encoding.
@@
 reflect functoriality. Since the Church encoding of a GADT that is not
-a nested type is not a functor, such a GADT must be viewed as its
-fixpoint representation. The functorial completion inherent in the
-fixpoint representation of a GADT adds to its syntax those, and only
-those, data elements needed to support a (functorial) $\mathsf{map}$
-function.  For example, let $\mathsf{1}$ be the unit type whose single
-element is also denoted by $\mathsf{1}$, and let $\mathsf{G}$ be the
-GADT defined by\looseness=-1
+a nested type does not denote a functor, to have a functorial
+semantics such a GADT must be viewed as its fixpoint
+representation. The functorial completion inherent in the fixpoint
+representation of a GADT adds to its syntax those, and only those,
+data elements needed to support a functorial $\mathsf{map}$ function.
+For example, let $\mathsf{1}$ be the unit type whose single element is
+also denoted by $\mathsf{1}$, and let $\mathsf{G}$ be the GADT defined
+by\looseness=-1
 \begin{equation}\label{eq:G}
@@
 functorial completion because it is not possible to define a function
-from $\mathsf{1}$ into $\mathsf{0}$. Note, however, that since not all
-instances of the fixpoint representation of a GADT need be inhabited,
-fixpoint representations of GADTs are not, in general, nested types.
+from $\mathsf{1}$ into $\mathsf{0}$.
+%Note, however, that since not all
+%instances of the fixpoint representation of a GADT need be inhabited,
+%fixpoint representations of GADTs are not, in general, nested types.
 
@@
 GADT programmer and the semanticist are both valid, the two are
-irreconcilable. Importantly, which representation of GADTs is adopted
-in any particular setting has significant consequences for the ways
-they can be used and reasoned about there. In particular, the way that
-GADTs are represented has deep implications for parametric reasoning
-about them. Specifically, a programmer who views GADTs as their Church
-encodings cannot safely use program transformations or reasoning
-principles that involve $\mathsf{map}$ functions for them, although
-they may be able to program with and reason about GADTs using other
-consequences of parametricity, such as type inhabitation results. On
-the other hand, a semanticist who views GADTs as fixpoints will have
-all naturality-based program transformations and reasoning principles
-for GADTs at their disposal since these all derive from
-functoriality. But since, as we show in Section~\ref{sec:par}, no
-parametric model can be constructed for fixpoint representations of
-GADTs, non-naturality consequences of parametricity will not
-necessarily hold for them.  Overall, we show that, as with software
-engineering's iron triangle~\cite{iron-triangle}, we can have any two
-of {\em GADTs}, {\em functoriality}, and {\em parametricity} we like,
-but we cannot have all three.
+irreconcilable. Importantly, which of our two representations of data
+types is adopted in any particular setting has significant
+consequences for the ways GADTs can be used and reasoned about
+there. In particular, the way that GADTs are represented has deep
+implications for parametric reasoning about them. Specifically, a
+programmer who views GADTs as their Church encodings cannot safely use
+program transformations or reasoning principles that involve
+$\mathsf{map}$ functions for them, although they may be able to
+program with and reason about GADTs using other consequences of
+parametricity, such as type inhabitation results. On the other hand, a
+semanticist who views GADTs as fixpoints will have all
+naturality-based program transformations and reasoning principles for
+GADTs at their disposal since these all derive from functoriality. But
+since, as we show in Section~\ref{sec:par}, no parametric model can be
+constructed for fixpoint representations of GADTs, non-naturality
+consequences of parametricity will not necessarily hold for them.
+Overall, we show that, as with software engineering's iron
+triangle~\cite{iron-triangle}, we can have any two of {\em GADTs},
+{\em functoriality}, and {\em parametricity} we like, but we cannot
+have all three.
 
 The goal of this paper is to show how the above observations can be
-made precise. We thereby answer the question we posed above in the
+made precise, and thereby to answer the question we posed above in the
 affirmative: Yes, it really does matter how we represent GADTs.
@@
 another type $\mathsf{b}$ simply by applying a given function
-$\mathsf{f \,:\,a \to b}$ to each of its elements. Indeed, every ADT
+$\mathsf{f \,::\,a \to b}$ to each of its elements. Moreover, every ADT
 $\mathsf{T}$ can be made an instance of Haskell's $\mathsf{Functor}$
@@
     (T\,a \to T\,b)}$ witnessing that a type constructor $\mathsf{T}$
-  is an instance of Haskell's $\mathsf{Functor}$ class. This
-  emphasizes that $\mathsf{fmap}$ functions in Haskell are intended to
+  is an instance of Haskell's $\mathsf{Functor}$ class. We
+  emphasize that $\mathsf{fmap}$ functions in Haskell are intended to
   satisfy syntactic reflections of the functor laws ---
@@
   is instead left to the good intentions of the programmer.}
-Moreover, given a type-independent way of rearranging an ADT
+Then, given a type-independent way of rearranging an ADT
 structure's shape $\mathsf{T\,a}$ into the shape for another ADT
@@
 $\mathsf{T\,b}$ and then rearrange that resulting structure into one
-of type $\mathsf{T'\,b}$. For example, if $\mathsf{f : a \to b}$,
-$\mathsf{t \,:\,Tree\,a}$, and $\mathsf{g : Tree\,a\to List\,a}$
+of type $\mathsf{T'\,b}$. For example, if $\mathsf{f :: a \to b}$,
+$\mathsf{t \,::\,Tree\,a}$, and $\mathsf{g :: Tree\,a\to List\,a}$
 arranges lists into trees in a type-independent way, then we have the
@@
 \[\mathsf{Tree\,a} = \mathsf{\forall b.\,(a \to b) \to (b \to a \to b
-  \to b) \to b}\] respectively. The argument types are abstractions of
-the types of ADTs' constructors. For instance, $\mathsf{b}$ abstracts
-the type $\mathsf{List\,a}$ of the constructor $\mathsf{Nil}$ for
-lists, and $\mathsf{a \to b \to b}$ abstracts the type $\mathsf{a \to
-  List\,a \to List\,a}$ of the constructor $\mathsf{Cons}$.
+  \to b) \to b}\] respectively. The argument types in a Church
+encoding of an ADT are abstractions of the types of the ADT's
+constructors. For instance, $\mathsf{b}$ abstracts the type
+$\mathsf{List\,a}$ of the constructor $\mathsf{Nil}$ for lists, and
+$\mathsf{a \to b \to b}$ abstracts the type $\mathsf{a \to List\,a \to
+  List\,a}$ of the constructor $\mathsf{Cons}$.
 
@@
 \[\mathsf{Tree} = \mathsf{\forall a. \forall b.\,(a \to b) \to (b \to
-  a \to b \to b) \to b}\] respectively. This is what allows an ADT's
-associated type constructor to be made an instance of Haskell's
-$\mathsf{Functor}$ class. For example, the
-$\mathsf{map}$ function
-from Haskell's standard library
-makes the type constructor $\mathsf{List}$ an
-instance of the $\mathsf{Functor}$ class, and for $\mathsf{Tree}$ we
-can define
+  a \to b \to b) \to b}\] respectively. This fact is what allows an
+ADT's associated type constructor to be made an instance of Haskell's
+$\mathsf{Functor}$ class. For example, the $\mathsf{map}$ function
+from Haskell's standard library makes the type constructor
+$\mathsf{List}$ an instance of the $\mathsf{Functor}$ class, and for
+$\mathsf{Tree}$ we can define
 \[\begin{array}{lll}
@@
 \end{array}\]
-Note, however, that the Church encoding of an ADT carries with it no
-expectation whatsoever that such a type-and-data-uniform,
-structure-preserving, data-changing $\mathsf{map}$ function can be
-defined or that, if one can, it will satisfy a rearrange-transform
-property. But such a $\mathsf{map}$ function can always be defined for
-any ADT precisely because its Church encoding and its fixpoint
-representation necessarily coincide, and this $\mathsf{map}$ function
-will, indeed, satisfy a rearrange-transform property.
+Note that the Church encoding of an ADT carries with it no expectation
+whatsoever that such a type-and-data-uniform, structure-preserving,
+data-changing $\mathsf{map}$ function can be defined or that, if one
+can, it will satisfy a rearrange-transform property. (Of course, such
+a $\mathsf{map}$ function can always be defined for any ADT precisely
+because its Church encoding and its fixpoint representation coincide,
+and this $\mathsf{map}$ function will necessarily satisfy a
+rearrange-transform property.)
 
@@
 Fixpoint representations capture in syntax the fact that ADTs can be
-considered as fixpoints of functors. For example $\mathsf{List\,a}$ is
+considered as fixpoints of functors. For example, $\mathsf{List\,a}$ is
 indeed a fixpoint of $\mathsf{F_{List\,a}}$, where
-$\mathsf{F_{List\,a}\,X} = \mathsf{1 + a \times X}$ as determined
-by~\eqref{eq:list}, i.e., is indeed such that
+$\mathsf{F_{List\,a}\,X} = \mathsf{1 + a \times X}$
+by~\eqref{eq:list}. That is,
 \[\mathsf{List\,a} = \mathsf{1 + a \times List\,a}\] 
@@
 F_{\mathit{List}\,a}$, where the functor $F_{\mathit{List}\,a}$ models
-the type constructor $\mathsf{F_{List\,a}}$.\footnote{Throughout, we
-  use $\mathtt{sans\; serif}$ font for program text and $\mathit{math
-    \; italic}$ font for semantic objects.}  Similarly, the ADT
-$\mathsf{Tree\,a}$ can be seen to be a fixpoint of
+the type constructor $\mathsf{F_{List\,a}}$.\footnote{Throughout this
+  paper, we use $\mathtt{sans\; serif}$ font for program text and
+  $\mathit{math \; italic}$ font for semantic objects.}  Similarly,
+the ADT $\mathsf{Tree\,a}$ can be seen to be a fixpoint of
 $\mathsf{F_{Tree\,a}}$, where $\mathsf{F_{Tree\,a}\,X} = \mathsf{a + X
   \times a \times X}$ by~\eqref{eq:tree}, so that $\mathsf{Tree\, a} =
-\mathsf{ \mu F_{Tree\,a}}$ is modeled by $\mu F_{\mathit{Tree}\,a}$,
-where $F_{\mathit{Tree}\,a} $ models $\mathsf{F_{Tree\,a}}$.
+\mathsf{ \mu F_{Tree\,a}}$ is modeled by $\mu F_{\mathit{Tree}\,a}$ if
+$F_{\mathit{Tree}\,a} $ models $\mathsf{F_{Tree\,a}}$.
 
@@
   to satisfy syntactic reflections of the functor laws in
-  $\mathit{Set}$ even though there is no mechanism in Haskell for
+  $\mathit{Set}$, even though there is no mechanism in Haskell for
   enforcing this.} $\mathsf{map}$ functions.
@@
 
-In any parametric model, the interpretations of the Church encoding
-and the fixpoint representation of an ADT or a nested type will
-necessarily be isomorphic. But whereas it is impossible even to {\em
+In any parametric model, the Church encoding and the fixpoint
+representation of an ADT or a nested type will necessarily be
+semantically equivalent. But whereas it is impossible even to {\em
   state} the rearrange-transform property for their Church encodings
-(unless $\mathsf{map}$ functions have somehow been defined for them),
-for their fixpoint representations such a property is simply the
-reflection back into syntax of the instance of naturality for the
-type-independent function that rearranges structures of type
+(unless functorial $\mathsf{map}$ functions have somehow been defined
+for them), for their fixpoint representations such a property is
+simply the reflection back into syntax of the instance of naturality
+for the type-independent function that rearranges structures of type
 $\mathsf{T\,a}$ into ones of type $\mathsf{T'\,a}$ and the
-content-transforming functions $\mathsf{map_T\,f}$ and
+structure-preserving, data-changing functions $\mathsf{map_T\,f}$ and
 $\mathsf{map_{T'}\,f}$ for a function $\mathsf{f :: a \to b}$, where
@@
 
-GADTs relax the restriction on the type instances appearing in a data
-type definition. The special form of GADTs known as {\em nested
+{\em Generalized algebraic data types} (GADTs)~\cite{pvww06} relax the
+restriction on the type instances appearing in a data type
+definition. The special form of GADTs known as {\em nested
   types}~\cite{bm98} allow the data constructors of a GADT to take as
@@
 $\mathsf{a}$ appears in both $\mathsf{PTree\,a}$ and $\mathsf{Seq\,a}$
-as a ``dummy'' parameter used only to give the kind \verb|* -> *| of
+as a ``dummy'' parameter used only to give the kind $\mathsf{* \to *}$ of
 the type constructors $\mathsf{PTree}$ and $\mathsf{Seq}$. This is
 explicitly captured in the alternative ``kind signature'' Haskell
-syntax
+syntax, which represents $\mathsf{PTree}$ and $\mathsf{Seq}$ as
+\[\begin{array}{l}
+\mathsf{data\, PTree :: * \to *\,where}\\
+\mathsf{\;\;\;\;\;\;\;\;PLeaf \,\,\,\;::\, a \to PTree\,a}\\
+\mathsf{\;\;\;\;\;\;\;\;PNode\,\, ::\, PTree \,(a \times a) \to
+  PTree\,a}\\ 
+\end{array}\]
+\noindent
+and
 \[\begin{array}{l}
@@
 \mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Seq\,a}\\
-\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, ::\, Seq \,a \to Seq\,b \to
+\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\; ::\, Seq \,a \to Seq\,b \to
   Seq\,(a \times b)}\\ 
 \end{array}\]
-A GADT --- even a nested type --- thus does not define a {\em family
+respectively. A GADT --- even a nested type --- thus does not define a {\em family
   of inductive types}, one for each type argument, like an ADT does,
@@
 
-GADTs are used in precisely those situations in which different
-behaviors at different instances of a data type are desired. This is
-achieved by allowing the programmer to give the type signatures of the
-GADT's data constructors independently, and then using pattern
-matching to force the desired type refinement. Applications of GADTs
-include generic programming, modeling programming languages via
+GADTs that are not nested types are used in precisely those situations
+in which different behaviors at different instances of a data type are
+desired. This is achieved by allowing the programmer to give the type
+signatures of the GADT's data constructors independently --- as is
+made explicit by the alternative syntax above --- and then using
+pattern matching to force the desired type refinement. Applications of
+GADTs include generic programming, modeling programming languages via
 higher-order abstract syntax, maintaining invariants in data
@@
 $\mathsf{Equal}$ must be the syntactic reflection of semantic equality
-function $\mathit{Equal}$. The type $\mathsf{Equal}$ can either be
-defined via GADT syntax, as in
+function $\mathit{Equal}$. The type $\mathsf{Equal}$ can be
+defined via GADT syntax as
 \[\begin{array}{l}
@@
 \noindent
-or, equivalently, represented by its Church encoding
+Its Church encoding (in a higher-kinded calculus such as
+$F_\omega$~\cite{atk12}) is
 \[\mathsf{Equal\,a\,b} = \mathsf{\forall f.\, (\forall c.\,
-  f\,c\,c) \to f\,a\,b}\] in a higher-kinded calculus such as
-$F_\omega$~\cite{atk12}. Following the technique in Section~4.5
-of~\cite{atk12}, we can represent the GADT $\mathsf{Seq}$ as its
-Church encoding
+  f\,c\,c) \to f\,a\,b}\] Following the technique in Section~4.5
+of~\cite{atk12}, we can rewrite its Church encoding as
 \[\mathsf{Seq\,a} = \mathsf{\forall f.\,(\forall d. \,(f\,d + \exists b
@@
   \to f \,d) \to f\, a}\]
-This is logically equivalent to the more intuitive encoding
+which is logically equivalent to the more intuitive encoding
 \[\mathsf{Seq\,a} = \mathsf{\forall f.\, (\forall d.\,d \to f\, d)
@@
 %evident from its Haskell definition as discussed in
-Section~\ref{sec:intro}.
+Section~\ref{sec:intro}. In that setting, given a function $\mathsf{f
+  :: (a \times b) \to e}$, the term $\mathsf{map_{Equal\,(a \times
+    b)}\,f\, Refl}$ would have type $\mathsf{Equal\,(a \times b)\,e}$.
+But we have no way to produce a term of this type in the absence of a
+functorial $\mathsf{map}$ function for $\mathsf{Equal}$, and thus no
+way to produce a term of type $\mathsf{Seq\,e}$ using the
+$\mathsf{Pair}$ constructor, as would be required by the clause of
+$\mathsf{map_{Seq}}$ for $\mathsf{Pair}$.
 
@@
 GADT's syntax. In this latter reading, GADTs can, like ADTs, be
-modeled as fixpoints of higher-order functors. Syntactically, the use
-of higher-order funct{\em ions} is essential; since the type arguments
-to the GADT being defined are not necessarily uniform across all of
-its instances in the types of its data constructors, GADTs cannot be
-seen as fixpoints of first-order functors the way ADTs
-can. Semantically, the use of higher-order funct{\em ors} is
-essential, as in the case of ADTs, to guarantee the existence of the
-fixpoints being computed~\cite{tfca}.
-
-Consider, again, the GADT $\mathsf{Seq}$. Because its type argument
-varies in the instances of $\mathsf{Seq}$ appearing in the types of
-its data constructor $\mathsf{Pair}$, $\mathsf{Seq}$ cannot be
-modeled as the fixpoint of any first-order functor. As shown
+modeled as fixpoints of higher-order functors. Syntactically, {\em
+  higher-orderness} is essential; since the type arguments to the GADT
+being defined are not necessarily uniform across all of its instances
+in the types of its data constructors, GADTs cannot be seen as
+first-order fixpoints the way ADTs can. Semantically, the use of
+(higher-order) {\em functors} is essential, as in the case of ADTs, to
+guarantee the existence of the (higher-order) fixpoints being
+computed~\cite{tfca}.
+
+To illustrate, consider again the GADT $\mathsf{Seq}$. Because its
+type argument varies in the instances of $\mathsf{Seq}$ appearing in
+the types of its data constructor $\mathsf{Pair}$, $\mathsf{Seq}$
+cannot be modeled as the fixpoint of any first-order functor. As shown
 in~\cite{jp19}, it can, however, be modeled as a solution to the
@@
 both extends the image of $K$ to $\cal D$ and agrees with $F$ on $\cal
-C$, in the sense that, for any other such functor $H$, there is a
+C$, in the sense that, for any other such functor $G$, there is a
 morphism of functors (i.e., a natural transformation) from $Lan_K \,F$
-to $H$. Formally, this is captured by the following
+to $G$. Formally, this is captured by the following
 definition~\cite{mac71}:
@@
 natural transformation $\eta : F \to (\mathit{Lan}_K\,F) \circ K$ such
-that, for every functor $H : {\cal E} \to {\cal D}$ and natural
-transformation $\gamma : F \to H \circ K$, there exists a unique
-natural transformation $\delta : \mathit{Lan}_K\,F \to H$ such that
+that, for every functor $G : {\cal E} \to {\cal D}$ and natural
+transformation $\gamma : F \to G \circ K$, there exists a unique
+natural transformation $\delta : \mathit{Lan}_K\,F \to G$ such that
 $(\delta K) \circ \eta = \gamma$. This is depicted in the diagram
@@
 \ar[ur, bend left, "{\mathit{Lan}_{K}\,F}"{name=Lan, description}]
-\ar[ur, bend right, "{H}"'{name=L, right}, ""'{name=Lr, right}]
+\ar[ur, bend right, "{G}"'{name=L, right}, ""'{name=Lr, right}]
 \ar[Rightarrow, bend left = 45, from=Fa, to=Lr, "{\gamma}" near start]
@@
 interpreted as sets, a calculus must support a primitive construct
-$\mathsf{Lan}$ such that the type constructor $\mathsf{Lan_k\,f}$ is
+$\mathsf{Lan}$ such that the type constructor $\mathsf{Lan_K\,F}$ is
 the syntactic reflection of the left Kan extension $\Lan_K \,F$ of the
-functor $F$ interpreting $\mathsf{f}$ along the functor $K$
-interpreting $\mathsf{k}$. In this setting, the categories
+functor $F$ interpreting $\mathsf{F}$ along the functor $K$
+interpreting $\mathsf{K}$. In this setting, the categories
 $\mathcal{C}$, $\mathcal{D}$, and $\mathcal{E}$ must all be of the
 form $\mathit{Set}^n$ for some $n$. Using $\mathsf{Lan}$ we can then
-rewrite the type of a constructor $\mathsf{C :: f a \to g(k\,a)}$ as
-$\mathsf{C :: (Lan_k\,f)\,a \to g\,a}$ since morphisms (i.e., natural
-transformations) from $F$ to $G \circ K$ are in one-to-one
+rewrite the type of a constructor $\mathsf{C :: F\, a \to G\,(K\,a)}$
+as $\mathsf{C :: (Lan_K\,F)\,a \to G\,a}$ since morphisms (i.e.,
+natural transformations) from $F$ to $G \circ K$ are in one-to-one
 correspondence with those from $\mathit{Lan}_K\,F$ to $G$ (see
@@
     d. c \times d} \lambda c d. \phi c \times \phi d)\,b)\,a}\] The
-fact that $Lan_K\,G$ is the best functorial approximation to $G$
+fact that $Lan_K\,F$ is the best functorial approximation to $F$
 factoring through $K$ means that the type constructor
-$\mathsf{Lan_k\,g}$ computes the smallest collection of data that is
+$\mathsf{Lan_K\,F}$ computes the smallest collection of data that is
 generated by the corresponding GADT data constructor's syntax and also
@@
 and~\ref{sec:gadts}. But whereas the two representations are always
-isomorphic for ADTs and nested types, they are not, in general, the
-same for GADTs. This will be proved in the next section.
+semantically equivalent for ADTs and nested types, they are not, in
+general, semantically equivalent for GADTs. This will be shown
+formally in the next section.
 
@@
 To see that the Church encoding of a GADT and its fixpoint
-representation need not be the same, consider again the GADT
-$\mathsf{G}$ defined in~\eqref{eq:G}. Despite its simplicity, this
-GADT serves as an informative case study highlighting the differences
-between a GADT's Church encoding and its fixpoint representation ---
-even if we consider only the data elements it contains, and ignore
-whether or not it supports a $\mathsf{map}$ function.
+representation need not be semantically equivalent, consider again the
+GADT $\mathsf{G}$ defined in~\eqref{eq:G}. Despite its simplicity,
+this GADT serves as an informative case study highlighting the
+differences between a GADT's Church encoding and its fixpoint
+representation --- even if we consider only the data elements it
+contains, and ignore whether or not it supports a $\mathsf{map}$
+function.
 
@@
 \[\mathsf{G\, a = Equal\,1\,a}\]
-As this makes clear,
+As the above equation makes clear,
 %Syntactically, the GADT $\mathsf{G}$ defined in~\eqref{eq:G} comprises
@@
 type $\mathsf{1}$. The interpretation of the Church encoding of
-$\mathsf{G}$ is thus, in accordance with Section~\ref{sec:CEs}, the
+$\mathsf{G}$ is thus, as explained in Section~\ref{sec:CEs}, the
 function whose value is a singleton set when the interpretation of
@@
 $\mathsf{C :: 1 \to G\,1}$ or, using~\eqref{eq:nat-transfs}, as
-$\mathsf{C :: (Lan_{\lambda *. 1}\,\lambda *. 1)\,a \to G\, a}$, where
-$\mathsf{\lambda *. 1}$ is the syntactic reflection of the constantly
+$\mathsf{C :: (Lan_{\lambda u. 1}\,\lambda u. 1)\,a \to G\, a}$, where
+$\mathsf{\lambda u. 1}$ is the syntactic reflection of the constantly
 $1$-valued functor from the category $\mathit{Set}^0$ with a single
-object $*$ to $\mathit{Set}$. We can therefore represent $\mathsf{G}$
-as
+object to $\mathit{Set}$. We can therefore represent $\mathsf{G}$ as
 \[\mathsf{G\,a} = \mathsf{(\mu \phi. \lambda b.
- (Lan_{\lambda *. 1} \,\lambda *.  1)\,b)\,a}\] The interpretation of
+ (Lan_{\lambda u. 1} \,\lambda u.  1)\,b)\,a}\] The interpretation of
 $\mathsf{G}$ is therefore obtained by computing the fixpoint of the
-interpretation of the body $\mathsf{\lambda b. (Lan_{\lambda *. 1}
-  \,\lambda *. 1)\,b}$ of the syntactic fixpoint $\mathsf{\mu
-  \phi. \lambda b.  (Lan_{\lambda *. 1} \,\lambda *.  1)\,b}$ and
+interpretation of the body $\mathsf{\lambda b. (Lan_{\lambda u. 1}
+  \,\lambda u. 1)\,b}$ of the syntactic fixpoint $\mathsf{\mu
+  \phi. \lambda b.  (Lan_{\lambda u. 1} \,\lambda u.  1)\,b}$ and
 applying the result to $\mathsf{a}$. But since the recursion variable
@@
 interpretation of $\mathsf{G\,a}$ is therefore $(\mathit{Lan}_{\lambda
-  *. 1} \lambda *. 1)\, A$, where $A$ interprets $\mathsf{a}$.
+  u. 1} \lambda u. 1)\, A$, where $A$ interprets $\mathsf{a}$.
 %and $1$ is the {\color{blue} constantly $1$-valued
@@
 It turns out, however, that, for any set $A$, $(\mathit{Lan}_{\lambda
-  *. 1} \lambda *. 1)\,A$ is, in fact, exactly $A$. Indeed,
+  u. 1} \lambda u. 1)\,A$ is, in fact, exactly $A$. Indeed,
 Proposition~7.1 of~\cite{blw03} gives that $(\mathit{Lan}_{\lambda
-  *. 1} \lambda *. 1)\,A$ can be computed as the set of all functions
+  u. 1} \lambda u. 1)\,A$ can be computed as the set of all functions
 $a : 1 \to A$ modulo the smallest equivalence relation $\sim$ such
@@
 
-In addition to not being semantically equivalent, the two
-representations of GADTs have very different implications for
-parametricity. We explore the differences in parametricity results for
-GADTs with respect to the two representations in the next section.
+Because they are not semantically equivalent, the two representations
+of GADTs have very different implications for parametricity. We
+explore the differences in parametricity results for the two
+representations of GADTs in the next section.
 
@@
 later popularized as Wadler's ``theorems for free''~\cite{wad89}, so
-called because it can deduce properties of programs in such languages
-solely from their types, i.e., with no knowledge whatsoever of the
-text of the programs involved.  Most of Wadler's free theorems are
-consequences of naturality for polymorphic list-processing
-functions. However, parametricity can also derive results that go
-beyond just naturality, such as inhabitation results, and prove the
-equivalence of Church encodings and fixpoint representations of ADTs
-and
+called because it can deduce properties of programs solely from their
+types, i.e., with no knowledge whatsoever of the text of the programs
+involved.  Most of Wadler's free theorems are consequences of
+naturality for polymorphic list-processing functions. However,
+parametricity can also derive results that go beyond just naturality,
+such as inhabitation results, and prove the equivalence of Church
+encodings and fixpoint representations of ADTs and
 %hereditarily polynomial
@@
 different. If, as intended, the set interpretation of
-$\mathsf{Lan_k\,f}$ is $\mathit{Lan}_K\,F$, where $K$ is the set
-interpretation of $\mathsf{k}$ and $F$ is the set interpretation of
-$\mathsf{f}$, then the exact same reasoning gives that the relational
-interpretation of $\mathsf{Lan_k\,f}$ is $\mathit{Lan}_K\,F$, where
-$K$ is the relational interpretation of $\mathsf{k}$ and $F$ is the
-relational interpretation of $\mathsf{f}$. But under these
+$\mathsf{Lan_K\,F}$ is $\mathit{Lan}_K\,F$, where $K$ is the set
+interpretation of $\mathsf{K}$ and $F$ is the set interpretation of
+$\mathsf{F}$, then the exact same reasoning gives that the relational
+interpretation of $\mathsf{Lan_K\,F}$ is $\mathit{Lan}_K\,F$, where
+$K$ is the relational interpretation of $\mathsf{K}$ and $F$ is the
+relational interpretation of $\mathsf{F}$. But under these
 interpretations there can be no parametric model. The following
@@
 %related by $R$ in order to be a well-defined morphism of relations.
-$(m_1\,c, m_2\,d)$ must be in $R$ whenever $(c,d)$ is in $S$ in order
-to be a well-defined morphism of relations.  The domain of $G_1\, R$
-thus depends on $A$ {\em and} $B$, rather than simply on $A$.
+$(m_1\,c, m_2\,d)$ must be in $R$ whenever $(c,d)$ is in $S$ for
+$(m_1,m_2)$ to be a well-defined morphism of relations.  The domain of
+$G_1\, R$ thus depends on both $A$ and $B$, rather than simply on $A$.
 Likewise, the codomain of $G_1\,R$ also depends on both $A$ and $B$.
@@
 $G_0\, B$, respectively.  This suggests that GADTs might fail to have
-relational interpretation, and might thus fail to have a parametric
-model, as described above.
+relational interpretation, and thus might fail to have a parametric
+model, as described in the previous paragraph.
 %But, as we now show, this is not generally possible when
@@
 by~\eqref{eq:G}. The set interpretation of $\mathsf{G}$ is
-$\mathit{Lan}_{\lambda *\!. 1}\,\lambda *\!. 1$, i.e., is, by the
+$\mathit{Lan}_{\lambda u\!. 1}\,\lambda u\!. 1$, i.e., is, by the
 reasoning of Example~\ref{ex:g1}, the identity functor on
@@
 interpretation of $\mathsf{G}$ is $\mathit{Lan}_{\lambda
-  *\!.\,\mathit{Eq}_1} \,\lambda *\!.\,\mathit{Eq}_1$, where $\lambda
-*\!.\,\mathit{Eq}_1$ is the constantly $\mathit{Eq}_1$-valued functor
-from the category $\mathit{Rel}\,^0$ with a single object (also called
-$*$) to $\mathit{Rel}$.  Indeed, the interpretation is still a left
-Kan extension, but now it is the left Kan extension determined by the
-functor interpreting the type $\mathsf{\lambda *\!. 1}$ in
+  u\!.\,\mathit{Eq}_1} \,\lambda u\!.\,\mathit{Eq}_1$, where $\lambda
+u\!.\,\mathit{Eq}_1$ is the constantly $\mathit{Eq}_1$-valued functor
+from the category $\mathit{Rel}\,^0$ with a single object to
+$\mathit{Rel}$.  Indeed, the interpretation is still a left Kan
+extension, but now it is the left Kan extension determined by the
+functor interpreting the type $\mathsf{\lambda u\!. 1}$ in
 $\mathit{Rel}$. For the Identity Extension Lemma to hold we would need
 that, for every relation $R : \mathit{Rel}\,(A, B)$, we have that
-$(\mathit{\Lan}_{\lambda *\!.\,\mathit{Eq}_1} \,\lambda
-*\!.\,\mathit{Eq}_1)\, R$ is a relation between the sets
-$(\mathit{Lan}_{\lambda *. 1}\,\lambda *\!. 1)\, A$ and
-$(\mathit{Lan}_{\lambda *. 1}\,\lambda *\!. 1)\,B$, i.e., between the
+$(\mathit{\Lan}_{\lambda u\!.\,\mathit{Eq}_1} \,\lambda
+u\!.\,\mathit{Eq}_1)\, R$ is a relation between the sets
+$(\mathit{Lan}_{\lambda u. 1}\,\lambda u\!. 1)\, A$ and
+$(\mathit{Lan}_{\lambda u. 1}\,\lambda u\!. 1)\,B$, i.e., between the
 sets $A$ and $B$. However,
@@
 relates the single element of $1$ to both elements of the two-element
-set $2$. We expect $(\mathit{Lan}_{\lambda *\!.\,\mathit{Eq}_1}
-\lambda *\!.\,\mathit{Eq}_1)\, R$ to be a relation with domain
+set $2$. We expect $(\mathit{Lan}_{\lambda u\!.\,\mathit{Eq}_1}
+\lambda u\!.\,\mathit{Eq}_1)\, R$ to be a relation with domain
 $1$. Since left Kan extensions preserve projections~\cite{rie16}, we
 can compute the domain as
-\[\pi_1\,\big( (\mathit{Lan}_{\lambda *\!.\,\mathit{Eq}_1} \lambda
-*\!.\,\mathit{Eq}_1) \, R \big)\;=\; (\mathit{Lan}_{\lambda
- *\!.\,\mathit{Eq}_1} \lambda *\!. \,1)\, R\] (Note that the left Kan
- extension of $\lambda *\!. \,1$ along $\lambda *\!.\,\mathit{Eq}_1$
+\[\pi_1\,\big( (\mathit{Lan}_{\lambda u\!.\,\mathit{Eq}_1} \lambda
+u\!.\,\mathit{Eq}_1) \, R \big)\;=\; (\mathit{Lan}_{\lambda
+u\!.\,\mathit{Eq}_1} \lambda u\!. \,1)\, R\] (Note that the left Kan
+ extension of $\lambda u\!. \,1$ along $\lambda u\!.\,\mathit{Eq}_1$
  is a functor from $\mathit{Rel}$ to $\mathit{Set}$.) By
@@
 \[ \big( \bigcup_{m\, :\, \mathit{Eq}_1 \to R} 1 \big) \; /\; \approx\]
-where $\approx$ is the smallest equivalence relation such that $(\ast,
-m) \approx (\ast, m')$ for $m, m' : \mathit{Eq}_1 \to R$ if there
-exists $f : \mathit{Eq}_1 \to \mathit{Eq}_1$ such that $f (\ast, \ast)
-= (\ast, \ast)$ and the following diagram commutes:
+Here, $*$ is the unique element of the singleton set $1$, and $\approx$
+is the smallest equivalence relation such that $(m,\ast) \approx
+(m',\ast)$ for $m, m' : \mathit{Eq}_1 \to R$ if there exists $f :
+\mathit{Eq}_1 \to \mathit{Eq}_1$ such that $f (\ast, \ast) = (\ast,
+\ast)$ and the following diagram commutes:
 \[
@@
 morphism does satisfy the condition $f (\ast, \ast) = (\ast, \ast)$,
-we have that $(\ast, m) \approx (\ast, m')$ if and only if $m =
+we have that $(m,\ast) \approx (m',\ast)$ if and only if $m =
 m'$. Thus
@@
 $\mathit{Set}$, respectively, and therefore $\pi_1\,\big(
-(\mathit{Lan}_{\lambda *\!.\,\mathit{Eq}_1} \lambda
-*\!.\,\mathit{Eq}_1) \, R \big)$ is not $1$, as would be needed for
+(\mathit{Lan}_{\lambda u\!.\,\mathit{Eq}_1} \lambda
+u\!.\,\mathit{Eq}_1) \, R \big)$ is not $1$, as would be needed for
 the Identity Extension Lemma to hold.  Since the Identity Extension
@@
 We have shown that GADTs can be considered as data types in two
-contrasting ways: as completely determined by their syntax, in which
-case they are represented as their Church encodings, or as the
-functorial completion of their syntax, in which case they are
-represented as fixpoints.  But regardless of which representation we
-choose, GADTs fail to have {\em all} expected parametricity
-properties. Indeed, we can read the results of Section~\ref{sec:par}
-as showing that {\em i}) inhabitation and other non-naturality results
-for calculi representing GADTs as Church encodings cannot be extended
-to calculi representing them as fixpoints, and {\em ii}) naturality
-results for calculi representing GADTs as fixpoints cannot be extended
-to calculi representing them as Church encodings. So, neither
-representation guarantees {\em both} naturality consequences {\em and}
-non-naturality consequences of parametricity. Intuitively, this is
-because a GADT is not a data type in the usual container-ish sense,
-and completing it so that it becomes one destroys its relationally
-uniform --- i.e., parametric --- behavior.  Which representation of
-GADTs a user chooses can therefore have significant consequences for
-programming with, and reasoning about, them.
+contrasting ways: as their Church encodings, in which case they are
+completely determined by their syntax, or as fixpoints, in which case
+they are regarded as the functorial completion of their syntax. But
+regardless of which representation we choose, GADTs fail to have {\em
+  all} expected parametricity properties.  Indeed, we can read the
+results of Section~\ref{sec:par} as showing that {\em i}) inhabitation
+and other non-naturality results for calculi representing GADTs as
+Church encodings cannot be extended to calculi representing them as
+fixpoints, and {\em ii}) naturality results for calculi representing
+GADTs as fixpoints cannot be extended to calculi representing them as
+Church encodings. So, neither representation guarantees {\em both}
+naturality consequences {\em and} non-naturality consequences of
+parametricity. Intuitively, this is because a GADT is not a data type
+in the usual container-ish sense, and completing it so that it becomes
+one destroys its relationally uniform --- i.e., parametric ---
+behavior.  Which representation of data types a user chooses can
+therefore have significant consequences for programming with, and
+reasoning about, GADTs.
 
@@
 encode GADTs, but this model still does not guarantee functoriality;
-significantly, the parametric properties of GADTs described in the
+accordingly, the parametric properties of GADTs described in the
 precursor work~\cite{vw06} to~\cite{vw10} are all inhabitation results
@@
 but, as with the models mentioned above, this model also does not
-guarantee that data types have functorial semantics.
+guarantee that GADTs have functorial semantics.
 

### 2021-05-07 induction-paper/ENTCS-GADT-induction.tex
gadt-induction ready for comments

@@
 and almost all non-trivial proofs of their properties are either
-proved by induction outright or rely on lemmas that are.  {\color{red}
-  Every time a new algebraic data type (ADT) is declared in
-  Coq}\footnote{{\color{red} Algebraic data types are called inductive
-    types in Coq.}}, an induction rule is automatically generated for
-it.\looseness=-1
-
-The data types handled by Coq are (possibly mutually inductive)
-polynomial ADTs, and the induction rules
-Coq generates for them are the expected ones for standard structural
-induction. It has long been understood, however, that these rules are
-too weak to be genuinely useful for so-called {\em deep
-  ADTs}~\cite{jp20}, i.e., ADTs that are (possibly mutually
-inductively) defined in terms of (other) such ADTs.
+proved by induction outright or rely on lemmas that are.  Every time a
+new inductive type is declared in Coq an induction rule is
+automatically generated for it.\looseness=-1
+
+The inductive data types handled by Coq include (possibly mutually
+inductive) polynomial algebraic data types (ADTs), and the induction
+rules Coq generates for them are the expected ones for standard
+structural induction. It has long been understood, however, that these
+rules are too weak to be genuinely useful for so-called deep
+ADTs~\cite{jp20}, i.e., ADTs that are (possibly mutually inductively)
+defined in terms of (other) such ADTs.
 %\footnote{Such data
@@
 Consider, for example, the following type of rose trees, here coded in
-{\color{red} Agda} and defined in terms of the standard type
-$\mathsf{List}$ of lists (see Section~\ref{sec:ADTs-and-nesteds}):
+Agda and defined in terms of the standard type $\mathsf{List}$ of
+lists (see Section~\ref{sec:ADTs-and-nesteds}):
 
@@
 \noindent
-The induction rule Coq automatically generates for rose trees is
+The induction rule Coq automatically generates for (the analogous Coq
+defintion of) rose trees is
 
@@
 \begin{array}{l}
-\mathsf{data\, Seq\,(A : Set) : Set\,where}\\
+\mathsf{data\, Seq\,: Set \to Set\,where}\\
 \mathsf{\;\;\;\;\;\;\;\;const :\, A \to Seq\,A}\\
@@
 of sequences only constructs sequences of pairs, rather than sequences
-of arbitrary type, as does $\mathsf{const}$. If all of the
-constructors for a GADT $\mathsf{G}$ return structured instances of
-$\mathsf{G}$, then some of $\mathsf{G}$'s instances might not be
-inhabited. GADTs therefore have two distinct, but equally natural,
-semantics: a functorial semantics interpreting them as left Kan
-extensions~\cite{mac71}, and a parametric semantics interpreting them
-as their Church encodings~\cite{atk12,vw10}. As detailed
-in~\cite{jgj21}, a key difference in the two semantics is that the
-former views GADTs as their {\em functorial completions}~\cite{jp19},
-and thus as containing more data than just those expressible in
-syntax. By contrast, the latter views them as what might be called
-{\em syntax-only} GADTs. Happily, these two views of GADTs coincide
-for those that are ADTs or other nested types. However, both they and
-their attendant properties differ greatly for proper GADTs. In fact,
-the views deriving from the functorial and parametric semantics for
-proper GADTs are sufficiently distinct that, by contrast with the
-situation for ADTs and other nested
-types~\cite{bfss90,gjfor15,jgj21f}, it is not actually possible to
-define a functorial parametric semantics for them~\cite{jgj21}.
+of arbitrary type, as does $\mathsf{const}$.\footnote{The type of
+  $\mathsf{Seq}$ is actually $\mathsf{Set \to \Set_1}$, but to aid
+  readability we elide the explicit tracking of universe levels in
+  this paper.} If all of the constructors for a GADT $\mathsf{G}$
+return structured instances of $\mathsf{G}$, then some of
+$\mathsf{G}$'s instances might not be inhabited. GADTs therefore have
+two distinct, but equally natural, semantics: a functorial semantics
+interpreting them as left Kan extensions~\cite{mac71}, and a
+parametric semantics interpreting them as their Church
+encodings~\cite{atk12,vw10}. As detailed in~\cite{jgj21}, a key
+difference in the two semantics is that the former views GADTs as
+their {\em functorial completions}~\cite{jp19}, and thus as containing
+more data than just those expressible in syntax. By contrast, the
+latter views them as what might be called {\em syntax-only}
+GADTs. Happily, these two views of GADTs coincide for those that are
+ADTs or other nested types. However, both they and their attendant
+properties differ greatly for proper GADTs. In fact, the views
+deriving from the functorial and parametric semantics for proper GADTs
+are sufficiently distinct that, by contrast with the situation for
+ADTs and other nested types~\cite{bfss90,gjfor15,jgj21f}, it is not
+actually possible to define a functorial parametric semantics for
+them~\cite{jgj21}.
 
@@
 
-\pagebreak
-
 \begin{equation*}\label{eq:ptree}
@@
 
-\vspace*{-0.1in}
+\vspace*{-0.075in}
 
@@
 below themselves. The quintessential example of such a type is that of
-bushes~\cite{bm98}:
+bushes\footnote{To define truly nested types in Agda we must use the
+  $\mathsf{NO\_POSITIVITY\_CHECK}$ flag. A similar flag is required in
+  Coq.}\cite{bm98}:
+
+\vspace*{-0.1in}
+
 \begin{equation*}\label{eq:bush}
@@
 
+\vspace*{0.025in}
+
 \noindent
-{\color{red} Flags to overcome positivity restrictions?}
 Even defining a structural induction rule for bushes requires that we
@@
 for bushes is
+
+\vspace*{-0.075in}
+
 \[\begin{array}{l}
@@
 \end{array}\]
+
+\vspace*{-0.05in}
+
+\noindent
 Here, $\mathsf{Bush^{\wedge} : \forall (A : Set) \to (A \to Set) \to
@@
 element of type $\mathsf{A}$ contained in its argument bush:
+
+\vspace*{-0.15in}
+
 \begin{align*}
@@
 \end{align*}
+
+\vspace*{-0.05in}
 
@@
   Set) \to Seq\,A \to Set}$ by
+
+\vspace*{-0.2in}
+
 \[\begin{array}{lll}
@@
 hypothesis
+
+\pagebreak
+
 \[\begin{array}{l}
@@
 \mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clist \, A\, Q_A\,
-    (app \,B \,\,t_{BA} \, t_B)\, (Q_B , list_{t_{BA}}, list_{t_B})} &
+    (app \,B \,\,t_{BA} \, t_B)\, (Q_B , lift_{t_{BA}}, lift_{t_B})} &
 = & \mathsf{capp\,A\,B\,Q_A\, Q_B\, t_{BA}\, t_B\, p_{BA} \, p_B}\\
-  \mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clistc \, A\, Q_A\,
+  \mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clist \, A\, Q_A\,
     (list \,B \,e \, ts) \, (Q_B , liftE', lift_{List})} & = & 
-  \mathsf{clistc \,A\,B\,Q_A\, Q_B\, e\, ts\, liftE'\, p_{List} }
+  \mathsf{clist \,A\,B\,Q_A\, Q_B\, e\, ts\, liftE'\, p_{List} }
 \end{array}\]}
@@
 
-\pagebreak
-
 \noindent
@@
 discussion. A further requirement that applies to all of the type
-constructors appearing in the right-hand sides of~\eqref{eq:gadts},
+constructors appearing in the right-hand side of the above grammar
 including those in $\ol{\mathsf{K}}$, is that they must all admit
@@
 type application admit predicate liftings, and so do GADTs constructed
-from the above grammar; in fact, we have seen such liftings for
-products and type application in Section~\ref{sec:deep-ind-GADTs}. A
-concrete way to define more general predicate liftings is, again, by
-induction on the structure of the types. We do not give a general
-definition of predicate liftings here, though, since that would
-require us to first design a full type calculus, which is beyond the
-scope of the present paper. {\color{red} Awkward paragraph? Rephrase?}
+from the grammar; in fact, we have seen such liftings for products and
+type application in Section~\ref{sec:deep-ind-GADTs}. A concrete way
+to define more general predicate liftings is, again, by induction on
+the structure of the types. We do not give a general definition of
+predicate liftings here, though, since that would require us to first
+design a full type calculus, which is beyond the scope of the present
+paper.
 
@@
 nested GADT because it is not possible to define a map function
-$\mathsf{Equal^\wedge Map}$ for $\mathsf{Eq^\wedge}$ when one of the
+$\mathsf{Equal^\wedge Map}$
+%for $\mathsf{Equal^\wedge}$
+when one of the
 type arguments to $\mathsf{Equal^\wedge}$ is structured (i.e., not a
@@
 \[ \mathsf{getTypeProof \,A \,t = 
-    dIndLTerm\, P \, cvar\, cabs\, capp\, clistc\, A\, K_\top\, t\,
+    dIndLTerm\, P \, cvar\, cabs\, capp\, clist\, A\, K_\top\, t\,
     (LTerm^\wedge KT\, A\, t) }\] where $\mathsf{t : LTerm\,A}$,
@@
 final argument is $\mathsf{nothing}$ and $\mathsf{cabs\, A \,B \,C
-  \,Q_A\, Q_B \,Q_C\, e \,s \,T_B\, t_C\, Eq^\wedge\,
+  \,Q_A\, Q_B \,Q_C\, e \,s \,T_B\, t_C\, liftE\;
 %  LLTy^\wedge B\,
-L\, (just \,T_C) = just \, (arr\, B\, C\, e\, T_B \,T_C)}$ otherwise.
+lift_{T_B}\; (just \,T_C) = just \, (arr\, B\, C\, e\, T_B \,T_C)}$ otherwise.
 In the third clause, $\mathsf{capp\, A \,B \, Q_A \, Q_B \, t_{BA}\,
@@
   B\, K_\top) \, ts }$.  As in that case we assume a proof
-  $\mathsf{EqualLListKT : Equal^{\wedge} \, A\, (List\,B)\, K_\top\,
+  $\mathsf{Equal^\wedge ListKT : Equal^{\wedge} \, A\, (List\,B)\, K_\top\,
   (List^{\wedge} \, B\, K_\top) \, e}$ for the first component.  We

### 2021-05-10 induction-paper/ENTCS-GADT-induction.tex
almost-final reading?

@@
 \begin{abstract} 
-  Deep data types those that are defined in terms of other data types,
+  Deep data types are those that are defined in terms of other data types,
   including, possibly, themselves.  In that case, they are said to be
@@
 
-%\begin{keyword}
-%  Please list keywords from your paper here, separated by commas.
-%\end{keyword}
-
 \end{frontmatter}
@@
 \section{Introduction}\label{sec:intro}
-
-%\vspace*{-0.09in}
 
@@
 structural induction. It has long been understood, however, that these
-rules are too weak to be genuinely useful for so-called deep
-ADTs~\cite{jp20}, i.e., ADTs that are (possibly mutually inductively)
-defined in terms of (other) such ADTs.
+rules are too weak to be genuinely useful for deep ADTs~\cite{jp20},
+i.e., ADTs that are (possibly mutually inductively) defined in terms
+of (other) such ADTs.\footnote{More generally, a {\em deep data type}
+  is one that is (possibly mutually inductively) defined in terms of
+  other such data types.}
 %\footnote{Such data
@@
 
-\vspace*{-0.075in}
+\vspace*{-0.2in}
 
@@
   $\mathsf{Seq}$ is actually $\mathsf{Set \to \Set_1}$, but to aid
-  readability we elide the explicit tracking of universe levels in
-  this paper.} If all of the constructors for a GADT $\mathsf{G}$
-return structured instances of $\mathsf{G}$, then some of
-$\mathsf{G}$'s instances might not be inhabited. GADTs therefore have
-two distinct, but equally natural, semantics: a functorial semantics
-interpreting them as left Kan extensions~\cite{mac71}, and a
-parametric semantics interpreting them as their Church
-encodings~\cite{atk12,vw10}. As detailed in~\cite{jgj21}, a key
-difference in the two semantics is that the former views GADTs as
-their {\em functorial completions}~\cite{jp19}, and thus as containing
-more data than just those expressible in syntax. By contrast, the
-latter views them as what might be called {\em syntax-only}
-GADTs. Happily, these two views of GADTs coincide for those that are
-ADTs or other nested types. However, both they and their attendant
-properties differ greatly for proper GADTs. In fact, the views
-deriving from the functorial and parametric semantics for proper GADTs
-are sufficiently distinct that, by contrast with the situation for
-ADTs and other nested types~\cite{bfss90,gjfor15,jgj21f}, it is not
-actually possible to define a functorial parametric semantics for
-them~\cite{jgj21}.
+  readability we elide the explicit tracking of Agda universe levels in
+  this paper.} If one or more of the constructors for a GADT
+$\mathsf{G}$ return structured instances of $\mathsf{G}$,
+%then some of
+%$\mathsf{G}$'s instances might not be inhabited.
+then the GADT will have two distinct, but equally natural, semantics:
+a functorial semantics interpreting them as left Kan
+extensions~\cite{mac71}, and a parametric semantics interpreting them
+as the interpretations of their Church encodings~\cite{atk12,vw10}. As
+detailed in~\cite{jgj21}, a key difference in the two semantics is
+that the former views GADTs as their {\em functorial
+  completions}~\cite{jp19}, and thus as containing more data than just
+those expressible in syntax. By contrast, the latter views them as
+what might be called {\em syntax-only} GADTs. Happily, these two views
+of GADTs coincide for those that are ADTs or other nested
+types. However, both they and their attendant properties differ
+greatly for proper GADTs. In fact, the views deriving from the
+functorial and parametric semantics for proper GADTs are sufficiently
+distinct that, by contrast with the situation for ADTs and other
+nested types~\cite{bfss90,gjfor15,jgj21f}, it is not actually possible
+to define a functorial parametric semantics for them~\cite{jgj21}.
 
@@
 functoriality of data types' interpretations from~\cite{jp19}, and
-functoriality is precisely what interpreting GADTs as their Church
-encodings fails to deliver. Our approach is to instead first give a
-predicate lifting styled after those of~\cite{jp20}, together with a
-(deep) induction rule, and for the simplest --- and arguably most
-important --- GADT, namely the equality GADT. (See
-Section~\ref{sec:ind-equal}.)  We then derive the deep induction rule
-a more complex GADT $\mathsf{G}$ by {\em i}) using the equality GADT
-to represent $\mathsf{G}$ as its so-called {\color{red} {\em Henry Ford
-  encoding}}~\cite{ch03,hin03,mcb99,sjsv09,sp04}, and {\em ii}) using
-the predicate liftings for the equality GADT, and any other GADTs
-appearing in the definition of $\mathsf{G}$, to appropriately thread
-the custom predicates for the primitive types appearing in
+functoriality is precisely what interpreting GADTs as the
+interpretations of their Church encodings fails to deliver. Our
+approach is to instead first give a predicate lifting styled after
+those of~\cite{jp20}, together with a (deep) induction rule, for the
+simplest --- and arguably most important --- GADT, namely the equality
+GADT. (See Section~\ref{sec:ind-equal}.)  We then derive the deep
+induction rule a more complex GADT $\mathsf{G}$ by {\em i}) using the
+equality GADT to represent $\mathsf{G}$ as its so-called {\em Henry
+  Ford encoding}~\cite{ch03,hin03,mcb99,sjsv09,sp04}, and {\em ii})
+using the predicate liftings for the equality GADT, and any other
+GADTs appearing in the definition of $\mathsf{G}$, to appropriately
+thread the custom predicates for the primitive types appearing in
 $\mathsf{G}$ throughout $\mathsf{G}$'s structure. This two-step
@@
 parametricity and true nesting still is not permitted.  As far as we
-know, no attempt has yet been made to extend either implementation
+know, no attempt has yet been made to extend either implementation to
 truly nested types or to proper GADTs. In fact, we know of no work
@@
 In this paper, we are interested in induction rules for proof-relevant
-predicates.  A proof-relevant predicate on a type $\mathsf{A : Set}$
-is a function $\mathsf{P\,:\,A \to Set}$ mapping each $\mathsf{a : A}$
-to the set of proofs that $\mathsf{P\,a}$ holds.  For example, the
+predicates.  A proof-relevant predicate on $\mathsf{A : Set}$ is a
+function $\mathsf{P\,:\,A \to Set}$ mapping each $\mathsf{a : A}$ to
+the set of proofs that $\mathsf{P\,a}$ holds.  For example, the
 structural induction rule for the list type
 
-\vspace*{-0.1in}
+\vspace*{-0.2in}
 
@@
 
-\vspace*{-0.2in}
+\vspace*{-0.3in}
 
@@
 \to P\,nil \to \big( \forall (a : A) (as: List\,A) \to Q\,a \to P\,as
-\to P\,(cons\,a\,as)\big)} \\ 
-\quad\mathsf{\to \forall (as : List\,A) \to List^{\wedge}\,A\,Q\,as
+\to P\,(cons\,a\,as)\big) \to} \\ 
+\quad\mathsf{\forall (as : List\,A) \to List^{\wedge}\,A\,Q\,as
   \to P\,as } 
@@
 \mathsf{\forall (P : \forall (A : Set) \to PTree\, A \to Set)
-\to \big( \forall (A : Set) (a : A) \to P\,A\,(pleaf\, a) \big)} \\
-\quad\mathsf{\to \big( \forall (A : Set) (pp : PTree\,(A \times A))
-  \to P\,(A \times A)\,pp \to P\,a\,(pnode\,pp)\big) \to \forall (A :
+\to \big( \forall (A : Set) (a : A) \to P\,A\,(pleaf\, a) \big) \to} \\
+\quad\mathsf{\big( \forall (A : Set) (pp : PTree\,(A \times A)) \to
+  P\,(A \times A)\,pp \to P\,a\,(pnode\,pp)\big) \to \forall (A :
   Set) (p : PTree\,A) \to P\,A\,p } 
@@
   \to Set) \to \big( \forall (A : Set) (Q : A \to Set) (a : A) \to
-  Q\,a \to P\,A\,Q\,(Pleaf\, a) \big)} \\ \quad \mathsf{\to \big(
+  Q\,a \to P\,A\,Q\,(Pleaf\, a) \big) \to} \\ \quad \mathsf{ \big(
   \forall (A : Set) (Q : A \to Set) (pp : PTree\,(A \times A)) \to
   P\,(A \times A)\,(Pair^{\wedge}\,A\,A\,Q\,Q)\,pp \to
-  P\,A\,Q\,(Pnode\,pp)\big)} \\ \quad \mathsf{\to \forall (A : Set) (Q
+  P\,A\,Q\,(Pnode\,pp) \big) \to} \\ \quad \mathsf{\forall (A : Set) (Q
   : A \to Set) (p : PTree\,A) \to PTree^{\wedge}\,A\,Q\,p \to
@@
 \mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to Bush\, A \to Set)
-\to \big( \forall (A : Set) \to P\,A\,bnil \big)} \\ 
-\quad\mathsf{\to \big( \forall (A : Set) (Q : A \to Set) (a : A) (bb :
+\to \big( \forall (A : Set) \to P\,A\,bnil \big) \to} \\ 
+\quad\mathsf{\big( \forall (A : Set) (Q : A \to Set) (a : A) (bb :
   Bush\,(Bush\,A)) \to Q\,a \to
   P\,(Bush\,A)\,(Bush^{\wedge}\,A\,Q)\,bb \to P\,A\,Q\,(bcons\,a\,bb)
-  \big)} \\ 
-\quad\mathsf{\to \forall (A : Set) (Q : A \to Set) (b : Bush\,A) \to
+  \big) \to} \\ 
+\quad\mathsf{\forall (A : Set) (Q : A \to Set) (b : Bush\,A) \to
   Bush^{\wedge}\,A\,Q\,b \to P\,A\,Q\,b } 
@@
 The equality constraints engendered by GADTs' data constructors makes
-deriving (deep) induction rules for then more involved than for ADTs
+deriving (deep) induction rules for them more involved than for ADTs
 and other nested types. Nevertheless, we show in this section how to
@@
 To define the (deep) induction rule for any (deep) GADT $\mathsf{G}$
-we first need to define a predicate lifting that maps a predicate on a
-type $\mathsf{A}$ to a predicate on $\mathsf{G\,A}$. Such a predicate
+we first need to define a predicate lifting that maps a predicate on
+a type $\mathsf{A}$ to a predicate on $\mathsf{G\,A}$. Such a predicate
 lifting $\mathsf{Equal^{\wedge} : \forall (A\,B : Set) \to (A \to Set)
@@
 \mathsf{\lambda (P : \forall (A\,B : Set) \to (A \to Set) \to (B \to
-  Set) \to Equal\,A\,B \to Set)} \\ 
-\quad\mathsf{\to \forall (C : Set) (Q\, Q' : C \to Set) \to
+  Set) \to Equal\,A\,B \to Set) \to} \\ 
+\quad\mathsf{\forall (C : Set) (Q\, Q' : C \to Set) \to
   Equal^{\wedge}\,C\,C\,Q\,Q'\,refl \to P\,C\,C\,Q\,Q'\,refl} 
@@
   \mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to
-  LTerm\,A \to Set)} \\ 
-  \quad\mathsf{\to 
+  LTerm\,A \to Set) \to} \\ 
+  \quad\mathsf{
   \forall (A\,B\,C: Set) (Q_A : A \to Set)  (Q_B : B \to Set) (Q_C : C
-  \to Set) (e : Equal\, A\, (B \to C)) (s : String) } \\ 
-  \quad\mathsf{ \to (T_B : LType\, B) \to (t_C : LTerm\, C)
+  \to Set) (e : Equal\, A\, (B \to C)) (s : String) \to } \\ 
+  \quad\mathsf{(T_B : LType\, B) \to (t_C : LTerm\, C)
   \to Equal^{\wedge}\,A\,(B \to C)\, Q_A \, (Arr^{\wedge} \, B\, C\,
-  Q_B \, Q_C) \, e } \\
+  Q_B \, Q_C) \, e \to  } \\
   \quad\mathsf{
-  \to LType^{\wedge}\, B\, Q_B\, T_B
+  LType^{\wedge}\, B\, Q_B\, T_B
   \to P\, C\, Q_C\, t_C\, 
@@
   \mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A
-    \to Set)} \\ 
-  \quad \mathsf{\to 
+    \to Set)\to} \\ 
+  \quad \mathsf{
   \forall (A \,B : Set) (Q_A : A \to Set)  (Q_B : B \to Set) 
-   (t_{BA} : LTerm\, (B \to A)) (t_B : LTerm\, B)} \\
+   (t_{BA} : LTerm\, (B \to A)) (t_B : LTerm\, B) \to} \\
   \quad \mathsf{
-  \to P\, (B \to A)\, (Arr^{\wedge} \, B\, A\, Q_B \, Q_A) \, t_{BA} \, 
+  P\, (B \to A)\, (Arr^{\wedge} \, B\, A\, Q_B \, Q_A) \, t_{BA} \, 
   \to P\, B\, Q_B\, t_B\, 
@@
   \mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A
-    \to Set)} \\ 
-  \quad \mathsf{\to 
+    \to Set) \to} \\ 
+  \quad \mathsf{
   \forall (A \,B : Set) (Q_A : A \to Set)  (Q_B : B \to Set) 
-    (e : Equal\, A\, (List\, B)) (ts : List\, (LTerm\, B))} \\ 
+    (e : Equal\, A\, (List\, B)) (ts : List\, (LTerm\, B)) \to} \\ 
   \quad \mathsf{
-    \to Equal^{\wedge}\, A\, (List\,B)\, Q_A\, (List^{\wedge}\, B\, Q_B)\, e 
+    Equal^{\wedge}\, A\, (List\,B)\, Q_A\, (List^{\wedge}\, B\, Q_B)\, e 
   \to List^{\wedge}\, (LTerm\,B) (P\, B\, Q_B)\, ts
@@
 \mathsf{dIndC = \lambda (P : \forall (A : Set) \to (A \to Set) \to
-  G\,A \to Set)}\\
-\quad \mathsf{\to \forall (A : Set)\, (\ol{B : Set})\, (Q_A : A \to Set)\,
+  G\,A \to Set) \to}\\
+\quad \mathsf{\forall (A : Set)\, (\ol{B : Set})\, (Q_A : A \to Set)\,
   (\ol{Q_B : B \to Set})\, (e : Equal\,A\,(K\,\ol{B}))\, (x :
-  F\,G\,\ol{B})} \\ 
-\quad \mathsf{\to
+  F\,G\,\ol{B}) \to} \\ 
+\quad \mathsf{
   Equal^{\wedge}\,A\,(K\,\ol{B})\,Q_A\,(K^{\wedge}\,\ol{B}\,\ol{Q_B})\,e
@@
 semantics at the same time~\cite{jgj21}. In this section we show how
-deep induction for truly nested GADTs nesting goes wrong by analyzing
-the following very simple nested proper GADT:\looseness=-1
+deep induction for truly nested GADTs goes wrong by analyzing the
+following very simple nested proper GADT:\looseness=-1
 
@@
 \mathsf{\lambda\, (P : \forall\, (A : Set) \to (A \to Set) \to G\,A
-  \to Set)} \\ 
-\quad\mathsf{\to \forall\, (A\;B : Set)\, (Q_A : A \to Set)\, (Q_B : B
-  \to Set)\, (e : Equal\,A\,(B \times B))\, (x : G\,(G\,B))} \\ 
-\quad\mathsf{\to Equal^{\wedge}\,A\,(B \times
+  \to Set)\to} \\ 
+\quad\mathsf{ \forall\, (A\;B : Set)\, (Q_A : A \to Set)\, (Q_B : B
+  \to Set)\, (e : Equal\,A\,(B \times B))\, (x : G\,(G\,B))\to} \\ 
+\quad\mathsf{Equal^{\wedge}\,A\,(B \times
   B)\,Q_A\,(Pair^{\wedge}\,B\,B\,Q_B\,Q_B)\,e \to P\,(G\,B)\,(P\,B\,Q_B)\,x 
@@
 from~\eqref{eq:eq_lterm} to extract the type from a lambda term.  The
-following predicate takes a lambda term as input and either returns
-its type if it is well-typed or indicates that it is not:
+following predicate either returns the type of its input lambda term
+if that type can be inferred or indicates that the type inference
+fails:
 
@@
 The predicate $\mathsf{GetType}$ uses the standard $\mathsf{Maybe}$
-data type to represent potential ill-typedness. It is defined by:
+data type to represent failure of type inference. It is defined by:\looseness=-1
 
@@
 & &    \quad\mathsf{(e : Equal\, A\, (B \to C))\, (s : String)\, (T_B
-   : LType\, B)\, (t_C : LTerm \, C)} \\  
-& & \quad\quad\mathsf{\to Equal^{\wedge}\, A\, (B \to C)\, Q_A\,
+   : LType\, B)\, (t_C : LTerm \, C)  \to} \\  
+& & \quad\mathsf{Equal^{\wedge}\, A\, (B \to C)\, Q_A\,
    (Arr^{\wedge}\, B\, C\, Q_B\, Q_C)\, e  \to LType^{\wedge} \, B\,
@@
   Set)\, (Q_B : B \to Set)\,  (t_{BA}: LTerm\, (B \to A))\, (t_B :
-  LTerm\, B)} \\  
-& & \quad\mathsf{\to Maybe\, (LType\, (B \to A))\to Maybe\, (LType\,
+  LTerm\, B)\to} \\  
+& & \quad\mathsf{ Maybe\, (LType\, (B \to A))\to Maybe\, (LType\,
       B) \to Maybe\, (LType\, A)}\\
@@
   Set)\, (Q_B : B \to Set)\, (e : Equal\, A\, (List\, B))\, (ts : 
-  List\, (LTerm\, B))}\\  
-& & \quad\mathsf{\to Equal^{\wedge}\, A\, (List\,B) \, Q_A\,
+  List\, (LTerm\, B)) \to}\\  
+& & \quad\mathsf{Equal^{\wedge}\, A\, (List\,B) \, Q_A\,
   (List^{\wedge}\, B\, Q_B)\, e \to List^{\wedge}\, (LTerm\, B)\,
@@
 $\mathsf{LType^\wedge KT\, B\, T}$ and $\mathsf{LTerm^\wedge KT\, C\,
-  t')}$ for the latter two proofs. For the former we note that, since
+  t'}$ for the latter two proofs. For the former we note that, since
 we are working with proof-relevant predicates, the lifting

### 2021-05-17 induction-paper/ENTCS-GADT-induction.tex
final MFPS

@@
 \begin{abstract} 
-  Deep data types are those that are defined in terms of other data types,
-  including, possibly, themselves.  In that case, they are said to be
-  truly nested.  Deep induction is an extension of structural
-  induction that traverses {\em all} of the structure in a deep data
-  type, propagating predicates on its primitive data throughout the
-  entire structure.  Deep induction was recently shown to prove
+  Deep data types are those that are defined in terms of other such
+  data types, including, possibly, themselves.  In that case, they are
+  said to be truly nested.  Deep induction is an extension of
+  structural induction that traverses {\em all} of the structure in a
+  deep data type, propagating predicates on its primitive data
+  throughout the entire structure.  Deep induction can be used to prove
   properties of nested types, including truly nested types, that
@@
 i.e., ADTs that are (possibly mutually inductively) defined in terms
-of (other) such ADTs.\footnote{More generally, a {\em deep data type}
-  is one that is (possibly mutually inductively) defined in terms of
-  other such data types.}
-%\footnote{Such data
-%  types are called nested inductive types by Chlipala~\cite{chl},
-%  reflecting the fact that ``inductive type'' means ``ADT''~in~Coq.}
-Consider, for example, the following type of rose trees, here coded in
-Agda and defined in terms of the standard type $\mathsf{List}$ of
-lists (see Section~\ref{sec:ADTs-and-nesteds}):
+of (other) such ADTs.\footnote{A data type is {\em deep} if it is
+  (possibly mutually inductively) defined in terms of other such data
+  types (including, possibly, itself).}  The following data type of
+rose trees, here coded in Agda and defined in terms of the standard
+type $\mathsf{List}$ of lists (see
+Section~\ref{sec:ADTs-and-nesteds}), is a deep ADT:
 
@@
 \mathsf{data\, Rose\, : Set \to Set\,where}\\
-\mathsf{\;\;\;\;\;\;\;\;empty\, :\, Rose\,A}\\
-\mathsf{\;\;\;\;\;\;\;\;\,node\,\,\,\, :\, A \to List\,(Rose\,A) \to Rose\,A} 
+\mathsf{\;\;\;\;\;\;\;\;empty\, :\,\forall \{A : Set\} \to Rose\,A}\\
+\mathsf{\;\;\;\;\;\;\;\;\,node\,\,\,\, :\, \forall \{A : Set\} \to A \to List\,(Rose\,A) \to Rose\,A} 
 \end{array}\]
@@
 for data types that go well beyond traditional ADTs. Such data types
-include nested types~\cite{bm98}, such as $\mathsf{Rose}$ above, as
-well as truly nested types\footnote{A truly nested type is a nested
-  type that is defined over itself. The data type $\mathsf{Bush}$ in
-  Section~\ref{sec:ADTs-and-nesteds} provides a concrete example.},
+include not just deep ADTs, but also (truly\footnote{A truly nested
+  type is a nested type that is defined over itself. The data type
+  $\mathsf{Bush}$ in Section~\ref{sec:ADTs-and-nesteds} provides a
+  concrete example.}) nested types~\cite{bm98},
+%, such as $\mathsf{Rose}$ above, as well as truly nested types
 generalized algebraic data types
@@
 
-Deep induction was shown in~\cite{jp20} to be the form of induction
-most appropriate to nested types (including ADTs) that are defined
-over, or mutually recursively with, other such types (including,
-possibly, themselves). Deep induction delivers the following genuinely
-useful induction rule for rose trees:
+Deep induction was shown in~\cite{jp20} to deliver induction rules
+appropriate to deep nested types, including deep ADTs.  For the deep
+ADT of rose trees, for example, it gives the following genuinely
+useful induction rule:
 
@@
 \mathsf{\hspace*{0.15in}(\forall \,(a : A)\, (ts :
-  List\,(Rose\,A))\to Q\,a \to List^\land\,P\,ts \to P\,(node\,a\,ts)) \to}\\
+  List\,(Rose\,A))\to Q\,a \to List^\land\,(Rose\,A)\, P\,ts \to P\,(node\,a\,ts)) \to}\\
 \mathsf{\hspace*{0.15in}\forall \,(x :
-  Rose\,A) \to Rose^\land\,Q\, x \to P\,x} 
+  Rose\,A) \to Rose^\land\,A\,Q\, x \to P\,x} 
 \end{array}
@@
 This paper shows how to extend deep induction to proper GADTs, i.e.,
-to GADTs that are not simply (truly) nested types (and thus are not
-ADTs).  A constructor for such a GADT $\mathsf{G}$ may, like a
-constructor for a nested type, take as arguments data whose types
-involve instances of $\mathsf{G}$ other than the one being defined ---
-including instances that involve $\mathsf{G}$ itself. But if
-$\mathsf{G}$ is a proper GADT then at least one of its constructors
-will also have such a structured instance of $\mathsf{G}$ --- albeit
-one not involving $\mathsf{G}$ itself --- as its codomain.
+to GADTs that are not simply nested types, and thus are not ADTs.  A
+constructor for such a GADT $\mathsf{G}$ may, like a constructor for a
+nested type, take as arguments data whose types involve instances of
+$\mathsf{G}$ other than the one being defined --- including instances
+that involve $\mathsf{G}$ itself. But if $\mathsf{G}$ is a proper GADT
+then at least one of its constructors will also have such a structured
+instance of $\mathsf{G}$ --- albeit one not involving $\mathsf{G}$
+itself --- as its codomain.
 %But while the types of the arguments to a GADT's constructor can
@@
 \mathsf{data\, Seq\,: Set \to Set\,where}\\
-\mathsf{\;\;\;\;\;\;\;\;const :\, A \to Seq\,A}\\
-\mathsf{\;\;\;\;\;\;\;\;pair\;\;\, :\, Seq \,A \to Seq\,B \to
+\mathsf{\;\;\;\;\;\;\;\;const :\, \forall \{A : Set\} \to A \to Seq\,A}\\
+\mathsf{\;\;\;\;\;\;\;\;pair\;\;\, :\,\forall \{A\,B : Set\} \to Seq \,A \to Seq\,B \to
   Seq\,(A \times B)}
@@
 \noindent
-of sequences only constructs sequences of pairs, rather than sequences
-of arbitrary type, as does $\mathsf{const}$.\footnote{The type of
-  $\mathsf{Seq}$ is actually $\mathsf{Set \to \Set_1}$, but to aid
-  readability we elide the explicit tracking of Agda universe levels in
-  this paper.} If one or more of the constructors for a GADT
-$\mathsf{G}$ return structured instances of $\mathsf{G}$,
-%then some of
-%$\mathsf{G}$'s instances might not be inhabited.
-then the GADT will have two distinct, but equally natural, semantics:
-a functorial semantics interpreting them as left Kan
-extensions~\cite{mac71}, and a parametric semantics interpreting them
-as the interpretations of their Church encodings~\cite{atk12,vw10}. As
+of sequences\footnote{The type of $\mathsf{Seq}$ is actually
+  $\mathsf{Set \to \Set_1}$, but to aid readability we elide the
+  explicit tracking of Agda universe levels in this paper.} only
+constructs sequences of pairs, rather than sequences of arbitrary
+type, as does $\mathsf{const}$. If one or more of the constructors for
+a GADT $\mathsf{G}$ return structured instances of $\mathsf{G}$, then
+the GADT will have two distinct, but equally natural, semantics: a
+functorial semantics interpreting it as a left Kan
+extension~\cite{mac71}, and a parametric semantics interpreting it as
+the interpretation of its Church encoding~\cite{atk12,vw10}. As
 detailed in~\cite{jgj21}, a key difference in the two semantics is
@@
 what might be called {\em syntax-only} GADTs. Happily, these two views
-of GADTs coincide for those that are ADTs or other nested
+of GADTs coincide for those GADTs that are ADTs or other nested
 types. However, both they and their attendant properties differ
-greatly for proper GADTs. In fact, the views deriving from the
-functorial and parametric semantics for proper GADTs are sufficiently
-distinct that, by contrast with the situation for ADTs and other
+greatly for proper GADTs. In fact, the functorial and parametric
+semantics for proper GADTs are sufficiently disparate that, by
+contrast with the semantics customarily given for ADTs and other
 nested types~\cite{bfss90,gjfor15,jgj21f}, it is not actually possible
@@
 determined by their Church encodings {\em does} support induction
-rules --- including deep induction rules --- for GADTs. Perhaps
-surprisingly, this paper gives the first-ever induction rules --- deep
-or otherwise --- for proper GADTs. But it actually delivers far more:
-it gives a general framework for deriving deep induction rules for
-deep GADTs directly from their syntax. This framework can serve as a
-basis for extending modern proof assistants' automatic generation of
-structural induction rules for ADTs to automatic generation of deep
-induction rules for GADTs. As for ADTs and other nested types, the
+rules --- including deep induction rules --- for GADTs.  Indeed, this
+paper gives the first-ever induction rules --- deep or otherwise ---
+for proper GADTs. But it actually delivers far more: it gives a
+general framework for deriving deep induction rules for deep GADTs
+directly from their syntax. This framework can serve as a basis for
+extending modern proof assistants' automatic generation of structural
+induction rules for ADTs to automatic generation of deep induction
+rules for GADTs. In addition, as for ADTs and other nested types, the
 structural induction rule for any GADT can be recovered from its deep
@@
 GADT. (See Section~\ref{sec:ind-equal}.)  We then derive the deep
-induction rule a more complex GADT $\mathsf{G}$ by {\em i}) using the
-equality GADT to represent $\mathsf{G}$ as its so-called {\em Henry
-  Ford encoding}~\cite{ch03,hin03,mcb99,sjsv09,sp04}, and {\em ii})
-using the predicate liftings for the equality GADT, and any other
-GADTs appearing in the definition of $\mathsf{G}$, to appropriately
-thread the custom predicates for the primitive types appearing in
-$\mathsf{G}$ throughout $\mathsf{G}$'s structure. This two-step
-process delivers deep induction rules for a very general class of deep
-GADTs. In Section~\ref{sec:GADTs} we introduce a series of
-increasingly complex GADTs as running examples, and in
+induction rule for a more complex GADT $\mathsf{G}$ by {\em i}) using
+the equality GADT to represent $\mathsf{G}$ as its so-called {\em
+  Henry Ford encoding}~\cite{ch03,hin03,mcb99,sjsv09,sp04}, and {\em
+  ii}) using the predicate liftings for the equality GADT, and any
+other GADTs appearing in the definition of $\mathsf{G}$, to
+appropriately thread the custom predicates for the primitive types
+appearing in $\mathsf{G}$ throughout $\mathsf{G}$'s structure. This
+two-step process delivers deep induction rules for a very general
+class of deep GADTs. In Section~\ref{sec:GADTs} we introduce a series
+of increasingly complex GADTs as running examples, and in
 Section~\ref{sec:deep-ind-GADTs} we derive a deep induction rule for
-each of them. In particular, we derive the deep induction rule for
-$\mathsf{Seq}$ in Section~\ref{sec:ind-seq}. We present our general
-framework for deriving (deep) induction rules for (deep) GADTs in
+each of them. In particular, we derive the deep induction rule for the
+$\mathsf{Seq}$ data type in~\eqref{eq:seq} in
+Section~\ref{sec:ind-seq}. We present our general framework for
+deriving (deep) induction rules for (deep) GADTs in
 Section~\ref{sec:framework}, and observe that the derivations in
@@
 not appear to be much of a restriction, however, since GADTs defined
-over themselves do not, to our knowledge, appear in applications or
-the literature. Section~\ref{sec:app} comprises a case study in using
-deep induction. All of the deep induction rules appearing in this
-paper have been derived by instantiating our general framework. Our
-Agda code implementing them is available at~\cite{web-page}.
+over themselves do not, to our knowledge, appear in practice or in the
+literature. Section~\ref{sec:app} comprises a case study in using deep
+induction. All of the deep induction rules appearing in this paper
+have been derived by instantiating our general framework. Our Agda
+code implementing them is available at~\cite{web-page}.
 
@@
 
-{\bf Related Work\/} Various techniques for deriving induction rules
-for data types that go beyond ADTs have been studied. For example, Fu
-and Selinger~\cite{fs18} show, via examples, how to derive induction
-rules for arbitrary nested types. Unfortunately, however, their
-technique is rather {\em ad hoc}, so is unclear how to generalize it
-to nested types other than the specific ones in their
-examples. Moreover,~\cite{fs18} actually derives induction rules for
-data types {\em related} to the original nested types rather than for
-the original nested types themselves, and it is unclear whether or not
-the derived rules are sufficiently expressive to prove all results
-about the original nested types that we would expect to be provable by
-induction. This latter point echoes the issue with Coq-derived
-induction rule for rose trees raised in Section~\ref{sec:intro}, which
-has the unfortunate effect of forcing users to manually write
-induction (and other) rules for such types for use in that
-system. Tassi~\cite{tas19} derives induction rules for data type
-definitions in Coq using unary parametricity. His technique seems to
-be essentially equivalent to that of~\cite{jp19} for nested types,
-although he does not permit true nesting. More recently,
+{\bf Additional Related Work\/} Various techniques for deriving
+induction rules for data types that go beyond ADTs have been
+studied. For example, Fu and Selinger~\cite{fs18} show, via examples,
+how to derive induction rules for arbitrary nested
+types. Unfortunately, however, their technique is rather {\em ad hoc},
+so is unclear how to generalize it to nested types other than the
+specific ones studied there. Moreover,~\cite{fs18} actually derives
+induction rules for data types {\em related} to the original nested
+types rather than for the original nested types themselves, and it is
+unclear whether or not the derived rules are sufficiently expressive
+to prove all results about the original nested types that we would
+expect to be provable by induction. This latter point echoes the issue
+with Coq-derived induction rule for rose trees raised in
+Section~\ref{sec:intro}, which has the unfortunate effect of forcing
+users to manually write induction (and other) rules for such types for
+use in that system. Tassi~\cite{tas19} derives induction rules for
+data type definitions in Coq using unary parametricity. His technique
+seems to be essentially equivalent to that of~\cite{jp19} for nested
+types, although he does not permit true nesting. More recently,
 Ullrich~\cite{ull20} has implemented a plugin in MetaCoq to generate
@@
 parametricity and true nesting still is not permitted.  As far as we
-know, no attempt has yet been made to extend either implementation to
-truly nested types or to proper GADTs. In fact, we know of no work
-other than that reported here that specifically addresses induction
-rules for (deep) GADTs.
+know, no attempts have been made to extend either implementation to
+truly nested types or proper GADTs or their deep variants. In fact, we
+know of no work other than that reported here that specifically
+addresses induction rules for such data types.
 
@@
 \mathsf{data\ List : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;nil\,\,\,\,\,\; :\, List\,A}\\
-\mathsf{\;\;\;\;\;\;\;cons\, :\, A \to List\,A \to List\,A} 
+\mathsf{\;\;\;\;\;\;\;nil\,\,\,\,\,\; :\, \forall \{A : Set\} \to List\,A}\\
+\mathsf{\;\;\;\;\;\;\;cons\, :\, \forall \{A : Set\} \to A \to List\,A \to List\,A} 
 \end{array}
@@
 \mathsf{data\ PTree : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;pleaf\,\,\; :\, A \to PTree\,A}\\
-\mathsf{\;\;\;\;\;\;\;pnode\, :\, PTree\,(A \times A) \to PTree\,A} 
+\mathsf{\;\;\;\;\;\;\;pleaf\,\,\; :\,\forall \{A : Set\} \to A \to PTree\,A}\\
+\mathsf{\;\;\;\;\;\;\;pnode\, :\,\forall \{A : Set\} \to PTree\,(A \times A) \to PTree\,A} 
 \end{array}
@@
 \quad\mathsf{\big( \forall (A : Set) (pp : PTree\,(A \times A)) \to
-  P\,(A \times A)\,pp \to P\,a\,(pnode\,pp)\big) \to \forall (A :
+  P\,(A \times A)\,pp \to P\,A\,(pnode\,pp)\big) \to \forall (A :
   Set) (p : PTree\,A) \to P\,A\,p } 
@@
   \to Set) \to \big( \forall (A : Set) (Q : A \to Set) (a : A) \to
-  Q\,a \to P\,A\,Q\,(Pleaf\, a) \big) \to} \\ \quad \mathsf{ \big(
+  Q\,a \to P\,A\,Q\,(pleaf\, a) \big) \to} \\ \quad \mathsf{ \big(
   \forall (A : Set) (Q : A \to Set) (pp : PTree\,(A \times A)) \to
   P\,(A \times A)\,(Pair^{\wedge}\,A\,A\,Q\,Q)\,pp \to
-  P\,A\,Q\,(Pnode\,pp) \big) \to} \\ \quad \mathsf{\forall (A : Set) (Q
+  P\,A\,Q\,(pnode\,pp) \big) \to} \\ \quad \mathsf{\forall (A : Set) (Q
   : A \to Set) (p : PTree\,A) \to PTree^{\wedge}\,A\,Q\,p \to
@@
 \mathsf{data\ Bush : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;bnil\,\,\,\,\,\; :\, Bush\,A}\\
-\mathsf{\;\;\;\;\;\;bcons\, :\, A \to Bush\,(Bush\,A) \to Bush\,A} 
+\mathsf{\;\;\;\;\;\;bnil\,\,\,\,\,\; :\, \forall \{A : Set\} \to Bush\,A}\\
+\mathsf{\;\;\;\;\;\;bcons\, :\, \forall \{A : Set\} \to A \to Bush\,(Bush\,A) \to Bush\,A} 
 \end{array}
@@
 \mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to Bush\, A \to Set)
-\to \big( \forall (A : Set) \to P\,A\,bnil \big) \to} \\ 
+\to \big( \forall (A : Set)\,(Q : A \to Set) \to P\,A\,Q\,bnil \big) \to} \\ 
 \quad\mathsf{\big( \forall (A : Set) (Q : A \to Set) (a : A) (bb :
   Bush\,(Bush\,A)) \to Q\,a \to
-  P\,(Bush\,A)\,(Bush^{\wedge}\,A\,Q)\,bb \to P\,A\,Q\,(bcons\,a\,bb)
+  P\,(Bush\,A)\,(P\,A\,Q)\,bb \to P\,A\,Q\,(bcons\,a\,bb)
   \big) \to} \\ 
@@
 resulting induction rule only inspects the top-level structure of its
-argument, rather than the contents of that structure, which exactly
-coincides with what structural induction should do.
+argument, rather than the contents of that structure, which is exactly
+what structural induction should do.
 %A concrete example of such a derivation is given in
@@
 the one being defined, its return type must still be at the (variable)
-type instance being defined. For example, every data constructor for
-$\mathsf{PTree\,A}$ returns an element of type
-$\mathsf{PTree\,A}$, regardless of the instances of $\mathsf{PTree}$
-appearing in the types of its arguments. GADTs relax this restriction,
-allowing their data constructors both to take as arguments \emph{and
-  return as results} data whose types involve instances other
-than the one being defined. That is, GADTs' constructors' return type
-instances can, like that of $\mathsf{pair}$ in~\eqref{eq:seq}, be
-structured.\looseness=-1
+type instance being defined. For example, each of $\mathsf{pleaf}$ and
+$\mathsf{pnode}$ returns an element of type $\mathsf{PTree\,A}$
+regardless of the instances of $\mathsf{PTree}$ appearing in the types
+of its arguments. GADTs relax this restriction, allowing their data
+constructors both to take as arguments \emph{and return as results}
+data whose types involve instances other than the one being
+defined. That is, GADTs' constructors' return type instances can, like
+that of $\mathsf{pair}$ in~\eqref{eq:seq}, be structured.\looseness=-1
 %these instances can be structured.
@@
 \mathsf{data\ Equal : Set \to Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;\;\;\;\;\,refl :\, Equal\,A\,A}
+\mathsf{\;\;\;\;\;\;\;\;\;\;\;\,refl :\, \forall \{A : Set\} \to Equal\,A\,A}
 \end{array}
@@
 is parameterized by two type indices, but it is only possible to
-construct data elements of type $\mathsf{Equal\,a\,b}$ if $\mathsf{a}$
-and $\mathsf{b}$ are instantiated at the same type. If the types
-$\mathsf{a}$ and $\mathsf{b}$ are syntactically identical then the
-type $\mathsf{Equal\,a\,b}$ contains the single data element
+construct data elements of type $\mathsf{Equal\,A\,B}$ if $\mathsf{A}$
+and $\mathsf{B}$ are instantiated at the same type. If the types
+$\mathsf{A}$ and $\mathsf{B}$ are syntactically identical then the
+type $\mathsf{Equal\,A\,B}$ contains the single data element
 $\mathsf{refl}$. It contains no data elements otherwise.
@@
 data in two already existing sequences. This GADT can be rewritten as
-its Henry Ford encoding, which makes critical use of the equality
-GADT, as follows:
+its Henry Ford encoding~\cite{ch03,hin03,mcb99,sjsv09,sp04}, which
+makes critical use of the equality GADT, as follows:
 \begin{equation}\label{eq:eq_seq}
@@
 \mathsf{data\ Seq : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;const :\, A \to Seq\,A}\\ 
-\mathsf{\;\;\;\;\;\;\;pair\,\;\; :\, \forall (B\,C : Set) \to
+\mathsf{\;\;\;\;\;\;\;const :\, \forall \{A : Set\} \to A \to Seq\,A}\\ 
+\mathsf{\;\;\;\;\;\;\;pair\,\;\; :\, \forall \{A : Set\} \to \forall (B\,C : Set) \to
   Equal\,A\,(B \times C) \to Seq\,B \to Seq\,C \to Seq\,A}\\ 
@@
 encodes terms of a simply typed lambda calculus. More robust
-variations on $\mathsf{LTerm}$ are, of course, possible. But since
-this variation is rich enough to illustrate all essential aspects of
-deep GADTs --- and later, in Section~\ref{sec:ind-lam}, their deep
+variations on $\mathsf{LTerm}$ are, of course, possible, but this
+variation is rich enough to illustrate all essential aspects of deep
+GADTs --- and later, in Section~\ref{sec:ind-lam}, their deep
 induction rules --- while still being small enough to ensure clarity
-of exposition, we keep it to a minimum.
+of exposition.
 
@@
 \mathsf{data\ LType : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;\;bool :\, \forall (B : Set) \to Equal\,A\,Bool
+\mathsf{\;\;\;\;\;\;\;\;bool :\, \forall \{A : Set\} \to \forall (B : Set) \to Equal\,A\,Bool
   \to LType\,A}\\ 
-\mathsf{\;\;\;\;\;\;\;\;arr\;\;\; :\, \forall (B\,C : Set) \to
+\mathsf{\;\;\;\;\;\;\;\;arr\;\;\; :\, \forall \{A : Set\} \to \forall (B\,C : Set) \to
   Equal\,A\,(B \to C) \to LType\,B \to LType\,C \to LType\,A}\\ 
-  \mathsf{\;\;\;\;\;\;\;\;list\,\;\; :\, \forall (B : Set) \to
+  \mathsf{\;\;\;\;\;\;\;\;list\,\;\; :\,  \forall \{A : Set\} \to\forall (B : Set) \to
     Equal\,A\,(\List\,B) \to LType\,B \to LType\,A} 
@@
 Terms are either variables, abstractions, applications, or lists of
-terms. They are represented by
+terms. They are similarly represented by
 \begin{equation}\label{eq:eq_lterm}
@@
 \mathsf{data\ LTerm : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;\;var\,\,\,:\, String \to LType\,A \to
+\mathsf{\;\;\;\;\;\;\;\;var\,\,\,:\,  \forall \{A : Set\} \to String \to LType\,A \to
   LTerm\,A} \\  
-\mathsf{\;\;\;\;\;\;\;\;abs\,\, :\, \forall (B\,C : Set) \to
+\mathsf{\;\;\;\;\;\;\;\;abs\,\, :\,  \forall \{A : Set\} \to \forall (B\,C : Set) \to
   Equal\,A\,(B \to C) \to String \to LType\,B \to LTerm\,C \to
   LTerm\,A}\\ 
-  \mathsf{\;\;\;\;\;\;\;\;app\; :\, \forall (B : Set) \to
+  \mathsf{\;\;\;\;\;\;\;\;app\; :\,  \forall \{A : Set\} \to \forall (B : Set) \to
     LTerm (B \to A) \to LTerm\,B \to LTerm\,A} \\ 
-  \mathsf{\;\;\;\;\;\;\;\;list\;\,\, :\, \forall (B : Set) \to
+  \mathsf{\;\;\;\;\;\;\;\;list\;\,\, :\,  \forall \{A : Set\} \to \forall (B : Set) \to
     Equal\,A\,(List\,B) \to List\,(LTerm\,B) \to LTerm\,A} 
@@
   \to Equal\,A\,B \to Set) \to \big( \forall (C : Set) \to
-  P\,C\,C\,refl \big) \to \forall (A\,B : Set) (e: Equal\,A\,B) \to
-  P\,A\,B\,e}
+  Q\,C\,C\,refl \big) \to \forall (A\,B : Set) (e: Equal\,A\,B) \to
+  Q\,A\,B\,e}
 \end{equation}
@@
 \mathsf{Seq^{\wedge}\,A\,Q_A\,(const\,a)} & = & \mathsf{Q_A\,a}\\
-\mathsf{Seq^{\wedge}\,A\,Q_A\,(sPair\,B\,C\,e\,s_B\,s_C)}
+\mathsf{Seq^{\wedge}\,A\,Q_A\,(pair\,B\,C\,e\,s_B\,s_C)}
 &=&\mathsf{\exists [Q_B] \exists [Q_C]\, Equal^{\wedge}\,A\, (B
-  \times C)\, Q_A\, (Q_B \times Q_C) \, e \times
+  \times C)\, Q_A\, (Pair^\wedge\,B\,C\, Q_B \, Q_C) \, e \times
   Seq^{\wedge}\,B\,Q_B\,s_B \times Seq^{\wedge}\,C\,Q_C\,s_C}
@@
   \,x}$ is syntactic sugar for the type of dependent pairs
-$\mathsf{(x,b)}$ where $\mathsf{x : A}$ and $\mathsf{b : F\, x}$ and
+$\mathsf{(x,b)}$, where $\mathsf{x : A}$, $\mathsf{b : F\, x}$, and
 $\mathsf{F : A \to Set}$.
@@
   (Pair^{\wedge}\,B\,C\,Q_B\,Q_C)\, e \to P\,B\,Q_B\,s_B \to
-  P\,C\,Q_C\,s_C \to P A Q_A ( pair\,B\,C\,e\,s_B\,s_C )}
+  P\,C\,Q_C\,s_C \to P\, A\, Q_A\, ( pair\,B\,C\,e\,s_B\,s_C )}
 \end{array}\]
@@
     = &  \mathsf{cconst\;A\;Q_A\;a\;liftA}\\
-\mathsf{dIndSeq\,P\,cconst\,cpair\,A\,Q_A\,(sPair\,B\,C\,e\,s_B\,s_C)\,
+\mathsf{dIndSeq\,P\,cconst\,cpair\,A\,Q_A\,(pair\,B\,C\,e\,s_B\,s_C)\,
     (Q_B,Q_C, liftE, liftB, liftC)}  & = &
@@
 $\mathsf{liftA : Seq^{\wedge}\,A\,Q_A\,(const\,a) = Q_A\,a}$. In the
-second, also $\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C : C \to Set}$,
-$\mathsf{e : Equal\,A\,(B \times C)}$, $\mathsf{s_B : Seq\,B}$,
-$\mathsf{s_C : Seq\,C}$, $\mathsf{liftE : Equal^{\wedge}\,A\, (B
-  \times C)\, Q_A\, (Q_B \times Q_C) \, e}$, $\mathsf{liftB :
-  Seq^{\wedge}\,B\,Q_B\,s_B}$, and $\mathsf{liftC :}$
-$\mathsf{Seq^{\wedge}\,C\,Q_C\,s_C}$ --- which together ensure that
-$\mathsf{(Q_B, Q_C, liftE, liftB, liftC) :
-  Seq^{\wedge}\,A\,Q\,(sPair\,B\,C\,e\,s_B\,s_C)}$ --- and
+second clause we also have $\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C : C \to
+  Set}$, $\mathsf{e : Equal\,A\,(B \times C)}$, $\mathsf{s_B :
+  Seq\,B}$, $\mathsf{s_C : Seq\,C}$, $\mathsf{liftE :
+  Equal^{\wedge}\,A\, (B \times C)\, Q_A\, (Pair^\wedge \,B\,C\,Q_B \,
+  Q_C) \, e}$, $\mathsf{liftB : Seq^{\wedge}\,B\,Q_B\,s_B}$, and
+$\mathsf{liftC :}$ $\mathsf{Seq^{\wedge}\,C\,Q_C\,s_C}$ --- which
+together ensure that $\mathsf{(Q_B, Q_C, liftE, liftB, liftC) :
+  Seq^{\wedge}\,A\,Q\,(pair\,B\,C\,e\,s_B\,s_C)}$ --- and
 $\mathsf{p_B} =\mathsf{dIndSeq\,P\,cconst\,cpair\,B\,Q_B\,s_B\,liftB :
@@
     \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;\;c\, :\, F\,G\,\ol{B} \to G (\ol{K\,\ol{B}})}
+\mathsf{\;\;\;\;\;\;\;\;c\, :\, \forall \{\ol{B : Set}\} \to F\,G\,\ol{B} \to G (\ol{K\,\ol{B}})}
 \end{array}
@@
 For brevity and clarity we indicate only one constructor $\mathsf{c}$
-in~\eqref{eq:gadts}, even though a GADT can, in general, have any
-finite number of them, each with a type the same form as
-$\mathsf{c}$'s. In~\eqref{eq:gadts}, $\mathsf{F}$ and $\mathsf{K}$ are
-type constructors with signatures $\mathsf{(Set^{\alpha} \to Set) \to
-  Set^{\beta} \to Set}$ and $\mathsf{Set^{\beta} \to Set}$,
-respectively. If $\mathsf{T}$ is a type constructor with signature
-$\mathsf{Set^{\gamma} \to Set}$ then we say that $\mathsf{T}$ has
-arity $\mathsf{\gamma}$.  The overline notation denotes a finite list
-whose length is exactly the arity of the type constructor being
-applied to it. The number of type constructors in $\ol{\mathsf{K}}$
-must therefore be $\alpha$. In addition, the type constructor
+in~\eqref{eq:gadts}, even though a GADT can have any finite number of
+them, each with a type of the same form as
+$\mathsf{c}$'s. In~\eqref{eq:gadts}, $\mathsf{F}$ and each
+$\mathsf{K}$ in $\ol{\mathsf{K}}$ are type constructors with
+signatures $\mathsf{(Set^{\alpha} \to Set) \to Set^{\beta} \to Set}$
+and $\mathsf{Set^{\beta} \to Set}$, respectively. If $\mathsf{T}$ is a
+type constructor with signature $\mathsf{Set^{\gamma} \to Set}$ then
+$\mathsf{T}$ has {\em arity} $\mathsf{\gamma}$.  The
+overline notation denotes a finite list whose length is exactly the
+arity of the type constructor being applied to it. The number of type
+constructors in $\ol{\mathsf{K}}$ (resp., $\ol{\mathsf{B}}$) is thus
+$\alpha$ (resp., $\beta$). In addition, the type constructor
 $\mathsf{F}$ must be constructed inductively according to the
@@
 discussion. A further requirement that applies to all of the type
-constructors appearing in the right-hand side of the above grammar
+constructors appearing in the right-hand side of the above grammar,
 including those in $\ol{\mathsf{K}}$, is that they must all admit
@@
 type application admit predicate liftings, and so do GADTs constructed
-from the grammar; in fact, we have seen such liftings for products and
+from the grammar. In fact, we have seen such liftings for products and
 type application in Section~\ref{sec:deep-ind-GADTs}. A concrete way
@@
 %\end{array}\]
-for $\mathsf{K}$, then the induction hypothesis associated with
-$\mathsf{c}$ is
+for $\mathsf{K}$, then the induction hypothesis $\mathsf{dIndC}$
+associated with $\mathsf{c}$ is
 \[\begin{array}{l}
@@
 $\mathsf{liftF_1 : F_1^{\wedge}\,\ol{B}\,\ol{Q_B}\,z}$. Note that
-$\mathsf{F_1}$ not containing $\mathsf{G}$ is a necessary restriction,
+$\mathsf{F_1}$ not containing $\mathsf{G}$ is a necessary restriction
 since the proof relies on
@@
 
-\section{Truly Nested GADTs Do Not Admit Deep Induction
+\section{Truly Nested GADTs Need Not Admit Deep Induction
     Rules}\label{sec:GADT-nested}
@@
 \mathsf{data\ G : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;\;\;c :\, G\,(G\,A) \to G\,(A \times A)}
+\mathsf{\;\;\;\;\;\;\;\;\;c :\,  \forall \{A : Set\} \to G\,(G\,A) \to G\,(A \times A)}
 \end{array}
@@
 is equal to $\mathsf{A'}$ does not provide a proof that $\mathsf{B}$
-is equal to $\mathsf{B'}$. This difficulty propagates to any truly
-nested GADT because it is not possible to define a map function
-$\mathsf{Equal^\wedge Map}$
-%for $\mathsf{Equal^\wedge}$
-when one of the
-type arguments to $\mathsf{Equal^\wedge}$ is structured (i.e., not a
-variable). And this is, of course, a key component in any approach to
-deep induction for truly nested GADTs.\looseness=-1
+is equal to $\mathsf{B'}$. And not being able to define
+$\mathsf{Equal^\wedge Map :}$ $\mathsf{\forall (A\, B : Set) (Q_A\,
+  Q_A' : A \to Set) (Q_B\, Q_B' : B \to Set) \to PredMap\,A\,Q_A\,Q_A'
+  \to PredMap\,B\,Q_B\,Q_B' \to
+  PredMap\,(Equal\,A\,B)}\\ \mathsf{(Equal^\wedge
+  A\,B\,Q_A\,Q_B)\,(Equal^\wedge A\,B\,Q_A'\,Q_B')}$ of course makes
+it impossible to define $\mathsf{G^\wedge Map}$ for more general
+$\mathsf{G}$.
 
@@
 \mathsf{data\ Maybe : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;nothing :\, Maybe\,A}\\
-\mathsf{\;\;\;\;\;\;just\;\;\;\;\;\;\; :\, A \to Maybe\,A}
+\mathsf{\;\;\;\;\;\;nothing :\,  \forall \{A : Set\} \to Maybe\,A}\\
+\mathsf{\;\;\;\;\;\;just\;\;\;\;\;\;\; :\,  \forall \{A : Set\} \to A \to Maybe\,A}
 \end{array}
@@
 \[ \mathsf{getTypeProof : \forall \, (A : Set)\, (t : LTerm\,A) \to
-   GetType \,A \,t}\] This property cannot be proved without deep
-induction, which is needed to apply the induction hypothesis to the
+   GetType \,A \,t}\] This property can be proved with deep
+induction, which is used to apply the induction hypothesis to the
 individual terms in the list of terms that the data constructor
-$\mathsf{list}$ takes as an argument. But using the deep induction
+$\mathsf{list}$ takes as an argument. Indeed, using the deep induction
 rule $\mathsf{dIndLTerm}$ from Section~\ref{sec:ind-lam} we can define
@@
   &\mathsf{clist\, A\, B\, Q\, Q'\, e\, (cons\, t\, ts') \, liftE\,
-    (just\, T' , lift_{ts'}) = just \, (TList\, B\, e\, T')} 
+    (just\, T' , lift_{ts'}) = just \, (list\, B\, e\, T')} 
 \end{align*}
@@
     ts) }$, we define $\mathsf{LTerm^\wedge KT\,A\, (list\, B\, e\,
-    ts) = (K_\top , \, Equal^\wedge ListKT , \, L_{ListL^\wedge
-      Term^\wedge KT} ) }$.
+    ts) = (K_\top , \, Equal^\wedge ListKT , \, L_{List^\wedge
+      LTerm^\wedge KT} ) }$.
 \end{itemize}
@@
 $\mathsf{G^{\wedge}\, A\, K_\top \, x}$ for every term $\mathsf{x :
-  G\, A}$, we need to know that if $\mathsf{G}$ has a constructor
+  G\, A}$, we need to know that, if $\mathsf{G}$ has a constructor
 $\mathsf{c : H \, (\ol{F\,G\, \ol{B}}) \to G\, (\ol{K\,\ol{B}})}$,
@@
 \bibitem{atk12} Atkey, R.  {\em Relational parametricity for higher
-  kinds}.  Computer Science Logic, pp.~46-61, 2012.
+  kinds}.  Proceedings, Computer Science Logic, pp.~46-61, 2012.
 
@@
 \bibitem{jgj21} Johann, P., Ghiorzi, E., and Jeffries, D. {\em
-  Functorial GADTs are not Parametric}. Submitted, 2021.
+  GADTs, Functoriality, Parametricity: Pick Two}. Submitted, 2021.
 

### 2021-06-08 20-gadts/21-LSFA/GADT-story.tex
final LSFA paper

@@
 $\mathsf{t \,::\,Tree\,a}$, and $\mathsf{g :: Tree\,a\to List\,a}$
-arranges lists into trees in a type-independent way, then we have the
+arranges trees into lists in a type-independent way, then we have the
 following rearrange-transform property:
@@
 associated with $\mathsf{T\,a}$ also supports its own $\mathsf{map}$
-function, we can require that $F$ also be the fixpoint $\mu H$ of a
-functor $H$. In this case, however, $H$ must be a functor on the
-category $\mathit{Set}^\mathit{Set}$ of functors and natural
-transformations on $\mathit{Set}$. That is, $H$ must be a {\em
-  higher-order} functor on $\mathit{Set}$. Reflecting this requirement
-back into syntax gives that $\mathsf{T\,a} = \mathsf{(\mu H)\,a}$ for
-the ``type constructor constructor'' $\mathsf{H}$ that supports
+function, we can require that $F$ be a functor on the category
+$\mathit{Set}^\mathit{Set}$ of functors and natural transformations on
+$\mathit{Set}$. That is, $F$ must be a {\em higher-order} functor on
+$\mathit{Set}$. Writing $H$ in place of $F$ to emphasize that it is
+higher-order, and reflecting this requirement back into syntax, we
+have that $\mathsf{T\,a} = \mathsf{(\mu H)\,a}$ for the ``type
+constructor constructor'' $\mathsf{H}$ that supports
 suitable\footnote{The $\mathsf{map}$ function for $\mathsf{H}$ is
@@
 Proposition~7.1 of~\cite{blw03} gives that $(\mathit{Lan}_{\lambda
-  u. 1} \lambda u. 1)\,A$ can be computed as the set of all functions
-$a : 1 \to A$ modulo the smallest equivalence relation $\sim$ such
-that, for any $f : 1 \to 1$, we have $a \sim a \circ f$.  But this set
-is isomorphic to $A$, because the only function $f : 1 \to 1$ is the
-identity function. This is different from the interpretation of
-$\mathsf{G}$'s Church encoding from the previous paragraph.
+  u. 1} \lambda u. 1)\,A$ can be computed as
+\[\big( \bigcup_{U : \mathit{Set}^0,\,f: (\lambda u.\,1)\,U \to A} (\lambda
+u. 1)\,U \;\big) \,/\,\sim
+\;\;\;\;=\;\;\;\;
+\big( \bigcup_{U :
+  \mathit{Set}^0,\,f: 1 \to A} 1 \big)\,/\,\sim\]
+%\;\;\;\;=\;\;\;\;
+%\big( \bigcup_{f: (\lambda u.1)\,U \to A} 1 \big)\,/\,\sim\]
+where $U$ is the unique object of $\mathit{Set}^0$, $*$ is the unique
+element of the singleton set $1$, and $\sim$ is the smallest
+equivalence relation such that $(U,f, *)$ and $(U,f',*)$ are related
+if
+\[
+\begin{tikzcd}
+(\lambda u.1)\,U \ar[rr, "{\mathit{(\lambda u. 1)\,id_U}}"] \ar[dr, "f"']
+&& (\lambda u.1)\,U \ar[dl, "f'"] \\
+&A
+\end{tikzcd} \;\;\; =\;\;\;
+\begin{tikzcd}
+1 \ar[rr, "{\mathit{id_1}}"] \ar[dr, "f"']
+&&1 \ar[dl, "f'"] \\
+&A
+\end{tikzcd}
+\]
+commutes, i.e., if $f = f'$. Since the relation generating $\sim$ is
+already an equivalence relation, we have that $(U,f,*) \sim (U,f', *)$
+iff $f = f'$. Thus, up to isomorphism, $(\mathit{Lan}_{\lambda u. 1}
+\lambda u. 1)\,A = \{ f : 1 \to A\}$, i.e., $(\mathit{Lan}_{\lambda
+  u. 1} \lambda u. 1)\,A = A$. This is different from the
+interpretation of $\mathsf{G}$'s Church encoding from the previous
+paragraph.
 
@@
 These additional data elements can be obtained by reflecting back into
-syntax the elements $\mathit{map}_G\,f_a\,c : G\,A$ resulting from
+syntax the elements $\mathit{map}_G\,f_a\,c \in G\,A$ resulting from
 applying the functorial action $\mathit{map}_G$ of $\mathsf{G}$'s
@@
 $\mathsf{map}$? From a semanticist's point of view, however, the
-prinitive representation of GADTs is entirely reasonable. Indeed, they
+primitive representation of GADTs is entirely reasonable. Indeed, they
 would likely find the nonfunctorial nature of a GADT's Church encoding
@@
 \section{Parametricity in the Presence of GADTs}\label{sec:par}
-
 {\em Relational parametricity} encodes a powerful notion of type
@@
 relations as well. The following definition is standard:
-\begin{definition}
+\begin{definition}\label{def:rel}
   The category $\mathit{Rel}$ has:
@@
 variable $\mathsf{x :: F[a]}$ as a function $t$ associating to each
-set $A$ a morphism $t \,A : F_0\,A \to G_0\,A$ in $\mathit{Set}$.
+set $A$ a morphism $t \,A : F_0\,A \to G_0\,A$ in
+$\mathit{Set}$. Here, $F_0$ is the set interpretation of $\mathsf{F}$.
 These interpretations are given inductively on the structures of
@@
 theorems are required for types and terms with any number of free
-variables.
+variables. In particular, if $\mathsf{t}$ is closed (i.e., has no free
+term variables) then $t\,A \in G_0\,A$.
 
@@
 elements that can be formed using its data constructors and whose type
-is that instance. In terms of syntax the parametricity property for a
-GADT's Church encoding gives that the GADT is inhabited at any
-instance for which data elements can be formed using its data
-constructors. In particular, the parametricity property for the GADT
-$\mathsf{G}$ from~\eqref{eq:G} gives that $\mathsf{G\,a}$ contains a
-single data element if $\mathsf{a}$ is semantically equivalent to
-$\mathsf{1}$ and none otherwise:
+is that instance. The parametricity property for the Church encoding
+of a GADT $\mathsf{G}$ gives that $\mathsf{G\,a}$ is inhabited iff data
+elements of the instance of $\mathsf{G}$ at $\mathsf{a}$ can be formed
+using $\mathsf{G}$'s data constructors.  In particular, the
+parametricity property for the GADT $\mathsf{G}$ from~\eqref{eq:G}
+gives that $\mathsf{G\,a}$ contains a single data element if
+$\mathsf{a}$ is semantically equivalent to $\mathsf{1}$ and none
+otherwise:
 
 \begin{example}\label{ex:CE-par}
-Let $\mathsf{t}$ be a term of type $\mathsf{G\,a}$ for the GADT
+Let $\mathsf{t}$ be a closed term of type $\mathsf{G\,a}$ for the GADT
 $\mathsf{G}$ defined in~\eqref{eq:G}, let $G = (G_0,G_1)$ be the
@@
 Intuitively, when $\mathsf{G}$ is viewed as a fixpoint, its data
-elements include those given by functorial completion. Thus, given any
-relation $S : \mathit{Rel}\,(C, D)$ and any morphism $m : S \to R$,
-$G_1\, R$ must contain all elements of the form $G_1\,m\,x$ for $x \in
-G_1\,S$.  But the two components $m_1 : C \to A$ and $m_2 : D \to B$
-of $m$ cannot be given independently of one another, since
+elements include those given by functorial completion. Since $G_1$ is
+a functor, given any relation $S : \mathit{Rel}\,(C, D)$ and any
+morphism $m : S \to R$, $G_1\, R$ must contain all elements of the
+form $G_1\,m\,x$ for $x \in G_1\,S$.  But the two components $m_1 : C
+\to A$ and $m_2 : D \to B$ of $m$ cannot be given independently of one
+another, since Definition~\ref{def:rel} entails that
 %$m$ has to send elements related by $S$ to elements
@@
   u\!.\,\mathit{Eq}_1} \lambda u\!. \,1)\, R\] (Note that the left Kan
- extension of $\lambda u\!. \,1$ along $\lambda u\!.\,\mathit{Eq}_1$
- is a functor from $\mathit{Rel}$ to $\mathit{Set}$.) By
- Proposition~7.1 of~\cite{blw03} the left Kan extension on the
- right-hand side can be computed as
-\[ \big( \bigcup_{m\, :\, \mathit{Eq}_1 \to R} 1 \big) \; /\; \approx\]
-Here, $*$ is the unique element of the singleton set $1$, and $\approx$
-is the smallest equivalence relation such that $(m,\ast) \approx
-(m',\ast)$ for $m, m' : \mathit{Eq}_1 \to R$ if there exists $f :
-\mathit{Eq}_1 \to \mathit{Eq}_1$ such that $f (\ast, \ast) = (\ast,
-\ast)$ and the following diagram commutes:
+extension of $\lambda u\!. \,1$ along $\lambda u\!.\,\mathit{Eq}_1$ is
+a functor from $\mathit{Rel}$ to $\mathit{Set}$.)  By the same
+reasoning as in Example~\ref{ex:g1}, Proposition~7.1 of~\cite{blw03}
+gives that $(\mathit{Lan}_{\lambda u. \mathit{Eq}_1} \lambda u. 1)\,R$
+can be computed as
+\[\big( \bigcup_{U : \mathit{Rel}^0,\,m: (\lambda u.\,\mathit{Eq}_1)\,U \to R} (\lambda
+u. 1)\,U \;\big) \,/\,\approx \;\;\;\;=\;\;\;\; \big( \bigcup_{U :
+  \mathit{Rel}^0,\,m: \mathit{Eq}_1 \to R} 1 \big)\,/\,\approx\] where
+$U$ is the unique object of $\mathit{Rel}^0$, $*$ is the unique
+element of the singleton set $1$, and $\approx$ is the smallest
+equivalence relation such that $(U,m,*)$ and $(U,m',*)$ are related if
 \[
 \begin{tikzcd}
-\mathit{Eq}_1 \ar[rr, "{f}"] \ar[dr, "m"']
+(\lambda u.\mathit{Eq}_1)\,U \ar[rr, "{\mathit{(\lambda
+        u.\mathit{Eq}_1)\,id_U}}"] \ar[dr, "m"'] 
+&& (\lambda u. \mathit{Eq}_1)\,U \ar[dl, "m'"] \\
+&R
+\end{tikzcd} \;\;\; =\;\;\;
+\begin{tikzcd}
+\mathit{Eq}_1 \ar[rr, "{\mathit{id_{\mathit{Eq}_1}}}"] \ar[dr, "m"']
 &&\mathit{Eq}_1 \ar[dl, "m'"] \\
@@
 \]
-The only morphism $f : \mathit{Eq}_1 \to \mathit{Eq}_1$ is $f = (!,
-!)$, where $!$ is the unique function from $1$ to $1$.  Since this
-morphism does satisfy the condition $f (\ast, \ast) = (\ast, \ast)$,
-we have that $(m,\ast) \approx (m',\ast)$ if and only if $m =
-m'$. Thus
-\[ \big( \bigcup_{m\, :\,
-  \mathit{Eq}_1 \to R} 1 \big) \; /\; \approx \;\;\; = \;\; \{m :
-\mathit{Eq}_1 \to R\} \;=\; \{(!, k_0), (!, k_1)\}\] where $k_0, k_1 :
-1 \to 2$ are the constantly $0$-valued and $1$-valued functions in
-$\mathit{Set}$, respectively, and therefore $\pi_1\,\big(
-(\mathit{Lan}_{\lambda u\!.\,\mathit{Eq}_1} \lambda
+commutes, i.e., if $m = m'$. Since the relation generating $\approx$
+is already an equivalence relation, we have that $(U,m,*) \approx
+(U,m', *)$ iff $m = m'$. Thus, up to isomorphism,
+$(\mathit{Lan}_{\lambda u\!.\,\mathit{Eq}_1} \lambda u\!. \,1)\, R =
+\{ m : \mathit{Eq}_1 \to R\}$.  But this set is $\{(!, k_0), (!,
+k_1)\}$, where $k_0, k_1 : 1 \to 2$ are the constantly $0$-valued and
+$1$-valued functions in $\mathit{Set}$, respectively, and therefore
+$\pi_1\,\big( (\mathit{Lan}_{\lambda u\!.\,\mathit{Eq}_1} \lambda
 u\!.\,\mathit{Eq}_1) \, R \big)$ is not $1$, as would be needed for

### 2021-06-23 21-lmcs/21-lmcs.tex
first LMCS revisions

--- 2021-01-12 free-theorems-nesteds/20-mscs.tex
@@
 Definition~\ref{def:wftypes} ensures that the expected weakening rules
-for well-formed types hold, although weakening does not change the
-contexts in which types can be formed. If $\Gamma;\emptyset \vdash F$
-and $\Gamma;\emptyset \vdash G$, then our rules allow formation of the
-types $\Gamma;\emptyset \vdash \Nat^\emptyset F \,G$ and $\Gamma;
-\emptyset \vdash \Nat^{\ol\alpha} \,\onet \,F$, which represent the
-System F types $\Gamma \vdash F \to G$ and $\forall$-type $\Gamma;
-\emptyset \vdash \forall \ol\alpha . F$, respectively. However, some
-System F types, such as $\forall \alpha. (\alpha \to \alpha) \to
-\alpha$, are not representable in our calculus. Note that, in the
-rule for $\mu$-types, no variables in $\Phi$ are applied to types
-involving variables from $\ol\alpha$ in the body $F$ of $\mu
-\phi. \lambda \ol\alpha.F$. This will be critical to proving the
-Identity Extension Lemma for our calculus.
+for well-formed types hold, i.e., if $\Gamma; \Phi \vdash F$ is
+well-formed, then both $\Gamma, \ol{\psi}; \Phi \vdash F$ and $\Gamma;
+\Phi, \ol{\phi} \vdash F$ are also well-formed. Note, however, that
+weakening does not change the contexts in which types can be formed.
+For example, the functorial context in which a type $\Nat^\emptyset F
+\,G$ is formed must be empty, although it can be weakened by a
+functorial variable $\beta$ {\em after} the type is formed; by
+contrast, we cannot {\em first} weaken the functorial contexts of $F$
+and $G$ to $\Gamma; \beta \vdash F$ and $\Gamma; \beta \vdash G$, and
+{\em then} form $\Nat^\emptyset F \,G$ in the weakened context
+$\Gamma; \beta$.  If $\Gamma;\emptyset \vdash F$ and $\Gamma;\emptyset
+\vdash G$, then our rules allow formation of the type
+$\Gamma;\emptyset \vdash \Nat^\emptyset F \,G$, which represents the
+System F type $\Gamma \vdash F \to G$. Similarly, if
+$\Gamma;\ol{\alpha} \vdash F$, then our rules allow formation of the
+type $\Gamma; \emptyset \vdash \Nat^{\ol\alpha} \,\onet \,F$, which
+represents the System F type $\Gamma; \emptyset \vdash \forall
+\ol\alpha . F$. However, some System F types, such as $\forall
+\alpha. (\alpha \to \alpha) \to \alpha$, are not representable in our
+calculus. Note that, in the rule for $\mu$-types, no variables in
+$\Phi$ are applied to types involving variables from $\ol\alpha$ in
+the body $F$ of $\mu \phi. \lambda \ol\alpha.F$. This will be critical
+to proving the Identity Extension Lemma for our calculus.
 
@@
 
-Note that since the body F of a type $(\mu \phi. \lambda
+Note that since the body $F$ of a type $(\mu \phi. \lambda
 \ol\alpha. F)\ol{G}$ can only be functorial in $\phi$ and the
@@
 $\phi$. It is not hard to see that $\Gamma;\Phi \vdash H[\phi
-  :=_{\ol{\alpha}} F]$. In addition, if \,$\Gamma, \phi^k; \Phi \vdash
-H$,\, $\Gamma; \ol\psi,\ol{\alpha} \vdash F$, $|\ol\alpha| = k$, and
-$\Phi \cap \ol\psi = \emptyset$, then $\Gamma,\ol\psi';\Phi \vdash
-H[\phi :=_{\ol{\alpha}} F[\ol{\psi :== \psi'}]]$.
+  :=_{\ol{\alpha}} F]$. {\color{red} REMOVE? In addition, if
+  \,$\Gamma, \phi^k; \Phi \vdash H$,\, $\Gamma; \ol\psi,\ol{\alpha}
+  \vdash F$, $|\ol\alpha| = k$, and $\Phi \cap \ol\psi = \emptyset$,
+  then $\Gamma,\ol\psi';\Phi \vdash H[\phi :=_{\ol{\alpha}} F[\ol{\psi
+        :== \psi'}]]$.}
 
@@
 each non-functorial monomorphic instantiation of $\mathit{flatten}$
-over a list of perfect trees is well-typed:
+over a list of perfect trees is well-formed:
 \[\begin{array}{l}
@@
 $\mathtt{reverseBush}$ function that reverses the data in a bush. This
-function maps, e.g., the bush
+function maps, e.g., the bush {\color{red} Change one 2 to 4? Find a
+  better name for this operation than reverse.}
 \[\begin{array}{l}
@@
 (\mathit{PTree}\,\alpha) \,(\Nat^\emptyset (\mathit{PTree}\,\alpha)\,
-(\mathit{PTree}\,(\alpha \times \alpha)))$ is not well-typed.
+(\mathit{PTree}\,(\alpha \times \alpha)))$ is not well-formed.
 Uncurrying gives $\Nat^\alpha (\mathit{PTree}\,\alpha \times
 \mathit{PTree}\,\alpha)\, (\mathit{PTree}\,(\alpha \times \alpha))$,
-which is well-typed, but $\mathsf{fold}$ cannot produce a term of this
+which is well-formed, but $\mathsf{fold}$ cannot produce a term of this
 type because $\mathit{PTree}\,\alpha \times \mathit{PTree}\,\alpha$ is
@@
 $i = 1,...,k$ and $\rho' \alpha = \rho \alpha$ if $\alpha \not \in
-\{\alpha_1,...,\alpha_k\}$.  If $\rho$ is a set environment we write
-$\Eq_\rho$ for the relation environment (see
-Definition~\ref{def:reln-env}) such that $\Eq_\rho v = \Eq_{\rho v}$
-for every type variable $v$.
-
-We can now define our set interpretations.  The relational
-interpretations appearing in the second clause of
-Definition~\ref{def:set-sem} are given in
-Definition~\ref{def:rel-sem}.
+\{\alpha_1,...,\alpha_k\}$.
+
+We can now define our set interpretations.  If $\rho$ is a set
+environment we write $\Eq_\rho$ for the equality relation environment
+such that $\Eq_\rho \phi = \Eq_{\rho \phi}$ for every type variable
+$\phi$; see Definitions~\ref{def:rel-transf} and~\ref{def:reln-env}
+for the definitions of a relation transformer and a relation
+environment, and Equation~\ref{def:eq-transf} for the definition of
+the relation transformer $\Eq_F$ on a functor $F$.  Equality relation
+environments appear in the third clause of
+Definition~\ref{def:set-sem}.  The relational interpretations also
+appearing in the third clause of Definition~\ref{def:set-sem} are
+given in Definition~\ref{def:rel-sem}.
 
@@
 transformer preserves equality relations.  The {\em equality relation
-  transformer} on $F : [\set^k,\set]$ is defined to be $\Eq_F =
-\graph{\id_{F}} = (F, F, \graph{\id_{F}}^*)$.
+  transformer} on $F : [\set^k,\set]$ is defined to be
+\begin{equation}\label{def:eq-transf}
+  \Eq_F = \graph{\id_{F}} = (F, F, \graph{\id_{F}}^*)
+\end{equation}  
 Lemma~\ref{lem:eq-reln-equalities} then gives that, for all $\ol{A :

### 2021-06-24 21-lmcs/21-lmcs.tex
second LMCS revisions

@@
 $i = 1,...,k$ and $\rho' \alpha = \rho \alpha$ if $\alpha \not \in
-\{\alpha_1,...,\alpha_k\}$.
+\{\alpha_1,...,\alpha_k\}$. We note that $\lambda \ol{A}.\,
+\rho[\ol{\alpha  := A}]$ is a functor in $\ol{A}$.
 
@@
 $\omega$-cocontinuity of $\setsem{\Gamma;\ol{\alpha} \vdash F}\rho$
-ensure that $\{\eta : \setsem{\Gamma;\ol{\alpha} \vdash F}\rho
-\Rightarrow \setsem{\Gamma;\ol{\alpha} \vdash G}\rho\}$ (which
-contains $\setsem{\Gamma;\emptyset \vdash
-  \Nat^{\ol{\alpha}}\,F\,G}\rho$) is a subset of
-$\big\{({\setsem{\Gamma;\ol{\alpha} \vdash G}\rho[\ol{\alpha :=
-      S}]})^{(\setsem{\Gamma;\ol{\alpha} \vdash F}\rho[\ol{\alpha :=
-      S}])}~\big|~ \ol{S} = (S_1,...,S_{|\ol{\alpha}|}), \mbox{ and }$
-$S_i \mbox{ is a finite set for } i =
-1,...,|\ol{\alpha}|\big\}$. There are countably many choices of tuples
-$\ol{S}$, each of which gives rise to a morphism from
-${\setsem{\Gamma;\ol{\alpha} \vdash F}\rho[ \ol{\alpha := S}]}$ to
-${\setsem{\Gamma;\ol{\alpha} \vdash G}\rho[\ol{\alpha := S}]}$, but
-only $\set$-many choices of morphisms between any two objects since
-$\set$ is locally small. Also, $\setsem{\Gamma; \emptyset \vdash
-  \Nat^{\ol\alpha} F\,G}$ is $\omega$-cocontinuous since it is
-constant on $\omega$-directed sets. Interpretations of $\Nat$-types
-ensure that $\setsem{\Gamma \vdash F \to G}$ and $\setsem{\Gamma
-  \vdash \forall \ol\alpha. F}$ are as expected in any parametric
-model.
+ensure that $\{\eta : \lambda \ol{A}.\,\setsem{\Gamma;\ol{\alpha}
+  \vdash F}\rho[\ol{\alpha := A}] \Rightarrow \lambda
+\ol{A}.\,\setsem{\Gamma;\ol{\alpha} \vdash G}\rho[\ol{\alpha := A}]\}$
+(containing $\setsem{\Gamma;\emptyset \vdash
+  \Nat^{\ol{\alpha}}\,F\,G}\rho$) can be embedded in the set
+\[ \prod_{\substack{\ol{S} =
+    (S_1,...,S_{|\ol{\alpha}|})\\ S_1,...,S_{|\ol{\alpha}|} \mbox{ are
+      finite cardinals}}} \hspace*{-0.4in}(\setsem{\Gamma;\ol{\alpha}
+  \vdash G}\rho[\ol{\alpha := S}])^{(\setsem{\Gamma;\ol{\alpha} \vdash
+    F}\rho[\ol{\alpha := S}])}\] Also, $\setsem{\Gamma; \emptyset
+  \vdash \Nat^{\ol\alpha} F\,G}$ is $\omega$-cocontinuous since it is
+constant on $\omega$-directed sets; indeed, it is constant on
+$\omega$-directed sets because Definition~\ref{def:set-env} ensures
+that restrictions to $\mathbb{T}^k$ of morphisms between set
+environments are identities.  Interpretations of $\Nat$-types ensure
+that $\setsem{\Gamma \vdash F \to G}$ and $\setsem{\Gamma \vdash
+  \forall \ol\alpha. F}$ are as expected in any parametric model.
 
@@
   \phi,\ol{\alpha} \vdash H}f[\phi := \id_F][\ol{\alpha := \id_A}]$.
+Note that this action of $T_{H,\rho}^\set$ on morphisms appears in the
+final clause of Definition~\ref{def:set-sem-funcs} below. It will
+indeed be well-defined there because the functorial action of
+$\setsem{\Gamma; \phi,\ol{\alpha} \vdash H}$ will already be available
+from the induction hypothesis on $H$.
 
@@
 \item An object of $RT_k$ is a relation transformer
-\item A morphism $\delta : (G^1,G^2,G^*) \to (H^1,H^2,H^*)$ in $RT_k$
+\item A morphism $\delta : (F^1,F^2,F^*) \to (G^1,G^2,G^*)$ in $RT_k$
   is a pair of natural transformations $(\delta^1, \delta^2)$, where
-  $\delta^1 : G^1 \to H^1$ and $\delta^2 : G^2 \to H^2$, such that,
-  for all $\ol{R : \rel(A, B)}$, if $(x, y) \in G^*\ol{R}$ then
-  $(\delta^1_{\ol{A}}x, \delta^2_{\ol{B}}y) \in H^*\ol{R}$
+  $\delta^1 : F^1 \to G^1$ and $\delta^2 : F^2 \to G^2$, such that,
+  for all $\ol{R : \rel(A, B)}$, if $(x, y) \in F^*\ol{R}$ then
+  $(\delta^1_{\ol{A}}x, \delta^2_{\ol{B}}y) \in G^*\ol{R}$
 \item Identity morphisms and composition are inherited from the
@@
   \Homrel(R_1,S_1),..., (\alpha_k, \beta_k)$ $\in \Homrel(R_k,S_k)$
-\item If $(\delta^1,\delta^2) : (F^1,F^2,F^*)\to (G^1,G^2,G^*)$ and
-  $R_1:\rel(A_1,B_1),...,R_k:\rel(A_k,B_k)$, then
+\item If $(\delta^1,\delta^2) : (F^1,F^2,F^*)\to (G^1,G^2,G^*)$ in
+  $RT_k$ and $R_1:\rel(A_1,B_1),...,R_k:\rel(A_k,B_k)$, then
   $((H^1\delta^1)_{\ol{A}}x, (H^2\delta^2)_{\ol{B}}y) \in
@@
 \begin{defi}\label{def:relenv-functor}
-For each $k$, an $\omega$-cocontinuous functor $H : [\relenv, RT_k]$
+For each $k$, an $\omega$-cocontinuous functor $H \in [\relenv, RT_k]$
 is a triple $H = (H^1,H^2,H^*)$, where
@@
 \item $H^1$ and $H^2$ are objects in $[\setenv,[\set^k,\set]]$
-\item $H^*$ is a an object in $[\relenv,[\rel^k,\rel]]$
+\item $H^*$ is an object in $[\relenv,[\rel^k,\rel]]$
 \item For all $\overline{R : \rel(A,B)}$ and morphisms $f$ in
@@
 
-Considering $\relenv$ as a product $\Pi_{\phi^k \in \tvars \cup
-  \fvars} RT_k$, computation of $\omega$-directed colimits in $RT_k$
-extends componentwise to $\relenv$. Similarly, $\omega$-cocontinuity
-for endofunctors on $RT_k$ extends componentwise to functors from
-$\relenv$ to $RT_k$.  Recalling from the start of this subsection that
-Definition~\ref{def:rel-sem} is given mutually inductively with
-Definition~\ref{def:set-sem} we can now define our relational
-interpretations.
+Considering $\relenv$ as a subcategory of the product $\Pi_{\phi^k \in
+  \tvars \cup \fvars} RT_k$, computation of $\omega$-directed colimits
+in $RT_k$ extends componentwise to $\relenv$. Similarly,
+$\omega$-cocontinuity for endofunctors on $RT_k$ extends componentwise
+to functors from $\relenv$ to $RT_k$.  Recalling from the start of
+this subsection that Definition~\ref{def:rel-sem} is given mutually
+inductively with Definition~\ref{def:set-sem} we can now define our
+relational interpretations.
 
@@
 for set interpretations, $\relsem{\Gamma; \emptyset \vdash
-  \Nat^{\ol\alpha} F\,G}$ is $\omega$-cocontinuous because it is
-constant on $\omega$-directed sets. If $\rho \in \relenv$ and $\vdash
-F$, then we write $\relsem{\vdash F}$ instead of $\relsem{\vdash
-  F}\rho$.  For the last clause in Definition~\ref{def:rel-sem} to be
-well-defined we need $T_{H,\rho}$ to be an $\omega$-cocontinuous
-endofunctor on $RT$ so that, by Lemma~\ref{lem:fp}, it admits a
-fixpoint. Since $T_{H,\rho}$ is defined in terms of
-$\relsem{\Gamma;\phi, \ol{\alpha} \vdash H}$, this means
-that relational interpretations of types must be $\omega$-cocontinuous
-functors from $\relenv$ to $RT_0$, which in turn entails that the
-actions of relational interpretations of types on objects and on
-morphisms in $\relenv$ are intertwined. As for set interpretations, we
-know from~\cite{jp19} that, for every $\Gamma; \ol{\alpha} \vdash F$,
-$\relsem{\Gamma; \ol{\alpha} \vdash F}$ is actually in $[\rel^k,\rel]$
-where $k = |\ol \alpha|$. We first define the actions of each of these
-functors on morphisms between environments, and then argue that they
-are well-defined and have the required properties. To do this, we
-extend $T_H$ to a {\em functor} from $\relenv$ to
-$[[\rel^k,\rel],[\rel^k,\rel]]$. Its action on an object $\rho \in
-\relenv$ is given by the higher-order functor $T_{H,\rho}$ whose
-actions on objects and morphisms are given in
+  \Nat^{\ol\alpha} F\,G}$ is $\omega$-cocontinuous; indeed, it is
+constant on $\omega$-directed sets because
+Definition~\ref{def:reln-env} ensures that restrictions to
+$\mathbb{T}^k$ of morphisms between relational environments are
+identities. If $\rho \in \relenv$ and $\vdash F$, then we write
+$\relsem{\vdash F}$ instead of $\relsem{\vdash F}\rho$.  For the last
+clause in Definition~\ref{def:rel-sem} to be well-defined we need
+$T_{H,\rho}$ to be an $\omega$-cocontinuous endofunctor on $RT$ so
+that, by Lemma~\ref{lem:fp}, it admits a fixpoint. Since $T_{H,\rho}$
+is defined in terms of $\relsem{\Gamma;\phi, \ol{\alpha} \vdash H}$,
+this means that relational interpretations of types must be
+$\omega$-cocontinuous functors from $\relenv$ to $RT_0$, which in turn
+entails that the actions of relational interpretations of types on
+objects and on morphisms in $\relenv$ are intertwined. As for set
+interpretations, we know from~\cite{jp19} that, for every $\Gamma;
+\ol{\alpha} \vdash F$, $\relsem{\Gamma; \ol{\alpha} \vdash F}$ is
+actually in $[\rel^k,\rel]$ where $k = |\ol \alpha|$. We first define
+the actions of each of these functors on morphisms between
+environments, and then argue that they are well-defined and have the
+required properties. To do this, we extend $T_H$ to a {\em functor}
+from $\relenv$ to $[[\rel^k,\rel],[\rel^k,\rel]]$. Its action on an
+object $\rho \in \relenv$ is given by the higher-order functor
+$T_{H,\rho}$ whose actions on objects and morphisms are given in
 Definition~\ref{def:rel-sem-funcs}. Its action on a morphism $f : \rho
@@
 T_{H,\rho'}\, F$ whose component at $\ol{R}$ is $(T_{H,f}\,
-F)_{\ol{R}} = \relsem{\Gamma; \phi,\ol{\alpha} \vdash
-  H}f[\phi := \id_F][\ol{\alpha := \id_R}]$.
+F)_{\ol{R}} = \relsem{\Gamma; \phi,\ol{\alpha} \vdash H}f[\phi :=
+  \id_F][\ol{\alpha := \id_R}]$.
 
@@
   $\rel$ for all ${R_1 : \rel(A_1,B_1)},...,{R_k : \rel(A_k,B_k)}$ if
-  and only if $t = t'$ and $(t_{\overline{A}}, t_{\overline{B}})$ is a
+  and only if $t = t'$.  We first show that if $(t_{\overline{A}},
+  t'_{\overline{B}})$ is a morphism from $\relsem{\Gamma;
+    \overline\alpha \vdash F} \Eq_{\rho}\overline{[\alpha := R]}$ to
+  $\relsem{\Gamma ; \overline\alpha \vdash G}
+  \Eq_{\rho}\overline{[\alpha := R]}$ for all ${R_1 : \rel(A_1,B_1),}$
+  $...,{R_k : \rel(A_k,B_k)}$ then $t = t'$.  By hypothesis, for all
+  $A_1\,...\,A_k : \set$, $(t_{\overline{A}}, t'_{\overline{A}})$ is a
   morphism from $\relsem{\Gamma; \overline\alpha \vdash F}
-  \Eq_{\rho}\overline{[\alpha := R]}$ to $\relsem{\Gamma ; \overline
-    \alpha \vdash G} \Eq_{\rho}\overline{[\alpha := R]}$ in $\rel$ for
-  all ${R_1 : \rel(A_1,B_1)}, ...,$ ${R_k : \rel(A_k,B_k)}$. The only
-  interesting part of this equivalence is to show that if
-  $(t_{\overline{A}}, t'_{\overline{B}})$ is a morphism from
-  $\relsem{\Gamma; \overline\alpha \vdash F}
-  \Eq_{\rho}\overline{[\alpha := R]}$ to $\relsem{\Gamma ;
-    \overline\alpha \vdash G} \Eq_{\rho}\overline{[\alpha := R]}$ for
-  all ${R_1 : \rel(A_1,B_1),}$ $...,{R_k : \rel(A_k,B_k)}$ then $t =
-  t'$.  By hypothesis, for all $A_1\,...\,A_k : \set$,
-  $(t_{\overline{A}}, t'_{\overline{A}})$ is a morphism from
-  $\relsem{\Gamma; \overline\alpha \vdash F}
   \Eq_{\rho}\overline{[\alpha := \Eq_{A}]}$ to $\relsem{\Gamma ;
@@
   $t_{\overline{A}}x = t'_{\overline{A}}x$, so $t = t'$ by
-  extensionality.
+  extensionality. For the converse, we observe that if $t = t'$ then
+  the desired conclusion is simply the extra condition in the
+  definition of $\setsem{\Gamma;\emptyset \vdash
+    \Nat^{\overline\alpha} \,F\,G} \rho$.
+
 \item The application case is proved by the following sequence of
@@
 every $n \in \nat$, the following two statements can be proved by
-simultaneous induction: and for any $H$, $\rho$, $A$, and subformula
-$J$ of $H$,
+simultaneous induction: for any $H$, $\rho$, and $A$, 
 \begin{equation}\label{eq:iel-fix-point-intermediate1}
@@
 \end{equation}
-and 
+and for any subformula $J$ of $H$,
 \begin{equation}\label{eq:iel-fix-point-intermediate2}
@@
 Equation~\ref{eq:helper} on $m$. The fourth equality holds because
-$\phi$ and the variables in $\ol\alpha$ do not appear in
-$G$. 
+$\phi$ and the variables in $\ol\alpha$ do not appear in $G$. (We note
+that this case of the IEL would fail to hold if the functorial context
+in which the body $F$ of a $\mu$-type $\mu \phi. \lambda
+\ol{\alpha}. F$ were formed in Definition~\ref{def:wftypes} were
+extended to contain variables other than $\phi$ and the $\alpha$s.
+This justifies our design of the $\mu$-type formation rule there.)
 \end{itemize}
+This concludes the fixpoint case.
 \item $\relsem{\Gamma; \Phi \vdash F + G} \Eq_{\rho} =
@@
   (L_{\ol{\alpha}}x.t)_{\ol{K}}s} = \dsem{\Gamma; \Phi~|~\Delta \vdash
-  t [\ol{\alpha := K}][x := s]}\] Term extensionality for both types
-and terms --- i.e., \[\dsem{\Gamma;\Phi~|~\Delta\vdash (L_\alpha x.t)_\alpha
-  \top : F} = \dsem{\Gamma;\Phi~|~\Delta \vdash t : F}\] and
-\[\dsem{\Gamma;\Phi~|~\Delta\vdash (L_\alpha x.t)_\alpha x : F} =
-\dsem{\Gamma;\Phi~|~\Delta \vdash t : F}\] --- then follow (assuming
-both sides of these equations are defined).
+  t [\ol{\alpha := K}][x := s]}\]
+       {\color{red} Add if true:
+$\dsem{\Gamma; \emptyset~|~\Delta \vdash L_{\ol{\alpha}} x.t_{\ol{\alpha}} x :
+   \Nat^{\ol{\alpha}}\, F\, G} =  \dsem{\Gamma; \emptyset~|~\Delta \vdash
+   t : \Nat^{\ol{\alpha}}\, F\, G}$}
+
+%Term extensionality for both types
+%and terms --- i.e., \[\dsem{\Gamma;\Phi~|~\Delta\vdash (L_\alpha x.t)_\alpha
+%  \top : F} = \dsem{\Gamma;\Phi~|~\Delta \vdash t : F}\] and
+%\[\dsem{\Gamma;\Phi~|~\Delta\vdash (L_\alpha x.t)_\alpha x : F} =
+%\dsem{\Gamma;\Phi~|~\Delta \vdash t : F}\] --- then follow (assuming
+%both sides of these equations are defined).
 
@@
   \emptyset \vdash F}$, and $\ol{\Gamma; \emptyset \vdash G}$, then,
-for all $\ol{f : \setsem{\Gamma;\emptyset \vdash \Nat^\emptyset
+for all $\ol{f \in \setsem{\Gamma;\emptyset \vdash \Nat^\emptyset
     F\,G}\rho}$,
@@
 $\lambda \ol{A}.\,\setsem{\Gamma; \ol{\alpha}\vdash H} \rho
-[\ol{\alpha := A}]$ is a functor in $\alpha$ and using
+[\ol{\alpha := A}]$ is a functor in $\ol{\alpha}$ and using
 $\mathit{map}_G\,\ol f$ to denote the action of the semantic functor
@@
 \noindent
-$\bullet$\; $\fold$ satisfies its universal property, i.e., if $\xi =
+$\bullet$\; $\fold$ behaves as expected, i.e., if $\xi =
 \Nat^{\overline{\beta}}\, H[\phi :=

### 2021-06-25 21-lmcs/21-lmcs.tex
third LMCS revisions

@@
 
-{\small
 \[\begin{array}{l}
@@
 \hspace*{0.4in}\mathtt{cons\;:\;A \rightarrow List\;A \rightarrow List\;A}
-\end{array}\]}
+\end{array}\]
 
@@
 properties that can be used for program verification and other
-applications.  A simple example of such a type is given by Bird and
-Meertens'~\cite{bm98} prototypical nested type
+applications.  A simple example of such a type is given by the type 
 \[\begin{array}{l}
@@
 of perfect trees, which can be thought of as constraining lists to
-have lengths that are powers of 2.  The above code makes clear that
-perfect trees at index type $\mathtt{A}$ are defined in terms of
-perfect trees at index type $\mathtt{A \times A}$. This is typical of
-nested types, one type instance of which can depend on others, so that
-the entire family of types must actually be defined at once. A nested
-type thus defines not a family of inductive types, but rather an {\em
-  inductive family of types}.  Nested types include simple nested
-types, like perfect trees, none of whose recursive occurrences occur
-below another type constructor; ``deep'' nested types~\cite{jp20},
-such as the nested type {\small
+have lengths that are powers of 2.  A similar data type
+$\mathtt{Nest}$ is given in the canonical paper~\cite{bm98} on nested
+types. The above code makes clear that perfect trees at index type
+$\mathtt{A}$ are defined in terms of perfect trees at index type
+$\mathtt{A \times A}$. This is typical of nested types, one type
+instance of which can depend on others, so that the entire family of
+types must actually be defined at once. A nested type thus defines not
+a family of inductive types, but rather an {\em inductive family of
+  types}.  Nested types include simple nested types, like perfect
+trees, none of whose recursive occurrences occur below another type
+constructor; ``deep'' nested types~\cite{jp20}, such as the nested
+type
 \[\begin{array}{l}
@@
 PForest\;A}
-\end{array}\]}
+\end{array}\]
 \hspace{-0.04in}of perfect forests, whose recursive occurrences appear
@@
   \rightarrow Bush\;A} 
-\end{array}\]}
+\end{array}\]
 \hspace{-0.04in}of bushes, whose recursive occurrences appear below
@@
 verifying in the just-constructed parametric model that each Church
-encoding is isomorphic to the data type it encodes.  The present paper
-draws on this rich tradition of categorical models of parametricity
-for System F, but modifies them to treat nested types (and therefore
-ADTs) as primitive data types.
+encoding is a solution for the fixpoint equation associated with its
+corresponding data type. The present paper draws on this rich
+tradition of categorical models of parametricity for System F, but
+modifies them to treat nested types (and therefore ADTs) as primitive
+data types.
 
@@
 of arity $k$, or $F$ is a subexpression of the form $\mu
-\phi^{k}.\lambda \alpha_1...\alpha_k.F'$. Moreover, if $F$ has arity
-$k$ then $F$ must be applied to exactly $k$ arguments.  Accordingly,
-an overbar indicates a sequence of types whose length matches the
-arity of the type applied to it.  Requiring that types are always in
-such \emph{$\eta$-long normal form} avoids having to consider
+\phi^{k}.\lambda \alpha_1...\alpha_k.F'$. Accordingly, an overbar
+indicates a sequence of types whose length matches the arity of the
+type applied to it.  Requiring that types are always in such
+\emph{$\eta$-long normal form} avoids having to consider
 $\beta$-conversion at the level of types. In a type
@@
 calculus. Note that, in the rule for $\mu$-types, no variables in
-$\Phi$ are applied to types involving variables from $\ol\alpha$ in
-the body $F$ of $\mu \phi. \lambda \ol\alpha.F$. This will be critical
-to proving the Identity Extension Lemma for our calculus.
+$\Phi$ appear in the body $F$ of $\mu \phi. \lambda \ol\alpha.F$. This
+will be critical to proving the Identity Extension Lemma for our
+calculus.
 
@@
 \mathit{List}\, \alpha & = & \mu \beta. \,\onet + \alpha \times
-\beta\; = \; (\mu \phi. \lambda \beta.\,\onet + \beta \times \phi
+\beta\; \mbox{ or } \; (\mu \phi. \lambda \beta.\,\onet + \beta \times \phi
 \beta)\,\alpha\\ 
@@
   second-order}. Substitution for first-order types is the usual
-capture-avoiding textual substitution. We write $F[\alpha := \sigma]$
-for the result of substituting $\sigma$ for $\alpha$ in $F$, and
-$F[\alpha_1 := F_1,...,\alpha_k := F_k]$, or $F[\ol{\alpha := F}]$
-when convenient, for $F[\alpha_1 := F_1][\alpha_2 := F_2,...,\alpha_k
+capture-avoiding textual substitution. We write $H[\alpha := F]$
+for the result of substituting $F$ for $\alpha$ in $H$, and
+$H[\alpha_1 := F_1,...,\alpha_k := F_k]$, or $H[\ol{\alpha := F}]$
+when convenient, for $H[\alpha_1 := F_1][\alpha_2 := F_2,...,\alpha_k
   := F_k]$. The operation of {\em second-order type substitution along
@@
   then $\Gamma,\ol\psi';\Phi \vdash H[\phi :=_{\ol{\alpha}} F[\ol{\psi
-        :== \psi'}]]$.}
+        :== \psi'}]]$. Ask Daniel.}
 
@@
 maps, e.g., the perfect tree
-\[\mathit{pnode}\,(\mathit{pnode}\,
-(\mathit{pleaf}\,1,\mathit{pleaf}\,2))\,(\mathit{pnode}\,
-(\mathit{pleaf}\,3,\mathit{pleaf}\,4) )\]
+\[\mathit{pnode}\,(\mathit{pnode}\,(\mathit{pleaf}\,
+((1, 2), (3, 4))))\]
 to the perfect tree
-\[\mathit{pnode}\,(\mathit{pnode}\,
-(\mathit{pleaf}\,4,\mathit{pleaf}\,3))\,(\mathit{pnode}\,
-(\mathit{pleaf}\,2,\mathit{pleaf}\,1) )\] i.e., maps $((1,2),(3,4))$
+\[\mathit{pnode}\,(\mathit{pnode}\,(\mathit{pleaf}\,
+((4, 3), (2, 1))))\]
+i.e., maps $((1,2),(3,4))$
 to $((4,3),(2,1))$.  It can be defined as
@@
 function maps, e.g., the bush {\color{red} Change one 2 to 4? Find a
-  better name for this operation than reverse.}
+  better name for this operation than reverse. Ask Daniel.}
 \[\begin{array}{l}
@@
 \onet \,\alpha$, can exist.  That is, our calculus admits no
-non-terminating terms, i.e., terms with the closed type $\Nat^\alpha
-\onet \,\alpha$ of the polymorphic bottom.
+(non-terminating) terms with the closed type $\Nat^\alpha \onet
+\,\alpha$ of the polymorphic bottom.
 
@@
 \id)\}$ That is, $G^\set$ is a natural transformation from the
-identity functor on $\set$ to itself. Now let $S$ be any set.  If $S =
-\emptyset$, then there is exactly one morphism $\id_S: S \to S$, so
-$G^\set_S : S \to S$ must be $\id_S$. If $S \not = \emptyset$, if $a$
+identity functor on $\set$ to itself. Now let $S$ be any set.  If $a$
 is any element of $S$, and if $K_a :S \to S$ is the constantly
@@
 \vdash C$ generalizes Theorem~\ref{thm:short-cut-nested} to deliver a
-more general a free theorem whose conclusion is $\mathit{fold}_{H}\, B
+more general free theorem whose conclusion is $\mathit{fold}_{H}\, B
 \; \circ \; G\; \mu H \; \mathit{in}_{H} = G
@@
 types of data constructors to be different instances of the datatype
-than the one being defined. For example, the GADT {\small
+than the one being defined. For example, the GADT 
 \[\begin{array}{l}
@@
   Seq\;(Nat \to A)}
-\end{array}\]}
+\end{array}\]
 \noindent
@@
 $\mathtt{Seq}$ to arrive at the following equivalent representation:
-{\small
 \[\begin{array}{l}
@@
   Seq\;A}
-\end{array}\]}
+\end{array}\]
 \noindent
@@
 
+\noindent
+Of course, we could have performed this replacement from the outset of
+the present paper. But since using the above rule instead of the one
+currently in Definition~\ref{def:wftypes} only makes the notation of
+our calculus heavier without increasing its expressivity, we have
+chosen not to.
+
 Finally, adding $\Lan$-types to the types of our calculus also
@@
 respectively. Note that both of these rules make essential use of our
-generalized $\Nat$-types.
+generalized $\Nat$-types. If $\Phi$ were required to be empty then we
+would not be able to express, e.g., the types
+\[\Nat^{\phi, \gamma,  \delta} (\phi \gamma \times
+\phi \delta) \; (\phi (\gamma \times \phi))\] and
+\[\Nat^{\phi, \beta}\,(\Lan^{\gamma,\delta}_{\gamma \times
+  \delta} \,(\phi \gamma \times \phi \delta))\beta \; (\phi \beta)\]
+associated with $\mathtt{Seq}$'s data constructor $\mathtt{spair}$.
 
@@
 \]
-where $\iota$ is the morphism mapping the cocone into the colimit.
-Since each projection $\pi_i$ is cocontinuous, we therefore have that 
+with $\iota$ mapping the cocone into this colimit.  Since each
+projection $\pi_i$ is cocontinuous, we therefore have that
 \[
@@
 \]
-with $\pi_i \iota$ mapping this cocone into this colimit.  On the
+with $\pi_i \iota$ mapping the cocone into this colimit.  On the
 other hand, Equation~\ref{eq:colim-form} also gives
@@
 \]
-where $j$ is the morphism mapping the cocone into the colimit.
+with $j$ mapping the cocone into this colimit.
 
@@
 modify the categories of relations and relation transformers so that
-these are always surjective on their domans and codomains; this would
-prohibit situations like that in Example~\ref{ex:no-rel-transf}, and
-might therefore allow us to recover our IEL and, ultimately, an
-Abstraction Theorem appropriate to GADTs.  If this turns out to be
-possible, then generalizing the resulting model construction to
-locally $\lambda$-presentable categories for $\lambda > \omega$ would
-make it possible to handle broader classes of GADTs. (As shown
-in~\cite{jp19}, $\lambda > \omega_1$ is required to interpret even
-common GADTs.) We could even attempt to carry out our construction in
-locally $\lambda$-presentable cartesian closed categories (lpcccs)
-$\cal C$ whose categories of (abstract) relations, obtained by
-pullback as in~\cite{jac99}, are also lpcccs and are appropriately
-fibred over $\cal C$. This would give a framework for constructing
-parametric models for calculi with primitive GADTs that is based on
-locally $\lambda$-presentable fibrations, for some appropriate
-definition thereof.
+if $(A,B,R)$ is a relation then $\pi_1 : R \to A$ and $\pi_2 : R \to
+B$ are always surjective; this would prohibit situations like that in
+Example~\ref{ex:no-rel-transf}, and might therefore allow us to
+recover our IEL and, ultimately, an Abstraction Theorem appropriate to
+GADTs.  If this turns out to be possible, then generalizing the
+resulting model construction to locally $\lambda$-presentable
+categories for $\lambda > \omega$ would make it possible to handle
+broader classes of GADTs. (As shown in~\cite{jp19}, $\lambda >
+\omega_1$ is required to interpret even common GADTs.) We could even
+attempt to carry out our construction in locally $\lambda$-presentable
+cartesian closed categories (lpcccs) $\cal C$ whose categories of
+(abstract) relations, obtained by pullback as in~\cite{jac99}, are
+also lpcccs and are appropriately fibred over $\cal C$. This would
+give a framework for constructing parametric models for calculi with
+primitive GADTs that is based on locally $\lambda$-presentable
+fibrations, for some appropriate definition thereof.
 

### 2021-06-28 21-lmcs/21-lmcs.tex
4th  revisions of LMCS paper

@@
 reason about nested types in System F. Indeed, such calculi support
-fundamentally new features that add complexity to their models that is
-entirely unnecessary for reasoning about nested types. This paper
-therefore pursues the second option above.  We first design a
-Hindley-Milner-style calculus supporting primitive nested types,
+fundamentally new features --- e.g., the full hierarchy of
+higher-kinds or term-indexed types --- adding complexity to their
+models that is entirely unnecessary for reasoning about nested types.
+In addition, to our knowledge no such models represent data types as
+primitive fixpoints. (See the past paragraph of this section for a
+more detailed discussion in the case of F$_\omega$.) Whether or not
+the results of this paper even hold for such calculi is an open
+question.
+
+This paper therefore pursues the second option above.  We first design
+a Hindley-Milner-style calculus supporting primitive nested types,
 together with primitive types of natural transformations representing
@@
 
-We can now define our set interpretations.  If $\rho$ is a set
-environment we write $\Eq_\rho$ for the equality relation environment
-such that $\Eq_\rho \phi = \Eq_{\rho \phi}$ for every type variable
-$\phi$; see Definitions~\ref{def:rel-transf} and~\ref{def:reln-env}
-for the definitions of a relation transformer and a relation
-environment, and Equation~\ref{def:eq-transf} for the definition of
-the relation transformer $\Eq_F$ on a functor $F$.  Equality relation
-environments appear in the third clause of
-Definition~\ref{def:set-sem}.  The relational interpretations also
-appearing in the third clause of Definition~\ref{def:set-sem} are
+We can now define our set interpretation. Its action on objects of
+$\setenv$ is given in Definition~\ref{def:set-sem}, and its action on
+morphisms of $\setenv$ is given in Definition~\ref{def:set-sem-funcs}.
+If $\rho$ is a set environment we write $\Eq_\rho$ for the equality
+relation environment such that $\Eq_\rho \phi = \Eq_{\rho \phi}$ for
+every type variable $\phi$; see Definitions~\ref{def:rel-transf}
+and~\ref{def:reln-env} for the definitions of a relation transformer
+and a relation environment, and Equation~\ref{def:eq-transf} for the
+definition of the relation transformer $\Eq_F$ on a functor $F$.
+Equality relation environments appear in the third clause of
+Definition~\ref{def:set-sem}.  The relational interpretation also
+appearing in the third clause of Definition~\ref{def:set-sem} is
 given in Definition~\ref{def:rel-sem}.
@@
 indeed define a set: local finite presentability of $\set$ and
-$\omega$-cocontinuity of $\setsem{\Gamma;\ol{\alpha} \vdash F}\rho$
+$\omega$-cocontinuity of $\setsem{\Gamma;\ol{\alpha} \vdash F}$
 ensure that $\{\eta : \lambda \ol{A}.\,\setsem{\Gamma;\ol{\alpha}
@@
   defined by
-  \[\begin{array}{ll} & (\mu T^\set_{H,f})\ol{\setsem{\Gamma;\Phi \vdash
-      G}\rho'} \circ (\mu T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi \vdash G}f}\\ = & (\mu T^\set_{H,\rho'})\ol{\setsem{\Gamma;\Phi
-      \vdash G}f} \circ (\mu T^\set_{H,f})\ol{\setsem{\Gamma;\Phi
-      \vdash G}\rho}\end{array}\]  The latter equality holds because $\mu
+  \[\begin{array}{ll} & (\mu T^\set_{H,f})_{\ol{\setsem{\Gamma;\Phi \vdash
+      G}\rho'}} \circ (\mu T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi
+      \vdash G}f}\\ = & (\mu T^\set_{H,\rho'})\ol{\setsem{\Gamma;\Phi  
+      \vdash G}f} \circ (\mu T^\set_{H,f})_{\ol{\setsem{\Gamma;\Phi
+      \vdash G}\rho}}\end{array}\]  The latter equality holds because $\mu
   T^\set_{H,\rho}$ and $\mu T^\set_{H,\rho'}$ are functors and $\mu
@@
 inductively with Definition~\ref{def:set-sem} we can now define our
-relational interpretations.
+relational interpretation.
 
@@
 relation transformer $F$ and every morphism $f : \rho \to \rho'$ in
-$\relenv$. This is an immediate consequence of
+$\relenv$. If we define $\sem{\Gamma;\Phi \vdash F}$ to be
+$(\setsem{\Gamma;\Phi \vdash F}, \setsem{\Gamma;\Phi \vdash
+  F},\relsem{\Gamma;\Phi \vdash F})$, then this is an immediate
+consequence of
 \begin{lem}\label{lem:rel-transf-morph}
-For every $\Gamma;\Phi \vdash F$,
-\[\sem{\Gamma;\Phi \vdash F} = (\setsem{\Gamma;\Phi \vdash F},
-\setsem{\Gamma;\Phi \vdash F},\relsem{\Gamma;\Phi \vdash F})
-\in[\relenv,RT_0]\] 
+For every $\Gamma;\Phi \vdash F$, $\sem{\Gamma;\Phi \vdash F}
+\in[\relenv,RT_0]$.
 \end{lem}
@@
 \end{figure*}
-
-\vspace*{-0.1in}
 
@@
 \begin{figure*}
-\hspace*{-3.1in}
-\resizebox{0.46\linewidth}{!}{
-\begin{minipage}[t]{0.5\textwidth}
+\begin{adjustbox}{varwidth=7.2in, max width=5.8in, fbox, center}
+%  \hspace*{-3.1in}
+%\resizebox{0.46\linewidth}{!}{
+%\begin{minipage}[t]{0.5\textwidth}
 \[\begin{array}{lll}
@@
 \dsem{\Gamma;\Phi \,|\, \Delta \vdash (s,t) : F \times G} \rho& = &
-\dsem{\Gamma;\Phi \,|\, \Delta \vdash s: F} \rho\times
-\dsem{\Gamma;\Phi \,|\, \Delta \vdash t : G} \rho\\
+\langle \dsem{\Gamma;\Phi \,|\, \Delta \vdash s: F} \rho,\,
+\dsem{\Gamma;\Phi \,|\, \Delta \vdash t : G} \rho\rangle\\
 \dsem{\Gamma;\Phi \,|\, \Delta \vdash \pi_1 t : F} \rho& = &
@@
 \caption{Term semantics}\label{fig:term-sem} 
-\end{minipage}}\vspace*{-0.05in}
+%\end{minipage}}
+\end{adjustbox}\vspace*{-0.05in}
 \end{figure*}
@@
 \[\begin{array}{lll}
-\dsem{\Gamma;\Phi~|~\Delta \vdash t : F}\rho & : & \dsem{\Gamma;
-  \Phi \vdash \Delta}\rho \to \dsem{\Gamma; \Phi \vdash F}\rho
+\setsem{\Gamma;\Phi~|~\Delta \vdash t : F}\rho & : & \setsem{\Gamma;
+  \Phi \vdash \Delta}\rho \to \setsem{\Gamma; \Phi \vdash F}\rho
 \end{array}\]
@@
       \setsem{\Gamma; \Phi \,|\, \Delta \vdash t : F } (\pi_2 \rho) \,b) \in 
-    \relsem{\Gamma; \Phi \vdash F} \rho$
+    \relsem{\Gamma; \Phi \vdash F} \rho$.
 \end{thm}
@@
 \id)\}$ That is, $G^\set$ is a natural transformation from the
-identity functor on $\set$ to itself. Now let $S$ be any set.  If $a$
-is any element of $S$, and if $K_a :S \to S$ is the constantly
-$a$-valued morphism on $S$, then instantiating the naturality square
-implied by the above equality gives that $G^\set_S \circ K_a = K_a
-\circ G^\set_S$, i.e., $G^\set_S \, a = a$, i.e., $G^\set_S = \id_S$.
+identity functor on $\set$ to itself. Now let $S$ be any set.  For any
+element $a$ of $S$, if $K_a :S \to S$ is the constantly $a$-valued
+morphism on $S$ then instantiating the naturality square implied by
+the above equality gives that $G^\set_S \circ K_a = K_a \circ
+G^\set_S$, i.e., $G^\set_S \, a = a$, i.e., $G^\set_S = \id_S$.
 Putting these two cases together we have that for every $S : \set$,
@@
 An alternative presentation characterizes the left Kan extension
-$(\mathit{Lan}_{\ol K} F, \eta)$ in terms of the adjunction between
+$(\mathit{Lan}_{\ol K} F, \eta)$ in terms of the bijection between
 natural transformations from $F$ to $G \circ \ol{K}$ and natural
@@
 $\eta$ is the unit. If Agda were to support a primitive
-$\mathsf{Lan}$ for left Kan extensions, we could use this adjunction
+$\mathsf{Lan}$ for left Kan extensions, we could use this bijection
 to rewrite the type of each data constructor for the GADT
@@
 using just unary $\Lan$-types, i.e, $\Lan$-types of the form
-$(\Lan^{\ol{\alpha^0}}_K F)\ol A$.
+$(\Lan^{\ol{\alpha^0}}_K F) A$.
 
@@
 \[\begin{array}{l}
-\mu_{\ol{N}\, d} : \lambda \ol{A}.\,
+\mu_{\ol{N}\, d} : 
 \mathit{Lan}_{\ol{\lambda \ol A.\,\setsem{\Gamma; \ol\alpha \vdash K}\rho 
@@
   \ol\alpha \vdash F}\rho {[\Phi, \ol \alpha := \ol N, \ol A ]}\\[2ex]
-\hspace*{0.8in}\to \lambda \ol B.\, \setsem{\Gamma; 
+\hspace*{0.8in} \to \lambda \ol B.\, \setsem{\Gamma; 
   \Phi, \ol\beta \vdash G} \rho[\Phi, \ol \beta := \ol N, \ol B]
@@
       \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha := R}] \to
-    \relsem{ \Gamma; \Phi \vdash A } \rho}}{ \relsem{ \Gamma; \Phi,
+    \relsem{ \Gamma; \Phi \vdash A } \rho}}{ \setsem{ \Gamma; \Phi,
     \ol\alpha \vdash F } (\pi_i \rho) [\ol{\alpha := \pi_i R}]}
@@
 colimit representation of $h$'s domain. That is, we take $h$ to be the
-unique morphism such that, for any $\ol{R : \rel_0}$ and $\ol{m :
+unique morphism such that, for any $\ol{R : \rel_0}$ and
+\[\ol{m :
   \relsem{ \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha := R}] \to
-  \relsem{ \Gamma; \Phi \vdash A } \rho}$, we have
+  \relsem{ \Gamma; \Phi \vdash A } \rho}\] we have
 \begin{equation}\label{eq:cocone-proj}

### 2021-06-29 21-lmcs/21-lmcs.tex
5th LMCS revisions

@@
 problem), which would allow the first-order representations of ADTs to
-be functorial. However, doing this requires some changes to the
+be functorial. However, since doing this requires some changes to the
 formation rule for $\mu$-types, as well as the delicate threading of
-some additional conditions throughout our model construction.  But
-since we can always use an ADT's (semantically equivalent)
-second-order representation when functoriality is needed, disallowing
-such ``extra'' variables does not negatively impact the expressivity
-of our calculus. We therefore pursue the simpler syntax here.
+some additional conditions throughout our model construction, we do
+not pursue this line of investigation here.
+%But since we can always use an ADT's (semantically equivalent)
+%second-order representation when functoriality is needed, disallowing
+%such ``extra'' variables does not negatively impact the expressivity
+%of our calculus. We therefore pursue the simpler syntax here.
 
@@
 below, it is not clear how we could define it for an arbitrary $F :
-[\rel^k,\rel]$. This observation will be especially critical in
-Section~\ref{sec:GADTs}.
+[\rel^k,\rel]$.
 
@@
 where $\mathtt{Nat} = \mu \alpha. 1 + \alpha$ and $\mathtt{Nat}
-\to C$ abbreviates $(\Lan^\alpha_{\mathtt{Nat}} 1)\,C$.  This
+\to C$ abbreviates $(\Lan^\emptyset_{\mathtt{Nat}} 1)\,C$.  This
 representation of $\mathit{Seq}$ corresponds exactly to the rewriting

### 2021-07-06 21-lmcs/21-lmcs.tex
6th  LMCS revisions
LMCS revisions 6

@@
 \EnableBpAbbreviations
+\def\ruleScoreFiller{\hrule height 0.8pt}
 
@@
 and working in a categorical setting we side-step such metatheoretic
-distractions. It is important to note that different consequences of
-parametricity are available in syntactic and semantic
-metatheories. Consequences of parametricity are possible for both
-closed and open System F terms in a syntactic metatheory --- although
-not all that can be formulated can be always proved; see, e.g., the
-end of Section~7 of~\cite{bm98}. By contrast, in a categorical
-metatheory consequences of parametricity are expressible only for {\em
-  closed} terms. For this reason, validating the standard consequences
-of parametricity for closed terms is --- going all the way back to
-Reynolds~\cite{rey83} --- all that is required for a model of
-parametricity to be considered good.
+distractions.
+%It is important to note that different consequences of
+%parametricity are available in syntactic and semantic
+%metatheories. Consequences of parametricity are possible for both
+%closed and open System F terms in a syntactic metatheory --- although
+%not all that can be formulated can be always proved; see, e.g., the
+%end of Section~7 of~\cite{bm98}. By contrast, in a semantic
+%metatheory consequences of parametricity are expressible only for {\em
+%  closed} terms. For this reason, validating the standard consequences
+%of parametricity for closed terms is --- going all the way back to
+%Reynolds~\cite{rey83} --- all that is required for a model of
+%parametricity to be considered good.
 
@@
 \DisplayProof
-\end{array}\]
+\end{array}\]\\[3ex]
 \[\begin{array}{c}
@@
 \DisplayProof
-\\[3ex]
+\\[5ex]
 \AXC{$\phi^k \in \Gamma \cup \Phi$}
@@
 \DisplayProof
-\\[3ex]
+\\[5ex]
 \AXC{$\Gamma;\ol{\alpha^0},\phi^k \vdash F$}
@@
 \DisplayProof
-\\[4ex]
+\\[6ex]
 \AXC{$\Gamma;\Phi \vdash F$}
@@
 \DisplayProof
-\\[3ex]
+\\[5ex]
 \AXC{$\Gamma;\Phi \vdash F$}
@@
 \BIC{$\Gamma; \Phi \vdash F \times G$}
-\DisplayProof
+\DisplayProof\\[2ex]
 \end{array}\]
@@
   \vdash \Nat^{\ol\alpha} F\,G}$ is $\omega$-cocontinuous since it is
-constant on $\omega$-directed sets; indeed, it is constant on
-$\omega$-directed sets because Definition~\ref{def:set-env} ensures
-that restrictions to $\mathbb{T}^k$ of morphisms between set
-environments are identities.  Interpretations of $\Nat$-types ensure
-that $\setsem{\Gamma \vdash F \to G}$ and $\setsem{\Gamma \vdash
-  \forall \ol\alpha. F}$ are as expected in any parametric model.
+constant (in particular, on $\omega$-directed sets). This is because
+Definition~\ref{def:set-env} ensures that restrictions to
+$\mathbb{T}^k$ of morphisms between set environments are identities.
+Interpretations of $\Nat$-types ensure that $\setsem{\Gamma \vdash F
+  \to G}$ and $\setsem{\Gamma \vdash \forall \ol\alpha. F}$ are as
+expected in any parametric model.
 
@@
   \Nat^{\ol\alpha} F\,G}$ is $\omega$-cocontinuous; indeed, it is
-constant on $\omega$-directed sets because
+constant (on $\omega$-directed sets) because
 Definition~\ref{def:reln-env} ensures that restrictions to
@@
 on the right commute, they must be equal.
-%\begin{figure*}[ht]
+\begin{figure*}[ht]
 %\vspace*{-0.1in}
-%  \hspace*{-0.85in}
+  \hspace*{-1in}
   \begin{minipage}[b]{0.25\linewidth}
@@
     \]}
-\end{minipage}
+\end{minipage}\hspace*{0.2in}
 \begin{minipage}[b]{0.25\linewidth}
@@
           F\overline{C}
-          &F\overline{C} \times F\overline{D}
+          &F\overline{C} \times G\overline{D}
           \ar[l, "{\pi_1}"'] \ar[r, "{\pi_2}"]
-          &F\overline{D}
-          \ar[r, "{\alpha_{\ol{D}}}"]
           &G\overline{D}\\
+%          \ar[r, "{\alpha_{\ol{D}}}"]
+%          &G\overline{D}\\
           &F(\overline{A \times B})
@@
           \ar[ul, "{F\ol{\pi_1} \circ F(\overline{\beta \times \beta'})}"]
-          \ar[urr, "{\alpha_{\ol{D}} \circ F\ol{\pi_2} \circ F(\overline{\beta \times \beta'})}"']
+          \ar[ur, "{\alpha_{\ol{D}} \circ F\ol{\pi_2} \circ F(\overline{\beta \times \beta'})}"']
       \end{tikzcd}
@@
 \end{minipage}
-%\end{figure*}
+\end{figure*}
 
@@
           &F\overline{A} \times G\overline{B}
-          \ar[d, "{F\overline{\beta} \times F\overline{\beta'}}"] \\
+          \ar[d, "{F\overline{\beta} \times G\overline{\beta'}}"] \\
           F\overline{S}
@@
 \noindent
-where $\mathtt{Nat} = \mu \alpha. 1 + \alpha$ and $\mathtt{Nat}
-\to C$ abbreviates $(\Lan^\emptyset_{\mathtt{Nat}} 1)\,C$.  This
-representation of $\mathit{Seq}$ corresponds exactly to the rewriting
-in Section~\ref{sec:lke}.  As explained in Section~IV.\,D
-of~\cite{jp19}, the more general construct $\Lan^{\ol{\alpha^0}}_{\ol
-  K} F$ allowing extensions along vectors of functors as depicted
-above makes it possible to represent GADTs with two or more type
-arguments that depend on one another. Such GADTs cannot be represented
-using just unary $\Lan$-types, i.e, $\Lan$-types of the form
-$(\Lan^{\ol{\alpha^0}}_K F) A$.
+where $\mathtt{Nat} = \mu \alpha. 1 + \alpha$ and $\mathtt{Nat} \to C$
+abbreviates $(\Lan^\emptyset_{\mathtt{Nat}} 1)\,C$, which equals
+$\colim{f : \mathtt{Nat} \to C}{1}$ by Equation~\ref{eq:colim-form}
+(here, $k = 0$). This representation of $\mathit{Seq}$ corresponds
+exactly to the rewriting in Section~\ref{sec:lke}.  As explained in
+Section~IV.\,D of~\cite{jp19}, the more general construct
+$\Lan^{\ol{\alpha^0}}_{\ol K} F$ allowing extensions along vectors of
+functors as depicted above makes it possible to represent GADTs with
+two or more type arguments that depend on one another. Such GADTs
+cannot be represented using just unary $\Lan$-types, i.e, $\Lan$-types
+of the form $(\Lan^{\ol{\alpha^0}}_K F) A$.
 

### 2021-07-07 21-lmcs/21-lmcs.tex
LMCS revisions 7

@@
 
-\author[D.~Jeffries]{Daniel Jeffries}	
-\address{Appalachian State University}	
-\email{jeffriesd@appstate.edu}
+%\author[D.~Jeffries]{Daniel Jeffries}	
+%\address{Appalachian State University}	
+%\email{jeffriesd@appstate.edu}
 
@@
 $\mathbb{T}^k$ of morphisms between set environments are identities.
-Interpretations of $\Nat$-types ensure that $\setsem{\Gamma \vdash F
-  \to G}$ and $\setsem{\Gamma \vdash \forall \ol\alpha. F}$ are as
-expected in any parametric model.
+Interpretations of $\Nat$-types ensure that the interpretations
+$\setsem{\Gamma \vdash F \to G}$ and $\setsem{\Gamma \vdash \forall
+  \ol\alpha. F}$ of the System F arrow types and $\forall$-types
+representable in our calculus are as expected in any parametric model.
 
@@
 and~\ref{def:rel-sem-funcs} below respect weakening, and also ensure
-that $\relsem{\Gamma \vdash F \to G}$ and $\relsem{\Gamma \vdash
-  \forall \ol \alpha. F}$ are as expected in any parametric model.  As
-for set interpretations, $\relsem{\Gamma; \emptyset \vdash
-  \Nat^{\ol\alpha} F\,G}$ is $\omega$-cocontinuous; indeed, it is
-constant (on $\omega$-directed sets) because
-Definition~\ref{def:reln-env} ensures that restrictions to
-$\mathbb{T}^k$ of morphisms between relational environments are
+that the interpretations $\relsem{\Gamma \vdash F \to G}$ and
+$\relsem{\Gamma \vdash \forall \ol\alpha. F}$ of the System F arrow
+types and $\forall$-types representable in our calculus are as
+expected in any parametric model.  As for set interpretations,
+$\relsem{\Gamma; \emptyset \vdash \Nat^{\ol\alpha} F\,G}$ is
+$\omega$-cocontinuous; indeed, it is constant (on $\omega$-directed
+sets) because Definition~\ref{def:reln-env} ensures that restrictions
+to $\mathbb{T}^k$ of morphisms between relational environments are
 identities. If $\rho \in \relenv$ and $\vdash F$, then we write
@@
 \[\dsem{\Gamma;\emptyset \,|\, \Delta
-    \vdash \lambda x.t : F \to G}\rho = \curry
+  \vdash L_\empty x. t : \Nat^\emptyset F\, G}\rho = \curry
   (\dsem{\Gamma;\emptyset \,|\, \Delta, x : F \vdash t :
@@
     G} \rho = \eval \circ \langle \dsem{\Gamma;\emptyset \,|\, \Delta
-  \vdash s: F \to G}\rho, \dsem{\Gamma;\emptyset \,|\, \Delta \vdash
+  \vdash s: \Nat^\emptyset F\, G}\rho, \dsem{\Gamma;\emptyset \,|\, \Delta \vdash
   t: F}\rho \rangle\]
@@
 \noindent
-so it specializes to the standard interpretations for System F terms.
-Term interpretation also respects substitution for both functorial and
-non-functorial type variables, as well as term substitution. That is,
-if $\Gamma,\alpha;\Phi \,|\, \Delta \vdash t : F$ and
-$\Gamma;\Phi,\alpha \,|\, \Delta \vdash t' : F$ and $\Gamma;\Phi
-\vdash G$ then
+so it specializes to the standard interpretations of those System F
+terms that are representable in our calculus.  Term interpretation
+also respects substitution for both functorial and non-functorial type
+variables, as well as term substitution. That is, if
+$\Gamma,\alpha;\Phi \,|\, \Delta \vdash t : F$ and $\Gamma;\Phi,\alpha
+\,|\, \Delta \vdash t' : F$ and $\Gamma;\Phi \vdash G$ then
 \[\dsem{\Gamma;\Phi \,|\, \Delta[\alpha := G] \vdash t[\alpha :=
@@
 theorems that are consequences of naturality, and thus do not require
-the full power of parametricity. In particular, we can prove all of
-the standard properties of the initial algebraic constructs $\map$,
-$\tin$, and $\fold$. We have, for example, that the interpretation of
-every $\map_H$ is a functor, i.e., if $\Gamma; \ol\alpha, \ol\gamma
-\vdash H$, $\ol{\Gamma; \emptyset \,|\, \Delta \vdash g :
-  \Nat^{\ol\gamma} F \, G}$, and $\ol{\Gamma; \emptyset \,|\, \Delta
-  \vdash f : \Nat^{\ol\gamma} G \, K}$, then
+the full power of parametricity. In particular,
+Definition~\ref{def:set-interp} alone ensures that standard properties
+of the initial algebraic constructs $\map$, $\tin$, and $\fold$
+hold. We have, for example, that the interpretation of every $\map_H$
+is a functor, i.e., if $\Gamma; \ol\alpha, \ol\gamma \vdash H$,
+$\ol{\Gamma; \emptyset \,|\, \Delta \vdash g : \Nat^{\ol\gamma} F \,
+  G}$, and $\ol{\Gamma; \emptyset \,|\, \Delta \vdash f :
+  \Nat^{\ol\gamma} G \, K}$, then
   \begin{align*}
@@
 consequences can be used to prove free theorems that go beyond mere
-naturality.  We also show that we can extend short cut short cut
-fusion for lists~\cite{glp93} to nested types, thereby formally
-proving correctness of the categorically inspired theorem
-from~\cite{jg10}.
+naturality.  We also show that we can extend short cut fusion for
+lists~\cite{glp93} to nested types, thereby formally proving
+correctness of the categorically inspired theorem from~\cite{jg10}.
 
@@
 
-Let $ \vdash g : \Nat^\alpha \,\onet\,\alpha$, let $G^\set =
-\setsem{\vdash g : \Nat^\alpha \,\onet\,\alpha}$ and $G^\rel =
+Let $ \vdash g : \Nat^\alpha \,\onet\,\alpha$, let $g^\set =
+\setsem{\vdash g : \Nat^\alpha \,\onet\,\alpha}$ and $g^\rel =
 \relsem{\vdash g : \Nat^\alpha \,\onet\,\alpha}$.  By
-Theorem~\ref{thm:at-gen}, $(G^\set(\pi_1\rho),G^\set(\pi_2\rho)) =
-G^\rel\rho$. Thus, for all $\rho \in \relenv$ and any $(a, b) \in
+Theorem~\ref{thm:at-gen}, $(g^\set(\pi_1\rho),g^\set(\pi_2\rho)) =
+g^\rel\rho$. Thus, for all $\rho \in \relenv$ and any $(a, b) \in
 \relsem{\vdash \emptyset}\rho = 1$, eliding the only possible
-instantiations of $a$ and $b$ gives that $(G^\set,G^\set) \;= \;
-(G^\set(\pi_1 \rho), G^\set (\pi_2 \rho))$ $ \in \relsem{\vdash
+instantiations of $a$ and $b$ gives that $(g^\set,g^\set) \;= \;
+(g^\set(\pi_1 \rho), g^\set (\pi_2 \rho))$ $ \in \relsem{\vdash
   \Nat^\alpha \,\onet\,\alpha}\rho$ $ = \{\eta : K_1 \Rightarrow
 \id\}$ $ = \{(\eta_1 : K_1 \Rightarrow \id, \eta_2 : K_1 \Rightarrow
-\id)\}$ That is, $G^\set$ is a natural transformation from the
+\id)\}$ That is, $g^\set$ is a natural transformation from the
 constantly $1$-valued functor to the identity functor in $\set$. In
-particular, for every $S : \set$, $G^\set_S : 1 \to S$. Note, however,
+particular, for every $S : \set$, $g^\set_S : 1 \to S$. Note, however,
 that if $S = \emptyset$, then there can be no such morphism, so no
@@
 
-Let $ \vdash g : \Nat^\alpha \,\alpha\,\alpha$, \/$G^\set =
-\setsem{\vdash g : \Nat^\alpha \,\alpha\,\alpha}$, and $G^\rel =
+Let $ \vdash g : \Nat^\alpha \,\alpha\,\alpha$, \/$g^\set =
+\setsem{\vdash g : \Nat^\alpha \,\alpha\,\alpha}$, and $g^\rel =
 \relsem{\vdash g : \Nat^\alpha \,\alpha\,\alpha}$.  By
-Theorem~\ref{thm:at-gen}, $(G^\set(\pi_1\rho),G^\set(\pi_2\rho)) =
-G^\rel\rho$. Thus, for all $\rho \in \relenv$ and any $(a, b) \in
+Theorem~\ref{thm:at-gen}, $(g^\set(\pi_1\rho),g^\set(\pi_2\rho)) =
+g^\rel\rho$. Thus, for all $\rho \in \relenv$ and any $(a, b) \in
 \relsem{\vdash \emptyset}\rho = 1$, eliding the only possible
-instantiations of $a$ and $b$ gives that $(G^\set, G^\set) \; = \;
-(G^\set(\pi_1 \rho), G^\set (\pi_2 \rho))$ $ \in \relsem{\vdash
+instantiations of $a$ and $b$ gives that $(g^\set, g^\set) \; = \;
+(g^\set(\pi_1 \rho), g^\set (\pi_2 \rho))$ $ \in \relsem{\vdash
   \Nat^\alpha \,\alpha\,\alpha}\rho$ $ = \{\eta : \id \Rightarrow
 \id\}$ $ = \{(\eta_1 : \id \Rightarrow \id, \eta_2 : \id \Rightarrow
-\id)\}$ That is, $G^\set$ is a natural transformation from the
+\id)\}$ That is, $g^\set$ is a natural transformation from the
 identity functor on $\set$ to itself. Now let $S$ be any set.  For any
@@
 morphism on $S$ then instantiating the naturality square implied by
-the above equality gives that $G^\set_S \circ K_a = K_a \circ
-G^\set_S$, i.e., $G^\set_S \, a = a$, i.e., $G^\set_S = \id_S$.
+the above equality gives that $g^\set_S \circ K_a = K_a \circ
+g^\set_S$, i.e., $g^\set_S \, a = a$, i.e., $g^\set_S = \id_S$.
 Putting these two cases together we have that for every $S : \set$,
-$G^\set_S = \id_S$, i.e., $G^\set$ is the identity natural
+$g^\set_S = \id_S$, i.e., $g^\set$ is the identity natural
 transformation for the identity functor on $\set$. So every closed
@@
 If \,$\vdash F$, $\vdash H$, and 
-$G \, = \, \setsem{\beta; \emptyset \,|\, \emptyset \vdash g :
+$\hat{g} \, = \, \setsem{\beta; \emptyset \,|\, \emptyset \vdash g :
   \Nat^{\emptyset} (\Nat^{\emptyset} (\onet + F \times \beta)\,
@@
 \setsem{\vdash H}$, then
-\[\mathit{fold}_{1 + \setsem{\vdash F} \times \_}\, n\, c\; (G\; (\mathit{List}\,
-\setsem{\vdash F})\,\mathit{nil} \,\mathit{cons}) = G \,\setsem{\vdash
+\[\mathit{fold}_{1 + \setsem{\vdash F} \times \_}\, n\, c\; (\hat{g}\; (\mathit{List}\,
+\setsem{\vdash F})\,\mathit{nil} \,\mathit{cons}) = \hat{g} \,\setsem{\vdash
   H}\, n\, c \]
@@
 \[\begin{array}{lll}
-(G \,(\pi_1 \rho), \,G\, (\pi_2 \rho)) & \in 
+(\hat{g} \,(\pi_1 \rho), \,\hat{g}\, (\pi_2 \rho)) & \in 
 & \relsem{\beta; \emptyset \vdash \Nat^{\emptyset} (\Nat^{\emptyset}
@@
 \rho\beta \to \rho\beta$ and $(n', n) \in \rho\beta$ then
-$(G \,(\pi_1 \rho)\, n'\, c', G\, (\pi_2 \rho)\, n\, c)
+$(\hat{g} \,(\pi_1 \rho)\, n'\, c', \hat{g}\, (\pi_2 \rho)\, n\, c)
 \in \rho \beta$.
@@
 \end{array}\]
-thus gives that $(G \;(\mathit{List}\,\setsem{\vdash
-  F})\,\mathit{nil} \,\mathit{cons}, G \,\setsem{\vdash H} \,
+thus gives that $(\hat{g}\;(\mathit{List}\,\setsem{\vdash
+  F})\,\mathit{nil} \,\mathit{cons}, \hat{g} \,\setsem{\vdash H} \,
 n\, c) \in \graph{\mathit{fold}_{1 + \setsem{\vdash F} \times \_}\, n\, c}$, i.e.,
-that $\mathit{fold}_{1 + \setsem{\vdash F} \times \_}\, n\, c\; (G\;
+that $\mathit{fold}_{1 + \setsem{\vdash F} \times \_}\, n\, c\; (\hat{g}\;
 (\mathit{List}\,\setsem{\vdash F})\, \mathit{nil}\, \mathit{cons})
-= G \,\setsem{\vdash H}\, n\, c$.
+= \hat{g}\,\setsem{\vdash H}\, n\, c$.
 \qed
@@
 and 
-\[G = \setsem{\phi;\emptyset\,|\,\emptyset \vdash g :
+\[\hat{g} = \setsem{\phi;\emptyset\,|\,\emptyset \vdash g :
 \Nat^\emptyset\,(\Nat^\alpha\,F\,(\phi\alpha))\,(\Nat^\alpha\,\onet \,
@@
   \alpha \vdash K}$,
-$$\mathit{fold}_{H}\, B \, (G\; \mu H \; \mathit{in}_{H}) = G
+$$\mathit{fold}_{H}\, B \, (\hat{g}\; \mu H \; \mathit{in}_{H}) = \hat{g}
 \,\setsem{\emptyset;\alpha \vdash K}\, B$$
@@
 \[\begin{array}{lll}
-(G \,(\pi_1 \rho), G\, (\pi_2 \rho)) & \in &
+(\hat{g} \,(\pi_1 \rho), \hat{g}\, (\pi_2 \rho)) & \in &
 \relsem{\phi;\emptyset\vdash \Nat^{\emptyset} (\Nat^\alpha 
@@
 so if $(A, B) \in \relsem{\phi;\emptyset\vdash \Nat^\alpha F\,
-  (\phi\alpha)}\rho$ then $(G \,(\pi_1 \rho)\, A, G\, (\pi_2 \rho)\,
+  (\phi\alpha)}\rho$ then $(\hat{g} \,(\pi_1 \rho)\, A, \hat{g}\, (\pi_2 \rho)\,
 B) \in \rho \phi$.
@@
 $\mathit{fold}_H$ as a (indeed, the unique) morphism from
-$\mathit{in}_H$ to $B$.  Thus, $(G \,(\pi_1 \rho)\, A, G\, (\pi_2
+$\mathit{in}_H$ to $B$.  Thus, $(\hat{g} \,(\pi_1 \rho)\, A,  \hat{g}\, (\pi_2
 \rho)\, B) \in \graph{\mathit{fold}_H\, B}$, i.e., $\mathit{fold}_H \,
-B \, (G\, (\pi_1 \rho) \, \mathit{in}_H) = G\,(\pi_2 \rho)\,B$.  But
-since $\phi$ is the only free variable in $G$, this simplifies to
-$\mathit{fold}_H\, B \, (G\, \mu H\, \mathit{in}_H) =
-G\,\setsem{\emptyset;\alpha\vdash K}\,B$. \qed
+B \, (\hat{g}\, (\pi_1 \rho) \, \mathit{in}_H) = \hat{g}\,(\pi_2 \rho)\,B$.  But
+since $\phi$ is the only free variable in $\hat{g}$, this simplifies to
+$\mathit{fold}_H\, B \, (\hat{g}\, \mu H\, \mathit{in}_H) =
+\hat{g}\,\setsem{\emptyset;\alpha\vdash K}\,B$. \qed
 
@@
 more general free theorem whose conclusion is $\mathit{fold}_{H}\, B
-\; \circ \; G\; \mu H \; \mathit{in}_{H} = G
+\; \circ \; \hat{g}\; \mu H \; \mathit{in}_{H} = \hat{g}
 \,\setsem{\emptyset;\alpha \vdash K}\, B$.
@@
 Generalized algebraic datatypes (GADTs) --- also known as guarded
-recursive data types or first-class phantom types --- generalize
-nested types by relaxing the above restriction to allow the return
-types of data constructors to be different instances of the datatype
-than the one being defined. For example, the GADT 
+recursive data types~\cite{xcc03} or first-class phantom
+types~\cite{ch03} --- generalize nested types by relaxing the above
+restriction to allow the return types of data constructors to be
+different instances of the datatype than the one being defined. For
+example, the GADT
 \[\begin{array}{l}
@@
 the (entirely standard) approach taken above for product, coproduct,
-and fixpoint types. In this section we outline this approach and show
-how our model from Section~\ref{sec:term-interp} can be extended to a
-parametric model when some classes of primitive GADTs are incorporated
-in this manner. However, as we argue at the end of this section, such
-extension is likely not possible.
+and fixpoint types. In this section we outline the obvious approach to
+extending our model from Section~\ref{sec:term-interp} to a parametric
+model when some classes of primitive GADTs are incorporated in this
+manner. However, as we argue at the end of this section, this naive
+approach fails because the IEL does not hold. Unfortunately, the IEL
+not holding derails more than just parametricity: it also affects the
+well-definedness of the term semantics (in particular, the semantics
+of $L$-terms).
 
@@
 The problem in Example~\ref{ex:no-rel-transf} lies in the fact that
-the type along whose intepretation we extend contains constants, i.e.,
+the type along whose interpretation we extend contains constants, i.e.,
 subtypes constructed from $\onet$. Indeed, if $\ol{\emptyset;

### 2021-07-08 21-lmcs/21-lmcs.tex
change bush reverse example
LMCS revisions on defs

@@
 $\mathtt{reverseBush}$ function that reverses the data in a bush. This
-function maps, e.g., the bush {\color{red} Change one 2 to 4? Find a
-  better name for this operation than reverse. Ask Daniel.}
+function maps, e.g., the bush 
 \[\begin{array}{l}
@@
 \mathit{bnil}))\\ \hspace*{0.93in}(\mathit{bcons}\, (\mathit{bcons}\,
-(\mathit{bcons}\, 2 \, (\mathit{bcons}\, (\mathit{bcons}\, 3\,
+(\mathit{bcons}\, 3 \, (\mathit{bcons}\, (\mathit{bcons}\, 4\,
 \mathit{bnil})\,  \mathit{bnil}))\, \mathit{bnil})\, \mathit{bnil}))
@@
 \[\begin{array}{c}
-\hspace*{-1.64in}\mathit{bcons}\,3\, (\mathit{bcons}\, (\mathit{bcons}\, 0
-\,(\mathit{bcons}\, (\mathit{bcons}\, 2 \, \mathit{bnil}) \,
+\hspace*{-1.64in}\mathit{bcons}\,4\, (\mathit{bcons}\, (\mathit{bcons}\, 0
+\,(\mathit{bcons}\, (\mathit{bcons}\, 3 \, \mathit{bnil}) \,
 \mathit{bnil}))\\ 
@@
 \begin{defi}\label{def:RT-functor}
-An endofunctor $H$ on $RT_k$ is a triple $H = (H^1,H^2,H^*)$, where
+A {\em higher-order relation transformer} $H$ on $RT_k$ is a triple $H
+= (H^1,H^2,H^*)$, where
 \begin{itemize}
@@
 \item $H^*$ is a functor from $RT_k$ to $[\rel^k,\rel]$
+\item $H^*(F^1,F^2,F^*) \ol{R} : \rel(H^1F^1 \ol{A}, H^2F^2 \ol{B})$
+  whenever $R_1:\rel(A_1,B_1),...,R_k:\rel(A_k,B_k)$
+\item $H^*(F^1,F^2,F^*)\, \ol{(\alpha, \beta)} = (H^1F^1\ol{\alpha},
+  H^2F^2 \ol{\beta})$ whenever $(\alpha_1, \beta_1) \in
+  \Homrel(R_1,S_1),..., (\alpha_k, \beta_k)$ $\in \Homrel(R_k,S_k)$
 \item For all $\overline{R : \rel(A,B)}$,
@@
   \delta^2)_{\overline{B}}$
+\end{itemize}
+\end{defi}
+\noindent
+Definition~\ref{def:RT-functor} entails that every higher-order
+relation transformer $H$ is an endofunctor on $RT_k$, where
+\begin{itemize}
 \item The action of $H$ on objects is given by $H\,(F^1,F^2,F^*) =
@@
 \end{itemize}  
+Note that the last condition of Definition~\ref{def:RT-functor}
+entails that if $(\delta^1,\delta^2) : (F^1,F^2,F^*)\to (G^1,G^2,G^*)$
+in $RT_k$, if $R_1:\rel(A_1,B_1),...,R_k:\rel(A_k,B_k)$, and if $(x,
+y) \in H^*(F^1,F^2,F^*)\ol{R}$, then
+\[((H^1\delta^1)_{\ol{A}}x, (H^2\delta^2)_{\ol{B}}y) \in
+H^*(G^1,G^2,G^*)\ol{R}\]
+
+\begin{defi}\label{def:RT-nat-trans}
+A {\em morphism} $\sigma : H \to K$ between higher-order relation
+transformers $H$ and $K$ on $RT_k$ is a pair $\sigma = (\sigma^1,
+\sigma^2)$, where $\sigma^1 : H^1 \to K^1$ and $\sigma^2 : H^2 \to
+K^2$ are natural transformations between endofunctors on
+$[\set^k,\set]$ such that $((\sigma^1_{F^1})_{\overline{A}},
+(\sigma^2_{F^2})_{\overline{B}})$ is a morphism from $H^*
+F\overline{R}$ to $K^* F\overline{R}$ in $\rel$ for any $F \in RT_k$
+and any $k$-tuple of relations $\overline{R : \rel(A, B)}$.
 \end{defi}
-Since the results of applying an endofunctor $H$ to $k$-ary relation
-transformers and morphisms between them must again be $k$-ary relation
-transformers and morphisms between them, respectively,
-Definition~\ref{def:RT-functor} implicitly requires that the following
-three conditions hold:
-\begin{itemize}
-\item $H^*(F^1,F^2,F^*) \ol{R} : \rel(H^1F^1 \ol{A}, H^2F^2 \ol{B})$
-  whenever $R_1:\rel(A_1,B_1),...,R_k:\rel(A_k,B_k)$
-\item $H^*(F^1,F^2,F^*)\, \ol{(\alpha, \beta)} = (H^1F^1\ol{\alpha},
-  H^2F^2 \ol{\beta})$ whenever $(\alpha_1, \beta_1) \in
-  \Homrel(R_1,S_1),..., (\alpha_k, \beta_k)$ $\in \Homrel(R_k,S_k)$
-\item If $(\delta^1,\delta^2) : (F^1,F^2,F^*)\to (G^1,G^2,G^*)$ in
-  $RT_k$ and $R_1:\rel(A_1,B_1),...,R_k:\rel(A_k,B_k)$, then
-  $((H^1\delta^1)_{\ol{A}}x, (H^2\delta^2)_{\ol{B}}y) \in
-  H^*(G^1,G^2,G^*)\ol{R}$ whenever $(x, y) \in H^*(F^1,F^2,F^*)\ol{R}$
-\end{itemize}
-  Note, however, that this last condition is
-  automatically satisfied because it is implied by the third bullet
-  point of Definition~\ref{def:RT-functor}.
-
-\begin{defi}\label{def:RT-nat-trans}
-For endofunctors $H$ and $K$ on $RT_k$, a {\em natural transformation}
-$\sigma : H \to K$ is a pair $\sigma = (\sigma^1, \sigma^2)$, where
-$\sigma^1 : H^1 \to K^1$ and $\sigma^2 : H^2 \to K^2$ are natural
-transformations between endofunctors on $[\set^k,\set]$ and the
-component of $\sigma$ at $F = (F^1,F^2,F^*) \in RT_k$ is given by
-$\sigma_F = (\sigma^1_{F^1}, \sigma^2_{F^2})$.
-\end{defi}
-Definition~\ref{def:RT-nat-trans} entails that $\sigma^i_{F^i}$ is
-natural in $F^i : [\set^k,\set]$, and, for every $F$, both
+
+Definition~\ref{def:RT-nat-trans} entails that a morphism $\sigma$
+between higher-order relation transformers is a natural transformation
+between endofunctors on $RT_k$ whose component at $F = (F^1,F^2,F^*)
+\in RT_k$ is given by $\sigma_F = (\sigma^1_{F^1}, \sigma^2_{F^2})$.
+Moreover, $\sigma^i_{F^i}$ is natural in $F^i : [\set^k,\set]$, and,
+for every $F = (F^1,F^2,F^*) \in RT_k$, both
 $(\sigma^1_{F^1})_{\overline{A}}$ and
 $(\sigma^2_{F^2})_{\overline{A}}$ are natural in $\overline{A}$.
-Moreover, since the results of applying $\sigma$ to $k$-ary relation
-transformers must be morphisms of $k$-ary relation transformers,
-Definition~\ref{def:RT-nat-trans} implicitly requires that
-$(\sigma_F)_{\overline{R}} = ( (\sigma^1_{F^1})_{\overline{A}},
-(\sigma^2_{F^2})_{\overline{B}})$ is a morphism in $\rel$ for any
-$k$-tuple of relations $\overline{R : \rel(A, B)}$, i.e., that if $(x,
-y) \in H^*F\overline{R}$, then $((\sigma^1_{F^1})_{\overline{A}} x,
-(\sigma^2_{F^2})_{\overline{B}} y) \in K^*F\overline{R}$.
-
-\vspace*{0.1in}
+
 Critically, we can compute $\omega$-directed colimits in
-$RT_k$. Indeed, if $\cal D$ is an $\omega$-directed set then
-$\colim{d \in {\cal D}}{(F^1_d, F^2_d,F^*_d)} = (\colim{d \in {\cal
+$RT_k$. Indeed, if $\cal D$ is an $\omega$-directed set then $\colim{d
+  \in {\cal D}}{(F^1_d, F^2_d,F^*_d)} = (\colim{d \in {\cal
     D}}{F^1_d}, \colim{d \in {\cal D}}{F^2_d}, \colim{d \in {\cal
-    D}}{F^*_d})$.  We define an endofunctor $T = (T^1,T^2,T^*)$ on
-$RT_k$ to be {\em $\omega$-cocontinuous} if $T^1$ and $T^2$ are
-$\omega$-cocontinuous endofunctors on $[\set^k,\set]$ and $T^*$ is an
-$\omega$-cocontinuous functor from $RT_k$ to $[\rel^k,\rel]$, i.e., is
-in $[RT_k,[\rel^k,\rel]]$.
+    D}}{F^*_d})$.  We define a higher-order relation transformer $T =
+(T^1,T^2,T^*)$ on $RT_k$ to be {\em $\omega$-cocontinuous} if $T^1$
+and $T^2$ are $\omega$-cocontinuous endofunctors on $[\set^k,\set]$
+and $T^*$ is an $\omega$-cocontinuous functor from $RT_k$ to
+$[\rel^k,\rel]$, i.e., is in $[RT_k,[\rel^k,\rel]]$.
 %\begin{lemma}\label{lem:colimits}
@@
 %\end{proof}
-\begin{comment}
-\begin{defi}\label{def:omega-cocont}
-An endofunctor $T = (T^1,T^2,T^*)$ on $RT_k$ is {\em
-$\omega$-cocontinuous} if $T^1$ and $T^2$ are $\omega$-cocontinuous
-endofunctors on $[\set^k,\set]$ and $T^*$ is an $\omega$-cocontinuous
-functor from $RT_k$ to $[\rel^k,\rel]$, i.e., is in
-$[RT_k,[\rel^k,\rel]]$.
-\end{defi}
-\end{comment}
 Now, for any $k$, any $A : \set$, and any $R : \rel(A, B)$, let
@@
 (K^\set_0,K^\set_0,K^\rel_0)$ is initial in $RT_k$. Thus, if $T =
-(T^1,T^2,T^*) : RT_k \to RT_k$ is an endofunctor on $RT_k$ then we can
-define the relation transformer $\mu T$ to be $\colim{n \in \nat}{T^n
-  K_0}$. It is not hard to see that $\mu T$ is given explicitly as
+(T^1,T^2,T^*) : RT_k \to RT_k$ is a higher-order relation transformer
+on $RT_k$ then we can define the relation transformer $\mu T$ to be
+$\colim{n \in \nat}{T^n K_0}$. It is not hard to see that $\mu T$ is
+given explicitly as
 \begin{equation}\label{eq:mu}
@@
 components of $\mu T$ and its third component, which reflects the
-important conceptual observation that the third component of an
-endofunctor on $RT_k$ need not be a functor on all of
-$[\rel^k,\rel]$. In particular, although we can define $T_{H,\rho}\,
-F$ for a relation transformer $F$ in Definition~\ref{def:rel-sem}
-below, it is not clear how we could define it for an arbitrary $F :
-[\rel^k,\rel]$.
+important conceptual observation that the third component of a
+higher-order relation transformer on $RT_k$ need not be a functor on
+all of $[\rel^k,\rel]$. In particular, although we can define
+$T_{H,\rho}\, F$ for a relation transformer $F$ in
+Definition~\ref{def:rel-sem} below, it is not clear how we could
+define it for an arbitrary $F : [\rel^k,\rel]$.
 
@@
 \begin{defi}\label{def:relenv-functor}
-For each $k$, an $\omega$-cocontinuous functor $H \in [\relenv, RT_k]$
-is a triple $H = (H^1,H^2,H^*)$, where
+  For each $k$, an {\em $\omega$-cocontinuous environment transformer}
+  $H$ is a triple $H = (H^1,H^2,H^*)$, where
 \begin{itemize}
@@
 \item $H^*$ is an object in $[\relenv,[\rel^k,\rel]]$
-\item For all $\overline{R : \rel(A,B)}$ and morphisms $f$ in
-  $\relenv$, $\pi_1(H^*f \,{\overline{R}}) = H^1 (\pi_1
-  f)\,{\overline{A}}$ and $\pi_2(H^*f \,{\overline{R}}) = H^2 (\pi_2
-  f)\,{\overline{B}}$
-\item The action of $H$ on $\rho$ in $\relenv$ is given by $H \rho = (H^1
-  (\pi_1 \rho),\,H^2 (\pi_2 \rho),\,H^*\rho)$
-\item The action of $H$ on morphisms $f : \rho \to \rho'$ in $\relenv$
-  is given by $Hf = (H^1 (\pi_1 f),H^2 (\pi_2 f))$
-\end{itemize}
-\end{defi}
-\noindent Spelling out the last two bullet points above gives the
-following analogues of the three conditions immediately following
-Definition~\ref{def:RT-functor}:
-\begin{itemize}
 \item $H^*\rho\, \ol{R} : \rel(H^1(\pi_1 \rho)\, \ol{A}, H^2(\pi_2
@@
   \Homrel(R_1,S_1),..., (\alpha_k, \beta_k) \in \Homrel(R_k,S_k)$
-\item If $f : \rho \to \rho'$ and
-  $R_1:\rel(A_1,B_1),...,R_k:\rel(A_k,B_k)$, then $(H^1(\pi_1
-  f)\,{\ol{A}}\,x, H^2(\pi_2 f)\,{\ol{B}}\,y) \in H^*\rho'\,\ol{R}$
-  whenever $(x, y) \in H^*\rho\,\ol{R}$.
+\item For all $\overline{R : \rel(A,B)}$, $\pi_1(H^*f
+  \,{\overline{R}}) = H^1 (\pi_1 f)\,{\overline{A}}$ and $\pi_2(H^*f
+  \,{\overline{R}}) = H^2 (\pi_2 f)\,{\overline{B}}$
 \end{itemize}
-As before, the last condition is automatically satisfied because it is
-implied by the third bullet point of
-Definition~\ref{def:relenv-functor}.
+\end{defi}
+\noindent
+Definition~\ref{def:relenv-functor} entails that every
+$\omega$-cocontinuous environment transformer $H$ is a
+$\omega$-cocontinuous functor from $\relenv$ to $RT_k$, where
+\begin{itemize}
+\item The action of $H$ on $\rho$ in $\relenv$ is given by $H \rho = (H^1
+  (\pi_1 \rho),\,H^2 (\pi_2 \rho),\,H^*\rho)$
+\item The action of $H$ on morphisms $f : \rho \to \rho'$ in $\relenv$
+  is given by $Hf = (H^1 (\pi_1 f),H^2 (\pi_2 f))$
+\end{itemize}
+\noindent
+Note that the last condition of Definition~\ref{def:relenv-functor}
+entails that if $f : \rho \to \rho'$, if
+$R_1:\rel(A_1,B_1),...,R_k:\rel(A_k,B_k)$, and if $(x, y) \in
+H^*\rho\,\ol{R}$, then \[(H^1(\pi_1 f)\,{\ol{A}}\,x, H^2(\pi_2
+f)\,{\ol{B}}\,y) \in H^*\rho'\,\ol{R}\]
 
@@
   \tvars \cup \fvars} RT_k$, computation of $\omega$-directed colimits
-in $RT_k$ extends componentwise to $\relenv$. Similarly,
-$\omega$-cocontinuity for endofunctors on $RT_k$ extends componentwise
-to functors from $\relenv$ to $RT_k$.  Recalling from the start of
-this subsection that Definition~\ref{def:rel-sem} is given mutually
+in $RT_k$ extends componentwise to $\relenv$. Recalling from the start
+of this subsection that Definition~\ref{def:rel-sem} is given mutually
 inductively with Definition~\ref{def:set-sem} we can now define our
@@
   \relsem{\Gamma;\Phi \vdash G}\rho\\  
-\end{align*}
-
-\vspace*{-0.5in}
-
-\begin{align*}
    \relsem{\Gamma;\Phi \vdash (\mu \phi.\lambda
@@
 clause in Definition~\ref{def:rel-sem} to be well-defined we need
-$T_{H,\rho}$ to be an $\omega$-cocontinuous endofunctor on $RT$ so
-that, by Lemma~\ref{lem:fp}, it admits a fixpoint. Since $T_{H,\rho}$
-is defined in terms of $\relsem{\Gamma;\phi, \ol{\alpha} \vdash H}$,
-this means that relational interpretations of types must be
-$\omega$-cocontinuous functors from $\relenv$ to $RT_0$, which in turn
-entails that the actions of relational interpretations of types on
-objects and on morphisms in $\relenv$ are intertwined. As for set
-interpretations, we know from~\cite{jp19} that, for every $\Gamma;
-\ol{\alpha} \vdash F$, $\relsem{\Gamma; \ol{\alpha} \vdash F}$ is
-actually in $[\rel^k,\rel]$ where $k = |\ol \alpha|$. We first define
-the actions of each of these functors on morphisms between
-environments, and then argue that they are well-defined and have the
-required properties. To do this, we extend $T_H$ to a {\em functor}
-from $\relenv$ to $[[\rel^k,\rel],[\rel^k,\rel]]$. Its action on an
-object $\rho \in \relenv$ is given by the higher-order functor
+$T_{H,\rho}$ to be an $\omega$-cocontinuous higher-order relation
+transformer on $RT_k$, where $k$ is the arity of $\phi$, so that, by
+Lemma~\ref{lem:fp}, it admits a fixpoint. Since $T_{H,\rho}$ is
+defined in terms of $\relsem{\Gamma;\phi, \ol{\alpha} \vdash H}$, this
+means that relational interpretations of types must be
+$\omega$-cocontinuous environment transformer from $\relenv$ to
+$RT_0$, which in turn entails that the actions of relational
+interpretations of types on objects and on morphisms in $\relenv$ are
+intertwined. As for set interpretations, we know from~\cite{jp19}
+that, for every $\Gamma; \ol{\alpha} \vdash F$, $\relsem{\Gamma;
+  \ol{\alpha} \vdash F}$ is actually in $[\rel^k,\rel]$ where $k =
+|\ol \alpha|$. We first define the actions of each of these functors
+on morphisms between environments, and then argue that they are
+well-defined and have the required properties. To do this, we extend
+$T_H$ to a {\em higher-order $\omega$-cocontinuous environment
+  transformer} from $\relenv$ to the category of $\omega$-cocontinuous
+higher-order relation transformers on $RT_k$:
+
+\begin{defi}\label{def:howcet}
+A {\em higher-order $\omega$-cocontinuous environment transformer}
+is a triple $H = (H^1,H^2,H^*)$, where
+\begin{itemize}
+\item $H^1$ and $H^2$ are objects in
+  $[\setenv,[[\set^k,\set],[\set^k,\set]]]$ 
+\item $H^*$ is an object in $[\relenv,[[\rel^k,\rel],[\rel^k,\rel]]]$
+\item $(H^1(\pi_1 \rho) F^1,H^2(\pi_2 \rho) F^2, H^* \rho F^*) \in
+  RT_k$ whenever $(F^1,F^2,F^*) \in RT_k$
+\item $H^*\rho\, (\delta^1, \delta^2)$ is a morphism in $RT_k$
+  whenever $(\delta^1, \delta^2)$ is a morphism in $RT_k$ and,
+  moreover, $H^*\rho\, (\delta^1, \delta^2) = (H^1(\pi_1
+  \rho)\,\delta^1, H^2(\pi_2 \rho)\, \delta^2)$
+\item For all $(F^1,F^2,F^*) \in RT_k$, $\pi_1(H^* f
+  \,F^*) = H^1 (\pi_1 f)\,F^1$ and $\pi_2(H^*f
+  \,F^*) = H^2 (\pi_2 f)\,F^2$
+\end{itemize}
+\end{defi}
+\noindent
+Definition~\ref{def:howcet} entails that every higher-order
+$\omega$-cocontinuous environment transformer $H$ is a
+$\omega$-cocontinuous functor from $\relenv$ to
+the category of $\omega$-cocontinuous higher-order
+relation transformers on $RT_k$, where
+\begin{itemize}
+\item The action of $H$ on $\rho$ in $\relenv$ is given by $H \rho = (H^1
+  (\pi_1 \rho),\,H^2 (\pi_2 \rho),\,H^*\rho)$
+\item The action of $H$ on morphisms $f : \rho \to \rho'$ in $\relenv$
+  is given by $Hf = (H^1 (\pi_1 f),H^2 (\pi_2 f))$
+\end{itemize}
+\noindent
+Note that the last condition of Definition~\ref{def:howcet} entails
+that if $f : \rho \to \rho'$, if $(F^1,F^2,F^*) \in RT_k$, if $R_1 :
+\rel(A_1,B_1),...,R_k : \rel(A_k,B_k)$, and if $(x, y) \in
+H^*\rho\,F^*\,\ol{R}$, then \[(H^1(\pi_1 f)\,F^1\,\ol{A}\,x, H^2(\pi_2
+f)\,F^2\,\ol{B}\,y) \in H^*\rho'\,F^*\,\ol{R}\]
+
+------
+
+{\color{red} HERE!}
+
+The action of $T_H$ on an object $\rho \in \relenv$ is
+
+{\color{blue}
+given by the higher-order functor
 $T_{H,\rho}$ whose actions on objects and morphisms are given in
-Definition~\ref{def:rel-sem-funcs}. Its action on a morphism $f : \rho
-\to \rho'$ is the higher-order natural transformation $T_{H,f} :
+Definition~\ref{def:rel-sem-funcs}.}
+
+The action of  $T_H$ on a morphism $f : \rho \to \rho'$ is
+
+{\color{blue}
+the higher-order natural transformation $T_{H,f} :
 T_{H,\rho} \to T_{H,\rho'}$ whose action on any $F : [\rel^k,\rel]$ is
@@
 F)_{\ol{R}} = \relsem{\Gamma; \phi,\ol{\alpha} \vdash H}f[\phi :=
-  \id_F][\ol{\alpha := \id_R}]$.
+  \id_F][\ol{\alpha := \id_R}]$.}
+
+CHECK CONDITIONS
 
@@
   is by the definition of application of relation transformers from
-  Definition~9, and the fourth is by Equation~\ref{eq:eq-pres-eq}:
+  Definition~\ref{def:rel-transf}, and the fourth is by
+  Equation~\ref{eq:eq-pres-eq}:
 \[

### 2021-07-12 21-lmcs/21-lmcs.tex
MFPS revisions #8

@@
     \ol{\alpha}. H)\ol{G}}\rho
-  &= (\mu T_{H,\rho})\ol{\relsem{\Gamma;\Phi \vdash G}\rho}\\
+  &= (----\mu T_{H,\rho})\ol{\relsem{\Gamma;\Phi \vdash G}\rho}\\
   \text{where }	T_{H,\rho}
@@
 on morphisms between environments, and then argue that they are
-well-defined and have the required properties. To do this, we extend
-$T_H$ to a {\em higher-order $\omega$-cocontinuous environment
-  transformer} from $\relenv$ to the category of $\omega$-cocontinuous
-higher-order relation transformers on $RT_k$:
+well-defined and have the required properties. To do this, we show
+that $T_H = (T^\set_H, T^\set_H, T^\rel_H) $ is a {\em higher-order
+  $\omega$-cocontinuous environment transformer}, as defined by
 
@@
   $[\setenv,[[\set^k,\set],[\set^k,\set]]]$ 
-\item $H^*$ is an object in $[\relenv,[[\rel^k,\rel],[\rel^k,\rel]]]$
-\item $(H^1(\pi_1 \rho) F^1,H^2(\pi_2 \rho) F^2, H^* \rho F^*) \in
-  RT_k$ whenever $(F^1,F^2,F^*) \in RT_k$
+\item $H^*$ is an object in $[\relenv,[RT_k,[\rel^k,\rel]]]$
+\item $(H^1(\pi_1 \rho) F^1,H^2(\pi_2 \rho) F^2, H^* \rho F) \in
+  RT_k$ whenever $F = (F^1,F^2,F^*) \in RT_k$
 \item $H^*\rho\, (\delta^1, \delta^2)$ is a morphism in $RT_k$
@@
   \rho)\,\delta^1, H^2(\pi_2 \rho)\, \delta^2)$
-\item For all $(F^1,F^2,F^*) \in RT_k$, $\pi_1(H^* f
-  \,F^*) = H^1 (\pi_1 f)\,F^1$ and $\pi_2(H^*f
-  \,F^*) = H^2 (\pi_2 f)\,F^2$
+\item For all $F = (F^1,F^2,F^*) \in RT_k$ and all $R_1 :
+  \rel(A_1,B_1),...,R_k : \rel(A_k,B_k)$, $\pi_1(H^* f \,F\,\ol{R}) =
+  H^1 (\pi_1 f)\,F^1\,\ol{A}$ and $\pi_2(H^*f \,F\,\ol{R}) = H^2
+  (\pi_2 f)\,F^2\,\ol{B}$
 \end{itemize}
@@
 Note that the last condition of Definition~\ref{def:howcet} entails
-that if $f : \rho \to \rho'$, if $(F^1,F^2,F^*) \in RT_k$, if $R_1 :
+that if $f : \rho \to \rho'$, if $F = (F^1,F^2,F^*) \in RT_k$, if $R_1 :
 \rel(A_1,B_1),...,R_k : \rel(A_k,B_k)$, and if $(x, y) \in
-H^*\rho\,F^*\,\ol{R}$, then \[(H^1(\pi_1 f)\,F^1\,\ol{A}\,x, H^2(\pi_2
-f)\,F^2\,\ol{B}\,y) \in H^*\rho'\,F^*\,\ol{R}\]
-
-------
-
-{\color{red} HERE!}
-
-The action of $T_H$ on an object $\rho \in \relenv$ is
-
-{\color{blue}
-given by the higher-order functor
-$T_{H,\rho}$ whose actions on objects and morphisms are given in
-Definition~\ref{def:rel-sem-funcs}.}
-
-The action of  $T_H$ on a morphism $f : \rho \to \rho'$ is
-
-{\color{blue}
-the higher-order natural transformation $T_{H,f} :
-T_{H,\rho} \to T_{H,\rho'}$ whose action on any $F : [\rel^k,\rel]$ is
-the natural transformation $T_{H,f}\, F : T_{H,\rho}\, F \to
-T_{H,\rho'}\, F$ whose component at $\ol{R}$ is $(T_{H,f}\,
-F)_{\ol{R}} = \relsem{\Gamma; \phi,\ol{\alpha} \vdash H}f[\phi :=
-  \id_F][\ol{\alpha := \id_R}]$.}
-
-CHECK CONDITIONS
-
-Using $T_H$, we can define the functorial action of relational
-interpretation.
+H^*\rho\,F\,\ol{R}$, then \[(H^1(\pi_1 f)\,F^1\,\ol{A}\,x, H^2(\pi_2
+f)\,F^2\,\ol{B}\,y) \in H^*\rho'\,F\,\ol{R}\]
+
+Now, the fact that $T_H = (T^\set_H, T^\set_H, T^\rel_H) $ is a
+higher-order $\omega$-cocontinuous environment transformer follows
+from the analogue for relations of the argument immediately preceding
+Definition~\ref{def:set-sem-funcs}, together with the observations
+that the action of $T^\rel_{H,\rho}$ on morphisms will be well-defined
+by Definition~\ref{def:rel-sem-funcs}, and the functorial action of
+$\relsem{\Gamma; \phi,\ol{\alpha} \vdash H}$ will already be available
+in the final clause of Definition~\ref{def:howcet} from the induction
+hypothesis on $H$ and Lemma~\ref{lem:rel-transf-morph} below. The
+action of $T_H$ on an object $\rho \in \relenv$ is given by the
+$\omega$-cocontinuous higher-order relation transformer $T_{H,\rho}$
+whose actions on objects and morphisms are given in
+Definition~\ref{def:howcet}. The action of $T_H$ on a morphism $f :
+\rho \to \rho'$ is the morphism $T_{H,f} : T_{H,\rho} \to T_{H,\rho'}$
+between higher-order relation transformers whose action on any $F \in
+RT_k$ is the morphism of relation transfomers $T_{H,f}\, F :
+T_{H,\rho}\, F \to T_{H,\rho'}\, F$ whose component at $\ol{R}$ is
+$(T_{H,f}\, F)_{\ol{R}} = \relsem{\Gamma; \phi,\ol{\alpha} \vdash
+  H}f[\phi := \id_F][\ol{\alpha := \id_R}]$, i.e., is
+$(\setsem{\Gamma; \phi,\ol{\alpha} \vdash H}\,(\pi_1 f)\,[\phi :=
+  \id_{F^1}][\ol{\alpha := \id_A}], \setsem{\Gamma; \phi,\ol{\alpha}
+  \vdash H}\,(\pi_2 f)[\phi := \id_{F^2}][\ol{\alpha := \id_B}])$.
+{\color{red} CHECK CONDITIONS} Using $T_H$, we can define the
+functorial action of relational interpretation.
 \begin{defi}\label{def:rel-sem-funcs}
@@
 
-For the functors given in Definitions~\ref{def:rel-sem}
-and~\ref{def:rel-sem-funcs} to be well-defined we need that, for every
-$H$, $T_{H,\rho}\,F$ is a relation transformer for any relation
-transformer $F$, and that $T_{H,f}\, F : T_{H,\rho}\, F \to
-T_{H,\rho'}\, F$ is a morphism of relation transformers for every
-relation transformer $F$ and every morphism $f : \rho \to \rho'$ in
-$\relenv$. If we define $\sem{\Gamma;\Phi \vdash F}$ to be
-$(\setsem{\Gamma;\Phi \vdash F}, \setsem{\Gamma;\Phi \vdash
-  F},\relsem{\Gamma;\Phi \vdash F})$, then this is an immediate
-consequence of
+If we define $\sem{\Gamma;\Phi \vdash F}$ to be $(\setsem{\Gamma;\Phi
+  \vdash F}, \setsem{\Gamma;\Phi \vdash F},\relsem{\Gamma;\Phi \vdash
+  F})$, then this is an immediate consequence of
 \begin{lem}\label{lem:rel-transf-morph}
-For every $\Gamma;\Phi \vdash F$, $\sem{\Gamma;\Phi \vdash F}
-\in[\relenv,RT_0]$.
+For every $\Gamma;\Phi \vdash F$, $\sem{\Gamma;\Phi \vdash F}$ is an
+$\omega$-cocontinuous environment transformer.
+
 \end{lem}
@@
 together with Lemma~\ref{lem:fp} and Equation~\ref{eq:mu} for
-$\mu$-types.
+$\mu$-types. We note that, for any relation environment $\rho$,  
+$\sem{\Gamma;\Phi \vdash F}\rho \in RT_0$.
 
@@
 Every well-formed term $\Gamma;\Phi~|~\Delta \vdash t : F$ induces
-a natural transformation from $\sem{\Gamma; \Phi \vdash \Delta}$ to
+a morphism from $\sem{\Gamma; \Phi \vdash \Delta}$ to
 $\sem{\Gamma; \Phi \vdash F}$, i.e., a triple of natural
@@
 \end{array}\]
+
 Moreover, for all $\rho : \relenv$, 
@@
 }, \setsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A} },
-\relsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A} }) \in
-[\relenv,RT_0]\]
+\relsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A} })\]
+is an $\omega$-cocontinuous environment transformer.
 \end{prop}

### 2021-07-13 21-lmcs/21-lmcs.tex
MFPS revisions #9

@@
 In addition, to our knowledge no such models represent data types as
-primitive fixpoints. (See the past paragraph of this section for a
-more detailed discussion in the case of F$_\omega$.) Whether or not
-the results of this paper even hold for such calculi is an open
-question.
+primitive fixpoints, so they cannot be used to obtain the results of
+this paper.  (See the past paragraph of this section for a more
+detailed discussion in the case of F$_\omega$.) Whether or not these
+results even hold for such calculi is an open question.
 
@@
 \AXC{$\Gamma;\ol{\alpha^0}  \vdash G$}
-\BIC{$\Gamma;\emptyset \vdash \Nat^{\ol{\alpha^0}}F \,G$}
+\BIC{$\Gamma;\Phi \vdash \Nat^{\ol{\alpha^0}}F \,G$}
 \DisplayProof
@@
 well-formed, then both $\Gamma, \ol{\psi}; \Phi \vdash F$ and $\Gamma;
-\Phi, \ol{\phi} \vdash F$ are also well-formed. Note, however, that
-weakening does not change the contexts in which types can be formed.
-For example, the functorial context in which a type $\Nat^\emptyset F
-\,G$ is formed must be empty, although it can be weakened by a
-functorial variable $\beta$ {\em after} the type is formed; by
-contrast, we cannot {\em first} weaken the functorial contexts of $F$
-and $G$ to $\Gamma; \beta \vdash F$ and $\Gamma; \beta \vdash G$, and
-{\em then} form $\Nat^\emptyset F \,G$ in the weakened context
-$\Gamma; \beta$.  If $\Gamma;\emptyset \vdash F$ and $\Gamma;\emptyset
-\vdash G$, then our rules allow formation of the type
-$\Gamma;\emptyset \vdash \Nat^\emptyset F \,G$, which represents the
-System F type $\Gamma \vdash F \to G$. Similarly, if
-$\Gamma;\ol{\alpha} \vdash F$, then our rules allow formation of the
-type $\Gamma; \emptyset \vdash \Nat^{\ol\alpha} \,\onet \,F$, which
-represents the System F type $\Gamma; \emptyset \vdash \forall
-\ol\alpha . F$. However, some System F types, such as $\forall
-\alpha. (\alpha \to \alpha) \to \alpha$, are not representable in our
-calculus. Note that, in the rule for $\mu$-types, no variables in
-$\Phi$ appear in the body $F$ of $\mu \phi. \lambda \ol\alpha.F$. This
-will be critical to proving the Identity Extension Lemma for our
-calculus.
+\Phi, \ol{\phi} \vdash F$ are also well-formed. Note that weakening
+does not change the contexts in which types can be formed.  For
+example, a $\mathsf{Nat}$-type --- i.e., a type of the form
+$\Nat^{\ol{\alpha}} F \,G$ --- can be formed in any functorial
+context. But since $\Nat$-types contain no functorial variables, we
+will form them in empty functorial contexts below whenever
+convenient. Of course, we cannot {\em first} weaken the functorial
+contexts of $F$ and $G$ to $\Gamma; \ol{\alpha}, \beta \vdash F$ and
+$\Gamma; \ol{\alpha}, \beta \vdash G$, and {\em then} form
+$\Nat^{\ol{\alpha}} F \,G$ in the weakened context $\Gamma; \beta$.
+If $\Gamma;\emptyset \vdash F$ and $\Gamma;\emptyset \vdash G$, then
+our rules allow formation of the type $\Gamma;\emptyset \vdash
+\Nat^\emptyset F \,G$, which represents the System F type $\Gamma
+\vdash F \to G$. Similarly, if $\Gamma;\ol{\alpha} \vdash F$, then our
+rules allow formation of the type $\Gamma; \emptyset \vdash
+\Nat^{\ol\alpha} \,\onet \,F$, which represents the System F type
+$\Gamma; \emptyset \vdash \forall \ol\alpha . F$. However, some System
+F types, such as $\forall \alpha. (\alpha \to \alpha) \to \alpha$, are
+not representable in our calculus. Note that, in the rule for
+$\mu$-types, no variables in $\Phi$ appear in the body $F$ of $\mu
+\phi. \lambda \ol\alpha.F$. This will be critical to proving the
+Identity Extension Lemma for our calculus.
 
@@
 $\phi$. It is not hard to see that $\Gamma;\Phi \vdash H[\phi
-  :=_{\ol{\alpha}} F]$. {\color{red} REMOVE? In addition, if
-  \,$\Gamma, \phi^k; \Phi \vdash H$,\, $\Gamma; \ol\psi,\ol{\alpha}
-  \vdash F$, $|\ol\alpha| = k$, and $\Phi \cap \ol\psi = \emptyset$,
-  then $\Gamma,\ol\psi';\Phi \vdash H[\phi :=_{\ol{\alpha}} F[\ol{\psi
-        :== \psi'}]]$. Ask Daniel.}
+  :=_{\ol{\alpha}} F]$.
 
@@
        \AXC{$\Gamma; \ol{\alpha} \,|\, \Delta, x : F \vdash t: G$} 
-       \TIC{$\Gamma; \emptyset
+       \TIC{$\Gamma; \Phi
          \,|\, \Delta \vdash L_{\ol{\alpha}} x.t : \Nat^{\ol{\alpha}} \,F \,G$}
@@
            G}$}
-       \TIC{$\Gamma; \emptyset
-         ~|~\emptyset
+       \TIC{$\Gamma; \Phi         ~|~\Delta
          \vdash \map^{\ol{F},\ol{G}}_H :
@@
        \AXC{$\Gamma; \phi, \ol{\alpha} \vdash H$}
-       \UIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \tin_H :
+       \UIC{$\Gamma; \Phi  \,|\, \Delta \vdash \tin_H :
          \Nat^{\ol{\beta}} H[\phi :=_{\ol{\beta}} (\mu
@@
        \AXC{$\Gamma; \ol\beta \vdash F$}
-       \BIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \fold^F_H :
+       \BIC{$\Gamma; \Phi  \,|\, \Delta \vdash \fold^F_H :
          \Nat^\emptyset\; (\Nat^{\ol{\beta}}\,H[\phi
@@
 well as all occurrences of $x$ in $t$. In the rule for $t_{\ol K} s$
-there is one type in $\ol K$ for every functorial
-variable in $\ol \alpha$. In the rule for $\map^{\ol{F},\ol{G}}_H$
-there is one type $F$ and one type
-$G$ for each functorial variable in $\ol\phi$. Moreover, for each
-$\phi^k$ in $\ol\phi$ the number of functorial variables in $\ol\beta$
-in the judgments for its corresponding type $F$ and
-$G$ is $k$. In the rules for $\tin_H$ and $\fold^F_H$, the functorial
-variables in $\ol{\beta}$ are fresh with respect to $H$, and there is
-one $\beta$ for every $\alpha$. Substitution for terms is the obvious
-extension of the usual capture-avoiding textual substitution, and the
-rules of Figure~\ref{fig:terms} ensure that weakening is respected.
+there is one type in $\ol K$ for every functorial variable in $\ol
+\alpha$. In the rule for $\map^{\ol{F},\ol{G}}_H$ there is one type
+$F$ and one type $G$ for each functorial variable in
+$\ol\phi$. Moreover, for each $\phi^k$ in $\ol\phi$ the number of
+functorial variables in $\ol\beta$ in the judgments for its
+corresponding type $F$ and $G$ is $k$. In the rules for $\tin_H$ and
+$\fold^F_H$, the functorial variables in $\ol{\beta}$ are fresh with
+respect to $H$, and there is one $\beta$ for every
+$\alpha$. Substitution for terms is the obvious extension of the usual
+capture-avoiding textual substitution, and the rules of
+Figure~\ref{fig:terms} ensure that weakening is respected.  Below we
+form $L$-terms, which contain no free functorial variables, in empty
+functorial contexts whenever convenient. We similarly form $\map$-,
+$\tin$-, and $\fold$-terms, which contain no free functorial variables
+or free term variables, in empty functorial contexts and empty term
+contexts whenever convenient.
 
@@
 \end{array}\]
-respectively. Our calculus can similarly define a
-$\mathtt{reverseBush}$ function that reverses the data in a bush. This
+respectively. Our calculus can similarly define a $\mathtt{rustle}$
+function that changes the placement of the data in a bush. This
 function maps, e.g., the bush
@@
   t [\ol{\alpha := K}][x := s]}\]
-       {\color{red} Add if true:
-$\dsem{\Gamma; \emptyset~|~\Delta \vdash L_{\ol{\alpha}} x.t_{\ol{\alpha}} x :
-   \Nat^{\ol{\alpha}}\, F\, G} =  \dsem{\Gamma; \emptyset~|~\Delta \vdash
-   t : \Nat^{\ol{\alpha}}\, F\, G}$}
+and
+\[\dsem{\Gamma; \emptyset~|~\Delta \vdash L_{\ol{\alpha}} x.t_{\ol{\alpha}} x :
+   \Nat^{\ol{\alpha}}\, F\, G} = \dsem{\Gamma; \emptyset~|~\Delta
+  \vdash t : \Nat^{\ol{\alpha}}\, F\, G}\]
 
@@
 In this section we show how Theorem~\ref{thm:at-gen} and its
-consequences can be used to prove free theorems that go beyond mere
-naturality.  We also show that we can extend short cut fusion for
-lists~\cite{glp93} to nested types, thereby formally proving
-correctness of the categorically inspired theorem from~\cite{jg10}.
+consequences can be used to prove free theorems. Those in
+Sections~\ref{sec:ft-adt},~\ref{sec:short-cut},
+and~\ref{sec:short-cut-nested} go beyond mere naturality.  We also
+show that we can extend short cut fusion for lists~\cite{glp93} to
+nested types, thereby formally proving correctness of the
+categorically inspired theorem from~\cite{jg10}.
 
@@
 
-Let $ \vdash g : \Nat^\alpha \,\onet\,\alpha$, let $g^\set =
-\setsem{\vdash g : \Nat^\alpha \,\onet\,\alpha}$ and $g^\rel =
-\relsem{\vdash g : \Nat^\alpha \,\onet\,\alpha}$.  By
-Theorem~\ref{thm:at-gen}, $(g^\set(\pi_1\rho),g^\set(\pi_2\rho)) =
-g^\rel\rho$. Thus, for all $\rho \in \relenv$ and any $(a, b) \in
-\relsem{\vdash \emptyset}\rho = 1$, eliding the only possible
-instantiations of $a$ and $b$ gives that $(g^\set,g^\set) \;= \;
-(g^\set(\pi_1 \rho), g^\set (\pi_2 \rho))$ $ \in \relsem{\vdash
-  \Nat^\alpha \,\onet\,\alpha}\rho$ $ = \{\eta : K_1 \Rightarrow
-\id\}$ $ = \{(\eta_1 : K_1 \Rightarrow \id, \eta_2 : K_1 \Rightarrow
-\id)\}$ That is, $g^\set$ is a natural transformation from the
-constantly $1$-valued functor to the identity functor in $\set$. In
-particular, for every $S : \set$, $g^\set_S : 1 \to S$. Note, however,
-that if $S = \emptyset$, then there can be no such morphism, so no
-such natural transformation, and thus no term $\vdash g : \Nat^\alpha
-\onet \,\alpha$, can exist.  That is, our calculus admits no
-(non-terminating) terms with the closed type $\Nat^\alpha \onet
-\,\alpha$ of the polymorphic bottom.
+Suppose $ \vdash g : \Nat^\alpha \,\onet\,\alpha$ and $g^\set =
+\setsem{\vdash g : \Nat^\alpha \,\onet\,\alpha}$. Then $g^\set$ is a
+natural transformation from the constantly $1$-valued functor to the
+identity functor in $\set$.  In particular, for every $S : \set$,
+$g^\set_S : 1 \to S$. Note, however, that if $S = \emptyset$, then
+there can be no such morphism, so no such natural transformation, and
+thus no term $\vdash g : \Nat^\alpha \onet \,\alpha$, can exist.  That
+is, our calculus admits no (non-terminating) terms with the closed
+type $\Nat^\alpha \onet \,\alpha$ of the polymorphic bottom.
 
@@
 
-Let $ \vdash g : \Nat^\alpha \,\alpha\,\alpha$, \/$g^\set =
-\setsem{\vdash g : \Nat^\alpha \,\alpha\,\alpha}$, and $g^\rel =
-\relsem{\vdash g : \Nat^\alpha \,\alpha\,\alpha}$.  By
-Theorem~\ref{thm:at-gen}, $(g^\set(\pi_1\rho),g^\set(\pi_2\rho)) =
-g^\rel\rho$. Thus, for all $\rho \in \relenv$ and any $(a, b) \in
-\relsem{\vdash \emptyset}\rho = 1$, eliding the only possible
-instantiations of $a$ and $b$ gives that $(g^\set, g^\set) \; = \;
-(g^\set(\pi_1 \rho), g^\set (\pi_2 \rho))$ $ \in \relsem{\vdash
-  \Nat^\alpha \,\alpha\,\alpha}\rho$ $ = \{\eta : \id \Rightarrow
-\id\}$ $ = \{(\eta_1 : \id \Rightarrow \id, \eta_2 : \id \Rightarrow
-\id)\}$ That is, $g^\set$ is a natural transformation from the
-identity functor on $\set$ to itself. Now let $S$ be any set.  For any
-element $a$ of $S$, if $K_a :S \to S$ is the constantly $a$-valued
-morphism on $S$ then instantiating the naturality square implied by
-the above equality gives that $g^\set_S \circ K_a = K_a \circ
-g^\set_S$, i.e., $g^\set_S \, a = a$, i.e., $g^\set_S = \id_S$.
-Putting these two cases together we have that for every $S : \set$,
-$g^\set_S = \id_S$, i.e., $g^\set$ is the identity natural
-transformation for the identity functor on $\set$. So every closed
-term $g$ of closed type $\Nat^\alpha\alpha\,\alpha$ always denotes the
-identity natural transformation for the identity functor on $\set$,
-i.e., every closed term $g$ of type $\Nat^\alpha\alpha\,\alpha$
-denotes the polymorphic identity function.
+Suppose $ \vdash g : \Nat^\alpha \,\alpha\,\alpha$ and $g^\set =
+\setsem{\vdash g : \Nat^\alpha \,\alpha\,\alpha}$.  Then $g^\set$ is a
+natural transformation from the identity functor on $\set$ to
+itself. If $S$ is any set, if $a$ is any element of $S$, and if $K_a
+:S \to S$ is the constantly $a$-valued morphism on $S$, then
+naturality of $g^\set$ gives that $g^\set_S \circ K_a = K_a \circ
+g^\set_S$, i.e., $g^\set_S \, a = a$, i.e., $g^\set_S = \id_S$.  That
+is, $g^\set$ is the identity natural transformation for the identity
+functor on $\set$. So every closed term $g$ of closed type
+$\Nat^\alpha\alpha\,\alpha$ always denotes the identity natural
+transformation for the identity functor on $\set$, i.e., every closed
+term $g$ of type $\Nat^\alpha\alpha\,\alpha$ denotes the polymorphic
+identity function.
 
@@
 We conclude this subsection by showing, as promised above, that no
-cutting down of set and relational interpretations of $\Lan$-types is
-possible. We have:
+cutting down of set and relational interpretations of $\Lan$-types by
+taking subsets is possible. We have:
 
@@
 Having extended our calculus with both $\Lan$-types and terms of such
-types, and having given the only sensible set and relational
-interpretations for these types and terms, we now need to verify that
-these interpretations give rise to a parametric model. The first step
-in this process is to extend Lemma~\ref{lem:rel-transf-morph} to
+types, and having given sensible set and relational interpretations
+for these types and terms, we now need to verify that these
+interpretations give rise to a parametric model. The first step in
+this process is to extend Lemma~\ref{lem:rel-transf-morph} to
 $\Lan$-types by adding a clause for $\Lan$-types to the proof.
@@
 our substantial and lengthy investigations into the issue lead us to
-conjecture that it is not possible to define a functorial semantics
-for $\Lan$-types that gives rise to parametric models for languages
-supporting even constant-free polynomial primitive GADTs.
+suspect that it may not actually be possible to define a functorial
+semantics for $\Lan$-types that gives rise to parametric models for
+languages supporting even constant-free polynomial primitive GADTs.
 

### 2021-07-14 21-lmcs/21-lmcs.tex
MFPS revisions final

@@
 \address{Appalachian State University}	
-\email{ghiorzie@appstate.edu}
+\email{enrico.ghiorzi@gmail.com}
 
@@
     (((\map_{\mathit{PTree}\,\alpha}^{\alpha \times \alpha, \alpha \times
-      \alpha})_\emptyset\, \mathit{swap})_\alpha\, t')}
-:  \Nat^{\alpha} (\alpha + \mathit{PTree}\,(\alpha \times
+      \alpha})_\emptyset\, \mathit{swap})_\alpha\, t')}\\
+\hspace*{0.5in}:  \Nat^{\alpha} (\alpha + \mathit{PTree}\,(\alpha \times
 \alpha))\; \mathit{PTree}\,\alpha
@@
     F}\rho} \to (\rho'\phi) \ol{\setsem{\Gamma;\Phi \vdash
-    F}\rho'}$ is defined by $\setsem{\Gamma;\Phi \vdash \phi
+    F}\rho'}$ is defined by \[\setsem{\Gamma;\Phi \vdash \phi
   \ol{F}} f = (f\phi)_{\ol{\setsem{\Gamma;\Phi \vdash
@@
       F}f}}\, \circ\, (f \phi)_{\ol{\setsem{\Gamma;\Phi \vdash
-      F}\rho}}$.  The latter equality holds because $\rho\phi$ and
+      F}\rho}}\]  The latter equality holds because $\rho\phi$ and
   $\rho'\phi$ are functors and $f\phi : \rho\phi \to \rho'\phi$ is a
@@
   \vdash H}\,(\pi_2 f)[\phi := \id_{F^2}][\ol{\alpha := \id_B}])$.
-{\color{red} CHECK CONDITIONS} Using $T_H$, we can define the
-functorial action of relational interpretation.
+Using $T_H$, we can define the functorial action of relational
+interpretation.
 \begin{defi}\label{def:rel-sem-funcs}
@@
 \end{equation}
-We prove Equation~\ref{eq:iel-fix-point-intermediate1} by induction on
-$n$.  The case $n=0$ is trivial, because $T^0_{H,\Eq_{\rho}} K_0 =
-K_0$ and $(T^\set_{H,\rho})^0 K_0 = K_0$; the inductive step is proved
-by the following sequence of equalities:
+The case $n=0$ is trivial:
+Equation~\ref{eq:iel-fix-point-intermediate1} and
+Equation~\ref{eq:iel-fix-point-intermediate2} both hold because
+$T^0_{H,\Eq_{\rho}} K_0 = K_0$, $(T^\set_{H,\rho})^0 K_0 = K_0$, and
+Equation~\ref{eq:eq-pres-eq} holds.
+
+The inductive case is proved as follows.  We prove
+Equation~\ref{eq:iel-fix-point-intermediate1} by the following
+sequence of equalities:
 \[
@@
 &= \relsem{\Gamma; \phi, \ol{\alpha} \vdash H} \Eq_{\rho} [\phi
-  := \Eq_{(T^\set_{H,\rho})^n K_0}] \overline{[\alpha :=
+  := \Eq_{(T^\set_{H,\rho})^{n} K_0}] \overline{[\alpha :=
     \Eq_A]} \\ 
 &= \relsem{\Gamma; \phi, \ol{\alpha} \vdash H} \Eq_{\rho [\phi
-    := (T^\set_{H,\rho})^n K_0] \overline{[\alpha :=
+    := (T^\set_{H,\rho})^{n} K_0] \overline{[\alpha :=
       A]}} \\ 
 &= \Eq_{\setsem{\Gamma; \phi, \ol{\alpha} \vdash H} \rho [\phi
-    := (T^\set_{H,\rho})^n K_0] \overline{[\alpha :=
+    := (T^\set_{H,\rho})^{n} K_0] \overline{[\alpha :=
       A]}} \\ 
@@
 \]
-Here, the third equality is by
+Here, the third equality is by the induction hypothesis of
 Equation~\ref{eq:iel-fix-point-intermediate2} for $J = H$, the fifth
@@
 & \relsem{\Gamma; \phi, \ol{\alpha} \vdash \phi
-    \ol{G}} \Eq_{\rho} [\phi := T^{n}_{H,\Eq_{\rho}} K_0]
+    \ol{G}} \Eq_{\rho} [\phi := T^{n+1}_{H,\Eq_{\rho}} K_0]
   \overline{[\alpha := \Eq_A]}
   \\
-&= T^{n}_{H,\Eq_{\rho}} K_0\, \overline{\relsem{\Gamma; 
+&= T^{n+1}_{H,\Eq_{\rho}} K_0\, \overline{\relsem{\Gamma; 
       \phi, \ol{\alpha} \vdash G} \Eq_{\rho} [\phi :=
-      T^{n}_{H,\Eq_{\rho}} K_0] \overline{[\alpha :=
+      T^{n+1}_{H,\Eq_{\rho}} K_0] \overline{[\alpha :=
         \Eq_A]}} \\ 
-&= T^{n}_{H,\Eq_{\rho}} K_0\, \overline{\relsem{\Gamma; 
+&= T^{n+1}_{H,\Eq_{\rho}} K_0\, \overline{\relsem{\Gamma; 
       \phi, \ol{\alpha} \vdash G} \Eq_{\rho} [\phi :=
-      \Eq_{(T^\set_{H,\rho})^{n} K_0}] \overline{[\alpha :=
+      \Eq_{(T^\set_{H,\rho})^{n+1} K_0}] \overline{[\alpha :=
         \Eq_A]}} \\ 
-&= T^{n}_{H,\Eq_{\rho}} K_0\, \overline{\relsem{\Gamma; 
-      \phi, \ol{\alpha} \vdash G} \Eq_{\rho [\phi := (T^\set_{H,\rho})^{n}
+&= T^{n+1}_{H,\Eq_{\rho}} K_0\, \overline{\relsem{\Gamma; 
+      \phi, \ol{\alpha} \vdash G} \Eq_{\rho [\phi := (T^\set_{H,\rho})^{n+1}
         K_0] \overline{[\alpha := A]}}} \\ 
-&= T^{n}_{H,\Eq_{\rho}} K_0\, \overline{\Eq_{\setsem{\Gamma;
+&= T^{n+1}_{H,\Eq_{\rho}} K_0\, \overline{\Eq_{\setsem{\Gamma;
         \phi, \ol{\alpha} \,\vdash G} \rho [\phi :=
-        (T^\set_{H,\rho})^{n} K_0] \overline{[\alpha :=
+        (T^\set_{H,\rho})^{n+1} K_0] \overline{[\alpha :=
           A]}}} \\ 
-&= (\Eq_{(T^\set_{H,\rho})^{n} K_0})^* \,\overline{\Eq_{\setsem{\Gamma;
+&= (\Eq_{(T^\set_{H,\rho})^{n+1} K_0})^* \,\overline{\Eq_{\setsem{\Gamma;
         \phi, \ol{\alpha} \,\vdash G} \rho [\phi :=
-        (T^\set_{H,\rho})^{n} K_0] \overline{[\alpha :=
+        (T^\set_{H,\rho})^{n+1} K_0] \overline{[\alpha :=
           A]}}} \\ 
-&= (\Eq_{(T^\set_{H,\rho})^{n} K_0})^* \overline{\relsem{\Gamma;
+&= (\Eq_{(T^\set_{H,\rho})^{n+1} K_0})^* \overline{\relsem{\Gamma;
       \phi, \ol{\alpha} \vdash G} \Eq_{\rho} [\phi :=
-      \Eq_{(T^\set_{H,\rho})^{n} K_0}] \overline{[\alpha :=
+      \Eq_{(T^\set_{H,\rho})^{n+1} K_0}] \overline{[\alpha :=
         \Eq_A]}} \\ 
 &= \relsem{\Gamma; \phi, \ol{\alpha} \vdash \phi \ol{G}}
-  \Eq_{\rho} [\phi := \Eq_{(T^\set_{H,\rho})^{n} K_0}]
+  \Eq_{\rho} [\phi := \Eq_{(T^\set_{H,\rho})^{n+1} K_0}]
   \overline{[\alpha := \Eq_A]} 
@@
 is by the induction hypothesis for the IEL on the $G$s, and the fifth
-is by the induction hypothesis on $n$ for
-Equation~\ref{eq:iel-fix-point-intermediate1}.
+is by Equation~\ref{eq:iel-fix-point-intermediate1}, which we have
+just proved.
 
@@
     \ol{\beta}.\, G) \ol{K}} 
-  \Eq_{\rho} [\phi := T^{n}_{H,\Eq_{\rho}} K_0]
+  \Eq_{\rho} [\phi := T^{n+1}_{H,\Eq_{\rho}} K_0]
   [\overline{\alpha := \Eq_A}] \\
-&= (\mu {T_{G, \Eq_{\rho}[\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]}})
+&= (\mu {T_{G, \Eq_{\rho}[\phi := T^{n+1}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]}})
   \,\ol{\relsem{\Gamma; \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
-  [\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]} \\ 
-&= \colim{m \in \nat} {T^m_{G, \Eq_{\rho}[\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]}}\,K_0
+  [\phi := T^{n+1}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]} \\ 
+&= \colim{m \in \nat} {T^m_{G, \Eq_{\rho}[\phi := T^{n+1}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]}}\,K_0
   \,(\ol{\relsem{\Gamma; \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
-  [\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]}) \\ 
-&= \colim{m \in \nat} {T^m_{G, \Eq_{\rho}[\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]}}\,K_0
+  [\phi := T^{n+1}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]}) \\ 
+&= \colim{m \in \nat} {T^m_{G, \Eq_{\rho}[\phi := T^{n+1}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]}}\,K_0
   \,(\ol{\relsem{\Gamma; \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
-  [\phi := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha := \Eq_A}]}) \\  %%
-&= \colim{m \in \nat} {T^m_{G, \Eq_{\rho}[\phi := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha := \Eq_A}]}}\,K_0
+  [\phi := \Eq_{(T^{\set}_{H,\rho})^{n+1} K_0}][\overline{\alpha := \Eq_A}]}) \\  %%
+&= \colim{m \in \nat} {T^m_{G, \Eq_{\rho}[\phi := \Eq_{(T^{\set}_{H,\rho})^{n+1} K_0}][\overline{\alpha := \Eq_A}]}}\,K_0
   \,(\ol{\relsem{\Gamma; \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
-  [\phi := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha := \Eq_A}]}) \\ 
-&= (\mu {T_{G, \Eq_{\rho}[\phi := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha := \Eq_A}]}})
+  [\phi := \Eq_{(T^{\set}_{H,\rho})^{n+1} K_0}][\overline{\alpha := \Eq_A}]}) \\ 
+&= (\mu {T_{G, \Eq_{\rho}[\phi := \Eq_{(T^{\set}_{H,\rho})^{n+1} K_0}][\overline{\alpha := \Eq_A}]}})
   \,\ol{\relsem{\Gamma; \phi, \ol{\alpha} \vdash K}\Eq_{\rho}
-  [\phi := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha := \Eq_A}]} \\ 
+  [\phi := \Eq_{(T^{\set}_{H,\rho})^{n+1} K_0}][\overline{\alpha := \Eq_A}]} \\ 
 &= \relsem{\Gamma; \phi, \ol{\alpha} \vdash (\mu \psi.\, \lambda \beta.\, G) \ol{K}}
-  \Eq_{\rho}[\phi := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha := \Eq_{A}}]
+  \Eq_{\rho}[\phi := \Eq_{(T^{\set}_{H,\rho})^{n+1} K_0}][\overline{\alpha := \Eq_{A}}]
 \end{split}
@@
 T^m_{G,
-\Eq_{\rho}[\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha :=
+\Eq_{\rho}[\phi := T^{n+1}_{H,\Eq_{\rho}} K_0][\overline{\alpha :=
     \Eq_A}]}\,K_0 \ = T^m_{G, \Eq_{\rho}[\phi :=
-    \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha :=
+    \Eq_{(T^{\set}_{H,\rho})^{n+1} K_0}][\overline{\alpha :=
       \Eq_A}]}\,K_0
@@
 \[T^0_{G,
-  \Eq_{\rho}[\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha :=
+  \Eq_{\rho}[\phi := T^{n+1}_{H,\Eq_{\rho}} K_0][\overline{\alpha :=
       \Eq_A}]}\,K_0 \ = K_0 = T^0_{G, \Eq_{\rho}[\phi :=
-    \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha :=
+    \Eq_{(T^{\set}_{H,\rho})^{n+1} K_0}][\overline{\alpha :=
       \Eq_A}]}\,K_0\]
@@
 \begin{align*}
-& T^{m+1}_{G, \Eq_{\rho}[\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]}\,K_0 \\
-&= T_{G, \Eq_{\rho}[\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha
-        := \Eq_A}]} (T^{m}_{G, \Eq_{\rho}[\phi := T^{n}_{H,\Eq_{\rho}}
+& T^{m+1}_{G, \Eq_{\rho}[\phi := T^{n+1}_{H,\Eq_{\rho}} K_0][\overline{\alpha := \Eq_A}]}\,K_0 \\
+&= T_{G, \Eq_{\rho}[\phi := T^{n+1}_{H,\Eq_{\rho}} K_0][\overline{\alpha
+        := \Eq_A}]} (T^{m}_{G, \Eq_{\rho}[\phi := T^{n+1}_{H,\Eq_{\rho}}
       K_0][\overline{\alpha := \Eq_A}]}\,K_0) \\
-&= T_{G, \Eq_{\rho}[\phi := T^{n}_{H,\Eq_{\rho}} K_0][\overline{\alpha
+&= T_{G, \Eq_{\rho}[\phi := T^{n+1}_{H,\Eq_{\rho}} K_0][\overline{\alpha
         := \Eq_A}]} (T^{m}_{G, \Eq_{\rho}[\phi :=
-      \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha :=
+      \Eq_{(T^{\set}_{H,\rho})^{n+1} K_0}][\overline{\alpha :=
         \Eq_A}]}\,K_0) \\
 &= \lambda \ol{R}. \relsem{\Gamma; \psi,
-    \ol\beta \vdash G} \Eq_{\rho} [\phi := T^{n}_{H,\Eq_{\rho}}
+    \ol\beta \vdash G} \Eq_{\rho} [\phi := T^{n+1}_{H,\Eq_{\rho}}
     K_0][\overline{\alpha := \Eq_A}][\psi := T^{m}_{G, \Eq_{\rho}[\phi
-        := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha :=
+        := \Eq_{(T^{\set}_{H,\rho})^{n+1} K_0}][\overline{\alpha :=
           \Eq_A}]}\,K_0][\overline{\beta := R}] \\
 &= \lambda \ol{R}. \relsem{\Gamma; \psi,
-    \ol\beta \vdash G} \Eq_{\rho} [\phi := \Eq_{(T^{\set}_{H,\rho})^n
+    \ol\beta \vdash G} \Eq_{\rho} [\phi := \Eq_{(T^{\set}_{H,\rho})^{n+1}
       K_0}][\overline{\alpha := \Eq_A}][\psi := T^{m}_{G,
-      \Eq_{\rho}[\phi := \Eq_{(T^{\set}_{H,\rho})^n
+      \Eq_{\rho}[\phi := \Eq_{(T^{\set}_{H,\rho})^{n+1}
           K_0}][\overline{\alpha := \Eq_A}]}\,K_0][\overline{\beta :=
       R}] \\
-&= T_{G, \Eq_{\rho}[\phi := \Eq_{(T^{\set}_{H,\rho})^n
+&= T_{G, \Eq_{\rho}[\phi := \Eq_{(T^{\set}_{H,\rho})^{n+1}
         K_0}][\overline{\alpha := \Eq_A}]} (T^{m}_{G, \Eq_{\rho}[\phi
-      := \Eq_{(T^{\set}_{H,\rho})^n K_0}][\overline{\alpha :=
+      := \Eq_{(T^{\set}_{H,\rho})^{n+1} K_0}][\overline{\alpha :=
         \Eq_A}]}\,K_0) \\
-&= T^{m+1}_{G, \Eq_{\rho}[\phi := \Eq_{(T^{\set}_{H,\rho})^n
+&= T^{m+1}_{G, \Eq_{\rho}[\phi := \Eq_{(T^{\set}_{H,\rho})^{n+1}
         K_0}][\overline{\alpha := \Eq_A}]}\,K_0
@@
     \setsem{\Gamma; \Phi \vdash G}\rho} = \Eq_{\setsem{\Gamma;
-      \Phi \vdash F \times G}\rho}$
+      \Phi \vdash F \times G}\rho}$\qedhere
 \end{itemize}
-\qed
+
+\vspace*{0.1in}
 
@@
 \id_{\setsem{\Gamma; \Phi \vdash F}\rho} x = x$, so for any $x \in
-\setsem{\Gamma; \Phi \vdash F}\rho$ we have that $(x, \setsem{\Gamma;
-  \Phi \vdash F}f\, x) \in \relsem{\Gamma; \Phi \vdash F}\graph{f}$.
+\setsem{\Gamma; \Phi \vdash F}\rho$ we have $(x, \setsem{\Gamma;
+  \Phi \vdash F}f\, x)$ $\in \relsem{\Gamma; \Phi \vdash F}\graph{f}$.
 Moreover, $x \in \setsem{\Gamma; \Phi \vdash F}\rho$ if and only if
@@
       A}]$, we have that
-  \begin{align*}
-      &(\setsem{\Gamma; \emptyset \,|\, \Delta \vdash L_{\overline{\alpha}}
-      x.t : \Nat^{\overline{\alpha}} \,F \,G}\rho\,d)_{\ol{A}} :  \\
-      &\hspace{1in} \setsem{\Gamma; \ol{\alpha} \vdash F}\rho[\overline{\alpha := A}]
+\[(\setsem{\Gamma; \emptyset \,|\, \Delta \vdash L_{\overline{\alpha}}
+      x.t : \Nat^{\overline{\alpha}} \,F \,G}\rho\,d)_{\ol{A}} :
+\setsem{\Gamma; \ol{\alpha} \vdash F}\rho[\overline{\alpha := A}]
     \to \setsem{\Gamma; \ol{\alpha} \vdash G}\rho[\overline{\alpha :=
-        A}]
-  \end{align*}
+        A}]\]
 but this follows easily from the induction hypothesis.
@@
 \setsem{\Gamma; \ol{\alpha} \vdash G}\rho[\overline{\alpha := \_}]$ is
-clear since $\eta_{\ol{A}} \, = \, \curry\,
+clear since \[\eta_{\ol{A}} \, = \, \curry\,
 (\setsem{\Gamma; \overline{\alpha} \,|\, \Delta, x : F \vdash t:
-  G}\rho[\overline{\alpha := A}])\,d$ is the component at $\ol{A}$ of
+  G}\rho[\overline{\alpha := A}])\,d\] is the component at $\ol{A}$ of
 the partial specialization to $d$ of the natural transformation
@@
   $\setsem{\Gamma;\Phi \vdash G [\overline{\alpha := K}]}$ we 
-  show that, for every $\rho : \setenv$, $\setsem{\Gamma;\Phi \,|\,
-    \Delta \vdash t_{\overline K} s: G [\overline{\alpha := K}]}\rho$
+  show that, for every $\rho : \setenv$, \[\setsem{\Gamma;\Phi \,|\,
+    \Delta \vdash t_{\overline K} s: G [\overline{\alpha := K}]}\rho\]
   is a morphism from $\setsem{\Gamma; \Phi \vdash \Delta}\rho$ to
@@
 \end{array}\]
-The induction hypothesis ensures that $(\setsem{\Gamma; \emptyset \,|\,
+The induction hypothesis ensures that \[(\setsem{\Gamma; \emptyset \,|\,
   \Delta \vdash t : \Nat^{\overline{\alpha}} \,F \,G}\rho\,
-d)_{\overline{\setsem{\Gamma;\Phi \vdash K}\rho}}$ has type
+d)_{\overline{\setsem{\Gamma;\Phi \vdash K}\rho}}\] has type
 $\setsem{\Gamma; \ol{\alpha} \vdash F}\rho[\ol{\alpha :=
@@
   $$\setsem{ \Gamma; \emptyset~|~\emptyset \vdash \fold^F_H :
-    \Nat^\emptyset\;(\Nat^{\ol{\beta}}\,H[\phi
-      :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\;
-    (\Nat^{{\ol{\beta}} }\,(\mu \phi.\lambda \overline
-    \alpha.H)\overline \beta \;F)}$$
-    is a natural transformation from
-  $\setsem{\Gamma;\emptyset \vdash \emptyset}$ to \[\setsem{\Gamma;
-    \emptyset \vdash \Nat^\emptyset\;(\Nat^{\ol{\beta}}\,H[\phi
-      :=_{\ol{\beta}} F][\ol{\alpha := 
-        \beta}]\,F)\; (\Nat^{{\ol{\beta}} }\,(\mu
-    \phi.\lambda \overline \alpha.H)\overline \beta\,F)}\] we need only
-  show that, for all $\rho : \setenv$, the unique $d :
-  \setsem{\Gamma;\emptyset \vdash \emptyset} \rho$, and all $\eta :
+    \Nat^\emptyset\;(\Nat^{\ol{\beta}}\,H[\phi :=_{\ol{\beta}}
+      F][\ol{\alpha := \beta}]\,F)\; (\Nat^{{\ol{\beta}} }\,(\mu
+    \phi.\lambda \overline \alpha.H)\overline \beta \;F)}$$ is a
+  natural transformation from $\setsem{\Gamma;\emptyset \vdash
+    \emptyset}$ to \[\setsem{\Gamma; \emptyset \vdash
+    \Nat^\emptyset\;(\Nat^{\ol{\beta}}\,H[\phi :=_{\ol{\beta}}
+      F][\ol{\alpha := \beta}]\,F)\; (\Nat^{{\ol{\beta}} }\,(\mu
+    \phi.\lambda \overline \alpha.H)\overline \beta\,F)}\] we need
+  only show that, for all $\rho : \setenv$, all $\eta :
   \setsem{\Gamma; \emptyset \vdash \Nat^{\ol{\beta}}\,H[\phi
-      :=_{\ol{\beta}} F][\ol{\alpha := 
-        \beta}]\,F} \rho$,
+      :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F} \rho$, and the
+  unique $d : \setsem{\Gamma;\emptyset \vdash \emptyset} \rho$, 
 \[ \setsem{\Gamma; \emptyset~|~\emptyset \vdash \fold^F_H :
@@
 \end{tikzcd}\]}
-by naturality of $\mathit{fold}_{T^\set_{H,\rho}}\eta$.
-To see that, for all $\rho : \setenv$, $d \in
-\setsem{\Gamma; \emptyset \vdash \emptyset}\rho$, and $\eta :
-\setsem{\Gamma; \emptyset \vdash \Nat^{\ol{\beta}}\,H[\phi :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F}
-\rho$,
+by naturality of $\mathit{fold}_{T^\set_{H,\rho}}\eta$.  To see that,
+for all $\rho : \setenv$, $\eta : \setsem{\Gamma; \emptyset \vdash
+  \Nat^{\ol{\beta}}\,H[\phi :=_{\ol{\beta}} F][\ol{\alpha :=
+      \beta}]\,F} \rho$, and $d \in \setsem{\Gamma; \emptyset \vdash
+  \emptyset}\rho$,
 \[\setsem{\Gamma; \emptyset~|~\emptyset
@@
   \beta\,F)}\,\rho\,d\,\eta\] satisfies the additional condition
-needed for it to be in $\setsem{\Gamma;\emptyset \vdash
+needed to be in $\setsem{\Gamma;\emptyset \vdash
   \Nat^{{\ol{\beta}} }\,(\mu \phi.\lambda \overline
@@
 Since $\eta$ satisfies the 
-additional condition needed for it to be in $\setsem{\Gamma; \emptyset
+additional condition needed to be in $\setsem{\Gamma; \emptyset
   \vdash \Nat^{\ol{\beta}}\,(H[\phi := F][\ol{\alpha :=
@@
 \end{array}\]
-thus gives that $(\hat{g}\;(\mathit{List}\,\setsem{\vdash
+thus gives that \[(\hat{g}\;(\mathit{List}\,\setsem{\vdash
   F})\,\mathit{nil} \,\mathit{cons}, \hat{g} \,\setsem{\vdash H} \,
-n\, c) \in \graph{\mathit{fold}_{1 + \setsem{\vdash F} \times \_}\, n\, c}$, i.e.,
-that $\mathit{fold}_{1 + \setsem{\vdash F} \times \_}\, n\, c\; (\hat{g}\;
+n\, c) \in \graph{\mathit{fold}_{1 + \setsem{\vdash F} \times \_}\,
+  n\, c}\]
+i.e.,
+\[\mathit{fold}_{1 + \setsem{\vdash F} \times \_}\, n\, c\; (\hat{g}\;
 (\mathit{List}\,\setsem{\vdash F})\, \mathit{nil}\, \mathit{cons})
-= \hat{g}\,\setsem{\vdash H}\, n\, c$.
+= \hat{g}\,\setsem{\vdash H}\, n\, c\]
 \qed
@@
 colimit representation of $k$'s domain. That is, we take $k$ to be the
-unique morphism such that, for any $\ol{S : \set_0}$ and $\ol{f :
+unique morphism such that, for any $\ol{S : \set_0}$ and \[\ol{f :
   \setsem{ \emptyset; \ol\alpha \vdash K } (\pi_1 \rho)[\ol{\alpha :=
-      S}] \to \setsem{ \Gamma; \Phi \vdash A } (\pi_1 \rho)}$, we have
+      S}] \to \setsem{ \Gamma; \Phi \vdash A } (\pi_1 \rho)}\] we have
 $k \circ j_{\ol{S},\, \ol{f}} = \pi_1 \iota_{\ol{(S, 0, 0)},\, \ol{(f,
@@
 
+\vspace*{0.1in}
+
+\noindent
+{\small \sc Acknowledgment} We thank Daniel Jeffries for his
+collaboration on~\cite{jgj21}, for his contributions to a preliminary
+version of this paper, and for supporting publication of the present
+version in his absence.
+
 \bibliographystyle{alpha}

### 2021-07-20 21-lmcs/21-lmcs.tex
MFPS revisions final(2)
just a test

@@
 and 
-$\mathit{pleaf}$,$\mathit{pnode}$, $\mathit{swap}$, and $s$ are the terms
+$\mathit{pleaf}$, $\mathit{pnode}$, $\mathit{swap}$, and $s$ are the terms
 \[\begin{array}{l}
@@
 inductively with Definition~\ref{def:set-sem} we can now define our
-relational interpretation.
+relational interpretation. As with the set interpretation, the
+relational interpretation is given in two parts, in
+Definitions~\ref{def:rel-sem} and~\ref{def:rel-sem-funcs}.
 
@@
     \ol{\alpha}. H)\ol{G}}\rho
-  &= (----\mu T_{H,\rho})\ol{\relsem{\Gamma;\Phi \vdash G}\rho}\\
+  &= (\mu T_{H,\rho})\ol{\relsem{\Gamma;\Phi \vdash G}\rho}\\
   \text{where }	T_{H,\rho}
@@
 $\lambda \ol{A}.\,\setsem{\Gamma; \ol{\alpha}\vdash H} \rho
-[\ol{\alpha := A}]$ is a functor in $\ol{\alpha}$ and using
+[\ol{\alpha := A}]$ is a functor in $\ol{A}$ and using
 $\mathit{map}_G\,\ol f$ to denote the action of the semantic functor
@@
 
-Generalized algebraic datatypes (GADTs) --- also known as guarded
+Generalized algebraic data types (GADTs) --- also known as guarded
 recursive data types~\cite{xcc03} or first-class phantom
@@
 restriction to allow the return types of data constructors to be
-different instances of the datatype than the one being defined. For
+different instances of the data type than the one being defined. For
 example, the GADT
@@
 \noindent
-\!\!has data constructors $\mathtt{spair}$ and $\mathtt{sseq}$ with
+has data constructors $\mathtt{spair}$ and $\mathtt{sseq}$ with
 return types $\mathtt{Seq\; (A \times B)}$ and $\mathtt{Seq\;(Nat \to
@@
 as given by the universal property of the left Kan extension.  The
-relational interpretations $\relsem{\Gamma;\emptyset~|~\emptyset
+relational interpretations\\ $\relsem{\Gamma;\emptyset~|~\emptyset
   \vdash \int_{\ol K,F} : \Nat^{\Phi, \ol\alpha}\, F\,
@@
 version of this paper, and for supporting publication of the present
-version in his absence.
+version in his absence. We also thank the anonymous reviewers for
+their detailed and thorough reviews, which have greatly improved the
+paper.
 

### 2021-08-16 induction-paper/revised-induction-paper.tex
added revised induction paper

--- 2021-05-17 induction-paper/ENTCS-GADT-induction.tex
@@
   \author{Enrico Ghiorzi~~~~}%\thanksref{GhiorziEmail}}
-  \author{Daniel Jeffries}%\thanksref{JeffriesEmail}}
+%  \author{Daniel Jeffries}%\thanksref{JeffriesEmail}}
 %  \thanks[JohannEmail]{Email: \href{mailto:johannp@appstate.edu} {\texttt{\normalshape johannp@appstate.edu}}}
@@
 %  \thanks[JeffriesEmail]{Email: \href{mailto:jeffriesd@appstate.edu} {\texttt{\normalshape jeffriesd@appstate.edu}}}
-  \address{$\mathtt{\{johannp,ghiorzie,jeffriesd\}@appstate.edu}$\\
+  \address{$\mathtt{\{johannp,ghiorzie\}@appstate.edu}$\\
     Appalachian State University}
@@
 The induction rule Coq automatically generates for (the analogous Coq
-defintion of) rose trees is
+definition of) rose trees is
 
@@
 (GADTs)~\cite{ch03,pvww06,sp04,xcc03}, more richly indexed
-families~\cite{ch88}, and deep variants of all of these.
+families~\cite{dyb94}, and deep variants of all of these.
 
@@
 following grammar:
-\[\mathsf{F\,G\,\ol{B}\; :=\;
-F_1\,G\,\ol{B} \times F_2\,G\,\ol{B} \ \vert\ F_1\,G\,\ol{B} +
+\[\begin{array}{lll}
+\mathsf{F\,G\,\ol{B}} & := &
+\mathsf{F_1\,G\,\ol{B} \times F_2\,G\,\ol{B} \ \vert\ F_1\,G\,\ol{B} +
 F_2\,G\,\ol{B} \ \vert\ F_1\,\ol{B} \to F_2\,G\,\ol{B}
 \ \vert\ G\,(\ol{F_1\,\ol{B}}) \ \vert\ H\,\ol{B} \ \vert\ H\,
-(\ol{F_1\,G\,\ol{B}}) }\]
+(\ol{F_1\,G\,\ol{B}}) }
+\end{array}\]
 
@@
 rule for $\mathsf{G}$; see Section~\ref{sec:GADT-nested} below. In the
-fifth clause, $\mathsf{H}$ does not contain $\mathsf{G}$. This clause
-therefore subsumes the cases in which $\mathsf{F\,G\,\ol{B}}$ is a
-closed type or one of the $\mathsf{B_i}$. In the sixth clause,
-$\mathsf{H : Set^\gamma \to Set}$ does not contain $\mathsf{G}$
-(although $\ol{\mathsf{F_1\,G\,\ol{B}}}$ can). Moreover, although
-$\mathsf{H}$ can construct any (truly) nested type, it must not
-construct a GADT. This ensures that $\mathsf{H}$ admits functorial
-semantics~\cite{jp20}, and thus has an associated map function. From
-the map function for $\mathsf{H}$ we can also construct a map function
+fifth and sixth clauses, $\mathsf{H : Set^\gamma \to Set}$ is the
+syntactic reflection of some functor, and thus has an associated map
+function. It is worth noting that the fifth clause subsumes the cases
+in which $\mathsf{F\,G\,\ol{B}}$ is a closed type or one of the
+$\mathsf{B_i}$, and that $\mathsf{H}$ can be the data type constructor
+for any (truly) nested type. From the map function for $\mathsf{H}$ we
+can also construct a map function
 
@@
 to define more general predicate liftings is, again, by induction on
-the structure of the types. We do not give a general definition of
-predicate liftings here, though, since that would require us to first
-design a full type calculus, which is beyond the scope of the present
-paper.
+the structure of the types. This ensures that the liftings satisfy the
+crucial property needed to derive deep induction rules, namely that of
+distributing over the type constructors. We do not give a general
+definition of predicate liftings here, though, since that would
+require us to first design a full type calculus, which is beyond the
+scope of the present paper.
 
@@
 deep induction rules, we might expect truly nested GADTs to admit them
-as well. Surprisingly, however, they do not. That is, our results from
-the previous section are the strongest possible. Indeed, the induction
-rule for a data type generally relies on (unary) parametricity of the
-model interpreting it, and deep induction for a truly nested type or a
-truly nested GADT also relies on this interpretation being functorial.
-But, whereas ADTs and nested types both admit functorial parametric
-semantics, proper GADTs cannot admit both functorial and parametric
-semantics at the same time~\cite{jgj21}. In this section we show how
-deep induction for truly nested GADTs goes wrong by analyzing the
-following very simple nested proper GADT:\looseness=-1
+as well. Unfortunately, however, the techniques developed in the
+previous sections do not extend to GADTs. Indeed, the induction rule
+for a data type generally relies on (unary) parametricity of the model
+interpreting it, and deep induction for a truly nested type or a truly
+nested GADT also relies on this interpretation being functorial.
+Whereas ADTs and nested types both admit functorial parametric
+semantics, proper GADTs admit parametric semantics but do not admit
+functorial semantics. In this section we show how the techniques
+developed in this paper for deriving deep induction rules go wrong for
+truly nested GADTs by analyzing the following very simple
+example:\looseness=-1
 
@@
 \end{equation*}
-The constructor $\mathsf{c}$ can be rewritten as $\mathsf{c :
-  \forall\, (B : Set) \to Equal\,A\,(B \times B) \to G\,(G\,B) \to
-  G\,A}$, so the predicate lifting $\mathsf{G^{\wedge} : \forall\, (A
-  : Set) \to (A \to Set) \to G\,A \to Set}$ for $\mathsf{G}$ is
+We acknowledge that $\mathsf{G}$ is semantically equivalent to the
+empty data type, and thus has a trivial (deep) induction principle. We
+could, of course, consider a more realistic counterexample, but this
+would only add notational overhead for no gain in conceptual
+clarity. Indeed, we need only exhibit a single GADT whose deep
+induction rule cannot be obtained using the techniques of this paper.
+
+To see this, we first note that the constructor $\mathsf{c}$ can be
+rewritten as $\mathsf{c : \forall\, (B : Set) \to}$ $\mathsf{Equal\,A\,(B \times
+  B) \to}$ $\mathsf{G\,(G\,B) \to G\,A}$. The predicate lifting
+$\mathsf{G^{\wedge} : \forall\, (A : Set) \to (A \to Set) \to G\,A \to
+  Set}$ for $\mathsf{G}$ is therefore
 \[
-\mathsf{G^{\wedge}\,A\,Q\,(c\,B\,e\,x)
-= \exists\, [Q_A]\,
+\mathsf{G^{\wedge}\,A\,Q_A\,(c\,B\,e\,x)
+= \exists\, [Q_B]\,
 Equal^{\wedge}\,A\,(B \times B)\,Q_A\,(Pair^{\wedge}\,B\,B\,Q_B\,Q_B)\,e
@@
 \[\mathsf{\forall\, (P : \forall\, (A : Set) \to (A \to Set) \to G\,A \to Set)
-\to dIndC\,P \to \forall\, (A : Set)\, (Q : A \to Set)\, (y : G\,a)
+\to dIndC\,P \to \forall\, (A : Set)\, (Q : A \to Set)\, (y : G\,A)
 \to G^{\wedge}\,A\,Q\,y \to P\,A\,Q\,y}\] But if we now try to show
@@
 
-\bibitem{ch88} Coquand, T. and Huet, G. {\em The calculus of
-  constructions}. Information and Computation 76(2/3), 1988.
-
 \bibitem{cis194}
@@
   $\mathtt{https://doi.org/10.5281/zenodo.3744225}$
+
+\bibitem{dyb94} Dybjer, P. {\em Inductive Families}. Formal Aspects of
+  Computing 6(4), pp. 440-465, 1994.
 
@@
   D. Accompanying Agda code for this paper.
-    $\mathtt{https://cs.appstate.edu/\tilde~johannp/FoSSaCS21Code.html}$
+    $\mathtt{https://cs.appstate.edu/\tilde~johannp/MFPS21Code.html}$
 
@@
 
-\bibitem{jgj21} Johann, P., Ghiorzi, E., and Jeffries, D. {\em
-  GADTs, Functoriality, Parametricity: Pick Two}. Submitted, 2021.
+\bibitem{jgj21} Johann, P., Ghiorzi, E., and Jeffries, D. {\em GADTs,
+  Functoriality, Parametricity: Pick Two}. Accepted for presentation
+  at Logical and Semantic Frameworks with Applications, 2021.
 

### 2021-08-17 induction-paper/revised-induction-paper.tex
revised induction paper

@@
 bushes\footnote{To define truly nested types in Agda we must use the
-  $\mathsf{NO\_POSITIVITY\_CHECK}$ flag. A similar flag is required in
-  Coq.}\cite{bm98}:
+  $\mathsf{NO\_POSITIVITY\_CHECK}$ flag, and to define functions over
+  them we must use the $\mathsf{TERMINATING}$ flag.  (Similar flags
+  are required in Coq.) Although as programmers we know from the
+  metatheory in~\cite{jp19} that $\mathsf{Bush}$ is well-defined and
+  the functions we define over them terminate, the flags are necessary
+  because Agda fails to infer these facts. Analogous comments apply at
+  several places below.}\cite{bm98}:
 
@@
 %Section~\ref{sec:ind-equal}.
+
+Under some circumstances deep induction can be mimicked by
+hand-threading applications of structural induction through the layers
+of data comprising a deep data type. But this is not the case when,
+e.g., one or more of the custom predicates in the data type's deep induction
+rule cannot be defined by structural induction.
 
@@
 In this section we use deep induction for the $\mathsf{LTerm}$ GADT
-from~\eqref{eq:eq_lterm} to extract the type from a lambda term.  The
+from~\eqref{eq:eq_lterm} to infer the type from a lambda term. The
 following predicate either returns the type of its input lambda term
@@
 \end{align*}
+Of course, since $\mathsf{GetType}$ is (trivially) defined by
+structural induction, we could perform type inference using
+hand-threaded applications of structural induction as observed at the
+end of Section~\ref{sec:ADTs-and-nesteds}. Nevertheless, the example
+as given nicely illustrates deep induction.
+
+By construction every lambda term in $\mathsf{LTerm}$ is well-typed,
+although that (necessarily unique) type cannot always be inferred.
 The predicate $\mathsf{GetType}$ uses the standard $\mathsf{Maybe}$
-data type to represent failure of type inference. It is defined by:\looseness=-1
+data type to represent failure of type inference. It is defined
+by:\looseness=-1
 

### 2021-08-18 induction-paper/revised-induction-paper.tex
9-18 revisions on mfps submission

@@
 
-It is not possible to extend structural induction to {\em truly}
-nested types, i.e., to nested types whose recursive occurrences appear
+It is not possible to extend structural induction to {\em truly
+nested types}, i.e., to nested types whose recursive occurrences appear
 below themselves. The quintessential example of such a type is that of
@@
 defined. That is, GADTs' constructors' return type instances can, like
-that of $\mathsf{pair}$ in~\eqref{eq:seq}, be structured.\looseness=-1
+that of $\mathsf{pair}$ in~\eqref{eq:seq}, be structured.  For every
+GADT in this paper, we require the instance of the return type for
+each of its data constructors to be a polynomial in that constructor's
+type arguments.\looseness=-1
 %these instances can be structured.
@@
 and the induction rule for $\mathsf{G}$ is
-\[
+\begin{equation}\label{eq:gen-ind-rule}
 \mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to G\,A \to
   Set) \to dIndC\,P \to \forall (A : Set)(Q_A : A \to Set)(y : G\,A)
-\to G^{\wedge}\,A\,Q_A\,y \to P\,A\,Q_A\,y} \]
+  \to G^{\wedge}\,A\,Q_A\,y \to P\,A\,Q_A\,y}
+\end{equation}
 
@@
 
+\vspace*{0.05in}
+
+Observing that the above development essentially uses the equality
+GADT and its predicate lifting in the discrete category of types to
+extend the lifting in~\cite{jp20} --- now specialized to the same
+category --- to GADTs, we have established the following theorem:
+
+\begin{theorem}
+A GADT $\mathsf{G}$ of the form in \eqref{eq:gadts} admits the deep
+induction rule in~\eqref{eq:gen-ind-rule}.
+\end{theorem}
+
 \section{Truly Nested GADTs Need Not Admit Deep Induction
@@
 
-In Sections~\ref{sec:deep-ind-GADTs} and~\ref{sec:framework} we
-derived deep induction rules for GADTs that are not truly
-nested. Since both nested types and GADTs without true nesting admit
-deep induction rules, we might expect truly nested GADTs to admit them
-as well. Unfortunately, however, the techniques developed in the
-previous sections do not extend to GADTs. Indeed, the induction rule
-for a data type generally relies on (unary) parametricity of the model
-interpreting it, and deep induction for a truly nested type or a truly
-nested GADT also relies on this interpretation being functorial.
-Whereas ADTs and nested types both admit functorial parametric
-semantics, proper GADTs admit parametric semantics but do not admit
-functorial semantics. In this section we show how the techniques
-developed in this paper for deriving deep induction rules go wrong for
-truly nested GADTs by analyzing the following very simple
-example:\looseness=-1
+{\em Truly nested GADTs} are proper GADTs whose recursive occurrences
+appear below themselves. In Sections~\ref{sec:deep-ind-GADTs}
+and~\ref{sec:framework} we derived deep induction rules for GADTs that
+are not truly nested GADTs.\footnote{Note carefully the distinction
+  between a GADT that is not a nested type --- i.e., a proper GADT ---
+  and a proper GADT that is not a truly nested GADT.  In fact, truly
+  nested types are not proper GADTs and truly nested GADTs are not
+  (truly) nested types. Confusion is understandable in light of the
+  original, and now well-established, use of the term ``nested type''
+  to refer to types that do not necessarily involve actual
+  nesting~\cite{bm98}.}  Since both nested types and GADTs without
+true nesting admit deep induction rules, we might expect truly nested
+GADTs to admit them as well. Unfortunately, however, the techniques
+developed in the previous sections do not extend to GADTs. Indeed, the
+induction rule for a data type generally relies on (unary)
+parametricity of the model interpreting it, and deep induction for a
+truly nested type or a truly nested GADT also relies on this
+interpretation being functorial.  Whereas ADTs and nested types both
+admit functorial parametric semantics, proper GADTs admit parametric
+semantics but do not admit functorial semantics. In this section we
+show how the techniques developed in this paper for deriving deep
+induction rules go wrong for truly nested GADTs by analyzing the
+following very simple example:\looseness=-1
 

### 2021-08-26 induction-paper/revised-induction-paper.tex
revisions 8-26

--- 2021-08-17 induction-paper/revised-induction-paper.tex
@@
 
-\def\lastname{Johann, Ghiorzi, and Jeffries}
+\def\lastname{Johann and Ghiorzi}
 
@@
 \begin{abstract} 
-  Deep data types are those that are defined in terms of other such
-  data types, including, possibly, themselves.  In that case, they are
-  said to be truly nested.  Deep induction is an extension of
-  structural induction that traverses {\em all} of the structure in a
-  deep data type, propagating predicates on its primitive data
-  throughout the entire structure.  Deep induction can be used to prove
-  properties of nested types, including truly nested types, that
-  cannot be proved via structural induction.  In this paper we show
-  how to extend deep induction to deep GADTs that are not truly
-  nested. We also show that deep induction cannot be extended to truly
-  nested GADTs.\looseness=-1
+  Deep data types are those that are contructed from other data types,
+  including, possibly, themselves.  In this case, they are said to be
+  truly nested.  Deep induction is an extension of structural
+  induction that traverses {\em all} of the structure in a deep data
+  type, propagating predicates on its primitive data throughout the
+  entire structure.  Deep induction can be used to prove properties of
+  nested types, including truly nested types, that cannot be proved
+  via structural induction.  In this paper we show how to extend deep
+  induction to deep GADTs that are not truly nested GADTs. We also
+  show that the techniques developed in this paper do not suffice for
+  extending deep induction to truly nested GADTs.\looseness=-1
 \end{abstract}
@@
 structural induction. It has long been understood, however, that these
-rules are too weak to be genuinely useful for deep ADTs~\cite{jp20},
-i.e., ADTs that are (possibly mutually inductively) defined in terms
-of (other) such ADTs.\footnote{A data type is {\em deep} if it is
-  (possibly mutually inductively) defined in terms of other such data
-  types (including, possibly, itself).}  The following data type of
-rose trees, here coded in Agda and defined in terms of the standard
-type $\mathsf{List}$ of lists (see
+rules are too weak to be genuinely useful for deep
+ADTs~\cite{jp20}.\footnote{An ADT/nested type/GADT is {\em deep} if it
+  is (possibly mutually inductively) defined in terms of other
+  ADTs/nested types/GADTs (including, possibly, itself).}  The
+following data type of rose trees, here coded in Agda and defined in
+terms of the standard type $\mathsf{List}$ of lists (see
 Section~\ref{sec:ADTs-and-nesteds}), is a deep ADT:
 
-\vspace*{-0.2in}
+\vspace*{-0.1in}
 
@@
 
-\pagebreak
-
 {\em Deep induction}~\cite{jp20} is a generalization of structural
@@
   Section~\ref{sec:ADTs-and-nesteds} below. The predicate lifting for
-  a container type like $\mathsf{List\,t}$ or $\mathsf{Rose\,t}$
+  a container type like $\mathsf{List\,A}$ or $\mathsf{Rose\,A}$
   simply traverses containers of that type and applies its predicate
-  argument pointwise to the constituent data of type $\mathsf{t}$.}
+  argument pointwise to the constituent data of type $\mathsf{A}$.
+  The ability to define predicate liftings for more general data types
+  will be critical to deriving their deep induction rules in
+  Section~\ref{sec:framework}.}
+
 Deep induction was also shown in~\cite{jp20} to
@@
 for the $\mathsf{Bush}$ data type~\cite{bm98} and other truly nested
-types. Deep induction for ADTs and nested types is reviewed in
+types. Deep induction for ADTs and (truly) nested types is reviewed in
 Section~\ref{sec:ADTs-and-nesteds} below.
@@
 This paper shows how to extend deep induction to proper GADTs, i.e.,
-to GADTs that are not simply nested types, and thus are not ADTs.  A
+GADTs that go beyond nested types (and thus also beyond ADTs).  A
 constructor for such a GADT $\mathsf{G}$ may, like a constructor for a
 nested type, take as arguments data whose types involve instances of
-$\mathsf{G}$ other than the one being defined --- including instances
-that involve $\mathsf{G}$ itself. But if $\mathsf{G}$ is a proper GADT
-then at least one of its constructors will also have such a structured
-instance of $\mathsf{G}$ --- albeit one not involving $\mathsf{G}$
-itself --- as its codomain.
+$\mathsf{G}$ other than the one being defined. These can even include
+instances involving $\mathsf{G}$ itself. If $\mathsf{G}$ is a proper
+GADT then at least one of its constructors will also have such a
+structured instance of $\mathsf{G}$ --- albeit one not involving
+$\mathsf{G}$ itself --- as its codomain.
 %But while the types of the arguments to a GADT's constructor can
@@
 contrast with the semantics customarily given for ADTs and other
-nested types~\cite{bfss90,gjfor15,jgj21f}, it is not actually possible
+nested types~\cite{bfss90,gjfor15,jgj21f}, it is not at all clear how
 to define a functorial parametric semantics for them~\cite{jgj21}.
@@
 of extending deep induction to GADTs. Indeed, induction can be seen as
-unary parametricity, so we quickly realize that GADTs viewed as their
-functorial completions cannot possibly support induction rules.  This
-makes sense intuitively: induction is a syntactic proof technique, so
-of course it cannot be used to prove properties of those elements of a
+unary parametricity, so GADTs viewed as their functorial completions
+do not obviously support induction rules.  This makes sense
+intuitively: induction is a syntactic proof technique, so it may not
+be possible to use it to prove properties of those elements of a
 GADT's functorial completion that are not expressible in syntax. All
@@
   Henry Ford encoding}~\cite{ch03,hin03,mcb99,sjsv09,sp04}, and {\em
-  ii}) using the predicate liftings for the equality GADT, and any
-other GADTs appearing in the definition of $\mathsf{G}$, to
+  ii}) using the predicate liftings for the equality GADT and the
+other GADTs appearing in the definition of $\mathsf{G}$ to
 appropriately thread the custom predicates for the primitive types
@@
 nested types, which do have a functorial semantics, syntax-only GADTs'
-lack of functoriality means that it is not possible to extend
-induction --- deep or otherwise --- to truly nested GADTs. This does
-not appear to be much of a restriction, however, since GADTs defined
-over themselves do not, to our knowledge, appear in practice or in the
+lack of functoriality means that it is not clear how to extend
+induction --- deep or otherwise --- to {\em truly nested GADTs}, i.e.,
+to proper GADTs whose recursive occurrences appear below
+themselves.\footnote{Note carefully the distinction between a GADT
+  that is not a nested type --- i.e., a proper GADT --- and a proper
+  GADT that is not a truly nested GADT.  In fact, truly nested types
+  are not proper GADTs and truly nested GADTs are not (truly) nested
+  types. Confusion is understandable in light of the original, and now
+  well-established, use of the term ``nested type'' to refer to types
+  that do not necessarily involve actual nesting~\cite{bm98}.}  This
+does not appear to be much of a restriction, however, since truly
+nested GADTs do not, to our knowledge, appear in practice or in the
 literature. Section~\ref{sec:app} comprises a case study in using deep
@@
 code implementing them is available at~\cite{web-page}.
-
-%\subsection{Related Work}
 
@@
 induction rules for nested types. This plugin is also based on unary
-parametricity and true nesting still is not permitted.  As far as we
+parametricity, and true nesting still is not permitted.  As far as we
 know, no attempts have been made to extend either implementation to
-truly nested types or proper GADTs or their deep variants. In fact, we
-know of no work other than that reported here that specifically
+truly nested types or to proper GADTs or their deep variants. In fact,
+we know of no work other than that reported here that specifically
 addresses induction rules for such data types.
@@
 
-It is not possible to extend structural induction to {\em truly}
-nested types, i.e., to nested types whose recursive occurrences appear
-below themselves. The quintessential example of such a type is that of
-bushes\footnote{To define truly nested types in Agda we must use the
-  $\mathsf{NO\_POSITIVITY\_CHECK}$ flag, and to define functions over
-  them we must use the $\mathsf{TERMINATING}$ flag.  (Similar flags
-  are required in Coq.) Although as programmers we know from the
-  metatheory in~\cite{jp19} that $\mathsf{Bush}$ is well-defined and
-  the functions we define over them terminate, the flags are necessary
-  because Agda fails to infer these facts. Analogous comments apply at
-  several places below.}\cite{bm98}:
+Using deep induction we can extend structural induction to {\em truly
+  nested types}, i.e., to nested types whose recursive occurrences
+appear below themselves. The quintessential example of such a type is
+that of bushes\footnote{To define truly nested types in Agda we must
+  use the $\mathsf{NO\_POSITIVITY\_CHECK}$ flag, and to define
+  functions over them we must use the $\mathsf{TERMINATING}$ flag.
+  (Similar flags are required in Coq.) Although as programmers we know
+  from the metatheory in~\cite{jp19} that $\mathsf{Bush}$ is
+  well-defined and the functions we define over them terminate, the
+  flags are necessary because Agda fails to infer these
+  facts. Analogous comments apply at several places
+  below.}\cite{bm98}:
 
@@
 be able to lift the rule's polymorphic predicate argument to
-$\mathsf{Bush}$ itself. The more general observation that an induction
-rule for any truly nested type must therefore necessarily be a deep
-induction rule was, in fact, the original motivation for the
-development of deep induction in~\cite{jp20}. The deep induction rule
-for bushes is
+$\mathsf{Bush}$ itself. This observation was, in fact, the original
+motivation for the development of deep induction in~\cite{jp20}. The
+deep induction rule for bushes is
 
@@
 
-\vspace*{-0.05in}
-
-Although a truly nested type admits only a single induction rule, it
-is worth noting that for those nested types that do admit distinct
-structural induction and deep induction rules, the latter always
-generalize the former. Indeed, the structural induction rule for each
-such nested type is recoverable from its deep induction rule by taking
-the custom predicates on its data of primitive types to be constantly
-$\mathsf{\top}$-valued predicates. This instantiation ensures that the
-resulting induction rule only inspects the top-level structure of its
-argument, rather than the contents of that structure, which is exactly
-what structural induction should do.
+We note that the structural induction rule for any (truly) nested type
+can be obtained as the special case of its deep induction rule in
+which the custom predicates are taken to be constantly
+$\mathsf{\top}$-valued predicates. This instantiation ensures that
+the resulting induction rule only inspects the top-level structure of
+its argument, rather than the contents of that structure, which is
+exactly what structural induction should do.
 %A concrete example of such a derivation is given in
@@
 hand-threading applications of structural induction through the layers
-of data comprising a deep data type. But this is not the case when,
-e.g., one or more of the custom predicates in the data type's deep induction
-rule cannot be defined by structural induction.
+of data comprising a deep data type. But this is not the case if,
+e.g., one or more of the custom predicates in the data type's deep
+induction rule cannot be defined by structural induction.
 
@@
 defined. That is, GADTs' constructors' return type instances can, like
-that of $\mathsf{pair}$ in~\eqref{eq:seq}, be structured.\looseness=-1
+that of $\mathsf{pair}$ in~\eqref{eq:seq}, be structured.  For every
+GADT in this paper, we require that the instance of the return type
+for each of its data constructors is a polynomial in that
+constructor's type arguments.\looseness=-1
 %these instances can be structured.
@@
 
-Neither $\mathsf{Equal}$ nor $\mathsf{Seq}$ is a deep GADT, but the
-following GADT $\mathsf{LTerm}$, inspired by~\cite{cis194}, is.  It
-encodes terms of a simply typed lambda calculus. More robust
-variations on $\mathsf{LTerm}$ are, of course, possible, but this
-variation is rich enough to illustrate all essential aspects of deep
-GADTs --- and later, in Section~\ref{sec:ind-lam}, their deep
-induction rules --- while still being small enough to ensure clarity
-of exposition.
+Although $\mathsf{Seq}$ does not at first glance appear to be a deep
+GADT, when written as its Henry Ford encoding, it, like all GADTs, can
+be regarded as ``deep over $\mathsf{Equal}$''.  By contrast, the
+%Neither $\mathsf{Equal}$ nor $\mathsf{Seq}$ is a deep GADT, but the
+ following GADT $\mathsf{LTerm}$, inspired by~\cite{cis194}, is {\em
+   inherently deep}. It encodes terms of a simply typed lambda
+ calculus. More robust variations on $\mathsf{LTerm}$ are, of course,
+ possible, but this variation is rich enough to illustrate all
+ essential aspects of deep GADTs --- and later, in
+ Section~\ref{sec:ind-lam}, their deep induction rules --- while still
+ being small enough to ensure clarity of exposition.
 
@@
 abstract these to a general framework that can be applied to any deep
-GADT that is not truly nested. As hinted above, the predicate lifting
-for the equality GADT plays a central role in deriving both structural
-and deep induction rules for more general GADTs.
+GADT that is not a truly nested GADT. As hinted above, the predicate
+lifting for the equality GADT plays a central role in deriving both
+structural and deep induction rules for more general GADTs.
 
@@
 hypothesis
-
-\pagebreak
 
@@
 Section~\ref{sec:deep-ind-GADTs} to a general framework for deriving
-deep induction rules for deep GADTs. We will treat deep GADTs of the
-form
+deep induction rules for deep GADTs that are not truly nested
+GADTs. We will treat deep GADTs of the form
 
@@
 predicate liftings. This is not an overly restrictive condition,
-though: all types constructed from sums, products, arrow types and
-type application admit predicate liftings, and so do GADTs constructed
-from the grammar. In fact, we have seen such liftings for products and
-type application in Section~\ref{sec:deep-ind-GADTs}. A concrete way
-to define more general predicate liftings is, again, by induction on
-the structure of the types. This ensures that the liftings satisfy the
-crucial property needed to derive deep induction rules, namely that of
-distributing over the type constructors. We do not give a general
-definition of predicate liftings here, though, since that would
-require us to first design a full type calculus, which is beyond the
-scope of the present paper.
+though: all
+%types constructed from sums, products, arrow types and
+%type application admit predicate liftings, and so do
+GADTs constructed from the above grammar admit predicate liftings.
+(In particular, the lifting for each type constructor $\mathsf{H}$ is
+constructed using its map function.)
+%In fact, we have seen such liftings for
+%products and type application in Section~\ref{sec:deep-ind-GADTs}.
+A concrete way to define more general predicate liftings is, again, by
+induction on the structure of the types in a suitable calculus. This
+will ensure that the liftings satisfy the crucial property needed to
+derive deep induction rules, namely that of distributing over the type
+constructors. We do not give a general definition of predicate
+liftings here, though, since that would require us to first design a
+full type calculus, which is beyond the scope of the present paper.
 
@@
 
+%{\em Truly nested GADTs} are proper GADTs whose recursive occurrences
+%appear below themselves.
 In Sections~\ref{sec:deep-ind-GADTs} and~\ref{sec:framework} we
-derived deep induction rules for GADTs that are not truly
-nested. Since both nested types and GADTs without true nesting admit
+derived deep induction rules for GADTs that are not truly nested
+GADTs. Since both nested types and GADTs without true nesting admit
 deep induction rules, we might expect truly nested GADTs to admit them
 as well. Unfortunately, however, the techniques developed in the
-previous sections do not extend to GADTs. Indeed, the induction rule
-for a data type generally relies on (unary) parametricity of the model
-interpreting it, and deep induction for a truly nested type or a truly
-nested GADT also relies on this interpretation being functorial.
-Whereas ADTs and nested types both admit functorial parametric
-semantics, proper GADTs admit parametric semantics but do not admit
-functorial semantics. In this section we show how the techniques
-developed in this paper for deriving deep induction rules go wrong for
-truly nested GADTs by analyzing the following very simple
-example:\looseness=-1
+previous sections do not extend to truly nested GADTs. Indeed, the
+induction rule for a data type generally relies on (unary)
+parametricity of the model interpreting it, and deep induction for a
+truly nested type or a truly nested GADT also relies on this
+interpretation being functorial.  Whereas ADTs and nested types both
+admit functorial parametric semantics, proper GADTs admit parametric
+semantics but do not admit functorial semantics. In this section we
+show how the techniques developed in this paper for deriving deep
+induction rules go wrong for truly nested GADTs by analyzing the
+following very simple example:\looseness=-1
 
@@
 clarity. Indeed, we need only exhibit a single GADT whose deep
-induction rule cannot be obtained using the techniques of this paper.
+induction rule cannot be obtained using the techniques of this paper,
+and for which more robust techniques will therefore be needed if deep
+induction rules are to be derived for them.
 
@@
   A\,B\,Q_A\,Q_B)\,(Equal^\wedge A\,B\,Q_A'\,Q_B')}$ of course makes
-it impossible to define $\mathsf{G^\wedge Map}$ for more general
+it unclear how to define $\mathsf{G^\wedge Map}$ for more general
 $\mathsf{G}$.
@@
 By construction every lambda term in $\mathsf{LTerm}$ is well-typed,
-although that (necessarily unique) type cannot always be inferred.
-The predicate $\mathsf{GetType}$ uses the standard $\mathsf{Maybe}$
-data type to represent failure of type inference. It is defined
+but that (necessarily unique) type cannot always be inferred.  The
+predicate $\mathsf{GetType}$ uses the standard $\mathsf{Maybe}$ data
+type to represent failure of type inference. It is defined
 by:\looseness=-1
@@
 nested. It also shows that truly nested GADTs, deep or not, do not
-admit (deep) induction rules. Our development is implemented in Agda,
-as is a case study showing how deep induction can prove properties of
-GADTs that are not provable by structural induction.\looseness=-1
+obviously admit (deep) induction rules. Our development is implemented
+in Agda, as is our case study from Section~\ref{sec:app}.
 

### 2021-08-29 induction-paper/revised-induction-paper.tex
revised induction paper 8-29

@@
   via structural induction.  In this paper we show how to extend deep
-  induction to deep GADTs that are not truly nested GADTs. We also
-  show that the techniques developed in this paper do not suffice for
+  induction to GADTs that are not truly nested GADTs. We also show
+  that the techniques developed in this paper do not suffice for
   extending deep induction to truly nested GADTs.\looseness=-1
@@
 structure, propagating suitable predicates to {\em all} of the data it
-contains. With data types becoming ever more advanced, and with deeply
-structured such types becoming increasingly ubiquitous in
-formalizations, it is critically important that proof assistants
-%{\color{red} (e.g., Coq, Agda, Lean)}
-are able to automatically generate genuinely useful induction rules
-for data types that go well beyond traditional ADTs. Such data types
+contains.  With data types becoming ever more advanced, and with
+deeply structured such types becoming increasingly ubiquitous in
+formalizations, such an enhanced notion of induction is essential if
+proof assistants are to be
+%it is critically important that proof assistants
+%%{\color{red} (e.g., Coq, Agda, Lean)}
+%are
+able to automatically generate genuinely useful induction rules for
+data types that go
+%well
+beyond traditional ADTs. These
+%Such data types
 include not just deep ADTs, but also (truly\footnote{A truly nested
@@
 Deep induction was shown in~\cite{jp20} to deliver induction rules
-appropriate to deep nested types, including deep ADTs.  For the deep
-ADT of rose trees, for example, it gives the following genuinely
-useful induction rule:
+appropriate to nested types, including ADTs.  For the (deep) ADT of
+rose trees, for example, it gives the following genuinely useful
+induction rule:
 
@@
   will be critical to deriving their deep induction rules in
-  Section~\ref{sec:framework}.}
-
-Deep induction was also shown in~\cite{jp20} to
+  Section~\ref{sec:framework}.} Deep induction was also shown
+in~\cite{jp20} to
 % be the missing piece making it possible to
@@
 This paper shows how to extend deep induction to proper GADTs, i.e.,
-GADTs that go beyond nested types (and thus also beyond ADTs).  A
-constructor for such a GADT $\mathsf{G}$ may, like a constructor for a
+GADTs that are not nested types (and thus are not ADTs).  A
+constructor for a GADT $\mathsf{G}$ may, like a constructor for a
 nested type, take as arguments data whose types involve instances of
 $\mathsf{G}$ other than the one being defined. These can even include
-instances involving $\mathsf{G}$ itself. If $\mathsf{G}$ is a proper
-GADT then at least one of its constructors will also have such a
+instances involving $\mathsf{G}$ itself. But if $\mathsf{G}$ is a
+proper GADT, then at least one of its constructors will also have a
 structured instance of $\mathsf{G}$ --- albeit one not involving
@@
   explicit tracking of Agda universe levels in this paper.} only
-constructs sequences of pairs, rather than sequences of arbitrary
-type, as does $\mathsf{const}$. If one or more of the constructors for
-a GADT $\mathsf{G}$ return structured instances of $\mathsf{G}$, then
-the GADT will have two distinct, but equally natural, semantics: a
-functorial semantics interpreting it as a left Kan
-extension~\cite{mac71}, and a parametric semantics interpreting it as
-the interpretation of its Church encoding~\cite{atk12,vw10}. As
-detailed in~\cite{jgj21}, a key difference in the two semantics is
-that the former views GADTs as their {\em functorial
-  completions}~\cite{jp19}, and thus as containing more data than just
-those expressible in syntax. By contrast, the latter views them as
-what might be called {\em syntax-only} GADTs. Happily, these two views
-of GADTs coincide for those GADTs that are ADTs or other nested
-types. However, both they and their attendant properties differ
-greatly for proper GADTs. In fact, the functorial and parametric
-semantics for proper GADTs are sufficiently disparate that, by
-contrast with the semantics customarily given for ADTs and other
-nested types~\cite{bfss90,gjfor15,jgj21f}, it is not at all clear how
-to define a functorial parametric semantics for them~\cite{jgj21}.
+constructs sequences of data whose types are pair-structured, rather
+than sequences of arbitrary type, as does $\mathsf{const}$.
+
+If one or more of the constructors for a GADT $\mathsf{G}$ return
+structured instances of $\mathsf{G}$, then the GADT will have two
+distinct, but equally natural, semantics: a functorial semantics
+interpreting it as a left Kan extension~\cite{mac71}, and a parametric
+semantics interpreting it as the interpretation of its Church
+encoding~\cite{atk12,vw10}. As detailed in~\cite{jgj21}, a key
+difference in the two semantics is that the former views GADTs as
+their {\em functorial completions}~\cite{jp19}, and thus as containing
+more data than just those expressible in syntax. By contrast, the
+latter views them as what might be called {\em syntax-only}
+GADTs. Happily, these two views of GADTs coincide for those GADTs that
+are ADTs or (other, including truly) nested types.  However, both they
+and their attendant properties differ greatly for proper GADTs. In
+fact, the functorial and parametric semantics for proper GADTs are
+sufficiently disparate that, by contrast with the semantics
+customarily given for ADTs and nested
+types~\cite{bfss90,gjfor15,jgj21f}, it is not at all clear how to
+define a functorial parametric semantics for GADTs~\cite{jgj21}.
 
@@
 rules --- including deep induction rules --- for GADTs. Indeed, this
-paper gives the first-ever induction rules --- deep or otherwise ---
-for proper GADTs. But it actually delivers far more: it gives a
-general framework for deriving deep induction rules for deep GADTs
-directly from their syntax. This framework can serve as a basis for
-extending modern proof assistants' automatic generation of structural
-induction rules for ADTs to automatic generation of deep induction
-rules for GADTs. In addition, as for ADTs and other nested types, the
-structural induction rule for any GADT can be recovered from its deep
-induction rule simply by taking the custom predicates in its deep
+paper gives the first-ever deep induction rules
+%--- deep or otherwise ---
+for proper GADTs.
+%It also shows how the structural induction rule for
+%a GADT can be derived as an instance of its deep induction rule.
+But it actually delivers far more: it gives a {\em general framework}
+for deriving deep induction rules for GADTs that can be instantiated
+to particular such data type of interest. This framework can serve as
+a basis for extending modern proof assistants' automatic generation of
+structural induction rules for ADTs to automatic generation of deep
+induction rules for GADTs. In addition, as for ADTs and nested types,
+the structural induction rule for any GADT can be recovered from its
+deep induction rule simply by taking the custom predicates in its deep
 induction rule to be constantly $\mathsf{True}$-valued (i.e.,
@@
 simplest --- and arguably most important --- GADT, namely the equality
-GADT. (See Section~\ref{sec:ind-equal}.)  We then derive the deep
-induction rule for a more complex GADT $\mathsf{G}$ by {\em i}) using
-the equality GADT to represent $\mathsf{G}$ as its so-called {\em
-  Henry Ford encoding}~\cite{ch03,hin03,mcb99,sjsv09,sp04}, and {\em
-  ii}) using the predicate liftings for the equality GADT and the
-other GADTs appearing in the definition of $\mathsf{G}$ to
-appropriately thread the custom predicates for the primitive types
-appearing in $\mathsf{G}$ throughout $\mathsf{G}$'s structure. This
-two-step process delivers deep induction rules for a very general
-class of deep GADTs. In Section~\ref{sec:GADTs} we introduce a series
-of increasingly complex GADTs as running examples, and in
-Section~\ref{sec:deep-ind-GADTs} we derive a deep induction rule for
-each of them. In particular, we derive the deep induction rule for the
-$\mathsf{Seq}$ data type in~\eqref{eq:seq} in
-Section~\ref{sec:ind-seq}. We present our general framework for
-deriving (deep) induction rules for (deep) GADTs in
+GADT~\eqref{eq:equal}. We then derive the deep induction rule for a
+more complex GADT $\mathsf{G}$ by {\em i}) using the equality GADT to
+represent $\mathsf{G}$ as its so-called {\em Henry Ford
+  encoding}~\cite{ch03,hin03,mcb99,sjsv09,sp04}, and {\em ii}) using
+the predicate liftings for the equality GADT and the other GADTs
+appearing in the definition of $\mathsf{G}$ to appropriately thread
+the custom predicates for the primitive types appearing in
+$\mathsf{G}$ throughout $\mathsf{G}$'s structure. This two-step
+process delivers deep induction rules for a very general class of
+GADTs. To illustrate, we introduce a series of increasingly complex
+GADTs as running examples in Section~\ref{sec:GADTs} and derive a deep
+induction rule for each of them in
+Section~\ref{sec:deep-ind-GADTs}. In particular, we derive the deep
+induction rules for the equality data type in
+Section~\ref{sec:ind-equal} and the $\mathsf{Seq}$ data type
+in~\eqref{eq:seq} in Section~\ref{sec:ind-seq}. We present our general
+framework for deriving (deep) induction rules for GADTs in
 Section~\ref{sec:framework}, and observe that the derivations in
@@
   are not proper GADTs and truly nested GADTs are not (truly) nested
-  types. Confusion is understandable in light of the original, and now
-  well-established, use of the term ``nested type'' to refer to types
-  that do not necessarily involve actual nesting~\cite{bm98}.}  This
-does not appear to be much of a restriction, however, since truly
-nested GADTs do not, to our knowledge, appear in practice or in the
+  types. There is ample scope for confusion in light of the original,
+  and now well-established, use of the term ``nested type'' to refer
+  to {\em any} type that allows non-variable instances in the domains
+  of its constructors, whether or not that type involves actual
+%  types that do not necessarily involve actual
+  nesting~\cite{bm98}.}  This does not appear to be much of a
+restriction, however, since truly nested GADTs do not, to our
+knowledge, appear in practice or in the
 literature. Section~\ref{sec:app} comprises a case study in using deep
@@
 we know of no work other than that reported here that specifically
-addresses induction rules for such data types.
+addresses induction rules for these data types.
 
@@
 e.g., one or more of the custom predicates in the data type's deep
-induction rule cannot be defined by structural induction.
+induction rule is not the characteristic function of an inductive data
+type.
+%cannot be defined by structural induction.
 
@@
 and $\mathsf{B}$ are instantiated at the same type. If the types
-$\mathsf{A}$ and $\mathsf{B}$ are syntactically identical then the
+$\mathsf{A}$ and $\mathsf{B}$ are {\color{red} propositionally equal}
+%syntactically identical
+then the
 type $\mathsf{Equal\,A\,B}$ contains the single data element
@@
 in Section~\ref{sec:GADTs}. Then, in Section~\ref{sec:framework}, we
-abstract these to a general framework that can be applied to any deep
-GADT that is not a truly nested GADT. As hinted above, the predicate
+abstract these to a general framework that can be applied to any GADT
+that is not a truly nested GADT. As hinted above, the predicate
 lifting for the equality GADT plays a central role in deriving both
@@
 
-To define the (deep) induction rule for any (deep) GADT $\mathsf{G}$
-we first need to define a predicate lifting that maps a predicate on
-a type $\mathsf{A}$ to a predicate on $\mathsf{G\,A}$. Such a predicate
+To define the (deep) induction rule for any GADT $\mathsf{G}$ we first
+need to define a predicate lifting that maps a predicate on a type
+$\mathsf{A}$ to a predicate on $\mathsf{G\,A}$. Such a predicate
 lifting $\mathsf{Equal^{\wedge} : \forall (A\,B : Set) \to (A \to Set)
@@
 \end{equation*}
-
 The deep induction rule for $\mathsf{G}$ now states that, if all of
@@
 Section~\ref{sec:deep-ind-GADTs} to a general framework for deriving
-deep induction rules for deep GADTs that are not truly nested
-GADTs. We will treat deep GADTs of the form
+deep induction rules for GADTs that are not truly nested GADTs. We
+will treat GADTs of the form
 
@@
 type constructors in $\mathsf{\ol{F_1}}$ contains $\mathsf{G}$. This
-prevents nesting, which would make it impossible to give an induction
-rule for $\mathsf{G}$; see Section~\ref{sec:GADT-nested} below. In the
-fifth and sixth clauses, $\mathsf{H : Set^\gamma \to Set}$ is the
-syntactic reflection of some functor, and thus has an associated map
-function. It is worth noting that the fifth clause subsumes the cases
-in which $\mathsf{F\,G\,\ol{B}}$ is a closed type or one of the
-$\mathsf{B_i}$, and that $\mathsf{H}$ can be the data type constructor
-for any (truly) nested type. From the map function for $\mathsf{H}$ we
-can also construct a map function
+prevents nesting, which would {\color{red} make it impossible to give}
+an induction rule for $\mathsf{G}$; see Section~\ref{sec:GADT-nested}
+below. In the fifth and sixth clauses, $\mathsf{H : Set^\gamma \to
+  Set}$ is the syntactic reflection of some functor, and thus has an
+associated map function. It is worth noting that the fifth clause
+subsumes the cases in which $\mathsf{F\,G\,\ol{B}}$ is a closed type
+or one of the $\mathsf{B_i}$, and that $\mathsf{H}$ can be the data
+type constructor for any (truly) nested type. From the map function
+for $\mathsf{H}$ we can also construct a map function
 
@@
 derived deep induction rules for GADTs that are not truly nested
-GADTs. Since both nested types and GADTs without true nesting admit
-deep induction rules, we might expect truly nested GADTs to admit them
-as well. Unfortunately, however, the techniques developed in the
-previous sections do not extend to truly nested GADTs. Indeed, the
-induction rule for a data type generally relies on (unary)
+GADTs. Since both (truly) nested types and GADTs without true nesting
+admit deep induction rules, we might expect truly nested GADTs to
+admit them as well. Unfortunately, however, the techniques developed
+in the previous sections do not extend to truly nested GADTs. Indeed,
+the induction rule for a data type generally relies on (unary)
 parametricity of the model interpreting it, and deep induction for a
@@
 
-This paper extends deep induction to deep GADTs that are not truly
-nested. It also shows that truly nested GADTs, deep or not, do not
-obviously admit (deep) induction rules. Our development is implemented
-in Agda, as is our case study from Section~\ref{sec:app}.
+This paper extends (deep) induction to GADTs that are not truly nested
+GADTs. It also shows that truly nested GADTs do not obviously admit
+(deep) induction rules. Our development is implemented in Agda, as is
+our case study from Section~\ref{sec:app}.
 

### 2021-08-29 induction-paper/revised-induction-paper.tex
small tweaks

@@
 and $\mathsf{B}$ are instantiated at the same type. If the types
-$\mathsf{A}$ and $\mathsf{B}$ are {\color{red} propositionally equal}
-%syntactically identical
-then the
+$\mathsf{A}$ and $\mathsf{B}$ are syntactically identical then the
 type $\mathsf{Equal\,A\,B}$ contains the single data element
@@
 type constructors in $\mathsf{\ol{F_1}}$ contains $\mathsf{G}$. This
-prevents nesting, which would {\color{red} make it impossible to give}
-an induction rule for $\mathsf{G}$; see Section~\ref{sec:GADT-nested}
-below. In the fifth and sixth clauses, $\mathsf{H : Set^\gamma \to
-  Set}$ is the syntactic reflection of some functor, and thus has an
-associated map function. It is worth noting that the fifth clause
-subsumes the cases in which $\mathsf{F\,G\,\ol{B}}$ is a closed type
-or one of the $\mathsf{B_i}$, and that $\mathsf{H}$ can be the data
-type constructor for any (truly) nested type. From the map function
-for $\mathsf{H}$ we can also construct a map function
+prevents nesting, which would make it impossible to give an induction
+rule for $\mathsf{G}$; see Section~\ref{sec:GADT-nested} below. In the
+fifth and sixth clauses, $\mathsf{H : Set^\gamma \to Set}$ is the
+syntactic reflection of some functor, and thus has an associated map
+function. It is worth noting that the fifth clause subsumes the cases
+in which $\mathsf{F\,G\,\ol{B}}$ is a closed type or one of the
+$\mathsf{B_i}$, and that $\mathsf{H}$ can be the data type constructor
+for any (truly) nested type. From the map function for $\mathsf{H}$ we
+can also construct a map function
 

### 2021-09-09 induction-paper/acm-rev-ind-paper.tex
added cpp ind paper

--- 2021-08-29 induction-paper/revised-induction-paper.tex
@@
-\documentclass[9pt]{entcs}
-\usepackage{entcsmacro}
+\documentclass[sigplan,10pt,anonymous,review]{acmart}
+\settopmatter{printfolios=true,printccs=false,printacmref=false}
+
 \usepackage{graphicx}
@@
 \usepackage{amscd}
-\usepackage{MnSymbol}
+%\usepackage{MnSymbol}
 \usepackage{xcolor}
@@
 \EnableBpAbbreviations
-
-\sloppy
-% The following is enclosed to allow easy detection of differences in
-% ascii coding.
-% Upper-case    A B C D E F G H I J K L M N O P Q R S T U V W X Y Z
-% Lower-case    a b c d e f g h i j k l m n o p q r s t u v w x y z
-% Digits        0 1 2 3 4 5 6 7 8 9
-% Exclamation   !           Double quote "          Hash (number) #
-% Dollar        $           Percent      %          Ampersand     &
-% Acute accent  '           Left paren   (          Right paren   )
-% Asterisk      *           Plus         +          Comma         ,
-% Minus         -           Point        .          Solidus       /
-% Colon         :           Semicolon    ;          Less than     <
-% Equals        =3D           Greater than >          Question mark ?
-% At            @           Left bracket [          Backslash     \
-% Right bracket ]           Circumflex   ^          Underscore    _
-% Grave accent  `           Left brace   {          Vertical bar  |
-% Right brace   }           Tilde        ~
-
-\bibliographystyle{entcs}
-
-\def\lastname{Johann and Ghiorzi}
-
 \input{macros}
@@
 \begin{document}
-\begin{frontmatter}
-  \vspace*{-0.1in}
-  \title{(Deep) Induction Rules for GADTs\vspace*{-0.1in}}
-  \author{Patricia Johann~~~~}%\thanksref{JohannEmail}}
-  \author{Enrico Ghiorzi~~~~}%\thanksref{GhiorziEmail}}
-%  \author{Daniel Jeffries}%\thanksref{JeffriesEmail}}
-%  \thanks[JohannEmail]{Email: \href{mailto:johannp@appstate.edu} {\texttt{\normalshape johannp@appstate.edu}}}
-%  \thanks[GhiorziEmail]{Email: \href{mailto:ghiorzie@appstate.edu} {\texttt{\normalshape ghiorzie@appstate.edu}}}
-%  \thanks[JeffriesEmail]{Email: \href{mailto:jeffriesd@appstate.edu} {\texttt{\normalshape jeffriesd@appstate.edu}}}
-  \address{$\mathtt{\{johannp,ghiorzie\}@appstate.edu}$\\
-    Appalachian State University}
-
-\vspace*{-0.1in}
+\title[(Deep) Induction Rules for GADTs]{(Deep) Induction Rules for GADTs}
+  \author{Patricia Johann, Enrico Ghiorzi}
+  \email{johannp,ghiorzie@appstate.edu}
 
@@
   via structural induction.  In this paper we show how to extend deep
-  induction to GADTs that are not truly nested GADTs. We also show
+  induction to GADTs that are not truly nested GADTs.  {\color{red}
+    This opens the way to incorporating automatic generation of (deep)
+    induction rules for them into proof assistants.}  We also show
   that the techniques developed in this paper do not suffice for
-  extending deep induction to truly nested GADTs.\looseness=-1
+  extending deep induction to truly nested GADTs, {\color{red} so more
+    sophisticated techniques are needed to derive deep induction rules
+    for them.}
 \end{abstract}
 
-\end{frontmatter}
+\maketitle
 
@@
 Section~\ref{sec:ADTs-and-nesteds}), is a deep ADT:
-
-\vspace*{-0.1in}
-
 \[\begin{array}{l}
 \mathsf{data\, Rose\, : Set \to Set\,where}\\
-\mathsf{\;\;\;\;\;\;\;\;empty\, :\,\forall \{A : Set\} \to Rose\,A}\\
-\mathsf{\;\;\;\;\;\;\;\;\,node\,\,\,\, :\, \forall \{A : Set\} \to A \to List\,(Rose\,A) \to Rose\,A} 
-\end{array}\]
-\noindent
+\mathsf{\;\;\;empty :\,\forall \{A : Set\} \to Rose\,A}\\
+\mathsf{\;\;\;node\,\,\,\, :\, \forall \{A : Set\} \to A \to List\,(Rose\,A) \to Rose\,A} 
+\end{array}\]
 The induction rule Coq automatically generates for (the analogous Coq
 definition of) rose trees is
-
-\vspace*{-0.075in}
-
-\[\begin{array}{l}
-\mathsf{\forall\, (A : Set)\,(P : Rose\,A \to Set) \to P\,empty \to}\\
+\[\begin{array}{l}
+\mathsf{\forall\, (A : Set)\,(P : Rose\,A \to Set) \to}\\
+\mathsf{\hspace*{0.15in}P\,empty \to}\\
 \mathsf{\hspace*{0.15in}
  (\forall\, (a : A)\,(ts :
-  List\,(Rose\,A)) \to P\,(node\,a\,ts)) \to 
-  \forall \,(x : Rose\,A) \to P\,x}
+  List\,(Rose\,A)) \to P\,(node\,a\,ts)) \to}\\ 
+\hspace*{0.15in}  \mathsf{\forall \,(x : Rose\,A) \to P\,x}
 \end{array}\]
@@
 induction rule:
-
-\vspace*{-0.075in}
-
 \begin{equation}\label{eq:rose}
@@
 \mathsf{\forall \,(A : Set)\,(P : Rose\,A \to Set)\,(Q : A \to Set)
-  \to P\,empty \to}\\ 
+  \to}\\
+\mathsf{\hspace*{0.15in}P\,empty \to}\\ 
 \mathsf{\hspace*{0.15in}(\forall \,(a : A)\, (ts :
-  List\,(Rose\,A))\to Q\,a \to List^\land\,(Rose\,A)\, P\,ts \to P\,(node\,a\,ts)) \to}\\
+  List\,(Rose\,A))\to Q\,a \to}\\
+\mathsf{\hspace*{0.3in}List^\land\,(Rose\,A)\, P\,ts \to P\,(node\,a\,ts)) \to}\\
 \mathsf{\hspace*{0.15in}\forall \,(x :
@@
 \end{equation}
-
-\noindent
-%Here, $\mathsf{List^\land}$ lifts its predicate argument $\mathsf{P}$
-%on data of type $\mathsf{Rose\,a}$ to a predicate on data of type
-%$\mathsf{List\,(Rose\,a)}$ asserting that $\mathsf{P}$ holds for every
-%element of its argument list, and similarly for the lifting
-%$\mathsf{Rose^\land}$ and the predicate $\mathsf{Q}$.
 Here, $\mathsf{List^\land}$ (resp., $\mathsf{Rose^\land}$) lifts its
@@
   associated data type definitions as described in
-  Section~\ref{sec:ADTs-and-nesteds} below. The predicate lifting for
-  a container type like $\mathsf{List\,A}$ or $\mathsf{Rose\,A}$
-  simply traverses containers of that type and applies its predicate
-  argument pointwise to the constituent data of type $\mathsf{A}$.
-  The ability to define predicate liftings for more general data types
-  will be critical to deriving their deep induction rules in
+  Section~\ref{sec:ADTs-and-nesteds}. The predicate lifting for a
+  container type like $\mathsf{List\,A}$ or $\mathsf{Rose\,A}$ simply
+  traverses containers of that type and applies its predicate argument
+  pointwise to the constituent data of type $\mathsf{A}$.  The ability
+  to define predicate liftings for more general data types will be
+  critical to deriving their deep induction rules in
   Section~\ref{sec:framework}.} Deep induction was also shown
@@
 types. Deep induction for ADTs and (truly) nested types is reviewed in
-Section~\ref{sec:ADTs-and-nesteds} below.
+Section~\ref{sec:ADTs-and-nesteds}.
 
@@
 \mathsf{data\, Seq\,: Set \to Set\,where}\\
-\mathsf{\;\;\;\;\;\;\;\;const :\, \forall \{A : Set\} \to A \to Seq\,A}\\
-\mathsf{\;\;\;\;\;\;\;\;pair\;\;\, :\,\forall \{A\,B : Set\} \to Seq \,A \to Seq\,B \to
+\mathsf{\;\;const :\, \forall \{A : Set\} \to A \to Seq\,A}\\
+\mathsf{\;\;pair\;\;\, :\,\forall \{A\,B : Set\} \to Seq \,A \to Seq\,B \to
   Seq\,(A \times B)}
@@
 
-If one or more of the constructors for a GADT $\mathsf{G}$ return
+If one or more of the data constructors for a GADT $\mathsf{G}$ return
 structured instances of $\mathsf{G}$, then the GADT will have two
@@
 the structural induction rule for any GADT can be recovered from its
-deep induction rule simply by taking the custom predicates in its deep
+deep induction rule by taking the custom predicates in its deep
 induction rule to be constantly $\mathsf{True}$-valued (i.e.,
@@
 structural induction rule for the list type
-
-\vspace*{-0.2in}
-
 \begin{equation*}\label{eq:list}
@@
 \mathsf{data\ List : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;nil\,\,\,\,\,\; :\, \forall \{A : Set\} \to List\,A}\\
-\mathsf{\;\;\;\;\;\;\;cons\, :\, \forall \{A : Set\} \to A \to List\,A \to List\,A} 
+\mathsf{\;\;nil\,\,\,\,\,\; :\, \forall \{A : Set\} \to List\,A}\\
+\mathsf{\;\;cons\, :\, \forall \{A : Set\} \to A \to List\,A \to List\,A} 
 \end{array}
 \end{equation*}
-
-\vspace*{-0.1in}
-
-\noindent
 is
-
-\vspace*{-0.3in}
-
 \begin{equation*}
-\mathsf{
-\forall (A : Set) (P : List\,A \to Set)
-\to P\,nil
-\to \big( \forall (a : A) (as: List\,A)
-\to P\,as
-\to P\,(cons\,a\,as)\big)
-\to \forall (as : List\,A)
-\to P\, as
-}
+  \begin{array}{l}
+  \mathsf{\forall (A : Set) (P : List\,A \to Set) \to}\\
+  \mathsf{\hspace*{0.15in}P\,nil \to}\\
+  \mathsf{\hspace*{0.15in}\big( \forall (a : A) (as: List\,A) \to P\,as
+    \to P\,(cons\,a\,as)\big) \to}\\
+  \mathsf{\hspace*{0.15in}\forall (as : List\,A) \to P\, as}
+\end{array}
 \end{equation*}
-
-\vspace*{-0.1in}
-
-\noindent
 As in Coq's induction rule for rose trees, the data inside a structure
@@
 introduction the deep induction rule for lists is
-
-\vspace*{-0.2in}
-
 \[\begin{array}{l}
 \mathsf{\forall (A : Set) (P : List\, A \to Set) (Q : A \to Set)
-\to P\,nil \to \big( \forall (a : A) (as: List\,A) \to Q\,a \to P\,as
+  \to}\\
+\mathsf{\hspace*{0.15in}P\,nil \to}\\
+\mathsf{\hspace*{0.15in}\big( \forall (a : A) (as: List\,A) \to Q\,a \to P\,as
 \to P\,(cons\,a\,as)\big) \to} \\ 
-\quad\mathsf{\forall (as : List\,A) \to List^{\wedge}\,A\,Q\,as
+\mathsf{\hspace*{0.15in}\forall (as : List\,A) \to List^{\wedge}\,A\,Q\,as
   \to P\,as } 
 \end{array}\]
-
-\vspace*{-0.1in}
 
@@
 following type of perfect trees~\cite{bm98}:
-
 \begin{equation*}\label{eq:ptree}
@@
 \mathsf{data\ PTree : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;pleaf\,\,\; :\,\forall \{A : Set\} \to A \to PTree\,A}\\
-\mathsf{\;\;\;\;\;\;\;pnode\, :\,\forall \{A : Set\} \to PTree\,(A \times A) \to PTree\,A} 
+\mathsf{\;\;pleaf\,\,\; :\,\forall \{A : Set\} \to A \to PTree\,A}\\
+\mathsf{\;\;pnode\, :\,\forall \{A : Set\} \to PTree\,(A \times A) \to PTree\,A} 
 \end{array}
@@
 perfect trees, for example, is
-
-\vspace*{-0.2in}
-
-\[\begin{array}{l}
-\mathsf{\forall (P : \forall (A : Set) \to PTree\, A \to Set)
-\to \big( \forall (A : Set) (a : A) \to P\,A\,(pleaf\, a) \big) \to} \\
-\quad\mathsf{\big( \forall (A : Set) (pp : PTree\,(A \times A)) \to
-  P\,(A \times A)\,pp \to P\,A\,(pnode\,pp)\big) \to \forall (A :
-  Set) (p : PTree\,A) \to P\,A\,p } 
-\end{array}\]
-
-\vspace*{-0.1in}
-
-\noindent
-The deep induction rule for perfect trees similarly uses polymorphic
-predicates but otherwise follows the now-familiar pattern:
-
-\vspace*{-0.2in}
-
+\[\begin{array}{l}
+\mathsf{\forall (P : \forall (A : Set) \to PTree\, A \to Set) \to}\\
+\mathsf{\hspace*{0.15in}\big( \forall (A : Set) (a : A) \to P\,A\,(pleaf\, a) \big) \to}\\
+\mathsf{\hspace*{0.15in}\big( \forall (A : Set) (pp : PTree\,(A \times A)) \to}\\
+\mathsf{\hspace*{0.3in}P\,(A \times A)\,pp \to P\,A\,(pnode\,pp)\big) \to}\\
+\mathsf{\hspace*{0.15in}\forall (A : Set) (p : PTree\,A) \to P\,A\,p} 
+\end{array}\]
+
+\begin{figure*}[t]
+
+\begin{adjustbox}{varwidth=7.3in, max width=7.1in, margin=-0.0in 0in
+      -0.2in 0in, fbox, center}
+
+{\small
 \[\begin{array}{l}
@@
   : A \to Set) (p : PTree\,A) \to PTree^{\wedge}\,A\,Q\,p \to
-  P\,A\,Q\,p }
-\end{array}\]
-
-\vspace*{-0.075in}
-
-\noindent
-Here, $\mathsf{Pair^{\wedge} : \forall (A\; B: Set) \to (A \to Set)
-  \to (B \to Set) \to A \times B \to Set}$ lifts predicates
-$\mathsf{Q_A}$ on data of type $\mathsf{A}$ and $\mathsf{Q_B}$ on data
-of type $\mathsf{B}$ to a predicate on pairs of type $\mathsf{A \times
-  B}$ in such a way that $\mathsf{Pair^{\wedge}\,A\,B\,Q_A\,Q_B\,(a,b)
- = Q_A\,a \times Q_B\,b}$. Similarly, $\mathsf{PTree^{\wedge} :
-  \forall (A : Set) \to (A \to Set) \to PTree\,A \to Set}$ lifts a
-predicate $\mathsf{Q}$ on data of type $\mathsf{A}$ to a predicate on
-data of type $\mathsf{PTree\,A}$ asserting that $\mathsf{Q}$ holds for
-every element of type $\mathsf{A}$ contained in its perfect tree
-argument.
-
-Using deep induction we can extend structural induction to {\em truly
-  nested types}, i.e., to nested types whose recursive occurrences
-appear below themselves. The quintessential example of such a type is
-that of bushes\footnote{To define truly nested types in Agda we must
-  use the $\mathsf{NO\_POSITIVITY\_CHECK}$ flag, and to define
-  functions over them we must use the $\mathsf{TERMINATING}$ flag.
-  (Similar flags are required in Coq.) Although as programmers we know
-  from the metatheory in~\cite{jp19} that $\mathsf{Bush}$ is
-  well-defined and the functions we define over them terminate, the
-  flags are necessary because Agda fails to infer these
-  facts. Analogous comments apply at several places
-  below.}\cite{bm98}:
-
-\vspace*{-0.1in}
-
-\begin{equation*}\label{eq:bush}
-\begin{array}{l}
-\mathsf{data\ Bush : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;bnil\,\,\,\,\,\; :\, \forall \{A : Set\} \to Bush\,A}\\
-\mathsf{\;\;\;\;\;\;bcons\, :\, \forall \{A : Set\} \to A \to Bush\,(Bush\,A) \to Bush\,A} 
-\end{array}
-\end{equation*}
-
-\vspace*{0.025in}
-
-\noindent
-Even defining a structural induction rule for bushes requires that we
-be able to lift the rule's polymorphic predicate argument to
-$\mathsf{Bush}$ itself. This observation was, in fact, the original
-motivation for the development of deep induction in~\cite{jp20}. The
-deep induction rule for bushes is
-
-\vspace*{-0.075in}
-
-\[\begin{array}{l}
+  P\,A\,Q\,p }\\
+ \\
 \mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to Bush\, A \to Set)
@@
   Bush^{\wedge}\,A\,Q\,b \to P\,A\,Q\,b } 
-\end{array}\]
-
-\vspace*{-0.05in}
-
-\noindent
-Here, $\mathsf{Bush^{\wedge} : \forall (A : Set) \to (A \to Set) \to
-  Bush\,A \to Set}$ is the following lifting of a predicate
-$\mathsf{Q}$ on data of type $\mathsf{A}$ to a predicate on data of
-type $\mathsf{Bush\,A}$ asserting that $\mathsf{Q}$ holds for every
-element of type $\mathsf{A}$ contained in its argument bush:
-
-\vspace*{-0.15in}
-
-\begin{align*}
-\mathsf{Bush^{\wedge}\,A\,Q\,bnil} &= \mathsf{\top} \\
-\mathsf{Bush^{\wedge}\,A\,Q\,(bcons\,a\,bb)} &= \mathsf{Q\,a \times
-  Bush^{\wedge}\,(Bush\,A)\,(Bush^{\wedge}\,A\,Q)\,bb}
-\end{align*}
+\end{array}\]}
+
+\vspace*{-0.1in}
+
+\caption{Deep induction rules for perfect trees and
+  bushes}\label{fig:ptree-and-bush}
+\end{adjustbox}
+\end{figure*}
+
+The deep induction rule for perfect trees similarly uses polymorphic
+predicates but otherwise follows the familiar pattern. It is given
+by the first expression in Figure~\ref{fig:ptree-and-bush}. There,
+$\mathsf{Pair^{\wedge} : \forall (A\; B: Set) \to (A \to Set) \to (B
+  \to Set)\; \to\;}$ $\mathsf{A \times B \to}$ $\mathsf{Set}$ lifts predicates
+$\mathsf{Q_A}$ on data of type $\mathsf{A}$ and $\mathsf{Q_B}$ on data
+of type $\mathsf{B}$ to a predicate on pairs of type $\mathsf{A \times
+  B}$ in such a way that $\mathsf{Pair^{\wedge}\,A\,B\,Q_A\,Q_B\,(a,b)
+  = Q_A\,a \times Q_B\,b}$. Similarly, $\mathsf{PTree^{\wedge}}$
+$\mathsf{\,: \,\forall (A : Set) \to (A \to Set) \to PTree\,A \to Set}$
+lifts a predicate $\mathsf{Q}$ on data of type $\mathsf{A}$ to a
+predicate on data of type $\mathsf{PTree\,A}$ asserting that
+$\mathsf{Q}$ holds for every element of type $\mathsf{A}$ contained in
+its perfect tree argument.\looseness=-1
+
+Using deep induction we can extend structural induction to {\em truly
+  nested types}, i.e., to nested types whose recursive occurrences
+appear below themselves. The quintessential example of such a type is
+that of bushes\footnote{To define truly nested types in Agda we must
+  use the $\mathsf{NO\_POSITIVITY\_CHECK}$ flag, and to define
+  functions over them we must use the $\mathsf{TERMINATING}$ flag.
+  {\color{red}(Similar flags are required in Coq.)} Although as
+  programmers we know from the metatheory in~\cite{jp19} that
+  $\mathsf{Bush}$ is well-defined and the functions we define over
+  them terminate, the flags are necessary because Agda fails to infer
+  these facts. Analogous comments apply at several places
+  below.}\cite{bm98}:
+\begin{equation*}\label{eq:bush}
+\begin{array}{l}
+\mathsf{data\ Bush : Set \to Set\ where}\\
+\mathsf{\;\;bnil\,\,\,\,\,\; :\, \forall \{A : Set\} \to Bush\,A}\\
+\mathsf{\;\;bcons\, :\, \forall \{A : Set\} \to A \to Bush\,(Bush\,A) \to Bush\,A} 
+\end{array}
+\end{equation*}
+Even defining a structural induction rule for bushes requires that we
+be able to lift the rule's polymorphic predicate argument to
+$\mathsf{Bush}$ itself. This observation was, in fact, the original
+motivation for the development of deep induction in~\cite{jp20}. The
+deep induction rule for bushes is given by the second expression in
+Figure~\ref{fig:ptree-and-bush}, where $\mathsf{Bush^{\wedge} :
+  \forall (A : Set) \to (A \to Set)}$ $\mathsf{\to Bush\,A \to Set}$
+is the following lifting of a predicate $\mathsf{Q}$ on data of type
+$\mathsf{A}$ to a predicate on data of type $\mathsf{Bush\,A}$
+asserting that $\mathsf{Q}$ holds for every element of type
+$\mathsf{A}$ contained in its argument bush:
+\begin{equation}
+\begin{array}{l}
+\mathsf{Bush^{\wedge}\,A\,Q\,bnil} \hspace*{0.49in} = \; \mathsf{\top} \\
+\mathsf{Bush^{\wedge}\,A\,Q\,(bcons\,a\,bb)} \; = \;\\
+\hspace*{0.3in}\mathsf{Q\,a \times Bush^{\wedge}\,(Bush\,A)\,(Bush^{\wedge}\,A\,Q)\,bb} 
+\end{array}
+\end{equation}
 
@@
 type.
-%cannot be defined by structural induction.
 
@@
 constructor's type arguments.\looseness=-1
-%these instances can be structured.
-%
-%Applications of GADTs include generic programming, modeling
-%programming languages via higher-order abstract syntax, maintaining
-%invariants in data structures, and expressing constraints in embedded
-%domain-specific languages. GADTs have also been used, e.g., to
-%implement tagless interpreters~\cite{pl04,pr06,pvww06}, to improve
-%memory performance~\cite{min15}, and to design APIs~\cite{pen20}.
 
@@
 achieved by allowing the programmer to give the type signatures of the
-GADT's data constructors independently, and then taking advantage of
-pattern matching to force the desired type refinement. For example,
-the {\em equality} GADT
+GADT's data constructors independently, and then using pattern
+matching to force the desired type refinement. For example, the {\em
+  equality} GADT
 \begin{equation}\label{eq:equal}
@@
 \mathsf{data\ Equal : Set \to Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;\;\;\;\;\,refl :\, \forall \{A : Set\} \to Equal\,A\,A}
+\mathsf{\;\;refl :\, \forall \{A : Set\} \to Equal\,A\,A}
 \end{array}
@@
 $\mathsf{refl}$. It contains no data elements otherwise.
+
+\begin{figure*}[t]
+
+\begin{adjustbox}{varwidth=7.3in, max width=7.1in, margin=-0.0in 0in
+      -0.2in 0in, fbox, center}
+
+{\small
+\[\begin{array}{l}
+\mathsf{data\ LType : Set \to Set\ where}\\
+\mathsf{\;\;bool :\, \forall \{A : Set\} \to \forall (B : Set) \to Equal\,A\,Bool
+  \to LType\,A}\\ 
+\mathsf{\;\;arr\;\;\; :\, \forall \{A : Set\} \to \forall (B\,C : Set) \to
+  Equal\,A\,(B \to C) \to LType\,B \to LType\,C \to LType\,A}\\ 
+  \mathsf{\;\;list\,\;\; :\,  \forall \{A : Set\} \to\forall (B : Set) \to
+    Equal\,A\,(\List\,B) \to LType\,B \to LType\,A}\\
+  \\
+\mathsf{data\ LTerm : Set \to Set\ where}\\
+\mathsf{\;\;var\,\,\,:\,  \forall \{A : Set\} \to String \to LType\,A \to
+  LTerm\,A} \\  
+\mathsf{\;\;abs\,\, :\,  \forall \{A : Set\} \to \forall (B\,C : Set) \to
+  Equal\,A\,(B \to C) \to String \to LType\,B \to LTerm\,C \to
+  LTerm\,A}\\ 
+  \mathsf{\;\;app\; :\,  \forall \{A : Set\} \to \forall (B : Set) \to
+    LTerm (B \to A) \to LTerm\,B \to LTerm\,A} \\ 
+  \mathsf{\;\;list\;\,\, :\,  \forall \{A : Set\} \to \forall (B : Set) \to
+    Equal\,A\,(List\,B) \to List\,(LTerm\,B) \to LTerm\,A} 
+\end{array}\]}
+
+\vspace*{-0.1in}
+
+\caption{The $\mathsf{LType}$ and $\mathsf{LTerm}$ data
+  types}\label{fig:type-and-term}
+\end{adjustbox}
+\end{figure*}
 
@@
 \mathsf{data\ Seq : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;const :\, \forall \{A : Set\} \to A \to Seq\,A}\\ 
-\mathsf{\;\;\;\;\;\;\;pair\,\;\; :\, \forall \{A : Set\} \to \forall (B\,C : Set) \to
-  Equal\,A\,(B \times C) \to Seq\,B \to Seq\,C \to Seq\,A}\\ 
+\mathsf{\;\;const :\, \forall \{A : Set\} \to A \to Seq\,A}\\ 
+\mathsf{\;\;pair\,\;\; :\, \forall \{A : Set\} \to \forall (B\,C : Set) \to
+  Equal\,A\,(B \times C) \to}\\
+\mathsf{\hspace*{0.7in}Seq\,B \to Seq\,C \to Seq\,A}\\ 
 \end{array}
@@
 %Neither $\mathsf{Equal}$ nor $\mathsf{Seq}$ is a deep GADT, but the
- following GADT $\mathsf{LTerm}$, inspired by~\cite{cis194}, is {\em
-   inherently deep}. It encodes terms of a simply typed lambda
- calculus. More robust variations on $\mathsf{LTerm}$ are, of course,
- possible, but this variation is rich enough to illustrate all
- essential aspects of deep GADTs --- and later, in
- Section~\ref{sec:ind-lam}, their deep induction rules --- while still
- being small enough to ensure clarity of exposition.
+GADT $\mathsf{LTerm}$ in Figure~\ref{fig:type-and-term}, which is
+inspired by~\cite{cis194}, is {\em inherently deep}.  It encodes terms
+of a simply typed lambda calculus. More robust variations on
+$\mathsf{LTerm}$ are, of course, possible, but this variation is rich
+enough to illustrate all essential aspects of deep GADTs --- and
+later, in Section~\ref{sec:ind-lam}, their deep induction rules ---
+while still being small enough to ensure clarity of exposition.
 
 Types are either booleans, arrow types, or list types. They are
-represented by the Henry Ford GADT
-\begin{equation}\label{eq:eq_ltype}
-\begin{array}{l}
-\mathsf{data\ LType : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;\;bool :\, \forall \{A : Set\} \to \forall (B : Set) \to Equal\,A\,Bool
-  \to LType\,A}\\ 
-\mathsf{\;\;\;\;\;\;\;\;arr\;\;\; :\, \forall \{A : Set\} \to \forall (B\,C : Set) \to
-  Equal\,A\,(B \to C) \to LType\,B \to LType\,C \to LType\,A}\\ 
-  \mathsf{\;\;\;\;\;\;\;\;list\,\;\; :\,  \forall \{A : Set\} \to\forall (B : Set) \to
-    Equal\,A\,(\List\,B) \to LType\,B \to LType\,A} 
-\end{array}
-\end{equation}
-Terms are either variables, abstractions, applications, or lists of
-terms. They are similarly represented by
-\begin{equation}\label{eq:eq_lterm}
-\begin{array}{l}
-\mathsf{data\ LTerm : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;\;\;var\,\,\,:\,  \forall \{A : Set\} \to String \to LType\,A \to
-  LTerm\,A} \\  
-\mathsf{\;\;\;\;\;\;\;\;abs\,\, :\,  \forall \{A : Set\} \to \forall (B\,C : Set) \to
-  Equal\,A\,(B \to C) \to String \to LType\,B \to LTerm\,C \to
-  LTerm\,A}\\ 
-  \mathsf{\;\;\;\;\;\;\;\;app\; :\,  \forall \{A : Set\} \to \forall (B : Set) \to
-    LTerm (B \to A) \to LTerm\,B \to LTerm\,A} \\ 
-  \mathsf{\;\;\;\;\;\;\;\;list\;\,\, :\,  \forall \{A : Set\} \to \forall (B : Set) \to
-    Equal\,A\,(List\,B) \to List\,(LTerm\,B) \to LTerm\,A} 
-\end{array}
-\end{equation}
-The type parameter for $\mathsf{LTerm}$ tracks the types of simply
-typed lambda calculus terms. For example, $\mathsf{LTerm\,A}$ is the
-type of simply typed lambda terms of type $\mathsf{A}$. Variables are
-tagged with their types by the data constructors $\mathsf{var}$ and
+represented by the Henry Ford GADT $\mathsf{LType}$.  Terms are either
+variables, abstractions, applications, or lists of terms. They are
+similarly represented by the Henry Ford GADT $\mathsf{LTerm}$.  The
+type parameter for $\mathsf{LTerm}$ tracks the types of simply typed
+lambda calculus terms. For example, $\mathsf{LTerm\,A}$ is the type of
+simply typed lambda terms of type $\mathsf{A}$. Variables are tagged
+with their types by the data constructors $\mathsf{var}$ and
 $\mathsf{abs}$, whose $\mathsf{LType}$ arguments ensure that their
-type tags are legal types. This ensures that all lambda terms
-produced by $\mathsf{var}$, $\mathsf{abs}$, $\mathsf{app}$, and
-$\mathsf{list}$ are well-typed.  We will revisit these GADTs in
+type tags are legal types. This ensures that all lambda terms produced
+by $\mathsf{var}$, $\mathsf{abs}$, $\mathsf{app}$, and $\mathsf{list}$
+are well-typed.  We will revisit these GADTs in
 Sections~\ref{sec:deep-ind-GADTs} and~\ref{sec:app}.
@@
 $\mathsf{A}$ to a predicate on $\mathsf{G\,A}$. Such a predicate
-lifting $\mathsf{Equal^{\wedge} : \forall (A\,B : Set) \to (A \to Set)
-  \to (B \to Set) \to Equal\,A\,B \to Set}$ for $\mathsf{Equal}$ is
-defined by $\mathsf{Equal^{\wedge}\,A\,A\,Q\,Q'\,refl = \forall (a :
-  A) \to Equal\,(Q\,a)(Q'\,a)}$.  It takes two predicates on the same
-type as input and tests them for extensional equality.
-%
-Next, we need to associate with each data constructor $\mathsf{c}$ of
-$\mathsf{G}$ an {\em induction hypothesis} asserting that, if the
-custom predicate arguments to a predicate $\mathsf{P}$ on $\mathsf{G}$
-can be lifted to $\mathsf{G}$ itself, then $\mathsf{c}$ {\em respects}
-$\mathsf{P}$, i.e., $\mathsf{c}$ constructs data satisfying the
-instance of $\mathsf{P}$ at those custom predicates. The following
-induction hypothesis $\mathsf{dIndRefl}$ is thus associated with the
-$\mathsf{refl}$ constructor for $\mathsf{Equal}$:
+lifting
+\[\begin{array}{l}
+\mathsf{Equal^{\wedge} : \forall (A\,B : Set) \to (A \to Set)
+  \to (B \to Set) \to}\\
+\mathsf{\hspace*{0.7in}Equal\,A\,B \to Set}
+\end{array}\]
+for $\mathsf{Equal}$ is defined by
+\[\mathsf{Equal^{\wedge}\,A\,A\,Q\,Q'\,refl = \forall (a :
+  A) \to Equal\,(Q\,a)(Q'\,a)}\]
+It takes two predicates on the same type as input and tests them for
+extensional equality.  Next, we need to associate with each data
+constructor $\mathsf{c}$ of $\mathsf{G}$ an {\em induction hypothesis}
+asserting that, if the custom predicate arguments to a predicate
+$\mathsf{P}$ on $\mathsf{G}$ can be lifted to $\mathsf{G}$ itself,
+then $\mathsf{c}$ {\em respects} $\mathsf{P}$, i.e., $\mathsf{c}$
+constructs data satisfying the instance of $\mathsf{P}$ at those
+custom predicates. The following induction hypothesis
+$\mathsf{dIndRefl}$ is thus associated with the $\mathsf{refl}$
+constructor for $\mathsf{Equal}$:
 \begin{equation*}\label{eq:ind-refl}
@@
 \mathsf{\lambda (P : \forall (A\,B : Set) \to (A \to Set) \to (B \to
-  Set) \to Equal\,A\,B \to Set) \to} \\ 
+  Set) \to}\\
+\mathsf{\hspace*{1.8in}Equal\,A\,B \to Set) \to} \\ 
 \quad\mathsf{\forall (C : Set) (Q\, Q' : C \to Set) \to
-  Equal^{\wedge}\,C\,C\,Q\,Q'\,refl \to P\,C\,C\,Q\,Q'\,refl} 
+  Equal^{\wedge}\,C\,C\,Q\,Q'\,refl \to}\\
+\mathsf{\hspace*{2in}P\,C\,C\,Q\,Q'\,refl} 
 \end{array}
@@
 \mathsf{\forall (P : \forall (A\,B : Set) \to (A \to Set) \to (B \to
-  Set) \to Equal\,A\,B \to Set) \to dIndRefl\,P \to}\\ \quad 
-  \mathsf{\forall (A\,B : Set) (Q_A : A \to Set) (Q_B : B \to Set) (e:
-  Equal\,A\,B) \to Equal^{\wedge}\,A\,B\,Q_A\,Q_B\,e \to
-  P\,A\,B\,Q_A\,Q_B\,e}
+  Set) \to}\\
+\mathsf{\hspace*{1in}Equal\,A\,B \to Set) \to dIndRefl\,P \to}\\ \quad 
+\mathsf{\forall (A\,B : Set) (Q_A : A \to Set) (Q_B : B \to Set)(e:
+  Equal\,A\,B)}\\
+\mathsf{\hspace*{1in}\to Equal^{\wedge}\,A\,B\,Q_A\,Q_B\,e \to
+  P\,A\,B\,Q_A\,Q_B\,e}\\ 
+\mathsf{\hspace*{2in} }
 \end{array}
@@
 $\mathsf{dIndEqual}$ by
-$\mathsf{dIndEqual\;P\;crefl\;A\;A\;Q_A\;Q_A'\;refl\;liftE =
-  crefl\;A\;Q_A\;Q_A'\,liftE}$. To recover $\mathsf{Equal}$'s structural
-induction rule
+\[\mathsf{dIndEqual\;P\;crefl\;A\;A\;Q_A\;Q_A'\;refl\;liftE =
+  crefl\;A\;Q_A\;Q_A'\,liftE}\]
+To recover $\mathsf{Equal}$'s structural induction rule
 \begin{equation}\label{eq:sind-equal}
+\begin{array}{l}
   \mathsf{\forall (Q : \forall (A\,B : Set)
-  \to Equal\,A\,B \to Set) \to \big( \forall (C : Set) \to
-  Q\,C\,C\,refl \big) \to \forall (A\,B : Set) (e: Equal\,A\,B) \to
+  \to Equal\,A\,B \to Set) \to}\\
+\mathsf{\hspace*{0.15in}\big( \forall (C : Set) \to
+  Q\,C\,C\,refl \big) \to}\\
+\mathsf{\hspace*{0.15in}\forall (A\,B : Set) (e: Equal\,A\,B) \to
   Q\,A\,B\,e}
+\end{array}
 \end{equation}
 we define a term $\mathsf{indEqual}$ of the type
-in~\eqref{eq:sind-equal} by $\mathsf{indEqual\;Q\;srefl\;A\;B\;refl =
-  dIndEqual\;P\;srefl'\;A\;B\;
-  K^A_\top\;K^B_\top\;refl\,sliftE}$. Here, $\mathsf{P : \forall (A\,B
-  : Set) \to (A \to Set) \to (B \to Set) \to Equal\,A\,B \to Set}$ is
-defined by $\mathsf{P\;A\;B\;Q_A\;Q_B\;e =}$ $\mathsf{Q\;A\;B\;e}$,
+in~\eqref{eq:sind-equal} by $\mathsf{indEqual\;Q}$
+$\mathsf{srefl\;A\;B\;refl = dIndEqual\;P\;srefl'\;A\;B\;
+  K^A_\top\;K^B_\top\;refl\,sliftE}$.\\
+Here,
+\[\begin{array}{l}
+\mathsf{P : \forall
+  (A\,B : Set) \to (A \to Set) \to (B \to Set) \to}\\
+\mathsf{\hspace*{2in}Equal\,A\,B} \mathsf{\to Set}
+\end{array}\]
+is defined by $\mathsf{P\;A\;B\;Q_A\;Q_B\;e =}$ $\mathsf{Q\;A\;B\;e}$,
 $\mathsf{K^A_\top}$ and $\mathsf{K^B_\top}$ are the constantly
@@
 respectively, $\mathsf{sliftE :}$ $\mathsf{Equal^{\wedge}\;A\;B\;
-  K^A_\top\;K^B_\top\;refl}$ is defined by $\mathsf{sliftE\,a = refl :
-  Equal\, \top\,\top}$ for every $\mathsf{a : A}$, and $\mathsf{srefl'
-  : \forall (C\,: Set) (Q_c\,Q'_c\, : C \to Set) \to
-  Equal^{\wedge}\,C\,C\,Q_c\,Q'_c\,refl \to \,Q\,C\,C\,refl}$ is
+  K^A_\top\;K^B_\top\;refl}$ is defined by \[\mathsf{sliftE\,a =}
+\mathsf{refl : Equal\, \top\,\top}\] for every $\mathsf{a : A}$, and
+\begin{equation*}
+\begin{array}{l}
+\mathsf{srefl' : \forall (C\,: Set) (Q_c\,Q'_c\, : C \to Set) \to}\\
+\mathsf{\hspace*{0.75in} Equal^{\wedge}\,C\,C\,Q_c\,Q'_c\,refl \to \,Q\,C\,C\,refl}
+\end{array}
+\end{equation*}
+is
 defined by $\mathsf{srefl'\,C\,Q_c\,Q'_c\,liftE'\,=srefl\, C}$.  The
@@
 
-To derive the deep induction rule for the GADT $\mathsf{Seq}$ we use
-its Henry Ford encoding from~\eqref{eq:eq_seq}.  We first define its
-predicate lifting $\mathsf{Seq^\wedge : \forall (A : Set) \to (A \to
-  Set) \to Seq\,A \to Set}$ by
-
-\vspace*{-0.2in}
-
+\begin{figure*}[t]
+
+\begin{adjustbox}{varwidth=7.3in, max width=7.1in, margin=-0.0in 0in
+      -0.2in 0in, fbox, center}
+
+{\small
 \[\begin{array}{lll}
@@
   \times C)\, Q_A\, (Pair^\wedge\,B\,C\, Q_B \, Q_C) \, e \times
-  Seq^{\wedge}\,B\,Q_B\,s_B \times Seq^{\wedge}\,C\,Q_C\,s_C}
-\end{array}\]
-Here, $\mathsf{a : A}$, $\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C : C
-  \to Set}$, $\mathsf{e : Equal\,A\,(B \times C)}$, $\mathsf{s_B :
+  Seq^{\wedge}\,B\,Q_B\,s_B \times Seq^{\wedge}\,C\,Q_C\,s_C}\\
+& & \\
+\mathsf{dIndConst} & = & \mathsf{\lambda (P : \forall (A : Set) \to (A
+  \to Set) \to Seq\,A \to Set) \to}\\
+ & & \mathsf{\hspace*{0.15in}
+\forall (A : Set) (Q_A : A \to Set) (a : A) \to Q_A\,a \to
+P\,A\,Q_A\,(const\,a)}\\
+& & \\
+\mathsf{dIndPair} & = & \mathsf{\lambda (P : \forall (A : Set) \to (A
+  \to Set) \to Seq\,A \to Set)} \to \\ 
+ & & \quad \mathsf{\forall (A\,B\,C : Set) (Q_A : A \to Set) (Q_B : B
+  \to Set) (Q_C : C \to Set)}\\
+& & \mathsf{\hspace*{1.5in}(s_B : Seq\,B) (s_C : Seq\,C) (e :
+  Equal\,A\,(B \times C)) \to} \\ 
+& & \quad \mathsf{Equal^{\wedge} A\, (B \times C)\, Q_A\,
+  (Pair^{\wedge}\,B\,C\,Q_B\,Q_C)\, e \to P\,B\,Q_B\,s_B \to}\\
+& & \mathsf{\hspace*{1.5in}  P\,C\,Q_C\,s_C \to P\, A\, Q_A\, (
+  pair\,B\,C\,e\,s_B\,s_C )} 
+\end{array}\]
+
+\vspace*{0.1in}
+
+\[\begin{array}{l}
+\hspace*{-2in}\mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to Seq\,A \to
+  Set)} \mathsf{\to dIndConst\,P \to dIndPair\,P \to} \\ \hspace*{-2in}\quad
+\mathsf{\forall (A : Set)(Q_A : A \to Set)(s_A : Seq\,A) \to
+  Seq^{\wedge}\,A\,Q_A\,s_A \to P\,A\,Q_A\,s_A}
+\end{array}\]}
+
+\vspace*{-0.1in}
+
+\caption{Deep induction rule for $\mathsf{Seq}$}\label{fig:seq}
+\end{adjustbox}
+\end{figure*}
+
+To derive the deep induction rule for the GADT $\mathsf{Seq}$ we use
+its Henry Ford encoding from~\eqref{eq:eq_seq}.  We first define its
+predicate lifting $\mathsf{Seq^\wedge : \forall (A : Set) \to (A \to
+  Set) \to Seq\,A \to Set}$ as in Figure~\ref{fig:seq}. There,
+$\mathsf{a : A}$, $\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C : C \to
+  Set}$, $\mathsf{e : Equal\,A\,(B \times C)}$, $\mathsf{s_B :
   Seq\,B}$, $\mathsf{s_C : Seq\,C}$, and $\mathsf{\exists [x]\, F
@@
 $\mathsf{(x,b)}$, where $\mathsf{x : A}$, $\mathsf{b : F\, x}$, and
-$\mathsf{F : A \to Set}$.
-
-Next, let $\mathsf{dIndConst}$ be the induction hypothesis
-\[\mathsf{
-\lambda (P : \forall (A : Set) \to (A \to Set) \to Seq\,A \to Set) \to
-\forall (A : Set) (Q_A : A \to Set) (a : A) \to Q_A\,a \to
-P\,A\,Q_A\,(const\,a)}\] associated with the constructor
-$\mathsf{const}$, and let $\mathsf{dIndPair}$ be the induction
-hypothesis
-
-\[\begin{array}{l}
-\mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to Seq\,A \to
-  Set)} \to \\ 
-\quad \mathsf{\forall (A\,B\,C : Set) (Q_A : A \to Set) (Q_B : B
-  \to Set) (Q_C : C \to Set)(s_B : Seq\,B) (s_C : Seq\,C) (e :
-  Equal\,A\,(B \times C)) \to} \\ 
-\quad \mathsf{Equal^{\wedge} A\, (B \times C)\, Q_A\,
-  (Pair^{\wedge}\,B\,C\,Q_B\,Q_C)\, e \to P\,B\,Q_B\,s_B \to
-  P\,C\,Q_C\,s_C \to P\, A\, Q_A\, ( pair\,B\,C\,e\,s_B\,s_C )}
-\end{array}\]
-associated with the constructor $\mathsf{pair}$. Then the deep
-induction rule for $\mathsf{Seq}$ is
-\begin{equation}\label{eq:ind-seq}
-\begin{array}{l}
-\mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to Seq\,A \to
-  Set)} \mathsf{\to dIndConst\,P \to dIndPair\,P \to} \\ \quad
-\mathsf{\forall (A : Set)(Q_A : A \to Set)(s_A : Seq\,A) \to
-  Seq^{\wedge}\,A\,Q_A\,s_A \to P\,A\,Q_A\,s_A}
-\end{array}
-\end{equation}
+$\mathsf{F : A \to Set}$. Next, let $\mathsf{dIndConst}$ and
+$\mathsf{dIndPair}$ be the induction hypotheses associated with the
+constructors $\mathsf{const}$ and $\mathsf{pair}$, respectively. These
+are given in Figure~\ref{fig:seq} well. Then the deep induction rule
+for $\mathsf{Seq}$ is given in the last two lines of
+Figure~\ref{fig:seq}.
 
 To prove that this rule is sound we provide a witness
-$\mathsf{dIndSeq}$ inhabiting the type in~\eqref{eq:ind-seq} as
-follows:
-\[\begin{array}{lll}
-\mathsf{dIndSeq\;P\;cconst\;cpair\;A\;Q_A\;(const\,a)\;liftA}  &
-    = &  \mathsf{cconst\;A\;Q_A\;a\;liftA}\\
-\mathsf{dIndSeq\,P\,cconst\,cpair\,A\,Q_A\,(pair\,B\,C\,e\,s_B\,s_C)\,
-    (Q_B,Q_C, liftE, liftB, liftC)}  & = &
-\mathsf{cpair\,A\,B\,C\,Q_A\,Q_B\,Q_C\,s_B\,s_C\,e\,liftE\,p_B\,p_C}
-\end{array}\]
-In the first clause, $\mathsf{a : A}$, $\mathsf{Q_A : A \to Set}$, and
-$\mathsf{liftA : Seq^{\wedge}\,A\,Q_A\,(const\,a) = Q_A\,a}$. In the
-second clause we also have $\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C : C \to
-  Set}$, $\mathsf{e : Equal\,A\,(B \times C)}$, $\mathsf{s_B :
-  Seq\,B}$, $\mathsf{s_C : Seq\,C}$, $\mathsf{liftE :
-  Equal^{\wedge}\,A\, (B \times C)\, Q_A\, (Pair^\wedge \,B\,C\,Q_B \,
-  Q_C) \, e}$, $\mathsf{liftB : Seq^{\wedge}\,B\,Q_B\,s_B}$, and
-$\mathsf{liftC :}$ $\mathsf{Seq^{\wedge}\,C\,Q_C\,s_C}$ --- which
-together ensure that $\mathsf{(Q_B, Q_C, liftE, liftB, liftC) :
-  Seq^{\wedge}\,A\,Q\,(pair\,B\,C\,e\,s_B\,s_C)}$ --- and
-$\mathsf{p_B} =\mathsf{dIndSeq\,P\,cconst\,cpair\,B\,Q_B\,s_B\,liftB :
-  P\,B\,Q_B\,s_B}$ and $\mathsf{p_C}
-=\mathsf{dIndSeq\,P\,cconst\,cpair\,C\,Q_C\,s_C\,liftC :
-  P\,C\,Q_C\,s_C}$.
+$\mathsf{dIndSeq}$ inhabiting the type in the last two lines of
+Figure~\ref{fig:seq} by
+\[\begin{array}{ll}
+& \!\!\mathsf{dIndSeq\;P\;cconst\;cpair\;A\;Q_A\;(const\,a)\;liftA}\\
+= & \!\!\mathsf{cconst\;A}$ $\mathsf{Q_A\;a\;liftA}
+\end{array}\]
+and
+\[\begin{array}{ll}
+ &\!\!\mathsf{dIndSeq\,P\,cconst\,cpair\,A\,Q_A\,(pair\,B\,C\,e\,s_B\,s_C)}\\
+ & \hspace*{1.4in}\mathsf{(Q_B, Q_C, liftE, liftB, liftC)}\\
+= &\!\! \mathsf{cpair\,A\,B\,C\,Q_A\,Q_B\,Q_C\,s_B\,s_C\,e\,liftE\,
+  p_B\,p_C}
+\end{array}\]
+In the first clause, $\mathsf{a : A}$, $\mathsf{Q_A : A
+  \to Set}$, and $\mathsf{liftA :}$
+$\mathsf{Seq^{\wedge}\,A\,Q_A}$ $\mathsf{(const\,a) = Q_A\,a}$. In the second
+clause we also have $\mathsf{Q_B : B \to}$ $\mathsf{Set}$, $\mathsf{Q_C : C \to
+  Set}$, $\mathsf{e \,:\, Equal\,A\,(B \times C)}$,\, $\mathsf{s_B :}$
+$\mathsf{Seq\,B}$, $\mathsf{s_C : Seq\,C}$,\, $\mathsf{liftE :
+  Equal^{\wedge}\,A\,}$ $\mathsf{(B \times C)\, Q_A\, (Pair^\wedge
+  \,B\,C\,Q_B}$ $\mathsf{Q_C) \, e}$, $\mathsf{liftB :
+  Seq^{\wedge}\,B}$ $\mathsf{Q_B\,s_B}$, and $\mathsf{liftC}$ $\mathsf{ :
+  Seq^{\wedge}\,C\,Q_C\,s_C}$ --- which together give that
+$\mathsf{(Q_B, Q_C, liftE,}$ $\mathsf{liftB, liftC) :
+  Seq^{\wedge}\,A\,Q\,(pair}$ $\mathsf{B\,C\,e\,s_B\,s_C)}$ --- and
+$\mathsf{p_B} = \mathsf{dIndSeq\,P}$
+$\mathsf{cconst\,cpair\,B\,Q_B\,s_B\,liftB :}$
+$\mathsf{P\,B\,Q_B\,s_B}$ and $\mathsf{p_C} =\mathsf{dIndSeq\,P}$
+$\mathsf{cconst\,cpair\,C\,Q_C\,s_C\,liftC : P}$
+$\mathsf{C\,Q_C\,s_C}$.
 
@@
 
-  \begin{adjustbox}{varwidth=6.8in, max width=6.6in, margin=-0.1in 0in
+%  \begin{adjustbox}{varwidth=6.8in, max width=6.6in, margin=-0.1in 0in
+%      -0.2in 0in, fbox, center}
+
+  \begin{adjustbox}{varwidth=7.3in, max width=7.1in, margin=-0.0in 0in
       -0.2in 0in, fbox, center}
+
   {\small
@@
 To derive the deep induction rule for the GADT $\mathsf{LTerm}$ we use
-its Henry Ford encoding from~\eqref{eq:eq_ltype}
-and~\eqref{eq:eq_lterm}. We first define the predicate lifting
-$\mathsf{Arr^{\wedge} : \forall (A\, B : Set) \to (A \to Set) \to (B
-  \to Set) \to (A \to B) \to Set}$ for arrow types, since arrow types
+its Henry Ford encoding from Figure~\ref{fig:type-and-term}.
+We first define the predicate lifting
+\[\begin{array}{l}
+\mathsf{Arr^{\wedge} : \forall (A\, B : Set) \to (A \to Set) \to (B
+  \to Set) \to}\\
+\mathsf{\hspace*{2.2in}(A \to B) \to Set}
+\end{array}\] for arrow types, since arrow types
 appear in $\mathsf{LType}$ and $\mathsf{LTerm}$.  It is given by
 $\mathsf{Arr^{\wedge}\, A\, B\, Q_A\, Q_B\, f = \forall (a : A) \to
-  Q_A\,a \to Q_B\, (f\,a)}$. The predicate liftings
-$\mathsf{LType^{\wedge} : \forall (A : Set) \to (A \to Set) \to
-  LType\,A \to Set}$ for $\mathsf{LType}$ and $\mathsf{LTerm^{\wedge}
-  :}$ $\mathsf{\forall (A : Set) \to (A \to Set) \to LTerm\,A \to
-  Set}$ for $\mathsf{LTerm}$ are defined in Figure~\ref{fig:liftings}.
-There, $\mathsf{s : String}$, $\mathsf{Q_A : A \to Set}$, $\mathsf{Q_B
-  : B \to Set}$, $\mathsf{Q_C : C \to Set}$,
+  Q_A\,a \to}$ $\mathsf{Q_B\, (f\,a)}$. The predicate liftings
+\[\mathsf{LType^{\wedge} : \forall (A : Set) \to (A \to Set) \to
+  LType\,A \to Set}\] for $\mathsf{LType}$ and
+\[\mathsf{LTerm^{\wedge}
+  : \forall (A : Set) \to (A \to Set) \to LTerm\,A \to Set}\] for
+$\mathsf{LTerm}$ are defined in Figure~\ref{fig:liftings}.  There,
+$\mathsf{s : String}$, $\mathsf{Q_A : A \to}$ $\mathsf{Set}$,
+$\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C : C \to Set}$,
 $\mathsf{K^{Bool}_{\top}}$ is the constantly $\mathsf{\top}$-valued
 predicate on $\mathsf{Bool}$, $\mathsf{T_A : LType\, A}$, $\mathsf{T_B
-  : LType \,B}$, $\mathsf{T_C : LType \,C}$, $\mathsf{t_B : LTerm \,
-  B}$, $\mathsf{t_C : LTerm \, C}$, and $\mathsf{t_{BA} : LTerm \, (B
-  \to A)}$.  Moreover, $\mathsf{e : Equal\,A\,Bool}$ in the first
-clause, $\mathsf{e : Equal\, A\, (B \to C)}$ in the second, $\mathsf{e
-  : Equal\, A\, (List\, B)}$ in the third, $\mathsf{e : Equal \, A \,
-  (B \to C)}$ in the fifth, and $\mathsf{e : Equal\, A\, (List \,B)}$,
+  : LType \,B}$, $\mathsf{T_C :}$ $\mathsf{LType \,C}$, $\mathsf{t_B :
+  LTerm \, B}$, $\mathsf{t_C : LTerm \, C}$, and $\mathsf{t_{BA} :
+  LTerm \, (B \to A)}$.  Also, $\mathsf{e : Equal\,A\,Bool}$ in the
+first clause, $\mathsf{e : Equal\, A\, (B \to C)}$ in the second and
+fifth, $\mathsf{e : Equal\, A\, (List\, B)}$ in the third,
+%$\mathsf{e : Equal \, A \, (B \to C)}$ in the fifth,
+and $\mathsf{e : Equal\, A\, (List \,B)}$,
 $\mathsf{ts : List\, (LTerm B)}$, and $\mathsf{List^\wedge}$ is the
@@
 $\mathsf{dIndList}$ associated with $\mathsf{LTerms}$'s data
-constructors. These are, respectively,
+constructors. These are given in Figure~\ref{fig:ind-hyps-lterm}.
+%These are, respectively
+\begin{comment}
 \[\begin{array}{l}
@@
 \end{array}\]
-
-\noindent
+\end{comment}
 The deep induction rule for $\mathsf{LTerm}$ is thus
-
-\vspace*{-0.225in}
-
 \begin{equation}\label{eq:ind-lam}
@@
 \mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A
-  \to Set) \to dIndVar\,P \to dIndAbs\,P \to dIndApp\,P \to
-  dIndList\,P \to} \\ \quad \mathsf{\forall (A : Set)(Q_A : A \to
-  Set)(t_A : LTerm\,A) \to LTerm^{\wedge}\,A\,Q_A\,t_A \to
+  \to Set) \to}\\
+\quad\mathsf{dIndVar\,P \to dIndAbs\,P \to dIndApp\,P \to
+  dIndList\,P \to}\\
+\quad\quad \mathsf{\forall (A : Set)(Q_A : A \to
+  Set)(t_A : LTerm\,A) \to}\\
+\quad\quad\quad\mathsf{LTerm^{\wedge}\,A\,Q_A\,t_A \to
   P\,A\,Q_A\,t_A}
@@
 
-  \begin{adjustbox}{varwidth=6.8in, max width=6.6in, margin=-0.1in 0in
+    \begin{adjustbox}{varwidth=7.3in, max width=7.1in, margin=-0.0in 0in
       -0.2in 0in, fbox, center}
+
+%  \begin{adjustbox}{varwidth=6.8in, max width=6.6in, margin=-0.1in 0in
+%      -0.2in 0in, fbox, center} 
 {\small
@@
 \end{array}\]}
-
-\vspace*{-0.1in}
-
-\caption{$\mathsf{dIndLTerm}$}\label{fig:dindlterm} \vspace*{0.1in} 
-\end{adjustbox}
-\end{figure*}
-
-\vspace*{-0.125in}
-
-To prove this rule is sound we define a witness
-$\mathsf{dIndLTerm}$ inhabiting the type in~\eqref{eq:ind-lam} as in
-Figure~\ref{fig:dindlterm}. There, $\mathsf{s : String}$, $\mathsf{Q_A
-  : A \to Set}$, $\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C : C \to
-  Set}$, $\mathsf{T_A : LType\,A}$, $\mathsf{T_B : LType\,B}$,
-$\mathsf{t_B : LTerm\,B}$, $\mathsf{t_C : LTerm\,C}$, $\mathsf{t_{BA}
-  : LTerm\,(B \to A)}$, $\mathsf{liftA : LTerm^{\wedge}\, A\,
-  Q_A\, (var\;s\,T_A) = LType^{\wedge}\,A\,Q_A\,T_A}$, $\mathsf{liftE
-  : Equal^{\wedge} \, A\, (B \to C)\, Q_A\, (Arr^{\wedge} \, B\, C\,
-  Q_B \, Q_C) \, e}$, $\mathsf{lift_{T_B}: LType^{\wedge} \, B\, Q_B\,
-  T_B}$, $\mathsf{lift_{t_C}: LTerm^{\wedge} \, C\, Q_C\,
-  T_C}$, $\mathsf{lift_{t_{BA}}: LTerm^{\wedge} \, (B \to A)\,
-  (Arr^{\wedge} \, B\, A\, Q_B \, Q_A)\,
-  t_{BA}}$, $\mathsf{lift_{t_B}: LTerm^{\wedge} \, B\, Q_B\,
-  t_B}$, $\mathsf{liftE' : Equal^{\wedge}\, A\, (List\,B)\, Q_A\,
-  (List^{\wedge}\, B\, Q_B)\, e}$, and $\mathsf{lift_{List}:
-  List^{\wedge} \, (LTerm\, B) \, (LTerm^{\wedge}\, B\, Q_B) \, ts}$.
-Moreover,
-
-\vspace*{-0.225in}
-
+\mbox{where}
+{\small
 \[\begin{array}{lll}
@@
   \, (P\,B\,Q_B)}
-\end{array}\]
+\end{array}\]}
 
@@
 
-\noindent
-where, in the final clause, $\mathsf{PredMap : \forall\, (A : Set) \to
-  (A \to Set) \to (A \to Set) \to Set }$ is the type constructor
-producing the type of morphisms between predicates defined by
-$\mathsf{PredMap \,A\, Q\,Q'\, = \forall\, (a : A) \to Q\,a \to
-  Q'\,a}$ and $\mathsf{liftListMap : \forall\, (A : Set) \to (Q \, Q'
-  : A \to Set) \to PredMap\,A\,Q\,Q' \to PredMap\,(List\,A)
-  \,(List^{\wedge}\, A\, Q)\, (List^{\wedge}\, A\, Q')}$, which takes
-a morphism $\mathsf{f}$ of predicates and produces a morphism of
-lifted predicates, is defined by $\mathsf{liftListMap\, A\, Q\, Q'\,
-  m\, nil\, tt = tt}$ (since $\mathsf{x : List^{\wedge}\, A\, Q\,
-  nil}$ must necessarily be the sole inhabitant $\mathsf{tt}$ of
-$\mathsf{\top}$), and by $\mathsf{liftListMap\, A\, Q\, Q'\, m\,
-  (cons\, a\, l')\, (y, x') = (m\,a\,y, liftListMap\, A\, Q\, Q'\, m\,
-  l'\, x')}$ (since $\mathsf{x : List^{\wedge}\, A\, Q\, (cons\, a\,
-  l')}$ must be of the form $\mathsf{x = (y, x')}$ where
-$\mathsf{y : Q\,a}$ and $\mathsf{x' : List^{\wedge}\, A\, Q\, l'}$).
+\caption{$\mathsf{dIndLTerm}$}\label{fig:dindlterm} \vspace*{0.1in} 
+\end{adjustbox}
+\end{figure*}
+
+\begin{figure*}[t]
+
+\begin{adjustbox}{varwidth=7.3in, max width=7.1in, margin=-0.0in 0in
+      -0.2in 0in, fbox, center}
+
+{\small
+\[\begin{array}{l}
+\mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to
+  LTerm\,A  \to Set) \to}\\
+  \quad\mathsf{\forall (A : Set) (Q_A : A \to Set) (s : String) (T_A :
+  LType\, A) \to LType^{\wedge} \, A\, Q_A\, T_A \to P \, A\, Q_A\,
+  (var \; s\, T_A)}\\[1ex]
+  \mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to
+  LTerm\,A \to Set) \to} \\ 
+  \quad\mathsf{
+  \forall (A\,B\,C: Set) (Q_A : A \to Set)  (Q_B : B \to Set) (Q_C : C
+  \to Set) (e : Equal\, A\, (B \to C)) (s : String) \to } \\ 
+  \quad\mathsf{(T_B : LType\, B) \to (t_C : LTerm\, C)
+  \to Equal^{\wedge}\,A\,(B \to C)\, Q_A \, (Arr^{\wedge} \, B\, C\,
+  Q_B \, Q_C) \, e \to  } \\
+  \quad\mathsf{
+  LType^{\wedge}\, B\, Q_B\, T_B
+  \to P\, C\, Q_C\, t_C\, 
+  \to P \, A\, Q_A\, (abs \,B \,C \, e \,s \,T_B \, t_C)}\\
+ \\
+  \mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A
+    \to Set)\to} \\ 
+  \quad \mathsf{
+  \forall (A \,B : Set) (Q_A : A \to Set)  (Q_B : B \to Set) 
+   (t_{BA} : LTerm\, (B \to A)) (t_B : LTerm\, B) \to} \\
+  \quad \mathsf{
+  P\, (B \to A)\, (Arr^{\wedge} \, B\, A\, Q_B \, Q_A) \, t_{BA} \, 
+  \to P\, B\, Q_B\, t_B\, 
+  \to P \, A\, Q_A\, (app \,B \,t_{BA} \, t_B) }\\[1ex]
+  \mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A
+    \to Set) \to} \\ 
+  \quad \mathsf{
+  \forall (A \,B : Set) (Q_A : A \to Set)  (Q_B : B \to Set) 
+    (e : Equal\, A\, (List\, B)) (ts : List\, (LTerm\, B)) \to} \\ 
+  \quad \mathsf{
+    Equal^{\wedge}\, A\, (List\,B)\, Q_A\, (List^{\wedge}\, B\, Q_B)\, e 
+  \to List^{\wedge}\, (LTerm\,B) (P\, B\, Q_B)\, ts
+  \to P \, A\, Q_A\, (list \,B \,e \, ts) }
+\end{array}\]}
+
+\vspace*{-0.1in}
+
+\caption{Induction hypotheses for $\mathsf{LTerm}$}\label{fig:ind-hyps-lterm}
+\end{adjustbox}
+\end{figure*}
+
+To prove this rule is sound we define a witness $\mathsf{dIndLTerm}$
+inhabiting the type in~\eqref{eq:ind-lam} as in
+Figure~\ref{fig:dindlterm}. There, $\mathsf{s : String}$, $\mathsf{Q_A
+  : A \to Set}$, $\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C : C \to
+  Set}$, $\mathsf{T_A : LType\,A}$, $\mathsf{T_B}$ $\mathsf{:
+  LType\,B}$, $\mathsf{t_B : LTerm\,B}$, $\mathsf{t_C : LTerm\,C}$,
+$\mathsf{t_{BA} : LTerm\,(B \to A)}$, $\mathsf{liftA :
+  LTerm^{\wedge}\, A\, Q_A\, (var\;s\,T_A) =
+  LType^{\wedge}\,A\,Q_A\,T_A}$, $\mathsf{liftE :}$
+$\mathsf{Equal^{\wedge}}$ $\mathsf{A\, (B \to C)\, Q_A\, (Arr^{\wedge}
+  \, B\, C\, Q_B \, Q_C) \, e}$, $\mathsf{lift_{T_B}: LType^{\wedge}
+  \, B\, Q_B\, T_B}$, $\mathsf{lift_{t_C}}$ $\mathsf{: LTerm^{\wedge}
+  \, C\, Q_C\, T_C}$, $\mathsf{lift_{t_{BA}}: LTerm^{\wedge} \, (B \to
+  A)\, (Arr^{\wedge} \, B\, A\, Q_B \, Q_A)}$ $\mathsf{t_{BA}}$,
+$\mathsf{lift_{t_B}: LTerm^{\wedge} \, B\, Q_B\, t_B}$,
+$\mathsf{liftE' : Equal^{\wedge}\, A\, (List\,B)\, Q_A}$ and
+$\mathsf{(List^{\wedge}\, B\, Q_B)\, e}$, and $\mathsf{lift_{List}:
+  List^{\wedge} \, (LTerm\, B) \, (LTerm^{\wedge}\, B\, Q_B) \, ts}$.
+Moreover, in the definition of $\mathsf{p_{ts}}$, $\mathsf{PredMap :
+  \forall\, (A : Set) \to}$ $\mathsf{(A \to Set)}$ $\mathsf{\to (A \to
+  Set) \to Set }$ is the type constructor producing the type of
+morphisms between predicates defined by
+\[\mathsf{PredMap \,A\, Q\,Q'\,
+  = \forall\, (a : A) \to Q\,a \to Q'\,a}\] and
+\[\begin{array}{l}
+\mathsf{liftListMap : \forall\, (A : Set) \to (Q \, Q' : A \to Set)
+  \to}\\\
+\hspace*{0.8in}\mathsf{PredMap\,A\,Q\,Q' \to}\\
+\hspace*{1in}\mathsf{PredMap\,(List\,A)
+  \,(List^{\wedge}\, A\, Q)\, (List^{\wedge}\, A\, Q')}
+\end{array}\]
+which takes a morphism $\mathsf{f}$ of predicates and produces a
+morphism of lifted predicates, is defined by
+\[\mathsf{liftListMap\, A\, Q\, Q'\, m\, nil\, tt = tt}\]
+(since $\mathsf{x : List^{\wedge}\,
+  A\, Q\, nil}$ must necessarily be the sole inhabitant $\mathsf{tt}$
+of $\mathsf{\top}$), and by $\mathsf{liftListMap\, A\, Q\, Q'\, m\,
+  (cons\, a\, l')\, (y, x') = (m\,a\,y,}$\\ $\mathsf{liftListMap\, A\,
+  Q\, Q'\, m\, l'\, x')}$ (since $\mathsf{x : List^{\wedge}\, A\,
+  Q\, (cons\, a\, l')}$ must be of the form $\mathsf{x = (y, x')}$
+where $\mathsf{y : Q\,a}$ and $\mathsf{x' : List^{\wedge}\, A\, Q\,
+  l'}$).
 
@@
 \mathsf{F_1\,G\,\ol{B} \times F_2\,G\,\ol{B} \ \vert\ F_1\,G\,\ol{B} +
-F_2\,G\,\ol{B} \ \vert\ F_1\,\ol{B} \to F_2\,G\,\ol{B}
+  F_2\,G\,\ol{B}}\\
+& & \!\!\!\!\mathsf{\vert\ F_1\,\ol{B} \to F_2\,G\,\ol{B}
 \ \vert\ G\,(\ol{F_1\,\ol{B}}) \ \vert\ H\,\ol{B} \ \vert\ H\,
@@
 \end{array}\]
-
-\noindent
 This grammar is subject to the following restrictions. In the third
@@
 can also construct a map function
-
-\vspace*{-0.05in}
-
 \begin{equation}\label{eq:hliftmap}
-\mathsf{H^\wedge Map : \forall (\ol{A : Set}) (\ol{Q\;Q' : A \to Set}) 
-\to \ol{PredMap\,A\,Q\,Q'} \to
-PredMap\,(H\,\ol{A})\,(H^{\wedge}\,\ol{A}\,\ol{Q})\, 
+\begin{array}{lll}
+\mathsf{H^\wedge Map} & : & \mathsf{\forall (\ol{A : Set}) (\ol{Q\;Q'
+    : A \to Set}) \to}\\
+& & \mathsf{\ol{PredMap\,A\,Q\,Q'} \to}\\
+& & \mathsf{PredMap\,(H\,\ol{A})\,(H^{\wedge}\,\ol{A}\,\ol{Q})\, 
 (H^{\wedge}\,\ol{A}\,\ol{Q'})}
+\end{array}
 \end{equation}
-
-\vspace*{0.05in}
-
-\noindent
 for $\mathsf{H^{\wedge}}$. A concrete way to define $\mathsf{H^\wedge
@@
 than heavier notation. In this case the type of $\mathsf{G}$'s single
-data constructor $\mathsf{c}$ can be rewritten as $\mathsf{c : \forall
-  (\ol{B : Set}) \to Equal\,A\,(K\,\ol{B}) \to F\,G\,\ol{B} \to
-  G\,A}$. The predicate lifting $\mathsf{G^{\wedge} : \forall (A :
-  Set) \to (A \to Set) \to G\,A \to Set}$ for $\mathsf{G}$ is
-therefore
-
-\vspace*{-0.05in}
-
-\[\mathsf{G^{\wedge}\,A\,Q_A\,(c\,\ol{B}\,e\,x)
-= \exists [\ol{Q_B}]\,
+data constructor $\mathsf{c}$ can be rewritten as
+\[\mathsf{c : \forall
+  (\ol{B : Set}) \to Equal\,A\,(K\,\ol{B})\to F\,G\,\ol{B} \to G\,A}\]
+The predicate lifting $\mathsf{G^{\wedge} : \forall (A : Set) \to (A
+  \to Set) \to G\,A \to}$\\ $\mathsf{Set}$ for $\mathsf{G}$ is therefore
+\[\begin{array}{l}
+\mathsf{G^{\wedge}\,A\,Q_A\,(c\,\ol{B}\,e\,x)
+  =}\\
+\mathsf{\hspace*{0.15in}\exists [\ol{Q_B}]\,
 Equal^{\wedge}\,A\,(K\,\ol{B})\,Q_A\,(K^{\wedge}\,\ol{B}\,\ol{Q_B})\,e
-\times F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,{\ol{Q_B}}\,x}\]
-
-\noindent
+\times F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,{\ol{Q_B}}\,x}
+\end{array}\]
 where $\mathsf{Q_A : A \to Set}$, $\ol{\mathsf{Q_B : B \to Set}}$,
-$\mathsf{e : Equal\,A\,(K\,\ol{B})}$, and $\mathsf{x : F\,G\,\ol{B}}$.
-If we have predicate liftings $\mathsf{F^{\wedge}} : \mathsf{\forall
-  (G : Set^{\alpha} \to Set) (\ol{B : Set}) \to (\forall (A : Set) \to
-  (A \to Set) \to G\,A \to Set) \to (\ol{B \to Set}) \to F\,G\,\ol{B}
-  \to Set}$ for $\mathsf{F}$ and $\mathsf{K^{\wedge}} :
+$\mathsf{e : Equal\,A\,(K\,\ol{B})}$, and $\mathsf{x :}$\\
+$\mathsf{F\,G\,\ol{B}}$.
+If we have predicate liftings
+\[\begin{array}{l}\mathsf{F^{\wedge}} : \mathsf{\forall
+  (G : Set^{\alpha} \to Set) (\ol{B : Set}) \to}\\
+\mathsf{\hspace*{0.3in}(\forall (A : Set) \to
+  (A \to Set) \to G\,A \to Set) \to}\\
+\mathsf{\hspace*{0.3in}(\ol{B \to Set}) \to F\,G\,\ol{B}
+  \to Set}
+\end{array}\] for $\mathsf{F}$ and \[\mathsf{K^{\wedge}} :
 \mathsf{\forall (\ol{B : Set}) \to (\ol{B \to Set}) \to K\,\ol{B} \to
-  Set}$
-%\[\begin{array}{lll}
-%\mathsf{F^{\wedge}} & : & \mathsf{\forall (G : Set^{\alpha} \to Set) (\ol{B : Set})
-%\to (\forall (A : Set) \to (A \to Set) \to G\,A \to Set)
-%\to (\ol{B \to Set}) \to F\,G\,\ol{B} \to Set}\\ 
-%\mathsf{K^{\wedge}} & : & \mathsf{\forall (\ol{B : Set}) \to (\ol{B \to Set}) \to
-%  K\,\ol{B} \to Set}
-%\end{array}\]
+  Set}\]
 for $\mathsf{K}$, then the induction hypothesis $\mathsf{dIndC}$
@@
   G\,A \to Set) \to}\\
-\quad \mathsf{\forall (A : Set)\, (\ol{B : Set})\, (Q_A : A \to Set)\,
-  (\ol{Q_B : B \to Set})\, (e : Equal\,A\,(K\,\ol{B}))\, (x :
-  F\,G\,\ol{B}) \to} \\ 
-\quad \mathsf{
-  Equal^{\wedge}\,A\,(K\,\ol{B})\,Q_A\,(K^{\wedge}\,\ol{B}\,\ol{Q_B})\,e
-  \to F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}\,x \to
+\quad \mathsf{\forall (A : Set)\, (\ol{B : Set})\, (Q_A : A \to
+  Set)\,(\ol{Q_B : B \to Set})}\\ 
+\quad \quad \mathsf{(e : Equal\,A\,(K\,\ol{B}))\, (x : F\,G\,\ol{B})
+  \to}\\
+\quad \quad \quad
+\mathsf{Equal^{\wedge}\,A\,(K\,\ol{B})\,Q_A\,(K^{\wedge}\,\ol{B}\,\ol{Q_B})\,e 
+  \to}\\
+\quad\quad\quad\quad\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}\,x \to
   P\,A\,Q_A\,(c\,\ol{B}\,e\,x)} 
@@
 \begin{equation}\label{eq:gen-ind-rule}
+\begin{array}{l}
   \mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to G\,A \to
-  Set) \to dIndC\,P \to \forall (A : Set)(Q_A : A \to Set)(y : G\,A)
-  \to G^{\wedge}\,A\,Q_A\,y \to P\,A\,Q_A\,y}
+  Set) \to}\\
+\quad \mathsf{dIndC\,P \to \forall (A : Set)(Q_A : A \to Set)(y : G\,A)
+  \to}\\
+\quad\quad\mathsf{G^{\wedge}\,A\,Q_A\,y \to P\,A\,Q_A\,y}
+\end{array}
 \end{equation}
@@
 $\mathsf{dIndG}$ inhabiting this type by
-\[\mathsf{dIndG\,P\,cc\,A\,Q_A\,(c\,\ol{B}\,e\,x)\,(\ol{Q_B}, liftE, liftF)
-= cc\,A\,\ol{B}\,Q_A\,\ol{Q_B}\,e\,x\,liftE\,(p\,x\,liftF)}\] Here,
-$\mathsf{cc : dIndC\,P}$, $\mathsf{e : Equal\,A\,(K\,\ol{B})}$,
+\[\begin{array}{ll}
+ & \mathsf{dIndG\,P\,cc\,A\,Q_A\,(c\,\ol{B}\,e\,x)\,(\ol{Q_B}, liftE, liftF)}\\
+= & \mathsf{cc\,A\,\ol{B}\,Q_A\,\ol{Q_B}\,e\,x\,liftE\,(p\,x\,liftF)}
+\end{array}\]
+Here, $\mathsf{cc : dIndC\,P}$, $\mathsf{e : Equal\,A\,(K\,\ol{B})}$,
 $\mathsf{x : F\,G\,\ol{B}}$, $\mathsf{Q_A : A \to Set}$,
@@
   Equal^{\wedge}\,A\,(K\,\ol{B})\,Q_A\,
-  (K^{\wedge}\,\ol{B}\,\ol{Q_B})\,e}$, and $\mathsf{liftF :
-  F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,{\ol{Q_B}}\,x}$. As a result,
+  (K^{\wedge}\,\ol{B}\,\ol{Q_B})\,e}$, and $\mathsf{liftF}$ $\mathsf{:
+  F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,{\ol{Q_B}}\,x}$, so
 $\mathsf{(\ol{Q_B}, liftE, liftF) : G^{\wedge}\,A\,Q_A (c\,\ol{B}\,
-  e\,x)}$ as expected.  Finally, the morphism of predicates $\mathsf{p
-  :
-  PredMap\,(F\,G\,\ol{B})\,(F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B})
-  (F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}$ is defined by structural
+  e\,x)}$ as expected.  Finally, the morphism of predicates
+\[\mathsf{p : PredMap\,(F\,G\,\ol{B}) (F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B})
+  (F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}\] is defined by structural
 induction on $\mathsf{F}$ as follows:
@@
 If $\mathsf{F\,G\,\ol{B} = F_1\,G\,\ol{B} \times F_2\,G\,\ol{B}}$ then
-$\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B} =
-  Pair^{\wedge}\,(F_1\,G\,\ol{B})\, (F_2\,G\,\ol{B})\,
+\[\begin{array}{ll}
+& \!\!\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}}\\
+\quad\quad= & \!\!\mathsf{Pair^{\wedge}\,(F_1\,G\,\ol{B})\, (F_2\,G\,\ol{B})\,
   (F_1^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})\,
-  (F_2^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}$. The induction hypothesis
+  (F_2^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}
+  \end{array}\] The induction hypothesis
   ensures
@@
 %  types) and}
-morphisms of predicates $\mathsf{p_1 :
-  PredMap\,(F_1\,G\,\ol{B})\,(F_1^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_BQ})
-  (F_1^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}$ and $\mathsf{p_2 :
+  morphisms of predicates
+  \[\quad\quad\;\;\mathsf{p_1
+  : PredMap\,(F_1\,G\,\ol{B})\,(F_1^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_BQ})
+  (F_1^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}\] and \[\quad\quad\mathsf{p_2 :
   PredMap\,(F_2\,G\,\ol{B})\,(F_2^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B})
-  (F_2^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}$.  For $\mathsf{x_1 :
+  (F_2^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}\] For $\mathsf{x_1 :
   F_1\,G\,\ol{B}}$, $\mathsf{liftF_1 :
   F_1^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}\,x_1}$, $\mathsf{x_2 :
-  F_2\,G\,\ol{B}}$ and $\mathsf{liftF_2 :
-  F_2^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}\,x_2}$ we therefore
-define $\mathsf{p\, (x_1, x_2)\, (liftF_1, liftF_2) =
-  (p_1\,x_1\,liftF_1,\, p_2\,x_2\,liftF_2)}$.
+  F_2\,G\,\ol{B}}$ and $\mathsf{liftF_2}$ $\mathsf{:
+  F_2^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}\,x_2}$ we then define
+\[\quad\mathsf{p\, (x_1, x_2)\, (liftF_1, liftF_2) =
+  (p_1\,x_1\,liftF_1,\, p_2\,x_2\,liftF_2)}\]
 \item The case $\mathsf{F\,G\,\ol{B} = F_1\,G\,\ol{B} +
@@
 \item If $\mathsf{F\,G\,\ol{B} = F_1\,\ol{B} \to F_2\,G\,\ol{B}}$ then
-  $\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}\,x = \forall (z :
+  \[\begin{array}{ll}
+  & \!\!\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}\,x}\\
+  \quad\quad= & \!\!\mathsf{\forall (z :
   F_1\,\ol{B}) \to F_1^{\wedge}\,\ol{B}\,\ol{Q_B}\,z \to
-  F_2^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}\,(x\,z)}$, where $\mathsf{x :
+  F_2^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}\,(x\,z)}
+  \end{array}\] where $\mathsf{x :
   F\,G\,\ol{B}}$. The induction hypothesis ensures
@@
 % liftings $\mathsf{F_1}$ and $\mathsf{F_2}$ (with types) and}
-a morphism of predicates $\mathsf{p_2 : PredMap\,(F_2\,G\,\ol{B})\,
+  a morphism of predicates
+  \[\quad\quad\;\;\mathsf{p_2 : PredMap\,(F_2\,G\,\ol{B})\,
   (F_2^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B})\,
-  (F_2^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}$.  We define
-$\mathsf{p\;x\;liftF : F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}\,x}$, where
-$\mathsf{liftF : F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}\,x}$, to
-be $\mathsf{p\;x\;liftF\;z\;liftF_1 = p_2\, (x\,z)\,
-  (liftF\,z\,liftF_1)}$ for $\mathsf{z : F_1\,\ol{B}}$ and
+  (F_2^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}\]  We therefore define
+$\mathsf{p\,x\,liftF : F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}\,x}$, where
+$\mathsf{liftF}$ $\mathsf{: F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}\,x}$, to
+be \[\mathsf{p\,x\,liftF\;z\;liftF_1 = p_2\, (x\,z)\,
+  (liftF\,z\,liftF_1)}\] for $\mathsf{z : F_1\,\ol{B}}$ and
 $\mathsf{liftF_1 : F_1^{\wedge}\,\ol{B}\,\ol{Q_B}\,z}$. Note that
@@
   does not contain $\mathsf{G}$, then
-  $\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B} = P \,(F_1\,\ol{B})\,
-    (F_1^{\wedge}\,\ol{B}\,\ol{Q_B})}$ for all $\mathsf{P : \forall (A
-    : Set)\to(A \to Set) \to G\,A \to Set}$.  We therefore define
-  $\mathsf{p = dIndG\,P\,
-    cc\,(F_1\,\ol{B})\,(F_1^{\wedge}\,\ol{B}\,\ol{Q_B})}$.
+  \[\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B} = P \,(F_1\,\ol{B})\,
+    (F_1^{\wedge}\,\ol{B}\,\ol{Q_B})}\] for all $\mathsf{P : \forall (A
+    : Set)\to}$ $\mathsf{(A \to Set) \to G\,A \to Set}$.  We then define
+  $\mathsf{p = dIndG\,P}$
+   $\mathsf{cc\,(F_1\,\ol{B})\,(F_1^{\wedge}\,\ol{B}\,\ol{Q_B})}$.
 \item If $\mathsf{F\,G\,\ol{B} = H\,\ol{B}}$ and $\mathsf{H}$ does not
   contain $\mathsf{G}$, then
-  $\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B} = H^\wedge\,\ol{Q_B}}$
-  for all $\mathsf{P : \forall (A : Set)\to(A \to Set) \to G\,A \to
-    Set}$.  We therefore define $\mathsf{p : PredMap\, (H\,\ol{B})\,
-    (H^{\wedge}\,\ol{B}\,\ol{Q_B})\, (H^{\wedge}\,\ol{B}\,\ol{Q_B})}$
+  \[\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B} = H^\wedge\,\ol{Q_B}}\]
+  for all $\mathsf{P : \forall (A : Set)\to(A \to Set) \to G\,A \to}$ $\mathsf{
+    Set}$.  We therefore define \[\mathsf{p : PredMap\, (H\,\ol{B})\,
+    (H^{\wedge}\,\ol{B}\,\ol{Q_B})\,(H^{\wedge}\,\ol{B}\,\ol{Q_B})}\]
   to be the identity morphism on predicates.
@@
   $\mathsf{H}$ does not contain $\mathsf{G}$, then
-  $\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B} = H^{\wedge}\,
+  \[\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B} = H^{\wedge}\,
   (\ol{F_k\,G\,\ol{B}})\,
-  (\ol{F_k^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}})}$ for all $\mathsf{P :
-  \forall (A : Set)\to(A \to Set) \to G\,A \to Set}$.  Since
+    (\ol{F_k^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}})}\] for all $\mathsf{P
+    : \forall (A : Set)\to(A \to Set) \to G\,A \to Set}$.  Since
   $\mathsf{H}$ is not a GADT, $\mathsf{H^\wedge}$ has a map function
   $\mathsf{H^\wedge Map}$ as in~\eqref{eq:hliftmap}.  The induction
-  hypothesis ensures morphisms of predicates $\ol{\mathsf{p_k :
+  hypothesis ensures morphisms of
+  predicates \[\quad\quad\;\;\ol{\mathsf{p_k :
       PredMap\,(F_k\,G\,\ol{B})
       (F_k^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B})
-      (F_k^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}}$. We therefore define
-  $\mathsf{p = H^\wedge Map\,(\ol{F_k\,G\,\ol{B}})\,
-    (\ol{F_k^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}})\,
-    (\ol{F_k^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}})\,\ol{p_k}}$.
+      (F_k^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}}\] We therefore define
+  \[\;\;\quad\quad\mathsf{p = H^\wedge Map\,(\ol{F_k\,G\,\ol{B}})\,
+    (\ol{F_k^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}})
+  \,(\ol{F_k^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}})\,\ol{p_k}}\]
 \end{itemize}
@@
 
-To see this, we first note that the constructor $\mathsf{c}$ can be
-rewritten as $\mathsf{c : \forall\, (B : Set) \to}$ $\mathsf{Equal\,A\,(B \times
-  B) \to}$ $\mathsf{G\,(G\,B) \to G\,A}$. The predicate lifting
-$\mathsf{G^{\wedge} : \forall\, (A : Set) \to (A \to Set) \to G\,A \to
-  Set}$ for $\mathsf{G}$ is therefore
-\[
-\mathsf{G^{\wedge}\,A\,Q_A\,(c\,B\,e\,x)
-= \exists\, [Q_B]\,
-Equal^{\wedge}\,A\,(B \times B)\,Q_A\,(Pair^{\wedge}\,B\,B\,Q_B\,Q_B)\,e
-\times G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q_B)\,x}
-\]
+To see this, we first rewrite the constructor $\mathsf{c}$'s type as
+\[\mathsf{c : \forall\, (B : Set) \to Equal\,A\,(B \times B)
+  \to G\,(G\,B) \to G\,A}\]
+The predicate lifting
+\[\mathsf{G^{\wedge} : \forall\, (A : Set) \to (A \to Set) \to G\,A \to
+  Set}\] for $\mathsf{G}$ is therefore
+\[\begin{array}{l}
+\mathsf{G^{\wedge}\,A\,Q_A\,(c\,B\,e\,x) =}\\
+\mathsf{\hspace*{0.15in}\exists\, [Q_B]\,
+Equal^{\wedge}\,A\,(B \times
+B)\,Q_A\,(Pair^{\wedge}\,B\,B\,Q_B\,Q_B)\,e}\\
+\mathsf{\hspace*{1.8in}\times \;G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q_B)\,x}
+\end{array}\]
 where $\mathsf{Q_A : A \to Set}$, $\mathsf{Q_B : B \to Set}$, $\mathsf{e
@@
 \quad\mathsf{ \forall\, (A\;B : Set)\, (Q_A : A \to Set)\, (Q_B : B
-  \to Set)\, (e : Equal\,A\,(B \times B))\, (x : G\,(G\,B))\to} \\ 
+  \to Set)}\\
+\hspace*{1in}\mathsf{(e : Equal\,A\,(B \times B))\, (x : G\,(G\,B))\to} \\ 
 \quad\mathsf{Equal^{\wedge}\,A\,(B \times
-  B)\,Q_A\,(Pair^{\wedge}\,B\,B\,Q_B\,Q_B)\,e \to P\,(G\,B)\,(P\,B\,Q_B)\,x 
+  B)\,Q_A\,(Pair^{\wedge}\,B\,B\,Q_B\,Q_B)\,e \to}\\
+\hspace*{1in}\mathsf{P\,(G\,B)\,(P\,B\,Q_B)\,x 
 	\to P\,A\,Q_A\,(c\,B\, e\,x)} 
@@
 so the deep induction rule for $\mathsf{G}$ is
-\[\mathsf{\forall\, (P : \forall\, (A : Set) \to (A \to Set) \to G\,A \to Set)
-\to dIndC\,P \to \forall\, (A : Set)\, (Q : A \to Set)\, (y : G\,A)
-\to G^{\wedge}\,A\,Q\,y \to P\,A\,Q\,y}\] But if we now try to show
+\[\begin{array}{l}
+\mathsf{\forall\, (P : \forall\, (A : Set) \to (A \to Set) \to G\,A \to Set)
+  \to}\\
+\quad\mathsf{dIndC\,P \to \forall\, (A : Set)\, (Q : A \to Set)\, (y : G\,A)
+  \to}\\
+\quad\quad\mathsf{G^{\wedge}\,A\,Q\,y \to P\,A\,Q\,y}
+\end{array}\] But if we now try to show
 that this rule is sound by constructing a witness $\mathsf{dIndG}$
 inhabiting this type we run into problems. We can define
-$\mathsf{dIndG\,P\,cc\,A\,Q\,(c\,B\,e\,x)\,(Q', liftE, liftG) =
-  cc\,A\,B\,Q\,Q'\,e\,x\,liftE\,p}$, where $\mathsf{cc : dIndC\,P}$,
-$\mathsf{Q : A \to Set}$, $\mathsf{Q' : B \to Set}$, $\mathsf{e :
-  Equal\,A\,(B \times B)}$, $\mathsf{x : G\,(G\,B)}$, $\mathsf{liftG :
-  G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q')\,x}$, and $\mathsf{liftE :
-  Equal^{\wedge}\,A\,(B \times
+$\mathsf{dIndG\,P\,cc\,A\,Q\,(c\,B\,e\,x)\,(Q', liftE, liftG)}$
+$\mathsf{= cc\,A\,B\,Q\,Q'\,e\,x\,liftE\,p}$, where $\mathsf{cc :
+  dIndC\,P}$, $\mathsf{Q : A \to Set}$,\\ $\mathsf{Q' : B \to Set}$,
+$\mathsf{e : Equal\,A\,(B \times B)}$, $\mathsf{x : G\,(G\,B)}$,
+$\mathsf{liftG : G^{\wedge}\,(G\,B)}$ $\mathsf{(G^{\wedge}
+  B\,Q')\,x}$, and $\mathsf{liftE : Equal^{\wedge}\,A\,(B \times
   B)\,Q\,(Pair^{\wedge}\,B\,B\,Q'\,Q')\,e}$, but we still need to
 define $\mathsf{p : P\,(G\,B)\,(P\,B\,Q')\,x}$.  For this we can use
-the induction rule and let $\mathsf{p =
-  dIndG\,P\,cc\,(G\,B)\,(P\,B\,Q')\,x\,q}$, but we still need to
-define $\mathsf{q : G^{\wedge}\,(G\,B)\,(P\,B\,Q')\,x}$.  If we had
-the map function $\mathsf{G^\wedge Map : \forall\, (A : Set)\, (Q\;Q'
-  : A \to Set) \to PredMap\,A\,Q\,Q' \to
-  PredMap\,(G\,A)\,(G^{\wedge}\,A\,Q)\,(G^{\wedge}\,A\,Q')}$ for
-$\mathsf{G^{\wedge}}$, then we could define $\mathsf{q = G^\wedge
-  Map\,(G\,B)\,(G^{\wedge}\,B\,Q')\,(P\,B\,Q')\,
-  (dIndG\,P\,cc\,B\,Q')\,x\,liftG}$.  Unfortunately, however, we
-cannot define $\mathsf{G^\wedge Map}$. Indeed, its definition would
-have to be $ \mathsf{G^\wedge Map\,A\,Q\,Q'\,m\,(c\,B\,e\,x)\,(Q_B,
-  liftE, liftG) = (Q'_B, liftE', liftG')}$ for some $\mathsf{Q'_B : B
+the induction rule and let
+\[\mathsf{p = dIndG\,P\,cc\,(G\,B)\,(P\,B\,Q')\,x\,q}\] but we still
+need to define \[\mathsf{q : G^{\wedge}\,(G\,B)\,(P\,B\,Q')\,x}\]  If
+we had the map function
+\[\begin{array}{l}
+\mathsf{G^\wedge Map : \forall\, (A : Set)\,
+  (Q\;Q' : A \to Set) \to}\\
+\hspace*{0.7in}\mathsf{PredMap\,A\,Q\,Q' \to}\\
+\hspace*{0.9in}\mathsf{PredMap\,(G\,A)\,(G^{\wedge}\,A\,Q)\,(G^{\wedge}\,A\,Q')} 
+\end{array}\]
+for $\mathsf{G^{\wedge}}$
+then we could define
+\[\mathsf{q = G^\wedge Map\,(G\,B)\,(G^{\wedge}\,B\,Q')\,(P\,B\,Q')\,
+(dIndG\,P\,cc\,B\,Q')\,x\,liftG}\]  Unfortunately, however,
+we cannot define $\mathsf{G^\wedge Map}$. Indeed, its definition would
+have to be
+\[\begin{array}{ll}
+ & \!\!\mathsf{G^\wedge Map\,A\,Q\,Q'\,m\,(c\,B\,e\,x)\,(Q_B,
+  liftE, liftG)}\\
+\quad\quad= & \!\! \mathsf{(Q'_B, liftE', liftG')}
+\end{array}\] for some $\mathsf{Q'_B : B
   \to Set}$, $\mathsf{liftE' : Equal^{\wedge}\,A\,(B \times
-  B)\,Q'\,(Pair^{\wedge}\,B\,B\,Q'_B\,Q'_B)\,e}$, and $\mathsf{liftG'
-  : G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q'_B)\,x}$, where $\mathsf{Q :
-  A \to Set}$, $\mathsf{Q' : A \to Set}$, $\mathsf{Q_B : B \to Set}$,
-$\mathsf{m : PredMap\,A\,Q\,Q'}$, $\mathsf{e : Equal\,A\,(B \times
-  B)}$, $\mathsf{x : G\,(G\,B)}$, $\mathsf{liftE :
-  Equal^{\wedge}\,A\,(B \times
-  B)\,Q\,(Pair^{\wedge}\,B\,B\,Q_B\,Q_B)\,e}$, and $\mathsf{liftG :
-  G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q_B)\,x}$. In other words, we
-would need to produce a proof $\mathsf{liftE'}$ of the (extensional)
-equality of the predicates $\mathsf{Q'}$ and
+  B)\,Q'\,(Pair^{\wedge}\,B}$ $\mathsf{B\,Q'_B\,Q'_B)\,e}$, and
+$\mathsf{liftG' : G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q'_B)\,x}$,
+where $\mathsf{Q : A \to}$ $\mathsf{Set}$, $\mathsf{Q' : A \to Set}$,
+$\mathsf{Q_B : B \to Set}$, $\mathsf{m : PredMap\,A\,Q\,Q'\,e :
+}$\\ $\mathsf{Equal}$ $\mathsf{A\,(B \times B)}$, $\mathsf{x :
+  G\,(G\,B)}$, $\mathsf{liftE : Equal^{\wedge}\,A\,(B \times
+  B)\,Q\,(Pair^{\wedge}}$ $\mathsf{B\,B\,Q_B\,Q_B)\,e}$, and
+$\mathsf{liftG : G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q_B)\,x}$. 
+That is, we would need to produce a proof $\mathsf{liftE'}$ of the
+(extensional) equality of the predicates $\mathsf{Q'}$ and
 $\mathsf{Pair^{\wedge}\,B\,B\,Q'_B\,Q'_B}$ from just a proof
@@
 is equal to $\mathsf{B'}$. And not being able to define
-$\mathsf{Equal^\wedge Map :}$ $\mathsf{\forall (A\, B : Set) (Q_A\,
-  Q_A' : A \to Set) (Q_B\, Q_B' : B \to Set) \to PredMap\,A\,Q_A\,Q_A'
-  \to PredMap\,B\,Q_B\,Q_B' \to
-  PredMap\,(Equal\,A\,B)}\\ \mathsf{(Equal^\wedge
-  A\,B\,Q_A\,Q_B)\,(Equal^\wedge A\,B\,Q_A'\,Q_B')}$ of course makes
-it unclear how to define $\mathsf{G^\wedge Map}$ for more general
-$\mathsf{G}$.
+\[\begin{array}{l}
+\mathsf{Equal^\wedge Map : \forall (A\, B : Set)}\\
+\hspace*{0.8in}\mathsf{(Q_A\, Q_A' : A \to Set)}\\
+\hspace*{0.8in}\mathsf{(Q_B\, Q_B' : B \to Set) \to}\\
+\hspace*{0.5in}\mathsf{PredMap\,A\,Q_A\,Q_A' \to}\\
+\hspace*{0.5in}\mathsf{PredMap\,B\,Q_B\,Q_B' \to}\\
+\hspace*{0.5in}\mathsf{PredMap\,(Equal\,A\,B)\,(Equal^\wedge
+  A\,B\,Q_A\,Q_B)}\\
+\hspace*{0.5in}\mathsf{(Equal^\wedge A\,B\,Q_A'\,Q_B')}
+\end{array}\]
+of course makes it unclear how to define $\mathsf{G^\wedge Map}$ for
+more general $\mathsf{G}$.
 
@@
 In this section we use deep induction for the $\mathsf{LTerm}$ GADT
-from~\eqref{eq:eq_lterm} to infer the type from a lambda term. The
-following predicate either returns the type of its input lambda term
-if that type can be inferred or indicates that the type inference
-fails:
-
-\vspace*{-0.3in}
-
+from Figure~\ref{fig:type-and-term} to infer the type from a lambda
+term. The following predicate either returns the type of its input
+lambda term if that type can be inferred or indicates that the type
+inference fails:
 \begin{align*}
@@
 
-By construction every lambda term in $\mathsf{LTerm}$ is well-typed,
-but that (necessarily unique) type cannot always be inferred.  The
-predicate $\mathsf{GetType}$ uses the standard $\mathsf{Maybe}$ data
-type to represent failure of type inference. It is defined
-by:\looseness=-1
-
-\vspace*{-0.1in}
-
-\begin{equation}\label{eq:maybe}
-\begin{array}{l}
-\mathsf{data\ Maybe : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;nothing :\,  \forall \{A : Set\} \to Maybe\,A}\\
-\mathsf{\;\;\;\;\;\;just\;\;\;\;\;\;\; :\,  \forall \{A : Set\} \to A \to Maybe\,A}
-\end{array}
-\end{equation}
-
-\noindent
-We want to show that $\mathsf{GetType\,A\,t}$ is satisfied by every
-element $\mathsf{t}$ in $\mathsf{LTerm\,A}$, i.e., we want to prove:
-\[ \mathsf{getTypeProof : \forall \, (A : Set)\, (t : LTerm\,A) \to
-   GetType \,A \,t}\] This property can be proved with deep
-induction, which is used to apply the induction hypothesis to the
-individual terms in the list of terms that the data constructor
-$\mathsf{list}$ takes as an argument. Indeed, using the deep induction
-rule $\mathsf{dIndLTerm}$ from Section~\ref{sec:ind-lam} we can define
-$\mathsf{getTypeProof}$ by
-\[ \mathsf{getTypeProof \,A \,t = 
-    dIndLTerm\, P \, cvar\, cabs\, capp\, clist\, A\, K_\top\, t\,
-    (LTerm^\wedge KT\, A\, t) }\] where $\mathsf{t : LTerm\,A}$,
-$\mathsf{P}$ is the polymorphic predicate $\mathsf{ \lambda \, (A:
-  Set)\, (Q : A \to Set)\, (t : LTerm\,A)\, \to Maybe \, (LType \,
-  A)}$, $\mathsf{K_\top}$ is the constantly $\mathsf{\top}$-valued
-predicate on $\mathsf{A}$, and $\mathsf{LTerm^\wedge KT : \forall\, (A
-  : Set)\, (t : LTerm A) \to LTerm^{\wedge}\, A\, K_\top\,t}$ is a
-term, to be defined below, witnessing that $\mathsf{K_\top}$ can be
-lifted to all terms. We also need the applications to $\mathsf{P}$ of
-each of the induction hypotheses from Section~\ref{sec:ind-lam}. These
-are:
+\begin{figure*}[t]
+
+\begin{adjustbox}{varwidth=7.3in, max width=7.1in, margin=-0.0in 0in
+      -0.2in 0in, fbox, center}
+
+{\small
 \[\begin{array}{lll}
@@
   (GetType\, B)\, ts \to Maybe\, (LType\, A)}
-\end{array}\]
-In the first clause, $\mathsf{cvar}$ returns $\mathsf{just\,T_A}$.  In
-the second clause, $\mathsf{cabs}$ returns $\mathsf{nothing}$ if its
-final argument is $\mathsf{nothing}$ and $\mathsf{cabs\, A \,B \,C
-  \,Q_A\, Q_B \,Q_C\, e \,s \,T_B\, t_C\, liftE\;
+\end{array}\]}
+
+\vspace*{-0.1in}
+
+\caption{Applies induction hypotheses for
+  $\mathsf{LTerm}$}\label{fig:ind-hyps-lterm-applied-to-P}
+\end{adjustbox}
+\end{figure*}
+
+By construction every lambda term in $\mathsf{LTerm}$ is well-typed,
+but that (necessarily unique) type cannot always be inferred.  The
+predicate $\mathsf{GetType}$ uses the standard $\mathsf{Maybe}$ data
+type to represent failure of type inference. It is defined
+by:\looseness=-1
+
+\vspace*{-0.1in}
+
+\begin{equation}\label{eq:maybe}
+\begin{array}{l}
+\mathsf{data\ Maybe : Set \to Set\ where}\\
+\mathsf{\;\;\;\;\;\;nothing :\,  \forall \{A : Set\} \to Maybe\,A}\\
+\mathsf{\;\;\;\;\;\;just\;\;\;\;\;\;\; :\,  \forall \{A : Set\} \to A \to Maybe\,A}
+\end{array}
+\end{equation}
+
+\noindent
+We want to show that $\mathsf{GetType\,A\,t}$ is satisfied by every
+element $\mathsf{t}$ in $\mathsf{LTerm\,A}$, i.e., we want to prove:
+\[ \mathsf{getTypeProof : \forall \, (A : Set)\, (t : LTerm\,A) \to
+   GetType \,A \,t}\] This property can be proved with deep
+induction, which is used to apply the induction hypothesis to the
+individual terms in the list of terms that the data constructor
+$\mathsf{list}$ takes as an argument. Indeed, using the deep induction
+rule $\mathsf{dIndLTerm}$ from Section~\ref{sec:ind-lam} we can define
+$\mathsf{getTypeProof}$ by
+\[\begin{array}{ll}
+ & \mathsf{getTypeProof \,A \,t}\\
+= & \mathsf{dIndLTerm\, P \, cvar\, cabs\, capp\, clist\, A\, K_\top\, t\,
+  (LTerm^\wedge KT\, A\, t) }
+\end{array}\] where $\mathsf{t : LTerm\,A}$,
+$\mathsf{P}$ is the polymorphic predicate \[\mathsf{ \lambda \, (A:
+  Set)\, (Q : A \to Set)\, (t : LTerm\,A)\, \to Maybe \, (LType \,
+  A)}\] and $\mathsf{K_\top}$ is the constantly $\mathsf{\top}$-valued
+predicate on $\mathsf{A}$, and \[\mathsf{LTerm^\wedge KT : \forall\,
+  (A : Set)\, (t : LTerm A) \to LTerm^{\wedge}\, A\, K_\top\,t}\] is a
+term, to be defined below, witnessing that $\mathsf{K_\top}$ can be
+lifted to all terms. We also need the applications to $\mathsf{P}$ of
+each of the induction hypotheses from Section~\ref{sec:ind-lam}. These
+are given in Figure~\ref{fig:ind-hyps-lterm-applied-to-P}. In the first
+clause, $\mathsf{cvar}$ returns $\mathsf{just\,T_A}$.  In the second
+clause, $\mathsf{cabs}$ returns $\mathsf{nothing}$ if its final
+argument is $\mathsf{nothing}$ and
+\[\begin{array}{ll}
+ & \!\!\mathsf{cabs\, A \,B \,C \,Q_A\, Q_B \,Q_C\, e \,s \,T_B\, t_C\, liftE\;
 %  LLTy^\wedge B\,
-lift_{T_B}\; (just \,T_C) = just \, (arr\, B\, C\, e\, T_B \,T_C)}$ otherwise.
-In the third clause, $\mathsf{capp\, A \,B \, Q_A \, Q_B \, t_{BA}\,
-  t_A \, (just \, (arr\, B \,A\, refl\, T_B\, T_A))\, mb = just\,
-  T_A}$ and $\mathsf{capp}$ returns $\mathsf{nothing}$ otherwise.  In
-the fourth clause, we must use $\mathsf{List^{\wedge}\, (LTerm\, B)\,
+  lift_{T_B}\; (just \,T_C)}\\
+  = & \!\!\mathsf{just \, (arr\, B\, C\, e\, T_B \,T_C)}
+\end{array}\] otherwise.
+In the third clause,
+\[\begin{array}{ll}
+ & \!\!\mathsf{capp\, A \,B \, Q_A \, Q_B \, t_{BA}\, t_A \, (just \,
+  (arr\, B \,A\, refl\, T_B\, T_A))\, mb}\\ 
+= & \!\!\mathsf{just\, T_A}
+\end{array}\]
+and $\mathsf{capp}$ returns $\mathsf{nothing}$ otherwise.  In the
+fourth clause, we must use $\mathsf{List^{\wedge}\, (LTerm\, B)\,
   (GetType\, B)\, ts}$ to extract the type of the head of
@@
 $\mathsf{ts} = \mathsf{nil}$ we define $\mathsf{clist\, A\, B\, Q\,
-  Q'\, e\, nil \, liftE\, lift_{ts} = nothing}$, where $\mathsf{liftE
-  : Equal^{\wedge}\, A\, (List\,B) \, Q\, (List^{\wedge}\, B\, Q')\,
-  e}$, and $\mathsf{lift_{ts} : List^{\wedge}\, (LTerm\, B)\,
-  (GetType\, B)\, ts}$.  When $\mathsf{ts} = \mathsf{cons\,t\,ts'}$
+  Q'\, e\, nil \, liftE\, lift_{ts}}$ $\mathsf{= nothing}$, where
+$\mathsf{liftE : Equal^{\wedge}\, A\, (List\,B) \, Q\,
+  (List^{\wedge}\, B\, Q')\, e}$, and $\mathsf{lift_{ts} :
+  List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, ts}$.  When
+$\mathsf{ts} = \mathsf{cons\,t\,ts'}$
 % The interesting case is
@@
 %the type we need.  The type of $\mathsf{list_{ts}}$ is
-the type of $\mathsf{list_{ts}}$ becomes $\mathsf{List^{\wedge}\,
-  (LTerm\, B)\, (GetType\, B)\, (cons\, t\, ts')} = \mathsf{GetType\,
-  B\, t \times List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, ts'} =
-\mathsf{Maybe\, (LType\, B) \times List^{\wedge}\, (LTerm\, B)\,
-  (GetType\, B)\, ts'}$. We pattern match on the first component
+the type of $\mathsf{list_{ts}}$ becomes
+\[\begin{array}{ll}
+& \!\!\mathsf{List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, (cons\, t}$
+$\mathsf{ts')}\\
+= & \!\!\mathsf{GetType\, B\, t \times List^{\wedge}\, (LTerm\, B)\,
+  (GetType\, B)\, ts'}\\
+= & \!\! \mathsf{Maybe\, (LType\, B) \times List^{\wedge}\, (LTerm\, B)\,
+  (GetType\, B)\, ts'}
+\end{array}\] We pattern match on the first component
 of the pair to define\looseness=-1
@@
 Here $\mathsf{e : Equal\, A\, (List\,B)}$, $\mathsf{T' : LType\, B}$,
-and $\mathsf{lift_{ts'} : List^{\wedge}\, (LTerm\, B)\, (GetType\,
-  B)\, ts'}$.
-
-\pagebreak
+and
+\[\mathsf{lift_{ts'} : List^{\wedge}\, (LTerm\, B)\, (GetType\,
+  B)\, ts'}\]
 
 To finish defining $\mathsf{getTypeProof}$ we still need a proof
-
-\vspace*{-0.1in}
-
 \[ \mathsf{LTerm^\wedge KT : \forall\, (A : Set)\, (t : LTerm\, A) \to
-  LTerm^{\wedge}\, A\, K_\top\,t} \]
-
-\vspace*{-0.05in}
-
-\noindent
-Since $\mathsf{LTerm^\wedge}$ is
+  LTerm^{\wedge}\, A\, K_\top\,t} \] Since $\mathsf{LTerm^\wedge}$ is
 defined in terms of $\mathsf{LType^\wedge}$ and $\mathsf{Arr^\wedge}$,
@@
 \item If $\mathsf{s : String}$ and $\mathsf{T : LType\,A}$ we define
-  $\mathsf{LTerm^\wedge KT\,A\,(var\,s\,T)} = \mathsf{LType^\wedge
-  KT\,A\,T}$.
+  \[\mathsf{LTerm^\wedge KT\,A\,(var\,s\,T)} = \mathsf{LType^\wedge
+  KT\,A\,T}\]
 \item If $\mathsf{e : Equal\,A\,(B \to C)}$, $\mathsf{s : String}$,
-  $\mathsf{T : LType\,B}$, and $\mathsf{t' : LTerm\,C}$ we need to
-  define $\mathsf{LTerm^\wedge KT\,A\, (abs\, B \,C \, e \,s \,T \,
-    t')}$ of type
-
-\vspace*{-0.15in}
-
-  \[\begin{array}{l}
-  \mathsf{LTerm^{\wedge}\,A\,K_\top\, (abs \,B \,C \,e \,s \,T
+  $\mathsf{T : LType\,B}$, and $\mathsf{t' :}$ $\mathsf{LTerm\,C}$ we
+  need to define $\mathsf{LTerm^\wedge KT\,A\, (abs\, B \,C \, e \,s
+    \,T \, t')}$ of type
+\[\begin{array}{ll}
+ & \!\!\mathsf{LTerm^{\wedge}\,A\,K_\top\, (abs \,B \,C \,e \,s \,T
   \,t')}\\
-\quad \mathsf{ = \exists [Q_B]\, [Q_C]\, Equal^{\wedge} \, A\, (B \to
-    C)\, K_\top\, (Arr^{\wedge} \, B\, C\, Q_B \, Q_C)\, e \times \,
-    LType^{\wedge}\, B\, Q_B\, T \times \, LTerm^{\wedge}\, C\, Q_C\,
-    t' }
+\quad\quad = & \!\!\mathsf{\exists [Q_B]\, [Q_C]\, Equal^{\wedge} \, A\, (B \to
+  C)\, K_\top\, (Arr^{\wedge} \, B\, C\, Q_B \, Q_C)\, e }\\
+ & \quad\quad\mathsf{\times \, LType^{\wedge}\, B\, Q_B\, T \times \,
+  LTerm^{\wedge}\, C\, Q_C\,t' }
 \end{array}\]
@@
 $\mathsf{Q_B}$ and $\mathsf{Q_C}$ be $\mathsf{K_\top}$, which means we
-need proofs of $\mathsf{Equal^{\wedge} \, A\, (B \to C)\, K_\top\,
+need proofs of
+$\mathsf{Equal^{\wedge} \, A\, (B \to C)}$ $\mathsf{K_\top\,
   (Arr^{\wedge} \, B\, C\, K_\top \, K_\top)\, e}$,
 $\mathsf{LType^{\wedge}\, B\, K_\top\, T}$ and
-$\mathsf{LTerm^{\wedge}\, C\, K_\top\, t'}$.  We take
+$\mathsf{LTerm^{\wedge}\, C}$ $\mathsf{K_\top\, t'}$.  We take
 $\mathsf{LType^\wedge KT\, B\, T}$ and $\mathsf{LTerm^\wedge KT\, C\,
@@
 issue in more detail at the end of the section, but for now we simply
-assume a proof $\mathsf{Equal^\wedge ArrKT : Equal^{\wedge} \, A\, (B
-  \to C)\, K_\top\, (Arr^{\wedge} \, B\, C\, K_\top \, K_\top)\, e}$
-and define $\mathsf{LTerm^\wedge KT\,A\, (abs\, B \,C \, e \,s \,T \,
-  t') = (K_\top , K_\top , Equal^\wedge ArrKT , LType^\wedge KT\, B\,
-  T , LTerm^\wedge KT\, C\, t') }$.
+assume a proof \[\begin{array}{ll}
+&\!\! \mathsf{Equal^\wedge ArrKT}\\
+\quad\quad  : & \!\!\mathsf{Equal^{\wedge} \, A\, (B
+  \to C)\, K_\top\, (Arr^{\wedge} \, B\, C\, K_\top \, K_\top)\, e}
+\end{array}\]
+and define
+\[\begin{array}{ll}
+ & \!\!\mathsf{LTerm^\wedge KT\,A\, (abs\, B \,C \, e \,s \,T \,
+  t')}\\
+\quad\quad= & \!\!\mathsf{(K_\top , K_\top , Equal^\wedge ArrKT ,
+  LType^\wedge KT\, B\, T , LTerm^\wedge KT\, C\, t') }
+\end{array}\]
 \item If $\mathsf{t_1 : LTerm\,(B \to A)}$ and $\mathsf{t_2 :
   LTerm\,B}$ then, by the same reasoning as in the previous case, we
-  need to define $\mathsf{LTerm^\wedge KT\,A\, (app\, B \, t_1\,t_2)}
-  : \mathsf{LTerm^{\wedge}\, (B \to A)\, (Arr^{\wedge} \, B\, A\,
-    K_\top \, K_\top)\, t_1 \times LTerm^{\wedge}\, B\, K_\top\,
-    t_2}$.  We define the second component of the pair to be
+  need to define
+  \[\begin{array}{ll}
+ & \!\! \mathsf{LTerm^\wedge KT\,A\, (app\, B \, t_1\,t_2)}\\
+\quad\quad : &\!\! \mathsf{LTerm^{\wedge}\, (B \to A)\, (Arr^{\wedge} \, B\, A\,
+  K_\top \, K_\top)\, t_1\; \times}\\
+ & \hspace*{1.5in} \mathsf{LTerm^{\wedge}\, B\, K_\top\, t_2}
+  \end{array}\]
+  We define the second component of the pair to be
   $\mathsf{LTerm^\wedge KT\,B\,t_2}$. We can define the first
   component from a proof of $\mathsf{LTerm^{\wedge}\, (B \to A)\,
-    K_\top\, t_1}$ and the function $\mathsf{LTerm^\wedge EqualMap :
-    \forall\, (A : Set)\, (Q\,Q' : A \to Set) \to
-    Equal^\wedge\,A\,A\,Q\,Q'\,refl \to}$
-  $\mathsf{PredMap\,(LTerm\,A)\,
-    (LTerm^{\wedge}\,A\,Q)\,(LTerm^{\wedge}\,A\,Q')}$ that takes two
+    K_\top\, t_1}$ and the function
+  \[\begin{array}{ll}
+  & \!\! \mathsf{LTerm^\wedge EqualMap}\\
+  \quad\quad : &
+  \!\!\mathsf{\forall\, (A : Set)\, (Q\,Q' : A \to
+    Set) \to}\\
+  & \;\;\mathsf{Equal^\wedge\,A\,A\,Q\,Q'\,refl \to}\\
+  &  \;\;\;\;\mathsf{PredMap\,(LTerm\,A)\,
+    (LTerm^{\wedge}\,A\,Q)\,(LTerm^{\wedge}\,A\,Q')}
+  \end{array}\] that takes two
   (extensionally) equal predicates with the same carrier and produces
@@
   $\mathsf{PredMap}$ is constructed using $\mathsf{var}$ and
-  $\mathsf{app}$, respectively. Taking $\mathsf{L_{K_\top} :
-    LTerm^{\wedge}\, (B \to A) \, K_\top\, t_1}$ to be the proof
-  $\mathsf{L_{K_\top} = LTerm^\wedge KT\, (B \to A)\, t_1}$ and taking
-  $\mathsf{ LTerm^\wedge Arr : LTerm^{\wedge}\, (B \to A)\,
-    (Arr^{\wedge} \, B\, A\, K_\top \, K_\top)\, t_1}$ to be the proof
-  $\mathsf{LTerm^\wedge Arr =}$ $\mathsf{LTerm^\wedge
-    EqualMap\,K_\top\,(Arr^{\wedge}\,B\,A\,K_\top\,K_\top)\,\,
-    Equal^\wedge ArrKT\, t_1\, L_{K_\top}}$, we define
-  $\mathsf{LTerm^\wedge KT\,A\, (app\, B\, t_1\, t_2) = (K_\top ,
-    LTerm^\wedge Arr , LTerm^\wedge KT\,B\,t_2)}$.
+  $\mathsf{app}$, respectively. If $\mathsf{L_{K_\top} :
+    LTerm^{\wedge}\, (B \to A) \, K_\top\, t_1}$ is the proof
+  $\mathsf{L_{K_\top} = LTerm^\wedge KT\, (B \to A)\, t_1}$ and
+  $\mathsf{ LTerm^\wedge Arr}$ $\mathsf{: LTerm^{\wedge}\, (B \to A)\,
+    (Arr^{\wedge} \, B\, A\, K_\top \, K_\top)\, t_1}$ is the proof
+  \[\begin{array}{l}
+   \quad\quad  \mathsf{LTerm^\wedge Arr = LTerm^\wedge
+    EqualMap\,K_\top\,(Arr^{\wedge}\,B\,A\,K_\top\,K_\top)}\\
+  \hspace*{1.5in} \mathsf{Equal^\wedge ArrKT\, t_1\, L_{K_\top}}
+  \end{array}\]
+  then we define
+  \[\begin{array}{ll}
+    & \mathsf{LTerm^\wedge KT\,A\, (app\, B\, t_1\, t_2)} \\
+  = & \mathsf{(K_\top ,  LTerm^\wedge Arr , LTerm^\wedge KT\,B\,t_2)}
+    \end{array}\]
 \item If $\mathsf{e : Equal\,A\,(List\,B)}$ and $\mathsf{ts :
   List\,(LTerm\, B)}$ then, as above, we need to define
-  $\mathsf{LTerm^\wedge KT\,A\, (list\, B \,e\, ts)}$ $\mathsf{:
-  Equal^{\wedge} \, A\, (List\,B)\, K_\top\, (List^{\wedge} \, B\,
-  K_\top) \, e \times List^{\wedge}\, (LTerm\,B) \, (LTerm^{\wedge} \,
-  B\, K_\top) \, ts }$.  As in that case we assume a proof
-  $\mathsf{Equal^\wedge ListKT : Equal^{\wedge} \, A\, (List\,B)\, K_\top\,
-  (List^{\wedge} \, B\, K_\top) \, e}$ for the first component.  We
+  \[\begin{array}{ll}
+  & \!\!\mathsf{LTerm^\wedge KT\,A\, (list\, B \,e\, ts)}\\
+\quad\quad : & \!\!\mathsf{Equal^{\wedge} \, A\, (List\,B)\, K_\top\,
+  (List^{\wedge} \, B\, K_\top) \, e\, \times}\\
+ & \quad\mathsf{List^{\wedge}\, (LTerm\,B)
+  \, (LTerm^{\wedge} \, B\, K_\top) \, ts }
+  \end{array}\]
+  As in that case we assume a proof
+  \[\;\;\quad\mathsf{Equal^\wedge ListKT : Equal^{\wedge} \, A\, (List\,B)\, K_\top\,
+  (List^{\wedge} \, B\, K_\top) \, e}\] for the first component.  We
   can define the second component using $\mathsf{liftListMap}$ from
@@
     (List^{\wedge}\,(LTerm\,B)\, (LTerm^{\wedge}\,B\,K_\top))}$ of
-  lifted predicates.  Taking $\mathsf{m_{K_\top} : PredMap\,
-    (LTerm\,B) \, (K_\top)\, (LTerm^{\wedge}\,B\,K_\top)}$ to be the
-  proof $\mathsf{m_{K_\top} \, t'\, tt\, = LTerm^\wedge KT\, B\, t'}$,
+  lifted predicates.  Taking
+  \[\mathsf{m_{K_\top} : PredMap\,
+    (LTerm\,B) \, (K_\top)\, (LTerm^{\wedge}\,B\,K_\top)}\] to be the
+  proof \[\mathsf{m_{K_\top} \, t'\, tt\, = LTerm^\wedge KT\, B\, t'}\]
   where $\mathsf{t' : LTerm\,B}$ and $\mathsf{tt}$ is the single
   element of $\mathsf{K_\top\, t'}$, and taking
-  $\mathsf{L_{List^\wedge LTerm^\wedge KT} : List^{\wedge}\,
-    (LTerm\,B) \, (LTerm^{\wedge} \, B\, K_\top) \, ts}$ to be the
-  proof $\mathsf{ L_{List^\wedge LTerm^\wedge KT} =}$
-  $\mathsf{liftListMap \, (LTerm\,B) \, K_\top \, (LTerm^{\wedge}\,
-    B\, K_\top) \, m_{K_\top} \, ts \, (List^\wedge KT\, (LTerm\, B)\,
-    ts) }$, we define $\mathsf{LTerm^\wedge KT\,A\, (list\, B\, e\,
-    ts) = (K_\top , \, Equal^\wedge ListKT , \, L_{List^\wedge
-      LTerm^\wedge KT} ) }$.
+  \[\mathsf{L_{List^\wedge LTerm^\wedge KT} : List^{\wedge}\,
+    (LTerm\,B) \, (LTerm^{\wedge} \, B\, K_\top) \, ts}\] to be the
+  proof
+\[\begin{array}{ll}
+& \!\!\mathsf{ L_{List^\wedge LTerm^\wedge KT}}\\
+\quad\quad= & 
+ \!\!\mathsf{liftListMap \, (LTerm\,B) \, K_\top \, (LTerm^{\wedge}\,
+   B\, K_\top) \, m_{K_\top} \, ts}\\
+ & \!\!\quad \mathsf{(List^\wedge KT\, (LTerm\, B)\,
+   ts) }
+\end{array}\]
+we define
+\[\begin{array}{ll}
+ & \!\!\mathsf{LTerm^\wedge KT\,A\, (list\, B\, e\,
+  ts)}\\
+\quad\quad = & \!\!\mathsf{(K_\top , \, Equal^\wedge ListKT , \, L_{List^\wedge
+    LTerm^\wedge KT} ) }
+\end{array}\]
 \end{itemize}
@@
 $\mathsf{G^\wedge KT : \forall\, (A : Set)\, (x : G\,A) \to
-  G^{\wedge}\, A\, K_\top\, x}$ for an arbitrary GADT $\mathsf{G}$ as
-defined in Section~\ref{sec:framework}.  To provide a proof of
+  G^{\wedge}\, A\, K_\top\, x}$ for any GADT $\mathsf{G}$ as defined
+in Section~\ref{sec:framework}.  To provide a proof of
 $\mathsf{G^{\wedge}\, A\, K_\top \, x}$ for every term $\mathsf{x :
@@
   , b) : A \times B}$, we have that
-$\mathsf{Pair^{\wedge}\,A\,B\,K_\top\,K_\top (a, b) = K_\top \, a
-  \times K_\top\, b = \top \times \top}$, whereas $\mathsf{K_\top\,
+\[\mathsf{Pair^{\wedge}\,A\,B\,K_\top\,K_\top (a, b) = K_\top \, a
+  \times K_\top\, b = \top \times \top}\] whereas $\mathsf{K_\top\,
   (a, b) = \top}$. While these types are not equal, they are clearly
@@
 (deep) induction rules. Our development is implemented in Agda, as is
-our case study from Section~\ref{sec:app}.
-
-\pagebreak
-
-\begin{thebibliography}{10}\label{bibliography}
-
-\bibitem{atk12} Atkey, R.  {\em Relational parametricity for higher
-  kinds}.  Proceedings, Computer Science Logic, pp.~46-61, 2012.
-
-\bibitem{bfss90} Bainbridge, E. S., Freyd, P. Scedrov, A., and Scott,
-  P. J. {\em Functorial polymorphism}. Theoretical Computer Science
-  70(1), pp. 35-64, 1990.
-
-\bibitem{bm98} Bird, R. and Meertens, L. {\em Nested
-  datatypes}. Proceedings, Mathematics of Program Construction,
-  pp. 52–67, 1998.
-
-\bibitem{ch03} Cheney, J. and Hinze, R. {\em First-class phantom
- types}. CUCIS TR2003-1901, Cornell University, 2003.
-
-\bibitem{cis194}
-Zilberstein, N. CIS194 homepage.
-$\mathtt{https://www.seas.upenn.edu/\tilde~cis194/spring15/lectures/11\!\!-\!\!stlc.html}$
-
-%\bibitem{chl} Chlipala, A. {\em Library Inductive
-%  Types}. $\mathtt{http://adam.chlipala.net/cpdt/html/InductiveTypes.html}$
-
-\bibitem{coq20} The Coq Development Team. {\em The Coq Proof
-  Assistant}, version 8.11.0, January 2020.
-  $\mathtt{https://doi.org/10.5281/zenodo.3744225}$
-
-\bibitem{dyb94} Dybjer, P. {\em Inductive Families}. Formal Aspects of
-  Computing 6(4), pp. 440-465, 1994.
-
-\bibitem{fs18} Fu, P. and Selinger, P.  Dependently typed folds for
-  nested data types, 2018. $\mathtt{https://arxiv.org/abs/1806.05230}$
-
-\bibitem{gjfor15} Ghani, N., Johann, P., Nordvall Forsberg, F.,
-  Orsanigo, F., and Revell, T. {\em Bifibrational functorial semantics
-    for parametric polymorphism}. Proceedings, Mathematical
-  Foundations of Program Semantics, pp. 165-181, 2015.
-
-\bibitem{hin03} Hinze, R. {\em Fun with phantom types}. Proceedings,
- The Fun of Programming, pp. 245–262, 2003.
-
-\bibitem{web-page} Johann, P., Ghiorzi, E., and Jeffries,
-  D. Accompanying Agda code for this paper.
-    $\mathtt{https://cs.appstate.edu/\tilde~johannp/MFPS21Code.html}$
-
-\bibitem{jgj21f} Johann, P., Ghiorzi, E., and Jeffries, D. {\em
-  Parametricity for primitive nested types}. Proceedings, Foundations
-  of Software Science and Computation Structures, pp. 324-343, 2021.
-
-\bibitem{jgj21} Johann, P., Ghiorzi, E., and Jeffries, D. {\em GADTs,
-  Functoriality, Parametricity: Pick Two}. Accepted for presentation
-  at Logical and Semantic Frameworks with Applications, 2021.
-
-\bibitem{jp19} Johann, P. and Polonsky, A. {\em Higher-kinded data
-  types: Syntax and semantics} Proceedings, Logic in Computer Science,
-  pp. 1-13, 2019.
-
-\bibitem{jp20} Johann, P. and Polonsky, A. {\em Deep induction:
-  Induction rules for (truly) nested types}.  Proceedings, Foundations
-  of Software Science and Computation Structures, pp. 339-358, 2020.
-
-\bibitem{mac71} MacLane, S. {\em Categories for the Working
-  Mathematician}. Springer, 1971.
-
-\bibitem{mcb99} McBride, C. {\em Dependently Typed Programs and their
-  Proofs}. PhD thesis, University of Edinburgh, 1999.
-
-%\bibitem{min15} Minsky, Y.  {\em Why {GADT}s matter for performance}.
-%  $\mathtt{https://blog.janestreet.com/why-gadts-matter-for-performance/}$,
-%  2015.
-
-%\bibitem{pl04} Pasalic, E., and Linger, N.  {\em Meta-programming with
-%  typed object-language representations}.  Generic Programming and
-%  Component Engineering, pp.~136-167, 2004.
-
-%\bibitem{pen20} Penner, C.  {\em Simpler and safer {API} design using
-%  {GADT}s}.  $\mathtt{https://chrispenner.ca/posts/gadt-design}$,
-%  2020.
-
-\bibitem{pvww06} Peyton Jones, S., Vytiniotis, D., Weirich, S., and
-  Washburn, G. {\em Simple unification-based type inference for
-    GADTs}. Proceedings, International Conference on Functional
-  Programming, pp. 50-61, 2006.
-
-%\bibitem{pr06} Pottier, F., and R{\'e}gis-Gianas, Y.  {\em Stratified
-%  type inference for generalized algebraic data types}.  Principles of
-%  Programming Languages, pp.~232-244, 2006.
-
-\bibitem{sjsv09} Schrijvers, T., Peyton Jones, S. L., Sulzmann, M., and
-  Vytiniotis, D. {\em Complete and decidable type inference for
-    GADTs}. Proceedings, International Conference on Functional
-  Programming, pp. 341– 352, 2009.
-
-\bibitem{sp04} Sheard, T., and Pasalic, E. {\em Meta-programming with
-  built-in type equality}. Proceedings, Workshop on Logical Frameworks
-  and Meta-languages, pp. 106-124, 2004.
-
-\bibitem{tas19} Tassi, E.: Deriving proved equality tests in Coq-elpi:
-  Stronger induction principles for containers in Coq. Proceedings,
-  Interactive Theorem Proving, pp. 1-18, 2019.
-
-\bibitem{vw10} Vytiniotis, D., and Weirich, S.  {\em Parametricity,
-  type equality, and higher-order polymorphism}.  Journal of
-  Functional Programming 20(2), pp.~175--210, 2010.
-
-\bibitem{xcc03} Xi, H., Chen, C. and Chen, G. {\em Guarded recursive
-  datatype constructors}. Proceedings, Principles of Programming
-  Languages, pp. 224–235, 2003.
-
-\bibitem{ull20} Ullrich, M. {\em Generating Induction Principles for
-Nested Induction Types in MetaCoq}. PhD thesis, Saarland University,
-  2020.
-
-\end{thebibliography}
+our case study from Section~\ref{sec:app}. {\color{red} Our
+  development opens the way to incorporating automatic generation of
+  (deep) induction rules for them into proof assistants.}
+
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{bibfile}
 

### 2021-09-09 induction-paper/acm-rev-ind-paper.tex
small changes to cpp ind paper

@@
 constructs sequences of data whose types are pair-structured, rather
-than sequences of arbitrary type, as does $\mathsf{const}$.
-
-If one or more of the data constructors for a GADT $\mathsf{G}$ return
+than sequences of arbitrary type, as does $\mathsf{const}$. If one or
+more of the data constructors for a GADT $\mathsf{G}$ return
 structured instances of $\mathsf{G}$, then the GADT will have two
@@
 $\mathsf{Pair^{\wedge} : \forall (A\; B: Set) \to (A \to Set) \to (B
-  \to Set)\; \to\;}$ $\mathsf{A \times B \to}$ $\mathsf{Set}$ lifts predicates
+  \to Set)\; \to\;}$ $\mathsf{A \times B \to}$ $\mathsf{Set}$
+lifts predicates
 $\mathsf{Q_A}$ on data of type $\mathsf{A}$ and $\mathsf{Q_B}$ on data
@@
 deep induction rule for bushes is given by the second expression in
-Figure~\ref{fig:ptree-and-bush}, where $\mathsf{Bush^{\wedge} :
-  \forall (A : Set) \to (A \to Set)}$ $\mathsf{\to Bush\,A \to Set}$
+Figure~\ref{fig:ptree-and-bush}, where
+\[\mathsf{Bush^{\wedge} :
+  \forall (A : Set) \to (A \to Set) \to Bush\,A \to Set}\]
 is the following lifting of a predicate $\mathsf{Q}$ on data of type
@@
 its Henry Ford encoding from~\eqref{eq:eq_seq}.  We first define its
-predicate lifting $\mathsf{Seq^\wedge : \forall (A : Set) \to (A \to
-  Set) \to Seq\,A \to Set}$ as in Figure~\ref{fig:seq}. There,
+predicate lifting
+\[\mathsf{Seq^\wedge : \forall (A : Set) \to (A \to
+  Set) \to Seq\,A \to Set}\] as in Figure~\ref{fig:seq}. There,
 $\mathsf{a : A}$, $\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C : C \to
@@
 
-\caption{Applies induction hypotheses for
+\caption{Applied induction hypotheses for
   $\mathsf{LTerm}$}\label{fig:ind-hyps-lterm-applied-to-P}
@@
 \mathsf{data\ Maybe : Set \to Set\ where}\\
-\mathsf{\;\;\;\;\;\;nothing :\,  \forall \{A : Set\} \to Maybe\,A}\\
-\mathsf{\;\;\;\;\;\;just\;\;\;\;\;\;\; :\,  \forall \{A : Set\} \to A \to Maybe\,A}
+\mathsf{\;\;nothing :\,  \forall \{A : Set\} \to Maybe\,A}\\
+\mathsf{\;\;just\;\;\;\;\;\;\; :\,  \forall \{A : Set\} \to A \to Maybe\,A}
 \end{array}
@@
   t')}\\
-\quad\quad= & \!\!\mathsf{(K_\top , K_\top , Equal^\wedge ArrKT ,
-  LType^\wedge KT\, B\, T , LTerm^\wedge KT\, C\, t') }
+\quad\quad= & \!\!\mathsf{(K_\top , K_\top , Equal^\wedge ArrKT ,}\\
+ & \hspace*{0.3in} \mathsf{LType^\wedge KT\, B\, T , LTerm^\wedge KT\, C\, t') }
 \end{array}\]

### 2021-09-14 induction-paper/acm-rev-ind-paper.tex
final version of CPP submission

@@
   via structural induction.  In this paper we show how to extend deep
-  induction to GADTs that are not truly nested GADTs.  {\color{red}
-    This opens the way to incorporating automatic generation of (deep)
-    induction rules for them into proof assistants.}  We also show
-  that the techniques developed in this paper do not suffice for
-  extending deep induction to truly nested GADTs, {\color{red} so more
-    sophisticated techniques are needed to derive deep induction rules
-    for them.}
+  induction to GADTs that are not truly nested GADTs.  This opens the
+  way to incorporating automatic generation of (deep) induction rules
+  for them into proof assistants.  We also show that the techniques
+  developed in this paper do not suffice for extending deep induction
+  to truly nested GADTs, so more sophisticated techniques are needed
+  to derive deep induction rules for them.
 \end{abstract}
@@
 This paper shows how to extend deep induction to proper GADTs, i.e.,
-GADTs that are not nested types (and thus are not ADTs).  A
-constructor for a GADT $\mathsf{G}$ may, like a constructor for a
-nested type, take as arguments data whose types involve instances of
-$\mathsf{G}$ other than the one being defined. These can even include
-instances involving $\mathsf{G}$ itself. But if $\mathsf{G}$ is a
-proper GADT, then at least one of its constructors will also have a
-structured instance of $\mathsf{G}$ --- albeit one not involving
-$\mathsf{G}$ itself --- as its codomain.
+GADTs that are not nested types (and thus are not ADTs).  Typical
+applications of such GADTs include generic programming, modeling
+programming languages via higher-order abstract syntax, maintaining
+invariants in data structures, and expressing constraints in embedded
+domain-specific languages. They have also been used to implement
+tagless interpreters~\cite{pl04,pr06,pvww06} by trading the definition
+of a universal value domain for a direct specification of the property
+of being a value. Other applications are described in,
+e.g.,~\cite{min15,rou06}. A constructor for a GADT $\mathsf{G}$ may,
+like a constructor for a nested type, take as arguments data whose
+types involve instances of $\mathsf{G}$ other than the one being
+defined. These can even include instances involving $\mathsf{G}$
+itself. But if $\mathsf{G}$ is a proper GADT, then at least one of its
+constructors will also have a structured instance of $\mathsf{G}$ ---
+albeit one not involving $\mathsf{G}$ itself --- as its codomain.
 %But while the types of the arguments to a GADT's constructor can
@@
 semantics interpreting it as the interpretation of its Church
-encoding~\cite{atk12,vw10}. As detailed in~\cite{jgj21}, a key
+encoding~\cite{atk12,vw10}. As detailed in~\cite{jg21}, a key
 difference in the two semantics is that the former views GADTs as
@@
 types~\cite{bfss90,gjfor15,jgj21f}, it is not at all clear how to
-define a functorial parametric semantics for GADTs~\cite{jgj21}.
+define a functorial parametric semantics for GADTs~\cite{jg21}.
 
@@
 have been derived by instantiating our general framework. Our Agda
-code implementing them is available at~\cite{web-page}.
+implemention of them is available in the accompanying code.\looseness=-1
+%at~\cite{web-page}.
 
@@
   functions over them we must use the $\mathsf{TERMINATING}$ flag.
-  {\color{red}(Similar flags are required in Coq.)} Although as
-  programmers we know from the metatheory in~\cite{jp19} that
-  $\mathsf{Bush}$ is well-defined and the functions we define over
-  them terminate, the flags are necessary because Agda fails to infer
-  these facts. Analogous comments apply at several places
+  (Similar flags are required in Coq.) Although as programmers we know
+  from the metatheory in~\cite{jp19} that $\mathsf{Bush}$ is
+  well-defined and the functions we define over them terminate, the
+  flags are necessary because Agda fails to infer these
+  facts. Analogous comments apply at several places
   below.}\cite{bm98}:
@@
 Types are either booleans, arrow types, or list types. They are
-represented by the Henry Ford GADT $\mathsf{LType}$.  Terms are either
-variables, abstractions, applications, or lists of terms. They are
-similarly represented by the Henry Ford GADT $\mathsf{LTerm}$.  The
-type parameter for $\mathsf{LTerm}$ tracks the types of simply typed
-lambda calculus terms. For example, $\mathsf{LTerm\,A}$ is the type of
-simply typed lambda terms of type $\mathsf{A}$. Variables are tagged
-with their types by the data constructors $\mathsf{var}$ and
+represented by the Henry Ford GADT $\mathsf{LType}$ in
+Figure~\ref{fig:type-and-term}.  Terms are either variables,
+abstractions, applications, or lists of terms. They are similarly
+represented by the Henry Ford GADT $\mathsf{LTerm}$.  The type
+parameter for $\mathsf{LTerm}$ tracks the types of simply typed lambda
+calculus terms. For example, $\mathsf{LTerm\,A}$ is the type of simply
+typed lambda terms of type $\mathsf{A}$. Variables are tagged with
+their types by the data constructors $\mathsf{var}$ and
 $\mathsf{abs}$, whose $\mathsf{LType}$ arguments ensure that their
@@
 {\small
-\[\begin{array}{l}
-\mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to
+\[\begin{array}{lll}
+\mathsf{dIndVar} & = & \mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to
   LTerm\,A  \to Set) \to}\\
-  \quad\mathsf{\forall (A : Set) (Q_A : A \to Set) (s : String) (T_A :
+& &   \quad\mathsf{\forall (A : Set) (Q_A : A \to Set) (s : String) (T_A :
   LType\, A) \to LType^{\wedge} \, A\, Q_A\, T_A \to P \, A\, Q_A\,
  (var \; s\, T_A)}\\[1ex]
-  \mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to
+\mathsf{dIndAbs} & = & \mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to
   LTerm\,A \to Set) \to} \\ 
-  \quad\mathsf{
+& &   \quad\mathsf{
   \forall (A\,B\,C: Set) (Q_A : A \to Set)  (Q_B : B \to Set) (Q_C : C
   \to Set) (e : Equal\, A\, (B \to C)) (s : String) \to } \\ 
-  \quad\mathsf{(T_B : LType\, B) \to (t_C : LTerm\, C)
+& &   \quad\mathsf{(T_B : LType\, B) \to (t_C : LTerm\, C)
   \to Equal^{\wedge}\,A\,(B \to C)\, Q_A \, (Arr^{\wedge} \, B\, C\,
   Q_B \, Q_C) \, e \to  } \\
-  \quad\mathsf{
+& &   \quad\mathsf{
   LType^{\wedge}\, B\, Q_B\, T_B
@@
  \\
-  \mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A
+\mathsf{dIndApp} & = &  \mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A
     \to Set)\to} \\ 
-  \quad \mathsf{
+& &  \quad \mathsf{
   \forall (A \,B : Set) (Q_A : A \to Set)  (Q_B : B \to Set) 
    (t_{BA} : LTerm\, (B \to A)) (t_B : LTerm\, B) \to} \\
-  \quad \mathsf{
+& &   \quad \mathsf{
   P\, (B \to A)\, (Arr^{\wedge} \, B\, A\, Q_B \, Q_A) \, t_{BA} \, 
@@
   \to P \, A\, Q_A\, (app \,B \,t_{BA} \, t_B) }\\[1ex]
-  \mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A
+\mathsf{dIndList} & = &   \mathsf{\lambda (P : \forall (A : Set) \to (A \to Set) \to LTerm\,A
     \to Set) \to} \\ 
-  \quad \mathsf{
+& &   \quad \mathsf{
   \forall (A \,B : Set) (Q_A : A \to Set)  (Q_B : B \to Set) 
     (e : Equal\, A\, (List\, B)) (ts : List\, (LTerm\, B)) \to} \\ 
-  \quad \mathsf{
+& &   \quad \mathsf{
     Equal^{\wedge}\, A\, (List\,B)\, Q_A\, (List^{\wedge}\, B\, Q_B)\, e 
@@
   Set) \to Set }$ is the type constructor producing the type of
-morphisms between predicates defined by
+morphisms between predicates defined by\looseness=-1
 \[\mathsf{PredMap \,A\, Q\,Q'\,
@@
 of $\mathsf{\top}$), and by $\mathsf{liftListMap\, A\, Q\, Q'\, m\,
-  (cons\, a\, l')\, (y, x') = (m\,a\,y,}$\\ $\mathsf{liftListMap\, A\,
+  (cons\, a\, l')\, (y, x') =}$\\ $\mathsf{ (m\,a\,y, \,liftListMap\, A\,
   Q\, Q'\, m\, l'\, x')}$ (since $\mathsf{x : List^{\wedge}\, A\,
@@
 
-We can generalize the approach taken in
-Section~\ref{sec:deep-ind-GADTs} to a general framework for deriving
-deep induction rules for GADTs that are not truly nested GADTs. We
-will treat GADTs of the form
+We can generalize the approach in Section~\ref{sec:deep-ind-GADTs} to
+a general framework for deriving deep induction rules for GADTs that
+are not truly nested GADTs. We will treat GADTs of the form
 
@@
 At a deeper level, the fundamental issue is that the $\mathsf{Equal}$
-type does not have functorial semantics~\cite{jgj21}, so that having
+type does not have functorial semantics~\cite{jg21}, so that having
 morphisms $\mathsf{A \to A'}$ and $\mathsf{B \to B'}$ (for any type
@@
 consequence of unary parametricity. At the object level, our Agda
-code~\cite{web-page} simply postulates each isomorphism needed since
+code
+%~\cite{web-page}
+simply postulates each isomorphism needed since
 an Agda implementation of full parametricity for some relevant
@@
 (deep) induction rules. Our development is implemented in Agda, as is
-our case study from Section~\ref{sec:app}. {\color{red} Our
-  development opens the way to incorporating automatic generation of
-  (deep) induction rules for them into proof assistants.}
+our case study from Section~\ref{sec:app}. Our development opens the
+way to incorporating automatic generation of (deep) induction rules
+for them into proof assistants.
 

### 2021-10-27 21-lmcs/21-lmcs.tex
final LMCS paper

@@
 primitive fixpoints, so they cannot be used to obtain the results of
-this paper.  (See the past paragraph of this section for a more
+this paper.  (See the last paragraph of this section for a more
 detailed discussion in the case of F$_\omega$.) Whether or not these
@@
 \end{array}\]
-respectively. Our calculus can similarly define a $\mathtt{rustle}$
-function that changes the placement of the data in a bush. This
-function maps, e.g., the bush
+respectively. Here, if $\Gamma; \emptyset \,|\, \Delta \vdash t:
+\Nat^{\overline{\alpha}} F\,G$ and $\Gamma; \emptyset \,|\, \Delta
+\vdash s: \Nat^{\overline{\alpha}} G\,H$ are terms then the {\em
+  composition} $s \circ t$ of $t$ and $s$ is defined by $s \circ t =
+\Gamma; \emptyset\,|\, \Delta \vdash L_{\overline{\alpha}}
+x. s_{\overline{\alpha}}(t_{\overline{\alpha}}x):
+\Nat^{\overline{\alpha}} F\,H$. Our calculus can similarly define a
+$\mathtt{rustle}$ function that changes the placement of the data in a
+bush. This function maps, e.g., the bush
 \[\begin{array}{l}
@@
 $\id_F = \Gamma;\emptyset~|~\emptyset \vdash L_{\ol{\alpha}}x.x :
-\Nat^{\ol{\alpha}} F\,F$, and if $\Gamma; \emptyset \,|\, \Delta
-\vdash t: \Nat^{\overline{\alpha}} F\,G$ and $\Gamma; \emptyset \,|\,
-\Delta \vdash s: \Nat^{\overline{\alpha}} G\,H$ are terms then define
-the {\em composition} $s \circ t$ of $t$ and $s$ by $s \circ t =
-\Gamma; \emptyset\,|\, \Delta \vdash L_{\overline{\alpha}}
+\Nat^{\ol{\alpha}} F\,F$. Also, recall from Section~\ref{sec:terms} that if
+$\Gamma; \emptyset \,|\, \Delta \vdash t: \Nat^{\overline{\alpha}}
+F\,G$ and $\Gamma; \emptyset \,|\, \Delta \vdash s:
+\Nat^{\overline{\alpha}} G\,H$ are terms then the composition $s \circ
+t$ of $t$ and $s$ is defined by $s \circ t = \Gamma; \emptyset\,|\,
+\Delta \vdash L_{\overline{\alpha}}
 x. s_{\overline{\alpha}}(t_{\overline{\alpha}}x):
@@
 
-\subsection{Extending the Type Semantics}
+\subsection{Extending the Type Semantics}\label{sec:type-sem}
 
@@
 $\Lan$-types, as well as of terms of both such types, in both $\set$
-and $\rel$. Recalling that the type constructor variables bound by
-$\Nat$ in the formation rule for generalized $\Nat$-types are
-interpreted in $\set$ as functors and in $\rel$ as relation
-transformers, we let $k_i$ be the arity of the $i^{th}$ element of the
-sequence $\Phi$ and extend the set and relational interpretations for
-$\Nat$ types from Section~\ref{sec:type-interp} as follows:
+and $\rel$. (Note that it is having both $\mathsf{Set}$ and
+$\mathsf{Rel}$ interpretations, and having these be interconnected in
+the specific way parametricity demands, that is at issue here; if we
+are only interested in having {\em some} model of the type calculus,
+then the one in~\cite{jp19} will do.) Recalling that the type
+constructor variables bound by $\Nat$ in the formation rule for
+generalized $\Nat$-types are interpreted in $\set$ as functors and in
+$\rel$ as relation transformers, we let $k_i$ be the arity of the
+$i^{th}$ element of the sequence $\Phi$ and extend the set and
+relational interpretations for $\Nat$ types from
+Section~\ref{sec:type-interp} as follows:
 \begin{align*}
@@
 \rightarrow \relsem{\Gamma; \Phi \vdash G}\Eq_{\rho}[\Phi := \ol{K}] \} \\
-%%%
+\end{align*}
+\begin{align*}
 \relsem{\Gamma; \emptyset \vdash \Nat^\Phi \,F\,G}\rho
@@
 Recalling the notation for the colimit representation of left Kan
-extensions from Section~\ref{sec:lke}, we define the set and
+extensions from Section~\ref{sec:lke}, we can define the set and
 relational interpretations of the terms from the introduction and
@@
   \Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}\,F) \ol \beta\;
-  G}$ are defined entirely analogously.
+  G}$ can then be defined entirely analogously.
 
@@
 %    \ol N, \ol A ]}$.
-An analogous result holds for relational
-interpretations of $\Lan$-types.
+An analogous result holds for relational interpretations of
+$\Lan$-types. This shows that the machinery of this section does not
+give a well-defined relational semantics and, given the expected
+properties of the term semantics, this cannot be
+fixed by restricting the relational semantics of $\mathsf{Lan}$-types
+from Section~\ref{sec:type-sem}.
 
@@
 Proposition~\ref{prop:full-lan}, suggests that the IEL may fail for
-them. We have been able neither to prove nor disprove it thus far, but
-our substantial and lengthy investigations into the issue lead us to
-suspect that it may not actually be possible to define a functorial
-semantics for $\Lan$-types that gives rise to parametric models for
-languages supporting even constant-free polynomial primitive GADTs.
+them. As noted above, this would derail not just parametricity, but
+the well-definedness of the term semantics as well.  We have been able
+neither to prove nor disprove the IEL thus far, but our substantial
+and lengthy investigations into the issue lead us to suspect that it
+may not actually be possible to define a functorial semantics for
+$\Lan$-types that gives rise to parametric models for languages
+supporting even constant-free polynomial primitive GADTs.
 

### 2021-11-30 induction-paper/cpp2022/cpp2022-paper-final.tex
added final cpp paper and bib file

--- 2021-09-14 induction-paper/acm-rev-ind-paper.tex
@@
-\documentclass[sigplan,10pt,anonymous,review]{acmart}
+\documentclass[sigplan,10pt]{acmart}
 \settopmatter{printfolios=true,printccs=false,printacmref=false}
@@
 \title[(Deep) Induction Rules for GADTs]{(Deep) Induction Rules for GADTs}
-  \author{Patricia Johann, Enrico Ghiorzi}
-  \email{johannp,ghiorzie@appstate.edu}
+  \author{Patricia Johann}
+  \email{johannp@appstate.edu}
+\affiliation{\institution{Appalachian State University}\country{USA}}
+  \author{Enrico Ghiorzi}
+  \authornote{New address: Istituto Italiano di Tecnologia, Italy,
+      enrico.ghiorzi@iit.it}
+  \email{ghiorzie@appstate.edu}
+\affiliation{\institution{Appalachian State University}\country{USA}}
 
 \begin{abstract} 
-  Deep data types are those that are contructed from other data types,
+  Deep data types are those that are constructed from other data types,
   including, possibly, themselves.  In this case, they are said to be
@@
 rules Coq generates for them are the expected ones for standard
-structural induction. It has long been understood, however, that these
-rules are too weak to be genuinely useful for deep
-ADTs~\cite{jp20}.\footnote{An ADT/nested type/GADT is {\em deep} if it
-  is (possibly mutually inductively) defined in terms of other
+structural induction. However, as discussed in~\cite{jp20}, it has
+long been understood that these rules are too weak to be genuinely
+useful for deep ADTs.\footnote{An ADT/nested type/GADT is {\em deep}
+  if it is (possibly mutually inductively) defined in terms of other
   ADTs/nested types/GADTs (including, possibly, itself).}  The
@@
 contains.  With data types becoming ever more advanced, and with
-deeply structured such types becoming increasingly ubiquitous in
+deeply structured types becoming increasingly ubiquitous in
 formalizations, such an enhanced notion of induction is essential if
@@
 (GADTs)~\cite{ch03,pvww06,sp04,xcc03}, more richly indexed
-families~\cite{dyb94}, and deep variants of all of these.
+families~\cite{dyb94}, and deep variants of all of these. A listing of
+the various data types considered in this paper is given in
+Table~\ref{fig:taxonomy}.
+
+\begin{table}[h!]
+\begin{tabular}{|l|l|l|}
+\hline
+Data types & Discussed & Examples \\
+          & in Sections &          \\
+\hline\hline
+ADTs & \ref{sec:intro} &
+$\mathsf{List}$, $\mathsf{Rose}$ \\ 
+Nested types & \ref{sec:ADTs-and-nesteds} & $\mathsf{PTree}$ \\  
+Truly nested types & \ref{sec:ADTs-and-nesteds} & $\mathsf{Bush}$\\
+GADT & \ref{sec:GADTs} & $\mathsf{Eq}$, $\mathsf{Seq}$,\\
+Truly nested GADTs & \ref{sec:GADTs} and~\ref{sec:GADT-nested}&
+$\mathsf{G}$\\ 
+\hline
+\end{tabular}
+\caption{Data types in this paper}\label{fig:taxonomy}
+\end{table}
 
@@
 more data than just those expressible in syntax. By contrast, the
-latter views them as what might be called {\em syntax-only}
-GADTs. Happily, these two views of GADTs coincide for those GADTs that
-are ADTs or (other, including truly) nested types.  However, both they
-and their attendant properties differ greatly for proper GADTs. In
-fact, the functorial and parametric semantics for proper GADTs are
-sufficiently disparate that, by contrast with the semantics
+latter views them as what might be called {\em syntax-only}\/
+GADTs. Fortunately, these two views of GADTs coincide for those GADTs
+that are ADTs or (other, including truly) nested types.  However, both
+they and their attendant properties differ greatly for proper
+GADTs. In fact, the functorial and parametric semantics for proper
+GADTs are sufficiently disparate that, by contrast with the semantics
 customarily given for ADTs and nested
@@
 for deriving deep induction rules for GADTs that can be instantiated
-to particular such data type of interest. This framework can serve as
-a basis for extending modern proof assistants' automatic generation of
+to particular GADTs of interest. This framework can serve as a basis
+for extending modern proof assistants' automatic generation of
 structural induction rules for ADTs to automatic generation of deep
@@
 functoriality is precisely what interpreting GADTs as the
-interpretations of their Church encodings fails to deliver. Our
-approach is to instead first give a predicate lifting styled after
-those of~\cite{jp20}, together with a (deep) induction rule, for the
-simplest --- and arguably most important --- GADT, namely the equality
-GADT~\eqref{eq:equal}. We then derive the deep induction rule for a
-more complex GADT $\mathsf{G}$ by {\em i}) using the equality GADT to
-represent $\mathsf{G}$ as its so-called {\em Henry Ford
-  encoding}~\cite{ch03,hin03,mcb99,sjsv09,sp04}, and {\em ii}) using
-the predicate liftings for the equality GADT and the other GADTs
-appearing in the definition of $\mathsf{G}$ to appropriately thread
-the custom predicates for the primitive types appearing in
-$\mathsf{G}$ throughout $\mathsf{G}$'s structure. This two-step
-process delivers deep induction rules for a very general class of
-GADTs. To illustrate, we introduce a series of increasingly complex
-GADTs as running examples in Section~\ref{sec:GADTs} and derive a deep
-induction rule for each of them in
+interpretations of their Church encodings fails to deliver;
+see~\cite{jg08} for a discussion of why $\mathsf{Seq}$, e.g., is not
+functorial. Our approach is to instead first give a predicate lifting
+styled after those of~\cite{jp20}, together with a (deep) induction
+rule, for the simplest --- and arguably most important --- GADT,
+namely the equality GADT~\eqref{eq:equal}. We then derive the deep
+induction rule for a more complex GADT $\mathsf{G}$ by {\em i}) using
+the equality GADT to represent $\mathsf{G}$ as its so-called {\em
+  Henry Ford encoding}~\cite{ch03,hin03,mcb99,sjsv09,sp04}, and {\em
+  ii}) using the predicate liftings for the equality GADT and the
+other GADTs appearing in the definition of $\mathsf{G}$ to
+appropriately thread the custom predicates for the primitive types
+appearing in $\mathsf{G}$ throughout $\mathsf{G}$'s structure. This
+two-step process delivers deep induction rules for a very general
+class of GADTs. To illustrate, we introduce a series of increasingly
+complex GADTs as running examples in Section~\ref{sec:GADTs} and
+derive a deep induction rule for each of them in
 Section~\ref{sec:deep-ind-GADTs}. In particular, we derive the deep
@@
 have been derived by instantiating our general framework. Our Agda
-implemention of them is available in the accompanying code.\looseness=-1
+implementation of them is available in the accompanying code.\looseness=-1
 %at~\cite{web-page}.
@@
 parametricity, and true nesting still is not permitted.  As far as we
-know, no attempts have been made to extend either implementation to
+know no attempts have been made to extend either implementation to
 truly nested types or to proper GADTs or their deep variants. In fact,
@@
 \[\mathsf{Equal^{\wedge}\,A\,A\,Q\,Q'\,refl = \forall (a :
-  A) \to Equal\,(Q\,a)(Q'\,a)}\]
-It takes two predicates on the same type as input and tests them for
-extensional equality.  Next, we need to associate with each data
-constructor $\mathsf{c}$ of $\mathsf{G}$ an {\em induction hypothesis}
-asserting that, if the custom predicate arguments to a predicate
-$\mathsf{P}$ on $\mathsf{G}$ can be lifted to $\mathsf{G}$ itself,
-then $\mathsf{c}$ {\em respects} $\mathsf{P}$, i.e., $\mathsf{c}$
-constructs data satisfying the instance of $\mathsf{P}$ at those
-custom predicates. The following induction hypothesis
-$\mathsf{dIndRefl}$ is thus associated with the $\mathsf{refl}$
-constructor for $\mathsf{Equal}$:
+  A) \to Equal\,(Q\,a)(Q'\,a)}\] It takes two predicates on the same
+type as input and is inhabited iff they are extensionally equal.
+Next, we need to associate with each data constructor $\mathsf{c}$ of
+$\mathsf{G}$ an {\em induction hypothesis} asserting that, if the
+custom predicate arguments to a predicate $\mathsf{P}$ on $\mathsf{G}$
+can be lifted to $\mathsf{G}$ itself, then $\mathsf{c}$ {\em respects}
+$\mathsf{P}$, i.e., $\mathsf{c}$ constructs data satisfying the
+instance of $\mathsf{P}$ at those custom predicates. The following
+induction hypothesis $\mathsf{dIndRefl}$ is thus associated with the
+$\mathsf{refl}$ constructor for $\mathsf{Equal}$:
 \begin{equation*}\label{eq:ind-refl}
@@
 constructors $\mathsf{const}$ and $\mathsf{pair}$, respectively. These
-are given in Figure~\ref{fig:seq} well. Then the deep induction rule
-for $\mathsf{Seq}$ is given in the last two lines of
+are given in Figure~\ref{fig:seq} as well. Then the deep induction
+rule for $\mathsf{Seq}$ is given in the last two lines of
 Figure~\ref{fig:seq}.
@@
 
-To prove this rule is sound we define a witness $\mathsf{dIndLTerm}$
-inhabiting the type in~\eqref{eq:ind-lam} as in
+To prove that this rule is sound we define a witness
+$\mathsf{dIndLTerm}$ inhabiting the type in~\eqref{eq:ind-lam} as in
 Figure~\ref{fig:dindlterm}. There, $\mathsf{s : String}$, $\mathsf{Q_A
@@
 $\mathsf{F}$ must be constructed inductively according to the
-following grammar:
+following grammar:\label{grammar}
 \[\begin{array}{lll}
@@
 GADTs constructed from the above grammar admit predicate liftings.
-(In particular, the lifting for each type constructor $\mathsf{H}$ is
-constructed using its map function.)
+(The fact that the domain of an arrow type is independent of
+$\mathsf{G}$ is crucial for this.) In particular, the lifting for each
+type constructor $\mathsf{H}$ is constructed using its map function.
 %In fact, we have seen such liftings for
@@
 constructors. We do not give a general definition of predicate
-liftings here, though, since that would require us to first design a
-full type calculus, which is beyond the scope of the present paper.
+liftings here, since that would require us to first design a full type
+calculus, which is beyond the scope of the present paper. We can
+however, define liftings for the type constructor $\mathsf{F}$ defined
+by the grammar on page~\pageref{grammar} by
+\begin{itemize}
+\item $\mathsf{F\,G\,\ol{B} = F_1\,G\,\ol{B} \times F_2\,G\,\ol{B}}$ then
+\[\begin{array}{ll}
+& \!\!\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}}\\
+\quad\quad= & \!\!\mathsf{Pair^{\wedge}\,(F_1\,G\,\ol{B})\, (F_2\,G\,\ol{B})\,
+  (F_1^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})\,
+  (F_2^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}
+  \end{array}\]
+\item $\mathsf{F\,G\,\ol{B} = F_1\,G\,\ol{B} + F_2\,G\,\ol{B}}$ then
+\[\begin{array}{ll}
+& \!\!\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}}\\
+\quad\quad= & \!\!\mathsf{Pair^{\wedge}\,(F_1\,G\,\ol{B})\, (F_2\,G\,\ol{B})\,
+  (F_1^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})\,
+  (F_2^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}
+  \end{array}\]
+\item If $\mathsf{F\,G\,\ol{B} = F_1\,\ol{B} \to F_2\,G\,\ol{B}}$ then
+  \[\begin{array}{ll}
+  & \!\!\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}\,x}\\
+  \quad\quad= & \!\!\mathsf{\forall (z :
+  F_1\,\ol{B}) \to F_1^{\wedge}\,\ol{B}\,\ol{Q_B}\,z \to
+  F_2^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}\,(x\,z)}
+  \end{array}\]
+\item If $\mathsf{F\,G\,\ol{B} = G\,(F_1\,\ol{B})}$ and $\mathsf{F_1}$
+  does not contain $\mathsf{G}$, then
+  \[\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B} = P \,(F_1\,\ol{B})\,
+    (F_1^{\wedge}\,\ol{B}\,\ol{Q_B})}\] for all $\mathsf{P : \forall (A
+    : Set)\to}$ $\mathsf{(A \to Set) \to G\,A \to Set}$.
+\item If $\mathsf{F\,G\,\ol{B} = H\,\ol{B}}$ and $\mathsf{H}$ does not
+  contain $\mathsf{G}$, then
+  \[\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B} = H^\wedge\,\ol{Q_B}}\]
+  for all $\mathsf{P : \forall (A : Set)\to(A \to Set) \to G\,A \to}$
+  $\mathsf{ Set}$.
+\item If $\mathsf{F\,G\,\ol{B} = H\, (\ol{F_k\,G\,\ol{B}})}$ and
+  $\mathsf{H}$ does not contain $\mathsf{G}$, then
+  \[\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B} = H^{\wedge}\,
+  (\ol{F_k\,G\,\ol{B}})\,
+    (\ol{F_k^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}})}\] for all $\mathsf{P
+    : \forall (A : Set)\to(A \to Set) \to G\,A \to Set}$.
+\end{itemize}
 
@@
 & \!\!\mathsf{F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}}\\
-\quad\quad= & \!\!\mathsf{Pair^{\wedge}\,(F_1\,G\,\ol{B})\, (F_2\,G\,\ol{B})\,
+\quad\quad= & \!\!\mathsf{Pair^{\wedge}\,(F_1\,G\,\ol{B})\, (F_2\,G\,\ol{B})%\,
   (F_1^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})\,
   (F_2^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}
-  \end{array}\] The induction hypothesis
-  ensures
+  \end{array}\] The
+induction hypothesis ensures
 %{\color{red} liftings $\mathsf{F_1}$ and $\mathsf{F_2}$ (with
@@
 in the previous sections do not extend to truly nested GADTs. Indeed,
-the induction rule for a data type generally relies on (unary)
-parametricity of the model interpreting it, and deep induction for a
-truly nested type or a truly nested GADT also relies on this
+while the induction rule for a data type generally relies on (unary)
+parametricity of the model interpreting it, deep induction for a truly
+nested type or a truly nested GADT crucially relies on this
 interpretation being functorial.  Whereas ADTs and nested types both
@@
 
-In this section we use deep induction for the $\mathsf{LTerm}$ GADT
+In this section, we use deep induction for the $\mathsf{LTerm}$ GADT
 from Figure~\ref{fig:type-and-term} to infer the type from a lambda
@@
 %the type we need.  The type of $\mathsf{list_{ts}}$ is
-the type of $\mathsf{list_{ts}}$ becomes
+the type of $\mathsf{lift_{ts}}$ becomes
 \[\begin{array}{ll}

### 2021-12-02 induction-paper/cpp2022/cpp2022-paper-final.tex
finalized cpp paper

@@
 (GADTs)~\cite{ch03,pvww06,sp04,xcc03}, more richly indexed
-families~\cite{dyb94}, and deep variants of all of these. A listing of
-the various data types considered in this paper is given in
-Table~\ref{fig:taxonomy}.
+families~\cite{dyb94}, and deep variants of all of these. A summary of
+the various classes of data types considered in this paper is given in
+Table~\ref{fig:taxonomy}.\looseness=-1
 
@@
 Truly nested types & \ref{sec:ADTs-and-nesteds} & $\mathsf{Bush}$\\
-GADT & \ref{sec:GADTs} & $\mathsf{Eq}$, $\mathsf{Seq}$,\\
+GADTs & \ref{sec:GADTs} & $\mathsf{Eq}$, $\mathsf{Seq}$\\
 Truly nested GADTs & \ref{sec:GADTs} and~\ref{sec:GADT-nested}&
-$\mathsf{G}$\\ 
+$\mathsf{G}$ in \eqref{gadt-nested}\\ 
 \hline
@@
 have been derived by instantiating our general framework. Our Agda
-implementation of them is available in the accompanying code.\looseness=-1
+implementation of them is available at
+\url{https://cs.appstate.edu/~johannp/CPP22Code.html}.
+\looseness=-1
 %at~\cite{web-page}.
 
-\vspace*{0.05in}
+%\vspace*{0.05in}
 
@@
 know no attempts have been made to extend either implementation to
-truly nested types or to proper GADTs or their deep variants. In fact,
-we know of no work other than that reported here that specifically
-addresses induction rules for these data types.
+truly nested types or to proper GADTs or their deep variants.  Other
+systems, including Isabelle and Lean, also derive induction rules for
+data types that go beyond ADTs.  But we know of no work other than
+that reported here that specifically addresses induction rules for the
+(deep) GADTs considered in this paper.
 
@@
 The deep induction rule for perfect trees similarly uses polymorphic
-predicates but otherwise follows the familiar pattern. It is given
-by the first expression in Figure~\ref{fig:ptree-and-bush}. There,
+predicates but otherwise follows the familiar pattern. It is given by
+the first expression in Figure~\ref{fig:ptree-and-bush}. There,
 $\mathsf{Pair^{\wedge} : \forall (A\; B: Set) \to (A \to Set) \to (B
-  \to Set)\; \to\;}$ $\mathsf{A \times B \to}$ $\mathsf{Set}$
-lifts predicates
-$\mathsf{Q_A}$ on data of type $\mathsf{A}$ and $\mathsf{Q_B}$ on data
-of type $\mathsf{B}$ to a predicate on pairs of type $\mathsf{A \times
-  B}$ in such a way that $\mathsf{Pair^{\wedge}\,A\,B\,Q_A\,Q_B\,(a,b)
-  = Q_A\,a \times Q_B\,b}$. Similarly, $\mathsf{PTree^{\wedge}}$
-$\mathsf{\,: \,\forall (A : Set) \to (A \to Set) \to PTree\,A \to Set}$
-lifts a predicate $\mathsf{Q}$ on data of type $\mathsf{A}$ to a
-predicate on data of type $\mathsf{PTree\,A}$ asserting that
-$\mathsf{Q}$ holds for every element of type $\mathsf{A}$ contained in
-its perfect tree argument.\looseness=-1
+  \to Set)\; \to\;}$ $\mathsf{A \times B \to}$ $\mathsf{Set}$ lifts
+predicates $\mathsf{Q_A}$ on data of type $\mathsf{A}$ and
+$\mathsf{Q_B}$ on data of type $\mathsf{B}$ to a predicate on pairs of
+type $\mathsf{A \times B}$ in such a way that
+$\mathsf{Pair^{\wedge}\,A\,B\,Q_A\,Q_B\,(a,b) = Q_A\,a \times
+  Q_B\,b}$. Similarly, $\mathsf{PTree^{\wedge}}$ $\mathsf{\,:
+  \,\forall (A : Set) \to (A \to Set) \to PTree\,A \to Set}$ lifts a
+predicate $\mathsf{Q}$ on data of type $\mathsf{A}$ to a predicate on
+data of type $\mathsf{PTree\,A}$ asserting that $\mathsf{Q}$ holds for
+every element of type $\mathsf{A}$ contained in its perfect tree
+argument.  A general definition of liftings for a robust class of
+GADTs including all those appearing in the literature is given in
+Section 5.
+%\looseness=-1
 
@@
 
-We note that the structural induction rule for any (truly) nested type
-can be obtained as the special case of its deep induction rule in
-which the custom predicates are taken to be constantly
-$\mathsf{\top}$-valued predicates. This instantiation ensures that
-the resulting induction rule only inspects the top-level structure of
-its argument, rather than the contents of that structure, which is
-exactly what structural induction should do.
+We note that, as for ADTs, the structural induction rule for any
+(truly) nested type can be obtained as the special case of its deep
+induction rule in which the custom predicates are taken to be
+constantly $\mathsf{\top}$-valued predicates. This instantiation
+ensures that the resulting induction rule only inspects the top-level
+structure of its argument, rather than the contents of that structure,
+which is exactly what structural induction should do.\looseness=-1
 %A concrete example of such a derivation is given in
@@
 \[\mathsf{Equal^{\wedge}\,A\,A\,Q\,Q'\,refl = \forall (a :
-  A) \to Equal\,(Q\,a)(Q'\,a)}\] It takes two predicates on the same
+  A) \to Equal\,(Q\,a)(Q'\,a)}\]
+It does exactly what we expect: it takes two predicates on the same
 type as input and is inhabited iff they are extensionally equal.
@@
 $\mathsf{(x,b)}$, where $\mathsf{x : A}$, $\mathsf{b : F\, x}$, and
-$\mathsf{F : A \to Set}$. Next, let $\mathsf{dIndConst}$ and
-$\mathsf{dIndPair}$ be the induction hypotheses associated with the
-constructors $\mathsf{const}$ and $\mathsf{pair}$, respectively. These
-are given in Figure~\ref{fig:seq} as well. Then the deep induction
-rule for $\mathsf{Seq}$ is given in the last two lines of
-Figure~\ref{fig:seq}.
+$\mathsf{F : A \to Set}$.  The lifting $\mathsf{Seq^\wedge}$ is
+derived as in Section~\ref{sec:framework}. Next, let
+$\mathsf{dIndConst}$ and $\mathsf{dIndPair}$ be the induction
+hypotheses associated with the constructors $\mathsf{const}$ and
+$\mathsf{pair}$, respectively. These are given in Figure~\ref{fig:seq}
+as well. Then the deep induction rule for $\mathsf{Seq}$ is given in
+the last two lines of Figure~\ref{fig:seq}.
 
@@
 $\mathsf{dIndSeq}$ inhabiting the type in the last two lines of
-Figure~\ref{fig:seq} by
+Figure~\ref{fig:seq} by\looseness=-1
 \[\begin{array}{ll}
@@
 \end{array}\]
-In the first clause, $\mathsf{a : A}$, $\mathsf{Q_A : A
-  \to Set}$, and $\mathsf{liftA :}$
-$\mathsf{Seq^{\wedge}\,A\,Q_A}$ $\mathsf{(const\,a) = Q_A\,a}$. In the second
-clause we also have $\mathsf{Q_B : B \to}$ $\mathsf{Set}$, $\mathsf{Q_C : C \to
-  Set}$, $\mathsf{e \,:\, Equal\,A\,(B \times C)}$,\, $\mathsf{s_B :}$
-$\mathsf{Seq\,B}$, $\mathsf{s_C : Seq\,C}$,\, $\mathsf{liftE :
-  Equal^{\wedge}\,A\,}$ $\mathsf{(B \times C)\, Q_A\, (Pair^\wedge
-  \,B\,C\,Q_B}$ $\mathsf{Q_C) \, e}$, $\mathsf{liftB :
-  Seq^{\wedge}\,B}$ $\mathsf{Q_B\,s_B}$, and $\mathsf{liftC}$ $\mathsf{ :
-  Seq^{\wedge}\,C\,Q_C\,s_C}$ --- which together give that
-$\mathsf{(Q_B, Q_C, liftE,}$ $\mathsf{liftB, liftC) :
-  Seq^{\wedge}\,A\,Q\,(pair}$ $\mathsf{B\,C\,e\,s_B\,s_C)}$ --- and
-$\mathsf{p_B} = \mathsf{dIndSeq\,P}$
-$\mathsf{cconst\,cpair\,B\,Q_B\,s_B\,liftB :}$
-$\mathsf{P\,B\,Q_B\,s_B}$ and $\mathsf{p_C} =\mathsf{dIndSeq\,P}$
-$\mathsf{cconst\,cpair\,C\,Q_C\,s_C\,liftC : P}$
-$\mathsf{C\,Q_C\,s_C}$.
+In the first clause, $\mathsf{a : A}$, $\mathsf{Q_A : A \to Set}$, and
+$\mathsf{liftA :}$ $\mathsf{Seq^{\wedge}\,A\,Q_A}$ $\mathsf{(const\,a)
+  = Q_A\,a}$. In the second clause we also have
+\[\begin{array}{lll}
+\mathsf{Q_B} & : & \mathsf{B \to Set}\\
+\mathsf{Q_C} & : & \mathsf{C \to Set}\\
+\mathsf{e}  & : & \mathsf{Equal\,A\,(B \times C)}\\
+\mathsf{s_B} & : & \mathsf{Seq\,B}\\
+\mathsf{s_C} & : & \mathsf{Seq\,C}\\
+\mathsf{liftE} & : & \mathsf{Equal^{\wedge}\,A\, (B \times C)\, Q_A\,
+  (Pair^\wedge \,B\,C\,Q_B\,Q_C) \, e}\\
+\mathsf{liftB} & : & \mathsf{Seq^{\wedge}\,B\,Q_B\,s_B}\\
+\mathsf{liftC} & : & \mathsf{Seq^{\wedge}\,C\,Q_C\,s_C}
+\end{array}\]
+%$\mathsf{Q_B : B \to}$ $\mathsf{Set}$, $\mathsf{Q_C : C \to Set}$,
+%$\mathsf{e \,:\, Equal\,A\,(B \times C)}$, \, $\mathsf{s_B :}$
+%$\mathsf{Seq\,B}$, $\mathsf{s_C : Seq\,C}$,\, $\mathsf{liftE :
+%Equal^{\wedge}\,A\,}$ $\mathsf{(B \times C)\, Q_A\, (Pair^\wedge
+%\,B\,C\,Q_B}$ $\mathsf{Q_C) \, e}$, $\mathsf{liftB :
+%Seq^{\wedge}\,B}$ $\mathsf{Q_B\,s_B}$, and $\mathsf{liftC}$ $\mathsf{
+%: Seq^{\wedge}\,C\,Q_C\,s_C}$
+Together these give that
+\[\mathsf{(Q_B, Q_C, liftE,liftB, liftC)
+\, :\,  Seq^{\wedge}\,A\,Q\,(pair\,B\,C\,e\,s_B\,s_C)}\]
+We therefore have
+\[\begin{array}{lllll}
+\mathsf{p_B}\!\! & = & \!\!\mathsf{dIndSeq\,P
+  \,cconst\,cpair\,B\,Q_B\,s_B\,liftB} \!\! & : & 
+\!\!\mathsf{P\,B\,Q_B\,s_B}\\
+\mathsf{p_C}\!\! & = & \!\!\mathsf{dIndSeq\,P\,
+  cconst\,cpair\,C\,Q_C\,s_C\,liftC}\!\! & : &\!\! \mathsf{P\,C\,Q_C\,s_C} 
+\end{array}\]
 
@@
 \mathsf{\hspace*{2.2in}(A \to B) \to Set}
-\end{array}\] for arrow types, since arrow types
-appear in $\mathsf{LType}$ and $\mathsf{LTerm}$.  It is given by
-$\mathsf{Arr^{\wedge}\, A\, B\, Q_A\, Q_B\, f = \forall (a : A) \to
-  Q_A\,a \to}$ $\mathsf{Q_B\, (f\,a)}$. The predicate liftings
+\end{array}\] for arrow types following the general framework in
+Section~\ref{sec:framework}, since 
+arrow types appear in $\mathsf{LType}$ and $\mathsf{LTerm}$.  It is
+given by \[\mathsf{Arr^{\wedge}\, A\, B\, Q_A\, Q_B\, f = \forall (a :
+  A) \to Q_A\,a \to Q_B\, (f\,a)}\] The predicate liftings
 \[\mathsf{LType^{\wedge} : \forall (A : Set) \to (A \to Set) \to
@@
   : \forall (A : Set) \to (A \to Set) \to LTerm\,A \to Set}\] for
-$\mathsf{LTerm}$ are defined in Figure~\ref{fig:liftings}.  There,
-$\mathsf{s : String}$, $\mathsf{Q_A : A \to}$ $\mathsf{Set}$,
-$\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C : C \to Set}$,
-$\mathsf{K^{Bool}_{\top}}$ is the constantly $\mathsf{\top}$-valued
-predicate on $\mathsf{Bool}$, $\mathsf{T_A : LType\, A}$, $\mathsf{T_B
-  : LType \,B}$, $\mathsf{T_C :}$ $\mathsf{LType \,C}$, $\mathsf{t_B :
-  LTerm \, B}$, $\mathsf{t_C : LTerm \, C}$, and $\mathsf{t_{BA} :
-  LTerm \, (B \to A)}$.  Also, $\mathsf{e : Equal\,A\,Bool}$ in the
-first clause, $\mathsf{e : Equal\, A\, (B \to C)}$ in the second and
-fifth, $\mathsf{e : Equal\, A\, (List\, B)}$ in the third,
+$\mathsf{LTerm}$ are defined in Figure~\ref{fig:liftings} following
+the general framework in Section~\ref{sec:framework}.  There,
+\[
+\begin{array}{lll}
+\mathsf{s} & : & \mathsf{String} \\
+\mathsf{Q_A} & : & \mathsf{A \to Set} \\
+\mathsf{Q_B} & : & \mathsf{B \to Set} \\
+\mathsf{Q_C} & : & \mathsf{C \to Set} \\
+\mathsf{T_A} & : & \mathsf{LType\, A} \\
+\mathsf{T_B} & : & \mathsf{LType\, B} \\
+\mathsf{T_C} & : & \mathsf{LType\, C} \\
+\mathsf{t_B} & : & \mathsf{LTerm \, B} \\
+\mathsf{t_C} & : & \mathsf{LTerm \, C} \\
+\mathsf{t_{BA}} & : & \mathsf{LTerm \, (B \to A)} \\
+\mathsf{ts} & : & \mathsf{List\, (LTerm\, B)}
+\end{array}
+\]
+and $\mathsf{K^{Bool}_{\top}}$ is the constantly
+$\mathsf{\top}$-valued predicate on $\mathsf{Bool}$ and
+$\mathsf{List^\wedge}$ is the predicate lifting for lists
+from~\eqref{eq:rose}.  Also,
+\[
+\begin{array}{llll}
+\mathsf{e} & : & \mathsf{Equal\,A\,Bool} & \text{in the first clause,} \\
+\mathsf{e} & : & \mathsf{Equal\, A\, (B \to C)} & \text{in the second and fifth clauses,} \\
+\mathsf{e} & : & \mathsf{Equal\, A\, (List\, B)} & \text{in the third clause,} \\
+\mathsf{e} & : & \text{Equal\, A\, (List \,B)} & \text{in the seventh clause.}
+\end{array}
+\]
+%$\mathsf{s : String}$, $\mathsf{Q_A : A \to}$ $\mathsf{Set}$,
+%$\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C : C \to Set}$,
+%$\mathsf{K^{Bool}_{\top}}$ is the constantly $\mathsf{\top}$-valued
+%predicate on $\mathsf{Bool}$, $\mathsf{T_A : LType\, A}$, $\mathsf{T_B
+ % : LType \,B}$, $\mathsf{T_C :}$ $\mathsf{LType \,C}$, $\mathsf{t_B :
+ % LTerm \, B}$, $\mathsf{t_C : LTerm \, C}$, and $\mathsf{t_{BA} :
+% LTerm \, (B \to A)}$.  Also, $\mathsf{e : Equal\,A\,Bool}$ in the
+%first clause, $\mathsf{e : Equal\, A\, (B \to C)}$ in the second and
+%fifth, $\mathsf{e : Equal\, A\, (List\, B)}$ in the third,
 %$\mathsf{e : Equal \, A \, (B \to C)}$ in the fifth,
-and $\mathsf{e : Equal\, A\, (List \,B)}$,
-$\mathsf{ts : List\, (LTerm B)}$, and $\mathsf{List^\wedge}$ is the
-predicate lifting for lists from~\eqref{eq:rose} in the seventh.
+%and $\mathsf{e : Equal\, A\, (List \,B)}$,
+%$\mathsf{ts : List\, (LTerm B)}$, and $\mathsf{List^\wedge}$ is the
+%predicate lifting for lists from~\eqref{eq:rose} in the seventh.
 
@@
 \end{equation}
-
-\begin{figure*}[t]
-
-    \begin{adjustbox}{varwidth=7.3in, max width=7.1in, margin=-0.0in 0in
-      -0.2in 0in, fbox, center}
-
-%  \begin{adjustbox}{varwidth=6.8in, max width=6.6in, margin=-0.1in 0in
-%      -0.2in 0in, fbox, center} 
-{\small
-\[\begin{array}{lll}
-\mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clist \, A\, Q_A\,
-  (var\;s\,T_A) \, liftA} & = & \mathsf{cvar \, A\, Q_A\, s\, T_A\,
-  liftA}\\ 
-\mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clist \, A\, Q_A\,
-  (abs \,B \,C \,e \,s \,T_B \, t_C) \, (Q_B , Q_C , liftE,
-  lift_{T_B}, lift_{t_C})} & = & \mathsf{cabs\,A\,B\,C\, Q_A\,
-  Q_B\, Q_C\, e\, s\, T_B\, t_C\, liftE\, lift_{T_B}\, p_C}\\
-\mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clist \, A\, Q_A\,
-    (app \,B \,\,t_{BA} \, t_B)\, (Q_B , lift_{t_{BA}}, lift_{t_B})} &
-= & \mathsf{capp\,A\,B\,Q_A\, Q_B\, t_{BA}\, t_B\, p_{BA} \, p_B}\\
-  \mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clist \, A\, Q_A\,
-    (list \,B \,e \, ts) \, (Q_B , liftE', lift_{List})} & = & 
-  \mathsf{clist \,A\,B\,Q_A\, Q_B\, e\, ts\, liftE'\, p_{List} }
-\end{array}\]}
-\mbox{where}
-{\small
-\[\begin{array}{lll}
-\mathsf{p_C} & = & \mathsf{dIndLTerm\,P\,cvar\,cabs \,capp \,clist\,
-  C\, Q_C\, t_C\, lift_{t_C} : P \, C\, Q_C \, t_C }\\
-\mathsf{p_B} & = & \mathsf{dIndLTerm\,P\,cvar\,cabs \,capp \,clist\,
-  B\, Q_B\, t_B\, lift_{t_B} : P \, B\, Q_B \, t_B }\\
-\mathsf{p_{BA}} & = & \mathsf{dIndLTerm\,P\,cvar\,cabs \,capp
-  \,clist\, (B \to A)\,(Arr^{\wedge} \, B\, A\, Q_B \, Q_A) \,
-  t_{BA}\, lift_{t_{BA}} : P \, (B \to A)\, (Arr^{\wedge} \, B\, A\,
-  Q_B \, Q_A) \, t_{BA}}\\ 
-\mathsf{p_{List}} & = &\mathsf{liftListMap \, (LTerm\, B) \,
-  (LTerm^{\wedge} \, B \, Q_B)\, (P\,B\,Q_B)\, p_{ts} \, ts\,
-  lift_{List} : List^{\wedge}\, (LTerm\,B) \, (P\,B\,Q_B) \, ts}\\
-\mathsf{p_{ts}} & = & \mathsf{dIndLTerm\, P\, cvar\, cabs\, capp\,
-  clist\, B\, Q_B : PredMap\,(LTerm\,B) \,(LTerm^{\wedge}\, B\, Q_B)
-  \, (P\,B\,Q_B)}
-\end{array}\]}
-
-\vspace*{-0.1in}
-
-\caption{$\mathsf{dIndLTerm}$}\label{fig:dindlterm} \vspace*{0.1in} 
-\end{adjustbox}
-\end{figure*}
 
@@
 
+\begin{figure*}[t]
+
+    \begin{adjustbox}{varwidth=7.3in, max width=7.1in, margin=-0.0in 0in
+      -0.2in 0in, fbox, center}
+
+%  \begin{adjustbox}{varwidth=6.8in, max width=6.6in, margin=-0.1in 0in
+%      -0.2in 0in, fbox, center} 
+{\small
+\[\begin{array}{lll}
+\mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clist \, A\, Q_A\,
+  (var\;s\,T_A) \, liftA} & = & \mathsf{cvar \, A\, Q_A\, s\, T_A\,
+  liftA}\\ 
+\mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clist \, A\, Q_A\,
+  (abs \,B \,C \,e \,s \,T_B \, t_C) \, (Q_B , Q_C , liftE,
+  lift_{T_B}, lift_{t_C})} & = & \mathsf{cabs\,A\,B\,C\, Q_A\,
+  Q_B\, Q_C\, e\, s\, T_B\, t_C\, liftE\, lift_{T_B}\, p_C}\\
+\mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clist \, A\, Q_A\,
+    (app \,B \,\,t_{BA} \, t_B)\, (Q_B , lift_{t_{BA}}, lift_{t_B})} &
+= & \mathsf{capp\,A\,B\,Q_A\, Q_B\, t_{BA}\, t_B\, p_{BA} \, p_B}\\
+  \mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clist \, A\, Q_A\,
+    (list \,B \,e \, ts) \, (Q_B , liftE', lift_{List})} & = & 
+  \mathsf{clist \,A\,B\,Q_A\, Q_B\, e\, ts\, liftE'\, p_{List} }
+\end{array}\]}
+\mbox{where}
+{\small
+\[\begin{array}{lll}
+\mathsf{p_C} & = & \mathsf{dIndLTerm\,P\,cvar\,cabs \,capp \,clist\,
+  C\, Q_C\, t_C\, lift_{t_C} : P \, C\, Q_C \, t_C }\\
+\mathsf{p_B} & = & \mathsf{dIndLTerm\,P\,cvar\,cabs \,capp \,clist\,
+  B\, Q_B\, t_B\, lift_{t_B} : P \, B\, Q_B \, t_B }\\
+\mathsf{p_{BA}} & = & \mathsf{dIndLTerm\,P\,cvar\,cabs \,capp
+  \,clist\, (B \to A)\,(Arr^{\wedge} \, B\, A\, Q_B \, Q_A) \,
+  t_{BA}\, lift_{t_{BA}} : P \, (B \to A)\, (Arr^{\wedge} \, B\, A\,
+  Q_B \, Q_A) \, t_{BA}}\\ 
+\mathsf{p_{List}} & = &\mathsf{liftListMap \, (LTerm\, B) \,
+  (LTerm^{\wedge} \, B \, Q_B)\, (P\,B\,Q_B)\, p_{ts} \, ts\,
+  lift_{List} : List^{\wedge}\, (LTerm\,B) \, (P\,B\,Q_B) \, ts}\\
+\mathsf{p_{ts}} & = & \mathsf{dIndLTerm\, P\, cvar\, cabs\, capp\,
+  clist\, B\, Q_B : PredMap\,(LTerm\,B) \,(LTerm^{\wedge}\, B\, Q_B)
+  \, (P\,B\,Q_B)}
+\end{array}\]}
+
+\vspace*{-0.1in}
+
+\caption{$\mathsf{dIndLTerm}$}\label{fig:dindlterm} \vspace*{0.1in} 
+\end{adjustbox}
+\end{figure*}
+
 To prove that this rule is sound we define a witness
 $\mathsf{dIndLTerm}$ inhabiting the type in~\eqref{eq:ind-lam} as in
-Figure~\ref{fig:dindlterm}. There, $\mathsf{s : String}$, $\mathsf{Q_A
-  : A \to Set}$, $\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C : C \to
-  Set}$, $\mathsf{T_A : LType\,A}$, $\mathsf{T_B}$ $\mathsf{:
-  LType\,B}$, $\mathsf{t_B : LTerm\,B}$, $\mathsf{t_C : LTerm\,C}$,
-$\mathsf{t_{BA} : LTerm\,(B \to A)}$, $\mathsf{liftA :
-  LTerm^{\wedge}\, A\, Q_A\, (var\;s\,T_A) =
-  LType^{\wedge}\,A\,Q_A\,T_A}$, $\mathsf{liftE :}$
-$\mathsf{Equal^{\wedge}}$ $\mathsf{A\, (B \to C)\, Q_A\, (Arr^{\wedge}
-  \, B\, C\, Q_B \, Q_C) \, e}$, $\mathsf{lift_{T_B}: LType^{\wedge}
-  \, B\, Q_B\, T_B}$, $\mathsf{lift_{t_C}}$ $\mathsf{: LTerm^{\wedge}
-  \, C\, Q_C\, T_C}$, $\mathsf{lift_{t_{BA}}: LTerm^{\wedge} \, (B \to
-  A)\, (Arr^{\wedge} \, B\, A\, Q_B \, Q_A)}$ $\mathsf{t_{BA}}$,
-$\mathsf{lift_{t_B}: LTerm^{\wedge} \, B\, Q_B\, t_B}$,
-$\mathsf{liftE' : Equal^{\wedge}\, A\, (List\,B)\, Q_A}$ and
-$\mathsf{(List^{\wedge}\, B\, Q_B)\, e}$, and $\mathsf{lift_{List}:
-  List^{\wedge} \, (LTerm\, B) \, (LTerm^{\wedge}\, B\, Q_B) \, ts}$.
-Moreover, in the definition of $\mathsf{p_{ts}}$, $\mathsf{PredMap :
-  \forall\, (A : Set) \to}$ $\mathsf{(A \to Set)}$ $\mathsf{\to (A \to
-  Set) \to Set }$ is the type constructor producing the type of
+Figure~\ref{fig:dindlterm}. There,
+
+\[
+\begin{array}{lll}
+\mathsf{s} & : & \mathsf{String} \\
+\mathsf{Q_A} & : & \mathsf{A \to Set} \\
+\mathsf{Q_B} & : & \mathsf{B \to Set} \\
+\mathsf{Q_C} & : & \mathsf{C \to Set} \\
+\mathsf{T_A} & : & \mathsf{LType\, A} \\
+\mathsf{T_B} & : & \mathsf{LType\, B} \\
+\mathsf{t_B} & : & \mathsf{LTerm \, B} \\
+\mathsf{t_C} & : & \mathsf{LTerm \, C} \\
+\mathsf{t_{BA}} & : & \mathsf{LTerm \, (B \to A)} \\
+\mathsf{ts} & : & \mathsf{List\, (LTerm\, B)}\\
+\mathsf{liftA} & : & \mathsf{LTerm^{\wedge}\, A\, Q_A\, (var\;s\,T_A) = LType^{\wedge}\,A\,Q_A\,T_A} \\
+\mathsf{liftE} & : & \mathsf{Equal^{\wedge}\, A\, (B \to C)\, Q_A\, (Arr^{\wedge} \, B\, C\, Q_B \, Q_C) \, e} \\
+\mathsf{lift_{T_B}} & : & \mathsf{LType^{\wedge} \, B\, Q_B\, T_B} \\
+\mathsf{lift_{t_C}} & : & \mathsf{LTerm^{\wedge} \, C\, Q_C\, t_C} \\
+\mathsf{lift_{t_{BA}}} & : & \mathsf{LTerm^{\wedge} \, (B \to A)\, (Arr^{\wedge} \, B\, A\, Q_B \, Q_A)\, t_{BA}} \\
+\mathsf{lift_{t_B}} & : & \mathsf{LTerm^{\wedge} \, B\, Q_B\, t_B} \\
+\mathsf{liftE'} & : & \mathsf{Equal^{\wedge}\, A\, (List\,B)\, Q_A\, (List^{\wedge}\, B\, Q_B)\, e} \\
+\mathsf{lift_{List}} & : & \mathsf{List^{\wedge} \, (LTerm\, B) \,
+  (LTerm^{\wedge}\, B\, Q_B) \, ts}
+\end{array}
+\]
+%$\mathsf{s : String}$, $\mathsf{Q_A
+%  : A \to Set}$, $\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C : C \to
+%  Set}$, $\mathsf{T_A : LType\,A}$, $\mathsf{T_B}$ $\mathsf{:
+%  LType\,B}$, $\mathsf{t_B : LTerm\,B}$, $\mathsf{t_C : LTerm\,C}$,
+%$\mathsf{t_{BA} : LTerm\,(B \to A)}$, $\mathsf{liftA :
+%  LTerm^{\wedge}\, A\, Q_A\, (var\;s\,T_A) =
+%  LType^{\wedge}\,A\,Q_A\,T_A}$, $\mathsf{liftE :}$
+%$\mathsf{Equal^{\wedge}}$ $\mathsf{A\, (B \to C)\, Q_A\, (Arr^{\wedge}
+%  \, B\, C\, Q_B \, Q_C) \, e}$, $\mathsf{lift_{T_B}: LType^{\wedge}
+%  \, B\, Q_B\, T_B}$, $\mathsf{lift_{t_C}}$ $\mathsf{: LTerm^{\wedge}
+%  \, C\, Q_C\, T_C}$, $\mathsf{lift_{t_{BA}}: LTerm^{\wedge} \, (B \to
+%  A)\, (Arr^{\wedge} \, B\, A\, Q_B \, Q_A)}$ $\mathsf{t_{BA}}$,
+%$\mathsf{lift_{t_B}: LTerm^{\wedge} \, B\, Q_B\, t_B}$,
+%$\mathsf{liftE' : Equal^{\wedge}\, A\, (List\,B)\, Q_A}$ and
+%$\mathsf{(List^{\wedge}\, B\, Q_B)\, e}$, and $\mathsf{lift_{List}:
+%  List^{\wedge} \, (LTerm\, B) \, (LTerm^{\wedge}\, B\, Q_B) \, ts}$.
+Moreover, in the definition of $\mathsf{p_{ts}}$, \[\mathsf{PredMap :
+  \forall\, (A : Set) \to (A \to Set) \to (A \to
+  Set) \to Set }\] is the type constructor producing the type of
 morphisms between predicates defined by\looseness=-1
@@
   A\, Q\, nil}$ must necessarily be the sole inhabitant $\mathsf{tt}$
-of $\mathsf{\top}$), and by $\mathsf{liftListMap\, A\, Q\, Q'\, m\,
-  (cons\, a\, l')\, (y, x') =}$\\ $\mathsf{ (m\,a\,y, \,liftListMap\, A\,
-  Q\, Q'\, m\, l'\, x')}$ (since $\mathsf{x : List^{\wedge}\, A\,
+of $\mathsf{\top}$), and by
+\[\begin{array}{ll}
+ & \mathsf{liftListMap\, A\, Q\, Q'\, m\,
+  (cons\, a\, l')\, (y, x')}\\
+= & \mathsf{(m\,a\,y, \,liftListMap\, A\,
+  Q\, Q'\, m\, l'\, x')}
+\end{array}\] (since $\mathsf{x : List^{\wedge}\, A\,
   Q\, (cons\, a\, l')}$ must be of the form $\mathsf{x = (y, x')}$
@@
 A concrete way to define more general predicate liftings is, again, by
-induction on the structure of the types in a suitable calculus. This
+induction on the structure of the types in a suitable calculus; this
 will ensure that the liftings satisfy the crucial property needed to
@@
 \end{array}\]
-Here, $\mathsf{cc : dIndC\,P}$, $\mathsf{e : Equal\,A\,(K\,\ol{B})}$,
-$\mathsf{x : F\,G\,\ol{B}}$, $\mathsf{Q_A : A \to Set}$,
-$\ol{\mathsf{Q_B : B \to Set}}$, $\mathsf{liftE :
-  Equal^{\wedge}\,A\,(K\,\ol{B})\,Q_A\,
-  (K^{\wedge}\,\ol{B}\,\ol{Q_B})\,e}$, and $\mathsf{liftF}$ $\mathsf{:
-  F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,{\ol{Q_B}}\,x}$, so
-$\mathsf{(\ol{Q_B}, liftE, liftF) : G^{\wedge}\,A\,Q_A (c\,\ol{B}\,
-  e\,x)}$ as expected.  Finally, the morphism of predicates
-\[\mathsf{p : PredMap\,(F\,G\,\ol{B}) (F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B})
+Here,
+\[\begin{array}{lll}
+\mathsf{cc} & : & \mathsf{dIndC\,P}\\
+\mathsf{e} & : & \mathsf{Equal\,A\,(K\,\ol{B})}\\
+\mathsf{x} & : & \mathsf{F\,G\,\ol{B}}\\
+\mathsf{Q_A} & : & \mathsf{A \to Set}\\
+\mathsf{liftE} & : & \mathsf{Equal^{\wedge}\,A\,(K\,\ol{B})\,Q_A\,
+  (K^{\wedge}\,\ol{B}\,\ol{Q_B})\,e}\\
+\mathsf{liftF} & : &
+\mathsf{F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,{\ol{Q_B}}\,x}
+\end{array}\]
+and $\ol{\mathsf{Q_B : B \to Set}}$,
+so
+\[\mathsf{(\ol{Q_B}, liftE, liftF) : G^{\wedge}\,A\,Q_A (c\,\ol{B}\,
+  e\,x)}\] as expected.  Finally, the morphism of predicates
+\[\mathsf{p : PredMap\,(F\,G\,\ol{B})
+  (F^{\wedge}\,G\,\ol{B}\,G^{\wedge}\,\ol{Q_B}) 
   (F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B})}\] is defined by structural
@@
 
-\begin{equation*}\label{gadt-nested}
+\begin{equation}\label{gadt-nested}
 \begin{array}{l}
@@
 \end{array}
-\end{equation*}
+\end{equation}
 We acknowledge that $\mathsf{G}$ is semantically equivalent to the
@@
 inhabiting this type we run into problems. We can define
-$\mathsf{dIndG\,P\,cc\,A\,Q\,(c\,B\,e\,x)\,(Q', liftE, liftG)}$
-$\mathsf{= cc\,A\,B\,Q\,Q'\,e\,x\,liftE\,p}$, where $\mathsf{cc :
-  dIndC\,P}$, $\mathsf{Q : A \to Set}$,\\ $\mathsf{Q' : B \to Set}$,
-$\mathsf{e : Equal\,A\,(B \times B)}$, $\mathsf{x : G\,(G\,B)}$,
-$\mathsf{liftG : G^{\wedge}\,(G\,B)}$ $\mathsf{(G^{\wedge}
-  B\,Q')\,x}$, and $\mathsf{liftE : Equal^{\wedge}\,A\,(B \times
-  B)\,Q\,(Pair^{\wedge}\,B\,B\,Q'\,Q')\,e}$, but we still need to
+\[\begin{array}{ll}
+  & \mathsf{dIndG\,P\,cc\,A\,Q\,(c\,B\,e\,x)\,(Q', liftE, liftG)}\\
+= & \mathsf{cc\,A\,B\,Q\,Q'\,e\,x\,liftE\,p}
+\end{array}\]
+where
+\[\begin{array}{lll}
+\mathsf{cc} & : & \mathsf{dIndC\,P}\\
+\mathsf{Q} & : & \mathsf{A \to Set}\\
+\mathsf{Q'} & : & \mathsf{B \to Set}\\
+\mathsf{e} & : & \mathsf{Equal\,A\,(B \times B)}\\
+\mathsf{x} & : & \mathsf{G\,(G\,B)}\\
+\mathsf{liftG} & : & \mathsf{G^{\wedge}\,(G\,B)\, (G^{\wedge}
+  B\,Q')\,x}\\
+\mathsf{liftE} & : & \mathsf{Equal^{\wedge}\,A\,(B \times 
+  B)\,Q\,(Pair^{\wedge}\,B\,B\,Q'\,Q')\,e}
+\end{array}\]
+but we still need to
 define $\mathsf{p : P\,(G\,B)\,(P\,B\,Q')\,x}$.  For this we can use
@@
 \quad\quad= & \!\! \mathsf{(Q'_B, liftE', liftG')}
-\end{array}\] for some $\mathsf{Q'_B : B
-  \to Set}$, $\mathsf{liftE' : Equal^{\wedge}\,A\,(B \times
-  B)\,Q'\,(Pair^{\wedge}\,B}$ $\mathsf{B\,Q'_B\,Q'_B)\,e}$, and
-$\mathsf{liftG' : G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q'_B)\,x}$,
-where $\mathsf{Q : A \to}$ $\mathsf{Set}$, $\mathsf{Q' : A \to Set}$,
-$\mathsf{Q_B : B \to Set}$, $\mathsf{m : PredMap\,A\,Q\,Q'\,e :
-}$\\ $\mathsf{Equal}$ $\mathsf{A\,(B \times B)}$, $\mathsf{x :
-  G\,(G\,B)}$, $\mathsf{liftE : Equal^{\wedge}\,A\,(B \times
-  B)\,Q\,(Pair^{\wedge}}$ $\mathsf{B\,B\,Q_B\,Q_B)\,e}$, and
-$\mathsf{liftG : G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q_B)\,x}$. 
+\end{array}\] for some
+\[\begin{array}{lll}
+\mathsf{Q'_B} & : & \mathsf{B \to Set}\\
+\mathsf{liftE'} & :  & \mathsf{Equal^{\wedge}\,A\,(B \times
+  B)\,Q'\,(Pair^{\wedge}\,B\,B\,Q'_B\,Q'_B)\,e}\\
+\mathsf{liftG'} & : &
+\mathsf{G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q'_B)\,x}
+\end{array}\]
+where
+\[\begin{array}{lll}
+\mathsf{Q} & : & \mathsf{A \to Set}\\
+\mathsf{Q'} & : & \mathsf{A \to Set}\\
+\mathsf{Q_B} & : & \mathsf{B \to Set}\\
+\mathsf{m} & : & \mathsf{PredMap\,A\,Q\,Q'}\\
+\mathsf{e} & : & \mathsf{Equal\,A\,(B \times B)}\\
+\mathsf{x} & : & \mathsf{G\,(G\,B)}\\
+\mathsf{liftE} & : & \mathsf{Equal^{\wedge}\,A\,(B \times
+  B)\,Q\,(Pair^{\wedge}\,B\,B\,Q_B\,Q_B)\,e}\\
+\mathsf{liftG} & : &
+\mathsf{G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q_B)\,x}
+\end{array}\]
 That is, we would need to produce a proof $\mathsf{liftE'}$ of the
@@
 $\mathsf{ts}$ (from which we can deduce the type of the list). When
-$\mathsf{ts} = \mathsf{nil}$ we define $\mathsf{clist\, A\, B\, Q\,
-  Q'\, e\, nil \, liftE\, lift_{ts}}$ $\mathsf{= nothing}$, where
+$\mathsf{ts} = \mathsf{nil}$ we define \[\mathsf{clist\, A\, B\, Q\,
+  Q'\, e\, nil \, liftE\, lift_{ts} = nothing}\] where
 $\mathsf{liftE : Equal^{\wedge}\, A\, (List\,B) \, Q\,
-  (List^{\wedge}\, B\, Q')\, e}$, and $\mathsf{lift_{ts} :
-  List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, ts}$.  When
+  (List^{\wedge}\, B\, Q')\, e}$, and $\mathsf{lift_{ts} :}$\\
+$\mathsf{List^{\wedge}\, (LTerm\, B)\, (GetType\, B)\, ts}$.  When
 $\mathsf{ts} = \mathsf{cons\,t\,ts'}$

### 2021-12-03 20-gadts/21-LSFA/GADT-story.tex
final LSFA paper for uploading

@@
 independence~\cite{adr09,dnb12}, equivalences between
-programs~\cite{hd11}, deep induction principles~\cite{jgj21s,jp20},
+programs~\cite{hd11}, deep induction principles~\cite{jg22,jp20},
 and useful (``free'') theorems about programs derived from their types

### 2021-12-08 induction-paper/cpp2022/cpp2022-paper-final.tex
final cpp and final lfsa

@@
-\documentclass[sigplan,10pt]{acmart}
-\settopmatter{printfolios=true,printccs=false,printacmref=false}
-
+\documentclass[sigplan,screen]{acmart}
+%\settopmatter{printfolios=true,printccs=false,printacmref=false}
+
+\usepackage{microtype}
 \usepackage{graphicx}
@@
 
+\copyrightyear{2022}
+\acmYear{2022}
+\setcopyright{acmlicensed}\acmConference[CPP '22]{Proceedings of the 11th
+ACM SIGPLAN International Conference on Certified Programs and
+Proofs}{January 17--18, 2022}{Philadelphia, PA, USA}
+\acmBooktitle{Proceedings of the 11th ACM SIGPLAN International Conference
+on Certified Programs and Proofs (CPP '22), January 17--18, 2022,
+Philadelphia, PA, USA}
+\acmPrice{15.00}
+\acmDOI{10.1145/3497775.3503680}
+\acmISBN{978-1-4503-9182-5/22/01}
+
 \begin{document}
@@
 
+\begin{CCSXML}
+<ccs2012>
+<concept>
+<concept_id>10003752</concept_id>
+<concept_desc>Theory of computation</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+<concept>
+<concept_id>10003752.10010124</concept_id>
+<concept_desc>Theory of computation~Semantics and reasoning</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+<concept>
+<concept_id>10003752.10010124.10010131.10010137</concept_id>
+<concept_desc>Theory of computation~Categorical semantics</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+</ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Theory of computation}
+\ccsdesc[500]{Theory of computation~Semantics and reasoning}
+\ccsdesc[500]{Theory of computation~Categorical semantics}
+
+%%
+%% Keywords. The author(s) should pick words that accurately describe
+%% the work being presented. Separate the keywords with commas.
+\keywords{GADTs, induction, proof assistants}
+
 \maketitle
-
-\vspace*{-0.1in}
 
@@
 
-%\vspace*{0.05in}
+\vspace*{0.05in}
 
@@
 
-\section{Deep induction for ADTs and nested types}\label{sec:ADTs-and-nesteds}
+\section{Deep Induction for ADTs and Nested Types}\label{sec:ADTs-and-nesteds}
 
@@
 \begin{adjustbox}{varwidth=7.3in, max width=7.1in, margin=-0.0in 0in
-      -0.2in 0in, fbox, center}
+      -0.2in 0in, center}
 
@@
 
-\section{(Deep) induction for GADTs}\label{sec:deep-ind-GADTs}
+\section{(Deep) Induction for GADTs}\label{sec:deep-ind-GADTs}
 
@@
 
-\subsection{(Deep) induction for $\mathsf{Equal}$}\label{sec:ind-equal}
+\subsection{(Deep) Induction for $\mathsf{Equal}$}\label{sec:ind-equal}
 
@@
 
-\subsection{(Deep) induction for $\mathsf{Seq}$}\label{sec:ind-seq}
+\subsection{(Deep) Induction for $\mathsf{Seq}$}\label{sec:ind-seq}
 
@@
 
-\subsection{(Deep) induction for $\mathsf{LTerm}$}\label{sec:ind-lam}
+\subsection{(Deep) Induction for $\mathsf{LTerm}$}\label{sec:ind-lam}
 
@@
   \begin{adjustbox}{varwidth=7.3in, max width=7.1in, margin=-0.0in 0in
-      -0.2in 0in, fbox, center}
+      -0.2in 0in, center}
 
@@
     \begin{adjustbox}{varwidth=7.3in, max width=7.1in, margin=-0.0in 0in
-      -0.2in 0in, fbox, center}
+      -0.2in 0in, center}
 
@@
 \mathsf{lift_{t_C}} & : & \mathsf{LTerm^{\wedge} \, C\, Q_C\, t_C} \\
-\mathsf{lift_{t_{BA}}} & : & \mathsf{LTerm^{\wedge} \, (B \to A)\, (Arr^{\wedge} \, B\, A\, Q_B \, Q_A)\, t_{BA}} \\
+\mathsf{lift_{t_{BA}}} & : & \mathsf{LTerm^{\wedge} \, (B \to A)\, (Arr^{\wedge} \, B\, A\, Q_B \, Q_A)\, t_{BA}}
+\end{array}\]
+
+\[\begin{array}{lll}
 \mathsf{lift_{t_B}} & : & \mathsf{LTerm^{\wedge} \, B\, Q_B\, t_B} \\
@@
 
-\section{The general framework}\label{sec:framework}
+\section{The General Framework}\label{sec:framework}
 
@@
   We define the second component of the pair to be
-  $\mathsf{LTerm^\wedge KT\,B\,t_2}$. We can define the first
+  $\mathsf{LTerm^\wedge KT\,B\,t_2}$. We define the first
   component from a proof of $\mathsf{LTerm^{\wedge}\, (B \to A)\,
-    K_\top\, t_1}$ and the function
+    K_\top\, t_1}$ and the function\looseness=-1
   \[\begin{array}{ll}
@@
   (extensionally) equal predicates with the same carrier and produces
-  a morphism of predicates between their liftings. The definition of
-  $\mathsf{LTerm^\wedge EqualMap}$ is straightforwardly given by
+  a morphism of predicates between their liftings. We define
+  $\mathsf{LTerm^\wedge EqualMap}$ straightforwardly by
   pattern matching on the first two arguments to $\mathsf{PredMap}$ in
   its return type, using transitivity and symmetry of the type
-  constructor $\mathsf{Equal}$, together with the analogously defined
+  constructor $\mathsf{Equal}$, together with the two analogously defined
   functions $\mathsf{LType^\wedge EqualMap}$ and $\mathsf{Arr^\wedge
@@
 we define
+
+\pagebreak
+
+\vspace*{-0.35in}
+
 \[\begin{array}{ll}
@@
 
+\subsection*{Acknowledgments}
+
+This work was supported by National Science Foundation award 1906388.
+
 \bibliographystyle{ACM-Reference-Format}

### 2021-12-14 induction-paper/cpp2022/cpp2022-paper-final.tex
minor tweaks to final version

@@
 
-\copyrightyear{2022}
-\acmYear{2022}
-\setcopyright{acmlicensed}\acmConference[CPP '22]{Proceedings of the 11th
-ACM SIGPLAN International Conference on Certified Programs and
-Proofs}{January 17--18, 2022}{Philadelphia, PA, USA}
-\acmBooktitle{Proceedings of the 11th ACM SIGPLAN International Conference
-on Certified Programs and Proofs (CPP '22), January 17--18, 2022,
-Philadelphia, PA, USA}
+%%% The following is specific to CPP '22 and the paper
+%%% '(Deep) Induction Rules for GADTs'
+%%% by Patricia Johann and Enrico Ghiorzi.
+%%%
+\setcopyright{acmlicensed}
 \acmPrice{15.00}
 \acmDOI{10.1145/3497775.3503680}
+\acmYear{2022}
+\copyrightyear{2022}
+\acmSubmissionID{poplws22cppmain-p17-p}
 \acmISBN{978-1-4503-9182-5/22/01}
+\acmConference[CPP '22]{Proceedings of the 11th ACM SIGPLAN International Conference on Certified Programs and Proofs}{January 17--18, 2022}{Philadelphia, PA, USA}
+\acmBooktitle{Proceedings of the 11th ACM SIGPLAN International Conference on Certified Programs and Proofs (CPP '22), January 17--18, 2022, Philadelphia, PA, USA}
 
@@
 \begin{table}[h!]
+\caption{Data types in this paper}\label{fig:taxonomy}
 \begin{tabular}{|l|l|l|}
@@
 \end{tabular}
-\caption{Data types in this paper}\label{fig:taxonomy}
 \end{table}
@@
 \begin{figure*}[t]
-
-\begin{adjustbox}{varwidth=7.3in, max width=7.1in, margin=-0.0in 0in
-      -0.2in 0in, center}
-
+\begin{adjustbox}{varwidth=\linewidth, max width=\linewidth, center} 
 {\small
@@
 \end{array}\]}
-
-\vspace*{-0.1in}
-
+\end{adjustbox}
 \caption{Deep induction rules for perfect trees and
   bushes}\label{fig:ptree-and-bush}
-\end{adjustbox}
 \end{figure*}
@@
 \end{array}\]}
-
-\vspace*{-0.1in}
-
+\end{adjustbox}
 \caption{The $\mathsf{LType}$ and $\mathsf{LTerm}$ data
   types}\label{fig:type-and-term}
-\end{adjustbox}
 \end{figure*}
@@
 \end{array}\]
-
-\vspace*{0.1in}
-
 \[\begin{array}{l}
@@
 \end{array}\]}
-
-\vspace*{-0.1in}
-
+\end{adjustbox}
 \caption{Deep induction rule for $\mathsf{Seq}$}\label{fig:seq}
-\end{adjustbox}
 \end{figure*}
@@
 \begin{figure*}[t]
-
-%  \begin{adjustbox}{varwidth=6.8in, max width=6.6in, margin=-0.1in 0in
-%      -0.2in 0in, fbox, center}
-
-  \begin{adjustbox}{varwidth=7.3in, max width=7.1in, margin=-0.0in 0in
-      -0.2in 0in, center}
-
+\begin{adjustbox}{varwidth=\linewidth, max width=\linewidth, center} 
   {\small
@@
 \end{array}   \]}
-
-\vspace*{-0.1in}
-
+\end{adjustbox}
 \caption{Predicate liftings for $\mathsf{LType}$ and
-  $\mathsf{LTerm}$}\label{fig:liftings} \vspace*{0.1in} 
-\end{adjustbox}
+  $\mathsf{LTerm}$}\label{fig:liftings}
 \end{figure*}
@@
 
-\begin{adjustbox}{varwidth=7.3in, max width=7.1in, margin=-0.0in 0in
-      -0.2in 0in, center}
-
+\begin{adjustbox}{varwidth=\linewidth, max width=\linewidth, center} 
 {\small
@@
 \end{array}\]}
-
-\vspace*{-0.1in}
-
+\end{adjustbox}
 \caption{Induction hypotheses for $\mathsf{LTerm}$}\label{fig:ind-hyps-lterm}
-\end{adjustbox}
 \end{figure*}
@@
 \begin{figure*}[t]
-
-    \begin{adjustbox}{varwidth=7.3in, max width=7.1in, margin=-0.0in 0in
-      -0.2in 0in, center}
-
+\begin{adjustbox}{varwidth=\linewidth, max width=\linewidth, center} 
 %  \begin{adjustbox}{varwidth=6.8in, max width=6.6in, margin=-0.1in 0in
@@
 \end{array}\]}
-
-\vspace*{-0.1in}
-
+\end{adjustbox}
 \caption{$\mathsf{dIndLTerm}$}\label{fig:dindlterm} \vspace*{0.1in} 
-\end{adjustbox}
 \end{figure*}
 
-To prove that this rule is sound we define a witness
+To prove this rule sound we define a witness
 $\mathsf{dIndLTerm}$ inhabiting the type in~\eqref{eq:ind-lam} as in
@@
 \mathsf{liftE} & : & \mathsf{Equal^{\wedge}\, A\, (B \to C)\, Q_A\, (Arr^{\wedge} \, B\, C\, Q_B \, Q_C) \, e} \\
-\mathsf{lift_{T_B}} & : & \mathsf{LType^{\wedge} \, B\, Q_B\, T_B} \\
+\mathsf{lift_{T_B}} & : & \mathsf{LType^{\wedge} \, B\, Q_B\, T_B}
+\end{array}\]
+
+\[\begin{array}{lll}
 \mathsf{lift_{t_C}} & : & \mathsf{LTerm^{\wedge} \, C\, Q_C\, t_C} \\
-\mathsf{lift_{t_{BA}}} & : & \mathsf{LTerm^{\wedge} \, (B \to A)\, (Arr^{\wedge} \, B\, A\, Q_B \, Q_A)\, t_{BA}}
-\end{array}\]
-
-\[\begin{array}{lll}
+\mathsf{lift_{t_{BA}}} & : & \mathsf{LTerm^{\wedge} \, (B \to A)\, (Arr^{\wedge} \, B\, A\, Q_B \, Q_A)\, t_{BA}} \\
 \mathsf{lift_{t_B}} & : & \mathsf{LTerm^{\wedge} \, B\, Q_B\, t_B} \\
@@
 for $\mathsf{G^{\wedge}}$
-then we could define
-\[\mathsf{q = G^\wedge Map\,(G\,B)\,(G^{\wedge}\,B\,Q')\,(P\,B\,Q')\,
+then we could define $\mathsf{q}$ as
+\[\mathsf{G^\wedge Map\,(G\,B)\,(G^{\wedge}\,B\,Q')\,(P\,B\,Q')\,
 (dIndG\,P\,cc\,B\,Q')\,x\,liftG}\]  Unfortunately, however,
@@
 \begin{figure*}[t]
-
-\begin{adjustbox}{varwidth=7.3in, max width=7.1in, margin=-0.0in 0in
-      -0.2in 0in, center}
-
+\begin{adjustbox}{varwidth=\linewidth, max width=\linewidth,
+	margin=-0.0in 0in -0.2in 0in, center}
 {\small
@@
 \end{array}\]}
-
-\vspace*{-0.1in}
-
-\caption{Applied induction hypotheses for
-  $\mathsf{LTerm}$}\label{fig:ind-hyps-lterm-applied-to-P}
 \end{adjustbox}
+\caption{Applied induction hypotheses for $\mathsf{LTerm}$}
+\label{fig:ind-hyps-lterm-applied-to-P}
 \end{figure*}
@@
 $\mathsf{P}$ is the polymorphic predicate \[\mathsf{ \lambda \, (A:
-  Set)\, (Q : A \to Set)\, (t : LTerm\,A)\, \to Maybe \, (LType \,
+  Set)\, (Q : A \to Set)\, (t : LTerm\,A) \to Maybe \, (LType \,
   A)}\] and $\mathsf{K_\top}$ is the constantly $\mathsf{\top}$-valued
@@
 %\end{array}\]
-\begin{align*}
-  &\mathsf{clist\, A\, B\, Q\, Q'\, e\, (cons\, t\, ts') \, liftE\,
+\begin{gather*}
+  \mathsf{clist\, A\, B\, Q\, Q'\, e\, (cons\, t\, ts') \, liftE\,
     (nothing , lift_{ts'}) = nothing} \\ 
-  &\mathsf{clist\, A\, B\, Q\, Q'\, e\, (cons\, t\, ts') \, liftE\,
-    (just\, T' , lift_{ts'}) = just \, (list\, B\, e\, T')} 
-\end{align*}
+  \mathsf{clist\;\! A\;\! B\;\! Q\;\! Q'\:\! e\;\! (cons\;\! t\;\! ts')\;\! liftE\;\!
+    (just\;\! T'\!, lift_{ts'}\!)\! = just (list\;\! B\;\! e\;\! T')} 
+\end{gather*}
 Here $\mathsf{e : Equal\, A\, (List\,B)}$, $\mathsf{T' : LType\, B}$,
@@
   \,t')}\\
-\quad\quad = & \!\!\mathsf{\exists [Q_B]\, [Q_C]\, Equal^{\wedge} \, A\, (B \to
-  C)\, K_\top\, (Arr^{\wedge} \, B\, C\, Q_B \, Q_C)\, e }\\
+\quad\quad\! = & \!\!\!\mathsf{\exists [Q_B] [Q_C]\;\! Equal^{\wedge}\;\! A\;\! (B\! \to\!
+  C)\;\! K_\top (Arr^{\wedge}\;\! B\;\! C\;\! Q_B\;\! Q_C)\;\! e }\\
  & \quad\quad\mathsf{\times \, LType^{\wedge}\, B\, Q_B\, T \times \,
@@
     Set) \to}\\
-  & \;\;\mathsf{Equal^\wedge\,A\,A\,Q\,Q'\,refl \to}\\
-  &  \;\;\;\;\mathsf{PredMap\,(LTerm\,A)\,
+  & \mathsf{Equal^\wedge\,A\,A\,Q\,Q'\,refl \to}\\
+  &  \;\mathsf{PredMap\,(LTerm\,A)\,
     (LTerm^{\wedge}\,A\,Q)\,(LTerm^{\wedge}\,A\,Q')}
@@
     LTerm^{\wedge}\, (B \to A) \, K_\top\, t_1}$ is the proof
-  $\mathsf{L_{K_\top} = LTerm^\wedge KT\, (B \to A)\, t_1}$ and
-  $\mathsf{ LTerm^\wedge Arr}$ $\mathsf{: LTerm^{\wedge}\, (B \to A)\,
+  $\mathsf{L_{K_\top}\!\! =\! LTerm^\wedge KT\, (B\! \to\! A)\, t_1}$ and
+  $\mathsf{LTerm^\wedge Arr}$ $\mathsf{: LTerm^{\wedge}\, (B \to A)\,
     (Arr^{\wedge} \, B\, A\, K_\top \, K_\top)\, t_1}$ is the proof
@@
 
-\subsection*{Acknowledgments}
+\section*{Acknowledgments}
 

### 2022-02-15 21-lmcs/21-lmcs.tex
small changes to final version of lmcs paper

@@
 below type constructors for other nested types; and truly nested
-types, such as the nested type {\small
+types, such as the nested type 
 \[\begin{array}{l}
@@
 \end{array}\]
-
+\noindent
 Moreover, for all $\rho : \relenv$, 
